### 2014-01-21 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing latex bug

\documentclass[11pt, oneside]{amsart}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage[latin9]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure} 
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}

%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}

\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
\usepackage[all]{hypcap}
\urlstyle{rm}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}

\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
\newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
\newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
\newcommand{\manishnote}[1]{  {\textcolor{violet}     { ***Manish: #1 }}}
\newcommand{\davidnote}[1]{  {\textcolor{darkgreen}      { ***David: #1 }}}
\newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}

% Don't use tt font for urls
\urlstyle{rm}

% 15 characters / 2.5 cm => 100 characters / line
% Using 11 pt => 94 characters / line
\setlength{\paperwidth}{216 mm}
% 6 lines / 2.5 cm => 55 lines / page
% Using 11pt => 48 lines / pages
\setlength{\paperheight}{279 mm}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
% You can use a baselinestretch of down to 0.9
\renewcommand{\baselinestretch}{0.96}

\sloppypar

\begin{document}

\title[]{First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE): Overview and Summary}

\author{TBD}

%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
%Manish Parashar$^{(4)}$, and David Proctor$^{(1)}$ }
%
%
%\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA}
%
%\thanks{{}$^{(2)}$ Skolkovo Institute of Science and Technology, Moscow, Russian Federation}
%
%\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
%  
%\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}

\begin{abstract}
This technical report discusses the First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE). ...
\end{abstract}

\maketitle

\section{Introduction}

%\katznote{example comment by Dan}
%
%\gabnote{example comment by Gabrielle}
%
%\nchnote{example comment by Neil}
%
%\manishnote{example comment by Manish}
%
%\davidnote{example comment by David}

The First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.

Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is now critical to address many challenges related to the development, deployment, and maintenance of reusable software. In addition, it is essential that scientists, researchers, and students are able to learn and adopt software-related skills and methodologies. Established researchers are already acquiring some of these skills, and in particular a specialized class of software developers is emerging in academic environments who are an integral and embedded part of successful research teams. The WSSSPE workshop is intended provide a forum for discussion of the challenges, including both positions and experiences. The short papers and discussion will be archived to provide a basis for continued discussion, and the workshop is intended to feed into the collaborative writing of one or more journal publications.

\section{Submissions}

The workshop call for paper included:

\begin{quote}
In practice, scientific software activities are part of an ecosystem where key roles are held by developers, users, and funders.  All three groups supply resources to the ecosystem, as well as requirements that bound it.  Roughly following the example of NSF's Vision and Strategy for Software (\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, the ecosystem may be viewed as having challenges related to:

\begin{itemize}
\item the development process that leads to new software
\begin{itemize}
\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
\item software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}
\item the support and maintenance of existing software, including software engineering
\begin{itemize}
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\end{itemize}
\item the role of open source communities or industry
\item use of the software
\begin{itemize}
\item growing communities
\item reproducibility, transparency needs that may be unique to science
\end{itemize}
\item policy issues, such as
\begin{itemize}
\item measuring usage and impact
\item software credit, attribution, incentive, and reward
\item career paths for developers and institutional roles
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}
\item education and training
\end{itemize}

\end{quote}

Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, the organizers wanted to make submission as easy as possible.  The call for papers stated:

\begin{quote}
We invite short (4-page) position/experience reports that will be used to organize panel and discussion sessions.  These papers will be archived by a third-party service, and provided DOIs. We encourage submitters to license their papers under a Creative Commons license that encourages sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop.  An interactive site will be created to link these papers and the workshop discussion, with options for later comments and contributions.  Contributions will be peer-reviewed for relevance and originality before the links are added to the workshop site; contributions will also be used to determine discussion topics and panelists.  We will also plan one or more papers to be collaboratively developed by the contributors, based on the panels and discussions.
\end{quote}

58 submissions were received, and almost all submitters used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.

\section{Peer-Review}

A peer review process followed the submissions, where the 58 papers received 181 reviews, an average of 3.12 reviews per paper.  Reviews were completed using a Google form, which allowed reviewers to provide scores on relevance and comments to the organizers, which were used to decide which papers to associate with the workshop, and comments to the authors, which were provided back to the authors to allow them to improve their papers.

The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact that the reports were already self-published.  The papers were also grouped into 3 areas, each of which was associated with a panel and discussion at the workshop.

\section{Results}

The contributed papers that were discussed at the workshop follow, listed by area.

\subsection{Developing, Deploying and Supporting Software}

\subsubsection{Development Experiences}

\begin{itemize}

\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith. Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}

\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}

\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}

\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}

\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs, Andrew E. Slaughter, Jason M. Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}

\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}

\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}

\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}

\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}

\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}

\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}

\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}

\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}

\end{itemize}

\subsubsection{Deployment, Support, and Maintenance of Existing Software}

\begin{itemize}

\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}

\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}

%\end{itemize}
%
%\subsubsection{Tools}

\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin. Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}

\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}

\end{itemize}

\subsubsection{Best Practices, Challenges, and Recommendations}

\begin{itemize}

\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}

\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland, D. Q. Lamb, F. L\:{o}ffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir, R. Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}

\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}

\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William K. Barnett, Von Welch. Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}

\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}

\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc. Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}

\end{itemize}

\subsection{Policy}

\subsubsection{Modeling Sustainability}

\begin{itemize}

\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}

\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}

\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}

\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}

\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak. Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}

\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}

\end{itemize}

\subsubsection{Credit, Citation, Impact}

\begin{itemize}

\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}

\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}

\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}

\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}

\end{itemize}

In addition, the following paper from another area will also be discussed in this area.

\begin{itemize}

\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}

\end{itemize}

\subsubsection{Reproducibility}

\begin{itemize}

\item Victoria Stodden, Sheila Miguez. Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research \cite{Stodden_WSSSPE}

\end{itemize}

\subsubsection{Implementing Policy}

\begin{itemize}

\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}

\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}

\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}

\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}

\end{itemize}

\subsection{Communities}

\subsubsection{Communities}

\begin{itemize}

\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}

\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}

\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}

\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}

\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}

\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}

\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}

\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}

\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak Mookherjee. Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}

\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione. Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}

\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}

\end{itemize}

In addition, the following paper from another area will also be discussed in this area.

\begin{itemize}

\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}

\end{itemize}

\subsubsection{Industry \& Economic Models}

\begin{itemize}

\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}

\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}

\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}

\end{itemize}

In addition, the following papers from other areas will also be discussed in this area.

\begin{itemize}

\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}

\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}

\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}

\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}

\end{itemize}

\subsubsection{Education \& Training}

\begin{itemize}

\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T. Brown. Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}

\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}

\end{itemize}

In addition, the following papers from other areas will also be discussed in this area.

\begin{itemize}

\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}

\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}

\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}

\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}

\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}

\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}

\end{itemize}

\section{Conclusions}

The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop agenda.  However, contributors also want to get credit for their participation in the process.  And the workshop organizers want to make sure that the workshop content and their efforts are recorded.  Ideally, there would be a service that would be able to index the contributions to the workshop, serving the authors, the organizers, and the larger community.  But since there isn't such a service today, the workshop organizers are writing this initial report and making use of arXiv as a partial solution to provide a record of the workshop.

After the workshop, one or more additional papers will be created that will include the discussions at the workshop.  These papers will likely have many authors, and may be submitted to peer-reviewed journals.

\section*{Acknowledgments}

\section*{Appendix - List of attendees}

The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:

Daniel S. Katz (@danielskatz)

Neil Chue Hong (@npch)

Gabrielle Allen

David Proctor (@djproctor)

Manish Parashar

Jay Alameda (alameda@illinois.edu)

David Andrs

Lorena A Barba (@LorenaABarba)

David Bernholdt (bernholdtde@ornl.gov)

Phil Bourne

Karl Broman (@kwbroman)

Maxine Brown, UIC (maxine@uic.edu)

David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)

Bruno Bzeznik

Jeffrey Carver, University of Alabama (carver@cs.ua.edu)

John W. Cobb (cobbjw@ornl.gov, @johnwcobb)

Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)

Rion Dooley

Anshu Dubey (adubey@lbl.gov)

Ian Foster, @ianfoster

Juliana Freire

Derek Gaston

Derek Groen (@whydoitweet on Twitter)

Edmund Hart (@distribecology)

James Hetherington @jamespjh

Simon Hettrick (s.hettrick@software.ac.uk)

Kenneth Hoste (@kehoste, Ghent University, Belgium)

James Howison, University of Texas at Austin (@jameshowison)

Daisie Huang

Kerk Kee (kerk.kee@gmail.com)

Arnaud Legrand

Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu

Louis Maddox (@lmmddx)

Ketan Maheshwari (@km\_9 on Twitter)

Suresh Marru (smarru@iu.edu, @sureshmarru)

Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek

James McClure

Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org

J. Yates Monteith, Clemson University

Pierre Neyron

Lucas Nussbaum (lucas.nussbaum@loria.fr)

Cody Permann

John Peterson

Quan Pham, University of Chicago

Marlon Pierce, marpierc@iu.edu, @marpierc

Heather Piwowar

Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places

Andreas Schreiber (@onyame)

William Scullin

Andrew Slaughter

Arfon Smith

James Spencer (@exp\_n)

Shel Swenson, University of Southern California (shel.swenson@usc.edu)

Fr\'{e}d\'{e}ric Suter

Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)

George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)

John Towns

Erik Trainer

Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)

Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk

Colin C. Venters, c.venters@hud.ac.uk

Von Welch, @VonWelch

Nancy Wilkins-Diehr, wilkinsn@sdsc.edu

Theresa Windus (twindus@iastate.edu)

Rich Wolski

Lynn Zentner

\bibliographystyle{plain}

\bibliography{wssspe}
\end{document}

### 2014-01-21 WSSSPE1_report/WSSSPE_pre-meeting_report.tex
fixing latex bug

--- 2014-01-21 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\title[]{First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE): Overview and Summary}
+\title[]{First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE): Submission and Peer-Review Process, and Results}
 
-\author{TBD}
+\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
+Manish Parashar$^{(4)}$, and David Proctor$^{(1)}$ }
 
-%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
-%Manish Parashar$^{(4)}$, and David Proctor$^{(1)}$ }
-%
-%
-%\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA}
-%
-%\thanks{{}$^{(2)}$ Skolkovo Institute of Science and Technology, Moscow, Russian Federation}
-%
-%\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
-%  
-%\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
+\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA}
+
+\thanks{{}$^{(2)}$ Skolkovo Institute of Science and Technology, Moscow, Russian Federation}
+
+\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+
+\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
 
 \begin{abstract}
-This technical report discusses the First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE). ...
+This technical report discusses the submission and peer-review process used by the First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE) and the results of that process.
+It is intended to record both this alternative model as well as the papers associated with the workshop that resulted from that process.
 \end{abstract}
@@
 
-The First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
+The First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE)\footnote{\url{http://wssspe.researchcomputing.org.uk}} will be held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
 
@@
 The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact that the reports were already self-published.
-The papers were also grouped into 3 areas, each of which was associated with a panel and discussion at the workshop.
+The papers were also grouped into 3 areas, each of which will be associated with a panel and discussion at the workshop.
 
@@
 
-The contributed papers that were discussed at the workshop follow, listed by area.
+The contributed papers that will be discussed at the workshop follow, listed by area.
 
@@
 
-\section*{Acknowledgments}
-
-\section*{Appendix - List of attendees}
-
-The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:
-
-Daniel S. Katz (@danielskatz)
-
-Neil Chue Hong (@npch)
-
-Gabrielle Allen
-
-David Proctor (@djproctor)
-
-Manish Parashar
-
-Jay Alameda (alameda@illinois.edu)
-
-David Andrs
-
-Lorena A Barba (@LorenaABarba)
-
-David Bernholdt (bernholdtde@ornl.gov)
-
-Phil Bourne
-
-Karl Broman (@kwbroman)
-
-Maxine Brown, UIC (maxine@uic.edu)
-
-David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)
-
-Bruno Bzeznik
-
-Jeffrey Carver, University of Alabama (carver@cs.ua.edu)
-
-John W. Cobb (cobbjw@ornl.gov, @johnwcobb)
-
-Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)
-
-Rion Dooley
-
-Anshu Dubey (adubey@lbl.gov)
-
-Ian Foster, @ianfoster
-
-Juliana Freire
-
-Derek Gaston
-
-Derek Groen (@whydoitweet on Twitter)
-
-Edmund Hart (@distribecology)
-
-James Hetherington @jamespjh
-
-Simon Hettrick (s.hettrick@software.ac.uk)
-
-Kenneth Hoste (@kehoste, Ghent University, Belgium)
-
-James Howison, University of Texas at Austin (@jameshowison)
-
-Daisie Huang
-
-Kerk Kee (kerk.kee@gmail.com)
-
-Arnaud Legrand
-
-Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu
-
-Louis Maddox (@lmmddx)
-
-Ketan Maheshwari (@km\_9 on Twitter)
-
-Suresh Marru (smarru@iu.edu, @sureshmarru)
-
-Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek
-
-James McClure
-
-Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org
-
-J. Yates Monteith, Clemson University
-
-Pierre Neyron
-
-Lucas Nussbaum (lucas.nussbaum@loria.fr)
-
-Cody Permann
-
-John Peterson
-
-Quan Pham, University of Chicago
-
-Marlon Pierce, marpierc@iu.edu, @marpierc
-
-Heather Piwowar
-
-Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places
-
-Andreas Schreiber (@onyame)
-
-William Scullin
-
-Andrew Slaughter
-
-Arfon Smith
-
-James Spencer (@exp\_n)
-
-Shel Swenson, University of Southern California (shel.swenson@usc.edu)
-
-Fr\'{e}d\'{e}ric Suter
-
-Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)
-
-George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)
-
-John Towns
-
-Erik Trainer
-
-Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)
-
-Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk
-
-Colin C. Venters, c.venters@hud.ac.uk
-
-Von Welch, @VonWelch
-
-Nancy Wilkins-Diehr, wilkinsn@sdsc.edu
-
-Theresa Windus (twindus@iastate.edu)
-
-Rich Wolski
-
-Lynn Zentner
-
 \bibliographystyle{plain}

### 2014-01-21 WSSSPE1_report/WSSSPE_pre-meeting_report.tex
small fixed - making more past tense

@@
 Established researchers are already acquiring some of these skills, and in particular a specialized class of software developers is emerging in academic environments who are an integral and embedded part of successful research teams.
-The WSSSPE workshop is intended provide a forum for discussion of the challenges, including both positions and experiences.
-The short papers and discussion will be archived to provide a basis for continued discussion, and the workshop is intended to feed into the collaborative writing of one or more journal publications.
+The WSSSPE workshop was intended to provide a forum for discussion of the challenges, including both positions and experiences.
+The short papers and discussion were archived to provide a basis for continued discussion, and the workshop was intended to feed into the collaborative writing of one or more journal publications.
 

### 2014-01-21 WSSSPE1_report/WSSSPE1_summary_report.tex
initial organization

@@
 
-\author{TBD}
+\author{Authors TBD}
 
@@
 
+People who have volunteered to work on this report:
+
+the workshop organizers:
+\begin{itemize}
+\item "Daniel S. Katz" <dkatz@nsf.gov>
+\item Gabrielle Allen <gdallen@illinois.edu>
+\item "Neil Chue Hong" <N.ChueHong@software.ac.uk>
+\item Manish Parashar <parashar@rutgers.edu>
+\item David Proctor <dproctor@nsf.gov>
+\end{itemize}
+
+and additional volunteers:
+\begin{itemize}
+\item Chris Mattmann <chris.a.mattmann@jpl.nasa.gov>
+\item Ketan Maheshwari <ketancmaheshwari@gmail.com>
+\item Marlon Pierce <marpierc@iu.edu>
+\item Colin Venters <C.Venters@hud.ac.uk>
+\item Suresh Marru <smarru@iu.edu>
+\item Lynn Zentner <lzentner@purdue.edu>
+\item Anne Elster <anne.elster@gmail.com>
+\item Shel Swenson <mdswenso@usc.edu>
+\item Andy Ray Terrel <andy.terrel@gmail.com>
+\item Abani Patra <abani.patra@gmail.com>
+\item Nancy Wilkins-Diehr <wilkinsn@sdsc.edu>
+\item James Spencer <j.spencer@imperial.ac.uk>
+\item Frank Loeffler <knarf@cct.lsu.edu>
+\item James Hetherington <j.hetherington@ucl.ac.uk>
+\item "Sou-Cheng (Terrya) Choi" <sou.cheng.terrya.choi@gmail.com>
+\item James Howison <james@howison.name>
+\item Bruce Berriman <gbb@ipac.caltech.edu>
+\item Hilmar Lapp <hlapp@nescent.org>
+\item "Marcus D. Hanwell" <marcus.hanwell@kitware.com>
+\item Lucas Nussbaum <lucas.nussbaum@loria.fr>
+\item Matthew Turk <matthewturk@gmail.com>
+\end{itemize}
+
+The original document is \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#} (but can no longer be edited).
+Note that the original document has comments in addition to text.
+
+\pagebreak
+
+\section*{Executive Summary}
+
+\pagebreak
+
 \section{Introduction}
 
-%\katznote{example comment by Dan}
-%
-%\gabnote{example comment by Gabrielle}
-%
-%\nchnote{example comment by Neil}
-%
-%\manishnote{example comment by Manish}
-%
-%\davidnote{example comment by David}
+The First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
+
+...
+
+\section{Agenda}
+
+\section{Keynotes}
+
+what was presented and what was discussed
+
+\section{Developing and Supporting Software}
+
+what was presented and what was discussed
+
+\section{Policy}
+
+what was presented and what was discussed
+
+\section{Communities}
+
+what was presented and what was discussed
+
+\section{Other Discussion}
+
+from the misc questions/answers/observations part of the google doc?
+
+\section{Cross-cutting Issues}
+
+\section{Conclusions}
+
+pull the discussion together
+
+\subsection{Recommendations or Lessons}
+
+if needed.
+
+\subsection{Follow up actions}
+
+or at least the discussion about them, and the current plans for future events.
+
+\section*{Acknowledgments}
+
+\section*{Appendix - List of attendees}
+
+The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:
+
+Daniel S. Katz (@danielskatz)
+
+Neil Chue Hong (@npch)
+
+Gabrielle Allen
+
+David Proctor (@djproctor)
+
+Manish Parashar
+
+Jay Alameda (alameda@illinois.edu)
+
+David Andrs
+
+Lorena A Barba (@LorenaABarba)
+
+David Bernholdt (bernholdtde@ornl.gov)
+
+Phil Bourne
+
+Karl Broman (@kwbroman)
+
+Maxine Brown, UIC (maxine@uic.edu)
+
+David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)
+
+Bruno Bzeznik
+
+Jeffrey Carver, University of Alabama (carver@cs.ua.edu)
+
+John W. Cobb (cobbjw@ornl.gov, @johnwcobb)
+
+Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)
+
+Rion Dooley
+
+Anshu Dubey (adubey@lbl.gov)
+
+Ian Foster, @ianfoster
+
+Juliana Freire
+
+Derek Gaston
+
+Derek Groen (@whydoitweet on Twitter)
+
+Edmund Hart (@distribecology)
+
+James Hetherington @jamespjh
+
+Simon Hettrick (s.hettrick@software.ac.uk)
+
+Kenneth Hoste (@kehoste, Ghent University, Belgium)
+
+James Howison, University of Texas at Austin (@jameshowison)
+
+Daisie Huang
+
+Kerk Kee (kerk.kee@gmail.com)
+
+Arnaud Legrand
+
+Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu
+
+Louis Maddox (@lmmddx)
+
+Ketan Maheshwari (@km\_9 on Twitter)
+
+Suresh Marru (smarru@iu.edu, @sureshmarru)
+
+Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek
+
+James McClure
+
+Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org
+
+J. Yates Monteith, Clemson University
+
+Pierre Neyron
+
+Lucas Nussbaum (lucas.nussbaum@loria.fr)
+
+Cody Permann
+
+John Peterson
+
+Quan Pham, University of Chicago
+
+Marlon Pierce, marpierc@iu.edu, @marpierc
+
+Heather Piwowar
+
+Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places
+
+Andreas Schreiber (@onyame)
+
+William Scullin
+
+Andrew Slaughter
+
+Arfon Smith
+
+James Spencer (@exp\_n)
+
+Shel Swenson, University of Southern California (shel.swenson@usc.edu)
+
+Fr\'{e}d\'{e}ric Suter
+
+Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)
+
+George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)
+
+John Towns
+
+Erik Trainer
+
+Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)
+
+Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk
+
+Colin C. Venters, c.venters@hud.ac.uk
+
+Von Welch, @VonWelch
+
+Nancy Wilkins-Diehr, wilkinsn@sdsc.edu
+
+Theresa Windus (twindus@iastate.edu)
+
+Rich Wolski
+
+Lynn Zentner
+
+\section{Pre-workshop section: Introduction}
 
@@
 
-\section{Submissions}
+\section{Pre-workshop section: Submissions}
 
@@
 
-\section{Peer-Review}
+\section{Pre-workshop section: Peer-Review}
 
@@
 
-\section{Results}
+\section{Pre-workshop section: Results}
 
@@
 
-\section{Conclusions}
+\section{Pre-workshop section: Conclusions}
 

### 2014-01-23 WSSSPE1_report/WSSSPE1_summary_report.tex
adding a Makefile, make now should generate WSSPE1_summary_report.pdf make clean should clean up after latex
Minor edits.

@@
 
-\title[]{First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE): Overview and Summary}
+\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)}
 
@@
 \begin{abstract}
-This technical report discusses the First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE). ...
+This technical report discusses the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE). ...
 \end{abstract}
@@
 
-the workshop organizers:
+The workshop organizers:
 \begin{itemize}
-\item "Daniel S. Katz" <dkatz@nsf.gov>
+\item Daniel S. Katz <dkatz@nsf.gov>
 \item Gabrielle Allen <gdallen@illinois.edu>
-\item "Neil Chue Hong" <N.ChueHong@software.ac.uk>
+\item Neil Chue Hong <N.ChueHong@software.ac.uk>
 \item Manish Parashar <parashar@rutgers.edu>
@@
 
-and additional volunteers:
+Additional volunteers:
 \begin{itemize}
@@
 \item James Hetherington <j.hetherington@ucl.ac.uk>
-\item "Sou-Cheng (Terrya) Choi" <sou.cheng.terrya.choi@gmail.com>
+\item Sou-Cheng (Terrya) Choi <sou.cheng.terrya.choi@gmail.com>
 \item James Howison <james@howison.name>
@@
 \item Hilmar Lapp <hlapp@nescent.org>
-\item "Marcus D. Hanwell" <marcus.hanwell@kitware.com>
+\item Marcus D. Hanwell <marcus.hanwell@kitware.com>
 \item Lucas Nussbaum <lucas.nussbaum@loria.fr>
@@
 
-The First Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
+The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
 
@@
 
-\section*{Appendix - List of attendees}
+\appendix
+\section{List of attendees}
 

### 2014-01-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Fix unreadable characters around email addresses.

@@
 \begin{itemize}
-\item Daniel S. Katz <dkatz@nsf.gov>
-\item Gabrielle Allen <gdallen@illinois.edu>
-\item Neil Chue Hong <N.ChueHong@software.ac.uk>
-\item Manish Parashar <parashar@rutgers.edu>
-\item David Proctor <dproctor@nsf.gov>
+\item Daniel S. Katz $\langle$\url{dkatz@nsf.gov}$\rangle$
+\item Gabrielle Allen $\langle$\url{gdallen@illinois.edu}$\rangle$
+\item Neil Chue Hong $\langle$\url{N.ChueHong@software.ac.uk}$\rangle$
+\item Manish Parashar $\langle$\url{parashar@rutgers.edu}$\rangle$
+\item David Proctor $\langle$\url{dproctor@nsf.gov}$\rangle$
 \end{itemize}
@@
 \begin{itemize}
-\item Chris Mattmann <chris.a.mattmann@jpl.nasa.gov>
-\item Ketan Maheshwari <ketancmaheshwari@gmail.com>
-\item Marlon Pierce <marpierc@iu.edu>
-\item Colin Venters <C.Venters@hud.ac.uk>
-\item Suresh Marru <smarru@iu.edu>
-\item Lynn Zentner <lzentner@purdue.edu>
-\item Anne Elster <anne.elster@gmail.com>
-\item Shel Swenson <mdswenso@usc.edu>
-\item Andy Ray Terrel <andy.terrel@gmail.com>
-\item Abani Patra <abani.patra@gmail.com>
-\item Nancy Wilkins-Diehr <wilkinsn@sdsc.edu>
-\item James Spencer <j.spencer@imperial.ac.uk>
-\item Frank Loeffler <knarf@cct.lsu.edu>
-\item James Hetherington <j.hetherington@ucl.ac.uk>
-\item Sou-Cheng (Terrya) Choi <sou.cheng.terrya.choi@gmail.com>
-\item James Howison <james@howison.name>
-\item Bruce Berriman <gbb@ipac.caltech.edu>
-\item Hilmar Lapp <hlapp@nescent.org>
-\item Marcus D. Hanwell <marcus.hanwell@kitware.com>
-\item Lucas Nussbaum <lucas.nussbaum@loria.fr>
-\item Matthew Turk <matthewturk@gmail.com>
+\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
+\item Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
+\item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
+\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
+\item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
+\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
+\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
+\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
+\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
+\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
+\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
+\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
+\item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
+\item Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
+\item James Howison $\langle$\url{james@howison.name}$\rangle$
+\item Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
+\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
+\item Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
+\item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
+\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
 \end{itemize}

### 2014-01-28 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 \section{Agenda}
+The workshop submissions were classified into various related categories.
+Representatives from each submission categories were appointed as penalists. The sessions were kickstarted with keynote talks followed by panel discussions.
 
@@
 what was presented and what was discussed
+
+Arfon Smith of github presented the first keynote. Following points summarize the keynote and the associated discussion:
+\begin{enumerate}
+    \item Socially driven software development.
+    \item Role of software in research.
+\end{enumerate}
 

### 2014-01-30 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 what was presented and what was discussed
+
+\subsection{What are communities?}
+\subsection{Challenges of community}
+include metrics for community success (e.g., more external contributors than internal).
+\subsection{Admirable scientific software communities}
+\begin{itemize}
+\item example communities
+\item their (quick) origin stories.
+\subsection{Resources for learning about software communities}
+\begin{itemize}
+\item academic fields studying communities (and software communities)
+\item courses on online communities
+\item books
+\item need for software carpentry module on organizing communities?
+\end{itemize}
 

### 2014-01-30 WSSSPE1_report/WSSSPE1_summary_report.tex
fix latex error
Updates to the development/support section
Began fleshing out what subsections/points might be made, with help
from the papers and shared Google document.

@@
 
+\subsection{Research or Reuse?}
+
+Discussion of software produced as a by-product versus software developed for reuse.
+How does this change the project, who develops the code, growth beyond 1--2 person projects to larger projects with diverse set of consumers and contributors.
+Modular and extensible code versus working for current research problem---tackled by same people, or a collaborative team?
+
+\subsection{Defining Sustainability for Scientific Software}
+
+What is sustainable? What are the best practices, can workshops fund meetings to help define and improve best practices in these areas? Software plans from the funding agencies?
+
+\subsection{Training Scientists to Develop and Support Software}
+
+Discussion of the evolving role of scientists, and/or others that fill these roles.
+
+\subsection{Software Process, Code Review, Automation, Reproducibility}
+
+There are a lot of tools out there, but few are currently used. Look at what some projects have found successful, and how to automate as much as possible to reduce additional overhead.
+
+\subsection{Software and Data Licensing}
+
+Its impact on how and where research products are used, who can collaborate and what patterns have worked/not worked in existing communities.
+
+\subsection{Funding, Sustainability Beyond the First Grant/Institution}
+
+How initial software development is funded, moving to follow up projects, maintenance, community growth, collaborating with other institutions, labs, industry, internationally.
+Contract relationship versus collaborative development between scientists and software developers.
+
+\subsection{Training Others to use Software}
+
+Who creates training materials, how are they distributed, integration into courses when projects see very wide application in research.
+
 \section{Policy}
@@
 what was presented and what was discussed
+
+\subsection{Career Tracks for Scientific Software Developers}
+
+How to ensure software is sustainable by ensuring there are career paths for developers.
 
@@
 \item their (quick) origin stories.
+\end{itemize}
 \subsection{Resources for learning about software communities}

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
adding use-case section
reorganizing

@@
 
-People who have volunteered to work on this report:
+People who have volunteered to work on this report: (with * for actual contributors - \note{if you contribute, add a * by your name})
 
@@
 \begin{itemize}
-\item Daniel S. Katz $\langle$\url{dkatz@nsf.gov}$\rangle$
+\item * Daniel S. Katz $\langle$\url{dkatz@nsf.gov}$\rangle$
 \item Gabrielle Allen $\langle$\url{gdallen@illinois.edu}$\rangle$
@@
 \item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
-\item Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
+\item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
 \item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
@@
 \item James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
-\item Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
+\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
 \item James Howison $\langle$\url{james@howison.name}$\rangle$
-\item Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
+\item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
 \item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
-\item Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
+\item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
 \item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
@@
 
+\todo{1 page summary goes here}
+
 \pagebreak
@@
 
-...
+Because progress in scientific research is dependent on the quality and accessibility
+of software at all levels, it is now critical to address many challenges related to the
+development, deployment, and maintenance of reusable software. In addition, it is
+essential that scientists, researchers, and students are able to learn and adopt
+software-related skills and methodologies. Established researchers are already
+acquiring some of these skills, and in particular a specialized class of software
+developers is emerging in academic environments who are an integral and embedded
+part of successful research teams. The WSSSPE workshop was intended provide a
+forum for discussion of the challenges, including both positions and experiences.
+The short papers and discussion have been archived to provide a basis for continued
+discussion, and the workshop has fed into the collaborative writing of this journal publication.
+Additional papers based on extended versions of workshop submissions are expected.
+The level of interest in the workshop has led the organizers, working with some of the submitters
+and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
+the workshop website into a community website that can be used as a focus for further
+discussion and progress.
+
+\section{Workshop Process}
+
+WSSSPE1 was organized via peer-review and organization of papers submitted in response to a call for papers.  The call for paper was created by the organizers and peer-review committee.
+
+The workshop call for paper included:
+
+\begin{quote}
+In practice, scientific software activities are part of an ecosystem where key roles are held by 
+developers, users, and funders.  All three groups supply resources to the ecosystem, as well as 
+requirements that bound it.  Roughly following the example of NSF's Vision and Strategy for Software 
+(\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, 
+the ecosystem may be viewed as having challenges related to:
+
+\begin{itemize}
+\item the development process that leads to new software
+\begin{itemize}
+\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
+\item software created as a by-product of research
+\item impact of computer science research on the development of scientific software
+\end{itemize}
+\item the support and maintenance of existing software, including software engineering
+\begin{itemize}
+\item governance, business, and sustainability models
+\item the role of community software repositories, their operation and sustainability
+\end{itemize}
+\item the role of open source communities or industry
+\item use of the software
+\begin{itemize}
+\item growing communities
+\item reproducibility, transparency needs that may be unique to science
+\end{itemize}
+\item policy issues, such as
+\begin{itemize}
+\item measuring usage and impact
+\item software credit, attribution, incentive, and reward
+\item career paths for developers and institutional roles
+\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
+\item mechanisms and venues for publishing software, and the role of publishers
+\end{itemize}
+\item education and training
+\end{itemize}
+
+\end{quote}
+
+Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, the organizers wanted to make submission as easy as possible.
+The call for papers stated:
+
+\begin{quote}
+We invite short (4-page) position/experience reports that will be used to organize panel and 
+discussion sessions.  These papers will be archived by a third-party service, and provided DOIs. We 
+encourage submitters to license their papers under a Creative Commons license that encourages 
+sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop.  
+An interactive site will be created to link these papers and the workshop discussion, with options for 
+later comments and contributions.  Contributions will be peer-reviewed for relevance and originality 
+before the links are added to the workshop site; contributions will also be used to determine 
+discussion topics and panelists.  We will also plan one or more papers to be collaboratively 
+developed by the contributors, based on the panels and discussions.
+\end{quote}
+
+58 submissions were received, and almost all submitters used either 
+arXiv\footnote{\url{http://arxiv.org}} or 
+figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+
+A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
+average of 3.12 reviews per paper.  Reviews were completed using a Google form, which allowed 
+reviewers to provide scores on relevance and comments to the organizers, which were used to 
+decide which papers to associate with the workshop, and comments to the authors, which were 
+provided back to the authors to allow them to improve their papers.
+
+The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
+high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
+that the reports were already self-published.  The papers were also grouped into 3 areas, each of 
+which was associated with a panel and discussion at the workshop.
 
 \section{Agenda}
-The workshop submissions were classified into various related categories.
-Representatives from each submission categories were appointed as penalists. The sessions were kickstarted with keynote talks followed by panel discussions.
+
+The workshop consisted of 2 keynote presentations and three panel/discussion sessions.  The panels 
+were organized based on a classification of the workshop submissions into three categories, following 
+the themes of the call for papers as modified by the areas of the submissions.  Representatives from 
+each submission categories were appointed as panelists, to discuss the papers in that category.
 
@@
 
-what was presented and what was discussed
+\todo{section should include what was presented and what was discussed}
 
@@
 
-what was presented and what was discussed
+\todo{section should include what was presented and what was discussed}
+
+\subsection{Papers}
+
+\subsubsection*{Development Experiences}
+
+\begin{itemize}
+
+\item Mark C.
+Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith.
+Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}
+
+\item Derek R.
+Gaston, John Peterson, Cody J.
+Permann, David Andrs, Andrew E.
+Slaughter, Jason M.
+Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}
+
+\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
+
+\begin{itemize}
+
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter.
+SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
+
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin.
+Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}
+
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Best Practices, Challenges, and Recommendations}
+
+\begin{itemize}
+
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}
+
+\item Anshu Dubey, S.
+Brandt, R.
+Brower, M.
+Giles, P.
+Hovland, D.
+Q.
+Lamb, F.
+L\:{o}ffler, B.
+Norris, B.
+O'Shea, C.
+Rebbi, M.
+Snir, R.
+Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
+
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}
+
+\item Craig A.
+Stewart, Julie Wernert, Eric A.
+Wernert, William K.
+Barnett, Von Welch.
+Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}
+
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}
+
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc.
+Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}
+
+\end{itemize}
 
@@
 
-What is sustainable? What are the best practices, can workshops fund meetings to help define and improve best practices in these areas? Software plans from the funding agencies?
+What is sustainable?
+What are the best practices, can workshops fund meetings to help define and improve best practices in these areas?
+Software plans from the funding agencies?
+\todo{mention as cross-cutting, since this came up in other parts of the workshop as well?}
 
@@
 
-what was presented and what was discussed
+\todo{section should include what was presented and what was discussed}
+
+\subsection{Papers}
+
+\subsubsection*{Modeling Sustainability}
+
+\begin{itemize}
+
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}
+
+\item Colin C.
+Venters, Lydia Lau, Michael K.
+Griffiths, Violeta Holmes, Rupert R.
+Ward, Jie Xu.
+The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak.
+Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Credit, Citation, Impact}
+
+\begin{itemize}
+
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}
+
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}
+
+\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}
+
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area were also discussed in this area.
+
+\begin{itemize}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Reproducibility}
+
+\begin{itemize}
+
+\item Victoria Stodden, Sheila Miguez. Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research \cite{Stodden_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Implementing Policy}
+
+\begin{itemize}
+
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+
+\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}
+
+\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}
+
+\end{itemize}
 
@@
 
-what was presented and what was discussed
+\todo{section should include what was presented and what was discussed}
+
+\subsection{Papers}
+
+\subsubsection*{Communities}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+
+\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Industry \& Economic Models}
+
+\begin{itemize}
+
+\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}
+
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed in this area.
+
+\begin{itemize}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Education \& Training}
+
+\begin{itemize}
+
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T.
+Brown.
+Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}
+
+\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed in this area.
+
+\begin{itemize}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
+
+\end{itemize}
 
@@
 
-from the misc questions/answers/observations part of the google doc?
+\note{from the misc questions/answers/observations part of the google doc?}
 
@@
 
+\subsection{Defining sustainability}
+
+\subsection{Licensing}
+
+\note{not sure if this was really cross-cutting...}
+
 \section{Conclusions}
 
-pull the discussion together
+\todo{pull the discussion together}
+
+\todo{add some analysis of the deficiencies and difficulties that are present in different fields, and those that are common?}
 
@@
 
-if needed.
+\note{if needed.}
 
@@
 
-or at least the discussion about them, and the current plans for future events.
+\note{or at least the discussion about them, and the current plans for future events.}
+
+\note{conclusions from pre-workshop paper follows}
+
+The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop 
+agenda.  However, contributors also want to get credit for their participation in the process.  And the 
+workshop organizers want to make sure that the workshop content and their efforts are recorded.  
+Ideally, there would be a service that would be able to index the contributions to the workshop, 
+serving the authors, the organizers, and the larger community.  But since there isn't such a service 
+today, the workshop organizers are writing this initial report and making use of arXiv as a partial 
+solution to provide a record of the workshop.
+
+After the workshop, one or more additional papers will be created that will include the discussions at 
+the workshop.  These papers will likely have many authors, and may be submitted to peer-reviewed 
+journals.
 
@@
 
-\section{Pre-workshop section: Introduction}
-
-The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
-
-Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is now critical to address many challenges related to the development, deployment, and maintenance of reusable software.
-In addition, it is essential that scientists, researchers, and students are able to learn and adopt software-related skills and methodologies.
-Established researchers are already acquiring some of these skills, and in particular a specialized class of software developers is emerging in academic environments who are an integral and embedded part of successful research teams.
-The WSSSPE workshop is intended provide a forum for discussion of the challenges, including both positions and experiences.
-The short papers and discussion will be archived to provide a basis for continued discussion, and the workshop is intended to feed into the collaborative writing of one or more journal publications.
-
-\section{Pre-workshop section: Submissions}
-
-The workshop call for paper included:
-
-\begin{quote}
-In practice, scientific software activities are part of an ecosystem where key roles are held by developers, users, and funders.
-All three groups supply resources to the ecosystem, as well as requirements that bound it.
-Roughly following the example of NSF's Vision and Strategy for Software (\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, the ecosystem may be viewed as having challenges related to:
-
-\begin{itemize}
-\item the development process that leads to new software
-\begin{itemize}
-\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
-\item software created as a by-product of research
-\item impact of computer science research on the development of scientific software
-\end{itemize}
-\item the support and maintenance of existing software, including software engineering
-\begin{itemize}
-\item governance, business, and sustainability models
-\item the role of community software repositories, their operation and sustainability
-\end{itemize}
-\item the role of open source communities or industry
-\item use of the software
-\begin{itemize}
-\item growing communities
-\item reproducibility, transparency needs that may be unique to science
-\end{itemize}
-\item policy issues, such as
-\begin{itemize}
-\item measuring usage and impact
-\item software credit, attribution, incentive, and reward
-\item career paths for developers and institutional roles
-\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of publishers
-\end{itemize}
-\item education and training
-\end{itemize}
-
-\end{quote}
-
-Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, the organizers wanted to make submission as easy as possible.
-The call for papers stated:
-
-\begin{quote}
-We invite short (4-page) position/experience reports that will be used to organize panel and discussion sessions.
-These papers will be archived by a third-party service, and provided DOIs.
-We encourage submitters to license their papers under a Creative Commons license that encourages sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop.
-An interactive site will be created to link these papers and the workshop discussion, with options for later comments and contributions.
-Contributions will be peer-reviewed for relevance and originality before the links are added to the workshop site; contributions will also be used to determine discussion topics and panelists.
-We will also plan one or more papers to be collaboratively developed by the contributors, based on the panels and discussions.
-\end{quote}
-
-58 submissions were received, and almost all submitters used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
-
-\section{Pre-workshop section: Peer-Review}
-
-A peer review process followed the submissions, where the 58 papers received 181 reviews, an average of 3.12 reviews per paper.
-Reviews were completed using a Google form, which allowed reviewers to provide scores on relevance and comments to the organizers, which were used to decide which papers to associate with the workshop, and comments to the authors, which were provided back to the authors to allow them to improve their papers.
-
-The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact that the reports were already self-published.
-The papers were also grouped into 3 areas, each of which was associated with a panel and discussion at the workshop.
-
-\section{Pre-workshop section: Results}
-
-The contributed papers that were discussed at the workshop follow, listed by area.
-
-\subsection{Developing, Deploying and Supporting Software}
-
-\subsubsection{Development Experiences}
-
-\begin{itemize}
-
-\item Mark C.
-Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith.
-Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
-
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}
-
-\item Derek R.
-Gaston, John Peterson, Cody J.
-Permann, David Andrs, Andrew E.
-Slaughter, Jason M.
-Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}
-
-\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
-
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
-
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Deployment, Support, and Maintenance of Existing Software}
-
-\begin{itemize}
-
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter.
-SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
-
-%\end{itemize}
-%
-%\subsubsection{Tools}
-
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin.
-Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}
-
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Best Practices, Challenges, and Recommendations}
-
-\begin{itemize}
-
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}
-
-\item Anshu Dubey, S.
-Brandt, R.
-Brower, M.
-Giles, P.
-Hovland, D.
-Q.
-Lamb, F.
-L\:{o}ffler, B.
-Norris, B.
-O'Shea, C.
-Rebbi, M.
-Snir, R.
-Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
-
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}
-
-\item Craig A.
-Stewart, Julie Wernert, Eric A.
-Wernert, William K.
-Barnett, Von Welch.
-Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}
-
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}
-
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc.
-Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}
-
-\end{itemize}
-
-\subsection{Policy}
-
-\subsubsection{Modeling Sustainability}
-
-\begin{itemize}
-
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}
-
-\item Colin C.
-Venters, Lydia Lau, Michael K.
-Griffiths, Violeta Holmes, Rupert R.
-Ward, Jie Xu.
-The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak.
-Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Credit, Citation, Impact}
-
-\begin{itemize}
-
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}
-
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}
-
-\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}
-
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area will also be discussed in this area.
-
-\begin{itemize}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Reproducibility}
-
-\begin{itemize}
-
-\item Victoria Stodden, Sheila Miguez. Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research \cite{Stodden_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Implementing Policy}
-
-\begin{itemize}
-
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
-
-\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}
-
-\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}
-
-\end{itemize}
-
-\subsection{Communities}
-
-\subsubsection{Communities}
-
-\begin{itemize}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
-
-\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area will also be discussed in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Industry \& Economic Models}
-
-\begin{itemize}
-
-\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}
-
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas will also be discussed in this area.
-
-\begin{itemize}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
-
-\end{itemize}
-
-\subsubsection{Education \& Training}
-
-\begin{itemize}
-
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T.
-Brown.
-Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}
-
-\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas will also be discussed in this area.
-
-\begin{itemize}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
-
-\end{itemize}
-
-\section{Pre-workshop section: Conclusions}
-
-The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop agenda.
-However, contributors also want to get credit for their participation in the process.
-And the workshop organizers want to make sure that the workshop content and their efforts are recorded.
-Ideally, there would be a service that would be able to index the contributions to the workshop, serving the authors, the organizers, and the larger community.
-But since there isn't such a service today, the workshop organizers are writing this initial report and making use of arXiv as a partial solution to provide a record of the workshop.
-
-After the workshop, one or more additional papers will be created that will include the discussions at the workshop.
-These papers will likely have many authors, and may be submitted to peer-reviewed journals.
-
 \bibliographystyle{plain}

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
First draft summary of keynotes

@@
 \item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
-\item James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
+\item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
 \item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
@@
 
-Arfon Smith of github presented the first keynote. Following points summarize the keynote and the associated discussion:
-\begin{enumerate}
-    \item Socially driven software development.
-    \item Role of software in research.
-\end{enumerate}
+\subsection{A recipe for sustainable software, Philip E. Bourne.}
+
+The first keynote
+\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delievered by Philip E. Bourne of UCSD. The slides
+are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+
+Philip is a basic biomedical scientist but has formed 4 software
+companies. He also helped found PLOS and PDB
+\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+
+He emphasised that Sustainability for software ``does not just mean more
+money from Government.''
+
+He noted that creating effective illustrations is a key function of
+software in the biomedical sciences: ``Automating Conceptualisation''
+(Visualising molecular structures is a problem that has now been solved,
+he is working on 3-d visualisations of the contents of cells.)
+
+He has looked into reproducibility in computational biology, and has
+concluded that ``I have proved I cannot reproduce research from my own
+lab'':
+(Computational
+Biology Resources Lack Persistence And Usability, Veretnik, Fink and
+Bourne.)\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1000136}}(
+
+He noted that in places, development in science has improved thanks to
+open source and Github, but for the most part remains arcane. He argues
+that there's a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and OpenScienceDataCloud
+\footnote{\url{https://www.opensciencedatacloud.org/}} as
+distribution channels.
+
+He noted a common evolutionary pathway for computational biology
+projects, from data archive to analytics platform to educational use. He
+suggests that use for outreach might be the final step.
+
+He has problems with staff retention with respect to private organisations which reward 
+those combining research and software expertise (the ``Google Bus''), however,
+he's a strong fan of private sector sustainability sources and PPP. He
+notes that making a successful business from scientific software alone
+is very rare: founders overvalue while customers undervalue. He noted
+that to last, an open source project needs a minimal funding requirement
+even with a vibrant community -- good will only goes so far if you're
+being paid to do something else.
+
+He talked about grant
+schemes
+ of relevance in the U.S., particularly with regard to technology transfer.
+\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
+
+He had problems with selling research software: his technology transfer
+office wanted wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the
+software. He was aware this was unrealistic but the TTO insisted for a
+while. He wants to push a much more one-click approach to purchasing
+university-written software.
+
+He then presented the usual arguments on directly valuing software as a
+research output alongside papers. Readers will be familiar with this
+debate. An interesting reference here is ``What
+does peer review mean when applied to computer code?, Rosemary Dickin.''
+\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}}
+
+He then expanded on the idea of the \textbf{digital enterprise}, where
+IT underpins the whole of organisational activities. He contends that
+universities are way behind the curve on this. In particular, he
+highlights the separation of research, teaching and administration into
+silos without a common IT framework as a blocker to many useful
+organisational innovations: ``University 2.0 is yet to happen.'' He
+spoke of a circumstance where someone had used an algorithm developed
+for computational biology in marketing.
+
+The role of an institution is important in this space. He argued that
+funders can help train institutions not just individuals.
+
+He concluded with a reference to his paper
+\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1002001}}``Ten
+simple rules for getting ahead as a computational biologist in
+academia'' and argued that computational scientists ``have a
+responsibility to convince their institutions and communities that
+software is scholarship,'' a point with which we strongly agree.
+
+\subsection{Scientifc Software and the Open Collaborative Web}
+
+The second keynote\footnote{\url{http://is.gd/wssspe}} was by
+ Arfon Smith\footnote{\url{http://arfon.org/}} of Github.
+
+His talk used an example from data reduction in Astronomy, where one
+attempts to remove interfering effects from the device. In this field, a
+``bad pixel mask'' is recalculated by each researcher, despite being
+persistent over time. He estimated 13 wasted person-years of waste.
+
+He asked, of this practice, ``why didn't we know better?'' and argued
+this was because we're taught to focus on immediate research outcomes
+and not on continuously improving and building on tools for research. He
+then asked, later in our working lives, when we do know better, why
+don't we act any different, and argued this was due to incentivisation:
+only the immediate products of research, not the software, are valued.
+
+He referenced Victoria Stodden's talk at OKCon
+\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},
+\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+which he said argued these points well.
+
+He cited the EMCEE \footnote{\url{https://github.com/dfm/emcee}} Markov Chain
+Monte Carlo code by Dan Forman Mackey \footnote{\url{http://dan.iel.fm/}} as a
+good example of a scientific project using GitHub. Amongst other
+examples of good scientific projects on GitHub he used an example from JoeZuntz
+\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}.
+
+C. Titus Brown \footnote{\url{http://ged.msu.edu/}} argues that with regard to
+reusable software, ``we should just start doing it''. Arfon notes that
+documentation should be ``treated as a first class entity.''
+
+He noted that some software fields have excellent cultures of module
+reuse, such as rubygems \footnote{\url{http://rubygems.org/}},
+PyPI \footnote{\url{https://pypi.python.org/pypi}}, and CPAN
+\footnote{\url{http://www.cpan.org/}}, where there is effectively
+ultra-low-friction collaboration through these repositories. This has
+not happened in highly numerical, compiled language scientific software.
+
+He argues that the GitHub Pull Request code review mechanism facilitates
+such collaboration, by allowing one to code first, and seek review and a
+merge back into the trunk later.
+
+Fernando Perez says that ``Open Source
+is reproducible by necessity'' - since reproducibility is a prerequisite
+for remote collaboration.
+
+Arfon therefore defines GitHub as part of the next stage in development
+of the web, following on from the ``social web'' of facebook et.al.:
+``The Collaborative Web.''
 
@@
 
+\section{Illustrative use-cases}
+
+Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+
 \section{Conclusions}

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
a few minor edits - a lot is still missing

--- 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
+
 \usepackage{color}
@@
 \usepackage{graphicx}
-\usepackage[latin9]{inputenc}
 \usepackage{multirow}
@@
 \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
-\item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
 \item James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
@@
 
-WSSSPE1 was organized via peer-review and organization of papers submitted in response to a call for papers.  The call for paper was created by the organizers and peer-review committee.
+WSSSPE1 was organized in collaboration of the relatively small group of organizers and
+a larger peer-review commitee. This commitee had early influence on the organization,
+e.g., on the specific call for papers.
 
-The workshop call for paper included:
+The workshop call for papers included:
 
@@
 Lamb, F.
-L\:{o}ffler, B.
+Löffler, B.
 Norris, B.
@@
 
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
 
@@
 \note{not sure if this was really cross-cutting...}
+
+\section{Illustrative use-cases}
+
+Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
 
@@
 
+Frank Löffler
+
 \bibliographystyle{plain}

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
minor changes in notes from James - keynote section

--- 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
+\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}
+
 People who have volunteered to work on this report: (with * for actual contributors - \note{if you contribute, add a * by your name})
@@
 
-Arfon Smith of github presented the first keynote. Following points summarize the keynote and the associated discussion:
-\begin{enumerate}
-    \item Socially driven software development.
-    \item Role of software in research.
-\end{enumerate}
+\subsection{A recipe for sustainable software, Philip E. Bourne.}
+
+The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delivered by Philip E. Bourne of UCSD. The slides
+are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+
+Philip is a basic biomedical scientist but has formed 4 software
+companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+
+He emphasized that Sustainability for software ``does not just mean more
+money from Government.''
+
+He noted that creating effective illustrations is a key function of
+software in the biomedical sciences: ``Automating Conceptualization''
+(Visualizing molecular structures is a problem that has now been solved,
+he is working on 3-d visualizations of the contents of cells.)
+
+He has looked into reproducibility in computational biology, and has
+concluded that ``I have proved I cannot reproduce research from my own
+lab'':
+(Computational
+Biology Resources Lack Persistence And Usability, Veretnik, Fink and
+Bourne.)\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1000136}} \todo{change this to a cite}
+
+He noted that in places, development in science has improved thanks to
+open source and Github, but for the most part remains arcane. He argues
+that there's a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
+OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
+distribution channels.
+
+He noted a common evolutionary pathway for computational biology
+projects, from data archive to analytics platform to educational use. He
+suggests that use for outreach might be the final step.
+
+He has problems with staff retention with respect to private organizations which reward 
+those combining research and software expertise (the ``Google Bus''), however,
+he's a strong fan of private sector sustainability sources and PPP. He
+notes that making a successful business from scientific software alone
+is very rare: founders overvalue while customers undervalue. He noted
+that to last, an open source project needs a minimal funding requirement
+even with a vibrant community -- good will only goes so far if you're
+being paid to do something else.
+
+He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}
+
+He had problems with selling research software: his technology transfer
+office wanted wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the
+software. He was aware this was unrealistic but the technology transfer
+office insisted for a
+while. He wants to push a much more one-click approach to purchasing
+university-written software.
+
+He then presented the usual arguments on directly valuing software as a
+research output alongside papers. Readers will be familiar with this
+debate. An interesting reference here is ``What
+does peer review mean when applied to computer code?, Rosemary Dickin.''
+\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}} \todo{change to cite}
+
+He then expanded on the idea of the \textbf{digital enterprise}, where
+IT underpins the whole of organizational activities. He contends that
+universities are way behind the curve on this. In particular, he
+highlights the separation of research, teaching and administration into
+silos without a common IT framework as a blocker to many useful
+organizational innovations: ``University 2.0 is yet to happen.'' He
+spoke of a circumstance where someone had used an algorithm developed
+for computational biology in marketing.
+
+The role of an institution is important in this space. He argued that
+funders can help train institutions not just individuals.
+
+He concluded with a reference to his paper
+\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1002001}}``Ten
+simple rules for getting ahead as a computational biologist in
+academia'' \todo{change to cite} and argued that computational scientists ``have a
+responsibility to convince their institutions and communities that
+software is scholarship,'' a point with which we strongly agree.
+
+\subsection{Scientific Software and the Open Collaborative Web}
+
+\todo{change references to papers or presentations to cites}
+
+The second keynote\footnote{\url{http://is.gd/wssspe}} was by
+ Arfon Smith\footnote{\url{http://arfon.org/}} of Github.
+
+His talk used an example from data reduction in Astronomy, where one
+attempts to remove interfering effects from the device. In this field, a
+``bad pixel mask'' is recalculated by each researcher, despite being
+persistent over time. He estimated 13 wasted person-years of waste.
+
+He asked, of this practice, ``why didn't we know better?'' and argued
+this was because we're taught to focus on immediate research outcomes
+and not on continuously improving and building on tools for research. He
+then asked, later in our working lives, when we do know better, why
+don't we act any different, and argued this was due to incentivization:
+only the immediate products of research, not the software, are valued.
+
+He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+which he said argued these points well.
+
+He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
+Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} as a
+good example of a scientific project using GitHub. Amongst other
+examples of good scientific projects on GitHub he used an example from Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}.
+
+C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
+reusable software, ``we should just start doing it''. Arfon notes that
+documentation should be ``treated as a first class entity.''
+
+He noted that some software fields have excellent cultures of module
+reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
+PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+ultra-low-friction collaboration through these repositories. This has
+not happened in highly numerical, compiled language scientific software.
+
+He argues that the GitHub Pull Request code review mechanism facilitates
+such collaboration, by allowing one to code first, and seek review and a
+merge back into the trunk later.
+
+Fernando Perez says that ``Open Source
+is reproducible by necessity'' - since reproducibility is a prerequisite
+for remote collaboration.
+
+Arfon therefore defines GitHub as part of the next stage in development
+of the web, following on from the ``social web'' of Facebook et.al.:
+``The Collaborative Web.''
 

### 2014-01-31 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
-\item Frank Loeffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
 \item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$

### 2014-02-01 WSSSPE1_report/WSSSPE1_summary_report.tex
changing some footnotes to cites, a few other small edits in the keynote section

@@
 concluded that ``I have proved I cannot reproduce research from my own
-lab'':
-(Computational
-Biology Resources Lack Persistence And Usability, Veretnik, Fink and
-Bourne.)\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1000136}} \todo{change this to a cite}
+lab''~\cite{Veretnik}.
 
@@
 does peer review mean when applied to computer code?, Rosemary Dickin.''
-\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}} \todo{change to cite}
+\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}}
 
@@
 
-He concluded with a reference to his paper
-\footnote{\url{http://www.ploscompbiol.org/article/info\%3Adoi\%2F10.1371\%2Fjournal.pcbi.1002001}}``Ten
-simple rules for getting ahead as a computational biologist in
-academia'' \todo{change to cite} and argued that computational scientists ``have a
+He concluded with a reference to his paper~\cite{bourne_ten} and argued that computational scientists ``have a
 responsibility to convince their institutions and communities that
@@
 
-His talk used an example from data reduction in Astronomy, where one
-attempts to remove interfering effects from the device. In this field, a
-``bad pixel mask'' is recalculated by each researcher, despite being
-persistent over time. He estimated 13 wasted person-years of waste.
+His talk used an example from data reduction in Astronomy, where he
+needed to remove interfering effects from the device. He built a
+``bad pixel mask'', and realized though while this is persistent, there
+was no way/practice of sharing this with the data, so many researchers
+repeated the same calculations. He estimated that 13 wasted person-years were wasted in
+this redundant calculation.
 
@@
 then asked, later in our working lives, when we do know better, why
-don't we act any different, and argued this was due to incentivization:
+don't we act any different, and argued this was due to incentives and their lack:
 only the immediate products of research, not the software, are valued.

### 2014-02-03 WSSSPE1_report/WSSSPE1_summary_report.tex
adding a note, fixing some spelling

@@
 
+\todo{mention arXiv:1311.3523 and explain what's different here}
+
 \section{Workshop Process}
@@
 WSSSPE1 was organized in collaboration of the relatively small group of organizers and
-a larger peer-review commitee. This commitee had early influence on the organization,
+a larger peer-review committee. This committee had early influence on the organization,
 e.g., on the specific call for papers.

### 2014-02-03 WSSSPE1_report/WSSSPE1_summary_report.tex
Added TODOs and refined Sections 1 to 3.

@@
 developers is emerging in academic environments who are an integral and embedded
-part of successful research teams. The WSSSPE workshop was intended provide a
+part of successful research teams. The WSSSPE workshop was intended to provide a
 forum for discussion of the challenges, including both positions and experiences.
@@
 
+\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July.
+It aims to further explore some of the key issues raised in WSSSPE.
+Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
+
 \todo{mention arXiv:1311.3523 and explain what's different here}
@@
 
-WSSSPE1 was organized in collaboration of the relatively small group of organizers and
-a larger peer-review committee. This committee had early influence on the organization,
+WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
+a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
 e.g., on the specific call for papers.
@@
 \begin{itemize}
-\item the development process that leads to new software
+\item the development process that leads to new (versions of) software
 \begin{itemize}
@@
 average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
-reviewers to provide scores on relevance and comments to the organizers, which were used to 
-decide which papers to associate with the workshop, and comments to the authors, which were 
-provided back to the authors to allow them to improve their papers.
+reviewers to provide scores on relevance and comments to the organizers and to the authors.
+The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.
 
@@
 high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
-that the reports were already self-published. The papers were also grouped into 3 areas, each of 
-which was associated with a panel and discussion at the workshop.
+that the reports were already self-published.
+The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}.
+Each subject area was associated with a panel and discussion at the workshop.
 
@@
 
-The workshop consisted of 2 keynote presentations and three panel/discussion sessions. The panels 
+The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
 were organized based on a classification of the workshop submissions into three categories, following 
-the themes of the call for papers as modified by the areas of the submissions. Representatives from 
+the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
 each submission categories were appointed as panelists, to discuss the papers in that category.
@@
 
-Philip is a basic biomedical scientist but has formed 4 software
+Philip is a basic biomedical scientist but has formed four software
 companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.

### 2014-02-03 WSSSPE1_report/WSSSPE1_summary_report.tex
Combined a few points to make sections 4.2 less fragmented.

@@
 open source and Github, but for the most part remains arcane. He argues
-that there's a lot we can learn from the app-store model. He
+that there is a lot we can learn from the app-store model. He
 mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
@@
 
-He has problems with staff retention with respect to private organizations which reward 
-those combining research and software expertise (the ``Google Bus''), however,
-he's a strong fan of private sector sustainability sources and PPP. He
-notes that making a successful business from scientific software alone
+He experienced problems with staff retention with respect to private organizations which reward 
+those combining research and software expertise (the ``Google Bus''). However,
+he is a strong fan of private sector sustainability sources and PPP. He
+noted that making a successful business from scientific software alone
 is very rare: founders overvalue while customers undervalue. He noted
 that to last, an open source project needs a minimal funding requirement
-even with a vibrant community -- good will only goes so far if you're
+even with a vibrant community -- good will only goes so far if you are
 being paid to do something else.
@@
 
-He had problems with selling research software: his technology transfer
-office wanted wanted huge IP reachthrough, whereby they would get a
+He had problems with selling research software: the technology transfer
+office in his university wanted huge IP reachthrough, whereby they would get a
 share of profits from drugs developed by pharma companies who use the
@@
 office insisted for a
-while. He wants to push a much more one-click approach to purchasing
+while. He wants to push a one-click approach to purchasing
 university-written software.
@@
 He asked, of this practice, ``why didn't we know better?'' and argued
-this was because we're taught to focus on immediate research outcomes
+this was because we were taught to focus on immediate research outcomes
 and not on continuously improving and building on tools for research. He
@@
 only the immediate products of research, not the software, are valued.
-
 He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
@@
 He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
-Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} as a
-good example of a scientific project using GitHub. Amongst other
-examples of good scientific projects on GitHub he used an example from Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}.
+Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} and Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}} as 
+good examples of  scientific projects using GitHub.
 
 C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
-reusable software, ``we should just start doing it''. Arfon notes that
+reusable software, ``we should just start doing it.'' Arfon noted that
 documentation should be ``treated as a first class entity.''
-
 He noted that some software fields have excellent cultures of module
@@
 
-He argues that the GitHub Pull Request code review mechanism facilitates
+He argued that the GitHub Pull Request code review mechanism facilitates
 such collaboration, by allowing one to code first, and seek review and a
@@
 for remote collaboration.
-
-Arfon therefore defines GitHub as part of the next stage in development
+Arfon therefore defined GitHub as part of the next stage in development
 of the web, following on from the ``social web'' of Facebook et.al.:

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex
Nancy's introductory work in the communities area.
Update WSSSPE1_summary_report.tex

@@
 
-\subsection{Papers}
+There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.
 
-\subsubsection*{Communities}
+\subsection{Communities Part 1}
+
+This collection of papers was summarized by Karen Cranston.
+
+\subsection{Communities Part 2}
+
+Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
+relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
+The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
+This is a 4-step iterative development process that incorporates Agile development principals.
 
@@
 
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+\item Step 1 - Design - thorough discussion of research questions
+\item Step 2 - Develop working code
+\item Step 3 - Refine based on new requirements
+\item Step 4 - Publish open source
+
+\end{itemize}
+
+The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
+code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
+build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
+Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
+In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
+Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
+Co‐locate rather than collaborating remotely whenever feasible.
+
+The Pierce paper described
+
+\subsection{Papers}
+
+\subsubsection*{Communities Part 1 Papers}
+
+\begin{itemize}
 
@@
 \item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
 
@@
 
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
-
 \item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
 Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
org change - removed agenda section started to change some URLs to cites fixed some todos

--- 2014-02-03 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}
+\note{papers and presentations and URLs should be references - see figshare-web in .bib file for URL style example}
 
@@
 
-\section{Workshop Process}
+\section{Workshop Process and Agenda}
 
-WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
-a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
+WSSSPE1 was organized in collaboration of the relatively small group of five organizers and
+a larger peer-review committee with 36 members. This committee had early influence on the organization,
 e.g., on the specific call for papers.
@@
 58 submissions were received, and almost all submitters used either 
-arXiv\footnote{\url{http://arxiv.org}} or 
-figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+arXiv~\cite{arXiv-web} or 
+figshare~\cite{figshare-web} to self-publish their papers.
 
@@
 
-\section{Agenda}
-
-The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
+The workshop itself consisted of two keynote presentations and the three panels/discussion sessions. The panels 
 were organized based on a classification of the workshop submissions into three categories, following 
-the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
-each submission categories were appointed as panelists, to discuss the papers in that category.
+the themes of the call for papers as modified by the areas of the submissions. Three to four representatives from 
+each submission category were appointed as panelists, and assigned to read a subset of the paper in that category and then discuss them in the panel.
 
@@
 
+The WSSSPE1 workshop began with two keynote presentations.
+
 \todo{section should include what was presented and what was discussed}
+
+\todo{change references to papers or presentations to cites}
 
@@
 
-He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}
+He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
 
@@
 
-\subsection{Scientific Software and the Open Collaborative Web}
-
-\todo{change references to papers or presentations to cites}
+\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}
 
@@
 
-\section{Illustrative use-cases}
+\section{Illustrative Use Cases}
 
 Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+
+\note{Dan: do we need this section?
+I think there's some overlap between this and the discussion in the Developing and Support Software section.
+Or perhaps this should go in the cross-cutting issues section?}
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Added one additional attendee I was aware of
footnote'd nsf vision link

@@
 
-The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
+The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1, \url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13, \url{http://sc13.supercomputing.org}).
 
@@
 The level of interest in the workshop has led the organizers, working with some of the submitters
-and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
+and some of the attendees, to plan two additional workshops (at the 2014 SciPy and SC conferences) and to turn
 the workshop website into a community website that can be used as a focus for further
 discussion and progress.
+
+Before the WSSSPE1 workshop took place, the organizers self-published a paper~\cite{WSSSPE1-pre-report} to document the process of organizing and advertising the workshop, collecting and reviewing the papers, and putting together the agenda.
+Section~\ref{sec:process} is based on that paper.
+The remainder of this paper is based on the workshop itself, as documented in a set of collaborative notes taken during the workshop~\cite{WSSSPE1-google-notes}, including discussion about the keynote presentations (\S\ref{sec:keynotes}), the three panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion (\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}), use cases (\S\ref{sec:use-cases}), and conclusions (\S\ref{sec:conclusions}).
 
@@
 
-\todo{mention arXiv:1311.3523 and explain what's different here}
-
-\section{Workshop Process and Agenda}
+\section{Workshop Process and Agenda} \label{sec:process}
 
@@
 requirements that bound it. Roughly following the example of NSF's Vision and Strategy for Software 
-(\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, 
+\footnote{\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113}}~\cite{NSF_software_vision}, 
 the ecosystem may be viewed as having challenges related to:
@@
 
-\section{Keynotes}
+\section{Keynotes} \label{sec:keynotes}
 
@@
 mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
-OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
+OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}} as
 distribution channels.
@@
 
-\section{Developing and Supporting Software}
+\section{Developing and Supporting Software} \label{sec:devel}
 
@@
 
-\section{Policy}
+\section{Policy} \label{sec:policy}
 
@@
 
-\section{Communities}
+\section{Communities} \label{sec:community}
 
@@
 
-\section{Other Discussion}
+\section{Other Discussion} \label{sec:other}
 
@@
 
-\section{Cross-cutting Issues}
+\section{Cross-cutting Issues}  \label{sec:cross-cutting}
 
@@
 
-\section{Illustrative Use Cases}
+\section{Illustrative Use Cases} \label{sec:use-cases}
 
@@
 
-\section{Conclusions}
+\section{Conclusions} \label{sec:conclusions}
 
@@
 
+Patrick O'Leary
+
 \bibliographystyle{plain}

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
rollback change in CFP *quotation* to original

@@
 requirements that bound it. Roughly following the example of NSF's Vision and Strategy for Software 
-\footnote{\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113}}~\cite{NSF_software_vision}, 
+(\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, 
 the ecosystem may be viewed as having challenges related to:

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
a note and some corrections

--- 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 those combining research and software expertise (the ``Google Bus''). However,
-he is a strong fan of private sector sustainability sources and PPP. He
+he is a strong fan of private sector sustainability sources and PPP\note{what does PPP mean?}. He
 noted that making a successful business from scientific software alone
@@
 needed to remove interfering effects from the device. He built a
-``bad pixel mask'', and realized though while this is persistent, there
+``bad pixel mask'', and realized that while this is persistent, there
 was no way/practice of sharing this with the data, so many researchers
-repeated the same calculations. He estimated that 13 wasted person-years were wasted in
+repeated the same calculations. He estimated that 13 person-years were wasted in
 this redundant calculation.

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
adding sustainability notes in 8.1 resolving a question

@@
 those combining research and software expertise (the ``Google Bus''). However,
-he is a strong fan of private sector sustainability sources and PPP\note{what does PPP mean?}. He
+he is a strong fan of private sector sustainability sources and public-private partnerships. He
 noted that making a successful business from scientific software alone
@@
 
+\begin{itemize}
+
+\item From discussion in Phil Bourne's keynote:
+\begin{itemize}
+\item Can sustainability be defined as the effort that happens to make the essential things continue?
+\item What is sustainability? What is it that we want to sustain? Is what we want to sustain valuable? Who cares if it goes away and how do you measure how much they care?
+\end{itemize}
+
+\item from discussion in developing panel
+
+\begin{itemize}
+\item what does sustainable mean?
+\begin{itemize}
+\item from http://oss-watch.ac.uk/resources/ssmm: When choosing software for procurement or development reuse - regardless of the license and development model you will use - you need to consider the future.
+While a software product may satisfy today's needs, will it satisfy tomorrow's needs?
+Will the supplier still be around in five years' time?
+Will the supplier still care for all its customers in five years' time?
+Will the supplier be responsive to bug reports and feature requests?
+In other words, is the software sustainable?
+\item I think this is a key question we as a community need to agree on (whats our problem statement?) and hope we come out today with some progress. - @VonWelch
+\item What is the goal of sustainability? Reproducible science, persistence, quality, or else?
+\item How do you measure success? How do you know when you have actually achieved sustainable software?
+\item Sustainability should be addressed throughout the life cycle (Daisie Huang)
+\end{itemize}
+
+\item What is the relationship between sustainability and provenance?
+
+\item Is usability separate from sustainability or a fundamental part?
+\end{itemize}
+
+\item from discussion in policy panel
+
+\begin{itemize}
+\item Chris Lenhardt's presentation of Venters paper~\cite{Venters_WSSSPE} included:
+
+\begin{itemize}
+\item What does software sustainability mean
+It ``is a rather ambiguous concept...''
+\item Lack of an accepted definition gets in the way of integrating the concept into software engineering
+\item Sustainability as a non-functional requirement
+\item Quality of software architectures determines sustainability
+\item Propose sustainability is a measure of central quality attributes:
+Extensibility,
+Interoperability,
+Maintainability,
+Portability,
+Reusability,
+Scalability
+\item Develop an architecture evaluation framework based on scenarios which help to illuminate how to measure quality/sustainability at the architectural level
+\end{itemize}
+
+\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE} included:
+\begin{itemize}
+\item Clarifying the notion of sustainability
+\begin{itemize}
+\item Provide same functionality going forward even if environment changes
+\item Is functionality and usability clear
+\item Will functionality be correct going into the future even if environment changes
+\item Is the functionality desired by current and future users
+\item Can it incorporate new things as they develop
+\end{itemize}
+\end{itemize}
+
+\item see slide 9 of \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
+
+\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
+
+\end{itemize}
+
+\end{itemize}
+
 \subsection{Licensing}

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
moving licensing from cross-cutting to note in conclusions
small updates in process section

@@
 average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
-reviewers to provide scores on relevance and comments to the organizers and to the authors.
-The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.
+reviewers to choose papers they wanted to review, then to provide scores on their relevance and general comments to the organizers and to the authors.
+The review reports enabled the organizers to decide which papers to associate with the workshop, and allowed the authors to improve their papers.
 
@@
 
-\section{Cross-cutting Issues}  \label{sec:cross-cutting}
-
-\subsection{Defining sustainability}
+\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
 
@@
 
-\subsection{Licensing}
-
-\note{not sure if this was really cross-cutting...}
-
 \section{Illustrative Use Cases} \label{sec:use-cases}
@@
 \todo{add some analysis of the deficiencies and difficulties that are present in different fields, and those that are common?}
+
+\todo{say something about licensing - lessons, advice, etc.?}
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
flag about sustainability discussion
List of Attendees Appendix: removing twitter/email/affil info

@@
 \newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
+
+\usepackage{array}
+\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 Software plans from the funding agencies?
-\todo{mention as cross-cutting, since this came up in other parts of the workshop as well?}
+This topic is expanded upon in \S\ref{sec:cross-cutting}, since it was discussed up in multiple parts of the WSSSPE1 workshop.
 
@@
 
-Daniel S. Katz (@danielskatz)
-
-Neil Chue Hong (@npch)
-
-Gabrielle Allen
-
-David Proctor (@djproctor)
-
-Manish Parashar
-
-Jay Alameda (alameda@illinois.edu)
-
-David Andrs
-
-Lorena A Barba (@LorenaABarba)
-
-David Bernholdt (bernholdtde@ornl.gov)
-
-Phil Bourne
-
-Karl Broman (@kwbroman)
-
-Maxine Brown, UIC (maxine@uic.edu)
-
-David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)
-
-Bruno Bzeznik
-
-Jeffrey Carver, University of Alabama (carver@cs.ua.edu)
-
-John W. Cobb (cobbjw@ornl.gov, @johnwcobb)
-
-Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)
-
-Rion Dooley
-
-Anshu Dubey (adubey@lbl.gov)
-
-Ian Foster, @ianfoster
-
-Juliana Freire
-
-Derek Gaston
-
-Derek Groen (@whydoitweet on Twitter)
-
-Edmund Hart (@distribecology)
-
-James Hetherington @jamespjh
-
-Simon Hettrick (s.hettrick@software.ac.uk)
-
-Kenneth Hoste (@kehoste, Ghent University, Belgium)
-
-James Howison, University of Texas at Austin (@jameshowison)
-
-Daisie Huang
-
-Kerk Kee (kerk.kee@gmail.com)
-
-Arnaud Legrand
-
-Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu
-
-Louis Maddox (@lmmddx)
-
-Ketan Maheshwari (@km\_9 on Twitter)
-
-Suresh Marru (smarru@iu.edu, @sureshmarru)
-
-Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek
-
-James McClure
-
-Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org
-
-J. Yates Monteith, Clemson University
-
-Pierre Neyron
-
-Lucas Nussbaum (lucas.nussbaum@loria.fr)
-
-Cody Permann
-
-John Peterson
-
-Quan Pham, University of Chicago
-
-Marlon Pierce, marpierc@iu.edu, @marpierc
-
-Heather Piwowar
-
-Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places
-
-Andreas Schreiber (@onyame)
-
-William Scullin
-
-Andrew Slaughter
-
-Arfon Smith
-
-James Spencer (@exp\_n)
-
-Shel Swenson, University of Southern California (shel.swenson@usc.edu)
-
-Fr\'{e}d\'{e}ric Suter
-
-Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)
-
-George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)
-
-John Towns
-
-Erik Trainer
-
-Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)
-
-Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk
-
-Colin C. Venters, c.venters@hud.ac.uk
-
-Von Welch, @VonWelch
-
-Nancy Wilkins-Diehr, wilkinsn@sdsc.edu
-
-Theresa Windus (twindus@iastate.edu)
-
-Rich Wolski
-
-Lynn Zentner
-
-Frank Löffler
-
-Patrick O'Leary
+\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
+   Daniel S. Katz
+&  Neil Chue Hong
+&  Gabrielle Allen
+&  David Proctor
+\\ Manish Parashar
+&  Jay Alameda
+&  David Andrs
+&  Lorena A Barba
+\\ David Bernholdt
+&  Phil Bourne
+&  Karl Broman
+&  Maxine Brown
+\\ David Bruhwiler
+&  Bruno Bzeznik
+&  Jeffrey Carver
+&  John W. Cobb
+\\ Karen Cranston
+&  Rion Dooley
+&  Anshu Dubey
+&  Ian Foster
+\\Juliana Freire
+& Derek Gaston
+& Derek Groen
+& Edmund Hart
+\\James Hetherington
+& Simon Hettrick
+& Kenneth Hoste
+& James Howison
+\\Daisie Huang
+& Kerk Kee
+& Arnaud Legrand
+& Paul Madden
+\\Louis Maddox
+& Ketan Maheshwari
+& Suresh Marru
+& Cezary Mazurek
+\\James McClure
+& Chris Mentzel
+& J. Yates Monteith
+& Pierre Neyron
+\\Lucas Nussbaum
+& Cody Permann
+& John Peterson
+& Quan Pham
+\\Marlon Pierce
+& Heather Piwowar
+& Jason Riedy
+& Andreas Schreiber
+\\William Scullin
+& Andrew Slaughter
+& Arfon Smith
+& James Spencer
+\\Shel Swenson
+& Fr\'{e}d\'{e}ric Suter
+& Andy Terrel
+& George  Thiruvathukal
+\\John Towns
+& Erik Trainer
+& Matthew Turk
+& Eric Van Wyk
+\\Colin C. Venters
+& Von Welch
+& Nancy Wilkins-Diehr
+& Theresa Windus
+\\Rich Wolski
+& Lynn Zentner
+& Frank Löffler
+& Patrick O'Leary
+\end{tabular}
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Introduced line breaks for easier diff.

@@
 
-\note{papers and presentations and URLs should be references - see figshare-web in .bib file for URL style example}
+\note{papers and presentations and URLs should be references - see
+  figshare-web in .bib file for URL style example}
 
-People who have volunteered to work on this report: (with * for actual contributors - \note{if you contribute, add a * by your name})
+People who have volunteered to work on this report: (with * for actual
+contributors - \note{if you contribute, add a * by your name})
 
@@
 
-The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1, \url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13, \url{http://sc13.supercomputing.org}).
+The First Workshop on Sustainable Software for Science: Practice and
+Experiences (WSSSPE1,
+\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
+Sunday, 17 November 2013, in conjunction with the 2013 International
+Conference for High Performance Computing, Networking, Storage and
+Analysis (SC13, \url{http://sc13.supercomputing.org}).
 
-Because progress in scientific research is dependent on the quality and accessibility
-of software at all levels, it is now critical to address many challenges related to the
-development, deployment, and maintenance of reusable software. In addition, it is
-essential that scientists, researchers, and students are able to learn and adopt
-software-related skills and methodologies. Established researchers are already
-acquiring some of these skills, and in particular a specialized class of software
-developers is emerging in academic environments who are an integral and embedded
-part of successful research teams. The WSSSPE workshop was intended to provide a
-forum for discussion of the challenges, including both positions and experiences.
-The short papers and discussion have been archived to provide a basis for continued
-discussion, and the workshop has fed into the collaborative writing of this journal publication.
-Additional papers based on extended versions of workshop submissions are expected.
-The level of interest in the workshop has led the organizers, working with some of the submitters
-and some of the attendees, to plan two additional workshops (at the 2014 SciPy and SC conferences) and to turn
-the workshop website into a community website that can be used as a focus for further
-discussion and progress.
+Because progress in scientific research is dependent on the quality
+and accessibility of software at all levels, it is now critical to
+address many challenges related to the development, deployment, and
+maintenance of reusable software. In addition, it is essential that
+scientists, researchers, and students are able to learn and adopt
+software-related skills and methodologies. Established researchers are
+already acquiring some of these skills, and in particular a
+specialized class of software developers is emerging in academic
+environments who are an integral and embedded part of successful
+research teams. The WSSSPE workshop was intended to provide a forum
+for discussion of the challenges, including both positions and
+experiences.  The short papers and discussion have been archived to
+provide a basis for continued discussion, and the workshop has fed
+into the collaborative writing of this journal publication.
+Additional papers based on extended versions of workshop submissions
+are expected.  The level of interest in the workshop has led the
+organizers, working with some of the submitters and some of the
+attendees, to plan two additional workshops (at the 2014 SciPy and SC
+conferences) and to turn the workshop website into a community website
+that can be used as a focus for further discussion and progress.
 
-Before the WSSSPE1 workshop took place, the organizers self-published a paper~\cite{WSSSPE1-pre-report} to document the process of organizing and advertising the workshop, collecting and reviewing the papers, and putting together the agenda.
-Section~\ref{sec:process} is based on that paper.
-The remainder of this paper is based on the workshop itself, as documented in a set of collaborative notes taken during the workshop~\cite{WSSSPE1-google-notes}, including discussion about the keynote presentations (\S\ref{sec:keynotes}), the three panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion (\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}), use cases (\S\ref{sec:use-cases}), and conclusions (\S\ref{sec:conclusions}).
+Before the WSSSPE1 workshop took place, the organizers self-published
+a paper~\cite{WSSSPE1-pre-report} to document the process of
+organizing and advertising the workshop, collecting and reviewing the
+papers, and putting together the agenda.  Section~\ref{sec:process} is
+based on that paper.  The remainder of this paper is based on the
+workshop itself, as documented in a set of collaborative notes taken
+during the workshop~\cite{WSSSPE1-google-notes}, including discussion
+about the keynote presentations (\S\ref{sec:keynotes}), the three
+panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
+(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
+use cases (\S\ref{sec:use-cases}), and conclusions
+(\S\ref{sec:conclusions}).
 
-\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July.
-It aims to further explore some of the key issues raised in WSSSPE.
-Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
+\todo{Mention a two-session minisymposium on ``Reliable Computational
+  Science'' in the SIAM Annual Meeting in July. It aims to further
+  explore some of the key issues raised in WSSSPE. Speakers include
+  Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
 
@@
 
-WSSSPE1 was organized in collaboration of the relatively small group of five organizers and
-a larger peer-review committee with 36 members. This committee had early influence on the organization,
-e.g., on the specific call for papers.
+WSSSPE1 was organized in collaboration of the relatively small group
+of five organizers and a larger peer-review committee with 36
+members. This committee had early influence on the organization, e.g.,
+on the specific call for papers.
 
@@
 \begin{quote}
-In practice, scientific software activities are part of an ecosystem where key roles are held by 
-developers, users, and funders. All three groups supply resources to the ecosystem, as well as 
-requirements that bound it. Roughly following the example of NSF's Vision and Strategy for Software 
+In practice, scientific software activities are part of an ecosystem
+where key roles are held by developers, users, and funders. All three
+groups supply resources to the ecosystem, as well as requirements that
+bound it. Roughly following the example of NSF's Vision and Strategy
+for Software
 (\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision},
@@
 \begin{itemize}
-\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
+\item how fundamental research in computer science or
+  science/engineering domains is turned into reusable software
 \item software created as a by-product of research
-\item impact of computer science research on the development of scientific software
+\item impact of computer science research on the development of
+  scientific software
 \end{itemize}
-\item the support and maintenance of existing software, including software engineering
+\item the support and maintenance of existing software, including
+  software engineering
 \begin{itemize}
 \item governance, business, and sustainability models
-\item the role of community software repositories, their operation and sustainability
+\item the role of community software repositories, their operation and
+  sustainability
 \end{itemize}
@@
 \item career paths for developers and institutional roles
-\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of publishers
+\item issues related to multiple organizations and multiple countries,
+  such as intellectual property, licensing, etc.
+\item mechanisms and venues for publishing software, and the role of
+  publishers
 \end{itemize}
@@
 
-Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, the organizers wanted to make submission as easy as possible.
-The call for papers stated:
+Based on the goal of encouraging a wide range of submissions from
+those involved in software practice, ranging from initial thoughts and
+partial studies to mature deployments, the organizers wanted to make
+submission as easy as possible. The call for papers stated:
 
 \begin{quote}
-We invite short (4-page) position/experience reports that will be used to organize panel and 
-discussion sessions. These papers will be archived by a third-party service, and provided DOIs. We 
-encourage submitters to license their papers under a Creative Commons license that encourages 
-sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop. 
-An interactive site will be created to link these papers and the workshop discussion, with options for 
-later comments and contributions. Contributions will be peer-reviewed for relevance and originality 
-before the links are added to the workshop site; contributions will also be used to determine 
-discussion topics and panelists. We will also plan one or more papers to be collaboratively 
-developed by the contributors, based on the panels and discussions.
+We invite short (4-page) position/experience reports that will be used
+to organize panel and discussion sessions. These papers will be
+archived by a third-party service, and provided DOIs. We encourage
+submitters to license their papers under a Creative Commons license
+that encourages sharing and remixing, as we will combine ideas (with
+attribution) into the outcomes of the workshop.  An interactive site
+will be created to link these papers and the workshop discussion, with
+options for later comments and contributions. Contributions will be
+peer-reviewed for relevance and originality before the links are added
+to the workshop site; contributions will also be used to determine
+discussion topics and panelists. We will also plan one or more papers
+to be collaboratively developed by the contributors, based on the
+panels and discussions.
 \end{quote}
@@
 58 submissions were received, and almost all submitters used either
-arXiv~\cite{arXiv-web} or 
-figshare~\cite{figshare-web} to self-publish their papers.
+arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
+their papers.
 
-A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
-average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
-reviewers to choose papers they wanted to review, then to provide scores on their relevance and general comments to the organizers and to the authors.
-The review reports enabled the organizers to decide which papers to associate with the workshop, and allowed the authors to improve their papers.
+A peer review process followed the submissions, where the 58 papers
+received 181 reviews, an average of 3.12 reviews per paper. Reviews
+were completed using a Google form, which allowed reviewers to choose
+papers they wanted to review, then to provide scores on their
+relevance and general comments to the organizers and to the
+authors. The review reports enabled the organizers to decide which
+papers to associate with the workshop, and allowed the authors to
+improve their papers.
 
-The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
-high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
-that the reports were already self-published.
-The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}.
-Each subject area was associated with a panel and discussion at the workshop.
+The organizers decided to list 54 of the papers as significantly
+contributing to the workshop, a very high acceptance rate, but one
+that is reasonable, given the goal of broad participation and the fact
+that the reports were already self-published. The papers were also
+grouped into three main categories, namely \emph{Developing and
+  Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
+subject area was associated with a panel and discussion at the
+workshop.
 
-The workshop itself consisted of two keynote presentations and the three panels/discussion sessions. The panels 
-were organized based on a classification of the workshop submissions into three categories, following 
-the themes of the call for papers as modified by the areas of the submissions. Three to four representatives from 
-each submission category were appointed as panelists, and assigned to read a subset of the paper in that category and then discuss them in the panel.
+The workshop itself consisted of two keynote presentations and the
+three panels/discussion sessions. The panels were organized based on a
+classification of the workshop submissions into three categories,
+following the themes of the call for papers as modified by the areas
+of the submissions. Three to four representatives from each submission
+category were appointed as panelists, and assigned to read a subset of
+the paper in that category and then discuss them in the panel.
 
@@
 
-The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
-was delivered by Philip E. Bourne of UCSD. The slides
-are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+The first
+keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delivered by Philip E. Bourne of UCSD. The slides are
+online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
 
 Philip is a basic biomedical scientist but has formed four software
-companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+companies. He also helped found PLOS and
+PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
 
-He emphasized that Sustainability for software ``does not just mean more
-money from Government.''
+He emphasized that Sustainability for software ``does not just mean
+more money from Government.''
 
@@
 software in the biomedical sciences: ``Automating Conceptualization''
-(Visualizing molecular structures is a problem that has now been solved,
-he is working on 3-d visualizations of the contents of cells.)
+(Visualizing molecular structures is a problem that has now been
+solved, he is working on 3-d visualizations of the contents of cells.)
 
@@
 He noted that in places, development in science has improved thanks to
-open source and Github, but for the most part remains arcane. He argues
-that there is a lot we can learn from the app-store model. He
-mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
-OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}} as
-distribution channels.
+open source and Github, but for the most part remains arcane. He
+argues that there is a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}}
+and
+OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
+as distribution channels.
 
 He noted a common evolutionary pathway for computational biology
-projects, from data archive to analytics platform to educational use. He
-suggests that use for outreach might be the final step.
+projects, from data archive to analytics platform to educational
+use. He suggests that use for outreach might be the final step.
 
-He experienced problems with staff retention with respect to private organizations which reward 
-those combining research and software expertise (the ``Google Bus''). However,
-he is a strong fan of private sector sustainability sources and public-private partnerships. He
+He experienced problems with staff retention with respect to private
+organizations which reward those combining research and software
+expertise (the ``Google Bus''). However, he is a strong fan of private
+sector sustainability sources and public-private partnerships. He
 noted that making a successful business from scientific software alone
 is very rare: founders overvalue while customers undervalue. He noted
-that to last, an open source project needs a minimal funding requirement
-even with a vibrant community -- good will only goes so far if you are
-being paid to do something else.
+that to last, an open source project needs a minimal funding
+requirement even with a vibrant community -- good will only goes so
+far if you are being paid to do something else.
 
-He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
+He talked about grant schemes of relevance in the U.S., particularly
+with regard to technology
+transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
 
-He had problems with selling research software: the technology transfer
-office in his university wanted huge IP reachthrough, whereby they would get a
-share of profits from drugs developed by pharma companies who use the
-software. He was aware this was unrealistic but the technology transfer
-office insisted for a
-while. He wants to push a one-click approach to purchasing
-university-written software.
+He had problems with selling research software: the technology
+transfer office in his university wanted huge IP reachthrough, whereby
+they would get a share of profits from drugs developed by pharma
+companies who use the software. He was aware this was unrealistic but
+the technology transfer office insisted for a while. He wants to push
+a one-click approach to purchasing university-written software.
 
@@
 
-He concluded with a reference to his paper~\cite{bourne_ten} and argued that computational scientists ``have a
-responsibility to convince their institutions and communities that
-software is scholarship,'' a point with which we strongly agree.
+He concluded with a reference to his paper~\cite{bourne_ten} and
+argued that computational scientists ``have a responsibility to
+convince their institutions and communities that software is
+scholarship,'' a point with which we strongly agree.
 
@@
 His talk used an example from data reduction in Astronomy, where he
-needed to remove interfering effects from the device. He built a
-``bad pixel mask'', and realized that while this is persistent, there
-was no way/practice of sharing this with the data, so many researchers
-repeated the same calculations. He estimated that 13 person-years were wasted in
-this redundant calculation.
+needed to remove interfering effects from the device. He built a ``bad
+pixel mask'', and realized that while this is persistent, there was no
+way/practice of sharing this with the data, so many researchers
+repeated the same calculations. He estimated that 13 person-years were
+wasted in this redundant calculation.
 
 He asked, of this practice, ``why didn't we know better?'' and argued
-this was because we were taught to focus on immediate research outcomes
-and not on continuously improving and building on tools for research. He
-then asked, later in our working lives, when we do know better, why
-don't we act any different, and argued this was due to incentives and their lack:
-only the immediate products of research, not the software, are valued.
-He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+this was because we were taught to focus on immediate research
+outcomes and not on continuously improving and building on tools for
+research. He then asked, later in our working lives, when we do know
+better, why don't we act any different, and argued this was due to
+incentives and their lack: only the immediate products of research,
+not the software, are valued.  He referenced Victoria Stodden's talk
+at
+OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
 which he said argued these points well.
 
-He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
-Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} and Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}} as 
-good examples of  scientific projects using GitHub.
+He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
+Chain Monte Carlo code by Dan Forman
+Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
+Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
+as good examples of scientific projects using GitHub.
 
-C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
-reusable software, ``we should just start doing it.'' Arfon noted that
-documentation should be ``treated as a first class entity.''
-He noted that some software fields have excellent cultures of module
-reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
-PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
+regard to reusable software, ``we should just start doing it.'' Arfon
+noted that documentation should be ``treated as a first class
+entity.''  He noted that some software fields have excellent cultures
+of module reuse, such as
+rubygems\footnote{\url{http://rubygems.org/}},
+PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
+CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
 ultra-low-friction collaboration through these repositories. This has
-not happened in highly numerical, compiled language scientific software.
+not happened in highly numerical, compiled language scientific
+software.
 
@@
 
-Fernando Perez says that ``Open Source
-is reproducible by necessity'' - since reproducibility is a prerequisite
-for remote collaboration.
-Arfon therefore defined GitHub as part of the next stage in development
-of the web, following on from the ``social web'' of Facebook et.al.:
-``The Collaborative Web.''
+Fernando Perez says that ``Open Source is reproducible by necessity''
+- since reproducibility is a prerequisite for remote collaboration.
+Arfon therefore defined GitHub as part of the next stage in
+development of the web, following on from the ``social web'' of
+Facebook et.al.: ``The Collaborative Web.''
 
@@
 
-\item Mark C.
-Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith.
-Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}
+\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
+  McInnes, Barry Smith. Package Management Practices Essential for
+  Interoperability: Lessons Learned and Strategies Developed for
+  FASTMath \cite{Miller_WSSSPE}
 
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable
+  \cite{Broman_WSSSPE}
 
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
 
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
+  Riihimaki. What Comes First, the OWL or the Bean?
+  \cite{Stephan_WSSSPE}
 
-\item Derek R.
-Gaston, John Peterson, Cody J.
-Permann, David Andrs, Andrew E.
-Slaughter, Jason M.
-Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}
+\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
+  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
+  Concurrent Computational Framework and Application Development
+  \cite{Gaston_WSSSPE}
 
-\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}
+\item Anshu Dubey, B. Van Straalen. Experiences from Software
+  Engineering of Large Scale AMR Multiphysics Code Frameworks
+  \cite{Dubey_WSSSPE}
 
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development \cite{Blatt_WSSSPE}
 
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
 
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software \cite{Ahern_WSSSPE}
 
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software
+  \cite{Choi_WSSSPE}
 
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
+  adapting sustainable scientific software development processes in a
+  small biology lab \cite{Crusoe_WSSSPE}
 
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. 
+The MVAPICH Project: Evolution and Sustainability of an Open Source 
+Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
 
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
+  H. Kellogg. Experiences with Automated Build and Test for
+  Geodynamics Simulation Codes \cite{Heien_WSSSPE}
 
@@
 
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter.
-SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
+  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
+  Versatile Simulation of Large Scale Distributed
+  Systems~\cite{Casanova_WSSSPE}
 
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software \cite{Trainer_WSSSPE}
 
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin.
-Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
+  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
+  Sherwin. Simplifying the Development, Use and Sustainability of HPC
+  Software \cite{Cohen_WSSSPE}
 
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
+  Deployment of Scientific and Engineering Applications
+  \cite{Slawinski_WSSSPE}
 
@@
 
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
+  Open Development of Scientific Software \cite{Prlic_WSSSPE}
 
-\item Anshu Dubey, S.
-Brandt, R.
-Brower, M.
-Giles, P.
-Hovland, D.
-Q.
-Lamb, F.
-Löffler, B.
-Norris, B.
-O'Shea, C.
-Rebbi, M.
-Snir, R.
-Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
+\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
+  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
+  R. Thakur, Software Abstractions and Methodologies for HPC
+  Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
 
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
+  Need Not Be Difficult \cite{Carver_WSSSPE}
 
-\item Craig A.
-Stewart, Julie Wernert, Eric A.
-Wernert, William K.
-Barnett, Von Welch.
-Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}
+\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
+  K. Barnett, Von Welch. Initial Findings from a Study of Best
+  Practices and Models for Cyberinfrastructure Software Sustainability
+  \cite{Stewart_WSSSPE}
 
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
+  anything less is unsustainable \cite{Brown_WSSSPE}
 
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc.
-Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
+  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
+  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
+  Platforms \cite{Swenson_WSSSPE}
 
@@
 
-Discussion of software produced as a by-product versus software developed for reuse.
-How does this change the project, who develops the code, growth beyond 1--2 person projects to larger projects with diverse set of consumers and contributors.
-Modular and extensible code versus working for current research problem---tackled by same people, or a collaborative team?
+Discussion of software produced as a by-product versus software
+developed for reuse. How does this change the project, who develops
+the code, growth beyond 1--2 person projects to larger projects with
+diverse set of consumers and contributors. Modular and extensible code
+versus working for current research problem---tackled by same people,
+or a collaborative team?
 
@@
 
-What is sustainable?
-What are the best practices, can workshops fund meetings to help define and improve best practices in these areas?
-Software plans from the funding agencies?
-This topic is expanded upon in \S\ref{sec:cross-cutting}, since it was discussed up in multiple parts of the WSSSPE1 workshop.
+What is sustainable? What are the best practices, can workshops fund
+meetings to help define and improve best practices in these areas?
+Software plans from the funding agencies? This topic is expanded upon
+in \S\ref{sec:cross-cutting}, since it was discussed up in multiple
+parts of the WSSSPE1 workshop.
 
@@
 
-Discussion of the evolving role of scientists, and/or others that fill these roles.
+Discussion of the evolving role of scientists, and/or others that fill
+these roles.
 
@@
 
-There are a lot of tools out there, but few are currently used. Look at what some projects have found successful, and how to automate as much as possible to reduce additional overhead.
+There are a lot of tools out there, but few are currently used. Look
+at what some projects have found successful, and how to automate as
+much as possible to reduce additional overhead.
 
@@
 
-Its impact on how and where research products are used, who can collaborate and what patterns have worked/not worked in existing communities.
+Its impact on how and where research products are used, who can
+collaborate and what patterns have worked/not worked in existing
+communities.
 
@@
 
-How initial software development is funded, moving to follow up projects, maintenance, community growth, collaborating with other institutions, labs, industry, internationally.
-Contract relationship versus collaborative development between scientists and software developers.
+How initial software development is funded, moving to follow up
+projects, maintenance, community growth, collaborating with other
+institutions, labs, industry, internationally. Contract relationship
+versus collaborative development between scientists and software
+developers.
 
@@
 
-Who creates training materials, how are they distributed, integration into courses when projects see very wide application in research.
+Who creates training materials, how are they distributed, integration
+into courses when projects see very wide application in research.
 
@@
 
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
+  Software Product Sustainability Model \cite{Calero_WSSSPE}
 
-\item Colin C.
-Venters, Lydia Lau, Michael K.
-Griffiths, Violeta Holmes, Rupert R.
-Ward, Jie Xu.
-The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}
+\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
+  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
+  Towards a Software Sustainability Architectural Evaluation Framework
+  \cite{Venters_WSSSPE}
 
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
+  Cyberinfrastructure Software Through Open Governance
+  \cite{Pierce_WSSSPE}
 
-\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
+\item Daniel S. Katz, David Proctor. A Framework for Discussing
+  e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
 
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak.
-Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
+  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
+  Lifecycle Management in the Context of Conducting Science
+  \cite{Lenhardt_WSSSPE}
 
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
+  Ecological Metaphors for Sustainable Software in Science
+  \cite{Weber_WSSSPE}
 
@@
 
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
+  Smith. Accurately Citing Software and Algorithms Used in
+  Publications \cite{Knepley_WSSSPE}
 
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
+  report for every software project \cite{Priem_WSSSPE}
 
-\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}
+\item Daniel S. Katz. Citation and Attribution of Digital Products:
+  Social and Technological Concerns \cite{Katz2_WSSSPE}
 
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
+  improving the reusability and sustainability of scientific software
+  \cite{Chue_Hong_WSSSPE}
 
@@
 
-In addition, the following paper from another area were also discussed in this area.
+In addition, the following paper from another area were also discussed
+in this area.
 
@@
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software
+  \cite{Loffler_WSSSPE}
 
@@
 
-\item Victoria Stodden, Sheila Miguez. Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research \cite{Stodden_WSSSPE}
+\item Victoria Stodden, Sheila Miguez. Best Practices for
+  Computational Science: Software Infrastructure and Environments for
+  Reproducible and Extensible Research \cite{Stodden_WSSSPE}
 
@@
 
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
+  Research Software Security Maturity Model \cite{Heiland_WSSSPE}
 
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software \cite{Blanton_WSSSPE}
 
-\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}
+\item Daisie Huang, Hilmar Lapp. Software Engineering as
+  Instrumentation for the Long Tail of Scientific Software
+  \cite{Huang_WSSSPE}
 
-\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}
+\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher. Developing Systems for API Governance
+  \cite{Wolski_WSSSPE}
 
@@
 
-How to ensure software is sustainable by ensuring there are career paths for developers.
+How to ensure software is sustainable by ensuring there are career
+paths for developers.
 
@@
 
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
+  Software through the Open Community Engagement Process
+  \cite{Christopherson_WSSSPE}
 
-\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
+\item Reagan Moore. Extensible Generic Data Management Software
+  \cite{Moore_WSSSPE}
 
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
+  grassroots approach to software sustainability
+  \cite{Cranston_WSSSPE}
 
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
+  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
+  Modeling Science Activities \cite{Vay_WSSSPE}
 
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
+  Borries Demeler. Science Gateway Operational Sustainability:
+  Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
 
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
+  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
+  Experiences and Challenges in Software Sustainability for a Large
+  Scientific Community \cite{Zentner_WSSSPE}
 
-\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+\item Andy Terrel. Sustaining the Python Scientific Software Community
+  \cite{Terrel_WSSSPE}
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software
+  \cite{Loffler_WSSSPE}
 
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
+  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
+  Mookherjee. Reusability in Science: From Initial User Engagement to
+  Dissemination of Results \cite{Maheshwari_WSSSPE}
 
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
+  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
+  Dooley, Dan Stanzione. Science Gateways and the Importance of
+  Sustainability \cite{Wilkins-Diehr_WSSSPE}
 
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
+  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
+  tools for facilitating Open Science \cite{Hart_WSSSPE}
 
@@
 
-In addition, the following paper from another area was also discussed in this area.
+In addition, the following paper from another area was also discussed
+in this area.
 
@@
 
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}
+\item Anne C. Elster. Software for Science: Some Personal Reflections
+  \cite{Elster_WSSSPE}
 
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
+  as a path to software sustainability \cite{Foster_WSSSPE}
 
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-In addition, the following papers from other areas were also discussed in this area.
+In addition, the following papers from other areas were also discussed
+in this area.
 
@@
 
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software \cite{Blanton_WSSSPE}
 
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development \cite{Blatt_WSSSPE}
 
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
+  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
+  Open Source Production Quality MPI Library for HPC
+  \cite{Panda_WSSSPE}
 
-\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+\item Andy Terrel. Sustaining the Python Scientific Software Community
+  \cite{Terrel_WSSSPE}
 
@@
 
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T.
-Brown.
-Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
+  T. Brown. Advanced Techniques for Scientific Programming and
+  Collaborative Development of Open Source Software Packages at the
+  International Centre for Theoretical Physics (ICTP)
+  \cite{Girotto_WSSSPE}
 
-\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}
+\item Thomas Crawford. On the Development of Sustainable Software for
+  Computational Chemistry \cite{Crawford_WSSSPE}
 
@@
 
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
 
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
 
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software \cite{Ahern_WSSSPE}
 
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software
+  \cite{Choi_WSSSPE}
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software
+  \cite{Loffler_WSSSPE}
 
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software \cite{Trainer_WSSSPE}
 
@@
 \subsection{Challenges of community}
-include metrics for community success (e.g., more external contributors than internal).
+include metrics for community success (e.g., more external
+contributors than internal).
 \subsection{Admirable scientific software communities}
@@
 \begin{itemize}
-\item Can sustainability be defined as the effort that happens to make the essential things continue?
-\item What is sustainability? What is it that we want to sustain? Is what we want to sustain valuable? Who cares if it goes away and how do you measure how much they care?
+\item Can sustainability be defined as the effort that happens to make
+  the essential things continue?
+\item What is sustainability? What is it that we want to sustain? Is
+  what we want to sustain valuable? Who cares if it goes away and how
+  do you measure how much they care?
 \end{itemize}
@@
 \begin{itemize}
-\item from http://oss-watch.ac.uk/resources/ssmm: When choosing software for procurement or development reuse - regardless of the license and development model you will use - you need to consider the future.
-While a software product may satisfy today's needs, will it satisfy tomorrow's needs?
-Will the supplier still be around in five years' time?
-Will the supplier still care for all its customers in five years' time?
-Will the supplier be responsive to bug reports and feature requests?
-In other words, is the software sustainable?
-\item I think this is a key question we as a community need to agree on (whats our problem statement?) and hope we come out today with some progress. - @VonWelch
-\item What is the goal of sustainability? Reproducible science, persistence, quality, or else?
-\item How do you measure success? How do you know when you have actually achieved sustainable software?
-\item Sustainability should be addressed throughout the life cycle (Daisie Huang)
+\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
+  software for procurement or development reuse - regardless of the
+  license and development model you will use - you need to consider
+  the future. While a software product may satisfy today's needs, will
+  it satisfy tomorrow's needs? Will the supplier still be around in
+  five years' time? Will the supplier still care for all its customers
+  in five years' time? Will the supplier be responsive to bug reports
+  and feature requests? In other words, is the software sustainable?
+\item I think this is a key question we as a community need to agree
+  on (whats our problem statement?) and hope we come out today with
+  some progress. - @VonWelch
+\item What is the goal of sustainability? Reproducible science,
+  persistence, quality, or else?
+\item How do you measure success? How do you know when you have
+  actually achieved sustainable software?
+\item Sustainability should be addressed throughout the life cycle
+  (Daisie Huang)
 \end{itemize}
@@
 \begin{itemize}
-\item Chris Lenhardt's presentation of Venters paper~\cite{Venters_WSSSPE} included:
+\item Chris Lenhardt's presentation of Venters
+  paper~\cite{Venters_WSSSPE} included:
 
@@
 It ``is a rather ambiguous concept...''
-\item Lack of an accepted definition gets in the way of integrating the concept into software engineering
+\item Lack of an accepted definition gets in the way of integrating
+  the concept into software engineering
 \item Sustainability as a non-functional requirement
 \item Quality of software architectures determines sustainability
-\item Propose sustainability is a measure of central quality attributes:
-Extensibility,
-Interoperability,
-Maintainability,
-Portability,
-Reusability,
-Scalability
-\item Develop an architecture evaluation framework based on scenarios which help to illuminate how to measure quality/sustainability at the architectural level
+\item Propose sustainability is a measure of central quality
+  attributes: Extensibility, Interoperability, Maintainability,
+  Portability, Reusability, Scalability
+\item Develop an architecture evaluation framework based on scenarios
+  which help to illuminate how to measure quality/sustainability at
+  the architectural level
 \end{itemize}
 
-\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE} included:
+\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
+  included:
 \begin{itemize}
@@
 \item Is functionality and usability clear
-\item Will functionality be correct going into the future even if environment changes
+\item Will functionality be correct going into the future even if
+  environment changes
 \item Is the functionality desired by current and future users
@@
 
-\item see slide 9 of \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
+\item see slide 9 of
+  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
 
@@
 
-Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+Perhaps discuss sustained software use-cases and issues surrounding
+them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
 
-\note{Dan: do we need this section?
-I think there's some overlap between this and the discussion in the Developing and Support Software section.
-Or perhaps this should go in the cross-cutting issues section?}
+\note{Dan: do we need this section?  I think there's some overlap
+  between this and the discussion in the Developing and Support
+  Software section.  Or perhaps this should go in the cross-cutting
+  issues section?}
 
@@
 
-\todo{add some analysis of the deficiencies and difficulties that are present in different fields, and those that are common?}
+\todo{add some analysis of the deficiencies and difficulties that are
+  present in different fields, and those that are common?}
 
@@
 
-\note{or at least the discussion about them, and the current plans for future events.}
+\note{or at least the discussion about them, and the current plans for
+  future events.}
 
@@
 
-The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop 
-agenda. However, contributors also want to get credit for their participation in the process. And the 
-workshop organizers want to make sure that the workshop content and their efforts are recorded. 
-Ideally, there would be a service that would be able to index the contributions to the workshop, 
-serving the authors, the organizers, and the larger community. But since there isn't such a service 
-today, the workshop organizers are writing this initial report and making use of arXiv as a partial 
-solution to provide a record of the workshop.
+The WSSSPE workshop has begun an experiment in how we can
+collaboratively build a workshop agenda. However, contributors also
+want to get credit for their participation in the process. And the
+workshop organizers want to make sure that the workshop content and
+their efforts are recorded.  Ideally, there would be a service that
+would be able to index the contributions to the workshop, serving the
+authors, the organizers, and the larger community. But since there
+isn't such a service today, the workshop organizers are writing this
+initial report and making use of arXiv as a partial solution to
+provide a record of the workshop.
 
-After the workshop, one or more additional papers will be created that will include the discussions at 
-the workshop. These papers will likely have many authors, and may be submitted to peer-reviewed 
-journals.
+After the workshop, one or more additional papers will be created that
+will include the discussions at the workshop. These papers will likely
+have many authors, and may be submitted to peer-reviewed journals.
 
@@
 
-The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:
+The following is a partial list of attendees who were recorded on the
+Google doc that was being used for live note taking at the workshop:
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Refined Section 1. Intro

@@
 are expected.  The level of interest in the workshop has led the
-organizers, working with some of the submitters and some of the
-attendees, to plan two additional workshops (at the 2014 SciPy and SC
-conferences) and to turn the workshop website into a community website
-that can be used as a focus for further discussion and progress.
+organizers, working with some of the submitters and attendees, to plan
+two additional workshops (at the 2014 SciPy and SC conferences) and to
+turn the workshop website into a community website that can be used as
+a focus for further discussion and progress.  A minisymposium on
+``Reliable Computational Science'' that aims to further explore some
+of the key issues raised in WSSSPE is co-organized by a participant in
+the 2014 SIAM Annual Meeting.
 
@@
 (\S\ref{sec:conclusions}).
-
-\todo{Mention a two-session minisymposium on ``Reliable Computational
-  Science'' in the SIAM Annual Meeting in July. It aims to further
-  explore some of the key issues raised in WSSSPE. Speakers include
-  Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
 

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Small changes to Section 2. Agenda

@@
 \usepackage{wrapfig}
+\usepackage{enumitem}
 
@@
 
-The original document is \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#} (but can no longer be edited).
-Note that the original document has comments in addition to text.
+The original document is
+\url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#}
+(but can no longer be edited).  Note that the original document has
+comments in addition to text.
 
@@
 
-\begin{itemize}
+\begin{itemize}[leftmargin=0.2in]
 \item the development process that leads to new (versions of) software
-\begin{itemize}
+\begin{itemize}[leftmargin=0.2in]
 \item how fundamental research in computer science or
@@
   software engineering
-\begin{itemize}
+\begin{itemize}[leftmargin=0.2in]
 \item governance, business, and sustainability models
@@
 \item use of the software
-\begin{itemize}
+\begin{itemize}[leftmargin=0.2in]
 \item growing communities
@@
 \item policy issues, such as
-\begin{itemize}
+\begin{itemize}[leftmargin=0.2in]
 \item measuring usage and impact
@@
 \item career paths for developers and institutional roles
-\item issues related to multiple organizations and multiple countries,
+\item issues related to multiple organizations and countries,
   such as intellectual property, licensing, etc.
@@
 arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
-their papers.
-
-A peer review process followed the submissions, where the 58 papers
-received 181 reviews, an average of 3.12 reviews per paper. Reviews
-were completed using a Google form, which allowed reviewers to choose
-papers they wanted to review, then to provide scores on their
-relevance and general comments to the organizers and to the
-authors. The review reports enabled the organizers to decide which
-papers to associate with the workshop, and allowed the authors to
-improve their papers.
-
-The organizers decided to list 54 of the papers as significantly
-contributing to the workshop, a very high acceptance rate, but one
+their papers.  A peer review process followed the submissions, where
+the 58 papers received 181 reviews, an average of 3.12 reviews per
+paper. Reviews were completed using a Google form, which allowed
+reviewers to choose papers they wanted to review, then to provide
+scores on their relevance and general comments to the organizers and
+to the authors. The review reports enabled the organizers to decide
+which papers to associate with the workshop, and allowed the authors
+to improve their papers.
+
+The organizers decided to list 54 of the papers as significant
+contributions to the workshop, a very high acceptance rate, but one
 that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. The papers were also
+that the reports were already self-published. The papers were
 grouped into three main categories, namely \emph{Developing and

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Changes made to improve the flow in Section 3.1 Keynote of Bourne.

@@
 keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
-was delivered by Philip E. Bourne of UCSD. The slides are
+was delivered by Philip E. Bourne of UCSD. The slides are available
 online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
-
 Philip is a basic biomedical scientist but has formed four software
 companies. He also helped found PLOS and
-PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
-
-He emphasized that Sustainability for software ``does not just mean
-more money from Government.''
-
-He noted that creating effective illustrations is a key function of
-software in the biomedical sciences: ``Automating Conceptualization''
-(Visualizing molecular structures is a problem that has now been
-solved, he is working on 3-d visualizations of the contents of cells.)
-
-He has looked into reproducibility in computational biology, and has
-concluded that ``I have proved I cannot reproduce research from my own
-lab''~\cite{Veretnik}.
-
-He noted that in places, development in science has improved thanks to
-open source and Github, but for the most part remains arcane. He
-argues that there is a lot we can learn from the app-store model. He
-mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}}
-and
+PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.  He is
+working on automating three-dimensional visualizations of cell
+contents and molecular structures, a problem that has not been solved
+and would serve as a key function of software in biomedical sciences.
+
+Philip emphasized that Sustainability for software ``does not just
+mean more money from Government.''  He noted that in places,
+development in science has improved thanks to open source and Github,
+but for the most part remains arcane. He argued that there is much we
+can learn from the app-store model. He mentioned
+BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
 OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
-as distribution channels.
-
-He noted a common evolutionary pathway for computational biology
-projects, from data archive to analytics platform to educational
-use. He suggests that use for outreach might be the final step.
-
-He experienced problems with staff retention with respect to private
-organizations which reward those combining research and software
-expertise (the ``Google Bus''). However, he is a strong fan of private
-sector sustainability sources and public-private partnerships. He
-noted that making a successful business from scientific software alone
-is very rare: founders overvalue while customers undervalue. He noted
-that to last, an open source project needs a minimal funding
-requirement even with a vibrant community -- good will only goes so
-far if you are being paid to do something else.
-
+as distribution channels.  On a related note, Philip observed a common
+evolutionary pathway for computational biology projects, from data
+archive to analytics platform to educational use, and suggested that
+use of scientific software for outreach might be the final step.
+
+Philip shared with the audience a few real challenges he
+encountered. First, he has looked into reproducibility in
+computational biology, and has concluded that ``I have proved I cannot
+reproduce research from my own lab''~\cite{Veretnik}.
+
+Another problem he experienced was staff retention with respect to
+private organizations which reward those combining research and
+software expertise (the ``Google Bus''). However, he is a strong fan
+of private sector sustainability sources and public-private
+partnerships. He noted that making a successful business from
+scientific software alone is very rare: founders overvalue while
+customers undervalue. He noted that to last, an open source project
+needs a minimal funding requirement even with a vibrant community --
+good will only goes so far if you are being paid to do something else.
 He talked about grant schemes of relevance in the U.S., particularly
@@
 
-He had problems with selling research software: the technology
+He also had problems with selling research software: the technology
 transfer office in his university wanted huge IP reachthrough, whereby
@@
 
-He then expanded on the idea of the \textbf{digital enterprise}, where
-IT underpins the whole of organizational activities. He contends that
+On the idea of the \textbf{digital enterprise}, where
+IT underpins the whole of organizational activities, he contended that
 universities are way behind the curve on this. In particular, he
-highlights the separation of research, teaching and administration into
+highlighted the separation of research, teaching and administration into
 silos without a common IT framework as a blocker to many useful

### 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
Added a TODO note.

@@
 (\S\ref{sec:conclusions}).
+
+\todo{Mention somewhere statistics of participants: There were 140 seats, I (Katz) think 149 people at the first keynote, and about 90 at the end of the workshop.
+They (SC13) counted people in the room 4 times during the day.
+But I don’t know how many unique people attended, which I agree they should know.}
 

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

--- 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 \usepackage{wrapfig}
-\usepackage{enumitem}
 
@@
 \newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
-
-\usepackage{array}
-\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
-\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
 
@@
 
-\note{papers and presentations and URLs should be references - see
-  figshare-web in .bib file for URL style example}
+\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}
 
-People who have volunteered to work on this report: (with * for actual
-contributors - \note{if you contribute, add a * by your name})
+People who have volunteered to work on this report: (with * for actual contributors - \note{if you contribute, add a * by your name})
 
@@
 
-The original document is
-\url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#}
-(but can no longer be edited).  Note that the original document has
-comments in addition to text.
+The original document is \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#} (but can no longer be edited).
+Note that the original document has comments in addition to text.
 
@@
 
-The First Workshop on Sustainable Software for Science: Practice and
-Experiences (WSSSPE1,
-\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
-Sunday, 17 November 2013, in conjunction with the 2013 International
-Conference for High Performance Computing, Networking, Storage and
-Analysis (SC13, \url{http://sc13.supercomputing.org}).
+The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
 
-Because progress in scientific research is dependent on the quality
-and accessibility of software at all levels, it is now critical to
-address many challenges related to the development, deployment, and
-maintenance of reusable software. In addition, it is essential that
-scientists, researchers, and students are able to learn and adopt
-software-related skills and methodologies. Established researchers are
-already acquiring some of these skills, and in particular a
-specialized class of software developers is emerging in academic
-environments who are an integral and embedded part of successful
-research teams. The WSSSPE workshop was intended to provide a forum
-for discussion of the challenges, including both positions and
-experiences.  The short papers and discussion have been archived to
-provide a basis for continued discussion, and the workshop has fed
-into the collaborative writing of this journal publication.
-Additional papers based on extended versions of workshop submissions
-are expected.  The level of interest in the workshop has led the
-organizers, working with some of the submitters and attendees, to plan
-two additional workshops (at the 2014 SciPy and SC conferences) and to
-turn the workshop website into a community website that can be used as
-a focus for further discussion and progress.  A minisymposium on
-``Reliable Computational Science'' that aims to further explore some
-of the key issues raised in WSSSPE is co-organized by a participant in
-the 2014 SIAM Annual Meeting.
+Because progress in scientific research is dependent on the quality and accessibility
+of software at all levels, it is now critical to address many challenges related to the
+development, deployment, and maintenance of reusable software. In addition, it is
+essential that scientists, researchers, and students are able to learn and adopt
+software-related skills and methodologies. Established researchers are already
+acquiring some of these skills, and in particular a specialized class of software
+developers is emerging in academic environments who are an integral and embedded
+part of successful research teams. The WSSSPE workshop was intended to provide a
+forum for discussion of the challenges, including both positions and experiences.
+The short papers and discussion have been archived to provide a basis for continued
+discussion, and the workshop has fed into the collaborative writing of this journal publication.
+Additional papers based on extended versions of workshop submissions are expected.
+The level of interest in the workshop has led the organizers, working with some of the submitters
+and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
+the workshop website into a community website that can be used as a focus for further
+discussion and progress.
 
-Before the WSSSPE1 workshop took place, the organizers self-published
-a paper~\cite{WSSSPE1-pre-report} to document the process of
-organizing and advertising the workshop, collecting and reviewing the
-papers, and putting together the agenda.  Section~\ref{sec:process} is
-based on that paper.  The remainder of this paper is based on the
-workshop itself, as documented in a set of collaborative notes taken
-during the workshop~\cite{WSSSPE1-google-notes}, including discussion
-about the keynote presentations (\S\ref{sec:keynotes}), the three
-panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
-(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
-use cases (\S\ref{sec:use-cases}), and conclusions
-(\S\ref{sec:conclusions}).
+\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July.
+It aims to further explore some of the key issues raised in WSSSPE.
+Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
 
-\section{Workshop Process and Agenda} \label{sec:process}
+\todo{mention arXiv:1311.3523 and explain what's different here}
 
-WSSSPE1 was organized in collaboration of the relatively small group
-of five organizers and a larger peer-review committee with 36
-members. This committee had early influence on the organization, e.g.,
-on the specific call for papers.
+\section{Workshop Process}
+
+WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
+a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
+e.g., on the specific call for papers.
 
@@
 \begin{quote}
-In practice, scientific software activities are part of an ecosystem
-where key roles are held by developers, users, and funders. All three
-groups supply resources to the ecosystem, as well as requirements that
-bound it. Roughly following the example of NSF's Vision and Strategy
-for Software
+In practice, scientific software activities are part of an ecosystem where key roles are held by 
+developers, users, and funders. All three groups supply resources to the ecosystem, as well as 
+requirements that bound it. Roughly following the example of NSF's Vision and Strategy for Software 
 (\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision}, 
@@
 
-\begin{itemize}[leftmargin=0.2in]
+\begin{itemize}
 \item the development process that leads to new (versions of) software
-\begin{itemize}[leftmargin=0.2in]
-\item how fundamental research in computer science or
-  science/engineering domains is turned into reusable software
+\begin{itemize}
+\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
 \item software created as a by-product of research
-\item impact of computer science research on the development of
-  scientific software
+\item impact of computer science research on the development of scientific software
 \end{itemize}
-\item the support and maintenance of existing software, including
-  software engineering
-\begin{itemize}[leftmargin=0.2in]
+\item the support and maintenance of existing software, including software engineering
+\begin{itemize}
 \item governance, business, and sustainability models
-\item the role of community software repositories, their operation and
-  sustainability
+\item the role of community software repositories, their operation and sustainability
 \end{itemize}
@@
 \item use of the software
-\begin{itemize}[leftmargin=0.2in]
+\begin{itemize}
 \item growing communities
@@
 \item policy issues, such as
-\begin{itemize}[leftmargin=0.2in]
+\begin{itemize}
 \item measuring usage and impact
@@
 \item career paths for developers and institutional roles
-\item issues related to multiple organizations and countries,
-  such as intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of
-  publishers
+\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
+\item mechanisms and venues for publishing software, and the role of publishers
 \end{itemize}
@@
 
-Based on the goal of encouraging a wide range of submissions from
-those involved in software practice, ranging from initial thoughts and
-partial studies to mature deployments, the organizers wanted to make
-submission as easy as possible. The call for papers stated:
+Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, the organizers wanted to make submission as easy as possible.
+The call for papers stated:
 
 \begin{quote}
-We invite short (4-page) position/experience reports that will be used
-to organize panel and discussion sessions. These papers will be
-archived by a third-party service, and provided DOIs. We encourage
-submitters to license their papers under a Creative Commons license
-that encourages sharing and remixing, as we will combine ideas (with
-attribution) into the outcomes of the workshop.  An interactive site
-will be created to link these papers and the workshop discussion, with
-options for later comments and contributions. Contributions will be
-peer-reviewed for relevance and originality before the links are added
-to the workshop site; contributions will also be used to determine
-discussion topics and panelists. We will also plan one or more papers
-to be collaboratively developed by the contributors, based on the
-panels and discussions.
+We invite short (4-page) position/experience reports that will be used to organize panel and 
+discussion sessions. These papers will be archived by a third-party service, and provided DOIs. We 
+encourage submitters to license their papers under a Creative Commons license that encourages 
+sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop. 
+An interactive site will be created to link these papers and the workshop discussion, with options for 
+later comments and contributions. Contributions will be peer-reviewed for relevance and originality 
+before the links are added to the workshop site; contributions will also be used to determine 
+discussion topics and panelists. We will also plan one or more papers to be collaboratively 
+developed by the contributors, based on the panels and discussions.
 \end{quote}
@@
 58 submissions were received, and almost all submitters used either 
-arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
-their papers.  A peer review process followed the submissions, where
-the 58 papers received 181 reviews, an average of 3.12 reviews per
-paper. Reviews were completed using a Google form, which allowed
-reviewers to choose papers they wanted to review, then to provide
-scores on their relevance and general comments to the organizers and
-to the authors. The review reports enabled the organizers to decide
-which papers to associate with the workshop, and allowed the authors
-to improve their papers.
+arXiv\footnote{\url{http://arxiv.org}} or 
+figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
 
-The organizers decided to list 54 of the papers as significant
-contributions to the workshop, a very high acceptance rate, but one
-that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. The papers were
-grouped into three main categories, namely \emph{Developing and
-  Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-subject area was associated with a panel and discussion at the
-workshop.
+A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
+average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
+reviewers to provide scores on relevance and comments to the organizers and to the authors.
+The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.
 
-The workshop itself consisted of two keynote presentations and the
-three panels/discussion sessions. The panels were organized based on a
-classification of the workshop submissions into three categories,
-following the themes of the call for papers as modified by the areas
-of the submissions. Three to four representatives from each submission
-category were appointed as panelists, and assigned to read a subset of
-the paper in that category and then discuss them in the panel.
+The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
+high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
+that the reports were already self-published.
+The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}.
+Each subject area was associated with a panel and discussion at the workshop.
 
-\section{Keynotes} \label{sec:keynotes}
+\section{Agenda}
 
-The WSSSPE1 workshop began with two keynote presentations.
+The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
+were organized based on a classification of the workshop submissions into three categories, following 
+the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
+each submission categories were appointed as panelists, to discuss the papers in that category.
+
+\section{Keynotes}
 
@@
 
-\todo{change references to papers or presentations to cites}
-
 \subsection{A recipe for sustainable software, Philip E. Bourne.}
 
-The first
-keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
-was delivered by Philip E. Bourne of UCSD. The slides are
-online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delivered by Philip E. Bourne of UCSD. The slides
+are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
 
 Philip is a basic biomedical scientist but has formed four software
-companies. He also helped found PLOS and
-PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
 
-He emphasized that Sustainability for software ``does not just mean
-more money from Government.''
+He emphasized that Sustainability for software ``does not just mean more
+money from Government.''
 
@@
 software in the biomedical sciences: ``Automating Conceptualization''
-(Visualizing molecular structures is a problem that has now been
-solved, he is working on 3-d visualizations of the contents of cells.)
+(Visualizing molecular structures is a problem that has now been solved,
+he is working on 3-d visualizations of the contents of cells.)
 
@@
 He noted that in places, development in science has improved thanks to
-open source and Github, but for the most part remains arcane. He
-argues that there is a lot we can learn from the app-store model. He
-mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}}
-and
-OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
-as distribution channels.
+open source and Github, but for the most part remains arcane. He argues
+that there is a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
+OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
+distribution channels.
 
 He noted a common evolutionary pathway for computational biology
-projects, from data archive to analytics platform to educational
-use. He suggests that use for outreach might be the final step.
+projects, from data archive to analytics platform to educational use. He
+suggests that use for outreach might be the final step.
 
-He experienced problems with staff retention with respect to private
-organizations which reward those combining research and software
-expertise (the ``Google Bus''). However, he is a strong fan of private
-sector sustainability sources and public-private partnerships. He
+He experienced problems with staff retention with respect to private organizations which reward 
+those combining research and software expertise (the ``Google Bus''). However,
+he is a strong fan of private sector sustainability sources and PPP. He
 noted that making a successful business from scientific software alone
 is very rare: founders overvalue while customers undervalue. He noted
-that to last, an open source project needs a minimal funding
-requirement even with a vibrant community -- good will only goes so
-far if you are being paid to do something else.
+that to last, an open source project needs a minimal funding requirement
+even with a vibrant community -- good will only goes so far if you are
+being paid to do something else.
 
-He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology
-transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
+He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}
 
-He had problems with selling research software: the technology
-transfer office in his university wanted huge IP reachthrough, whereby
-they would get a share of profits from drugs developed by pharma
-companies who use the software. He was aware this was unrealistic but
-the technology transfer office insisted for a while. He wants to push
-a one-click approach to purchasing university-written software.
+He had problems with selling research software: the technology transfer
+office in his university wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the
+software. He was aware this was unrealistic but the technology transfer
+office insisted for a
+while. He wants to push a one-click approach to purchasing
+university-written software.
 
@@
 
-He concluded with a reference to his paper~\cite{bourne_ten} and
-argued that computational scientists ``have a responsibility to
-convince their institutions and communities that software is
-scholarship,'' a point with which we strongly agree.
+He concluded with a reference to his paper~\cite{bourne_ten} and argued that computational scientists ``have a
+responsibility to convince their institutions and communities that
+software is scholarship,'' a point with which we strongly agree.
 
-\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}
+\subsection{Scientific Software and the Open Collaborative Web}
+
+\todo{change references to papers or presentations to cites}
 
@@
 His talk used an example from data reduction in Astronomy, where he
-needed to remove interfering effects from the device. He built a ``bad
-pixel mask'', and realized that while this is persistent, there was no
-way/practice of sharing this with the data, so many researchers
-repeated the same calculations. He estimated that 13 person-years were
-wasted in this redundant calculation.
+needed to remove interfering effects from the device. He built a
+``bad pixel mask'', and realized though while this is persistent, there
+was no way/practice of sharing this with the data, so many researchers
+repeated the same calculations. He estimated that 13 wasted person-years were wasted in
+this redundant calculation.
 
 He asked, of this practice, ``why didn't we know better?'' and argued
-this was because we were taught to focus on immediate research
-outcomes and not on continuously improving and building on tools for
-research. He then asked, later in our working lives, when we do know
-better, why don't we act any different, and argued this was due to
-incentives and their lack: only the immediate products of research,
-not the software, are valued.  He referenced Victoria Stodden's talk
-at
-OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+this was because we were taught to focus on immediate research outcomes
+and not on continuously improving and building on tools for research. He
+then asked, later in our working lives, when we do know better, why
+don't we act any different, and argued this was due to incentives and their lack:
+only the immediate products of research, not the software, are valued.
+He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
 which he said argued these points well.
 
-He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
-Chain Monte Carlo code by Dan Forman
-Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
-Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
-as good examples of scientific projects using GitHub.
+He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
+Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} and Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}} as 
+good examples of  scientific projects using GitHub.
 
-C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
-regard to reusable software, ``we should just start doing it.'' Arfon
-noted that documentation should be ``treated as a first class
-entity.''  He noted that some software fields have excellent cultures
-of module reuse, such as
-rubygems\footnote{\url{http://rubygems.org/}},
-PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
-CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
+reusable software, ``we should just start doing it.'' Arfon noted that
+documentation should be ``treated as a first class entity.''
+He noted that some software fields have excellent cultures of module
+reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
+PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
 ultra-low-friction collaboration through these repositories. This has
-not happened in highly numerical, compiled language scientific
-software.
+not happened in highly numerical, compiled language scientific software.
 
@@
 
-Fernando Perez says that ``Open Source is reproducible by necessity''
-- since reproducibility is a prerequisite for remote collaboration.
-Arfon therefore defined GitHub as part of the next stage in
-development of the web, following on from the ``social web'' of
-Facebook et.al.: ``The Collaborative Web.''
+Fernando Perez says that ``Open Source
+is reproducible by necessity'' - since reproducibility is a prerequisite
+for remote collaboration.
+Arfon therefore defined GitHub as part of the next stage in development
+of the web, following on from the ``social web'' of Facebook et.al.:
+``The Collaborative Web.''
 
-\section{Developing and Supporting Software} \label{sec:devel}
+\section{Developing and Supporting Software}
 
@@
 
-\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
-  McInnes, Barry Smith. Package Management Practices Essential for
-  Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath \cite{Miller_WSSSPE}
+\item Mark C.
+Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith.
+Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}
 
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable
-  \cite{Broman_WSSSPE}
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}
 
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
 
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?
-  \cite{Stephan_WSSSPE}
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}
 
-\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
-  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development
-  \cite{Gaston_WSSSPE}
+\item Derek R.
+Gaston, John Peterson, Cody J.
+Permann, David Andrs, Andrew E.
+Slaughter, Jason M.
+Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}
 
-\item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks
-  \cite{Dubey_WSSSPE}
+\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}
 
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
 
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
 
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
 
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
 
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
-  adapting sustainable scientific software development processes in a
-  small biology lab \cite{Crusoe_WSSSPE}
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}
 
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
-The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
 
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
-  H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes \cite{Heien_WSSSPE}
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}
 
@@
 
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
-  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
-  Versatile Simulation of Large Scale Distributed
-  Systems~\cite{Casanova_WSSSPE}
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter.
+SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}
 
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software \cite{Trainer_WSSSPE}
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
 
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
-  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
-  Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software \cite{Cohen_WSSSPE}
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin.
+Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}
 
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications
-  \cite{Slawinski_WSSSPE}
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}
 
@@
 
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software \cite{Prlic_WSSSPE}
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}
 
-\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
-  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
-  R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
+\item Anshu Dubey, S.
+Brandt, R.
+Brower, M.
+Giles, P.
+Hovland, D.
+Q.
+Lamb, F.
+Löffler, B.
+Norris, B.
+O'Shea, C.
+Rebbi, M.
+Snir, R.
+Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
 
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult \cite{Carver_WSSSPE}
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}
 
-\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
-  K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability
-  \cite{Stewart_WSSSPE}
+\item Craig A.
+Stewart, Julie Wernert, Eric A.
+Wernert, William K.
+Barnett, Von Welch.
+Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}
 
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable \cite{Brown_WSSSPE}
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}
 
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
-  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
-  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms \cite{Swenson_WSSSPE}
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc.
+Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}
 
@@
 
-Discussion of software produced as a by-product versus software
-developed for reuse. How does this change the project, who develops
-the code, growth beyond 1--2 person projects to larger projects with
-diverse set of consumers and contributors. Modular and extensible code
-versus working for current research problem---tackled by same people,
-or a collaborative team?
+Discussion of software produced as a by-product versus software developed for reuse.
+How does this change the project, who develops the code, growth beyond 1--2 person projects to larger projects with diverse set of consumers and contributors.
+Modular and extensible code versus working for current research problem---tackled by same people, or a collaborative team?
 
@@
 
-What is sustainable? What are the best practices, can workshops fund
-meetings to help define and improve best practices in these areas?
-Software plans from the funding agencies? This topic is expanded upon
-in \S\ref{sec:cross-cutting}, since it was discussed up in multiple
-parts of the WSSSPE1 workshop.
+What is sustainable?
+What are the best practices, can workshops fund meetings to help define and improve best practices in these areas?
+Software plans from the funding agencies?
+\todo{mention as cross-cutting, since this came up in other parts of the workshop as well?}
 
@@
 
-Discussion of the evolving role of scientists, and/or others that fill
-these roles.
+Discussion of the evolving role of scientists, and/or others that fill these roles.
 
@@
 
-There are a lot of tools out there, but few are currently used. Look
-at what some projects have found successful, and how to automate as
-much as possible to reduce additional overhead.
+There are a lot of tools out there, but few are currently used. Look at what some projects have found successful, and how to automate as much as possible to reduce additional overhead.
 
@@
 
-Its impact on how and where research products are used, who can
-collaborate and what patterns have worked/not worked in existing
-communities.
+Its impact on how and where research products are used, who can collaborate and what patterns have worked/not worked in existing communities.
 
@@
 
-How initial software development is funded, moving to follow up
-projects, maintenance, community growth, collaborating with other
-institutions, labs, industry, internationally. Contract relationship
-versus collaborative development between scientists and software
-developers.
+How initial software development is funded, moving to follow up projects, maintenance, community growth, collaborating with other institutions, labs, industry, internationally.
+Contract relationship versus collaborative development between scientists and software developers.
 
@@
 
-Who creates training materials, how are they distributed, integration
-into courses when projects see very wide application in research.
+Who creates training materials, how are they distributed, integration into courses when projects see very wide application in research.
 
-\section{Policy} \label{sec:policy}
+\section{Policy}
 
@@
 
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model \cite{Calero_WSSSPE}
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}
 
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework
-  \cite{Venters_WSSSPE}
+\item Colin C.
+Venters, Lydia Lau, Michael K.
+Griffiths, Violeta Holmes, Rupert R.
+Ward, Jie Xu.
+The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}
 
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance
-  \cite{Pierce_WSSSPE}
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}
 
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
+\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
 
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science
-  \cite{Lenhardt_WSSSPE}
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak.
+Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}
 
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science
-  \cite{Weber_WSSSPE}
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}
 
@@
 
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
-  Smith. Accurately Citing Software and Algorithms Used in
-  Publications \cite{Knepley_WSSSPE}
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}
 
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project \cite{Priem_WSSSPE}
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}
 
-\item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns \cite{Katz2_WSSSPE}
+\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}
 
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software
-  \cite{Chue_Hong_WSSSPE}
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}
 
@@
 
-In addition, the following paper from another area were also discussed
-in this area.
+In addition, the following paper from another area were also discussed in this area.
 
@@
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
 
@@
 
-\item Victoria Stodden, Sheila Miguez. Best Practices for
-  Computational Science: Software Infrastructure and Environments for
-  Reproducible and Extensible Research \cite{Stodden_WSSSPE}
+\item Victoria Stodden, Sheila Miguez. Best Practices for Computational Science: Software Infrastructure and Environments for Reproducible and Extensible Research \cite{Stodden_WSSSPE}
 
@@
 
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model \cite{Heiland_WSSSPE}
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}
 
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
 
-\item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software
-  \cite{Huang_WSSSPE}
+\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}
 
-\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher. Developing Systems for API Governance
-  \cite{Wolski_WSSSPE}
+\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}
 
@@
 
-How to ensure software is sustainable by ensuring there are career
-paths for developers.
+How to ensure software is sustainable by ensuring there are career paths for developers.
 
-\section{Communities} \label{sec:community}
+\section{Communities}
 
@@
 
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
-  Software through the Open Community Engagement Process
-  \cite{Christopherson_WSSSPE}
-
-\item Reagan Moore. Extensible Generic Data Management Software
-  \cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability
-  \cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities \cite{Vay_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
-  Borries Demeler. Science Gateway Operational Sustainability:
-  Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community \cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
-
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
-  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
-  Mookherjee. Reusability in Science: From Initial User Engagement to
-  Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
-  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
-  Dooley, Dan Stanzione. Science Gateways and the Importance of
-  Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
-  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
-  tools for facilitating Open Science \cite{Hart_WSSSPE}
+\item Step 1 - Design - thorough discussion of research questions
+\item Step 2 - Develop working code
+\item Step 3 - Refine based on new requirements
+\item Step 4 - Publish open source
 
@@
 
-In addition, the following paper from another area was also discussed
-in this area.
+The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
+code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
+build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
+Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
+In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
+Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
+Co‐locate rather than collaborating remotely whenever feasible.
+
+The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure and they are much
+more than a fertile ground for grid computing research. The Pierce paper summarizes the explosion of use of just a few gateways 
+over the last 4 years -
+
+\subsection{Papers}
+
+\subsubsection*{Communities Part 1 Papers}
 
@@
 
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-\item Anne C. Elster. Software for Science: Some Personal Reflections
-  \cite{Elster_WSSSPE}
+\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}
 
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability \cite{Foster_WSSSPE}
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}
 
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-In addition, the following papers from other areas were also discussed
-in this area.
+In addition, the following papers from other areas were also discussed in this area.
 
@@
 
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
 
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
 
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
-  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC
-  \cite{Panda_WSSSPE}
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
 
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
 
@@
 
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
-  T. Brown. Advanced Techniques for Scientific Programming and
-  Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)
-  \cite{Girotto_WSSSPE}
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T.
+Brown.
+Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}
 
-\item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry \cite{Crawford_WSSSPE}
+\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}
 
@@
 
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
 
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
 
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
 
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
 
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software \cite{Trainer_WSSSPE}
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
 
@@
 \subsection{Challenges of community}
-include metrics for community success (e.g., more external
-contributors than internal).
+include metrics for community success (e.g., more external contributors than internal).
 \subsection{Admirable scientific software communities}
@@
 
-\section{Other Discussion} \label{sec:other}
+\section{Other Discussion}
 
@@
 
-\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
+\section{Cross-cutting Issues}
 
-\begin{itemize}
+\subsection{Defining sustainability}
 
-\item From discussion in Phil Bourne's keynote:
-\begin{itemize}
-\item Can sustainability be defined as the effort that happens to make
-  the essential things continue?
-\item What is sustainability? What is it that we want to sustain? Is
-  what we want to sustain valuable? Who cares if it goes away and how
-  do you measure how much they care?
-\end{itemize}
+\subsection{Licensing}
 
-\item from discussion in developing panel
+\note{not sure if this was really cross-cutting...}
 
-\begin{itemize}
-\item what does sustainable mean?
-\begin{itemize}
-\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
-  software for procurement or development reuse - regardless of the
-  license and development model you will use - you need to consider
-  the future. While a software product may satisfy today's needs, will
-  it satisfy tomorrow's needs? Will the supplier still be around in
-  five years' time? Will the supplier still care for all its customers
-  in five years' time? Will the supplier be responsive to bug reports
-  and feature requests? In other words, is the software sustainable?
-\item I think this is a key question we as a community need to agree
-  on (whats our problem statement?) and hope we come out today with
-  some progress. - @VonWelch
-\item What is the goal of sustainability? Reproducible science,
-  persistence, quality, or else?
-\item How do you measure success? How do you know when you have
-  actually achieved sustainable software?
-\item Sustainability should be addressed throughout the life cycle
-  (Daisie Huang)
-\end{itemize}
+\section{Illustrative use-cases}
 
-\item What is the relationship between sustainability and provenance?
+Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
 
-\item Is usability separate from sustainability or a fundamental part?
-\end{itemize}
-
-\item from discussion in policy panel
-
-\begin{itemize}
-\item Chris Lenhardt's presentation of Venters
-  paper~\cite{Venters_WSSSPE} included:
-
-\begin{itemize}
-\item What does software sustainability mean
-It ``is a rather ambiguous concept...''
-\item Lack of an accepted definition gets in the way of integrating
-  the concept into software engineering
-\item Sustainability as a non-functional requirement
-\item Quality of software architectures determines sustainability
-\item Propose sustainability is a measure of central quality
-  attributes: Extensibility, Interoperability, Maintainability,
-  Portability, Reusability, Scalability
-\item Develop an architecture evaluation framework based on scenarios
-  which help to illuminate how to measure quality/sustainability at
-  the architectural level
-\end{itemize}
-
-\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
-  included:
-\begin{itemize}
-\item Clarifying the notion of sustainability
-\begin{itemize}
-\item Provide same functionality going forward even if environment changes
-\item Is functionality and usability clear
-\item Will functionality be correct going into the future even if
-  environment changes
-\item Is the functionality desired by current and future users
-\item Can it incorporate new things as they develop
-\end{itemize}
-\end{itemize}
-
-\item see slide 9 of
-  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
-
-\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
-
-\end{itemize}
-
-\end{itemize}
-
-\section{Illustrative Use Cases} \label{sec:use-cases}
-
-Perhaps discuss sustained software use-cases and issues surrounding
-them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
-
-\note{Dan: do we need this section?  I think there's some overlap
-  between this and the discussion in the Developing and Support
-  Software section.  Or perhaps this should go in the cross-cutting
-  issues section?}
-
-\section{Conclusions} \label{sec:conclusions}
+\section{Conclusions}
 
@@
 
-\todo{add some analysis of the deficiencies and difficulties that are
-  present in different fields, and those that are common?}
-
-\todo{say something about licensing - lessons, advice, etc.?}
+\todo{add some analysis of the deficiencies and difficulties that are present in different fields, and those that are common?}
 
@@
 
-\note{or at least the discussion about them, and the current plans for
-  future events.}
+\note{or at least the discussion about them, and the current plans for future events.}
 
@@
 
-The WSSSPE workshop has begun an experiment in how we can
-collaboratively build a workshop agenda. However, contributors also
-want to get credit for their participation in the process. And the
-workshop organizers want to make sure that the workshop content and
-their efforts are recorded.  Ideally, there would be a service that
-would be able to index the contributions to the workshop, serving the
-authors, the organizers, and the larger community. But since there
-isn't such a service today, the workshop organizers are writing this
-initial report and making use of arXiv as a partial solution to
-provide a record of the workshop.
+The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop 
+agenda. However, contributors also want to get credit for their participation in the process. And the 
+workshop organizers want to make sure that the workshop content and their efforts are recorded. 
+Ideally, there would be a service that would be able to index the contributions to the workshop, 
+serving the authors, the organizers, and the larger community. But since there isn't such a service 
+today, the workshop organizers are writing this initial report and making use of arXiv as a partial 
+solution to provide a record of the workshop.
 
-After the workshop, one or more additional papers will be created that
-will include the discussions at the workshop. These papers will likely
-have many authors, and may be submitted to peer-reviewed journals.
+After the workshop, one or more additional papers will be created that will include the discussions at 
+the workshop. These papers will likely have many authors, and may be submitted to peer-reviewed 
+journals.
 
@@
 
-The following is a partial list of attendees who were recorded on the
-Google doc that was being used for live note taking at the workshop:
+The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:
 
-\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
-   Daniel S. Katz
-&  Neil Chue Hong
-&  Gabrielle Allen
-&  David Proctor
-\\ Manish Parashar
-&  Jay Alameda
-&  David Andrs
-&  Lorena A Barba
-\\ David Bernholdt
-&  Phil Bourne
-&  Karl Broman
-&  Maxine Brown
-\\ David Bruhwiler
-&  Bruno Bzeznik
-&  Jeffrey Carver
-&  John W. Cobb
-\\ Karen Cranston
-&  Rion Dooley
-&  Anshu Dubey
-&  Ian Foster
-\\Juliana Freire
-& Derek Gaston
-& Derek Groen
-& Edmund Hart
-\\James Hetherington
-& Simon Hettrick
-& Kenneth Hoste
-& James Howison
-\\Daisie Huang
-& Kerk Kee
-& Arnaud Legrand
-& Paul Madden
-\\Louis Maddox
-& Ketan Maheshwari
-& Suresh Marru
-& Cezary Mazurek
-\\James McClure
-& Chris Mentzel
-& J. Yates Monteith
-& Pierre Neyron
-\\Lucas Nussbaum
-& Cody Permann
-& John Peterson
-& Quan Pham
-\\Marlon Pierce
-& Heather Piwowar
-& Jason Riedy
-& Andreas Schreiber
-\\William Scullin
-& Andrew Slaughter
-& Arfon Smith
-& James Spencer
-\\Shel Swenson
-& Fr\'{e}d\'{e}ric Suter
-& Andy Terrel
-& George  Thiruvathukal
-\\John Towns
-& Erik Trainer
-& Matthew Turk
-& Eric Van Wyk
-\\Colin C. Venters
-& Von Welch
-& Nancy Wilkins-Diehr
-& Theresa Windus
-\\Rich Wolski
-& Lynn Zentner
-& Frank Löffler
-& Patrick O'Leary
-\end{tabular}
+Daniel S. Katz (@danielskatz)
+
+Neil Chue Hong (@npch)
+
+Gabrielle Allen
+
+David Proctor (@djproctor)
+
+Manish Parashar
+
+Jay Alameda (alameda@illinois.edu)
+
+David Andrs
+
+Lorena A Barba (@LorenaABarba)
+
+David Bernholdt (bernholdtde@ornl.gov)
+
+Phil Bourne
+
+Karl Broman (@kwbroman)
+
+Maxine Brown, UIC (maxine@uic.edu)
+
+David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)
+
+Bruno Bzeznik
+
+Jeffrey Carver, University of Alabama (carver@cs.ua.edu)
+
+John W. Cobb (cobbjw@ornl.gov, @johnwcobb)
+
+Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)
+
+Rion Dooley
+
+Anshu Dubey (adubey@lbl.gov)
+
+Ian Foster, @ianfoster
+
+Juliana Freire
+
+Derek Gaston
+
+Derek Groen (@whydoitweet on Twitter)
+
+Edmund Hart (@distribecology)
+
+James Hetherington @jamespjh
+
+Simon Hettrick (s.hettrick@software.ac.uk)
+
+Kenneth Hoste (@kehoste, Ghent University, Belgium)
+
+James Howison, University of Texas at Austin (@jameshowison)
+
+Daisie Huang
+
+Kerk Kee (kerk.kee@gmail.com)
+
+Arnaud Legrand
+
+Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu
+
+Louis Maddox (@lmmddx)
+
+Ketan Maheshwari (@km\_9 on Twitter)
+
+Suresh Marru (smarru@iu.edu, @sureshmarru)
+
+Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek
+
+James McClure
+
+Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org
+
+J. Yates Monteith, Clemson University
+
+Pierre Neyron
+
+Lucas Nussbaum (lucas.nussbaum@loria.fr)
+
+Cody Permann
+
+John Peterson
+
+Quan Pham, University of Chicago
+
+Marlon Pierce, marpierc@iu.edu, @marpierc
+
+Heather Piwowar
+
+Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places
+
+Andreas Schreiber (@onyame)
+
+William Scullin
+
+Andrew Slaughter
+
+Arfon Smith
+
+James Spencer (@exp\_n)
+
+Shel Swenson, University of Southern California (shel.swenson@usc.edu)
+
+Fr\'{e}d\'{e}ric Suter
+
+Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)
+
+George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)
+
+John Towns
+
+Erik Trainer
+
+Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)
+
+Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk
+
+Colin C. Venters, c.venters@hud.ac.uk
+
+Von Welch, @VonWelch
+
+Nancy Wilkins-Diehr, wilkinsn@sdsc.edu
+
+Theresa Windus (twindus@iastate.edu)
+
+Rich Wolski
+
+Lynn Zentner
+
+Frank Löffler
 

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
adding leads and volunteers, also adding Katz NSF ack

--- 2014-02-04 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
-\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
 \item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
@@
 
+\note{lead: Sou-Cheng (Terrya) Choi}
+
 The WSSSPE1 workshop began with two keynote presentations.
@@
 
+\note{lead: Marcus Hanwell, contributors: Suresh Marru}
+
 \todo{section should include what was presented and what was discussed}
@@
 
+\note{lead: Colin Venters}
+
 \todo{section should include what was presented and what was discussed}
@@
 
-\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
   Schnetter. Cactus: Issues for Sustainable Simulation Software
@@
 \section{Communities} \label{sec:community}
+
+\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
 
@@
 
+\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
+
+\todo{perhaps remove this section}
+
 \section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
+
+\note{lead: Daniel S. Katz}
 
@@
 
+\note{lead: Ketan Maheshwari, contributors: Andy Terrel}
+
 Perhaps discuss sustained software use-cases and issues surrounding
@@
   Software section.  Or perhaps this should go in the cross-cutting
-  issues section?}
+  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
 
@@
 
+\todo{anyone who needs to put something in here should}
+
+Some of the work by Katz was
+supported by the National Science Foundation while working at the
+Foundation; any opinion, finding, and conclusions or recommendations
+expressed in this material are those of the author(s) and do not
+necessarily reflect the views of the National Science Foundation.
+
 \appendix
 \section{List of attendees}
+
+\note{lead: Shel Swenson}
+
+\note{should alphabetize all, but if we get more from the SC13 folks, it's not worth spending
+time doing this now.  I wonder if there's a way to just have a simple list that we can keep ordered, and let LaTeX do the formatting, rather than manually creating a table as this is now}
 
@@
 & Lynn Zentner
-& Frank Löffler
+& Frank L\"{o}ffler
 & Patrick O'Leary

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes in process section

@@
 \item career paths for developers and institutional roles
-\item issues related to multiple organizations and countries,
+\item issues related to multiple organizations and multiple countries,
   such as intellectual property, licensing, etc.
@@
 arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
-their papers.  A peer review process followed the submissions, where
+their papers.
+
+A peer review process followed the submissions, where
 the 58 papers received 181 reviews, an average of 3.12 reviews per
@@
 reviewers to choose papers they wanted to review, then to provide
-scores on their relevance and general comments to the organizers and
+general comments and scores on relevance to the organizers and
 to the authors. The review reports enabled the organizers to decide

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
work in cross-cutting section

@@
 
-\begin{itemize}
-
-\item From discussion in Phil Bourne's keynote:
-\begin{itemize}
-\item Can sustainability be defined as the effort that happens to make
-  the essential things continue?
-\item What is sustainability? What is it that we want to sustain? Is
-  what we want to sustain valuable? Who cares if it goes away and how
-  do you measure how much they care?
-\end{itemize}
-
-\item from discussion in developing panel
+The question of what was meant by ``sustainability'' in the context of software
+came up in many different parts of the workshop, specifically in one of the
+keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}
+
+In his keynote, Phil Bourne suggested that perhaps sustainability can be defined
+as the effort that happens to make the essential things continue. This leads to
+having to decide what it is that we want to sustain, if what we want to sustain is
+valuable, and finally, who would care if it went away, and how does one measure
+how much they care.
+
+\note{rest of section will be turned from bullets into text...}
+
+In the Developing and Maintaining Software panel,
 
@@
 
-\item from discussion in policy panel
+In the Policy panel,
 
@@
 \item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
-
-\end{itemize}
 

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
listing software projects
adding to software usecase list
Section 3 Keynote: turn presentations and URLs into references

@@
 
-\section{Keynotes} \label{sec:keynotes}
-
-\note{lead: Sou-Cheng (Terrya) Choi}
+\section{Keynotes\note{lead: Sou-Cheng (Terrya) Choi}} \label{sec:keynotes}
 
@@
 The first
-keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
-was delivered by Philip E. Bourne of UCSD. The slides are available
-online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
-Philip is a basic biomedical scientist but has formed four software
+keynote~\cite{WSSSPE1-keynote1}
+was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
 companies. He also helped found PLOS and
@@
 
-Philip emphasized that Sustainability for software ``does not just
+Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
+He emphasized that sustainability for software ``does not just
 mean more money from Government.''  He noted that in places,
@@
 but for the most part remains arcane. He argued that there is much we
-can learn from the app-store model. He mentioned
+can learn from the App Store model. He mentioned
 BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
 OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
-as distribution channels.  On a related note, Philip observed a common
+as distribution channels.  On a related note, Bourne observed a common
 evolutionary pathway for computational biology projects, from data
@@
 
-Philip shared with the audience a few real challenges he
+Bourne shared with the audience a few real challenges he
 encountered. First, he has looked into reproducibility in
@@
 
-The second keynote\footnote{\url{http://is.gd/wssspe}} was by
+The second keynote~\cite{WSSSPE1-keynote2}\footnote{\url{}} was by
  Arfon Smith\footnote{\url{http://arfon.org/}} of Github.
@@
 Perhaps discuss sustained software use-cases and issues surrounding
-them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+them. Following is a list of software projects discussed throughout the workshop:
+
+\begin{itemize}
+    \item github
+    \item mvapich
+    \item Cactus
+    \item visTrails
+    \item MINRES/QLP
+    \item SimGrid
+    \item Swift
+    \item Galaxy
+    \item FastMath and package managers
+    \item R/qtl
+    \item OWL/RDF
+    \item AMR/multiphysics
+    \item DUNE
+    \item PAAS
+    \item Gateways
+    \item Globus-Genomics
+\end{itemize}
 

### 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3 Keynote: turn presentations and URLs into references
Section 3 Keynote: turn a URL into a reference

@@
 companies. He also helped found PLOS and
-PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.  He is
+PDB~\cite{PDB}.  He is
 working on automating three-dimensional visualizations of cell
@@
 
-The second keynote~\cite{WSSSPE1-keynote2}\footnote{\url{}} was by
- Arfon Smith\footnote{\url{http://arfon.org/}} of Github.
+The second keynote~\cite{WSSSPE1-keynote2} was delivered by
+ Arfon Smith of GitHub.
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

--- 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\note{papers and presentations and URLs should be references - see
-  figshare-web in .bib file for URL style example}
-
-People who have volunteered to work on this report: (with * for actual
-contributors - \note{if you contribute, add a * by your name})
+\note{papers and presentations should be references - URLs perhaps can just be footnotes for now, but may want to think about this again.}
+
+People who have volunteered to work on this report: (with * for actual contributors - \note{if you contribute, add a * by your name})
 
@@
 \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
-\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
 \item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
@@
 
-The First Workshop on Sustainable Software for Science: Practice and
-Experiences (WSSSPE1,
-\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
-Sunday, 17 November 2013, in conjunction with the 2013 International
-Conference for High Performance Computing, Networking, Storage and
-Analysis (SC13, \url{http://sc13.supercomputing.org}).
-
-Because progress in scientific research is dependent on the quality
-and accessibility of software at all levels, it is now critical to
-address many challenges related to the development, deployment, and
-maintenance of reusable software. In addition, it is essential that
-scientists, researchers, and students are able to learn and adopt
-software-related skills and methodologies. Established researchers are
-already acquiring some of these skills, and in particular a
-specialized class of software developers is emerging in academic
-environments who are an integral and embedded part of successful
-research teams. The WSSSPE workshop was intended to provide a forum
-for discussion of the challenges, including both positions and
-experiences.  The short papers and discussion have been archived to
-provide a basis for continued discussion, and the workshop has fed
-into the collaborative writing of this journal publication.
-Additional papers based on extended versions of workshop submissions
-are expected.  The level of interest in the workshop has led the
-organizers, working with some of the submitters and attendees, to plan
-two additional workshops (at the 2014 SciPy and SC conferences) and to
-turn the workshop website into a community website that can be used as
-a focus for further discussion and progress.  A minisymposium on
-``Reliable Computational Science'' that aims to further explore some
-of the key issues raised in WSSSPE is co-organized by a participant in
-the 2014 SIAM Annual Meeting.
-
-Before the WSSSPE1 workshop took place, the organizers self-published
-a paper~\cite{WSSSPE1-pre-report} to document the process of
-organizing and advertising the workshop, collecting and reviewing the
-papers, and putting together the agenda.  Section~\ref{sec:process} is
-based on that paper.  The remainder of this paper is based on the
-workshop itself, as documented in a set of collaborative notes taken
-during the workshop~\cite{WSSSPE1-google-notes}, including discussion
-about the keynote presentations (\S\ref{sec:keynotes}), the three
-panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
-(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
-use cases (\S\ref{sec:use-cases}), and conclusions
-(\S\ref{sec:conclusions}).
-
-\todo{Mention somewhere statistics of participants: There were 140 seats, I (Katz) think 149 people at the first keynote, and about 90 at the end of the workshop.
-They (SC13) counted people in the room 4 times during the day.
-But I don’t know how many unique people attended, which I agree they should know.}
-
-\section{Workshop Process and Agenda} \label{sec:process}
-
-WSSSPE1 was organized in collaboration of the relatively small group
-of five organizers and a larger peer-review committee with 36
-members. This committee had early influence on the organization, e.g.,
-on the specific call for papers.
+The First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)\footnote{\url{http://wssspe.researchcomputing.org.uk/WSSSPE1}} was held on Sunday, 17 November 2013, in conjunction with the 2013 International Conference for High Performance Computing, Networking, Storage and Analysis (SC13)\footnote{\url{http://sc13.supercomputing.org}}.
+
+Because progress in scientific research is dependent on the quality and accessibility
+of software at all levels, it is now critical to address many challenges related to the
+development, deployment, and maintenance of reusable software. In addition, it is
+essential that scientists, researchers, and students are able to learn and adopt
+software-related skills and methodologies. Established researchers are already
+acquiring some of these skills, and in particular a specialized class of software
+developers is emerging in academic environments who are an integral and embedded
+part of successful research teams. The WSSSPE workshop was intended to provide a
+forum for discussion of the challenges, including both positions and experiences.
+The short papers and discussion have been archived to provide a basis for continued
+discussion, and the workshop has fed into the collaborative writing of this journal publication.
+Additional papers based on extended versions of workshop submissions are expected.
+The level of interest in the workshop has led the organizers, working with some of the submitters
+and some of the attendees, to plan two additional workshops (\todo{details}) and to turn
+the workshop website into a community website that can be used as a focus for further
+discussion and progress.
+
+\todo{Mention a two-session minisymposium on ``Reliable Computational Science'' in the SIAM Annual Meeting in July.
+It aims to further explore some of the key issues raised in WSSSPE.
+Speakers include Tim Davis, Barry Smith, Lorena Barba, Tanu Malik, Bill Rider.}
+
+\todo{mention arXiv:1311.3523 and explain what's different here}
+
+\section{Workshop Process}
+
+WSSSPE1 was organized in collaboration of the relatively small group of \todo{how many} organizers and
+a larger peer-review committee consisting of \todo{how many members}. This committee had early influence on the organization,
+e.g., on the specific call for papers.
 
@@
 
-\begin{itemize}[leftmargin=0.2in]
+\begin{itemize}
 \item the development process that leads to new (versions of) software
-\begin{itemize}[leftmargin=0.2in]
-\item how fundamental research in computer science or
-  science/engineering domains is turned into reusable software
+\begin{itemize}
+\item how fundamental research in computer science or science/engineering domains is turned  into reusable software
 \item software created as a by-product of research
-\item impact of computer science research on the development of
-  scientific software
-\end{itemize}
-\item the support and maintenance of existing software, including
-  software engineering
-\begin{itemize}[leftmargin=0.2in]
+\item impact of computer science research on the development of scientific software
+\end{itemize}
+\item the support and maintenance of existing software, including software engineering
+\begin{itemize}
 \item governance, business, and sustainability models
-\item the role of community software repositories, their operation and
-  sustainability
+\item the role of community software repositories, their operation and sustainability
 \end{itemize}
@@
 \item career paths for developers and institutional roles
-\item issues related to multiple organizations and multiple countries,
-  such as intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of
-  publishers
+\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
+\item mechanisms and venues for publishing software, and the role of publishers
 \end{itemize}
@@
 58 submissions were received, and almost all submitters used either 
-arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
-their papers.
-
-A peer review process followed the submissions, where
-the 58 papers received 181 reviews, an average of 3.12 reviews per
-paper. Reviews were completed using a Google form, which allowed
-reviewers to choose papers they wanted to review, then to provide
-general comments and scores on relevance to the organizers and
-to the authors. The review reports enabled the organizers to decide
-which papers to associate with the workshop, and allowed the authors
-to improve their papers.
-
-The organizers decided to list 54 of the papers as significant
-contributions to the workshop, a very high acceptance rate, but one
-that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. The papers were
-grouped into three main categories, namely \emph{Developing and
-Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-subject area was associated with a panel and discussion at the
-workshop.
-
-The workshop itself consisted of two keynote presentations and the
-three panels/discussion sessions. The panels were organized based on a
-classification of the workshop submissions into three categories,
-following the themes of the call for papers as modified by the areas
-of the submissions. Three to four representatives from each submission
-category were appointed as panelists, and assigned to read a subset of
-the paper in that category and then discuss them in the panel.
-
-\section{Keynotes} \label{sec:keynotes}
-
-\note{lead: Sou-Cheng (Terrya) Choi}
-
-The WSSSPE1 workshop began with two keynote presentations.
+arXiv\footnote{\url{http://arxiv.org}} or 
+figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+
+A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
+average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
+reviewers to provide scores on relevance and comments to the organizers and to the authors.
+The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.
+
+The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
+high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
+that the reports were already self-published.
+The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}.
+Each subject area was associated with a panel and discussion at the workshop.
+
+\section{Agenda}
+
+The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
+were organized based on a classification of the workshop submissions into three categories, following 
+the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
+each submission categories were appointed as panelists, to discuss the papers in that category.
+
+\section{Keynotes}
 
@@
 
-\todo{change references to papers or presentations to cites}
-
 \subsection{A recipe for sustainable software, Philip E. Bourne.}
 
-The first
-keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
-was delivered by Philip E. Bourne of UCSD. The slides are available
-online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delivered by Philip E. Bourne of UCSD. The slides
+are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+
 Philip is a basic biomedical scientist but has formed four software
-companies. He also helped found PLOS and
-PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.  He is
-working on automating three-dimensional visualizations of cell
-contents and molecular structures, a problem that has not been solved
-and would serve as a key function of software in biomedical sciences.
-
-Philip emphasized that Sustainability for software ``does not just
-mean more money from Government.''  He noted that in places,
-development in science has improved thanks to open source and Github,
-but for the most part remains arcane. He argued that there is much we
-can learn from the app-store model. He mentioned
-BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
-OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
-as distribution channels.  On a related note, Philip observed a common
-evolutionary pathway for computational biology projects, from data
-archive to analytics platform to educational use, and suggested that
-use of scientific software for outreach might be the final step.
-
-Philip shared with the audience a few real challenges he
-encountered. First, he has looked into reproducibility in
-computational biology, and has concluded that ``I have proved I cannot
-reproduce research from my own lab''~\cite{Veretnik}.
-
-Another problem he experienced was staff retention with respect to
-private organizations which reward those combining research and
-software expertise (the ``Google Bus''). However, he is a strong fan
-of private sector sustainability sources and public-private
-partnerships. He noted that making a successful business from
-scientific software alone is very rare: founders overvalue while
-customers undervalue. He noted that to last, an open source project
-needs a minimal funding requirement even with a vibrant community --
-good will only goes so far if you are being paid to do something else.
-He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology
-transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
-
-He also had problems with selling research software: the technology
-transfer office in his university wanted huge IP reachthrough, whereby
-they would get a share of profits from drugs developed by pharma
-companies who use the software. He was aware this was unrealistic but
-the technology transfer office insisted for a while. He wants to push
-a one-click approach to purchasing university-written software.
+companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+
+He emphasized that Sustainability for software ``does not just mean more
+money from Government.''
+
+He noted that creating effective illustrations is a key function of
+software in the biomedical sciences: ``Automating Conceptualization''
+(Visualizing molecular structures is a problem that has now been solved,
+he is working on 3-d visualizations of the contents of cells.)
+
+He has looked into reproducibility in computational biology, and has
+concluded that ``I have proved I cannot reproduce research from my own
+lab''~\cite{Veretnik}.
+
+He noted that in places, development in science has improved thanks to
+open source and Github, but for the most part remains arcane. He argues
+that there is a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
+OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
+distribution channels.
+
+He noted a common evolutionary pathway for computational biology
+projects, from data archive to analytics platform to educational use. He
+suggests that use for outreach might be the final step.
+
+He experienced problems with staff retention with respect to private organizations which reward 
+those combining research and software expertise (the ``Google Bus''). However,
+he is a strong fan of private sector sustainability sources and PPP. He
+noted that making a successful business from scientific software alone
+is very rare: founders overvalue while customers undervalue. He noted
+that to last, an open source project needs a minimal funding requirement
+even with a vibrant community -- good will only goes so far if you are
+being paid to do something else.
+
+He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}
+
+He had problems with selling research software: the technology transfer
+office in his university wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the
+software. He was aware this was unrealistic but the technology transfer
+office insisted for a
+while. He wants to push a one-click approach to purchasing
+university-written software.
 
@@
 
-On the idea of the \textbf{digital enterprise}, where
-IT underpins the whole of organizational activities, he contended that
+He then expanded on the idea of the \textbf{digital enterprise}, where
+IT underpins the whole of organizational activities. He contends that
 universities are way behind the curve on this. In particular, he
-highlighted the separation of research, teaching and administration into
+highlights the separation of research, teaching and administration into
 silos without a common IT framework as a blocker to many useful
@@
 
-He concluded with a reference to his paper~\cite{bourne_ten} and
-argued that computational scientists ``have a responsibility to
-convince their institutions and communities that software is
-scholarship,'' a point with which we strongly agree.
-
-\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}
+He concluded with a reference to his paper~\cite{bourne_ten} and argued that computational scientists ``have a
+responsibility to convince their institutions and communities that
+software is scholarship,'' a point with which we strongly agree.
+
+\subsection{Scientific Software and the Open Collaborative Web}
+
+\todo{change references to papers or presentations to cites}
 
@@
 His talk used an example from data reduction in Astronomy, where he
-needed to remove interfering effects from the device. He built a ``bad
-pixel mask'', and realized that while this is persistent, there was no
-way/practice of sharing this with the data, so many researchers
-repeated the same calculations. He estimated that 13 person-years were
-wasted in this redundant calculation.
+needed to remove interfering effects from the device. He built a
+``bad pixel mask'', and realized though while this is persistent, there
+was no way/practice of sharing this with the data, so many researchers
+repeated the same calculations. He estimated that 13 wasted person-years were wasted in
+this redundant calculation.
 
 He asked, of this practice, ``why didn't we know better?'' and argued
-this was because we were taught to focus on immediate research
-outcomes and not on continuously improving and building on tools for
-research. He then asked, later in our working lives, when we do know
-better, why don't we act any different, and argued this was due to
-incentives and their lack: only the immediate products of research,
-not the software, are valued.  He referenced Victoria Stodden's talk
-at
-OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+this was because we were taught to focus on immediate research outcomes
+and not on continuously improving and building on tools for research. He
+then asked, later in our working lives, when we do know better, why
+don't we act any different, and argued this was due to incentives and their lack:
+only the immediate products of research, not the software, are valued.
+He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
 which he said argued these points well.
 
-He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
-Chain Monte Carlo code by Dan Forman
-Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
-Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
-as good examples of scientific projects using GitHub.
-
-C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
-regard to reusable software, ``we should just start doing it.'' Arfon
-noted that documentation should be ``treated as a first class
-entity.''  He noted that some software fields have excellent cultures
-of module reuse, such as
-rubygems\footnote{\url{http://rubygems.org/}},
-PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
-CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
+Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} and Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}} as 
+good examples of  scientific projects using GitHub.
+
+C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
+reusable software, ``we should just start doing it.'' Arfon noted that
+documentation should be ``treated as a first class entity.''
+He noted that some software fields have excellent cultures of module
+reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
+PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
 ultra-low-friction collaboration through these repositories. This has
-not happened in highly numerical, compiled language scientific
-software.
+not happened in highly numerical, compiled language scientific software.
 
@@
 
-Fernando Perez says that ``Open Source is reproducible by necessity''
-- since reproducibility is a prerequisite for remote collaboration.
-Arfon therefore defined GitHub as part of the next stage in
-development of the web, following on from the ``social web'' of
-Facebook et.al.: ``The Collaborative Web.''
-
-\section{Developing and Supporting Software} \label{sec:devel}
-
-\note{lead: Marcus Hanwell, contributors: Suresh Marru}
+Fernando Perez says that ``Open Source
+is reproducible by necessity'' - since reproducibility is a prerequisite
+for remote collaboration.
+Arfon therefore defined GitHub as part of the next stage in development
+of the web, following on from the ``social web'' of Facebook et.al.:
+``The Collaborative Web.''
+
+\section{Developing and Supporting Software}
 
@@
 
-\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
-  McInnes, Barry Smith. Package Management Practices Essential for
-  Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath \cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable
-  \cite{Broman_WSSSPE}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
-
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?
-  \cite{Stephan_WSSSPE}
-
-\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
-  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development
-  \cite{Gaston_WSSSPE}
-
-\item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks
-  \cite{Dubey_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
-
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
-  adapting sustainable scientific software development processes in a
-  small biology lab \cite{Crusoe_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
-The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
-
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
-  H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes \cite{Heien_WSSSPE}
+\item Mark C.
+Miller, Lori Diachin, Satish Balay, Lois Curfman McInnes, Barry Smith.
+Package Management Practices Essential for Interoperability: Lessons Learned and Strategies Developed for FASTMath \cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable \cite{Broman_WSSSPE}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura Riihimaki. What Comes First, the OWL or the Bean? \cite{Stephan_WSSSPE}
+
+\item Derek R.
+Gaston, John Peterson, Cody J.
+Permann, David Andrs, Andrew E.
+Slaughter, Jason M.
+Miller, Continuous Integration for Concurrent Computational Framework and Application Development \cite{Gaston_WSSSPE}
+
+\item Anshu Dubey, B. Van Straalen. Experiences from Software Engineering of Large Scale AMR Multiphysics Code Frameworks \cite{Dubey_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and adapting sustainable scientific software development processes in a small biology lab \cite{Crusoe_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise H. Kellogg. Experiences with Automated Build and Test for Geodynamics Simulation Codes \cite{Heien_WSSSPE}
 
@@
 
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
-  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
-  Versatile Simulation of Large Scale Distributed
-  Systems~\cite{Casanova_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software \cite{Trainer_WSSSPE}
-
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
-  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
-  Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software \cite{Cohen_WSSSPE}
-
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications
-  \cite{Slawinski_WSSSPE}
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson, Fr\'{e}d\'{e}ric Suter.
+SimGrid: a Sustained Effort for the Versatile Simulation of Large Scale Distributed Systems~\cite{Casanova_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
+
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey, Malcolm Illingworth, Andrew Turner, John Darlington, Spencer Sherwin.
+Simplifying the Development, Use and Sustainability of HPC Software \cite{Cohen_WSSSPE}
+
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic Deployment of Scientific and Engineering Applications \cite{Slawinski_WSSSPE}
 
@@
 
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software \cite{Prlic_WSSSPE}
-
-\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
-  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
-  R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
-
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult \cite{Carver_WSSSPE}
-
-\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
-  K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability
-  \cite{Stewart_WSSSPE}
-
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable \cite{Brown_WSSSPE}
-
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
-  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
-  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms \cite{Swenson_WSSSPE}
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the Open Development of Scientific Software \cite{Prlic_WSSSPE}
+
+\item Anshu Dubey, S.
+Brandt, R.
+Brower, M.
+Giles, P.
+Hovland, D.
+Q.
+Lamb, F.
+Löffler, B.
+Norris, B.
+O'Shea, C.
+Rebbi, M.
+Snir, R.
+Thakur, Software Abstractions and Methodologies for HPC Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
+
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering Need Not Be Difficult \cite{Carver_WSSSPE}
+
+\item Craig A.
+Stewart, Julie Wernert, Eric A.
+Wernert, William K.
+Barnett, Von Welch.
+Initial Findings from a Study of Best Practices and Models for Cyberinfrastructure Software Sustainability \cite{Stewart_WSSSPE}
+
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility: anything less is unsustainable \cite{Brown_WSSSPE}
+
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar, Jason Riedy, David Bader, Richard Vuduc.
+Sustainable Software Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging Platforms \cite{Swenson_WSSSPE}
 
@@
 
-Who creates training materials, how are they distributed, integration
-into courses when projects see very wide application in research.
-
-\section{Policy} \label{sec:policy}
-
-\note{lead: Colin Venters}
+Who creates training materials, how are they distributed, integration into courses when projects see very wide application in research.
+
+\section{Policy}
 
@@
 
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model \cite{Calero_WSSSPE}
-
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework
-  \cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance
-  \cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science
-  \cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science
-  \cite{Weber_WSSSPE}
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a Software Product Sustainability Model \cite{Calero_WSSSPE}
+
+\item Colin C.
+Venters, Lydia Lau, Michael K.
+Griffiths, Violeta Holmes, Rupert R.
+Ward, Jie Xu.
+The Blind Men and the Elephant: Towards a Software Sustainability Architectural Evaluation Framework \cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable Cyberinfrastructure Software Through Open Governance \cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura Christopherson, Ray Idaszak.
+Data Management Lifecycle and Software Lifecycle Management in the Context of Conducting Science \cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling: Ecological Metaphors for Sustainable Software in Science \cite{Weber_WSSSPE}
 
@@
 
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
-  Smith. Accurately Citing Software and Algorithms Used in
-  Publications \cite{Knepley_WSSSPE}
-
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project \cite{Priem_WSSSPE}
-
-\item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns \cite{Katz2_WSSSPE}
-
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software
-  \cite{Chue_Hong_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry Smith. Accurately Citing Software and Algorithms Used in Publications \cite{Knepley_WSSSPE}
+
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact report for every software project \cite{Priem_WSSSPE}
+
+\item Daniel S. Katz. Citation and Attribution of Digital Products: Social and Technological Concerns \cite{Katz2_WSSSPE}
+
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers: improving the reusability and sustainability of scientific software \cite{Chue_Hong_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area were also discussed in this area.
+
+\begin{itemize}
+
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
 
@@
 
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model \cite{Heiland_WSSSPE}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
-
-\item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software
-  \cite{Huang_WSSSPE}
-
-\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher. Developing Systems for API Governance
-  \cite{Wolski_WSSSPE}
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a Research Software Security Maturity Model \cite{Heiland_WSSSPE}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+
+\item Daisie Huang, Hilmar Lapp. Software Engineering as Instrumentation for the Long Tail of Scientific Software \cite{Huang_WSSSPE}
+
+\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos Dimopoulos, Alexander Pucher. Developing Systems for API Governance \cite{Wolski_WSSSPE}
 
@@
 
-How to ensure software is sustainable by ensuring there are career
-paths for developers.
-
-\section{Communities} \label{sec:community}
-
-\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
+How to ensure software is sustainable by ensuring there are career paths for developers.
+
+\section{Communities}
 
@@
 
+There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.
+
+\subsection{Communities Part 1}
+
+This collection of papers was summarized by Karen Cranston.
+
+\subsection{Communities Part 2}
+
+Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
+relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
+The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
+This is a 4-step iterative development process that incorporates Agile development principals.
+
+\begin{itemize}
+
+\item Step 1 - Design - thorough discussion of research questions
+\item Step 2 - Develop working code
+\item Step 3 - Refine based on new requirements
+\item Step 4 - Publish open source
+
+\end{itemize}
+
+The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
+code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
+build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
+Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
+In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
+Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
+Co‐locate rather than collaborating remotely whenever feasible.
+
+The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure. The paper opens with a description of the explosion of use of just a few gateways. 
+Over 3.5 years, more than 7,000 biologists have run phylogenetic codes on supercomputers using the CIPRES Science Gateway.
+In 1 year, over 120 scientists from 50 institutions used the UltraScan Science Gateway,
+increasing the sophistication of analytical ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG) registered 100+ 
+users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often make use of common components. Science Gateway Platform as a Service (Sci-GaP)
+delivers middleware as a hosted, scalable third ­party service while domain developers focus on user interfaces and domain-specific data in the gateway.
+The middleware handles things like authentication, application installation and reliable execution and help desk support.
+
+One key to SciGaP is its openness.
+While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
+The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
+stakeholder requirements. The project is leveraging Internet2’s Net+ mechanisms for converting SciGaP and its gateways into commodity services.
+
+The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as 
+a “body of persons of common and especially professional interests scattered through a larger society.” This makes support
+challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
+whether they are indirectly alienating part of the community or encouraging particular types of use.
+
+nanoHUB's 10-year history with over 260,000 users annually provides a lot of data to analyze. 4000 resources contributed by
+1000 authors. nanoHUB serves both the research and education community and the contribution model allows researchers to get
+their codes out into the community and in use in education very rapidly. The primary software challenges are twofold - support
+for the HUBzero framework and challenges related to the software contributed by the community.
+
+The group has learned that community contributions are maximized with a tolerant licensing approach. HUBzero uses an 
+LGPLv3 license so contributors can create unique components and license as they choose. If they make changes to source 
+code, the original license must be maintained for redistribution. As far as contributed resources, these must be open access,
+but not necessarily open source. This allows contributors to meet the requirements of their institutions and funding agencies.
+Quality is maintained via user ratings. Documentation is encouraged and nanoHUB supplies regression test capabilities, but 
+the user community provides ratings, poses questions and contributes to wishlists and citation counts - all of which incentivize
+code authors.
+
+The Terrel paper describes support for the Python scientific community through two major efforts - the SciPy conference and the
+NumFOCUS foundation. Reliance on software in science has driven a huge demand for development, but this development is typically done as a side effort
+and often in a rush to publish without documentation and testing.
+While created by academics, software support often falls to industrial institutions. SciPy brings together industry, government, 
+and academics to share their code and experience in an open environment.
+
+NumFOCUS promotes open, usable scientific software while sustaining the community. Specific activities include educational programs, 
+collaborative research tools and documentation and 
+promotion of high­-level languages, reproducible scientific research, and open­code developmet.
+Governance of the non-profit is a loose grantor-­grantee relationship with projects allowing for monies to be placed in the groups accounts.
+This has raised money to hire developers for open code development, maintain testing machines, organize the PyData conference series, and sponsor community members to attend conferences.
+
+Software sustainability relies on contributions from all sectors of user community.
+Together SciPy and NumFOCUS support these sectors.
+By maximizing contributions growing the user base they help develop and mature Python.
+
+The Loffler paper
+
 \subsection{Papers}
 
-\subsubsection*{Communities}
-
-\begin{itemize}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
-  Software through the Open Community Engagement Process
-  \cite{Christopherson_WSSSPE}
-
-\item Reagan Moore. Extensible Generic Data Management Software
-  \cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability
-  \cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities \cite{Vay_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
-  Borries Demeler. Science Gateway Operational Sustainability:
-  Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community \cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
-  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
-  Mookherjee. Reusability in Science: From Initial User Engagement to
-  Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
-  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
-  Dooley, Dan Stanzione. Science Gateways and the Importance of
-  Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
-  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
-  tools for facilitating Open Science \cite{Hart_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area was also discussed
-in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\subsubsection*{Communities Part 1 Papers}
+
+\begin{itemize}
+
+\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-\item Anne C. Elster. Software for Science: Some Personal Reflections
-  \cite{Elster_WSSSPE}
-
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability \cite{Foster_WSSSPE}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
-  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC
-  \cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
+\item Anne C. Elster. Software for Science: Some Personal Reflections \cite{Elster_WSSSPE}
+
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service as a path to software sustainability \cite{Foster_WSSSPE}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed in this area.
+
+\begin{itemize}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable Scientific Software \cite{Blanton_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source Scientific Software Development \cite{Blatt_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar. The MVAPICH Project: Evolution and Sustainability of an Open Source Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
 
@@
 
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
-  T. Brown. Advanced Techniques for Scientific Programming and
-  Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)
-  \cite{Girotto_WSSSPE}
-
-\item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry \cite{Crawford_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software \cite{Trainer_WSSSPE}
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn T.
+Brown.
+Advanced Techniques for Scientific Programming and Collaborative Development of Open Source Software Packages at the International Centre for Theoretical Physics (ICTP) \cite{Girotto_WSSSPE}
+
+\item Thomas Crawford. On the Development of Sustainable Software for Computational Chemistry \cite{Crawford_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed in this area.
+
+\begin{itemize}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith, Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences with Sustainable Software \cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable Reproducible Research via Staunch Scientific Software \cite{Choi_WSSSPE}
+
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big Effects of Short-term Efforts: A Catalyst for Community Engagement in Scientific Software \cite{Trainer_WSSSPE}
 
@@
 
-\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
-
-\todo{perhaps remove this section}
-
-\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
-
-\note{lead: Daniel S. Katz}
-
-The question of what was meant by ``sustainability'' in the context of software
-came up in many different parts of the workshop, specifically in one of the
-keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}
-
-In his keynote, Phil Bourne suggested that perhaps sustainability can be defined
-as the effort that happens to make the essential things continue. This leads to
-having to decide what it is that we want to sustain, if what we want to sustain is
-valuable, and finally, who would care if it went away, and how does one measure
-how much they care.
-
-\note{rest of section will be turned from bullets into text...}
-
-In the Developing and Maintaining Software panel,
-
-\begin{itemize}
-\item what does sustainable mean?
-\begin{itemize}
-\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
-  software for procurement or development reuse - regardless of the
-  license and development model you will use - you need to consider
-  the future. While a software product may satisfy today's needs, will
-  it satisfy tomorrow's needs? Will the supplier still be around in
-  five years' time? Will the supplier still care for all its customers
-  in five years' time? Will the supplier be responsive to bug reports
-  and feature requests? In other words, is the software sustainable?
-\item I think this is a key question we as a community need to agree
-  on (whats our problem statement?) and hope we come out today with
-  some progress. - @VonWelch
-\item What is the goal of sustainability? Reproducible science,
-  persistence, quality, or else?
-\item How do you measure success? How do you know when you have
-  actually achieved sustainable software?
-\item Sustainability should be addressed throughout the life cycle
-  (Daisie Huang)
-\end{itemize}
-
-\item What is the relationship between sustainability and provenance?
-
-\item Is usability separate from sustainability or a fundamental part?
-\end{itemize}
-
-In the Policy panel,
-
-\begin{itemize}
-\item Chris Lenhardt's presentation of Venters
-  paper~\cite{Venters_WSSSPE} included:
-
-\begin{itemize}
-\item What does software sustainability mean
-It ``is a rather ambiguous concept...''
-\item Lack of an accepted definition gets in the way of integrating
-  the concept into software engineering
-\item Sustainability as a non-functional requirement
-\item Quality of software architectures determines sustainability
-\item Propose sustainability is a measure of central quality
-  attributes: Extensibility, Interoperability, Maintainability,
-  Portability, Reusability, Scalability
-\item Develop an architecture evaluation framework based on scenarios
-  which help to illuminate how to measure quality/sustainability at
-  the architectural level
-\end{itemize}
-
-\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
-  included:
-\begin{itemize}
-\item Clarifying the notion of sustainability
-\begin{itemize}
-\item Provide same functionality going forward even if environment changes
-\item Is functionality and usability clear
-\item Will functionality be correct going into the future even if
-  environment changes
-\item Is the functionality desired by current and future users
-\item Can it incorporate new things as they develop
-\end{itemize}
-\end{itemize}
-
-\item see slide 9 of
-  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
-
-\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
-
-\end{itemize}
-
-\section{Illustrative Use Cases} \label{sec:use-cases}
-
-\note{lead: Ketan Maheshwari, contributors: Andy Terrel}
-
-Perhaps discuss sustained software use-cases and issues surrounding
-them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
-
-\note{Dan: do we need this section?  I think there's some overlap
-  between this and the discussion in the Developing and Support
-  Software section.  Or perhaps this should go in the cross-cutting
-  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
-
-\section{Conclusions} \label{sec:conclusions}
+\section{Cross-cutting Issues}
+
+\subsection{Defining sustainability}
+
+\subsection{Licensing}
+
+\note{not sure if this was really cross-cutting...}
+
+\section{Illustrative use-cases}
+
+Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+
+\section{Conclusions}
 
@@
 
-The WSSSPE workshop has begun an experiment in how we can
-collaboratively build a workshop agenda. However, contributors also
-want to get credit for their participation in the process. And the
-workshop organizers want to make sure that the workshop content and
-their efforts are recorded.  Ideally, there would be a service that
-would be able to index the contributions to the workshop, serving the
-authors, the organizers, and the larger community. But since there
-isn't such a service today, the workshop organizers are writing this
-initial report and making use of arXiv as a partial solution to
-provide a record of the workshop.
-
-After the workshop, one or more additional papers will be created that
-will include the discussions at the workshop. These papers will likely
-have many authors, and may be submitted to peer-reviewed journals.
+The WSSSPE workshop has begun an experiment in how we can collaboratively build a workshop 
+agenda. However, contributors also want to get credit for their participation in the process. And the 
+workshop organizers want to make sure that the workshop content and their efforts are recorded. 
+Ideally, there would be a service that would be able to index the contributions to the workshop, 
+serving the authors, the organizers, and the larger community. But since there isn't such a service 
+today, the workshop organizers are writing this initial report and making use of arXiv as a partial 
+solution to provide a record of the workshop.
+
+After the workshop, one or more additional papers will be created that will include the discussions at 
+the workshop. These papers will likely have many authors, and may be submitted to peer-reviewed 
+journals.
 
 \section*{Acknowledgments}
-
-\todo{anyone who needs to put something in here should}
-
-Some of the work by Katz was
-supported by the National Science Foundation while working at the
-Foundation; any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
 
@@
 
-\note{lead: Shel Swenson}
-
-\note{should alphabetize all, but if we get more from the SC13 folks, it's not worth spending
-time doing this now.  I wonder if there's a way to just have a simple list that we can keep ordered, and let LaTeX do the formatting, rather than manually creating a table as this is now}
-
-The following is a partial list of attendees who were recorded on the
-Google doc that was being used for live note taking at the workshop:
-
-\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
-   Daniel S. Katz
-&  Neil Chue Hong
-&  Gabrielle Allen
-&  David Proctor
-\\ Manish Parashar
-&  Jay Alameda
-&  David Andrs
-&  Lorena A Barba
-\\ David Bernholdt
-&  Phil Bourne
-&  Karl Broman
-&  Maxine Brown
-\\ David Bruhwiler
-&  Bruno Bzeznik
-&  Jeffrey Carver
-&  John W. Cobb
-\\ Karen Cranston
-&  Rion Dooley
-&  Anshu Dubey
-&  Ian Foster
-\\Juliana Freire
-& Derek Gaston
-& Derek Groen
-& Edmund Hart
-\\James Hetherington
-& Simon Hettrick
-& Kenneth Hoste
-& James Howison
-\\Daisie Huang
-& Kerk Kee
-& Arnaud Legrand
-& Paul Madden
-\\Louis Maddox
-& Ketan Maheshwari
-& Suresh Marru
-& Cezary Mazurek
-\\James McClure
-& Chris Mentzel
-& J. Yates Monteith
-& Pierre Neyron
-\\Lucas Nussbaum
-& Cody Permann
-& John Peterson
-& Quan Pham
-\\Marlon Pierce
-& Heather Piwowar
-& Jason Riedy
-& Andreas Schreiber
-\\William Scullin
-& Andrew Slaughter
-& Arfon Smith
-& James Spencer
-\\Shel Swenson
-& Fr\'{e}d\'{e}ric Suter
-& Andy Terrel
-& George  Thiruvathukal
-\\John Towns
-& Erik Trainer
-& Matthew Turk
-& Eric Van Wyk
-\\Colin C. Venters
-& Von Welch
-& Nancy Wilkins-Diehr
-& Theresa Windus
-\\Rich Wolski
-& Lynn Zentner
-& Frank L\"{o}ffler
-& Patrick O'Leary
-\end{tabular}
+The following is a partial list of attendees who were recorded on the Google doc that was being used for live note taking at the workshop:
+
+Daniel S. Katz (@danielskatz)
+
+Neil Chue Hong (@npch)
+
+Gabrielle Allen
+
+David Proctor (@djproctor)
+
+Manish Parashar
+
+Jay Alameda (alameda@illinois.edu)
+
+David Andrs
+
+Lorena A Barba (@LorenaABarba)
+
+David Bernholdt (bernholdtde@ornl.gov)
+
+Phil Bourne
+
+Karl Broman (@kwbroman)
+
+Maxine Brown, UIC (maxine@uic.edu)
+
+David Bruhwiler   (bruhwiler@radiasoft.net    @radiasoft)
+
+Bruno Bzeznik
+
+Jeffrey Carver, University of Alabama (carver@cs.ua.edu)
+
+John W. Cobb (cobbjw@ornl.gov, @johnwcobb)
+
+Karen Cranston (@kcranstn, National Evolutionary Synthesis Center)
+
+Rion Dooley
+
+Anshu Dubey (adubey@lbl.gov)
+
+Ian Foster, @ianfoster
+
+Juliana Freire
+
+Derek Gaston
+
+Derek Groen (@whydoitweet on Twitter)
+
+Edmund Hart (@distribecology)
+
+James Hetherington @jamespjh
+
+Simon Hettrick (s.hettrick@software.ac.uk)
+
+Kenneth Hoste (@kehoste, Ghent University, Belgium)
+
+James Howison, University of Texas at Austin (@jameshowison)
+
+Daisie Huang
+
+Kerk Kee (kerk.kee@gmail.com)
+
+Arnaud Legrand
+
+Paul Madden, NOAA / University of Colorado, maddenp@colorado.edu
+
+Louis Maddox (@lmmddx)
+
+Ketan Maheshwari (@km\_9 on Twitter)
+
+Suresh Marru (smarru@iu.edu, @sureshmarru)
+
+Cezary Mazurek, mazurek@man.poznan.pl, @cezary\_mazurek
+
+James McClure
+
+Chris Mentzel @chrismentzel - firstname dot lastname at moore dot org
+
+J. Yates Monteith, Clemson University
+
+Pierre Neyron
+
+Lucas Nussbaum (lucas.nussbaum@loria.fr)
+
+Cody Permann
+
+John Peterson
+
+Quan Pham, University of Chicago
+
+Marlon Pierce, marpierc@iu.edu, @marpierc
+
+Heather Piwowar
+
+Jason Riedy, Georgia Tech, http://fmrl.me/jasonriedy and many other places
+
+Andreas Schreiber (@onyame)
+
+William Scullin
+
+Andrew Slaughter
+
+Arfon Smith
+
+James Spencer (@exp\_n)
+
+Shel Swenson, University of Southern California (shel.swenson@usc.edu)
+
+Fr\'{e}d\'{e}ric Suter
+
+Andy Terrel (@aterrel, andy.terrel@gmail.com, NumFOCUS and Continuum Analytics)
+
+George K. Thiruvathukal, Loyola University Chicago and Computing in Science and Engineering (IEEE CS/AIP) (gkt@cs.luc.edu)
+
+John Towns
+
+Erik Trainer
+
+Matthew Turk (@powersoffour on Twitter, matthewturk@gmail.com)
+
+Eric Van Wyk, University of Minnesota, evw@cs.umn.edu, @ericvanwyk
+
+Colin C. Venters, c.venters@hud.ac.uk
+
+Von Welch, @VonWelch
+
+Nancy Wilkins-Diehr, wilkinsn@sdsc.edu
+
+Theresa Windus (twindus@iastate.edu)
+
+Rich Wolski
+
+Lynn Zentner
+
+Frank Löffler
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
more work in sustainability section (8)

--- 2014-02-05 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\note{rest of section will be turned from bullets into text...}
-
-In the Developing and Maintaining Software panel,
-
-\begin{itemize}
-\item what does sustainable mean?
-\begin{itemize}
-\item from http://oss-watch.ac.uk/resources/ssmm: When choosing
+In the Developing and Maintaining Software panel, there was some discussion
+on this topic: what does sustainability mean? It was pointed out that
+OSS Watch proposes a Software Sustainability Maturity Model to address
+the issue of how sustainability a particular element of software is, and says
+that this sustainability is important. ``When choosing
 software for procurement or development reuse - regardless of the
@@
 in five years' time? Will the supplier be responsive to bug reports
-  and feature requests? In other words, is the software sustainable?
-\item I think this is a key question we as a community need to agree
-  on (whats our problem statement?) and hope we come out today with
-  some progress. - @VonWelch
-\item What is the goal of sustainability? Reproducible science,
-  persistence, quality, or else?
-\item How do you measure success? How do you know when you have
-  actually achieved sustainable software?
-\item Sustainability should be addressed throughout the life cycle
-  (Daisie Huang)
-\end{itemize}
-
-\item What is the relationship between sustainability and provenance?
-
-\item Is usability separate from sustainability or a fundamental part?
-\end{itemize}
+and feature requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
+
+Attendees suggested that a key question that the definition of sustainability is
+one issue on which the community needs to agree, and that ideally, an initial
+definition would be determined during the workshop, or at least some progress
+would be made towards this goal.  Another topic that came up is what the
+goal of sustainability is.  Perhaps it's reproducible science, or
+persistence, or quality, or something else.  Similarly, some attendees want
+to understand how success in sustainability is measured.  How does a set of
+developers know when they have actually achieved sustainable software?
+This led to a comment that sustainability should be addressed throughout the
+full software life cycle.
+
+Another topic that came up during the Developing and Maintaining Software
+panel is the relationship of sustainability to other software attributes.  Attendees
+asked: what is the relationship between sustainability and provenance? And:
+is usability separate from sustainability or a fundamental part of it?
+
+\note{rest of section will be turned from bullets into text...}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
bullet list of what we want to discuss in use-cases

@@
 
-Perhaps discuss sustained software use-cases and issues surrounding
-them. Following is a list of software projects discussed throughout the workshop:
+Following is a list of software projects discussed throughout the workshop:
 
@@
 
+In this section, we discuss the following:
+\begin{itemize}
+    \item Categorize the real-world software projects into workshop themes. This is based on the aspect of a software project that dominates and best fits one of the workshop themes.
+    \item Discuss how the theme has impacted software projects over its lifetime.
+    \item Discuss similarities and differences between these projects.
+    \item What is a takeaway for a new project? What are the lessons learned? How can the course of current and future development projects be altered to implement lessons learned.
+\end{itemize}
+
 \note{Dan: do we need this section?  I think there's some overlap
@@
   issues section?  Let's write it down here - then we can decide if it goes elsewhere}
+\note{Ketan: yes, may be this section will split and merge into other sections}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
more work in section 8 - defining sustainability

@@
 came up in many different parts of the workshop, specifically in one of the
-keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}
+keynotes, the Developing and Maintaining Software panel, and the Policy panel.
 
@@
 panel is the relationship of sustainability to other software attributes.  Attendees
-asked: what is the relationship between sustainability and provenance? And:
-is usability separate from sustainability or a fundamental part of it?
-
-\note{rest of section will be turned from bullets into text...}
-
-In the Policy panel,
-
-\begin{itemize}
-\item Chris Lenhardt's presentation of Venters
-  paper~\cite{Venters_WSSSPE} included:
-
-\begin{itemize}
-\item What does software sustainability mean
-It ``is a rather ambiguous concept...''
-\item Lack of an accepted definition gets in the way of integrating
-  the concept into software engineering
-\item Sustainability as a non-functional requirement
-\item Quality of software architectures determines sustainability
-\item Propose sustainability is a measure of central quality
-  attributes: Extensibility, Interoperability, Maintainability,
-  Portability, Reusability, Scalability
-\item Develop an architecture evaluation framework based on scenarios
-  which help to illuminate how to measure quality/sustainability at
-  the architectural level
-\end{itemize}
-
-\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
-  included:
-\begin{itemize}
-\item Clarifying the notion of sustainability
-\begin{itemize}
-\item Provide same functionality going forward even if environment changes
-\item Is functionality and usability clear
-\item Will functionality be correct going into the future even if
-  environment changes
-\item Is the functionality desired by current and future users
-\item Can it incorporate new things as they develop
-\end{itemize}
-\end{itemize}
-
-\item see slide 9 of
-  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
-
-\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
-
-\end{itemize}
+asked ``what is the relationship between sustainability and provenance?'' And,
+``is usability separate from sustainability or a fundamental part of it?''
+
+In addition, the Policy panel had a large amount of discussion about
+defining sustainability, as one of the subtopics in that panel was
+Modeling Sustainability, and modeling requires defining what will be modeled.
+
+In this subtopic, two papers included discussion of the definition of
+sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned
+that this is a rather ambiguous concept, and that the lack of an accepted
+definition gets in the way of integrating the concept into software engineering.
+They suggest that sustainability is a non-functional requirement, and that the
+quality of software architectures determines sustainability.  They then
+propose that sustainability is a measure of a set of central quality attributes:
+extensibility, interoperability, maintainability, portability, reusability, and scalability.
+Finally, they develop an architecture evaluation framework based on scenarios
+that helps to illuminate how to measure quality/sustainability at
+the architectural level.
+
+Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
+be used to measure software sustainability, and though these questions might falsely lead
+to yes or no answers, the complete set would determine a range of values for
+sustainability. These question are:
+Will the software provide same functionality going forward even if environment changes?
+Are the functionality and usability clear?
+Will functionality be correct going into the future even if environment changes?
+Is the functionality that which is desired by current and future users?
+Can the software incorporate new things as they develop?
+\todo{check paper for wording - might be better there than in this version, which comes from bullets}
+
+\todo{See slide 9 of \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table - incorporate/include table?}
+
+\todo{Also see discussion at:} \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

--- 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 58 submissions were received, and almost all submitters used either 
-arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
-their papers.
-
-A peer review process followed the submissions, where
-the 58 papers received 181 reviews, an average of 3.12 reviews per
-paper. Reviews were completed using a Google form, which allowed
-reviewers to choose papers they wanted to review, then to provide
-general comments and scores on relevance to the organizers and
-to the authors. The review reports enabled the organizers to decide
-which papers to associate with the workshop, and allowed the authors
-to improve their papers.
-
-The organizers decided to list 54 of the papers as significant
-contributions to the workshop, a very high acceptance rate, but one
-that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. The papers were
-grouped into three main categories, namely \emph{Developing and
-Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-subject area was associated with a panel and discussion at the
-workshop.
-
-The workshop itself consisted of two keynote presentations and the
-three panels/discussion sessions. The panels were organized based on a
-classification of the workshop submissions into three categories,
-following the themes of the call for papers as modified by the areas
-of the submissions. Three to four representatives from each submission
-category were appointed as panelists, and assigned to read a subset of
-the paper in that category and then discuss them in the panel.
-
-\section{Keynotes\note{lead: Sou-Cheng (Terrya) Choi}} \label{sec:keynotes}
-
-The WSSSPE1 workshop began with two keynote presentations.
+arXiv\footnote{\url{http://arxiv.org}} or 
+figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+
+A peer review process followed the submissions, where the 58 papers received 181 reviews, an 
+average of 3.12 reviews per paper. Reviews were completed using a Google form, which allowed 
+reviewers to provide scores on relevance and comments to the organizers and to the authors.
+The review reports enabled the organizers to decide which papers to associate with the workshop, and the authors to improve their papers.
+
+The organizers decided to list 54 of the papers as significantly contributing to the workshop, a very 
+high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact 
+that the reports were already self-published.
+The papers were also grouped into three main categories, namely \emph{Developing and Supporting Software}, \emph{Policy}, and \emph{Communities}.
+Each subject area was associated with a panel and discussion at the workshop.
+
+\section{Agenda}
+
+The workshop consisted of two keynote presentations and three panels/discussion sessions. The panels 
+were organized based on a classification of the workshop submissions into three categories, following 
+the themes of the call for papers as modified by the areas of the submissions. Four \todo{Is it five?} representatives from 
+each submission categories were appointed as panelists, to discuss the papers in that category.
+
+\section{Keynotes}
 
@@
 
-\todo{change references to papers or presentations to cites}
-
 \subsection{A recipe for sustainable software, Philip E. Bourne.}
 
-The first
-keynote~\cite{WSSSPE1-keynote1}
-was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
-companies. He also helped found PLOS and
-PDB~\cite{PDB}.  He is
-working on automating three-dimensional visualizations of cell
-contents and molecular structures, a problem that has not been solved
-and would serve as a key function of software in biomedical sciences.
-
-Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
-He emphasized that sustainability for software ``does not just
-mean more money from Government.''  He noted that in places,
-development in science has improved thanks to open source and Github,
-but for the most part remains arcane. He argued that there is much we
-can learn from the App Store model. He mentioned
-BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
-OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
-as distribution channels.  On a related note, Bourne observed a common
-evolutionary pathway for computational biology projects, from data
-archive to analytics platform to educational use, and suggested that
-use of scientific software for outreach might be the final step.
-
-Bourne shared with the audience a few real challenges he
-encountered. First, he has looked into reproducibility in
-computational biology, and has concluded that ``I have proved I cannot
-reproduce research from my own lab''~\cite{Veretnik}.
-
-Another problem he experienced was staff retention with respect to
-private organizations which reward those combining research and
-software expertise (the ``Google Bus''). However, he is a strong fan
-of private sector sustainability sources and public-private
-partnerships. He noted that making a successful business from
-scientific software alone is very rare: founders overvalue while
-customers undervalue. He noted that to last, an open source project
-needs a minimal funding requirement even with a vibrant community --
-good will only goes so far if you are being paid to do something else.
-He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology
-transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
-
-He also had problems with selling research software: the technology
-transfer office in his university wanted huge IP reachthrough, whereby
-they would get a share of profits from drugs developed by pharma
-companies who use the software. He was aware this was unrealistic but
-the technology transfer office insisted for a while. He wants to push
-a one-click approach to purchasing university-written software.
+The first keynote\footnote{\url{http://wssspe.researchcomputing.org.uk/keynotes/\#bourne}}
+was delivered by Philip E. Bourne of UCSD. The slides
+are online.\footnote{\url{www.slideshare.net/pebourne/a-recipe-for-sustainable-software}}
+
+Philip is a basic biomedical scientist but has formed four software
+companies. He also helped found PLOS and PDB\footnote{\url{http://www.rcsb.org/pdb/home/home.do}}.
+
+He emphasized that Sustainability for software ``does not just mean more
+money from Government.''
+
+He noted that creating effective illustrations is a key function of
+software in the biomedical sciences: ``Automating Conceptualization''
+(Visualizing molecular structures is a problem that has now been solved,
+he is working on 3-d visualizations of the contents of cells.)
+
+He has looked into reproducibility in computational biology, and has
+concluded that ``I have proved I cannot reproduce research from my own
+lab''~\cite{Veretnik}.
+
+He noted that in places, development in science has improved thanks to
+open source and Github, but for the most part remains arcane. He argues
+that there is a lot we can learn from the app-store model. He
+mentioned BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
+OpenScienceDataCloud \footnote{\url{https://www.opensciencedatacloud.org/}} as
+distribution channels.
+
+He noted a common evolutionary pathway for computational biology
+projects, from data archive to analytics platform to educational use. He
+suggests that use for outreach might be the final step.
+
+He experienced problems with staff retention with respect to private organizations which reward 
+those combining research and software expertise (the ``Google Bus''). However,
+he is a strong fan of private sector sustainability sources and PPP. He
+noted that making a successful business from scientific software alone
+is very rare: founders overvalue while customers undervalue. He noted
+that to last, an open source project needs a minimal funding requirement
+even with a vibrant community -- good will only goes so far if you are
+being paid to do something else.
+
+He talked about grant schemes of relevance in the U.S., particularly with regard to technology transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}} \todo{change at least first to cite}
+
+He had problems with selling research software: the technology transfer
+office in his university wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the
+software. He was aware this was unrealistic but the technology transfer
+office insisted for a
+while. He wants to push a one-click approach to purchasing
+university-written software.
 
@@
 
-He concluded with a reference to his paper~\cite{bourne_ten} and
-argued that computational scientists ``have a responsibility to
-convince their institutions and communities that software is
-scholarship,'' a point with which we strongly agree.
-
-\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}
-
-The second keynote~\cite{WSSSPE1-keynote2} was delivered by
- Arfon Smith of GitHub.
+He concluded with a reference to his paper~\cite{bourne_ten} and argued that computational scientists ``have a
+responsibility to convince their institutions and communities that
+software is scholarship,'' a point with which we strongly agree.
+
+\subsection{Scientific Software and the Open Collaborative Web}
+
+\todo{change references to papers or presentations to cites}
+
+The second keynote\footnote{\url{http://is.gd/wssspe}} was by
+ Arfon Smith\footnote{\url{http://arfon.org/}} of Github.
 
 His talk used an example from data reduction in Astronomy, where he
-needed to remove interfering effects from the device. He built a ``bad
-pixel mask'', and realized that while this is persistent, there was no
-way/practice of sharing this with the data, so many researchers
-repeated the same calculations. He estimated that 13 person-years were
-wasted in this redundant calculation.
+needed to remove interfering effects from the device. He built a
+``bad pixel mask'', and realized though while this is persistent, there
+was no way/practice of sharing this with the data, so many researchers
+repeated the same calculations. He estimated that 13 wasted person-years were wasted in
+this redundant calculation.
 
 He asked, of this practice, ``why didn't we know better?'' and argued
-this was because we were taught to focus on immediate research
-outcomes and not on continuously improving and building on tools for
-research. He then asked, later in our working lives, when we do know
-better, why don't we act any different, and argued this was due to
-incentives and their lack: only the immediate products of research,
-not the software, are valued.  He referenced Victoria Stodden's talk
-at
-OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+this was because we were taught to focus on immediate research outcomes
+and not on continuously improving and building on tools for research. He
+then asked, later in our working lives, when we do know better, why
+don't we act any different, and argued this was due to incentives and their lack:
+only the immediate products of research, not the software, are valued.
+He referenced Victoria Stodden's talk at OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
 which he said argued these points well.
 
-He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
-Chain Monte Carlo code by Dan Forman
-Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
-Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
-as good examples of scientific projects using GitHub.
-
-C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
-regard to reusable software, ``we should just start doing it.'' Arfon
-noted that documentation should be ``treated as a first class
-entity.''  He noted that some software fields have excellent cultures
-of module reuse, such as
-rubygems\footnote{\url{http://rubygems.org/}},
-PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
-CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov Chain
+Monte Carlo code by Dan Forman Mackey\footnote{\url{http://dan.iel.fm/}} and Joe Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}} as 
+good examples of  scientific projects using GitHub.
+
+C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with regard to
+reusable software, ``we should just start doing it.'' Arfon noted that
+documentation should be ``treated as a first class entity.''
+He noted that some software fields have excellent cultures of module
+reuse, such as rubygems\footnote{\url{http://rubygems.org/}},
+PyPI\footnote{\url{https://pypi.python.org/pypi}}, and CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
 ultra-low-friction collaboration through these repositories. This has
-not happened in highly numerical, compiled language scientific
-software.
+not happened in highly numerical, compiled language scientific software.
 
@@
 
+There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.
+
+\subsection{Communities Part 1}
+
+This collection of papers was summarized by Karen Cranston.
+
+\subsection{Communities Part 2}
+
+Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
+relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
+The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
+This is a 4-step iterative development process that incorporates Agile development principals.
+
+\begin{itemize}
+
+\item Step 1 - Design - thorough discussion of research questions
+\item Step 2 - Develop working code
+\item Step 3 - Refine based on new requirements
+\item Step 4 - Publish open source
+
+\end{itemize}
+
+The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
+code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
+build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
+Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
+In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
+Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
+Co‐locate rather than collaborating remotely whenever feasible.
+
+The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure. The paper opens with a description of the explosion of use of just a few gateways. 
+Over 3.5 years, more than 7,000 biologists have run phylogenetic codes on supercomputers using the CIPRES Science Gateway.
+In 1 year, over 120 scientists from 50 institutions used the UltraScan Science Gateway,
+increasing the sophistication of analytical ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG) registered 100+ 
+users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often make use of common components. Science Gateway Platform as a Service (Sci-GaP)
+delivers middleware as a hosted, scalable third ­party service while domain developers focus on user interfaces and domain-specific data in the gateway.
+The middleware handles things like authentication, application installation and reliable execution and help desk support.
+
+One key to SciGaP is its openness.
+While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
+The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
+stakeholder requirements. The project is leveraging Internet2’s Net+ mechanisms for converting SciGaP and its gateways into commodity services.
+
+The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as 
+a “body of persons of common and especially professional interests scattered through a larger society.” This makes support
+challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
+whether they are indirectly alienating part of the community or encouraging particular types of use.
+
+nanoHUB's 10-year history with over 260,000 users annually provides a lot of data to analyze. 4000 resources contributed by
+1000 authors. nanoHUB serves both the research and education community and the contribution model allows researchers to get
+their codes out into the community and in use in education very rapidly. The primary software challenges are twofold - support
+for the HUBzero framework and challenges related to the software contributed by the community.
+
+The group has learned that community contributions are maximized with a tolerant licensing approach. HUBzero uses an 
+LGPLv3 license so contributors can create unique components and license as they choose. If they make changes to source 
+code, the original license must be maintained for redistribution. As far as contributed resources, these must be open access,
+but not necessarily open source. This allows contributors to meet the requirements of their institutions and funding agencies.
+Quality is maintained via user ratings. Documentation is encouraged and nanoHUB supplies regression test capabilities, but 
+the user community provides ratings, poses questions and contributes to wishlists and citation counts - all of which incentivize
+code authors.
+
+The Terrel paper describes support for the Python scientific community through two major efforts - the SciPy conference and the
+NumFOCUS foundation. Reliance on software in science has driven a huge demand for development, but this development is typically done as a side effort
+and often in a rush to publish without documentation and testing.
+While created by academics, software support often falls to industrial institutions. SciPy brings together industry, government, 
+and academics to share their code and experience in an open environment.
+
+NumFOCUS promotes open, usable scientific software while sustaining the community. Specific activities include educational programs, 
+collaborative research tools and documentation and 
+promotion of high­-level languages, reproducible scientific research, and open­code developmet.
+Governance of the non-profit is a loose grantor-­grantee relationship with projects allowing for monies to be placed in the groups accounts.
+This has raised money to hire developers for open code development, maintain testing machines, organize the PyData conference series, and sponsor community members to attend conferences.
+
+Software sustainability relies on contributions from all sectors of user community.
+Together SciPy and NumFOCUS support these sectors.
+By maximizing contributions growing the user base they help develop and mature Python.
+
+The Loffler paper describes the Cactus project which was started in 1996 by participants in USA Binary Black Hole Alliance 
+Challenge. Cactus has a flesh and thorns model - a community-oriented framework that allows researchers to easily work 
+together with reusable and extendable software elements. Modules are compiled into an executable and can remain dormant, becoming
+active only when parameters and simulation data dictate. Users can  use modules written by others or can write their own 
+modules without changing other code. The community has grown and diversified beyond the original science areas.
+
+The paper points out 4 keys to sustaining the community - modular design, growing a collaborative community, career paths and credit.
+In modular design, the Cactus project went far beyond standard practices of APIs. 
+Domain specific languages (DSLs) allow decoupling of components - for example I/O, mesh refinement, PAPI counters, and
+boundary conditions abstracted from science code. In academia, publications are the main currency of credit. Because the 
+project connects code developments to science, the work is 
+publishable and modules are citable. Because of the open source, modular approach, programmers can see the impact of their contributions and often continue
+work after graduation. Career paths remain a challenge, however. Tasks that are essential from a software engineering perspective
+are often not rewarded in academia. The best programmers in a science environment often have multidisciplinary expertise.
+That also is not rewarded in academia.
+
+The Wilkins-Diehr paper
+
 \subsection{Papers}
 
-\subsubsection*{Communities}
-
-\begin{itemize}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
-  Software through the Open Community Engagement Process
-  \cite{Christopherson_WSSSPE}
-
-\item Reagan Moore. Extensible Generic Data Management Software
-  \cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability
-  \cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities \cite{Vay_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
-  Borries Demeler. Science Gateway Operational Sustainability:
-  Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community \cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
-  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
-  Mookherjee. Reusability in Science: From Initial User Engagement to
-  Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
-  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
-  Dooley, Dan Stanzione. Science Gateways and the Importance of
-  Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
-  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
-  tools for facilitating Open Science \cite{Hart_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area was also discussed
-in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+\subsubsection*{Communities Part 1 Papers}
+
+\begin{itemize}
+
+\item Reagan Moore. Extensible Generic Data Management Software \cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A grassroots approach to software sustainability \cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman, D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator Modeling Science Activities \cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org: Experiences and Challenges in Software Sustainability for a Large Scientific Community \cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community \cite{Terrel_WSSSPE}
+
+\item Frank Löffler, Steven R. Brandt, Gabrielle Allen and Erik Schnetter. Cactus: Issues for Sustainable Simulation Software \cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary, Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software Ecosystems for Open Science \cite{Hanwell_WSSSPE}
 
@@
 
-\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
-
-\todo{perhaps remove this section}
-
-\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
-
-\note{lead: Daniel S. Katz}
-
-The question of what was meant by ``sustainability'' in the context of software
-came up in many different parts of the workshop, specifically in one of the
-keynotes, the Developing and Maintaining Software panel, and the Policy panel. \todo{check closing discussion notes for this topic}
-
-In his keynote, Phil Bourne suggested that perhaps sustainability can be defined
-as the effort that happens to make the essential things continue. This leads to
-having to decide what it is that we want to sustain, if what we want to sustain is
-valuable, and finally, who would care if it went away, and how does one measure
-how much they care.
-
-In the Developing and Maintaining Software panel, there was some discussion
-on this topic: what does sustainability mean? It was pointed out that
-OSS Watch proposes a Software Sustainability Maturity Model to address
-the issue of how sustainability a particular element of software is, and says
-that this sustainability is important. ``When choosing
-software for procurement or development reuse - regardless of the
-license and development model you will use - you need to consider
-the future. While a software product may satisfy today's needs, will
-it satisfy tomorrow's needs? Will the supplier still be around in
-five years' time? Will the supplier still care for all its customers
-in five years' time? Will the supplier be responsive to bug reports
-and feature requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
-
-Attendees suggested that a key question that the definition of sustainability is
-one issue on which the community needs to agree, and that ideally, an initial
-definition would be determined during the workshop, or at least some progress
-would be made towards this goal.  Another topic that came up is what the
-goal of sustainability is.  Perhaps it's reproducible science, or
-persistence, or quality, or something else.  Similarly, some attendees want
-to understand how success in sustainability is measured.  How does a set of
-developers know when they have actually achieved sustainable software?
-This led to a comment that sustainability should be addressed throughout the
-full software life cycle.
-
-Another topic that came up during the Developing and Maintaining Software
-panel is the relationship of sustainability to other software attributes.  Attendees
-asked: what is the relationship between sustainability and provenance? And:
-is usability separate from sustainability or a fundamental part of it?
-
-\note{rest of section will be turned from bullets into text...}
-
-In the Policy panel,
-
-\begin{itemize}
-\item Chris Lenhardt's presentation of Venters
-  paper~\cite{Venters_WSSSPE} included:
-
-\begin{itemize}
-\item What does software sustainability mean
-It ``is a rather ambiguous concept...''
-\item Lack of an accepted definition gets in the way of integrating
-  the concept into software engineering
-\item Sustainability as a non-functional requirement
-\item Quality of software architectures determines sustainability
-\item Propose sustainability is a measure of central quality
-  attributes: Extensibility, Interoperability, Maintainability,
-  Portability, Reusability, Scalability
-\item Develop an architecture evaluation framework based on scenarios
-  which help to illuminate how to measure quality/sustainability at
-  the architectural level
-\end{itemize}
-
-\item Chris Lenhardt's presentation of Katz paper~\cite{Katz_WSSSPE}
-  included:
-\begin{itemize}
-\item Clarifying the notion of sustainability
-\begin{itemize}
-\item Provide same functionality going forward even if environment changes
-\item Is functionality and usability clear
-\item Will functionality be correct going into the future even if
-  environment changes
-\item Is the functionality desired by current and future users
-\item Can it incorporate new things as they develop
-\end{itemize}
-\end{itemize}
-
-\item see slide 9 of
-  \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table
-
-\item also see discussion at \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
-
-\end{itemize}
-
-\section{Illustrative Use Cases} \label{sec:use-cases}
-
-\note{lead: Ketan Maheshwari, contributors: Andy Terrel}
-
-Perhaps discuss sustained software use-cases and issues surrounding
-them. Following is a list of software projects discussed throughout the workshop:
-
-\begin{itemize}
-    \item github
-    \item mvapich
-    \item Cactus
-    \item visTrails
-    \item MINRES/QLP
-    \item SimGrid
-    \item Swift
-    \item Galaxy
-    \item FastMath and package managers
-    \item R/qtl
-    \item OWL/RDF
-    \item AMR/multiphysics
-    \item DUNE
-    \item PAAS
-    \item Gateways
-    \item Globus-Genomics
-\end{itemize}
-
-\note{Dan: do we need this section?  I think there's some overlap
-  between this and the discussion in the Developing and Support
-  Software section.  Or perhaps this should go in the cross-cutting
-  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
-
-\section{Conclusions} \label{sec:conclusions}
+\section{Cross-cutting Issues}
+
+\subsection{Defining sustainability}
+
+\subsection{Licensing}
+
+\note{not sure if this was really cross-cutting...}
+
+\section{Illustrative use-cases}
+
+Perhaps discuss sustained software use-cases and issues surrounding them. Some examples could be mvapich, Galaxy, Swift, Cactus, etc.
+
+\section{Conclusions}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
more work is sustainability section - still in progress

--- 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 sustainability. These question are:
-Will the software provide same functionality going forward even if environment changes?
-Are the functionality and usability clear?
-Will functionality be correct going into the future even if environment changes?
-Is the functionality that which is desired by current and future users?
-Can the software incorporate new things as they develop?
-\todo{check paper for wording - might be better there than in this version, which comes from bullets}
-
-\todo{See slide 9 of \url{http://dx.doi.org/10.6084/m9.figshare.853817} for summary table - incorporate/include table?}
+\begin{itemize}
+\item Will the software continue to provide the same functionality in the future, even if the environment (other parts of the infrastructure) changes?
+\item Is the functionality and usability clearly explained to new users? Do users have a mechanism to ask questions and to learn about the element?
+\item Will the functionality be correct in the future, even if the environment changes?
+\item Does it provide the functionality that existing and future users want?
+\item Does it incorporate new science/theory/tools as they develop?
+\end{itemize}
+
+Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling Sustainability papers as in Table~\ref{tab:defining-sustainability}. \todo{discussion of this?}
+
+\begin{table}[h]
+  \begin{scriptsize}
+    \begin{center}
+      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}}
+      \label{tab:defining-sustainability}
+      \begin{tabular}{|p{2.3cm}|p{3.6cm}|p{4.4cm}|p{4.8cm}|}
+		\hline
+{\bf Authors}
+& {\bf Software}
+& {\bf Sustainability}
+& {\bf Approach to Understand or Evaluate Sustainability} \\
+		\hline
+Calero, et al. \cite{Calero_WSSSPE}
+& General notion of software. Not explicitly defined.
+& Sustainability is linked to quality.
+& Add to ISO \\
+		\hline
+Venters, et al. \cite{Venters_WSSSPE}
+& Software as science software; increasingly complex; service-oriented computing
+& Extensibility, interoperability, maintainability, portability, reusability, scalability, efficiency
+& Use various architecture evaluation approaches to assess sustainability \\
+		\hline
+Pierce, et al. \cite{Pierce_WSSSPE}
+& Cyberinfrastructure software
+& Sustainable to the extent to which there is a community to support it
+& Open community governance \\
+		\hline
+Katz and Proctor \cite{Katz_WSSSPE}
+& E-research infrastructures (i.e. cyberinfrastructure)
+& Persisting over time, meeting original needs and projected needs
+& Equates models for the creation of software with sustaining software \\
+		\hline
+Lenhardt, et al. \cite{Lenhardt_WSSSPE}
+& Broadly defined as software supporting science
+& Re-use; reproducible science
+& Comparing data management life cycle to software development life cycle \\
+		\hline
+Weber, et al. \cite{Weber_WSSSPE}
+& Software broadly defined; a software ecosystem
+& Software niches
+& Ecological analysis and ecosystem \\
+		\hline
+     \end{tabular}
+    \end{center}
+  \end{scriptsize}
+\end{table}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing order of attendees in a way that should be easy to add new people later

@@
 \usepackage{enumitem}
+\usepackage{multicol}
 
@@
 
-\note{should alphabetize all, but if we get more from the SC13 folks, it's not worth spending
-time doing this now.  I wonder if there's a way to just have a simple list that we can keep ordered, and let LaTeX do the formatting, rather than manually creating a table as this is now}
-
 The following is a partial list of attendees who were recorded on the
-Google doc that was being used for live note taking at the workshop:
-
-\begin{tabular}{ L{4cm}  L{4cm}  L{4cm}  L{4cm} }
+Google doc~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, with some additions made by the authors of this paper.
+
+\begin{multicols}{4}
+\setlength{\parindent}{0pt}
+%in theory, should save the old value then set it back after this section, but since this is the end...
+
+Jay Alameda
+
+Gabrielle Allen
+
+David Andrs
+
+Lorena A Barba
+
+David Bernholdt
+
+Phil Bourne
+
+Karl Broman
+
+Maxine Brown
+
+David Bruhwiler
+
+Bruno Bzeznik
+
+Jeffrey Carver
+
+Neil Chue Hong
+
+John W. Cobb
+
+Karen Cranston
+
+Rion Dooley
+
+Anshu Dubey
+
+Ian Foster
+
+Juliana Freire
+
+Derek Gaston
+
+Derek Groen
+
+Edmund Hart
+
+James Hetherington
+
+Simon Hettrick
+
+Kenneth Hoste
+
+James Howison
+
+Daisie Huang
+
 Daniel S. Katz
-&  Neil Chue Hong
-&  Gabrielle Allen
-&  David Proctor
-\\ Manish Parashar
-&  Jay Alameda
-&  David Andrs
-&  Lorena A Barba
-\\ David Bernholdt
-&  Phil Bourne
-&  Karl Broman
-&  Maxine Brown
-\\ David Bruhwiler
-&  Bruno Bzeznik
-&  Jeffrey Carver
-&  John W. Cobb
-\\ Karen Cranston
-&  Rion Dooley
-&  Anshu Dubey
-&  Ian Foster
-\\Juliana Freire
-& Derek Gaston
-& Derek Groen
-& Edmund Hart
-\\James Hetherington
-& Simon Hettrick
-& Kenneth Hoste
-& James Howison
-\\Daisie Huang
-& Kerk Kee
-& Arnaud Legrand
-& Paul Madden
-\\Louis Maddox
-& Ketan Maheshwari
-& Suresh Marru
-& Cezary Mazurek
-\\James McClure
-& Chris Mentzel
-& J. Yates Monteith
-& Pierre Neyron
-\\Lucas Nussbaum
-& Cody Permann
-& John Peterson
-& Quan Pham
-\\Marlon Pierce
-& Heather Piwowar
-& Jason Riedy
-& Andreas Schreiber
-\\William Scullin
-& Andrew Slaughter
-& Arfon Smith
-& James Spencer
-\\Shel Swenson
-& Fr\'{e}d\'{e}ric Suter
-& Andy Terrel
-& George  Thiruvathukal
-\\John Towns
-& Erik Trainer
-& Matthew Turk
-& Eric Van Wyk
-\\Colin C. Venters
-& Von Welch
-& Nancy Wilkins-Diehr
-& Theresa Windus
-\\Rich Wolski
-& Lynn Zentner
-& Frank L\"{o}ffler
-& Patrick O'Leary
-\end{tabular}
+
+Kerk Kee
+
+Arnaud Legrand
+
+Frank L\"{o}ffler
+
+Paul Madden
+
+Louis Maddox
+
+Ketan Maheshwari
+
+Suresh Marru
+
+Cezary Mazurek
+
+James McClure
+
+Chris Mentzel
+
+J. Yates Monteith
+
+Pierre Neyron
+
+Lucas Nussbaum
+
+Patrick O'Leary
+
+Manish Parashar
+
+Cody Permann
+
+John Peterson
+
+Quan Pham
+
+Marlon Pierce
+
+Heather Piwowar
+
+David Proctor
+
+Jason Riedy
+
+Andreas Schreiber
+
+William Scullin
+
+Andrew Slaughter
+
+Arfon Smith
+
+James Spencer
+
+Shel Swenson
+
+Fr\'{e}d\'{e}ric Suter
+
+Andy Terrel
+
+George  Thiruvathukal
+
+John Towns
+
+Erik Trainer
+
+Matthew Turk
+
+Eric Van Wyk
+
+Colin C. Venters
+
+Von Welch
+
+Nancy Wilkins-Diehr
+
+Theresa Windus
+
+Rich Wolski
+
+Lynn Zentner
+
+\end{multicols}
 

### 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

--- 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-The Wilkins-Diehr paper
+The Wilkins-Diehr paper describes an NSF software institute effort to build a community of those creating science gateways or
+web portals for science. These gateways, as described in some of the other papers in this section, can be quite capable, having a 
+remarkable impact on science.
+
+This paper mentioned challenges highlighted by other papers in this section - mainly the conflict between funding for research
+vs infrastructure and the challenges around getting academic credit for infrastructure. Because the authors have studied
+many projects, they've also observed how development is often done in an isolated hobbyist environment. Developers are unable
+to take advantage of similar work done by others. isolation even when projects have common needs.
+But often projects struggle for sustainable funding because they provide infrastructure to conduct
+research and many times only the research is funded. Gateways also may start as small group research project, taking off in popularity
+once they go live without any long term plans for sustainability or without resources in the project to plan for such.
+Subsequent disruptions in service can limit effectiveness and test the limits of the research community's trust. The impact
+of gateways can be increased substantially if we understand what makes them successful.
+
+Recommendations from an early study of successful gateways include the following. Leadership and management teams should 
+design governance to represent multiple strengths and perspectives, plan for change and turnover in the future, 
+recruit a development team that understands both the technical and domain-related issues, consider sustainability and measure success early and often.
+Successful projects recognize the benefits and costs of hiring a team of professionals, demonstrate credibility through stability and 
+clarity of purpose, leverage the work of others and plan for flexibility.
+Successful projects identify an existing community and understand that communities' needs before beginning. Projects adapt
+as the communities' needs evolve. For funders, the lifecycle of technology projects must be considered. Solicitations should
+be designed to reward effective planning, recognize the benefits and limitations of both technology innovation and reuse,
+expect adjustments during the production process, copy effective models from other industries and sectors, and encourage partnerships that support gateway sustainability.
+
+Through a business incubator type approach, the institute plans to provide a variety of services that could be shared amongst
+projects. Consultation and resources on topics such as business plan development, software engineering practices, 
+software licensing options, usability, security and project management as well as a software repository and hosting 
+service will be available. Experts will be available for multi-month assignments to help research teams build their own gateways,
+teaching them how to maintain and add to the work after the collaboration ends. Forums, symposia and an annual conference
+will connect members of the development community.A modular, layered framework that supports community contributions and 
+allows developers to choose components will be delivered and finally 
+workforce development activities will help train the next generation for careers in this cross- disciplinary area and 
+build pools of institutional expertise that many projects can leverage. Shared services and forums can dramatically reduce the cost
+of building and sustaining gateways. Workforce development can encourage technology professionals to remain in the sciences.
 

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex

--- 2014-02-06 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
+There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.
+
+\subsection{Communities Part 1}
+
+This collection of papers was summarized by Karen Cranston.
+
+\subsection{Communities Part 2}
+
+Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
+relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
+The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
+This is a 4-step iterative development process that incorporates Agile development principals.
+
+\begin{itemize}
+
+\item Step 1 - Design - thorough discussion of research questions
+\item Step 2 - Develop working code
+\item Step 3 - Refine based on new requirements
+\item Step 4 - Publish open source
+
+\end{itemize}
+
+The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of 
+code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually 
+build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
+Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons. 
+In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
+Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
+Co-locate rather than collaborating remotely whenever feasible.
+
+The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure. The paper opens with a description of the explosion of use of just a few gateways. 
+Over 3.5 years, more than 7,000 biologists have run phylogenetic codes on supercomputers using the CIPRES Science Gateway.
+In 1 year, over 120 scientists from 50 institutions used the UltraScan Science Gateway,
+increasing the sophistication of analytical ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG) registered 100+ 
+users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often make use of common components. Science Gateway Platform as a Service (Sci-GaP)
+delivers middleware as a hosted, scalable third-party service while domain developers focus on user interfaces and domain-specific data in the gateway.
+The middleware handles things like authentication, application installation and reliable execution and help desk support.
+
+One key to SciGaP is its openness.
+While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
+The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
+stakeholder requirements. The project is leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways into commodity services.
+
+The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as 
+a ``body of persons of common and especially professional interests scattered through a larger society.'' This makes support
+challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
+whether they are indirectly alienating part of the community or encouraging particular types of use.
+
+nanoHUB's 10-year history with over 260,000 users annually provides a lot of data to analyze. 4000 resources contributed by
+1000 authors. nanoHUB serves both the research and education community and the contribution model allows researchers to get
+their codes out into the community and in use in education very rapidly. The primary software challenges are twofold - support
+for the HUBzero framework and challenges related to the software contributed by the community.
+
+The group has learned that community contributions are maximized with a tolerant licensing approach. HUBzero uses an 
+LGPLv3 license so contributors can create unique components and license as they choose. If they make changes to source 
+code, the original license must be maintained for redistribution. As far as contributed resources, these must be open access,
+but not necessarily open source. This allows contributors to meet the requirements of their institutions and funding agencies.
+Quality is maintained via user ratings. Documentation is encouraged and nanoHUB supplies regression test capabilities, but 
+the user community provides ratings, poses questions and contributes to wishlists and citation counts - all of which incentivize
+code authors.
+
+The Terrel paper describes support for the Python scientific community through two major efforts - the SciPy conference and the
+NumFOCUS foundation. Reliance on software in science has driven a huge demand for development, but this development is typically done as a side effort
+and often in a rush to publish without documentation and testing.
+While created by academics, software support often falls to industrial institutions. SciPy brings together industry, government, 
+and academics to share their code and experience in an open environment.
+
+NumFOCUS promotes open, usable scientific software while sustaining the community. Specific activities include educational programs, 
+collaborative research tools and documentation and 
+promotion of high-level languages, reproducible scientific research, and open-code development.
+Governance of the non-profit is a loose grantor-grantee relationship with projects allowing for monies to be placed in the groups accounts.
+This has raised money to hire developers for open code development, maintain testing machines, organize the PyData conference series, and sponsor community members to attend conferences.
+
+Software sustainability relies on contributions from all sectors of user community.
+Together SciPy and NumFOCUS support these sectors.
+By maximizing contributions growing the user base they help develop and mature Python.
+
+The L\"{o}ffler paper describes the Cactus project which was started in 1996 by participants in USA Binary Black Hole Alliance 
+Challenge. Cactus has a flesh and thorns model - a community-oriented framework that allows researchers to easily work 
+together with reusable and extendable software elements. Modules are compiled into an executable and can remain dormant, becoming
+active only when parameters and simulation data dictate. Users can  use modules written by others or can write their own 
+modules without changing other code. The community has grown and diversified beyond the original science areas.
+
+The paper points out 4 keys to sustaining the community - modular design, growing a collaborative community, career paths and credit.
+In modular design, the Cactus project went far beyond standard practices of APIs. 
+Domain specific languages (DSLs) allow decoupling of components - for example I/O, mesh refinement, PAPI counters, and
+boundary conditions abstracted from science code. In academia, publications are the main currency of credit. Because the 
+project connects code developments to science, the work is 
+publishable and modules are citable. Because of the open source, modular approach, programmers can see the impact of their contributions and often continue
+work after graduation. Career paths remain a challenge, however. Tasks that are essential from a software engineering perspective
+are often not rewarded in academia. The best programmers in a science environment often have multidisciplinary expertise.
+That also is not rewarded in academia.
+
+The Wilkins-Diehr paper describes an NSF software institute effort to build a community of those creating science gateways or
+web portals for science. These gateways, as described in some of the other papers in this section, can be quite capable, having a 
+remarkable impact on science.
+
+This paper mentioned challenges highlighted by other papers in this section - mainly the conflict between funding for research
+vs infrastructure and the challenges around getting academic credit for infrastructure. Because the authors have studied
+many projects, they've also observed how development is often done in an isolated hobbyist environment. Developers are unable
+to take advantage of similar work done by others. isolation even when projects have common needs.
+But often projects struggle for sustainable funding because they provide infrastructure to conduct
+research and many times only the research is funded. Gateways also may start as small group research project, taking off in popularity
+once they go live without any long term plans for sustainability or without resources in the project to plan for such.
+Subsequent disruptions in service can limit effectiveness and test the limits of the research community's trust. The impact
+of gateways can be increased substantially if we understand what makes them successful.
+
+Recommendations from an early study of successful gateways include the following. Leadership and management teams should 
+design governance to represent multiple strengths and perspectives, plan for change and turnover in the future, 
+recruit a development team that understands both the technical and domain-related issues, consider sustainability and measure success early and often.
+Successful projects recognize the benefits and costs of hiring a team of professionals, demonstrate credibility through stability and 
+clarity of purpose, leverage the work of others and plan for flexibility.
+Successful projects identify an existing community and understand that communities' needs before beginning. Projects adapt
+as the communities' needs evolve. For funders, the lifecycle of technology projects must be considered. Solicitations should
+be designed to reward effective planning, recognize the benefits and limitations of both technology innovation and reuse,
+expect adjustments during the production process, copy effective models from other industries and sectors, and encourage partnerships that support gateway sustainability.
+
+Through a business incubator type approach, the institute plans to provide a variety of services that could be shared amongst
+projects. Consultation and resources on topics such as business plan development, software engineering practices, 
+software licensing options, usability, security and project management as well as a software repository and hosting 
+service will be available. Experts will be available for multi-month assignments to help research teams build their own gateways,
+teaching them how to maintain and add to the work after the collaboration ends. Forums, symposia and an annual conference
+will connect members of the development community. A modular, layered framework that supports community contributions and 
+allows developers to choose components will be delivered and finally 
+workforce development activities will help train the next generation for careers in this cross- disciplinary area and 
+build pools of institutional expertise that many projects can leverage. Shared services and forums can dramatically reduce the cost
+of building and sustaining gateways. Workforce development can encourage technology professionals to remain in the sciences.
+
 \subsection{Papers}
 
-\subsubsection*{Communities}
-
-\begin{itemize}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
-  Software through the Open Community Engagement Process
-  \cite{Christopherson_WSSSPE}
+\subsubsection*{Communities Part 1 Papers}
+
+\begin{itemize}
 
@@
 
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
-  Borries Demeler. Science Gateway Operational Sustainability:
-  Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
 
@@
 
-\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
-  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
-  Mookherjee. Reusability in Science: From Initial User Engagement to
-  Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
-  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
-  Dooley, Dan Stanzione. Science Gateways and the Importance of
-  Sustainability \cite{Wilkins-Diehr_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
-  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
-  tools for facilitating Open Science \cite{Hart_WSSSPE}
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
 

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
finished first draft of cross-cutting section

@@
 
-In his keynote, Phil Bourne suggested that perhaps sustainability can be defined
+In the opening keynote, Phil Bourne suggested that perhaps sustainability can be defined
 as the effort that happens to make the essential things continue. This leads to
@@
 
-Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling Sustainability papers as in Table~\ref{tab:defining-sustainability}. \todo{discussion of this?}
+Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling
+Sustainability papers in the panel.
+As shown in Table~\ref{tab:defining-sustainability}, for each paper he
+discussed what was meant by software, if there was a definition of sustainability, and
+what the approach was to either understand or evaluate sustainability.
 
@@
 		\hline
-{\bf Authors}
+{\bf Paper/Authors}
 & {\bf Software}
@@
 
-\todo{Also see discussion at:} \url{https://docs.google.com/document/d/1eVfioGNlihXG_1Y8BgdCI6tXZKrybZgz5XuQHjT1oKU/edit?pli=1#bookmark=id.2pxw5yey8jl0}
+Finally, in the workshop's closing session, one of the discussion topics was 
+what success would look like for the set of WSSSPE activities beyond just the workshop.
+One of the answers that was suggested was
+having an agreed version of what we mean by sustainability.
 

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing typo, mainly for commit message testing
small change to test final commits
Section 3 Keynote: turn URLs into a references

@@
 
-\todo{change references to papers or presentations to cites}
-
 \subsection{A recipe for sustainable software, Philip E. Bourne.}
@@
 companies. He also helped found PLOS and
-PDB~\cite{PDB}.  He is
+PDB~\cite{pdb-web}.  He is
 working on automating three-dimensional visualizations of cell
@@
 can learn from the App Store model. He mentioned
-BioJava\footnote{\url{http://biojava.org/wiki/Main_Page}} and
-OpenScienceDataCloud\footnote{\url{https://www.opensciencedatacloud.org/}}
+BioJava~\cite{biojava-web} and
+OpenScienceDataCloud~\cite{osdc-web}
 as distribution channels.  On a related note, Bourne observed a common
@@
 He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology
-transfer.\footnote{\url{http://www.nsf.gov/pubs/policydocs/pappguide/nsf09_1/gpg_2.jsp\#IID2}} \footnote{\url{http://www.nsf.gov/eng/iip/sbir/}}
+with regard to technology transfer.~\cite{sbir-web, fased-web}.
 
@@
 research output alongside papers. Readers will be familiar with this
-debate. An interesting reference here is ``What
-does peer review mean when applied to computer code?, Rosemary Dickin.''
-\footnote{\url{http://blogs.plos.org/biologue/2013/08/08/what-does-peer-review-mean-when-applied-to-computer-code/}}
+debate. An interesting reference here is~\cite{peer-review-code}.
 
@@
 not the software, are valued.  He referenced Victoria Stodden's talk
-at
-OKCon\footnote{\url{http://okcon.org/main-speakers/victoria-stodden/}},\footnote{\url{http://www.stanford.edu/~vcs/talks/OKcon2013-Sept172013-STODDEN.pdf}}
+at OKCon~\cite{okcon-stodden-talk}
 which he said argued these points well.
 
-He cited the EMCEE\footnote{\url{https://github.com/dfm/emcee}} Markov
-Chain Monte Carlo code by Dan Forman
-Mackey\footnote{\url{http://dan.iel.fm/}} and Joe
-Zuntz\footnote{\url{http://www.jb.man.ac.uk/~zuntz/Joe_Zuntz/Welcome.html}}
+He cited the EMCEE Markov
+Chain Monte Carlo code~\cite{emcee-web} by Dan Foreman-Mackey and contributors
 as good examples of scientific projects using GitHub.
 
-C. Titus Brown\footnote{\url{http://ged.msu.edu/}} argues that with
+C. Titus Brown~\cite{ged-web} argues that with
 regard to reusable software, ``we should just start doing it.'' Arfon
@@
 of module reuse, such as
-rubygems\footnote{\url{http://rubygems.org/}},
-PyPI\footnote{\url{https://pypi.python.org/pypi}}, and
-CPAN\footnote{\url{http://www.cpan.org/}}, where there is effectively
+RubyGems~\cite{rubygems-web},
+PyPI~\cite{pypi-web}, and
+CPAN~\cite{cpan-web}, where there is effectively
 ultra-low-friction collaboration through these repositories. This has

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3 Keynote:  Minor changes.

@@
 contents and molecular structures, a problem that has not been solved
-and would serve as a key function of software in biomedical sciences.
-
-Bourne's keynote presentation is entiled ``A Recipe for Sustainable Software''.
+and when done, would serve as a key function of software in biomedical sciences.
+
+Bourne's keynote presentation is entitled ``A Recipe for Sustainable Software''.
 He emphasized that sustainability for software ``does not just
@@
 He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology transfer.~\cite{sbir-web, fased-web}.
+with regard to technology transfer~\cite{sbir-web, fased-web}.
 
@@
 needed to remove interfering effects from the device. He built a ``bad
-pixel mask'', and realized that while this is persistent, there was no
+pixel mask'', and realized that while it was persistent, there was no
 way/practice of sharing this with the data, so many researchers
@@
 
-C. Titus Brown~\cite{ged-web} argues that with
-regard to reusable software, ``we should just start doing it.'' Arfon
+C. Titus Brown~\cite{ged-web} argued that with
+regard to reusable software, ``we should just start doing it.'' Smith
 noted that documentation should be ``treated as a first class
@@
 
-Fernando Perez says that ``Open Source is reproducible by necessity''
+Fernando Perez emphasize that ``open source is reproducible by necessity''
 - since reproducibility is a prerequisite for remote collaboration.
-Arfon therefore defined GitHub as part of the next stage in
-development of the web, following on from the ``social web'' of
-Facebook et.al.: ``The Collaborative Web.''
+Smith therefore defined GitHub as part of the next stage in web
+development, following on from the social web of
+Facebook et al.: ``The Collaborative Web.''
 

### 2014-02-07 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3 Keynote:  Minor changes and introduce line breaks.

@@
 
-The first
-keynote~\cite{WSSSPE1-keynote1}
-was delivered by Philip E. Bourne of University of California, San Diego.  Bourne is a basic biomedical scientist but has formed four software
-companies. He also helped found PLOS and
-PDB~\cite{pdb-web}.  He is
-working on automating three-dimensional visualizations of cell
-contents and molecular structures, a problem that has not been solved
-and when done, would serve as a key function of software in biomedical sciences.
-
-Bourne's keynote presentation is entitled ``A Recipe for Sustainable Software''.
-He emphasized that sustainability for software ``does not just
-mean more money from Government.''  He noted that in places,
+The first keynote~\cite{WSSSPE1-keynote1} was delivered by Philip
+E. Bourne of University of California, San Diego.  Bourne is a basic
+biomedical scientist but has formed four software companies. He also
+helped found PLOS and PDB~\cite{pdb-web}.  He is working on automating
+three-dimensional visualizations of cell contents and molecular
+structures, a problem that has not been solved and when done, would
+serve as a key function of software in biomedical sciences.
+
+Bourne's keynote presentation is entitled ``A Recipe for Sustainable
+Software''.  He emphasized that sustainability for software ``does not
+just mean more money from Government.''  He noted that in places,
 development in science has improved thanks to open source and Github,
@@
 can learn from the App Store model. He mentioned
-BioJava~\cite{biojava-web} and
-OpenScienceDataCloud~\cite{osdc-web}
-as distribution channels.  On a related note, Bourne observed a common
+BioJava~\cite{biojava-web} and OpenScienceDataCloud~\cite{osdc-web} as
+distribution channels.  On a related note, Bourne observed a common
 evolutionary pathway for computational biology projects, from data
@@
 
-He then presented the usual arguments on directly valuing software as a
-research output alongside papers. Readers will be familiar with this
+He then presented the usual arguments on directly valuing software as
+a research output alongside papers. Readers will be familiar with this
 debate. An interesting reference here is~\cite{peer-review-code}.
@@
 
-The second keynote~\cite{WSSSPE1-keynote2} was delivered by
- Arfon Smith of GitHub.
+The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
+Smith of GitHub.
 
@@
 
-He cited the EMCEE Markov
-Chain Monte Carlo code~\cite{emcee-web} by Dan Foreman-Mackey and contributors
-as good examples of scientific projects using GitHub.
-
-C. Titus Brown~\cite{ged-web} argued that with
-regard to reusable software, ``we should just start doing it.'' Smith
-noted that documentation should be ``treated as a first class
-entity.''  He noted that some software fields have excellent cultures
-of module reuse, such as
-RubyGems~\cite{rubygems-web},
-PyPI~\cite{pypi-web}, and
-CPAN~\cite{cpan-web}, where there is effectively
-ultra-low-friction collaboration through these repositories. This has
-not happened in highly numerical, compiled language scientific
-software.
-
-He argued that the GitHub Pull Request code review mechanism facilitates
-such collaboration, by allowing one to code first, and seek review and a
-merge back into the trunk later.
-
-Fernando Perez emphasize that ``open source is reproducible by necessity''
-- since reproducibility is a prerequisite for remote collaboration.
-Smith therefore defined GitHub as part of the next stage in web
-development, following on from the social web of
-Facebook et al.: ``The Collaborative Web.''
+He cited the EMCEE Markov Chain Monte Carlo code~\cite{emcee-web} by
+Dan Foreman-Mackey and contributors as good examples of scientific
+projects using GitHub.
+
+C. Titus Brown~\cite{ged-web} argued that with regard to reusable
+software, ``we should just start doing it.'' Smith noted that
+documentation should be ``treated as a first class entity.''  He noted
+that some software fields have excellent cultures of module reuse,
+such as RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
+CPAN~\cite{cpan-web}, where there is effectively ultra-low-friction
+collaboration through these repositories. This has not happened in
+highly numerical, compiled language scientific software.
+
+He argued that the GitHub Pull Request code review mechanism
+facilitates such collaboration, by allowing one to code first, and
+seek review and a merge back into the trunk later.
+
+Fernando Perez emphasized that ``open source is reproducible by
+necessity'' - since reproducibility is a prerequisite for remote
+collaboration.  Smith therefore defined GitHub as part of the next
+stage in web development, following on from the social web of
+Facebook, i.e., ``The Collaborative Web.''
 

### 2014-02-08 WSSSPE1_report/WSSSPE1_summary_report.tex
small comments on initial sections

@@
 
-Because progress in scientific research is dependent on the quality
-and accessibility of software at all levels, it is now critical to
-address many challenges related to the development, deployment, and
-maintenance of reusable software. In addition, it is essential that
-scientists, researchers, and students are able to learn and adopt
-software-related skills and methodologies. Established researchers are
-already acquiring some of these skills, and in particular a
-specialized class of software developers is emerging in academic
-environments who are an integral and embedded part of successful
-research teams. The WSSSPE workshop was intended to provide a forum
-for discussion of the challenges, including both positions and
-experiences.  The short papers and discussion have been archived to
-provide a basis for continued discussion, and the workshop has fed
-into the collaborative writing of this journal publication.
-Additional papers based on extended versions of workshop submissions
-are expected.  The level of interest in the workshop has led the
-organizers, working with some of the submitters and attendees, to plan
-two additional workshops (at the 2014 SciPy and SC conferences) and to
-turn the workshop website into a community website that can be used as
-a focus for further discussion and progress.  A minisymposium on
-``Reliable Computational Science'' that aims to further explore some
-of the key issues raised in WSSSPE is co-organized by a participant in
-the 2014 SIAM Annual Meeting.
+Because progress in scientific research is dependent on the quality and
+accessibility of software at all levels, it is now critical to address many
+challenges related to the development, deployment, and maintenance of reusable
+software.\note{Ketan: What are the levels in scientific process where software
+is required. It might be interesting to outline these stages and types of
+software required in each stage.} In addition, it is essential that scientists,
+researchers, and students are able to learn and adopt software-related skills
+and methodologies. Established researchers are already acquiring some of these
+skills, and in particular a specialized class of software developers is
+emerging in academic environments who are an integral and embedded part of
+successful research teams. The WSSSPE workshop was intended to provide a forum
+for discussion of the challenges, including both positions and experiences. The
+short papers and discussion have been archived to provide a basis for continued
+discussion, and the workshop has fed into the collaborative writing of this
+journal publication.  Additional papers based on extended versions of workshop
+submissions are expected.  The level of interest in the workshop has led the
+organizers, working with some of the submitters and attendees, to plan two
+additional workshops (at the 2014 SciPy and SC conferences) and to turn the
+workshop website into a community website that can be used as a focus for
+further discussion and progress.  A minisymposium on ``Reliable Computational
+Science'' that aims to further explore some of the key issues raised in WSSSPE
+is co-organized by a participant in the 2014 SIAM Annual Meeting.
 
@@
 \item impact of computer science research on the development of
-  scientific software
+    scientific software and vice versa
 \end{itemize}
@@
 needs a minimal funding requirement even with a vibrant community --
-good will only goes so far if you are being paid to do something else.
+goodwill only goes so far if you are being paid to do something else.
 He talked about grant schemes of relevance in the U.S., particularly

### 2014-02-08 WSSSPE1_report/WSSSPE1_summary_report.tex
adding my paper desc to community section

@@
 
+What are the possible career paths for a specialist software developer working as part of a scientific research group?
+
 \section{Communities} \label{sec:community}
@@
 This collection of papers was summarized by Karen Cranston.
+
+One of the paper by Maheshwari et al~\cite{Maheshwari_WSSSPE} that was
+discussed was about the role of technological catalysts in overall e-research
+process.  The paper focuses on ``technology catalysts" and their role in modern
+scientific research process. Technology catalyst could be defined as a role
+played by an individual with a knowhow of technological advancements tasked
+with user engagement with a goal of enacting scientific or engineering
+applications using suitable tools and techniques to take advantage of
+technological capabilities and benefitting applications.
+
+One of the tasks of catalysts is to seek community collaborations for new
+applications and user engagement thus benefitting both: science, by effective
+running of scientific codes over computational infrastructure and technology,
+by conducting research and seeking findings to improve technology.
+
+The particular engagements described in the paper came up from authors work as
+postdoctoral researcher at Cornell and Argonne. Interaction with the scientific
+communities in both institutions resulted in these collaborations.
 

### 2014-02-08 WSSSPE1_report/WSSSPE1_summary_report.tex
minor changes in a few places

@@
 \item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
-\item Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
+\item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
 \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
-\item * Frank Löffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
+\item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
 \item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
@@
 challenges related to the development, deployment, and maintenance of reusable
-software.\note{Ketan: What are the levels in scientific process where software
-is required. It might be interesting to outline these stages and types of
-software required in each stage.} In addition, it is essential that scientists,
+software.
+%\note{Ketan: What are the levels in scientific process where software
+%is required. It might be interesting to outline these stages and types of
+%software required in each stage.}
+In addition, it is essential that scientists,
 researchers, and students are able to learn and adopt software-related skills
@@
 
-Bourne's keynote presentation is entitled ``A Recipe for Sustainable
-Software''.  He emphasized that sustainability for software ``does not
+Bourne's keynote presentation was entitled ``A Recipe for Sustainable
+Software.''  He emphasized that sustainability for software ``does not
 just mean more money from Government.''  He noted that in places,
@@
 computational biology, and has concluded that ``I have proved I cannot
-reproduce research from my own lab''~\cite{Veretnik}.
+reproduce research from my own lab.''~\cite{Veretnik}
 
@@
 He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology transfer~\cite{sbir-web, fased-web}.
+with regard to technology transfer.~\cite{sbir-web, fased-web}
 
@@
 
-One of the paper by Maheshwari et al~\cite{Maheshwari_WSSSPE} that was
+One paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} that was
 discussed was about the role of technological catalysts in overall e-research
@@
 
-Nancy Wilkins-Diehr summarized 6 papers in "Communities Part 2". The Christopherson paper outlines the degree to which research
+Nancy Wilkins-Diehr summarized 6 papers in ``Communities Part 2.'' The Christopherson paper outlines the degree to which research
 relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
@@
 The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
-Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hack-athon to develop working code.
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through.
+Step 2 was a 5-day hack-athon \todo{we spell hackathon three different ways in two paragraphs - Pick one and be consistent.} to develop working code.
 Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of

### 2014-02-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3 Keynote: Merge minor changes.

@@
 
-\section{Keynotes\note{lead: Sou-Cheng (Terrya) Choi}} \label{sec:keynotes}
+\section{Keynotes\scriptsize{\note{lead: Sou-Cheng (Terrya) Choi}, \todo{include what was presented and discussed}}} \label{sec:keynotes}
 
@@
 He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology transfer.~\cite{sbir-web, fased-web}
+with regard to technology transfer~\cite{sbir-web, fased-web}.
 
@@
 
-His talk used an example from data reduction in Astronomy, where he
+Smith's talk started with an example from data reduction in Astronomy, where he
 needed to remove interfering effects from the device. He built a ``bad
 pixel mask'', and realized that while it was persistent, there was no
-way/practice of sharing this with the data, so many researchers
+way/practice of sharing this with the data. Consequently many researchers
 repeated the same calculations. He estimated that 13 person-years were
@@
 
-C. Titus Brown~\cite{ged-web} argued that with regard to reusable
+C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant, argued that with regard to reusable
 software, ``we should just start doing it.'' Smith noted that
@@
 
-Fernando Perez emphasized that ``open source is reproducible by
+Fernando Perez from the audience emphasized that ``open source is reproducible by
 necessity'' - since reproducibility is a prerequisite for remote
-collaboration.  Smith therefore defined GitHub as part of the next
+collaboration.   In response, Smith defined GitHub as part of the next
 stage in web development, following on from the social web of

### 2014-02-11 WSSSPE1_report/WSSSPE1_summary_report.tex
minor format and added some content to use-case section

@@
 Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software.''  He emphasized that sustainability for software ``does not
-just mean more money from Government.''  He noted that in places,
-development in science has improved thanks to open source and Github,
-but for the most part remains arcane. He argued that there is much we
-can learn from the App Store model. He mentioned
-BioJava~\cite{biojava-web} and OpenScienceDataCloud~\cite{osdc-web} as
-distribution channels.  On a related note, Bourne observed a common
-evolutionary pathway for computational biology projects, from data
-archive to analytics platform to educational use, and suggested that
-use of scientific software for outreach might be the final step.
-
-Bourne shared with the audience a few real challenges he
-encountered. First, he has looked into reproducibility in
-computational biology, and has concluded that ``I have proved I cannot
-reproduce research from my own lab.''~\cite{Veretnik}
-
-Another problem he experienced was staff retention with respect to
-private organizations which reward those combining research and
-software expertise (the ``Google Bus''). However, he is a strong fan
-of private sector sustainability sources and public-private
-partnerships. He noted that making a successful business from
-scientific software alone is very rare: founders overvalue while
-customers undervalue. He noted that to last, an open source project
-needs a minimal funding requirement even with a vibrant community --
-goodwill only goes so far if you are being paid to do something else.
-He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology transfer~\cite{sbir-web, fased-web}.
-
-He also had problems with selling research software: the technology
-transfer office in his university wanted huge IP reachthrough, whereby
-they would get a share of profits from drugs developed by pharma
-companies who use the software. He was aware this was unrealistic but
-the technology transfer office insisted for a while. He wants to push
-a one-click approach to purchasing university-written software.
-
-He then presented the usual arguments on directly valuing software as
-a research output alongside papers. Readers will be familiar with this
-debate. An interesting reference here is~\cite{peer-review-code}.
-
-On the idea of the \textbf{digital enterprise}, where
-IT underpins the whole of organizational activities, he contended that
-universities are way behind the curve on this. In particular, he
-highlighted the separation of research, teaching and administration into
-silos without a common IT framework as a blocker to many useful
-organizational innovations: ``University 2.0 is yet to happen.'' He
-spoke of a circumstance where someone had used an algorithm developed
-for computational biology in marketing.
-
-The role of an institution is important in this space. He argued that
-funders can help train institutions not just individuals.
-
-He concluded with a reference to his paper~\cite{bourne_ten} and
-argued that computational scientists ``have a responsibility to
-convince their institutions and communities that software is
-scholarship,'' a point with which we strongly agree.
+Software.''  He emphasized that sustainability for software ``does not just
+mean more money from Government.''  He noted that in places, development in
+science has improved thanks to open source and Github, but for the most part
+remains arcane. He argued that there is much we can learn from the App Store
+model. He mentioned BioJava~\cite{biojava-web} and
+OpenScienceDataCloud~\cite{osdc-web} as distribution channels.  On a related
+note, Bourne observed a common evolutionary pathway for computational biology
+projects, from data archive to analytics platform to educational use, and
+suggested that use of scientific software for outreach might be the final step.
+
+Bourne shared with the audience a few real challenges he encountered. First, he
+has looked into reproducibility in computational biology, and has concluded
+that ``I have proved I cannot reproduce research from my own
+lab.''~\cite{Veretnik}
+
+Another problem he experienced was staff retention with respect to private
+organizations which reward those combining research and software expertise (the
+``Google Bus''). However, he is a strong fan of private sector sustainability
+sources and public-private partnerships. He noted that making a successful
+business from scientific software alone is very rare: founders overvalue while
+customers undervalue. He noted that to last, an open source project needs a
+minimal funding requirement even with a vibrant community -- goodwill only goes
+so far if you are being paid to do something else.  He talked about grant
+schemes of relevance in the U.S., particularly with regard to technology
+transfer~\cite{sbir-web, fased-web}.
+
+He also had problems with selling research software: the technology transfer
+office in his university wanted huge IP reachthrough, whereby they would get a
+share of profits from drugs developed by pharma companies who use the software.
+He was aware this was unrealistic but the technology transfer office insisted
+for a while. He wants to push a one-click approach to purchasing
+university-written software.
+
+He then presented the usual arguments on directly valuing software as a
+research output alongside papers. Readers will be familiar with this debate. An
+interesting reference here is~\cite{peer-review-code}.
+
+On the idea of the \textbf{digital enterprise}, where IT underpins the whole of
+organizational activities, he contended that universities are way behind the
+curve on this. In particular, he highlighted the separation of research,
+teaching and administration into silos without a common IT framework as a
+blocker to many useful organizational innovations: ``University 2.0 is yet to
+happen.'' He spoke of a circumstance where someone had used an algorithm
+developed for computational biology in marketing.
+
+The role of an institution is important in this space. He argued that funders
+can help train institutions not just individuals.
+
+He concluded with a reference to his paper~\cite{bourne_ten} and argued that
+computational scientists ``have a responsibility to convince their institutions
+and communities that software is scholarship,'' a point with which we strongly
+agree.
 
@@
 
+The software projects discussed in the workshop can be classified as follows:
+
+- General Purpose
+    Software that enables other software to run by providing suitable environment. eg. Collaborative development, Workflow and scripting frameworks
+
+- Special Purpose
+Software that was developed specifically for a target use-case scenario or keeping some fixed requirements in mind.
+
+--Science Oriented
+      Software developed to solve domain science problems eg. MINRES/QLP
+
+--Infrastructure Oriented
+      Software developed to better utilize new infrastructure, eg. MVAPICH
+
 \note{Dan: do we need this section?  I think there's some overlap

### 2014-02-12 WSSSPE1_report/WSSSPE1_summary_report.tex
Mainly Section 3: Keynote and also other sections: fixed typos, added references.

@@
 biomedical scientist but has formed four software companies. He also
-helped found PLOS and PDB~\cite{pdb-web}.  He is working on automating
+helped found PLOS~\cite{plos-web} and PDB~\cite{pdb-web}.  He is working on automating
 three-dimensional visualizations of cell contents and molecular
@@
 mean more money from Government.''  He noted that in places, development in
-science has improved thanks to open source and Github, but for the most part
+science has improved thanks to open source and GitHub~\cite{github-web}, but for the most part
 remains arcane. He argued that there is much we can learn from the App Store
 model. He mentioned BioJava~\cite{biojava-web} and
-OpenScienceDataCloud~\cite{osdc-web} as distribution channels.  On a related
+Open Science Data Cloud~\cite{osdc-web} as distribution channels.  On a related
 note, Bourne observed a common evolutionary pathway for computational biology
@@
 He also had problems with selling research software: the technology transfer
-office in his university wanted huge IP reachthrough, whereby they would get a
-share of profits from drugs developed by pharma companies who use the software.
+office in his university wanted huge IP reach through, whereby they would get a
+share of profits from drugs developed by pharmaceutical companies who use the software.
 He was aware this was unrealistic but the technology transfer office insisted
@@
 
-On the idea of the \textbf{digital enterprise}, where IT underpins the whole of
+On the idea of the \emph{digital enterprise}, where IT underpins the whole of
 organizational activities, he contended that universities are way behind the
@@
 needed to remove interfering effects from the device. He built a ``bad
-pixel mask'', and realized that while it was persistent, there was no
+pixel mask,'' and realized that while it was persistent, there was no
 way/practice of sharing this with the data. Consequently many researchers
@@
 as the effort that happens to make the essential things continue. This leads to
-having to decide what it is that we want to sustain, if what we want to sustain is
+having to decide what it is that we want to sustain, whether what we want to sustain is
 valuable, and finally, who would care if it went away, and how does one measure
@@
     \item visTrails
-    \item MINRES/QLP
+    \item MINRES-QLP
     \item SimGrid
@@
 --Science Oriented
-      Software developed to solve domain science problems eg. MINRES/QLP
+      Software developed to solve domain science problems eg. MINRES-QLP
 

### 2014-02-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Mainly Section 3.2: Keynote 2. Better flow.

@@
 pixel mask,'' and realized that while it was persistent, there was no
-way/practice of sharing this with the data. Consequently many researchers
+way or practice of sharing this with the data. Consequently many researchers
 repeated the same calculations. He estimated that 13 person-years were
@@
 
-He asked, of this practice, ``why didn't we know better?'' and argued
+``Why didn't we do better?''  Smith asked of this practice. He argued
 this was because we were taught to focus on immediate research
 outcomes and not on continuously improving and building on tools for
-research. He then asked, later in our working lives, when we do know
-better, why don't we act any different, and argued this was due to
+research. He then asked,  when we do know
+better, why we do not act any different. He argued that it was due to
 incentives and their lack: only the immediate products of research,
@@
 
-He cited the EMCEE Markov Chain Monte Carlo code~\cite{emcee-web} by
-Dan Foreman-Mackey and contributors as good examples of scientific
-projects using GitHub.
-
 C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant, argued that with regard to reusable
-software, ``we should just start doing it.'' Smith noted that
+software, ``we should just start doing it.'' Smith put forward that
 documentation should be ``treated as a first class entity.''  He noted
-that some software fields have excellent cultures of module reuse,
-such as RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
-CPAN~\cite{cpan-web}, where there is effectively ultra-low-friction
-collaboration through these repositories. This has not happened in
+that the open source community has excellent cultures of code reuse, for example,
+RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
+CPAN~\cite{cpan-web}, where there is effectively low-friction
+collaboration through the use of repositories. This has not happened in
 highly numerical, compiled language scientific software.
+An exception he cited as a good example of scientific
+projects using GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web} by
+Dan Foreman-Mackey and contributors.
 

### 2014-02-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3:  Keynote: Add line breaks.
Section 3.2:  Keynote 2 strengthened and lengthened.

@@
 biomedical scientist but has formed four software companies. He also
-helped found PLOS~\cite{plos-web} and PDB~\cite{pdb-web}.  He is working on automating
-three-dimensional visualizations of cell contents and molecular
-structures, a problem that has not been solved and when done, would
-serve as a key function of software in biomedical sciences.
+helped found PLOS~\cite{plos-web} and PDB~\cite{pdb-web}.  He is
+working on automating three-dimensional visualizations of cell
+contents and molecular structures, a problem that has not been solved
+and when done, would serve as a key function of software in biomedical
+sciences.
 
 Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software.''  He emphasized that sustainability for software ``does not just
-mean more money from Government.''  He noted that in places, development in
-science has improved thanks to open source and GitHub~\cite{github-web}, but for the most part
-remains arcane. He argued that there is much we can learn from the App Store
-model. He mentioned BioJava~\cite{biojava-web} and
-Open Science Data Cloud~\cite{osdc-web} as distribution channels.  On a related
-note, Bourne observed a common evolutionary pathway for computational biology
-projects, from data archive to analytics platform to educational use, and
-suggested that use of scientific software for outreach might be the final step.
-
-Bourne shared with the audience a few real challenges he encountered. First, he
-has looked into reproducibility in computational biology, and has concluded
-that ``I have proved I cannot reproduce research from my own
-lab.''~\cite{Veretnik}
-
-Another problem he experienced was staff retention with respect to private
-organizations which reward those combining research and software expertise (the
-``Google Bus''). However, he is a strong fan of private sector sustainability
-sources and public-private partnerships. He noted that making a successful
-business from scientific software alone is very rare: founders overvalue while
-customers undervalue. He noted that to last, an open source project needs a
-minimal funding requirement even with a vibrant community -- goodwill only goes
-so far if you are being paid to do something else.  He talked about grant
-schemes of relevance in the U.S., particularly with regard to technology
-transfer~\cite{sbir-web, fased-web}.
-
-He also had problems with selling research software: the technology transfer
-office in his university wanted huge IP reach through, whereby they would get a
-share of profits from drugs developed by pharmaceutical companies who use the software.
-He was aware this was unrealistic but the technology transfer office insisted
-for a while. He wants to push a one-click approach to purchasing
+Software.''  He emphasized that sustainability for software ``does not
+just mean more money from Government.''  He noted that in places,
+development in science has improved thanks to open source and
+GitHub~\cite{github-web}, but for the most part remains arcane. He
+argued that there is much we can learn from the App Store model. He
+mentioned BioJava~\cite{biojava-web} and Open Science Data
+Cloud~\cite{osdc-web} as distribution channels.  On a related note,
+Bourne observed a common evolutionary pathway for computational
+biology projects, from data archive to analytics platform to
+educational use, and suggested that use of scientific software for
+outreach might be the final step.
+
+Bourne shared with the audience a few real challenges he
+encountered. First, he has looked into reproducibility in
+computational biology, and has concluded that ``I have proved I cannot
+reproduce research from my own lab.''~\cite{Veretnik}
+
+Another problem he experienced was staff retention with respect to
+private organizations which reward those combining research and
+software expertise (the ``Google Bus''). However, he is a strong fan
+of private sector sustainability sources and public-private
+partnerships. He noted that making a successful business from
+scientific software alone is very rare: founders overvalue while
+customers undervalue. He noted that to last, an open source project
+needs a minimal funding requirement even with a vibrant community --
+goodwill only goes so far if you are being paid to do something else.
+He talked about grant schemes of relevance in the U.S., particularly
+with regard to technology transfer~\cite{sbir-web, fased-web}.
+
+He also had problems with selling research software: the technology
+transfer office in his university wanted huge IP reach through,
+whereby they would get a share of profits from drugs developed by
+pharmaceutical companies who use the software.  He was aware this was
+unrealistic but the technology transfer office insisted for a
+while. He wants to push a one-click approach to purchasing
 university-written software.
 
-He then presented the usual arguments on directly valuing software as a
-research output alongside papers. Readers will be familiar with this debate. An
-interesting reference here is~\cite{peer-review-code}.
-
-On the idea of the \emph{digital enterprise}, where IT underpins the whole of
-organizational activities, he contended that universities are way behind the
-curve on this. In particular, he highlighted the separation of research,
-teaching and administration into silos without a common IT framework as a
-blocker to many useful organizational innovations: ``University 2.0 is yet to
-happen.'' He spoke of a circumstance where someone had used an algorithm
-developed for computational biology in marketing.
-
-The role of an institution is important in this space. He argued that funders
-can help train institutions not just individuals.
-
-He concluded with a reference to his paper~\cite{bourne_ten} and argued that
-computational scientists ``have a responsibility to convince their institutions
-and communities that software is scholarship,'' a point with which we strongly
-agree.
+He then presented the usual arguments on directly valuing software as
+a research output alongside papers. Readers will be familiar with this
+debate. An interesting reference here is~\cite{peer-review-code}.
+
+On the idea of the \emph{digital enterprise}, where IT underpins the
+whole of organizational activities, he contended that universities are
+way behind the curve on this. In particular, he highlighted the
+separation of research, teaching and administration into silos without
+a common IT framework as a blocker to many useful organizational
+innovations: ``University 2.0 is yet to happen.'' He spoke of a
+circumstance where someone had used an algorithm developed for
+computational biology in marketing.
+
+The role of an institution is important in this space. He argued that
+funders can help train institutions not just individuals.
+
+He concluded with a reference to his paper~\cite{bourne_ten} and
+argued that computational scientists ``have a responsibility to
+convince their institutions and communities that software is
+scholarship,'' a point with which we strongly agree.
 
@@
 
-Smith's talk started with an example from data reduction in Astronomy, where he
-needed to remove interfering effects from the device. He built a ``bad
-pixel mask,'' and realized that while it was persistent, there was no
-way or practice of sharing this with the data. Consequently many researchers
-repeated the same calculations. He estimated that 13 person-years were
-wasted in this redundant calculation.
+Smith's talk started with an example from data reduction in Astronomy,
+where he needed to remove interfering effects from the device. He
+built a ``bad pixel mask,'' and realized that while it was persistent,
+there was no way or practice of sharing this with the
+data. Consequently many researchers repeated the same calculations. He
+estimated that 13 person-years were wasted in this redundant
+calculation.
 
@@
 outcomes and not on continuously improving and building on tools for
-research. He then asked,  when we do know
-better, why we do not act any different. He argued that it was due to
-incentives and their lack: only the immediate products of research,
-not the software, are valued.  He referenced Victoria Stodden's talk
-at OKCon~\cite{okcon-stodden-talk}
-which he said argued these points well.
-
-C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant, argued that with regard to reusable
-software, ``we should just start doing it.'' Smith put forward that
-documentation should be ``treated as a first class entity.''  He noted
-that the open source community has excellent cultures of code reuse, for example,
+research. He then asked, when we do know better, why we do not act any
+different. He argued that it was due to incentives and their lack:
+only the immediate products of research, not the software, are valued.
+He referenced Victoria Stodden's talk at
+OKCon~\cite{okcon-stodden-talk} which he said argued these points
+well.
+
+C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant,
+argued that with regard to reusable software, ``we should just start
+doing it.'' Smith replied that documentation should be ``treated
+as a first class entity.''  He noted that the open source community
+has excellent cultures of code reuse, for example,
 RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
 CPAN~\cite{cpan-web}, where there is effectively low-friction
-collaboration through the use of repositories. This has not happened in
-highly numerical, compiled language scientific software.
-An exception he cited as a good example of scientific
-projects using GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web} by
-Dan Foreman-Mackey and contributors.
-
-He argued that the GitHub Pull Request code review mechanism
+collaboration through the use of repositories. This has not happened
+in highly numerical, compiled language scientific software.  An
+exception he cited as a good example of scientific projects using
+GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
+by Dan Foreman-Mackey and contributors.
+
+He argued that  GitHub's \emph{Pull Request} code review mechanism
 facilitates such collaboration, by allowing one to code first, and
-seek review and a merge back into the trunk later.
-
-Fernando Perez from the audience emphasized that ``open source is reproducible by
-necessity'' - since reproducibility is a prerequisite for remote
-collaboration. In response, Smith defined GitHub as part of the next
-stage in web development, following on from the social web of
-Facebook, i.e., ``The Collaborative Web.''
+seek review and merge back into the trunk later.
+
+``Open source is \ldots reproducible by necessity,''
+Smith quoted  Fernando Perez~\cite{perez-open-src-reproducible}, explaining that 
+reproducibility is a prerequisite for remote collaboration.  He pointed out that GitHub could be the propeller
+of the next stage in web development, i.e., ``the Collaborative Web,'' following on from the social web of
+Facebook.
+
+In conclusion Smith reiterated the importance of establishing incentive models 
+for open contributions and tool builders, for example, meaningful metrics 
+and research grants such as \cite{NSF_software_vision}. He urged computational scientists to 
+collabroate and share often  their research reports, teaching materials, 
+code, as well as data by attaching proper licenses.
 

### 2014-02-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3.2:  Keynote 2 strengthened and lengthened.

@@
 
-``Open source is \ldots reproducible by necessity,''
-Smith quoted  Fernando Perez~\cite{perez-open-src-reproducible}, explaining that 
-reproducibility is a prerequisite for remote collaboration.  He pointed out that GitHub could be the propeller
-of the next stage in web development, i.e., ``the Collaborative Web,'' following on from the social web of
-Facebook.
-
-In conclusion Smith reiterated the importance of establishing incentive models 
-for open contributions and tool builders, for example, meaningful metrics 
-and research grants such as \cite{NSF_software_vision}. He urged computational scientists to 
-collabroate and share often  their research reports, teaching materials, 
-code, as well as data by attaching proper licenses.
+``Open source is \ldots reproducible by necessity,'' Smith quoted
+Fernando Perez~\cite{perez-open-src-reproducible}, explaining that
+reproducibility is a prerequisite for remote collaboration.  He
+pointed out that GitHub could propel the next stage of web
+development, i.e., ``the collaborative web,'' following on from the
+social web of Facebook.
+
+In conclusion Smith reiterated the importance of establishing
+incentive models for open contributions and tool builders, for
+example, meaningful metrics and research grants such as
+\cite{NSF_software_vision}. He urged computational scientists to
+collabroate and share often their research reports, teaching
+materials, code, as well as data by attaching proper licenses.
 

### 2014-02-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3.2:  Keynote 2 refined a bit.

@@
 argued that with regard to reusable software, ``we should just start
-doing it.'' Smith replied that documentation should be ``treated
-as a first class entity.''  He noted that the open source community
-has excellent cultures of code reuse, for example,
+doing it.'' In this regard Smith replied that documentation should be
+``treated as a first class entity.''  He noted that the open source
+community has excellent cultures of code reuse, for example,
 RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and

### 2014-02-14 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes

@@
 \begin{itemize}
-    \item github
-    \item mvapich
-    \item Cactus
-    \item visTrails
-    \item MINRES-QLP
-    \item SimGrid
-    \item Swift
-    \item Galaxy
-    \item FastMath and package managers
-    \item R/qtl
-    \item OWL/RDF
-    \item AMR/multiphysics
-    \item DUNE
-    \item PAAS
-    \item Gateways
-    \item Globus-Genomics
+    \item mvapich, Cactus, DUNE
+    \item gateways, github, VisIT, visTrails, PAAS
+    \item SimGrid, Swift, Galaxy
+    \item FastMath, R/qtl, MINRES-QLP
+    \item OWL/RDF, AMR/multiphysics, Globus-Genomics
 \end{itemize}
@@
     \item Categorize the real-world software projects into workshop themes. This is based on the aspect of a software project that dominates and best fits one of the workshop themes.
-    \item Discuss how the theme has impacted software projects over its lifetime.
-    \item Discuss similarities and differences between these projects.
+    \item How the worshop themes are applicable to and has impacted software projects over its lifetime.
+    \item Similarities and differences between these projects.
     \item What is a takeaway for a new project? What are the lessons learned? How can the course of current and future development projects be altered to implement lessons learned.
@@
 The software projects discussed in the workshop can be classified as follows:
-
-- General Purpose
-    Software that enables other software to run by providing suitable environment. eg. Collaborative development, Workflow and scripting frameworks
-
-- Special Purpose
-Software that was developed specifically for a target use-case scenario or keeping some fixed requirements in mind.
-
---Science Oriented
-      Software developed to solve domain science problems eg. MINRES-QLP
-
---Infrastructure Oriented
-      Software developed to better utilize new infrastructure, eg. MVAPICH
+\begin{itemize}
+\item General Purpose Software applicable to a generic task and/or enabling other software to run by providing suitable environment.
+eg.
+Collaborative development, Workflow and scripting frameworks, visualization frameworks.
+\item Special Purpose Software that was developed specifically for a target use-case scenario or keeping some fixed/frozen requirements in mind.
+\begin{itemize}
+  \item Domain science oriented software developed to solve domain science problems, eg. MINRES-QLP.
+  \item Infrastructure oriented software developed to better utilize new infrastructure, eg. MVAPICH.
+\end{itemize}
+\end{itemize}
+
+A recurrent theme of the workshop was following best practices while developing software. Many submissions discuss software development best practices.
 

### 2014-02-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3.1: Keynote 1: Good, better, best, never let it rest.

@@
 
-\todo{section should include what was presented and what was discussed}
-
-\subsection{A recipe for sustainable software, Philip E. Bourne.}
+\subsection{A Recipe for Sustainable Software, Philip E. Bourne.}
 
@@
 Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software.''  He emphasized that sustainability for software ``does not
-just mean more money from Government.''  He noted that in places,
-development in science has improved thanks to open source and
-GitHub~\cite{github-web}, but for the most part remains arcane. He
-argued that there is much we can learn from the App Store model. He
-mentioned BioJava~\cite{biojava-web} and Open Science Data
-Cloud~\cite{osdc-web} as distribution channels.  On a related note,
-Bourne observed a common evolutionary pathway for computational
-biology projects, from data archive to analytics platform to
-educational use, and suggested that use of scientific software for
-outreach might be the final step.
+Software,'' developed based on his experiences.  He emphasized that
+sustainability for software ``does not just mean more money from
+Government'' (see also Section~\ref{sec:cross-cutting}).  Other
+factors to consider, he mentioned, include costs of production, ease
+of maintainence, community involvement, and distribution channels.
+
+In places, Bourne said, development in science has improved thanks to
+open source and GitHub~\cite{github-web}, but for the most part
+remains arcane. He argued that we can learn much from the App Store
+model about interfaces, ratings, and so on. He also mentioned
+BioJava~\cite{biojava-web} and Open Science Data Cloud~\cite{osdc-web}
+as distribution channels.  On a related note, Bourne observed a common
+evolutionary pathway for computational biology projects, from data
+archive to analytics platform to educational use, and suggested that
+use of scientific software for outreach might be the final step.
 
 Bourne shared with the audience a few real challenges he
-encountered. First, he has looked into reproducibility in
-computational biology, and has concluded that ``I have proved I cannot
-reproduce research from my own lab.''~\cite{Veretnik}
+encountered. First, also an anecdote, he has looked into
+reproducibility in computational biology, but has concluded that ``I
+have proved I cannot reproduce research from my own
+lab.''~\cite{Veretnik}
 
@@
 software expertise (the ``Google Bus''). However, he is a strong fan
-of private sector sustainability sources and public-private
-partnerships. He noted that making a successful business from
-scientific software alone is very rare: founders overvalue while
-customers undervalue. He noted that to last, an open source project
-needs a minimal funding requirement even with a vibrant community --
-goodwill only goes so far if you are being paid to do something else.
-He talked about grant schemes of relevance in the U.S., particularly
-with regard to technology transfer~\cite{sbir-web, fased-web}.
+of software sustainability through public-private partnerships. He
+noted that making a successful business from scientific software alone
+is very rare: founders overvalue while customers undervalue. He noted
+that to last, an open source project needs a minimal funding
+requirement even with a vibrant community --- goodwill only goes so
+far if one is being paid to do something else.  He talked about grant
+schemes of relevance in the U.S., particularly with regard to
+technology transfer~\cite{sbir-web, fased-web}.
 
 He also had problems with selling research software: the technology
-transfer office in his university wanted huge IP reach through,
-whereby they would get a share of profits from drugs developed by
-pharmaceutical companies who use the software.  He was aware this was
-unrealistic but the technology transfer office insisted for a
-while. He wants to push a one-click approach to purchasing
-university-written software.
+transfer office in his university wanted huge intellectural property
+reach through, whereby they would get a share of profits from drugs
+developed by pharmaceutical companies who use the software.  He was
+aware this was unrealistic but the technology transfer office insisted
+for a while. He wants to push a one-click approach for customers to
+purchase university-written software.
 
@@
 a research output alongside papers. Readers will be familiar with this
-debate. An interesting reference here is~\cite{peer-review-code}.
-
-On the idea of the \emph{digital enterprise}, where IT underpins the
-whole of organizational activities, he contended that universities are
-way behind the curve on this. In particular, he highlighted the
-separation of research, teaching and administration into silos without
-a common IT framework as a blocker to many useful organizational
-innovations: ``University 2.0 is yet to happen.'' He spoke of a
-circumstance where someone had used an algorithm developed for
-computational biology in marketing.
-
+debate.  An interesting reference provided by him
+is~\cite{peer-review-code}, which explores involving software
+engineers in the review process of scientific code.
+
+On the notion of \emph{digital enterprise}, where information
+technology (IT) underpins the whole of organizational activities, he
+contended that universities are way behind the curve. In
+particular, he highlighted the separation of research, teaching, and
+administration into silos without a common IT framework as a blocker
+to many useful organizational innovations: ``University 2.0 is yet to
+happen.'' He spoke of a circumstance where someone had used an
+algorithm developed for computational biology in marketing.
 The role of an institution is important in this space. He argued that
-funders can help train institutions not just individuals.
+funders can help train institutions, not just individuals, in this regard.
 
@@
 argued that computational scientists ``have a responsibility to
-convince their institutions and communities that software is
-scholarship,'' a point with which we strongly agree.
+convince their institutions, reviewers, and communities that software is
+scholarship, frequently
+more valuable than a research article,'' a point with which we strongly agree.
 

### 2014-02-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: Cross-cutting issue. Try to help.

@@
 
-\subsection{A Recipe for Sustainable Software, Philip E. Bourne.}
+\subsection{A Recipe for Sustainable Software, Philip E. Bourne.} \label{sec:keynote1}
 
@@
 
-\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith}
+\subsection{Scientific Software and the Open Collaborative Web, Arfon Smith} \label{sec:keynote2}
 
@@
 
-\section{Cross-cutting Issue: Defining Sustainability}  \label{sec:cross-cutting}
-
-\note{lead: Daniel S. Katz}
+\section{Cross-cutting Issue: Defining Sustainability \scriptsize{\note{lead: Daniel S. Katz}}}  \label{sec:cross-cutting}
 
 The question of what was meant by ``sustainability'' in the context of software
-came up in many different parts of the workshop, specifically in one of the
-keynotes, the Developing and Maintaining Software panel, and the Policy panel.
-
-In the opening keynote, Phil Bourne suggested that perhaps sustainability can be defined
+came up in many different parts of the workshop, specifically in the first
+keynote (Section~\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.
+
+In the opening keynote, Philip Bourne suggested that perhaps sustainability can be defined
 as the effort that happens to make the essential things continue. This leads to
 having to decide what it is that we want to sustain, whether what we want to sustain is
-valuable, and finally, who would care if it went away, and how does one measure
+valuable, and finally, who would care if it went away, and how one measures
 how much they care.
@@
 would be made towards this goal.  Another topic that came up is what the
-goal of sustainability is.  Perhaps it's reproducible science, or
+goal of sustainability is.  Perhaps it is reproducible science, or
 persistence, or quality, or something else.  Similarly, some attendees want
-to understand how success in sustainability is measured.  How does a set of
+to understand how success in sustainability is measured.  How does a group of
 developers know when they have actually achieved sustainable software?
@@
 Finally, they develop an architecture evaluation framework based on scenarios
-that helps to illuminate how to measure quality/sustainability at
+that help to illuminate how to measure quality or sustainability at
 the architectural level.
@@
 \begin{itemize}
-\item Will the software continue to provide the same functionality in the future, even if the environment (other parts of the infrastructure) changes?
-\item Is the functionality and usability clearly explained to new users? Do users have a mechanism to ask questions and to learn about the element?
+\item Will the software continue to provide the same functionality in the future, 
+      even if the environment (other parts of the infrastructure) changes?
+\item Is the functionality and usability clearly explained to new users? 
+\item Do users have a mechanism to ask questions and to learn about the element?
 \item Will the functionality be correct in the future, even if the environment changes?
@@
 As shown in Table~\ref{tab:defining-sustainability}, for each paper he
-discussed what was meant by software, if there was a definition of sustainability, and
+discussed what software meant, whether there was a definition of sustainability, and
 what the approach was to either understand or evaluate sustainability.
 
-\begin{table}[h]
+\begin{table}[t]
   \begin{scriptsize}
     \begin{center}
-      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}}
+      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}.}
       \label{tab:defining-sustainability}

### 2014-02-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Removed redundant lines.
Section 1: Intro. Add statistics of participants.

@@
 discussion, and the workshop has fed into the collaborative writing of this
-journal publication.  Additional papers based on extended versions of workshop
+journal publication. An estimate of 90 to 150 paricipants were present 
+throughout the day. Additional papers based on extended versions of workshop
 submissions are expected.  The level of interest in the workshop has led the
@@
 (\S\ref{sec:conclusions}).
-
-\todo{Mention somewhere statistics of participants: There were 140 seats, I (Katz) think 149 people at the first keynote, and about 90 at the end of the workshop.
-They (SC13) counted people in the room 4 times during the day.
-But I don’t know how many unique people attended, which I agree they should know.}
 

### 2014-02-16 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 3.1: Keynote 1:  Made changes the speaker requested.
text improvement

@@
 E. Bourne of University of California, San Diego.  Bourne is a basic
-biomedical scientist but has formed four software companies. He also
-helped found PLOS~\cite{plos-web} and PDB~\cite{pdb-web}.  He is
-working on automating three-dimensional visualizations of cell
+biomedical scientist but has formed four software companies. He co-founded 
+PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB PDB~\cite{pdb-web}.  
+He is working on automating three-dimensional visualizations of cell
 contents and molecular structures, a problem that has not been solved
@@
 The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
-Step 1 was realized as a 2-day, in person specifications meeting and code walk-through.
-Step 2 was a 5-day hack-athon \todo{we spell hackathon three different ways in two paragraphs - Pick one and be consistent.} to develop working code.
-Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hackathon to develop working code.
+Step 3 was a 3-day hackathon to refine the code based on new requirements. The team worked on small, low-risk unit of
 code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
@@
 
-\section{Illustrative Use Cases} \label{sec:use-cases}
-
+\section{Case Studies} \label{sec:use-cases}
 \note{lead: Ketan Maheshwari, contributors: Andy Terrel}
 
-Following is a list of software projects discussed throughout the workshop:
-
-\begin{itemize}
-    \item mvapich, Cactus, DUNE
-    \item gateways, github, VisIT, visTrails, PAAS
-    \item SimGrid, Swift, Galaxy
-    \item FastMath, R/qtl, MINRES-QLP
-    \item OWL/RDF, AMR/multiphysics, Globus-Genomics
-\end{itemize}
-
-In this section, we discuss the following:
-\begin{itemize}
-    \item Categorize the real-world software projects into workshop themes. This is based on the aspect of a software project that dominates and best fits one of the workshop themes.
-    \item How the worshop themes are applicable to and has impacted software projects over its lifetime.
-    \item Similarities and differences between these projects.
-    \item What is a takeaway for a new project? What are the lessons learned? How can the course of current and future development projects be altered to implement lessons learned.
-\end{itemize}
-
-The software projects discussed in the workshop can be classified as follows:
-\begin{itemize}
-\item General Purpose Software applicable to a generic task and/or enabling other software to run by providing suitable environment.
-eg.
-Collaborative development, Workflow and scripting frameworks, visualization frameworks.
-\item Special Purpose Software that was developed specifically for a target use-case scenario or keeping some fixed/frozen requirements in mind.
-\begin{itemize}
-  \item Domain science oriented software developed to solve domain science problems, eg. MINRES-QLP.
-  \item Infrastructure oriented software developed to better utilize new infrastructure, eg. MVAPICH.
-\end{itemize}
-\end{itemize}
-
-A recurrent theme of the workshop was following best practices while developing software. Many submissions discuss software development best practices.
+For the purpose of studying software case-studies we classify the software
+projects discussed in the workshop into three broad categories. First, the
+`utility software', software tools that enable and/or facilitate the
+development of other tools and techniques to carry out scientific work. Second,
+the `scientific software', software that was developed with an aim to solve a
+class of scientific problems. Third, the `infrastructure software', software
+developed to efficiently utilize new research infrastructure.
+
+Although we anticipate overlap in categories and point it out in our
+discussion. The classification is based on two criterions: First, the character
+of a software project that dominates and best fits and the theme the paper
+discussing the project was placed in the workshop discussion.
+
+We discuss the case studies with an aim of understanding the practicalities of
+the many points discussed in the workshop and hope to extract the best
+practices and associate them with the projects. We discuss how the cases are
+applicable to and has impacted user communities over their lifetime isolating
+similarities and differences between projects. What are the takeaways for a new
+project and the lessons learned?  How can the course of current and
+future development projects be altered to implement lessons learned.
+
+\subsection{Utility Software}
+The General Purpose Software applicable to a generic task and/or enabling other
+software to run by providing suitable environment. For example, collaborative
+development, Workflow and scripting frameworks, visualization frameworks.
+\subsection{Scientific Software}
+Special Purpose Software that was developed specifically for a target use-case
+scenario or keeping some fixed/frozen requirements in mind.  
+Domain science
+oriented software developed to solve domain science problems, eg MINRES-QLP.
+\subsection{Infrastructure Software}
+Infrastructure oriented software developed to better utilize new
+infrastructure, eg. MVAPICH.
+%
+%\begin{itemize}
+%  \item mvapich, Cactus, DUNE
+%  \item gateways, github, VisIT, visTrails, PAAS
+%  \item SimGrid, Swift, Galaxy
+%  \item FastMath, R/qtl, MINRES-QLP
+%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
+%\end{itemize}
+%
 

### 2014-02-17 WSSSPE1_report/WSSSPE1_summary_report.tex
some more text to section 9

@@
 \subsection{Utility Software}
-The General Purpose Software applicable to a generic task and/or enabling other
-software to run by providing suitable environment. For example, collaborative
-development, Workflow and scripting frameworks, visualization frameworks.
+Utility software has a potential of broadest community impact. This is
+especially relavent in the modern era of social computing. In the context of
+this paper, a utility software could be defined as a general purpose software
+applicable to one or many generic tasks and/or enabling other software to run
+by providing suitable environment. For example, collaborative development
+framework such as \emph{github} and \emph{bitbucket}, Workflow and scripting
+frameworks such as \emph{Galaxy}, \emph{Swift} and \emph{VisTrails},
+visualization frameworks such as \emph{visIT}. Often projects start with
+specific, small set of requirements and users in mind but expand to provide
+generic functionalities and appeal to a wider user community.
+
 \subsection{Scientific Software}
 Special Purpose Software that was developed specifically for a target use-case
-scenario or keeping some fixed/frozen requirements in mind.  
-Domain science
+scenario or keeping some fixed/frozen requirements in mind.  Domain science
 oriented software developed to solve domain science problems, eg MINRES-QLP.
+
 \subsection{Infrastructure Software}

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
small updates and smoothing in sections 1 and 9

--- 2014-02-16 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 discussion, and the workshop has fed into the collaborative writing of this
-journal publication. An estimate of 90 to 150 paricipants were present 
+journal publication. An estimate of 90 to 150 participants were present 
 throughout the day. Additional papers based on extended versions of workshop
@@
 organizers, working with some of the submitters and attendees, to plan two
-additional workshops (at the 2014 SciPy and SC conferences) and to turn the
+additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
 workshop website into a community website that can be used as a focus for
-further discussion and progress.  A minisymposium on ``Reliable Computational
+further discussion and progress.  Additionally, a minisymposium at 2014 SIAM
+Annual Meeting on ``Reliable Computational
 Science'' that aims to further explore some of the key issues raised in WSSSPE
-is co-organized by a participant in the 2014 SIAM Annual Meeting.
+is co-organized by a WSSSPE1 participant.
 
@@
 
-For the purpose of studying software case-studies we classify the software
-projects discussed in the workshop into three broad categories. First, the
+For the purpose of studying software case studies, we classify the software
+projects discussed in the workshop in three broad categories. First, 
 `utility software', software tools that enable and/or facilitate the
 development of other tools and techniques to carry out scientific work. Second,
-the `scientific software', software that was developed with an aim to solve a
-class of scientific problems. Third, the `infrastructure software', software
+`scientific software', software that was developed with an aim to solve a
+class of scientific problems. Third, `infrastructure software', software
 developed to efficiently utilize new research infrastructure.
 
-Although we anticipate overlap in categories and point it out in our
-discussion. The classification is based on two criterions: First, the character
-of a software project that dominates and best fits and the theme the paper
-discussing the project was placed in the workshop discussion.
-
-We discuss the case studies with an aim of understanding the practicalities of
-the many points discussed in the workshop and hope to extract the best
+These categories have some overlap, and in some \note{most?} cases, the
+case studies don't cleanly fit into a single category.
+Our classification is based on two criteria: first, the character
+of a software project that dominates and best fits, and, second,
+the theme of the workshop discussion in which the paper discussing
+the project was placed.
+
+%and point it out in our
+%discussion.
+
+Our discussion of the case studies is aimed at understanding the practicalities of
+the many points discussed in the workshop, and we hope to extract the best
 practices and associate them with the projects. We discuss how the cases are
-applicable to and has impacted user communities over their lifetime isolating
-similarities and differences between projects. What are the takeaways for a new
-project and the lessons learned?  How can the course of current and
-future development projects be altered to implement lessons learned.
+applicable to and have impacted user communities over their lifetime, isolating
+similarities and differences between projects. We try to define the takeaways for a new
+project and the lessons learned. We try to understand how the course of current and
+future development projects might be altered to implement lessons learned.
 
 \subsection{Utility Software}
-The General Purpose Software applicable to a generic task and/or enabling other
-software to run by providing suitable environment. For example, collaborative
-development, Workflow and scripting frameworks, visualization frameworks.
+Utility software potentially has the broadest community impact of the three
+categories. This is
+especially relevant in the modern era of social computing. In the context of
+this paper, a utility software project could be defined as general purpose software
+applicable to one or many generic tasks and/or enabling other software to run
+by providing a suitable environment. Examples are collaborative development
+framework such as \emph{GitHub} and \emph{Bitbucket}, Workflow and scripting
+frameworks such as \emph{Galaxy}, \emph{Swift} and \emph{VisTrails},
+and visualization frameworks such as \emph{visIT}. Often projects start with
+specific, small set of requirements and users in mind but expand to provide
+more generic functionality and appeal to a wider user community.
+
 \subsection{Scientific Software}
-Special Purpose Software that was developed specifically for a target use-case
-scenario or keeping some fixed/frozen requirements in mind.  
-Domain science
-oriented software developed to solve domain science problems, eg MINRES-QLP.
+Special purpose software that was developed specifically for a target use-case
+scenario or keeping some fixed/frozen requirements in mind.  Domain science
+oriented software developed to solve domain science problems, e.g., MINRES-QLP.
+
 \subsection{Infrastructure Software}
 Infrastructure oriented software developed to better utilize new
-infrastructure, eg. MVAPICH.
+infrastructure, e.g., MVAPICH.
 %
@@
 
-\note{Dan: do we need this section?  I think there's some overlap
+\note{Dan: I think there's some overlap
   between this and the discussion in the Developing and Support
-  Software section.  Or perhaps this should go in the cross-cutting
+  Software section.  Perhaps this should go in the cross-cutting
   issues section?  Let's write it down here - then we can decide if it goes elsewhere}

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

--- 2014-02-16 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 E. Bourne of University of California, San Diego.  Bourne is a basic
-biomedical scientist but has formed four software companies. He co-founded 
-PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB PDB~\cite{pdb-web}.  
-He is working on automating three-dimensional visualizations of cell
+biomedical scientist but has formed four software companies. He also
+helped found PLOS~\cite{plos-web} and PDB~\cite{pdb-web}.  He is
+working on automating three-dimensional visualizations of cell
 contents and molecular structures, a problem that has not been solved
@@
 The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
-Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hackathon to develop working code.
-Step 3 was a 3-day hackathon to refine the code based on new requirements. The team worked on small, low-risk unit of
+Step 1 was realized as a 2-day, in person specifications meeting and code walk-through.
+Step 2 was a 5-day hack-athon \todo{we spell hackathon three different ways in two paragraphs - Pick one and be consistent.} to develop working code.
+Step 3 was a 3-day hack-a-thon to refine the code based on new requirements. The team worked on small, low-risk unit of
 code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
@@
 
-\section{Case Studies} \label{sec:use-cases}
+\section{Illustrative Use Cases} \label{sec:use-cases}
+
 \note{lead: Ketan Maheshwari, contributors: Andy Terrel}
 
-For the purpose of studying software case-studies we classify the software
-projects discussed in the workshop into three broad categories. First, the
-`utility software', software tools that enable and/or facilitate the
-development of other tools and techniques to carry out scientific work. Second,
-the `scientific software', software that was developed with an aim to solve a
-class of scientific problems. Third, the `infrastructure software', software
-developed to efficiently utilize new research infrastructure.
-
-Although we anticipate overlap in categories and point it out in our
-discussion. The classification is based on two criterions: First, the character
-of a software project that dominates and best fits and the theme the paper
-discussing the project was placed in the workshop discussion.
-
-We discuss the case studies with an aim of understanding the practicalities of
-the many points discussed in the workshop and hope to extract the best
-practices and associate them with the projects. We discuss how the cases are
-applicable to and has impacted user communities over their lifetime isolating
-similarities and differences between projects. What are the takeaways for a new
-project and the lessons learned?  How can the course of current and
-future development projects be altered to implement lessons learned.
-
-\subsection{Utility Software}
-The General Purpose Software applicable to a generic task and/or enabling other
-software to run by providing suitable environment. For example, collaborative
-development, Workflow and scripting frameworks, visualization frameworks.
-\subsection{Scientific Software}
-Special Purpose Software that was developed specifically for a target use-case
-scenario or keeping some fixed/frozen requirements in mind.  
-Domain science
-oriented software developed to solve domain science problems, eg MINRES-QLP.
-\subsection{Infrastructure Software}
-Infrastructure oriented software developed to better utilize new
-infrastructure, eg. MVAPICH.
-%
-%\begin{itemize}
-%  \item mvapich, Cactus, DUNE
-%  \item gateways, github, VisIT, visTrails, PAAS
-%  \item SimGrid, Swift, Galaxy
-%  \item FastMath, R/qtl, MINRES-QLP
-%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
-%\end{itemize}
-%
+Following is a list of software projects discussed throughout the workshop:
+
+\begin{itemize}
+    \item mvapich, Cactus, DUNE
+    \item gateways, github, VisIT, visTrails, PAAS
+    \item SimGrid, Swift, Galaxy
+    \item FastMath, R/qtl, MINRES-QLP
+    \item OWL/RDF, AMR/multiphysics, Globus-Genomics
+\end{itemize}
+
+In this section, we discuss the following:
+\begin{itemize}
+    \item Categorize the real-world software projects into workshop themes. This is based on the aspect of a software project that dominates and best fits one of the workshop themes.
+    \item How the worshop themes are applicable to and has impacted software projects over its lifetime.
+    \item Similarities and differences between these projects.
+    \item What is a takeaway for a new project? What are the lessons learned? How can the course of current and future development projects be altered to implement lessons learned.
+\end{itemize}
+
+The software projects discussed in the workshop can be classified as follows:
+\begin{itemize}
+\item General Purpose Software applicable to a generic task and/or enabling other software to run by providing suitable environment.
+eg.
+Collaborative development, Workflow and scripting frameworks, visualization frameworks.
+\item Special Purpose Software that was developed specifically for a target use-case scenario or keeping some fixed/frozen requirements in mind.
+\begin{itemize}
+  \item Domain science oriented software developed to solve domain science problems, eg. MINRES-QLP.
+  \item Infrastructure oriented software developed to better utilize new infrastructure, eg. MVAPICH.
+\end{itemize}
+\end{itemize}
+
+A recurrent theme of the workshop was following best practices while developing software. Many submissions discuss software development best practices.
 
@@
 
+Brian Austin
+
 Lorena A Barba
@@
 
+Sharon Broude Geva
+
+Jed Brown
+
 Maxine Brown
@@
 
+Alexandru Calotoiu
+
 Jeffrey Carver
 
+Shreyas Cholia
+
+Peng Huat Chua
+
 Neil Chue Hong
@@
 
+Timothy Cockerill
+
 Karen Cranston
@@
 
+Marat Dukhan
+
 Ian Foster
@@
 
+Jeffrey Frey
+
 Derek Gaston
 
+AllisonGehrke
+
+Brian Glendenning
+
+Christian Godenschwager
+
 Derek Groen
@@
 
+Magne Haveraaen
+
+Steven Heaton
+
+Oscar Hernandez
+
 James Hetherington
 
+James Hetherington
+
 Simon Hettrick
 
+Jonathan Hicks
+
 Kenneth Hoste
@@
 
+Shao-Ching Huang
+
+Tsutomu Ikegami
+
+Kaxuya Ishimura
+
+Christian Iwainsky
+
+Craig Jackson
+
+Wesley Jones
+
+Randall Judd
+
+Shusuke Kasamatsu
+
 Daniel S. Katz
@@
 
+Kellie Kercher
+
+Mads Kristensen
+
+Carolyn Lauzon
+
 Arnaud Legrand
 
+Chris Lenhardt
+
+Michael Levy
+
 Frank L\"{o}ffler
 
+Monica Luecke
+
+Simon A. F. Lund
+
+Arthur Maccabe
+
 Paul Madden
@@
 
+Philip Maechling
+
 Ketan Maheshwari
 
+Brian Marker
+
 Suresh Marru
@@
 
+Matt McKenzie
+
 Chris Mentzel
 
+Paul Messina
+
+Mike Mikailov
+
 J. Yates Monteith
 
+Reagan More
+
+Rafael Morizawa
+
 Pierre Neyron
@@
 
+Jack Perdue
+
 John Peterson
@@
 
+Sara Rambacher
+
+Nicolas Renon
+
 Jason Riedy
 
+Todd Rimer
+
+Bill Sacks
+
 Andreas Schreiber
@@
 
+Jaraslaw Slawinski
+
 Arfon Smith
 
+Spencer Smith
+
 James Spencer
 
+Eric Stahlberg
+
+Timothy Stitt
+
+Hyoshin Sung
+
+Fr\'{e}d\'{e}ric Suter
+
 Shel Swenson
 
-Fr\'{e}d\'{e}ric Suter
+Yoshio Tanaka
 
@@
 
+Keren Tomko
+
 John Towns
@@
 
+Satori Tsuzuki
+
 Matthew Turk
@@
 
+Brice Videau
+
+Tajendra Vir Singh
+
 Von Welch
@@
 Theresa Windus
+
+Felix Wolf
 

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
minor changes

--- 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 \item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
-\item Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
+\item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
 \item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
@@
 by providing a suitable environment. Examples are collaborative development
-framework such as \emph{GitHub} and \emph{Bitbucket}, Workflow and scripting
-frameworks such as \emph{Galaxy}, \emph{Swift} and \emph{VisTrails},
+framework such as \emph{GitHub} and \emph{Bitbucket}, workflow and scripting
+frameworks such as \emph{Galaxy}, \emph{Swift}, and \emph{VisTrails},
 and visualization frameworks such as \emph{visIT}. Often projects start with
@@
 The following is a partial list of attendees who were recorded on the
-Google doc~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, with some additions made by the authors of this paper.
-
-\begin{multicols}{4}
+Google doc~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, or by the SC13 student volunteers, with some additions also made by the authors of this report.
+
+\begin{multicols}{3}
 \setlength{\parindent}{0pt}

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
small update

@@
 These categories have some overlap, and in some \note{most?} cases, the
-case studies don't cleanly fit into a single category.
+case studies do not cleanly fit into a single category.
 Our classification is based on two criteria: first, the character
@@
 Utility software potentially has the broadest community impact of the three
-categories. This is
-especially relevant in the modern era of social computing. In the context of
+categories. This is especially relevant in the modern era of social computing. In the context of
 this paper, a utility software project could be defined as general purpose software
@@
 more generic functionality and appeal to a wider user community.
+Owing to its generic nature, utility software finds a wider users communities which in turn leads to a higher participation in development.
+Consequently, the process becomes user-driven and self sustaining.
+Software developed in such scenario often benefits from community driven development, deployment and maintenance best practices.
+
+For instance the \emph{Galaxy} project has significantly benefitted from user community participation in its development.
+It follows agile software development practices and implements standard practices such as test-driven development and socialized bug managing practices via \emph{trello}.
+Galaxy \emph{histories} and \emph{toolshed} offers easy community sharing of data and tools further promoting a collaborative environment.
 

### 2014-02-20 WSSSPE1_report/WSSSPE1_summary_report.tex
more text to sec 9 subsections

@@
 framework such as \emph{GitHub} and \emph{Bitbucket}, workflow and scripting
-frameworks such as \emph{Galaxy}, \emph{Swift}, and \emph{VisTrails},
+frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and \emph{VisTrails},
 and visualization frameworks such as \emph{visIT}. Often projects start with
@@
 Software developed in such scenario often benefits from community driven development, deployment and maintenance best practices.
-
 For instance the \emph{Galaxy} project has significantly benefitted from user community participation in its development.
@@
 
+Another paradigm in utility software is the software delivered as service over the web.
+
 \subsection{Scientific Software}
 Special purpose software that was developed specifically for a target use-case
-scenario or keeping some fixed/frozen requirements in mind.  Domain science
-oriented software developed to solve domain science problems, e.g., MINRES-QLP.
+scenario or keeping some fixed/frozen requirements in mind.
+Software projects pertaining to specific scientific domain often tend to be in a niche and the user community tends to be small to medium.
+Accuracy and optimization are some of the paramount requirements of most scientific software.
+Advantages of a smaller codebase and requirements include stability, ease of installation, configurability and integration into larger systems.
+While the software can stay stable and require relatively low maintenance, the responsibility is often on the shoulders of a very few developers.
+Community participation in the development tends to be low.
+Development is linear and simplistic with a limited scope to follow software best practices.
+Domain science oriented software developed to solve  specific problems, e.g., \emph{MINRES-QLP}, \emph{FASTMath}.
 
@@
 infrastructure, e.g., MVAPICH.
+Explores uncharted territories.
+Highest risk.
+Long requirement-development-test cycles.
+Wide community support but tends to be handled by specialists rather then end users.
+
 %

### 2014-02-21 WSSSPE1_report/WSSSPE1_summary_report.tex
small edit
minor grammar fix
small edits

@@
 to yes or no answers, the complete set would determine a range of values for
-sustainability. These question are:
+sustainability. These questions are:
 \begin{itemize}
@@
 Galaxy \emph{histories} and \emph{toolshed} offers easy community sharing of data and tools further promoting a collaborative environment.
+The project very closely follows the guidelines described in~\cite{Carver_WSSSPE} and many from~\cite{Prlic_WSSSPE}.
 
@@
 Software projects pertaining to specific scientific domain often tend to be in a niche and the user community tends to be small to medium.
-Accuracy and optimization are some of the paramount requirements of most scientific software.
+Numerical accuracy and algorithmic optimization are some of the paramount requirements of most scientific software.
 Advantages of a smaller codebase and requirements include stability, ease of installation, configurability and integration into larger systems.
@@
 Development is linear and simplistic with a limited scope to follow software best practices.
-Domain science oriented software developed to solve  specific problems, e.g., \emph{MINRES-QLP}, \emph{FASTMath}.
+Domain science oriented software developed to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
+One exception is \emph{Kitware}.
 
@@
 Infrastructure oriented software developed to better utilize new
-infrastructure, e.g., MVAPICH.
+infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}.
 Explores uncharted territories.
 Highest risk.
+Overlap with the utility software as it must serve as a facilitator to run existing software.
 Long requirement-development-test cycles.
 Wide community support but tends to be handled by specialists rather then end users.
-
-%
-%\begin{itemize}
-%  \item mvapich, Cactus, DUNE
-%  \item gateways, github, VisIT, visTrails, PAAS
-%  \item SimGrid, Swift, Galaxy
-%  \item FastMath, R/qtl, MINRES-QLP
-%  \item OWL/RDF, AMR/multiphysics, Globus-Genomics
-%\end{itemize}
-%
+Promotes collaborations, one of the key ingredient of a sustainable process.
 

### 2014-02-22 WSSSPE1_report/WSSSPE1_summary_report.tex
adding some text

@@
 Utility software potentially has the broadest community impact of the three
-categories. This is especially relevant in the modern era of social computing. In the context of
-this paper, a utility software project could be defined as general purpose software
-applicable to one or many generic tasks and/or enabling other software to run
-by providing a suitable environment. Examples are collaborative development
-framework such as \emph{GitHub} and \emph{Bitbucket}, workflow and scripting
-frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and \emph{VisTrails},
-and visualization frameworks such as \emph{visIT}. Often projects start with
-specific, small set of requirements and users in mind but expand to provide
-more generic functionality and appeal to a wider user community.
-Owing to its generic nature, utility software finds a wider users communities which in turn leads to a higher participation in development.
-Consequently, the process becomes user-driven and self sustaining.
-Software developed in such scenario often benefits from community driven development, deployment and maintenance best practices.
-For instance the \emph{Galaxy} project has significantly benefitted from user community participation in its development.
-It follows agile software development practices and implements standard practices such as test-driven development and socialized bug managing practices via \emph{trello}.
-Galaxy \emph{histories} and \emph{toolshed} offers easy community sharing of data and tools further promoting a collaborative environment.
-The project very closely follows the guidelines described in~\cite{Carver_WSSSPE} and many from~\cite{Prlic_WSSSPE}.
-
-Another paradigm in utility software is the software delivered as service over the web.
+categories. This is especially relevant in the modern era of social computing.
+In the context of this paper, a utility software project could be defined as
+general purpose software applicable to one or many generic tasks and/or
+enabling other software to run by providing a suitable environment. Examples
+are collaborative development framework such as \emph{GitHub} and
+\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
+\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
+such as \emph{visIT}. Often projects start with specific, small set of
+requirements and users in mind but expand to provide more generic functionality
+and appeal to a wider user community. Owing to its generic nature, utility
+software finds a wider users communities which in turn leads to a higher
+participation in development. Consequently, the process becomes user-driven and
+self sustaining. Software developed in such scenario often benefits from
+community driven development, deployment and maintenance best practices. For
+instance the \emph{Galaxy} project has significantly benefitted from user
+community participation in its development. It follows agile software
+development practices and implements standard practices such as test-driven
+development and socialized bug managing practices via \emph{trello}. Galaxy
+\emph{histories} and \emph{toolshed} offers easy community sharing of data and
+tools further promoting a collaborative environment. The project very closely
+follows the guidelines described in~\cite{Carver_WSSSPE} and many
+from~\cite{Prlic_WSSSPE}.
+
+Another paradigm in utility software is the software delivered as service over
+the web. With increasing popularity of cloud-based computational environments,
+many users are leaning towards tools used as service. \emph{GitHub} can be
+arguably considered one such tool, catering to collaborative development. For
+scientific users \emph{Globus} based tools are case of service based utility
+software discussed during the workshop. The data movement, authentication and
+sharing services offered by Globus can be easily used over the web by
+collaborating researcher.
 
@@
 Special purpose software that was developed specifically for a target use-case
-scenario or keeping some fixed/frozen requirements in mind.
-Software projects pertaining to specific scientific domain often tend to be in a niche and the user community tends to be small to medium.
-Numerical accuracy and algorithmic optimization are some of the paramount requirements of most scientific software.
-Advantages of a smaller codebase and requirements include stability, ease of installation, configurability and integration into larger systems.
-While the software can stay stable and require relatively low maintenance, the responsibility is often on the shoulders of a very few developers.
-Community participation in the development tends to be low.
-Development is linear and simplistic with a limited scope to follow software best practices.
-Domain science oriented software developed to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
-One exception is \emph{Kitware}.
+scenario or keeping some fixed/frozen requirements in mind. Software projects
+pertaining to specific scientific domain often tend to be in a niche and the
+user community tends to be small to medium. Numerical accuracy and algorithmic
+optimization are some of the paramount requirements of most scientific
+software. Advantages of a smaller codebase and requirements include stability,
+ease of installation, configurability and integration into larger systems.
+While the software can stay stable and require relatively low maintenance, the
+responsibility is often on the shoulders of a very few developers. Community
+participation in the development tends to be low. Development is linear and
+simplistic with a limited scope to follow software best practices. Domain
+science oriented software developed to solve  specific problems, e.g.,
+\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. One exception is
+\emph{Kitware}. Sustainability of such software is often a significant
+challenge. Many submissions report the software considered as a `byproduct` of
+actual research. Others contend that the software was not the main funded part
+of their research.
 
@@
 Infrastructure oriented software developed to better utilize new
-infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}.
-Explores uncharted territories.
-Highest risk.
-Overlap with the utility software as it must serve as a facilitator to run existing software.
-Long requirement-development-test cycles.
-Wide community support but tends to be handled by specialists rather then end users.
-Promotes collaborations, one of the key ingredient of a sustainable process.
+infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted
+territories. Highest risk. Overlap with the utility software as it must serve
+as a facilitator to run existing software. Long requirement-development-test
+cycles. Wide community support but tends to be handled by specialists rather
+then end users. Promotes collaborations, one of the key ingredient of a
+sustainable process.
 

### 2014-02-24 WSSSPE1_report/WSSSPE1_summary_report.tex
small change
Updating the developing and supporting software section
Turning the notes into somewhat structured paragraphs with common
themes, once complete I will start organizing them into subsections
and adding more references to the papers that were discussed.

@@
 \todo{section should include what was presented and what was discussed}
+
+The panel on developing and supporting software examined the challenges
+around scientific software development/support mainly focused on research
+groups that also produce code in various forms. There was widespread agreement
+that developing and maintaining software is hard, but best practices can help.
+Several stated that documentation is not just for users, and paying attention
+to API documentation, tutorials for building and deploying software, along with
+documented development practices can be very helpful in bringing  new developers
+into a project efficiently.
+
+There was discussion that backward compatibility is not always desirable, and it
+can be very costly. This must be balanced with the aims of a given project, how
+many other projects depend on and use the code when backwards incompatible
+changes are to be made. There are many examples in the wider open source
+software world of strategies for dealing with this, and again best practices
+can go a long way to mitigating issues around backwards compatibility. Many
+projects live with sub-optimal code for a while, and make allow backwards
+compatibility to be broken at agreed upon development points, such as a major
+release for a software library.
+
+Communities are extremely important in software projects, and both their
+building and continued engagement need attention during the project life cycle.
+Several of the papers submitted discussed how communities have been built around
+projects, and what is needed to enable a project to grow. Among these are public
+source code hosting, mailing lists, documentation, wikis, bug trackers, software
+downloads, continuous integration, software quality dashboards, and of course
+a general web presence to tie all of these things together. Questions were posed
+around users not answering the questions of other users. Several participants
+offered counterexamples, such as lists where developers do not participate in
+mailing lists as much due to users being more active, or whether the ``core
+team'' can end up setting unrealistic expectations by doing too much. Team Geek
+and Turk's paper on scaling code in the human dimension discuss how development
+lists that are welcoming to people actually have many more people contributing.
+
+Discussion around recruiting and/or retaining personnel in this area is hard,
+with one of the major reasons being no long-term career paths (especially when
+compared with industry). How should software development be balanced with
+research? It is apparent that things are beginning to improve, such as
+incentives for software development in the form of altmetrics, tenure committee
+consideration, NSF Products vs Publications is was noted it can be very
+difficult to measure where people are using small bits of your code.
+
+There were 13 articles about different experiences in this area, but little
+about GUI testing, performance, scalability, or agile development practices.
+There were several unique perspectives about issues such as managing API
+changes, using the same best practices for software as data, and going beyond
+simply ``slapping an OSI-approved license on code''.
+
+The question of what ``sustainable'' means in the context of software was
+raised. The resources at http://oss-watch.ac.uk/resources/ssmm discuss what
+to look for when choosing software for procurement, or reuse in further software
+development. Regardless of the license and development model that will be used
+the future of the project must be considered. Even if a particular piece of
+software satisfies today's requirements, will it continue to do so in the
+future? Questions such as whether the supplier will still be around in five
+years time, will it still care for its customers needs, will it be responsive to
+bug reports and feature requests? Should you tie your investment to a single
+supplier using a proprietary product, or ensure the project uses an OSI-approved
+license, and can outlive any single entity if the software is still important.
+
+What is the overarching goal of sustainability? Is it reproducible science,
+persistence, quality, something else? How should success be measured in this
+context? Is there some metric that can be used to determine when you have
+achieved sustainable software, or is this an ongoing process with no clear
+endpoint. For truly sustainable software there should be no endpoint, as the
+software products continue to be used and useful beyond the single institution,
+grant, and developer/development team. Sustainability must be addressed
+throughout the project life cycle.
+
+What about actual software engineering principles, such as modularity and
+extensibility? This is how industry maintains software, and ensures it continues
+to be useful. Everyone thinks that you can't rewrite your software, but if it is
+modular then you can keep systems up to date. Extensibility will keep you
+relevant if it is built into the project. One counterpoint raised by Jason Riedy
+was that trying to take advantage of the latest and greatest hardware often
+makes this painful, hence the lack of papers mentioning ``GPUs and exotic
+hardware''.
+
+The question of whether funders, such as the NSF, can mandate software plans in
+much the same way as they do data management plans. Daniel Katz responded that
+software is supposed to be described as part of the NSF data management plan,
+and that in their definition data includes software. A comment from Twitter
+(@biomickwatson) raised the issue that this requires reviewers and funders who
+understand the answers that are given in these plans. Daniel Katz responded
+that in programs focused on software or data this can be done effectively, but
+conceded that in more general programs this is indeed a problem.
+
+The papers submitted to this panel, and several others, include lots of general
+recommendations for processes, practices, tools, etc. One of the papers
+suggested that a ``Software Sustainability Institute'' should be vested with
+the authority to develop standardized quality processes, a central common
+repository, central resources for services and consulting, a think tank of
+sorts, and a software orphanage center. The idea received some resistance,
+stating difficulties in acting as a central common repository with so many
+compelling alternatives, and as an orphanage. The point for centralization of
+communication/point of contact might be reasonable, with the statement that
+``vested with authority'' is perhaps too strong, but ``providing tools if
+needed'' might be more appropriate. The wider open source community is moving
+to a model of distributed version control (Git, Mercurial, Bazaar, etc), rather
+than more centralized, as discussed in the second keynote.
+
+Some of the questions raised after the panel discussion.
+
+Rather than teaching developers about domain science, or domain scientists about
+software development practices, why don't we teach both communities to
+collaborate more effectively? Can this be done without teaching each side a
+little of the other to enable communication, with a response that it really is
+not two binary communities, but a spectrum with useful roles in the middle.
+Someone else asked if a developer can be effective without being part of the
+domain community, with responses that this really depends on the specific
+problem---translators and T-shaped people are important. Why aren't academic
+communities taught how to evaluate cross-disciplinary work well?
+
+Is there a role for the growing science of the team science field in this? There
+is overlap between the communities, with support for virtual organizations,
+tool development, etc. How can we make time in an already crowded schedule to
+introduce these topics to students? Should they be introduced through lab
+classes as in ``real sciences''?
 
@@
 infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted
-territories. Highest risk. Overlap with the utility software as it must serve
-as a facilitator to run existing software. Long requirement-development-test
-cycles. Wide community support but tends to be handled by specialists rather
-then end users. Promotes collaborations, one of the key ingredient of a
-sustainable process.
+territories. Highest risk since the projects are experimental. Overlap with the
+utility software as it must serve as a facilitator to run existing software.
+Long requirement-development-test cycles. Wide community support but tends to
+be handled by specialists rather then end users. Promotes collaborations, one
+of the key ingredient of a sustainable process.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
+associated with software projects dealing with new architectures.
 

### 2014-02-24 WSSSPE1_report/WSSSPE1_summary_report.tex
A few more paragraphs from the questions
This will move into more cohesive sections soon!

@@
 classes as in ``real sciences''?
+
+Are there significant differences in projects that have been running for 1, 3,
+5, or 10+ years? Are there shared experiences for projects of a similar stage
+of maturity? It was noted that computing and communication have changed
+significantly over the past decade, and many of the experiences are tied to
+the history of computing and communication. See the history of GCC, Emacs, or
+the Visualization Toolkit for examples. Others felt that computing has changed
+less, but communication and the widespread availability of tools has. It was
+noted that email lists, websites, chat rooms, version control, virtual and
+physical meetings are all over 20 years old.
+
+There was debate that while some of the basics of computing may be fairly
+similar, that the tools commonly used for computing have changed quite
+significantly. Reference was made to Perl, which was commonly used, giving way
+to whole new languages, such as Python, for gluing things together and how this
+induces many students into entirely rewriting the scaffolding, leaving the old
+to rot and the experiments to become non-reproducible as the tools change.
+Jason Riedy stated that he had been guilty of this in the past. There was
+discussion of this tendency along with the enormous differences in the speed and
+ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
+the immediate sharing of the latest development in CVS, Subversion, Git,
+Mercurial, Bazaar, etc.
+
+The question was also posed as to whether the distinction between researcher
+and developer is sensible, with James Hetherington commenting that in the UK
+they have been examining a more nuanced view of research software engineers
+and researcher developers. Should this be less of a contract relationship, and
+more of a collaborative relationship. This is also at the core of the business
+model that Kitware presented in its submission to the workshop. Are other
+ingredients missing such as applied mathematicians? Should this be defined more
+in terms of skill sets rather than roles and/or identities? This builds on the
+comments from Vaidy Sunderam that scientists are generally good writers, and
+have mathematical skills, so why can't they learn software engineering
+principles?
+
+Miller commented that all of the infrastructure that sits around a new
+algorithm that we need to make it useful and sustainable requires different
+skill sets than the algorithm developer. Friere commented that there are no
+good career paths for people with broad skills, no incentives for them to
+continue in these roles. There was debate around people doing what interests
+them, and learning computing leaves people cold, but is it that it leaves the
+people who find career paths in academia cold versus the full spectrum of
+people involved in research? Is this also caused by poor teaching, or that the
+perceived benefits for doing this are too small? It could also be attributed to
+their focus being on science, not software engineering, or do the people with
+the passion for software engineering in science simply have no viable career
+path and either adapt or seek out alternate career paths.
 

### 2014-02-24 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes

@@
 
-\section{Keynotes\scriptsize{\note{lead: Sou-Cheng (Terrya) Choi}, \todo{include what was presented and discussed}}} \label{sec:keynotes}
+\section{Keynotes \note{lead: Sou-Cheng (Terrya) Choi} } \label{sec:keynotes}
 
@@
 
-\section{Developing and Supporting Software} \label{sec:devel}
-
-\note{lead: Marcus Hanwell, contributors: Suresh Marru}
+\section{Developing and Supporting Software \note{lead: Marcus Hanwell, contributors: Suresh Marru}} \label{sec:devel}
 
@@
 
-\section{Policy} \label{sec:policy}
-
-\note{lead: Colin Venters}
+\section{Policy \note{lead: Colin Venters}} \label{sec:policy}
 
@@
 
-\section{Communities} \label{sec:community}
-
-\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
+\section{Communities \note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}} \label{sec:community}
 
@@
 
-\section{Case Studies} \label{sec:use-cases}
-\note{lead: Ketan Maheshwari, contributors: Andy Terrel}
+\section{Case Studies \note{lead: Ketan Maheshwari, contributors: Andy Terrel}} \label{sec:use-cases}
 
@@
 \appendix
-\section{List of attendees}
-
-\note{lead: Shel Swenson}
+\section{List of attendees \note{lead: Shel Swenson}}
 

### 2014-02-24 WSSSPE1_report/WSSSPE1_summary_report.tex
adding text on kitware

@@
 science oriented software developed to solve  specific problems, e.g.,
-\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. One exception is
-\emph{Kitware}. Sustainability of such software is often a significant
-challenge. Many submissions report the software considered as a `byproduct` of
-actual research. Others contend that the software was not the main funded part
-of their research.
+\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
+of such software is often a significant challenge. Many submissions report the
+software considered as a `byproduct` of actual research.  Others contend that
+the software was not the main funded part of their research. One exception is
+\emph{Kitware}. While being a software product specializing in scientific
+process, developing communities around software process is a core development
+focus. One example of this process is the development of the \emph{CMake} build
+utility which started out as a building tool for \emph{ITK} but grown to become
+a generic build utility for C++ projects.
 
 \subsection{Infrastructure Software}
-Infrastructure oriented software developed to better utilize new
-infrastructure, e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted
-territories. Highest risk since the projects are experimental. Overlap with the
-utility software as it must serve as a facilitator to run existing software.
-Long requirement-development-test cycles. Wide community support but tends to
-be handled by specialists rather then end users. Promotes collaborations, one
-of the key ingredient of a sustainable process.
-Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
-associated with software projects dealing with new architectures.
+Software developed to better utilize new infrastructure, e.g., \emph{MVAPICH},
+\emph{visIT}. Explores uncharted territories. Highest risk since the projects
+are experimental. Successful projects reap high rewards and have longer usage
+span. By nature, they overlap with the utility software as it must serve as a
+facilitator to run existing software. Long requirement-development-test cycles.
+Wide community support but tends to be handled by specialists rather then end
+users. Promotes collaborations across the breadth (eg. different departments)
+and depth (eg. stakeholders within a department) of community, one of the key
+ingredient of a sustainable process. Ferenbaugh~\cite{Ferenbaugh_WSSSPE}
+discusses the risks and challenges associated with software projects dealing
+with new architectures.
 

### 2014-02-25 WSSSPE1_report/WSSSPE1_summary_report.tex
adding comments and making small changes in 4, partly done but stopping for now

@@
 The panel on developing and supporting software examined the challenges
-around scientific software development/support mainly focused on research
+around scientific software development/support, mainly focused on research
 groups that also produce code in various forms. There was widespread agreement
@@
 There was discussion that backward compatibility is not always desirable, and it
-can be very costly. This must be balanced with the aims of a given project, how
+can be very costly. This must be balanced with the aims of a given project, and how
 many other projects depend on and use the code when backwards incompatible
@@
 can go a long way to mitigating issues around backwards compatibility. Many
-projects live with sub-optimal code for a while, and make allow backwards
+projects live with sub-optimal code for a while, and may allow backwards
 compatibility to be broken at agreed upon development points, such as a major
@@
 building and continued engagement need attention during the project life cycle.
-Several of the papers submitted discussed how communities have been built around
+Several of the submitted papers discussed how communities have been built around
 projects, and what is needed to enable a project to grow. Among these are public
 source code hosting, mailing lists, documentation, wikis, bug trackers, software
-downloads, continuous integration, software quality dashboards, and of course
+downloads, continuous integration, software quality dashboards, and of course,
 a general web presence to tie all of these things together. Questions were posed
@@
 mailing lists as much due to users being more active, or whether the ``core
-team'' can end up setting unrealistic expectations by doing too much. Team Geek
-and Turk's paper on scaling code in the human dimension discuss how development
+team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
+and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
 lists that are welcoming to people actually have many more people contributing.
 
-Discussion around recruiting and/or retaining personnel in this area is hard,
+Recruiting and/or retaining personnel in this area is hard,
 with one of the major reasons being no long-term career paths (especially when
@@
 incentives for software development in the form of altmetrics, tenure committee
-consideration, NSF Products vs Publications is was noted it can be very
+consideration, and NSF ``products'' vs ``publications''. It was noted it can be very
 difficult to measure where people are using small bits of your code.
@@
 changes, using the same best practices for software as data, and going beyond
-simply ``slapping an OSI-approved license on code''.
+simply ``slapping an OSI-approved license on code.''
 
 The question of what ``sustainable'' means in the context of software was
-raised. The resources at http://oss-watch.ac.uk/resources/ssmm discuss what
+raised (see \S\ref{sec:cross-cutting}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
 to look for when choosing software for procurement, or reuse in further software
-development. Regardless of the license and development model that will be used
+development. Regardless of the license and development model that will be used,
 the future of the project must be considered. Even if a particular piece of
 software satisfies today's requirements, will it continue to do so in the
-future? Questions such as whether the supplier will still be around in five
+future? Important questions include whether the supplier will still be around in five
 years time, will it still care for its customers needs, will it be responsive to
@@
 extensibility? This is how industry maintains software, and ensures it continues
-to be useful. Everyone thinks that you can't rewrite your software, but if it is
+to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
 modular then you can keep systems up to date. Extensibility will keep you
-relevant if it is built into the project. One counterpoint raised by Jason Riedy
+relevant if it is built into the project. One counterpoint raised by Jason Riedy\todo{add cite}
 was that trying to take advantage of the latest and greatest hardware often

### 2014-02-25 WSSSPE1_report/WSSSPE1_summary_report.tex
some additions to text in sec 9

@@
 of such software is often a significant challenge. Many submissions report the
-software considered as a `byproduct` of actual research.  Others contend that
+software considered as a `byproduct' of actual research.  Others contend that
 the software was not the main funded part of their research. One exception is
@@
 \subsection{Infrastructure Software}
-Software developed to better utilize new infrastructure, e.g., \emph{MVAPICH},
-\emph{visIT}. Explores uncharted territories. Highest risk since the projects
-are experimental. Successful projects reap high rewards and have longer usage
-span. By nature, they overlap with the utility software as it must serve as a
-facilitator to run existing software. Long requirement-development-test cycles.
-Wide community support but tends to be handled by specialists rather then end
-users. Promotes collaborations across the breadth (eg. different departments)
-and depth (eg. stakeholders within a department) of community, one of the key
-ingredient of a sustainable process. Ferenbaugh~\cite{Ferenbaugh_WSSSPE}
-discusses the risks and challenges associated with software projects dealing
-with new architectures.
+Software developed to better utilize new infrastructure and architectures,
+e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest
+risk since the projects are experimental. Successful projects reap high rewards
+and have longer usage span. By nature, they overlap with the utility software
+as they must serve as a facilitator to run ordinary codes. Long
+requirement-development-test cycles.  Wide community support but tends to be
+handled by specialists rather then end users. Promotes collaborations across
+the breadth (eg. different departments) and depth (eg. stakeholders within a
+department) of community, one of the key ingredient of a sustainable process.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
+associated with software projects dealing with new architectures. To leverage
+the power of accelerators such as GPUs and MICs new code and libraries are
+required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
+met with a limited success but nonetheless with many invaluable lessons were
+learned about cultural and technical aspects in sustainable software
+development practices.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
_

@@
 \subsection{Infrastructure Software}
+%TODO: expand
 Software developed to better utilize new infrastructure and architectures,
@@
 met with a limited success but nonetheless with many invaluable lessons were
-learned about cultural and technical aspects in sustainable software
+learned about influential cultural and technical aspects in sustainable software
 development practices.
+
+Summary of above discussed cases: lessons, trends, commonalities, is there an
+ideal sustainability scenario?
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
changes/edits/comments in sections 4 and 9

@@
 modular then you can keep systems up to date. Extensibility will keep you
-relevant if it is built into the project. One counterpoint raised by Jason Riedy\todo{add cite}
+relevant if it is built into the project. One counterpoint raised by Jason Riedy
 was that trying to take advantage of the latest and greatest hardware often
@@
 software is supposed to be described as part of the NSF data management plan,
-and that in their definition data includes software. A comment from Twitter
+and that in their definition, data includes software. A comment from Twitter
 (@biomickwatson) raised the issue that this requires reviewers and funders who
@@
 that in programs focused on software or data this can be done effectively, but
-conceded that in more general programs this is indeed a problem.
+agreed that in more general programs this is indeed a problem.
 
@@
 
-Some of the questions raised after the panel discussion.
+Some of the questions raised after the panel discussion are:
 
@@
 
-\subsection{Papers}
+\subsection*{Papers}
 
@@
 
-\section{Cross-cutting Issue: Defining Sustainability \scriptsize{\note{lead: Daniel S. Katz}}}  \label{sec:cross-cutting}
+\section{Cross-cutting Issue: Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:cross-cutting}
 
@@
 came up in many different parts of the workshop, specifically in the first
-keynote (Section~\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.
+keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.
 
@@
 For the purpose of studying software case studies, we classify the software
-projects discussed in the workshop in three broad categories. First, 
+projects discussed in the workshop in three broad categories. The first is 
 `utility software', software tools that enable and/or facilitate the
-development of other tools and techniques to carry out scientific work. Second,
+development of other tools and techniques to carry out scientific work. The second is
 `scientific software', software that was developed with an aim to solve a
-class of scientific problems. Third, `infrastructure software', software
+class of scientific problems. The third is `infrastructure software', software
 developed to efficiently utilize new research infrastructure.
 
-These categories have some overlap, and in some \note{most?} cases, the
+These categories have some overlap, and in some cases, the
 case studies do not cleanly fit into a single category.
 Our classification is based on two criteria: first, the character
-of a software project that dominates and best fits, and, second,
-the theme of the workshop discussion in which the paper discussing
-the project was placed.
+of a software project that dominates and where it `seems' to be the best fit, and, second,
+the the workshop discussion session where the paper about
+the project was discussed.
 
@@
 
-\subsection{Scientific Software}
-Special purpose software that was developed specifically for a target use-case
+\subsection{Scientific software}
+Scientific software consists primarily of special purpose software that was developed for a target use-case
 scenario or keeping some fixed/frozen requirements in mind. Software projects
@@
 \emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
-of such software is often a significant challenge. Many submissions report the
-software considered as a `byproduct' of actual research.  Others contend that
-the software was not the main funded part of their research. One exception is
-\emph{Kitware}. While being a software product specializing in scientific
-process, developing communities around software process is a core development
-focus. One example of this process is the development of the \emph{CMake} build
-utility which started out as a building tool for \emph{ITK} but grown to become
+of such software is often a significant challenge. Many submissions reported the
+software considered a `byproduct' of the actual research.  Others contended that
+the software was not the main funded part of their research. One exception was
+\emph{Kitware}, which while being a software product specializing in scientific
+process, has a core focus of developing communities around software processes.
+One example of this process is the development of the \emph{CMake} build
+utility, which started out as a building tool for \emph{ITK} but grew to become
 a generic build utility for C++ projects.
 
-\subsection{Infrastructure Software}
+\subsection{Infrastructure software}
 %TODO: expand
-Software developed to better utilize new infrastructure and architectures,
+Infrastructure software is developed to better utilize particular, often new, infrastructures and architectures,
 e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest
-risk since the projects are experimental. Successful projects reap high rewards
+risk since the projects are experimental. \note{Dan: so is the only difference here that
+this category is aimed at new systems, where utility software is not?
+If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.
+Something in the framing of this subsection doesn't feel right to me.} Successful projects reap high rewards
 and have longer usage span. By nature, they overlap with the utility software
@@
 handled by specialists rather then end users. Promotes collaborations across
-the breadth (eg. different departments) and depth (eg. stakeholders within a
+the breadth (e.g., different departments) and depth (e.g., stakeholders within a
 department) of community, one of the key ingredient of a sustainable process.

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
fixed a few typos

@@
 factors to consider, he mentioned, include costs of production, ease
-of maintainence, community involvement, and distribution channels.
+of maintenance, community involvement, and distribution channels.
 
@@
 He also had problems with selling research software: the technology
-transfer office in his university wanted huge intellectural property
+transfer office in his university wanted huge intellectual property
 reach through, whereby they would get a share of profits from drugs
@@
 \cite{NSF_software_vision}. He urged computational scientists to
-collabroate and share often their research reports, teaching
+collaborate and share often their research reports, teaching
 materials, code, as well as data by attaching proper licenses.
@@
 scientific research process. Technology catalyst could be defined as a role
-played by an individual with a knowhow of technological advancements tasked
+played by an individual with a know-how of technological advancements tasked
 with user engagement with a goal of enacting scientific or engineering
 applications using suitable tools and techniques to take advantage of
-technological capabilities and benefitting applications.
+technological capabilities and benefiting applications.
 
 One of the tasks of catalysts is to seek community collaborations for new
-applications and user engagement thus benefitting both: science, by effective
+applications and user engagement thus benefiting both: science, by effective
 running of scientific codes over computational infrastructure and technology,
@@
 The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
-Step 1 was realized as a 2-day, in person specifications meeting and code walk-through. Step 2 was a 5-day hackathon to develop working code.
+Step 1 was realized as a 2-day, in person specifications meeting and code walkthrough. Step 2 was a 5-day hackathon to develop working code.
 Step 3 was a 3-day hackathon to refine the code based on new requirements. The team worked on small, low-risk unit of
@@
 community driven development, deployment and maintenance best practices. For
-instance the \emph{Galaxy} project has significantly benefitted from user
+instance the \emph{Galaxy} project has significantly benefited from user
 community participation in its development. It follows agile software

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
use WSSSPE1 in title and abstract (not just WSSSPE)

@@
 
-\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE)}
+\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
 
@@
 \begin{abstract}
-This technical report discusses the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE). ...
+This technical report discusses the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1). ...
 \end{abstract}

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
avoid double-usage of 'discussion'

@@
 for discussion of the challenges, including both positions and experiences. The
-short papers and discussion have been archived to provide a basis for continued
+short papers and debates have been archived to provide a basis for continued
 discussion, and the workshop has fed into the collaborative writing of this

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
restructure sentence

@@
 workshop website into a community website that can be used as a focus for
-further discussion and progress.  Additionally, a minisymposium at 2014 SIAM
-Annual Meeting on ``Reliable Computational
-Science'' that aims to further explore some of the key issues raised in WSSSPE
-is co-organized by a WSSSPE1 participant.
+further discussion and progress. Additionally, a minisymposium that aims to
+further explore some of the key issues raised in WSSSPE is co-organized by a
+WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
+Computational Science''.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
avoid duplicate 'paper'
don't show URL _and_ cite, only cite

@@
 papers, and putting together the agenda. Section~\ref{sec:process} is
-based on that paper. The remainder of this paper is based on the
+based on that publication. The remainder of this paper is based on the
 workshop itself, as documented in a set of collaborative notes taken
@@
 bound it. Roughly following the example of NSF's Vision and Strategy
-for Software
-(\url{http://www.nsf.gov/publications/pub_summ.jsp?ods_key=nsf12113})~\cite{NSF_software_vision},
+for Software~\cite{NSF_software_vision},
 the ecosystem may be viewed as having challenges related to:

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
spell out PDB

@@
 E. Bourne of University of California, San Diego.  Bourne is a basic
-biomedical scientist but has formed four software companies. He co-founded 
-PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB PDB~\cite{pdb-web}.  
+biomedical scientist but has also formed four software companies. He co-founded 
+PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB
+Protein Data Bank~\cite{pdb-web}.  
 He is working on automating three-dimensional visualizations of cell

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
fixed a couple of minor things in first keynote section

@@
 Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software,'' developed based on his experiences.  He emphasized that
+Software'', and developed based on his experiences.  He emphasized that
 sustainability for software ``does not just mean more money from
@@
 In places, Bourne said, development in science has improved thanks to
-open source and GitHub~\cite{github-web}, but for the most part
+open source and hosting services like GitHub~\cite{github-web}, but for the most part
 remains arcane. He argued that we can learn much from the App Store
@@
 have proved I cannot reproduce research from my own
-lab.''~\cite{Veretnik}
+lab''~\cite{Veretnik}.
 
@@
 
-He then presented the usual arguments on directly valuing software as
-a research output alongside papers. Readers will be familiar with this
-debate.  An interesting reference provided by him
+He then presented arguments on directly valuing software as
+a research output alongside papers, a common discussion within this field.
+An interesting reference provided by him
 is~\cite{peer-review-code}, which explores involving software
@@
 scholarship, frequently
-more valuable than a research article,'' a point with which we strongly agree.
+more valuable than a research article,'' a point with which all authors
+strongly agree.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
use unsorted bibliography. Might be controvertial. Please revert if there is disagreement that this is better.
put punctuation inside of quoatation marks. It looks odd to me, but apparently has to be like this in American English (not so in British)

@@
 WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
-Computational Science''.
+Computational Science.''
 
@@
 Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software'', and developed based on his experiences.  He emphasized that
+Software,'' and developed based on his experiences.  He emphasized that
 sustainability for software ``does not just mean more money from
@@
 incentives for software development in the form of altmetrics, tenure committee
-consideration, and NSF ``products'' vs ``publications''. It was noted it can be very
+consideration, and NSF ``products'' vs ``publications.'' It was noted it can be very
 difficult to measure where people are using small bits of your code.
@@
 makes this painful, hence the lack of papers mentioning ``GPUs and exotic
-hardware''.
+hardware.''
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{unsrt}
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
add a note

@@
 The panel on developing and supporting software examined the challenges
-around scientific software development/support, mainly focused on research
+around scientific software development and support, mainly focused on research
 groups that also produce code in various forms. There was widespread agreement
 that developing and maintaining software is hard, but best practices can help.
-Several stated that documentation is not just for users, and paying attention
+Several participants stated that documentation is not just for users, and paying attention
 to API documentation, tutorials for building and deploying software, along with
@@
 projects live with sub-optimal code for a while, and may allow backwards
-compatibility to be broken at agreed upon development points, such as a major
+compatibility to be broken at agreed-upon development points, such as a major
 release for a software library.
@@
 around users not answering the questions of other users. Several participants
-offered counterexamples, such as lists where developers do not participate in
-mailing lists as much due to users being more active, or whether the ``core
+offered counterexamples, such as mailing lists where developers do not participate
+as much due to users being more active, or whether the ``core
 team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
@@
 extensibility? This is how industry maintains software, and ensures it continues
-to be useful. Everyone \note{everyone in the audience?  paper authors?  software developers in general?} thinks that you can't rewrite your software, but if it is
-modular then you can keep systems up to date. Extensibility will keep you
-relevant if it is built into the project. One counterpoint raised by Jason Riedy
+to be useful. Often, rewriting software is considered to be too costly, but
+with a modular design it can be kept up to date. Extensibility is expected to keep
+it relevant, if built into the project. One counterpoint raised by Jason Riedy
 was that trying to take advantage of the latest and greatest hardware often
@@
 The question of whether funders, such as the NSF, can mandate software plans in
-much the same way as they do data management plans. Daniel Katz responded that
+much the same way as they do data management plans, was raised. Daniel Katz responded that
 software is supposed to be described as part of the NSF data management plan,
-and that in their definition, data includes software. A comment from Twitter
+and that in their \note{who is ``they''? OCI? NSF?} definition, data includes software. A comment from Twitter
 (@biomickwatson) raised the issue that this requires reviewers and funders who

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
clarify some sentences - remove statement about version control systems, as this didn't add to the relevant discussion

@@
 repository, central resources for services and consulting, a think tank of
-sorts, and a software orphanage center. The idea received some resistance,
-stating difficulties in acting as a central common repository with so many
-compelling alternatives, and as an orphanage. The point for centralization of
-communication/point of contact might be reasonable, with the statement that
-``vested with authority'' is perhaps too strong, but ``providing tools if
-needed'' might be more appropriate. The wider open source community is moving
-to a model of distributed version control (Git, Mercurial, Bazaar, etc), rather
-than more centralized, as discussed in the second keynote.
+sorts, and a software orphanage center. The idea of one common repository received
+some resistance, with so many compelling alternatives available, e.g. BitBucket or
+GitHub. The point for centralization of communication/point of contact was seen as
+reasonable, with the statement that ``vested with authority'' is perhaps too strong,
+but ``providing tools if needed'' might be more appropriate.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
reformulate sentence

@@
 not two binary communities, but a spectrum with useful roles in the middle.
-Someone else asked if a developer can be effective without being part of the
+The question was raised if a developer can be effective without being part of the
 domain community, with responses that this really depends on the specific

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
reformulate sentence
extend sentence a bit

@@
 
-Is there a role for the growing science of the team science field in this? There
+What is the role for the growing field of team science? There
 is overlap between the communities, with support for virtual organizations,
@@
 There was debate that while some of the basics of computing may be fairly
-similar, that the tools commonly used for computing have changed quite
+similar tools commonly used for computing have changed quite
 significantly. Reference was made to Perl, which was commonly used, giving way
@@
 ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
-the immediate sharing of the latest development in CVS, Subversion, Git,
-Mercurial, Bazaar, etc.
+the immediate sharing of the latest development online, using revision control
+systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
use passive in sentence to avoid 'they'

@@
 and developer is sensible, with James Hetherington commenting that in the UK
-they have been examining a more nuanced view of research software engineers
-and researcher developers. Should this be less of a contract relationship, and
+a more nuanced view of research software engineers
+and researcher developers is examined. Should this be less of a contract relationship, and
 more of a collaborative relationship. This is also at the core of the business

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
minor rewordings

@@
 and researcher developers is examined. Should this be less of a contract relationship, and
-more of a collaborative relationship. This is also at the core of the business
+more of a collaborative relationship? This is also at the core of the business
 model that Kitware presented in its submission to the workshop. Are other
@@
 people who find career paths in academia cold versus the full spectrum of
-people involved in research? Is this also caused by poor teaching, or that the
-perceived benefits for doing this are too small? It could also be attributed to
-their focus being on science, not software engineering, or do the people with
+people involved in research? Is this also caused by poor teaching, or because the
+benefits for doing this are perceived as too small? It could also be attributed to
+their focus being on science, not software engineering, or do people with
 the passion for software engineering in science simply have no viable career
-path and either adapt or seek out alternate career paths.
+path and either adapt or seek out alternate career paths?
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
minor rewording

@@
 
-One paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} that was
-discussed was about the role of technological catalysts in overall e-research
-process.  The paper focuses on ``technology catalysts" and their role in modern
+The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on
+``technology catalysts" and their role in the modern
 scientific research process. Technology catalyst could be defined as a role
-played by an individual with a know-how of technological advancements tasked
+played by an individual with a know-how of technological advancements, tasked
 with user engagement with a goal of enacting scientific or engineering
-applications using suitable tools and techniques to take advantage of
+applications, and using suitable tools and techniques to take advantage of
 technological capabilities and benefiting applications.
@@
 applications and user engagement thus benefiting both: science, by effective
-running of scientific codes over computational infrastructure and technology,
-by conducting research and seeking findings to improve technology.
+running of scientific codes on computational infrastructure, and technology,
+by conducting research and seeking findings for technology improvement.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
add note
minor changes

@@
 by conducting research and seeking findings for technology improvement.
-
 The particular engagements described in the paper came up from authors work as
@@
 
+\note{More papers are needed here.}
+
 \subsection{Communities Part 2}
 
-Nancy Wilkins-Diehr summarized 6 papers in ``Communities Part 2.'' The Christopherson paper outlines the degree to which research
-relies on high quality software. There are often barriers and a lack of suitable incentives for researchers to embrace software engineering principals.
-The Water Science Software Institute is working to lower some of these barriers through an Open Community Engagement Process.
-This is a 4-step iterative development process that incorporates Agile development principals.
-
-\begin{itemize}
-
-\item Step 1 - Design - thorough discussion of research questions
-\item Step 2 - Develop working code
-\item Step 3 - Refine based on new requirements
-\item Step 4 - Publish open source
-
-\end{itemize}
-
-The Christopherson paper reports on the application of Steps 1-3 to a computational modeling framework developed in the 1990s.
-Step 1 was realized as a 2-day, in person specifications meeting and code walkthrough. Step 2 was a 5-day hackathon to develop working code.
-Step 3 was a 3-day hackathon to refine the code based on new requirements. The team worked on small, low-risk unit of
-code. It was challenging, revealed unanticipated obstacles, programmers had to work together, and experimentation was encouraged.
-
-The paper summarized recommendations to others wishing to engage in this or a similar process. Start small and gradually
-build toward more complex objectives. This is consistent with Agile development. Refactor before adding new functionality.
-Approach development as a learning experience. Welcome experimentation, and treat mistakes as a natural part of the learning process.
-Repeat Step 1 activities before all hackathons to develop consensus before coding. This allows coding to be the focus of subsequent hackathons.
-In higher risk situations, provide additional time for Step 1 activities. The Christopherson team recommends a minimum of two months.
-Ensure any newcomers receive some form of orientation prior to the hackathon, such as a code walkthrough or system documentation.
-Co-locate rather than collaborating remotely whenever feasible.
+Nancy Wilkins-Diehr summarized 6 papers in this part.
+
+The Christopherson paper outlines the degree to which research relies on high
+quality software. There are often barriers and a lack of suitable incentives
+for researchers to embrace software engineering principals. The Water Science
+Software Institute is working to lower some of these barriers through an Open
+Community Engagement Process. This is a 4-step iterative development process
+that incorporates Agile development principals.
+
+\begin{itemize}
+\item Step 1: Design - thorough discussion of research questions
+\item Step 2: Develop working code
+\item Step 3: Refine based on new requirements
+\item Step 4: Publish open source
+\end{itemize}
+
+The Christopherson paper reports on the application of Steps 1-3 to a
+computational modeling framework developed in the 1990s. Step 1 was realized as
+a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
+5-day hackathon to develop working code, and Step 3 a 3-day hackathon to refine
+the code based on new requirements. The team worked on small, low-risk units of
+code. It was challenging, revealed unanticipated obstacles, programmers had to
+work together, and experimentation was encouraged.
+
+The paper summarized recommendations to others wishing to engage in this or a
+similar process, starting small and gradually building toward more complex
+objectives. This is consistent with Agile development. Refactor before adding
+new functionality. Approach development as a learning experience. Welcome
+experimentation, and treat mistakes as a natural part of the learning process.
+Repeat Step 1 activities before all hackathons to develop consensus before
+coding. This allows coding to be the focus of subsequent hackathons. In higher
+risk situations, provide additional time for Step 1 activities. The
+Christopherson team recommends a minimum of two months. Ensure any newcomers
+receive some form of orientation prior to the hackathon, such as a code
+walkthrough or system documentation. Co-locate rather than collaborating
+remotely whenever feasible.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
mostly minor changes, but also shortened a lot of overly long lines in the latex source

@@
 
-The Pierce paper described how science gateways can provide a user-friendly entry to complex cyberinfrastructure. The paper opens with a description of the explosion of use of just a few gateways.
-Over 3.5 years, more than 7,000 biologists have run phylogenetic codes on supercomputers using the CIPRES Science Gateway.
-In 1 year, over 120 scientists from 50 institutions used the UltraScan Science Gateway,
-increasing the sophistication of analytical ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG) registered 100+
-users who used 250,000 CPU hours in only a few months.
-
-Gateways, however, need to keep operational costs low and can often make use of common components. Science Gateway Platform as a Service (Sci-GaP)
-delivers middleware as a hosted, scalable third-party service while domain developers focus on user interfaces and domain-specific data in the gateway.
-The middleware handles things like authentication, application installation and reliable execution and help desk support.
-
-One key to SciGaP is its openness.
-While based on Apache Airavata project and the CIPRES Workbench Framework, community contributions are encouraged because of its open source, open governance and open operations policies.
-The goal is robust, sustainable infrastructure with a cycle of development that improves reliability and prioritizes
-stakeholder requirements. The project is leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways into commodity services.
-
-The Zentner paper describes experiences and challenges with the large nanoHUB.org community. They define community as
-a ``body of persons of common and especially professional interests scattered through a larger society.'' This makes support
-challenging because of the diversity of viewpoints and needs. The group constantly examines its policies to determine
-whether they are indirectly alienating part of the community or encouraging particular types of use.
-
-nanoHUB's 10-year history with over 260,000 users annually provides a lot of data to analyze. 4000 resources contributed by
-1000 authors. nanoHUB serves both the research and education community and the contribution model allows researchers to get
-their codes out into the community and in use in education very rapidly. The primary software challenges are twofold - support
-for the HUBzero framework and challenges related to the software contributed by the community.
-
-The group has learned that community contributions are maximized with a tolerant licensing approach. HUBzero uses an
-LGPLv3 license so contributors can create unique components and license as they choose. If they make changes to source
-code, the original license must be maintained for redistribution. As far as contributed resources, these must be open access,
-but not necessarily open source. This allows contributors to meet the requirements of their institutions and funding agencies.
-Quality is maintained via user ratings. Documentation is encouraged and nanoHUB supplies regression test capabilities, but
-the user community provides ratings, poses questions and contributes to wishlists and citation counts - all of which incentivize
-code authors.
-
-The Terrel paper describes support for the Python scientific community through two major efforts - the SciPy conference and the
-NumFOCUS foundation. Reliance on software in science has driven a huge demand for development, but this development is typically done as a side effort
-and often in a rush to publish without documentation and testing.
-While created by academics, software support often falls to industrial institutions. SciPy brings together industry, government,
-and academics to share their code and experience in an open environment.
-
-NumFOCUS promotes open, usable scientific software while sustaining the community. Specific activities include educational programs,
-collaborative research tools and documentation and
-promotion of high-level languages, reproducible scientific research, and open-code development.
-Governance of the non-profit is a loose grantor-grantee relationship with projects allowing for monies to be placed in the groups accounts.
-This has raised money to hire developers for open code development, maintain testing machines, organize the PyData conference series, and sponsor community members to attend conferences.
-
-Software sustainability relies on contributions from all sectors of user community.
-Together SciPy and NumFOCUS support these sectors.
-By maximizing contributions growing the user base they help develop and mature Python.
-
-The L\"{o}ffler paper describes the Cactus project which was started in 1996 by participants in USA Binary Black Hole Alliance
-Challenge. Cactus has a flesh and thorns model - a community-oriented framework that allows researchers to easily work
-together with reusable and extendable software elements. Modules are compiled into an executable and can remain dormant, becoming
-active only when parameters and simulation data dictate. Users can  use modules written by others or can write their own
-modules without changing other code. The community has grown and diversified beyond the original science areas.
-
-The paper points out 4 keys to sustaining the community - modular design, growing a collaborative community, career paths and credit.
-In modular design, the Cactus project went far beyond standard practices of APIs.
-Domain specific languages (DSLs) allow decoupling of components - for example I/O, mesh refinement, PAPI counters, and
-boundary conditions abstracted from science code. In academia, publications are the main currency of credit. Because the
-project connects code developments to science, the work is
-publishable and modules are citable. Because of the open source, modular approach, programmers can see the impact of their contributions and often continue
-work after graduation. Career paths remain a challenge, however. Tasks that are essential from a software engineering perspective
-are often not rewarded in academia. The best programmers in a science environment often have multidisciplinary expertise.
-That also is not rewarded in academia.
-
-The Wilkins-Diehr paper describes an NSF software institute effort to build a community of those creating science gateways or
-web portals for science. These gateways, as described in some of the other papers in this section, can be quite capable, having a
-remarkable impact on science.
-
-This paper mentioned challenges highlighted by other papers in this section - mainly the conflict between funding for research
-vs infrastructure and the challenges around getting academic credit for infrastructure. Because the authors have studied
-many projects, they've also observed how development is often done in an isolated hobbyist environment. Developers are unable
-to take advantage of similar work done by others. isolation even when projects have common needs.
-But often projects struggle for sustainable funding because they provide infrastructure to conduct
-research and many times only the research is funded. Gateways also may start as small group research project, taking off in popularity
-once they go live without any long term plans for sustainability or without resources in the project to plan for such.
-Subsequent disruptions in service can limit effectiveness and test the limits of the research community's trust. The impact
-of gateways can be increased substantially if we understand what makes them successful.
-
-Recommendations from an early study of successful gateways include the following. Leadership and management teams should
-design governance to represent multiple strengths and perspectives, plan for change and turnover in the future,
-recruit a development team that understands both the technical and domain-related issues, consider sustainability and measure success early and often.
-Successful projects recognize the benefits and costs of hiring a team of professionals, demonstrate credibility through stability and
-clarity of purpose, leverage the work of others and plan for flexibility.
-Successful projects identify an existing community and understand that communities' needs before beginning. Projects adapt
-as the communities' needs evolve. For funders, the lifecycle of technology projects must be considered. Solicitations should
-be designed to reward effective planning, recognize the benefits and limitations of both technology innovation and reuse,
-expect adjustments during the production process, copy effective models from other industries and sectors, and encourage partnerships that support gateway sustainability.
-
-Through a business incubator type approach, the institute plans to provide a variety of services that could be shared amongst
-projects. Consultation and resources on topics such as business plan development, software engineering practices,
-software licensing options, usability, security and project management as well as a software repository and hosting
-service will be available. Experts will be available for multi-month assignments to help research teams build their own gateways,
-teaching them how to maintain and add to the work after the collaboration ends. Forums, symposia and an annual conference
-will connect members of the development community. A modular, layered framework that supports community contributions and
-allows developers to choose components will be delivered and finally
-workforce development activities will help train the next generation for careers in this cross- disciplinary area and
-build pools of institutional expertise that many projects can leverage. Shared services and forums can dramatically reduce the cost
-of building and sustaining gateways. Workforce development can encourage technology professionals to remain in the sciences.
+The Pierce paper described how science gateways can provide a user-friendly
+entry to complex cyberinfrastructure. The paper opens with a description of the
+explosion of use of just a few gateways. Over 3.5 years, more than 7,000
+biologists have run phylogenetic codes on supercomputers using the CIPRES
+Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
+UltraScan Science Gateway, increasing the sophistication of analytical
+ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG)
+registered 100+ users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often make use of
+common components. Science Gateway Platform as a Service (Sci-GaP) delivers
+middleware as a hosted, scalable third-party service while domain developers
+focus on user interfaces and domain-specific data in the gateway. The
+middleware handles things like authentication, application installation and
+reliable execution and help desk support.
+
+One key to SciGaP is its openness. While based on the Apache Airavata project
+and the CIPRES Workbench Framework, community contributions are encouraged
+because of its open source, open governance and open operations policies. The
+goal is robust, sustainable infrastructure with a cycle of development that
+improves reliability and prioritizes stakeholder requirements. The project is
+leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways
+into commodity services.
+
+The Zentner paper describes experiences and challenges with the large
+nanoHUB.org community. The authors define community as a ``body of persons of
+common and especially professional interests scattered through a larger
+society.'' This makes support challenging because of the diversity of
+viewpoints and needs. The group constantly examines its policies to determine
+whether they are indirectly alienating part of the community or encouraging
+particular types of use.
+
+nanoHUB's 10-year history with over 260,000 users annually provides a lot of
+data to analyze. 4000 resources contributed by 1000 authors. nanoHUB serves
+both the research and education community and the contribution model allows
+researchers to get their codes out into the community and in use in education
+very rapidly. The primary software challenges are twofold - support for the
+HUBzero framework and challenges related to the software contributed by the
+community.
+
+The group has learned that community contributions are maximized with a
+tolerant licensing approach. HUBzero uses an LGPLv3 license so contributors can
+create unique components and license as they choose. If they make changes to
+source code, the original license must be maintained for redistribution. As far
+as contributed resources, these must be open access, but not necessarily open
+source. This allows contributors to meet the requirements of their institutions
+and funding agencies. Quality is maintained via user ratings. Documentation is
+encouraged and nanoHUB supplies regression test capabilities, but the user
+community provides ratings, poses questions and contributes to wishlists and
+citation counts - all of which incentivize code authors.
+
+The Terrel paper describes support for the Python scientific community through
+two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
+on software in science has driven a huge demand for development, but this
+development is typically done as a side effort and often in a rush to publish
+without documentation and testing. While created by academics, software support
+often falls to industrial institutions. SciPy brings together industry,
+government, and academics to share their code and experience in an open
+environment.
+
+NumFOCUS promotes open, usable scientific software while sustaining the
+community. Specific activities include educational programs, collaborative
+research tools and documentation and promotion of high-level languages,
+reproducible scientific research, and open-code development. Governance of the
+non-profit is a loose grantor-grantee relationship with projects allowing for
+monies to be placed in the groups accounts. This has raised money to hire
+developers for open code development, maintain testing machines, organize the
+PyData conference series, and sponsor community members to attend conferences.
+
+Software sustainability relies on contributions from all sectors of the user
+community. Together SciPy and NumFOCUS support these sectors. By maximizing
+contributions growing the user base they help develop and mature Python.
+
+The L\"{o}ffler paper describes the Cactus project which was started in 1996 by
+participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
+and thorns, core and module, model - a community-oriented framework that allows
+researchers to easily work together with reusable and extendable software
+elements. Modules are compiled into an executable and can remain dormant,
+becoming active only when parameters and simulation data dictate. Users can
+use modules written by others or can write their own modules without changing
+other code. The community has grown and diversified beyond the original science
+areas.
+
+The paper points out 4 keys to sustaining the community - modular design,
+growing a collaborative community, career paths and credit. In modular design,
+the Cactus project went far beyond standard practices of APIs. Domain specific
+languages (DSLs) allow decoupling of components - for example I/O, mesh
+refinement, PAPI counters, and boundary conditions abstracted from science
+code. In academia, publications are the main currency of credit. Because the
+project connects code developments to science, the work is publishable and
+modules are citable. Because of the open source, modular approach, programmers
+can see the impact of their contributions and often continue work after
+graduation. Career paths remain a challenge, however. Tasks that are essential
+from a software engineering perspective are often not rewarded in academia. The
+best programmers in a science environment often have multidisciplinary
+expertise. That also is not rewarded in academia.
+
+The Wilkins-Diehr paper describes an NSF software institute effort to build a
+community of those creating science gateways or web portals for science. These
+gateways, as described in some of the other papers in this section, can be
+quite capable, having a remarkable impact on some parts of science.
+
+This paper mentioned challenges highlighted by other papers in this section -
+mainly the conflict between funding for research vs infrastructure and the
+challenges around getting academic credit for infrastructure. Because the
+authors have studied many projects, they've also observed how development is
+often done in an isolated hobbyist environment. Developers are unable to take
+advantage of similar work done by others, isolation even when projects have
+common needs. But often projects struggle for sustainable funding because they
+provide infrastructure to conduct research and many times only the research is
+funded. Gateways also may start as small group research project, taking off in
+popularity once they go live without any long term plans for sustainability or
+without resources in the project to plan for such. Subsequent disruptions in
+service can limit effectiveness and test the limits of the research community's
+trust. The impact of gateways can be increased substantially if we understand
+what makes them successful.
+
+Recommendations from an early study of successful gateways include the
+following. Leadership and management teams should design governance to
+represent multiple strengths and perspectives, plan for change and turnover in
+the future, recruit a development team that understands both the technical and
+domain-related issues, consider sustainability and measure success early and
+often. Successful projects recognize the benefits and costs of hiring a team of
+professionals, demonstrate credibility through stability and clarity of
+purpose, leverage the work of others and plan for flexibility. Successful
+projects identify an existing community and understand the communities' needs
+before beginning. Projects adapt as the communities' needs evolve. For funders,
+the lifecycle of technology projects must be considered. Solicitations should
+be designed to reward effective planning, recognize the benefits and
+limitations of both technology innovation and reuse, expect adjustments during
+the production process, copy effective models from other industries and
+sectors, and encourage partnerships that support gateway sustainability.
+
+Through a business incubator type approach, the institute plans to provide a
+variety of services that could be shared amongst projects. Consultation and
+resources on topics such as business plan development, software engineering
+practices, software licensing options, usability, security and project
+management as well as a software repository and hosting service will be
+available. Experts will be available for multi-month assignments to help
+research teams build their own gateways, teaching them how to maintain and add
+to the work after the collaboration ends. Forums, symposia and an annual
+conference will connect members of the development community. A modular,
+layered framework that supports community contributions and allows developers
+to choose components will be delivered and finally workforce development
+activities will help train the next generation for careers in this cross-
+disciplinary area and build pools of institutional expertise that many projects
+can leverage. Shared services and forums can dramatically reduce the cost of
+building and sustaining gateways. Workforce development can encourage
+technology professionals to remain in the sciences.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
shortened a lot of overly long lines in the latex source
add note

@@
 
+\note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
 The question of what was meant by ``sustainability'' in the context of software
 came up in many different parts of the workshop, specifically in the first
-keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel, and the Policy panel.
-
-In the opening keynote, Philip Bourne suggested that perhaps sustainability can be defined
-as the effort that happens to make the essential things continue. This leads to
-having to decide what it is that we want to sustain, whether what we want to sustain is
-valuable, and finally, who would care if it went away, and how one measures
-how much they care.
-
-In the Developing and Maintaining Software panel, there was some discussion
-on this topic: what does sustainability mean? It was pointed out that
-OSS Watch proposes a Software Sustainability Maturity Model to address
-the issue of how sustainability a particular element of software is, and says
-that this sustainability is important. ``When choosing
-software for procurement or development reuse - regardless of the
-license and development model you will use - you need to consider
-the future. While a software product may satisfy today's needs, will
-it satisfy tomorrow's needs? Will the supplier still be around in
-five years' time? Will the supplier still care for all its customers
-in five years' time? Will the supplier be responsive to bug reports
-and feature requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
-
-Attendees suggested that a key question that the definition of sustainability is
-one issue on which the community needs to agree, and that ideally, an initial
-definition would be determined during the workshop, or at least some progress
-would be made towards this goal.  Another topic that came up is what the
-goal of sustainability is.  Perhaps it is reproducible science, or
-persistence, or quality, or something else.  Similarly, some attendees want
-to understand how success in sustainability is measured.  How does a group of
-developers know when they have actually achieved sustainable software?
-This led to a comment that sustainability should be addressed throughout the
-full software life cycle.
-
-Another topic that came up during the Developing and Maintaining Software
-panel is the relationship of sustainability to other software attributes.  Attendees
+keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel,
+and the Policy panel.
+
+In the opening keynote, Philip Bourne suggested that perhaps sustainability can
+be defined as the effort that happens to make the essential things continue.
+This leads to having to decide what it is that we want to sustain, whether what
+we want to sustain is valuable, and finally, who would care if it went away,
+and how one measures how much they care.
+
+In the Developing and Maintaining Software panel, there was some discussion on
+this topic: what does sustainability mean? It was pointed out that OSS Watch
+proposes a Software Sustainability Maturity Model to address the issue of how
+sustainability a particular element of software is, and says that this
+sustainability is important. ``When choosing software for procurement or
+development reuse - regardless of the license and development model you will
+use - you need to consider the future. While a software product may satisfy
+today's needs, will it satisfy tomorrow's needs? Will the supplier still be
+around in five years' time? Will the supplier still care for all its customers
+in five years' time? Will the supplier be responsive to bug reports and feature
+requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
+
+Attendees suggested that a key question that the definition of sustainability
+is one issue on which the community needs to agree, and that ideally, an
+initial definition would be determined during the workshop, or at least some
+progress would be made towards this goal.  Another topic that came up is what
+the goal of sustainability is.  Perhaps it is reproducible science, or
+persistence, or quality, or something else.  Similarly, some attendees want to
+understand how success in sustainability is measured.  How does a group of
+developers know when they have actually achieved sustainable software? This led
+to a comment that sustainability should be addressed throughout the full
+software life cycle.
+
+Another topic that came up during the Developing and Maintaining Software panel
+is the relationship of sustainability to other software attributes.  Attendees
 asked ``what is the relationship between sustainability and provenance?'' And,
@@
 
-In addition, the Policy panel had a large amount of discussion about
-defining sustainability, as one of the subtopics in that panel was
-Modeling Sustainability, and modeling requires defining what will be modeled.
+In addition, the Policy panel had a large amount of discussion about defining
+sustainability, as one of the subtopics in that panel was Modeling
+Sustainability, and modeling requires defining what will be modeled.
 
 In this subtopic, two papers included discussion of the definition of
-sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned
-that this is a rather ambiguous concept, and that the lack of an accepted
-definition gets in the way of integrating the concept into software engineering.
-They suggest that sustainability is a non-functional requirement, and that the
-quality of software architectures determines sustainability.  They then
-propose that sustainability is a measure of a set of central quality attributes:
-extensibility, interoperability, maintainability, portability, reusability, and scalability.
-Finally, they develop an architecture evaluation framework based on scenarios
-that help to illuminate how to measure quality or sustainability at
-the architectural level.
-
-Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could
-be used to measure software sustainability, and though these questions might falsely lead
-to yes or no answers, the complete set would determine a range of values for
-sustainability. These questions are:
+sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
+rather ambiguous concept, and that the lack of an accepted definition gets in
+the way of integrating the concept into software engineering. They suggest that
+sustainability is a non-functional requirement, and that the quality of
+software architectures determines sustainability.  They then propose that
+sustainability is a measure of a set of central quality attributes:
+extensibility, interoperability, maintainability, portability, reusability, and
+scalability. Finally, they develop an architecture evaluation framework based
+on scenarios that help to illuminate how to measure quality or sustainability
+at the architectural level.
+
+Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could be
+used to measure software sustainability, and though these questions might
+falsely lead to yes or no answers, the complete set would determine a range of
+values for sustainability. These questions are:
 \begin{itemize}
@@
 
-Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the Modeling
-Sustainability papers in the panel.
-As shown in Table~\ref{tab:defining-sustainability}, for each paper he
-discussed what software meant, whether there was a definition of sustainability, and
-what the approach was to either understand or evaluate sustainability.
+Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the
+Modeling Sustainability papers in the panel. As shown in
+Table~\ref{tab:defining-sustainability}, for each paper he discussed what
+software meant, whether there was a definition of sustainability, and what the
+approach was to either understand or evaluate sustainability.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
lots of small text improvements

@@
 sorts, and a software orphanage center. The idea of one common repository received
-some resistance, with so many compelling alternatives available, e.g. BitBucket or
+some resistance, with so many compelling alternatives available, e.g. Bitbucket or
 GitHub. The point for centralization of communication/point of contact was seen as
@@
 
+\note{We should use either single or double quotations consistently throughout the paper.}
 For the purpose of studying software case studies, we classify the software
@@
 
-These categories have some overlap, and in some cases, the
+These categories have some overlap, showing in some cases where
 case studies do not cleanly fit into a single category.
@@
 enabling other software to run by providing a suitable environment. Examples
-are collaborative development framework such as \emph{GitHub} and
+are collaborative development frameworks such as \emph{GitHub} and
 \emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
 \emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
-such as \emph{visIT}. Often projects start with specific, small set of
+such as \emph{VisIT}. Often projects start with a specific, small set of
 requirements and users in mind but expand to provide more generic functionality
 and appeal to a wider user community. Owing to its generic nature, utility
-software finds a wider users communities which in turn leads to a higher
+software finds a wider user community which in turn leads to a higher
 participation in development. Consequently, the process becomes user-driven and
-self sustaining. Software developed in such scenario often benefits from
-community driven development, deployment and maintenance best practices. For
-instance the \emph{Galaxy} project has significantly benefited from user
-community participation in its development. It follows agile software
+self-sustaining.
+
+Software developed in such a scenario often benefits from
+community driven development, deployment and maintenance best practices. One
+such example is the \emph{Galaxy} project. It follows agile software
 development practices and implements standard practices such as test-driven
@@
 the web. With increasing popularity of cloud-based computational environments,
-many users are leaning towards tools used as service. \emph{GitHub} can be
-arguably considered one such tool, catering to collaborative development. For
-scientific users \emph{Globus} based tools are case of service based utility
-software discussed during the workshop. The data movement, authentication and
-sharing services offered by Globus can be easily used over the web by
-collaborating researcher.
+many users are leaning towards tools used as service. \emph{GitHub} and
+\emph{Bitbucket} can be arguably considered to be such tools, catering to
+collaborative development. For scientific users \emph{Globus}-based tools are
+a case of service-based utility software discussed during the workshop. The
+data movement, authentication and sharing services offered by Globus can be
+easily used over the web by collaborating researchers.
 
 \subsection{Scientific software}
-Scientific software consists primarily of special purpose software that was developed for a target use-case
-scenario or keeping some fixed/frozen requirements in mind. Software projects
-pertaining to specific scientific domain often tend to be in a niche and the
-user community tends to be small to medium. Numerical accuracy and algorithmic
-optimization are some of the paramount requirements of most scientific
-software. Advantages of a smaller codebase and requirements include stability,
-ease of installation, configurability and integration into larger systems.
-While the software can stay stable and require relatively low maintenance, the
-responsibility is often on the shoulders of a very few developers. Community
-participation in the development tends to be low. Development is linear and
-simplistic with a limited scope to follow software best practices. Domain
-science oriented software developed to solve  specific problems, e.g.,
-\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability
-of such software is often a significant challenge. Many submissions reported the
-software considered a `byproduct' of the actual research.  Others contended that
-the software was not the main funded part of their research. One exception was
-\emph{Kitware}, which while being a software product specializing in scientific
-process, has a core focus of developing communities around software processes.
-One example of this process is the development of the \emph{CMake} build
-utility, which started out as a building tool for \emph{ITK} but grew to become
-a generic build utility for C++ projects.
+Scientific software consists primarily of special-purpose software that was
+developed for a target use-case scenario or keeping some fixed/frozen
+requirements in mind. Software projects pertaining to specific scientific
+domain often tend to be in a niche and the user community tends to be small to
+medium. Numerical accuracy and algorithmic optimization are some of the
+paramount requirements of most scientific software. Advantages of a smaller
+codebase and requirements include stability, ease of installation,
+configurability and integration into larger systems. While the software can
+stay stable and require relatively low maintenance, the responsibility is often
+on the shoulders of a very few developers. Community participation in the
+development tends to be low. Development is linear and simplistic with a
+limited scope to follow software best practices. Domain science oriented
+software is developed to solve  specific problems, e.g., \emph{DUNE},
+\emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such
+software is often a significant challenge. Many submissions reported the
+software is considered a `byproduct' of the actual research.  Others contended
+that the software was not the main funded part of their research. One exception
+was \emph{Kitware}, which while being a software product specializing in the
+scientific process, has a core focus of developing communities around software
+processes. One example of this process is the development of the \emph{CMake}
+build utility, which started out as a building tool for \emph{ITK} but grew to
+become a generic build utility for C++ projects.
 

### 2014-02-26 WSSSPE1_report/WSSSPE1_summary_report.tex
add comment

@@
 %TODO: expand
-Infrastructure software is developed to better utilize particular, often new, infrastructures and architectures,
-e.g., \emph{MVAPICH}, \emph{visIT}. Explores uncharted territories. Highest
-risk since the projects are experimental. \note{Dan: so is the only difference here that
-this category is aimed at new systems, where utility software is not?
-If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.
-Something in the framing of this subsection doesn't feel right to me.} Successful projects reap high rewards
-and have longer usage span. By nature, they overlap with the utility software
-as they must serve as a facilitator to run ordinary codes. Long
-requirement-development-test cycles.  Wide community support but tends to be
-handled by specialists rather then end users. Promotes collaborations across
-the breadth (e.g., different departments) and depth (e.g., stakeholders within a
-department) of community, one of the key ingredient of a sustainable process.
-Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
-associated with software projects dealing with new architectures. To leverage
-the power of accelerators such as GPUs and MICs new code and libraries are
-required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
-met with a limited success but nonetheless with many invaluable lessons were
-learned about influential cultural and technical aspects in sustainable software
-development practices.
+Infrastructure software is developed to better utilize particular, often new,
+infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}. Explores
+uncharted territories. Highest risk since the projects are experimental.
+\note{Dan: so is the only difference here that this category is aimed at new
+systems, where utility software is not?  If so, perhaps this should be
+incorporated into that subsection, so there would just be two subsections.
+Something in the framing of this subsection doesn't feel right to me.
+FL: I respectfully disagree the the only reason to develop frameworks is
+to better utilize (new or old) infrastructure. Another really important reason
+is to modularize development, making it much easier for an individual developer/user
+to integrate their own code.}
+Successful projects reap high rewards and have longer usage span. By nature,
+they overlap with the utility software as they must serve as a facilitator to
+run ordinary codes. Long requirement-development-test cycles.  Wide community
+support but tends to be handled by specialists rather then end users. Promotes
+collaborations across the breadth (e.g., different departments) and depth
+(e.g., stakeholders within a department) of community, one of the key
+ingredient of a sustainable process. Ferenbaugh~\cite{Ferenbaugh_WSSSPE}
+discusses the risks and challenges associated with software projects dealing
+with new architectures. To leverage the power of accelerators such as GPUs and
+MICs new code and libraries are required. The experience of efforts as
+described in~\cite{Ferenbaugh_WSSSPE} met with a limited success but
+nonetheless with many invaluable lessons were learned about influential
+cultural and technical aspects in sustainable software development practices.
 

### 2014-02-27 WSSSPE1_report/WSSSPE1_summary_report.tex
reorder/fix some names, hopefully not messing anything up in the process
small fixes
resolving a couple of notes, adding a new one

@@
 software is supposed to be described as part of the NSF data management plan,
-and that in their \note{who is ``they''? OCI? NSF?} definition, data includes software. A comment from Twitter
+and that in NSF's definition, data includes software. A comment from Twitter
 (@biomickwatson) raised the issue that this requires reviewers and funders who
@@
 \note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
+\note{Dan: assuming the previous note is from FL, what do you propose?}
 The question of what was meant by ``sustainability'' in the context of software
@@
 
-\note{We should use either single or double quotations consistently throughout the paper.}
 For the purpose of studying software case studies, we classify the software
 projects discussed in the workshop in three broad categories. The first is 
-`utility software', software tools that enable and/or facilitate the
+utility software, software tools that enable and/or facilitate the
 development of other tools and techniques to carry out scientific work. The second is
-`scientific software', software that was developed with an aim to solve a
-class of scientific problems. The third is `infrastructure software', software
+scientific software, software that was developed with an aim to solve a
+class of scientific problems. The third is infrastructure software, software
 developed to efficiently utilize new research infrastructure.
@@
 
-Lorena A Barba
+Lorena A. Barba
 
@@
 
-Neil Chue Hong
+Neil Chue Hong %surname is Chue Hong, this is the correct order.
 
@@
 
-AllisonGehrke
+Allison Gehrke
 
@@
 
-James Hetherington
-
 Simon Hettrick
@@
 
-Monica Luecke
+Monica L\"{u}cke
 
@@
 
-Eric Van Wyk
+Eric van Wyk
 

### 2014-02-27 WSSSPE1_report/WSSSPE1_summary_report.tex
Added discussion in the policy section, particularly for credit and citation.  Hopefully it is useful.

@@
 
-\section{Policy \note{lead: Colin Venters}} \label{sec:policy}
+\section{Policy \note{lead: Colin Venters, contributors: James Howison}} \label{sec:policy}
 
 \todo{section should include what was presented and what was discussed}
+\todo{Question of extent to which this section (indeed the whole document) should call out to existing work on these questions, or stick to summarizing the papers and the discussion.
+I think we should call out as much as possible --JamesHowison}
 
@@
 
+The policy panel and discussion addressed the question of recognition of work on scientific software and linked that recognition to questions of reward and thus motivation for particular kinds of work on scientific software.
+In short, software work in science was seen to be inadequately visible in ways that ``count'' within the reputation system underlying science.
+In his paper for this workshop, Katz placed software work along with other ``activities that facilitate science but are not currently rewarded or recognized'' \cite{Katz_WSSSPE}.
+Priem and Piwowar argued for the need to ``support all researchers in presenting meaningful impact evidence in tenure, promotion, and funding applications.'' \cite{Priem_WSSSPE}.
+Kneply et al.
+argued that the visibility of software that supported a piece of science ``can have detrimental effects on funding, future library development, and even scientific careers.'' \cite{Knepley_WSSSPE}.
+
+These papers, and the discussion at the workshop, join a nascent literature seeking to understand what drives software work in science and how the reward systems of science thereby shape the type of software work undertaken, including the extent to which developers are motivated to build software for the long-term, for the use of others and whether to work collaboratively or separately \cite{howison_incentives_2013, howison_scientific_2011, bietz_synergizing_2010}.
+Software work is not only motivated by direct citations, but the visibility of software work in the literature is important to those who write software used in science.
+
+Papers and discussion concentrated in three areas: How then ought software work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
+
+Most of the papers in this area focused on visibility of software in scientific papers, since scientific papers are the most widely accepted documentation of achievement in science.
+It was noted that there appear to be no widely accepted standards on how the use of software towards a paper ought to be mentioned, that journals, citation style guides and other guides to scientific conduct are vague about how to describe software.
+To address this, papers spoke of a need for a fixed identifier for software, either directly through a mechanism such as a Digital Object Identifier \cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper written to document the software (and perhaps its creation), a ``software paper'' \cite{Chue_Hong_WSSSPE}.
+
+Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software.
+Knepley et al approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
+They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface \cite{Knepley_WSSSPE}.
+Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited.
+Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency.
+In stead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies.
+Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used \cite{Priem_WSSSPE}.
+Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact.
+Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.
+Work since the workshop has described this process at publications.wikia.com.
+
+Acknowledgement in science does not come only in publications, of course.
+A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting.
+Some progress has been made here.
+Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them.
+Nonetheless there was skepticism that peer review panels would value these contributions in the same was as publications).
+
+Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond) \cite{Priem_WSSSPE}.
+They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.
+
+\todo{convert this sketch to discussion of policy options}	
+	Opportunities for Policy interventions:
+		Funding agencies
+		Journals.  Success of data policies, growth of software policies.
+		Promotion and Tenure. Follow lead of Science of Team Science in surveying policies at universities and other institutes.
+
+\todo{review PiratePad notes for other aspects of credit-giving discussed.}
 \begin{itemize}

### 2014-02-27 WSSSPE1_report/WSSSPE1_summary_report.tex
adding notes

@@
 \item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
-\item James Howison $\langle$\url{james@howison.name}$\rangle$
+\item * James Howison $\langle$\url{james@howison.name}$\rangle$
 \item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
@@
 
-\section{Policy \note{lead: Colin Venters, contributors: James Howison}} \label{sec:policy}
+\section{Policy \note{lead: Colin Venters, contributors: James Howison, Hilmar Lapp}} \label{sec:policy}
 
 \todo{section should include what was presented and what was discussed}
-\todo{Question of extent to which this section (indeed the whole document) should call out to existing work on these questions, or stick to summarizing the papers and the discussion.
+
+\note{Question of extent to which this section (indeed the whole document) should call out to existing work on these questions, or stick to summarizing the papers and the discussion.
 I think we should call out as much as possible --JamesHowison}
+
+\note{Dan: I think this is a report on the workshop, and as we need to refer to other things, that's fine, but the focus should be on what was submitted and discussed.}
 

### 2014-02-28 WSSSPE1_report/WSSSPE1_summary_report.tex
comment

@@
 \subsection{Infrastructure software}
-%TODO: expand
+\todo{expand}
 Infrastructure software is developed to better utilize particular, often new,
@@
 incorporated into that subsection, so there would just be two subsections.
-Something in the framing of this subsection doesn't feel right to me.
-FL: I respectfully disagree the the only reason to develop frameworks is
-to better utilize (new or old) infrastructure. Another really important reason
-is to modularize development, making it much easier for an individual developer/user
-to integrate their own code.}
-Successful projects reap high rewards and have longer usage span. By nature,
-they overlap with the utility software as they must serve as a facilitator to
-run ordinary codes. Long requirement-development-test cycles.  Wide community
-support but tends to be handled by specialists rather then end users. Promotes
-collaborations across the breadth (e.g., different departments) and depth
-(e.g., stakeholders within a department) of community, one of the key
-ingredient of a sustainable process. Ferenbaugh~\cite{Ferenbaugh_WSSSPE}
-discusses the risks and challenges associated with software projects dealing
-with new architectures. To leverage the power of accelerators such as GPUs and
-MICs new code and libraries are required. The experience of efforts as
-described in~\cite{Ferenbaugh_WSSSPE} met with a limited success but
-nonetheless with many invaluable lessons were learned about influential
-cultural and technical aspects in sustainable software development practices.
+Something in the framing of this subsection doesn't feel right to me.  FL:
+I respectfully disagree the the only reason to develop frameworks is to
+better utilize (new or old) infrastructure. Another really important reason
+is to modularize development, making it much easier for an individual
+developer/user to integrate their own code. Ketan: I think in this section
+I wanted to bring up a class of software projects which are undertaken with the
+main aim of tackling new systems and architectures in mind. New infrastructures
+such as GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I
+see many submissions around them in the workshop.}
+Successful projects reap
+high rewards and have longer usage span. By nature, they overlap with the
+utility software as they must serve as a facilitator to run ordinary codes.
+Long requirement-development-test cycles.  Wide community support but tends to
+be handled by specialists rather then end users. Promotes collaborations across
+the breadth (e.g., different departments) and depth (e.g., stakeholders within
+a department) of community, one of the key ingredient of a sustainable process.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
+associated with software projects dealing with new architectures. To leverage
+the power of accelerators such as GPUs and MICs new code and libraries are
+required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
+met with a limited success but nonetheless with many invaluable lessons were
+learned about influential cultural and technical aspects in sustainable
+software development practices.
 

### 2014-02-28 WSSSPE1_report/WSSSPE1_summary_report.tex
merge, todo: smoothen and expand

--- 2014-02-27 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 For the purpose of studying software case studies, we classify the software
-projects discussed in the workshop in three broad categories. The first is 
-utility software, software tools that enable and/or facilitate the
-development of other tools and techniques to carry out scientific work. The second is
-scientific software, software that was developed with an aim to solve a
-class of scientific problems. The third is infrastructure software, software
-developed to efficiently utilize new research infrastructure.
-
-These categories have some overlap, showing in some cases where
-case studies do not cleanly fit into a single category.
-Our classification is based on two criteria: first, the character
-of a software project that dominates and where it `seems' to be the best fit, and, second,
-the the workshop discussion session where the paper about
-the project was discussed.
-
-%and point it out in our
-%discussion.
+projects discussed in the workshop in two broad categories. The first is
+utility software, software tools that enable and/or facilitate the development
+of other tools and techniques to carry out scientific work. This includes the
+software developed to efficiently utilize new research infrastructures. The
+second is scientific software, software that was developed with an aim to solve
+a class of scientific problems.
+
+These categories have some overlap, showing in some cases where case studies do
+not cleanly fit into a single category.  Our classification is based on two
+criteria: first, the character of a software project that dominates and where
+it `seems' to be the best fit, and, second, the the workshop discussion session
+where the paper about the project was discussed.
 
@@
 from~\cite{Prlic_WSSSPE}.
+
+%\subsection{Infrastructure software}
+\todo{expand}
+Utility software is often developed aimed at better utilizing particular, new
+infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
+Development in many cases is often a high risk/reward undertaking and explores
+uncharted territories.
+
+%
+%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?
+If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.
+Something in the framing of this subsection doesn't feel right to me.
+FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure.
+Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code.
+Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind.
+New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc.
+were on my mind as I see many submissions around them in the workshop.}
+%
+Successful projects reap high rewards and have longer usage span. Long
+requirement-development-test cycles.  Wide community support but tends to be
+handled by specialists rather then end users. Promotes collaborations across
+the breadth (e.g., different departments) and depth (e.g., stakeholders within
+a department) of community, one of the key ingredient of a sustainable process.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
+associated with software projects dealing with new architectures. To leverage
+the power of accelerators such as GPUs and MICs new code and libraries are
+required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
+met with a limited success but nonetheless with many invaluable lessons were
+learned about influential cultural and technical aspects in sustainable
+software development practices.
 
@@
 
-\subsection{Infrastructure software}
-%TODO: expand
-Infrastructure software is developed to better utilize particular, often new,
-infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}. Explores
-uncharted territories. Highest risk since the projects are experimental.
-\note{Dan: so is the only difference here that this category is aimed at new
-systems, where utility software is not?  If so, perhaps this should be
-incorporated into that subsection, so there would just be two subsections.
-Something in the framing of this subsection doesn't feel right to me.
-FL: I respectfully disagree the the only reason to develop frameworks is
-to better utilize (new or old) infrastructure. Another really important reason
-is to modularize development, making it much easier for an individual developer/user
-to integrate their own code.}
-Successful projects reap high rewards and have longer usage span. By nature,
-they overlap with the utility software as they must serve as a facilitator to
-run ordinary codes. Long requirement-development-test cycles.  Wide community
-support but tends to be handled by specialists rather then end users. Promotes
-collaborations across the breadth (e.g., different departments) and depth
-(e.g., stakeholders within a department) of community, one of the key
-ingredient of a sustainable process. Ferenbaugh~\cite{Ferenbaugh_WSSSPE}
-discusses the risks and challenges associated with software projects dealing
-with new architectures. To leverage the power of accelerators such as GPUs and
-MICs new code and libraries are required. The experience of efforts as
-described in~\cite{Ferenbaugh_WSSSPE} met with a limited success but
-nonetheless with many invaluable lessons were learned about influential
-cultural and technical aspects in sustainable software development practices.
-
 Summary of above discussed cases: lessons, trends, commonalities, is there an

### 2014-03-06 WSSSPE1_report/WSSSPE1_summary_report.tex
improve text

@@
 For the purpose of studying software case studies, we classify the software
-projects discussed in the workshop in two broad categories. The first is
-utility software, software tools that enable and/or facilitate the development
-of other tools and techniques to carry out scientific work. This includes the
-software developed to efficiently utilize new research infrastructures. The
-second is scientific software, software that was developed with an aim to solve
-a class of scientific problems.
-
-These categories have some overlap, showing in some cases where case studies do
-not cleanly fit into a single category.  Our classification is based on two
-criteria: first, the character of a software project that dominates and where
-it `seems' to be the best fit, and, second, the the workshop discussion session
-where the paper about the project was discussed.
-
-Our discussion of the case studies is aimed at understanding the practicalities of
-the many points discussed in the workshop, and we hope to extract the best
-practices and associate them with the projects. We discuss how the cases are
-applicable to and have impacted user communities over their lifetime, isolating
-similarities and differences between projects. We try to define the takeaways for a new
-project and the lessons learned. We try to understand how the course of current and
-future development projects might be altered to implement lessons learned.
+projects discussed in the workshop in two broad categories. First, the utility
+software, comprising software tools that enable and/or facilitate the
+development of other tools and techniques to carry out scientific work. This
+includes the software developed to efficiently utilize new research
+infrastructures.  Second, the scientific software, comprising software that was
+developed with an aim to solve a class of scientific problems.
+
+These categories have overlap in some cases where case studies do not cleanly
+fit into a single category.  Our classification is based on two criteria:
+first, the character of a software project that dominates and where it `seems'
+to be the best fit, and, second, the workshop discussion session where the
+paper about the project was discussed.
+
+Our discussion is aimed at understanding the practicalities of the points
+discussed in the workshop, and we hope to extract the best practices and
+associate them with the projects. We discuss how the cases are applicable to
+and have impacted user communities over their lifetime, isolating similarities
+and differences between projects. We discuss the takeaways for a new
+project and the lessons learned. We try to understand how the course of current
+and future development projects might be altered to implement lessons learned.
 
 \subsection{Utility Software}
-Utility software potentially has the broadest community impact of the three
+Utility software has the potential of the broadest community impact of the two
 categories. This is especially relevant in the modern era of social computing.
-In the context of this paper, a utility software project could be defined as
+In the context of research, a utility software project could be defined as
 general purpose software applicable to one or many generic tasks and/or
@@
 
-%\subsection{Infrastructure software}
-\todo{expand}
 Utility software is often developed aimed at better utilizing particular, new
@@
 uncharted territories. 
-
 %
@@
 Scientific software consists primarily of special-purpose software that was
-developed for a target use-case scenario or keeping some fixed/frozen
-requirements in mind. Software projects pertaining to specific scientific
-domain often tend to be in a niche and the user community tends to be small to
-medium. Numerical accuracy and algorithmic optimization are some of the
-paramount requirements of most scientific software. Advantages of a smaller
-codebase and requirements include stability, ease of installation,
-configurability and integration into larger systems. While the software can
-stay stable and require relatively low maintenance, the responsibility is often
-on the shoulders of a very few developers. Community participation in the
-development tends to be low. Development is linear and simplistic with a
-limited scope to follow software best practices. Domain science oriented
-software is developed to solve  specific problems, e.g., \emph{DUNE},
-\emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such
-software is often a significant challenge. Many submissions reported the
-software is considered a `byproduct' of the actual research.  Others contended
-that the software was not the main funded part of their research. One exception
-was \emph{Kitware}, which while being a software product specializing in the
-scientific process, has a core focus of developing communities around software
-processes. One example of this process is the development of the \emph{CMake}
-build utility, which started out as a building tool for \emph{ITK} but grew to
-become a generic build utility for C++ projects.
+developed for a target use-case scenario or fixed/frozen requirements in mind.
+Software projects pertaining to specific scientific domain often tend to be in
+a niche and the user community tends to be small to medium. Specific
+requirements such as numerical accuracy and algorithmic optimization are some
+of the paramount requirements of most scientific software. The resulting
+codebase is relatiely small. A smaller codebase and fixed requirements result
+in stability, ease of installation, and configurability. Many such projects are
+treated as libraries used into larger systems such as some of the utility
+software discussed in the previous section. While the software can stay stable
+and require relatively low maintenance, the responsibility is often on the
+shoulders of a very few developers. Community participation in the development
+tends to be low. Development is linear and simplistic with a limited scope to
+follow software best practices. Domain science oriented software is developed
+to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc},
+\emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such software is often a
+significant challenge. Many submissions reported the software is considered a
+`byproduct' of the actual research.  Others contended that the software was not
+the main funded part of their research. One exception was \emph{Kitware}, which
+while being a software product specializing in the scientific process, has a
+core focus of developing communities around software processes. One example of
+this process is the development of the \emph{CMake} build utility, which
+started out as a building tool for \emph{ITK} but grew to become a generic
+build utility for C++ projects.
 

### 2014-03-14 WSSSPE1_report/WSSSPE1_summary_report.tex
commenting a note
Adding more papers from Karen's summary.

@@
 
-There were a number of papers categorized under the Communities banner and so two presenters each summarized half of the submissions in this area.
+There were a number of papers categorized under the Communities banner and so
+two presenters each summarized half of the submissions in this area.
 
@@
 
-The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on
-``technology catalysts" and their role in the modern
-scientific research process. Technology catalyst could be defined as a role
-played by an individual with a know-how of technological advancements, tasked
-with user engagement with a goal of enacting scientific or engineering
-applications, and using suitable tools and techniques to take advantage of
-technological capabilities and benefiting applications.
+Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
+developing teams of technical specialists able to overcome a lack of
+coordination between projects.  Maximization of scientific output requires
+maximizing the usability of the scientific codes while minimizing the cost of
+developing and supporting those codes.  This included targeting different
+architectures for their software to be deployed, as well as coordination
+between technically-focused individuals and usage of a common scripting
+language between projects.  Instead of fragmenting the development of
+simulation codes across institutions, the paper suggests that a cohesive
+strategy reducing duplication and increasing coordination will broadly increase
+the efficiency across institutions.  The approach proposed is of de-fragmenting
+the existing ecosystem in a non-disruptive way.
+
+The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
+catalysts" and their role in the modern scientific research process. Technology
+catalyst could be defined as a role played by an individual with a know-how of
+technological advancements, tasked with user engagement with a goal of enacting
+scientific or engineering applications, and using suitable tools and techniques
+to take advantage of technological capabilities and benefiting applications.
 
@@
 
-\note{More papers are needed here.}
+In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
+source software development at Kitware.  In particular, they focus on their
+success at growing their community of users through enabling multiple channels
+of communication, directly reaching out to individuals, and lowering the
+barrier to entry for contributions.  This involves providing clear,
+test-oriented and review-based mechanisms for evaluating contributions,
+permissive licenses, and a service-based model for sustaining development.
+This model enables Kitware to review both public funding, as well as private
+funding to support improvements and targeted developments of the software.
+
+At NESCent, a combination of in-house informatics individuals and domain
+scientists collaborate to develop software to study evolutionary science.  The
+report, \cite{Cranston_WSSSPE}, studied the success of a ``hackathon'' model
+for development, where short-form, hands-on events combining users,
+researcher-developers and software engineers targeted specific code
+improvements.  From this experiment, the authors identified sweveral key
+outcomes as well as lessons-learned -- specifically, the co-localization of
+developers was seen as having a strong impact, enabling casual conversation
+that led to discrete outcomes.  The formation of the discussion mailing list,
+specifically in response to the social capital built at the hackathon, was seen
+as spurring on longer-term benefits to the community and fostering
+sustainability.
+
+\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
+collaborations supporting tool development for Open Science.  This software
+collective, organized around the statistical programming environment R,
+develops access mechanisms for data repositories and attempts to reduce the
+barrier to entry for individuals wanting to access data repositories and study
+the data contained therein.  The collective fosters direct collaboration
+between individuals and data providers, designed to ``train academics in
+reproducible science workflows focused around R.''  The two central challenges
+to this goal were identified as engagement of existing users within ecology and
+evolutionary biology (EEB), and how can the community make inroads and traction
+in other disciplines?  Currently, the collective is exploring addressing these
+challenges through use of social media and holding workshops and hackathons.
+This helps to both raise the profile of the collective within EEB and in other
+domains.  However, the overarching challenge identified in the paper was that
+of incentivizing maintenance of software, which is difficult in academia.
 
@@
 
-Summary of above discussed cases: lessons, trends, commonalities, is there an
-ideal sustainability scenario?
+%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
 

### 2014-03-14 WSSSPE1_report/WSSSPE1_summary_report.tex
make sure citations don't break lines

@@
 incentive models for open contributions and tool builders, for
-example, meaningful metrics and research grants such as
-\cite{NSF_software_vision}. He urged computational scientists to
+example, meaningful metrics and research grants such as~\cite{NSF_software_vision}. He urged computational scientists to
 collaborate and share often their research reports, teaching
@@
   Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath \cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable
-  \cite{Broman_WSSSPE}
+  FASTMath~\cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
 
 \item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
 
 \item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?
-  \cite{Stephan_WSSSPE}
+  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
 
@@
   Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development
-  \cite{Gaston_WSSSPE}
+  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
 
 \item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks
-  \cite{Dubey_WSSSPE}
+  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
 
 \item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
+  Scientific Software Development~\cite{Blatt_WSSSPE}
 
@@
   Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
+  with Sustainable Software~\cite{Ahern_WSSSPE}
 
 \item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
 
@@
   adapting sustainable scientific software development processes in a
-  small biology lab \cite{Crusoe_WSSSPE}
+  small biology lab~\cite{Crusoe_WSSSPE}
 
@@
 The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC \cite{Panda_WSSSPE}
+Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
 
@@
   H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes \cite{Heien_WSSSPE}
+  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
 
@@
   Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software \cite{Trainer_WSSSPE}
+  in Scientific Software~\cite{Trainer_WSSSPE}
 
@@
   Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software \cite{Cohen_WSSSPE}
+  Software~\cite{Cohen_WSSSPE}
 
 \item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications
-  \cite{Slawinski_WSSSPE}
+  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
 
@@
 \item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software \cite{Prlic_WSSSPE}
+  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
 
@@
   R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures \cite{Dubey2_WSSSPE}
+  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
 
 \item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult \cite{Carver_WSSSPE}
+  Need Not Be Difficult~\cite{Carver_WSSSPE}
 
@@
   K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability
-  \cite{Stewart_WSSSPE}
+  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
 
 \item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable \cite{Brown_WSSSPE}
+  anything less is unsustainable~\cite{Brown_WSSSPE}
 
@@
   Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms \cite{Swenson_WSSSPE}
+  Platforms~\cite{Swenson_WSSSPE}
 
@@
 \item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model \cite{Calero_WSSSPE}
+  Software Product Sustainability Model~\cite{Calero_WSSSPE}
 
@@
   Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework
-  \cite{Venters_WSSSPE}
+  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
 
 \item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance
-  \cite{Pierce_WSSSPE}
+  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
 
 \item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability \cite{Katz_WSSSPE}
+  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
 
@@
   Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science
-  \cite{Lenhardt_WSSSPE}
+  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
 
 \item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science
-  \cite{Weber_WSSSPE}
+  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
 
@@
 In short, software work in science was seen to be inadequately visible in ways that ``count'' within the reputation system underlying science.
-In his paper for this workshop, Katz placed software work along with other ``activities that facilitate science but are not currently rewarded or recognized'' \cite{Katz_WSSSPE}.
-Priem and Piwowar argued for the need to ``support all researchers in presenting meaningful impact evidence in tenure, promotion, and funding applications.'' \cite{Priem_WSSSPE}.
+In his paper for this workshop, Katz placed software work along with other ``activities that facilitate science but are not currently rewarded or recognized''~\cite{Katz_WSSSPE}.
+Priem and Piwowar argued for the need to ``support all researchers in presenting meaningful impact evidence in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
 Kneply et al.
-argued that the visibility of software that supported a piece of science ``can have detrimental effects on funding, future library development, and even scientific careers.'' \cite{Knepley_WSSSPE}.
-
-These papers, and the discussion at the workshop, join a nascent literature seeking to understand what drives software work in science and how the reward systems of science thereby shape the type of software work undertaken, including the extent to which developers are motivated to build software for the long-term, for the use of others and whether to work collaboratively or separately \cite{howison_incentives_2013, howison_scientific_2011, bietz_synergizing_2010}.
+argued that the visibility of software that supported a piece of science ``can have detrimental effects on funding, future library development, and even scientific careers.''~\cite{Knepley_WSSSPE}.
+
+These papers, and the discussion at the workshop, join a nascent literature seeking to understand what drives software work in science and how the reward systems of science thereby shape the type of software work undertaken, including the extent to which developers are motivated to build software for the long-term, for the use of others and whether to work collaboratively or separately~\cite{howison_incentives_2013, howison_scientific_2011, bietz_synergizing_2010}.
 Software work is not only motivated by direct citations, but the visibility of software work in the literature is important to those who write software used in science.
@@
 It was noted that there appear to be no widely accepted standards on how the use of software towards a paper ought to be mentioned, that journals, citation style guides and other guides to scientific conduct are vague about how to describe software.
-To address this, papers spoke of a need for a fixed identifier for software, either directly through a mechanism such as a Digital Object Identifier \cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper written to document the software (and perhaps its creation), a ``software paper'' \cite{Chue_Hong_WSSSPE}.
+To address this, papers spoke of a need for a fixed identifier for software, either directly through a mechanism such as a Digital Object Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper written to document the software (and perhaps its creation), a ``software paper''~\cite{Chue_Hong_WSSSPE}.
 
@@
 Knepley et al approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
-They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface \cite{Knepley_WSSSPE}.
+They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}.
 Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited.
@@
 In stead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies.
-Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used \cite{Priem_WSSSPE}.
+Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.
 Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact.
@@
 
-Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond) \cite{Priem_WSSSPE}.
+Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
 They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.
@@
   Smith. Accurately Citing Software and Algorithms Used in
-  Publications \cite{Knepley_WSSSPE}
+  Publications~\cite{Knepley_WSSSPE}
 
 \item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project \cite{Priem_WSSSPE}
+  report for every software project~\cite{Priem_WSSSPE}
 
 \item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns \cite{Katz2_WSSSPE}
+  Social and Technological Concerns~\cite{Katz2_WSSSPE}
 
 \item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software
-  \cite{Chue_Hong_WSSSPE}
+  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
 
@@
 \item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
 
@@
   Computational Science: Software Infrastructure and Environments for
-  Reproducible and Extensible Research \cite{Stodden_WSSSPE}
+  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
 
@@
 \item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model \cite{Heiland_WSSSPE}
+  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
 
 \item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
+  Scientific Software~\cite{Blanton_WSSSPE}
 
 \item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software
-  \cite{Huang_WSSSPE}
+  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
 
 \item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher. Developing Systems for API Governance
-  \cite{Wolski_WSSSPE}
+  Dimopoulos, Alexander Pucher. Developing Systems for API Governance~\cite{Wolski_WSSSPE}
 
@@
 
-Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
+Drawing on experiences from high-energy physics,~\cite{Vay_WSSSPE} proposed
 developing teams of technical specialists able to overcome a lack of
@@
 
-In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
+In~\cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
 source software development at Kitware.  In particular, they focus on their
@@
 scientists collaborate to develop software to study evolutionary science.  The
-report, \cite{Cranston_WSSSPE}, studied the success of a ``hackathon'' model
+report~\cite{Cranston_WSSSPE} studied the success of a ``hackathon'' model
 for development, where short-form, hands-on events combining users,
@@
 
-\item Reagan Moore. Extensible Generic Data Management Software
-  \cite{Moore_WSSSPE}
+\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
 
 \item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability
-  \cite{Cranston_WSSSPE}
+  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
 
@@
   D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities \cite{Vay_WSSSPE}
+  Modeling Science Activities~\cite{Vay_WSSSPE}
 
@@
 Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results \cite{Maheshwari_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science \cite{Hart_WSSSPE}
+Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
 
@@
 
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process \cite{Christopherson_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach \cite{Pierce2_WSSSPE}
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
 
@@
   Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community \cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
+  Scientific Community~\cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
 
 \item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
 
 \item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability \cite{Wilkins-Diehr_WSSSPE}
+Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
 
@@
   Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science \cite{Hanwell_WSSSPE}
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
 
@@
 
-\item Anne C. Elster. Software for Science: Some Personal Reflections
-  \cite{Elster_WSSSPE}
+\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
 
 \item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability \cite{Foster_WSSSPE}
+  as a path to software sustainability~\cite{Foster_WSSSPE}
 
@@
 \item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software \cite{Blanton_WSSSPE}
+  Scientific Software~\cite{Blanton_WSSSPE}
 
 \item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development \cite{Blatt_WSSSPE}
+  Scientific Software Development~\cite{Blatt_WSSSPE}
 
@@
   Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC
-  \cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community
-  \cite{Terrel_WSSSPE}
+  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
 
@@
   Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)
-  \cite{Girotto_WSSSPE}
+  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
 
 \item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry \cite{Crawford_WSSSPE}
+  Computational Chemistry~\cite{Crawford_WSSSPE}
 
@@
 \item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures \cite{Ferenbaugh_WSSSPE}
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
 
@@
   Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software \cite{Ahern_WSSSPE}
+  with Sustainable Software~\cite{Ahern_WSSSPE}
 
 \item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software
-  \cite{Choi_WSSSPE}
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
 
 \item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software
-  \cite{Loffler_WSSSPE}
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
 
@@
     \begin{center}
-      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from \cite{lenhardt-wssspe1-panel}.}
+      \caption{Summary of Modeling Sustainability papers from Policy Panel.  Adapted from~\cite{lenhardt-wssspe1-panel}.}
       \label{tab:defining-sustainability}
@@
                 \hline
-Calero, et al. \cite{Calero_WSSSPE}
+Calero, et al.~\cite{Calero_WSSSPE}
 & General notion of software. Not explicitly defined.
@@
                 \hline
-Venters, et al. \cite{Venters_WSSSPE}
+Venters, et al.~\cite{Venters_WSSSPE}
 & Software as science software; increasingly complex; service-oriented computing
@@
                 \hline
-Pierce, et al. \cite{Pierce_WSSSPE}
+Pierce, et al.~\cite{Pierce_WSSSPE}
 & Cyberinfrastructure software
@@
                 \hline
-Katz and Proctor \cite{Katz_WSSSPE}
+Katz and Proctor~\cite{Katz_WSSSPE}
 & E-research infrastructures (i.e. cyberinfrastructure)
@@
                 \hline
-Lenhardt, et al. \cite{Lenhardt_WSSSPE}
+Lenhardt, et al.~\cite{Lenhardt_WSSSPE}
 & Broadly defined as software supporting science
@@
                 \hline
-Weber, et al. \cite{Weber_WSSSPE}
+Weber, et al.~\cite{Weber_WSSSPE}
 & Software broadly defined; a software ecosystem

### 2014-03-22 WSSSPE1_report/WSSSPE1_summary_report.tex
review and small changes to sec 9

@@
 Software projects pertaining to specific scientific domain often tend to be in
-a niche and the user community tends to be small to medium. Specific
-requirements such as numerical accuracy and algorithmic optimization are some
-of the paramount requirements of most scientific software. The resulting
-codebase is relatiely small. A smaller codebase and fixed requirements result
-in stability, ease of installation, and configurability. Many such projects are
-treated as libraries used into larger systems such as some of the utility
-software discussed in the previous section. While the software can stay stable
-and require relatively low maintenance, the responsibility is often on the
-shoulders of a very few developers. Community participation in the development
-tends to be low. Development is linear and simplistic with a limited scope to
-follow software best practices. Domain science oriented software is developed
-to solve  specific problems, e.g., \emph{DUNE}, \emph{PETSc},
-\emph{MINRES-QLP}, \emph{FASTMath}. Sustainability of such software is often a
-significant challenge. Many submissions reported the software is considered a
-`byproduct' of the actual research.  Others contended that the software was not
-the main funded part of their research. One exception was \emph{Kitware}, which
-while being a software product specializing in the scientific process, has a
-core focus of developing communities around software processes. One example of
-this process is the development of the \emph{CMake} build utility, which
-started out as a building tool for \emph{ITK} but grew to become a generic
-build utility for C++ projects.
+a niche and the user community tends to be small to medium.
+%
+Specific needs such as numerical accuracy and algorithmic optimization are some
+of the paramount requirements of most scientific software.
+%
+The resulting codebase is relatively small. A smaller codebase and fixed
+requirements result in stability, ease of installation, and configurability.
+Many such projects are treated as libraries to be used into larger systems such as
+some of the utility software discussed in the previous section.
+%
+While the
+software can stay stable and require relatively low maintenance, the
+responsibility is often on the shoulders of a very few developers.
+%
+Community
+participation in the development tends to be low. Development is linear and
+simplistic with a limited scope to follow software best practices.
+%
+Some examples of such software discussed as part of the workshop are,
+\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
+%
+Sustainability of such software is often a significant challenge. Many
+submissions reported the software is considered a `byproduct' of the actual
+research. Others contended that the software was not the main funded part of
+their research. One exception was \emph{Kitware}, which while being a software
+product specializing in the scientific process, has a core focus of developing
+communities around software processes. An instance of this process is the
+development of the \emph{CMake} build utility, which started out as a building
+tool for \emph{ITK} but grew to become a generic build utility for C++
+projects.
 

### 2014-03-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Draft for the abstract.

@@
 \begin{abstract}
-This technical report discusses the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1). ...
+This technical report discusses the First Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE1).  The content
+is based on the workshop itself, as documented in a set of
+collaborative notes taken during the workshop, including discussion
+about the keynote presentations, three themed panels that focus on
+papers accepted by the workshop, other discussion, cross-cutting
+issues, use cases, and conclusions.  It joins a nascent literature
+seeking to understand what drives software work in science and how the
+reward systems of science thereby shape the type of software work
+undertaken, including the extent to which developers are motivated to
+build software for the long-term, for the use of others, and whether
+to work collaboratively or separately. Unique perspectives were
+captured about issues such as documentation of APIs, software
+deployment, and development practices; licenses on software; career
+paths for scientific software writers; identifiers for software
+through a mechanism such as a Digital Object Identifier or
+publications of ``software papers''; drawing on evidence of software
+contribution, and impact, recorded in online systems such as source
+code repositories like GitHub.  For truly sustainable software there
+should be no endpoint, as the software products continue to be used
+and useful beyond the single institution, grant, and developer or
+development team.
+
 \end{abstract}

### 2014-03-25 WSSSPE1_report/WSSSPE1_summary_report.tex
removing section 7 for now
small wording change

@@
 
-\section{Other Discussion} \label{sec:other}
-
-\note{from the misc questions/answers/observations part of the google doc?}
-
-\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
-
-\todo{perhaps remove this section}
+%\section{Other Discussion} \label{sec:other}
+%
+%\note{from the misc questions/answers/observations part of the google doc?}
+%
+%\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
+%
+%\todo{perhaps remove this section}
 
@@
 For the purpose of studying software case studies, we classify the software
-projects discussed in the workshop in two broad categories. First, the utility
+projects discussed in the workshop in two broad categories. First, utility
 software, comprising software tools that enable and/or facilitate the
@@
 includes the software developed to efficiently utilize new research
-infrastructures.  Second, the scientific software, comprising software that was
+infrastructures.  Second, scientific software, comprising software that was
 developed with an aim to solve a class of scientific problems.

### 2014-03-26 WSSSPE1_report/WSSSPE1_summary_report.tex
minor fixes in communities

@@
 \item Lucas Nussbaum $\langle$\url{lucas.nussbaum@loria.fr}$\rangle$
-\item Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
+\item * Matthew Turk $\langle$\url{matthewturk@gmail.com}$\rangle$
 \end{itemize}
@@
 
-Drawing on experiences from high-energy physics,~\cite{Vay_WSSSPE} proposed
+Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
 developing teams of technical specialists able to overcome a lack of
@@
 
-In~\cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
+In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
 source software development at Kitware.  In particular, they focus on their
@@
 researcher-developers and software engineers targeted specific code
-improvements.  From this experiment, the authors identified sweveral key
+improvements.  From this experiment, the authors identified several key
 outcomes as well as lessons-learned -- specifically, the co-localization of
@@
 to this goal were identified as engagement of existing users within ecology and
-evolutionary biology (EEB), and how can the community make inroads and traction
-in other disciplines?  Currently, the collective is exploring addressing these
+evolutionary biology (EEB), and how the community could make inroads and traction
+in other disciplines.  Currently, the collective is exploring addressing these
 challenges through use of social media and holding workshops and hackathons.
@@
 
-The Christopherson paper outlines the degree to which research relies on high
+Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree to which research relies on high
 quality software. There are often barriers and a lack of suitable incentives
@@
 
-The Christopherson paper reports on the application of Steps 1-3 to a
+Christopherson reports on the application of Steps 1-3 to a
 computational modeling framework developed in the 1990s. Step 1 was realized as
 a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
-5-day hackathon to develop working code, and Step 3 a 3-day hackathon to refine
+5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
 the code based on new requirements. The team worked on small, low-risk units of
@@
 
-The Pierce paper described how science gateways can provide a user-friendly
+Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
 entry to complex cyberinfrastructure. The paper opens with a description of the
@@
 
-The Zentner paper describes experiences and challenges with the large
+Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
 nanoHUB.org community. The authors define community as a ``body of persons of
@@
 
-The Terrel paper describes support for the Python scientific community through
+Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
 two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
@@
 
-The L\"{o}ffler paper describes the Cactus project which was started in 1996 by
+L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project which was started in 1996 by
 participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
@@
 
-The Wilkins-Diehr paper describes an NSF software institute effort to build a
+Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
 community of those creating science gateways or web portals for science. These

### 2014-03-26 WSSSPE1_report/WSSSPE1_summary_report.tex
shuffling of some sentences in sec 9

@@
 Some examples of such software discussed as part of the workshop are,
-\emph{DUNE}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
+\emph{DUNE}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
+Most of the above efforts are focused on one scientific and/or applied
+mathematics domain. One exception was \emph{Kitware}, which while being a
+software product specializing in the scientific process, has a core focus of
+developing communities around software processes. An instance of this process
+is the development of the \emph{CMake} build utility, which started out as a
+building tool for \emph{ITK} but grew to become a generic build utility for C++
+projects.
 %
-Sustainability of such software is often a significant challenge. Many
-submissions reported the software is considered a `byproduct' of the actual
-research. Others contended that the software was not the main funded part of
-their research. One exception was \emph{Kitware}, which while being a software
-product specializing in the scientific process, has a core focus of developing
-communities around software processes. An instance of this process is the
-development of the \emph{CMake} build utility, which started out as a building
-tool for \emph{ITK} but grew to become a generic build utility for C++
-projects.
+Sustainability of scientific research software is often a significant
+challenge. Many submissions reported that software is practically considered a
+`byproduct' of the actual research. Others contended that the software was not
+the main funded part of their research.
 

### 2014-03-27 WSSSPE1_report/WSSSPE1_summary_report.tex
reading and some small changes

@@
 deployment, and development practices; licenses on software; career
-paths for scientific software writers; identifiers for software
+paths for scientific software developers; identifiers for software
 through a mechanism such as a Digital Object Identifier or
@@
 discussion, and the workshop has fed into the collaborative writing of this
-journal publication. An estimate of 90 to 150 participants were present
+journal publication \note{Ketan: we say report in abstract}. An estimate of 90 to 150 participants were present
 throughout the day. Additional papers based on extended versions of workshop
@@
 category were appointed as panelists, and assigned to read a subset of
-the paper in that category and then discuss them in the panel.
+the papers in that category and then discuss them in the panel.
 

### 2014-03-27 WSSSPE1_report/WSSSPE1_summary_report.tex
more reading and small changes

@@
 
-Papers and discussion concentrated in three areas: How then ought software work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
+Papers and discussion concentrated in three areas: How then \note{something missing here in this sentence --ketan} ought software  work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
 
@@
 Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software.
-Knepley et al approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
+Knepley et.
+al.
+approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
 They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}.
@@
 Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency.
-In stead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies.
+Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies.
 Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.
@@
 Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.
-Work since the workshop has described this process at publications.wikia.com.
+Work since the workshop has described this process at \textit{publications.wikia.com}.
 
@@
 Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them.
-Nonetheless there was skepticism that peer review panels would value these contributions in the same was as publications).
+Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications).
 
@@
 The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
-catalysts" and their role in the modern scientific research process. Technology
-catalyst could be defined as a role played by an individual with a know-how of
-technological advancements, tasked with user engagement with a goal of enacting
+catalysts" and their role in the modern scientific research process. A technology
+catalyst can be defined as a role played by an individual with a know-how of
+technological advancements, tasked with user-engagement with a goal of enacting
 scientific or engineering applications, and using suitable tools and techniques
-to take advantage of technological capabilities and benefiting applications.
-
-One of the tasks of catalysts is to seek community collaborations for new
+to take advantage of technological capabilities thus benefiting applications.
+
+One of the tasks of technology catalysts is to seek community collaborations for new
 applications and user engagement thus benefiting both: science, by effective
@@
 quality software. There are often barriers and a lack of suitable incentives
-for researchers to embrace software engineering principals. The Water Science
+for researchers to embrace software engineering principles. The Water Science
 Software Institute is working to lower some of these barriers through an Open
 Community Engagement Process. This is a 4-step iterative development process
-that incorporates Agile development principals.
+that incorporates Agile development principles.
 
@@
 
-One key to SciGaP is its openness. While based on the Apache Airavata project
+One key to Sci-GaP is its openness. While based on the Apache Airavata project
 and the CIPRES Workbench Framework, community contributions are encouraged

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
reading and list of notes and todos
addressing Ketan's comments, and a few other small changes

@@
 \begin{abstract}
-This technical report discusses the First Workshop on Sustainable
+This document discusses the First Workshop on Sustainable
 Software for Science: Practice and Experiences (WSSSPE1).  The content
-is based on the workshop itself, as documented in a set of
+is based on the workshop itself, as recorded in a set of
 collaborative notes taken during the workshop, including discussion
@@
 captured about issues such as documentation of APIs, software
-deployment, and development practices; licenses on software; career
+deployment, and development practices; software licenses; career
 paths for scientific software developers; identifiers for software
 through a mechanism such as a Digital Object Identifier or
-publications of ``software papers''; drawing on evidence of software
-contribution, and impact, recorded in online systems such as source
+publication of a ``software paper''; and evidence of software
+contribution and impact, recorded in online systems such as source
 code repositories like GitHub. For truly sustainable software there
@@
 discussion, and the workshop has fed into the collaborative writing of this
-journal publication \note{Ketan: we say report in abstract}. An estimate of 90 to 150 participants were present
+document. An estimate of 90 to 150 participants were present
 throughout the day. Additional papers based on extended versions of workshop
@@
 about the keynote presentations (\S\ref{sec:keynotes}), the three
-panels (\S\ref{sec:devel}-\ref{sec:community}), other discussion
-(\S\ref{sec:other}), cross-cutting issues (\S\ref{sec:cross-cutting}),
+panels (\S\ref{sec:devel}-\ref{sec:community}), 
+cross-cutting issues (\S\ref{sec:cross-cutting}),
 use cases (\S\ref{sec:use-cases}), and conclusions
@@
 
-Papers and discussion concentrated in three areas: How then \note{something missing here in this sentence --ketan} ought software  work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
+Papers and discussion concentrated in three areas: How ought software work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
 
@@
 \item Is the functionality and usability clearly explained to new users?
-\item Do users have a mechanism to ask questions and to learn about the element?
+\item Do users have a mechanism to ask questions and to learn about the software?
 \item Will the functionality be correct in the future, even if the environment changes?
@@
 
+% == Notes and ToDos ==
+% 1.
+Why do we classify software cases into general purpose and special purpose?
+what is different?
+What are the sustainability challenges each face?
+Are they radically different?
+What kind of research do each category support?
+Are there separate questions to ask and answer for each category?
+Can we assume that the general purpose software has a wider scope and user community and challenges specific to these kinds of environments are not relavent in the special purpose scientific software which tend to be developed in the small?
+
+% 2.
+What are the factors that has driven the development of these cases?
+How was the development started?
+What was the spark?
+Was there a design approach followed?
+Is the software modular/extensible?
+Can we identify unique and common factors and the reasons behind them.
+
+% 3. Can we characterize the process itself for the selected cases: collaborative, seperate.
+
+% 4.
+ToDo: summarize and report on an inclusive list (might not be exhaustive) of all the software products discussed in the workshop.
+(If possible color code the representaion according to a scheme based on the different dimensions pertaining to them.)
+
+% 5. Has one or more of the following best practices played a role in the development: 
+%     a) hosted community source-code and other artefacts repository
+%     b) documentation, wikis and training material
+%     c) mailing lists and web-presence
+%     d) development process: agile, hackathons, collaborative?
+%     e) regression tests, continuous integration, quality dashboard, code review
+
+% 6. Is there a conscious effort to deal with issues such as backward compatibility, ...
+
+% 7. Career path: are the original developer(s) still around? Is most of the discussion around career path issues a myth after all?
+
+% 8. Is there a software paper for this software product? Cite?
+
+% 9.
+Has the project proved to be sustainable so far?
+Does the developers, managers make conscious efforts to make sure the project is sustainable?
+What determines the sustainability of the project in question: science, publications, usability, funding, reusability?
+Perhaps all factors to some extent?
+What factors are dominating?
+Can we send out a message to current scientific project to emphasize on one or more of these or other aspects?
+% ====
+
 For the purpose of studying software case studies, we classify the software

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
polishing sec8

@@
 % == Notes and ToDos ==
-% 1.
-Why do we classify software cases into general purpose and special purpose?
-what is different?
-What are the sustainability challenges each face?
-Are they radically different?
-What kind of research do each category support?
-Are there separate questions to ask and answer for each category?
-Can we assume that the general purpose software has a wider scope and user community and challenges specific to these kinds of environments are not relavent in the special purpose scientific software which tend to be developed in the small?
-
-% 2.
-What are the factors that has driven the development of these cases?
-How was the development started?
-What was the spark?
-Was there a design approach followed?
-Is the software modular/extensible?
-Can we identify unique and common factors and the reasons behind them.
-
-% 3. Can we characterize the process itself for the selected cases: collaborative, seperate.
-
-% 4.
-ToDo: summarize and report on an inclusive list (might not be exhaustive) of all the software products discussed in the workshop.
-(If possible color code the representaion according to a scheme based on the different dimensions pertaining to them.)
+% 1. Why do we classify software cases into general purpose and special
+% purpose? what is different? What are the sustainability challenges each face?
+% Are they radically different? What kind of research do each category support?
+% Are there separate questions to ask and answer for each category? Can we
+% assume that the general purpose software has a wider scope and user community
+% and challenges specific to these kinds of environments are not relavent in
+% the special purpose scientific software which tend to be developed in the
+% small?
+
+% 2. What are the factors that has driven the development of these cases? How
+% was the development started? What was the spark? Was there a design approach
+% followed? Is the software modular/extensible? Can we identify unique and
+% common factors and the reasons behind them.
+
+% 3. Can we characterize the process itself for the selected cases:
+% collaborative, seperate.
+
+% 4. ToDo: summarize and report on an inclusive list (might not be exhaustive)
+% of all the software products discussed in the workshop. (If possible color
+% code the representaion according to a scheme based on the different
+% dimensions pertaining to them.)
 
@@
 
-% 6. Is there a conscious effort to deal with issues such as backward compatibility, ...
-
-% 7. Career path: are the original developer(s) still around? Is most of the discussion around career path issues a myth after all?
+% 6. Is there a conscious effort to deal with issues such as backward
+% compatibility, ...
+
+% 7. Career path: are the original developer(s) still around? Is most of the
+% discussion around career path issues a myth after all?
 
@@
 
-% 9.
-Has the project proved to be sustainable so far?
-Does the developers, managers make conscious efforts to make sure the project is sustainable?
-What determines the sustainability of the project in question: science, publications, usability, funding, reusability?
-Perhaps all factors to some extent?
-What factors are dominating?
-Can we send out a message to current scientific project to emphasize on one or more of these or other aspects?
+% 9. Has the project proved to be sustainable so far? Does the developers,
+% managers make conscious efforts to make sure the project is sustainable? What
+% determines the sustainability of the project in question: science,
+% publications, usability, funding, reusability? Perhaps all factors to some
+% extent? What factors are dominating? Can we send out a message to current
+% scientific project to emphasize on one or more of these or other aspects?
 % ====
 
-For the purpose of studying software case studies, we classify the software
-projects discussed in the workshop in two broad categories. First, utility
-software, comprising software tools that enable and/or facilitate the
-development of other tools and techniques to carry out scientific work. This
-includes the software developed to efficiently utilize new research
-infrastructures.  Second, scientific software, comprising software that was
-developed with an aim to solve a class of scientific problems.
-
-These categories have overlap in some cases where case studies do not cleanly
-fit into a single category.  Our classification is based on two criteria:
-first, the character of a software project that dominates and where it `seems'
-to be the best fit, and, second, the workshop discussion session where the
-paper about the project was discussed.
-
-Our discussion is aimed at understanding the practicalities of the points
+% TODO: Why are we doing case studies?
+Many important issues related to scientific software sustainability were
+discussed in the previous sections. In this section we discuss some of the
+software projects as case-studies to better understand the relevance of those
+issues and how are they connected to the software sustainability and related
+factors.
+
+We classify the software projects discussed in the workshop in two broad
+categories. First, the utility software, comprising of software that enable
+and/or facilitate the development of other tools and techniques to carry out
+scientific work. This includes the software developed to efficiently utilize
+new research infrastructures. Second, scientific software, comprising software
+that was developed with an aim to solve a specific scientific problem.
+
+% Why are we classifying software like above?
+This classification is motivated by our argument that the software projects
+wildly vary in factors such as scope, purpose and usage. The development and
+management of software is significantly influenced by these factors.
+Consequently, the sustainability challenges faced by software differ and must
+be treated separately. For instance, the challenges faced by a gateway software
+development project such as \textit{CIPRES} are distinct to a niche software
+for ab initio modeling and simulation such as \textit{VASP}.
+
+% Below: What are the goals of this discussion? TODO: Better motivate. May be make a bried list of items discussed in the workshop.
+The discussion is aimed at understanding the practicalities of the key points
 discussed in the workshop, and we hope to extract the best practices and
-associate them with the projects. We discuss how the cases are applicable to
+associated with the case studies. We discuss how the cases are applicable to
 and have impacted user communities over their lifetime, isolating similarities
-and differences between projects. We discuss the takeaways for a new
-project and the lessons learned. We try to understand how the course of current
+and differences between projects. We discuss the takeaways for new
+projects and the lessons learned. We try to understand how the course of current
 and future development projects might be altered to implement lessons learned.
@@
 \subsection{Utility Software}
-Utility software has the potential of the broadest community impact of the two
-categories. This is especially relevant in the modern era of social computing.
-In the context of research, a utility software project could be defined as
-general purpose software applicable to one or many generic tasks and/or
-enabling other software to run by providing a suitable environment. Examples
-are collaborative development frameworks such as \emph{GitHub} and
-\emph{Bitbucket}, workflow and scripting frameworks such as \emph{Galaxy},
-\emph{Swift}, \emph{Globus} and \emph{VisTrails}, and visualization frameworks
-such as \emph{VisIT}. Often projects start with a specific, small set of
-requirements and users in mind but expand to provide more generic functionality
-and appeal to a wider user community. Owing to its generic nature, utility
-software finds a wider user community which in turn leads to a higher
-participation in development. Consequently, the process becomes user-driven and
-self-sustaining.
+Utility software has the potential of a broader user base. This is especially
+relevant in the modern era of social computing. In the context of research, a
+utility software project could be defined as general purpose system applicable
+to one or many generic tasks and/or enabling other software to run by providing
+a suitable environment. Examples are collaborative development frameworks such
+as \emph{GitHub} and \emph{Bitbucket}, distributed and workflow computing
+frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and
+\emph{VisTrails}, and visualization frameworks such as \emph{VisIT}. Often
+projects start with a specific, small set of requirements and users in mind but
+expand to provide functionality and appeal to a wider user community. Owing to
+its generic nature, utility software finds a wider user community which in turn
+leads to a higher participation in development.  Consequently, the process
+becomes user-driven and self-sustaining.
 
@@
 %
+%
 %\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?
@@
 were on my mind as I see many submissions around them in the workshop.}
+%
 %

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
adding context and content

@@
 
-% Below: What are the goals of this discussion? TODO: Better motivate. May be make a bried list of items discussed in the workshop.
-The discussion is aimed at understanding the practicalities of the key points
-discussed in the workshop, and we hope to extract the best practices and
-associated with the case studies. We discuss how the cases are applicable to
-and have impacted user communities over their lifetime, isolating similarities
-and differences between projects. We discuss the takeaways for new
-projects and the lessons learned. We try to understand how the course of current
-and future development projects might be altered to implement lessons learned.
+% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
+We aim to understand the practicalities of the key points discussed in the
+workshop, and hope to extract the best practices associated with the case
+studies. We discuss how the issues are applicable to the cases and try to
+understand how the course of current and future development projects might be
+altered to implement lessons learned. In particular, we discuss the following issues:
+\begin{itemize}
+    \item How closely are the best practices discussed in the workshop were followed?
+    \item Career path: Are the original developers still around?
+    \item What conscious sustainability efforts were undertaken by stakeholders?
+    \item What is the dominant sustainability factor: science, publication, (re)usability?
+\end{itemize}
 
 \subsection{Utility Software}
-Utility software has the potential of a broader user base. This is especially
-relevant in the modern era of social computing. In the context of research, a
-utility software project could be defined as general purpose system applicable
-to one or many generic tasks and/or enabling other software to run by providing
-a suitable environment. Examples are collaborative development frameworks such
-as \emph{GitHub} and \emph{Bitbucket}, distributed and workflow computing
-frameworks such as \emph{Galaxy}, \emph{Swift}, \emph{Globus} and
-\emph{VisTrails}, and visualization frameworks such as \emph{VisIT}. Often
-projects start with a specific, small set of requirements and users in mind but
-expand to provide functionality and appeal to a wider user community. Owing to
-its generic nature, utility software finds a wider user community which in turn
-leads to a higher participation in development.  Consequently, the process
-becomes user-driven and self-sustaining.
-
-Software developed in such a scenario often benefits from
-community driven development, deployment and maintenance best practices. One
-such example is the \emph{Galaxy} project. It follows agile software
+
+We classify software developed with a potentially wider audience and general
+purpose usage in mind as Utility software. In the context of research, a
+utility software could be defined as general purpose system applicable to one
+or many generic tasks and/or enabling other software to run by providing a
+suitable environment. Utility software typically do not address a fundamental
+research challenge in a given scientific domain. Examples are collaborative
+development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
+and workflow computing frameworks such as \emph{Galaxy}, \emph{Swift},
+\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
+\emph{VisIT}.
+
+% What are the sustainability challenges? What are the oppurtunities?
+Development is often a high risk/reward undertaking and explores uncharted
+territories and is largely dominated by (re)usability factors. Owing to
+relatively large number of features, the development and prototype process is
+also lengthy which poses a significant survival risk.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} further discusses the risks and challenges
+associated with software projects dealing with new architectures.
+
+On the other hand, owing to its generic nature, utility software presents
+oppurtunities to be usable by a larger community. This makes it more visible
+which in turn leads to a broader and deeper participation. For
+instance, promoting collaborations across the breadth (e.g. different
+departments) and depth (e.g. stakeholders within a department) of community,
+one of the key ingredient of a sustainable process. Successful projects reap
+high rewards and have longer usage span. Development process becomes
+user-driven and self-sustaining.
+
+One such example is the \emph{Galaxy} project. It follows agile software
 development practices and implements standard practices such as test-driven
@@
 follows the guidelines described in~\cite{Carver_WSSSPE} and many
-from~\cite{Prlic_WSSSPE}.
-
-Utility software is often developed aimed at better utilizing particular, new
-infrastructures and architectures, e.g., \emph{MVAPICH}, \emph{VisIT}.
-Development in many cases is often a high risk/reward undertaking and explores
-uncharted territories.
+from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
+aiming better utilizing a particular, new infrastructure and architecture,
+e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
+emerging architectures such as CPU accelerators new code and libraries are
+required. The experience of one such effort as described
+in~\cite{Ferenbaugh_WSSSPE} met with a limited success but nonetheless with
+many invaluable lessons were learned about influential cultural and technical
+aspects in sustainable software development practices.
+
+A relatively new paradigm in utility software is the software delivered as
+service over the web. With increasing popularity of cloud-based storage and
+computational environments, many users are leaning towards tools used as
+services. \emph{GitHub} and \emph{Bitbucket} can be arguably considered to be
+such tools, catering to collaborative development. For scientific users
+\emph{Globus}-based tools are a case of service-based utility software
+discussed during the workshop. The data movement and sharing services offered
+by Globus can be easily used over the web by collaborating researchers.
 %
 %
-%\note{Dan: so is the only difference here that this category is aimed at new systems, where utility software is not?
-If so, perhaps this should be incorporated into that subsection, so there would just be two subsections.
-Something in the framing of this subsection doesn't feel right to me.
-FL: I respectfully disagree the the only reason to develop frameworks is to better utilize (new or old) infrastructure.
-Another really important reason is to modularize development, making it much easier for an individual developer/user to integrate their own code.
-Ketan: I think in this section I wanted to bring up a class of software projects which are undertaken with the main aim of tackling new systems and architectures in mind.
-New infrastructures such as GPU, Cloud, fast IB, federal science networks, etc.
-were on my mind as I see many submissions around them in the workshop.}
+%\note{Dan: so is the only difference here that this category is aimed at new
+%systems, where utility software is not?  If so, perhaps this should be
+%incorporated into that subsection, so there would just be two subsections.
+%Something in the framing of this subsection doesn't feel right to me.  FL: I
+%respectfully disagree the the only reason to develop frameworks is to better
+%utilize (new or old) infrastructure. Another really important reason is to
+%modularize development, making it much easier for an individual developer/user
+%to integrate their own code. Ketan: I think in this section I wanted to bring
+%up a class of software projects which are undertaken with the main aim of
+%tackling new systems and architectures in mind. New infrastructures such as
+%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
+%many submissions around them in the workshop.}
 %
 %
-Successful projects reap high rewards and have longer usage span. Long
-requirement-development-test cycles.  Wide community support but tends to be
-handled by specialists rather then end users. Promotes collaborations across
-the breadth (e.g., different departments) and depth (e.g., stakeholders within
-a department) of community, one of the key ingredient of a sustainable process.
-Ferenbaugh~\cite{Ferenbaugh_WSSSPE} discusses the risks and challenges
-associated with software projects dealing with new architectures. To leverage
-the power of accelerators such as GPUs and MICs new code and libraries are
-required. The experience of efforts as described in~\cite{Ferenbaugh_WSSSPE}
-met with a limited success but nonetheless with many invaluable lessons were
-learned about influential cultural and technical aspects in sustainable
-software development practices.
-
-Another paradigm in utility software is the software delivered as service over
-the web. With increasing popularity of cloud-based computational environments,
-many users are leaning towards tools used as service. \emph{GitHub} and
-\emph{Bitbucket} can be arguably considered to be such tools, catering to
-collaborative development. For scientific users \emph{Globus}-based tools are
-a case of service-based utility software discussed during the workshop. The
-data movement, authentication and sharing services offered by Globus can be
-easily used over the web by collaborating researchers.
 
 \subsection{Scientific software}
+
 Scientific software consists primarily of special-purpose software that was
-developed for a target use-case scenario or fixed/frozen requirements in mind.
-Software projects pertaining to specific scientific domain often tend to be in
-a niche and the user community tends to be small to medium.
-%
-Specific needs such as numerical accuracy and algorithmic optimization are some
-of the paramount requirements of most scientific software.
-%
-The resulting codebase is relatively small. A smaller codebase and fixed
-requirements result in stability, ease of installation, and configurability.
-Many such projects are treated as libraries to be used into larger systems such as
-some of the utility software discussed in the previous section.
-%
-While the
-software can stay stable and require relatively low maintenance, the
-responsibility is often on the shoulders of a very few developers.
-%
-Community
-participation in the development tends to be low. Development is linear and
-simplistic with a limited scope to follow software best practices.
-%
+purpose-built for a target use-case scenario or fixed/frozen requirements in
+mind.  Software projects pertaining to specific scientific domain often tend to
+be in a niche and the user community tends to be small to medium. They are
+mostly driven by the science and specific needs of a research group.
+
+% What are the challenges?
+Long-term sustainability of scientific research software is often a significant
+challenge. Many submissions reported that software is practically considered a
+`byproduct' of the actual research. Others contended that the software was not
+the main funded part of their research.  Specific needs such as numerical
+accuracy and algorithmic optimization are some of the paramount requirements of
+most scientific software.  The resulting codebase is relatively small. A
+smaller codebase and fixed requirements result in stability, ease of
+installation, and configurability.  Many such projects mature and are treated
+as libraries to be used into larger systems such as some of the utility
+software discussed in the previous section.  While the software can stay stable
+and require relatively low maintenance, the responsibility is often on the
+shoulders of a very few developers who might not be specialists in software
+development. Development tends to be linear and simplistic with a limited
+scope to follow software best practices.
+
+% TODO: What are the oppurtunities?
+
 Some examples of such software discussed as part of the workshop are,
@@
 Most of the above efforts are focused on one scientific and/or applied
-mathematics domain. One exception was \emph{Kitware}, which while being a
+mathematics domain. However, sometimes such projects grow beyond the initial
+vision of developers. One such example is \emph{Kitware}, which while being a
 software product specializing in the scientific process, has a core focus of
@@
 %
-Sustainability of scientific research software is often a significant
-challenge. Many submissions reported that software is practically considered a
-`byproduct' of the actual research. Others contended that the software was not
-the main funded part of their research.
 
 %Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
+It appears that the career paths for utility software developers are broader
+compared to the developers of niche software. However, as projects as a whole,
+the small-scale, specific software have a lower risk and cost in terms of
+sustaining as compared to utility software. The development of scientific
+software requires more scientific skills compared with those of utility
+software. This diminishes the need of specialist programmers and practices in
+many cases where the bulk of development is carried out by a domain scientist.
 

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
adding career path discussion

@@
 Some examples of such software discussed as part of the workshop are,
-\emph{DUNE}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP}, \emph{FASTMath}.
-Most of the above efforts are focused on one scientific and/or applied
-mathematics domain. However, sometimes such projects grow beyond the initial
-vision of developers. One such example is \emph{Kitware}, which while being a
-software product specializing in the scientific process, has a core focus of
-developing communities around software processes. An instance of this process
-is the development of the \emph{CMake} build utility, which started out as a
-building tool for \emph{ITK} but grew to become a generic build utility for C++
-projects.
+\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
+\emph{FASTMath}.  Most of the above efforts are focused on one scientific
+and/or applied mathematics domain. However, sometimes such projects grow beyond
+the initial vision of developers. One such example is \emph{Kitware}, which
+while being a software product specializing in the scientific process, has a
+core focus of developing communities around software processes. An instance of
+this process is the development of the \emph{CMake} build utility, which
+started out as a building tool for \emph{ITK} but grew to become a generic
+build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
+becoming a general purpose utility system usable for solving variety of
+scientific problems.
 %
@@
 %Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
-It appears that the career paths for utility software developers are broader
-compared to the developers of niche software. However, as projects as a whole,
-the small-scale, specific software have a lower risk and cost in terms of
-sustaining as compared to utility software. The development of scientific
-software requires more scientific skills compared with those of utility
-software. This diminishes the need of specialist programmers and practices in
-many cases where the bulk of development is carried out by a domain scientist.
+In order to investigate the much discussed issue of career paths for developers
+of scientific software, we conducted a small survey of the use-cases discussed
+above with respect to the current status of its core developers. We looked at
+the workshop submissions and the original cited papers for the respective
+software. We looked at the webpages and source codes of the earliest versions
+where available to find the original authors. we find that in a majority of
+scientific software more than one original developers are still around, in some
+cases for more than 10 years. Whereas, in the case of utility software
+projects, it appears that at least one original developer is still around often
+in a part-time role. In other cases, it appears that the original developers
+have went on to assume a larger role, again with the same project.
+
+From this survey, it appears that the career paths for utility software
+developers are broader compared to the developers of niche software. However,
+as projects as a whole, the small-scale, specific software have a lower risk
+and cost in terms of sustaining as compared to utility software. Often the
+sustainability of such projects are achieved by the fact that the core
+developer or team heavily utilizes the software for their science, eg.
+\emph{R/qtl, PETSc}. The development of scientific software requires more
+scientific skills compared with those of utility software. This diminishes the
+need of specialist programmers and practices in many cases where the bulk of
+development is carried out by a domain scientist.
 

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes

@@
 development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
-and workflow computing frameworks such as \emph{Galaxy}, \emph{Swift},
-\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
-\emph{VisIT}.
+and workflow and generic computing frameworks such as \emph{Galaxy},
+\emph{nanoHUB}, \emph{SimGrid}, \emph{Swift}, \emph{Globus} and \emph{VisTrails}, and
+visualization frameworks such as \emph{VisIT}.
 
@@
 scientific software more than one original developers are still around, in some
-cases for more than 10 years. Whereas, in the case of utility software
-projects, it appears that at least one original developer is still around often
-in a part-time role. In other cases, it appears that the original developers
-have went on to assume a larger role, again with the same project.
+cases for more than 10 years, (eg. \emph{R/qtl}). Whereas, in the case of
+utility software projects, it appears that at least one original developer is
+still around often in a part-time role. In other cases, it appears that the
+original developers have went on to assume a larger role, again with the same
+project.
 

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
more edits

@@
 the workshop submissions and the original cited papers for the respective
-software. We looked at the webpages and source codes of the earliest versions
-where available to find the original authors. we find that in a majority of
-scientific software more than one original developers are still around, in some
-cases for more than 10 years, (eg. \emph{R/qtl}). Whereas, in the case of
-utility software projects, it appears that at least one original developer is
-still around often in a part-time role. In other cases, it appears that the
-original developers have went on to assume a larger role, again with the same
-project.
+software. Where available, we looked at the webpages and source code of the
+earliest versions to find the original authors. We excluded relatively new
+projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
+that in a majority of scientific software more than one original developers are
+still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
+Whereas, in the case of utility software projects, it appears that at least one
+original developer is still around often in a part-time role. In other cases,
+it appears that the original developers have went on to assume a larger role,
+again with the same project.
 
@@
 and cost in terms of sustaining as compared to utility software. Often the
-sustainability of such projects are achieved by the fact that the core
+sustainability of such projects is achieved by the fact that the core
 developer or team heavily utilizes the software for their science, eg.
-\emph{R/qtl, PETSc}. The development of scientific software requires more
+\emph{R/qtl, PETSc}. Furthermore, the development of scientific software requires more
 scientific skills compared with those of utility software. This diminishes the

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
small change

@@
 be treated separately. For instance, the challenges faced by a gateway software
-development project such as \textit{CIPRES} are distinct to a niche software
-for ab initio modeling and simulation such as \textit{VASP}.
+development project such as \textit{CIPRES} or \textit{VisIT} are distinct to a
+niche software for ab initio modeling and simulation such as \textit{VASP} or
+\textit{Quantum Espresso}.
 

### 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
Just some quick summaries of papers in prep for more systematic summary

@@
 
-\todo{section should include what was presented and what was discussed}
-
-\note{Question of extent to which this section (indeed the whole document) should call out to existing work on these questions, or stick to summarizing the papers and the discussion.
-I think we should call out as much as possible --JamesHowison}
-
-\note{Dan: I think this is a report on the workshop, and as we need to refer to other things, that's fine, but the focus should be on what was submitted and discussed.}
-
 \subsection*{Papers}
@@
   Software Product Sustainability Model~\cite{Calero_WSSSPE}
+
+They examine existing ISO standards about software quality.
+These standards include attributes of the software but currently do not discuss sustainability, in terms of both meanings of sustainability (energy efficiency and what they call perduability, the software "lasting over time".
+In terms of lasting over time, they highlight characteristics of reliabiity, maintainability (e.g., modularity, the degree to which the code is understandable, testable and modifiable) and the degree to which the code can be adapted.
 
@@
 
+They argue that the way that a project is run relates to whether it can continue to serve its users.
+In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input and thus have the potential to "transform passive users into active contributors".
+
+They also argue that open source principles have been mis-applied by the research software community.
+
+Their framework puts together three parts supported by open software: operations and support, escience research and scientific collaboration.
+
+They define open governance as "project deliberations on open, archivable forums ...
+with votes carried out asynchronously over a period of time ..."  This, presumably, contrasts with how governance is done in research software projects at present, through offline meetings amongst grant funded groups, for the very large part.
+
 \item Daniel S. Katz, David Proctor. A Framework for Discussing
   e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
+
+Their framework looks at (for each software element): technical (how does it work with other components), social (who is building it, who is using it) and political context (who is funding it).
+
+The propose three axes on which to locate particular pieces of software: temporal extent over which software should persist, spatial extent (particular lab to global context), Purpose (particular use to general use.).
+They also discuss resourcing models: 1) commercial for profit, 2) open source, 3) closed partnerships and 4) funded projects.
+An additional model can function during a sustaining phase: open source with paid support.
+Finally they introduce two models of governance, top-down and bottom-up.
+
+They hypothesize that each of these models will be appropriate for a different location on their axes and look to additional research to make the connections (for example, government funding might be most appropriate for longer time scales).
 

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
Adding a few paragraphs on "what are communities"

--- 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 \subsection{What are communities?}
+
+The workshop did not directly answer the question ``What are communities?'' but
+instead a number of different answers were indirectly presented, through the
+depiction of individuals and stakeholders in different aspects of the
+scientific software lifecycle.  In broad strokes, however, scientific software
+communities were generally accepted as consisting of individuals, often but not
+always composed of scientist practitioners, that were working with some degree
+of coordination toward a common goal enabled by software.
+
+The discussion of development-focused communities centered around describing
+methods of interaction between individuals and the scientific software.  The
+first type of interaction was the development of a specific piece of software,
+the second was a particular domain or discipline, and the final primary
+type of interaction was around the development of applications built on a
+particular piece of software that was perhaps developed by another group.  As
+an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
+both the for-profit company Kitware and the users and contributors to their
+software packages such as VTK.  This structure, of the centralized development
+of core infrastructure around which communities of individuals applying that
+infrastructure and developing applications utilizing it, was similarly
+reflected in \cite{Terrel_WSSSPE}, where the core scientific python ecosystem
+is supported by a non-profit entity that fosters community investment in that
+ecosystem.  In many ways, these two organizations (Kitware and NumFOCUS)
+attempt to cross domain boundaries and provide support for both the
+infrastructure and application sides of community building.
+
 \subsection{Challenges of community}

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
Updating bibliography.
completed summaries in the frameworks area

--- 2014-03-28 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
+``paper considers how software sustainability can be addressed in the design and engineering process of software systems...it must be considered a composite, non- functional requirement, which can be reasoned over at an architectural level by building and evaluating appropriate architectural models.''
+
+``Software Sustainability Institute define sustainability as ‘software you use today will be available - and continue to be improved and supported - in the future’ which implicitly suggests that sustainability is concerned with concepts of availability, extensibility, and the maintainability of the software [9]''
+
+``software sustainability can be defined as ‘a measure of a systems extensibility, interoperability, maintainability, portability, reusability and scalability''
+
+``For software to be sustainable it needs to be both useful and adaptable as stakeholders requirements, technology and environments evolve and change.''
+
+``Architectural representations of systems are therefore effective in understanding broader system concerns by abstracting away from system details, hence the trend for reasoning about quality attributes at the architectural level''
+
+They look to architectural level analysis to support sustainability, especially using scenario planning
+
+They talk about sustainability going beyond the artifact itself, but then pivot to environmental sustainability.
+Hard to see a framework coming from software engineering dealing with the range of contextual conditions relevant to whether and how a piece of software will continue to be improved and supported in the future.
+They don't mention resource availability at all.
 \item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
@@
 
+Compares the data lifecycle with a software lifecycle, ``suggesting that the well­accepted and well­documented digital science data lifecycle management best practices may inform the development of sustainable scientific software.'' In particular they focus on adding metadata about software through its development, adding the idea of preservation (most likely via validation tests) to the software lifecycle, and assessing ``data in/data out'' connections.
+
 \item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
   Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
+
+``providing some history and provocations for the use of an ecological metaphor to describe software development environments ...
+we want to ask if it can indeed be a productive and thought provoking one.''  They examine historical uses of the ecosystem metaphor (centered around feedback and self-regulation) across domains.
+They note the limits of the metaphor as it encounters human agency within the broad landscape of software work.
+In essence they propose that study and best practices should (in addition to a focus on individual projects) focus on the relationships/dependencies/dynamics across and between projects/components.
+In particular they argue for examining ``niches'' which sustain particular pieces of software, they define ``a software niche as the set of technical requirements, organizational conditions and cultural mores that support its maintenance and use over time.'' They call for better understanding and modeling of niches.
 

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex

--- 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\todo{section should include what was presented and what was discussed}
-
-\note{Question of extent to which this section (indeed the whole document) should call out to existing work on these questions, or stick to summarizing the papers and the discussion.
-I think we should call out as much as possible --JamesHowison}
-
-\note{Dan: I think this is a report on the workshop, and as we need to refer to other things, that's fine, but the focus should be on what was submitted and discussed.}
-
 \subsection*{Papers}
 
+% note that paper summaries are available in an earlier commit 
+% in James Howison's fork:
+% https://github.com/jameshowison/WSSSPE/blob/587705cb6f12f46951e822fdbf5eb72b104fa397/WSSSPE1_report/WSSSPE1_summary_report.tex
+
 \subsubsection*{Modeling Sustainability}
 
-\begin{itemize}
-
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model~\cite{Calero_WSSSPE}
-
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
-
-\end{itemize}
+Frameworks provide a scaffold for thinking; they help us understand which elements are important and how they relate.
+They also help us break down a problem into chunks that can be pursued by different groups or perhaps disciplines.
+
+A group of papers submitted to the workshop provide frameworks with these intentions.
+The frameworks have substantial overlap but with different emphases and extents.
+Accordingly we present them starting with the area of most overlap and proceed to show the different emphases and the range of factors included in the frameworks.
+
+\todo{seems like definitions will be mentioned elsewhere in the paper?}
+Each of the framework papers included a definition of sustainability, with many overlaps.
+In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}.
+This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
+
+One area in which there was not complete overlap was whether the word (and thus the effort called for by the WSSSPE workshop) involved environmental sustainability.
+Of course the word sustainability has strong connotations from consideration of environmental issues, evoking some mention of the areas in which software interacts with overall environmental resource usage, particularly energy efficiency.
+The two papers in this area which mentioned this did so without integrating that analysis into the question of software being around long-term, suggesting that questions of environmental impact of scientific software is a conceptually distinct area of inquiry.
+
+One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality.
+This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.
+Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
+This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al.
+argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact.
+Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
+In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.
+
+A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability.
+The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project).
+For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
+In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''.
+They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.
+Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.
+They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).
+
+Frameworks also addressed themselves to the context in which software projects exist (thus moving upwards in abstraction, from the artifact itself, to the organization of its production and, in these frameworks, the shape of the space which an artifact or project exists in).
+These frameworks take the form of contingency theories in that they outline a different set of challenges and argue that different project organizations (and presumably artifact attributes) are necessary to persist longterm in spaces with particular characteristics.
+Katz and Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in terms of three axes: temporal (long or short term needs), spatial (local or global use) and purpose (specific to general).
+They propose that different project organizations will be needed in different locations and argue that we should concentrate research to understand those connections.
+Weber et al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural ecosystems as ``niches'' which sustain particular pieces of software.
+They define ``a software niche as the set of technical requirements, organizational conditions and cultural mores that support its maintenance and use over time.'' They call for better understanding and modeling of niches (as well as further exploration of the usefulness of ecosystem metaphors).
 
@@
 \subsection{What are communities?}
-
-The workshop did not directly answer the question ``What are communities?'' but
-instead a number of different answers were indirectly presented, through the
-depiction of individuals and stakeholders in different aspects of the
-scientific software lifecycle.  In broad strokes, however, scientific software
-communities were generally accepted as consisting of individuals, often but not
-always composed of scientist practitioners, that were working with some degree
-of coordination toward a common goal enabled by software.
-
-The discussion of development-focused communities centered around describing
-methods of interaction between individuals and the scientific software.  The
-first type of interaction was the development of a specific piece of software,
-the second was a particular domain or discipline, and the final primary
-type of interaction was around the development of applications built on a
-particular piece of software that was perhaps developed by another group.  As
-an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
-both the for-profit company Kitware and the users and contributors to their
-software packages such as VTK.  This structure, of the centralized development
-of core infrastructure around which communities of individuals applying that
-infrastructure and developing applications utilizing it, was similarly
-reflected in \cite{Terrel_WSSSPE}, where the core scientific python ecosystem
-is supported by a non-profit entity that fosters community investment in that
-ecosystem.  In many ways, these two organizations (Kitware and NumFOCUS)
-attempt to cross domain boundaries and provide support for both the
-infrastructure and application sides of community building.
-
 \subsection{Challenges of community}

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 
-\subsection*{Papers}
-
-% note that paper summaries are available in an earlier commit 
-% in James Howison's fork:
-% https://github.com/jameshowison/WSSSPE/blob/587705cb6f12f46951e822fdbf5eb72b104fa397/WSSSPE1_report/WSSSPE1_summary_report.tex
-
-\subsubsection*{Modeling Sustainability}
+\subsection*{Modeling Sustainability}
 
@@
 
-\subsubsection*{Credit, Citation, Impact}
+\subsection*{Credit, Citation, Impact}
 
@@
 \todo{convert this sketch to discussion of policy options}
+\begin{verbatim}
         Opportunities for Policy interventions:
@@
                 Promotion and Tenure. Follow lead of Science of Team Science in surveying policies at universities and other institutes.
-
+\end{verbatim}
 \todo{review PiratePad notes for other aspects of credit-giving discussed.}
+
+\subsection*{Implementing Policy}
+
+\subsection*{Papers}
+
+% note that paper summaries are available in tag jameshowison-paper-summaries: 
+% https://github.com/danielskatz/WSSSPE/blob/jameshowison-paper-summaries/WSSSPE1_report/WSSSPE1_summary_report.tex
+
+\subsubsection*{Modeling Sustainability}
+
+\begin{itemize}
+
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
+  Software Product Sustainability Model~\cite{Calero_WSSSPE}
+
+\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
+  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
+  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
+  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing
+  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
+  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
+  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
+  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Credit, Citation, Impact}
+
 \begin{itemize}

### 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
Very minor edits to Policy section.
Mostly typos and citations.

@@
 This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.
-Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
+Thus Calero and colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perdurability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
 This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al.
 argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact.
-Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
+Lenhardt et al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered).
+In addition, their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
 In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.
@@
 The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project).
-For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
+For example Pierce et al~\cite{Pierce_WSSSPE} focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
 In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''.
 They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.
-Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.
+Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance.
 They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).
@@
 Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software.
-Knepley et.
-al.
-approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
+Knepley et al~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
 They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}.
@@
 Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them.
-Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications).
-
-Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
-They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.
+Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications.
+
+Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like Github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
+They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``altmetrics'' approach.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Adds opening parapraph to Implementing Policy.
Adds Daisie Huang's speaker notes.
Adding them as comments for now. These summarize the papers in this section
(Implementing Policy).
First paragraph in Implementing Policy.

@@
 \subsection*{Implementing Policy}
+
+The Workshop contributions in this group were concerned with the aspect of how implementation of best practices and other recommendations for improving scientific software sustainability could be promoted.
+Specifically, if scientific software is to become more sustainable, corresponding policies and guidelines need to be such that the scientific community can follow and implement them.
+This is considerably more challenging that it might seem at first, because in the reality of science today resources, both financial and personnel, that could be devoted to implementation are very limited, and the reward system does not encourage scientists to do so.
+Furthermore, implementing sustainability-targeting policies and guidelines often takes a variety of specialized software engineering expertises.
+
+% To create sustainable software, we need to look at some key issues.
+% First, we need to acknowledge that as a software package matures, 
+% we will face new types of problems, and 
+% we need to plan for these throughout a package’s life cycle. 
+% Therefore, making scientific software sustainable means that we define policies and guidelines that the scientific community can follow and implement.
+% But the reality of science today is that we have limited resources and rewards to encourage people to follow these policies and guidelines.
+% Implementing policy often takes specialized expertise in software engineering.
+
+% In light of these issues, I’ll be discussing several papers that were contributed to the workshop. 
+% Some of these papers focus on specific facets of software design that are not often addressed in scientific software development, such as 
+% API governance and 
+% Software Security, and 
+% some of the papers discuss strategies to implement all of the different facets of sustainable software design in the framework of scientific software.
+
+% First, Krintz et al from The University of California at Santa Barbara’s Department of Computer Science discuss one of these important issues: developing systems for API governance.
+
+% The authors make the point that scientific research is moving away from local hardware environments towards cloud computing. 
+% Therefore, instead of focusing on access to hardware, we will need to focus on access to the digital assets: the code and the data. 
+% APIs—programming interfaces—are the main link governing interactions between these assets.
+% Because APIs are the main interface between different digital assets, they have to be maintained in a sustainable way.
+
+% The authors focus on understanding the portability and consistency of APIs used to connect these data archives, 
+% because changes in APIs affect accessibility of data. 
+% They define two different types of compatibility, 
+% semantic compatibility and 
+% syntactic compatibility. 
+% They demonstrate an algorithmic method for categorizing a particular API port as “hard” or “easy,” at least for semantic compatibility.
+
+% Next, we’ll look at issues of software security.
+
+% Heiland et al from the Center for Trustworthy Scientific Infrastructure discussed issues related to implementing strong security measures for scientific software. 
+% They point out that cybersecurity is rarely addressed in scientific software design.
+
+% Security considerations for software vary depending on the maturity level of the software package.
+% But when we initially develop scientific software, we generally don’t know what the final maturity level will be.
+% Scientific software developers are probably not aware of best practices for cybersecurity.
+% So the authors introduce the concept of Software Security Maturity Models, such as OpenSAMM and BSI-MM. 
+% These are used in industry to identify and define security vulnerabilities at different stages of the software life cycle.
+
+% They suggest that a similar Software Security Maturity Model can formalize this process:
+% It provides classification of software security practices.
+% It provides a path for tightening security practices as a package’s maturity level increases.
+% It emphasizes understandability over complexity.
+
+% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.
+
+% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.
+
+% The authors focus on a point that has been brought up many times in this context:
+% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
+% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
+% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
+% We don’t have a way to validate that the software used in a paper is actually done right.
+% The best way to get software designed correctly is to make sure best practices are considered from the start.
+
+% The authors highlight two models for sustainable software, at different extremes:
+% One is what they call “co-funding”:
+% In these projects, usually large, multi-year collaborations, 
+% there is equal emphasis on both the science and the software development. 
+% Both are planned into the project from inception. 
+% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.
+
+% At the other extreme, they discuss “software carpentry”: 
+% in this model, it’s assumed that the scientists themselves will write and maintain their code.
+% Groups like Software Carpentry and ROpenSci assume that 
+% scientists won’t have access to dedicated software engineering, 
+% so they try to give them tools to use best practices in their own software development.
+
+% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
+% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.
+
+% What do we mean when we refer to the “long tail” of scientific software?
+% Think of the distribution of resources in scientific software.
+Most are focused on big projects with lots of community buy-in and funding.
+But a lot of scientific software exists away from this model.
+% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
+% Look at MacClade: it was originally released in 1986 and last updated in 2005,
+% but it was still cited over 400 times in 2013!
+% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.
+
+% Another dimension of the long tail can also be found in my particular research domain. 
+% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
+% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
+% Most of these programs are developed by academic scientists…
+% They generally have limited training in software engineering
+% Limited time or career incentive to improve software
+% Limited funding
+
+% So, to summarize a bit:
+% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
+% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
+% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.
+
+% Then we have to consider that the users of scientific software are scientists, 
+% so the developers need to understand the users and the science.
+% This is the idea of a “t-skilled” person: 
+% one who is both well-versed in a scientific domain 
+% and deeply experienced in one or more facets of software engineering. 
+% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.
+
+% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
+% Think of it as analogous to DNA sequencing:
+% Sequencers used to be something that individual labs and institutions had to 
+% buy, maintain, and operate themselves, 
+% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
+% But now, core facilities provide the instrumentation and service to labs of any size. 
+% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.
+
+% We propose that software engineering can be “instrumented” in a similar way. 
+% Let’s create a nonprofit center for scientific software engineering.
+% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
+% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
+% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.
+
+% So, to conclude…
+% Implementing policies to encourage sustainability in scientific software 
+% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
+% But most of them aren’t addressed in the status quo.
+% We’ve highlighted some of these facets today and suggested some possible solutions.
+% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
+% Grassroots developer groups can provide guidance to scientists about best practices in software development.
+% We think there is a place for a software engineering center that can provide 
+% both engineering expertise and guidance 
+% with a contract-driven instrumentation model 
+% to the scientific software in the long tail.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex

--- 2014-03-29 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 \begin{itemize}
-\item academic fields studying communities (and software communities)
+\item academic fields studying communities (and software communities).
 \item courses on online communities
-\item books
 \item need for software carpentry module on organizing communities?
+\item \cite{howison_scientific_2011}
+\item \cite{citeulike:11831265}
+\item \cite{citeulike:7888211}
+\item \cite{citeulike:478633}
+\item \cite{opac-b1134063}
+\item \cite{DBLP:conf/cscw/HowisonH11}
+\item \cite{DBLP:conf/cscw/2011}
+\item \cite{Trapani:2011}
+\item \cite{Allsopp:2012}
+\item \cite{1749-4699-6-1-015010}
+\item \cite{Turk:2013:SCH:2484762.2484782}
 \end{itemize}

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Second paragraph in Implementing Policy.

--- 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-The Workshop contributions in this group were concerned with the aspect of how implementation of best practices and other recommendations for improving scientific software sustainability could be promoted.
+The workshop contributions in this group were concerned with the aspect of how implementation of best practices and other recommendations for improving scientific software sustainability could be promoted.
 Specifically, if scientific software is to become more sustainable, corresponding policies and guidelines need to be such that the scientific community can follow and implement them.
 This is considerably more challenging that it might seem at first, because in the reality of science today resources, both financial and personnel, that could be devoted to implementation are very limited, and the reward system does not encourage scientists to do so.
-Furthermore, implementing sustainability-targeting policies and guidelines often takes a variety of specialized software engineering expertises.
-
-% To create sustainable software, we need to look at some key issues.
-% First, we need to acknowledge that as a software package matures, 
-% we will face new types of problems, and 
-% we need to plan for these throughout a package’s life cycle. 
-% Therefore, making scientific software sustainable means that we define policies and guidelines that the scientific community can follow and implement.
-% But the reality of science today is that we have limited resources and rewards to encourage people to follow these policies and guidelines.
-% Implementing policy often takes specialized expertise in software engineering.
-
-% In light of these issues, I’ll be discussing several papers that were contributed to the workshop. 
-% Some of these papers focus on specific facets of software design that are not often addressed in scientific software development, such as 
-% API governance and 
-% Software Security, and 
-% some of the papers discuss strategies to implement all of the different facets of sustainable software design in the framework of scientific software.
+Furthermore, implementing sustainability-targeting policies and guidelines often takes a variety of specialized software engineering expertises, which are not necessarily found in a single engineer, and much less so in a domain scientist cross-trained in programming.
+Adding to the policy implementation challenges, applicable sustainability-promoting practices and guidelines will change through a software project's lifecycle, in particular as it gains maturity.
+
+Two of the papers in this group focus on specific facets of software design that are important factors in a project's sustainability but are often addressed only late in the scientific software development cycle, if at all: Wolski et al~\cite{} look at API governance, and Heiland et al~\cite{} discuss maturity models for software security.
+The other two papers discuss implementation strategies for science from the perspective of facilitating many or all facets of sustainanability-oriented software design for science:  Blanton and Lenhardt\cite{} contrast large projects that have software infrastructue development built-in with cross-training domain scientist PIs in software engineering best practices.
+Huang and Lapp~\cite{} discuss how various specialized software engineering skills could be turned into shared instrumentation with low barriers to access.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Author order change for Krintz et al.

@@
 
-Two of the papers in this group focus on specific facets of software design that are important factors in a project's sustainability but are often addressed only late in the scientific software development cycle, if at all: Wolski et al~\cite{} look at API governance, and Heiland et al~\cite{} discuss maturity models for software security.
-The other two papers discuss implementation strategies for science from the perspective of facilitating many or all facets of sustainanability-oriented software design for science:  Blanton and Lenhardt\cite{} contrast large projects that have software infrastructue development built-in with cross-training domain scientist PIs in software engineering best practices.
-Huang and Lapp~\cite{} discuss how various specialized software engineering skills could be turned into shared instrumentation with low barriers to access.
+Two of the papers in this group focus on specific facets of software design that are important factors in a project's sustainability but are often addressed only late in the scientific software development cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity models for software security.
+The other two papers discuss implementation strategies for science from the perspective of facilitating many or all facets of sustainanability-oriented software design for science:  Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast large projects that have software infrastructue development built-in with cross-training domain scientist PIs in software engineering best practices.
+Huang and Lapp~\cite{Huang_WSSSPE} discuss how various specialized software engineering skills could be turned into shared instrumentation with low barriers to access.
 
@@
 
-\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher. Developing Systems for API Governance~\cite{Wolski_WSSSPE}
+\item Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Summarized paper 1 in Implementing Policy.

@@
 
-% First, Krintz et al from The University of California at Santa Barbara’s Department of Computer Science discuss one of these important issues: developing systems for API governance.
-
-% The authors make the point that scientific research is moving away from local hardware environments towards cloud computing. 
-% Therefore, instead of focusing on access to hardware, we will need to focus on access to the digital assets: the code and the data. 
-% APIs—programming interfaces—are the main link governing interactions between these assets.
-% Because APIs are the main interface between different digital assets, they have to be maintained in a sustainable way.
-
-% The authors focus on understanding the portability and consistency of APIs used to connect these data archives, 
-% because changes in APIs affect accessibility of data. 
-% They define two different types of compatibility, 
-% semantic compatibility and 
-% syntactic compatibility. 
-% They demonstrate an algorithmic method for categorizing a particular API port as “hard” or “easy,” at least for semantic compatibility.
+Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which computing more and more frequently takes place in a distributed cloud, the control over digital assets is increasingly shifting from physical infrastructure to APIs, in particular web-service APIs.
+Yet, as Krintz et al observe, although science communities have  typically developed solid practices for managing physical IT infrastructure in data centers, corresponding practices and technology for API governance, refered to by Krintz et al as the ``combined policy, implementation, and deployment control'', is far less developed.
+As Krintz et al demonstrate, APIs to digital assets can and do change, sometimes quite frequently.
+raising the need to port dependent applications.
+The effort required for porting is notoriously difficult to estimate, making it nearly impossible for IT organizations to assess and thus properly manage their impact.
+To address this, Krintz et al propose a mechanism that evaluates the porting effort between two versions of a web-service API in a formal and automated way.
+To analyze the porting effort, their method divides API changes into \emph{syntactic similarity}, the changes in inputs and outputs, and into \emph{semantic similarity}, the changes in the API's behavior.
+In initial tests, the method showed good congruence with human developers in scoring porting effort, offering the potential that API governance can become as solid a part of scientific IT management as data center infrastructure management is today.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Small fixes in Implementing Policy.

--- 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-% First, Krintz et al from The University of California at Santa Barbara’s Department of Computer Science discuss one of these important issues: developing systems for API governance.
-
-% The authors make the point that scientific research is moving away from local hardware environments towards cloud computing. 
-% Therefore, instead of focusing on access to hardware, we will need to focus on access to the digital assets: the code and the data. 
-% APIs—programming interfaces—are the main link governing interactions between these assets.
-% Because APIs are the main interface between different digital assets, they have to be maintained in a sustainable way.
-
-% The authors focus on understanding the portability and consistency of APIs used to connect these data archives, 
-% because changes in APIs affect accessibility of data. 
-% They define two different types of compatibility, 
-% semantic compatibility and 
-% syntactic compatibility. 
-% They demonstrate an algorithmic method for categorizing a particular API port as “hard” or “easy,” at least for semantic compatibility.
+Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which computing frequently takes place in a distributed cloud, the control over digital resources is increasingly shifting from physical infrastructure to APIs, in particular web-service APIs.
+Yet, as Krintz et al observe, although science communities have (for the most part) developed solid practices for managing physical IT infrastructure in data centers, corresponding practices and technology for API governance, refered to by Krintz et al as the ``combined policy, implementation, and deployment control'', is far less developed.
+As Krintz et al demonstrate, web APIs can and do change, sometimes quite frequently.
+raising the need to port dependent applications.
+The effort required for porting is notoriously difficult to estimate, making it nearly impossible for IT organizations to assess and thus properly manage the impact of API changes.
+To address this, Krintz et al propose a mechanism that evaluates the porting effort between two versions of a web-service API in a formal and automated way.
+To analyze the porting effort, they divide API changes into \emph{syntactic similarity}, the changes in inputs and outputs, and into \emph{semantic similarity}, the changes in the API's behavior.
+In initial tests, their method showed good congruence with human developers in scoring porting effort, offering the possibility that API governance can become as solid a part of scientific IT management as data center infrastructure management is today.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Reformat to break lines in paragraph.
This is for easier readability in online interface to VCS - I thought Github
already breaks lines in text, but apparently not.

@@
 
-The workshop contributions in this group were concerned with the aspect of how implementation of best practices and other recommendations for improving scientific software sustainability could be promoted.
-Specifically, if scientific software is to become more sustainable, corresponding policies and guidelines need to be such that the scientific community can follow and implement them.
-This is considerably more challenging that it might seem at first, because in the reality of science today resources, both financial and personnel, that could be devoted to implementation are very limited, and the reward system does not encourage scientists to do so.
-Furthermore, implementing sustainability-targeting policies and guidelines often takes a variety of specialized software engineering expertises, which are not necessarily found in a single engineer, and much less so in a domain scientist cross-trained in programming.
-Adding to the policy implementation challenges, applicable sustainability-promoting practices and guidelines will change through a software project's lifecycle, in particular as it gains maturity.
-
-Two of the papers in this group focus on specific facets of software design that are important factors in a project's sustainability but are often addressed only late in the scientific software development cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity models for software security.
-The other two papers discuss implementation strategies for science from the perspective of facilitating many or all facets of sustainanability-oriented software design for science:  Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast large projects that have software infrastructue development built-in with cross-training domain scientist PIs in software engineering best practices.
-Huang and Lapp~\cite{Huang_WSSSPE} discuss how various specialized software engineering skills could be turned into shared instrumentation with low barriers to access.
-
-Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which computing frequently takes place in a distributed cloud, the control over digital resources is increasingly shifting from physical infrastructure to APIs, in particular web-service APIs.
-Yet, as Krintz et al observe, although science communities have (for the most part) developed solid practices for managing physical IT infrastructure in data centers, corresponding practices and technology for API governance, refered to by Krintz et al as the ``combined policy, implementation, and deployment control'', is far less developed.
-As Krintz et al demonstrate, web APIs can and do change, sometimes quite frequently.
-raising the need to port dependent applications.
-The effort required for porting is notoriously difficult to estimate, making it nearly impossible for IT organizations to assess and thus properly manage the impact of API changes.
-To address this, Krintz et al propose a mechanism that evaluates the porting effort between two versions of a web-service API in a formal and automated way.
-To analyze the porting effort, they divide API changes into \emph{syntactic similarity}, the changes in inputs and outputs, and into \emph{semantic similarity}, the changes in the API's behavior.
-In initial tests, their method showed good congruence with human developers in scoring porting effort, offering the possibility that API governance can become as solid a part of scientific IT management as data center infrastructure management is today.
+The workshop contributions in this group were concerned with the
+aspect of how implementation of best practices and other
+recommendations for improving scientific software sustainability could
+be promoted. Specifically, if scientific software is to become more
+sustainable, corresponding policies and guidelines need to be such
+that the scientific community can follow and implement them. This is
+considerably more challenging that it might seem at first, because in
+the reality of science today resources, both financial and personnel,
+that could be devoted to implementation are very limited, and the
+reward system does not encourage scientists to do so. Furthermore,
+implementing sustainability-targeting policies and guidelines often
+takes a variety of specialized software engineering expertises, which
+are not necessarily found in a single engineer, and much less so in a
+domain scientist cross-trained in programming. Adding to the policy
+implementation challenges, applicable sustainability-promoting
+practices and guidelines will change through a software project's
+lifecycle, in particular as it gains maturity.
+
+Two of the papers in this group focus on specific facets of software
+design that are important factors in a project's sustainability but
+are often addressed only late in the scientific software development
+cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
+governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
+models for software security. The other two papers discuss
+implementation strategies for science from the perspective of
+facilitating many or all facets of sustainanability-oriented software
+design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
+large projects that have software infrastructue development built-in
+with cross-training domain scientist PIs in software engineering best
+practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
+specialized software engineering skills could be turned into shared
+instrumentation with low barriers to access.
+
+Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
+computing frequently takes place in a distributed cloud, the control
+over digital resources is increasingly shifting from physical
+infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
+et al observe, although science communities have (for the most part)
+developed solid practices for managing physical IT infrastructure in
+data centers, corresponding practices and technology for API
+governance, refered to by Krintz et al as the ``combined policy,
+implementation, and deployment control'', is far less developed. As
+Krintz et al demonstrate, web APIs can and do change, sometimes quite
+frequently. raising the need to port dependent applications. The
+effort required for porting is notoriously difficult to estimate,
+making it nearly impossible for IT organizations to assess and thus
+properly manage the impact of API changes. To address this, Krintz et
+al propose a mechanism that evaluates the porting effort between two
+versions of a web-service API in a formal and automated way. To
+analyze the porting effort, they divide API changes into
+\emph{syntactic similarity}, the changes in inputs and outputs, and
+into \emph{semantic similarity}, the changes in the API's behavior. In
+initial tests, their method showed good congruence with human
+developers in scoring porting effort, offering the possibility that
+API governance can become as solid a part of scientific IT management
+as data center infrastructure management is today.
 

### 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
Heiland et al summary for Implementing Policy.

--- 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-The workshop contributions in this group were concerned with the aspect of how implementation of best practices and other recommendations for improving scientific software sustainability could be promoted.
-Specifically, if scientific software is to become more sustainable, corresponding policies and guidelines need to be such that the scientific community can follow and implement them.
-This is considerably more challenging that it might seem at first, because in the reality of science today resources, both financial and personnel, that could be devoted to implementation are very limited, and the reward system does not encourage scientists to do so.
-Furthermore, implementing sustainability-targeting policies and guidelines often takes a variety of specialized software engineering expertises, which are not necessarily found in a single engineer, and much less so in a domain scientist cross-trained in programming.
-Adding to the policy implementation challenges, applicable sustainability-promoting practices and guidelines will change through a software project's lifecycle, in particular as it gains maturity.
-
-Two of the papers in this group focus on specific facets of software design that are important factors in a project's sustainability but are often addressed only late in the scientific software development cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity models for software security.
-The other two papers discuss implementation strategies for science from the perspective of facilitating many or all facets of sustainanability-oriented software design for science:  Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast large projects that have software infrastructue development built-in with cross-training domain scientist PIs in software engineering best practices.
-Huang and Lapp~\cite{Huang_WSSSPE} discuss how various specialized software engineering skills could be turned into shared instrumentation with low barriers to access.
-
-Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which computing frequently takes place in a distributed cloud, the control over digital resources is increasingly shifting from physical infrastructure to APIs, in particular web-service APIs.
-Yet, as Krintz et al observe, although science communities have (for the most part) developed solid practices for managing physical IT infrastructure in data centers, corresponding practices and technology for API governance, refered to by Krintz et al as the ``combined policy, implementation, and deployment control'', is far less developed.
-As Krintz et al demonstrate, web APIs can and do change, sometimes quite frequently.
-raising the need to port dependent applications.
-The effort required for porting is notoriously difficult to estimate, making it nearly impossible for IT organizations to assess and thus properly manage the impact of API changes.
-To address this, Krintz et al propose a mechanism that evaluates the porting effort between two versions of a web-service API in a formal and automated way.
-To analyze the porting effort, they divide API changes into \emph{syntactic similarity}, the changes in inputs and outputs, and into \emph{semantic similarity}, the changes in the API's behavior.
-In initial tests, their method showed good congruence with human developers in scoring porting effort, offering the possibility that API governance can become as solid a part of scientific IT management as data center infrastructure management is today.
-
-% Next, we’ll look at issues of software security.
-
-% Heiland et al from the Center for Trustworthy Scientific Infrastructure discussed issues related to implementing strong security measures for scientific software. 
-% They point out that cybersecurity is rarely addressed in scientific software design.
-
-% Security considerations for software vary depending on the maturity level of the software package.
-% But when we initially develop scientific software, we generally don’t know what the final maturity level will be.
-% Scientific software developers are probably not aware of best practices for cybersecurity.
-% So the authors introduce the concept of Software Security Maturity Models, such as OpenSAMM and BSI-MM. 
-% These are used in industry to identify and define security vulnerabilities at different stages of the software life cycle.
-
-% They suggest that a similar Software Security Maturity Model can formalize this process:
-% It provides classification of software security practices.
-% It provides a path for tightening security practices as a package’s maturity level increases.
-% It emphasizes understandability over complexity.
+The workshop contributions in this group were concerned with the
+aspect of how implementation of best practices and other
+recommendations for improving scientific software sustainability could
+be promoted. Specifically, if scientific software is to become more
+sustainable, corresponding policies and guidelines need to be such
+that the scientific community can follow and implement them. This is
+considerably more challenging that it might seem at first, because in
+the reality of science today resources, both financial and personnel,
+that could be devoted to implementation are very limited, and the
+reward system does not encourage scientists to do so. Furthermore,
+implementing sustainability-targeting policies and guidelines often
+takes a variety of specialized software engineering expertises, which
+are not necessarily found in a single engineer, and much less so in a
+domain scientist cross-trained in programming. Adding to the policy
+implementation challenges, applicable sustainability-promoting
+practices and guidelines will change through a software project's
+lifecycle, in particular as it gains maturity.
+
+Two of the papers in this group focus on specific facets of software
+design that are important factors in a project's sustainability but
+are often addressed only late in the scientific software development
+cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
+governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
+models for software security. The other two papers discuss
+implementation strategies for science from the perspective of
+facilitating many or all facets of sustainanability-oriented software
+design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
+large projects that have software infrastructue development built-in
+with cross-training domain scientist PIs in software engineering best
+practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
+specialized software engineering skills could be turned into shared
+instrumentation with low barriers to access.
+
+Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
+computing frequently takes place in a distributed cloud, the control
+over digital resources is increasingly shifting from physical
+infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
+et al observe, unlike for physical IT infrastructure in data centers,
+science communities have developed very little in the way of practices and
+technology for API governance, refered to by Krintz et al as the
+``combined policy, implementation, and deployment control''. Web APIs
+can and do change, sometimes quite frequently. raising the need to
+port dependent applications. The effort required for porting is
+notoriously difficult to estimate, making it nearly impossible for IT
+organizations to assess and thus properly manage the impact of API
+changes. To address this, Krintz et al propose a mechanism that
+evaluates the porting effort between two versions of a web-service API
+in a formal and automated way. To analyze the porting effort, they
+divide API changes into \emph{syntactic similarity}, the changes in
+inputs and outputs, and into \emph{semantic similarity}, the changes
+in the API's behavior. In initial tests, their method showed good
+congruence with human developers in scoring porting effort, offering
+the possibility that API governance can become as solid a part of
+scientific IT management as data center infrastructure management is
+today.
+
+Another aspect of sustainable software that is rarely given due
+diligence in scientific software design is software security. Heiland
+et al~\cite{Heiland_WSSSPE} discuss how the security measures
+appropriate for scientific software strongly depend on the maturity
+level of the software. However, the maturity level of a software
+project meant to be sustained changes tremendously over its
+development life cycle, and the eventual maturity level is often
+difficult to predict during initial development. Best practice models
+for identifying and defining cybersecurity vulnerabilities at
+different stages of the software life cycle have been formalized as
+Software Security Maturity Models. Although these are widely used in
+industry, awareness among scientific software development communities
+is low. Developing such models for scientific software would align
+well with the objective of providing implementation approaches that
+the scientific community can actually follow: A Software Security
+Maturity Model provides a path to tightening security practices as
+software matures, emphasizes understandability over complexity, and
+classifies cybersecurity-pertinent practices for easy recognition of
+what level of cybersecurity a software addresses.
 

### 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex
some text changes

--- 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.
-Thus Calero and colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perdurability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
+Thus Calero and colleagues propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perduability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
 This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al.
 argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact.
-Lenhardt et al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered).
-In addition, their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
+Lenhardt et al compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered.) In addition their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
 In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.
@@
 The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project).
-For example Pierce et al~\cite{Pierce_WSSSPE} focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
+For example Pierce et al focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
 In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''.
 They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.
-Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance.
+Katz and Proctor also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance~\cite{Katz_WSSSPE}.
 They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).
@@
 Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software.
-Knepley et al~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
+Knepley et.
+al.
+approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
 They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}.
@@
 Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them.
-Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications.
-
-Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like Github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
-They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``altmetrics'' approach.
+Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications).
+
+Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
+They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``alt.metrics'' approach.
 
@@
 
-The workshop contributions in this group were concerned with the
-aspect of how implementation of best practices and other
-recommendations for improving scientific software sustainability could
-be promoted. Specifically, if scientific software is to become more
-sustainable, corresponding policies and guidelines need to be such
-that the scientific community can follow and implement them. This is
-considerably more challenging that it might seem at first, because in
-the reality of science today resources, both financial and personnel,
-that could be devoted to implementation are very limited, and the
-reward system does not encourage scientists to do so. Furthermore,
-implementing sustainability-targeting policies and guidelines often
-takes a variety of specialized software engineering expertises, which
-are not necessarily found in a single engineer, and much less so in a
-domain scientist cross-trained in programming. Adding to the policy
-implementation challenges, applicable sustainability-promoting
-practices and guidelines will change through a software project's
-lifecycle, in particular as it gains maturity.
-
-Two of the papers in this group focus on specific facets of software
-design that are important factors in a project's sustainability but
-are often addressed only late in the scientific software development
-cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
-governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
-models for software security. The other two papers discuss
-implementation strategies for science from the perspective of
-facilitating many or all facets of sustainanability-oriented software
-design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
-large projects that have software infrastructue development built-in
-with cross-training domain scientist PIs in software engineering best
-practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
-specialized software engineering skills could be turned into shared
-instrumentation with low barriers to access.
-
-Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
-computing frequently takes place in a distributed cloud, the control
-over digital resources is increasingly shifting from physical
-infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
-et al observe, although science communities have (for the most part)
-developed solid practices for managing physical IT infrastructure in
-data centers, corresponding practices and technology for API
-governance, refered to by Krintz et al as the ``combined policy,
-implementation, and deployment control'', is far less developed. As
-Krintz et al demonstrate, web APIs can and do change, sometimes quite
-frequently. raising the need to port dependent applications. The
-effort required for porting is notoriously difficult to estimate,
-making it nearly impossible for IT organizations to assess and thus
-properly manage the impact of API changes. To address this, Krintz et
-al propose a mechanism that evaluates the porting effort between two
-versions of a web-service API in a formal and automated way. To
-analyze the porting effort, they divide API changes into
-\emph{syntactic similarity}, the changes in inputs and outputs, and
-into \emph{semantic similarity}, the changes in the API's behavior. In
-initial tests, their method showed good congruence with human
-developers in scoring porting effort, offering the possibility that
-API governance can become as solid a part of scientific IT management
-as data center infrastructure management is today.
-
-% Next, we’ll look at issues of software security.
-
-% Heiland et al from the Center for Trustworthy Scientific Infrastructure discussed issues related to implementing strong security measures for scientific software. 
-% They point out that cybersecurity is rarely addressed in scientific software design.
-
-% Security considerations for software vary depending on the maturity level of the software package.
-% But when we initially develop scientific software, we generally don’t know what the final maturity level will be.
-% Scientific software developers are probably not aware of best practices for cybersecurity.
-% So the authors introduce the concept of Software Security Maturity Models, such as OpenSAMM and BSI-MM. 
-% These are used in industry to identify and define security vulnerabilities at different stages of the software life cycle.
-
-% They suggest that a similar Software Security Maturity Model can formalize this process:
-% It provides classification of software security practices.
-% It provides a path for tightening security practices as a package’s maturity level increases.
-% It emphasizes understandability over complexity.
-
-% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.
-
-% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.
-
-% The authors focus on a point that has been brought up many times in this context:
-% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
-% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
-% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
-% We don’t have a way to validate that the software used in a paper is actually done right.
-% The best way to get software designed correctly is to make sure best practices are considered from the start.
-
-% The authors highlight two models for sustainable software, at different extremes:
-% One is what they call “co-funding”:
-% In these projects, usually large, multi-year collaborations, 
-% there is equal emphasis on both the science and the software development. 
-% Both are planned into the project from inception. 
-% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.
-
-% At the other extreme, they discuss “software carpentry”: 
-% in this model, it’s assumed that the scientists themselves will write and maintain their code.
-% Groups like Software Carpentry and ROpenSci assume that 
-% scientists won’t have access to dedicated software engineering, 
-% so they try to give them tools to use best practices in their own software development.
-
-% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
-% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.
-
-% What do we mean when we refer to the “long tail” of scientific software?
-% Think of the distribution of resources in scientific software.
-Most are focused on big projects with lots of community buy-in and funding.
-But a lot of scientific software exists away from this model.
-% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
-% Look at MacClade: it was originally released in 1986 and last updated in 2005,
-% but it was still cited over 400 times in 2013!
-% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.
-
-% Another dimension of the long tail can also be found in my particular research domain. 
-% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
-% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
-% Most of these programs are developed by academic scientists…
-% They generally have limited training in software engineering
-% Limited time or career incentive to improve software
-% Limited funding
-
-% So, to summarize a bit:
-% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
-% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
-% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.
-
-% Then we have to consider that the users of scientific software are scientists, 
-% so the developers need to understand the users and the science.
-% This is the idea of a “t-skilled” person: 
-% one who is both well-versed in a scientific domain 
-% and deeply experienced in one or more facets of software engineering. 
-% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.
-
-% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
-% Think of it as analogous to DNA sequencing:
-% Sequencers used to be something that individual labs and institutions had to 
-% buy, maintain, and operate themselves, 
-% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
-% But now, core facilities provide the instrumentation and service to labs of any size. 
-% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.
-
-% We propose that software engineering can be “instrumented” in a similar way. 
-% Let’s create a nonprofit center for scientific software engineering.
-% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
-% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
-% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.
-
-% So, to conclude…
-% Implementing policies to encourage sustainability in scientific software 
-% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
-% But most of them aren’t addressed in the status quo.
-% We’ve highlighted some of these facets today and suggested some possible solutions.
-% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
-% Grassroots developer groups can provide guidance to scientists about best practices in software development.
-% We think there is a place for a software engineering center that can provide 
-% both engineering expertise and guidance 
-% with a contract-driven instrumentation model 
-% to the scientific software in the long tail.
-
 \subsection*{Papers}
@@
 
-\item Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
+\item Rich Wolski, Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher. Developing Systems for API Governance~\cite{Wolski_WSSSPE}
 
@@
 understand how the course of current and future development projects might be
-altered to implement lessons learned. In particular, we discuss the following issues:
-\begin{itemize}
-    \item How closely are the best practices discussed in the workshop were followed?
-    \item Career path: Are the original developers still around?
-    \item What conscious sustainability efforts were undertaken by stakeholders?
-    \item What is the dominant sustainability factor: science, publication, (re)usability?
-\end{itemize}
+altered to implement lessons learned. We discuss vatious issues including
+career paths of original developers and investigate if the original developers
+are still around?
 
@@
 % 
-
-%Summary of above discussed cases: lessons, trends, commonalities, is there an ideal sustainability scenario?
 In order to investigate the much discussed issue of career paths for developers
@@
 
-From this survey, it appears that the career paths for utility software
-developers are broader compared to the developers of niche software. However,
-as projects as a whole, the small-scale, specific software have a lower risk
-and cost in terms of sustaining as compared to utility software. Often the
-sustainability of such projects is achieved by the fact that the core
+In conclusion, there are clear distinctions in the characteristics and the
+kinds of challenges faced between generic utility oriented and specific
+scientific domain oriented software projects. In terms of career paths of
+original developers, it appears that the career paths for utility software
+developers are broader compared to the developers of scientific software.
+However, it appears that in the case of scientific software, core competencies
+of original developers tend to be domain science who happen to be the coders of
+their software. As projects as a whole, the small-scale, specific software have
+a lower risk and cost in terms of sustaining as compared to utility software.
+Often the sustainability of such projects is achieved by the fact that the core
 developer or team heavily utilizes the software for their science, eg.
-\emph{R/qtl, PETSc}. Furthermore, the development of scientific software requires more
-scientific skills compared with those of utility software. This diminishes the
-need of specialist programmers and practices in many cases where the bulk of
-development is carried out by a domain scientist.
+\emph{R/qtl, PETSc}. Furthermore, the development of scientific software
+requires more scientific skills compared with those of utility software. This
+diminishes the need of specialist programmers and practices in many cases where
+the bulk of development is carried out by a domain scientist. With these
+inferences, we believe that separate guidelines and/or principles for
+sustainability be defined for each category of software projects in science and
+they be given separate treatment.
 

### 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex
some text changes

@@
 
-In conclusion, there are clear distinctions in the characteristics and the
-kinds of challenges faced between generic utility oriented and specific
-scientific domain oriented software projects. In terms of career paths of
-original developers, it appears that the career paths for utility software
-developers are broader compared to the developers of scientific software.
-However, it appears that in the case of scientific software, core competencies
-of original developers tend to be domain science who happen to be the coders of
-their software. As projects as a whole, the small-scale, specific software have
-a lower risk and cost in terms of sustaining as compared to utility software.
-Often the sustainability of such projects is achieved by the fact that the core
-developer or team heavily utilizes the software for their science, eg.
-\emph{R/qtl, PETSc}. Furthermore, the development of scientific software
-requires more scientific skills compared with those of utility software. This
-diminishes the need of specialist programmers and practices in many cases where
-the bulk of development is carried out by a domain scientist. With these
-inferences, we believe that separate guidelines and/or principles for
-sustainability be defined for each category of software projects in science and
-they be given separate treatment.
+In conclusion, we find that there are distinctions in the characteristics and
+the kinds of challenges faced between generic utility and specific scientific
+domain software projects. In terms of career paths of original developers, it
+appears that the career paths for utility software developers are broader
+compared to the developers of scientific software.  However, it appears that in
+the case of scientific software, core competencies of original developers tend
+to be domain science who happen to be the coders of their software. As projects
+as a whole, the small-scale, specific software have a lower risk and cost in
+terms of sustaining as compared to utility software.  Often the sustainability
+of such projects is achieved by the fact that the core developer or team
+heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
+Furthermore, the development of scientific software requires more scientific
+skills compared with those of utility software. This diminishes the need of
+specialist programmers and practices in many cases where the bulk of
+development is carried out by a domain scientist. With these inferences, we
+believe that separate guidelines and/or principles for sustainability be
+defined and practiced for each category of scientific software and they be
+given separate treatment.
 

### 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex

--- 2014-03-31 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-From this survey, it appears that the career paths for utility software
-developers are broader compared to the developers of niche software. However,
-as projects as a whole, the small-scale, specific software have a lower risk
-and cost in terms of sustaining as compared to utility software. Often the
-sustainability of such projects is achieved by the fact that the core
-developer or team heavily utilizes the software for their science, eg.
-\emph{R/qtl, PETSc}. Furthermore, the development of scientific software requires more
-scientific skills compared with those of utility software. This diminishes the
-need of specialist programmers and practices in many cases where the bulk of
-development is carried out by a domain scientist.
+In conclusion, we find that there are distinctions in the characteristics and
+the kinds of challenges faced between generic utility and specific scientific
+domain software projects. In terms of career paths of original developers, it
+appears that the career paths for utility software developers are broader
+compared to the developers of scientific software.  However, it appears that in
+the case of scientific software, core competencies of original developers tend
+to be domain science who happen to be the coders of their software. As projects
+as a whole, the small-scale, specific software have a lower risk and cost in
+terms of sustaining as compared to utility software.  Often the sustainability
+of such projects is achieved by the fact that the core developer or team
+heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
+Furthermore, the development of scientific software requires more scientific
+skills compared with those of utility software. This diminishes the need of
+specialist programmers and practices in many cases where the bulk of
+development is carried out by a domain scientist. With these inferences, we
+believe that separate guidelines and/or principles for sustainability be
+defined and practiced for each category of scientific software and they be
+given separate treatment.
 

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
small change in opening

@@
 % TODO: Why are we doing case studies?
-Many important issues related to scientific software sustainability were
-discussed in the previous sections. In this section we discuss some of the
-software projects as case-studies to better understand the relevance of those
-issues and how are they connected to the software sustainability and related
-factors.
+In this section, we discuss some of the software projects as case-studies to
+better understand the points discussed during the workshop and described in the
+previous sections. We dive deeper to find how are they affected by the
+sustainability issues in practice.
 

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
Blanton summary and Huang summary for Implementing Policy.

--- 2014-04-01 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 facilitating many or all facets of sustainanability-oriented software
-design for science: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
-large projects that have software infrastructue development built-in
+design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
+large projects that have software infrastructue development built-in,
 with cross-training domain scientist PIs in software engineering best
@@
 
-% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.
-
-% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.
-
-% The authors focus on a point that has been brought up many times in this context:
-% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
-% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
-% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
-% We don’t have a way to validate that the software used in a paper is actually done right.
-% The best way to get software designed correctly is to make sure best practices are considered from the start.
-
-% The authors highlight two models for sustainable software, at different extremes:
-% One is what they call “co-funding”:
-% In these projects, usually large, multi-year collaborations, 
-% there is equal emphasis on both the science and the software development. 
-% Both are planned into the project from inception. 
-% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.
-
-% At the other extreme, they discuss “software carpentry”: 
-% in this model, it’s assumed that the scientists themselves will write and maintain their code.
-% Groups like Software Carpentry and ROpenSci assume that 
-% scientists won’t have access to dedicated software engineering, 
-% so they try to give them tools to use best practices in their own software development.
-
-% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
-% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.
+API governance and cybersecurity measures appropriate for a project's
+maturity are all but two facets of sustainability-oriented software
+development. Others include user-centered interface design, test
+engineering, dependency management, and deployment operations. Each of
+these facets require specialized skills and training in software
+engineering. How can the implementation of best practices and
+guidelines along many or all of these different facets be facilitated
+in scientific software development? Blanton and
+Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
+and personnel devoted to software engineering is ``co-funded'' with
+the driving science project. This typically implies large multi-year
+collaborative projects that to succeed require significant software
+infrastructure to be built, and which thus have the funding to support
+one or several software engineers. The sustainability of such projects
+then depends on sustaining the funding. In the other model, domain
+scientists also take the role of software developers, whether by
+necessity such as funding limitations, or due to cross-disciplinary
+professional interests. For this to result in sustainable software
+products, the domain scientists need to be (or become) cross-trained
+in software engineering standards and best practices.
+
+In practice, there are example for both extremes of the spectrum. For
+example, in the life sciences the iPlant Collaborative~\cite{iPlant},
+the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
+multi-year projects with significant software infrastructure funding
+and deliverables. Typically though these are the exception rather than
+the rule, and particularly in the long tail of science the
+scientist-developer predominates. For providing these with better
+training in software engineering best practices, initiatives such as
+Software Carpentry~\cite{SoftwareCarpentry} have been especially
+valuable, and Software Carpentry also shown how such training could
+scale.
+
+However, there may also be a middle ground between the two
+extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
+Excellence model that leverages economies of scale to make software
+engineering experts and their skills accessible to the long tail of
+science. As Huang and Lapp discuss, this model could effectively turn
+the utilization of software engineering expertise from a complex human
+resource recruitment and management challenge, to buying time, when
+and to the extent needed, on shared instrumentation. There is
+precedence to using such a model to lower access barriers for long
+tail science, in particular for new experimental technologies.  For
+example, although the acquisition and operation of next-generation
+high-throughput DNA sequencers requires substantial investments of
+money, time, and expertise, the establishment of ``core facilities''
+on many university campuses has made these technologies accessible to
+a wide swath of scientists, with transformative results for science.
+
+One of the important conclusions from this group of papers is that creating sustainable software requires paying attention not to one or two, but to several different facets of software engineering, each with corresponding best practices and standards of excellence.
+Even if a science project requires and has funding for a full software engineering FTE, what the project really needs could be fractions of different engineers with different specialty training.
+The vast majority of long tail science projects lacks the funding for even one full software engineer, let alone one who combines expertise in all the applicable facets of engineering.
+Some scientists in the long tail will have the professional interests to cross-train enough in software engineering to be successful with creating sustainable software, but it is likely unrealistic to expect that of all or even the majority.
+This is where a software engineering center of excellence can provide a critical service.
+
+% So, to summarize a bit:
+% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
+% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
+% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.
+
+% So, to conclude…
+% Implementing policies to encourage sustainability in scientific software 
+% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
+% But most of them aren’t addressed in the status quo.
+% We’ve highlighted some of these facets today and suggested some possible solutions.
+% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
+% Grassroots developer groups can provide guidance to scientists about best practices in software development.
+% We think there is a place for a software engineering center that can provide 
+% both engineering expertise and guidance 
+% with a contract-driven instrumentation model 
+% to the scientific software in the long tail.
 

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
small change

--- 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-% TODO: Why are we doing case studies?
 In this section, we discuss some of the software projects as case-studies to
@@
 We aim to understand the practicalities of the key points discussed in the
-workshop, and hope to extract the best practices associated with the case
-studies. We discuss how the issues are applicable to the cases and try to
-understand how the course of current and future development projects might be
-altered to implement lessons learned. We discuss vatious issues including
-career paths of original developers and investigate if the original developers
-are still around?
+workshop, and discuss the practices associated with the use-cases. We discuss
+how the issues are applicable to the cases and try to understand how the course
+of current and future development projects might be altered to implement
+lessons learned. We discuss various issues including career paths of original
+developers and investigate if the original developers are still around?
 

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
removing redundant content

@@
 previous sections. We dive deeper to find how are they affected by the
-sustainability issues in practice.
+sustainability issues in practice. Additionally, We discuss various issues
+including career paths of original developers and investigate if the original
+developers are still around?
 
@@
 \textit{Quantum Espresso}.
-
-% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
-We aim to understand the practicalities of the key points discussed in the
-workshop, and discuss the practices associated with the use-cases. We discuss
-how the issues are applicable to the cases and try to understand how the course
-of current and future development projects might be altered to implement
-lessons learned. We discuss various issues including career paths of original
-developers and investigate if the original developers are still around?
 

### 2014-04-03 WSSSPE1_report/WSSSPE1_summary_report.tex
addressing Berriman comments

@@
 
-Because progress in scientific research is dependent on the quality and
-accessibility of software at all levels, it is now critical to address many
+Because progress in scientific research is dependent on the quality of and
+accessibility to software at all levels, it is now critical to address many
 challenges related to the development, deployment, and maintenance of reusable
@@
 further explore some of the key issues raised in WSSSPE is co-organized by a
-WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
+WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
+(SIAM) Annual Meeting on ``Reliable
 Computational Science.''
@@
 
+The goal of this paper is to summarize the contents of the workshop papers
+and discussion, and identify issues that future WSSSPE events will address.
+
 \section{Workshop Process and Agenda} \label{sec:process}
@@
 \begin{quote}
-In practice, scientific software activities are part of an ecosystem
+``In practice, scientific software activities are part of an ecosystem
 where key roles are held by developers, users, and funders. All three
@@
 \end{itemize}
-\item education and training
+\item education and training''
 \end{itemize}
@@
 \begin{quote}
-We invite short (4-page) position/experience reports that will be used
+``We invite short (4-page) position/experience reports that will be used
 to organize panel and discussion sessions. These papers will be
-archived by a third-party service, and provided DOIs. We encourage
+archived by a third-party service, and provided DOIs [Digital Object Identifiers].
+We encourage
 submitters to license their papers under a Creative Commons license
@@
 to be collaboratively developed by the contributors, based on the
-panels and discussions.
+panels and discussions.''
 \end{quote}
@@
 software architectures determines sustainability. They then propose that
-sustainability is a measure of a set of central quality attributes:
+sustainability is a consequence of a set of central quality attributes:
 extensibility, interoperability, maintainability, portability, reusability, and
@@
 of such projects is achieved by the fact that the core developer or team
-heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
+heavily utilizes the software for their science, e.g.,  \emph{R/qtl, PETSc}.
 Furthermore, the development of scientific software requires more scientific
@@
 \todo{say something about licensing - lessons, advice, etc.?}
+
+\subsection{Top Issues}
 

### 2014-04-04 WSSSPE1_report/WSSSPE1_summary_report.tex
changing bib to vancouver style
updates to structure overall, a few other changes scattered

@@
 seeking to understand what drives software work in science and how the
-reward systems of science thereby shape the type of software work
+reward systems of science thereby shapes the type of software work
 undertaken, including the extent to which developers are motivated to
@@
 
-\note{papers and presentations and URLs should be references - see
-  figshare-web in .bib file for URL style example}
+%\note{papers and presentations and URLs should be references - see
+%  figshare-web in .bib file for URL style example}
 
@@
 
-\pagebreak
-
-\section*{Executive Summary}
-
-\todo{1 page summary goes here}
-
-\pagebreak
+%\pagebreak
+%
+%\section*{Executive Summary}
+%
+%\todo{1 page summary goes here}
+%
+%\pagebreak
 
@@
 organizers, working with some of the submitters and attendees, to plan two
-additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
+additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
+funds from the National Science Foundation and the Gordon and Betty Moore
+Foundation to support these workshops, and to turn the
 workshop website into a community website that can be used as a focus for
@@
 
+The goal of this paper is to summarize the contents of the workshop papers
+and discussion, and identify issues that future WSSSPE events will address.
 Before the WSSSPE1 workshop took place, the organizers self-published
@@
 (\S\ref{sec:conclusions}).
-
-The goal of this paper is to summarize the contents of the workshop papers
-and discussion, and identify issues that future WSSSPE events will address.
 
@@
 The organizers decided to list 54 of the papers as significant
-contributions to the workshop, a very high acceptance rate, but one
+contributions to the workshop, a very high acceptance rate but one
 that is reasonable, given the goal of broad participation and the fact
@@
 
-\todo{section should include what was presented and what was discussed}
-
 The panel on developing and supporting software examined the challenges
@@
 as much due to users being more active, or whether the ``core
-team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
-and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
+team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
+and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
 lists that are welcoming to people actually have many more people contributing.
@@
 
-\subsection*{Papers}
-
-\subsubsection*{Development Experiences}
-
-\begin{itemize}
-
-\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
-  McInnes, Barry Smith. Package Management Practices Essential for
-  Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath~\cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
-
-\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
-  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
-
-\item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
-  adapting sustainable scientific software development processes in a
-  small biology lab~\cite{Crusoe_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
-The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
-  H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
-
-\begin{itemize}
-
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
-  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
-  Versatile Simulation of Large Scale Distributed
-  Systems~\cite{Casanova_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
-  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
-  Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software~\cite{Cohen_WSSSPE}
-
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Best Practices, Challenges, and Recommendations}
-
-\begin{itemize}
-
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
-
-\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
-  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
-  R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
-
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult~\cite{Carver_WSSSPE}
-
-\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
-  K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
-
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable~\cite{Brown_WSSSPE}
-
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
-  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
-  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms~\cite{Swenson_WSSSPE}
-
-\end{itemize}
-
 \subsection{Research or Reuse?}
@@
 
-\subsection*{Modeling Sustainability}
+\todo{some text is needed here before the first subsection}
+
+\subsection{Modeling Sustainability}
 
@@
 
-\todo{seems like definitions will be mentioned elsewhere in the paper?}
+%\todo{seems like definitions will be mentioned elsewhere in the paper?}
 Each of the framework papers included a definition of sustainability, with many overlaps.
@@
 This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
+See \S\ref{sec:cross-cutting} for more on the discussion of definitions of sustainability in the workshop.
 
@@
 
-\subsection*{Credit, Citation, Impact}
+\subsection{Credit, Citation, Impact}
 
@@
 
-\subsection*{Implementing Policy}
+\subsection{Implementing Policy}
 
@@
 
-\subsection*{Papers}
-
 % note that paper summaries are available in tag jameshowison-paper-summaries: 
@@
 
-\subsubsection*{Modeling Sustainability}
-
-\begin{itemize}
-
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model~\cite{Calero_WSSSPE}
-
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Credit, Citation, Impact}
-
-\begin{itemize}
-
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
-  Smith. Accurately Citing Software and Algorithms Used in
-  Publications~\cite{Knepley_WSSSPE}
-
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project~\cite{Priem_WSSSPE}
-
-\item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns~\cite{Katz2_WSSSPE}
-
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Reproducibility}
-
-\begin{itemize}
-
-\item Victoria Stodden, Sheila Miguez. Best Practices for
-  Computational Science: Software Infrastructure and Environments for
-  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Implementing Policy}
-
-\begin{itemize}
-
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
-
-\item Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
-
-\end{itemize}
-
 \subsection{Career Tracks for Scientific Software Developers}
@@
 
-\todo{section should include what was presented and what was discussed}
-
-There were a number of papers categorized under the Communities banner and so
-two presenters each summarized half of the submissions in this area.
-
-\subsection{Communities Part 1}
-
-This collection of papers was summarized by Karen Cranston.
+%There were a number of papers categorized under the Communities banner and so
+%two presenters each summarized half of the submissions in this area.
+\todo{an intro to the whole section is needed}
+
+\subsection{Communities}
+
+%This collection of papers was summarized by Karen Cranston.
 
@@
 
-\subsection{Communities Part 2}
-
-Nancy Wilkins-Diehr summarized 6 papers in this part.
+%\subsection{Communities Part 2}
+
+%Nancy Wilkins-Diehr summarized 6 papers in this part.
 
@@
 
-\subsection*{Papers}
-
-\subsubsection*{Communities Part 1 Papers}
-
-\begin{itemize}
-
-\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities~\cite{Vay_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Communities Part 2 Papers}
-
-\begin{itemize}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community~\cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area was also discussed
-in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Industry \& Economic Models}
-
-\begin{itemize}
-
-\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
-
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability~\cite{Foster_WSSSPE}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
-  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Education \& Training}
-
-\begin{itemize}
-
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
-  T. Brown. Advanced Techniques for Scientific Programming and
-  Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
-
-\item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry~\cite{Crawford_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\end{itemize}
-
-\subsection{What are communities?}
+\subsubsection{What are communities?}
 
@@
 
-\subsection{Challenges of community}
+\subsubsection{Challenges of community}
 include metrics for community success (e.g., more external
 contributors than internal).
-\subsection{Admirable scientific software communities}
+
+\subsubsection{Admirable scientific software communities}
 \begin{itemize}
@@
 \end{itemize}
-\subsection{Resources for learning about software communities}
+
+\subsubsection{Resources for learning about software communities}
 \begin{itemize}
@@
 \end{itemize}
+
+\subsection{Industry \& Economic Models}
+
+\todo{text about papers and discussion in this part of the workshop}
+
+\subsection{Education \& Training}
+
+\todo{text about papers and discussion in this part of the workshop}
 
@@
 \appendix
-\section{List of attendees \note{lead: Shel Swenson}}
+\section{Papers Accepted and Discussed at WSSSPE1}
+
+\section*{Developing and Supporting Software}
+
+\subsubsection*{Development Experiences}
+
+\begin{itemize}
+
+\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
+  McInnes, Barry Smith. Package Management Practices Essential for
+  Interoperability: Lessons Learned and Strategies Developed for
+  FASTMath~\cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
+  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
+
+\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
+  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
+  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
+
+\item Anshu Dubey, B. Van Straalen. Experiences from Software
+  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
+  adapting sustainable scientific software development processes in a
+  small biology lab~\cite{Crusoe_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
+The MVAPICH Project: Evolution and Sustainability of an Open Source
+Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
+  H. Kellogg. Experiences with Automated Build and Test for
+  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
+
+\begin{itemize}
+
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
+  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
+  Versatile Simulation of Large Scale Distributed
+  Systems~\cite{Casanova_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
+  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
+  Sherwin. Simplifying the Development, Use and Sustainability of HPC
+  Software~\cite{Cohen_WSSSPE}
+
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
+  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Best Practices, Challenges, and Recommendations}
+
+\begin{itemize}
+
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
+  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
+
+\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
+  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
+  R. Thakur, Software Abstractions and Methodologies for HPC
+  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
+
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
+  Need Not Be Difficult~\cite{Carver_WSSSPE}
+
+\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
+  K. Barnett, Von Welch. Initial Findings from a Study of Best
+  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
+
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
+  anything less is unsustainable~\cite{Brown_WSSSPE}
+
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
+  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
+  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
+  Platforms~\cite{Swenson_WSSSPE}
+
+\end{itemize}
+
+\section*{Policy}
+
+\subsubsection*{Modeling Sustainability}
+
+\begin{itemize}
+
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
+  Software Product Sustainability Model~\cite{Calero_WSSSPE}
+
+\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
+  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
+  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
+  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing
+  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
+  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
+  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
+  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Credit, Citation, Impact}
+
+\begin{itemize}
+
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
+  Smith. Accurately Citing Software and Algorithms Used in
+  Publications~\cite{Knepley_WSSSPE}
+
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
+  report for every software project~\cite{Priem_WSSSPE}
+
+\item Daniel S. Katz. Citation and Attribution of Digital Products:
+  Social and Technological Concerns~\cite{Katz2_WSSSPE}
+
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
+  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Reproducibility}
+
+\begin{itemize}
+
+\item Victoria Stodden, Sheila Miguez. Best Practices for
+  Computational Science: Software Infrastructure and Environments for
+  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Implementing Policy}
+
+\begin{itemize}
+
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
+  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Daisie Huang, Hilmar Lapp. Software Engineering as
+  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
+
+\item Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
+
+\end{itemize}
+
+\section*{Communities, Models, and Education}
+
+\subsubsection*{Communities}
+
+\begin{itemize}
+
+\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
+  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
+  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
+  Modeling Science Activities~\cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
+  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
+  Experiences and Challenges in Software Sustainability for a Large
+  Scientific Community~\cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed
+in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Industry \& Economic Models}
+
+\begin{itemize}
+
+\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
+
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
+  as a path to software sustainability~\cite{Foster_WSSSPE}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
+  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
+  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Education \& Training}
+
+\begin{itemize}
+
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
+  T. Brown. Advanced Techniques for Scientific Programming and
+  Collaborative Development of Open Source Software Packages at the
+  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
+
+\item Thomas Crawford. On the Development of Sustainable Software for
+  Computational Chemistry~\cite{Crawford_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\end{itemize}
+
+\section{Attendees \note{lead: Shel Swenson}}
 
@@
 
-\bibliographystyle{unsrt}
-
-\bibliography{wssspe}
+\bibliographystyle{vancouver}
+
+\bibliography{wssspe_paper}
 \end{document}

### 2014-04-04 WSSSPE1_report/WSSSPE1_summary_report.tex
removing some authors who haven't been involved

@@
 \begin{itemize}
-\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
+% \item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
 \item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
 \item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
-\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
+\item * Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
 \item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
-\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
+% \item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
 \item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
 \item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
-\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
-\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
+\item * Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
+% \item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
 \item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
-\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
+% \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
 \item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
@@
 \item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
-\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
+\item * Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
 \item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
improve the flow of text and cleanup

@@
 
-In this section, we discuss some of the software projects as case-studies to
+In this section, we discuss some of the software projects as case studies to
 better understand the points discussed during the workshop and described in the
 previous sections. We dive deeper to find how are they affected by the
-sustainability issues in practice. Additionally, We discuss various issues
-including career paths of original developers and investigate if the original
-developers are still around?
+sustainability related issues in practice. Additionally, we discuss issues
+surrounding people, especially the career paths of original developers and
+investigate to answer the question: are the original developers are still
+around?
 
 We classify the software projects discussed in the workshop in two broad
-categories. First, the utility software, comprising of software that enable
-and/or facilitate the development of other tools and techniques to carry out
-scientific work. This includes the software developed to efficiently utilize
-new research infrastructures. Second, scientific software, comprising software
-that was developed with an aim to solve a specific scientific problem.
+categories. First, the \emph{utility} software, comprising of general purpose
+software. Utility software is often used as enabler and/or facilitator for the
+development of other tools and techniques to carry out scientific work. This
+includes the software developed to efficiently utilize new research
+infrastructures. Second, the \emph{scientific} software, comprising the
+software that was originally developed with an aim to solve a specific
+scientific problem.
 
 % Why are we classifying software like above?
-This classification is motivated by our argument that the software projects
-wildly vary in factors such as scope, purpose and usage. The development and
-management of software is significantly influenced by these factors.
-Consequently, the sustainability challenges faced by software differ and must
-be treated separately. For instance, the challenges faced by a gateway software
-development project such as \textit{CIPRES} or \textit{VisIT} are distinct to a
-niche software for ab initio modeling and simulation such as \textit{VASP} or
-\textit{Quantum Espresso}.
+This classification is motivated by our argument that the two kinds of software
+projects wildly vary in factors such as scope, purpose and usage. The
+development and management of each kind is significantly different.
+Consequently, the sustainability challenges faced by them differ and must be
+treated separately. For instance, the challenges faced by a gateway software
+development project such as \textit{CIPRES} or a visualization software product
+such as \textit{VisIT} are distinct to a niche software for \textit{ab initio}
+modeling and simulation such as \textit{VASP} or \textit{Quantum Espresso}.
 
 \subsection{Utility Software}
-
-We classify software developed with a potentially wider audience and general
-purpose usage in mind as Utility software. In the context of research, a
-utility software could be defined as general purpose system applicable to one
-or many generic tasks and/or enabling other software to run by providing a
-suitable environment. Utility software typically do not address a fundamental
-research challenge in a given scientific domain. Examples are collaborative
-development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
-and workflow and generic computing frameworks such as \emph{Galaxy},
-\emph{nanoHUB}, \emph{SimGrid}, \emph{Swift}, \emph{Globus} and \emph{VisTrails}, and
-visualization frameworks such as \emph{VisIT}.
+Software developed with a potentially wider audience and general purpose usage
+in mind is utility software. Utility software typically does not address
+fundamental research problems for a given scientific domain. Examples are
+collaborative development frameworks such as \emph{GitHub} and
+\emph{Bitbucket}, distributed workflow and generic computing frameworks
+such as \emph{Galaxy}, \emph{nanoHUB}, \emph{SimGrid}, \emph{Swift},
+\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
+\emph{VisIT}.
 
 % What are the sustainability challenges? What are the oppurtunities?
-Development is often a high risk/reward undertaking and explores uncharted
-territories and is largely dominated by (re)usability factors. Owing to
+Development is often a high risk/reward undertaking exploring uncharted
+territories and is largely influenced by (re)usability factors. Owing to
 relatively large number of features, the development and prototype process is
-also lengthy which poses a significant survival risk.
-Ferenbaugh~\cite{Ferenbaugh_WSSSPE} further discusses the risks and challenges
-associated with software projects dealing with new architectures.
-
-On the other hand, owing to its generic nature, utility software presents
-oppurtunities to be usable by a larger community. This makes it more visible
-which in turn leads to a broader and deeper participation. For
-instance, promoting collaborations across the breadth (e.g. different
-departments) and depth (e.g. stakeholders within a department) of community,
-one of the key ingredient of a sustainable process. Successful projects reap
-high rewards and have longer usage span. Development process becomes
-user-driven and self-sustaining.
+also lengthy which poses a significant survival risk. Challenges on a class of
+utility software for new architectures is well discussed
+in~\cite{Ferenbaugh_WSSSPE}.
+
+On the other hand, utility software presents oppurtunities to be usable by a
+larger community making its undertaking and development an attractive pursuit.
+It is generally more visible in community which in turn leads to a broader and
+deeper participation. For instance, promoting collaborations across the breadth
+(e.g.  different departments) and depth (e.g. stakeholders within a department)
+of community, one of the key ingredient of a sustainable process. Successful
+utility projects reap high rewards and have a longer usage span. Development
+process becomes user-driven and self-sustaining.
 
 One such example is the \emph{Galaxy} project. It follows agile software
-development practices and implements standard practices such as test-driven
-development and socialized bug managing practices via \emph{trello}. Galaxy
+development practices and implements standards such as test-driven development
+and socialized bug managing practices via \emph{trello}. Galaxy
 \emph{histories} and \emph{toolshed} offers easy community sharing of data and
@@
 e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
-emerging architectures such as CPU accelerators new code and libraries are
+emerging architectures such as accelerators, new code and libraries are
 required. The experience of one such effort as described
-in~\cite{Ferenbaugh_WSSSPE} met with a limited success but nonetheless with
-many invaluable lessons were learned about influential cultural and technical
-aspects in sustainable software development practices.
+in~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
+with many invaluable lessons were learned about influential cultural and
+technical aspects in sustainable software development practices.
 
@@
 computational environments, many users are leaning towards tools used as
-services. \emph{GitHub} and \emph{Bitbucket} can be arguably considered to be
+services. \emph{GitHub} and \emph{Bitbucket} can be argued to be
 such tools, catering to collaborative development. For scientific users
@@
 \subsection{Scientific software}
-
 Scientific software consists primarily of special-purpose software that was
-purpose-built for a target use-case scenario or fixed/frozen requirements in
-mind.  Software projects pertaining to specific scientific domain often tend to
-be in a niche and the user community tends to be small to medium. They are
-mostly driven by the science and specific needs of a research group.
+purpose-built for a target use-case scenario,  fixed/frozen requirements in
+mind or solving a specific problem. Software projects pertaining to specific
+scientific domain tend to be in a niche and the user community tends to be
+small to medium. They are mostly driven by the science and specific needs of a
+research group. Specific needs such as numerical accuracy and
+algorithmic optimization are some of the paramount requirements of most
+scientific software.
 
 % What are the challenges?
-Long-term sustainability of scientific research software is often a significant
-challenge. Many submissions reported that software is practically considered a
-`byproduct' of the actual research. Others contended that the software was not
-the main funded part of their research.  Specific needs such as numerical
-accuracy and algorithmic optimization are some of the paramount requirements of
-most scientific software.  The resulting codebase is relatively small. A
-smaller codebase and fixed requirements result in stability, ease of
-installation, and configurability.  Many such projects mature and are treated
-as libraries to be used into larger systems such as some of the utility
-software discussed in the previous section.  While the software can stay stable
-and require relatively low maintenance, the responsibility is often on the
-shoulders of a very few developers who might not be specialists in software
-development. Development tends to be linear and simplistic with a limited
-scope to follow software best practices.
-
-% TODO: What are the oppurtunities?
+Long-term sustainability of scientific software is often a significant
+challenge and face radically different issues compared to utility software.
+Many submissions reported that software is practically considered a `byproduct'
+of the actual research. Others contended that the software was not the main
+funded part of their research. A smaller codebase and fixed requirements result
+in stability, ease of installation, and configurability.  Many such projects
+mature and are treated as libraries to be used into larger systems such as some
+of the utility software discussed in the previous section. While the software
+can stay stable and require relatively low maintenance, the responsibility is
+often on the shoulders of a very few developers who might not be specialists in
+software development. Development tends to be linear and simplistic with a
+limited scope to follow software best practices.
 
@@
 %
-In order to investigate the much discussed issue of career paths for developers
-of scientific software, we conducted a small survey of the use-cases discussed
-above with respect to the current status of its core developers. We looked at
-the workshop submissions and the original cited papers for the respective
-software. Where available, we looked at the webpages and source code of the
-earliest versions to find the original authors. We excluded relatively new
-projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
-that in a majority of scientific software more than one original developers are
-still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
-Whereas, in the case of utility software projects, it appears that at least one
-original developer is still around often in a part-time role. In other cases,
-it appears that the original developers have went on to assume a larger role,
-again with the same project.
+
+Finally, it appears that the career options of developers from different kinds
+of software projects differ significantly. In order to further investigate the
+much discussed issue of career paths for developers of these software, we
+conducted a small survey of the use-cases discussed above with respect to the
+current status of its core developers. We looked at the workshop submissions
+and the original cited papers for the respective software. Where available, we
+looked at the webpages and source code of the earliest versions to find the
+original developers. We included projects older than 5 years and excluded
+relatively new projects such as \emph{SciPy} and \emph{numFOCUS} from this
+survey. we find that in a majority of scientific software more than one
+original developers are still around, in some cases for more than 10 years,
+(eg. \emph{R/qtl}).  Whereas, in the case of utility software projects, it
+appears that at least one original developer is still around, often in a
+part-time or ``benefactor" role. In other cases, it appears that the original
+developers have went on to assume a larger role, with the same project.
 
 In conclusion, we find that there are distinctions in the characteristics and
-the kinds of challenges faced between generic utility and specific scientific
-domain software projects. In terms of career paths of original developers, it
-appears that the career paths for utility software developers are broader
-compared to the developers of scientific software.  However, it appears that in
-the case of scientific software, core competencies of original developers tend
-to be domain science who happen to be the coders of their software. As projects
-as a whole, the small-scale, specific software have a lower risk and cost in
-terms of sustaining as compared to utility software.  Often the sustainability
-of such projects is achieved by the fact that the core developer or team
-heavily utilizes the software for their science, e.g.,  \emph{R/qtl, PETSc}.
-Furthermore, the development of scientific software requires more scientific
-skills compared with those of utility software. This diminishes the need of
-specialist programmers and practices in many cases where the bulk of
-development is carried out by a domain scientist. With these inferences, we
-believe that separate guidelines and/or principles for sustainability be
-defined and practiced for each category of scientific software and they be
-given separate treatment.
+the kinds of challenges faced between utility and scientific software projects.
+In terms of career paths of original developers, it appears that the career
+choices for utility software developers are broader compared to the developers
+of scientific software. However, it appears that in the case of scientific
+software, core competencies of original developers tend to be domain science
+who happen to be the coders of their software. As projects as a whole, the
+small-scale, specific software have a lower risk and cost in terms of
+sustaining as compared to utility software. We also find that the utility
+software tend to adhere more loyally to the many best practices discussed
+during the workshop. Often the sustainability of scientific software projects
+is achieved by the fact that the core developer or team heavily utilizes the
+software for their science, e.g.,  \emph{R/qtl, PETSc}.  Furthermore, the
+development of scientific software requires more scientific skills compared
+with those of utility software. This diminishes the need of specialist
+programmers and practices in many cases where the bulk of development is
+carried out by a domain scientist. With these inferences, we believe that
+separate guidelines and/or principles for sustainability be defined and
+practiced for each category of scientific software and they be given separate
+treatment.
 

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
quotes
Adding introduction to section 6.

@@
 %two presenters each summarized half of the submissions in this area.
-\todo{an intro to the whole section is needed}
+Across all talks and papers submitted, authors implicitly and explicitly
+recognized the concept of ``communities'' as a driving and unifying force
+within software projects.  Despite that, the actual nature of the communities,
+the incentive structures and bind them together, the infrastructure they
+utilize for communication and even the types of individuals that make up those
+communities were different.  Some of these communities were primarily composed
+of members of industry, some were funded and driven by individuals focused on
+developing software as opposed to utilizing it, and others were primarily
+composed of scientist practitioners, and were true communities of practice.
+
+These varying structures and compositions result in differing modes of
+interaction within communities, styles of development, and the structure of
+planning for future development of software.  In this section, we summarize
+the different types of communities in scientific software as well as the
+resultant impact on sustainability and development of functionality.
 
@@
 challenge and face radically different issues compared to utility software.
-Many submissions reported that software is practically considered a `byproduct'
+Many submissions reported that software is practically considered a ``byproduct"
 of the actual research. Others contended that the software was not the main

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Adding text to bibliography for communities.

@@
 \subsubsection{Resources for learning about software communities}
+
+Scientific software communities were viewed as a subset of software communities
+as whole.  As such, resources applicable to generic software communities --
+such as open source and proprietary technology companies -- can be used as
+input and as guiding understanding of how to steward and develop scientific
+software communities.  Because incentive structures are different in industry
+and volunteer-based open source communities, these can provide rough guidelines
+but not necessarily identically applicable.  The analogy between corporations
+and scientific investigators (particularly in terms of competition, cooperation
+and competitive advantage) has been explored in the literature below, but
+because of the different incentive structure the analogy is not universally
+true.
+
+The literature below, suggested by attendees, addresses both non-scientific
+software projects, as well as scientific projects.  The selections address both
+descriptive and prescriptive approaches to communities.
+
 \begin{itemize}
-\item academic fields studying communities (and software communities).
-\item courses on online communities
-\item need for software carpentry module on organizing communities?
 \item \cite{howison_scientific_2011}

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Remaining paper summaries for Implementing Policy.

--- 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 seeking to understand what drives software work in science and how the
-reward systems of science thereby shapes the type of software work
+reward systems of science thereby shape the type of software work
 undertaken, including the extent to which developers are motivated to
@@
 
-%\note{papers and presentations and URLs should be references - see
-%  figshare-web in .bib file for URL style example}
+\note{papers and presentations and URLs should be references - see
+  figshare-web in .bib file for URL style example}
 
@@
 \begin{itemize}
-% \item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
+\item Chris Mattmann $\langle$\url{chris.a.mattmann@jpl.nasa.gov}$\rangle$
 \item * Ketan Maheshwari $\langle$\url{ketancmaheshwari@gmail.com}$\rangle$
 \item Marlon Pierce $\langle$\url{marpierc@iu.edu}$\rangle$
-\item * Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
+\item Colin Venters $\langle$\url{C.Venters@hud.ac.uk}$\rangle$
 \item Suresh Marru $\langle$\url{smarru@iu.edu}$\rangle$
-% \item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
+\item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
 \item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
 \item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
-\item * Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
-% \item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
+\item Andy Ray Terrel $\langle$\url{andy.terrel@gmail.com}$\rangle$
+\item Abani Patra $\langle$\url{abani.patra@gmail.com}$\rangle$
 \item * Nancy Wilkins-Diehr $\langle$\url{wilkinsn@sdsc.edu}$\rangle$
-% \item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
+\item James Spencer $\langle$\url{j.spencer@imperial.ac.uk}$\rangle$
 \item * Frank L\"{o}ffler $\langle$\url{knarf@cct.lsu.edu}$\rangle$
@@
 \item * Bruce Berriman $\langle$\url{gbb@ipac.caltech.edu}$\rangle$
-\item * Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
+\item Hilmar Lapp $\langle$\url{hlapp@nescent.org}$\rangle$
 \item * Marcus D. Hanwell $\langle$\url{marcus.hanwell@kitware.com}$\rangle$
@@
 
-%\pagebreak
-%
-%\section*{Executive Summary}
-%
-%\todo{1 page summary goes here}
-%
-%\pagebreak
+\pagebreak
+
+\section*{Executive Summary}
+
+\todo{1 page summary goes here}
+
+\pagebreak
 
@@
 
-Because progress in scientific research is dependent on the quality of and
-accessibility to software at all levels, it is now critical to address many
+Because progress in scientific research is dependent on the quality and
+accessibility of software at all levels, it is now critical to address many
 challenges related to the development, deployment, and maintenance of reusable
@@
 organizers, working with some of the submitters and attendees, to plan two
-additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
-funds from the National Science Foundation and the Gordon and Betty Moore
-Foundation to support these workshops, and to turn the
+additional workshops (at the 2014 SciPy and SC14 conferences) and to turn the
 workshop website into a community website that can be used as a focus for
@@
 further explore some of the key issues raised in WSSSPE is co-organized by a
-WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
-(SIAM) Annual Meeting on ``Reliable
+WSSSPE1 participant at the 2014 SIAM Annual Meeting on ``Reliable
 Computational Science.''
 
-The goal of this paper is to summarize the contents of the workshop papers
-and discussion, and identify issues that future WSSSPE events will address.
 Before the WSSSPE1 workshop took place, the organizers self-published
@@
 \begin{quote}
-``In practice, scientific software activities are part of an ecosystem
+In practice, scientific software activities are part of an ecosystem
 where key roles are held by developers, users, and funders. All three
@@
 \end{itemize}
-\item education and training''
+\item education and training
 \end{itemize}
@@
 \begin{quote}
-``We invite short (4-page) position/experience reports that will be used
+We invite short (4-page) position/experience reports that will be used
 to organize panel and discussion sessions. These papers will be
-archived by a third-party service, and provided DOIs [Digital Object Identifiers].
-We encourage
+archived by a third-party service, and provided DOIs. We encourage
 submitters to license their papers under a Creative Commons license
@@
 to be collaboratively developed by the contributors, based on the
-panels and discussions.''
+panels and discussions.
 \end{quote}
@@
 The organizers decided to list 54 of the papers as significant
-contributions to the workshop, a very high acceptance rate but one
+contributions to the workshop, a very high acceptance rate, but one
 that is reasonable, given the goal of broad participation and the fact
@@
 
+\todo{section should include what was presented and what was discussed}
+
 The panel on developing and supporting software examined the challenges
@@
 as much due to users being more active, or whether the ``core
-team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
-and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
+team'' can end up setting unrealistic expectations by doing too much. Team Geek\todo{add cite}
+and Turk's paper on scaling code in the human dimension\todo{add cite}  discuss how development
 lists that are welcoming to people actually have many more people contributing.
@@
 
+\subsection*{Papers}
+
+\subsubsection*{Development Experiences}
+
+\begin{itemize}
+
+\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
+  McInnes, Barry Smith. Package Management Practices Essential for
+  Interoperability: Lessons Learned and Strategies Developed for
+  FASTMath~\cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
+  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
+
+\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
+  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
+  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
+
+\item Anshu Dubey, B. Van Straalen. Experiences from Software
+  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
+  adapting sustainable scientific software development processes in a
+  small biology lab~\cite{Crusoe_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
+The MVAPICH Project: Evolution and Sustainability of an Open Source
+Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
+  H. Kellogg. Experiences with Automated Build and Test for
+  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
+
+\begin{itemize}
+
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
+  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
+  Versatile Simulation of Large Scale Distributed
+  Systems~\cite{Casanova_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
+  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
+  Sherwin. Simplifying the Development, Use and Sustainability of HPC
+  Software~\cite{Cohen_WSSSPE}
+
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
+  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Best Practices, Challenges, and Recommendations}
+
+\begin{itemize}
+
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
+  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
+
+\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
+  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
+  R. Thakur, Software Abstractions and Methodologies for HPC
+  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
+
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
+  Need Not Be Difficult~\cite{Carver_WSSSPE}
+
+\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
+  K. Barnett, Von Welch. Initial Findings from a Study of Best
+  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
+
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
+  anything less is unsustainable~\cite{Brown_WSSSPE}
+
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
+  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
+  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
+  Platforms~\cite{Swenson_WSSSPE}
+
+\end{itemize}
+
 \subsection{Research or Reuse?}
@@
 
-\todo{some text is needed here before the first subsection}
-
-\subsection{Modeling Sustainability}
+\subsection*{Modeling Sustainability}
 
@@
 
-%\todo{seems like definitions will be mentioned elsewhere in the paper?}
+\todo{seems like definitions will be mentioned elsewhere in the paper?}
 Each of the framework papers included a definition of sustainability, with many overlaps.
@@
 This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
-See \S\ref{sec:cross-cutting} for more on the discussion of definitions of sustainability in the workshop.
 
@@
 
-\subsection{Credit, Citation, Impact}
+\subsection*{Credit, Citation, Impact}
 
@@
 
-\subsection{Implementing Policy}
+\subsection*{Implementing Policy}
 
@@
 
-% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.
-
-% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.
-
-% The authors focus on a point that has been brought up many times in this context:
-% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
-% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
-% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
-% We don’t have a way to validate that the software used in a paper is actually done right.
-% The best way to get software designed correctly is to make sure best practices are considered from the start.
-
-% The authors highlight two models for sustainable software, at different extremes:
-% One is what they call “co-funding”:
-% In these projects, usually large, multi-year collaborations, 
-% there is equal emphasis on both the science and the software development. 
-% Both are planned into the project from inception. 
-% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.
-
-% At the other extreme, they discuss “software carpentry”: 
-% in this model, it’s assumed that the scientists themselves will write and maintain their code.
-% Groups like Software Carpentry and ROpenSci assume that 
-% scientists won’t have access to dedicated software engineering, 
-% so they try to give them tools to use best practices in their own software development.
-
-% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
-% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.
-
-% What do we mean when we refer to the “long tail” of scientific software?
-% Think of the distribution of resources in scientific software.
-Most are focused on big projects with lots of community buy-in and funding.
-But a lot of scientific software exists away from this model.
-% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
-% Look at MacClade: it was originally released in 1986 and last updated in 2005,
-% but it was still cited over 400 times in 2013!
-% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.
-
-% Another dimension of the long tail can also be found in my particular research domain. 
-% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
-% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
-% Most of these programs are developed by academic scientists…
-% They generally have limited training in software engineering
-% Limited time or career incentive to improve software
-% Limited funding
-
-% So, to summarize a bit:
-% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
-% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
-% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.
-
-% Then we have to consider that the users of scientific software are scientists, 
-% so the developers need to understand the users and the science.
-% This is the idea of a “t-skilled” person: 
-% one who is both well-versed in a scientific domain 
-% and deeply experienced in one or more facets of software engineering. 
-% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.
-
-% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
-% Think of it as analogous to DNA sequencing:
-% Sequencers used to be something that individual labs and institutions had to 
-% buy, maintain, and operate themselves, 
-% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
-% But now, core facilities provide the instrumentation and service to labs of any size. 
-% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.
-
-% We propose that software engineering can be “instrumented” in a similar way. 
-% Let’s create a nonprofit center for scientific software engineering.
-% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
-% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
-% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.
-
-% So, to conclude…
-% Implementing policies to encourage sustainability in scientific software 
-% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
-% But most of them aren’t addressed in the status quo.
-% We’ve highlighted some of these facets today and suggested some possible solutions.
-% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
-% Grassroots developer groups can provide guidance to scientists about best practices in software development.
-% We think there is a place for a software engineering center that can provide 
-% both engineering expertise and guidance 
-% with a contract-driven instrumentation model 
-% to the scientific software in the long tail.
+API governance and cybersecurity measures appropriate for a project's
+maturity are all but two facets of sustainability-oriented software
+development. Others include user-centered interface design, test
+engineering, dependency management, and deployment operations. Each of
+these facets require specialized skills and training in software
+engineering. How can the implementation of best practices and
+guidelines along many or all of these different facets be facilitated
+in scientific software development? Blanton and
+Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
+and personnel devoted to software engineering is ``co-funded'' with
+the driving science project. This typically implies large multi-year
+collaborative projects that to succeed require significant software
+infrastructure to be built, and which thus have the funding to support
+one or several software engineers. The sustainability of such projects
+then depends on sustaining the funding. In the other model, domain
+scientists also take the role of software developers, whether by
+necessity such as funding limitations, or due to cross-disciplinary
+professional interests. For this to result in sustainable software
+products, the domain scientists need to be (or become) cross-trained
+in software engineering standards and best practices.
+
+In practice, there are example for both extremes of the spectrum. For
+example, in the life sciences the iPlant Collaborative~\cite{iPlant},
+the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
+multi-year projects with significant software infrastructure funding
+and deliverables. Typically though these are the exception rather than
+the rule, and particularly in the long tail of science the
+scientist-developer predominates. For providing these with better
+training in software engineering best practices, initiatives such as
+Software Carpentry~\cite{SoftwareCarpentry} have been especially
+valuable, and Software Carpentry also shown how such training could
+scale.
+
+However, there may also be a middle ground between the two
+extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
+Excellence model that leverages economies of scale to make software
+engineering experts and their skills accessible to the long tail of
+science. As Huang and Lapp discuss, this model could effectively turn
+the utilization of software engineering expertise from a complex human
+resource recruitment and management challenge, to buying time, when
+and to the extent needed, on shared instrumentation. There is
+precedence to using such a model to lower access barriers for long
+tail science, in particular for new experimental technologies.  For
+example, although the acquisition and operation of next-generation
+high-throughput DNA sequencers requires substantial investments of
+money, time, and expertise, the establishment of ``core facilities''
+on many university campuses has made these technologies accessible to
+a wide swath of scientists, with transformative results for science.
+
+One of the important conclusions from this group of papers is that
+creating sustainable software requires paying attention not to one or
+two, but to several different facets of software engineering, each
+with corresponding best practices and standards of excellence. Even if
+a science project requires and has funding for a full software
+engineering FTE, what the project really needs could be fractions of
+different engineers with different specialty training. The vast
+majority of long tail science projects lacks the funding for even one
+full software engineer, let alone one who combines expertise in all of
+the applicable facets of engineering. Some scientists in the long tail
+will have the professional interests to cross-train enough in software
+engineering to be successful with creating sustainable software, but
+it is unrealistic to expect this expertise and interest of all or even
+the majority. This is where a software engineering center of
+excellence could provide a critical resource by enabling scientists in
+the long tail to complement their resources and expertise with facets
+that are missing, but which, if applied at the right time, would
+improve the chances of a software product to become sustainable. Such
+complementary expertise also need not be restricted to software
+engineering in the strict sense; for example, it could consist of
+community building, leadership, and support for some period of
+transition to sustainability.
+
+In summary, implementing software sustainability practices on a
+broader basis requires on the one hand the development of guidelines
+and practices that are suitable for the exploratory research context
+in which most scientific software is created, and on the other hand a
+skilled workforce trained in a variety of software engineering facets
+and community building. The capabilities afforded by such a workforce
+need to be accessible not only to large projects with sufficient
+funding to provide competitive employment, but also to the many
+smaller projects in the long tail of science. Sufficiently
+cross-training domain scientists could be one way to achieve this;
+another, complementary approach is to instrument the necessary
+capabilities so they can be shared.
+
+\subsection*{Papers}
 
@@
 
+\subsubsection*{Modeling Sustainability}
+
+\begin{itemize}
+
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
+  Software Product Sustainability Model~\cite{Calero_WSSSPE}
+
+\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
+  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
+  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
+  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing
+  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
+  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
+  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
+  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Credit, Citation, Impact}
+
+\begin{itemize}
+
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
+  Smith. Accurately Citing Software and Algorithms Used in
+  Publications~\cite{Knepley_WSSSPE}
+
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
+  report for every software project~\cite{Priem_WSSSPE}
+
+\item Daniel S. Katz. Citation and Attribution of Digital Products:
+  Social and Technological Concerns~\cite{Katz2_WSSSPE}
+
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
+  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Reproducibility}
+
+\begin{itemize}
+
+\item Victoria Stodden, Sheila Miguez. Best Practices for
+  Computational Science: Software Infrastructure and Environments for
+  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Implementing Policy}
+
+\begin{itemize}
+
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
+  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Daisie Huang, Hilmar Lapp. Software Engineering as
+  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
+
+\item Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
+
+\end{itemize}
+
 \subsection{Career Tracks for Scientific Software Developers}
@@
 
-%There were a number of papers categorized under the Communities banner and so
-%two presenters each summarized half of the submissions in this area.
-\todo{an intro to the whole section is needed}
-
-\subsection{Communities}
-
-%This collection of papers was summarized by Karen Cranston.
+\todo{section should include what was presented and what was discussed}
+
+There were a number of papers categorized under the Communities banner and so
+two presenters each summarized half of the submissions in this area.
+
+\subsection{Communities Part 1}
+
+This collection of papers was summarized by Karen Cranston.
 
@@
 
-%\subsection{Communities Part 2}
-
-%Nancy Wilkins-Diehr summarized 6 papers in this part.
+\subsection{Communities Part 2}
+
+Nancy Wilkins-Diehr summarized 6 papers in this part.
 
@@
 
-\subsubsection{What are communities?}
+\subsection*{Papers}
+
+\subsubsection*{Communities Part 1 Papers}
+
+\begin{itemize}
+
+\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
+  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
+  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
+  Modeling Science Activities~\cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J.
+Krieder, Justin M.
+Wozniak, Daniel S.
+Katz, Mei Zhi-Gang, Mainak Mookherjee.
+Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Communities Part 2 Papers}
+
+\begin{itemize}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
+  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
+  Experiences and Challenges in Software Sustainability for a Large
+  Scientific Community~\cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
+Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed
+in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Industry \& Economic Models}
+
+\begin{itemize}
+
+\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
+
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
+  as a path to software sustainability~\cite{Foster_WSSSPE}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
+  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
+  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Education \& Training}
+
+\begin{itemize}
+
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
+  T. Brown. Advanced Techniques for Scientific Programming and
+  Collaborative Development of Open Source Software Packages at the
+  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
+
+\item Thomas Crawford. On the Development of Sustainable Software for
+  Computational Chemistry~\cite{Crawford_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\end{itemize}
+
+\subsection{What are communities?}
 
@@
 
-\subsubsection{Challenges of community}
+\subsection{Challenges of community}
 include metrics for community success (e.g., more external
 contributors than internal).
-
-\subsubsection{Admirable scientific software communities}
+\subsection{Admirable scientific software communities}
 \begin{itemize}
@@
 \end{itemize}
-
-\subsubsection{Resources for learning about software communities}
+\subsection{Resources for learning about software communities}
 \begin{itemize}
@@
 
-\subsection{Industry \& Economic Models}
-
-\todo{text about papers and discussion in this part of the workshop}
-
-\subsection{Education \& Training}
-
-\todo{text about papers and discussion in this part of the workshop}
-
 %\section{Other Discussion} \label{sec:other}
@@
 software architectures determines sustainability.  They then propose that
-sustainability is a consequence of a set of central quality attributes:
+sustainability is a measure of a set of central quality attributes:
 extensibility, interoperability, maintainability, portability, reusability, and
@@
 
-In this section, we discuss some of the software projects as case studies to
-better understand the points discussed during the workshop and described in the
-previous sections. We dive deeper to find how are they affected by the
-sustainability related issues in practice. Additionally, we discuss issues
-surrounding people, especially the career paths of original developers and
-investigate to answer the question: are the original developers are still
-around?
+% TODO: Why are we doing case studies?
+Many important issues related to scientific software sustainability were
+discussed in the previous sections. In this section we discuss some of the
+software projects as case-studies to better understand the relevance of those
+issues and how are they connected to the software sustainability and related
+factors.
 
 We classify the software projects discussed in the workshop in two broad
-categories. First, the \emph{utility} software, comprising of general purpose
-software. Utility software is often used as enabler and/or facilitator for the
-development of other tools and techniques to carry out scientific work. This
-includes the software developed to efficiently utilize new research
-infrastructures. Second, the \emph{scientific} software, comprising the
-software that was originally developed with an aim to solve a specific
-scientific problem.
+categories. First, the utility software, comprising of software that enable
+and/or facilitate the development of other tools and techniques to carry out
+scientific work. This includes the software developed to efficiently utilize
+new research infrastructures. Second, scientific software, comprising software
+that was developed with an aim to solve a specific scientific problem.
 
 % Why are we classifying software like above?
-This classification is motivated by our argument that the two kinds of software
-projects wildly vary in factors such as scope, purpose and usage. The
-development and management of each kind is significantly different.
-Consequently, the sustainability challenges faced by them differ and must be
-treated separately. For instance, the challenges faced by a gateway software
-development project such as \textit{CIPRES} or a visualization software product
-such as \textit{VisIT} are distinct to a niche software for \textit{ab initio}
-modeling and simulation such as \textit{VASP} or \textit{Quantum Espresso}.
+This classification is motivated by our argument that the software projects
+wildly vary in factors such as scope, purpose and usage. The development and
+management of software is significantly influenced by these factors.
+Consequently, the sustainability challenges faced by software differ and must
+be treated separately. For instance, the challenges faced by a gateway software
+development project such as \textit{CIPRES} or \textit{VisIT} are distinct to a
+niche software for ab initio modeling and simulation such as \textit{VASP} or
+\textit{Quantum Espresso}.
+
+% Below: What are the goals of this discussion? TODO: Better motivate. May be make a brief list of items discussed in the workshop.
+We aim to understand the practicalities of the key points discussed in the
+workshop, and hope to extract the best practices associated with the case
+studies. We discuss how the issues are applicable to the cases and try to
+understand how the course of current and future development projects might be
+altered to implement lessons learned. We discuss vatious issues including
+career paths of original developers and investigate if the original developers
+are still around?
 
 \subsection{Utility Software}
-Software developed with a potentially wider audience and general purpose usage
-in mind is utility software. Utility software typically does not address
-fundamental research problems for a given scientific domain. Examples are
-collaborative development frameworks such as \emph{GitHub} and
-\emph{Bitbucket}, distributed workflow and generic computing frameworks
-such as \emph{Galaxy}, \emph{nanoHUB}, \emph{SimGrid}, \emph{Swift},
-\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
-\emph{VisIT}.
+
+We classify software developed with a potentially wider audience and general
+purpose usage in mind as Utility software. In the context of research, a
+utility software could be defined as general purpose system applicable to one
+or many generic tasks and/or enabling other software to run by providing a
+suitable environment. Utility software typically do not address a fundamental
+research challenge in a given scientific domain. Examples are collaborative
+development frameworks such as \emph{GitHub} and \emph{Bitbucket}, distributed
+and workflow and generic computing frameworks such as \emph{Galaxy},
+\emph{nanoHUB}, \emph{SimGrid}, \emph{Swift}, \emph{Globus} and \emph{VisTrails}, and
+visualization frameworks such as \emph{VisIT}.
 
 % What are the sustainability challenges? What are the oppurtunities?
-Development is often a high risk/reward undertaking exploring uncharted
-territories and is largely influenced by (re)usability factors. Owing to
+Development is often a high risk/reward undertaking and explores uncharted
+territories and is largely dominated by (re)usability factors. Owing to
 relatively large number of features, the development and prototype process is
-also lengthy which poses a significant survival risk. Challenges on a class of
-utility software for new architectures is well discussed
-in~\cite{Ferenbaugh_WSSSPE}.
-
-On the other hand, utility software presents oppurtunities to be usable by a
-larger community making its undertaking and development an attractive pursuit.
-It is generally more visible in community which in turn leads to a broader and
-deeper participation. For instance, promoting collaborations across the breadth
-(e.g.  different departments) and depth (e.g. stakeholders within a department)
-of community, one of the key ingredient of a sustainable process. Successful
-utility projects reap high rewards and have a longer usage span. Development
-process becomes user-driven and self-sustaining.
+also lengthy which poses a significant survival risk.
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} further discusses the risks and challenges
+associated with software projects dealing with new architectures.
+
+On the other hand, owing to its generic nature, utility software presents
+oppurtunities to be usable by a larger community. This makes it more visible
+which in turn leads to a broader and deeper participation. For
+instance, promoting collaborations across the breadth (e.g. different
+departments) and depth (e.g. stakeholders within a department) of community,
+one of the key ingredient of a sustainable process. Successful projects reap
+high rewards and have longer usage span. Development process becomes
+user-driven and self-sustaining.
 
 One such example is the \emph{Galaxy} project. It follows agile software
-development practices and implements standards such as test-driven development
-and socialized bug managing practices via \emph{trello}. Galaxy
+development practices and implements standard practices such as test-driven
+development and socialized bug managing practices via \emph{trello}. Galaxy
 \emph{histories} and \emph{toolshed} offers easy community sharing of data and
@@
 e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
-emerging architectures such as accelerators, new code and libraries are
+emerging architectures such as CPU accelerators new code and libraries are
 required. The experience of one such effort as described
-in~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
-with many invaluable lessons were learned about influential cultural and
-technical aspects in sustainable software development practices.
+in~\cite{Ferenbaugh_WSSSPE} met with a limited success but nonetheless with
+many invaluable lessons were learned about influential cultural and technical
+aspects in sustainable software development practices.
 
@@
 computational environments, many users are leaning towards tools used as
-services. \emph{GitHub} and \emph{Bitbucket} can be argued to be
+services. \emph{GitHub} and \emph{Bitbucket} can be arguably considered to be
 such tools, catering to collaborative development. For scientific users
@@
 \subsection{Scientific software}
+
 Scientific software consists primarily of special-purpose software that was
-purpose-built for a target use-case scenario,  fixed/frozen requirements in
-mind or solving a specific problem. Software projects pertaining to specific
-scientific domain tend to be in a niche and the user community tends to be
-small to medium. They are mostly driven by the science and specific needs of a
-research group. Specific needs such as numerical accuracy and
-algorithmic optimization are some of the paramount requirements of most
-scientific software.
+purpose-built for a target use-case scenario or fixed/frozen requirements in
+mind.  Software projects pertaining to specific scientific domain often tend to
+be in a niche and the user community tends to be small to medium. They are
+mostly driven by the science and specific needs of a research group.
 
 % What are the challenges?
-Long-term sustainability of scientific software is often a significant
-challenge and face radically different issues compared to utility software.
-Many submissions reported that software is practically considered a `byproduct'
-of the actual research. Others contended that the software was not the main
-funded part of their research. A smaller codebase and fixed requirements result
-in stability, ease of installation, and configurability.  Many such projects
-mature and are treated as libraries to be used into larger systems such as some
-of the utility software discussed in the previous section. While the software
-can stay stable and require relatively low maintenance, the responsibility is
-often on the shoulders of a very few developers who might not be specialists in
-software development. Development tends to be linear and simplistic with a
-limited scope to follow software best practices.
+Long-term sustainability of scientific research software is often a significant
+challenge. Many submissions reported that software is practically considered a
+`byproduct' of the actual research. Others contended that the software was not
+the main funded part of their research.  Specific needs such as numerical
+accuracy and algorithmic optimization are some of the paramount requirements of
+most scientific software.  The resulting codebase is relatively small. A
+smaller codebase and fixed requirements result in stability, ease of
+installation, and configurability.  Many such projects mature and are treated
+as libraries to be used into larger systems such as some of the utility
+software discussed in the previous section.  While the software can stay stable
+and require relatively low maintenance, the responsibility is often on the
+shoulders of a very few developers who might not be specialists in software
+development. Development tends to be linear and simplistic with a limited
+scope to follow software best practices.
+
+% TODO: What are the oppurtunities?
 
@@
 % 
-
-Finally, it appears that the career options of developers from different kinds
-of software projects differ significantly. In order to further investigate the
-much discussed issue of career paths for developers of these software, we
-conducted a small survey of the use-cases discussed above with respect to the
-current status of its core developers. We looked at the workshop submissions
-and the original cited papers for the respective software. Where available, we
-looked at the webpages and source code of the earliest versions to find the
-original developers. We included projects older than 5 years and excluded
-relatively new projects such as \emph{SciPy} and \emph{numFOCUS} from this
-survey. we find that in a majority of scientific software more than one
-original developers are still around, in some cases for more than 10 years,
-(eg. \emph{R/qtl}).  Whereas, in the case of utility software projects, it
-appears that at least one original developer is still around, often in a
-part-time or ``benefactor" role. In other cases, it appears that the original
-developers have went on to assume a larger role, with the same project.
+In order to investigate the much discussed issue of career paths for developers
+of scientific software, we conducted a small survey of the use-cases discussed
+above with respect to the current status of its core developers. We looked at
+the workshop submissions and the original cited papers for the respective
+software. Where available, we looked at the webpages and source code of the
+earliest versions to find the original authors. We excluded relatively new
+projects such as \emph{SciPy} and \emph{numFOCUS} from this survey. we find
+that in a majority of scientific software more than one original developers are
+still around, in some cases for more than 10 years, (eg. \emph{R/qtl}).
+Whereas, in the case of utility software projects, it appears that at least one
+original developer is still around often in a part-time role. In other cases,
+it appears that the original developers have went on to assume a larger role,
+again with the same project.
 
 In conclusion, we find that there are distinctions in the characteristics and
-the kinds of challenges faced between utility and scientific software projects.
-In terms of career paths of original developers, it appears that the career
-choices for utility software developers are broader compared to the developers
-of scientific software. However, it appears that in the case of scientific
-software, core competencies of original developers tend to be domain science
-who happen to be the coders of their software. As projects as a whole, the
-small-scale, specific software have a lower risk and cost in terms of
-sustaining as compared to utility software. We also find that the utility
-software tend to adhere more loyally to the many best practices discussed
-during the workshop. Often the sustainability of scientific software projects
-is achieved by the fact that the core developer or team heavily utilizes the
-software for their science, e.g.,  \emph{R/qtl, PETSc}.  Furthermore, the
-development of scientific software requires more scientific skills compared
-with those of utility software. This diminishes the need of specialist
-programmers and practices in many cases where the bulk of development is
-carried out by a domain scientist. With these inferences, we believe that
-separate guidelines and/or principles for sustainability be defined and
-practiced for each category of scientific software and they be given separate
-treatment.
+the kinds of challenges faced between generic utility and specific scientific
+domain software projects. In terms of career paths of original developers, it
+appears that the career paths for utility software developers are broader
+compared to the developers of scientific software.  However, it appears that in
+the case of scientific software, core competencies of original developers tend
+to be domain science who happen to be the coders of their software. As projects
+as a whole, the small-scale, specific software have a lower risk and cost in
+terms of sustaining as compared to utility software.  Often the sustainability
+of such projects is achieved by the fact that the core developer or team
+heavily utilizes the software for their science, eg.  \emph{R/qtl, PETSc}.
+Furthermore, the development of scientific software requires more scientific
+skills compared with those of utility software. This diminishes the need of
+specialist programmers and practices in many cases where the bulk of
+development is carried out by a domain scientist. With these inferences, we
+believe that separate guidelines and/or principles for sustainability be
+defined and practiced for each category of scientific software and they be
+given separate treatment.
 
@@
 \todo{say something about licensing - lessons, advice, etc.?}
-
-\subsection{Top Issues}
 
@@
 \appendix
-\section{Papers Accepted and Discussed at WSSSPE1}
-
-\section*{Developing and Supporting Software}
-
-\subsubsection*{Development Experiences}
-
-\begin{itemize}
-
-\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
-  McInnes, Barry Smith. Package Management Practices Essential for
-  Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath~\cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
-
-\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
-  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
-
-\item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
-  adapting sustainable scientific software development processes in a
-  small biology lab~\cite{Crusoe_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
-The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
-  H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
-
-\begin{itemize}
-
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
-  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
-  Versatile Simulation of Large Scale Distributed
-  Systems~\cite{Casanova_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
-  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
-  Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software~\cite{Cohen_WSSSPE}
-
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Best Practices, Challenges, and Recommendations}
-
-\begin{itemize}
-
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
-
-\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
-  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
-  R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
-
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult~\cite{Carver_WSSSPE}
-
-\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
-  K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
-
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable~\cite{Brown_WSSSPE}
-
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
-  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
-  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms~\cite{Swenson_WSSSPE}
-
-\end{itemize}
-
-\section*{Policy}
-
-\subsubsection*{Modeling Sustainability}
-
-\begin{itemize}
-
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model~\cite{Calero_WSSSPE}
-
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Credit, Citation, Impact}
-
-\begin{itemize}
-
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
-  Smith. Accurately Citing Software and Algorithms Used in
-  Publications~\cite{Knepley_WSSSPE}
-
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project~\cite{Priem_WSSSPE}
-
-\item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns~\cite{Katz2_WSSSPE}
-
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Reproducibility}
-
-\begin{itemize}
-
-\item Victoria Stodden, Sheila Miguez. Best Practices for
-  Computational Science: Software Infrastructure and Environments for
-  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Implementing Policy}
-
-\begin{itemize}
-
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
-
-\item Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
-
-\end{itemize}
-
-\section*{Communities, Models, and Education}
-
-\subsubsection*{Communities}
-
-\begin{itemize}
-
-\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities~\cite{Vay_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community~\cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area was also discussed
-in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Industry \& Economic Models}
-
-\begin{itemize}
-
-\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
-
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability~\cite{Foster_WSSSPE}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
-  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Education \& Training}
-
-\begin{itemize}
-
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
-  T. Brown. Advanced Techniques for Scientific Programming and
-  Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
-
-\item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry~\cite{Crawford_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\end{itemize}
-
-\section{Attendees \note{lead: Shel Swenson}}
+\section{List of attendees \note{lead: Shel Swenson}}
 
@@
 
-\bibliographystyle{vancouver}
-
-\bibliography{wssspe_paper}
+\bibliographystyle{unsrt}
+
+\bibliography{wssspe}
 \end{document}

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex

--- 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-% Finally, we’ll look at some papers that discuss implementing sustainability in scientific software.
-
-% Blanton and Lenhardt from the Renaissance Computing Institute discuss these issues from a user perspective.
-
-% The authors focus on a point that has been brought up many times in this context:
-% There is a tension between writing code that is good enough just to “get it done,” i.e. to publish a paper about scientific results obtained using software, 
-% and “getting it right,” that is, developing software that is comprehensible to future users and reviewers.
-% Just because the elevator panel works like this doesn’t mean it’s sustainable for the long run.
-% We don’t have a way to validate that the software used in a paper is actually done right.
-% The best way to get software designed correctly is to make sure best practices are considered from the start.
-
-% The authors highlight two models for sustainable software, at different extremes:
-% One is what they call “co-funding”:
-% In these projects, usually large, multi-year collaborations, 
-% there is equal emphasis on both the science and the software development. 
-% Both are planned into the project from inception. 
-% In the life sciences, the iPlant Collaborative, Galaxy Project, and Qiime are good examples of these sorts of large, well-designed projects.
-
-% At the other extreme, they discuss “software carpentry”: 
-% in this model, it’s assumed that the scientists themselves will write and maintain their code.
-% Groups like Software Carpentry and ROpenSci assume that 
-% scientists won’t have access to dedicated software engineering, 
-% so they try to give them tools to use best practices in their own software development.
-
-% There might be a middle ground here: a way to get the engineering expertise that large co-funded projects have to individual scientist-developers.
-% Hilmar Lapp of NESCent and I discuss one such possibility in our paper, Software Engineering as Instrumentation for the Long Tail of Scientific Software.
-
-% What do we mean when we refer to the “long tail” of scientific software?
-% Think of the distribution of resources in scientific software.
-Most are focused on big projects with lots of community buy-in and funding.
-But a lot of scientific software exists away from this model.
-% For example, scientific software can be used long after the original developer has moved on or the funding runs out.
-% Look at MacClade: it was originally released in 1986 and last updated in 2005,
-% but it was still cited over 400 times in 2013!
-% The scientists who developed it have a newer package, Mesquite, that was meant to replace MacClade, but they haven’t had sufficient time or resources to maintain either package fully, let alone both of them.
-
-% Another dimension of the long tail can also be found in my particular research domain. 
-% In the field of phylogenetics, we have a lot of programs that implement different computational methods in slightly different ways.
-% Here, Joe Felsenstein has listed some (but not anywhere near all) phylogenetics packages available online.
-% Most of these programs are developed by academic scientists…
-% They generally have limited training in software engineering
-% Limited time or career incentive to improve software
-% Limited funding
-
-% So, to summarize a bit:
-% Making sustainable software means we have to pay attention to many facets of software design, like APIs, security, user experience, testing, etc.
-% A single project that requires one full­-time software engineer may actually require fractions of different kinds of engineers. 
-% But long-tail projects can’t even fund one FTE, let alone one that can address all these facets.
-
-% Then we have to consider that the users of scientific software are scientists, 
-% so the developers need to understand the users and the science.
-% This is the idea of a “t-skilled” person: 
-% one who is both well-versed in a scientific domain 
-% and deeply experienced in one or more facets of software engineering. 
-% These people are pretty rare in the first place and difficult to retain in academia, because the academic career structure doesn’t incentivize this.
-
-% We should look at software engineering as an expensive resource, but one that needs to be accessible to scientists at all levels.
-% Think of it as analogous to DNA sequencing:
-% Sequencers used to be something that individual labs and institutions had to 
-% buy, maintain, and operate themselves, 
-% so only highly-funded operations had them and probably didn’t use them to their full capacity even when they had one.
-% But now, core facilities provide the instrumentation and service to labs of any size. 
-% Anyone can pay a core facility to sequence their samples for them and provide quality control and bioinformatics advice as additional services.
-
-% We propose that software engineering can be “instrumented” in a similar way. 
-% Let’s create a nonprofit center for scientific software engineering.
-% This center can hire these t-skilled personnel and provide access to them for projects at contracted cost.
-% Because the center is focused on providing development services to scientific projects, it is not tied to the long-term success or failure of any individual project.
-% It would emphasize the centrality of doing good science by making functional software tools as envisioned by scientists.
-
-% So, to conclude…
-% Implementing policies to encourage sustainability in scientific software 
-% requires that many facets of good software design are addressed throughout the lifecycle of these projects. 
-% But most of them aren’t addressed in the status quo.
-% We’ve highlighted some of these facets today and suggested some possible solutions.
-% Large projects can afford to hire software engineers with the expertise to implement these facets correctly.
-% Grassroots developer groups can provide guidance to scientists about best practices in software development.
-% We think there is a place for a software engineering center that can provide 
-% both engineering expertise and guidance 
-% with a contract-driven instrumentation model 
-% to the scientific software in the long tail.
+API governance and cybersecurity measures appropriate for a project's
+maturity are all but two facets of sustainability-oriented software
+development. Others include user-centered interface design, test
+engineering, dependency management, and deployment operations. Each of
+these facets require specialized skills and training in software
+engineering. How can the implementation of best practices and
+guidelines along many or all of these different facets be facilitated
+in scientific software development? Blanton and
+Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
+and personnel devoted to software engineering is ``co-funded'' with
+the driving science project. This typically implies large multi-year
+collaborative projects that to succeed require significant software
+infrastructure to be built, and which thus have the funding to support
+one or several software engineers. The sustainability of such projects
+then depends on sustaining the funding. In the other model, domain
+scientists also take the role of software developers, whether by
+necessity such as funding limitations, or due to cross-disciplinary
+professional interests. For this to result in sustainable software
+products, the domain scientists need to be (or become) cross-trained
+in software engineering standards and best practices.
+
+In practice, there are example for both extremes of the spectrum. For
+example, in the life sciences the iPlant Collaborative~\cite{iPlant},
+the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
+multi-year projects with significant software infrastructure funding
+and deliverables. Typically though these are the exception rather than
+the rule, and particularly in the long tail of science the
+scientist-developer predominates. For providing these with better
+training in software engineering best practices, initiatives such as
+Software Carpentry~\cite{SoftwareCarpentry} have been especially
+valuable, and Software Carpentry also shown how such training could
+scale.
+
+However, there may also be a middle ground between the two
+extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
+Excellence model that leverages economies of scale to make software
+engineering experts and their skills accessible to the long tail of
+science. As Huang and Lapp discuss, this model could effectively turn
+the utilization of software engineering expertise from a complex human
+resource recruitment and management challenge, to buying time, when
+and to the extent needed, on shared instrumentation. There is
+precedence to using such a model to lower access barriers for long
+tail science, in particular for new experimental technologies.  For
+example, although the acquisition and operation of next-generation
+high-throughput DNA sequencers requires substantial investments of
+money, time, and expertise, the establishment of ``core facilities''
+on many university campuses has made these technologies accessible to
+a wide swath of scientists, with transformative results for science.
+
+One of the important conclusions from this group of papers is that
+creating sustainable software requires paying attention not to one or
+two, but to several different facets of software engineering, each
+with corresponding best practices and standards of excellence. Even if
+a science project requires and has funding for a full software
+engineering FTE, what the project really needs could be fractions of
+different engineers with different specialty training. The vast
+majority of long tail science projects lacks the funding for even one
+full software engineer, let alone one who combines expertise in all of
+the applicable facets of engineering. Some scientists in the long tail
+will have the professional interests to cross-train enough in software
+engineering to be successful with creating sustainable software, but
+it is unrealistic to expect this expertise and interest of all or even
+the majority. This is where a software engineering center of
+excellence could provide a critical resource by enabling scientists in
+the long tail to complement their resources and expertise with facets
+that are missing, but which, if applied at the right time, would
+improve the chances of a software product to become sustainable. Such
+complementary expertise also need not be restricted to software
+engineering in the strict sense; for example, it could consist of
+community building, leadership, and support for some period of
+transition to sustainability.
+
+In summary, implementing software sustainability practices on a
+broader basis requires on the one hand the development of guidelines
+and practices that are suitable for the exploratory research context
+in which most scientific software is created, and on the other hand a
+skilled workforce trained in a variety of software engineering facets
+and community building. The capabilities afforded by such a workforce
+need to be accessible not only to large projects with sufficient
+funding to provide competitive employment, but also to the many
+smaller projects in the long tail of science. Sufficiently
+cross-training domain scientists could be one way to achieve this;
+another, complementary approach is to instrument the necessary
+capabilities so they can be shared.
 

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Breaks lines after 70 chars in Policy.
This makes it much easier to read online (such as on Github), compared to if
paragraphs are just a single line.
Changes presentation of Maturity Models.
These changes are in response to Von Welch's comment in the Google Doc tha a
"maturity model for scientific software would be generally useful, with
security as one example."

@@
 
-Frameworks provide a scaffold for thinking; they help us understand which elements are important and how they relate.
-They also help us break down a problem into chunks that can be pursued by different groups or perhaps disciplines.
-
-A group of papers submitted to the workshop provide frameworks with these intentions.
-The frameworks have substantial overlap but with different emphases and extents.
-Accordingly we present them starting with the area of most overlap and proceed to show the different emphases and the range of factors included in the frameworks.
-
-%\todo{seems like definitions will be mentioned elsewhere in the paper?}
-Each of the framework papers included a definition of sustainability, with many overlaps.
-In essence the papers related the question to change over time, an idea well captured in the definition used by the UK's Scientific Software Sustainability Institute: ``software you use today will be available---and continue to be improved and supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}.
-This idea was also expressed as software that continues to serve its users~\cite{Pierce_WSSSPE}.
-See \S\ref{sec:cross-cutting} for more on the discussion of definitions of sustainability in the workshop.
-
-One area in which there was not complete overlap was whether the word (and thus the effort called for by the WSSSPE workshop) involved environmental sustainability.
-Of course the word sustainability has strong connotations from consideration of environmental issues, evoking some mention of the areas in which software interacts with overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned this did so without integrating that analysis into the question of software being around long-term, suggesting that questions of environmental impact of scientific software is a conceptually distinct area of inquiry.
-
-One group of papers presented frameworks that were primarily about characteristics of software artifacts, connecting with the long discourse on software quality.
-This approach is realized in adjectives that can be applied to pieces of software but might also extend to describe software projects.
-Thus Calero and colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO standards for measuring software quality: proposing an additional dimension of quality they call perdurability or ability to ``last over time'' made up of three characteristics: reliability, maintainability and adaptability.
-This shows some overlap with the framework employed by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of ``extensibility, interoperability, maintainability, portability, reusability and scalability,'' anticipating the sorts of work that people would need to do with a software artifact in the future, ``as stakeholders requirements, technology and environments evolve and change.'' Venters et al.
-argue that these choices are really made very early because they relate to the architecture of the software and involve trade offs that ought to be analyzed alongside other attributes of the software artifact.
-Lenhardt et al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data lifecycle to argue for the inclusion of metadata throughout a piece of software's life (discussing, for example, how it has been built and tested and what ``data in'' and ``data out'' has been considered).
-In addition, their analogy suggests that the software lifecycle might add a phase of ``preservation'' and draw on understandings of what that involves from studies of data.
-In sum, then, these frameworks focused on what needs to be accomplished to have more sustainable software.
-
-A second theme in these papers was the continued availability of resources to accomplish the goals of sustainability.
-The elements of these frameworks focused far more on the organization of a software project than they did on characteristics of the artifact itself (although it is certainly true that the adjectives discussed above could be applied to a software project).
-For example Pierce et al~\cite{Pierce_WSSSPE} focus on the way that the project is run, particularly in terms of how those involved communicate and jointly set priorities, a process they call governance.
-In particular they argue that because sustainability relates to having ongoing resources, governance must be open to receive diverse input (by occurring online, asynchronously, and publicly) and thus have the potential to ``transform passive users into active contributors''.
-They argue that the Apache Software Foundation's incubation process teaches this and could be learned from by projects throughout scientific software.
-Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance, describing two modes: ``top-down'' and ``bottom-up'' governance.
-They place governance alongside questions of the artifact (``technical''), and questions of the origins of resources, in terms of who is funding the work surrounding a piece of software (``the political'') and the manner in which resources come to the project initially (commercial, open source, closed partnerships, grant funded) and long-term (all four plus paid support).
-
-Frameworks also addressed themselves to the context in which software projects exist (thus moving upwards in abstraction, from the artifact itself, to the organization of its production and, in these frameworks, the shape of the space which an artifact or project exists in).
-These frameworks take the form of contingency theories in that they outline a different set of challenges and argue that different project organizations (and presumably artifact attributes) are necessary to persist longterm in spaces with particular characteristics.
-Katz and Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in terms of three axes: temporal (long or short term needs), spatial (local or global use) and purpose (specific to general).
-They propose that different project organizations will be needed in different locations and argue that we should concentrate research to understand those connections.
-Weber et al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural ecosystems as ``niches'' which sustain particular pieces of software.
-They define ``a software niche as the set of technical requirements, organizational conditions and cultural mores that support its maintenance and use over time.'' They call for better understanding and modeling of niches (as well as further exploration of the usefulness of ecosystem metaphors).
+Frameworks provide a scaffold for thinking; they help us understand
+which elements are important and how they relate. They also help us
+break down a problem into chunks that can be pursued by different
+groups or perhaps disciplines.
+
+A group of papers submitted to the workshop provide frameworks with
+these intentions. The frameworks have substantial overlap but with
+different emphases and extents.  Accordingly we present them starting
+with the area of most overlap and proceed to show the different
+emphases and the range of factors included in the frameworks.
+
+% \todo{seems like definitions will be mentioned elsewhere in the
+% paper?}
+Each of the framework papers included a definition of sustainability,
+with many overlaps.  In essence the papers related the question to
+change over time, an idea well captured in the definition used by the
+UK's Scientific Software Sustainability Institute: ``software you use
+today will be available---and continue to be improved and
+supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
+idea was also expressed as software that continues to serve its
+users~\cite{Pierce_WSSSPE}.  See \S\ref{sec:cross-cutting} for more on
+the discussion of definitions of sustainability in the workshop.
+
+One area in which there was not complete overlap was whether the word
+(and thus the effort called for by the WSSSPE workshop) involved
+environmental sustainability. Of course the word sustainability has
+strong connotations from consideration of environmental issues,
+evoking some mention of the areas in which software interacts with
+overall environmental resource usage, particularly energy efficiency.
+The two papers in this area which mentioned this did so without
+integrating that analysis into the question of software being around
+long-term, suggesting that questions of environmental impact of
+scientific software is a conceptually distinct area of inquiry.
+
+One group of papers presented frameworks that were primarily about
+characteristics of software artifacts, connecting with the long
+discourse on software quality. This approach is realized in adjectives
+that can be applied to pieces of software but might also extend to
+describe software projects.  Thus Calero and
+colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO
+standards for measuring software quality: proposing an additional
+dimension of quality they call perdurability or ability to ``last over
+time'' made up of three characteristics: reliability, maintainability
+and adaptability. This shows some overlap with the framework employed
+by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
+``extensibility, interoperability, maintainability, portability,
+reusability and scalability,'' anticipating the sorts of work that
+people would need to do with a software artifact in the future, ``as
+stakeholders requirements, technology and environments evolve and
+change.'' Venters et al. argue that these choices are really made very
+early because they relate to the architecture of the software and
+involve trade offs that ought to be analyzed alongside other
+attributes of the software artifact. Lenhardt et
+al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
+lifecycle to argue for the inclusion of metadata throughout a piece of
+software's life (discussing, for example, how it has been built and
+tested and what ``data in'' and ``data out'' has been considered). In
+addition, their analogy suggests that the software lifecycle might add
+a phase of ``preservation'' and draw on understandings of what that
+involves from studies of data.  In sum, then, these frameworks focused
+on what needs to be accomplished to have more sustainable software.
+
+A second theme in these papers was the continued availability of
+resources to accomplish the goals of sustainability. The elements of
+these frameworks focused far more on the organization of a software
+project than they did on characteristics of the artifact itself
+(although it is certainly true that the adjectives discussed above
+could be applied to a software project). For example Pierce et
+al~\cite{Pierce_WSSSPE} focus on the way that the project is run,
+particularly in terms of how those involved communicate and jointly
+set priorities, a process they call governance. In particular they
+argue that because sustainability relates to having ongoing resources,
+governance must be open to receive diverse input (by occurring online,
+asynchronously, and publicly) and thus have the potential to
+``transform passive users into active contributors''. They argue that
+the Apache Software Foundation's incubation process teaches this and
+could be learned from by projects throughout scientific software.
+Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
+describing two modes: ``top-down'' and ``bottom-up'' governance.  They
+place governance alongside questions of the artifact (``technical''),
+and questions of the origins of resources, in terms of who is funding
+the work surrounding a piece of software (``the political'') and the
+manner in which resources come to the project initially (commercial,
+open source, closed partnerships, grant funded) and long-term (all
+four plus paid support).
+
+Frameworks also addressed themselves to the context in which software
+projects exist (thus moving upwards in abstraction, from the artifact
+itself, to the organization of its production and, in these
+frameworks, the shape of the space which an artifact or project exists
+in). These frameworks take the form of contingency theories in that
+they outline a different set of challenges and argue that different
+project organizations (and presumably artifact attributes) are
+necessary to persist longterm in spaces with particular
+characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
+of this idea of space in terms of three axes: temporal (long or short
+term needs), spatial (local or global use) and purpose (specific to
+general).  They propose that different project organizations will be
+needed in different locations and argue that we should concentrate
+research to understand those connections. Weber et
+al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
+ecosystems as ``niches'' which sustain particular pieces of
+software. They define ``a software niche as the set of technical
+requirements, organizational conditions and cultural mores that
+support its maintenance and use over time.'' They call for better
+understanding and modeling of niches (as well as further exploration
+of the usefulness of ecosystem metaphors).
 
@@
 
-The policy panel and discussion addressed the question of recognition of work on scientific software and linked that recognition to questions of reward and thus motivation for particular kinds of work on scientific software.
-In short, software work in science was seen to be inadequately visible in ways that ``count'' within the reputation system underlying science.
-In his paper for this workshop, Katz placed software work along with other ``activities that facilitate science but are not currently rewarded or recognized''~\cite{Katz_WSSSPE}.
-Priem and Piwowar argued for the need to ``support all researchers in presenting meaningful impact evidence in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
-Kneply et al.
-argued that the visibility of software that supported a piece of science ``can have detrimental effects on funding, future library development, and even scientific careers.''~\cite{Knepley_WSSSPE}.
-
-These papers, and the discussion at the workshop, join a nascent literature seeking to understand what drives software work in science and how the reward systems of science thereby shape the type of software work undertaken, including the extent to which developers are motivated to build software for the long-term, for the use of others and whether to work collaboratively or separately~\cite{howison_incentives_2013, howison_scientific_2011, bietz_synergizing_2010}.
-Software work is not only motivated by direct citations, but the visibility of software work in the literature is important to those who write software used in science.
-
-Papers and discussion concentrated in three areas: How ought software work to be visible, what are the barriers to its visibility, and what can be done to make it more visible?
-
-Most of the papers in this area focused on visibility of software in scientific papers, since scientific papers are the most widely accepted documentation of achievement in science.
-It was noted that there appear to be no widely accepted standards on how the use of software towards a paper ought to be mentioned, that journals, citation style guides and other guides to scientific conduct are vague about how to describe software.
-To address this, papers spoke of a need for a fixed identifier for software, either directly through a mechanism such as a Digital Object Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper written to document the software (and perhaps its creation), a ``software paper''~\cite{Chue_Hong_WSSSPE}.
-
-Another approach is to try to reduce the difficulty of citing software for authors, acknowledging that authors are often working with software that itself wraps other software and therefore hides that software.
-Knepley et al~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by which the software itself, after it has run, provides the user with a set of citations that are relevant to the code actually run.
-They describe a prototype implementation whereby the citations are embedded in libraries and reported along with the results, via a commandline interface~\cite{Knepley_WSSSPE}.
-Discussion highlighted the difficulty that attempting to acknowledge the contributions of all pieces of dependent code within a paper faces the difficulty of creating very long citation lists, straining the analogy of code used to papers cited.
-Katz approaches this issue by proposing a system of ``transitive credit,'' recording dependencies and relative contributions outside particular papers, relieving authors from the responsibility of acknowledging each and every dependency.
-Instead authors would acknowledge the percentage contribution of the software they used directly and an external system would then be able to recursively allocate that credit to those who had provided dependencies.
-Finally Priem and Piwowar argued that machine learning techniques could examine the body of published literature and extract mentions of software, coping with the multitude of informal ways in which authors mention software they have used~\cite{Priem_WSSSPE}.
-Discussion included turning the emphasis from asking users to improve their citation of software contributions, to ask how projects producing software might monitor the literature to improve their ability to show impact.
-Michael McLennan described the approach taken by the NanoHub project to scan the literature using keywords and names of known users to discover papers that are likely to have used their software and platform, then to have graduate students read each paper, highlighting any mention and perhaps following up with the authors to locate stories of impact.
-Work since the workshop has described this process at \textit{publications.wikia.com}.
-
-Acknowledgement in science does not come only in publications, of course.
-A key location for visibility might be in the grant funding process, both in bio-sketches and in grant project reporting.
-Some progress has been made here.
-Representatives of the NSF at the meeting emphasized these opportunities and encouraged participants to take advantage of them.
-Nonetheless, there was skepticism that peer review panels would value these contributions in the same ways as publications.
-
-Priem and Piwowar argued for additionally looking beyond publications and drawing on evidence of contribution, and impact, recorded in other online systems, such as source code repositories like Github, both code contributions, downloads of code and dependencies as well as conversations about software (mailing lists, twitter and beyond)~\cite{Priem_WSSSPE}.
-They argue for providing scholars with flexible resources so that they can tell their own stories in the manner most appropriate for them and their audiences, a principle of the ``altmetrics'' approach.
+The policy panel and discussion addressed the question of recognition
+of work on scientific software and linked that recognition to
+questions of reward and thus motivation for particular kinds of work
+on scientific software. In short, software work in science was seen to
+be inadequately visible in ways that ``count'' within the reputation
+system underlying science. In his paper for this workshop, Katz placed
+software work along with other ``activities that facilitate science
+but are not currently rewarded or
+recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
+to ``support all researchers in presenting meaningful impact evidence
+in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
+Kneply et al. argued that the visibility of software that supported a
+piece of science ``can have detrimental effects on funding, future
+library development, and even scientific
+careers.''~\cite{Knepley_WSSSPE}.
+
+These papers, and the discussion at the workshop, join a nascent
+literature seeking to understand what drives software work in science
+and how the reward systems of science thereby shape the type of
+software work undertaken, including the extent to which developers are
+motivated to build software for the long-term, for the use of others
+and whether to work collaboratively or
+separately~\cite{howison_incentives_2013, howison_scientific_2011,
+  bietz_synergizing_2010}. Software work is not only motivated by
+direct citations, but the visibility of software work in the
+literature is important to those who write software used in science.
+
+Papers and discussion concentrated in three areas: How ought software
+work to be visible, what are the barriers to its visibility, and what
+can be done to make it more visible?
+
+Most of the papers in this area focused on visibility of software in
+scientific papers, since scientific papers are the most widely
+accepted documentation of achievement in science. It was noted that
+there appear to be no widely accepted standards on how the use of
+software towards a paper ought to be mentioned, that journals,
+citation style guides and other guides to scientific conduct are vague
+about how to describe software. To address this, papers spoke of a
+need for a fixed identifier for software, either directly through a
+mechanism such as a Digital Object
+Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
+written to document the software (and perhaps its creation), a
+``software paper''~\cite{Chue_Hong_WSSSPE}.
+
+Another approach is to try to reduce the difficulty of citing software
+for authors, acknowledging that authors are often working with
+software that itself wraps other software and therefore hides that
+software. Knepley et al~\cite{Knepley_WSSSPE} approach this by
+proposing a mechanism by which the software itself, after it has run,
+provides the user with a set of citations that are relevant to the
+code actually run. They describe a prototype implementation whereby
+the citations are embedded in libraries and reported along with the
+results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion
+highlighted the difficulty that attempting to acknowledge the
+contributions of all pieces of dependent code within a paper faces the
+difficulty of creating very long citation lists, straining the analogy
+of code used to papers cited. Katz approaches this issue by proposing
+a system of ``transitive credit,'' recording dependencies and relative
+contributions outside particular papers, relieving authors from the
+responsibility of acknowledging each and every dependency. Instead
+authors would acknowledge the percentage contribution of the software
+they used directly and an external system would then be able to
+recursively allocate that credit to those who had provided
+dependencies. Finally Priem and Piwowar argued that machine learning
+techniques could examine the body of published literature and extract
+mentions of software, coping with the multitude of informal ways in
+which authors mention software they have used~\cite{Priem_WSSSPE}.
+Discussion included turning the emphasis from asking users to improve
+their citation of software contributions, to ask how projects
+producing software might monitor the literature to improve their
+ability to show impact. Michael McLennan described the approach taken
+by the NanoHub project to scan the literature using keywords and names
+of known users to discover papers that are likely to have used their
+software and platform, then to have graduate students read each paper,
+highlighting any mention and perhaps following up with the authors to
+locate stories of impact.  Work since the workshop has described this
+process at \textit{publications.wikia.com}.
+
+Acknowledgement in science does not come only in publications, of
+course. A key location for visibility might be in the grant funding
+process, both in bio-sketches and in grant project reporting. Some
+progress has been made here. Representatives of the NSF at the meeting
+emphasized these opportunities and encouraged participants to take
+advantage of them. Nonetheless, there was skepticism that peer review
+panels would value these contributions in the same ways as
+publications.
+
+Priem and Piwowar argued for additionally looking beyond publications
+and drawing on evidence of contribution, and impact, recorded in other
+online systems, such as source code repositories like Github, both
+code contributions, downloads of code and dependencies as well as
+conversations about software (mailing lists, twitter and
+beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with
+flexible resources so that they can tell their own stories in the
+manner most appropriate for them and their audiences, a principle of
+the ``altmetrics'' approach.
 
@@
 
-Another aspect of sustainable software that is rarely given due
-diligence in scientific software design is software security. Heiland
-et al~\cite{Heiland_WSSSPE} discuss how the security measures
-appropriate for scientific software strongly depend on the maturity
-level of the software. However, the maturity level of a software
-project meant to be sustained changes tremendously over its
+Many facets of engineering for software sustainability strongly depend
+on the maturity level of the software. However, the maturity level of
+a software project meant to be sustained changes tremendously over its
 development life cycle, and the eventual maturity level is often
-difficult to predict during initial development. Best practice models
-for identifying and defining cybersecurity vulnerabilities at
-different stages of the software life cycle have been formalized as
-Software Security Maturity Models. Although these are widely used in
-industry, awareness among scientific software development communities
-is low. Developing such models for scientific software would align
-well with the objective of providing implementation approaches that
-the scientific community can actually follow: A Software Security
-Maturity Model provides a path to tightening security practices as
-software matures, emphasizes understandability over complexity, and
-classifies cybersecurity-pertinent practices for easy recognition of
-what level of cybersecurity a software addresses.
+difficult to predict during initial development. Using software
+security as their case sudy, Heiland et al~\cite{Heiland_WSSSPE}
+discuss how Maturity Models can be used to formalize best practices
+appropriate for the different levels of maturity that a software
+project may go through over its lifecycle. Cybersecurity is also an
+example of a sustainability-relevant aspect in software design that is
+rarely given due diligence in science. In particular in industry,
+cybersecurity best practices for different stages of life cycle and
+maturity have been formalized as Software Security Maturity Models,
+and are widely used, yet awareness of these among scientific software
+development communities remains low. In providing a path to tightening
+security practices as software matures, such models align with the
+objective of providing implementation approaches that the scientific
+community can actually follow.
 

### 2014-04-05 WSSSPE1_report/WSSSPE1_summary_report.tex
Includes rOpenSci example.

@@
 the rule, and particularly in the long tail of science the
-scientist-developer predominates. For providing these with better
-training in software engineering best practices, initiatives such as
-Software Carpentry~\cite{SoftwareCarpentry} have been especially
-valuable, and Software Carpentry also shown how such training could
-scale.
+scientist-developer predominates, even for software that is widely
+used or crosses domain boundaries such as
+rOpenSci~\cite{rOpenSci}. For better training domain scientists at
+least in basic software engineering best practices, initiatives such
+as Software Carpentry~\cite{SoftwareCarpentry} have demonstrated how
+this could be achieved at scale.
 

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Removing subsections from software section
Looking at integrating several subsections as themes, but the
intent for these sections was to be integrated into the text.

@@
 path and either adapt or seek out alternate career paths?
-
-\subsection{Research or Reuse?}
-
-Discussion of software produced as a by-product versus software
-developed for reuse. How does this change the project, who develops
-the code, growth beyond 1--2 person projects to larger projects with
-diverse set of consumers and contributors. Modular and extensible code
-versus working for current research problem---tackled by same people,
-or a collaborative team?
-
-\subsection{Defining Sustainability for Scientific Software}
-
-What is sustainable? What are the best practices, can workshops fund
-meetings to help define and improve best practices in these areas?
-Software plans from the funding agencies? This topic is expanded upon
-in \S\ref{sec:cross-cutting}, since it was discussed up in multiple
-parts of the WSSSPE1 workshop.
-
-\subsection{Training Scientists to Develop and Support Software}
-
-Discussion of the evolving role of scientists, and/or others that fill
-these roles.
-
-\subsection{Software Process, Code Review, Automation, Reproducibility}
-
-There are a lot of tools out there, but few are currently used. Look
-at what some projects have found successful, and how to automate as
-much as possible to reduce additional overhead.
-
-\subsection{Software and Data Licensing}
-
-Its impact on how and where research products are used, who can
-collaborate and what patterns have worked/not worked in existing
-communities.
-
-\subsection{Funding, Sustainability Beyond the First Grant/Institution}
-
-How initial software development is funded, moving to follow up
-projects, maintenance, community growth, collaborating with other
-institutions, labs, industry, internationally. Contract relationship
-versus collaborative development between scientists and software
-developers.
-
-\subsection{Training Others to use Software}
-
-Who creates training materials, how are they distributed, integration
-into courses when projects see very wide application in research.
 

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Adding back in some of the subsections to software
Some reordering of text and discussion, fitting in the main text
with the overall themes of the discussion in order to make it easier
to read through the section.

@@
 that developing and maintaining software is hard, but best practices can help.
-Several participants stated that documentation is not just for users, and paying attention
-to API documentation, tutorials for building and deploying software, along with
+Several participants added that documentation is not just for users, and writing
+API documentation, tutorials for building and deploying software, along with
 documented development practices can be very helpful in bringing  new developers
 into a project efficiently.
+
+\subsection{Research or Reuse?}
 
@@
 
+There were 13 articles about different experiences in this area, but little
+about GUI testing, performance, scalability, or agile development practices.
+There were several unique perspectives about issues such as managing API
+changes, using the same best practices for software as data, and going beyond
+simply ``slapping an OSI-approved license on code.''
+
+\subsection{Communities and Career Paths}
+
 Communities are extremely important in software projects, and both their
@@
 downloads, continuous integration, software quality dashboards, and of course,
-a general web presence to tie all of these things together. Questions were posed
-around users not answering the questions of other users. Several participants
+a general web presence to tie all of these things together.
+
+There was extended discussion about users not answering the questions of other
+users. Several participants
 offered counterexamples, such as mailing lists where developers do not participate
@@
 
-There were 13 articles about different experiences in this area, but little
-about GUI testing, performance, scalability, or agile development practices.
-There were several unique perspectives about issues such as managing API
-changes, using the same best practices for software as data, and going beyond
-simply ``slapping an OSI-approved license on code.''
+\subsection{Defining Sustainability for Scientific Software}
 
@@
 
-What about actual software engineering principles, such as modularity and
-extensibility? This is how industry maintains software, and ensures it continues
-to be useful. Often, rewriting software is considered to be too costly, but
-with a modular design it can be kept up to date. Extensibility is expected to keep
-it relevant, if built into the project. One counterpoint raised by Jason Riedy
-was that trying to take advantage of the latest and greatest hardware often
-makes this painful, hence the lack of papers mentioning ``GPUs and exotic
-hardware.''
-
-The question of whether funders, such as the NSF, can mandate software plans in
-much the same way as they do data management plans, was raised. Daniel Katz responded that
-software is supposed to be described as part of the NSF data management plan,
-and that in NSF's definition, data includes software. A comment from Twitter
-(@biomickwatson) raised the issue that this requires reviewers and funders who
-understand the answers that are given in these plans. Daniel Katz responded
-that in programs focused on software or data this can be done effectively, but
-agreed that in more general programs this is indeed a problem.
+\subsection{Software Process, Code Review, Automation, Reproducibility}
 
@@
 
+What about actual software engineering principles, such as modularity and
+extensibility? This is how industry maintains software, and ensures it continues
+to be useful. Often, rewriting software is considered to be too costly, but
+with a modular design it can be kept up to date. Extensibility is expected to keep
+it relevant, if built into the project. One counterpoint raised by Jason Riedy
+was that trying to take advantage of the latest and greatest hardware often
+makes this painful, hence the lack of papers mentioning ``GPUs and exotic
+hardware.''
+
+The question of whether funders, such as the NSF, can mandate software plans in
+much the same way as they do data management plans, was raised. Daniel Katz responded that
+software is supposed to be described as part of the NSF data management plan,
+and that in NSF's definition, data includes software. A comment from Twitter
+(@biomickwatson) raised the issue that this requires reviewers and funders who
+understand the answers that are given in these plans. Daniel Katz responded
+that in programs focused on software or data this can be done effectively, but
+agreed that in more general programs this is indeed a problem.
+
+\subsection{Training Scientists to Develop and Support Software}
+
 Some of the questions raised after the panel discussion are:
@@
 classes as in ``real sciences''?
+
+\subsection{Funding, Sustainability Beyond the First Grant/Institution}
 

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Added some introduction to backward compatibility

@@
 \subsection{Research or Reuse?}
+
+Software is developed for many different purposes, and the requirements can
+vary significantly depending on the intended audience. Most end-users make use
+of either a graphical user interface of some kind, or a command line that may
+offer input and output formats for running the code and analyzing its output(s).
+When discussing backward compatibility it is these various interfaces that are
+discussed. For software that builds on other software frameworks it is the
+API (application programming interface) that is most important, and this can
+encompass issues such as the source and binary interfaces to the software
+libraries developed---with each potentially having a high maintenance cost if
+they are to remain compatible over many years. When using command line programs
+it is generally the command line switches as well as the input/output formats
+that could incur significant costs if they are changed.
 

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Various edits in Policy.
Includes adding one aspect from discussion documented in Google Doc. The one
other aspect that is missing is "Is sustainability designed for or an emergent
property". However, I think this best goes into the "What does sustainability
mean" cross-cutting issue (if it isn't there already).

@@
 
-Frameworks provide a scaffold for thinking; they help us understand
-which elements are important and how they relate. They also help us
-break down a problem into chunks that can be pursued by different
-groups or perhaps disciplines.
-
-A group of papers submitted to the workshop provide frameworks with
-these intentions. The frameworks have substantial overlap but with
-different emphases and extents.  Accordingly we present them starting
-with the area of most overlap and proceed to show the different
-emphases and the range of factors included in the frameworks.
-
-% \todo{seems like definitions will be mentioned elsewhere in the
-% paper?}
-Each of the framework papers included a definition of sustainability,
+The workshop submissions grouped under this section provide frameworks
+for thinking about, researching and understanding which elements of sustainability
+are important and how they relate. Although there is substantial
+overlap between the frameworks, they have different emphases and
+extents.  Each paper in this group included a definition of sustainability,
 with many overlaps.  In essence the papers related the question to
@@
 idea was also expressed as software that continues to serve its
-users~\cite{Pierce_WSSSPE}.  See \S\ref{sec:cross-cutting} for more on
-the discussion of definitions of sustainability in the workshop.
+users~\cite{Pierce_WSSSPE}.  A more thorough discussion of definitions
+of software sustainability is presented in \S\ref{sec:cross-cutting}.
 
@@
 overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned this did so without
+The two papers in this area which mentioned this\todo{which are these
+  papers? Should cite here.} did so without
 integrating that analysis into the question of software being around
@@
 that can be applied to pieces of software but might also extend to
-describe software projects.  Thus Calero and
-colleagues~\cite{Calero_WSSSPE} propose adding elements to the ISO
-standards for measuring software quality: proposing an additional
+describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE} propose adding elements to the ISO
+standards for measuring software quality. Specifically, they propose an additional
 dimension of quality they call perdurability or ability to ``last over
@@
 could be applied to a software project). For example Pierce et
-al~\cite{Pierce_WSSSPE} focus on the way that the project is run,
+al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
 particularly in terms of how those involved communicate and jointly
@@
 
-Frameworks also addressed themselves to the context in which software
-projects exist (thus moving upwards in abstraction, from the artifact
-itself, to the organization of its production and, in these
-frameworks, the shape of the space which an artifact or project exists
-in). These frameworks take the form of contingency theories in that
+Frameworks also concerned themselves with the context in which software
+projects exist, moving in abstraction from the software artifact
+itself to the organization of its production and
+the shape of the space in which an artifact or project exists.
+These frameworks take the form of contingency theories in that
 they outline a different set of challenges and argue that different
@@
 research to understand those connections. Weber et
-al~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
+al.~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
 ecosystems as ``niches'' which sustain particular pieces of
@@
 
-The policy panel and discussion addressed the question of recognition
-of work on scientific software and linked that recognition to
-questions of reward and thus motivation for particular kinds of work
-on scientific software. In short, software work in science was seen to
-be inadequately visible in ways that ``count'' within the reputation
-system underlying science. In his paper for this workshop, Katz placed
+How work on scientific software is recognized and rewarded strongly influences the motivation for particular kinds of work
+on scientific software. A recurring theme of the panel discussion was
+that software work in science is
+inadequately visible within the reputation
+system underlying science; in other words it often doesn't ``count''. In his paper for this workshop, Katz placed
 software work along with other ``activities that facilitate science
@@
 
-Papers and discussion concentrated in three areas: How ought software
+Papers and discussion concentrated on three overarching questions: How ought software
 work to be visible, what are the barriers to its visibility, and what
@@
 there appear to be no widely accepted standards on how the use of
-software towards a paper ought to be mentioned, that journals,
+software towards a paper ought to be mentioned, and that journals,
 citation style guides and other guides to scientific conduct are vague
-about how to describe software. To address this, papers spoke of a
+about how to describe software. To address this, papers advocated the
 need for a fixed identifier for software, either directly through a
@@
 written to document the software (and perhaps its creation), a
-``software paper''~\cite{Chue_Hong_WSSSPE}.
-
-Another approach is to try to reduce the difficulty of citing software
-for authors, acknowledging that authors are often working with
-software that itself wraps other software and therefore hides that
-software. Knepley et al~\cite{Knepley_WSSSPE} approach this by
+``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
+out during the panel discussion, one of the problems with papers as
+the cited product is that their author list is fixed in time, which
+discourages potential contributors who are not on the original author
+list from designing incremental improvements as integration work
+rather than separate (and hence possibly rewritten) software
+products~\cite{howison_incentives_2013}.
+
+Another approach is to reduce the difficulty of citing all software
+underlying a research paper. For example, scientists often work with
+software that itself wraps other software, leading to attribution
+stacking that can make it non-obvious or even difficult to determine
+what attribitions would be appropriate. Knepley et al.~\cite{Knepley_WSSSPE} approach this by
 proposing a mechanism by which the software itself, after it has run,
-provides the user with a set of citations that are relevant to the
-code actually run. They describe a prototype implementation whereby
+provides the user with a set of citations, according to the pieces of
+code actually executed. They describe a prototype implementation whereby
 the citations are embedded in libraries and reported along with the
@@
 mentions of software, coping with the multitude of informal ways in
-which authors mention software they have used~\cite{Priem_WSSSPE}.
-Discussion included turning the emphasis from asking users to improve
-their citation of software contributions, to ask how projects
-producing software might monitor the literature to improve their
-ability to show impact. Michael McLennan described the approach taken
-by the NanoHub project to scan the literature using keywords and names
+which authors mention software they used~\cite{Priem_WSSSPE}.
+A point raised in the panel discussion was that instead of asking users to improve
+their software citation practices, one can also ask how software
+projects can better monitor the literature to improve their
+ability to show impact. For example, the NanoHub project scans the literature using keywords and names
 of known users to discover papers that are likely to have used their
-software and platform, then to have graduate students read each paper,
-highlighting any mention and perhaps following up with the authors to
-locate stories of impact.  Work since the workshop has described this
-process at \textit{publications.wikia.com}.
-
-Acknowledgement in science does not come only in publications, of
-course. A key location for visibility might be in the grant funding
-process, both in bio-sketches and in grant project reporting. Some
-progress has been made here. Representatives of the NSF at the meeting
-emphasized these opportunities and encouraged participants to take
-advantage of them. Nonetheless, there was skepticism that peer review
-panels would value these contributions in the same ways as
-publications.
-
-Priem and Piwowar argued for additionally looking beyond publications
-and drawing on evidence of contribution, and impact, recorded in other
-online systems, such as source code repositories like Github, both
-code contributions, downloads of code and dependencies as well as
-conversations about software (mailing lists, twitter and
-beyond)~\cite{Priem_WSSSPE}. They argue for providing scholars with
-flexible resources so that they can tell their own stories in the
-manner most appropriate for them and their audiences, a principle of
-the ``altmetrics'' approach.
-
-\todo{convert this sketch to discussion of policy options}
-\begin{verbatim}
-        Opportunities for Policy interventions:
-                Funding agencies
-                Journals.  Success of data policies, growth of software policies.
-                Promotion and Tenure. Follow lead of Science of Team Science in surveying policies at universities and other institutes.
-\end{verbatim}
-\todo{review PiratePad notes for other aspects of credit-giving discussed.}
+software and platform, and assigns graduate students to read each paper,
+highlighting mentions of software use and sometimes following up with the authors to
+identify stories for demonstrating impact.  A process for tracking
+software-using publications with the goal of increasing impact visibility is now described at \url{http://publications.wikia.com}.
+
+Potential visibility, and thus acknowledgement of scientific software
+products is not restricted to publications. Another key location for
+visibility is in the grant funding
+process, and as emphasized by NSF representatives at the meeting,
+recent changes to grant proposal and reporting formats now allow both
+applicants and awardees to report and highlight software products as
+much as publications. Nonetheless, whether peer review
+panels would value these contributions in the same way as
+publications remains to be seen.
+
+Priem and Piwowar argued that assessing the impact of software work
+requires looking beyond publications, including evidence of contribution and impact recorded in social
+coding-oriented resources such as Github, and conversations about software
+in issue trackers, mailing lists, twitter and
+beyond~\cite{Priem_WSSSPE}. In keeping with a principle of
+the ``altmetrics'' approach, they advocate that scholars should have
+resources that empower them to tell their own stories in the
+manner most appropriate for them and their audiences.
 
@@
 are often addressed only late in the scientific software development
-cycle, if at all: Krintz et al~\cite{Krintz_WSSSPE} look at API
-governance, and Heiland et al~\cite{Heiland_WSSSPE} discuss maturity
+cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
+governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
 models for software security. The other two papers discuss
@@
 
-Krintz et al~\cite{Krintz_WSSSPE} describe how in an era in which
+Krintz et al.~\cite{Krintz_WSSSPE} describe how in an era in which
 computing frequently takes place in a distributed cloud, the control
@@
 infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
-et al observe, unlike for physical IT infrastructure in data centers,
+et al. observe, unlike for physical IT infrastructure in data centers,
 science communities have developed very little in the way of practices and
-technology for API governance, refered to by Krintz et al as the
+technology for API governance, refered to by Krintz et al. as the
 ``combined policy, implementation, and deployment control''. Web APIs
@@
 organizations to assess and thus properly manage the impact of API
-changes. To address this, Krintz et al propose a mechanism that
+changes. To address this, Krintz et al. propose a mechanism that
 evaluates the porting effort between two versions of a web-service API
@@
 difficult to predict during initial development. Using software
-security as their case sudy, Heiland et al~\cite{Heiland_WSSSPE}
+security as their case sudy, Heiland et al.~\cite{Heiland_WSSSPE}
 discuss how Maturity Models can be used to formalize best practices
@@
 capabilities so they can be shared.
-
-% note that paper summaries are available in tag jameshowison-paper-summaries: 
-% https://github.com/danielskatz/WSSSPE/blob/jameshowison-paper-summaries/WSSSPE1_report/WSSSPE1_summary_report.tex
 

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Moving Career Track subsection under Cross-cutting issues.
Also adjusts section level, label, and refs for/to Defining Sustainability.

@@
 sustainability for software ``does not just mean more money from
-Government'' (see also Section~\ref{sec:cross-cutting}).  Other
+Government'' (see also Section~\ref{sec:defining-sustainability}).  Other
 factors to consider, he mentioned, include costs of production, ease
@@
 The question of what ``sustainable'' means in the context of software was
-raised (see \S\ref{sec:cross-cutting}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
+raised (see \S\ref{sec:defining-sustainability}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
 to look for when choosing software for procurement, or reuse in further software
@@
 users~\cite{Pierce_WSSSPE}.  A more thorough discussion of definitions
-of software sustainability is presented in \S\ref{sec:cross-cutting}.
+of software sustainability is presented in \S\ref{sec:defining-sustainability}.
 
@@
 
-\subsection{Career Tracks for Scientific Software Developers}
-
-How to ensure software is sustainable by ensuring there are career
-paths for developers.
-
-What are the possible career paths for a specialist software developer working as part of a scientific research group?
-
 \section{Communities \note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}} \label{sec:community}
@@
 
-\section{Cross-cutting Issue: Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:cross-cutting}
+\section{Cross-cutting Issues} \label{sec:cross-cutting}
+
+\subsection{Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:defining-sustainability}
 
@@
 
+\subsection{Career Tracks for Scientific Software Developers} \label{sec:career-tracks}
+
+How to ensure software is sustainable by ensuring there are career
+paths for developers.
+
+What are the possible career paths for a specialist software developer working as part of a scientific research group?
+
 \section{Case Studies \note{lead: Ketan Maheshwari, contributors: Andy Terrel}} \label{sec:use-cases}

### 2014-04-06 WSSSPE1_report/WSSSPE1_summary_report.tex
Introductory paragraph for Policy.

@@
 
-\todo{some text is needed here before the first subsection}
+The Policy panel was designated to discuss the group of workshop
+contributions dealing with the wide range of software sustainability
+aspects that relate to establishing, promoting, and implementing
+policies. Six papers presented frameworks for defining, modeling, and
+evaluating software sustainability, the basis of establishing
+policies. Four papers advocated mechanisms for more properly assessing
+the impact of scientific software, and for crediting and recognizing
+work that promotes software sustainability, all of which are
+instrumental in effectively promoting policies that aim to change
+current practices. Four papers discuss facets of implementing software
+sustainability, and models of implementation across different facets.
 
@@
 idea was also expressed as software that continues to serve its
-users~\cite{Pierce_WSSSPE}.  A more thorough discussion of definitions
-of software sustainability is presented in \S\ref{sec:defining-sustainability}.
+users~\cite{Pierce_WSSSPE}.  For a more thorough discussion of definitions
+of software sustainability see \S\ref{sec:defining-sustainability}.
 

### 2014-04-07 WSSSPE1_report/WSSSPE1_summary_report.tex
Various small edits across document.
Corrects double quotes, spelling errors, and makes consistent the formatting
of tool/project names.

@@
 \usepackage[all]{hypcap}
+% Don't use tt font for urls
 \urlstyle{rm}
@@
 
-% Don't use tt font for urls
-\urlstyle{rm}
+% formatting commands
+%   - consistent formatting of names of software tools and/or projects:
+\newcommand{\toolname}[1] {\textsf{#1}}
 
@@
 58 submissions were received, and almost all submitters used either
-arXiv~\cite{arXiv-web} or figshare~\cite{figshare-web} to self-publish
+arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to self-publish
 their papers.
@@
 stacking that can make it non-obvious or even difficult to determine
-what attribitions would be appropriate. Knepley et al.~\cite{Knepley_WSSSPE} approach this by
+what attributions would be appropriate. Knepley et al.~\cite{Knepley_WSSSPE} approach this by
 proposing a mechanism by which the software itself, after it has run,
@@
 the citations are embedded in libraries and reported along with the
-results, via a commandline interface~\cite{Knepley_WSSSPE}. Discussion
+results, via a command-line interface~\cite{Knepley_WSSSPE}. Discussion
 highlighted the difficulty that attempting to acknowledge the
@@
 design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
-large projects that have software infrastructue development built-in,
+large projects that have software infrastructure development built-in,
 with cross-training domain scientist PIs in software engineering best
@@
 science communities have developed very little in the way of practices and
-technology for API governance, refered to by Krintz et al. as the
+technology for API governance, referred to by Krintz et al. as the
 ``combined policy, implementation, and deployment control''. Web APIs
@@
 difficult to predict during initial development. Using software
-security as their case sudy, Heiland et al.~\cite{Heiland_WSSSPE}
+security as their case study, Heiland et al.~\cite{Heiland_WSSSPE}
 discuss how Maturity Models can be used to formalize best practices
@@
 example, in the life sciences the iPlant Collaborative~\cite{iPlant},
-the Galaxy Project~\cite{Galaxy}, and Qiime~\cite{Qiime} are large
+the Galaxy Project~\cite{Galaxy}, and \toolname{Qiime}~\cite{Qiime} are large
 multi-year projects with significant software infrastructure funding
@@
 used or crosses domain boundaries such as
-rOpenSci~\cite{rOpenSci}. For better training domain scientists at
+\toolname{rOpenSci}~\cite{rOpenSci}. For better training domain scientists at
 least in basic software engineering best practices, initiatives such
@@
 The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
-catalysts" and their role in the modern scientific research process. A technology
+catalysts'' and their role in the modern scientific research process. A technology
 catalyst can be defined as a role played by an individual with a know-how of
@@
 explosion of use of just a few gateways. Over 3.5 years, more than 7,000
-biologists have run phylogenetic codes on supercomputers using the CIPRES
+biologists have run phylogenetic codes on supercomputers using the\toolname{CIPRES}
 Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
 UltraScan Science Gateway, increasing the sophistication of analytical
-ultracentrifugation experiments worldwide. The new Neuroscience Gateway (NSG)
+ultracentrifugation experiments worldwide. The new Neuroscience Gateway (\toolname{NSG})
 registered 100+ users who used 250,000 CPU hours in only a few months.
@@
 Gateways, however, need to keep operational costs low and can often make use of
-common components. Science Gateway Platform as a Service (Sci-GaP) delivers
+common components. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
 middleware as a hosted, scalable third-party service while domain developers
@@
 
-One key to Sci-GaP is its openness. While based on the Apache Airavata project
-and the CIPRES Workbench Framework, community contributions are encouraged
+One key to \toolname{SciGaP} is its openness. While based on the \toolname{Apache Airavata} project
+and the \toolname{CIPRES} Workbench Framework, community contributions are encouraged
 because of its open source, open governance and open operations policies. The
@@
 improves reliability and prioritizes stakeholder requirements. The project is
-leveraging Internet2's Net+ mechanisms for converting SciGaP and its gateways
+leveraging Internet2's \toolname{Net+} mechanisms for converting \toolname{SciGaP} and its gateways
 into commodity services.
@@
 an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
-both the for-profit company Kitware and the users and contributors to their
-software packages such as VTK.  This structure, of the centralized development
+both the for-profit company \toolname{Kitware} and the users and contributors to their
+software packages such as \toolname{VTK}.  This structure, of the centralized development
 of core infrastructure around which communities of individuals applying that
@@
 is supported by a non-profit entity that fosters community investment in that
-ecosystem.  In many ways, these two organizations (Kitware and NumFOCUS)
+ecosystem.  In many ways, these two organizations (\toolname{Kitware} and \toolname{NumFOCUS})
 attempt to cross domain boundaries and provide support for both the
@@
 % 3. Can we characterize the process itself for the selected cases:
-% collaborative, seperate.
+% collaborative, separate.
 
@@
 % of all the software products discussed in the workshop. (If possible color
-% code the representaion according to a scheme based on the different
+% code the representation according to a scheme based on the different
 % dimensions pertaining to them.)
@@
 treated separately. For instance, the challenges faced by a gateway software
-development project such as \textit{CIPRES} or a visualization software product
-such as \textit{VisIT} are distinct to a niche software for \textit{ab initio}
-modeling and simulation such as \textit{VASP} or \textit{Quantum Espresso}.
+development project such as \toolname{CIPRES} or a visualization software product
+such as \toolname{VisIT} are distinct to a niche software for \textit{ab initio}
+modeling and simulation such as \toolname{VASP} or \toolname{Quantum Espresso}.
 
@@
 fundamental research problems for a given scientific domain. Examples are
-collaborative development frameworks such as \emph{GitHub} and
-\emph{Bitbucket}, distributed workflow and generic computing frameworks
-such as \emph{Galaxy}, \emph{nanoHUB}, \emph{SimGrid}, \emph{Swift},
-\emph{Globus} and \emph{VisTrails}, and visualization frameworks such as
-\emph{VisIT}.
-
-% What are the sustainability challenges? What are the oppurtunities?
+collaborative development frameworks such as \toolname{GitHub} and
+\toolname{Bitbucket}, distributed workflow and generic computing frameworks
+such as \toolname{Galaxy}, \toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
+\toolname{Globus} and \toolname{VisTrails}, and visualization frameworks such as
+\toolname{VisIT}.
+
+% What are the sustainability challenges? What are the opportunities?
 Development is often a high risk/reward undertaking exploring uncharted
@@
 
-On the other hand, utility software presents oppurtunities to be usable by a
+On the other hand, utility software presents opportunities to be usable by a
 larger community making its undertaking and development an attractive pursuit.
@@
 
-One such example is the \emph{Galaxy} project. It follows agile software
+One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It follows agile software
 development practices and implements standards such as test-driven development
-and socialized bug managing practices via \emph{trello}. Galaxy
+and socialized bug managing practices via \toolname{trello}. Galaxy
 \emph{histories} and \emph{toolshed} offers easy community sharing of data and
 tools further promoting a collaborative environment. The project very closely
-follows the guidelines described in~\cite{Carver_WSSSPE} and many
-from~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
+follows the guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE} and many
+from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
 aiming better utilizing a particular, new infrastructure and architecture,
-e.g., \emph{MVAPICH}, \emph{VisIT}. Similarly, to leverage the power of
+e.g., \toolname{MVAPICH}, \toolname{VisIT}. Similarly, to leverage the power of
 emerging architectures such as accelerators, new code and libraries are
 required. The experience of one such effort as described
-in~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
+in Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
 with many invaluable lessons were learned about influential cultural and
@@
 computational environments, many users are leaning towards tools used as
-services. \emph{GitHub} and \emph{Bitbucket} can be argued to be
+services. \toolname{GitHub} and \toolname{Bitbucket} can be argued to be
 such tools, catering to collaborative development. For scientific users
-\emph{Globus}-based tools are a case of service-based utility software
+\toolname{Globus}-based tools are a case of service-based utility software
 discussed during the workshop. The data movement and sharing services offered
@@
 challenge and face radically different issues compared to utility software.
-Many submissions reported that software is practically considered a ``byproduct"
+Many submissions reported that software is practically considered a ``byproduct''
 of the actual research. Others contended that the software was not the main
@@
 Some examples of such software discussed as part of the workshop are,
-\emph{DUNE}, \emph{R/qtl}, \emph{Kitware}, \emph{PETSc}, \emph{MINRES-QLP},
-\emph{FASTMath}.  Most of the above efforts are focused on one scientific
+\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware}, \toolname{PETSc}, \toolname{MINRES-QLP},
+\toolname{FASTMath}.  Most of the above efforts are focused on one scientific
 and/or applied mathematics domain. However, sometimes such projects grow beyond
-the initial vision of developers. One such example is \emph{Kitware}, which
+the initial vision of developers. One such example is \toolname{Kitware}, which
 while being a software product specializing in the scientific process, has a
 core focus of developing communities around software processes. An instance of
-this process is the development of the \emph{CMake} build utility, which
-started out as a building tool for \emph{ITK} but grew to become a generic
-build utility for C++ projects. Similarly, \emph{PETSc} is growing towards
+this process is the development of the \toolname{CMake} build utility, which
+started out as a building tool for \toolname{ITK} but grew to become a generic
+build utility for C++ projects. Similarly, \toolname{PETSc} is growing towards
 becoming a general purpose utility system usable for solving variety of
 scientific problems.
-%
 
@@
 original developers. We included projects older than 5 years and excluded
-relatively new projects such as \emph{SciPy} and \emph{numFOCUS} from this
+relatively new projects such as \toolname{SciPy} and \toolname{numFOCUS} from this
 survey. we find that in a majority of scientific software more than one
 original developers are still around, in some cases for more than 10 years,
-(eg. \emph{R/qtl}).  Whereas, in the case of utility software projects, it
+(eg. \toolname{R/qtl}).  Whereas, in the case of utility software projects, it
 appears that at least one original developer is still around, often in a
-part-time or ``benefactor" role. In other cases, it appears that the original
+part-time or ``benefactor'' role. In other cases, it appears that the original
 developers have went on to assume a larger role, with the same project.
@@
 is achieved by the fact that the core developer or team heavily utilizes the
-software for their science, e.g.,  \emph{R/qtl, PETSc}.  Furthermore, the
+software for their science, e.g.,  \toolname{R/qtl, PETSc}.  Furthermore, the
 development of scientific software requires more scientific skills compared

### 2014-04-07 WSSSPE1_report/WSSSPE1_summary_report.tex
small edit

@@
 surrounding people, especially the career paths of original developers and
-investigate to answer the question: are the original developers are still
-around?
+investigate to answer the question: \emph{are the original developers still
+around?}
 

### 2014-04-08 WSSSPE1_report/WSSSPE1_summary_report.tex
Changes to mainly first page. Minor changes to section titles.

@@
 \usepackage{multicol}
+\usepackage{verbatim}
 
@@
 
-\author{Authors TBD}
-
-%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
-%Manish Parashar$^{(4)}$, and David Proctor$^{(1)}$ }
+\author{Bruce Berriman$^{(1)}$, Sou-Cheng T.
+Choi$^{(2)}$, Marcus D.
+Hanwell$^{(3)}$, James Hetherington$^{(4)}$, James Howison$^{(5)}$, Daniel S.
+Katz$^{(6)}$, Hilmar Lapp$^{(7)}$, Frank L\"{o}ffler$^{(8)}$, Ketan Maheshwari$^{(9)}$, Shel Swenson$^{(10)}$,  Andy R.
+Terrel$^{(11)}$, Matthew Turk$^{(12)}$,  Colin Venters$^{(13)}$, Nancy Wilkins-Diehr$^{(14)}$}
+\note{Authors, please add/check middle initials and affliations.  For now, the order is alphabetical for now  because Choi does not know how to count...}
+
+\thanks{{}$^{(1)}$ xxx}
 %
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA (\url{sctchoi@uchicago.edu}}
 % 
-%\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA}
+\thanks{{}$^{(3)}$ xxx}
 %
-%\thanks{{}$^{(2)}$ Skolkovo Institute of Science and Technology, Moscow, Russian Federation}
+\thanks{{}$^{(4)}$ xxx}
 %
-%\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+\thanks{{}$^{(5)}$ xxx}
 %
-%\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
-
-\begin{abstract}
+\thanks{{}$^{(6)}$ National Science Foundation, Arlington, VA, USA}
+%
+\thanks{{}$^{(7)}$ xxx}
+%
+\thanks{{}$^{(8)}$ xxx}
+%
+\thanks{{}$^{(9)}$ xxx}
+%
+\thanks{{}$^{(10)}$ xxx}
+%
+\thanks{{}$^{(11)}$ xxx}
+%
+\thanks{{}$^{(12)}$ xxx}
+%
+\thanks{{}$^{(13)}$ xxx}
+%
+\thanks{{}$^{(14)}$ xxx}
+
+\begin{abstract} \note{Lead: Sou-Cheng T. Choi}
 This document discusses the First Workshop on Sustainable
@@
 
+\begin{comment}
 People who have volunteered to work on this report: (with * for actual
@@
 \item * James Hetherington $\langle$\url{j.hetherington@ucl.ac.uk}$\rangle$
-\item * Sou-Cheng (Terrya) Choi $\langle$\url{sou.cheng.terrya.choi@gmail.com}$\rangle$
+\item * Sou-Cheng (Terrya) Choi $\langle$\url{sctchoi@uchicago.edu}$\rangle$
 \item * James Howison $\langle$\url{james@howison.name}$\rangle$
@@
 comments in addition to text.
+\end{comment}
 
@@
 
-\section{Keynotes \note{lead: Sou-Cheng (Terrya) Choi} } \label{sec:keynotes}
+\section{Keynotes \note{lead: Sou-Cheng (Terrya) Choi}} \label{sec:keynotes}
 
@@
 
-\section{Developing and Supporting Software \note{lead: Marcus Hanwell, contributors: Suresh Marru}} \label{sec:devel}
+\section{Developing and Supporting Software \note{\scriptsize lead: Marcus Hanwell, contributor: Suresh Marru}} \label{sec:devel}
 
@@
 
-\section{Communities \note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}} \label{sec:community}
+\section{Communities \note{\vspace{-1ex}\tiny lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}} \label{sec:community}
 
@@
 descriptive and prescriptive approaches to communities.
+\note{Choi: Do we need to itemize the reference numbers below?
+Please collapse them into a comma seperated list with a short description for all or each of the papers.
+}
 

### 2014-04-08 WSSSPE1_report/WSSSPE1_summary_report.tex
Fix typo in note.
ordering authors, other misc things

@@
 
-\author{Bruce Berriman$^{(1)}$, Sou-Cheng T.
-Choi$^{(2)}$, Marcus D.
-Hanwell$^{(3)}$, James Hetherington$^{(4)}$, James Howison$^{(5)}$, Daniel S.
-Katz$^{(6)}$, Hilmar Lapp$^{(7)}$, Frank L\"{o}ffler$^{(8)}$, Ketan Maheshwari$^{(9)}$, Shel Swenson$^{(10)}$,  Andy R.
-Terrel$^{(11)}$, Matthew Turk$^{(12)}$,  Colin Venters$^{(13)}$, Nancy Wilkins-Diehr$^{(14)}$}
-\note{Authors, please add/check middle initials and affliations.  For now, the order is alphabetical for now  because Choi does not know how to count...}
-
-\thanks{{}$^{(1)}$ xxx}
+\author{Daniel S.
+Katz$^{(1)}$, Hilmar Lapp$^{(2)}$, Ketan Maheshwari$^{(3)}$, Frank L\"{o}ffler$^{(4)}$, Matthew Turk$^{(5)}$, Marcus D.
+Hanwell$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$, James Hetherington$^{(9)}$, James Howison$^{(9)}$, Sou-Cheng T.
+Choi$^{(10)}$, Shel Swenson$^{(11)}$,  Bruce Berriman$^{(12)}$, Andy R.
+Terrel$^{(13)}$, Colin Venters$^{(14)}$}
+\note{Authors, please add/check middle initials and affiliations.}
+
+\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{d.katz@ieee.org}}
 %
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA (\url{sctchoi@uchicago.edu}}
+\thanks{{}$^{(2)}$ xxx}
 %
@@
 %
-\thanks{{}$^{(6)}$ National Science Foundation, Arlington, VA, USA}
-%
 \thanks{{}$^{(7)}$ xxx}
@@
 %
-\thanks{{}$^{(10)}$ xxx}
+\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; (\url{sctchoi@uchicago.edu}}
 % 
@@
 descriptive and prescriptive approaches to communities. 
-\note{Choi: Do we need to itemize the reference numbers below?
-Please collapse them into a comma seperated list with a short description for all or each of the papers.
-}
-
 \begin{itemize}
@@
 
+Two issues were raised in a number of the workshops sessions, defining sustainability and career paths.  This section brings those discussions together.
+
 \subsection{Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:defining-sustainability}
 
-\note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
-\note{Dan: assuming the previous note is from FL, what do you propose?}
+%\note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
+%\note{Dan: assuming the previous note is from FL, what do you propose?}
 The question of what was meant by ``sustainability'' in the context of software
@@
 
-\subsection{Career Tracks for Scientific Software Developers} \label{sec:career-tracks}
+\subsection{Career Tracks for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:defining-sustainability}\label{sec:career-tracks}
 
@@
 
+\katznote{discuss sustainability survey}
+
 \section*{Acknowledgments}

### 2014-04-08 WSSSPE1_report/WSSSPE1_summary_report.tex
minor fix

@@
 standards for measuring software quality. Specifically, they propose an additional
-dimension of quality they call perdurability or ability to ``last over
+dimension of quality they call ``perdurability" or ability to ``last over
 time'' made up of three characteristics: reliability, maintainability
@@
 ``combined policy, implementation, and deployment control''. Web APIs
-can and do change, sometimes quite frequently. raising the need to
+can and do change, sometimes quite frequently. Raising the need to
 port dependent applications. The effort required for porting is

### 2014-04-08 WSSSPE1_report/WSSSPE1_summary_report.tex
adding affiliation
add email
proofread pass on abstract

@@
 %
-\thanks{{}$^{(3)}$ xxx}
+\thanks{{}$^{(3)}$ Argonne National Laboratory, Chicago, IL, USA; \url{ketan@anl.gov}}
 %
@@
 \begin{abstract} \note{Lead: Sou-Cheng T. Choi}
-This document discusses the First Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE1).  The content
-is based on the workshop itself, as recorded in a set of
-collaborative notes taken during the workshop, including discussion
-about the keynote presentations, three themed panels that focus on
-papers accepted by the workshop, other discussion, cross-cutting
-issues, use cases, and conclusions.  It joins a nascent literature
-seeking to understand what drives software work in science and how the
-reward systems of science thereby shapes the type of software work
-undertaken, including the extent to which developers are motivated to
-build software for the long-term, for the use of others, and whether
-to work collaboratively or separately. Unique perspectives were
-captured about issues such as documentation of APIs, software
-deployment, and development practices; software licenses; career
-paths for scientific software developers; identifiers for software
-through a mechanism such as a Digital Object Identifier or
-publication of a ``software paper''; and evidence of software
-contribution and impact, recorded in online systems such as source
-code repositories like GitHub. For truly sustainable software there
-should be no endpoint, as the software products continue to be used
-and useful beyond the single institution, grant, and developer or
+This document discusses the First Workshop on Sustainable Software for Science:
+Practice and Experiences (WSSSPE1). The content is based on the workshop
+itself, as recorded in a set of collaborative notes taken during the workshop,
+including discussion about the keynote presentations, three themed panels that
+focused on papers accepted at the workshop, other discussion, cross-cutting
+issues, use cases, and conclusions. It joins a nascent literature seeking to
+understand what drives software work in science and how the reward systems of
+science work, thereby shaping the type of software work undertaken, including the
+extent to which developers are motivated to build software for the long-term,
+for the use of others, and whether to work collaboratively or separately.
+Unique perspectives were captured about issues such as documentation of APIs,
+software deployment, and development practices; software licenses; career paths
+for developers; identifiers for software through a
+mechanism such as a Digital Object Identifier or publication of a ``software
+paper''; and evidence of software contribution and impact, recorded in online
+systems such as source code repositories like GitHub. For truly sustainable
+software, there should be no endpoint, as the software products continue to be
+used and useful beyond a single institution, grant, and developer or
 development team.

### 2014-04-09 WSSSPE1_report/WSSSPE1_summary_report.tex
Adds affiliation, email, and support ack for HL.
Fixes double quote error, and doubly defined label.

@@
 %
-\thanks{{}$^{(2)}$ xxx}
+\thanks{{}$^{(2)}$ National Evolutionary Synthesis Center (NESCent),
+  Durham, NC, USA; \url{hlapp@nescent.org}}
 %
@@
 standards for measuring software quality. Specifically, they propose an additional
-dimension of quality they call ``perdurability" or ability to ``last over
+dimension of quality they call ``perdurability'' or ability to ``last over
 time'' made up of three characteristics: reliability, maintainability
@@
 
-\subsection{Career Tracks for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:defining-sustainability}\label{sec:career-tracks}
+\subsection{Career Tracks for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:career-tracks}
 
@@
 expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation.
+necessarily reflect the views of the National Science Foundation. HL
+was supported by the National Evolutionary Synthesis Center (NESCent), 
+NSF #EF-0905606.
 

### 2014-04-09 WSSSPE1_report/WSSSPE1_summary_report.tex
Hetherington affiliation
SSI funding acknowledged.
Adding my affiliation

@@
 %
-\thanks{{}$^{(5)}$ xxx}
+\thanks{{}$^{(5)}$ Columbia Astrophysics Laboratory, Columbia University, New
+York, NY, USA; \url{matthewturk@gmail.com}}
 %
@@
 %
-\thanks{{}$^{(9)}$ xxx}
+\thanks{{}$^{(9)}$ Leader, Research Software Development Team, University College London; \url {j.hetherington@ucl.ac.uk}}
 %
@@
 
+Chue-Hong and Hetherington were supported by the UK Engineering and 
+Physical Sciences Research Council (EPSRC) Grant EP/H043160/1 for the 
+UK Software Sustainability Institute.
+
 \appendix

### 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
Fix a compilation error.
Adding MINRES-QLP paper to the relevant sections.
fix author problem

@@
 Katz$^{(1)}$, Hilmar Lapp$^{(2)}$, Ketan Maheshwari$^{(3)}$, Frank L\"{o}ffler$^{(4)}$, Matthew Turk$^{(5)}$, Marcus D.
-Hanwell$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$, James Hetherington$^{(9)}$, James Howison$^{(9)}$, Sou-Cheng T.
+Hanwell$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$, James Hetherington$^{(8)}$, James Howison$^{(9)}$, Sou-Cheng T.
 Choi$^{(10)}$, Shel Swenson$^{(11)}$,  Bruce Berriman$^{(12)}$, Andy R.
@@
 %
-\thanks{{}$^{(8)}$ xxx}
+\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London; \url {j.hetherington@ucl.ac.uk}}
 %
-\thanks{{}$^{(9)}$ Leader, Research Software Development Team, University College London; \url {j.hetherington@ucl.ac.uk}}
+\thanks{{}$^{(9)}$ xxx}
 %
@@
 was supported by the National Evolutionary Synthesis Center (NESCent), 
-NSF #EF-0905606.
+NSF \#EF-0905606.
 
@@
 \begin{itemize}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
 
@@
 
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
 \end{itemize}

### 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
add my affiliation
starting to work on cross-cutting Career Track section - in progress

@@
 %
-\thanks{{}$^{(4)}$ xxx}
+\thanks{{}$^{(4)}$ Center for Computation and Technology, Louisiana State University, Baton Rouge, LA, USA; \url{knarf@cct.lsu.edu}}
 %
@@
 a research output alongside papers, a common discussion within this field.
-An interesting reference provided by him
-is~\cite{peer-review-code}, which explores involving software
-engineers in the review process of scientific code.
+He mentioned an interesting exploration of involving software
+engineers in the review process of scientific code~\cite{peer-review-code}.
 
@@
 with one of the major reasons being no long-term career paths (especially when
-compared with industry). How should software development be balanced with
-research? It is apparent that things are beginning to improve, such as
-incentives for software development in the form of altmetrics, tenure committee
-consideration, and NSF ``products'' vs ``publications.'' It was noted it can be very
-difficult to measure where people are using small bits of your code.
+compared with industry). This is further discussed in \S\ref{sec:career-tracks}.
+
+%moved to cross-cutting issues
+%
+%How should software development be balanced with
+%research? It is apparent that things are beginning to improve, such as
+%incentives for software development in the form of altmetrics, tenure committee
+%consideration, and NSF ``products'' vs ``publications.'' It was noted it can be very
+%difficult to measure where people are using small bits of your code.
 
@@
 
-How to ensure software is sustainable by ensuring there are career
-paths for developers.
+Career track issues also came up repeatedly, starting in the the first keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the Google Bus'' to describe the issue of staff in his lab in San Diego who had chosen to leave science and work for Google (but since they didn't want to physically move, they take the Google bus to the Google office each day.)
+
+The lack of career paths also came up in the Developing and Supporting Software discussion (\S\ref{sec:devel}), as an issue that led to difficulties in established projects retaining as well as recruiting staff.
+The issue there was that for software to be sustainable, there need to be career paths for software developers.
+There was also some discussion about how should software development be balanced with research, with the idea that faculty are hired based on their research qualifications, not on their software development qualification, and as such, specializing in software development was likely to not lead to a career in science at a university, though clearly, national laboratories are a different case.
+
+In terms of possible solutions, the Developing and Supporting Software discussion (\S\ref{sec:devel}) said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketches of asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
+However, it was also noted it can be very difficult to measure where people are using small bits of your code.
+
+Open questions:
 

### 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
small label fixes for appendices

@@
 use cases (\S\ref{sec:use-cases}), and conclusions
-(\S\ref{sec:conclusions}).
+(\S\ref{sec:conclusions}).  Additionally, the accepted workshop papers are listed in
+Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
+in Appendix~\ref{sec:attendees}.
 
@@
 
-The organizers decided to list 54 of the papers as significant
+The organizers decided to list 54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
 contributions to the workshop, a very high acceptance rate but one
@@
 \appendix
-\section{Papers Accepted and Discussed at WSSSPE1}
+\section{Papers Accepted and Discussed at WSSSPE1} \label{sec:papers}
 
@@
 
-\section{Attendees \note{lead: Shel Swenson}}
+\section{Attendees \note{lead: Shel Swenson}} \label{sec:attendees}
 

### 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
adding Nancy's affil and email
Added cites for two papers that mention enviornment sustaiability

@@
 %
-\thanks{{}$^{(7)}$ xxx}
+\thanks{{}$^{(7)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
 %
@@
 overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned this\todo{which are these
-  papers? Should cite here.} did so without
+The two papers in this area which mentioned this \cite{Venters_WSSSPE,Weber_WSSSPE} did so without
 integrating that analysis into the question of software being around

### 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
Added Contact details for James Howison.
Added non-breaking spaces in author names to avoid line breaks between first and last names.

@@
 
-\author{Daniel S.
-Katz$^{(1)}$, Hilmar Lapp$^{(2)}$, Ketan Maheshwari$^{(3)}$, Frank L\"{o}ffler$^{(4)}$, Matthew Turk$^{(5)}$, Marcus D.
-Hanwell$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$, James Hetherington$^{(8)}$, James Howison$^{(9)}$, Sou-Cheng T.
-Choi$^{(10)}$, Shel Swenson$^{(11)}$,  Bruce Berriman$^{(12)}$, Andy R.
-Terrel$^{(13)}$, Colin Venters$^{(14)}$}
+\author{Daniel~S.~Katz$^{(1)}$, Hilmar~Lapp$^{(2)}$, Ketan~Maheshwari$^{(3)}$, Frank~L\"{o}ffler$^{(4)}$, Matthew~Turk$^{(5)}$, Marcus~D.~Hanwell$^{(6)}$, Nancy~Wilkins-Diehr$^{(7)}$, James~Hetherington$^{(8)}$, James~Howison$^{(9)}$, Sou-Cheng~T.~Choi$^{(10)}$, Shel~Swenson$^{(11)}$,  Bruce~Berriman$^{(12)}$, Andy~R.~Terrel$^{(13)}$, Colin~Venters$^{(14)}$}
 \note{Authors, please add/check middle initials and affiliations.}
@@
 %
-\thanks{{}$^{(9)}$ xxx}
+\thanks{{}$^{(9)}$ University of Texas at Austin; \url{jhowison@ischool.utexas.edu}}
 %

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Proposing a paragraph to connect with ongoing work about measurement in scientific software communities, including a working paper and website that outlines appropriate measures and practices. Perhaps too post-workshop but I hate to see things unconnected.  Certianly the cite to the Millington book is appropriate, given that it came up in the discussion (well, I brought it up) and was in the notes

--- 2014-04-10 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\subsubsection{Challenges of community}
-include metrics for community success (e.g., more external
-contributors than internal).
+\subsubsection{Measuring community}
+How might a project know when it has built a sustainable community?
+How might an outsider be able to assess the activity and sustainability of that community?
+These questions have been partially addressed in the literature.
+For example, a key metric in online communities in general is the cross-over point where there are more external contributors than internal ones.
+Richard Millington's book ``Buzzing Communities'' does an excellent job of outlining these measures, drawing on online communities research in an accessible manner~\cite{millington_buzzing_2012}.
+Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Cleaned up later part of the Communities section; the intent of those sections definitely met by the Kitware story and the cases presented in the cross-cutting section.
Fix typo.

@@
 %
-\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; (\url{sctchoi@uchicago.edu}}
+\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
 % 
@@
 
-\subsubsection{Admirable scientific software communities}
-\begin{itemize}
-\item example communities
-\item their (quick) origin stories.
-\end{itemize}
-
-\subsubsection{Resources for learning about software communities}
+\subsubsection{Additional Resources for learning about software communities}
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Refine abstract.

@@
 
-\begin{abstract} \note{Lead: Sou-Cheng T. Choi}
-This document discusses the First Workshop on Sustainable Software for Science:
+\begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
+This document summarizes the First Workshop on Sustainable Software for Science:
 Practice and Experiences (WSSSPE1). The content is based on the workshop
@@
 including discussion about the keynote presentations, three themed panels that
-focused on papers accepted at the workshop, other discussion, cross-cutting
+discussed papers accepted at the workshop, other discussion, cross-cutting
 issues, use cases, and conclusions. It joins a nascent literature seeking to
@@
 for the use of others, and whether to work collaboratively or separately.
-Unique perspectives were captured about issues such as documentation of APIs,
+Unique perspectives were captured about issues such as comprehensive documentation,
 software deployment, and development practices; software licenses; career paths

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Make sure a name does not split into two lines.

@@
 
-\author{Daniel S.
-Katz$^{(1)}$, Hilmar Lapp$^{(2)}$, Ketan Maheshwari$^{(3)}$, Frank L\"{o}ffler$^{(4)}$, Matthew Turk$^{(5)}$, Marcus D.
-Hanwell$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$, James Hetherington$^{(8)}$, James Howison$^{(9)}$, Sou-Cheng T.
-Choi$^{(10)}$, Shel Swenson$^{(11)}$,  Bruce Berriman$^{(12)}$, Andy R.
-Terrel$^{(13)}$, Colin Venters$^{(14)}$}
+\author{Daniel~S.~Katz$^{(1)}$, Hilmar~Lapp$^{(2)}$, Ketan~Maheshwari$^{(3)}$, Frank~L\"{o}ffler$^{(4)}$, 
+Matthew~Turk$^{(5)}$, Marcus~D.~Hanwell$^{(6)}$, Nancy~Wilkins-Diehr$^{(7)}$, James~Hetherington$^{(8)}$, 
+James~Howison$^{(9)}$, Sou-Cheng~T.~Choi$^{(10)}$, Shel~Swenson$^{(11)}$,  Bruce~Berriman$^{(12)}$, Andy~R.~Terrel$^{(13)}$, Colin~Venters$^{(14)}$}
 \note{Authors, please add/check middle initials and affiliations.}

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Squeeze the note.

@@
 
-\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
+\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) 
+\note{\scriptsize Authors, please add/check middle initials and affiliations}}
 
@@
 James~Howison$^{(9)}$, Sou-Cheng~T.~Choi$^{(10)}$, Shel~Swenson$^{(11)}$,  Bruce~Berriman$^{(12)}$, Andy~R.~Terrel$^{(13)}$, Colin~Venters$^{(14)}$}
-\note{Authors, please add/check middle initials and affiliations.}
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Moved Kitware story and added Globus summary.

@@
 
-In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
-source software development at Kitware.  In particular, they focus on their
-success at growing their community of users through enabling multiple channels
-of communication, directly reaching out to individuals, and lowering the
-barrier to entry for contributions. This involves providing clear,
-test-oriented and review-based mechanisms for evaluating contributions,
-permissive licenses, and a service-based model for sustaining development.
-This model enables Kitware to review both public funding, as well as private
-funding to support improvements and targeted developments of the software.
-
 At NESCent, a combination of in-house informatics individuals and domain
@@
 
-\todo{text about papers and discussion in this part of the workshop}
+Several papers presented discussed the connection between industry and
+scientific software, from the perspective of both integrating efforts between
+the two and sustaining long-term development.
+
+In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
+source software development at Kitware.  In particular, they focus on their
+success at growing their community of users through enabling multiple channels
+of communication, directly reaching out to individuals, and lowering the
+barrier to entry for contributions. This involves providing clear,
+test-oriented and review-based mechanisms for evaluating contributions,
+permissive licenses, and a service-based model for sustaining development.
+This model enables Kitware to review both public funding, as well as private
+funding to support improvements and targeted developments of the software.
+
+\cite{Foster_WSSSPE} discussed the approach of developing sustainability models
+around Software-as-a-Service (SaaS) platforms, with the target example being
+that of GlobusOnline.  The authors build a case that both grant-based and
+volunteer-based development fall short in sustaining software, resulting in
+software that is disproportionately difficult to use compared to its
+functionality, which they note directly impacts the overall scientific
+productivity of its users.  In contrast, by charging a subscription fee for
+hosted, centrally-managed software (similar to offerings by Dropbox, EverNote,
+GMail), the authors propose to manage the funding cycle and enable a greater
+focus on the aspects of software that directly impact individuals, rather than
+funders.  Globus has deployed such a service, for which they have attempted to
+develop a sustainable economic model that reduces institutional obstacles to
+funding and subscription.  However, they do identify that cultural obstacles
+do still remain, and they note a particular difference in culture between NSF-
+and NIH-funded researchers.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes in communities & finishing career path section

--- 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 
-\subsection{Credit, Citation, Impact}
+\subsection{Credit, Citation, and Impact}
 
@@
 facilitating many or all facets of sustainanability-oriented software
-design: Blanton and Lenhardt\cite{Blanton_WSSSPE} contrast
+design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast
 large projects that have software infrastructure development built-in,
@@
 
-Drawing on experiences from high-energy physics, \cite{Vay_WSSSPE} proposed
+Drawing on experiences from high-energy physics, Vay et al.~\cite{Vay_WSSSPE} proposed
 developing teams of technical specialists able to overcome a lack of
@@
 
-In \cite{Hanwell_WSSSPE}, the authors reflect on the 15-year history of open
+Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
 source software development at Kitware.  In particular, they focus on their
@@
 
-\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
+Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
 collaborations supporting tool development for Open Science.  This software
@@
 
-In terms of possible solutions, the Developing and Supporting Software discussion (\S\ref{sec:devel}) said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketches of asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
+In the Policy discussion (\S\ref{sec:policy}), career paths were not discussed explicitly, though the discussion about credit, citation, and impact focused on how those who create or contribute to software are recognized; this recognition (or the current lack thereof) is tied to career paths (or their lack) for software developers.
+
+The Communities discussion (\S\ref{sec:community}) did not directly address career paths, though one of the papers~\cite{Loffler_WSSSPE} that was part of this section also makes a similar policy point, in a section about career paths.
+Specifically, it says ``The most severe problem for developers in most computational sciences currently is that while most of the work is done creating hopefully well-written, sustainable software, the academic success is often exclusively tied to the solution of the scientific problem the software was designed for.
+Tasks that from a software engineering standpoint are essential, e.g., high usability, well-written and updated documentation, or porting infrastructure to new platforms, are not rewarded within this system.''
+
+In terms of possible solutions, the Developing and Supporting Software discussion said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketch requirement: asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
 However, it was also noted it can be very difficult to measure where people are using small bits of your code.
-
-Open questions:
-
-What are the possible career paths for a specialist software developer working as part of a scientific research group?
+And the possible career paths for a specialist software developer working as part of a scientific research group in a university are certainly not yet clear.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 
-Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
-source software development at Kitware.  In particular, they focus on their
-success at growing their community of users through enabling multiple channels
-of communication, directly reaching out to individuals, and lowering the
-barrier to entry for contributions. This involves providing clear,
-test-oriented and review-based mechanisms for evaluating contributions,
-permissive licenses, and a service-based model for sustaining development.
-This model enables Kitware to review both public funding, as well as private
-funding to support improvements and targeted developments of the software.
-
 At NESCent, a combination of in-house informatics individuals and domain
@@
 
-\todo{text about papers and discussion in this part of the workshop}
+Several papers presented discussed the connection between industry and
+scientific software, from the perspective of both integrating efforts between
+the two and sustaining long-term development.
+
+Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
+source software development at Kitware.  In particular, they focus on their
+success at growing their community of users through enabling multiple channels
+of communication, directly reaching out to individuals, and lowering the
+barrier to entry for contributions. This involves providing clear,
+test-oriented and review-based mechanisms for evaluating contributions,
+permissive licenses, and a service-based model for sustaining development.
+This model enables Kitware to review both public funding, as well as private
+funding to support improvements and targeted developments of the software.
+
+Foster et al.~\cite{Foster_WSSSPE} discussed the approach of developing sustainability models
+around Software-as-a-Service (SaaS) platforms, with the target example being
+that of GlobusOnline.  The authors build a case that both grant-based and
+volunteer-based development fall short in sustaining software, resulting in
+software that is disproportionately difficult to use compared to its
+functionality, which they note directly impacts the overall scientific
+productivity of its users.  In contrast, by charging a subscription fee for
+hosted, centrally-managed software (similar to offerings by Dropbox, EverNote,
+GMail), the authors propose to manage the funding cycle and enable a greater
+focus on the aspects of software that directly impact individuals, rather than
+funders.  Globus has deployed such a service, for which they have attempted to
+develop a sustainable economic model that reduces institutional obstacles to
+funding and subscription.  However, they do identify that cultural obstacles
+do still remain, and they note a particular difference in culture between NSF-
+and NIH-funded researchers.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Adding Elster paper summary.

--- 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 and NIH-funded researchers.
+
+While many papers briefly discussed licensing issues, \cite{Elster_WSSSPE}
+directly discussed the impact of different licenses on obtaining industrial
+funding for scientific software projects.  In particular, the author describes
+experiences with researchers unwilling to utilize GPL (copyleft) code, as it
+adds restrictions to reuse that some industries find unacceptable for future
+commercialization.  (This was discussed in \cite{Hanwell_WSSSPE} as well.)
+Further obstacles to sustainability identified by the author include patenting
+of software, a lack of open access, and even more broadly, a lack of open
+source codes even in open access journals.  Finally, a lack of awareness on the
+part of scientific software developers of commodity libraries for common tasks
+reduces their ability to reuse code.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
minor edits to new text

--- 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
-source software development at Kitware.  In particular, they focus on their
+source software development at \toolname{Kitware}.  In particular, they focus on their
 success at growing their community of users through enabling multiple channels
@@
 permissive licenses, and a service-based model for sustaining development.
-This model enables Kitware to review both public funding, as well as private
+This model enables \toolname{Kitware} to review both public funding, as well as private
 funding to support improvements and targeted developments of the software.
@@
 around Software-as-a-Service (SaaS) platforms, with the target example being
-that of GlobusOnline.  The authors build a case that both grant-based and
+that of \toolname{GlobusOnline}.  The authors build a case that both grant-based and
 volunteer-based development fall short in sustaining software, resulting in
@@
 productivity of its users.  In contrast, by charging a subscription fee for
-hosted, centrally-managed software (similar to offerings by Dropbox, EverNote,
-GMail), the authors propose to manage the funding cycle and enable a greater
+hosted, centrally-managed software (similar to offerings by \toolname{Dropbox}, \toolname{EverNote},
+\toolname{GMail}), the authors propose to manage the funding cycle and enable a greater
 focus on the aspects of software that directly impact individuals, rather than
-funders.  Globus has deployed such a service, for which they have attempted to
+funders.  \toolname{Globus} has deployed such a service, for which they have attempted to
 develop a sustainable economic model that reduces institutional obstacles to
@@
 and NIH-funded researchers.
+
+While many papers briefly discussed licensing issues, Elster~\cite{Elster_WSSSPE}
+directly discussed the impact of different licenses on obtaining industrial
+funding for scientific software projects.  In particular, she describes
+experiences with researchers unwilling to utilize GPL (copyleft) code, as it
+adds restrictions to reuse that some industries find unacceptable for future
+commercialization.  (This was discussed by Hanwell et al.~\cite{Hanwell_WSSSPE} as well.)
+Further obstacles to sustainability identified by the author include patenting
+of software, a lack of open access, and even more broadly, a lack of open
+source codes even in open access journals.  Finally, a lack of awareness on the
+part of scientific software developers of commodity libraries for common tasks
+reduces their ability to reuse code.
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
light editing in communities section - in progress

@@
 developing teams of technical specialists able to overcome a lack of
-coordination between projects.  Maximization of scientific output requires
+coordination between projects.  Maximizing scientific output requires
 maximizing the usability of the scientific codes while minimizing the cost of
@@
 
-The paper by Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
+Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
 catalysts'' and their role in the modern scientific research process. A technology
-catalyst can be defined as a role played by an individual with a know-how of
-technological advancements, tasked with user-engagement with a goal of enacting
-scientific or engineering applications, and using suitable tools and techniques
-to take advantage of technological capabilities thus benefiting applications.
-
+catalyst is an individual with knowledge of
+technological advancements, tasked with user engagement to create
+scientific or engineering applications, using suitable tools and techniques
+to take advantage of current technological capabilities.
+%
 One of the tasks of technology catalysts is to seek community collaborations for new
-applications and user engagement thus benefiting both: science, by effective
+applications and engage users, thus benefiting both science, by effective
 running of scientific codes on computational infrastructure, and technology,
 by conducting research and seeking findings for technology improvement.
-The particular engagements described in the paper came up from authors work as
-postdoctoral researcher at Cornell and Argonne. Interaction with the scientific
+The particular engagements described in the paper came up from the lead author's work as
+a postdoctoral researcher at Cornell and Argonne, where interaction with the scientific
 communities in both institutions resulted in these collaborations.
@@
 improvements. From this experiment, the authors identified several key
-outcomes as well as lessons-learned -- specifically, the co-localization of
+outcomes as well as lessons-learned: specifically, the co-localization of
 developers was seen as having a strong impact, enabling casual conversation
@@
 between individuals and data providers, designed to ``train academics in
-reproducible science workflows focused around R.''  The two central challenges
-to this goal were identified as engagement of existing users within ecology and
-evolutionary biology (EEB), and how the community could make inroads and traction
+reproducible science workflows focused around R.''  Two central challenges
+are engagement of existing users within ecology and
+evolutionary biology (EEB), and how the community can make inroads and traction
 in other disciplines. Currently, the collective is exploring addressing these
-challenges through use of social media and holding workshops and hackathons.
+challenges through use of social media, holding workshops, and hackathons.
 This helps to both raise the profile of the collective within EEB and in other
@@
 \begin{itemize}
-\item Step 1: Design - thorough discussion of research questions
-\item Step 2: Develop working code
-\item Step 3: Refine based on new requirements
-\item Step 4: Publish open source
+\setlength{\itemindent}{0.5in}
+\item[Step 1:] Design - thorough discussion of research questions
+\item[Step 2:] Develop working code
+\item[Step 3:] Refine based on new requirements
+\item[Step 4:] Publish open source
 \end{itemize}

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
edits in communities text

@@
 Christopherson reports on the application of Steps 1-3 to a
-computational modeling framework developed in the 1990s. Step 1 was realized as
-a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
+computational modeling framework developed in the 1990s. Step 1 was
+a 2-day, in-person specifications meeting and code walkthrough. Step 2 was a
 5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
@@
 work together, and experimentation was encouraged.
-
-The paper summarized recommendations to others wishing to engage in this or a
-similar process, starting small and gradually building toward more complex
-objectives. This is consistent with Agile development. Refactor before adding
-new functionality. Approach development as a learning experience. Welcome
-experimentation, and treat mistakes as a natural part of the learning process.
-Repeat Step 1 activities before all hackathons to develop consensus before
-coding. This allows coding to be the focus of subsequent hackathons. In higher
-risk situations, provide additional time for Step 1 activities. The
-Christopherson team recommends a minimum of two months. Ensure any newcomers
-receive some form of orientation prior to the hackathon, such as a code
-walkthrough or system documentation. Co-locate rather than collaborating
-remotely whenever feasible.
+%
+The paper recommended: start small and gradually building toward more complex
+objectives, consistent with Agile development; develop consensus before
+coding, by repeating step 1 before all hackathons; ensure newcomers
+receive orientation prior to the hackathon, such as a code
+walkthrough or system documentation; and co-locate collaborators
+whenever feasible.
+
+%Refactor before adding
+%new functionality. Approach development as a learning experience. Welcome
+%experimentation, and treat mistakes as a natural part of the learning process.
+%Repeat Step 1 activities before all hackathons to develop consensus before
+%coding. This allows coding to be the focus of subsequent hackathons. In higher
+%risk situations, provide additional time for Step 1 activities. The
+%Christopherson team recommends a minimum of two months. 
+%Ensure any newcomers
+%receive some form of orientation prior to the hackathon, such as a code
+%walkthrough or system documentation. Co-locate rather than collaborating
+%remotely whenever feasible.
 
 Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
-entry to complex cyberinfrastructure. The paper opens with a description of the
-explosion of use of just a few gateways. Over 3.5 years, more than 7,000
-biologists have run phylogenetic codes on supercomputers using the\toolname{CIPRES}
-Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
-UltraScan Science Gateway, increasing the sophistication of analytical
+entry to complex cyberinfrastructure. For example, more than 7,000
+biologists have run phylogenetic codes on supercomputers using the \toolname{CIPRES}
+Science Gateway in 3 1/2 years. Over 120 scientists from 50 institutions used the
+\toolname{UltraScan Science Gateway} in one year, increasing the sophistication of analytical
 ultracentrifugation experiments worldwide. The new Neuroscience Gateway (\toolname{NSG})
@@
 Gateways, however, need to keep operational costs low and can often make use of
-common components. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
+common components, such as authentication, application installation and
+reliable execution and help desk support. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
 middleware as a hosted, scalable third-party service while domain developers
-focus on user interfaces and domain-specific data in the gateway. The
-middleware handles things like authentication, application installation and
-reliable execution and help desk support.
-
-One key to \toolname{SciGaP} is its openness. While based on the \toolname{Apache Airavata} project
+focus on user interfaces and domain-specific data in the gateway. 
+While \toolname{SciGaP} is based on the \toolname{Apache Airavata} project
 and the \toolname{CIPRES} Workbench Framework, community contributions are encouraged
-because of its open source, open governance and open operations policies. The
+by its open source, open governance and open operations policies. The
 goal is robust, sustainable infrastructure with a cycle of development that
@@
 Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
-nanoHUB.org community. The authors define community as a ``body of persons of
+\toolname{nanoHUB.org} community, where community is defined as a ``body of persons of
 common and especially professional interests scattered through a larger
-society.'' This makes support challenging because of the diversity of
+society.'' Support is challenging because of the diversity of
 viewpoints and needs. The group constantly examines its policies to determine
@@
 particular types of use.
-
-nanoHUB's 10-year history with over 260,000 users annually provides a lot of
-data to analyze. 4000 resources contributed by 1000 authors. nanoHUB serves
+%
+\toolname{nanoHUB}'s 10-year history with over 260,000 users annually provides a lot of
+data to analyze. 4000 resources contributed by 1000 authors. \toolname{nanoHUB} serves
 both the research and education community and the contribution model allows
@@
 very rapidly. The primary software challenges are twofold - support for the
-HUBzero framework and challenges related to the software contributed by the
+\toolname{HUBzero} framework and challenges related to the software contributed by the
 community.
@@
 The group has learned that community contributions are maximized with a
-tolerant licensing approach. HUBzero uses an LGPLv3 license so contributors can
+tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3 license so contributors can
 create unique components and license as they choose. If they make changes to
@@
 community provides ratings, poses questions and contributes to wishlists and
-citation counts - all of which incentivize code authors.
+citation counts, all of which incentivize code authors.
 
 Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
-two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
+two major efforts: the SciPy conference and the NumFOCUS foundation. 
+Since software sustainability relies on contributions from all sectors of the user
+community, these efforts support these sectors, and help develop and mature Python.
+%
+The reliance
 on software in science has driven a huge demand for development, but this
 development is typically done as a side effort and often in a rush to publish
-without documentation and testing. While created by academics, software support
-often falls to industrial institutions. SciPy brings together industry,
+without documentation and testing. While the software is often created by academics, software support
+can fall to industrial institutions. SciPy brings together industry,
 government, and academics to share their code and experience in an open
 environment.
-
-NumFOCUS promotes open, usable scientific software while sustaining the
-community. Specific activities include educational programs, collaborative
-research tools and documentation and promotion of high-level languages,
-reproducible scientific research, and open-code development. Governance of the
-non-profit is a loose grantor-grantee relationship with projects allowing for
-monies to be placed in the groups accounts. This has raised money to hire
+%
+NumFOCUS in a non-profit that promotes open, usable scientific software while sustaining the
+community through educational programs; collaborative
+research tools and documentation; and promotion of high-level languages,
+reproducible scientific research, and open-code development. Governance
+is a loose grantor-grantee relationship with projects, allowing funds
+to be placed in the groups' accounts. This has raised money to hire
 developers for open code development, maintain testing machines, organize the
@@
 
-Software sustainability relies on contributions from all sectors of the user
-community. Together SciPy and NumFOCUS support these sectors. By maximizing
-contributions growing the user base they help develop and mature Python.
-
-L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project which was started in 1996 by
-participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
-and thorns, core and module, model - a community-oriented framework that allows
+L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
+ which was started in 1996 by
+participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh (core)
+and thorns (modules) model, a community-oriented framework that allows
 researchers to easily work together with reusable and extendable software
@@
 
-The paper points out 4 keys to sustaining the community - modular design,
+The paper points out 4 keys to sustaining the community: modular design,
 growing a collaborative community, career paths and credit. In modular design,
@@
 best programmers in a science environment often have multidisciplinary
-expertise. That also is not rewarded in academia.
+expertise. This also is not rewarded in academia.
 
 Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
-community of those creating science gateways or web portals for science. These
+community of those creating science gateways for science. These
 gateways, as described in some of the other papers in this section, can be
-quite capable, having a remarkable impact on some parts of science.
-
-This paper mentioned challenges highlighted by other papers in this section -
-mainly the conflict between funding for research vs infrastructure and the
-challenges around getting academic credit for infrastructure. Because the
-authors have studied many projects, they've also observed how development is
+quite capable and can have strong scientific impact.
+Challenges are similar to those highlighted by other papers in this section:
+the conflict between funding for research vs infrastructure and the
+challenges around getting academic credit for infrastructure. 
+%
+The authors observe that  development is
 often done in an isolated hobbyist environment. Developers are unable to take
@@
 provide infrastructure to conduct research and many times only the research is
-funded. Gateways also may start as small group research project, taking off in
-popularity once they go live without any long term plans for sustainability or
-without resources in the project to plan for such. Subsequent disruptions in
+funded. Gateways also may start as a small group research project, taking off in
+popularity once they go live, without any long term plans for sustainability.
+Subsequent disruptions in
 service can limit effectiveness and test the limits of the research community's
-trust. The impact of gateways can be increased substantially if we understand
-what makes them successful.
-
-Recommendations from an early study of successful gateways include the
-following. Leadership and management teams should design governance to
+trust. %The impact of gateways can be increased substantially if we understand
+%what makes them successful.
+
+Recommendations from an early study of successful gateways include:
+1) Leadership and management teams should design governance to
 represent multiple strengths and perspectives, plan for change and turnover in
@@
 domain-related issues, consider sustainability and measure success early and
-often. Successful projects recognize the benefits and costs of hiring a team of
+often. 2) Projects should hire a team of
 professionals, demonstrate credibility through stability and clarity of
-purpose, leverage the work of others and plan for flexibility. Successful
-projects identify an existing community and understand the communities' needs
-before beginning. Projects adapt as the communities' needs evolve. For funders,
-the lifecycle of technology projects must be considered. Solicitations should
-be designed to reward effective planning, recognize the benefits and
+purpose, leverage the work of others, and plan for flexibility. 3) Projects should
+identify an existing community and understand the communities' needs
+before beginning, then adapt as the communities' needs evolve. 4) Funders,
+should consider the technology project lifecycle, and design solicitations 
+to reward effective planning, recognize the benefits and
 limitations of both technology innovation and reuse, expect adjustments during
@@
 
-Through a business incubator type approach, the institute plans to provide a
-variety of services that could be shared amongst projects. Consultation and
-resources on topics such as business plan development, software engineering
-practices, software licensing options, usability, security and project
-management as well as a software repository and hosting service will be
-available. Experts will be available for multi-month assignments to help
-research teams build their own gateways, teaching them how to maintain and add
-to the work after the collaboration ends. Forums, symposia and an annual
-conference will connect members of the development community. A modular,
-layered framework that supports community contributions and allows developers
-to choose components will be delivered and finally workforce development
-activities will help train the next generation for careers in this cross-
-disciplinary area and build pools of institutional expertise that many projects
-can leverage. Shared services and forums can dramatically reduce the cost of
-building and sustaining gateways. Workforce development can encourage
-technology professionals to remain in the sciences.
+%Through a business incubator type approach, the institute plans to provide a
+%variety of services that could be shared amongst projects. Consultation and
+%resources on topics such as business plan development, software engineering
+%practices, software licensing options, usability, security and project
+%management as well as a software repository and hosting service will be
+%available. Experts will be available for multi-month assignments to help
+%research teams build their own gateways, teaching them how to maintain and add
+%to the work after the collaboration ends. Forums, symposia and an annual
+%conference will connect members of the development community. A modular,
+%layered framework that supports community contributions and allows developers
+%to choose components will be delivered and finally workforce development
+%activities will help train the next generation for careers in this cross-
+%disciplinary area and build pools of institutional expertise that many projects
+%can leverage. Shared services and forums can dramatically reduce the cost of
+%building and sustaining gateways. Workforce development can encourage
+%technology professionals to remain in the sciences.
 
@@
 Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}.
+\katznote{please change to cite}
 

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Author details for James H; tidied up double citation; fixed cite re environmental sustainability

@@
 %
-\thanks{{}$^{(9)}$ xxx}
+\thanks{{}$^{(9)}$ University of Texas at Austin; jhowison@ischool.utexas.edu}
 %
@@
 overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned this\todo{which are these
-  papers? Should cite here.} did so without
+The two papers in this area which mentioned this~\cite{Venters_WSSSPE,Calero_WSSSPE} did so without
 integrating that analysis into the question of software being around
@@
 \begin{itemize}
-\item \cite{howison_scientific_2011}
+\item \cite{howison_scientific_2011, howison_incentives_2013}
 \item \cite{citeulike:11831265}
@@
 \item \cite{opac-b1134063}
-\item \cite{DBLP:conf/cscw/HowisonH11}
-\item \cite{DBLP:conf/cscw/2011}
 \item \cite{Trapani:2011}

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
Added url around my email address
minor edits

@@
 %
-\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London; \url {j.hetherington@ucl.ac.uk}}
-%
-\thanks{{}$^{(9)}$ University of Texas at Austin; jhowison@ischool.utexas.edu}
+\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
+%
+\thanks{{}$^{(9)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
 %
@@
 
-\note{Dan: I think there's some overlap
-  between this and the discussion in the Developing and Support
-  Software section.  Perhaps this should go in the cross-cutting
-  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
-\note{Ketan: yes, may be this section will split and merge into other sections}
+%\note{Dan: I think there's some overlap
+%  between this and the discussion in the Developing and Support
+%  Software section.  Perhaps this should go in the cross-cutting
+%  issues section?  Let's write it down here - then we can decide if it goes elsewhere}
+%\note{Ketan: yes, may be this section will split and merge into other sections}
 
@@
 expressed in this material are those of the author(s) and do not
-necessarily reflect the views of the National Science Foundation. HL
-was supported by the National Evolutionary Synthesis Center (NESCent), 
-NSF \#EF-0905606.
-
-Chue-Hong and Hetherington were supported by the UK Engineering and 
+necessarily reflect the views of the National Science Foundation.
+
+Lapp was supported by the National Evolutionary Synthesis Center (NESCent), 
+NSF EF-0905606.
+
+Hetherington was supported by the UK Engineering and 
 Physical Sciences Research Council (EPSRC) Grant EP/H043160/1 for the 

### 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
Refine abstract.

@@
 This document summarizes the First Workshop on Sustainable Software for Science:
-Practice and Experiences (WSSSPE1). The content is based on the workshop
-itself, as recorded in a set of collaborative notes taken during the workshop,
+Practice and Experiences (WSSSPE1). The content is based on 
+a set of collaborative notes taken during the workshop,
 including discussion about the keynote presentations, three themed panels that
@@
 understand what drives software work in science and how the reward systems of
-science work, thereby shaping the type of software work undertaken, including the
+science function, thereby shaping the types of software projects undertaken, including the
 extent to which developers are motivated to build software for the long-term,
@@
 Unique perspectives were captured about issues such as comprehensive documentation, 
-software deployment, and development practices; software licenses; career paths
+development and deployment practices; software licenses; career paths
 for developers; identifiers for software through a
@@
 skills, and in particular a specialized class of software developers is
-emerging in academic environments who are an integral and embedded part of
+emerging in academic environments as an integral part of
 successful research teams. The WSSSPE workshop was intended to provide a forum

### 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 1. Add non-breaking spaces. A minor change.

@@
 discussion, and the workshop has fed into the collaborative writing of this
-document. An estimate of 90 to 150 participants were present
+document. Estimates of 90 to 150 participants were present
 throughout the day. Additional papers based on extended versions of workshop
@@
 during the workshop~\cite{WSSSPE1-google-notes}, including discussion
-about the keynote presentations (\S\ref{sec:keynotes}), the three
-panels (\S\ref{sec:devel}-\ref{sec:community}), 
-cross-cutting issues (\S\ref{sec:cross-cutting}),
-use cases (\S\ref{sec:use-cases}), and conclusions
-(\S\ref{sec:conclusions}).  Additionally, the accepted workshop papers are listed in
+about the keynote presentations~(\S\ref{sec:keynotes}), the three
+panels~(\S\ref{sec:devel}-\ref{sec:community}), 
+cross-cutting issues~(\S\ref{sec:cross-cutting}),
+use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).  
+Additionally, the accepted workshop papers are listed in
 Appendix~\ref{sec:papers}, and a partial list of workshop attendees is

### 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
Adding prose about resources listed.

--- 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 descriptive and prescriptive approaches to communities.
-\begin{itemize}
-\item \cite{howison_scientific_2011, howison_incentives_2013}
-\item \cite{citeulike:11831265}
-\item \cite{citeulike:7888211}
-\item \cite{citeulike:478633}
-\item \cite{opac-b1134063}
-\item \cite{Trapani:2011}
-\item \cite{Allsopp:2012}
-\item \cite{1749-4699-6-1-015010}
-\item \cite{Turk:2013:SCH:2484762.2484782}
-\end{itemize}
+
+Both \cite{howison_scientific_2011} and \cite{howison_incentives_2013} 
+study how scientific software collectives self-organize and address issues of
+incentive, long-term support, and development of infrastructure as well as new
+features.  As noted elsewhere in this summary,
+\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
+communities in astrophysics.
+
+From the perspective of developing prescriptions for successful scientific
+software development, both \cite{citeulike:11831265} and
+\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
+sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
+for developing open source scientific software, focusing on both the choices
+made during development and the sustainability of practices in the long term.  
+\cite{1749-4699-6-1-015010} describes the development and long-term growth of
+the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
+applications and domains has shaped its development and community trajectory.
+
+From more traditional open source development, resources were shared that
+developed communities explicitly, such as 
+\cite{citeulike:7888211} and \cite{Trapani:2011}, focusing on
+large-scale projects such as the Ubuntu linux distribution and smaller-scale
+volunteer-developed projects like such as ThinkUp, respectively.
+The process of open source development, while less explicitly focused on
+community building, outlined in \cite{citeulike:478633} was seen as a valuable
+resource, particularly when combined with the management and personal
+interaction techniques outlined in \cite{opac-b1134063}.  Growing diversity in
+communities was directly addressed in \cite{Allsopp:2012}, where experiences
+growing the diversity of technical conferences in open source were described.
 

### 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
Added affiliation information for myself
Summarizing education and training papers.

@@
 
-\todo{text about papers and discussion in this part of the workshop}
+The papers describing education and training were focused primarily on how
+these aspects of community development impact on the long-term sustainability
+of software projects.  \cite{Girotto_WSSSPE} described the impact of the
+mandate within the International Centre for Theoretical Physics (ICTP) to
+foster resources and competences in software development and HPC, resulting in
+the development of educational curricula directed in this area.  The paper
+itself described the changes made in these curricula as a result of the current
+changes in the HPC and scientific software landscape due to the advent of
+scripting languages, new programming paradigms and new types of hardware such
+as accelerator technologies.  The development of a workshop, with carefully
+selected participants and an immersive approach to learning, was identified as
+a major success for educating and developing new scientific software developers
+from targeted domains.
+
+\cite{Crawford_WSSSPE} addressed education and training within computational
+chemistry frameworks, particularly as these frameworks attempt to address
+next-generation computer hardware and software and as chemistry courses
+emphasize lab work over computational education.  The authors identify this
+lack of computational awareness and training as the primary challenge to future
+advances in computational chemistry.  The authors propose a new institute for
+computational chemistry, emphasizing collaboration (and a licensing structure,
+such as LGPL or more permissive) and education of future generations of
+chemistry researchers.
 

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Fine tune sections 1, 2, 3.

@@
 York, NY, USA; \url{matthewturk@gmail.com}}
+%
+\thanks{{}$^{(6)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
 %
@@
 (SIAM) Annual Meeting on ``Reliable
-Computational Science.''
+Computational Science'' (SIAM AN14, \url{http://meetings.siam.org}).
 
@@
 The workshop itself consisted of two keynote presentations and the
-three panels/discussion sessions. The panels were organized based on a
+three panels with discussion sessions. The panels were organized based on a
 classification of the workshop submissions into three categories,
@@
 
-Another problem he experienced was staff retention with respect to
+Another problem Bourne experienced was staff retention with respect to
 private organizations which reward those combining research and
@@
 
-He also had problems with selling research software: the technology
+Bourne also had problems with selling research software: the technology
 transfer office in his university wanted huge intellectual property

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Fine tune sections 3 and 4.

@@
 
-He concluded with a reference to his paper~\cite{bourne_ten} and
+Bourne concluded with a reference to his paper~\cite{bourne_ten} and
 argued that computational scientists ``have a responsibility to
@@
 
-In conclusion Smith reiterated the importance of establishing
+In conclusion Smith reiterated the importance of establishing effective
 incentive models for open contributions and tool builders, for
@@
 Several participants added that documentation is not just for users, and writing
-API documentation, tutorials for building and deploying software, along with
+API documentation, tutorials for building and deploying software, together with
 documented development practices can be very helpful in bringing  new developers
@@
 
-Recruiting and/or retaining personnel in this area is hard,
+Recruiting and retaining personnel in this area is hard,
 with one of the major reasons being no long-term career paths (especially when
-compared with industry). This is further discussed in \S\ref{sec:career-tracks}.
+compared with industry). This is further discussed in~\S\ref{sec:career-tracks}.
 
@@
 The question of what ``sustainable'' means in the context of software was
-raised (see \S\ref{sec:defining-sustainability}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
+raised (see~\S\ref{sec:defining-sustainability}). The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
 to look for when choosing software for procurement, or reuse in further software
@@
 future? Important questions include whether the supplier will still be around in five
-years time, will it still care for its customers needs, will it be responsive to
+years' time, will it still care for its customers' needs, will it be responsive to
 bug reports and feature requests? Should you tie your investment to a single
@@
 some resistance, with so many compelling alternatives available, e.g. Bitbucket or
-GitHub. The point for centralization of communication/point of contact was seen as
+GitHub. The centralized communication or point of contact was seen as
 reasonable, with the statement that ``vested with authority'' is perhaps too strong,

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex
Fine tune section 4.
Small changes to first half of paper.

--- 2014-04-12 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 This document summarizes the First Workshop on Sustainable Software for Science:
-Practice and Experiences (WSSSPE1). The content is based on 
-a set of collaborative notes taken during the workshop,
+Practice and Experiences (WSSSPE1). The content is based on the workshop
+itself, as recorded in a set of collaborative notes taken during the workshop,
 including discussion about the keynote presentations, three themed panels that
-discussed papers accepted at the workshop, other discussion, cross-cutting
+discussed papers accepted at the workshop`, other discussion, cross-cutting
 issues, use cases, and conclusions. It joins a nascent literature seeking to
 understand what drives software work in science and how the reward systems of
-science function, thereby shaping the types of software projects undertaken, including the
+science work, thereby shaping the type of software work undertaken, including the
 extent to which developers are motivated to build software for the long-term,
@@
 Unique perspectives were captured about issues such as comprehensive documentation,
-development and deployment practices; software licenses; career paths
+software deployment, and development practices; software licenses; career paths
 for developers; identifiers for software through a
@@
 skills, and in particular a specialized class of software developers is
-emerging in academic environments as an integral part of
-successful research teams. The WSSSPE workshop was intended to provide a forum
-for discussion of the challenges, including both positions and experiences. The
-short papers and debates have been archived to provide a basis for continued
-discussion, and the workshop has fed into the collaborative writing of this
-document. Estimates of 90 to 150 participants were present
-throughout the day. Additional papers based on extended versions of workshop
-submissions are expected. The level of interest in the workshop has led the
-organizers, working with some of the submitters and attendees, to plan two
-additional workshops (at the 2014 SciPy and SC14 conferences), to obtain
-funds from the National Science Foundation and the Gordon and Betty Moore
-Foundation to support these workshops, and to turn the
-workshop website into a community website that can be used as a focus for
-further discussion and progress. Additionally, a minisymposium that aims to
+emerging in academic environments who are an integral and embedded part of
+successful research teams. The WSSSPE workshop  provided a forum
+for discussion of the challenges around scientific software, including both positions and experiences. The
+workshop short papers and debates have been archived to provide a basis for continued
+discussion, and have fed into the collaborative writing of this
+document. Additional papers based on extended versions of workshop
+submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved 
+an estimated 90 to 150 participants during the day has led to further activities:
+Two additional WSSSPE workshops are planned  (at the 2014 SciPy and SC14 conferences), 
+funds to support the workshops have been obtained from 
+the National Science Foundation and the Gordon and Betty Moore
+Foundation, and the
+original workshop website has been turned into a community website for
+further discussion and progress. Additionally, a minisymposium  to
 further explore some of the key issues raised in WSSSPE is co-organized by a
@@
 
-The goal of this paper is to summarize the contents of the workshop papers
-and discussion, and identify issues that future WSSSPE events will address.
+This paper summarizes the  workshop papers
+and discussion, and identifies issues that future WSSSPE events will address.
 Before the WSSSPE1 workshop took place, the organizers self-published
@@
 during the workshop~\cite{WSSSPE1-google-notes}, including discussion
-about the keynote presentations~(\S\ref{sec:keynotes}), the three
-panels~(\S\ref{sec:devel}-\ref{sec:community}), 
-cross-cutting issues~(\S\ref{sec:cross-cutting}),
-use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).  
-Additionally, the accepted workshop papers are listed in
+about the keynote presentations (\S\ref{sec:keynotes}), the three
+panels (\S\ref{sec:devel}-\ref{sec:community}), 
+cross-cutting issues (\S\ref{sec:cross-cutting}),
+use cases (\S\ref{sec:use-cases}), and conclusions
+(\S\ref{sec:conclusions}).  Additionally, the accepted workshop papers are listed in
 Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
@@
 that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. The papers were
+that the reports were already self-published. Based on the submissions, the papers were
 grouped into three main categories, namely \emph{Developing and
 Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-subject area was associated with a panel and discussion at the
-workshop.
-
-The workshop itself consisted of two keynote presentations and the
-three panels/discussion sessions. The panels were organized based on a
-classification of the workshop submissions into three categories,
-following the themes of the call for papers as modified by the areas
-of the submissions. Three to four representatives from each submission
+category was associated with a panel and discussion at the
+workshop.  Three to four representatives from each submission
 category were appointed as panelists, and assigned to read a subset of
 the papers in that category and then discuss them in the panel.
+The workshop itself consisted of two keynote presentations followed by the 
+three panel/discussion sessions.
 
@@
 The first keynote~\cite{WSSSPE1-keynote1} was delivered by Philip
-E. Bourne of University of California, San Diego.  Bourne is a basic
-biomedical scientist but has also formed four software companies. He co-founded
+E. Bourne of University of California, San Diego.  Bourne is a 
+biomedical scientist who has also formed four software companies. He co-founded
 PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB
@@
 
-Bourne's keynote presentation was entitled ``A Recipe for Sustainable
-Software,'' and developed based on his experiences.  He emphasized that
+Bourne's  presentation was based on his own software experiences.  He emphasized that
 sustainability for software ``does not just mean more money from
@@
 Bourne shared with the audience a few real challenges he
-encountered. First, also an anecdote, he has looked into
+encountered. His first anecdote was that he has looked into
 reproducibility in computational biology, but has concluded that ``I
@@
 
-He also had problems with selling research software: the technology
-transfer office in his university wanted huge intellectual property
+He also had problems with selling research software: the university technology
+transfer office wanted huge and unrealistic intellectual property
 reach through, whereby they would get a share of profits from drugs
-developed by pharmaceutical companies who use the software.  He was
-aware this was unrealistic but the technology transfer office insisted
-for a while. He wants to push a one-click approach for customers to
+developed by pharmaceutical companies who use the software. 
+He advocated for  a one-click approach for customers to
 purchase university-written software.
@@
 a research output alongside papers, a common discussion within this field.
-He mentioned an interesting exploration of involving software
-engineers in the review process of scientific code~\cite{peer-review-code}.
+He mentioned an  exploration of involving software
+engineers in the review process of scientific code~\cite{peer-review-code}, and discussed 
+how publishing software reviews could change attitudes.
 
@@
 to many useful organizational innovations: ``University 2.0 is yet to
-happen.'' He spoke of a circumstance where someone had used an
-algorithm developed for computational biology in marketing.
-The role of an institution is important in this space. He argued that
-funders can help train institutions, not just individuals, in this regard.
-
-He concluded with a reference to his paper~\cite{bourne_ten} and
+happen.''  He argued that
+funders such as NSF and NIH can help train institutions, not just individuals, in this regard.
+
+He concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
 argued that computational scientists ``have a responsibility to
@@
 scholarship, frequently
-more valuable than a research article,'' a point with which all authors
-strongly agree.
+more valuable than a research article''.
 
@@
 The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
-Smith of GitHub.
-
-Smith's talk started with an example from data reduction in Astronomy,
-where he needed to remove interfering effects from the device. He
+Smith of GitHub. Smith's talk started with an example from data reduction in Astronomy,
+where he needed to remove interfering effects from the device (XXXX? which device). He
 built a ``bad pixel mask,'' and realized that while it was persistent,
@@
 estimated that 13 person-years were wasted in this redundant
-calculation.
+calculation. (XXXX whole paragraph not clear)
 
@@
 research. He then asked, when we do know better, why we do not act any
-different. He argued that it was due to incentives and their lack:
+different. He argued that it was due to the lack of incentives:
 only the immediate products of research, not the software, are valued.
@@
 
-C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor and participant,
+C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor,
 argued that with regard to reusable software, ``we should just start
@@
 The question of what ``sustainable'' means in the context of software was
-raised (see \S\ref{sec:defining-sustainability}.) The resources at \url{http://oss-watch.ac.uk/resources/ssmm} discuss what
+raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch}   discuss what
 to look for when choosing software for procurement, or reuse in further software
@@
 
-The papers submitted to this panel, and several others, include lots of general
+The papers submitted to this panel included many general
 recommendations for processes, practices, tools, etc. One of the papers
@@
 \subsection{Training Scientists to Develop and Support Software}
-
-Some of the questions raised after the panel discussion are:
-
-Rather than teaching developers about domain science, or domain scientists about
-software development practices, why don't we teach both communities to
-collaborate more effectively? Can this be done without teaching each side a
-little of the other to enable communication, with a response that it really is
-not two binary communities, but a spectrum with useful roles in the middle.
-The question was raised if a developer can be effective without being part of the
-domain community, with responses that this really depends on the specific
-problem---translators and T-shaped people are important. Why aren't academic
-communities taught how to evaluate cross-disciplinary work well?
-
-What is the role for the growing field of team science? There
-is overlap between the communities, with support for virtual organizations,
-tool development, etc. How can we make time in an already crowded schedule to
-introduce these topics to students? Should they be introduced through lab
-classes as in ``real sciences''?
+Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work. 
+One question raised was whether a software developer can be effective if they are not part
+of the appropriate domain community, with responses that this depends on the 
+specific problem and situation, and that "T-shaped" people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
+The discussion focused on  whether we could teach software developers and domain scientists to 
+collaborate together more effectively rather than trying to teach software developers
+about domain science and domain scientists about software development practices. 
+The end goal of this would be to have a single community with a spectrum of expertise across
+domain science and software development rather than two separate communities.
+
+The role of the growing field of team science with software development  was discussed. The science of team science
+deals with understanding and improving collaborative and team-based scientific research, and such issues as virtual organizations, 
+and tool development cut across software development communities. Further, how should these skills and best practices then be 
+introduced to students?
 
@@
 to rot and the experiments to become non-reproducible as the tools change.
-Jason Riedy stated that he had been guilty of this in the past. There was
+ There was
 discussion of this tendency along with the enormous differences in the speed and
@@
 
-The Policy panel was designated to discuss the group of workshop
+The  panel on policy  discussed workshop
 contributions dealing with the wide range of software sustainability
@@
 in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
-Kneply et al. argued that the visibility of software that supported a
+Knepley et al. argued that the lack of visibility of software that supported a
 piece of science ``can have detrimental effects on funding, future
@@
 descriptive and prescriptive approaches to communities. 
-
-Both \cite{howison_scientific_2011} and \cite{howison_incentives_2013} 
-study how scientific software collectives self-organize and address issues of
-incentive, long-term support, and development of infrastructure as well as new
-features.  As noted elsewhere in this summary,
-\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
-communities in astrophysics.
-
-From the perspective of developing prescriptions for successful scientific
-software development, both \cite{citeulike:11831265} and
-\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
-sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
-for developing open source scientific software, focusing on both the choices
-made during development and the sustainability of practices in the long term.  
-\cite{1749-4699-6-1-015010} describes the development and long-term growth of
-the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
-applications and domains has shaped its development and community trajectory.
-
-From more traditional open source development, resources were shared that
-developed communities explicitly, such as 
-\cite{citeulike:7888211} and \cite{Trapani:2011}, focusing on
-large-scale projects such as the Ubuntu linux distribution and smaller-scale
-volunteer-developed projects like such as ThinkUp, respectively.
-The process of open source development, while less explicitly focused on
-community building, outlined in \cite{citeulike:478633} was seen as a valuable
-resource, particularly when combined with the management and personal
-interaction techniques outlined in \cite{opac-b1134063}.  Growing diversity in
-communities was directly addressed in \cite{Allsopp:2012}, where experiences
-growing the diversity of technical conferences in open source were described.
+\begin{itemize}
+\item \cite{howison_scientific_2011, howison_incentives_2013}
+\item \cite{citeulike:11831265}
+\item \cite{citeulike:7888211}
+\item \cite{citeulike:478633}
+\item \cite{opac-b1134063}
+\item \cite{Trapani:2011}
+\item \cite{Allsopp:2012}
+\item \cite{1749-4699-6-1-015010}
+\item \cite{Turk:2013:SCH:2484762.2484782}
+\end{itemize}
 
@@
 
-The papers describing education and training were focused primarily on how
-these aspects of community development impact on the long-term sustainability
-of software projects.  \cite{Girotto_WSSSPE} described the impact of the
-mandate within the International Centre for Theoretical Physics (ICTP) to
-foster resources and competences in software development and HPC, resulting in
-the development of educational curricula directed in this area.  The paper
-itself described the changes made in these curricula as a result of the current
-changes in the HPC and scientific software landscape due to the advent of
-scripting languages, new programming paradigms and new types of hardware such
-as accelerator technologies.  The development of a workshop, with carefully
-selected participants and an immersive approach to learning, was identified as
-a major success for educating and developing new scientific software developers
-from targeted domains.
-
-\cite{Crawford_WSSSPE} addressed education and training within computational
-chemistry frameworks, particularly as these frameworks attempt to address
-next-generation computer hardware and software and as chemistry courses
-emphasize lab work over computational education.  The authors identify this
-lack of computational awareness and training as the primary challenge to future
-advances in computational chemistry.  The authors propose a new institute for
-computational chemistry, emphasizing collaboration (and a licensing structure,
-such as LGPL or more permissive) and education of future generations of
-chemistry researchers.
+\todo{text about papers and discussion in this part of the workshop}
 

### 2014-04-13 WSSSPE1_report/WSSSPE1_summary_report.tex

@@
 %
+\thanks{{}$^{(6)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
+%
 \thanks{{}$^{(7)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
@@
 skills, and in particular a specialized class of software developers is
-emerging in academic environments who are an integral and embedded part of
+emerging in academic environments as an integral part of
 successful research teams. The WSSSPE workshop  provided a forum
@@
 
-He also had problems with selling research software: the university technology
+Bourne also had problems with selling research software: the university technology
 transfer office wanted huge and unrealistic intellectual property
@@
 
-He concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
+Bourne concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
 argued that computational scientists ``have a responsibility to
@@
 
-There was debate that while some of the basics of computing may be fairly
+There was debate that while some of the basics of computing may be 
 similar tools commonly used for computing have changed quite

### 2014-04-14 WSSSPE1_report/WSSSPE1_summary_report.tex
Acknowledgement.
Fix et al.

@@
 stakeholders requirements, technology and environments evolve and
-change.'' Venters et al. argue that these choices are really made very
+change.'' Venters et al.~argue that these choices are really made very
 early because they relate to the architecture of the software and
@@
 attributes of the software artifact. Lenhardt et
-al~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
+al.~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
 lifecycle to argue for the inclusion of metadata throughout a piece of
@@
 in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
-Knepley et al. argued that the lack of visibility of software that supported a
+Knepley et al.~argued that the lack of visibility of software that supported a
 piece of science ``can have detrimental effects on funding, future
@@
 infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
-et al. observe, unlike for physical IT infrastructure in data centers,
+et al.~observe, unlike for physical IT infrastructure in data centers,
 science communities have developed very little in the way of practices and
-technology for API governance, referred to by Krintz et al. as the
+technology for API governance, referred to by Krintz et al.~as the
 ``combined policy, implementation, and deployment control''. Web APIs
@@
 organizations to assess and thus properly manage the impact of API
-changes. To address this, Krintz et al. propose a mechanism that
+changes. To address this, Krintz et al.~propose a mechanism that
 evaluates the porting effort between two versions of a web-service API
@@
 
+Choi thanks Ian Foster, Director of the Computation Institute, University of Chicago,
+for encouraging her work in reliable reproducible research and 
+supporting her trip to participate in the WSSSPE1 conference as a contributor.
+
 \appendix

### 2014-04-14 WSSSPE1_report/WSSSPE1_summary_report.tex
Add non-breaking spaces and other minor changes.

@@
 of the appropriate domain community, with responses that this depends on the 
-specific problem and situation, and that "T-shaped" people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
+specific problem and situation, and that ``T-shaped'' people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important. 
 The discussion focused on  whether we could teach software developers and domain scientists to 
@@
 There was debate that while some of the basics of computing may be 
-similar tools commonly used for computing have changed quite
+similar, tools commonly used for computing have changed quite
 significantly. Reference was made to Perl, which was commonly used, giving way
@@
 users~\cite{Pierce_WSSSPE}.  For a more thorough discussion of definitions
-of software sustainability see \S\ref{sec:defining-sustainability}.
+of software sustainability see~\S\ref{sec:defining-sustainability}.
 
@@
 project organizations (and presumably artifact attributes) are
-necessary to persist longterm in spaces with particular
+necessary to persist long term in spaces with particular
 characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
@@
 particular piece of software that was perhaps developed by another group.  As
-an example, in \cite{Hanwell_WSSSPE}, the community described is comprised of
+an example, in~\cite{Hanwell_WSSSPE}, the community described is comprised of
 both the for-profit company \toolname{Kitware} and the users and contributors to their
@@
 infrastructure and developing applications utilizing it, was similarly
-reflected in \cite{Terrel_WSSSPE}, where the core scientific python ecosystem
+reflected in~\cite{Terrel_WSSSPE}, where the core scientific python ecosystem
 is supported by a non-profit entity that fosters community investment in that
@@
 
-Both \cite{howison_scientific_2011} and \cite{howison_incentives_2013} 
+Both~\cite{howison_scientific_2011} and~\cite{howison_incentives_2013} 
 study how scientific software collectives self-organize and address issues of
@@
 From the perspective of developing prescriptions for successful scientific
-software development, both \cite{citeulike:11831265} and
+software development, both~\cite{citeulike:11831265} and
 \cite{1749-4699-6-1-015010} share experiences and suggestions for developing
@@
 From more traditional open source development, resources were shared that
-developed communities explicitly, such as 
-\cite{citeulike:7888211} and \cite{Trapani:2011}, focusing on
+developed communities explicitly, such as~\cite{citeulike:7888211} 
+and~\cite{Trapani:2011}, focusing on
 large-scale projects such as the Ubuntu linux distribution and smaller-scale
@@
 The process of open source development, while less explicitly focused on
-community building, outlined in \cite{citeulike:478633} was seen as a valuable
+community building, outlined in~\cite{citeulike:478633} was seen as a valuable
 resource, particularly when combined with the management and personal
-interaction techniques outlined in \cite{opac-b1134063}.  Growing diversity in
-communities was directly addressed in \cite{Allsopp:2012}, where experiences
+interaction techniques outlined in~\cite{opac-b1134063}.  Growing diversity in
+communities was directly addressed in~\cite{Allsopp:2012}, where experiences
 growing the diversity of technical conferences in open source were described.

### 2014-04-14 WSSSPE1_report/WSSSPE1_summary_report.tex
Fix typos.

@@
 implementation strategies for science from the perspective of
-facilitating many or all facets of sustainanability-oriented software
+facilitating many or all facets of sustainability-oriented software
 design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast
@@
 ``combined policy, implementation, and deployment control''. Web APIs
-can and do change, sometimes quite frequently. Raising the need to
+can and do change, sometimes quite frequently, raising the need to
 port dependent applications. The effort required for porting is
@@
 engineering, dependency management, and deployment operations. Each of
-these facets require specialized skills and training in software
+these facets requires specialized skills and training in software
 engineering. How can the implementation of best practices and

### 2014-04-14 WSSSPE1_report/WSSSPE1_summary_report.tex
fix some typos

@@
 
-\author{Daniel~S.~Katz$^{(1)}$, Hilmar~Lapp$^{(2)}$, Ketan~Maheshwari$^{(3)}$, Frank~L\"{o}ffler$^{(4)}$, 
+\author{Daniel~S.~Katz$^{(1)}$, Hilmar~Lapp$^{(2)}$, Ketan~Maheshwari$^{(3)}$, Frank~Löffler$^{(4)}$, 
 Matthew~Turk$^{(5)}$, Marcus~D.~Hanwell$^{(6)}$, Nancy~Wilkins-Diehr$^{(7)}$, James~Hetherington$^{(8)}$, 
@@
 
-The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
-Smith of GitHub. Smith's talk started with an example from data reduction in Astronomy,
-where he needed to remove interfering effects from the device (XXXX? which device). He
-built a ``bad pixel mask,'' and realized that while it was persistent,
-there was no way or practice of sharing this with the
-data. Consequently many researchers repeated the same calculations. He
-estimated that 13 person-years were wasted in this redundant
-calculation. (XXXX whole paragraph not clear)
+The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon Smith of
+GitHub. Smith's talk started with an example from his own past in data
+reduction in Astronomy, where he needed to remove interfering effects from the
+experimental apparatuses. He built a ``bad pixel mask,'' and realized that
+while it was persistent, there was no way or practice of sharing these data
+among the scientists. Consequently many researchers repeated the same
+calculations. Smith estimated that 13 person-years were wasted by this
+repetition.
 
@@
 
-Priem and Piwowar argued that assessing the impact of software work
-requires looking beyond publications, including evidence of contribution and impact recorded in social
-coding-oriented resources such as Github, and conversations about software
-in issue trackers, mailing lists, twitter and
-beyond~\cite{Priem_WSSSPE}. In keeping with a principle of
-the ``altmetrics'' approach, they advocate that scholars should have
-resources that empower them to tell their own stories in the
-manner most appropriate for them and their audiences.
+Priem and Piwowar argued that assessing the impact of software work requires
+looking beyond publications, including evidence of contribution and impact
+recorded in social coding-oriented resources such as GitHub, and conversations
+about software in issue trackers, mailing lists, twitter and
+beyond~\cite{Priem_WSSSPE}. In keeping with a principle of the ``altmetrics''
+approach, they advocate that scholars should have resources that empower them
+to tell their own stories in the manner most appropriate for them and their
+audiences.
 
@@
 
-Christopherson reports on the application of Steps 1-3 to a
-computational modeling framework developed in the 1990s. Step 1 was
-a 2-day, in-person specifications meeting and code walkthrough. Step 2 was a
-5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
-the code based on new requirements. The team worked on small, low-risk units of
-code. It was challenging, revealed unanticipated obstacles, programmers had to
-work together, and experimentation was encouraged.
+Christopherson reports on the application of Steps 1-3 to a computational
+modeling framework developed in the 1990s. Step 1 was a 2-day, in-person
+specifications meeting and code walk-through. Step 2 was a 5-day hackathon to
+develop working code, and Step 3 was a 3-day hackathon to refine the code based
+on new requirements. The team worked on small, low-risk units of code. It was
+challenging, revealed unanticipated obstacles, programmers had to work
+together, and experimentation was encouraged.
 %
 The paper recommended: start small and gradually building toward more complex
-objectives, consistent with Agile development; develop consensus before
-coding, by repeating step 1 before all hackathons; ensure newcomers
-receive orientation prior to the hackathon, such as a code
-walkthrough or system documentation; and co-locate collaborators
-whenever feasible.
+objectives, consistent with Agile development; develop consensus before coding,
+by repeating step 1 before all hackathons; ensure newcomers receive orientation
+prior to the hackathon, such as a code walk-through or system documentation;
+and co-locate collaborators whenever feasible.
 
@@
 developed communities explicitly, such as~\cite{citeulike:7888211}
-and~\cite{Trapani:2011}, focusing on
-large-scale projects such as the Ubuntu linux distribution and smaller-scale
-volunteer-developed projects like such as ThinkUp, respectively.
-The process of open source development, while less explicitly focused on
-community building, outlined in~\cite{citeulike:478633} was seen as a valuable
-resource, particularly when combined with the management and personal
-interaction techniques outlined in~\cite{opac-b1134063}.  Growing diversity in
-communities was directly addressed in~\cite{Allsopp:2012}, where experiences
-growing the diversity of technical conferences in open source were described.
+and~\cite{Trapani:2011}, focusing on large-scale projects such as the Ubuntu
+Linux distribution and smaller-scale volunteer-developed projects like such as
+ThinkUp, respectively. The process of open source development, while less
+explicitly focused on community building, outlined in~\cite{citeulike:478633}
+was seen as a valuable resource, particularly when combined with the management
+and personal interaction techniques outlined in~\cite{opac-b1134063}.  Growing
+diversity in communities was directly addressed in~\cite{Allsopp:2012}, where
+experiences growing the diversity of technical conferences in open source were
+described.
 
@@
 and the original cited papers for the respective software. Where available, we
-looked at the webpages and source code of the earliest versions to find the
+looked at the web pages and source code of the earliest versions to find the
 original developers. We included projects older than 5 years and excluded
-relatively new projects such as \toolname{SciPy} and \toolname{numFOCUS} from this
-survey. we find that in a majority of scientific software more than one
+relatively new projects such as \toolname{SciPy} and \toolname{numFOCUS} from
+this survey. we find that in a majority of scientific software more than one
 original developers are still around, in some cases for more than 10 years,

### 2014-04-15 count.sh
Script for counting files, insertions, deletions per author.

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+# Example:
+#   ./count.sh "ketancmaheshwari"
+# returns
+#   56, 1030, 787
+# 
+# of files, insertions, deletions, respectively
+#
+# Cf. 
+#   git shortlog -s -n
+# only returns number of files
+  
+for i in "Katz" "ketancmaheshwari" "Terrya" "Lapp" "Loeffler" "Turk" "jameshowison" "Hanwell" "Frank" "wilkins" "Wilkins" "Hetherington" "Allen" "shelswenson"
+do
+	git log --author="$i" --shortstat $BRANCH | \
+	awk '/^ [0-9]/ { f += $1; i += $4; d += $6 } \
+	END { printf("%d, %d, %d\n", f, i, d) }'
+done
+

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
small fix in intro, fix of cite later

@@
 The workshop itself consisted of two keynote presentations followed by the 
-three panel/discussion sessions.
+three panel/discussion sessions.  Notes were collaboratively taken
+by the group during the discussion, using a web document~\cite{WSSSPE1-google-notes}.
 
@@
 The question of what ``sustainable'' means in the context of software was
-raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch}   discuss what
+raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch-web}   discuss what
 to look for when choosing software for procurement, or reuse in further software
@@
 The following is a partial list of attendees who were recorded on the
-Google doc~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, or by the SC13 student volunteers, with some additions also made by the authors of this report.
+document~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, or by the SC13 student volunteers, with some additions also made by the authors of this report.
 

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
changing author order to match lines committed

@@
 
-\author{Daniel~S.~Katz$^{(1)}$, Hilmar~Lapp$^{(2)}$, Ketan~Maheshwari$^{(3)}$, Frank~Löffler$^{(4)}$, 
-Matthew~Turk$^{(5)}$, Marcus~D.~Hanwell$^{(6)}$, Nancy~Wilkins-Diehr$^{(7)}$, James~Hetherington$^{(8)}$, 
-James~Howison$^{(9)}$, Sou-Cheng~T.~Choi$^{(10)}$, Shel~Swenson$^{(11)}$,  Bruce~Berriman$^{(12)}$, Andy~R.~Terrel$^{(13)}$, Colin~Venters$^{(14)}$}
+\author{Daniel~S.~Katz$^{(1)}$,
+Sou-Cheng~T.~Choi$^{(2)}$,
+Hilmar~Lapp$^{(3)}$,
+Ketan~Maheshwari$^{(4)}$,
+Frank~L\:{o}ffler$^{(5)}$, 
+Matthew~Turk$^{(6)}$,
+Marcus~D.~Hanwell$^{(7)}$,
+Nancy~Wilkins-Diehr$^{(8)}$,
+James~Hetherington$^{(9)}$, 
+James~Howison$^{(10)}$,
+Shel~Swenson$^{(11)}$, 
+Gabrielle~D.~Allen$^{(12)}$,
+Bruce~Berriman$^{(13)}$,
+Andy~R.~Terrel$^{(14)}$,
+Colin~Venters$^{(15)}$
+}
 
@@
 %
-\thanks{{}$^{(2)}$ National Evolutionary Synthesis Center (NESCent),
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+%
+\thanks{{}$^{(3)}$ National Evolutionary Synthesis Center (NESCent),
   Durham, NC, USA; \url{hlapp@nescent.org}}
 %
-\thanks{{}$^{(3)}$ Argonne National Laboratory, Chicago, IL, USA; \url{ketan@anl.gov}}
-%
-\thanks{{}$^{(4)}$ Center for Computation and Technology, Louisiana State University, Baton Rouge, LA, USA; \url{knarf@cct.lsu.edu}}
-%
-\thanks{{}$^{(5)}$ Columbia Astrophysics Laboratory, Columbia University, New
+\thanks{{}$^{(4)}$ Argonne National Laboratory, Chicago, IL, USA; \url{ketan@anl.gov}}
+%
+\thanks{{}$^{(5)}$ Center for Computation and Technology, Louisiana State University, Baton Rouge, LA, USA; \url{knarf@cct.lsu.edu}}
+%
+\thanks{{}$^{(6)}$ Columbia Astrophysics Laboratory, Columbia University, New
 York, NY, USA; \url{matthewturk@gmail.com}}
 %
-\thanks{{}$^{(6)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
-%
-\thanks{{}$^{(7)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
-%
-\thanks{{}$^{(8)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
-%
-\thanks{{}$^{(9)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
-%
-\thanks{{}$^{(10)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+\thanks{{}$^{(7)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
+%
+\thanks{{}$^{(8)}$ San Diego Supercomputer Center, University of California San Diego, La Jolla, CA, USA, \url{wilkinsn@sdsc.edu}}
+%
+\thanks{{}$^{(9)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
+%
+\thanks{{}$^{(10)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
 % 
@@
 \thanks{{}$^{(13)}$ xxx}
+%
+\thanks{{}$^{(14)}$ xxx}
 %

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Fix typo in a name

@@
 Ketan~Maheshwari$^{(4)}$,
-Frank~L\:{o}ffler$^{(5)}$, 
+Frank~Löffler$^{(5)}$, 
 Matthew~Turk$^{(6)}$,

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Shorten my aflliation.
Minor changes to Section 1.

@@
 %
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and the Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
 %
@@
 
-58 submissions were received, and almost all submitters used either
+In total, 58 submissions were received, and almost all submitters used either
 arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to self-publish
@@
 
-The organizers decided to list 54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
+The organizers decided to list~54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
 contributions to the workshop, a very high acceptance rate but one

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Eliminate a title in affliction.
Eliminate street in affliction.

@@
 %
-\thanks{{}$^{(7)}$ Scientific Computing Group, Kitware, Inc., 28 Corporate Drive, Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
+\thanks{{}$^{(7)}$ Scientific Computing Group, Kitware, Inc.,  Clifton Park, NY, USA; \url{marcus.hanwell@kitware.com}}
 %
@@
 %
-\thanks{{}$^{(9)}$ Leader, Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
+\thanks{{}$^{(9)}$ Research Software Development Team, University College London, London, UK; \url {j.hetherington@ucl.ac.uk}}
 %

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
Add non-breaking spaces and other minor changes.

@@
 \url{http://wssspe.researchcomputing.org.uk/WSSSPE1}) was held on
-Sunday, 17 November 2013, in conjunction with the 2013 International
+Sunday, 17 November~2013, in conjunction with the~2013 International
 Conference for High Performance Computing, Networking, Storage and
@@
 submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved 
-an estimated 90 to 150 participants during the day has led to further activities:
-Two additional WSSSPE workshops are planned  (at the 2014 SciPy and SC14 conferences), 
+an estimated~90 to~150 participants during the day has led to further activities:
+Two additional WSSSPE workshops are planned  (at the~2014 SciPy and SC14 conferences), 
 funds to support the workshops have been obtained from 
@@
 further explore some of the key issues raised in WSSSPE is co-organized by a
-WSSSPE1 participant at the 2014 Society for Industrial and Applied Mathematics
+WSSSPE1 participant at the~2014 Society for Industrial and Applied Mathematics
 (SIAM) Annual Meeting on ``Reliable
@@
 WSSSPE1 was organized in collaboration of the relatively small group
-of five organizers and a larger peer-review committee with 36
+of five organizers and a larger peer-review committee with~36
 members. This committee had early influence on the organization, e.g.,
@@
 A peer review process followed the submissions, where
-the 58 papers received 181 reviews, an average of 3.12 reviews per
+the~58 papers received~181 reviews, an average of~3.12 reviews per
 paper. Reviews were completed using a Google form, which allowed
@@
 Government'' (see also Section~\ref{sec:defining-sustainability}).  Other
-factors to consider, he mentioned, include costs of production, ease
+factors to consider, he mentioned, encompass costs of production, ease
 of maintenance, community involvement, and distribution channels.

### 2014-04-15 WSSSPE1_report/WSSSPE1_summary_report.tex
removing Andy Terrel as an author on his request

@@
 Bruce~Berriman$^{(13)}$,
-Andy~R.~Terrel$^{(14)}$,
-Colin~Venters$^{(15)}$
+Colin~Venters$^{(14)}$
 }
@@
 \thanks{{}$^{(13)}$ xxx}
-%
-\thanks{{}$^{(14)}$ xxx}
 %

### 2014-04-16 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 4: Minor changes.

@@
 Several participants added that documentation is not just for users, and writing
-API documentation, tutorials for building and deploying software, together with
+application programming interface (API) documentation, tutorials for building and deploying software, together with
 documented development practices can be very helpful in bringing  new developers
@@
 discussed. For software that builds on other software frameworks it is the
-API (application programming interface) that is most important, and this can
+API that is most important, and this can
 encompass issues such as the source and binary interfaces to the software
 libraries developed---with each potentially having a high maintenance cost if
-they are to remain compatible over many years. When using command line programs
-it is generally the command line switches as well as the input/output formats
+they are to remain compatible over many years. When using command-line programs
+it is generally the command-line switches as well as the input/output formats
 that could incur significant costs if they are changed.
@@
 Several of the submitted papers discussed how communities have been built around
-projects, and what is needed to enable a project to grow. Among these are public
-source code hosting, mailing lists, documentation, wikis, bug trackers, software
+projects, and what is needed to enable a project to grow. Among these are 
+public source-code hosting, mailing lists, documentation, wikis, bug trackers, software
 downloads, continuous integration, software quality dashboards, and of course,
@@
 repository, central resources for services and consulting, a think tank of
-sorts, and a software orphanage center. The idea of one common repository received
+sorts, and a software orphanage center. \note{What is software orphanage center?} The idea of one common repository received
 some resistance, with so many compelling alternatives available, e.g. Bitbucket or
@@
 Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work. 
-One question raised was whether a software developer can be effective if they are not part
+One question raised was whether software developers can be effective if they are not part
 of the appropriate domain community, with responses that this depends on the 

### 2014-04-16 WSSSPE1_report/WSSSPE1_summary_report.tex
work on conclusions

@@
 
-\todo{pull the discussion together}
-
-\todo{add some analysis of the deficiencies and difficulties that are
+\note{add some analysis of the deficiencies and difficulties that are
   present in different fields, and those that are common?}
 
-\todo{say something about licensing - lessons, advice, etc.?}
+\note{say something about licensing - lessons, advice, etc.?}
 
@@
 
+\todo{what are the main issues discussed in each discussion section?, and what are the main common issues from the cross-cutting section - Dan to do that latter part}
+
 \subsection{Recommendations or Lessons}
 
-\note{if needed.}
-
-\subsection{Follow up actions}
-
-\note{or at least the discussion about them, and the current plans for
-  future events.}
-
-\note{conclusions from pre-workshop paper follows}
-
-The WSSSPE workshop has begun an experiment in how we can
-collaboratively build a workshop agenda. However, contributors also
-want to get credit for their participation in the process. And the
-workshop organizers want to make sure that the workshop content and
-their efforts are recorded.  Ideally, there would be a service that
+\note{lessons from the 3 discussion sessions go here, and lessons from the cross-cutting section as well?}
+
+\subsubsection{Workshop process}
+
+The WSSSPE workshop can be viewed as an experiment in how we can
+collaboratively and inclusively build a workshop agenda, without asking
+large numbers of people to submit papers that will be rejected so that the
+workshop can have a low acceptance rate.
+
+Contributors also want to get credit for their participation in the process.
+And the workshop organizers want to make sure that the workshop content and
+their efforts are recorded.  The methods used in the WSSSPE1 workshop
+were successful; we had good participation, many contributors have a
+report they can cite, and the record of the workshop is open and available,
+through the self-published reports, the workshop web site and notes site,
+and this paper.  In addition, many additional papers are being created that
+will include the discussions at the workshop, including extended versions
+of many of the self published reports. such as those that are in this
+special issue.
+
+Ideally, there would be a service that
 would be able to index the contributions to the workshop, serving the
-authors, the organizers, and the larger community. But since there
-isn't such a service today, the workshop organizers are writing this
-initial report and making use of arXiv as a partial solution to
-provide a record of the workshop.
-
-After the workshop, one or more additional papers will be created that
-will include the discussions at the workshop. These papers will likely
-have many authors, and may be submitted to peer-reviewed journals.
-
-\katznote{discuss sustainability survey}
+authors, the organizers, and the larger community.
+
+\subsection{Future activities}
+
+The organizers of the workshop have begun a survey to understand how
+the community define software sustainability.  It is expected that this survey
+will gather one or more consensus definitions, and lead to a short paper
+discussing them, as well as the level of consensus.
+
+Additional activities that are being planned include
+two additional WSSSPE workshops at the 2014 SciPy and SC14 conferences.
+The SciPy workshop (WSSSPE1.1) will focus on how some software projects
+from the SciPy community have dealt with software sustainability issues, both
+successfully and unsuccessfully, while the SC14 workshop (WSSSPE2) will
+be more general, and will likely focus on determining and publicizing specific
+activities that the overall scientific software community can take to move forward.
+In addition, there will be a minisymposium on ``Reliable Computational Science''
+at the 2014 Society for Industrial and Applied Mathematics (SIAM)
+Annual Meeting (SIAM AN14, http://meetings.siam.org)
+to further explore some of the key issues raised in
+WSSSPE .
 

### 2014-04-17 WSSSPE1_report/WSSSPE1_summary_report.tex
changes to conclusions, mostly

@@
 
-\subsection{Career Tracks for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:career-tracks}
-
-Career track issues also came up repeatedly, starting in the the first keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the Google Bus'' to describe the issue of staff in his lab in San Diego who had chosen to leave science and work for Google (but since they didn't want to physically move, they take the Google bus to the Google office each day.)
+\subsection{Career Paths for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:career-tracks}
+
+Career path issues also came up repeatedly, starting in the the first keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the Google Bus'' to describe the issue of staff in his lab in San Diego who had chosen to leave science and work for Google (but since they didn't want to physically move, they take the Google bus to the Google office each day.)
 
@@
 And the possible career paths for a specialist software developer working as part of a scientific research group in a university are certainly not yet clear.
+
+\note{add a subsection about licensing - lessons, advice, etc.?}
 
@@
 
+To conclude, we discuss the major concerns and issues that were raised in
+the workshop, what we have learned, and what we plan to do going forward.
+
 \note{add some analysis of the deficiencies and difficulties that are
@@
 
-\note{say something about licensing - lessons, advice, etc.?}
-
-\subsection{Top Issues}
-
-\todo{what are the main issues discussed in each discussion section?, and what are the main common issues from the cross-cutting section - Dan to do that latter part}
-
-\subsection{Recommendations or Lessons}
+\subsection{Concerns and issues}
+
+\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}
+
+The issues that came up repeatedly in different parts of the workshop are, as described in
+\S\ref{sec:cross-cutting}, the need for a definition of sustainability, and the need for
+career paths for scientific software developers.
+
+\subsection{Recommendations or lessons}
 
@@
 
-\subsubsection{Workshop process}
+\subsubsection*{Workshop process}
 

### 2014-04-18 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 9: Adding a few best practices and issues from Section 3.

@@
 
+\subsubsection*{Developing and Supporting Software}
+
+\subsubsection*{Policy}
+
+\subsubsection*{Communities, Models, and Education}
+
+Recognition of scholarship in scientific software over research 
+articles is overdue.
+
 \subsection{Recommendations or lessons}
@@
 
+\subsubsection*{Developing and Supporting Software}
+
+Successful collaborative projects conducted via the use of shared
+repositories ensures reproducibility in computational science.
+
+\subsubsection*{Policy}
+
+\subsubsection*{Communities, Models, and Education}
+A sustainability model for scientific software is to build a pipeline
+from construction to consumption as found in the most efficient
+information technology enterprises.
+
 \subsubsection*{Workshop process}
-
 The WSSSPE workshop can be viewed as an experiment in how we can
@@
 at the 2014 Society for Industrial and Applied Mathematics (SIAM)
-Annual Meeting (SIAM AN14, http://meetings.siam.org)
+Annual Meeting (SIAM AN14, \url{http://meetings.siam.org})
 to further explore some of the key issues raised in
-WSSSPE .
+WSSSPE.
 

### 2014-04-20 WSSSPE1_report/WSSSPE1_summary_report.tex
changed a footnote to bib

@@
 Richard Millington's book ``Buzzing Communities'' does an excellent job of outlining these measures, drawing on online communities research in an accessible manner~\cite{millington_buzzing_2012}.
-Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}.
-\katznote{please change to cite}
+Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem~\cite{metrics-web}.
+%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please change to cite}
 

### 2014-04-20 WSSSPE1_report/WSSSPE1_summary_report.tex
concl on sec 8

@@
 much discussed issue of career paths for developers of these software, we
-conducted a small survey of the use-cases discussed above with respect to the
+conducted a small survey of the cases discussed above with respect to the
 current status of its core developers. We looked at the workshop submissions
@@
 relatively new projects such as \toolname{SciPy} and \toolname{numFOCUS} from
-this survey. we find that in a majority of scientific software more than one
-original developers are still around, in some cases for more than 10 years,
+this survey. We find that in a majority of scientific software, more than one
+original developer are still around, in some cases after more than 10 years,
 (eg. \toolname{R/qtl}).  Whereas, in the case of utility software projects, it
@@
 
+% Conclusion of S8 : Use Cases, not sure where to put
+\note{conclusions from S8, not sure of location --Ketan}
+
+Two distinct class of scientific software projects and products were discussed
+in the workshop: 1) generic, large-scale utility software and 2) niche, medium-
+and small-scale scientific software. Each class faces different and significant
+challenges. New undertakings should recognize the differences in advance and
+identify such challenges within the development and sustaining efforts. In
+particular, the dynamics associated with developers, scope, life cycle,
+users-community, (re)usability, funding support and career paths vary widely
+among the two classes of software.
+
+% end conclusion S8
+
 \subsubsection*{Workshop process}

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Removes sustainability defs from Sec 5.

@@
 extents.  Each paper in this group included a definition of sustainability,
-with many overlaps.  In essence the papers related the question to
-change over time, an idea well captured in the definition used by the
-UK's Scientific Software Sustainability Institute: ``software you use
-today will be available---and continue to be improved and
-supported---in the future,'' as quoted in~\cite{Venters_WSSSPE}. This
-idea was also expressed as software that continues to serve its
-users~\cite{Pierce_WSSSPE}.  For a more thorough discussion of definitions
-of software sustainability see~\S\ref{sec:defining-sustainability}.
+with many overlaps between them  (see Table~\ref{tab:defining-sustainability}).
+Perhaps unsurprisingly, the issue of how to define sustainability came
+to the fore multiple times during the workshop, and it is thus
+summarized in depth separately in ~\S\ref{sec:defining-sustainability}.
 

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Rework of cross-cutting issue sust. def.

@@
 
-Two issues were raised in a number of the workshops sessions, defining sustainability and career paths.  This section brings those discussions together.
+Two issues, how to define software sustainability, and how career
+paths (or their lack) interact with achieving it, were raised across
+the workshop's panels.  This section aims to synthesize these
+discussions from different perspectives.
 
@@
 
-%\note{So far this reads a lot like a collection of thoughts only, without ``red line''.}
-%\note{Dan: assuming the previous note is from FL, what do you propose?}
-The question of what was meant by ``sustainability'' in the context of software
-came up in many different parts of the workshop, specifically in the first
-keynote (\S\ref{sec:keynote1}), the Developing and Maintaining Software panel,
-and the Policy panel.
-
-In the opening keynote, Philip Bourne suggested that perhaps sustainability can
-be defined as the effort that happens to make the essential things continue.
-This leads to having to decide what it is that we want to sustain, whether what
-we want to sustain is valuable, and finally, who would care if it went away,
-and how one measures how much they care.
-
-In the Developing and Maintaining Software panel, there was some discussion on
-this topic: what does sustainability mean? It was pointed out that OSS Watch
-proposes a Software Sustainability Maturity Model to address the issue of how
-sustainability a particular element of software is, and says that this
-sustainability is important. ``When choosing software for procurement or
-development reuse - regardless of the license and development model you will
-use - you need to consider the future. While a software product may satisfy
+What is, or should be meant by ``sustainability'' in the context of
+software came up in many different parts of the workshop, specifically
+in the first keynote (\S\ref{sec:keynote1}), the Developing and
+Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
+(\S\ref{sec:policy}). It quickly became clear that at present there is
+no consensus among the community, whether within or accross
+disciplines, on what this definition should be, and that a variety of
+different defintiions were being assumed, used, or sometime expressly
+articulated by contributors and attendees. However, some concepts,
+particularly relating sustainability to change over time, were also
+evidently held in common. This common notion is, for example, captured
+in the definition used by the UK's Scientific Software Sustainability
+Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
+will be available---and continue to be improved and supported---in the
+future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
+software that continues to serve its users.
+
+Philip Bourne, too, used the relation to change over time when he
+suggested in his opening keynote that sustainability can perhaps be
+defined as the effort needed to make the essential things continue.
+This leads to having to decide what it is that we want to sustain,
+whether what we want to sustain is valuable, and finally, who would
+care and how much if it went away. As was pointed out during a
+discussion session, OSS Watch, an advisory service for issues relating
+to free and open source software in the UK, proposes a Software
+Sustainability Maturity Model to address the issue of what level of
+sustainability a particular element of software needs to have, and
+where this is important. It, too, expresses sustainability in relation
+to change over time:
+\begin{quote}``When choosing software for procurement or
+development reuse --- regardless of the license and development model you will
+use --- you need to consider the future. While a software product may satisfy
 today's needs, will it satisfy tomorrow's needs? Will the supplier still be
@@
 requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
-
-Attendees suggested that a key question that the definition of sustainability
-is one issue on which the community needs to agree, and that ideally, an
-initial definition would be determined during the workshop, or at least some
-progress would be made towards this goal.  Another topic that came up is what
-the goal of sustainability is.  Perhaps it is reproducible science, or
-persistence, or quality, or something else.  Similarly, some attendees want to
-understand how success in sustainability is measured.  How does a group of
-developers know when they have actually achieved sustainable software? This led
-to a comment that sustainability should be addressed throughout the full
-software life cycle.
-
-Another topic that came up during the Developing and Maintaining Software panel
-is the relationship of sustainability to other software attributes.  Attendees
-asked ``what is the relationship between sustainability and provenance?'' And,
-``is usability separate from sustainability or a fundamental part of it?''
-
-In addition, the Policy panel had a large amount of discussion about defining
-sustainability, as one of the subtopics in that panel was Modeling
-Sustainability, and modeling requires defining what will be modeled.
-
-In this subtopic, two papers included discussion of the definition of
-sustainability.  Venters et al.~\cite{Venters_WSSSPE} mentioned that this is a
-rather ambiguous concept, and that the lack of an accepted definition gets in
-the way of integrating the concept into software engineering. They suggest that
-sustainability is a non-functional requirement, and that the quality of
-software architectures determines sustainability. They then propose that
-sustainability is a consequence of a set of central quality attributes:
-extensibility, interoperability, maintainability, portability, reusability, and
-scalability. Finally, they develop an architecture evaluation framework based
-on scenarios that help to illuminate how to measure quality or sustainability
-at the architectural level.
-
-Katz and Proctor~\cite{Katz_WSSSPE} included a set of questions that could be
-used to measure software sustainability, and though these questions might
-falsely lead to yes or no answers, the complete set would determine a range of
-values for sustainability. These questions are:
+\end{quote}
+
+Attendees suggested that having a definition of sustainability on
+which the community can agree is key.  A related question that was
+raised is what the goal of sustainability should be, with a wide range
+of possible answers, including more reproducible science, software
+persistence, and quality.  And given a goal of sustainability, how
+would success in achieving it be measured?  How would one know that a
+piece of software has reached sustainability? As a commenter
+suggested, this may mean that sustainability should be addressed
+throughout the full software life cycle.  It was also pointed out that
+software sustainability is not isolated from other attributes of
+scientific software and its use, such as usability, and provenance.
+
+Because part of the Policy panel (\S\ref{sec:policy}) was about
+modeling sustainability, and modeling requires defining what will be
+modeled, this panel saw particular attention to the questions
+surrounding the definition of sustainability. Two papers, Venters et
+al.~\cite{Venters_WSSSPE} and Katz and Proctor~\cite{Katz_WSSSPE},
+specifically discuss the issue.
+
+According to Venters et al.~\cite{Venters_WSSSPE}, sustainability is a
+rather ambiguous concept, and the lack of an accepted definition
+hampers integrating the concept into software engineering. They
+suggest that sustainability falls under the category of non-functional
+requirements, and that a software's sustainability is a consequence of
+a set of central software architecture quality attributes, including
+extensibility, interoperability, maintainability, portability,
+reusability, and scalability. They also propose an evaluation
+framework with which quality and sustainability could be measured at
+the architectural level.
+
+Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that could be
+used to measure software sustainability:
 \begin{itemize}
@@
 \item Does it provide the functionality that existing and future users want?
-\item Does it incorporate new science/theory/tools as they develop?
+\item Will it incorporate new science, theory, or tools as they develop?
 \end{itemize}
-
-Lenhardt~\cite{lenhardt-wssspe1-panel} summarized the contributions of the
-Modeling Sustainability papers in the panel. As shown in
-Table~\ref{tab:defining-sustainability}, for each paper he discussed what
-software meant, whether there was a definition of sustainability, and what the
-approach was to either understand or evaluate sustainability.
-
+Despite their phrasing, these questions are not intended to be given
+simplistic yes or no answers, and it is the complete set rather than
+any individual one that would determine where in the range of
+sustainability a software falls.
+
+Table~\ref{tab:defining-sustainability} reproduces a summary created
+by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each contribution to the
+Modeling Sustainability panel what it meant when referring to
+software, how it defined software sustainability, and which
+approach it suggested to understand or evaluate sustainability.
 \begin{table}[t]

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Adds closing recommendation to define sustainability.

@@
 
-Finally, in the workshop's closing session, one of the discussion topics was
-what success would look like for the set of WSSSPE activities beyond just the workshop.
-One of the answers that was suggested was
-having an agreed version of what we mean by sustainability.
-
 \subsection{Career Paths for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:career-tracks}
@@
 
+One of the subjects that contributed papers, attendees, and speakers
+wrestled with repeatedly is defining sustainability such that the
+community can get behind it. Although some had hoped that at least an
+initial consensus could be reached in the course of the workshop, this
+proved elusive. However, in the absence of such a definition it will
+remain difficult to define exactly what the goals should be towards
+achieving, or even only improving software sustainability, and hence
+what practices should be followed and implemented when. Thus, to be
+seen successful, activities emerging from or following this first
+WSSSPE workshop should include working out a community consensus on
+what we mean by sustainability.
+
 \subsubsection*{Communities, Models, and Education}

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Fixes dashes that should be em-dashes.

@@
 researchers to get their codes out into the community and in use in education
-very rapidly. The primary software challenges are twofold - support for the
+very rapidly. The primary software challenges are twofold --- support for the
 \toolname{HUBzero} framework and challenges related to the software contributed by the
@@
 the Cactus project went far beyond standard practices of APIs. Domain specific
-languages (DSLs) allow decoupling of components - for example I/O, mesh
+languages (DSLs) allow decoupling of components --- for example I/O, mesh
 refinement, PAPI counters, and boundary conditions abstracted from science

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Removes duplicated word.

@@
 In places, Bourne said, development in science has improved thanks to
-open source and hosting services like GitHub~\cite{github-web}, but for the most part
+open source and hosting services like GitHub~\cite{github-web}, but
+for the most part it
 remains arcane. He argued that we can learn much from the App Store

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
Revises Career Paths section.

@@
 
-Career path issues also came up repeatedly, starting in the the first keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the Google Bus'' to describe the issue of staff in his lab in San Diego who had chosen to leave science and work for Google (but since they didn't want to physically move, they take the Google bus to the Google office each day.)
-
-The lack of career paths also came up in the Developing and Supporting Software discussion (\S\ref{sec:devel}), as an issue that led to difficulties in established projects retaining as well as recruiting staff.
-The issue there was that for software to be sustainable, there need to be career paths for software developers.
-There was also some discussion about how should software development be balanced with research, with the idea that faculty are hired based on their research qualifications, not on their software development qualification, and as such, specializing in software development was likely to not lead to a career in science at a university, though clearly, national laboratories are a different case.
-
-In the Policy discussion (\S\ref{sec:policy}), career paths were not discussed explicitly, though the discussion about credit, citation, and impact focused on how those who create or contribute to software are recognized; this recognition (or the current lack thereof) is tied to career paths (or their lack) for software developers.
-
-The Communities discussion (\S\ref{sec:community}) did not directly address career paths, though one of the papers~\cite{Loffler_WSSSPE} that was part of this section also makes a similar policy point, in a section about career paths.
-Specifically, it says ``The most severe problem for developers in most computational sciences currently is that while most of the work is done creating hopefully well-written, sustainable software, the academic success is often exclusively tied to the solution of the scientific problem the software was designed for.
-Tasks that from a software engineering standpoint are essential, e.g., high usability, well-written and updated documentation, or porting infrastructure to new platforms, are not rewarded within this system.''
-
-In terms of possible solutions, the Developing and Supporting Software discussion said that things are beginning to improve, for example, with incentives for software development in the form of altmetrics, tenure committee consideration, and NSF's change in its biosketch requirement: asking for proposer's ``products'' rather than their ``publications,'' as had been done previously.
-However, it was also noted it can be very difficult to measure where people are using small bits of your code.
-And the possible career paths for a specialist software developer working as part of a scientific research group in a university are certainly not yet clear.
+Career path issues also came up repeatedly, starting in the first
+keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the
+Google Bus'' to describe the issue of well-trained software development staff in
+academic labs choosing to leave science and to work instead for technology
+firms, especially Google, which happens in large enough numbers that
+Google operates a bus every day to its nearest offices (and hence
+staff who leaves academia in this way does not even have  to physically move).
+
+The career path issue emerged repeatedly across panels because for
+scientific software to be(come) sustainable, projects trying to create
+sustainable software need to be able to recruit and retain software
+developers trained in the various requisite software engineering
+facets.  However, a career path in research means faculty at most
+universitities, and as was noted repeatedly in discussions, faculty
+are hired based on their scientific qualifications, not on their
+software development skills or track record. Consequently, developing
+special software development skills is unlikely to further a career in
+science at a university, although national laboratories were
+acknowledged as a different case. Loffler et
+al.~\cite{Loffler_WSSSPE}, one of the papers in the Communities panel
+(\S\ref{sec:community}), brought the problem to the point:
+\begin{quote}
+  ``The most severe problem for developers in most computational
+  sciences currently is that while most of the work is done creating
+  hopefully well-written, sustainable software, the academic success
+  is often exclusively tied to the solution of the scientific problem
+  the software was designed for. Tasks that from a software
+  engineering standpoint are essential, e.g., high usability,
+  well-written and updated documentation, or porting infrastructure to
+  new platforms, are not rewarded within this system.'' ~\cite{Loffler_WSSSPE}
+\end{quote}
+
+Clearly, improving the recognition of software engineering work is
+connected to addressing the career path problem. As was noted in the
+Developing and Supporting Software (\S\ref{sec:devel}) and the Policy
+(\S\ref{sec:policy}) panel discussions, there are encouraging signs of
+improvement, including some altmetric services (such as Impactstory,
+\url{http://impactstory.org}) collecting metrics for software source
+code, and the fact that NSF now asks to list ``products'' rather than
+their ``publications'' in an investigator's biosketch or results from
+prior NSF support. However, how software, let alone parts of software
+are reused by others can be very difficult to measure, and better
+recognizing software products for principle investigators by itself
+does not create career paths for specialist software developers
+working as part of a university research group. Huang and
+Lapp~\cite{Huang_WSSSPE}, a contribution to the Policy panel, offer
+one possible solution in which a software engineering center of
+excellence offers a career path for a correspondingly trained
+workforce, and increased recognition of the resulting more sustainable
+software would in a virtuous cycle heighten the value of the center's
+services.
 

### 2014-04-21 WSSSPE1_report/WSSSPE1_summary_report.tex
finishing conclusions for first draft

@@
 repository, central resources for services and consulting, a think tank of
-sorts, and a software orphanage center. \note{What is software orphanage center?} The idea of one common repository received
+sorts, and a software orphanage center (i.e., a place to `take care' of software
+when the original developers have stopped doing so).
+The idea of one common repository received
 some resistance, with so many compelling alternatives available, e.g. Bitbucket or
@@
 
-To conclude, we discuss the major concerns and issues that were raised in
-the workshop, what we have learned, and what we plan to do going forward.
-
-\note{add some analysis of the deficiencies and difficulties that are
-  present in different fields, and those that are common?}
-
-\subsection{Concerns and issues}
-
-\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}
-
-The issues that came up repeatedly in different parts of the workshop are, as described in
-\S\ref{sec:cross-cutting}, the need for a definition of sustainability, and the need for
-career paths for scientific software developers.
-
-\subsubsection*{Developing and Supporting Software}
-
-\subsubsection*{Policy}
-
-\subsubsection*{Communities, Models, and Education}
-
-Recognition of scholarship in scientific software over research 
-articles is overdue.
-
-\subsection{Recommendations or lessons}
-
-\note{lessons from the 3 discussion sessions go here, and lessons from the cross-cutting section as well?}
-
-\subsubsection*{Developing and Supporting Software}
-
-Successful collaborative projects conducted via the use of shared
-repositories ensures reproducibility in computational science.
-
-\subsubsection*{Policy}
-
-One of the subjects that contributed papers, attendees, and speakers
-wrestled with repeatedly is defining sustainability such that the
+To conclude, we discuss what we have learned from the workshop and what we plan to do going forward.
+
+%\note{add some analysis of the deficiencies and difficulties that are
+%  present in different fields, and those that are common?}
+
+%\subsection{Concerns and issues}
+%
+%\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}
+
+\subsection{Issues and lessons}
+
+%\note{lessons from the 3 discussion sessions go here, and lessons from the cross-cutting section as well?}
+
+Three issues came up repeatedly in different parts of the workshop:
+
+\begin{enumerate}
+\item The need for a definition of sustainability such that the
 community can get behind it. Although some had hoped that at least an
@@
 achieving, or even only improving software sustainability, and hence
-what practices should be followed and implemented when. Thus, to be
-seen successful, activities emerging from or following this first
-WSSSPE workshop should include working out a community consensus on
-what we mean by sustainability.
-
-\subsubsection*{Communities, Models, and Education}
-A sustainability model for scientific software is to build a pipeline
-from construction to consumption as found in the most efficient
+what practices should be followed and implemented when. As described
+in the next subsection (\S\ref{sec:future}), the workshop organizers have begun an effort
+to address this.
+
+\item The need for academic
+career paths for scientific software developers.  
+Unfortunately, it is not clear how to ensure that these career paths become available,
+other than repeatedly talking about this issue. The recent Moore and Sloan initiative
+in data science~\cite{moore_sloan} are trying to address this, to some extent, by providing funds and
+incentives to universities in the US that work towards this goal.
+
+%
+%\subsubsection*{Developing and Supporting Software}
+%
+%\subsubsection*{Policy}
+%
+%\subsubsection*{Communities, Models, and Education}
+
+\item The need for recognition of scholarship in scientific software over research 
+articles. This need probably is the most addressed of the three, today, with
+efforts underway such as the Mozilla Science Lab, GitHub, and figshare
+``Code as a research object'' project~\cite{code_as_a_research_object} among others.
+
+\end{enumerate}
+
+%\subsubsection*{Developing and Supporting Software}
+
+\subsubsection*{Discussion sessions}
+
+Two strong lessons came out of the three discussion sessions:
+
+\begin{enumerate}
+\item
+Successful collaborative projects conducted via the use of shared
+repositories are a means towards reproducibility in computational science.
+\katznote{this seems rather limited - what else comes out of `successful collaborative projects conducted via the use of shared
+repositories'?}
+
+%\subsubsection*{Policy}
+
+%One of the subjects that contributed papers, attendees, and speakers
+%wrestled with repeatedly is defining sustainability such that the
+%community can get behind it. Although some had hoped that at least an
+%initial consensus could be reached in the course of the workshop, this
+%proved elusive. However, in the absence of such a definition it will
+%remain difficult to define exactly what the goals should be towards
+%achieving, or even only improving software sustainability, and hence
+%what practices should be followed and implemented when. As described
+%in the next subsection, a subset of the organizers have begun an effort
+%to address this.
+
+%Thus, to be
+%seen successful, activities emerging from or following this first
+%WSSSPE workshop should include working out a community consensus on
+%what we mean by sustainability.
+
+%\subsubsection*{Communities, Models, and Education}
+\item A sustainability model for scientific software is to build a pipeline
+from construction to consumption, as found in the most efficient
 information technology enterprises.
 
+\end{enumerate}
+
 % Conclusion of S8 : Use Cases, not sure where to put
-\note{conclusions from S8, not sure of location --Ketan}
-
+%\note{conclusions from S8, not sure of location --Ketan}
+
+\subsubsection*{Use cases}
 Two distinct class of scientific software projects and products were discussed
@@
 
-\subsection{Future activities}
+\subsection{Future activities} \label{sec:future}
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
adding Berriman affiliation
Career paths and definition issues as refs to cross-cutting.

@@
 
-The panel on developing and supporting software examined the challenges
+The panel on Developing and Supporting Software examined the challenges
 around scientific software development and support, mainly focused on research
-groups that also produce code in various forms. There was widespread agreement
+groups that in addition to pursuing research also produce code in various forms. There was widespread agreement
 that developing and maintaining software is hard, but best practices can help.
@@
 documented development practices can be very helpful in bringing  new developers
-into a project efficiently.
+into a project.
+
+Two subjects that prominently surfaced in this panel also came up
+throughout other parts of the workshop, and are therefore deferred to
+the section on Cross-cutting Issues (\S\ref{sec:cross-cutting}). These
+are the lack of long-term career paths for specialists in the various
+software development and support areas (see
+\S\ref{sec:career-tracks}), and the question of what ``sustainable''
+should mean in the context of software (see
+\S\ref{sec:defining-sustainability}).
 
@@
 
-\subsection{Communities and Career Paths}
+\subsection{The Importance of Communities}
 
@@
 building and continued engagement need attention during the project life cycle.
-Several of the submitted papers discussed how communities have been built around
-projects, and what is needed to enable a project to grow. Among these are
+Several of the submitted papers \todo{which ones? should cite here.} discussed how communities have been built around
+projects, and what is needed to enable a project to grow. The latter
+includes 
 public source-code hosting, mailing lists, documentation, wikis, bug trackers, software
 downloads, continuous integration, software quality dashboards, and of course,
-a general web presence to tie all of these things together.
-
-There was extended discussion about users not answering the questions of other
-users. Several participants
-offered counterexamples, such as mailing lists where developers do not participate
-as much due to users being more active, or whether the ``core
-team'' can end up setting unrealistic expectations by doing too much. Team Geek~\cite{opac-b1134063}
+a general web presence to tie a project's channels and artifacts together.
+
+There was extended discussion about the challenge of fostering
+communities in which users help each other, rather than always
+deferring to the developers of project to answer user queries. Participants
+offered examples that this is indeed possible, such as mailing lists
+in which developers do not participate
+much because users actively respond to questions from other users, but
+also asked whether by doing too much the ``core
+team'' could end up setting unrealistic expectations. Team Geek~\cite{opac-b1134063}
 and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
-lists that are welcoming to people actually have many more people contributing.
-
-Recruiting and retaining personnel in this area is hard,
-with one of the major reasons being no long-term career paths (especially when
-compared with industry). This is further discussed in~\S\ref{sec:career-tracks}.
-
-%moved to cross-cutting issues
-%
-%How should software development be balanced with
-%research? It is apparent that things are beginning to improve, such as
-%incentives for software development in the form of altmetrics, tenure committee
-%consideration, and NSF ``products'' vs ``publications.'' It was noted it can be very
-%difficult to measure where people are using small bits of your code.
-
-\subsection{Defining Sustainability for Scientific Software}
-
-The question of what ``sustainable'' means in the context of software was
-raised (see \S\ref{sec:defining-sustainability}.) The resources at OSS Watch~\cite{osswatch-web}   discuss what
-to look for when choosing software for procurement, or reuse in further software
-development. Regardless of the license and development model that will be used,
-the future of the project must be considered. Even if a particular piece of
-software satisfies today's requirements, will it continue to do so in the
-future? Important questions include whether the supplier will still be around in five
-years' time, will it still care for its customers' needs, will it be responsive to
-bug reports and feature requests? Should you tie your investment to a single
-supplier using a proprietary product, or ensure the project uses an OSI-approved
-license, and can outlive any single entity if the software is still important.
-
-What is the overarching goal of sustainability? Is it reproducible science,
-persistence, quality, something else? How should success be measured in this
-context? Is there some metric that can be used to determine when you have
-achieved sustainable software, or is this an ongoing process with no clear
-endpoint. For truly sustainable software there should be no endpoint, as the
-software products continue to be used and useful beyond the single institution,
-grant, and developer/development team. Sustainability must be addressed
-throughout the project life cycle.
+lists tend to  have many more people contributing when they are welcoming to people.
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Adds endpoint and license comments to sust. def.
These were in the Developing Software section, but hadn't fully carried over
here yet.

@@
 would success in achieving it be measured?  How would one know that a
-piece of software has reached sustainability? As a commenter
-suggested, this may mean that sustainability should be addressed
-throughout the full software life cycle.  It was also pointed out that
-software sustainability is not isolated from other attributes of
-scientific software and its use, such as usability, and provenance.
+piece of software has reached sustainability? Participants pointed out
+that for truly sustainable software there should be no endpoint at
+which sustainability can be claimed, because the software products
+would continue to be used and useful beyond the single institution,
+grant, and developer or development team that created them. This may
+mean that sustainability needs to be addressed throughout the full
+software life cycle.  It was also pointed out that software
+sustainability is not isolated from other attributes of scientific
+software and its use, such as usability, and provenance. Similarly,
+the question was considered, albeit only briefly, whether proprietary
+versus open-source license plays a role in the context softwara
+sustainability. For example, should a project ensure that it uses an
+OSI-approved license so that software products can outlive any single
+entity if they remain important.
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Clean up notes.

@@
 
-\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)
-\note{\scriptsize Authors, please add/check middle initials and affiliations}}
+\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
+%\note{\scriptsize Authors, please add/check middle initials and affiliations}
 
@@
 %
-\thanks{{}$^{(13)}$ xxx}
+\thanks{{}$^{(13)}$ Infrared Processing and Analysis Center, California Institute of Technology, Pasadena, CA, USA; \url{gbb@ipac.caltech.edu}}
 %
@@
 
-\section{Keynotes \note{lead: Sou-Cheng (Terrya) Choi}} \label{sec:keynotes}
+\section{Keynotes } \label{sec:keynotes} %\note{lead: Sou-Cheng (Terrya) Choi}
 
@@
 
-\section{Developing and Supporting Software \note{\scriptsize lead: Marcus Hanwell, contributor: Suresh Marru}} \label{sec:devel}
+\section{Developing and Supporting Software} \label{sec:devel}%\note{\scriptsize lead: Marcus Hanwell, contributor: Suresh Marru}
 
@@
 
-\section{Policy \note{lead: Colin Venters, contributors: James Howison, Hilmar Lapp}} \label{sec:policy}
+\section{Policy} \label{sec:policy}%\note{lead: Colin Venters, contributors: James Howison, Hilmar Lapp}
 
@@
 
-\section{Communities \note{\vspace{-1ex}\tiny lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}} \label{sec:community}
+\section{Communities} \label{sec:community}%\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
 
@@
 
-\subsection{Defining Sustainability \note{lead: Daniel S. Katz}}  \label{sec:defining-sustainability}
+\subsection{Defining Sustainability}  \label{sec:defining-sustainability}%\note{lead: Daniel S. Katz}
 
@@
                 \hline
-Lenhardt, et al.~\cite{Lenhardt_WSSSPE}
+Lenhardt,~et~al.~\cite{Lenhardt_WSSSPE}
 & Broadly defined as software supporting science
@@
 
-\subsection{Career Paths for Scientific Software Developers} \note{lead: Daniel S. Katz} \label{sec:career-tracks}
+\subsection{Career Paths for Scientific Software Developers}\label{sec:career-tracks} %\note{lead: Daniel S. Katz}
 
@@
 
-\todo{anyone who needs to put something in here should}
+%\todo{anyone who needs to put something in here should}
 
@@
 
-\section{Attendees \note{lead: Shel Swenson}} \label{sec:attendees}
+\section{Attendees} \label{sec:attendees}  %\note{lead: Shel Swenson}
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Minor changes.

@@
 The idea of one common repository received
-some resistance, with so many compelling alternatives available, e.g. Bitbucket or
+some resistance, with so many compelling alternatives available, e.g., Bitbucket or
 GitHub. The centralized communication or point of contact was seen as
@@
 deeper participation. For instance, promoting collaborations across the breadth
-(e.g.  different departments) and depth (e.g. stakeholders within a department)
+(e.g., different departments) and depth (e.g., stakeholders within a department)
 of community, one of the key ingredient of a sustainable process. Successful
@@
 this survey. We find that in a majority of scientific software, more than one
-original developer are still around, in some cases after more than 10 years,
-(eg. \toolname{R/qtl}).  Whereas, in the case of utility software projects, it
+original developer are still around, in some cases after more than 10 years
+(e.g., \toolname{R/qtl}).  Whereas, in the case of utility software projects, it
 appears that at least one original developer is still around, often in a

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
More minor changes.

@@
 
-\title[]{Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
+\title[]{Overview \& Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
 %\note{\scriptsize Authors, please add/check middle initials and affiliations}
@@
 discussed. For software that builds on other software frameworks it is the
-API that is most important, and this can
+APIs that are most important, and this can
 encompass issues such as the source and binary interfaces to the software
@@
 The papers submitted to this panel included many general
-recommendations for processes, practices, tools, etc. One of the papers
+recommendations for processes, practices, tools, etc. One of the papers \note{cite?}
 suggested that a ``Software Sustainability Institute'' should be vested with
-the authority to develop standardized quality processes, a central common
+the authority to develop standardized quality processes, a common
 repository, central resources for services and consulting, a think tank of
@@
 
-The role of the growing field of team science with software development  was discussed. The science of team science
-deals with understanding and improving collaborative and team-based scientific research, and such issues as virtual organizations,
-and tool development cut across software development communities. Further, how should these skills and best practices then be
+The role of the growing field of team science with software development  was discussed. Team science
+deals with understanding and improving collaborative and team-based scientific research, and  issues such as virtual organizations,
+and tool development across software development communities. Further, how should these skills and best practices then be
 introduced to students?
@@
 In addition, there will be a minisymposium on ``Reliable Computational Science''
-at the 2014 Society for Industrial and Applied Mathematics (SIAM)
+at the 2014 Society for Industrial and Applied Mathematics  
 Annual Meeting (SIAM AN14, \url{http://meetings.siam.org})
-to further explore some of the key issues raised in
-WSSSPE.
+to further explore some of the key issues raised here.
 
@@
 Foundation; any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not
+expressed in this material are those of the author and do not
 necessarily reflect the views of the National Science Foundation.
+
+Choi thanks Ian Foster, Director of the Computation Institute, University of Chicago,
+for encouraging her work in reliable reproducible research and
+supporting her trip to participate in the WSSSPE1 conference as a contributor.
 
@@
 UK Software Sustainability Institute.
-
-Choi thanks Ian Foster, Director of the Computation Institute, University of Chicago,
-for encouraging her work in reliable reproducible research and
-supporting her trip to participate in the WSSSPE1 conference as a contributor.
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Clean up and revision of Introduction.

--- 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 emerging in academic environments as an integral part of
-successful research teams. The WSSSPE workshop  provided a forum
-for discussion of the challenges around scientific software, including both positions and experiences. The
-workshop short papers and debates have been archived to provide a basis for continued
+successful research teams. This first WSSSPE workshop  provided a forum
+for discussion of the challenges around sustaining scientific software, including
+contributed short papers in the form of both positions and experience
+reports. These short papers, as well as notes from the debates around
+them, have been archived to provide a basis for continued
 discussion, and have fed into the collaborative writing of this
-document. Additional papers based on extended versions of workshop
-submissions have been developed as part of this special journal edition. The level of interest in the workshop which involved
-an estimated~90 to~150 participants during the day has led to further activities:
-Two additional WSSSPE workshops are planned  (at the~2014 SciPy and SC14 conferences),
+report.  Some of the workshop submissions have been extended to full
+papers, which form part of the same special journal edition in which this report
+appears. The workshop generated a high level of interest; an
+estimated~90 to~150 participants were in attendance at different times
+of the day. The interest and discussions have already led to follow-up activities:
+A smaller Python-specific workshop is planned to be held at the~2014
+SciPy conference, and a follow-on WSSSPE2 workshop has been proposed
+to, and meanwhile been
+accepted for the SC14 conference. In addition, 
 funds to support the workshops have been obtained from
-the National Science Foundation and the Gordon and Betty Moore
+the US National Science Foundation (NSF) and the Gordon and Betty Moore
 Foundation, and the
-original workshop website has been turned into a community website for
-further discussion and progress. Additionally, a minisymposium  to
-further explore some of the key issues raised in WSSSPE is co-organized by a
-WSSSPE1 participant at the~2014 Society for Industrial and Applied Mathematics
+original workshop website has been turned into a community website to engender
+further discussion and progress. Additionally, a minisymposium  at the~2014 Society for Industrial and Applied Mathematics
 (SIAM) Annual Meeting on ``Reliable
-Computational Science'' (SIAM AN14, \url{http://meetings.siam.org}).
-
-This paper summarizes the  workshop papers
-and discussion, and identifies issues that future WSSSPE events will address.
-Before the WSSSPE1 workshop took place, the organizers self-published
-a paper~\cite{WSSSPE1-pre-report} to document the process of
-organizing and advertising the workshop, collecting and reviewing the
-papers, and putting together the agenda. Section~\ref{sec:process} is
-based on that publication. The remainder of this paper is based on the
-workshop itself, as documented in a set of collaborative notes taken
-during the workshop~\cite{WSSSPE1-google-notes}, including discussion
-about the keynote presentations~(\S\ref{sec:keynotes}), the three
-panels~(\S\ref{sec:devel}-\ref{sec:community}),
-cross-cutting issues~(\S\ref{sec:cross-cutting}),
-use cases~(\S\ref{sec:use-cases}), and conclusions~(\S\ref{sec:conclusions}).
-Additionally, the accepted workshop papers are listed in
-Appendix~\ref{sec:papers}, and a partial list of workshop attendees is
-in Appendix~\ref{sec:attendees}.
+Computational Science'' (SIAM AN14, \url{http://meetings.siam.org})  is being co-organized by a
+WSSSPE1 participant to
+further explore some of the key issues raised at the workshop.
+
+This report attempts to summarize the various aspects of the
+workshop.  The remainder of this paper first gives an overview of the
+process with which the workshop was put together
+(\S\ref{sec:process}),  then proceeds with summaries of the two
+keynote presentations~(\S\ref{sec:keynotes}), followed by summaries of
+the workshop papers grouped by the three thematic workshop panels to which they
+were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two broader
+cross-cutting issues surfaced repeatedly, and are
+discussed separately~(\S\ref{sec:cross-cutting}), as are
+use cases for sustainable software~(\S\ref{sec:use-cases}). The
+summaries are based not only on the papers and panel presentations, but
+also on the many comments raised in both the onsite and online discussions
+that accompanied the workshop elements, as documented by collaborative
+note taking during the workshop~\cite{WSSSPE1-google-notes}. We
+conclude with issues and lessons learned, as well as plans for future activities~(\S\ref{sec:conclusions}).
+The short papers accepted to the workshop are listed in
+Appendix~\ref{sec:papers}, and a partial list of workshop attendees
+can be found in Appendix~\ref{sec:attendees}.
 

### 2014-04-22 WSSSPE1_report/WSSSPE1_summary_report.tex
Move CfP to Appendix and revise Process.

@@
 
-WSSSPE1 was organized in collaboration of the relatively small group
-of five organizers and a larger peer-review committee with~36
-members. This committee had early influence on the organization, e.g.,
-on the specific call for papers.
-
-The workshop call for papers included:
-
-\begin{quote}
-``In practice, scientific software activities are part of an ecosystem
-where key roles are held by developers, users, and funders. All three
-groups supply resources to the ecosystem, as well as requirements that
-bound it. Roughly following the example of NSF's Vision and Strategy
-for Software~\cite{NSF_software_vision},
-the ecosystem may be viewed as having challenges related to:
-
-\begin{itemize}[leftmargin=0.2in]
-\item the development process that leads to new (versions of) software
-\begin{itemize}[leftmargin=0.2in]
-\item how fundamental research in computer science or
-  science/engineering domains is turned into reusable software
-\item software created as a by-product of research
-\item impact of computer science research on the development of
-    scientific software and vice versa
-\end{itemize}
-\item the support and maintenance of existing software, including
-  software engineering
-\begin{itemize}[leftmargin=0.2in]
-\item governance, business, and sustainability models
-\item the role of community software repositories, their operation and
-  sustainability
-\end{itemize}
-\item the role of open source communities or industry
-\item use of the software
-\begin{itemize}[leftmargin=0.2in]
-\item growing communities
-\item reproducibility, transparency needs that may be unique to science
-\end{itemize}
-\item policy issues, such as
-\begin{itemize}[leftmargin=0.2in]
-\item measuring usage and impact
-\item software credit, attribution, incentive, and reward
-\item career paths for developers and institutional roles
-\item issues related to multiple organizations and multiple countries,
-  such as intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of
-  publishers
-\end{itemize}
-\item education and training''
-\end{itemize}
-
-\end{quote}
-
-Based on the goal of encouraging a wide range of submissions from
+WSSSPE1 was organized by a relatively small group
+of five organizers and a larger program committee of~36
+members. The program committee peer-reviewed submissions, but also had
+early influence in the workshop's organization, such as articulating
+the Call for Papers (see Appendix~\ref{sec:cfp}).
+
+Aside from setting the stage for the relevance of software
+sustainability and corresponding training and workforce issues to
+science,  the call for papers enumerated the topics it was interested
+in as challenges to the ecosystem of which scientific software is a
+part, and in which software developers, users, and funders hold
+roles. These challenges roughly followed NSF's Vision and Strategy
+for Software~\cite{NSF_software_vision}, and specifically included the
+development and research process that leads to new (or new versions of existing)
+software; the support, community infrastructure, and engineering for
+maintenance of existing software; the role of open source communities
+and industry; aspects of the use of software, such as
+reproducibility, that may be unique to science;  policy issues
+related to software sustainability such as measuring impact,
+giving credit, and incentivizing best practices; and education and training.
+
+The workshop's goal was to encourage a wide range of submissions from
 those involved in software practice, ranging from initial thoughts and
-partial studies to mature deployments, the organizers wanted to make
-submission as easy as possible. The call for papers stated:
-
-\begin{quote}
-``We invite short (4-page) position/experience reports that will be used
-to organize panel and discussion sessions. These papers will be
-archived by a third-party service, and provided DOIs [Digital Object Identifiers].
-We encourage
-submitters to license their papers under a Creative Commons license
-that encourages sharing and remixing, as we will combine ideas (with
-attribution) into the outcomes of the workshop.  An interactive site
-will be created to link these papers and the workshop discussion, with
-options for later comments and contributions. Contributions will be
-peer-reviewed for relevance and originality before the links are added
-to the workshop site; contributions will also be used to determine
-discussion topics and panelists. We will also plan one or more papers
-to be collaboratively developed by the contributors, based on the
-panels and discussions.''
-\end{quote}
-
-In total, 58 submissions were received, and almost all submitters used either
-arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to self-publish
-their papers.
-
-A peer review process followed the submissions, where
-the~58 papers received~181 reviews, an average of~3.12 reviews per
+partial studies to mature deployments. Consequently, the organizers
+aimed to make submission as easy as possible. Rather than requiring
+adherence to a formal submission system and a full research
+paper-style template, submissions were intentionally limited to short
+4-page papers, articulating either a position on one or more of the
+topics, or reporting experiences related to them. Furthermore, for
+submission authors were asked to self-archive (and thus self-publish)
+their papers with a third-party service that issues persistent
+identifiers, such as Digital Object Identifiers (DOIs), and to then
+submit the URL to the archived paper by email. This had the side
+effect that every submitter would also have a publicly available and
+citable version of their workshop contribution.
+
+This process resulted in a total of 58 submissions. Almost all submitters used either
+arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to archive
+their papers. The submissions were then subjected to peer review by
+the program committee, resulting in~181 reviews, an average of~3.12 reviews per
 paper. Reviews were completed using a Google form, which allowed
-reviewers to choose papers they wanted to review, then to provide
-general comments and scores on relevance to the organizers and
-to the authors. The review reports enabled the organizers to decide
-which papers to associate with the workshop, and allowed the authors
-to improve their papers.
-
-The organizers decided to list~54 of the papers (see Appendix~\ref{sec:papers} for a list of these papers) as significant
-contributions to the workshop, a very high acceptance rate but one
-that is reasonable, given the goal of broad participation and the fact
-that the reports were already self-published. Based on the submissions, the papers were
+reviewers to choose papers they wanted to review, and to provide
+general comments as well as relevance scores to the organizers and
+to the authors. Based on the review reports, the organizers decided to
+list~54 of the papers (see Appendix~\ref{sec:papers} for a full list) as significant
+contributions to the workshop. The high acceptance rate may come as a
+surprise, but it is nonetheless consistent with the goal of
+fostering broad participation, and as a corollary of the chosen submission
+process paper acceptance was no longer a means to filter
+the papers' public availability.
+
+In roughly following the aforementioned scientific software ecosystem
+challenges, the accepted submissions were
 grouped into three main categories, namely \emph{Developing and
 Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-category was associated with a panel and discussion at the
-workshop.  Three to four representatives from each submission
-category were appointed as panelists, and assigned to read a subset of
-the papers in that category and then discuss them in the panel.
+category was assigned a panel, with three to four panelists 
+drawn from authors of the associated submissions, who were each
+assigned to read and summarily present a subset of
+the papers associated with the panel. The process from organizing and
+advertising the workshop, to collecting and reviewing the
+papers, and putting together the agenda was documented by the
+organizers in a report~\cite{WSSSPE1-pre-report}, which they
+self-archived in the same way as contributed papers.
+
+For each panel, the paper summaries were followed by active discussion
+involving panelists, onsite attendees, and often enough online
+participants. The latter was facilitated by having a shared Google
+Doc~\cite{WSSSPE1-google-notes} for collaborative note taking. Some of
+the online discussion also took place on Twitter (hashtag
+\texttt{\#wssspe}).
+
 The workshop itself consisted of two keynote presentations followed by the
-three panel/discussion sessions.  Notes were collaboratively taken
-by the group during the discussion, using a web document~\cite{WSSSPE1-google-notes}.
+three panel sessions.  Remote participation was
+facilitated by a live-cast of all keynotes and panels on Ustream.tv (\url{http://ustream.tv}).
 
@@
 
+\section{Call For Papers} \label{sec:cfp}
+
+The full Call for Papers for WSSSPE1 can be found online at
+\url{http://wssspe.researchcomputing.org.uk/wssspe1/cfp/}. We include
+below the part that explains scope and topics for which submissions
+were solicited (see \S\ref{sec:process} for number of submissions
+received, accepted, and the corresponding process).
+
+\begin{quote}
+``Progress in scientific research is dependent on the quality and
+accessibility of software at all levels and it is now critical to
+address many new challenges related to the development, deployment,
+and maintenance of reusable software. In addition, it is essential
+that scientists, researchers, and students are able to learn and adopt
+a new set of software-related skills and methodologies. Established
+researchers are already acquiring some of these skills, and in
+particular a specialized class of software developers is emerging in
+academic environments who are an integral and embedded part of
+successful research teams. This workshop will provide a forum for
+discussion of the challenges, including both positions and
+experiences. The short papers and discussion will be archived as a
+basis for continued discussion, and we intend the workshop to feed
+into the collaborative writing of one or more journal publications.
+
+In practice, scientific software activities are part of an ecosystem
+where key roles are held by developers, users, and funders. All three
+groups supply resources to the ecosystem, as well as requirements that
+bound it. Roughly following the example of NSF's Vision and Strategy
+for Software~\cite{NSF_software_vision},
+the ecosystem may be viewed as having challenges related to:
+
+\begin{itemize}[leftmargin=0.2in]
+\item the development process that leads to new (versions of) software
+\begin{itemize}[leftmargin=0.2in]
+\item how fundamental research in computer science or
+  science/engineering domains is turned into reusable software
+\item software created as a by-product of research
+\item impact of computer science research on the development of
+    scientific software and vice versa
+\end{itemize}
+\item the support and maintenance of existing software, including
+  software engineering
+\begin{itemize}[leftmargin=0.2in]
+\item governance, business, and sustainability models
+\item the role of community software repositories, their operation and
+  sustainability
+\end{itemize}
+\item the role of open source communities or industry
+\item use of the software
+\begin{itemize}[leftmargin=0.2in]
+\item growing communities
+\item reproducibility, transparency needs that may be unique to science
+\end{itemize}
+\item policy issues, such as
+\begin{itemize}[leftmargin=0.2in]
+\item measuring usage and impact
+\item software credit, attribution, incentive, and reward
+\item career paths for developers and institutional roles
+\item issues related to multiple organizations and multiple countries,
+  such as intellectual property, licensing, etc.
+\item mechanisms and venues for publishing software, and the role of
+  publishers
+\end{itemize}
+\item education and training
+\end{itemize}
+
+This workshop is interested in all of the above topics.  We invite short (4-page) position/experience reports that will be used
+to organize panel and discussion sessions. These papers will be
+archived by a third-party service, and provided DOIs [Digital Object Identifiers].
+We encourage
+submitters to license their papers under a Creative Commons license
+that encourages sharing and remixing, as we will combine ideas (with
+attribution) into the outcomes of the workshop.  An interactive site
+will be created to link these papers and the workshop discussion, with
+options for later comments and contributions. Contributions will be
+peer-reviewed for relevance and originality before the links are added
+to the workshop site; contributions will also be used to determine
+discussion topics and panelists. We will also plan one or more papers
+to be collaboratively developed by the contributors, based on the
+panels and discussions.''
+\end{quote}
+
 \section{Attendees} \label{sec:attendees}  %\note{lead: Shel Swenson}

### 2014-04-23 WSSSPE1_report/WSSSPE1_summary_report.tex
Shorten title.
fiddling.

@@
 
-\title[]{Overview \& Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
+\title[]{Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)}
 %\note{\scriptsize Authors, please add/check middle initials and affiliations}
@@
 software, there should be no endpoint, as the software products continue to be
-used and useful beyond a single institution, grant, and developer or
+used beyond a single institution, grant, and developer or
 development team.
@@
 
-Another problem Bourne experienced was staff retention with respect to
+Another problem Bourne experienced was staff retention from
 private organizations which reward those combining research and
@@
 noted that making a successful business from scientific software alone
-is very rare: founders overvalue while customers undervalue. He noted
+is  rare: founders overvalue while customers undervalue. He noted
 that to last, an open source project needs a minimal funding
-requirement even with a vibrant community --- goodwill only goes so
+requirement even with a vibrant community---goodwill only goes so
 far if one is being paid to do something else.  He talked about grant
@@
 The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon Smith of
-GitHub. Smith's talk started with an example from his own past in data
+GitHub. Smith started with an example from his past in data
 reduction in Astronomy, where he needed to remove interfering effects from the
@@
 while it was persistent, there was no way or practice of sharing these data
-among the scientists. Consequently many researchers repeated the same
+among scientists. Consequently many researchers repeated the same
 calculations. Smith estimated that 13 person-years were wasted by this
@@
 argued that with regard to reusable software, ``we should just start
-doing it.'' In this regard Smith replied that documentation should be
+doing it.''  Smith replied that documentation should be
 ``treated as a first class entity.''  He noted that the open source

### 2014-04-23 WSSSPE1_report/WSSSPE1_summary_report.tex
Response to a remark on use of repositories.

@@
 building and continued engagement need attention during the project life cycle.
-Several of the submitted papers \todo{which ones? should cite here.} discussed how communities have been built around
+Several of the submitted papers \todo{cite} discussed how communities have been built around
 projects, and what is needed to enable a project to grow. The latter
@@
 
-\section{Case Studies \note{lead: Ketan Maheshwari, contributors: Andy Terrel}} \label{sec:use-cases}
+\section{Case Studies} \label{sec:use-cases} % \note{lead: Ketan Maheshwari, contributors: Andy Terrel}
 
@@
 \begin{enumerate}
-\item
-Successful collaborative projects conducted via the use of shared
-repositories are a means towards reproducibility in computational science.
-\katznote{this seems rather limited - what else comes out of `successful collaborative projects conducted via the use of shared
-repositories'?}
+\item Use of shared repositories in the development of collaborative 
+projects facilitates collaboration, reproducibility, and sustainability 
+in computational science. However it represents a barrier in some scientific fields and 
+has yet to be more widely adopted.
 

### 2014-04-23 WSSSPE1_report/WSSSPE1_summary_report.tex
Commenting out authors with no affiliations at this point.

@@
 James~Howison$^{(10)}$,
-Shel~Swenson$^{(11)}$,
-Gabrielle~D.~Allen$^{(12)}$,
-Bruce~Berriman$^{(13)}$,
-Colin~Venters$^{(14)}$
+%Shel~Swenson$^{(11)}$,
+%Gabrielle~D.~Allen$^{(12)}$,
+Bruce~Berriman$^{(11)}$
+%Colin~Venters$^{(14)}$
 }
@@
 %
-\thanks{{}$^{(11)}$ xxx}
-%
-\thanks{{}$^{(12)}$ xxx}
-%
-\thanks{{}$^{(13)}$ Infrared Processing and Analysis Center, California Institute of Technology, Pasadena, CA, USA; \url{gbb@ipac.caltech.edu}}
-%
-\thanks{{}$^{(14)}$ xxx}
+%\thanks{{}$^{(11)}$ xxx}
+%
+%\thanks{{}$^{(12)}$ xxx}
+%
+\thanks{{}$^{(11)}$ Infrared Processing and Analysis Center, California Institute of Technology, Pasadena, CA, USA; \url{gbb@ipac.caltech.edu}}
+%
+%\thanks{{}$^{(14)}$ xxx}
 

### 2014-04-23 WSSSPE1_report/WSSSPE1_summary_report.tex
No change to content. Formatting tex.

@@
 \begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
-This document summarizes the First Workshop on Sustainable Software for Science:
-Practice and Experiences (WSSSPE1). The content is based on
-a set of collaborative notes taken during the workshop,
-including discussion about the keynote presentations, three themed panels that
-discussed papers accepted at the workshop`, other discussion, cross-cutting
-issues, use cases, and conclusions. It joins a nascent literature seeking to
-understand what drives software work in science and how the reward systems of
-science function, thereby shaping the types of software projects undertaken, including the
-extent to which developers are motivated to build software for the long-term,
-for the use of others, and whether to work collaboratively or separately.
-Unique perspectives were captured about issues such as comprehensive documentation,
-development and deployment practices; software licenses; career paths
-for developers; identifiers for software through a
-mechanism such as a Digital Object Identifier or publication of a ``software
-paper''; and evidence of software contribution and impact, recorded in online
-systems such as source code repositories like GitHub. For truly sustainable
-software, there should be no endpoint, as the software products continue to be
-used beyond a single institution, grant, and developer or
-development team.
+This document summarizes the First Workshop on Sustainable Software
+for Science: Practice and Experiences (WSSSPE1). The content is based
+on a set of collaborative notes taken during the workshop, including
+discussion about the keynote presentations, three themed panels that
+discussed papers accepted at the workshop`, other discussion,
+cross-cutting issues, use cases, and conclusions. It joins a nascent
+literature seeking to understand what drives software work in science
+and how the reward systems of science function, thereby shaping the
+types of software projects undertaken, including the extent to which
+developers are motivated to build software for the long-term, for the
+use of others, and whether to work collaboratively or separately.
+Unique perspectives were captured about issues such as comprehensive
+documentation, development and deployment practices; software
+licenses; career paths for developers; identifiers for software
+through a mechanism such as a Digital Object Identifier or publication
+of a ``software paper''; and evidence of software contribution and
+impact, recorded in online systems such as source code repositories
+like GitHub. For truly sustainable software, there should be no
+endpoint, as the software products continue to be used beyond a single
+institution, grant, and developer or development team.
 
@@
 
-Because progress in scientific research is dependent on the quality of and
-accessibility to software at all levels, it is now critical to address many
-challenges related to the development, deployment, and maintenance of reusable
-software.
+Because progress in scientific research is dependent on the quality of
+and accessibility to software at all levels, it is now critical to
+address many challenges related to the development, deployment, and
+maintenance of reusable software.
 %\note{Ketan: What are the levels in scientific process where software
@@
 %software required in each stage.}
-In addition, it is essential that scientists,
-researchers, and students are able to learn and adopt software-related skills
-and methodologies. Established researchers are already acquiring some of these
-skills, and in particular a specialized class of software developers is
-emerging in academic environments as an integral part of
+In addition, it is essential that scientists, researchers, and
+students are able to learn and adopt software-related skills and
+methodologies. Established researchers are already acquiring some of
+these skills, and in particular a specialized class of software
+developers is emerging in academic environments as an integral part of
 successful research teams. This first WSSSPE workshop provided a forum
-for discussion of the challenges around sustaining scientific software, including
-contributed short papers in the form of both positions and experience
-reports. These short papers, as well as notes from the debates around
-them, have been archived to provide a basis for continued
-discussion, and have fed into the collaborative writing of this
-report.  Some of the workshop submissions have been extended to full
-papers, which form part of the same special journal edition in which this report
-appears. The workshop generated a high level of interest; an
-estimated~90 to~150 participants were in attendance at different times
-of the day. The interest and discussions have already led to follow-up activities:
-A smaller Python-specific workshop is planned to be held at the~2014
-SciPy conference, and a follow-on WSSSPE2 workshop has been proposed
-to, and meanwhile been
-accepted for the SC14 conference. In addition, 
-funds to support the workshops have been obtained from
-the US National Science Foundation (NSF) and the Gordon and Betty Moore
-Foundation, and the
-original workshop website has been turned into a community website to engender
-further discussion and progress. Additionally, a minisymposium  at the~2014 Society for Industrial and Applied Mathematics
-(SIAM) Annual Meeting on ``Reliable
-Computational Science'' (SIAM AN14, \url{http://meetings.siam.org})  is being co-organized by a
-WSSSPE1 participant to
-further explore some of the key issues raised at the workshop.
-
-This report attempts to summarize the various aspects of the
-workshop.  The remainder of this paper first gives an overview of the
-process with which the workshop was put together
-(\S\ref{sec:process}),  then proceeds with summaries of the two
-keynote presentations~(\S\ref{sec:keynotes}), followed by summaries of
-the workshop papers grouped by the three thematic workshop panels to which they
-were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two broader
-cross-cutting issues surfaced repeatedly, and are
-discussed separately~(\S\ref{sec:cross-cutting}), as are
-use cases for sustainable software~(\S\ref{sec:use-cases}). The
-summaries are based not only on the papers and panel presentations, but
-also on the many comments raised in both the onsite and online discussions
-that accompanied the workshop elements, as documented by collaborative
-note taking during the workshop~\cite{WSSSPE1-google-notes}. We
-conclude with issues and lessons learned, as well as plans for future activities~(\S\ref{sec:conclusions}).
-The short papers accepted to the workshop are listed in
-Appendix~\ref{sec:papers}, and a partial list of workshop attendees
-can be found in Appendix~\ref{sec:attendees}.
+for discussion of the challenges around sustaining scientific
+software, including contributed short papers in the form of both
+positions and experience reports. These short papers, as well as notes
+from the debates around them, have been archived to provide a basis
+for continued discussion, and have fed into the collaborative writing
+of this report.  Some of the workshop submissions have been extended
+to full papers, which form part of the same special journal edition in
+which this report appears. The workshop generated a high level of
+interest; an estimated~90 to~150 participants were in attendance at
+different times of the day. The interest and discussions have already
+led to follow-up activities: A smaller Python-specific workshop is
+planned to be held at the~2014 SciPy conference, and a follow-on
+WSSSPE2 workshop has been proposed to, and meanwhile been accepted for
+the SC14 conference. In addition, funds to support the workshops have
+been obtained from the US National Science Foundation (NSF) and the
+Gordon and Betty Moore Foundation, and the original workshop website
+has been turned into a community website to engender further
+discussion and progress. Additionally, a minisymposium at the~2014
+Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
+on ``Reliable Computational Science'' (SIAM AN14,
+\url{http://meetings.siam.org}) is being co-organized by a WSSSPE1
+participant to further explore some of the key issues raised at the
+workshop.
+
+This report attempts to summarize the various aspects of the workshop.
+The remainder of this paper first gives an overview of the process
+with which the workshop was put together (\S\ref{sec:process}), then
+proceeds with summaries of the two keynote
+presentations~(\S\ref{sec:keynotes}), followed by summaries of the
+workshop papers grouped by the three thematic workshop panels to which
+they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two
+broader cross-cutting issues surfaced repeatedly, and are discussed
+separately~(\S\ref{sec:cross-cutting}), as are use cases for
+sustainable software~(\S\ref{sec:use-cases}). The summaries are based
+not only on the papers and panel presentations, but also on the many
+comments raised in both the onsite and online discussions that
+accompanied the workshop elements, as documented by collaborative note
+taking during the workshop~\cite{WSSSPE1-google-notes}. We conclude
+with issues and lessons learned, as well as plans for future
+activities~(\S\ref{sec:conclusions}).  The short papers accepted to
+the workshop are listed in Appendix~\ref{sec:papers}, and a partial
+list of workshop attendees can be found in
+Appendix~\ref{sec:attendees}.
 
@@
 
-WSSSPE1 was organized by a relatively small group
-of five organizers and a larger program committee of~36
-members. The program committee peer-reviewed submissions, but also had
-early influence in the workshop's organization, such as articulating
-the Call for Papers (see Appendix~\ref{sec:cfp}).
+WSSSPE1 was organized by a relatively small group of five organizers
+and a larger program committee of~36 members. The program committee
+peer-reviewed submissions, but also had early influence in the
+workshop's organization, such as articulating the Call for Papers (see
+Appendix~\ref{sec:cfp}).
 
@@
 part, and in which software developers, users, and funders hold
-roles. These challenges roughly followed NSF's Vision and Strategy
-for Software~\cite{NSF_software_vision}, and specifically included the
-development and research process that leads to new (or new versions of existing)
-software; the support, community infrastructure, and engineering for
-maintenance of existing software; the role of open source communities
-and industry; aspects of the use of software, such as
-reproducibility, that may be unique to science;  policy issues
-related to software sustainability such as measuring impact,
-giving credit, and incentivizing best practices; and education and training.
+roles. These challenges roughly followed NSF's Vision and Strategy for
+Software~\cite{NSF_software_vision}, and specifically included the
+development and research process that leads to new (or new versions of
+existing) software; the support, community infrastructure, and
+engineering for maintenance of existing software; the role of open
+source communities and industry; aspects of the use of software, such
+as reproducibility, that may be unique to science; policy issues
+related to software sustainability such as measuring impact, giving
+credit, and incentivizing best practices; and education and training.
 
@@
 
-This process resulted in a total of 58 submissions. Almost all submitters used either
-arXiv~\cite{arXiv-web} or Figshare~\cite{figshare-web} to archive
-their papers. The submissions were then subjected to peer review by
-the program committee, resulting in~181 reviews, an average of~3.12 reviews per
-paper. Reviews were completed using a Google form, which allowed
-reviewers to choose papers they wanted to review, and to provide
-general comments as well as relevance scores to the organizers and
-to the authors. Based on the review reports, the organizers decided to
-list~54 of the papers (see Appendix~\ref{sec:papers} for a full list) as significant
+This process resulted in a total of 58 submissions. Almost all
+submitters used either arXiv~\cite{arXiv-web} or
+Figshare~\cite{figshare-web} to archive their papers. The submissions
+were then subjected to peer review by the program committee, resulting
+in~181 reviews, an average of~3.12 reviews per paper. Reviews were
+completed using a Google form, which allowed reviewers to choose
+papers they wanted to review, and to provide general comments as well
+as relevance scores to the organizers and to the authors. Based on the
+review reports, the organizers decided to list~54 of the papers (see
+Appendix~\ref{sec:papers} for a full list) as significant
 contributions to the workshop. The high acceptance rate may come as a
-surprise, but it is nonetheless consistent with the goal of
-fostering broad participation, and as a corollary of the chosen submission
-process paper acceptance was no longer a means to filter
-the papers' public availability.
+surprise, but it is nonetheless consistent with the goal of fostering
+broad participation, and as a corollary of the chosen submission
+process paper acceptance was no longer a means to filter the papers'
+public availability.
 
 In roughly following the aforementioned scientific software ecosystem
-challenges, the accepted submissions were
-grouped into three main categories, namely \emph{Developing and
-Supporting Software}, \emph{Policy}, and \emph{Communities}. Each
-category was assigned a panel, with three to four panelists 
-drawn from authors of the associated submissions, who were each
-assigned to read and summarily present a subset of
-the papers associated with the panel. The process from organizing and
-advertising the workshop, to collecting and reviewing the
-papers, and putting together the agenda was documented by the
-organizers in a report~\cite{WSSSPE1-pre-report}, which they
+challenges, the accepted submissions were grouped into three main
+categories, namely \emph{Developing and Supporting Software},
+\emph{Policy}, and \emph{Communities}. Each category was assigned a
+panel, with three to four panelists drawn from authors of the
+associated submissions, who were each assigned to read and summarily
+present a subset of the papers associated with the panel. The process
+from organizing and advertising the workshop, to collecting and
+reviewing the papers, and putting together the agenda was documented
+by the organizers in a report~\cite{WSSSPE1-pre-report}, which they
 self-archived in the same way as contributed papers.
@@
 
-The workshop itself consisted of two keynote presentations followed by the
-three panel sessions.  Remote participation was
-facilitated by a live-cast of all keynotes and panels on Ustream.tv (\url{http://ustream.tv}).
+The workshop itself consisted of two keynote presentations followed by
+the three panel sessions.  Remote participation was facilitated by a
+live-cast of all keynotes and panels on Ustream.tv
+(\url{http://ustream.tv}).
 
@@
 E. Bourne of University of California, San Diego.  Bourne is a
-biomedical scientist who has also formed four software companies. He co-founded
-PLOS Computational Biology~\cite{plos-web} and helped develop the RCSB
-Protein Data Bank~\cite{pdb-web}.
-He is working on automating three-dimensional visualizations of cell
-contents and molecular structures, a problem that has not been solved
-and when done, would serve as a key function of software in biomedical
+biomedical scientist who has also formed four software companies. He
+co-founded PLOS Computational Biology~\cite{plos-web} and helped
+develop the RCSB Protein Data Bank~\cite{pdb-web}.  He is working on
+automating three-dimensional visualizations of cell contents and
+molecular structures, a problem that has not been solved and when
+done, would serve as a key function of software in biomedical
 sciences.
 
-Bourne's  presentation was based on his own software experiences.  He emphasized that
-sustainability for software ``does not just mean more money from
-Government'' (see also Section~\ref{sec:defining-sustainability}).  Other
-factors to consider, he mentioned, encompass costs of production, ease
-of maintenance, community involvement, and distribution channels.
+Bourne's presentation was based on his own software experiences.  He
+emphasized that sustainability for software ``does not just mean more
+money from Government'' (see also
+Section~\ref{sec:defining-sustainability}).  Other factors to
+consider, he mentioned, encompass costs of production, ease of
+maintenance, community involvement, and distribution channels.
 
@@
 open source and hosting services like GitHub~\cite{github-web}, but
-for the most part it
-remains arcane. He argued that we can learn much from the App Store
-model about interfaces, ratings, and so on. He also mentioned
-BioJava~\cite{biojava-web} and Open Science Data Cloud~\cite{osdc-web}
-as distribution channels.  On a related note, Bourne observed a common
-evolutionary pathway for computational biology projects, from data
-archive to analytics platform to educational use, and suggested that
-use of scientific software for outreach might be the final step.
+for the most part it remains arcane. He argued that we can learn much
+from the App Store model about interfaces, ratings, and so on. He also
+mentioned BioJava~\cite{biojava-web} and Open Science Data
+Cloud~\cite{osdc-web} as distribution channels.  On a related note,
+Bourne observed a common evolutionary pathway for computational
+biology projects, from data archive to analytics platform to
+educational use, and suggested that use of scientific software for
+outreach might be the final step.
 
@@
 
-Another problem Bourne experienced was staff retention from
-private organizations which reward those combining research and
-software expertise (the ``Google Bus''). However, he is a strong fan
-of software sustainability through public-private partnerships. He
-noted that making a successful business from scientific software alone
-is  rare: founders overvalue while customers undervalue. He noted
-that to last, an open source project needs a minimal funding
-requirement even with a vibrant community---goodwill only goes so
-far if one is being paid to do something else.  He talked about grant
-schemes of relevance in the U.S., particularly with regard to
-technology transfer~\cite{sbir-web, fased-web}.
-
-Bourne also had problems with selling research software: the university technology
-transfer office wanted huge and unrealistic intellectual property
-reach through, whereby they would get a share of profits from drugs
-developed by pharmaceutical companies who use the software.
-He advocated for  a one-click approach for customers to
+Another problem Bourne experienced was staff retention from private
+organizations which reward those combining research and software
+expertise (the ``Google Bus''). However, he is a strong fan of
+software sustainability through public-private partnerships. He noted
+that making a successful business from scientific software alone is
+rare: founders overvalue while customers undervalue. He noted that to
+last, an open source project needs a minimal funding requirement even
+with a vibrant community---goodwill only goes so far if one is being
+paid to do something else.  He talked about grant schemes of relevance
+in the U.S., particularly with regard to technology
+transfer~\cite{sbir-web, fased-web}.
+
+Bourne also had problems with selling research software: the
+university technology transfer office wanted huge and unrealistic
+intellectual property reach through, whereby they would get a share of
+profits from drugs developed by pharmaceutical companies who use the
+software.  He advocated for a one-click approach for customers to
 purchase university-written software.
 
-He then presented arguments on directly valuing software as
-a research output alongside papers, a common discussion within this field.
-He mentioned an  exploration of involving software
-engineers in the review process of scientific code~\cite{peer-review-code}, and discussed
-how publishing software reviews could change attitudes.
+He then presented arguments on directly valuing software as a research
+output alongside papers, a common discussion within this field.  He
+mentioned an exploration of involving software engineers in the review
+process of scientific code~\cite{peer-review-code}, and discussed how
+publishing software reviews could change attitudes.
 
@@
 technology (IT) underpins the whole of organizational activities, he
-contended that universities are way behind the curve. In
-particular, he highlighted the separation of research, teaching, and
+contended that universities are way behind the curve. In particular,
+he highlighted the separation of research, teaching, and
 administration into silos without a common IT framework as a blocker
 to many useful organizational innovations: ``University 2.0 is yet to
-happen.''  He argued that
-funders such as NSF and NIH can help train institutions, not just individuals, in this regard.
-
-Bourne concluded by discussing his 2011 paper ``Ten Simple Rules for Getting Ahead as a Computational Biologist in Academia''~\cite{bourne_ten} and
-argued that computational scientists ``have a responsibility to
-convince their institutions, reviewers, and communities that software is
-scholarship, frequently
-more valuable than a research article''.
+happen.''  He argued that funders such as NSF and NIH can help train
+institutions, not just individuals, in this regard.
+
+Bourne concluded by discussing his 2011 paper ``Ten Simple Rules for
+Getting Ahead as a Computational Biologist in
+Academia''~\cite{bourne_ten} and argued that computational scientists
+``have a responsibility to convince their institutions, reviewers, and
+communities that software is scholarship, frequently more valuable
+than a research article''.
 
@@
 
-The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon Smith of
-GitHub. Smith started with an example from his past in data
-reduction in Astronomy, where he needed to remove interfering effects from the
-experimental apparatuses. He built a ``bad pixel mask,'' and realized that
-while it was persistent, there was no way or practice of sharing these data
-among scientists. Consequently many researchers repeated the same
-calculations. Smith estimated that 13 person-years were wasted by this
-repetition.
+The second keynote~\cite{WSSSPE1-keynote2} was delivered by Arfon
+Smith of GitHub. Smith started with an example from his past in data
+reduction in Astronomy, where he needed to remove interfering effects
+from the experimental apparatuses. He built a ``bad pixel mask,'' and
+realized that while it was persistent, there was no way or practice of
+sharing these data among scientists. Consequently many researchers
+repeated the same calculations. Smith estimated that 13 person-years
+were wasted by this repetition.
 
@@
 research. He then asked, when we do know better, why we do not act any
-different. He argued that it was due to the lack of incentives:
-only the immediate products of research, not the software, are valued.
-He referenced Victoria Stodden's talk at
-OKCon~\cite{okcon-stodden-talk} which he said argued these points
-well.
-
-C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor,
-argued that with regard to reusable software, ``we should just start
-doing it.''  Smith replied that documentation should be
-``treated as a first class entity.''  He noted that the open source
-community has excellent cultures of code reuse, for example,
-RubyGems~\cite{rubygems-web}, PyPI~\cite{pypi-web}, and
-CPAN~\cite{cpan-web}, where there is effectively low-friction
-collaboration through the use of repositories. This has not happened
-in highly numerical, compiled language scientific software.  An
-exception he cited as a good example of scientific projects using
-GitHub is the EMCEE Markov Chain Monte Carlo project~\cite{emcee-web}
-by Dan Foreman-Mackey and contributors.
+different. He argued that it was due to the lack of incentives: only
+the immediate products of research, not the software, are valued.  He
+referenced Victoria Stodden's talk at OKCon~\cite{okcon-stodden-talk}
+which he said argued these points well.
+
+C. Titus Brown~\cite{ged-web}, a WSSSPE1 contributor, argued that with
+regard to reusable software, ``we should just start doing it.''  Smith
+replied that documentation should be ``treated as a first class
+entity.''  He noted that the open source community has excellent
+cultures of code reuse, for example, RubyGems~\cite{rubygems-web},
+PyPI~\cite{pypi-web}, and CPAN~\cite{cpan-web}, where there is
+effectively low-friction collaboration through the use of
+repositories. This has not happened in highly numerical, compiled
+language scientific software.  An exception he cited as a good example
+of scientific projects using GitHub is the EMCEE Markov Chain Monte
+Carlo project~\cite{emcee-web} by Dan Foreman-Mackey and contributors.
 
@@
 
-In conclusion Smith reiterated the importance of establishing effective
-incentive models for open contributions and tool builders, for
-example, meaningful metrics and research grants such as~\cite{NSF_software_vision}. He urged computational scientists to
+In conclusion Smith reiterated the importance of establishing
+effective incentive models for open contributions and tool builders,
+for example, meaningful metrics and research grants such
+as~\cite{NSF_software_vision}. He urged computational scientists to
 collaborate and share often their research reports, teaching
@@
 
-The panel on Developing and Supporting Software examined the challenges
-around scientific software development and support, mainly focused on research
-groups that in addition to pursuing research also produce code in various forms. There was widespread agreement
-that developing and maintaining software is hard, but best practices can help.
-Several participants added that documentation is not just for users, and writing
-application programming interface (API) documentation, tutorials for building and deploying software, together with
-documented development practices can be very helpful in bringing  new developers
-into a project.
+The panel on Developing and Supporting Software examined the
+challenges around scientific software development and support, mainly
+focused on research groups that in addition to pursuing research also
+produce code in various forms. There was widespread agreement that
+developing and maintaining software is hard, but best practices can
+help.  Several participants added that documentation is not just for
+users, and writing application programming interface (API)
+documentation, tutorials for building and deploying software, together
+with documented development practices can be very helpful in bringing
+new developers into a project.
 
@@
 
-Software is developed for many different purposes, and the requirements can
-vary significantly depending on the intended audience. Most end-users make use
-of either a graphical user interface of some kind, or a command line that may
-offer input and output formats for running the code and analyzing its output(s).
-When discussing backward compatibility it is these various interfaces that are
-discussed. For software that builds on other software frameworks it is the
-APIs that are most important, and this can
-encompass issues such as the source and binary interfaces to the software
-libraries developed---with each potentially having a high maintenance cost if
-they are to remain compatible over many years. When using command-line programs
-it is generally the command-line switches as well as the input/output formats
-that could incur significant costs if they are changed.
-
-There was discussion that backward compatibility is not always desirable, and it
-can be very costly. This must be balanced with the aims of a given project, and how
-many other projects depend on and use the code when backwards incompatible
-changes are to be made. There are many examples in the wider open source
-software world of strategies for dealing with this, and again best practices
-can go a long way to mitigating issues around backwards compatibility. Many
-projects live with sub-optimal code for a while, and may allow backwards
-compatibility to be broken at agreed-upon development points, such as a major
-release for a software library.
-
-There were 13 articles about different experiences in this area, but little
-about GUI testing, performance, scalability, or agile development practices.
-There were several unique perspectives about issues such as managing API
-changes, using the same best practices for software as data, and going beyond
-simply ``slapping an OSI-approved license on code.''
+Software is developed for many different purposes, and the
+requirements can vary significantly depending on the intended
+audience. Most end-users make use of either a graphical user interface
+of some kind, or a command line that may offer input and output
+formats for running the code and analyzing its output(s).  When
+discussing backward compatibility it is these various interfaces that
+are discussed. For software that builds on other software frameworks
+it is the APIs that are most important, and this can encompass issues
+such as the source and binary interfaces to the software libraries
+developed---with each potentially having a high maintenance cost if
+they are to remain compatible over many years. When using command-line
+programs it is generally the command-line switches as well as the
+input/output formats that could incur significant costs if they are
+changed.
+
+There was discussion that backward compatibility is not always
+desirable, and it can be very costly. This must be balanced with the
+aims of a given project, and how many other projects depend on and use
+the code when backwards incompatible changes are to be made. There are
+many examples in the wider open source software world of strategies
+for dealing with this, and again best practices can go a long way to
+mitigating issues around backwards compatibility. Many projects live
+with sub-optimal code for a while, and may allow backwards
+compatibility to be broken at agreed-upon development points, such as
+a major release for a software library.
+
+There were 13 articles about different experiences in this area, but
+little about GUI testing, performance, scalability, or agile
+development practices.  There were several unique perspectives about
+issues such as managing API changes, using the same best practices for
+software as data, and going beyond simply ``slapping an OSI-approved
+license on code.''
 
@@
 
-Communities are extremely important in software projects, and both their
-building and continued engagement need attention during the project life cycle.
-Several of the submitted papers \todo{cite} discussed how communities have been built around
-projects, and what is needed to enable a project to grow. The latter
-includes 
-public source-code hosting, mailing lists, documentation, wikis, bug trackers, software
-downloads, continuous integration, software quality dashboards, and of course,
-a general web presence to tie a project's channels and artifacts together.
+Communities are extremely important in software projects, and both
+their building and continued engagement need attention during the
+project life cycle.  Several of the submitted papers \todo{cite}
+discussed how communities have been built around projects, and what is
+needed to enable a project to grow. The latter includes public
+source-code hosting, mailing lists, documentation, wikis, bug
+trackers, software downloads, continuous integration, software quality
+dashboards, and of course, a general web presence to tie a project's
+channels and artifacts together.
 
@@
 communities in which users help each other, rather than always
-deferring to the developers of project to answer user queries. Participants
-offered examples that this is indeed possible, such as mailing lists
-in which developers do not participate
-much because users actively respond to questions from other users, but
-also asked whether by doing too much the ``core
-team'' could end up setting unrealistic expectations. Team Geek~\cite{opac-b1134063}
-and Turk's paper on scaling code in the human dimension~\cite{Turk:2013:SCH:2484762.2484782}  discuss how development
-lists tend to  have many more people contributing when they are welcoming to people.
+deferring to the developers of project to answer user
+queries. Participants offered examples that this is indeed possible,
+such as mailing lists in which developers do not participate much
+because users actively respond to questions from other users, but also
+asked whether by doing too much the ``core team'' could end up setting
+unrealistic expectations. Team Geek~\cite{opac-b1134063} and Turk's
+paper on scaling code in the human
+dimension~\cite{Turk:2013:SCH:2484762.2484782} discuss how development
+lists tend to have many more people contributing when they are
+welcoming to people.
 
@@
 The papers submitted to this panel included many general
-recommendations for processes, practices, tools, etc. One of the papers \note{cite?}
-suggested that a ``Software Sustainability Institute'' should be vested with
-the authority to develop standardized quality processes, a common
-repository, central resources for services and consulting, a think tank of
-sorts, and a software orphanage center (i.e., a place to `take care' of software
-when the original developers have stopped doing so).
-The idea of one common repository received
-some resistance, with so many compelling alternatives available, e.g., Bitbucket or
-GitHub. The centralized communication or point of contact was seen as
-reasonable, with the statement that ``vested with authority'' is perhaps too strong,
-but ``providing tools if needed'' might be more appropriate.
-
-What about actual software engineering principles, such as modularity and
-extensibility? This is how industry maintains software, and ensures it continues
-to be useful. Often, rewriting software is considered to be too costly, but
-with a modular design it can be kept up to date. Extensibility is expected to keep
-it relevant, if built into the project. One counterpoint raised by Jason Riedy
-was that trying to take advantage of the latest and greatest hardware often
-makes this painful, hence the lack of papers mentioning ``GPUs and exotic
-hardware.''
-
-The question of whether funders, such as the NSF, can mandate software plans in
-much the same way as they do data management plans, was raised. Daniel Katz responded that
-software is supposed to be described as part of the NSF data management plan,
-and that in NSF's definition, data includes software. A comment from Twitter
-(@biomickwatson) raised the issue that this requires reviewers and funders who
-understand the answers that are given in these plans. Daniel Katz responded
-that in programs focused on software or data this can be done effectively, but
-agreed that in more general programs this is indeed a problem.
+recommendations for processes, practices, tools, etc. One of the
+papers \note{cite?}  suggested that a ``Software Sustainability
+Institute'' should be vested with the authority to develop
+standardized quality processes, a common repository, central resources
+for services and consulting, a think tank of sorts, and a software
+orphanage center (i.e., a place to `take care' of software when the
+original developers have stopped doing so).  The idea of one common
+repository received some resistance, with so many compelling
+alternatives available, e.g., Bitbucket or GitHub. The centralized
+communication or point of contact was seen as reasonable, with the
+statement that ``vested with authority'' is perhaps too strong, but
+``providing tools if needed'' might be more appropriate.
+
+What about actual software engineering principles, such as modularity
+and extensibility? This is how industry maintains software, and
+ensures it continues to be useful. Often, rewriting software is
+considered to be too costly, but with a modular design it can be kept
+up to date. Extensibility is expected to keep it relevant, if built
+into the project. One counterpoint raised by Jason Riedy was that
+trying to take advantage of the latest and greatest hardware often
+makes this painful, hence the lack of papers mentioning ``GPUs and
+exotic hardware.''
+
+The question of whether funders, such as the NSF, can mandate software
+plans in much the same way as they do data management plans, was
+raised. Daniel Katz responded that software is supposed to be
+described as part of the NSF data management plan, and that in NSF's
+definition, data includes software. A comment from Twitter
+(@biomickwatson) raised the issue that this requires reviewers and
+funders who understand the answers that are given in these
+plans. Daniel Katz responded that in programs focused on software or
+data this can be done effectively, but agreed that in more general
+programs this is indeed a problem.
 
 \subsection{Training Scientists to Develop and Support Software}
-Part of the panel discussion focused around community structures and how academic communities are not taught how to evaluate cross-disciplinary work.
-One question raised was whether software developers can be effective if they are not part
-of the appropriate domain community, with responses that this depends on the
-specific problem and situation, and that ``T-shaped'' people who have both disciplinary depth as well as interdisciplinary and collaboration skills are important.
-The discussion focused on  whether we could teach software developers and domain scientists to
-collaborate together more effectively rather than trying to teach software developers
-about domain science and domain scientists about software development practices.
-The end goal of this would be to have a single community with a spectrum of expertise across
-domain science and software development rather than two separate communities.
-
-The role of the growing field of team science with software development  was discussed. Team science
-deals with understanding and improving collaborative and team-based scientific research, and  issues such as virtual organizations,
-and tool development across software development communities. Further, how should these skills and best practices then be
-introduced to students?
+Part of the panel discussion focused around community structures and
+how academic communities are not taught how to evaluate
+cross-disciplinary work.  One question raised was whether software
+developers can be effective if they are not part of the appropriate
+domain community, with responses that this depends on the specific
+problem and situation, and that ``T-shaped'' people who have both
+disciplinary depth as well as interdisciplinary and collaboration
+skills are important.  The discussion focused on whether we could
+teach software developers and domain scientists to collaborate
+together more effectively rather than trying to teach software
+developers about domain science and domain scientists about software
+development practices.  The end goal of this would be to have a single
+community with a spectrum of expertise across domain science and
+software development rather than two separate communities.
+
+The role of the growing field of team science with software
+development was discussed. Team science deals with understanding and
+improving collaborative and team-based scientific research, and issues
+such as virtual organizations, and tool development across software
+development communities. Further, how should these skills and best
+practices then be introduced to students?
 
@@
 
-Are there significant differences in projects that have been running for 1, 3,
-5, or 10+ years? Are there shared experiences for projects of a similar stage
-of maturity? It was noted that computing and communication have changed
-significantly over the past decade, and many of the experiences are tied to
-the history of computing and communication. See the history of GCC, Emacs, or
-the Visualization Toolkit for examples. Others felt that computing has changed
-less, but communication and the widespread availability of tools has. It was
-noted that email lists, websites, chat rooms, version control, virtual and
-physical meetings are all over 20 years old.
+Are there significant differences in projects that have been running
+for 1, 3, 5, or 10+ years? Are there shared experiences for projects
+of a similar stage of maturity? It was noted that computing and
+communication have changed significantly over the past decade, and
+many of the experiences are tied to the history of computing and
+communication. See the history of GCC, Emacs, or the Visualization
+Toolkit for examples. Others felt that computing has changed less, but
+communication and the widespread availability of tools has. It was
+noted that email lists, websites, chat rooms, version control, virtual
+and physical meetings are all over 20 years old.
 
@@
 similar, tools commonly used for computing have changed quite
-significantly. Reference was made to Perl, which was commonly used, giving way
-to whole new languages, such as Python, for gluing things together and how this
-induces many students into entirely rewriting the scaffolding, leaving the old
-to rot and the experiments to become non-reproducible as the tools change.
- There was
-discussion of this tendency along with the enormous differences in the speed and
-ease of sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
-the immediate sharing of the latest development online, using revision control
-systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.
-
-The question was also posed as to whether the distinction between researcher
-and developer is sensible, with James Hetherington commenting that in the UK
-a more nuanced view of research software engineers
-and researcher developers is examined. Should this be less of a contract relationship, and
-more of a collaborative relationship? This is also at the core of the business
-model that Kitware presented in its submission to the workshop. Are other
-ingredients missing such as applied mathematicians? Should this be defined more
-in terms of skill sets rather than roles and/or identities? This builds on the
-comments from Vaidy Sunderam that scientists are generally good writers, and
-have mathematical skills, so why can't they learn software engineering
-principles?
+significantly. Reference was made to Perl, which was commonly used,
+giving way to whole new languages, such as Python, for gluing things
+together and how this induces many students into entirely rewriting
+the scaffolding, leaving the old to rot and the experiments to become
+non-reproducible as the tools change.  There was discussion of this
+tendency along with the enormous differences in the speed and ease of
+sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
+the immediate sharing of the latest development online, using revision
+control systems like CVS, Subversion, Git, Mercurial, Bazaar, etc.
+
+The question was also posed as to whether the distinction between
+researcher and developer is sensible, with James Hetherington
+commenting that in the UK a more nuanced view of research software
+engineers and researcher developers is examined. Should this be less
+of a contract relationship, and more of a collaborative relationship?
+This is also at the core of the business model that Kitware presented
+in its submission to the workshop. Are other ingredients missing such
+as applied mathematicians? Should this be defined more in terms of
+skill sets rather than roles and/or identities? This builds on the
+comments from Vaidy Sunderam that scientists are generally good
+writers, and have mathematical skills, so why can't they learn
+software engineering principles?
 
 Miller commented that all of the infrastructure that sits around a new
-algorithm that we need to make it useful and sustainable requires different
-skill sets than the algorithm developer. Friere commented that there are no
-good career paths for people with broad skills, no incentives for them to
-continue in these roles. There was debate around people doing what interests
-them, and learning computing leaves people cold, but is it that it leaves the
-people who find career paths in academia cold versus the full spectrum of
-people involved in research? Is this also caused by poor teaching, or because the
-benefits for doing this are perceived as too small? It could also be attributed to
-their focus being on science, not software engineering, or do people with
-the passion for software engineering in science simply have no viable career
-path and either adapt or seek out alternate career paths?
+algorithm that we need to make it useful and sustainable requires
+different skill sets than the algorithm developer. Friere commented
+that there are no good career paths for people with broad skills, no
+incentives for them to continue in these roles. There was debate
+around people doing what interests them, and learning computing leaves
+people cold, but is it that it leaves the people who find career paths
+in academia cold versus the full spectrum of people involved in
+research? Is this also caused by poor teaching, or because the
+benefits for doing this are perceived as too small? It could also be
+attributed to their focus being on science, not software engineering,
+or do people with the passion for software engineering in science
+simply have no viable career path and either adapt or seek out
+alternate career paths?
 
@@
 
-The  panel on policy  discussed workshop
-contributions dealing with the wide range of software sustainability
-aspects that relate to establishing, promoting, and implementing
-policies. Six papers presented frameworks for defining, modeling, and
-evaluating software sustainability, the basis of establishing
-policies. Four papers advocated mechanisms for more properly assessing
-the impact of scientific software, and for crediting and recognizing
-work that promotes software sustainability, all of which are
-instrumental in effectively promoting policies that aim to change
-current practices. Four papers discuss facets of implementing software
+The panel on policy discussed workshop contributions dealing with the
+wide range of software sustainability aspects that relate to
+establishing, promoting, and implementing policies. Six papers
+presented frameworks for defining, modeling, and evaluating software
+sustainability, the basis of establishing policies. Four papers
+advocated mechanisms for more properly assessing the impact of
+scientific software, and for crediting and recognizing work that
+promotes software sustainability, all of which are instrumental in
+effectively promoting policies that aim to change current
+practices. Four papers discuss facets of implementing software
 sustainability, and models of implementation across different facets.
@@
 The workshop submissions grouped under this section provide frameworks
-for thinking about, researching and understanding which elements of sustainability
-are important and how they relate. Although there is substantial
-overlap between the frameworks, they have different emphases and
-extents.  Each paper in this group included a definition of sustainability,
-with many overlaps between them  (see Table~\ref{tab:defining-sustainability}).
-Perhaps unsurprisingly, the issue of how to define sustainability came
-to the fore multiple times during the workshop, and it is thus
-summarized in depth separately in ~\S\ref{sec:defining-sustainability}.
+for thinking about, researching and understanding which elements of
+sustainability are important and how they relate. Although there is
+substantial overlap between the frameworks, they have different
+emphases and extents.  Each paper in this group included a definition
+of sustainability, with many overlaps between them (see
+Table~\ref{tab:defining-sustainability}).  Perhaps unsurprisingly, the
+issue of how to define sustainability came to the fore multiple times
+during the workshop, and it is thus summarized in depth separately in
+~\S\ref{sec:defining-sustainability}.
 
@@
 overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned this~\cite{Venters_WSSSPE,Calero_WSSSPE} did so without
-integrating that analysis into the question of software being around
-long-term, suggesting that questions of environmental impact of
-scientific software is a conceptually distinct area of inquiry.
+The two papers in this area which mentioned
+this~\cite{Venters_WSSSPE,Calero_WSSSPE} did so without integrating
+that analysis into the question of software being around long-term,
+suggesting that questions of environmental impact of scientific
+software is a conceptually distinct area of inquiry.
 
@@
 that can be applied to pieces of software but might also extend to
-describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE} propose adding elements to the ISO
-standards for measuring software quality. Specifically, they propose an additional
-dimension of quality they call ``perdurability'' or ability to ``last over
-time'' made up of three characteristics: reliability, maintainability
-and adaptability. This shows some overlap with the framework employed
-by Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
+describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE}
+propose adding elements to the ISO standards for measuring software
+quality. Specifically, they propose an additional dimension of quality
+they call ``perdurability'' or ability to ``last over time'' made up
+of three characteristics: reliability, maintainability and
+adaptability. This shows some overlap with the framework employed by
+Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
 ``extensibility, interoperability, maintainability, portability,
@@
 
-Frameworks also concerned themselves with the context in which software
-projects exist, moving in abstraction from the software artifact
-itself to the organization of its production and
-the shape of the space in which an artifact or project exists.
-These frameworks take the form of contingency theories in that
-they outline a different set of challenges and argue that different
-project organizations (and presumably artifact attributes) are
-necessary to persist long term in spaces with particular
-characteristics.  Katz and Proctor~\cite{Katz_WSSSPE} propose thinking
-of this idea of space in terms of three axes: temporal (long or short
-term needs), spatial (local or global use) and purpose (specific to
-general).  They propose that different project organizations will be
-needed in different locations and argue that we should concentrate
-research to understand those connections. Weber et
-al.~\cite{Weber_WSSSPE} describe their spaces by analogy with natural
-ecosystems as ``niches'' which sustain particular pieces of
-software. They define ``a software niche as the set of technical
-requirements, organizational conditions and cultural mores that
-support its maintenance and use over time.'' They call for better
-understanding and modeling of niches (as well as further exploration
-of the usefulness of ecosystem metaphors).
+Frameworks also concerned themselves with the context in which
+software projects exist, moving in abstraction from the software
+artifact itself to the organization of its production and the shape of
+the space in which an artifact or project exists.  These frameworks
+take the form of contingency theories in that they outline a different
+set of challenges and argue that different project organizations (and
+presumably artifact attributes) are necessary to persist long term in
+spaces with particular characteristics.  Katz and
+Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in
+terms of three axes: temporal (long or short term needs), spatial
+(local or global use) and purpose (specific to general).  They propose
+that different project organizations will be needed in different
+locations and argue that we should concentrate research to understand
+those connections. Weber et al.~\cite{Weber_WSSSPE} describe their
+spaces by analogy with natural ecosystems as ``niches'' which sustain
+particular pieces of software. They define ``a software niche as the
+set of technical requirements, organizational conditions and cultural
+mores that support its maintenance and use over time.'' They call for
+better understanding and modeling of niches (as well as further
+exploration of the usefulness of ecosystem metaphors).
 
@@
 
-How work on scientific software is recognized and rewarded strongly influences the motivation for particular kinds of work
-on scientific software. A recurring theme of the panel discussion was
-that software work in science is
-inadequately visible within the reputation
-system underlying science; in other words it often doesn't ``count''. In his paper for this workshop, Katz placed
-software work along with other ``activities that facilitate science
-but are not currently rewarded or
+How work on scientific software is recognized and rewarded strongly
+influences the motivation for particular kinds of work on scientific
+software. A recurring theme of the panel discussion was that software
+work in science is inadequately visible within the reputation system
+underlying science; in other words it often doesn't ``count''. In his
+paper for this workshop, Katz placed software work along with other
+``activities that facilitate science but are not currently rewarded or
 recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
@@
 in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
-Knepley et al.~argued that the lack of visibility of software that supported a
-piece of science ``can have detrimental effects on funding, future
-library development, and even scientific
+Knepley et al.~argued that the lack of visibility of software that
+supported a piece of science ``can have detrimental effects on
+funding, future library development, and even scientific
 careers.''~\cite{Knepley_WSSSPE}.
@@
 
-Papers and discussion concentrated on three overarching questions: How ought software
-work to be visible, what are the barriers to its visibility, and what
-can be done to make it more visible?
+Papers and discussion concentrated on three overarching questions: How
+ought software work to be visible, what are the barriers to its
+visibility, and what can be done to make it more visible?
 
@@
 stacking that can make it non-obvious or even difficult to determine
-what attributions would be appropriate. Knepley et al.~\cite{Knepley_WSSSPE} approach this by
-proposing a mechanism by which the software itself, after it has run,
-provides the user with a set of citations, according to the pieces of
-code actually executed. They describe a prototype implementation whereby
-the citations are embedded in libraries and reported along with the
-results, via a command-line interface~\cite{Knepley_WSSSPE}. Discussion
-highlighted the difficulty that attempting to acknowledge the
-contributions of all pieces of dependent code within a paper faces the
-difficulty of creating very long citation lists, straining the analogy
-of code used to papers cited. Katz approaches this issue by proposing
-a system of ``transitive credit,'' recording dependencies and relative
+what attributions would be appropriate. Knepley et
+al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
+which the software itself, after it has run, provides the user with a
+set of citations, according to the pieces of code actually
+executed. They describe a prototype implementation whereby the
+citations are embedded in libraries and reported along with the
+results, via a command-line
+interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
+that attempting to acknowledge the contributions of all pieces of
+dependent code within a paper faces the difficulty of creating very
+long citation lists, straining the analogy of code used to papers
+cited. Katz approaches this issue by proposing a system of
+``transitive credit,'' recording dependencies and relative
 contributions outside particular papers, relieving authors from the
@@
 mentions of software, coping with the multitude of informal ways in
-which authors mention software they used~\cite{Priem_WSSSPE}.
-A point raised in the panel discussion was that instead of asking users to improve
-their software citation practices, one can also ask how software
-projects can better monitor the literature to improve their
-ability to show impact. For example, the NanoHub project scans the literature using keywords and names
-of known users to discover papers that are likely to have used their
-software and platform, and assigns graduate students to read each paper,
-highlighting mentions of software use and sometimes following up with the authors to
-identify stories for demonstrating impact.  A process for tracking
-software-using publications with the goal of increasing impact visibility is now described at \url{http://publications.wikia.com}.
+which authors mention software they used~\cite{Priem_WSSSPE}.  A point
+raised in the panel discussion was that instead of asking users to
+improve their software citation practices, one can also ask how
+software projects can better monitor the literature to improve their
+ability to show impact. For example, the NanoHub project scans the
+literature using keywords and names of known users to discover papers
+that are likely to have used their software and platform, and assigns
+graduate students to read each paper, highlighting mentions of
+software use and sometimes following up with the authors to identify
+stories for demonstrating impact.  A process for tracking
+software-using publications with the goal of increasing impact
+visibility is now described at \url{http://publications.wikia.com}.
 
@@
 products is not restricted to publications. Another key location for
-visibility is in the grant funding
-process, and as emphasized by NSF representatives at the meeting,
-recent changes to grant proposal and reporting formats now allow both
-applicants and awardees to report and highlight software products as
-much as publications. Nonetheless, whether peer review
-panels would value these contributions in the same way as
-publications remains to be seen.
-
-Priem and Piwowar argued that assessing the impact of software work requires
-looking beyond publications, including evidence of contribution and impact
-recorded in social coding-oriented resources such as GitHub, and conversations
-about software in issue trackers, mailing lists, twitter and
-beyond~\cite{Priem_WSSSPE}. In keeping with a principle of the ``altmetrics''
-approach, they advocate that scholars should have resources that empower them
-to tell their own stories in the manner most appropriate for them and their
-audiences.
+visibility is in the grant funding process, and as emphasized by NSF
+representatives at the meeting, recent changes to grant proposal and
+reporting formats now allow both applicants and awardees to report and
+highlight software products as much as publications. Nonetheless,
+whether peer review panels would value these contributions in the same
+way as publications remains to be seen.
+
+Priem and Piwowar argued that assessing the impact of software work
+requires looking beyond publications, including evidence of
+contribution and impact recorded in social coding-oriented resources
+such as GitHub, and conversations about software in issue trackers,
+mailing lists, twitter and beyond~\cite{Priem_WSSSPE}. In keeping with
+a principle of the ``altmetrics'' approach, they advocate that
+scholars should have resources that empower them to tell their own
+stories in the manner most appropriate for them and their audiences.
 
@@
 facilitating many or all facets of sustainability-oriented software
-design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast
-large projects that have software infrastructure development built-in,
-with cross-training domain scientist PIs in software engineering best
+design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast large
+projects that have software infrastructure development built-in, with
+cross-training domain scientist PIs in software engineering best
 practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
@@
 et al.~observe, unlike for physical IT infrastructure in data centers,
-science communities have developed very little in the way of practices and
-technology for API governance, referred to by Krintz et al.~as the
+science communities have developed very little in the way of practices
+and technology for API governance, referred to by Krintz et al.~as the
 ``combined policy, implementation, and deployment control''. Web APIs
@@
 example, in the life sciences the iPlant Collaborative~\cite{iPlant},
-the Galaxy Project~\cite{Galaxy}, and \toolname{Qiime}~\cite{Qiime} are large
-multi-year projects with significant software infrastructure funding
-and deliverables. Typically though these are the exception rather than
-the rule, and particularly in the long tail of science the
+the Galaxy Project~\cite{Galaxy}, and \toolname{Qiime}~\cite{Qiime}
+are large multi-year projects with significant software infrastructure
+funding and deliverables. Typically though these are the exception
+rather than the rule, and particularly in the long tail of science the
 scientist-developer predominates, even for software that is widely
 used or crosses domain boundaries such as
-\toolname{rOpenSci}~\cite{rOpenSci}. For better training domain scientists at
-least in basic software engineering best practices, initiatives such
-as Software Carpentry~\cite{SoftwareCarpentry} have demonstrated how
-this could be achieved at scale.
+\toolname{rOpenSci}~\cite{rOpenSci}. For better training domain
+scientists at least in basic software engineering best practices,
+initiatives such as Software Carpentry~\cite{SoftwareCarpentry} have
+demonstrated how this could be achieved at scale.
 
@@
 %two presenters each summarized half of the submissions in this area.
-Across all talks and papers submitted, authors implicitly and explicitly
-recognized the concept of ``communities'' as a driving and unifying force
-within software projects.  Despite that, the actual nature of the communities,
-the incentive structures and bind them together, the infrastructure they
-utilize for communication and even the types of individuals that make up those
-communities were different.  Some of these communities were primarily composed
-of members of industry, some were funded and driven by individuals focused on
-developing software as opposed to utilizing it, and others were primarily
-composed of scientist practitioners, and were true communities of practice.
+Across all talks and papers submitted, authors implicitly and
+explicitly recognized the concept of ``communities'' as a driving and
+unifying force within software projects.  Despite that, the actual
+nature of the communities, the incentive structures and bind them
+together, the infrastructure they utilize for communication and even
+the types of individuals that make up those communities were
+different.  Some of these communities were primarily composed of
+members of industry, some were funded and driven by individuals
+focused on developing software as opposed to utilizing it, and others
+were primarily composed of scientist practitioners, and were true
+communities of practice.
 
 These varying structures and compositions result in differing modes of
-interaction within communities, styles of development, and the structure of
-planning for future development of software.  In this section, we summarize
-the different types of communities in scientific software as well as the
-resultant impact on sustainability and development of functionality.
+interaction within communities, styles of development, and the
+structure of planning for future development of software.  In this
+section, we summarize the different types of communities in scientific
+software as well as the resultant impact on sustainability and
+development of functionality.
 
@@
 
-Drawing on experiences from high-energy physics, Vay et al.~\cite{Vay_WSSSPE} proposed
-developing teams of technical specialists able to overcome a lack of
-coordination between projects.  Maximizing scientific output requires
-maximizing the usability of the scientific codes while minimizing the cost of
-developing and supporting those codes.  This included targeting different
-architectures for their software to be deployed, as well as coordination
-between technically-focused individuals and usage of a common scripting
-language between projects.  Instead of fragmenting the development of
-simulation codes across institutions, the paper suggests that a cohesive
-strategy reducing duplication and increasing coordination will broadly increase
-the efficiency across institutions.  The approach proposed is of de-fragmenting
-the existing ecosystem in a non-disruptive way.
+Drawing on experiences from high-energy physics, Vay et
+al.~\cite{Vay_WSSSPE} proposed developing teams of technical
+specialists able to overcome a lack of coordination between projects.
+Maximizing scientific output requires maximizing the usability of the
+scientific codes while minimizing the cost of developing and
+supporting those codes.  This included targeting different
+architectures for their software to be deployed, as well as
+coordination between technically-focused individuals and usage of a
+common scripting language between projects.  Instead of fragmenting
+the development of simulation codes across institutions, the paper
+suggests that a cohesive strategy reducing duplication and increasing
+coordination will broadly increase the efficiency across institutions.
+The approach proposed is of de-fragmenting the existing ecosystem in a
+non-disruptive way.
 
 Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
-catalysts'' and their role in the modern scientific research process. A technology
-catalyst is an individual with knowledge of
+catalysts'' and their role in the modern scientific research
+process. A technology catalyst is an individual with knowledge of
 technological advancements, tasked with user engagement to create
-scientific or engineering applications, using suitable tools and techniques
-to take advantage of current technological capabilities.
+scientific or engineering applications, using suitable tools and
+techniques to take advantage of current technological capabilities.
 %
-One of the tasks of technology catalysts is to seek community collaborations for new
-applications and engage users, thus benefiting both science, by effective
-running of scientific codes on computational infrastructure, and technology,
-by conducting research and seeking findings for technology improvement.
-The particular engagements described in the paper came up from the lead author's work as
-a postdoctoral researcher at Cornell and Argonne, where interaction with the scientific
-communities in both institutions resulted in these collaborations.
-
-At NESCent, a combination of in-house informatics individuals and domain
-scientists collaborate to develop software to study evolutionary science.  The
-report~\cite{Cranston_WSSSPE} studied the success of a ``hackathon'' model
-for development, where short-form, hands-on events combining users,
-researcher-developers and software engineers targeted specific code
-improvements. From this experiment, the authors identified several key
-outcomes as well as lessons-learned: specifically, the co-localization of
-developers was seen as having a strong impact, enabling casual conversation
-that led to discrete outcomes.  The formation of the discussion mailing list,
-specifically in response to the social capital built at the hackathon, was seen
-as spurring on longer-term benefits to the community and fostering
-sustainability.
-
-Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci project in developing
-collaborations supporting tool development for Open Science.  This software
-collective, organized around the statistical programming environment R,
-develops access mechanisms for data repositories and attempts to reduce the
-barrier to entry for individuals wanting to access data repositories and study
-the data contained therein.  The collective fosters direct collaboration
-between individuals and data providers, designed to ``train academics in
-reproducible science workflows focused around R.''  Two central challenges
-are engagement of existing users within ecology and
-evolutionary biology (EEB), and how the community can make inroads and traction
-in other disciplines. Currently, the collective is exploring addressing these
-challenges through use of social media, holding workshops, and hackathons.
-This helps to both raise the profile of the collective within EEB and in other
-domains.  However, the overarching challenge identified in the paper was that
-of incentivizing maintenance of software, which is difficult in academia.
+One of the tasks of technology catalysts is to seek community
+collaborations for new applications and engage users, thus benefiting
+both science, by effective running of scientific codes on
+computational infrastructure, and technology, by conducting research
+and seeking findings for technology improvement.  The particular
+engagements described in the paper came up from the lead author's work
+as a postdoctoral researcher at Cornell and Argonne, where interaction
+with the scientific communities in both institutions resulted in these
+collaborations.
+
+At NESCent, a combination of in-house informatics individuals and
+domain scientists collaborate to develop software to study
+evolutionary science.  The report~\cite{Cranston_WSSSPE} studied the
+success of a ``hackathon'' model for development, where short-form,
+hands-on events combining users, researcher-developers and software
+engineers targeted specific code improvements. From this experiment,
+the authors identified several key outcomes as well as
+lessons-learned: specifically, the co-localization of developers was
+seen as having a strong impact, enabling casual conversation that led
+to discrete outcomes.  The formation of the discussion mailing list,
+specifically in response to the social capital built at the hackathon,
+was seen as spurring on longer-term benefits to the community and
+fostering sustainability.
+
+Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
+project in developing collaborations supporting tool development for
+Open Science.  This software collective, organized around the
+statistical programming environment R, develops access mechanisms for
+data repositories and attempts to reduce the barrier to entry for
+individuals wanting to access data repositories and study the data
+contained therein.  The collective fosters direct collaboration
+between individuals and data providers, designed to ``train academics
+in reproducible science workflows focused around R.''  Two central
+challenges are engagement of existing users within ecology and
+evolutionary biology (EEB), and how the community can make inroads and
+traction in other disciplines. Currently, the collective is exploring
+addressing these challenges through use of social media, holding
+workshops, and hackathons.  This helps to both raise the profile of
+the collective within EEB and in other domains.  However, the
+overarching challenge identified in the paper was that of
+incentivizing maintenance of software, which is difficult in academia.
 
@@
 
-Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree to which research relies on high
-quality software. There are often barriers and a lack of suitable incentives
-for researchers to embrace software engineering principles. The Water Science
-Software Institute is working to lower some of these barriers through an Open
-Community Engagement Process. This is a 4-step iterative development process
+Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree
+to which research relies on high quality software. There are often
+barriers and a lack of suitable incentives for researchers to embrace
+software engineering principles. The Water Science Software Institute
+is working to lower some of these barriers through an Open Community
+Engagement Process. This is a 4-step iterative development process
 that incorporates Agile development principles.
@@
 
-Christopherson reports on the application of Steps 1-3 to a computational
-modeling framework developed in the 1990s. Step 1 was a 2-day, in-person
-specifications meeting and code walk-through. Step 2 was a 5-day hackathon to
-develop working code, and Step 3 was a 3-day hackathon to refine the code based
-on new requirements. The team worked on small, low-risk units of code. It was
-challenging, revealed unanticipated obstacles, programmers had to work
-together, and experimentation was encouraged.
+Christopherson reports on the application of Steps 1-3 to a
+computational modeling framework developed in the 1990s. Step 1 was a
+2-day, in-person specifications meeting and code walk-through. Step 2
+was a 5-day hackathon to develop working code, and Step 3 was a 3-day
+hackathon to refine the code based on new requirements. The team
+worked on small, low-risk units of code. It was challenging, revealed
+unanticipated obstacles, programmers had to work together, and
+experimentation was encouraged.
 %
-The paper recommended: start small and gradually building toward more complex
-objectives, consistent with Agile development; develop consensus before coding,
-by repeating step 1 before all hackathons; ensure newcomers receive orientation
-prior to the hackathon, such as a code walk-through or system documentation;
-and co-locate collaborators whenever feasible.
+The paper recommended: start small and gradually building toward more
+complex objectives, consistent with Agile development; develop
+consensus before coding, by repeating step 1 before all hackathons;
+ensure newcomers receive orientation prior to the hackathon, such as a
+code walk-through or system documentation; and co-locate collaborators
+whenever feasible.
 
@@
 
-Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
-entry to complex cyberinfrastructure. For example, more than 7,000
-biologists have run phylogenetic codes on supercomputers using the \toolname{CIPRES}
-Science Gateway in 3 1/2 years. Over 120 scientists from 50 institutions used the
-\toolname{UltraScan Science Gateway} in one year, increasing the sophistication of analytical
-ultracentrifugation experiments worldwide. The new Neuroscience Gateway (\toolname{NSG})
-registered 100+ users who used 250,000 CPU hours in only a few months.
-
-Gateways, however, need to keep operational costs low and can often make use of
-common components, such as authentication, application installation and
-reliable execution and help desk support. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
-middleware as a hosted, scalable third-party service while domain developers
+Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can
+provide a user-friendly entry to complex cyberinfrastructure. For
+example, more than 7,000 biologists have run phylogenetic codes on
+supercomputers using the \toolname{CIPRES} Science Gateway in 3 1/2
+years. Over 120 scientists from 50 institutions used the
+\toolname{UltraScan Science Gateway} in one year, increasing the
+sophistication of analytical ultracentrifugation experiments
+worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
+100+ users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often
+make use of common components, such as authentication, application
+installation and reliable execution and help desk support. Science
+Gateway Platform as a Service (\toolname{SciGaP}) delivers middleware
+as a hosted, scalable third-party service while domain developers
 focus on user interfaces and domain-specific data in the gateway.
-While \toolname{SciGaP} is based on the \toolname{Apache Airavata} project
-and the \toolname{CIPRES} Workbench Framework, community contributions are encouraged
-by its open source, open governance and open operations policies. The
-goal is robust, sustainable infrastructure with a cycle of development that
-improves reliability and prioritizes stakeholder requirements. The project is
-leveraging Internet2's \toolname{Net+} mechanisms for converting \toolname{SciGaP} and its gateways
-into commodity services.
-
-Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
-\toolname{nanoHUB.org} community, where community is defined as a ``body of persons of
-common and especially professional interests scattered through a larger
-society.'' Support is challenging because of the diversity of
-viewpoints and needs. The group constantly examines its policies to determine
-whether they are indirectly alienating part of the community or encouraging
-particular types of use.
+While \toolname{SciGaP} is based on the \toolname{Apache Airavata}
+project and the \toolname{CIPRES} Workbench Framework, community
+contributions are encouraged by its open source, open governance and
+open operations policies. The goal is robust, sustainable
+infrastructure with a cycle of development that improves reliability
+and prioritizes stakeholder requirements. The project is leveraging
+Internet2's \toolname{Net+} mechanisms for converting
+\toolname{SciGaP} and its gateways into commodity services.
+
+Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and
+challenges with the large \toolname{nanoHUB.org} community, where
+community is defined as a ``body of persons of common and especially
+professional interests scattered through a larger society.'' Support
+is challenging because of the diversity of viewpoints and needs. The
+group constantly examines its policies to determine whether they are
+indirectly alienating part of the community or encouraging particular
+types of use.
 %
-\toolname{nanoHUB}'s 10-year history with over 260,000 users annually provides a lot of
-data to analyze. 4000 resources contributed by 1000 authors. \toolname{nanoHUB} serves
-both the research and education community and the contribution model allows
-researchers to get their codes out into the community and in use in education
-very rapidly. The primary software challenges are twofold --- support for the
-\toolname{HUBzero} framework and challenges related to the software contributed by the
-community.
-
-The group has learned that community contributions are maximized with a
-tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3 license so contributors can
-create unique components and license as they choose. If they make changes to
-source code, the original license must be maintained for redistribution. As far
-as contributed resources, these must be open access, but not necessarily open
-source. This allows contributors to meet the requirements of their institutions
-and funding agencies. Quality is maintained via user ratings. Documentation is
-encouraged and nanoHUB supplies regression test capabilities, but the user
-community provides ratings, poses questions and contributes to wishlists and
-citation counts, all of which incentivize code authors.
-
-Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
-two major efforts: the SciPy conference and the NumFOCUS foundation.
-Since software sustainability relies on contributions from all sectors of the user
-community, these efforts support these sectors, and help develop and mature Python.
+\toolname{nanoHUB}'s 10-year history with over 260,000 users annually
+provides a lot of data to analyze. 4000 resources contributed by 1000
+authors. \toolname{nanoHUB} serves both the research and education
+community and the contribution model allows researchers to get their
+codes out into the community and in use in education very rapidly. The
+primary software challenges are twofold --- support for the
+\toolname{HUBzero} framework and challenges related to the software
+contributed by the community.
+
+The group has learned that community contributions are maximized with
+a tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3
+license so contributors can create unique components and license as
+they choose. If they make changes to source code, the original license
+must be maintained for redistribution. As far as contributed
+resources, these must be open access, but not necessarily open
+source. This allows contributors to meet the requirements of their
+institutions and funding agencies. Quality is maintained via user
+ratings. Documentation is encouraged and nanoHUB supplies regression
+test capabilities, but the user community provides ratings, poses
+questions and contributes to wishlists and citation counts, all of
+which incentivize code authors.
+
+Terrel~\cite{Terrel_WSSSPE} describes support for the Python
+scientific community through two major efforts: the SciPy conference
+and the NumFOCUS foundation.  Since software sustainability relies on
+contributions from all sectors of the user community, these efforts
+support these sectors, and help develop and mature Python.
 %
-The reliance
-on software in science has driven a huge demand for development, but this
-development is typically done as a side effort and often in a rush to publish
-without documentation and testing. While the software is often created by academics, software support
-can fall to industrial institutions. SciPy brings together industry,
-government, and academics to share their code and experience in an open
-environment.
+The reliance on software in science has driven a huge demand for
+development, but this development is typically done as a side effort
+and often in a rush to publish without documentation and
+testing. While the software is often created by academics, software
+support can fall to industrial institutions. SciPy brings together
+industry, government, and academics to share their code and experience
+in an open environment.
 %
-NumFOCUS in a non-profit that promotes open, usable scientific software while sustaining the
-community through educational programs; collaborative
-research tools and documentation; and promotion of high-level languages,
-reproducible scientific research, and open-code development. Governance
-is a loose grantor-grantee relationship with projects, allowing funds
-to be placed in the groups' accounts. This has raised money to hire
-developers for open code development, maintain testing machines, organize the
-PyData conference series, and sponsor community members to attend conferences.
+NumFOCUS in a non-profit that promotes open, usable scientific
+software while sustaining the community through educational programs;
+collaborative research tools and documentation; and promotion of
+high-level languages, reproducible scientific research, and open-code
+development. Governance is a loose grantor-grantee relationship with
+projects, allowing funds to be placed in the groups' accounts. This
+has raised money to hire developers for open code development,
+maintain testing machines, organize the PyData conference series, and
+sponsor community members to attend conferences.
 
 L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
- which was started in 1996 by
-participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh (core)
-and thorns (modules) model, a community-oriented framework that allows
-researchers to easily work together with reusable and extendable software
-elements. Modules are compiled into an executable and can remain dormant,
-becoming active only when parameters and simulation data dictate. Users can
-use modules written by others or can write their own modules without changing
-other code. The community has grown and diversified beyond the original science
-areas.
-
-The paper points out 4 keys to sustaining the community: modular design,
-growing a collaborative community, career paths and credit. In modular design,
-the Cactus project went far beyond standard practices of APIs. Domain specific
-languages (DSLs) allow decoupling of components --- for example I/O, mesh
-refinement, PAPI counters, and boundary conditions abstracted from science
-code. In academia, publications are the main currency of credit. Because the
-project connects code developments to science, the work is publishable and
-modules are citable. Because of the open source, modular approach, programmers
-can see the impact of their contributions and often continue work after
-graduation. Career paths remain a challenge, however. Tasks that are essential
-from a software engineering perspective are often not rewarded in academia. The
-best programmers in a science environment often have multidisciplinary
-expertise. This also is not rewarded in academia.
-
-Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
-community of those creating science gateways for science. These
-gateways, as described in some of the other papers in this section, can be
-quite capable and can have strong scientific impact.
-Challenges are similar to those highlighted by other papers in this section:
-the conflict between funding for research vs infrastructure and the
-challenges around getting academic credit for infrastructure.
+which was started in 1996 by participants in USA Binary Black Hole
+Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
+model, a community-oriented framework that allows researchers to
+easily work together with reusable and extendable software
+elements. Modules are compiled into an executable and can remain
+dormant, becoming active only when parameters and simulation data
+dictate. Users can use modules written by others or can write their
+own modules without changing other code. The community has grown and
+diversified beyond the original science areas.
+
+The paper points out 4 keys to sustaining the community: modular
+design, growing a collaborative community, career paths and credit. In
+modular design, the Cactus project went far beyond standard practices
+of APIs. Domain specific languages (DSLs) allow decoupling of
+components --- for example I/O, mesh refinement, PAPI counters, and
+boundary conditions abstracted from science code. In academia,
+publications are the main currency of credit. Because the project
+connects code developments to science, the work is publishable and
+modules are citable. Because of the open source, modular approach,
+programmers can see the impact of their contributions and often
+continue work after graduation. Career paths remain a challenge,
+however. Tasks that are essential from a software engineering
+perspective are often not rewarded in academia. The best programmers
+in a science environment often have multidisciplinary expertise. This
+also is not rewarded in academia.
+
+Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF
+software institute effort to build a community of those creating
+science gateways for science. These gateways, as described in some of
+the other papers in this section, can be quite capable and can have
+strong scientific impact.  Challenges are similar to those highlighted
+by other papers in this section: the conflict between funding for
+research vs infrastructure and the challenges around getting academic
+credit for infrastructure.
 %
-The authors observe that  development is
-often done in an isolated hobbyist environment. Developers are unable to take
-advantage of similar work done by others, isolation even when projects have
-common needs. But often projects struggle for sustainable funding because they
-provide infrastructure to conduct research and many times only the research is
-funded. Gateways also may start as a small group research project, taking off in
-popularity once they go live, without any long term plans for sustainability.
-Subsequent disruptions in
-service can limit effectiveness and test the limits of the research community's
-trust. %The impact of gateways can be increased substantially if we understand
+The authors observe that development is often done in an isolated
+hobbyist environment. Developers are unable to take advantage of
+similar work done by others, isolation even when projects have common
+needs. But often projects struggle for sustainable funding because
+they provide infrastructure to conduct research and many times only
+the research is funded. Gateways also may start as a small group
+research project, taking off in popularity once they go live, without
+any long term plans for sustainability.  Subsequent disruptions in
+service can limit effectiveness and test the limits of the research
+community's trust. %The impact of gateways can be increased
+substantially if we understand
 %what makes them successful.
 
-Recommendations from an early study of successful gateways include:
-1) Leadership and management teams should design governance to
-represent multiple strengths and perspectives, plan for change and turnover in
-the future, recruit a development team that understands both the technical and
-domain-related issues, consider sustainability and measure success early and
-often. 2) Projects should hire a team of
-professionals, demonstrate credibility through stability and clarity of
-purpose, leverage the work of others, and plan for flexibility. 3) Projects should
-identify an existing community and understand the communities' needs
-before beginning, then adapt as the communities' needs evolve. 4) Funders,
-should consider the technology project lifecycle, and design solicitations
-to reward effective planning, recognize the benefits and
-limitations of both technology innovation and reuse, expect adjustments during
-the production process, copy effective models from other industries and
-sectors, and encourage partnerships that support gateway sustainability.
-
-%Through a business incubator type approach, the institute plans to provide a
-%variety of services that could be shared amongst projects. Consultation and
-%resources on topics such as business plan development, software engineering
-%practices, software licensing options, usability, security and project
-%management as well as a software repository and hosting service will be
-%available. Experts will be available for multi-month assignments to help
-%research teams build their own gateways, teaching them how to maintain and add
-%to the work after the collaboration ends. Forums, symposia and an annual
-%conference will connect members of the development community. A modular,
-%layered framework that supports community contributions and allows developers
-%to choose components will be delivered and finally workforce development
-%activities will help train the next generation for careers in this cross-
-%disciplinary area and build pools of institutional expertise that many projects
-%can leverage. Shared services and forums can dramatically reduce the cost of
-%building and sustaining gateways. Workforce development can encourage
-%technology professionals to remain in the sciences.
+Recommendations from an early study of successful gateways include: 1)
+Leadership and management teams should design governance to represent
+multiple strengths and perspectives, plan for change and turnover in
+the future, recruit a development team that understands both the
+technical and domain-related issues, consider sustainability and
+measure success early and often. 2) Projects should hire a team of
+professionals, demonstrate credibility through stability and clarity
+of purpose, leverage the work of others, and plan for flexibility. 3)
+Projects should identify an existing community and understand the
+communities' needs before beginning, then adapt as the communities'
+needs evolve. 4) Funders, should consider the technology project
+lifecycle, and design solicitations to reward effective planning,
+recognize the benefits and limitations of both technology innovation
+and reuse, expect adjustments during the production process, copy
+effective models from other industries and sectors, and encourage
+partnerships that support gateway sustainability.
+
+%Through a business incubator type approach, the institute plans to
+%provide a variety of services that could be shared amongst
+%projects. Consultation and resources on topics such as business plan
+%development, software engineering practices, software licensing
+%options, usability, security and project management as well as a
+%software repository and hosting service will be available. Experts
+%will be available for multi-month assignments to help research teams
+%build their own gateways, teaching them how to maintain and add to
+%the work after the collaboration ends. Forums, symposia and an annual
+%conference will connect members of the development community. A
+%modular, layered framework that supports community contributions and
+%allows developers to choose components will be delivered and finally
+%workforce development activities will help train the next generation
+%for careers in this cross- disciplinary area and build pools of
+%institutional expertise that many projects can leverage. Shared
+%services and forums can dramatically reduce the cost of building and
+%sustaining gateways. Workforce development can encourage technology
+%professionals to remain in the sciences.
 
@@
 
-The workshop did not directly answer the question ``What are communities?'' but
-instead a number of different answers were indirectly presented, through the
-depiction of individuals and stakeholders in different aspects of the
-scientific software lifecycle.  In broad strokes, however, scientific software
-communities were generally accepted as consisting of individuals, often but not
-always composed of scientist practitioners, that were working with some degree
-of coordination toward a common goal enabled by software.
-
-The discussion of development-focused communities centered around describing
-methods of interaction between individuals and the scientific software.  The
-first type of interaction was the development of a specific piece of software,
-the second was a particular domain or discipline, and the final primary
-type of interaction was around the development of applications built on a
-particular piece of software that was perhaps developed by another group.  As
-an example, in~\cite{Hanwell_WSSSPE}, the community described is comprised of
-both the for-profit company \toolname{Kitware} and the users and contributors to their
-software packages such as \toolname{VTK}.  This structure, of the centralized development
-of core infrastructure around which communities of individuals applying that
-infrastructure and developing applications utilizing it, was similarly
-reflected in~\cite{Terrel_WSSSPE}, where the core scientific python ecosystem
-is supported by a non-profit entity that fosters community investment in that
-ecosystem.  In many ways, these two organizations (\toolname{Kitware} and \toolname{NumFOCUS})
-attempt to cross domain boundaries and provide support for both the
-infrastructure and application sides of community building.
+The workshop did not directly answer the question ``What are
+communities?'' but instead a number of different answers were
+indirectly presented, through the depiction of individuals and
+stakeholders in different aspects of the scientific software
+lifecycle.  In broad strokes, however, scientific software communities
+were generally accepted as consisting of individuals, often but not
+always composed of scientist practitioners, that were working with
+some degree of coordination toward a common goal enabled by software.
+
+The discussion of development-focused communities centered around
+describing methods of interaction between individuals and the
+scientific software.  The first type of interaction was the
+development of a specific piece of software, the second was a
+particular domain or discipline, and the final primary type of
+interaction was around the development of applications built on a
+particular piece of software that was perhaps developed by another
+group.  As an example, in~\cite{Hanwell_WSSSPE}, the community
+described is comprised of both the for-profit company
+\toolname{Kitware} and the users and contributors to their software
+packages such as \toolname{VTK}.  This structure, of the centralized
+development of core infrastructure around which communities of
+individuals applying that infrastructure and developing applications
+utilizing it, was similarly reflected in~\cite{Terrel_WSSSPE}, where
+the core scientific python ecosystem is supported by a non-profit
+entity that fosters community investment in that ecosystem.  In many
+ways, these two organizations (\toolname{Kitware} and
+\toolname{NumFOCUS}) attempt to cross domain boundaries and provide
+support for both the infrastructure and application sides of community
+building.
 
@@
 How might a project know when it has built a sustainable community?
-How might an outsider be able to assess the activity and sustainability of that community?
-These questions have been partially addressed in the literature.
-For example, a key metric in online communities in general is the cross-over point where there are more external contributors than internal ones.
-Richard Millington's book ``Buzzing Communities'' does an excellent job of outlining these measures, drawing on online communities research in an accessible manner~\cite{millington_buzzing_2012}.
-Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem~\cite{metrics-web}.
-%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please change to cite}
+How might an outsider be able to assess the activity and
+sustainability of that community?  These questions have been partially
+addressed in the literature.  For example, a key metric in online
+communities in general is the cross-over point where there are more
+external contributors than internal ones. Richard Millington's book
+``Buzzing Communities'' does an excellent job of outlining these
+measures, drawing on online communities research in an accessible
+manner~\cite{millington_buzzing_2012}. Some participants in the
+workshop have since prepared online materials outlining current and
+future practices in measurement of scientific software and its
+ecosystem~\cite{metrics-web}.
+%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please
+%change to cite}
 
@@
 
-Scientific software communities were viewed as a subset of software communities
-as whole.  As such, resources applicable to generic software communities --
-such as open source and proprietary technology companies -- can be used as
-input and as guiding understanding of how to steward and develop scientific
-software communities.  Because incentive structures are different in industry
-and volunteer-based open source communities, these can provide rough guidelines
-but not necessarily identically applicable.  The analogy between corporations
-and scientific investigators (particularly in terms of competition, cooperation
-and competitive advantage) has been explored in the literature below, but
-because of the different incentive structure the analogy is not universally
-true.
-
-The literature below, suggested by attendees, addresses both non-scientific
-software projects, as well as scientific projects.  The selections address both
-descriptive and prescriptive approaches to communities.
+Scientific software communities were viewed as a subset of software
+communities as whole.  As such, resources applicable to generic
+software communities -- such as open source and proprietary technology
+companies -- can be used as input and as guiding understanding of how
+to steward and develop scientific software communities.  Because
+incentive structures are different in industry and volunteer-based
+open source communities, these can provide rough guidelines but not
+necessarily identically applicable.  The analogy between corporations
+and scientific investigators (particularly in terms of competition,
+cooperation and competitive advantage) has been explored in the
+literature below, but because of the different incentive structure the
+analogy is not universally true.
+
+The literature below, suggested by attendees, addresses both
+non-scientific software projects, as well as scientific projects.  The
+selections address both descriptive and prescriptive approaches to
+communities.
 
 Both~\cite{howison_scientific_2011} and~\cite{howison_incentives_2013}
-study how scientific software collectives self-organize and address issues of
-incentive, long-term support, and development of infrastructure as well as new
-features.  As noted elsewhere in this summary,
-\cite{Turk:2013:SCH:2484762.2484782} shared prescriptions from two software
-communities in astrophysics.
-
-From the perspective of developing prescriptions for successful scientific
-software development, both~\cite{citeulike:11831265} and
-\cite{1749-4699-6-1-015010} share experiences and suggestions for developing
-sustainable practices.  \cite{citeulike:11831265} proposes "ten simple rules"
-for developing open source scientific software, focusing on both the choices
-made during development and the sustainability of practices in the long term.
-\cite{1749-4699-6-1-015010} describes the development and long-term growth of
-the \texttt{deal.II} library, and how its place in its ecosystem of libraries,
-applications and domains has shaped its development and community trajectory.
-
-From more traditional open source development, resources were shared that
-developed communities explicitly, such as~\cite{citeulike:7888211}
-and~\cite{Trapani:2011}, focusing on large-scale projects such as the Ubuntu
-Linux distribution and smaller-scale volunteer-developed projects like such as
-ThinkUp, respectively. The process of open source development, while less
-explicitly focused on community building, outlined in~\cite{citeulike:478633}
-was seen as a valuable resource, particularly when combined with the management
-and personal interaction techniques outlined in~\cite{opac-b1134063}.  Growing
-diversity in communities was directly addressed in~\cite{Allsopp:2012}, where
-experiences growing the diversity of technical conferences in open source were
-described.
+study how scientific software collectives self-organize and address
+issues of incentive, long-term support, and development of
+infrastructure as well as new features.  As noted elsewhere in this
+summary, \cite{Turk:2013:SCH:2484762.2484782} shared prescriptions
+from two software communities in astrophysics.
+
+From the perspective of developing prescriptions for successful
+scientific software development, both~\cite{citeulike:11831265} and
+\cite{1749-4699-6-1-015010} share experiences and suggestions for
+developing sustainable practices.  \cite{citeulike:11831265} proposes
+"ten simple rules" for developing open source scientific software,
+focusing on both the choices made during development and the
+sustainability of practices in the long term.
+\cite{1749-4699-6-1-015010} describes the development and long-term
+growth of the \texttt{deal.II} library, and how its place in its
+ecosystem of libraries, applications and domains has shaped its
+development and community trajectory.
+
+From more traditional open source development, resources were shared
+that developed communities explicitly, such
+as~\cite{citeulike:7888211} and~\cite{Trapani:2011}, focusing on
+large-scale projects such as the Ubuntu Linux distribution and
+smaller-scale volunteer-developed projects like such as ThinkUp,
+respectively. The process of open source development, while less
+explicitly focused on community building, outlined
+in~\cite{citeulike:478633} was seen as a valuable resource,
+particularly when combined with the management and personal
+interaction techniques outlined in~\cite{opac-b1134063}.  Growing
+diversity in communities was directly addressed
+in~\cite{Allsopp:2012}, where experiences growing the diversity of
+technical conferences in open source were described.
 
@@
 Several papers presented discussed the connection between industry and
-scientific software, from the perspective of both integrating efforts between
-the two and sustaining long-term development.
-
-Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of open
-source software development at \toolname{Kitware}.  In particular, they focus on their
-success at growing their community of users through enabling multiple channels
-of communication, directly reaching out to individuals, and lowering the
-barrier to entry for contributions. This involves providing clear,
-test-oriented and review-based mechanisms for evaluating contributions,
-permissive licenses, and a service-based model for sustaining development.
-This model enables \toolname{Kitware} to review both public funding, as well as private
-funding to support improvements and targeted developments of the software.
-
-Foster et al.~\cite{Foster_WSSSPE} discussed the approach of developing sustainability models
-around Software-as-a-Service (SaaS) platforms, with the target example being
-that of \toolname{GlobusOnline}.  The authors build a case that both grant-based and
-volunteer-based development fall short in sustaining software, resulting in
-software that is disproportionately difficult to use compared to its
-functionality, which they note directly impacts the overall scientific
-productivity of its users.  In contrast, by charging a subscription fee for
-hosted, centrally-managed software (similar to offerings by \toolname{Dropbox}, \toolname{EverNote},
-\toolname{GMail}), the authors propose to manage the funding cycle and enable a greater
-focus on the aspects of software that directly impact individuals, rather than
-funders.  \toolname{Globus} has deployed such a service, for which they have attempted to
-develop a sustainable economic model that reduces institutional obstacles to
-funding and subscription.  However, they do identify that cultural obstacles
-do still remain, and they note a particular difference in culture between NSF-
-and NIH-funded researchers.
-
-While many papers briefly discussed licensing issues, Elster~\cite{Elster_WSSSPE}
-directly discussed the impact of different licenses on obtaining industrial
-funding for scientific software projects.  In particular, she describes
-experiences with researchers unwilling to utilize GPL (copyleft) code, as it
-adds restrictions to reuse that some industries find unacceptable for future
-commercialization.  (This was discussed by Hanwell et al.~\cite{Hanwell_WSSSPE} as well.)
-Further obstacles to sustainability identified by the author include patenting
-of software, a lack of open access, and even more broadly, a lack of open
-source codes even in open access journals.  Finally, a lack of awareness on the
-part of scientific software developers of commodity libraries for common tasks
-reduces their ability to reuse code.
+scientific software, from the perspective of both integrating efforts
+between the two and sustaining long-term development.
+
+Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of
+open source software development at \toolname{Kitware}.  In
+particular, they focus on their success at growing their community of
+users through enabling multiple channels of communication, directly
+reaching out to individuals, and lowering the barrier to entry for
+contributions. This involves providing clear, test-oriented and
+review-based mechanisms for evaluating contributions, permissive
+licenses, and a service-based model for sustaining development.  This
+model enables \toolname{Kitware} to review both public funding, as
+well as private funding to support improvements and targeted
+developments of the software.
+
+Foster et al.~\cite{Foster_WSSSPE} discussed the approach of
+developing sustainability models around Software-as-a-Service (SaaS)
+platforms, with the target example being that of
+\toolname{GlobusOnline}.  The authors build a case that both
+grant-based and volunteer-based development fall short in sustaining
+software, resulting in software that is disproportionately difficult
+to use compared to its functionality, which they note directly impacts
+the overall scientific productivity of its users.  In contrast, by
+charging a subscription fee for hosted, centrally-managed software
+(similar to offerings by \toolname{Dropbox}, \toolname{EverNote},
+\toolname{GMail}), the authors propose to manage the funding cycle and
+enable a greater focus on the aspects of software that directly impact
+individuals, rather than funders.  \toolname{Globus} has deployed such
+a service, for which they have attempted to develop a sustainable
+economic model that reduces institutional obstacles to funding and
+subscription.  However, they do identify that cultural obstacles do
+still remain, and they note a particular difference in culture between
+NSF- and NIH-funded researchers.
+
+While many papers briefly discussed licensing issues,
+Elster~\cite{Elster_WSSSPE} directly discussed the impact of different
+licenses on obtaining industrial funding for scientific software
+projects.  In particular, she describes experiences with researchers
+unwilling to utilize GPL (copyleft) code, as it adds restrictions to
+reuse that some industries find unacceptable for future
+commercialization.  (This was discussed by Hanwell et
+al.~\cite{Hanwell_WSSSPE} as well.)  Further obstacles to
+sustainability identified by the author include patenting of software,
+a lack of open access, and even more broadly, a lack of open source
+codes even in open access journals.  Finally, a lack of awareness on
+the part of scientific software developers of commodity libraries for
+common tasks reduces their ability to reuse code.
 
@@
 
-The papers describing education and training were focused primarily on how
-these aspects of community development impact on the long-term sustainability
-of software projects.  \cite{Girotto_WSSSPE} described the impact of the
-mandate within the International Centre for Theoretical Physics (ICTP) to
-foster resources and competences in software development and HPC, resulting in
-the development of educational curricula directed in this area.  The paper
-itself described the changes made in these curricula as a result of the current
-changes in the HPC and scientific software landscape due to the advent of
-scripting languages, new programming paradigms and new types of hardware such
-as accelerator technologies.  The development of a workshop, with carefully
-selected participants and an immersive approach to learning, was identified as
-a major success for educating and developing new scientific software developers
-from targeted domains.
-
-\cite{Crawford_WSSSPE} addressed education and training within computational
-chemistry frameworks, particularly as these frameworks attempt to address
-next-generation computer hardware and software and as chemistry courses
-emphasize lab work over computational education.  The authors identify this
-lack of computational awareness and training as the primary challenge to future
-advances in computational chemistry.  The authors propose a new institute for
-computational chemistry, emphasizing collaboration (and a licensing structure,
-such as LGPL or more permissive) and education of future generations of
+The papers describing education and training were focused primarily on
+how these aspects of community development impact on the long-term
+sustainability of software projects.  \cite{Girotto_WSSSPE} described
+the impact of the mandate within the International Centre for
+Theoretical Physics (ICTP) to foster resources and competences in
+software development and HPC, resulting in the development of
+educational curricula directed in this area.  The paper itself
+described the changes made in these curricula as a result of the
+current changes in the HPC and scientific software landscape due to
+the advent of scripting languages, new programming paradigms and new
+types of hardware such as accelerator technologies.  The development
+of a workshop, with carefully selected participants and an immersive
+approach to learning, was identified as a major success for educating
+and developing new scientific software developers from targeted
+domains.
+
+\cite{Crawford_WSSSPE} addressed education and training within
+computational chemistry frameworks, particularly as these frameworks
+attempt to address next-generation computer hardware and software and
+as chemistry courses emphasize lab work over computational education.
+The authors identify this lack of computational awareness and training
+as the primary challenge to future advances in computational
+chemistry.  The authors propose a new institute for computational
+chemistry, emphasizing collaboration (and a licensing structure, such
+as LGPL or more permissive) and education of future generations of
 chemistry researchers.
@@
 \begin{quote}``When choosing software for procurement or
-development reuse --- regardless of the license and development model you will
-use --- you need to consider the future. While a software product may satisfy
-today's needs, will it satisfy tomorrow's needs? Will the supplier still be
-around in five years' time? Will the supplier still care for all its customers
-in five years' time? Will the supplier be responsive to bug reports and feature
-requests? In other words, is the software sustainable?''~\cite{OSS-ssmm-web}
+development reuse --- regardless of the license and development model
+you will use --- you need to consider the future. While a software
+product may satisfy today's needs, will it satisfy tomorrow's needs?
+Will the supplier still be around in five years' time? Will the
+supplier still care for all its customers in five years' time? Will
+the supplier be responsive to bug reports and feature requests? In
+other words, is the software sustainable?''~\cite{OSS-ssmm-web}
 \end{quote}
@@
 
-Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that could be
-used to measure software sustainability:
+Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that
+could be used to measure software sustainability:
 \begin{itemize}
-\item Will the software continue to provide the same functionality in the future,
-      even if the environment (other parts of the infrastructure) changes?
-\item Is the functionality and usability clearly explained to new users?
-\item Do users have a mechanism to ask questions and to learn about the software?
-\item Will the functionality be correct in the future, even if the environment changes?
-\item Does it provide the functionality that existing and future users want?
-\item Will it incorporate new science, theory, or tools as they develop?
+\item Will the software continue to provide the same functionality in
+  the future, even if the environment (other parts of the
+  infrastructure) changes?
+\item Is the functionality and usability clearly explained to new
+  users?
+\item Do users have a mechanism to ask questions and to learn about
+  the software?
+\item Will the functionality be correct in the future, even if the
+  environment changes?
+\item Does it provide the functionality that existing and future users
+  want?
+\item Will it incorporate new science, theory, or tools as they
+  develop?
 \end{itemize}
@@
 Table~\ref{tab:defining-sustainability} reproduces a summary created
-by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each contribution to the
-Modeling Sustainability panel what it meant when referring to
-software, how it defined software sustainability, and which
-approach it suggested to understand or evaluate sustainability.
+by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each
+contribution to the Modeling Sustainability panel what it meant when
+referring to software, how it defined software sustainability, and
+which approach it suggested to understand or evaluate sustainability.
 \begin{table}[t]
@@
 keynote (\S\ref{sec:keynote1}), where Phil Bourne used the term ``the
-Google Bus'' to describe the issue of well-trained software development staff in
-academic labs choosing to leave science and to work instead for technology
-firms, especially Google, which happens in large enough numbers that
-Google operates a bus every day to its nearest offices (and hence
-staff who leaves academia in this way does not even have  to physically move).
+Google Bus'' to describe the issue of well-trained software
+development staff in academic labs choosing to leave science and to
+work instead for technology firms, especially Google, which happens in
+large enough numbers that Google operates a bus every day to its
+nearest offices (and hence staff who leaves academia in this way does
+not even have to physically move).
 
@@
 
-% 9. Has the project proved to be sustainable so far? Does the developers,
-% managers make conscious efforts to make sure the project is sustainable? What
-% determines the sustainability of the project in question: science,
-% publications, usability, funding, reusability? Perhaps all factors to some
-% extent? What factors are dominating? Can we send out a message to current
-% scientific project to emphasize on one or more of these or other aspects?
-% ====
-
-In this section, we discuss some of the software projects as case studies to
-better understand the points discussed during the workshop and described in the
-previous sections. We dive deeper to find how are they affected by the
-sustainability related issues in practice. Additionally, we discuss issues
-surrounding people, especially the career paths of original developers and
-investigate to answer the question: \emph{are the original developers still
-around?}
-
-We classify the software projects discussed in the workshop in two broad
-categories. First, the \emph{utility} software, comprising of general purpose
-software. Utility software is often used as enabler and/or facilitator for the
-development of other tools and techniques to carry out scientific work. This
-includes the software developed to efficiently utilize new research
-infrastructures. Second, the \emph{scientific} software, comprising the
-software that was originally developed with an aim to solve a specific
-scientific problem.
+% 9. Has the project proved to be sustainable so far? Does the
+% developers, managers make conscious efforts to make sure the project
+% is sustainable? What determines the sustainability of the project in
+% question: science, publications, usability, funding, reusability?
+% Perhaps all factors to some extent? What factors are dominating? Can
+% we send out a message to current scientific project to emphasize on
+% one or more of these or other aspects?
+%====
+
+In this section, we discuss some of the software projects as case
+studies to better understand the points discussed during the workshop
+and described in the previous sections. We dive deeper to find how are
+they affected by the sustainability related issues in
+practice. Additionally, we discuss issues surrounding people,
+especially the career paths of original developers and investigate to
+answer the question: \emph{are the original developers still around?}
+
+We classify the software projects discussed in the workshop in two
+broad categories. First, the \emph{utility} software, comprising of
+general purpose software. Utility software is often used as enabler
+and/or facilitator for the development of other tools and techniques
+to carry out scientific work. This includes the software developed to
+efficiently utilize new research infrastructures. Second, the
+\emph{scientific} software, comprising the software that was
+originally developed with an aim to solve a specific scientific
+problem.
 
 % Why are we classifying software like above?
-This classification is motivated by our argument that the two kinds of software
-projects wildly vary in factors such as scope, purpose and usage. The
-development and management of each kind is significantly different.
-Consequently, the sustainability challenges faced by them differ and must be
-treated separately. For instance, the challenges faced by a gateway software
-development project such as \toolname{CIPRES} or a visualization software product
-such as \toolname{VisIT} are distinct to a niche software for \textit{ab initio}
-modeling and simulation such as \toolname{VASP} or \toolname{Quantum Espresso}.
+This classification is motivated by our argument that the two kinds of
+software projects wildly vary in factors such as scope, purpose and
+usage. The development and management of each kind is significantly
+different.  Consequently, the sustainability challenges faced by them
+differ and must be treated separately. For instance, the challenges
+faced by a gateway software development project such as
+\toolname{CIPRES} or a visualization software product such as
+\toolname{VisIT} are distinct to a niche software for \textit{ab
+  initio} modeling and simulation such as \toolname{VASP} or
+\toolname{Quantum Espresso}.
 
 \subsection{Utility Software}
-Software developed with a potentially wider audience and general purpose usage
-in mind is utility software. Utility software typically does not address
-fundamental research problems for a given scientific domain. Examples are
-collaborative development frameworks such as \toolname{GitHub} and
-\toolname{Bitbucket}, distributed workflow and generic computing frameworks
-such as \toolname{Galaxy}, \toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
-\toolname{Globus} and \toolname{VisTrails}, and visualization frameworks such as
-\toolname{VisIT}.
+Software developed with a potentially wider audience and general
+purpose usage in mind is utility software. Utility software typically
+does not address fundamental research problems for a given scientific
+domain. Examples are collaborative development frameworks such as
+\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
+generic computing frameworks such as \toolname{Galaxy},
+\toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
+\toolname{Globus} and \toolname{VisTrails}, and visualization
+frameworks such as \toolname{VisIT}.
 
 % What are the sustainability challenges? What are the opportunities?
-Development is often a high risk/reward undertaking exploring uncharted
-territories and is largely influenced by (re)usability factors. Owing to
-relatively large number of features, the development and prototype process is
-also lengthy which poses a significant survival risk. Challenges on a class of
-utility software for new architectures is well discussed
-in~\cite{Ferenbaugh_WSSSPE}.
-
-On the other hand, utility software presents opportunities to be usable by a
-larger community making its undertaking and development an attractive pursuit.
-It is generally more visible in community which in turn leads to a broader and
-deeper participation. For instance, promoting collaborations across the breadth
-(e.g., different departments) and depth (e.g., stakeholders within a department)
-of community, one of the key ingredient of a sustainable process. Successful
-utility projects reap high rewards and have a longer usage span. Development
-process becomes user-driven and self-sustaining.
-
-One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It follows agile software
-development practices and implements standards such as test-driven development
-and socialized bug managing practices via \toolname{trello}. Galaxy
-\emph{histories} and \emph{toolshed} offers easy community sharing of data and
-tools further promoting a collaborative environment. The project very closely
-follows the guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE} and many
-from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
-aiming better utilizing a particular, new infrastructure and architecture,
-e.g., \toolname{MVAPICH}, \toolname{VisIT}. Similarly, to leverage the power of
-emerging architectures such as accelerators, new code and libraries are
-required. The experience of one such effort as described
-in Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success but nonetheless
-with many invaluable lessons were learned about influential cultural and
-technical aspects in sustainable software development practices.
-
-A relatively new paradigm in utility software is the software delivered as
-service over the web. With increasing popularity of cloud-based storage and
-computational environments, many users are leaning towards tools used as
-services. \toolname{GitHub} and \toolname{Bitbucket} can be argued to be
-such tools, catering to collaborative development. For scientific users
-\toolname{Globus}-based tools are a case of service-based utility software
-discussed during the workshop. The data movement and sharing services offered
-by Globus can be easily used over the web by collaborating researchers.
-%
-%
-%\note{Dan: so is the only difference here that this category is aimed at new
-%systems, where utility software is not?  If so, perhaps this should be
-%incorporated into that subsection, so there would just be two subsections.
-%Something in the framing of this subsection doesn't feel right to me.  FL: I
-%respectfully disagree the the only reason to develop frameworks is to better
-%utilize (new or old) infrastructure. Another really important reason is to
-%modularize development, making it much easier for an individual developer/user
-%to integrate their own code. Ketan: I think in this section I wanted to bring
-%up a class of software projects which are undertaken with the main aim of
-%tackling new systems and architectures in mind. New infrastructures such as
-%GPU, Cloud, fast IB, federal science networks, etc. were on my mind as I see
-%many submissions around them in the workshop.}
-%
-%
+Development is often a high risk/reward undertaking exploring
+uncharted territories and is largely influenced by (re)usability
+factors. Owing to relatively large number of features, the development
+and prototype process is also lengthy which poses a significant
+survival risk. Challenges on a class of utility software for new
+architectures is well discussed in~\cite{Ferenbaugh_WSSSPE}.
+
+On the other hand, utility software presents opportunities to be
+usable by a larger community making its undertaking and development an
+attractive pursuit.  It is generally more visible in community which
+in turn leads to a broader and deeper participation. For instance,
+promoting collaborations across the breadth (e.g., different
+departments) and depth (e.g., stakeholders within a department) of
+community, one of the key ingredient of a sustainable
+process. Successful utility projects reap high rewards and have a
+longer usage span. Development process becomes user-driven and
+self-sustaining.
+
+One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
+follows agile software development practices and implements standards
+such as test-driven development and socialized bug managing practices
+via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
+offers easy community sharing of data and tools further promoting a
+collaborative environment. The project very closely follows the
+guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
+and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
+software projects are often developed aiming better utilizing a
+particular, new infrastructure and architecture, e.g.,
+\toolname{MVAPICH}, \toolname{VisIT}. Similarly, to leverage the power
+of emerging architectures such as accelerators, new code and libraries
+are required. The experience of one such effort as described in
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
+but nonetheless with many invaluable lessons were learned about
+influential cultural and technical aspects in sustainable software
+development practices.
+
+A relatively new paradigm in utility software is the software
+delivered as service over the web. With increasing popularity of
+cloud-based storage and computational environments, many users are
+leaning towards tools used as services. \toolname{GitHub} and
+\toolname{Bitbucket} can be argued to be such tools, catering to
+collaborative development. For scientific users
+\toolname{Globus}-based tools are a case of service-based utility
+software discussed during the workshop. The data movement and sharing
+services offered by Globus can be easily used over the web by
+collaborating researchers.
+% \note{Dan: so is the only difference here that this category is
+%aimed at new systems, where utility software is not?  If so, perhaps
+%this should be incorporated into that subsection, so there would just
+%be two subsections.  Something in the framing of this subsection
+%doesn't feel right to me.  FL: I respectfully disagree the the only
+%reason to develop frameworks is to better utilize (new or old)
+%infrastructure. Another really important reason is to modularize
+%development, making it much easier for an individual developer/user
+%to integrate their own code. Ketan: I think in this section I wanted
+%to bring up a class of software projects which are undertaken with
+%the main aim of tackling new systems and architectures in mind. New
+%infrastructures such as GPU, Cloud, fast IB, federal science
+%networks, etc. were on my mind as I see many submissions around them
+%in the workshop.}
 
 \subsection{Scientific software}
-Scientific software consists primarily of special-purpose software that was
-purpose-built for a target use-case scenario,  fixed/frozen requirements in
-mind or solving a specific problem. Software projects pertaining to specific
-scientific domain tend to be in a niche and the user community tends to be
-small to medium. They are mostly driven by the science and specific needs of a
-research group. Specific needs such as numerical accuracy and
-algorithmic optimization are some of the paramount requirements of most
-scientific software.
+Scientific software consists primarily of special-purpose software
+that was purpose-built for a target use-case scenario, fixed/frozen
+requirements in mind or solving a specific problem. Software projects
+pertaining to specific scientific domain tend to be in a niche and the
+user community tends to be small to medium. They are mostly driven by
+the science and specific needs of a research group. Specific needs
+such as numerical accuracy and algorithmic optimization are some of
+the paramount requirements of most scientific software.
 
@@
 Long-term sustainability of scientific software is often a significant
-challenge and face radically different issues compared to utility software.
-Many submissions reported that software is practically considered a ``byproduct''
-of the actual research. Others contended that the software was not the main
-funded part of their research. A smaller codebase and fixed requirements result
-in stability, ease of installation, and configurability.  Many such projects
-mature and are treated as libraries to be used into larger systems such as some
-of the utility software discussed in the previous section. While the software
-can stay stable and require relatively low maintenance, the responsibility is
-often on the shoulders of a very few developers who might not be specialists in
-software development. Development tends to be linear and simplistic with a
-limited scope to follow software best practices.
+challenge and face radically different issues compared to utility
+software.  Many submissions reported that software is practically
+considered a ``byproduct'' of the actual research. Others contended
+that the software was not the main funded part of their research. A
+smaller codebase and fixed requirements result in stability, ease of
+installation, and configurability.  Many such projects mature and are
+treated as libraries to be used into larger systems such as some of
+the utility software discussed in the previous section. While the
+software can stay stable and require relatively low maintenance, the
+responsibility is often on the shoulders of a very few developers who
+might not be specialists in software development. Development tends to
+be linear and simplistic with a limited scope to follow software best
+practices.
 
 Some examples of such software discussed as part of the workshop are,
-\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware}, \toolname{PETSc}, \toolname{MINRES-QLP},
-\toolname{FASTMath}.  Most of the above efforts are focused on one scientific
-and/or applied mathematics domain. However, sometimes such projects grow beyond
-the initial vision of developers. One such example is \toolname{Kitware}, which
-while being a software product specializing in the scientific process, has a
-core focus of developing communities around software processes. An instance of
-this process is the development of the \toolname{CMake} build utility, which
-started out as a building tool for \toolname{ITK} but grew to become a generic
-build utility for C++ projects. Similarly, \toolname{PETSc} is growing towards
-becoming a general purpose utility system usable for solving variety of
+\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware},
+\toolname{PETSc}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
+the above efforts are focused on one scientific and/or applied
+mathematics domain. However, sometimes such projects grow beyond the
+initial vision of developers. One such example is \toolname{Kitware},
+which while being a software product specializing in the scientific
+process, has a core focus of developing communities around software
+processes. An instance of this process is the development of the
+\toolname{CMake} build utility, which started out as a building tool
+for \toolname{ITK} but grew to become a generic build utility for C++
+projects. Similarly, \toolname{PETSc} is growing towards becoming a
+general purpose utility system usable for solving variety of
 scientific problems.
 
-Finally, it appears that the career options of developers from different kinds
-of software projects differ significantly. In order to further investigate the
-much discussed issue of career paths for developers of these software, we
-conducted a small survey of the cases discussed above with respect to the
-current status of its core developers. We looked at the workshop submissions
-and the original cited papers for the respective software. Where available, we
-looked at the web pages and source code of the earliest versions to find the
-original developers. We included projects older than 5 years and excluded
-relatively new projects such as \toolname{SciPy} and \toolname{numFOCUS} from
-this survey. We find that in a majority of scientific software, more than one
-original developer are still around, in some cases after more than 10 years
-(e.g., \toolname{R/qtl}).  Whereas, in the case of utility software projects, it
-appears that at least one original developer is still around, often in a
-part-time or ``benefactor'' role. In other cases, it appears that the original
-developers have went on to assume a larger role, with the same project.
-
-In conclusion, we find that there are distinctions in the characteristics and
-the kinds of challenges faced between utility and scientific software projects.
-In terms of career paths of original developers, it appears that the career
-choices for utility software developers are broader compared to the developers
-of scientific software. However, it appears that in the case of scientific
-software, core competencies of original developers tend to be domain science
-who happen to be the coders of their software. As projects as a whole, the
+Finally, it appears that the career options of developers from
+different kinds of software projects differ significantly. In order to
+further investigate the much discussed issue of career paths for
+developers of these software, we conducted a small survey of the cases
+discussed above with respect to the current status of its core
+developers. We looked at the workshop submissions and the original
+cited papers for the respective software. Where available, we looked
+at the web pages and source code of the earliest versions to find the
+original developers. We included projects older than 5 years and
+excluded relatively new projects such as \toolname{SciPy} and
+\toolname{numFOCUS} from this survey. We find that in a majority of
+scientific software, more than one original developer are still
+around, in some cases after more than 10 years (e.g.,
+\toolname{R/qtl}).  Whereas, in the case of utility software projects,
+it appears that at least one original developer is still around, often
+in a part-time or ``benefactor'' role. In other cases, it appears that
+the original developers have went on to assume a larger role, with the
+same project.
+
+In conclusion, we find that there are distinctions in the
+characteristics and the kinds of challenges faced between utility and
+scientific software projects.  In terms of career paths of original
+developers, it appears that the career choices for utility software
+developers are broader compared to the developers of scientific
+software. However, it appears that in the case of scientific software,
+core competencies of original developers tend to be domain science who
+happen to be the coders of their software. As projects as a whole, the
 small-scale, specific software have a lower risk and cost in terms of
-sustaining as compared to utility software. We also find that the utility
-software tend to adhere more loyally to the many best practices discussed
-during the workshop. Often the sustainability of scientific software projects
-is achieved by the fact that the core developer or team heavily utilizes the
-software for their science, e.g.,  \toolname{R/qtl, PETSc}.  Furthermore, the
-development of scientific software requires more scientific skills compared
-with those of utility software. This diminishes the need of specialist
-programmers and practices in many cases where the bulk of development is
-carried out by a domain scientist. With these inferences, we believe that
-separate guidelines and/or principles for sustainability be defined and
-practiced for each category of scientific software and they be given separate
-treatment.
+sustaining as compared to utility software. We also find that the
+utility software tend to adhere more loyally to the many best
+practices discussed during the workshop. Often the sustainability of
+scientific software projects is achieved by the fact that the core
+developer or team heavily utilizes the software for their science,
+e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
+scientific software requires more scientific skills compared with
+those of utility software. This diminishes the need of specialist
+programmers and practices in many cases where the bulk of development
+is carried out by a domain scientist. With these inferences, we
+believe that separate guidelines and/or principles for sustainability
+be defined and practiced for each category of scientific software and
+they be given separate treatment.
 
@@
 
-%\subsection{Concerns and issues}
-%
-%\todo{what are the main issues discussed in each discussion section? - and, if you want, and recommendations on what should be done about them}
+%\subsection{Concerns and issues} \todo{what are the main issues
+%discussed in each discussion section? - and, if you want, and
+%recommendations on what should be done about them}
 
@@
 \item The need for a definition of sustainability such that the
-community can get behind it. Although some had hoped that at least an
-initial consensus could be reached in the course of the workshop, this
-proved elusive. However, in the absence of such a definition it will
-remain difficult to define exactly what the goals should be towards
-achieving, or even only improving software sustainability, and hence
-what practices should be followed and implemented when. As described
-in the next subsection (\S\ref{sec:future}), the workshop organizers have begun an effort
-to address this.
-
-\item The need for academic
-career paths for scientific software developers.  
-Unfortunately, it is not clear how to ensure that these career paths become available,
-other than repeatedly talking about this issue. The recent Moore and Sloan initiative
-in data science~\cite{moore_sloan} are trying to address this, to some extent, by providing funds and
-incentives to universities in the US that work towards this goal.
+  community can get behind it. Although some had hoped that at least
+  an initial consensus could be reached in the course of the workshop,
+  this proved elusive. However, in the absence of such a definition it
+  will remain difficult to define exactly what the goals should be
+  towards achieving, or even only improving software sustainability,
+  and hence what practices should be followed and implemented when. As
+  described in the next subsection (\S\ref{sec:future}), the workshop
+  organizers have begun an effort to address this.
+
+\item The need for academic career paths for scientific software
+  developers.  Unfortunately, it is not clear how to ensure that these
+  career paths become available, other than repeatedly talking about
+  this issue. The recent Moore and Sloan initiative in data
+  science~\cite{moore_sloan} are trying to address this, to some
+  extent, by providing funds and incentives to universities in the US
+  that work towards this goal.
 
@@
 
-\item The need for recognition of scholarship in scientific software over research 
-articles. This need probably is the most addressed of the three, today, with
-efforts underway such as the Mozilla Science Lab, GitHub, and figshare
-``Code as a research object'' project~\cite{code_as_a_research_object} among others.
+\item The need for recognition of scholarship in scientific software
+  over research articles. This need probably is the most addressed of
+  the three, today, with efforts underway such as the Mozilla Science
+  Lab, GitHub, and figshare ``Code as a research object''
+  project~\cite{code_as_a_research_object} among others.
 
@@
 \item Use of shared repositories in the development of collaborative
-projects facilitates collaboration, reproducibility, and sustainability 
-in computational science. However it represents a barrier in some scientific fields and 
-has yet to be more widely adopted.
+  projects facilitates collaboration, reproducibility, and
+  sustainability in computational science. However it represents a
+  barrier in some scientific fields and has yet to be more widely
+  adopted.
 
@@
 %community can get behind it. Although some had hoped that at least an
-%initial consensus could be reached in the course of the workshop, this
-%proved elusive. However, in the absence of such a definition it will
-%remain difficult to define exactly what the goals should be towards
-%achieving, or even only improving software sustainability, and hence
-%what practices should be followed and implemented when. As described
-%in the next subsection, a subset of the organizers have begun an effort
-%to address this.
-
-%Thus, to be
-%seen successful, activities emerging from or following this first
-%WSSSPE workshop should include working out a community consensus on
-%what we mean by sustainability.
+%initial consensus could be reached in the course of the workshop,
+%this proved elusive. However, in the absence of such a definition it
+%will remain difficult to define exactly what the goals should be
+%towards achieving, or even only improving software sustainability,
+%and hence what practices should be followed and implemented when. As
+%described in the next subsection, a subset of the organizers have
+%begun an effort to address this.
+
+%Thus, to be seen successful, activities emerging from or following
+%this first WSSSPE workshop should include working out a community
+%consensus on what we mean by sustainability.
 
 %\subsubsection*{Communities, Models, and Education}
-\item A sustainability model for scientific software is to build a pipeline
-from construction to consumption, as found in the most efficient
-information technology enterprises.
+\item A sustainability model for scientific software is to build a
+  pipeline from construction to consumption, as found in the most
+  efficient information technology enterprises.
 
@@
 \subsubsection*{Use cases}
-Two distinct class of scientific software projects and products were discussed
-in the workshop: 1) generic, large-scale utility software and 2) niche, medium-
-and small-scale scientific software. Each class faces different and significant
-challenges. New undertakings should recognize the differences in advance and
-identify such challenges within the development and sustaining efforts. In
-particular, the dynamics associated with developers, scope, life cycle,
-users-community, (re)usability, funding support and career paths vary widely
-among the two classes of software.
+Two distinct class of scientific software projects and products were
+discussed in the workshop: 1) generic, large-scale utility software
+and 2) niche, medium- and small-scale scientific software. Each class
+faces different and significant challenges. New undertakings should
+recognize the differences in advance and identify such challenges
+within the development and sustaining efforts. In particular, the
+dynamics associated with developers, scope, life cycle,
+users-community, (re)usability, funding support and career paths vary
+widely among the two classes of software.
 
@@
 The WSSSPE workshop can be viewed as an experiment in how we can
-collaboratively and inclusively build a workshop agenda, without asking
-large numbers of people to submit papers that will be rejected so that the
-workshop can have a low acceptance rate.
-
-Contributors also want to get credit for their participation in the process.
-And the workshop organizers want to make sure that the workshop content and
-their efforts are recorded.  The methods used in the WSSSPE1 workshop
-were successful; we had good participation, many contributors have a
-report they can cite, and the record of the workshop is open and available,
-through the self-published reports, the workshop web site and notes site,
-and this paper.  In addition, many additional papers are being created that
-will include the discussions at the workshop, including extended versions
-of many of the self published reports. such as those that are in this
-special issue.
-
-Ideally, there would be a service that
-would be able to index the contributions to the workshop, serving the
-authors, the organizers, and the larger community.
+collaboratively and inclusively build a workshop agenda, without
+asking large numbers of people to submit papers that will be rejected
+so that the workshop can have a low acceptance rate.
+
+Contributors also want to get credit for their participation in the
+process.  And the workshop organizers want to make sure that the
+workshop content and their efforts are recorded.  The methods used in
+the WSSSPE1 workshop were successful; we had good participation, many
+contributors have a report they can cite, and the record of the
+workshop is open and available, through the self-published reports,
+the workshop web site and notes site, and this paper.  In addition,
+many additional papers are being created that will include the
+discussions at the workshop, including extended versions of many of
+the self published reports. such as those that are in this special
+issue.
+
+Ideally, there would be a service that would be able to index the
+contributions to the workshop, serving the authors, the organizers,
+and the larger community.
 
@@
 The organizers of the workshop have begun a survey to understand how
-the community define software sustainability.  It is expected that this survey
-will gather one or more consensus definitions, and lead to a short paper
-discussing them, as well as the level of consensus.
-
-Additional activities that are being planned include
-two additional WSSSPE workshops at the 2014 SciPy and SC14 conferences.
-The SciPy workshop (WSSSPE1.1) will focus on how some software projects
-from the SciPy community have dealt with software sustainability issues, both
-successfully and unsuccessfully, while the SC14 workshop (WSSSPE2) will
-be more general, and will likely focus on determining and publicizing specific
-activities that the overall scientific software community can take to move forward.
-In addition, there will be a minisymposium on ``Reliable Computational Science''
-at the 2014 Society for Industrial and Applied Mathematics  
-Annual Meeting (SIAM AN14, \url{http://meetings.siam.org})
-to further explore some of the key issues raised here.
+the community define software sustainability.  It is expected that
+this survey will gather one or more consensus definitions, and lead to
+a short paper discussing them, as well as the level of consensus.
+
+Additional activities that are being planned include two additional
+WSSSPE workshops at the 2014 SciPy and SC14 conferences.  The SciPy
+workshop (WSSSPE1.1) will focus on how some software projects from the
+SciPy community have dealt with software sustainability issues, both
+successfully and unsuccessfully, while the SC14 workshop (WSSSPE2)
+will be more general, and will likely focus on determining and
+publicizing specific activities that the overall scientific software
+community can take to move forward.  In addition, there will be a
+minisymposium on ``Reliable Computational Science'' at the 2014
+Society for Industrial and Applied Mathematics Annual Meeting (SIAM
+AN14, \url{http://meetings.siam.org}) to further explore some of the
+key issues raised here.
 
@@
 
-Some of the work by Katz was
-supported by the National Science Foundation while working at the
-Foundation; any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author and do not
-necessarily reflect the views of the National Science Foundation.
-
-Choi thanks Ian Foster, Director of the Computation Institute, University of Chicago,
-for encouraging her work in reliable reproducible research and
-supporting her trip to participate in the WSSSPE1 conference as a contributor.
-
-Lapp was supported by the National Evolutionary Synthesis Center (NESCent),
-NSF EF-0905606.
-
-Hetherington was supported by the UK Engineering and
-Physical Sciences Research Council (EPSRC) Grant EP/H043160/1 for the
-UK Software Sustainability Institute.
+Some of the work by Katz was supported by the National Science
+Foundation while working at the Foundation; any opinion, finding, and
+conclusions or recommendations expressed in this material are those of
+the author and do not necessarily reflect the views of the National
+Science Foundation.
+
+Choi thanks Ian Foster, Director of the Computation Institute,
+University of Chicago, for encouraging her work in reliable
+reproducible research and supporting her trip to participate in the
+WSSSPE1 conference as a contributor.
+
+Lapp was supported by the National Evolutionary Synthesis Center
+(NESCent), NSF EF-0905606.
+
+Hetherington was supported by the UK Engineering and Physical Sciences
+Research Council (EPSRC) Grant EP/H043160/1 for the UK Software
+Sustainability Institute.
 
@@
 
-\item Ketan Maheshwari, David Kelly, Scott J.
-Krieder, Justin M.
-Wozniak, Daniel S.
-Katz, Mei Zhi-Gang, Mainak Mookherjee.
-Reusability in Science: From Initial User Engagement to Dissemination of Results~\cite{Maheshwari_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott Chamberlain. rOpenSci -- a collaborative effort to develop R-based tools for facilitating Open Science~\cite{Hart_WSSSPE}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific Software through the Open Community Engagement Process~\cite{Christopherson_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar, Borries Demeler. Science Gateway Operational Sustainability: Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
+\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
+  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
+  Mookherjee. Reusability in Science: From Initial User Engagement to
+  Dissemination of Results~\cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
+  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
+  tools for facilitating Open Science~\cite{Hart_WSSSPE}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
+  Software through the Open Community Engagement
+  Process~\cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
+  Borries Demeler. Science Gateway Operational Sustainability:
+  Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
 
@@
 
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+\item Andy Terrel. Sustaining the Python Scientific Software
+  Community~\cite{Terrel_WSSSPE}
 
 \item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion Dooley, Dan Stanzione.
-Science Gateways and the Importance of Sustainability~\cite{Wilkins-Diehr_WSSSPE}
+  Schnetter. Cactus: Issues for Sustainable Simulation
+  Software~\cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
+  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
+  Dooley, Dan Stanzione. Science Gateways and the Importance of
+  Sustainability~\cite{Wilkins-Diehr_WSSSPE}
 
@@
 
-\item Anne C. Elster. Software for Science: Some Personal Reflections~\cite{Elster_WSSSPE}
+\item Anne C. Elster. Software for Science: Some Personal
+  Reflections~\cite{Elster_WSSSPE}
 
@@
   Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software Community~\cite{Terrel_WSSSPE}
+  Open Source Production Quality MPI Library for
+  HPC~\cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software
+  Community~\cite{Terrel_WSSSPE}
 
@@
 \item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+  Schnetter. Cactus: Issues for Sustainable Simulation
+  Software~\cite{Loffler_WSSSPE}
 
@@
 
-This workshop is interested in all of the above topics.  We invite short (4-page) position/experience reports that will be used
-to organize panel and discussion sessions. These papers will be
-archived by a third-party service, and provided DOIs [Digital Object Identifiers].
-We encourage
-submitters to license their papers under a Creative Commons license
-that encourages sharing and remixing, as we will combine ideas (with
-attribution) into the outcomes of the workshop.  An interactive site
-will be created to link these papers and the workshop discussion, with
-options for later comments and contributions. Contributions will be
-peer-reviewed for relevance and originality before the links are added
-to the workshop site; contributions will also be used to determine
-discussion topics and panelists. We will also plan one or more papers
-to be collaboratively developed by the contributors, based on the
-panels and discussions.''
+This workshop is interested in all of the above topics.  We invite
+short (4-page) position/experience reports that will be used to
+organize panel and discussion sessions. These papers will be archived
+by a third-party service, and provided DOIs [Digital Object
+  Identifiers].  We encourage submitters to license their papers under
+a Creative Commons license that encourages sharing and remixing, as we
+will combine ideas (with attribution) into the outcomes of the
+workshop.  An interactive site will be created to link these papers
+and the workshop discussion, with options for later comments and
+contributions. Contributions will be peer-reviewed for relevance and
+originality before the links are added to the workshop site;
+contributions will also be used to determine discussion topics and
+panelists. We will also plan one or more papers to be collaboratively
+developed by the contributors, based on the panels and discussions.''
 \end{quote}
@@
 The following is a partial list of attendees who were recorded on the
-document~\cite{WSSSPE1-google-notes} that was being used for live note taking at the workshop, or by the SC13 student volunteers, with some additions also made by the authors of this report.
+document~\cite{WSSSPE1-google-notes} that was being used for live note
+taking at the workshop, or by the SC13 student volunteers, with some
+additions also made by the authors of this report.
 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Some minor corrections of typos, sentences
Added in a few additional tool examples where they seemed appropriate.

@@
 licenses, and a service-based model for sustaining development.  This
-model enables \toolname{Kitware} to review both public funding, as
+model enables \toolname{Kitware} to receive both public funding, as
 well as private funding to support improvements and targeted
@@
 (\S\ref{sec:policy}). It quickly became clear that at present there is
-no consensus among the community, whether within or accross
+no consensus among the community, whether within or across
 disciplines, on what this definition should be, and that a variety of
-different defintiions were being assumed, used, or sometime expressly
+different definitions were being assumed, used, or sometime expressly
 articulated by contributors and attendees. However, some concepts,
@@
 the question was considered, albeit only briefly, whether proprietary
-versus open-source license plays a role in the context softwara
+versus open-source license plays a role in the context of software
 sustainability. For example, should a project ensure that it uses an
@@
 facets.  However, a career path in research means faculty at most
-universitities, and as was noted repeatedly in discussions, faculty
+universities, and as was noted repeatedly in discussions, faculty
 are hired based on their scientific qualifications, not on their
@@
 faced by a gateway software development project such as
-\toolname{CIPRES} or a visualization software product such as
-\toolname{VisIT} are distinct to a niche software for \textit{ab
-  initio} modeling and simulation such as \toolname{VASP} or
+\toolname{CIPRES} or visualization software products such as
+\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
+\textit{ab initio} modeling and simulation such as \toolname{VASP} or
 \toolname{Quantum Espresso}.
@@
 \toolname{Globus} and \toolname{VisTrails}, and visualization
-frameworks such as \toolname{VisIT}.
+frameworks such as \toolname{VTK}, \toolname{VisIT}, and \toolname{ParaView}.
 
@@
 particular, new infrastructure and architecture, e.g.,
-\toolname{MVAPICH}, \toolname{VisIT}. Similarly, to leverage the power
+\toolname{MVAPICH}, \toolname{VisIT}, toolname{ParaView}. Similarly, to leverage the power
 of emerging architectures such as accelerators, new code and libraries

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Very minor typo

@@
 large enough numbers that Google operates a bus every day to its
-nearest offices (and hence staff who leaves academia in this way does
+nearest offices (and hence staff who leaves academia in this way do
 not even have to physically move).

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Minor change.
Section 9: refine.

@@
                 \hline
-Katz and Proctor~\cite{Katz_WSSSPE}
+Katz, et al.~\cite{Katz_WSSSPE}
 & E-research infrastructures (i.e. cyberinfrastructure)
@@
 dynamics associated with developers, scope, life cycle,
-users-community, (re)usability, funding support and career paths vary
+users-community, (re)usability, funding support, and career paths vary
 widely among the two classes of software.
@@
 collaboratively and inclusively build a workshop agenda, without
-asking large numbers of people to submit papers that will be rejected
+asking a large number of people to submit papers that will be rejected
 so that the workshop can have a low acceptance rate.
@@
 workshop content and their efforts are recorded.  The methods used in
-the WSSSPE1 workshop were successful; we had good participation, many
-contributors have a report they can cite, and the record of the
-workshop is open and available, through the self-published reports,
-the workshop web site and notes site, and this paper.  In addition,
+the WSSSPE1 workshop were successful: we had good participation;  
+contributors have a report they can cite;   the record of the
+workshop is open and available through the self-published reports,
+the workshop website and notes site, and this paper.  In addition,
 many additional papers are being created that will include the
 discussions at the workshop, including extended versions of many of
-the self published reports. such as those that are in this special
+the self-published reports such as those that are in this special
 issue.
@@
 community can take to move forward.  In addition, there will be a
-minisymposium on ``Reliable Computational Science'' at the 2014
+two-session minisymposium on ``Reliable Computational Science'' at the 2014
 Society for Industrial and Applied Mathematics Annual Meeting (SIAM

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: minor changes.

@@
 studies to better understand the points discussed during the workshop
-and described in the previous sections. We dive deeper to find how are
-they affected by the sustainability related issues in
+and described in the previous sections. We dive deeper to find how 
+they are affected by the sustainability related issues in
 practice. Additionally, we discuss issues surrounding people,
@@
 general purpose software. Utility software is often used as enabler
-and/or facilitator for the development of other tools and techniques
+or facilitator for the development of other tools and techniques
 to carry out scientific work. This includes the software developed to
@@
 faced by a gateway software development project such as
-\toolname{CIPRES} or visualization software products such as
+\toolname{CIPRES} or visualization software pinroducts such as
 \toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
 \textit{ab initio} modeling and simulation such as \toolname{VASP} or
-\toolname{Quantum Espresso}.
+\toolname{Quantum Espresso}. \todo{cite}
 
@@
 uncharted territories and is largely influenced by (re)usability
-factors. Owing to relatively large number of features, the development
+factors. Owing to a relatively large number of features, the development
 and prototype process is also lengthy which poses a significant
@@
 departments) and depth (e.g., stakeholders within a department) of
-community, one of the key ingredient of a sustainable
+community, one of the key ingredients of a sustainable
 process. Successful utility projects reap high rewards and have a
@@
 offers easy community sharing of data and tools further promoting a
-collaborative environment. The project very closely follows the
+collaborative environment. The project closely follows the
 guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
@@
 particular, new infrastructure and architecture, e.g.,
-\toolname{MVAPICH}, \toolname{VisIT}, toolname{ParaView}. Similarly, to leverage the power
+\toolname{MVAPICH}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to leverage the power
 of emerging architectures such as accelerators, new code and libraries
@@
 Scientific software consists primarily of special-purpose software
-that was purpose-built for a target use-case scenario, fixed/frozen
-requirements in mind or solving a specific problem. Software projects
+that was purpose-built for a target use-case scenario, fixed 
+requirements in mind, or solving a specific problem. Software projects
 pertaining to specific scientific domain tend to be in a niche and the
@@
 installation, and configurability.  Many such projects mature and are
-treated as libraries to be used into larger systems such as some of
+treated as libraries to be integrated into larger systems such as some of
 the utility software discussed in the previous section. While the
 software can stay stable and require relatively low maintenance, the
-responsibility is often on the shoulders of a very few developers who
-might not be specialists in software development. Development tends to
+responsibility is often on the shoulders of a few developers who
+are often not specialists in software development. Development tends to
 be linear and simplistic with a limited scope to follow software best
@@
 \toolname{PETSc}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
-the above efforts are focused on one scientific and/or applied
+the above efforts are focused on one scientific or applied
 mathematics domain. However, sometimes such projects grow beyond the
@@
 process, has a core focus of developing communities around software
-processes. An instance of this process is the development of the
+processes. Another instance of this process is the development of the
 \toolname{CMake} build utility, which started out as a building tool
@@
 projects. Similarly, \toolname{PETSc} is growing towards becoming a
-general purpose utility system usable for solving variety of
+general purpose utility system usable for solving a variety of
 scientific problems.
@@
 Finally, it appears that the career options of developers from
-different kinds of software projects differ significantly. In order to
+two kinds of software projects differ significantly. In order to
 further investigate the much discussed issue of career paths for
@@
 in a part-time or ``benefactor'' role. In other cases, it appears that
-the original developers have went on to assume a larger role, with the
+the original developers have went on to assume a larger role with the
 same project.
@@
 is carried out by a domain scientist. With these inferences, we
-believe that separate guidelines and/or principles for sustainability
+believe that separate guidelines and principles for sustainability are to
 be defined and practiced for each category of scientific software and

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: make a small note.
Sections 1-3: Fiddling.

@@
 list of workshop attendees can be found in
-Appendix~\ref{sec:attendees}.
+Appendix~\ref{sec:attendees}. The original call for papers is included in 
+Appendix~\ref{sec:cfp}.
 
@@
 
-The WSSSPE1 workshop began with two keynote presentations.
+The WSSSPE1 workshop began with two keynote presentations, which 
+resonated with the audience and spurred a number of topics  
+discussed throughout the meeting.
 
@@
 organizations which reward those combining research and software
-expertise (the ``Google Bus''). However, he is a strong fan of
+expertise (the ``Google Bus''). However, he is a strong supporter of
 software sustainability through public-private partnerships. He noted
@@
 last, an open source project needs a minimal funding requirement even
-with a vibrant community---goodwill only goes so far if one is being
+with a vibrant community --- goodwill only goes so far if one is being
 paid to do something else.  He talked about grant schemes of relevance
@@
 sharing these data among scientists. Consequently many researchers
-repeated the same calculations. Smith estimated that 13 person-years
+repeated the same calculations. Smith estimated that plausibly 13 person-years
 were wasted by this repetition.
@@
 of scientific projects using GitHub is the EMCEE Markov Chain Monte
-Carlo project~\cite{emcee-web} by Dan Foreman-Mackey and contributors.
+Carlo project~\cite{emcee-web} developed by Dan Foreman-Mackey and contributors.
 
@@
 \textit{ab initio} modeling and simulation such as \toolname{VASP} or
-\toolname{Quantum Espresso}. \todo{cite}
+\toolname{Quantum Espresso}. \note{cite SW}
 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
finishing authors

@@
 Ketan~Maheshwari$^{(4)}$,
-Frank~Löffler$^{(5)}$,
+Frank~L\"{o}ffler$^{(5)}$,
 Matthew~Turk$^{(6)}$,
@@
 James~Howison$^{(10)}$,
-%Shel~Swenson$^{(11)}$,
-%Gabrielle~D.~Allen$^{(12)}$,
+Shel~Swenson$^{(11)}$,
+Gabrielle~D.~Allen$^{(12)}$,
 Bruce~Berriman$^{(11)}$
-%Colin~Venters$^{(14)}$
+Colin~Venters$^{(14)}$
 }
@@
 %
-%\thanks{{}$^{(11)}$ xxx}
+\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA; \url{gdallen@illinois.edu}}
 %
-%\thanks{{}$^{(12)}$ xxx}
+\thanks{{}$^{(12)}$ University of Southern California, Los Angeles, CA, USA; \url{mdswenso@usc.edu}}
 %
@@
 %
-%\thanks{{}$^{(14)}$ xxx}
+\thanks{{}$^{(14)}$ University of Huddersfield, Huddersfield, West Yorkshire, UK; \url{C.Venters@hud.ac.uk}}
 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 6: refine.

@@
 unifying force within software projects.  Despite that, the actual
-nature of the communities, the incentive structures and bind them
+nature of the communities, the incentive structures that bind them
 together, the infrastructure they utilize for communication and even
@@
 al.~\cite{Vay_WSSSPE} proposed developing teams of technical
-specialists able to overcome a lack of coordination between projects.
+specialists   to overcome a lack of coordination between projects.
 Maximizing scientific output requires maximizing the usability of the
-scientific codes while minimizing the cost of developing and
-supporting those codes.  This included targeting different
+scientific code while minimizing the cost of developing and
+supporting those code.  This included targeting different
 architectures for their software to be deployed, as well as
@@
 Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
-project in developing collaborations supporting tool development for
+project in developing collaboration supporting tools for
 Open Science.  This software collective, organized around the
@@
 traction in other disciplines. Currently, the collective is exploring
-addressing these challenges through use of social media, holding
-workshops, and hackathons.  This helps to both raise the profile of
+addressing these challenges through the use of social media, holding
+workshops and hackathons.  This helps to both raise the profile of
 the collective within EEB and in other domains.  However, the
@@
 is working to lower some of these barriers through an Open Community
-Engagement Process. This is a 4-step iterative development process
+Engagement Process. This is a four-step iterative development process
 that incorporates Agile development principles.
@@
 
-Christopherson reports on the application of Steps 1-3 to a
-computational modeling framework developed in the 1990s. Step 1 was a
+Christopherson reports on the application of Steps~1-3 to a
+computational modeling framework developed in the 1990s. Step~1 was a
 2-day, in-person specifications meeting and code walk-through. Step 2
@@
 complex objectives, consistent with Agile development; develop
-consensus before coding, by repeating step 1 before all hackathons;
+consensus before coding, by repeating step~1 before all hackathons;
 ensure newcomers receive orientation prior to the hackathon, such as a
@@
 
-Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can
+Pierce et al.~\cite{Pierce2_WSSSPE} describes how science gateways can
 provide a user-friendly entry to complex cyberinfrastructure. For
 example, more than 7,000 biologists have run phylogenetic codes on
-supercomputers using the \toolname{CIPRES} Science Gateway in 3 1/2
+supercomputers using the CIPRES Science Gateway in $3 \tfrac{1}{2}$
 years. Over 120 scientists from 50 institutions used the
@@
 worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
-100+ users who used 250,000 CPU hours in only a few months.
+more than 100 users who used 250,000 CPU hours in only a few months.
 
@@
 \toolname{nanoHUB}'s 10-year history with over 260,000 users annually
-provides a lot of data to analyze. 4000 resources contributed by 1000
+provides a lot of data to analyze: 4000 resources contributed by 1000
 authors. \toolname{nanoHUB} serves both the research and education
@@
 L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
-which was started in 1996 by participants in USA Binary Black Hole
+which was started in 1996 by participants in the USA Binary Black Hole
 Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
@@
 
-The paper points out 4 keys to sustaining the community: modular
-design, growing a collaborative community, career paths and credit. In
+The paper points out four keys to sustaining the community: modular
+design, growing a collaborative community, career paths, and credit. In
 modular design, the Cactus project went far beyond standard practices
@@
 hobbyist environment. Developers are unable to take advantage of
-similar work done by others, isolation even when projects have common
-needs. But often projects struggle for sustainable funding because
+similar work done by others, finding themselves in isolation even when their projects have common
+goals. But often projects struggle for sustainable funding because
 they provide infrastructure to conduct research and many times only
@@
 community's trust. %The impact of gateways can be increased
-substantially if we understand
+%substantially if we understand
 %what makes them successful.
@@
 of purpose, leverage the work of others, and plan for flexibility. 3)
-Projects should identify an existing community and understand the
+Projects should identify one ore more existing communities and understand the
 communities' needs before beginning, then adapt as the communities'
@@
 ``Buzzing Communities'' does an excellent job of outlining these
-measures, drawing on online communities research in an accessible
+measures, drawing on communities research in an accessible
 manner~\cite{millington_buzzing_2012}. Some participants in the
-workshop have since prepared online materials outlining current and
+workshop have since prepared materials outlining current and
 future practices in measurement of scientific software and its
@@
 respectively. The process of open source development, while less
-explicitly focused on community building, outlined
+explicitly focused on community building, sketched
 in~\cite{citeulike:478633} was seen as a valuable resource,

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
some fixes

@@
 faced by a gateway software development project such as
-\toolname{CIPRES} or visualization software pinroducts such as
+\toolname{CIPRES} or visualization software products such as
 \toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
@@
 
-On the other hand, utility software presents opportunities to be
-usable by a larger community making its undertaking and development an
-attractive pursuit.  It is generally more visible in community which
-in turn leads to a broader and deeper participation. For instance,
-promoting collaborations across the breadth (e.g., different
-departments) and depth (e.g., stakeholders within a department) of
-community, one of the key ingredients of a sustainable
-process. Successful utility projects reap high rewards and have a
-longer usage span. Development process becomes user-driven and
-self-sustaining.
+On the other hand, utility software presents opportunities to be usable by a
+larger community making its undertaking and development an attractive pursuit.
+It is generally more visible in community which in turn leads to a broader and
+deeper participation. For instance, it helps promoting collaborations across
+the breadth (e.g., different departments) and depth (e.g., stakeholders within
+a department) of community, one of the key ingredients of a sustainable
+process. Successful utility projects reap high rewards and have a longer usage
+span. Development process becomes user-driven and self-sustaining.
 
@@
 developers of these software, we conducted a small survey of the cases
-discussed above with respect to the current status of its core
-developers. We looked at the workshop submissions and the original
-cited papers for the respective software. Where available, we looked
-at the web pages and source code of the earliest versions to find the
-original developers. We included projects older than 5 years and
-excluded relatively new projects such as \toolname{SciPy} and
-\toolname{numFOCUS} from this survey. We find that in a majority of
-scientific software, more than one original developer are still
-around, in some cases after more than 10 years (e.g.,
-\toolname{R/qtl}).  Whereas, in the case of utility software projects,
-it appears that at least one original developer is still around, often
-in a part-time or ``benefactor'' role. In other cases, it appears that
-the original developers have went on to assume a larger role with the
-same project.
+discussed above with respect to the current status of its core developers. We
+looked at the workshop submissions and the original cited papers for the
+respective software. Where available, we looked at the web pages and source
+code of the earliest versions to find the original developers. We included
+projects older than 5 years and excluded relatively new projects such as
+\toolname{SciPy} and \toolname{numFOCUS} from this survey. We find that in a
+majority of scientific software, more than one original developer are still
+around, in some cases after more than 10 years (e.g., \toolname{R/qtl}).
+Whereas, in the case of utility software projects, it appears that at least one
+original developer is still around, often in a part-time or ``benefactor''
+role. In other cases, it appears that the original developers have went on to
+assume a larger role (eg. manager) within the same project.
 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
comment a note
minor fix in abstract

@@
 discussion about the keynote presentations, three themed panels that
-discussed papers accepted at the workshop`, other discussion,
+discussed papers accepted at the workshop, other discussion,
 cross-cutting issues, use cases, and conclusions. It joins a nascent
@@
 \textit{ab initio} modeling and simulation such as \toolname{VASP} or
-\toolname{Quantum Espresso}. \note{cite SW}
+\toolname{Quantum Espresso}. %\note{cite SW}
 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing author bug

@@
 Gabrielle~D.~Allen$^{(12)}$,
-Bruce~Berriman$^{(11)}$
+Bruce~Berriman$^{(13)}$
 Colin~Venters$^{(14)}$

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
small fix

--- 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 software discussed during the workshop. The data movement and sharing
-services offered by Globus can be easily used over the web by
+services offered by \toolname{Globus} can be easily used over the web by
 collaborating researchers.
@@
 smaller codebase and fixed requirements result in stability, ease of
-installation, and configurability.  Many such projects mature and are
+installation, and configuration. Many such projects mature and are
 treated as libraries to be integrated into larger systems such as some of

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex

@@
 and how the reward systems of science function, thereby shaping the
-types of software projects undertaken, including the extent to which
+types of software projects undertaken. These incentives can determine the extent to which
 developers are motivated to build software for the long-term, for the
@@
 Unique perspectives were captured about issues such as comprehensive
-documentation, development and deployment practices; software
-licenses; career paths for developers; identifiers for software
-through a mechanism such as a Digital Object Identifier or publication
-of a ``software paper''; and evidence of software contribution and
-impact, recorded in online systems such as source code repositories
+documentation, development and deployment practices, software
+licenses and career paths for developers. Attribution systems that capture
+evidence of software contribution and impact were also discussed. These include
+mechanisms such as Digital Object Identifiers, publication
+of ``software papers'' and the use of online systems, for example source code repositories
 like GitHub. For truly sustainable software, there should be no
@@
 planned to be held at the~2014 SciPy conference, and a follow-on
-WSSSPE2 workshop has been proposed to, and meanwhile been accepted for
+WSSSPE2 workshop has been accepted for
 the SC14 conference. In addition, funds to support the workshops have
@@
 
+Wilkins-Diehr was supported by the Extreme Science and Engineering Discovery Environment (XSEDE), 
+NSF ACI-1053575.
+
 \appendix

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Added support info for Howison and tidied one or two sentences in the Policy section

@@
 and how the reward systems of science thereby shape the type of
-software work undertaken, including the extent to which developers are
-motivated to build software for the long-term, for the use of others
+software work undertaken. This study includes the extent to which developers are
+motivated to build software for the long-term, for the use of others,
 and whether to work collaboratively or
@@
 Papers and discussion concentrated on three overarching questions: How
-ought software work to be visible, what are the barriers to its
+ought software work be visible, what are the barriers to its
 visibility, and what can be done to make it more visible?
@@
 
+Howison was supported by NSF SBE-1064209 and NSF ACI-0943168.
+
 Wilkins-Diehr was supported by the Extreme Science and Engineering Discovery Environment (XSEDE), 

### 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex
This commit includes suggested updates for the abstract by Anne C. Elster
Further discussion on software licenses and some suggested updates of the 
conclusions will follow in subsequent commits.

@@
 Gabrielle~D.~Allen$^{(12)}$,
-Bruce~Berriman$^{(13)}$
-Colin~Venters$^{(14)}$
+Anne~C.~Elster$^{(13)}$,
+Bruce~Berriman$^{(14)}$
+Colin~Venters$^{(15)}$
 }
@@
 %
-\thanks{{}$^{(11)}$ Infrared Processing and Analysis Center, California Institute of Technology, Pasadena, CA, USA; \url{gbb@ipac.caltech.edu}}
+\thanks{{}$^{(13)}$ Norwegian University of Science and Technology, Trondheim, Norway; \url{elster@ntnu.no}}
 %
-\thanks{{}$^{(14)}$ University of Huddersfield, Huddersfield, West Yorkshire, UK; \url{C.Venters@hud.ac.uk}}
+\thanks{{}$^{(14)}$ Infrared Processing and Analysis Center, California Institute of Technology, Pasadena, CA, USA; \url{gbb@ipac.caltech.edu}}
+%
+\thanks{{}$^{(15)}$ University of Huddersfield, Huddersfield, West Yorkshire, UK; \url{C.Venters@hud.ac.uk}}
 
 \begin{abstract} %\note{Lead: Sou-Cheng T. Choi}
-This document summarizes the First Workshop on Sustainable Software
-for Science: Practice and Experiences (WSSSPE1). The content is based
-on a set of collaborative notes taken during the workshop, including
-discussion about the keynote presentations, three themed panels that
-discussed papers accepted at the workshop, other discussion,
-cross-cutting issues, use cases, and conclusions. It joins a nascent
-literature seeking to understand what drives software work in science
-and how the reward systems of science function, thereby shaping the
-types of software projects undertaken. These incentives can determine the extent to which
+% Anne: Added a lead-in motivator to address the submission as a Experience paper
+% and tried to improve writing... ( I am assumeing no abstract word count limit...)
+Challenges related to the development, deployment, and maintenance of 
+reusable software for science is becoming an increasing concern. 
+There are now many scientists whose research increasingly depends on 
+the quality and availability of software they can use and build on. 
+At the same time, software specialists have more than enough to do 
+without having to "reinvent the wheel". To highlight some of these 
+issues and share experiences, The First Workshop on Sustainable 
+Software for Science: Practice and Experiences (WSSSPE1) was held 
+in November 2013 at SC13.
+The workshop featured keynote presentations and based its discussion 
+them, as well as on a large number (54) of solicited extended abstracts 
+that were grouped into three themes presented via panels that encouraged 
+audience participation. A set of collaborative notes was taken during the workshop.
+
+This document 
+% article? 
+summarizes 
+the issues and shared experiences discussed atthe workshop, including
+cross-cutting issues and use cases. It joins a nascent
+literature seeking to understand what drives software work in science,
+and how it is impacted by the reward systems of science. 
+These incentives can determine the extent to which
 developers are motivated to build software for the long-term, for the
 use of others, and whether to work collaboratively or separately.
+
 Unique perspectives were captured about issues such as comprehensive
@@
 
+% Original:
+% This document summarizes the First Workshop on Sustainable Software
+% for Science: Practice and Experiences (WSSSPE1). The content is based
+% on a set of collaborative notes taken during the workshop, including
+% discussion about the keynote presentations, three themed panels that
+% discussed papers accepted at the workshop, other discussion,
+% cross-cutting issues, use cases, and conclusions. It joins a nascent
+% literature seeking to understand what drives software work in science
+% and how the reward systems of science function, thereby shaping the
+% types of software projects undertaken. These incentives can determine the extent to which
+% developers are motivated to build software for the long-term, for the
+% use of others, and whether to work collaboratively or separately.
+% Unique perspectives were captured about issues such as comprehensive
+% documentation, development and deployment practices, software
+% licenses and career paths for developers. Attribution systems that capture
+% evidence of software contribution and impact were also discussed. These include
+% mechanisms such as Digital Object Identifiers, publication
+% of ``software papers'' and the use of online systems, for example source code repositories
+% like GitHub. For truly sustainable software, there should be no
+% endpoint, as the software products continue to be used beyond a single
+% institution, grant, and developer or development team.
+
 \end{abstract}
@@
 % \item Lynn Zentner $\langle$\url{lzentner@purdue.edu}$\rangle$
-\item Anne Elster $\langle$\url{anne.elster@gmail.com}$\rangle$
+\item Anne C. Elster $\langle$\url{elster@ntnu.no}$\rangle$
 \item * Shel Swenson $\langle$\url{mdswenso@usc.edu}$\rangle$
@@
 
+%Anne did not attend the whole workshop as she was busy as a SC14 Deouty Panels chair, but definately was there :-)
+Anne C. Elster
+
 Ian Foster

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
proof read pass over abstract

@@
 % and tried to improve writing... ( I am assumeing no abstract word count limit...)
-Challenges related to the development, deployment, and maintenance of 
-reusable software for science is becoming an increasing concern. 
-There are now many scientists whose research increasingly depends on 
-the quality and availability of software they can use and build on. 
-At the same time, software specialists have more than enough to do 
-without having to "reinvent the wheel". To highlight some of these 
-issues and share experiences, The First Workshop on Sustainable 
-Software for Science: Practice and Experiences (WSSSPE1) was held 
-in November 2013 at SC13.
-The workshop featured keynote presentations and based its discussion 
-them, as well as on a large number (54) of solicited extended abstracts 
-that were grouped into three themes presented via panels that encouraged 
-audience participation. A set of collaborative notes was taken during the workshop.
-
-This document 
-% article? 
-summarizes 
-the issues and shared experiences discussed atthe workshop, including
-cross-cutting issues and use cases. It joins a nascent
-literature seeking to understand what drives software work in science,
-and how it is impacted by the reward systems of science. 
-These incentives can determine the extent to which
-developers are motivated to build software for the long-term, for the
-use of others, and whether to work collaboratively or separately.
+Challenges related to the development, deployment, and maintenance of reusable
+software for science is becoming an increasing concern.  There are now many
+scientists whose research increasingly depends on the quality and availability
+of software they can use and build on.  At the same time, software specialists
+have more than enough to do without having to "reinvent the wheel". To
+highlight some of these issues and share experiences, The First Workshop on
+Sustainable Software for Science: Practice and Experiences (WSSSPE1) was held
+in November 2013 at SC13.  The workshop featured keynote presentations and
+based its discussion on them, as well as on a large number (54) of solicited
+extended abstracts that were grouped into three themes presented via panels
+that encouraged audience participation. A set of collaborative notes was taken
+during the workshop.
+
+This document % article? 
+summarizes the issues and shared experiences discussed at the workshop,
+including cross-cutting issues and use cases. It joins a nascent literature
+seeking to understand what drives software work in science, and how it is
+impacted by the reward systems of science.  These incentives can determine the
+extent to which developers are motivated to build software for the long-term,
+for the use of others, and whether to work collaboratively or separately.
 
 Unique perspectives were captured about issues such as comprehensive
-documentation, development and deployment practices, software
-licenses and career paths for developers. Attribution systems that capture
-evidence of software contribution and impact were also discussed. These include
-mechanisms such as Digital Object Identifiers, publication
-of ``software papers'' and the use of online systems, for example source code repositories
-like GitHub. For truly sustainable software, there should be no
-endpoint, as the software products continue to be used beyond a single
-institution, grant, and developer or development team.
+documentation, development and deployment practices, software licenses and
+career paths for developers. Attribution systems that capture evidence of
+software contribution and impact were also discussed. These include mechanisms
+such as Digital Object Identifiers, publication of ``software papers'' and the
+use of online systems, for example source code repositories like GitHub. For
+truly sustainable software, there should be no endpoint, as the software
+products continue to be used beyond a single institution, grant, and developer
+or development team.
 

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Added citations that were missing

@@
 their building and continued engagement need attention during the
-project life cycle.  Several of the submitted papers \todo{cite}
+project life cycle.  Several of the submitted papers
 discussed how communities have been built around projects, and what is
-needed to enable a project to grow. The latter includes public
+needed to enable a project to
+grow~\cite{Crusoe_WSSSPE, Hanwell_WSSSPE, Christopherson_WSSSPE, Cranston_WSSSPE, Terrel_WSSSPE}.
+The latter includes public
 source-code hosting, mailing lists, documentation, wikis, bug
@@
 recommendations for processes, practices, tools, etc. One of the
-papers \note{cite?}  suggested that a ``Software Sustainability
+papers suggested that a ``Software Sustainability
 Institute'' should be vested with the authority to develop
@@
 orphanage center (i.e., a place to `take care' of software when the
-original developers have stopped doing so).  The idea of one common
+original developers have stopped doing so)~\cite{Crawford_WSSSPE}.
+The idea of one common
 repository received some resistance, with so many compelling

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Update WSSSPE1_summary_report.tex
Added subsubsection on Licensing and Patents (expanded on what was there 
under "Industry").

Added short statement from my paper on how multi- and many core
computing forces the need for awareness of locality and synch.

Added short statement on licensing and patents and how they impact funding for SW

--- 2014-04-24 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 their building and continued engagement need attention during the
-project life cycle.  Several of the submitted papers \todo{cite}
+project life cycle.  Several of the submitted papers 
+% need help here...
+\todo{cite}
 discussed how communities have been built around projects, and what is
@@
 recommendations for processes, practices, tools, etc. One of the
-papers \note{cite?}  suggested that a ``Software Sustainability
+papers 
+%N a todo!
+\note{cite?}  
+suggested that a ``Software Sustainability
 Institute'' should be vested with the authority to develop
@@
 communication or point of contact was seen as reasonable, with the
-statement that ``vested with authority'' is perhaps too strong, but
+statement that ``vested with authority'' is perhaps too strong. However,
 ``providing tools if needed'' might be more appropriate.
@@
 
+%Anne: The folowing sentence is verbatim from the original Abstract..
+% Se preivious commit for suggestion on hwo to shorten this sentence
 These papers, and the discussion at the workshop, join a nascent
@@
 
+\subsubsection{Licensing and Software Patents}
+Software is no longer either open or closed (only binaries available),
+but also licensed and patented, which clearly also impact 
+software sustainability.
 While many papers briefly discussed licensing issues,
-Elster~\cite{Elster_WSSSPE} directly discussed the impact of different
-licenses on obtaining industrial funding for scientific software
-projects.  In particular, she describes experiences with researchers
+Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
+on obtaining industrial funding for scientific software
+projects.  In particular, she describes her experiences with researchers
 unwilling to utilize GPL (copyleft) code, as it adds restrictions to
-reuse that some industries find unacceptable for future
+reuse that they themselve as we well as some industries find unacceptable for future
 commercialization.  (This was discussed by Hanwell et
-al.~\cite{Hanwell_WSSSPE} as well.)  Further obstacles to
-sustainability identified by the author include patenting of software,
-a lack of open access, and even more broadly, a lack of open source
+al.~\cite{Hanwell_WSSSPE} as well.)  US ICT companies funding 
+academics will thus often insist on BSD licensing on software 
+since they then can legally include the code into their commercial
+codes.
+
+On the other hand, there are companies
+that fund larger GPL-licensed software projects \cite{Blatt_WSSSPE}
+that insists that the university projects they fund 
+also produce code with GPL licensing.
+They do not accept BSD-like licenses since they argue that
+other companies then may choose to build closed commercial
+codes on they software they funded, rather than encouraging 
+the community to contribute freely and thus ensuring software 
+sustainability for the community.
+
+In either case, the university researchers are not given much choice
+if they want these much sought after funds in a increasingly competitive
+grant world.
+
+Another obstacles to
+sustainability identified by the author include patenting of software.
+Most countries place some limits on software patents. The European 
+Union outright forbids them, while US patent law excludes “abstract ideas” 
+that has been used to refuse some software patents.
+
+Further obstacles to
+sustainability is a lack of open access, and even more broadly, a lack of open source
 codes even in open access journals.  Finally, a lack of awareness on
@@
 domains.
+
+Elster \cite{Elster_WSSSPE} also points out how the 
+prevalence and rapid growth of multi and many-core systems forces 
+awareness of data locality and synchronization issues
+if one want to teach people how to develop high-performing scientific codes.
 
@@
 
-To conclude, we discuss what we have learned from the workshop and what we plan to do going forward.
+To conclude, we highlight what we have learned from the workshop, and what we plan to do going forward.
 
@@
   projects facilitates collaboration, reproducibility, and
-  sustainability in computational science. However it represents a
+  sustainability in computational science. However, it represents a
   barrier in some scientific fields and has yet to be more widely
@@
 \end{enumerate}
-
+In addition, licensing and patents and how they impact research funding
+for software developement, was also discussed.
 % Conclusion of S8 : Use Cases, not sure where to put
@@
 
+Elster would like to thank Statoil ASA for supporting her travel
+and her research group through a grant related to the OPM/DUNE open source projects.
+
 \appendix
@@
 \item Anne C. Elster. Software for Science: Some Personal
-  Reflections~\cite{Elster_WSSSPE}
+  Reflections on Funding, Licensing, Publishing and Teaching~\cite{Elster_WSSSPE}
 

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
fix a typo, add some non-linebreaking spaces

@@
 unwilling to utilize GPL (copyleft) code, as it adds restrictions to
-reuse that they themselve as we well as some industries find unacceptable for future
+reuse that they themselve as well as some industries find unacceptable for future
 commercialization.  (This was discussed by Hanwell et
@@
 On the other hand, there are companies
-that fund larger GPL-licensed software projects \cite{Blatt_WSSSPE}
+that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
 that insists that the university projects they fund 
@@
 
-Elster \cite{Elster_WSSSPE} also points out how the 
+Elster~\cite{Elster_WSSSPE} also points out how the 
 prevalence and rapid growth of multi and many-core systems forces 

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Abstract: refine.

@@
 Challenges related to the development, deployment, and maintenance of 
-reusable software for science is becoming an increasing concern. 
+reusable software for science are becoming a growing concern. 
 There are now many scientists whose research increasingly depends on 
@@
 At the same time, software specialists have more than enough to do 
-without having to "reinvent the wheel". To highlight some of these 
+without having to ``reinvent the wheel.'' To highlight some of these 
 issues and share experiences, The First Workshop on Sustainable 
@@
 The workshop featured keynote presentations and based its discussion 
-them, as well as on a large number (54) of solicited extended abstracts 
+on them, as well as on a large number (54) of solicited extended abstracts 
 that were grouped into three themes presented via panels that encouraged 
@@
 
-This document 
-% article? 
-summarizes 
-the issues and shared experiences discussed atthe workshop, including
+This document summarizes 
+the issues and shared experiences discussed at the workshop, including
 cross-cutting issues and use cases. It joins a nascent
@@
 developers are motivated to build software for the long-term, for the
-use of others, and whether to work collaboratively or separately.
+use of others, and whether to work collaboratively or separately. We 
+also elabroate on community building, leadership, and dynamics leading to 
+successful scientific software.
 
@@
 documentation, development and deployment practices, software
-licenses and career paths for developers. Attribution systems that capture
+licenses and career paths for developers. Attribution systems that accounts for 
 evidence of software contribution and impact were also discussed. These include
 mechanisms such as Digital Object Identifiers, publication
-of ``software papers'' and the use of online systems, for example source code repositories
+of ``software papers'', and the use of online systems, for example source code repositories
 like GitHub. For truly sustainable software, there should be no

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Abstract: refine.

@@
 % and tried to improve writing... ( I am assumeing no abstract word count limit...)
-Challenges related to the development, deployment, and maintenance of 
+Challenges related to development, deployment, and maintenance of 
 reusable software for science are becoming a growing concern. 
-There are now many scientists whose research increasingly depends on 
-the quality and availability of software they can use and build on. 
+Research of many scientists now increasingly depends  on 
+the quality and availability of software upon which their works are built. 
 At the same time, software specialists have more than enough to do 
 without having to ``reinvent the wheel.'' To highlight some of these 
-issues and share experiences, The First Workshop on Sustainable 
+issues and share experiences, the First Workshop on Sustainable 
 Software for Science: Practice and Experiences (WSSSPE1) was held 
-in November 2013 at SC13.
+in November 2013 in conjunction with the Supercomputing Conference SC13.
 The workshop featured keynote presentations and based its discussion 
@@
 use of others, and whether to work collaboratively or separately. We 
-also elabroate on community building, leadership, and dynamics leading to 
+also explore on community building, leadership, and dynamics in relation to
 successful scientific software.

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Minor change.

@@
 that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
-that insists that the university projects they fund 
+and insist that the university projects they fund 
 also produce code with GPL licensing.

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
moving licensing discussion to cross-cutting

@@
 workshop papers grouped by the three thematic workshop panels to which
-they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Two
+they were assigned~(\S\ref{sec:devel}-\ref{sec:community}). Three
 broader cross-cutting issues surfaced repeatedly, and are discussed
@@
 
-\subsubsection{Licensing and Software Patents}
-Software is no longer either open or closed (only binaries available),
-but also licensed and patented, which clearly also impact 
-software sustainability.
-While many papers briefly discussed licensing issues,
-Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
-on obtaining industrial funding for scientific software
-projects.  In particular, she describes her experiences with researchers
-unwilling to utilize GPL (copyleft) code, as it adds restrictions to
-reuse that they themselve as well as some industries find unacceptable for future
-commercialization.  (This was discussed by Hanwell et
-al.~\cite{Hanwell_WSSSPE} as well.)  US ICT companies funding 
-academics will thus often insist on BSD licensing on software 
-since they then can legally include the code into their commercial
-codes.
-
-On the other hand, there are companies
-that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
-and insist that the university projects they fund 
-also produce code with GPL licensing.
-They do not accept BSD-like licenses since they argue that
-other companies then may choose to build closed commercial
-codes on they software they funded, rather than encouraging 
-the community to contribute freely and thus ensuring software 
-sustainability for the community.
-
-In either case, the university researchers are not given much choice
-if they want these much sought after funds in a increasingly competitive
-grant world.
-
-Another obstacles to
-sustainability identified by the author include patenting of software.
-Most countries place some limits on software patents. The European 
-Union outright forbids them, while US patent law excludes “abstract ideas” 
-that has been used to refuse some software patents.
-
-Further obstacles to
-sustainability is a lack of open access, and even more broadly, a lack of open source
-codes even in open access journals.  Finally, a lack of awareness on
-the part of scientific software developers of commodity libraries for
-common tasks reduces their ability to reuse code.
-
 \subsection{Education \& Training}
@@
 
-Two issues, how to define software sustainability, and how career
-paths (or their lack) interact with achieving it, were raised across
+Three issues, how to define software sustainability, how career
+paths (or their lack) interact with achieving it, and the impact of
+software licenses, were raised across
 the workshop's panels.  This section aims to synthesize these
@@
 
-\note{add a subsection about licensing - lessons, advice, etc.?}
+\subsection{Licensing and Software Patents}
+Issues related to licensing and patents primarily was discussed in the Communities
+session, but licensing was also a concern of many of the other contributors
+in other sessions.
+%
+Software is no longer just open or closed (only binaries available),
+but also licensed and patented, which clearly also impacts 
+software sustainability.
+While many papers briefly discussed licensing issues,
+Elster~\cite{Elster_WSSSPE} directly discussed the impact of software licenses 
+on obtaining industrial funding for scientific software
+projects.  In particular, she described her experiences with researchers
+unwilling to utilize GPL (copyleft) code, as it adds restrictions to
+reuse that they themselves as well as some industries find unacceptable for future
+commercialization.  (This was discussed by Hanwell et
+al.~\cite{Hanwell_WSSSPE} as well.)
+
+US information technology companies funding 
+academics will thus often insist on BSD licensing on software 
+since they then can legally include the code into their commercial
+codes.
+%
+On the other hand, there are companies
+that fund larger GPL-licensed software projects~\cite{Blatt_WSSSPE}
+and insist that the university projects they fund 
+also produce code with GPL licensing.
+They do not accept BSD-like licenses since they argue that
+other companies then may choose to build closed commercial
+codes on they software they funded, rather than encouraging 
+the community to contribute freely and thus ensuring software 
+sustainability for the community.
+%
+In either case, the university researchers are not given much choice
+if they want these much sought after funds in a increasingly competitive
+grant world.
+
+Another obstacle to
+sustainability identified by Elster include patenting of software.
+Most countries place some limits on software patents. The European 
+Union outright forbids them, while US patent law excludes ``abstract ideas,'' 
+which has been used to refuse some software patents.
+%
+Further obstacles to
+sustainability include a lack of open access, and even more broadly, a lack of open source
+codes even in open access journals.  Finally, a lack of awareness on
+the part of scientific software developers of commodity libraries for
+common tasks reduces their ability to reuse code.
 
@@
 
-Three issues came up repeatedly in different parts of the workshop:
+Three major issues came up repeatedly in different parts of the workshop:
 
@@
 \end{enumerate}
+%
+In addition, licensing and patents, and how they impact research funding
+for software development, were also discussed.
 
@@
 \end{enumerate}
-In addition, licensing and patents and how they impact research funding
-for software developement, was also discussed.
+
 % Conclusion of S8 : Use Cases, not sure where to put

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
remaking Hanwell changes that were somewhow undone

@@
 project life cycle.  Several of the submitted papers 
-% need help here...
-\todo{cite}
 discussed how communities have been built around projects, and what is
-needed to enable a project to grow. The latter includes public
+needed to enable a project to grow~\cite{Crusoe_WSSSPE, Hanwell_WSSSPE, Christopherson_WSSSPE, Cranston_WSSSPE, Terrel_WSSSPE}. The latter includes public
 source-code hosting, mailing lists, documentation, wikis, bug
@@
 recommendations for processes, practices, tools, etc. One of the
-papers 
-%N a todo!
-\note{cite?}  
+papers~\cite{Crawford_WSSSPE}
 suggested that a ``Software Sustainability

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
edits to abstract and intro

@@
 reusable software for science are becoming a growing concern. 
-Research of many scientists now increasingly depends  on 
+Many scientists' research increasingly depends  on 
 the quality and availability of software upon which their works are built. 
-At the same time, software specialists have more than enough to do 
-without having to ``reinvent the wheel.'' To highlight some of these 
+To highlight some of these 
 issues and share experiences, the First Workshop on Sustainable 
 Software for Science: Practice and Experiences (WSSSPE1) was held 
-in November 2013 in conjunction with the Supercomputing Conference SC13.
-The workshop featured keynote presentations and based its discussion 
-on them, as well as on a large number (54) of solicited extended abstracts 
-that were grouped into three themes presented via panels that encouraged 
-audience participation. A set of collaborative notes was taken during the workshop.
-
-This document summarizes 
-the issues and shared experiences discussed at the workshop, including
-cross-cutting issues and use cases. It joins a nascent
-literature seeking to understand what drives software work in science,
-and how it is impacted by the reward systems of science. 
-These incentives can determine the extent to which
-developers are motivated to build software for the long-term, for the
-use of others, and whether to work collaboratively or separately. We 
-also explore on community building, leadership, and dynamics in relation to
-successful scientific software.
+in November 2013 in conjunction with the SC13 Conference.
+The workshop featured keynote presentations and a large number (54)
+of solicited extended abstracts 
+that were grouped into three themes and presented via panels.
+A set of collaborative notes of the presentations and discussion
+was taken during the workshop.
 
@@
 institution, grant, and developer or development team.
+
+This paper summarizes 
+the issues and shared experiences that were discussed, including
+cross-cutting issues and use cases. It joins a nascent
+literature seeking to understand what drives software work in science,
+and how it is impacted by the reward systems of science. 
+These incentives can determine the extent to which
+developers are motivated to build software for the long-term, for the
+use of others, and whether to work collaboratively or separately. It 
+also explores community building, leadership, and dynamics in relation to
+successful scientific software.
 
@@
 has been turned into a community website to engender further
-discussion and progress. Additionally, a minisymposium at the~2014
+discussion and progress. Additionally, a minisymposium at the 2014
 Society for Industrial and Applied Mathematics (SIAM) Annual Meeting
@@
 activities~(\S\ref{sec:conclusions}).  The short papers accepted to
-the workshop are listed in Appendix~\ref{sec:papers}, and a partial
+the workshop are listed in Appendix~\ref{sec:papers}, part of he original
+call for papers is included in 
+Appendix~\ref{sec:cfp}, and a partial
 list of workshop attendees can be found in
-Appendix~\ref{sec:attendees}. The original call for papers is included in 
-Appendix~\ref{sec:cfp}.
+Appendix~\ref{sec:attendees}.
 
@@
 
-In roughly following the aforementioned scientific software ecosystem
-challenges, the accepted submissions were grouped into three main
+Roughly following the call for papers topics, the accepted submissions
+were grouped into three main
 categories, namely \emph{Developing and Supporting Software},
@@
 
-For each panel, the paper summaries were followed by active discussion
-involving panelists, onsite attendees, and often enough online
+The workshop also included two keynote presentations.
+Remote participation was facilitated by a
+live-cast of keynotes and panels via Ustream.tv
+(\url{http://ustream.tv}).
+In each panel, the paper summary presentations were followed by active discussion
+involving panelists, onsite attendees, and often online
 participants. The latter was facilitated by having a shared Google
@@
 \texttt{\#wssspe}).
-
-The workshop itself consisted of two keynote presentations followed by
-the three panel sessions.  Remote participation was facilitated by a
-live-cast of all keynotes and panels on Ustream.tv
-(\url{http://ustream.tv}).
 

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
(1) Shuffle Appendix A and Appendix B and change related mentions. (2) Move Table 1 from section 7 to section 5 where it is first mentioned. (3) Minor changes to Section 5.

@@
 with issues and lessons learned, as well as plans for future
-activities~(\S\ref{sec:conclusions}).  The short papers accepted to
-the workshop are listed in Appendix~\ref{sec:papers}, part of he original
-call for papers is included in 
-Appendix~\ref{sec:cfp}, and a partial
+activities~(\S\ref{sec:conclusions}).  The original call for papers is included in 
+Appendix~\ref{sec:cfp}.  The short papers accepted to
+the workshop are listed in Appendix~\ref{sec:papers}, and a partial
 list of workshop attendees can be found in
@@
 The workshop submissions grouped under this section provide frameworks
-for thinking about, researching and understanding which elements of
-sustainability are important and how they relate. Although there is
+for thinking about, researching, and understanding which elements of
+sustainability are important and how they are related to each other. Although there is
 substantial overlap between the frameworks, they have different
@@
 ~\S\ref{sec:defining-sustainability}.
-
-One area in which there was not complete overlap was whether the word
-(and thus the effort called for by the WSSSPE workshop) involved
-environmental sustainability. Of course the word sustainability has
-strong connotations from consideration of environmental issues,
-evoking some mention of the areas in which software interacts with
-overall environmental resource usage, particularly energy efficiency.
-The two papers in this area which mentioned
-this~\cite{Venters_WSSSPE,Calero_WSSSPE} did so without integrating
-that analysis into the question of software being around long-term,
-suggesting that questions of environmental impact of scientific
-software is a conceptually distinct area of inquiry.
-
-One group of papers presented frameworks that were primarily about
-characteristics of software artifacts, connecting with the long
-discourse on software quality. This approach is realized in adjectives
-that can be applied to pieces of software but might also extend to
-describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE}
-propose adding elements to the ISO standards for measuring software
-quality. Specifically, they propose an additional dimension of quality
-they call ``perdurability'' or ability to ``last over time'' made up
-of three characteristics: reliability, maintainability and
-adaptability. This shows some overlap with the framework employed by
-Venters et al.~\cite{Venters_WSSSPE} who employ the adjectives of
-``extensibility, interoperability, maintainability, portability,
-reusability and scalability,'' anticipating the sorts of work that
-people would need to do with a software artifact in the future, ``as
-stakeholders requirements, technology and environments evolve and
-change.'' Venters et al.~argue that these choices are really made very
-early because they relate to the architecture of the software and
-involve trade offs that ought to be analyzed alongside other
-attributes of the software artifact. Lenhardt et
-al.~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
-lifecycle to argue for the inclusion of metadata throughout a piece of
-software's life (discussing, for example, how it has been built and
-tested and what ``data in'' and ``data out'' has been considered). In
-addition, their analogy suggests that the software lifecycle might add
-a phase of ``preservation'' and draw on understandings of what that
-involves from studies of data.  In sum, then, these frameworks focused
-on what needs to be accomplished to have more sustainable software.
-
-A second theme in these papers was the continued availability of
-resources to accomplish the goals of sustainability. The elements of
-these frameworks focused far more on the organization of a software
-project than they did on characteristics of the artifact itself
-(although it is certainly true that the adjectives discussed above
-could be applied to a software project). For example Pierce et
-al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
-particularly in terms of how those involved communicate and jointly
-set priorities, a process they call governance. In particular they
-argue that because sustainability relates to having ongoing resources,
-governance must be open to receive diverse input (by occurring online,
-asynchronously, and publicly) and thus have the potential to
-``transform passive users into active contributors''. They argue that
-the Apache Software Foundation's incubation process teaches this and
-could be learned from by projects throughout scientific software.
-Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
-describing two modes: ``top-down'' and ``bottom-up'' governance.  They
-place governance alongside questions of the artifact (``technical''),
-and questions of the origins of resources, in terms of who is funding
-the work surrounding a piece of software (``the political'') and the
-manner in which resources come to the project initially (commercial,
-open source, closed partnerships, grant funded) and long-term (all
-four plus paid support).
-
-Frameworks also concerned themselves with the context in which
-software projects exist, moving in abstraction from the software
-artifact itself to the organization of its production and the shape of
-the space in which an artifact or project exists.  These frameworks
-take the form of contingency theories in that they outline a different
-set of challenges and argue that different project organizations (and
-presumably artifact attributes) are necessary to persist long term in
-spaces with particular characteristics.  Katz and
-Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in
-terms of three axes: temporal (long or short term needs), spatial
-(local or global use) and purpose (specific to general).  They propose
-that different project organizations will be needed in different
-locations and argue that we should concentrate research to understand
-those connections. Weber et al.~\cite{Weber_WSSSPE} describe their
-spaces by analogy with natural ecosystems as ``niches'' which sustain
-particular pieces of software. They define ``a software niche as the
-set of technical requirements, organizational conditions and cultural
-mores that support its maintenance and use over time.'' They call for
-better understanding and modeling of niches (as well as further
-exploration of the usefulness of ecosystem metaphors).
-
-\subsection{Credit, Citation, and Impact}
-
-How work on scientific software is recognized and rewarded strongly
-influences the motivation for particular kinds of work on scientific
-software. A recurring theme of the panel discussion was that software
-work in science is inadequately visible within the reputation system
-underlying science; in other words it often doesn't ``count''. In his
-paper for this workshop, Katz placed software work along with other
-``activities that facilitate science but are not currently rewarded or
-recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
-to ``support all researchers in presenting meaningful impact evidence
-in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
-Knepley et al.~argued that the lack of visibility of software that
-supported a piece of science ``can have detrimental effects on
-funding, future library development, and even scientific
-careers.''~\cite{Knepley_WSSSPE}.
-
-%Anne: The folowing sentence is verbatim from the original Abstract..
-% Se preivious commit for suggestion on hwo to shorten this sentence
-These papers, and the discussion at the workshop, join a nascent
-literature seeking to understand what drives software work in science
-and how the reward systems of science thereby shape the type of
-software work undertaken. This study includes the extent to which developers are
-motivated to build software for the long-term, for the use of others,
-and whether to work collaboratively or
-separately~\cite{howison_incentives_2013, howison_scientific_2011,
-  bietz_synergizing_2010}. Software work is not only motivated by
-direct citations, but the visibility of software work in the
-literature is important to those who write software used in science.
-
-Papers and discussion concentrated on three overarching questions: How
-ought software work be visible, what are the barriers to its
-visibility, and what can be done to make it more visible?
-
-Most of the papers in this area focused on visibility of software in
-scientific papers, since scientific papers are the most widely
-accepted documentation of achievement in science. It was noted that
-there appear to be no widely accepted standards on how the use of
-software towards a paper ought to be mentioned, and that journals,
-citation style guides and other guides to scientific conduct are vague
-about how to describe software. To address this, papers advocated the
-need for a fixed identifier for software, either directly through a
-mechanism such as a Digital Object
-Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
-written to document the software (and perhaps its creation), a
-``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
-out during the panel discussion, one of the problems with papers as
-the cited product is that their author list is fixed in time, which
-discourages potential contributors who are not on the original author
-list from designing incremental improvements as integration work
-rather than separate (and hence possibly rewritten) software
-products~\cite{howison_incentives_2013}.
-
-Another approach is to reduce the difficulty of citing all software
-underlying a research paper. For example, scientists often work with
-software that itself wraps other software, leading to attribution
-stacking that can make it non-obvious or even difficult to determine
-what attributions would be appropriate. Knepley et
-al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
-which the software itself, after it has run, provides the user with a
-set of citations, according to the pieces of code actually
-executed. They describe a prototype implementation whereby the
-citations are embedded in libraries and reported along with the
-results, via a command-line
-interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
-that attempting to acknowledge the contributions of all pieces of
-dependent code within a paper faces the difficulty of creating very
-long citation lists, straining the analogy of code used to papers
-cited. Katz approaches this issue by proposing a system of
-``transitive credit,'' recording dependencies and relative
-contributions outside particular papers, relieving authors from the
-responsibility of acknowledging each and every dependency. Instead
-authors would acknowledge the percentage contribution of the software
-they used directly and an external system would then be able to
-recursively allocate that credit to those who had provided
-dependencies. Finally Priem and Piwowar argued that machine learning
-techniques could examine the body of published literature and extract
-mentions of software, coping with the multitude of informal ways in
-which authors mention software they used~\cite{Priem_WSSSPE}.  A point
-raised in the panel discussion was that instead of asking users to
-improve their software citation practices, one can also ask how
-software projects can better monitor the literature to improve their
-ability to show impact. For example, the NanoHub project scans the
-literature using keywords and names of known users to discover papers
-that are likely to have used their software and platform, and assigns
-graduate students to read each paper, highlighting mentions of
-software use and sometimes following up with the authors to identify
-stories for demonstrating impact.  A process for tracking
-software-using publications with the goal of increasing impact
-visibility is now described at \url{http://publications.wikia.com}.
-
-Potential visibility, and thus acknowledgement of scientific software
-products is not restricted to publications. Another key location for
-visibility is in the grant funding process, and as emphasized by NSF
-representatives at the meeting, recent changes to grant proposal and
-reporting formats now allow both applicants and awardees to report and
-highlight software products as much as publications. Nonetheless,
-whether peer review panels would value these contributions in the same
-way as publications remains to be seen.
-
-Priem and Piwowar argued that assessing the impact of software work
-requires looking beyond publications, including evidence of
-contribution and impact recorded in social coding-oriented resources
-such as GitHub, and conversations about software in issue trackers,
-mailing lists, twitter and beyond~\cite{Priem_WSSSPE}. In keeping with
-a principle of the ``altmetrics'' approach, they advocate that
-scholars should have resources that empower them to tell their own
-stories in the manner most appropriate for them and their audiences.
-
-\subsection{Implementing Policy}
-
-The workshop contributions in this group were concerned with the
-aspect of how implementation of best practices and other
-recommendations for improving scientific software sustainability could
-be promoted. Specifically, if scientific software is to become more
-sustainable, corresponding policies and guidelines need to be such
-that the scientific community can follow and implement them. This is
-considerably more challenging that it might seem at first, because in
-the reality of science today resources, both financial and personnel,
-that could be devoted to implementation are very limited, and the
-reward system does not encourage scientists to do so. Furthermore,
-implementing sustainability-targeting policies and guidelines often
-takes a variety of specialized software engineering expertises, which
-are not necessarily found in a single engineer, and much less so in a
-domain scientist cross-trained in programming. Adding to the policy
-implementation challenges, applicable sustainability-promoting
-practices and guidelines will change through a software project's
-lifecycle, in particular as it gains maturity.
-
-Two of the papers in this group focus on specific facets of software
-design that are important factors in a project's sustainability but
-are often addressed only late in the scientific software development
-cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
-governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
-models for software security. The other two papers discuss
-implementation strategies for science from the perspective of
-facilitating many or all facets of sustainability-oriented software
-design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast large
-projects that have software infrastructure development built-in, with
-cross-training domain scientist PIs in software engineering best
-practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
-specialized software engineering skills could be turned into shared
-instrumentation with low barriers to access.
-
-Krintz et al.~\cite{Krintz_WSSSPE} describe how in an era in which
-computing frequently takes place in a distributed cloud, the control
-over digital resources is increasingly shifting from physical
-infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
-et al.~observe, unlike for physical IT infrastructure in data centers,
-science communities have developed very little in the way of practices
-and technology for API governance, referred to by Krintz et al.~as the
-``combined policy, implementation, and deployment control''. Web APIs
-can and do change, sometimes quite frequently, raising the need to
-port dependent applications. The effort required for porting is
-notoriously difficult to estimate, making it nearly impossible for IT
-organizations to assess and thus properly manage the impact of API
-changes. To address this, Krintz et al.~propose a mechanism that
-evaluates the porting effort between two versions of a web-service API
-in a formal and automated way. To analyze the porting effort, they
-divide API changes into \emph{syntactic similarity}, the changes in
-inputs and outputs, and into \emph{semantic similarity}, the changes
-in the API's behavior. In initial tests, their method showed good
-congruence with human developers in scoring porting effort, offering
-the possibility that API governance can become as solid a part of
-scientific IT management as data center infrastructure management is
-today.
-
-Many facets of engineering for software sustainability strongly depend
-on the maturity level of the software. However, the maturity level of
-a software project meant to be sustained changes tremendously over its
-development life cycle, and the eventual maturity level is often
-difficult to predict during initial development. Using software
-security as their case study, Heiland et al.~\cite{Heiland_WSSSPE}
-discuss how Maturity Models can be used to formalize best practices
-appropriate for the different levels of maturity that a software
-project may go through over its lifecycle. Cybersecurity is also an
-example of a sustainability-relevant aspect in software design that is
-rarely given due diligence in science. In particular in industry,
-cybersecurity best practices for different stages of life cycle and
-maturity have been formalized as Software Security Maturity Models,
-and are widely used, yet awareness of these among scientific software
-development communities remains low. In providing a path to tightening
-security practices as software matures, such models align with the
-objective of providing implementation approaches that the scientific
-community can actually follow.
-
-API governance and cybersecurity measures appropriate for a project's
-maturity are all but two facets of sustainability-oriented software
-development. Others include user-centered interface design, test
-engineering, dependency management, and deployment operations. Each of
-these facets requires specialized skills and training in software
-engineering. How can the implementation of best practices and
-guidelines along many or all of these different facets be facilitated
-in scientific software development? Blanton and
-Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
-and personnel devoted to software engineering is ``co-funded'' with
-the driving science project. This typically implies large multi-year
-collaborative projects that to succeed require significant software
-infrastructure to be built, and which thus have the funding to support
-one or several software engineers. The sustainability of such projects
-then depends on sustaining the funding. In the other model, domain
-scientists also take the role of software developers, whether by
-necessity such as funding limitations, or due to cross-disciplinary
-professional interests. For this to result in sustainable software
-products, the domain scientists need to be (or become) cross-trained
-in software engineering standards and best practices.
-
-In practice, there are example for both extremes of the spectrum. For
-example, in the life sciences the iPlant Collaborative~\cite{iPlant},
-the Galaxy Project~\cite{Galaxy}, and \toolname{Qiime}~\cite{Qiime}
-are large multi-year projects with significant software infrastructure
-funding and deliverables. Typically though these are the exception
-rather than the rule, and particularly in the long tail of science the
-scientist-developer predominates, even for software that is widely
-used or crosses domain boundaries such as
-\toolname{rOpenSci}~\cite{rOpenSci}. For better training domain
-scientists at least in basic software engineering best practices,
-initiatives such as Software Carpentry~\cite{SoftwareCarpentry} have
-demonstrated how this could be achieved at scale.
-
-However, there may also be a middle ground between the two
-extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
-Excellence model that leverages economies of scale to make software
-engineering experts and their skills accessible to the long tail of
-science. As Huang and Lapp discuss, this model could effectively turn
-the utilization of software engineering expertise from a complex human
-resource recruitment and management challenge, to buying time, when
-and to the extent needed, on shared instrumentation. There is
-precedence to using such a model to lower access barriers for long
-tail science, in particular for new experimental technologies.  For
-example, although the acquisition and operation of next-generation
-high-throughput DNA sequencers requires substantial investments of
-money, time, and expertise, the establishment of ``core facilities''
-on many university campuses has made these technologies accessible to
-a wide swath of scientists, with transformative results for science.
-
-One of the important conclusions from this group of papers is that
-creating sustainable software requires paying attention not to one or
-two, but to several different facets of software engineering, each
-with corresponding best practices and standards of excellence. Even if
-a science project requires and has funding for a full software
-engineering FTE, what the project really needs could be fractions of
-different engineers with different specialty training. The vast
-majority of long tail science projects lacks the funding for even one
-full software engineer, let alone one who combines expertise in all of
-the applicable facets of engineering. Some scientists in the long tail
-will have the professional interests to cross-train enough in software
-engineering to be successful with creating sustainable software, but
-it is unrealistic to expect this expertise and interest of all or even
-the majority. This is where a software engineering center of
-excellence could provide a critical resource by enabling scientists in
-the long tail to complement their resources and expertise with facets
-that are missing, but which, if applied at the right time, would
-improve the chances of a software product to become sustainable. Such
-complementary expertise also need not be restricted to software
-engineering in the strict sense; for example, it could consist of
-community building, leadership, and support for some period of
-transition to sustainability.
-
-In summary, implementing software sustainability practices on a
-broader basis requires on the one hand the development of guidelines
-and practices that are suitable for the exploratory research context
-in which most scientific software is created, and on the other hand a
-skilled workforce trained in a variety of software engineering facets
-and community building. The capabilities afforded by such a workforce
-need to be accessible not only to large projects with sufficient
-funding to provide competitive employment, but also to the many
-smaller projects in the long tail of science. Sufficiently
-cross-training domain scientists could be one way to achieve this;
-another, complementary approach is to instrument the necessary
-capabilities so they can be shared.
-
-\section{Communities} \label{sec:community}%\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
-
-%There were a number of papers categorized under the Communities banner and so
-%two presenters each summarized half of the submissions in this area.
-Across all talks and papers submitted, authors implicitly and
-explicitly recognized the concept of ``communities'' as a driving and
-unifying force within software projects.  Despite that, the actual
-nature of the communities, the incentive structures that bind them
-together, the infrastructure they utilize for communication and even
-the types of individuals that make up those communities were
-different.  Some of these communities were primarily composed of
-members of industry, some were funded and driven by individuals
-focused on developing software as opposed to utilizing it, and others
-were primarily composed of scientist practitioners, and were true
-communities of practice.
-
-These varying structures and compositions result in differing modes of
-interaction within communities, styles of development, and the
-structure of planning for future development of software.  In this
-section, we summarize the different types of communities in scientific
-software as well as the resultant impact on sustainability and
-development of functionality.
-
-\subsection{Communities}
-
-%This collection of papers was summarized by Karen Cranston.
-
-Drawing on experiences from high-energy physics, Vay et
-al.~\cite{Vay_WSSSPE} proposed developing teams of technical
-specialists   to overcome a lack of coordination between projects.
-Maximizing scientific output requires maximizing the usability of the
-scientific code while minimizing the cost of developing and
-supporting those code.  This included targeting different
-architectures for their software to be deployed, as well as
-coordination between technically-focused individuals and usage of a
-common scripting language between projects.  Instead of fragmenting
-the development of simulation codes across institutions, the paper
-suggests that a cohesive strategy reducing duplication and increasing
-coordination will broadly increase the efficiency across institutions.
-The approach proposed is of de-fragmenting the existing ecosystem in a
-non-disruptive way.
-
-Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
-catalysts'' and their role in the modern scientific research
-process. A technology catalyst is an individual with knowledge of
-technological advancements, tasked with user engagement to create
-scientific or engineering applications, using suitable tools and
-techniques to take advantage of current technological capabilities.
-%
-One of the tasks of technology catalysts is to seek community
-collaborations for new applications and engage users, thus benefiting
-both science, by effective running of scientific codes on
-computational infrastructure, and technology, by conducting research
-and seeking findings for technology improvement.  The particular
-engagements described in the paper came up from the lead author's work
-as a postdoctoral researcher at Cornell and Argonne, where interaction
-with the scientific communities in both institutions resulted in these
-collaborations.
-
-At NESCent, a combination of in-house informatics individuals and
-domain scientists collaborate to develop software to study
-evolutionary science.  The report~\cite{Cranston_WSSSPE} studied the
-success of a ``hackathon'' model for development, where short-form,
-hands-on events combining users, researcher-developers and software
-engineers targeted specific code improvements. From this experiment,
-the authors identified several key outcomes as well as
-lessons-learned: specifically, the co-localization of developers was
-seen as having a strong impact, enabling casual conversation that led
-to discrete outcomes.  The formation of the discussion mailing list,
-specifically in response to the social capital built at the hackathon,
-was seen as spurring on longer-term benefits to the community and
-fostering sustainability.
-
-Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
-project in developing collaboration supporting tools for
-Open Science.  This software collective, organized around the
-statistical programming environment R, develops access mechanisms for
-data repositories and attempts to reduce the barrier to entry for
-individuals wanting to access data repositories and study the data
-contained therein.  The collective fosters direct collaboration
-between individuals and data providers, designed to ``train academics
-in reproducible science workflows focused around R.''  Two central
-challenges are engagement of existing users within ecology and
-evolutionary biology (EEB), and how the community can make inroads and
-traction in other disciplines. Currently, the collective is exploring
-addressing these challenges through the use of social media, holding
-workshops and hackathons.  This helps to both raise the profile of
-the collective within EEB and in other domains.  However, the
-overarching challenge identified in the paper was that of
-incentivizing maintenance of software, which is difficult in academia.
-
-%\subsection{Communities Part 2}
-
-%Nancy Wilkins-Diehr summarized 6 papers in this part.
-
-Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree
-to which research relies on high quality software. There are often
-barriers and a lack of suitable incentives for researchers to embrace
-software engineering principles. The Water Science Software Institute
-is working to lower some of these barriers through an Open Community
-Engagement Process. This is a four-step iterative development process
-that incorporates Agile development principles.
-
-\begin{itemize}
-\setlength{\itemindent}{0.5in}
-\item[Step 1:] Design - thorough discussion of research questions
-\item[Step 2:] Develop working code
-\item[Step 3:] Refine based on new requirements
-\item[Step 4:] Publish open source
-\end{itemize}
-
-Christopherson reports on the application of Steps~1-3 to a
-computational modeling framework developed in the 1990s. Step~1 was a
-2-day, in-person specifications meeting and code walk-through. Step 2
-was a 5-day hackathon to develop working code, and Step 3 was a 3-day
-hackathon to refine the code based on new requirements. The team
-worked on small, low-risk units of code. It was challenging, revealed
-unanticipated obstacles, programmers had to work together, and
-experimentation was encouraged.
-%
-The paper recommended: start small and gradually building toward more
-complex objectives, consistent with Agile development; develop
-consensus before coding, by repeating step~1 before all hackathons;
-ensure newcomers receive orientation prior to the hackathon, such as a
-code walk-through or system documentation; and co-locate collaborators
-whenever feasible.
-
-%Refactor before adding
-%new functionality. Approach development as a learning experience. Welcome
-%experimentation, and treat mistakes as a natural part of the learning process.
-%Repeat Step 1 activities before all hackathons to develop consensus before
-%coding. This allows coding to be the focus of subsequent hackathons. In higher
-%risk situations, provide additional time for Step 1 activities. The
-%Christopherson team recommends a minimum of two months.
-%Ensure any newcomers
-%receive some form of orientation prior to the hackathon, such as a code
-%walkthrough or system documentation. Co-locate rather than collaborating
-%remotely whenever feasible.
-
-Pierce et al.~\cite{Pierce2_WSSSPE} describes how science gateways can
-provide a user-friendly entry to complex cyberinfrastructure. For
-example, more than 7,000 biologists have run phylogenetic codes on
-supercomputers using the CIPRES Science Gateway in $3 \tfrac{1}{2}$
-years. Over 120 scientists from 50 institutions used the
-\toolname{UltraScan Science Gateway} in one year, increasing the
-sophistication of analytical ultracentrifugation experiments
-worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
-more than 100 users who used 250,000 CPU hours in only a few months.
-
-Gateways, however, need to keep operational costs low and can often
-make use of common components, such as authentication, application
-installation and reliable execution and help desk support. Science
-Gateway Platform as a Service (\toolname{SciGaP}) delivers middleware
-as a hosted, scalable third-party service while domain developers
-focus on user interfaces and domain-specific data in the gateway.
-While \toolname{SciGaP} is based on the \toolname{Apache Airavata}
-project and the \toolname{CIPRES} Workbench Framework, community
-contributions are encouraged by its open source, open governance and
-open operations policies. The goal is robust, sustainable
-infrastructure with a cycle of development that improves reliability
-and prioritizes stakeholder requirements. The project is leveraging
-Internet2's \toolname{Net+} mechanisms for converting
-\toolname{SciGaP} and its gateways into commodity services.
-
-Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and
-challenges with the large \toolname{nanoHUB.org} community, where
-community is defined as a ``body of persons of common and especially
-professional interests scattered through a larger society.'' Support
-is challenging because of the diversity of viewpoints and needs. The
-group constantly examines its policies to determine whether they are
-indirectly alienating part of the community or encouraging particular
-types of use.
-%
-\toolname{nanoHUB}'s 10-year history with over 260,000 users annually
-provides a lot of data to analyze: 4000 resources contributed by 1000
-authors. \toolname{nanoHUB} serves both the research and education
-community and the contribution model allows researchers to get their
-codes out into the community and in use in education very rapidly. The
-primary software challenges are twofold --- support for the
-\toolname{HUBzero} framework and challenges related to the software
-contributed by the community.
-
-The group has learned that community contributions are maximized with
-a tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3
-license so contributors can create unique components and license as
-they choose. If they make changes to source code, the original license
-must be maintained for redistribution. As far as contributed
-resources, these must be open access, but not necessarily open
-source. This allows contributors to meet the requirements of their
-institutions and funding agencies. Quality is maintained via user
-ratings. Documentation is encouraged and nanoHUB supplies regression
-test capabilities, but the user community provides ratings, poses
-questions and contributes to wishlists and citation counts, all of
-which incentivize code authors.
-
-Terrel~\cite{Terrel_WSSSPE} describes support for the Python
-scientific community through two major efforts: the SciPy conference
-and the NumFOCUS foundation.  Since software sustainability relies on
-contributions from all sectors of the user community, these efforts
-support these sectors, and help develop and mature Python.
-%
-The reliance on software in science has driven a huge demand for
-development, but this development is typically done as a side effort
-and often in a rush to publish without documentation and
-testing. While the software is often created by academics, software
-support can fall to industrial institutions. SciPy brings together
-industry, government, and academics to share their code and experience
-in an open environment.
-%
-NumFOCUS in a non-profit that promotes open, usable scientific
-software while sustaining the community through educational programs;
-collaborative research tools and documentation; and promotion of
-high-level languages, reproducible scientific research, and open-code
-development. Governance is a loose grantor-grantee relationship with
-projects, allowing funds to be placed in the groups' accounts. This
-has raised money to hire developers for open code development,
-maintain testing machines, organize the PyData conference series, and
-sponsor community members to attend conferences.
-
-L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
-which was started in 1996 by participants in the USA Binary Black Hole
-Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
-model, a community-oriented framework that allows researchers to
-easily work together with reusable and extendable software
-elements. Modules are compiled into an executable and can remain
-dormant, becoming active only when parameters and simulation data
-dictate. Users can use modules written by others or can write their
-own modules without changing other code. The community has grown and
-diversified beyond the original science areas.
-
-The paper points out four keys to sustaining the community: modular
-design, growing a collaborative community, career paths, and credit. In
-modular design, the Cactus project went far beyond standard practices
-of APIs. Domain specific languages (DSLs) allow decoupling of
-components --- for example I/O, mesh refinement, PAPI counters, and
-boundary conditions abstracted from science code. In academia,
-publications are the main currency of credit. Because the project
-connects code developments to science, the work is publishable and
-modules are citable. Because of the open source, modular approach,
-programmers can see the impact of their contributions and often
-continue work after graduation. Career paths remain a challenge,
-however. Tasks that are essential from a software engineering
-perspective are often not rewarded in academia. The best programmers
-in a science environment often have multidisciplinary expertise. This
-also is not rewarded in academia.
-
-Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF
-software institute effort to build a community of those creating
-science gateways for science. These gateways, as described in some of
-the other papers in this section, can be quite capable and can have
-strong scientific impact.  Challenges are similar to those highlighted
-by other papers in this section: the conflict between funding for
-research vs infrastructure and the challenges around getting academic
-credit for infrastructure.
-%
-The authors observe that development is often done in an isolated
-hobbyist environment. Developers are unable to take advantage of
-similar work done by others, finding themselves in isolation even when their projects have common
-goals. But often projects struggle for sustainable funding because
-they provide infrastructure to conduct research and many times only
-the research is funded. Gateways also may start as a small group
-research project, taking off in popularity once they go live, without
-any long term plans for sustainability.  Subsequent disruptions in
-service can limit effectiveness and test the limits of the research
-community's trust. %The impact of gateways can be increased
-%substantially if we understand
-%what makes them successful.
-
-Recommendations from an early study of successful gateways include: 1)
-Leadership and management teams should design governance to represent
-multiple strengths and perspectives, plan for change and turnover in
-the future, recruit a development team that understands both the
-technical and domain-related issues, consider sustainability and
-measure success early and often. 2) Projects should hire a team of
-professionals, demonstrate credibility through stability and clarity
-of purpose, leverage the work of others, and plan for flexibility. 3)
-Projects should identify one ore more existing communities and understand the
-communities' needs before beginning, then adapt as the communities'
-needs evolve. 4) Funders, should consider the technology project
-lifecycle, and design solicitations to reward effective planning,
-recognize the benefits and limitations of both technology innovation
-and reuse, expect adjustments during the production process, copy
-effective models from other industries and sectors, and encourage
-partnerships that support gateway sustainability.
-
-%Through a business incubator type approach, the institute plans to
-%provide a variety of services that could be shared amongst
-%projects. Consultation and resources on topics such as business plan
-%development, software engineering practices, software licensing
-%options, usability, security and project management as well as a
-%software repository and hosting service will be available. Experts
-%will be available for multi-month assignments to help research teams
-%build their own gateways, teaching them how to maintain and add to
-%the work after the collaboration ends. Forums, symposia and an annual
-%conference will connect members of the development community. A
-%modular, layered framework that supports community contributions and
-%allows developers to choose components will be delivered and finally
-%workforce development activities will help train the next generation
-%for careers in this cross- disciplinary area and build pools of
-%institutional expertise that many projects can leverage. Shared
-%services and forums can dramatically reduce the cost of building and
-%sustaining gateways. Workforce development can encourage technology
-%professionals to remain in the sciences.
-
-\subsubsection{What are communities?}
-
-The workshop did not directly answer the question ``What are
-communities?'' but instead a number of different answers were
-indirectly presented, through the depiction of individuals and
-stakeholders in different aspects of the scientific software
-lifecycle.  In broad strokes, however, scientific software communities
-were generally accepted as consisting of individuals, often but not
-always composed of scientist practitioners, that were working with
-some degree of coordination toward a common goal enabled by software.
-
-The discussion of development-focused communities centered around
-describing methods of interaction between individuals and the
-scientific software.  The first type of interaction was the
-development of a specific piece of software, the second was a
-particular domain or discipline, and the final primary type of
-interaction was around the development of applications built on a
-particular piece of software that was perhaps developed by another
-group.  As an example, in~\cite{Hanwell_WSSSPE}, the community
-described is comprised of both the for-profit company
-\toolname{Kitware} and the users and contributors to their software
-packages such as \toolname{VTK}.  This structure, of the centralized
-development of core infrastructure around which communities of
-individuals applying that infrastructure and developing applications
-utilizing it, was similarly reflected in~\cite{Terrel_WSSSPE}, where
-the core scientific python ecosystem is supported by a non-profit
-entity that fosters community investment in that ecosystem.  In many
-ways, these two organizations (\toolname{Kitware} and
-\toolname{NumFOCUS}) attempt to cross domain boundaries and provide
-support for both the infrastructure and application sides of community
-building.
-
-\subsubsection{Measuring community}
-How might a project know when it has built a sustainable community?
-How might an outsider be able to assess the activity and
-sustainability of that community?  These questions have been partially
-addressed in the literature.  For example, a key metric in online
-communities in general is the cross-over point where there are more
-external contributors than internal ones. Richard Millington's book
-``Buzzing Communities'' does an excellent job of outlining these
-measures, drawing on communities research in an accessible
-manner~\cite{millington_buzzing_2012}. Some participants in the
-workshop have since prepared materials outlining current and
-future practices in measurement of scientific software and its
-ecosystem~\cite{metrics-web}.
-%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please
-%change to cite}
-
-\subsubsection{Additional Resources for learning about software communities}
-
-Scientific software communities were viewed as a subset of software
-communities as whole.  As such, resources applicable to generic
-software communities -- such as open source and proprietary technology
-companies -- can be used as input and as guiding understanding of how
-to steward and develop scientific software communities.  Because
-incentive structures are different in industry and volunteer-based
-open source communities, these can provide rough guidelines but not
-necessarily identically applicable.  The analogy between corporations
-and scientific investigators (particularly in terms of competition,
-cooperation and competitive advantage) has been explored in the
-literature below, but because of the different incentive structure the
-analogy is not universally true.
-
-The literature below, suggested by attendees, addresses both
-non-scientific software projects, as well as scientific projects.  The
-selections address both descriptive and prescriptive approaches to
-communities.
-
-Both~\cite{howison_scientific_2011} and~\cite{howison_incentives_2013}
-study how scientific software collectives self-organize and address
-issues of incentive, long-term support, and development of
-infrastructure as well as new features.  As noted elsewhere in this
-summary, \cite{Turk:2013:SCH:2484762.2484782} shared prescriptions
-from two software communities in astrophysics.
-
-From the perspective of developing prescriptions for successful
-scientific software development, both~\cite{citeulike:11831265} and
-\cite{1749-4699-6-1-015010} share experiences and suggestions for
-developing sustainable practices.  \cite{citeulike:11831265} proposes
-"ten simple rules" for developing open source scientific software,
-focusing on both the choices made during development and the
-sustainability of practices in the long term.
-\cite{1749-4699-6-1-015010} describes the development and long-term
-growth of the \texttt{deal.II} library, and how its place in its
-ecosystem of libraries, applications and domains has shaped its
-development and community trajectory.
-
-From more traditional open source development, resources were shared
-that developed communities explicitly, such
-as~\cite{citeulike:7888211} and~\cite{Trapani:2011}, focusing on
-large-scale projects such as the Ubuntu Linux distribution and
-smaller-scale volunteer-developed projects like such as ThinkUp,
-respectively. The process of open source development, while less
-explicitly focused on community building, sketched
-in~\cite{citeulike:478633} was seen as a valuable resource,
-particularly when combined with the management and personal
-interaction techniques outlined in~\cite{opac-b1134063}.  Growing
-diversity in communities was directly addressed
-in~\cite{Allsopp:2012}, where experiences growing the diversity of
-technical conferences in open source were described.
-
-\subsection{Industry \& Economic Models}
-
-Several papers presented discussed the connection between industry and
-scientific software, from the perspective of both integrating efforts
-between the two and sustaining long-term development.
-
-Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of
-open source software development at \toolname{Kitware}.  In
-particular, they focus on their success at growing their community of
-users through enabling multiple channels of communication, directly
-reaching out to individuals, and lowering the barrier to entry for
-contributions. This involves providing clear, test-oriented and
-review-based mechanisms for evaluating contributions, permissive
-licenses, and a service-based model for sustaining development.  This
-model enables \toolname{Kitware} to receive both public funding, as
-well as private funding to support improvements and targeted
-developments of the software.
-
-Foster et al.~\cite{Foster_WSSSPE} discussed the approach of
-developing sustainability models around Software-as-a-Service (SaaS)
-platforms, with the target example being that of
-\toolname{GlobusOnline}.  The authors build a case that both
-grant-based and volunteer-based development fall short in sustaining
-software, resulting in software that is disproportionately difficult
-to use compared to its functionality, which they note directly impacts
-the overall scientific productivity of its users.  In contrast, by
-charging a subscription fee for hosted, centrally-managed software
-(similar to offerings by \toolname{Dropbox}, \toolname{EverNote},
-\toolname{GMail}), the authors propose to manage the funding cycle and
-enable a greater focus on the aspects of software that directly impact
-individuals, rather than funders.  \toolname{Globus} has deployed such
-a service, for which they have attempted to develop a sustainable
-economic model that reduces institutional obstacles to funding and
-subscription.  However, they do identify that cultural obstacles do
-still remain, and they note a particular difference in culture between
-NSF- and NIH-funded researchers.
-
-\subsection{Education \& Training}
-
-The papers describing education and training were focused primarily on
-how these aspects of community development impact on the long-term
-sustainability of software projects.  \cite{Girotto_WSSSPE} described
-the impact of the mandate within the International Centre for
-Theoretical Physics (ICTP) to foster resources and competences in
-software development and HPC, resulting in the development of
-educational curricula directed in this area.  The paper itself
-described the changes made in these curricula as a result of the
-current changes in the HPC and scientific software landscape due to
-the advent of scripting languages, new programming paradigms and new
-types of hardware such as accelerator technologies.  The development
-of a workshop, with carefully selected participants and an immersive
-approach to learning, was identified as a major success for educating
-and developing new scientific software developers from targeted
-domains.
-
-Elster~\cite{Elster_WSSSPE} also points out how the 
-prevalence and rapid growth of multi and many-core systems forces 
-awareness of data locality and synchronization issues
-if one want to teach people how to develop high-performing scientific codes.
-
-\cite{Crawford_WSSSPE} addressed education and training within
-computational chemistry frameworks, particularly as these frameworks
-attempt to address next-generation computer hardware and software and
-as chemistry courses emphasize lab work over computational education.
-The authors identify this lack of computational awareness and training
-as the primary challenge to future advances in computational
-chemistry.  The authors propose a new institute for computational
-chemistry, emphasizing collaboration (and a licensing structure, such
-as LGPL or more permissive) and education of future generations of
-chemistry researchers.
-
-%\section{Other Discussion} \label{sec:other}
-%
-%\note{from the misc questions/answers/observations part of the google doc?}
-%
-%\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
-%
-%\todo{perhaps remove this section}
-
-\section{Cross-cutting Issues} \label{sec:cross-cutting}
-
-Three issues, how to define software sustainability, how career
-paths (or their lack) interact with achieving it, and the impact of
-software licenses, were raised across
-the workshop's panels.  This section aims to synthesize these
-discussions from different perspectives.
-
-\subsection{Defining Sustainability}  \label{sec:defining-sustainability}%\note{lead: Daniel S. Katz}
-
-What is, or should be meant by ``sustainability'' in the context of
-software came up in many different parts of the workshop, specifically
-in the first keynote (\S\ref{sec:keynote1}), the Developing and
-Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
-(\S\ref{sec:policy}). It quickly became clear that at present there is
-no consensus among the community, whether within or across
-disciplines, on what this definition should be, and that a variety of
-different definitions were being assumed, used, or sometime expressly
-articulated by contributors and attendees. However, some concepts,
-particularly relating sustainability to change over time, were also
-evidently held in common. This common notion is, for example, captured
-in the definition used by the UK's Scientific Software Sustainability
-Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
-will be available---and continue to be improved and supported---in the
-future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
-software that continues to serve its users.
-
-Philip Bourne, too, used the relation to change over time when he
-suggested in his opening keynote that sustainability can perhaps be
-defined as the effort needed to make the essential things continue.
-This leads to having to decide what it is that we want to sustain,
-whether what we want to sustain is valuable, and finally, who would
-care and how much if it went away. As was pointed out during a
-discussion session, OSS Watch, an advisory service for issues relating
-to free and open source software in the UK, proposes a Software
-Sustainability Maturity Model to address the issue of what level of
-sustainability a particular element of software needs to have, and
-where this is important. It, too, expresses sustainability in relation
-to change over time:
-\begin{quote}``When choosing software for procurement or
-development reuse --- regardless of the license and development model
-you will use --- you need to consider the future. While a software
-product may satisfy today's needs, will it satisfy tomorrow's needs?
-Will the supplier still be around in five years' time? Will the
-supplier still care for all its customers in five years' time? Will
-the supplier be responsive to bug reports and feature requests? In
-other words, is the software sustainable?''~\cite{OSS-ssmm-web}
-\end{quote}
-
-Attendees suggested that having a definition of sustainability on
-which the community can agree is key.  A related question that was
-raised is what the goal of sustainability should be, with a wide range
-of possible answers, including more reproducible science, software
-persistence, and quality.  And given a goal of sustainability, how
-would success in achieving it be measured?  How would one know that a
-piece of software has reached sustainability? Participants pointed out
-that for truly sustainable software there should be no endpoint at
-which sustainability can be claimed, because the software products
-would continue to be used and useful beyond the single institution,
-grant, and developer or development team that created them. This may
-mean that sustainability needs to be addressed throughout the full
-software life cycle.  It was also pointed out that software
-sustainability is not isolated from other attributes of scientific
-software and its use, such as usability, and provenance. Similarly,
-the question was considered, albeit only briefly, whether proprietary
-versus open-source license plays a role in the context of software
-sustainability. For example, should a project ensure that it uses an
-OSI-approved license so that software products can outlive any single
-entity if they remain important.
-
-Because part of the Policy panel (\S\ref{sec:policy}) was about
-modeling sustainability, and modeling requires defining what will be
-modeled, this panel saw particular attention to the questions
-surrounding the definition of sustainability. Two papers, Venters et
-al.~\cite{Venters_WSSSPE} and Katz and Proctor~\cite{Katz_WSSSPE},
-specifically discuss the issue.
-
-According to Venters et al.~\cite{Venters_WSSSPE}, sustainability is a
-rather ambiguous concept, and the lack of an accepted definition
-hampers integrating the concept into software engineering. They
-suggest that sustainability falls under the category of non-functional
-requirements, and that a software's sustainability is a consequence of
-a set of central software architecture quality attributes, including
-extensibility, interoperability, maintainability, portability,
-reusability, and scalability. They also propose an evaluation
-framework with which quality and sustainability could be measured at
-the architectural level.
-
-Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that
-could be used to measure software sustainability:
-\begin{itemize}
-\item Will the software continue to provide the same functionality in
-  the future, even if the environment (other parts of the
-  infrastructure) changes?
-\item Is the functionality and usability clearly explained to new
-  users?
-\item Do users have a mechanism to ask questions and to learn about
-  the software?
-\item Will the functionality be correct in the future, even if the
-  environment changes?
-\item Does it provide the functionality that existing and future users
-  want?
-\item Will it incorporate new science, theory, or tools as they
-  develop?
-\end{itemize}
-Despite their phrasing, these questions are not intended to be given
-simplistic yes or no answers, and it is the complete set rather than
-any individual one that would determine where in the range of
-sustainability a software falls.
 
@@
 
+One area in which there was not complete overlap was whether the word
+(and thus the effort called for by the WSSSPE workshop) involved
+environmental sustainability. Of course the word sustainability has
+strong connotations from consideration of environmental issues,
+evoking some mention of the areas in which software interacts with
+overall environmental resource usage, particularly energy efficiency.
+The two papers in this area which mentioned
+this~\cite{Venters_WSSSPE,Calero_WSSSPE} did so without integrating
+that analysis into the question of software being around long-term,
+suggesting that questions of environmental impact of scientific
+software is a conceptually distinct area of inquiry.
+
+One group of papers presented frameworks that were primarily about
+characteristics of software artifacts, connecting with the long
+discourse on software quality. This approach is realized in adjectives
+that can be applied to pieces of software but might also extend to
+describe software projects.  Thus Calero et al.~\cite{Calero_WSSSPE}
+propose adding elements to the ISO standards for measuring software
+quality. Specifically, they propose an additional dimension of quality
+they call ``perdurability'' with
+three defining characteristics: reliability, maintainability, and
+adaptability. This overlaps with the framework   by
+Venters et al.~\cite{Venters_WSSSPE} who employ the features
+``extensibility, interoperability, maintainability, portability,
+reusability and scalability,'' anticipating the sorts of work that
+people would need to do with a software artifact in the future, ``as
+stakeholders requirements, technology and environments evolve and
+change.'' Venters et al.~argue that these choices need to be made 
+early because they are related to the architecture of the software and
+involve trade offs that ought to be analyzed alongside each other. Lenhardt et
+al.~\cite{Lenhardt_WSSSPE} compare the software lifecycle to the data
+lifecycle to argue for the inclusion of metadata throughout a piece of
+software's life (discussing, for example, how it has been built and
+tested and what ``data in'' and ``data out'' has been considered). In
+addition, their analogy suggests that the software lifecycle might add
+a phase of ``preservation'' and draw on the understanding of what that
+involves from studies of data.  In sum, then, these frameworks focused
+on what needs to be accomplished to have more sustainable software.
+
+A second theme in these papers was the continued availability of
+resources to accomplish the goals of sustainability. The elements of
+these frameworks focused far more on the organization of a software
+project than they did on characteristics of the artifact itself
+(although it is certainly true that the adjectives discussed above
+could be applied to a software project). For example Pierce et
+al.~\cite{Pierce_WSSSPE} focus on the way that the project is run,
+particularly in terms of how those involved communicate and jointly
+set priorities, a process they call governance. In particular they
+argue that because sustainability is related to having ongoing resources,
+governance must be open to receive diverse input (by occurring online,
+asynchronously, or publicly) and thus have the potential to
+``transform passive users into active contributors.'' They argue that
+the Apache Software Foundation's incubation process teaches this and
+could be learned from by projects throughout scientific software.
+Katz and Proctor~\cite{Katz_WSSSPE} also discuss governance,
+describing two modes: ``top-down'' and ``bottom-up'' governance.  They
+place governance alongside questions of the artifact (``technical''),
+and questions of the origins of resources, in terms of who is funding
+the work surrounding a piece of software (``the political'') and the
+manner in which resources come to the project initially (commercial,
+open source, closed partnerships, grant funded) and long-term (all
+four plus paid support).
+
+Frameworks also concerned themselves with the context in which
+software projects exist, moving in abstraction from the software
+artifact itself to the organization of its production and the shape of
+the space in which an artifact or project exists.  These frameworks
+take the form of contingency theories in that they outline a different
+set of challenges and argue that different project organizations (and
+presumably artifact attributes) are necessary to persist long term in
+spaces with particular characteristics.  Katz and
+Proctor~\cite{Katz_WSSSPE} propose thinking of this idea of space in
+terms of three axes: temporal (long or short term needs), spatial
+(local or global use) and purpose (specific to general).  They propose
+that different project organizations will be needed in different
+locations and argue that we should concentrate research to understand
+those connections. Weber et al.~\cite{Weber_WSSSPE} describe their
+spaces by analogy with natural ecosystems as ``niches'' which sustain
+particular pieces of software. They define ``a software niche as the
+set of technical requirements, organizational conditions and cultural
+mores that support its maintenance and use over time.'' They call for
+better understanding and modeling of niches (as well as further
+exploration of the usefulness of ecosystem metaphors).
+
+\subsection{Credit, Citation, and Impact}
+
+How work on scientific software is recognized and rewarded strongly
+influences the motivation for particular kinds of work on scientific
+software. A recurring theme of the panel discussion was that software
+work in science is inadequately visible within the reputation system
+underlying science; in other words it often doesn't ``count''. In his
+paper for this workshop, Katz placed software work along with other
+``activities that facilitate science but are not currently rewarded or
+recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
+to ``support all researchers in presenting meaningful impact evidence
+in tenure, promotion, and funding applications.''~\cite{Priem_WSSSPE}.
+Knepley et al.~argued that the lack of visibility of software that
+supported a piece of science ``can have detrimental effects on
+funding, future library development, and even scientific
+careers.''~\cite{Knepley_WSSSPE}.
+
+%Anne: The folowing sentence is verbatim from the original Abstract..
+% Se preivious commit for suggestion on hwo to shorten this sentence
+These papers, and the discussion at the workshop, join a nascent
+literature seeking to understand what drives software work in science
+and how the reward systems of science thereby shape the type of
+software work undertaken. This study includes the extent to which developers are
+motivated to build software for the long-term, for the use of others,
+and whether to work collaboratively or
+separately~\cite{howison_incentives_2013, howison_scientific_2011,
+  bietz_synergizing_2010}. Software work is not only motivated by
+direct citations, but the visibility of software work in the
+literature is important to those who write software used in science.
+
+Papers and discussion concentrated on three overarching questions: How
+ought software work be visible, what are the barriers to its
+visibility, and what can be done to make it more visible?
+
+Most of the papers in this area focused on visibility of software in
+scientific papers, since scientific papers are the most widely
+accepted documentation of achievement in science. It was noted that
+there appear to be no widely accepted standards on how the use of
+software towards a paper ought to be mentioned, and that journals,
+citation style guides and other guides to scientific conduct are vague
+about how to describe software. To address this, papers advocated the
+need for a fixed identifier for software, either directly through a
+mechanism such as a Digital Object
+Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
+written to document the software (and perhaps its creation), a
+``software paper''~\cite{Chue_Hong_WSSSPE}. However, as was pointed
+out during the panel discussion, one of the problems with papers as
+the cited product is that their author list is fixed in time, which
+discourages potential contributors who are not on the original author
+list from designing incremental improvements as integration work
+rather than separate (and hence possibly rewritten) software
+products~\cite{howison_incentives_2013}.
+
+Another approach is to reduce the difficulty of citing all software
+underlying a research paper. For example, scientists often work with
+software that itself wraps other software, leading to attribution
+stacking that can make it non-obvious or even difficult to determine
+what attributions would be appropriate. Knepley et
+al.~\cite{Knepley_WSSSPE} approach this by proposing a mechanism by
+which the software itself, after it has run, provides the user with a
+set of citations, according to the pieces of code actually
+executed. They describe a prototype implementation whereby the
+citations are embedded in libraries and reported along with the
+results, via a command-line
+interface~\cite{Knepley_WSSSPE}. Discussion highlighted the difficulty
+that attempting to acknowledge the contributions of all pieces of
+dependent code within a paper faces the difficulty of creating very
+long citation lists, straining the analogy of code used to papers
+cited. Katz approaches this issue by proposing a system of
+``transitive credit,'' recording dependencies and relative
+contributions outside particular papers, relieving authors from the
+responsibility of acknowledging each and every dependency. Instead
+authors would acknowledge the percentage contribution of the software
+they used directly and an external system would then be able to
+recursively allocate that credit to those who had provided
+dependencies. Finally Priem and Piwowar argued that machine learning
+techniques could examine the body of published literature and extract
+mentions of software, coping with the multitude of informal ways in
+which authors mention software they used~\cite{Priem_WSSSPE}.  A point
+raised in the panel discussion was that instead of asking users to
+improve their software citation practices, one can also ask how
+software projects can better monitor the literature to improve their
+ability to show impact. For example, the NanoHub project scans the
+literature using keywords and names of known users to discover papers
+that are likely to have used their software and platform, and assigns
+graduate students to read each paper, highlighting mentions of
+software use and sometimes following up with the authors to identify
+stories for demonstrating impact.  A process for tracking
+software-using publications with the goal of increasing impact
+visibility is now described at \url{http://publications.wikia.com}.
+
+Potential visibility, and thus acknowledgement of scientific software
+products is not restricted to publications. Another key location for
+visibility is in the grant funding process, and as emphasized by NSF
+representatives at the meeting, recent changes to grant proposal and
+reporting formats now allow both applicants and awardees to report and
+highlight software products as much as publications. Nonetheless,
+whether peer review panels would value these contributions in the same
+way as publications remains to be seen.
+
+Priem and Piwowar argued that assessing the impact of software work
+requires looking beyond publications, including evidence of
+contribution and impact recorded in social coding-oriented resources
+such as GitHub, and conversations about software in issue trackers,
+mailing lists, twitter and beyond~\cite{Priem_WSSSPE}. In keeping with
+a principle of the ``altmetrics'' approach, they advocate that
+scholars should have resources that empower them to tell their own
+stories in the manner most appropriate for them and their audiences.
+
+\subsection{Implementing Policy}
+
+The workshop contributions in this group were concerned with the
+aspect of how implementation of best practices and other
+recommendations for improving scientific software sustainability could
+be promoted. Specifically, if scientific software is to become more
+sustainable, corresponding policies and guidelines need to be such
+that the scientific community can follow and implement them. This is
+considerably more challenging that it might seem at first, because in
+the reality of science today resources, both financial and personnel,
+that could be devoted to implementation are very limited, and the
+reward system does not encourage scientists to do so. Furthermore,
+implementing sustainability-targeting policies and guidelines often
+takes a variety of specialized software engineering expertises, which
+are not necessarily found in a single engineer, and much less so in a
+domain scientist cross-trained in programming. Adding to the policy
+implementation challenges, applicable sustainability-promoting
+practices and guidelines will change through a software project's
+lifecycle, in particular as it gains maturity.
+
+Two of the papers in this group focus on specific facets of software
+design that are important factors in a project's sustainability but
+are often addressed only late in the scientific software development
+cycle, if at all: Krintz et al.~\cite{Krintz_WSSSPE} look at API
+governance, and Heiland et al.~\cite{Heiland_WSSSPE} discuss maturity
+models for software security. The other two papers discuss
+implementation strategies for science from the perspective of
+facilitating many or all facets of sustainability-oriented software
+design: Blanton and Lenhardt~\cite{Blanton_WSSSPE} contrast large
+projects that have software infrastructure development built-in, with
+cross-training domain scientist PIs in software engineering best
+practices. Huang and Lapp~\cite{Huang_WSSSPE} discuss how various
+specialized software engineering skills could be turned into shared
+instrumentation with low barriers to access.
+
+Krintz et al.~\cite{Krintz_WSSSPE} describe how in an era in which
+computing frequently takes place in a distributed cloud, the control
+over digital resources is increasingly shifting from physical
+infrastructure to APIs, in particular web-service APIs. Yet, as Krintz
+et al.~observe, unlike for physical IT infrastructure in data centers,
+science communities have developed very little in the way of practices
+and technology for API governance, referred to by Krintz et al.~as the
+``combined policy, implementation, and deployment control''. Web APIs
+can and do change, sometimes quite frequently, raising the need to
+port dependent applications. The effort required for porting is
+notoriously difficult to estimate, making it nearly impossible for IT
+organizations to assess and thus properly manage the impact of API
+changes. To address this, Krintz et al.~propose a mechanism that
+evaluates the porting effort between two versions of a web-service API
+in a formal and automated way. To analyze the porting effort, they
+divide API changes into \emph{syntactic similarity}, the changes in
+inputs and outputs, and into \emph{semantic similarity}, the changes
+in the API's behavior. In initial tests, their method showed good
+congruence with human developers in scoring porting effort, offering
+the possibility that API governance can become as solid a part of
+scientific IT management as data center infrastructure management is
+today.
+
+Many facets of engineering for software sustainability strongly depend
+on the maturity level of the software. However, the maturity level of
+a software project meant to be sustained changes tremendously over its
+development life cycle, and the eventual maturity level is often
+difficult to predict during initial development. Using software
+security as their case study, Heiland et al.~\cite{Heiland_WSSSPE}
+discuss how Maturity Models can be used to formalize best practices
+appropriate for the different levels of maturity that a software
+project may go through over its lifecycle. Cybersecurity is also an
+example of a sustainability-relevant aspect in software design that is
+rarely given due diligence in science. In particular in industry,
+cybersecurity best practices for different stages of life cycle and
+maturity have been formalized as Software Security Maturity Models,
+and are widely used, yet awareness of these among scientific software
+development communities remains low. In providing a path to tightening
+security practices as software matures, such models align with the
+objective of providing implementation approaches that the scientific
+community can actually follow.
+
+API governance and cybersecurity measures appropriate for a project's
+maturity are all but two facets of sustainability-oriented software
+development. Others include user-centered interface design, test
+engineering, dependency management, and deployment operations. Each of
+these facets requires specialized skills and training in software
+engineering. How can the implementation of best practices and
+guidelines along many or all of these different facets be facilitated
+in scientific software development? Blanton and
+Lenhardt~\cite{Blanton_WSSSPE} contrast two models. In one, the time
+and personnel devoted to software engineering is ``co-funded'' with
+the driving science project. This typically implies large multi-year
+collaborative projects that to succeed require significant software
+infrastructure to be built, and which thus have the funding to support
+one or several software engineers. The sustainability of such projects
+then depends on sustaining the funding. In the other model, domain
+scientists also take the role of software developers, whether by
+necessity such as funding limitations, or due to cross-disciplinary
+professional interests. For this to result in sustainable software
+products, the domain scientists need to be (or become) cross-trained
+in software engineering standards and best practices.
+
+In practice, there are example for both extremes of the spectrum. For
+example, in the life sciences the iPlant Collaborative~\cite{iPlant},
+the Galaxy Project~\cite{Galaxy}, and \toolname{Qiime}~\cite{Qiime}
+are large multi-year projects with significant software infrastructure
+funding and deliverables. Typically though these are the exception
+rather than the rule, and particularly in the long tail of science the
+scientist-developer predominates, even for software that is widely
+used or crosses domain boundaries such as
+\toolname{rOpenSci}~\cite{rOpenSci}. For better training domain
+scientists at least in basic software engineering best practices,
+initiatives such as Software Carpentry~\cite{SoftwareCarpentry} have
+demonstrated how this could be achieved at scale.
+
+However, there may also be a middle ground between the two
+extremes. Huang and Lapp~\cite{Huang_WSSSPE} propose a Center of
+Excellence model that leverages economies of scale to make software
+engineering experts and their skills accessible to the long tail of
+science. As Huang and Lapp discuss, this model could effectively turn
+the utilization of software engineering expertise from a complex human
+resource recruitment and management challenge, to buying time, when
+and to the extent needed, on shared instrumentation. There is
+precedence to using such a model to lower access barriers for long
+tail science, in particular for new experimental technologies.  For
+example, although the acquisition and operation of next-generation
+high-throughput DNA sequencers requires substantial investments of
+money, time, and expertise, the establishment of ``core facilities''
+on many university campuses has made these technologies accessible to
+a wide swath of scientists, with transformative results for science.
+
+One of the important conclusions from this group of papers is that
+creating sustainable software requires paying attention not to one or
+two, but to several different facets of software engineering, each
+with corresponding best practices and standards of excellence. Even if
+a science project requires and has funding for a full software
+engineering FTE, what the project really needs could be fractions of
+different engineers with different specialty training. The vast
+majority of long tail science projects lacks the funding for even one
+full software engineer, let alone one who combines expertise in all of
+the applicable facets of engineering. Some scientists in the long tail
+will have the professional interests to cross-train enough in software
+engineering to be successful with creating sustainable software, but
+it is unrealistic to expect this expertise and interest of all or even
+the majority. This is where a software engineering center of
+excellence could provide a critical resource by enabling scientists in
+the long tail to complement their resources and expertise with facets
+that are missing, but which, if applied at the right time, would
+improve the chances of a software product to become sustainable. Such
+complementary expertise also need not be restricted to software
+engineering in the strict sense; for example, it could consist of
+community building, leadership, and support for some period of
+transition to sustainability.
+
+In summary, implementing software sustainability practices on a
+broader basis requires on the one hand the development of guidelines
+and practices that are suitable for the exploratory research context
+in which most scientific software is created, and on the other hand a
+skilled workforce trained in a variety of software engineering facets
+and community building. The capabilities afforded by such a workforce
+need to be accessible not only to large projects with sufficient
+funding to provide competitive employment, but also to the many
+smaller projects in the long tail of science. Sufficiently
+cross-training domain scientists could be one way to achieve this;
+another, complementary approach is to instrument the necessary
+capabilities so they can be shared.
+
+\section{Communities} \label{sec:community}%\note{lead: Matthew Turk, contributors: Nancy Wilkins-Diehr, Chris Mattmann, Suresh Marru, Frank L\"{o}ffler, Andy Terrel}
+
+%There were a number of papers categorized under the Communities banner and so
+%two presenters each summarized half of the submissions in this area.
+Across all talks and papers submitted, authors implicitly and
+explicitly recognized the concept of ``communities'' as a driving and
+unifying force within software projects.  Despite that, the actual
+nature of the communities, the incentive structures that bind them
+together, the infrastructure they utilize for communication and even
+the types of individuals that make up those communities were
+different.  Some of these communities were primarily composed of
+members of industry, some were funded and driven by individuals
+focused on developing software as opposed to utilizing it, and others
+were primarily composed of scientist practitioners, and were true
+communities of practice.
+
+These varying structures and compositions result in differing modes of
+interaction within communities, styles of development, and the
+structure of planning for future development of software.  In this
+section, we summarize the different types of communities in scientific
+software as well as the resultant impact on sustainability and
+development of functionality.
+
+\subsection{Communities}
+
+%This collection of papers was summarized by Karen Cranston.
+
+Drawing on experiences from high-energy physics, Vay et
+al.~\cite{Vay_WSSSPE} proposed developing teams of technical
+specialists   to overcome a lack of coordination between projects.
+Maximizing scientific output requires maximizing the usability of the
+scientific code while minimizing the cost of developing and
+supporting those code.  This included targeting different
+architectures for their software to be deployed, as well as
+coordination between technically-focused individuals and usage of a
+common scripting language between projects.  Instead of fragmenting
+the development of simulation codes across institutions, the paper
+suggests that a cohesive strategy reducing duplication and increasing
+coordination will broadly increase the efficiency across institutions.
+The approach proposed is of de-fragmenting the existing ecosystem in a
+non-disruptive way.
+
+Maheshwari et al.~\cite{Maheshwari_WSSSPE} focuses on ``technology
+catalysts'' and their role in the modern scientific research
+process. A technology catalyst is an individual with knowledge of
+technological advancements, tasked with user engagement to create
+scientific or engineering applications, using suitable tools and
+techniques to take advantage of current technological capabilities.
+%
+One of the tasks of technology catalysts is to seek community
+collaborations for new applications and engage users, thus benefiting
+both science, by effective running of scientific codes on
+computational infrastructure, and technology, by conducting research
+and seeking findings for technology improvement.  The particular
+engagements described in the paper came up from the lead author's work
+as a postdoctoral researcher at Cornell and Argonne, where interaction
+with the scientific communities in both institutions resulted in these
+collaborations.
+
+At NESCent, a combination of in-house informatics individuals and
+domain scientists collaborate to develop software to study
+evolutionary science.  The report~\cite{Cranston_WSSSPE} studied the
+success of a ``hackathon'' model for development, where short-form,
+hands-on events combining users, researcher-developers and software
+engineers targeted specific code improvements. From this experiment,
+the authors identified several key outcomes as well as
+lessons-learned: specifically, the co-localization of developers was
+seen as having a strong impact, enabling casual conversation that led
+to discrete outcomes.  The formation of the discussion mailing list,
+specifically in response to the social capital built at the hackathon,
+was seen as spurring on longer-term benefits to the community and
+fostering sustainability.
+
+Hart et al.~\cite{Hart_WSSSPE} addresses the success of the rOpenSci
+project in developing collaboration supporting tools for
+Open Science.  This software collective, organized around the
+statistical programming environment R, develops access mechanisms for
+data repositories and attempts to reduce the barrier to entry for
+individuals wanting to access data repositories and study the data
+contained therein.  The collective fosters direct collaboration
+between individuals and data providers, designed to ``train academics
+in reproducible science workflows focused around R.''  Two central
+challenges are engagement of existing users within ecology and
+evolutionary biology (EEB), and how the community can make inroads and
+traction in other disciplines. Currently, the collective is exploring
+addressing these challenges through the use of social media, holding
+workshops and hackathons.  This helps to both raise the profile of
+the collective within EEB and in other domains.  However, the
+overarching challenge identified in the paper was that of
+incentivizing maintenance of software, which is difficult in academia.
+
+%\subsection{Communities Part 2}
+
+%Nancy Wilkins-Diehr summarized 6 papers in this part.
+
+Christopherson et al.~\cite{Christopherson_WSSSPE} outlines the degree
+to which research relies on high quality software. There are often
+barriers and a lack of suitable incentives for researchers to embrace
+software engineering principles. The Water Science Software Institute
+is working to lower some of these barriers through an Open Community
+Engagement Process. This is a four-step iterative development process
+that incorporates Agile development principles.
+
+\begin{itemize}
+\setlength{\itemindent}{0.5in}
+\item[Step 1:] Design - thorough discussion of research questions
+\item[Step 2:] Develop working code
+\item[Step 3:] Refine based on new requirements
+\item[Step 4:] Publish open source
+\end{itemize}
+
+Christopherson reports on the application of Steps~1-3 to a
+computational modeling framework developed in the 1990s. Step~1 was a
+2-day, in-person specifications meeting and code walk-through. Step 2
+was a 5-day hackathon to develop working code, and Step 3 was a 3-day
+hackathon to refine the code based on new requirements. The team
+worked on small, low-risk units of code. It was challenging, revealed
+unanticipated obstacles, programmers had to work together, and
+experimentation was encouraged.
+%
+The paper recommended: start small and gradually building toward more
+complex objectives, consistent with Agile development; develop
+consensus before coding, by repeating step~1 before all hackathons;
+ensure newcomers receive orientation prior to the hackathon, such as a
+code walk-through or system documentation; and co-locate collaborators
+whenever feasible.
+
+%Refactor before adding
+%new functionality. Approach development as a learning experience. Welcome
+%experimentation, and treat mistakes as a natural part of the learning process.
+%Repeat Step 1 activities before all hackathons to develop consensus before
+%coding. This allows coding to be the focus of subsequent hackathons. In higher
+%risk situations, provide additional time for Step 1 activities. The
+%Christopherson team recommends a minimum of two months.
+%Ensure any newcomers
+%receive some form of orientation prior to the hackathon, such as a code
+%walkthrough or system documentation. Co-locate rather than collaborating
+%remotely whenever feasible.
+
+Pierce et al.~\cite{Pierce2_WSSSPE} describes how science gateways can
+provide a user-friendly entry to complex cyberinfrastructure. For
+example, more than 7,000 biologists have run phylogenetic codes on
+supercomputers using the CIPRES Science Gateway in $3 \tfrac{1}{2}$
+years. Over 120 scientists from 50 institutions used the
+\toolname{UltraScan Science Gateway} in one year, increasing the
+sophistication of analytical ultracentrifugation experiments
+worldwide. The new Neuroscience Gateway (\toolname{NSG}) registered
+more than 100 users who used 250,000 CPU hours in only a few months.
+
+Gateways, however, need to keep operational costs low and can often
+make use of common components, such as authentication, application
+installation and reliable execution and help desk support. Science
+Gateway Platform as a Service (\toolname{SciGaP}) delivers middleware
+as a hosted, scalable third-party service while domain developers
+focus on user interfaces and domain-specific data in the gateway.
+While \toolname{SciGaP} is based on the \toolname{Apache Airavata}
+project and the \toolname{CIPRES} Workbench Framework, community
+contributions are encouraged by its open source, open governance and
+open operations policies. The goal is robust, sustainable
+infrastructure with a cycle of development that improves reliability
+and prioritizes stakeholder requirements. The project is leveraging
+Internet2's \toolname{Net+} mechanisms for converting
+\toolname{SciGaP} and its gateways into commodity services.
+
+Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and
+challenges with the large \toolname{nanoHUB.org} community, where
+community is defined as a ``body of persons of common and especially
+professional interests scattered through a larger society.'' Support
+is challenging because of the diversity of viewpoints and needs. The
+group constantly examines its policies to determine whether they are
+indirectly alienating part of the community or encouraging particular
+types of use.
+%
+\toolname{nanoHUB}'s 10-year history with over 260,000 users annually
+provides a lot of data to analyze: 4000 resources contributed by 1000
+authors. \toolname{nanoHUB} serves both the research and education
+community and the contribution model allows researchers to get their
+codes out into the community and in use in education very rapidly. The
+primary software challenges are twofold --- support for the
+\toolname{HUBzero} framework and challenges related to the software
+contributed by the community.
+
+The group has learned that community contributions are maximized with
+a tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3
+license so contributors can create unique components and license as
+they choose. If they make changes to source code, the original license
+must be maintained for redistribution. As far as contributed
+resources, these must be open access, but not necessarily open
+source. This allows contributors to meet the requirements of their
+institutions and funding agencies. Quality is maintained via user
+ratings. Documentation is encouraged and nanoHUB supplies regression
+test capabilities, but the user community provides ratings, poses
+questions and contributes to wishlists and citation counts, all of
+which incentivize code authors.
+
+Terrel~\cite{Terrel_WSSSPE} describes support for the Python
+scientific community through two major efforts: the SciPy conference
+and the NumFOCUS foundation.  Since software sustainability relies on
+contributions from all sectors of the user community, these efforts
+support these sectors, and help develop and mature Python.
+%
+The reliance on software in science has driven a huge demand for
+development, but this development is typically done as a side effort
+and often in a rush to publish without documentation and
+testing. While the software is often created by academics, software
+support can fall to industrial institutions. SciPy brings together
+industry, government, and academics to share their code and experience
+in an open environment.
+%
+NumFOCUS in a non-profit that promotes open, usable scientific
+software while sustaining the community through educational programs;
+collaborative research tools and documentation; and promotion of
+high-level languages, reproducible scientific research, and open-code
+development. Governance is a loose grantor-grantee relationship with
+projects, allowing funds to be placed in the groups' accounts. This
+has raised money to hire developers for open code development,
+maintain testing machines, organize the PyData conference series, and
+sponsor community members to attend conferences.
+
+L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
+which was started in 1996 by participants in the USA Binary Black Hole
+Alliance Challenge. Cactus has a flesh (core) and thorns (modules)
+model, a community-oriented framework that allows researchers to
+easily work together with reusable and extendable software
+elements. Modules are compiled into an executable and can remain
+dormant, becoming active only when parameters and simulation data
+dictate. Users can use modules written by others or can write their
+own modules without changing other code. The community has grown and
+diversified beyond the original science areas.
+
+The paper points out four keys to sustaining the community: modular
+design, growing a collaborative community, career paths, and credit. In
+modular design, the Cactus project went far beyond standard practices
+of APIs. Domain specific languages (DSLs) allow decoupling of
+components --- for example I/O, mesh refinement, PAPI counters, and
+boundary conditions abstracted from science code. In academia,
+publications are the main currency of credit. Because the project
+connects code developments to science, the work is publishable and
+modules are citable. Because of the open source, modular approach,
+programmers can see the impact of their contributions and often
+continue work after graduation. Career paths remain a challenge,
+however. Tasks that are essential from a software engineering
+perspective are often not rewarded in academia. The best programmers
+in a science environment often have multidisciplinary expertise. This
+also is not rewarded in academia.
+
+Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF
+software institute effort to build a community of those creating
+science gateways for science. These gateways, as described in some of
+the other papers in this section, can be quite capable and can have
+strong scientific impact.  Challenges are similar to those highlighted
+by other papers in this section: the conflict between funding for
+research vs infrastructure and the challenges around getting academic
+credit for infrastructure.
+%
+The authors observe that development is often done in an isolated
+hobbyist environment. Developers are unable to take advantage of
+similar work done by others, finding themselves in isolation even when their projects have common
+goals. But often projects struggle for sustainable funding because
+they provide infrastructure to conduct research and many times only
+the research is funded. Gateways also may start as a small group
+research project, taking off in popularity once they go live, without
+any long term plans for sustainability.  Subsequent disruptions in
+service can limit effectiveness and test the limits of the research
+community's trust. %The impact of gateways can be increased
+%substantially if we understand
+%what makes them successful.
+
+Recommendations from an early study of successful gateways include: 1)
+Leadership and management teams should design governance to represent
+multiple strengths and perspectives, plan for change and turnover in
+the future, recruit a development team that understands both the
+technical and domain-related issues, consider sustainability and
+measure success early and often. 2) Projects should hire a team of
+professionals, demonstrate credibility through stability and clarity
+of purpose, leverage the work of others, and plan for flexibility. 3)
+Projects should identify one ore more existing communities and understand the
+communities' needs before beginning, then adapt as the communities'
+needs evolve. 4) Funders, should consider the technology project
+lifecycle, and design solicitations to reward effective planning,
+recognize the benefits and limitations of both technology innovation
+and reuse, expect adjustments during the production process, copy
+effective models from other industries and sectors, and encourage
+partnerships that support gateway sustainability.
+
+%Through a business incubator type approach, the institute plans to
+%provide a variety of services that could be shared amongst
+%projects. Consultation and resources on topics such as business plan
+%development, software engineering practices, software licensing
+%options, usability, security and project management as well as a
+%software repository and hosting service will be available. Experts
+%will be available for multi-month assignments to help research teams
+%build their own gateways, teaching them how to maintain and add to
+%the work after the collaboration ends. Forums, symposia and an annual
+%conference will connect members of the development community. A
+%modular, layered framework that supports community contributions and
+%allows developers to choose components will be delivered and finally
+%workforce development activities will help train the next generation
+%for careers in this cross- disciplinary area and build pools of
+%institutional expertise that many projects can leverage. Shared
+%services and forums can dramatically reduce the cost of building and
+%sustaining gateways. Workforce development can encourage technology
+%professionals to remain in the sciences.
+
+\subsubsection{What are communities?}
+
+The workshop did not directly answer the question ``What are
+communities?'' but instead a number of different answers were
+indirectly presented, through the depiction of individuals and
+stakeholders in different aspects of the scientific software
+lifecycle.  In broad strokes, however, scientific software communities
+were generally accepted as consisting of individuals, often but not
+always composed of scientist practitioners, that were working with
+some degree of coordination toward a common goal enabled by software.
+
+The discussion of development-focused communities centered around
+describing methods of interaction between individuals and the
+scientific software.  The first type of interaction was the
+development of a specific piece of software, the second was a
+particular domain or discipline, and the final primary type of
+interaction was around the development of applications built on a
+particular piece of software that was perhaps developed by another
+group.  As an example, in~\cite{Hanwell_WSSSPE}, the community
+described is comprised of both the for-profit company
+\toolname{Kitware} and the users and contributors to their software
+packages such as \toolname{VTK}.  This structure, of the centralized
+development of core infrastructure around which communities of
+individuals applying that infrastructure and developing applications
+utilizing it, was similarly reflected in~\cite{Terrel_WSSSPE}, where
+the core scientific python ecosystem is supported by a non-profit
+entity that fosters community investment in that ecosystem.  In many
+ways, these two organizations (\toolname{Kitware} and
+\toolname{NumFOCUS}) attempt to cross domain boundaries and provide
+support for both the infrastructure and application sides of community
+building.
+
+\subsubsection{Measuring community}
+How might a project know when it has built a sustainable community?
+How might an outsider be able to assess the activity and
+sustainability of that community?  These questions have been partially
+addressed in the literature.  For example, a key metric in online
+communities in general is the cross-over point where there are more
+external contributors than internal ones. Richard Millington's book
+``Buzzing Communities'' does an excellent job of outlining these
+measures, drawing on communities research in an accessible
+manner~\cite{millington_buzzing_2012}. Some participants in the
+workshop have since prepared materials outlining current and
+future practices in measurement of scientific software and its
+ecosystem~\cite{metrics-web}.
+%\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please
+%change to cite}
+
+\subsubsection{Additional Resources for learning about software communities}
+
+Scientific software communities were viewed as a subset of software
+communities as whole.  As such, resources applicable to generic
+software communities -- such as open source and proprietary technology
+companies -- can be used as input and as guiding understanding of how
+to steward and develop scientific software communities.  Because
+incentive structures are different in industry and volunteer-based
+open source communities, these can provide rough guidelines but not
+necessarily identically applicable.  The analogy between corporations
+and scientific investigators (particularly in terms of competition,
+cooperation and competitive advantage) has been explored in the
+literature below, but because of the different incentive structure the
+analogy is not universally true.
+
+The literature below, suggested by attendees, addresses both
+non-scientific software projects, as well as scientific projects.  The
+selections address both descriptive and prescriptive approaches to
+communities.
+
+Both~\cite{howison_scientific_2011} and~\cite{howison_incentives_2013}
+study how scientific software collectives self-organize and address
+issues of incentive, long-term support, and development of
+infrastructure as well as new features.  As noted elsewhere in this
+summary, \cite{Turk:2013:SCH:2484762.2484782} shared prescriptions
+from two software communities in astrophysics.
+
+From the perspective of developing prescriptions for successful
+scientific software development, both~\cite{citeulike:11831265} and
+\cite{1749-4699-6-1-015010} share experiences and suggestions for
+developing sustainable practices.  \cite{citeulike:11831265} proposes
+"ten simple rules" for developing open source scientific software,
+focusing on both the choices made during development and the
+sustainability of practices in the long term.
+\cite{1749-4699-6-1-015010} describes the development and long-term
+growth of the \texttt{deal.II} library, and how its place in its
+ecosystem of libraries, applications and domains has shaped its
+development and community trajectory.
+
+From more traditional open source development, resources were shared
+that developed communities explicitly, such
+as~\cite{citeulike:7888211} and~\cite{Trapani:2011}, focusing on
+large-scale projects such as the Ubuntu Linux distribution and
+smaller-scale volunteer-developed projects like such as ThinkUp,
+respectively. The process of open source development, while less
+explicitly focused on community building, sketched
+in~\cite{citeulike:478633} was seen as a valuable resource,
+particularly when combined with the management and personal
+interaction techniques outlined in~\cite{opac-b1134063}.  Growing
+diversity in communities was directly addressed
+in~\cite{Allsopp:2012}, where experiences growing the diversity of
+technical conferences in open source were described.
+
+\subsection{Industry \& Economic Models}
+
+Several papers presented discussed the connection between industry and
+scientific software, from the perspective of both integrating efforts
+between the two and sustaining long-term development.
+
+Hanwell et al.~\cite{Hanwell_WSSSPE} reflect on the 15-year history of
+open source software development at \toolname{Kitware}.  In
+particular, they focus on their success at growing their community of
+users through enabling multiple channels of communication, directly
+reaching out to individuals, and lowering the barrier to entry for
+contributions. This involves providing clear, test-oriented and
+review-based mechanisms for evaluating contributions, permissive
+licenses, and a service-based model for sustaining development.  This
+model enables \toolname{Kitware} to receive both public funding, as
+well as private funding to support improvements and targeted
+developments of the software.
+
+Foster et al.~\cite{Foster_WSSSPE} discussed the approach of
+developing sustainability models around Software-as-a-Service (SaaS)
+platforms, with the target example being that of
+\toolname{GlobusOnline}.  The authors build a case that both
+grant-based and volunteer-based development fall short in sustaining
+software, resulting in software that is disproportionately difficult
+to use compared to its functionality, which they note directly impacts
+the overall scientific productivity of its users.  In contrast, by
+charging a subscription fee for hosted, centrally-managed software
+(similar to offerings by \toolname{Dropbox}, \toolname{EverNote},
+\toolname{GMail}), the authors propose to manage the funding cycle and
+enable a greater focus on the aspects of software that directly impact
+individuals, rather than funders.  \toolname{Globus} has deployed such
+a service, for which they have attempted to develop a sustainable
+economic model that reduces institutional obstacles to funding and
+subscription.  However, they do identify that cultural obstacles do
+still remain, and they note a particular difference in culture between
+NSF- and NIH-funded researchers.
+
+\subsection{Education \& Training}
+
+The papers describing education and training were focused primarily on
+how these aspects of community development impact on the long-term
+sustainability of software projects.  \cite{Girotto_WSSSPE} described
+the impact of the mandate within the International Centre for
+Theoretical Physics (ICTP) to foster resources and competences in
+software development and HPC, resulting in the development of
+educational curricula directed in this area.  The paper itself
+described the changes made in these curricula as a result of the
+current changes in the HPC and scientific software landscape due to
+the advent of scripting languages, new programming paradigms and new
+types of hardware such as accelerator technologies.  The development
+of a workshop, with carefully selected participants and an immersive
+approach to learning, was identified as a major success for educating
+and developing new scientific software developers from targeted
+domains.
+
+Elster~\cite{Elster_WSSSPE} also points out how the 
+prevalence and rapid growth of multi and many-core systems forces 
+awareness of data locality and synchronization issues
+if one want to teach people how to develop high-performing scientific codes.
+
+\cite{Crawford_WSSSPE} addressed education and training within
+computational chemistry frameworks, particularly as these frameworks
+attempt to address next-generation computer hardware and software and
+as chemistry courses emphasize lab work over computational education.
+The authors identify this lack of computational awareness and training
+as the primary challenge to future advances in computational
+chemistry.  The authors propose a new institute for computational
+chemistry, emphasizing collaboration (and a licensing structure, such
+as LGPL or more permissive) and education of future generations of
+chemistry researchers.
+
+%\section{Other Discussion} \label{sec:other}
+%
+%\note{from the misc questions/answers/observations part of the google doc?}
+%
+%\note{all to put notes here of stuff that doesn't fit elsewhere, if any}
+%
+%\todo{perhaps remove this section}
+
+\section{Cross-cutting Issues} \label{sec:cross-cutting}
+
+Three issues, how to define software sustainability, how career
+paths (or their lack) interact with achieving it, and the impact of
+software licenses, were raised across
+the workshop's panels.  This section aims to synthesize these
+discussions from different perspectives.
+
+\subsection{Defining Sustainability}  \label{sec:defining-sustainability}%\note{lead: Daniel S. Katz}
+
+What is, or should be meant by ``sustainability'' in the context of
+software came up in many different parts of the workshop, specifically
+in the first keynote (\S\ref{sec:keynote1}), the Developing and
+Maintaining Software panel (\S\ref{sec:devel}), and the Policy panel
+(\S\ref{sec:policy}). It quickly became clear that at present there is
+no consensus among the community, whether within or across
+disciplines, on what this definition should be, and that a variety of
+different definitions were being assumed, used, or sometime expressly
+articulated by contributors and attendees. However, some concepts,
+particularly relating sustainability to change over time, were also
+evidently held in common. This common notion is, for example, captured
+in the definition used by the UK's Scientific Software Sustainability
+Institute, quoted in~\cite{Venters_WSSSPE}: ``software you use today
+will be available---and continue to be improved and supported---in the
+future''. Pierce et al~\cite{Pierce_WSSSPE} express this idea as
+software that continues to serve its users.
+
+Philip Bourne, too, used the relation to change over time when he
+suggested in his opening keynote that sustainability can perhaps be
+defined as the effort needed to make the essential things continue.
+This leads to having to decide what it is that we want to sustain,
+whether what we want to sustain is valuable, and finally, who would
+care and how much if it went away. As was pointed out during a
+discussion session, OSS Watch, an advisory service for issues relating
+to free and open source software in the UK, proposes a Software
+Sustainability Maturity Model to address the issue of what level of
+sustainability a particular element of software needs to have, and
+where this is important. It, too, expresses sustainability in relation
+to change over time:
+\begin{quote}``When choosing software for procurement or
+development reuse --- regardless of the license and development model
+you will use --- you need to consider the future. While a software
+product may satisfy today's needs, will it satisfy tomorrow's needs?
+Will the supplier still be around in five years' time? Will the
+supplier still care for all its customers in five years' time? Will
+the supplier be responsive to bug reports and feature requests? In
+other words, is the software sustainable?''~\cite{OSS-ssmm-web}
+\end{quote}
+
+Attendees suggested that having a definition of sustainability on
+which the community can agree is key.  A related question that was
+raised is what the goal of sustainability should be, with a wide range
+of possible answers, including more reproducible science, software
+persistence, and quality.  And given a goal of sustainability, how
+would success in achieving it be measured?  How would one know that a
+piece of software has reached sustainability? Participants pointed out
+that for truly sustainable software there should be no endpoint at
+which sustainability can be claimed, because the software products
+would continue to be used and useful beyond the single institution,
+grant, and developer or development team that created them. This may
+mean that sustainability needs to be addressed throughout the full
+software life cycle.  It was also pointed out that software
+sustainability is not isolated from other attributes of scientific
+software and its use, such as usability, and provenance. Similarly,
+the question was considered, albeit only briefly, whether proprietary
+versus open-source license plays a role in the context of software
+sustainability. For example, should a project ensure that it uses an
+OSI-approved license so that software products can outlive any single
+entity if they remain important.
+
+Because part of the Policy panel (\S\ref{sec:policy}) was about
+modeling sustainability, and modeling requires defining what will be
+modeled, this panel saw particular attention to the questions
+surrounding the definition of sustainability. Two papers, Venters et
+al.~\cite{Venters_WSSSPE} and Katz and Proctor~\cite{Katz_WSSSPE},
+specifically discuss the issue.
+
+According to Venters et al.~\cite{Venters_WSSSPE}, sustainability is a
+rather ambiguous concept, and the lack of an accepted definition
+hampers integrating the concept into software engineering. They
+suggest that sustainability falls under the category of non-functional
+requirements, and that a software's sustainability is a consequence of
+a set of central software architecture quality attributes, including
+extensibility, interoperability, maintainability, portability,
+reusability, and scalability. They also propose an evaluation
+framework with which quality and sustainability could be measured at
+the architectural level.
+
+Katz and Proctor~\cite{Katz_WSSSPE} propose a set of questions that
+could be used to measure software sustainability:
+\begin{itemize}
+\item Will the software continue to provide the same functionality in
+  the future, even if the environment (other parts of the
+  infrastructure) changes?
+\item Is the functionality and usability clearly explained to new
+  users?
+\item Do users have a mechanism to ask questions and to learn about
+  the software?
+\item Will the functionality be correct in the future, even if the
+  environment changes?
+\item Does it provide the functionality that existing and future users
+  want?
+\item Will it incorporate new science, theory, or tools as they
+  develop?
+\end{itemize}
+Despite their phrasing, these questions are not intended to be given
+simplistic yes or no answers, and it is the complete set rather than
+any individual one that would determine where in the range of
+sustainability a software falls.
+
 \subsection{Career Paths for Scientific Software Developers}\label{sec:career-tracks} %\note{lead: Daniel S. Katz}
@@
 \appendix
-\section{Papers Accepted and Discussed at WSSSPE1} \label{sec:papers}
-
-\section*{Developing and Supporting Software}
-
-\subsubsection*{Development Experiences}
-
-\begin{itemize}
-
-\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
-  McInnes, Barry Smith. Package Management Practices Essential for
-  Interoperability: Lessons Learned and Strategies Developed for
-  FASTMath~\cite{Miller_WSSSPE}
-
-\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
-  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
-
-\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
-  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
-  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
-
-\item Anshu Dubey, B. Van Straalen. Experiences from Software
-  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
-  adapting sustainable scientific software development processes in a
-  small biology lab~\cite{Crusoe_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
-The MVAPICH Project: Evolution and Sustainability of an Open Source
-Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
-
-\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
-  H. Kellogg. Experiences with Automated Build and Test for
-  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
-
-\begin{itemize}
-
-\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
-  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
-  Versatile Simulation of Large Scale Distributed
-  Systems~\cite{Casanova_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
-  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
-  Sherwin. Simplifying the Development, Use and Sustainability of HPC
-  Software~\cite{Cohen_WSSSPE}
-
-\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
-  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Best Practices, Challenges, and Recommendations}
-
-\begin{itemize}
-
-\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
-  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
-
-\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
-  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
-  R. Thakur, Software Abstractions and Methodologies for HPC
-  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
-
-\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
-  Need Not Be Difficult~\cite{Carver_WSSSPE}
-
-\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
-  K. Barnett, Von Welch. Initial Findings from a Study of Best
-  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
-
-\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
-  anything less is unsustainable~\cite{Brown_WSSSPE}
-
-\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
-  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
-  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
-  Platforms~\cite{Swenson_WSSSPE}
-
-\end{itemize}
-
-\section*{Policy}
-
-\subsubsection*{Modeling Sustainability}
-
-\begin{itemize}
-
-\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
-  Software Product Sustainability Model~\cite{Calero_WSSSPE}
-
-\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
-  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
-  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
-  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
-
-\item Daniel S. Katz, David Proctor. A Framework for Discussing
-  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
-
-\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
-  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
-  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
-
-\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
-  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Credit, Citation, Impact}
-
-\begin{itemize}
-
-\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
-  Smith. Accurately Citing Software and Algorithms Used in
-  Publications~\cite{Knepley_WSSSPE}
-
-\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
-  report for every software project~\cite{Priem_WSSSPE}
-
-\item Daniel S. Katz. Citation and Attribution of Digital Products:
-  Social and Technological Concerns~\cite{Katz2_WSSSPE}
-
-\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
-  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Reproducibility}
-
-\begin{itemize}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Victoria Stodden, Sheila Miguez. Best Practices for
-  Computational Science: Software Infrastructure and Environments for
-  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Implementing Policy}
-
-\begin{itemize}
-
-\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
-  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Daisie Huang, Hilmar Lapp. Software Engineering as
-  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
-
-\item Chandra Krintz, Hiranya Jayathilaka, Stratos
-  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
-
-\end{itemize}
-
-\section*{Communities, Models, and Education}
-
-\subsubsection*{Communities}
-
-\begin{itemize}
-
-\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
-
-\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
-  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
-
-\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
-  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
-  Modeling Science Activities~\cite{Vay_WSSSPE}
-
-\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
-  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
-  Mookherjee. Reusability in Science: From Initial User Engagement to
-  Dissemination of Results~\cite{Maheshwari_WSSSPE}
-
-\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
-  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
-  tools for facilitating Open Science~\cite{Hart_WSSSPE}
-
-\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
-  Software through the Open Community Engagement
-  Process~\cite{Christopherson_WSSSPE}
-
-\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
-  Borries Demeler. Science Gateway Operational Sustainability:
-  Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
-
-\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
-  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
-  Experiences and Challenges in Software Sustainability for a Large
-  Scientific Community~\cite{Zentner_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software
-  Community~\cite{Terrel_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation
-  Software~\cite{Loffler_WSSSPE}
-
-\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
-  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
-  Dooley, Dan Stanzione. Science Gateways and the Importance of
-  Sustainability~\cite{Wilkins-Diehr_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\end{itemize}
-
-In addition, the following paper from another area was also discussed
-in this area.
-
-\begin{itemize}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Industry \& Economic Models}
-
-\begin{itemize}
-
-\item Anne C. Elster. Software for Science: Some Personal
-  Reflections on Funding, Licensing, Publishing and Teaching~\cite{Elster_WSSSPE}
-
-\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
-  as a path to software sustainability~\cite{Foster_WSSSPE}
-
-\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
-  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
-  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
-  Scientific Software~\cite{Blanton_WSSSPE}
-
-\item Markus Blatt. DUNE as an Example of Sustainable Open Source
-  Scientific Software Development~\cite{Blatt_WSSSPE}
-
-\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
-  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
-  Open Source Production Quality MPI Library for
-  HPC~\cite{Panda_WSSSPE}
-
-\item Andy Terrel. Sustaining the Python Scientific Software
-  Community~\cite{Terrel_WSSSPE}
-
-\end{itemize}
-
-\subsubsection*{Education \& Training}
-
-\begin{itemize}
-
-\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
-  T. Brown. Advanced Techniques for Scientific Programming and
-  Collaborative Development of Open Source Software Packages at the
-  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
-
-\item Thomas Crawford. On the Development of Sustainable Software for
-  Computational Chemistry~\cite{Crawford_WSSSPE}
-
-\end{itemize}
-
-In addition, the following papers from other areas were also discussed
-in this area.
-
-\begin{itemize}
-
-\item Charles R. Ferenbaugh, Experiments in Sustainable Software
-  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
-
-\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
-  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
-
-\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
-  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
-  with Sustainable Software~\cite{Ahern_WSSSPE}
-
-\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
-  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
-
-\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
-  Schnetter. Cactus: Issues for Sustainable Simulation
-  Software~\cite{Loffler_WSSSPE}
-
-\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
-  Effects of Short-term Efforts: A Catalyst for Community Engagement
-  in Scientific Software~\cite{Trainer_WSSSPE}
-
-\end{itemize}
 
@@
 
+\section{Papers Accepted and Discussed at WSSSPE1} \label{sec:papers}
+
+\section*{Developing and Supporting Software}
+
+\subsubsection*{Development Experiences}
+
+\begin{itemize}
+
+\item Mark C. Miller, Lori Diachin, Satish Balay, Lois Curfman
+  McInnes, Barry Smith. Package Management Practices Essential for
+  Interoperability: Lessons Learned and Strategies Developed for
+  FASTMath~\cite{Miller_WSSSPE}
+
+\item Karl W. Broman, Thirteen years of R/qtl: Just barely sustainable~\cite{Broman_WSSSPE}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item Eric G Stephan, Todd O Elsethagen, Kerstin Kleese van Dam, Laura
+  Riihimaki. What Comes First, the OWL or the Bean?~\cite{Stephan_WSSSPE}
+
+\item Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs,
+  Andrew E. Slaughter, Jason M. Miller, Continuous Integration for
+  Concurrent Computational Framework and Application Development~\cite{Gaston_WSSSPE}
+
+\item Anshu Dubey, B. Van Straalen. Experiences from Software
+  Engineering of Large Scale AMR Multiphysics Code Frameworks~\cite{Dubey_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Michael Crusoe, C. Titus Brown. Walking the talk: adopting and
+  adapting sustainable scientific software development processes in a
+  small biology lab~\cite{Crusoe_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava Majumdar.
+The MVAPICH Project: Evolution and Sustainability of an Open Source
+Production Quality MPI Library for HPC~\cite{Panda_WSSSPE}
+
+\item Eric M. Heien, Todd L. Miller, Becky Gietzel, Louise
+  H. Kellogg. Experiences with Automated Build and Test for
+  Geodynamics Simulation Codes~\cite{Heien_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Deployment, Support, and Maintenance of Existing Software}
+
+\begin{itemize}
+
+\item Henri Casanova, Arnaud Giersch, Arnaud Legrand, Martin Quinson,
+  Fr\'{e}d\'{e}ric Suter. SimGrid: a Sustained Effort for the
+  Versatile Simulation of Large Scale Distributed
+  Systems~\cite{Casanova_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\item Jeremy Cohen, Chris Cantwell, Neil Chue Hong, David Moxey,
+  Malcolm Illingworth, Andrew Turner, John Darlington, Spencer
+  Sherwin. Simplifying the Development, Use and Sustainability of HPC
+  Software~\cite{Cohen_WSSSPE}
+
+\item Jaroslaw Slawinski, Vaidy Sunderam. Towards Semi-Automatic
+  Deployment of Scientific and Engineering Applications~\cite{Slawinski_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Best Practices, Challenges, and Recommendations}
+
+\begin{itemize}
+
+\item Andreas Prli\'{c}, James B. Procter. Ten Simple Rules for the
+  Open Development of Scientific Software~\cite{Prlic_WSSSPE}
+
+\item Anshu Dubey, S. Brandt, R. Brower, M. Giles, P. Hovland,
+  D. Q. Lamb, F. Löffler, B. Norris, B. O'Shea, C. Rebbi, M. Snir,
+  R. Thakur, Software Abstractions and Methodologies for HPC
+  Simulation Codes on Future Architectures~\cite{Dubey2_WSSSPE}
+
+\item Jeffrey Carver, George K. Thiruvathukal. Software Engineering
+  Need Not Be Difficult~\cite{Carver_WSSSPE}
+
+\item Craig A. Stewart, Julie Wernert, Eric A. Wernert, William
+  K. Barnett, Von Welch. Initial Findings from a Study of Best
+  Practices and Models for Cyberinfrastructure Software Sustainability~\cite{Stewart_WSSSPE}
+
+\item Jed Brown, Matthew Knepley, Barry Smith. Run-time extensibility:
+  anything less is unsustainable~\cite{Brown_WSSSPE}
+
+\item Shel Swenson, Yogesh Simmhan, Viktor Prasanna, Manish Parashar,
+  Jason Riedy, David Bader, Richard Vuduc. Sustainable Software
+  Development for Next-Gen Sequencing (NGS) Bioinformatics on Emerging
+  Platforms~\cite{Swenson_WSSSPE}
+
+\end{itemize}
+
+\section*{Policy}
+
+\subsubsection*{Modeling Sustainability}
+
+\begin{itemize}
+
+\item Coral Calero, M. Angeles Moraga, Manuel F. Bertoa. Towards a
+  Software Product Sustainability Model~\cite{Calero_WSSSPE}
+
+\item Colin C. Venters, Lydia Lau, Michael K. Griffiths, Violeta
+  Holmes, Rupert R. Ward, Jie Xu. The Blind Men and the Elephant:
+  Towards a Software Sustainability Architectural Evaluation Framework~\cite{Venters_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Chris Mattmann. Sustainable
+  Cyberinfrastructure Software Through Open Governance~\cite{Pierce_WSSSPE}
+
+\item Daniel S. Katz, David Proctor. A Framework for Discussing
+  e-Research Infrastructure Sustainability~\cite{Katz_WSSSPE}
+
+\item Christopher Lenhardt, Stanley Ahalt, Brian Blanton, Laura
+  Christopherson, Ray Idaszak. Data Management Lifecycle and Software
+  Lifecycle Management in the Context of Conducting Science~\cite{Lenhardt_WSSSPE}
+
+\item Nicholas Weber, Andrea Thomer, Michael Twidale. Niche Modeling:
+  Ecological Metaphors for Sustainable Software in Science~\cite{Weber_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Credit, Citation, Impact}
+
+\begin{itemize}
+
+\item Matthew Knepley, Jed Brown, Lois Curfman McInnes, Barry
+  Smith. Accurately Citing Software and Algorithms Used in
+  Publications~\cite{Knepley_WSSSPE}
+
+\item Jason Priem, Heather Piwowar. Toward a comprehensive impact
+  report for every software project~\cite{Priem_WSSSPE}
+
+\item Daniel S. Katz. Citation and Attribution of Digital Products:
+  Social and Technological Concerns~\cite{Katz2_WSSSPE}
+
+\item Neil Chue Hong, Brian Hole, Samuel Moore. Software Papers:
+  improving the reusability and sustainability of scientific software~\cite{Chue_Hong_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation Software~\cite{Loffler_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Reproducibility}
+
+\begin{itemize}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Victoria Stodden, Sheila Miguez. Best Practices for
+  Computational Science: Software Infrastructure and Environments for
+  Reproducible and Extensible Research~\cite{Stodden_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Implementing Policy}
+
+\begin{itemize}
+
+\item Randy Heiland, Betsy Thomas, Von Welch, Craig Jackson. Toward a
+  Research Software Security Maturity Model~\cite{Heiland_WSSSPE}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Daisie Huang, Hilmar Lapp. Software Engineering as
+  Instrumentation for the Long Tail of Scientific Software~\cite{Huang_WSSSPE}
+
+\item Chandra Krintz, Hiranya Jayathilaka, Stratos
+  Dimopoulos, Alexander Pucher, Rich Wolski. Developing Systems for API Governance~\cite{Krintz_WSSSPE}
+
+\end{itemize}
+
+\section*{Communities, Models, and Education}
+
+\subsubsection*{Communities}
+
+\begin{itemize}
+
+\item Reagan Moore. Extensible Generic Data Management Software~\cite{Moore_WSSSPE}
+
+\item Karen Cranston, Todd Vision, Brian O'Meara, Hilmar Lapp. A
+  grassroots approach to software sustainability~\cite{Cranston_WSSSPE}
+
+\item J.-L. Vay, C. G. R. Geddes, A. Koniges, A. Friedman,
+  D. P. Grote, D. L. Bruhwiler. White Paper on DOE-HEP Accelerator
+  Modeling Science Activities~\cite{Vay_WSSSPE}
+
+\item Ketan Maheshwari, David Kelly, Scott J. Krieder, Justin
+  M. Wozniak, Daniel S. Katz, Mei Zhi-Gang, Mainak
+  Mookherjee. Reusability in Science: From Initial User Engagement to
+  Dissemination of Results~\cite{Maheshwari_WSSSPE}
+
+\item Edmund Hart, Carl Boettiger, Karthik Ram, Scott
+  Chamberlain. rOpenSci -- a collaborative effort to develop R-based
+  tools for facilitating Open Science~\cite{Hart_WSSSPE}
+
+\item L. Christopherson, R. Idaszak, S. Ahalt. Developing Scientific
+  Software through the Open Community Engagement
+  Process~\cite{Christopherson_WSSSPE}
+
+\item Marlon Pierce, Suresh Marru, Mark A. Miller, Amit Majumdar,
+  Borries Demeler. Science Gateway Operational Sustainability:
+  Adopting a Platform-as-a-Service Approach~\cite{Pierce2_WSSSPE}
+
+\item Lynn Zentner, Michael Zentner, Victoria Farnsworth, Michael
+  McLennan, Krishna Madhavan, and Gerhard Klimeck, nanoHUB.org:
+  Experiences and Challenges in Software Sustainability for a Large
+  Scientific Community~\cite{Zentner_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software
+  Community~\cite{Terrel_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation
+  Software~\cite{Loffler_WSSSPE}
+
+\item Nancy Wilkins-Diehr, Katherine Lawrence, Linda Hayden, Marlon
+  Pierce, Suresh Marru, Michael McLennan, Michael Zentner, Rion
+  Dooley, Dan Stanzione. Science Gateways and the Importance of
+  Sustainability~\cite{Wilkins-Diehr_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\end{itemize}
+
+In addition, the following paper from another area was also discussed
+in this area.
+
+\begin{itemize}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Industry \& Economic Models}
+
+\begin{itemize}
+
+\item Anne C. Elster. Software for Science: Some Personal
+  Reflections on Funding, Licensing, Publishing and Teaching~\cite{Elster_WSSSPE}
+
+\item Ian Foster, Vas Vasiliadis, Steven Tuecke. Software as a Service
+  as a path to software sustainability~\cite{Foster_WSSSPE}
+
+\item Marcus Hanwell, Amitha Perera, Wes Turner, Patrick O'Leary,
+  Katie Osterdahl, Bill Hoffman, Will Schroeder. Sustainable Software
+  Ecosystems for Open Science~\cite{Hanwell_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Brian Blanton, Chris Lenhardt, A User Perspective on Sustainable
+  Scientific Software~\cite{Blanton_WSSSPE}
+
+\item Markus Blatt. DUNE as an Example of Sustainable Open Source
+  Scientific Software Development~\cite{Blatt_WSSSPE}
+
+\item Dhabaleswar K. Panda, Karen Tomko, Karl Schulz, Amitava
+  Majumdar. The MVAPICH Project: Evolution and Sustainability of an
+  Open Source Production Quality MPI Library for
+  HPC~\cite{Panda_WSSSPE}
+
+\item Andy Terrel. Sustaining the Python Scientific Software
+  Community~\cite{Terrel_WSSSPE}
+
+\end{itemize}
+
+\subsubsection*{Education \& Training}
+
+\begin{itemize}
+
+\item Ivan Girotto, Axel Kohlmeyer, David Grellscheid, Shawn
+  T. Brown. Advanced Techniques for Scientific Programming and
+  Collaborative Development of Open Source Software Packages at the
+  International Centre for Theoretical Physics (ICTP)~\cite{Girotto_WSSSPE}
+
+\item Thomas Crawford. On the Development of Sustainable Software for
+  Computational Chemistry~\cite{Crawford_WSSSPE}
+
+\end{itemize}
+
+In addition, the following papers from other areas were also discussed
+in this area.
+
+\begin{itemize}
+
+\item Charles R. Ferenbaugh, Experiments in Sustainable Software
+  Practices for Future Architectures~\cite{Ferenbaugh_WSSSPE}
+
+\item David Koop, Juliana Freiere, Cl\'{a}udio T. Silva, Enabling
+  Reproducible Science with VisTrails~\cite{Koop_WSSSPE}
+
+\item Sean Ahern, Eric Brugger, Brad Whitlock, Jeremy S. Meredith,
+  Kathleen Biagas, Mark C. Miller, Hank Childs, VisIt: Experiences
+  with Sustainable Software~\cite{Ahern_WSSSPE}
+
+\item Sou-Cheng (Terrya) Choi. MINRES-QLP Pack and Reliable
+  Reproducible Research via Staunch Scientific Software~\cite{Choi_WSSSPE}
+
+\item Frank L\"{o}ffler, Steven R. Brandt, Gabrielle Allen and Erik
+  Schnetter. Cactus: Issues for Sustainable Simulation
+  Software~\cite{Loffler_WSSSPE}
+
+\item Erik Trainer, Chalalai Chaihirunkarn, James Herbsleb. The Big
+  Effects of Short-term Efforts: A Catalyst for Community Engagement
+  in Scientific Software~\cite{Trainer_WSSSPE}
+
+\end{itemize}
+
 \section{Attendees} \label{sec:attendees}  %\note{lead: Shel Swenson}

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
Section 8: Fiddling.
changes to 5

@@
 
-Table~\ref{tab:defining-sustainability} reproduces a summary created
-by C.~Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each
+Table~\ref{tab:defining-sustainability} is based on a summary 
+by Lenhardt ~\cite{lenhardt-wssspe1-panel}, which shows for each
 contribution to the Modeling Sustainability panel what it meant when
@@
 describing two modes: ``top-down'' and ``bottom-up'' governance.  They
-place governance alongside questions of the artifact (``technical''),
-and questions of the origins of resources, in terms of who is funding
-the work surrounding a piece of software (``the political'') and the
-manner in which resources come to the project initially (commercial,
+place governance alongside technical questions about the software,
+political questions about who is funding
+the work surrounding the software, and the
+manner in which resources come to the project both initially (commercial,
 open source, closed partnerships, grant funded) and long-term (all
@@
 such as test-driven development and socialized bug managing practices
-via \toolname{trello}. Galaxy \emph{histories} and \emph{toolshed}
-offers easy community sharing of data and tools further promoting a
+via \toolname{trello}. \toolname{Galaxy} \emph{histories} and \emph{toolshed}
+offer easy community sharing of data and tools further promoting a
 collaborative environment. The project closely follows the

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
read and just one minor change
editing of the end of 8

@@
 Most countries place some limits on software patents. The European 
-Union outright forbids them, while US patent law excludes ``abstract ideas,'' 
+Union outright forbids them, while US patent law excludes ``abstract ideas'', 
 which has been used to refuse some software patents.
@@
 
-Finally, it appears that the career options of developers from
-two kinds of software projects differ significantly. In order to
-further investigate the much discussed issue of career paths for
-developers of these software, we conducted a small survey of the cases
-discussed above with respect to the current status of its core developers. We
-looked at the workshop submissions and the original cited papers for the
-respective software. Where available, we looked at the web pages and source
-code of the earliest versions to find the original developers. We included
-projects older than 5 years and excluded relatively new projects such as
-\toolname{SciPy} and \toolname{numFOCUS} from this survey. We find that in a
-majority of scientific software, more than one original developer are still
-around, in some cases after more than 10 years (e.g., \toolname{R/qtl}).
-Whereas, in the case of utility software projects, it appears that at least one
-original developer is still around, often in a part-time or ``benefactor''
-role. In other cases, it appears that the original developers have went on to
-assume a larger role (eg. manager) within the same project.
+%Finally, it appears that the career options of developers from
+%two kinds of software projects differ significantly. 
+%Examining the workshop submissions for 
+%projects older than 5 years (excluding relatively new projects such as
+%\toolname{SciPy} and \toolname{numFOCUS}) and the original cited papers for the
+%respective software, and, where available, the web pages and source
+%code of the earliest versions to find the original developers, it appears
+%that in a majority of scientific software, more than of the one original developer are still
+%associated with the project, in some cases after more than 10 years later (e.g., \toolname{R/qtl}).
+%For the utility software projects, it appears that at least one
+%original developer is still
+%associated with the project, often in a part-time or ``benefactor''
+%role, or in a larger role (e.g., manager).
 
 In conclusion, we find that there are distinctions in the
-characteristics and the kinds of challenges faced between utility and
-scientific software projects.  In terms of career paths of original
-developers, it appears that the career choices for utility software
-developers are broader compared to the developers of scientific
-software. However, it appears that in the case of scientific software,
-core competencies of original developers tend to be domain science who
-happen to be the coders of their software. As projects as a whole, the
-small-scale, specific software have a lower risk and cost in terms of
-sustaining as compared to utility software. We also find that the
-utility software tend to adhere more loyally to the many best
-practices discussed during the workshop. Often the sustainability of
+characteristics and challenges faced between utility and
+scientific software projects.  
+%In terms of career paths of original
+%developers, it appears that the career choices for utility software
+%developers are broader compared to the developers of scientific
+%software. However, it appears that in the case of scientific software,
+%core competencies of original developers tend to be domain science who
+%happen to be the coders of their software. 
+%As projects as a whole, the
+%The
+%small-scale, specific software have a lower risk and cost in terms of
+%sustaining as compared to utility software. 
+We %also 
+find that the
+utility software packages are more likely to use the best
+practices discussed during the workshop. Often, sustainability of
 scientific software projects is achieved by the fact that the core
-developer or team heavily utilizes the software for their science,
+developer or team heavily utilizes the software for their own science,
 e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
-scientific software requires more scientific skills compared with
-those of utility software. This diminishes the need of specialist
-programmers and practices in many cases where the bulk of development
-is carried out by a domain scientist. With these inferences, we
-believe that separate guidelines and principles for sustainability are to
-be defined and practiced for each category of scientific software and
-they be given separate treatment.
+scientific software requires more scientific background compared with
+utility software, thus in many cases, the bulk of development
+is carried out by a domain scientist. For these reasons, we
+believe that separate guidelines and sustainability principles could
+be defined for these two software categories.
 

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing two mixed up affiliations

@@
 %
-\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA; \url{gdallen@illinois.edu}}
-%
-\thanks{{}$^{(12)}$ University of Southern California, Los Angeles, CA, USA; \url{mdswenso@usc.edu}}
+\thanks{{}$^{(11)}$ University of Southern California, Los Angeles, CA, USA; \url{mdswenso@usc.edu}}
+%
+\thanks{{}$^{(12)}$ University of Illinois, Champaign, IL, USA; \url{gdallen@illinois.edu}}
 %

### 2014-04-27 WSSSPE1_report/WSSSPE1_summary_report.tex
small changes in various places - will submit this morning

@@
 The remainder of this paper first gives an overview of the process
-with which the workshop was put together (\S\ref{sec:process}), then
+with which the workshop was organized (\S\ref{sec:process}), then
 proceeds with summaries of the two keynote
@@
 categories, namely \emph{Developing and Supporting Software},
-\emph{Policy}, and \emph{Communities}. Each category was assigned a
+\emph{Policy}, and \emph{Communities}. Each category was assigned to a
 panel, with three to four panelists drawn from authors of the
@@
 studies to better understand the points discussed during the workshop
-and described in the previous sections. We dive deeper to find how 
-they are affected by the sustainability related issues in
-practice. Additionally, we discuss issues surrounding people,
-especially the career paths of original developers and investigate to
-answer the question: \emph{are the original developers still around?}
+and described in the previous sections, and to find how 
+they are affected by sustainability issues in
+practice. 
+%Additionally, we discuss issues surrounding people,
+%especially the career paths of original developers and investigate to
+%answer the question: \emph{are the original developers still around?}
 
@@
 problem.
-
+%
 % Why are we classifying software like above?
@@
 %role, or in a larger role (e.g., manager).
+
+\subsection{Distinctions}
 
@@
 
-\section*{Developing and Supporting Software}
+\subsection*{Developing and Supporting Software}
 
@@
 
-\section*{Policy}
+\subsection*{Policy}
 
@@
 
-\section*{Communities, Models, and Education}
+\subsection*{Communities, Models, and Education}
 

### 2014-04-28 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing a bad cite

@@
 ``activities that facilitate science but are not currently rewarded or
-recognized''~\cite{Katz_WSSSPE}. Priem and Piwowar argued for the need
+recognized''~\cite{Katz2_WSSSPE}. Priem and Piwowar argued for the need
 to ``support all researchers in presenting meaningful impact evidence

### 2014-04-28 WSSSPE1_report/WSSSPE1_summary_report.tex
2 more small fixes

@@
 mechanism such as a Digital Object
-Identifier~\cite{Katz_WSSSPE,Knepley_WSSSPE} or via a published paper
+Identifier~\cite{Katz2_WSSSPE,Knepley_WSSSPE} or via a published paper
 written to document the software (and perhaps its creation), a
@@
 recursively allocate that credit to those who had provided
-dependencies. Finally Priem and Piwowar argued that machine learning
+dependencies~\cite{Katz2_WSSSPE}. Finally Priem and Piwowar argued that machine learning
 techniques could examine the body of published literature and extract

### 2014-06-03 WSSSPE1_report/WSSSPE1_summary_report.tex
partial changes to reviewer comments

@@
 % Anne: Added a lead-in motivator to address the submission as a Experience paper
-% and tried to improve writing... ( I am assumeing no abstract word count limit...)
+% and tried to improve writing... ( I am assuming no abstract word count limit...)
 Challenges related to development, deployment, and maintenance of 
@@
 like GitHub. For truly sustainable software, there should be no
-endpoint, as the software products continue to be used beyond a single
+end, as the software products will continue to be used beyond a single
 institution, grant, and developer or development team.
@@
 
-There was debate that while some of the basics of computing may be
-similar, tools commonly used for computing have changed quite
+It appears that while some of the basics of computing may be
+similar, the tools commonly used for computing have changed quite
 significantly. Reference was made to Perl, which was commonly used,
@@
 software projects can better monitor the literature to improve their
-ability to show impact. For example, the NanoHub project scans the
+ability to show impact. For example, the nanoHUB project scans the
 literature using keywords and names of known users to discover papers
@@
 of purpose, leverage the work of others, and plan for flexibility. 3)
-Projects should identify one ore more existing communities and understand the
+Projects should identify one or more existing communities and understand the
 communities' needs before beginning, then adapt as the communities'
-needs evolve. 4) Funders, should consider the technology project
+needs evolve. 4) Funders should consider the technology project
 lifecycle, and design solicitations to reward effective planning,
@@
 generic computing frameworks such as \toolname{Galaxy},
-\toolname{nanoHUB}, \toolname{SimGrid}, \toolname{Swift},
+\toolname{HUBzero}, \toolname{SimGrid}, \toolname{Swift},
 \toolname{Globus} and \toolname{VisTrails}, and visualization

### 2014-06-03 WSSSPE1_report/WSSSPE1_summary_report.tex
removing sentence in abstract

@@
 documentation, development and deployment practices, software
-licenses and career paths for developers. Attribution systems that accounts for 
+licenses and career paths for developers. Attribution systems that account for 
 evidence of software contribution and impact were also discussed. These include
@@
 of ``software papers'', and the use of online systems, for example source code repositories
-like GitHub. For truly sustainable software, there should be no
-end, as the software products will continue to be used beyond a single
-institution, grant, and developer or development team.
+like GitHub. %For truly sustainable software, there should be no
+%end, as the software products will continue to be used beyond a single
+%institution, grant, and developer or development team.
 

### 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
added a sentence about age of software projects and added launch years for some of the discussed software

@@
 %answer the question: \emph{are the original developers still around?}
+%
+In terms of age, most of the software discussed here were publicly launched 10
+or more years ago. To get a sense of their age, we note release year in the
+parenthesis at first mention of the software where available.
 
@@
 % Why are we classifying software like above?
-This classification is motivated by our argument that the two kinds of
-software projects wildly vary in factors such as scope, purpose and
-usage. The development and management of each kind is significantly
-different. Consequently, the sustainability challenges faced by them
-differ and must be treated separately. For instance, the challenges
-faced by a gateway software development project such as
-\toolname{CIPRES} or visualization software products such as
-\toolname{VisIT} or \toolname{ParaView} are distinct to a niche software for
-\textit{ab initio} modeling and simulation such as \toolname{VASP} or
-\toolname{Quantum Espresso}. %\note{cite SW}
+This classification is motivated by our argument that the two kinds of software
+projects wildly vary in factors such as scope, purpose and usage. The
+development and management of each kind is significantly different.
+Consequently, the sustainability challenges faced by them differ and must be
+treated separately. For instance, the challenges faced by a gateway software
+development project such as \toolname{CIPRES (r2007)} or visualization software
+products such as \toolname{VisIT (r2001)} or \toolname{ParaView (r2002)} are
+distinct to a niche software for \textit{ab initio} modeling and simulation
+such as \toolname{VASP (r1992)} or \toolname{Quantum Espresso}. %\note{cite SW}
 
 \subsection{Utility Software}
-Software developed with a potentially wider audience and general
-purpose usage in mind is utility software. Utility software typically
-does not address fundamental research problems for a given scientific
-domain. Examples are collaborative development frameworks such as
-\toolname{GitHub} and \toolname{Bitbucket}, distributed workflow and
-generic computing frameworks such as \toolname{Galaxy},
-\toolname{HUBzero}, \toolname{SimGrid}, \toolname{Swift},
-\toolname{Globus} and \toolname{VisTrails}, and visualization
-frameworks such as \toolname{VTK}, \toolname{VisIT}, and \toolname{ParaView}.
+Software developed with a potentially wider audience and general purpose usage
+in mind is utility software. Utility software typically does not address
+fundamental research problems for a given scientific domain. Examples are
+collaborative development frameworks such as \toolname{GitHub (r2008)} and
+\toolname{Bitbucket (r2008)}, distributed workflow and generic computing
+frameworks such as \toolname{Galaxy (r2006)}, \toolname{HUBzero},
+\toolname{SimGrid}, \toolname{Swift (r2007)}, \toolname{Globus (r2000)} and
+\toolname{VisTrails}, and visualization frameworks such as \toolname{VTK},
+\toolname{VisIT}, and \toolname{ParaView}.
 
@@
 
-One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It
-follows agile software development practices and implements standards
-such as test-driven development and socialized bug managing practices
-via \toolname{trello}. \toolname{Galaxy} \emph{histories} and \emph{toolshed}
-offer easy community sharing of data and tools further promoting a
-collaborative environment. The project closely follows the
-guidelines described in Carver and Thiruvathukal~\cite{Carver_WSSSPE}
-and many from Prli\'{c} and Procter~\cite{Prlic_WSSSPE}. Many utility
-software projects are often developed aiming better utilizing a
-particular, new infrastructure and architecture, e.g.,
-\toolname{MVAPICH}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to leverage the power
-of emerging architectures such as accelerators, new code and libraries
-are required. The experience of one such effort as described in
-Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success
-but nonetheless with many invaluable lessons were learned about
-influential cultural and technical aspects in sustainable software
-development practices.
+One such example is the \toolname{Galaxy} project~\cite{Galaxy}. It follows
+agile software development practices and implements standards such as
+test-driven development and socialized bug managing practices via
+\toolname{trello}. \toolname{Galaxy} \emph{histories} and \emph{toolshed} offer
+easy community sharing of data and tools further promoting a collaborative
+environment. The project closely follows the guidelines described in Carver and
+Thiruvathukal~\cite{Carver_WSSSPE} and many from Prli\'{c} and
+Procter~\cite{Prlic_WSSSPE}. Many utility software projects are often developed
+aiming better utilizing a particular, new infrastructure and architecture,
+e.g., \toolname{MVAPICH (r2002)}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to
+leverage the power of emerging architectures such as accelerators, new code and
+libraries are required. The experience of one such effort as described in
+Ferenbaugh~\cite{Ferenbaugh_WSSSPE} which met with a limited success but
+nonetheless with many invaluable lessons were learned about influential
+cultural and technical aspects in sustainable software development practices.
 
@@
 Some examples of such software discussed as part of the workshop are,
-\toolname{DUNE}, \toolname{R/qtl}, \toolname{Kitware},
-\toolname{PETSc}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
+\toolname{DUNE (r2008)}, \toolname{R/qtl (r2002)}, \toolname{Kitware (r1998)},
+\toolname{PETSc (r1995)}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
 the above efforts are focused on one scientific or applied

### 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
edits in project years and style for showing them

@@
 %
-In terms of age, most of the software discussed here were publicly launched 10
-or more years ago. To get a sense of their age, we note release year in the
-parenthesis at first mention of the software where available.
+Most of the software projects discussed here were publicly launched 10
+or more years ago. We generally note the original release (o.r.) year of each project in 
+parenthesis in its first mention.
 
@@
 treated separately. For instance, the challenges faced by a gateway software
-development project such as \toolname{CIPRES (r2007)} or visualization software
-products such as \toolname{VisIT (r2001)} or \toolname{ParaView (r2002)} are
+development project such as \toolname{CIPRES} (o.r. 2007) or visualization software
+products such as \toolname{VisIT (o.r. 2001)} or \toolname{ParaView} (o.r.  2002) are
 distinct to a niche software for \textit{ab initio} modeling and simulation
-such as \toolname{VASP (r1992)} or \toolname{Quantum Espresso}. %\note{cite SW}
+such as \toolname{VASP} (o.r. 1992) or \toolname{Quantum Espresso} (o.r. 2001). %\note{cite SW}
 
@@
 fundamental research problems for a given scientific domain. Examples are
-collaborative development frameworks such as \toolname{GitHub (r2008)} and
-\toolname{Bitbucket (r2008)}, distributed workflow and generic computing
-frameworks such as \toolname{Galaxy (r2006)}, \toolname{HUBzero},
-\toolname{SimGrid}, \toolname{Swift (r2007)}, \toolname{Globus (r2000)} and
-\toolname{VisTrails}, and visualization frameworks such as \toolname{VTK},
+collaborative development frameworks such as \toolname{GitHub} (o.r. 2008) and
+\toolname{Bitbucket} (o.r.  2008), distributed workflow and generic computing
+frameworks such as \toolname{Galaxy} (o.r. 2006), \toolname{HUBzero} (o.r. 2010),
+\toolname{SimGrid} (o.r. 2001), \toolname{Swift} (o.r. 2007), \toolname{Globus} (o.r. 2000) and
+\toolname{VisTrails} (o.r. 2007), and visualization frameworks such as \toolname{VTK},
 \toolname{VisIT}, and \toolname{ParaView}.
@@
 aiming better utilizing a particular, new infrastructure and architecture,
-e.g., \toolname{MVAPICH (r2002)}, \toolname{VisIT}, \toolname{ParaView}. Similarly, to
+e.g., \toolname{MVAPICH} (original release 2002), \toolname{VisIT}, \toolname{ParaView}. Similarly, to
 leverage the power of emerging architectures such as accelerators, new code and
@@
 Some examples of such software discussed as part of the workshop are,
-\toolname{DUNE (r2008)}, \toolname{R/qtl (r2002)}, \toolname{Kitware (r1998)},
-\toolname{PETSc (r1995)}, \toolname{MINRES-QLP}, \toolname{FASTMath}.  Most of
+\toolname{DUNE (o.r. 2008)}, \toolname{R/qtl} (o.r. 2002), \toolname{Kitware} (o.r. 1998),
+\toolname{PETSc} (o.r.
+1995), \toolname{MINRES-QLP} (o.r.
+2007), \toolname{FASTMath} (o.r.
+\textcolor{red}{TBD - Is FASTMath really a software project, or a collection of different packages?}).
+Most of
 the above efforts are focused on one scientific or applied

### 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
style error corrected

@@
 aiming better utilizing a particular, new infrastructure and architecture,
-e.g., \toolname{MVAPICH} (original release 2002), \toolname{VisIT}, \toolname{ParaView}. Similarly, to
+e.g., \toolname{MVAPICH} (o.r. 2002), \toolname{VisIT}, \toolname{ParaView}. Similarly, to
 leverage the power of emerging architectures such as accelerators, new code and

### 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
FastMATH is a packaging and configuring system for Math libs for HPC, removing it for now, looking to include it in a different place
fixing grammar

@@
 
-Some examples of such software discussed as part of the workshop are,
+Some examples of such software discussed as part of the workshop are
 \toolname{DUNE (o.r. 2008)}, \toolname{R/qtl} (o.r. 2002), \toolname{Kitware} (o.r. 1998),
-\toolname{PETSc} (o.r.
-1995), \toolname{MINRES-QLP} (o.r.
-2007), \toolname{FASTMath} (o.r.
-\textcolor{red}{TBD - Is FASTMath really a software project, or a collection of different packages?}).
-Most of
-the above efforts are focused on one scientific or applied
+\toolname{PETSc} (o.r. 1995), and \toolname{MINRES-QLP} (o.r. 2007),
+%\toolname{FASTMath} (o.r. \textcolor{red}{TBD - Is FASTMath really a software project, or a collection of different packages?}).
+most of which are focused on one scientific or applied
 mathematics domain. However, sometimes such projects grow beyond the

### 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
adding information about submission regions and domain/discipline in section 2

@@
 self-archived in the same way as contributed papers.
+
+The workshop received submissions from eight countries across the North
+American and European regions. In some instances authors collaborated across
+the US and European regions contributing jointly authored papers. A majority of
+contributions came from the US with 42 papers where at least one author was
+affiliated to a US institution. A total of 10 submissions were from the
+European region and 4 were from Canada.
+
+In terms of subject of the papers, the submissions were dominated by the domain
+of practice of sustainable software engineering and management with about 32
+papers based on these themes. These papers were further based on a variety of
+disciplines including infrastructure and architecture, user engagement, and
+governance. Additionally, 18 papers were based on the pure sciences and
+applied mathematics domains with disciplines including High Energy Physics,
+Bioinformatics, Nanotechnology, Chemistry, and material sciences. Others were
+dominated by miscellaneous topics such as science gateways and visualization.
 

### 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
minor wording changes in next text in S2

--- 2014-06-04 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 self-archived in the same way as contributed papers.
+
+The workshop received submissions from eight North
+American and European countries. In some instances authors collaborated across
+multiple countries towards jointly authored papers. A majority of
+contributions came from the US with 42 papers where at least one author was
+affiliated with a US institution. A total of 10 submissions were from 
+Europe and 4 were from Canada.  This is not surprising for a workshop
+being held in the US.  We believe future versions of the workshop will
+have contributions from more countries and more continents.
+
+In terms of subject of the papers, the submissions were dominated by the domain
+of practice of sustainable software engineering and management with about 32
+papers based on these themes. These papers were further based on a variety of
+disciplines including infrastructure and architecture, user engagement, and
+governance. Additionally, 18 papers were based on the sciences and
+applied mathematics domains with disciplines including High Energy Physics,
+Bioinformatics, Nanotechnology, Chemistry, and material sciences. Others were
+included topics such as science gateways and visualization.  Again, given that
+this workshop was held with a computer and computational science conference,
+these numbers are not surprising.
 

### 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
Added some references to relevant papers discussing training
I think these papers sufficiently elaborate on the points made here.

@@
 \subsection{Training Scientists to Develop and Support Software}
+
 Part of the panel discussion focused around community structures and
@@
 disciplinary depth as well as interdisciplinary and collaboration
-skills are important.  The discussion focused on whether we could
+skills are important~\cite{Trainer_WSSSPE, Dubey_WSSSPE, Girotto_WSSSPE}.
+The discussion focused on whether we could
 teach software developers and domain scientists to collaborate
@@
 community with a spectrum of expertise across domain science and
-software development rather than two separate communities.
+software development, rather than two separate communities~\cite{Prlic_WSSSPE}.
 
@@
 such as virtual organizations, and tool development across software
-development communities. Further, how should these skills and best
-practices then be introduced to students?
+development communities~\cite{Christopherson_WSSSPE, Cranston_WSSSPE}.
+Further, how should these skills and best practices then be introduced to students?
 

### 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
Made it clearer that this refers to the earl years of development
Review queried tapes with regards to GCC, Emacs etc as they are still
developed today. Made the sentence more complete, making it clear that
this simply refers to their early development (not the present day,
where they also benefit from immediate sharing).

@@
 tendency along with the enormous differences in the speed and ease of
-sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
+sharing---having to ship tapes around in the early days of software development
+(shapnig development of GCC, Emacs, in their formative years) as opposed to
 the immediate sharing of the latest development online, using revision

### 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
fixing spelling, small change

--- 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 tendency along with the enormous differences in the speed and ease of
-sharing---having to ship tapes around (GCC, Emacs, etc) as opposed to
+sharing---having to ship tapes around in the early days of software development
+(which shaped development of GCC and Emacs in their formative years) as opposed to
 the immediate sharing of the latest development online, using revision

### 2014-06-10 WSSSPE1_report/WSSSPE1_summary_report.tex
Added a paragraph to address optional improvement
Added an additional paragraph discussing long term projects that
were discussed in the workshop, and explore what that means. Tried
to keep to projects where it was obvvious they have been active for
over a decade - sorry if I missed any!

@@
 license on code.''
+
+It should be noted that several articles that discussed long-term projects, that
+could be said to have reached a sustainable period. The Visualization Toolkit
+(VTK) was described~\cite{Hanwell_WSSSPE}, being one of the oldest projects
+serving as a basis for several other tools such as ParaView~\cite{Hanwell_WSSSPE}, VisIt~\cite{Ahern_WSSSPE}, and VisTrails~\cite{Koop_WSSSPE}. Other examples of
+long-term, sustainable projects included MVAPICH~\cite{Panda_WSSSPE} which began
+development in 2000, R/qtl~\cite{Broman_WSSSPE} which also began development
+in 2000, and DUNE~\cite{Blatt_WSSSPE} which is also over a decade old. In
+addition to how long a project has been active other metrics are important,
+such as number of developers, number of institutions, and whether there are
+active developers acting as advocates for the continued viability of a project
+beyond individual projects and/or institutions.
 

### 2014-06-10 WSSSPE1_report/WSSSPE1_summary_report.tex
small edits to new text

--- 2014-06-09 WSSSPE1_report/WSSSPE1_summary_report.tex
@@
 license on code.''
+
+It should be noted that several articles that discussed long-term projects, that
+could be said to have reached a sustainable period. The Visualization Toolkit
+(VTK) was described~\cite{Hanwell_WSSSPE} as being one of the oldest projects
+serving as a basis for several other tools such as ParaView~\cite{Hanwell_WSSSPE}, VisIt~\cite{Ahern_WSSSPE}, and VisTrails~\cite{Koop_WSSSPE}. Other examples of
+long-term, sustainable projects included MVAPICH~\cite{Panda_WSSSPE} and R/qtl~\cite{Broman_WSSSPE}, which both began development
+in 2000, and DUNE~\cite{Blatt_WSSSPE}, which is also over a decade old. In
+addition to how long a project has been active, other metrics are important,
+such as number of developers, number of institutions, and whether there are
+active developers acting as advocates for the continued viability of a project
+beyond individual projects and/or institutions.
 

### 2014-10-10 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
adding initial pre-workshop report for WSSSPE2

\documentclass[11pt, oneside]{amsart}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float}
\usepackage{graphicx}
\usepackage[latin9]{inputenc}
\usepackage{multirow}
\usepackage{rotating}
\usepackage{subfigure} 
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}

%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}

\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
\usepackage[all]{hypcap}
\urlstyle{rm}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}

\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
\newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
\newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
\newcommand{\manishnote}[1]{  {\textcolor{violet}     { ***Manish: #1 }}}
\newcommand{\davidnote}[1]{  {\textcolor{darkgreen}      { ***David: #1 }}}
\newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}

% Don't use tt font for urls
\urlstyle{rm}

% 15 characters / 2.5 cm => 100 characters / line
% Using 11 pt => 94 characters / line
\setlength{\paperwidth}{216 mm}
% 6 lines / 2.5 cm => 55 lines / page
% Using 11pt => 48 lines / pages
\setlength{\paperheight}{279 mm}
\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
% You can use a baselinestretch of down to 0.9
\renewcommand{\baselinestretch}{0.96}

\sloppypar

\begin{document}

\title[]{Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review, and Sorting Process, and Results}

\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
Karen Cranston$^{(affil?)}$, Manish Parashar$^{(4)}$, David Proctor$^{(affil?)}$, \\
Matthew Turk$^{(2)}$, Colin Venters$^{(affil?)}$, Nancy Wilkins-Diehr$^{(5)}$}

\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}

\thanks{{}$^{(2)}$ University of Illinois, Champaign, IL, USA}

\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}

\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}

\thanks{{}$^{(5)}$ University of California-San Diego, San Diego, CA, USA}

\begin{abstract}
This technical report discusses the submission and peer-review process used by the Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.  It is intended to record both this alternative model as well as the papers associated with the workshop that resulted from that process.
\end{abstract}

\maketitle

\section{Introduction}

%\katznote{example comment by Dan}
%
%\gabnote{example comment by Gabrielle}
%
%\nchnote{example comment by Neil}
%
%\manishnote{example comment by Manish}
%
%\davidnote{example comment by David}

The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} will be held on Sunday, 16 November 2014, in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
WSSSPE2 follows an initial general workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.

Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is critical to address challenges related to the development, deployment, and maintenance of reusable software as well as education around software practices. These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities). The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas of interest for sustainable scientific software. WSSSPE2 invites the community to propose and discuss specific mechanisms to move towards an imagined future practice of software development and usage in science and engineering. The workshop will include multiple mechanisms for participation, encourage team building around solutions, and identify risky solutions with potentially transformative outcomes. Participation by early career students and postdoctoral researchers is strongly encouraged.

\section{Submissions}

The workshop call for paper included the following areas of interest:
\begin{quote}
\begin{itemize}

\item defining software sustainability in the context of science and engineering software
\begin{itemize}
\item how to evaluate software sustainability
\end{itemize}

\item improving the development process that leads to new software
\begin{itemize}
\item methods to develop sustainable software from the outset
\item effective approaches to reusable software created as a by-product of research
\item impact of computer science research on the development of scientific software
\end{itemize}

\item recommendations for the support and maintenance of existing software
\begin{itemize}
\item software engineering best practices
\item governance, business, and sustainability models
\item the role of community software repositories, their operation and sustainability
\item reproducibility, transparency needs that may be unique to science
\end{itemize}

\item successful open source software implementations
\begin{itemize}
\item incentives for using and contributing to open source software
\item transitioning users into contributing developers
\end{itemize}

\item building large and engaged user communities
\begin{itemize}
\item developing strong advocates
\item measurement of usage and impact
\end{itemize}

\item encouraging industry's role in sustainability
\begin{itemize}
\item engagement of industry with volunteer communities
\item incentives for industry
\item incentives for community to contribute to industry-driven projects
\end{itemize}

\item recommending policy changes
\begin{itemize}
\item software credit, attribution, incentive, and reward
\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
\item mechanisms and venues for publishing software, and the role of publishers
\end{itemize}

\item improving education and training
\begin{itemize}
\item best practices for providing graduate students and postdoctoral researchers in domain communities with sufficient training in software development
\item novel uses of sustainable software in education (K-20)
\item case studies from students on issues around software development in the undergraduate or graduate curricula
\end{itemize}

\item careers and profession
\begin{itemize}
\item successful examples of career paths for developers
\item institutional changes to support sustainable software such as promotion and tenure metrics, job categories, etc.
\end{itemize}

\end{itemize}

\end{quote}

Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, but focusing on papers that are intended to lead to changes, the organizers wanted to make submission as easy as possible.  The call for papers stated:

\begin{quote}
We invite short (4-page) \textbf{actionable} papers that will lead to improvements for sustainable software science. These papers could be a call to action, or could provide position or experience reports on sustainable software activities. The papers will be used by the organizing committee to design sessions that will be highly interactive and targeted towards facilitating action. Submitted papers should be archived by a third-party service that provides DOIs. We encourage submitters to license their papers under a Creative Commons license that encourages sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop.\end{quote}

31 submissions were received, and almost all submitters used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.

\section{Peer-Review}

The review process was fairly standard, where reviewers bid for papers, then an automated system
matched bids to determine assignments, and reviewers then completed their assigned
reviews (with an average of 4.9 reviews per paper, and 4.1 reviews per reviewer)
This process was done through EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide scores on relevance and comments to the organizers, which were used to decide which papers to associate with the workshop, and comments to the authors, which were provided back to the authors to allow them to improve their papers.

The organizers decided to list 28 of the papers as significantly contributing to the workshop, a very high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact that the reports were already self-published.

\todo{discuss grouping - well-sorted... and breakouts at WSSSPE2}

\section{Results}

The contributed papers that will be discussed at the workshop follow, listed by \todo{update when possible - current list is just alphabetic order, not grouped}

\begin{itemize}

\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software Development in Practice ~\cite{wssspe2_adams}

\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software registry~\cite{wssspe2_allen}

\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}

\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}

\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}

\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}

\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}

\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}

\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to scientific software: a hackathon experience~\cite{wssspe2_crusoe}

\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}

\item Yolanda Gil,  Eunyoung Moon and James Howison. No Science Software is an Island: Collaborative Software Development Needs in Geosciences~\cite{wssspe2_gil}

\item ed Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for Sustainable Data and Software~\cite{wssspe2_habermann}

\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating for Science~\cite{wssspe2_hanwell}

\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}

\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski. Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}

\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}

\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}

\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton, S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-??WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}

\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}

\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar. Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}

\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}

\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}

\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}

\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}

\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}

\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}

\item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann, David Andrs and Jason M. Miller. Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}

\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W. M. C. Foulkes, James J. Shepherd and Alex J. W. Thom. The {H}ighly {A}ccurate {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic diagonalisation for quantum chemistry~\cite{wssspe2_spencer}

\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward and David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}

\end{itemize}

\section{Conclusions}

The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can collaboratively build a workshop agenda.  The differences from WSSSPE1 to WSSSPE2 are in using an existing service (EasyChair) to handle submissions and reviews, rather than an ad hoc process, and using an existing service (well-sorted) to allow collaborative grouping of papers into themes by all authors, reviewers, and the community, rather than this being done in an ad hoc manner just by the organizers.

The fact remains that contributors also want to get credit for their participation in the process.  And the workshop organizers still want to make sure that the workshop content and their efforts are recorded.  Ideally, there would be a service that would be able to index the contributions to the workshop, serving the authors, the organizers, and the larger community.  But since there still isn't such a service today, the workshop organizers are writing this initial report and making use of arXiv as a partial solution to provide a record of the workshop.

After the workshop, one or more additional papers will be created that will include the discussions at the workshop.  These papers will likely have many authors, and may be submitted to peer-reviewed journals.

\bibliographystyle{plain}

\bibliography{wssspe}
\end{document}

### 2014-10-10 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
fixing a bunch of typos

@@
 \maketitle
+
+\todo{fix affils}
 
@@
 
-31 submissions were received, and almost all submitters used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+31 submissions were received, and all but one used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
 
@@
 Schildhauer.
-{ISEES-??WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
+{ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
 

### 2014-10-13 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex

@@
 This technical report discusses the submission and peer-review process used by the Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
-It is intended to record both this alternative model as well as the papers associated with the workshop that resulted from that process.
+It is intended to record both, the alternative review model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
 \end{abstract}
@@
 The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} will be held on Sunday, 16 November 2014, in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 follows an initial general workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
+WSSSPE2 follows a more general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
 
-Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is critical to address challenges related to the development, deployment, and maintenance of reusable software as well as education around software practices.
-These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities).
+Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is critical to address challenges related to the development, deployment, maintenance and overall sustainability of reusable software as well as education around software practices.
+These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), software engnieering researchers, and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities).
 The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas of interest for sustainable scientific software.
-WSSSPE2 invites the community to propose and discuss specific mechanisms to move towards an imagined future practice of software development and usage in science and engineering.
+WSSSPE2 invites the community to propose and discuss specific mechanisms to move towards an imagined future practice for software development and usage in science and engineering.
 The workshop will include multiple mechanisms for participation, encourage team building around solutions, and identify risky solutions with potentially transformative outcomes.
@@
 
-The workshop call for paper included the following areas of interest:
+The workshop call for papers included the following areas of interest:
 \begin{quote}
@@
 
-\item encouraging industry's role in sustainability
+\item encouraging industry's role in the sustainability of scientific software
 \begin{itemize}
@@
 The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can collaboratively build a workshop agenda.
-The differences from WSSSPE1 to WSSSPE2 are in using an existing service (EasyChair) to handle submissions and reviews, rather than an ad hoc process, and using an existing service (well-sorted) to allow collaborative grouping of papers into themes by all authors, reviewers, and the community, rather than this being done in an ad hoc manner just by the organizers.
+The differences in WSSSPE2 from WSSSPE1 are in using an existing service (EasyChair) to handle submissions and reviews, rather than an ad hoc process, and using an existing service (well-sorted) to allow collaborative grouping of papers into themes by all authors, reviewers, and the community, rather than this being done in an ad hoc manner by the organizers alone.
 

### 2014-10-13 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
small changes to Manish's changes

@@
 This technical report discusses the submission and peer-review process used by the Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
-It is intended to record both, the alternative review model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
+It is intended to record both the alternative review model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
 \end{abstract}
@@
 The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} will be held on Sunday, 16 November 2014, in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 follows a more general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
+WSSSPE2 follows a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
 
 Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is critical to address challenges related to the development, deployment, maintenance and overall sustainability of reusable software as well as education around software practices.
-These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), software engnieering researchers, and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities).
+These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), software engineering researchers, and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities).
 The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas of interest for sustainable scientific software.
@@
 
-\item encouraging industry's role in the sustainability of scientific software
+\item encouraging industry's role in sustainability
 \begin{itemize}

### 2014-10-13 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
addressing Manish's email comment

@@
 This technical report discusses the submission and peer-review process used by the Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
-It is intended to record both the alternative review model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
+It is intended to record both the alternative submission and program organization model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
 \end{abstract}

### 2014-10-15 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
adding some well-sorted material

@@
 
-\section{Peer-Review}
+\section{Peer-Review and Peer-Grouping}
 
@@
 
-\todo{discuss grouping - well-sorted... and breakouts at WSSSPE2}
+WSSSPE1 was organized into sessions, each of which was aimed at discussing one or more of the themes from the call for papers, with a few paper authors invited to summarize the other papers in that them as a panel, followed by general discussion about that theme.
+The mapping of papers to themes was done by the organizers.
+
+For WSSSPE2, the organizers wanted to increase the interactivity of the sessions, and to open the process of creating the sessions to the full program committee, the paper authors, and others who might attend the workshop.
+In order to do this, the organizers decided to use a breakout format for two sessions, and to use an open process to determine the breakout topics.
+Specifically, well-sorted\footnote{\url{http://www.well-sorted.org}} was used as follows:
+\begin{enumerate}
+\item Authors were asked to create well-sorted ``cards'' for the papers.  These cards have a title (50 characters maximum) and a body (255 characters maximum).
+\item Authors, program committee members, and members of the WSSSPE mailing list were asked to sort the cards.
+Each person drags the cards, one by one, into groups.
+A group can have as many cards as the person wants it to have, and it can have whatever meaning makes sense to that person.
+\item Well-sorted then produces an average of all the sorts, which the organizers \todo{need to wait to decide what's next}
+(\end{enumerate}
 

### 2014-11-04 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
adding well-sorted info and results

@@
 A group can have as many cards as the person wants it to have, and it can have whatever meaning makes sense to that person.
-\item Well-sorted then produces an average of all the sorts, which the organizers \todo{need to wait to decide what's next}
-(\end{enumerate}
+\item Well-sorted then produces a set of averages of all the sorts, with various numbers of clusters of cards.
+\end{enumerate}
+
+The organizers then chose the soft with five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
+\begin{itemize}
+\item Exploring Sustainability
+\item Software Development Experiences
+\item Policy
+\item Reproducibility \& Reuse \& Sharing
+\item Code Testing \& Code Review
+\end{itemize}
+
+Finally, since some of the papers were not represented by cards in the process, they were not placed in groups, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups, as listed in the next section of this report.
 
@@
 \begin{itemize}
+\item Exploring Sustainability
+\begin{itemize}
+\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}
+\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}
+\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar.
+Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}
+\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}
+\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}
+\item Colin C.
+Venters, Michael K.
+Griffiths, Violeta Holmes, Rupert R.
+Ward and David J.
+Cooke.
+The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}
+\end{itemize}
 
+\item Software Development Experiences
+\begin{itemize}
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software Development in Practice ~\cite{wssspe2_adams}
-
 \item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software registry~\cite{wssspe2_allen}
-
-\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel.
-The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}
-
 \item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
-
-\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}
-
-\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}
-
-\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
-
-\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
-
 \item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to scientific software: a hackathon experience~\cite{wssspe2_crusoe}
-
-\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}
-
 \item Yolanda Gil,  Eunyoung Moon and James Howison. No Science Software is an Island: Collaborative Software Development Needs in Geosciences~\cite{wssspe2_gil}
-
-\item ed Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna.
+\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna.
 The Hierarchical Data Format ({HDF}): A Foundation for Sustainable Data and Software~\cite{wssspe2_habermann}
-
 \item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating for Science~\cite{wssspe2_hanwell}
-
-\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}
-
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski. Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
-
-\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}
-
-\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}
-
 \item W.
@@
 {ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
-
-\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
-
-\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar.
-Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}
-
-\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}
-
-\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}
-
-\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}
-
-\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}
-
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
-
-\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}
-
-\item Andrew E.
-Slaughter, Derek R.
-Gaston, John Peterson, Cody J.
-Permann, David Andrs and Jason M.
-Miller.
-Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
-
 \item James S.
@@
 The {H}ighly {A}ccurate {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
+\end{itemize}
 
-\item Colin C.
-Venters, Michael K.
-Griffiths, Violeta Holmes, Rupert R.
-Ward and David J.
-Cooke.
-The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}
+\item Policy
+\begin{itemize}
+\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}
+\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}
+\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}
+\end{itemize}
+
+\item Reproducibility \& Reuse \& Sharing
+\begin{itemize}
+\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel.
+The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}
+\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}
+\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}
+\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
+\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
+\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}
+\end{itemize}
+
+\item Code Testing \& Code Review
+\begin{itemize}
+\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
+\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}
+\item Andrew E.
+Slaughter, Derek R.
+Gaston, John Peterson, Cody J.
+Permann, David Andrs and Jason M.
+Miller.
+Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
+\end{itemize}
+
+\end{itemize}
+
+Uncategorized
+
+\begin{itemize}
+
+\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
 

### 2014-11-04 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
adding section on other parts of the meeting

@@
 
+\section{Other parts of the workshop}
+
+WSSSPE2 will include two keynote presentations, and lightening talks from accepted paper authors, in addition to a set of breakout sessions.
+The breakouts will allow attendees to participate in discussions about sustainability, future actions, and two of the five areas that resulted from the community submission, review, and grouping process.
+There will also be reporting from the breakout groups to collect information for all attendees, as well as people who were not able to attend, to understand the discussions.
+
 \section{Conclusions}

### 2014-11-04 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
Minor update
Added organization

@@
 \newcommand{\davidnote}[1]{  {\textcolor{darkgreen}      { ***David: #1 }}}
-\newcommand{\note}[1]{ {\textcolor{red}    { #1 }}}
+\newcommand{\colinnote}[1]{ {\textcolor{red}    {***Colin #1 }}}
 
@@
 Karen Cranston$^{(affil?)}$, Manish Parashar$^{(4)}$, David Proctor$^{(affil?)}$, \\
-Matthew Turk$^{(2)}$, Colin Venters$^{(affil?)}$, Nancy Wilkins-Diehr$^{(5)}$}
+Matthew Turk$^{(2)}$, Colin C. Venters$^{(6)}$, Nancy Wilkins-Diehr$^{(5)}$}
 
@@
 \thanks{{}$^{(5)}$ University of California-San Diego, San Diego, CA, USA}
+
+\thanks{{}$^{(6)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
 

### 2014-11-11 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
moving the last uncategorized paper to a category
small changes
fixing Karen's affil

@@
 \author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
-Karen Cranston$^{(affil?)}$, Manish Parashar$^{(4)}$, David Proctor$^{(affil?)}$, \\
-Matthew Turk$^{(2)}$, Colin C. Venters$^{(6)}$, Nancy Wilkins-Diehr$^{(5)}$}
+Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(affil?)}$, \\
+Matthew Turk$^{(2)}$, Colin C. Venters$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$}
 
@@
 
-\thanks{{}$^{(4)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
+\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
 
-\thanks{{}$^{(5)}$ University of California-San Diego, San Diego, CA, USA}
+\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
 
 \thanks{{}$^{(6)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+
+\thanks{{}$^{(7)}$ University of California-San Diego, San Diego, CA, USA}
 
@@
 
-\todo{fix affils}
+\todo{fix David's affil}
 
@@
 \item Software Development Experiences
-\item Policy
+\item Credit \& Incentives
 \item Reproducibility \& Reuse \& Sharing
@@
 
-The contributed papers that will be discussed at the workshop follow, listed by \todo{update when possible - current list is just alphabetic order, not grouped}
+The contributed papers that will be discussed at the workshop follow, listed by groups (determined as described in the previous section.)
 
@@
 {ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
+\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
 \item James S.
@@
 
-\item Policy
+\item Credit \& Incentives
 \begin{itemize}
@@
 
-Uncategorized
-
-\begin{itemize}
-
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
-
-\end{itemize}
-
 \section{Other parts of the workshop}

### 2014-11-12 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
fixing David's affil and the typos he caught

@@
 \author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
-Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(affil?)}$, \\
-Matthew Turk$^{(2)}$, Colin C. Venters$^{(6)}$, Nancy Wilkins-Diehr$^{(7)}$}
+Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
+Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$}
 
@@
 
-\thanks{{}$^{(6)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
 
-\thanks{{}$^{(7)}$ University of California-San Diego, San Diego, CA, USA}
+\thanks{{}$^{(7)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+
+\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
 
 \begin{abstract}
-This technical report discusses the submission and peer-review process used by the Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
+This technical report discusses the submission and peer-review process used by the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
 It is intended to record both the alternative submission and program organization model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
@@
 \maketitle
-
-\todo{fix David's affil}
 
@@
 
-WSSSPE2 will include two keynote presentations, and lightening talks from accepted paper authors, in addition to a set of breakout sessions.
+WSSSPE2 will include two keynote presentations, and lightning talks from accepted paper authors, in addition to a set of breakout sessions.
 The breakouts will allow attendees to participate in discussions about sustainability, future actions, and two of the five areas that resulted from the community submission, review, and grouping process.

### 2015-02-06 WSSSPE2_report/WSSSPE2_review_report.tex
initial commit of WSSSPE2 report - just with pre-workshop report content for now

--- 2014-11-12 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
@@
 
-\title[]{Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review, and Sorting Process, and Results}
+\title[]{Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review and Sorting Process, and Results}
 

### 2015-02-06 WSSSPE2_preworkshop_report/WSSSPE2_review_report.tex
removing extra word in title

@@
 
-\title[]{Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review, and Sorting Process, and Results}
+\title[]{Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review and Sorting Process, and Results}
 

### 2015-02-06 WSSSPE2_report/WSSSPE2_review.tex
renaming main file

--- 2015-02-06 WSSSPE2_report/WSSSPE2_review_report.tex
@@
 
-\title[]{Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2): Submission, Peer-Review and Sorting Process, and Results}
+\title[]{Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)}
 
-\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
-Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$}
-
-\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}
-
-\thanks{{}$^{(2)}$ University of Illinois, Champaign, IL, USA}
-
-\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
-
-\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
-
-\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
-
-\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
-
-\thanks{{}$^{(7)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
-
-\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
+\author{TBD by contributions}
+%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
+%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
+%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$}
+%
+%
+%\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}
+%
+%\thanks{{}$^{(2)}$ University of Illinois, Champaign, IL, USA}
+%
+%
+%\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+%  
+%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
+%
+%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
+%
+%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
+%
+%\thanks{{}$^{(7)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+%
+%\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
 

### 2015-02-06 WSSSPE2_report/WSSSPE2_review.tex
maybe now ready for others to work

@@
 
-The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} will be held on Sunday, 16 November 2014, in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 follows a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
+\todo{update this section with intro to full report - currently just intro for pre-workshop stuff}
+
+link to live notes - \url{https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing}
+
+The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was held on Sunday, 16 November 2014 in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
+WSSSPE2 followed a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
 
@@
 
-\section{Submissions}
+\section{Submissions, Peer-Review, and Peer-Grouping}
+
+\todo{shorten this section? - cite~\cite{WSSSPE2-pre-report} somewhere here}
 
@@
 
-\section{Peer-Review and Peer-Grouping}
-
 The review process was fairly standard, where reviewers bid for papers, then an automated system
@@
 
-The organizers then chose the soft with five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
+The organizers then chose the sort that contained five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
 \begin{itemize}
@@
 
-Finally, since some of the papers were not represented by cards in the process, they were not placed in groups, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups, as listed in the next section of this report.
+Finally, since some of the papers were not represented by cards in the process, they were not placed in groups by the peer-grouping system, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups.
+The later sections of this paper that discuss the breakout groups include a list of the papers associated with each group.
 
-\section{Results}
+\section{Keynotes}
 
-The contributed papers that will be discussed at the workshop follow, listed by groups (determined as described in the previous section.)
+\todo{VOLUNTEER NEEDED to write summaries for Kaitlin and Neil's keynotes}
 
-\begin{itemize}
-\item Exploring Sustainability
+\section{Defining Sustainability}
+
+\todo{Dan to discuss initial breakout and results}
+
+\section{Exploring Sustainability}
+
+\todo{Colin to lead writing of this section}
+
+Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit}
+
+\todo{short intro to group here}
+
+\subsection{Papers}
+
 \begin{itemize}
@@
 
-\item Software Development Experiences
+\subsection{Discussion and Actions}
+
+\section{Software Development Experiences}
+
+\todo{Nancy to lead writing of this section}
+
+Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit}
+
+\todo{short intro to group here}
+
+\subsection{Papers}
+
 \begin{itemize}
@@
 
-\item Credit \& Incentives
+\subsection{Discussion and Actions}
+
+\section{Credit \& Incentives}
+
+\todo{Dan (OR VOLUNTEER) to lead writing of this section}
+
+Google doc notes - \url{https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q}
+
+\todo{short intro to group here}
+
+\subsection{Papers}
+
 \begin{itemize}
@@
 
-\item Reproducibility \& Reuse \& Sharing
+\subsection{Discussion and Actions}
+
+citation ecosystems sub-breakout: \url{https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit}
+
+Taxonomy of Contributorship/Guidelines for Software for Tenure Review sub-breakout: \url{https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit}
+
+Prizes sub-breakout: \url{https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit} (but as of 2/6, nothing is here and need to look at main breakout doc - Suresh is moving text, I think)
+
+\section{Reproducibility \& Reuse \& Sharing}
+
+\todo{Neil OR VOLUNTEER to lead writing of this section}
+
+google doc notes - \url{https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9}
+
+\todo{short intro to group here}
+
+\subsection{Papers}
+
 \begin{itemize}
@@
 
-\item Code Testing \& Code Review
+\subsection{Discussion and Actions}
+
+\section{Code Testing \& Code Review}
+
+\todo{Karen to lead writing of this section}
+
+google notes doc - 
+\url{https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur}
+
+\todo{short intro to group here}
+
+\subsection{Papers}
+
 \begin{itemize}
@@
 
-\end{itemize}
-
-\section{Other parts of the workshop}
-
-WSSSPE2 will include two keynote presentations, and lightning talks from accepted paper authors, in addition to a set of breakout sessions.
-The breakouts will allow attendees to participate in discussions about sustainability, future actions, and two of the five areas that resulted from the community submission, review, and grouping process.
-There will also be reporting from the breakout groups to collect information for all attendees, as well as people who were not able to attend, to understand the discussions.
+\subsection{Discussion and Actions}
 
 \section{Conclusions}
+
+\todo{currently just pre-workshop stuff - needs to be updated}
 
@@
 
+\appendix
+
+\section{Attendees}
+
+\todo{VOLUNTEER NEEDED to neaten this - alphabetize, columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+
+Daniel S. Katz, U Chicago \& Argonne
+Gabrielle Allen, U Illinois
+Alice Allen, Astrophysics Source Code Library (ASCL)
+Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh, N 
+Tom Crick, Cardiff Metropolitan University, UK
+Samin Ishtiaq, Microsoft Research Cambridge, UK
+Matt Turk, NCSA/UIUC
+Colin C. Venters, University of Huddersfield
+Wolfgang Bangerth
+Carl Boettiger, UCSC/ropensci,
+Chris Bogart, ISR/CMU
+Abani Patra, Univ @ Buffalo
+Kyle Marcus, Univ @ Buffalo
+Karen Cranston, NESCent
+Andre Luckow, Rutgers
+Marlon Pierce, Indiana University, 
+Frank L\"{o}ffler, Louisiana State University,
+Steven R. Brandt, Louisiana State University, 
+James Howison, University of Texas
+Ray Idaszak, RENCI/UNC,
+Chris Lenhardt, 
+Robert H. McDonald - Indiana University - Data to Insight Center/Libraries - 
+Marcus D. Hanwell, Kitware, 
+Tom Clune - NASA GSFC - 
+Mike Rilee - NASA/GSFC, Rilee Systems Technologies - 
+Sarah Harris, University of Leeds, 
+James Spencer, Imperial College London,
+John W. Peterson, Idaho National Laboratory,
+Cody J Permann, Idaho National Laboratory, 
+Richard Littauer, MIT/University of Saarland 
+Aleksandra Pawlik, Software Sustainability Institute, 
+Jordan Adams, Tulane University, 
+Jory Schossau, Michigan State University, 
+Robert R Downs, CIESIN, Columbia University, 
+Ethan Davis, UCAR Unidata, 
+Nancy Wilkins-Diehr, SDSC, 
+Suresh Marru, Indiana University 
+Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University, 
+Bryan Marker, The University of Texas at Austin,  
+James Shepherd, Rice University, 
+Ethan White, University of Florida/Utah State University, 
+Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
+Carol Song, Purdue University, 
+John W. Cobb, 
+Matt Jones,  University of California nia Santa Barbara
+Erin Robinson, Foundation for Earth Science 
+Nick Jones, New Zealand eScience Infrastructure, University of Auckland, 
+Yolanda Gil, Information Sciences Institute, University of Southern California, 
+Karthik Ram, , University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
+Sushil Prasad, Georgia State University, Atlanta 
+Nicole Gasparini, Tulane University, New Orleans
+Stephen Piccolo,  Brigham Young University, Utah
+David Henty,  EPCC, The University of Edinburgh
+Alan Simpson,  EPCC, The University of Edinburgh
+Berkin Malkoc, Istanbul Technical University
+Pierre-Yves Aquilanti TOTAL E\&P R\&T USA
+Greg Wilson  Software Carpentry
+Anshu Dubey  Lawrence Berkeley National Laboratory
+Sol Greenspan  National Science Foundation
+Tracy Teal Data Carpentry
+Frank Seinstra Netherlands eScience Center
+Dirk Colbry, Michigan State University
+Kaitlin Thaney,  Mozilla Science Lab, 
+Ted Habermann, The HDF Group
+Benjamin Pharr, University of Mississippi, 
+Eric Hutton,  University of Colorado
+Ray Plante, NCSA/University of Illinois Urbana-Champaign
+Kurt Glaesemann Pacific northwest national lab
+Ardita Shkurti University of Nottingham
+Andy Nutter-Upham, Whitehead Institute,
+David Bernholdt, Oak Ridge National Laboratory
+
+Names below are from working groups - probably duplicates of above but perhaps not all...
+
+Colin C. Venters, University of Huddersfield
+Erin Robinson, Foundation for Earth Science, 
+Robert Downs, CIESIN 
+Justin Shi
+Ethan Davis 
+Kyle Marcus
+Andy Nutter-Upham, Whitehead Institute, 
+Abani Patra
+
+Chris Lenhardt , Nancy Wilkins-Diehr , Marcus Hanwell , Tracy Teal , Jory Schossau , Andre Luckow
+
+Ian Kelley 
+Ray Idaszak 
+Ray Plante 
+Anshu Dubey
+Dirk Colbry 
+Dan Katz
+Chris Bogart
+Sol Greenspan
+
+Kaitlin Thaney
+Abani Patri
+Frank Seinstra
+Frank Loffler
+Richard
+Robert McDonald
+Steven Brandt
+
+Samin Ishtaq, Nathan Weeks, Rupert Nash, Stephen Piccolo, Hilmar Lapp, Tom Crick, Jakob Blomer, Neil Chue Hong
+
+Sarah Harris, Carl Boettiger, Bryan Marker, Berkin Malkoc, Carol Song, Mark Schildhauer, Benjamin Pharr
+
+Karen Cranston, National Evolutionary Synthesis Center,
+Greg Wilson, Software Carpentry, 
+Wolfgang Bangerth 
+Ethan White, Utah State Univ/University of Florida, 
+John Peterson, Idaho National Laboratory, jwpeterson@gmail.com, 
+David Bernholdt
+Aleksandra Pawlik, Software Sustainability Institute, 
+Kurt Glaesemann
+Cody Permann
+Maxime Hughes
+Anshu Dubey
+Tom Clune
+Mike Rilee
+
 \bibliographystyle{plain}

### 2015-02-09 WSSSPE2_report/WSSSPE2_review.tex

@@
 John W. Cobb, 
-Matt Jones,  University of California nia Santa Barbara
+Matt Jones,  University of California Santa Barbara
 Erin Robinson, Foundation for Earth Science 

### 2015-02-10 WSSSPE2_report/WSSSPE2_review.tex
adding some text for understanding sustainability section

--- 2015-02-06 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-\todo{Dan to discuss initial breakout and results}
+\todo{writing led by Dan - other contributions welcome}
+
+In the first interactive session, the attendees divided themselves into groups to discuss software sustainability.  They were asked to
+\begin{enumerate}
+\item discuss what the term ``software sustainability'' meant to them
+\item determine three things they considered to be significant enablers of software sustainability
+\item determine three things they considered to be significant barriers to software sustainability
+\end{enumerate}
+Once each group had come up with answers, all the answers were compiled, and the attendees voted on which they thought were important by a show of hands.
+
+The general responses to what software sustainability meant were:
+
+\begin{itemize}
+\item keeping software scientifically useful
+\item separating techniques in code from knowledge in code
+\item that an adequately large community finds value in software and is willing to sustain it
+\end{itemize}
+
+\katznote{any discussion of this?}
+
+The enablers of and barriers to software sustainability, roughly ranked by attendee votes are shown in Tables~\ref{tb:software_sustainability_enablers} and \ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other items were suggested as barriers, but were not voted on due to lack of time in the session:
+layering up dependencies;
+using software past its sustainable life;
+using software past its usable life;
+inertia for accepted answers vs wrong/right answers;
+monolithic or poor code; and
+need to restructure code when hardware/software/libraries change.
+}
+
+\katznote{any discussion of these tables?}
+
+\begin{table*}[ht]
+\begin{center}
+    \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
+    \begin{scriptsize}
+    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+    \hline
+  Importance & item \\ \hline \hline
+********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
+********** & designing for growth and extension - open development \\ \hline
+******* & culture in community for reuse \\ \hline
+**** & portability \\ \hline
+**** & culture in dev community to support transition between developers \\ \hline
+*** & interdisciplinary people : science + IT experience \\ \hline
+** & planning for end of life \\ \hline
+** & make smart choices about dependencies \\ \hline
+* & thinking of software as product lines - long term vs short term view \\ \hline
+ & not all communities need new software \\ \hline
+ & converting use into resources \\ \hline
+    \end{tabular}
+    \end{scriptsize}
+    \label{tb:software_sustainability_enablers}
+\end{center}   
+\end{table*}
+
+\begin{table*}[ht]
+\begin{center}
+    \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
+    \begin{scriptsize}
+    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+    \hline
+  Importance & item \\ \hline \hline
+******* & lack of incentives, including promotion and tenure process; promotion and tenure process in academic is incompatible \\ \hline
+*****  & absent / poor documentation \\ \hline
+***** & funding difficult \\ \hline
+**** & developers are not computer scientists; don't have software engineering practices (in particular, those needed to scale-up projects to support and be developed by a large sustainable community) \\ \hline
+*** & overreliance on one or two people - bus test \\ \hline
+** & rate of change of underlying technologies \\ \hline
+** & lack of business models for sustainability \\ \hline
+* & lack of training for how to build sustainability into the system \\ \hline
+ & maintenance needed for software is not visible, appears to ``just happen'' \\ \hline
+ & licensing issues \\ \hline
+ & staff turnover - lack of continuity \\ \hline
+    \end{tabular}
+    \end{scriptsize}
+    \label{tb:software_sustainability_barriers}
+\end{center}   
+\end{table*}
 

### 2015-02-10 WSSSPE2_report/WSSSPE2_review.tex
Alphabetized names and removed duplicates
Also removed trailing commas. Added my name to contributors.

@@
 %Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$}
-%
+%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$,}
+%Richard Littauer$^{(9)}$ \\
 %
@@
 %\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
+%
+%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
+%
 
@@
 
-\todo{VOLUNTEER NEEDED to neaten this - alphabetize, columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+\todo{VOLUNTEER NEEDED to neaten this - columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
 
+Abani Patra, Univ @ Buffalo
+Alan Simpson,  EPCC, The University of Edinburgh
+Aleksandra Pawlik, Software Sustainability Institute
+Alice Allen, Astrophysics Source Code Library (ASCL)
+Andre Luckow, Rutgers
+Andy Nutter-Upham, Whitehead Institute
+Anshu Dubey  Lawrence Berkeley National Laboratory
+Ardita Shkurti University of Nottingham
+Benjamin Pharr, University of Mississippi
+Berkin Malkoc, Istanbul Technical University
+Bryan Marker, The University of Texas at Austin
+Carl Boettiger, UCSC/ropensci
+Carol Song, Purdue University
+Chris Bogart, ISR/CMU
+Chris Lenhardt
+Cody J Permann, Idaho National Laboratory
+Colin C. Venters, University of Huddersfield
 Daniel S. Katz, U Chicago \& Argonne
+David Bernholdt, Oak Ridge National Laboratory
+David Henty,  EPCC, The University of Edinburgh
+Dirk Colbry, Michigan State University
+Eric Hutton,  University of Colorado
+Erin Robinson, Foundation for Earth Science 
+Ethan Davis, UCAR Unidata
+Ethan White, University of Florida/Utah State University
+Frank L\"{o}ffler, Louisiana State University
+Frank Seinstra Netherlands eScience Center
 Gabrielle Allen, U Illinois
-Alice Allen, Astrophysics Source Code Library (ASCL)
+Greg Wilson  Software Carpentry
+Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University
+James Howison, University of Texas
+James Shepherd, Rice University
+James Spencer, Imperial College London
+John W. Cobb
+John W. Peterson, Idaho National Laboratory
+Jordan Adams, Tulane University
+Jory Schossau, Michigan State University
+Kaitlin Thaney,  Mozilla Science Lab
+Karen Cranston, NESCent
+Karthik Ram, , University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
+Kurt Glaesemann Pacific northwest national lab
+Kyle Marcus, University of Buffalo
+Marcus D. Hanwell, Kitware
+Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
+Marlon Pierce, Indiana University
+Matt Jones,  University of California Santa Barbara
+Matt Turk, NCSA/UIUC
+Mike Rilee - NASA/GSFC, Rilee Systems Technologies - 
+Nancy Wilkins-Diehr, SDSC
 Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh, N 
+Nick Jones, New Zealand eScience Infrastructure, University of Auckland
+Nicole Gasparini, Tulane University, New Orleans
+Pierre-Yves Aquilanti TOTAL E\&P R\&T USA
+Ray Idaszak, RENCI/UNC
+Ray Plante, NCSA/University of Illinois Urbana-Champaign
+Richard Littauer, MIT/University of Saarland 
+Robert H. McDonald - Indiana University - Data to Insight Center/Libraries - 
+Robert R Downs, CIESIN, Columbia University
+Samin Ishtiaq, Microsoft Research Cambridge, UK
+Sarah Harris, University of Leeds
+Sol Greenspan  National Science Foundation
+Stephen Piccolo,  Brigham Young University, Utah
+Steven R. Brandt, Louisiana State University
+Suresh Marru, Indiana University 
+Sushil Prasad, Georgia State University, Atlanta 
+Ted Habermann, The HDF Group
+Tom Clune - NASA GSFC - 
 Tom Crick, Cardiff Metropolitan University, UK
-Samin Ishtiaq, Microsoft Research Cambridge, UK
-Matt Turk, NCSA/UIUC
-Colin C. Venters, University of Huddersfield
+Tracy Teal Data Carpentry
 Wolfgang Bangerth
-Carl Boettiger, UCSC/ropensci,
-Chris Bogart, ISR/CMU
-Abani Patra, Univ @ Buffalo
-Kyle Marcus, Univ @ Buffalo
-Karen Cranston, NESCent
-Andre Luckow, Rutgers
-Marlon Pierce, Indiana University, 
-Frank L\"{o}ffler, Louisiana State University,
-Steven R. Brandt, Louisiana State University, 
-James Howison, University of Texas
-Ray Idaszak, RENCI/UNC,
-Chris Lenhardt, 
-Robert H. McDonald - Indiana University - Data to Insight Center/Libraries - 
-Marcus D. Hanwell, Kitware, 
-Tom Clune - NASA GSFC - 
-Mike Rilee - NASA/GSFC, Rilee Systems Technologies - 
-Sarah Harris, University of Leeds, 
-James Spencer, Imperial College London,
-John W. Peterson, Idaho National Laboratory,
-Cody J Permann, Idaho National Laboratory, 
-Richard Littauer, MIT/University of Saarland 
-Aleksandra Pawlik, Software Sustainability Institute, 
-Jordan Adams, Tulane University, 
-Jory Schossau, Michigan State University, 
-Robert R Downs, CIESIN, Columbia University, 
-Ethan Davis, UCAR Unidata, 
-Nancy Wilkins-Diehr, SDSC, 
-Suresh Marru, Indiana University 
-Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University, 
-Bryan Marker, The University of Texas at Austin,  
-James Shepherd, Rice University, 
-Ethan White, University of Florida/Utah State University, 
-Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
-Carol Song, Purdue University, 
-John W. Cobb, 
-Matt Jones,  University of California Santa Barbara
-Erin Robinson, Foundation for Earth Science 
-Nick Jones, New Zealand eScience Infrastructure, University of Auckland, 
-Yolanda Gil, Information Sciences Institute, University of Southern California, 
-Karthik Ram, , University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
-Sushil Prasad, Georgia State University, Atlanta 
-Nicole Gasparini, Tulane University, New Orleans
-Stephen Piccolo,  Brigham Young University, Utah
-David Henty,  EPCC, The University of Edinburgh
-Alan Simpson,  EPCC, The University of Edinburgh
-Berkin Malkoc, Istanbul Technical University
-Pierre-Yves Aquilanti TOTAL E\&P R\&T USA
-Greg Wilson  Software Carpentry
-Anshu Dubey  Lawrence Berkeley National Laboratory
-Sol Greenspan  National Science Foundation
-Tracy Teal Data Carpentry
-Frank Seinstra Netherlands eScience Center
-Dirk Colbry, Michigan State University
-Kaitlin Thaney,  Mozilla Science Lab, 
-Ted Habermann, The HDF Group
-Benjamin Pharr, University of Mississippi, 
-Eric Hutton,  University of Colorado
-Ray Plante, NCSA/University of Illinois Urbana-Champaign
-Kurt Glaesemann Pacific northwest national lab
-Ardita Shkurti University of Nottingham
-Andy Nutter-Upham, Whitehead Institute,
-David Bernholdt, Oak Ridge National Laboratory
+Yolanda Gil, Information Sciences Institute, University of Southern California
 
-Names below are from working groups - probably duplicates of above but perhaps not all...
-
-Colin C. Venters, University of Huddersfield
-Erin Robinson, Foundation for Earth Science, 
-Robert Downs, CIESIN 
-Justin Shi
-Ethan Davis 
-Kyle Marcus
-Andy Nutter-Upham, Whitehead Institute, 
-Abani Patra
-
-Chris Lenhardt , Nancy Wilkins-Diehr , Marcus Hanwell , Tracy Teal , Jory Schossau , Andre Luckow
+% Names below were in the working groups but were not present in the above list
 
 Ian Kelley 
-Ray Idaszak 
-Ray Plante 
-Anshu Dubey
-Dirk Colbry 
-Dan Katz
-Chris Bogart
-Sol Greenspan
-
-Kaitlin Thaney
-Abani Patri
-Frank Seinstra
-Frank Loffler
-Richard
-Robert McDonald
-Steven Brandt
-
-Samin Ishtaq, Nathan Weeks, Rupert Nash, Stephen Piccolo, Hilmar Lapp, Tom Crick, Jakob Blomer, Neil Chue Hong
-
-Sarah Harris, Carl Boettiger, Bryan Marker, Berkin Malkoc, Carol Song, Mark Schildhauer, Benjamin Pharr
-
-Karen Cranston, National Evolutionary Synthesis Center,
-Greg Wilson, Software Carpentry, 
-Wolfgang Bangerth 
-Ethan White, Utah State Univ/University of Florida, 
-John Peterson, Idaho National Laboratory, jwpeterson@gmail.com, 
-David Bernholdt
-Aleksandra Pawlik, Software Sustainability Institute, 
-Kurt Glaesemann
-Cody Permann
+Jakob Blomer
+Justin Shi
 Maxime Hughes
-Anshu Dubey
-Tom Clune
 Mike Rilee
+Nathan Weeks
+Rupert Nash
 

### 2015-02-11 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex

@@
 
-\todo{VOLUNTEER NEEDED to neaten this - columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+\todo{VOLUNTEER NEEDED to neaten this - alphabetize by surname, columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
 
-Abani Patra, Univ @ Buffalo
-Alan Simpson,  EPCC, The University of Edinburgh
-Aleksandra Pawlik, Software Sustainability Institute
 Alice Allen, Astrophysics Source Code Library (ASCL)
+David Bernholdt, Oak Ridge National Laboratory
+Jakob Blomer
+Carl Boettiger, UCSC/ropensci
+Chris Bogart, ISR/CMU
+Dirk Colbry, Michigan State University
+Anshu Dubey  Lawrence Berkeley National Laboratory
+David Henty,  EPCC, The University of Edinburgh
+Maxime Hughes
+Eric Hutton,  University of Colorado
+Daniel S. Katz, U Chicago \& Argonne
+Ian Kelley 
+Chris Lenhardt
 Andre Luckow, Rutgers
-Andy Nutter-Upham, Whitehead Institute
-Anshu Dubey  Lawrence Berkeley National Laboratory
-Ardita Shkurti University of Nottingham
-Benjamin Pharr, University of Mississippi
 Berkin Malkoc, Istanbul Technical University
 Bryan Marker, The University of Texas at Austin
-Carl Boettiger, UCSC/ropensci
+Rupert Nash
+Andy Nutter-Upham, Whitehead Institute
+Abani Patra, Univ @ Buffalo
+Aleksandra Pawlik, Software Sustainability Institute
+Cody J Permann, Idaho National Laboratory
+Benjamin Pharr, University of Mississippi
+Mike Rilee
+Erin Robinson, Foundation for Earth Science 
+Alan Simpson,  EPCC, The University of Edinburgh
+Justin Shi
+Ardita Shkurti University of Nottingham
 Carol Song, Purdue University
-Chris Bogart, ISR/CMU
-Chris Lenhardt
-Cody J Permann, Idaho National Laboratory
 Colin C. Venters, University of Huddersfield
-Daniel S. Katz, U Chicago \& Argonne
-David Bernholdt, Oak Ridge National Laboratory
-David Henty,  EPCC, The University of Edinburgh
-Dirk Colbry, Michigan State University
-Eric Hutton,  University of Colorado
-Erin Robinson, Foundation for Earth Science 
+Nathan Weeks
+Ethan White, University of Florida/Utah State University
+
 Ethan Davis, UCAR Unidata
-Ethan White, University of Florida/Utah State University
 Frank L\"{o}ffler, Louisiana State University
@@
 
-% Names below were in the working groups but were not present in the above list
-
-Ian Kelley 
-Jakob Blomer
-Justin Shi
-Maxime Hughes
-Mike Rilee
-Nathan Weeks
-Rupert Nash
-
 \bibliographystyle{plain}

### 2015-02-11 WSSSPE2_report/WSSSPE2_review.tex
change to use macro

--- 2015-02-10 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-The enablers of and barriers to software sustainability, roughly ranked by attendee votes are shown in Tables~\ref{tb:software_sustainability_enablers} and \ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other items were suggested as barriers, but were not voted on due to lack of time in the session:
+The enablers of and barriers to software sustainability, roughly ranked by attendee votes are shown in {\tablename}s~\ref{tb:software_sustainability_enablers} and \ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other items were suggested as barriers, but were not voted on due to lack of time in the session:
 layering up dependencies;

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
Add section label and use comma to separate affiliation

@@
 \section{Attendees}
+\label{sec:attendees}
 
@@
 Dirk Colbry, Michigan State University
-Anshu Dubey  Lawrence Berkeley National Laboratory
+Anshu Dubey, Lawrence Berkeley National Laboratory
 David Henty,  EPCC, The University of Edinburgh
@@
 Aleksandra Pawlik, Software Sustainability Institute
-Cody J Permann, Idaho National Laboratory
+Cody J. Permann, Idaho National Laboratory
 Benjamin Pharr, University of Mississippi
@@
 Justin Shi
-Ardita Shkurti University of Nottingham
+Ardita Shkurti, University of Nottingham
 Carol Song, Purdue University
@@
 Ethan White, University of Florida/Utah State University
-
 Ethan Davis, UCAR Unidata
 Frank L\"{o}ffler, Louisiana State University
-Frank Seinstra Netherlands eScience Center
+Frank Seinstra, Netherlands eScience Center
 Gabrielle Allen, U Illinois
-Greg Wilson  Software Carpentry
+Greg Wilson,  Software Carpentry
 Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University
@@
 Karen Cranston, NESCent
-Karthik Ram, , University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
-Kurt Glaesemann Pacific northwest national lab
+Karthik Ram, University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
+Kurt Glaesemann, Pacific northwest national lab
 Kyle Marcus, University of Buffalo
@@
 Matt Turk, NCSA/UIUC
-Mike Rilee - NASA/GSFC, Rilee Systems Technologies - 
+Mike Rilee, NASA/GSFC, Rilee Systems Technologies - 
 Nancy Wilkins-Diehr, SDSC
-Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh, N 
+Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh
 Nick Jones, New Zealand eScience Infrastructure, University of Auckland
 Nicole Gasparini, Tulane University, New Orleans
-Pierre-Yves Aquilanti TOTAL E\&P R\&T USA
+Pierre-Yves Aquilanti, TOTAL E\&P R\&T USA
 Ray Idaszak, RENCI/UNC
@@
 Richard Littauer, MIT/University of Saarland 
-Robert H. McDonald - Indiana University - Data to Insight Center/Libraries - 
+Robert H. McDonald, Indiana University - Data to Insight Center/Libraries - 
 Robert R Downs, CIESIN, Columbia University
@@
 Sarah Harris, University of Leeds
-Sol Greenspan  National Science Foundation
+Sol Greenspan,  National Science Foundation
 Stephen Piccolo,  Brigham Young University, Utah
@@
 Ted Habermann, The HDF Group
-Tom Clune - NASA GSFC - 
+Tom Clune, NASA GSFC - 
 Tom Crick, Cardiff Metropolitan University, UK
-Tracy Teal Data Carpentry
+Tracy Teal, Data Carpentry
 Wolfgang Bangerth

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
alphabetize attendees

@@
 
-\todo{VOLUNTEER NEEDED to neaten this - alphabetize by surname, columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+\todo{VOLUNTEER NEEDED to neaten this - columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
 
+Jordan Adams, Tulane University
 Alice Allen, Astrophysics Source Code Library (ASCL)
+Gabrielle Allen, U Illinois
+Pierre-Yves Aquilanti, TOTAL E\&P R\&T USA
+Wolfgang Bangerth
 David Bernholdt, Oak Ridge National Laboratory
@@
 Chris Bogart, ISR/CMU
+Steven R. Brandt, Louisiana State University
+Tom Clune, NASA GSFC - 
+John W. Cobb
 Dirk Colbry, Michigan State University
+Karen Cranston, NESCent
+Tom Crick, Cardiff Metropolitan University, UK
+Ethan Davis, UCAR Unidata
+Robert R Downs, CIESIN, Columbia University
 Anshu Dubey, Lawrence Berkeley National Laboratory
+Nicole Gasparini, Tulane University, New Orleans
+Yolanda Gil, Information Sciences Institute, University of Southern California
+Kurt Glaesemann, Pacific northwest national lab
+Sol Greenspan,  National Science Foundation
+Ted Habermann, The HDF Group
+Marcus D. Hanwell, Kitware
+Sarah Harris, University of Leeds
 David Henty,  EPCC, The University of Edinburgh
+Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh
+James Howison, University of Texas
 Maxime Hughes
 Eric Hutton,  University of Colorado
+Ray Idaszak, RENCI/UNC
+Samin Ishtiaq, Microsoft Research Cambridge, UK
+Matt Jones,  University of California Santa Barbara
+Nick Jones, New Zealand eScience Infrastructure, University of Auckland
 Daniel S. Katz, U Chicago \& Argonne
 Ian Kelley
+Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University
 Chris Lenhardt
+Richard Littauer, MIT/University of Saarland 
+Frank L\"{o}ffler, Louisiana State University
 Andre Luckow, Rutgers
 Berkin Malkoc, Istanbul Technical University
+Kyle Marcus, University of Buffalo
 Bryan Marker, The University of Texas at Austin
+Suresh Marru, Indiana University 
+Robert H. McDonald, Indiana University - Data to Insight Center/Libraries - 
 Rupert Nash
@@
 Cody J. Permann, Idaho National Laboratory
+John W. Peterson, Idaho National Laboratory
 Benjamin Pharr, University of Mississippi
+Stephen Piccolo,  Brigham Young University, Utah
+Marlon Pierce, Indiana University
+Ray Plante, NCSA/University of Illinois Urbana-Champaign
+Sushil Prasad, Georgia State University, Atlanta 
+Karthik Ram, University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
 Mike Rilee
+Mike Rilee, NASA/GSFC, Rilee Systems Technologies - 
 Erin Robinson, Foundation for Earth Science 
-Alan Simpson,  EPCC, The University of Edinburgh
+Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
+Jory Schossau, Michigan State University
+Frank Seinstra, Netherlands eScience Center
+James Shepherd, Rice University
 Justin Shi
 Ardita Shkurti, University of Nottingham
+Alan Simpson,  EPCC, The University of Edinburgh
 Carol Song, Purdue University
+James Spencer, Imperial College London
+Tracy Teal, Data Carpentry
+Kaitlin Thaney,  Mozilla Science Lab
+Matt Turk, NCSA/UIUC
 Colin C. Venters, University of Huddersfield
@@
 Ethan White, University of Florida/Utah State University
-Ethan Davis, UCAR Unidata
-Frank L\"{o}ffler, Louisiana State University
-Frank Seinstra, Netherlands eScience Center
-Gabrielle Allen, U Illinois
+Nancy Wilkins-Diehr, SDSC
 Greg Wilson,  Software Carpentry
-Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University
-James Howison, University of Texas
-James Shepherd, Rice University
-James Spencer, Imperial College London
-John W. Cobb
-John W. Peterson, Idaho National Laboratory
-Jordan Adams, Tulane University
-Jory Schossau, Michigan State University
-Kaitlin Thaney,  Mozilla Science Lab
-Karen Cranston, NESCent
-Karthik Ram, University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
-Kurt Glaesemann, Pacific northwest national lab
-Kyle Marcus, University of Buffalo
-Marcus D. Hanwell, Kitware
-Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
-Marlon Pierce, Indiana University
-Matt Jones,  University of California Santa Barbara
-Matt Turk, NCSA/UIUC
-Mike Rilee, NASA/GSFC, Rilee Systems Technologies - 
-Nancy Wilkins-Diehr, SDSC
-Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh
-Nick Jones, New Zealand eScience Infrastructure, University of Auckland
-Nicole Gasparini, Tulane University, New Orleans
-Pierre-Yves Aquilanti, TOTAL E\&P R\&T USA
-Ray Idaszak, RENCI/UNC
-Ray Plante, NCSA/University of Illinois Urbana-Champaign
-Richard Littauer, MIT/University of Saarland 
-Robert H. McDonald, Indiana University - Data to Insight Center/Libraries - 
-Robert R Downs, CIESIN, Columbia University
-Samin Ishtiaq, Microsoft Research Cambridge, UK
-Sarah Harris, University of Leeds
-Sol Greenspan,  National Science Foundation
-Stephen Piccolo,  Brigham Young University, Utah
-Steven R. Brandt, Louisiana State University
-Suresh Marru, Indiana University 
-Sushil Prasad, Georgia State University, Atlanta 
-Ted Habermann, The HDF Group
-Tom Clune, NASA GSFC - 
-Tom Crick, Cardiff Metropolitan University, UK
-Tracy Teal, Data Carpentry
-Wolfgang Bangerth
-Yolanda Gil, Information Sciences Institute, University of Southern California
 

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
Use a tabular environment with two columns

@@
 \usepackage{wrapfig}
+\usepackage{xtab}
 
@@
 
-\todo{VOLUNTEER NEEDED to neaten this - columnize, add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+\todo{VOLUNTEER NEEDED to neaten this - add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
 
-Jordan Adams, Tulane University
-Alice Allen, Astrophysics Source Code Library (ASCL)
-Gabrielle Allen, U Illinois
-Pierre-Yves Aquilanti, TOTAL E\&P R\&T USA
-Wolfgang Bangerth
-David Bernholdt, Oak Ridge National Laboratory
-Jakob Blomer
-Carl Boettiger, UCSC/ropensci
-Chris Bogart, ISR/CMU
-Steven R. Brandt, Louisiana State University
-Tom Clune, NASA GSFC - 
-John W. Cobb
-Dirk Colbry, Michigan State University
-Karen Cranston, NESCent
-Tom Crick, Cardiff Metropolitan University, UK
-Ethan Davis, UCAR Unidata
-Robert R Downs, CIESIN, Columbia University
-Anshu Dubey, Lawrence Berkeley National Laboratory
-Nicole Gasparini, Tulane University, New Orleans
-Yolanda Gil, Information Sciences Institute, University of Southern California
-Kurt Glaesemann, Pacific northwest national lab
-Sol Greenspan,  National Science Foundation
-Ted Habermann, The HDF Group
-Marcus D. Hanwell, Kitware
-Sarah Harris, University of Leeds
-David Henty,  EPCC, The University of Edinburgh
-Neil Chue Hong, Software Sustainability Institute \& University of Edinburgh
-James Howison, University of Texas
-Maxime Hughes
-Eric Hutton,  University of Colorado
-Ray Idaszak, RENCI/UNC
-Samin Ishtiaq, Microsoft Research Cambridge, UK
-Matt Jones,  University of California Santa Barbara
-Nick Jones, New Zealand eScience Infrastructure, University of Auckland
-Daniel S. Katz, U Chicago \& Argonne
-Ian Kelley
-Hilmar Lapp, National Evolutionary Synthesis Center (NESCent) and Duke University
-Chris Lenhardt
-Richard Littauer, MIT/University of Saarland 
-Frank L\"{o}ffler, Louisiana State University
-Andre Luckow, Rutgers
-Berkin Malkoc, Istanbul Technical University
-Kyle Marcus, University of Buffalo
-Bryan Marker, The University of Texas at Austin
-Suresh Marru, Indiana University 
-Robert H. McDonald, Indiana University - Data to Insight Center/Libraries - 
-Rupert Nash
-Andy Nutter-Upham, Whitehead Institute
-Abani Patra, Univ @ Buffalo
-Aleksandra Pawlik, Software Sustainability Institute
-Cody J. Permann, Idaho National Laboratory
-John W. Peterson, Idaho National Laboratory
-Benjamin Pharr, University of Mississippi
-Stephen Piccolo,  Brigham Young University, Utah
-Marlon Pierce, Indiana University
-Ray Plante, NCSA/University of Illinois Urbana-Champaign
-Sushil Prasad, Georgia State University, Atlanta 
-Karthik Ram, University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
-Mike Rilee
-Mike Rilee, NASA/GSFC, Rilee Systems Technologies - 
-Erin Robinson, Foundation for Earth Science 
-Mark Schildhauer,  NCEAS, Univ. Calif, Santa Barbara
-Jory Schossau, Michigan State University
-Frank Seinstra, Netherlands eScience Center
-James Shepherd, Rice University
-Justin Shi
-Ardita Shkurti, University of Nottingham
-Alan Simpson,  EPCC, The University of Edinburgh
-Carol Song, Purdue University
-James Spencer, Imperial College London
-Tracy Teal, Data Carpentry
-Kaitlin Thaney,  Mozilla Science Lab
-Matt Turk, NCSA/UIUC
-Colin C. Venters, University of Huddersfield
-Nathan Weeks
-Ethan White, University of Florida/Utah State University
-Nancy Wilkins-Diehr, SDSC
-Greg Wilson,  Software Carpentry
+\begin{xtabular}{ll}
+Jordan Adams & Tulane University\\
+Alice Allen & Astrophysics Source Code Library (ASCL)\\
+Gabrielle Allen & U Illinois\\
+Pierre-Yves Aquilanti & TOTAL E\&P R\&T USA\\
+Wolfgang Bangerth\\
+David Bernholdt & Oak Ridge National Laboratory\\
+Jakob Blomer\\
+Carl Boettiger & UCSC/ropensci\\
+Chris Bogart & ISR/CMU\\
+Steven R. Brandt & Louisiana State University\\
+Tom Clune & NASA GSFC - \\
+John W. Cobb\\
+Dirk Colbry & Michigan State University\\
+Karen Cranston & NESCent\\
+Tom Crick & Cardiff Metropolitan University, UK\\
+Ethan Davis & UCAR Unidata\\
+Robert R Downs & CIESIN, Columbia University\\
+Anshu Dubey & Lawrence Berkeley National Laboratory\\
+Nicole Gasparini & Tulane University, New Orleans\\
+Yolanda Gil & Information Sciences Institute, University of Southern California\\
+Kurt Glaesemann & Pacific northwest national lab\\
+Sol Greenspan &  National Science Foundation\\
+Ted Habermann & The HDF Group\\
+Marcus D. Hanwell & Kitware\\
+Sarah Harris & University of Leeds\\
+David Henty &  EPCC, The University of Edinburgh\\
+Neil Chue Hong & Software Sustainability Institute \& University of Edinburgh\\
+James Howison & University of Texas\\
+Maxime Hughes\\
+Eric Hutton &  University of Colorado\\
+Ray Idaszak & RENCI/UNC\\
+Samin Ishtiaq & Microsoft Research Cambridge, UK\\
+Matt Jones &  University of California Santa Barbara\\
+Nick Jones & New Zealand eScience Infrastructure, University of Auckland\\
+Daniel S. Katz & U Chicago \& Argonne\\
+Ian Kelley\\
+Hilmar Lapp & National Evolutionary Synthesis Center (NESCent) and Duke University\\
+Chris Lenhardt\\
+Richard Littauer & MIT/University of Saarland \\
+Frank L\"{o}ffler & Louisiana State University\\
+Andre Luckow & Rutgers\\
+Berkin Malkoc & Istanbul Technical University\\
+Kyle Marcus & University of Buffalo\\
+Bryan Marker & The University of Texas at Austin\\
+Suresh Marru & Indiana University \\
+Robert H. McDonald & Indiana University - Data to Insight Center/Libraries - \\
+Rupert Nash\\
+Andy Nutter-Upham & Whitehead Institute\\
+Abani Patra & Univ @ Buffalo\\
+Aleksandra Pawlik & Software Sustainability Institute\\
+Cody J. Permann & Idaho National Laboratory\\
+John W. Peterson & Idaho National Laboratory\\
+Benjamin Pharr & University of Mississippi\\
+Stephen Piccolo &  Brigham Young University, Utah\\
+Marlon Pierce & Indiana University\\
+Ray Plante & NCSA/University of Illinois Urbana-Champaign\\
+Sushil Prasad & Georgia State University, Atlanta \\
+Karthik Ram & University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci\\
+Mike Rilee\\
+Mike Rilee & NASA/GSFC, Rilee Systems Technologies - \\
+Erin Robinson & Foundation for Earth Science \\
+Mark Schildhauer &  NCEAS, Univ. Calif, Santa Barbara\\
+Jory Schossau & Michigan State University\\
+Frank Seinstra & Netherlands eScience Center\\
+James Shepherd & Rice University\\
+Justin Shi\\
+Ardita Shkurti & University of Nottingham\\
+Alan Simpson &  EPCC, The University of Edinburgh\\
+Carol Song & Purdue University\\
+James Spencer & Imperial College London\\
+Tracy Teal & Data Carpentry\\
+Kaitlin Thaney &  Mozilla Science Lab\\
+Matt Turk & NCSA/UIUC\\
+Colin C. Venters & University of Huddersfield\\
+Nathan Weeks\\
+Ethan White & University of Florida/Utah State University\\
+Nancy Wilkins-Diehr & SDSC\\
+Greg Wilson &  Software Carpentry\\
+\end{xtabular}
 

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
Minor changes to affiliations

@@
 Steven R. Brandt & Louisiana State University\\
-Tom Clune & NASA GSFC - \\
+Tom Clune & NASA GSFC\\
 John W. Cobb\\
@@
 Suresh Marru & Indiana University \\
-Robert H. McDonald & Indiana University - Data to Insight Center/Libraries - \\
+Robert H. McDonald & Indiana University - Data to Insight Center/Libraries \\
 Rupert Nash\\
@@
 Mike Rilee\\
-Mike Rilee & NASA/GSFC, Rilee Systems Technologies - \\
+Mike Rilee & NASA/GSFC, Rilee Systems Technologies \\
 Erin Robinson & Foundation for Earth Science \\
-Mark Schildhauer &  NCEAS, Univ. Calif, Santa Barbara\\
+Mark Schildhauer &  NCEAS, Univ. California, Santa Barbara\\
 Jory Schossau & Michigan State University\\

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
Align tabular table

@@
 \begin{xtabular}{ll}
-Jordan Adams & Tulane University\\
-Alice Allen & Astrophysics Source Code Library (ASCL)\\
-Gabrielle Allen & U Illinois\\
-Pierre-Yves Aquilanti & TOTAL E\&P R\&T USA\\
-Wolfgang Bangerth\\
-David Bernholdt & Oak Ridge National Laboratory\\
-Jakob Blomer\\
-Carl Boettiger & UCSC/ropensci\\
-Chris Bogart & ISR/CMU\\
-Steven R. Brandt & Louisiana State University\\
-Tom Clune & NASA GSFC\\
-John W. Cobb\\
-Dirk Colbry & Michigan State University\\
-Karen Cranston & NESCent\\
-Tom Crick & Cardiff Metropolitan University, UK\\
-Ethan Davis & UCAR Unidata\\
-Robert R Downs & CIESIN, Columbia University\\
-Anshu Dubey & Lawrence Berkeley National Laboratory\\
-Nicole Gasparini & Tulane University, New Orleans\\
-Yolanda Gil & Information Sciences Institute, University of Southern California\\
-Kurt Glaesemann & Pacific northwest national lab\\
-Sol Greenspan &  National Science Foundation\\
-Ted Habermann & The HDF Group\\
-Marcus D. Hanwell & Kitware\\
-Sarah Harris & University of Leeds\\
-David Henty &  EPCC, The University of Edinburgh\\
-Neil Chue Hong & Software Sustainability Institute \& University of Edinburgh\\
-James Howison & University of Texas\\
-Maxime Hughes\\
-Eric Hutton &  University of Colorado\\
-Ray Idaszak & RENCI/UNC\\
-Samin Ishtiaq & Microsoft Research Cambridge, UK\\
-Matt Jones &  University of California Santa Barbara\\
-Nick Jones & New Zealand eScience Infrastructure, University of Auckland\\
-Daniel S. Katz & U Chicago \& Argonne\\
-Ian Kelley\\
-Hilmar Lapp & National Evolutionary Synthesis Center (NESCent) and Duke University\\
-Chris Lenhardt\\
+Jordan Adams          &  Tulane University                       \\
+Alice Allen           &  Astrophysics Source Code Library (ASCL) \\
+Gabrielle Allen       &  U Illinois                              \\
+Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA                     \\
+Wolfgang Bangerth     \\
+David Bernholdt       &  Oak Ridge National Laboratory           \\
+Jakob Blomer          \\
+Carl Boettiger        &  UCSC/ropensci                           \\
+Chris Bogart          &  ISR/CMU                                 \\
+Steven R. Brandt      &  Louisiana State University              \\
+Tom Clune             &  NASA GSFC                               \\
+John W. Cobb          \\
+Dirk Colbry           &  Michigan State University               \\
+Karen Cranston        &  NESCent                                 \\
+Tom Crick             &  Cardiff Metropolitan University, UK     \\
+Ethan Davis           &  UCAR Unidata                            \\
+Robert R Downs        &  CIESIN, Columbia University             \\
+Anshu Dubey           &  Lawrence Berkeley National Laboratory                                \\
+Nicole Gasparini      &  Tulane University, New Orleans                                       \\
+Yolanda Gil           &  Information Sciences Institute, University of Southern California    \\
+Kurt Glaesemann       &  Pacific northwest national lab                                       \\
+Sol Greenspan         &  National Science Foundation                                          \\
+Ted Habermann         &  The HDF Group                                                        \\
+Marcus D. Hanwell     &  Kitware                                                              \\
+Sarah Harris          &  University of Leeds                                                  \\
+David Henty           &  EPCC, The University of Edinburgh                                    \\
+Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh         \\
+James Howison         &  University of Texas                                                  \\
+Maxime Hughes         \\
+Eric Hutton           &  University of Colorado                                               \\
+Ray Idaszak           &  RENCI/UNC                                                            \\
+Samin Ishtiaq         &  Microsoft Research Cambridge, UK                                     \\
+Matt Jones            &  University of California Santa Barbara                               \\
+Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland          \\
+Daniel S. Katz        &  U Chicago \& Argonne                                                 \\
+Ian Kelley            \\
+Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) and Duke University \\
+Chris Lenhardt        \\
 Richard Littauer      &  MIT/University of Saarland                                           \\
-Frank L\"{o}ffler & Louisiana State University\\
-Andre Luckow & Rutgers\\
-Berkin Malkoc & Istanbul Technical University\\
-Kyle Marcus & University of Buffalo\\
-Bryan Marker & The University of Texas at Austin\\
+Frank L\"{o}ffler     &  Louisiana State University                                           \\
+Andre Luckow          &  Rutgers                                                              \\
+Berkin Malkoc         &  Istanbul Technical University                                        \\
+Kyle Marcus           &  University of Buffalo                                                \\
+Bryan Marker          &  The University of Texas at Austin                                    \\
 Suresh Marru          &  Indiana University                                                   \\
 Robert H. McDonald    &  Indiana University - Data to Insight Center/Libraries                \\
-Rupert Nash\\
-Andy Nutter-Upham & Whitehead Institute\\
-Abani Patra & Univ @ Buffalo\\
-Aleksandra Pawlik & Software Sustainability Institute\\
-Cody J. Permann & Idaho National Laboratory\\
-John W. Peterson & Idaho National Laboratory\\
-Benjamin Pharr & University of Mississippi\\
-Stephen Piccolo &  Brigham Young University, Utah\\
-Marlon Pierce & Indiana University\\
-Ray Plante & NCSA/University of Illinois Urbana-Champaign\\
+Rupert Nash           \\
+Andy Nutter-Upham     &  Whitehead Institute                                                  \\
+Abani Patra           &  Univ @ Buffalo                                                       \\
+Aleksandra Pawlik     &  Software Sustainability Institute                                    \\
+Cody J. Permann       &  Idaho National Laboratory                                            \\
+John W. Peterson      &  Idaho National Laboratory                                            \\
+Benjamin Pharr        &  University of Mississippi                                            \\
+Stephen Piccolo       &  Brigham Young University, Utah                                       \\
+Marlon Pierce         &  Indiana University                                                   \\
+Ray Plante            &  NCSA/University of Illinois Urbana-Champaign                         \\
 Sushil Prasad         &  Georgia State University, Atlanta                                    \\
-Karthik Ram & University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci\\
-Mike Rilee\\
+Karthik Ram                   &  University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci \\
+Mike Rilee                    \\
 Mike Rilee                    &  NASA/GSFC, Rilee Systems Technologies                                \\
 Erin Robinson                 &  Foundation for Earth Science                                         \\
-Mark Schildhauer &  NCEAS, Univ. California, Santa Barbara\\
-Jory Schossau & Michigan State University\\
-Frank Seinstra & Netherlands eScience Center\\
-James Shepherd & Rice University\\
-Justin Shi\\
-Ardita Shkurti & University of Nottingham\\
-Alan Simpson &  EPCC, The University of Edinburgh\\
-Carol Song & Purdue University\\
-James Spencer & Imperial College London\\
-Tracy Teal & Data Carpentry\\
-Kaitlin Thaney &  Mozilla Science Lab\\
-Matt Turk & NCSA/UIUC\\
-Colin C. Venters & University of Huddersfield\\
-Nathan Weeks\\
-Ethan White & University of Florida/Utah State University\\
-Nancy Wilkins-Diehr & SDSC\\
-Greg Wilson &  Software Carpentry\\
+Mark Schildhauer              &  NCEAS, Univ. California, Santa Barbara                               \\
+Jory Schossau                 &  Michigan State University                                            \\
+Frank Seinstra                &  Netherlands eScience Center                                          \\
+James Shepherd                &  Rice University                                                      \\
+Justin Shi                    \\
+Ardita Shkurti                &  University of Nottingham                                             \\
+Alan Simpson                  &  EPCC, The University of Edinburgh                                    \\
+Carol Song                    &  Purdue University                                                    \\
+James Spencer                 &  Imperial College London                                              \\
+Tracy Teal                    &  Data Carpentry                                                       \\
+Kaitlin Thaney                &  Mozilla Science Lab                                                  \\
+Matt Turk                     &  NCSA/UIUC                                                            \\
+Colin C. Venters              &  University of Huddersfield                                           \\
+Nathan Weeks                  \\
+Ethan White                   &  University of Florida/Utah State University                          \\
+Nancy Wilkins-Diehr           &  SDSC                                                                 \\
+Greg Wilson                   &  Software Carpentry                                                   \\
 \end{xtabular}

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
Remove duplicate
Add reference to WSSSPE2 collaborative notes

@@
 
-\todo{VOLUNTEER NEEDED to neaten this - add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+The following is a partial list of workshop attendees who registered on the
+collaborative notes document~\cite{WSSSPE2-google-notes} that was being used
+for live note taking at the workshop, with some additions from
+attendees to breakout groups.
 
@@
 Karthik Ram                   &  University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci \\
-Mike Rilee                    \\
 Mike Rilee                    &  NASA/GSFC, Rilee Systems Technologies                                \\

### 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex

--- 2015-02-15 WSSSPE2_report/WSSSPE2_review.tex
@@
 \usepackage{wrapfig}
-\usepackage{xtab}
+\usepackage{longtable}
 
@@
 
-\todo{VOLUNTEER NEEDED to neaten this - add short intro sentence that these were generated from people who signed the google doc saying they were there and are not complete, also added names from breakout groups}
+The following is a partial list of workshop attendees who registered on the
+collaborative notes document~\cite{WSSSPE2-google-notes} that was used
+for shared note-taking at the meeting. Some additional
+participants in breakout groups are included in the list.
 
-\begin{xtabular}{ll}
+\begin{longtable}{ll}
 Jordan Adams          &  Tulane University                       \\
@@
 Greg Wilson                   &  Software Carpentry                                                   \\
-\end{xtabular}
+\end{longtable}
 

### 2015-02-16 WSSSPE2_report/WSSSPE2_review.tex
align longtable for attendees

@@
 collaborative notes document~\cite{WSSSPE2-google-notes} that was used
-for shared note-taking at the meeting. Some additional
-participants in breakout groups are included in the list.
+for shared note-taking at the meeting, or who participated in a breakout groups
+and were noted in that group's notes.
 

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
Use inter-word space after middle initials
minor change in wording
Add name to contributors list
Add line breaks.
Use short url for TODO. Fix typos.

@@
 
-link to live notes - \url{https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing}
+link to live notes -
+\url{tinyurl.com/q6ew45v}%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing}
 
-The Second Workshop on on Sustainable Software for Science: Practice and Experiences (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was held on Sunday, 16 November 2014 in conjunction with the 2014 International Conference for High Performance Computing, Networking, Storage and Analysis (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 followed a general initial workshop, WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1}, that was held in 2013 in conjunction with the SC13 conference, and a focused workshop, WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, that was held in 2014 in conjunction with the SciPy conference.
+The Second Workshop on Sustainable Software for Science: Practice and
+Experiences
+(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
+held on Sunday, 16 November 2014 in conjunction with the 2014 International
+Conference for High Performance Computing, Networking, Storage and Analysis
+(SC14)\footnote{\url{http://sc14.supercomputing.org}}.
+WSSSPE2 followed a general initial workshop,
+WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
+that was held in 2013 in conjunction with the SC13 conference, and a focused
+workshop,
+WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
+that was held in 2014 in conjunction with the SciPy conference.
 
-Progress in scientific research is dependent on the quality and accessibility of software at all levels and it is critical to address challenges related to the development, deployment, maintenance and overall sustainability of reusable software as well as education around software practices.
-These challenges can be technological, policy based, organizational, and educational, and are of interest to developers (the software community), users (science disciplines), software engineering researchers, and researchers studying the conduct of science (science of team science, science of organizations, science of science and innovation policy, and social science communities).
-The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas of interest for sustainable scientific software.
-WSSSPE2 invites the community to propose and discuss specific mechanisms to move towards an imagined future practice for software development and usage in science and engineering.
-The workshop will include multiple mechanisms for participation, encourage team building around solutions, and identify risky solutions with potentially transformative outcomes.
-Participation by early career students and postdoctoral researchers is strongly encouraged.
+Progress in scientific research is dependent on the quality and accessibility of
+software at all levels and it is critical to address challenges related to the
+development, deployment, maintenance, and overall sustainability of reusable
+software as well as education around software practices. These challenges can be
+technological, policy based, organizational, and educational, and are of
+interest to developers (the software community), users (science disciplines),
+software-engineering researchers, and researchers studying the conduct of
+science (science of team science, science of organizations, science of science
+and innovation policy, and social science communities). The WSSSPE1 workshop
+engaged the broad scientific community to identify challenges and best practices
+in areas of interest for sustainable scientific software. WSSSPE2 invites the
+community to propose and discuss specific mechanisms to move towards an imagined
+future practice for software development and usage in science and engineering.
+The workshop includes multiple mechanisms for participation, encourages team
+building around solutions, and identifies risky solutions with potentially
+transformative outcomes. It strongly encourages participation of early-career
+students and postdoctoral researchers.
 

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
Front matter and Section 1:  minor changes.

@@
 \begin{abstract}
-This technical report discusses the submission and peer-review process used by the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2) and the results of that process.
-It is intended to record both the alternative submission and program organization model used by WSSSPE2 as well as the papers associated with the workshop that resulted from that process.
+This technical report discusses the submission and peer-review process used by
+the Second Workshop on Sustainable Software for Science: Practice and
+Experiences (WSSSPE2) and the results of that process. It is intended to record
+both the alternative submission and program organization model used by WSSSPE2
+as well as the papers associated with the workshop that resulted from that
+process.
 \end{abstract}
@@
 
-\todo{update this section with intro to full report - currently just intro for pre-workshop stuff}
-
-link to live notes -
-\url{tinyurl.com/q6ew45v}%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing}
+\todo{update this section with intro to full report - currently just intro for pre-workshop stuff. Link to live notes -
+\url{tinyurl.com/q6ew45v}}
+%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
 

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
Section 2: Change list margin.
Section 2: line breaks.

@@
 \usepackage{longtable}
+
+\usepackage{enumitem}
+\setlist{leftmargin=3mm}
 
@@
 
-\item defining software sustainability in the context of science and engineering software
+\item defining software sustainability in the context of science and engineering
+software
 \begin{itemize}
@@
 \item methods to develop sustainable software from the outset
-\item effective approaches to reusable software created as a by-product of research
-\item impact of computer science research on the development of scientific software
+\item effective approaches to reusable software created as a by-product of
+research
+\item impact of computer science research on the development of scientific
+software
 \end{itemize}
@@
 \item governance, business, and sustainability models
-\item the role of community software repositories, their operation and sustainability
+\item the role of community software repositories, their operation and
+sustainability
 \item reproducibility, transparency needs that may be unique to science
@@
 \item software credit, attribution, incentive, and reward
-\item issues related to multiple organizations and multiple countries, such as intellectual property, licensing, etc.
+\item issues related to multiple organizations and multiple countries, such as
+intellectual property, licensing, etc.
 \item mechanisms and venues for publishing software, and the role of publishers
@@
 \begin{itemize}
-\item best practices for providing graduate students and postdoctoral researchers in domain communities with sufficient training in software development
+\item best practices for providing graduate students and postdoctoral
+researchers in domain communities with sufficient training in software
+development
 \item novel uses of sustainable software in education (K-20)
-\item case studies from students on issues around software development in the undergraduate or graduate curricula
+\item case studies from students on issues around software development in the
+undergraduate or graduate curricula
 \end{itemize}
@@
 \item successful examples of career paths for developers
-\item institutional changes to support sustainable software such as promotion and tenure metrics, job categories, etc.
+\item institutional changes to support sustainable software such as promotion
+and tenure metrics, job categories, etc.
 \end{itemize}
@@
 
-Based on the goal of encouraging a wide range of submissions from those involved in software practice, ranging from initial thoughts and partial studies to mature deployments, but focusing on papers that are intended to lead to changes, the organizers wanted to make submission as easy as possible.
-The call for papers stated:
+Based on the goal of encouraging a wide range of submissions from those involved
+in software practice, ranging from initial thoughts and partial studies to
+mature deployments, but focusing on papers that are intended to lead to changes,
+the organizers wanted to make submission as easy as possible. The call for
+papers stated:
 
 \begin{quote}
-We invite short (4-page) \textbf{actionable} papers that will lead to improvements for sustainable software science.
-These papers could be a call to action, or could provide position or experience reports on sustainable software activities.
-The papers will be used by the organizing committee to design sessions that will be highly interactive and targeted towards facilitating action.
-Submitted papers should be archived by a third-party service that provides DOIs.
-We encourage submitters to license their papers under a Creative Commons license that encourages sharing and remixing, as we will combine ideas (with attribution) into the outcomes of the workshop.\end{quote}
+We invite short (4-page) \textbf{actionable} papers that will lead to
+improvements for sustainable software science. These papers could be a call to
+action, or could provide position or experience reports on sustainable software
+activities. The papers will be used by the organizing committee to design
+sessions that will be highly interactive and targeted towards facilitating
+action. Submitted papers should be archived by a third-party service that
+provides DOIs. We encourage submitters to license their papers under a Creative
+Commons license that encourages sharing and remixing, as we will combine ideas
+(with attribution) into the outcomes of the workshop.
+\end{quote}
 
-31 submissions were received, and all but one used either arXiv\footnote{\url{http://arxiv.org}} or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+31 submissions were received, and all but one used either
+arXiv\footnote{\url{http://arxiv.org}} or
+figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
 
-The review process was fairly standard, where reviewers bid for papers, then an automated system
-matched bids to determine assignments, and reviewers then completed their assigned
-reviews (with an average of 4.9 reviews per paper, and 4.1 reviews per reviewer)
-This process was done through EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide scores on relevance and comments to the organizers, which were used to decide which papers to associate with the workshop, and comments to the authors, which were provided back to the authors to allow them to improve their papers.
+The review process was fairly standard, where reviewers bid for papers, then an
+automated system
+matched bids to determine assignments, and reviewers then completed their
+assigned
+reviews (with an average of 4.9 reviews per paper, and 4.1 reviews per
+reviewer).
+This process was done through EasyChair\footnote{http://easychair.org/}, which
+allowed reviewers to provide scores on relevance and comments to the organizers,
+which were used to decide which papers to associate with the workshop, and
+comments to the authors, which were provided back to the authors to allow them
+to improve their papers.
 
-The organizers decided to list 28 of the papers as significantly contributing to the workshop, a very high acceptance rate, but one that is reasonable, given the goal of broad participation and the fact that the reports were already self-published.
+The organizers decided to list 28 of the papers as significantly contributing to
+the workshop, a very high acceptance rate, but one that is reasonable, given the
+goal of broad participation and the fact that the reports were already
+self-published.
 
-WSSSPE1 was organized into sessions, each of which was aimed at discussing one or more of the themes from the call for papers, with a few paper authors invited to summarize the other papers in that them as a panel, followed by general discussion about that theme.
-The mapping of papers to themes was done by the organizers.
+WSSSPE1 was organized into sessions, each of which was aimed at discussing one
+or more of the themes from the call for papers, with a few paper authors invited
+to summarize the other papers in that them as a panel, followed by general
+discussion about that theme. The mapping of papers to themes was done by the
+organizers.
 
-For WSSSPE2, the organizers wanted to increase the interactivity of the sessions, and to open the process of creating the sessions to the full program committee, the paper authors, and others who might attend the workshop.
-In order to do this, the organizers decided to use a breakout format for two sessions, and to use an open process to determine the breakout topics.
-Specifically, well-sorted\footnote{\url{http://www.well-sorted.org}} was used as follows:
+For WSSSPE2, the organizers wanted to increase the interactivity of the
+sessions, and to open the process of creating the sessions to the full program
+committee, the paper authors, and others who might attend the workshop. In order
+to do this, the organizers decided to use a breakout format for two sessions,
+and to use an open process to determine the breakout topics. Specifically,
+well-sorted\footnote{\url{http://www.well-sorted.org}} was used as follows:
 \begin{enumerate}
-\item Authors were asked to create well-sorted ``cards'' for the papers.  These cards have a title (50 characters maximum) and a body (255 characters maximum).
-\item Authors, program committee members, and members of the WSSSPE mailing list were asked to sort the cards.
-Each person drags the cards, one by one, into groups.
-A group can have as many cards as the person wants it to have, and it can have whatever meaning makes sense to that person.
-\item Well-sorted then produces a set of averages of all the sorts, with various numbers of clusters of cards.
+\item Authors were asked to create well-sorted ``cards'' for the papers. These
+cards have a title (50 characters maximum) and a body (255 characters maximum).
+\item Authors, program committee members, and members of the WSSSPE mailing list
+were asked to sort the cards. Each person drags the cards, one by one, into
+groups. A group can have as many cards as the person wants it to have, and it
+can have whatever meaning makes sense to that person.
+\item Well-sorted then produces a set of averages of all the sorts, with various
+numbers of clusters of cards.
 \end{enumerate}
 
-The organizers then chose the sort that contained five groups as the one that felt most meaningful, then decided on names for the five groups, namely:
+The organizers then chose the sort that contained five groups as the one that
+felt most meaningful, then decided on names for the five groups, namely:
 \begin{itemize}
@@
 
-Finally, since some of the papers were not represented by cards in the process, they were not placed in groups by the peer-grouping system, so the authors of these papers were asked which group seemed the best for their papers, and those papers were then placed in those groups.
-The later sections of this paper that discuss the breakout groups include a list of the papers associated with each group.
+Finally, since some of the papers were not represented by cards in the process,
+they were not placed in groups by the peer-grouping system, so the authors of
+these papers were asked which group seemed the best for their papers, and those
+papers were then placed in those groups. The later sections of this paper that
+discuss the breakout groups include a list of the papers associated with each
+group.
 

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
Section headers: add visual aid.
Subsection headers: add visual aids.
Shorten URLs.

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Introduction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Submissions, Peer-Review, and Peer-Grouping}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Keynotes}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{VOLUNTEER NEEDED to write summaries for Kaitlin and Neil's keynotes}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Defining Sustainability}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{writing led by Dan - other contributions welcome}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Exploring Sustainability}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{Colin to lead writing of this section.}
 
-\todo{Colin to lead writing of this section}
-
-Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit}
+Google doc notes - \url{http://tinyurl.com/mpbhvyb}
+%https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Papers}
-
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \begin{itemize}
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Software Development Experiences}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{Nancy to lead writing of this section}
 
-Google doc notes - \url{https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit}
+Google doc notes - \url{http://tinyurl.com/pn4eq8z}
+%https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Credit \& Incentives}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{Dan (OR VOLUNTEER) to lead writing of this section}
 
-Google doc notes - \url{https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q}
+Google doc notes - \url{http://tinyurl.com/k8ruyn9}
+%https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+citation ecosystems sub-breakout: \url{http://tinyurl.com/l4tc6a4}
+%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
 
-citation ecosystems sub-breakout: \url{https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit}
+Taxonomy of Contributorship/Guidelines for Software for Tenure Review sub-breakout: \url{http://tinyurl.com/mx58tmx}
+%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
 
-Taxonomy of Contributorship/Guidelines for Software for Tenure Review sub-breakout: \url{https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit}
+Prizes sub-breakout: \url{http://tinyurl.com/mjok4o5}
+%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit 
+(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh is moving text, I think)
 
-Prizes sub-breakout: \url{https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit} (but as of 2/6, nothing is here and need to look at main breakout doc - Suresh is moving text, I think)
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Reproducibility \& Reuse \& Sharing}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{Neil OR VOLUNTEER to lead writing of this section}
 
-google doc notes - \url{https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9}
+google doc notes - \url{http://tinyurl.com/kqpe87z}
+%https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Code Testing \& Code Review}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{Karen to lead writing of this section}
@@
 google notes doc - 
-\url{https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur}
+\url{http://tinyurl.com/l5t5h45}
+%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Papers}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 \subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \todo{currently just pre-workshop stuff - needs to be updated}
@@
 \appendix
-
-\section{Attendees}
-\label{sec:attendees}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Attendees}  \label{sec:attendees}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The following is a partial list of workshop attendees who registered on the

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
All long lines: added breaks.

@@
 %Richard Littauer$^{(9)}$ \\
+%Miguel de Val-Borro$^{(10)}$
 %
@@
 %
+%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
 
@@
 
-In the first interactive session, the attendees divided themselves into groups to discuss software sustainability.  They were asked to
+In the first interactive session, the attendees divided themselves into groups
+to discuss software sustainability. They were asked to
 \begin{enumerate}
 \item discuss what the term ``software sustainability'' meant to them
-\item determine three things they considered to be significant enablers of software sustainability
-\item determine three things they considered to be significant barriers to software sustainability
+
+\item determine three things they considered to be significant enablers of
+software sustainability
+
+\item determine three things they considered to be significant barriers to
+software sustainability
 \end{enumerate}
-Once each group had come up with answers, all the answers were compiled, and the attendees voted on which they thought were important by a show of hands.
+Once each group had come up with answers, all the answers were compiled, and the
+attendees voted on which they thought were important by a show of hands.
 
 The general responses to what software sustainability meant were:
-
 \begin{itemize}
@@
 \item separating techniques in code from knowledge in code
-\item that an adequately large community finds value in software and is willing to sustain it
+\item that an adequately large community finds value in software and is willing
+to sustain it
 \end{itemize}
@@
 
-The enablers of and barriers to software sustainability, roughly ranked by attendee votes are shown in {\tablename}s~\ref{tb:software_sustainability_enablers} and \ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other items were suggested as barriers, but were not voted on due to lack of time in the session:
+The enablers of and barriers to software sustainability, roughly ranked by
+attendee votes are shown in
+{\tablename}s~\ref{tb:software_sustainability_enablers} and
+\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
+items were suggested as barriers, but were not voted on due to lack of time in
+the session:
 layering up dependencies;
@@
 \begin{center}
-    \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
+\caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
+indicating the fraction of attendees who voted for an item as important}
     \begin{scriptsize}
@@
 \begin{center}
-    \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly indicating the fraction of attendees who voted for an item as important}
+\caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
+indicating the fraction of attendees who voted for an item as important}
     \begin{scriptsize}
@@
   Importance & item \\ \hline \hline
-******* & lack of incentives, including promotion and tenure process; promotion and tenure process in academic is incompatible \\ \hline
+******* & lack of incentives, including promotion and tenure process; promotion
+and tenure process in academic is incompatible \\ \hline
 *****  & absent / poor documentation \\ \hline
 ***** & funding difficult \\ \hline
-**** & developers are not computer scientists; don't have software engineering practices (in particular, those needed to scale-up projects to support and be developed by a large sustainable community) \\ \hline
+**** & developers are not computer scientists; don't have software engineering
+practices (in particular, those needed to scale-up projects to support and be
+developed by a large sustainable community) \\ \hline
 *** & overreliance on one or two people - bus test \\ \hline
@@
 \begin{itemize}
-\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining Sustainability through Developers' Eyes: Recommendations from an Interview Study~\cite{wssspe2_rosada_de_souza}
-\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, Nicholas Weber. Community Recommendations for Sustainable Scientific Software~\cite{wssspe2_downs}
+\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining
+Sustainability through Developers' Eyes: Recommendations from an Interview
+Study~\cite{wssspe2_rosada_de_souza}
+
+\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis,
+Nicholas Weber. Community Recommendations for Sustainable Scientific
+Software~\cite{wssspe2_downs}
+
 \item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar.
-Role of Online Platforms, Communications and Workflows in Developing Sustainable Software for Science Communities~\cite{wssspe2_patra}
-\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up, Pulling It Forward~\cite{wssspe2_pierce}
-\item Justin Shi. Seeking the Principles of Sustainable Software Engineering~\cite{wssspe2_shi}
-\item Colin C.
-Venters, Michael K.
-Griffiths, Violeta Holmes, Rupert R.
-Ward and David J.
-Cooke.
-The Nebuchadnezzar Effect: Dreaming of Sustainable Software through Sustainable Software Architectures~\cite{wssspe2_venters}
+Role of Online Platforms, Communications and Workflows in Developing Sustainable
+Software for Science Communities~\cite{wssspe2_patra}
+
+\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up,
+Pulling It Forward~\cite{wssspe2_pierce}
+
+\item Justin Shi. Seeking the Principles of Sustainable Software
+Engineering~\cite{wssspe2_shi}
+
+\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward and
+David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures~\cite{wssspe2_venters}
 \end{itemize}
@@
 \begin{itemize}
-\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software Development in Practice ~\cite{wssspe2_adams}
-\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software registry~\cite{wssspe2_allen}
-\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
-\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to scientific software: a hackathon experience~\cite{wssspe2_crusoe}
-\item Yolanda Gil,  Eunyoung Moon and James Howison. No Science Software is an Island: Collaborative Software Development Needs in Geosciences~\cite{wssspe2_gil}
-\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe Rnc Maia and Suren Byna.
-The Hierarchical Data Format ({HDF}): A Foundation for Sustainable Data and Software~\cite{wssspe2_habermann}
-\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating for Science~\cite{wssspe2_hanwell}
-\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski. Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
-\item W.
-Christopher Lenhardt, Stanley Ahalt, Matt Jones, J.
-Aukema, S.
-Hampton, S.
-R.
-Hespanh, R.
-Idaszak and M.
-Schildhauer.
-{ISEES-WSSI} Lessons for Sustainable Science Software from an Early Career Training Institute on Open Science Synthesis~\cite{wssspe2_lenhardt}
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do Scientists Find Most Useful?~\cite{wssspe2_schossau}
-\item James S.
-Spencer, Nicholas S.
-Blunt, William A.
-Vigor, Fionn D.
-Malone, W.
-M.
-C.
-Foulkes, James J.
-Shepherd and Alex J.
-W.
-Thom.
-The {H}ighly {A}ccurate {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
+\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
+Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software
+Development in Practice ~\cite{wssspe2_adams}
+
+\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software
+registry~\cite{wssspe2_allen}
+
+\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building
+software, building community: lessons from the {ROpenSci}
+project~\cite{wssspe2_boettiger}
+
+\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
+scientific software: a hackathon experience~\cite{wssspe2_crusoe}
+
+\item Yolanda Gil, Eunyoung Moon and James Howison. No Science Software is an
+Island: Collaborative Software Development Needs in
+Geosciences~\cite{wssspe2_gil}
+
+\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay
+Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
+Rnc Maia and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
+Sustainable Data and Software~\cite{wssspe2_habermann}
+
+\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software
+Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
+for Science~\cite{wssspe2_hanwell}
+
+\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski.
+Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
+
+\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
+S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for
+Sustainable Science Software from an Early Career Training Institute on Open
+Science Synthesis~\cite{wssspe2_lenhardt}
+
+\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do
+Scientists Find Most Useful?~\cite{wssspe2_schossau}
+
+\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W.
+M. C. Foulkes, James J. Shepherd and Alex J. W. Thom. The {H}ighly {A}ccurate
+{N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic
+diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
 \end{itemize}
@@
 \begin{itemize}
-\item James Howison. Retract bit-rotten publications: Aligning incentives for sustaining scientific software~\cite{wssspe2_howison}
-\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with {JSON-LD}~\cite{wssspe2_katz}
-\item Ian Kelley. Publish or perish: the credit deficit to making software and generating data~\cite{wssspe2_kelley}
+\item James Howison. Retract bit-rotten publications: Aligning incentives for
+sustaining scientific software~\cite{wssspe2_howison}
+
+\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with
+{JSON-LD}~\cite{wssspe2_katz}
+
+\item Ian Kelley. Publish or perish: the credit deficit to making software and
+generating data~\cite{wssspe2_kelley}
 \end{itemize}
@@
 
-Taxonomy of Contributorship/Guidelines for Software for Tenure Review sub-breakout: \url{http://tinyurl.com/mx58tmx}
+Taxonomy of Contributorship/Guidelines for Software for Tenure Review
+sub-breakout: \url{http://tinyurl.com/mx58tmx}
 %https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
@@
 %https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit 
-(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh is moving text, I think)
+(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh
+is moving text, I think)
 
@@
 \begin{itemize}
-\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis, Gerardo Ganis, George Lestaris and Ren\'{e} Meusel.
-The Need for a Versioned Data Analysis Software Environment~\cite{wssspe2_blomer}
-\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support Reproducible Research~\cite{wssspe2_chamberlain}
-\item Neil Chue Hong. Minimal information for reusable scientific software~\cite{wssspe2_chue_hong}
-\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
-\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
-\item Stephen Piccolo. Building Portable Analytical Environments to improve sustainability of computational-analysis pipelines in the sciences~\cite{wssspe2_piccolo}
+\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis,
+Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned
+Data Analysis Software Environment~\cite{wssspe2_blomer}
+
+\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support
+Reproducible Research~\cite{wssspe2_chamberlain}
+
+\item Neil Chue Hong. Minimal information for reusable scientific
+software~\cite{wssspe2_chue_hong}
+
+\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your
+Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
+
+\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making
+Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
+
+\item Stephen Piccolo. Building Portable Analytical Environments to improve
+sustainability of computational-analysis pipelines in the
+sciences~\cite{wssspe2_piccolo}
 \end{itemize}
@@
 \begin{itemize}
-\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
-\item Marian Petre and Greg Wilson. Code Review For and By Scientists~\cite{wssspe2_petre}
-\item Andrew E.
-Slaughter, Derek R.
-Gaston, John Peterson, Cody J.
-Permann, David Andrs and Jason M.
-Miller.
-Continuous Integration for Concurrent {MOOSE} Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
+\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential
+Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
+
+\item Marian Petre and Greg Wilson. Code Review For and By
+Scientists~\cite{wssspe2_petre}
+
+\item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann,
+David Andrs and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
+Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
 \end{itemize}
@@
 
-The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can collaboratively build a workshop agenda.
-The differences in WSSSPE2 from WSSSPE1 are in using an existing service (EasyChair) to handle submissions and reviews, rather than an ad hoc process, and using an existing service (well-sorted) to allow collaborative grouping of papers into themes by all authors, reviewers, and the community, rather than this being done in an ad hoc manner by the organizers alone.
+The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can
+collaboratively build a workshop agenda. The differences in WSSSPE2 from WSSSPE1
+are in using an existing service (EasyChair) to handle submissions and reviews,
+rather than an ad hoc process, and using an existing service (well-sorted) to
+allow collaborative grouping of papers into themes by all authors, reviewers,
+and the community, rather than this being done in an ad hoc manner by the
+organizers alone.
 
-The fact remains that contributors also want to get credit for their participation in the process.
-And the workshop organizers still want to make sure that the workshop content and their efforts are recorded.
-Ideally, there would be a service that would be able to index the contributions to the workshop, serving the authors, the organizers, and the larger community.
-But since there still isn't such a service today, the workshop organizers are writing this initial report and making use of arXiv as a partial solution to provide a record of the workshop.
+The fact remains that contributors also want to get credit for their
+participation in the process. And the workshop organizers still want to make
+sure that the workshop content and their efforts are recorded. Ideally, there
+would be a service that would be able to index the contributions to the
+workshop, serving the authors, the organizers, and the larger community. But
+since there still isn't such a service today, the workshop organizers are
+writing this initial report and making use of arXiv as a partial solution to
+provide a record of the workshop.
 
-After the workshop, one or more additional papers will be created that will include the discussions at the workshop.
-These papers will likely have many authors, and may be submitted to peer-reviewed journals.
+After the workshop, one or more additional papers will be created that will
+include the discussions at the workshop. These papers will likely have many
+authors, and may be submitted to peer-reviewed journals.
 
@@
 Chris Bogart          &  ISR/CMU                                                                              \\
-Steven R. Brandt      &  Louisiana State University              \\
+Steven R.~Brandt      &  Louisiana State University                                                           \\
 Tom Clune             &  NASA GSFC                                                                            \\
-John W. Cobb          \\
+John W.~Cobb          \\
 Dirk Colbry           &  Michigan State University                                                            \\
@@
 Ted Habermann         &  The HDF Group                                                                        \\
-Marcus D. Hanwell     &  Kitware                                                              \\
+Marcus D.~Hanwell     &  Kitware                                                                              \\
 Sarah Harris          &  University of Leeds                                                                  \\
@@
 Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland                          \\
-Daniel S. Katz        &  U Chicago \& Argonne                                                 \\
+Daniel S.~Katz        &  U Chicago \& Argonne                                                                 \\
 Ian Kelley            \\
@@
 Suresh Marru          &  Indiana University                                                                   \\
-Robert H. McDonald    &  Indiana University - Data to Insight Center/Libraries                \\
+Robert H.~McDonald    &  Indiana University - Data to Insight Center/Libraries                                \\
 Rupert Nash           \\
@@
 Aleksandra Pawlik     &  Software Sustainability Institute                                                    \\
-Cody J. Permann       &  Idaho National Laboratory                                            \\
-John W. Peterson      &  Idaho National Laboratory                                            \\
+Cody J.~Permann       &  Idaho National Laboratory                                                            \\
+John W.~Peterson      &  Idaho National Laboratory                                                            \\
 Benjamin Pharr        &  University of Mississippi                                                            \\
@@
 Matt Turk             &  NCSA/UIUC                                                                            \\
-Colin C. Venters              &  University of Huddersfield                                           \\
+Colin C.~Venters      &  University of Huddersfield                                                           \\
 Nathan Weeks          \\

### 2015-02-17 WSSSPE2_report/WSSSPE2_review.tex
Appendix A: format.

@@
 
+{\small
 \begin{longtable}{ll}
-Jordan Adams          &  Tulane University                                                                    \\
-Alice Allen           &  Astrophysics Source Code Library (ASCL)                                              \\
-Gabrielle Allen       &  U Illinois                                                                           \\
-Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA                                                                  \\
-Wolfgang Bangerth     \\
-David Bernholdt       &  Oak Ridge National Laboratory                                                        \\
-Jakob Blomer          \\
-Carl Boettiger        &  UCSC/ropensci                                                                        \\
-Chris Bogart          &  ISR/CMU                                                                              \\
-Steven R.~Brandt      &  Louisiana State University                                                           \\
-Tom Clune             &  NASA GSFC                                                                            \\
-John W.~Cobb          \\
-Dirk Colbry           &  Michigan State University                                                            \\
-Karen Cranston        &  NESCent                                                                              \\
-Tom Crick             &  Cardiff Metropolitan University, UK                                                  \\
-Ethan Davis           &  UCAR Unidata                                                                         \\
-Robert R Downs        &  CIESIN, Columbia University                                                          \\
-Anshu Dubey           &  Lawrence Berkeley National Laboratory                                                \\
-Nicole Gasparini      &  Tulane University, New Orleans                                                       \\
-Yolanda Gil           &  Information Sciences Institute, University of Southern California                    \\
-Kurt Glaesemann       &  Pacific northwest national lab                                                       \\
-Sol Greenspan         &  National Science Foundation                                                          \\
-Ted Habermann         &  The HDF Group                                                                        \\
-Marcus D.~Hanwell     &  Kitware                                                                              \\
-Sarah Harris          &  University of Leeds                                                                  \\
-David Henty           &  EPCC, The University of Edinburgh                                                    \\
-Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh                         \\
-James Howison         &  University of Texas                                                                  \\
-Maxime Hughes         \\
-Eric Hutton           &  University of Colorado                                                               \\
-Ray Idaszak           &  RENCI/UNC                                                                            \\
-Samin Ishtiaq         &  Microsoft Research Cambridge, UK                                                     \\
-Matt Jones            &  University of California Santa Barbara                                               \\
-Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland                          \\
-Daniel S.~Katz        &  U Chicago \& Argonne                                                                 \\
-Ian Kelley            \\
-Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) and Duke University                 \\
-Chris Lenhardt        \\
-Richard Littauer      &  MIT/University of Saarland                                                           \\
-Frank L\"{o}ffler     &  Louisiana State University                                                           \\
-Andre Luckow          &  Rutgers                                                                              \\
-Berkin Malkoc         &  Istanbul Technical University                                                        \\
-Kyle Marcus           &  University of Buffalo                                                                \\
-Bryan Marker          &  The University of Texas at Austin                                                    \\
-Suresh Marru          &  Indiana University                                                                   \\
-Robert H.~McDonald    &  Indiana University - Data to Insight Center/Libraries                                \\
-Rupert Nash           \\
-Andy Nutter-Upham     &  Whitehead Institute                                                                  \\
-Abani Patra           &  Univ @ Buffalo                                                                       \\
-Aleksandra Pawlik     &  Software Sustainability Institute                                                    \\
-Cody J.~Permann       &  Idaho National Laboratory                                                            \\
-John W.~Peterson      &  Idaho National Laboratory                                                            \\
-Benjamin Pharr        &  University of Mississippi                                                            \\
-Stephen Piccolo       &  Brigham Young University, Utah                                                       \\
-Marlon Pierce         &  Indiana University                                                                   \\
-Ray Plante            &  NCSA/University of Illinois Urbana-Champaign                                         \\
-Sushil Prasad         &  Georgia State University, Atlanta                                                    \\
-Karthik Ram           &  University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci \\
-Mike Rilee            &  NASA/GSFC, Rilee Systems Technologies                                                \\
-Erin Robinson         &  Foundation for Earth Science                                                         \\
-Mark Schildhauer      &  NCEAS, Univ. California, Santa Barbara                                               \\
-Jory Schossau         &  Michigan State University                                                            \\
-Frank Seinstra        &  Netherlands eScience Center                                                          \\
-James Shepherd        &  Rice University                                                                      \\
-Justin Shi            \\
-Ardita Shkurti        &  University of Nottingham                                                             \\
-Alan Simpson          &  EPCC, The University of Edinburgh                                                    \\
-Carol Song            &  Purdue University                                                                    \\
-James Spencer         &  Imperial College London                                                              \\
-Tracy Teal            &  Data Carpentry                                                                       \\
-Kaitlin Thaney        &  Mozilla Science Lab                                                                  \\
-Matt Turk             &  NCSA/UIUC                                                                            \\
-Colin C.~Venters      &  University of Huddersfield                                                           \\
-Nathan Weeks          \\
-Ethan White           &  University of Florida/Utah State University                                          \\
-Nancy Wilkins-Diehr   &  SDSC                                                                                 \\
-Greg Wilson           &  Software Carpentry                                                                   \\
+   Jordan Adams          &  Tulane University                                                                    
+\\ Alice Allen           &  Astrophysics Source Code Library (ASCL)                                              
+\\ Gabrielle Allen       &  U Illinois                                                                           
+\\ Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA                                                                  
+\\ Wolfgang Bangerth     
+\\ David Bernholdt       &  Oak Ridge National Laboratory                                                        
+\\ Jakob Blomer          
+\\ Carl Boettiger        &  UCSC/ropensci                                                                        
+\\ Chris Bogart          &  ISR/CMU                                                                              
+\\ Steven R.~Brandt      &  Louisiana State University                                                           
+\\ Tom Clune             &  NASA GSFC                                                                            
+\\ John W.~Cobb          
+\\ Dirk Colbry           &  Michigan State University                                                            
+\\ Karen Cranston        &  NESCent                                                                              
+\\ Tom Crick             &  Cardiff Metropolitan University, UK                                                  
+\\ Ethan Davis           &  UCAR Unidata                                                                         
+\\ Robert R Downs        &  CIESIN, Columbia University                                                          
+\\ Anshu Dubey           &  Lawrence Berkeley National Laboratory                                                
+\\ Nicole Gasparini      &  Tulane University, New Orleans                                                       
+\\ Yolanda Gil           &  Information Sciences Institute, University of Southern California                    
+\\ Kurt Glaesemann       &  Pacific northwest national lab                                                       
+\\ Sol Greenspan         &  National Science Foundation                                                          
+\\ Ted Habermann         &  The HDF Group                                                                        
+\\ Marcus D.~Hanwell     &  Kitware                                                                              
+\\ Sarah Harris          &  University of Leeds                                                                  
+\\ David Henty           &  EPCC, The University of Edinburgh                                                    
+\\ Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh                         
+\\ James Howison         &  University of Texas                                                                  
+\\ Maxime Hughes         
+\\ Eric Hutton           &  University of Colorado                                                               
+\\ Ray Idaszak           &  RENCI/UNC                                                                            
+\\ Samin Ishtiaq         &  Microsoft Research Cambridge, UK                                                     
+\\ Matt Jones            &  University of California Santa Barbara                                               
+\\ Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland                          
+\\ Daniel S.~Katz        &  U Chicago \& Argonne                                                                 
+\\ Ian Kelley            
+\\ Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) and Duke University                 
+\\ Chris Lenhardt        
+\\ Richard Littauer      &  MIT/University of Saarland                                                           
+\\ Frank L\"{o}ffler     &  Louisiana State University                                                           
+\\ Andre Luckow          &  Rutgers                                                                              
+\\ Berkin Malkoc         &  Istanbul Technical University                                                        
+\\ Kyle Marcus           &  University of Buffalo                                                                
+\\ Bryan Marker          &  The University of Texas at Austin                                                    
+\\ Suresh Marru          &  Indiana University                                                                   
+\\ Robert H.~McDonald    &  Indiana University - Data to Insight Center/Libraries                                
+\\ Rupert Nash           
+\\ Andy Nutter-Upham     &  Whitehead Institute                                                                  
+\\ Abani Patra           &  Univ @ Buffalo                                                                       
+\\ Aleksandra Pawlik     &  Software Sustainability Institute                                                    
+\\ Cody J.~Permann       &  Idaho National Laboratory                                                            
+\\ John W.~Peterson      &  Idaho National Laboratory                                                            
+\\ Benjamin Pharr        &  University of Mississippi                                                            
+\\ Stephen Piccolo       &  Brigham Young University, Utah                                                       
+\\ Marlon Pierce         &  Indiana University                                                                   
+\\ Ray Plante            &  NCSA/University of Illinois Urbana-Champaign                                         
+\\ Sushil Prasad         &  Georgia State University, Atlanta                                                    
+\\ Karthik Ram           &  University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci 
+\\ Mike Rilee            &  NASA/GSFC, Rilee Systems Technologies                                                
+\\ Erin Robinson         &  Foundation for Earth Science                                                         
+\\ Mark Schildhauer      &  NCEAS, Univ. California, Santa Barbara                                               
+\\ Jory Schossau         &  Michigan State University                                                            
+\\ Frank Seinstra        &  Netherlands eScience Center                                                          
+\\ James Shepherd        &  Rice University                                                                      
+\\ Justin Shi            
+\\ Ardita Shkurti        &  University of Nottingham                                                             
+\\ Alan Simpson          &  EPCC, The University of Edinburgh                                                    
+\\ Carol Song            &  Purdue University                                                                    
+\\ James Spencer         &  Imperial College London                                                              
+\\ Tracy Teal            &  Data Carpentry                                                                       
+\\ Kaitlin Thaney        &  Mozilla Science Lab                                                                  
+\\ Matt Turk             &  NCSA/UIUC                                                                            
+\\ Colin C.~Venters      &  University of Huddersfield                                                           
+\\ Nathan Weeks          
+\\ Ethan White           &  University of Florida/Utah State University                                          
+\\ Nancy Wilkins-Diehr   &  SDSC                                                                                 
+\\ Greg Wilson           &  Software Carpentry                                                                   
 \end{longtable}
+}
 

### 2015-02-18 WSSSPE2_report/WSSSPE2_review.tex
Front matter: Debug and add myself in comment.
Fix list space.
Fix URL link and volunteer for section 3.
First keynote from live notes.

@@
 \usepackage{enumitem}
-\setlist{leftmargin=3mm}
+\setlist{leftmargin=7mm}
 
@@
 %Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$,}
-%Richard Littauer$^{(9)}$ \\
-%Miguel de Val-Borro$^{(10)}$
+%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$,
+%Richard Littauer$^{(9)}$,
+%Miguel de Val-Borro$^{(10)}$, 
+%Sou-Cheng T. Choi$^{(11)}$}
 %
@@
 %\thanks{{}$^{(2)}$ University of Illinois, Champaign, IL, USA}
-%
 %
@@
 %\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
+%
+%\thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
 
@@
 \todo{update this section with intro to full report - currently just intro for pre-workshop stuff. Link to live notes -
-\url{tinyurl.com/q6ew45v}}
+\url{http://tinyurl.com/q6ew45v}}
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 \begin{itemize} 
-
+\renewcommand{\labelenumi}{\textbf{\theenumi}.}
+\setlength{\rightmargin}{1em}
 \item defining software sustainability in the context of science and engineering
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{VOLUNTEER NEEDED to write summaries for Kaitlin and Neil's keynotes}
+\todo{Lead: Choi. Volunteers welcome---contact lead.  See \url{http://tinyurl.com/q6ew45v}}
+%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\begin{enumerate}
+\item Why is Mozilla interested in Science: Mozilla is a non-profit interested in openness, in news, in site creation, and in Science, taking advantage of the open web.
+\item Current systems designed to create friction --- social issues hindering us
+\item case example: National Cancer Institute's caBIG
+\begin{itemize}
+\item $350m$ spent, $60m+$ for management, $70 + $ tools created, but still seen as a failure (Kaitlin didn't really say why?
+Were the tools used?
+Did they immediately bit-rot?
+Was their development unsustained?).
+- should ask question about this
+\item Those that had the least investment were the most used, the largest investments were the least used.
+\item Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like there is: 
+\url{http://tinyurl.com/maf6dz2}%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
+\end{itemize}
+
+\item What do we mean by open research? In regard to community/technology/practices
+\item Inefficiency: \url{http://www.xkcd.com/1445/}
+\item Need research tools (e.g. software repositories), social capital (e.g. incentives) and capacity (eg training and mentorship) for efficient reproducible research
+\item Our systems need to talk to each other but Q for speaker is that as systems become less monolithic, paradoxically it may be harder to sustain the links between them see: 
+\begin{itemize}
+\item \url{http://tinyurl.com/l76tba2}%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
+(but does Anon Grizzly have other cites/info for that?)
+\item Note, for example, that creative commons licensed works typically don't have dependencies (a book, a photo, an artwork).
+\end{itemize}
+
+\item Software Discovery Index: \url{http://softwarediscoveryindex.org/report/}
+\item Final points:
+\begin{itemize}
+\item Design with the audience not the individual in mind
+\item Design to unlock latent potential of our systems
+\item Rethink how we reward researchers and support roles
+\item By mindful of jargon/semantics traps
+\end{itemize}
+\item Questions:
+\begin{itemize}
+\item Ted Habermann: if the metadata is minimal less onerous for data provider, but more onerous for user. How to work with data providers to show that users need more.
+\item  Find balance between things that have to be captured and what would be ideal.
+Don't want to get to point where like old Harvard Dataverse, three pages of metadata capture, only first four things ever filled out.
+Don't want to lose engagement.
+\end{itemize}
+\end{enumerate}
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subsection{Neil }
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \todo{writing led by Dan - other contributions welcome}

### 2015-02-18 WSSSPE2_report/WSSSPE2_review.tex
Format.

@@
 \begin{enumerate}
-\item Why is Mozilla interested in Science: Mozilla is a non-profit interested in openness, in news, in site creation, and in Science, taking advantage of the open web.
+\item Why is Mozilla interested in Science: Mozilla is a non-profit interested
+in openness, in news, in site creation, and in Science, taking advantage of the
+open web.
 \item Current systems designed to create friction --- social issues hindering us
@@
 \begin{itemize}
-\item $350m$ spent, $60m+$ for management, $70 + $ tools created, but still seen as a failure (Kaitlin didn't really say why?
-Were the tools used?
-Did they immediately bit-rot?
-Was their development unsustained?).
-- should ask question about this
-\item Those that had the least investment were the most used, the largest investments were the least used.
-\item Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like there is: 
+\item $350m$ spent, $60m+$ for management, $70 + $ tools created, but still seen
+as a failure (Kaitlin didn't really say why? Were the tools used? Did they
+immediately bit-rot? Was their development unsustained?). - should ask question
+about this
+\item Those that had the least investment were the most used, the largest
+investments were the least used.
+\item Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray,
+looks like there is:
 \url{http://tinyurl.com/maf6dz2}%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
@@
 
-\item What do we mean by open research? In regard to community/technology/practices
+\item What do we mean by open research? In regard to
+community/technology/practices
 \item Inefficiency: \url{http://www.xkcd.com/1445/}
-\item Need research tools (e.g. software repositories), social capital (e.g. incentives) and capacity (eg training and mentorship) for efficient reproducible research
-\item Our systems need to talk to each other but Q for speaker is that as systems become less monolithic, paradoxically it may be harder to sustain the links between them see: 
+\item Need research tools (e.g. software repositories), social capital (e.g.
+incentives) and capacity (eg training and mentorship) for efficient reproducible
+research
+\item Our systems need to talk to each other but Q for speaker is that as
+systems become less monolithic, paradoxically it may be harder to sustain the
+links between them see:
 \begin{itemize}
-\item \url{http://tinyurl.com/l76tba2}%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
+\item
+\url{http://tinyurl.com/l76tba2}%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
 (but does Anon Grizzly have other cites/info for that?)
-\item Note, for example, that creative commons licensed works typically don't have dependencies (a book, a photo, an artwork).
+\item Note, for example, that creative commons licensed works typically don't
+have dependencies (a book, a photo, an artwork).
 \end{itemize}
@@
 \begin{itemize}
-\item Ted Habermann: if the metadata is minimal less onerous for data provider, but more onerous for user. How to work with data providers to show that users need more.
-\item  Find balance between things that have to be captured and what would be ideal.
-Don't want to get to point where like old Harvard Dataverse, three pages of metadata capture, only first four things ever filled out.
-Don't want to lose engagement.
+\item Ted Habermann: if the metadata is minimal less onerous for data provider,
+but more onerous for user. How to work with data providers to show that users
+need more.
+\item Find balance between things that have to be captured and what would be
+ideal. Don't want to get to point where like old Harvard Dataverse, three pages
+of metadata capture, only first four things ever filled out. Don't want to lose
+engagement.
 \end{itemize}

### 2015-02-18 WSSSPE2_report/WSSSPE2_review.tex
Add dates.

@@
 
-\author{TBD by contributions}
+\author{TBD by contributions \\{\scriptsize\today \quad  (Due:  $\backsim$ April 1)}}
 %\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
@@
 
-\begin{abstract}
+\begin{abstract} \todo{TBD due date above}       
 This technical report discusses the submission and peer-review process used by

### 2015-02-18 count.sh
Added TODOs

@@ -11,7 +11,7 @@
 #   git shortlog -s -n
 # only returns number of files
   
-for i in "Katz" "ketancmaheshwari" "Terrya" "Lapp" "Loeffler" "Turk" "jameshowison" "Hanwell" "Frank" "wilkins" "Wilkins" "Hetherington" "Allen" "shelswenson"
+for i in "Katz" "ketancmaheshwari" "Terrya" "Lapp" "Loeffler" "Turk" "jameshowison" "Hanwell" "Frank" "wilkins" "Wilkins" "Hetherington" "Allen" "shelswenson" "Anne"
 do
 	git log --author="$i" --shortstat $BRANCH | \
 	awk '/^ [0-9]/ { f += $1; i += $4; d += $6 } \

### 2015-02-24 WSSSPE2_report/WSSSPE2_review.tex
Added TODOs
minor cleanup of affiliations
modify description of notes URLs

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes}
+\section{Keynotes} \label{keynotes}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Lead: Choi. Volunteers welcome---contact lead.  See \url{http://tinyurl.com/q6ew45v}}
+\todo{Lead: Choi. Volunteers welcome---contact lead}
+
+\todo{Live notes at \url{http://tinyurl.com/q6ew45v}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
+Abstracts at \url{http://tinyurl.com/mnenzms}}
 
@@
 \subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability}
+\label{keynote1}
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\todo{Slides: \url{http://tinyurl.com/las8tsq}}
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Neil }
+\subsection{Neil Chue Hong, We are the 92$\%$} 
+\label{keynote2}
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+\todo{Slides: \url{http://tinyurl.com/pd9tmh3}}
 
@@
 
-google doc notes - \url{http://tinyurl.com/kqpe87z}
+Google doc notes - 
+\href{http://tinyurl.com/kqpe87z}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 
-google notes doc - 
+Google doc notes - 
 \url{http://tinyurl.com/l5t5h45}
@@
 \\ Berkin Malkoc         &  Istanbul Technical University
-\\ Kyle Marcus           &  University of Buffalo                                                                
+\\ Kyle Marcus           &  University at Buffalo
 \\ Bryan Marker          &  The University of Texas at Austin
@@
 \\ Andy Nutter-Upham     &  Whitehead Institute
-\\ Abani Patra           &  Univ @ Buffalo                                                                       
+\\ Abani Patra           &  University at Buffalo
 \\ Aleksandra Pawlik     &  Software Sustainability Institute

### 2015-02-24 WSSSPE2_report/WSSSPE2_review.tex
Use href command for external links to google docs notes
The string "google doc notes" can be clicked to open the corresponding
section of the document in a browser

@@
 
-Google doc notes - \url{http://tinyurl.com/mpbhvyb}
+\href{http://tinyurl.com/mpbhvyb}{Google doc notes}
 %https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit
@@
 
-Google doc notes - \url{http://tinyurl.com/pn4eq8z}
+\href{http://tinyurl.com/pn4eq8z}{Google doc notes}
 %https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit
@@
 
-Google doc notes - \url{http://tinyurl.com/k8ruyn9}
+\href{http://tinyurl.com/k8ruyn9}{Google doc notes}
 %https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q
@@
 
-Google doc notes - 
-\href{http://tinyurl.com/kqpe87z}
+\href{http://tinyurl.com/kqpe87z}{Google doc notes}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 
-Google doc notes - 
-\url{http://tinyurl.com/l5t5h45}
+\href{http://tinyurl.com/l5t5h45}{Google doc notes}
 %https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur

### 2015-03-03 WSSSPE2_report/WSSSPE2_review.tex
Section 3: Intro.

@@
 
-\todo{Live notes at \url{http://tinyurl.com/q6ew45v}.
+\todo{Live notes at \href{http://tinyurl.com/q6ew45v}{Google Notes}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-Abstracts at \url{http://tinyurl.com/mnenzms}}
+Abstracts at \href{http://tinyurl.com/mnenzms}{WSSSPE2}.}
+
+The workshop featured two keynote addresses. In the opening keynote
+presentation, Kaitlin Thaney of Mozilla Science talked about her organization's
+work and policy to enable and support sustainable and reproducible scientific
+research. Our second keynote speaker was Neil Chue Hong of Software
+Sustainability Institute. He shined the light on how scientific software is
+prevalently driving advances in many sciences and engineering fields. Following
+the keynotes, participants broke into groups to define and discuss the crucial
+notion ``sustanability'' spawned by the speeches of our keynote speakers as well
+as the theme of our workshop.
 

### 2015-03-03 WSSSPE2_report/WSSSPE2_review.tex
small wording change

--- 2015-02-24 WSSSPE2_report/WSSSPE2_review.tex
@@
 ******* & lack of incentives, including promotion and tenure process; promotion
-and tenure process in academic is incompatible \\ \hline
-*****  & absent / poor documentation \\ \hline
-***** & funding difficult \\ \hline
+and tenure process in academic is incompatible with sustainability \\ \hline
+*****  & absent or poor documentation \\ \hline
+***** & funding to ensure sustainability is difficult to obtain \\ \hline
 **** & developers are not computer scientists; don't have software engineering
@@
 
-Google doc notes - 
-\href{http://tinyurl.com/kqpe87z}
+google doc notes - \url{http://tinyurl.com/kqpe87z}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 
-Google doc notes - 
+google notes doc - 
 \url{http://tinyurl.com/l5t5h45}
@@
 \\ Berkin Malkoc         &  Istanbul Technical University
-\\ Kyle Marcus           &  University at Buffalo
+\\ Kyle Marcus           &  University of Buffalo
 \\ Bryan Marker          &  The University of Texas at Austin
@@
 \\ Andy Nutter-Upham     &  Whitehead Institute
-\\ Abani Patra           &  University at Buffalo
+\\ Abani Patra           &  Univ @ Buffalo
 \\ Aleksandra Pawlik     &  Software Sustainability Institute

### 2015-03-03 WSSSPE2_report/WSSSPE2_review.tex

@@
 
-google doc notes - \url{http://tinyurl.com/kqpe87z}
+\href{http://tinyurl.com/kqpe87z}{Google doc notes}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 
-google notes doc - 
-\url{http://tinyurl.com/l5t5h45}
+\href{http://tinyurl.com/l5t5h45}{Google doc notes}
 %https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
@@
 \\ Berkin Malkoc         &  Istanbul Technical University
-\\ Kyle Marcus           &  University of Buffalo
+\\ Kyle Marcus           &  University at Buffalo
 \\ Bryan Marker          &  The University of Texas at Austin
@@
 \\ Andy Nutter-Upham     &  Whitehead Institute
-\\ Abani Patra           &  Univ @ Buffalo
+\\ Abani Patra           &  University at Buffalo
 \\ Aleksandra Pawlik     &  Software Sustainability Institute

### 2015-03-04 WSSSPE2_report/WSSSPE2_review.tex
working in intro, style, credit section

@@
 \newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
+\newcommand{\note}[1]{ {\textcolor{blueish}    { ***Note:      #1 }}}
 \newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
@@
 \begin{abstract} \todo{TBD due date above}       
+\todo{update abstract}
 This technical report discusses the submission and peer-review process used by
@@
 
-\todo{update this section with intro to full report - currently just intro for pre-workshop stuff. Link to live notes -
-\url{http://tinyurl.com/q6ew45v}}
+\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 
+This report extends a previous report that discussed the submission, peer-review, and peer-grouping processes in detail.~\cite{WSSSPE2-pre-report}.  It is also based on a collaborative set
+of notes taken through a google doc during the workshop.~\cite{WSSSPE2-google-notes} 
+Overall, it discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), \todo{discuss rest of report here...} ...
+exploring sustainability (\S\ref{sec:exploring}) ...
+code testing and code review (\S\ref{sec:code_testing})
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Submissions, Peer-Review, and Peer-Grouping}
+\section{Submissions, Peer-Review, and Peer-Grouping} \label{sec:preworkshop}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\todo{shorten this section? - cite~\cite{WSSSPE2-pre-report} somewhere here}
+\note{this section is taken from \cite{WSSSPE2-google-notes}. It could be shortened.}
 
-The workshop call for papers included the following areas of interest:
+WSSSPE2 began with a call for papers.
+Based on the goal of encouraging a wide range of submissions from those involved
+in software practice, ranging from initial thoughts and partial studies to
+mature deployments, but focusing on papers that are intended to lead to changes,
+the organizers wanted to make submission as easy as possible. The call for
+papers stated:
+
+\begin{quote}
+We invite short (4-page) \textbf{actionable} papers that will lead to
+improvements for sustainable software science. These papers could be a call to
+action, or could provide position or experience reports on sustainable software
+activities. The papers will be used by the organizing committee to design
+sessions that will be highly interactive and targeted towards facilitating
+action. Submitted papers should be archived by a third-party service that
+provides DOIs. We encourage submitters to license their papers under a Creative
+Commons license that encourages sharing and remixing, as we will combine ideas
+(with attribution) into the outcomes of the workshop.
+\end{quote}
+
+The call included the following areas of interest:
 \begin{quote}
@@
 
-Based on the goal of encouraging a wide range of submissions from those involved
-in software practice, ranging from initial thoughts and partial studies to
-mature deployments, but focusing on papers that are intended to lead to changes,
-the organizers wanted to make submission as easy as possible. The call for
-papers stated:
-
-\begin{quote}
-We invite short (4-page) \textbf{actionable} papers that will lead to
-improvements for sustainable software science. These papers could be a call to
-action, or could provide position or experience reports on sustainable software
-activities. The papers will be used by the organizing committee to design
-sessions that will be highly interactive and targeted towards facilitating
-action. Submitted papers should be archived by a third-party service that
-provides DOIs. We encourage submitters to license their papers under a Creative
-Commons license that encourages sharing and remixing, as we will combine ideas
-(with attribution) into the outcomes of the workshop.
-\end{quote}
-
 31 submissions were received, and all but one used either
@@
 these papers were asked which group seemed the best for their papers, and those
-papers were then placed in those groups. The later sections of this paper that
-discuss the breakout groups include a list of the papers associated with each
+papers were then placed in those groups. Sections~\ref{sec:exploring}-\ref{sec:code_testing}
+discuss the breakout groups, including a list of the papers associated with each
 group.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes} \label{keynotes}
+\section{Keynotes} \label{sec:keynotes}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 the keynotes, participants broke into groups to define and discuss the crucial
-notion ``sustanability'' spawned by the speeches of our keynote speakers as well
+notion ``sustainability'' spawned by the speeches of our keynote speakers as well
 as the theme of our workshop.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Exploring Sustainability}
+\section{Exploring Sustainability} \label{sec:exploring}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\href{http://tinyurl.com/mpbhvyb}{Google doc notes}
+\note{\href{http://tinyurl.com/mpbhvyb}{Google doc notes}}
 %https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit
@@
 \todo{short intro to group here}
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\href{http://tinyurl.com/pn4eq8z}{Google doc notes}
+\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
 %https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit
@@
 
-\href{http://tinyurl.com/k8ruyn9}{Google doc notes}
+\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
 %https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q
 
-\todo{short intro to group here}
+This group, with just three papers but good interest and participation from attendees,
+focused on the institutional, social, and cultural mechanisms encourage the creation
+and maintenance of shared software, in the context of what now exists but focusing
+on what mechanisms are desired and how we might achieve them.
+
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subsection{Discussion and Actions}
+%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+The group had some initial discussion ...
+
+The group then broke into three subgroups
+\begin{itemize}
+\item Citation ecosystems
+\item Taxonomy of contributorship/Guidelines for software for tenure review
+\item Prizes
+\end{itemize}
+
+citation ecosystems sub-breakout: \url{http://tinyurl.com/l4tc6a4}
+%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
+
+Taxonomy of Contributorship/Guidelines for Software for Tenure Review
+sub-breakout: \url{http://tinyurl.com/mx58tmx}
+%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
+
+Prizes sub-breakout: \url{http://tinyurl.com/mjok4o5}
+%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit 
+(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh
+is moving text, I think)
 
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-citation ecosystems sub-breakout: \url{http://tinyurl.com/l4tc6a4}
-%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
-
-Taxonomy of Contributorship/Guidelines for Software for Tenure Review
-sub-breakout: \url{http://tinyurl.com/mx58tmx}
-%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
-
-Prizes sub-breakout: \url{http://tinyurl.com/mjok4o5}
-%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit 
-(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh
-is moving text, I think)
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\href{http://tinyurl.com/kqpe87z}{Google doc notes}
+\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Code Testing \& Code Review} \label{sec:code_testing}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{Karen to lead writing of this section}
+
+\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
+%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
+
+\todo{short intro to group here}
+
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Code Testing \& Code Review}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Karen to lead writing of this section}
-
-\href{http://tinyurl.com/l5t5h45}{Google doc notes}
-%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
-
-\todo{short intro to group here}
 
@@
 \end{itemize}
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
 
@@
 \\ Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA
-\\ Wolfgang Bangerth
+\\ Wolfgang Bangerth & Texas A\&M
 \\ David Bernholdt       &  Oak Ridge National Laboratory

### 2015-03-04 WSSSPE2_report/WSSSPE2_review.tex
bit more on section 1 - organization of paper

@@
 of notes taken through a google doc during the workshop.~\cite{WSSSPE2-google-notes} 
-Overall, it discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), \todo{discuss rest of report here...} ...
-exploring sustainability (\S\ref{sec:exploring}) ...
-code testing and code review (\S\ref{sec:code_testing})
+Overall, the report discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), a session on defining sustainability (\S\ref{sec:defining}), the five breakout sessions---exploring sustainability (\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit}); reproducibility \& reuse \& sharing (\S\ref{sec:reproduce}); code testing \& code review (\S\ref{sec:code_testing})---and some conclusions (\S\ref{sec:conclusions}).
+The report also includes an incomplete list of attendees (Appendix \ref{sec:attendees}).
 
@@
 
-\note{this section is taken from \cite{WSSSPE2-google-notes}. It could be shortened.}
+\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Defining Sustainability}
+\section{Defining Sustainability} \label{sec:defining}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Software Development Experiences}
+\section{Software Development Experiences} \label{sec:devel}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Credit \& Incentives}
+\section{Credit \& Incentives} \label{sec:credit}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Reproducibility \& Reuse \& Sharing}
+\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions}
+\section{Conclusions} \label{sec:conclusions}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-03-04 WSSSPE2_report/WSSSPE2_review.tex
Front matter and sections 1-2: tiny changes. Section 3: re-organizing...to be cont'd.

@@
 
-\begin{abstract} \todo{TBD due date above}       
+\begin{abstract}      
 \todo{update abstract}
@@
 
-31 submissions were received, and all but one used either
+31 submissions were received; all but one used  
 arXiv\footnote{\url{http://arxiv.org}} or
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Lead: Choi. Volunteers welcome---contact lead}
-
-\todo{Live notes at \href{http://tinyurl.com/q6ew45v}{Google Notes}.
+\todo{\footnotesize{Lead: Choi. Volunteers contact lead. Live notes: \href{http://tinyurl.com/q6ew45v}{Google Notes}}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-Abstracts at \href{http://tinyurl.com/mnenzms}{WSSSPE2}.}
+Abstracts: \href{http://tinyurl.com/mnenzms}{WSSSPE2}. Slides: \href{http://tinyurl.com/las8tsq}{here}.}
 
 The workshop featured two keynote addresses. In the opening keynote
-presentation, Kaitlin Thaney of Mozilla Science talked about her organization's
-work and policy to enable and support sustainable and reproducible scientific
-research. Our second keynote speaker was Neil Chue Hong of Software
-Sustainability Institute. He shined the light on how scientific software is
-prevalently driving advances in many sciences and engineering fields. Following
-the keynotes, participants broke into groups to define and discuss the crucial
-notion ``sustainability'' spawned by the speeches of our keynote speakers as well
-as the theme of our workshop.
+presentation, Kaitlin Thaney of the Mozilla Science Lab talked about her
+organization's work and policy to enable and support sustainable and
+reproducible scientific research. Our second keynote speaker was Neil Chue Hong
+of Software Sustainability Institute. He shined the light on how scientific
+software is prevalently driving advances in many science and engineering fields.
+Both keynote speeches spawned further discussion among workshop participants on
+the crucial notion ``sustainability'' in the theme of our workshop.
 
@@
 
-\todo{Slides: \url{http://tinyurl.com/las8tsq}}
+Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla).
+Mozilla is a non-profit organization interested in openness, in news, in website
+creation, and in Science, taking advantage of the open web.
 
-\begin{enumerate}
-\item Why is Mozilla interested in Science: Mozilla is a non-profit interested
-in openness, in news, in site creation, and in Science, taking advantage of the
-open web.
-\item Current systems designed to create friction --- social issues hindering us
-\item case example: National Cancer Institute's caBIG
-\begin{itemize}
-\item $350m$ spent, $60m+$ for management, $70 + $ tools created, but still seen
+Thaney started noting the unfortunate fact that many current systems have been designed to create friction hindering us, despite the original intentions.%---social issues hindering us
+An example is the National Cancer Institute's caBIG. 
+A total of $\$350m$ was spent, more than $\$60m$ for management, $70+$ tools created, but still seen
 as a failure (Kaitlin didn't really say why? Were the tools used? Did they
-immediately bit-rot? Was their development unsustained?). - should ask question
-about this
-\item Those that had the least investment were the most used, the largest
+immediately bit-rot? Was their development unsustained?)-should ask question
+about this.
+ Those that had the least investment were the most used, the largest
 investments were the least used.
-\item Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray,
+ Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray,
 looks like there is:
-\url{http://tinyurl.com/maf6dz2}%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
-\end{itemize}
+\url{http://tinyurl.com/maf6dz2}.%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
 
-\item What do we mean by open research? In regard to
+What do we mean by open research? In regard to
 community/technology/practices
-\item Inefficiency: \url{http://www.xkcd.com/1445/}
-\item Need research tools (e.g. software repositories), social capital (e.g.
+
+Inefficiency: \url{http://www.xkcd.com/1445/}. Need research tools (e.g. software repositories), social capital (e.g.
 incentives) and capacity (eg training and mentorship) for efficient reproducible
-research
-\item Our systems need to talk to each other but Q for speaker is that as
+research.
+
+Our systems need to talk to each other but Q for speaker is that as
 systems become less monolithic, paradoxically it may be harder to sustain the
 links between them see:
-\begin{itemize}
-\item
 \url{http://tinyurl.com/l76tba2}%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
-(but does Anon Grizzly have other cites/info for that?)
-\item Note, for example, that creative commons licensed works typically don't
+(but does Anon Grizzly have other cites/info for that?).
+Note, for example, that creative commons licensed works typically do not 
 have dependencies (a book, a photo, an artwork).
-\end{itemize}
+Software Discovery Index: \url{http://softwarediscoveryindex.org/report/}.
 
-\item Software Discovery Index: \url{http://softwarediscoveryindex.org/report/}
-\item Final points:
+ Final points:
 \begin{itemize}
@@
 \end{itemize}
-\item Questions:
+
+Questions:
 \begin{itemize}
@@
 \end{itemize}
-\end{enumerate}
 

### 2015-03-05 WSSSPE2_report/WSSSPE2_review.tex
working in section 7 - in progress

@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+The papers that were discussed in the Exploring Sustainability group are:
 \begin{itemize}
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+The papers that were discussed in the Software Development Experiences group are:
 \begin{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan (OR VOLUNTEER) to lead writing of this section}
+\todo{Dan leading writing of this section}
 
@@
 
-This group, with just three papers but good interest and participation from attendees,
+This group, with just three papers but a large amount of interest and participation from attendees,
 focused on the institutional, social, and cultural mechanisms encourage the creation
@@
 
-The group had some initial discussion ...
+In the first discussion session, this group decided to break into two smaller groups,
+each independently working through the same general topic: credit and incentives.
 
-The group then broke into three subgroups
+\subsubsection{First breakout discussion: Group A}
+The first sub-group discussed around the current system for credit and current incentives,
+which it called ``hacking the incentive structure''.  The group considered four potential
+points of leverage:
+
+First, that we currently have systems that collect information, and these could be modified
+to collect different information, then map that information to actions. We could
+initially build a proof-of-concept for a new use of a given system, then determine
+what actions would be needed to make this use more common.
+
+Second that we could create entirely new systems, perhaps because the existing
+systems are too tied to what they measure, and modifying them is not practical.
+
+Third that we could change academic culture, rather than worrying about the systems.
+This was mostly focused on citations, because they matter for hiring, promotion,
+and tenure decisions.  The group discussed how we could weight the citations
+within papers better than we now do? How could we identify the five citations
+that really matter for a paper, distinguishing them from the related works and
+general background that are also cited.  Perhaps we could break these out in
+the reference list, working with publishers to implement this.  Or maybe we
+could also break out categories of citations, such as the most important software used,
+the previous publication that we are building from, the data that we actually used, etc.
+The Moore Foundation's award in data science was given as an example, asking
+proposers: What are the five canonical citations that are most important to your work?
+This would be a way of giving credit and assigning importance to these works,
+differently from how we just count citations today.  A possible action that the group
+discussed was conducting a longitudinal study of most useful \{software, data, etc.\}
+in a discipline
+
+Fourth, that we could change the ways funders make decisions, and use these funding
+policies as incentives.
+
+After this discussion, group A brainstormed about incentives,
+with the following items suggested:
 \begin{itemize}
-\item Citation ecosystems
-\item Taxonomy of contributorship/Guidelines for software for tenure review
-\item Prizes
+\item running programming contests, creating bounties for contributing to open source software, etc.
+\item augment author lists to give credit to people who don’t now get credit (and making them machine readable)
+\item developing a microcitation standard and mechanism (for both software and data)
+\item developing a well-defined standard for author contribution -- what level of contribution rises to the level of authorship?
+\item leveraging social media for citation and reviewing of content -- then using social media to bring more people into the review process than is traditional
+\item determining where else software can be cited and recorded (e.g., acknowledgments sections of papers)
+\item developing a taxonomy of contributors, tied to places that these metrics are already stored (e.g., ORCID)
+\item making metadata more easy to add for software, creating incentive for providing software metadata -- note that this cannot be centralized
+\item creating something like the h-index that tenure committees can make use of - simplify a way of measuring and documenting the overall credit given to an individual over different projects
+\item thinking about publishing software vs journal - software does not have to be novel
+\item determining guidelines for recommending software characteristics for tenure - perhaps draft guidelines then get ACM or IEEE agreement
 \end{itemize}
 
-citation ecosystems sub-breakout: \url{http://tinyurl.com/l4tc6a4}
+\subsubsection{First breakout discussion: Group B}
+\note{writing in progress}
+Discussion Extracts
+
+Who to incentivize: 
+Those in science (to do better, more shareable, more sustainable work).
+Those in industry (but are interested in science).
+
+Giving credit in the traditional way:  software papers (JORS?).
+Difficulty of adding new authors to publications
+ - original authors disproportionately benefit
+ - how can you add authors?
+Can we control the end-users?
+ - Yes, but we prefer something not so "confrontational"
+Versioning.  How can we add that to publications? To add on people.  JSON-LD percentages.
+John Towns: What do you want to encourage?
+ - "good computational work" to advance their careers.
+ - People to realize that I'm doing good software.
+ - Empirical Q: Most of the developers are not going into traditional scientific careers.
+    - Why are people going to Industry?  Money? Or working without worrying about careers paths
+Portable reputation.
+Trying to support career paths of "those in the grey area (between science and software - in development of software)"
+ - How do you ensure that those "engineers" do not block the path "back into science".  Force the PIs to make programmers co-authors.
+
+Prizes: reframe software as "instrument building"
+	will prizes be good or bad, how can we
+Converting reputation by joining apache software foundation.
+	Google etc.  Inviting ASF and open source people to contribute to scientific code (at least on questions of overlapping
+
+Template for assessing scientific contributions made through software.
+
+making decisions about who to fund from funder perspective - who did what in the project?
+
+someone writes paper in some area, adds references at end to make sure it’s clear that the paper fits in a field, but maybe don't think about how we use citations to give credit - how to rank/say what citations really matter, different from those that just explain the field (required background)
+
+\subsubsection{Group merger and redivision}
+
+After the first breakout session, the groups A and B came together and discussed 
+a rollup of the ideas from the subgroups at a high level:
+\begin{itemize}
+\item citation ecosystem - traces of usage (metrics)
+\item taxonomy of contributorship, understanding roles
+\item prizes
+\item new metrics (for people's activities in software)
+\item guidelines for evaluating scientific contribution through software (perhaps using new metrics)
+\end{itemize}
+
+In the remaining discussion sessions, the group chose to split into three subgroups to discuss
+a version of these topics: citation ecosystems,
+taxonomy of contributorship/guidelines for software for tenure review, and prizes.
+
+\subsubsection{Remaining breakout discussions: Citation ecosystems}
+\note{writing to be done}
+
+\url{http://tinyurl.com/l4tc6a4}
 %https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
 
-Taxonomy of Contributorship/Guidelines for Software for Tenure Review
+\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
+\note{writing to be done}
+
 sub-breakout: \url{http://tinyurl.com/mx58tmx}
@@
 
-Prizes sub-breakout: \url{http://tinyurl.com/mjok4o5}
+\subsubsection{Remaining breakout discussions: Prizes}
+\note{writing to be done}
+
+\url{http://tinyurl.com/mjok4o5}
 %https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit
-(but as of 2/6, nothing is here and need to look at main breakout doc - Suresh
-is moving text, I think)
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+The papers that were discussed in the Citation \& Incentives group are:
 \begin{itemize}
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+The papers that were discussed in the Reproducibility \& Reuse \& Sharing group are:
 \begin{itemize}
@@
 
+The papers that were discussed in the Code Testing \& Code Review group are:
 \begin{itemize}

### 2015-03-07 WSSSPE2_report/WSSSPE2_review.tex
small work in 7.1.3
very small work in 7.1.2

@@
 \note{writing in progress}
-Discussion Extracts
 
-Who to incentivize: 
-Those in science (to do better, more shareable, more sustainable work).
-Those in industry (but are interested in science).
+This group started by discussing who should be incentivized, thinking of two categories of people:  
+those in science (who could be incentivized to do better, more shareable, more sustainable work), and
+those in industry but interested in science (who could be incentivized to contribute to science.)
 
@@
 a version of these topics: citation ecosystems,
-taxonomy of contributorship/guidelines for software for tenure review, and prizes.
+taxonomy of contributorship/guidelines for software for tenure review, and prizes.  The subgroups were asked to clearly identify
+\begin{itemize}
+\item the problem to be solved
+\item steps towards a solution
+\end{itemize}
 

### 2015-03-07 WSSSPE2_report/WSSSPE2_review.tex
text written in 7.1.4

@@
 \subsubsection{Remaining breakout discussions: Citation ecosystems}
-\note{writing to be done}
 
-\url{http://tinyurl.com/l4tc6a4}
+\note{for reference notes are in \url{http://tinyurl.com/l4tc6a4}}
 %https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
+
+This group defined its goal as creating a low-barrier-of-entry method for recording names and roles (and in a second phase, optionally including weights) of contributors (coders and other intellectual contributions) to a software package in a machine readable way (to be called a credit file), then encouraging the scientific community to adopt this practice.
+
+The following general points were initially discussed: 
+\begin{itemize}
+\item The FLASH project was suggested as an example of how something like this has been done.
+\item This data could be a file that can be associated with a citation to the software, either through use of a DOI for the credit file, or by uploading the credit files as associated with the paper.
+\item This idea could also be applied to data
+\item This data (the credit file) should be part of the metadata that is outside of paywalls, like citations are now.
+\item It was suggested that there should also be a separate file to track software dependencies, but this is not the same file.
+\end{itemize}
+
+The group came up with the following actions to be performed:
+\begin{enumerate}
+\item Build a tool that can automatically determine who the contributors are (from a git or other repo), then allows the user to manually edit the output to add/remove people, define roles. 
+\item Work with repositories to encourage them to provide the information we need based on what they already store.
+\item Define what a citation file should look like and what it should be called.
+\item Test adoption, for example, with LBNL.
+\item Create credit file for a set of software.
+\item Build a validator (and perhaps a visualizer) for credit files.
+\item Write a tool to collect files and visualize/output interconnections (which software is used with which), based on an existing project.
+\item When we (the group members) write papers, we should track the software we use, and encourage the software developers to make their software citable and create credit files.
+\item Build a tool to export the credit file to BibTex and other citation styles.
+\item Make sure the BibTex entries (exported from internal data) are somewhat standardized so that they can be imported into papers.
+Also make sure that standard LaTex style files understand and accept these entries.
+\end{enumerate}
 

### 2015-03-08 WSSSPE2_report/WSSSPE2_review.tex
small changes

@@
 Mozilla is a non-profit organization interested in openness, in news, in website
-creation, and in Science, taking advantage of the open web.
+creation, and in Science, all taking advantage of the open web.
 
-Thaney started noting the unfortunate fact that many current systems have been designed to create friction hindering us, despite the original intentions.%---social issues hindering us
+Thaney started noting the unfortunate fact that many current systems have been designed to create friction hindering us, despite the original intentions.
+\katznote{prev sentence is confusing}%---social issues hindering us
 An example is the National Cancer Institute's caBIG. 
-A total of $\$350m$ was spent, more than $\$60m$ for management, $70+$ tools created, but still seen
+A total of \$350m was spent, including more than \$60m for management.  $70+$ tools were created, but caBIG is still seen
 as a failure (Kaitlin didn't really say why? Were the tools used? Did they
@@
 looks like there is:
-\url{http://tinyurl.com/maf6dz2}.%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
+\url{http://tinyurl.com/maf6dz2}. %http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
 
@@
 
-Inefficiency: \url{http://www.xkcd.com/1445/}. Need research tools (e.g. software repositories), social capital (e.g.
-incentives) and capacity (eg training and mentorship) for efficient reproducible
+Inefficiency: \url{http://www.xkcd.com/1445/}. Need research tools (e.g., software repositories), social capital (e.g.,
+incentives) and capacity (e.g., training and mentorship) for efficient reproducible
 research.
@@
 links between them see:
-\url{http://tinyurl.com/l76tba2}%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
+\url{http://tinyurl.com/l76tba2} %http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
 (but does Anon Grizzly have other cites/info for that?).
@@
 
-\item Daniel Katz and Arfon Smith. Implementing Transitive Credit with
+\item Daniel S. Katz and Arfon M. Smith. Implementing Transitive Credit with
 {JSON-LD}~\cite{wssspe2_katz}

### 2015-03-09 WSSSPE2_report/WSSSPE2_review.tex
adding author names to subsections in 7

@@
 \item running programming contests, creating bounties for contributing to open source software, etc.
-\item augment author lists to give credit to people who don’t now get credit (and making them machine readable)
+\item augment author lists to give credit to people who don't now get credit (and making them machine readable)
 \item developing a microcitation standard and mechanism (for both software and data)
@@
 \subsubsection{First breakout discussion: Group B}
-\note{writing in progress}
+\note{writing in progress by James Howison - expected 9 March 2015}
 
@@
 \subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
-\note{writing to be done}
+\note{writing in progress by Frank Seinstra - expected by 22 March 2015}
 
@@
 \subsubsection{Remaining breakout discussions: Prizes}
-\note{writing to be done}
+\note{writing in progress by Suresh Marru - in a GitHub branch to be merged?}
 

### 2015-03-12 WSSSPE2_report/WSSSPE2_review.tex
expanded notes on first breakout into full text, included some elements from memory.

@@
 or more of the themes from the call for papers, with a few paper authors invited
-to summarize the other papers in that them as a panel, followed by general
+to summarize the other papers in them as a panel, followed by general
 discussion about that theme. The mapping of papers to themes was done by the
@@
 \subsubsection{First breakout discussion: Group B}
-\note{writing in progress by James Howison - expected 9 March 2015}
 
@@
 those in science (who could be incentivized to do better, more shareable, more sustainable work), and
-those in industry but interested in science (who could be incentivized to contribute to science.)
+those in industry but interested in science (who could be incentivized to contribute to science.)  John Towns made the point that we are not yet clear enough on exactly what we want to incentivize, suggesting that we need to have a clearer picture of ``good computational work" and what sort of contributions are truly generative for science.
 
-Giving credit in the traditional way:  software papers (JORS?).
-Difficulty of adding new authors to publications
- - original authors disproportionately benefit
- - how can you add authors?
-Can we control the end-users?
- - Yes, but we prefer something not so "confrontational"
-Versioning.  How can we add that to publications? To add on people.  JSON-LD percentages.
-John Towns: What do you want to encourage?
- - "good computational work" to advance their careers.
- - People to realize that I'm doing good software.
- - Empirical Q: Most of the developers are not going into traditional scientific careers.
-    - Why are people going to Industry?  Money? Or working without worrying about careers paths
-Portable reputation.
-Trying to support career paths of "those in the grey area (between science and software - in development of software)"
- - How do you ensure that those "engineers" do not block the path "back into science".  Force the PIs to make programmers co-authors.
+The discussion of incentivizing those in science acknowledged that the publications system was far from perfect for incentivizing good software work.
+Nonetheless discussion focused on bringing software people into publications.
+Two main suggestions: the first is to focus on end-users of software and encourage them to cite software the ``right" way.
+James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway); he suggested making access to the software conditional on a license that requires citation.
+Others found this ``too confrontational" and preferred to concentrate on making it easier to do the right thing.
+The group was more enthusiastic about ``forcing" PIs to include their ``software people" on publications, although there were few ideas on how exactly to do this.
+Another technique mentioned was that when scientific software projects are hosted in organizations like Apache, the scientific contributors can benefit from building their reputations, perhaps yielding job offers that they can use to negotiate better job and career packages.
 
-Prizes: reframe software as "instrument building"
-	will prizes be good or bad, how can we
-Converting reputation by joining apache software foundation.
-	Google etc.  Inviting ASF and open source people to contribute to scientific code (at least on questions of overlapping
+The discussion on incentivizing those outside science focused on accessing the well of affection that those working in software have for scientific research.
+How can the interest and skills of this group be marshaled towards sustainable contributions?
+The is evidence that the migration of scientific software projects to the Apache Foundation has created opportunities for those not employed in the scientific center to contribute to projects initiated for scientists (especially where there is cross-over with industry needs, such as provenance and workflow).
 
-Template for assessing scientific contributions made through software.
-
-making decisions about who to fund from funder perspective - who did what in the project?
-
-someone writes paper in some area, adds references at end to make sure it’s clear that the paper fits in a field, but maybe don't think about how we use citations to give credit - how to rank/say what citations really matter, different from those that just explain the field (required background)
+The group centered on a discussion of developing ``software prizes" arguing that while it is hard to ``mint" other new sources of reputation, prizes are possible without getting too many others on board.
+The prize criteria can form a template for describing what we mean by scientific contributions made through software, particularly focusing on building active communities, not only writing great code.
+Prizes were identified as something worthy of more discussion, prompting a later break-out group, discussed in Section \ref{sec:prizes}.
 
@@
 
-\subsubsection{Remaining breakout discussions: Prizes}
+\subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
 \note{writing in progress by Suresh Marru - in a GitHub branch to be merged?}

### 2015-03-12 WSSSPE2_report/WSSSPE2_review.tex
light editing in 7.1.2-3 and comments

@@
 those in science (who could be incentivized to do better, more shareable, more sustainable work), and
-those in industry but interested in science (who could be incentivized to contribute to science.)  John Towns made the point that we are not yet clear enough on exactly what we want to incentivize, suggesting that we need to have a clearer picture of ``good computational work" and what sort of contributions are truly generative for science.
+those in industry but interested in science (who could be incentivized to contribute to science.)  It was pointed out that we are not yet clear enough on exactly what we want to incentivize, suggesting that we need to have a clearer picture of ``good computational work" and what sort of contributions are truly generative for science.
 
 The discussion of incentivizing those in science acknowledged that the publications system was far from perfect for incentivizing good software work.
-Nonetheless discussion focused on bringing software people into publications.
-Two main suggestions: the first is to focus on end-users of software and encourage them to cite software the ``right" way.
-James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway); he suggested making access to the software conditional on a license that requires citation.
+Nonetheless discussion focused on bringing software people into publications. There were two main suggestions.
+The first is to focus on end users of software and encourage them to cite software the ``right" way. \katznote{what is the second?}
+James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway) \katznote{add cite?}; he suggested making access to the software conditional on a license that requires citation.
 Others found this ``too confrontational" and preferred to concentrate on making it easier to do the right thing.
@@
 How can the interest and skills of this group be marshaled towards sustainable contributions?
-The is evidence that the migration of scientific software projects to the Apache Foundation has created opportunities for those not employed in the scientific center to contribute to projects initiated for scientists (especially where there is cross-over with industry needs, such as provenance and workflow).
+The is evidence that the migration of scientific software projects to the Apache Foundation has created opportunities for those not employed in the scientific center to contribute to projects initiated by scientists (especially where there is cross-over with industry needs, such as provenance and workflow).
 
-The group centered on a discussion of developing ``software prizes" arguing that while it is hard to ``mint" other new sources of reputation, prizes are possible without getting too many others on board.
+The group also discussed developing ``software prizes" arguing that while it is hard to ``mint" other new sources of reputation, prizes are possible without getting too many others on board.
 The prize criteria can form a template for describing what we mean by scientific contributions made through software, particularly focusing on building active communities, not only writing great code.
-Prizes were identified as something worthy of more discussion, prompting a later break-out group, discussed in Section \ref{sec:prizes}.
 
@@
 In the remaining discussion sessions, the group chose to split into three subgroups to discuss
-a version of these topics: citation ecosystems,
-taxonomy of contributorship/guidelines for software for tenure review, and prizes.  The subgroups were asked to clearly identify
+a version of these topics: {\em citation ecosystems},
+{\em taxonomy of contributorship/guidelines for software for tenure review}, and {\em prizes}.  The subgroups were asked to clearly identify
 \begin{itemize}

### 2015-03-12 WSSSPE2_report/WSSSPE2_review.tex
added citation and fiex flow in incentives section

@@
 Nonetheless discussion focused on bringing software people into publications. There were two main suggestions.
-The first is to focus on end users of software and encourage them to cite software the ``right" way. \katznote{what is the second?}
-James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway) \katznote{add cite?}; he suggested making access to the software conditional on a license that requires citation.
+The first is to focus on end users of software and encourage them to cite software the ``right" way.
+James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway) \cite{howison2015jasist} he suggested making access to the software conditional on a license that requires citation.
 Others found this ``too confrontational" and preferred to concentrate on making it easier to do the right thing.
-The group was more enthusiastic about ``forcing" PIs to include their ``software people" on publications, although there were few ideas on how exactly to do this.
+The second was focusing on those leading software projects, and the group was more enthusiastic about ``forcing" PIs to include their ``software people" on publications, although there were few ideas on how exactly to do this.
 Another technique mentioned was that when scientific software projects are hosted in organizations like Apache, the scientific contributors can benefit from building their reputations, perhaps yielding job offers that they can use to negotiate better job and career packages.

### 2015-03-13 WSSSPE2_report/WSSSPE2_review.tex
adding initial prizes text - more to be done by Suresh

@@
 \subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
-\note{writing in progress by Suresh Marru - in a GitHub branch to be merged?}
+\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
+%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit
 
-\url{http://tinyurl.com/mjok4o5}
-%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit
+This group discussed the idea of prizes.  Prizes are expected to reframe software
+as ``instrument building'' but will prizes be good or bad, and how can we make sure
+there are no negative affects and the process cannot be gamed?
+
+Prizes in different categories were discussed (like Academy Awards), for example:
+best contribution (non-founder),
+broadest diversity of contributions,
+best tutorials/documentation,
+best leadership transition (award ex leader and new leader),
+best generalization (taking something that was limited and making it more general), and
+best mentorship of contributors (bringing others into the community).
+
+\katznote{I don't understand this at all.  Can some explain it?}
+There are also other indirect forms of incentives
+(though these are not a direct prize, they still can incentivize someone):
+\begin{itemize}
+\item Converting reputation by joining apache software foundation, Google etc. 
+\item Inviting ASF and open source people to contribute to scientific code (at least on questions of overlapping
+\item Template for assessing scientific contributions made through software.   
+\end{itemize}
+
+To determine who should give prizes, perhaps we should find those who we think would
+be awarded prizes, then ask them who they would want to receive a prize from, then
+ask those organizations to see if they are willing to be involved in the process.
+
+One idea is to create a funding program for disciplines or other organizations to
+create a prize program.  We provide a framework, for example: awards to individuals, must
+award in 5 or 6 areas, must have a jury and/or objective criteria that includes
+senior/junior domain people and technology people, must have the recipients awarded
+at a relevant event, must provide citations that explain why the prizes should be awarded.
+A concern is that having many organizations award prizes may reduce the impact.
+
+Possible awarders are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology Society of America, etc.
+Or perhaps this should be a joint technology/science partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary society] prize?
+
+Some potential criteria for prizes are:
+community engagement, helping out others;
+number of unique contributors;
+adding new pieces of functionality to software;
+integrating software into broader ecosystem; championing broad principles of sustainability, open science, open source, etc.;
+improving accessibility to software, to scientific software (perhaps championing inclusiveness or making software accessible?);
+documentation;
+tutorials;
+commits / patching;
+leadership transition; and
+best contribution by a non-founder.
+
+Perhaps there should be different criteria for ``established'' members of the community versus junior members?  Or perhaps prizes should be restricted to junior members?
+
+An important point is that we don't want to give prizes just to reward people who are really good at this one thing, but rather we want to reward people who are building the culture we want as scientists.
 

### 2015-03-18 WSSSPE2_report/WSSSPE2_review.tex
Section 1: minor changes.

@@
 (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on Sunday, 16 November 2014 in conjunction with the 2014 International
+held on Sunday, 16 November 2014 in conjunction with the  International
 Conference for High Performance Computing, Networking, Storage and Analysis
 (SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 followed a general initial workshop,
+WSSSPE2 followed the model of a general initial workshop,
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-that was held in 2013 in conjunction with the SC13 conference, and a focused
+which co-occurred with the SC13 conference, and a focused
 workshop,
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
-that was held in 2014 in conjunction with the SciPy conference.
+which was organized in 2014 jointly with the SciPy conference.
 
 Progress in scientific research is dependent on the quality and accessibility of
-software at all levels and it is critical to address challenges related to the
+software at all levels. Hence it is critical to address challenges related to the
 development, deployment, maintenance, and overall sustainability of reusable
@@
 engaged the broad scientific community to identify challenges and best practices
-in areas of interest for sustainable scientific software. WSSSPE2 invites the
+in areas of interest to creating sustainable scientific software. WSSSPE2 invited the
 community to propose and discuss specific mechanisms to move towards an imagined
 future practice for software development and usage in science and engineering.
-The workshop includes multiple mechanisms for participation, encourages team
-building around solutions, and identifies risky solutions with potentially
-transformative outcomes. It strongly encourages participation of early-career
-students and postdoctoral researchers.
+The workshop included multiple mechanisms for participation, encouraged team
+building around solutions, and identified risky solutions with potentially
+transformative outcomes. It strongly encouraged participation of early-career
+scientists, postdoctoral researchers, and graduate students.
 
 This report extends a previous report that discussed the submission, peer-review, and peer-grouping processes in detail.~\cite{WSSSPE2-pre-report}.  It is also based on a collaborative set
-of notes taken through a google doc during the workshop.~\cite{WSSSPE2-google-notes} 
-Overall, the report discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), a session on defining sustainability (\S\ref{sec:defining}), the five breakout sessions---exploring sustainability (\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit}); reproducibility \& reuse \& sharing (\S\ref{sec:reproduce}); code testing \& code review (\S\ref{sec:code_testing})---and some conclusions (\S\ref{sec:conclusions}).
+of notes taken through a google doc during the workshop~\cite{WSSSPE2-google-notes}.
+Overall, the report discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), a session on defining sustainability (\S\ref{sec:defining}), the five breakout sessions---exploring sustainability (\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit}); reproducibility, reuse, \& sharing (\S\ref{sec:reproduce}); code testing \& code review (\S\ref{sec:code_testing})---and some conclusions (\S\ref{sec:conclusions}).
 The report also includes an incomplete list of attendees (Appendix \ref{sec:attendees}).

### 2015-03-18 WSSSPE2_report/WSSSPE2_review.tex
Section 1: Warp lines.
no message

@@
 Conference for High Performance Computing, Networking, Storage and Analysis
-(SC14)\footnote{\url{http://sc14.supercomputing.org}}.
-WSSSPE2 followed the model of a general initial workshop,
+(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
+model of a general initial workshop,
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-which co-occurred with the SC13 conference, and a focused
-workshop,
+which co-occurred with the SC13 conference, and a focused workshop,
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
@@
 Progress in scientific research is dependent on the quality and accessibility of
-software at all levels. Hence it is critical to address challenges related to the
-development, deployment, maintenance, and overall sustainability of reusable
+software at all levels. Hence it is critical to address challenges related to
+the development, deployment, maintenance, and overall sustainability of reusable
 software as well as education around software practices. These challenges can be
@@
 engaged the broad scientific community to identify challenges and best practices
-in areas of interest to creating sustainable scientific software. WSSSPE2 invited the
-community to propose and discuss specific mechanisms to move towards an imagined
-future practice for software development and usage in science and engineering.
-The workshop included multiple mechanisms for participation, encouraged team
-building around solutions, and identified risky solutions with potentially
-transformative outcomes. It strongly encouraged participation of early-career
-scientists, postdoctoral researchers, and graduate students.
-
-This report extends a previous report that discussed the submission, peer-review, and peer-grouping processes in detail.~\cite{WSSSPE2-pre-report}.  It is also based on a collaborative set
-of notes taken through a google doc during the workshop~\cite{WSSSPE2-google-notes}.
-Overall, the report discusses the organization work done before the workshop (\S\ref{sec:preworkshop}), the keynotes (\S\ref{sec:keynotes}), a session on defining sustainability (\S\ref{sec:defining}), the five breakout sessions---exploring sustainability (\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit}); reproducibility, reuse, \& sharing (\S\ref{sec:reproduce}); code testing \& code review (\S\ref{sec:code_testing})---and some conclusions (\S\ref{sec:conclusions}).
-The report also includes an incomplete list of attendees (Appendix \ref{sec:attendees}).
+in areas of interest to creating sustainable scientific software. WSSSPE2
+invited the community to propose and discuss specific mechanisms to move towards
+an imagined future practice for software development and usage in science and
+engineering. The workshop included multiple mechanisms for participation,
+encouraged team building around solutions, and identified risky solutions with
+potentially transformative outcomes. It strongly encouraged participation of
+early-career scientists, postdoctoral researchers, and graduate students.
+
+This report extends a previous report that discussed the submission,
+peer-review, and peer-grouping processes in detail~\cite{WSSSPE2-pre-report}.
+It is also based on a collaborative set of notes taken through a google doc
+during the workshop~\cite{WSSSPE2-google-notes}. Overall, the report discusses
+the organization work done before the workshop (\S\ref{sec:preworkshop}); the
+keynotes (\S\ref{sec:keynotes}); a session on defining sustainability
+(\S\ref{sec:defining}); five breakout sessions that explored sustainability
+(\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel});
+credit \& incentives (\S\ref{sec:credit}); reproducibility, reuse, \& sharing
+(\S\ref{sec:reproduce}); code testing \& code review
+(\S\ref{sec:code_testing}); and some conclusions (\S\ref{sec:conclusions}). The
+report also includes an incomplete list of attendees (Appendix~\ref{sec:attendees}).
 

### 2015-03-18 WSSSPE2_report/WSSSPE2_review.tex
Abstract: refine.

@@
 \todo{update abstract}
-This technical report discusses the submission and peer-review process used by
+This technical report records and discusses the alternative submission and peer-review process used by
 the Second Workshop on Sustainable Software for Science: Practice and
 Experiences (WSSSPE2) and the results of that process. It is intended to record
-both the alternative submission and program organization model used by WSSSPE2
-as well as the papers associated with the workshop that resulted from that
-process.
+both the program-organization model adapted by WSSSPE2
+as well as the papers accepted by the workshop.
 \end{abstract}

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Absract: line breaks.
Abstract: change one word.
updated abstract

@@
 \begin{abstract}      
-\todo{update abstract}
-This technical report records and discusses the alternative submission and peer-review process used by
-the Second Workshop on Sustainable Software for Science: Practice and
-Experiences (WSSSPE2) and the results of that process. It is intended to record
-both the program-organization model adapted by WSSSPE2
-as well as the papers accepted by the workshop.
+This technical report records and discusses the Second Workshop on Sustainable Software for
+Science: Practice and Experiences (WSSSPE2).  The workshop used an alternative submission
+and peer-review process, which led to a set of papers divided across five topic areas:
+exploring sustainability;
+software development experiences;
+credit \& incentives;
+reproducibility \& reuse \& sharing; and
+code testing \& code review.
+The report includes a description of the submission and review process, the workshop
+keynote presentations,
+a discussion on sustainability, and the five discussions from the topic areas.  The report also
+contains a description of potential actions that were proposed in the the five discussions.
 \end{abstract}

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
adding ack section

@@
 authors, and may be submitted to peer-reviewed journals.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgments} \label{sec:acks}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{feel free to add stuff here}
+
+Work by Katz was supported by the National
+Science Foundation while working at the Foundation.  Any opinion, finding, and
+conclusions or recommendations expressed in this material are those of the
+author(s) and do not necessarily reflect the views of the National Science
+Foundation.
 

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Section 1: minor changes.

@@
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-which co-occurred with the SC13 conference, and a focused workshop,
+which co-occurred with SC13, and a focused workshop,
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
-which was organized in 2014 jointly with the SciPy conference.
+which was organized in July 2014 jointly with the SciPy 
+conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.
 

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Section 2: touch up.

@@
 \item governance, business, and sustainability models
-\item the role of community software repositories, their operation and
-sustainability
+\item the role, operation, and
+sustainability of community software repositories 
 \item reproducibility, transparency needs that may be unique to science
@@
 allowed reviewers to provide scores on relevance and comments to the organizers,
-which were used to decide which papers to associate with the workshop, and
-comments to the authors, which were provided back to the authors to allow them
-to improve their papers.
+who used the information to decide which papers to associate with the workshop, and
+provided the comments to the authors to help them improve their papers.
 
@@
 and to use an open process to determine the breakout topics. Specifically,
-well-sorted\footnote{\url{http://www.well-sorted.org}} was used as follows:
+Well Sorted\footnote{\url{http://www.well-sorted.org}} was used in the following steps:
 \begin{enumerate}
-\item Authors were asked to create well-sorted ``cards'' for the papers. These
+\item Authors were asked to create Well Sorted ``cards'' for the papers. These
 cards have a title (50 characters maximum) and a body (255 characters maximum).
@@
 groups. A group can have as many cards as the person wants it to have, and it
-can have whatever meaning makes sense to that person.
-\item Well-sorted then produces a set of averages of all the sorts, with various
-numbers of clusters of cards.
+can have whatever meaning that makes sense to that person.
+\item Well Sorted  produces a set of averages of all the sorts, with various
+numbers of card clusters.
 \end{enumerate}
@@
 The organizers then chose the sort that contained five groups as the one that
-felt most meaningful, then decided on names for the five groups, namely:
+felt most meaningful, and decided on names for the five groups, namely:
 \begin{itemize}
@@
 \item Reproducibility \& Reuse \& Sharing
-\item Code Testing \& Code Review
+\item Code Testing \& Code Review.
 \end{itemize}
@@
 Finally, since some of the papers were not represented by cards in the process,
-they were not placed in groups by the peer-grouping system, so the authors of
-these papers were asked which group seemed the best for their papers, and those
+they were not placed in groups by the peer-grouping system; the authors of
+these papers were asked which groups seemed the best for their papers---these
 papers were then placed in those groups. Sections~\ref{sec:exploring}-\ref{sec:code_testing}

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Section 2: touch up.

@@
 
-The organizers then chose the sort that contained five groups as the one that
-felt most meaningful, and decided on names for the five groups, namely:
+The organizers then chose a sort that contained five groups and that
+felt most meaningful. After that, they decided on themes for the five groups, namely:
 \begin{itemize}

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Section 2: line breaks.
Section 3: intro : add a few words.

@@
 
-WSSSPE2 began with a call for papers.
-Based on the goal of encouraging a wide range of submissions from those involved
-in software practice, ranging from initial thoughts and partial studies to
-mature deployments, but focusing on papers that are intended to lead to changes,
-the organizers wanted to make submission as easy as possible. The call for
-papers stated:
-
-\begin{quote}
-We invite short (4-page) \textbf{actionable} papers that will lead to
-improvements for sustainable software science. These papers could be a call to
-action, or could provide position or experience reports on sustainable software
-activities. The papers will be used by the organizing committee to design
-sessions that will be highly interactive and targeted towards facilitating
-action. Submitted papers should be archived by a third-party service that
-provides DOIs. We encourage submitters to license their papers under a Creative
-Commons license that encourages sharing and remixing, as we will combine ideas
-(with attribution) into the outcomes of the workshop.
+WSSSPE2 began with a call for papers. Based on the goal of encouraging a wide
+range of submissions from those involved in software practice, ranging from
+initial thoughts and partial studies to mature deployments, but focusing on
+papers that are intended to lead to changes, the organizers wanted to make
+submission as easy as possible. The call for papers stated:
+
+\begin{quote} We invite short (4-page) \textbf{actionable} papers that will lead
+to improvements for sustainable software science. These papers could be a call
+to action, or could provide position or experience reports on sustainable
+software activities. The papers will be used by the organizing committee to
+design sessions that will be highly interactive and targeted towards
+facilitating action. Submitted papers should be archived by a third-party
+service that provides DOIs. We encourage submitters to license their papers
+under a Creative Commons license that encourages sharing and remixing, as we
+will combine ideas (with attribution) into the outcomes of the workshop.
 \end{quote}
@@
 The review process was fairly standard, where reviewers bid for papers, then an
-automated system
-matched bids to determine assignments, and reviewers then completed their
-assigned
-reviews (with an average of 4.9 reviews per paper, and 4.1 reviews per
-reviewer).
-This process was done through EasyChair\footnote{http://easychair.org/}, which
-allowed reviewers to provide scores on relevance and comments to the organizers,
-who used the information to decide which papers to associate with the workshop, and
-provided the comments to the authors to help them improve their papers.
+automated system matched bids to determine assignments, and reviewers then
+completed their assigned reviews (with an average of 4.9 reviews per paper, and
+4.1 reviews per reviewer). This process was done through
+EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide
+scores on relevance and comments to the organizers, who used the information to
+decide which papers to associate with the workshop, and provided the comments to
+the authors to help them improve their papers.
 
@@
 to do this, the organizers decided to use a breakout format for two sessions,
-and to use an open process to determine the breakout topics. Specifically,
-Well Sorted\footnote{\url{http://www.well-sorted.org}} was used in the following steps:
+and to use an open process to determine the breakout topics. Specifically, Well
+Sorted\footnote{\url{http://www.well-sorted.org}} was used in the following
+steps:
 \begin{enumerate}
@@
 
-The organizers then chose a sort that contained five groups and that
-felt most meaningful. After that, they decided on themes for the five groups, namely:
+The organizers then chose a sort that contained five groups and that felt most
+meaningful. After that, they decided on themes for the five groups, namely:
 \begin{itemize}
@@
 organization's work and policy to enable and support sustainable and
-reproducible scientific research. Our second keynote speaker was Neil Chue Hong
+reproducible scientific research through the open web. Our second keynote speaker was Neil Chue Hong
 of Software Sustainability Institute. He shined the light on how scientific

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
adding section of 7

@@
 
-\note{for reference notes are in \url{http://tinyurl.com/l4tc6a4}}
+\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
 %https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
@@
 \subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
-\note{writing in progress by Frank Seinstra - expected by 22 March 2015}
-
-sub-breakout: \url{http://tinyurl.com/mx58tmx}
+\note{by Frank Seinstra}
+
+\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
 %https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
+
+At the start of the discussion, the breakout group brought forth the
+important observation of the wide disparity in commonly accepted habits of
+publication in different research fields.
+In domains which have, historically, relied on large groups of researchers
+collaborating towards a common goal (e.g., high-energy physics,
+astronomy), publications often have tens or even hundreds of co-authors (with
+some papers in experimental particle physics having over 3000.)
+In other domains, the number of co-authors is typically much smaller,
+with, in some cases, even a preference for single-author papers.
+Similarly, the various platforms for publication are valued differently in
+different domains.
+Most commonly, publications in peer-reviewed scientific journals are regarded
+as the most important and most impactful.
+However, in certain domains, especially in Computer Science, many researchers
+typically regard conference proceedings as their prime publication target.
+It is often suggested that this difference is due to the rapid developments
+in ICT science, a pace that can not be upheld by traditional peer-reviewed
+journals.
+Whatever the causes, any useful taxonomy of contributorship or guideline for
+tenure review should take such differences into account.
+
+Despite these differences, and despite the fact that software often
+has taken the role of a proper, albeit less tangible, scientific research
+instrument, neither the software nor its creators are commonly credited as part
+of a scientific publication.
+The group acknowledged the need for more recognition for the creators of such
+software instruments, and indicated a number of possible pathways.
+First and foremost, domain scientists must be made aware of the important role
+of software, and include the developers as co-authors of papers.
+A second approach is to fully embrace an open badging infrastructure (such as
+Mozilla's Open Badges), where a {\em badge\/} is a free, transferrable, 
+evidence-based indicator of an accomplishment, skill, quality, or interest.
+A third approach is for the scientific community to support the increasing
+momentum of peer-reviewed journals specialized in the open source/open
+access publication of scientific research software, such as Computer Physics
+Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.
+
+Recognizing publication of research software as a proper scientific
+contribution raises several important but currently unsolved questions,
+however.
+For example, is the number of users of the software a relevant measure of
+impact?
+What standards of coding quality must be followed in order to justify
+publication and hence recognition?
+Should the release of a new version of the software be eligible for a new
+publication; if so, under what conditions?
+And above all: should software publications be valued in the same way as
+traditional scientific publications?
+Or is there a need for new measures of productivity and impact?
+
+In part, the answers will come from the scientific community at large, as a
+natural consequence of growing awareness and mindset change.
+Some of the answers, however, also should be based on decades of experience
+in (and developing standards for) implementing, maintaining, refactoring,
+documenting, testing, and deploying software instruments in scientific
+research.
+Care should be taken, however, not to impose such standards for all domains
+in equal ways right from the start.
+Forerunners should serve as an example, but should not scare away domains that
+have based their progress on much less advanced methods of {\em software
+carpentry}.
+Nevertheless, proper guidelines are needed, which eventually should be followed
+across all domains.
+The group also recognized that funding bodies, universities, and publishers
+eventually should demand that research projects follow such guidelines, and to
+implement a proper software sustainability plan.
+
+To enable a form of standardized crediting for developers of research
+software, the group proposed to work towards a taxonomy for software-based
+contributorship.
+The taxonomy should be derived from, or extend, existing taxonomies for
+research impact and contributorship such as defined by CASRAI (in particular
+based on the Wellcome-Harvard contributorship taxonomy\katznote{Dan to add cite or url}), VIVO, or ISNI.
+An interesting measure of impact raised by the group was the {\em betweenness
+centrality}, an indicator of a person's centrality (and hence, importance)
+in a scientific collaboration.
+It is expected that developers of research software often play such a central
+role.
+
+The group defined the following actions to be performed:
+\begin{enumerate}
+\item Investigate existing taxonomies for roles and contributorships.
+\item Investigate prototype badging initiatives.
+\item Investigate journals focusing on publishing peer-reviewed research
+software.
+\item Investigate guidelines and checklists of best practices.
+\item Communicate the results of the above investigations to the WSSSPE
+community, and to decision-making bodies (funders, publishers, universities,
+and tenure committee representatives).
+\item Ensure engagement of the broader research community in this discussion.
+\end{enumerate}
 

### 2015-03-19 WSSSPE2_report/WSSSPE2_review.tex
Section 3.1: First draft.

@@
 \usepackage{longtable}
+\usepackage{verbatim}
 
@@
 \newcommand{\colinnote}[1]{ {\textcolor{red}    {***Colin #1 }}}
+\newcommand{\choinote}[1]{ {\textcolor{orange}    {***Choi #1 }}}
 
@@
 organization's work and policy to enable and support sustainable and
-reproducible scientific research through the open web. Our second keynote speaker was Neil Chue Hong
-of Software Sustainability Institute. He shined the light on how scientific
-software is prevalently driving advances in many science and engineering fields.
-Both keynote speeches spawned further discussion among workshop participants on
-the crucial notion ``sustainability'' in the theme of our workshop.
+reproducible scientific research through the open web. Our second keynote
+speaker was Neil Chue Hong of Software Sustainability Institute. He shined the
+light on how scientific software is prevalently driving advances in many science
+and engineering fields. Both keynote speeches spawned further discussion among
+workshop participants on the crucial notion ``sustainability'' in the theme of
+our workshop.
 
@@
 
-Thaney started noting the unfortunate fact that many current systems have been designed to create friction hindering us, despite the original intentions.
-\katznote{prev sentence is confusing}%---social issues hindering us
-An example is the National Cancer Institute's caBIG. 
-A total of \$350m was spent, including more than \$60m for management.  $70+$ tools were created, but caBIG is still seen
-as a failure (Kaitlin didn't really say why? Were the tools used? Did they
-immediately bit-rot? Was their development unsustained?)-should ask question
-about this.
- Those that had the least investment were the most used, the largest
-investments were the least used.
- Is there a caBIG report available? Quotes from Andrea Califano/Joe Gray,
-looks like there is:
-\url{http://tinyurl.com/maf6dz2}. %http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068?
-
-What do we mean by open research? In regard to
-community/technology/practices
-
-Inefficiency: \url{http://www.xkcd.com/1445/}. Need research tools (e.g., software repositories), social capital (e.g.,
-incentives) and capacity (e.g., training and mentorship) for efficient reproducible
-research.
-
-Our systems need to talk to each other but Q for speaker is that as
-systems become less monolithic, paradoxically it may be harder to sustain the
-links between them see:
-\url{http://tinyurl.com/l76tba2} %http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
-(but does Anon Grizzly have other cites/info for that?).
-Note, for example, that creative commons licensed works typically do not 
+Thaney started noting the unfortunate fact that many current systems suffer the
+unintended consequence of creating friction that hinders users, despite
+designers' original purposes to do good. An example is the National Cancer
+Institute's caBIG. A total of \$350 million was spent, including more than \$60
+million for management. More than $70$ tools were created, but caBIG is still
+seen as a failure\footnote{\url{http://tinyurl.com/maf6dz2}}. Those that had the
+least investment were the most used; the most invested software were the least
+utilized.
+\begin{comment}
+(Kaitlin didn't really say why? Were the tools used? Did they immediately
+bit-rot? Was their development unsustained?)-should ask question about this. Is
+there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
+there is: \url{tinyurl.com/qdodbo7}.
+%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068
+
+What do we mean by open research? In regard to community/technology/practices
+
+Inefficiency cartoon:  \url{http://www.xkcd.com/1445/}. 
+\end{comment}
+
+Thaney emphasized that for efficient reproducible open research, we would need
+research tools (e.g., software repositories), social capital (e.g., incentives),
+and capacity (e.g., training and mentorship). Our systems would need to
+communicate with each other. A point was made by a member of the auidence that
+as systems become less monolithic, it often becomes harder to sustain the links between
+them\footnote{See, for example, \url{http://tinyurl.com/l76tba2}.}.
+%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
+%(but does Anon Grizzly have other cites/info for that?).
+In contrast, works licensed by Creative Commons typically do not 
 have dependencies (a book, a photo, an artwork).
-Software Discovery Index: \url{http://softwarediscoveryindex.org/report/}.
-
- Final points:
-\begin{itemize}
-\item Design with the audience not the individual in mind
-\item Design to unlock latent potential of our systems
-\item Rethink how we reward researchers and support roles
-\item By mindful of jargon/semantics traps
-\end{itemize}
-
-Questions:
-\begin{itemize}
-\item Ted Habermann: if the metadata is minimal less onerous for data provider,
-but more onerous for user. How to work with data providers to show that users
-need more.
-\item Find balance between things that have to be captured and what would be
-ideal. Don't want to get to point where like old Harvard Dataverse, three pages
-of metadata capture, only first four things ever filled out. Don't want to lose
-engagement.
-\end{itemize}
+
+Thaney introduced Mozilla's Software Discovery
+Index\footnote{\url{http://softwarediscoveryindex.org/report/}}, which
+identifies scientific software by archiving and standardizing metadata for
+software and hence help connect both developers and users. Ted Habermann in the
+audience commented that if the metadata is minimal, it would be less onerous for
+data providers, but more burdensome for users---it could be challenging to keep
+a balance between what have to be captured and what would be ideal if we do not
+want to lose user engagement as in the case of the old Harvard Dataverse, finding often
+only the first four fields of three pages of metadata were filled out.
+
+The speaker concluded her talk urging the audience to design scientific software
+with the general community, not an individual, in mind; and to design to unlock
+latent potential of our systems. In addition, she encouraged everyone to rethink
+how we reward researchers and support roles. Lastly, she cautioned the community
+to be mindful of jargon or semantics traps.
 

### 2015-03-21 WSSSPE2_report/WSSSPE2_review.tex
minor changes in 3

@@
 organization's work and policy to enable and support sustainable and
-reproducible scientific research through the open web. Our second keynote
-speaker was Neil Chue Hong of Software Sustainability Institute. He shined the
+reproducible scientific research through the open web. The second keynote
+speaker was Neil Chue Hong of Software Sustainability Institute. He shined a
 light on how scientific software is prevalently driving advances in many science

### 2015-03-21 WSSSPE2_report/WSSSPE2_review.tex
Section 7: Add line breaks. No change in content.
Section 7: Intro: add a word.

@@
 
-This group, with just three papers but a large amount of interest and participation from attendees,
-focused on the institutional, social, and cultural mechanisms encourage the creation
-and maintenance of shared software, in the context of what now exists but focusing
-on what mechanisms are desired and how we might achieve them.
+This group, with just three papers but a large amount of interest and
+participation from attendees, focused on the institutional, social, and cultural
+mechanisms that encourage the creation and maintenance of shared software, in the
+context of what now exists but focusing on what mechanisms are desired and how
+we might achieve them.
 
@@
 \subsubsection{First breakout discussion: Group A}
-The first sub-group discussed around the current system for credit and current incentives,
-which it called ``hacking the incentive structure''.  The group considered four potential
-points of leverage:
-
-First, that we currently have systems that collect information, and these could be modified
-to collect different information, then map that information to actions. We could
-initially build a proof-of-concept for a new use of a given system, then determine
-what actions would be needed to make this use more common.
+The first sub-group discussed around the current system for credit and current
+incentives, which it called ``hacking the incentive structure''. The group
+considered four potential points of leverage:
+
+First, that we currently have systems that collect information, and these could
+be modified to collect different information, then map that information to
+actions. We could initially build a proof-of-concept for a new use of a given
+system, then determine what actions would be needed to make this use more
+common.
 
@@
 
-Third that we could change academic culture, rather than worrying about the systems.
-This was mostly focused on citations, because they matter for hiring, promotion,
-and tenure decisions.  The group discussed how we could weight the citations
-within papers better than we now do? How could we identify the five citations
-that really matter for a paper, distinguishing them from the related works and
-general background that are also cited.  Perhaps we could break these out in
-the reference list, working with publishers to implement this.  Or maybe we
-could also break out categories of citations, such as the most important software used,
-the previous publication that we are building from, the data that we actually used, etc.
-The Moore Foundation's award in data science was given as an example, asking
-proposers: What are the five canonical citations that are most important to your work?
-This would be a way of giving credit and assigning importance to these works,
-differently from how we just count citations today.  A possible action that the group
-discussed was conducting a longitudinal study of most useful \{software, data, etc.\}
-in a discipline
+Third that we could change academic culture, rather than worrying about the
+systems. This was mostly focused on citations, because they matter for hiring,
+promotion, and tenure decisions. The group discussed how we could weight the
+citations within papers better than we now do? How could we identify the five
+citations that really matter for a paper, distinguishing them from the related
+works and general background that are also cited. Perhaps we could break these
+out in the reference list, working with publishers to implement this. Or maybe
+we could also break out categories of citations, such as the most important
+software used, the previous publication that we are building from, the data that
+we actually used, etc. The Moore Foundation's award in data science was given as
+an example, asking proposers: What are the five canonical citations that are
+most important to your work? This would be a way of giving credit and assigning
+importance to these works, differently from how we just count citations today. A
+possible action that the group discussed was conducting a longitudinal study of
+most useful \{software, data, etc.\} in a discipline
 
@@
 \begin{itemize}
-\item running programming contests, creating bounties for contributing to open source software, etc.
-\item augment author lists to give credit to people who don't now get credit (and making them machine readable)
-\item developing a microcitation standard and mechanism (for both software and data)
-\item developing a well-defined standard for author contribution -- what level of contribution rises to the level of authorship?
-\item leveraging social media for citation and reviewing of content -- then using social media to bring more people into the review process than is traditional
-\item determining where else software can be cited and recorded (e.g., acknowledgments sections of papers)
-\item developing a taxonomy of contributors, tied to places that these metrics are already stored (e.g., ORCID)
-\item making metadata more easy to add for software, creating incentive for providing software metadata -- note that this cannot be centralized
-\item creating something like the h-index that tenure committees can make use of - simplify a way of measuring and documenting the overall credit given to an individual over different projects
-\item thinking about publishing software vs journal - software does not have to be novel
-\item determining guidelines for recommending software characteristics for tenure - perhaps draft guidelines then get ACM or IEEE agreement
+\item running programming contests, creating bounties for contributing to open
+source software, etc.
+\item augment author lists to give credit to people who don't now get credit
+(and making them machine readable)
+\item developing a microcitation standard and mechanism (for both software and
+data)
+\item developing a well-defined standard for author contribution -- what level
+of contribution rises to the level of authorship?
+\item leveraging social media for citation and reviewing of content -- then
+using social media to bring more people into the review process than is
+traditional
+\item determining where else software can be cited and recorded (e.g.,
+acknowledgments sections of papers)
+\item developing a taxonomy of contributors, tied to places that these metrics
+are already stored (e.g., ORCID)
+\item making metadata more easy to add for software, creating incentive for
+providing software metadata -- note that this cannot be centralized
+\item creating something like the h-index that tenure committees can make use of
+- simplify a way of measuring and documenting the overall credit given to an
+individual over different projects
+\item thinking about publishing software vs journal - software does not have to
+be novel
+\item determining guidelines for recommending software characteristics for
+tenure - perhaps draft guidelines then get ACM or IEEE agreement
 \end{itemize}
@@
 
-This group started by discussing who should be incentivized, thinking of two categories of people:  
-those in science (who could be incentivized to do better, more shareable, more sustainable work), and
-those in industry but interested in science (who could be incentivized to contribute to science.)  It was pointed out that we are not yet clear enough on exactly what we want to incentivize, suggesting that we need to have a clearer picture of ``good computational work" and what sort of contributions are truly generative for science.
-
-The discussion of incentivizing those in science acknowledged that the publications system was far from perfect for incentivizing good software work.
-Nonetheless discussion focused on bringing software people into publications. There were two main suggestions.
-The first is to focus on end users of software and encourage them to cite software the ``right" way.
-James Howison suggested that his research showed that few projects were making a formal request for citation (but that authors weren't necessarily following those suggestions anyway) \cite{howison2015jasist} he suggested making access to the software conditional on a license that requires citation.
-Others found this ``too confrontational" and preferred to concentrate on making it easier to do the right thing.
-The second was focusing on those leading software projects, and the group was more enthusiastic about ``forcing" PIs to include their ``software people" on publications, although there were few ideas on how exactly to do this.
-Another technique mentioned was that when scientific software projects are hosted in organizations like Apache, the scientific contributors can benefit from building their reputations, perhaps yielding job offers that they can use to negotiate better job and career packages.
-
-The discussion on incentivizing those outside science focused on accessing the well of affection that those working in software have for scientific research.
-How can the interest and skills of this group be marshaled towards sustainable contributions?
-The is evidence that the migration of scientific software projects to the Apache Foundation has created opportunities for those not employed in the scientific center to contribute to projects initiated by scientists (especially where there is cross-over with industry needs, such as provenance and workflow).
-
-The group also discussed developing ``software prizes" arguing that while it is hard to ``mint" other new sources of reputation, prizes are possible without getting too many others on board.
-The prize criteria can form a template for describing what we mean by scientific contributions made through software, particularly focusing on building active communities, not only writing great code.
+This group started by discussing who should be incentivized, thinking of two
+categories of people: those in science (who could be incentivized to do better,
+more shareable, more sustainable work), and those in industry but interested in
+science (who could be incentivized to contribute to science.) It was pointed out
+that we are not yet clear enough on exactly what we want to incentivize,
+suggesting that we need to have a clearer picture of ``good computational work"
+and what sort of contributions are truly generative for science.
+
+The discussion of incentivizing those in science acknowledged that the
+publications system was far from perfect for incentivizing good software work.
+Nonetheless discussion focused on bringing software people into publications.
+There were two main suggestions. The first is to focus on end users of software
+and encourage them to cite software the ``right" way. James Howison suggested
+that his research showed that few projects were making a formal request for
+citation (but that authors weren't necessarily following those suggestions
+anyway) \cite{howison2015jasist} he suggested making access to the software
+conditional on a license that requires citation. Others found this ``too
+confrontational" and preferred to concentrate on making it easier to do the
+right thing. The second was focusing on those leading software projects, and the
+group was more enthusiastic about ``forcing" PIs to include their ``software
+people" on publications, although there were few ideas on how exactly to do
+this. Another technique mentioned was that when scientific software projects are
+hosted in organizations like Apache, the scientific contributors can benefit
+from building their reputations, perhaps yielding job offers that they can use
+to negotiate better job and career packages.
+
+The discussion on incentivizing those outside science focused on accessing the
+well of affection that those working in software have for scientific research.
+How can the interest and skills of this group be marshaled towards sustainable
+contributions? The is evidence that the migration of scientific software
+projects to the Apache Foundation has created opportunities for those not
+employed in the scientific center to contribute to projects initiated by
+scientists (especially where there is cross-over with industry needs, such as
+provenance and workflow).
+
+The group also discussed developing ``software prizes" arguing that while it is
+hard to ``mint" other new sources of reputation, prizes are possible without
+getting too many others on board. The prize criteria can form a template for
+describing what we mean by scientific contributions made through software,
+particularly focusing on building active communities, not only writing great
+code.
 
@@
 \item new metrics (for people's activities in software)
-\item guidelines for evaluating scientific contribution through software (perhaps using new metrics)
-\end{itemize}
-
-In the remaining discussion sessions, the group chose to split into three subgroups to discuss
-a version of these topics: {\em citation ecosystems},
-{\em taxonomy of contributorship/guidelines for software for tenure review}, and {\em prizes}.  The subgroups were asked to clearly identify
+\item guidelines for evaluating scientific contribution through software (perhaps 
+using new metrics)
+\end{itemize}
+
+In the remaining discussion sessions, the group chose to split into three
+subgroups to discuss a version of these topics: {\em citation ecosystems}, {\em
+taxonomy of contributorship/guidelines for software for tenure review}, and {\em
+prizes}. The subgroups were asked to clearly identify
 \begin{itemize}
@@
 
-This group defined its goal as creating a low-barrier-of-entry method for recording names and roles (and in a second phase, optionally including weights) of contributors (coders and other intellectual contributions) to a software package in a machine readable way (to be called a credit file), then encouraging the scientific community to adopt this practice.
+This group defined its goal as creating a low-barrier-of-entry method for
+recording names and roles (and in a second phase, optionally including weights)
+of contributors (coders and other intellectual contributions) to a software
+package in a machine readable way (to be called a credit file), then encouraging
+the scientific community to adopt this practice.
 
@@
 \begin{itemize}
-\item The FLASH project was suggested as an example of how something like this has been done.
-\item This data could be a file that can be associated with a citation to the software, either through use of a DOI for the credit file, or by uploading the credit files as associated with the paper.
+\item The FLASH project was suggested as an example of how something 
+like this has been done.
+\item This data could be a file that can be associated with a citation to the 
+software, either through use of a DOI for the credit file, or by uploading the 
+credit files as associated with the paper.
 \item This idea could also be applied to data
-\item This data (the credit file) should be part of the metadata that is outside of paywalls, like citations are now.
-\item It was suggested that there should also be a separate file to track software dependencies, but this is not the same file.
+\item This data (the credit file) should be part of the metadata that is outside
+of paywalls, like citations are now.
+\item It was suggested that there should also be a separate file to track
+software dependencies, but this is not the same file. 
 \end{itemize}
@@
 \begin{enumerate}
-\item Build a tool that can automatically determine who the contributors are (from a git or other repo), then allows the user to manually edit the output to add/remove people, define roles. 
-\item Work with repositories to encourage them to provide the information we need based on what they already store.
+\item Build a tool that can automatically determine who the contributors are 
+(from a git or other repo), then allows the user to manually edit the output to 
+add/remove people, define roles. 
+\item Work with repositories to encourage them to provide the information we 
+need based on what they already store.
 \item Define what a citation file should look like and what it should be called.
@@
 \item Build a validator (and perhaps a visualizer) for credit files.
-\item Write a tool to collect files and visualize/output interconnections (which software is used with which), based on an existing project.
-\item When we (the group members) write papers, we should track the software we use, and encourage the software developers to make their software citable and create credit files.
+\item Write a tool to collect files and visualize/output interconnections (which 
+software is used with which), based on an existing project.
+\item When we (the group members) write papers, we should track the software 
+we use, and encourage the software developers to make their software citable 
+and create credit files.
 \item Build a tool to export the credit file to BibTex and other citation styles.
-\item Make sure the BibTex entries (exported from internal data) are somewhat standardized so that they can be imported into papers.
-Also make sure that standard LaTex style files understand and accept these entries.
+\item Make sure the BibTex entries (exported from internal data) are somewhat 
+standardized so that they can be imported into papers.  Also make sure that 
+standard LaTex style files understand and accept these entries.
 \end{enumerate}
@@
 
-At the start of the discussion, the breakout group brought forth the
-important observation of the wide disparity in commonly accepted habits of
-publication in different research fields.
-In domains which have, historically, relied on large groups of researchers
-collaborating towards a common goal (e.g., high-energy physics,
-astronomy), publications often have tens or even hundreds of co-authors (with
-some papers in experimental particle physics having over 3000.)
-In other domains, the number of co-authors is typically much smaller,
-with, in some cases, even a preference for single-author papers.
-Similarly, the various platforms for publication are valued differently in
-different domains.
-Most commonly, publications in peer-reviewed scientific journals are regarded
-as the most important and most impactful.
-However, in certain domains, especially in Computer Science, many researchers
-typically regard conference proceedings as their prime publication target.
-It is often suggested that this difference is due to the rapid developments
-in ICT science, a pace that can not be upheld by traditional peer-reviewed
-journals.
-Whatever the causes, any useful taxonomy of contributorship or guideline for
-tenure review should take such differences into account.
-
-Despite these differences, and despite the fact that software often
-has taken the role of a proper, albeit less tangible, scientific research
-instrument, neither the software nor its creators are commonly credited as part
-of a scientific publication.
-The group acknowledged the need for more recognition for the creators of such
-software instruments, and indicated a number of possible pathways.
-First and foremost, domain scientists must be made aware of the important role
-of software, and include the developers as co-authors of papers.
+At the start of the discussion, the breakout group brought forth the important
+observation of the wide disparity in commonly accepted habits of publication in
+different research fields. In domains which have, historically, relied on large
+groups of researchers collaborating towards a common goal (e.g., high-energy
+physics, astronomy), publications often have tens or even hundreds of co-authors
+(with some papers in experimental particle physics having over 3000.) In other
+domains, the number of co-authors is typically much smaller, with, in some
+cases, even a preference for single-author papers. Similarly, the various
+platforms for publication are valued differently in different domains. Most
+commonly, publications in peer-reviewed scientific journals are regarded as the
+most important and most impactful. However, in certain domains, especially in
+Computer Science, many researchers typically regard conference proceedings as
+their prime publication target. It is often suggested that this difference is
+due to the rapid developments in ICT science, a pace that can not be upheld by
+traditional peer-reviewed journals. Whatever the causes, any useful taxonomy of
+contributorship or guideline for tenure review should take such differences into
+account.
+
+Despite these differences, and despite the fact that software often has taken
+the role of a proper, albeit less tangible, scientific research instrument,
+neither the software nor its creators are commonly credited as part of a
+scientific publication. The group acknowledged the need for more recognition for
+the creators of such software instruments, and indicated a number of possible
+pathways. First and foremost, domain scientists must be made aware of the
+important role of software, and include the developers as co-authors of papers.
 A second approach is to fully embrace an open badging infrastructure (such as
 Mozilla's Open Badges), where a {\em badge\/} is a free, transferrable,
-evidence-based indicator of an accomplishment, skill, quality, or interest.
-A third approach is for the scientific community to support the increasing
-momentum of peer-reviewed journals specialized in the open source/open
-access publication of scientific research software, such as Computer Physics
+evidence-based indicator of an accomplishment, skill, quality, or interest. A
+third approach is for the scientific community to support the increasing
+momentum of peer-reviewed journals specialized in the open source/open access
+publication of scientific research software, such as Computer Physics
 Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.
 
-Recognizing publication of research software as a proper scientific
-contribution raises several important but currently unsolved questions,
-however.
-For example, is the number of users of the software a relevant measure of
-impact?
-What standards of coding quality must be followed in order to justify
-publication and hence recognition?
-Should the release of a new version of the software be eligible for a new
-publication; if so, under what conditions?
-And above all: should software publications be valued in the same way as
-traditional scientific publications?
-Or is there a need for new measures of productivity and impact?
+Recognizing publication of research software as a proper scientific contribution
+raises several important but currently unsolved questions, however. For example,
+is the number of users of the software a relevant measure of impact? What
+standards of coding quality must be followed in order to justify publication and
+hence recognition? Should the release of a new version of the software be
+eligible for a new publication; if so, under what conditions? And above all:
+should software publications be valued in the same way as traditional scientific
+publications? Or is there a need for new measures of productivity and impact?
 
 In part, the answers will come from the scientific community at large, as a
-natural consequence of growing awareness and mindset change.
-Some of the answers, however, also should be based on decades of experience
-in (and developing standards for) implementing, maintaining, refactoring,
-documenting, testing, and deploying software instruments in scientific
-research.
-Care should be taken, however, not to impose such standards for all domains
-in equal ways right from the start.
-Forerunners should serve as an example, but should not scare away domains that
-have based their progress on much less advanced methods of {\em software
-carpentry}.
-Nevertheless, proper guidelines are needed, which eventually should be followed
-across all domains.
-The group also recognized that funding bodies, universities, and publishers
-eventually should demand that research projects follow such guidelines, and to
-implement a proper software sustainability plan.
-
-To enable a form of standardized crediting for developers of research
-software, the group proposed to work towards a taxonomy for software-based
-contributorship.
-The taxonomy should be derived from, or extend, existing taxonomies for
-research impact and contributorship such as defined by CASRAI (in particular
-based on the Wellcome-Harvard contributorship taxonomy\katznote{Dan to add cite or url}), VIVO, or ISNI.
-An interesting measure of impact raised by the group was the {\em betweenness
-centrality}, an indicator of a person's centrality (and hence, importance)
-in a scientific collaboration.
-It is expected that developers of research software often play such a central
-role.
+natural consequence of growing awareness and mindset change. Some of the
+answers, however, also should be based on decades of experience in (and
+developing standards for) implementing, maintaining, refactoring, documenting,
+testing, and deploying software instruments in scientific research. Care should
+be taken, however, not to impose such standards for all domains in equal ways
+right from the start. Forerunners should serve as an example, but should not
+scare away domains that have based their progress on much less advanced methods
+of {\em software carpentry}. Nevertheless, proper guidelines are needed, which
+eventually should be followed across all domains. The group also recognized that
+funding bodies, universities, and publishers eventually should demand that
+research projects follow such guidelines, and to implement a proper software
+sustainability plan.
+
+To enable a form of standardized crediting for developers of research software,
+the group proposed to work towards a taxonomy for software-based
+contributorship. The taxonomy should be derived from, or extend, existing
+taxonomies for research impact and contributorship such as defined by CASRAI (in
+particular based on the Wellcome-Harvard contributorship taxonomy\katznote{Dan
+to add cite or url}), VIVO, or ISNI. An interesting measure of impact raised by
+the group was the {\em betweenness centrality}, an indicator of a person's
+centrality (and hence, importance) in a scientific collaboration. It is expected
+that developers of research software often play such a central role.
 
@@
 \item Converting reputation by joining apache software foundation, Google etc. 
-\item Inviting ASF and open source people to contribute to scientific code (at least on questions of overlapping
+\item Inviting ASF and open source people to contribute to scientific code (at
+least on questions of overlapping
 \item Template for assessing scientific contributions made through software.   
@@
 
-Possible awarders are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology Society of America, etc.
-Or perhaps this should be a joint technology/science partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary society] prize?
+Possible awarders are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
+Society of America, etc. Or perhaps this should be a joint technology/science
+partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
+society] prize?
 
@@
 adding new pieces of functionality to software;
-integrating software into broader ecosystem; championing broad principles of sustainability, open science, open source, etc.;
-improving accessibility to software, to scientific software (perhaps championing inclusiveness or making software accessible?);
+integrating software into broader ecosystem; championing broad principles of
+sustainability, open science, open source, etc.;
+improving accessibility to software, to scientific software (perhaps championing
+inclusiveness or making software accessible?);
 documentation;
@@
 
-Perhaps there should be different criteria for ``established'' members of the community versus junior members?  Or perhaps prizes should be restricted to junior members?
-
-An important point is that we don't want to give prizes just to reward people who are really good at this one thing, but rather we want to reward people who are building the culture we want as scientists.
+Perhaps there should be different criteria for ``established'' members of the
+community versus junior members? Or perhaps prizes should be restricted to
+junior members?
+
+An important point is that we don't want to give prizes just to reward people
+who are really good at this one thing, but rather we want to reward people who
+are building the culture we want as scientists.
 

### 2015-03-21 WSSSPE2_report/WSSSPE2_review.tex
Section 7.1: minor changes.

@@
 \newcommand{\davidnote}[1]{  {\textcolor{darkgreen}      { ***David: #1 }}}
-\newcommand{\colinnote}[1]{ {\textcolor{red}    {***Colin #1 }}}
-\newcommand{\choinote}[1]{ {\textcolor{orange}    {***Choi #1 }}}
+\newcommand{\colinnote}[1]{ {\textcolor{red}    {***Colin: #1 }}}
+\newcommand{\choinote}[1]{ {\textcolor{orange}    {***Choi: #1 }}}
 
@@
 \subsubsection{First breakout discussion: Group A}
-The first sub-group discussed around the current system for credit and current
-incentives, which it called ``hacking the incentive structure''. The group
+The first sub-group discussed about the current system for credit and current
+incentives, which it called ``hacking the incentive structure.'' The group
 considered four potential points of leverage:
@@
 systems. This was mostly focused on citations, because they matter for hiring,
-promotion, and tenure decisions. The group discussed how we could weight the
+promotion, and tenure decisions. The group discussed how we could weigh the
 citations within papers better than we now do? How could we identify the five
@@
 software used, the previous publication that we are building from, the data that
-we actually used, etc. The Moore Foundation's award in data science was given as
+we actually used, etc. The Moore Foundation's award in data science \choinote{cite} was given as
 an example, asking proposers: What are the five canonical citations that are
@@
 possible action that the group discussed was conducting a longitudinal study of
-most useful \{software, data, etc.\} in a discipline
+most useful \{software, data, etc.\} in a discipline.
 
@@
 source software, etc.
-\item augment author lists to give credit to people who don't now get credit
+\item augment author lists to give credit to people who do not now get credit
 (and making them machine readable)
@@
 \item developing a taxonomy of contributors, tied to places that these metrics
-are already stored (e.g., ORCID)
-\item making metadata more easy to add for software, creating incentive for
+are already stored (e.g., ORCID) \choinote{cite}
+\item making metadata easier to add for software, creating incentive for
 providing software metadata -- note that this cannot be centralized
@@
 individual over different projects
-\item thinking about publishing software vs journal - software does not have to
+\item thinking about publishing software versus journal - software does not have to
 be novel

### 2015-03-22 WSSSPE2_report/WSSSPE2_review.tex
Section 7: More minor changes.

@@
 that his research showed that few projects were making a formal request for
-citation (but that authors weren't necessarily following those suggestions
-anyway) \cite{howison2015jasist} he suggested making access to the software
+citation (but that authors were not necessarily following those suggestions
+anyway) \cite{howison2015jasist}. He suggested making access to the software
 conditional on a license that requires citation. Others found this ``too
@@
 How can the interest and skills of this group be marshaled towards sustainable
-contributions? The is evidence that the migration of scientific software
+contributions? There is evidence that the migration of scientific software
 projects to the Apache Foundation has created opportunities for those not
@@
 \item the problem to be solved
-\item steps towards a solution
+\item steps towards a solution.
 \end{itemize}
@@
 \begin{itemize}
-\item The FLASH project was suggested as an example of how something 
+\item The FLASH \choinote{cite} project was suggested as an example of how something 
 like this has been done.
@@
 standardized so that they can be imported into papers.  Also make sure that 
-standard LaTex style files understand and accept these entries.
+standard \LaTeX style files understand and accept these entries.
 \end{enumerate}

### 2015-03-22 WSSSPE2_report/WSSSPE2_review.tex
undoing a few edits, most are good, will add requested cites later

@@
 \subsubsection{First breakout discussion: Group A}
-The first sub-group discussed about the current system for credit and current
+The first sub-group discussed issues around the current system for credit and current
 incentives, which it called ``hacking the incentive structure.'' The group
@@
 source software, etc.
-\item augment author lists to give credit to people who do not now get credit
+\item augment author lists to give credit to people who don't now get credit
 (and making them machine readable)
@@
 that his research showed that few projects were making a formal request for
-citation (but that authors were not necessarily following those suggestions
+citation (but that authors weren't necessarily following those suggestions
 anyway) \cite{howison2015jasist}. He suggested making access to the software
@@
 standardized so that they can be imported into papers.  Also make sure that 
-standard \LaTeX style files understand and accept these entries.
+standard LaTeX style files understand and accept these entries.
 \end{enumerate}

### 2015-03-22 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: Add some content.

@@
 \todo{Slides: \url{http://tinyurl.com/pd9tmh3}}
+
+In a recent survey of UK research-intensive universities, 92\% of researchers
+said they used research software and 68\% said their research would be impossible
+without software. Yet 71\% have had no formal software training, and few are
+ready to apply many of the things we take for granted such as testing or
+virtualisation. WSSSPE represents the pinnacle of what we understand to be the
+best practice around scientific software in our community. My talk will
+challenge the workshop participants to come up with ways of taking this best
+practice to those 92\% of researchers in a way that will lead to maximum benefit
+to the scientific community.
+
+%\url{http://dx.doi.org/10.6084/m9.ﬁgshare.1243288}
+
+Where indicated slides licensed under World-leading research relies on software
+
+``Today there are very few science areas left which do not rely on IT and thus
+software for the majority of their research work. More importantly key
+scientific advances in experimental and observational science would have been
+impossible without better software.'' Kersten Kleese van Dam Pacific Northwestern
+National Laboratory via change.org campaign
+
+``Scientific discovery and innovation are advancing along fundamentally new
+pathways opened by development of increasingly sophisticated software. Software
+is an integral enabler of computation, experiment and theory, and directly
+responsible for increased scientific productivity and enhancement of
+researchers' capabilities.'' Dan Katz SI2 Program Director, National Science
+Foundation.
+
+We do not recognise these hundreds of thousands of researchers because we think
+they are the long tail and forget that they are the mainstream and we are the
+elite minority
+
+Software is not special it is mainstream software?
+
+What would happen to your research without software
+
+Survey of researchers from 15 Russell Group unis conducted by SSI between Aug- Oct 2014.
+406 respondents covering representative range of funders, discipline and seniority.
+
+This is not just about the ``traditional'' sciences
+
+And 	attitudes 	are	changing
+
+Overall
+
+ which 	is	changing 	the	situation
+
+PhD students
+
+Survey of  PhD students from Russell Group university conducted by SSI between Aug - Sep 2014.
+~100 respondents covering representative range of discipline.
+
+Even	though 	a	lot	is	still	the	same
+
+Survey of 138 University professionals conducted by SSI in Feb-Apr 2013.
+
+The cost of UK research that relies on software is 
+£840m!
+
+Investment in 2013-2014 financial year, an amount that has risen by
+3\% on average over last four years
+
+30\%
+
+Of total research investment has been spent on research which relies on software
+over the last four financial years
+
+Analysis of data from 49,650 grant titles and abstracts published on Gateway to
+Research covering 2010-2014.
+
+The UK research community is just starting to understand the magnitude of the issue
+
+56\%
+Of UK researchers develop their
+own research software
+
+71\%
+Of UK researchers have had no formal
+software development training
+
+140,000
+UK researchers are relying on
+their own coding skills
+
+4\%
+Of jobs advertised in UK
+universities were software related
+
+Survey of researchers from 15 Russell Group unis conducted by SSI between Aug - Oct 2014.
+
+If 92\% of researchers care about software and WSSSPE is about caring about
+software---what are we going to do to help researchers?
+
+Sustainability
+
+With apologies to Greg Wilson:
+\url{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
+
+We	can 	talk	all	day 	about 	the best 	ways	of	achieving sustainability
+
+But	what	are	we	going 	to	do to	make	it	actually sustainable?
+
+Careers outside academic sector
+
+Non-university Research (industry, government etc.)
+
+UK	STEM graduate career paths
+
+Early Career
+Research
+
+Permanent
+Research Staff
+
+Professor
+
+\begin{verbatim}
+Source: The Scientific Century, Royal Society, 2010 (revised to reﬂect ﬁrst stage
+clariﬁcaFon from ``What Do PhD’s Do?'' study)
+\end{verbatim}
+
+Who	are	the	people 	we	need to	persuade?
+
+What		are	the	incentives 	we need 	to	put 	in	place?
+
+How are we going to show that the work we present here has benefit to the 92\%?
+
+We	are	all	the 92\%	and	we can	change	the world	if	we	want!
+
+\url{http://dx.doi.org/10.6084/m9.ﬁgshare.1243288}
 

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: temp fix for compilation errors.
Section 3: A handful minor changes.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{\footnotesize{Lead: Choi. Volunteers contact lead. Live notes: \href{http://tinyurl.com/q6ew45v}{Google Notes}}.
+\todo{\footnotesize{Lead: Choi. Live notes: \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 
-Thaney introduced Mozilla's Software Discovery
-Index\footnote{\url{http://softwarediscoveryindex.org/report/}}, which
-identifies scientific software by archiving and standardizing metadata for
-software and hence help connect both developers and users. Ted Habermann in the
-audience commented that if the metadata is minimal, it would be less onerous for
-data providers, but more burdensome for users---it could be challenging to keep
-a balance between what have to be captured and what would be ideal if we do not
-want to lose user engagement as in the case of the old Harvard Dataverse, finding often
-only the first four fields of three pages of metadata were filled out.
+Thaney introduced Mozilla's Software Discovery Index
+(SDI)\footnote{\url{http://softwarediscoveryindex.org/report/}}, which
+facilitates more efficient scientific research. SDI identifies scientific
+software by archiving and standardizing metadata for software and hence help
+connect both developers and users. Ted Habermann in the audience commented that
+if the metadata is minimal, it would be less onerous for data providers, but
+more burdensome for users---it could be challenging to keep a balance between
+what have to be captured and what would be ideal if we do not want to lose user
+engagement as in the case of the old Harvard Dataverse, finding often only the
+first four fields of three pages of metadata were filled out.
 
@@
 
-\todo{Slides: \url{http://tinyurl.com/pd9tmh3}}
-
 In a recent survey of UK research-intensive universities, 92\% of researchers
@@
 ready to apply many of the things we take for granted such as testing or
-virtualisation. WSSSPE represents the pinnacle of what we understand to be the
+virtualization. WSSSPE represents the pinnacle of what we understand to be the
 best practice around scientific software in our community. My talk will
@@
 they are the long tail and forget that they are the mainstream and we are the
-elite minority
+elite minority.
 
@@
 
-Survey of  PhD students from Russell Group university conducted by SSI between Aug - Sep 2014.
-~100 respondents covering representative range of discipline.
+Survey of  PhD students from Russell Group university conducted by SSI between Aug -- Sep 2014.
+approximately 100 respondents covering representative range of discipline.
 
@@
 
-Investment in 2013-2014 financial year, an amount that has risen by
+Investment in 2013--2014 financial year, an amount that has risen by
 3\% on average over last four years
@@
 
-56\%
-Of UK researchers develop their
-own research software
-
-71\%
-Of UK researchers have had no formal
-software development training
-
-140,000
-UK researchers are relying on
-their own coding skills
-
-4\%
-Of jobs advertised in UK
-universities were software related
+56\% Of UK researchers develop their own research software
+
+71\% Of UK researchers have had no formal software development training
+
+140,000 UK researchers are relying on their own coding skills
+
+4\% Of jobs advertised in UK universities were software related
 
@@
 
-\begin{verbatim}
+  \begin{comment}
 Source: The Scientific Century, Royal Society, 2010 (revised to reﬂect ﬁrst stage
 clariﬁcaFon from ``What Do PhD’s Do?'' study)
-\end{verbatim}
 
@@
 \url{http://dx.doi.org/10.6084/m9.ﬁgshare.1243288}
+\end{comment}
 

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: Minor changes.

@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
+\begin{comment}
 In a recent survey of UK research-intensive universities, 92\% of researchers
@@
 to the scientific community.
+ \end{comment}
 
@@
 
-Where indicated slides licensed under World-leading research relies on software
+World-leading research relies on software
 
@@
 
-Software is not special it is mainstream software?
+Software is not special; it is mainstream software
 
@@
 
-And 	attitudes 	are	changing
-
-Overall
-
- which 	is	changing 	the	situation
-
-PhD students
+And 	attitudes 	are	changing, which 	is	changing 	the	situation
 
@@
 
-The cost of UK research that relies on software is 
-£840m!
-
-Investment in 2013--2014 financial year, an amount that has risen by
+The cost of UK research that relies on software
+
+\textsterling 840 million of investment in 2013--2014 financial year, an amount that has risen by
 3\% on average over last four years
 
-30\%
-
-Of total research investment has been spent on research which relies on software
+30\% Of total research investment has been spent on research which relies on software
 over the last four financial years
@@
 We	can 	talk	all	day 	about 	the best 	ways	of	achieving sustainability
-
 But	what	are	we	going 	to	do to	make	it	actually sustainable?
 
-Careers outside academic sector
-
-Non-university Research (industry, government etc.)
-
-UK	STEM graduate career paths
-
-Early Career
-Research
-
-Permanent
-Research Staff
-
+Careers outside academic sector: 
+Non-university Research (industry, government, etc.)
+
+UK STEM graduate career paths: Early Career Research, Permanent Research Staff,
 Professor
@@
 Who	are	the	people 	we	need to	persuade?
-
 What		are	the	incentives 	we need 	to	put 	in	place?
-
 How are we going to show that the work we present here has benefit to the 92\%?

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Section 3.1: Changes due to feedback from Thaney.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes} \label{sec:keynotes}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{\footnotesize{Lead: Choi. Live notes: \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
+\section{Keynotes\todo{\footnotesize{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-Abstracts: \href{http://tinyurl.com/mnenzms}{WSSSPE2}. Slides: \href{http://tinyurl.com/las8tsq}{here}.}
+\href{http://tinyurl.com/mnenzms}{Abstracts}.  \href{http://tinyurl.com/las8tsq}{Slides}.}
+} \label{sec:keynotes}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 utilized.
+
 \begin{comment}
@@
 and capacity (e.g., training and mentorship). Our systems would need to
-communicate with each other. A point was made by a member of the auidence that
-as systems become less monolithic, it often becomes harder to sustain the links between
-them\footnote{See, for example, \url{http://tinyurl.com/l76tba2}.}.
+communicate with each other. A point was made by a member of the audience that
+as systems become less monolithic, it often becomes harder to sustain the links
+between them\footnote{See, for example, \url{http://tinyurl.com/l76tba2}.}.
 %http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
 %(but does Anon Grizzly have other cites/info for that?).
-In contrast, works licensed by Creative Commons typically do not 
-have dependencies (a book, a photo, an artwork).
-
-Thaney introduced Mozilla's Software Discovery Index
-(SDI)\footnote{\url{http://softwarediscoveryindex.org/report/}}, which
-facilitates more efficient scientific research. SDI identifies scientific
-software by archiving and standardizing metadata for software and hence help
-connect both developers and users. Ted Habermann in the audience commented that
-if the metadata is minimal, it would be less onerous for data providers, but
-more burdensome for users---it could be challenging to keep a balance between
-what have to be captured and what would be ideal if we do not want to lose user
-engagement as in the case of the old Harvard Dataverse, finding often only the
-first four fields of three pages of metadata were filled out.
+%In contrast, works licensed by Creative Commons typically do not 
+%have dependencies (a book, a photo, an artwork).
+
+Thaney spoke about Mozilla's work around code citation, through a collaboration
+and prototype crafted between Mozilla, Github, figshare and Zenodo. This work
+was presented at a closed meeting in May 2014 at the National Institutes of
+Health (NIH) around these issues, sparking a conversation from that meeting
+around what a \emph{Software Discovery
+Index}\footnote{\url{http://softwarediscoveryindex.org/report/}} might look
+like. The meeting included a number of publishers, researchers, and those behind
+major scientific software efforts such as Bioconductor, Galaxy, and
+nanoHUB. 
+%facilitates more efficient scientific research. SDI identifies scientific
+%software by archiving and standardizing metadata for software and hence help
+%connect both developers and users.
+Ted Habermann in the audience commented that if the metadata is minimal, it
+would be less onerous for data providers, but more burdensome for users---it
+could be challenging to keep a balance between what have to be captured and what
+would be ideal if we do not want to lose user engagement as in the case of the
+old Harvard Dataverse, finding often only the first four fields of three pages
+of metadata were filled out.
 
@@
 latent potential of our systems. In addition, she encouraged everyone to rethink
-how we reward researchers and support roles. Lastly, she cautioned the community
-to be mindful of jargon or semantics traps.
+how we reward researchers and support roles.
+%Lastly, she cautioned the community to be mindful of jargon or semantics traps.
 
@@
 
-We do not recognise these hundreds of thousands of researchers because we think
+We do not recognize these hundreds of thousands of researchers because we think
 they are the long tail and forget that they are the mainstream and we are the

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Sections 1-3: compressed headers

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction}
+\section{Introduction \note{\href{http://tinyurl.com/q6ew45v}{Google Notes}}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Submissions, Peer-Review, and Peer-Grouping} \label{sec:preworkshop}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
-
-WSSSPE2 began with a call for papers. Based on the goal of encouraging a wide
+\section{Submissions, Peer-Review, and Peer-Grouping \todo{Shorten}} \label{sec:preworkshop}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the goal of encouraging a wide
 range of submissions from those involved in software practice, ranging from

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Section 3: tiny changes.
Section 7: minor changes.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes\todo{\footnotesize{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
+\section{Keynotes\note{\footnotesize{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 and engineering fields. Both keynote speeches spawned further discussion among
-workshop participants on the crucial notion ``sustainability'' in the theme of
+workshop participants on the crucial notion \emph{sustainability} in the theme of
 our workshop.
@@
 broadest diversity of contributions,
-best tutorials/documentation,
+best tutorials or documentation,
 best leadership transition (award ex leader and new leader),
@@
 
-\katznote{I don't understand this at all.  Can some explain it?}
+\katznote{I don't understand this at all.  Can someone explain it?}
 There are also other indirect forms of incentives
@@
 \begin{itemize}
-\item Converting reputation by joining apache software foundation, Google etc. 
+\item Converting reputation by joining the Apache Software Foundation, Google, etc. 
 \item Inviting ASF and open source people to contribute to scientific code (at

### 2015-03-23 WSSSPE2_report/WSSSPE2_review.tex
Sections 1-3: Restore notes in response to Dan's comment.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction \note{\href{http://tinyurl.com/q6ew45v}{Google Notes}}}
+\section{Introduction} \label{sec:intro}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Submissions, Peer-Review, and Peer-Grouping \todo{Shorten}} \label{sec:preworkshop}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Submissions, Peer-Review, and Peer-Grouping} \label{sec:preworkshop}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes\note{\footnotesize{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes} and \href{http://tinyurl.com/pd9tmh3}{here}}.
+\section{Keynotes} \label{sec:keynotes}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-\href{http://tinyurl.com/mnenzms}{Abstracts}.  \href{http://tinyurl.com/las8tsq}{Slides}.}
-} \label{sec:keynotes}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\href{http://tinyurl.com/mnenzms}{Abstracts and slides}}
 

### 2015-03-24 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: reshaping content. To be continued.

@@
 
-World-leading research relies on software
-
-``Today there are very few science areas left which do not rely on IT and thus
-software for the majority of their research work. More importantly key
-scientific advances in experimental and observational science would have been
-impossible without better software.'' Kersten Kleese van Dam Pacific Northwestern
-National Laboratory via change.org campaign
-
-``Scientific discovery and innovation are advancing along fundamentally new
-pathways opened by development of increasingly sophisticated software. Software
-is an integral enabler of computation, experiment and theory, and directly
-responsible for increased scientific productivity and enhancement of
-researchers' capabilities.'' Dan Katz SI2 Program Director, National Science
-Foundation.
-
-We do not recognize these hundreds of thousands of researchers because we think
-they are the long tail and forget that they are the mainstream and we are the
-elite minority.
-
-Software is not special; it is mainstream software
-
-What would happen to your research without software
-
-Survey of researchers from 15 Russell Group unis conducted by SSI between Aug- Oct 2014.
-406 respondents covering representative range of funders, discipline and seniority.
-
-This is not just about the ``traditional'' sciences
-
-And 	attitudes 	are	changing, which 	is	changing 	the	situation
-
-Survey of  PhD students from Russell Group university conducted by SSI between Aug -- Sep 2014.
-approximately 100 respondents covering representative range of discipline.
-
-Even	though 	a	lot	is	still	the	same
-
-Survey of 138 University professionals conducted by SSI in Feb-Apr 2013.
-
-The cost of UK research that relies on software
-
-\textsterling 840 million of investment in 2013--2014 financial year, an amount that has risen by
-3\% on average over last four years
-
-30\% Of total research investment has been spent on research which relies on software
-over the last four financial years
-
-Analysis of data from 49,650 grant titles and abstracts published on Gateway to
-Research covering 2010-2014.
-
-The UK research community is just starting to understand the magnitude of the issue
-
-56\% Of UK researchers develop their own research software
-
-71\% Of UK researchers have had no formal software development training
-
-140,000 UK researchers are relying on their own coding skills
-
-4\% Of jobs advertised in UK universities were software related
-
-Survey of researchers from 15 Russell Group unis conducted by SSI between Aug - Oct 2014.
-
-If 92\% of researchers care about software and WSSSPE is about caring about
-software---what are we going to do to help researchers?
-
-Sustainability
-
+Neil Chue Hong is currently serving as the Director of the Software
+Sustainability Institute (SSI), United Kingdom. Hong started with making the
+point that world-leading research relies on software, quoting Kersten Kleese Van
+Dam of the Pacific Northwestern National Laboratory via the campaign
+\href{http://change.org}{change.org}, ``\emph{Today there are very few science
+areas left which do not rely on IT and thus software for the majority of the
+their research work. More importantly key scientific advances in experimental
+and observational science would have been impossible without better software.}''
+He also cited Daniel Katz, Sustained Innovation (SI2) Program Director of the
+National Science Foundation, ``\emph{Scientific discovery and innovation are
+advancing along fundamentally new pathways opened by development of increasingly
+sophisticated software. Software is an integral enabler of computation,
+experiment and theory, and directly responsible for increased scientific
+productivity and enhancement of researchers' capabilities.}''
+
+Hong drew attention to the misconception that we too often disregard hundreds of
+thousands of software researchers thinking they are the long tail. We forget
+that actually they are the mainstream and we are the elite minority. He
+emphasized that software is no longer special; it is prevailing in scientific
+research. Survey of researchers from 15 Russell Group universities conducted by
+the SSI between Aug--Oct 2014 included a total of 406 respondents covering a
+representative range of funders, disciplines, and seniority: 92\% confirmed the
+use of research software such as MATLAB, SPSS, Excel, so on and so forth; 89\%
+affirmed that it would be impossible or difficult to conduct research without
+software. %Researchers often discover new software predominatly through
+%recommendation of colleagues or web searches.
+Nevertheless the British research community is just starting to understand the
+magnitude of the issue. In the aforementioned SSI survey, 56\% of  
+researchers develop their own research software; 71\%  have had
+no formal software -development training; 140,000 researchers rely on
+their own coding skills; 4\% of jobs advertised in British universities were software
+related.
+
+According to Hong, costs of software-reliant research in the United Kingdom include
+\textsterling 840 million of investment in the financial year 2013--2014, an
+amount that has risen by 3\% on average over the past four years. About 30\% of
+total research investment has been spent on research which relies on software
+over the last four financial years. These numbers stemmed from an analysis of
+data from $49,650$ grant titles and abstracts published on Gateway to Research
+covering years 2010--2014.
+
+Hong led the audience to ponder on the questions that if 92\% of researchers and WSSSPE
+care about scientific software, what are we going to \emph{do} to help researchers and 
+to achieve software sustainability?
+\begin{comment}
 With apologies to Greg Wilson:
-\url{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
-
+\href{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
+{You Keep Using That Word}
 We can talk all day about the best ways of achieving sustainability
 But what are we going to do to make it actually sustainable?
-
+\end{comment}
 Careers outside academic sector: 

### 2015-03-24 WSSSPE2_report/WSSSPE2_review.tex
Section 3.1: Use past tense.

@@
 magnitude of the issue. In the aforementioned SSI survey, 56\% of  
-researchers develop their own research software; 71\%  have had
-no formal software -development training; 140,000 researchers rely on
+researchers developed their own research software; 71\%    had
+no formal software-development training; 140,000 researchers relied on
 their own coding skills; 4\% of jobs advertised in British universities were software

### 2015-03-24 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: Fixed compilation errors.

@@
 
-Hong led the audience to ponder on the questions that if 92\% of researchers and WSSSPE
-care about scientific software, what are we going to \emph{do} to help researchers and 
-to achieve software sustainability?
+Hong led the audience to ponder the questions of what we are going to \emph{do}
+to help researchers and to achieve software sustainability if 92\% of
+researchers and WSSSPE care about scientific software.
 \begin{comment}
@@
 Non-university Research (industry, government, etc.)
-
 UK STEM graduate career paths: Early Career Research, Permanent Research Staff,
-Professor
-
-\begin{comment}
-Source: The Scientific Century, Royal Society, 2010 (revised to reﬂect ﬁrst stage
-clariﬁcaFon from ``What Do PhD’s Do?'' study)
-
+Professor.
+The Scientific Century, Royal Society, 2010 (revised to reflect first stage
+clarification from ``What Do PhD's Do?'' study.)
 Who	are the people  we need to persuade?
@@
 How are we going to show that the work we present here has benefit to the 92\%?
-
 We are all the 92\%	and we can change the world if we	want!
-
-\url{http://dx.doi.org/10.6084/m9.ﬁgshare.1243288}
-\end{comment}
 

### 2015-03-25 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: first draft.

@@
 
-According to Hong, costs of software-reliant research in the United Kingdom include
+According to Hong, costs of software-reliant research in the United Kingdom included
 \textsterling 840 million of investment in the financial year 2013--2014, an
@@
 data from $49,650$ grant titles and abstracts published on Gateway to Research
-covering years 2010--2014.
-
-Hong led the audience to ponder the questions of what we are going to \emph{do}
-to help researchers and to achieve software sustainability if 92\% of
-researchers and WSSSPE care about scientific software.
+between years 2010 and 2014.
+
+Hong led the audience to ponder the following questions: What are we going to
+\emph{do} to help and benefit software researchers and to achieve software
+sustainability if~92\% of researchers and WSSSPE care about scientific software?
+Who we need to persuade? What are the incentives we need to put in place? He
+challenged the workshop participants to change the current deficient practices
+in research and academia.
 \begin{comment}
@@
 But what are we going to do to make it actually sustainable?
-\end{comment}
+
 Careers outside academic sector: 
@@
 We are all the 92\%	and we can change the world if we	want!
+\end{comment}
 

### 2015-03-25 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: clean up.

@@
 affirmed that it would be impossible or difficult to conduct research without
-software. %Researchers often discover new software predominatly through
-%recommendation of colleagues or web searches.
-Nevertheless the British research community is just starting to understand the
-magnitude of the issue. In the aforementioned SSI survey, 56\% of  
-researchers developed their own research software; 71\%    had
-no formal software-development training; 140,000 researchers relied on
-their own coding skills; 4\% of jobs advertised in British universities were software
-related.
-
-According to Hong, costs of software-reliant research in the United Kingdom included
-\textsterling 840 million of investment in the financial year 2013--2014, an
-amount that has risen by 3\% on average over the past four years. About 30\% of
-total research investment has been spent on research which relies on software
-over the last four financial years. These numbers stemmed from an analysis of
-data from $49,650$ grant titles and abstracts published on Gateway to Research
-between years 2010 and 2014.
+software. Nevertheless the British research community is just starting to
+understand the magnitude of the issue. In the aforementioned SSI survey, 56\% of
+researchers developed their own research software; 71\% had no formal
+software-development training; 140,000 researchers relied on their own coding
+skills; 4\% of jobs advertised in British universities were software related.
+
+According to Hong, costs of software-reliant research in the United Kingdom
+included \textsterling 840 million of investment in the financial year
+2013--2014, an amount that has risen by 3\% on average over the past four years.
+About 30\% of total research investment has been spent on research which relies
+on software over the last four financial years. These numbers stemmed from an
+analysis of data from $49,650$ grant titles and abstracts published on Gateway
+to Research between years 2010 and 2014.
 
@@
 in research and academia.
+
 \begin{comment}

### 2015-03-25 WSSSPE2_report/WSSSPE2_review.tex
minor fixes, mostly names, in 3.2

@@
 Neil Chue Hong is currently serving as the Director of the Software
-Sustainability Institute (SSI), United Kingdom. Hong started with making the
+Sustainability Institute (SSI), United Kingdom. Chue Hong started with making the
 point that world-leading research relies on software, quoting Kersten Kleese Van
@@
 and observational science would have been impossible without better software.}''
-He also cited Daniel Katz, Sustained Innovation (SI2) Program Director of the
+He also cited Daniel Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of the
 National Science Foundation, ``\emph{Scientific discovery and innovation are
@@
 
-Hong drew attention to the misconception that we too often disregard hundreds of
+Chue Hong drew attention to the fact that we too often disregard hundreds of
 thousands of software researchers thinking they are the long tail. We forget
 that actually they are the mainstream and we are the elite minority. He
-emphasized that software is no longer special; it is prevailing in scientific
-research. Survey of researchers from 15 Russell Group universities conducted by
-the SSI between Aug--Oct 2014 included a total of 406 respondents covering a
-representative range of funders, disciplines, and seniority: 92\% confirmed the
-use of research software such as MATLAB, SPSS, Excel, so on and so forth; 89\%
+emphasized that software is no longer special; it is both essential to and common in scientific
+research. A survey of researchers from 15 Russell Group universities conducted by
+the SSI between Aug--Oct 2014, which included a total of 406 respondents covering a
+representative range of funders, disciplines, and seniority, reported that 92\% confirmed the
+use of research software such as MATLAB, SPSS, Excel, etc. and 89\%
 affirmed that it would be impossible or difficult to conduct research without
@@
 understand the magnitude of the issue. In the aforementioned SSI survey, 56\% of
-researchers developed their own research software; 71\% had no formal
-software-development training; 140,000 researchers relied on their own coding
-skills; 4\% of jobs advertised in British universities were software related.
-
-According to Hong, costs of software-reliant research in the United Kingdom
+researchers developed their own research software, 71\% had no formal
+software-development training, 140,000 researchers relied on their own coding
+skills, and only 4\% of jobs advertised in British universities were software related.
+
+According to Chue Hong, costs of software-reliant research in the United Kingdom
 included \textsterling 840 million of investment in the financial year
-2013--2014, an amount that has risen by 3\% on average over the past four years.
-About 30\% of total research investment has been spent on research which relies
+2013--2014, and this amount has risen by 3\% on average over the past four years.
+About 30\% of total research investment has been spent on research that relies
 on software over the last four financial years. These numbers stemmed from an
@@
 
-Hong led the audience to ponder the following questions: What are we going to
+Chue Hong led the audience in discussing the following questions: What are we going to
 \emph{do} to help and benefit software researchers and to achieve software
 sustainability if~92\% of researchers and WSSSPE care about scientific software?
-Who we need to persuade? What are the incentives we need to put in place? He
+Who do we need to persuade? What are the incentives we need to put in place? He
 challenged the workshop participants to change the current deficient practices

### 2015-03-28 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Minor changes + 1 note.

@@
 
-\begin{table*}[ht]
+\begin{table*}[th]
 \begin{center}
 \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important}
+indicating the fraction of attendees who voted for an item as important.}
     \begin{scriptsize}
@@
 ********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
-********** & designing for growth and extension - open development \\ \hline
+********** & designing for growth and extension --- open development \\ \hline
 ******* & culture in community for reuse \\ \hline
@@
 ** & make smart choices about dependencies \\ \hline
-* & thinking of software as product lines - long term vs short term view \\ \hline
+* & thinking of software as product lines --- long term vs.~short term view \\ \hline
  & not all communities need new software \\ \hline
@@
 \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important}
+indicating the fraction of attendees who voted for an item as important.}
     \begin{scriptsize}
@@
 developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people - bus test \\ \hline
+*** & overreliance on one or two people - bus test \choinote{What does ``bus test'' mean?}\\ \hline
 ** & rate of change of underlying technologies \\ \hline

### 2015-03-28 WSSSPE2_report/WSSSPE2_review.tex
Section 10: Add a question in note.

@@
 and the community, rather than this being done in an ad hoc manner by the
-organizers alone.
+organizers alone.\choinote{Are lightning talks new features in WSSSPE2?}
 

### 2015-03-28 WSSSPE2_report/WSSSPE2_review.tex
Section 10: Add a question in note.
edits in response to edits and comments

--- 2015-03-28 WSSSPE2_report/WSSSPE2_review.tex
@@
 ********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
-********** & designing for growth and extension --- open development \\ \hline
+********** & designing for growth and extension -- open development \\ \hline
 ******* & culture in community for reuse \\ \hline
@@
 ** & make smart choices about dependencies \\ \hline
-* & thinking of software as product lines --- long term vs.~short term view \\ \hline
+* & thinking of software as product lines -- long term vs.~short term view \\ \hline
  & not all communities need new software \\ \hline
@@
 \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.}
+indicating the fraction of attendees who voted for an item as important.\katznote{note that there's a footnote within this table -- Where has LaTeX put it?  How do we get it to show up near here?}}
     \begin{scriptsize}
@@
 developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people - bus test \choinote{What does ``bus test'' mean?}\\ \hline
+*** & overreliance on one or two people - the `bus test'\footnote{how few key people would the project need to lose to become non-viable?  (larger is better)} \\ \hline
 ** & rate of change of underlying technologies \\ \hline
@@
 taxonomies for research impact and contributorship such as defined by CASRAI (in
-particular based on the Wellcome-Harvard contributorship taxonomy\katznote{Dan
-to add cite or url}), VIVO, or ISNI. An interesting measure of impact raised by
+particular based on the Wellcome-Harvard contributorship taxonomy\footnote{Project CRediT, \url{http://credit.casrai.org/}}, VIVO, or ISNI. An interesting measure of impact raised by
 the group was the {\em betweenness centrality}, an indicator of a person's
@@
 organizers alone.
+\choinote{Are lightning talks a new feature in WSSSPE2?} \katznote{yes, along with a change to breakout groups that were intended to define actions, then potentially become longer lasting working groups to perform those actions}
 

### 2015-03-31 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Made footnote appear in Table 2. Touched up.

@@
 \usepackage{verbatim}
+
+% The following three lines are used for displaying footnote in tables.
+\usepackage{footnote}
+\makesavenoteenv{tabular}
+\makesavenoteenv{table}
 
@@
 \item that an adequately large community finds value in software and is willing
-to sustain it
-\end{itemize}
-
+to sustain it.
+\end{itemize}
 \katznote{any discussion of this?}
@@
 }
-
 \katznote{any discussion of these tables?}
@@
 \begin{table*}[th]
-\begin{center}
+ \centering%\begin{center}
 \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
 indicating the fraction of attendees who voted for an item as important.}
-    \begin{scriptsize}
-    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+\label{tb:software_sustainability_enablers}
+    %\begin{scriptsize}
+    %\begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+     \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
     \hline
-  Importance & item \\ \hline \hline
+  Importance & Item \\ \hline \hline
 ********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
@@
 **** & culture in dev community to support transition between developers \\ \hline
-*** & interdisciplinary people : science + IT experience \\ \hline
+*** & interdisciplinary people: science + IT experience \\ \hline
 ** & planning for end of life \\ \hline
@@
     \end{tabular}
-    \end{scriptsize}
-    \label{tb:software_sustainability_enablers}
-\end{center}   
+    %\end{scriptsize}
+%\end{center}   
 \end{table*}
 
-\begin{table*}[ht]
-\begin{center}
+\begin{table}[th]
 \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.\katznote{note that there's a footnote within this table -- Where has LaTeX put it?  How do we get it to show up near here?}}
-    \begin{scriptsize}
-    \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+indicating the fraction of attendees who voted for an item as important.}  
+\label{tb:software_sustainability_barriers} 
+ \centering
+  %\begin{scriptsize}
+  %\begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+ \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
    \hline
-  Importance & item \\ \hline \hline
+  Importance & Item \\ \hline \hline
 ******* & lack of incentives, including promotion and tenure process; promotion
@@
 developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people - the `bus test'\footnote{how few key people would the project need to lose to become non-viable?  (larger is better)} \\ \hline
+*** & overreliance on one or two people -- the `bus test'\footnote{how few key people would the project need to lose to become non-viable?  (larger is better)} \\ \hline
 ** & rate of change of underlying technologies \\ \hline
@@
 * & lack of training for how to build sustainability into the system \\ \hline
- & maintenance needed for software is not visible, appears to ``just happen'' \\ \hline
+ & maintenance for software is not visible, appears to ``just happen'' \\ \hline
  & licensing issues \\ \hline
- & staff turnover - lack of continuity \\ \hline
+ & staff turnover -- lack of continuity \\ \hline
 \end{tabular}
-    \end{scriptsize}
-    \label{tb:software_sustainability_barriers}
-\end{center}   
-\end{table*}
+% \end{scriptsize}
+\end{table}
 

### 2015-03-31 WSSSPE2_report/WSSSPE2_review.tex
Section 4 : Table 2: Refine keynote.

@@
 developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people -- the `bus test'\footnote{how few key people would the project need to lose to become non-viable?  (larger is better)} \\ \hline
+*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test'' in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number of key people a project would need to lose to become non-viable (the larger the number, the healthier the project).} \\ \hline
 ** & rate of change of underlying technologies \\ \hline

### 2015-03-31 WSSSPE2_report/WSSSPE2_review.tex
Section 7 and Section 10: Add two notes.

@@
 taxonomies for research impact and contributorship such as defined by CASRAI (in
-particular based on the Wellcome-Harvard contributorship taxonomy\footnote{Project CRediT, \url{http://credit.casrai.org/}}, VIVO, or ISNI. An interesting measure of impact raised by
-the group was the {\em betweenness centrality}, an indicator of a person's
-centrality (and hence, importance) in a scientific collaboration. It is expected
-that developers of research software often play such a central role.
+particular based on the Wellcome-Harvard contributorship
+taxonomy\footnote{Project CRediT, \url{http://credit.casrai.org/}}, VIVO, or
+ISNI. An interesting measure of impact raised by the group was the {\em
+betweenness centrality}, an indicator of a person's centrality (and hence,
+importance) in a scientific collaboration. It is expected that developers of
+research software often play such a central role. \choinote{cite VIVO or ISNI?}
 
@@
 and the community, rather than this being done in an ad hoc manner by the
-organizers alone.
-\choinote{Are lightning talks a new feature in WSSSPE2?} \katznote{yes, along with a change to breakout groups that were intended to define actions, then potentially become longer lasting working groups to perform those actions}
+organizers alone. \choinote{Are lightning talks a new feature in WSSSPE2?}
+\katznote{yes, along with a change to breakout groups that were intended to
+define actions, then potentially become longer lasting working groups to perform
+those actions} \choinote{Since lightning talks is confirmed a new feature in
+WSSSPE2, shall it be recorded somewhere in an earlier section?}
 

### 2015-03-31 WSSSPE2_report/WSSSPE2_review.tex
Front matter: add a note.

@@
 
+\choinote{To extend deadline to after April 1?}
+
 \begin{abstract}      
-This technical report records and discusses the Second Workshop on Sustainable Software for
-Science: Practice and Experiences (WSSSPE2).  The workshop used an alternative submission
-and peer-review process, which led to a set of papers divided across five topic areas:
-exploring sustainability;
-software development experiences;
-credit \& incentives;
-reproducibility \& reuse \& sharing; and
-code testing \& code review.
-The report includes a description of the submission and review process, the workshop
-keynote presentations,
-a discussion on sustainability, and the five discussions from the topic areas.  The report also
-contains a description of potential actions that were proposed in the the five discussions.
+This technical report records and discusses the Second Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE2). The workshop used an
+alternative submission and peer-review process, which led to a set of papers
+divided across five topic areas: exploring sustainability; software development
+experiences; credit \& incentives; reproducibility \& reuse \& sharing; and code
+testing \& code review. The report includes a description of the submission and
+review process, the workshop keynote presentations, a discussion on
+sustainability, and the five discussions from the topic areas. The report also
+contains a description of potential actions that were proposed in the the five
+discussions.
 \end{abstract}
@@
 
-WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the goal of encouraging a wide
-range of submissions from those involved in software practice, ranging from
-initial thoughts and partial studies to mature deployments, but focusing on
-papers that are intended to lead to changes, the organizers wanted to make
-submission as easy as possible. The call for papers stated:
+WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the
+goal of encouraging a wide range of submissions from those involved in software
+practice, ranging from initial thoughts and partial studies to mature
+deployments, but focusing on papers that are intended to lead to changes, the
+organizers wanted to make submission as easy as possible. The call for papers
+stated:
 
@@
 Neil Chue Hong is currently serving as the Director of the Software
-Sustainability Institute (SSI), United Kingdom. Chue Hong started with making the
-point that world-leading research relies on software, quoting Kersten Kleese Van
-Dam of the Pacific Northwestern National Laboratory via the campaign
+Sustainability Institute (SSI), United Kingdom. Chue Hong started with making
+the point that world-leading research relies on software, quoting Kersten Kleese
+Van Dam of the Pacific Northwestern National Laboratory via the campaign
 \href{http://change.org}{change.org}, ``\emph{Today there are very few science
@@
 and observational science would have been impossible without better software.}''
-He also cited Daniel Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of the
-National Science Foundation, ``\emph{Scientific discovery and innovation are
-advancing along fundamentally new pathways opened by development of increasingly
-sophisticated software. Software is an integral enabler of computation,
-experiment and theory, and directly responsible for increased scientific
-productivity and enhancement of researchers' capabilities.}''
+He also cited Daniel Katz, Software Infrastructure for Sustained Innovation
+(SI2) Program Director of the National Science Foundation, ``\emph{Scientific
+discovery and innovation are advancing along fundamentally new pathways opened
+by development of increasingly sophisticated software. Software is an integral
+enabler of computation, experiment and theory, and directly responsible for
+increased scientific productivity and enhancement of researchers'
+capabilities.}''
 
@@
 that actually they are the mainstream and we are the elite minority. He
-emphasized that software is no longer special; it is both essential to and common in scientific
-research. A survey of researchers from 15 Russell Group universities conducted by
-the SSI between Aug--Oct 2014, which included a total of 406 respondents covering a
-representative range of funders, disciplines, and seniority, reported that 92\% confirmed the
-use of research software such as MATLAB, SPSS, Excel, etc. and 89\%
-affirmed that it would be impossible or difficult to conduct research without
-software. Nevertheless the British research community is just starting to
-understand the magnitude of the issue. In the aforementioned SSI survey, 56\% of
-researchers developed their own research software, 71\% had no formal
-software-development training, 140,000 researchers relied on their own coding
-skills, and only 4\% of jobs advertised in British universities were software related.
+emphasized that software is no longer special; it is both essential to and
+common in scientific research. A survey of researchers from 15 Russell Group
+universities conducted by the SSI between Aug--Oct 2014, which included a total
+of 406 respondents covering a representative range of funders, disciplines, and
+seniority, reported that 92\% confirmed the use of research software such as
+MATLAB, SPSS, Excel, etc. and 89\% affirmed that it would be impossible or
+difficult to conduct research without software. Nevertheless the British
+research community is just starting to understand the magnitude of the issue. In
+the aforementioned SSI survey, 56\% of researchers developed their own research
+software, 71\% had no formal software-development training, 140,000 researchers
+relied on their own coding skills, and only 4\% of jobs advertised in British
+universities were software related.
 
@@
 included \textsterling 840 million of investment in the financial year
-2013--2014, and this amount has risen by 3\% on average over the past four years.
-About 30\% of total research investment has been spent on research that relies
-on software over the last four financial years. These numbers stemmed from an
-analysis of data from $49,650$ grant titles and abstracts published on Gateway
-to Research between years 2010 and 2014.
-
-Chue Hong led the audience in discussing the following questions: What are we going to
-\emph{do} to help and benefit software researchers and to achieve software
-sustainability if~92\% of researchers and WSSSPE care about scientific software?
-Who do we need to persuade? What are the incentives we need to put in place? He
-challenged the workshop participants to change the current deficient practices
-in research and academia.
+2013--2014, and this amount has risen by 3\% on average over the past four
+years. About 30\% of total research investment has been spent on research that
+relies on software over the last four financial years. These numbers stemmed
+from an analysis of data from $49,650$ grant titles and abstracts published on
+Gateway to Research between years 2010 and 2014.
+
+Chue Hong led the audience in discussing the following questions: What are we
+going to \emph{do} to help and benefit software researchers and to achieve
+software sustainability if~92\% of researchers and WSSSPE care about scientific
+software? Who do we need to persuade? What are the incentives we need to put in
+place? He challenged the workshop participants to change the current deficient
+practices in research and academia.
 

### 2015-03-31 WSSSPE2_report/WSSSPE2_review.tex
new comment

@@
 importance) in a scientific collaboration. It is expected that developers of
-research software often play such a central role. \choinote{cite VIVO or ISNI?}
+research software often play such a central role. %\choinote{cite VIVO or ISNI?}
 
@@
 those actions} \choinote{Since lightning talks is confirmed a new feature in
-WSSSPE2, shall it be recorded somewhere in an earlier section?}
+WSSSPE2, shall it be recorded somewhere in an earlier section?} \katznote{yes, please feel free to add something in the abstract and intro - maybe we also need a new section after keynotes for lightning talks?}
 

### 2015-04-01 WSSSPE2_report/WSSSPE2_review.tex
adding lightning talk section

@@
 the organization work done before the workshop (\S\ref{sec:preworkshop}); the
-keynotes (\S\ref{sec:keynotes}); a session on defining sustainability
+keynotes (\S\ref{sec:keynotes}); a set of lightning talks (\S\ref{sec:lightning}),
+intended to give an opportunity for attendees to quickly highlight an issue
+or potential solution; a session on defining sustainability
 (\S\ref{sec:defining}); five breakout sessions that explored sustainability
@@
 \end{comment}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Lightning Talks} \label{sec:lightning}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2015-04-01 WSSSPE2_report/WSSSPE2_review.tex
Update due date. Section 4:  Add a few notes from Dan.

@@
 
-\author{TBD by contributions \\{\scriptsize\today \quad  (Due:  $\backsim$ April 1)}} 
+\author{TBD by contributions \\{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
 %\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
@@
 %\thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
-
-\choinote{To extend deadline to after April 1?}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Lead: Choi. Volunteers welcome.
+\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides}}
+
+Lightning talks is a new feature in WSSSPE2. Every talk in it were given only 2
+minutes.
+
+\katznote{You could write to everyone and ask for a short summary of their talks
+- perhaps 1-3 sentences}
+
+\katznote{yes, along with a change to breakout groups that were intended to
+define actions, then potentially become longer lasting working groups to perform
+those actions}
+
+\katznote{yes, please feel free to add something in the abstract and intro}
 
@@
 and the community, rather than this being done in an ad hoc manner by the
-organizers alone. \choinote{Are lightning talks a new feature in WSSSPE2?}
-\katznote{yes, along with a change to breakout groups that were intended to
-define actions, then potentially become longer lasting working groups to perform
-those actions} \choinote{Since lightning talks is confirmed a new feature in
-WSSSPE2, shall it be recorded somewhere in an earlier section?} \katznote{yes, please feel free to add something in the abstract and intro - maybe we also need a new section after keynotes for lightning talks?}
+organizers alone.
 

### 2015-04-01 WSSSPE2_report/WSSSPE2_review.tex
Section 4: microscopic changes.
minor bib + other fixes

@@
 
-Lightning talks is a new feature in WSSSPE2. Every talk in it were given only 2
+Lightning talks is a new feature in WSSSPE2. Every talk in it were given only two
 minutes.
@@
 software used, the previous publication that we are building from, the data that
-we actually used, etc. The Moore Foundation's award in data science \choinote{cite} was given as
+we actually used, etc. The Moore Foundation's award in data science was given as
 an example, asking proposers: What are the five canonical citations that are
-most important to your work? This would be a way of giving credit and assigning
+most important to your work?~\cite{moore-canonical} This would be a way of giving credit and assigning
 importance to these works, differently from how we just count citations today. A
@@
 acknowledgments sections of papers)
-\item developing a taxonomy of contributors, tied to places that these metrics
-are already stored (e.g., ORCID) \choinote{cite}
+\item developing a taxonomy of contributors (e.g., Project Credit~\cite{projectcredit}) tied to places that these metrics
+are already stored (e.g., ORCID) \cite{orcid}
 \item making metadata easier to add for software, creating incentive for
@@
 \begin{itemize}
-\item The FLASH \choinote{cite} project was suggested as an example of how something 
+\item The FLASH~\cite{flash} project was suggested as an example of how something 
 like this has been done.

### 2015-04-02 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Outline.

@@
 
-Lightning talks is a new feature in WSSSPE2. Every talk in it were given only two
-minutes.
-
-\katznote{You could write to everyone and ask for a short summary of their talks
-- perhaps 1-3 sentences}
+Lightning talks is a new feature in WSSSPE2. Every talk in it were given only
+two minutes and the space of one slide. A total of 18 lightning talks were
+selected in reverse alphabetic order of speakers' last names.
+\begin{enumerate}
+\item Greg Wilson: Close Enough for Scientific Work
+
+\item Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures
+
+\item James Spencer: Developing New Developers
+
+\item Erin Robinson:  Sustaining Science Bazaars
+
+\item Marlon Pierce: Patching It Up, Pulling It Forward
+
+\item John W Peterson: Continuous Integration for Concurrent MOOSE Framework and
+Application Development on GitHub
+
+\item Abani Patra: Execute it
+
+\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gap in
+Early Career Training
+
+\item Daniel S. Katz: Implementing Transitive Credit with JSON-LD
+
+\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+
+\item James Howison: Retract all Bit-Rotten Publications
+
+\item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
+Perspective
+
+\item Marcus D. Hanwell: We Need Tool Builders for Sustainable Scientific
+Software
+
+\item Ted Habermann: Communities as the Whole Product
+
+\item Robert R. Downs: Community Recommendations for Improving Sustainable
+Scientific Software Practices
+
+\item Carl Boettiger: rOpenSci: building sustainable software by fostering a
+diverse community
+
+\item Jakob Blomer: The Need for a Versioned Data Analysis Environment
+
+\item Alice Allen: Find it! Cite it!
+\end{enumerate}
 

### 2015-04-02 WSSSPE2_report/WSSSPE2_review.tex
Sections 4 and 5: format change.
putting tables back
adding stuff about lightning talks in abstract, section 4, conclusions fixing some affils in appendix adding note in 3.2

@@
 testing \& code review. The report includes a description of the submission and
-review process, the workshop keynote presentations, a discussion on
+review process, the workshop keynote presentations, a set of lightning talks, a discussion on
 sustainability, and the five discussions from the topic areas. The report also
@@
 
-%\url{http://dx.doi.org/10.6084/m9.ﬁgshare.1243288}
+%\url{http://dx.doi.org/10.6084/m9.figshare.1243288}
 
@@
 the point that world-leading research relies on software, quoting Kersten Kleese
-Van Dam of the Pacific Northwestern National Laboratory via the campaign
-\href{http://change.org}{change.org}, ``\emph{Today there are very few science
+Van Dam of the Pacific Northwestern National Laboratory via a petition campaign at \katznote{add cite for the petition}
+change.org, ``\emph{Today there are very few science
 areas left which do not rely on IT and thus software for the majority of the
@@
 \href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides}}
-
-Lightning talks is a new feature in WSSSPE2. Every talk in it were given only
-two minutes and the space of one slide. A total of 18 lightning talks were
-selected in reverse alphabetic order of speakers' last names.
+ \todo{in the abstract and intro}
+
+Lightning talks were a new feature in WSSSPE2. Since the workshop program was
+mostly dedicated to discussions, the organizers wanted to give the attendees a
+chance to also `make a pitch' for an idea, either representing a contributed paper
+or something different.  18 attendees volunteered to participate in the lightning
+talks, each given two minutes to talk, and one slide if they wanted to use one.
+The talked were presented in the reverse alphabetic order of speakers' last names.
+
 \begin{enumerate}
@@
 \end{enumerate}
-
-\katznote{yes, along with a change to breakout groups that were intended to
-define actions, then potentially become longer lasting working groups to perform
-those actions}
-
-\katznote{yes, please feel free to add something in the abstract and intro}
 
@@
 
-\begin{table*}[th]
- \centering%\begin{center}
+\begin{table*}[ht]
+ \centering
 \caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
@@
 \label{tb:software_sustainability_enablers}
-    %\begin{scriptsize}
-    %\begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
-     \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
+  \begin{scriptsize}
+  \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+   %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
     \hline
@@
     \end{tabular}
-    %\end{scriptsize}
-%\end{center}   
+    \end{scriptsize}
 \end{table*}
 
-\begin{table}[th]
+\begin{table}[ht]
 \caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
@@
  \centering
-  %\begin{scriptsize}
-  %\begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
- \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
+  \begin{scriptsize}
+  \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+% \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
    \hline
@@
 \end{tabular}
-% \end{scriptsize}
+ \end{scriptsize}
 \end{table}
@@
 
+\katznote{add discussion of keynotes, lightning talks, charge to working groups about building plans, etc.}
+
 The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can
@@
 \\ Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Gabrielle Allen       &  U Illinois
+\\ Gabrielle Allen       & University of Illinois Urbana-Champaign
 \\ Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA
-\\ Wolfgang Bangerth & Texas A\&M
+\\ Wolfgang Bangerth & Texas A\&M University
 \\ David Bernholdt       &  Oak Ridge National Laboratory
 \\ Jakob Blomer
-\\ Carl Boettiger        &  UCSC/ropensci
+\\ Carl Boettiger        &  University of California Santa Cruz \& rOpenSci
 \\ Chris Bogart          &  ISR/CMU
-\\ Steven R.~Brandt      &  Louisiana State University
+\\ Steven R. Brandt      &  Louisiana State University
+\\ Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh
 \\ Tom Clune             &  NASA GSFC
-\\ John W.~Cobb
+\\ John W. Cobb
 \\ Dirk Colbry           &  Michigan State University
@@
 \\ Ted Habermann         &  The HDF Group
-\\ Marcus D.~Hanwell     &  Kitware
+\\ Marcus D. Hanwell     &  Kitware
 \\ Sarah Harris          &  University of Leeds
 \\ David Henty           &  EPCC, The University of Edinburgh
-\\ Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh
 \\ James Howison         &  University of Texas
@@
 \\ Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland
-\\ Daniel S.~Katz        &  U Chicago \& Argonne
+\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
 \\ Ian Kelley
-\\ Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) and Duke University
+\\ Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) \& Duke University
 \\ Chris Lenhardt
-\\ Richard Littauer      &  MIT/University of Saarland
+\\ Richard Littauer      &  MIT \& University of Saarland
 \\ Frank L\"{o}ffler     &  Louisiana State University
@@
 \\ Suresh Marru          &  Indiana University
-\\ Robert H.~McDonald    &  Indiana University - Data to Insight Center/Libraries
+\\ Robert H. McDonald    &  Data to Insight Center/Libraries, Indiana University
 \\ Rupert Nash
@@
 \\ Aleksandra Pawlik     &  Software Sustainability Institute
-\\ Cody J.~Permann       &  Idaho National Laboratory
-\\ John W.~Peterson      &  Idaho National Laboratory
+\\ Cody J. Permann       &  Idaho National Laboratory
+\\ John W. Peterson      &  Idaho National Laboratory
 \\ Benjamin Pharr        &  University of Mississippi
@@
 \\ Marlon Pierce         &  Indiana University
-\\ Ray Plante            &  NCSA/University of Illinois Urbana-Champaign
+\\ Ray Plante            &  NCSA, University of Illinois Urbana-Champaign
 \\ Sushil Prasad         &  Georgia State University, Atlanta
-\\ Karthik Ram           &  University of California, Berkeley. Berkeley Institute for Data Science and rOpenSci
-\\ Mike Rilee            &  NASA/GSFC, Rilee Systems Technologies
+\\ Karthik Ram           &  Berkeley Institute for Data Science, University of California Berkeley \& rOpenSci
+\\ Mike Rilee            &  NASA/GSFC \& Rilee Systems Technologies
 \\ Erin Robinson         &  Foundation for Earth Science
@@
 \\ Kaitlin Thaney        &  Mozilla Science Lab
-\\ Matt Turk             &  NCSA/UIUC
-\\ Colin C.~Venters      &  University of Huddersfield
+\\ Matt Turk             &  NCSA, University of Illinois Urbana-Champaign
+\\ Colin C. Venters      &  University of Huddersfield
 \\ Nathan Weeks
 \\ Ethan White           &  University of Florida/Utah State University
-\\ Nancy Wilkins-Diehr   &  SDSC
+\\ Nancy Wilkins-Diehr   &  San Diego Supercomputer Center, University of California San Diego
 \\ Greg Wilson           &  Software Carpentry

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Mini-summaries.

@@
 
+James sought to provoke discussion by proposing that papers whose workflows are
+not kept current with the changing software ecosystem should be automatically
+retracted. This would create the right incentive for authors to keep their
+software current and usable, rather than the current situation in which every
+potential user has to do this individually. A softer version of the proposal
+would identify papers whose software workflow has become bit-rotten and allow
+others to keep the code up to date, either adding them as new authors of the
+paper or providing credit in some other form for their academic service.
+
 \item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
@@
 
+I described how the rOpenSci project has been successful by focusing not just on
+building software but on building a community of researchers who learn and adopt
+our approaches to reproducible research and sustainable software practice.
+Through outreach, mentoring, workshops and hackathons, we have not only reached
+new users, but turned users into co-developers of software and practices to
+support data science research across a growing set of disciplines.
+
 \item Jakob Blomer: The Need for a Versioned Data Analysis Environment
@@
 \item Alice Allen: Find it! Cite it!
+
+The Astrophysics Source Code Library (ASCL) is an online registry of
+scientist-written software used in astronomy research. Our primary interest is
+making research more transparent by making this software more discoverable for
+examination. The ASCL is treated like a publication by the indexing resource for
+astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
+indexes, including citations to ASCL entries. Increasing rewards for writing
+software, whether through citation, transitive credit or other methods, gives
+software authors a powerful reason to take the time to build sustainability into
+their software and is an excellent way to drive community change. The ASCL
+actively looks for opportunities to use change management strategies to move
+astronomy to a more transparent discipline; these strategies can be used in
+other disciplines to further change in them. 
 \end{enumerate}

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Section 4: add email for reference.

@@
 \todo{in the abstract and intro}
+
+\begin{comment}  
+gvwilson@software-carpentry.org, Colin Venters <C.Venters@hud.ac.uk>,
+j.spencer@imperial.ac.uk, erinrobinson@esipfed.org, marpierc@iu.edu,
+jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org, dsk@uchicago.edu,
+Samin.Ishtiaq@microsoft.com, james@howison.name, s.a.harris@leeds.ac.uk,
+marcus.hanwell@kitware.com, thabermann@hdfgroup.org, rdowns@ciesin.columbia.edu,
+cboettig@gmail.com, jakob.blomer@cern.ch, editor@ascl.net 
+\end{comment}
 

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Sections 1 and 4: touch up.

--- 2015-04-02 WSSSPE2_report/WSSSPE2_review.tex
@@
 testing \& code review. The report includes a description of the submission and
-review process, the workshop keynote presentations, a set of lightning talks, a discussion on
+review process, the workshop keynote presentations, a series of lightning talks, a discussion on
 sustainability, and the five discussions from the topic areas. The report also
@@
 the organization work done before the workshop (\S\ref{sec:preworkshop}); the
-keynotes (\S\ref{sec:keynotes}); a set of lightning talks (\S\ref{sec:lightning}),
-intended to give an opportunity for attendees to quickly highlight an issue
-or potential solution; a session on defining sustainability
+keynotes (\S\ref{sec:keynotes}); a series of lightning talks (\S\ref{sec:lightning}),
+intended to give an opportunity for attendees to quickly highlight an important issue
+or a potential solution; a session on defining sustainability
 (\S\ref{sec:defining}); five breakout sessions that explored sustainability
@@
 
+\begin{comment}  
+gvwilson@software-carpentry.org, Colin Venters <C.Venters@hud.ac.uk>,
+j.spencer@imperial.ac.uk, erinrobinson@esipfed.org, marpierc@iu.edu,
+jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org, dsk@uchicago.edu,
+Samin.Ishtiaq@microsoft.com, james@howison.name, s.a.harris@leeds.ac.uk,
+marcus.hanwell@kitware.com, thabermann@hdfgroup.org, rdowns@ciesin.columbia.edu,
+cboettig@gmail.com, jakob.blomer@cern.ch, editor@ascl.net 
+\end{comment}
+
 Lightning talks were a new feature in WSSSPE2. Since the workshop program was
@@
 chance to also `make a pitch' for an idea, either representing a contributed paper
-or something different.  18 attendees volunteered to participate in the lightning
-talks, each given two minutes to talk, and one slide if they wanted to use one.
-The talked were presented in the reverse alphabetic order of speakers' last names.
+or something different.  Eighteen attendees volunteered to participate in the lightning
+talks, each given only two minutes to speak and at most one slide to show.
+The talks were presented in reverse alphabetic order of speakers' last names. 
+In the rest of this session, we highlight the gist of the speakers' key messages.
 
@@
 
+Howison sought to provoke discussion by proposing that papers whose workflows are
+not kept current with the changing software ecosystem should be automatically
+retracted. This would create an incentive for authors to keep their
+software current and usable, rather than the current situation in which every
+potential user has to do this individually. A softer version of the proposal
+would identify papers whose software workflow has become bit-rotten and allow
+others to keep the code up to date, either adding them as new authors of the
+paper or providing credit in some other form for their academic service.
+
 \item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
@@
 
+Boettiger described how the rOpenSci project has been successful by focusing not just on
+building software but also on building a community of researchers who learn and adopt
+their approaches to reproducible research and sustainable software practice.
+Through outreach, mentoring, workshops, and hackathons, they have not only reached
+new users, but also turned users into co-developers of robust software and good practices to
+support data science research across a growing set of disciplines.
+
 \item Jakob Blomer: The Need for a Versioned Data Analysis Environment
@@
 \item Alice Allen: Find it! Cite it!
+
+The Astrophysics Source Code Library (ASCL) is an online registry of
+scientist-written software used in astronomy research. Our primary interest is
+rendering research more transparent by making this software more discoverable for
+examination. The ASCL is treated like a publication by an indexing resource for
+astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
+indexes, including citations to ASCL entries. Increasing rewards for writing
+software, whether through citation, transitive credit or other methods, gives
+software authors a powerful reason to take the time to build sustainability into
+their software and is an excellent way to drive community change. 
+\begin{comment}The ASCL
+actively looks for opportunities to use change management strategies to move
+astronomy to a more transparent discipline; these strategies can be used in
+other disciplines to further change in them. 
+\end{comment}
 \end{enumerate}

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Add mini-summary from speaker.

@@
 \item Marlon Pierce: Patching It Up, Pulling It Forward
+
+How open is open source? Open software needs a diverse, openly governed
+community behind it, just as it needs open licensing and a public code
+repository. To probe the level of governance within open source projects, we
+suggest a contest to encourage individual developers to submit patches and pull
+request to projects that are important to them. This simple mechanism will
+expose several governance mechanisms, such as how easy it is for independent
+developers to communicate with project leadership, how projects accept and
+properly license third party contributions, and how projects make decisions such
+as granting source tree write access.
 

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Appendix A: Add a TODO.
Section 4: Add a note from discussion with Dan.

@@
 
+\note{
+Lighting talks, breakout groups, working groups are all different. Lightning
+talks were new in WSSSPE2. They were offered to all paper contributors, and some
+other attendees, those who got travel awards. Breakout groups in WSSSPE1 were
+just discussions about topics; in WSSSPE2, the breakout groups were supposed to
+come up with plans for actions in specific areas. The idea of working groups is
+that they will come out of breakout groups as a way for the groups to actually
+do the actions over time, not at the workshops.
+}
+
 \begin{comment}  
+Email addresses of speakers:
 gvwilson@software-carpentry.org, Colin Venters <C.Venters@hud.ac.uk>,
@@
 and were noted in that group's notes.
+
+\choinote{Add missing speaker names}
 

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
update for Pierce lighning talk

@@
 
-How open is open source? Open software needs a diverse, openly governed
+Pierce discussed how open open source is. Open software needs a diverse, openly governed
 community behind it, just as it needs open licensing and a public code
-repository. To probe the level of governance within open source projects, we
-suggest a contest to encourage individual developers to submit patches and pull
+repository. To probe the level of governance within open source projects, he and
+his co-authors (Suresh Marru and Chris Mattman)
+suggested a contest to encourage individual developers to submit patches and pull
 request to projects that are important to them. This simple mechanism will

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
Section 4: removed a TODO item done by Dan.
fix bib entry, adding note
Update WSSSPE2_review.tex
Lightning talk update

@@
 \note{Lead: Choi. Volunteers welcome.
-\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides}}
-\todo{in the abstract and intro}
+\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}
 
@@
 through Sustainable Software Architectures
+
+Venters proposed that technically sustainable software is a composite, first-class, non-functional requirement (NFR) 
+that is at a minimum a measure of a systems maintainability and extensibility but may also include other NFR such as 
+efficiency (energy and cost), interoperability, portability, reusability, scalability, and usability. To achieve 
+technically sustainable software it was suggested that software architectures are fundamental as they are the primary 
+carrier of system qualities (NFR) i.e. pre-system understanding, and influence how developers are able to understand, 
+analyze, extend, test and maintain a software system i.e. post-deployment system understanding. In addition, it was 
+highligted that the sustainability of software architectures also needs to be addressed to endure different types of 
+change and evolution in order to mitigate architectural drift and erosion, and architectural knowledge vaporization.
 
@@
 
-\choinote{Add missing speaker names}
+\choinote{Add missing speaker names} \katznote{they are all there}
 
@@
 \\ Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) \& Duke University
-\\ Chris Lenhardt
+\\ Christopher Lenhardt
 \\ Richard Littauer      &  MIT \& University of Saarland

### 2015-04-03 WSSSPE2_report/WSSSPE2_review.tex
rewording a little of new entry, and setting line endings back to Unix style

@@
 
-Venters proposed that technically sustainable software is a composite, first-class, non-functional requirement (NFR) 
-that is at a minimum a measure of a systems maintainability and extensibility but may also include other NFR such as 
-efficiency (energy and cost), interoperability, portability, reusability, scalability, and usability. To achieve 
-technically sustainable software it was suggested that software architectures are fundamental as they are the primary 
-carrier of system qualities (NFR) i.e. pre-system understanding, and influence how developers are able to understand, 
-analyze, extend, test and maintain a software system i.e. post-deployment system understanding. In addition, it was 
-highligted that the sustainability of software architectures also needs to be addressed to endure different types of 
-change and evolution in order to mitigate architectural drift and erosion, and architectural knowledge vaporization.
+Venters proposed that technically sustainable software is a composite, first-class, 
+non-functional requirement (NFR) that is at a minimum a measure of a systems 
+maintainability and extensibility but may also include other NFR such as efficiency 
+(energy and cost), interoperability, portability, reusability, scalability, and usability. To 
+achieve technically sustainable software Venters suggested that software 
+architectures are fundamental as they are the primary carrier of system qualities 
+(NFR), i.e., pre-system understanding, and influence how developers are able to 
+understand, analyze, extend, test and maintain a software system, i.e., post-
+deployment system understanding. In addition, Venters highlighted that the 
+sustainability of software architectures also needs to be addressed to endure 
+different types of change and evolution in order to mitigate architectural drift and 
+erosion, and architectural knowledge vaporization.
 

### 2015-04-05 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Comment out for now talks that have no summaries.

@@
 The talks were presented in reverse alphabetic order of speakers' last names. 
-In the rest of this session, we highlight the gist of the speakers' key messages.
+In the rest of this session, we highlight the gist of some of the speakers' messages.
 
 \begin{enumerate}
-\item Greg Wilson: Close Enough for Scientific Work
+%\item Greg Wilson: Close Enough for Scientific Work
 
@@
 
-\item James Spencer: Developing New Developers
-
-\item Erin Robinson:  Sustaining Science Bazaars
+%\item James Spencer: Developing New Developers
+
+%\item Erin Robinson:  Sustaining Science Bazaars
 
@@
 
-\item John W Peterson: Continuous Integration for Concurrent MOOSE Framework and
-Application Development on GitHub
-
-\item Abani Patra: Execute it
-
-\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gap in
-Early Career Training
-
-\item Daniel S. Katz: Implementing Transitive Credit with JSON-LD
-
-\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+%\item John W Peterson: Continuous Integration for Concurrent MOOSE Framework and
+%Application Development on GitHub
+
+%\item Abani Patra: Execute it
+
+%\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gap in
+%Early Career Training
+
+%\item Daniel S. Katz: Implementing Transitive Credit with JSON-LD
+
+%\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
 
@@
 
-\item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
-Perspective
-
-\item Marcus D. Hanwell: We Need Tool Builders for Sustainable Scientific
-Software
-
-\item Ted Habermann: Communities as the Whole Product
-
-\item Robert R. Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices
+%\item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
+%Perspective
+
+%\item Marcus D. Hanwell: We Need Tool Builders for Sustainable Scientific
+%Software
+
+%\item Ted Habermann: Communities as the Whole Product
+
+%\item Robert R. Downs: Community Recommendations for Improving Sustainable
+%Scientific Software Practices
 
@@
 
-\item Jakob Blomer: The Need for a Versioned Data Analysis Environment
+% \item Jakob Blomer: The Need for a Versioned Data Analysis Environment
 

### 2015-04-06 WSSSPE2_report/WSSSPE2_review.tex
adding stuff from Nancy

@@
 
-\todo{short intro to group here}
+\katznote{need some short general intro to Software Development Experiences here}
 
@@
 
+Because of the large number of submitted papers categorized as Software
+Development Experiences, papers were split into two subgroups prior to the
+workshop.
+
+\katznote{need something about subgroup A too}
+
+Subgroup B discussed four papers and had six
+participants. Because of the nature of the papers, training emerged as a
+common theme, however the conversation was wide-ranging, including
+incentives, reproducibility and funding to promote sustainability.
+
+\subsubsection{Subgroup A Discussion}
+
+\subsubsection{Subgroup B Discussion}
+The statement ``applied computer science is being attempted in academia
+without any formal training'' kicked off our discussion. The group brought
+expertise in several different training models, from a 2-day Software
+Carpentry workshop, to a 3-week Open Science for Synthesis (OSS) program
+to semester-long programs. We discussed the pros and cons of different
+training approaches, also touching on informal learning, for example where
+people learning the necessary skills because they had cross-disciplinary
+people (``boundary scientists'') in their work environments of whom they
+could ask questions.
+
+Some papers explored gaps in training of early career scientists. Industry
+participants in our group confirm this. We asked ourselves, ``Are
+traditional courses failing?''. We think yes. Changes to undergraduate
+curricula requirements are difficult. But as programming models become
+more complex, we have to raise the skill levels. Skills must be improved
+not only in traditional programming - learning languages and algorithms -
+but around professional software development. We want to get to a point
+where ``Everyone has a new minimum now - everyone knows git''. Developers
+also need training on licensing choices. Just because a code is on git
+doesn't mean it's open.
+
+"How do you get people to look for the training they need?'' wondered one
+participant. People seek out opportunities like Software Carpentry to
+augment skills. While some instruction can be done in an institutional
+curricula, independent groups (non-profits, institutes) have more
+flexibility. Some asked whether Silicon Valley be interested in funding
+training so that people are better prepared to enter the workforce? Some
+felt that companies were reluctant to deliver training because employees
+then leave the company. While others felt that pushing all training to
+industry could lead to good technical people ``getting on the Google bus''
+and leaving the sciences.
+
+"How will we know when people are trained effectively in these new
+skills?" We discussed certification. It can be hard to build the
+recognition of Java, database certifications amongst all technologies. OSS
+offers badges to those completing training, however we need to demonstrate
+proficiency, not just completion. Google Summer of Code is a big CV
+augmentation for participants, can we duplicate something like that?
+
+The group also discussed how to fund training. On participant observed
+that NYU runs a 6-week data science training, companies grab graduates but
+pay for those they hire. OSS also used the NSF Software Institutes program
+as a vehicle to fund training activities. Software Carpentry uses a
+collaborative teaching approach where  people publish open teaching
+materials and receive credit for their use.
+
+We then discussed career paths for those supporting sustainable software.
+While tenure track is not the only option for students, the challenges for
+those who remain in academia can be large. Research scientists are
+entirely dependent on soft money, which can be unpredictable. Postdocs and
+those who do pursue tenure track positions need to publish and see no
+rewards in software development. These challenges were all identified at
+WSSSPE1. What actions would we recommend to improve things? Altmetrics and
+download statistics may slowly change the system and improving a
+developer's ability receive credit for time invested in software
+development. NSF recognition of data products (including software) as well
+as publications is also helping.
+
+We asked ourselves, ``Where are there examples of things that are changing
+because of this and how can we build momentum?''. One example demonstrates
+change over time. In 2007, nanoHUB listed the academic reward structure as
+a problem in an EDUCAUSE report where the authors note, ``In the future,
+nanoHUB researchers are hoping to change the research culture. While they
+recognize that young faculty members are unlikely to get tenure based on
+their nanoHUB contributions, they hope to encourage faculty to think
+beyond their own research needs to consider publishing tutorials and other
+Web content in their field on the nanoHUB site''. Fast forward seven years
+later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
+former student of mine published eight tools on nanoHUB, serving over
+6,000 people with his tools. He then joined a university as a professor
+and introduced nanoHUB. Use of the gateway from that university
+skyrocketed; he used nanoHUB in existing classes, created new classes, and
+infused it in his research. Ultimately, the professor's department head
+associates his two-year rise to tenure with the notoriety and innovation
+he gained through nanoHUB.''
+
+The group also discussed ``attribution trees'', an idea put forward by Dan
+Katz and Arfon Smith where a chain of attributions can, for example, give
+credit to those developing libraries and building blocks that other pieces
+of software use. The group considered where to push the use of this, which
+journals? On participant noted that Dryad works with the journals. If a
+paper is accepted, the supporting data must be submitted. We discussed
+reproducibility as a component of the journal review process and ``active
+papers'', with immediate links to the data and software.
+
+We then discussed variations amongst scientific domain areas and wondered
+"Are some communities more or less open than others?'' To some members of
+the group biology seems to be open, while physics less so. Some felt with
+the more recent development of bioinformatics as a field there were fewer
+historical practices to undo. Physics has pre-print philosophy to
+overcome. Environmental sciences may be mixed. The group felt the
+biomedical area, however, was very competitive and closed. One participant
+mentioned blueprints for going open source (like NWChem recently did)
+where authors outline how this helps, what you do and what the next steps
+are.
+
+We then moved beyond our training discussion to address funding that
+encourages sustainable software -- funding of both people and of projects
+that create a true system to support sustainability. We called this
+``institutionalized serendipity''. Because science is increasingly reliant
+on software, one participant observed that ``software development can have
+much broader impact than publishing individual research, but it is not
+viewed that way''. Because of this centrality, one participant mentioned
+that training ought to be called Science Carpentry rather than Software
+Carpentry. We felt we were beginning to see changes in the research
+community as a result of the NSF's data management plan requirements. PIs
+are thinking more about data and some university libraries are offering
+data repositories. We wondered if a software management plan might be
+effective. "How might funding programs need to adapt to reward good
+software development practices?'', we asked ourselves. We thought about
+best practices such as version control, test harnesses, mailing lists, bug
+tracking, community contribution, and reuse where appropriate. We thought
+about measuring success through usage statistics (downloads, altmetrics).
+"Should funders demand reproducibility?'' we wondered? In order for results
+to be reproduced, software would need to be carefully curated.
+
+Again, our industry participants contributed unique viewpoints. Partnering
+with industry was seen one path to sustainability. The unique partnerships
+Kitware engages in promote academic freedom while creating an income
+stream from certain portions of the software. This type of approach to
+sustainable software frees researchers from performing tasks that do not
+offer the rewards their institution values''. We talked some about
+successful models for industry partnerships that preserve open science.
+Participants noted that there are some NSF programs that prohibit
+partnerships with for profit companies (but there are others where this is
+encouraged).
+
+\subsubsection{Subgroup A Actions}
+
+\subsubsection{Subgroup B Actions}
+
+Subgroup B then focused on actions it could take. It discussed development
+of a white paper that describes a matrixed approach to training
+(multi-day, multi-week, semester). The white paper might include a survey
+of existing techniques. There are many, some dating back many years, for
+example the Interuniversity Consortium for Political and Social Research
+(ICPSR) and various summer institutes. The white paper could Include a
+call for a comprehensive assessment of these techniques. We need to think
+carefully about the right venue for such a white paper, where would it
+have the most impact. The subgroup believes it would need to approach editors
+directly to ascertain this.
+
+The group felt that training in techniques that promote sustainability has
+a range of benefits: career paths, educated reviewers, reproducible
+science and more. There is some information on how that has been
+approached and assessed, but more is needed. The group felt that this
+training is undervalued and that it's important to communicate the return
+on investment - both individual ROI (skills that make scientists more
+effective and more marketable) and funder ROI (better use of taxpayer
+funds, research more likely to be reproducible because sustainable
+software exists, better trained reviewers).
+
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Software Development Experiences group are:
-\begin{itemize}
+The papers that were discussed in the Software Development Experiences Subgroup A are:
+
+\begin{itemize}
+\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
+scientific software: a hackathon experience~\cite{wssspe2_crusoe}
+
+\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software
+Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
+for Science~\cite{wssspe2_hanwell}
+
+\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
+S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for
+Sustainable Science Software from an Early Career Training Institute on Open
+Science Synthesis~\cite{wssspe2_lenhardt}
+
+\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do
+Scientists Find Most Useful?~\cite{wssspe2_schossau}
+
+\end{itemize}
+
+The papers that were discussed in the Software Development Experiences Subgroup B are:
+
+\begin{itemize}
+
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
@@
 project~\cite{wssspe2_boettiger}
-
-\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
-scientific software: a hackathon experience~\cite{wssspe2_crusoe}
 
@@
 
-\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software
-Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
-for Science~\cite{wssspe2_hanwell}
-
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski.
 Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
-
-\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
-S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for
-Sustainable Science Software from an Early Career Training Institute on Open
-Science Synthesis~\cite{wssspe2_lenhardt}
-
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do
-Scientists Find Most Useful?~\cite{wssspe2_schossau}
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section 1. Outline modified.

@@
 the organization work done before the workshop (\S\ref{sec:preworkshop}); the
-keynotes (\S\ref{sec:keynotes}); a series of lightning talks (\S\ref{sec:lightning}),
-intended to give an opportunity for attendees to quickly highlight an important issue
-or a potential solution; a session on defining sustainability
-(\S\ref{sec:defining}); five breakout sessions that explored sustainability
-(\S\ref{sec:exploring}); software development experiences (\S\ref{sec:devel});
-credit \& incentives (\S\ref{sec:credit}); reproducibility, reuse, \& sharing
-(\S\ref{sec:reproduce}); code testing \& code review
-(\S\ref{sec:code_testing}); and some conclusions (\S\ref{sec:conclusions}). The
-report also includes an incomplete list of attendees (Appendix~\ref{sec:attendees}).
+keynotes (\S\ref{sec:keynotes}); a series of lightning talks
+(\S\ref{sec:lightning}), intended to give an opportunity for attendees to
+quickly highlight an important issue or a potential solution; a session on
+defining sustainability (\S\ref{sec:defining}). The report also gives summaries
+of action plans proposed by five breakout sessions, which explored in specific
+areas including sustainability (\S\ref{sec:exploring}); software development
+experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit});
+reproducibility, reuse, \& sharing (\S\ref{sec:reproduce}); code testing \& code
+review (\S\ref{sec:code_testing}). Lastly, the report also includes some
+conclusions (\S\ref{sec:conclusions}) and an incomplete list of attendees
+(Appendix~\ref{sec:attendees}).
 
@@
 
-\note{
+\begin{comment}
 Lighting talks, breakout groups, working groups are all different. Lightning
@@
 do the actions over time, not at the workshops.
-}
+\end{comment{
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section 4: raw summary from speaker.

@@
 
-%\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+%  Microsoft Research Cambridge
+
+The reproduction and replication of novel results has become a major issue in
+computer science, systems biology and other computational disciplines. These
+include both the inability to re-implement novel algorithms and approaches, and
+also an agreement on how and what to benchmark these algorithms on.
+
+In the talk, we discussed these problems and suggested several suggestions to
+address them.
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Slight touch up.

--- 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-%\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+%  Microsoft Research Cambridge
+
+The reproduction and replication of novel results has become a major issue in
+computer science, systems biology, and other computational disciplines. These
+include both the inability to re-implement novel algorithms and approaches, and
+lack of an agreement on how and what to benchmark these algorithms on.
+Ishtiaq pointed out these problems and made several suggestions to address them.
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section 4. More minor changes.

@@
 do the actions over time, not at the workshops.
-\end{comment{
-
-\begin{comment}  
+
 Email addresses of speakers:
@@
 
-\item Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
-through Sustainable Software Architectures
-
+\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures.}
 Venters proposed that technically sustainable software is a composite, first-class, 
@@
 
-\item Marlon Pierce: Patching It Up, Pulling It Forward
-
+\item \textbf{Marlon Pierce: Patching It Up, Pulling It Forward.}
 Pierce discussed how open open source is. Open software needs a diverse, openly governed
@@
 
-\item Samin Ishtiaq: Daemons, Notifications and Sustaining Software
+\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software.}
 %  Microsoft Research Cambridge
-
 The reproduction and replication of novel results has become a major issue in
@@
 lack of an agreement on how and what to benchmark these algorithms on.
-Ishtiaq pointed out these problems and made several suggestions to address them.
-
-\item James Howison: Retract all Bit-Rotten Publications
-
+Ishtiaq from Microsoft Research Cambridge pointed out these problems and made several suggestions to address them.
+
+\item \textbf{James Howison: Retract all Bit-Rotten Publications.}
 Howison sought to provoke discussion by proposing that papers whose workflows are
@@
 
-\item Carl Boettiger: rOpenSci: building sustainable software by fostering a
-diverse community
-
+\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
+Diverse Community.}
 Boettiger described how the rOpenSci project has been successful by focusing not just on
@@
 
-\item Alice Allen: Find it! Cite it!
+\item \textbf{Alice Allen: Find it! Cite it!}
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section 3: make placeholder for citation.

@@
 the point that world-leading research relies on software, quoting Kersten Kleese
-Van Dam of the Pacific Northwestern National Laboratory via a petition campaign at \katznote{add cite for the petition} 
+Van Dam of the Pacific Northwestern National Laboratory via a petition campaign\footnote{\katznote{add cite for the petition}} at
 change.org, ``\emph{Today there are very few science

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Abstract: fix typo.
Section 1: Add a TODO about travel award.
Updated NCH's keynote description

--- 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
@@
 sustainability, and the five discussions from the topic areas. The report also
-contains a description of potential actions that were proposed in the the five
+contains a description of potential actions that were proposed in the five
 discussions.
@@
 
-Neil Chue Hong is currently serving as the Director of the Software
-Sustainability Institute (SSI), United Kingdom. Chue Hong started with making
-the point that world-leading research relies on software, quoting Kersten Kleese
-Van Dam of the Pacific Northwestern National Laboratory via a petition campaign at \katznote{add cite for the petition} 
-change.org, ``\emph{Today there are very few science
-areas left which do not rely on IT and thus software for the majority of the
-their research work. More importantly key scientific advances in experimental
-and observational science would have been impossible without better software.}''
-He also cited Daniel Katz, Software Infrastructure for Sustained Innovation
-(SI2) Program Director of the National Science Foundation, ``\emph{Scientific
-discovery and innovation are advancing along fundamentally new pathways opened
-by development of increasingly sophisticated software. Software is an integral
-enabler of computation, experiment and theory, and directly responsible for
-increased scientific productivity and enhancement of researchers'
-capabilities.}''
-
-Chue Hong drew attention to the fact that we too often disregard hundreds of
-thousands of software researchers thinking they are the long tail. We forget
-that actually they are the mainstream and we are the elite minority. He
+Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
+United Kingdom. The SSI was founded to support the UK?s research software 
+community by cultivating better, more sustainable, research software to enable 
+world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
+ started by making the point of that the use of 
+- and reliance on - software is pervasive in all areas of world-leading research, 
+showing examples from disciplines as diverse as the humanities and high-energy 
+physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
+National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
+at change.org, ``\emph{Today there are very few science areas left 
+which do not rely on IT and thus software for the majority of their research work. 
+More importantly key scientific advances in experimental and observational science 
+would have been impossible without better software.}'' He also cited Daniel Katz, 
+Software Infrastructure for Sustained Innovation(SI2) Program Director of the 
+National Science Foundation, ``\emph{Scientific discovery and innovation are 
+advancing along fundamentally new pathways opened by development of 
+increasingly sophisticated software. Software is an integral enabler of 
+computation, experiment and theory, and directly responsible for increased 
+scientific productivity and enhancement of researchers' capabilities.}''
+
+Chue Hong drew attention to the issue that in the cyber-infrastructure and high 
+performance community, we all too often disregard the hundreds of thousands 
+of researchers developing software thinking they are the long tail. In fact, the 
+numbers point to the fact that they are actually the mainstream and we (referring 
+to people including the audience of the talk) are the elite minority. He
 emphasized that software is no longer special; it is both essential to and
-common in scientific research. A survey of researchers from 15 Russell Group
-universities conducted by the SSI between Aug--Oct 2014, which included a total
-of 406 respondents covering a representative range of funders, disciplines, and
-seniority, reported that 92\% confirmed the use of research software such as
-MATLAB, SPSS, Excel, etc. and 89\% affirmed that it would be impossible or
-difficult to conduct research without software. Nevertheless the British
-research community is just starting to understand the magnitude of the issue. In
-the aforementioned SSI survey, 56\% of researchers developed their own research
-software, 71\% had no formal software-development training, 140,000 researchers
-relied on their own coding skills, and only 4\% of jobs advertised in British
-universities were software related.
-
-According to Chue Hong, costs of software-reliant research in the United Kingdom
+common in scientific research. A 2014 survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers} 
+conducted by the SSI polled researchers 
+from 15 research-intensive UK universities (406 respondents covering a 
+representative range of funders, disciplines, and seniority) reported that 92\% 
+confirmed the use of research software  and 89\% affirmed that it would be 
+impossible or difficult to conduct research without software. Nevertheless the British
+research community is just starting to understand the magnitude of the issue. 
+Whilst many researchers make use of software such as MATLAB, SPSS, and 
+Excel, data from the aforementioned SSI survey shows that over half (56\%) of 
+respondents developed their own research software (which equates to over 
+140,000 researchers if extrapolated across the UK) and yet 71\% had no formal 
+software-development training, having to rely on their own coding skills.
+
+Examining another aspect of the size of the research software community, 
+Chue Hong noted that the costs of software-reliant research in the United Kingdom
 included \textsterling 840 million of investment in the financial year
@@
 relies on software over the last four financial years. These numbers stemmed
-from an analysis of data from $49,650$ grant titles and abstracts published on
-Gateway to Research between years 2010 and 2014.
-
-Chue Hong led the audience in discussing the following questions: What are we
-going to \emph{do} to help and benefit software researchers and to achieve
-software sustainability if~92\% of researchers and WSSSPE care about scientific
-software? Who do we need to persuade? What are the incentives we need to put in
-place? He challenged the workshop participants to change the current deficient
-practices in research and academia.
-
-\begin{comment}
-With apologies to Greg Wilson:
-\href{http://software-carpentry.org/blog/2013/10/you-keep-using-that-word.html}
-{You Keep Using That Word}
-We can talk all day about the best ways of achieving sustainability
-But what are we going to do to make it actually sustainable?
-
-Careers outside academic sector: 
-Non-university Research (industry, government, etc.)
-UK STEM graduate career paths: Early Career Research, Permanent Research Staff,
-Professor.
-The Scientific Century, Royal Society, 2010 (revised to reflect first stage
-clarification from ``What Do PhD's Do?'' study.)
-Who	are the people  we need to persuade?
-What	 are	the incentives we need to put in place?
-How are we going to show that the work we present here has benefit to the 92\%?
-We are all the 92\%	and we can change the world if we	want!
-\end{comment}
+from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
+Gateway to Research between years 2010 and 2014. A similar analysis of university 
+jobs advertised in the same period discovered that despite this investment, only 4\% of 
+positions were software development related, and of these only 17\% were explicitly 
+named as a software developer / software engineer position: the vast majority being 
+advertised as research associate / research assistant positions. This in turn leads to
+the issue of career paths for those bridging the research and software worlds, who
+are essential to support the use and further development of research software, a point 
+highlighted by a graphic showing UK STEM graduate career paths\footnote{Source: The Scientific Century, Royal Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's Do?'' study)}
+ showing that only 3.5\% were able to secure permanent positions.
+
+To conclude, Chue Hong led the audience in discussing the following questions: 
+What are \emph{we} going to \emph{do} to help and benefit the 92\% of researchers 
+who rely on software? Who do we need to persuade? What are the incentives we 
+need to put in place? Finally, he challenged the workshop participants to change 
+the current deficient practices in research and academia.
 

### 2015-04-07 WSSSPE2_report/WSSSPE2_review.tex
Section4: raw summary.

@@
 early-career scientists, postdoctoral researchers, and graduate students.
+\todo{record how/number the travel awards were created and recipients selected, etc?}
 
@@
 physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
-National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
+National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
 at change.org, ``\emph{Today there are very few science areas left 
@@
 
-%\item Robert R. Downs: Community Recommendations for Improving Sustainable
-%Scientific Software Practices
+\item \textbf{Robert R. Downs: Community Recommendations for Improving Sustainable
+Scientific Software Practices.}
+Robert R. Downs, of the Columbia University Center for International Earth
+Science Information Network (CIESIN), presented Community Recommendations for
+Improving Sustainable Scientific Software Practices (2014), and described a
+focus group study that was conducted with other members of the Science Software
+Cluster of the Federation of Earth Science Information Partners (ESIP),
+including W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
+Weber. For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
+invited to participate in simultaneous roundtable discussions, on the
+sustainability of science software. Over two-thirds of the roundtable focus
+groups provided responses to a semi-structured questionnaire that contained
+three sets of questions. Initial analysis of the participants' responses to the
+set of questions eliciting recommendations for near-term actions of the
+community to improve sustainable software practices revealed several
+suggestions. The recommendations for improving scientific software
+sustainability included improving community engagement and collaborative
+activities, increasing understanding and awareness, and creating incentives to
+motivate sustainable science software practices. Based on the suggestions
+offered by the focus group respondents for near-term actions, the ESIP Science
+Software Cluster plans to engage the community to achieve progress in the near
+future on the recommended activities for improving sustainable scientific
+software practices.
+
+Reference
+
+Downs RR, Lenhardt WC, Robinson E, Davis E, Weber N. 2014. Community
+Recommendations for Improving Sustainable Scientific Software Practices. 2nd
+Workshop on Sustainable Software for Science: Practice and Experiences
+(WSSSPE2), New Orleans, Louisiana, 16 November 2014.
+http://wssspe.researchcomputing.org.uk/wp-content/uploads/2014/11/Downs.pdf and
+http://dx.doi.org/10.7916/D8Q52NBC
+\choinote{Choi to revise.}
 

### 2015-04-09 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Touch up

@@
 early-career scientists, postdoctoral researchers, and graduate students.
-\todo{record how/number the travel awards were created and recipients selected, etc?}
+\todo{record how/number of  travel awards were created \& recipients selected?}
 
@@
 Robert R. Downs, of the Columbia University Center for International Earth
-Science Information Network (CIESIN), presented Community Recommendations for
-Improving Sustainable Scientific Software Practices (2014), and described a
-focus group study that was conducted with other members of the Science Software
-Cluster of the Federation of Earth Science Information Partners (ESIP),
-including W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
-Weber. For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
-invited to participate in simultaneous roundtable discussions, on the
+Science Information Network (CIESIN), 
+%presented Community Recommendations for
+%Improving Sustainable Scientific Software Practices (2014), and 
+described a
+focus group study that was conducted with W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
+Weber of the Science Software
+Cluster of the Federation of Earth Science Information Partners (ESIP). For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
+invited to participate in simultaneous roundtable discussions on  
 sustainability of science software. Over two-thirds of the roundtable focus
-groups provided responses to a semi-structured questionnaire that contained
-three sets of questions. Initial analysis of the participants' responses to the
-set of questions eliciting recommendations for near-term actions of the
-community to improve sustainable software practices revealed several
-suggestions. The recommendations for improving scientific software
-sustainability included improving community engagement and collaborative
+groups   responded to a semi-structured survey that contained
+three sets of questions eliciting recommendations for near-term actions of the
+community to improve sustainable software practices. Initial analysis of the participants' responses to the questionnaire
+ revealed several suggestions, which  included improving community engagement and collaborative
 activities, increasing understanding and awareness, and creating incentives to
-motivate sustainable science software practices. Based on the suggestions
-offered by the focus group respondents for near-term actions, the ESIP Science
+motivate sustainable science software practices. 
+%Based on the recommendations
+%offered by the focus group respondents for near-term actions, 
+The ESIP Science
 Software Cluster plans to engage the community to achieve progress in the near
@@
 
-Reference
-
+ \begin{comment}
 Downs RR, Lenhardt WC, Robinson E, Davis E, Weber N. 2014. Community
@@
 http://dx.doi.org/10.7916/D8Q52NBC
-\choinote{Choi to revise.}
+ \end{comment}
 

### 2015-04-10 WSSSPE2_report/WSSSPE2_review.tex
Section 3.2: fix typos.

@@
 Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
-United Kingdom. The SSI was founded to support the UK?s research software 
-community by cultivating better, more sustainable, research software to enable 
+United Kingdom. The SSI was founded to support the UK's research software 
+community by cultivating better, more sustainable research software to enable 
 world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
  started by making the point of that the use of 
-- and reliance on - software is pervasive in all areas of world-leading research, 
+-- and reliance on -- software is pervasive in all areas of world-leading research, 
 showing examples from disciplines as diverse as the humanities and high-energy 
@@
 
-Chue Hong drew attention to the issue that in the cyber-infrastructure and high 
-performance community, we all too often disregard the hundreds of thousands 
+Chue Hong drew attention to the issue that in the cyber-infrastructure and 
+high-performance community, we all too often disregard the hundreds of thousands 
 of researchers developing software thinking they are the long tail. In fact, the 

### 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
adding some travel support stuff in intro, finishing 8.5 (prizes)

@@
 potentially transformative outcomes. It strongly encouraged participation of
-early-career scientists, postdoctoral researchers, and graduate students.
-\todo{record how/number of  travel awards were created \& recipients selected?}
+early-career scientists, postdoctoral researchers, and graduate students, with
+funds provided to the conference organizers by the Moore Foundation and the
+National Science Foundation, to support the travel of potential participants who
+would not otherwise be able to attend, and young participants and those from
+underrepresented groups, respectively.  These funds allowed \todo{number}
+additional participants to attend, and each was offered the chance to present a
+lightning talk.
 
@@
 participants in our group confirm this. We asked ourselves, ``Are
-traditional courses failing?''. We think yes. Changes to undergraduate
+traditional courses failing?'' We think yes. Changes to undergraduate
 curricula requirements are difficult. But as programming models become
@@
 but around professional software development. We want to get to a point
-where ``Everyone has a new minimum now - everyone knows git''. Developers
+where ``Everyone has a new minimum now - everyone knows git.'' Developers
 also need training on licensing choices. Just because a code is on git
@@
 We asked ourselves, ``Where are there examples of things that are changing
-because of this and how can we build momentum?''. One example demonstrates
+because of this and how can we build momentum?'' One example demonstrates
 change over time. In 2007, nanoHUB listed the academic reward structure as
@@
 beyond their own research needs to consider publishing tutorials and other
-Web content in their field on the nanoHUB site''. Fast forward seven years
+Web content in their field on the nanoHUB site.'' Fast forward seven years
 later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
@@
 
-The group also discussed ``attribution trees'', an idea put forward by Dan
+The group also discussed ``attribution trees,'' an idea put forward by Dan
 Katz and Arfon Smith where a chain of attributions can, for example, give
@@
 reproducibility as a component of the journal review process and ``active
-papers'', with immediate links to the data and software.
+papers,'' with immediate links to the data and software.
 
@@
 that create a true system to support sustainability. We called this
-``institutionalized serendipity''. Because science is increasingly reliant
+``institutionalized serendipity.'' Because science is increasingly reliant
 on software, one participant observed that ``software development can have
 much broader impact than publishing individual research, but it is not
-viewed that way''. Because of this centrality, one participant mentioned
+viewed that way.'' Because of this centrality, one participant mentioned
 that training ought to be called Science Carpentry rather than Software
@@
 effective. "How might funding programs need to adapt to reward good
-software development practices?'', we asked ourselves. We thought about
+software development practices?'' we asked ourselves. We thought about
 best practices such as version control, test harnesses, mailing lists, bug
@@
 about measuring success through usage statistics (downloads, altmetrics).
-"Should funders demand reproducibility?'' we wondered? In order for results
+"Should funders demand reproducibility?'' we wondered. In order for results
 to be reproduced, software would need to be carefully curated.
@@
 sustainable software frees researchers from performing tasks that do not
-offer the rewards their institution values''. We talked some about
+offer the rewards their institution values.'' We talked some about
 successful models for industry partnerships that preserve open science.
@@
 
-\katznote{I don't understand this at all.  Can someone explain it?}
-There are also other indirect forms of incentives
-(though these are not a direct prize, they still can incentivize someone):
+Other (non-prize) forms of incentives are
 \begin{itemize}
@@
 
-To determine who should give prizes, perhaps we should find those who we think would
+The group suggested that to determine who should give out prizes, perhaps we should find those who we think would
 be awarded prizes, then ask them who they would want to receive a prize from, then
@@
 
-One idea is to create a funding program for disciplines or other organizations to
-create a prize program.  We provide a framework, for example: awards to individuals, must
-award in 5 or 6 areas, must have a jury and/or objective criteria that includes
-senior/junior domain people and technology people, must have the recipients awarded
-at a relevant event, must provide citations that explain why the prizes should be awarded.
-A concern is that having many organizations award prizes may reduce the impact.
-
-Possible awarders are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
-Society of America, etc. Or perhaps this should be a joint technology/science
+One of the group's ideas was to create a funding program for disciplines or other organizations to
+create a prize program.  We would provide a framework and a set of requirements,
+for example: awards to individuals; must
+award in 5 or 6 areas; must have a jury and/or objective criteria that includes
+senior/junior domain people and technology people; must have the recipients awarded
+at a relevant event; and must provide citations that explain why the prizes should be awarded.
+Different organizations could then decide to sign up to the framework and give out
+awards under this general brand.  However, there was a concern that having many
+organizations award prizes may reduce the impact of the prizes.
+
+Possible groups that might give out awards, either under our framework, or more generally,
+are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
+Society of America, etc. Perhaps this could be a joint technology/science
 partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
-society] prize?
-
-Some potential criteria for prizes are:
+society] prize.
+
+Some potential criteria for prizes the group suggested are:
 community engagement, helping out others;
@@
 
-Perhaps there should be different criteria for ``established'' members of the
-community versus junior members? Or perhaps prizes should be restricted to
-junior members?
-
-An important point is that we don't want to give prizes just to reward people
+The group discussion if there should be different criteria for ``established'' members of the
+community versus junior members, and if prizes should be restricted to
+junior members, but left these as open questions.
+
+A point the group considered important is that we don't want to give prizes just to reward people
 who are really good at this one thing, but rather we want to reward people who
@@
 and were noted in that group's notes.
-
-\choinote{Add missing speaker names} \katznote{they are all there}
 

### 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
fixing a typo - also, need to remind other lightning talk speakers to contribute summaries
Section 4: Add raw summary.

@@
 The talks were presented in reverse alphabetic order of speakers' last names. 
-In the rest of this session, we highlight the gist of some of the speakers' messages.
+In the rest of this section, we highlight the gist of some of the speakers' messages.
 
@@
 
-%\item John W Peterson: Continuous Integration for Concurrent MOOSE Framework and
-%Application Development on GitHub
+\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
+Application Development on GitHub}
+
+%Affiliation: Idaho National Laboratory
+
+In March 2014, the MOOSE framework was released under an open
+source license on GitHub, significantly expanding and diversifying the pool of
+current active and potential future contributors on the project. The MOOSE team
+employs an extensive continuous integration test suite to ensure that both the
+framework, and the applications based on the framework, are verified before any
+code changes are accepted into the repository. We use a combination of built-in
+git features such as branching and submodules, GitHub API integration
+capabilities, and in-house developed testing software to both perform this
+verification and update the dependent applications in a relatively seamless
+manner for users.
 

### 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
Section 4: touch up.

--- 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-%\item John W Peterson: Continuous Integration for Concurrent MOOSE Framework and
-%Application Development on GitHub
+\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
+Application Development on GitHub}
+
+%Affiliation: Idaho National Laboratory
+
+Peterson from the the Idaho National Laboratory reported that in March 2014, the MOOSE framework was released under an open
+source license on GitHub, significantly expanding and diversifying the pool of
+current active and potential future contributors on the project. The MOOSE team
+employs an extensive continuous integration test suite to ensure that both the
+framework and the applications based on the framework are verified before any
+code changes are accepted into the repository. They use a combination of built-in
+git features such as branching and submodules, GitHub API integration
+capabilities, and in-house developed testing software to perform this
+verification and update the dependent applications in a relatively seamless
+manner for users.
 

### 2015-04-14 WSSSPE2_report/WSSSPE2_review.tex
Section Section 4: touch up.

@@
 
-\item \textbf{Robert R. Downs: Community Recommendations for Improving Sustainable
+\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
 Scientific Software Practices.}
-Robert R. Downs, of the Columbia University Center for International Earth
+Robert Downs, of the Columbia University Center for International Earth
 Science Information Network (CIESIN), 

### 2015-04-15 WSSSPE2_report/WSSSPE2_review.tex
Section 4: touch up.
adding intro to software dev section from Nancy

@@
 Application Development on GitHub} 
-
 %Affiliation: Idaho National Laboratory
-
 Peterson from the the Idaho National Laboratory reported that in March 2014, the MOOSE framework was released under an open
@@
 
-\katznote{need some short general intro to Software Development Experiences here}
+Of the short actionable papers that would lead to improvements for sustainable
+software science, 11 submissions were categorized in the  Software Development Experiences
+group. Because of the large number, we split these into two subgroups
+prior to the event. Some common themes helped in this division, for
+example several papers that addressed education and training issues,
+including best practices and case studies, were grouped
+together. Others discussed experiences with registries, developer
+collectives and specific examples of successful, sustainable software
+(including a valuable industry perspective).
 
@@
 
-Because of the large number of submitted papers categorized as Software
-Development Experiences, papers were split into two subgroups prior to the
-workshop.
-
-\katznote{need something about subgroup A too}
+\katznote{need something about subgroup A here}
 

### 2015-04-15 WSSSPE2_report/WSSSPE2_review.tex
adding some authors

@@
 
-\author{TBD by contributions \\{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
-%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$, Neil Chue Hong$^{(3)}$, \\
+\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(11)}$, Nancy Wilkins-Diehr$^{(8)}$, Neil Chue Hong$^{(3)}$, \\
+James Howison  (affil etc. needed), Miguel de Val-Borro (affill etc. needed), Richard Littauer (affil etc. needed) \\
+more TBD by contributions \\
+{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
+%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$,  \\
 %Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
@@
 %
-%\thanks{{}$^{(1)}$ National Science Foundation, Arlington, VA, USA; Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}
+\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}
 %
@@
 %
-%\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
 %  
@@
 %
-%\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
+\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
 %
@@
 %
-%\thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+\thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA}
 

### 2015-04-15 WSSSPE2_report/WSSSPE2_review.tex
Authors: Minor clean up.

@@
 \author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(11)}$, Nancy Wilkins-Diehr$^{(8)}$, Neil Chue Hong$^{(3)}$, \\
-James Howison  (affil etc. needed), Miguel de Val-Borro (affill etc. needed), Richard Littauer (affil etc. needed) \\
+James Howison$^{(12)}$, Miguel de Val-Borro$^{(13)}$, Richard Littauer$^{(14)}$ \\
 more TBD by contributions \\
@@
 \thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA}
+\thanks{{}$^{(12)}$ TBD}
+\thanks{{}$^{(14)}$ TBD}
 

### 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
Authors: touch up.

--- 2015-04-15 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(11)}$, Nancy Wilkins-Diehr$^{(8)}$, Neil Chue Hong$^{(3)}$, \\
-James Howison  (affil etc. needed), Miguel de Val-Borro (affill etc. needed), Richard Littauer (affil etc. needed) \\
-more TBD by contributions \\
+\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
+\\ James Howison$^{(5)}$, Miguel de Val-Borro$^{(6)}$, Richard Littauer$^{(7)}$ 
+\\ more TBD by contributions 
+\\ 
 {\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
-%\author{Daniel S. Katz$^{(1)}$, Gabrielle Allen$^{(2)}$,  \\
+%\author{Gabrielle Allen$^{(11)}$,  \\
 %Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$, Nancy Wilkins-Diehr$^{(8)}$,
-%Richard Littauer$^{(9)}$,
-%Miguel de Val-Borro$^{(10)}$, 
-%Sou-Cheng T. Choi$^{(11)}$}
+%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$,  Miguel de Val-Borro$^{(10)}$}
 %
@@
 %
-%\thanks{{}$^{(2)}$ University of Illinois, Champaign, IL, USA}
-%
-\thanks{{}$^{(3)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA}
+%
+\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA}
+%
+\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
+%
+\thanks{{}$^{(5)}$ TBD}
+%
+\thanks{{}$^{(6)}$ TBD}  
+%
+\thanks{{}$^{(7)}$ TBD}   
 %  
@@
 %
-\thanks{{}$^{(8)}$ University of California-San Diego, San Diego, CA, USA}
-%
 %\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
@@
 %
-\thanks{{}$^{(11)}$ NORC at the University of Chicago and   Illinois Institute of Technology, Chicago, IL, USA}
+%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}
 

### 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
Sections 1-4: fiddling.

@@
 (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on Sunday, 16 November 2014 in conjunction with the International
+held on  16 November 2014 in conjunction with the International
 Conference for High Performance Computing, Networking, Storage and Analysis
@@
 funds provided to the conference organizers by the Moore Foundation and the
-National Science Foundation, to support the travel of potential participants who
+National Science Foundation (NSF), to support the travel of potential participants who
 would not otherwise be able to attend, and young participants and those from
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\begin{comment}
 \note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
@@
 \href{http://tinyurl.com/mnenzms}{Abstracts and slides}}
+\end{comment}
 
@@
 -- and reliance on -- software is pervasive in all areas of world-leading research, 
-showing examples from disciplines as diverse as the humanities and high-energy 
+showing examples from disciplines as diverse as humanities and high-energy 
 physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
@@
 would have been impossible without better software.}'' He also cited Daniel Katz, 
-Software Infrastructure for Sustained Innovation(SI2) Program Director of the 
-National Science Foundation, ``\emph{Scientific discovery and innovation are 
+Software Infrastructure for Sustained Innovation (SI2) Program Director of the 
+NSF, ``\emph{Scientific discovery and innovation are 
 advancing along fundamentally new pathways opened by development of 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
+\begin{comment}
 \note{Lead: Choi. Volunteers welcome.
 \href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}
+\end{comment}
 
@@
 through Sustainable Software Architectures.}
-Venters proposed that technically sustainable software is a composite, first-class, 
-non-functional requirement (NFR) that is at a minimum a measure of a systems 
-maintainability and extensibility but may also include other NFR such as efficiency 
-(energy and cost), interoperability, portability, reusability, scalability, and usability. To 
-achieve technically sustainable software Venters suggested that software 
-architectures are fundamental as they are the primary carrier of system qualities 
-(NFR), i.e., pre-system understanding, and influence how developers are able to 
-understand, analyze, extend, test and maintain a software system, i.e., post-
-deployment system understanding. In addition, Venters highlighted that the 
-sustainability of software architectures also needs to be addressed to endure 
-different types of change and evolution in order to mitigate architectural drift and 
-erosion, and architectural knowledge vaporization.
+Venters proposed that sustainable software is a composite, first-class, 
+non-functional requirement (NFR) that is at a minimum a measure of a system's 
+maintainability and extensibility, but may also include other NFRs such as efficiency 
+(e.g., energy, cost), interoperability, portability, reusability, scalability, and usability. To 
+achieve technically sustainable software, Venters suggested that software 
+architectures are fundamental as they are the primary carrier of system   
+NFRs, i.e., pre-system understanding; and influence how developers are able to 
+understand, analyze, extend, test, and maintain a software system, i.e., 
+post-deployment system understanding. In addition, Venters highlighted that   
+sustainability of software architectures needs to be addressed to endure 
+different types of change and evolution in order to mitigate architectural drift, 
+erosion, and knowledge vaporization.
 

### 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
Section 4: fiddling.

@@
 \item \textbf{Alice Allen: Find it! Cite it!}
-
+%
 The Astrophysics Source Code Library (ASCL) is an online registry of
-scientist-written software used in astronomy research. Our primary interest is
+scientist-written software used in astronomy research. Their primary interest is
 rendering research more transparent by making this software more discoverable for
-examination. The ASCL is treated like a publication by an indexing resource for
+examination. The ASCL is treated as a publication by an indexing resource for
 astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it

### 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Simplify a summary a bit.

@@
 described a
-focus group study that was conducted with W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
-Weber of the Science Software
-Cluster of the Federation of Earth Science Information Partners (ESIP). For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
+focus group study conducted with W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
+Weber of the Science Software Cluster (SSC) of the Federation of Earth Science Information Partners (ESIP). For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
 invited to participate in simultaneous roundtable discussions on  
@@
 %offered by the focus group respondents for near-term actions, 
-The ESIP Science
-Software Cluster plans to engage the community to achieve progress in the near
-future on the recommended activities for improving sustainable scientific
+The ESIP SSC plans to engage the community %to achieve progress in the near future 
+in the recommended activities for improving sustainable scientific
 software practices.

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 10: Add top #WSSSPE tweets (to be continued)

@@
 \author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
-\\ James Howison$^{(5)}$, Miguel de Val-Borro$^{(6)}$, Richard Littauer$^{(7)}$ 
-\\ more TBD by contributions 
-\\ 
-{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
+\\James Howison$^{(5)}$, Miguel de Val-Borro$^{(6)}$, Richard Littauer$^{(7)}$, more TBD %by contributions 
+\\{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
 %\author{Gabrielle Allen$^{(11)}$,  \\
@@
 provide a record of the workshop.
+
+WSSSPE has developed an engaging online mechanism via the online social network
+Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets during
+WSSSPE2 than WSSSPE1 or WSSSPE1.1. The WSSSPE community counts 178 tweets and
+61 folllowers between May 16, 2013 to April 18, 2015. Some tweets were re-tweeeted or like more than the others.
+Table~\ref{tab:tweets} summarizes the top posts and their statistics.
+
+ \begin{table*}[t]
+ \centering
+\caption{Top tweets of \#WSSSPE.}\label{tab:tweets}
+  \begin{scriptsize}
+  \begin{tabular}{ l|l|l|r|r}
+    Source & Date & Tweet  & Retweets &  Favorites
+\\ \hline Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
+\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
+%
+\\ Neil P Chue Hong  & Nov 16 &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
+\\     & & management plan used by EPSRC software grants    &    &  
+\\     & & http://www.software.ac.uk/resources/guides/software-management-plans  &    & 
+%
+\\ Neil P Chue Hong  & Nov 16 &  @jameshowison as well as software plans   & 4 & 7
+\\   & &  http://www.software.ac.uk/resources/guides/software-management-plans &    & 
+\\   & &  we provide a software evaluation tool:   &    & 
+\\   & &  http://www.software.ac.uk/online-sustainability-evaluation  &    & 
+%
+\\ Dr Tom Crick & Nov 16 & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
+\\ & & UK researchers write research software w$/$out any formal training &    & 
+%
+\\ Karthik Ram &  Nov 16 &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+%
+\\ Josh Greenberg &  Nov 16 &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
+\\   & &  ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
+%
+\\ Ethan White &  Nov 16 & ``@rOpenSci is all about community... our measures of success  & 9 & 3
+\\ & &  [include] how many faces are up on our community page''   &    & 
+%
+\\ Ethan White &  Nov 16 & Daniel Katz talking about implementing transitive credit for  & 9 & 7
+\\ & & software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
+% 
+\\ Aleksandra Pawlik &  Nov 16 &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
+% 
+\\ Aleksandra Pawlik &  Nov 16 & Lack of training as one of the main barriers for sustainable software & 7 & 2
+\\ & &   at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
+%
+\\ Kaitlin Thaney &  Nov 16 & My slides from this morning's keynote at & 11 & 12
+\\ & & WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
+\\ & &  http://www.slideshare.net/kaythaney/     & & 
+\\ & & designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
+%
+\\ Neil P Chue Hong  & Nov 16 & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
+\\ & & @rOpenSci @stilettofiend around open training activities for sustainability  & & 
+%
+\\ Richard Littauer  & Nov 16 &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
+\\
+    \end{tabular}
+    \end{scriptsize}
+\end{table*}
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 11:  more tweets.

--- 2015-04-18 WSSSPE2_report/WSSSPE2_review.tex
@@
 provide a record of the workshop.
+
+WSSSPE has developed an engaging online mechanism via the online social network
+Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets during
+WSSSPE2 than WSSSPE1 or WSSSPE1.1. The WSSSPE community counts 178 tweets and
+61 folllowers between May 16, 2013 to April 18, 2015. Some tweets were re-tweeeted or like more than the others.
+Table~\ref{tab:tweets} summarizes the top posts and their statistics.
+
+ \begin{table*}[t]
+ \centering
+\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
+  \begin{scriptsize}
+  \begin{tabular}{l|l|r|r}
+ \hline
+    Source &   Tweet  & Retweets &  Favorites
+\\ \hline 
+% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
+%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
+%
+% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
+%\\     &  management plan used by EPSRC software grants    &    &  
+% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
+%
+ Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
+\\   &   http://www.software.ac.uk/resources/guides/software-management-plans &    & 
+\\   &   we provide a software evaluation tool:   &    & 
+\\   &   http://www.software.ac.uk/online-sustainability-evaluation  &    & 
+%
+\\ Dr Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
+\\ & UK researchers write research software w$/$out any formal training &    & 
+%
+\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+%
+\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
+\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
+%
+\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
+\\ &   [include] how many faces are up on our community page''   &    & 
+%
+\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
+\\ &  software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
+% 
+%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
+% 
+%\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 7 & 2
+%\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
+%
+\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
+\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
+\\ &   http://www.slideshare.net/kaythaney/     & & 
+\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
+%
+\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
+\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
+%
+%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
+%
+\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
+\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
+\\ & http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/  & & 
+%
+\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
+\\ & Bioinformatics Manifesto: https://github.com/pjotrp/bioinformatics   & 
+%
+\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
+\\ & but because influential people are averse to change...''  C. Sunstein     & & 
+% 
+\\ Dr Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
+\\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
+%
+\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
+\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
+%
+\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
+\\ &   $+$ sustainability. Tune in! Live stream: http://ustre.am/17ddh & & 
+%
+\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
+\\ & http://wssspe.researchcomputing.org.uk/wssspe2/agenda/   &  &
+\\ & URL for live stream of keynotes \& lightning talks: http://ustre.am/17ddh   &  &
+\\ \hline
+    \end{tabular}
+    \end{scriptsize}
+\end{table*}
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 11: minor changes.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \todo{currently just pre-workshop stuff - needs to be updated}
@@
 \katznote{add discussion of keynotes, lightning talks, charge to working groups about building plans, etc.}
+
+\begin{table*}[t]
+\centering
+\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
+  \begin{scriptsize}
+  \begin{tabular}{l|l|r|r}
+ \hline
+    Author  &   Tweet  & Retweets &  Favorites
+\\ \hline 
+% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
+%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
+%
+% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
+%\\     &  management plan used by EPSRC software grants    &    &  
+% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
+%
+ Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
+\\   &   http://www.software.ac.uk/resources/guides/software-management-plans &    & 
+\\   &   we provide a software evaluation tool:   &    & 
+\\   &   http://www.software.ac.uk/online-sustainability-evaluation  &    & 
+%
+\\ Dr Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
+\\ & UK researchers write research software w$/$out any formal training &    & 
+%
+\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+%
+\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
+\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
+%
+\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
+\\ &   [include] how many faces are up on our community page''   &    & 
+%
+\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
+\\ &  software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
+% 
+%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
+% 
+\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
+\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
+%
+\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
+\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
+\\ &   http://www.slideshare.net/kaythaney/     & & 
+\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
+%
+\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
+\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
+%
+%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
+%
+\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
+\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
+\\ & http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/  & & 
+%
+\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
+\\ & Bioinformatics Manifesto: https://github.com/pjotrp/bioinformatics   & 
+%
+\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
+\\ & but because influential people are averse to change...''  C. Sunstein     & & 
+% 
+\\ Dr Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
+\\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
+%
+\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
+\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
+%
+\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
+\\ &   $+$ sustainability. Tune in! Live stream: http://ustre.am/17ddh & & 
+%
+\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
+\\ & http://wssspe.researchcomputing.org.uk/wssspe2/agenda/   &  &
+\\ & URL for live stream of keynotes \& lightning talks: http://ustre.am/17ddh   &  &
+\\ \hline
+    \end{tabular}
+    \end{scriptsize}
+\end{table*}
 
@@
 WSSSPE has developed an engaging online mechanism via the online social network
-Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets during
+Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets (messages) during
 WSSSPE2 than WSSSPE1 or WSSSPE1.1. The WSSSPE community counts 178 tweets and
-61 folllowers between May 16, 2013 to April 18, 2015. Some tweets were re-tweeeted or like more than the others.
-Table~\ref{tab:tweets} summarizes the top posts and their statistics.
-
- \begin{table*}[t]
- \centering
-\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
-  \begin{scriptsize}
-  \begin{tabular}{l|l|r|r}
- \hline
-    Source &   Tweet  & Retweets &  Favorites
-\\ \hline 
-% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
-%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    & 
-%
-% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
-%\\     &  management plan used by EPSRC software grants    &    &  
-% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    & 
-%
- Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
-\\   &   http://www.software.ac.uk/resources/guides/software-management-plans &    & 
-\\   &   we provide a software evaluation tool:   &    & 
-\\   &   http://www.software.ac.uk/online-sustainability-evaluation  &    & 
-%
-\\ Dr Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
-\\ & UK researchers write research software w$/$out any formal training &    & 
-%
-\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
-%
-\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
-\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    & 
-%
-\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
-\\ &   [include] how many faces are up on our community page''   &    & 
-%
-\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
-\\ &  software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
-% 
-%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
-% 
-%\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 7 & 2
-%\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & & 
-%
-\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
-\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
-\\ &   http://www.slideshare.net/kaythaney/     & & 
-\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
-%
-\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
-\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & & 
-%
-%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 & 
-%
-\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
-\\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
-\\ & http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/  & & 
-%
-\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
-\\ & Bioinformatics Manifesto: https://github.com/pjotrp/bioinformatics   & 
-%
-\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3 
-\\ & but because influential people are averse to change...''  C. Sunstein     & & 
-% 
-\\ Dr Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
-\\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
-%
-\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
-\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
-%
-\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
-\\ &   $+$ sustainability. Tune in! Live stream: http://ustre.am/17ddh & & 
-%
-\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
-\\ & http://wssspe.researchcomputing.org.uk/wssspe2/agenda/   &  &
-\\ & URL for live stream of keynotes \& lightning talks: http://ustre.am/17ddh   &  &
-\\ \hline
-    \end{tabular}
-    \end{scriptsize}
-\end{table*}
+61 followers between May 16, 2013 and April 18, 2015. Some tweets were re-tweeeted or liked more than the others.
+Table~\ref{tab:tweets} summarizes the top tweets and their statistics on the day of workshop, Nov 16, 2014,
+reverberating some of the main points and highlights in the meeting.
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 11: touch up.

@@
 WSSSPE has developed an engaging online mechanism via the online social network
-Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets (messages) during
-WSSSPE2 than WSSSPE1 or WSSSPE1.1. The WSSSPE community counts 178 tweets and
-61 followers between May 16, 2013 and April 18, 2015. Some tweets were re-tweeeted or liked more than the others.
-Table~\ref{tab:tweets} summarizes the top tweets and their statistics on the day of workshop, Nov 16, 2014,
-reverberating some of the main points and highlights in the meeting.
+Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets (messages)
+during WSSSPE2 than WSSSPE1 or WSSSPE1.1. Some of the main points and highlights
+in the meeting reverberated in Twitter. Table~\ref{tab:tweets} summarizes the
+top tweets happened on the day of workshop, Nov 16, 2014,
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Format. No changes.

@@
 testing \& code review. The report includes a description of the submission and
-review process, the workshop keynote presentations, a series of lightning talks, a discussion on
-sustainability, and the five discussions from the topic areas. The report also
-contains a description of potential actions that were proposed in the five
+review process, the workshop keynote presentations, a series of lightning talks,
+a discussion on sustainability, and the five discussions from the topic areas.
+The report also contains a description of potential actions that were proposed
+in the five
 discussions.
@@
 community by cultivating better, more sustainable research software to enable
-world-class research. Chue Hong's keynote\footnote{Slides available at: http://dx.doi.org/10.6084/m9.figshare.1243288}
- started by making the point of that the use of 
--- and reliance on -- software is pervasive in all areas of world-leading research, 
-showing examples from disciplines as diverse as humanities and high-energy 
-physics, and quoting Kersten Kleese Van Dam of the Pacific Northwestern 
-National Laboratory via a petition campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
-at change.org, ``\emph{Today there are very few science areas left 
-which do not rely on IT and thus software for the majority of their research work. 
-More importantly key scientific advances in experimental and observational science 
-would have been impossible without better software.}'' He also cited Daniel Katz, 
-Software Infrastructure for Sustained Innovation (SI2) Program Director of the 
-NSF, ``\emph{Scientific discovery and innovation are 
-advancing along fundamentally new pathways opened by development of 
-increasingly sophisticated software. Software is an integral enabler of 
-computation, experiment and theory, and directly responsible for increased 
-scientific productivity and enhancement of researchers' capabilities.}''
+world-class research. Chue Hong's keynote\footnote{Slides available at:
+http://dx.doi.org/10.6084/m9.figshare.1243288} started by making the point of
+that the use of -- and reliance on -- software is pervasive in all areas of
+world-leading research, showing examples from disciplines as diverse as
+humanities and high-energy physics, and quoting Kersten Kleese Van Dam of the
+Pacific Northwestern National Laboratory via a petition
+campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
+at change.org, ``\emph{Today there are very few science areas left which do not
+rely on IT and thus software for the majority of their research work. More
+importantly key scientific advances in experimental and observational science
+would have been impossible without better software.}'' He also cited Daniel
+Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
+the NSF, ``\emph{Scientific discovery and innovation are advancing along
+fundamentally new pathways opened by development of increasingly sophisticated
+software. Software is an integral enabler of computation, experiment and theory,
+and directly responsible for increased scientific productivity and enhancement
+of researchers' capabilities.}''
 
@@
 of researchers developing software thinking they are the long tail. In fact, the
-numbers point to the fact that they are actually the mainstream and we (referring 
-to people including the audience of the talk) are the elite minority. He
-emphasized that software is no longer special; it is both essential to and
-common in scientific research. A 2014 survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers} 
-conducted by the SSI polled researchers 
-from 15 research-intensive UK universities (406 respondents covering a 
-representative range of funders, disciplines, and seniority) reported that 92\% 
-confirmed the use of research software  and 89\% affirmed that it would be 
-impossible or difficult to conduct research without software. Nevertheless the British
-research community is just starting to understand the magnitude of the issue. 
-Whilst many researchers make use of software such as MATLAB, SPSS, and 
-Excel, data from the aforementioned SSI survey shows that over half (56\%) of 
-respondents developed their own research software (which equates to over 
-140,000 researchers if extrapolated across the UK) and yet 71\% had no formal 
-software-development training, having to rely on their own coding skills.
+numbers point to the fact that they are actually the mainstream and we
+(referring to people including the audience of the talk) are the elite minority.
+He emphasized that software is no longer special; it is both essential to and
+common in scientific research. A 2014
+survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}
+conducted by the SSI polled researchers from 15 research-intensive UK
+universities (406 respondents covering a representative range of funders,
+disciplines, and seniority) reported that 92\% confirmed the use of research
+software and 89\% affirmed that it would be impossible or difficult to conduct
+research without software. Nevertheless the British research community is just
+starting to understand the magnitude of the issue. Whilst many researchers make
+use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
+SSI survey shows that over half (56\%) of respondents developed their own
+research software (which equates to over 140,000 researchers if extrapolated
+across the UK) and yet 71\% had no formal software-development training, having
+to rely on their own coding skills.
 
@@
 Gateway to Research between years 2010 and 2014. A similar analysis of university 
-jobs advertised in the same period discovered that despite this investment, only 4\% of 
-positions were software development related, and of these only 17\% were explicitly 
-named as a software developer / software engineer position: the vast majority being 
-advertised as research associate / research assistant positions. This in turn leads to
-the issue of career paths for those bridging the research and software worlds, who
-are essential to support the use and further development of research software, a point 
-highlighted by a graphic showing UK STEM graduate career paths\footnote{Source: The Scientific Century, Royal Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's Do?'' study)}
- showing that only 3.5\% were able to secure permanent positions.
+jobs advertised in the same period discovered that despite this investment, only
+4\% of positions were software development related, and of these only 17\% were
+explicitly named as a software developer / software engineer position: the vast
+majority being advertised as research associate / research assistant positions.
+This in turn leads to the issue of career paths for those bridging the research
+and software worlds, who are essential to support the use and further
+development of research software, a point highlighted by a graphic showing UK
+STEM graduate career paths\footnote{Source: The Scientific Century, Royal
+Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's
+Do?'' study)} showing that only 3.5\% were able to secure permanent positions.
 
@@
 non-functional requirement (NFR) that is at a minimum a measure of a system's
-maintainability and extensibility, but may also include other NFRs such as efficiency 
-(e.g., energy, cost), interoperability, portability, reusability, scalability, and usability. To 
-achieve technically sustainable software, Venters suggested that software 
-architectures are fundamental as they are the primary carrier of system   
-NFRs, i.e., pre-system understanding; and influence how developers are able to 
-understand, analyze, extend, test, and maintain a software system, i.e., 
-post-deployment system understanding. In addition, Venters highlighted that   
-sustainability of software architectures needs to be addressed to endure 
-different types of change and evolution in order to mitigate architectural drift, 
-erosion, and knowledge vaporization.
+maintainability and extensibility, but may also include other NFRs such as
+efficiency (e.g., energy, cost), interoperability, portability, reusability,
+scalability, and usability. To achieve technically sustainable software, Venters
+suggested that software architectures are fundamental as they are the primary
+carrier of system NFRs, i.e., pre-system understanding; and influence how
+developers are able to understand, analyze, extend, test, and maintain a
+software system, i.e., post-deployment system understanding. In addition,
+Venters highlighted that sustainability of software architectures needs to be
+addressed to endure different types of change and evolution in order to mitigate
+architectural drift, erosion, and knowledge vaporization.
 
@@
 \item \textbf{Marlon Pierce: Patching It Up, Pulling It Forward.}
-Pierce discussed how open open source is. Open software needs a diverse, openly governed
-community behind it, just as it needs open licensing and a public code
+Pierce discussed how open open source is. Open software needs a diverse, openly
+governed community behind it, just as it needs open licensing and a public code
 repository. To probe the level of governance within open source projects, he and
-his co-authors (Suresh Marru and Chris Mattman)
-suggested a contest to encourage individual developers to submit patches and pull
-request to projects that are important to them. This simple mechanism will
-expose several governance mechanisms, such as how easy it is for independent
-developers to communicate with project leadership, how projects accept and
-properly license third party contributions, and how projects make decisions such
-as granting source tree write access.
+his co-authors (Suresh Marru and Chris Mattman) suggested a contest to encourage
+individual developers to submit patches and pull request to projects that are
+important to them. This simple mechanism will expose several governance
+mechanisms, such as how easy it is for independent developers to communicate
+with project leadership, how projects accept and properly license third party
+contributions, and how projects make decisions such as granting source tree
+write access.
 
@@
 %Affiliation: Idaho National Laboratory
-Peterson from the the Idaho National Laboratory reported that in March 2014, the MOOSE framework was released under an open
-source license on GitHub, significantly expanding and diversifying the pool of
-current active and potential future contributors on the project. The MOOSE team
-employs an extensive continuous integration test suite to ensure that both the
-framework and the applications based on the framework are verified before any
-code changes are accepted into the repository. They use a combination of built-in
-git features such as branching and submodules, GitHub API integration
-capabilities, and in-house developed testing software to perform this
-verification and update the dependent applications in a relatively seamless
-manner for users.
+Peterson from the the Idaho National Laboratory reported that in March 2014, the
+MOOSE framework was released under an open source license on GitHub,
+significantly expanding and diversifying the pool of current active and
+potential future contributors on the project. The MOOSE team employs an
+extensive continuous integration test suite to ensure that both the framework
+and the applications based on the framework are verified before any code changes
+are accepted into the repository. They use a combination of built-in git
+features such as branching and submodules, GitHub API integration capabilities,
+and in-house developed testing software to perform this verification and update
+the dependent applications in a relatively seamless manner for users.
 
@@
 described a
-focus group study conducted with W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and Nicholas
-Weber of the Science Software Cluster (SSC) of the Federation of Earth Science Information Partners (ESIP). For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were
-invited to participate in simultaneous roundtable discussions on  
-sustainability of science software. Over two-thirds of the roundtable focus
-groups   responded to a semi-structured survey that contained
-three sets of questions eliciting recommendations for near-term actions of the
-community to improve sustainable software practices. Initial analysis of the participants' responses to the questionnaire
- revealed several suggestions, which  included improving community engagement and collaborative
-activities, increasing understanding and awareness, and creating incentives to
-motivate sustainable science software practices. 
+focus group study conducted with %W. Christopher Lenhardt, Erin Robinson, Ethan
+% Davis, and Nicholas Weber of 
+the Science Software Cluster (SSC) of the
+Federation of Earth Science Information Partners (ESIP). For the study, almost
+300 attendees of the 2014 Summer ESIP Meeting were invited to participate in
+simultaneous roundtable discussions on sustainability of science software. Over
+two-thirds of the roundtable focus groups responded to a semi-structured survey
+that contained three sets of questions eliciting recommendations for near-term
+actions of the community to improve sustainable software practices. Initial
+analysis of the
+participants' responses to the questionnaire revealed several suggestions, which
+included improving community engagement and collaborative activities, increasing
+understanding and awareness, and creating incentives to motivate sustainable
+science software practices.
 %Based on the recommendations
@@
 Diverse Community.}
-Boettiger described how the rOpenSci project has been successful by focusing not just on
-building software but also on building a community of researchers who learn and adopt
-their approaches to reproducible research and sustainable software practice.
-Through outreach, mentoring, workshops, and hackathons, they have not only reached
-new users, but also turned users into co-developers of robust software and good practices to
-support data science research across a growing set of disciplines.
+Boettiger described how the rOpenSci project has been successful by focusing not
+just on building software but also on building a community of researchers who
+learn and adopt their approaches to reproducible research and sustainable
+software practice. Through outreach, mentoring, workshops, and hackathons, they
+have not only reached new users, but also turned users into co-developers of
+robust software and good practices to support data science research across a
+growing set of disciplines.
 
@@
 their software and is an excellent way to drive community change.
-\begin{comment}The ASCL
-actively looks for opportunities to use change management strategies to move
-astronomy to a more transparent discipline; these strategies can be used in
+
+\begin{comment}
+The ASCL actively looks for opportunities to use change management strategies to
+move astronomy to a more transparent discipline; these strategies can be used in
 other disciplines to further change in them.
@@
 developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test'' in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number of key people a project would need to lose to become non-viable (the larger the number, the healthier the project).} \\ \hline
+*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test''
+in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number
+of key people a project would need to lose to become non-viable (the larger the
+number, the healthier the project).} \\ \hline
 ** & rate of change of underlying technologies \\ \hline
@@
 
-The papers that were discussed in the Software Development Experiences Subgroup B are:
+The papers that were discussed in the Software Development Experiences Subgroup
+B are:
 
@@
 \subsubsection{First breakout discussion: Group A}
-The first sub-group discussed issues around the current system for credit and current
-incentives, which it called ``hacking the incentive structure.'' The group
-considered four potential points of leverage:
+The first sub-group discussed issues around the current system for credit and
+current incentives, which it called ``hacking the incentive structure.'' The
+group considered four potential points of leverage:
 
@@
 an example, asking proposers: What are the five canonical citations that are
-most important to your work?~\cite{moore-canonical} This would be a way of giving credit and assigning
-importance to these works, differently from how we just count citations today. A
-possible action that the group discussed was conducting a longitudinal study of
+most important to your work?~\cite{moore-canonical} This would be a way of
+giving credit and assigning importance to these works, differently from how we
+just count citations today. A possible action that the group discussed was
+conducting a longitudinal study of
 most useful \{software, data, etc.\} in a discipline.
 
-Fourth, that we could change the ways funders make decisions, and use these funding
-policies as incentives.
+Fourth, that we could change the ways funders make decisions, and use these
+funding policies as incentives.
 
@@
 acknowledgments sections of papers)
-\item developing a taxonomy of contributors (e.g., Project Credit~\cite{projectcredit}) tied to places that these metrics
-are already stored (e.g., ORCID) \cite{orcid}
+\item developing a taxonomy of contributors (e.g., Project
+Credit~\cite{projectcredit}) tied to places that these metrics are already
+stored (e.g., ORCID) \cite{orcid}
 \item making metadata easier to add for software, creating incentive for
@@
 This group discussed the idea of prizes. Prizes are expected to reframe software
-as ``instrument building'' but will prizes be good or bad, and how can we make sure
-there are no negative affects and the process cannot be gamed?
-
-Prizes in different categories were discussed (like Academy Awards), for example:
-best contribution (non-founder),
-broadest diversity of contributions,
-best tutorials or documentation,
-best leadership transition (award ex leader and new leader),
-best generalization (taking something that was limited and making it more general), and
-best mentorship of contributors (bringing others into the community).
+as ``instrument building'' but will prizes be good or bad, and how can we make
+sure there are no negative affects and the process cannot be gamed?
+
+Prizes in different categories were discussed (like Academy Awards), for
+example: best contribution (non-founder), broadest diversity of contributions,
+best tutorials or documentation, best leadership transition (award ex leader and
+new leader), best generalization (taking something that was limited and making
+it more general), and best mentorship of contributors (bringing others into the
+community).
 
@@
 
-The group suggested that to determine who should give out prizes, perhaps we should find those who we think would
-be awarded prizes, then ask them who they would want to receive a prize from, then
-ask those organizations to see if they are willing to be involved in the process.
-
-One of the group's ideas was to create a funding program for disciplines or other organizations to
-create a prize program.  We would provide a framework and a set of requirements,
-for example: awards to individuals; must
-award in 5 or 6 areas; must have a jury and/or objective criteria that includes
-senior/junior domain people and technology people; must have the recipients awarded
-at a relevant event; and must provide citations that explain why the prizes should be awarded.
-Different organizations could then decide to sign up to the framework and give out
-awards under this general brand.  However, there was a concern that having many
-organizations award prizes may reduce the impact of the prizes.
+The group suggested that to determine who should give out prizes, perhaps we
+should find those who we think would be awarded prizes, then ask them who they
+would want to receive a prize from, then ask those organizations to see if they
+are willing to be involved in the process.
+
+One of the group's ideas was to create a funding program for disciplines or
+other organizations to create a prize program. We would provide a framework and
+a set of requirements, for example: awards to individuals; must award in 5 or 6
+areas; must have a jury and/or objective criteria that includes senior/junior
+domain people and technology people; must have the recipients awarded at a
+relevant event; and must provide citations that explain why the prizes should be
+awarded. Different organizations could then decide to sign up to the framework
+and give out awards under this general brand. However, there was a concern that
+having many organizations award prizes may reduce the impact of the prizes.
 
@@
 
-The group discussion if there should be different criteria for ``established'' members of the
-community versus junior members, and if prizes should be restricted to
-junior members, but left these as open questions.
-
-A point the group considered important is that we don't want to give prizes just to reward people
-who are really good at this one thing, but rather we want to reward people who
-are building the culture we want as scientists.
+The group discussion if there should be different criteria for ``established''
+members of the community versus junior members, and if prizes should be
+restricted to junior members, but left these as open questions.
+
+A point the group considered important is that we don't want to give prizes just
+to reward people who are really good at this one thing, but rather we want to
+reward people who are building the culture we want as scientists.
 
@@
 
-Work by Katz was supported by the National
-Science Foundation while working at the Foundation.  Any opinion, finding, and
-conclusions or recommendations expressed in this material are those of the
-author(s) and do not necessarily reflect the views of the National Science
-Foundation.
+Work by Katz was supported by the National Science Foundation while working at
+the Foundation. Any opinion, finding, and conclusions or recommendations
+expressed in this material are those of the author(s) and do not necessarily
+reflect the views of the National Science Foundation.
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Add a raw summary.

@@
 \item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
-Application Development on GitHub} 
+Application Development on GitHub.} 
 %Affiliation: Idaho National Laboratory
@@
 
-%\item Abani Patra: Execute it
+\item \textbf{Abani Patra: Execute it.}
+%Univ at Buffalo, SUNY
+The primary focus of this presentation was to reinforce the value of
+an available and easily accessible platform for executing scientific software
+e.g. using platforms like Hubzero to access XSEDE or other computing resources.
+Availability of such a platform for executing benchmark problems (even at small
+scale) allows the developer community access to a reference implementation and
+an easy way to train the larger user community. A second focus of this
+presentation was the idea that much more attention needs to be paid to the
+integrated use of simulation tools inside complex workflows. True usability
+needs supporting the user community workflow. \todo{revise}
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
edits in 4 and 11

@@
 repository. To probe the level of governance within open source projects, he and
-his co-authors (Suresh Marru and Chris Mattman) suggested a contest to encourage
+his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
 individual developers to submit patches and pull request to projects that are
@@
 %Univ at Buffalo, SUNY
-The primary focus of this presentation was to reinforce the value of
-an available and easily accessible platform for executing scientific software
-e.g. using platforms like Hubzero to access XSEDE or other computing resources.
-Availability of such a platform for executing benchmark problems (even at small
+The primary purpose of this presentation was to reinforce the value of
+an available and easily accessible platform for executing scientific software,
+e.g., HUBzero, to access XSEDE or other computing resources.
+Such a platform for executing benchmark problems (even at small
 scale) allows the developer community access to a reference implementation and
-an easy way to train the larger user community. A second focus of this
+provides an easy way to train the larger user community. A second focus of this
 presentation was the idea that much more attention needs to be paid to the
@@
 %
-\\ Dr Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
+\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
 \\ & UK researchers write research software w$/$out any formal training &    & 
@@
 % 
-\\ Dr Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
+\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
 \\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
@@
 
-WSSSPE has developed an engaging online mechanism via the online social network
-Twitter (hashtag ``\#WSSSPE''). There were substantially more tweets (messages)
+WSSSPE actively used the online social network
+Twitter, with hashtag ``\#WSSSPE''. There were substantially more tweets (messages)
 during WSSSPE2 than WSSSPE1 or WSSSPE1.1. Some of the main points and highlights
-in the meeting reverberated in Twitter. Table~\ref{tab:tweets} summarizes the
-top tweets happened on the day of workshop, Nov 16, 2014,
+in the meeting are shown in Table~\ref{tab:tweets}, which summarizes the
+top \#WSSSPE tweets from the day of workshop.
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
Section 11: minor changes.

@@
 
-WSSSPE actively used the online social network
-Twitter, with hashtag ``\#WSSSPE''. There were substantially more tweets (messages)
-during WSSSPE2 than WSSSPE1 or WSSSPE1.1. Some of the main points and highlights
-in the meeting are shown in Table~\ref{tab:tweets}, which summarizes the
-top \#WSSSPE tweets from the day of workshop.
+WSSSPE actively used the online social network Twitter, with hashtag
+``\#WSSSPE''. There were substantially more tweets (messages) during the days of
+the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
+18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
+day of the workshop. Some of the main points and highlights in the meeting are
+shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
+the day of workshop, selected by the metrics that number of retweets or
+favourites larger than five and the sum of two measures greater than ten.
 

### 2015-04-19 WSSSPE2_report/WSSSPE2_review.tex
minor fix and adding lightning talk for myself

@@
 
-%\item Daniel S. Katz: Implementing Transitive Credit with JSON-LD
+\item \textbf{Daniel S. Katz: Implementing Transitive Credit with JSON-LD.}
+Science and engineering research increasingly relies on activities that facilitate research but are not currently rewarded or recognized, such as: data sharing; developing common data resources, software and methodologies; and annotating data and publications.
+To promote and advance these activities, we must develop mechanisms for assigning credit, facilitate the appropriate attribution of research outcomes, devise incentives for activities that facilitate research, and allocate funds to maximize return on investment.
+Katz discussed the issue of assigning credit for both direct and indirect contributions by using JSON-LD to implement a prototype transitive credit system~\cite{Katz_transitive_credit_poster}.
 
@@
 Credit~\cite{projectcredit}) tied to places that these metrics are already
-stored (e.g., ORCID) \cite{orcid}
+stored (e.g., ORCID~\cite{orcid})
 \item making metadata easier to add for software, creating incentive for
@@
 citation (but that authors weren't necessarily following those suggestions
-anyway) \cite{howison2015jasist}. He suggested making access to the software
+anyway)~\cite{howison2015jasist}. He suggested making access to the software
 conditional on a license that requires citation. Others found this ``too

### 2015-04-20 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Dan's summary : Format and removed one word "currently".

@@
 \item \textbf{Daniel S. Katz: Implementing Transitive Credit with JSON-LD.}
-Science and engineering research increasingly relies on activities that facilitate research but are not currently rewarded or recognized, such as: data sharing; developing common data resources, software and methodologies; and annotating data and publications.
-To promote and advance these activities, we must develop mechanisms for assigning credit, facilitate the appropriate attribution of research outcomes, devise incentives for activities that facilitate research, and allocate funds to maximize return on investment.
-Katz discussed the issue of assigning credit for both direct and indirect contributions by using JSON-LD to implement a prototype transitive credit system~\cite{Katz_transitive_credit_poster}.
+Science and engineering research increasingly relies on activities that
+facilitate research but are not rewarded or recognized, such as: data
+sharing; developing common data resources, software and methodologies; and
+annotating data and publications. To promote and advance these activities, we
+must develop mechanisms for assigning credit, facilitate the appropriate
+attribution of research outcomes, devise incentives for activities that
+facilitate research, and allocate funds to maximize return on investment. Katz
+discussed the issue of assigning credit for both direct and indirect
+contributions by using JSON-LD to implement a prototype transitive credit
+system~\cite{Katz_transitive_credit_poster}.
 

### 2015-04-20 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Add a raw summary.

@@
 lack of an agreement on how and what to benchmark these algorithms on.
-Ishtiaq from Microsoft Research Cambridge pointed out these problems and made several suggestions to address them.
+Ishtiaq from Microsoft Research Cambridge pointed out these problems and 
+made several suggestions to address them.
 
@@
 
-% \item Jakob Blomer: The Need for a Versioned Data Analysis Environment
+\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment}
+%
+Large-scale scientific endeavors, such as the discovery of the Higgs
+boson at the Large Hadron Collider (LHC), often rely on complex software
+stacks.  Having to maintain thousands of dependencies of software
+libraries and operating system versions, it turns out that despite
+source code availability the setup and the validation of a minimal
+usable analysis environment can easily become prohibitively expensive.
+There is a substantial gap between merely having access to open source
+software and the ability to create a data analysis runtime environment.
+ In high-energy physics, very positive experience was made with a
+special purpose, open-source, versioning and snapshotting file system
+(CernVM-FS), that is used to capture and distribute entire software
+stacks.  It provides the instant access to ready-to-run data analysis
+environments for the global scientific community as well as for
+interested citizens, for instance through the CERN Open Data Portal
+(http://opendata.cern.ch). \todo{revise}
 

### 2015-04-21 WSSSPE2_report/WSSSPE2_review.tex
edits to lightning talks

--- 2015-04-20 WSSSPE2_report/WSSSPE2_review.tex
@@
 %Univ at Buffalo, SUNY
-The primary purpose of this presentation was to reinforce the value of
+Patra discussed the value of
 an available and easily accessible platform for executing scientific software,
@@
 
-% \item Jakob Blomer: The Need for a Versioned Data Analysis Environment
+\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment}
+%
+Large-scale scientific endeavors, such as the discovery of the Higgs
+boson at the Large Hadron Collider (LHC), often rely on complex software
+stacks.  Maintaining thousands of dependencies of software
+libraries and operating system versions has shown that despite
+source code availability, the setup and the validation of a minimal
+usable analysis environment can easily become prohibitively expensive.
+%There is a substantial gap between merely having access to open source
+%software and the ability to create a data analysis runtime environment.
+ In high-energy physics, a
+special purpose, open-source, versioning and snapshotting file system
+(CernVM-FS) that is used to capture and distribute entire software stacks,
+providing instant access to ready-to-run data analysis environments,
+proved to be very useful.
+% for the global scientific community as well as for
+%interested citizens, for instance through the CERN Open Data Portal
+%(http://opendata.cern.ch). \todo{revise}
 

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 1: touch up.
Section 3: touch up.

@@
 peer-review, and peer-grouping processes in detail~\cite{WSSSPE2-pre-report}.
-It is also based on a collaborative set of notes taken through a google doc
+It is also based on a collaborative set of notes taken with Google Docs 
 during the workshop~\cite{WSSSPE2-google-notes}. Overall, the report discusses
@@
 Thaney spoke about Mozilla's work around code citation, through a collaboration
-and prototype crafted between Mozilla, Github, figshare and Zenodo. This work
+and prototype crafted between Mozilla, GitHub, figshare, and Zenodo. This work
 was presented at a closed meeting in May 2014 at the National Institutes of

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 4: rewrite some parts.

@@
 %Affiliation: Idaho National Laboratory
-Peterson from the the Idaho National Laboratory reported that in March 2014, the
+Peterson from the Idaho National Laboratory reported that in March 2014, the
 MOOSE framework was released under an open source license on GitHub,
@@
 and the applications based on the framework are verified before any code changes
-are accepted into the repository. They use a combination of built-in git
+are accepted into the repository. They use a combination of built-in Git
 features such as branching and submodules, GitHub API integration capabilities,
@@
 scale) allows the developer community access to a reference implementation and
-provides an easy way to train the larger user community. A second focus of this
-presentation was the idea that much more attention needs to be paid to the
-integrated use of simulation tools inside complex workflows. True usability
-needs supporting the user community workflow. \todo{revise}
+provides an easy way to train the larger user community. A second idea of this
+presentation was that for true usability, much more attention and support needs to be given to the
+integrated use of simulation tools inside complex workflows.
+%needs supporting the user community workflow.
 
@@
 
-\item \textbf{Daniel S. Katz: Implementing Transitive Credit with JSON-LD.}
+\item \textbf{Daniel Katz: Implementing Transitive Credit with JSON-LD.}
 Science and engineering research increasingly relies on activities that

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 5: touch up.

@@
 using software past its usable life;
-inertia for accepted answers vs wrong/right answers;
+inertia for accepted answers versus wrong or right answers;
 monolithic or poor code; and
@@
 **** & portability \\ \hline
-**** & culture in dev community to support transition between developers \\ \hline
+**** & culture in development community to support transition between developers \\ \hline
 *** & interdisciplinary people: science + IT experience \\ \hline

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 6: minor changes. Other sections: change a double quote to two single quotes.

@@
 without any formal training'' kicked off our discussion. The group brought
-expertise in several different training models, from a 2-day Software
-Carpentry workshop, to a 3-week Open Science for Synthesis (OSS) program
+expertise in several different training models, from a two-day Software
+Carpentry workshop, to a three-week Open Science for Synthesis (OSS) program
 to semester-long programs. We discussed the pros and cons of different
@@
 more complex, we have to raise the skill levels. Skills must be improved
-not only in traditional programming - learning languages and algorithms -
+not only in traditional programming -- learning languages and algorithms --
 but around professional software development. We want to get to a point
-where ``Everyone has a new minimum now - everyone knows git.'' Developers
-also need training on licensing choices. Just because a code is on git
-doesn't mean it's open.
-
-"How do you get people to look for the training they need?'' wondered one
+where ``Everyone has a new minimum now -- everyone knows Git.'' Developers
+also need training on licensing choices. Just because a code is on Git
+does not mean that it is open.
+
+``How do you get people to look for the training they need?'' wondered one
 participant. People seek out opportunities like Software Carpentry to
@@
 
-"How will we know when people are trained effectively in these new
-skills?" We discussed certification. It can be hard to build the
-recognition of Java, database certifications amongst all technologies. OSS
-offers badges to those completing training, however we need to demonstrate
+``How will we know when people are trained effectively in these new
+skills?'' We discussed certification. It can be hard to build the
+recognition of Java or database certifications amongst all technologies. OSS
+offers badges to those completing training. However we need to demonstrate
 proficiency, not just completion. Google Summer of Code is a big CV
-augmentation for participants, can we duplicate something like that?
+augmentation for participants---can we duplicate something similar?
 
 The group also discussed how to fund training. On participant observed
-that NYU runs a 6-week data science training, companies grab graduates but
+that NYU runs a six-week data-science training; companies grab the graduates and
 pay for those they hire. OSS also used the NSF Software Institutes program
@@
 We then discussed variations amongst scientific domain areas and wondered
-"Are some communities more or less open than others?'' To some members of
+``Are some communities more or less open than others?'' To some members of
 the group biology seems to be open, while physics less so. Some felt with
@@
 data repositories. We wondered if a software management plan might be
-effective. "How might funding programs need to adapt to reward good
+effective. ``How might funding programs need to adapt to reward good
 software development practices?'' we asked ourselves. We thought about
@@
 about measuring success through usage statistics (downloads, altmetrics).
-"Should funders demand reproducibility?'' we wondered. In order for results
+``Should funders demand reproducibility?'' we wondered. In order for results
 to be reproduced, software would need to be carefully curated.
@@
 Again, our industry participants contributed unique viewpoints. Partnering
-with industry was seen one path to sustainability. The unique partnerships
+with industry was seen as one path to sustainability. The unique partnerships
 Kitware engages in promote academic freedom while creating an income
@@
 sustainable software frees researchers from performing tasks that do not
-offer the rewards their institution values.'' We talked some about
+offer  the rewards their institutions value. We also discussed  about
 successful models for industry partnerships that preserve open science.
 Participants noted that there are some NSF programs that prohibit
-partnerships with for profit companies (but there are others where this is
+partnerships with for-profit companies (but there are other programs in which this is
 encouraged).
@@
 Subgroup B then focused on actions it could take. It discussed development
-of a white paper that describes a matrixed approach to training
+of a white paper that describes a matrix approach to training
 (multi-day, multi-week, semester). The white paper might include a survey
@@
 approached and assessed, but more is needed. The group felt that this
-training is undervalued and that it's important to communicate the return
-on investment - both individual ROI (skills that make scientists more
+training is undervalued and that it is important to communicate the return
+on investment -- both individual ROI (skills that make scientists more
 effective and more marketable) and funder ROI (better use of taxpayer
@@
 \begin{itemize}
-\item Michael R. Crusoe and C.Titus Brown. Channeling community contributions to
+\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
 scientific software: a hackathon experience~\cite{wssspe2_crusoe}
@@
 that we are not yet clear enough on exactly what we want to incentivize,
-suggesting that we need to have a clearer picture of ``good computational work"
+suggesting that we need to have a clearer picture of ``good computational work''
 and what sort of contributions are truly generative for science.
@@
 There were two main suggestions. The first is to focus on end users of software
-and encourage them to cite software the ``right" way. James Howison suggested
+and encourage them to cite software the ``right'' way. James Howison suggested
 that his research showed that few projects were making a formal request for
@@
 conditional on a license that requires citation. Others found this ``too
-confrontational" and preferred to concentrate on making it easier to do the
+confrontational'' and preferred to concentrate on making it easier to do the
 right thing. The second was focusing on those leading software projects, and the
-group was more enthusiastic about ``forcing" PIs to include their ``software
-people" on publications, although there were few ideas on how exactly to do
+group was more enthusiastic about ``forcing'' PIs to include their ``software
+people'' on publications, although there were few ideas on how exactly to do
 this. Another technique mentioned was that when scientific software projects are
@@
 
-The group also discussed developing ``software prizes" arguing that while it is
-hard to ``mint" other new sources of reputation, prizes are possible without
+The group also discussed developing ``software prizes'' arguing that while it is
+hard to ``mint'' other new sources of reputation, prizes are possible without
 getting too many others on board. The prize criteria can form a template for

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Section 8: minor changes.

@@
 \item This data (the credit file) should be part of the metadata that is outside
-of paywalls, like citations are now.
+of paywalls, like citations are now. \choinote{what are ``paywalls''?}
 \item It was suggested that there should also be a separate file to track
@@
 \item Build a tool that can automatically determine who the contributors are 
-(from a git or other repo), then allows the user to manually edit the output to 
+(from a Git or other repository), then allows the user to manually edit the output to 
 add/remove people, define roles. 
@@
 
-A point the group considered important is that we don't want to give prizes just
+A point the group considered important is that we do not want to give prizes just
 to reward people who are really good at this one thing, but rather we want to

### 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
Add a comma before 'and' for papers with more than two authors.

@@
 \begin{itemize}
-\item Mario {Rosado de Souza} and Robert Haines and Caroline Jay. Defining
+\item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
 Sustainability through Developers' Eyes: Recommendations from an Interview
@@
 
-\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis,
+\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
 Nicholas Weber. Community Recommendations for Sustainable Scientific
@@
 
-\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus and Tevfik Kosar.
+\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
 Role of Online Platforms, Communications and Workflows in Developing Sustainable
@@
 
-\item Marlon Pierce, Suresh Marru and Chris Mattmann.{WSSSPE2}: Patching It Up,
+\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
 Pulling It Forward~\cite{wssspe2_pierce}
@@
 
-\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward and
+\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
 David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Software Development Experiences Subgroup A are:
+The papers that were discussed in the Software Development Experiences Subgroup~A are:
 
@@
 
-\item Marcus Hanwell, Patrick O'Leary and Bob O'Bara. Sustainable Software
+\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable Software
 Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
@@
 \item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
-S. R. Hespanh, R. Idaszak and M. Schildhauer. {ISEES-WSSI} Lessons for
+S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
 Sustainable Science Software from an Early Career Training Institute on Open
@@
 
-The papers that were discussed in the Software Development Experiences Subgroup
-B are:
+The papers that were discussed in the Software Development Experiences Subgroup~B are:
 
@@
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
-Hutton, Gregory Tucker and Erkan Istanbulluoglu. Landlab: Sustainable Software
+Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
 Development in Practice ~\cite{wssspe2_adams}
@@
 
-\item Carl Boettiger, Ted Hart, Scott Chamberlain and Karthik Ram. Building
+\item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
 software, building community: lessons from the {ROpenSci}
@@
 
-\item Yolanda Gil, Eunyoung Moon and James Howison. No Science Software is an
+\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software is an
 Island: Collaborative Software Development Needs in
@@
 Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
-Rnc Maia and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
+Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
 Sustainable Data and Software~\cite{wssspe2_habermann}
 
-\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner and James Syvitski.
+\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
 Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
@@
 \item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W.
-M. C. Foulkes, James J. Shepherd and Alex J. W. Thom. The {H}ighly {A}ccurate
+M. C. Foulkes, James J. Shepherd, and Alex J. W. Thom. The {H}ighly {A}ccurate
 {N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic
@@
 
-\item Tom Crick, Benjamin A. Hall and Samin Ishtiaq. ``Can I Implement Your
+\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I Implement Your
 Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
 
-\item Bryan Marker, Don Batory, Field G. Van Zee and Robert van de Geijn. Making
+\item Bryan Marker, Don Batory, Field G. Van Zee, and Robert van de Geijn. Making
 Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
@@
 \begin{itemize}
-\item Thomas Clune, Michael Rilee and Damian Rouson. Testing as an Essential
+\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an Essential
 Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
@@
 \item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann,
-David Andrs and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
+David Andrs, and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
 Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}

### 2015-04-27 WSSSPE2_report/WSSSPE2_review.tex
Sections 3 and 4: Add two bib items. Modified a third bib item.

@@
 The workshop featured two keynote addresses. In the opening keynote
-presentation, Kaitlin Thaney of the Mozilla Science Lab talked about her
+presentation~\cite{Thaney_slides}, Kaitlin Thaney of the Mozilla Science Lab talked about her
 organization's work and policy to enable and support sustainable and
 reproducible scientific research through the open web. The second keynote
-speaker was Neil Chue Hong of Software Sustainability Institute. He shined a
+speaker was Neil Chue Hong of Software Sustainability Institute. His talk~\cite{Hong_slides} shined a
 light on how scientific software is prevalently driving advances in many science

### 2015-04-27 WSSSPE2_report/WSSSPE2_review.tex
Section 4: Added references for all talks.

--- 2015-04-25 WSSSPE2_report/WSSSPE2_review.tex
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability}
+\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability~\cite{Thaney_slides}}
 \label{keynote1}
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Neil Chue Hong, We are the 92$\%$}
+\subsection{Neil Chue Hong, We are the 92$\%$~\cite{Hong_slides}}
 \label{keynote2}
@@
 community by cultivating better, more sustainable research software to enable
-world-class research. Chue Hong's keynote\footnote{Slides available at:
-http://dx.doi.org/10.6084/m9.figshare.1243288} started by making the point of
+world-class research. Chue Hong's keynote started by making the point of
 that the use of -- and reliance on -- software is pervasive in all areas of
@@
 
-\begin{enumerate}
 %\item Greg Wilson: Close Enough for Scientific Work
 
-\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
-through Sustainable Software Architectures.}
+\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures~\cite{Venters_poster}}
 Venters proposed that sustainable software is a composite, first-class,
@@
 
-\item \textbf{Marlon Pierce: Patching It Up, Pulling It Forward.}
+\subsection{Marlon Pierce: Patching It Up, Pulling It Forward~\cite{Pierce_poster}}
 Pierce discussed how open open source is. Open software needs a diverse, openly
@@
 
-\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
-Application Development on GitHub.} 
+\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
+Application Development on GitHub~\cite{Peterson_poster}} 
 %Affiliation: Idaho National Laboratory
@@
 
-\item \textbf{Abani Patra: Execute it.}
+\subsection{Abani Patra: Execute it~\cite{Patra_poster}}
 %Univ at Buffalo, SUNY
@@
 
-\item \textbf{Daniel Katz: Implementing Transitive Credit with JSON-LD.}
+\subsection{Daniel Katz: Implementing Transitive Credit with JSON-LD~\cite{Katz_transitive_credit_poster}}
 Science and engineering research increasingly relies on activities that
@@
 contributions by using JSON-LD to implement a prototype transitive credit
-system~\cite{Katz_transitive_credit_poster}.
-
-\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software.}
+system.
+
+\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software~\cite{Ishtiaq_poster}}
 %  Microsoft Research Cambridge
@@
 
-\item \textbf{James Howison: Retract all Bit-Rotten Publications.}
+\subsection{James Howison: Retract all Bit-Rotten Publications~\cite{Howison_poster}}
 Howison sought to provoke discussion by proposing that papers whose workflows are
@@
 
-\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices.}
+\subsection{Robert Downs: Community Recommendations for Improving Sustainable
+Scientific Software Practices~\cite{Downs_poster}}
 Robert Downs, of the Columbia University Center for International Earth
@@
 
-\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
-Diverse Community.}
+\subsection{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
+Diverse Community~\cite{Boettiger_poster}}
 Boettiger described how the rOpenSci project has been successful by focusing not
@@
 
-\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment}
+\subsection{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}}
 %
@@
 
-\item \textbf{Alice Allen: Find it! Cite it!}
+\subsection{Alice Allen: Find it! Cite it!~\cite{Allen_poster}}
 %
@@
 \end{comment}
-\end{enumerate}
 

### 2015-04-28 WSSSPE2_report/WSSSPE2_review.tex
BIb and Section 4: Simplified.

@@
 
-\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software~\cite{Ishtiaq_poster}}
+\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software}%~\cite{Ishtiaq_poster}}
 %  Microsoft Research Cambridge
@@
 
-\subsection{James Howison: Retract all Bit-Rotten Publications~\cite{Howison_poster}}
+\subsection{James Howison: Retract all Bit-Rotten Publications}%~\cite{Howison_poster}}
 Howison sought to provoke discussion by proposing that papers whose workflows are
@@
 
-\subsection{Alice Allen: Find it! Cite it!~\cite{Allen_poster}}
+\subsection{Alice Allen: Find it! Cite it!}%~\cite{Allen_poster}}
 %

### 2015-04-28 WSSSPE2_report/WSSSPE2_review.tex
Added authors affliations and emails.

@@
 %
-\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA}
-%
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA}
-%
-\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA}
-%
-\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK}
-%
-\thanks{{}$^{(5)}$ TBD}
-%
-\thanks{{}$^{(6)}$ TBD}  
-%
-\thanks{{}$^{(7)}$ TBD}   
+\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
+%
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+%
+\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
+%
+\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
+%
+\thanks{{}$^{(5)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
+%
+\thanks{{}$^{(6)}$ Princeton University, Princeton, NJ, USA; \url{todo}}  
+%
+\thanks{{}$^{(7)}$ TODO; \url{todo}}   
 %  

### 2015-04-28 WSSSPE2_report/WSSSPE2_review.tex
Add in missing periods at the end of two lists.

@@
 \item determining guidelines for recommending software characteristics for
-tenure - perhaps draft guidelines then get ACM or IEEE agreement
+tenure - perhaps draft guidelines then get ACM or IEEE agreement.
 \end{itemize}
@@
 \item guidelines for evaluating scientific contribution through software (perhaps 
-using new metrics)
+using new metrics).
 \end{itemize}

### 2015-04-30 WSSSPE2_report/WSSSPE2_review.tex
Added First breakout discussion notes from Reproducibility group

@@
 
-\todo{short intro to group here}
+This group discussed five papers with a wide variety ideas of how to support reproducibility and reuse.
+It focussed on identifying concrete that the attendees could work together on, which would have a positive effect on the community.
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+In the first discussion session, this group broke up into two smaller groups working on different topics: Reproducibility, and Reuse and Sharing.
+
+\subsubsection{First breakout discussion: Reproducibility group}
+
+The first group discussed ways in which reproducibility of papers could be improved.
+A consensus surrounded the provision of examples: demonstrating to others how to achieve reproducibility.
+Major public funding investments go into research that heavily relies on software to be reproducible, hence lack thereof is raising major concerns.
+
+Two main avenues could be used to implement policy that would improve reproducibility and drive top-down culture change: 
+\begin{itemize}
+\item Funders can aim to get more software expertise on to funding review panels and provide more guidance on what is required of software related to research (cf. data management plans)
+\item Journals can define and publish journal policies to improve reproducibility, and reviewers can insist that authors provide sufficient information and access to data and software to allow them reproduce the results in the paper.
+Stronger policies even for some high-impact journals have recently come into place, for example Nature Publishing journals.
+\end{itemize}
+
+If journals do enforce greater reproducibility constraints, it is important to lower the barriers to reviewers attempting to verifying the correctness of the software used to generate the results.
+A major issue is that a lot of software only builds on certain systems.
+Should journals provide more tools / support for reviewers, and if so, what is it?
+An idea that came from the groups was to define a set of support services that should be available to software paper reviewers.
+Another was to provide the ability for reviewers to flag the requirement for a ?software verification?, similar to ability to flag that a paper needs to be seen by a statistician.
+
+Other discussion focussed on ways in which researchers themselves could improve the reproducibility of research.
+One way would be to establish tracks at conferences that subject papers to reproduction, which for those that pass would lend them an additional badge (similar to OOPSLA).
+Another is simply to get more people to use your software: for instance by outreach to high school students - can your software be used by them?
+Finally, there is a role for community-curated benchmarks to validate the performance and capabilities of tools.
+
+\subsubsection{First breakout discussion: Reuse and Sharing group}
 

### 2015-04-30 WSSSPE2_report/WSSSPE2_review.tex
Added notes on discussion group on reuse and sharing

@@
 In the first discussion session, this group broke up into two smaller groups working on different topics: Reproducibility, and Reuse and Sharing.
+In the second discussion session, three groups worked on creating specific pieces of guidance.
 
@@
 \subsubsection{First breakout discussion: Reuse and Sharing group}
+
+The second group discussed ways in which software could be more easily reused and shared.
+From the perspective of both user and developer, any solutions must be 1) easy; 2) cheap; 3) not take much time.
+
+Reuse and sharing were considered to be distinct but linked.
+In many cases, pre-existing software does not exist, so new software is written but even then it is not shared afterwards.
+The principal barriers to reuse are the difficulty of finding out what software is available, understanding if it is usable, and then of installing and running software if it is located.
+
+Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable.
+Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of "the community"?
+Some communities have had significant initiatives to improve software discovery, e.g.
+the NIH Software Discovery Index\footnote{NIH Software Discovery Index: http://softwarediscoveryindex.org/}.
+Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.
+
+An issue around the sustainability of software catalogues is that their usefulness often depends on the domain.
+For instance, in the bio-sciences there is more homogenous data, and more standard shred codes.
+In areas like ecology, codes are very specific to the problem, meaning that the level of re-use might be at a very general statistical level of abstraction, but then every research use is highly customized.
+
+Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website.
+Using code repositories like GitHub\footnote{GitHub: http://github.com/} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: http://figshare.com/} or Zenodo\footnote{Zenodo: http://zenodo.org/}.
+Docker\footnote{Docker: htp://www.docker.com/} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.
+
+The key enabler for reuse and sharing was to get domain scientists more effectively connected with programmers/analysts.
+Both have skills and experience which is necessary to make the right decisions for improving the reusability and discoverability of software, and to apply community pressure to change practice.
+
+\subsubsection{Categorisation of software journals}
+
+\subsubsection{What should journals provide reviewers of software papers}
+
+\subsubsection{Reproducibility Meta-track toolkit}
 

### 2015-04-30 WSSSPE2_report/WSSSPE2_review.tex
Added notes on second discussion sessions of Reproducibility, Reuse and Sharing breakout

@@
 
-\subsubsection{Categorisation of software journals}
-
-\subsubsection{What should journals provide reviewers of software papers}
-
-\subsubsection{Reproducibility Meta-track toolkit}
+\subsubsection{Second breakout discussion: Categorisation of software journals}
+
+This group aimed to come up with a categorisation of journals which published software papers.
+Starting from the list of journals\footnote{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers.
+These were the Journal of Open Research Software\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}, PLoS ONE\footnote{http://www.plosone.org/static/guidelines\#software}, Journal of Statistical Software\footnote{http://www.jstatsoft.org/instructions}.
+Methods in Ecology and Evolution\footnote{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}, Transactions of Mathematical Software\footnote{http://toms.acm.org/Authors.html}, GigaScience\footnote{http://www.gigasciencejournal.com/about}, and PLoS Computational Biology\footnote{http://www.ploscompbiol.org/static/guidelines\#software}.
+
+From these a set of common categories were synthesized, against which all journals could be compared:
+
+\begin{itemize}
+\item Journal Policies
+\begin{itemize}
+	\item Accessibility of papers
+	\begin{itemize}
+		\item Open Access
+		\item ``Freely'' available
+	\end{itemize}
+	\item Repositories
+	\begin{itemize}
+		\item Provides suggestions for recommended repositories
+		\item Provides its own repository
+	\end{itemize}
+	\item Review
+	\begin{itemize}
+		\item Reviewing software is mandatory
+		\begin{itemize}
+			\item Must check that software runs
+			\item Must check quality of code
+			\item Must check performance of code if paper makes claims on relative performance
+		\end{itemize}
+	\end{itemize}	
+	\item Supporting Data
+	\begin{itemize}
+		\item Must be publicly available
+		\item Must be in an Open Access repository
+		\item must have a DOI
+	\end{itemize}
+	\item Article Processing Charge 
+	\begin{itemize}
+		\item APC charge is transparent
+		\item APC waiver programme
+	\end{itemize}
+\end{itemize}
+
+\item Paper Policies
+\begin{itemize}
+	\item Required Sections
+	\item Keywords
+	\begin{itemize}
+		\item Paper provides keywords to help describe software
+	\end{itemize}
+	\item Papers can be updated when new releases of software are releases
+	\begin{itemize}
+		\item At no extra cost / at significantly reduced cost
+	\end{itemize}
+\end{itemize}
+
+\item Software Policies
+\begin{itemize}
+	\item Software must have a license
+	\begin{itemize}
+		\item Software must have an open source license
+	\end{itemize}
+	\item Availability
+	\begin{itemize}
+		\item Software must be openly available and accessible
+	\end{itemize}
+	\item Deposit policies
+	\begin{itemize}
+		\item Software should be in a public repository
+		\begin{itemize}
+			\item Of particular stature / with preservation plan
+		\end{itemize}
+		\item Software should have a permanent identifier
+		\item Software deposit doesn?t count as a prior publication
+	\end{itemize}
+	\item Runnability and dependencies
+	\begin{itemize}
+		\item Provide documentation to understand how to run
+		\item Provide sample data
+		\item Provide dependency information
+	\end{itemize}
+\end{itemize}
+\end{itemize}
+
+The follow-up actions to this work are to use this set of categories on all the journals in the list, refining the categories if necessary, then identify if any of the categories are seen to be more useful to promote reproducibility, reuse and sharing.
+
+\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers}
+
+This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a``Software Paper''.
+
+Journals should provide guidelines about what to consider when reviewing a software submission.
+A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies}.
+Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions.
+Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.
+
+Journals should provide mechanisms to enable and track communication between reviewers and authors.
+For anonymous peer review journals, maintain anonymity of reviewers.
+If communication is necessary, that may mean that software is not that well documented.
+Journals should also provide a set of simple metrics for software evaluation that reviewers can use for ratings, similar to Consumer Reports.
+
+Journals should provide guidelines about requirements for documentation of code: both in-lined in code, and manuals/web pages, etc.
+Journal editors could provide documentation checks before it goes out for review.
+This should include a requirement for good Use Cases specified for the software, with references to executable test cases that demonstrate each use-case is met (at least in the form of the test case).
+
+Journals should support mechanisms to run software.
+Sometimes this may be very hard to accomplish, despite best efforts.
+Journals could provide an Execution environment for any Software submitted for review, perhaps via a Docker container or a virtual machine.
+If not, instructions must be adequate to compile and execute the software; and interpret the results (output files, formats, etc.), and the full source code must be accessible to the reviewer.
+Mechanisms to quantify what has changed compared to a previously published version would assist version comparison.
+However, this cannot simply take lines of code into account.
+For example, a speedup of an algorithm may not result in a huge code difference, but may nonetheless provide a break-through.
+Journals should require software submissions to also provide ?Test materials?-- sample data, parameters to validate that code is working as intended.
+In certain cases, well-selected benchmark datasets may be required to assess performance and accuracy.
+
+Journals should ensure minimal metadata are provided, similarly as is already the case for certain kinds of data (though the latter is often enforced by data repositories)-- Dublin core-ish (creator, owner), and more specific (platform and compiler dependencies, sample benchmarks of performance, etc.).
+Journals should provide guidance or constraints as to software licensing conditions.
+
+The follow-up actions to this work are to work with journals and reviewers to identify whether any of these suggestions can be easily provided, perhaps for a range of journals.
+
+\subsubsection{Second breakout discussion: Reproducibility Meta-track toolkit}
+
+This group worked on defining a ``toolkit'' for running a reproducibility meta-track at a conference. They decided to take the work done during the workshop and publish it as a paper.
 

### 2015-05-05 WSSSPE2_report/WSSSPE2_review.tex
Minor changes.

@@
 (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on  16 November 2014 in conjunction with the International
+held on  16 November, 2014 in conjunction with the International
 Conference for High Performance Computing, Networking, Storage and Analysis
@@
 
-The review process was fairly standard, where reviewers bid for papers, then an
-automated system matched bids to determine assignments, and reviewers then
+The review process was fairly standard. First, reviewers bid for papers. Then an
+automated system matched the bids to determine assignments. Next, reviewers 
 completed their assigned reviews (with an average of 4.9 reviews per paper, and
@@
 scores on relevance and comments to the organizers, who used the information to
-decide which papers to associate with the workshop, and provided the comments to
+decide which papers to associate with the workshop and provided the comments to
 the authors to help them improve their papers.
@@
 mechanisms, such as how easy it is for independent developers to communicate
-with project leadership, how projects accept and properly license third party
+with project leadership, how projects accept and properly license third-party
 contributions, and how projects make decisions such as granting source tree
@@
 group. Because of the large number, we split these into two subgroups
-prior to the event. Some common themes helped in this division, for
-example several papers that addressed education and training issues,
+prior to the event. Some common themes helped in this division. For
+example, several papers that addressed education and training issues,
 including best practices and case studies, were grouped
@@
 Web content in their field on the nanoHUB site.'' Fast forward seven years
-later where, in a 2014 iSGTW article, quote nanoHUB PI Gerhard Klimeck, ``A
+later where, in a 2014 iSGTW article, quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A
 former student of mine published eight tools on nanoHUB, serving over
@@
 We then moved beyond our training discussion to address funding that
-encourages sustainable software -- funding of both people and of projects
+encourages sustainable software -- funding of both people and projects
 that create a true system to support sustainability. We called this
-``institutionalized serendipity.'' Because science is increasingly reliant
+``institutionalized serendipity.'' As science is increasingly reliant
 on software, one participant observed that ``software development can have

### 2015-05-06 WSSSPE2_report/WSSSPE2_review.tex
Break down a long sentence.

@@
 The review process was fairly standard. First, reviewers bid for papers. Then an
-automated system matched the bids to determine assignments. Next, reviewers 
-completed their assigned reviews (with an average of 4.9 reviews per paper, and
-4.1 reviews per reviewer). This process was done through
-EasyChair\footnote{http://easychair.org/}, which allowed reviewers to provide
-scores on relevance and comments to the organizers, who used the information to
+automated system matched the bids to determine assignments. After the reviewers 
+completed their assigned reviews (with an average of 4.9 reviews per paper and
+4.1 reviews per reviewer), they used
+EasyChair\footnote{http://easychair.org/} to record
+scores on relevance and comments. Finally, the organizers accessed the information to
 decide which papers to associate with the workshop and provided the comments to

### 2015-05-06 WSSSPE2_report/WSSSPE2_review.tex
Reformat. No content change.

@@
 funds provided to the conference organizers by the Moore Foundation and the
-National Science Foundation (NSF), to support the travel of potential participants who
-would not otherwise be able to attend, and young participants and those from
-underrepresented groups, respectively.  These funds allowed \todo{number}
-additional participants to attend, and each was offered the chance to present a
-lightning talk.
+National Science Foundation (NSF), to support the travel of potential
+participants who would not otherwise be able to attend, and young participants
+and those from underrepresented groups, respectively. These funds allowed
+\todo{number} additional participants to attend, and each was offered the chance
+to present a lightning talk.
 
 This report extends a previous report that discussed the submission,
-peer-review, and peer-grouping processes in detail~\cite{WSSSPE2-pre-report}.
-It is also based on a collaborative set of notes taken with Google Docs 
-during the workshop~\cite{WSSSPE2-google-notes}. Overall, the report discusses
-the organization work done before the workshop (\S\ref{sec:preworkshop}); the
+peer-review, and peer-grouping processes in detail~\cite{WSSSPE2-pre-report}. It
+is also based on a collaborative set of notes taken with Google Docs during the
+workshop~\cite{WSSSPE2-google-notes}. Overall, the report discusses the
+organization work done before the workshop (\S\ref{sec:preworkshop}); the
 keynotes (\S\ref{sec:keynotes}); a series of lightning talks
@@
 completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used
-EasyChair\footnote{http://easychair.org/} to record
-scores on relevance and comments. Finally, the organizers accessed the information to
-decide which papers to associate with the workshop and provided the comments to
-the authors to help them improve their papers.
+4.1 reviews per reviewer), they used EasyChair\footnote{http://easychair.org/}
+to record scores on relevance and comments. Finally, the organizers accessed the
+information to decide which papers to associate with the workshop and provided
+the comments to the authors to help them improve their papers.
 
@@
 or more of the themes from the call for papers, with a few paper authors invited
-to summarize the other papers in them as a panel, followed by general
-discussion about that theme. The mapping of papers to themes was done by the
-organizers.
+to summarize the other papers in them as a panel, followed by general discussion
+about that theme. The mapping of papers to themes was done by the organizers.
 
@@
 Sorted\footnote{\url{http://www.well-sorted.org}} was used in the following
-steps:
-\begin{enumerate}
-\item Authors were asked to create Well Sorted ``cards'' for the papers. These
-cards have a title (50 characters maximum) and a body (255 characters maximum).
-\item Authors, program committee members, and members of the WSSSPE mailing list
-were asked to sort the cards. Each person drags the cards, one by one, into
-groups. A group can have as many cards as the person wants it to have, and it
-can have whatever meaning that makes sense to that person.
-\item Well Sorted  produces a set of averages of all the sorts, with various
-numbers of card clusters.
-\end{enumerate}
+steps: \begin{enumerate} \item Authors were asked to create Well Sorted
+``cards'' for the papers. These cards have a title (50 characters maximum) and a
+body (255 characters maximum). \item Authors, program committee members, and
+members of the WSSSPE mailing list were asked to sort the cards. Each person
+drags the cards, one by one, into groups. A group can have as many cards as the
+person wants it to have, and it can have whatever meaning that makes sense to
+that person. \item Well Sorted produces a set of averages of all the sorts, with
+various numbers of card clusters. \end{enumerate}
 
@@
 and engineering fields. Both keynote speeches spawned further discussion among
-workshop participants on the crucial notion \emph{sustainability} in the theme of
-our workshop.
+workshop participants on the crucial notion \emph{sustainability} in the theme
+of our workshop.
 

### 2015-05-06 WSSSPE2_report/WSSSPE2_review.tex
Reformat. No content change.

@@
 \href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}
-\end{comment}
-
-\begin{comment}
+
 Lighting talks, breakout groups, working groups are all different. Lightning
@@
 
-Email addresses of speakers:
-gvwilson@software-carpentry.org, Colin Venters <C.Venters@hud.ac.uk>,
-j.spencer@imperial.ac.uk, erinrobinson@esipfed.org, marpierc@iu.edu,
-jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org, dsk@uchicago.edu,
-Samin.Ishtiaq@microsoft.com, james@howison.name, s.a.harris@leeds.ac.uk,
-marcus.hanwell@kitware.com, thabermann@hdfgroup.org, rdowns@ciesin.columbia.edu,
-cboettig@gmail.com, jakob.blomer@cern.ch, editor@ascl.net 
+Email addresses of speakers: gvwilson@software-carpentry.org, Colin Venters
+<C.Venters@hud.ac.uk>, j.spencer@imperial.ac.uk, erinrobinson@esipfed.org,
+marpierc@iu.edu, jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org,
+dsk@uchicago.edu, Samin.Ishtiaq@microsoft.com, james@howison.name,
+s.a.harris@leeds.ac.uk, marcus.hanwell@kitware.com, thabermann@hdfgroup.org,
+rdowns@ciesin.columbia.edu, cboettig@gmail.com, jakob.blomer@cern.ch,
+editor@ascl.net
 \end{comment}
@@
 mostly dedicated to discussions, the organizers wanted to give the attendees a
-chance to also `make a pitch' for an idea, either representing a contributed paper
-or something different.  Eighteen attendees volunteered to participate in the lightning
-talks, each given only two minutes to speak and at most one slide to show.
-The talks were presented in reverse alphabetic order of speakers' last names. 
-In the rest of this section, we highlight the gist of some of the speakers' messages.
+chance to also `make a pitch' for an idea, either representing a contributed
+paper or something different. Eighteen attendees volunteered to participate in
+the lightning talks, each given only two minutes to speak and at most one slide
+to show. The talks were presented in reverse alphabetic order of speakers' last
+names. In the rest of this section, we highlight the gist of some of the
+speakers' messages.
 
@@
 
-\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable Software
-through Sustainable Software Architectures~\cite{Venters_poster}}
+\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
+Software through Sustainable Software Architectures~\cite{Venters_poster}}
 Venters proposed that sustainable software is a composite, first-class,
@@
 
-\subsection{Marlon Pierce: Patching It Up, Pulling It Forward~\cite{Pierce_poster}}
-Pierce discussed how open open source is. Open software needs a diverse, openly
-governed community behind it, just as it needs open licensing and a public code
-repository. To probe the level of governance within open source projects, he and
-his co-authors (Suresh Marru and Chris Mattmann) suggested a contest to encourage
-individual developers to submit patches and pull request to projects that are
-important to them. This simple mechanism will expose several governance
-mechanisms, such as how easy it is for independent developers to communicate
-with project leadership, how projects accept and properly license third-party
-contributions, and how projects make decisions such as granting source tree
-write access.
-
-\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework and
-Application Development on GitHub~\cite{Peterson_poster}} 
-%Affiliation: Idaho National Laboratory
-Peterson from the Idaho National Laboratory reported that in March 2014, the
-MOOSE framework was released under an open source license on GitHub,
-significantly expanding and diversifying the pool of current active and
-potential future contributors on the project. The MOOSE team employs an
-extensive continuous integration test suite to ensure that both the framework
+\subsection{Marlon Pierce: Patching It Up, Pulling It
+Forward~\cite{Pierce_poster}} Pierce discussed how open open source is. Open
+software needs a diverse, openly governed community behind it, just as it needs
+open licensing and a public code repository. To probe the level of governance
+within open source projects, he and his co-authors (Suresh Marru and Chris
+Mattmann) suggested a contest to encourage individual developers to submit
+patches and pull request to projects that are important to them. This simple
+mechanism will expose several governance mechanisms, such as how easy it is for
+independent developers to communicate with project leadership, how projects
+accept and properly license third-party contributions, and how projects make
+decisions such as granting source tree write access.
+
+\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework
+and Application Development on GitHub~\cite{Peterson_poster}} %Affiliation:
+Idaho National Laboratory Peterson from the Idaho National Laboratory reported
+that in March 2014, the MOOSE framework was released under an open source
+license on GitHub, significantly expanding and diversifying the pool of current
+active and potential future contributors on the project. The MOOSE team employs
+an extensive continuous integration test suite to ensure that both the framework
 and the applications based on the framework are verified before any code changes
@@
 %Univ at Buffalo, SUNY
-Patra discussed the value of
-an available and easily accessible platform for executing scientific software,
-e.g., HUBzero, to access XSEDE or other computing resources.
-Such a platform for executing benchmark problems (even at small
+Patra discussed the value of an available and easily accessible platform for
+executing scientific software, e.g., HUBzero, to access XSEDE or other computing
+resources. Such a platform for executing benchmark problems (even at small
 scale) allows the developer community access to a reference implementation and
 provides an easy way to train the larger user community. A second idea of this
-presentation was that for true usability, much more attention and support needs to be given to the
-integrated use of simulation tools inside complex workflows.
+presentation was that for true usability, much more attention and support needs
+to be given to the integrated use of simulation tools inside complex workflows.
 %needs supporting the user community workflow.
 
-%\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gap in
-%Early Career Training
-
-\subsection{Daniel Katz: Implementing Transitive Credit with JSON-LD~\cite{Katz_transitive_credit_poster}}
-Science and engineering research increasingly relies on activities that
-facilitate research but are not rewarded or recognized, such as: data
-sharing; developing common data resources, software and methodologies; and
-annotating data and publications. To promote and advance these activities, we
-must develop mechanisms for assigning credit, facilitate the appropriate
-attribution of research outcomes, devise incentives for activities that
-facilitate research, and allocate funds to maximize return on investment. Katz
-discussed the issue of assigning credit for both direct and indirect
-contributions by using JSON-LD to implement a prototype transitive credit
+%\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gapin %Early Career Training
+
+\subsection{Daniel Katz: Implementing Transitive Credit with
+JSON-LD~\cite{Katz_transitive_credit_poster}} Science and engineering research
+increasingly relies on activities that facilitate research but are not rewarded
+or recognized, such as: data sharing; developing common data resources, software
+and methodologies; and annotating data and publications. To promote and advance
+these activities, we must develop mechanisms for assigning credit, facilitate
+the appropriate attribution of research outcomes, devise incentives for
+activities that facilitate research, and allocate funds to maximize return on
+investment. Katz discussed the issue of assigning credit for both direct and
+indirect contributions by using JSON-LD to implement a prototype transitive
+credit
 system.
@@
 include both the inability to re-implement novel algorithms and approaches, and
-lack of an agreement on how and what to benchmark these algorithms on.
-Ishtiaq from Microsoft Research Cambridge pointed out these problems and 
-made several suggestions to address them.
+lack of an agreement on how and what to benchmark these algorithms on. Ishtiaq
+from Microsoft Research Cambridge pointed out these problems and made several
+suggestions to address them.
 
 \subsection{James Howison: Retract all Bit-Rotten Publications}%~\cite{Howison_poster}}
-Howison sought to provoke discussion by proposing that papers whose workflows are
-not kept current with the changing software ecosystem should be automatically
-retracted. This would create an incentive for authors to keep their
-software current and usable, rather than the current situation in which every
-potential user has to do this individually. A softer version of the proposal
-would identify papers whose software workflow has become bit-rotten and allow
-others to keep the code up to date, either adding them as new authors of the
-paper or providing credit in some other form for their academic service.
+Howison sought to provoke discussion by proposing that papers whose workflows
+are not kept current with the changing software ecosystem should be
+automatically retracted. This would create an incentive for authors to keep
+their software current and usable, rather than the current situation in which
+every potential user has to do this individually. A softer version of the
+proposal would identify papers whose software workflow has become bit-rotten and
+allow others to keep the code up to date, either adding them as new authors of
+the paper or providing credit in some other form for their academic service.
 
@@
 Scientific Software Practices~\cite{Downs_poster}}
-Robert Downs, of the Columbia University Center for International Earth
-Science Information Network (CIESIN), 
+Robert Downs, of the Columbia University Center for International Earth Science
+Information Network (CIESIN),
 %presented Community Recommendations for
@@
 % Davis, and Nicholas Weber of 
-the Science Software Cluster (SSC) of the
-Federation of Earth Science Information Partners (ESIP). For the study, almost
-300 attendees of the 2014 Summer ESIP Meeting were invited to participate in
-simultaneous roundtable discussions on sustainability of science software. Over
-two-thirds of the roundtable focus groups responded to a semi-structured survey
-that contained three sets of questions eliciting recommendations for near-term
-actions of the community to improve sustainable software practices. Initial
-analysis of the
+the Science Software Cluster (SSC) of the Federation of Earth Science
+Information Partners (ESIP). For the study, almost 300 attendees of the 2014
+Summer ESIP Meeting were invited to participate in simultaneous roundtable
+discussions on sustainability of science software. Over two-thirds of the
+roundtable focus groups responded to a semi-structured survey that contained
+three sets of questions eliciting recommendations for near-term actions of the
+community to improve sustainable software practices. Initial analysis of the
 participants' responses to the questionnaire revealed several suggestions, which
@@
 understanding and awareness, and creating incentives to motivate sustainable
-science software practices.
-%Based on the recommendations
-%offered by the focus group respondents for near-term actions, 
-The ESIP SSC plans to engage the community %to achieve progress in the near future 
-in the recommended activities for improving sustainable scientific
-software practices.
+science software practices. The ESIP SSC plans to engage the community in the
+recommended activities for improving sustainable scientific software practices.
 
@@
 %
-Large-scale scientific endeavors, such as the discovery of the Higgs
-boson at the Large Hadron Collider (LHC), often rely on complex software
-stacks.  Maintaining thousands of dependencies of software
-libraries and operating system versions has shown that despite
-source code availability, the setup and the validation of a minimal
-usable analysis environment can easily become prohibitively expensive.
-%There is a substantial gap between merely having access to open source
-%software and the ability to create a data analysis runtime environment.
- In high-energy physics, a
-special purpose, open-source, versioning and snapshotting file system
-(CernVM-FS) that is used to capture and distribute entire software stacks,
-providing instant access to ready-to-run data analysis environments,
-proved to be very useful.
+Large-scale scientific endeavors, such as the discovery of the Higgs boson at
+the Large Hadron Collider (LHC), often rely on complex software stacks.
+Maintaining thousands of dependencies of software libraries and operating system
+versions has shown that despite source code availability, the setup and the
+validation of a minimal usable analysis environment can easily become
+prohibitively expensive. In high-energy physics, a special purpose, open-source,
+versioning and snapshotting file system (CernVM-FS) that is used to capture and
+distribute entire software stacks, providing instant access to ready-to-run data
+analysis environments, proved to be very useful.
 % for the global scientific community as well as for

### 2015-05-07 WSSSPE2_report/WSSSPE2_review.tex
Section 2: A few minor edits. Section 3: format change only.

@@
 information to decide which papers to associate with the workshop and provided
-the comments to the authors to help them improve their papers.
+authors with the comments to help them improve their papers.
 
@@
 or more of the themes from the call for papers, with a few paper authors invited
-to summarize the other papers in them as a panel, followed by general discussion
+to summarize the other papers in the session as a panel, followed by a general discussion
 about that theme. The mapping of papers to themes was done by the organizers.
@@
 Finally, since some of the papers were not represented by cards in the process,
-they were not placed in groups by the peer-grouping system; the authors of
+they were not placed in groups by the peer-grouping system. The authors of
 these papers were asked which groups seemed the best for their papers---these
@@
 \subsection{Robert Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices~\cite{Downs_poster}}
-Robert Downs, of the Columbia University Center for International Earth Science
-Information Network (CIESIN),
-%presented Community Recommendations for
-%Improving Sustainable Scientific Software Practices (2014), and 
-described a
-focus group study conducted with %W. Christopher Lenhardt, Erin Robinson, Ethan
-% Davis, and Nicholas Weber of 
-the Science Software Cluster (SSC) of the Federation of Earth Science
-Information Partners (ESIP). For the study, almost 300 attendees of the 2014
-Summer ESIP Meeting were invited to participate in simultaneous roundtable
-discussions on sustainability of science software. Over two-thirds of the
-roundtable focus groups responded to a semi-structured survey that contained
-three sets of questions eliciting recommendations for near-term actions of the
-community to improve sustainable software practices. Initial analysis of the
-participants' responses to the questionnaire revealed several suggestions, which
-included improving community engagement and collaborative activities, increasing
-understanding and awareness, and creating incentives to motivate sustainable
-science software practices. The ESIP SSC plans to engage the community in the
-recommended activities for improving sustainable scientific software practices.
+Scientific Software Practices~\cite{Downs_poster}} Robert Downs, of the Columbia
+University Center for International Earth Science Information Network (CIESIN),
+described a focus group study conducted with the Science Software Cluster (SSC)
+of the Federation of Earth Science Information Partners (ESIP). For the study,
+almost 300 attendees of the 2014 Summer ESIP Meeting were invited to participate
+in simultaneous roundtable discussions on sustainability of science software.
+Over two-thirds of the roundtable focus groups responded to a semi-structured
+survey that contained three sets of questions eliciting recommendations for
+near-term actions of the community to improve sustainable software practices.
+Initial analysis of the participants' responses to the questionnaire revealed
+several suggestions, which included improving community engagement and
+collaborative activities, increasing understanding and awareness, and creating
+incentives to motivate sustainable science software practices. The ESIP SSC
+plans to engage the community in the recommended activities for improving
+sustainable scientific software practices.
 

### 2015-05-07 WSSSPE2_report/WSSSPE2_review.tex
Section 4: clean up, small edits, and format changes.

@@
 
-\subsection{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
-Software through Sustainable Software Architectures~\cite{Venters_poster}}
+\begin{enumerate}
+\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
+Software through Sustainable Software Architectures~\cite{Venters_poster}.}
 Venters proposed that sustainable software is a composite, first-class,
@@
 
-\subsection{Marlon Pierce: Patching It Up, Pulling It
-Forward~\cite{Pierce_poster}} Pierce discussed how open open source is. Open
+\item \textbf{Marlon Pierce: Patching It Up, Pulling It
+Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
 software needs a diverse, openly governed community behind it, just as it needs
 open licensing and a public code repository. To probe the level of governance
-within open source projects, he and his co-authors (Suresh Marru and Chris
-Mattmann) suggested a contest to encourage individual developers to submit
-patches and pull request to projects that are important to them. This simple
-mechanism will expose several governance mechanisms, such as how easy it is for
+within open source projects, he and his co-authors (S.~Marru and C.~Mattmann) 
+suggested a contest to encourage individual developers to submit
+patches and requests to projects that are important to them. This simple
+mechanism shall expose several governance mechanisms, such as how easy it is for
 independent developers to communicate with project leadership, how projects
-accept and properly license third-party contributions, and how projects make
+accept and license third-party contributions, and how projects make
 decisions such as granting source tree write access.
 
-\subsection{John Peterson: Continuous Integration for Concurrent MOOSE Framework
-and Application Development on GitHub~\cite{Peterson_poster}} %Affiliation:
-Idaho National Laboratory Peterson from the Idaho National Laboratory reported
+\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework
+and Application Development on GitHub~\cite{Peterson_poster}.}   
+Peterson from the Idaho National Laboratory reported
 that in March 2014, the MOOSE framework was released under an open source
@@
 
-\subsection{Abani Patra: Execute it~\cite{Patra_poster}}
+\item \textbf{Abani Patra: Execute it~\cite{Patra_poster}.}
 %Univ at Buffalo, SUNY
-Patra discussed the value of an available and easily accessible platform for
-executing scientific software, e.g., HUBzero, to access XSEDE or other computing
-resources. Such a platform for executing benchmark problems (even at small
-scale) allows the developer community access to a reference implementation and
+Patra discussed the value of an easily accessible platform for
+executing scientific software, e.g., HUBzero to access XSEDE or other computing
+resources. Such a platform for executing benchmark problems (even at a small
+scale) allows the developer community access a reference implementation and
 provides an easy way to train the larger user community. A second idea of this
-presentation was that for true usability, much more attention and support needs
+talk was that for true usability, much more attention and support needs
 to be given to the integrated use of simulation tools inside complex workflows.
-%needs supporting the user community workflow.
 
@@
 
-\subsection{Daniel Katz: Implementing Transitive Credit with
-JSON-LD~\cite{Katz_transitive_credit_poster}} Science and engineering research
+\item \textbf{Daniel Katz: Implementing Transitive Credit with
+JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
 increasingly relies on activities that facilitate research but are not rewarded
@@
 indirect contributions by using JSON-LD to implement a prototype transitive
-credit
-system.
-
-\subsection{Samin Ishtiaq: Daemons, Notifications and Sustaining Software}%~\cite{Ishtiaq_poster}}
+credit system.
+
+\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software. }%~\cite{Ishtiaq_poster}}
 %  Microsoft Research Cambridge
@@
 
-\subsection{James Howison: Retract all Bit-Rotten Publications}%~\cite{Howison_poster}}
+\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
 Howison sought to provoke discussion by proposing that papers whose workflows
@@
 
-\subsection{Robert Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices~\cite{Downs_poster}} Robert Downs, of the Columbia
+\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
+Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the Columbia
 University Center for International Earth Science Information Network (CIESIN),
@@
 
-\begin{comment}
-Downs RR, Lenhardt WC, Robinson E, Davis E, Weber N. 2014. Community
-Recommendations for Improving Sustainable Scientific Software Practices. 2nd
-Workshop on Sustainable Software for Science: Practice and Experiences
-(WSSSPE2), New Orleans, Louisiana, 16 November 2014.
-http://wssspe.researchcomputing.org.uk/wp-content/uploads/2014/11/Downs.pdf and
-http://dx.doi.org/10.7916/D8Q52NBC
-\end{comment}
-
-\subsection{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
-Diverse Community~\cite{Boettiger_poster}}
+\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
+Diverse Community~\cite{Boettiger_poster}.}
 Boettiger described how the rOpenSci project has been successful by focusing not
@@
 
-\subsection{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}}
-%
+\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}.}
 Large-scale scientific endeavors, such as the discovery of the Higgs boson at
@@
 validation of a minimal usable analysis environment can easily become
-prohibitively expensive. In high-energy physics, a special purpose, open-source,
-versioning and snapshotting file system (CernVM-FS) that is used to capture and
-distribute entire software stacks, providing instant access to ready-to-run data
-analysis environments, proved to be very useful.
+prohibitively expensive. In high-energy physics, CernVM-FS, a special-purpose, open-source,
+versioning, and snapshotting file system used to capture and
+distribute entire software stacks, proved to be useful for providing instant access to ready-to-run data
+analysis environments.
 % for the global scientific community as well as for
@@
 
-\subsection{Alice Allen: Find it! Cite it!}%~\cite{Allen_poster}}
-%
+\item \textbf{Alice Allen: Find it! Cite it! }%~\cite{Allen_poster}}
 The Astrophysics Source Code Library (ASCL) is an online registry of
@@
 their software and is an excellent way to drive community change.
-
 \begin{comment}
@@
 \end{comment}
+\end{enumerate}
+
+\smallskip
 
@@
 Of the short actionable papers that would lead to improvements for sustainable
-software science, 11 submissions were categorized in the  Software Development Experiences
-group. Because of the large number, we split these into two subgroups
-prior to the event. Some common themes helped in this division. For
-example, several papers that addressed education and training issues,
-including best practices and case studies, were grouped
-together. Others discussed experiences with registries, developer
-collectives and specific examples of successful, sustainable software
-(including a valuable industry perspective).
+software science, 11 submissions were categorized in the Software Development
+Experiences group. Because of the large number, we split these into two
+subgroups prior to the event. Some common themes helped in this division. For
+example, several papers that addressed education and training issues, including
+best practices and case studies, were grouped together. Others discussed
+experiences with registries, developer collectives and specific examples of
+successful, sustainable software (including a valuable industry perspective).
 
@@
 
-Subgroup B discussed four papers and had six
-participants. Because of the nature of the papers, training emerged as a
-common theme, however the conversation was wide-ranging, including
-incentives, reproducibility and funding to promote sustainability.
+Subgroup B discussed four papers and had six participants. Because of the nature
+of the papers, training emerged as a common theme, however the conversation was
+wide-ranging, including incentives, reproducibility and funding to promote
+sustainability.
 
@@
 \subsubsection{Subgroup B Discussion}
-The statement ``applied computer science is being attempted in academia
-without any formal training'' kicked off our discussion. The group brought
-expertise in several different training models, from a two-day Software
-Carpentry workshop, to a three-week Open Science for Synthesis (OSS) program
-to semester-long programs. We discussed the pros and cons of different
-training approaches, also touching on informal learning, for example where
-people learning the necessary skills because they had cross-disciplinary
-people (``boundary scientists'') in their work environments of whom they
-could ask questions.
+The statement ``applied computer science is being attempted in academia without
+any formal training'' kicked off our discussion. The group brought expertise in
+several different training models, from a two-day Software Carpentry workshop,
+to a three-week Open Science for Synthesis (OSS) program to semester-long
+programs. We discussed the pros and cons of different training approaches, also
+touching on informal learning, for example where people learning the necessary
+skills because they had cross-disciplinary people (``boundary scientists'') in
+their work environments of whom they could ask questions.
 
 Some papers explored gaps in training of early career scientists. Industry
-participants in our group confirm this. We asked ourselves, ``Are
-traditional courses failing?'' We think yes. Changes to undergraduate
-curricula requirements are difficult. But as programming models become
-more complex, we have to raise the skill levels. Skills must be improved
-not only in traditional programming -- learning languages and algorithms --
-but around professional software development. We want to get to a point
-where ``Everyone has a new minimum now -- everyone knows Git.'' Developers
-also need training on licensing choices. Just because a code is on Git
-does not mean that it is open.
+participants in our group confirm this. We asked ourselves, ``Are traditional
+courses failing?'' We think yes. Changes to undergraduate curricula requirements
+are difficult. But as programming models become more complex, we have to raise
+the skill levels. Skills must be improved not only in traditional programming --
+learning languages and algorithms -- but around professional software
+development. We want to get to a point where ``Everyone has a new minimum now --
+everyone knows Git.'' Developers also need training on licensing choices. Just
+because a code is on Git does not mean that it is open.
 
 ``How do you get people to look for the training they need?'' wondered one
-participant. People seek out opportunities like Software Carpentry to
-augment skills. While some instruction can be done in an institutional
-curricula, independent groups (non-profits, institutes) have more
-flexibility. Some asked whether Silicon Valley be interested in funding
-training so that people are better prepared to enter the workforce? Some
-felt that companies were reluctant to deliver training because employees
-then leave the company. While others felt that pushing all training to
-industry could lead to good technical people ``getting on the Google bus''
-and leaving the sciences.
-
-``How will we know when people are trained effectively in these new
-skills?'' We discussed certification. It can be hard to build the
-recognition of Java or database certifications amongst all technologies. OSS
-offers badges to those completing training. However we need to demonstrate
-proficiency, not just completion. Google Summer of Code is a big CV
-augmentation for participants---can we duplicate something similar?
-
-The group also discussed how to fund training. On participant observed
-that NYU runs a six-week data-science training; companies grab the graduates and
-pay for those they hire. OSS also used the NSF Software Institutes program
-as a vehicle to fund training activities. Software Carpentry uses a
-collaborative teaching approach where  people publish open teaching
-materials and receive credit for their use.
-
-We then discussed career paths for those supporting sustainable software.
-While tenure track is not the only option for students, the challenges for
-those who remain in academia can be large. Research scientists are
-entirely dependent on soft money, which can be unpredictable. Postdocs and
-those who do pursue tenure track positions need to publish and see no
-rewards in software development. These challenges were all identified at
-WSSSPE1. What actions would we recommend to improve things? Altmetrics and
-download statistics may slowly change the system and improving a
-developer's ability receive credit for time invested in software
-development. NSF recognition of data products (including software) as well
-as publications is also helping.
+participant. People seek out opportunities like Software Carpentry to augment
+skills. While some instruction can be done in an institutional curricula,
+independent groups (non-profits, institutes) have more flexibility. Some asked
+whether Silicon Valley be interested in funding training so that people are
+better prepared to enter the workforce? Some felt that companies were reluctant
+to deliver training because employees then leave the company. While others felt
+that pushing all training to industry could lead to good technical people
+``getting on the Google bus'' and leaving the sciences.
+
+``How will we know when people are trained effectively in these new skills?'' We
+discussed certification. It can be hard to build the recognition of Java or
+database certifications amongst all technologies. OSS offers badges to those
+completing training. However we need to demonstrate proficiency, not just
+completion. Google Summer of Code is a big CV augmentation for
+participants---can we duplicate something similar?
+
+The group also discussed how to fund training. On participant observed that NYU
+runs a six-week data-science training; companies grab the graduates and pay for
+those they hire. OSS also used the NSF Software Institutes program as a vehicle
+to fund training activities. Software Carpentry uses a collaborative teaching
+approach where people publish open teaching materials and receive credit for
+their use.
+
+We then discussed career paths for those supporting sustainable software. While
+tenure track is not the only option for students, the challenges for those who
+remain in academia can be large. Research scientists are entirely dependent on
+soft money, which can be unpredictable. Postdocs and those who do pursue tenure
+track positions need to publish and see no rewards in software development.
+These challenges were all identified at WSSSPE1. What actions would we recommend
+to improve things? Altmetrics and download statistics may slowly change the
+system and improving a developer's ability receive credit for time invested in
+software development. NSF recognition of data products (including software) as
+well as publications is also helping.
 
 We asked ourselves, ``Where are there examples of things that are changing
-because of this and how can we build momentum?'' One example demonstrates
-change over time. In 2007, nanoHUB listed the academic reward structure as
-a problem in an EDUCAUSE report where the authors note, ``In the future,
-nanoHUB researchers are hoping to change the research culture. While they
-recognize that young faculty members are unlikely to get tenure based on
-their nanoHUB contributions, they hope to encourage faculty to think
-beyond their own research needs to consider publishing tutorials and other
-Web content in their field on the nanoHUB site.'' Fast forward seven years
-later where, in a 2014 iSGTW article, quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A
-former student of mine published eight tools on nanoHUB, serving over
-6,000 people with his tools. He then joined a university as a professor
-and introduced nanoHUB. Use of the gateway from that university
-skyrocketed; he used nanoHUB in existing classes, created new classes, and
-infused it in his research. Ultimately, the professor's department head
-associates his two-year rise to tenure with the notoriety and innovation
-he gained through nanoHUB.''
-
-The group also discussed ``attribution trees,'' an idea put forward by Dan
-Katz and Arfon Smith where a chain of attributions can, for example, give
-credit to those developing libraries and building blocks that other pieces
-of software use. The group considered where to push the use of this, which
-journals? On participant noted that Dryad works with the journals. If a
-paper is accepted, the supporting data must be submitted. We discussed
-reproducibility as a component of the journal review process and ``active
-papers,'' with immediate links to the data and software.
-
-We then discussed variations amongst scientific domain areas and wondered
-``Are some communities more or less open than others?'' To some members of
-the group biology seems to be open, while physics less so. Some felt with
-the more recent development of bioinformatics as a field there were fewer
-historical practices to undo. Physics has pre-print philosophy to
-overcome. Environmental sciences may be mixed. The group felt the
-biomedical area, however, was very competitive and closed. One participant
-mentioned blueprints for going open source (like NWChem recently did)
-where authors outline how this helps, what you do and what the next steps
-are.
-
-We then moved beyond our training discussion to address funding that
-encourages sustainable software -- funding of both people and projects
-that create a true system to support sustainability. We called this
-``institutionalized serendipity.'' As science is increasingly reliant
-on software, one participant observed that ``software development can have
-much broader impact than publishing individual research, but it is not
-viewed that way.'' Because of this centrality, one participant mentioned
-that training ought to be called Science Carpentry rather than Software
-Carpentry. We felt we were beginning to see changes in the research
-community as a result of the NSF's data management plan requirements. PIs 
-are thinking more about data and some university libraries are offering
-data repositories. We wondered if a software management plan might be
-effective. ``How might funding programs need to adapt to reward good
-software development practices?'' we asked ourselves. We thought about
-best practices such as version control, test harnesses, mailing lists, bug
-tracking, community contribution, and reuse where appropriate. We thought
-about measuring success through usage statistics (downloads, altmetrics).
-``Should funders demand reproducibility?'' we wondered. In order for results
-to be reproduced, software would need to be carefully curated.
-
-Again, our industry participants contributed unique viewpoints. Partnering
-with industry was seen as one path to sustainability. The unique partnerships
-Kitware engages in promote academic freedom while creating an income
-stream from certain portions of the software. This type of approach to
-sustainable software frees researchers from performing tasks that do not
-offer  the rewards their institutions value. We also discussed  about
-successful models for industry partnerships that preserve open science.
-Participants noted that there are some NSF programs that prohibit
-partnerships with for-profit companies (but there are other programs in which this is
-encouraged).
+because of this and how can we build momentum?'' One example demonstrates change
+over time. In 2007, nanoHUB listed the academic reward structure as a problem in
+an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
+are hoping to change the research culture. While they recognize that young
+faculty members are unlikely to get tenure based on their nanoHUB contributions,
+they hope to encourage faculty to think beyond their own research needs to
+consider publishing tutorials and other Web content in their field on the
+nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
+quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
+mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
+He then joined a university as a professor and introduced nanoHUB. Use of the
+gateway from that university skyrocketed; he used nanoHUB in existing classes,
+created new classes, and infused it in his research. Ultimately, the professor's
+department head associates his two-year rise to tenure with the notoriety and
+innovation he gained through nanoHUB.''
+
+The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
+and Arfon Smith where a chain of attributions can, for example, give credit to
+those developing libraries and building blocks that other pieces of software
+use. The group considered where to push the use of this, which journals? On
+participant noted that Dryad works with the journals. If a paper is accepted,
+the supporting data must be submitted. We discussed reproducibility as a
+component of the journal review process and ``active papers,'' with immediate
+links to the data and software.
+
+We then discussed variations amongst scientific domain areas and wondered ``Are
+some communities more or less open than others?'' To some members of the group
+biology seems to be open, while physics less so. Some felt with the more recent
+development of bioinformatics as a field there were fewer historical practices
+to undo. Physics has pre-print philosophy to overcome. Environmental sciences
+may be mixed. The group felt the biomedical area, however, was very competitive
+and closed. One participant mentioned blueprints for going open source (like
+NWChem recently did) where authors outline how this helps, what you do and what
+the next steps are.
+
+We then moved beyond our training discussion to address funding that encourages
+sustainable software -- funding of both people and projects that create a true
+system to support sustainability. We called this ``institutionalized
+serendipity.'' As science is increasingly reliant on software, one participant
+observed that ``software development can have much broader impact than
+publishing individual research, but it is not viewed that way.'' Because of this
+centrality, one participant mentioned that training ought to be called Science
+Carpentry rather than Software Carpentry. We felt we were beginning to see
+changes in the research community as a result of the NSF's data management plan
+requirements. PIs are thinking more about data and some university libraries are
+offering data repositories. We wondered if a software management plan might be
+effective. ``How might funding programs need to adapt to reward good software
+development practices?'' we asked ourselves. We thought about best practices
+such as version control, test harnesses, mailing lists, bug tracking, community
+contribution, and reuse where appropriate. We thought about measuring success
+through usage statistics (downloads, altmetrics). ``Should funders demand
+reproducibility?'' we wondered. In order for results to be reproduced, software
+would need to be carefully curated.
+
+Again, our industry participants contributed unique viewpoints. Partnering with
+industry was seen as one path to sustainability. The unique partnerships Kitware
+engages in promote academic freedom while creating an income stream from certain
+portions of the software. This type of approach to sustainable software frees
+researchers from performing tasks that do not offer the rewards their
+institutions value. We also discussed about successful models for industry
+partnerships that preserve open science. Participants noted that there are some
+NSF programs that prohibit partnerships with for-profit companies (but there are
+other programs in which this is encouraged).
 
@@
 
-Subgroup B then focused on actions it could take. It discussed development
-of a white paper that describes a matrix approach to training
-(multi-day, multi-week, semester). The white paper might include a survey
-of existing techniques. There are many, some dating back many years, for
-example the Interuniversity Consortium for Political and Social Research
-(ICPSR) and various summer institutes. The white paper could Include a
-call for a comprehensive assessment of these techniques. We need to think
-carefully about the right venue for such a white paper, where would it
-have the most impact. The subgroup believes it would need to approach editors
-directly to ascertain this.
-
-The group felt that training in techniques that promote sustainability has
-a range of benefits: career paths, educated reviewers, reproducible
-science and more. There is some information on how that has been
-approached and assessed, but more is needed. The group felt that this
-training is undervalued and that it is important to communicate the return
-on investment -- both individual ROI (skills that make scientists more
-effective and more marketable) and funder ROI (better use of taxpayer
-funds, research more likely to be reproducible because sustainable
-software exists, better trained reviewers).
+Subgroup B then focused on actions it could take. It discussed development of a
+white paper that describes a matrix approach to training (multi-day, multi-week,
+semester). The white paper might include a survey of existing techniques. There
+are many, some dating back many years, for example the Interuniversity
+Consortium for Political and Social Research (ICPSR) and various summer
+institutes. The white paper could Include a call for a comprehensive assessment
+of these techniques. We need to think carefully about the right venue for such a
+white paper, where would it have the most impact. The subgroup believes it would
+need to approach editors directly to ascertain this.
+
+The group felt that training in techniques that promote sustainability has a
+range of benefits: career paths, educated reviewers, reproducible science and
+more. There is some information on how that has been approached and assessed,
+but more is needed. The group felt that this training is undervalued and that it
+is important to communicate the return on investment -- both individual ROI
+(skills that make scientists more effective and more marketable) and funder ROI
+(better use of taxpayer funds, research more likely to be reproducible because
+sustainable software exists, better trained reviewers).
 

### 2015-05-07 WSSSPE2_report/WSSSPE2_review.tex
Section 1: add location of WSSSPE2.

@@
 (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on  16 November, 2014 in conjunction with the International
+held on  16 November, 2014 in the city of New Orleans, Louisiana, USA, in conjunction with the International
 Conference for High Performance Computing, Networking, Storage and Analysis

### 2015-05-08 WSSSPE2_report/WSSSPE2_review.tex
Section 1: Wrap lines.
Sections 3 to 7:  touch up.

@@
 (WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on  16 November, 2014 in the city of New Orleans, Louisiana, USA, in conjunction with the International
-Conference for High Performance Computing, Networking, Storage and Analysis
+held on 16 November, 2014 in the city of New Orleans, Louisiana, USA, in
+conjunction with the International Conference for High Performance Computing,
+Networking, Storage and Analysis
 (SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
@@
 Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the 
-United Kingdom. The SSI was founded to support the UK's research software 
+United Kingdom (UK). The SSI was founded to support the UK's research software 
 community by cultivating better, more sustainable research software to enable
-world-class research. Chue Hong's keynote started by making the point of
+world-class research. Chue Hong's keynote started by making the point 
 that the use of -- and reliance on -- software is pervasive in all areas of
 world-leading research, showing examples from disciplines as diverse as
-humanities and high-energy physics, and quoting Kersten Kleese Van Dam of the
+humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
 Pacific Northwestern National Laboratory via a petition
@@
 Chue Hong drew attention to the issue that in the cyber-infrastructure and
-high-performance community, we all too often disregard the hundreds of thousands
-of researchers developing software thinking they are the long tail. In fact, the
-numbers point to the fact that they are actually the mainstream and we
-(referring to people including the audience of the talk) are the elite minority.
+high-performance community, hundreds of thousands
+of researchers developing software are all too often disregarded or considered the long tail. 
+Actually, the numbers point to the fact that they are the mainstream. %and we
+%(referring to people including the audience of the talk) are the elite minority.
 He emphasized that software is no longer special; it is both essential to and
@@
 universities (406 respondents covering a representative range of funders,
-disciplines, and seniority) reported that 92\% confirmed the use of research
+disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
 software and 89\% affirmed that it would be impossible or difficult to conduct
@@
 research software (which equates to over 140,000 researchers if extrapolated
-across the UK) and yet 71\% had no formal software-development training, having
+across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
 to rely on their own coding skills.
@@
 Examining another aspect of the size of the research software community, 
-Chue Hong noted that the costs of software-reliant research in the United Kingdom
+Chue Hong noted that the costs of software-reliant research in the UK
 included \textsterling 840 million of investment in the financial year
@@
 4\% of positions were software development related, and of these only 17\% were
-explicitly named as a software developer / software engineer position: the vast
-majority being advertised as research associate / research assistant positions.
+explicitly named as a software developer or software engineer positions: the vast
+majority being advertised as research associate or research assistant positions.
 This in turn leads to the issue of career paths for those bridging the research
@@
 \item determine three things they considered to be significant barriers to
-software sustainability
+software sustainability.
 \end{enumerate}
@@
 Subgroup B discussed four papers and had six participants. Because of the nature
-of the papers, training emerged as a common theme, however the conversation was
-wide-ranging, including incentives, reproducibility and funding to promote
+of the papers, training emerged as a common theme. However the conversation was
+wide-ranging, including incentives, reproducibility, and funding to promote
 sustainability.
@@
 several different training models, from a two-day Software Carpentry workshop,
-to a three-week Open Science for Synthesis (OSS) program to semester-long
-programs. We discussed the pros and cons of different training approaches, also
-touching on informal learning, for example where people learning the necessary
-skills because they had cross-disciplinary people (``boundary scientists'') in
-their work environments of whom they could ask questions.
-
-Some papers explored gaps in training of early career scientists. Industry
-participants in our group confirm this. We asked ourselves, ``Are traditional
+a three-week Open Science for Synthesis (OSS) program, to a semester-long
+programs. We discussed the pros and cons of different training approaches, 
+touching on informal learning, for example, where people learning the necessary
+skills by asking questions of cross-disciplinary people (``boundary scientists'') in
+their work environments.
+
+Some papers explored gaps in training of early-career scientists. Industry
+participants in our group confirmed this observation. We asked ourselves, ``Are traditional
 courses failing?'' We think yes. Changes to undergraduate curricula requirements
@@
 the skill levels. Skills must be improved not only in traditional programming --
-learning languages and algorithms -- but around professional software
+learning languages and algorithms -- but also around professional software
 development. We want to get to a point where ``Everyone has a new minimum now --
@@
 participant. People seek out opportunities like Software Carpentry to augment
-skills. While some instruction can be done in an institutional curricula,
+skills. While some instructions can be done in institutional curricula,
 independent groups (non-profits, institutes) have more flexibility. Some asked
-whether Silicon Valley be interested in funding training so that people are
-better prepared to enter the workforce? Some felt that companies were reluctant
-to deliver training because employees then leave the company. While others felt
+whether Silicon Valley would be interested in funding training so that people are
+better prepared to enter the workforce. Some felt that companies were reluctant
+to deliver training for fear that their employees might then leave them. While others felt
 that pushing all training to industry could lead to good technical people
-``getting on the Google bus'' and leaving the sciences.
+``getting on the Google bus'' and leaving sciences.
 
@@
 completion. Google Summer of Code is a big CV augmentation for
-participants---can we duplicate something similar?
+participants---can we create something similar?
 
@@
 to fund training activities. Software Carpentry uses a collaborative teaching
-approach where people publish open teaching materials and receive credit for
+approach where people publish open teaching materials and receive credits for
 their use.
@@
 to improve things? Altmetrics and download statistics may slowly change the
-system and improving a developer's ability receive credit for time invested in
-software development. NSF recognition of data products (including software) as
+system and improve a developer's ability to receive credits for time invested in
+software development. NSF's recognition of data products (including software) as
 well as publications is also helping.
@@
 they hope to encourage faculty to think beyond their own research needs to
-consider publishing tutorials and other Web content in their field on the
+consider publishing tutorials and other content in their fields on the
 nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,

### 2015-05-09 WSSSPE2_report/WSSSPE2_review.tex
Section 3.1: Add URLs as footnotes.  Section 7: minor changes. Add a note.

@@
 
-Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla).
-Mozilla is a non-profit organization interested in openness, in news, in website
-creation, and in Science, all taking advantage of the open web.
+Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla), which
+is a non-profit organization interested in openness, news, website
+creation, and Science, all taking advantage of the open web.
 
@@
 million for management. More than $70$ tools were created, but caBIG is still
-seen as a failure\footnote{\url{http://tinyurl.com/maf6dz2}}. Those that had the
-least investment were the most used; the most invested software were the least
-utilized.
+seen as a failure\footnote{Report Blasts Problem-Plagued Cancer Research Grid,
+\url{http://tinyurl.com/maf6dz2}}. Those that had the least investment were the
+most used; the most invested software were the least utilized.
 
@@
 around what a \emph{Software Discovery
-Index}\footnote{\url{http://softwarediscoveryindex.org/report/}} might look
+Index}\footnote{Software Discovery Index, \url{http://softwarediscoveryindex.org/report/}} might look
 like. The meeting included a number of publishers, researchers, and those behind
-major scientific software efforts such as Bioconductor, Galaxy, and
-nanoHUB. 
+major scientific software efforts such as 
+Bioconductor{\footnote{Bioconductor, \url{http://www.bioconductor.org}}, 
+Galaxy\footnote{Galaxy, \url{http://galaxyproject.org}}, and
+nanoHUB\footnote{nanoHUB, \url{https://nanohub.org}}. 
 %facilitates more efficient scientific research. SDI identifies scientific
@@
 We then discussed career paths for those supporting sustainable software. While
-tenure track is not the only option for students, the challenges for those who
+tenure track is not the only option for graduate students, the challenges for those who
 remain in academia can be large. Research scientists are entirely dependent on
@@
 system and improve a developer's ability to receive credits for time invested in
-software development. NSF's recognition of data products (including software) as
-well as publications is also helping.
+software development. NSF's recognition of scientific products including datasets, 
+software, and publications is also helping.
 
@@
 created new classes, and infused it in his research. Ultimately, the professor's
-department head associates his two-year rise to tenure with the notoriety and
+department head attributed his two-year rise to tenure with the reputation and
 innovation he gained through nanoHUB.''
@@
 those developing libraries and building blocks that other pieces of software
-use. The group considered where to push the use of this, which journals? On
-participant noted that Dryad works with the journals. If a paper is accepted,
-the supporting data must be submitted. We discussed reproducibility as a
-component of the journal review process and ``active papers,'' with immediate
-links to the data and software.
+use. The group considered potential journals and medium to push the application
+of this idea. One participant noted that Dryad\footnote{The Dryad Digital
+Repository, \url{http://datadryad.org}.} works with journals. If a paper is
+accepted, the supporting data must be submitted. We discussed reproducibility as
+a component of the journal review process and ``active papers,'' with immediate
+links to the data and software. \choinote{Connect the last two sentences to the preceeding materials.}
 

### 2015-05-22 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex

@@
 \begin{enumerate}
-\item \textbf{Colin Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
+\item \textbf{Colin C. Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
 Software through Sustainable Software Architectures~\cite{Venters_poster}.}
@@
 \todo{short intro to group here}
-
+Six papers were categorized under the theme of Exploring Sustainability.
+The group included four of the authors from the six papers submitted and an additional number of participants who expressed an interest in the theme at the workshop.
+Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of communities to defining sustainability as a first-class, composite non-functional requirement.
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

### 2015-05-22 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex

@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
+\subsubsection{Discussion}
+
+Each author was invited to outline the key action from their paper as a potential discussion point for the group; in their absence, the group facilitators outlined the actionable points from papers where the authors were not present.
+The key actions from the six papers included recommendations for improving practice in sustainable software engineering ~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service as a mechanism for fostering sustainable science communities ~\cite{wssspe2_patra}, developer incentives for code contributions to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in scalability, reproducibility, and energy efficiency ~\cite{wssspe2_shi}, and applying software architectures as a mechanism for architectural-level reasoning about sustainability ~\cite{wssspe2_venters}.
+
+The group took a position to view sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of reusable software~cite{}.
+
+The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
+It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research programme would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software.
+This type of initiative would provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime of the project.
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

### 2015-05-22 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex
lots of little changes throughout

--- 2015-05-22 WSSSPE2_report/WSSSPE2_review.tex
@@
 \author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
-\\James Howison$^{(5)}$, Miguel de Val-Borro$^{(6)}$, Richard Littauer$^{(7)}$, more TBD %by contributions 
-\\{\scriptsize\today \quad  (Due:  $\backsim$ April 30)}} 
+\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Miguel de Val-Borro$^{(8)}$, Richard Littauer$^{(9)}$, more TBD by contributions}
 %\author{Gabrielle Allen$^{(11)}$,  \\
 %Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-%Matthew Turk$^{(2)}$, Colin C. Venters$^{(7)}$,  Miguel de Val-Borro$^{(10)}$}
+%Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
 %
@@
 %
-\thanks{{}$^{(5)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
-%
-\thanks{{}$^{(6)}$ Princeton University, Princeton, NJ, USA; \url{todo}}  
-%
-\thanks{{}$^{(7)}$ TODO; \url{todo}}   
+\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+%
+\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
+%
+\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands \url{F.Seinstra@esciencecenter.nl}}   
+%
+\thanks{{}$^{(8)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
+%
+\thanks{{}$^{(9)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
 %  
@@
 %\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
-%
-%\thanks{{}$^{(7)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
 %
@@
 and those from underrepresented groups, respectively. These funds allowed
-\todo{number} additional participants to attend, and each was offered the chance
+17 additional participants to attend, and each was offered the chance
 to present a lightning talk.
@@
 
-\item \textbf{Daniel Katz: Implementing Transitive Credit with
+\item \textbf{Daniel S. Katz: Implementing Transitive Credit with
 JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Colin to lead writing of this section.}
+\todo{Colin led writing of this section.}
 
@@
 
-\todo{short intro to group here}
-Six papers were categorized under the theme of Exploring Sustainability.
-The group included four of the authors from the six papers submitted and an additional number of participants who expressed an interest in the theme at the workshop.
-Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of communities to defining sustainability as a first-class, composite non-functional requirement.
+Six papers were categorized under the theme of Exploring Sustainability. The group included four of the authors 
+from the six papers submitted and an additional number of participants who expressed an interest in the theme at the 
+workshop. Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of 
+communities to defining sustainability as a first-class, composite non-functional requirement.
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+\subsubsection{Discussion}
+
+Each author was invited to outline the key action from their paper as a potential discussion point for the group; in 
+their absence, the group facilitators outlined the actionable points from papers where the authors were not present. 
+The key actions from the six papers included recommendations for improving practice in sustainable software 
+engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service 
+as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions 
+to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in 
+scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism 
+for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took a position to view 
+sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of 
+reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards 
+developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies 
+funding research projects where software was an intrinsic part of enabling the research program would be an additional 
+element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would 
+provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime 
+of the project.
+
+\subsubsection{Actions}
+
+The main action to come from the group was a proposal to identify the ten best software engineering practices 
+similar to Philip Bourne's approach of Ten Simple Rules~\cite{bourne-ten-simple-rules}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Nancy to lead writing of this section}
+\todo{Nancy led writing of this section}
 
@@
 Repository, \url{http://datadryad.org}.} works with journals. If a paper is
-accepted, the supporting data must be submitted. We discussed reproducibility as
+accepted to one of those journals, the supporting data must be submitted to Dryad.
+The group also discussed reproducibility as
 a component of the journal review process and ``active papers,'' with immediate
-links to the data and software. \choinote{Connect the last two sentences to the preceeding materials.}
+links to the data and software.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan leading writing of this section}
+\todo{Dan led writing of this section}
 
@@
 \item This idea could also be applied to data
-\item This data (the credit file) should be part of the metadata that is outside
-of paywalls, like citations are now. \choinote{what are ``paywalls''?}
+\item This data (the credit file) should be part of the metadata that are freely
+available to those who have not paid for access to the journal, like citations are now in most cases.
 \item It was suggested that there should also be a separate file to track
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Neil OR VOLUNTEER to lead writing of this section}
+\todo{Neil led writing of this section}
 

### 2015-05-26 WSSSPE2_report/WSSSPE2_review.tex

@@
 
-\katznote{need something about subgroup A here}
-
+Subgroup A consisted of ten participants who discussed papers surrounding training
+and successful community software initiatives.
 Subgroup B discussed four papers and had six participants. Because of the nature
@@
 \subsubsection{Subgroup A Discussion}
+
+\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}
+
+Group discussion started with a position statement by each person surrounding what they had
+learned form their software development experiences and how those lessons might be translated
+into actionable outcomes.  Participants came from a range of backgrounds, and represented
+multiple software and training initiatives, including Software Carpentry, Data 
+Carpentry, Open Science for Synthesis, the Community Surface Dynamics Modeling Systems group, 
+ROpenSci, DataONE, the HDF Group, and others. Some of these software experiences were focused on development
+of new products for use in the sciences (e.g., ROpenSci, CSDMS), and these recounted the 
+difficulties of engaging with disciplinary scientists in writing software.  Software was clearly
+utilized broadly across the various science disciplines represented, and it was developed within those
+disciplines as well.  Some researchers created software for statistical analysis and modeling, while
+others used it to control instrumentation, collate data across networks, collect information from
+respondents, and many other uses.  There was broad consensus that, within the disciplines represented,
+formal training in any type of software development or engineering was rare among the practitioners;
+most are self-taught, and develop software to get another job done. Any ancillary utility of the software
+outside of the specific science target was generally unplanned and few researchers would want to invest more
+time to make their own software more re-usable.
+
+Despite agreement that this body of disciplinary software needed improvement, there was general agreement that
+such improvement needed to be understood in terms of the scientific productivity that could be acheived. A
+software maturity model is needed for science software, but it needs to be introduced in a way that fits
+in with the culture of science, which largely thinks of software as a tool, rather than a product itself. 
+The group was in general consensus that more widespread training in software practice is needed within the
+domain sciences, and several of the participants were involved in efforts along these lines.  Participants felt
+that projects that build a sense of community via training in software and technical practices would have the
+most success in changing practices in that community, but that there was a need for a managed introduction of
+these practices. Participants also recognized that these could not be one-time, one-off training opportunities,
+as software and technologies change rapidly over time.  For example, while today Software Carpentry is focused
+on teaching version control via Git, there has been a rapid evolution from RCS to CVS to SVN to Git over a short
+time frame, and thus communities should expect to need to train for adaptibility and a changing technology tool
+chain, rather than assume these technologoes stay fixed.  Thus, although short term training that introduces
+specific tools was considered highly valuable, these trainins were also not considered sufficient to engender the changes
+in software practice that were deemed neccessary.  Combining the need for changes in practice to be introduced
+incrementally with the need to to minimize the divergence of training from direct science goals and the 
+difficulties of training for a rapidly evolving technology space, participants concluded that multiple training
+efforts that targeted different parts of the spectrum were needed.  Short-term courses introducing immediately
+useful skills needs to be offered alongside more in-depth courses on software engineering and practice that allow
+students to adapt to a changing landscape.
+
+Finally, after agreeing that these complementary training models were needed, the group discussed sustainability of
+training, and how the leading groups in this space are teaching only a small fraction of the community that
+needs and wants training.  Most graduate programs in the sciences do not currently incorporate these appraoches
+in their graduate curricula, although there are an increasing number of quantitatively oriented courses around
+analysis and modeling.  These still, however, generally omit ngineering practices such as version control, unit
+and regression testing, and software modularity and abstraction, often because the instructors themselves in the 
+domain sciences are not familiar with these techniques.  Thus, students who are being trained in these appraches
+are doing so through short 2-3 day training workshops such as Software and Data Carpentry, rather than through 
+semester-long graduate education courses at their University, which tend to focus on statistical and modeling 
+techniques.  Hybrid programs like the three week Open Science for Synthesis (OSS) training that combine the three 
+(science, quantitiative techniques, and software engineering) into an holistic course serve part of the need but 
+reach only a few researchers at this time.  Thus, the group concluded that significant strides needed to be made
+in coordinating these trainings so that they are complementary, on increasing their size and scope, and on
+better integrating them directly into graduate education programs.
 
@@
 \subsubsection{Subgroup A Actions}
+
+Three main actions were identified by Subgroup A that would be of interest to participants and that
+would benefit the research community.  These focused on the desire to amplify the 
+current community efforts in training and software engineering by conecting the current training initiatives.
+( e.g., software/data carpentry, rOpenSci, OSS) of different durations at appropriate stages. Generally it 
+was felt that a modicum of interaction between short term workshops (SwC, DC), medium term trainings (e.g., OSS),
+and longer-term courses (e.g., BIDS Data Science) would benefit from coordinating curriculum,
+discussing and aligning prerequisites, and coordinating timing of courses.  A training
+coordination effort would go a long ways towards amplifying the value of the individual
+efforts and make them all more effective.
+
+\begin{itemize}
+    \item Action 1: Create a roadmap of research software training initiatives. Such a roadmap would
+        provide a taxonomy of training opportunities: what they deliver, and what is needed to know 
+        going in (prerequisites). It would also show which recommended roadmap actions would have a clear 
+        and immediate payoff, and which will have long-term payoffs.  The training roadmap would emphasize 
+        time savings and efficiency gains to be had from each training.
+
+    \item Action 2: Build a report card characterizing use of best practices in sceintific software. Generally,
+        people felt that researchers would be very willing to migrate practices if they could identify
+        where they needed to improve. Such a report card would ask simple questions to charactize use of best
+        practices in science software, and could be structured similarly to Joel Spolsky's 'Software Maturity'
+        questionairre (http://www.joelonsoftware.com/articles/fog0000000043.html).  The survey would create a 
+        report card that shows areas where a project could improve, and then link those areas to specific 
+        training offerings from the training taxonomy from Action 1.
+
+    \item Action 3: Create science software review forum.  It was generally acknowledged that a little code review
+        can have a tremendous impact on the quality of software in a project, but that sites for science code
+        review are lacking.  While people can ask questions on the Stack Exchange sites, they are generally not open
+        to questions of style, approach, or appropriateness, as they try to avoid subjective commentary.   Instead, 
+        we need a site where code gets discussed/summarized/described (in small bites) by the science software community.
+        The target audience would be graduate and undergraduate students in the sciences, and there would need to be
+        mechanisms to keep the review positive and constructive, and not get pedantic or judgemental.  
+        This could be tied to a software registry (such as the nascent GeoSoft project), or to language repositories like 
+        R's CRAN repository, and could lead to the report card discussed in Action 2.
+\end{itemize}
 

### 2015-05-26 WSSSPE2_report/WSSSPE2_review.tex
spelling and other small changes in 7.1.1, still to do 7.1.3

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Nancy led writing of this section}
+\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}
 
@@
 Group discussion started with a position statement by each person surrounding what they had
-learned form their software development experiences and how those lessons might be translated
+learned from their software development experiences and how those lessons might be translated
 into actionable outcomes.  Participants came from a range of backgrounds, and represented
@@
 Despite agreement that this body of disciplinary software needed improvement, there was general agreement that
-such improvement needed to be understood in terms of the scientific productivity that could be acheived. A
+such improvement needed to be understood in terms of the scientific productivity that could be achieved. A
 software maturity model is needed for science software, but it needs to be introduced in a way that fits
-in with the culture of science, which largely thinks of software as a tool, rather than a product itself. 
+the culture of science, which largely thinks of software as a tool, rather than a product itself. 
 The group was in general consensus that more widespread training in software practice is needed within the
@@
 on teaching version control via Git, there has been a rapid evolution from RCS to CVS to SVN to Git over a short
-time frame, and thus communities should expect to need to train for adaptibility and a changing technology tool
-chain, rather than assume these technologoes stay fixed.  Thus, although short term training that introduces
-specific tools was considered highly valuable, these trainins were also not considered sufficient to engender the changes
-in software practice that were deemed neccessary.  Combining the need for changes in practice to be introduced
+time frame, and thus communities should expect to need to train for adaptability and a changing technology tool
+chain, rather than assume these technologies stay fixed.  Thus, although short term training that introduces
+specific tools was considered highly valuable, these trainings were also not considered sufficient to engender the changes
+in software practice that were deemed necessary.  Combining the need for changes in practice to be introduced
 incrementally with the need to to minimize the divergence of training from direct science goals and the 
@@
 training, and how the leading groups in this space are teaching only a small fraction of the community that
-needs and wants training.  Most graduate programs in the sciences do not currently incorporate these appraoches
+needs and wants training.  Most graduate programs in the sciences do not currently incorporate these approaches
 in their graduate curricula, although there are an increasing number of quantitatively oriented courses around
-analysis and modeling.  These still, however, generally omit ngineering practices such as version control, unit
+analysis and modeling.  These still, however, generally omit engineering practices such as version control, unit
 and regression testing, and software modularity and abstraction, often because the instructors themselves in the 
-domain sciences are not familiar with these techniques.  Thus, students who are being trained in these appraches
+domain sciences are not familiar with these techniques.  Thus, students who are being trained in these approaches
 are doing so through short 2-3 day training workshops such as Software and Data Carpentry, rather than through 
@@
 techniques.  Hybrid programs like the three week Open Science for Synthesis (OSS) training that combine the three 
-(science, quantitiative techniques, and software engineering) into an holistic course serve part of the need but 
+(science, quantitative techniques, and software engineering) into an holistic course serve part of the need but 
 reach only a few researchers at this time.  Thus, the group concluded that significant strides needed to be made

### 2015-05-26 WSSSPE2_report/WSSSPE2_review.tex
mostly changes in 7.1.3 - now done with 7

@@
 and successful community software initiatives.
-Subgroup B discussed four papers and had six participants. Because of the nature
+Subgroup B had six participants who discussed four papers. Because of the nature
 of the papers, training emerged as a common theme. However the conversation was
 wide-ranging, including incentives, reproducibility, and funding to promote
-sustainability.
+sustainability.  In the end, both groups discussed training, though from somewhat different
+points of view and resulting in somewhat different suggested actions.
 
@@
 would benefit the research community.  These focused on the desire to amplify the 
-current community efforts in training and software engineering by conecting the current training initiatives.
-( e.g., software/data carpentry, rOpenSci, OSS) of different durations at appropriate stages. Generally it 
+current community efforts in training and software engineering by connecting the current training initiatives
+(e.g., software/data carpentry, rOpenSci, OSS) that are of different durations at appropriate stages. Generally it 
 was felt that a modicum of interaction between short term workshops (SwC, DC), medium term trainings (e.g., OSS),
@@
     \item Action 1: Create a roadmap of research software training initiatives. Such a roadmap would
-        provide a taxonomy of training opportunities: what they deliver, and what is needed to know 
-        going in (prerequisites). It would also show which recommended roadmap actions would have a clear 
+        provide a taxonomy of training opportunities: what they deliver, and what attendees need to know 
+        going in (prerequisites). It would also show which recommended roadmap actions will have a clear 
         and immediate payoff, and which will have long-term payoffs.  The training roadmap would emphasize 
@@
 
-    \item Action 2: Build a report card characterizing use of best practices in sceintific software. Generally,
+    \item Action 2: Build a report card characterizing use of best practices in scientific software. Generally,
         people felt that researchers would be very willing to migrate practices if they could identify
-        where they needed to improve. Such a report card would ask simple questions to charactize use of best
-        practices in science software, and could be structured similarly to Joel Spolsky's 'Software Maturity'
-        questionairre (http://www.joelonsoftware.com/articles/fog0000000043.html).  The survey would create a 
+        where they needed to improve. Such a report card would ask simple questions to characterize use of best
+        practices in science software, and could be structured similarly to Joel Spolsky's Software Maturity
+        questionnaire (also known as \href{http://www.joelonsoftware.com/articles/fog0000000043.html}{the Joel Test}~\cite{joel-test}).  The survey would create a 
         report card that shows areas where a project could improve, and then link those areas to specific 
@@
 
-    \item Action 3: Create science software review forum.  It was generally acknowledged that a little code review
+    \item Action 3: Create a science software review forum.  It was generally acknowledged that a little code review
         can have a tremendous impact on the quality of software in a project, but that sites for science code
@@
         The target audience would be graduate and undergraduate students in the sciences, and there would need to be
-        mechanisms to keep the review positive and constructive, and not get pedantic or judgemental.  
+        mechanisms to keep the review positive and constructive, and not get pedantic or judgmental.  
         This could be tied to a software registry (such as the nascent GeoSoft project), or to language repositories like 

### 2015-05-27 WSSSPE2_report/WSSSPE2_review.tex
adding Matt Jones as author

@@
 \author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$, 
-\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Miguel de Val-Borro$^{(8)}$, Richard Littauer$^{(9)}$, more TBD by contributions}
+\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
+\\Miguel de Val-Borro$^{(9)}$, Richard Littauer$^{(10)}$, more TBD by contributions}
 %\author{Gabrielle Allen$^{(11)}$,  \\
@@
 %
-\thanks{{}$^{(8)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
-%
-\thanks{{}$^{(9)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
+\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
+%
+\thanks{{}$^{(9)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
+%
+\thanks{{}$^{(10)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
 %  

### 2015-05-27 WSSSPE2_report/WSSSPE2_review.tex
starting to work on Section 10 - currently in an ugly intermediate state

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Karen to lead writing of this section}
+\todo{Dan leading writing of this section}
 
@@
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
+
+This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group is to think about concrete things that people in the group are willing to commit to, to improve sustainability of software.
+
+Greg Wilson suggested that these concrete things should be small things so that the group members were likely to do them.
+
+An example was the Architecture of Open Source Applications book~\cite{aosa}, containing contributed chapters on open source software; the group wondering if it could do the same thing for testing?
+
+The group asked if it could produce a 100-line example of how to test neutron transport (or other specific scientific examples) targeted at a sophomore?
+This would be a demonstration of ``how do I do this right''
+
+Another question was ``How do I test so that I know that the code is not the source of my problems?''
+A challenge is stripping things down to that level / that condensed
+Test for infrastructure easier than models
+cf. Lorena Barba’s “12 Steps to Navier Stokes” http://lorenabarba.com/blog/cfd-python-12-steps-to-navier-stokes/
+And explanation of why we don’t teach testing in Software Carpentry any more http://software-carpentry.org/blog/2014/10/why-we-dont-teach-testing.html 
+need to teach people how to test short bits of code, can’t wait until we have 1000s of lines of code
+Distinction between 100 lines of input files vs. 100 lines of source code
+Paper on multiphysics software verificationt01
+"If we can't teach people how to test code until they're developing 1000s of lines of code, we can't be surprised when the don't write tests" - Greg Wilson
+Don’t accept patches that don’t have tests
+Setting up infrastructure to test well is difficult
+Want a standard way that runs out of the box
+Q: what about Jenkins?
+A: sure, but you have to set it up, and that’s not easy for most scientists
+People who’ve worked with a testing framework once (even if they have to be forced to do so) won’t work without it afterward
+Q: but how to get them to do it the first time?
+Two important pieces of the barrier:
+Picking/implementing testing frameworks (technical barrier)
+What are the actual tests that I write?
+Ask students “how is what you’re doing different from typing random keys on the keyboard?” Turn their answer into the concept of tests.
+How many scientists are aware of the fact that “testing the code” and “testing the science” are separate?
+“Every scientist should go through a programming class”
+That’s probably too big a goal for this group for the next year
+Software Carpentry exists because university curricula are full; can’t add extra classes
+Validation questions: how do you know that your method produces a result that is ‘close enough’
+Tom Clune wants a piece of code that badly engineered/lacks unit tests so that he can add the unit tests and show people what it should look like
+The collaboration must be with someone that either wrote the code or is intimately familiar with the topic.
+The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert
+Code does not have to be “badly engineered”, but is either representative or is believed to be impenetrable from a unit-testing perspective.
+
+The group discussed how they learned testing.
+One attendee learned from a TDD tutorial at a Best Practices conference.
+The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving it could be done with Fortran.
+Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.
+The difference in his work is that he now knows about frameworks that let him do that systematically.
+He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
+most of those tests are ephemeral.  They live long enough to give the author of the test confidence in code, then the author moves on; the tests are not handed on and made available to next user.
+
+Attendees joked that they started testing when their skepticism got bigger than their arrogance, or when their guilt got bigger than my arrogance.
+
+Useful reading:
+Daniel Hook, Diane Kelly, "Mutation Sensitivity Testing", Computing in Science \& Engineering, vol.11, no. 6, pp. 40-47, November/December 2009, doi:10.1109/MCSE.2009.200
+Daniel Hook, Diane Kelly, "Testing for trustworthiness in scientific software", ICSE Workshop on Software Engineering for Computational Science and Engineering (SECSE 2009), 2009 pp.
+59-64, doi:10.1109/SECSE.2009.5069163
+
+See the “13 dwarves” from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html
+
+For each problem do:
+Write your name and email address so we can get back in touch with you
+Write a paragraph or two explaining the scientific problem the code addresses
+Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
+Give us a point-form list of the test cases you would use
+
+Tom Clune Problem 1<Thomas.L.Clune@nasa.gov>
+Runge-Kutta 4th order
+O(100 lines of code)
+point-form list of interesting tests
+%Test substep:      k’ = F(x + k h, t + h)
+%F (x) = C; h = 1; k = 1;   x = 0; expect k’ = C
+%F (x) = x^2; h = 2; k = ?;  x = 3; expect k’ = 16
+%Test that correct sequence of substeps are used:  use unrealistic F
+%F(x) =  {1,4,2,5} on nth call;  h = 1 (unused); x0 = 2;  expect x1 = ?(1*1 + 2*4 + 2*2 + 1*5) = 3
+<test with 2D/3D>
+
+Tom Clune Problem 2<Thomas.L.Clune@nasa.gov>
+Interpolation with unevenly spaced abscissae
+O(200 lines of code)
+point-form list of interesting tests
+Find bracket
+%Trivial case:  nodes @ {1,3}; x = 2;  interval = 1
+%nodes@{1,3,5};  x = 4; interval = 2
+%How to handle edges? nodes@{1,3,5}; x = 3
+%If/how to handle nodes that are not increasing?
+%If/how to handle duplicate node?
+%Find weights
+%End pt. - Interval = {1,2} ; x= 1;  w1 = 1
+%Other end pt. - Interval = {1,2}; x = 2; w1 = 0
+%middle - Interval = {1,2} x = 1.5; w1 = 0.5
+%other sizeInterval = {1, 5}; x = 4; w1 = 0.25
+%Combine weights:   
+%End pt - w1 = 1;   y1 = 1; y2 = 2;  y = 1
+%End pt - w1 = 0;  y1 = 1; y2 = 2; y =2
+%Middle - w1 = 0.5;   y1 = 1, y2 = 3;  y = 2
+Put it together
+At this point really checking that steps are called in order
+%interval = bracket(x)
+%find weights = (x, interval)
+%combine  y = sum(w_i, y_i)
+A few simple cases expressed as a table with expected values.
+
+Ethan White Problem 1<ethan@weecology.org>
+Logistic growth models are used to model the dynamics of populations of individuals both in isolation and when interacting with other species.
+At each time step the population of the focal species increases or decreases depending on its reproductive rate, competition from other species, and competition with other individuals of its own species.
+This growth can be either deterministic or stochastic.
+Stochastic growth involves some noise in the reproductive rate.
+25-100 lines
+point-form list of interesting tests
+%stochasticity == 0
+%r = 0, S=1, x_t = x_0
+%r >> 0, S=1, x_1000 = K
+%S=2, r1 > r2, K1=K2, x1_1000 = K1, x2_1000 = 0
+%S=2 r1=r2, k1=k2, x1_100 = x2_100 = analytical equilibrium 
+%stochasticity > 0
+%K = 0, x_t = 0
+%r1 >> r2, x1_1000 = K1, x2_1000 = K2
+pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes
+e
+
+Ethan White Problem 1<ethan@weecology.org>
+Neutral model (based on https://github.com/rueuntal/species-associations/)
+100 lines
+point-form list of interesting tests
+%disp_function == no dispersal
+%m == 0, self.immigrants == 0 at any t
+pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes
+
+Karen Cranston <karen.cranston@nescent.org>
+existing SWC example: http://software-carpentry.org/v4/setdict/phylogen.html 
+Given DNA sequences for a set of species, we can infer a tree that describes the evolutionary history of the species.
+We assume that the best tree is the one that involves the fewest number of changes in the DNA (the “maximum parsimony” tree).
+This code will calculate the parsimony score for a given tree and given DNA alignment.
+http://telliott99.blogspot.com/2010/03/fitch-and-sankoff-algorithms-for.html
+hmmm…
+list of tests
+DNA sequences are identical; score = 0
+TLC:  Use Fitch’s algorithm for parsimony score.  Assuming your languages does not have built in type for Set.   (E.g. python does, but Fortran does not.) 
+%Test Set pseudo-Intersection  “^” (empty intersection is union)
+%Single nucleotide tests
+%{a} ^ {a} = {a}
+%{b} ^ {b} = {b}
+%{a} ^ {b} = {a,b
+%{a,b} ^ {b,c} = {b}
+%Multiple nucleotide tests
+%({a},{b})^({b},{b}) = ({a,b},{b})
+%Node tests:   Node v has tuple of sets (X_i(v)) and integer score S(v)
+%N1 = <({a});s=2>  N2 = <({a});s=3>    ->   Nparent = <({a}; s=2+3>
+%N1 = <({a},{a,b});s=2>  N2 = <({b},{b}});s=3>    ->   
+%Nparent = <({a,b},{b}); s=2+3+1>
+Probably should create a table of N1, N2, Nparent and write a generic test that checks each row of the table.
+Tree tests:  Assuming already have post-order binary tree traversal
+Tree with 2 leaves and one parent node.   Use table from 3.4.3 to check parent status after traversal with scoring rule
+Create more complex tree with 3-4 leaves.  Manually determine status of each top node.
+Define parsimony score as s-value of top node
+
+Greg Wilson <gvwilson@software-carpentry.org
+Invasion percolation is used to study the spread of pollutants in fractured rock.
+Model the rock as a 2D grid filled with random integers between 0 and K; mark the central cell as “filled”; and then repeatedly select the cell on the boundary with the lowest value (i.e., the most permeable) and fill it.
+The density of the resulting fractal tells you how quickly the pollutant spreads.
+Simple implementation is ~100 lines of Python.
+Interesting tests are:
+3x3 grid with a single ‘0’ on the boundary (all other cells are ‘1’) - should fill to boundary in a single step, and the cell with ‘0’ should be the one that fills.
+5x5 grid with a straight line of 0’s between the center and the boundary (same case as above, but expanded by one layer)
+5x5 grid: cells immediately surrounding the center are all ‘0’, all of boundary is ‘1’ except for one cell, which is 0 - all of the interior should fill, then the single cell on the boundary that’s a ‘1’
+
+For example, look at the testing of the invasion percolation code in this sequence of slides:
+The problem: http://software-carpentry.org/v4/invperc/intro.html 
+The testing: http://software-carpentry.org/v4/invperc/test.html
 

### 2015-05-27 WSSSPE2_report/WSSSPE2_review.tex
still working on 10

@@
 
-This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group is to think about concrete things that people in the group are willing to commit to, to improve sustainability of software.
-
-Greg Wilson suggested that these concrete things should be small things so that the group members were likely to do them.
+This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group was really to think about concrete things that people in the group would be willing to commit to, in order to improve the sustainability of software.
+It was suggested that these concrete things should be small things so that the group members were likely to do them.
 
@@
 The group asked if it could produce a 100-line example of how to test neutron transport (or other specific scientific examples) targeted at a sophomore?
-This would be a demonstration of ``how do I do this right''
+This would be a demonstration of ``How do I do this right?''
 
 Another question was ``How do I test so that I know that the code is not the source of my problems?''
-A challenge is stripping things down to that level / that condensed
-Test for infrastructure easier than models
-cf. Lorena Barba’s “12 Steps to Navier Stokes” http://lorenabarba.com/blog/cfd-python-12-steps-to-navier-stokes/
-And explanation of why we don’t teach testing in Software Carpentry any more http://software-carpentry.org/blog/2014/10/why-we-dont-teach-testing.html 
-need to teach people how to test short bits of code, can’t wait until we have 1000s of lines of code
-Distinction between 100 lines of input files vs. 100 lines of source code
-Paper on multiphysics software verificationt01
-"If we can't teach people how to test code until they're developing 1000s of lines of code, we can't be surprised when the don't write tests" - Greg Wilson
-Don’t accept patches that don’t have tests
-Setting up infrastructure to test well is difficult
-Want a standard way that runs out of the box
-Q: what about Jenkins?
-A: sure, but you have to set it up, and that’s not easy for most scientists
-People who’ve worked with a testing framework once (even if they have to be forced to do so) won’t work without it afterward
-Q: but how to get them to do it the first time?
-Two important pieces of the barrier:
-Picking/implementing testing frameworks (technical barrier)
-What are the actual tests that I write?
-Ask students “how is what you’re doing different from typing random keys on the keyboard?” Turn their answer into the concept of tests.
-How many scientists are aware of the fact that “testing the code” and “testing the science” are separate?
-“Every scientist should go through a programming class”
-That’s probably too big a goal for this group for the next year
-Software Carpentry exists because university curricula are full; can’t add extra classes
-Validation questions: how do you know that your method produces a result that is ‘close enough’
-Tom Clune wants a piece of code that badly engineered/lacks unit tests so that he can add the unit tests and show people what it should look like
+A challenge is stripping things down to an appropriate level.
+
+Testing for infrastructure is easier than testing for models. \katznote{to delete? - don't see how it fits}
+
+The group referred to Lorena Barba's ``12 Steps to Navier Stokes~\cite{12steps},'' an online CFD lesson repository created as a set of iPython notebooks.
+
+Grid Wilson pointed to an explanation of why Software Carpentry doesn't teach testing any more~\cite{SCtesting}, which boils down to two issues.
+First, that scientific computing doesn't (yet) have the cultural norms for error bars that experimental sciences have, and second, that there is a breathtaking diversity of scientific code; scientific research is highly specialized, which means that the tests scientists write are much less transferable or reusable than those found in other fields.
+
+The group recognized that we need to teach people how to test short bits of code, we can't wait until we have thousands of lines of code.
+Greg commented, ``if we can't teach people how to test code until they're developing thousands of lines of code, we can't be surprised when they don't write tests.''
+
+Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
+
+in progress....
+
+%Paper on multiphysics software verificationt01 - \url{http://onlinelibrary.wiley.com/doi/10.1002/spe.2220/abstract}
+%
+%Don't accept patches that don't have tests
+%
+%Setting up infrastructure to test well is difficult
+%
+%Want a standard way that runs out of the box
+%Q: what about Jenkins?
+%A: sure, but you have to set it up, and that's not easy for most scientists
+%
+%People who've worked with a testing framework once (even if they have to be forced to do so) won't work without it afterward
+%Q: but how to get them to do it the first time?
+%
+%Two important pieces of the barrier:
+%a. Picking/implementing testing frameworks (technical barrier)
+%b. What are the actual tests that I write?
+%Ask students ``how is what you're doing different from typing random keys on the keyboard?'' Turn their answer into the concept of tests.
+%
+%How many scientists are aware of the fact that ``testing the code'' and ``testing the science'' are separate?
+%
+%``Every scientist should go through a programming class''
+%a. That's probably too big a goal for this group for the next year
+%b. Software Carpentry exists because university curricula are full; can't add extra classes
+%
+%
+%Validation questions: how do you know that your method produces a result that is ‘close enough’
+%
+%Tom Clune wants a piece of code that badly engineered/lacks unit tests so that he can add the unit tests and show people what it should look like
+%a.
 The collaboration must be with someone that either wrote the code or is intimately familiar with the topic.
 The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert
-Code does not have to be “badly engineered”, but is either representative or is believed to be impenetrable from a unit-testing perspective.
+%b. Code does not have to be ``badly engineered'', but is either representative or is believed to be impenetrable from a unit-testing perspective.
 

### 2015-05-27 WSSSPE2_report/WSSSPE2_review.tex
more work in 10

@@
 major scientific software efforts such as 
-Bioconductor{\footnote{Bioconductor, \url{http://www.bioconductor.org}}, 
+Bioconductor\footnote{Bioconductor, \url{http://www.bioconductor.org}}, 
 Galaxy\footnote{Galaxy, \url{http://galaxyproject.org}}, and
@@
 
-\todo{short intro to group here}
+This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group was really to think about concrete things that people in the group would be willing to commit to, in order to improve the sustainability of software.
 
@@
 
-This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group was really to think about concrete things that people in the group would be willing to commit to, in order to improve the sustainability of software.
 It was suggested that these concrete things should be small things so that the group members were likely to do them.
@@
 
-in progress....
-
-%Paper on multiphysics software verificationt01 - \url{http://onlinelibrary.wiley.com/doi/10.1002/spe.2220/abstract}
-%
-%Don't accept patches that don't have tests
-%
-%Setting up infrastructure to test well is difficult
-%
-%Want a standard way that runs out of the box
-%Q: what about Jenkins?
-%A: sure, but you have to set it up, and that's not easy for most scientists
-%
-%People who've worked with a testing framework once (even if they have to be forced to do so) won't work without it afterward
-%Q: but how to get them to do it the first time?
-%
-%Two important pieces of the barrier:
-%a. Picking/implementing testing frameworks (technical barrier)
-%b. What are the actual tests that I write?
-%Ask students ``how is what you're doing different from typing random keys on the keyboard?'' Turn their answer into the concept of tests.
-%
-%How many scientists are aware of the fact that ``testing the code'' and ``testing the science'' are separate?
-%
-%``Every scientist should go through a programming class''
-%a. That's probably too big a goal for this group for the next year
-%b. Software Carpentry exists because university curricula are full; can't add extra classes
-%
-%
-%Validation questions: how do you know that your method produces a result that is ‘close enough’
-%
-%Tom Clune wants a piece of code that badly engineered/lacks unit tests so that he can add the unit tests and show people what it should look like
-%a.
-The collaboration must be with someone that either wrote the code or is intimately familiar with the topic.
-The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert
-%b. Code does not have to be ``badly engineered'', but is either representative or is believed to be impenetrable from a unit-testing perspective.
+Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.
+
+One general rule that was suggested is that projects shouldn't accept patches that don't have associated tests.
+This led to discussion about the fact that setting up infrastructure to test well is difficult.
+The group would like a standard way that runs out of the box.
+Jenkins was suggested as
+a solution, but setting it up is considered difficult for most scientists.
+
+Group members also comments that people who've worked with a testing framework once (even if they have to be forced to do so) won't work without it afterward.
+But how to get them to do this the first time is not clear.
+
+Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?
+A possible method to solve the latter is to ask students ``how is what you're doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.
+
+Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are separate.
+A related question that came up is ``How do you know that your method produces a result that is `close enough'?''
+
+Some group members thought that every scientist should go through a programming class, but that was considered too big a goal for the group for the next year.
+Also, it's not clear where these classes could be offered.
+Software Carpentry exists because university curricula are full; they can't add extra classes.
+
+Tom Clune suggested it would be useful to have a piece of code that is badly engineered and lacks unit tests so that he can add the unit tests and show people what it should look like.
+However, such a collaboration must be with someone that either wrote the code or is intimately familiar with the topic.
+The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert.
+Others suggested that the code does not have to be ``badly engineered'', but is either representative or is believed to be impenetrable from a unit-testing perspective.
 
@@
 
-Useful reading:
-Daniel Hook, Diane Kelly, "Mutation Sensitivity Testing", Computing in Science \& Engineering, vol.11, no. 6, pp. 40-47, November/December 2009, doi:10.1109/MCSE.2009.200
-Daniel Hook, Diane Kelly, "Testing for trustworthiness in scientific software", ICSE Workshop on Software Engineering for Computational Science and Engineering (SECSE 2009), 2009 pp.
-59-64, doi:10.1109/SECSE.2009.5069163
-
-See the “13 dwarves” from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html
+The group suggested two papers by Hook and Kelly that would be useful for those interested in software testing to read~\cite{hook1,hook2}
+
+See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
+
+\subsection{Testing examples}
+
+The group took some time to define
+
+\katznote{... working from here still...}
 

### 2015-05-29 WSSSPE2_report/WSSSPE2_review.tex
Minor changes - mostly checking to ensure I can push to github.

@@
 \newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
+\newcommand{\clunenote}[1]{ {\textcolor{orange}    { ***Tom:      #1 }}}
 \newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
@@
 Testing for infrastructure is easier than testing for models. \katznote{to delete? - don't see how it fits}
+\clunenote{Do not know if I can answer whether it fits, but can expand upon the intent of the sentence.
+The idea was to ground people's expectations.
+All software can be tested.
+But some types of software are much easier to test.
+In particular numerical argorithms introduce new issues that increase the difficulty and can represent a high barrier for starting to unit test.
+The issues that arise are (1) dealing with numerical error (and in particular roundoff error that accumulates through multiple operations and/or loop iterations, and (2)  the \emph{apparent} lack of oracles for nontrivial cases.)
 
@@
 Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
+\clunenote{I no longer recall the context for this statement.}
 
@@
 See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
+\clunenote{No opinion - someone thought it was a useful reference for the topic.}
 

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex

--- 2015-05-27 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-It was suggested that these concrete things should be small things so that the group members were likely to do them.
-
-An example was the Architecture of Open Source Applications book~\cite{aosa}, containing contributed chapters on open source software; the group wondering if it could do the same thing for testing?
-
-The group asked if it could produce a 100-line example of how to test neutron transport (or other specific scientific examples) targeted at a sophomore?
-This would be a demonstration of ``How do I do this right?''
-
-Another question was ``How do I test so that I know that the code is not the source of my problems?''
-A challenge is stripping things down to an appropriate level.
-
-Testing for infrastructure is easier than testing for models. \katznote{to delete? - don't see how it fits}
-
-The group referred to Lorena Barba's ``12 Steps to Navier Stokes~\cite{12steps},'' an online CFD lesson repository created as a set of iPython notebooks.
+The consensus of the group was that small, concrete actions would have
+a greater chance of proper follow-through by mpmebers of the group.
+An example of this approach was the Architecture of Open Source
+Applications book~\cite{aosa}, containing contributed chapters on open
+source software.  
+The group contemplated creating an analogous book/web-resource for
+testing of scientific software.
+
+Several members of the discussion indicated that while they valued the
+concpet software testing in theory, they were unaware how to test
+certain kinds of software.  
+Or, more importantly, they lacked relevant examples of software
+testing that would be suitable for new members to their team.
+The group asked if it could produce a 100-line example of how to test
+neutron transport (or other specific scientific examples) targeted at
+a sophomore? This would be a demonstration of ``How do I do this
+right?''
+
+Another question was ``How do I test so that I know that the code is
+not the source of my problems?''  A common challenge is that of
+stripping things down to an appropriate level for tests.  The group
+recognized that the difficulty is different for so-called software
+infrastructure than for numerical/scientific layers of software
+applications.
+
+The group referred to Lorena Barba's ``12 Steps to Navier
+Stokes~\cite{12steps},'' an online CFD lesson repository created as a
+set of iPython notebooks.  \clunenote{I'm not sure of the context
+  here, except possibly as an analog for something under discussion.}
 
@@
 Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
+\clunenote{I no longer recall the context for this statement; does not seem important to me.}
 
@@
 
-One general rule that was suggested is that projects shouldn't accept patches that don't have associated tests.
-This led to discussion about the fact that setting up infrastructure to test well is difficult.
+One general rule that was suggested is that projects should not accept
+software patches that don't have associated tests.  
+This led to discussion about the fact that setting up appropriate 
+infrastructure for testing can be difficult.
 The group would like a standard way that runs out of the box.
-Jenkins was suggested as
-a solution, but setting it up is considered difficult for most scientists.
-
-Group members also comments that people who've worked with a testing framework once (even if they have to be forced to do so) won't work without it afterward.
-But how to get them to do this the first time is not clear.
+Jenkins was suggested as a common open-source solution, but initial
+configuration is considered to be challenging for typical scientists.
+
+Several members of the group commented that once a developer has been
+exposed to a testing framework in a nontrivial fashion, he/she will
+subsequently insist upon using a framework for further work.
+The acknowledged challenge was how to create that crucial first exposure
+to such frameworks.
 
 Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?
-A possible method to solve the latter is to ask students ``how is what you're doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.
-
-Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are separate.
+A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.
+I.e.
+expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''.
+
+Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are distinct issues.
 A related question that came up is ``How do you know that your method produces a result that is `close enough'?''
@@
 
-Tom Clune suggested it would be useful to have a piece of code that is badly engineered and lacks unit tests so that he can add the unit tests and show people what it should look like.
-However, such a collaboration must be with someone that either wrote the code or is intimately familiar with the topic.
-The exercise will be to extract unit-tests through reverse-engineering {\b and} questioning the developer/expert.
-Others suggested that the code does not have to be ``badly engineered'', but is either representative or is believed to be impenetrable from a unit-testing perspective.
+Tom Clune suggested that a useful demonstration could be generate by
+beginning with a (smallish) piece of badly engineered code that lacks
+appropriate unit tests.   
+The code would then be refactored into a more presentable form once
+unit test were created/derived.
+However, such a collaboration must be with someone that either wrote
+the code or is intimately familiar with the subject matter.  The exercise would
+be to extract unit-tests through reverse-engineering {\b and}
+questioning the developer/expert.  Others suggested that the code does
+not have to be ``badly engineered'', but is either representative or
+is believed to be impenetrable from a unit-testing perspective.
 
 The group discussed how they learned testing.
-One attendee learned from a TDD tutorial at a Best Practices conference.
-The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving it could be done with Fortran.
+One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference.
+The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.
 Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.
 The difference in his work is that he now knows about frameworks that let him do that systematically.
-He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.  He also commented that
-most of those tests are ephemeral.  They live long enough to give the author of the test confidence in code, then the author moves on; the tests are not handed on and made available to next user.
-
-Attendees joked that they started testing when their skepticism got bigger than their arrogance, or when their guilt got bigger than my arrogance.
+He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.
+He also commented that
+most of those initial checks were ephemeral.
+They live long enough to give the author of the test confidence in code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.
+
+Attendees joked that they started testing when their skepticism/guilt became larger than their arrogance.
 

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
finished with section 10 - thanks Tom!

@@
 \\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
-\\Miguel de Val-Borro$^{(9)}$, Richard Littauer$^{(10)}$, more TBD by contributions}
+\\Thomas L. Clune$^{(9)}$, Miguel de Val-Borro$^{(10)}$, Richard Littauer$^{(11)}$,
+\\more TBD by contributions}
 %\author{Gabrielle Allen$^{(11)}$,  \\
@@
 %
-\thanks{{}$^{(9)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
-%
-\thanks{{}$^{(10)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
+\thanks{{}$^{(9)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}  
+%
+\thanks{{}$^{(10)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}  
+%
+\thanks{{}$^{(11)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}   
 %  
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan leading writing of this section}
+\todo{Dan leading writing of this section, with help from Tom Clune}
 
@@
 The consensus of the group was that small, concrete actions would have
-a greater chance of proper follow-through by mpmebers of the group.
+the greatest chance of proper followthrough by members of the group.
 An example of this approach was the Architecture of Open Source
@@
 source software.  
-The group contemplated creating an analogous book/web-resource for
+The group contemplated creating an analogous book or web resource for
 testing of scientific software.
@@
 Several members of the discussion indicated that while they valued the
-concpet software testing in theory, they were unaware how to test
+concept of software testing in theory, they were unaware how to test
 certain kinds of software.  
@@
 
-The group referred to Lorena Barba's ``12 Steps to Navier
-Stokes~\cite{12steps},'' an online CFD lesson repository created as a
-set of iPython notebooks.  \clunenote{I'm not sure of the context
-  here, except possibly as an analog for something under discussion.}
+%The group referred to Lorena Barba's ``12 Steps to Navier
+%Stokes~\cite{12steps},'' an online CFD lesson repository created as a
+%set of iPython notebooks.  \clunenote{I'm not sure of the context
+%  here, except possibly as an analog for something under discussion.}
 
@@
 Greg commented, ``if we can't teach people how to test code until they're developing thousands of lines of code, we can't be surprised when they don't write tests.''
-
-Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
-\clunenote{I no longer recall the context for this statement; does not seem important to me.}
-
+%
+%Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
+%\clunenote{I no longer recall the context for this statement; does not seem important to me.}
+%
 Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.
@@
 A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.
-I.e.
-expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''.
+That is to say, expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''
 
@@
 most of those initial checks were ephemeral.
-They live long enough to give the author of the test confidence in code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.
+They live long enough to give the author of the test confidence in the code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.
 
@@
 
-See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
-\clunenote{No opinion - someone thought it was a useful reference for the topic.}
-
-\subsection{Testing examples}
-
-The group took some time to define
-
-\katznote{... working from here still...}
-
-For each problem do:
-Write your name and email address so we can get back in touch with you
-Write a paragraph or two explaining the scientific problem the code addresses
-Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
-Give us a point-form list of the test cases you would use
-
-Tom Clune Problem 1<Thomas.L.Clune@nasa.gov>
-Runge-Kutta 4th order
-O(100 lines of code)
-point-form list of interesting tests
-%Test substep:      k’ = F(x + k h, t + h)
-%F (x) = C; h = 1; k = 1;   x = 0; expect k’ = C
-%F (x) = x^2; h = 2; k = ?;  x = 3; expect k’ = 16
-%Test that correct sequence of substeps are used:  use unrealistic F
-%F(x) =  {1,4,2,5} on nth call;  h = 1 (unused); x0 = 2;  expect x1 = ?(1*1 + 2*4 + 2*2 + 1*5) = 3
-<test with 2D/3D>
-
-Tom Clune Problem 2<Thomas.L.Clune@nasa.gov>
-Interpolation with unevenly spaced abscissae
-O(200 lines of code)
-point-form list of interesting tests
-Find bracket
-%Trivial case:  nodes @ {1,3}; x = 2;  interval = 1
-%nodes@{1,3,5};  x = 4; interval = 2
-%How to handle edges? nodes@{1,3,5}; x = 3
-%If/how to handle nodes that are not increasing?
-%If/how to handle duplicate node?
-%Find weights
-%End pt. - Interval = {1,2} ; x= 1;  w1 = 1
-%Other end pt. - Interval = {1,2}; x = 2; w1 = 0
-%middle - Interval = {1,2} x = 1.5; w1 = 0.5
-%other sizeInterval = {1, 5}; x = 4; w1 = 0.25
-%Combine weights:   
-%End pt - w1 = 1;   y1 = 1; y2 = 2;  y = 1
-%End pt - w1 = 0;  y1 = 1; y2 = 2; y =2
-%Middle - w1 = 0.5;   y1 = 1, y2 = 3;  y = 2
-Put it together
-At this point really checking that steps are called in order
-%interval = bracket(x)
-%find weights = (x, interval)
-%combine  y = sum(w_i, y_i)
-A few simple cases expressed as a table with expected values.
-
-Ethan White Problem 1<ethan@weecology.org>
-Logistic growth models are used to model the dynamics of populations of individuals both in isolation and when interacting with other species.
-At each time step the population of the focal species increases or decreases depending on its reproductive rate, competition from other species, and competition with other individuals of its own species.
-This growth can be either deterministic or stochastic.
-Stochastic growth involves some noise in the reproductive rate.
-25-100 lines
-point-form list of interesting tests
-%stochasticity == 0
-%r = 0, S=1, x_t = x_0
-%r >> 0, S=1, x_1000 = K
-%S=2, r1 > r2, K1=K2, x1_1000 = K1, x2_1000 = 0
-%S=2 r1=r2, k1=k2, x1_100 = x2_100 = analytical equilibrium 
-%stochasticity > 0
-%K = 0, x_t = 0
-%r1 >> r2, x1_1000 = K1, x2_1000 = K2
-pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes
-e
-
-Ethan White Problem 1<ethan@weecology.org>
-Neutral model (based on https://github.com/rueuntal/species-associations/)
-100 lines
-point-form list of interesting tests
-%disp_function == no dispersal
-%m == 0, self.immigrants == 0 at any t
-pass a “stochasticity” function that changes things in a known way, either always returning the same value or with a set of values that has variation, but whose values are deterministic and then test against known outcomes
-
-Karen Cranston <karen.cranston@nescent.org>
-existing SWC example: http://software-carpentry.org/v4/setdict/phylogen.html 
-Given DNA sequences for a set of species, we can infer a tree that describes the evolutionary history of the species.
-We assume that the best tree is the one that involves the fewest number of changes in the DNA (the “maximum parsimony” tree).
-This code will calculate the parsimony score for a given tree and given DNA alignment.
-http://telliott99.blogspot.com/2010/03/fitch-and-sankoff-algorithms-for.html
-hmmm…
-list of tests
-DNA sequences are identical; score = 0
-TLC:  Use Fitch’s algorithm for parsimony score.  Assuming your languages does not have built in type for Set.   (E.g. python does, but Fortran does not.) 
-%Test Set pseudo-Intersection  “^” (empty intersection is union)
-%Single nucleotide tests
-%{a} ^ {a} = {a}
-%{b} ^ {b} = {b}
-%{a} ^ {b} = {a,b
-%{a,b} ^ {b,c} = {b}
-%Multiple nucleotide tests
-%({a},{b})^({b},{b}) = ({a,b},{b})
-%Node tests:   Node v has tuple of sets (X_i(v)) and integer score S(v)
-%N1 = <({a});s=2>  N2 = <({a});s=3>    ->   Nparent = <({a}; s=2+3>
-%N1 = <({a},{a,b});s=2>  N2 = <({b},{b}});s=3>    ->   
-%Nparent = <({a,b},{b}); s=2+3+1>
-Probably should create a table of N1, N2, Nparent and write a generic test that checks each row of the table.
-Tree tests:  Assuming already have post-order binary tree traversal
-Tree with 2 leaves and one parent node.   Use table from 3.4.3 to check parent status after traversal with scoring rule
-Create more complex tree with 3-4 leaves.  Manually determine status of each top node.
-Define parsimony score as s-value of top node
-
-Greg Wilson <gvwilson@software-carpentry.org
-Invasion percolation is used to study the spread of pollutants in fractured rock.
-Model the rock as a 2D grid filled with random integers between 0 and K; mark the central cell as “filled”; and then repeatedly select the cell on the boundary with the lowest value (i.e., the most permeable) and fill it.
-The density of the resulting fractal tells you how quickly the pollutant spreads.
-Simple implementation is ~100 lines of Python.
-Interesting tests are:
-3x3 grid with a single ‘0’ on the boundary (all other cells are ‘1’) - should fill to boundary in a single step, and the cell with ‘0’ should be the one that fills.
-5x5 grid with a straight line of 0’s between the center and the boundary (same case as above, but expanded by one layer)
-5x5 grid: cells immediately surrounding the center are all ‘0’, all of boundary is ‘1’ except for one cell, which is 0 - all of the interior should fill, then the single cell on the boundary that’s a ‘1’
-
-For example, look at the testing of the invasion percolation code in this sequence of slides:
-The problem: http://software-carpentry.org/v4/invperc/intro.html 
-The testing: http://software-carpentry.org/v4/invperc/test.html
+The group then decided that a useful action would be to create a set of scientific codes
+with associated tests. For each, they suggested:
+\begin{enumerate}
+\item Write your name and email address so we can get back in touch with you
+\item Write a paragraph or two explaining the scientific problem the code addresses
+\item Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
+\item Give us a point-form list of the test cases you would use
+\end{enumerate}
+
+%See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
+%\clunenote{No opinion - someone thought it was a useful reference for the topic.}
 

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
editing in intro and conclusions

@@
 
-WSSSPE1 was organized into sessions, each of which was aimed at discussing one
-or more of the themes from the call for papers, with a few paper authors invited
-to summarize the other papers in the session as a panel, followed by a general discussion
-about that theme. The mapping of papers to themes was done by the organizers.
-
-For WSSSPE2, the organizers wanted to increase the interactivity of the
-sessions, and to open the process of creating the sessions to the full program
+The organizers wanted very interactive
+sessions, with the process of creating the sessions open to the full program
 committee, the paper authors, and others who might attend the workshop. In order
-to do this, the organizers decided to use a breakout format for two sessions,
-and to use an open process to determine the breakout topics. Specifically, Well
-Sorted\footnote{\url{http://www.well-sorted.org}} was used in the following
-steps: \begin{enumerate} \item Authors were asked to create Well Sorted
+to do this, the organizers used Well
+Sorted\footnote{\url{http://www.well-sorted.org}}, with the following
+steps:
+\begin{enumerate}
+\item Authors were asked to create Well Sorted
 ``cards'' for the papers. These cards have a title (50 characters maximum) and a
-body (255 characters maximum). \item Authors, program committee members, and
+body (255 characters maximum).
+\item Authors, program committee members, and
 members of the WSSSPE mailing list were asked to sort the cards. Each person
@@
 person wants it to have, and it can have whatever meaning that makes sense to
-that person. \item Well Sorted produces a set of averages of all the sorts, with
-various numbers of card clusters. \end{enumerate}
-
-The organizers then chose a sort that contained five groups and that felt most
-meaningful. After that, they decided on themes for the five groups, namely:
+that person.
+\item Well Sorted produced a set of averages of all the sorts, with
+various numbers of card clusters.
+\end{enumerate}
+
+The organizers then chose a sort that contained five groups that felt most
+meaningful. After that, they decided on names for the five groups:
 \begin{itemize}
@@
 they were not placed in groups by the peer-grouping system. The authors of
-these papers were asked which groups seemed the best for their papers---these
+these papers were asked which groups seemed the best for their papers; these
 papers were then placed in those groups. Sections~\ref{sec:exploring}-\ref{sec:code_testing}
@@
 
-\todo{currently just pre-workshop stuff - needs to be updated}
-
-\katznote{add discussion of keynotes, lightning talks, charge to working groups about building plans, etc.}
+The WSSSPE2 workshop continued our experiment from WSSSPE1 in how we can
+collaboratively build a workshop agenda, and we began a new experiment in
+how to build a series of workshops into an ongoing community activity.
+
+The differences in workshop organization in WSSSPE2 from WSSSPE1
+are in using an existing service (EasyChair) to handle submissions and reviews,
+rather than an ad hoc process, and using an existing service (well-sorted) to
+allow collaborative grouping of papers into themes by all authors, reviewers,
+and the community, rather than this being done in an ad hoc manner by the
+organizers alone.
+
+The fact remains that contributors also want to get credit for their
+participation in the process. And the workshop organizers still want to make
+sure that the workshop content and their efforts are recorded. Ideally, there
+would be a service that would be able to index the contributions to the
+workshop, serving the authors, the organizers, and the larger community. But
+since there still isn't such a service today, the workshop organizers are
+writing this initial report and making use of arXiv as a partial solution to
+provide a record of the workshop.
+
+WSSSPE actively used the online social network Twitter, with hashtag
+``\#WSSSPE''. There were substantially more tweets (messages) during the days of
+the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
+18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
+day of the workshop. Some of the main points and highlights in the meeting are
+shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
+the day of workshop, selected by the metrics that number of retweets or
+favorites larger than five and the sum of two measures greater than ten.
+
+In terms of building community activities, we wanted to focus primarily on
+working groups, which we were able to do, as discussed above, but we
+also wanted to make sure that attendees felt they had a chance to get their
+ideas across to the whole group, which was the purpose of the lightning talks.
+Overall, this seemed to be successful at the time, in terms of both the lightning
+talks and the breakout groups, and the discussion of sustainability also led
+to interesting and useful results, but the challenge that we have discovered
+since WSSSPE2 is that it is very hard to continue the breakout groups'
+activities.  The WSSSPE2 participants were willing to dedicate their time to
+the groups while they were at the meeting, but since then, they have gone
+back to their (paid) jobs.  We need to determine how to tie the WSSSPE
+breakout activities to people's jobs, so that they feel that continuing them
+is a higher priority that it is now, perhaps through funding the participants,
+or through funding coordinators for each activity, or perhaps by getting
+the workshop participants to agree to a specific schedule of activities during the
+workshop.
 
@@
 
-The WSSSPE2 workshop continues our experiment from WSSSPE1 in how we can
-collaboratively build a workshop agenda. The differences in WSSSPE2 from WSSSPE1
-are in using an existing service (EasyChair) to handle submissions and reviews,
-rather than an ad hoc process, and using an existing service (well-sorted) to
-allow collaborative grouping of papers into themes by all authors, reviewers,
-and the community, rather than this being done in an ad hoc manner by the
-organizers alone.
-
-The fact remains that contributors also want to get credit for their
-participation in the process. And the workshop organizers still want to make
-sure that the workshop content and their efforts are recorded. Ideally, there
-would be a service that would be able to index the contributions to the
-workshop, serving the authors, the organizers, and the larger community. But
-since there still isn't such a service today, the workshop organizers are
-writing this initial report and making use of arXiv as a partial solution to
-provide a record of the workshop.
-
-WSSSPE actively used the online social network Twitter, with hashtag
-``\#WSSSPE''. There were substantially more tweets (messages) during the days of
-the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
-18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
-day of the workshop. Some of the main points and highlights in the meeting are
-shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
-the day of workshop, selected by the metrics that number of retweets or
-favourites larger than five and the sum of two measures greater than ten.
-
-After the workshop, one or more additional papers will be created that will
-include the discussions at the workshop. These papers will likely have many
-authors, and may be submitted to peer-reviewed journals.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{feel free to add stuff here}
 
@@
 reflect the views of the National Science Foundation.
+
+\todo{feel free to add stuff here}
 

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
Replace non-ascii character
Add url commands

@@
 completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used EasyChair\footnote{http://easychair.org/}
+4.1 reviews per reviewer), they used EasyChair\footnote{\url{http://easychair.org/}}
 to record scores on relevance and comments. Finally, the organizers accessed the
@@
 there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
-there is: \url{tinyurl.com/qdodbo7}.
+there is: \url{http://tinyurl.com/qdodbo7}.
 %http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068
@@
 Pacific Northwestern National Laboratory via a petition
-campaign\footnote{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}
+campaign\footnote{\url{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
 at change.org, ``\emph{Today there are very few science areas left which do not
@@
 common in scientific research. A 2014
-survey\footnote{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}
+survey\footnote{\url{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
 conducted by the SSI polled researchers from 15 research-intensive UK
@@
 Some communities have had significant initiatives to improve software discovery, e.g.
-the NIH Software Discovery Index\footnote{NIH Software Discovery Index: http://softwarediscoveryindex.org/}.
+the NIH Software Discovery Index\footnote{NIH Software Discovery Index: \url{http://softwarediscoveryindex.org/}}.
 Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.
@@
 Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website.
-Using code repositories like GitHub\footnote{GitHub: http://github.com/} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: http://figshare.com/} or Zenodo\footnote{Zenodo: http://zenodo.org/}.
-Docker\footnote{Docker: htp://www.docker.com/} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.
+Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}.
+Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.
 
@@
 This group aimed to come up with a categorisation of journals which published software papers.
-Starting from the list of journals\footnote{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers.
-These were the Journal of Open Research Software\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}, PLoS ONE\footnote{http://www.plosone.org/static/guidelines\#software}, Journal of Statistical Software\footnote{http://www.jstatsoft.org/instructions}.
-Methods in Ecology and Evolution\footnote{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}, Transactions of Mathematical Software\footnote{http://toms.acm.org/Authors.html}, GigaScience\footnote{http://www.gigasciencejournal.com/about}, and PLoS Computational Biology\footnote{http://www.ploscompbiol.org/static/guidelines\#software}.
+Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers.
+These were the Journal of Open Research Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}}, PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}}, Journal of Statistical Software\footnote{\url{http://www.jstatsoft.org/instructions}}.
+Methods in Ecology and Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}}, Transactions of Mathematical Software\footnote{\url{http://toms.acm.org/Authors.html}}, GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS Computational Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
 
@@
 Journals should provide guidelines about what to consider when reviewing a software submission.
-A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{http://openresearchsoftware.metajnl.com/about/editorialPolicies}.
+A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
 Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions.
@@
 The difference in his work is that he now knows about frameworks that let him do that systematically.
-He realized he was always building on top of other people’s software, and eventually realized that he should adopt those good software habits.
+He realized he was always building on top of other people's software, and eventually realized that he should adopt those good software habits.
 He also commented that
@@
  Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
-\\   &   http://www.software.ac.uk/resources/guides/software-management-plans &    & 
+\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    & 
 \\   &   we provide a software evaluation tool:   &    & 
-\\   &   http://www.software.ac.uk/online-sustainability-evaluation  &    & 
+\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    & 
 %
@@
 \\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
-\\ &  software http://arxiv.org/abs/1407.5117  Work with @arfon  &    & 
+\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    & 
 % 
@@
 \\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & & 
-\\ &   http://www.slideshare.net/kaythaney/     & & 
+\\ &   \url{http://www.slideshare.net/kaythaney/}     & & 
 \\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & & 
@@
 \\ & but want SWORD/DSpace/Fedora/ePrints see:  & & 
-\\ & http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/  & & 
+\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & & 
 %
 \\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
-\\ & Bioinformatics Manifesto: https://github.com/pjotrp/bioinformatics   & 
+\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   & 
 %
@@
 \\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
-\\   &  Research Software http://arxiv.org/abs/1407.5981  & & 
+\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & & 
 %
@@
 \\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
-\\ &   $+$ sustainability. Tune in! Live stream: http://ustre.am/17ddh & & 
+\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & & 
 %
 \\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
-\\ & http://wssspe.researchcomputing.org.uk/wssspe2/agenda/   &  &
-\\ & URL for live stream of keynotes \& lightning talks: http://ustre.am/17ddh   &  &
+\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
+\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
 \\ \hline

### 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
minor fixes, American English, quote marks, etc.

@@
 world-leading research, showing examples from disciplines as diverse as
-humanities and high-energy physics, quoting Kersten Kleese Van Dam of the
+humanities and high-energy physics, quoting Kerstin Kleese Van Dam of the
 Pacific Northwestern National Laboratory via a petition
@@
 A second approach is to fully embrace an open badging infrastructure (such as
-Mozilla's Open Badges), where a {\em badge\/} is a free, transferrable,
+Mozilla's Open Badges), where a {\em badge} is a free, transferrable,
 evidence-based indicator of an accomplishment, skill, quality, or interest. A
@@
 Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable.
-Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of "the community"?
+Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of ``the community''?
 Some communities have had significant initiatives to improve software discovery, e.g.
@@
 Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website.
-Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a licence or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}.
+Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a license or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}.
 Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.
@@
 
-\subsubsection{Second breakout discussion: Categorisation of software journals}
-
-This group aimed to come up with a categorisation of journals which published software papers.
+\subsubsection{Second breakout discussion: Categorization of software journals}
+
+This group aimed to come up with a categorization of journals which published software papers.
 Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers.
@@
 		\item Software should have a permanent identifier
-		\item Software deposit doesn?t count as a prior publication
+		\item Software deposit doesn't count as a prior publication
 	\end{itemize}
@@
 A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
-Journals might also learn from organisations such as the Apache Foundation as to what is good practice for software submissions.
+Journals might also learn from organizations such as the Apache Foundation as to what is good practice for software submissions.
 Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.

### 2015-06-04 WSSSPE2_report/WSSSPE2_review.tex
Update WSSSPE2_review.tex
Additions and updates to Exploring Sustainability section.

@@
 engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service
-as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions
-to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
+as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributionsto open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
 scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism
@@
 sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
-reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards
-developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies
-funding research projects where software was an intrinsic part of enabling the research program would be an additional
-element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would
-provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime
-of the project.
+reusable software~\cite{XXX}.
+\katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
+It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research program would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software.
+This type of initiative would provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime of the project.
+
+The discussion then focused on a need for a common language and a definition on the concept of sustainability that moved beyond the current fuzzy definitions where time was the simple measurement of sustainment.
+It was suggested that there was a need to identify tangible actions that underpin sustainability that developers could incorporate into the development stream of their software.
+This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{}.
+The focus of the discussion was based on the paper by Venters et.
+al., ~\cite{wssspe2_venters}, which suggested that sustainability is a first-class, composite, non-functional requirement composed of a number of sub-characteristics.
+It was generally agreed that maintainability and extensibility were key qualities underlying sustainability.
+In addition, the group also discussed what other non-functional requirements would contribute to the development of sustainable software e.g.
+reusability and scalability.
+However, it was recognized that there was a need to identify appropriate metrics and measures.
+
+The group also discussed whether the concept of sustainability was a barrier in itself to achieving sustainable software.
+It was suggested that "the first rule of software sustainability is, do not to talk about software sustainability".
+Instead there should be a focus on best software engineering practice.
+In order to play devils advocate, it was suggested that if the focus on the concept of sustainability was ignored, what current software engineering practices and principles could be utilized by software developers and domain scientists to achieve sustainability.
+This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{} was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community.
+As a result, there is a need to identify best practice and reach a consensus beyond the theories that underpinned the discipline of software engineering.
+Similarly, how could we translate or distill some of the key building blocks that underpin software engineering practice.
+
+A final key point of the discussion was the role that software design and patterns play before committing to writing a line of code.
+It was suggested that modeling must play a major role in attaining sustainable software.
+The point was raised that design involves making decisions and a mechanism would be necessary for capturing these decision points.
+This introduced the idea of software provenance that moved beyond commits in software repositories to how to capture and maintain relationships between sources and design decisions to prevent knowledge vaporization.
+Whether this could be achieved through software architectures was a open research challenge.
 

### 2015-06-04 WSSSPE2_report/WSSSPE2_review.tex
minor wording changes, flagging missing references

--- 2015-06-01 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-Each author was invited to outline the key action from their paper as a potential discussion point for the group; in
-their absence, the group facilitators outlined the actionable points from papers where the authors were not present.
+Each author was invited to outline the key action from their paper as a potential discussion point for the group; where the authors were not present,
+the group facilitators outlined the actionable points from their papers.
 The key actions from the six papers included recommendations for improving practice in sustainable software
 engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service
-as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions
-to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
+as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
 scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism
-for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took a position to view
+for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took the position of viewing
 sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
-reusable software~\cite{XXX}. \katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards
-developing sustainable software in academic environments. It was suggested that a new requirement driven by the agencies
-funding research projects where software was an intrinsic part of enabling the research program would be an additional
-element within the grant proposal to provide a sustainability plan for sustaining the software. This type of initiative would
-provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime
-of the project.
+reusable software~\cite{XXX}.
+\katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
+It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research program would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software.
+This type of initiative would provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime of the project.
+
+The discussion then focused on a need for a common language and a definition on the concept of sustainability that moved beyond the current fuzzy definitions where time was the simple measurement of sustainment.
+It was suggested that there was a need to identify tangible actions that underpin sustainability that developers could incorporate into the development stream of their software.
+This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{XXX}.
+\katznote{fix cite} The focus of the discussion was based on the paper by Venters et.
+al.~\cite{wssspe2_venters}, that suggested that sustainability is a first-class, composite, non-functional requirement composed of a number of sub-characteristics.
+It was generally agreed that maintainability and extensibility were key qualities underlying sustainability.
+In addition, the group also discussed what other non-functional requirements would contribute to the development of sustainable software e.g.
+reusability and scalability.
+However, it was recognized that there was a need to identify appropriate metrics and measures.
+
+The group also discussed whether the concept of sustainability itself was a barrier to achieving sustainable software.
+It was suggested that ``the first rule of software sustainability is do not to talk about software sustainability.'' Instead there should be a focus on best software engineering practice.
+Playing devil's advocate, it was asked that if the focus on the concept of sustainability was ignored, what current software engineering practices and principles could be utilized by software developers and domain scientists to achieve sustainability?
+This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{XXX} \katznote{fix cite}, was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community.
+As a result, there is a need to identify best practices and reach a consensus beyond the theories that underpin the discipline of software engineering.
+Similarly, how could we translate or distill some of the key building blocks that underpin software engineering practice?
+
+A final key point of the discussion was the role that software design and patterns play before committing to writing a line of code.
+It was suggested that modeling must play a major role in attaining sustainable software.
+The point was raised that design involves making decisions and a mechanism would be necessary for capturing these decision points.
+This introduced the idea of software provenance that moved beyond commits in software repositories to how to capture and maintain relationships between sources and design decisions to prevent knowledge vaporization.
+Whether this could be achieved through software architectures is a open research challenge.
 

### 2015-06-13 WSSSPE2_report/WSSSPE2_review.tex
major editing of Testing and Review section

@@
 %
-\thanks{{}$^{(10)}$ Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
+\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
 %
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan leading writing of this section, with help from Tom Clune}
+\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}
 
@@
 
-Several members of the discussion indicated that while they valued the
+A key point of discussion was that there are two major challenges to software
+testing in science. The first is convincing developers to incorporate testing. This 
+challenge is both social and technical. We need to communicate the value of testing
+and also teach developers how to choose and use testing frameworks. The second 
+challenge - and the one that the group felt was more difficult - is choosing  
+appropriate tests for scientific software. This is the difference between learning how to use Python 
+\texttt{unittest} and knowing how to test that your code correctly implements, for example, a 
+Lattice-Boltzmann model in computational fluid dynamics.
+
+\subsubsection{Choosing and implementing testing frameworks}
+There was general consensus that once a developer has been
+exposed to a testing framework in a nontrivial fashion, he/she will
+subsequently insist upon using a framework for further work.
+The acknowledged challenge was how to create that crucial first exposure
+to such frameworks.
+
+Within the group, there were several different paths by which participants had their 
+first exposure to code testing:
+\begin{itemize}
+\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and / or the domain of the participant.
+\item As a way to be confident in other software: When breaking out pieces of software from a larger application, a developer wants to ensure the software works as expected.
+These types of tests may be ephemeral, though, living long enough to give the author of the test confidence in the code, but not handed on and made available to next user.
+These experiences can lead to more systematic testing.
+\item Through experience with coding: As you write more software, your confidence in code is reduced and the importance of testing becomes clearer.
+Attendees joked that they started testing ``when their skepticism/guilt became larger than their arrogance''.
+\item From other projects: When building on top of projects with good testing frameworks, you realize that you should also adopt those good software engineering habits.  
+\end{itemize}
+
+The two non-self-taught paths were direct learning and indirect through other software (learning by example).
+The group recognized that teaching about testing needs to start early in projects and careers.
+We need to teach people how to test short bits of code, rather than waiting until they have thousands of lines of code and being surprised when they don't write tests.
+The group generated a set of practical suggestions for improving adoption of code testing.
+
+Established software projects can encourage the latter through a simple rule of accepting only software patches with associated tests and through good documentation of testing practices and requirements.
+
+Simpler, standard ways of setting up appropriate testing infrastructure will be important for adoption by scientists.
+Jenkins \footnote{\url{https://jenkins-ci.org/}} was suggested as a common open-source solution, but the initial configuration was considered to be challenging for typical scientists.
+
+Teach testing be beginning with a (smallish) 
+piece of code that lacks appropriate tests and develop an exercise that involves 
+refactoring the code into a more presentable form through creating / deriving unit 
+tests. Students would extract unit-tests through reverse-engineering and / or 
+questioning a partner who is a developer/expert.
+
+This lead into a more general discussion about whether programming courses should be part of the standard curriculum for science students and the challenges of fitting additional material into university curricula.
+This is the reasons that short workshops such as Software Carpentry and others exist.
+
+Ironically, Software Carpentry no longer teaches testing~\cite{SCtesting}.
+This boils down to two issues.
+First, that scientific computing doesn't (yet) have the cultural norms for error bars that experimental sciences have, and second, that there is a breathtaking diversity of scientific code; scientific research is highly specialized, which means that the tests scientists write are much less transferable or reusable than those found in other fields.
+
+\subsubsection{Writing tests for scientific software}
+The second major discussion thread in the group addressed the issues identified by Software Carpentry - the difficulties in writing tests appropriate for scientific software.
+Even among this group, while many What tests are appropriate to ensure that an complex method is working correctly?
+When is the result of a numerical computation 'close enough' to pass a test?
+
+Several members of the group indicated that while they valued the
 concept of software testing in theory, they were unaware how to test
@@
 testing that would be suitable for new members to their team.
-The group asked if it could produce a 100-line example of how to test
+As an example, the group asked if it could produce a 100-line example of how to test
 neutron transport (or other specific scientific examples) targeted at
@@
 
-Grid Wilson pointed to an explanation of why Software Carpentry doesn't teach testing any more~\cite{SCtesting}, which boils down to two issues.
-First, that scientific computing doesn't (yet) have the cultural norms for error bars that experimental sciences have, and second, that there is a breathtaking diversity of scientific code; scientific research is highly specialized, which means that the tests scientists write are much less transferable or reusable than those found in other fields.
-
-The group recognized that we need to teach people how to test short bits of code, we can't wait until we have thousands of lines of code.
-Greg commented, ``if we can't teach people how to test code until they're developing thousands of lines of code, we can't be surprised when they don't write tests.''
 %
@@
 %\clunenote{I no longer recall the context for this statement; does not seem important to me.}
-%
+
 Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.
-
-One general rule that was suggested is that projects should not accept
-software patches that don't have associated tests.
-This led to discussion about the fact that setting up appropriate
-infrastructure for testing can be difficult.
-The group would like a standard way that runs out of the box.
-Jenkins was suggested as a common open-source solution, but initial
-configuration is considered to be challenging for typical scientists.
-
-Several members of the group commented that once a developer has been
-exposed to a testing framework in a nontrivial fashion, he/she will
-subsequently insist upon using a framework for further work.
-The acknowledged challenge was how to create that crucial first exposure
-to such frameworks.
 
@@
 
-Some group members thought that every scientist should go through a programming class, but that was considered too big a goal for the group for the next year.
-Also, it's not clear where these classes could be offered.
-Software Carpentry exists because university curricula are full; they can't add extra classes.
-
-Tom Clune suggested that a useful demonstration could be generate by
-beginning with a (smallish) piece of badly engineered code that lacks
-appropriate unit tests.
-The code would then be refactored into a more presentable form once
-unit test were created/derived.
-However, such a collaboration must be with someone that either wrote
-the code or is intimately familiar with the subject matter.  The exercise would
-be to extract unit-tests through reverse-engineering {\b and}
-questioning the developer/expert.  Others suggested that the code does
-not have to be ``badly engineered'', but is either representative or
-is believed to be impenetrable from a unit-testing perspective.
-
-The group discussed how they learned testing.
-One attendee learned from a tutorial on test driven development (TDD) at a Software Developers Best Practices conference.
-The examples were {\b not} relevant, partly since he was working with Fortran code in his job, so the remainder of his learning was in the form of proving the process could be applied with Fortran.
-Another attendee said that most scientists have broken out pieces of software from a larger application and done some checking in an informal way.
-The difference in his work is that he now knows about frameworks that let him do that systematically.
-He realized he was always building on top of other people's software, and eventually realized that he should adopt those good software habits.
-He also commented that
-most of those initial checks were ephemeral.
-They live long enough to give the author of the test confidence in the code, then the author moves on; the checks  are not handed on and made available to next user as permanent tests.
-
-Attendees joked that they started testing when their skepticism/guilt became larger than their arrogance.
-
-The group suggested two papers by Hook and Kelly that would be useful for those interested in software testing to read~\cite{hook1,hook2}
-
 The group then decided that a useful action would be to create a set of scientific codes
-with associated tests. For each, they suggested:
+with associated tests. They developed the following basic structure for these examples 
+and sketched out six specific examples:
 \begin{enumerate}
-\item Write your name and email address so we can get back in touch with you
-\item Write a paragraph or two explaining the scientific problem the code addresses
-\item Tell us how big the simplest piece of relevant, interesting code would be (very rough estimate of lines of code)
-\item Give us a point-form list of the test cases you would use
+\item A paragraph or two explaining the scientific problem the code addresses
+\item The size of the simplest piece of relevant, interesting code (i.e. an estimate of lines of code)
+\item A point-form list of the test cases you would use
 \end{enumerate}
+
+The follow-up actions from this work include creating a compilation of testing examples in scientific software.
+Some of the examples from the workshop have become the starting seed of a collaborative book (https://github.com/swcarpentry/close-enough-for-scientific-work) in which scientists provide concrete examples of testing scientific software.
+The goal is a set of testing examples, aimed at sophomores in science and engineering, that cover a broad range of domains and problems and could be easily incorporated into other workshops or courses.
 
@@
 Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
+
+\item Daniel Hook and Diane Kelly. Mutation sensitivity testing. Computing in 
+science and engineering 11, no. 6 (2009): 40-47.~\cite{hook1}
+
+\item Daniel Hook and Diane Kelly. Testing for trustworthiness in scientific software.
+In Software Engineering for Computational Science and Engineering, 2009. 
+SECSE'09. ICSE Workshop on, pp. 59-64. IEEE, 2009.~\cite{hook2}
 

### 2015-06-13 WSSSPE2_report/WSSSPE2_review.tex
adding Karen as author

@@
 \\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
-\\Thomas L. Clune$^{(9)}$, Miguel de Val-Borro$^{(10)}$, Richard Littauer$^{(11)}$,
+\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$,
 \\more TBD by contributions}
 %\author{Gabrielle Allen$^{(11)}$,  \\
-%Karen Cranston$^{(4)}$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
+%$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
 %Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
@@
 %
-\thanks{{}$^{(9)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
-%
-\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
-%
-\thanks{{}$^{(11)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}
-%
-%\thanks{{}$^{(4)}$ National Evolutionary Synthesis Center, Durham, NC, USA}
+\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
+%
+\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
+%
+\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
+%
+\thanks{{}$^{(12)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}
 %
@@
 
+There is some discussion of the issue in the computer science literature~\cite{hook1,hook2} but less in domain-specific journals.
 Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.
@@
 Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
-
-\item Daniel Hook and Diane Kelly. Mutation sensitivity testing. Computing in 
-science and engineering 11, no. 6 (2009): 40-47.~\cite{hook1}
-
-\item Daniel Hook and Diane Kelly. Testing for trustworthiness in scientific software.
-In Software Engineering for Computational Science and Engineering, 2009. 
-SECSE'09. ICSE Workshop on, pp. 59-64. IEEE, 2009.~\cite{hook2}
 

### 2015-06-16 WSSSPE2_report/WSSSPE2_review.tex
Use en-dash for number range

@@
 reproducible scientific research through the open web. The second keynote
-speaker was Neil Chue Hong of Software Sustainability Institute. He shined a
+speaker was Neil Chue Hong of Software Sustainability Institute. He shed
 light on how scientific software is prevalently driving advances in many science
 and engineering fields. Both keynote speeches spawned further discussion among
-workshop participants on the crucial notion \emph{sustainability} in the theme
-of our workshop.
+workshop participants on the crucial notion of \emph{software sustainability}
+in the theme of our workshop.
 
@@
 domain sciences are not familiar with these techniques.  Thus, students who are being trained in these approaches
-are doing so through short 2-3 day training workshops such as Software and Data Carpentry, rather than through
+are doing so through short 2--3 day training workshops such as Software and Data Carpentry, rather than through
 semester-long graduate education courses at their University, which tend to focus on statistical and modeling

### 2015-06-19 WSSSPE2_report/WSSSPE2_review.tex
Use en-dash

--- 2015-06-16 WSSSPE2_report/WSSSPE2_review.tex
@@
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
-Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
+Building Sustainable Software -- The {CSDMS} Approach~\cite{wssspe2_hutton}
 
@@
 \begin{itemize}
-\item citation ecosystem - traces of usage (metrics)
+\item citation ecosystem -- traces of usage (metrics)
 \item taxonomy of contributorship, understanding roles
@@
 One way would be to establish tracks at conferences that subject papers to reproduction, which for those that pass would lend them an additional badge (similar to OOPSLA).
-Another is simply to get more people to use your software: for instance by outreach to high school students - can your software be used by them?
+Another is simply to get more people to use your software: for instance by outreach to high school students -- can your software be used by them?
 Finally, there is a role for community-curated benchmarks to validate the performance and capabilities of tools.
@@
 For example, a speedup of an algorithm may not result in a huge code difference, but may nonetheless provide a break-through.
-Journals should require software submissions to also provide ?Test materials?-- sample data, parameters to validate that code is working as intended.
+Journals should require software submissions to also provide ?Test materials?
+-- sample data, parameters to validate that code is working as intended.
 In certain cases, well-selected benchmark datasets may be required to assess performance and accuracy.
 
-Journals should ensure minimal metadata are provided, similarly as is already the case for certain kinds of data (though the latter is often enforced by data repositories)-- Dublin core-ish (creator, owner), and more specific (platform and compiler dependencies, sample benchmarks of performance, etc.).
+Journals should ensure minimal metadata are provided, similarly as is already the case for certain kinds of data (though the latter is often enforced by data repositories) -- Dublin core-ish (creator, owner), and more specific (platform and compiler dependencies, sample benchmarks of performance, etc.).
 Journals should provide guidance or constraints as to software licensing conditions.
@@
 and also teach developers how to choose and use testing frameworks. The second 
-challenge - and the one that the group felt was more difficult - is choosing  
+challenge -- and the one that the group felt was more difficult -- is choosing  
 appropriate tests for scientific software. This is the difference between learning how to use Python 

### 2015-06-19 WSSSPE2_report/WSSSPE2_review.tex
Use emph command for words between question marks

@@
 An idea that came from the groups was to define a set of support services that should be available to software paper reviewers.
-Another was to provide the ability for reviewers to flag the requirement for a ?software verification?, similar to ability to flag that a paper needs to be seen by a statistician.
+Another was to provide the ability for reviewers to flag the requirement for a \emph{software verification}, similar to ability to flag that a paper needs to be seen by a statistician.
 
@@
 For example, a speedup of an algorithm may not result in a huge code difference, but may nonetheless provide a break-through.
-Journals should require software submissions to also provide ?Test materials?
--- sample data, parameters to validate that code is working as intended.
+Journals should require software submissions to also provide \emph{Test materials} -- sample data, parameters to validate that code is working as intended.
 In certain cases, well-selected benchmark datasets may be required to assess performance and accuracy.

### 2015-06-19 WSSSPE2_report/WSSSPE2_review.tex
Typeset space as normal inter-word space

@@
 \begin{itemize}
-\item Funders can aim to get more software expertise on to funding review panels and provide more guidance on what is required of software related to research (cf. data management plans)
+\item Funders can aim to get more software expertise on to funding review panels and provide more guidance on what is required of software related to research (cf.\ data management plans)
 \item Journals can define and publish journal policies to improve reproducibility, and reviewers can insist that authors provide sufficient information and access to data and software to allow them reproduce the results in the paper.

### 2015-06-21 WSSSPE2_report/WSSSPE2_review.tex
undoing/changing some small edits

@@
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
-Building Sustainable Software -- The {CSDMS} Approach~\cite{wssspe2_hutton}
+Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
 
@@
 An idea that came from the groups was to define a set of support services that should be available to software paper reviewers.
-Another was to provide the ability for reviewers to flag the requirement for a \emph{software verification}, similar to ability to flag that a paper needs to be seen by a statistician.
-
-Other discussion focussed on ways in which researchers themselves could improve the reproducibility of research.
+Another was to provide the ability for reviewers to flag the requirement for a `software verification,' similar to ability to flag that a paper needs to be seen by a statistician.
+
+Other discussion focused on ways in which researchers themselves could improve the reproducibility of research.
 One way would be to establish tracks at conferences that subject papers to reproduction, which for those that pass would lend them an additional badge (similar to OOPSLA).
@@
 An issue around the sustainability of software catalogues is that their usefulness often depends on the domain.
-For instance, in the bio-sciences there is more homogenous data, and more standard shred codes.
+For instance, in the bio-sciences there is more homogeneous data, and more standard shred codes.
 In areas like ecology, codes are very specific to the problem, meaning that the level of re-use might be at a very general statistical level of abstraction, but then every research use is highly customized.
@@
 For example, a speedup of an algorithm may not result in a huge code difference, but may nonetheless provide a break-through.
-Journals should require software submissions to also provide \emph{Test materials} -- sample data, parameters to validate that code is working as intended.
+Journals should require software submissions to also provide `test materials' -- sample data, parameters to validate that code is working as intended.
 In certain cases, well-selected benchmark datasets may be required to assess performance and accuracy.
@@
 and also teach developers how to choose and use testing frameworks. The second 
-challenge -- and the one that the group felt was more difficult -- is choosing  
+challenge---and the one that the group felt was more difficult---is choosing  
 appropriate tests for scientific software. This is the difference between learning how to use Python 

### 2015-06-24 WSSSPE2_report/WSSSPE2_review.tex
Replace a -> an
Update WSSSPE2_review.tex
Missing reference tags added to Section 6

@@
 sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
-reusable software~\cite{XXX}.
-\katznote{fix cite} The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
+reusable software~\cite{katzJORS2014}.
+The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
 It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research program would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software.
@@
 It was suggested that there was a need to identify tangible actions that underpin sustainability that developers could incorporate into the development stream of their software.
-This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{XXX}.
-\katznote{fix cite} The focus of the discussion was based on the paper by Venters et.
+This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{iso25010}.
+The focus of the discussion was based on the paper by Venters et.
 al.~\cite{wssspe2_venters}, that suggested that sustainability is a first-class, composite, non-functional requirement composed of a number of sub-characteristics.
@@
 Playing devil's advocate, it was asked that if the focus on the concept of sustainability was ignored, what current software engineering practices and principles could be utilized by software developers and domain scientists to achieve sustainability?
-This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{XXX} \katznote{fix cite}, was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community.
+This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community.
 As a result, there is a need to identify best practices and reach a consensus beyond the theories that underpin the discipline of software engineering.
@@
 This introduced the idea of software provenance that moved beyond commits in software repositories to how to capture and maintain relationships between sources and design decisions to prevent knowledge vaporization.
-Whether this could be achieved through software architectures is a open research challenge.
+Whether this could be achieved through software architectures is an open-research challenge.
 

### 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
removing extra author info
Expand ICT abbreviation

@@
 \\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
-\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$,
-\\more TBD by contributions}
+\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
 %\author{Gabrielle Allen$^{(11)}$,  \\
@@
 their prime publication target. It is often suggested that this difference is
-due to the rapid developments in ICT science, a pace that can not be upheld by
+due to the rapid developments in information technology, a pace that can not be upheld by
 traditional peer-reviewed journals. Whatever the causes, any useful taxonomy of

### 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
old change, got lost...

--- 2015-06-24 WSSSPE2_report/WSSSPE2_review.tex
@@
 %
-\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}; \url{C.Venters@hud.ac.uk}}
 %

### 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
Remove extra bracket and add missing semicolons

--- 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
@@
 %
-\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}
+\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
 %
@@
 %
-\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands \url{F.Seinstra@esciencecenter.nl}}
+\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
 %
@@
 %
-\thanks{{}$^{(12)}$ MIT, Boston, MA, USA \& University of Saarland, Germany \url{richard.littauer@gmail.com}}
+\thanks{{}$^{(12)}$ MIT, Boston, MA, USA \& University of Saarland, Germany; \url{richard.littauer@gmail.com}}
 %

### 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex

--- 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
@@
 %
-% Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 1
-%\\     &  management plan used by EPSRC software grants    &    &
-% \\     &  http://w%ww.software.ac.uk/resources/guides/software-management-plans  &    &
-%
- Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
+Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 4
+\\     &  management plan used by EPSRC software grants    &    &
+\\     &  \url{http://www.software.ac.uk/resources/guides/software-management-plans}  &    &
+%
+\\  Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
 \\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    &
@@
 %
-\\ Karthik Ram  &  ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
-%
-\\ Josh Greenberg  &  ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
+\\ Karthik Ram  &  OH: ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+%
+\\ Josh Greenberg  &  @jameshowison: ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
 \\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    &
@@
 \\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    &
+%
+\\ Kaitlin Thaney  & Great point by @tracykteal about planning for ``end of life'' with scientific & 4 & 8
+\\ &  software projects and sustainability.  & &
 %

### 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex

@@
 %
-\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK}; \url{C.Venters@hud.ac.uk}}
+\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
 %

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
Keeping only organizations, but not departments, in author info.

@@
 %
-\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
+\thanks{{}$^{(1)}$ %Computation Institute, 
+University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
 %
@@
 %
-\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
+\thanks{{}$^{(4)}$ %Software Sustainability Institute, 
+University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
 %
@@
 %
-\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
+\thanks{{}$^{(11)}$ %Department of Astrophysical Sciences, 
+Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
 %

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
First page: clean up. No change in content.
re-adding units
Clean up section 2. No content change.
Section 2: minor changes.

--- 2015-06-25 WSSSPE2_report/WSSSPE2_review.tex
@@
 \\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
-%\author{Gabrielle Allen$^{(11)}$,  \\
-%$, Manish Parashar$^{(5)}$, David Proctor$^{(6)}$, \\
-%Matthew Turk$^{(2)}$,  Miguel de Val-Borro$^{(10)}$}
-%
-\thanks{{}$^{(1)}$ Computation Institute, University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
+%
+\thanks{{}$^{(1)}$ Computation Institute, 
+University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
 %
@@
 %
-\thanks{{}$^{(4)}$ Software Sustainability Institute, University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
+\thanks{{}$^{(4)}$ Software Sustainability Institute, 
+University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
 %
@@
 %
-\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
+\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
+Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
 %
@@
 %
-%\thanks{{}$^{(5)}$ Rutgers Discovery Informatics Institute, Rutgers University, New Brunswick, NJ, USA}
-%
-%\thanks{{}$^{(6)}$ International Consortium of Research Staff Associations, Dublin, Ireland}
-%
-%\thanks{{}$^{(9)}$ Saarland University, Saarbr\"{u}cken, Germany}
-%
-%\thanks{{}$^{(10)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA}
-%
-%\thanks{{}$^{(11)}$ University of Illinois, Champaign, IL, USA}
 
@@
 The report also contains a description of potential actions that were proposed
-in the five
-discussions.
+in the five discussions.
 \end{abstract}
@@
 
-\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
+%\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
 %https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
@@
 participants who would not otherwise be able to attend, and young participants
-and those from underrepresented groups, respectively. These funds allowed
-17 additional participants to attend, and each was offered the chance
-to present a lightning talk.
+and those from underrepresented groups, respectively. These funds allowed 17
+additional participants to attend, and each was offered the chance to present a
+lightning talk.
 
@@
 
-\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
+%\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
 
@@
 completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used EasyChair\footnote{\url{http://easychair.org/}}
-to record scores on relevance and comments. Finally, the organizers accessed the
-information to decide which papers to associate with the workshop and provided
-authors with the comments to help them improve their papers.
+4.1 reviews per reviewer), they used
+EasyChair\footnote{\url{http://easychair.org/}} to record scores on relevance
+and comments. Finally, the organizers accessed the information to decide which
+papers to associate with the workshop and provided authors with the comments to
+help them improve their papers.
 
@@
 
-The organizers wanted very interactive
-sessions, with the process of creating the sessions open to the full program
-committee, the paper authors, and others who might attend the workshop. In order
-to do this, the organizers used Well
-Sorted\footnote{\url{http://www.well-sorted.org}}, with the following
-steps:
+The organizers wanted very interactive sessions, with the process of creating
+the sessions open to the full program committee, the paper authors, and others
+who might attend the workshop. In order to do this, the organizers used Well
+Sorted\footnote{\url{http://www.well-sorted.org}}, with the following steps:
 \begin{enumerate}
-\item Authors were asked to create Well Sorted
-``cards'' for the papers. These cards have a title (50 characters maximum) and a
-body (255 characters maximum).
-\item Authors, program committee members, and
-members of the WSSSPE mailing list were asked to sort the cards. Each person
-drags the cards, one by one, into groups. A group can have as many cards as the
-person wants it to have, and it can have whatever meaning that makes sense to
-that person.
+%
+\item Authors were asked to create Well Sorted ``cards'' for the papers. These
+cards have a title (50 characters maximum) and a body (255 characters maximum).
+%
+\item Authors, members of the WSSSPE program committee and mailing list
+were asked to sort the cards. Each person dragged the cards into
+groups. A group could have as many cards as the person wanted it to have, and it
+could have whatever meaning that made sense to that person.
+%
 \item Well Sorted produced a set of averages of all the sorts, with
 various numbers of card clusters.
+%
 \end{enumerate}

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
Section 2: minor changes.

@@
 %
-\item Authors, members of the WSSSPE program committee and mailing list
+\item Authors as well as members of the WSSSPE program committee and mailing list
 were asked to sort the cards. Each person dragged the cards into
 groups. A group could have as many cards as the person wanted it to have, and it
-could have whatever meaning that made sense to that person.
+could have any meaning that made sense to that person.
 %

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
small change in 2

@@
 %
-\item Authors as well as members of the WSSSPE program committee and mailing list
-were asked to sort the cards. Each person dragged the cards into
+\item Authors, members of the WSSSPE program committee, and mailing list subscribers
+were asked to sort the cards. Each person dragged the cards, one by one, into
 groups. A group could have as many cards as the person wanted it to have, and it

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
Section 4: shorten urls. Section 5: Comment a note.

--- 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
@@
 United Kingdom (UK). The SSI was founded to support the UK's research software
-community by cultivating better, more sustainable research software to enable
+community by cultivating better, more sustainable software to enable
 world-class research. Chue Hong's keynote started by making the point
@@
 Pacific Northwestern National Laboratory via a petition
-campaign\footnote{\url{https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
+campaign\footnote{\url{http://tinyurl.com/nkn5tnv}}
+%https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
 at change.org, ``\emph{Today there are very few science areas left which do not
@@
 common in scientific research. A 2014
-survey\footnote{\url{http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
+survey\footnote{\url{http://tinyurl.com/ooajs7m}}
+%http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
 conducted by the SSI polled researchers from 15 research-intensive UK
@@
 
-\todo{writing led by Dan - other contributions welcome}
+%\todo{writing led by Dan - other contributions welcome}
 

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
Secion 6: Added line breaks. Commented out two notes.

@@
 \item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the Columbia
-University Center for International Earth Science Information Network (CIESIN),
-described a focus group study conducted with the Science Software Cluster (SSC)
-of the Federation of Earth Science Information Partners (ESIP). For the study,
-almost 300 attendees of the 2014 Summer ESIP Meeting were invited to participate
-in simultaneous roundtable discussions on sustainability of science software.
-Over two-thirds of the roundtable focus groups responded to a semi-structured
-survey that contained three sets of questions eliciting recommendations for
-near-term actions of the community to improve sustainable software practices.
-Initial analysis of the participants' responses to the questionnaire revealed
-several suggestions, which included improving community engagement and
-collaborative activities, increasing understanding and awareness, and creating
-incentives to motivate sustainable science software practices. The ESIP SSC
-plans to engage the community in the recommended activities for improving
-sustainable scientific software practices.
-
-\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by Fostering a
-Diverse Community~\cite{Boettiger_poster}.}
-Boettiger described how the rOpenSci project has been successful by focusing not
-just on building software but also on building a community of researchers who
-learn and adopt their approaches to reproducible research and sustainable
-software practice. Through outreach, mentoring, workshops, and hackathons, they
-have not only reached new users, but also turned users into co-developers of
-robust software and good practices to support data science research across a
-growing set of disciplines.
-
-\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis Environment~\cite{Blomer_poster}.}
-Large-scale scientific endeavors, such as the discovery of the Higgs boson at
-the Large Hadron Collider (LHC), often rely on complex software stacks.
-Maintaining thousands of dependencies of software libraries and operating system
-versions has shown that despite source code availability, the setup and the
-validation of a minimal usable analysis environment can easily become
-prohibitively expensive. In high-energy physics, CernVM-FS, a special-purpose, open-source,
-versioning, and snapshotting file system used to capture and
-distribute entire software stacks, proved to be useful for providing instant access to ready-to-run data
-analysis environments.
+Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the
+Columbia University Center for International Earth Science Information Network
+(CIESIN), described a focus group study conducted with the Science Software
+Cluster (SSC) of the Federation of Earth Science Information Partners (ESIP).
+For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were invited
+to participate in simultaneous roundtable discussions on sustainability of
+science software. Over two-thirds of the roundtable focus groups responded to a
+semi-structured survey that contained three sets of questions eliciting
+recommendations for near-term actions of the community to improve sustainable
+software practices. Initial analysis of the participants' responses to the
+questionnaire revealed several suggestions, which included improving community
+engagement and collaborative activities, increasing understanding and awareness,
+and creating incentives to motivate sustainable science software practices. The
+ESIP SSC plans to engage the community in the recommended activities for
+improving sustainable scientific software practices.
+
+\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by
+Fostering a Diverse Community~\cite{Boettiger_poster}.} Boettiger described how
+the rOpenSci project has been successful by focusing not just on building
+software but also on building a community of researchers who learn and adopt
+their approaches to reproducible research and sustainable software practice.
+Through outreach, mentoring, workshops, and hackathons, they have not only
+reached new users, but also turned users into co-developers of robust software
+and good practices to support data science research across a growing set of
+disciplines.
+
+\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis
+Environment~\cite{Blomer_poster}.} Large-scale scientific endeavors, such as the
+discovery of the Higgs boson at the Large Hadron Collider (LHC), often rely on
+complex software stacks. Maintaining thousands of dependencies of software
+libraries and operating system versions has shown that despite source code
+availability, the setup and the validation of a minimal usable analysis
+environment can easily become prohibitively expensive. In high-energy physics,
+CernVM-FS, a special-purpose, open-source, versioning, and snapshotting file
+system used to capture and distribute entire software stacks, proved to be
+useful for providing instant access to ready-to-run data analysis environments.
 % for the global scientific community as well as for
@@
 their software and is an excellent way to drive community change.
-\begin{comment}
-The ASCL actively looks for opportunities to use change management strategies to
-move astronomy to a more transparent discipline; these strategies can be used in
-other disciplines to further change in them.
-\end{comment}
 \end{enumerate}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Colin led writing of this section.}
-
-\note{\href{http://tinyurl.com/mpbhvyb}{Google doc notes}}
+%\todo{Colin led writing of this section.}
+
+%\note{\href{http://tinyurl.com/mpbhvyb}{Google doc notes}}
 %https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit
 
-Six papers were categorized under the theme of Exploring Sustainability. The group included four of the authors
-from the six papers submitted and an additional number of participants who expressed an interest in the theme at the
-workshop. Each paper had a different perspective on the concept of sustainability, which ranged from the sustainment of
-communities to defining sustainability as a first-class, composite non-functional requirement.
+Six papers were categorized under the theme of Exploring Sustainability. The
+group included four of the authors from the six papers submitted and an
+additional number of participants who expressed an interest in the theme at the
+workshop. Each paper had a different perspective on the concept of
+sustainability, which ranged from the sustainment of communities to defining
+sustainability as a first-class, composite non-functional requirement.
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 
-Each author was invited to outline the key action from their paper as a potential discussion point for the group; where the authors were not present,
-the group facilitators outlined the actionable points from their papers.
-The key actions from the six papers included recommendations for improving practice in sustainable software
-engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs}, the development of Software and Infrastructure as a Service
-as a mechanism for fostering sustainable science communities~\cite{wssspe2_patra}, developer incentives for code contributions to open source projects~\cite{wssspe2_pierce}, the establishment of a set software engineering principles based in
-scalability, reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software architectures as a mechanism
-for architectural-level reasoning about sustainability~\cite{wssspe2_venters}. The group took the position of viewing
-sustainability from the perspective of addressing the challenges related to the development, deployment, and maintenance of
-reusable software~\cite{katzJORS2014}.
-The principal focus of the initial discussion considered how to foster cultural change towards developing sustainable software in academic environments.
-It was suggested that a new requirement driven by the agencies funding research projects where software was an intrinsic part of enabling the research program would be an additional element within the grant proposal to provide a sustainability plan for sustaining the software.
-This type of initiative would provide the necessary incentive and motivation for researchers to consider how to sustain their software beyond the lifetime of the project.
-
-The discussion then focused on a need for a common language and a definition on the concept of sustainability that moved beyond the current fuzzy definitions where time was the simple measurement of sustainment.
-It was suggested that there was a need to identify tangible actions that underpin sustainability that developers could incorporate into the development stream of their software.
-This prompted a debate regarding whether sustainability should be considered as a non-functional requirement or software quality as defined within ISO/IEC 25010~\cite{iso25010}.
-The focus of the discussion was based on the paper by Venters et.
-al.~\cite{wssspe2_venters}, that suggested that sustainability is a first-class, composite, non-functional requirement composed of a number of sub-characteristics.
-It was generally agreed that maintainability and extensibility were key qualities underlying sustainability.
-In addition, the group also discussed what other non-functional requirements would contribute to the development of sustainable software e.g.
-reusability and scalability.
-However, it was recognized that there was a need to identify appropriate metrics and measures.
-
-The group also discussed whether the concept of sustainability itself was a barrier to achieving sustainable software.
-It was suggested that ``the first rule of software sustainability is do not to talk about software sustainability.'' Instead there should be a focus on best software engineering practice.
-Playing devil's advocate, it was asked that if the focus on the concept of sustainability was ignored, what current software engineering practices and principles could be utilized by software developers and domain scientists to achieve sustainability?
-This raised the question of why existing software engineering knowledge, such as that contained in SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the environment has a strong influence on practice within the scientific and engineering community.
-As a result, there is a need to identify best practices and reach a consensus beyond the theories that underpin the discipline of software engineering.
-Similarly, how could we translate or distill some of the key building blocks that underpin software engineering practice?
-
-A final key point of the discussion was the role that software design and patterns play before committing to writing a line of code.
-It was suggested that modeling must play a major role in attaining sustainable software.
-The point was raised that design involves making decisions and a mechanism would be necessary for capturing these decision points.
-This introduced the idea of software provenance that moved beyond commits in software repositories to how to capture and maintain relationships between sources and design decisions to prevent knowledge vaporization.
-Whether this could be achieved through software architectures is an open-research challenge.
+Each author was invited to outline the key action from their paper as a
+potential discussion point for the group; where the authors were not present,
+the group facilitators outlined the actionable points from their papers. The key
+actions from the six papers included recommendations for improving practice in
+sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs},
+the development of Software and Infrastructure as a Service as a mechanism for
+fostering sustainable science communities~\cite{wssspe2_patra}, developer
+incentives for code contributions to open source projects~\cite{wssspe2_pierce},
+the establishment of a set software engineering principles based in scalability,
+reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software
+architectures as a mechanism for architectural-level reasoning about
+sustainability~\cite{wssspe2_venters}. The group took the position of viewing
+sustainability from the perspective of addressing the challenges related to the
+development, deployment, and maintenance of reusable
+software~\cite{katzJORS2014}. The principal focus of the initial discussion
+considered how to foster cultural change towards developing sustainable software
+in academic environments. It was suggested that a new requirement driven by the
+agencies funding research projects where software was an intrinsic part of
+enabling the research program would be an additional element within the grant
+proposal to provide a sustainability plan for sustaining the software. This type
+of initiative would provide the necessary incentive and motivation for
+researchers to consider how to sustain their software beyond the lifetime of the
+project.
+
+The discussion then focused on a need for a common language and a definition on
+the concept of sustainability that moved beyond the current fuzzy definitions
+where time was the simple measurement of sustainment. It was suggested that
+there was a need to identify tangible actions that underpin sustainability that
+developers could incorporate into the development stream of their software. This
+prompted a debate regarding whether sustainability should be considered as a
+non-functional requirement or software quality as defined within ISO/IEC
+25010~\cite{iso25010}. The focus of the discussion was based on the paper by
+Venters et. al.~\cite{wssspe2_venters}, that suggested that sustainability is a
+first-class, composite, non-functional requirement composed of a number of
+sub-characteristics. It was generally agreed that maintainability and
+extensibility were key qualities underlying sustainability. In addition, the
+group also discussed what other non-functional requirements would contribute to
+the development of sustainable software e.g. reusability and scalability.
+However, it was recognized that there was a need to identify appropriate metrics
+and measures.
+
+The group also discussed whether the concept of sustainability itself was a
+barrier to achieving sustainable software. It was suggested that ``the first
+rule of software sustainability is do not to talk about software
+sustainability.'' Instead there should be a focus on best software engineering
+practice. Playing devil's advocate, it was asked that if the focus on the
+concept of sustainability was ignored, what current software engineering
+practices and principles could be utilized by software developers and domain
+scientists to achieve sustainability? This raised the question of why existing
+software engineering knowledge, such as that contained in
+SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the
+environment has a strong influence on practice within the scientific and
+engineering community. As a result, there is a need to identify best practices
+and reach a consensus beyond the theories that underpin the discipline of
+software engineering. Similarly, how could we translate or distill some of the
+key building blocks that underpin software engineering practice?
+
+A final key point of the discussion was the role that software design and
+patterns play before committing to writing a line of code. It was suggested that
+modeling must play a major role in attaining sustainable software. The point was
+raised that design involves making decisions and a mechanism would be necessary
+for capturing these decision points. This introduced the idea of software
+provenance that moved beyond commits in software repositories to how to capture
+and maintain relationships between sources and design decisions to prevent
+knowledge vaporization. Whether this could be achieved through software
+architectures is an open-research challenge.
 
@@
 
-The main action to come from the group was a proposal to identify the ten best software engineering practices
-similar to Philip Bourne's approach of Ten Simple Rules~\cite{bourne-ten-simple-rules}.
+The main action to come from the group was a proposal to identify the ten best
+software engineering practices similar to Philip Bourne's approach of Ten Simple
+Rules~\cite{bourne-ten-simple-rules}.
 

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex
undoing a change

@@
 United Kingdom (UK). The SSI was founded to support the UK's research software
-community by cultivating better, more sustainable software to enable
+community by cultivating better, more sustainable research software to enable
 world-class research. Chue Hong's keynote started by making the point

### 2015-06-26 WSSSPE2_report/WSSSPE2_review.tex

@@
 \\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & &
-\\ &   \url{http://www.slideshare.net/kaythaney/}     & &
-\\ &  designing-for-truth-scale-and-sustainability-wssspe2-keynote   & &
+\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{http://www.slideshare.net/kaythaney/}     & &
+\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{designing-for-truth-scale-and-sustainability-wssspe2-keynote} & &
 %

### 2015-06-27 WSSSPE2_report/WSSSPE2_review.tex
Section 2: Prototype: Footnote URLs as in-text hyperlinks.

@@
 %
-\thanks{{}$^{(12)}$ MIT, Boston, MA, USA \& University of Saarland, Germany; \url{richard.littauer@gmail.com}}
+\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
 %
@@
 
-31 submissions were received; all but one used
-arXiv\footnote{\url{http://arxiv.org}} or
-figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
+31 submissions were received; all but one used \href{http://arxiv.org}{arxiv}
+%arXiv\footnote{\url{http://arxiv.org}}  or figshare\footnote{\url{http://figshare.com}} 
+or \href{http://figshare.com}{figshare} to self-publish their papers.
 
@@
 completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used
-EasyChair\footnote{\url{http://easychair.org/}} to record scores on relevance
+4.1 reviews per reviewer), they used \href{http://easychair.org}{EasyChair} 
+%EasyChair\footnote{\url{http://easychair.org/}} 
+to record scores on relevance
 and comments. Finally, the organizers accessed the information to decide which
@@
 the sessions open to the full program committee, the paper authors, and others
-who might attend the workshop. In order to do this, the organizers used Well
-Sorted\footnote{\url{http://www.well-sorted.org}}, with the following steps:
+who might attend the workshop. In order to do this, the organizers used 
+\href{http://www.well-sorted.org}{Well Sorted},
+%Well Sorted\footnote{\url{http://www.well-sorted.org}}
+with the following steps:
 \begin{enumerate}
@@
 \\ Christopher Lenhardt
-\\ Richard Littauer      &  MIT \& University of Saarland
+\\ Richard Littauer      &  University of Saarland
 \\ Frank L\"{o}ffler     &  Louisiana State University

### 2015-06-27 WSSSPE2_report/WSSSPE2_review.tex
Fix a few typos.

@@
 contribution, and reuse where appropriate. We thought about measuring success
-through usage statistics (downloads, altmetrics). ``Should funders demand
+through usage statistics (downloads, Altmetrics). ``Should funders demand
 reproducibility?'' we wondered. In order for results to be reproduced, software
@@
 Society of America, etc. Perhaps this could be a joint technology/science
-partnership, for example, the [Apache $|$ Mozilla]--[AAAS $|$ disciplinary
+partnership, for example, the [Apache $|$ Mozilla]-[AAAS $|$ disciplinary
 society] prize.
@@
 tutorials;
-commits / patching;
+commits/patching;
 leadership transition; and
@@
 This group discussed five papers with a wide variety ideas of how to support reproducibility and reuse.
-It focussed on identifying concrete that the attendees could work together on, which would have a positive effect on the community.
+It focused on identifying concrete that the attendees could work together on, which would have a positive effect on the community.
 
@@
 A major issue is that a lot of software only builds on certain systems.
-Should journals provide more tools / support for reviewers, and if so, what is it?
+Should journals provide more tools/support for reviewers, and if so, what is it?
 An idea that came from the groups was to define a set of support services that should be available to software paper reviewers.
@@
 		\item APC charge is transparent
-		\item APC waiver programme
+		\item APC waiver program
 	\end{itemize}
@@
 	\end{itemize}
-	\item Papers can be updated when new releases of software are releases
-	\begin{itemize}
-		\item At no extra cost / at significantly reduced cost
+	\item Papers can be updated when new releases of software are releasesat
+	\begin{itemize}
+		\item At no extra cost/at significantly reduced cost
 	\end{itemize}
@@
 		\begin{itemize}
-			\item Of particular stature / with preservation plan
+			\item Of particular stature/with a preservation plan
 		\end{itemize}
@@
 
-This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a``Software Paper''.
+This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a ``Software Paper''.
 
@@
 \begin{itemize}
-\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and / or the domain of the participant.
+\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and/or the domain of the participant.
 \item As a way to be confident in other software: When breaking out pieces of software from a larger application, a developer wants to ensure the software works as expected.
@@
 piece of code that lacks appropriate tests and develop an exercise that involves 
-refactoring the code into a more presentable form through creating / deriving unit 
-tests. Students would extract unit-tests through reverse-engineering and / or 
+refactoring the code into a more presentable form through creating/deriving unit 
+tests. Students would extract unit-tests through reverse-engineering and/or 
 questioning a partner who is a developer/expert.

### 2015-06-27 WSSSPE2_report/WSSSPE2_review.tex
Added line breaks to long lines for readability.

@@
 
-Group discussion started with a position statement by each person surrounding what they had
-learned from their software development experiences and how those lessons might be translated
-into actionable outcomes.  Participants came from a range of backgrounds, and represented
-multiple software and training initiatives, including Software Carpentry, Data
-Carpentry, Open Science for Synthesis, the Community Surface Dynamics Modeling Systems group,
-ROpenSci, DataONE, the HDF Group, and others. Some of these software experiences were focused on development
-of new products for use in the sciences (e.g., ROpenSci, CSDMS), and these recounted the
-difficulties of engaging with disciplinary scientists in writing software.  Software was clearly
-utilized broadly across the various science disciplines represented, and it was developed within those
-disciplines as well.  Some researchers created software for statistical analysis and modeling, while
-others used it to control instrumentation, collate data across networks, collect information from
-respondents, and many other uses.  There was broad consensus that, within the disciplines represented,
-formal training in any type of software development or engineering was rare among the practitioners;
-most are self-taught, and develop software to get another job done. Any ancillary utility of the software
-outside of the specific science target was generally unplanned and few researchers would want to invest more
+Group discussion started with a position statement by each person surrounding
+what they had learned from their software development experiences and how those
+lessons might be translated into actionable outcomes. Participants came from a
+range of backgrounds, and represented multiple software and training
+initiatives, including Software Carpentry, Data Carpentry, Open Science for
+Synthesis, the Community Surface Dynamics Modeling Systems group, ROpenSci,
+DataONE, the HDF Group, and others. Some of these software experiences were
+focused on development of new products for use in the sciences (e.g., ROpenSci,
+CSDMS), and these recounted the difficulties of engaging with disciplinary
+scientists in writing software. Software was clearly utilized broadly across the
+various science disciplines represented, and it was developed within those
+disciplines as well. Some researchers created software for statistical analysis
+and modeling, while others used it to control instrumentation, collate data
+across networks, collect information from respondents, and many other uses.
+There was broad consensus that, within the disciplines represented, formal
+training in any type of software development or engineering was rare among the
+practitioners; most are self-taught, and develop software to get another job
+done. Any ancillary utility of the software outside of the specific science
+target was generally unplanned and few researchers would want to invest more
 time to make their own software more re-usable.
 
-Despite agreement that this body of disciplinary software needed improvement, there was general agreement that
-such improvement needed to be understood in terms of the scientific productivity that could be achieved. A
-software maturity model is needed for science software, but it needs to be introduced in a way that fits
-the culture of science, which largely thinks of software as a tool, rather than a product itself.
-The group was in general consensus that more widespread training in software practice is needed within the
-domain sciences, and several of the participants were involved in efforts along these lines.  Participants felt
-that projects that build a sense of community via training in software and technical practices would have the
-most success in changing practices in that community, but that there was a need for a managed introduction of
-these practices. Participants also recognized that these could not be one-time, one-off training opportunities,
-as software and technologies change rapidly over time.  For example, while today Software Carpentry is focused
-on teaching version control via Git, there has been a rapid evolution from RCS to CVS to SVN to Git over a short
-time frame, and thus communities should expect to need to train for adaptability and a changing technology tool
-chain, rather than assume these technologies stay fixed.  Thus, although short term training that introduces
-specific tools was considered highly valuable, these trainings were also not considered sufficient to engender the changes
-in software practice that were deemed necessary.  Combining the need for changes in practice to be introduced
-incrementally with the need to to minimize the divergence of training from direct science goals and the
-difficulties of training for a rapidly evolving technology space, participants concluded that multiple training
-efforts that targeted different parts of the spectrum were needed.  Short-term courses introducing immediately
-useful skills needs to be offered alongside more in-depth courses on software engineering and practice that allow
-students to adapt to a changing landscape.
-
-Finally, after agreeing that these complementary training models were needed, the group discussed sustainability of
-training, and how the leading groups in this space are teaching only a small fraction of the community that
-needs and wants training.  Most graduate programs in the sciences do not currently incorporate these approaches
-in their graduate curricula, although there are an increasing number of quantitatively oriented courses around
-analysis and modeling.  These still, however, generally omit engineering practices such as version control, unit
-and regression testing, and software modularity and abstraction, often because the instructors themselves in the
-domain sciences are not familiar with these techniques.  Thus, students who are being trained in these approaches
-are doing so through short 2--3 day training workshops such as Software and Data Carpentry, rather than through
-semester-long graduate education courses at their University, which tend to focus on statistical and modeling
-techniques.  Hybrid programs like the three week Open Science for Synthesis (OSS) training that combine the three
-(science, quantitative techniques, and software engineering) into an holistic course serve part of the need but
-reach only a few researchers at this time.  Thus, the group concluded that significant strides needed to be made
-in coordinating these trainings so that they are complementary, on increasing their size and scope, and on
-better integrating them directly into graduate education programs.
+Despite agreement that this body of disciplinary software needed improvement,
+there was general agreement that such improvement needed to be understood in
+terms of the scientific productivity that could be achieved. A software maturity
+model is needed for science software, but it needs to be introduced in a way
+that fits the culture of science, which largely thinks of software as a tool,
+rather than a product itself. The group was in general consensus that more
+widespread training in software practice is needed within the domain sciences,
+and several of the participants were involved in efforts along these lines.
+Participants felt that projects that build a sense of community via training in
+software and technical practices would have the most success in changing
+practices in that community, but that there was a need for a managed
+introduction of these practices. Participants also recognized that these could
+not be one-time, one-off training opportunities, as software and technologies
+change rapidly over time. For example, while today Software Carpentry is focused
+on teaching version control via Git, there has been a rapid evolution from RCS
+to CVS to SVN to Git over a short time frame, and thus communities should expect
+to need to train for adaptability and a changing technology tool chain, rather
+than assume these technologies stay fixed. Thus, although short term training
+that introduces specific tools was considered highly valuable, these trainings
+were also not considered sufficient to engender the changes in software practice
+that were deemed necessary. Combining the need for changes in practice to be
+introduced incrementally with the need to to minimize the divergence of training
+from direct science goals and the difficulties of training for a rapidly
+evolving technology space, participants concluded that multiple training efforts
+that targeted different parts of the spectrum were needed. Short-term courses
+introducing immediately useful skills needs to be offered alongside more
+in-depth courses on software engineering and practice that allow students to
+adapt to a changing landscape.
+
+Finally, after agreeing that these complementary training models were needed,
+the group discussed sustainability of training, and how the leading groups in
+this space are teaching only a small fraction of the community that needs and
+wants training. Most graduate programs in the sciences do not currently
+incorporate these approaches in their graduate curricula, although there are an
+increasing number of quantitatively oriented courses around analysis and
+modeling. These still, however, generally omit engineering practices such as
+version control, unit and regression testing, and software modularity and
+abstraction, often because the instructors themselves in the domain sciences are
+not familiar with these techniques. Thus, students who are being trained in
+these approaches are doing so through short 2--3 day training workshops such as
+Software and Data Carpentry, rather than through semester-long graduate
+education courses at their University, which tend to focus on statistical and
+modeling techniques. Hybrid programs like the three week Open Science for
+Synthesis (OSS) training that combine the three (science, quantitative
+techniques, and software engineering) into an holistic course serve part of the
+need but reach only a few researchers at this time. Thus, the group concluded
+that significant strides needed to be made in coordinating these trainings so
+that they are complementary, on increasing their size and scope, and on better
+integrating them directly into graduate education programs.
 
@@
 
-Three main actions were identified by Subgroup A that would be of interest to participants and that
-would benefit the research community.  These focused on the desire to amplify the
-current community efforts in training and software engineering by connecting the current training initiatives
-(e.g., software/data carpentry, rOpenSci, OSS) that are of different durations at appropriate stages. Generally it
-was felt that a modicum of interaction between short term workshops (SwC, DC), medium term trainings (e.g., OSS),
-and longer-term courses (e.g., BIDS Data Science) would benefit from coordinating curriculum,
-discussing and aligning prerequisites, and coordinating timing of courses.  A training
-coordination effort would go a long ways towards amplifying the value of the individual
-efforts and make them all more effective.
-
-\begin{itemize}
-    \item Action 1: Create a roadmap of research software training initiatives. Such a roadmap would
-        provide a taxonomy of training opportunities: what they deliver, and what attendees need to know
-        going in (prerequisites). It would also show which recommended roadmap actions will have a clear
-        and immediate payoff, and which will have long-term payoffs.  The training roadmap would emphasize
-        time savings and efficiency gains to be had from each training.
-
-    \item Action 2: Build a report card characterizing use of best practices in scientific software. Generally,
-        people felt that researchers would be very willing to migrate practices if they could identify
-        where they needed to improve. Such a report card would ask simple questions to characterize use of best
-        practices in science software, and could be structured similarly to Joel Spolsky's Software Maturity
-        questionnaire (also known as \href{http://www.joelonsoftware.com/articles/fog0000000043.html}{the Joel Test}~\cite{joel-test}).  The survey would create a
-        report card that shows areas where a project could improve, and then link those areas to specific
-        training offerings from the training taxonomy from Action 1.
-
-    \item Action 3: Create a science software review forum.  It was generally acknowledged that a little code review
-        can have a tremendous impact on the quality of software in a project, but that sites for science code
-        review are lacking.  While people can ask questions on the Stack Exchange sites, they are generally not open
-        to questions of style, approach, or appropriateness, as they try to avoid subjective commentary.   Instead,
-        we need a site where code gets discussed/summarized/described (in small bites) by the science software community.
-        The target audience would be graduate and undergraduate students in the sciences, and there would need to be
-        mechanisms to keep the review positive and constructive, and not get pedantic or judgmental.
-        This could be tied to a software registry (such as the nascent GeoSoft project), or to language repositories like
-        R's CRAN repository, and could lead to the report card discussed in Action 2.
+Three main actions were identified by Subgroup A that would be of interest to
+participants and that would benefit the research community. These focused on the
+desire to amplify the current community efforts in training and software
+engineering by connecting the current training initiatives (e.g., software/data
+carpentry, rOpenSci, OSS) that are of different durations at appropriate stages.
+Generally it was felt that a modicum of interaction between short term workshops
+(SwC, DC), medium term trainings (e.g., OSS), and longer-term courses (e.g.,
+BIDS Data Science) would benefit from coordinating curriculum, discussing and
+aligning prerequisites, and coordinating timing of courses. A training
+coordination effort would go a long ways towards amplifying the value of the
+individual efforts and make them all more effective.
+
+\begin{itemize}
+\item Action 1: Create a roadmap of research software training initiatives. Such
+a roadmap would provide a taxonomy of training opportunities: what they deliver,
+and what attendees need to know going in (prerequisites). It would also show
+which recommended roadmap actions will have a clear and immediate payoff, and
+which will have long-term payoffs. The training roadmap would emphasize time
+savings and efficiency gains to be had from each training.
+
+\item Action 2: Build a report card characterizing use of best practices in
+scientific software. Generally, people felt that researchers would be very
+willing to migrate practices if they could identify where they needed to
+improve. Such a report card would ask simple questions to characterize use of
+best practices in science software, and could be structured similarly to Joel
+Spolsky's Software Maturity questionnaire (also known as
+\href{http://www.joelonsoftware.com/articles/fog0000000043.html}{the Joel
+Test}~\cite{joel-test}). The survey would create a report card that shows areas
+where a project could improve, and then link those areas to specific training
+offerings from the training taxonomy from Action 1.
+
+\item Action 3: Create a science software review forum. It was generally
+acknowledged that a little code review can have a tremendous impact on the
+quality of software in a project, but that sites for science code review are
+lacking. While people can ask questions on the Stack Exchange sites, they are
+generally not open to questions of style, approach, or appropriateness, as they
+try to avoid subjective commentary. Instead, we need a site where code gets
+discussed/summarized/described (in small bites) by the science software
+community. The target audience would be graduate and undergraduate students in
+the sciences, and there would need to be mechanisms to keep the review positive
+and constructive, and not get pedantic or judgmental. This could be tied to a
+software registry (such as the nascent GeoSoft project), or to language
+repositories like R's CRAN repository, and could lead to the report card
+discussed in Action 2.
 \end{itemize}
@@
 just count citations today. A possible action that the group discussed was
-conducting a longitudinal study of
-most useful \{software, data, etc.\} in a discipline.
+conducting a longitudinal study of most useful \{software, data, etc.\} in a
+discipline.
 
@@
 \item This data (the credit file) should be part of the metadata that are freely
-available to those who have not paid for access to the journal, like citations are now in most cases.
+available to those who have not paid for access to the journal, like citations
+are now in most cases.
 \item It was suggested that there should also be a separate file to track
@@
 their prime publication target. It is often suggested that this difference is
-due to the rapid developments in information technology, a pace that can not be upheld by
-traditional peer-reviewed journals. Whatever the causes, any useful taxonomy of
-contributorship or guideline for tenure review should take such differences into
-account.
+due to the rapid developments in information technology, a pace that can not be
+upheld by traditional peer-reviewed journals. Whatever the causes, any useful
+taxonomy of contributorship or guideline for tenure review should take such
+differences into account.
 
@@
 
-Possible groups that might give out awards, either under our framework, or more generally,
-are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology
-Society of America, etc. Perhaps this could be a joint technology/science
-partnership, for example, the [Apache $|$ Mozilla]-[AAAS $|$ disciplinary
-society] prize.
+Possible groups that might give out awards, either under our framework, or more
+generally, are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology Society
+of America, etc. Perhaps this could be a joint technology/science partnership,
+for example, the [Apache $|$ Mozilla]-[AAAS $|$ disciplinary society] prize.
 
@@
 
-This group discussed five papers with a wide variety ideas of how to support reproducibility and reuse.
-It focused on identifying concrete that the attendees could work together on, which would have a positive effect on the community.
+This group discussed five papers with a wide variety ideas of how to support
+reproducibility and reuse. It focused on identifying concrete that the attendees
+could work together on, which would have a positive effect on the community.
 
@@
 
-In the first discussion session, this group broke up into two smaller groups working on different topics: Reproducibility, and Reuse and Sharing.
-In the second discussion session, three groups worked on creating specific pieces of guidance.
+In the first discussion session, this group broke up into two smaller groups
+working on different topics: Reproducibility, and Reuse and Sharing. In the
+second discussion session, three groups worked on creating specific pieces of
+guidance.
 
@@
 
-The first group discussed ways in which reproducibility of papers could be improved.
-A consensus surrounded the provision of examples: demonstrating to others how to achieve reproducibility.
-Major public funding investments go into research that heavily relies on software to be reproducible, hence lack thereof is raising major concerns.
-
-Two main avenues could be used to implement policy that would improve reproducibility and drive top-down culture change:
-\begin{itemize}
-\item Funders can aim to get more software expertise on to funding review panels and provide more guidance on what is required of software related to research (cf.\ data management plans)
-\item Journals can define and publish journal policies to improve reproducibility, and reviewers can insist that authors provide sufficient information and access to data and software to allow them reproduce the results in the paper.
-Stronger policies even for some high-impact journals have recently come into place, for example Nature Publishing journals.
-\end{itemize}
-
-If journals do enforce greater reproducibility constraints, it is important to lower the barriers to reviewers attempting to verifying the correctness of the software used to generate the results.
-A major issue is that a lot of software only builds on certain systems.
-Should journals provide more tools/support for reviewers, and if so, what is it?
-An idea that came from the groups was to define a set of support services that should be available to software paper reviewers.
-Another was to provide the ability for reviewers to flag the requirement for a `software verification,' similar to ability to flag that a paper needs to be seen by a statistician.
-
-Other discussion focused on ways in which researchers themselves could improve the reproducibility of research.
-One way would be to establish tracks at conferences that subject papers to reproduction, which for those that pass would lend them an additional badge (similar to OOPSLA).
-Another is simply to get more people to use your software: for instance by outreach to high school students -- can your software be used by them?
-Finally, there is a role for community-curated benchmarks to validate the performance and capabilities of tools.
+The first group discussed ways in which reproducibility of papers could be
+improved. A consensus surrounded the provision of examples: demonstrating to
+others how to achieve reproducibility. Major public funding investments go into
+research that heavily relies on software to be reproducible, hence lack thereof
+is raising major concerns.
+
+Two main avenues could be used to implement policy that would improve
+reproducibility and drive top-down culture change:
+\begin{itemize}
+\item Funders can aim to get more software expertise on to funding review panels
+and provide more guidance on what is required of software related to research
+(cf.\ data management plans)
+
+\item Journals can define and publish journal policies to improve
+reproducibility, and reviewers can insist that authors provide sufficient
+information and access to data and software to allow them reproduce the results
+in the paper. Stronger policies even for some high-impact journals have recently
+come into place, for example Nature Publishing journals.
+\end{itemize}
+
+If journals do enforce greater reproducibility constraints, it is important to
+lower the barriers to reviewers attempting to verifying the correctness of the
+software used to generate the results. A major issue is that a lot of software
+only builds on certain systems. Should journals provide more tools/support for
+reviewers, and if so, what is it? An idea that came from the groups was to
+define a set of support services that should be available to software paper
+reviewers. Another was to provide the ability for reviewers to flag the
+requirement for a `software verification,' similar to ability to flag that a
+paper needs to be seen by a statistician.
+
+Other discussion focused on ways in which researchers themselves could improve
+the reproducibility of research. One way would be to establish tracks at
+conferences that subject papers to reproduction, which for those that pass would
+lend them an additional badge (similar to OOPSLA). Another is simply to get more
+people to use your software: for instance by outreach to high school students --
+can your software be used by them? Finally, there is a role for
+community-curated benchmarks to validate the performance and capabilities of
+tools.
 
@@
 
-The second group discussed ways in which software could be more easily reused and shared.
-From the perspective of both user and developer, any solutions must be 1) easy; 2) cheap; 3) not take much time.
-
-Reuse and sharing were considered to be distinct but linked.
-In many cases, pre-existing software does not exist, so new software is written but even then it is not shared afterwards.
-The principal barriers to reuse are the difficulty of finding out what software is available, understanding if it is usable, and then of installing and running software if it is located.
-
-Discovery of relevant software is still fundamental issue: we need standard vocabularies and metadata, better tools, and approaches that are sustainable.
-Publications are an easy entry point for locating suitable software, but should the publishers lead the way, or is this the responsibility of ``the community''?
-Some communities have had significant initiatives to improve software discovery, e.g.
-the NIH Software Discovery Index\footnote{NIH Software Discovery Index: \url{http://softwarediscoveryindex.org/}}.
-Likewise, there were examples of journals which had made software more discoverable: ACM Transactions on Software offer a reproducibility review; the Journal of Biostatistics has an opt-in to provide code and a certification mark if it can be run; the Journal of Open Research Software requires software to be deposited in suitable repositories and referenced.
-
-An issue around the sustainability of software catalogues is that their usefulness often depends on the domain.
-For instance, in the bio-sciences there is more homogeneous data, and more standard shred codes.
-In areas like ecology, codes are very specific to the problem, meaning that the level of re-use might be at a very general statistical level of abstraction, but then every research use is highly customized.
-
-Finally, it was clear from discussion that there were many ways in which software could be made available in more reusable ways that just a tarball sitting on a personal website.
-Using code repositories like GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared platform and improve the reusability of your software incrementally, for instance by adding a license or by archiving (with a DOI) in Figshare\footnote{FigShare: \url{http://figshare.com/}} or Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}.
-Docker\footnote{Docker: \url{htp://www.docker.com/}} might be a solution to the issue of dependencies, to allow binaries and libraries to be bundled in a more lightweight fashion than a virtual machine image.
-
-The key enabler for reuse and sharing was to get domain scientists more effectively connected with programmers/analysts.
-Both have skills and experience which is necessary to make the right decisions for improving the reusability and discoverability of software, and to apply community pressure to change practice.
+The second group discussed ways in which software could be more easily reused
+and shared. From the perspective of both user and developer, any solutions must
+be 1) easy; 2) cheap; 3) not take much time.
+
+Reuse and sharing were considered to be distinct but linked. In many cases,
+pre-existing software does not exist, so new software is written but even then
+it is not shared afterwards. The principal barriers to reuse are the difficulty
+of finding out what software is available, understanding if it is usable, and
+then of installing and running software if it is located.
+
+Discovery of relevant software is still fundamental issue: we need standard
+vocabularies and metadata, better tools, and approaches that are sustainable.
+Publications are an easy entry point for locating suitable software, but should
+the publishers lead the way, or is this the responsibility of ``the community''?
+Some communities have had significant initiatives to improve software discovery,
+e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
+\url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of
+journals which had made software more discoverable: ACM Transactions on Software
+offer a reproducibility review; the Journal of Biostatistics has an opt-in to
+provide code and a certification mark if it can be run; the Journal of Open
+Research Software requires software to be deposited in suitable repositories and
+referenced.
+
+An issue around the sustainability of software catalogues is that their
+usefulness often depends on the domain. For instance, in the bio-sciences there
+is more homogeneous data, and more standard shred codes. In areas like ecology,
+codes are very specific to the problem, meaning that the level of re-use might
+be at a very general statistical level of abstraction, but then every research
+use is highly customized.
+
+Finally, it was clear from discussion that there were many ways in which
+software could be made available in more reusable ways that just a tarball
+sitting on a personal website. Using code repositories like
+GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared
+platform and improve the reusability of your software incrementally, for
+instance by adding a license or by archiving (with a DOI) in
+Figshare\footnote{FigShare: \url{http://figshare.com/}} or
+Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}. Docker\footnote{Docker:
+\url{htp://www.docker.com/}} might be a solution to the issue of dependencies,
+to allow binaries and libraries to be bundled in a more lightweight fashion than
+a virtual machine image.
+
+The key enabler for reuse and sharing was to get domain scientists more
+effectively connected with programmers/analysts. Both have skills and experience
+which is necessary to make the right decisions for improving the reusability and
+discoverability of software, and to apply community pressure to change practice.
 
@@
 
-This group aimed to come up with a categorization of journals which published software papers.
-Starting from the list of journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}} maintained by the Software Sustainability Institute, the group chose seven journals and looked at their advice to authors and reviewers.
-These were the Journal of Open Research Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}}, PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}}, Journal of Statistical Software\footnote{\url{http://www.jstatsoft.org/instructions}}.
-Methods in Ecology and Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}}, Transactions of Mathematical Software\footnote{\url{http://toms.acm.org/Authors.html}}, GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS Computational Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
-
-From these a set of common categories were synthesized, against which all journals could be compared:
+This group aimed to come up with a categorization of journals which published
+software papers. Starting from the list of
+journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}}
+maintained by the Software Sustainability Institute, the group chose seven
+journals and looked at their advice to authors and reviewers. These were the
+Journal of Open Research
+Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}},
+PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}},
+Journal of Statistical
+Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in
+Ecology and
+Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}},
+Transactions of Mathematical
+Software\footnote{\url{http://toms.acm.org/Authors.html}},
+GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS
+Computational
+Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
+
+From these a set of common categories were synthesized, against which all
+journals could be compared:
 
@@
 
-The follow-up actions to this work are to use this set of categories on all the journals in the list, refining the categories if necessary, then identify if any of the categories are seen to be more useful to promote reproducibility, reuse and sharing.
+The follow-up actions to this work are to use this set of categories on all the
+journals in the list, refining the categories if necessary, then identify if any
+of the categories are seen to be more useful to promote reproducibility, reuse
+and sharing.
 
@@
 
-This group discussed whether they could come up with a list of things that a journal should provide its reviewers to make it easier to review software submitted for publication as a ``Software Paper''.
-
-Journals should provide guidelines about what to consider when reviewing a software submission.
-A good example for a relatively comprehensive guidelines for reviewers (and thus in turn authors) are those of JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
-Journals might also learn from organizations such as the Apache Foundation as to what is good practice for software submissions.
-Guidance is needed on what constitutes an incremental improvement that is significant enough to qualify for publication, otherwise this assessment can be very subjective.
-
-Journals should provide mechanisms to enable and track communication between reviewers and authors.
-For anonymous peer review journals, maintain anonymity of reviewers.
-If communication is necessary, that may mean that software is not that well documented.
-Journals should also provide a set of simple metrics for software evaluation that reviewers can use for ratings, similar to Consumer Reports.
-
-Journals should provide guidelines about requirements for documentation of code: both in-lined in code, and manuals/web pages, etc.
-Journal editors could provide documentation checks before it goes out for review.
-This should include a requirement for good Use Cases specified for the software, with references to executable test cases that demonstrate each use-case is met (at least in the form of the test case).
-
-Journals should support mechanisms to run software.
-Sometimes this may be very hard to accomplish, despite best efforts.
-Journals could provide an Execution environment for any Software submitted for review, perhaps via a Docker container or a virtual machine.
-If not, instructions must be adequate to compile and execute the software; and interpret the results (output files, formats, etc.), and the full source code must be accessible to the reviewer.
-Mechanisms to quantify what has changed compared to a previously published version would assist version comparison.
-However, this cannot simply take lines of code into account.
-For example, a speedup of an algorithm may not result in a huge code difference, but may nonetheless provide a break-through.
-Journals should require software submissions to also provide `test materials' -- sample data, parameters to validate that code is working as intended.
-In certain cases, well-selected benchmark datasets may be required to assess performance and accuracy.
-
-Journals should ensure minimal metadata are provided, similarly as is already the case for certain kinds of data (though the latter is often enforced by data repositories) -- Dublin core-ish (creator, owner), and more specific (platform and compiler dependencies, sample benchmarks of performance, etc.).
-Journals should provide guidance or constraints as to software licensing conditions.
-
-The follow-up actions to this work are to work with journals and reviewers to identify whether any of these suggestions can be easily provided, perhaps for a range of journals.
+This group discussed whether they could come up with a list of things that a
+journal should provide its reviewers to make it easier to review software
+submitted for publication as a ``Software Paper''.
+
+Journals should provide guidelines about what to consider when reviewing a
+software submission. A good example for a relatively comprehensive guidelines
+for reviewers (and thus in turn authors) are those of
+JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
+Journals might also learn from organizations such as the Apache Foundation as to
+what is good practice for software submissions. Guidance is needed on what
+constitutes an incremental improvement that is significant enough to qualify for
+publication, otherwise this assessment can be very subjective.
+
+Journals should provide mechanisms to enable and track communication between
+reviewers and authors. For anonymous peer review journals, maintain anonymity of
+reviewers. If communication is necessary, that may mean that software is not
+that well documented. Journals should also provide a set of simple metrics for
+software evaluation that reviewers can use for ratings, similar to Consumer
+Reports.
+
+Journals should provide guidelines about requirements for documentation of code:
+both in-lined in code, and manuals/web pages, etc. Journal editors could provide
+documentation checks before it goes out for review. This should include a
+requirement for good Use Cases specified for the software, with references to
+executable test cases that demonstrate each use-case is met (at least in the
+form of the test case).
+
+Journals should support mechanisms to run software. Sometimes this may be very
+hard to accomplish, despite best efforts. Journals could provide an Execution
+environment for any Software submitted for review, perhaps via a Docker
+container or a virtual machine. If not, instructions must be adequate to compile
+and execute the software; and interpret the results (output files, formats,
+etc.), and the full source code must be accessible to the reviewer. Mechanisms
+to quantify what has changed compared to a previously published version would
+assist version comparison. However, this cannot simply take lines of code into
+account. For example, a speedup of an algorithm may not result in a huge code
+difference, but may nonetheless provide a break-through. Journals should require
+software submissions to also provide `test materials' -- sample data, parameters
+to validate that code is working as intended. In certain cases, well-selected
+benchmark datasets may be required to assess performance and accuracy.
+
+Journals should ensure minimal metadata are provided, similarly as is already
+the case for certain kinds of data (though the latter is often enforced by data
+repositories) -- Dublin core-ish (creator, owner), and more specific (platform
+and compiler dependencies, sample benchmarks of performance, etc.). Journals
+should provide guidance or constraints as to software licensing conditions.
+
+The follow-up actions to this work are to work with journals and reviewers to
+identify whether any of these suggestions can be easily provided, perhaps for a
+range of journals.
 
@@
 
-This group worked on defining a ``toolkit'' for running a reproducibility meta-track at a conference. They decided to take the work done during the workshop and publish it as a paper.
+This group worked on defining a ``toolkit'' for running a reproducibility
+meta-track at a conference. They decided to take the work done during the
+workshop and publish it as a paper.
 
@@
 
-This group began by recognizing that they know what the code testing and code review problems are, so the purpose of the group was really to think about concrete things that people in the group would be willing to commit to, in order to improve the sustainability of software.
+This group began by recognizing that they know what the code testing and code
+review problems are, so the purpose of the group was really to think about
+concrete things that people in the group would be willing to commit to, in order
+to improve the sustainability of software.
 
@@
 
-The consensus of the group was that small, concrete actions would have
-the greatest chance of proper followthrough by members of the group.
-An example of this approach was the Architecture of Open Source
-Applications book~\cite{aosa}, containing contributed chapters on open
-source software.
-The group contemplated creating an analogous book or web resource for
-testing of scientific software.
+The consensus of the group was that small, concrete actions would have the
+greatest chance of proper followthrough by members of the group. An example of
+this approach was the Architecture of Open Source Applications book~\cite{aosa},
+containing contributed chapters on open source software. The group contemplated
+creating an analogous book or web resource for testing of scientific software.
 
 A key point of discussion was that there are two major challenges to software
-testing in science. The first is convincing developers to incorporate testing. This 
-challenge is both social and technical. We need to communicate the value of testing
-and also teach developers how to choose and use testing frameworks. The second 
-challenge---and the one that the group felt was more difficult---is choosing  
-appropriate tests for scientific software. This is the difference between learning how to use Python 
-\texttt{unittest} and knowing how to test that your code correctly implements, for example, a 
-Lattice-Boltzmann model in computational fluid dynamics.
-
-\subsubsection{Choosing and implementing testing frameworks}
-There was general consensus that once a developer has been
-exposed to a testing framework in a nontrivial fashion, he/she will
-subsequently insist upon using a framework for further work.
-The acknowledged challenge was how to create that crucial first exposure
-to such frameworks.
+testing in science. The first is convincing developers to incorporate testing.
+This challenge is both social and technical. We need to communicate the value of
+testing and also teach developers how to choose and use testing frameworks. The
+second challenge---and the one that the group felt was more difficult---is
+choosing appropriate tests for scientific software. This is the difference
+between learning how to use Python \texttt{unittest} and knowing how to test
+that your code correctly implements, for example, a Lattice-Boltzmann model in
+computational fluid dynamics.
+
+\subsubsection{Choosing and implementing testing frameworks} There was general
+consensus that once a developer has been exposed to a testing framework in a
+nontrivial fashion, he/she will subsequently insist upon using a framework for
+further work. The acknowledged challenge was how to create that crucial first
+exposure to such frameworks.
 
@@
 \begin{itemize}
-\item Formal tutorials at conferences (for example, a test driven development (TDD) at a Software Developers Best Practices conference): These often require self-learning following the tutorial, and it can be hard to find a tutorial that matches the programming language and/or the domain of the participant.
-\item As a way to be confident in other software: When breaking out pieces of software from a larger application, a developer wants to ensure the software works as expected.
-These types of tests may be ephemeral, though, living long enough to give the author of the test confidence in the code, but not handed on and made available to next user.
-These experiences can lead to more systematic testing.
-\item Through experience with coding: As you write more software, your confidence in code is reduced and the importance of testing becomes clearer.
-Attendees joked that they started testing ``when their skepticism/guilt became larger than their arrogance''.
-\item From other projects: When building on top of projects with good testing frameworks, you realize that you should also adopt those good software engineering habits.  
-\end{itemize}
-
-The two non-self-taught paths were direct learning and indirect through other software (learning by example).
-The group recognized that teaching about testing needs to start early in projects and careers.
-We need to teach people how to test short bits of code, rather than waiting until they have thousands of lines of code and being surprised when they don't write tests.
-The group generated a set of practical suggestions for improving adoption of code testing.
-
-Established software projects can encourage the latter through a simple rule of accepting only software patches with associated tests and through good documentation of testing practices and requirements.
-
-Simpler, standard ways of setting up appropriate testing infrastructure will be important for adoption by scientists.
-Jenkins \footnote{\url{https://jenkins-ci.org/}} was suggested as a common open-source solution, but the initial configuration was considered to be challenging for typical scientists.
-
-Teach testing be beginning with a (smallish) 
-piece of code that lacks appropriate tests and develop an exercise that involves 
-refactoring the code into a more presentable form through creating/deriving unit 
-tests. Students would extract unit-tests through reverse-engineering and/or 
-questioning a partner who is a developer/expert.
-
-This lead into a more general discussion about whether programming courses should be part of the standard curriculum for science students and the challenges of fitting additional material into university curricula.
-This is the reasons that short workshops such as Software Carpentry and others exist.
-
-Ironically, Software Carpentry no longer teaches testing~\cite{SCtesting}.
-This boils down to two issues.
-First, that scientific computing doesn't (yet) have the cultural norms for error bars that experimental sciences have, and second, that there is a breathtaking diversity of scientific code; scientific research is highly specialized, which means that the tests scientists write are much less transferable or reusable than those found in other fields.
+
+\item Formal tutorials at conferences (for example, a test driven development
+(TDD) at a Software Developers Best Practices conference): These often require
+self-learning following the tutorial, and it can be hard to find a tutorial that
+matches the programming language and/or the domain of the participant.
+
+\item As a way to be confident in other software: When breaking out pieces of
+software from a larger application, a developer wants to ensure the software
+works as expected. These types of tests may be ephemeral, though, living long
+enough to give the author of the test confidence in the code, but not handed on
+and made available to next user. These experiences can lead to more systematic
+testing.
+
+\item Through experience with coding: As you write more software, your
+confidence in code is reduced and the importance of testing becomes clearer.
+Attendees joked that they started testing ``when their skepticism/guilt became
+larger than their arrogance''.
+
+\item From other projects: When building on top of projects with good testing
+frameworks, you realize that you should also adopt those good software
+engineering habits.
+
+\end{itemize}
+
+The two non-self-taught paths were direct learning and indirect through other
+software (learning by example). The group recognized that teaching about testing
+needs to start early in projects and careers. We need to teach people how to
+test short bits of code, rather than waiting until they have thousands of lines
+of code and being surprised when they don't write tests. The group generated a
+set of practical suggestions for improving adoption of code testing.
+
+Established software projects can encourage the latter through a simple rule of
+accepting only software patches with associated tests and through good
+documentation of testing practices and requirements.
+
+Simpler, standard ways of setting up appropriate testing infrastructure will be
+important for adoption by scientists. Jenkins
+\footnote{\url{https://jenkins-ci.org/}} was suggested as a common open-source
+solution, but the initial configuration was considered to be challenging for
+typical scientists.
+
+Teach testing be beginning with a (smallish) piece of code that lacks
+appropriate tests and develop an exercise that involves refactoring the code
+into a more presentable form through creating/deriving unit tests. Students
+would extract unit-tests through reverse-engineering and/or questioning a
+partner who is a developer/expert.
+
+This lead into a more general discussion about whether programming courses
+should be part of the standard curriculum for science students and the
+challenges of fitting additional material into university curricula. This is the
+reasons that short workshops such as Software Carpentry and others exist.
+
+Ironically, Software Carpentry no longer teaches testing~\cite{SCtesting}. This
+boils down to two issues. First, that scientific computing doesn't (yet) have
+the cultural norms for error bars that experimental sciences have, and second,
+that there is a breathtaking diversity of scientific code; scientific research
+is highly specialized, which means that the tests scientists write are much less
+transferable or reusable than those found in other fields.
 
 \subsubsection{Writing tests for scientific software}
-The second major discussion thread in the group addressed the issues identified by Software Carpentry - the difficulties in writing tests appropriate for scientific software.
-Even among this group, while many What tests are appropriate to ensure that an complex method is working correctly?
-When is the result of a numerical computation 'close enough' to pass a test?
-
-Several members of the group indicated that while they valued the
-concept of software testing in theory, they were unaware how to test
-certain kinds of software.
-Or, more importantly, they lacked relevant examples of software
-testing that would be suitable for new members to their team.
-As an example, the group asked if it could produce a 100-line example of how to test
-neutron transport (or other specific scientific examples) targeted at
-a sophomore? This would be a demonstration of ``How do I do this
-right?''
-
-Another question was ``How do I test so that I know that the code is
-not the source of my problems?''  A common challenge is that of
-stripping things down to an appropriate level for tests.  The group
-recognized that the difficulty is different for so-called software
-infrastructure than for numerical/scientific layers of software
-applications.
+The second major discussion thread in the group addressed the issues identified
+by Software Carpentry - the difficulties in writing tests appropriate for
+scientific software. Even among this group, while many What tests are
+appropriate to ensure that an complex method is working correctly? When is the
+result of a numerical computation 'close enough' to pass a test?
+
+Several members of the group indicated that while they valued the concept of
+software testing in theory, they were unaware how to test certain kinds of
+software. Or, more importantly, they lacked relevant examples of software
+testing that would be suitable for new members to their team. As an example, the
+group asked if it could produce a 100-line example of how to test neutron
+transport (or other specific scientific examples) targeted at a sophomore? This
+would be a demonstration of ``How do I do this right?''
+
+Another question was ``How do I test so that I know that the code is not the
+source of my problems?'' A common challenge is that of stripping things down to
+an appropriate level for tests. The group recognized that the difficulty is
+different for so-called software infrastructure than for numerical/scientific
+layers of software applications.
 
@@
 
-There is some discussion of the issue in the computer science literature~\cite{hook1,hook2} but less in domain-specific journals.
-Some lessons on multiphysics software verification can be found in a recent paper~\cite{dubey-FLASH}.
-
-Two important pieces of the barrier are a) picking/implementing testing frameworks (a technical barrier) and b) deciding what are the actual tests that I need to write?
-A possible method to solve the latter is to ask students ``how is what you are doing different from typing random keys on the keyboard?'' and then, turn their answer into the concept of tests.
-That is to say, expand the process of creating code from one that is solely about ``how?'' into one that also include ``why?''
-
-Another point of discussion is that many scientists are not aware that ``testing the code'' and ``testing the science'' are distinct issues.
-A related question that came up is ``How do you know that your method produces a result that is `close enough'?''
-
-The group then decided that a useful action would be to create a set of scientific codes
-with associated tests. They developed the following basic structure for these examples 
-and sketched out six specific examples:
+There is some discussion of the issue in the computer science
+literature~\cite{hook1,hook2} but less in domain-specific journals. Some lessons
+on multiphysics software verification can be found in a recent
+paper~\cite{dubey-FLASH}.
+
+Two important pieces of the barrier are a) picking/implementing testing
+frameworks (a technical barrier) and b) deciding what are the actual tests that
+I need to write? A possible method to solve the latter is to ask students ``how
+is what you are doing different from typing random keys on the keyboard?'' and
+then, turn their answer into the concept of tests. That is to say, expand the
+process of creating code from one that is solely about ``how?'' into one that
+also include ``why?''
+
+Another point of discussion is that many scientists are not aware that ``testing
+the code'' and ``testing the science'' are distinct issues. A related question
+that came up is ``How do you know that your method produces a result that is
+`close enough'?''
+
+The group then decided that a useful action would be to create a set of
+scientific codes with associated tests. They developed the following basic
+structure for these examples and sketched out six specific examples:
 \begin{enumerate}
+
 \item A paragraph or two explaining the scientific problem the code addresses
-\item The size of the simplest piece of relevant, interesting code (i.e. an estimate of lines of code)
+
+\item The size of the simplest piece of relevant, interesting code (i.e. an
+estimate of lines of code)
+
 \item A point-form list of the test cases you would use
+
 \end{enumerate}
 
-The follow-up actions from this work include creating a compilation of testing examples in scientific software.
-Some of the examples from the workshop have become the starting seed of a collaborative book (https://github.com/swcarpentry/close-enough-for-scientific-work) in which scientists provide concrete examples of testing scientific software.
-The goal is a set of testing examples, aimed at sophomores in science and engineering, that cover a broad range of domains and problems and could be easily incorporated into other workshops or courses.
+The follow-up actions from this work include creating a compilation of testing
+examples in scientific software. Some of the examples from the workshop have
+become the starting seed of a collaborative book
+(https://github.com/swcarpentry/close-enough-for-scientific-work) in which
+scientists provide concrete examples of testing scientific software. The goal is
+a set of testing examples, aimed at sophomores in science and engineering, that
+cover a broad range of domains and problems and could be easily incorporated
+into other workshops or courses.
 

### 2015-06-27 WSSSPE2_report/WSSSPE2_review.tex
Remove unncessary curly brackets.
undoing some changes

@@
 
-31 submissions were received; all but one used \href{http://arxiv.org}{arxiv}
-%arXiv\footnote{\url{http://arxiv.org}}  or figshare\footnote{\url{http://figshare.com}} 
-or \href{http://figshare.com}{figshare} to self-publish their papers.
+31 submissions were received; all but one used arXiv\footnote{\url{http://arxiv.org}}
+or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
 
@@
 completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used \href{http://easychair.org}{EasyChair} 
-%EasyChair\footnote{\url{http://easychair.org/}} 
+4.1 reviews per reviewer), they used EasyChair\footnote{\url{http://easychair.org/}} 
 to record scores on relevance
@@
 who might attend the workshop. In order to do this, the organizers used 
-\href{http://www.well-sorted.org}{Well Sorted},
-%Well Sorted\footnote{\url{http://www.well-sorted.org}}
-with the following steps:
+Well Sorted\footnote{\url{http://www.well-sorted.org}} with the following steps:
 \begin{enumerate}
@@
 contribution, and reuse where appropriate. We thought about measuring success
-through usage statistics (downloads, Altmetrics). ``Should funders demand
+through usage statistics (downloads, altmetrics). ``Should funders demand
 reproducibility?'' we wondered. In order for results to be reproduced, software
@@
 just count citations today. A possible action that the group discussed was
-conducting a longitudinal study of most useful \{software, data, etc.\} in a
+conducting a longitudinal study of most useful software, data, etc.\ in a
 discipline.
@@
 	\end{itemize}
-	\item Papers can be updated when new releases of software are releasesat
+	\item Papers can be updated when new releases of software are released
 	\begin{itemize}

### 2015-06-29 WSSSPE2_report/WSSSPE2_review.tex
Use ASF abbreviation on first occurrence

@@
 contributions? There is evidence that the migration of scientific software
-projects to the Apache Foundation has created opportunities for those not
+projects to the Apache Software Foundation (ASF) has created opportunities for those not
 employed in the scientific center to contribute to projects initiated by
@@
 \begin{itemize}
-\item Converting reputation by joining the Apache Software Foundation, Google, etc.
+\item Converting reputation by joining the ASF, Google, etc.
 \item Inviting ASF and open source people to contribute to scientific code (at
@@
 JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
-Journals might also learn from organizations such as the Apache Foundation as to
+Journals might also learn from organizations such as the ASF as to
 what is good practice for software submissions. Guidance is needed on what

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 6: Used semicolons instead of commas. Added line breaks. Fixed some typos.

@@
 sustainability as a first-class, composite non-functional requirement.
+
 %~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
@@
 actions from the six papers included recommendations for improving practice in
-sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs},
-the development of Software and Infrastructure as a Service as a mechanism for
-fostering sustainable science communities~\cite{wssspe2_patra}, developer
-incentives for code contributions to open source projects~\cite{wssspe2_pierce},
-the establishment of a set software engineering principles based in scalability,
-reproducibility, and energy efficiency~\cite{wssspe2_shi}, and applying software
+sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs};
+development of \emph{Software and Infrastructure as a Service} as a mechanism for
+fostering sustainable science communities~\cite{wssspe2_patra}; developer
+incentives for code contributions to open source projects~\cite{wssspe2_pierce};
+establishment of a set of software engineering principles based on scalability,
+reproducibility, and energy efficiency~\cite{wssspe2_shi}; and applying software
 architectures as a mechanism for architectural-level reasoning about
@@
 development, deployment, and maintenance of reusable
-software~\cite{katzJORS2014}. The principal focus of the initial discussion
+software~\cite{katzJORS2014}.
+
+The principal focus of the initial discussion
 considered how to foster cultural change towards developing sustainable software
@@
 25010~\cite{iso25010}. The focus of the discussion was based on the paper by
-Venters et. al.~\cite{wssspe2_venters}, that suggested that sustainability is a
+Venters et al.~\cite{wssspe2_venters}, which suggested that sustainability is a
 first-class, composite, non-functional requirement composed of a number of
@@
 group also discussed what other non-functional requirements would contribute to
-the development of sustainable software e.g. reusability and scalability.
+the development of sustainable software, e.g., reusability and scalability.
 However, it was recognized that there was a need to identify appropriate metrics
@@
 barrier to achieving sustainable software. It was suggested that ``the first
-rule of software sustainability is do not to talk about software
+rule of software sustainability is do not talk about software
 sustainability.'' Instead there should be a focus on best software engineering

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7: minor changes.

@@
 
-Despite agreement that this body of disciplinary software needed improvement,
-there was general agreement that such improvement needed to be understood in
+%Despite agreement that this body of disciplinary software needed improvement,
+There was general agreement that this body of disciplinary software improvement needed to be understood in
 terms of the scientific productivity that could be achieved. A software maturity
@@
 to CVS to SVN to Git over a short time frame, and thus communities should expect
-to need to train for adaptability and a changing technology tool chain, rather
-than assume these technologies stay fixed. Thus, although short term training
+the need to train for adaptability and a changing technology tool chain, rather
+than assume that these technologies will stay fixed. Thus, although short term training
 that introduces specific tools was considered highly valuable, these trainings
@@
 that were deemed necessary. Combining the need for changes in practice to be
-introduced incrementally with the need to to minimize the divergence of training
+introduced incrementally with the need to minimize the divergence of training
 from direct science goals and the difficulties of training for a rapidly

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7: Line breaks.
Section 7.1.1: minor changes.

@@
 %Despite agreement that this body of disciplinary software needed improvement,
-There was general agreement that this body of disciplinary software improvement needed to be understood in
-terms of the scientific productivity that could be achieved. A software maturity
-model is needed for science software, but it needs to be introduced in a way
-that fits the culture of science, which largely thinks of software as a tool,
-rather than a product itself. The group was in general consensus that more
-widespread training in software practice is needed within the domain sciences,
-and several of the participants were involved in efforts along these lines.
-Participants felt that projects that build a sense of community via training in
-software and technical practices would have the most success in changing
-practices in that community, but that there was a need for a managed
-introduction of these practices. Participants also recognized that these could
-not be one-time, one-off training opportunities, as software and technologies
-change rapidly over time. For example, while today Software Carpentry is focused
-on teaching version control via Git, there has been a rapid evolution from RCS
-to CVS to SVN to Git over a short time frame, and thus communities should expect
-the need to train for adaptability and a changing technology tool chain, rather
-than assume that these technologies will stay fixed. Thus, although short term training
-that introduces specific tools was considered highly valuable, these trainings
-were also not considered sufficient to engender the changes in software practice
-that were deemed necessary. Combining the need for changes in practice to be
-introduced incrementally with the need to minimize the divergence of training
-from direct science goals and the difficulties of training for a rapidly
-evolving technology space, participants concluded that multiple training efforts
-that targeted different parts of the spectrum were needed. Short-term courses
-introducing immediately useful skills needs to be offered alongside more
-in-depth courses on software engineering and practice that allow students to
-adapt to a changing landscape.
+There was general agreement that this body of disciplinary software improvement
+needed to be understood in terms of the scientific productivity that could be
+achieved. A software maturity model is needed for science software, but it needs
+to be introduced in a way that fits the culture of science, which largely thinks
+of software as a tool, rather than a product itself. The group was in general
+consensus that more widespread training in software practice is needed within
+the domain sciences, and several of the participants were involved in efforts
+along these lines. Participants felt that projects that build a sense of
+community via training in software and technical practices would have the most
+success in changing practices in that community, but that there was a need for a
+managed introduction of these practices. Participants also recognized that these
+could not be one-time, one-off training opportunities, as software and
+technologies change rapidly over time. For example, while today Software
+Carpentry is focused on teaching version control via Git, there has been a rapid
+evolution from RCS to CVS to SVN to Git over a short time frame, and thus
+communities should expect the need to train for adaptability and a changing
+technology tool chain, rather than assume that these technologies will stay
+fixed. Thus, although short term training that introduces specific tools was
+considered highly valuable, these trainings were also not considered sufficient
+to engender the changes in software practice that were deemed necessary.
+Combining the need for changes in practice to be introduced incrementally with
+the need to minimize the divergence of training from direct science goals and
+the difficulties of training for a rapidly evolving technology space,
+participants concluded that multiple training efforts that targeted different
+parts of the spectrum were needed. Short-term courses introducing immediately
+useful skills needs to be offered alongside more in-depth courses on software
+engineering and practice that allow students to adapt to a changing landscape.
 
@@
 wants training. Most graduate programs in the sciences do not currently
-incorporate these approaches in their graduate curricula, although there are an
+incorporate these approaches in their graduate curricula, although there is an
 increasing number of quantitatively oriented courses around analysis and
@@
 Software and Data Carpentry, rather than through semester-long graduate
-education courses at their University, which tend to focus on statistical and
-modeling techniques. Hybrid programs like the three week Open Science for
+education courses at their universities, which tend to focus on statistical and
+modeling techniques. Hybrid programs like the three-week Open Science for
 Synthesis (OSS) training that combine the three (science, quantitative

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
no message

@@
 several different training models, from a two-day Software Carpentry workshop,
-a three-week Open Science for Synthesis (OSS) program, to a semester-long
+a three-week Open Science for Synthesis (OSS) program, to semester-long
 programs. We discussed the pros and cons of different training approaches,
-touching on informal learning, for example, where people learning the necessary
+touching on informal learning, for example, where people learn the necessary
 skills by asking questions of cross-disciplinary people (``boundary scientists'') in
@@
 everyone knows Git.'' Developers also need training on licensing choices. Just
-because a code is on Git does not mean that it is open.
+because some source code is on Git does not mean that it is open.
 
@@
 
-We asked ourselves, ``Where are there examples of things that are changing
+We asked ourselves, ``Are there examples of things that are changing
 because of this and how can we build momentum?'' One example demonstrates change

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7.1.2: minor changes.

@@
 discussed certification. It can be hard to build the recognition of Java or
-database certifications amongst all technologies. OSS offers badges to those
+database certifications among all technologies. OSS offers badges to those
 completing training. However we need to demonstrate proficiency, not just
@@
 
-We then discussed variations amongst scientific domain areas and wondered ``Are
-some communities more or less open than others?'' To some members of the group
-biology seems to be open, while physics less so. Some felt with the more recent
-development of bioinformatics as a field there were fewer historical practices
-to undo. Physics has pre-print philosophy to overcome. Environmental sciences
-may be mixed. The group felt the biomedical area, however, was very competitive
+We then discussed variations among scientific domain areas and wondered ``Are
+some communities more or less open than others?'' To some members of the group,
+biology seems to be more open, while physics less so. Some felt, with the more recent
+development of bioinformatics as a field, there were fewer historical practices
+to undo in biology. Physics has preprint philosophy to overcome. Environmental sciences
+may be mixed. The group felt that the biomedical area, however, was very competitive
 and closed. One participant mentioned blueprints for going open source (like

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7: minor changes. Not done yet. Commit before taking a short break.

@@
 range of backgrounds, and represented multiple software and training
-initiatives, including Software Carpentry, Data Carpentry, Open Science for
-Synthesis, the Community Surface Dynamics Modeling Systems group, ROpenSci,
+initiatives, including Software Carpentry (SwC), Data Carpentry (DC), Open Science for
+Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
 DataONE, the HDF Group, and others. Some of these software experiences were
@@
 Three main actions were identified by Subgroup A that would be of interest to
-participants and that would benefit the research community. These focused on the
+participants and benefit the research community. These focused on the
 desire to amplify the current community efforts in training and software
-engineering by connecting the current training initiatives (e.g., software/data
-carpentry, rOpenSci, OSS) that are of different durations at appropriate stages.
+engineering by connecting the current training initiatives (e.g., SwC, DC, 
+rOpenSci, OSS) that are of different durations at appropriate stages.
 Generally it was felt that a modicum of interaction between short term workshops
@@
 aligning prerequisites, and coordinating timing of courses. A training
-coordination effort would go a long ways towards amplifying the value of the
+coordination effort would go a long way towards amplifying the value of the
 individual efforts and make them all more effective.
@@
 Consortium for Political and Social Research (ICPSR) and various summer
-institutes. The white paper could Include a call for a comprehensive assessment
+institutes. The white paper could include a call for a comprehensive assessment
 of these techniques. We need to think carefully about the right venue for such a

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 6: Technical report titles do not need first letter capital.

@@
 \item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
-Sustainability through Developers' Eyes: Recommendations from an Interview
-Study~\cite{wssspe2_rosada_de_souza}
+sustainability through developers' eyes: Recommendations from an interview
+study~\cite{wssspe2_rosada_de_souza}
 
 \item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
-Nicholas Weber. Community Recommendations for Sustainable Scientific
-Software~\cite{wssspe2_downs}
+Nicholas Weber. Community recommendations for sustainable scientific
+software~\cite{wssspe2_downs}
 
 \item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
-Role of Online Platforms, Communications and Workflows in Developing Sustainable
-Software for Science Communities~\cite{wssspe2_patra}
-
-\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
-Pulling It Forward~\cite{wssspe2_pierce}
-
-\item Justin Shi. Seeking the Principles of Sustainable Software
-Engineering~\cite{wssspe2_shi}
+Role of online platforms, communications and workflows in developing sustainable
+software for science communities~\cite{wssspe2_patra}
+
+\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching it up,
+pulling it forward~\cite{wssspe2_pierce}
+
+\item Justin Shi. Seeking the principles of sustainable software
+engineering~\cite{wssspe2_shi}
 
 \item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
-David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
-through Sustainable Software Architectures~\cite{wssspe2_venters}
+David J. Cooke. The Nebuchadnezzar effect: dreaming of sustainable software
+through sustainable software architectures~\cite{wssspe2_venters}
 \end{itemize}

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7: Technical report titles do not need first letter capital.

@@
 \item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
-scientific software: a hackathon experience~\cite{wssspe2_crusoe}
-
-\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable Software
-Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
-for Science~\cite{wssspe2_hanwell}
+scientific software: A hackathon experience~\cite{wssspe2_crusoe}
+
+\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
+ecosystems: Software engineers, domain scientists, and engineers collaborating
+for science~\cite{wssspe2_hanwell}
 
@@
 S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
-Sustainable Science Software from an Early Career Training Institute on Open
-Science Synthesis~\cite{wssspe2_lenhardt}
-
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices Do
-Scientists Find Most Useful?~\cite{wssspe2_schossau}
+sustainable science software from an early career training institute on open
+science synthesis~\cite{wssspe2_lenhardt}
+
+\item Jory Schossau and Greg Wilson. Which sustainable software practices do
+scientists find most useful?~\cite{wssspe2_schossau}
 
@@
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
-Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
-Development in Practice ~\cite{wssspe2_adams}
+Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable software
+development in practice ~\cite{wssspe2_adams}
 
@@
 \item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
-software, building community: lessons from the {ROpenSci}
-project~\cite{wssspe2_boettiger}
-
-\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software is an
-Island: Collaborative Software Development Needs in
-Geosciences~\cite{wssspe2_gil}
+software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
+
+\item Yolanda Gil, Eunyoung Moon, and James Howison. No science software is an
+island: Collaborative software development needs in
+geosciences~\cite{wssspe2_gil}
 
@@
 Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
-Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
-Sustainable Data and Software~\cite{wssspe2_habermann}
+Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A foundation for
+sustainable data and software~\cite{wssspe2_habermann}
 
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
-Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
+Building sustainable software - the {CSDMS} approach~\cite{wssspe2_hutton}
 

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 7.1.4: minor changes.

@@
 of these techniques. We need to think carefully about the right venue for such a
-white paper, where would it have the most impact. The subgroup believes it would
+white paper, where it would have the most impact. The subgroup believes it would
 need to approach editors directly to ascertain this.
@@
 The group felt that training in techniques that promote sustainability has a
-range of benefits: career paths, educated reviewers, reproducible science and
+range of benefits: career paths, educated reviewers, reproducible science, and
 more. There is some information on how that has been approached and assessed,
 but more is needed. The group felt that this training is undervalued and that it
-is important to communicate the return on investment -- both individual ROI
+is important to communicate the return on investment (ROI) -- both individual ROI
 (skills that make scientists more effective and more marketable) and funder ROI
-(better use of taxpayer funds, research more likely to be reproducible because
-sustainable software exists, better trained reviewers).
+(better use of taxpayer funds; research more likely to be reproducible because
+sustainable software exists; better trained reviewers).
 

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
undoing some changes

@@
 \item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
-sustainability through developers' eyes: Recommendations from an interview
-study~\cite{wssspe2_rosada_de_souza}
+Sustainability through Developers' Eyes: Recommendations from an Interview
+Study~\cite{wssspe2_rosada_de_souza}
 
 \item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
-Nicholas Weber. Community recommendations for sustainable scientific
-software~\cite{wssspe2_downs}
+Nicholas Weber. Community Recommendations for Sustainable Scientific
+Software~\cite{wssspe2_downs}
 
 \item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
-Role of online platforms, communications and workflows in developing sustainable
-software for science communities~\cite{wssspe2_patra}
-
-\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching it up,
-pulling it forward~\cite{wssspe2_pierce}
-
-\item Justin Shi. Seeking the principles of sustainable software
-engineering~\cite{wssspe2_shi}
+Role of Online Platforms, Communications and Workflows in Developing Sustainable
+Software for Science Communities~\cite{wssspe2_patra}
+
+\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
+Pulling It Forward~\cite{wssspe2_pierce}
+
+\item Justin Shi. Seeking the Principles of Sustainable Software
+Engineering~\cite{wssspe2_shi}
 
 \item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
-David J. Cooke. The Nebuchadnezzar effect: dreaming of sustainable software
-through sustainable software architectures~\cite{wssspe2_venters}
+David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures~\cite{wssspe2_venters}
 \end{itemize}
@@
 range of backgrounds, and represented multiple software and training
-initiatives, including Software Carpentry (SwC), Data Carpentry (DC), Open Science for
-Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
+initiatives, including Software Carpentry, Data Carpentry, Open Science for
+Synthesis, the Community Surface Dynamics Modeling Systems (CSDMS) group, ROpenSci,
 DataONE, the HDF Group, and others. Some of these software experiences were
@@
 Three main actions were identified by Subgroup A that would be of interest to
-participants and benefit the research community. These focused on the
+participants and would benefit the research community. These focused on the
 desire to amplify the current community efforts in training and software
-engineering by connecting the current training initiatives (e.g., SwC, DC, 
+engineering by connecting the current training initiatives (e.g., Software and Data Carpentry, 
 rOpenSci, OSS) that are of different durations at appropriate stages.

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
undoing more changes

@@
 (skills that make scientists more effective and more marketable) and funder ROI
-(better use of taxpayer funds; research more likely to be reproducible because
-sustainable software exists; better trained reviewers).
+(better use of taxpayer funds, research more likely to be reproducible because
+sustainable software exists, better trained reviewers).
 
@@
 \begin{itemize}
-\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
-scientific software: A hackathon experience~\cite{wssspe2_crusoe}
-
-\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
-ecosystems: Software engineers, domain scientists, and engineers collaborating
-for science~\cite{wssspe2_hanwell}
+\item Michael R. Crusoe and C. Titus Brown. Channeling Community Contributions to
+Scientific Software: A Hackathon Experience~\cite{wssspe2_crusoe}
+
+\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable Software
+Ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
+for Science~\cite{wssspe2_hanwell}
 
@@
 S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
-sustainable science software from an early career training institute on open
-science synthesis~\cite{wssspe2_lenhardt}
-
-\item Jory Schossau and Greg Wilson. Which sustainable software practices do
-scientists find most useful?~\cite{wssspe2_schossau}
+Sustainable Science Software from an Early Career Training Institute on Open
+Science Synthesis~\cite{wssspe2_lenhardt}
+
+\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices do
+Scientists Find Most Useful?~\cite{wssspe2_schossau}
 
@@
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
-Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable software
-development in practice ~\cite{wssspe2_adams}
-
-\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software
-registry~\cite{wssspe2_allen}
+Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
+Development in Practice~\cite{wssspe2_adams}
+
+\item Alice Allen and Judy Schmidt. Looking before Leaping: Creating a Software
+Registry~\cite{wssspe2_allen}
 
 \item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
-software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
-
-\item Yolanda Gil, Eunyoung Moon, and James Howison. No science software is an
-island: Collaborative software development needs in
-geosciences~\cite{wssspe2_gil}
+Software, Building Community: Lessons from the {ROpenSci} Project~\cite{wssspe2_boettiger}
+
+\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software Is an
+Island: Collaborative Software Development Needs in
+Geosciences~\cite{wssspe2_gil}
 
@@
 Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
-Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A foundation for
-sustainable data and software~\cite{wssspe2_habermann}
+Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
+Sustainable Data and Software~\cite{wssspe2_habermann}
 
 \item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
-Building sustainable software - the {CSDMS} approach~\cite{wssspe2_hutton}
+Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
 
@@
 M. C. Foulkes, James J. Shepherd, and Alex J. W. Thom. The {H}ighly {A}ccurate
-{N-DE}terminant ({HANDE}) quantum {Monte Carlo} project: Open-source stochastic
-diagonalisation for quantum chemistry~\cite{wssspe2_spencer}
+{N-DE}terminant ({HANDE}) Quantum {Monte Carlo} Project: Open-source Stochastic
+Diagonalisation for Quantum Chemistry~\cite{wssspe2_spencer}
 \end{itemize}

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Section 8: small changes.

--- 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
@@
 
-This group, with just three papers but a large amount of interest and
+This group, with just three papers but a high level of interest and
 participation from attendees, focused on the institutional, social, and cultural
 mechanisms that encourage the creation and maintenance of shared software, in the
-context of what now exists but focusing on what mechanisms are desired and how
+context of what now exists, what mechanisms are desired, and how
 we might achieve them.
@@
 The first sub-group discussed issues around the current system for credit and
-current incentives, which it called ``hacking the incentive structure.'' The
+incentives, which it called ``hacking the incentive structure.'' The
 group considered four potential points of leverage:
@@
 promotion, and tenure decisions. The group discussed how we could weigh the
-citations within papers better than we now do? How could we identify the five
+citations within papers better than we now do. How could we identify the five
 citations that really matter for a paper, distinguishing them from the related
@@
 more shareable, more sustainable work), and those in industry but interested in
-science (who could be incentivized to contribute to science.) It was pointed out
+science (who could be incentivized to contribute to science). It was pointed out
 that we are not yet clear enough on exactly what we want to incentivize,
@@
 publications system was far from perfect for incentivizing good software work.
-Nonetheless discussion focused on bringing software people into publications.
+Nonetheless, the discussion focused on bringing software people into publications.
 There were two main suggestions. The first is to focus on end users of software
@@
 that his research showed that few projects were making a formal request for
-citation (but that authors weren't necessarily following those suggestions
+citation (but that authors were not necessarily following those suggestions
 anyway)~\cite{howison2015jasist}. He suggested making access to the software
@@
 In the remaining discussion sessions, the group chose to split into three
-subgroups to discuss a version of these topics: {\em citation ecosystems}, {\em
+subgroups to discuss these topics: {\em citation ecosystems}, {\em
 taxonomy of contributorship/guidelines for software for tenure review}, and {\em

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Bigger font size but still smaller than regular to make text more readable.
Sections 8-10: Technical report titles do not need first letter capital.

@@
 \label{tb:software_sustainability_enablers}
-  \begin{scriptsize}
-  \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+  \begin{small}
+  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
    %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
@@
     \end{tabular}
-    \end{scriptsize}
+    \end{small}
 \end{table*}
@@
  \centering
-  \begin{scriptsize}
-  \begin{tabular}{ | p{1.65cm} | p{10.0cm} |}
+  \begin{small}
+  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
 % \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
@@
 \end{tabular}
- \end{scriptsize}
+ \end{small}
 \end{table}
@@
 
-\item Daniel S. Katz and Arfon M. Smith. Implementing Transitive Credit with
+\item Daniel S. Katz and Arfon M. Smith. Implementing transitive credit with
 {JSON-LD}~\cite{wssspe2_katz}
@@
 \item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis,
-Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned
-Data Analysis Software Environment~\cite{wssspe2_blomer}
-
-\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support
-Reproducible Research~\cite{wssspe2_chamberlain}
+Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The need for a versioned
+data analysis software environment~\cite{wssspe2_blomer}
+
+\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to support
+reproducible research~\cite{wssspe2_chamberlain}
 
@@
 
-\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I Implement Your
-Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
+\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I implement your
+algorithm?'': A model for reproducible research software~\cite{wssspe2_crick}
 
 \item Bryan Marker, Don Batory, Field G. Van Zee, and Robert van de Geijn. Making
-Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
-
-\item Stephen Piccolo. Building Portable Analytical Environments to improve
+scientific computing libraries forward compatible~\cite{wssspe2_marker}
+
+\item Stephen Piccolo. Building portable analytical environments to improve
 sustainability of computational-analysis pipelines in the
@@
 \begin{itemize}
-\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an Essential
-Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
-
-\item Marian Petre and Greg Wilson. Code Review For and By
-Scientists~\cite{wssspe2_petre}
+\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an essential
+process for developing and maintaining scientific software~\cite{wssspe2_clune}
+
+\item Marian Petre and Greg Wilson. Code review for and by
+scientists~\cite{wssspe2_petre}
 
 \item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann,
-David Andrs, and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
-Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
+David Andrs, and Jason M. Miller. Continuous integration for concurrent {MOOSE}
+framework and application development on {GitHub}~\cite{wssspe2_slaughter}
 \end{itemize}

### 2015-07-04 WSSSPE2_report/WSSSPE2_review.tex
Remove duplicated citation to WSSSPE1 report.

@@
 development, deployment, and maintenance of reusable
-software~\cite{katzJORS2014}.
+software~\cite{WSSSPE1}.
 

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
undoing changes

@@
 \item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
-sustainability through developers' eyes: Recommendations from an interview
-study~\cite{wssspe2_rosada_de_souza}
+sustainability through Developers' Eyes: Recommendations from an Interview
+Study~\cite{wssspe2_rosada_de_souza}
 
 \item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
-Nicholas Weber. Community recommendations for sustainable scientific
-software~\cite{wssspe2_downs}
+Nicholas Weber. Community Recommendations for Sustainable Scientific
+Software~\cite{wssspe2_downs}
 
 \item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
-Role of online platforms, communications and workflows in developing sustainable
-software for science communities~\cite{wssspe2_patra}
-
-\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching it up,
-pulling it forward~\cite{wssspe2_pierce}
+Role of Online Platforms, Communications and Workflows in Developing Sustainable
+Software for Science Communities~\cite{wssspe2_patra}
+
+\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
+Pulling It Forward~\cite{wssspe2_pierce}
 
@@
 \item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
-David J. Cooke. The Nebuchadnezzar effect: dreaming of sustainable software
-through sustainable software architectures~\cite{wssspe2_venters}
+David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
+through Sustainable Software Architectures~\cite{wssspe2_venters}
 \end{itemize}
@@
 range of backgrounds, and represented multiple software and training
-initiatives, including Software Carpentry (SwC), Data Carpentry (DC), Open Science for
+initiatives, including Software Carpentry, Data Carpentry, Open Science for
 Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
@@
 desire to amplify the current community efforts in training and software
-engineering by connecting the current training initiatives (e.g., SwC, DC, 
+engineering by connecting the current training initiatives (e.g., Software and Data Carpentry,
 rOpenSci, OSS) that are of different durations at appropriate stages.
@@
 \item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
-scientific software: A hackathon experience~\cite{wssspe2_crusoe}
+scientific Software: A Hackathon Experience~\cite{wssspe2_crusoe}
 
 \item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
-ecosystems: Software engineers, domain scientists, and engineers collaborating
-for science~\cite{wssspe2_hanwell}
+ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
+for Science~\cite{wssspe2_hanwell}
 
@@
 \item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
-Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable software
-development in practice ~\cite{wssspe2_adams}
-
-\item Alice Allen and Judy Schmidt. Looking before leaping: Creating a software
-registry~\cite{wssspe2_allen}
+Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
+Development in Practice ~\cite{wssspe2_adams}
+
+\item Alice Allen and Judy Schmidt. Looking before Leaping: Creating a Software
+Registry~\cite{wssspe2_allen}
 
 \item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
-software, building community: lessons from the {ROpenSci} project~\cite{wssspe2_boettiger}
-
-\item Yolanda Gil, Eunyoung Moon, and James Howison. No science software is an
-island: Collaborative software development needs in
-geosciences~\cite{wssspe2_gil}
+Software, Building Community: Lessons from the {ROpenSci} Project~\cite{wssspe2_boettiger}
+
+\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software Is an
+Island: Collaborative Software Development Needs in
+Geosciences~\cite{wssspe2_gil}
 
@@
 
-This group, with just three papers but a high level of interest and
+This group, with just three papers but a large amount of interest and
 participation from attendees, focused on the institutional, social, and cultural
@@
 citations that really matter for a paper, distinguishing them from the related
-works and general background that are also cited. Perhaps we could break these
+works and general background that are also cited? Perhaps we could break these
 out in the reference list, working with publishers to implement this. Or maybe
@@
 more shareable, more sustainable work), and those in industry but interested in
-science (who could be incentivized to contribute to science). It was pointed out
+science (who could be incentivized to contribute to science.) It was pointed out
 that we are not yet clear enough on exactly what we want to incentivize,
@@
 that his research showed that few projects were making a formal request for
-citation (but that authors were not necessarily following those suggestions
+citation (but that authors weren't necessarily following those suggestions
 anyway)~\cite{howison2015jasist}. He suggested making access to the software
@@
 In the remaining discussion sessions, the group chose to split into three
-subgroups to discuss these topics: {\em citation ecosystems}, {\em
+subgroups to discuss a version of these topics: {\em citation ecosystems}, {\em
 taxonomy of contributorship/guidelines for software for tenure review}, and {\em

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
undoing more changes

@@
 \begin{itemize}
-\item James Howison. Retract bit-rotten publications: Aligning incentives for
-sustaining scientific software~\cite{wssspe2_howison}
-
-\item Daniel S. Katz and Arfon M. Smith. Implementing transitive credit with
+\item James Howison. Retract Bit-rotten Publications: Aligning Incentives for
+Sustaining Scientific Software~\cite{wssspe2_howison}
+
+\item Daniel S. Katz and Arfon M. Smith. Implementing Transitive Credit with
 {JSON-LD}~\cite{wssspe2_katz}
 
-\item Ian Kelley. Publish or perish: the credit deficit to making software and
-generating data~\cite{wssspe2_kelley}
+\item Ian Kelley. Publish or Perish: The Credit Deficit to Making Software and
+Generating Data~\cite{wssspe2_kelley}
 \end{itemize}
@@
 \item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis,
-Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The need for a versioned
-data analysis software environment~\cite{wssspe2_blomer}
-
-\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to support
-reproducible research~\cite{wssspe2_chamberlain}
-
-\item Neil Chue Hong. Minimal information for reusable scientific
-software~\cite{wssspe2_chue_hong}
-
-\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I implement your
-algorithm?'': A model for reproducible research software~\cite{wssspe2_crick}
+Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned
+Data Analysis Software Environment~\cite{wssspe2_blomer}
+
+\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support
+Reproducible Research~\cite{wssspe2_chamberlain}
+
+\item Neil Chue Hong. Minimal Information for Reusable Scientific
+Software~\cite{wssspe2_chue_hong}
+
+\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I Implement your
+Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
 
 \item Bryan Marker, Don Batory, Field G. Van Zee, and Robert van de Geijn. Making
-scientific computing libraries forward compatible~\cite{wssspe2_marker}
-
-\item Stephen Piccolo. Building portable analytical environments to improve
-sustainability of computational-analysis pipelines in the
-sciences~\cite{wssspe2_piccolo}
+Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
+
+\item Stephen Piccolo. Building Portable Analytical Environments to Improve
+Sustainability of Computational-Analysis Pipelines in the
+Sciences~\cite{wssspe2_piccolo}
 \end{itemize}
@@
 \begin{itemize}
-\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an essential
-process for developing and maintaining scientific software~\cite{wssspe2_clune}
-
-\item Marian Petre and Greg Wilson. Code review for and by
-scientists~\cite{wssspe2_petre}
+\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an Essential
+Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
+
+\item Marian Petre and Greg Wilson. Code Review for and by
+Scientists~\cite{wssspe2_petre}
 
 \item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann,
-David Andrs, and Jason M. Miller. Continuous integration for concurrent {MOOSE}
-framework and application development on {GitHub}~\cite{wssspe2_slaughter}
+David Andrs, and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
+Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
 \end{itemize}

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Added a question.
Section 8.1.4: minor but hopefully better changes.

@@
 
-This group defined its goal as creating a low-barrier-of-entry method for
+This group defined its goal as creating a low-barrier-of-entry method, to be called a credit file,  for
 recording names and roles (and in a second phase, optionally including weights)
 of contributors (coders and other intellectual contributions) to a software
-package in a machine readable way (to be called a credit file), then encouraging
+package in a machine readable way, then encouraging
 the scientific community to adopt this practice.
@@
 \item This data could be a file that can be associated with a citation to the
-software, either through use of a DOI for the credit file, or by uploading the
+software, either through use of a DOI (digital object identifier) for the credit file, or by uploading the
 credit files as associated with the paper.
-\item This idea could also be applied to data
-\item This data (the credit file) should be part of the metadata that are freely
+\item This idea could also be applied to data.
+\item The credit file should be part of the metadata that are freely
 available to those who have not paid for access to the journal, like citations
 are now in most cases.
-\item It was suggested that there should also be a separate file to track
-software dependencies, but this is not the same file.
+\item It was suggested that there should also be a \emph{separate} file to track
+software dependencies.%, but this is not the same file.
 \end{itemize}
@@
 \item Define what a citation file should look like and what it should be called.
-\item Test adoption, for example, with LBNL.
+\item Test adoption, for example, with LBNL. \choinote{what is LBNL?}
 \item Create credit file for a set of software.

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Clarify LBNL.

@@
 \item Define what a citation file should look like and what it should be called.
-\item Test adoption, for example, with LBNL. \choinote{what is LBNL?}
+\item Test adoption with a substantial scientific organization such as LBNL (Lawrence Berkeley National Laboratory).
 \item Create credit file for a set of software.
@@
 physics, astronomy), publications often have tens or even hundreds of co-authors
-(with some papers in experimental particle physics having over 3000.) In other
-domains, the number of co-authors is typically much smaller, with, in some
+(with some papers in experimental particle physics having over 3000). In other
+domains, the number of co-authors is typically much smaller with, in some
 cases, even a preference for single-author papers. Similarly, the various
@@
 their prime publication target. It is often suggested that this difference is
-due to the rapid developments in information technology, a pace that can not be
+due to the rapid developments in information technology, a pace that cannot be
 upheld by traditional peer-reviewed journals. Whatever the causes, any useful

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Clarify LBNL.
Section 8: various minor changes.

@@
 \item Define what a citation file should look like and what it should be called.
-\item Test adoption with a substantial scientific organization such as LBNL (Lawrence Berkeley National Laboratory).
+\item Test adoption with a substantial scientific organization such as the Lawrence Berkeley National Laboratory (LBNL).
 \item Create credit file for a set of software.
@@
 \item Communicate the results of the above investigations to the WSSSPE
-community, and to decision-making bodies (funders, publishers, universities,
+community and decision-making bodies (funders, publishers, universities,
 and tenure committee representatives).
@@
 example: best contribution (non-founder), broadest diversity of contributions,
-best tutorials or documentation, best leadership transition (award ex leader and
+best tutorials or documentation, best leadership transition (award ex-leader and
 new leader), best generalization (taking something that was limited and making
@@
 \item Inviting ASF and open source people to contribute to scientific code (at
-least on questions of overlapping
+least on overlapping questions)
 \item Template for assessing scientific contributions made through software.
@@
 should find those who we think would be awarded prizes, then ask them who they
-would want to receive a prize from, then ask those organizations to see if they
+would want to receive a prize from, and subsequently contact those organizations to see if they
 are willing to be involved in the process.
 
-One of the group's ideas was to create a funding program for disciplines or
+One of the groups's ideas was to create a funding program for disciplines or
 other organizations to create a prize program. We would provide a framework and
 a set of requirements, for example: awards to individuals; must award in 5 or 6
-areas; must have a jury and/or objective criteria that includes senior/junior
-domain people and technology people; must have the recipients awarded at a
+areas; must have objective criteria and a jury that includes senior/junior
+domain experts and technologists; must have the recipients awarded at a
 relevant event; and must provide citations that explain why the prizes should be
@@
 Possible groups that might give out awards, either under our framework, or more
-generally, are AAAS, Nature Publishing Group, ACM/IEEE, Astro, Ecology Society
+generally, are AAAS, Nature Publishing Group, ACM, IEEE, Astro, Ecology Society
 of America, etc. Perhaps this could be a joint technology/science partnership,
@@
 
-The group discussion if there should be different criteria for ``established''
+The group discussed if there should be different criteria for ``established''
 members of the community versus junior members, and if prizes should be

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Section 9.1: minor changes.

@@
 This group discussed five papers with a wide variety ideas of how to support
-reproducibility and reuse. It focused on identifying concrete that the attendees
+reproducibility and reuse. It focused on identifying concrete practices that the attendees
 could work together on, which would have a positive effect on the community.
@@
 working on different topics: Reproducibility, and Reuse and Sharing. In the
-second discussion session, three groups worked on creating specific pieces of
+second discussion session, three subgroups worked on creating specific pieces of
 guidance.
@@
 reviewers. Another was to provide the ability for reviewers to flag the
-requirement for a `software verification,' similar to ability to flag that a
+requirement for a `software verification,' similar to the ability to flag that a
 paper needs to be seen by a statistician.
@@
 lend them an additional badge (similar to OOPSLA). Another is simply to get more
-people to use your software: for instance by outreach to high school students --
+people to use your software: for instance by outreach to high school students ---
 can your software be used by them? Finally, there is a role for
@@
 and shared. From the perspective of both user and developer, any solutions must
-be 1) easy; 2) cheap; 3) not take much time.
+be 1) easy; 2) cheap; 3) not too time-consuming.
 
@@
 Some communities have had significant initiatives to improve software discovery,
-e.g. the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
+e.g., the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
 \url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of
 journals which had made software more discoverable: ACM Transactions on Software
-offer a reproducibility review; the Journal of Biostatistics has an opt-in to
+offers a reproducibility review; the Journal of Biostatistics has an opt-in to
 provide code and a certification mark if it can be run; the Journal of Open
@@
 An issue around the sustainability of software catalogues is that their
-usefulness often depends on the domain. For instance, in the bio-sciences there
-is more homogeneous data, and more standard shred codes. In areas like ecology,
+usefulness often depends on the domain. For instance, in the bio-sciences, there
+are more homogeneous data and standard shred code. In areas like ecology,
 codes are very specific to the problem, meaning that the level of re-use might
@@
 
-Finally, it was clear from discussion that there were many ways in which
-software could be made available in more reusable ways that just a tarball
+Finally, it was clear from the discussion that there were many ways in which
+software could be made available in more reusable ways than just a tarball
 sitting on a personal website. Using code repositories like

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
Section 9.1.3: minor changes.
Section 9-11: Minor changes.

@@
 	\begin{itemize}
-		\item Open Access
+		\item Open access
 		\item ``Freely'' available
@@
 	\end{itemize}	
-	\item Supporting Data
+	\item Supporting data
 	\begin{itemize}
 		\item Must be publicly available
-		\item Must be in an Open Access repository
+		\item Must be in an open access repository
 		\item must have a DOI
 	\end{itemize}
-	\item Article Processing Charge
+	\item Article processing charge (APC)
 	\begin{itemize}
@@
 \begin{itemize}
-	\item Required Sections
+	\item Required sections
 	\item Keywords
@@
 	\end{itemize}
-	\item Papers can be updated when new releases of software are released
+	\item Papers can be updated when new releases of software are available
 	\begin{itemize}
@@
 
-\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers}
+\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers?}
 
@@
 journal should provide its reviewers to make it easier to review software
-submitted for publication as a ``Software Paper''.
+submitted for publication as a ``Software Paper.''
 
@@
 Journals should provide mechanisms to enable and track communication between
-reviewers and authors. For anonymous peer review journals, maintain anonymity of
-reviewers. If communication is necessary, that may mean that software is not
+reviewers and authors. For anonymous peer review journals, anonymity of
+reviewers should be maintained. If communication is necessary, that may mean that software is not
 that well documented. Journals should also provide a set of simple metrics for
@@
 Journals should support mechanisms to run software. Sometimes this may be very
-hard to accomplish, despite best efforts. Journals could provide an Execution
-environment for any Software submitted for review, perhaps via a Docker
+hard to accomplish, despite best efforts. Journals could provide an execution
+environment for any software submitted for review, perhaps via a Docker
 container or a virtual machine. If not, instructions must be adequate to compile
-and execute the software; and interpret the results (output files, formats,
-etc.), and the full source code must be accessible to the reviewer. Mechanisms
+and execute the software; to interpret the results (output files, formats,
+etc.); and the full source code must be accessible to the reviewer. Mechanisms
 to quantify what has changed compared to a previously published version would
@@
 account. For example, a speedup of an algorithm may not result in a huge code
-difference, but may nonetheless provide a break-through. Journals should require
+difference, but may nonetheless provide a breakthrough. Journals should require
 software submissions to also provide `test materials' -- sample data, parameters
@@
 
-This group began by recognizing that they know what the code testing and code
-review problems are, so the purpose of the group was really to think about
+This group began by recognizing that they knew what the problems of code testing and code
+review are, so the purpose of the group was really to think about
 concrete things that people in the group would be willing to commit to, in order
@@
 choosing appropriate tests for scientific software. This is the difference
-between learning how to use Python \texttt{unittest} and knowing how to test
-that your code correctly implements, for example, a Lattice-Boltzmann model in
+between learning how to use, say, Python \texttt{unittest} and knowing how to test
+that one's code correctly implements, for example, a Lattice-Boltzmann model in
 computational fluid dynamics.
@@
 (TDD) at a Software Developers Best Practices conference): These often require
-self-learning following the tutorial, and it can be hard to find a tutorial that
-matches the programming language and/or the domain of the participant.
+self-learning following a tutorial, but it can be hard to find a tutorial that
+matches the programming language and the domain of the participant.
 
@@
 software from a larger application, a developer wants to ensure the software
-works as expected. These types of tests may be ephemeral, though, living long
+works as expected. These types of tests may be ephemeral though, living long
 enough to give the author of the test confidence in the code, but not handed on
@@
 Attendees joked that they started testing ``when their skepticism/guilt became
-larger than their arrogance''.
+larger than their arrogance.''
 
@@
 Simpler, standard ways of setting up appropriate testing infrastructure will be
-important for adoption by scientists. Jenkins
-\footnote{\url{https://jenkins-ci.org/}} was suggested as a common open-source
+important for adoption by scientists. Jenkins\footnote{\url{https://jenkins-ci.org/}} 
+was suggested as a common open-source
 solution, but the initial configuration was considered to be challenging for
@@
 
-Teach testing be beginning with a (smallish) piece of code that lacks
+Teach testing by beginning with a (smallish) piece of code that lacks
 appropriate tests and develop an exercise that involves refactoring the code
 into a more presentable form through creating/deriving unit tests. Students
-would extract unit-tests through reverse-engineering and/or questioning a
+would extract unit-tests through reverse-engineering or questioning a
 partner who is a developer/expert.
 
-This lead into a more general discussion about whether programming courses
+This leads to a more general discussion about whether programming courses
 should be part of the standard curriculum for science students and the
@@
 The second major discussion thread in the group addressed the issues identified
-by Software Carpentry - the difficulties in writing tests appropriate for
-scientific software. Even among this group, while many What tests are
+by Software Carpentry --- the difficulties in writing tests appropriate for
+scientific software: %Even among this group, while many 
+What tests are
 appropriate to ensure that an complex method is working correctly? When is the
-result of a numerical computation 'close enough' to pass a test?
+result of a numerical computation `close enough' to pass a test?
 
 Several members of the group indicated that while they valued the concept of
-software testing in theory, they were unaware how to test certain kinds of
+software testing in theory, they were unaware of how to test certain kinds of
 software. Or, more importantly, they lacked relevant examples of software
-testing that would be suitable for new members to their team. As an example, the
+testing that would be suitable for members new to their teams. As an example, the
 group asked if it could produce a 100-line example of how to test neutron
@@
 layers of software applications.
-
+%
 %The group referred to Lorena Barba's ``12 Steps to Navier
@@
 %  here, except possibly as an analog for something under discussion.}
-
+%
 %
@@
 %\clunenote{I no longer recall the context for this statement; does not seem important to me.}
-
+%
 There is some discussion of the issue in the computer science
@@
 process of creating code from one that is solely about ``how?'' into one that
-also include ``why?''
+also includes ``why?''
 
@@
 
-\item The size of the simplest piece of relevant, interesting code (i.e. an
+\item The size of the simplest piece of relevant, interesting code (i.e., an
 estimate of lines of code)
 
-\item A point-form list of the test cases you would use
+\item A point-form list of the test cases you would use.
 
@@
 are in using an existing service (EasyChair) to handle submissions and reviews,
-rather than an ad hoc process, and using an existing service (well-sorted) to
+rather than an ad hoc process, and using an existing service (Well Sorted) to
 allow collaborative grouping of papers into themes by all authors, reviewers,
@@
 
-The fact remains that contributors also want to get credit for their
-participation in the process. And the workshop organizers still want to make
+The fact remains that contributors want to get credit for their
+participation in the process. The workshop organizers also want to make
 sure that the workshop content and their efforts are recorded. Ideally, there
-would be a service that would be able to index the contributions to the
-workshop, serving the authors, the organizers, and the larger community. But
-since there still isn't such a service today, the workshop organizers are
+would be a service that could index the contributions to the
+workshop, serving the authors, the organizers, and the larger community. 
+Since there still isn't such a service today, the workshop organizers are
 writing this initial report and making use of arXiv as a partial solution to
@@
 talks and the breakout groups, and the discussion of sustainability also led
-to interesting and useful results, but the challenge that we have discovered
+to interesting and useful results. However, the challenge that we have discovered
 since WSSSPE2 is that it is very hard to continue the breakout groups'
 activities.  The WSSSPE2 participants were willing to dedicate their time to
-the groups while they were at the meeting, but since then, they have gone
+the groups while they were at the meeting, but afterwards, they have gone
 back to their (paid) jobs.  We need to determine how to tie the WSSSPE
 breakout activities to people's jobs, so that they feel that continuing them
-is a higher priority that it is now, perhaps through funding the participants,
+is a higher priority than it is now, perhaps through funding the participants,
 or through funding coordinators for each activity, or perhaps by getting

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
My attempt to strengthen the abstract.

@@
 This technical report records and discusses the Second Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE2). The workshop used an
-alternative submission and peer-review process, which led to a set of papers
-divided across five topic areas: exploring sustainability; software development
-experiences; credit \& incentives; reproducibility \& reuse \& sharing; and code
-testing \& code review. The report includes a description of the submission and
-review process, the workshop keynote presentations, a series of lightning talks,
-a discussion on sustainability, and the five discussions from the topic areas.
-The report also contains a description of potential actions that were proposed
-in the five discussions.
+Software for Science: Practice and Experiences (WSSSPE2). 
+%The workshop used an
+%alternative submission and peer-review process, which led to a set of papers
+%divided across five topic areas: 
+The report includes a description of the alternative, experimental submission
+and review process, two workshop keynote presentations, a series of lightning
+talks, a discussion on sustainability, and five discussions from the topic areas
+of exploring sustainability; software development experiences; credit \&
+incentives; reproducibility \& reuse \& sharing; and code testing \& code
+review. For each topic, the report includes a list of tangible actions that were
+proposed and would lead to potential changes.
+
+First and foremost, the workshop recognized that reliance on scientific software
+is pervasive in all areas of world-leading research today. Then the workshop
+participants proceeded to explore different perspectives on the concept of
+sustainability. Key enablers and barriers of sustainable scientific software
+were identified from the experiences of the workshop participants. In addition,
+recommendations with new requirements such as software credit files and software
+prize frameworks were outlined for improving practices in sustainable software
+engineering.
+
+There was also broad consensus that, within the disciplines represented, formal
+training in software development or engineering was rare among the
+practitioners. Significant strides are needed to be made in building a sense of
+community via training in software and technical practices, on increasing their
+size and scope, and on better integrating them directly into graduate education
+programs.
+
+Journals can define and publish policies to improve reproducibility, whereas
+reviewers can insist that authors provide sufficient information and access to
+data and software to allow them reproduce the results in the paper. Hence a list of
+critieria is compiled for  journals to provide to reviewers so as to make it easier to
+review software submitted for publication as a ``Software Paper.''
+
 \end{abstract}

### 2015-07-05 WSSSPE2_report/WSSSPE2_review.tex
pagebreak after abstract
Some fiddling.

@@
 \maketitle
+\newpage
 
@@
 open licensing and a public code repository. To probe the level of governance
-within open source projects, he and his co-authors (S.~Marru and C.~Mattmann)
+within open source projects, he and his collaborators %(S.~Marru and C.~Mattmann)
 suggested a contest to encourage individual developers to submit
@@
 allow others to keep the code up to date, either adding them as new authors of
-the paper or providing credit in some other form for their academic service.
+the paper or providing credit in some other form. %for their academic service.
 
@@
 engagement and collaborative activities, increasing understanding and awareness,
-and creating incentives to motivate sustainable science software practices. The
-ESIP SSC plans to engage the community in the recommended activities for
+and creating incentives to motivate sustainable science software practices. 
+The ESIP SSC plans to engage the community in the recommended activities for
 improving sustainable scientific software practices.
@@
 others how to achieve reproducibility. Major public funding investments go into
-research that heavily relies on software to be reproducible, hence lack thereof
+research that heavily relies on reproducible software, hence lack thereof
 is raising major concerns.
@@
 
-Discovery of relevant software is still fundamental issue: we need standard
+Discovery of relevant software is still a fundamental issue: we need standard
 vocabularies and metadata, better tools, and approaches that are sustainable.
@@
 are more homogeneous data and standard shred code. In areas like ecology,
-codes are very specific to the problem, meaning that the level of re-use might
-be at a very general statistical level of abstraction, but then every research
+code is often specific to a problem, meaning that the level of re-use might
+be at a general statistical level of abstraction, but then every research
 use is highly customized.
@@
 Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
-
+%
 From these a set of common categories were synthesized, against which all
@@
 set of practical suggestions for improving adoption of code testing.
-
+%
 Established software projects can encourage the latter through a simple rule of

### 2015-07-06 WSSSPE2_report/WSSSPE2_review.tex
bib style is vancouver now (JSOR's)
fix typo
lots of little things in response to previous changes and other fixes

@@
 review. For each topic, the report includes a list of tangible actions that were
-proposed and would lead to potential changes.
+proposed and that would lead to potential change.
 
 First and foremost, the workshop recognized that reliance on scientific software
-is pervasive in all areas of world-leading research today. Then the workshop
-participants proceeded to explore different perspectives on the concept of
+is pervasive in all areas of world-leading research today. The workshop
+participants then proceeded to explore different perspectives on the concept of
 sustainability. Key enablers and barriers of sustainable scientific software
-were identified from the experiences of the workshop participants. In addition,
+were identified from their experiences. In addition,
 recommendations with new requirements such as software credit files and software
@@
 training in software development or engineering was rare among the
-practitioners. Significant strides are needed to be made in building a sense of
+practitioners. Significant strides need to be made in building a sense of
 community via training in software and technical practices, on increasing their
@@
 data and software to allow them reproduce the results in the paper. Hence a list of
-critieria is compiled for  journals to provide to reviewers so as to make it easier to
+criteria is compiled for  journals to provide to reviewers so as to make it easier to
 review software submitted for publication as a ``Software Paper.''
@@
 open licensing and a public code repository. To probe the level of governance
-within open source projects, he and his collaborators %(S.~Marru and C.~Mattmann)
+within open source projects, he and his co-authors (Marru and Mattmann)
 suggested a contest to encourage individual developers to submit
@@
 allow others to keep the code up to date, either adding them as new authors of
-the paper or providing credit in some other form. %for their academic service.
+the paper or providing credit for their academic service in some other form.
 
@@
 \end{itemize}
-\katznote{any discussion of this?}
+%\katznote{any discussion of this?}
 
@@
 }
-\katznote{any discussion of these tables?}
+%\katznote{any discussion of these tables?}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}
-
-\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
+%\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}
+
+%\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
 %https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit
@@
 
-\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}
+%\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan led writing of this section}
-
-\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
+%\todo{Dan led writing of this section}
+
+%\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
 %https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q
@@
 
-\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
+%\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
 %https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
 
-This group defined its goal as creating a low-barrier-of-entry method, to be called a credit file,  for
+This group defined its goal as creating a low-barrier-of-entry method for
 recording names and roles (and in a second phase, optionally including weights)
 of contributors (coders and other intellectual contributions) to a software
-package in a machine readable way, then encouraging
+package in a machine readable way (to be called a credit file), then encouraging
 the scientific community to adopt this practice.
@@
 \subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
-\note{by Frank Seinstra}
-
-\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
+%\note{by Frank Seinstra}
+
+%\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
 %https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
@@
 \subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
-\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
+%\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
 %https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit
@@
 \item Inviting ASF and open source people to contribute to scientific code (at
-least on overlapping questions)
+least in areas where there is overlapping interest).
 \item Template for assessing scientific contributions made through software.
@@
 
-One of the groups's ideas was to create a funding program for disciplines or
+One of the group's ideas was to create a funding program for disciplines or
 other organizations to create a prize program. We would provide a framework and
 a set of requirements, for example: awards to individuals; must award in 5 or 6
-areas; must have objective criteria and a jury that includes senior/junior
-domain experts and technologists; must have the recipients awarded at a
+areas; must have a jury that includes senior/junior
+domain experts and technologists (and should have objective criteria); must have the recipients awarded at a
 relevant event; and must provide citations that explain why the prizes should be
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Neil led writing of this section}
-
-\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
+%\todo{Neil led writing of this section}
+
+%\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
 %https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
@@
 others how to achieve reproducibility. Major public funding investments go into
-research that heavily relies on reproducible software, hence lack thereof
+research that heavily relies on reproducible software, hence the lack thereof
 is raising major concerns.
@@
 lend them an additional badge (similar to OOPSLA). Another is simply to get more
-people to use your software: for instance by outreach to high school students ---
+people to use your software: for instance by outreach to high school students --
 can your software be used by them? Finally, there is a role for
@@
 An issue around the sustainability of software catalogues is that their
-usefulness often depends on the domain. For instance, in the bio-sciences, there
-are more homogeneous data and standard shred code. In areas like ecology,
-code is often specific to a problem, meaning that the level of re-use might
+usefulness often depends on the domain. For instance, in the biosciences, there
+is more homogeneous data and standard shared code. In areas like ecology,
+code is often very specific to a problem, meaning that the level of re-use might
 be at a general statistical level of abstraction, but then every research
@@
 	\end{itemize}
-	\item Papers can be updated when new releases of software are available
+	\item Papers can be updated when new releases of software are made
 	\begin{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}
-
-\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
+%\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}
+
+%\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
 %https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
@@
 self-learning following a tutorial, but it can be hard to find a tutorial that
-matches the programming language and the domain of the participant.
+matches the programming language and/or the domain of the participant.
 
@@
 into a more presentable form through creating/deriving unit tests. Students
-would extract unit-tests through reverse-engineering or questioning a
+would extract unit-tests through reverse-engineering and/or questioning a
 partner who is a developer/expert.
 
-This leads to a more general discussion about whether programming courses
+This led into a more general discussion about whether programming courses
 should be part of the standard curriculum for science students and the
@@
 The second major discussion thread in the group addressed the issues identified
-by Software Carpentry --- the difficulties in writing tests appropriate for
+by Software Carpentry -- the difficulties in writing tests appropriate for
 scientific software: %Even among this group, while many 
@@
 software. Or, more importantly, they lacked relevant examples of software
-testing that would be suitable for members new to their teams. As an example, the
+testing that would be suitable for new members of their teams. As an example, the
 group asked if it could produce a 100-line example of how to test neutron
@@
 
-The fact remains that contributors want to get credit for their
-participation in the process. The workshop organizers also want to make
+The fact remains that contributors also want to get credit for their
+participation in the process. And the workshop organizers will want to make
 sure that the workshop content and their efforts are recorded. Ideally, there
-would be a service that could index the contributions to the
+would be a service that would index the contributions to the
 workshop, serving the authors, the organizers, and the larger community. 
@@
 provide a record of the workshop.
+
+\begin{table*}[t]
+\centering
+\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
+  \begin{scriptsize}
+  \begin{tabular}{l|l|r|r}
+ \hline
+    Author  &   Tweet  & Retweets &  Favorites
+\\ \hline
+% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
+%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    &
+%
+Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 4
+\\     &  management plan used by EPSRC software grants    &    &
+\\     &  \url{http://www.software.ac.uk/resources/guides/software-management-plans}  &    &
+%
+\\  Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
+\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    &
+\\   &   we provide a software evaluation tool:   &    &
+\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    &
+%
+\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
+\\ & UK researchers write research software w$/$out any formal training &    &
+%
+\\ Karthik Ram  &  OH: ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+%
+\\ Josh Greenberg  &  @jameshowison: ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
+\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    &
+%
+\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
+\\ &   [include] how many faces are up on our community page''   &    &
+%
+\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
+\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    &
+%
+\\ Kaitlin Thaney  & Great point by @tracykteal about planning for ``end of life'' with scientific & 4 & 8
+\\ &  software projects and sustainability.  & &
+%
+%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
+%
+\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
+\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & &
+%
+\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
+\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & &
+\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{http://www.slideshare.net/kaythaney/}     & &
+\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{designing-for-truth-scale-and-sustainability-wssspe2-keynote} & &
+%
+\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
+\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & &
+%
+%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 &
+%
+\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
+\\ & but want SWORD/DSpace/Fedora/ePrints see:  & &
+\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & &
+%
+\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
+\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   &
+%
+\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3
+\\ & but because influential people are averse to change...''  C. Sunstein     & &
+%
+\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
+\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & &
+%
+\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
+\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
+%
+\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
+\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & &
+%
+\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
+\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
+\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
+\\ \hline
+    \end{tabular}
+    \end{scriptsize}
+\end{table*}
 
@@
 
-\begin{table*}[t]
-\centering
-\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
-  \begin{scriptsize}
-  \begin{tabular}{l|l|r|r}
- \hline
-    Author  &   Tweet  & Retweets &  Favorites
-\\ \hline
-% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
-%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    &
-%
-Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 4
-\\     &  management plan used by EPSRC software grants    &    &
-\\     &  \url{http://www.software.ac.uk/resources/guides/software-management-plans}  &    &
-%
-\\  Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
-\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    &
-\\   &   we provide a software evaluation tool:   &    &
-\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    &
-%
-\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
-\\ & UK researchers write research software w$/$out any formal training &    &
-%
-\\ Karthik Ram  &  OH: ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
-%
-\\ Josh Greenberg  &  @jameshowison: ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
-\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    &
-%
-\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
-\\ &   [include] how many faces are up on our community page''   &    &
-%
-\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
-\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    &
-%
-\\ Kaitlin Thaney  & Great point by @tracykteal about planning for ``end of life'' with scientific & 4 & 8
-\\ &  software projects and sustainability.  & &
-%
-%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
-%
-\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
-\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & &
-%
-\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
-\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & &
-\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{http://www.slideshare.net/kaythaney/}     & &
-\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{designing-for-truth-scale-and-sustainability-wssspe2-keynote} & &
-%
-\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
-\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & &
-%
-%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 &
-%
-\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
-\\ & but want SWORD/DSpace/Fedora/ePrints see:  & &
-\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & &
-%
-\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
-\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   &
-%
-\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3
-\\ & but because influential people are averse to change...''  C. Sunstein     & &
-%
-\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
-\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & &
-%
-\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
-\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
-%
-\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
-\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & &
-%
-\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
-\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
-\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
-\\ \hline
-    \end{tabular}
-    \end{scriptsize}
-\end{table*}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\todo{feel free to add stuff here}
+%\todo{feel free to add stuff here}
 
@@
 
-\bibliographystyle{plain}
+\bibliographystyle{vancouver}
 

### 2015-07-07 WSSSPE2_report/WSSSPE2_review.tex
adding comma for authors, shortening abstract and making 1 paragraph, for arXiv

@@
 \author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
-\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$
+\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
 \\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
@@
 proposed and that would lead to potential change.
-
-First and foremost, the workshop recognized that reliance on scientific software
+%
+The workshop recognized that reliance on scientific software
 is pervasive in all areas of world-leading research today. The workshop
@@
 engineering.
-
-There was also broad consensus that, within the disciplines represented, formal
+%
+There was also broad consensus that formal
 training in software development or engineering was rare among the
@@
 programs.
-
-Journals can define and publish policies to improve reproducibility, whereas
+%
+Finally, journals can define and publish policies to improve reproducibility, whereas
 reviewers can insist that authors provide sufficient information and access to

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
starting WSSSPE3 report

--- 2015-07-07 WSSSPE2_report/WSSSPE2_review.tex
@@
 \documentclass[11pt, oneside]{amsart}
+\pdfoutput=1
 
@@
 
-\title[]{Report on the Second Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE2)}
-
-\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
-\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
-\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
-%
-\thanks{{}$^{(1)}$ Computation Institute, 
-University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
-%
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
-%
-\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
-%
-\thanks{{}$^{(4)}$ Software Sustainability Institute, 
-University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
-%
-\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
-%
-\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
-%
-\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
-%
-\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
-%
-\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
-%
-\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
-%
-\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
-Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
-%
-\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
-%
+\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)}
+
+\author{tbd by writing and organizing}
+
+%\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
+%\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
+%\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
+%%
+%\thanks{{}$^{(1)}$ Computation Institute, 
+%University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
+%%
+%\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
+%%
+%\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
+%%
+%\thanks{{}$^{(4)}$ Software Sustainability Institute, 
+%University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
+%%
+%\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
+%%
+%\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
+%%
+%\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
+%%
+%\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
+%%
+%\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
+%%
+%\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
+%%
+%\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
+%Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
+%%
+%\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
+%%
 
 \begin{abstract}
+\todo{need to update for WSSSPE3}
 This technical report records and discusses the Second Workshop on Sustainable
@@
 \maketitle
-\newpage
+%\newpage
 
@@
 
+\todo{need to update for WSSSPE3}
 The Second Workshop on Sustainable Software for Science: Practice and
@@
 
+\todo{need to update for WSSSPE3}
 WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynotes} \label{sec:keynotes}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{comment}
-\note{Lead: Choi. \href{http://tinyurl.com/q6ew45v}{Google Notes}.
-%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-\href{http://tinyurl.com/mnenzms}{Abstracts and slides}}
-\end{comment}
-
+\section{Keynote} \label{sec:keynote}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{need to update for WSSSPE3}
 The workshop featured two keynote addresses. In the opening keynote
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Kaitlin Thaney, Designing for Truth, Scale, and Sustainability~\cite{Thaney_slides}}
-\label{keynote1}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
 Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla), which
@@
 most used; the most invested software were the least utilized.
-
-\begin{comment}
-(Kaitlin didn't really say why? Were the tools used? Did they immediately
-bit-rot? Was their development unsustained?)-should ask question about this. Is
-there a caBIG report available? Quotes from Andrea Califano/Joe Gray, looks like
-there is: \url{http://tinyurl.com/qdodbo7}.
-%http://www.informationweek.com/architecture/report-blasts-problem-plagued-cancer-research-grid/d/d-id/1097068
-
-What do we mean by open research? In regard to community/technology/practices
-
-Inefficiency cartoon:  \url{http://www.xkcd.com/1445/}.
-\end{comment}
 
@@
 
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Neil Chue Hong, We are the 92$\%$~\cite{Hong_slides}}
-\label{keynote2}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\begin{comment}
-In a recent survey of UK research-intensive universities, 92\% of researchers
-said they used research software and 68\% said their research would be impossible
-without software. Yet 71\% have had no formal software training, and few are
-ready to apply many of the things we take for granted such as testing or
-virtualization. WSSSPE represents the pinnacle of what we understand to be the
-best practice around scientific software in our community. My talk will
-challenge the workshop participants to come up with ways of taking this best
-practice to those 92\% of researchers in a way that will lead to maximum benefit
-to the scientific community.
- \end{comment}
-
-%\url{http://dx.doi.org/10.6084/m9.figshare.1243288}
-
-Neil Chue Hong is the Director of the Software Sustainability Institute (SSI) in the
-United Kingdom (UK). The SSI was founded to support the UK's research software
-community by cultivating better, more sustainable research software to enable
-world-class research. Chue Hong's keynote started by making the point
-that the use of -- and reliance on -- software is pervasive in all areas of
-world-leading research, showing examples from disciplines as diverse as
-humanities and high-energy physics, quoting Kerstin Kleese Van Dam of the
-Pacific Northwestern National Laboratory via a petition
-campaign\footnote{\url{http://tinyurl.com/nkn5tnv}}
-%https://www.change.org/p/everyone-in-the-research-community-we-must-accept-that-software-is-fundamental-to-research-or-we-will-lose-our-ability-to-make-groundbreaking-discoveries}}
-at change.org, ``\emph{Today there are very few science areas left which do not
-rely on IT and thus software for the majority of their research work. More
-importantly key scientific advances in experimental and observational science
-would have been impossible without better software.}'' He also cited Daniel
-Katz, Software Infrastructure for Sustained Innovation (SI2) Program Director of
-the NSF, ``\emph{Scientific discovery and innovation are advancing along
-fundamentally new pathways opened by development of increasingly sophisticated
-software. Software is an integral enabler of computation, experiment and theory,
-and directly responsible for increased scientific productivity and enhancement
-of researchers' capabilities.}''
-
-Chue Hong drew attention to the issue that in the cyber-infrastructure and
-high-performance community, hundreds of thousands
-of researchers developing software are all too often disregarded or considered the long tail.
-Actually, the numbers point to the fact that they are the mainstream. %and we
-%(referring to people including the audience of the talk) are the elite minority.
-He emphasized that software is no longer special; it is both essential to and
-common in scientific research. A 2014
-survey\footnote{\url{http://tinyurl.com/ooajs7m}}
-%http://www.software.ac.uk/blog/2014-12-04-its-impossible-conduct-research-without-software-say-7-out-10-uk-researchers}}
-conducted by the SSI polled researchers from 15 research-intensive UK
-universities (406 respondents covering a representative range of funders,
-disciplines, and seniority). The survey reported that 92\% respondents confirmed the use of research
-software and 89\% affirmed that it would be impossible or difficult to conduct
-research without software. Nevertheless the British research community is just
-starting to understand the magnitude of the issue. Whilst many researchers make
-use of software such as MATLAB, SPSS, and Excel, data from the aforementioned
-SSI survey shows that over half (56\%) of respondents developed their own
-research software (which equates to over 140,000 researchers if extrapolated
-across the UK) and yet 71\% of all UK researchers had no formal software-development training, having
-to rely on their own coding skills.
-
-Examining another aspect of the size of the research software community,
-Chue Hong noted that the costs of software-reliant research in the UK
-included \textsterling 840 million of investment in the financial year
-2013--2014, and this amount has risen by 3\% on average over the past four
-years. About 30\% of total research investment has been spent on research that
-relies on software over the last four financial years. These numbers stemmed
-from an analysis by the SSI of data from $49,650$ grant titles and abstracts published on
-Gateway to Research between years 2010 and 2014. A similar analysis of university
-jobs advertised in the same period discovered that despite this investment, only
-4\% of positions were software development related, and of these only 17\% were
-explicitly named as a software developer or software engineer positions: the vast
-majority being advertised as research associate or research assistant positions.
-This in turn leads to the issue of career paths for those bridging the research
-and software worlds, who are essential to support the use and further
-development of research software, a point highlighted by a graphic showing UK
-STEM graduate career paths\footnote{Source: The Scientific Century, Royal
-Society, 2010 (revised to reflect first stage clarification from ``What Do PhD's
-Do?'' study)} showing that only 3.5\% were able to secure permanent positions.
-
-To conclude, Chue Hong led the audience in discussing the following questions:
-What are \emph{we} going to \emph{do} to help and benefit the 92\% of researchers
-who rely on software? Who do we need to persuade? What are the incentives we
-need to put in place? Finally, he challenged the workshop participants to change
-the current deficient practices in research and academia.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{need to update for WSSSPE3}
 \begin{comment}
@@
 
-\smallskip
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Defining Sustainability} \label{sec:defining}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\todo{writing led by Dan - other contributions welcome}
-
-In the first interactive session, the attendees divided themselves into groups
-to discuss software sustainability. They were asked to
-\begin{enumerate}
-\item discuss what the term ``software sustainability'' meant to them
-
-\item determine three things they considered to be significant enablers of
-software sustainability
-
-\item determine three things they considered to be significant barriers to
-software sustainability.
-\end{enumerate}
-Once each group had come up with answers, all the answers were compiled, and the
-attendees voted on which they thought were important by a show of hands.
-
-The general responses to what software sustainability meant were:
-\begin{itemize}
-\item keeping software scientifically useful
-\item separating techniques in code from knowledge in code
-\item that an adequately large community finds value in software and is willing
-to sustain it.
-\end{itemize}
-%\katznote{any discussion of this?}
-
-The enablers of and barriers to software sustainability, roughly ranked by
-attendee votes are shown in
-{\tablename}s~\ref{tb:software_sustainability_enablers} and
-\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
-items were suggested as barriers, but were not voted on due to lack of time in
-the session:
-layering up dependencies;
-using software past its sustainable life;
-using software past its usable life;
-inertia for accepted answers versus wrong or right answers;
-monolithic or poor code; and
-need to restructure code when hardware/software/libraries change.
-}
-%\katznote{any discussion of these tables?}
-
-\begin{table*}[ht]
-\centering
-\caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.}
-\label{tb:software_sustainability_enablers}
-  \begin{small}
-  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
-   %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
-    \hline
-  Importance & Item \\ \hline \hline
-********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
-********** & designing for growth and extension -- open development \\ \hline
-******* & culture in community for reuse \\ \hline
-**** & portability \\ \hline
-**** & culture in development community to support transition between developers \\ \hline
-*** & interdisciplinary people: science + IT experience \\ \hline
-** & planning for end of life \\ \hline
-** & make smart choices about dependencies \\ \hline
-* & thinking of software as product lines -- long term vs.~short term view \\ \hline
- & not all communities need new software \\ \hline
- & converting use into resources \\ \hline
-    \end{tabular}
-    \end{small}
-\end{table*}
-
-\begin{table}[ht]
-\caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.}
-\label{tb:software_sustainability_barriers}
- \centering
-  \begin{small}
-  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
-% \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
-   \hline
-  Importance & Item \\ \hline \hline
-******* & lack of incentives, including promotion and tenure process; promotion
-and tenure process in academic is incompatible with sustainability \\ \hline
-*****  & absent or poor documentation \\ \hline
-***** & funding to ensure sustainability is difficult to obtain \\ \hline
-**** & developers are not computer scientists; don't have software engineering
-practices (in particular, those needed to scale-up projects to support and be
-developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test''
-in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number
-of key people a project would need to lose to become non-viable (the larger the
-number, the healthier the project).} \\ \hline
-** & rate of change of underlying technologies \\ \hline
-** & lack of business models for sustainability \\ \hline
-* & lack of training for how to build sustainability into the system \\ \hline
- & maintenance for software is not visible, appears to ``just happen'' \\ \hline
- & licensing issues \\ \hline
- & staff turnover -- lack of continuity \\ \hline
-\end{tabular}
- \end{small}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Exploring Sustainability} \label{sec:exploring}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Colin led writing of this section.}
-
-%\note{\href{http://tinyurl.com/mpbhvyb}{Google doc notes}}
-%https://docs.google.com/a/uchicago.edu/document/d/10XkshP3YjXnA5JQgP-UFHbOxshWczzehWGSo6V7bNH4/edit
-
-Six papers were categorized under the theme of Exploring Sustainability. The
-group included four of the authors from the six papers submitted and an
-additional number of participants who expressed an interest in the theme at the
-workshop. Each paper had a different perspective on the concept of
-sustainability, which ranged from the sustainment of communities to defining
-sustainability as a first-class, composite non-functional requirement.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsubsection{Discussion}
-
-Each author was invited to outline the key action from their paper as a
-potential discussion point for the group; where the authors were not present,
-the group facilitators outlined the actionable points from their papers. The key
-actions from the six papers included recommendations for improving practice in
-sustainable software engineering~\cite{wssspe2_rosada_de_souza, wssspe2_downs};
-development of \emph{Software and Infrastructure as a Service} as a mechanism for
-fostering sustainable science communities~\cite{wssspe2_patra}; developer
-incentives for code contributions to open source projects~\cite{wssspe2_pierce};
-establishment of a set of software engineering principles based on scalability,
-reproducibility, and energy efficiency~\cite{wssspe2_shi}; and applying software
-architectures as a mechanism for architectural-level reasoning about
-sustainability~\cite{wssspe2_venters}. The group took the position of viewing
-sustainability from the perspective of addressing the challenges related to the
-development, deployment, and maintenance of reusable
-software~\cite{WSSSPE1}.
-
-The principal focus of the initial discussion
-considered how to foster cultural change towards developing sustainable software
-in academic environments. It was suggested that a new requirement driven by the
-agencies funding research projects where software was an intrinsic part of
-enabling the research program would be an additional element within the grant
-proposal to provide a sustainability plan for sustaining the software. This type
-of initiative would provide the necessary incentive and motivation for
-researchers to consider how to sustain their software beyond the lifetime of the
-project.
-
-The discussion then focused on a need for a common language and a definition on
-the concept of sustainability that moved beyond the current fuzzy definitions
-where time was the simple measurement of sustainment. It was suggested that
-there was a need to identify tangible actions that underpin sustainability that
-developers could incorporate into the development stream of their software. This
-prompted a debate regarding whether sustainability should be considered as a
-non-functional requirement or software quality as defined within ISO/IEC
-25010~\cite{iso25010}. The focus of the discussion was based on the paper by
-Venters et al.~\cite{wssspe2_venters}, which suggested that sustainability is a
-first-class, composite, non-functional requirement composed of a number of
-sub-characteristics. It was generally agreed that maintainability and
-extensibility were key qualities underlying sustainability. In addition, the
-group also discussed what other non-functional requirements would contribute to
-the development of sustainable software, e.g., reusability and scalability.
-However, it was recognized that there was a need to identify appropriate metrics
-and measures.
-
-The group also discussed whether the concept of sustainability itself was a
-barrier to achieving sustainable software. It was suggested that ``the first
-rule of software sustainability is do not talk about software
-sustainability.'' Instead there should be a focus on best software engineering
-practice. Playing devil's advocate, it was asked that if the focus on the
-concept of sustainability was ignored, what current software engineering
-practices and principles could be utilized by software developers and domain
-scientists to achieve sustainability? This raised the question of why existing
-software engineering knowledge, such as that contained in
-SWEBOK~\cite{Bourque2014}, was largely ignored and to what extent the
-environment has a strong influence on practice within the scientific and
-engineering community. As a result, there is a need to identify best practices
-and reach a consensus beyond the theories that underpin the discipline of
-software engineering. Similarly, how could we translate or distill some of the
-key building blocks that underpin software engineering practice?
-
-A final key point of the discussion was the role that software design and
-patterns play before committing to writing a line of code. It was suggested that
-modeling must play a major role in attaining sustainable software. The point was
-raised that design involves making decisions and a mechanism would be necessary
-for capturing these decision points. This introduced the idea of software
-provenance that moved beyond commits in software repositories to how to capture
-and maintain relationships between sources and design decisions to prevent
-knowledge vaporization. Whether this could be achieved through software
-architectures is an open-research challenge.
-
-\subsubsection{Actions}
-
-The main action to come from the group was a proposal to identify the ten best
-software engineering practices similar to Philip Bourne's approach of Ten Simple
-Rules~\cite{bourne-ten-simple-rules}.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Papers}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Exploring Sustainability group are:
-\begin{itemize}
-\item Mario Rosado de Souza, Robert Haines, and Caroline Jay. Defining
-sustainability through Developers' Eyes: Recommendations from an Interview
-Study~\cite{wssspe2_rosada_de_souza}
-
-\item Robert Downs, W. Christopher Lenhardt, Erin Robinson, Ethan Davis, and
-Nicholas Weber. Community Recommendations for Sustainable Scientific
-Software~\cite{wssspe2_downs}
-
-\item Abani Patra, Matthew Jones, Steven Gallo, Kyle Marcus, and Tevfik Kosar.
-Role of Online Platforms, Communications and Workflows in Developing Sustainable
-Software for Science Communities~\cite{wssspe2_patra}
-
-\item Marlon Pierce, Suresh Marru, and Chris Mattmann. {WSSSPE2}: Patching It Up,
-Pulling It Forward~\cite{wssspe2_pierce}
-
-\item Justin Shi. Seeking the principles of sustainable software
-engineering~\cite{wssspe2_shi}
-
-\item Colin C. Venters, Michael K. Griffiths, Violeta Holmes, Rupert R. Ward, and
-David J. Cooke. The Nebuchadnezzar Effect: Dreaming of Sustainable Software
-through Sustainable Software Architectures~\cite{wssspe2_venters}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Software Development Experiences} \label{sec:devel}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Nancy led writing of this section, with Subgroup A part from Matt Jones}
-
-%\note{\href{http://tinyurl.com/pn4eq8z}{Google doc notes}}
-%https://docs.google.com/a/uchicago.edu/document/d/1_q0jmiEPNFRGAEtGLk_QxQ1vVokgTDYruTGURowNGkY/edit
-
-Of the short actionable papers that would lead to improvements for sustainable
-software science, 11 submissions were categorized in the Software Development
-Experiences group. Because of the large number, we split these into two
-subgroups prior to the event. Some common themes helped in this division. For
-example, several papers that addressed education and training issues, including
-best practices and case studies, were grouped together. Others discussed
-experiences with registries, developer collectives and specific examples of
-successful, sustainable software (including a valuable industry perspective).
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-Subgroup A consisted of ten participants who discussed papers surrounding training
-and successful community software initiatives.
-Subgroup B had six participants who discussed four papers. Because of the nature
-of the papers, training emerged as a common theme. However the conversation was
-wide-ranging, including incentives, reproducibility, and funding to promote
-sustainability.  In the end, both groups discussed training, though from somewhat different
-points of view and resulting in somewhat different suggested actions.
-
-\subsubsection{Subgroup A Discussion}
-
-%\note{\href{https://docs.google.com/document/d/1n413NcFPDyDGb27tEp-ujryiRhz-ZaAhmUapWRYmh1g/edit}{Google doc notes}}
-
-Group discussion started with a position statement by each person surrounding
-what they had learned from their software development experiences and how those
-lessons might be translated into actionable outcomes. Participants came from a
-range of backgrounds, and represented multiple software and training
-initiatives, including Software Carpentry, Data Carpentry, Open Science for
-Synthesis (OSS), the Community Surface Dynamics Modeling Systems group, ROpenSci,
-DataONE, the HDF Group, and others. Some of these software experiences were
-focused on development of new products for use in the sciences (e.g., ROpenSci,
-CSDMS), and these recounted the difficulties of engaging with disciplinary
-scientists in writing software. Software was clearly utilized broadly across the
-various science disciplines represented, and it was developed within those
-disciplines as well. Some researchers created software for statistical analysis
-and modeling, while others used it to control instrumentation, collate data
-across networks, collect information from respondents, and many other uses.
-There was broad consensus that, within the disciplines represented, formal
-training in any type of software development or engineering was rare among the
-practitioners; most are self-taught, and develop software to get another job
-done. Any ancillary utility of the software outside of the specific science
-target was generally unplanned and few researchers would want to invest more
-time to make their own software more re-usable.
-
-%Despite agreement that this body of disciplinary software needed improvement,
-There was general agreement that this body of disciplinary software improvement
-needed to be understood in terms of the scientific productivity that could be
-achieved. A software maturity model is needed for science software, but it needs
-to be introduced in a way that fits the culture of science, which largely thinks
-of software as a tool, rather than a product itself. The group was in general
-consensus that more widespread training in software practice is needed within
-the domain sciences, and several of the participants were involved in efforts
-along these lines. Participants felt that projects that build a sense of
-community via training in software and technical practices would have the most
-success in changing practices in that community, but that there was a need for a
-managed introduction of these practices. Participants also recognized that these
-could not be one-time, one-off training opportunities, as software and
-technologies change rapidly over time. For example, while today Software
-Carpentry is focused on teaching version control via Git, there has been a rapid
-evolution from RCS to CVS to SVN to Git over a short time frame, and thus
-communities should expect the need to train for adaptability and a changing
-technology tool chain, rather than assume that these technologies will stay
-fixed. Thus, although short term training that introduces specific tools was
-considered highly valuable, these trainings were also not considered sufficient
-to engender the changes in software practice that were deemed necessary.
-Combining the need for changes in practice to be introduced incrementally with
-the need to minimize the divergence of training from direct science goals and
-the difficulties of training for a rapidly evolving technology space,
-participants concluded that multiple training efforts that targeted different
-parts of the spectrum were needed. Short-term courses introducing immediately
-useful skills needs to be offered alongside more in-depth courses on software
-engineering and practice that allow students to adapt to a changing landscape.
-
-Finally, after agreeing that these complementary training models were needed,
-the group discussed sustainability of training, and how the leading groups in
-this space are teaching only a small fraction of the community that needs and
-wants training. Most graduate programs in the sciences do not currently
-incorporate these approaches in their graduate curricula, although there is an
-increasing number of quantitatively oriented courses around analysis and
-modeling. These still, however, generally omit engineering practices such as
-version control, unit and regression testing, and software modularity and
-abstraction, often because the instructors themselves in the domain sciences are
-not familiar with these techniques. Thus, students who are being trained in
-these approaches are doing so through short 2--3 day training workshops such as
-Software and Data Carpentry, rather than through semester-long graduate
-education courses at their universities, which tend to focus on statistical and
-modeling techniques. Hybrid programs like the three-week Open Science for
-Synthesis (OSS) training that combine the three (science, quantitative
-techniques, and software engineering) into an holistic course serve part of the
-need but reach only a few researchers at this time. Thus, the group concluded
-that significant strides needed to be made in coordinating these trainings so
-that they are complementary, on increasing their size and scope, and on better
-integrating them directly into graduate education programs.
-
-\subsubsection{Subgroup B Discussion}
-The statement ``applied computer science is being attempted in academia without
-any formal training'' kicked off our discussion. The group brought expertise in
-several different training models, from a two-day Software Carpentry workshop,
-a three-week Open Science for Synthesis (OSS) program, to semester-long
-programs. We discussed the pros and cons of different training approaches,
-touching on informal learning, for example, where people learn the necessary
-skills by asking questions of cross-disciplinary people (``boundary scientists'') in
-their work environments.
-
-Some papers explored gaps in training of early-career scientists. Industry
-participants in our group confirmed this observation. We asked ourselves, ``Are traditional
-courses failing?'' We think yes. Changes to undergraduate curricula requirements
-are difficult. But as programming models become more complex, we have to raise
-the skill levels. Skills must be improved not only in traditional programming --
-learning languages and algorithms -- but also around professional software
-development. We want to get to a point where ``Everyone has a new minimum now --
-everyone knows Git.'' Developers also need training on licensing choices. Just
-because some source code is on Git does not mean that it is open.
-
-``How do you get people to look for the training they need?'' wondered one
-participant. People seek out opportunities like Software Carpentry to augment
-skills. While some instructions can be done in institutional curricula,
-independent groups (non-profits, institutes) have more flexibility. Some asked
-whether Silicon Valley would be interested in funding training so that people are
-better prepared to enter the workforce. Some felt that companies were reluctant
-to deliver training for fear that their employees might then leave them. While others felt
-that pushing all training to industry could lead to good technical people
-``getting on the Google bus'' and leaving sciences.
-
-``How will we know when people are trained effectively in these new skills?'' We
-discussed certification. It can be hard to build the recognition of Java or
-database certifications among all technologies. OSS offers badges to those
-completing training. However we need to demonstrate proficiency, not just
-completion. Google Summer of Code is a big CV augmentation for
-participants---can we create something similar?
-
-The group also discussed how to fund training. On participant observed that NYU
-runs a six-week data-science training; companies grab the graduates and pay for
-those they hire. OSS also used the NSF Software Institutes program as a vehicle
-to fund training activities. Software Carpentry uses a collaborative teaching
-approach where people publish open teaching materials and receive credits for
-their use.
-
-We then discussed career paths for those supporting sustainable software. While
-tenure track is not the only option for graduate students, the challenges for those who
-remain in academia can be large. Research scientists are entirely dependent on
-soft money, which can be unpredictable. Postdocs and those who do pursue tenure
-track positions need to publish and see no rewards in software development.
-These challenges were all identified at WSSSPE1. What actions would we recommend
-to improve things? Altmetrics and download statistics may slowly change the
-system and improve a developer's ability to receive credits for time invested in
-software development. NSF's recognition of scientific products including datasets,
-software, and publications is also helping.
-
-We asked ourselves, ``Are there examples of things that are changing
-because of this and how can we build momentum?'' One example demonstrates change
-over time. In 2007, nanoHUB listed the academic reward structure as a problem in
-an EDUCAUSE report where the authors note, ``In the future, nanoHUB researchers
-are hoping to change the research culture. While they recognize that young
-faculty members are unlikely to get tenure based on their nanoHUB contributions,
-they hope to encourage faculty to think beyond their own research needs to
-consider publishing tutorials and other content in their fields on the
-nanoHUB site.'' Fast forward seven years later where, in a 2014 iSGTW article,
-quote nanoHUB Principal Investigator (PI) Gerhard Klimeck, ``A former student of
-mine published eight tools on nanoHUB, serving over 6,000 people with his tools.
-He then joined a university as a professor and introduced nanoHUB. Use of the
-gateway from that university skyrocketed; he used nanoHUB in existing classes,
-created new classes, and infused it in his research. Ultimately, the professor's
-department head attributed his two-year rise to tenure with the reputation and
-innovation he gained through nanoHUB.''
-
-The group also discussed ``attribution trees,'' an idea put forward by Dan Katz
-and Arfon Smith where a chain of attributions can, for example, give credit to
-those developing libraries and building blocks that other pieces of software
-use. The group considered potential journals and medium to push the application
-of this idea. One participant noted that Dryad\footnote{The Dryad Digital
-Repository, \url{http://datadryad.org}.} works with journals. If a paper is
-accepted to one of those journals, the supporting data must be submitted to Dryad.
-The group also discussed reproducibility as
-a component of the journal review process and ``active papers,'' with immediate
-links to the data and software.
-
-We then discussed variations among scientific domain areas and wondered ``Are
-some communities more or less open than others?'' To some members of the group,
-biology seems to be more open, while physics less so. Some felt, with the more recent
-development of bioinformatics as a field, there were fewer historical practices
-to undo in biology. Physics has preprint philosophy to overcome. Environmental sciences
-may be mixed. The group felt that the biomedical area, however, was very competitive
-and closed. One participant mentioned blueprints for going open source (like
-NWChem recently did) where authors outline how this helps, what you do and what
-the next steps are.
-
-We then moved beyond our training discussion to address funding that encourages
-sustainable software -- funding of both people and projects that create a true
-system to support sustainability. We called this ``institutionalized
-serendipity.'' As science is increasingly reliant on software, one participant
-observed that ``software development can have much broader impact than
-publishing individual research, but it is not viewed that way.'' Because of this
-centrality, one participant mentioned that training ought to be called Science
-Carpentry rather than Software Carpentry. We felt we were beginning to see
-changes in the research community as a result of the NSF's data management plan
-requirements. PIs are thinking more about data and some university libraries are
-offering data repositories. We wondered if a software management plan might be
-effective. ``How might funding programs need to adapt to reward good software
-development practices?'' we asked ourselves. We thought about best practices
-such as version control, test harnesses, mailing lists, bug tracking, community
-contribution, and reuse where appropriate. We thought about measuring success
-through usage statistics (downloads, altmetrics). ``Should funders demand
-reproducibility?'' we wondered. In order for results to be reproduced, software
-would need to be carefully curated.
-
-Again, our industry participants contributed unique viewpoints. Partnering with
-industry was seen as one path to sustainability. The unique partnerships Kitware
-engages in promote academic freedom while creating an income stream from certain
-portions of the software. This type of approach to sustainable software frees
-researchers from performing tasks that do not offer the rewards their
-institutions value. We also discussed about successful models for industry
-partnerships that preserve open science. Participants noted that there are some
-NSF programs that prohibit partnerships with for-profit companies (but there are
-other programs in which this is encouraged).
-
-\subsubsection{Subgroup A Actions}
-
-Three main actions were identified by Subgroup A that would be of interest to
-participants and benefit the research community. These focused on the
-desire to amplify the current community efforts in training and software
-engineering by connecting the current training initiatives (e.g., Software and Data Carpentry,
-rOpenSci, OSS) that are of different durations at appropriate stages.
-Generally it was felt that a modicum of interaction between short term workshops
-(SwC, DC), medium term trainings (e.g., OSS), and longer-term courses (e.g.,
-BIDS Data Science) would benefit from coordinating curriculum, discussing and
-aligning prerequisites, and coordinating timing of courses. A training
-coordination effort would go a long way towards amplifying the value of the
-individual efforts and make them all more effective.
-
-\begin{itemize}
-\item Action 1: Create a roadmap of research software training initiatives. Such
-a roadmap would provide a taxonomy of training opportunities: what they deliver,
-and what attendees need to know going in (prerequisites). It would also show
-which recommended roadmap actions will have a clear and immediate payoff, and
-which will have long-term payoffs. The training roadmap would emphasize time
-savings and efficiency gains to be had from each training.
-
-\item Action 2: Build a report card characterizing use of best practices in
-scientific software. Generally, people felt that researchers would be very
-willing to migrate practices if they could identify where they needed to
-improve. Such a report card would ask simple questions to characterize use of
-best practices in science software, and could be structured similarly to Joel
-Spolsky's Software Maturity questionnaire (also known as
-\href{http://www.joelonsoftware.com/articles/fog0000000043.html}{the Joel
-Test}~\cite{joel-test}). The survey would create a report card that shows areas
-where a project could improve, and then link those areas to specific training
-offerings from the training taxonomy from Action 1.
-
-\item Action 3: Create a science software review forum. It was generally
-acknowledged that a little code review can have a tremendous impact on the
-quality of software in a project, but that sites for science code review are
-lacking. While people can ask questions on the Stack Exchange sites, they are
-generally not open to questions of style, approach, or appropriateness, as they
-try to avoid subjective commentary. Instead, we need a site where code gets
-discussed/summarized/described (in small bites) by the science software
-community. The target audience would be graduate and undergraduate students in
-the sciences, and there would need to be mechanisms to keep the review positive
-and constructive, and not get pedantic or judgmental. This could be tied to a
-software registry (such as the nascent GeoSoft project), or to language
-repositories like R's CRAN repository, and could lead to the report card
-discussed in Action 2.
-\end{itemize}
-
-\subsubsection{Subgroup B Actions}
-
-Subgroup B then focused on actions it could take. It discussed development of a
-white paper that describes a matrix approach to training (multi-day, multi-week,
-semester). The white paper might include a survey of existing techniques. There
-are many, some dating back many years, for example the Interuniversity
-Consortium for Political and Social Research (ICPSR) and various summer
-institutes. The white paper could include a call for a comprehensive assessment
-of these techniques. We need to think carefully about the right venue for such a
-white paper, where it would have the most impact. The subgroup believes it would
-need to approach editors directly to ascertain this.
-
-The group felt that training in techniques that promote sustainability has a
-range of benefits: career paths, educated reviewers, reproducible science, and
-more. There is some information on how that has been approached and assessed,
-but more is needed. The group felt that this training is undervalued and that it
-is important to communicate the return on investment (ROI) -- both individual ROI
-(skills that make scientists more effective and more marketable) and funder ROI
-(better use of taxpayer funds, research more likely to be reproducible because
-sustainable software exists, better trained reviewers).
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Papers}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Software Development Experiences Subgroup~A are:
-
-\begin{itemize}
-\item Michael R. Crusoe and C. Titus Brown. Channeling community contributions to
-scientific Software: A Hackathon Experience~\cite{wssspe2_crusoe}
-
-\item Marcus Hanwell, Patrick O'Leary, and Bob O'Bara. Sustainable software
-ecosystems: Software Engineers, Domain Scientists, and Engineers Collaborating
-for Science~\cite{wssspe2_hanwell}
-
-\item W. Christopher Lenhardt, Stanley Ahalt, Matt Jones, J. Aukema, S. Hampton,
-S. R. Hespanh, R. Idaszak, and M. Schildhauer. {ISEES-WSSI} Lessons for
-Sustainable Science Software from an Early Career Training Institute on Open
-Science Synthesis~\cite{wssspe2_lenhardt}
-
-\item Jory Schossau and Greg Wilson. Which Sustainable Software Practices do
-Scientists Find Most Useful?~\cite{wssspe2_schossau}
-
-\end{itemize}
-
-The papers that were discussed in the Software Development Experiences Subgroup~B are:
-
-\begin{itemize}
-
-\item Jordan Adams, Sai Nudurupati, Nicole Gasparini, Daniel Hobley, Eric
-Hutton, Gregory Tucker, and Erkan Istanbulluoglu. Landlab: Sustainable Software
-Development in Practice ~\cite{wssspe2_adams}
-
-\item Alice Allen and Judy Schmidt. Looking before Leaping: Creating a Software
-Registry~\cite{wssspe2_allen}
-
-\item Carl Boettiger, Ted Hart, Scott Chamberlain, and Karthik Ram. Building
-Software, Building Community: Lessons from the {ROpenSci} Project~\cite{wssspe2_boettiger}
-
-\item Yolanda Gil, Eunyoung Moon, and James Howison. No Science Software Is an
-Island: Collaborative Software Development Needs in
-Geosciences~\cite{wssspe2_gil}
-
-\item Ted Habermann, Andrew Collette, Steve Vincena, Werner Benger, Jay Jay
-Billings, Matt Gerring, Konrad Hinsen, Pierre de Buyl, Mark K\"{o}nnecke, Filipe
-Rnc Maia, and Suren Byna. The Hierarchical Data Format ({HDF}): A Foundation for
-Sustainable Data and Software~\cite{wssspe2_habermann}
-
-\item Eric Hutton, Mark Piper, Irina Overeem, Albert Kettner, and James Syvitski.
-Building Sustainable Software - The {CSDMS} Approach~\cite{wssspe2_hutton}
-
-\item James S. Spencer, Nicholas S. Blunt, William A. Vigor, Fionn D. Malone, W.
-M. C. Foulkes, James J. Shepherd, and Alex J. W. Thom. The {H}ighly {A}ccurate
-{N-DE}terminant ({HANDE}) Quantum {Monte Carlo} Project: Open-source Stochastic
-Diagonalisation for Quantum Chemistry~\cite{wssspe2_spencer}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Credit \& Incentives} \label{sec:credit}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Dan led writing of this section}
-
-%\note{\href{http://tinyurl.com/k8ruyn9}{Google doc notes}}
-%https://docs.google.com/document/d/1Jsi_trBT5cjEXVnGjGHJpcIhMEcVpAn6XmYNxfIHXv0/edit#heading=h.gk1dzxu9wi6q
-
-This group, with just three papers but a large amount of interest and
-participation from attendees, focused on the institutional, social, and cultural
-mechanisms that encourage the creation and maintenance of shared software, in the
-context of what now exists, what mechanisms are desired, and how
-we might achieve them.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-In the first discussion session, this group decided to break into two smaller groups,
-each independently working through the same general topic: credit and incentives.
-
-\subsubsection{First breakout discussion: Group A}
-The first sub-group discussed issues around the current system for credit and
-incentives, which it called ``hacking the incentive structure.'' The
-group considered four potential points of leverage:
-
-First, that we currently have systems that collect information, and these could
-be modified to collect different information, then map that information to
-actions. We could initially build a proof-of-concept for a new use of a given
-system, then determine what actions would be needed to make this use more
-common.
-
-Second that we could create entirely new systems, perhaps because the existing
-systems are too tied to what they measure, and modifying them is not practical.
-
-Third that we could change academic culture, rather than worrying about the
-systems. This was mostly focused on citations, because they matter for hiring,
-promotion, and tenure decisions. The group discussed how we could weigh the
-citations within papers better than we now do. How could we identify the five
-citations that really matter for a paper, distinguishing them from the related
-works and general background that are also cited? Perhaps we could break these
-out in the reference list, working with publishers to implement this. Or maybe
-we could also break out categories of citations, such as the most important
-software used, the previous publication that we are building from, the data that
-we actually used, etc. The Moore Foundation's award in data science was given as
-an example, asking proposers: What are the five canonical citations that are
-most important to your work?~\cite{moore-canonical} This would be a way of
-giving credit and assigning importance to these works, differently from how we
-just count citations today. A possible action that the group discussed was
-conducting a longitudinal study of most useful software, data, etc.\ in a
-discipline.
-
-Fourth, that we could change the ways funders make decisions, and use these
-funding policies as incentives.
-
-After this discussion, group A brainstormed about incentives,
-with the following items suggested:
-\begin{itemize}
-\item running programming contests, creating bounties for contributing to open
-source software, etc.
-\item augment author lists to give credit to people who do not now get credit
-(and making them machine readable)
-\item developing a microcitation standard and mechanism (for both software and
-data)
-\item developing a well-defined standard for author contribution -- what level
-of contribution rises to the level of authorship?
-\item leveraging social media for citation and reviewing of content -- then
-using social media to bring more people into the review process than is
-traditional
-\item determining where else software can be cited and recorded (e.g.,
-acknowledgments sections of papers)
-\item developing a taxonomy of contributors (e.g., Project
-Credit~\cite{projectcredit}) tied to places that these metrics are already
-stored (e.g., ORCID~\cite{orcid})
-\item making metadata easier to add for software, creating incentive for
-providing software metadata -- note that this cannot be centralized
-\item creating something like the h-index that tenure committees can make use of
-- simplify a way of measuring and documenting the overall credit given to an
-individual over different projects
-\item thinking about publishing software versus journal - software does not have to
-be novel
-\item determining guidelines for recommending software characteristics for
-tenure - perhaps draft guidelines then get ACM or IEEE agreement.
-\end{itemize}
-
-\subsubsection{First breakout discussion: Group B}
-
-This group started by discussing who should be incentivized, thinking of two
-categories of people: those in science (who could be incentivized to do better,
-more shareable, more sustainable work), and those in industry but interested in
-science (who could be incentivized to contribute to science.) It was pointed out
-that we are not yet clear enough on exactly what we want to incentivize,
-suggesting that we need to have a clearer picture of ``good computational work''
-and what sort of contributions are truly generative for science.
-
-The discussion of incentivizing those in science acknowledged that the
-publications system was far from perfect for incentivizing good software work.
-Nonetheless, the discussion focused on bringing software people into publications.
-There were two main suggestions. The first is to focus on end users of software
-and encourage them to cite software the ``right'' way. James Howison suggested
-that his research showed that few projects were making a formal request for
-citation (but that authors weren't necessarily following those suggestions
-anyway)~\cite{howison2015jasist}. He suggested making access to the software
-conditional on a license that requires citation. Others found this ``too
-confrontational'' and preferred to concentrate on making it easier to do the
-right thing. The second was focusing on those leading software projects, and the
-group was more enthusiastic about ``forcing'' PIs to include their ``software
-people'' on publications, although there were few ideas on how exactly to do
-this. Another technique mentioned was that when scientific software projects are
-hosted in organizations like Apache, the scientific contributors can benefit
-from building their reputations, perhaps yielding job offers that they can use
-to negotiate better job and career packages.
-
-The discussion on incentivizing those outside science focused on accessing the
-well of affection that those working in software have for scientific research.
-How can the interest and skills of this group be marshaled towards sustainable
-contributions? There is evidence that the migration of scientific software
-projects to the Apache Software Foundation (ASF) has created opportunities for those not
-employed in the scientific center to contribute to projects initiated by
-scientists (especially where there is cross-over with industry needs, such as
-provenance and workflow).
-
-The group also discussed developing ``software prizes'' arguing that while it is
-hard to ``mint'' other new sources of reputation, prizes are possible without
-getting too many others on board. The prize criteria can form a template for
-describing what we mean by scientific contributions made through software,
-particularly focusing on building active communities, not only writing great
-code.
-
-\subsubsection{Group merger and redivision}
-
-After the first breakout session, the groups A and B came together and discussed
-a rollup of the ideas from the subgroups at a high level:
-\begin{itemize}
-\item citation ecosystem -- traces of usage (metrics)
-\item taxonomy of contributorship, understanding roles
-\item prizes
-\item new metrics (for people's activities in software)
-\item guidelines for evaluating scientific contribution through software (perhaps
-using new metrics).
-\end{itemize}
-
-In the remaining discussion sessions, the group chose to split into three
-subgroups to discuss a version of these topics: {\em citation ecosystems}, {\em
-taxonomy of contributorship/guidelines for software for tenure review}, and {\em
-prizes}. The subgroups were asked to clearly identify
-\begin{itemize}
-\item the problem to be solved
-\item steps towards a solution.
-\end{itemize}
-
-\subsubsection{Remaining breakout discussions: Citation ecosystems}
-
-%\note{for reference, notes are in \url{http://tinyurl.com/l4tc6a4}}
-%https://docs.google.com/document/d/1SOja7k0mXwvOB2tQD3ZW3O38Qv4NwfA21zdMQ9JlUQc/edit
-
-This group defined its goal as creating a low-barrier-of-entry method for
-recording names and roles (and in a second phase, optionally including weights)
-of contributors (coders and other intellectual contributions) to a software
-package in a machine readable way (to be called a credit file), then encouraging
-the scientific community to adopt this practice.
-
-The following general points were initially discussed:
-\begin{itemize}
-\item The FLASH~\cite{flash} project was suggested as an example of how something
-like this has been done.
-\item This data could be a file that can be associated with a citation to the
-software, either through use of a DOI (digital object identifier) for the credit file, or by uploading the
-credit files as associated with the paper.
-\item This idea could also be applied to data.
-\item The credit file should be part of the metadata that are freely
-available to those who have not paid for access to the journal, like citations
-are now in most cases.
-\item It was suggested that there should also be a \emph{separate} file to track
-software dependencies.%, but this is not the same file.
-\end{itemize}
-
-The group came up with the following actions to be performed:
-\begin{enumerate}
-\item Build a tool that can automatically determine who the contributors are
-(from a Git or other repository), then allows the user to manually edit the output to
-add/remove people, define roles.
-\item Work with repositories to encourage them to provide the information we
-need based on what they already store.
-\item Define what a citation file should look like and what it should be called.
-\item Test adoption with a substantial scientific organization such as the Lawrence Berkeley National Laboratory (LBNL).
-\item Create credit file for a set of software.
-\item Build a validator (and perhaps a visualizer) for credit files.
-\item Write a tool to collect files and visualize/output interconnections (which
-software is used with which), based on an existing project.
-\item When we (the group members) write papers, we should track the software
-we use, and encourage the software developers to make their software citable
-and create credit files.
-\item Build a tool to export the credit file to BibTex and other citation styles.
-\item Make sure the BibTex entries (exported from internal data) are somewhat
-standardized so that they can be imported into papers.  Also make sure that
-standard LaTeX style files understand and accept these entries.
-\end{enumerate}
-
-\subsubsection{Remaining breakout discussions: Taxonomy of contributorship/guidelines for software for tenure review}
-%\note{by Frank Seinstra}
-
-%\note{for reference, notes are in \url{http://tinyurl.com/mx58tmx}}
-%https://docs.google.com/document/d/1sDZLaBR2elZ9Tj6eLIwEtf5VqQ_x_AN_SetKtJlFUpI/edit
-
-At the start of the discussion, the breakout group brought forth the important
-observation of the wide disparity in commonly accepted habits of publication in
-different research fields. In domains which have, historically, relied on large
-groups of researchers collaborating towards a common goal (e.g., high-energy
-physics, astronomy), publications often have tens or even hundreds of co-authors
-(with some papers in experimental particle physics having over 3000). In other
-domains, the number of co-authors is typically much smaller with, in some
-cases, even a preference for single-author papers. Similarly, the various
-platforms for publication are valued differently in different domains. Most
-commonly, publications in peer-reviewed scientific journals are regarded as the
-most important and most impactful. However, in certain domains, especially in
-Computer Science, many researchers typically regard conference proceedings as
-their prime publication target. It is often suggested that this difference is
-due to the rapid developments in information technology, a pace that cannot be
-upheld by traditional peer-reviewed journals. Whatever the causes, any useful
-taxonomy of contributorship or guideline for tenure review should take such
-differences into account.
-
-Despite these differences, and despite the fact that software often has taken
-the role of a proper, albeit less tangible, scientific research instrument,
-neither the software nor its creators are commonly credited as part of a
-scientific publication. The group acknowledged the need for more recognition for
-the creators of such software instruments, and indicated a number of possible
-pathways. First and foremost, domain scientists must be made aware of the
-important role of software, and include the developers as co-authors of papers.
-A second approach is to fully embrace an open badging infrastructure (such as
-Mozilla's Open Badges), where a {\em badge} is a free, transferrable,
-evidence-based indicator of an accomplishment, skill, quality, or interest. A
-third approach is for the scientific community to support the increasing
-momentum of peer-reviewed journals specialized in the open source/open access
-publication of scientific research software, such as Computer Physics
-Communication, F1000 Research, Journal of Open Research Software, and SoftwareX.
-
-Recognizing publication of research software as a proper scientific contribution
-raises several important but currently unsolved questions, however. For example,
-is the number of users of the software a relevant measure of impact? What
-standards of coding quality must be followed in order to justify publication and
-hence recognition? Should the release of a new version of the software be
-eligible for a new publication; if so, under what conditions? And above all:
-should software publications be valued in the same way as traditional scientific
-publications? Or is there a need for new measures of productivity and impact?
-
-In part, the answers will come from the scientific community at large, as a
-natural consequence of growing awareness and mindset change. Some of the
-answers, however, also should be based on decades of experience in (and
-developing standards for) implementing, maintaining, refactoring, documenting,
-testing, and deploying software instruments in scientific research. Care should
-be taken, however, not to impose such standards for all domains in equal ways
-right from the start. Forerunners should serve as an example, but should not
-scare away domains that have based their progress on much less advanced methods
-of {\em software carpentry}. Nevertheless, proper guidelines are needed, which
-eventually should be followed across all domains. The group also recognized that
-funding bodies, universities, and publishers eventually should demand that
-research projects follow such guidelines, and to implement a proper software
-sustainability plan.
-
-To enable a form of standardized crediting for developers of research software,
-the group proposed to work towards a taxonomy for software-based
-contributorship. The taxonomy should be derived from, or extend, existing
-taxonomies for research impact and contributorship such as defined by CASRAI (in
-particular based on the Wellcome-Harvard contributorship
-taxonomy\footnote{Project CRediT, \url{http://credit.casrai.org/}}, VIVO, or
-ISNI. An interesting measure of impact raised by the group was the {\em
-betweenness centrality}, an indicator of a person's centrality (and hence,
-importance) in a scientific collaboration. It is expected that developers of
-research software often play such a central role. %\choinote{cite VIVO or ISNI?}
-
-The group defined the following actions to be performed:
-\begin{enumerate}
-\item Investigate existing taxonomies for roles and contributorships.
-\item Investigate prototype badging initiatives.
-\item Investigate journals focusing on publishing peer-reviewed research
-software.
-\item Investigate guidelines and checklists of best practices.
-\item Communicate the results of the above investigations to the WSSSPE
-community and decision-making bodies (funders, publishers, universities,
-and tenure committee representatives).
-\item Ensure engagement of the broader research community in this discussion.
-\end{enumerate}
-
-\subsubsection{Remaining breakout discussions: Prizes}\label{sec:prizes}
-%\note{sub-breakout notes: \url{http://tinyurl.com/mjok4o5}}
-%https://docs.google.com/document/d/1foSppeHQeQ-qOjd-Rp_CUCNy8JMLxd_vrCDMmqvG2-w/edit
-
-This group discussed the idea of prizes. Prizes are expected to reframe software
-as ``instrument building'' but will prizes be good or bad, and how can we make
-sure there are no negative affects and the process cannot be gamed?
-
-Prizes in different categories were discussed (like Academy Awards), for
-example: best contribution (non-founder), broadest diversity of contributions,
-best tutorials or documentation, best leadership transition (award ex-leader and
-new leader), best generalization (taking something that was limited and making
-it more general), and best mentorship of contributors (bringing others into the
-community).
-
-Other (non-prize) forms of incentives are
-\begin{itemize}
-\item Converting reputation by joining the ASF, Google, etc.
-\item Inviting ASF and open source people to contribute to scientific code (at
-least in areas where there is overlapping interest).
-\item Template for assessing scientific contributions made through software.
-\end{itemize}
-
-The group suggested that to determine who should give out prizes, perhaps we
-should find those who we think would be awarded prizes, then ask them who they
-would want to receive a prize from, and subsequently contact those organizations to see if they
-are willing to be involved in the process.
-
-One of the group's ideas was to create a funding program for disciplines or
-other organizations to create a prize program. We would provide a framework and
-a set of requirements, for example: awards to individuals; must award in 5 or 6
-areas; must have a jury that includes senior/junior
-domain experts and technologists (and should have objective criteria); must have the recipients awarded at a
-relevant event; and must provide citations that explain why the prizes should be
-awarded. Different organizations could then decide to sign up to the framework
-and give out awards under this general brand. However, there was a concern that
-having many organizations award prizes may reduce the impact of the prizes.
-
-Possible groups that might give out awards, either under our framework, or more
-generally, are AAAS, Nature Publishing Group, ACM, IEEE, Astro, Ecology Society
-of America, etc. Perhaps this could be a joint technology/science partnership,
-for example, the [Apache $|$ Mozilla]-[AAAS $|$ disciplinary society] prize.
-
-Some potential criteria for prizes the group suggested are:
-community engagement, helping out others;
-number of unique contributors;
-adding new pieces of functionality to software;
-integrating software into broader ecosystem; championing broad principles of
-sustainability, open science, open source, etc.;
-improving accessibility to software, to scientific software (perhaps championing
-inclusiveness or making software accessible?);
-documentation;
-tutorials;
-commits/patching;
-leadership transition; and
-best contribution by a non-founder.
-
-The group discussed if there should be different criteria for ``established''
-members of the community versus junior members, and if prizes should be
-restricted to junior members, but left these as open questions.
-
-A point the group considered important is that we do not want to give prizes just
-to reward people who are really good at this one thing, but rather we want to
-reward people who are building the culture we want as scientists.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Papers}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Citation \& Incentives group are:
-\begin{itemize}
-\item James Howison. Retract Bit-rotten Publications: Aligning Incentives for
-Sustaining Scientific Software~\cite{wssspe2_howison}
-
-\item Daniel S. Katz and Arfon M. Smith. Implementing Transitive Credit with
-{JSON-LD}~\cite{wssspe2_katz}
-
-\item Ian Kelley. Publish or Perish: The Credit Deficit to Making Software and
-Generating Data~\cite{wssspe2_kelley}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Reproducibility \& Reuse \& Sharing} \label{sec:reproduce}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Neil led writing of this section}
-
-%\note{\href{http://tinyurl.com/kqpe87z}{Google doc notes}}
-%https://docs.google.com/document/d/1Mivadj16_9tPrw8Nxmhp72AtNC0BEVO-f7M84FXTbSU/edit#heading=h.elaktnt2tep9
-
-This group discussed five papers with a wide variety ideas of how to support
-reproducibility and reuse. It focused on identifying concrete practices that the attendees
-could work together on, which would have a positive effect on the community.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-In the first discussion session, this group broke up into two smaller groups
-working on different topics: Reproducibility, and Reuse and Sharing. In the
-second discussion session, three subgroups worked on creating specific pieces of
-guidance.
-
-\subsubsection{First breakout discussion: Reproducibility group}
-
-The first group discussed ways in which reproducibility of papers could be
-improved. A consensus surrounded the provision of examples: demonstrating to
-others how to achieve reproducibility. Major public funding investments go into
-research that heavily relies on reproducible software, hence the lack thereof
-is raising major concerns.
-
-Two main avenues could be used to implement policy that would improve
-reproducibility and drive top-down culture change:
-\begin{itemize}
-\item Funders can aim to get more software expertise on to funding review panels
-and provide more guidance on what is required of software related to research
-(cf.\ data management plans)
-
-\item Journals can define and publish journal policies to improve
-reproducibility, and reviewers can insist that authors provide sufficient
-information and access to data and software to allow them reproduce the results
-in the paper. Stronger policies even for some high-impact journals have recently
-come into place, for example Nature Publishing journals.
-\end{itemize}
-
-If journals do enforce greater reproducibility constraints, it is important to
-lower the barriers to reviewers attempting to verifying the correctness of the
-software used to generate the results. A major issue is that a lot of software
-only builds on certain systems. Should journals provide more tools/support for
-reviewers, and if so, what is it? An idea that came from the groups was to
-define a set of support services that should be available to software paper
-reviewers. Another was to provide the ability for reviewers to flag the
-requirement for a `software verification,' similar to the ability to flag that a
-paper needs to be seen by a statistician.
-
-Other discussion focused on ways in which researchers themselves could improve
-the reproducibility of research. One way would be to establish tracks at
-conferences that subject papers to reproduction, which for those that pass would
-lend them an additional badge (similar to OOPSLA). Another is simply to get more
-people to use your software: for instance by outreach to high school students --
-can your software be used by them? Finally, there is a role for
-community-curated benchmarks to validate the performance and capabilities of
-tools.
-
-\subsubsection{First breakout discussion: Reuse and Sharing group}
-
-The second group discussed ways in which software could be more easily reused
-and shared. From the perspective of both user and developer, any solutions must
-be 1) easy; 2) cheap; 3) not too time-consuming.
-
-Reuse and sharing were considered to be distinct but linked. In many cases,
-pre-existing software does not exist, so new software is written but even then
-it is not shared afterwards. The principal barriers to reuse are the difficulty
-of finding out what software is available, understanding if it is usable, and
-then of installing and running software if it is located.
-
-Discovery of relevant software is still a fundamental issue: we need standard
-vocabularies and metadata, better tools, and approaches that are sustainable.
-Publications are an easy entry point for locating suitable software, but should
-the publishers lead the way, or is this the responsibility of ``the community''?
-Some communities have had significant initiatives to improve software discovery,
-e.g., the NIH Software Discovery Index\footnote{NIH Software Discovery Index:
-\url{http://softwarediscoveryindex.org/}}. Likewise, there were examples of
-journals which had made software more discoverable: ACM Transactions on Software
-offers a reproducibility review; the Journal of Biostatistics has an opt-in to
-provide code and a certification mark if it can be run; the Journal of Open
-Research Software requires software to be deposited in suitable repositories and
-referenced.
-
-An issue around the sustainability of software catalogues is that their
-usefulness often depends on the domain. For instance, in the biosciences, there
-is more homogeneous data and standard shared code. In areas like ecology,
-code is often very specific to a problem, meaning that the level of re-use might
-be at a general statistical level of abstraction, but then every research
-use is highly customized.
-
-Finally, it was clear from the discussion that there were many ways in which
-software could be made available in more reusable ways than just a tarball
-sitting on a personal website. Using code repositories like
-GitHub\footnote{GitHub: \url{http://github.com/}} gets you an archived, shared
-platform and improve the reusability of your software incrementally, for
-instance by adding a license or by archiving (with a DOI) in
-Figshare\footnote{FigShare: \url{http://figshare.com/}} or
-Zenodo\footnote{Zenodo: \url{http://zenodo.org/}}. Docker\footnote{Docker:
-\url{htp://www.docker.com/}} might be a solution to the issue of dependencies,
-to allow binaries and libraries to be bundled in a more lightweight fashion than
-a virtual machine image.
-
-The key enabler for reuse and sharing was to get domain scientists more
-effectively connected with programmers/analysts. Both have skills and experience
-which is necessary to make the right decisions for improving the reusability and
-discoverability of software, and to apply community pressure to change practice.
-
-\subsubsection{Second breakout discussion: Categorization of software journals}
-
-This group aimed to come up with a categorization of journals which published
-software papers. Starting from the list of
-journals\footnote{\url{http://www.software.ac.uk/resources/guides/which-journals-should-i-publish-my-software}}
-maintained by the Software Sustainability Institute, the group chose seven
-journals and looked at their advice to authors and reviewers. These were the
-Journal of Open Research
-Software\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies\#peerReviewProcess}},
-PLoS ONE\footnote{\url{http://www.plosone.org/static/guidelines\#software}},
-Journal of Statistical
-Software\footnote{\url{http://www.jstatsoft.org/instructions}}. Methods in
-Ecology and
-Evolution\footnote{\url{http://www.methodsinecologyandevolution.org/view/0/authorGuidelines.html}},
-Transactions of Mathematical
-Software\footnote{\url{http://toms.acm.org/Authors.html}},
-GigaScience\footnote{\url{http://www.gigasciencejournal.com/about}}, and PLoS
-Computational
-Biology\footnote{\url{http://www.ploscompbiol.org/static/guidelines\#software}}.
-%
-From these a set of common categories were synthesized, against which all
-journals could be compared:
-
-\begin{itemize}
-\item Journal Policies
-\begin{itemize}
-	\item Accessibility of papers
-	\begin{itemize}
-		\item Open access
-		\item ``Freely'' available
-	\end{itemize}
-	\item Repositories
-	\begin{itemize}
-		\item Provides suggestions for recommended repositories
-		\item Provides its own repository
-	\end{itemize}
-	\item Review
-	\begin{itemize}
-		\item Reviewing software is mandatory
-		\begin{itemize}
-			\item Must check that software runs
-			\item Must check quality of code
-			\item Must check performance of code if paper makes claims on relative performance
-		\end{itemize}
-	\end{itemize}	
-	\item Supporting data
-	\begin{itemize}
-		\item Must be publicly available
-		\item Must be in an open access repository
-		\item must have a DOI
-	\end{itemize}
-	\item Article processing charge (APC)
-	\begin{itemize}
-		\item APC charge is transparent
-		\item APC waiver program
-	\end{itemize}
-\end{itemize}
-
-\item Paper Policies
-\begin{itemize}
-	\item Required sections
-	\item Keywords
-	\begin{itemize}
-		\item Paper provides keywords to help describe software
-	\end{itemize}
-	\item Papers can be updated when new releases of software are made
-	\begin{itemize}
-		\item At no extra cost/at significantly reduced cost
-	\end{itemize}
-\end{itemize}
-
-\item Software Policies
-\begin{itemize}
-	\item Software must have a license
-	\begin{itemize}
-		\item Software must have an open source license
-	\end{itemize}
-	\item Availability
-	\begin{itemize}
-		\item Software must be openly available and accessible
-	\end{itemize}
-	\item Deposit policies
-	\begin{itemize}
-		\item Software should be in a public repository
-		\begin{itemize}
-			\item Of particular stature/with a preservation plan
-		\end{itemize}
-		\item Software should have a permanent identifier
-		\item Software deposit doesn't count as a prior publication
-	\end{itemize}
-	\item Runnability and dependencies
-	\begin{itemize}
-		\item Provide documentation to understand how to run
-		\item Provide sample data
-		\item Provide dependency information
-	\end{itemize}
-\end{itemize}
-\end{itemize}
-
-The follow-up actions to this work are to use this set of categories on all the
-journals in the list, refining the categories if necessary, then identify if any
-of the categories are seen to be more useful to promote reproducibility, reuse
-and sharing.
-
-\subsubsection{Second breakout discussion: What should journals provide reviewers of software papers?}
-
-This group discussed whether they could come up with a list of things that a
-journal should provide its reviewers to make it easier to review software
-submitted for publication as a ``Software Paper.''
-
-Journals should provide guidelines about what to consider when reviewing a
-software submission. A good example for a relatively comprehensive guidelines
-for reviewers (and thus in turn authors) are those of
-JORS\footnote{\url{http://openresearchsoftware.metajnl.com/about/editorialPolicies}}.
-Journals might also learn from organizations such as the ASF as to
-what is good practice for software submissions. Guidance is needed on what
-constitutes an incremental improvement that is significant enough to qualify for
-publication, otherwise this assessment can be very subjective.
-
-Journals should provide mechanisms to enable and track communication between
-reviewers and authors. For anonymous peer review journals, anonymity of
-reviewers should be maintained. If communication is necessary, that may mean that software is not
-that well documented. Journals should also provide a set of simple metrics for
-software evaluation that reviewers can use for ratings, similar to Consumer
-Reports.
-
-Journals should provide guidelines about requirements for documentation of code:
-both in-lined in code, and manuals/web pages, etc. Journal editors could provide
-documentation checks before it goes out for review. This should include a
-requirement for good Use Cases specified for the software, with references to
-executable test cases that demonstrate each use-case is met (at least in the
-form of the test case).
-
-Journals should support mechanisms to run software. Sometimes this may be very
-hard to accomplish, despite best efforts. Journals could provide an execution
-environment for any software submitted for review, perhaps via a Docker
-container or a virtual machine. If not, instructions must be adequate to compile
-and execute the software; to interpret the results (output files, formats,
-etc.); and the full source code must be accessible to the reviewer. Mechanisms
-to quantify what has changed compared to a previously published version would
-assist version comparison. However, this cannot simply take lines of code into
-account. For example, a speedup of an algorithm may not result in a huge code
-difference, but may nonetheless provide a breakthrough. Journals should require
-software submissions to also provide `test materials' -- sample data, parameters
-to validate that code is working as intended. In certain cases, well-selected
-benchmark datasets may be required to assess performance and accuracy.
-
-Journals should ensure minimal metadata are provided, similarly as is already
-the case for certain kinds of data (though the latter is often enforced by data
-repositories) -- Dublin core-ish (creator, owner), and more specific (platform
-and compiler dependencies, sample benchmarks of performance, etc.). Journals
-should provide guidance or constraints as to software licensing conditions.
-
-The follow-up actions to this work are to work with journals and reviewers to
-identify whether any of these suggestions can be easily provided, perhaps for a
-range of journals.
-
-\subsubsection{Second breakout discussion: Reproducibility Meta-track toolkit}
-
-This group worked on defining a ``toolkit'' for running a reproducibility
-meta-track at a conference. They decided to take the work done during the
-workshop and publish it as a paper.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Papers}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-The papers that were discussed in the Reproducibility \& Reuse \& Sharing group are:
-\begin{itemize}
-\item Jakob Blomer, Dario Berzano, Predrag Buncic, Ioannis Charalampidis,
-Gerardo Ganis, George Lestaris and Ren\'{e} Meusel. The Need for a Versioned
-Data Analysis Software Environment~\cite{wssspe2_blomer}
-
-\item Ryan Chamberlain and Jennifer Schommer. Using {Docker} to Support
-Reproducible Research~\cite{wssspe2_chamberlain}
-
-\item Neil Chue Hong. Minimal Information for Reusable Scientific
-Software~\cite{wssspe2_chue_hong}
-
-\item Tom Crick, Benjamin A. Hall, and Samin Ishtiaq. ``Can I Implement your
-Algorithm?'': A Model for Reproducible Research Software~\cite{wssspe2_crick}
-
-\item Bryan Marker, Don Batory, Field G. Van Zee, and Robert van de Geijn. Making
-Scientific Computing Libraries Forward Compatible~\cite{wssspe2_marker}
-
-\item Stephen Piccolo. Building Portable Analytical Environments to Improve
-Sustainability of Computational-Analysis Pipelines in the
-Sciences~\cite{wssspe2_piccolo}
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Code Testing \& Code Review} \label{sec:code_testing}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Dan leading writing of this section, with help from Tom Clune and Karen Cranston}
-
-%\note{\href{http://tinyurl.com/l5t5h45}{Google doc notes}}
-%https://docs.google.com/document/d/1G1_U5-hkAvDKtTgT4DvUhIYvyYUY4cBBvQHJIdgADa0/edit#heading=h.v7g0nqpn1fur
-
-This group began by recognizing that they knew what the problems of code testing and code
-review are, so the purpose of the group was really to think about
-concrete things that people in the group would be willing to commit to, in order
-to improve the sustainability of software.
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Discussion and Actions}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-The consensus of the group was that small, concrete actions would have the
-greatest chance of proper followthrough by members of the group. An example of
-this approach was the Architecture of Open Source Applications book~\cite{aosa},
-containing contributed chapters on open source software. The group contemplated
-creating an analogous book or web resource for testing of scientific software.
-
-A key point of discussion was that there are two major challenges to software
-testing in science. The first is convincing developers to incorporate testing.
-This challenge is both social and technical. We need to communicate the value of
-testing and also teach developers how to choose and use testing frameworks. The
-second challenge---and the one that the group felt was more difficult---is
-choosing appropriate tests for scientific software. This is the difference
-between learning how to use, say, Python \texttt{unittest} and knowing how to test
-that one's code correctly implements, for example, a Lattice-Boltzmann model in
-computational fluid dynamics.
-
-\subsubsection{Choosing and implementing testing frameworks} There was general
-consensus that once a developer has been exposed to a testing framework in a
-nontrivial fashion, he/she will subsequently insist upon using a framework for
-further work. The acknowledged challenge was how to create that crucial first
-exposure to such frameworks.
-
-Within the group, there were several different paths by which participants had their 
-first exposure to code testing:
-\begin{itemize}
-
-\item Formal tutorials at conferences (for example, a test driven development
-(TDD) at a Software Developers Best Practices conference): These often require
-self-learning following a tutorial, but it can be hard to find a tutorial that
-matches the programming language and/or the domain of the participant.
-
-\item As a way to be confident in other software: When breaking out pieces of
-software from a larger application, a developer wants to ensure the software
-works as expected. These types of tests may be ephemeral though, living long
-enough to give the author of the test confidence in the code, but not handed on
-and made available to next user. These experiences can lead to more systematic
-testing.
-
-\item Through experience with coding: As you write more software, your
-confidence in code is reduced and the importance of testing becomes clearer.
-Attendees joked that they started testing ``when their skepticism/guilt became
-larger than their arrogance.''
-
-\item From other projects: When building on top of projects with good testing
-frameworks, you realize that you should also adopt those good software
-engineering habits.
-
-\end{itemize}
-
-The two non-self-taught paths were direct learning and indirect through other
-software (learning by example). The group recognized that teaching about testing
-needs to start early in projects and careers. We need to teach people how to
-test short bits of code, rather than waiting until they have thousands of lines
-of code and being surprised when they don't write tests. The group generated a
-set of practical suggestions for improving adoption of code testing.
-%
-Established software projects can encourage the latter through a simple rule of
-accepting only software patches with associated tests and through good
-documentation of testing practices and requirements.
-
-Simpler, standard ways of setting up appropriate testing infrastructure will be
-important for adoption by scientists. Jenkins\footnote{\url{https://jenkins-ci.org/}} 
-was suggested as a common open-source
-solution, but the initial configuration was considered to be challenging for
-typical scientists.
-
-Teach testing by beginning with a (smallish) piece of code that lacks
-appropriate tests and develop an exercise that involves refactoring the code
-into a more presentable form through creating/deriving unit tests. Students
-would extract unit-tests through reverse-engineering and/or questioning a
-partner who is a developer/expert.
-
-This led into a more general discussion about whether programming courses
-should be part of the standard curriculum for science students and the
-challenges of fitting additional material into university curricula. This is the
-reasons that short workshops such as Software Carpentry and others exist.
-
-Ironically, Software Carpentry no longer teaches testing~\cite{SCtesting}. This
-boils down to two issues. First, that scientific computing doesn't (yet) have
-the cultural norms for error bars that experimental sciences have, and second,
-that there is a breathtaking diversity of scientific code; scientific research
-is highly specialized, which means that the tests scientists write are much less
-transferable or reusable than those found in other fields.
-
-\subsubsection{Writing tests for scientific software}
-The second major discussion thread in the group addressed the issues identified
-by Software Carpentry -- the difficulties in writing tests appropriate for
-scientific software: %Even among this group, while many 
-What tests are
-appropriate to ensure that an complex method is working correctly? When is the
-result of a numerical computation `close enough' to pass a test?
-
-Several members of the group indicated that while they valued the concept of
-software testing in theory, they were unaware of how to test certain kinds of
-software. Or, more importantly, they lacked relevant examples of software
-testing that would be suitable for new members of their teams. As an example, the
-group asked if it could produce a 100-line example of how to test neutron
-transport (or other specific scientific examples) targeted at a sophomore? This
-would be a demonstration of ``How do I do this right?''
-
-Another question was ``How do I test so that I know that the code is not the
-source of my problems?'' A common challenge is that of stripping things down to
-an appropriate level for tests. The group recognized that the difficulty is
-different for so-called software infrastructure than for numerical/scientific
-layers of software applications.
-%
-%The group referred to Lorena Barba's ``12 Steps to Navier
-%Stokes~\cite{12steps},'' an online CFD lesson repository created as a
-%set of iPython notebooks.  \clunenote{I'm not sure of the context
-%  here, except possibly as an analog for something under discussion.}
-%
-%
-%Distinction between 100 lines of input files vs. 100 lines of source code \katznote{to delete? - don't see how it fits}
-%\clunenote{I no longer recall the context for this statement; does not seem important to me.}
-%
-There is some discussion of the issue in the computer science
-literature~\cite{hook1,hook2} but less in domain-specific journals. Some lessons
-on multiphysics software verification can be found in a recent
-paper~\cite{dubey-FLASH}.
-
-Two important pieces of the barrier are a) picking/implementing testing
-frameworks (a technical barrier) and b) deciding what are the actual tests that
-I need to write? A possible method to solve the latter is to ask students ``how
-is what you are doing different from typing random keys on the keyboard?'' and
-then, turn their answer into the concept of tests. That is to say, expand the
-process of creating code from one that is solely about ``how?'' into one that
-also includes ``why?''
-
-Another point of discussion is that many scientists are not aware that ``testing
-the code'' and ``testing the science'' are distinct issues. A related question
-that came up is ``How do you know that your method produces a result that is
-`close enough'?''
-
-The group then decided that a useful action would be to create a set of
-scientific codes with associated tests. They developed the following basic
-structure for these examples and sketched out six specific examples:
-\begin{enumerate}
-
-\item A paragraph or two explaining the scientific problem the code addresses
-
-\item The size of the simplest piece of relevant, interesting code (i.e., an
-estimate of lines of code)
-
-\item A point-form list of the test cases you would use.
-
-\end{enumerate}
-
-The follow-up actions from this work include creating a compilation of testing
-examples in scientific software. Some of the examples from the workshop have
-become the starting seed of a collaborative book
-(https://github.com/swcarpentry/close-enough-for-scientific-work) in which
-scientists provide concrete examples of testing scientific software. The goal is
-a set of testing examples, aimed at sophomores in science and engineering, that
-cover a broad range of domains and problems and could be easily incorporated
-into other workshops or courses.
-
-%See the ``13 dwarves'' from Berkeley http://www.eecs.berkeley.edu/Pubs/TechRpts/2006/EECS-2006-183.html \katznote{to delete? - don't see how it fits}
-%\clunenote{No opinion - someone thought it was a useful reference for the topic.}
-
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-\subsection{Papers}
-%~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
-
-The papers that were discussed in the Code Testing \& Code Review group are:
-\begin{itemize}
-\item Thomas Clune, Michael Rilee, and Damian Rouson. Testing as an Essential
-Process for Developing and Maintaining Scientific Software~\cite{wssspe2_clune}
-
-\item Marian Petre and Greg Wilson. Code Review for and by
-Scientists~\cite{wssspe2_petre}
-
-\item Andrew E. Slaughter, Derek R. Gaston, John Peterson, Cody J. Permann,
-David Andrs, and Jason M. Miller. Continuous Integration for Concurrent {MOOSE}
-Framework and Application Development on {GitHub}~\cite{wssspe2_slaughter}
-\end{itemize}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Working Groups} \label{sec:WGs}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
+\todo{need to update for WSSSPE3}
 The WSSSPE2 workshop continued our experiment from WSSSPE1 in how we can
@@
 \centering
-\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014.}\label{tab:tweets}
+\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014. \todo{update this for WSSSPE3?}}\label{tab:tweets}
   \begin{scriptsize}
@@
 
-%\todo{feel free to add stuff here}
+\todo{feel free to add stuff here}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{need to update for WSSSPE3}
 The following is a partial list of workshop attendees who registered on the

### 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
Rename wg_appendix_software_credit.tex to wg_appendix_template.tex

\section{Software Credit Working Group Discussion}

\subsection{Group Members}
{\small
\begin{longtable}{ll}
   Alice Allen           &  Astrophysics Source Code Library (ASCL)
\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
\end{longtable}
}

\subsection{Summary of Discussion}

\subsection{Description of Opportunity, Challenges, and Obstacles}

\subsection{Key Next Steps}

\subsection{Plan for Future Organization}

\subsection{What Else is Needed?}

\subsection{Key Milestones and Responsible Parties}

\subsection{Description of Funding Needed}

### 2015-09-30 WSSSPE3/report/wg_main_legacy_SW.tex
For main report
Former-commit-id: a5b6ce61a25ea1be2db5109d7f96e0cbf386a995

\subsection{Topic} \todo{change the title to your topic}

\subsubsection{Why it is important}
\todo{short text here}

\subsubsection{Fit with related activities}
\todo{short text here - can include links/cites}

\subsubsection{Discussion}
\todo{short-ish text here}

\subsubsection{Plans}
\todo{short text here - not bullets}

\subsubsection{Landing Page}
\todo{link to landing page}

### 2015-09-30 WSSSPE3/report/wg_appendix_software_credit.tex
Create wg_appendix_software_credit.tex
Update wg_appendix_software_credit.tex
Completed group list

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Software Credit Working Group Discussion}
+\label{sec:appendix_SW_credit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
    Alice Allen            &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
+\\ Sou-Cheng Choi         &  NORC at the University of Chicago and Illinois Institute of Technology
+\\ James Hetherington     &  University College London
+\\ Lorraine Hwang         &  University of California, Davis
+\\ Daniel S.\ Katz        &  University of Chicago \& Argonne National Laboratory
+\\ Frank Löffler          &  Louisiana State University
+\\ Abby Cabunoc Mayes     &  Mozilla Science Lab
+\\ Kyle E.\ Niemeyer      &  Oregon State University
+\\ Grace Peng             &  National Center for Atmospheric Research
+\\ Ilian Todorov          &  Science \& Technology Facilities Council, UK
 \end{longtable}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Summary of Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Key Next Steps}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Plan for Future Organization}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{What Else is Needed?}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Key Milestones and Responsible Parties}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Description of Funding Needed}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-09-30 WSSSPE3/report/wg_main_user_community.tex

\subsection{Building Sustainable User Communities for Scientific Software}
%\todo{change the title to your topic}

\subsubsection{Why it is important}
%\todo{short text here}

User communities are the lifeblood of sustainable scientific software. The user community includes the developers, 
both internal and external, of the software; direct users of the software; other software projects that depend on
the software; and any other groups that create or consume data that is specific to the software. Together these
groups provide both the reason for sustaining the software and, collectively, the requirements that drive its continued
evolution and improvement.

\subsubsection{Fit with related activities}
\todo{short text here - can include links/cites}

\subsubsection{Discussion}
\todo{short-ish text here}

\subsubsection{Plans}
\todo{short text here - not bullets}

\subsubsection{Landing Page}
\todo{link to landing page}

### 2015-09-30 WSSSPE3/report/wg_appendix_publishing_SW.tex
Create wg_appendix_publishing_SW.tex
Content for the appendix on Executable Papers
Former-commit-id: eb9f7144117ea6dcd54deee46dbae986146aeef1

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Publishing Software Working Group Discussion}
+\label{sec:appendix_publishing_SW}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
+   Steven R. Brandt & Louisiana State University
+\\ Daniel Gunter    & 
+\\ Yuhan Ding       & 
+\\ Neil Chue Hong   & Software Sustainability Institute
 \end{longtable}
@@
 
+This group explored the value of executable papers (papers whose content includes
+the code needed to produce their own results), and other forms of publishing which
+include dynamic electronic content. Transitioning to this type of publication offers
+possibilities of addressing, or partially addressing sustainability concerns 
+such as reproducibility (the paper contains all the artifacts needed to verify its
+results), transitive credit (modules an executable paper depends on must be explicitly
+loaded, making it more feasible to identify them), and improving documentation (an executable
+paper must explain what its code does).
+
+While the group did not feel it had a way to influence what is currently happening
+in these experimental publishing venues, it felt that creating and curating a list of
+such efforts would have value by attracting interest to these activities.
+The Software Sustainability Institute agreed to host this list.
+
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+
+The opportunity is to collect a list of current executable papers and
+shine a light on the experiments and development efforts currently underway.
+
+The only obstacle to this is the difficulty in finding and identifying such
+publications. The Software Sustainability Institute was able to do something similar
+for publications about software by making a public page containing a catalog
+of these publications and enlisting the help of the community to grow the list.
 
@@
 
+Create the first version of the web page to be displayed on the Software Sustainability
+Institute's website.
+
 \subsection{Plan for Future Organization}
+
+None at this time.
 
@@
 
+Nothing else at this time.
+
 \subsection{Key Milestones and Responsible Parties}
 
+Steven R. Brandt will create a first version of the page within a week or so of the WSSSPE3 conference.
+
+Neil Chue Hong will take responsibility for the page once it is up.
+
 \subsection{Description of Funding Needed}
+
+None.

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
use utf8
Added includes for every wg_main file
Former-commit-id: b07a1db0988ce04a85a5a6860db6a4c13a79e2d7
add names and titles of lightning talks
Former-commit-id: 5478422906ae768e27dceb886d50b047e87af0f7

@@
 \usepackage{graphicx}
-\usepackage[latin9]{inputenc}
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+\usepackage{lmodern}
 \usepackage{multirow}
@@
 \begin{comment}
-\note{Lead: Choi. Volunteers welcome.
-\href{http://wssspe.researchcomputing.org.uk/wssspe2/lightning-talks/}{Slides.}}
-
-Lighting talks, breakout groups, working groups are all different. Lightning
-talks were new in WSSSPE2. They were offered to all paper contributors, and some
-other attendees, those who got travel awards. Breakout groups in WSSSPE1 were
-just discussions about topics; in WSSSPE2, the breakout groups were supposed to
-come up with plans for actions in specific areas. The idea of working groups is
-that they will come out of breakout groups as a way for the groups to actually
-do the actions over time, not at the workshops.
-
-Email addresses of speakers: gvwilson@software-carpentry.org, Colin Venters
-<C.Venters@hud.ac.uk>, j.spencer@imperial.ac.uk, erinrobinson@esipfed.org,
-marpierc@iu.edu, jwpeterson@gmail.com, abani@buffalo.edu, clenhardt@renci.org,
-dsk@uchicago.edu, Samin.Ishtiaq@microsoft.com, james@howison.name,
-s.a.harris@leeds.ac.uk, marcus.hanwell@kitware.com, thabermann@hdfgroup.org,
-rdowns@ciesin.columbia.edu, cboettig@gmail.com, jakob.blomer@cern.ch,
-editor@ascl.net
+\note{
+\href{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}{Slides.}}
 \end{comment}
 
-Lightning talks were a new feature in WSSSPE2. Since the workshop program was
-mostly dedicated to discussions, the organizers wanted to give the attendees a
-chance to also `make a pitch' for an idea, either representing a contributed
-paper or something different. Eighteen attendees volunteered to participate in
-the lightning talks, each given only two minutes to speak and at most one slide
-to show. The talks were presented in reverse alphabetic order of speakers' last
-names. In the rest of this section, we highlight the gist of some of the
-speakers' messages.
+\begin{enumerate}
+\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
+Facilitating Research Training While Supporting Scientific Research}}
 
-%\item Greg Wilson: Close Enough for Scientific Work
+\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
+Betz, Ruzanna Chitchyan, Letícia Duboc, Steve Easterbrook, Guillermo
+Rodriguez-Navas and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
+Sustainability}}
 
-\begin{enumerate}
-\item \textbf{Colin C. Venters: The Nebuchadnezzar Effect: Dreaming of Sustainable
-Software through Sustainable Software Architectures~\cite{Venters_poster}.}
-Venters proposed that sustainable software is a composite, first-class,
-non-functional requirement (NFR) that is at a minimum a measure of a system's
-maintainability and extensibility, but may also include other NFRs such as
-efficiency (e.g., energy, cost), interoperability, portability, reusability,
-scalability, and usability. To achieve technically sustainable software, Venters
-suggested that software architectures are fundamental as they are the primary
-carrier of system NFRs, i.e., pre-system understanding; and influence how
-developers are able to understand, analyze, extend, test, and maintain a
-software system, i.e., post-deployment system understanding. In addition,
-Venters highlighted that sustainability of software architectures needs to be
-addressed to endure different types of change and evolution in order to mitigate
-architectural drift, erosion, and knowledge vaporization.
+\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
+and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
+Improve Productivity of Legacy Tools}}
 
-%\item James Spencer: Developing New Developers
+\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins and Kaitlin
+Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
+from Open Source}}
 
-%\item Erin Robinson:  Sustaining Science Bazaars
+\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
+through Best Practices in Open Source Software: Computational Infrastructure
+for Geodynamics}}
 
-\item \textbf{Marlon Pierce: Patching It Up, Pulling It
-Forward~\cite{Pierce_poster}.} Pierce discussed how open open source is. Open
-software needs a diverse, openly governed community behind it, just as it needs
-open licensing and a public code repository. To probe the level of governance
-within open source projects, he and his co-authors (Marru and Mattmann)
-suggested a contest to encourage individual developers to submit
-patches and requests to projects that are important to them. This simple
-mechanism shall expose several governance mechanisms, such as how easy it is for
-independent developers to communicate with project leadership, how projects
-accept and license third-party contributions, and how projects make
-decisions such as granting source tree write access.
+\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
+Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
+Computational Infrastructure for Geodynamics}}
 
-\item \textbf{John Peterson: Continuous Integration for Concurrent MOOSE Framework
-and Application Development on GitHub~\cite{Peterson_poster}.}
-Peterson from the Idaho National Laboratory reported
-that in March 2014, the MOOSE framework was released under an open source
-license on GitHub, significantly expanding and diversifying the pool of current
-active and potential future contributors on the project. The MOOSE team employs
-an extensive continuous integration test suite to ensure that both the framework
-and the applications based on the framework are verified before any code changes
-are accepted into the repository. They use a combination of built-in Git
-features such as branching and submodules, GitHub API integration capabilities,
-and in-house developed testing software to perform this verification and update
-the dependent applications in a relatively seamless manner for users.
+\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
+Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for Open
+Science (DASPOS)}}
 
-\item \textbf{Abani Patra: Execute it~\cite{Patra_poster}.}
-%Univ at Buffalo, SUNY
-Patra discussed the value of an easily accessible platform for
-executing scientific software, e.g., HUBzero to access XSEDE or other computing
-resources. Such a platform for executing benchmark problems (even at a small
-scale) allows the developer community access a reference implementation and
-provides an easy way to train the larger user community. A second idea of this
-talk was that for true usability, much more attention and support needs
-to be given to the integrated use of simulation tools inside complex workflows.
+\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
+Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods and
+Susheel Varma: \textit{Research Software Engineering Groups in Universities: The Story
+from the UK}}
 
-%\item Christopher Lenhardt: Open Science for Synthesis (OSS): Filling the Gapin %Early Career Training
+\item \textbf{Dan Gunter, Sarah Poon and Lavanya Ramakrishnan: \textit{Bringing the
+User into Building Sustainable Software for Science}}
 
-\item \textbf{Daniel S. Katz: Implementing Transitive Credit with
-JSON-LD~\cite{Katz_transitive_credit_poster}.} Science and engineering research
-increasingly relies on activities that facilitate research but are not rewarded
-or recognized, such as: data sharing; developing common data resources, software
-and methodologies; and annotating data and publications. To promote and advance
-these activities, we must develop mechanisms for assigning credit, facilitate
-the appropriate attribution of research outcomes, devise incentives for
-activities that facilitate research, and allocate funds to maximize return on
-investment. Katz discussed the issue of assigning credit for both direct and
-indirect contributions by using JSON-LD to implement a prototype transitive
-credit system.
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and
+Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
+Biology: Lessons Learned on the KBase Project}}
 
-\item \textbf{Samin Ishtiaq: Daemons, Notifications and Sustaining Software. }%~\cite{Ishtiaq_poster}}
-%  Microsoft Research Cambridge
-The reproduction and replication of novel results has become a major issue in
-computer science, systems biology, and other computational disciplines. These
-include both the inability to re-implement novel algorithms and approaches, and
-lack of an agreement on how and what to benchmark these algorithms on. Ishtiaq
-from Microsoft Research Cambridge pointed out these problems and made several
-suggestions to address them.
+\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson and Karan
+Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
+Scientists in Best Practices of Software Sharing}}
 
-\item \textbf{James Howison: Retract all Bit-Rotten Publications. }%~\cite{Howison_poster}
-Howison sought to provoke discussion by proposing that papers whose workflows
-are not kept current with the changing software ecosystem should be
-automatically retracted. This would create an incentive for authors to keep
-their software current and usable, rather than the current situation in which
-every potential user has to do this individually. A softer version of the
-proposal would identify papers whose software workflow has become bit-rotten and
-allow others to keep the code up to date, either adding them as new authors of
-the paper or providing credit for their academic service in some other form.
+\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software Accreditation
+Framework}}
 
-%\item Sarah Harris: Sustainability of Multidisciplinary Software: A Science
-%Perspective
+\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
+Sustainable Scientic Software}}
 
-%\item Marcus D. Hanwell: We Need Tool Builders for Sustainable Scientific
-%Software
+\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
 
-%\item Ted Habermann: Communities as the Whole Product
+\item \textbf{Alice Allen: \textit{We’re giving away the store! (Merchandise not
+included)}}
 
-\item \textbf{Robert Downs: Community Recommendations for Improving Sustainable
-Scientific Software Practices~\cite{Downs_poster}.} Robert Downs, of the
-Columbia University Center for International Earth Science Information Network
-(CIESIN), described a focus group study conducted with the Science Software
-Cluster (SSC) of the Federation of Earth Science Information Partners (ESIP).
-For the study, almost 300 attendees of the 2014 Summer ESIP Meeting were invited
-to participate in simultaneous roundtable discussions on sustainability of
-science software. Over two-thirds of the roundtable focus groups responded to a
-semi-structured survey that contained three sets of questions eliciting
-recommendations for near-term actions of the community to improve sustainable
-software practices. Initial analysis of the participants' responses to the
-questionnaire revealed several suggestions, which included improving community
-engagement and collaborative activities, increasing understanding and awareness,
-and creating incentives to motivate sustainable science software practices. 
-The ESIP SSC plans to engage the community in the recommended activities for
-improving sustainable scientific software practices.
+\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
+Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
+Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
+Bo Zhang: \textit{Scientific Software Success: Developing Metrics While Developing
+Community}}
 
-\item \textbf{Carl Boettiger: rOpenSci: Building Sustainable Software by
-Fostering a Diverse Community~\cite{Boettiger_poster}.} Boettiger described how
-the rOpenSci project has been successful by focusing not just on building
-software but also on building a community of researchers who learn and adopt
-their approaches to reproducible research and sustainable software practice.
-Through outreach, mentoring, workshops, and hackathons, they have not only
-reached new users, but also turned users into co-developers of robust software
-and good practices to support data science research across a growing set of
-disciplines.
-
-\item \textbf{Jakob Blomer: The Need for a Versioned Data Analysis
-Environment~\cite{Blomer_poster}.} Large-scale scientific endeavors, such as the
-discovery of the Higgs boson at the Large Hadron Collider (LHC), often rely on
-complex software stacks. Maintaining thousands of dependencies of software
-libraries and operating system versions has shown that despite source code
-availability, the setup and the validation of a minimal usable analysis
-environment can easily become prohibitively expensive. In high-energy physics,
-CernVM-FS, a special-purpose, open-source, versioning, and snapshotting file
-system used to capture and distribute entire software stacks, proved to be
-useful for providing instant access to ready-to-run data analysis environments.
-% for the global scientific community as well as for
-%interested citizens, for instance through the CERN Open Data Portal
-%(http://opendata.cern.ch). \todo{revise}
-
-\item \textbf{Alice Allen: Find it! Cite it! }%~\cite{Allen_poster}}
-The Astrophysics Source Code Library (ASCL) is an online registry of
-scientist-written software used in astronomy research. Their primary interest is
-rendering research more transparent by making this software more discoverable for
-examination. The ASCL is treated as a publication by an indexing resource for
-astronomy, the Astrophysics Data System (ADS). ADS tracks citations to what it
-indexes, including citations to ASCL entries. Increasing rewards for writing
-software, whether through citation, transitive credit or other methods, gives
-software authors a powerful reason to take the time to build sustainability into
-their software and is an excellent way to drive community change.
 \end{enumerate}

### 2015-09-30 WSSSPE3/report/wg_appendix_publishing_SW.tex

@@
 The Software Sustainability Institute agreed to host this list.
+
+A tentative first cut at the list contains the following:
+\begin{enumeration}
+\item ACM Transactions on Mathematical Software (TOMS) - which provides the extra step
+ of having reviewers validate the code which was submitted with the publication.
+\item The Mathematica Journal - which publishes Mathematica notebooks (with equations,
+figures, etc.) directly.
+\item O'Reily Media has announced that it plans to make IPython Notebooks a first-class
+ authoring environment for their publishing program alongside their existing mechanisms.
+\item Nature is offering a list of notebooks published alongside more traditional articles,
+ and is looking at ways to make these documents more official. There are, in fact, a
+ number of journals that offer "electronic supplements" to the more traditionally published
+ static articles.
+\item There is also a list of "reproducible academic publications" maintained here:
+  \url{https://github.com/ipython/ipython/wiki/A-gallery-of-interesting-IPython-Notebooks#reproducible-academic-publications}
+\end{enumeration}
 

### 2015-09-30 WSSSPE3/report/wg_main_publishing_SW.tex
In progress working on main.
Former-commit-id: c9378e230ad21eee30b9b39952b47c052784b58a

\subsection{Publishing Software Working Group Discussion}

\subsubsection{Why it is important}

This group explored the value of executable papers (papers whose content includes
the code needed to produce their own results), and other forms of publishing which
include dynamic electronic content. Transitioning to this type of publication offers
possibilities of addressing, or partially addressing sustainability concerns 
such as reproducibility (the paper contains all the artifacts needed to verify its
results), transitive credit (modules an executable paper depends on must be explicitly
loaded, making it more feasible to identify them), and improving documentation (an executable
paper must explain what its code does).

Reproducibility: Part of the purpose of these venues is to (at least partially)
address the reproducibility issue by making the paper itself recompute its own
results.

Transitive Credit: Since these forms of publishing must make their sources explicit,
they should be easier to trace even if appropriately worded credit for software
is not provided. In addition, these notebooks make it possible to provide/define
additional metadata to make the tracing of credit clearer. In addition, attributions
could be added to citations to identify whether a paper extends a result, verifies it,
contradicts it, etc.

Best Practices: Because an executable paper showcases the code

\subsubsection{Fit with related activities}
\todo{short text here - can include links/cites}

\subsubsection{Discussion}
\todo{short-ish text here}

\subsubsection{Plans}
\todo{short text here - not bullets}

\subsubsection{Landing Page}

A page will be made available on the Software Sustainability Institute website.

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
adding stuff
Former-commit-id: 98823e9caf05f70abb8fd365541a163b57a9beb1
create participant list in latex from registrants.txt'
Former-commit-id: 60c7ece9b4ff817bc533c7a4237925958ac6cdfb

@@
 
+\include{wg_main_best_practices}
+\include{wg_main_catalogs}
+\include{wg_main_eng_design}
+\include{wg_main_funding_spec_expert}
+\include{wg_main_funding_spec_expert}
+\include{wg_main_industry_interaction}
+\include{wg_main_industry_interaction}
+\include{wg_main_legacy_SW}
+\include{wg_main_metrics}
+\include{wg_main_publishing_SW}
+\include{wg_main_software_credit}
+\include{wg_main_training}
+\include{wg_main_user_community}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{longtable}{ll}
-   Jordan Adams          &  Tulane University
-\\ Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Gabrielle Allen       & University of Illinois Urbana-Champaign
-\\ Pierre-Yves Aquilanti &  TOTAL E\&P R\&T USA
-\\ Wolfgang Bangerth & Texas A\&M University
-\\ David Bernholdt       &  Oak Ridge National Laboratory
-\\ Jakob Blomer
-\\ Carl Boettiger        &  University of California Santa Cruz \& rOpenSci
-\\ Chris Bogart          &  ISR/CMU
-\\ Steven R. Brandt      &  Louisiana State University
-\\ Neil Chue Hong        &  Software Sustainability Institute \& University of Edinburgh
-\\ Tom Clune             &  NASA GSFC
-\\ John W. Cobb
-\\ Dirk Colbry           &  Michigan State University
-\\ Karen Cranston        &  NESCent
-\\ Tom Crick             &  Cardiff Metropolitan University, UK
-\\ Ethan Davis           &  UCAR Unidata
-\\ Robert R Downs        &  CIESIN, Columbia University
-\\ Anshu Dubey           &  Lawrence Berkeley National Laboratory
-\\ Nicole Gasparini      &  Tulane University, New Orleans
-\\ Yolanda Gil           &  Information Sciences Institute, University of Southern California
-\\ Kurt Glaesemann       &  Pacific northwest national lab
-\\ Sol Greenspan         &  National Science Foundation
-\\ Ted Habermann         &  The HDF Group
-\\ Marcus D. Hanwell     &  Kitware
-\\ Sarah Harris          &  University of Leeds
-\\ David Henty           &  EPCC, The University of Edinburgh
-\\ James Howison         &  University of Texas
-\\ Maxime Hughes
-\\ Eric Hutton           &  University of Colorado
-\\ Ray Idaszak           &  RENCI/UNC
-\\ Samin Ishtiaq         &  Microsoft Research Cambridge, UK
-\\ Matt Jones            &  University of California Santa Barbara
-\\ Nick Jones            &  New Zealand eScience Infrastructure, University of Auckland
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\\ Ian Kelley
-\\ Hilmar Lapp           &  National Evolutionary Synthesis Center (NESCent) \& Duke University
-\\ Christopher Lenhardt
-\\ Richard Littauer      &  University of Saarland
-\\ Frank L\"{o}ffler     &  Louisiana State University
-\\ Andre Luckow          &  Rutgers
-\\ Berkin Malkoc         &  Istanbul Technical University
-\\ Kyle Marcus           &  University at Buffalo
-\\ Bryan Marker          &  The University of Texas at Austin
-\\ Suresh Marru          &  Indiana University
-\\ Robert H. McDonald    &  Data to Insight Center/Libraries, Indiana University
-\\ Rupert Nash
-\\ Andy Nutter-Upham     &  Whitehead Institute
-\\ Abani Patra           &  University at Buffalo
-\\ Aleksandra Pawlik     &  Software Sustainability Institute
-\\ Cody J. Permann       &  Idaho National Laboratory
-\\ John W. Peterson      &  Idaho National Laboratory
-\\ Benjamin Pharr        &  University of Mississippi
-\\ Stephen Piccolo       &  Brigham Young University, Utah
-\\ Marlon Pierce         &  Indiana University
-\\ Ray Plante            &  NCSA, University of Illinois Urbana-Champaign
-\\ Sushil Prasad         &  Georgia State University, Atlanta
-\\ Karthik Ram           &  Berkeley Institute for Data Science, University of California Berkeley \& rOpenSci
-\\ Mike Rilee            &  NASA/GSFC \& Rilee Systems Technologies
-\\ Erin Robinson         &  Foundation for Earth Science
-\\ Mark Schildhauer      &  NCEAS, Univ. California, Santa Barbara
-\\ Jory Schossau         &  Michigan State University
-\\ Frank Seinstra        &  Netherlands eScience Center
-\\ James Shepherd        &  Rice University
-\\ Justin Shi
-\\ Ardita Shkurti        &  University of Nottingham
-\\ Alan Simpson          &  EPCC, The University of Edinburgh
-\\ Carol Song            &  Purdue University
-\\ James Spencer         &  Imperial College London
-\\ Tracy Teal            &  Data Carpentry
-\\ Kaitlin Thaney        &  Mozilla Science Lab
-\\ Matt Turk             &  NCSA, University of Illinois Urbana-Champaign
-\\ Colin C. Venters      &  University of Huddersfield
-\\ Nathan Weeks
-\\ Ethan White           &  University of Florida/Utah State University
-\\ Nancy Wilkins-Diehr   &  San Diego Supercomputer Center, University of California San Diego
-\\ Greg Wilson           &  Software Carpentry
+\input{registrants}
 \end{longtable}

### 2015-09-30 WSSSPE3/report/registrants.tex
create participant list in latex from registrants.txt'
Former-commit-id: 60c7ece9b4ff817bc533c7a4237925958ac6cdfb

Alice Allen & Astrophysics Source Code Library\\
Gabrielle Allen & NCSA\\
Janine Aquino & UCAR/NCAR Earth Observing Laboratory\\
Steven Brandt & Louisiana State University\\
Jed Brown & CU Boulder\\
Matthias Bussonnier & UC Berkeley\\
Jeffrey Carver & University of Alabama\\
Emily Chen & NCSA\\
Sou Cheng Choi & NORC at UChicago / IIT\\
Nancy Collins & NCAR\\
Ethan Davis & UCAR Unidata\\
Davide DelVento & NCAR/CISL\\
Yuhan Ding & Illinois Institute of Technology\\
Tim Dunne & KnowInnovation \\
Ward Fisher & UCAR/Unidata\\
Sandra Gesing & University of Notre Dame\\
Josh Greenberg & Sloan Foundation\\
Dan Gunter & LBNL\\
Ted Habermann & The HDF Group\\
James Hetherington & University College London\\
Neil Chue Hong & Software Sustainability Institute\\
Elisabeth Huffer & Lingua Logica/NASA \\
Lorraine Hwang & UC Davis - CIG\\
Raymond Idaszak & RENCI; University of North Carolina at Chapel Hill\\
Elizabeth Jessup & University of Colorado Boulder\\
Nick Jones & New Zealand eScience Infrastructure (NeSI)\\
Daniel Katz & U Chicago & Argonne\\
Iain Larmour & EPSRC (UK)\\
Frank Löffler & Louisiana State University\\
Suresh Marru & Indiana University\\
Ryan May & UCAR/Unidata\\
Abigail Cabunoc Mayes & Mozilla Foundation\\
Jeff McWhirter & Geode Systems\\
Constantinos Michailidis & Knowinnovation\\
Don Middleton & NCAR\\
Mark Miller & SDSC\\
Pate Motter & University of Colorado\\
Jaroslaw Nabrzyski & University of Notre Dame\\
Patrick Nichols & National Center for Atmospheric Research\\
Kyle Niemeyer & Oregon State University\\
Laura Owen & NCSA\\
Abani Patra & Univ at Buffalo\\
Grace Peng & National Center for Atmospheric Research\\
Birgit Penzenstadler & California State University Long Beach\\
Lindsay Powers & The HDF Group\\
Bernie Randles & UCLA\\
Erin Robinson & Foundation for Earth Science\\
Daniel Sellars & CANARIE Inc\\
Nikolay Simakov & SUNY University at Buffalo\\
Ian Taylor & Cardiff University\\
Ilian Todorov & Science & Technology Facilities Council, UK\\
Benjamin Tovar & University of Notre Dame\\
Gregory Tucker & University of Colorado at Boulder\\
Matthew Turk & NCSA\\
Colin Venters & University of Huddersfield\\
Alexander Vyushkov & University of Notre Dame\\
Fraser Watson & National Solar Observatory\\
Nic Weber & University of Washington\\
Daniel Ziskin & NCAR - ACOM\\

### 2015-09-30 WSSSPE3/report/insert_registrants.py

@@ -1,0 +1,21 @@
+#!/usr/bin/env python
+import re, operator
+
+with open("registrants.txt", 'r') as f:
+  people = []
+  person = {}
+  for line in f:
+    if 'institution: ' in line:
+      person['institution'] = re.search('institution: (.*)', line).group(1).replace('&', r'\&')
+    elif 'email: ' in line:
+      person['email'] = re.search('email: (.*)', line).group(1)
+    elif len(line) > 2:
+      person['name'] = line[:-1].replace('&', r'\&')
+    else:
+      people.append(person)
+      person = {}
+if 'name' in person:
+  people.append(person)
+
+for person in sorted(people, key=lambda x: x['name'].split()[-1]):
+  print(person['name'] + ' & ' + person['institution'] + r'\\')

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
done with participants
Former-commit-id: f35b1beca166d8fb50eea687378e71b0a1f8f0af

@@
 
-\include{wg_main_best_practices}
-\include{wg_main_catalogs}
-\include{wg_main_eng_design}
-\include{wg_main_funding_spec_expert}
-\include{wg_main_funding_spec_expert}
-\include{wg_main_industry_interaction}
-\include{wg_main_industry_interaction}
-\include{wg_main_legacy_SW}
-\include{wg_main_metrics}
-\include{wg_main_publishing_SW}
-\include{wg_main_software_credit}
-\include{wg_main_training}
-\include{wg_main_user_community}
+\input{wg_main_best_practices}
+\input{wg_main_funding_spec_expert}
+\input{wg_main_catalogs}
+\input{wg_main_industry_interaction}
+\input{wg_main_legacy_SW}
+\input{wg_main_eng_design}
+\input{wg_main_metrics}
+\input{wg_main_training}
+\input{wg_main_software_credit}
+\input{wg_main_publishing_SW}
+\input{wg_main_user_community}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{need to update for WSSSPE3}
-The following is a partial list of workshop attendees who registered on the
-collaborative notes document~\cite{WSSSPE2-google-notes} that was used
-for shared note-taking at the meeting, or who participated in a breakout groups
-and were noted in that group's notes.
+The following is list of participants registered for the WSSSPE3 workshop.
 
@@
 
+\input{wg_appendix_best_practices}
+\input{wg_appendix_funding_spec_expert}
+\input{wg_appendix_catalogs}
+\input{wg_appendix_industry_interaction}
+\input{wg_appendix_legacy_SW}
+\input{wg_appendix_eng_design}
+\input{wg_appendix_metrics}
+\input{wg_appendix_training}
+\input{wg_appendix_software_credit}
+\input{wg_appendix_publishing_SW}
+\input{wg_appendix_user_community}
+
 \bibliographystyle{vancouver}

### 2015-09-30 WSSSPE3/report/wg_appendix_publishing_SW.tex
Update member affiliations.
Former-commit-id: 6d0b2140887bc8c5403be56880d5009b438a21cd
Fix compile problem.
Former-commit-id: 79bddccfbde22d98d3af9e93bcb9f9581bf36a81
Fleshed out main section.
Former-commit-id: 88d5f396cbab1b88397c525cf3d98b376faff1e3

@@
    Steven R. Brandt & Louisiana State University
-\\ Daniel Gunter    & 
-\\ Yuhan Ding       & 
+\\ Daniel Gunter    & LBNL
+\\ Yuhan Ding       & Illinois Institute of Technology
 \\ Neil Chue Hong   & Software Sustainability Institute
@@
 
-This group explored the value of executable papers (papers whose content includes
-the code needed to produce their own results), and other forms of publishing which
-include dynamic electronic content. Transitioning to this type of publication offers
-possibilities of addressing, or partially addressing sustainability concerns 
-such as reproducibility (the paper contains all the artifacts needed to verify its
-results), transitive credit (modules an executable paper depends on must be explicitly
-loaded, making it more feasible to identify them), and improving documentation (an executable
-paper must explain what its code does).
-
-While the group did not feel it had a way to influence what is currently happening
-in these experimental publishing venues, it felt that creating and curating a list of
-such efforts would have value by attracting interest to these activities.
-The Software Sustainability Institute agreed to host this list.
-
 A tentative first cut at the list contains the following:
-\begin{enumeration}
+\begin{itemize}
 \item ACM Transactions on Mathematical Software (TOMS) - which provides the extra step
@@
   \url{https://github.com/ipython/ipython/wiki/A-gallery-of-interesting-IPython-Notebooks#reproducible-academic-publications}
-\end{enumeration}
+\end{itemize}
 

### 2015-09-30 WSSSPE3/report/wg_main_publishing_SW.tex

@@
 possibilities of addressing, or partially addressing sustainability concerns 
-such as reproducibility (the paper contains all the artifacts needed to verify its
-results), transitive credit (modules an executable paper depends on must be explicitly
-loaded, making it more feasible to identify them), and improving documentation (an executable
-paper must explain what its code does).
+such as reproducibility, transitive credit, and best practices.
+
+\subsubsection{Fit with related activities}
 
@@
 
-Best Practices: Because an executable paper showcases the code
-
-\subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+Best Practices: Because an executable paper showcases the code, and the code itself
+is subject to the review process, authors are more likely to pay attention to coding
+practices. In addition, because the paper must explain what the code does, better
+documentation is likely to be achieved.
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
+
+The group felt that the best way to encourage the use of these new publishing
+concepts would be to create and curate a list identifying publishing venues
+that support them. The Software Sustainability Institute agreed to host this list.
 
 \subsubsection{Plans}
-\todo{short text here - not bullets}
+
+The plan is to create a web page describing executable papers, their value, and
+a list of what publishers support them.
 
@@
 
-A page will be made available on the Software Sustainability Institute website.
+The afforementioned page to be published on SSI.

### 2015-09-30 WSSSPE3/report/registrants.tex
fixing & issues
Former-commit-id: 82394150cd431397cff054c3cbcb0ec53c458527

@@
 Nick Jones & New Zealand eScience Infrastructure (NeSI)\\
-Daniel Katz & U Chicago & Argonne\\
+Daniel Katz & U Chicago \& Argonne\\
 Iain Larmour & EPSRC (UK)\\
@@
 Ian Taylor & Cardiff University\\
-Ilian Todorov & Science & Technology Facilities Council, UK\\
+Ilian Todorov & Science \& Technology Facilities Council, UK\\
 Benjamin Tovar & University of Notre Dame\\

### 2015-09-30 WSSSPE3/report/wg_appendix_publishing_SW.tex
Fleshed out appendix.
Former-commit-id: 003d4d20a8350b03d32864245f08748ce674c802

@@
 
-A tentative first cut at the list contains the following:
+A tentative first cut at the list containing executable papers identified the following:
+
 \begin{itemize}
@@
   \url{https://github.com/ipython/ipython/wiki/A-gallery-of-interesting-IPython-Notebooks#reproducible-academic-publications}
+\item KBase narratives build on IPython/Jupyter notebooks to build publications that are
+  reproducible, and can be commented or annotated.
+\end{itemize}
+
+The group also discussed future possibilities for this new publication format which might
+provide advantages.
+
+\begin{itemize}
+\item Journals could be built around an existing, widely-used framework thereby reducing
+  the burden of studying code on the part of reviewers (common bits of infrastructure
+  which aren't relevant to the science would be automatically excluded).
+\item Journals might be encouraged to use more metadata, making them easier to mine
+  for various analytical purposes.
+\item RIO is an effort to publish fragmentary results that can subsequently be combined
+  into a single content item.
+\item Papers could be made more understandable. Each equation or technical term could
+  offer a link to a document/tutorial explaining its origin and/or derivation.
+\item So many options for publication currently exist that good science may be getting
+  lost in the noise. Would some sort of "upvote" mechanism be of value?
+\item Some sort of Replicated Computation Results badge could be made available to
+  publications that have undergone greater scrutiny (this is already done by TOMS).
 \end{itemize}

### 2015-09-30 WSSSPE3/report/wg_appendix_software_credit.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Software Credit Working Group Discussion}
+\label{sec:appendix_SW_credit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+Kyle Niemeyer\footnote{email: \href{mailto:kyle.niemeyer@oregonstate.edu}{kyle.niemeyer@oregonstate.edu}} will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+\begin{itemize}
+\item Alice Allen -- Astrophysics Source Code Library
+\item Sou-Cheng Choi -- NORC at Univ.\ Chicago, Illinois Institute of Technology
+\item James Hetherington -- University College London
+\item Lorraine Hwang -- University of California, Davis
+\item Daniel S.\ Katz -- University of Chicago \& Argonne National Laboratory
+\item Frank Löffler -- Louisiana State University
+\item Abby Cabunoc Mayes -- Mozilla Science Lab
+\item Kyle E.\ Niemeyer -- Oregon State University
+\item Grace Peng -- National Center for Atmospheric Research
+\item Ilian Todorov -- Science \& Technology Facilities Council, UK
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Summary of Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Key Next Steps}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Plan for Future Organization}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{What Else is Needed?}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Key Milestones and Responsible Parties}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Description of Funding Needed}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-09-30 WSSSPE3/report/WSSSPE3_review.tex
for the moment include email addresses, but added note
Former-commit-id: 512da7c50731ee50269c56d6b76e1fc2e055e939
changes in first paragraph
Former-commit-id: 88b38b64d2e81f3f5c57f978ee82800ff04ba81b

@@
 
-%\note{google doc of notes for reference: \url{http://tinyurl.com/q6ew45v}}
-%https://docs.google.com/document/d/1-BxkYWDQ6nNNBXBStUL0xcKF9qCTlEALwf928J_MemI/edit?usp=sharing
-
 \todo{need to update for WSSSPE3}
-The Second Workshop on Sustainable Software for Science: Practice and
+The Third Workshop on Sustainable Software for Science: Practice and
 Experiences
-(WSSSPE2)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}} was
-held on 16 November, 2014 in the city of New Orleans, Louisiana, USA, in
-conjunction with the International Conference for High Performance Computing,
-Networking, Storage and Analysis
-(SC14)\footnote{\url{http://sc14.supercomputing.org}}. WSSSPE2 followed the
-model of a general initial workshop,
+(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
+held on 28-29 September 2015 in Boulder, Colorado, USA.
+Previous events in the WSSSPE series are
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-which co-occurred with SC13, and a focused workshop,
+held in conjunction with SC13, 
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
-which was organized in July 2014 jointly with the SciPy
-conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}.
+a focused workshop organized in July 2014 jointly with the SciPy
+conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}}, 
+WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
+held in conjunction with SC14, and
+WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}},
+a focused workshop organized in July 2015 again jointly with SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{this section is taken from \cite{WSSSPE2-pre-report}. It could be shortened.}
 
@@
 between them\footnote{See, for example, \url{http://tinyurl.com/l76tba2}.}.
-%http://www.slideshare.net/jameshowison/scientific-software-sustainability-and-ecosystem-complexity
-%(but does Anon Grizzly have other cites/info for that?).
-%In contrast, works licensed by Creative Commons typically do not
-%have dependencies (a book, a photo, an artwork).
 
@@
 nanoHUB\footnote{nanoHUB, \url{https://nanohub.org}}.
-%facilitates more efficient scientific research. SDI identifies scientific
-%software by archiving and standardizing metadata for software and hence help
-%connect both developers and users.
 Ted Habermann in the audience commented that if the metadata is minimal, it
@@
 how we reward researchers and support roles.
-%Lastly, she cautioned the community to be mindful of jargon or semantics traps.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\todo{Do we want email addresses here?}
 The following is list of participants registered for the WSSSPE3 workshop.
 
-{\small
-\begin{longtable}{ll}
+{\scriptsize
+\begin{longtable}{lll}
 \input{registrants}

### 2015-10-01 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Main body of the section added.

Former-commit-id: 2c37b96651f126371fc527ab84f64dbbd669e8e3

\subsection{Principles for Engineering Design for Sustainable Software} \todo{change the title to your topic}

\subsubsection{Why it is important}
\todo{short text here}

\subsubsection{Fit with related activities}
\todo{short text here - can include links/cites}

\subsubsection{Discussion}
\todo{short-ish text here}
This group discussed the principles of engineering design for sustainable software and their application in various domains. The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy. Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion. There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community. Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members. The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK):

Software design principles:
\begin{itemize}
\item Abstraction;
\item Coupling and cohesion;
\item Decomposition and modularization;
\item Encapsulation and information hiding;
\item Separation of interface and implementation;
\item Sufficiency completeness & primitiveness;
\item Separation of concerns.
\end{itemize}

User interface design principles:
\begin{itemize}
\item Learnability;
\item User familiarity;
\item Consistency;
\item Minimal surprise;
\item Recoverability;
\item User guidance;
\item User diversity.
\end{itemize}

The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design:
\begin{itemize}
\item Sustainability is systemic;
\item ...is multidimensional
\item ...is interdisciplinary;
\item ...transcends the system’s purpose;
\item ...applies to both a system and its wider contexts;
\item ...requires action on multiple levels;
\item ...requires multiple timescales;
\item Changing design to take into account long-term effects doesn’t automatically imply sacrifices;
\item System visibility is a precondition for and enabler of sustainability design.
\end{itemize}

This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing. The group identified two example systems to discuss the application of the principles. The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids. The second one was a modeling system for Malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets. They then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference. The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community. In addition, two challenges were pointed out: mapping of the principles to best practice, and  demonstrating the return on investment of those best practices.

\subsubsection{Plans}
The next steps in this endeavor are to (1) Systematically analyse a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.

\subsubsection{Landing Page}
\todo{link to landing page}

### 2015-10-01 WSSSPE3/report/wg_main_eng_design.tex
fixing tex errors, adding note about citation
Former-commit-id: 4241176b695b2d87e96b5e76d055070a2a6dbb81

@@
-\subsection{Principles for Engineering Design for Sustainable Software} \todo{change the title to your topic}
+\subsection{Principles for Engineering Design for Sustainable Software}
 
@@
 \item Separation of interface and implementation;
-\item Sufficiency completeness & primitiveness;
+\item Sufficiency completeness \& primitiveness;
 \item Separation of concerns.
@@
 
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design:
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design \todo{add reference}:
 \begin{itemize}
@@
 \item ...is interdisciplinary;
-\item ...transcends the system’s purpose;
+\item ...transcends the system's purpose;
 \item ...applies to both a system and its wider contexts;
@@
 \item ...requires multiple timescales;
-\item Changing design to take into account long-term effects doesn’t automatically imply sacrifices;
+\item Changing design to take into account long-term effects doesn't automatically imply sacrifices;
 \item System visibility is a precondition for and enabler of sustainability design.
@@
 
-This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
-The group identified two example systems to discuss the application of the principles.
-The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
-The second one was a modeling system for Malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
-They then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community.
-In addition, two challenges were pointed out: mapping of the principles to best practice, and  demonstrating the return on investment of those best practices.
+This congregated list is an initial collection of principles that could be extended by adding from further 
+related work form separate disciplines within the field of software engineering, including requirements 
+engineering, software architecture, and testing. The group identified two example systems to discuss 
+the application of the principles. The first one was a quantum chemistry system that allows the analysis 
+of the characteristics and capabilities of molecules and solids. The second one was a modeling system 
+for malaria that permitted biologists to analyze a range of datasets across geography, biology, and 
+epidemiology, and add their own datasets. They then examined the principles and took a retrospective 
+analysis of what the developers did in practice against how the principles could have made a 
+difference. The opportunity was identified to distill existing software engineering and sustainability 
+design knowledge into ``bite sized'' chunks for the Computational Science and Engineering 
+Community. In addition, two challenges were pointed out: mapping of the principles to best practice, 
+and  demonstrating the return on investment of those best practices.
 
 \subsubsection{Plans}
-The next steps in this endeavor are to (1) Systematically analyse a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
+The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
 

### 2015-10-06 WSSSPE3/report/wg_appendix_software_credit.tex
Software credit appendix in progress; added some relevant pubs
Former-commit-id: d4910da9ce59d01dbc76b0bd5c003886b00bda70

@@
 \item Alice Allen -- Astrophysics Source Code Library
-\item Sou-Cheng Choi -- NORC at Univ.\ Chicago, Illinois Institute of Technology
+\item Sou-Cheng Choi -- NORC at University of Chicago, Illinois Institute of Technology
 \item James Hetherington -- University College London
 \item Lorraine Hwang -- University of California, Davis
-\item Daniel S.\ Katz -- University of Chicago \& Argonne National Laboratory
+\item Daniel S.\ Katz -- University of Chicago, Argonne National Laboratory
 \item Frank Löffler -- Louisiana State University
@@
 
+The following section summarizes the working group's discussion based on contributions prior to the meeting~\cite{WSSSPE3-SC-github-issues} and the collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
+Please refer to the original sources for the unedited discussions.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+There currently is no standard/accepted mechanism for citing software or receiving credit for software (akin to publications/citations).
+Software is eligible for DOI assignment, but DOI metadata fields are not well tuned for software (vs. publications).
+Some software providers apply for DOIs but it is still not a widely adopted.
+Also, no mechanism to cite software dependencies within software.
+
+Obstacle: indexers (e.g., Scopus) don’t currently support software/data document types or DataCite DOIs.
+
+Obstacle: no standard mechanism for software to cite other software.
+
+Obstacle: for tenure, there is no standard for software products to be included in policy,
+sometimes not even within one university: hard to change globally.
 
@@
 
+\begin{enumerate}
+\item Hold virtual meeting to determine group members responsible\slash willing to work on the following tasks, to be organized within one month of the workshop
+\item Compile best practices of software citation across multiple disciplines, including journals and communities of interest\slash practice in the research world, to begin by December 2015
+\item Compile examples of including other products in promotion and tenure dossier, to begin by December 2015
+\item Draft the Software Citation Principles document (including citation metadata file), by April 2016
+\item Publish\slash release the Software Citation Principles document, by August 2016
+\item Reach out to journals, publishers, teachers\slash educators, indexers, and professional societies---likely through meetings with key groups, to begin by September 2016
+
+\end{enumerate}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+The WSSSPE breakout group plans to join efforts related to citing softare with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
+However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
+These activites include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo\footnote{Zenodo, \url{https://zenodo.org}}, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
+
+The group will primarily communicate electronically, with Kyle Niemeyer responsible for ensuring regular progress.
 
@@
 
+The near-term actions of the group, focused mainly on software citation, do not require any additional resources.
+However, connections with publishers and indexers will be needed to pursue related activities, although the FORCE11-SCWG may satisfy this need; in addition, some members of the group already reached out to relevant contacts.
+Funding may be needed to organize meetings or for group members to attend relevant meetings, as discussed further below.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
+Following the meeting, Kyle Niemeyer formally requested the merging of software citation activities with FORCE11-SCWG.
+Within a month of the meeting, Kyle will organize a virtual meeting of the group and lead the division of responsibilities for compiling existing practices of software citation and including software\slash products in promotion and tenure dossiers.
+Building off of these efforts, the next major milestone is drafting the Software Citation Principles document in collaboration with the SCWG, targeted for April 2016.
+While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the efforts of that group torwards the Software Citation Principles document, Kyle will help coordinate contributions from the WSSSPE group members.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Some funding would be useful to support primarily travel to conferences for group meetings (e.g., FORCE2016\footnote{FORCE2016, \url{https://www.force11.org/meetings/force2016}}), and to hold meetings to bring together both group members and key stakeholders (e.g., journals, publishers, professional societies, indexers).
+In addition, funding would be desired to support group members' time to perform work towards the key steps described previously.

### 2015-10-07 WSSSPE3/report/wg_appendix_software_credit.tex
minor changes and fixes
Former-commit-id: bd71f53fc5aed4fe0bbfe0eb6f5d0d1e5b6ec452

@@
 There currently is no standard/accepted mechanism for citing software or receiving credit for software (akin to publications/citations).
-Software is eligible for DOI assignment, but DOI metadata fields are not well tuned for software (vs. publications).
-Some software providers apply for DOIs but it is still not a widely adopted.
-Also, no mechanism to cite software dependencies within software.
+Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
+Some software providers apply for DOIs but it is still not widely adopted.
+Also, there is no mechanism to cite software dependencies within software.
 
-Obstacle: indexers (e.g., Scopus) don’t currently support software/data document types or DataCite DOIs.
+Obstacle: indexers (e.g., Scopus, Web of Science, Google Scholar) don't currently support software/data document types or DataCite DOIs.
 
@@
 
-The WSSSPE breakout group plans to join efforts related to citing softare with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
+The WSSSPE breakout group plans to join efforts related to citing software with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
 However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
-These activites include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo\footnote{Zenodo, \url{https://zenodo.org}}, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
+These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo\footnote{Zenodo, \url{https://zenodo.org}}, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
 
@@
 Building off of these efforts, the next major milestone is drafting the Software Citation Principles document in collaboration with the SCWG, targeted for April 2016.
-While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the efforts of that group torwards the Software Citation Principles document, Kyle will help coordinate contributions from the WSSSPE group members.
+While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the efforts of that group towards the Software Citation Principles document, Kyle will help coordinate contributions from the WSSSPE group members.
 

### 2015-10-07 WSSSPE3/report/WSSSPE3_review.tex
updates in intro mostly
Former-commit-id: 9ed8ed718e6be83f8a7e0c9a701e18e6de4dd5fc

@@
 \begin{abstract}
-\todo{need to update for WSSSPE3}
+\todo{need to update this abstract for WSSSPE3}
 This technical report records and discusses the Second Workshop on Sustainable
@@
 
-\todo{need to update for WSSSPE3}
 The Third Workshop on Sustainable Software for Science: Practice and
@@
 an imagined future practice for software development and usage in science and
-engineering. The workshop included multiple mechanisms for participation,
+engineering, but WSSSPE2 didn't have a good way to enact those mechanisms,
+or to encourage the attendees to follow through on their intentions.
+
+\todo{update this paragraph for WSSSPE3}
+The workshop included multiple mechanisms for participation,
 encouraged team building around solutions, and identified risky solutions with
@@
 early-career scientists, postdoctoral researchers, and graduate students, with
-funds provided to the conference organizers by the Moore Foundation and the
-National Science Foundation (NSF), to support the travel of potential
-participants who would not otherwise be able to attend, and young participants
-and those from underrepresented groups, respectively. These funds allowed 17
-additional participants to attend, and each was offered the chance to present a
-lightning talk.
+funds provided to the conference organizers by the Moore Foundation, the
+National Science Foundation (NSF), and the Software Sustainability Institute,
+to support the travel of potential
+participants who would not otherwise be able to attend and young participants
+and those from underrepresented groups. These funds allowed \todo{update}17
+additional participants to attend.
 
-This report extends a previous report that discussed the submission,
-peer-review, and peer-grouping processes in detail~\cite{WSSSPE2-pre-report}. It
-is also based on a collaborative set of notes taken with Google Docs during the
-workshop~\cite{WSSSPE2-google-notes}. Overall, the report discusses the
+This report is based on collaborative notes taken with during the
+workshop, which were linked from the GitHub issues that represented the
+potential and actual working groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}. 
+Overall, the report discusses the
 organization work done before the workshop (\S\ref{sec:preworkshop}); the
-keynotes (\S\ref{sec:keynotes}); a series of lightning talks
+keynotes (\S\ref{sec:keynote}); a series of lightning talks
 (\S\ref{sec:lightning}), intended to give an opportunity for attendees to
-quickly highlight an important issue or a potential solution; a session on
-defining sustainability (\S\ref{sec:defining}). The report also gives summaries
-of action plans proposed by five breakout sessions, which explored in specific
-areas including sustainability (\S\ref{sec:exploring}); software development
-experiences (\S\ref{sec:devel}); credit \& incentives (\S\ref{sec:credit});
-reproducibility, reuse, \& sharing (\S\ref{sec:reproduce}); code testing \& code
-review (\S\ref{sec:code_testing}). Lastly, the report also includes some
-conclusions (\S\ref{sec:conclusions}) and an incomplete list of attendees
-(Appendix~\ref{sec:attendees}).
+quickly highlight an important issue or a potential solution.
+The report also gives summaries
+of action plans proposed by the eleven working groups (\S\ref{sec:WGs}), and some
+conclusions (\S\ref{sec:conclusions}). Finally, the report includes a list of
+registered attendees (Appendix~\ref{sec:attendees}) and longer descriptions
+of the activities that occurred in each the working groups (Appendices~\ref{sec:appendix_best_practices}-\ref{sec:appendix_user_community}.)
 
@@
 
-\todo{need to update for WSSSPE3}
+\todo{need to update this section for WSSSPE3}
 WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the

### 2015-10-07 WSSSPE3/report/WSSSPE3_review.tex
more tweaks in intro section
Former-commit-id: 9f9e958b75387421aa64ca30005f23e91d8c0ce8

@@
 
-\todo{update this paragraph for WSSSPE3}
-The workshop included multiple mechanisms for participation,
-encouraged team building around solutions, and identified risky solutions with
-potentially transformative outcomes. It strongly encouraged participation of
+The workshop included multiple mechanisms for participation and 
+encouraged team building around solutions. It strongly encouraged participation of
 early-career scientists, postdoctoral researchers, and graduate students, with
@@
 to support the travel of potential
-participants who would not otherwise be able to attend and young participants
-and those from underrepresented groups. These funds allowed \todo{update}17
+participants who would not otherwise be able to attend, younger participants
+and those from underrepresented groups. These funds allowed 16
 additional participants to attend.
@@
 \todo{need to update this section for WSSSPE3}
-WSSSPE2 began with a call for papers~\cite{WSSSPE2-pre-report}. Based on the
+WSSSPE2 began with a call for papers. Based on the
 goal of encouraging a wide range of submissions from those involved in software

### 2015-10-09 WSSSPE3/report/wg_appendix_software_credit.tex
SC: added some discussion to appendix, other subsections completed
Former-commit-id: 2b2bb8d4d76e58b73579a1e60e21905b9c2c381b

@@
 Please refer to the original sources for the unedited discussions.
+
+Initial discussions focused on both various mechanisms for, and the philosophical approach behind, crediting software in scientific papers.
+These began with proposals for various ways to credit software (or other research products including data) that contributed more significantly than a generic citation, including
+\begin{itemize}
+    \item A hierarchy of citations, with a ``substantial'' citation category to indicate software or data that played a more significant role in the research;
+    \item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns contriponents (contributors and components) various weights depending on their level of importance; and
+    \item Project CRediT~\cite{projectcredit}, which assigns roles to paper authors based on their specific contributions; and
+    \item Mozilla Science Lab's recently introduced Contributorship Badges for Science, which provide a badge---associated with ORCID~\cite{orcid}---based on author contributions in a similar manner to Project CRediT.
+\end{itemize}
+However, as of this writing, only the Project CRediT roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship Badges~\cite{} have been implemented in published papers, though both of these only provide a single ``Software'' or ``Computation'' category associated with software.
+In addition, neither of these options allow for the citation of software itself, but only provide an author contribution related to software.
+The discussion quickly focused on transtitive credit as a more quantitative measure of allocating credit to both authors and software, although there were some concerns about authors overestimating their own contributions compared to prior work.
+
+The discussion then evolved into philosophical questions about the importantance or reliance of a particular work on prior science, materials, or software---in other words, is there a difference between depending on prior scientific advances and depending on particular software (or experimental equipment)?
+Multiple contributors converged on the conclusion that unique capabilities require some additional credit.
+The---albeit limited---consensus was that if a particular study relied on the unique capabilities of software, data, or an experimental apparatus, then the authors or developers that created this capability should be credited somehow.
+
+The group also agreed on the fact that additional data was required to support the assertion that software was not being sufficiently cited in the literature.
+In particular, this issue seemed to be field-dependent.
+For example, as shown by study of Howison and Bullard~\cite{Howison2015}, in the field of biology the most-cited papers appear to be describing scientific software.
+However, this may not---and likely is not---the case in other fields, nor is it clear whether developers of scientific software, even in the case of the biology field, are receiving sufficient credit for their efforts.
 

### 2015-10-12 WSSSPE3/report/wg_main_best_practices.tex

\subsection{White paper/journal paper about best practices in developing sustainable software}

\subsubsection{Why it is important}
\todo{short text here}

\subsubsection{Fit with related activities}
\todo{short text here - can include links/cites}

\subsubsection{Discussion}
\todo{short-ish text here}

\subsubsection{Plans}
\todo{short text here - not bullets}

[15 Nov] Introduction and scope finished
[15 Nov] Sections assigned
[31 Jan] Analysing funding possibilities for survey
[31 Jan] First versions of section
[15 Feb] Distribution to WSSSPE community
[31 Mar] Final version of white paper
[30 Apr] Submission of peer-reviewed paper?

\begin{enumerate}
\item Introduction and Scope of White Paper 
\item Related Work
\item Case Studies
\begin{enumerate} 
\item PeTSC
\item NWCHEM
\item CIG
\end{enumerate}
\item Community Related Practices
\begin{enumerate} 
\item Findings
\item Recommendations
\end{enumerate}
\item Governance and management
\begin{enumerate} 
\item Findings
\item Recommendations
\end{enumerate}
\item Funding Related
\begin{enumerate} 
\item Findings
\item Recommendations
\end{enumerate}
\item Metrics for sustainability
\item Tools
\item Conclusions
\end{enumerate}

\subsubsection{Landing Page}
\todo{link to landing page}

### 2015-10-12 WSSSPE3/report/wg_appendix_software_credit.tex
Removed large photos from repo history

@@
 In particular, this issue seemed to be field-dependent.
-For example, as shown by study of Howison and Bullard~\cite{Howison2015}, in the field of biology the most-cited papers appear to be describing scientific software.
+For example, as shown by study of Howison and Bullard~\cite{Howison2015}, in the field of biology the most-cited papers appear to be those describing scientific software.
 However, this may not---and likely is not---the case in other fields, nor is it clear whether developers of scientific software, even in the case of the biology field, are receiving sufficient credit for their efforts.
+
+During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} as a potential model for scientific software.
+That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
+One important use of this system is to track rights and credit for contributors to entertainment works in order to distrubute revenues---similar to the proposed transitive credit concept.
+
+The group also discussed separating quantitative measures (e.g., number of citations) from the value of a work in order to give credit, moving towards qualitative or anecdotal evidence of value.
+
+PageRank for citations, based on number of mentions
+
+market penetration or adoption rate in a community as a metric, although it was not clear how this would be measured.
+
+Finally, the concept a software tool's uniqueness or indispensability to a community was mentioned, with value being characterized by a particular piece of software either offering unique capabilities or doing something better, faster, or with less computational requirements than other offerings.
 

### 2015-10-12 WSSSPE1_report/install-deps.sh

@@ -1,1 +1,65 @@
-check-deps.sh
+# This is free and unencumbered software released into the public domain.
+#
+# Anyone is free to copy, modify, publish, use, compile, sell, or
+# distribute this software, either in source code form or as a compiled
+# binary, for any purpose, commercial or non-commercial, and by any
+# means.
+#
+# In jurisdictions that recognize copyright laws, the author or authors
+# of this software dedicate any and all copyright interest in the
+# software to the public domain. We make this dedication for the benefit
+# of the public at large and to the detriment of our heirs and
+# successors. We intend this dedication to be an overt act of
+# relinquishment in perpetuity of all present and future rights to this
+# software under copyright law.
+#
+# THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+# EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+# MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND NONINFRINGEMENT.
+# IN NO EVENT SHALL THE AUTHORS BE LIABLE FOR ANY CLAIM, DAMAGES OR
+# OTHER LIABILITY, WHETHER IN AN ACTION OF CONTRACT, TORT OR OTHERWISE,
+# ARISING FROM, OUT OF OR IN CONNECTION WITH THE SOFTWARE OR THE USE OR
+# OTHER DEALINGS IN THE SOFTWARE.
+#
+# For more information, please refer to <http://unlicense.org>
+
+#########################################################################
+# Determines the package dependencies for the .tex file given as the
+# first argument (or the first .tex file in a listing of the current
+# working directory, if no argument is provided), and lists for each
+# whether it is installed or not. If invoked with the name
+# install-deps.sh, it will also attempt to install packages that
+# aren't installed.
+#
+# Caveats: This script assumes a TeXLive installation, and uses tlmgr
+# to query installation status as well as to install a package. On the
+# first run you may see complaints that tlmgr itself needs to be
+# updated. Execute the command suggested in the error message to fix
+# this; this script will intentionally not it for you. Installation of
+# a package will invoke sudo, and will thus fail if you do not have
+# sudo privileges on the machine on which you are trying to do this.
+#########################################################################
+
[…]

### 2015-10-12 WSSSPE3/report/WSSSPE3_review.tex
draft of section 2

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Submissions, Peer-Review, and Peer-Grouping} \label{sec:preworkshop}
+\section{Calls for Participation} \label{sec:preworkshop}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\todo{need to update this section for WSSSPE3}
-WSSSPE2 began with a call for papers. Based on the
-goal of encouraging a wide range of submissions from those involved in software
-practice, ranging from initial thoughts and partial studies to mature
-deployments, but focusing on papers that are intended to lead to changes, the
-organizers wanted to make submission as easy as possible. The call for papers
-stated:
+WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting a process to make progress in sustainable software, as the calls for participation said:
 
-\begin{quote} We invite short (4-page) \textbf{actionable} papers that will lead
-to improvements for sustainable software science. These papers could be a call
-to action, or could provide position or experience reports on sustainable
-software activities. The papers will be used by the organizing committee to
-design sessions that will be highly interactive and targeted towards
-facilitating action. Submitted papers should be archived by a third-party
-service that provides DOIs. We encourage submitters to license their papers
-under a Creative Commons license that encourages sharing and remixing, as we
-will combine ideas (with attribution) into the outcomes of the workshop.
+ \begin{quote} The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas relevant to sustainable scientific software.
+WSSSPE2 invited the community to propose and discuss specific mechanisms to move towards an imagined future practice of software development and usage in science and engineering.
+WSSSPE3 will organize self-directed teams that will collaborate prior to and during the workshop to create vision documents, proposals, papers, and action plans that will help the scientific software community produce software that is more sustainable, including developing sustainable career paths for community members.
+These teams are intended to lead into working groups that will be active after the workshop, if appropriate, working collaboratively to achieve their goals, and seeking funding to do so if needed.
 \end{quote}
 
-The call included the following areas of interest:
+The first call for participation requested lightning talks, where authors could make a brief statement about work that either had been done or was needed, with the goal of contributing to the discussion of one or more working groups.
+There was 24 lightning talks submitted, and after a peer-review process, 16 of these were accepted, as discussed further in Section~\ref{sec:lightning}.
+
+The first call also discussed the potential action topics that came out of WSSSPE2, and requested additional suggestions.
+The combination of existing and new topics led to the following 18 potential topics that were advertised in the call for participation:
+
 \begin{quote}
@@
 \setlength{\rightmargin}{1em}
-\item defining software sustainability in the context of science and engineering
-software
+
+\item Development and Community
 \begin{itemize}
-\item how to evaluate software sustainability
+\item Writing a white paper/review paper about best practices in developing sustainable software
+\item Documenting successful models for funding specialist expertise in software collaborations
+\item Creating and curating catalogs for software tools that aid sustainability (perhaps categorized by domain, programming languages, architectures, and/or functions, e.g., for code testing, documentation)
+\item Documenting case studies for academia/industry interaction
+\item Determining effective strategies for refactoring/improving legacy scientific software
+\item Determining principles for engineering design for sustainable software
+\item Create a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges/pitfalls; then publish this as a white paper
 \end{itemize}
 
-\item improving the development process that leads to new software
+\item Training
 \begin{itemize}
-\item methods to develop sustainable software from the outset
-\item effective approaches to reusable software created as a by-product of
-research
-\item impact of computer science research on the development of scientific
-software
+\item Writing a white paper on training for developing sustainable software, and coordinating multiple ongoing training-oriented projects
+\item Developing curriculum for software sustainability, and ideas about where such curriculum would be presented, such as a summer training institute
 \end{itemize}
 
-\item recommendations for the support and maintenance of existing software
+\item Credit
 \begin{itemize}
-\item software engineering best practices
-\item governance, business, and sustainability models
-\item the role, operation, and
-sustainability of community software repositories
-\item reproducibility, transparency needs that may be unique to science
+\item Hacking the credit and citation ecosystem (making it work, or work better, for software)
+\item Developing a taxonomy of contributorship/guidelines for including software contributions in tenure review
+\item Documenting case studies of receiving credit for software contributions
+\item Developing a system of awards and recognitions to encourage sustainable software
 \end{itemize}
 
-\item successful open source software implementations
+\item Publishing
 \begin{itemize}
-\item incentives for using and contributing to open source software
-\item transitioning users into contributing developers
+\item Developing a categorization of journals that publish software papers (building on existing work), and case studies of alternative publishing mechanisms that have been shown to improve software discoverability/reuse e.g., popular blogs/websites
+\item Determining what journals that publish software paper should provide to their reviewers (e.g., guidelines, mechanisms, metadata standards, etc.)
 \end{itemize}
 
-\item building large and engaged user communities
+\item Reproducibility and Testing
 \begin{itemize}
-\item developing strong advocates
-\item measurement of usage and impact
+\item Building a toolkit that could allow conference organizers to easily add a reproducibility track
+\item Documenting best practices for code testing and code review
 \end{itemize}
 
-\item encouraging industry's role in sustainability
+\item Documentation
 \begin{itemize}
-\item engagement of industry with volunteer communities
-\item incentives for industry
-\item incentives for community to contribute to industry-driven projects
-\end{itemize}
-
-\item recommending policy changes
-\begin{itemize}
-\item software credit, attribution, incentive, and reward
-\item issues related to multiple organizations and multiple countries, such as
-intellectual property, licensing, etc.
-\item mechanisms and venues for publishing software, and the role of publishers
-\end{itemize}
-
-\item improving education and training
-\begin{itemize}
-\item best practices for providing graduate students and postdoctoral
-researchers in domain communities with sufficient training in software
-development
-\item novel uses of sustainable software in education (K-20)
-\item case studies from students on issues around software development in the
-undergraduate or graduate curricula
-\end{itemize}
-
-\item careers and profession
-\begin{itemize}
-\item successful examples of career paths for developers
-\item institutional changes to support sustainable software such as promotion
-and tenure metrics, job categories, etc.
+\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the community to easily find up-to-date information on a WSSSPE topic (e.g., software credit, scientific software metrics, testing scientific software)
 \end{itemize}
@@
 \end{itemize}
-
 \end{quote}
-
-31 submissions were received; all but one used arXiv\footnote{\url{http://arxiv.org}}
-or figshare\footnote{\url{http://figshare.com}} to self-publish their papers.
-
-The review process was fairly standard. First, reviewers bid for papers. Then an
-automated system matched the bids to determine assignments. After the reviewers
-completed their assigned reviews (with an average of 4.9 reviews per paper and
-4.1 reviews per reviewer), they used EasyChair\footnote{\url{http://easychair.org/}} 
-to record scores on relevance
-and comments. Finally, the organizers accessed the information to decide which
-papers to associate with the workshop and provided authors with the comments to
-help them improve their papers.
-
-The organizers decided to list 28 of the papers as significantly contributing to
-the workshop, a very high acceptance rate, but one that is reasonable, given the
-goal of broad participation and the fact that the reports were already
-self-published.
-
-The organizers wanted very interactive sessions, with the process of creating
-the sessions open to the full program committee, the paper authors, and others
-who might attend the workshop. In order to do this, the organizers used 
-Well Sorted\footnote{\url{http://www.well-sorted.org}} with the following steps:
-\begin{enumerate}
-%
-\item Authors were asked to create Well Sorted ``cards'' for the papers. These
-cards have a title (50 characters maximum) and a body (255 characters maximum).
-%
-\item Authors, members of the WSSSPE program committee, and mailing list subscribers
-were asked to sort the cards. Each person dragged the cards, one by one, into
-groups. A group could have as many cards as the person wanted it to have, and it
-could have any meaning that made sense to that person.
-%
-\item Well Sorted produced a set of averages of all the sorts, with
-various numbers of card clusters.
-%
-\end{enumerate}
-
-The organizers then chose a sort that contained five groups that felt most
-meaningful. After that, they decided on names for the five groups:
-\begin{itemize}
-\item Exploring Sustainability
-\item Software Development Experiences
-\item Credit \& Incentives
-\item Reproducibility \& Reuse \& Sharing
-\item Code Testing \& Code Review.
-\end{itemize}
-
-Finally, since some of the papers were not represented by cards in the process,
-they were not placed in groups by the peer-grouping system. The authors of
-these papers were asked which groups seemed the best for their papers; these
-papers were then placed in those groups. Sections~\ref{sec:exploring}-\ref{sec:code_testing}
-discuss the breakout groups, including a list of the papers associated with each
-group.
 

### 2015-10-19 WSSSPE3/report/WSSSPE3_review.tex
adding a crossref and removing a comment
Add TODO notes. Removed old content.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{need to update for WSSSPE3}
-The workshop featured two keynote addresses. In the opening keynote
-presentation, Kaitlin Thaney of the Mozilla Science Lab talked about her
-organization's work and policy to enable and support sustainable and
-reproducible scientific research through the open web. The second keynote
-speaker was Neil Chue Hong of Software Sustainability Institute. He shed
-light on how scientific software is prevalently driving advances in many science
-and engineering fields. Both keynote speeches spawned further discussion among
-workshop participants on the crucial notion of \emph{software sustainability}
-in the theme of our workshop.
-
-Kaitlin Thaney is the Director of the Mozilla Science Lab (hereafter Mozilla), which
-is a non-profit organization interested in openness, news, website
-creation, and Science, all taking advantage of the open web.
-
-Thaney started noting the unfortunate fact that many current systems suffer the
-unintended consequence of creating friction that hinders users, despite
-designers' original purposes to do good. An example is the National Cancer
-Institute's caBIG. A total of \$350 million was spent, including more than \$60
-million for management. More than $70$ tools were created, but caBIG is still
-seen as a failure\footnote{Report Blasts Problem-Plagued Cancer Research Grid,
-\url{http://tinyurl.com/maf6dz2}}. Those that had the least investment were the
-most used; the most invested software were the least utilized.
-
-Thaney emphasized that for efficient reproducible open research, we would need
-research tools (e.g., software repositories), social capital (e.g., incentives),
-and capacity (e.g., training and mentorship). Our systems would need to
-communicate with each other. A point was made by a member of the audience that
-as systems become less monolithic, it often becomes harder to sustain the links
-between them\footnote{See, for example, \url{http://tinyurl.com/l76tba2}.}.
-
-Thaney spoke about Mozilla's work around code citation, through a collaboration
-and prototype crafted between Mozilla, GitHub, figshare, and Zenodo. This work
-was presented at a closed meeting in May 2014 at the National Institutes of
-Health (NIH) around these issues, sparking a conversation from that meeting
-around what a \emph{Software Discovery
-Index}\footnote{Software Discovery Index, \url{http://softwarediscoveryindex.org/report/}} might look
-like. The meeting included a number of publishers, researchers, and those behind
-major scientific software efforts such as
-Bioconductor\footnote{Bioconductor, \url{http://www.bioconductor.org}},
-Galaxy\footnote{Galaxy, \url{http://galaxyproject.org}}, and
-nanoHUB\footnote{nanoHUB, \url{https://nanohub.org}}.
-Ted Habermann in the audience commented that if the metadata is minimal, it
-would be less onerous for data providers, but more burdensome for users---it
-could be challenging to keep a balance between what have to be captured and what
-would be ideal if we do not want to lose user engagement as in the case of the
-old Harvard Dataverse, finding often only the first four fields of three pages
-of metadata were filled out.
-
-The speaker concluded her talk urging the audience to design scientific software
-with the general community, not an individual, in mind; and to design to unlock
-latent potential of our systems. In addition, she encouraged everyone to rethink
-how we reward researchers and support roles.
+\todo{need to update for WSSSPE3. 
+Lead: S.-C. Choi. 
+PDF slides at \url{http://tinyurl.com/qbbqgsj}.
+Youtube Video at \url{http://tinyurl.com/q45kfcn}.
+}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{Do we want email addresses here?}
+%\todo{Do we want email addresses here?}
 The following is list of participants registered for the WSSSPE3 workshop.

### 2015-10-19 WSSSPE3/report/wg_appendix_publishing_SW.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Added myself as POC

@@
 
+Steven R. Brandt\footnote{email: \href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}} will serve as the point of contact for this working group.
+
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Steven R. Brandt & Louisiana State University
-\\ Daniel Gunter    & LBNL
-\\ Yuhan Ding       & Illinois Institute of Technology
-\\ Neil Chue Hong   & Software Sustainability Institute
-\end{longtable}
-}
+
+\begin{itemize}
+\item Steven R. Brandt -- Louisiana State University
+\item Daniel Gunter -- LBNL
+\item Yuhan Ding -- Illinois Institute of Technology
+\item Neil Chue Hong -- Software Sustainability Institute
+\end{itemize}
 

### 2015-10-20 WSSSPE3/report/WSSSPE3_review.tex
Keynote: intro.
Keynote: intro revised.

@@
 
+A highlight of WSSSPE3 was the keynote speech delivered by Professor Matthew
+Turk from the Department of Astronomy, University of Illinois on ``\emph{Why
+Sustain Scientific Software?}''. Turk is a prolific practioner of scientific
+software and has extensive experiences working on large collaborative projects
+employing modern computing tools. He is also a co-organizer and champion of all
+the WSSSPE conferences so far.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
small edits in keynote text

--- 2015-10-19 WSSSPE3/report/WSSSPE3_review.tex
@@
 
+WSSSPE3 began with a keynote speech delivered by Professor Matthew
+Turk from the Department of Astronomy, University of Illinois titled \emph{Why
+Sustain Scientific Software?}. Turk is a prolific practitioner of scientific
+software and has extensive experiences working on large collaborative projects
+employing modern computing tools. He is also a co-organizer and champion of all
+the WSSSPE events so far.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Sections 1-2: line breaks. Section 3. Recap of WSSSPE workshops.

@@
 
-The Third Workshop on Sustainable Software for Science: Practice and
-Experiences
+The Third Workshop on Sustainable Software for Science: Practice and Experiences
 (WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
-held on 28-29 September 2015 in Boulder, Colorado, USA.
-Previous events in the WSSSPE series are
+held on 28-29 September 2015 in Boulder, Colorado, USA. Previous events in the
+WSSSPE series are
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
 held in conjunction with SC13,
-WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}},
-a focused workshop organized in July 2014 jointly with the SciPy
+WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
+focused workshop organized in July 2014 jointly with the SciPy
 conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}},
@@
 held in conjunction with SC14, and
-WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}},
-a focused workshop organized in July 2015 again jointly with SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
+WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
+focused workshop organized in July 2015 again jointly with
+SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
 
@@
 an imagined future practice for software development and usage in science and
-engineering, but WSSSPE2 didn't have a good way to enact those mechanisms,
-or to encourage the attendees to follow through on their intentions.
+engineering, but WSSSPE2 didn't have a good way to enact those mechanisms, or to
+encourage the attendees to follow through on their intentions.
 
-The workshop included multiple mechanisms for participation and 
-encouraged team building around solutions. It strongly encouraged participation of
-early-career scientists, postdoctoral researchers, and graduate students, with
-funds provided to the conference organizers by the Moore Foundation, the
-National Science Foundation (NSF), and the Software Sustainability Institute,
-to support the travel of potential
-participants who would not otherwise be able to attend, younger participants
-and those from underrepresented groups. These funds allowed 16
-additional participants to attend.
+The workshop included multiple mechanisms for participation and encouraged team
+building around solutions. It strongly encouraged participation of early-career
+scientists, postdoctoral researchers, and graduate students, with funds provided
+to the conference organizers by the Moore Foundation, the National Science
+Foundation (NSF), and the Software Sustainability Institute, to support the
+travel of potential participants who would not otherwise be able to attend,
+younger participants and those from underrepresented groups. These funds allowed
+16 additional participants to attend.
 
-This report is based on collaborative notes taken with during the
-workshop, which were linked from the GitHub issues that represented the
-potential and actual working groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}. 
-Overall, the report discusses the
-organization work done before the workshop (\S\ref{sec:preworkshop}); the
-keynotes (\S\ref{sec:keynote}); a series of lightning talks
-(\S\ref{sec:lightning}), intended to give an opportunity for attendees to
-quickly highlight an important issue or a potential solution.
-The report also gives summaries
-of action plans proposed by the eleven working groups (\S\ref{sec:WGs}), and some
-conclusions (\S\ref{sec:conclusions}). Finally, the report includes a list of
-registered attendees (Appendix~\ref{sec:attendees}) and longer descriptions
-of the activities that occurred in each the working groups (Appendices~\ref{sec:appendix_best_practices}-\ref{sec:appendix_user_community}.)
+This report is based on collaborative notes taken with during the workshop,
+which were linked from the GitHub issues that represented the potential and
+actual working
+groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
+Overall, the report discusses the organization work done before the workshop
+(\S\ref{sec:preworkshop}); the keynotes (\S\ref{sec:keynote}); a series of
+lightning talks (\S\ref{sec:lightning}), intended to give an opportunity for
+attendees to quickly highlight an important issue or a potential solution. The
+report also gives summaries of action plans proposed by the eleven working
+groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
+Finally, the report includes a list of registered attendees
+(Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
+occurred in each the working groups
+(Appendices~\ref{sec:appendix_best_practices}-\ref{sec:appendix_user_community}.)
 
@@
 
-WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting a process to make progress in sustainable software, as the calls for participation said:
+WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting
+a process to make progress in sustainable software, as the calls for
+participation said:
 
- \begin{quote} The WSSSPE1 workshop engaged the broad scientific community to identify challenges and best practices in areas relevant to sustainable scientific software.
-WSSSPE2 invited the community to propose and discuss specific mechanisms to move towards an imagined future practice of software development and usage in science and engineering.
-WSSSPE3 will organize self-directed teams that will collaborate prior to and during the workshop to create vision documents, proposals, papers, and action plans that will help the scientific software community produce software that is more sustainable, including developing sustainable career paths for community members.
-These teams are intended to lead into working groups that will be active after the workshop, if appropriate, working collaboratively to achieve their goals, and seeking funding to do so if needed.
-\end{quote}
+\begin{quote} The WSSSPE1 workshop engaged the broad scientific community to
+identify challenges and best practices in areas relevant to sustainable
+scientific software. WSSSPE2 invited the community to propose and discuss
+specific mechanisms to move towards an imagined future practice of software
+development and usage in science and engineering. WSSSPE3 will organize
+self-directed teams that will collaborate prior to and during the workshop to
+create vision documents, proposals, papers, and action plans that will help the
+scientific software community produce software that is more sustainable,
+including developing sustainable career paths for community members. These teams
+are intended to lead into working groups that will be active after the workshop,
+if appropriate, working collaboratively to achieve their goals, and seeking
+funding to do so if needed. \end{quote}
 
-The first call for participation requested lightning talks, where authors could make a brief statement about work that either had been done or was needed, with the goal of contributing to the discussion of one or more working groups.
-There was 24 lightning talks submitted, and after a peer-review process, 16 of these were accepted, as discussed further in Section~\ref{sec:lightning}.
+The first call for participation requested lightning talks, where authors could
+make a brief statement about work that either had been done or was needed, with
+the goal of contributing to the discussion of one or more working groups. There
+was 24 lightning talks submitted, and after a peer-review process, 16 of these
+were accepted, as discussed further in Section~\ref{sec:lightning}.
 
-The first call also discussed the potential action topics that came out of WSSSPE2, and requested additional suggestions.
-The combination of existing and new topics led to the following 18 potential topics that were advertised in the call for participation:
+The first call also discussed the potential action topics that came out of
+WSSSPE2, and requested additional suggestions. The combination of existing and
+new topics led to the following 18 potential topics that were advertised in the
+call for participation:
 
@@
 \begin{itemize}
-\item Writing a white paper/review paper about best practices in developing sustainable software
-\item Documenting successful models for funding specialist expertise in software collaborations
-\item Creating and curating catalogs for software tools that aid sustainability (perhaps categorized by domain, programming languages, architectures, and/or functions, e.g., for code testing, documentation)
+\item Writing a white paper/review paper about best practices in developing
+sustainable software
+\item Documenting successful models for funding specialist expertise in software
+collaborations
+\item Creating and curating catalogs for software tools that aid sustainability
+(perhaps categorized by domain, programming languages, architectures, and/or
+functions, e.g., for code testing, documentation)
 \item Documenting case studies for academia/industry interaction
-\item Determining effective strategies for refactoring/improving legacy scientific software
+\item Determining effective strategies for refactoring/improving legacy
+scientific software
 \item Determining principles for engineering design for sustainable software
-\item Create a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges/pitfalls; then publish this as a white paper
+\item Create a set of guidance giving examples of specific metrics for the
+success of scientific software in use, why they were chosen, what they are
+useful to measure, and any challenges/pitfalls; then publish this as a white
+paper
 \end{itemize}
@@
 \begin{itemize}
-\item Writing a white paper on training for developing sustainable software, and coordinating multiple ongoing training-oriented projects
-\item Developing curriculum for software sustainability, and ideas about where such curriculum would be presented, such as a summer training institute
+\item Writing a white paper on training for developing sustainable software, and
+coordinating multiple ongoing training-oriented projects
+\item Developing curriculum for software sustainability, and ideas about where
+such curriculum would be presented, such as a summer training institute
 \end{itemize}
@@
 \begin{itemize}
-\item Hacking the credit and citation ecosystem (making it work, or work better, for software)
-\item Developing a taxonomy of contributorship/guidelines for including software contributions in tenure review
+\item Hacking the credit and citation ecosystem (making it work, or work better,
+for software)
+\item Developing a taxonomy of contributorship/guidelines for including software
+contributions in tenure review
 \item Documenting case studies of receiving credit for software contributions
@@
 \begin{itemize}
-\item Developing a categorization of journals that publish software papers (building on existing work), and case studies of alternative publishing mechanisms that have been shown to improve software discoverability/reuse e.g., popular blogs/websites
-\item Determining what journals that publish software paper should provide to their reviewers (e.g., guidelines, mechanisms, metadata standards, etc.)
+\item Developing a categorization of journals that publish software papers
+(building on existing work), and case studies of alternative publishing
+mechanisms that have been shown to improve software discoverability/reuse e.g.,
+popular blogs/websites
+\item Determining what journals that publish software paper should provide to
+their reviewers (e.g., guidelines, mechanisms, metadata standards, etc.)
 \end{itemize}
@@
 \begin{itemize}
-\item Building a toolkit that could allow conference organizers to easily add a reproducibility track
+\item Building a toolkit that could allow conference organizers to easily add a
+reproducibility track
 \item Documenting best practices for code testing and code review
@@
 \begin{itemize}
-\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the community to easily find up-to-date information on a WSSSPE topic (e.g., software credit, scientific software metrics, testing scientific software)
+\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the
+community to easily find up-to-date information on a WSSSPE topic (e.g.,
+software credit, scientific software metrics, testing scientific software)
 \end{itemize}
@@
 the WSSSPE events so far.
+
+In his keynote address, Turk recapped the course of developments of WSSSPE
+workshops alongside with his career over the past few years. The first WSSSPE
+workshop was at SuperComputing in 2013, but the notion of sustainable scientific
+software drew in an audience beyond supercomputing. A year later WSSSPE1.1 at
+SciPy had speakers talking about how software has been sustained inside the
+scientific Python community. WSSSPE2 at SuperComputing in 2014 had breakout
+group discussions coming up with actionable items, and WSSSPE2.1 at SciPy 2015
+was similar. Turk noted that WSSSPE3 broke free of SuperComputing this time. He
+noted the different atmosphere and perspectives of the surrounding large
+conferences, despite similar WSSSPE participants.
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Keynote: recap continued.

--- 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
@@
 the WSSSPE events so far.
+
+In his keynote address, Turk recapped the course of developments of WSSSPE
+workshops alongside with his career over the past few years. The first WSSSPE
+workshop was at the Supercomputing Conference (SC13) in 2013, but he observed
+that the notion of sustainable scientific software drew in an audience beyond
+supercomputing. In the following year, WSSSPE1.1 at SciPy had speakers talking
+about how software has been sustained inside the scientific Python community.
+WSSSPE2 at SC14 had breakout group discussions coming up with actionable items,
+and WSSSPE2.1 at SciPy~2015 was similar. Turk noted the different atmosphere and
+perspectives of the surrounding large conferences, despite similar WSSSPE
+participants. WSSSPE3 broke free of the Supercomputing Conference environment
+this time, and in Turk's words, it spoke to the fact that scientific software
+comes from many different types of inquiries, deployment, strategies for
+maintenance, users, and ways of measuring the value of a piece of software.
+
+%preconceptions
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Keynote: fix typos and add more details.

--- 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
@@
 the WSSSPE events so far.
+
+In his keynote address, Turk recapped the course of development of WSSSPE
+workshops alongside with his career from a postdoc to an academic over the past
+few years. The first WSSSPE workshop was at the Supercomputing Conference (SC13)
+in 2013, but he observed that the notion of sustainable scientific software drew
+in an audience beyond supercomputing. In the following year, WSSSPE1.1 at SciPy
+had speakers talking about how software has been sustained inside the scientific
+Python community. WSSSPE2 at SC14 had breakout group discussions coming up with
+actionable items, and WSSSPE2.1 at SciPy 2015 was similar. Turk noted the
+different atmosphere and approaches of the surrounding large conferences,
+despite similar WSSSPE participants. WSSSPE3 broke free of the Supercomputing
+Conference environment this time, and in Turk's words, it spoke to the fact that
+scientific software comes from many different types of inquiries, deployment,
+strategies for maintenance, users; and ways of measuring the value of a piece of
+software. It appeared to Turk that the supercomputing conferences generally
+adopt some top-down approaches, whereas SciPy bottom-up systems; the essential
+messages perceived were also often circumpolar: software getting harder in
+supercomputing communities, but software becoming better in the SciPy groups.
+
+%preconceptions
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Keynote: fix typos and add more details.

@@
 actionable items, and WSSSPE2.1 at SciPy 2015 was similar. Turk noted the
-different atmosphere and approaches of the surrounding large conferences,
-despite similar WSSSPE participants. WSSSPE3 broke free of the Supercomputing
-Conference environment this time, and in Turk's words, it spoke to the fact that
-scientific software comes from many different types of inquiries, deployment,
-strategies for maintenance, users; and ways of measuring the value of a piece of
-software. It appeared to Turk that the supercomputing conferences generally
-adopt some top-down approaches, whereas SciPy bottom-up systems; the essential
-messages perceived were also often circumpolar: software getting harder in
-supercomputing communities, but software becoming better in the SciPy groups.
+different atmosphere of the surrounding large conferences, despite similar
+WSSSPE participants. WSSSPE3 broke free of the Supercomputing Conference
+environment this time, and in Turk's words, it spoke to the fact that scientific
+software comes from many different types of inquiries, deployment, strategies
+for maintenance, users; and ways of measuring the value of a piece of software.
+It appeared to Turk that the supercomputing conferences generally adopt some
+top-down approaches, whereas SciPy bottom-up systems; the essential messages
+perceived were also often circumpolar: software getting harder in supercomputing
+communities with exascale computing in mind, but software becoming better in the
+SciPy groups with new tools such as Jupiter and packages for research workflow.
+Admitting such comparisons are somewhat unfair generalization, Turk reminded the
+audience that the different approaches bring different types of ideas to the
+table, but he welcomed that WSSSPE3 was outside existing preconceptions.
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Removed a comment.
Section 3 Keynote: a new paragraph.

@@
 different atmosphere of the surrounding large conferences, despite similar
-WSSSPE participants. WSSSPE3 broke free of the Supercomputing Conference
-environment this time, and in Turk's words, it spoke to the fact that scientific
+WSSSPE participants.
+
+WSSSPE3 broke free of the traditional Supercomputing Conference
+environment this year, and in Turk's words, it spoke to the fact that scientific
 software comes from many different types of inquiries, deployment, strategies
@@
 perceived were also often circumpolar: software getting harder in supercomputing
-communities with exascale computing in mind, but software becoming better in the
-SciPy groups with new tools such as Jupiter and packages for research workflow.
+communities with exascale computing and optimization issues in mind, but software becoming better in the
+SciPy groups with emerging tools such as Jupiter and productivity packages for research workflow.
 Admitting such comparisons are somewhat unfair generalization, Turk reminded the
@@
 table, but he welcomed that WSSSPE3 was outside existing preconceptions.
-
-%preconceptions
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
Section 3 Keynote: minor changes.

@@
 audience that the different approaches bring different types of ideas to the
-table, but he welcomed that WSSSPE3 was outside existing preconceptions.
+table, but he welcomed that WSSSPE3 being conducted outside existing preconceptions.
 

### 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
edits and comment in keynote section

--- 2015-10-21 WSSSPE3/report/WSSSPE3_review.tex
@@
 workshops alongside with his career from a postdoc to an academic over the past
-few years. The first WSSSPE workshop was at the Supercomputing Conference (SC13)
+few years. The first WSSSPE workshop was at the Supercomputing conference (SC13)
 in 2013, but he observed that the notion of sustainable scientific software drew
@@
 It appeared to Turk that the supercomputing conferences generally adopt some
-top-down approaches, whereas SciPy bottom-up systems; the essential messages
-perceived were also often circumpolar: software getting harder in supercomputing
+top-down approaches, whereas SciPy generally uses more bottom-up systems.
+The essential messages
+perceived were also often circumpolar: \katznote{this means ``situated around or inhabiting one of the earth's poles'' - not sure this is the right word} software getting harder in supercomputing
 communities with exascale computing and optimization issues in mind, but software becoming better in the
-SciPy groups with emerging tools such as Jupiter and productivity packages for research workflow.
-Admitting such comparisons are somewhat unfair generalization, Turk reminded the
+SciPy group with emerging tools such as Jupiter and productivity packages for research workflow.
+Admitting such comparisons are somewhat unfair generalizations, Turk reminded the
 audience that the different approaches bring different types of ideas to the
-table, but he welcomed that WSSSPE3 was outside existing preconceptions.
+table, but he welcomed WSSSPE3 being conducted outside existing preconceptions.
 

### 2015-10-22 WSSSPE3/report/wg_main_publishing_SW.tex
adding a crossref and removing a comment
Added label, fixed minor typo

@@
-\subsection{Publishing Software Working Group Discussion}
+\subsection{Publishing Software Working Group Discussion} \label{sec:publishing-software}
 
@@
 
-Transitive Credit: Since these forms of publishing must make their sources explicit,
+Software Credit (\S\ref{sec:software-credit}): Since these forms of publishing must make their sources explicit,
 they should be easier to trace even if appropriately worded credit for software
@@
 
-The afforementioned page to be published on SSI.
+The aforementioned page to be published on SSI.

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote. Dropped 'circumpolar'. Too colorful.

@@
 
-WSSSPE3 began with a keynote speech delivered by Professor Matthew
-Turk from the Department of Astronomy, University of Illinois titled \emph{Why
-Sustain Scientific Software?}. Turk is a prolific practitioner of scientific
-software and has extensive experiences working on large collaborative projects
-employing modern computing tools. He is also a co-organizer and champion of all
-the WSSSPE events so far.
+WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
+Department of Astronomy, University of Illinois titled \emph{Why Sustain
+Scientific Software?}. Turk is a prolific practitioner of scientific software
+and has extensive experiences working on large collaborative projects employing
+modern computing tools. He is also a co-organizer and champion of all the WSSSPE
+events so far.
 
@@
 
-WSSSPE3 broke free of the traditional Supercomputing Conference
-environment this year, and in Turk's words, it spoke to the fact that scientific
-software comes from many different types of inquiries, deployment, strategies
-for maintenance, users; and ways of measuring the value of a piece of software.
-It appeared to Turk that the supercomputing conferences generally adopt some
-top-down approaches, whereas SciPy generally uses more bottom-up systems.
-The essential messages
-perceived were also often circumpolar: \katznote{this means ``situated around or inhabiting one of the earth's poles'' - not sure this is the right word} software getting harder in supercomputing
-communities with exascale computing and optimization issues in mind, but software becoming better in the
-SciPy group with emerging tools such as Jupiter and productivity packages for research workflow.
-Admitting such comparisons are somewhat unfair generalizations, Turk reminded the
-audience that the different approaches bring different types of ideas to the
-table, but he welcomed WSSSPE3 being conducted outside existing preconceptions.
+WSSSPE3 broke free of the traditional Supercomputing Conference environment this
+year, and in Turk's words, it spoke to the fact that scientific software comes
+from many different types of inquiries, deployment, strategies for maintenance,
+users; and ways of measuring the value of a piece of software. It appeared to
+Turk that the supercomputing conferences generally adopt some top-down
+approaches, whereas SciPy more often than not uses more bottom-up systems. The essential
+messages perceived were also often polar: software getting harder in
+supercomputing communities with exascale computing and optimization issues in
+mind, but software becoming better in the SciPy group with emerging tools such
+as Jupiter and productivity packages for research workflow. Admitting such
+comparisons are somewhat unfair generalizations, Turk reminded the audience that
+the different approaches bring different types of ideas to the table, but he
+welcomed WSSSPE3 being conducted outside existing preconceptions.
 

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
small wording fix

@@
 approaches, whereas SciPy more often than not uses more bottom-up systems. The essential
-messages perceived were also often polar: software getting harder in
+messages perceived were also often bipolar: software getting harder in
 supercomputing communities with exascale computing and optimization issues in

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote: Back to the topic.

--- 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
@@
 Turk that the supercomputing conferences generally adopt some top-down
-approaches, whereas SciPy more often than not uses more bottom-up systems. The essential
-messages perceived were also often polar: software getting harder in
+approaches, whereas SciPy more often than not uses more bottom-up systems. The
+essential messages perceived were also often polar: software getting harder in
 supercomputing communities with exascale computing and optimization issues in
@@
 welcomed WSSSPE3 being conducted outside existing preconceptions.
+
+Returning to the topic of his talk, Turk invited the audience to picture
+scientific software as a flower on a landscape under the Sun, which may
+represent in this analogy a number of measurable factors such as number of
+citations; growth of a community and number of contributors; amount of funding;
+prestigious prizes awarded; stability of the community in terms of leadership
+transitions, serving community needs, not breaking test suites, and performance
+on new architectures. But all these metrics are strictly speaking
+\emph{proxies} for the values and the impact scientific software bears. What we
+can measure does not give us direct insight---it just gives us proxies of
+insight.
 

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote: Bipolar is better than polar.

@@
 approaches, whereas SciPy more often than not uses more bottom-up systems. The
-essential messages perceived were also often polar: software getting harder in
+essential messages perceived were also often bipolar: software getting harder in
 supercomputing communities with exascale computing and optimization issues in

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote: Definitions of sustainability.

@@
 transitions, serving community needs, not breaking test suites, and performance
-on new architectures. But all these metrics are strictly speaking
-\emph{proxies} for the values and the impact scientific software bears. What we
-can measure does not give us direct insight---it just gives us proxies of
-insight.
+on new architectures. But all these metrics are strictly speaking \emph{proxies}
+for the values and the impact scientific software bears. What we can measure
+does not give us direct insight---it just gives us proxies of insight.
+
+Turk then moved onto various different definitions of sustainability. His most
+favourite one was keeping up with bug reports, in which case even if no new
+features were added, the software remains sustainable. Another definition of
+sustainability Turk mentioned was adding on new features, or maintaining the
+software for a long period of time such as the cases of TeX or LaTeX with
+community help. A notion Turk heard often at supercomputing conferences was that
+sustainable software continued to work on new architectures. Yet another metric
+was people continuing to be able to learn how to use and apply the software.
+Some funder Turk heard talked about sustainability as getting funded. Turk also
+recalled that Greg Wilson said in WSSSPE 1.1 in his view sustainable software
+was software that continued to give the same results over time. A last measure
+of sustainability Turk presented was the ability to transition between different people
+developing and using a piece of software.
 

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
edits in new text

@@
 scientific software as a flower on a landscape under the Sun, which may
-represent in this analogy a number of measurable factors such as number of
+represent a number of measurable factors such as number of
 citations; growth of a community and number of contributors; amount of funding;
@@
 
-Turk then moved onto various different definitions of sustainability. His most
-favourite one was keeping up with bug reports, in which case even if no new
+Turk then moved onto various different definitions of sustainability. His 
+favorite one was keeping up with bug reports, in which case even if no new
 features were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was adding on new features, or maintaining the
+sustainability Turk mentioned was adding of new features, or maintaining the
 software for a long period of time such as the cases of TeX or LaTeX with
@@
 was people continuing to be able to learn how to use and apply the software.
-Some funder Turk heard talked about sustainability as getting funded. Turk also
-recalled that Greg Wilson said in WSSSPE 1.1 in his view sustainable software
+A funder Turk heard talked about sustainability as continuing to get funded. Turk also
+recalled that Greg Wilson said in WSSSPE 1.1 that his view of sustainable software
 was software that continued to give the same results over time. A last measure

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Minor formatting fixes

--- 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
@@
 (WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
-held on 28-29 September 2015 in Boulder, Colorado, USA. Previous events in the
+held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
 WSSSPE series are
@@
 travel of potential participants who would not otherwise be able to attend,
-younger participants and those from underrepresented groups. These funds allowed
+younger participants, and those from underrepresented groups. These funds allowed
 16 additional participants to attend.
@@
 occurred in each the working groups
-(Appendices~\ref{sec:appendix_best_practices}-\ref{sec:appendix_user_community}.)
+(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.)
 
@@
 the goal of contributing to the discussion of one or more working groups. There
-was 24 lightning talks submitted, and after a peer-review process, 16 of these
+was 24 lightning talks submitted, and, after a peer-review process, 16 of these
 were accepted, as discussed further in Section~\ref{sec:lightning}.
@@
 
-WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
-Department of Astronomy, University of Illinois titled \emph{Why Sustain
-Scientific Software?}. Turk is a prolific practitioner of scientific software
-and has extensive experiences working on large collaborative projects employing
-modern computing tools. He is also a co-organizer and champion of all the WSSSPE
-events so far.
+WSSSPE3 began with a keynote speech delivered by Professor Matthew
+Turk from the Department of Astronomy, University of Illinois titled \emph{Why
+Sustain Scientific Software?}. Turk is a prolific practitioner of scientific
+software and has extensive experiences working on large collaborative projects
+employing modern computing tools. He is also a co-organizer and champion of all
+the WSSSPE events so far.
 
@@
 
-WSSSPE3 broke free of the traditional Supercomputing Conference environment this
-year, and in Turk's words, it spoke to the fact that scientific software comes
-from many different types of inquiries, deployment, strategies for maintenance,
-users; and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing conferences generally adopt some top-down
-approaches, whereas SciPy more often than not uses more bottom-up systems. The
-essential messages perceived were also often bipolar: software getting harder in
-supercomputing communities with exascale computing and optimization issues in
-mind, but software becoming better in the SciPy group with emerging tools such
-as Jupiter and productivity packages for research workflow. Admitting such
-comparisons are somewhat unfair generalizations, Turk reminded the audience that
-the different approaches bring different types of ideas to the table, but he
-welcomed WSSSPE3 being conducted outside existing preconceptions.
-
-Returning to the topic of his talk, Turk invited the audience to picture
-scientific software as a flower on a landscape under the Sun, which may
-represent in this analogy a number of measurable factors such as number of
-citations; growth of a community and number of contributors; amount of funding;
-prestigious prizes awarded; stability of the community in terms of leadership
-transitions, serving community needs, not breaking test suites, and performance
-on new architectures. But all these metrics are strictly speaking
-\emph{proxies} for the values and the impact scientific software bears. What we
-can measure does not give us direct insight---it just gives us proxies of
-insight.
+WSSSPE3 broke free of the traditional Supercomputing Conference
+environment this year, and in Turk's words, this spoke to the fact that scientific
+software comes from many different types of inquiries, deployment, strategies
+for maintenance, users; and ways of measuring the value of a piece of software.
+It appeared to Turk that the supercomputing conferences generally adopt some
+top-down approaches, whereas SciPy generally uses more bottom-up systems.
+The essential messages
+perceived were also often circumpolar: \katznote{this means ``situated around or inhabiting one of the earth's poles'' - not sure this is the right word} software getting harder in supercomputing
+communities with exascale computing and optimization issues in mind, but software becoming better in the
+SciPy group with emerging tools such as Jupiter and productivity packages for research workflow.
+Admitting such comparisons are somewhat unfair generalizations, Turk reminded the
+audience that the different approaches bring different types of ideas to the
+table, but he welcomed WSSSPE3 being conducted outside existing preconceptions.
 

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex

@@
 
-WSSSPE3 broke free of the traditional Supercomputing Conference
-environment this year, and in Turk's words, this spoke to the fact that scientific
-software comes from many different types of inquiries, deployment, strategies
-for maintenance, users; and ways of measuring the value of a piece of software.
-It appeared to Turk that the supercomputing conferences generally adopt some
-top-down approaches, whereas SciPy generally uses more bottom-up systems.
-The essential messages
-perceived were also often circumpolar: \katznote{this means ``situated around or inhabiting one of the earth's poles'' - not sure this is the right word} software getting harder in supercomputing
-communities with exascale computing and optimization issues in mind, but software becoming better in the
-SciPy group with emerging tools such as Jupiter and productivity packages for research workflow.
-Admitting such comparisons are somewhat unfair generalizations, Turk reminded the
-audience that the different approaches bring different types of ideas to the
-table, but he welcomed WSSSPE3 being conducted outside existing preconceptions.
+WSSSPE3 broke free of the traditional Supercomputing Conference environment this
+year, and in Turk's words, this spoke to the fact that scientific software comes
+from many different types of inquiries, deployment, strategies for maintenance,
+users; and ways of measuring the value of a piece of software. It appeared to
+Turk that the supercomputing conferences generally adopt some top-down
+approaches, whereas SciPy more often than not uses more bottom-up systems. The
+essential messages perceived were also often bipolar: software getting harder in
+supercomputing communities with exascale computing and optimization issues in
+mind, but software becoming better in the SciPy group with emerging tools such
+as Jupiter and productivity packages for research workflow. Admitting such
+comparisons are somewhat unfair generalizations, Turk reminded the audience that
+the different approaches bring different types of ideas to the table, but he
+welcomed WSSSPE3 being conducted outside existing preconceptions.
+
+Returning to the topic of his talk, Turk invited the audience to picture
+scientific software as a flower on a landscape under the Sun, which may
+represent a number of measurable factors such as number of
+citations; growth of a community and number of contributors; amount of funding;
+prestigious prizes awarded; stability of the community in terms of leadership
+transitions, serving community needs, not breaking test suites, and performance
+on new architectures. But all these metrics are strictly speaking \emph{proxies}
+for the values and the impact scientific software bears. What we can measure
+does not give us direct insight---it just gives us proxies of insight.
+
+Turk then moved onto various different definitions of sustainability. His 
+favorite one was keeping up with bug reports, in which case even if no new
+features were added, the software remains sustainable. Another definition of
+sustainability Turk mentioned was adding of new features, or maintaining the
+software for a long period of time such as the cases of TeX or LaTeX with
+community help. A notion Turk heard often at supercomputing conferences was that
+sustainable software continued to work on new architectures. Yet another metric
+was people continuing to be able to learn how to use and apply the software.
+A funder Turk heard talked about sustainability as continuing to get funded. Turk also
+recalled that Greg Wilson said in WSSSPE 1.1 that his view of sustainable software
+was software that continued to give the same results over time. A last measure
+of sustainability Turk presented was the ability to transition between different people
+developing and using a piece of software.
 

### 2015-10-22 WSSSPE3/report/wg_main_publishing_SW.tex
adding a note about an uncertain change

@@
 
-Software Credit (\S\ref{sec:software-credit}): Since these forms of publishing must make their sources explicit,
+\todo{this originally was ``Transitive Credit:''  Kyle and Dan are not sure if this, or the original, is correct.}
+Software Credit (\S\ref{sec:software-credit}):
+Since these forms of publishing must make their sources explicit,
 they should be easier to trace even if appropriately worded credit for software

### 2015-10-22 WSSSPE3/report/wg_main_software_credit.tex
Software credit appendix in progress; added some relevant pubs
Former-commit-id: d4910da9ce59d01dbc76b0bd5c003886b00bda70
adding a crossref and removing a comment
Added label
Added SC landing pages

--- 2015-09-30 WSSSPE3/report/wg_main_legacy_SW.tex
@@
-\subsection{Topic} \todo{change the title to your topic}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Software Credit Working Group}
+\label{sec:software-credit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \todo{short text here - can include links/cites}
+
+Publishing Software Working Group (\S\ref{sec:publishing-software}): publishing a software paper offers one mechanism for receiving credit
 
@@
 \subsubsection{Landing Page}
-\todo{link to landing page}
+
+Since near-term efforts will be shifting to the FORCE11 Software Citation Working Group (SCWG), we direct interested readers to that group's existing landing page\footnote{FORCE11-SCWG landing page, \url{https://www.force11.org/group/software-citation-working-group}} and GitHub repository\footnote{FORCE11-SCWG GitHub page,  \url{https://github.com/force11/force11-scwg}}.

### 2015-10-22 WSSSPE3/report/WSSSPE3_review.tex
Replaced WSSSPE2 tweets with those from WSSSPE3; could be trimmed

@@
 \usepackage{verbatim}
+\usepackage{booktabs,multicol}
 
@@
 \todo{need to update for WSSSPE3}
-The WSSSPE2 workshop continued our experiment from WSSSPE1 in how we can
+The WSSSPE3 workshop continued our experiment from WSSSPE1 and WSSSPE2 in how we can
 collaboratively build a workshop agenda, and we began a new experiment in
@@
 \centering
-\caption{Top tweets tagged \#WSSSPE on Nov 16, 2014. \todo{update this for WSSSPE3?}}\label{tab:tweets}
+\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015. \todo{update this for WSSSPE3?}}\label{tab:tweets}
   \begin{scriptsize}
-  \begin{tabular}{l|l|r|r}
- \hline
+  \begin{tabular}{@{}l l l l@{}}
+ \toprule
     Author  &   Tweet  & Retweets &  Favorites
-\\ \hline
-% Software Carpentry & Nov 17 & You call it ``project planning'' if it hasn't started yet,          & 8 & 1
-%\\                          & &  and``software sustainability'' if it has and wasn't planned.     &    &
+\\ \midrule
+Neil P Chue Hong  &  Getting ready for the start of the \#WSSSPE workshop & 2 & 1
+\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}  & &
+\\ & Video stream: \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & &
 %
-Neil P Chue Hong   &  Here's @SoftwareSaved guidance on Writing and using a software & 7 & 4
-\\     &  management plan used by EPSRC software grants    &    &
-\\     &  \url{http://www.software.ac.uk/resources/guides/software-management-plans}  &    &
+\\ August Muench & Astronomy \& Software for a totally distracting Monday (2/2) => & 1 & 3
+\\ & Workshop on Sustainable SW for Science \#WSSSPE3 & & 
+\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/} & &
 %
-\\  Neil P Chue Hong  &  @jameshowison as well as software plans   & 4 & 7
-\\   &   \url{http://www.software.ac.uk/resources/guides/software-management-plans} &    &
-\\   &   we provide a software evaluation tool:   &    &
-\\   &   \url{http://www.software.ac.uk/online-sustainability-evaluation}  &    &
+\\ Neil P Chue Hong & My \#WSSSPE lightning talk on Building a Scientific Software & 6 & 3
+\\ &  Accreditation Framework: \url{http://dx.doi.org/10.6084/m9.figshare.1555925} & & 
 %
-\\ Tom Crick  & $56\%$ of UK researchers develop their own software $\rightarrow  140,000$   &  14 & 8
-\\ & UK researchers write research software w$/$out any formal training &    &
+\\ Mozilla Science Lab & At \#wssspe3 and want to learn more about our Working Open guide & 2 & 1 
+\\ &  or Contributorship badges pilot? Keep an eye out for @abbycabs! & & 
 %
-\\ Karthik Ram  &  OH: ``Institutionalize metadata before metadata institutionalizes you'' & 8 & 6
+\\ Daniel S.\ Katz & \#wssspe starting! & 1 & 3
 %
-\\ Josh Greenberg  &  @jameshowison: ``1. retract any paper with bitrotten dependencies'' *mic drop* & 13 & 8
-\\   &   ``2. add anyone who fixes bitrot as an author'' *mic drop*  &    &
+\\ Daniel S.\ Katz & @powersoffour starting \#wssspe keynote & 1 & 2
 %
-\\ Ethan White &  ``@rOpenSci is all about community... our measures of success  & 9 & 3
-\\ &   [include] how many faces are up on our community page''   &    &
+\\ Neil P Chue Hong   &  The difference between software presented at SC \& SciPy & 1 & 4
+\\  &  according to @powersoffour straw person \#wssspe & &
 %
-\\ Ethan White & Daniel Katz talking about implementing transitive credit for  & 9 & 7
-\\ &  software \url{http://arxiv.org/abs/1407.5117}  Work with @arfon  &    &
+\\ Daniel S.\ Katz & sustainability may mean: keeps up with bug reports, adds new features, & 3 & 1
+\\ &  works on new architectures (1/2)  -- @powersoffour at \#wssspe (1/2) & & 
 %
-\\ Kaitlin Thaney  & Great point by @tracykteal about planning for ``end of life'' with scientific & 4 & 8
-\\ &  software projects and sustainability.  & &
+\\ Daniel S.\ Katz & sustainability may mean: people use it, keeps getting funded, produces & 2 & 
+\\ &  same results, transits over people (2/2)  -- @powersoffour at \#wssspe & &
 %
-%\\ Aleksandra Pawlik  &  ``Tell us how you test your scientific software'' @gvwilson @swcarpentry  & 7 & 2
+\\ Nic Weber & Rec, by Matt Turk: Ash Dryden - Ethics of Unpaid Labor & 3 & 2 
+\\ & \url{http://t.co/KfMwS4VG4N} & &
 %
-\\ Aleksandra Pawlik  & Lack of training as one of the main barriers for sustainable software & 10 & 4
-\\ &    at @Supercomputing. @swcarpentry @datacarpentry can fix that!  & &
+\\ Nic Weber & My .02 - many more models (than 3) for sustaining sci. software. & 3 & 2
+\\ & See this pub on open-source:  & & 
+\\ & \url{http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2568185} & &
 %
-\\ Kaitlin Thaney  & My slides from this morning's keynote at & 11 & 12
-\\ &  WSSSPE on Designing for Truth, Scale $+$ Sustainability:  & &
-\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{http://www.slideshare.net/kaythaney/}     & &
-\\ &  \href{http://www.slideshare.net/kaythaney/designing-for-truth-scale-and-sustainability-wssspe2-keynote}{designing-for-truth-scale-and-sustainability-wssspe2-keynote} & &
+\\ Erin Robinson & \#WSSSPE - Software is like a vampire. It's only dead if you really kill it. & 2 & 2
 %
-\\ Neil P Chue Hong & @kaythaney shout out for @swcarpentry @datacarpentry & 9 & 4
-\\ &  @rOpenSci @stilettofiend around open training activities for sustainability  & &
+\\ Neil P Chue Hong & Absolutely great \#wssspe keynote from @powersoffour to inspire & 2 & 1 
+\\ & the workshop. Should be archived via & & 
+\\ & \url{http://ucarconnect.ucar.edu/live?room=cg1aud#.VgmpvCdIrCQ} soon & & 
 %
-%\\ Richard Littauer &  Sweet! @swcarpentry has had 4000+ learners in the past year. & 6 &
+\\ Daniel S.\ Katz & \#wssspe lightning talks starting - streaming at  & 5  &  1
+\\ &  \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & & 
 %
-\\ Neil P Chue Hong & For those interested in Github - Figshare/Zenodo integration, & 5 & 12
-\\ & but want SWORD/DSpace/Fedora/ePrints see:  & &
-\\ & \url{http://blog.stuartlewis.com/2014/09/09/github-to-repository-deposit/}  & &
+\\ Kyle Niemeyer & Participatory open source means better documentation  & 2 & 1 
+\\ & (readme, roadmap, contributing) ? @abbycabs at \#wssspe & &
 %
-\\ Hilmar Lapp & Re: adopting the unix philosophy, consider signing the Small Tools in & 7 & 6
-\\ & Bioinformatics Manifesto: \url{https://github.com/pjotrp/bioinformatics}   &
+\\ Erin Robinson & Interesting resource: \url{http://mozillascience.github.io/leadership-training/} & 2 & 
+\\ &  \#WSSSPE cc: @abbycabs & & 
 %
-\\Andre Luckow &  ``Traditions last not because they are excellent, & 12 & 3
-\\ & but because influential people are averse to change...''  C. Sunstein     & &
+\\ Nic Weber & @MozillaScience + @abbycabs giving look at cool resources & 2 & 3
+\\ & for leading an open source community & & 
+\\ & \url{http://mozillascience.github.io/leadership-training/} & & 
 %
-\\ Tom Crick &  ``Can I Implement Your Algorithm?'':  A Model for  Reproducible & 9 & 8
-\\   &  Research Software \url{http://arxiv.org/abs/1407.5981}  & &
+\\ Nic Weber & Some interesting sepeartion of best practices for Geosoftware dev & 2 & 
+\\ & into categories \url{https://geodynamics.org/cig/dev/best-practices} & &
 %
-\\Mozilla Science Lab & At a loose end this Sunday? Care about reproducibility, software  $+$ & 10 & 5
-\\ &  \#openscience? Follow the    \#WSSSPE hashtag for more, live from New Orleans.  &  &
+\\ Neil P Chue Hong & Some statistics of software citation in geodynamics & 2 & 
 %
-\\Kaitlin Thaney & I'm in New Orleans at \#WSSSPE , speaking at 9:50 ET on  scientific software & 9 & 9
-\\ &   $+$ sustainability. Tune in! Live stream: \url{http://ustre.am/17ddh} & &
+\\ Daniel S.\ Katz & @jamespjh goal: make research software less rubbish; solution is RSEs & 3 & 
 %
-\\   Daniel S. Katz & \#WSSSPE Agenda (Sunday):  & 10 & 1
-\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe2/agenda/}   &  &
-\\ & URL for live stream of keynotes \& lightning talks: \url{http://ustre.am/17ddh}   &  &
-\\ \hline
+\\ Nic Weber & Highly relevant to \#wssspe crowd - @Impactstory team is launching & 7 & 3
+\\ & a impact of research software project \url{http://depsy.org/}  \#verycool & & 
+%
+\\ Neil P Chue Hong & @jamespjh talking about Research Software Engineer groups in the UK & 2 & 1
+\\ &  \#wssspe @ResearchSoftEng @SoftwareSaved & & 
+%
+\\ Abby Cabunoc Mayes & Wonderful time @ \#WSSSPE 3! Encouraging 2 see a diverse community & 1 & 2 
+\\ & working 2gether (\& making real plans!) \url{http://gif.co/s6v4.gif} & & 
+%
+\\ \bottomrule
     \end{tabular}

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
small edits in conclusion notes

@@
 
-\todo{need to update for WSSSPE3}
 The WSSSPE3 workshop continued our experiment from WSSSPE1 and WSSSPE2 in how we can
@@
 
+\todo{need to update rest of this section for WSSSPE3, except the tweet table has already been updated}
 The differences in workshop organization in WSSSPE2 from WSSSPE1
@@
 \centering
-\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015. \todo{update this for WSSSPE3?}}\label{tab:tweets}
+\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015.}\label{tab:tweets}
   \begin{scriptsize}

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
Keynote: 3 models of sustainability

@@
 sustainable software continued to work on new architectures. Yet another metric
-was people continuing to be able to learn how to use and apply the software.
-A funder Turk heard talked about sustainability as continuing to get funded. Turk also
-recalled that Greg Wilson said in WSSSPE 1.1 that his view of sustainable software
-was software that continued to give the same results over time. A last measure
-of sustainability Turk presented was the ability to transition between different people
-developing and using a piece of software.
+was people continuing to be able to learn how to use and apply the software. A
+funder Turk heard talked about sustainability as continuing to get funded. Turk
+also recalled that Greg Wilson. among others, said in WSSSPE1.1 that his view of
+sustainable software was software that continued to give the same results over
+time. A last measure of sustainability Turk presented was the ability to
+transition between different people developing and using a piece of software.
+
+At WSSSPE1, several different models were presented for ensuring sustainability.
+Turk considered a familiar one was a funded piece of software where an external
+agency provided funds to a group that was not necessarily exclusively working on
+and developing a piece of software, keeps it going, and provides that out to the
+scientific community. The model of productized software, in which a piece of
+software has grown into something that individuals rely on to the point that
+research groups or people are willing to support it at some pay rate, for
+instance, subscription to use cloud services that deploy a piece of software, or
+purchase of a piece of software. A last model Turk felt conflicted about is a
+volunteer model, which is traditional old school---not modern-day open
+source---developement.
 

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
removing catalog group from README and main document
Section 3: keynote: More discussion on productization  and volunteer models.

--- 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
@@
 scientific software as a flower on a landscape under the Sun, which may
-represent a number of measurable factors such as number of
-citations; growth of a community and number of contributors; amount of funding;
-prestigious prizes awarded; stability of the community in terms of leadership
-transitions, serving community needs, not breaking test suites, and performance
-on new architectures. But all these metrics are strictly speaking \emph{proxies}
-for the values and the impact scientific software bears. What we can measure
-does not give us direct insight---it just gives us proxies of insight.
+represent a number of measurable factors such as number of citations; growth of
+a community and number of contributors; amount of funding; prestigious prizes
+awarded; stability of the community in terms of leadership transitions, serving
+community needs, not breaking test suites, and performance on new architectures.
+But all these metrics are strictly speaking \emph{proxies} for the values and
+the impact scientific software bears. What we can measure does not give us
+direct insight---it just gives us proxies of insight.
 
@@
 sustainable software continued to work on new architectures. Yet another metric
-was people continuing to be able to learn how to use and apply the software.
-A funder Turk heard talked about sustainability as continuing to get funded. Turk also
-recalled that Greg Wilson said in WSSSPE 1.1 that his view of sustainable software
-was software that continued to give the same results over time. A last measure
-of sustainability Turk presented was the ability to transition between different people
-developing and using a piece of software.
+was people continuing to be able to learn how to use and apply the software. A
+funder Turk heard talked about sustainability as continuing to get funded. Turk
+also recalled that Greg Wilson, among others, said in WSSSPE1.1 that his view of
+sustainable software was software that continued to give the same results over
+time. A last measure of sustainability Turk presented was the ability to
+transition between different people developing and using a piece of software.
+
+At WSSSPE1, several models were presented for ensuring sustainability. Turk
+considered that a familiar one was a funded piece of software where an external
+agency provided funds to a group that was not necessarily exclusively working on
+and developing a piece of software, keeps it going, and provides that out to the
+scientific community. The model of productized software, in which a piece of
+software has grown into something that individuals rely on to the point that
+research groups or people are willing to support it at some pay rate, for
+instance, subscription to use cloud services that deploy a piece of software, or
+purchase of a piece of software. A final model Turk felt conflicted about is a
+volunteer model that is traditional old-school---not modern-day open
+source---developement.
+
+Turk discussed whether productizing scientific software was synonoymous with
+being sustsainable and self-sufficient. He thought it was not necessarily the
+case and furthermore, it could lead to a divergence of interest between users
+and developers.
+
+The volunteer model meant unpaid labor. On this note, Turk recommended Ashe
+Dryden's blog
+post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
+on the ethics of unpaid labor and the open source software community, which
+mentions GitHub as one's resume, for example. Often times, a person funded to
+work on some scientific project full time has a small amount of time for working
+on a piece of software necessary for the project. However, researchers's ability
+to participate in that volunteer community is not always the same and it is not
+always aligned with their research projects. From Turk's experience, we cannot
+always rely on unpaid labor and volunteer time to sustain a piece of software---
+that came down to the notions of the top-down and the bottom-up approaches, the
+funded versus the grassroot.
 
@@
 \input{wg_main_funding_spec_expert}
-\input{wg_main_catalogs}
+%\input{wg_main_catalogs}
 \input{wg_main_industry_interaction}
@@
 \input{wg_appendix_funding_spec_expert}
-\input{wg_appendix_catalogs}
+%\input{wg_appendix_catalogs}
 \input{wg_appendix_industry_interaction}

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
Section 3: keynote: remove a space

@@
 always aligned with their research projects. From Turk's experience, we cannot
-always rely on unpaid labor and volunteer time to sustain a piece of software---
-that came down to the notions of the top-down and the bottom-up approaches, the
-funded versus the grassroot.
+always rely on unpaid labor and volunteer time to sustain a piece of
+software---that came down to the notions of the top-down and the bottom-up
+approaches, the funded versus the grassroot.
 

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
Section 3: keynote: tiny changes.

@@
 work on some scientific project full time has a small amount of time for working
-on a piece of software necessary for the project. However, researchers's ability
+on a piece of software necessary for the project. However, a researcher's ability
 to participate in that volunteer community is not always the same and it is not
-always aligned with their research projects. From Turk's experience, we cannot
+always aligned with the research projects. From Turk's experience, we cannot
 always rely on unpaid labor and volunteer time to sustain a piece of

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
edits/comment in keynote section

@@
 Department of Astronomy, University of Illinois titled \emph{Why Sustain
-Scientific Software?}. Turk is a prolific practitioner of scientific software
+Scientific Software?}. Turk is a prolific scientific software practitioner
 and has extensive experiences working on large collaborative projects employing
-modern computing tools. He is also a co-organizer and champion of all the WSSSPE
-events so far.
+modern computing tools. He also co-organizes and champions WSSSPE
+events.
 
 In his keynote address, Turk recapped the course of development of WSSSPE
-workshops alongside with his career from a postdoc to an academic over the past
-few years. The first WSSSPE workshop was at the Supercomputing conference (SC13)
+workshops over the past
+few years, alongside his career development from a postdoc to an academic. The first WSSSPE workshop was at the Supercomputing conference (SC13)
 in 2013, but he observed that the notion of sustainable scientific software drew
@@
 
-WSSSPE3 broke free of the traditional Supercomputing Conference environment this
+WSSSPE3 left the traditional Supercomputing Conference environment this
 year, and in Turk's words, this spoke to the fact that scientific software comes
 from many different types of inquiries, deployment, strategies for maintenance,
-users; and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing conferences generally adopt some top-down
-approaches, whereas SciPy more often than not uses more bottom-up systems. The
-essential messages perceived were also often bipolar: software getting harder in
-supercomputing communities with exascale computing and optimization issues in
-mind, but software becoming better in the SciPy group with emerging tools such
+users, and ways of measuring the value of a piece of software. It appeared to
+Turk that the supercomputing community generally adopt some top-down
+approaches, whereas the SciPy community more often than not uses more bottom-up systems. The
+essential messages perceived were also often bipolar: the supercomputing community thinks software is getting harder, with exascale computing and optimization issues in
+mind, but the SciPy community thinks software becoming better, with emerging tools such
 as Jupiter and productivity packages for research workflow. Admitting such
 comparisons are somewhat unfair generalizations, Turk reminded the audience that
-the different approaches bring different types of ideas to the table, but he
+the different approaches bring different types of ideas to the table, and he
 welcomed WSSSPE3 being conducted outside existing preconceptions.
@@
 Turk then moved onto various different definitions of sustainability. His
-favorite one was keeping up with bug reports, in which case even if no new
+favorite one was `keeping up with bug reports,' where even if no new
 features were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was adding of new features, or maintaining the
-software for a long period of time such as the cases of TeX or LaTeX with
+sustainability Turk mentioned was `adding of new features', or `maintaining the
+software for a long period of time' such as the cases of TeX or LaTeX with
 community help. A notion Turk heard often at supercomputing conferences was that
-sustainable software continued to work on new architectures. Yet another metric
-was people continuing to be able to learn how to use and apply the software. A
-funder Turk heard talked about sustainability as continuing to get funded. Turk
+sustainable software `continues to work on new architectures'. Yet another metric
+was `people continuing to be able to learn how to use and apply the software'. A
+funder Turk heard talked about sustainability as `continuing to get funded'. Turk
 also recalled that Greg Wilson, among others, said in WSSSPE1.1 that his view of
-sustainable software was software that continued to give the same results over
-time. A last measure of sustainability Turk presented was the ability to
-transition between different people developing and using a piece of software.
+sustainable software was software that `continued to give the same results over
+time'. A last measure of sustainability Turk presented was `the ability to
+transition between different people developing and using a piece of software.'
 
@@
 considered that a familiar one was a funded piece of software where an external
-agency provided funds to a group that was not necessarily exclusively working on
-and developing a piece of software, keeps it going, and provides that out to the
+agency provided funds to a group who are not necessarily exclusively working on
+and developing a piece of software, keeps it going, and provides it to the
 scientific community. The model of productized software, in which a piece of
 software has grown into something that individuals rely on to the point that
-research groups or people are willing to support it at some pay rate, for
-instance, subscription to use cloud services that deploy a piece of software, or
+research groups or people are willing to support it with some amount of funding, for
+instance, a subscription to use cloud services that deploy a piece of software, or
 purchase of a piece of software. A final model Turk felt conflicted about is a
 volunteer model that is traditional old-school---not modern-day open
-source---developement.
+source---development.
 
-Turk discussed whether productizing scientific software was synonoymous with
-being sustsainable and self-sufficient. He thought it was not necessarily the
+Turk discussed whether productizing scientific software was synonymous with
+being sustainable and self-sufficient. He thought it was not necessarily the
 case and furthermore, it could lead to a divergence of interest between users
@@
 
-The volunteer model meant unpaid labor. On this note, Turk recommended Ashe
+Turk reminded the audience that the volunteer model means unpaid labor. On this note, he recommended Ashe
 Dryden's blog
@@
 on the ethics of unpaid labor and the open source software community, which
-mentions GitHub as one's resume, for example. Often times, a person funded to
-work on some scientific project full time has a small amount of time for working
-on a piece of software necessary for the project. However, a researcher's ability
-to participate in that volunteer community is not always the same and it is not
-always aligned with the research projects. From Turk's experience, we cannot
+mentions GitHub as one's resume, for example.
+\katznote{I don't think I understand the point in the previous sentence - maybe there are multiple points and this should be broken in multiple sentences?} Often times, a person funded to
+work full time on some scientific project can spend a small amount of time for working
+on a piece of software necessary for the project. However, all researchers' ability
+to participate in that volunteer community are not always the same and may not
+always be aligned with the research projects. From Turk's experience, we cannot
 always rely on unpaid labor and volunteer time to sustain a piece of
-software---that came down to the notions of the top-down and the bottom-up
-approaches, the funded versus the grassroot.
+software---this came down to the notions of the top-down and the bottom-up
+approaches, the funded versus the grassroots.
 

### 2015-10-23 WSSSPE3/report/WSSSPE3_review.tex
Section 3: keynote: Response to Dan's note to make merely one point.

@@
 Department of Astronomy, University of Illinois titled \emph{Why Sustain
-Scientific Software?}. Turk is a prolific scientific software practitioner
-and has extensive experiences working on large collaborative projects employing
-modern computing tools. He also co-organizes and champions WSSSPE
-events.
+Scientific Software?}. Turk is a prolific scientific software practitioner and
+has extensive experiences working on large collaborative projects employing
+modern computing tools. He also co-organizes and champions WSSSPE events.
 
 In his keynote address, Turk recapped the course of development of WSSSPE
-workshops over the past
-few years, alongside his career development from a postdoc to an academic. The first WSSSPE workshop was at the Supercomputing conference (SC13)
-in 2013, but he observed that the notion of sustainable scientific software drew
-in an audience beyond supercomputing. In the following year, WSSSPE1.1 at SciPy
-had speakers talking about how software has been sustained inside the scientific
-Python community. WSSSPE2 at SC14 had breakout group discussions coming up with
-actionable items, and WSSSPE2.1 at SciPy 2015 was similar. Turk noted the
-different atmosphere of the surrounding large conferences, despite similar
-WSSSPE participants.
+workshops over the past few years, alongside his career development from a
+postdoc to an academic. The first WSSSPE workshop was at the Supercomputing
+conference (SC13) in 2013, but he observed that the notion of sustainable
+scientific software drew in an audience beyond supercomputing. In the following
+year, WSSSPE1.1 at SciPy had speakers talking about how software has been
+sustained inside the scientific Python community. WSSSPE2 at SC14 had breakout
+group discussions coming up with actionable items, and WSSSPE2.1 at SciPy 2015
+was similar. Turk noted the different atmosphere of the surrounding large
+conferences, despite similar WSSSPE participants.
 
-WSSSPE3 left the traditional Supercomputing Conference environment this
-year, and in Turk's words, this spoke to the fact that scientific software comes
-from many different types of inquiries, deployment, strategies for maintenance,
+WSSSPE3 left the traditional Supercomputing Conference environment this year,
+and in Turk's words, this spoke to the fact that scientific software comes from
+many different types of inquiries, deployment, strategies for maintenance,
 users, and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing community generally adopt some top-down
-approaches, whereas the SciPy community more often than not uses more bottom-up systems. The
-essential messages perceived were also often bipolar: the supercomputing community thinks software is getting harder, with exascale computing and optimization issues in
-mind, but the SciPy community thinks software becoming better, with emerging tools such
-as Jupiter and productivity packages for research workflow. Admitting such
-comparisons are somewhat unfair generalizations, Turk reminded the audience that
-the different approaches bring different types of ideas to the table, and he
-welcomed WSSSPE3 being conducted outside existing preconceptions.
+Turk that the supercomputing community generally adopt some top-down approaches,
+whereas the SciPy community more often than not uses more bottom-up systems. The
+essential messages perceived were also often bipolar: the supercomputing
+community thinks software is getting harder, with exascale computing and
+optimization issues in mind, but the SciPy community thinks software becoming
+better, with emerging tools such as Jupiter and productivity packages for
+research workflow. Admitting such comparisons are somewhat unfair
+generalizations, Turk reminded the audience that the different approaches bring
+different types of ideas to the table, and he welcomed WSSSPE3 being conducted
+outside existing preconceptions.
 
@@
 Turk then moved onto various different definitions of sustainability. His
-favorite one was `keeping up with bug reports,' where even if no new
-features were added, the software remains sustainable. Another definition of
+favorite one was `keeping up with bug reports,' where even if no new features
+were added, the software remains sustainable. Another definition of
 sustainability Turk mentioned was `adding of new features', or `maintaining the
@@
 community help. A notion Turk heard often at supercomputing conferences was that
-sustainable software `continues to work on new architectures'. Yet another metric
-was `people continuing to be able to learn how to use and apply the software'. A
-funder Turk heard talked about sustainability as `continuing to get funded'. Turk
-also recalled that Greg Wilson, among others, said in WSSSPE1.1 that his view of
-sustainable software was software that `continued to give the same results over
-time'. A last measure of sustainability Turk presented was `the ability to
-transition between different people developing and using a piece of software.'
+sustainable software `continues to work on new architectures'. Yet another
+metric was `people continuing to be able to learn how to use and apply the
+software'. A funder Turk heard talked about sustainability as `continuing to get
+funded'. Turk also recalled that Greg Wilson, among others, said in WSSSPE1.1
+that his view of sustainable software was software that `continued to give the
+same results over time'. A last measure of sustainability Turk presented was
+`the ability to transition between different people developing and using a piece
+of software.'
 
@@
 software has grown into something that individuals rely on to the point that
-research groups or people are willing to support it with some amount of funding, for
-instance, a subscription to use cloud services that deploy a piece of software, or
-purchase of a piece of software. A final model Turk felt conflicted about is a
-volunteer model that is traditional old-school---not modern-day open
+research groups or people are willing to support it with some amount of funding,
+for instance, a subscription to use cloud services that deploy a piece of
+software, or purchase of a piece of software. A final model Turk felt conflicted
+about is a volunteer model that is traditional old-school---not modern-day open
 source---development.
@@
 
-Turk reminded the audience that the volunteer model means unpaid labor. On this note, he recommended Ashe
-Dryden's blog
+Turk reminded the audience that the volunteer model means unpaid labor. On this
+note, he recommended Ashe Dryden's blog
 post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
-on the ethics of unpaid labor and the open source software community, which
-mentions GitHub as one's resume, for example.
-\katznote{I don't think I understand the point in the previous sentence - maybe there are multiple points and this should be broken in multiple sentences?} Often times, a person funded to
-work full time on some scientific project can spend a small amount of time for working
-on a piece of software necessary for the project. However, all researchers' ability
-to participate in that volunteer community are not always the same and may not
-always be aligned with the research projects. From Turk's experience, we cannot
-always rely on unpaid labor and volunteer time to sustain a piece of
-software---this came down to the notions of the top-down and the bottom-up
-approaches, the funded versus the grassroots.
+on the ethics of unpaid labor and the open source software community. Often
+times, a person funded to work full time on some scientific project can spend a
+small amount of time for working on a piece of software necessary for the
+project. However, all researchers' ability to participate in that volunteer
+community are not always the same and may not always be aligned with the
+research projects. From Turk's experience, we cannot always rely on unpaid labor
+and volunteer time to sustain a piece of software---this came down to the
+notions of the top-down and the bottom-up approaches, the funded versus the
+grassroots.
 

### 2015-10-24 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote first draft completed.

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote} \label{sec:keynote} 
+\section{Keynote \note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}} \label{sec:keynote} 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{need to update for WSSSPE3. 
-Lead: S.-C. Choi. 
-PDF slides at \url{http://tinyurl.com/qbbqgsj}.
-Youtube Video at \url{http://tinyurl.com/q45kfcn}.
-}
 
@@
 Turk reminded the audience that the volunteer model means unpaid labor. On this
-note, he recommended Ashe Dryden's blog
-post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
-on the ethics of unpaid labor and the open source software community. Often
-times, a person funded to work full time on some scientific project can spend a
-small amount of time for working on a piece of software necessary for the
-project. However, all researchers' ability to participate in that volunteer
-community are not always the same and may not always be aligned with the
-research projects. From Turk's experience, we cannot always rely on unpaid labor
-and volunteer time to sustain a piece of software---this came down to the
-notions of the top-down and the bottom-up approaches, the funded versus the
-grassroots.
+note, he recommended Ashe Dryden's blog post on the
+\href{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}{ethics
+of unpaid labor and the open source software community}. Often times, a person
+funded to work full time on some scientific project can spend a small amount of
+time for working on a piece of software necessary for the project. However, all
+researchers' ability to participate in that volunteer community are not always
+the same and may not always be aligned with the research projects. From Turk's
+experience, we cannot always rely on unpaid labor and volunteer time to sustain
+a piece of software---this came down to the notions of the top-down and the
+bottom-up approaches, i.e., the funded versus the grassroots. However, Turk
+pointed out that bottom-up volunteers driven projects can be just as large scale
+as a top-down model software development.
+
+When Turk talked about sustaining scientific software, it really meant to him as
+conducting scientific inquiries by often some specific software and sustaining
+the people we care about, our careers, and the future of our fields. According
+to Turk, we all have an invested stake in sustaining scientific software. Again
+it came down to the top-down versus bottom-up notion that we can have anointed
+specific software packages to be sustained at the expense of new grassroot
+developments.
+
+To close, Turk talked about possibilities of why we want to sustain scientific
+software: devotion to science and interests in pursuing the next stage of
+research; fun and creative thrill in writing code and paper; usefulness with
+measurable impacts, for example, LINPACK and HDF groups providing of data
+storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
+presented his wishlist of questions to be answered: 
+%
+\begin{itemize}
+
+\item How do we ship a product on time when dealing with a mix of funding models
+and motivations especially when we rely on volunteers?
+
+\item how do we know when it is time to end some software and move on? For
+example, shall we stop sustaining Python and switch to Julia and Javascript?
+
+\item How can productized software balance its future versus its past, or the new
+needs of the customers versus the existing needs of the development community?
+
+\item How can we help avoid burnout and retain the joy in the communities?
+
+\item How can we reduce systemic bias, which goes back to the blog post of
+Dryden especially on how ethics of unpaid labor disproportionately affect
+underrepresented communities?
+
+\end{itemize}
 

### 2015-10-25 WSSSPE3/report/WSSSPE3_review.tex
edits/comment in keynote

@@
 note, he recommended Ashe Dryden's blog post on the
-\href{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}{ethics
-of unpaid labor and the open source software community}. Often times, a person
-funded to work full time on some scientific project can spend a small amount of
-time for working on a piece of software necessary for the project. However, all
+ethics of unpaid labor and the open source software
+community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
+Often times, a person
+funded to work full time on a scientific project can spend a small amount of
+time for working on a piece of software necessary for that project. However, all
 researchers' ability to participate in that volunteer community are not always
-the same and may not always be aligned with the research projects. From Turk's
+the same and may not always be aligned with their research projects. From Turk's
 experience, we cannot always rely on unpaid labor and volunteer time to sustain
@@
 bottom-up approaches, i.e., the funded versus the grassroots. However, Turk
-pointed out that bottom-up volunteers driven projects can be just as large scale
-as a top-down model software development.
+pointed out that bottom-up, volunteer-driven projects can be just as large
+as a top-down software development project.
 
-When Turk talked about sustaining scientific software, it really meant to him as
-conducting scientific inquiries by often some specific software and sustaining
+Turk said that sustaining scientific software really meant to him
+conducting scientific inquiries, often by some specific software, and sustaining
 the people we care about, our careers, and the future of our fields. According
@@
 it came down to the top-down versus bottom-up notion that we can have anointed
-specific software packages to be sustained at the expense of new grassroot
-developments.
+specific software packages to be sustained at the expense of new grassroots
+developments. \katznote{I don't understand this last sentence.}
 
-To close, Turk talked about possibilities of why we want to sustain scientific
+Turk talked about possible reasons why we want to sustain scientific
 software: devotion to science and interests in pursuing the next stage of
-research; fun and creative thrill in writing code and paper; usefulness with
+research; fun and creative thrill in writing codes and papers; usefulness with
 measurable impacts, for example, LINPACK and HDF groups providing of data
@@
 
-\item how do we know when it is time to end some software and move on? For
-example, shall we stop sustaining Python and switch to Julia and Javascript?
+\item How do we know when it is time to end some software and move on? For
+example, should we stop sustaining Python and switch to Julia and Javascript?
 

### 2015-10-25 WSSSPE3/report/wg_appendix_software_credit.tex
finished discussion of day 1 in SC appendix

@@
 The group also discussed separating quantitative measures (e.g., number of citations) from the value of a work in order to give credit, moving towards qualitative or anecdotal evidence of value.
+Other topics brought up included a form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and using market penetration or adoption rate in a community as a metric, although it was not clear how this would be measured.
+Finally, the concept a software tool's uniqueness or indispensability to a community was mentioned, with value being characterized by a particular piece of software either offering unique capabilities or doing something better, faster, or with less computational requirements than other offerings.
 
-PageRank for citations, based on number of mentions
-
-market penetration or adoption rate in a community as a metric, although it was not clear how this would be measured.
-
-Finally, the concept a software tool's uniqueness or indispensability to a community was mentioned, with value being characterized by a particular piece of software either offering unique capabilities or doing something better, faster, or with less computational requirements than other offerings.
+On the second day of WSSSPE3, the working group decided to put aside the taxonomy of contributions and focus on software citations to ensure developers receive credit (regardless of contribution).
+Eventually, once standardizing software citations, the goal would be to return to establishing different roles\slash contributions for this credit.
 

### 2015-10-25 WSSSPE3/report/wg_appendix_software_credit.tex
added missing ref for Mozilla Contributorship Badges

@@
     \item Project CRediT~\cite{projectcredit}, which assigns roles to paper authors based on their specific contributions; and
-    \item Mozilla Science Lab's recently introduced Contributorship Badges for Science, which provide a badge---associated with ORCID~\cite{orcid}---based on author contributions in a similar manner to Project CRediT.
+    \item Mozilla Science Lab's recently introduced Contributorship Badges for Science~\cite{Mozilla_badges}, which provide a badge---associated with ORCID~\cite{orcid}---based on author contributions in a similar manner to Project CRediT.
 \end{itemize}
-However, as of this writing, only the Project CRediT roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship Badges~\cite{} have been implemented in published papers, though both of these only provide a single ``Software'' or ``Computation'' category associated with software.
+However, as of this writing, only the Project CRediT roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship Badges~\cite{Mozilla_badges} have been implemented in published papers, though both of these only provide a single ``Software'' or ``Computation'' category associated with software.
 In addition, neither of these options allow for the citation of software itself, but only provide an author contribution related to software.

### 2015-10-27 WSSSPE3/report/WSSSPE3_review.tex
removing the tweet table and text - not sure they really add anything alternative views welcome

@@
 
+\todo{need to update this section for WSSSPE3}
+
 The WSSSPE3 workshop continued our experiment from WSSSPE1 and WSSSPE2 in how we can
@@
 
-\todo{need to update rest of this section for WSSSPE3, except the tweet table has already been updated}
 The differences in workshop organization in WSSSPE2 from WSSSPE1
@@
 
-\begin{table*}[t]
-\centering
-\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015.}\label{tab:tweets}
-  \begin{scriptsize}
-  \begin{tabular}{@{}l l l l@{}}
- \toprule
-    Author  &   Tweet  & Retweets &  Favorites
-\\ \midrule
-Neil P Chue Hong  &  Getting ready for the start of the \#WSSSPE workshop & 2 & 1
-\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}  & &
-\\ & Video stream: \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & &
+%\begin{table*}[t]
+%\centering
+%\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015.}\label{tab:tweets}
+%  \begin{scriptsize}
+%  \begin{tabular}{@{}l l l l@{}}
+% \toprule
+%    Author  &   Tweet  & Retweets &  Favorites
+%\\ \midrule
+%Neil P Chue Hong  &  Getting ready for the start of the \#WSSSPE workshop & 2 & 1
+%\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}  & &
+%\\ & Video stream: \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & &
+%%
+%\\ August Muench & Astronomy \& Software for a totally distracting Monday (2/2) => & 1 & 3
+%\\ & Workshop on Sustainable SW for Science \#WSSSPE3 & & 
+%\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/} & &
+%%
+%\\ Neil P Chue Hong & My \#WSSSPE lightning talk on Building a Scientific Software & 6 & 3
+%\\ &  Accreditation Framework: \url{http://dx.doi.org/10.6084/m9.figshare.1555925} & & 
+%%
+%\\ Mozilla Science Lab & At \#wssspe3 and want to learn more about our Working Open guide & 2 & 1 
+%\\ &  or Contributorship badges pilot? Keep an eye out for @abbycabs! & & 
+%%
+%\\ Daniel S.\ Katz & \#wssspe starting! & 1 & 3
+%%
+%\\ Daniel S.\ Katz & @powersoffour starting \#wssspe keynote & 1 & 2
+%%
+%\\ Neil P Chue Hong   &  The difference between software presented at SC \& SciPy & 1 & 4
+%\\  &  according to @powersoffour straw person \#wssspe & &
+%%
+%\\ Daniel S.\ Katz & sustainability may mean: keeps up with bug reports, adds new features, & 3 & 1
+%\\ &  works on new architectures (1/2)  -- @powersoffour at \#wssspe (1/2) & & 
+%%
+%\\ Daniel S.\ Katz & sustainability may mean: people use it, keeps getting funded, produces & 2 & 
+%\\ &  same results, transits over people (2/2)  -- @powersoffour at \#wssspe & &
+%%
+%\\ Nic Weber & Rec, by Matt Turk: Ash Dryden - Ethics of Unpaid Labor & 3 & 2 
+%\\ & \url{http://t.co/KfMwS4VG4N} & &
+%%
+%\\ Nic Weber & My .02 - many more models (than 3) for sustaining sci. software. & 3 & 2
+%\\ & See this pub on open-source:  & & 
+%\\ & \url{http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2568185} & &
+%%
+%\\ Erin Robinson & \#WSSSPE - Software is like a vampire. It's only dead if you really kill it. & 2 & 2
+%%
+%\\ Neil P Chue Hong & Absolutely great \#wssspe keynote from @powersoffour to inspire & 2 & 1 
+%\\ & the workshop. Should be archived via & & 
+%\\ & \url{http://ucarconnect.ucar.edu/live?room=cg1aud#.VgmpvCdIrCQ} soon & & 
+%%
+%\\ Daniel S.\ Katz & \#wssspe lightning talks starting - streaming at  & 5  &  1
+%\\ &  \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & & 
+%%
+%\\ Kyle Niemeyer & Participatory open source means better documentation  & 2 & 1 
+%\\ & (readme, roadmap, contributing) ? @abbycabs at \#wssspe & &
+%%
+%\\ Erin Robinson & Interesting resource: \url{http://mozillascience.github.io/leadership-training/} & 2 & 
+%\\ &  \#WSSSPE cc: @abbycabs & & 
+%%
+%\\ Nic Weber & @MozillaScience + @abbycabs giving look at cool resources & 2 & 3
+%\\ & for leading an open source community & & 
+%\\ & \url{http://mozillascience.github.io/leadership-training/} & & 
+%%
+%\\ Nic Weber & Some interesting sepeartion of best practices for Geosoftware dev & 2 & 
+%\\ & into categories \url{https://geodynamics.org/cig/dev/best-practices} & &
+%%
+%\\ Neil P Chue Hong & Some statistics of software citation in geodynamics & 2 & 
+%%
+%\\ Daniel S.\ Katz & @jamespjh goal: make research software less rubbish; solution is RSEs & 3 & 
+%%
+%\\ Nic Weber & Highly relevant to \#wssspe crowd - @Impactstory team is launching & 7 & 3
+%\\ & a impact of research software project \url{http://depsy.org/}  \#verycool & & 
+%%
+%\\ Neil P Chue Hong & @jamespjh talking about Research Software Engineer groups in the UK & 2 & 1
+%\\ &  \#wssspe @ResearchSoftEng @SoftwareSaved & & 
+%%
+%\\ Abby Cabunoc Mayes & Wonderful time @ \#WSSSPE 3! Encouraging 2 see a diverse community & 1 & 2 
+%\\ & working 2gether (\& making real plans!) \url{http://gif.co/s6v4.gif} & & 
+%%
+%\\ \bottomrule
+%    \end{tabular}
+%    \end{scriptsize}
+%\end{table*}
 %
-\\ August Muench & Astronomy \& Software for a totally distracting Monday (2/2) => & 1 & 3
-\\ & Workshop on Sustainable SW for Science \#WSSSPE3 & & 
-\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/} & &
-%
-\\ Neil P Chue Hong & My \#WSSSPE lightning talk on Building a Scientific Software & 6 & 3
-\\ &  Accreditation Framework: \url{http://dx.doi.org/10.6084/m9.figshare.1555925} & & 
-%
-\\ Mozilla Science Lab & At \#wssspe3 and want to learn more about our Working Open guide & 2 & 1 
-\\ &  or Contributorship badges pilot? Keep an eye out for @abbycabs! & & 
-%
-\\ Daniel S.\ Katz & \#wssspe starting! & 1 & 3
-%
-\\ Daniel S.\ Katz & @powersoffour starting \#wssspe keynote & 1 & 2
-%
-\\ Neil P Chue Hong   &  The difference between software presented at SC \& SciPy & 1 & 4
-\\  &  according to @powersoffour straw person \#wssspe & &
-%
-\\ Daniel S.\ Katz & sustainability may mean: keeps up with bug reports, adds new features, & 3 & 1
-\\ &  works on new architectures (1/2)  -- @powersoffour at \#wssspe (1/2) & & 
-%
-\\ Daniel S.\ Katz & sustainability may mean: people use it, keeps getting funded, produces & 2 & 
-\\ &  same results, transits over people (2/2)  -- @powersoffour at \#wssspe & &
-%
-\\ Nic Weber & Rec, by Matt Turk: Ash Dryden - Ethics of Unpaid Labor & 3 & 2 
-\\ & \url{http://t.co/KfMwS4VG4N} & &
-%
-\\ Nic Weber & My .02 - many more models (than 3) for sustaining sci. software. & 3 & 2
-\\ & See this pub on open-source:  & & 
-\\ & \url{http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2568185} & &
-%
-\\ Erin Robinson & \#WSSSPE - Software is like a vampire. It's only dead if you really kill it. & 2 & 2
-%
-\\ Neil P Chue Hong & Absolutely great \#wssspe keynote from @powersoffour to inspire & 2 & 1 
-\\ & the workshop. Should be archived via & & 
-\\ & \url{http://ucarconnect.ucar.edu/live?room=cg1aud#.VgmpvCdIrCQ} soon & & 
-%
-\\ Daniel S.\ Katz & \#wssspe lightning talks starting - streaming at  & 5  &  1
-\\ &  \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & & 
-%
-\\ Kyle Niemeyer & Participatory open source means better documentation  & 2 & 1 
-\\ & (readme, roadmap, contributing) ? @abbycabs at \#wssspe & &
-%
-\\ Erin Robinson & Interesting resource: \url{http://mozillascience.github.io/leadership-training/} & 2 & 
-\\ &  \#WSSSPE cc: @abbycabs & & 
-%
-\\ Nic Weber & @MozillaScience + @abbycabs giving look at cool resources & 2 & 3
-\\ & for leading an open source community & & 
-\\ & \url{http://mozillascience.github.io/leadership-training/} & & 
-%
-\\ Nic Weber & Some interesting sepeartion of best practices for Geosoftware dev & 2 & 
-\\ & into categories \url{https://geodynamics.org/cig/dev/best-practices} & &
-%
-\\ Neil P Chue Hong & Some statistics of software citation in geodynamics & 2 & 
-%
-\\ Daniel S.\ Katz & @jamespjh goal: make research software less rubbish; solution is RSEs & 3 & 
-%
-\\ Nic Weber & Highly relevant to \#wssspe crowd - @Impactstory team is launching & 7 & 3
-\\ & a impact of research software project \url{http://depsy.org/}  \#verycool & & 
-%
-\\ Neil P Chue Hong & @jamespjh talking about Research Software Engineer groups in the UK & 2 & 1
-\\ &  \#wssspe @ResearchSoftEng @SoftwareSaved & & 
-%
-\\ Abby Cabunoc Mayes & Wonderful time @ \#WSSSPE 3! Encouraging 2 see a diverse community & 1 & 2 
-\\ & working 2gether (\& making real plans!) \url{http://gif.co/s6v4.gif} & & 
-%
-\\ \bottomrule
-    \end{tabular}
-    \end{scriptsize}
-\end{table*}
-
-WSSSPE actively used the online social network Twitter, with hashtag
-``\#WSSSPE''. There were substantially more tweets (messages) during the days of
-the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
-18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
-day of the workshop. Some of the main points and highlights in the meeting are
-shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
-the day of workshop, selected by the metrics that number of retweets or
-favorites larger than five and the sum of two measures greater than ten.
+%WSSSPE actively used the online social network Twitter, with hashtag
+%``\#WSSSPE''. There were substantially more tweets (messages) during the days of
+%the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
+%18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
+%day of the workshop. Some of the main points and highlights in the meeting are
+%shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
+%the day of workshop, selected by the metrics that number of retweets or
+%favorites larger than five and the sum of two measures greater than ten.
 

### 2015-10-27 WSSSPE3/report/WSSSPE3_review.tex
working on conclusions

@@
 
-\todo{need to update this section for WSSSPE3}
+In WSSSPE3, we attempted to take what we learned from WSSSPE1 and WSSSPE2
+in how we can collaboratively build a workshop agenda and turn that into
+an ongoing community activity.  The success or failure of this effort
+of this will only become apparent over time.
 
-The WSSSPE3 workshop continued our experiment from WSSSPE1 and WSSSPE2 in how we can
-collaboratively build a workshop agenda, and we began a new experiment in
-how to build a series of workshops into an ongoing community activity.
+The workshop had two components, presentations and working groups.  The presentations,
+in the first half day of the workshop, including a inspirational keynote and a set of lightning
+talks.
+We used lightning talks for two reasons: first, the need of some participants to have a slot
+on the agenda to justify their attendance, and second, as a way to get new ideas across
+to all the attendees.  We broke with the tradition of requiring the lightning talk submitters
+to self-publish their papers, and instead used a common peer-review platform\footnote{\url{http://easychair.org}}, choosing to publish the speaker slides on the workshop web site
+instead.
 
-The differences in workshop organization in WSSSPE2 from WSSSPE1
-are in using an existing service (EasyChair) to handle submissions and reviews,
-rather than an ad hoc process, and using an existing service (Well Sorted) to
-allow collaborative grouping of papers into themes by all authors, reviewers,
-and the community, rather than this being done in an ad hoc manner by the
-organizers alone.
+The working groups met for a small part of the first half day and all of the second day,
+excepting some times for the groups to report back to the collected workshop attendees.
+\todo{need to say some more about what came of this}
 
-The fact remains that contributors also want to get credit for their
-participation in the process. And the workshop organizers will want to make
-sure that the workshop content and their efforts are recorded. Ideally, there
-would be a service that would index the contributions to the
-workshop, serving the authors, the organizers, and the larger community. 
-Since there still isn't such a service today, the workshop organizers are
-writing this initial report and making use of arXiv as a partial solution to
-provide a record of the workshop.
+\todo{this paragraph is from the wssspe2 report - to be updated...}
+However, the challenge that we have discovered
+since WSSSPE2 is that it is very hard to continue the breakout groups'
+activities.  The WSSSPE2 participants were willing to dedicate their time to
+the groups while they were at the meeting, but afterwards, they have gone
+back to their (paid) jobs.  We need to determine how to tie the WSSSPE
+breakout activities to people's jobs, so that they feel that continuing them
+is a higher priority than it is now, perhaps through funding the participants,
+or through funding coordinators for each activity, or perhaps by getting
+the workshop participants to agree to a specific schedule of activities during the
+workshop.
 
@@
 
-In terms of building community activities, we wanted to focus primarily on
-working groups, which we were able to do, as discussed above, but we
-also wanted to make sure that attendees felt they had a chance to get their
-ideas across to the whole group, which was the purpose of the lightning talks.
-Overall, this seemed to be successful at the time, in terms of both the lightning
-talks and the breakout groups, and the discussion of sustainability also led
-to interesting and useful results. However, the challenge that we have discovered
-since WSSSPE2 is that it is very hard to continue the breakout groups'
-activities.  The WSSSPE2 participants were willing to dedicate their time to
-the groups while they were at the meeting, but afterwards, they have gone
-back to their (paid) jobs.  We need to determine how to tie the WSSSPE
-breakout activities to people's jobs, so that they feel that continuing them
-is a higher priority than it is now, perhaps through funding the participants,
-or through funding coordinators for each activity, or perhaps by getting
-the workshop participants to agree to a specific schedule of activities during the
-workshop.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2015-10-27 WSSSPE3/report/wg_appendix_software_credit.tex
Finished draft of credit appendix

@@
 Eventually, once standardizing software citations, the goal would be to return to establishing different roles\slash contributions for this credit.
+Following this decision, the group identified two necessary actions to move forward:
+\begin{enumerate}
+    \item standardizing a citation file or some other form of metadata associated with software, and
+    \item standardizing the way to cite software (used directly) in papers.
+\end{enumerate}
+For both of these actions, a number of ongoing efforts were identified and discussed.
+
+\subsubsection{Software Citation Metadata}
+
+At a minimum, the metadata needed for software citation (aside from the DOI for the software itself) includes:
+\begin{itemize}
+    \item Contributors, in the form of name and ORCID,
+    \item Software dependencies, in the form of DOI, and
+    \item Other people and artifacts that would be cited or acknowledged in a paper.
+\end{itemize}
+This information would then be contained in a citation file as part of 
+the, e.g., GitHub repository.
+The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
+metadata schemes in JSON and XML for scientific software and code, and 
+implementing transitive credit via JSON-LD~\cite{wssspe2_katz}.
+Some questions arose about how this information would be stored for 
+closed-source software.
+
+As one mechanism for constructing accurate contributor lists from existing 
+project contributors, the group discussed associating GitHub accounts---as 
+well as accounts on Bitbucket, CodePlex, and other repositories for 
+open-source scientific software---accounts with ORCID.
+However, a (quick) response from GitHub (via Arfon Smith) indicated that 
+this might not be possible in the near future: ``GitHub doesn't have any 
+plans to allow ORCID accounts to be associated with GitHub user accounts.''
+
+\subsubsection{Citing Software in Publications}
+
+Although far from a standard practice, examples of citing software in 
+publications can be found in various scientific communities---notably, 
+representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
+The group recommended collecting similar examples from other communities, then 
+developing a software citation principles document in concert with the 
+FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details) 
+following the model of the FORCE11 Data Citation Principles 
+document~\cite{DataCitation2014}.
+
+The group also discussioned briefly whether software used directly in 
+a publication---whether to perform simulation, analysis, or as a dependency 
+for new software developed---should be distinguished from other references 
+due to the dependence of the study on these research artifacts.
+Suggestions included a separate list of citations (with DOIs) for software 
+and other research objects that serve this sort of ``vital'' role.
+Similar recommendations were made by the credit breakout group at 
+WSSSPE2~\cite{WSSSPE2}.
+
+Finally, although a discrete task from the software citation, significant 
+discussion focused on ensuring software citations are indexed in the same 
+manner as publications---allowing the construction of a corresponding 
+software citation network.
+Currently, software releases can receive citable DOIs via 
+Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
+FigShare\footnote{FigShare: \url{https://figshare.com}}; however, these 
+citations are not processed by indexers such as Web of Science, Scopus, 
+or Google Scholar.
+Thus, either in parallel or following the primary task, the group will 
+need to reach out to these organizations 
+Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
+during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
+also does not yet have an internal identifier for software or data (but 
+needs\slash plans to add this support).
+According to Taylor, they prefer a ``software article'' with the usual 
+article metadata (e.g., authors, citations), and mentioned Zenodo as an 
+example---this proposal seemed to align with group discussions.
+Taylor also mentioned another benefit of the software and associated DOI 
+on GitHub: in addition to a citation, could obtain statistics on 
+usage/downloads/forks.
 
@@
 
-There currently is no standard/accepted mechanism for citing software or receiving credit for software (akin to publications/citations).
+Modern scientific research relies considerably on software, and ensuring 
+that developers of such scientific software receive credit for their efforts 
+will encourage additional creation and maintenance.
+There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
 Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
@@
 
-Obstacle: indexers (e.g., Scopus, Web of Science, Google Scholar) don't currently support software/data document types or DataCite DOIs.
+Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, 
+Google Scholar) do not currently support software/data document types or 
+DataCite DOIs.
+Therefore, even with univeral association of scientific software with DOIs 
+and standardizing practices for citing software in publications
 
-Obstacle: no standard mechanism for software to cite other software.
-
-Obstacle: for tenure, there is no standard for software products to be included in policy,
-sometimes not even within one university: hard to change globally.
+Although the working group discussions at WSSSPE3 did not focus much on the 
+topic of tenure, the group recognized that there is no standard policy---even 
+within a single university---for software products to be included in 
+promotion and tenure dossiers.
+Thus, it may be difficult to encourage valuing software contributions across 
+the United States or United Kingdom, and globally; furthermore, stakeholders 
+are typically not tenured and thus may not be influential enough.
 
@@
 \subsection{Key Next Steps}
+\label{SC:next-steps}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \subsection{Plan for Future Organization}
+\label{SC:plan}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
-These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo\footnote{Zenodo, \url{https://zenodo.org}}, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
+These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo or FigShare, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
 

### 2015-10-28 WSSSPE3/report/wg_main_software_credit.tex
draft SC main section completed

--- 2015-09-30 WSSSPE3/report/wg_main_legacy_SW.tex
@@
-\subsection{Topic} \todo{change the title to your topic}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Software Credit Working Group}
+\label{sec:software-credit}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
 \subsubsection{Why it is important}
-\todo{short text here}
+
+Science and engineering research increasingly relies on software, but there is currently no standard mechanism for citing software or receiving credit for developing software akin to receiving credit via citations for writing papers.
+Standardizing software citations offers one route to establishing such a mechanism.
+Software is currently eligible for DOI assignment, but DOI metadata fields are not well tuned for software compared to publications. 
+Some software providers apply for DOIs but it is still not widely adopted. 
+Also, there is no mechanism to cite software dependencies within software in the same way papers cite supporting prior work.
 
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+
+Publishing Software Working Group (\S\ref{sec:publishing-software}): publishing a software paper offers one existing mechanism for receiving credit, and further developing new publishing concepts for software will strengthen our activities.
+
+A number of groups external to WSSSPE (although with some overlapping members) are also focused on aspects of software credit, including the FORCE11 Software Citation Working Group (see plans for coordination below).
+In addition, a Software Credit workshop\footnote{London Software Credit workshop: \url{www.software.ac.uk/software-credit}} convened in London on October 19, following the conclusion of WSSSPE3.
+See \S\ref{sec:appendix_SW_credit} for more detailed discussion of related activities.
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
+
+The group discussed a number of topics related to software credit, including a contributorship taxonomy, software citation metadata, standards for citing software in publications, and increasing the value of software in academic promotion and tenure reviews.
+Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} used in the entertainment industry, the group decided to prioritize software citation.
+This decision was motivated by the idea that standardizing citations for software would introduce some initial credit for developers, and later the quantification of credit could be refined based on concepts such as transitive credit~\cite{wssspe2_katz,Katz:2014_tc}.
+
+The majority of the remaining discussion focused on standardizing (1) the metadata necessary for software to be cited and (2) the mechanism for citing software in publications.
+In addition, discussions also oriented around the indexing of software citations necessary for establishing a software citation network either integrated with the existing paper citation ecosystem or complementary to it.
+See \S\ref{sec:appendix_SW_credit} for a more detailed summary of the working group's discussion on these topics.
 
 \subsubsection{Plans}
-\todo{short text here - not bullets}
+
+The group already merged with the FORCE11 Software Citation Working Group (SCWG), and their efforts will focus (over the next six to nine months) on developing a document describing principles for software citation.
+Following publication of that document, the working group will focus on outreach to key groups (e.g., journals, publishers, indexers, professional societies).
+Longer-term plans include working with indexers to ensure that software citations are indexed and pursuing an open\slash community indexer; these activities may be organized by future FORCE11 working groups.
 
 \subsubsection{Landing Page}
-\todo{link to landing page}
+
+Since near-term efforts will be shifting to the FORCE11-SCWG, we direct interested readers to that group's existing landing page\footnote{FORCE11-SCWG landing page, \url{https://www.force11.org/group/software-citation-working-group}} and GitHub repository\footnote{FORCE11-SCWG GitHub page,  \url{https://github.com/force11/force11-scwg}}.

### 2015-10-28 WSSSPE3/report/wg_appendix_software_credit.tex
small fixes/comments to software credit appendix

@@
 \end{itemize}
-This information would then be contained in a citation file as part of 
-the, e.g., GitHub repository.
+\katznote{is a name/title also needed?}
+This information would then be contained in a citation file, e.g.,  as part of 
+the GitHub repository.
 The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
@@
 developing a software citation principles document in concert with the 
-FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details) 
+FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details), 
 following the model of the FORCE11 Data Citation Principles 
@@
 
-The group also discussioned briefly whether software used directly in 
+The group also discussed briefly whether software used directly in 
 a publication---whether to perform simulation, analysis, or as a dependency 
@@
 
-Finally, although a discrete task from the software citation, significant 
+Finally, although a discrete task from software citations themselves, significant 
 discussion focused on ensuring software citations are indexed in the same 
-manner as publications---allowing the construction of a corresponding 
+manner as publications, allowing the construction of a corresponding 
 software citation network.
@@
 Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
-FigShare\footnote{FigShare: \url{https://figshare.com}}; however, these 
+figshare\footnote{figshare: \url{https://figshare.com}}; however, these 
 citations are not processed by indexers such as Web of Science, Scopus, 
@@
 needs\slash plans to add this support).
-According to Taylor, they prefer a ``software article'' with the usual 
+Taylor said they prefer a ``software article'' with the usual 
 article metadata (e.g., authors, citations), and mentioned Zenodo as an 
-example---this proposal seemed to align with group discussions.
+example -- this proposal seemed to align with group discussions.
 Taylor also mentioned another benefit of the software and associated DOI 
 on GitHub: in addition to a citation, could obtain statistics on 
-usage/downloads/forks.
+usage/downloads/forks, which happens to be what Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.
 
@@
 that developers of such scientific software receive credit for their efforts 
-will encourage additional creation and maintenance.
+will encourage additional creation and maintenance. \katznote{the prev sentence is good, but feels a bit odd in the middle of the appendix, rather than at the beginning}
 There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
@@
 DataCite DOIs.
-Therefore, even with univeral association of scientific software with DOIs 
-and standardizing practices for citing software in publications
+Therefore, even with universal association of scientific software with DOIs 
+and standardizing practices for citing software in publications \katznote{this sentence is missing its ed}
 
-Although the working group discussions at WSSSPE3 did not focus much on the 
+Although this working group's discussions at WSSSPE3 did not focus much on the 
 topic of tenure, the group recognized that there is no standard policy---even 
@@
 Thus, it may be difficult to encourage valuing software contributions across 
-the United States or United Kingdom, and globally; furthermore, stakeholders 
-are typically not tenured and thus may not be influential enough.
+the United States or United Kingdom and globally; furthermore, stakeholders 
+are typically not tenured and thus may not be influential enough. \katznote{to do what?}
 

### 2015-10-28 WSSSPE3/report/wg_appendix_software_credit.tex
Minor changes to SC appendix in response to Dan notes

@@
 \begin{itemize}
+    \item Name of software and DOI,
     \item Contributors, in the form of name and ORCID,
@@
 \end{itemize}
-\katznote{is a name/title also needed?}
 This information would then be contained in a citation file, e.g.,  as part of 
@@
 
-Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, 
-Google Scholar) do not currently support software/data document types or 
-DataCite DOIs.
-Therefore, even with universal association of scientific software with DOIs 
-and standardizing practices for citing software in publications \katznote{this sentence is missing its ed}
+Major obstacles include the fact that indexers (e.g., Scopus, Web of Science,  Google Scholar) do not currently support software/data document types or  DataCite DOIs.
+Therefore, even with universal association of scientific software with DOIs  and standardizing practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
 
-Although this working group's discussions at WSSSPE3 did not focus much on the 
-topic of tenure, the group recognized that there is no standard policy---even 
-within a single university---for software products to be included in 
-promotion and tenure dossiers.
-Thus, it may be difficult to encourage valuing software contributions across 
-the United States or United Kingdom and globally; furthermore, stakeholders 
-are typically not tenured and thus may not be influential enough. \katznote{to do what?}
+Although this working group's discussions at WSSSPE3 did not focus much on the topic of tenure, the group recognized that there is no standard policy---even within a single university---for software products to be included in promotion and tenure dossiers.
+Thus, it may be difficult to encourage valuing software contributions across the United States or United Kingdom and globally; furthermore, stakeholders are typically not tenured and thus may not be influential enough to change the status quo.
 

### 2015-10-28 WSSSPE3/report/wg_main_software_credit.tex
Fixed URL in SC main section

@@
 A number of groups external to WSSSPE (although with some overlapping members) are also focused on aspects of software credit, including the FORCE11 Software Citation Working Group (see plans for coordination below).
-In addition, a Software Credit workshop\footnote{London Software Credit workshop: \url{www.software.ac.uk/software-credit}} convened in London on October 19, following the conclusion of WSSSPE3.
+In addition, a Software Credit workshop\footnote{London Software Credit workshop: \url{http://www.software.ac.uk/software-credit}} convened in London on October 19, following the conclusion of WSSSPE3.
 See \S\ref{sec:appendix_SW_credit} for more detailed discussion of related activities.

### 2015-10-28 WSSSPE3/report/wg_main_software_credit.tex
Fixed appendix refs in SC main section

@@
 In addition, a Software Credit workshop\footnote{London Software Credit workshop: \url{http://www.software.ac.uk/software-credit}} convened in London on October 19, following the conclusion of WSSSPE3.
-See \S\ref{sec:appendix_SW_credit} for more detailed discussion of related activities.
+See Appendix~\ref{sec:appendix_SW_credit} for more detailed discussion of related activities.
 
@@
 In addition, discussions also oriented around the indexing of software citations necessary for establishing a software citation network either integrated with the existing paper citation ecosystem or complementary to it.
-See \S\ref{sec:appendix_SW_credit} for a more detailed summary of the working group's discussion on these topics.
+See Appendix~\ref{sec:appendix_SW_credit} for a more detailed summary of the working group's discussion on these topics.
 

### 2015-10-28 WSSSPE3/report/wg_main_industry_interaction.tex
new draft of WSSSPE 3 case statement
added Industry / Academic Interaction case statement

\subsection{Topic}

Transition Pathways to Sustainable Software: Industry \& Academic Collaboration

\subsubsection{Why it is important}

Most scientific software is produced as a part of grant-funded research projects sponsored by federal governments. If we are interested in the sustainability of scientific software then we need to understand what exactly happens when that sponsorship ends. More than likely, the project and its resulting software will need to undergo some kind of transition in funding and consequently management.

There are a number of potential funding transitions that may occur:

\begin{itemize}
\item A project could be \textbf{refunded}, and development or maintenance of the software continues as planned.
\item A project might located a \textbf{new source of funding} in which case the software may be further developed or simply maintained as before. 
\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source. (e.g. see Howison 2015)
\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc. may change significantly. 
\end{itemize}

We characterize each of these potential changes in funding as "transition pathways" to sustainable software (see similar work by Geels and Schot, 2007).

At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to "transition" from government-funded research projects to industry sponsorship. (This may be an initially awkward phrase - because some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. In such cases, there is still a need to understand how IP and how maintenance of the software is sustained over time)

Although transitions iare often studied under the broad umbrella of 'technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved. In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes / failures.

\subsubsection{Fit with related activities}

In looking at existing literature that addresses industry transitions many reports focus on benefits that accrue to the private sector, or to a government that originally sponsored the research project. This literature does not address the impact that these transitions have on the accessibility, or usability of the software, or the impact that these transitions have on the career of the researchers involved.

Examples of the former scenario (benefits accruing to private sector) are as follows:

\begin{itemize}
\item REF Impact Case Studies: http://impact.ref.ac.uk/CaseStudies/
\item Background of projects funded in the UK: http://gtr.rcuk.ac.uk/ 
\item Dowling Review from the UK - addresses complexity of work between these two communities: http://www.raeng.org.uk/policy/dowling-review 
\item Pathway to Impact - UK report  - 2 pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic )
\end{itemize}

\subsubsection{Discussion}

Our work at WSSSPE 3 included the following activities (described in detail below): 1. Brainstorming goals for this type of research, 2. Imagining potential outcomes of completing a set of case studies on this topic, and 3. Generating a set of working definitions for some of the broad concepts we are describing.

\textbf{GOAL}
\emph{What is the goal of doing research on transition pathways?}

Can we identify collaborations that have occurred and try to understand which were successful, which were unsuccessful, and what factors contributed to these successes/failures?

Can we determine what each partner wants to get out of such a collaboration? Why would industry be interested in collaborating with academia?  Why would academia be interested in collaborating with industry?

How could we design a study that focused on the impact of the software in undergoing this type of transition?

\textbf{Potential outcomes}

Complete a set of case studies which look at successful and unsuccessful transitions between academic researchers and industry. This might address one of each of the transition types (described below). Succesful transitions are described as those that lead to either weak or strong sustainability (also defined below).

Create a generalizable framework that might allow for the study of different transition pathways (other than academia -> industry).

\textbf{General Definitions}

We characterize transitions in the following way:

\begin{itemize}
\item Handoff model - i.e. academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
\item Co-Production Model - i.e. industry and academia interact throughout development of the project.
\item Sponsorship Model  - i.e. academia writes and maintains the software, the industry contributes funding for the development/maintenance of software. In this example industry is also likley a user of the software.
\item Spinoff model- transition to For-profit company owned by or in collaboration original developers; or, Non-profit formed by or in collaboration with developers. 
\end{itemize}

We characterize sustainability in the following ways:

\begin{itemize}
\item Weak Sustainability: Software continues to be accessible, useful, and usable. 
\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e. issued non-restrictive open-source license). (see Becker et al., 2014 for extended discussion of weak vs. strong sustainability.)
\end{itemize}

\subsubsection{Plans}

Plans for carrying forward are currently unclear - this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.

\subsubsection{Landing Page}

\subsection{Works Referrenced}

\begin{itemize}
\item Becker, C., Chitchyan, R., Duboc, L., Easterbrook, S., Mahaux, M., Penzenstadler, B., ... \& Betz, S. (2014). The Karlskrona manifesto for sustainability design. arXiv preprint arXiv:1410.6968.
\item Geels, F. W., \& Schot, J. (2007). Typology of sociotechnical transition pathways. Research policy, 36(3), 399-417.
\item Howison, J. (2015). Sustaining scientific infrastructures: transitioning from grants to peer production. Proceedings of the 2015 iConference. 
\end{itemize}

### 2015-10-28 WSSSPE3/report/wg_appendix_software_credit.tex
minor tweaks in credit appendix

@@
 \item Daniel S.\ Katz -- University of Chicago, Argonne National Laboratory
-\item Frank Löffler -- Louisiana State University
+\item Frank L\"{o}ffler -- Louisiana State University
 \item Abby Cabunoc Mayes -- Mozilla Science Lab
@@
 \begin{itemize}
-    \item Name of software and DOI,
-    \item Contributors, in the form of name and ORCID,
-    \item Software dependencies, in the form of DOI, and
+    \item Name of software,
+    \item DOI for software,
+    \item Contributors, in the form of names and ORCIDs,
+    \item Software dependencies, in the form of DOIs, and
     \item Other people and artifacts that would be cited or acknowledged in a paper.
@@
 Major obstacles include the fact that indexers (e.g., Scopus, Web of Science,  Google Scholar) do not currently support software/data document types or  DataCite DOIs.
-Therefore, even with universal association of scientific software with DOIs  and standardizing practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
+Therefore, even with universal association of scientific software with DOIs  and standized practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
 
-Although this working group's discussions at WSSSPE3 did not focus much on the topic of tenure, the group recognized that there is no standard policy---even within a single university---for software products to be included in promotion and tenure dossiers.
+Although this working group's discussions at WSSSPE3 did not focus much on the topic of tenure, the group recognized that there is no standard policy---generally even within a single university---for software products to be included in promotion and tenure dossiers.
 Thus, it may be difficult to encourage valuing software contributions across the United States or United Kingdom and globally; furthermore, stakeholders are typically not tenured and thus may not be influential enough to change the status quo.

### 2015-10-28 WSSSPE3/report/wg_main_industry_interaction.tex
small fixes and comments in industry interactions section

@@
 
-We characterize each of these potential changes in funding as "transition pathways" to sustainable software (see similar work by Geels and Schot, 2007).
+We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot, 2007).
 
-At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to "transition" from government-funded research projects to industry sponsorship.
+At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship.
 (This may be an initially awkward phrase - because some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia.
@@
 
-Although transitions iare often studied under the broad umbrella of 'technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved.
+Although transitions are often studied under the broad umbrella of `technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved.
 In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes / failures.
@@
 \begin{itemize}
-\item REF Impact Case Studies: http://impact.ref.ac.uk/CaseStudies/
-\item Background of projects funded in the UK: http://gtr.rcuk.ac.uk/ 
-\item Dowling Review from the UK - addresses complexity of work between these two communities: http://www.raeng.org.uk/policy/dowling-review 
+\item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
+\item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
+\item Dowling Review from the UK - addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
 \item Pathway to Impact - UK report  - 2 pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic )
@@
 
-Our work at WSSSPE 3 included the following activities (described in detail below): 1.
+Our work at WSSSPE3 included the following activities (described in detail below): 1.
 Brainstorming goals for this type of research, 2.
@@
 This might address one of each of the transition types (described below).
-Succesful transitions are described as those that lead to either weak or strong sustainability (also defined below).
+Successful transitions are described as those that lead to either weak or strong sustainability (also defined below).
 
@@
 \begin{itemize}
-\item Handoff model - i.e. academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
-\item Co-Production Model - i.e. industry and academia interact throughout development of the project.
-\item Sponsorship Model  - i.e.
-academia writes and maintains the software, the industry contributes funding for the development/maintenance of software.
-In this example industry is also likley a user of the software.
-\item Spinoff model- transition to For-profit company owned by or in collaboration original developers; or, Non-profit formed by or in collaboration with developers. 
+\item Handoff model - i.e., academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
+\item Co-Production Model - i.e., industry and academia interact throughout development of the project.
+\item Sponsorship Model  - i.e., academia writes and maintains the software, the industry contributes funding for the development/maintenance of software.
+In this example industry is also likely a user of the software.
+\item Spinoff model- transition to for-profit company owned by or in collaboration original developers; or, non-profit formed by or in collaboration with developers. 
 \end{itemize}
@@
 \item Weak Sustainability: Software continues to be accessible, useful, and usable. 
-\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e.
-issued non-restrictive open-source license).
+\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e., issued non-restrictive open-source license).
 (see Becker et al., 2014 for extended discussion of weak vs.
@@
 
-\subsection{Works Referrenced}
+\todo{where should someone go who want to know more about this and perhaps wants to contribute?}
+
+\subsection{Works Referenced}
+
+\todo{please add these to the bib file and cite them from there.}
 

### 2015-10-28 WSSSPE3/report/wg_appendix_software_credit.tex
Minor fix to minor tweak

@@
 Major obstacles include the fact that indexers (e.g., Scopus, Web of Science,  Google Scholar) do not currently support software/data document types or  DataCite DOIs.
-Therefore, even with universal association of scientific software with DOIs  and standized practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
+Therefore, even with universal association of scientific software with DOIs and standardized practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
 

### 2015-10-28 WSSSPE3/report/wg_main_legacy_SW.tex
adding brief summary of legacy group, and removing corresponding appendix

@@
-\subsection{Topic} \todo{change the title to your topic}
+\subsection{Legacy Software}
 
-\subsubsection{Why it is important}
-\todo{short text here}
+This group met only briefly, for one period on the first day.
+They discussed that it is difficult to define legacy code because there is so much stigma associated with the term.
+At some point there will be more difficulty and resources wasted trying to keep legacy software supported, but it will eventually be too expensive compared to how much it would be to just rebuild the software or kill it.
+Most of the group members were not able to attend on the second day, and those who were able to attend joined other groups.
 
-\subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
-
-\subsubsection{Discussion}
-\todo{short-ish text here}
-
-\subsubsection{Plans}
-\todo{short text here - not bullets}
-
-\subsubsection{Landing Page}
-\todo{link to landing page}
+%\subsubsection{Why it is important}
+%\todo{short text here}
+%
+%\subsubsection{Fit with related activities}
+%\todo{short text here - can include links/cites}
+%
+%\subsubsection{Discussion}
+%\todo{short-ish text here}
+%
+%\subsubsection{Plans}
+%\todo{short text here - not bullets}
+%
+%\subsubsection{Landing Page}
+%\todo{link to landing page}

### 2015-10-28 WSSSPE3/report/WSSSPE3_review.tex
adding brief summary of legacy group, and removing corresponding appendix
fixes for the fact that some parts of 5 will not have corresponding appendices

@@
 (Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
-occurred in each the working groups
+occurred in each of the working groups that made substantial progress.
 (Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.)
@@
 
+\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}
+
 \input{wg_main_best_practices}
@@
 \input{wg_appendix_industry_interaction}
-\input{wg_appendix_legacy_SW}
+%\input{wg_appendix_legacy_SW}
 \input{wg_appendix_eng_design}

### 2015-10-29 WSSSPE3/report/awardees.tex
Add travel awardees.

Alice Allen & Astrophysics Source Code Library\\
Steven Brandt &  Louisiana State University\\
Jeffrey Carver &  University of Alabama\\
Emily Chen & NCSA, University of Illinois\\
Sou-Cheng Choi &  Illinois Institute of Technology\\
%Yuhan Ding &  Illinois Institute of Technology\\
Lorraine Hwang &  CIG,  UC Davis\\
Ray Idaszak &  RENCI, University of North Carolina at Chapel Hill\\
%Frank Löffler &  Louisiana State University\\
%Abigail Cabunoc Mayes &  Mozilla Science Lab\\
Pate Motter &  University of Colorado\\
Kyle Niemeyer &  Oregon State University\\
Birgit Penzenstadler &  California State University Long Beach\\
Bernadette Randles &  UCLA\\
%Ilian Todorov &  STFC Daresbury Laboratory\\
Nic Weber &  University of Washington iSchool

### 2015-10-29 WSSSPE3/report/WSSSPE3_review.tex
Add travel awardees.
Fixed some punctuation in Intro

@@
 (Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
-occurred in each of the working groups that made substantial progress.
-(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.)
+occurred in each of the working groups that made substantial progress 
+(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).
 
@@
 storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
-presented his wishlist of questions to be answered: 
+presented his wishlist of questions to be answered in the future: 
 %
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Travel Award Recipients}  \label{sec:awardees}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\todo{Do we want email addresses here?}
+The following is list of travel award recipients for the WSSSPE3 workshop.
+
+{\scriptsize
+\begin{longtable}{lll}
+\input{awardees}
+\end{longtable}
+}
+
 \input{wg_appendix_best_practices}

### 2015-10-29 WSSSPE3/report/wg_appendix_publishing_SW.tex
Some formatting fixes to SW pub appendix

@@
 
-Steven R. Brandt\footnote{email: \href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}} will serve as the point of contact for this working group.
+Steven R.\ Brandt\footnote{email: \href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}} will serve as the point of contact for this working group.
 
@@
 A tentative first cut at the list containing executable papers identified the following:
-
 \begin{itemize}
-\item ACM Transactions on Mathematical Software (TOMS) - which provides the extra step
+\item ACM Transactions on Mathematical Software (TOMS): provides the extra step
  of having reviewers validate the code which was submitted with the publication.
-\item The Mathematica Journal - which publishes Mathematica notebooks (with equations,
+\item The Mathematica Journal: publishes Mathematica notebooks (with equations,
 figures, etc.) directly.
-\item O'Reily Media has announced that it plans to make IPython Notebooks a first-class
+\item O'Reily Media: announced that it plans to make IPython Notebooks a first-class
  authoring environment for their publishing program alongside their existing mechanisms.
-\item Nature is offering a list of notebooks published alongside more traditional articles,
+\item Nature: offering a list of notebooks published alongside more traditional articles,
  and is looking at ways to make these documents more official. There are, in fact, a
- number of journals that offer "electronic supplements" to the more traditionally published
+ number of journals that offer ``electronic supplements'' to the more traditionally published
  static articles.
-\item There is also a list of "reproducible academic publications" maintained here:
-  \url{https://github.com/ipython/ipython/wiki/A-gallery-of-interesting-IPython-Notebooks#reproducible-academic-publications}
-\item KBase narratives build on IPython/Jupyter notebooks to build publications that are
+\item IPython: maintained list of ``reproducible academic publications''~\cite{ipython-pubs}.
+\item KBase: narratives build on IPython\slash Jupyter notebooks to build publications that are
   reproducible, and can be commented or annotated.
@@
 The group also discussed future possibilities for this new publication format which might
-provide advantages.
-
+provide advantages:
 \begin{itemize}
-\item Journals could be built around an existing, widely-used framework thereby reducing
+\item Journals could be built around an existing, widely used framework thereby reducing
   the burden of studying code on the part of reviewers (common bits of infrastructure
@@
 \item So many options for publication currently exist that good science may be getting
-  lost in the noise. Would some sort of "upvote" mechanism be of value?
+  lost in the noise. Would some sort of ``upvote'' mechanism be of value?
 \item Some sort of Replicated Computation Results badge could be made available to

### 2015-10-29 WSSSPE3/report/orgcom.tex

Daniel S. Katz &  University of Chicago \& Argonne National Laboratory, USA\\
Gabrielle Allen &  University of Illinois Urbana-Champaign, USA\\
Neil Chue Hong &  Software Sustainability Institute.  University of Edinburgh, UK\\
Sou-Cheng (Terrya) Choi &  Illinois Institute of Technology \& University of Chicago, USA\\
Sandra Gesing &  University of Notre Dame,  USA\\
Lorraine Hwang &   University of California, Davis, USA\\
Manish Parashar &  Rutgers University, USA\\
Erin Robinson &  Foundation for Earth Science, USA (local organizer)\\
Matthew Turk &  University of Illinois Urbana-Champaign, USA\\
Colin C. Venters &  University of Huddersfield, UK

### 2015-10-29 WSSSPE3/report/wg_appendix_software_credit.tex
software credit typo

@@
 
-The discussion then evolved into philosophical questions about the importantance or reliance of a particular work on prior science, materials, or software---in other words, is there a difference between depending on prior scientific advances and depending on particular software (or experimental equipment)?
+The discussion then evolved into philosophical questions about the importance or reliance of a particular work on prior science, materials, or software---in other words, is there a difference between depending on prior scientific advances and depending on particular software (or experimental equipment)?
 Multiple contributors converged on the conclusion that unique capabilities require some additional credit.

### 2015-10-29 WSSSPE3/report/WSSSPE3_review.tex
Organizers.
adding info about the new appendices in the intro

@@
 groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
-Finally, the report includes a list of registered attendees
-(Appendix~\ref{sec:attendees}) and longer descriptions of the activities that
+Lists of the organizing committee (Appendix~\ref{sec:orgcom}), the registered
+attendees (Appendix~\ref{sec:attendees}), and the travel award recipients
+(Appendix~\ref{sec:awardees}) are provided. Finally, the report includes longer
+descriptions of the activities that
 occurred in each of the working groups that made substantial progress 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Organizing Committee}  \label{sec:orgcom}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\todo{Do we want email addresses here?}
+The following is the list of organizers of WSSSPE3.
+
+{\scriptsize
+\begin{longtable}{lll}
+\input{orgcom}
+\end{longtable}
+}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Attendees}  \label{sec:attendees}

### 2015-10-29 WSSSPE3/report/wg_appendix_software_credit.tex

@@
 
-During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} as a potential model for scientific software.
+During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for scientific software.
 That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
@@
 publications can be found in various scientific communities---notably, 
-representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
+representative samples can be found in astronomy~\cite{astronomy_SW_examples} and biology~\cite{Howison2015}.
 The group recommended collecting similar examples from other communities, then 
@@
 Currently, software releases can receive citable DOIs via 
-Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
-figshare\footnote{figshare: \url{https://figshare.com}}; however, these 
+Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these 
 citations are not processed by indexers such as Web of Science, Scopus, 

### 2015-10-29 WSSSPE3/report/wg_main_industry_interaction.tex
fixing a subsection that should have been a subsubsection

@@
-\subsection{Topic}
-
-Transition Pathways to Sustainable Software: Industry \& Academic Collaboration
+\subsection{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration}
 
@@
 \item A project might located a \textbf{new source of funding} in which case the software may be further developed or simply maintained as before. 
-\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source.
-(e.g.
-see Howison 2015)
-\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc. may change significantly. 
+\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source (e.g., see Howison~\cite{howison_sustaining_2015}).
+\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc.\ may change significantly. 
 \end{itemize}
 
-We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot, 2007).
+We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot~\cite{Geels:2007}).
 
 At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. 
-(This may be an initially awkward phrase - because some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia.
+(This may be an initially awkward phrase---some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. 
 In such cases, there is still a need to understand how IP and how maintenance of the software is sustained over time)
@@
 Although transitions are often studied under the broad umbrella of `technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved.
-In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes / failures.
+In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes\slash failures.
 
@@
 Examples of the former scenario (benefits accruing to private sector) are as follows:
-
 \begin{itemize}
@@
 \item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
-\item Dowling Review from the UK - addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
-\item Pathway to Impact - UK report  - 2 pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic )
+\item Dowling Review from the UK: addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
+\item Pathway to Impact - UK report: two pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic)
 \end{itemize}
@@
 
-Our work at WSSSPE3 included the following activities (described in detail below): 1.
-Brainstorming goals for this type of research, 2.
-Imagining potential outcomes of completing a set of case studies on this topic, and 3.
-Generating a set of working definitions for some of the broad concepts we are describing.
+Our work at WSSSPE3 included the following activities (described in detail below): (1) brainstorming goals for this type of research, (2) imagining potential outcomes of completing a set of case studies on this topic, and (3) generating a set of working definitions for some of the broad concepts we are describing.
 
@@
 
-We characterize transitions in the following way:
-
+We characterize transitions in the following ways:
 \begin{itemize}
-\item Handoff model - i.e., academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
-\item Co-Production Model - i.e., industry and academia interact throughout development of the project.
-\item Sponsorship Model  - i.e., academia writes and maintains the software, the industry contributes funding for the development/maintenance of software.
+\item Handoff model: academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
+\item Co-Production Model: industry and academia interact throughout development of the project.
+\item Sponsorship Model: academia writes and maintains the software, the industry contributes funding for the development\slash maintenance of software.
 In this example industry is also likely a user of the software.
-\item Spinoff model- transition to for-profit company owned by or in collaboration original developers; or, non-profit formed by or in collaboration with developers. 
+\item Spinoff model: transition to for- or non-profit company owned by or in collaboration with original developers. 
 \end{itemize}
@@
 We characterize sustainability in the following ways:
-
 \begin{itemize}
@@
 \item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e., issued non-restrictive open-source license).
-(see Becker et al., 2014 for extended discussion of weak vs.
-strong sustainability.)
 \end{itemize}
+See Becker et al.~\cite{Becker:2014} for extended discussion of weak vs.\ strong sustainability.
 
@@
 
-Plans for carrying forward are currently unclear - this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.
+Plans for carrying forward are currently unclear---this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.
 
@@
 \todo{where should someone go who want to know more about this and perhaps wants to contribute?}
-
-\subsection{Works Referenced}
-
-\todo{please add these to the bib file and cite them from there.}
-
-\begin{itemize}
-\item Becker, C., Chitchyan, R., Duboc, L., Easterbrook, S., Mahaux, M., Penzenstadler, B., ... \& Betz, S. (2014). The Karlskrona manifesto for sustainability design. arXiv preprint arXiv:1410.6968.
-\item Geels, F. W., \& Schot, J. (2007). Typology of sociotechnical transition pathways. Research policy, 36(3), 399-417.
-\item Howison, J. (2015). Sustaining scientific infrastructures: transitioning from grants to peer production. Proceedings of the 2015 iConference. 
-\end{itemize}

### 2015-10-29 WSSSPE3/report/wg_main_software_credit.tex

@@
 The group discussed a number of topics related to software credit, including a contributorship taxonomy, software citation metadata, standards for citing software in publications, and increasing the value of software in academic promotion and tenure reviews.
-Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry\footnote{Entertainment Identifier Registry: \url{http://eidr.org}} used in the entertainment industry, the group decided to prioritize software citation.
+Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry~\cite{EIDR} used in the entertainment industry, the group decided to prioritize software citation.
 This decision was motivated by the idea that standardizing citations for software would introduce some initial credit for developers, and later the quantification of credit could be refined based on concepts such as transitive credit~\cite{wssspe2_katz,Katz:2014_tc}.

### 2015-10-29 WSSSPE3/report/awardees.tex
travel update

@@
 Sou-Cheng Choi &  Illinois Institute of Technology\\
-%Yuhan Ding &  Illinois Institute of Technology\\
+Yuhan Ding &  Illinois Institute of Technology\\
 Lorraine Hwang &  CIG,  UC Davis\\
 Ray Idaszak &  RENCI, University of North Carolina at Chapel Hill\\
-%Frank Löffler &  Louisiana State University\\
+%Frank L\"{o}ffler &  Louisiana State University\\
 %Abigail Cabunoc Mayes &  Mozilla Science Lab\\

### 2015-10-29 WSSSPE3/report/wg_main_metrics.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Useful Metrics for Scientific Software}
\label{sec:software-metrics}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsubsection{Why it is important}

Metrics for scientific software are important for tenure and promotion, scientific impact, discovery, reducing duplication, serving as a basis for potential industrial interest in adopting software, and making a case for new or continued funding.  However, there is no commonly-used standard for collecting or presenting metrics, nor is it known if there is a common set of metrics for scientific software.  It is imperative that scientific software stakeholders understand that it is useful to collect metrics.

\subsubsection{Fit with related activities}

The group discussion here focused on identifying existing frameworks and activities for scientific software metrics.  The group identified the following related activities:

\begin{itemize}

\item
\href{https://geodynamics.org/cig/dev/best-practices/}{(Computational Infrastructure for Geodynamics: Software Development Best Practices)}

\item
\href{https://docs.google.com/document/d/1cgUDH3RxrfsLotWhKKOrXUnaYFhrtjcV1TDRkFtwQKI/edit}{(WSSSPE3 Breakout Session: How can we measure the impact of a code on research, and its value to the community?)}

\item
\href{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/edit#heading=h.5lah0hp73q99}{(2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics)}

\item
\href{https://docs.google.com/document/d/1uDim5bw8rBuubmtaUrz5Eh35NxzDgivmmdXhVzDs3tc/edit}{(2015 NSF SI2 PI Workshop Breakout Session on Software Metrics)}

\item
\href{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/edit#slide=id.g5e66ec9f2_027}{(NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics)}

\item
\href{http://www.software.ac.uk/software-evaluation-guide}{(U.K. Software Sustainability Institute Software Evaluation Guide)}

\item
\href{http://www.software.ac.uk/blog/2013-04-09-five-stars-research-software}{(U.K. Software Sustainability Institute Blog post: The five stars of research software)}

\item
\href{http://figshare.com/articles/Minimal_information_for_reusable_scientific_software/1112528}{(Minimal information for reusable scientific software)}

\item
\href{http://equipment.data.ac.uk/}{(EPSRC-funded Equipment Data Search Site)}

\item
\href{http://www.canarie.ca/software/}{(Canarie Research Software: Software to accelerate discovery)}

\item
\href{https://science.canarie.ca/researchmiddleware/platforms/list/main.html}{(Canarie Research Software: Research Software Platform Registry)}

\item
\href{https://collaboration.canarie.ca/elgg/file/view/2471/research-platform-support-for-the-canarie-registry-and-monitoring-system-revision-3}{(collaboration@CANARIE post: platform support for Canarie registry and monitoring system)}

\item
\href{https://collaboration.canarie.ca/elgg/file/view/2453/research-service-support-for-canarie-registry-and-monitoring-system-revision-7}{(collaboration@CANARIE post: service support for Canarie registry and monitoring system)}

\item
\href{https://www.openhub.net/}{(BlackDuck Open HUB)}

\item
\href{https://www.innovationpolicyplatform.org/frontpage}{(Innovation Policy Platform)}

\end{itemize}

\subsubsection{Discussion}

The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper.  The group discussed many questions related to useful metrics for scientific software including addressing if there is a common set of metrics that can be filtered in some way, can metrics be fit into a common template, which metrics would be the most useful for each stakeholder, which metrics are the most helpful and how would we assess this, how are metrics monitored, and many more.  A more complete bulleted list of these questions can be found in Appendix H.  Next, a roadmap for how to proceed was discussed including creating a set of milestones and tasks.  The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in order to email out a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.  This survey would be created by one of the student group members.  Similarly, it was suggested that a software metrics survey be sent to the UK SFTF and TRDF software projects to ask them what metrics would be useful to report.  The remainder of the discussion focused mainly on the creation of a white paper on this topic.  This resulted in a paper outline and writing assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE, or JORS.

\subsubsection{Plans}

The main plans for the group going forward are the creation of a white paper on the topic of useful metrics for scientific software.  The authoring of this white paper would happen in parallel with the creation of a survey by the group with the survey results to be incorporated in the white paper.  The timeline for completion of the white paper is approximately one year targeting venues discussed in the previous section.

\subsubsection{Landing Page}

The group will use Google Docs to proceed with the authoring of the white paper in lieu of a group project landing page.  Links to the resulting white paper will be provided when completed.

### 2015-10-29 WSSSPE3/report/awardees.tex
update on awardees

@@
 %Frank L\"{o}ffler &  Louisiana State University\\
-%Abigail Cabunoc Mayes &  Mozilla Science Lab\\
+Abigail Cabunoc Mayes &  Mozilla Science Lab\\
 Pate Motter &  University of Colorado\\

### 2015-10-29 WSSSPE3/report/awardees.tex
Updating my own affliations.

@@
 Emily Chen & NCSA, University of Illinois\\
-Sou-Cheng Choi &  Illinois Institute of Technology\\
+Sou-Cheng Choi & NORC at the University of Chicago \&  Illinois Institute of Technology\\
 Yuhan Ding &  Illinois Institute of Technology\\

### 2015-10-29 WSSSPE3/report/orgcom.tex
Updating my own affliations.

@@
 Neil Chue Hong &  Software Sustainability Institute.  University of Edinburgh, UK\\
-Sou-Cheng (Terrya) Choi &  Illinois Institute of Technology \& University of Chicago, USA\\
+Sou-Cheng (Terrya) Choi &  NORC at the University of Chicago \& Illinois Institute of Technology, USA\\
 Sandra Gesing &  University of Notre Dame,  USA\\

### 2015-10-30 WSSSPE3/report/wg_main_publishing_SW.tex
Fixed to do, and tweaked wording.

--- 2015-10-22 WSSSPE3/report/wg_main_publishing_SW.tex
@@
 
-This group explored the value of executable papers (papers whose content includes
-the code needed to produce their own results), and other forms of publishing which
-include dynamic electronic content. Transitioning to this type of publication offers
-possibilities of addressing, or partially addressing sustainability concerns 
-such as reproducibility, transitive credit, and best practices.
+This group explored the value of executable papers (papers whose content
+includes the code needed to produce their own results), and other forms of
+publishing which include dynamic electronic content. Transitioning to this type
+of publication offers possibilities of addressing, or partially addressing
+sustainability concerns such as reproducibility, software credit, and best
+practices.
 
@@
 
-Reproducibility: Part of the purpose of these venues is to (at least partially)
-address the reproducibility issue by making the paper itself recompute its own
-results.
+\textbf{Reproducibility}: Part of the purpose of these executable paper venues
+is to (at least partially) address the reproducibility issue by making the
+paper itself recompute its own results.
 
-Software Credit (\S\ref{sec:software-credit}): Since these forms of publishing must make their sources explicit,
-they should be easier to trace even if appropriately worded credit for software
-is not provided. In addition, these notebooks make it possible to provide/define
-additional metadata to make the tracing of credit clearer. In addition, attributions
-could be added to citations to identify whether a paper extends a result, verifies it,
-contradicts it, etc.
+\textbf{Software Credit (\S\ref{sec:software-credit})}: Since these forms of
+publishing must make their sources explicit in order to execute, they should be
+easier to trace even if appropriately worded credit for software is not
+provided. In addition, these notebooks make it possible to provide/define
+additional metadata to make the tracing of credit clearer. Finally,
+attributions could be added to citations to identify whether a paper extends a
+result, verifies it, contradicts it, etc.
 
-Best Practices: Because an executable paper showcases the code, and the code itself
-is subject to the review process, authors are more likely to pay attention to coding
-practices. In addition, because the paper must explain what the code does, better
-documentation is likely to be achieved.
+\textbf{Best Practices (\S\ref{sec:best-practices})}: Because an executable paper
+showcases the code, and the code itself is subject to the review process,
+authors are more likely to pay attention to coding practices. In addition,
+because the paper must explain what the code does, better documentation is
+likely to be achieved.
 

### 2015-11-02 WSSSPE3/report/wg_main_funding_spec_expert.tex
Make minor changes to remind me which files to editå

--- 2015-10-12 WSSSPE3/report/wg_main_best_practices.tex
@@
-\subsection{White paper/journal paper about best practices in developing sustainable software}
+\subsection{Research Programming Expertise -- how to fund it and sustain it.}
 
@@
 
-[15 Nov] Introduction and scope finished
-[15 Nov] Sections assigned
-[31 Jan] Analysing funding possibilities for survey
-[31 Jan] First versions of section
-[15 Feb] Distribution to WSSSPE community
-[31 Mar] Final version of white paper
-[30 Apr] Submission of peer-reviewed paper?
-
-\begin{enumerate}
-\item Introduction and Scope of White Paper 
-\item Related Work
-\item Case Studies
-\begin{enumerate} 
-\item PeTSC
-\item NWCHEM
-\item CIG
-\end{enumerate}
-\item Community Related Practices
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Governance and management
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Funding Related
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Metrics for sustainability
-\item Tools
-\item Conclusions
-\end{enumerate}
-
 \subsubsection{Landing Page}

### 2015-11-02 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Create wg_appendix_funding_spec_expert.tex
Former-commit-id: 02a53b0fb3c9753516884b9b5cbeccaa1ec38281
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Make minor changes to remind me which files to editå
Temporarily paste-in Mark's notes from the workshop as-is.

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Funding Specialist Expertise Group Discussion}
+\label{sec:appendix_funding_spec_expert}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{add POC here}
+
+This is where the bit I need to write will go.
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+
+Don Middleton
+Josh Greenburg: SLOAN
+James Hetherington: University College London j.hetherington@ucl.ac.uk
+Lindsay Powers: The HDF Group, lpowers@hdfgroup.org
+Mark Miller
+Dan Sellars: CANARIE
+
+\begin{itemize}
+\item name -- affiliation
+\item name2 -- affiliation2
+\end{itemize}
+
+\subsection{Temporary direct paste-in of notes from Workshop}
+
+Our charge: document successful funding models as a means of creating a conversation that can be used in finding funding for new efforts;
+Don Middleton
+Josh Greenburg: SLOAN
+James Hetherington: University College London j.hetherington@ucl.ac.uk
+Lindsay Powers: The HDF Group, lpowers@hdfgroup.org
+Mark Miller
+Dan Sellars: CANARIE: Canada; like internet 2, they make some funding available to do projects.
+Fund a PI /team to produce a platform to do the work.
+Cultural barrier is steep for communicating across barriers, and scientists seeing the developers as “other”.
+SO collaboration, and tools to fund/encourage/motivate collaboration are extremely important.
+Projects that engage communities on an “at will” basis
+One model is to engage scientists in the development process as part of a community effort, and that may be largely unfunded.
+Models that are community driven can be very organic, so new contributions are possible from unexpected places.
+Funded engagement Models
+It is usually more effective and efficient to do development where there is funding to support the collaboration. 
+Co-funding on a single proposal.
+One way to do this is by co-funding individuals from both camps, that is domain scientists and developers are funded together in a single proposal.
+At NCAR, University partnerships are required for submission of proposals, so collaboration is an essential part of grant submission, and this will tend to bring developers and scientists together.
+This is certainly an effective method for promoting project based/project scale work.
+We aren’t sure how this requirement could be imposed globally, but it should and probably often is required by any given RFP for SW development.
+One example of a success is the development of semantics and linked data in support of Ocean Sciences.
+An Earthcube funded project pairs domain scientists with RSEs and has been successful as the semantics attached have increased data use and discovery significantly.
+Developer support from a resource provider to augment existing funding.
+We discussed another fundamental strategy used by several organizations, where a group that wants to promote use of a new resource, be it internet (Canarie) or one or more compute engines (XSEDE (US) and Archer (UK)) fund a developer team that works directly with scientists from outside the organization to promote the use of the new resource.
+The Canarie group (Canada) accepts proposals for providing services to broad communities, integrate people who are doing things that are complementary, the goal is to make the available stack more robust and richer for everyone.
+They offer short rounds of funding that can have as a key metric creating some useful functionality that shows a diversity of input and draws from across disciplines, then more funding could follow.
+This could apply within or across institutions.
+The UK allows this kind of collaboration to support the creation of scientific software for the large supercomputing resource called Archer.
+The Archer project fund a group of developers, and domain scientists can apply to the group for developer time to help them launch a new tool on Archer.
+The US XSEDE project does essentially the same thing.
+The extended collaboration service (ECS) is a set of developers who are paid with XSEDE funding, and are on “permanent” staff.
+When PIs request allocations on XSEDE resources, there is a finite pool of developer time that can be awarded, typically for one year only, and at partial effort, typically 20% or so.
+The finite time allowed provides motivation for the scientist and their group to work closely the developer and to become educated in what the developer is doing, so they can sustain the effort, once the ECS period is over.
+This funding mechanism is highly efficient for scientific problems, because the developer pool assembled by the research providers are, by definition, expert in the characteristics of their specific resource, and can very quickly assess the scientists needs, and what it will take to implement software that meets the user’s needs.
+Moreover, when an organization assembles a developer pool, there will be diversity in the talent pool, so issues that require multiple kinds of expertise can be addressed by distinct individuals, again with maximum efficiency.
+Promoting institutional support for research science developers
+The Sloan Foundation is currently funding data science engineers, who work in the context of other SW developers at University of Washington.
+These scientists work in the e-Science Studio/dataScience Studio, and they help a group of graduate students in solving their problems in data science/data management.
+During Fall and Spring, a 10-week incubator program allows students to work two days a week to work on a data-intensive science project.
+Some fraction of the developer time is dedicated to the developers personal interests as well as instruction.
+Piloting and documenting what works and what doesn’t.
+The goal for Sloan is to provide success stories, to provide demonstrable value in the presence of data scientists on university staff.
+These stories are the basis for arguments to the host organization.
+This is an effort to create awareness of the value of research scientist developers.
+Embedding with scientists, and adding spare capacity is critical to making the innovation possible.
+This model is essentially to argue for permanent budget lines to support data scientists as part of university staff hires, just as with core facilities.
+This could become a fee-for-service model requested by grant funding, just as DNA sequencing is for core facilities, if it becomes apparent that this gives competitive advantage to the University’s research effort.
+Researchers wont be able to do this without seed money to get it going.
+James succeeded at his institution which was building cyberinsfrastructure.
+Three posts were funded by the university.
+This is budget line for professional developers.
+Currently developer support for individual projects is awarded by a panel.
+Researchers can contribute funds that are left over from other awards where people leave prematurely.
+This is happening currently.
+Some people committed to using funds in an existing award.
+There is a UK fellowship for replicating this basic concept: the award is for salary for one position as team leader,  and a first team member; these awards are for five years duration (woohoo!).
+There are other RSEs in other UK institutions.
+These may engage in labor exchange via  informal work sharing agreements between institutions.
+Three month engagement preferred, but not everyone works that way some prefer quick in and done modes.
+To make such a program work, it seems it  must be university wide, to be broad enough to have the ftes needed funded.
+But there is not one mind on this, it maybe that the broadest approach is not idea.
+Perhaps a specialization focus on developers to support physical or biological sciences may be preferable, if the customer base is large enough.
+How to aggregate enough work to make it sustainable.
+How do you make sure you have the right staff skills?
+This will be an issue in a smaller group, and it will be exacerbated if the approach is general rather than focused on a domain emphasis where the number of possible skills required is less.
+The optimum use of resources seems to be 40% effort, on a project; two projects per developer.
+Two developers per project seems to be ideal, in the sense that software development is enhanced by two pairs of eyes..
+Collaborative computational projects (CCP) to support community networking: domain specific communities put forward proposals that are a priority of the community as a whole: biosimulation plasma PHYSICS QUANTUM PHYSICS.
+% years of funding.
+Custodians of community codes.
+Software support.
+help these are incubators.
+Add new features.
+Software of the future, open to all.
+Ccp submits.
+15 ftes for development across all CCPs.
+GET IAIN.
+4% success rate for software of the future project, high end computing consortia, in domains of CCPs.
+Co-design; the art of bringing it into existence together.
+The mission is codesigning 5 projects: the collaboration did work, but not as intimately as it might have.
+Incentive structures are still based on getting papers out, and the benefits to the postdocs are career focused.
+RSE fellowships: 204 applications, 5 funded.
+Intention was to fund applicants that demonstrated broad benefit to community/ies.
+How to better incentivize the collaboration to serve a broader community?
+Challenge: Scientists do not appreciate the engineering issues, whereas the cs person does not necessarily have the understanding of the domain.
+Possibility of a training course for research software engineering, a course could be run to teach people to manage a grant that is focused on software development.
+Non-hands on software carpentry for PIs.
+What are the next steps:
+Have we actually been successful?
+What are the metrics for success?
+Publications are tracked in James organization.
+How to evaluate the productivity gains that are made as a result of this engagement with James’ group?
+How to evaluate what did not go wrong because of this intervention?
+Notre Dame and LSU groups who are RSE groups.
+NCEAS/SOCSYNC; some RSE support; NESCENT/NIMBIOS also.
+The power of the label? Getting the word out about RSE support using the label.
+
+Will research science developers be required, in the long run?
+One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.
+And this brings up the flip side question, “Do scientists really want to be writing code?”
+Career Path.
+We also had a little discussion about how to make a career path for research developers.
+It need not be solely an academic enterprise, but tenure is always problematic for people of this class.
+There is a need for more senior people to have a track.
+Retention is hard, because these people are highly in demand by the economy, so retention is hard.
+People who have a PhD in science and enter industry, may return for diverse reasons.
+Things we haven’t started talking about:
+Other above the campus models for funding;
+Industry:academic partnerships
+
+Follow-on Discussion at GCE15
+rse.ac.uk
+People: Lorraine Hwang, Smon Trigger, Nancy Wilkins-Diehr,  Alex Vyshkov, Sandra, Keynote person from Zooniverse
+
+Good example: Notre Dame (14 research programmer), LSU, Princeton, 
+SDSC -- actually even in national labs there are inefficiencies with per-project jobs rather than develop pool -- potential issue is degree of control. 
+Matt Trunnel - CIO of Broad institute boston - Chris Dwan - they had individual RSEs, then aggregated. 
+James Taylor’s team  - John’s Hopkins
+Org models - OVPR, CS, Physics, IT, Library
+IT Coalescence issues.
+Maybe Yale?
 

### 2015-11-02 WSSSPE3/report/wg_main_funding_spec_expert.tex
Temporary direct paste-in of Mark's notes from the Workshop as-is

@@
 \subsection{Research Programming Expertise -- how to fund it and sustain it.}
+
+\subsubsection{Temporary direct paste-in}
+
+What is the summary of the opportunity/challenge and key obstacles?
+The issue is how to fund and create a career path for Research Software Engineer’s.
+RSE’s are going to be key to promoting co-development, which seems like a highly efficient mechanism for scientific software development.
+Funding is “hard” for individual PIs to find support for RSEs. How can we make it easier for folks to access RSEs, and in so doing create an RSE niche in the University.
+Goal: Demonstrate value of RSEs to both funding agencies and institutions, leading to creation of positions for RSEs.
+ We have defined some models where RSEs have been successfully funded, and where they have contributed to scientific software development, and believe this can be extended more broadly.
+
+Next Steps?
+White paper, add groups outside the UK that are RSE groups.
+Poll for where there are existing RSE groups; how they are funded.
+James will post existing RSE groups on University College London - RSDG website and ask others to identify themselves.
+Use this information to collect more information on successful funding models.
+Promoting communication and even resource sharing among RSE groups.
+Recognition that this is a spectrum, and subject to interpretation, i.e. professionals versus development out of hobby or necessity.
+Will we work on the next steps?
+Lindsay, Mark, and James will continue to push this agenda forward by turning the notes into a white paper. We will work as peers. 
+How will we organize ourselves going forward?
+Mark will be the scribe, and will work to be sure the notes are translates into a white paper draft as we go forward.
+What else do we need?
+We would like to create a list of other RSE organizations. How do we define and identify these organizations? How will we support this activity?
+How can we credibly document the impact of RSEs on research progress; not just how things go faster, but what discoveries are enabled faster, and what setbacks were avoided because good engineering practices were followed.
+What are key Milestones?
+Prepare a white paper to solidify our understanding of the issue, and to bring it more into the public consciousness.
+Create a forum to promote information and resource sharing across RSEs.
+Maybe host an RSE meeting for exchange of thoughts and experiences.
+What is the size shape and structure of funding needed?
+Support would be required to create a resource sharing structure for RSEs
+Support would be required to hold a meeting of RSE groups.
 

### 2015-11-02 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Some work towards a draft of RSE funding section

@@
 
-\todo{add POC here}
-
-This is where the bit I need to write will go.
+James Hetherington\footnote{email: \href{mailto:j.hetherington@ucl.ac.uk}{j.hetherington@ucl.ac.uk}}
+will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
 
@@
 
-Don Middleton
-Josh Greenburg: SLOAN
-James Hetherington: University College London j.hetherington@ucl.ac.uk
-Lindsay Powers: The HDF Group, lpowers@hdfgroup.org
+Don Middleton -- National Center for Atmospheric Research
+Joshua Greenberg -- Alfred P. Sloan Foundation
+James Hetherington -- University College London
+Lindsay Powers -- The HDF Group
 Mark Miller
@@
 \begin{itemize}
-\item name -- affiliation
-\item name2 -- affiliation2
+\item Don Middleton -- National Center for Atmospheric Research
+\item Josh Greenburg -- affiliation2
 \end{itemize}

### 2015-11-02 WSSSPE3/report/wg_main_funding_spec_expert.tex
Some work towards a draft of RSE funding section

@@
 \subsubsection{Why it is important}
-\todo{short text here}
+
+Research Software Engineers -- those who contribute to science and scholarship
+through computer programming -- are an important part of the team
+needed to deliver 21st century research. However, existing academic structures
+and systems of funding do not effectively fund and sustain these skills.
+The resulting high levels of turnover, and inappropriate incentivisation,
+are a significant contributing factor to low levels of reliability and
+readability observed in scientific software.
 
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+
+One solution to this problem would be a successful solution to the problems of
+obtaining academic credit for programming work. If this were solved, traditional
+academic funding models and career tracks would properly sustain this
+contribution to research.
 

### 2015-11-03 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Complete first draft of main report on RSE funding

@@
 
-Don Middleton -- National Center for Atmospheric Research
-Joshua Greenberg -- Alfred P. Sloan Foundation
-James Hetherington -- University College London
-Lindsay Powers -- The HDF Group
-Mark Miller
-Dan Sellars: CANARIE
-
 \begin{itemize}
 \item Don Middleton -- National Center for Atmospheric Research
-\item Josh Greenburg -- affiliation2
+\item Joshua Greenberg -- Alfred P. Sloan Foundation
+\item James Hetherington -- University College London
+\item Lindsay Powers -- The HDF Group
+\item Mark A. Miller -- Sand Diego Supercomputer Center
+\item Dan Sellars -- CANARIE
 \end{itemize}
@@
 \subsection{Temporary direct paste-in of notes from Workshop}
+
+What is the summary of the opportunity/challenge and key obstacles?
+The issue is how to fund and create a career path for Research Software Engineer’s.
+Goal: Demonstrate value of RSEs to both funding agencies and institutions, leading to creation of positions for RSEs.
+ We have defined some models where RSEs have been successfully funded, and where they have contributed to scientific software development, and believe this can be extended more broadly.
+
+Next Steps?
+White paper, add groups outside the UK that are RSE groups.
+Poll for where there are existing RSE groups; how they are funded.
+James will post existing RSE groups on University College London - RSDG website and ask others to identify themselves.
+Use this information to collect more information on successful funding models.
+Promoting communication and even resource sharing among RSE groups.
+Recognition that this is a spectrum, and subject to interpretation, i.e. professionals versus development out of hobby or necessity.
+Will we work on the next steps?
+Lindsay, Mark, and James will continue to push this agenda forward by turning the notes into a white paper. We will work as peers.
+How will we organize ourselves going forward?
+Mark will be the scribe, and will work to be sure the notes are translates into a white paper draft as we go forward.
+What else do we need?
+We would like to create a list of other RSE organizations. How do we define and identify these organizations? How will we support this activity?
+How can we credibly document the impact of RSEs on research progress; not just how things go faster, but what discoveries are enabled faster, and what setbacks were avoided because good engineering practices were followed.
+What are key Milestones?
+Prepare a white paper to solidify our understanding of the issue, and to bring it more into the public consciousness.
+Create a forum to promote information and resource sharing across RSEs.
+Maybe host an RSE meeting for exchange of thoughts and experiences.
+What is the size shape and structure of funding needed?
+Support would be required to create a resource sharing structure for RSEs
+Support would be required to hold a meeting of RSE groups.
 

### 2015-11-03 WSSSPE3/report/wg_main_funding_spec_expert.tex
Complete first draft of main report on RSE funding

--- 2015-11-02 WSSSPE3/report/wg_main_funding_spec_expert.tex
@@
-\subsection{Research Programming Expertise -- how to fund it and sustain it.}
+\subsection{Research Programming Expertise -- how to fund it?}
 
 \subsubsection{Why it is important}
-\todo{short text here}
+
+Research Software Engineers -- those who contribute to science and scholarship
+through computer programming -- are an important part of the team
+needed to deliver 21st century research. However, existing academic structures
+and systems of funding do not effectively fund and sustain these skills.
+The resulting high levels of turnover, and inappropriate incentivisation,
+are a significant contributing factor to low levels of reliability and
+readability observed in scientific software.
+
+For effective development of software for advanced research, it is necessary
+that researchers can work with scientific programmers who are able to understand
+the research domain sufficiently to be able to build meaningful software. This
+requires a collaborative approach -- programmers fully engaged in the research
+context, co-developing software with domain academics.
 
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+
+One solution to this problem would be a successful solution to the problems of
+obtaining academic credit for programming work. If this were solved, traditional
+academic funding models and career tracks would properly sustain this
+contribution to research. However, in the absence of guaranteed success in that
+stream, this working party seeks to look to an alternate solution; the
+development and funding of dedicated research programming roles. We also noted
+that measures of impact and success for research programming
+groups, as well as for individual research software engineers, will be required
+in order to make the case for continued funding.
+
+A contributing factor to the problem is the devastating effect on an academic
+publication record resulting from time in industry; such postings often
+develop exactly the skills that research programmers need, yet returns to
+university positions following an industry role are penalised by the current
+structures.
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
+
+It is hard for an individual PI in a University or College
+to support dedicated research programming
+resources, as the need for, and funding for, these activities is intermittent
+within the research cycle. To sustain this capacity, therefore, it is
+necessary to aggregate this work across multiple research groups.
+
+One solution is to fund dedicated programming roles for major research software
+projects at national laboratories or other non-educational institutions. This
+solution is working well for many well-used scientific codebases.
+However, much of the body of software for which this expertise is needed sits
+within the longer tail of research in universities. Therefore, we argue that
+institutions need to develop hybrid academic-technical roles for this capacity
+who work with more than one PI, rather than the traditional RA role within a
+single group. This could be centrally, perhaps within research computing
+organisations which have traditionally supported university cyberinfrastructure,
+or more locally.
+
+Having convinced ourselves that this would be a positive innovation, we were
+then faced with the specific question of how to fund the initiation of this
+activity. A self-sustaining research software group will support itself
+through collaborations with PIs in the normal grant process, with PIs choosing
+to fund some amount of research programming effort through grants in the usual
+way. However, to bootstrap such a function to a level where it has the
+reputation and client base to be self-sustaining requires seed investment.
+
+This might come from unversities themselves (this was the model that led to the
+creation of the group in University College London), but more likely, seed
+funding needs to come from research councils (as with the Research Software
+Engineering Fellowship provided by the UK Engineering and Physical Sciences
+Research Council). We therefore recommend that funding organisations consider
+how they might provide such seed funding.
+
+We finally noted that the institutional case for such groups is made easier
+by having successful examples to point to. In the UK, a collective effort to
+identify this community, with individuals clearly stating ``I'm a research
+software engineer'', has been important to the campaign. It will be useful to
+the global effort to similarly identify emerging research software organisations,
+and also, importantly, to identify longer-running research software groups,
+which have in some cases had a long running \emph{sui-generis} existence, but which
+now can be identified as part of a wider solution.
 
 \subsubsection{Plans}
-\todo{short text here - not bullets}
+
+We will therefore seek to gather a list of groups which are prepared to
+self-identify as research software engineering groups. We will collect information
+as to the organisational models under which these groups function, and how they
+are funded.
 
 \subsubsection{Landing Page}
-\todo{link to landing page}
+
+TBD, probably into the UKRSE site.

### 2015-11-04 WSSSPE3/report/WSSSPE3_review.tex
question about lightning talks section
Acknowledge support.
Section 3: Keynote: Response to Dan's comment after Matt explained.

@@
 
-Turk said that sustaining scientific software really meant to him
-conducting scientific inquiries, often by some specific software, and sustaining
-the people we care about, our careers, and the future of our fields. According
-to Turk, we all have an invested stake in sustaining scientific software. Again
-it came down to the top-down versus bottom-up notion that we can have anointed
-specific software packages to be sustained at the expense of new grassroots
-developments. \katznote{I don't understand this last sentence.}
+Turk said that sustaining scientific software really meant to him conducting
+scientific inquiries, often by some specific software, and sustaining the people
+we care about, our careers, and the future of our fields. According to Turk, we
+all have an invested stake in sustaining scientific software. Hence, having
+``sustained'' projects can suffocate new projects, for which we need to keep a
+lookout.
+%This is a straw man concept, but it's something to look out for, since organizations will act in their own interest.
 
@@
 \todo{need to update for WSSSPE3}
+\todo{KEN: Should these be bolded?}
 \begin{comment}
@@
 expressed in this material are those of the author(s) and do not necessarily
-reflect the views of the National Science Foundation.
+reflect the views of the National Science Foundation. Choi's work is supported
+in part by the National Science Foundation research grant DMS-1522687 and
+a WSSSPE3 travel award.
 

### 2015-11-05 WSSSPE3/report/wg_appendix_metrics.tex
Create wg_appendix_metrics.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Update wg_appendix_metrics.tex

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Metrics Working Group Discussion}
+\label{sec:appendix_metrics}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
+Gabrielle Allen \footnote{email: \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} will serve as the point of contact for this working group.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{itemize}
+\item Gabrielle Allen -- National Center for Supercomputing Applications
+\item Emily Chen -- University of Illinois at Urbana-Champaign
+\item Neil Chue Hong -- U.K. Software Sustainability Institute
+\item Ray Idaszak -- RENCI, University of North Carolina at Chapel Hill
+\item Iain Larmou -- Engineering and Physical Sciences Research Council
+\item Bernie Randles -- University of California, Los Angeles
+\item Dan Sellars -- Canarie
+\item Fraser Watson -- National Solar Observatory
+\end{itemize}
 
@@
 
+The following summary of the group’s discussion represents the Useful Metrics for Scientific Software working group’s discussion during the WSSSPE3 workshop.
+The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper.
+The group discussed many questions related to useful metrics for scientific software as follows:
+
+\begin{itemize}
+
+\item
+Is there a common set of metrics, that can be filtered in some way
+
+\begin{itemize}
+\item
+        Does this create a large cost
+\end{itemize}
+
+\item
+Can we fit metrics into a common template (i.e. for collection, for description)
+
+\item
+Which would be the most useful ones
+
+\begin{itemize}
+\item
+        Which ones would be most useful for each stakeholder
+\end{itemize}
+
+\item
+Which ones are the most helpful, and how would we assess this
+
+\item
+How do you monitor
+
+\begin{itemize}
+\item
+        Self-checking - if monitoring is done in the open, then people will call out cheats
+\end{itemize}
+
+\item
+Should this be published with the software metadata
+
+\begin{itemize}
+\item
+        This would make it easier for public to see the metadata
+
+\item
+        However, there is no commonly used standard (DOAP is a good standard but not widely adopted)
+
+\item
+        The Open Directory Project (ODP) metadata is available for UK infrastructure
+\end{itemize}
+
+\item
+Intersection of most useful and easiest to collect should be explored
+
+\item
+How can students/curricula be used as part of a solution
+
+\item
+Number of users could be affected by other metrics e.g. by accessibility
+
+\item
+Assume metrics are collected properly, but guidance should be provided none-the-less
+
+\item
+Continuum for each metric
+
+\begin{itemize}
+\item
+        Ideal situation is the absolute minimum, so that people can decide on their own what the cost versus usefulness tipping point is
+\end{itemize}
+
+\item
+Maturity plays a part
+
+\begin{itemize}
+\item
+        Consider different metrics brackets for different maturity levels
+\end{itemize}
+
+\item
+What are we using metrics for
+
+\begin{itemize}
+\item
+        What software should I use if I have a choice
+
+\item
+        Where should funders place funding for best impact (e.g. funding two-star software versus three-star) and where there are gaps
+
+\item
+        How to promote reduction of code proliferation
+
+\item
+        Metrics used for software panels to provide information
+
+\item
+        Metrics used for finding problems in their systems
+
+\end{itemize}
+
+\item
+Can we use metrics to help people identify the best codes as part of a community effort
+
+\end{itemize}
+
+\smallskip
+\noindent
+Next, a roadmap for how to proceed was discussed including creating a set of milestones and tasks as follows:
+
+\begin{itemize}
+\item
+Can we create a roadmap and milestones for this activity
+
+\item
+Need to come up with a set of tasks
+
+\item
+Go to NSF Software Infrastructure for Sustained Innovation (SI2) projects asking them what metrics they defined, and how useful they were
+
+\begin{itemize}
+\item
+        Milestone: Create report which assesses the metrics that SI2 projects used
+
+\begin{itemize}
+\item
+                Ask SI2 PIs to say what metrics they said they would use (copied from proposal)
+
+\item
+                Ask SI2 PIs what numbers they reported
+
+\item
+                Ask SI2 PIs what they would have changed
+
+\item
+                A UIUC student on the project will work on this
+\end{itemize}
+
+\item
+        Tentatively aim for March 2016
+\end{itemize}
+
+\item
+Do something similar for UK SFTF and TRDF software projects to ask them what would be useful metrics to report; also eCSE projects
+
+\begin{itemize}
+\item
+        Compare these to understand if there were any implications for including metrics
+\end{itemize}
+
+\item
+Collaboratively create plan and documentation for doing this
+
+\begin{itemize}
+\item
+        Give some examples from group members projects, and aim to build out some of the measurement continuum
+
+\item
+        Road-test at the WSSSPE4 meeting
+\end{itemize}
+
+\item
+Collect the various frameworks together and do a comparison summary
+
+\end{itemize}
+
+\smallskip
+\noindent
+The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in order to email out a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.
+This survey would be created by one of the student group members.
+Similarly, it was suggested that a software metrics survey be sent to the UK SFTF and TRDF software projects to ask them what metrics would be useful to report.
+The remainder of the discussion focused mainly on the creation of a white paper on this topic.
+This resulted in a paper outline and writing assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE, or JORS.
+
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+
+The following opportunies, challenges, and obstacles were discussed:
+
+\begin{itemize}
+\item
+Metrics are important for:
+
+\begin{itemize}
+\item
+        Tenure and promotion
+
+\item
+        Scientific impact
+
+\item
+        Discovery
+
+\item
+        Reducing duplication
+
+\item
+        Basis for potential industrial interest in adopting software
+
+\item
+        Make case for funding
+\end{itemize}
+
+\item
+No commonly-used standard for collecting or presenting metrics
+
+\item
+We don’t know if there is a common set of metrics
+
+\item
+We have to persuade projects that it is useful to collect metrics
+
+\end{itemize}
 
@@
 
+The following next steps were discussed:
+
+\begin{itemize}
+\item
+Skype phone call to coordinate shortly after the conclusion of the WSSSPE3 workshop
+
+\item
+Get started on IRB at University of Illinois Urbana-Champaign in anticipation of SI2 project survey (may need more thought into survey)
+
+\item
+Get started on white paper and associated survey
+
+\end{itemize}
+
 \subsection{Plan for Future Organization}
+
+The following plan for future organization was discussed:
+
+\begin{itemize}
+\item
+Our group has created a white paper outline with sections assigned to the above individuals, plus see Section 2 response above for timeline
+
+\item
+Organizing coordinating phone calls
+
+\end{itemize}
 
@@
 
+The following list of what else is needed was discussed:
+
+\begin{itemize}
+\item
+IRB approval/exemption needed for surveys, collecting data
+
+\item
+Coordination with 2016 NSF SI2 PI workshop organizing committee to possibly piggyback on this event to offer survey to attendees in advance
+
+\item
+Coordination (mail communication, info page etc), via WSSSPE github or?
+
+\end{itemize}
+
 \subsection{Key Milestones and Responsible Parties}
 
+The following items were discussed as a roadmap for the production of a white paper:
+
+\begin{enumerate}
+\item
+October – November 2015: IRB paperwork as appropriate completed (Gabrielle Allen and Emily Chen)
+
+\item
+October – December 2015: Draft white paper sections 1-3 (the paper outline has initial writing assignments)
+
+\item
+October – December 2015: Run surveys and collect information
+
+\begin{enumerate}
+\item
+        Piggy back on planning for 2016 NSF SI2 PIs meeting to be held Feb 16-17, 2016
+\end{enumerate}
+
+\item
+January – February 2016: Analyze results of data collection from projects
+
+\item
+March – April 2016: Draft sections 4-7 of the white paper
+
+\item
+May 2016: Draft section 8-9 of the white paper
+
+\item
+May – June 2016: Get initial feedback from members of the community and revise
+
+\item
+Est. July 2016: By time of next CFP for WSSSPE have complete draft of white paper
+
+\item
+Est. Sept – Oct 2016: Responses to white paper submitted to WSSSPE4
+
+\end{enumerate}
+
 \subsection{Description of Funding Needed}
+
+Funding needs were not discussed in this working group and it was thought that this could potentially be revisited down the road.

### 2015-11-05 WSSSPE3/report/wg_appendix_metrics.tex
small edits

@@
 
-Gabrielle Allen \footnote{email: \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} will serve as the point of contact for this working group.
+Gabrielle Allen\footnote{email: \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} will serve as the point of contact for this working group.
 
@@
 
-The following summary of the group’s discussion represents the Useful Metrics for Scientific Software working group’s discussion during the WSSSPE3 workshop.
+The following summary of the group's discussion represents the Useful Metrics for Scientific Software working group's discussion during the WSSSPE3 workshop.
 The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper.
@@
 
-The following opportunies, challenges, and obstacles were discussed:
+The following opportunities, challenges, and obstacles were discussed:
 
@@
 \item
-We don’t know if there is a common set of metrics
+We don't know if there is a common set of metrics
 
@@
 \item
-October – November 2015: IRB paperwork as appropriate completed (Gabrielle Allen and Emily Chen)
+October -- November 2015: IRB paperwork as appropriate completed (Gabrielle Allen and Emily Chen)
 
 \item
-October – December 2015: Draft white paper sections 1-3 (the paper outline has initial writing assignments)
+October -- December 2015: Draft white paper sections 1-3 (the paper outline has initial writing assignments)
 
 \item
-October – December 2015: Run surveys and collect information
+October -- December 2015: Run surveys and collect information
 
@@
 \item
-        Piggy back on planning for 2016 NSF SI2 PIs meeting to be held Feb 16-17, 2016
+        Piggyback on planning for 2016 NSF SI2 PIs meeting to be held Feb 16-17, 2016
 \end{enumerate}
@@
 \item
-January – February 2016: Analyze results of data collection from projects
+January -- February 2016: Analyze results of data collection from projects
 
 \item
-March – April 2016: Draft sections 4-7 of the white paper
+March -- April 2016: Draft sections 4-7 of the white paper
 
@@
 \item
-May – June 2016: Get initial feedback from members of the community and revise
+May -- June 2016: Get initial feedback from members of the community and revise
 
@@
 \item
-Est. Sept – Oct 2016: Responses to white paper submitted to WSSSPE4
+Est. Sept -- Oct 2016: Responses to white paper submitted to WSSSPE4
 

### 2015-11-05 WSSSPE3/report/awardees.tex
adding an awardee

--- 2015-10-29 WSSSPE3/report/awardees.tex
@@
 Ray Idaszak &  RENCI, University of North Carolina at Chapel Hill\\
-%Frank L\"{o}ffler &  Louisiana State University\\
+Frank L\"{o}ffler &  Louisiana State University\\
 Abigail Cabunoc Mayes &  Mozilla Science Lab\\

### 2015-11-05 WSSSPE3/report/wg_main_user_community.tex
Update wg_main_user_community.tex
added some related activities

@@
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+%\todo{short text here - can include links/cites}
+
+Mozilla Science maintains a "'Working Open' Project Guide" (http://mozillascience.github.io/leadership-training). From the introduction:
+\begin{quote}
+Working openly with contributors enables your
+    community to learn how to build and collaborate together. This
+    document is a guideline on how to work openly and involve others
+    in your projects with Mozilla. We want to help you engage your
+    community in a way that encourages contributors and builds other
+    leaders.
+  \end{quote}
+
+Several books have been written about software communities:
+\begin{itemize}
+\item "Art of Community" by Jono Bacon. We could consider distilling this for scientific software.
 

### 2015-11-05 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Close to complete on RSE sections

--- 2015-11-02 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
@@
 
-Don Middleton -- National Center for Atmospheric Research
-Joshua Greenberg -- Alfred P. Sloan Foundation
-James Hetherington -- University College London
-Lindsay Powers -- The HDF Group
-Mark Miller
-Dan Sellars: CANARIE
+The group at WSSPE:
 
@@
 \item Don Middleton -- National Center for Atmospheric Research
-\item Josh Greenburg -- affiliation2
+\item Joshua Greenberg -- Alfred P. Sloan Foundation
+\item James Hetherington -- University College London
+\item Lindsay Powers -- The HDF Group
+\item Mark A. Miller -- San Diego Supercomputer Center
+\item Dan Sellars -- CANARIE
 \end{itemize}
 
-\subsection{Temporary direct paste-in of notes from Workshop}
+This was further enhanced by additional discussions at the following
+GCE15 conference:
 
-Our charge: document successful funding models as a means of creating a conversation that can be used in finding funding for new efforts;
-Don Middleton
-Josh Greenburg: SLOAN
-James Hetherington: University College London j.hetherington@ucl.ac.uk
-Lindsay Powers: The HDF Group, lpowers@hdfgroup.org
-Mark Miller
-Dan Sellars: CANARIE: Canada; like internet 2, they make some funding available to do projects.
-Fund a PI /team to produce a platform to do the work.
-Cultural barrier is steep for communicating across barriers, and scientists seeing the developers as “other”.
-SO collaboration, and tools to fund/encourage/motivate collaboration are extremely important.
-Projects that engage communities on an “at will” basis
-One model is to engage scientists in the development process as part of a community effort, and that may be largely unfunded.
-Models that are community driven can be very organic, so new contributions are possible from unexpected places.
-Funded engagement Models
-It is usually more effective and efficient to do development where there is funding to support the collaboration.
-Co-funding on a single proposal.
-One way to do this is by co-funding individuals from both camps, that is domain scientists and developers are funded together in a single proposal.
-At NCAR, University partnerships are required for submission of proposals, so collaboration is an essential part of grant submission, and this will tend to bring developers and scientists together.
-This is certainly an effective method for promoting project based/project scale work.
-We aren’t sure how this requirement could be imposed globally, but it should and probably often is required by any given RFP for SW development.
-One example of a success is the development of semantics and linked data in support of Ocean Sciences.
-An Earthcube funded project pairs domain scientists with RSEs and has been successful as the semantics attached have increased data use and discovery significantly.
-Developer support from a resource provider to augment existing funding.
-We discussed another fundamental strategy used by several organizations, where a group that wants to promote use of a new resource, be it internet (Canarie) or one or more compute engines (XSEDE (US) and Archer (UK)) fund a developer team that works directly with scientists from outside the organization to promote the use of the new resource.
-The Canarie group (Canada) accepts proposals for providing services to broad communities, integrate people who are doing things that are complementary, the goal is to make the available stack more robust and richer for everyone.
-They offer short rounds of funding that can have as a key metric creating some useful functionality that shows a diversity of input and draws from across disciplines, then more funding could follow.
-This could apply within or across institutions.
-The UK allows this kind of collaboration to support the creation of scientific software for the large supercomputing resource called Archer.
-The Archer project fund a group of developers, and domain scientists can apply to the group for developer time to help them launch a new tool on Archer.
-The US XSEDE project does essentially the same thing.
-The extended collaboration service (ECS) is a set of developers who are paid with XSEDE funding, and are on “permanent” staff.
-When PIs request allocations on XSEDE resources, there is a finite pool of developer time that can be awarded, typically for one year only, and at partial effort, typically 20% or so.
-The finite time allowed provides motivation for the scientist and their group to work closely the developer and to become educated in what the developer is doing, so they can sustain the effort, once the ECS period is over.
-This funding mechanism is highly efficient for scientific problems, because the developer pool assembled by the research providers are, by definition, expert in the characteristics of their specific resource, and can very quickly assess the scientists needs, and what it will take to implement software that meets the user’s needs.
-Moreover, when an organization assembles a developer pool, there will be diversity in the talent pool, so issues that require multiple kinds of expertise can be addressed by distinct individuals, again with maximum efficiency.
-Promoting institutional support for research science developers
-The Sloan Foundation is currently funding data science engineers, who work in the context of other SW developers at University of Washington.
-These scientists work in the e-Science Studio/dataScience Studio, and they help a group of graduate students in solving their problems in data science/data management.
-During Fall and Spring, a 10-week incubator program allows students to work two days a week to work on a data-intensive science project.
-Some fraction of the developer time is dedicated to the developers personal interests as well as instruction.
-Piloting and documenting what works and what doesn’t.
-The goal for Sloan is to provide success stories, to provide demonstrable value in the presence of data scientists on university staff.
-These stories are the basis for arguments to the host organization.
-This is an effort to create awareness of the value of research scientist developers.
-Embedding with scientists, and adding spare capacity is critical to making the innovation possible.
-This model is essentially to argue for permanent budget lines to support data scientists as part of university staff hires, just as with core facilities.
-This could become a fee-for-service model requested by grant funding, just as DNA sequencing is for core facilities, if it becomes apparent that this gives competitive advantage to the University’s research effort.
-Researchers wont be able to do this without seed money to get it going.
-James succeeded at his institution which was building cyberinsfrastructure.
-Three posts were funded by the university.
-This is budget line for professional developers.
-Currently developer support for individual projects is awarded by a panel.
-Researchers can contribute funds that are left over from other awards where people leave prematurely.
-This is happening currently.
-Some people committed to using funds in an existing award.
-There is a UK fellowship for replicating this basic concept: the award is for salary for one position as team leader,  and a first team member; these awards are for five years duration (woohoo!).
-There are other RSEs in other UK institutions.
-These may engage in labor exchange via  informal work sharing agreements between institutions.
-Three month engagement preferred, but not everyone works that way some prefer quick in and done modes.
-To make such a program work, it seems it  must be university wide, to be broad enough to have the ftes needed funded.
-But there is not one mind on this, it maybe that the broadest approach is not idea.
-Perhaps a specialization focus on developers to support physical or biological sciences may be preferable, if the customer base is large enough.
-How to aggregate enough work to make it sustainable.
-How do you make sure you have the right staff skills?
-This will be an issue in a smaller group, and it will be exacerbated if the approach is general rather than focused on a domain emphasis where the number of possible skills required is less.
-The optimum use of resources seems to be 40% effort, on a project; two projects per developer.
-Two developers per project seems to be ideal, in the sense that software development is enhanced by two pairs of eyes..
-Collaborative computational projects (CCP) to support community networking: domain specific communities put forward proposals that are a priority of the community as a whole: biosimulation plasma PHYSICS QUANTUM PHYSICS.
-% years of funding.
-Custodians of community codes.
-Software support.
-help these are incubators.
-Add new features.
-Software of the future, open to all.
-Ccp submits.
-15 ftes for development across all CCPs.
-GET IAIN.
-4% success rate for software of the future project, high end computing consortia, in domains of CCPs.
-Co-design; the art of bringing it into existence together.
-The mission is codesigning 5 projects: the collaboration did work, but not as intimately as it might have.
-Incentive structures are still based on getting papers out, and the benefits to the postdocs are career focused.
-RSE fellowships: 204 applications, 5 funded.
-Intention was to fund applicants that demonstrated broad benefit to community/ies.
-How to better incentivize the collaboration to serve a broader community?
-Challenge: Scientists do not appreciate the engineering issues, whereas the cs person does not necessarily have the understanding of the domain.
-Possibility of a training course for research software engineering, a course could be run to teach people to manage a grant that is focused on software development.
-Non-hands on software carpentry for PIs.
-What are the next steps:
-Have we actually been successful?
-What are the metrics for success?
-Publications are tracked in James organization.
-How to evaluate the productivity gains that are made as a result of this engagement with James’ group?
-How to evaluate what did not go wrong because of this intervention?
-Notre Dame and LSU groups who are RSE groups.
-NCEAS/SOCSYNC; some RSE support; NESCENT/NIMBIOS also.
-The power of the label? Getting the word out about RSE support using the label.
+\begin{itemize}
+\item Lorraine Hwang -- UC Davis
+\item Simon Trigger
+\item Nancy Wilkins-Diehr -- San Diego Supercomputer Center
+\item Alexander Vyushkov -- Notre Dame
+\item Sandra Gesing - Notre Dame
+\item Ali Swanson -- University of Oxford
+
+\end{itemize}
+
+\subsection{Summary of Discussion}
+
+In addition to the points noted in the main discussion~\ref{RSE}, we also
+discussed the following:
+
+``Are you an RSE or a RA?'' -- this is not properly a binary question. Most of
+us sit at different points on that spectrum, and move along it during our
+careers. (Ususally from RA to RSE -- examples of movement in the other direction
+from readers would be welcomed.)
+Either way, the label ``Research Software Engineer'' is now starting to
+have some power. Many scientists do not want to be writing code; some do, to
+varying degrees. These groups can usefully support each other.
+
+he power of the label? Getting the word out about RSE support using the label.
 
@@
 Career Path.
-We also had a little discussion about how to make a career path for research developers.
-It need not be solely an academic enterprise, but tenure is always problematic for people of this class.
-There is a need for more senior people to have a track.
-Retention is hard, because these people are highly in demand by the economy, so retention is hard.
-People who have a PhD in science and enter industry, may return for diverse reasons.
-Things we haven’t started talking about:
-Other above the campus models for funding;
-Industry:academic partnerships
+We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but tenure is always problematic for people of this class.
 
-Follow-on Discussion at GCE15
-rse.ac.uk
-People: Lorraine Hwang, Smon Trigger, Nancy Wilkins-Diehr,  Alex Vyshkov, Sandra, Keynote person from Zooniverse
+Skills and resources may vary between teams. To help resolve this, maintaining
+high levels of communication between groups will be valuable. In the UK, there
+are plans to permit resource sharing between institutional RSE groups.
 
-Good example: Notre Dame (14 research programmer), LSU, Princeton,
-SDSC -- actually even in national labs there are inefficiencies with per-project jobs rather than develop pool -- potential issue is degree of control.
-Matt Trunnel - CIO of Broad institute boston - Chris Dwan - they had individual RSEs, then aggregated.
-James Taylor’s team  - John’s Hopkins
-Org models - OVPR, CS, Physics, IT, Library
-IT Coalescence issues.
-Maybe Yale?
+Collaborative funding can be important for RSE groups, to ensure that research
+leadership remains with the domain scientists. At NCAR, University partnerships
+are required for submission of proposals, so collaboration is an essential part
+of grant submission, and this will tend to bring developers and
+scientists together. The UCL group also follows this approach, with all bids
+requiring an academic collaborator.
+Domain scientists and developers are funded together in a single proposal.
+Another example of a success is the development of semantics and linked data in
+support of Ocean Sciences. An Earthcube funded project pairs domain scientists
+with RSEs and has been successful as the semantics attached have increased data
+use and discovery significantly.
 
-\subsection{Summary of Discussion}
+An alternative approach has been the provision of programming expertise as
+part of national compute services. The US XSEDE project's extended collaboration
+ service (ECS) is a set of developers who are paid with XSEDE funding,
+ and are on “permanent” staff.  When PIs request allocations on XSEDE resources,
+ there is a finite pool of developer time that can be awarded,
+typically for one year only, and at partial effort, typically 20 percent or so.
+The finite time allowed provides motivation for the scientist and their group to
+work closely the developer and to become educated in what the developer is doing,
+so they can sustain the effort, once the ECS period is over. This funding
+mechanism can be highly efficient for scientific problems, because the developer
+pool assembled by the research providers are, by definition, expert in the
+characteristics of their specific resource, and can very quickly assess the
+scientists needs, and what it will take to implement software that meets the
+user’s needs. However, it does not develop capacity within institutions,
+and indeed can have a ``chilling effect''.
 
-\subsection{Description of Opportunity, Challenges, and Obstacles}
+The UK allows this kind of collaboration to support the creation of scientific
+software for the large supercomputing resource called ARCHER.
+However, as well as resourcing this from the staff of the Edinburgh Parallel
+Computing Centre, who host the computer, this ``embedded CSE'' resource also
+allows the programming to come from local groups. This has been very helpful
+in providing funding to establish local groups. These work best when they
+develop good collaborations with national cyberinfrastucture pools.
+When an organization assembles
+a developer pool, diversity is developed and skills can be transferred.
+
+We would like to see these models applied outside High Performance Computing;
+most scientific software is not destined to run on national cyberinfrastructure,
+but needs similar support. The argument regarding making better use of expensive
+hardware through software improvements has been useful politically, (and many
+RSE groups are cited in organisations which host clusters for this reason), but
+the time has come to make the case that software itself is a critical
+cyberinfrastructure, and, with a much longer shelf-life than hardware, is itself
+a capital investment.
+
+The Canarie group (Canada) accepts proposals
+for providing services to broad communities, integrate people who are doing
+things that are complementary, the goal is to make the available stack more
+robust and richer for everyone. They offer short rounds of funding that can
+have as a key metric creating some useful functionality that shows a diversity
+of input and draws from across disciplines, then more funding could follow. This
+could apply within or across institutions.
+
+There can be problems communicating across cultural barriers, with domain
+scientists seeing developers as “other”.
+Collaboration, and tools to fund, encourage or motivate collaboration are
+extremely important.
+
+We think support from non-governmental organisations will be important if RSE
+groups will become established.
+The Sloan Foundation is currently funding data science engineers, who work in
+the context of other SW developers at University of Washington.
+These scientists work in the e-Science Studio/dataScience Studio, and they
+help a group of graduate students in solving their problems in data
+science and data management. During Fall and Spring, a 10-week incubator program
+allows students to work two days a week to work on a data-intensive science
+project. Some fraction of the developer time is dedicated to the developers'
+personal interests as well as instruction.
+
+The goal for Sloan is to provide success stories, to provide demonstrable value
+in the presence of data scientists on university staff. These stories are the
+basis for arguments to the host organization. This is an effort to create
+awareness of the value of research scientist developers. Embedding with
+scientists, and adding spare capacity is critical to making the innovation
+possible. This model is essentially to argue for permanent budget lines to
+support data scientists as part of university staff hires, just as with core
+facilities. This could become a fee-for-service model requested by grant
+funding, just as DNA sequencing is for core facilities, if it becomes apparent
+that this gives competitive advantage to the University’s research effort.
+
+One model that has been helpful in finding funding for RSE groups is the use of
+funds left over on research grants when RAs have left prematurely -- PIs like
+this arrangement as it is hard to find good staff for short-term positions, so
+having a pool of research programming staff on hand resolves this problem.
+We recommend that funders give explicit guidance to grant holders and
+institutions that such arrangement are favourable. Framework agreements
+permitting this to go ahead without
+checking back every time with funders and/or grant panels would further
+smooth this.
+(This is also provides a more stable job for those who hold these skills, but
+arguments about making life nicer for postdocs will not help persuade funders
+or PIs!)
+
+There is some question about the most effective
+duration and percent of full time for a programmer's work on a project.
+At least three months is necessary for the programmer to read into the science
+(RSEs must not become so disengaged from research that they don't have time to
+read a few papers -- this will result in code which doesn't meet scientific
+needs), but too long could result in an RSE losing their flexibility, becoming
+so engaged in one project that when that project ends, they find it hard to
+transfer. For this reason, we also recommend that 40 percent is ideal;
+two projects per developer, with some time for training and infrastructure work.
+Two developers per project seems to be ideal, in the sense that software
+development is enhanced by two pairs of eyes.
+
+There is as yet no clear answer as to the scale of aggregation needed to
+make such a program work. A university wide programme allows enough scale to
+be robust to fluctuations of funding within one field.
+But a specialization focus on developers to support, for example,
+physical or biological
+sciences may be preferable, if the customer base is large enough.
+The desire to aggregate enough work to make it sustainable, and the need to
+have domain-relevant research programming skils, are in tension.
+
+In the UK, another source of funding for research software has been the
+Collaborative computational projects (CCPs): domain specific communities put
+forward proposals that are a priority of the community as a whole, for example,
+biosimulation or plasma physics. These bodies act as custodians of community codes,
+and a central team also provides software engineering support support.
+
+However this area develops, the need for funding to secure software as a
+cyberinfrastructure component is clear. Funding which permits code to be
+refactored, tidied and optimised is rare; this is often done ``on the sly''
+in a scientifically focused grant. The UK EPSRC's ``software for the future''
+call, which really permits explicit investiment in software as an infrastructure,
+is so oversubscribed as to have a FOUR PERCENT success rate; the demand is clear!
+
+One opportunity is the idea of co-design; where infrastructural libraries are
+developed alongside the scientific codes that will call them. However,
+collaboration is hard to foster here; as incentive structures are still focused
+on short-term papers. This can cause infrastructure developers to
+focus more on publications
+in their areas of mathematics and CS, the domain developers on the shorter-term
+needs of their own fields. Genuine collaborative co-construction
+is harder to foster.
+
+It can be a more difficult
+to help leading domain scientists see the value of engineering effort than those
+in their teams who are forced to work with difficult-to-use or unreliable
+software tools, as they do not see the pain. Perhaps a version of
+``software carpentry'' targeted at those PIs
+who are awarded or apply for software heavy grants could be valuable here.
+
+RSEs provide a useful contribution to their Universities' *teaching* missions,
+as well as for research, as they are well placed to deliver the research programming
+training that many scientists now need. In the
+longer term, with programming skills taught to all through their career,
+we hope specialist
+scientific developers will be less needed.
 
@@
 
-\subsection{Plan for Future Organization}
+We will seek to identify and approach existing research programming organisations,
+to get their permission to list them on a list of research software groups.
+We will also look for examples of groups which have successfully become self-
+sustaining following
 
-\subsection{What Else is Needed?}
+\subsection{Plan for Future Organization and Future Needs}
 
-\subsection{Key Milestones and Responsible Parties}
+The UKRSE community will provide initial facilities to host this list, and
+continue to work to spread the initiative, but local leadership in the US
+is needed if this campaign is to succeed. This will require an initial gathering
+of identified research software organisations in the US to this end.
 
 \subsection{Description of Funding Needed}
+
+Financial support for an initial conference to bring together research software
+groups to form an organisation and create a resource sharing structure would
+help to further this campaign.
+
+In the longer term, funding organisations, especially non-governmental organisations
+with the capability to effect innovation through seed funding, could provide
+support to nucleate the creation of research software groups.

### 2015-11-05 WSSSPE3/report/wg_main_funding_spec_expert.tex
Close to complete on RSE sections

@@
 \subsection{Research Programming Expertise -- how to fund it?}
+\label{RSE}
 
@@
 groups, as well as for individual research software engineers, will be required
-in order to make the case for continued funding.
+in order to make the case for continued funding. Research programmers will not
+be measured by publications, we hope, but by other measures. Middle-author
+publications are common for RSEs. Most RSEs welcome co-authorship on papers
+where the PIs consider the contribution deserves it.
 
@@
 university positions following an industry role are penalised by the current
-structures.
+structures. Retention of senior programmers is hard, because these people
+are highly in demand by the economy. However,
+people who have a PhD in science and enter industry,
+may desire to return for diverse reasons, and should be welcomed back.
 
@@
 organisations which have traditionally supported university cyberinfrastructure,
-or more locally.
+library organisations, or research offices. Alternatively, these groups could
+be organisationally closer to research groups, sitting within academic
+departments. The most
+effective model will vary from institution to institution,
+but the mandate and ways of working should be similar.
 
@@
 how they might provide such seed funding.
+
+Success, appropriately measured, will help make the case for further investment.
+However, measurement of code cleanliness, and the resulting productivity gains,
+is an unsolved problem in empirical software engineering. To measure
+``what did not go wrong'' because of an intervention is particularly hard.
 
@@
 
-TBD, probably into the UKRSE site.
+http://www.rse.ac.uk/groups

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
new language for old point in keynote
Jeffrey Carver's talk

@@
 all have an invested stake in sustaining scientific software. Hence, having
-``sustained'' projects can suffocate new projects, for which we need to keep a
-lookout.
+``sustained'' projects can suffocate new projects, so we need to make sure
+we don't cause novel ideas and packages to suffer at the hands of the status quo.
 %This is a straw man concept, but it's something to look out for, since organizations will act in their own interest.
@@
 Sustainable Scientic Software}}
+
+I argue that for scientific software to be truly sustainable, there is a need
+for developers to use appropriate software engineering practices. My experience
+interacting with scientific teams indicates that choosing and tailoring these
+practices is not a trivial exercise. There is a general culture clash between
+software engineering and science that hinders our ability to communicate and
+choose appropriate methods. In addition, many experience scientific software
+developers appear to be unaware of software engineering practices that may be
+beneficial to them. The most appropriate software engineering practices are
+those that are lightweight, properly tailored, and focus on the key software
+development problems faced by scientists. In order to increase the use of
+software engineering in science, we need more documented success stories. These
+successes need to be socialized within the scientific community through
+workshops like the Software Engineering for Science workshop
+series~\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
+and the new Software Engineering track in
+Computing in Science and Engineering magazine.
 

### 2015-11-05 WSSSPE3/report/wg_main_user_community.tex
Update wg_main_user_community.tex
added discussion section

@@
 \item "Art of Community" by Jono Bacon. We could consider distilling this for scientific software.
+\item Iain Larmour, from EPSRC inthe UK [[https://www.epsrc.ac.uk/][EPSRC]] (not sure who from mentioned UK Collaborative Computational Projects ([[http://ccp.ac.uk][CCP]])
+\end{itemize}
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
+%\todo{short-ish text here}
+
+Discussion revolved around a few questions: what is the benefit of having a "community" for software sustainability, what practices and circumstances lead to having a community, how can funding help or hinder this process, and perhaps most importantly how can best practices be described and distilled into a document that can help new projects.
+
+The benefits of having a community that were brought up were considered largely obvious.
+In addition to having advocates for the software, and a possible source of ``free'' contributions to the codebase, the community becomes a good source for requirements, feedback, and metrics.
+The software community can also act as "cheerleaders" who convince funders or other potential users to fund/use the software, and thus help sustain the software.
+
+Practices and circumstances that lead to a community are first, that the software offers value.
+But in addition to this, a community will be much more likely to form if they receive (expert) support when they have questions.
+Additional contributing factors are good usability (not always needed), and an open development process such as IPython developer meetings on YouTube.
+It was also pointed out that an evangelist for the project, not necessarily but often one of the developers, can often make a big difference.
+
+Funding can help the process by encouraging both value to the community and high-quality user support.
+Only providing funding for the software development may create good software, but with less likelihood to have a real community.
+It was discussed that federal laboratories are a good incubator for software communities, and that a general facility like EarthCube is too dispersed to really make a community.
+Also, domain-specific groups within laboratories or universities might provide as an incubator for software communitites.
+
+In describing best practices, the group discussed the different modes for starting a scientific software project: building on an existing product that needs improving, recognizing an unsatisfied need of an existing community, or creating a new solution to a need not yet recognized by the community.
+The group also thought that the existing books on software communities would need to be evaluated in light of differences between Science Software projets and general OSS projects in terms of scale, science, acknowledgement and credit, and funding models.
 

### 2015-11-05 WSSSPE3/report/wg_main_user_community.tex
added plans

@@
 \subsubsection{Plans}
-\todo{short text here - not bullets}
+%\todo{short text here - not bullets}
+
+The most important next steps is a ``Best Practice'' document, which would describe what successful projects with engaged communities look like, how to replicate this type of project, and look at end-of-life on a community project.
+Inputs to this document would include a software community survey of high functioning communities such as R Open Science, Python SciPy, OPeNDAP, and Unidata, with analysis of factors that feed into their success.
+Also references lik the "Art of Community" could be adapted and summarized for the science software community.
+
+Another next step would be increasing recognition of need for science software projects to focus on building and supporting their user community.
+Good Software Engineering practices are not enough, and popular training like Software Carpentry does not currently address this issue head on.
 

### 2015-11-05 WSSSPE3/report/wg_main_user_community.tex
added line breaks
so it looks better on github in non-edit mode
typo
fixed some quotes

@@
 
-Mozilla Science maintains a "'Working Open' Project Guide" (http://mozillascience.github.io/leadership-training). From the introduction:
+Mozilla Science maintains a ``Working Open' Project Guide'' (http://mozillascience.github.io/leadership-training). From the introduction:
 \begin{quote}
@@
 \begin{itemize}
-\item "Art of Community" by Jono Bacon. We could consider distilling this for scientific software.
-\item Iain Larmour, from EPSRC inthe UK [[https://www.epsrc.ac.uk/][EPSRC]] (not sure who from mentioned UK Collaborative Computational Projects ([[http://ccp.ac.uk][CCP]])
+\item ``Art of Community'' by Jono Bacon. We could consider distilling this for scientific software.
+\item Iain Larmour, from EPSRC inthe UK (EPSRC: https://www.epsrc.ac.uk/) (not sure who from mentioned UK Collaborative Computational Projects (CCP: http://ccp.ac.uk)
 \end{itemize}
@@
 
-Discussion revolved around a few questions: what is the benefit of having a "community" for software sustainability, what practices and circumstances lead to having a community, how can funding help or hinder this process, and perhaps most importantly how can best practices be described and distilled into a document that can help new projects.
+Discussion revolved around a few questions: what is the benefit of having a ``community'' for software sustainability, what
+practices and circumstances lead to having a community, how can funding help or hinder this process, and perhaps most
+importantly how can best practices be described and distilled into a document that can help new projects.
 
-The benefits of having a community that were brought up were considered largely obvious.
-In addition to having advocates for the software, and a possible source of ``free'' contributions to the codebase, the community becomes a good source for requirements, feedback, and metrics.
-The software community can also act as "cheerleaders" who convince funders or other potential users to fund/use the software, and thus help sustain the software.
+The benefits of having a community that were brought up were considered largely obvious. In addition to having advocates for
+the software, and a possible source of ``free'' contributions to the codebase, the community becomes a good source for
+requirements, feedback, and metrics. The software community can also act as ``cheerleaders'' who convince funders or other
+potential users to fund/use the software, and thus help sustain the software.
 
-Practices and circumstances that lead to a community are first, that the software offers value.
-But in addition to this, a community will be much more likely to form if they receive (expert) support when they have questions.
-Additional contributing factors are good usability (not always needed), and an open development process such as IPython developer meetings on YouTube.
-It was also pointed out that an evangelist for the project, not necessarily but often one of the developers, can often make a big difference.
+Practices and circumstances that lead to a community are first, that the software offers value. But in addition to this, a
+community will be much more likely to form if they receive (expert) support when they have questions. Additional contributing
+factors are good usability (not always needed), and an open development process such as IPython developer meetings on YouTube.
+It was also pointed out that an evangelist for the project, not necessarily but often one of the developers, can often make a
+big difference.
 
-Funding can help the process by encouraging both value to the community and high-quality user support.
-Only providing funding for the software development may create good software, but with less likelihood to have a real community.
-It was discussed that federal laboratories are a good incubator for software communities, and that a general facility like EarthCube is too dispersed to really make a community.
-Also, domain-specific groups within laboratories or universities might provide as an incubator for software communitites.
+Funding can help the process by encouraging both value to the community and high-quality user support. Only providing funding
+for the software development may create good software, but with less likelihood to have a real community. It was discussed
+that federal laboratories are a good incubator for software communities, and that a general facility like EarthCube is too
+dispersed to really make a community. Also, domain-specific groups within laboratories or universities might provide as an
+incubator for software communities.
 
-In describing best practices, the group discussed the different modes for starting a scientific software project: building on an existing product that needs improving, recognizing an unsatisfied need of an existing community, or creating a new solution to a need not yet recognized by the community.
-The group also thought that the existing books on software communities would need to be evaluated in light of differences between Science Software projets and general OSS projects in terms of scale, science, acknowledgement and credit, and funding models.
+In describing best practices, the group discussed the different modes for starting a scientific software project: building on
+an existing product that needs improving, recognizing an unsatisfied need of an existing community, or creating a new solution
+to a need not yet recognized by the community. The group also thought that the existing books on software communities would
+need to be evaluated in light of differences between Science Software projets and general OSS projects in terms of scale,
+science, acknowledgement and credit, and funding models.
 
@@
 
-The most important next steps is a ``Best Practice'' document, which would describe what successful projects with engaged communities look like, how to replicate this type of project, and look at end-of-life on a community project.
-Inputs to this document would include a software community survey of high functioning communities such as R Open Science, Python SciPy, OPeNDAP, and Unidata, with analysis of factors that feed into their success.
-Also references lik the "Art of Community" could be adapted and summarized for the science software community.
+The most important next steps is a ``Best Practice'' document, which would describe what successful projects with engaged
+communities look like, how to replicate this type of project, and look at end-of-life on a community project. Inputs to this
+document would include a software community survey of high functioning communities such as R Open Science, Python SciPy,
+OPeNDAP, and Unidata, with analysis of factors that feed into their success. Also references lik the ``Art of Community'' could
+be adapted and summarized for the science software community.
 
-Another next step would be increasing recognition of need for science software projects to focus on building and supporting their user community.
-Good Software Engineering practices are not enough, and popular training like Software Carpentry does not currently address this issue head on.
+Another next step would be increasing recognition of need for science software projects to focus on building and supporting
+their user community. Good Software Engineering practices are not enough, and popular training like Software Carpentry does
+not currently address this issue head on.
 

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
adding a comment
add Birgit's talk summary

@@
 \todo{need to update for WSSSPE3}
-\todo{KEN: Should these be bolded?}
+\todo{KEN: Should these be bolded? - dsk: yes, titles can be bolded, but contents under each should not}
 \begin{comment}
@@
 Sustainability}}
+The concept of sustainability has become a topic of interest in the field of computing, which is
+evidenced by the increase in the number of events that focus on the topic - nevertheless, it isn’t
+well understood yet. Penzenstadler argued that we often define sustainability too narrowly. Instead,
+sustainability at its heart is a systemic concept and must be viewed from a range of different
+dimensions including environmental, economic, individual, social and technical. She introduced the
+Karlskrona Manifesto on Software Design, which distills knowledge from a broad range of related work
+on the topic of sustainability into a set of (mis-)perceptions and principles. The manifesto does
+not proclaim that there is an easy, one size fits all solution around the corner, but rather points
+out that sustainability is a ``wicked problem'' and is often misunderstood.
+Due to these misperceptions, even though sustainability's importance is increasingly recognized,
+many software systems are unsustainable. Even more alarming is that most software systems'
+broader impacts on sustainability are unknown. To change this, the Karlskrona Manifesto proposes
+nine principles and commitments. These commitments are not dogmatic laws, but rather a commitment to
+rethink, to move beyond the silo mentality, and analyze in more depth. As such, they do not
+restrict, but rather open up a space for discussion.
 
@@
 \input{wg_appendix_eng_design}
-\input{wg_appendix_metrics}
+%\input{wg_appendix_metrics}
 \input{wg_appendix_training}

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
edits in lightning talk texts

@@
 The concept of sustainability has become a topic of interest in the field of computing, which is
-evidenced by the increase in the number of events that focus on the topic - nevertheless, it isn’t
+evidenced by the increase in the number of events that focus on the topic. Nevertheless, it isn't
 well understood yet. Penzenstadler argued that we often define sustainability too narrowly. Instead,
@@
 dimensions including environmental, economic, individual, social and technical. She introduced the
-Karlskrona Manifesto on Software Design, which distills knowledge from a broad range of related work
+Karlskrona Manifesto on Software Design~\cite{Becker:2014}, which distills knowledge from a broad range of related work
 on the topic of sustainability into a set of (mis-)perceptions and principles. The manifesto does
-not proclaim that there is an easy, one size fits all solution around the corner, but rather points
+not proclaim that there is an easy, one-size-fits-all solution around the corner, but rather points
 out that sustainability is a ``wicked problem'' and is often misunderstood.
@@
 
-I argue that for scientific software to be truly sustainable, there is a need
-for developers to use appropriate software engineering practices. My experience
+Jeff argued that for scientific software to be truly sustainable, there is a need
+for developers to use appropriate software engineering practices. His experience
 interacting with scientific teams indicates that choosing and tailoring these
@@
 workshops like the Software Engineering for Science workshop
-series~\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
+series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
 and the new Software Engineering track in
-Computing in Science and Engineering magazine.
+\textit{Computing in Science and Engineering} magazine.
 

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
add Birgit's talk summary

--- 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
@@
 Facilitating Research Training While Supporting Scientific Research}}
+A case study: ``The Cooperative Computing Lab'' at the University of Notre Dame
+is a small group of individuals with the main tasks of collaborating with
+people that have large-scale computing problems, operating various parallel
+computer systems, conducting computer science research, and developing open
+source software. One of the main challenges they face is finding the balance
+between flexibility/training and stability/quality. Their current solution for
+ensuring the latter was to add a software engineer (the presenter) to the
+existing team of faculty and students, who now also serves as ``spring''
+between flexibility and stability.
 
@@
 Sustainability}}
-The concept of sustainability has become a topic of interest in the field of computing, which is
-evidenced by the increase in the number of events that focus on the topic - nevertheless, it isn’t
-well understood yet. Penzenstadler argued that we often define sustainability too narrowly. Instead,
-sustainability at its heart is a systemic concept and must be viewed from a range of different
-dimensions including environmental, economic, individual, social and technical. She introduced the
-Karlskrona Manifesto on Software Design, which distills knowledge from a broad range of related work
-on the topic of sustainability into a set of (mis-)perceptions and principles. The manifesto does
-not proclaim that there is an easy, one size fits all solution around the corner, but rather points
-out that sustainability is a ``wicked problem'' and is often misunderstood.
-Due to these misperceptions, even though sustainability's importance is increasingly recognized,
-many software systems are unsustainable. Even more alarming is that most software systems'
-broader impacts on sustainability are unknown. To change this, the Karlskrona Manifesto proposes
-nine principles and commitments. These commitments are not dogmatic laws, but rather a commitment to
-rethink, to move beyond the silo mentality, and analyze in more depth. As such, they do not
-restrict, but rather open up a space for discussion.
+The concept of sustainability has become a topic of interest in the field of
+computing, which is evidenced by the increase in the number of events that
+focus on the topic - nevertheless, it isn’t well understood yet. Penzenstadler
+argued that we often define sustainability too narrowly. Instead,
+sustainability at its heart is a systemic concept and must be viewed from a
+range of different dimensions including environmental, economic, individual,
+social and technical. She introduced the Karlskrona Manifesto on Software
+Design, which distills knowledge from a broad range of related work on the
+topic of sustainability into a set of (mis-)perceptions and principles. The
+manifesto does not proclaim that there is an easy, one size fits all solution
+around the corner, but rather points out that sustainability is a ``wicked
+problem'' and is often misunderstood. Due to these misperceptions, even though
+sustainability's importance is increasingly recognized, many software systems
+are unsustainable. Even more alarming is that most software systems' broader
+impacts on sustainability are unknown. To change this, the Karlskrona Manifesto
+proposes nine principles and commitments. These commitments are not dogmatic
+laws, but rather a commitment to rethink, to move beyond the silo mentality,
+and analyze in more depth. As such, they do not restrict, but rather open up a
+space for discussion.
 
@@
 Sustainable Scientic Software}}
-
 I argue that for scientific software to be truly sustainable, there is a need
@@
 series~\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in
-Computing in Science and Engineering magazine.
+and the new Software Engineering track in Computing in Science and Engineering
+magazine.
 

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex

@@
 Sustainable Scientic Software}}
-I argue that for scientific software to be truly sustainable, there is a need
-for developers to use appropriate software engineering practices. My experience
+Jeff argued that for scientific software to be truly sustainable, there is a need
+for developers to use appropriate software engineering practices. His experience
 interacting with scientific teams indicates that choosing and tailoring these
@@
 workshops like the Software Engineering for Science workshop
-series~\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in Computing in Science and Engineering
-magazine.
+series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
+and the new Software Engineering track in
+\textit{Computing in Science and Engineering} magazine.
 

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
more small edits in lightning talk section

@@
 A case study: ``The Cooperative Computing Lab'' at the University of Notre Dame
-is a small group of individuals with the main tasks of collaborating with
+is a small group of individuals whose main tasks are collaborating with
 people that have large-scale computing problems, operating various parallel
@@
 \item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
-Betz, Ruzanna Chitchyan, Letícia Duboc, Steve Easterbrook, Guillermo
+Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
 Rodriguez-Navas and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
@@
 computing, which is evidenced by the increase in the number of events that
-focus on the topic - nevertheless, it isn’t well understood yet. Penzenstadler
+focus on the topic - nevertheless, it isn't well understood yet. Penzenstadler
 argued that we often define sustainability too narrowly. Instead,
@@
 social and technical. She introduced the Karlskrona Manifesto on Software
-Design, which distills knowledge from a broad range of related work on the
+Design~\cite{Becker:2014}, which distills knowledge from a broad range of related work on the
 topic of sustainability into a set of (mis-)perceptions and principles. The

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
a few more changes

@@
 computing, which is evidenced by the increase in the number of events that
-focus on the topic - nevertheless, it isn't well understood yet. Penzenstadler
+focus on the topic. Nevertheless, it isn't well understood yet. Penzenstadler
 argued that we often define sustainability too narrowly. Instead,
@@
 topic of sustainability into a set of (mis-)perceptions and principles. The
-manifesto does not proclaim that there is an easy, one size fits all solution
+manifesto does not proclaim that there is an easy, one-size-fits-all solution
 around the corner, but rather points out that sustainability is a ``wicked

### 2015-11-05 WSSSPE3/report/wg_main_metrics.tex
small changes, line breaks

@@
 
-Metrics for scientific software are important for tenure and promotion, scientific impact, discovery, reducing duplication, serving as a basis for potential industrial interest in adopting software, and making a case for new or continued funding.
-However, there is no commonly-used standard for collecting or presenting metrics, nor is it known if there is a common set of metrics for scientific software.
+Metrics for scientific software are important for tenure and promotion, scientific impact, discovery, reducing duplication, serving as a basis for potential industrial interest in adopting software, prioritizing develop and support towards
+strategic objectives and making a case for new or continued funding.  However, there is no commonly-used 
+standard for collecting or presenting metrics, nor is it known if there is a common set of metrics for scientific software. 
  It is imperative that scientific software stakeholders understand that it is useful to collect metrics.
@@
 
-The group discussion here focused on identifying existing frameworks and activities for scientific software metrics.  The group identified the following related activities:
+The group discussion  focused on identifying existing frameworks and activities for scientific software metrics.  
+The group identified the following related activities:
 
@@
 
-The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper.
-The group discussed many questions related to useful metrics for scientific software including addressing if there is a common set of metrics that can be filtered in some way, can metrics be fit into a common template, which metrics would be the most useful for each stakeholder, which metrics are the most helpful and how would we assess this, how are metrics monitored, and many more.
-A more complete bulleted list of these questions can be found in Appendix H.
-Next, a roadmap for how to proceed was discussed including creating a set of milestones and tasks.
-The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in order to email out a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.
-This survey would be created by one of the student group members.
-Similarly, it was suggested that a software metrics survey be sent to the UK SFTF and TRDF software projects to ask them what metrics would be useful to report.
-The remainder of the discussion focused mainly on the creation of a white paper on this topic.
-This resulted in a paper outline and writing assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE, or JORS.
+The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific 
+software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper. 
+ The group discussed many questions related to useful metrics for scientific software including addressing if there is a common set of metrics that 
+ can be filtered in some way, can metrics be fit into a common template, which metrics would be the most useful for each stakeholder, 
+ which metrics are the most helpful and how would we assess this, how are metrics monitored, and many more.  
+ A more complete bulleted list of these questions can be found in Appendix H.  Next, a roadmap for how to proceed was discussed including
+ creating a set of milestones and tasks.  The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure
+  for Sustained Innovation (SI2) PI workshop in order to send a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.  
+ The five solicitations for software elements released  under the NSF SI2 program all included metrics as a required component with submitters requested to include 
+ {\it "a list of tangible metrics, with end user involvement, to be used to measure the success of the software element developed, ...'}.
+These metrics are then reported as part of annual reports to NSF by the projects.
+Although neither the proposal text describing the metrics nor the reported metric results are publicly available, there is reason to believe that the community will be willing to provide this information through a survey mechanism.
+This survey would be created by one of the student group members.  Similarly, it was suggested that a software metrics survey be sent to the 
+UK SFTF and TRDF software projects to ask them what metrics would be useful to report.  The remainder of the discussion focused mainly on the 
+creation of a white paper on this topic.  This resulted in a paper outline and writing assignments with the goal of publishing in venues including 
+WSSSPE4, IEEE CISE, or JORS.
 

### 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
small changes, line breaks

--- 2015-11-05 WSSSPE3/report/WSSSPE3_review.tex
@@
 travel of potential participants who would not otherwise be able to attend,
-younger participants, and those from underrepresented groups. These funds allowed
+early-stage researchers, and those from underrepresented groups. These funds allowed
 16 additional participants to attend.
@@
 \todo{need to update for WSSSPE3}
-\todo{KEN: Should these be bolded? - dsk: yes, titles can be bolded, but contents under each should not}
+\todo{KEN: Should these be bolded?}
 \begin{comment}
@@
 Facilitating Research Training While Supporting Scientific Research}}
-A case study: ``The Cooperative Computing Lab'' at the University of Notre Dame
-is a small group of individuals with the main tasks of collaborating with
-people that have large-scale computing problems, operating various parallel
-computer systems, conducting computer science research, and developing open
-source software. One of the main challenges they face is finding the balance
-between flexibility/training and stability/quality. Their current solution for
-ensuring the latter was to add a software engineer (the presenter) to the
-existing team of faculty and students, who now also serves as ``spring''
-between flexibility and stability.
 
@@
 Sustainability}}
-The concept of sustainability has become a topic of interest in the field of
-computing, which is evidenced by the increase in the number of events that
-focus on the topic - nevertheless, it isn’t well understood yet. Penzenstadler
-argued that we often define sustainability too narrowly. Instead,
-sustainability at its heart is a systemic concept and must be viewed from a
-range of different dimensions including environmental, economic, individual,
-social and technical. She introduced the Karlskrona Manifesto on Software
-Design, which distills knowledge from a broad range of related work on the
-topic of sustainability into a set of (mis-)perceptions and principles. The
-manifesto does not proclaim that there is an easy, one size fits all solution
-around the corner, but rather points out that sustainability is a ``wicked
-problem'' and is often misunderstood. Due to these misperceptions, even though
-sustainability's importance is increasingly recognized, many software systems
-are unsustainable. Even more alarming is that most software systems' broader
-impacts on sustainability are unknown. To change this, the Karlskrona Manifesto
-proposes nine principles and commitments. These commitments are not dogmatic
-laws, but rather a commitment to rethink, to move beyond the silo mentality,
-and analyze in more depth. As such, they do not restrict, but rather open up a
-space for discussion.
 
@@
 Sustainable Scientic Software}}
-Jeff argued that for scientific software to be truly sustainable, there is a need
-for developers to use appropriate software engineering practices. His experience
-interacting with scientific teams indicates that choosing and tailoring these
-practices is not a trivial exercise. There is a general culture clash between
-software engineering and science that hinders our ability to communicate and
-choose appropriate methods. In addition, many experience scientific software
-developers appear to be unaware of software engineering practices that may be
-beneficial to them. The most appropriate software engineering practices are
-those that are lightweight, properly tailored, and focus on the key software
-development problems faced by scientists. In order to increase the use of
-software engineering in science, we need more documented success stories. These
-successes need to be socialized within the scientific community through
-workshops like the Software Engineering for Science workshop
-series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in
-\textit{Computing in Science and Engineering} magazine.
 
@@
 \input{wg_appendix_eng_design}
-%\input{wg_appendix_metrics}
+\input{wg_appendix_metrics}
 \input{wg_appendix_training}

### 2015-11-06 WSSSPE3/report/awardees.tex
final travel awardee

--- 2015-10-29 WSSSPE3/report/awardees.tex
@@
 Bernadette Randles &  UCLA\\
-%Ilian Todorov &  STFC Daresbury Laboratory\\
+Ilian Todorov &  STFC Daresbury Laboratory\\
 Nic Weber &  University of Washington iSchool

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
small fixes
add Abani Patra's talk

@@
 Improve Productivity of Legacy Tools}}
+Abani Patra presented an example how the community using Titan2d, a Geoflow
+Simulation Software, increased the productivity of their tools by improving
+both code and data layout. Main obstacles in this change were the non-existence
+of a common version control system for the source code, coupled with multiple
+versions of the same code base, the fixed format of input files, that many
+input values used to be set as compilation flags, and that the internal data
+layout was not suitable for modern technologies (vectorization, accelerators).
+The approach of the Titan2d developers included reinforcing the code structure
+using a multiple laters of Python and C++, and a redesign of the data layout to
+be better suitable for modern CPUs and accelerators.
 
@@
 
-\item \textbf{Alice Allen: \textit{We’re giving away the store! (Merchandise not
+\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
 included)}}

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add Abigail's talk

@@
 from Open Source}}
+Abigail Cabunoc Mayes combines two properties of open source that, together,
+form a suitable habitat for sustainable software. The first of the two
+properties, being public, does not only mean public code. It also includes
+public discussions, a public process of including contributions, and an open
+license. The second property, participatory, stresses the importance of
+reaching out to the community and helping potential new members by providing
+better documenation and learning experiences, like code review and examples of
+good first bug reports. Together, Abigail concludes, these two properties not
+only lead to higher quality, reusability and ease of understanding, but
+eventually, to sustainablity.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
Lorraine's talk

@@
 Computational Infrastructure for Geodynamics}}
+Lorraine Hwang presented experiences with CIG: the computational infrastructure
+for Geodynamics, a community software with a worldwide user base. As others,
+their main goals include high usability, sustainability, and reproducibility.
+In order to achieve this, communication channels have been developed, like
+mailing lists, wikis, workshops, hackathons, tutorials, and webinars. In order
+to contribute to the infrastructure, codes have to adhere to specified minimum
+standards, but ideals are also given to encourage reaching the aforementioned
+goals. This includes, e.g., the use of version control, certain coding styles,
+the presence and nature of code tests and documentation and certain user
+workflows. In addition, various ways to cite efforts within the framework are
+provided, including science papers, code papers, user manuals, and the website.
+In an effort creating statistics of the resulting citations, $80\%$ of papers
+that use CIG codes mention the code name, and about the same number includes a
+citation. Only about $20\%$ acknowledge CIG. Within the same sample of papers,
+about one fifth of use an URL to cite codes (including non-CIG codes), and only
+about one eigth specifies the version used, showing that compared to other
+codes, CIG seems to be much better cited. This, in part, is attributed to ideal
+citation forms communicated to the community, and runtime tools that help
+collecting a relevant citation list automatically.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
split Lorraine's talks

@@
 for Geodynamics}}
-
-\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
-Computational Infrastructure for Geodynamics}}
 Lorraine Hwang presented experiences with CIG: the computational infrastructure
@@
 the presence and nature of code tests and documentation and certain user
-workflows. In addition, various ways to cite efforts within the framework are
-provided, including science papers, code papers, user manuals, and the website.
-In an effort creating statistics of the resulting citations, $80\%$ of papers
-that use CIG codes mention the code name, and about the same number includes a
-citation. Only about $20\%$ acknowledge CIG. Within the same sample of papers,
-about one fifth of use an URL to cite codes (including non-CIG codes), and only
-about one eigth specifies the version used, showing that compared to other
-codes, CIG seems to be much better cited. This, in part, is attributed to ideal
-citation forms communicated to the community, and runtime tools that help
-collecting a relevant citation list automatically.
+workflows.
+
+\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
+Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
+Computational Infrastructure for Geodynamics}}
+In addition to the previous talk \textit{Advancing Earth Science
+through Best Practices in Open Source Software: Computational Infrastructure
+for Geodynamics}, Lorraine Hwang mentions in a second talk various ways to cite
+efforts within the framework, including science papers, code papers, user
+manuals, and the website. In an effort creating statistics of the resulting
+citations, $80\%$ of papers that use CIG codes mention the code name, and about
+the same number includes a citation. Only about $20\%$ acknowledge CIG. Within
+the same sample of papers, about one fifth of use an URL to cite codes
+(including non-CIG codes), and only about one eigth specifies the version used,
+showing that compared to other codes, CIG seems to be much better cited. This,
+in part, is attributed to ideal citation forms communicated to the community,
+and runtime tools that help collecting a relevant citation list automatically.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add Mike Hildreth's talk

@@
 Science (DASPOS)}}
+DASPOS is an NSF-funded multi-disciplinary effort, with main host at Notre
+Dame, Chicago, and links the HEP effort to Bioligy, Astrophysics, Digital
+Curation, and other disciplines. It includes physicists, digital librarians, as
+well as computer scientists, and aims to achieve some commonality across
+disciplines. Examples are meta-data descriptions of archived data,
+computational descriptions, descriptions of how data was processed, but also
+includes questions like whether computation replication can be automated, of
+what the impact of access policies on the preservation infrastructure is. One
+of the products of this effort is a suite of tools that hels with this
+preservation, and the questions was posed wether that software itself is
+sustainable. Points that were brought up include the need for a user community
+depending on a given software, and the need to provide added value for its
+users. An important aspect to achieve this to plan to work with the user
+community from the start, and to budget that way. For the specific example of
+the preservation software this means that besides adding value to the
+community, it needs to be transparent to their workflow, i.e., not requiring
+additional effort to preserve ``research objects''.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
edits in lightning talks

--- 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 Improve Productivity of Legacy Tools}}
-Abani Patra presented an example how the community using Titan2d, a Geoflow
+Abani Patra presented an example how the community using Titan2D, a Geoflow
 Simulation Software, increased the productivity of their tools by improving
-both code and data layout. Main obstacles in this change were the non-existence
+both code and data layout. The main obstacles in this change were the non-existence
 of a common version control system for the source code, coupled with multiple
@@
 layout was not suitable for modern technologies (vectorization, accelerators).
-The approach of the Titan2d developers included reinforcing the code structure
+The approach of the Titan2D developers included reinforcing the code structure
 using a multiple laters of Python and C++, and a redesign of the data layout to
@@
 Abigail Cabunoc Mayes combines two properties of open source that, together,
-form a suitable habitat for sustainable software. The first of the two
-properties, being public, does not only mean public code. It also includes
+create a suitable habitat for sustainable software. The first of the two
+properties, public, does not only mean public code. It also includes
 public discussions, a public process of including contributions, and an open
@@
 reaching out to the community and helping potential new members by providing
-better documenation and learning experiences, like code review and examples of
-good first bug reports. Together, Abigail concludes, these two properties not
+better documentation and learning experiences, like code review and examples of
+good first bug reports. Together, Abigail concluded, these two properties not
 only lead to higher quality, reusability and ease of understanding, but
-eventually, to sustainablity.
+eventually, to sustainability.
 
@@
 for Geodynamics}}
-
-\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
-Computational Infrastructure for Geodynamics}}
-Lorraine Hwang presented experiences with CIG: the computational infrastructure
-for Geodynamics, a community software with a worldwide user base. As others,
+Lorraine Hwang presented experiences with CIG: the Computational Infrastructure
+for Geodynamics, a community software with a worldwide user base. Like others,
 their main goals include high usability, sustainability, and reproducibility.
-In order to achieve this, communication channels have been developed, like
+In order to achieve this, communication channels have been developed, such as
 mailing lists, wikis, workshops, hackathons, tutorials, and webinars. In order
@@
 the presence and nature of code tests and documentation and certain user
-workflows. In addition, various ways to cite efforts within the framework are
-provided, including science papers, code papers, user manuals, and the website.
-In an effort creating statistics of the resulting citations, $80\%$ of papers
-that use CIG codes mention the code name, and about the same number includes a
-citation. Only about $20\%$ acknowledge CIG. Within the same sample of papers,
-about one fifth of use an URL to cite codes (including non-CIG codes), and only
-about one eigth specifies the version used, showing that compared to other
-codes, CIG seems to be much better cited. This, in part, is attributed to ideal
-citation forms communicated to the community, and runtime tools that help
-collecting a relevant citation list automatically.
+workflows.
+
+\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
+Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
+Computational Infrastructure for Geodynamics}}
+In a second talk, Lorraine Hwang mentioned various ways to cite
+efforts within the framework, including science papers, code papers, user
+manuals, and the website. An analysis of the resulting
+citations showed that $80\%$ of papers that use CIG codes mention the code name, and about
+the same number includes a citation. Only about $20\%$ acknowledge CIG. Within
+the same sample of papers, about one fifth use an URL to cite codes
+(including non-CIG codes), and only about one eighth specify the version used,
+showing that compared to other codes, CIG seems to be much better cited. This,
+in part, is attributed to suggested citation forms communicated to the community,
+and runtime tools that help collecting a relevant citation list automatically.
 
@@
 Science (DASPOS)}}
+DASPOS is an NSF-funded multi-disciplinary effort, located at Notre
+Dame and Chicago, that links the HEP effort to Biology, Astrophysics, Digital
+Curation, and other disciplines. It includes physicists, digital librarians, as
+well as computer scientists, and aims to achieve some commonality across
+disciplines. Examples are meta-data descriptions of archived data,
+computational descriptions, descriptions of how data was processed, but also
+includes questions such as whether computation replication can be automated, and
+what the impact of access policies on the preservation infrastructure is. One
+of the products of this effort is a suite of tools that deals with this
+preservation, and the questions was posed whether that software itself is
+sustainable. Points that were brought up included the need for a user community
+depending on a given software, and the need to provide added value for its
+users. An important method to achieve this was to work with the user
+community from the start, and to budget that way. For the specific example of
+the preservation software, this means that besides adding value to the
+community, it needs to be transparent to their workflow, i.e., not requiring
+additional effort to preserve ``research objects''.
 
@@
 \item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
-Sustainable Scientic Software}}
+Sustainable Scientific Software}}
 Jeff argued that for scientific software to be truly sustainable, there is a need

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add James Hetherington's talk

--- 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods and
-Susheel Varma: \textit{Research Software Engineering Groups in Universities: The Story
-from the UK}}
+Susheel Varma: \textit{An update from UK Research Software Engineers}}
+James Hetherington started by listing some of the problems research software
+faces, which include low levels of reuse and poor standard of verification. For
+a long time, technical solutions to such technical problems were focus of the
+e-Research community, and include research software distributions, grids,
+middleware and workflows. Some limited adoption can be seen today in research
+communities, but the main problems have not been solved to a sufficient level.
+James Hetherington hypothesises that instead of technical solutions, social
+innovation is needed: a new role in the academic system focused on research
+software that combines the best of a craftsperson and a scholar. However,
+social innovation in centuries-old institutions is hard. Alternatives to such a
+new role would have to include rewards for good research software, recognition
+of software as academic output, and rejecting submissions based on
+unreproducible computational results. Advantages of research software
+engineering (RSE) groups include the possibility of training in reproducible
+computational research, providing collaborations for for researchers who don't
+want to be programmers, and synergies with research computing platforms. The
+success of such a group would be measured by the output and quality of the
+research software, and could be part of Research Computing or faculty. Members
+would not be independent researchers, but would have to have a research
+background. An attempt to form a community of RSE groups within the UK has been
+under way for several years, including funding from the UK Research Council. An
+open question is whether this approach can be adopted also in other countries,
+including the USA.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
edits to Hetherington LT

@@
 a long time, technical solutions to such technical problems were focus of the
-e-Research community, and include research software distributions, grids,
+eResearch community, including research software distributions, grids,
 middleware and workflows. Some limited adoption can be seen today in research
 communities, but the main problems have not been solved to a sufficient level.
-James Hetherington hypothesises that instead of technical solutions, social
+James Hetherington hypothesizes that instead of technical solutions, social
 innovation is needed: a new role in the academic system focused on research
-software that combines the best of a craftsperson and a scholar. However,
+software that combines the best parts of a craftsperson and a scholar. However,
 social innovation in centuries-old institutions is hard. Alternatives to such a
@@
 of software as academic output, and rejecting submissions based on
-unreproducible computational results. Advantages of research software
+unreproducible computational results. Some advantages of research software
 engineering (RSE) groups include the possibility of training in reproducible
@@
 background. An attempt to form a community of RSE groups within the UK has been
-under way for several years, including funding from the UK Research Council. An
+underway for several years, including funding from the UK Research Council. An
 open question is whether this approach can be adopted also in other countries,

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add Dan Gunter's talk

--- 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 User into Building Sustainable Software for Science}}
+``What is needed to develop sustainable software'' is the main question in Dan
+Gunter's talk. Above the usual suspects of funding, great developers, good
+design, and software engineering practices, however, Dan Gunter places the
+users. He explains this using a traditional software development model starting
+from gathering requirements, and reaching release through a design and
+development and testing phase. The main defficiency that with this approach is
+pointed out to be the too-late interaction with users. Instead, an alternative
+approach is proposed that, at first, skips the development phase and repeatedly
+goes through requirement gatherning, design and user interaction/learning
+phases, and only eventually starts development once an agreement is reached,
+leading to an increased user satisfaction, higher adoption, and eventually to
+systained software.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
edits in new LT

--- 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 User into Building Sustainable Software for Science}}
+Dan Gunter's main question was, "What is needed to develop sustainable software?".
+Beyond the usual suspects of funding, great developers, good
+design, and software engineering practices, however, Dan placed the
+users. He explained this using a traditional software development model starting
+from gathering requirements, and reaching release through a design and
+development and testing phase. The main deficiency that with this approach was
+pointed out to be the too-late interaction with users. Instead, an alternative
+approach was proposed that, at first, skips the development phase and repeatedly
+goes through requirement gathering, design and user interaction/learning
+phases, and only eventually starts development once an agreement is reached,
+leading to an increased user satisfaction, higher adoption, and eventually to
+sustained software.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add the second talk by Dan Gunter

--- 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 Biology: Lessons Learned on the KBase Project}}
+In this talk, Dan Gunter presented experiences and lessons learned as part of
+KBase, an open software and data platform for addressing the grand challenge of
+systems biology: predicting and designing biological function. KBase is a
+unified system that integrates data and analytical tools for comparative
+functional genomics of microbes, plants, and their communities. However,
+it is also a collaborative environment for sharing methods and results, and
+placing those results in the context of knowledge in the field. Being a large,
+multiinstutional project, one the big challenges is to agree on standard to
+enable a single, maintainable system. Working in isolation does not work (anymore)
+within this field, and no standards for software engineering existed either.
+This is contrasted to computer science research, where these approaches can lead to
+shorter design cycles, leading to more time for highly rewarded activities like
+publishing, performance studies, graduating students, or protecting ideas
+before publication. Sustainability is not of big concern there, compared to
+these. Instead of this more traditional approach, the following is used within
+KBase for large, distributed teams. After picking projects and team members, a
+number of 4-5 teams uses about two weeks to work on their project before a
+1-week evaluation by the executive committee, and after which new teams and
+projects might be chosen to restart the ``agile'' development cycle. This
+process is intentionally open and documented.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
add Yolanda Gil's talk

@@
 Scientists in Best Practices of Software Sharing}}
+Yolanda Gil presented an approach to overcome some hurdles in scientific
+publishing: the disconnect between experimental data, research software, and
+publications. A current effort in the Geosciences is the ``Geoscience Paper of
+the Future'', which includes four elements. Firstly, it forms a modern paper
+including both text, data and pointers to supplementary matertial. Secondly, is
+reproducible, including data processing, workflow and visualization tools. As
+third element, it is part of open science, which includes being publicly
+available under open licenses, and providing meta-data. Last, but not least, it
+uses digital scholarship elements like persisent identifiers for software,
+data, and authors, and it cites both data and software. OntoSoft is an effort
+to help this effort, providing software stewardship for the Geosciences. As
+part of this inituative, a special issue of a journal in a geoscience areas is
+planned that only includes geoscience papers of the future, with submissions
+open until the end of 2015. In addition, training sessions are provided to
+geoscientists to learn best practices in software and data sharing, provenance
+documentation, and scholarly publication.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
Neil's talk

@@
 Framework}}
+Neil Chue Hong presented a proposal to build a scientific software
+accreditation framework. One of the aims of such a framework would be to
+measure how ``good' a given software is, and to evaluate how this can be
+effectively measured in the first place. This can be compared to the effort of
+the standardized, easy to read and understand nutritional labelling of goods,
+which only contain a small set of categories. However, such a framework for
+software would have to be able to evaluate differently due to different
+existing community norms. The challenges such a framework faces include that
+many measurements are subjective, many metrics are too costly, and that
+self-assessment needs to be encouraged. Possible categories would include
+availability, usability, transferability, among others. Such a framework could
+enable both improvement of specific software, as well as comparisons of similar
+software. An accreditation by such a framework could then be part of software
+management plans, ensuring that software is accessible and reusable throughout
+the proposed project and beyond.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
Matthias Bussonnier's talk

@@
 \item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
+This talk highlighted an attempt to solve the common problem for open source
+development that it is difficult to collect information about how many people
+use the software, how often, which version, which parts of the software, or on
+which operating system. Current solutions include surveys, but these have high
+uncertainties. A different approach is based on automatic ``call-backs'' that
+collect these information at run-time and send it to a central place for
+analysis. Problems with this approach include obtaining agreement from the
+user, legal issues, increased maintenance (servers), ethical questions, but
+also the lack of a common infrastructure. Some of these problems are of social
+nature and have to be solved as such, but especially the last problem of a
+missing common infrastructure is attacked byt the sempervirens project, which
+provides common APIs and a library implementation for common, repeating tasks
+like obtaining user consent. Results are uploaded not directly to project
+servers, but to neutral third parties that only publish aggregated statistics
+to projects. Sempervirens can be found on github~\footnote{github:njsmith/sempervirens}.
 

### 2015-11-06 WSSSPE3/report/WSSSPE3_review.tex
fix typo

@@
 nature and have to be solved as such, but especially the last problem of a
-missing common infrastructure is attacked byt the sempervirens project, which
+missing common infrastructure is attacked by the sempervirens project, which
 provides common APIs and a library implementation for common, repeating tasks

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
add Alica Allen's talk

@@
 included)}}
+Alice Allen described the Astrophysics Source Code Library (ascl.net), an increasingly
+used way to obtain a unique ID for astrophysics software that is indexed by indexing
+services and can be cited. ASCL offers clones of existing infrastructure, provides server
+space and computing resources, shares innovations, and permits moves elsewhere. In turn,
+users have to provide a domain nam, control and configure their site and use it as
+intended, gather their codes to their liking, share innovations and protect the
+provided computing environment.
 

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
add Stan Ahalt's talk

@@
 included)}}
-Alice Allen described the Astrophysics Source Code Library (ascl.net), an increasingly
-used way to obtain a unique ID for astrophysics software that is indexed by indexing
-services and can be cited. ASCL offers clones of existing infrastructure, provides server
-space and computing resources, shares innovations, and permits moves elsewhere. In turn,
-users have to provide a domain nam, control and configure their site and use it as
-intended, gather their codes to their liking, share innovations and protect the
-provided computing environment.
+Alice Allen described the Astrophysics Source Code Library (ascl.net), an
+increasingly used way to obtain a unique ID for astrophysics software that is
+indexed by indexing services and can be cited. ASCL offers clones of existing
+infrastructure, provides server space and computing resources, shares
+innovations, and permits moves elsewhere. In turn, users have to provide a
+domain nam, control and configure their site and use it as intended, gather
+their codes to their liking, share innovations and protect the provided
+computing environment.
 
@@
 Community}}
+The effort behind this talk started out as break-out team of an NSF SI2
+workshop in 2015, and centeres around creating a framework for creation of
+metrics for scientific software. This framework would improve both the metrics,
+and the software it evaluates, and tries to build a community around this idea.
+With especially the latter idea in mind, a software ``peer review group'' would
+be created, respresenting stakeholders who will self-review software created by
+their respective communities, and will concurrently develop metrics. The whole
+project should be community-goverend, without a single institution overseeing
+the activity or infrastructure, with the hope to evolving community-generated
+and adopted standards. The generation of metrics would be tied to the actual
+evaluation of software, creating incentive by improving the evaluated software
+itself during this process. Infrastructure for the creationg of metrics and
+evaluation would be provided by the framework code itself, but also provided
+code reviews of the evaluated software, as well as socially through forums for
+generation of software success metrics. With enough examples one of the open
+questions is whether it is possible to fit the resulting metrics in a common
+template. So far, this is still in a design phase, with a white paper at the
+CSESSP workshop 2015 and this talk, but WSSPE workshop are seen as a forum for
+the community to assemble and act, and are planned to be used also in the
+future to build this community and framework.
 

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
fix a lot of typos, mostly added by myself

--- 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
@@
 storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
-presented his wishlist of questions to be answered in the future: 
+presented his wish-list of questions to be answered in the future: 
 %
@@
 The approach of the Titan2D developers included reinforcing the code structure
-using a multiple laters of Python and C++, and a redesign of the data layout to
+using a multiple layers of Python and C++, and a redesign of the data layout to
 be better suitable for modern CPUs and accelerators.
@@
 placing those results in the context of knowledge in the field. Being a large,
-multiinstutional project, one the big challenges is to agree on standard to
+multi-institutional project, one the big challenges is to agree on standard to
 enable a single, maintainable system. Working in isolation does not work (anymore)
@@
 the Future'', which includes four elements. Firstly, it forms a modern paper
-including both text, data and pointers to supplementary matertial. Secondly, is
+including both text, data and pointers to supplementary material. Secondly, is
 reproducible, including data processing, workflow and visualization tools. As
@@
 available under open licenses, and providing meta-data. Last, but not least, it
-uses digital scholarship elements like persisent identifiers for software,
+uses digital scholarship elements like persistent identifiers for software,
 data, and authors, and it cites both data and software. OntoSoft is an effort
 to help this effort, providing software stewardship for the Geosciences. As
-part of this inituative, a special issue of a journal in a geoscience areas is
-planned that only includes geoscience papers of the future, with submissions
+part of this initiative, a special issue of a journal in a Geoscience areas is
+planned that only includes Geoscience papers of the future, with submissions
 open until the end of 2015. In addition, training sessions are provided to
@@
 effectively measured in the first place. This can be compared to the effort of
-the standardized, easy to read and understand nutritional labelling of goods,
+the standardized, easy to read and understand nutritional labeling of goods,
 which only contain a small set of categories. However, such a framework for
@@
 included)}}
-Alice Allen described the Astrophysics Source Code Library (ascl.net), an increasingly
-used way to obtain a unique ID for astrophysics software that is indexed by indexing
-services and can be cited. ASCL offers clones of existing infrastructure, provides server
-space and computing resources, shares innovations, and permits moves elsewhere. In turn,
-users have to provide a domain nam, control and configure their site and use it as
-intended, gather their codes to their liking, share innovations and protect the
-provided computing environment.
+Alice Allen described the Astrophysics Source Code Library (ascl.net), an
+increasingly used way to obtain a unique ID for astrophysics software that is
+indexed by indexing services and can be cited. ASCL offers clones of existing
+infrastructure, provides server space and computing resources, shares
+innovations, and permits moves elsewhere. In turn, users have to provide a
+domain name, control and configure their site and use it as intended, gather
+their codes to their liking, share innovations and protect the provided
+computing environment.
 
@@
 Community}}
+The effort behind this talk started out as break-out team of an NSF SI2
+workshop in 2015, and centers around creating a framework for creation of
+metrics for scientific software. This framework would improve both the metrics,
+and the software it evaluates, and tries to build a community around this idea.
+With especially the latter idea in mind, a software ``peer review group'' would
+be created, representing stakeholders who will self-review software created by
+their respective communities, and will concurrently develop metrics. The whole
+project should be community-governed, without a single institution overseeing
+the activity or infrastructure, with the hope to evolving community-generated
+and adopted standards. The generation of metrics would be tied to the actual
+evaluation of software, creating incentive by improving the evaluated software
+itself during this process. Infrastructure for the creation of metrics and
+evaluation would be provided by the framework code itself, but also provided
+code reviews of the evaluated software, as well as socially through forums for
+generation of software success metrics. With enough examples one of the open
+questions is whether it is possible to fit the resulting metrics in a common
+template. So far, this is still in a design phase, with a white paper at the
+CSESSP workshop 2015 and this talk, but WSSSPE workshop are seen as a forum for
+the community to assemble and act, and are planned to be used also in the
+future to build this community and framework.
 
@@
 on the agenda to justify their attendance, and second, as a way to get new ideas across
-to all the attendees.  We broke with the tradition of requiring the lightning talk submitters
+to all the attendees.  We broke with the tradition of requiring the lightning talk submitter
 to self-publish their papers, and instead used a common peer-review platform\footnote{\url{http://easychair.org}}, choosing to publish the speaker slides on the workshop web site

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
edits in lightning talks

@@
 placing those results in the context of knowledge in the field. Being a large,
-multi-institutional project, one the big challenges is to agree on standard to
+multi-institutional project, one of the big challenges is to agree on standard to
 enable a single, maintainable system. Working in isolation does not work (anymore)
-within this field, and no standards for software engineering existed either.
-This is contrasted to computer science research, where these approaches can lead to
-shorter design cycles, leading to more time for highly rewarded activities like
+within this field, and they also didn't have standards for software engineering.
+This is contrasted to computer science research, where these approaches 
+shorten design cycles, leading to more time for highly rewarded activities like
 publishing, performance studies, graduating students, or protecting ideas
 before publication. Sustainability is not of big concern there, compared to
-these. Instead of this more traditional approach, the following is used within
-KBase for large, distributed teams. After picking projects and team members, a
-number of 4-5 teams uses about two weeks to work on their project before a
+these. \katznote{prev sentence is unclear} Instead of this more traditional approach,
+KBase uses the following for large, distributed teams. After picking projects and team members,
+4-5 teams work on projects about two weeks before a
 1-week evaluation by the executive committee, and after which new teams and
@@
 Scientists in Best Practices of Software Sharing}}
-Yolanda Gil presented an approach to overcome some hurdles in scientific
+Erin Robinson presented an approach to overcome some hurdles in scientific
 publishing: the disconnect between experimental data, research software, and
 publications. A current effort in the Geosciences is the ``Geoscience Paper of
-the Future'', which includes four elements. Firstly, it forms a modern paper
-including both text, data and pointers to supplementary material. Secondly, is
-reproducible, including data processing, workflow and visualization tools. As
-third element, it is part of open science, which includes being publicly
+the Future,'' which includes four elements. First, it forms a modern paper
+including text, data, and pointers to supplementary material. Second, it is
+reproducible, including data processing, workflow, and visualization tools. 
+Third, it is part of open science, which includes being publicly
 available under open licenses, and providing meta-data. Last, but not least, it
@@
 accreditation framework. One of the aims of such a framework would be to
-measure how ``good' a given software is, and to evaluate how this can be
+measure how ``good'' a given piece of software is, and to evaluate how this can be
 effectively measured in the first place. This can be compared to the effort of
-the standardized, easy to read and understand nutritional labeling of goods,
+the standardized, easy to read, and understandable nutritional labeling of goods,
 which only contain a small set of categories. However, such a framework for
-software would have to be able to evaluate differently due to different
+software would be more difficult due to different
 existing community norms. The challenges such a framework faces include that
@@
 This talk highlighted an attempt to solve the common problem for open source
-development that it is difficult to collect information about how many people
-use the software, how often, which version, which parts of the software, or on
+development: it is difficult to collect information about how many people
+use particular software, how often, which version, which parts of the software, or on
 which operating system. Current solutions include surveys, but these have high
 uncertainties. A different approach is based on automatic ``call-backs'' that
-collect these information at run-time and send it to a central place for
+collect these information at runtime and send it to a central place for
 analysis. Problems with this approach include obtaining agreement from the
 user, legal issues, increased maintenance (servers), ethical questions, but
-also the lack of a common infrastructure. Some of these problems are of social
-nature and have to be solved as such, but especially the last problem of a
-missing common infrastructure is attacked by the sempervirens project, which
-provides common APIs and a library implementation for common, repeating tasks
-like obtaining user consent. Results are uploaded not directly to project
+also the lack of a common infrastructure. Some of these problems are of a social
+nature and have to be solved as such, but the last problem (a
+missing common infrastructure) is attacked by the sempervirens project~\cite{sempervirens}, which
+is developing common APIs and a library implementation for common, repeating tasks
+such as obtaining user consent. Results are uploaded not directly to project
 servers, but to neutral third parties that only publish aggregated statistics
-to projects. Sempervirens can be found on github~\footnote{github:njsmith/sempervirens}.
+to projects.
 
@@
 infrastructure, provides server space and computing resources, shares
-innovations, and permits moves elsewhere. In turn, users have to provide a
-domain name, control and configure their site and use it as intended, gather
-their codes to their liking, share innovations and protect the provided
+innovations, and permits moves elsewhere. Users provide a
+domain name, then control and configure their site and use it as intended, gather
+their codes as they wish, share innovations, and protect the provided
 computing environment.
@@
 Community}}
-The effort behind this talk started out as break-out team of an NSF SI2
+The effort behind this talk by Ray Idaszak started from a break-out group at an NSF SI2
 workshop in 2015, and centers around creating a framework for creation of
 metrics for scientific software. This framework would improve both the metrics,
-and the software it evaluates, and tries to build a community around this idea.
+and the software it evaluates, and is trying to build a community around this idea.
 With especially the latter idea in mind, a software ``peer review group'' would
@@
 and adopted standards. The generation of metrics would be tied to the actual
-evaluation of software, creating incentive by improving the evaluated software
-itself during this process. Infrastructure for the creation of metrics and
-evaluation would be provided by the framework code itself, but also provided
-code reviews of the evaluated software, as well as socially through forums for
-generation of software success metrics. With enough examples one of the open
-questions is whether it is possible to fit the resulting metrics in a common
+evaluation of software, creating an incentive by improving the evaluated software
+itself during this process. The framework code would provide infrastructure for the
+creation of metrics and evaluation, and forums for
+generation of software success metrics.  It would also support
+code reviews of the evaluated software. An open
+question is whether it is possible to fit the resulting metrics in a common
 template. So far, this is still in a design phase, with a white paper at the
-CSESSP workshop 2015 and this talk, but WSSSPE workshop are seen as a forum for
-the community to assemble and act, and are planned to be used also in the
+CSESSP workshop 2015 and this talk, but WSSSPE workshop is seen as a forum for
+the community to assemble and act, and is planned to be used also in the
 future to build this community and framework.

### 2015-11-07 WSSSPE3/report/wg_main_funding_spec_expert.tex
Americanizing (and more fixing) spelling, edits, notes in RSE main section

@@
 
-\subsubsection{Why it is important}
+\katznote{I don't think this section title is exactly right - are there some other options?
+Also, the title of Appendix~\ref{sec:appendix_funding_spec_expert} should match, whatever is chosen.
+Maybe `Funding Research Programmer Expertise'?}
+
+%\subsubsection{Why it is important}
 
@@
 and systems of funding do not effectively fund and sustain these skills.
-The resulting high levels of turnover, and inappropriate incentivisation,
+The resulting high levels of turnover, and inappropriate incentives,
 are a significant contributing factor to low levels of reliability and
@@
 develop exactly the skills that research programmers need, yet returns to
-university positions following an industry role are penalised by the current
+university positions following an industry role are penalized by the current
 structures. Retention of senior programmers is hard, because these people
@@
 
-It is hard for an individual PI in a University or College
+It is hard for an individual PI in a university or college
 to support dedicated research programming
@@
 single group. This could be centrally, perhaps within research computing
-organisations which have traditionally supported university cyberinfrastructure,
-library organisations, or research offices. Alternatively, these groups could
-be organisationally closer to research groups, sitting within academic
+organizations which have traditionally supported university cyberinfrastructure,
+library organizations, or research offices. Alternatively, these groups could
+be organizationally closer to research groups, sitting within academic
 departments. The most
@@
 
-This might come from unversities themselves (this was the model that led to the
+This might come from universities themselves (this was the model that led to the
 creation of the group in University College London), but more likely, seed
@@
 Engineering Fellowship provided by the UK Engineering and Physical Sciences
-Research Council). We therefore recommend that funding organisations consider
+Research Council). We therefore recommend that funding organizations consider
 how they might provide such seed funding.
@@
 software engineer'', has been important to the campaign. It will be useful to
-the global effort to similarly identify emerging research software organisations,
+the global effort to similarly identify emerging research software organizations,
 and also, importantly, to identify longer-running research software groups,
@@
 now can be identified as part of a wider solution.
+
+For more detail on the discussion, see Appendix~\ref{sec:appendix_funding_spec_expert}.
 
@@
 self-identify as research software engineering groups. We will collect information
-as to the organisational models under which these groups function, and how they
+as to the organizational models under which these groups function, and how they
 are funded.
@@
 
-http://www.rse.ac.uk/groups
+To find more information about this group, or join it, see \url{http://www.rse.ac.uk/groups}.

### 2015-11-07 WSSSPE3/report/wg_main_industry_interaction.tex
notes/changes in Industry Interaction section

@@
 
-\subsubsection{Why it is important}
+%\subsubsection{Why it is important}
 
@@
 
-At WSSSPE 3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. 
+At WSSSPE3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. 
 (This may be an initially awkward phrase---some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. 
@@
 See Becker et al.~\cite{Becker:2014} for extended discussion of weak vs.\ strong sustainability.
+For more detail on the group's discussion, see Appendix~\ref{sec:appendix_industry_interaction}.
 
@@
 
-\todo{where should someone go who want to know more about this and perhaps wants to contribute?}
+To find more information about this group, or join it, see \todo{where should someone go who want to know more about this and perhaps wants to contribute?}

### 2015-11-07 WSSSPE3/report/wg_main_user_community.tex
edits & suggestions in user community section

@@
 \subsection{Building Sustainable User Communities for Scientific Software}
-%\todo{change the title to your topic}
 
-\subsubsection{Why it is important}
-%\todo{short text here}
+%\subsubsection{Why it is important}
 
@@
 \subsubsection{Fit with related activities}
-%\todo{short text here - can include links/cites}
 
-Mozilla Science maintains a ``Working Open' Project Guide'' (http://mozillascience.github.io/leadership-training). From the introduction:
+Mozilla Science maintains a ``Working Open Project Guide'' (\url{http://mozillascience.github.io/leadership-training}). \katznote{should get changed to a cite} From the introduction:
 \begin{quote}
@@
 \begin{itemize}
-\item ``Art of Community'' by Jono Bacon. We could consider distilling this for scientific software.
-\item Iain Larmour, from EPSRC inthe UK (EPSRC: https://www.epsrc.ac.uk/) (not sure who from mentioned UK Collaborative Computational Projects (CCP: http://ccp.ac.uk)
+\item ``Art of Community'' by Jono Bacon. \katznote{cite?} We could consider distilling this for scientific software.
+\item Iain Larmour, from EPSRC in the UK (EPSRC: https://www.epsrc.ac.uk/) (not sure who from mentioned UK Collaborative Computational Projects (CCP: http://ccp.ac.uk)
 \end{itemize}
@@
 \subsubsection{Discussion}
-%\todo{short-ish text here}
 
@@
 to a need not yet recognized by the community. The group also thought that the existing books on software communities would
-need to be evaluated in light of differences between Science Software projets and general OSS projects in terms of scale,
+need to be evaluated in light of differences between Science Software projects and general OSS projects in terms of scale,
 science, acknowledgement and credit, and funding models.
 
+Additional details on the group's discussion can be found in Appendix~\ref{sec:appendix_user_community}.
+
 \subsubsection{Plans}
-%\todo{short text here - not bullets}
 

### 2015-11-07 WSSSPE3/report/WSSSPE3_review.tex
minor change - formatting
removing number of groups in intro

@@
 attendees to quickly highlight an important issue or a potential solution. The
-report also gives summaries of action plans proposed by the eleven working
+report also gives summaries of action plans proposed by the working
 groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote \note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}} \label{sec:keynote} 
+\section{Keynote \label{sec:keynote}}
+%\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\todo{need to update for WSSSPE3}
-\todo{KEN: Should these be bolded? - dsk: yes, titles can be bolded, but contents under each should not}
 \begin{comment}

### 2015-11-09 WSSSPE3/report/WSSSPE3_review.tex
removing comment
Minor changes.

@@
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with SC13,
+held in conjunction with SC13;
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
 focused workshop organized in July 2014 jointly with the SciPy
-conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}},
+conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
 WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
-held in conjunction with SC14, and
+held in conjunction with SC14; and
 WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
@@
 community to easily find up-to-date information on a WSSSPE topic (e.g.,
-software credit, scientific software metrics, testing scientific software)
+software credit, scientific software metrics, testing scientific software).
 \end{itemize}
@@
 
-\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}
+%\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}
 

### 2015-11-09 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Fix typos

@@
 in providing funding to establish local groups. These work best when they
-develop good collaborations with national cyberinfrastucture pools.
+develop good collaborations with national cyberinfrastructure pools.
 When an organization assembles
@@
 hardware through software improvements has been useful politically, (and many
-RSE groups are cited in organisations which host clusters for this reason), but
+RSE groups are cited in organizations which host clusters for this reason), but
 the time has come to make the case that software itself is a critical
@@
 
-We think support from non-governmental organisations will be important if RSE
+We think support from non-governmental organizations will be important if RSE
 groups will become established.
@@
 We recommend that funders give explicit guidance to grant holders and
-institutions that such arrangement are favourable. Framework agreements
+institutions that such arrangement are favorable. Framework agreements
 permitting this to go ahead without
@@
 There is as yet no clear answer as to the scale of aggregation needed to
-make such a program work. A university wide programme allows enough scale to
+make such a program work. A university wide program allows enough scale to
 be robust to fluctuations of funding within one field.
@@
 The desire to aggregate enough work to make it sustainable, and the need to
-have domain-relevant research programming skils, are in tension.
+have domain-relevant research programming skills, are in tension.
 
@@
 biosimulation or plasma physics. These bodies act as custodians of community codes,
-and a central team also provides software engineering support support.
+and a central team also provides software engineering support.
 
@@
 cyberinfrastructure component is clear. Funding which permits code to be
-refactored, tidied and optimised is rare; this is often done ``on the sly''
+refactored, tidied and optimized  is rare; this is often done ``on the sly''
 in a scientifically focused grant. The UK EPSRC's ``software for the future''
-call, which really permits explicit investiment in software as an infrastructure,
+call, which really permits explicit investment in software as an infrastructure,
 is so oversubscribed as to have a FOUR PERCENT success rate; the demand is clear!
@@
 
-We will seek to identify and approach existing research programming organisations,
+We will seek to identify and approach existing research programming organizations,
 to get their permission to list them on a list of research software groups.
@@
 is needed if this campaign is to succeed. This will require an initial gathering
-of identified research software organisations in the US to this end.
+of identified research software organizations in the US to this end.
 
@@
 Financial support for an initial conference to bring together research software
-groups to form an organisation and create a resource sharing structure would
+groups to form an organization and create a resource sharing structure would
 help to further this campaign.
 
-In the longer term, funding organisations, especially non-governmental organisations
+In the longer term, funding organizations, especially non-governmental organizations
 with the capability to effect innovation through seed funding, could provide

### 2015-11-09 WSSSPE3/report/wg_appendix_software_credit.tex
Fix typos

@@
 In addition, neither of these options allow for the citation of software itself, but only provide an author contribution related to software.
-The discussion quickly focused on transtitive credit as a more quantitative measure of allocating credit to both authors and software, although there were some concerns about authors overestimating their own contributions compared to prior work.
+The discussion quickly focused on transitive credit as a more quantitative measure of allocating credit to both authors and software, although there were some concerns about authors overestimating their own contributions compared to prior work.
 
@@
 That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
-One important use of this system is to track rights and credit for contributors to entertainment works in order to distrubute revenues---similar to the proposed transitive credit concept.
+One important use of this system is to track rights and credit for contributors to entertainment works in order to distribute revenues---similar to the proposed transitive credit concept.
 

### 2015-11-09 WSSSPE3/report/wg_appendix_publishing_SW.tex
minor changes/comments in publishing sections moved some credit text to be a general intro to the section
Fix typos

@@
 figures, etc.) directly.
-\item O'Reily Media: announced that it plans to make IPython Notebooks a first-class
+\item O'Reilly Media: announced that it plans to make IPython Notebooks a first-class
  authoring environment for their publishing program alongside their existing mechanisms.
@@
 Create the first version of the web page to be displayed on the Software Sustainability
-Institute's website.
+Institute's website: \url{http://www.software.ac.uk}. \katznote{when?}
 

### 2015-11-09 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Fix typos

@@
 us sit at different points on that spectrum, and move along it during our
-careers. (Ususally from RA to RSE -- examples of movement in the other direction
+careers. (Usually from RA to RSE -- examples of movement in the other direction
 from readers would be welcomed.)
@@
 Another example of a success is the development of semantics and linked data in
-support of Ocean Sciences. An Earthcube funded project pairs domain scientists
+support of Ocean Sciences. An EarthCube funded project pairs domain scientists
 with RSEs and has been successful as the semantics attached have increased data
@@
 the context of other SW developers at University of Washington.
-These scientists work in the e-Science Studio/dataScience Studio, and they
+These scientists work in the e-Science Studio/Data Science Studio, and they
 help a group of graduate students in solving their problems in data

### 2015-11-09 WSSSPE3/report/WSSSPE3_review.tex
Fix typos

@@
 of software as academic output, and rejecting submissions based on
-unreproducible computational results. Some advantages of research software
+irreproducible computational results. Some advantages of research software
 engineering (RSE) groups include the possibility of training in reproducible

### 2015-11-09 WSSSPE3/report/wg_main_metrics.tex
comments

@@
 \item
-\href{https://geodynamics.org/cig/dev/best-practices/}{(Computational Infrastructure for Geodynamics: Software Development Best Practices)}
+Computational Infrastructure for Geodynamics: Software Development Best Practices\footnote{\url{https://geodynamics.org/cig/dev/best-practices/}}
 
 \item
-\href{https://docs.google.com/document/d/1cgUDH3RxrfsLotWhKKOrXUnaYFhrtjcV1TDRkFtwQKI/edit}{(WSSSPE3 Breakout Session: How can we measure the impact of a code on research, and its value to the community?)}
+WSSSPE3 Breakout Session: How can we measure the impact of a code on research, and its value to the community?\footnote{\url{https://docs.google.com/document/d/1cgUDH3RxrfsLotWhKKOrXUnaYFhrtjcV1TDRkFtwQKI/edit}}
 
 \item
-\href{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/edit#heading=h.5lah0hp73q99}{(2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics)}
+2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics\footnote{\url{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/edit#heading=h.5lah0hp73q99}}
 
 \item
-\href{https://docs.google.com/document/d/1uDim5bw8rBuubmtaUrz5Eh35NxzDgivmmdXhVzDs3tc/edit}{(2015 NSF SI2 PI Workshop Breakout Session on Software Metrics)}
+2015 NSF SI2 PI Workshop Breakout Session on Software Metrics\footnote{\url{https://docs.google.com/document/d/1uDim5bw8rBuubmtaUrz5Eh35NxzDgivmmdXhVzDs3tc/edit}}
 
 \item
-\href{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/edit#slide=id.g5e66ec9f2_027}{(NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics)}
+NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics\footnote{\url{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/edit#slide=id.g5e66ec9f2_027
+}}
 
 \item
-\href{http://www.software.ac.uk/software-evaluation-guide}{(U.K. Software Sustainability Institute Software Evaluation Guide)}
+U.K. Software Sustainability Institute Software Evaluation Guide\footnote{\url{http://www.software.ac.uk/software-evaluation-guide}}
 
 \item
-\href{http://www.software.ac.uk/blog/2013-04-09-five-stars-research-software}{(U.K. Software Sustainability Institute Blog post: The five stars of research software)}
+U.K. Software Sustainability Institute Blog post: The five stars of research software\footnote{\url{http://www.software.ac.uk/blog/2013-04-09-five-stars-research-software}}
 
 \item
-\href{http://figshare.com/articles/Minimal_information_for_reusable_scientific_software/1112528}{(Minimal information for reusable scientific software)}
+Minimal information for reusable scientific software\footnote{\url{http://figshare.com/articles/Minimal_information_for_reusable_scientific_software/1112528}}
 
 \item
-\href{http://equipment.data.ac.uk/}{(EPSRC-funded Equipment Data Search Site)}
+EPSRC-funded Equipment Data Search Site\footnote{\url{http://equipment.data.ac.uk/}}
 
 \item
-\href{http://www.canarie.ca/software/}{(Canarie Research Software: Software to accelerate discovery)}
+Canarie Research Software: Software to accelerate discovery\footnote{\url{http://www.canarie.ca/software/}}
 
 \item
-\href{https://science.canarie.ca/researchmiddleware/platforms/list/main.html}{(Canarie Research Software: Research Software Platform Registry)}
+Canarie Research Software: Research Software Platform Registry\footnote{\url{https://science.canarie.ca/researchmiddleware/platforms/list/main.html}}
 
 \item
-\href{https://collaboration.canarie.ca/elgg/file/view/2471/research-platform-support-for-the-canarie-registry-and-monitoring-system-revision-3}{(collaboration@CANARIE post: platform support for Canarie registry and monitoring system)}
+BlackDuck Open HUB\footnote{\url{https://www.openhub.net/}}
 
 \item
-\href{https://collaboration.canarie.ca/elgg/file/view/2453/research-service-support-for-canarie-registry-and-monitoring-system-revision-7}{(collaboration@CANARIE post: service support for Canarie registry and monitoring system)}
-
-\item
-\href{https://www.openhub.net/}{(BlackDuck Open HUB)}
-
-\item
-\href{https://www.innovationpolicyplatform.org/frontpage}{(Innovation Policy Platform)}
+Innovation Policy Platform\footnote{\url{https://www.innovationpolicyplatform.org/frontpage}}
 
@@
 creation of a white paper on this topic.  This resulted in a paper outline and writing assignments with the goal of publishing in venues including 
-WSSSPE4, IEEE CISE, or JORS.
+WSSSPE4, IEEE CISE, or JORS. More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
 
@@
 
-The group will use Google Docs to proceed with the authoring of the white paper in lieu of a group project landing page.  Links to the resulting white paper will be provided when completed.
+In lieu of a landing page, the Useful Metrics for Scientific Software working group requests an email be sent to Gabrielle Allen \href{mailto:gdallen@illinois.edu}{(gdallen@illinois.edu)} to find out more about the group's efforts and how to participate.

### 2015-11-09 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Update wg_appendix_funding_spec_expert.tex

@@
 
-he power of the label? Getting the word out about RSE support using the label.
+The power of the label? Getting the word out about RSE support using the label.
 
@@
 And this brings up the flip side question, “Do scientists really want to be writing code?”
+
 Career Path.
-We also had a little discussion about how to make a career path for research developers. It need not be solely an academic enterprise, but tenure is always problematic for people of this class.
+We also had a little discussion about how to make a career path for research developers.
+It need not be solely an academic enterprise, but in the past tenure has often been problematic for people of this class.
 
@@
 high levels of communication between groups will be valuable. In the UK, there
-are plans to permit resource sharing between institutional RSE groups.
+are plans to permit resource sharing between institutional RSE groups. Perhaps 
+there are circumstances under which an RSE skill exchange could be arranged, either
+formally or informally.
 
@@
 requiring an academic collaborator.
+
 Domain scientists and developers are funded together in a single proposal.
 Another example of a success is the development of semantics and linked data in
-support of Ocean Sciences. An EarthCube funded project pairs domain scientists
-with RSEs and has been successful as the semantics attached have increased data
+support of Ocean Sciences. An EarthCube-funded project pairs domain scientists
+with RSEs and has been successful; the semantics attached have increased data
 use and discovery significantly.
@@
 part of national compute services. The US XSEDE project's extended collaboration
- service (ECS) is a set of developers who are paid with XSEDE funding,
- and are on “permanent” staff.  When PIs request allocations on XSEDE resources,
- there is a finite pool of developer time that can be awarded,
-typically for one year only, and at partial effort, typically 20 percent or so.
-The finite time allowed provides motivation for the scientist and their group to
-work closely the developer and to become educated in what the developer is doing,
-so they can sustain the effort, once the ECS period is over. This funding
-mechanism can be highly efficient for scientific problems, because the developer
-pool assembled by the research providers are, by definition, expert in the
-characteristics of their specific resource, and can very quickly assess the
-scientists needs, and what it will take to implement software that meets the
-user’s needs. However, it does not develop capacity within institutions,
-and indeed can have a ``chilling effect''.
+service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
+“permanent” staff.  When PIs request allocations on XSEDE resources, there is a 
+finite pool of developer time that can be awarded, typically for one year only, 
+and at partial effort, typically 20 percent or so. The finite time allowed provides
+motivation for the scientist and their group to work closely the developer and to 
+become educated in what the developer is doing, so they can sustain the effort, once 
+the ECSS period is over. This funding mechanism can be highly efficient for scientific
+problems, because the developer pool assembled by the research providers are, by definition, 
+expert in the characteristics of their specific resource, and can very quickly assess the
+scientists needs, and what it will take to implement software that meets the user’s needs. 
+However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
+program, it should not be relied upon as a long term solution to acquiring this type of capacity.
 
@@
 develop good collaborations with national cyberinfrastructure pools.
-When an organization assembles
-a developer pool, diversity is developed and skills can be transferred.
+When an organization assembles a developer pool, diversity is developed and 
+skills can be transferred.
 
-We would like to see these models applied outside High Performance Computing;
-most scientific software is not destined to run on national cyberinfrastructure,
+We would like to see these models applied outside High Performance Computing.
+Most scientific software is not destined to run on national cyberinfrastructure,
 but needs similar support. The argument regarding making better use of expensive
@@
 
-The Canarie group (Canada) accepts proposals
-for providing services to broad communities, integrate people who are doing
-things that are complementary, the goal is to make the available stack more
-robust and richer for everyone. They offer short rounds of funding that can
+The Canarie group (Canada) accepts proposals for providing services to broad communities, 
+integrate people who are doing things that are complementary, the goal is to make the available 
+stack more robust and richer for everyone. They offer short rounds of funding that can
 have as a key metric creating some useful functionality that shows a diversity
@@
 
-There can be problems communicating across cultural barriers, with domain
-scientists seeing developers as “other”.
-Collaboration, and tools to fund, encourage or motivate collaboration are
+There can be problems communicating across cultural barriers, with domain scientists seeing developers 
+as “other”. Collaboration, and tools to fund, encourage or motivate collaboration are
 extremely important.
 
-We think support from non-governmental organizations will be important if RSE
-groups will become established.
-The Sloan Foundation is currently funding data science engineers, who work in
-the context of other SW developers at University of Washington.
-These scientists work in the e-Science Studio/Data Science Studio, and they
-help a group of graduate students in solving their problems in data
-science and data management. During Fall and Spring, a 10-week incubator program
-allows students to work two days a week to work on a data-intensive science
-project. Some fraction of the developer time is dedicated to the developers'
-personal interests as well as instruction.
+We think support from non-governmental organizations will be important if RSE groups will become established.
+The Sloan Foundation is currently funding data science engineers, who work in the context of other SW developers
+at University of Washington. These scientists work in the e-Science Studio/Data Science Studio, and they
+help a group of graduate students in solving their problems in data science and data management. During Fall and
+Spring, a 10-week incubator program allows students to work two days a week to work on a data-intensive science
+project. Some fraction of the developer time is dedicated to the developers' personal interests as well as instruction.
 
@@
 
-There is some question about the most effective
-duration and percent of full time for a programmer's work on a project.
-At least three months is necessary for the programmer to read into the science
-(RSEs must not become so disengaged from research that they don't have time to
-read a few papers -- this will result in code which doesn't meet scientific
-needs), but too long could result in an RSE losing their flexibility, becoming
-so engaged in one project that when that project ends, they find it hard to
-transfer. For this reason, we also recommend that 40 percent is ideal;
-two projects per developer, with some time for training and infrastructure work.
-Two developers per project seems to be ideal, in the sense that software
-development is enhanced by two pairs of eyes.
+There is some question about the most effective duration and percent of full time for a 
+programmer's work on a project. At least three months is necessary for the programmer to 
+read into the science (RSEs must not become so disengaged from research that they don't have
+time to read a few papers -- this will result in code which doesn't meet scientific needs), 
+but too long could result in an RSE losing their flexibility, becoming so engaged in one project
+that when that project ends, they find it hard to transfer. For this reason, we also recommend
+that 40 percent is ideal; two projects per developer, with some time for training and infrastructure work.
+Two developers per project seems to be ideal, in the sense that software development is enhanced by 
+two pairs of eyes.
 
@@
 to get their permission to list them on a list of research software groups.
+Casual conversation during the meeting made it clear that although the title is not
+widely used in the US, the this position is not rare. We spoke with several individuals
+who, at distinct Universities, had RSEs (in effect if not in name) and these were funded
+and under differing models.
+
 We will also look for examples of groups which have successfully become self-
-sustaining following
+sustaining following.
+
+In this respect, information gatheirng via a survey subsequent analysis could be very useful.
+We would need to assemble a list of targetted individuals (what position and rank is
+likely to know and care enough to respond?) Perhaps the Science Gateway Insitute has already acquire
+information that could be helpful to advance this issue, and/or craft the proper survey and target
+individuals.
 
@@
 groups to form an organization and create a resource sharing structure would
-help to further this campaign.
+help to further this campaign. Funding to conduct and analyze a survey could also be quite useful
+as knowing where we stand today, and what models are in use could fuel the ideas for further
+development of developers in this category.
 
@@
 with the capability to effect innovation through seed funding, could provide
-support to nucleate the creation of research software groups.
+support to nucleate the creation of research software groups. As noted above, Sloan has already
+initiated one such program, and collaboration with Slaon or at least study of their methods and 
+success or failure could be extremely useful in approaching Universities and other insitutions 
+in funding this development track. It seems clear that if the value proposition can be made to
+University adminstrators, this track could flourish with buy-in at the adminstrative level.

### 2015-11-09 WSSSPE3/report/wg_main_funding_spec_expert.tex
Update wg_main_funding_spec_expert.tex

@@
-\subsection{Research Programming Expertise -- how to fund it?}
+\subsection{Funding Research Programmer Expertise}
 \label{RSE}
@@
 Research Software Engineers -- those who contribute to science and scholarship
-through computer programming -- are an important part of the team
+through software development -- are an important part of the team
 needed to deliver 21st century research. However, existing academic structures
 and systems of funding do not effectively fund and sustain these skills.
-The resulting high levels of turnover, and inappropriate incentives,
+The resulting high levels of turnover and inappropriate incentives
 are a significant contributing factor to low levels of reliability and
-readability observed in scientific software.
+readability observed in scientific software. Moreover, the absence of skilled and experienced 
+developers retards progress in key projects, and at times causes important projects to fail completely.
 
-For effective development of software for advanced research, it is necessary
-that researchers can work with scientific programmers who are able to understand
-the research domain sufficiently to be able to build meaningful software. This
-requires a collaborative approach -- programmers fully engaged in the research
-context, co-developing software with domain academics.
+Effective development of software for advanced research requires
+that researchers work closely with scientific software developers who understand
+the research domain sufficiently to build meaningful software at a reasonable pace. 
+This requires a collaborative approach -- where developers who are fully engaged/invested 
+in the research context are co-developing software with domain academics.
 
@@
 
-One solution to this problem would be a successful solution to the problems of
-obtaining academic credit for programming work. If this were solved, traditional
-academic funding models and career tracks would properly sustain this
-contribution to research. However, in the absence of guaranteed success in that
-stream, this working party seeks to look to an alternate solution; the
-development and funding of dedicated research programming roles. We also noted
-that measures of impact and success for research programming
-groups, as well as for individual research software engineers, will be required
-in order to make the case for continued funding. Research programmers will not
-be measured by publications, we hope, but by other measures. Middle-author
-publications are common for RSEs. Most RSEs welcome co-authorship on papers
-where the PIs consider the contribution deserves it.
+The solution we envision entails creating an environment where software developers 
+are a stable part of the research team. Such an environment mitigates the risk
+of losing a key developer at a critical moment in a projects lifetime, and provides
+the benefits of building a store of institutional knowledge about specific projects as well as about software 
+development for today's research. Our vision is to find a way to promote a University/research institute environment where
+software developers are stable components of research project teams.
 
-A contributing factor to the problem is the devastating effect on an academic
-publication record resulting from time in industry; such postings often
-develop exactly the skills that research programmers need, yet returns to
-university positions following an industry role are penalized by the current
-structures. Retention of senior programmers is hard, because these people
-are highly in demand by the economy. However,
-people who have a PhD in science and enter industry,
-may desire to return for diverse reasons, and should be welcomed back.
+One strategy to promote stability is implementing a mechansim for developers to obtain academic
+credit for software development work. With such a mechanism in place, traditional academic funding
+models and career tracks could properly sustain individuals for whom software development is their
+primary contribution to research. A contributing factor to the problem with the current academic reward system is the
+devastating effect on an academic
+publication record resulting from time in industry; such postings often develop exactly the skills that research programmers
+need, yet returns to university positions following an industry role are penalized by the current structures. Retention of
+senior programmers is hard, because these people are highly in demand by the economy. However, people who have a PhD in
+science and enter industry, may desire to return for diverse reasons, and should be welcomed back.
+
+While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
+change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
+sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
+developing dedicated research programming roles within the University, and finding stable funding for those individuals is
+the most promising mechanism for creating a stable software development staff.
+
+Measures of impact and success for research programming groups, as well as for individual research software engineers, will
+be required in order to make the case to the University for continued funding.
+Research programmers will not be measured by publications, we hope, but by other measures.
+(what measures?) Middle-author publications are common for RSEs.
+Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
+(these last two statements seem contradictory)
 
@@
 
-It is hard for an individual PI in a university or college
-to support dedicated research programming
-resources, as the need for, and funding for, these activities is intermittent
-within the research cycle. To sustain this capacity, therefore, it is
+It is hard for an individual PI in a university or college to support dedicated research programming resources, as the need
+for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is
 necessary to aggregate this work across multiple research groups.
 
-One solution is to fund dedicated programming roles for major research software
-projects at national laboratories or other non-educational institutions. This
-solution is working well for many well-used scientific codebases.
-However, much of the body of software for which this expertise is needed sits
-within the longer tail of research in universities. Therefore, we argue that
-institutions need to develop hybrid academic-technical roles for this capacity
-who work with more than one PI, rather than the traditional RA role within a
-single group. This could be centrally, perhaps within research computing
-organizations which have traditionally supported university cyberinfrastructure,
-library organizations, or research offices. Alternatively, these groups could
-be organizationally closer to research groups, sitting within academic
-departments. The most
-effective model will vary from institution to institution,
-but the mandate and ways of working should be similar.
+One solution is to fund dedicated programming roles for major research software projects at national laboratories
+or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
+However, this strategy has limited application, as much of the body of software is created and maintained in research
+universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
+capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
+. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
+traditionally supported university cyberinfrastructure, libraryorganizations, or research offices. Alternatively, these
+groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
+will vary from institution to institution, but the mandate and ways of working should be similar.
 
-Having convinced ourselves that this would be a positive innovation, we were
-then faced with the specific question of how to fund the initiation of this
-activity. A self-sustaining research software group will support itself
-through collaborations with PIs in the normal grant process, with PIs choosing
-to fund some amount of research programming effort through grants in the usual
-way. However, to bootstrap such a function to a level where it has the
-reputation and client base to be self-sustaining requires seed investment.
+Having convinced ourselves that this would be a positive innovation, we were then faced with the specific question of how to
+fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
+with PIs in the normal grant process, with PIs choosing to fund some amount of research programming effort through grants in
+the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
+-sustaining will generally require seed investment.
 
-This might come from universities themselves (this was the model that led to the
-creation of the group in University College London), but more likely, seed
-funding needs to come from research councils (as with the Research Software
-Engineering Fellowship provided by the UK Engineering and Physical Sciences
-Research Council). We therefore recommend that funding organizations consider
-how they might provide such seed funding.
+This might come from universities themselves (this was the model that led to the creation of the group in University College
+London), but more likely, seed funding needs to come from research councils (as with the Research Software
+Engineering Fellowship provided by the UK Engineering and Physical Sciences Research Council). We therefore recommend that
+funding organizations consider how they might provide such seed funding.
 
-Success, appropriately measured, will help make the case for further investment.
-However, measurement of code cleanliness, and the resulting productivity gains,
-is an unsolved problem in empirical software engineering. To measure
-``what did not go wrong'' because of an intervention is particularly hard.
+Success, appropriately measured, will help make the case to such funding bodies for further investment.
+One might expect that metrics such as improved productivity, software adoption rates, and grant success rates would be sufficient arguments in favor of such a model.
+However, useful measurement of code cleanliness, and the resulting productivity gains, is an unsolved problem in empirical software engineering.
+To measure ``what did not go wrong'' because of an intervention is particularly hard.
 
-We finally noted that the institutional case for such groups is made easier
-by having successful examples to point to. In the UK, a collective effort to
-identify this community, with individuals clearly stating ``I'm a research
-software engineer'', has been important to the campaign. It will be useful to
-the global effort to similarly identify emerging research software organizations,
-and also, importantly, to identify longer-running research software groups,
-which have in some cases had a long running \emph{sui-generis} existence, but which
-now can be identified as part of a wider solution.
+We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
+UK, a collective effort to identify this community, with individuals clearly stating ``I'm a research software engineer'',
+has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software
+organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a
+long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution.
+There remains the problem of how to "sell" the value of this investment to investigators in the university will be problem left to the individual universities, once the initial investment is made.
+This is an issue best addressed by th eindividual organizations that embark on the plan.
 
@@
 
-We will therefore seek to gather a list of groups which are prepared to
-self-identify as research software engineering groups. We will collect information
-as to the organizational models under which these groups function, and how they
-are funded.
+The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSE's who do not identify themselves as such at this time.
+We will collect information as to the organizational models under which these groups function, and how they are funded.
+For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not.
+Are these developers paid by the University or through a program supported by research grants/individual PIs?
+How did they bootstrap the developer track to get this started?
+How successful is the university in getting investigators to pay for fractional RSEs?
+We will author a report describing our findings, should funding be available to conduct the investigation.
 

### 2015-11-10 WSSSPE3/report/wg_appendix_training.tex
Create wg_appendix_training.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Final draft of full content

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Training Working Group Discussion}
+\label{sec:appendix_training}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Nick Jones\footnote{email: \href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}} will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+\begin{itemize}
+\item Nick Jones -- New Zealand eScience Infrastructure
+\item Iain Larmour -- Engineering \& Physical Sciences Research Council (UK)
+\item Erin Robinson -- Foundation for Earth Science
+\end{itemize}
 
@@
 
+\paragraph{}
+While little training focuses specifically on this outcome, a variety of training activities grow researcher awareness of and engagement with software professionals and software engineering practices.
+Research Software Engineers are being recognised as critical contributors to high quality research; the pathway to acquire and master the relevant skills isn’t yet clear; equally those skills required by researchers in general are also not commonly understood nor routinely developed.
+\paragraph{}
+Discussion explored a rapidly growing array of training which is seen to contribute to sustainable software.
+Offerings are diverse, including: self paced online modules focused around specific tools; single and multiple day training workshops that raise awareness of a tool chain to support collaborative and shared software development within a research workflow; block courses specialising on particular methods, technologies, and applications; academic programmes at undergraduate and masters levels; doctoral training programmes that in part contain requisite skills training activities.
+\paragraph{}
+While some of this training focuses on applying software engineering practices within the context of research, meeting the values and goals of research are less often incorporated as explicit learning outcomes.
+With software (and similarly, data) often the only tangible artefact of a research method or protocol, the dependency between software applications, and the quality of research adds complexity to the learners journey.
+In recognition of the longer term investment required by researchers to integrate such skills into their research practices, many activities are focusing on emotionally engaging researchers and cohorts, to build a sense of shared purpose beyond the obvious goal of technical skill acquisition.
+\paragraph{}
+In reviewing current training activities, the group identified a variety of perspectives seen as useful in positioning activities in ways to better communicate why and when best to apply each activity.
+Training can be categorised on a variety of spectra, with content and delivery focused from: programming to research; basic to advanced; technical to emotional; informal to formal; self paced to participative.
+A few attempts have been made to situate a cross section of training activities within such dimensions, creating easier means of communicating the value of any specific opportunity and the pathways across opportunities over time.
+\paragraph{}
+Evaluation of training delivery and outcomes is seen as a weakness common to most non academic training activities.
+Opportunities for measuring success in delivering training start simply with collecting a Net Promoter Score, which lets those delivering training know whether attendees are likely to recommend the training to others.
+In looking at the longer term outcomes for the learner, frameworks such as Blooms Taxonomy and Kirkpatrick Evaluations offer possible approaches.
+\paragraph{}
+In this latter case of formal evaluation, ownership of evaluation as a component of career development for any researcher appears mostly absent.
+While academic research institutions have professional development centres to support research staff, the skills taught which might impact on sustainable software are limited at best, and lack a clear and coherent development pathway.
+\paragraph{}
+Coordination of these training projects will depend on buy in from a broad range of training programme and activity leaders, suggesting a key opportunity lies in identifying and bringing together these people on a regular basis.
+
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+\paragraph{}
+Software skills are needed by an increasing array of researchers and fields.
+The training arc is not well-defined, with a sometimes baffling array of training opportunities responding to various facets of skill deficit and need.
+Given this current complexity, coordination across training projects would create common frames of reference, communicating and integrating activities to better serve the needs of researchers.
+\paragraph{}
+Building this community could lift the maturity of training projects and capabilities, enabling more advanced approaches to address key gaps in evaluation, career development, and a lift in the standard of research practices.
+\paragraph{}
+In aiming at these opportunities it will be necessary to find the means to support those involved in leading training activities to allocate time to coordination activities, which will often sit beyond their current scope of responsibility.
+\paragraph{}
+These activities are also distributed globally, with no single country or region offering a comprehensive set of capabilities and initiatives.
+Any coordination activity will therefore need to raise the profile of the opportunity gap with relevant research funders and policy makers.
 
 \subsection{Key Next Steps}
+\paragraph{}
+The goal of these next steps is to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects. The following:
+\begin{enumerate}
+    \item Hold a virtual meeting by December 2015, to bring together a broader group of interest in this topic, with specific goals to:
+    \begin{enumerate}
+        \item Identify programmes with existing activity aimed at integrating across training projects.
+        \item Identify training projects with an interest in participating in coordination efforts.
+        \item Identify funding opportunities to bring together training programme and project leaders to identify shared goals for future coordination of activities.
+        \item Agree a communications plan to qualify whether programmes, projects, and funders are interested in engaging and committing to ongoing activities.
+    \end{enumerate}
+    \item Review progress within 3 months, to establish next steps, if any.
+\end{enumerate}
 
 \subsection{Plan for Future Organization}
+\paragraph{}
+Continue to track progress by posting comments to WSSSPE3 issue.
 
 \subsection{What Else is Needed?}
+\paragraph{}
+If the group moves from early stage formation into working towards shared goals, expertise will likely be required in pedagogy and training evaluation
 
 \subsection{Key Milestones and Responsible Parties}
+\begin{enumerate}
+    \item During October, Nick Jones and Erin Robinson to draft WSSSPE3 report back. 
+    \item Before December 2015, Nick Jones and Erin Robinson to call a meeting of the broader group, to review key next steps. 
+    \item 1st quarter 2016 - if willing parties identified, draft workshop proposal and identify relevant forum, including future WSSSPE events
+\end{enumerate}
 
 \subsection{Description of Funding Needed}
+\paragraph{}
+Workshop/RCN travel funding to bring together key programme, project, and funder representatives from across North America, EU, UK, Australasia.
+In addition, funding to support work on better defining the landscape of training activities, the useful perspectives in communicating the value of the varied training projects, and the possible pathways through training activities over time.

### 2015-11-10 WSSSPE3/report/wg_main_training.tex

\subsection{Topic} \todo{Training Working Group Discussion}

\subsubsection{Why it is important}

This group explored a rapidly growing array of training which is seen to contribute to sustainable software. Offerings are diverse, providing training more or less directly relevant to sustinable software. While research institutions  support professional development for research staff, the skills taught which might impact on sustainable software are limited at best, often lacking a clear and coherent development pathway. This growing array of training opportunities could usefully be coordinated, by bringing together those involved in leading relevant initiatives, on a regular basis.

\subsubsection{Fit with related activities}
Two existing venues for discussion of related activities are identified:

\begin{itemize}

\item
\item
WSSSPE Workshop on Sustainable Software for Science: Practice and Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}}

\item
SEHPCCSE International Workshop on Software Engineering for High Performance Computing in Computational Science and Engineering \footnote{\url{http://wssspe.researchcomputing.org.uk/}}

\end{itemize}

\subsubsection{Discussion}

Next steps have been identified to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects.

See Appendix~\ref{sec:appendix_training} for more details about the discussion.

\subsubsection{Plans}

The main plans for the group are to convene discussion to explore bringing together a regular meeting of those involved in leading relavent training projects.

\subsubsection{Landing Page}

The Training working group requests an email be sent to Nick Jones \href{mailto:nick.jones@nesi.org.nz}{(nick.jones@nesi.org.nz)} to find out more about the group's efforts and how to participate.

### 2015-11-10 WSSSPE3/report/wg_main_metrics.tex
fixing a build problem

@@
 \item
-2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics\footnote{\url{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/edit#heading=h.5lah0hp73q99}}
+2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics\footnote{\url{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/}}
 
@@
 \item
-NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics\footnote{\url{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/edit#slide=id.g5e66ec9f2_027
-}}
+NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics\footnote{\url{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/}}
 

### 2015-11-10 WSSSPE3/report/wg_appendix_training.tex
updates to training - spelling, US/UK changes

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
 
+While little training focuses specifically on this outcome \katznote{sustainable software?}, a variety of training activities increase researcher awareness of and engagement with software professionals and software engineering practices.
+Research Software Engineers are being recognized as critical contributors to high quality research; the pathway to acquire and master the relevant skills isn't yet clear; equally those skills required by researchers in general are also not commonly understood nor routinely developed.
+
+The group's discussion explored a rapidly growing array of training that is seen to contribute to sustainable software.
+The offerings are diverse, including: self paced online modules focused around specific tools; single and multiple day training workshops that raise awareness of a tool chain to support collaborative and shared software development within a research workflow; block courses specializing on particular methods, technologies, and applications; academic programs at undergraduate and masters levels; doctoral training programs that in part contain requisite skills training activities.
+
+While some of this training focuses on applying software engineering practices within the context of research, meeting the values and goals of research are less often incorporated as explicit learning outcomes.
+With software (and similarly, data) often the only tangible artifact of a research method or protocol, the dependency between software applications, and the quality of research adds complexity to the learners journey.
+In recognition of the longer term investment required by researchers to integrate such skills into their research practices, many activities are focusing on emotionally engaging researchers and cohorts, to build a sense of shared purpose beyond the obvious goal of technical skill acquisition.
+
+In reviewing current training activities, the group identified a variety of perspectives seen as useful in positioning activities in ways to better communicate why and when best to apply each activity.
+Training can be categorized on a variety of spectra, with content and delivery focused from: programming to research; basic to advanced; technical to emotional; informal to formal; self paced to participative.
+A few attempts have been made to situate a cross section of training activities within such dimensions, creating easier means of communicating the value of any specific opportunity and the pathways across opportunities over time.
+
+Evaluation of training delivery and outcomes is seen as a weakness common to most non academic training activities.
+Opportunities for measuring success in delivering training start simply with collecting a Net Promoter Score, which lets those delivering training know whether attendees are likely to recommend the training to others.
+In looking at the longer term outcomes for the learner, frameworks such as Blooms Taxonomy and Kirkpatrick Evaluations offer possible approaches.
+
+In this latter case of formal evaluation, ownership of evaluation as a component of career development for any researcher appears mostly absent.
+While academic research institutions have professional development centers to support research staff, the skills taught which might impact on sustainable software are limited at best, and lack a clear and coherent development pathway.
+
+Coordination of these training projects will depend on buy in from a broad range of training program and activity leaders, suggesting a key opportunity lies in identifying and bringing together these people on a regular basis.
+
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+
+Software skills are needed by an increasing array of researchers and fields.
+The training arc is not well-defined, with a sometimes baffling array of training opportunities responding to various facets of skill deficit and need.
+Given this current complexity, coordination across training projects would create common frames of reference, communicating and integrating activities to better serve the needs of researchers.
+
+Building this community could lift the maturity of training projects and capabilities, enabling more advanced approaches to address key gaps in evaluation, career development, and a lift in the standard of research practices.
+
+In aiming at these opportunities it will be necessary to find the means to support those involved in leading training activities to allocate time to coordination activities, which will often sit beyond their current scope of responsibility.
+
+These activities are also distributed globally, with no single country or region offering a comprehensive set of capabilities and initiatives.
+Any coordination activity will therefore need to raise the profile of the opportunity gap with relevant research funders and policy makers.
 
@@
 
+The goal of the following next steps is to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects.
+
+\begin{enumerate}
+    \item Hold a virtual meeting by December 2015, to bring together a broader group of interest in this topic, with specific goals to:
+    \begin{enumerate}
+        \item Identify programs with existing activity aimed at integrating across training projects.
+        \item Identify training projects with an interest in participating in coordination efforts.
+        \item Identify funding opportunities to bring together training program and project leaders to identify shared goals for future coordination of activities.
+        \item Agree a communications plan to qualify whether programs, projects, and funders are interested in engaging and committing to ongoing activities.
+    \end{enumerate}
+    \item Review progress within 3 months, to establish next steps, if any.
+\end{enumerate}
+
 \subsection{Plan for Future Organization}
+
+Continue to track progress by posting comments to WSSSPE3 issue.
 
@@
 
+If the group moves from early stage formation into working towards shared goals, expertise will likely be required in pedagogy and training evaluation
+
 \subsection{Key Milestones and Responsible Parties}
+\begin{enumerate}
+    \item During October, Nick Jones and Erin Robinson to draft WSSSPE3 report back. 
+    \item Before December 2015, Nick Jones and Erin Robinson to call a meeting of the broader group, to review key next steps. 
+    \item 1st quarter 2016 - if willing parties identified, draft workshop proposal and identify relevant forum, including future WSSSPE events
+\end{enumerate}
 
 \subsection{Description of Funding Needed}
+
+Workshop/RCN travel funding to bring together key program, project, and funder representatives from across North America, EU, UK, Australasia.
+In addition, funding to support work on better defining the landscape of training activities, the useful perspectives in communicating the value of the varied training projects, and the possible pathways through training activities over time.

### 2015-11-10 WSSSPE3/report/wg_main_training.tex
updates to training - spelling, US/UK changes

@@
-\subsection{Topic} \todo{Training Working Group Discussion}
+\subsection{Training}
 
-\subsubsection{Why it is important}
+%\subsubsection{Why it is important}
 
-This group explored a rapidly growing array of training which is seen to contribute to sustainable software.
-Offerings are diverse, providing training more or less directly relevant to sustinable software.
+This group explored a rapidly growing array of training that is seen to contribute to sustainable software.
+The offerings are diverse, providing training that is more or less directly relevant to sustainable software.
 While research institutions  support professional development for research staff, the skills taught which might impact on sustainable software are limited at best, often lacking a clear and coherent development pathway.
-This growing array of training opportunities could usefully be coordinated, by bringing together those involved in leading relevant initiatives, on a regular basis.
+This growing array of training opportunities could usefully be coordinated by bringing together those involved in leading relevant initiatives on a regular basis.
 
@@
 \item
-\item
-WSSSPE Workshop on Sustainable Software for Science: Practice and Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+WSSSPE Workshop on Sustainable Software for Science: Practice and Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}} \katznote{change to cite}
 
 \item
-SEHPCCSE International Workshop on Software Engineering for High Performance Computing in Computational Science and Engineering \footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+SEHPCCSE International Workshop on Software Engineering for High Performance Computing in Computational Science and Engineering\footnote{\url{http://wssspe.researchcomputing.org.uk/}} \katznote{fix URL and change to cite}
 
@@
 
-The main plans for the group are to convene discussion to explore bringing together a regular meeting of those involved in leading relavent training projects.
+The main plans for the group are to convene discussion to explore bringing together a regular meeting of those involved in leading relevant training projects.
 

### 2015-11-10 WSSSPE3/report/WSSSPE3_review.tex
Section 4:  Line breaks. No change in conetent.

@@
 \item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
-Facilitating Research Training While Supporting Scientific Research}}
-A case study: ``The Cooperative Computing Lab'' at the University of Notre Dame
-is a small group of individuals whose main tasks are collaborating with
-people that have large-scale computing problems, operating various parallel
-computer systems, conducting computer science research, and developing open
-source software. One of the main challenges they face is finding the balance
-between flexibility/training and stability/quality. Their current solution for
-ensuring the latter was to add a software engineer (the presenter) to the
-existing team of faculty and students, who now also serves as ``spring''
-between flexibility and stability.
+Facilitating Research Training While Supporting Scientific Research}} A case
+study: ``The Cooperative Computing Lab'' at the University of Notre Dame is a
+small group of individuals whose main tasks are collaborating with people that
+have large-scale computing problems, operating various parallel computer
+systems, conducting computer science research, and developing open source
+software. One of the main challenges they face is finding the balance between
+flexibility/training and stability/quality. Their current solution for ensuring
+the latter was to add a software engineer (the presenter) to the existing team
+of faculty and students, who now also serves as ``spring'' between flexibility
+and stability.
 
@@
 Rodriguez-Navas and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
-Sustainability}}
-The concept of sustainability has become a topic of interest in the field of
-computing, which is evidenced by the increase in the number of events that
-focus on the topic. Nevertheless, it isn't well understood yet. Penzenstadler
-argued that we often define sustainability too narrowly. Instead,
+Sustainability}} The concept of sustainability has become a topic of interest in
+the field of computing, which is evidenced by the increase in the number of
+events that focus on the topic. Nevertheless, it isn't well understood yet.
+Penzenstadler argued that we often define sustainability too narrowly. Instead,
 sustainability at its heart is a systemic concept and must be viewed from a
@@
 social and technical. She introduced the Karlskrona Manifesto on Software
-Design~\cite{Becker:2014}, which distills knowledge from a broad range of related work on the
-topic of sustainability into a set of (mis-)perceptions and principles. The
-manifesto does not proclaim that there is an easy, one-size-fits-all solution
-around the corner, but rather points out that sustainability is a ``wicked
-problem'' and is often misunderstood. Due to these misperceptions, even though
-sustainability's importance is increasingly recognized, many software systems
-are unsustainable. Even more alarming is that most software systems' broader
-impacts on sustainability are unknown. To change this, the Karlskrona Manifesto
-proposes nine principles and commitments. These commitments are not dogmatic
-laws, but rather a commitment to rethink, to move beyond the silo mentality,
-and analyze in more depth. As such, they do not restrict, but rather open up a
-space for discussion.
+Design~\cite{Becker:2014}, which distills knowledge from a broad range of
+related work on the topic of sustainability into a set of (mis-)perceptions and
+principles. The manifesto does not proclaim that there is an easy,
+one-size-fits-all solution around the corner, but rather points out that
+sustainability is a ``wicked problem'' and is often misunderstood. Due to these
+misperceptions, even though sustainability's importance is increasingly
+recognized, many software systems are unsustainable. Even more alarming is that
+most software systems' broader impacts on sustainability are unknown. To change
+this, the Karlskrona Manifesto proposes nine principles and commitments. These
+commitments are not dogmatic laws, but rather a commitment to rethink, to move
+beyond the silo mentality, and analyze in more depth. As such, they do not
+restrict, but rather open up a space for discussion.
 
 \item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
-and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces to
-Improve Productivity of Legacy Tools}}
-Abani Patra presented an example how the community using Titan2D, a Geoflow
-Simulation Software, increased the productivity of their tools by improving
-both code and data layout. The main obstacles in this change were the non-existence
-of a common version control system for the source code, coupled with multiple
-versions of the same code base, the fixed format of input files, that many
-input values used to be set as compilation flags, and that the internal data
-layout was not suitable for modern technologies (vectorization, accelerators).
-The approach of the Titan2D developers included reinforcing the code structure
-using a multiple layers of Python and C++, and a redesign of the data layout to
-be better suitable for modern CPUs and accelerators.
+and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
+to Improve Productivity of Legacy Tools}} Abani Patra presented an example how
+the community using Titan2D, a Geoflow Simulation Software, increased the
+productivity of their tools by improving both code and data layout. The main
+obstacles in this change were the non-existence of a common version control
+system for the source code, coupled with multiple versions of the same code
+base, the fixed format of input files, that many input values used to be set as
+compilation flags, and that the internal data layout was not suitable for modern
+technologies (vectorization, accelerators). The approach of the Titan2D
+developers included reinforcing the code structure using a multiple layers of
+Python and C++, and a redesign of the data layout to be better suitable for
+modern CPUs and accelerators.
 
@@
 \item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
-through Best Practices in Open Source Software: Computational Infrastructure
-for Geodynamics}}
-Lorraine Hwang presented experiences with CIG: the Computational Infrastructure
-for Geodynamics, a community software with a worldwide user base. Like others,
-their main goals include high usability, sustainability, and reproducibility.
-In order to achieve this, communication channels have been developed, such as
-mailing lists, wikis, workshops, hackathons, tutorials, and webinars. In order
-to contribute to the infrastructure, codes have to adhere to specified minimum
-standards, but ideals are also given to encourage reaching the aforementioned
-goals. This includes, e.g., the use of version control, certain coding styles,
-the presence and nature of code tests and documentation and certain user
-workflows.
+through Best Practices in Open Source Software: Computational Infrastructure for
+Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
+Infrastructure for Geodynamics, a community software with a worldwide user base.
+Like others, their main goals include high usability, sustainability, and
+reproducibility. In order to achieve this, communication channels have been
+developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
+webinars. In order to contribute to the infrastructure, codes have to adhere to
+specified minimum standards, but ideals are also given to encourage reaching the
+aforementioned goals. This includes, e.g., the use of version control, certain
+coding styles, the presence and nature of code tests and documentation and
+certain user workflows.
 
 \item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications in the
-Computational Infrastructure for Geodynamics}}
-In a second talk, Lorraine Hwang mentioned various ways to cite
-efforts within the framework, including science papers, code papers, user
-manuals, and the website. An analysis of the resulting
-citations showed that $80\%$ of papers that use CIG codes mention the code name, and about
-the same number includes a citation. Only about $20\%$ acknowledge CIG. Within
-the same sample of papers, about one fifth use an URL to cite codes
-(including non-CIG codes), and only about one eighth specify the version used,
-showing that compared to other codes, CIG seems to be much better cited. This,
-in part, is attributed to suggested citation forms communicated to the community,
-and runtime tools that help collecting a relevant citation list automatically.
+Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications
+in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
+Hwang mentioned various ways to cite efforts within the framework, including
+science papers, code papers, user manuals, and the website. An analysis of the
+resulting citations showed that $80\%$ of papers that use CIG codes mention the
+code name, and about the same number includes a citation. Only about $20\%$
+acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
+cite codes (including non-CIG codes), and only about one eighth specify the
+version used, showing that compared to other codes, CIG seems to be much better
+cited. This, in part, is attributed to suggested citation forms communicated to
+the community, and runtime tools that help collecting a relevant citation list
+automatically.
 
 \item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
-Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for Open
-Science (DASPOS)}}
-DASPOS is an NSF-funded multi-disciplinary effort, located at Notre
-Dame and Chicago, that links the HEP effort to Biology, Astrophysics, Digital
-Curation, and other disciplines. It includes physicists, digital librarians, as
-well as computer scientists, and aims to achieve some commonality across
-disciplines. Examples are meta-data descriptions of archived data,
-computational descriptions, descriptions of how data was processed, but also
-includes questions such as whether computation replication can be automated, and
-what the impact of access policies on the preservation infrastructure is. One
-of the products of this effort is a suite of tools that deals with this
-preservation, and the questions was posed whether that software itself is
-sustainable. Points that were brought up included the need for a user community
-depending on a given software, and the need to provide added value for its
-users. An important method to achieve this was to work with the user
+Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for
+Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
+located at Notre Dame and Chicago, that links the HEP effort to Biology,
+Astrophysics, Digital Curation, and other disciplines. It includes physicists,
+digital librarians, as well as computer scientists, and aims to achieve some
+commonality across disciplines. Examples are meta-data descriptions of archived
+data, computational descriptions, descriptions of how data was processed, but
+also includes questions such as whether computation replication can be
+automated, and what the impact of access policies on the preservation
+infrastructure is. One of the products of this effort is a suite of tools that
+deals with this preservation, and the questions was posed whether that software
+itself is sustainable. Points that were brought up included the need for a user
+community depending on a given software, and the need to provide added value for
+its users. An important method to achieve this was to work with the user
 community from the start, and to budget that way. For the specific example of
@@
 Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods and
-Susheel Varma: \textit{An update from UK Research Software Engineers}}
-James Hetherington started by listing some of the problems research software
-faces, which include low levels of reuse and poor standard of verification. For
-a long time, technical solutions to such technical problems were focus of the
-eResearch community, including research software distributions, grids,
-middleware and workflows. Some limited adoption can be seen today in research
-communities, but the main problems have not been solved to a sufficient level.
-James Hetherington hypothesizes that instead of technical solutions, social
-innovation is needed: a new role in the academic system focused on research
-software that combines the best parts of a craftsperson and a scholar. However,
-social innovation in centuries-old institutions is hard. Alternatives to such a
-new role would have to include rewards for good research software, recognition
-of software as academic output, and rejecting submissions based on
-irreproducible computational results. Some advantages of research software
-engineering (RSE) groups include the possibility of training in reproducible
-computational research, providing collaborations for for researchers who don't
-want to be programmers, and synergies with research computing platforms. The
-success of such a group would be measured by the output and quality of the
-research software, and could be part of Research Computing or faculty. Members
-would not be independent researchers, but would have to have a research
-background. An attempt to form a community of RSE groups within the UK has been
-underway for several years, including funding from the UK Research Council. An
-open question is whether this approach can be adopted also in other countries,
-including the USA.
+Susheel Varma: \textit{An update from UK Research Software Engineers}} James
+Hetherington started by listing some of the problems research software faces,
+which include low levels of reuse and poor standard of verification. For a long
+time, technical solutions to such technical problems were focus of the eResearch
+community, including research software distributions, grids, middleware and
+workflows. Some limited adoption can be seen today in research communities, but
+the main problems have not been solved to a sufficient level. James Hetherington
+hypothesizes that instead of technical solutions, social innovation is needed: a
+new role in the academic system focused on research software that combines the
+best parts of a craftsperson and a scholar. However, social innovation in
+centuries-old institutions is hard. Alternatives to such a new role would have
+to include rewards for good research software, recognition of software as
+academic output, and rejecting submissions based on irreproducible computational
+results. Some advantages of research software engineering (RSE) groups include
+the possibility of training in reproducible computational research, providing
+collaborations for for researchers who don't want to be programmers, and
+synergies with research computing platforms. The success of such a group would
+be measured by the output and quality of the research software, and could be
+part of Research Computing or faculty. Members would not be independent
+researchers, but would have to have a research background. An attempt to form a
+community of RSE groups within the UK has been underway for several years,
+including funding from the UK Research Council. An open question is whether this
+approach can be adopted also in other countries, including the USA.
 
-\item \textbf{Dan Gunter, Sarah Poon and Lavanya Ramakrishnan: \textit{Bringing the
-User into Building Sustainable Software for Science}}
-Dan Gunter's main question was, "What is needed to develop sustainable software?".
-Beyond the usual suspects of funding, great developers, good
-design, and software engineering practices, however, Dan placed the
-users. He explained this using a traditional software development model starting
-from gathering requirements, and reaching release through a design and
-development and testing phase. The main deficiency that with this approach was
-pointed out to be the too-late interaction with users. Instead, an alternative
-approach was proposed that, at first, skips the development phase and repeatedly
-goes through requirement gathering, design and user interaction/learning
-phases, and only eventually starts development once an agreement is reached,
-leading to an increased user satisfaction, higher adoption, and eventually to
-sustained software.
+\item \textbf{Dan Gunter, Sarah Poon and Lavanya Ramakrishnan: \textit{Bringing
+the User into Building Sustainable Software for Science}} Dan Gunter's main
+question was, "What is needed to develop sustainable software?". Beyond the
+usual suspects of funding, great developers, good design, and software
+engineering practices, however, Dan placed the users. He explained this using a
+traditional software development model starting from gathering requirements, and
+reaching release through a design and development and testing phase. The main
+deficiency that with this approach was pointed out to be the too-late
+interaction with users. Instead, an alternative approach was proposed that, at
+first, skips the development phase and repeatedly goes through requirement
+gathering, design and user interaction/learning phases, and only eventually
+starts development once an agreement is reached, leading to an increased user
+satisfaction, higher adoption, and eventually to sustained software.
 
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and
-Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
-Biology: Lessons Learned on the KBase Project}}
-In this talk, Dan Gunter presented experiences and lessons learned as part of
-KBase, an open software and data platform for addressing the grand challenge of
-systems biology: predicting and designing biological function. KBase is a
-unified system that integrates data and analytical tools for comparative
-functional genomics of microbes, plants, and their communities. However,
-it is also a collaborative environment for sharing methods and results, and
-placing those results in the context of knowledge in the field. Being a large,
-multi-institutional project, one of the big challenges is to agree on standard to
-enable a single, maintainable system. Working in isolation does not work (anymore)
-within this field, and they also didn't have standards for software engineering.
-This is contrasted to computer science research, where these approaches 
-shorten design cycles, leading to more time for highly rewarded activities like
-publishing, performance studies, graduating students, or protecting ideas
-before publication. Sustainability is not of big concern there, compared to
-these. \katznote{prev sentence is unclear} Instead of this more traditional approach,
-KBase uses the following for large, distributed teams. After picking projects and team members,
-4-5 teams work on projects about two weeks before a
-1-week evaluation by the executive committee, and after which new teams and
-projects might be chosen to restart the ``agile'' development cycle. This
-process is intentionally open and documented.
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and Sergei
+Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
+Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
+presented experiences and lessons learned as part of KBase, an open software and
+data platform for addressing the grand challenge of systems biology: predicting
+and designing biological function. KBase is a unified system that integrates
+data and analytical tools for comparative functional genomics of microbes,
+plants, and their communities. However, it is also a collaborative environment
+for sharing methods and results, and placing those results in the context of
+knowledge in the field. Being a large, multi-institutional project, one of the
+big challenges is to agree on standard to enable a single, maintainable system.
+Working in isolation does not work (anymore) within this field, and they also
+didn't have standards for software engineering. This is contrasted to computer
+science research, where these approaches shorten design cycles, leading to more
+time for highly rewarded activities like publishing, performance studies,
+graduating students, or protecting ideas before publication. Sustainability is
+not of big concern there, compared to these. \katznote{prev sentence is unclear}
+Instead of this more traditional approach, KBase uses the following for large,
+distributed teams. After picking projects and team members, 4-5 teams work on
+projects about two weeks before a 1-week evaluation by the executive committee,
+and after which new teams and projects might be chosen to restart the ``agile''
+development cycle. This process is intentionally open and documented.
 
@@
 Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
-Scientists in Best Practices of Software Sharing}}
-Erin Robinson presented an approach to overcome some hurdles in scientific
-publishing: the disconnect between experimental data, research software, and
-publications. A current effort in the Geosciences is the ``Geoscience Paper of
-the Future,'' which includes four elements. First, it forms a modern paper
-including text, data, and pointers to supplementary material. Second, it is
-reproducible, including data processing, workflow, and visualization tools. 
-Third, it is part of open science, which includes being publicly
-available under open licenses, and providing meta-data. Last, but not least, it
-uses digital scholarship elements like persistent identifiers for software,
-data, and authors, and it cites both data and software. OntoSoft is an effort
-to help this effort, providing software stewardship for the Geosciences. As
-part of this initiative, a special issue of a journal in a Geoscience areas is
-planned that only includes Geoscience papers of the future, with submissions
-open until the end of 2015. In addition, training sessions are provided to
-geoscientists to learn best practices in software and data sharing, provenance
-documentation, and scholarly publication.
+Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
+approach to overcome some hurdles in scientific publishing: the disconnect
+between experimental data, research software, and publications. A current effort
+in the Geosciences is the ``Geoscience Paper of the Future,'' which includes
+four elements. First, it forms a modern paper including text, data, and pointers
+to supplementary material. Second, it is reproducible, including data
+processing, workflow, and visualization tools. Third, it is part of open
+science, which includes being publicly available under open licenses, and
+providing meta-data. Last, but not least, it uses digital scholarship elements
+like persistent identifiers for software, data, and authors, and it cites both
+data and software. OntoSoft is an effort to help this effort, providing software
+stewardship for the Geosciences. As part of this initiative, a special issue of
+a journal in a Geoscience areas is planned that only includes Geoscience papers
+of the future, with submissions open until the end of 2015. In addition,
+training sessions are provided to geoscientists to learn best practices in
+software and data sharing, provenance documentation, and scholarly publication.
 
-\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software Accreditation
-Framework}}
-Neil Chue Hong presented a proposal to build a scientific software
-accreditation framework. One of the aims of such a framework would be to
-measure how ``good'' a given piece of software is, and to evaluate how this can be
-effectively measured in the first place. This can be compared to the effort of
-the standardized, easy to read, and understandable nutritional labeling of goods,
-which only contain a small set of categories. However, such a framework for
-software would be more difficult due to different
-existing community norms. The challenges such a framework faces include that
-many measurements are subjective, many metrics are too costly, and that
+\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
+Accreditation Framework}} Neil Chue Hong presented a proposal to build a
+scientific software accreditation framework. One of the aims of such a framework
+would be to measure how ``good'' a given piece of software is, and to evaluate
+how this can be effectively measured in the first place. This can be compared to
+the effort of the standardized, easy to read, and understandable nutritional
+labeling of goods, which only contain a small set of categories. However, such a
+framework for software would be more difficult due to different existing
+community norms. The challenges such a framework faces include that many
+measurements are subjective, many metrics are too costly, and that
 self-assessment needs to be encouraged. Possible categories would include
@@
 
-\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering Support for
-Sustainable Scientific Software}}
-Jeff argued that for scientific software to be truly sustainable, there is a need
-for developers to use appropriate software engineering practices. His experience
-interacting with scientific teams indicates that choosing and tailoring these
-practices is not a trivial exercise. There is a general culture clash between
-software engineering and science that hinders our ability to communicate and
-choose appropriate methods. In addition, many experience scientific software
-developers appear to be unaware of software engineering practices that may be
-beneficial to them. The most appropriate software engineering practices are
-those that are lightweight, properly tailored, and focus on the key software
-development problems faced by scientists. In order to increase the use of
-software engineering in science, we need more documented success stories. These
-successes need to be socialized within the scientific community through
-workshops like the Software Engineering for Science workshop
+\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
+Support for Sustainable Scientific Software}} Jeff argued that for scientific
+software to be truly sustainable, there is a need for developers to use
+appropriate software engineering practices. His experience interacting with
+scientific teams indicates that choosing and tailoring these practices is not a
+trivial exercise. There is a general culture clash between software engineering
+and science that hinders our ability to communicate and choose appropriate
+methods. In addition, many experience scientific software developers appear to
+be unaware of software engineering practices that may be beneficial to them. The
+most appropriate software engineering practices are those that are lightweight,
+properly tailored, and focus on the key software development problems faced by
+scientists. In order to increase the use of software engineering in science, we
+need more documented success stories. These successes need to be socialized
+within the scientific community through workshops like the Software Engineering
+for Science workshop
 series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in
-\textit{Computing in Science and Engineering} magazine.
+and the new Software Engineering track in \textit{Computing in Science and
+Engineering} magazine.
 
@@
 This talk highlighted an attempt to solve the common problem for open source
-development: it is difficult to collect information about how many people
-use particular software, how often, which version, which parts of the software, or on
-which operating system. Current solutions include surveys, but these have high
-uncertainties. A different approach is based on automatic ``call-backs'' that
-collect these information at runtime and send it to a central place for
-analysis. Problems with this approach include obtaining agreement from the
-user, legal issues, increased maintenance (servers), ethical questions, but
-also the lack of a common infrastructure. Some of these problems are of a social
-nature and have to be solved as such, but the last problem (a
-missing common infrastructure) is attacked by the sempervirens project~\cite{sempervirens}, which
-is developing common APIs and a library implementation for common, repeating tasks
-such as obtaining user consent. Results are uploaded not directly to project
-servers, but to neutral third parties that only publish aggregated statistics
-to projects.
+development: it is difficult to collect information about how many people use
+particular software, how often, which version, which parts of the software, or
+on which operating system. Current solutions include surveys, but these have
+high uncertainties. A different approach is based on automatic ``call-backs''
+that collect these information at runtime and send it to a central place for
+analysis. Problems with this approach include obtaining agreement from the user,
+legal issues, increased maintenance (servers), ethical questions, but also the
+lack of a common infrastructure. Some of these problems are of a social nature
+and have to be solved as such, but the last problem (a missing common
+infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
+which is developing common APIs and a library implementation for common,
+repeating tasks such as obtaining user consent. Results are uploaded not
+directly to project servers, but to neutral third parties that only publish
+aggregated statistics to projects.
 
 \item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}}
-Alice Allen described the Astrophysics Source Code Library (ascl.net), an
-increasingly used way to obtain a unique ID for astrophysics software that is
-indexed by indexing services and can be cited. ASCL offers clones of existing
-infrastructure, provides server space and computing resources, shares
-innovations, and permits moves elsewhere. Users provide a
-domain name, then control and configure their site and use it as intended, gather
-their codes as they wish, share innovations, and protect the provided
+included)}} Alice Allen described the Astrophysics Source Code Library
+(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
+software that is indexed by indexing services and can be cited. ASCL offers
+clones of existing infrastructure, provides server space and computing
+resources, shares innovations, and permits moves elsewhere. Users provide a
+domain name, then control and configure their site and use it as intended,
+gather their codes as they wish, share innovations, and protect the provided
 computing environment.
@@
 Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
-Bo Zhang: \textit{Scientific Software Success: Developing Metrics While Developing
-Community}}
-The effort behind this talk by Ray Idaszak started from a break-out group at an NSF SI2
-workshop in 2015, and centers around creating a framework for creation of
-metrics for scientific software. This framework would improve both the metrics,
-and the software it evaluates, and is trying to build a community around this idea.
-With especially the latter idea in mind, a software ``peer review group'' would
-be created, representing stakeholders who will self-review software created by
-their respective communities, and will concurrently develop metrics. The whole
-project should be community-governed, without a single institution overseeing
-the activity or infrastructure, with the hope to evolving community-generated
-and adopted standards. The generation of metrics would be tied to the actual
-evaluation of software, creating an incentive by improving the evaluated software
-itself during this process. The framework code would provide infrastructure for the
-creation of metrics and evaluation, and forums for
-generation of software success metrics.  It would also support
-code reviews of the evaluated software. An open
-question is whether it is possible to fit the resulting metrics in a common
-template. So far, this is still in a design phase, with a white paper at the
-CSESSP workshop 2015 and this talk, but WSSSPE workshop is seen as a forum for
-the community to assemble and act, and is planned to be used also in the
-future to build this community and framework.
+Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
+Developing Community}} The effort behind this talk by Ray Idaszak started from a
+break-out group at an NSF SI2 workshop in 2015, and centers around creating a
+framework for creation of metrics for scientific software. This framework would
+improve both the metrics, and the software it evaluates, and is trying to build
+a community around this idea. With especially the latter idea in mind, a
+software ``peer review group'' would be created, representing stakeholders who
+will self-review software created by their respective communities, and will
+concurrently develop metrics. The whole project should be community-governed,
+without a single institution overseeing the activity or infrastructure, with the
+hope to evolving community-generated and adopted standards. The generation of
+metrics would be tied to the actual evaluation of software, creating an
+incentive by improving the evaluated software itself during this process. The
+framework code would provide infrastructure for the creation of metrics and
+evaluation, and forums for generation of software success metrics. It would also
+support code reviews of the evaluated software. An open question is whether it
+is possible to fit the resulting metrics in a common template. So far, this is
+still in a design phase, with a white paper at the CSESSP workshop 2015 and this
+talk, but WSSSPE workshop is seen as a forum for the community to assemble and
+act, and is planned to be used also in the future to build this community and
+framework.
 

### 2015-11-10 WSSSPE3/report/WSSSPE3_review.tex
Section 1-3:  Line breaks. No change in conetent.
Section 4: Fix open quotes.

@@
 travel of potential participants who would not otherwise be able to attend,
-early-stage researchers, and those from underrepresented groups. These funds allowed
-16 additional participants to attend.
+early-stage researchers, and those from underrepresented groups. These funds
+allowed 16 additional participants to attend.
 
@@
 attendees to quickly highlight an important issue or a potential solution. The
-report also gives summaries of action plans proposed by the working
-groups (\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}).
-Lists of the organizing committee (Appendix~\ref{sec:orgcom}), the registered
-attendees (Appendix~\ref{sec:attendees}), and the travel award recipients
+report also gives summaries of action plans proposed by the working groups
+(\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}). Lists of the
+organizing committee (Appendix~\ref{sec:orgcom}), the registered attendees
+(Appendix~\ref{sec:attendees}), and the travel award recipients
 (Appendix~\ref{sec:awardees}) are provided. Finally, the report includes longer
-descriptions of the activities that
-occurred in each of the working groups that made substantial progress 
+descriptions of the activities that occurred in each of the working groups that
+made substantial progress
 (Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).
@@
 Turk reminded the audience that the volunteer model means unpaid labor. On this
-note, he recommended Ashe Dryden's blog post on the
-ethics of unpaid labor and the open source software
+note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
+the open source software
 community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
-Often times, a person
-funded to work full time on a scientific project can spend a small amount of
-time for working on a piece of software necessary for that project. However, all
-researchers' ability to participate in that volunteer community are not always
-the same and may not always be aligned with their research projects. From Turk's
-experience, we cannot always rely on unpaid labor and volunteer time to sustain
-a piece of software---this came down to the notions of the top-down and the
-bottom-up approaches, i.e., the funded versus the grassroots. However, Turk
-pointed out that bottom-up, volunteer-driven projects can be just as large
-as a top-down software development project.
+Often times, a person funded to work full time on a scientific project can spend
+a small amount of time for working on a piece of software necessary for that
+project. However, all researchers' ability to participate in that volunteer
+community are not always the same and may not always be aligned with their
+research projects. From Turk's experience, we cannot always rely on unpaid labor
+and volunteer time to sustain a piece of software---this came down to the
+notions of the top-down and the bottom-up approaches, i.e., the funded versus
+the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
+projects can be just as large as a top-down software development project.
 
@@
 all have an invested stake in sustaining scientific software. Hence, having
-``sustained'' projects can suffocate new projects, so we need to make sure
-we don't cause novel ideas and packages to suffer at the hands of the status quo.
-%This is a straw man concept, but it's something to look out for, since organizations will act in their own interest.
+``sustained'' projects can suffocate new projects, so we need to make sure we
+don't cause novel ideas and packages to suffer at the hands of the status quo.
+%This is a straw man concept, but it's something to look out for, since
+organizations will act in their own interest.
 
-Turk talked about possible reasons why we want to sustain scientific
-software: devotion to science and interests in pursuing the next stage of
-research; fun and creative thrill in writing codes and papers; usefulness with
-measurable impacts, for example, LINPACK and HDF groups providing of data
-storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
-presented his wish-list of questions to be answered in the future: 
+Turk talked about possible reasons why we want to sustain scientific software:
+devotion to science and interests in pursuing the next stage of research; fun
+and creative thrill in writing codes and papers; usefulness with measurable
+impacts, for example, LINPACK and HDF groups providing of data storage to
+satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
+his wish-list of questions to be answered in the future:
 %
@@
 
-\item How can productized software balance its future versus its past, or the new
-needs of the customers versus the existing needs of the development community?
+\item How can productized software balance its future versus its past, or the
+new needs of the customers versus the existing needs of the development
+community?
 
@@
 the User into Building Sustainable Software for Science}} Dan Gunter's main
-question was, "What is needed to develop sustainable software?". Beyond the
+question was, ``What is needed to develop sustainable software?''. Beyond the
 usual suspects of funding, great developers, good design, and software

### 2015-11-10 WSSSPE3/report/WSSSPE3_review.tex
Section 4: Touch up.

@@
 Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
-Rodriguez-Navas and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
+Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
 Sustainability}} The concept of sustainability has become a topic of interest in
@@
 
-\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones
+\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones,
 and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
@@
 
-\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins and Kaitlin
+\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
 Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
@@
 \item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith and Laura Soito: \textit{Software Attribution for Geoscience Applications
+Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
 in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
@@
 \item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
-Douglas Thain and Charles Vardeman: \textit{Data And Software Preservation for
+Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
 Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
@@
 \item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
-Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods and
+Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
 Susheel Varma: \textit{An update from UK Research Software Engineers}} James
@@
 
-\item \textbf{Dan Gunter, Sarah Poon and Lavanya Ramakrishnan: \textit{Bringing
+\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
 the User into Building Sustainable Software for Science}} Dan Gunter's main
@@
 
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham and Sergei
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
 Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
@@
 
-\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson and Karan
+\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
 Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex

@@
 devastating effect on an academic
-publication record resulting from time in industry; such postings often develop exactly the skills that research programmers
-need, yet returns to university positions following an industry role are penalized by the current structures. Retention of
-senior programmers is hard, because these people are highly in demand by the economy. However, people who have a PhD in
-science and enter industry, may desire to return for diverse reasons, and should be welcomed back.
+publication record resulting from time in industry; such postings often develop exactly the skills that research software
+engineers need, yet returns to university positions following an industry role are penalized by the current structures.
+Retention of senior developers is hard, because these people are highly in demand by the economy. However, people who have a
+PhD in science and enter industry, may desire to return for diverse reasons, and should be welcomed back.
 
@@
 sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
-developing dedicated research programming roles within the University, and finding stable funding for those individuals is
-the most promising mechanism for creating a stable software development staff.
+developing dedicated research software engineering roles within the University, and finding stable funding for those individuals is the most promising mechanism for creating a stable software development staff.
 
@@
 be required in order to make the case to the University for continued funding.
-Research programmers will not be measured by publications, we hope, but by other measures.
+Research software engineers will not be measured by publications, we hope, but by other measures.
 (what measures?) Middle-author publications are common for RSEs.
 Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
-(these last two statements seem contradictory)
+(these last two statements seem contradictory, not sure which message is to be used?)
 
@@
 
-It is hard for an individual PI in a university or college to support dedicated research programming resources, as the need
-for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is
-necessary to aggregate this work across multiple research groups.
+It is hard for an individual PI in a university or college to support dedicated research software engineering resources, as
+the need for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is necessary to aggregate this work across multiple research groups.
 
-One solution is to fund dedicated programming roles for major research software projects at national laboratories
+One solution is to fund dedicated software engineering roles for major research software projects at national laboratories
 or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
@@
 fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
-with PIs in the normal grant process, with PIs choosing to fund some amount of research programming effort through grants in
+with PIs in the normal grant process, with PIs choosing to fund some amount of research software engineering effort through grants in
 the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
@@
 We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
-UK, a collective effort to identify this community, with individuals clearly stating ``I'm a research software engineer'',
-has been important to the campaign. It will be useful to the global effort to similarly identify emerging research software
-organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a
-long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution.
-There remains the problem of how to "sell" the value of this investment to investigators in the university will be problem left to the individual universities, once the initial investment is made.
-This is an issue best addressed by th eindividual organizations that embark on the plan.
+UK, a collective effort to identify the research software engineering community, with individuals clearly stating "I'm a research software engineer", has been important to the campaign.
+It will be useful to the global effort to similarly identify emerging research software organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution.
+There remains the problem of how to "sell" the value of this investment to investigators within the university.
+This is an issue best addressed by the individual organizations that embark on the plan.
 

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex
Update wg_main_funding_spec_expert.tex

@@
 
-One strategy to promote stability is implementing a mechansim for developers to obtain academic
+One strategy to promote stability is implementing a mechanism for developers to obtain academic
 credit for software development work. With such a mechanism in place, traditional academic funding
@@
 . This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
-traditionally supported university cyberinfrastructure, libraryorganizations, or research offices. Alternatively, these
+traditionally supported university cyberinfrastructure, library organizations, or research offices. Alternatively, these
 groups could be organizationally closer to research groups, sitting within academic departments. The most effective model

### 2015-11-10 WSSSPE3/report/wg_appendix_funding_spec_expert.tex

@@
 
-In this respect, information gatheirng via a survey subsequent analysis could be very useful.
-We would need to assemble a list of targetted individuals (what position and rank is
-likely to know and care enough to respond?) Perhaps the Science Gateway Insitute has already acquire
+In this respect, information gathering via a survey subsequent analysis could be very useful.
+We would need to assemble a list of targeted individuals (what position and rank is
+likely to know and care enough to respond?) Perhaps the Science Gateway Institute has already acquire
 information that could be helpful to advance this issue, and/or craft the proper survey and target
@@
 support to nucleate the creation of research software groups. As noted above, Sloan has already
-initiated one such program, and collaboration with Slaon or at least study of their methods and 
-success or failure could be extremely useful in approaching Universities and other insitutions 
+initiated one such program, and collaboration with Sloan or at least study of their methods and 
+success or failure could be extremely useful in approaching Universities and other institutions 
 in funding this development track. It seems clear that if the value proposition can be made to
-University adminstrators, this track could flourish with buy-in at the adminstrative level.
+University administrators, this track could flourish with buy-in at the administrative level.

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex

@@
 with PIs in the normal grant process, with PIs choosing to fund some amount of research software engineering effort through grants in
-the usual way. However, to bootstrap such a function to a level where it has sufficent reputation and client base to be self
+the usual way. However, to bootstrap such a function to a level where it has sufficient reputation and client base to be self
 -sustaining will generally require seed investment.

### 2015-11-10 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
small formatting/quotes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Funding Specialist Expertise Group Discussion}
+\section{Funding Research Programmer Expertise Group Discussion}
 \label{sec:appendix_funding_spec_expert}
@@
 One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.
-And this brings up the flip side question, “Do scientists really want to be writing code?”
+And this brings up the flip side question, ``Do scientists really want to be writing code?''
 
@@
 service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
-“permanent” staff.  When PIs request allocations on XSEDE resources, there is a 
+``permanent'' staff.  When PIs request allocations on XSEDE resources, there is a 
 finite pool of developer time that can be awarded, typically for one year only, 
@@
 expert in the characteristics of their specific resource, and can very quickly assess the
-scientists needs, and what it will take to implement software that meets the user’s needs. 
+scientists needs, and what it will take to implement software that meets the user's needs. 
 However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
@@
 
-The Canarie group (Canada) accepts proposals for providing services to broad communities, 
+The CANARIE group (Canada) accepts proposals for providing services to broad communities, 
 integrate people who are doing things that are complementary, the goal is to make the available 
@@
 There can be problems communicating across cultural barriers, with domain scientists seeing developers 
-as “other”. Collaboration, and tools to fund, encourage or motivate collaboration are
+as ``other''. Collaboration, and tools to fund, encourage or motivate collaboration are
 extremely important.
@@
 funding, just as DNA sequencing is for core facilities, if it becomes apparent
-that this gives competitive advantage to the University’s research effort.
+that this gives competitive advantage to the University's research effort.
 

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex
small formatting/quotes

@@
 
-\katznote{I don't think this section title is exactly right - are there some other options?
+%\katznote{I don't think this section title is exactly right - are there some other options?
 Also, the title of Appendix~\ref{sec:appendix_funding_spec_expert} should match, whatever is chosen.
@@
 universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
-capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group
-. This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
+capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group.
+This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
 traditionally supported university cyberinfrastructure, library organizations, or research offices. Alternatively, these
@@
 
-The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSE's who do not identify themselves as such at this time.
+The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSEs who do not identify themselves as such at this time.
 We will collect information as to the organizational models under which these groups function, and how they are funded.

### 2015-11-10 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
a few more small changes

@@
 
-In addition to the points noted in the main discussion~\ref{RSE}, we also
+In addition to the points noted in the main discussion (\S\ref{RSE}), we also
 discussed the following:
@@
 call, which really permits explicit investment in software as an infrastructure,
-is so oversubscribed as to have a FOUR PERCENT success rate; the demand is clear!
+is so oversubscribed as to have a 4\% success rate; the demand is clear!
 
@@
 
-RSEs provide a useful contribution to their Universities' *teaching* missions,
+RSEs provide a useful contribution to their Universities' \textbf{teaching} missions,
 as well as for research, as they are well placed to deliver the research programming
@@
 
-The UKRSE community will provide initial facilities to host this list, and
+The UK RSE community will provide initial facilities to host this list, and
 continue to work to spread the initiative, but local leadership in the US

### 2015-11-10 WSSSPE3/report/wg_main_funding_spec_expert.tex
a few more small changes

@@
 Research software engineers will not be measured by publications, we hope, but by other measures.
-(what measures?) Middle-author publications are common for RSEs.
+\todo{what measures?} Middle-author publications are common for RSEs.
 Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
@@
 We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
-UK, a collective effort to identify the research software engineering community, with individuals clearly stating "I'm a research software engineer", has been important to the campaign.
+UK, a collective effort to identify the research software engineering community, with individuals clearly stating ``I'm a research software engineer,'' has been important to the campaign.
 It will be useful to the global effort to similarly identify emerging research software organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution.

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
Sections 1-4: minor enhancements.

@@
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
-focused workshop organized in July 2014 jointly with the SciPy
+focused workshop organized jointly with the SciPy
 conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
@@
 WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
-focused workshop organized in July 2015 again jointly with
+focused workshop organized again jointly with
 SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
@@
 software at all levels. Hence it is critical to address challenges related to
-the development, deployment, maintenance, and overall sustainability of reusable
+development, deployment, maintenance, and overall sustainability of reusable
 software as well as education around software practices. These challenges can be
-technological, policy based, organizational, and educational, and are of
+technological, policy based, organizational, and educational; and are of
 interest to developers (the software community), users (science disciplines),
@@
 and innovation policy, and social science communities). The WSSSPE1 workshop
-engaged the broad scientific community to identify challenges and best practices
+engaged a broad scientific community to identify challenges and best practices
 in areas of interest to creating sustainable scientific software. WSSSPE2
 invited the community to propose and discuss specific mechanisms to move towards
-an imagined future practice for software development and usage in science and
-engineering, but WSSSPE2 didn't have a good way to enact those mechanisms, or to
+an imagined future for software development and usage in science and
+engineering. But WSSSPE2 didn't have a good way to enact those mechanisms, or to
 encourage the attendees to follow through on their intentions.
 
-The workshop included multiple mechanisms for participation and encouraged team
+The WSSSPE3 workshop included multiple mechanisms for participation and encouraged team
 building around solutions. It strongly encouraged participation of early-career
@@
 
-This report is based on collaborative notes taken with during the workshop,
+This report is based on collaborative notes taken during the workshop,
 which were linked from the GitHub issues that represented the potential and
@@
 Overall, the report discusses the organization work done before the workshop
-(\S\ref{sec:preworkshop}); the keynotes (\S\ref{sec:keynote}); a series of
-lightning talks (\S\ref{sec:lightning}), intended to give an opportunity for
-attendees to quickly highlight an important issue or a potential solution. The
+(\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
+lightning talks (\S\ref{sec:lightning}). The
 report also gives summaries of action plans proposed by the working groups
@@
 (Appendix~\ref{sec:attendees}), and the travel award recipients
-(Appendix~\ref{sec:awardees}) are provided. Finally, the report includes longer
+(Appendix~\ref{sec:awardees}) are compiled. Finally, the report includes longer
 descriptions of the activities that occurred in each of the working groups that
@@
 
-The first call for participation requested lightning talks, where authors could
+The first call for participation requested lightning talks, where each author could
 make a brief statement about work that either had been done or was needed, with
 the goal of contributing to the discussion of one or more working groups. There
-was 24 lightning talks submitted, and, after a peer-review process, 16 of these
+were~24 lightning talks submitted, and, after a peer-review process, 16 of those
 were accepted, as discussed further in Section~\ref{sec:lightning}.
@@
 WSSSPE2, and requested additional suggestions. The combination of existing and
-new topics led to the following 18 potential topics that were advertised in the
-call for participation:
+new topics led to the following 18 potential topics that were advertised in the subsequent
+calls for participation:
 
@@
 \item Create a set of guidance giving examples of specific metrics for the
-success of scientific software in use, why they were chosen, what they are
+success of scientific software in use, why they are chosen, what they are
 useful to measure, and any challenges/pitfalls; then publish this as a white
@@
 (building on existing work), and case studies of alternative publishing
-mechanisms that have been shown to improve software discoverability/reuse e.g.,
+mechanisms that have been shown to improve software discoverability/reuse, e.g.,
 popular blogs/websites
 \item Determining what journals that publish software paper should provide to
-their reviewers (e.g., guidelines, mechanisms, metadata standards, etc.)
+their reviewers (e.g., guidelines, mechanisms, metadata standards)
 \end{itemize}
@@
 WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
-Department of Astronomy, University of Illinois titled \emph{Why Sustain
+Department of Astronomy, University of Illinois, titled \emph{Why Sustain
 Scientific Software?}. Turk is a prolific scientific software practitioner and
@@
 users, and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing community generally adopt some top-down approaches,
+Turk that the supercomputing community generally adopts some top-down approaches,
 whereas the SciPy community more often than not uses more bottom-up systems. The
 essential messages perceived were also often bipolar: the supercomputing
-community thinks software is getting harder, with exascale computing and
-optimization issues in mind, but the SciPy community thinks software becoming
+community thinks that software is getting harder, with exascale computing and
+optimization issues in mind; but the SciPy community thinks that software is becoming
 better, with emerging tools such as Jupiter and productivity packages for
-research workflow. Admitting such comparisons are somewhat unfair
+research workflows. Admitting such comparisons are somewhat unfair
 generalizations, Turk reminded the audience that the different approaches bring
@@
 agency provided funds to a group who are not necessarily exclusively working on
-and developing a piece of software, keeps it going, and provides it to the
+and developing the software, keeps it going, and provides it to the
 scientific community. The model of productized software, in which a piece of
-software has grown into something that individuals rely on to the point that
+software has grown to the point that
 research groups or people are willing to support it with some amount of funding,
@@
 being sustainable and self-sufficient. He thought it was not necessarily the
-case and furthermore, it could lead to a divergence of interest between users
+case and furthermore, it could lead to a divergence of interests between users
 and developers.
@@
 a small amount of time for working on a piece of software necessary for that
-project. However, all researchers' ability to participate in that volunteer
+project. However, researchers' abilities to participate in that volunteer
 community are not always the same and may not always be aligned with their
@@
 the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
-projects can be just as large as a top-down software development project.
+projects can be just as large-scale as a top-down software development project.
 
@@
 don't cause novel ideas and packages to suffer at the hands of the status quo.
-%This is a straw man concept, but it's something to look out for, since
-organizations will act in their own interest.
 
@@
 and creative thrill in writing codes and papers; usefulness with measurable
-impacts, for example, LINPACK and HDF groups providing of data storage to
+impacts, for example, LINPACK and HDF groups providing data storage to
 satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
@@
 
+The lightning talks were intended to give an opportunity for
+attendees to quickly highlight an important issue or a potential solution. 
 \begin{enumerate}

### 2015-11-12 WSSSPE3/report/wg_main_best_practices.tex
Fixed to do, and tweaked wording.
Update wg_main_best_practices.tex

@@
 \subsection{White paper/journal paper about best practices in developing sustainable software}
+\label{sec:best-practices}
 
 \subsubsection{Why it is important}
-\todo{short text here}
+Reviewing multiple past articles and talks at different meetings like WSSPEx\cite{WSSPE papers, heroux paper, ...} and 
+analyzing and promoting sustainable scientific software makes it clear that there 
+are several common and recurring ideas that underpin success. However, outside of a small community, this knowledge is not widely
+shared. This is  especially true for the large community of scientists who generate most of the software used by scientists but are not primarily software developers. In this
+scenario a clear and precise exposition of these best practices collected from many sources and open collaboration among all in the community
+in a single source (e.g. a journal paper, tutorial etc.)  that can be widely disseminated is necessary and likely to be very valuable.
 
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+
+The creation of such a ``best practices" document will build upon the range of activities and topics discussed at WSSPE3 and associated prior meeting. We will attempt to
+distill the emerging body of knowledge into this document.
+The large number of  articles from the NSF funded S2I2 projects (SSE and SSI) ``lightning talks", ``white papers"  and reports from different workshops point have created a large if somewhat diffuse source for this report.
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
+
+Core questions that will need to be explored are in knowledge management, 
+(transitions between people), reliability (reproducibility), usability and how a software tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability. Ideas 
+that may need to be explored include:
+\begin{itemize}
+\item Requirements engineering to create tools with immediate uptake;
+\item When should software "die"?
+\item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
+\item Dimensions of sustainability - economic, technical, environmental, 
+declining interest in primary application area), social.
+\item 
+\end{itemize}
+
+Sustainability requires community participation in code development and/or a wide adoption of software.
+The larger the community base is using a software, the better are the funding possibilities and thus also the sustainability options.
+Additionally developer commitment to an application is essential and experience shows that especially software packages with an evangelist imposing strong inspiration and discipline is required to achieve sustainability.
+While a single person can push sustainability to a certain level, open source software also needs sustained commitment from the developer community.
+Such sustained commitments include diverse tasks and roles, which can be fulfilled by diverse developers with different knowledge levels.
+Besides developing software and appropriate software management with measures for extensibility and scalability of the software, active (expertise) support for users via a user forum with a quick turnaround is crucial.
+The barrier to entry the community as user as well as developer has to be as low as possible.
 
 \subsubsection{Plans}
-\todo{short text here - not bullets}
+%\todo{short text here - not bullets}
+The creation of a best-practices document needs a large and diverse community involved. We have enlisted over ten contributors from the attendees at the WSSPE3 and 
+those on the mailing list. The primary mechanism for developing these will be to examine and analyze the success of several well known community scientific
+software and organizations supporting scientific software. We will attempt then to abstract general principles and best practices. Some of the tools identified for such analysis are
+the general purpose PeTSC toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.
+We also established a timeline and a rough outline for the report.
 
-[15 Nov] Introduction and scope finished
-[15 Nov] Sections assigned
-[31 Jan] Analysing funding possibilities for survey
-[31 Jan] First versions of section
-[15 Feb] Distribution to WSSSPE community
-[31 Mar] Final version of white paper
-[30 Apr] Submission of peer-reviewed paper?
+\noindent{\bf Timeline:}
+\begin{itemize}
+%
+\item [15 Nov] Introduction and scope finished
+\item [15 Nov] Sections assigned
+\item [31 Jan] Analysing funding possibilities for survey
+\item [31 Jan] First versions of section
+\item [15 Feb] Distribution to WSSSPE community
+\item [31 Mar] Final version of white paper
+\item [30 Apr] Submission of peer-reviewed paper?
+\end{itemize}
 
+\noindent{\bf Outline:}
 \begin{enumerate}
@@
 \item PeTSC
-\item NWCHEM
+\item NWChem
 \item CIG
@@
 \subsubsection{Landing Page}
-\todo{link to landing page}
+The landing page with instructions, timeline and the white paper is here: https://drive.google.com/drive/folders/0B7KZv1TRi06fbnFkZjQ0ZEJKckk
+Discussions can be also continued in https://github.com/WSSSPE/meetings/issues/42

### 2015-11-12 WSSSPE3/report/wg_main_best_practices.tex
fixing WSSSPE spelling

@@
 \subsubsection{Why it is important}
-Reviewing multiple past articles and talks at different meetings like WSSPEx\cite{WSSPE papers, heroux paper, ...} and 
+Reviewing multiple past articles and talks at different meetings like WSSSPEx\cite{WSSSPE papers, heroux paper, ...} and 
 analyzing and promoting sustainable scientific software makes it clear that there 
@@
 
-The creation of such a ``best practices" document will build upon the range of activities and topics discussed at WSSPE3 and associated prior meeting. We will attempt to
+The creation of such a ``best practices" document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meeting. We will attempt to
 distill the emerging body of knowledge into this document.
@@
 %\todo{short text here - not bullets}
-The creation of a best-practices document needs a large and diverse community involved. We have enlisted over ten contributors from the attendees at the WSSPE3 and 
+The creation of a best-practices document needs a large and diverse community involved. We have enlisted over ten contributors from the attendees at the WSSSPE3 and 
 those on the mailing list. The primary mechanism for developing these will be to examine and analyze the success of several well known community scientific

### 2015-11-12 WSSSPE3/report/wg_appendix_best_practices.tex
Create wg_appendix_best_practices.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Update wg_appendix_best_practices.tex
Former-commit-id: f934153c37aef1cf5a20b528497e0002c91c4419
Minor changes.
Update wg_appendix_best_practices.tex

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Best Practices Group Discussion}
+\label{sec:appendix_best_practices}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{add POC here}
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+
+\begin{itemize}
+\item Abani Patra 
+\item Sandra Gesing -- University of Notre Dame
+\item Neil Chue Hong 
+\item Greg Tucker 
+\item Birgit Penzens 
+\item Abigail Cabunoc Mayes 
+\item Jeff Carver 
+\item Frank L\"offler 
+\item Colin Venter 
+\item Lorraine Hwang 
+\item Sou-Cheng Choi
+\item Suresh Marru 
+\item Don Middleton 
+\item Daniel Katz  
+\item Kyle Niemeyer 
+\item Jeff Carver 
+\item Dan Gunter 
+\item Alexander Konovalov 
+\item Tom Crick
+
+\end{itemize}
 
 \subsection{Summary of Discussion}
+Core questions that will need to be explored are in knowledge management, 
+(transitions between people), reliability (reproducibility), usability and how a software tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability. Ideas 
+that may need to be explored include:
+\begin{itemize}
+\item Requirements engineering to create tools with immediate uptake;
+\item When should software "die"?
+\item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
+\item Dimensions of sustainability - economic, technical, environmental, 
+declining interest in primary application area), social.
+\item 
+\end{itemize}
+
+Sustainability requires community participation in code development and/or a wide adoption of software.
+The larger the community base is using a software, the better are the funding possibilities and thus also the sustainability options.
+Additionally developer commitment to an application is essential and experience shows that especially software packages with an evangelist imposing strong inspiration and discipline is required to achieve sustainability.
+While a single person can push sustainability to a certain level, open source software also needs sustained commitment from the developer community.
+Such sustained commitments include diverse tasks and roles, which can be fulfilled by diverse developers with different knowledge levels.
+Besides developing software and appropriate software management with measures for extensibility and scalability of the software, active (expertise) support for users via a user forum with a quick turnaround is crucial.
+The barrier to entry the community as user as well as developer has to be as low as possible.
 
@@
 
+The opportunity lays in the collaboration on a white paper, which will be revisited regularly for further improvements and enhances the state on best practices resulting in a peer-reviewed paper.
+This way we would like to reach a wide community.
+But these are also the challenges and obstacles - to get everyone to write on the paper and reach the community.
+
 \subsection{Key Next Steps}
+
+The key next steps are to write the introduction, reach out to the co-authors and agree on a scope of the white paper.
 
@@
 
+Sandra Gesing and Abani Patra are the main editors and will organize the overall communication and paper. Sections will be assigned to diverse co-authors.
+
 \subsection{What Else is Needed?}
 
+At the moment we don't see any further requirements.
+
 \subsection{Key Milestones and Responsible Parties}
+\item [15 Nov] Introduction and scope finished (Abani Patra/Sandra Gesing)
+\item [15 Nov] Sections assigned (Abani Patra/Sandra Gesing)
+\item [31 Jan] Analysing funding possibilities for survey
+\item [31 Jan] First versions of section
+\item [15 Feb] Distribution to WSSSPE community
+\item [31 Mar] Final version of white paper
+\item [30 Apr] Submission of peer-reviewed paper?
 
 \subsection{Description of Funding Needed}
+We might need funding for a journal publication (open-access options).

### 2015-11-12 WSSSPE3/report/wg_appendix_best_practices.tex
small formatting/etc issues

@@
 \item Jeff Carver 
-\item Frank L\"offler 
+\item Frank L\"{o}ffler 
 \item Colin Venter 
@@
 \item Requirements engineering to create tools with immediate uptake;
-\item When should software "die"?
+\item When should software ``die''?
 \item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
@@
 declining interest in primary application area), social.
-\item 
 \end{itemize}
@@
 \subsection{Key Milestones and Responsible Parties}
-\item [15 Nov] Introduction and scope finished (Abani Patra/Sandra Gesing)
-\item [15 Nov] Sections assigned (Abani Patra/Sandra Gesing)
-\item [31 Jan] Analysing funding possibilities for survey
-\item [31 Jan] First versions of section
-\item [15 Feb] Distribution to WSSSPE community
-\item [31 Mar] Final version of white paper
-\item [30 Apr] Submission of peer-reviewed paper?
+\begin{itemize}
+\item 15 Nov: Introduction and scope finished (Abani Patra/Sandra Gesing)
+\item 15 Nov: Sections assigned (Abani Patra/Sandra Gesing)
+\item 31 Jan: Analyzing funding possibilities for survey
+\item 31 Jan: First versions of section
+\item 15 Feb: Distribution to WSSSPE community
+\item 31 Mar: Final version of white paper
+\item 30 Apr: Submission of peer-reviewed paper?
+\end{itemize}
 

### 2015-11-12 WSSSPE3/report/wg_main_best_practices.tex
small formatting/etc issues

@@
 
-\subsubsection{Why it is important}
+%\subsubsection{Why it is important}
 Reviewing multiple past articles and talks at different meetings like WSSSPEx\cite{WSSSPE papers, heroux paper, ...} and 
 analyzing and promoting sustainable scientific software makes it clear that there 
-are several common and recurring ideas that underpin success. However, outside of a small community, this knowledge is not widely
+are several common and recurring ideas that underpin success in developing sustainable software. However, outside of a small community, this knowledge is not widely
 shared. This is  especially true for the large community of scientists who generate most of the software used by scientists but are not primarily software developers. In this
@@
 
-The creation of such a ``best practices" document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meeting. We will attempt to
+The creation of such a ``best practices'' document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meeting. We will attempt to
 distill the emerging body of knowledge into this document.
-The large number of  articles from the NSF funded S2I2 projects (SSE and SSI) ``lightning talks", ``white papers"  and reports from different workshops point have created a large if somewhat diffuse source for this report.
+The large number of  articles from the NSF funded S2I2 projects (SSE and SSI) ``lightning talks'', ``white papers''  and reports from different workshops point have created a large if somewhat diffuse source for this report.
 
@@
 The barrier to entry the community as user as well as developer has to be as low as possible.
+
+For additional information about the discussion, see Appendix~\ref{sec:appendix_best_practices}.
 
@@
 %
-\item [15 Nov] Introduction and scope finished
-\item [15 Nov] Sections assigned
-\item [31 Jan] Analysing funding possibilities for survey
-\item [31 Jan] First versions of section
-\item [15 Feb] Distribution to WSSSPE community
-\item [31 Mar] Final version of white paper
-\item [30 Apr] Submission of peer-reviewed paper?
+\item 15 Nov: Introduction and scope finished
+\item 15 Nov: Sections assigned
+\item 31 Jan: Analyzing funding possibilities for survey
+\item 31 Jan: First versions of section
+\item 15 Feb: Distribution to WSSSPE community
+\item 31 Mar: Final version of white paper
+\item 30 Apr: Submission of peer-reviewed paper?
 \end{itemize}
@@
 \subsubsection{Landing Page}
-The landing page with instructions, timeline and the white paper is here: https://drive.google.com/drive/folders/0B7KZv1TRi06fbnFkZjQ0ZEJKckk
-Discussions can be also continued in https://github.com/WSSSPE/meetings/issues/42
+The landing page with instructions, timeline and the white paper is here: \url{https://drive.google.com/drive/folders/0B7KZv1TRi06fbnFkZjQ0ZEJKckk}
+Discussions can be also continued in \url{https://github.com/WSSSPE/meetings/issues/42}

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
Section 4: minor changes and added questions.

@@
 
-The WSSSPE3 workshop included multiple mechanisms for participation and encouraged team
-building around solutions. It strongly encouraged participation of early-career
-scientists, postdoctoral researchers, and graduate students, with funds provided
-to the conference organizers by the Moore Foundation, the National Science
-Foundation (NSF), and the Software Sustainability Institute, to support the
-travel of potential participants who would not otherwise be able to attend,
-early-stage researchers, and those from underrepresented groups. These funds
-allowed 16 additional participants to attend.
+The WSSSPE3 workshop included multiple mechanisms for participation and
+encouraged team building around solutions. It strongly encouraged participation
+of early-career scientists, postdoctoral researchers, and graduate students,
+with funds provided to the conference organizers by the Moore Foundation, the
+National Science Foundation (NSF), and the Software Sustainability Institute, to
+support the travel of potential participants who would not otherwise be able to
+attend, early-stage researchers, and those from underrepresented groups. These
+funds allowed 16 additional participants to attend.
 
-This report is based on collaborative notes taken during the workshop,
-which were linked from the GitHub issues that represented the potential and
-actual working
+This report is based on collaborative notes taken during the workshop, which
+were linked from the GitHub issues that represented the potential and actual
+working
 groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
@@
 (\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
-lightning talks (\S\ref{sec:lightning}). The
-report also gives summaries of action plans proposed by the working groups
-(\S\ref{sec:WGs}), and some conclusions (\S\ref{sec:conclusions}). Lists of the
-organizing committee (Appendix~\ref{sec:orgcom}), the registered attendees
+lightning talks (\S\ref{sec:lightning}). The report also gives summaries of
+action plans proposed by the working groups (\S\ref{sec:WGs}), and some
+conclusions (\S\ref{sec:conclusions}). Lists of the organizing committee
+(Appendix~\ref{sec:orgcom}), the registered attendees
 (Appendix~\ref{sec:attendees}), and the travel award recipients
@@
 
-The first call for participation requested lightning talks, where each author could
-make a brief statement about work that either had been done or was needed, with
-the goal of contributing to the discussion of one or more working groups. There
-were~24 lightning talks submitted, and, after a peer-review process, 16 of those
-were accepted, as discussed further in Section~\ref{sec:lightning}.
+The first call for participation requested lightning talks, where each author
+could make a brief statement about work that either had been done or was needed,
+with the goal of contributing to the discussion of one or more working groups.
+There were~24 lightning talks submitted, and, after a peer-review process, 16 of
+those were accepted, as discussed further in Section~\ref{sec:lightning}.
 
@@
 WSSSPE2, and requested additional suggestions. The combination of existing and
-new topics led to the following 18 potential topics that were advertised in the subsequent
-calls for participation:
+new topics led to the following 18 potential topics that were advertised in the
+subsequent calls for participation:
 
@@
 \item Documenting case studies of receiving credit for software contributions
-\item Developing a system of awards and recognitions to encourage sustainable software
+\item Developing a system of awards and recognitions to encourage sustainable
+software
 \end{itemize}
@@
 users, and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing community generally adopts some top-down approaches,
-whereas the SciPy community more often than not uses more bottom-up systems. The
-essential messages perceived were also often bipolar: the supercomputing
-community thinks that software is getting harder, with exascale computing and
-optimization issues in mind; but the SciPy community thinks that software is becoming
-better, with emerging tools such as Jupiter and productivity packages for
-research workflows. Admitting such comparisons are somewhat unfair
-generalizations, Turk reminded the audience that the different approaches bring
-different types of ideas to the table, and he welcomed WSSSPE3 being conducted
-outside existing preconceptions.
+Turk that the supercomputing community generally adopts some top-down
+approaches, whereas the SciPy community more often than not uses more bottom-up
+systems. The essential messages perceived were also often bipolar: the
+supercomputing community thinks that software is getting harder, with exascale
+computing and optimization issues in mind; but the SciPy community thinks that
+software is becoming better, with emerging tools such as Jupiter and
+productivity packages for research workflows. Admitting such comparisons are
+somewhat unfair generalizations, Turk reminded the audience that the different
+approaches bring different types of ideas to the table, and he welcomed WSSSPE3
+being conducted outside existing preconceptions.
 
@@
 agency provided funds to a group who are not necessarily exclusively working on
-and developing the software, keeps it going, and provides it to the
-scientific community. The model of productized software, in which a piece of
-software has grown to the point that
-research groups or people are willing to support it with some amount of funding,
-for instance, a subscription to use cloud services that deploy a piece of
-software, or purchase of a piece of software. A final model Turk felt conflicted
-about is a volunteer model that is traditional old-school---not modern-day open
-source---development.
+and developing the software, keeps it going, and provides it to the scientific
+community. The model of productized software, in which a piece of software has
+grown to the point that research groups or people are willing to support it with
+some amount of funding, for instance, a subscription to use cloud services that
+deploy a piece of software, or purchase of a piece of software. A final model
+Turk felt conflicted about is a volunteer model that is traditional
+old-school---not modern-day open source---development.
 
@@
 
-The lightning talks were intended to give an opportunity for
-attendees to quickly highlight an important issue or a potential solution. 
+The lightning talks were intended to give an opportunity for attendees to
+quickly highlight an important issue or a potential solution.
+%
 \begin{enumerate}
-\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs. Stability:
-Facilitating Research Training While Supporting Scientific Research}} A case
-study: ``The Cooperative Computing Lab'' at the University of Notre Dame is a
+\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
+Facilitating Research Training While Supporting Scientific Research}} Benjamin Tovar presented a case
+study of ``The Cooperative Computing Lab'' at the University of Notre Dame, which is a
 small group of individuals whose main tasks are collaborating with people that
@@
 systems, conducting computer science research, and developing open source
-software. One of the main challenges they face is finding the balance between
+software. One of the main challenges they face is finding a balance between
 flexibility/training and stability/quality. Their current solution for ensuring
 the latter was to add a software engineer (the presenter) to the existing team
-of faculty and students, who now also serves as ``spring'' between flexibility
+of faculty and students, who now also serves as a ``spring'' between flexibility
 and stability.
@@
 events that focus on the topic. Nevertheless, it isn't well understood yet.
-Penzenstadler argued that we often define sustainability too narrowly. Instead,
+Birgit Penzenstadler argued that we often define sustainability too narrowly. Instead,
 sustainability at its heart is a systemic concept and must be viewed from a
 range of different dimensions including environmental, economic, individual,
-social and technical. She introduced the Karlskrona Manifesto on Software
+social, and technical. She introduced the Karlskrona Manifesto on Software
 Design~\cite{Becker:2014}, which distills knowledge from a broad range of
@@
 this, the Karlskrona Manifesto proposes nine principles and commitments. These
-commitments are not dogmatic laws, but rather a commitment to rethink, to move
-beyond the silo mentality, and analyze in more depth. As such, they do not
+commitments are not dogmatic laws, but rather commitments to rethink, to move
+beyond the silo mentality, and to analyze in more depth. As such, they do not
 restrict, but rather open up a space for discussion.
@@
 and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
-to Improve Productivity of Legacy Tools}} Abani Patra presented an example how
+to Improve Productivity of Legacy Tools}} Abani Patra presented an example of how
 the community using Titan2D, a Geoflow Simulation Software, increased the
@@
 system for the source code, coupled with multiple versions of the same code
-base, the fixed format of input files, that many input values used to be set as
+base, the fixed format of input files, that many input values were set as
 compilation flags, and that the internal data layout was not suitable for modern
 technologies (vectorization, accelerators). The approach of the Titan2D
-developers included reinforcing the code structure using a multiple layers of
-Python and C++, and a redesign of the data layout to be better suitable for
+developers included reinforcing the code structure using  multiple layers of
+Python and C++ interfaces, and a redesign of the data layout to be more suitable for
 modern CPUs and accelerators.
@@
 good first bug reports. Together, Abigail concluded, these two properties not
-only lead to higher quality, reusability and ease of understanding, but
+only lead to higher quality, reusability, and ease of understanding, but also
 eventually, to sustainability.
@@
 Like others, their main goals include high usability, sustainability, and
-reproducibility. In order to achieve this, communication channels have been
+reproducibility. In order to achieve these goals, communication channels have been
 developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
 webinars. In order to contribute to the infrastructure, codes have to adhere to
-specified minimum standards, but ideals are also given to encourage reaching the
-aforementioned goals. This includes, e.g., the use of version control, certain
-coding styles, the presence and nature of code tests and documentation and
+specified minimum standards, but ideals \choinote{ideals or ideas?} are also given to encourage reaching the
+aforementioned goals. These include, e.g., the use of version control, certain
+coding styles, the presence and nature of code tests and documentation, and
 certain user workflows.
@@
 Hwang mentioned various ways to cite efforts within the framework, including
-science papers, code papers, user manuals, and the website. An analysis of the
+science papers, code papers, user manuals, and the CIG website. An analysis of the
 resulting citations showed that $80\%$ of papers that use CIG codes mention the
@@
 cite codes (including non-CIG codes), and only about one eighth specify the
-version used, showing that compared to other codes, CIG seems to be much better
+version used. Compared to other codes, CIG seems to be much better
 cited. This, in part, is attributed to suggested citation forms communicated to
@@
 Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
-located at Notre Dame and Chicago, that links the HEP effort to Biology,
+located at Notre Dame and Chicago, that links the HEP \choinote{What is HEP?} effort to Biology,
 Astrophysics, Digital Curation, and other disciplines. It includes physicists,
@@
 commonality across disciplines. Examples are meta-data descriptions of archived
-data, computational descriptions, descriptions of how data was processed, but
-also includes questions such as whether computation replication can be
+data, computational descriptions, descriptions of how data was processed, 
+questions such as whether computation replication can be
 automated, and what the impact of access policies on the preservation
@@
 the preservation software, this means that besides adding value to the
-community, it needs to be transparent to their workflow, i.e., not requiring
+community, it needs to be transparent to their workflows, i.e., not requiring
 additional effort to preserve ``research objects''.
@@
 workflows. Some limited adoption can be seen today in research communities, but
-the main problems have not been solved to a sufficient level. James Hetherington
-hypothesizes that instead of technical solutions, social innovation is needed: a
+the main problems have not been solved to a sufficient level.  Hetherington
+hypothesized that instead of technical solutions, social innovation is needed: a
 new role in the academic system focused on research software that combines the
@@
 the possibility of training in reproducible computational research, providing
-collaborations for for researchers who don't want to be programmers, and
+collaborations for researchers who don't want to be programmers, and
 synergies with research computing platforms. The success of such a group would
@@
 including funding from the UK Research Council. An open question is whether this
-approach can be adopted also in other countries, including the USA.
+approach can be adopted in other countries, including the USA.
 
@@
 usual suspects of funding, great developers, good design, and software
-engineering practices, however, Dan placed the users. He explained this using a
-traditional software development model starting from gathering requirements, and
-reaching release through a design and development and testing phase. The main
-deficiency that with this approach was pointed out to be the too-late
+engineering practices, Dan placed the users. He explained this using a
+traditional software-development model starting from gathering requirements, and
+reaching release through a design and development and testing phase \choinote{very complicated phrase of the phase}. The main
+deficiency with this approach was pointed out to be the too-late
 interaction with users. Instead, an alternative approach was proposed that, at
@@
 big challenges is to agree on standard to enable a single, maintainable system.
-Working in isolation does not work (anymore) within this field, and they also
+Working in isolation does not work (anymore) within this field, and they \choinote{who are they?} also
 didn't have standards for software engineering. This is contrasted to computer
@@
 Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
-approach to overcome some hurdles in scientific publishing: the disconnect
+approach to overcome some hurdles in scientific publishing: disconnects
 between experimental data, research software, and publications. A current effort
@@
 four elements. First, it forms a modern paper including text, data, and pointers
-to supplementary material. Second, it is reproducible, including data
+to supplementary materials. Second, it is reproducible, including data
 processing, workflow, and visualization tools. Third, it is part of open
@@
 like persistent identifiers for software, data, and authors, and it cites both
-data and software. OntoSoft is an effort to help this effort, providing software
+data and software. OntoSoft is a tool to help this effort, providing software
 stewardship for the Geosciences. As part of this initiative, a special issue of
-a journal in a Geoscience areas is planned that only includes Geoscience papers
+a journal in Geoscience areas is planned to include only Geoscience papers
 of the future, with submissions open until the end of 2015. In addition,
@@
 the effort of the standardized, easy to read, and understandable nutritional
-labeling of goods, which only contain a small set of categories. However, such a
+labeling of food, which only contain a small set of categories. However, such a
 framework for software would be more difficult due to different existing
 community norms. The challenges such a framework faces include that many
-measurements are subjective, many metrics are too costly, and that
+measurements are subjective, that many metrics are too costly, and that
 self-assessment needs to be encouraged. Possible categories would include

### 2015-11-12 WSSSPE3/report/WSSSPE3_review.tex
answering questions, fixing some text

@@
 webinars. In order to contribute to the infrastructure, codes have to adhere to
-specified minimum standards, but ideals \choinote{ideals or ideas?} are also given to encourage reaching the
+specified minimum standards, but ideals %\choinote{ideals or ideas?} \katznote{ideals}
+are also given to encourage reaching the
 aforementioned goals. These include, e.g., the use of version control, certain
@@
 Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
-located at Notre Dame and Chicago, that links the HEP \choinote{What is HEP?} effort to Biology,
-Astrophysics, Digital Curation, and other disciplines. It includes physicists,
+located at Notre Dame and Chicago, that links the high energy physics effort to biology,
+astrophysics, digital curation, and other disciplines. It includes physicists,
 digital librarians, as well as computer scientists, and aims to achieve some
@@
 traditional software-development model starting from gathering requirements, and
-reaching release through a design and development and testing phase \choinote{very complicated phrase of the phase}. The main
+reaching release through a design, development, and testing phase. The main
 deficiency with this approach was pointed out to be the too-late
@@
 big challenges is to agree on standard to enable a single, maintainable system.
-Working in isolation does not work (anymore) within this field, and they \choinote{who are they?} also
+Working in isolation does not work (anymore) within this field, and the community
+in the field also
 didn't have standards for software engineering. This is contrasted to computer

### 2015-11-12 WSSSPE3/report/wg_main_best_practices.tex
edits/comments

@@
 shared. This is  especially true for the large community of scientists who generate most of the software used by scientists but are not primarily software developers. In this
-scenario a clear and precise exposition of these best practices collected from many sources and open collaboration among all in the community
-in a single source (e.g. a journal paper, tutorial etc.)  that can be widely disseminated is necessary and likely to be very valuable.
+scenario, a clear and precise exposition of these best practices collected from many sources and open collaboration among all in the community
+in a single source (e.g., journal paper, tutorial)  that can be widely disseminated is necessary and likely to be very valuable.
 
@@
 
-The creation of such a ``best practices'' document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meeting. We will attempt to
+The creation of such a ``best practices'' document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meetings. We will attempt to
 distill the emerging body of knowledge into this document.
-The large number of  articles from the NSF funded S2I2 projects (SSE and SSI) ``lightning talks'', ``white papers''  and reports from different workshops point have created a large if somewhat diffuse source for this report.
+The large number of  articles from the NSF funded SI2 projects (SSE and SSI), ``lightning talks'', ``white papers,'' and reports from different workshops have created a large if somewhat diffuse source for this report.
 
@@
 Core questions that will need to be explored are in knowledge management, 
-(transitions between people), reliability (reproducibility), usability and how a software tool becomes part of the core workflow of well identified users (stakeholders)
-relating to tool success and hence sustainability. Ideas 
+(transitions between people), reliability (reproducibility), usability, and how a software tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability. \katznote{prev sentence is complex and awkward} Ideas 
 that may need to be explored include:
@@
 \item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
-\item Dimensions of sustainability - economic, technical, environmental, 
-declining interest in primary application area), social.
+\item Dimensions of sustainability -- economic, technical, environmental, 
+declining interest in primary application area), \katznote{not sure what the prev. comment goes with} social.
 \end{itemize}
@@
 Sustainability requires community participation in code development and/or a wide adoption of software.
-The larger the community base is using a software, the better are the funding possibilities and thus also the sustainability options.
-Additionally developer commitment to an application is essential and experience shows that especially software packages with an evangelist imposing strong inspiration and discipline is required to achieve sustainability.
+The larger the community base is using a piece of software, the better are the funding possibilities and thus also the sustainability options.
+Additionally developer commitment to an application is essential and experience shows that software packages with an evangelist imposing strong inspiration and discipline are more likely to achieve sustainability.
 While a single person can push sustainability to a certain level, open source software also needs sustained commitment from the developer community.
@@
 Besides developing software and appropriate software management with measures for extensibility and scalability of the software, active (expertise) support for users via a user forum with a quick turnaround is crucial.
-The barrier to entry the community as user as well as developer has to be as low as possible.
+The barrier to entry for the community as users as well as developers has to be as low as possible.
 
@@
 The creation of a best-practices document needs a large and diverse community involved. We have enlisted over ten contributors from the attendees at the WSSSPE3 and 
-those on the mailing list. The primary mechanism for developing these will be to examine and analyze the success of several well known community scientific
-software and organizations supporting scientific software. We will attempt then to abstract general principles and best practices. Some of the tools identified for such analysis are
+those on the mailing list.
+The primary mechanism for developing this document will be to examine and analyze the success of several well known community scientific
+software and organizations supporting scientific software.
+We will attempt then to abstract general principles and best practices.
+Some of the tools identified for such analysis are
 the general purpose PeTSC toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.

### 2015-11-12 WSSSPE3/report/wg_appendix_best_practices.tex
more comments/edits

@@
 \subsection{Group Members}
+\katznote{add affils for all, please}
 
@@
 \subsection{Summary of Discussion}
+
 Core questions that will need to be explored are in knowledge management, 
-(transitions between people), reliability (reproducibility), usability and how a software tool becomes part of the core workflow of well identified users (stakeholders)
-relating to tool success and hence sustainability. Ideas 
+(transitions between people), reliability (reproducibility), usability, and how a software tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability. \katznote{prev sentence is complex and awkward} Ideas 
 that may need to be explored include:
@@
 
-The opportunity lays in the collaboration on a white paper, which will be revisited regularly for further improvements and enhances the state on best practices resulting in a peer-reviewed paper.
-This way we would like to reach a wide community.
-But these are also the challenges and obstacles - to get everyone to write on the paper and reach the community.
+The opportunity lies in collaboration on a white paper, which will be revisited regularly for further improvements, to enhance knowledge of the state of best practices, resulting in a peer-reviewed paper.
+We would like to reach a wide community by doing this.
+But these are also the challenges and obstacles -- to get everyone to contribute to the paper and to reach the community.
 
@@
 
-The key next steps are to write the introduction, reach out to the co-authors and agree on a scope of the white paper.
+The key next steps are to write the introduction, reach out to the co-authors, and to agree on the scope of the white paper.
 

### 2015-11-13 WSSSPE3/report/wg_main_funding_spec_expert.tex
edits in RSE main section

@@
 Research software engineers will not be measured by publications, we hope, but by other measures.
-\todo{what measures?} Middle-author publications are common for RSEs.
-Most RSEs welcome co-authorship on papers where the PIs consider the contribution deserves it.
-(these last two statements seem contradictory, not sure which message is to be used?)
+Middle-author publications are common for RSEs.
+Most RSEs welcome co-authorship on papers where the PIs think that the contribution deserves it.
 
@@
 
-For more detail on the discussion, see Appendix~\ref{sec:appendix_funding_spec_expert}.
+For more details on the discussion, see Appendix~\ref{sec:appendix_funding_spec_expert}.
 

### 2015-11-13 WSSSPE3/report/wg_main_user_community.tex
adding cite

@@
 
-Mozilla Science maintains a ``Working Open Project Guide'' (\url{http://mozillascience.github.io/leadership-training}). \katznote{should get changed to a cite} From the introduction:
+Mozilla Science maintains a ``Working Open Project Guide'' (\url{http://mozillascience.github.io/leadership-training}), the introduction of which states: \katznote{should get changed to a cite}
 \begin{quote}
@@
 \begin{itemize}
-\item ``Art of Community'' by Jono Bacon. \katznote{cite?} We could consider distilling this for scientific software.
+\item ``Art of Community'' by Jono Bacon~\cite{art-of-community}. We could consider distilling this for scientific software.
 \item Iain Larmour, from EPSRC in the UK (EPSRC: https://www.epsrc.ac.uk/) (not sure who from mentioned UK Collaborative Computational Projects (CCP: http://ccp.ac.uk)

### 2015-11-13 WSSSPE3/report/WSSSPE3_review.tex
Section 4: minor changes. Added one question.

@@
 \item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
-Support for Sustainable Scientific Software}} Jeff argued that for scientific
+Support for Sustainable Scientific Software}} Carver argued that for scientific
 software to be truly sustainable, there is a need for developers to use
@@
 and science that hinders our ability to communicate and choose appropriate
-methods. In addition, many experience scientific software developers appear to
+methods. In addition, many experienced scientific software developers appear to
 be unaware of software engineering practices that may be beneficial to them. The
@@
 need more documented success stories. These successes need to be socialized 
+\choinote{not sure if ``socialize'' can be used in such a way}
 within the scientific community through workshops like the Software Engineering
-for Science workshop
-series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in \textit{Computing in Science and
+for Science Workshop
+Series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
+and the new Software Engineering track in the \textit{Computing in Science and
 Engineering} magazine.
@@
 analysis. Problems with this approach include obtaining agreement from the user,
-legal issues, increased maintenance (servers), ethical questions, but also the
+legal issues, increased maintenance (servers), ethical questions, and also the
 lack of a common infrastructure. Some of these problems are of a social nature
@@
 Developing Community}} The effort behind this talk by Ray Idaszak started from a
-break-out group at an NSF SI2 workshop in 2015, and centers around creating a
+break-out group at an NSF SI2 workshop in 2015, and centerd around creating a
 framework for creation of metrics for scientific software. This framework would
-improve both the metrics, and the software it evaluates, and is trying to build
-a community around this idea. With especially the latter idea in mind, a
-software ``peer review group'' would be created, representing stakeholders who
+improve both the metrics and the software it evaluates, and could also serve as a tool for building
+a community around the idea. With especially the latter idea (building
+a community) in mind, a software ``peer review group'' would be created, representing stakeholders who
 will self-review software created by their respective communities, and will
 concurrently develop metrics. The whole project should be community-governed,
-without a single institution overseeing the activity or infrastructure, with the
+without a single institution overseeing the activities or infrastructure, with the
 hope to evolving community-generated and adopted standards. The generation of
@@
 still in a design phase, with a white paper at the CSESSP workshop 2015 and this
-talk, but WSSSPE workshop is seen as a forum for the community to assemble and
+talk. The WSSSPE workshop is seen as a forum for the community to assemble and
 act, and is planned to be used also in the future to build this community and

### 2015-11-13 WSSSPE3/report/WSSSPE3_review.tex
changes to changes, some undoing

@@
 within the scientific community through workshops like the Software Engineering
-for Science Workshop
-Series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in the \textit{Computing in Science and
+for Science workshop
+series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
+and the new Software Engineering track in \textit{Computing in Science and
 Engineering} magazine.
@@
 Developing Community}} The effort behind this talk by Ray Idaszak started from a
-break-out group at an NSF SI2 workshop in 2015, and centerd around creating a
+break-out group at an NSF SI2 workshop in 2015, and centers around creating a
 framework for creation of metrics for scientific software. This framework would
@@
 still in a design phase, with a white paper at the CSESSP workshop 2015 and this
-talk. The WSSSPE workshop is seen as a forum for the community to assemble and
+talk, but the WSSSPE workshops are seen as a forum for the community to assemble and
 act, and is planned to be used also in the future to build this community and

### 2015-11-15 WSSSPE3/report/wg_main_funding_spec_expert.tex
Section 5.2: formatting. No content change.
Section 5.2: Minor changes.

@@
 
-Research Software Engineers -- those who contribute to science and scholarship
-through software development -- are an important part of the team
-needed to deliver 21st century research. However, existing academic structures
-and systems of funding do not effectively fund and sustain these skills.
-The resulting high levels of turnover and inappropriate incentives
-are a significant contributing factor to low levels of reliability and
-readability observed in scientific software. Moreover, the absence of skilled and experienced 
-developers retards progress in key projects, and at times causes important projects to fail completely.
+Research Software Engineers (RSEs) -- those who contribute to science and scholarship
+through software development -- are an important part of the team needed to
+deliver 21st century research. However, existing academic structures and systems
+of funding do not effectively fund and sustain these skills. The resulting high
+levels of turnover and inappropriate incentives are significant contributing
+factors to low levels of reliability and readability observed in scientific
+software. Moreover, the absence of skilled and experienced developers retards
+progress in key projects, and at times causes important projects to fail
+completely.
 
-Effective development of software for advanced research requires
-that researchers work closely with scientific software developers who understand
-the research domain sufficiently to build meaningful software at a reasonable pace. 
-This requires a collaborative approach -- where developers who are fully engaged/invested 
-in the research context are co-developing software with domain academics.
+Effective development of software for advanced research requires that
+researchers work closely with scientific software developers who understand the
+research domain sufficiently to build meaningful software at a reasonable pace.
+This requires a collaborative approach -- where developers who are fully
+engaged or invested in the research context are co-developing software with domain
+academics.
 
@@
 
-The solution we envision entails creating an environment where software developers 
-are a stable part of the research team. Such an environment mitigates the risk
-of losing a key developer at a critical moment in a projects lifetime, and provides
-the benefits of building a store of institutional knowledge about specific projects as well as about software 
-development for today's research. Our vision is to find a way to promote a University/research institute environment where
-software developers are stable components of research project teams.
+The solution we envision entails creating an environment where software
+developers are a stable part of a research team. Such an environment mitigates
+the risk of losing a key developer at a critical moment in a projects lifetime,
+and provides the benefits of building a store of institutional knowledge about
+specific projects as well as about software development for today's research.
+Our vision is to find a way to promote a university/research institute
+environment where software developers are stable components of research project
+teams.
 
-One strategy to promote stability is implementing a mechanism for developers to obtain academic
-credit for software development work. With such a mechanism in place, traditional academic funding
-models and career tracks could properly sustain individuals for whom software development is their
-primary contribution to research. A contributing factor to the problem with the current academic reward system is the
-devastating effect on an academic
-publication record resulting from time in industry; such postings often develop exactly the skills that research software
-engineers need, yet returns to university positions following an industry role are penalized by the current structures.
-Retention of senior developers is hard, because these people are highly in demand by the economy. However, people who have a
-PhD in science and enter industry, may desire to return for diverse reasons, and should be welcomed back.
+One strategy to promote stability is implementing a mechanism for developers to
+obtain academic credit for software development work. With such a mechanism in
+place, traditional academic funding models and career tracks could properly
+sustain individuals for whom software development is their primary contribution
+to research. A contributing factor to the problem with the current academic
+reward system is the devastating effect on an academic publication record
+resulting from time in industry; such postings often develop exactly the skills
+that research software engineers need, yet returns to university positions
+following an industry role are penalized by the current structures. Retention of
+senior developers is hard, because these people are high in demand by the
+economy. However, people who have a PhD in science and enter industry, may
+desire to return for diverse reasons, and should be welcomed back.
 
-While development of new mechanisms in the current academic reward system is a worthy aspirational goal, such a dramatic
-change in this structure does not seem likely in a time scale relevant to this working group. Accordingly, our working party
-sought alternate solutions that may be achievable within the context of existing academic structures. The group felt that
-developing dedicated research software engineering roles within the University, and finding stable funding for those individuals is the most promising mechanism for creating a stable software development staff.
+While development of new mechanisms in the current academic reward system is a
+worthy aspirational goal, such a dramatic change in this structure does not seem
+likely in a time scale relevant to this working group. Accordingly, our working
+party sought alternative solutions that may be achievable within the context of
+existing academic structures. The group felt that developing dedicated research
+software engineering roles within the university, and finding stable funding for
+those individuals is the most promising mechanism for creating a stable software
+development staff.
 
-Measures of impact and success for research programming groups, as well as for individual research software engineers, will
-be required in order to make the case to the University for continued funding.
-Research software engineers will not be measured by publications, we hope, but by other measures.
-Middle-author publications are common for RSEs.
-Most RSEs welcome co-authorship on papers where the PIs think that the contribution deserves it.
+Measures of impact and success for research programming groups, as well as for
+individual research software engineers, will be required in order to make the
+case to the university for continued funding. Research software engineers will
+not be measured by publications, we hope, but by other metrics. Middle-author
+publications are common for RSEs. Most RSEs welcome co-authorship on papers
+where the PIs think that the contribution deserves it.
 
@@
 
-It is hard for an individual PI in a university or college to support dedicated research software engineering resources, as
-the need for, and funding for, these activities is intermittent within the research cycle. To sustain this capacity, therefore, it is necessary to aggregate this work across multiple research groups.
+It is hard for an individual PI in a university or college to support dedicated
+research software engineering resources, as the need and funding for these
+activities are intermittent within a research cycle. To sustain this capacity,
+therefore, it is necessary to aggregate work across multiple research
+groups.
 
-One solution is to fund dedicated software engineering roles for major research software projects at national laboratories
-or other non-educational institutions. This solution is in place and working well for many well-used scientific codebases.
-However, this strategy has limited application, as much of the body of software is created and maintained in research
-universities. Therefore, we argue that research institutions should develop hybrid academic-technical tracks for this
-capacity, where employees in this track work with more than one PI, rather than the traditional RA role within a single group.
-This could be coordinated centrally, as a core facility, perhaps within research computing organizations which have
-traditionally supported university cyberinfrastructure, library organizations, or research offices. Alternatively, these
-groups could be organizationally closer to research groups, sitting within academic departments. The most effective model
-will vary from institution to institution, but the mandate and ways of working should be similar.
+One solution is to fund dedicated software engineering roles for major research
+software projects at national laboratories or other non-educational
+institutions. This solution is in place and working well for many well-used
+scientific codebases. However, this strategy has limited application, as much of
+the body of software is created and maintained in research universities.
+Therefore, we argue that research institutions should develop hybrid
+academic-technical tracks for this capacity, where employees in this track work
+with more than one PI, rather than the traditional RA role within a single
+group. This could be coordinated centrally, as a core facility, perhaps within
+research computing organizations which have traditionally supported university
+cyberinfrastructure, library organizations, or research offices. Alternatively,
+these groups could be organizationally closer to research groups, sitting within
+academic departments. The most effective model will vary from institution to
+institution, but the mandate and ways of working should be similar.
 
-Having convinced ourselves that this would be a positive innovation, we were then faced with the specific question of how to
-fund the initiation of this activity. A self-sustaining research software group will support itself through collaborations
-with PIs in the normal grant process, with PIs choosing to fund some amount of research software engineering effort through grants in
-the usual way. However, to bootstrap such a function to a level where it has sufficient reputation and client base to be self
--sustaining will generally require seed investment.
+Having convinced ourselves that this would be a positive innovation, we were
+then faced with the specific question of how to fund the initiation of this
+activity. A self-sustaining research software group will support itself through
+collaborations with PIs in the normal grant process, with PIs choosing to fund
+some amount of research software engineering effort through grants in the usual
+way. However, to bootstrap such a function to a level where it has sufficient
+reputation and client base to be self -sustaining will generally require seed
+investment.
 
-This might come from universities themselves (this was the model that led to the creation of the group in University College
-London), but more likely, seed funding needs to come from research councils (as with the Research Software
-Engineering Fellowship provided by the UK Engineering and Physical Sciences Research Council). We therefore recommend that
-funding organizations consider how they might provide such seed funding.
+This might come from universities themselves (this was the model that led to the
+creation of the group in University College London), but more likely, seed
+funding needs to come from research councils (as with the Research Software
+Engineering Fellowship provided by the UK Engineering and Physical Sciences
+Research Council). We therefore recommend that funding organizations consider
+how they might provide such seed funding.
 
-Success, appropriately measured, will help make the case to such funding bodies for further investment.
-One might expect that metrics such as improved productivity, software adoption rates, and grant success rates would be sufficient arguments in favor of such a model.
-However, useful measurement of code cleanliness, and the resulting productivity gains, is an unsolved problem in empirical software engineering.
-To measure ``what did not go wrong'' because of an intervention is particularly hard.
+Success, appropriately measured, will help make the case to such funding bodies
+for further investment. One might expect that metrics such as improved
+productivity, software adoption rates, and grant success rates would be
+sufficient arguments in favor of such a model. However, useful measurement of
+code cleanliness and the resulting productivity gains is an unsolved problem
+in empirical software engineering. To measure ``what did not go wrong'' because
+of an intervention is particularly hard.
 
-We finally noted that the institutional case for such groups is made easier by having successful examples to point to. In the
-UK, a collective effort to identify the research software engineering community, with individuals clearly stating ``I'm a research software engineer,'' has been important to the campaign.
-It will be useful to the global effort to similarly identify emerging research software organizations, and also, importantly, to identify longer-running research software groups, which have in some cases had a long running \emph{sui-generis} existence, but which now can be identified as part of a wider solution.
-There remains the problem of how to "sell" the value of this investment to investigators within the university.
-This is an issue best addressed by the individual organizations that embark on the plan.
+We finally noted that the institutional case for such groups is made easier by
+having successful examples to point to. In the UK, a collective effort to
+identify the research software engineering community, with individuals clearly
+stating ``I am a research software engineer,'' has been important to the
+campaign. It will be useful to the global effort to similarly identify emerging
+research software organizations, and also, importantly, to identify
+longer-running research software groups, which have in some cases had a long
+running \emph{sui-generis} existence, but which now can be identified as part of
+a wider solution. There remains the problem of how to ``sell'' the value of this
+investment to investigators within a university. This is an issue best
+addressed by the individual organizations that embark on the plan.
 
@@
 
-The first step in moving this strategy forward is to gather a list of groups that self-identify as research software engineering groups, and to reach out to other organizations to see if there may be a widespread community of RSEs who do not identify themselves as such at this time.
-We will collect information as to the organizational models under which these groups function, and how they are funded.
-For example, how many research universities currently fund people in the RSE track, whether they bear the the RSE moniker or not.
-Are these developers paid by the University or through a program supported by research grants/individual PIs?
-How did they bootstrap the developer track to get this started?
-How successful is the university in getting investigators to pay for fractional RSEs?
-We will author a report describing our findings, should funding be available to conduct the investigation.
+The first step in moving this strategy forward is to gather a list of groups
+that self-identify as research software engineering groups, and to reach out to
+other organizations to see if there may be a widespread community of RSEs who do
+not identify themselves as such at this time. We will collect information about
+the organizational models under which these groups function, and how they are
+funded. For example, how many research universities currently fund people in the
+RSE track, whether they bear the RSE moniker or not. Are these developers
+paid by the university or through a program supported by research
+grants/individual PIs? How did they bootstrap the developer track to get this
+started? How successful is the university in getting investigators to pay for
+fractional RSEs? We will author a report describing our findings, should funding
+be available to conduct the investigation.
 

### 2015-11-15 WSSSPE3/report/WSSSPE3_review.tex
Section 4: Clean up. No content change.

@@
 \item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
-Facilitating Research Training While Supporting Scientific Research}} Benjamin Tovar presented a case
-study of ``The Cooperative Computing Lab'' at the University of Notre Dame, which is a
-small group of individuals whose main tasks are collaborating with people that
-have large-scale computing problems, operating various parallel computer
-systems, conducting computer science research, and developing open source
-software. One of the main challenges they face is finding a balance between
-flexibility/training and stability/quality. Their current solution for ensuring
-the latter was to add a software engineer (the presenter) to the existing team
-of faculty and students, who now also serves as a ``spring'' between flexibility
-and stability.
+Facilitating Research Training While Supporting Scientific Research}} Benjamin
+Tovar presented a case study of ``The Cooperative Computing Lab'' at the
+University of Notre Dame, which is a small group of individuals whose main tasks
+are collaborating with people that have large-scale computing problems,
+operating various parallel computer systems, conducting computer science
+research, and developing open source software. One of the main challenges they
+face is finding a balance between flexibility/training and stability/quality.
+Their current solution for ensuring the latter was to add a software engineer
+(the presenter) to the existing team of faculty and students, who now also
+serves as a ``spring'' between flexibility and stability.
 
@@
 events that focus on the topic. Nevertheless, it isn't well understood yet.
-Birgit Penzenstadler argued that we often define sustainability too narrowly. Instead,
-sustainability at its heart is a systemic concept and must be viewed from a
-range of different dimensions including environmental, economic, individual,
-social, and technical. She introduced the Karlskrona Manifesto on Software
-Design~\cite{Becker:2014}, which distills knowledge from a broad range of
-related work on the topic of sustainability into a set of (mis-)perceptions and
-principles. The manifesto does not proclaim that there is an easy,
+Birgit Penzenstadler argued that we often define sustainability too narrowly.
+Instead, sustainability at its heart is a systemic concept and must be viewed
+from a range of different dimensions including environmental, economic,
+individual, social, and technical. She introduced the Karlskrona Manifesto on
+Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
+of related work on the topic of sustainability into a set of (mis-)perceptions
+and principles. The manifesto does not proclaim that there is an easy,
 one-size-fits-all solution around the corner, but rather points out that
@@
 
-\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew D. Jones,
+\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
 and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
-to Improve Productivity of Legacy Tools}} Abani Patra presented an example of how
-the community using Titan2D, a Geoflow Simulation Software, increased the
+to Improve Productivity of Legacy Tools}} Abani Patra presented an example of
+how the community using Titan2D, a Geoflow Simulation Software, increased the
 productivity of their tools by improving both code and data layout. The main
@@
 developers included reinforcing the code structure using multiple layers of
-Python and C++ interfaces, and a redesign of the data layout to be more suitable for
-modern CPUs and accelerators.
+Python and C++ interfaces, and a redesign of the data layout to be more suitable
+for modern CPUs and accelerators.
 
 \item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
-Thaney: \textit{Collaborative Software Development as Sustainable Software: Lessons
-from Open Source}}
-Abigail Cabunoc Mayes combines two properties of open source that, together,
-create a suitable habitat for sustainable software. The first of the two
-properties, public, does not only mean public code. It also includes
-public discussions, a public process of including contributions, and an open
-license. The second property, participatory, stresses the importance of
+Thaney: \textit{Collaborative Software Development as Sustainable Software:
+Lessons from Open Source}} Abigail Cabunoc Mayes combines two properties of open
+source that, together, create a suitable habitat for sustainable software. The
+first of the two properties, public, does not only mean public code. It also
+includes public discussions, a public process of including contributions, and an
+open license. The second property, participatory, stresses the importance of
 reaching out to the community and helping potential new members by providing
@@
 Like others, their main goals include high usability, sustainability, and
-reproducibility. In order to achieve these goals, communication channels have been
-developed, such as mailing lists, wikis, workshops, hackathons, tutorials, and
-webinars. In order to contribute to the infrastructure, codes have to adhere to
-specified minimum standards, but ideals %\choinote{ideals or ideas?} \katznote{ideals}
-are also given to encourage reaching the
-aforementioned goals. These include, e.g., the use of version control, certain
-coding styles, the presence and nature of code tests and documentation, and
-certain user workflows.
+reproducibility. In order to achieve these goals, communication channels have
+been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
+and webinars. In order to contribute to the infrastructure, codes have to adhere
+to specified minimum standards, but ideals are also given to encourage reaching
+the aforementioned goals. These include, e.g., the use of version control,
+certain coding styles, the presence and nature of code tests and documentation,
+and certain user workflows.
 
@@
 Hwang mentioned various ways to cite efforts within the framework, including
-science papers, code papers, user manuals, and the CIG website. An analysis of the
-resulting citations showed that $80\%$ of papers that use CIG codes mention the
-code name, and about the same number includes a citation. Only about $20\%$
+science papers, code papers, user manuals, and the CIG website. An analysis of
+the resulting citations showed that $80\%$ of papers that use CIG codes mention
+the code name, and about the same number includes a citation. Only about $20\%$
 acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
 cite codes (including non-CIG codes), and only about one eighth specify the
-version used. Compared to other codes, CIG seems to be much better
-cited. This, in part, is attributed to suggested citation forms communicated to
-the community, and runtime tools that help collecting a relevant citation list
+version used. Compared to other codes, CIG seems to be much better cited. This,
+in part, is attributed to suggested citation forms communicated to the
+community, and runtime tools that help collecting a relevant citation list
 automatically.
@@
 Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
-located at Notre Dame and Chicago, that links the high energy physics effort to biology,
-astrophysics, digital curation, and other disciplines. It includes physicists,
-digital librarians, as well as computer scientists, and aims to achieve some
-commonality across disciplines. Examples are meta-data descriptions of archived
-data, computational descriptions, descriptions of how data was processed, 
-questions such as whether computation replication can be
-automated, and what the impact of access policies on the preservation
-infrastructure is. One of the products of this effort is a suite of tools that
-deals with this preservation, and the questions was posed whether that software
-itself is sustainable. Points that were brought up included the need for a user
-community depending on a given software, and the need to provide added value for
-its users. An important method to achieve this was to work with the user
-community from the start, and to budget that way. For the specific example of
-the preservation software, this means that besides adding value to the
-community, it needs to be transparent to their workflows, i.e., not requiring
-additional effort to preserve ``research objects''.
+located at Notre Dame and Chicago, that links the high energy physics effort to
+biology, astrophysics, digital curation, and other disciplines. It includes
+physicists, digital librarians, as well as computer scientists, and aims to
+achieve some commonality across disciplines. Examples are meta-data descriptions
+of archived data, computational descriptions, descriptions of how data was
+processed, questions such as whether computation replication can be automated,
+and what the impact of access policies on the preservation infrastructure is.
+One of the products of this effort is a suite of tools that deals with this
+preservation, and the questions was posed whether that software itself is
+sustainable. Points that were brought up included the need for a user community
+depending on a given software, and the need to provide added value for its
+users. An important method to achieve this was to work with the user community
+from the start, and to budget that way. For the specific example of the
+preservation software, this means that besides adding value to the community, it
+needs to be transparent to their workflows, i.e., not requiring additional
+effort to preserve ``research objects''.
 
@@
 the possibility of training in reproducible computational research, providing
-collaborations for researchers who don't want to be programmers, and
-synergies with research computing platforms. The success of such a group would
-be measured by the output and quality of the research software, and could be
-part of Research Computing or faculty. Members would not be independent
-researchers, but would have to have a research background. An attempt to form a
-community of RSE groups within the UK has been underway for several years,
-including funding from the UK Research Council. An open question is whether this
-approach can be adopted in other countries, including the USA.
+collaborations for researchers who don't want to be programmers, and synergies
+with research computing platforms. The success of such a group would be measured
+by the output and quality of the research software, and could be part of
+Research Computing or faculty. Members would not be independent researchers, but
+would have to have a research background. An attempt to form a community of RSE
+groups within the UK has been underway for several years, including funding from
+the UK Research Council. An open question is whether this approach can be
+adopted in other countries, including the USA.
 
@@
 reaching release through a design, development, and testing phase. The main
-deficiency with this approach was pointed out to be the too-late
-interaction with users. Instead, an alternative approach was proposed that, at
-first, skips the development phase and repeatedly goes through requirement
-gathering, design and user interaction/learning phases, and only eventually
-starts development once an agreement is reached, leading to an increased user
-satisfaction, higher adoption, and eventually to sustained software.
+deficiency with this approach was pointed out to be the too-late interaction
+with users. Instead, an alternative approach was proposed that, at first, skips
+the development phase and repeatedly goes through requirement gathering, design
+and user interaction/learning phases, and only eventually starts development
+once an agreement is reached, leading to an increased user satisfaction, higher
+adoption, and eventually to sustained software.
 
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and Sergei
-Maslov: \textit{Challenges of a Sustainable Software Platform for Predictive
-Biology: Lessons Learned on the KBase Project}} In this talk, Dan Gunter
-presented experiences and lessons learned as part of KBase, an open software and
-data platform for addressing the grand challenge of systems biology: predicting
-and designing biological function. KBase is a unified system that integrates
-data and analytical tools for comparative functional genomics of microbes,
-plants, and their communities. However, it is also a collaborative environment
-for sharing methods and results, and placing those results in the context of
-knowledge in the field. Being a large, multi-institutional project, one of the
-big challenges is to agree on standard to enable a single, maintainable system.
-Working in isolation does not work (anymore) within this field, and the community
-in the field also
-didn't have standards for software engineering. This is contrasted to computer
-science research, where these approaches shorten design cycles, leading to more
-time for highly rewarded activities like publishing, performance studies,
-graduating students, or protecting ideas before publication. Sustainability is
-not of big concern there, compared to these. \katznote{prev sentence is unclear}
-Instead of this more traditional approach, KBase uses the following for large,
-distributed teams. After picking projects and team members, 4-5 teams work on
-projects about two weeks before a 1-week evaluation by the executive committee,
-and after which new teams and projects might be chosen to restart the ``agile''
-development cycle. This process is intentionally open and documented.
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and
+Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for
+Predictive Biology: Lessons Learned on the KBase Project}} In this talk, Dan
+Gunter presented experiences and lessons learned as part of KBase, an open
+software and data platform for addressing the grand challenge of systems
+biology: predicting and designing biological function. KBase is a unified system
+that integrates data and analytical tools for comparative functional genomics of
+microbes, plants, and their communities. However, it is also a collaborative
+environment for sharing methods and results, and placing those results in the
+context of knowledge in the field. Being a large, multi-institutional project,
+one of the big challenges is to agree on standard to enable a single,
+maintainable system. Working in isolation does not work (anymore) within this
+field, and the community in the field also didn't have standards for software
+engineering. This is contrasted to computer science research, where these
+approaches shorten design cycles, leading to more time for highly rewarded
+activities like publishing, performance studies, graduating students, or
+protecting ideas before publication. Sustainability is not of big concern there,
+compared to these. \katznote{prev sentence is unclear} Instead of this more
+traditional approach, KBase uses the following for large, distributed teams.
+After picking projects and team members, 4-5 teams work on projects about two
+weeks before a 1-week evaluation by the executive committee, and after which new
+teams and projects might be chosen to restart the ``agile'' development cycle.
+This process is intentionally open and documented.
 
@@
 Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
-approach to overcome some hurdles in scientific publishing: disconnects
-between experimental data, research software, and publications. A current effort
-in the Geosciences is the ``Geoscience Paper of the Future,'' which includes
-four elements. First, it forms a modern paper including text, data, and pointers
-to supplementary materials. Second, it is reproducible, including data
-processing, workflow, and visualization tools. Third, it is part of open
-science, which includes being publicly available under open licenses, and
-providing meta-data. Last, but not least, it uses digital scholarship elements
-like persistent identifiers for software, data, and authors, and it cites both
-data and software. OntoSoft is a tool to help this effort, providing software
-stewardship for the Geosciences. As part of this initiative, a special issue of
-a journal in Geoscience areas is planned to include only Geoscience papers
-of the future, with submissions open until the end of 2015. In addition,
-training sessions are provided to geoscientists to learn best practices in
-software and data sharing, provenance documentation, and scholarly publication.
+approach to overcome some hurdles in scientific publishing: disconnects between
+experimental data, research software, and publications. A current effort in the
+Geosciences is the ``Geoscience Paper of the Future,'' which includes four
+elements. First, it forms a modern paper including text, data, and pointers to
+supplementary materials. Second, it is reproducible, including data processing,
+workflow, and visualization tools. Third, it is part of open science, which
+includes being publicly available under open licenses, and providing meta-data.
+Last, but not least, it uses digital scholarship elements like persistent
+identifiers for software, data, and authors, and it cites both data and
+software. OntoSoft is a tool to help this effort, providing software stewardship
+for the Geosciences. As part of this initiative, a special issue of a journal in
+Geoscience areas is planned to include only Geoscience papers of the future,
+with submissions open until the end of 2015. In addition, training sessions are
+provided to geoscientists to learn best practices in software and data sharing,
+provenance documentation, and scholarly publication.
 
@@
 framework for creation of metrics for scientific software. This framework would
-improve both the metrics and the software it evaluates, and could also serve as a tool for building
-a community around the idea. With especially the latter idea (building
-a community) in mind, a software ``peer review group'' would be created, representing stakeholders who
-will self-review software created by their respective communities, and will
-concurrently develop metrics. The whole project should be community-governed,
-without a single institution overseeing the activities or infrastructure, with the
-hope to evolving community-generated and adopted standards. The generation of
-metrics would be tied to the actual evaluation of software, creating an
-incentive by improving the evaluated software itself during this process. The
-framework code would provide infrastructure for the creation of metrics and
-evaluation, and forums for generation of software success metrics. It would also
-support code reviews of the evaluated software. An open question is whether it
-is possible to fit the resulting metrics in a common template. So far, this is
-still in a design phase, with a white paper at the CSESSP workshop 2015 and this
-talk, but the WSSSPE workshops are seen as a forum for the community to assemble and
-act, and is planned to be used also in the future to build this community and
-framework.
+improve both the metrics and the software it evaluates, and could also serve as
+a tool for building a community around the idea. With especially the latter idea
+(building a community) in mind, a software ``peer review group'' would be
+created, representing stakeholders who will self-review software created by
+their respective communities, and will concurrently develop metrics. The whole
+project should be community-governed, without a single institution overseeing
+the activities or infrastructure, with the hope to evolving community-generated
+and adopted standards. The generation of metrics would be tied to the actual
+evaluation of software, creating an incentive by improving the evaluated
+software itself during this process. The framework code would provide
+infrastructure for the creation of metrics and evaluation, and forums for
+generation of software success metrics. It would also support code reviews of
+the evaluated software. An open question is whether it is possible to fit the
+resulting metrics in a common template. So far, this is still in a design phase,
+with a white paper at the CSESSP workshop 2015 and this talk, but the WSSSPE
+workshops are seen as a forum for the community to assemble and act, and is
+planned to be used also in the future to build this community and framework.
 

### 2015-11-15 WSSSPE3/report/wg_main_funding_spec_expert.tex
small updates

@@
 activities are intermittent within a research cycle. To sustain this capacity,
-therefore, it is necessary to aggregate work across multiple research
+therefore, it is necessary to aggregate this work across multiple research
 groups.
@@
 sufficient arguments in favor of such a model. However, useful measurement of
-code cleanliness and the resulting productivity gains is an unsolved problem
+code cleanliness, and the resulting productivity gains, is an unsolved problem
 in empirical software engineering. To measure ``what did not go wrong'' because

### 2015-11-15 WSSSPE3/report/wg_main_industry_interaction.tex
Section 5.2 and Section 5.3: No content change
Section 5.3: tiny changes

@@
 
-Most scientific software is produced as a part of grant-funded research projects sponsored by federal governments.
-If we are interested in the sustainability of scientific software then we need to understand what exactly happens when that sponsorship ends.
-More than likely, the project and its resulting software will need to undergo some kind of transition in funding and consequently management.
+Most scientific software is produced as a part of grant-funded research projects
+sponsored by federal governments. If we are interested in the sustainability of
+scientific software, then we need to understand what exactly happens when that
+sponsorship ends. More than likely, the project and its resulting software will
+need to undergo some kind of transition in funding and consequently management.
 
 There are a number of potential funding transitions that may occur:  
+%
+\begin{itemize}
 
-\begin{itemize}
-\item A project could be \textbf{refunded}, and development or maintenance of the software continues as planned.
-\item A project might located a \textbf{new source of funding} in which case the software may be further developed or simply maintained as before. 
-\item The project could transition to a \textbf{community supported model} whereby ownership, maintenance and stewardship of the software becomes similar to peer-production models in open-source (e.g., see Howison~\cite{howison_sustaining_2015}).
-\item The project could receive some form of industry sponsorship in which case ownership of the intellectual property, licensing, maintenance activities, hosting, etc.\ may change significantly. 
+\item A project could be \textbf{refunded}, and development or maintenance of
+the software continue as planned.
+
+\item A project might locate a \textbf{new source of funding} in which case the
+software may be further developed or simply maintained as before.
+
+\item The project could transition to a \textbf{community supported model}
+whereby ownership, maintenance, and stewardship of the software become similar
+to peer-production models in open-source (e.g., see
+Howison~\cite{howison_sustaining_2015}).
+
+\item The project could receive some form of industry sponsorship in which case
+ownership of the intellectual property, licensing, maintenance activities,
+hosting, etc.\ may change significantly.
+
 \end{itemize}
 
-We characterize each of these potential changes in funding as ``transition pathways'' to sustainable software (see similar work by Geels and Schot~\cite{Geels:2007}).
+We characterize each of these potential changes in funding as ``transition
+pathways'' to sustainable software (see similar work by Geels and
+Schot~\cite{Geels:2007}).
 
-At WSSSPE3 our group was interested in better understanding successful pathways for scientific software to ``transition'' from government-funded research projects to industry sponsorship. 
-(This may be an initially awkward phrase---some software projects will begin their life being sponsored by industry, or result in collaboration between industry and academia. 
-In such cases, there is still a need to understand how IP and how maintenance of the software is sustained over time)
+At WSSSPE3 our group was interested in better understanding successful pathways
+for scientific software to ``transition'' from government-funded research
+projects to industry sponsorship. (This may be an initially awkward
+phrase---some software projects will begin their life being sponsored by
+industry, or result in collaboration between industry and academia. In such
+cases, there is still a need to understand how IP and how maintenance of the
+software is sustained over time.)
 
-Although transitions are often studied under the broad umbrella of `technology transfer' we believe there are likely to be a number of different ways in which a pathway from initial production to long-term maintenance and secure funding is achieved.
-In short, industry sponsorship is an important aspect of sustaining scientific software, but our current understanding of these transitions focuses narrowly on commercial successes\slash failures.
+Although transitions are often studied under the broad umbrella of ``technology
+transfer'', we believe there are likely to be a number of different ways in which
+a pathway from initial production to long-term maintenance and secure funding are
+achieved. In short, industry sponsorship is an important aspect of sustaining
+scientific software, but our current understanding of these transitions focuses
+narrowly on commercial successes\slash failures.
 
@@
 
-In looking at existing literature that addresses industry transitions many reports focus on benefits that accrue to the private sector, or to a government that originally sponsored the research project.
-This literature does not address the impact that these transitions have on the accessibility, or usability of the software, or the impact that these transitions have on the career of the researchers involved.
+In looking at existing literature that addresses industry transitions, many
+reports focus on benefits that accrue to the private sector, or to a government
+that originally sponsored the research project. This literature does not address
+the impact that these transitions have on the accessibility or usability of the
+software, or the impact that these transitions have on the career of the
+researchers involved.
 
-Examples of the former scenario (benefits accruing to private sector) are as follows:
+Examples of the former scenario (benefits accruing to private sector) are as
+follows:
 \begin{itemize}
+
 \item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
+
 \item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
-\item Dowling Review from the UK: addresses complexity of work between these two communities: \url{http://www.raeng.org.uk/policy/dowling-review}
-\item Pathway to Impact - UK report: two pages of grant proposals are asked to forecast how they might impact (includes environmental, academic, economic)
+
+\item Dowling Review from the UK: addresses complexity of work between these two
+communities: \url{http://www.raeng.org.uk/policy/dowling-review}
+
+\item Pathway to Impact - UK report: two pages of grant proposals are asked to
+forecast how they might impact (includes environmental, academic, economic) \choinote{say impact on what?}.
+
 \end{itemize}
@@
 
-Our work at WSSSPE3 included the following activities (described in detail below): (1) brainstorming goals for this type of research, (2) imagining potential outcomes of completing a set of case studies on this topic, and (3) generating a set of working definitions for some of the broad concepts we are describing.
+Our work at WSSSPE3 included the following activities (described in detail
+below): (1) brainstorming goals for this type of research, (2) imagining
+potential outcomes of completing a set of case studies on this topic, and (3)
+generating a set of working definitions for some of the broad concepts we are
+describing.
 
-\textbf{GOAL}
+\textbf{Goal}
+
 \emph{What is the goal of doing research on transition pathways?}
@@
 
-How could we design a study that focused on the impact of the software in undergoing this type of transition?
+How could we design a study that focused on the impact of the software in
+undergoing this type of transition?
 
@@
 
-Complete a set of case studies which look at successful and unsuccessful transitions between academic researchers and industry.
-This might address one of each of the transition types (described below).
-Successful transitions are described as those that lead to either weak or strong sustainability (also defined below).
+Complete a set of case studies which look at successful and unsuccessful
+transitions of researchers between academia and industry. This might address 
+each of the transition types (described below). Successful transitions are
+described as those that lead to either weak or strong sustainability (also
+defined below).
 
-Create a generalizable framework that might allow for the study of different transition pathways (other than academia -> industry).
+Create a generalizable framework that might allow for the study of different
+transition pathways (other than academia to industry).
 
@@
 \begin{itemize}
-\item Handoff model: academia initially writes the software, industry (for profit or nonprofit) then takes over project. 
-\item Co-Production Model: industry and academia interact throughout development of the project.
-\item Sponsorship Model: academia writes and maintains the software, the industry contributes funding for the development\slash maintenance of software.
-In this example industry is also likely a user of the software.
-\item Spinoff model: transition to for- or non-profit company owned by or in collaboration with original developers. 
+
+\item Handoff model: academia initially writes the software, industry (for-profit 
+or nonprofit) then takes over the project.
+
+\item Co-Production Model: industry and academia interact throughout development
+of the project.
+
+\item Sponsorship Model: academia writes and maintains the software; the
+industry contributes funding for the development\slash maintenance of software.
+In this example, industry is also likely a user of the software.
+
+\item Spinoff model: transition to for- or non-profit company owned by or in
+collaboration with original developers.
+
 \end{itemize}
@@
 \begin{itemize}
-\item Weak Sustainability: Software continues to be accessible, useful, and usable. 
-\item Strong Sustainability: Software meets criteria above, but is also able to be reused for further innovation (i.e., issued non-restrictive open-source license).
+
+\item Weak Sustainability: Software continues to be accessible, useful, and
+usable.
+
+\item Strong Sustainability: Software meets criteria above, but is also able to
+be reused for further innovation (i.e., issued non-restrictive open-source
+license).
+
 \end{itemize}
-See Becker et al.~\cite{Becker:2014} for extended discussion of weak vs.\ strong sustainability.
-For more detail on the group's discussion, see Appendix~\ref{sec:appendix_industry_interaction}.
+We refer readers to Becker et al.~\cite{Becker:2014} for extended discussion of weak versus strong
+sustainability. For more detail on the group's discussion, see
+Appendix~\ref{sec:appendix_industry_interaction}.
 
@@
 
-Plans for carrying forward are currently unclear---this project would require sustained attention and effort from our team, and at least some amount of funding in order for us to be involved for extended periods of time.
+Plans for carrying forward are currently unclear---this project would require
+sustained attention and effort from our team, and at least some amount of
+funding in order for us to be involved for extended periods of time.
 
@@
 
-To find more information about this group, or join it, see \todo{where should someone go who want to know more about this and perhaps wants to contribute?}
+To find more information about this group, or join it, see \todo{where should
+someone go who want to know more about this and perhaps wants to contribute?}

### 2015-11-15 WSSSPE3/report/wg_main_industry_interaction.tex
very tiny changes

@@
 Although transitions are often studied under the broad umbrella of ``technology
-transfer'', we believe there are likely to be a number of different ways in which
+transfer,'' we believe there are likely to be a number of different ways in which
 a pathway from initial production to long-term maintenance and secure funding are
@@
 \item Pathway to Impact - UK report: two pages of grant proposals are asked to
-forecast how they might impact (includes environmental, academic, economic) \choinote{say impact on what?}.
+forecast what impact they might have (including environmental, academic, economic).
 

### 2015-11-16 WSSSPE3/report/wg_main_industry_interaction.tex
Section 5.3: Added a question and request for review.
small edits to edits and other in 5.3

@@
 Can we determine what each partner wants to get out of such a collaboration?
-Why would industry be interested in collaborating with academia? 
-Why would academia be interested in collaborating with industry?
+For example, why would industry be interested in collaborating with academia? 
+Or why would academia be interested in collaborating with industry?
 
@@
 
-Complete a set of case studies which look at successful and unsuccessful
+A set of case studies that look at successful and unsuccessful
 transitions of researchers between academia and industry. This might address 
@@
 
-\item Sponsorship Model: academia writes and maintains the software; the
+\item Sponsorship Model: academia writes and maintains the software; 
 industry contributes funding for the development\slash maintenance of software.
@@
 
-\item Spinoff model: transition to for- or non-profit company owned by or in
+\item Spinoff model: transition to a for-profit or non-profit company owned by or in
 collaboration with original developers.
@@
 \end{itemize}
-We refer readers to Becker et al.~\cite{Becker:2014} for extended discussion of weak versus strong
+We refer readers to Becker et al.~\cite{Becker:2014} for an extended discussion of weak versus strong
 sustainability. For more detail on the group's discussion, see

### 2015-11-16 WSSSPE3/report/wg_main_user_community.tex
adding cite for mozilla working open guide

@@
 
-Mozilla Science maintains a ``Working Open Project Guide'' (\url{http://mozillascience.github.io/leadership-training}), the introduction of which states: \katznote{should get changed to a cite}
+Mozilla Science maintains a ``Working Open Project Guide''~\cite{working-open-wssspe3}, the introduction of which states:
 \begin{quote}

### 2015-11-17 WSSSPE3/report/WSSSPE3_review.tex
No content change.
Section 6: microscopic changes.

@@
 
-In WSSSPE3, we attempted to take what we learned from WSSSPE1 and WSSSPE2
-in how we can collaboratively build a workshop agenda and turn that into
-an ongoing community activity.  The success or failure of this effort
-of this will only become apparent over time.
+In WSSSPE3, we attempted to take what we learned from WSSSPE1 and WSSSPE2 in how
+we can collaboratively build a workshop agenda and turn that into an ongoing
+community activity. The success or failure of this effort will only
+become apparent over time.
 
-The workshop had two components, presentations and working groups.  The presentations,
-in the first half day of the workshop, including a inspirational keynote and a set of lightning
-talks.
-We used lightning talks for two reasons: first, the need of some participants to have a slot
-on the agenda to justify their attendance, and second, as a way to get new ideas across
-to all the attendees.  We broke with the tradition of requiring the lightning talk submitter
-to self-publish their papers, and instead used a common peer-review platform\footnote{\url{http://easychair.org}}, choosing to publish the speaker slides on the workshop web site
-instead.
+The workshop had two components, presentations and working groups. The
+presentations, in the first half day of the workshop, included an inspirational
+keynote and a set of lightning talks. We used lightning talks for two reasons:
+first, the need of some participants to have a slot on the agenda to justify
+their attendance; and second, as a way to get new ideas across to all the
+attendees. We broke with the tradition of requiring the lightning talk submitters
+to self-publish their papers, and instead used a common peer-review
+platform\footnote{\url{http://easychair.org}}, choosing to publish their
+slides on the workshop website instead.
 
-The working groups met for a small part of the first half day and all of the second day,
-excepting some times for the groups to report back to the collected workshop attendees.
-\todo{need to say some more about what came of this}
+The working groups met for a small part of the first half day and all of the
+second day, excepting some times for the groups to report back to the collected
+workshop attendees. \todo{need to say some more about what came of this}
 
-\todo{this paragraph is from the wssspe2 report - to be updated...}
-However, the challenge that we have discovered
-since WSSSPE2 is that it is very hard to continue the breakout groups'
-activities.  The WSSSPE2 participants were willing to dedicate their time to
-the groups while they were at the meeting, but afterwards, they have gone
-back to their (paid) jobs.  We need to determine how to tie the WSSSPE
-breakout activities to people's jobs, so that they feel that continuing them
-is a higher priority than it is now, perhaps through funding the participants,
-or through funding coordinators for each activity, or perhaps by getting
-the workshop participants to agree to a specific schedule of activities during the
-workshop.
-
-%\begin{table*}[t]
-%\centering
-%\caption{Top tweets tagged \#WSSSPE or \#WSSSPE3 on Sep.\ 28--29, 2015.}\label{tab:tweets}
-%  \begin{scriptsize}
-%  \begin{tabular}{@{}l l l l@{}}
-% \toprule
-%    Author  &   Tweet  & Retweets &  Favorites
-%\\ \midrule
-%Neil P Chue Hong  &  Getting ready for the start of the \#WSSSPE workshop & 2 & 1
-%\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}  & &
-%\\ & Video stream: \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & &
-%%
-%\\ August Muench & Astronomy \& Software for a totally distracting Monday (2/2) => & 1 & 3
-%\\ & Workshop on Sustainable SW for Science \#WSSSPE3 & & 
-%\\ & \url{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/} & &
-%%
-%\\ Neil P Chue Hong & My \#WSSSPE lightning talk on Building a Scientific Software & 6 & 3
-%\\ &  Accreditation Framework: \url{http://dx.doi.org/10.6084/m9.figshare.1555925} & & 
-%%
-%\\ Mozilla Science Lab & At \#wssspe3 and want to learn more about our Working Open guide & 2 & 1 
-%\\ &  or Contributorship badges pilot? Keep an eye out for @abbycabs! & & 
-%%
-%\\ Daniel S.\ Katz & \#wssspe starting! & 1 & 3
-%%
-%\\ Daniel S.\ Katz & @powersoffour starting \#wssspe keynote & 1 & 2
-%%
-%\\ Neil P Chue Hong   &  The difference between software presented at SC \& SciPy & 1 & 4
-%\\  &  according to @powersoffour straw person \#wssspe & &
-%%
-%\\ Daniel S.\ Katz & sustainability may mean: keeps up with bug reports, adds new features, & 3 & 1
-%\\ &  works on new architectures (1/2)  -- @powersoffour at \#wssspe (1/2) & & 
-%%
-%\\ Daniel S.\ Katz & sustainability may mean: people use it, keeps getting funded, produces & 2 & 
-%\\ &  same results, transits over people (2/2)  -- @powersoffour at \#wssspe & &
-%%
-%\\ Nic Weber & Rec, by Matt Turk: Ash Dryden - Ethics of Unpaid Labor & 3 & 2 
-%\\ & \url{http://t.co/KfMwS4VG4N} & &
-%%
-%\\ Nic Weber & My .02 - many more models (than 3) for sustaining sci. software. & 3 & 2
-%\\ & See this pub on open-source:  & & 
-%\\ & \url{http://papers.ssrn.com/sol3/papers.cfm?abstract_id=2568185} & &
-%%
-%\\ Erin Robinson & \#WSSSPE - Software is like a vampire. It's only dead if you really kill it. & 2 & 2
-%%
-%\\ Neil P Chue Hong & Absolutely great \#wssspe keynote from @powersoffour to inspire & 2 & 1 
-%\\ & the workshop. Should be archived via & & 
-%\\ & \url{http://ucarconnect.ucar.edu/live?room=cg1aud#.VgmpvCdIrCQ} soon & & 
-%%
-%\\ Daniel S.\ Katz & \#wssspe lightning talks starting - streaming at  & 5  &  1
-%\\ &  \url{http://ucarconnect.ucar.edu/live?room=cg1aud} & & 
-%%
-%\\ Kyle Niemeyer & Participatory open source means better documentation  & 2 & 1 
-%\\ & (readme, roadmap, contributing) ? @abbycabs at \#wssspe & &
-%%
-%\\ Erin Robinson & Interesting resource: \url{http://mozillascience.github.io/leadership-training/} & 2 & 
-%\\ &  \#WSSSPE cc: @abbycabs & & 
-%%
-%\\ Nic Weber & @MozillaScience + @abbycabs giving look at cool resources & 2 & 3
-%\\ & for leading an open source community & & 
-%\\ & \url{http://mozillascience.github.io/leadership-training/} & & 
-%%
-%\\ Nic Weber & Some interesting sepeartion of best practices for Geosoftware dev & 2 & 
-%\\ & into categories \url{https://geodynamics.org/cig/dev/best-practices} & &
-%%
-%\\ Neil P Chue Hong & Some statistics of software citation in geodynamics & 2 & 
-%%
-%\\ Daniel S.\ Katz & @jamespjh goal: make research software less rubbish; solution is RSEs & 3 & 
-%%
-%\\ Nic Weber & Highly relevant to \#wssspe crowd - @Impactstory team is launching & 7 & 3
-%\\ & a impact of research software project \url{http://depsy.org/}  \#verycool & & 
-%%
-%\\ Neil P Chue Hong & @jamespjh talking about Research Software Engineer groups in the UK & 2 & 1
-%\\ &  \#wssspe @ResearchSoftEng @SoftwareSaved & & 
-%%
-%\\ Abby Cabunoc Mayes & Wonderful time @ \#WSSSPE 3! Encouraging 2 see a diverse community & 1 & 2 
-%\\ & working 2gether (\& making real plans!) \url{http://gif.co/s6v4.gif} & & 
-%%
-%\\ \bottomrule
-%    \end{tabular}
-%    \end{scriptsize}
-%\end{table*}
-%
-%WSSSPE actively used the online social network Twitter, with hashtag
-%``\#WSSSPE''. There were substantially more tweets (messages) during the days of
-%the workshops WSSSPE2, WSSSPE1.1, and WSSSPE1. Out of about 670 tweets as of Apr
-%18, 2015, more than 225 were about WSSSPE2 and about 180 were posted during the
-%day of the workshop. Some of the main points and highlights in the meeting are
-%shown in Table~\ref{tab:tweets}, which summarizes the top \#WSSSPE tweets from
-%the day of workshop, selected by the metrics that number of retweets or
-%favorites larger than five and the sum of two measures greater than ten.
+\todo{this paragraph is from the wssspe2 report - to be updated...} However, the
+challenge that we have discovered since WSSSPE2 is that it is very hard to
+continue the breakout groups' activities. The WSSSPE2 participants were willing
+to dedicate their time to the groups while they were at the meeting, but
+afterwards, they went back to their (paid) jobs. We need to determine how
+to tie the WSSSPE breakout activities to people's jobs, so that they feel that
+continuing them is a higher priority than it is now, perhaps through funding the
+participants, or through funding coordinators for each activity, or perhaps by
+getting the workshop participants to agree to a specific schedule of activities
+during the workshop.
 

### 2015-11-17 WSSSPE3/report/wg_main_user_community.tex
Section 5.10: No content change.
Section 5.10: itsy-bitsy mutations

@@
 
-User communities are the lifeblood of sustainable scientific software. The user community includes the developers, 
-both internal and external, of the software; direct users of the software; other software projects that depend on
-the software; and any other groups that create or consume data that is specific to the software. Together these
-groups provide both the reason for sustaining the software and, collectively, the requirements that drive its continued
-evolution and improvement.
+User communities are the lifeblood of sustainable scientific software. The user
+community includes the developers, both internal and external, of the software;
+direct users of the software; other software projects that depend on the
+software; and any other groups that create or consume data that is specific to
+the software. Together these groups provide both the reason for sustaining the
+software and, collectively, the requirements that drive its continued evolution
+and improvement.
 
@@
 
-Mozilla Science maintains a ``Working Open Project Guide''~\cite{working-open-wssspe3}, the introduction of which states:
+Mozilla Science maintains a ``Working Open Project
+Guide''~\cite{working-open-wssspe3}, the introduction of which states:
 \begin{quote}
-Working openly with contributors enables your
-    community to learn how to build and collaborate together. This
-    document is a guideline on how to work openly and involve others
-    in your projects with Mozilla. We want to help you engage your
-    community in a way that encourages contributors and builds other
-    leaders.
+Working openly with contributors enables your community to learn how to build
+and collaborate together. This document is a guideline on how to work openly and
+involve others in your projects with Mozilla. We want to help you engage your
+community in a way that encourages contributors and builds other leaders.
  \end{quote}
 
-Several books have been written about software communities:
+Several books have been written about software communities: \choinote{only 1 book below}
 \begin{itemize}
-\item ``Art of Community'' by Jono Bacon~\cite{art-of-community}. We could consider distilling this for scientific software.
-\item Iain Larmour, from EPSRC in the UK (EPSRC: https://www.epsrc.ac.uk/) (not sure who from mentioned UK Collaborative Computational Projects (CCP: http://ccp.ac.uk)
+
+\item ``Art of Community'' by Jono Bacon~\cite{art-of-community}. We could
+consider distilling this for scientific software.
+
+\item Iain Larmour, from EPSRC in the UK (EPSRC: https://www.epsrc.ac.uk/) (not
+sure who from mentioned UK Collaborative Computational Projects (CCP:
+http://ccp.ac.uk) \choinote{Who is ``not sure who'' supposed to be?}
+
 \end{itemize}
@@
 
-Discussion revolved around a few questions: what is the benefit of having a ``community'' for software sustainability, what
-practices and circumstances lead to having a community, how can funding help or hinder this process, and perhaps most
-importantly how can best practices be described and distilled into a document that can help new projects.
+Discussion revolved around a few questions: what is the benefit of having a
+``community'' for software sustainability, what practices and circumstances lead
+to having a community, how can funding help or hinder this process, and perhaps
+most importantly, how can best practices be described and distilled into a
+document that can help new projects.
 
-The benefits of having a community that were brought up were considered largely obvious. In addition to having advocates for
-the software, and a possible source of ``free'' contributions to the codebase, the community becomes a good source for
-requirements, feedback, and metrics. The software community can also act as ``cheerleaders'' who convince funders or other
-potential users to fund/use the software, and thus help sustain the software.
+The benefits of having a community that were brought up were considered largely
+obvious. In addition to having advocates for the software, and a possible source
+of ``free'' contributions to the codebase, the community becomes a good source
+for requirements, feedback, and metrics. The software community can also act as
+``cheerleaders'' who convince funders or other potential users to fund/use the
+software, and thus help sustain the software.
 
-Practices and circumstances that lead to a community are first, that the software offers value. But in addition to this, a
-community will be much more likely to form if they receive (expert) support when they have questions. Additional contributing
-factors are good usability (not always needed), and an open development process such as IPython developer meetings on YouTube.
-It was also pointed out that an evangelist for the project, not necessarily but often one of the developers, can often make a
-big difference.
+Practices and circumstances that lead to a community are first, that the
+software offers value. But in addition to this, a community will be much more
+likely to form if they receive (expert) support when they have questions.
+Additional contributing factors are good usability (not always needed), and an
+open development process such as IPython developer meetings on YouTube. It was
+also pointed out that an evangelist for the project, not necessarily but often
+one of the developers, can often make a big difference.
 
-Funding can help the process by encouraging both value to the community and high-quality user support. Only providing funding
-for the software development may create good software, but with less likelihood to have a real community. It was discussed
-that federal laboratories are a good incubator for software communities, and that a general facility like EarthCube is too
-dispersed to really make a community. Also, domain-specific groups within laboratories or universities might provide as an
-incubator for software communities.
+Funding can help the process by encouraging both value to the community and
+high-quality user support. Only providing funding for the software development
+may create good software, but with less likelihood to have a real community. It
+was discussed that federal laboratories are a good incubator for software
+communities, and that a general facility like EarthCube is too dispersed to
+really make a community. Also, domain-specific groups within laboratories or
+universities might provide as an incubator for software communities.
 
-In describing best practices, the group discussed the different modes for starting a scientific software project: building on
-an existing product that needs improving, recognizing an unsatisfied need of an existing community, or creating a new solution
-to a need not yet recognized by the community. The group also thought that the existing books on software communities would
-need to be evaluated in light of differences between Science Software projects and general OSS projects in terms of scale,
-science, acknowledgement and credit, and funding models.
+In describing best practices, the group discussed the different modes for
+starting a scientific software project: building on an existing product that
+needs improving, recognizing an unsatisfied need of an existing community, or
+creating a new solution to a need not yet recognized by the community. The group
+also thought that the existing books on software communities would need to be
+evaluated in light of differences between Science Software projects and general
+OSS projects in terms of scale, science, acknowledgement and credit, and funding
+models.
 
-Additional details on the group's discussion can be found in Appendix~\ref{sec:appendix_user_community}.
+Additional details on the group's discussion can be found in
+Appendix~\ref{sec:appendix_user_community}.
 
@@
 
-The most important next steps is a ``Best Practice'' document, which would describe what successful projects with engaged
-communities look like, how to replicate this type of project, and look at end-of-life on a community project. Inputs to this
-document would include a software community survey of high functioning communities such as R Open Science, Python SciPy,
-OPeNDAP, and Unidata, with analysis of factors that feed into their success. Also references lik the ``Art of Community'' could
-be adapted and summarized for the science software community.
+The most important next steps is a ``Best Practice'' document, which would
+describe what successful projects with engaged communities look like, how to
+replicate this type of project, and look at end-of-life on a community project.
+Inputs to this document would include a software community survey of highly
+functioning communities such as R Open Science, Python SciPy, OPeNDAP, and
+Unidata, with analysis of factors that feed into their success. Also references
+lik the ``Art of Community'' could be adapted and summarized for the science
+software community.
 
-Another next step would be increasing recognition of need for science software projects to focus on building and supporting
-their user community. Good Software Engineering practices are not enough, and popular training like Software Carpentry does
-not currently address this issue head on.
+Another next step would be increasing recognition of need for science software
+projects to focus on building and supporting their user communities. Good software
+engineering practices are not enough, and popular training like Software
+Carpentry does not currently address this issue head on.
 

### 2015-11-17 WSSSPE3/report/wg_main_publishing_SW.tex
minor changes/comments in publishing sections moved some credit text to be a general intro to the section
minor change to start of publishing section
Section 5.9: No content change.
Section 5.9:  teeny-weeny transformation

@@
 \subsection{Publishing Software Working Group Discussion} \label{sec:publishing-software}
-
-\subsubsection{Why it is important}
 
@@
 includes the code needed to produce their own results), and other forms of
-publishing which include dynamic electronic content. Transitioning to this type
-of publication offers possibilities of addressing, or partially addressing
-sustainability concerns such as reproducibility, software credit, and best
-practices.
+publishing which include dynamic electronic content.
+%
+%\subsubsection{Why it is important}
+%
+Transitioning to this type of publication offers possibilities of addressing, or
+partially addressing, sustainability concerns such as reproducibility, software
+credit, and best practices.
 
@@
 \textbf{Reproducibility}: Part of the purpose of these executable paper venues
-is to (at least partially) address the reproducibility issue by making the
-paper itself recompute its own results.
+is to (at least partially) address the reproducibility issue by making a paper
+itself recompute its own results.
 
@@
 provided. In addition, these notebooks make it possible to provide/define
-additional metadata to make the tracing of credit clearer. Finally,
-attributions could be added to citations to identify whether a paper extends a
-result, verifies it, contradicts it, etc.
+additional metadata to make the tracing of credit clearer. Finally, attributions
+could be added to citations to identify whether a paper extends a result,
+verifies it, contradicts it, etc.
 
-\textbf{Best Practices (\S\ref{sec:best-practices})}: Because an executable paper
-showcases the code, and the code itself is subject to the review process,
+\textbf{Best Practices (\S\ref{sec:best-practices})}: Because an executable
+paper showcases the code, and the code itself is subject to the review process,
 authors are more likely to pay attention to coding practices. In addition,
 because the paper must explain what the code does, better documentation is
-likely to be achieved.
+more likely to be achieved.
 
@@
 The group felt that the best way to encourage the use of these new publishing
-concepts would be to create and curate a list identifying publishing venues
-that support them. The Software Sustainability Institute agreed to host this list.
+concepts would be to create and curate a list identifying publishing venues that
+support them. The Software Sustainability Institute agreed to host this list.
+
+See Appendix~\ref{sec:appendix_publishing_SW} for more details about the
+discussion.
 
@@
 
-The plan is to create a web page describing executable papers, their value, and
-a list of what publishers support them.
+The plan is to create a web page describing executable papers, their values, and
+a list of what publishers support them. \katznote{when?}
 
@@
 
-The aforementioned page to be published on SSI.
+The aforementioned page will be published on the Software Sustainability
+Institute website: \url{http://www.software.ac.uk}.

### 2015-11-17 WSSSPE3/report/WSSSPE3_review.tex
Appendices A-C:  teensy-weensy revision

@@
 %\todo{Do we want email addresses here?}
-The following is list of participants registered for the WSSSPE3 workshop.
+The following is a list of participants registered for the WSSSPE3 workshop.
 
@@
 %\todo{Do we want email addresses here?}
-The following is list of travel award recipients for the WSSSPE3 workshop.
+The following is the list of travel award recipients for the WSSSPE3 workshop.
 

### 2015-11-18 WSSSPE3/report/wg_main_publishing_SW.tex
small edits

@@
 \textbf{Reproducibility}: Part of the purpose of these executable paper venues
-is to (at least partially) address the reproducibility issue by making a paper
-itself recompute its own results.
+is to (at least partially) address the reproducibility issue by making paper
+recompute their own results.
 
@@
 
-The plan is to create a web page describing executable papers, their values, and
+The plan is to create a web page describing executable papers, their value, and
 a list of what publishers support them. \katznote{when?}

### 2015-11-19 WSSSPE3/report/wg_main_metrics.tex
Sections 5.6-5.7: no content change.

@@
 
-Metrics for scientific software are important for tenure and promotion, scientific impact, discovery, reducing duplication, serving as a basis for potential industrial interest in adopting software, prioritizing develop and support towards
-strategic objectives and making a case for new or continued funding.  However, there is no commonly-used 
-standard for collecting or presenting metrics, nor is it known if there is a common set of metrics for scientific software. 
- It is imperative that scientific software stakeholders understand that it is useful to collect metrics.
+Metrics for scientific software are important for tenure and promotion,
+scientific impact, discovery, reducing duplication, serving as a basis for
+potential industrial interest in adopting software, prioritizing develop and
+support towards strategic objectives and making a case for new or continued
+funding. However, there is no commonly-used standard for collecting or
+presenting metrics, nor is it known if there is a common set of metrics for
+scientific software. It is imperative that scientific software stakeholders
+understand that it is useful to collect metrics.
 
@@
 
-The group discussion  focused on identifying existing frameworks and activities for scientific software metrics.  
-The group identified the following related activities:
+The group discussion focused on identifying existing frameworks and activities
+for scientific software metrics. The group identified the following related
+activities:
 
@@
 
-\item
-Computational Infrastructure for Geodynamics: Software Development Best Practices\footnote{\url{https://geodynamics.org/cig/dev/best-practices/}}
+\item Computational Infrastructure for Geodynamics: Software Development Best
+Practices\footnote{\url{https://geodynamics.org/cig/dev/best-practices/}}
 
-\item
-WSSSPE3 Breakout Session: How can we measure the impact of a code on research, and its value to the community?\footnote{\url{https://docs.google.com/document/d/1cgUDH3RxrfsLotWhKKOrXUnaYFhrtjcV1TDRkFtwQKI/edit}}
+\item WSSSPE3 Breakout Session: How can we measure the impact of a code on
+research, and its value to the
+community?\footnote{\url{https://docs.google.com/document/d/1cgUDH3RxrfsLotWhKKOrXUnaYFhrtjcV1TDRkFtwQKI/edit}}
 
-\item
-2015 NSF SI2 PI Workshop Breakout Session on Framing Success Metrics\footnote{\url{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/}}
+\item 2015 NSF SI2 PI Workshop Breakout Session on Framing Success
+Metrics\footnote{\url{https://docs.google.com/document/d/10yj7MYEjvrg__t522XR41ogASYMp647-l-BpFTsqEV4/}}
 
-\item
-2015 NSF SI2 PI Workshop Breakout Session on Software Metrics\footnote{\url{https://docs.google.com/document/d/1uDim5bw8rBuubmtaUrz5Eh35NxzDgivmmdXhVzDs3tc/edit}}
+\item 2015 NSF SI2 PI Workshop Breakout Session on Software
+Metrics\footnote{\url{https://docs.google.com/document/d/1uDim5bw8rBuubmtaUrz5Eh35NxzDgivmmdXhVzDs3tc/edit}}
 
-\item
-NSF Workshop on Software and Data Citation Breakout Group on Useful Metrics\footnote{\url{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/}}
+\item NSF Workshop on Software and Data Citation Breakout Group on Useful
+Metrics\footnote{\url{https://docs.google.com/presentation/d/1PPLVL6uoOmisqnHTlwhsVKJBTFFK1IVzvr8FdEEIvAE/}}
 
-\item
-U.K. Software Sustainability Institute Software Evaluation Guide\footnote{\url{http://www.software.ac.uk/software-evaluation-guide}}
+\item U.K. Software Sustainability Institute Software Evaluation
+Guide\footnote{\url{http://www.software.ac.uk/software-evaluation-guide}}
 
-\item
-U.K. Software Sustainability Institute Blog post: The five stars of research software\footnote{\url{http://www.software.ac.uk/blog/2013-04-09-five-stars-research-software}}
+\item U.K. Software Sustainability Institute Blog post: The five stars of
+research
+software\footnote{\url{http://www.software.ac.uk/blog/2013-04-09-five-stars-research-software}}
 
-\item
-Minimal information for reusable scientific software\footnote{\url{http://figshare.com/articles/Minimal_information_for_reusable_scientific_software/1112528}}
+\item Minimal information for reusable scientific
+software\footnote{\url{http://figshare.com/articles/Minimal_information_for_reusable_scientific_software/1112528}}
 
-\item
-EPSRC-funded Equipment Data Search Site\footnote{\url{http://equipment.data.ac.uk/}}
+\item EPSRC-funded Equipment Data Search
+Site\footnote{\url{http://equipment.data.ac.uk/}}
 
-\item
-Canarie Research Software: Software to accelerate discovery\footnote{\url{http://www.canarie.ca/software/}}
+\item Canarie Research Software: Software to accelerate
+discovery\footnote{\url{http://www.canarie.ca/software/}}
 
-\item
-Canarie Research Software: Research Software Platform Registry\footnote{\url{https://science.canarie.ca/researchmiddleware/platforms/list/main.html}}
+\item Canarie Research Software: Research Software Platform
+Registry\footnote{\url{https://science.canarie.ca/researchmiddleware/platforms/list/main.html}}
 
-\item
-BlackDuck Open HUB\footnote{\url{https://www.openhub.net/}}
-
-\item
-Innovation Policy Platform\footnote{\url{https://www.innovationpolicyplatform.org/frontpage}}
+\item BlackDuck Open HUB\footnote{\url{https://www.openhub.net/}}
+\item Innovation Policy
+Platform\footnote{\url{https://www.innovationpolicyplatform.org/frontpage}}
 
@@
 
-The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific 
-software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper. 
- The group discussed many questions related to useful metrics for scientific software including addressing if there is a common set of metrics that 
- can be filtered in some way, can metrics be fit into a common template, which metrics would be the most useful for each stakeholder, 
- which metrics are the most helpful and how would we assess this, how are metrics monitored, and many more.  
- A more complete bulleted list of these questions can be found in Appendix H.  Next, a roadmap for how to proceed was discussed including
- creating a set of milestones and tasks.  The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure
-  for Sustained Innovation (SI2) PI workshop in order to send a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.  
- The five solicitations for software elements released  under the NSF SI2 program all included metrics as a required component with submitters requested to include 
- {\it "a list of tangible metrics, with end user involvement, to be used to measure the success of the software element developed, ...'}.
-These metrics are then reported as part of annual reports to NSF by the projects.
-Although neither the proposal text describing the metrics nor the reported metric results are publicly available, there is reason to believe that the community will be willing to provide this information through a survey mechanism.
-This survey would be created by one of the student group members.  Similarly, it was suggested that a software metrics survey be sent to the 
-UK SFTF and TRDF software projects to ask them what metrics would be useful to report.  The remainder of the discussion focused mainly on the 
-creation of a white paper on this topic.  This resulted in a paper outline and writing assignments with the goal of publishing in venues including 
-WSSSPE4, IEEE CISE, or JORS. More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
+The group discussion began by agreeing on the common purpose of creating a set
+of guidance giving examples of specific metrics for the success of scientific
+software in use, why they were chosen, what they are useful to measure, and any
+challenges and pitfalls; then publish this as a white paper. The group discussed
+many questions related to useful metrics for scientific software including
+addressing if there is a common set of metrics that can be filtered in some way,
+can metrics be fit into a common template, which metrics would be the most
+useful for each stakeholder, which metrics are the most helpful and how would we
+assess this, how are metrics monitored, and many more. A more complete bulleted
+list of these questions can be found in Appendix H. Next, a roadmap for how to
+proceed was discussed including creating a set of milestones and tasks. The idea
+was put forth for the group to interact with the organizing committee of the
+2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in
+order to send a software metrics survey to all SI2 and related awardees as a
+targeted and relevant set of stakeholders. The five solicitations for software
+elements released under the NSF SI2 program all included metrics as a required
+component with submitters requested to include {\it "a list of tangible metrics,
+with end user involvement, to be used to measure the success of the software
+element developed, ...'}. These metrics are then reported as part of annual
+reports to NSF by the projects. Although neither the proposal text describing
+the metrics nor the reported metric results are publicly available, there is
+reason to believe that the community will be willing to provide this information
+through a survey mechanism. This survey would be created by one of the student
+group members. Similarly, it was suggested that a software metrics survey be
+sent to the UK SFTF and TRDF software projects to ask them what metrics would be
+useful to report. The remainder of the discussion focused mainly on the creation
+of a white paper on this topic. This resulted in a paper outline and writing
+assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE,
+or JORS. More information about the group discussion is available in
+Appendix~\ref{sec:appendix_metrics}.
 
@@
 
-The main plans for the group going forward are the creation of a white paper on the topic of useful metrics for scientific software.
-The authoring of this white paper would happen in parallel with the creation of a survey by the group with the survey results to be incorporated in the white paper.
-The timeline for completion of the white paper is approximately one year targeting venues discussed in the previous section.
+The main plans for the group going forward are the creation of a white paper on
+the topic of useful metrics for scientific software. The authoring of this white
+paper would happen in parallel with the creation of a survey by the group with
+the survey results to be incorporated in the white paper. The timeline for
+completion of the white paper is approximately one year targeting venues
+discussed in the previous section.
 
@@
 
-In lieu of a landing page, the Useful Metrics for Scientific Software working group requests an email be sent to Gabrielle Allen \href{mailto:gdallen@illinois.edu}{(gdallen@illinois.edu)} to find out more about the group's efforts and how to participate.
+In lieu of a landing page, the Useful Metrics for Scientific Software working
+group requests an email be sent to Gabrielle Allen
+\href{mailto:gdallen@illinois.edu}{(gdallen@illinois.edu)} to find out more
+about the group's efforts and how to participate.

### 2015-11-19 WSSSPE3/report/wg_main_training.tex
Sections 5.6-5.7: no content change.

@@
 
-This group explored a rapidly growing array of training that is seen to contribute to sustainable software.
-The offerings are diverse, providing training that is more or less directly relevant to sustainable software.
-While research institutions  support professional development for research staff, the skills taught which might impact on sustainable software are limited at best, often lacking a clear and coherent development pathway.
-This growing array of training opportunities could usefully be coordinated by bringing together those involved in leading relevant initiatives on a regular basis.
+This group explored a rapidly growing array of training that is seen to
+contribute to sustainable software. The offerings are diverse, providing
+training that is more or less directly relevant to sustainable software. While
+research institutions support professional development for research staff, the
+skills taught which might impact on sustainable software are limited at best,
+often lacking a clear and coherent development pathway. This growing array of
+training opportunities could usefully be coordinated by bringing together those
+involved in leading relevant initiatives on a regular basis.
 
-\subsubsection{Fit with related activities}
-Two existing venues for discussion of related activities are identified:
+\subsubsection{Fit with related activities} Two existing venues for discussion
+of related activities are identified:
 
@@
 
-\item
-WSSSPE Workshop on Sustainable Software for Science: Practice and Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}} \katznote{change to cite}
+\item WSSSPE Workshop on Sustainable Software for Science: Practice and
+Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+\katznote{change to cite}
 
-\item
-SEHPCCSE International Workshop on Software Engineering for High Performance Computing in Computational Science and Engineering\footnote{\url{http://wssspe.researchcomputing.org.uk/}} \katznote{fix URL and change to cite}
+\item SEHPCCSE International Workshop on Software Engineering for High
+Performance Computing in Computational Science and
+Engineering\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+\katznote{fix URL and change to cite}
 
@@
 
-Next steps have been identified to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects.
+Next steps have been identified to quickly test whether there is interest in
+establishing a community committed to increasing the degree of coordination
+across training projects.
 
@@
 
-The main plans for the group are to convene discussion to explore bringing together a regular meeting of those involved in leading relevant training projects.
+The main plans for the group are to convene discussion to explore bringing
+together a regular meeting of those involved in leading relevant training
+projects.
 
@@
 
-The Training working group requests an email be sent to Nick Jones \href{mailto:nick.jones@nesi.org.nz}{(nick.jones@nesi.org.nz)} to find out more about the group's efforts and how to participate.
+The Training working group requests an email be sent to Nick Jones
+\href{mailto:nick.jones@nesi.org.nz}{(nick.jones@nesi.org.nz)} to find out more
+about the group's efforts and how to participate.

### 2015-11-19 WSSSPE3/report/wg_main_metrics.tex
Section 5.6-5.8: nanoscopic editing.

@@
 scientific impact, discovery, reducing duplication, serving as a basis for
-potential industrial interest in adopting software, prioritizing develop and
-support towards strategic objectives and making a case for new or continued
+potential industrial interest in adopting software, prioritizing development and
+support towards strategic objectives, and making a case for new or continued
 funding. However, there is no commonly-used standard for collecting or
@@
 
-\item WSSSPE3 Breakout Session: How can we measure the impact of a code on
+\item WSSSPE3 Breakout Session: How can we measure the impact of a piece of code on
 research, and its value to the
@@
 elements released under the NSF SI2 program all included metrics as a required
-component with submitters requested to include {\it "a list of tangible metrics,
+component with submitters requested to include {\it ``a list of tangible metrics,
 with end user involvement, to be used to measure the success of the software
-element developed, ...'}. These metrics are then reported as part of annual
+element developed, \dots''}. These metrics are then reported as part of annual
 reports to NSF by the projects. Although neither the proposal text describing
@@
 or JORS. More information about the group discussion is available in
-Appendix~\ref{sec:appendix_metrics}.
+Appendix~\ref{sec:appendix_metrics}. \choinote{Spell out SFTF, TRDF, CISE, JORS?}
 
@@
 
-The main plans for the group going forward are the creation of a white paper on
+The main plan for the group going forward is the creation of a white paper on
 the topic of useful metrics for scientific software. The authoring of this white

### 2015-11-19 WSSSPE3/report/wg_main_software_credit.tex
minor changes/comments in publishing sections moved some credit text to be a general intro to the section
minor change in start of software credit section
Section 5.6-5.8: nanoscopic editing.

@@
 
-\subsubsection{Why it is important}
+%\subsubsection{Why it is important}
 
-Science and engineering research increasingly relies on software, but there is currently no standard mechanism for citing software or receiving credit for developing software akin to receiving credit via citations for writing papers.
-Standardizing software citations offers one route to establishing such a mechanism.
-Software is currently eligible for DOI assignment, but DOI metadata fields are not well tuned for software compared to publications. 
-Some software providers apply for DOIs but it is still not widely adopted. 
-Also, there is no mechanism to cite software dependencies within software in the same way papers cite supporting prior work.
+Science and engineering research increasingly relies on software, but there is
+currently no standard mechanism for citing software or receiving credit for
+developing software akin to receiving credit via citations for writing papers.
+Standardizing software citations offers one route to establishing such a
+citation and credit mechanism. Software is currently eligible for DOI
+assignment, but DOI metadata fields are not well tuned for software compared to
+publications. Some software providers apply for DOIs but it is still not widely
+adopted. Also, there is no mechanism to cite software dependencies within
+software in the same way papers cite supporting prior work.
 
@@
 
-Publishing Software Working Group (\S\ref{sec:publishing-software}): publishing a software paper offers one existing mechanism for receiving credit, and further developing new publishing concepts for software will strengthen our activities.
+Publishing Software Working Group (\S\ref{sec:publishing-software}): publishing
+a software paper offers one existing mechanism for receiving credit, and further
+developing new publishing concepts for software will strengthen our activities.
 
-A number of groups external to WSSSPE (although with some overlapping members) are also focused on aspects of software credit, including the FORCE11 Software Citation Working Group (see plans for coordination below).
-In addition, a Software Credit workshop\footnote{London Software Credit workshop: \url{http://www.software.ac.uk/software-credit}} convened in London on October 19, following the conclusion of WSSSPE3.
-See Appendix~\ref{sec:appendix_SW_credit} for more detailed discussion of related activities.
+A number of groups external to WSSSPE (although with some overlapping members)
+also focused on aspects of software credit, including the FORCE11 Software
+Citation Working Group (see plans for coordination below). In addition, a
+Software Credit workshop\footnote{London Software Credit workshop:
+\url{http://www.software.ac.uk/software-credit}} convened in London on October
+19, following the conclusion of WSSSPE3. See
+Appendix~\ref{sec:appendix_SW_credit} for more detailed discussion of related
+activities.
 
@@
 
-The group discussed a number of topics related to software credit, including a contributorship taxonomy, software citation metadata, standards for citing software in publications, and increasing the value of software in academic promotion and tenure reviews.
-Although initial discussions both prior to and during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing as an example the Entertainment Identifier Registry~\cite{EIDR} used in the entertainment industry, the group decided to prioritize software citation.
-This decision was motivated by the idea that standardizing citations for software would introduce some initial credit for developers, and later the quantification of credit could be refined based on concepts such as transitive credit~\cite{wssspe2_katz,Katz:2014_tc}.
+The group discussed a number of topics related to software credit, including a
+contributorship taxonomy, software citation metadata, standards for citing
+software in publications, and increasing the value of software in academic
+promotion and tenure reviews. Although initial discussions both prior to and
+during WSSSPE3 focused on contribution taxonomy and dividing credit, discussing
+as an example the Entertainment Identifier Registry~\cite{EIDR} used in the
+entertainment industry, the group decided to prioritize software citation. This
+decision was motivated by the idea that standardizing citations for software
+would introduce some initial credit for developers, and later the quantification
+of credit could be refined based on concepts such as transitive
+credit~\cite{wssspe2_katz,Katz:2014_tc}.
 
-The majority of the remaining discussion focused on standardizing (1) the metadata necessary for software to be cited and (2) the mechanism for citing software in publications.
-In addition, discussions also oriented around the indexing of software citations necessary for establishing a software citation network either integrated with the existing paper citation ecosystem or complementary to it.
-See Appendix~\ref{sec:appendix_SW_credit} for a more detailed summary of the working group's discussion on these topics.
+The majority of the remaining discussion focused on standardizing (1) the
+metadata necessary for software to be cited and (2) the mechanism for citing
+software in publications. In addition, discussions also oriented around the
+indexing of software citations necessary for establishing a software citation
+network either integrated with the existing paper citation ecosystem or
+complementary to it. See Appendix~\ref{sec:appendix_SW_credit} for a more
+detailed summary of the working group's discussion on these topics.
 
@@
 
-The group already merged with the FORCE11 Software Citation Working Group (SCWG), and their efforts will focus (over the next six to nine months) on developing a document describing principles for software citation.
-Following publication of that document, the working group will focus on outreach to key groups (e.g., journals, publishers, indexers, professional societies).
-Longer-term plans include working with indexers to ensure that software citations are indexed and pursuing an open\slash community indexer; these activities may be organized by future FORCE11 working groups.
+The group already merged with the FORCE11 Software Citation Working Group
+(SCWG), and their efforts will focus (over the next six to nine months) on
+developing a document describing principles for software citation. Following
+publication of that document, the working group will focus on outreach to key
+groups (e.g., journals, publishers, indexers, professional societies).
+Longer-term plans include working with indexers to ensure that software
+citations are indexed and pursuing an open\slash community indexer; these
+activities may be organized by future FORCE11 working groups.
 
@@
 
-Since near-term efforts will be shifting to the FORCE11-SCWG, we direct interested readers to that group's existing landing page\footnote{FORCE11-SCWG landing page, \url{https://www.force11.org/group/software-citation-working-group}} and GitHub repository\footnote{FORCE11-SCWG GitHub page,  \url{https://github.com/force11/force11-scwg}}.
+Since near-term efforts will be shifting to the FORCE11-SCWG, we direct
+interested readers to that group's existing landing page\footnote{FORCE11-SCWG
+landing page,
+\url{https://www.force11.org/group/software-citation-working-group}} and GitHub
+repository\footnote{FORCE11-SCWG GitHub page,
+\url{https://github.com/force11/force11-scwg}}.

### 2015-11-19 WSSSPE3/report/wg_main_training.tex
Section 5.6-5.8: nanoscopic editing.

@@
 
-\item WSSSPE Workshop on Sustainable Software for Science: Practice and
-Experiences \footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+\item Workshop on Sustainable Software for Science: Practice and
+Experiences (WSSSPE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
 \katznote{change to cite}
 
-\item SEHPCCSE International Workshop on Software Engineering for High
+\item  International Workshop on Software Engineering for High
 Performance Computing in Computational Science and
-Engineering\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+Engineering(SEHPCCSE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
 \katznote{fix URL and change to cite}
@@
 
-The main plans for the group are to convene discussion to explore bringing
-together a regular meeting of those involved in leading relevant training
+The main plan for the group is to convene a discussion to explore bringing
+together regular meetings of those involved in leading relevant training
 projects.

### 2015-11-19 WSSSPE3/report/wg_main_software_credit.tex
undoing an edit

@@
 A number of groups external to WSSSPE (although with some overlapping members)
-also focused on aspects of software credit, including the FORCE11 Software
+are also focused on aspects of software credit, including the FORCE11 Software
 Citation Working Group (see plans for coordination below). In addition, a

### 2015-11-19 WSSSPE3/report/wg_main_training.tex
edit to edit

@@
 Performance Computing in Computational Science and
-Engineering(SEHPCCSE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+Engineering (SEHPCCSE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
 \katznote{fix URL and change to cite}

### 2015-11-19 WSSSPE3/report/WSSSPE3_review.tex
removing comment

@@
 need more documented success stories. These successes need to be socialized 
-\choinote{not sure if ``socialize'' can be used in such a way}
+%\choinote{not sure if ``socialize'' can be used in such a way}
 within the scientific community through workshops like the Software Engineering

### 2015-11-23 WSSSPE3/report/wg_appendix_best_practices.tex
Appendix D: Added in affiliations except for two.

@@
 \begin{itemize}
-\item Abani Patra 
-\item Sandra Gesing -- University of Notre Dame
-\item Neil Chue Hong 
-\item Greg Tucker 
-\item Birgit Penzens 
-\item Abigail Cabunoc Mayes 
-\item Jeff Carver 
-\item Frank L\"{o}ffler 
-\item Colin Venter 
-\item Lorraine Hwang 
-\item Sou-Cheng Choi
-\item Suresh Marru 
-\item Don Middleton 
-\item Daniel Katz  
-\item Kyle Niemeyer 
-\item Jeff Carver 
-\item Dan Gunter 
-\item Alexander Konovalov 
-\item Tom Crick
+\item Abani Patra, University at Buffalo
+\item Sandra Gesing, University of Notre Dame
+\item Neil Chue Hong, Software Sustainability Institute
+\item Gregory Tucker, University of Colorado at Boulder
+\item Birgit Penzens, California State University Long Beach
+\item Abigail Cabunoc Mayes, Mozilla Foundation
+\item Jeff Carver, University of Alabama
+\item Frank L\"{o}ffler, Louisiana State University 
+\item Colin Venter,  University of Huddersfield
+\item Lorraine Hwang, UC Davis - CIG
+\item Sou-Cheng Choi, NORC at the University of Chicago \&  Illinois Institute of Technology
+\item Suresh Marru, Indiana University
+\item Don Middleton, NCAR 
+\item Daniel Katz,  U Chicago \& Argonne
+\item Kyle Niemeyer, Oregon State University
+\item Jeffrey Carver, University of Alabama
+\item Dan Gunter, LBNL
+\item Alexander Konovalov, \choinote{TBD}
+\item Tom Crick,  \choinote{TBD}
 

### 2015-11-23 WSSSPE3/report/wg_appendix_best_practices.tex
Appendix D: Fix typos.
Appendix D: Added line breaks.
Appendix D:  Small changes.

@@
 \begin{itemize}
+
 \item Requirements engineering to create tools with immediate uptake;
+
 \item When should software ``die''?
-\item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
+
+\item Catering to disruptive developments in environment, e.g., new hardware,
+new methodology;
+
 \item Dimensions of sustainability -- economic, technical, environmental,
-declining interest in primary application area), \katznote{not sure what the prev. comment goes with} social.
+declining interest in primary application area), \katznote{not sure what the
+prev. comment goes with} social.
+
 \end{itemize}
 
-Sustainability requires community participation in code development and/or a wide adoption of software.
-The larger the community base is using a piece of software, the better are the funding possibilities and thus also the sustainability options.
-Additionally developer commitment to an application is essential and experience shows that software packages with an evangelist imposing strong inspiration and discipline are more likely to achieve sustainability.
-While a single person can push sustainability to a certain level, open source software also needs sustained commitment from the developer community.
-Such sustained commitments include diverse tasks and roles, which can be fulfilled by diverse developers with different knowledge levels.
-Besides developing software and appropriate software management with measures for extensibility and scalability of the software, active (expertise) support for users via a user forum with a quick turnaround is crucial.
-The barrier to entry for the community as users as well as developers has to be as low as possible.
+Sustainability requires community participation in code development and/or a
+wide adoption of software. The larger the community base is using a piece of
+software, the better are the funding possibilities and thus also the
+sustainability options. Additionally,  developers’ commitment to an application is
+essential and experience shows that software packages with an evangelist
+imposing strong inspiration and discipline are more likely to achieve
+sustainability. While a single person can push sustainability to a certain
+level, open source software also needs sustained commitment from the developer
+community. Such sustained commitments include diverse tasks and roles, which can
+be fulfilled by diverse developers with different knowledge levels. Besides
+developing software and appropriate software management with measures for
+extensibility and scalability of the software, active (expertise) support for
+users via a user forum with a quick turnaround is crucial. The barrier to entry
+for the community as users as well as developers has to be as low as possible.
 
@@
 
-The opportunity lies in collaboration on a white paper, which will be revisited regularly for further improvements, to enhance knowledge of the state of best practices, resulting in a peer-reviewed paper.
-We would like to reach a wide community by doing this.
-But these are also the challenges and obstacles -- to get everyone to contribute to the paper and to reach the community.
+The opportunity lies in collaboration on a white paper, which will be revisited
+regularly for further improvements, to enhance knowledge of the state of best
+practices, resulting in a peer-reviewed paper. We would like to reach a wide
+community by doing this. But these are also the challenges and obstacles -- to
+get everyone to contribute to the paper and to reach the community.
 
@@
 
-The key next steps are to write the introduction, reach out to the co-authors, and to agree on the scope of the white paper.
+The key next steps are to write an introduction, reach out to the co-authors,
+and to agree on the scope of the white paper.
 
@@
 
-Sandra Gesing and Abani Patra are the main editors and will organize the overall communication and paper. Sections will be assigned to diverse co-authors.
+Sandra Gesing and Abani Patra are the main editors and will organize the overall
+communication and the paper. Sections will be assigned to diverse co-authors.
 
@@
 
-At the moment we don't see any further requirements.
+At the moment we do not see any further requirements.
 
@@
 \item 31 Jan: Analyzing funding possibilities for survey
-\item 31 Jan: First versions of section
-\item 15 Feb: Distribution to WSSSPE community
-\item 31 Mar: Final version of white paper
-\item 30 Apr: Submission of peer-reviewed paper?
+\item 31 Jan: First version of each section
+\item 15 Feb: Distribution to the WSSSPE community
+\item 31 Mar: Final version of the white paper
+\item 30 Apr: Submission to a peer-reviewed journal?
 \end{itemize}

### 2015-11-23 WSSSPE3/report/wg_appendix_best_practices.tex
Appendix D: Added in affiliations except for two.

@@
 \begin{itemize}
-\item Abani Patra, University at Buffalo
-\item Sandra Gesing, University of Notre Dame
-\item Neil Chue Hong, Software Sustainability Institute
-\item Gregory Tucker, University of Colorado at Boulder
-\item Birgit Penzens, California State University Long Beach
-\item Abigail Cabunoc Mayes, Mozilla Foundation
-\item Jeff Carver, University of Alabama
-\item Frank L\"{o}ffler, Louisiana State University 
-\item Colin Venter,  University of Huddersfield
-\item Lorraine Hwang, UC Davis - CIG
-\item Sou-Cheng Choi, NORC at the University of Chicago \&  Illinois Institute of Technology
-\item Suresh Marru, Indiana University
-\item Don Middleton, NCAR 
-\item Daniel Katz,  U Chicago \& Argonne
-\item Kyle Niemeyer, Oregon State University
-\item Jeffrey Carver, University of Alabama
-\item Dan Gunter, LBNL
-\item Alexander Konovalov, \choinote{TBD}
-\item Tom Crick,  \choinote{TBD}
+\item Abani Patra -- University at Buffalo
+\item Sandra Gesing -- University of Notre Dame
+\item Neil Chue Hong -- Software Sustainability Institute
+\item Gregory Tucker -- University of Colorado at Boulder
+\item Birgit Penzens -- California State University Long Beach
+\item Abigail Cabunoc Mayes -- Mozilla Foundation
+\item Jeff Carver -- University of Alabama
+\item Frank L\"{o}ffler -- Louisiana State University 
+\item Colin Venter --  University of Huddersfield
+\item Lorraine Hwang -- UC Davis 
+\item Sou-Cheng Choi -- NORC at the University of Chicago \&  Illinois Institute of Technology
+\item Suresh Marru -- Indiana University
+\item Don Middleton -- NCAR 
+\item Daniel Katz --  University of Chicago \& Argonne National Laboratory
+\item Kyle Niemeyer -- Oregon State University
+\item Jeffrey Carver -- University of Alabama
+\item Dan Gunter -- LBNL
+\item Alexander Konovalov -- \choinote{TBD}
+\item Tom Crick --  \choinote{TBD}
 

### 2015-11-23 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
fixing WSSSPE spelling again
Appendix E: Touch up affiliations.
Appendix E: Added line breaks.

@@
 
-James Hetherington\footnote{email: \href{mailto:j.hetherington@ucl.ac.uk}{j.hetherington@ucl.ac.uk}}
-will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+James Hetherington\footnote{email:
+\href{mailto:j.hetherington@ucl.ac.uk}{j.hetherington@ucl.ac.uk}} will serve as
+the point of contact for this working group, and be responsible for ensuring
+timely progress of the planned actions.
 
@@
 
-The group at WSSPE:
+The group at WSSSPE:
 
@@
 \item Nancy Wilkins-Diehr -- San Diego Supercomputer Center
-\item Alexander Vyushkov -- Notre Dame
-\item Sandra Gesing - Notre Dame
+\item Alexander Vyushkov -- University of Notre Dame
+\item Sandra Gesing -- University of Notre Dame
 \item Ali Swanson -- University of Oxford
@@
 careers. (Usually from RA to RSE -- examples of movement in the other direction
-from readers would be welcomed.)
-Either way, the label ``Research Software Engineer'' is now starting to
-have some power. Many scientists do not want to be writing code; some do, to
-varying degrees. These groups can usefully support each other.
+from readers would be welcomed.) Either way, the label ``Research Software
+Engineer'' is now starting to have some power. Many scientists do not want to be
+writing code; some do, to varying degrees. These groups can usefully support
+each other.
 
@@
 
-Will research science developers be required, in the long run?
-One issue that came up was whether the need for developers was a time bounded one; is it the case that the new generation of computer and software savvy scientists will be so comfortable in developing their own code that the professional developer will not be needed.
-And this brings up the flip side question, ``Do scientists really want to be writing code?''
+Will research science developers be required, in the long run? One issue that
+came up was whether the need for developers was a time bounded one; is it the
+case that the new generation of computer and software savvy scientists will be
+so comfortable in developing their own code that the professional developer will
+not be needed. And this brings up the flip side question, ``Do scientists really
+want to be writing code?''
 
-Career Path.
-We also had a little discussion about how to make a career path for research developers.
-It need not be solely an academic enterprise, but in the past tenure has often been problematic for people of this class.
+Career Path. We also had a little discussion about how to make a career path for
+research developers. It need not be solely an academic enterprise, but in the
+past tenure has often been problematic for people of this class.
 
@@
 are plans to permit resource sharing between institutional RSE groups. Perhaps
-there are circumstances under which an RSE skill exchange could be arranged, either
-formally or informally.
+there are circumstances under which an RSE skill exchange could be arranged,
+either formally or informally.
 
@@
 are required for submission of proposals, so collaboration is an essential part
-of grant submission, and this will tend to bring developers and
-scientists together. The UCL group also follows this approach, with all bids
-requiring an academic collaborator.
+of grant submission, and this will tend to bring developers and scientists
+together. The UCL group also follows this approach, with all bids requiring an
+academic collaborator.
 
@@
 
-An alternative approach has been the provision of programming expertise as
-part of national compute services. The US XSEDE project's extended collaboration
-service (ECSS) is a set of developers who are paid with XSEDE funding, and are on 
-``permanent'' staff.  When PIs request allocations on XSEDE resources, there is a 
-finite pool of developer time that can be awarded, typically for one year only, 
-and at partial effort, typically 20 percent or so. The finite time allowed provides
-motivation for the scientist and their group to work closely the developer and to 
-become educated in what the developer is doing, so they can sustain the effort, once 
-the ECSS period is over. This funding mechanism can be highly efficient for scientific
-problems, because the developer pool assembled by the research providers are, by definition, 
-expert in the characteristics of their specific resource, and can very quickly assess the
-scientists needs, and what it will take to implement software that meets the user's needs. 
-However, it does not develop capacity within institutions, and since XSEDE is a time-bounded
-program, it should not be relied upon as a long term solution to acquiring this type of capacity.
+An alternative approach has been the provision of programming expertise as part
+of national compute services. The US XSEDE project's extended collaboration
+service (ECSS) is a set of developers who are paid with XSEDE funding, and are
+on ``permanent'' staff. When PIs request allocations on XSEDE resources, there
+is a finite pool of developer time that can be awarded, typically for one year
+only, and at partial effort, typically 20 percent or so. The finite time allowed
+provides motivation for the scientist and their group to work closely the
+developer and to become educated in what the developer is doing, so they can
+sustain the effort, once the ECSS period is over. This funding mechanism can be
+highly efficient for scientific problems, because the developer pool assembled
+by the research providers are, by definition, expert in the characteristics of
+their specific resource, and can very quickly assess the scientists needs, and
+what it will take to implement software that meets the user's needs. However, it
+does not develop capacity within institutions, and since XSEDE is a time-bounded
+program, it should not be relied upon as a long term solution to acquiring this
+type of capacity.
 
 The UK allows this kind of collaboration to support the creation of scientific
-software for the large supercomputing resource called ARCHER.
-However, as well as resourcing this from the staff of the Edinburgh Parallel
-Computing Centre, who host the computer, this ``embedded CSE'' resource also
-allows the programming to come from local groups. This has been very helpful
-in providing funding to establish local groups. These work best when they
-develop good collaborations with national cyberinfrastructure pools.
-When an organization assembles a developer pool, diversity is developed and 
-skills can be transferred.
+software for the large supercomputing resource called ARCHER. However, as well
+as resourcing this from the staff of the Edinburgh Parallel Computing Centre,
+who host the computer, this ``embedded CSE'' resource also allows the
+programming to come from local groups. This has been very helpful in providing
+funding to establish local groups. These work best when they develop good
+collaborations with national cyberinfrastructure pools. When an organization
+assembles a developer pool, diversity is developed and skills can be
+transferred.
 
@@
 
-The CANARIE group (Canada) accepts proposals for providing services to broad communities, 
-integrate people who are doing things that are complementary, the goal is to make the available 
-stack more robust and richer for everyone. They offer short rounds of funding that can
-have as a key metric creating some useful functionality that shows a diversity
-of input and draws from across disciplines, then more funding could follow. This
-could apply within or across institutions.
+The CANARIE group (Canada) accepts proposals for providing services to broad
+communities, integrate people who are doing things that are complementary, the
+goal is to make the available stack more robust and richer for everyone. They
+offer short rounds of funding that can have as a key metric creating some useful
+functionality that shows a diversity of input and draws from across disciplines,
+then more funding could follow. This could apply within or across institutions.
 
-There can be problems communicating across cultural barriers, with domain scientists seeing developers 
-as ``other''. Collaboration, and tools to fund, encourage or motivate collaboration are
-extremely important.
+There can be problems communicating across cultural barriers, with domain
+scientists seeing developers as ``other''. Collaboration, and tools to fund,
+encourage or motivate collaboration are extremely important.
 
-We think support from non-governmental organizations will be important if RSE groups will become established.
-The Sloan Foundation is currently funding data science engineers, who work in the context of other SW developers
-at University of Washington. These scientists work in the e-Science Studio/Data Science Studio, and they
-help a group of graduate students in solving their problems in data science and data management. During Fall and
-Spring, a 10-week incubator program allows students to work two days a week to work on a data-intensive science
-project. Some fraction of the developer time is dedicated to the developers' personal interests as well as instruction.
+We think support from non-governmental organizations will be important if RSE
+groups will become established. The Sloan Foundation is currently funding data
+science engineers, who work in the context of other SW developers at University
+of Washington. These scientists work in the e-Science Studio/Data Science
+Studio, and they help a group of graduate students in solving their problems in
+data science and data management. During Fall and Spring, a 10-week incubator
+program allows students to work two days a week to work on a data-intensive
+science project. Some fraction of the developer time is dedicated to the
+developers' personal interests as well as instruction.
 
@@
 this arrangement as it is hard to find good staff for short-term positions, so
-having a pool of research programming staff on hand resolves this problem.
-We recommend that funders give explicit guidance to grant holders and
-institutions that such arrangement are favorable. Framework agreements
-permitting this to go ahead without
-checking back every time with funders and/or grant panels would further
-smooth this.
-(This is also provides a more stable job for those who hold these skills, but
-arguments about making life nicer for postdocs will not help persuade funders
-or PIs!)
+having a pool of research programming staff on hand resolves this problem. We
+recommend that funders give explicit guidance to grant holders and institutions
+that such arrangement are favorable. Framework agreements permitting this to go
+ahead without checking back every time with funders and/or grant panels would
+further smooth this. (This is also provides a more stable job for those who hold
+these skills, but arguments about making life nicer for postdocs will not help
+persuade funders or PIs!)
 
-There is some question about the most effective duration and percent of full time for a 
-programmer's work on a project. At least three months is necessary for the programmer to 
-read into the science (RSEs must not become so disengaged from research that they don't have
-time to read a few papers -- this will result in code which doesn't meet scientific needs), 
-but too long could result in an RSE losing their flexibility, becoming so engaged in one project
-that when that project ends, they find it hard to transfer. For this reason, we also recommend
-that 40 percent is ideal; two projects per developer, with some time for training and infrastructure work.
-Two developers per project seems to be ideal, in the sense that software development is enhanced by 
-two pairs of eyes.
+There is some question about the most effective duration and percent of full
+time for a programmer's work on a project. At least three months is necessary
+for the programmer to read into the science (RSEs must not become so disengaged
+from research that they don't have time to read a few papers -- this will result
+in code which doesn't meet scientific needs), but too long could result in an
+RSE losing their flexibility, becoming so engaged in one project that when that
+project ends, they find it hard to transfer. For this reason, we also recommend
+that 40 percent is ideal; two projects per developer, with some time for
+training and infrastructure work. Two developers per project seems to be ideal,
+in the sense that software development is enhanced by two pairs of eyes.
 
-There is as yet no clear answer as to the scale of aggregation needed to
-make such a program work. A university wide program allows enough scale to
-be robust to fluctuations of funding within one field.
-But a specialization focus on developers to support, for example,
-physical or biological
-sciences may be preferable, if the customer base is large enough.
-The desire to aggregate enough work to make it sustainable, and the need to
-have domain-relevant research programming skills, are in tension.
+There is as yet no clear answer as to the scale of aggregation needed to make
+such a program work. A university wide program allows enough scale to be robust
+to fluctuations of funding within one field. But a specialization focus on
+developers to support, for example, physical or biological sciences may be
+preferable, if the customer base is large enough. The desire to aggregate enough
+work to make it sustainable, and the need to have domain-relevant research
+programming skills, are in tension.
 
@@
 forward proposals that are a priority of the community as a whole, for example,
-biosimulation or plasma physics. These bodies act as custodians of community codes,
-and a central team also provides software engineering support.
+biosimulation or plasma physics. These bodies act as custodians of community
+codes, and a central team also provides software engineering support.
 
@@
 cyberinfrastructure component is clear. Funding which permits code to be
-refactored, tidied and optimized  is rare; this is often done ``on the sly''
-in a scientifically focused grant. The UK EPSRC's ``software for the future''
-call, which really permits explicit investment in software as an infrastructure,
-is so oversubscribed as to have a 4\% success rate; the demand is clear!
+refactored, tidied and optimized is rare; this is often done ``on the sly'' in a
+scientifically focused grant. The UK EPSRC's ``software for the future'' call,
+which really permits explicit investment in software as an infrastructure, is so
+oversubscribed as to have a 4\% success rate; the demand is clear!
 
@@
 collaboration is hard to foster here; as incentive structures are still focused
-on short-term papers. This can cause infrastructure developers to
-focus more on publications
-in their areas of mathematics and CS, the domain developers on the shorter-term
-needs of their own fields. Genuine collaborative co-construction
-is harder to foster.
+on short-term papers. This can cause infrastructure developers to focus more on
+publications in their areas of mathematics and CS, the domain developers on the
+shorter-term needs of their own fields. Genuine collaborative co-construction is
+harder to foster.
 
-It can be a more difficult
-to help leading domain scientists see the value of engineering effort than those
-in their teams who are forced to work with difficult-to-use or unreliable
-software tools, as they do not see the pain. Perhaps a version of
-``software carpentry'' targeted at those PIs
-who are awarded or apply for software heavy grants could be valuable here.
+It can be a more difficult to help leading domain scientists see the value of
+engineering effort than those in their teams who are forced to work with
+difficult-to-use or unreliable software tools, as they do not see the pain.
+Perhaps a version of ``software carpentry'' targeted at those PIs who are
+awarded or apply for software heavy grants could be valuable here.
 
-RSEs provide a useful contribution to their Universities' \textbf{teaching} missions,
-as well as for research, as they are well placed to deliver the research programming
-training that many scientists now need. In the
-longer term, with programming skills taught to all through their career,
-we hope specialist
+RSEs provide a useful contribution to their Universities' \textbf{teaching}
+missions, as well as for research, as they are well placed to deliver the
+research programming training that many scientists now need. In the longer term,
+with programming skills taught to all through their career, we hope specialist
 scientific developers will be less needed.
@@
 to get their permission to list them on a list of research software groups.
-Casual conversation during the meeting made it clear that although the title is not
-widely used in the US, the this position is not rare. We spoke with several individuals
-who, at distinct Universities, had RSEs (in effect if not in name) and these were funded
-and under differing models.
+Casual conversation during the meeting made it clear that although the title is
+not widely used in the US, the this position is not rare. We spoke with several
+individuals who, at distinct Universities, had RSEs (in effect if not in name)
+and these were funded and under differing models.
 
@@
 
-In this respect, information gathering via a survey subsequent analysis could be very useful.
-We would need to assemble a list of targeted individuals (what position and rank is
-likely to know and care enough to respond?) Perhaps the Science Gateway Institute has already acquire
-information that could be helpful to advance this issue, and/or craft the proper survey and target
-individuals.
+In this respect, information gathering via a survey subsequent analysis could be
+very useful. We would need to assemble a list of targeted individuals (what
+position and rank is likely to know and care enough to respond?) Perhaps the
+Science Gateway Institute has already acquire information that could be helpful
+to advance this issue, and/or craft the proper survey and target individuals.
 
@@
 The UK RSE community will provide initial facilities to host this list, and
-continue to work to spread the initiative, but local leadership in the US
-is needed if this campaign is to succeed. This will require an initial gathering
-of identified research software organizations in the US to this end.
+continue to work to spread the initiative, but local leadership in the US is
+needed if this campaign is to succeed. This will require an initial gathering of
+identified research software organizations in the US to this end.
 
@@
 groups to form an organization and create a resource sharing structure would
-help to further this campaign. Funding to conduct and analyze a survey could also be quite useful
-as knowing where we stand today, and what models are in use could fuel the ideas for further
-development of developers in this category.
+help to further this campaign. Funding to conduct and analyze a survey could
+also be quite useful as knowing where we stand today, and what models are in use
+could fuel the ideas for further development of developers in this category.
 
-In the longer term, funding organizations, especially non-governmental organizations
-with the capability to effect innovation through seed funding, could provide
-support to nucleate the creation of research software groups. As noted above, Sloan has already
-initiated one such program, and collaboration with Sloan or at least study of their methods and 
-success or failure could be extremely useful in approaching Universities and other institutions 
-in funding this development track. It seems clear that if the value proposition can be made to
-University administrators, this track could flourish with buy-in at the administrative level.
+In the longer term, funding organizations, especially non-governmental
+organizations with the capability to effect innovation through seed funding,
+could provide support to nucleate the creation of research software groups. As
+noted above, Sloan has already initiated one such program, and collaboration
+with Sloan or at least study of their methods and success or failure could be
+extremely useful in approaching Universities and other institutions in funding
+this development track. It seems clear that if the value proposition can be made
+to University administrators, this track could flourish with buy-in at the
+administrative level.

### 2015-11-23 WSSSPE3/report/wg_appendix_best_practices.tex
small changes in affils and other

@@
 \subsection{Group Members}
-\katznote{add affils for all, please}
 
@@
 \item Don Middleton -- NCAR 
-\item Daniel Katz --  University of Chicago \& Argonne National Laboratory
+\item Daniel S. Katz --  University of Chicago \& Argonne National Laboratory
 \item Kyle Niemeyer -- Oregon State University
@@
 \item Dan Gunter -- LBNL
-\item Alexander Konovalov -- \choinote{TBD}
-\item Tom Crick --  \choinote{TBD}
+\item Alexander Konovalov -- University of St Andrews
+\item Tom Crick --  Cardiff Metropolitan University
 
@@
 software, the better are the funding possibilities and thus also the
-sustainability options. Additionally,  developers’ commitment to an application is
+sustainability options. Additionally, the developers’ commitment to an application is
 essential and experience shows that software packages with an evangelist

### 2015-11-24 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Appendix E: Fix typos and added comments.

@@
 \item Lorraine Hwang -- UC Davis
-\item Simon Trigger
+\item Simon Trigger \choinote{TBD}
 \item Nancy Wilkins-Diehr -- San Diego Supercomputer Center
@@
 
-``Are you an RSE or a RA?'' -- this is not properly a binary question. Most of
+``Are you an RSE or a RA?'' is not properly a binary question. Most of
 us sit at different points on that spectrum, and move along it during our
-careers. (Usually from RA to RSE -- examples of movement in the other direction
-from readers would be welcomed.) Either way, the label ``Research Software
+careers (usually from RA to RSE -- examples of movement in the other direction
+from readers would be welcomed). Either way, the label ``Research Software
 Engineer'' is now starting to have some power. Many scientists do not want to be
@@
 The power of the label? Getting the word out about RSE support using the label. 
+\choinote{Please review this short, fragmented paragraph.}
 
-Will research science developers be required, in the long run? One issue that
+Will research science developers be required in the long run? One issue that
 came up was whether the need for developers was a time bounded one; is it the
 case that the new generation of computer and software savvy scientists will be
-so comfortable in developing their own code that the professional developer will
-not be needed. And this brings up the flip side question, ``Do scientists really
+so comfortable in developing their own code that the professional developers will
+not be needed? And this brings up the flip side question, ``Do scientists really
 want to be writing code?''
 
-Career Path. We also had a little discussion about how to make a career path for
+We also had a little discussion about how to make a career path for
 research developers. It need not be solely an academic enterprise, but in the
@@
 Skills and resources may vary between teams. To help resolve this, maintaining
-high levels of communication between groups will be valuable. In the UK, there
+high levels of communication between groups will be valuable. In the United Kingdom (UK), there
 are plans to permit resource sharing between institutional RSE groups. Perhaps
@@
 Collaborative funding can be important for RSE groups, to ensure that research
-leadership remains with the domain scientists. At NCAR, University partnerships
+leadership remains with the domain scientists. At NCAR, university partnerships
 are required for submission of proposals, so collaboration is an essential part
@@
 Another example of a success is the development of semantics and linked data in
-support of Ocean Sciences. An EarthCube-funded project pairs domain scientists
+support of ocean sciences. An EarthCube-funded project pairs domain scientists
 with RSEs and has been successful; the semantics attached have increased data
@@
 An alternative approach has been the provision of programming expertise as part
-of national compute services. The US XSEDE project's extended collaboration
-service (ECSS) is a set of developers who are paid with XSEDE funding, and are
+of national computing services. The US XSEDE project's Extended Collaboration Support 
+Services (ECSS) is a set of developers who are paid with XSEDE funding, and are
 on ``permanent'' staff. When PIs request allocations on XSEDE resources, there
@@
 only, and at partial effort, typically 20 percent or so. The finite time allowed
-provides motivation for the scientist and their group to work closely the
+provides motivation for the scientists and their group to work closely with a
 developer and to become educated in what the developer is doing, so they can
-sustain the effort, once the ECSS period is over. This funding mechanism can be
+sustain the effort once the ECSS period is over. This funding mechanism can be
 highly efficient for scientific problems, because the developer pool assembled
 by the research providers are, by definition, expert in the characteristics of
-their specific resource, and can very quickly assess the scientists needs, and
+their specific resource, and can very quickly assess the scientists' needs, and
 what it will take to implement software that meets the user's needs. However, it
 does not develop capacity within institutions, and since XSEDE is a time-bounded
-program, it should not be relied upon as a long term solution to acquiring this
+program, it should not be relied upon as a long-term solution to acquiring this
 type of capacity.
 
-The UK allows this kind of collaboration to support the creation of scientific
-software for the large supercomputing resource called ARCHER. However, as well
-as resourcing this from the staff of the Edinburgh Parallel Computing Centre,
+The UK allows this kind of collaboration, called ARCHER, to support the creation of scientific
+software for the large supercomputing resource. However, as well
+as resourcing this from the staff of the Edinburgh Parallel Computing Centre, \choinote{not sure what the previous part means}
 who host the computer, this ``embedded CSE'' resource also allows the
@@
 
-We would like to see these models applied outside High Performance Computing.
+We would like to see these models applied outside high performance computing.
 Most scientific software is not destined to run on national cyberinfrastructure,
@@
 The CANARIE group (Canada) accepts proposals for providing services to broad
-communities, integrate people who are doing things that are complementary, the
+communities, integrating people who are doing things that are complementary. The
 goal is to make the available stack more robust and richer for everyone. They
@@
 functionality that shows a diversity of input and draws from across disciplines,
-then more funding could follow. This could apply within or across institutions.
+then more funding could follow. \choinote{I find the last sentence unclear.} This could apply within or across institutions.
 
-There can be problems communicating across cultural barriers, with domain
+There can be communication problems across cultural barriers, with domain
 scientists seeing developers as ``other''. Collaboration, and tools to fund,
 encourage or motivate collaboration are extremely important. 
+\choinote{Last sentence have at least two verbs.}
 
@@
 groups will become established. The Sloan Foundation is currently funding data
-science engineers, who work in the context of other SW developers at University
+science engineers, who work in the context of other SW developers at the University
 of Washington. These scientists work in the e-Science Studio/Data Science
@@
 data science and data management. During Fall and Spring, a 10-week incubator
-program allows students to work two days a week to work on a data-intensive
+program allows students to work two days a week on a data-intensive
 science project. Some fraction of the developer time is dedicated to the
-developers' personal interests as well as instruction.
+developers' personal interests as well as instructions.
 
-The goal for Sloan is to provide success stories, to provide demonstrable value
+The goal for Sloan is to provide success stories and demonstrable value
 in the presence of data scientists on university staff. These stories are the
@@
 funding, just as DNA sequencing is for core facilities, if it becomes apparent
-that this gives competitive advantage to the University's research effort.
+that this gives competitive advantages to a university's research effort.
 
@@
 ahead without checking back every time with funders and/or grant panels would
-further smooth this. (This is also provides a more stable job for those who hold
-these skills, but arguments about making life nicer for postdocs will not help
-persuade funders or PIs!)
+further smooth this. (This also provides more stable jobs for those who hold
+the skills, but arguments about making life nicer for postdocs will not help
+persuade the funders or the PIs.)
 
-There is some question about the most effective duration and percent of full
+There is some questions about the most effective duration and percent of full
 time for a programmer's work on a project. At least three months is necessary
 for the programmer to read into the science (RSEs must not become so disengaged
-from research that they don't have time to read a few papers -- this will result
-in code which doesn't meet scientific needs), but too long could result in an
+from research that they do not have time to read a few papers -- this will result
+in code which does not meet scientific needs), but too long could result in an
 RSE losing their flexibility, becoming so engaged in one project that when that
@@
 
-There is as yet no clear answer as to the scale of aggregation needed to make
+There is not yet clear answer to the scale of aggregation needed to make
 such a program work. A university wide program allows enough scale to be robust
@@
 In the UK, another source of funding for research software has been the
-Collaborative computational projects (CCPs): domain specific communities put
+Collaborative Computational Projects (CCPs): domain specific communities put
 forward proposals that are a priority of the community as a whole, for example,
@@
 However this area develops, the need for funding to secure software as a
-cyberinfrastructure component is clear. Funding which permits code to be
-refactored, tidied and optimized is rare; this is often done ``on the sly'' in a
+cyberinfrastructure component is clear. \choinote{last sentence unclear} Funding which permits code to be
+refactored, tidied and optimized is rare; this is often done ``on the sly'' \choinote{on the fly?} in a
 scientifically focused grant. The UK EPSRC's ``software for the future'' call,
@@
 
-One opportunity is the idea of co-design; where infrastructural libraries are
+One opportunity is the idea of co-design, where infrastructural libraries are
 developed alongside the scientific codes that will call them. However,
@@
 on short-term papers. This can cause infrastructure developers to focus more on
-publications in their areas of mathematics and CS, the domain developers on the
+publications in their areas of mathematics and computer science, the domain developers on the
 shorter-term needs of their own fields. Genuine collaborative co-construction is
@@
 Perhaps a version of ``software carpentry'' targeted at those PIs who are
-awarded or apply for software heavy grants could be valuable here.
+awarded or apply for software grants could be valuable here.
 
-RSEs provide a useful contribution to their Universities' \textbf{teaching}
-missions, as well as for research, as they are well placed to deliver the
+RSEs provide a useful contribution to their universities' teaching
+missions, as well as research, as they are well placed to deliver the
 research programming training that many scientists now need. In the longer term,
-with programming skills taught to all through their career, we hope specialist
+with programming skills taught to all through their careers, we hope specialist
 scientific developers will be less needed.
@@
 Casual conversation during the meeting made it clear that although the title is
-not widely used in the US, the this position is not rare. We spoke with several
-individuals who, at distinct Universities, had RSEs (in effect if not in name)
-and these were funded and under differing models.
+not widely used in the US,  this position is not rare. We spoke with several
+individuals who, at distinct universities, had RSEs (in effect if not in name)
+and these were funded under differing models.
 
 We will also look for examples of groups which have successfully become self-
-sustaining following.
+sustaining following. \choinote{following what?}
 
-In this respect, information gathering via a survey subsequent analysis could be
+In this respect, information gathering via a survey and subsequent analysis could be
 very useful. We would need to assemble a list of targeted individuals (what
-position and rank is likely to know and care enough to respond?) Perhaps the
-Science Gateway Institute has already acquire information that could be helpful
-to advance this issue, and/or craft the proper survey and target individuals.
+positions and ranks are likely to know and care enough to respond?). Perhaps the
+Science Gateway Institute has already acquired information that could be helpful
+to advance this issue, and/or craft a proper survey and target some individuals.
 
@@
 with Sloan or at least study of their methods and success or failure could be
-extremely useful in approaching Universities and other institutions in funding
+extremely useful in approaching universities and other institutions in funding
 this development track. It seems clear that if the value proposition can be made
-to University administrators, this track could flourish with buy-in at the
+to university administrators, this track could flourish with buy-in at the
 administrative level.

### 2015-11-24 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
changes, mostly small

@@
 
-The power of the label? Getting the word out about RSE support using the label. 
-\choinote{Please review this short, fragmented paragraph.}
+What is the power of the label? How can we get the word out about RSE support using the label?
 
@@
 An alternative approach has been the provision of programming expertise as part
-of national computing services. The US XSEDE project's Extended Collaboration Support 
+of national compute services. The US XSEDE project's Extended Collaboration Support 
 Services (ECSS) is a set of developers who are paid with XSEDE funding, and are
@@
 only, and at partial effort, typically 20 percent or so. The finite time allowed
-provides motivation for the scientists and their group to work closely with a
+provides motivation for the scientist and the scientist's group to work closely with the
 developer and to become educated in what the developer is doing, so they can
@@
 by the research providers are, by definition, expert in the characteristics of
-their specific resource, and can very quickly assess the scientists' needs, and
+their specific resource, and can very quickly assess the scientist's needs, and
 what it will take to implement software that meets the user's needs. However, it
@@
 
-The UK allows this kind of collaboration, called ARCHER, to support the creation of scientific
-software for the large supercomputing resource. However, as well
-as resourcing this from the staff of the Edinburgh Parallel Computing Centre, \choinote{not sure what the previous part means}
-who host the computer, this ``embedded CSE'' resource also allows the
-programming to come from local groups. This has been very helpful in providing
-funding to establish local groups. These work best when they develop good
+The UK allows this kind of collaboration to support the creation of scientific
+software for the large supercomputing resource (ARCHER). However, while the
+support can come the staff of the Edinburgh Parallel Computing Centre,
+who host the computer, this ``embedded CSE'' resource also funds the
+programming coming from local groups. This has been very helpful in providing
+funding to establish local groups. These work best when these groups develop good
 collaborations with national cyberinfrastructure pools. When an organization
@@
 goal is to make the available stack more robust and richer for everyone. They
-offer short rounds of funding that can have as a key metric creating some useful
-functionality that shows a diversity of input and draws from across disciplines,
-then more funding could follow. \choinote{I find the last sentence unclear.} This could apply within or across institutions.
+offer short cycles of funding that can have use creating some useful
+functionality that shows a diversity of input and draws from across disciplines as a key metric,  If this metric is met successfully, 
+then more funding may follow. This can apply within or across institutions.
 
-There can be communication problems across cultural barriers, with domain
-scientists seeing developers as ``other''. Collaboration, and tools to fund,
-encourage or motivate collaboration are extremely important. 
-\choinote{Last sentence have at least two verbs.}
+There can be problems communicating across cultural barriers, with domain
+scientists seeing developers as ``other''. Both collaboration and tools to fund,
+encourage, or motivate collaboration are extremely important.
 
@@
 groups will become established. The Sloan Foundation is currently funding data
-science engineers, who work in the context of other SW developers at the University
+science engineers, who work in the context of other software developers at the University
 of Washington. These scientists work in the e-Science Studio/Data Science
@@
 science project. Some fraction of the developer time is dedicated to the
-developers' personal interests as well as instructions.
+developers' personal interests as well as instruction.
 
-The goal for Sloan is to provide success stories and demonstrable value
-in the presence of data scientists on university staff. These stories are the
+The goal for Sloan is to obtain success stories and demonstrable value
+of the presence of data scientists on university staff. These stories are the
 basis for arguments to the host organization. This is an effort to create
@@
 funding, just as DNA sequencing is for core facilities, if it becomes apparent
-that this gives competitive advantages to a university's research effort.
+that this gives competitive advantage to a university's research effort.
 
@@
 further smooth this. (This also provides more stable jobs for those who hold
-the skills, but arguments about making life nicer for postdocs will not help
-persuade the funders or the PIs.)
+these skills, but arguments about making life nicer for postdocs will not help
+persuade funders or PIs!)
 
-There is some questions about the most effective duration and percent of full
+There is some question about the most effective duration and percent of full
 time for a programmer's work on a project. At least three months is necessary
@@
 
-There is not yet clear answer to the scale of aggregation needed to make
+There is, as yet, no clear answer as to the scale of aggregation needed to make
 such a program work. A university wide program allows enough scale to be robust
@@
 
-However this area develops, the need for funding to secure software as a
-cyberinfrastructure component is clear. \choinote{last sentence unclear} Funding which permits code to be
-refactored, tidied and optimized is rare; this is often done ``on the sly'' \choinote{on the fly?} in a
+However this area develops, the need for funding for software as a
+cyberinfrastructure component is clear.  Funding that permits code to be
+refactored, tidied, and optimized is rare; this is often done ``on the sly'' in a
 scientifically focused grant. The UK EPSRC's ``software for the future'' call,
@@
 Perhaps a version of ``software carpentry'' targeted at those PIs who are
-awarded or apply for software grants could be valuable here.
+awarded or apply for software-intensive grants could be valuable here.
 
@@
 individuals who, at distinct universities, had RSEs (in effect if not in name)
-and these were funded under differing models.
+who were funded under differing models.
 
@@
 In this respect, information gathering via a survey and subsequent analysis could be
-very useful. We would need to assemble a list of targeted individuals (what
-positions and ranks are likely to know and care enough to respond?). Perhaps the
+very useful. We would need to assemble a list of targeted individuals. (what
+positions and ranks are likely to know and care enough to respond?) Perhaps the
 Science Gateway Institute has already acquired information that could be helpful
-to advance this issue, and/or craft a proper survey and target some individuals.
+to advance this issue, and/or craft a proper survey and suggest target individuals.
 

### 2015-11-25 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Appendix H:  A few changes.

@@
 software for the large supercomputing resource (ARCHER). However, while the
-support can come the staff of the Edinburgh Parallel Computing Centre,
-who host the computer, this ``embedded CSE'' resource also funds the
+support can come from the staff of the Edinburgh Parallel Computing Centre,
+who hosts the computer, this ``embedded CSE'' resource also funds the
 programming coming from local groups. This has been very helpful in providing
-funding to establish local groups. These work best when these groups develop good
+funding to establish local groups. It works best when these groups develop good
 collaborations with national cyberinfrastructure pools. When an organization
@@
 goal is to make the available stack more robust and richer for everyone. They
-offer short cycles of funding that can have use creating some useful
+offer short cycles of funding for creating some useful
 functionality that shows a diversity of input and draws from across disciplines as a key metric,  If this metric is met successfully, 
@@
 In this respect, information gathering via a survey and subsequent analysis could be
-very useful. We would need to assemble a list of targeted individuals. (what
+very useful. We would need to assemble a list of targeted individuals. (What
 positions and ranks are likely to know and care enough to respond?) Perhaps the

### 2015-11-25 WSSSPE3/report/wg_appendix_training.tex
Appendix I:  Reformat. No content change.

@@
 
-Nick Jones\footnote{email: \href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}} will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+Nick Jones\footnote{email:
+\href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}} will serve as the
+point of contact for this working group, and be responsible for ensuring timely
+progress of the planned actions.
 
@@
 
-While little training focuses specifically on this outcome \katznote{sustainable software?}, a variety of training activities increase researcher awareness of and engagement with software professionals and software engineering practices.
-Research Software Engineers are being recognized as critical contributors to high quality research; the pathway to acquire and master the relevant skills isn't yet clear; equally those skills required by researchers in general are also not commonly understood nor routinely developed.
+While little training focuses specifically on this outcome \katznote{sustainable
+software?}, a variety of training activities increase researcher awareness of
+and engagement with software professionals and software engineering practices.
+Research Software Engineers are being recognized as critical contributors to
+high quality research; the pathway to acquire and master the relevant skills
+isn't yet clear; equally those skills required by researchers in general are
+also not commonly understood nor routinely developed.
 
-The group's discussion explored a rapidly growing array of training that is seen to contribute to sustainable software.
-The offerings are diverse, including: self paced online modules focused around specific tools; single and multiple day training workshops that raise awareness of a tool chain to support collaborative and shared software development within a research workflow; block courses specializing on particular methods, technologies, and applications; academic programs at undergraduate and masters levels; doctoral training programs that in part contain requisite skills training activities.
+The group's discussion explored a rapidly growing array of training that is seen
+to contribute to sustainable software. The offerings are diverse, including:
+self paced online modules focused around specific tools; single and multiple day
+training workshops that raise awareness of a tool chain to support collaborative
+and shared software development within a research workflow; block courses
+specializing on particular methods, technologies, and applications; academic
+programs at undergraduate and masters levels; doctoral training programs that in
+part contain requisite skills training activities.
 
-While some of this training focuses on applying software engineering practices within the context of research, meeting the values and goals of research are less often incorporated as explicit learning outcomes.
-With software (and similarly, data) often the only tangible artifact of a research method or protocol, the dependency between software applications, and the quality of research adds complexity to the learners journey.
-In recognition of the longer term investment required by researchers to integrate such skills into their research practices, many activities are focusing on emotionally engaging researchers and cohorts, to build a sense of shared purpose beyond the obvious goal of technical skill acquisition.
+While some of this training focuses on applying software engineering practices
+within the context of research, meeting the values and goals of research are
+less often incorporated as explicit learning outcomes. With software (and
+similarly, data) often the only tangible artifact of a research method or
+protocol, the dependency between software applications, and the quality of
+research adds complexity to the learners journey. In recognition of the longer
+term investment required by researchers to integrate such skills into their
+research practices, many activities are focusing on emotionally engaging
+researchers and cohorts, to build a sense of shared purpose beyond the obvious
+goal of technical skill acquisition.
 
-In reviewing current training activities, the group identified a variety of perspectives seen as useful in positioning activities in ways to better communicate why and when best to apply each activity.
-Training can be categorized on a variety of spectra, with content and delivery focused from: programming to research; basic to advanced; technical to emotional; informal to formal; self paced to participative.
-A few attempts have been made to situate a cross section of training activities within such dimensions, creating easier means of communicating the value of any specific opportunity and the pathways across opportunities over time.
+In reviewing current training activities, the group identified a variety of
+perspectives seen as useful in positioning activities in ways to better
+communicate why and when best to apply each activity. Training can be
+categorized on a variety of spectra, with content and delivery focused from:
+programming to research; basic to advanced; technical to emotional; informal to
+formal; self paced to participative. A few attempts have been made to situate a
+cross section of training activities within such dimensions, creating easier
+means of communicating the value of any specific opportunity and the pathways
+across opportunities over time.
 
-Evaluation of training delivery and outcomes is seen as a weakness common to most non academic training activities.
-Opportunities for measuring success in delivering training start simply with collecting a Net Promoter Score, which lets those delivering training know whether attendees are likely to recommend the training to others.
-In looking at the longer term outcomes for the learner, frameworks such as Blooms Taxonomy and Kirkpatrick Evaluations offer possible approaches.
+Evaluation of training delivery and outcomes is seen as a weakness common to
+most non academic training activities. Opportunities for measuring success in
+delivering training start simply with collecting a Net Promoter Score, which
+lets those delivering training know whether attendees are likely to recommend
+the training to others. In looking at the longer term outcomes for the learner,
+frameworks such as Blooms Taxonomy and Kirkpatrick Evaluations offer possible
+approaches.
 
-In this latter case of formal evaluation, ownership of evaluation as a component of career development for any researcher appears mostly absent.
-While academic research institutions have professional development centers to support research staff, the skills taught which might impact on sustainable software are limited at best, and lack a clear and coherent development pathway.
+In this latter case of formal evaluation, ownership of evaluation as a component
+of career development for any researcher appears mostly absent. While academic
+research institutions have professional development centers to support research
+staff, the skills taught which might impact on sustainable software are limited
+at best, and lack a clear and coherent development pathway.
 
-Coordination of these training projects will depend on buy in from a broad range of training program and activity leaders, suggesting a key opportunity lies in identifying and bringing together these people on a regular basis.
+Coordination of these training projects will depend on buy in from a broad range
+of training program and activity leaders, suggesting a key opportunity lies in
+identifying and bringing together these people on a regular basis.
 
@@
 
-Software skills are needed by an increasing array of researchers and fields.
-The training arc is not well-defined, with a sometimes baffling array of training opportunities responding to various facets of skill deficit and need.
-Given this current complexity, coordination across training projects would create common frames of reference, communicating and integrating activities to better serve the needs of researchers.
+Software skills are needed by an increasing array of researchers and fields. The
+training arc is not well-defined, with a sometimes baffling array of training
+opportunities responding to various facets of skill deficit and need. Given this
+current complexity, coordination across training projects would create common
+frames of reference, communicating and integrating activities to better serve
+the needs of researchers.
 
-Building this community could lift the maturity of training projects and capabilities, enabling more advanced approaches to address key gaps in evaluation, career development, and a lift in the standard of research practices.
+Building this community could lift the maturity of training projects and
+capabilities, enabling more advanced approaches to address key gaps in
+evaluation, career development, and a lift in the standard of research
+practices.
 
-In aiming at these opportunities it will be necessary to find the means to support those involved in leading training activities to allocate time to coordination activities, which will often sit beyond their current scope of responsibility.
+In aiming at these opportunities it will be necessary to find the means to
+support those involved in leading training activities to allocate time to
+coordination activities, which will often sit beyond their current scope of
+responsibility.
 
-These activities are also distributed globally, with no single country or region offering a comprehensive set of capabilities and initiatives.
-Any coordination activity will therefore need to raise the profile of the opportunity gap with relevant research funders and policy makers.
+These activities are also distributed globally, with no single country or region
+offering a comprehensive set of capabilities and initiatives. Any coordination
+activity will therefore need to raise the profile of the opportunity gap with
+relevant research funders and policy makers.
 
@@
 
-The goal of the following next steps is to quickly test whether there is interest in establishing a community committed to increasing the degree of coordination across training projects.
+The goal of the following next steps is to quickly test whether there is
+interest in establishing a community committed to increasing the degree of
+coordination across training projects.
 
 \begin{enumerate}
-    \item Hold a virtual meeting by December 2015, to bring together a broader group of interest in this topic, with specific goals to:
+
+\item Hold a virtual meeting by December 2015, to bring together a broader group
+of interest in this topic, with specific goals to:
+
 	\begin{enumerate}
-        \item Identify programs with existing activity aimed at integrating across training projects.
-        \item Identify training projects with an interest in participating in coordination efforts.
-        \item Identify funding opportunities to bring together training program and project leaders to identify shared goals for future coordination of activities.
-        \item Agree a communications plan to qualify whether programs, projects, and funders are interested in engaging and committing to ongoing activities.
+
+	\item Identify programs with existing activity aimed at integrating across
+	training projects.
+
+	\item Identify training projects with an interest in participating in
+	coordination efforts.
+
+	\item Identify funding opportunities to bring together training program and
+	project leaders to identify shared goals for future coordination of activities.
+
+	\item Agree a communications plan to qualify whether programs, projects, and
+	funders are interested in engaging and committing to ongoing activities.
+
 	\end{enumerate}
+
 \item Review progress within 3 months, to establish next steps, if any.
+
 \end{enumerate}
@@
 
-If the group moves from early stage formation into working towards shared goals, expertise will likely be required in pedagogy and training evaluation
+If the group moves from early stage formation into working towards shared goals,
+expertise will likely be required in pedagogy and training evaluation
 
@@
 \begin{enumerate}
+
 \item During October, Nick Jones and Erin Robinson to draft WSSSPE3 report back.
-    \item Before December 2015, Nick Jones and Erin Robinson to call a meeting of the broader group, to review key next steps. 
-    \item 1st quarter 2016 - if willing parties identified, draft workshop proposal and identify relevant forum, including future WSSSPE events
+
+\item Before December 2015, Nick Jones and Erin Robinson to call a meeting of
+the broader group, to review key next steps.
+
+\item 1st quarter 2016 - if willing parties identified, draft workshop proposal
+and identify relevant forum, including future WSSSPE events
+
 \end{enumerate}
@@
 
-Workshop/RCN travel funding to bring together key program, project, and funder representatives from across North America, EU, UK, Australasia.
-In addition, funding to support work on better defining the landscape of training activities, the useful perspectives in communicating the value of the varied training projects, and the possible pathways through training activities over time.
+Workshop/RCN travel funding to bring together key program, project, and funder
+representatives from across North America, EU, UK, Australasia. In addition,
+funding to support work on better defining the landscape of training activities,
+the useful perspectives in communicating the value of the varied training
+projects, and the possible pathways through training activities over time.

### 2015-11-25 WSSSPE3/report/wg_appendix_training.tex
Appendix I:  Minor changes.

@@
 \item Nick Jones -- New Zealand eScience Infrastructure
-\item Iain Larmour -- Engineering \& Physical Sciences Research Council (UK)
+\item Iain Larmour -- Engineering \& Physical Sciences Research Council, UK
 \item Erin Robinson -- Foundation for Earth Science
@@
 high quality research; the pathway to acquire and master the relevant skills
-isn't yet clear; equally those skills required by researchers in general are
+is not yet clear; equally those skills required by researchers in general are
 also not commonly understood nor routinely developed.
@@
 to contribute to sustainable software. The offerings are diverse, including:
-self paced online modules focused around specific tools; single and multiple day
+self-paced online modules that focused around specific tools; single and multiple day
 training workshops that raise awareness of a tool chain to support collaborative
@@
 less often incorporated as explicit learning outcomes. With software (and
-similarly, data) often the only tangible artifact of a research method or
+similarly, data), often the only tangible artifact of a research method or
 protocol, the dependency between software applications, and the quality of
-research adds complexity to the learners journey. In recognition of the longer
+research add complexity to the learners journey. In recognition of the longer
 term investment required by researchers to integrate such skills into their
@@
 communicate why and when best to apply each activity. Training can be
-categorized on a variety of spectra, with content and delivery focused from:
+categorized on a variety of spectra, with content and delivery focused on:
 programming to research; basic to advanced; technical to emotional; informal to
-formal; self paced to participative. A few attempts have been made to situate a
+formal; self-paced to participative. A few attempts have been made to situate a
 cross section of training activities within such dimensions, creating easier
@@
 Evaluation of training delivery and outcomes is seen as a weakness common to
-most non academic training activities. Opportunities for measuring success in
+most non-academic training activities. Opportunities for measuring success in
 delivering training start simply with collecting a Net Promoter Score, which
-lets those delivering training know whether attendees are likely to recommend
+let those delivering training know whether attendees are likely to recommend
 the training to others. In looking at the longer term outcomes for the learner,
-frameworks such as Blooms Taxonomy and Kirkpatrick Evaluations offer possible
+frameworks such as Bloom's taxonomy and Kirkpatrick's evaluation model offer possible
 approaches.
@@
 
-Coordination of these training projects will depend on buy in from a broad range
+Coordination of these training projects will depend on buy-in from a broad range
 of training program and activity leaders, suggesting a key opportunity lies in
@@
 
-In aiming at these opportunities it will be necessary to find the means to
+In aiming at these opportunities, it will be necessary to find the means to
 support those involved in leading training activities to allocate time to
@@
 
-	\item Identify programs with existing activity aimed at integrating across
+	\item Identify programs with existing activities aimed at integrating across
 	training projects.
@@
 
-	\item Identify funding opportunities to bring together training program and
+	\item Identify funding opportunities to bring together training programs and
 	project leaders to identify shared goals for future coordination of activities.
 
-	\item Agree a communications plan to qualify whether programs, projects, and
+	\item Agree on a communications plan to qualify whether programs, projects, and
 	funders are interested in engaging and committing to ongoing activities.
@@
 
-If the group moves from early stage formation into working towards shared goals,
-expertise will likely be required in pedagogy and training evaluation
+If the group moves from early-stage formation into working towards shared goals,
+expertise will likely be required in pedagogy and training evaluation.
 
@@
 
-\item 1st quarter 2016 - if willing parties identified, draft workshop proposal
-and identify relevant forum, including future WSSSPE events
+\item First quarter 2016 -- if willing parties identified, draft workshop proposal
+and identify a relevant forum, including future WSSSPE events.
 

### 2015-11-25 WSSSPE3/report/wg_appendix_training.tex
small edits to edits

@@
 to contribute to sustainable software. The offerings are diverse, including:
-self-paced online modules that focused around specific tools; single and multiple day
+self-paced online modules focused around specific tools; single and multiple day
 training workshops that raise awareness of a tool chain to support collaborative
@@
 less often incorporated as explicit learning outcomes. With software (and
-similarly, data), often the only tangible artifact of a research method or
-protocol, the dependency between software applications, and the quality of
-research add complexity to the learners journey. In recognition of the longer
+similarly, data) often the only tangible artifact of a research method or
+protocol, the dependency between software applications and the quality of
+research adds complexity to the learner's journey. In recognition of the longer
 term investment required by researchers to integrate such skills into their
@@
 communicate why and when best to apply each activity. Training can be
-categorized on a variety of spectra, with content and delivery focused on:
+categorized on a variety of spectra, with content and delivery ranging across them, for example:
 programming to research; basic to advanced; technical to emotional; informal to
-formal; self-paced to participative. A few attempts have been made to situate a
+formal; and self-paced to participative. A few attempts have been made to situate a
 cross section of training activities within such dimensions, creating easier
@@
 delivering training start simply with collecting a Net Promoter Score, which
-let those delivering training know whether attendees are likely to recommend
+lets those delivering training know whether attendees are likely to recommend
 the training to others. In looking at the longer term outcomes for the learner,
@@
 
-	\item Identify funding opportunities to bring together training programs and
+	\item Identify funding opportunities to bring together training program and
 	project leaders to identify shared goals for future coordination of activities.
@@
 
-\item First quarter 2016 -- if willing parties identified, draft workshop proposal
+\item First quarter 2016 -- if willing parties are identified, draft workshop proposal
 and identify a relevant forum, including future WSSSPE events.

### 2015-11-26 WSSSPE3/report/wg_appendix_software_credit.tex
Appendix J:  Format. No content change.

@@
 
-Kyle Niemeyer\footnote{email: \href{mailto:kyle.niemeyer@oregonstate.edu}{kyle.niemeyer@oregonstate.edu}} will serve as the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+Kyle Niemeyer\footnote{email:
+\href{mailto:kyle.niemeyer@oregonstate.edu}{kyle.niemeyer@oregonstate.edu}} will
+serve as the point of contact for this working group, and be responsible for
+ensuring timely progress of the planned actions.
 
@@
 
-The following section summarizes the working group's discussion based on contributions prior to the meeting~\cite{WSSSPE3-SC-github-issues} and the collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
+The following section summarizes the working group's discussion based on
+contributions prior to the meeting~\cite{WSSSPE3-SC-github-issues} and the
+collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
 Please refer to the original sources for the unedited discussions.
 
-Initial discussions focused on both various mechanisms for, and the philosophical approach behind, crediting software in scientific papers.
-These began with proposals for various ways to credit software (or other research products including data) that contributed more significantly than a generic citation, including
+Initial discussions focused on both various mechanisms for, and the
+philosophical approach behind, crediting software in scientific papers. These
+began with proposals for various ways to credit software (or other research
+products including data) that contributed more significantly than a generic
+citation, including
 \begin{itemize}
-    \item A hierarchy of citations, with a ``substantial'' citation category to indicate software or data that played a more significant role in the research;
-    \item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns contriponents (contributors and components) various weights depending on their level of importance; and
-    \item Project CRediT~\cite{projectcredit}, which assigns roles to paper authors based on their specific contributions; and
-    \item Mozilla Science Lab's recently introduced Contributorship Badges for Science~\cite{Mozilla_badges}, which provide a badge---associated with ORCID~\cite{orcid}---based on author contributions in a similar manner to Project CRediT.
+
+\item A hierarchy of citations, with a ``substantial'' citation category to
+indicate software or data that played a more significant role in the research;
+
+\item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns
+contriponents (contributors and components) various weights depending on their
+level of importance; and
+
+\item Project CRediT~\cite{projectcredit}, which assigns roles to paper authors
+based on their specific contributions; and
+
+\item Mozilla Science Lab's recently introduced Contributorship Badges for
+Science~\cite{Mozilla_badges}, which provide a badge---associated with
+ORCID~\cite{orcid}---based on author contributions in a similar manner to
+Project CRediT.
+
 \end{itemize}
-However, as of this writing, only the Project CRediT roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship Badges~\cite{Mozilla_badges} have been implemented in published papers, though both of these only provide a single ``Software'' or ``Computation'' category associated with software.
-In addition, neither of these options allow for the citation of software itself, but only provide an author contribution related to software.
-The discussion quickly focused on transitive credit as a more quantitative measure of allocating credit to both authors and software, although there were some concerns about authors overestimating their own contributions compared to prior work.
+However, as of this writing, only the Project CRediT
+roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship
+Badges~\cite{Mozilla_badges} have been implemented in published papers, though
+both of these only provide a single ``Software'' or ``Computation'' category
+associated with software. In addition, neither of these options allow for the
+citation of software itself, but only provide an author contribution related to
+software. The discussion quickly focused on transitive credit as a more
+quantitative measure of allocating credit to both authors and software, although
+there were some concerns about authors overestimating their own contributions
+compared to prior work.
 
-The discussion then evolved into philosophical questions about the importance or reliance of a particular work on prior science, materials, or software---in other words, is there a difference between depending on prior scientific advances and depending on particular software (or experimental equipment)?
-Multiple contributors converged on the conclusion that unique capabilities require some additional credit.
-The---albeit limited---consensus was that if a particular study relied on the unique capabilities of software, data, or an experimental apparatus, then the authors or developers that created this capability should be credited somehow.
+The discussion then evolved into philosophical questions about the importance or
+reliance of a particular work on prior science, materials, or software---in
+other words, is there a difference between depending on prior scientific
+advances and depending on particular software (or experimental equipment)?
+Multiple contributors converged on the conclusion that unique capabilities
+require some additional credit. The---albeit limited---consensus was that if a
+particular study relied on the unique capabilities of software, data, or an
+experimental apparatus, then the authors or developers that created this
+capability should be credited somehow.
 
-The group also agreed on the fact that additional data was required to support the assertion that software was not being sufficiently cited in the literature.
-In particular, this issue seemed to be field-dependent.
-For example, as shown by study of Howison and Bullard~\cite{Howison2015}, in the field of biology the most-cited papers appear to be those describing scientific software.
-However, this may not---and likely is not---the case in other fields, nor is it clear whether developers of scientific software, even in the case of the biology field, are receiving sufficient credit for their efforts.
+The group also agreed on the fact that additional data was required to support
+the assertion that software was not being sufficiently cited in the literature.
+In particular, this issue seemed to be field-dependent. For example, as shown by
+study of Howison and Bullard~\cite{Howison2015}, in the field of biology the
+most-cited papers appear to be those describing scientific software. However,
+this may not---and likely is not---the case in other fields, nor is it clear
+whether developers of scientific software, even in the case of the biology
+field, are receiving sufficient credit for their efforts.
 
-During the first day of WSSSPE3 breakout sessions, the group discussed the Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for scientific software.
-That system assigns unique Digital Object Identifiers (DOIs)---the same system used for scientific publications---to all content (e.g., movies, television shows) and contributors, along with relevant metadata.
-One important use of this system is to track rights and credit for contributors to entertainment works in order to distribute revenues---similar to the proposed transitive credit concept.
+During the first day of WSSSPE3 breakout sessions, the group discussed the
+Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
+scientific software. That system assigns unique Digital Object Identifiers
+(DOIs)---the same system used for scientific publications---to all content
+(e.g., movies, television shows) and contributors, along with relevant metadata.
+One important use of this system is to track rights and credit for contributors
+to entertainment works in order to distribute revenues---similar to the proposed
+transitive credit concept.
 
-The group also discussed separating quantitative measures (e.g., number of citations) from the value of a work in order to give credit, moving towards qualitative or anecdotal evidence of value.
-Other topics brought up included a form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and using market penetration or adoption rate in a community as a metric, although it was not clear how this would be measured.
-Finally, the concept a software tool's uniqueness or indispensability to a community was mentioned, with value being characterized by a particular piece of software either offering unique capabilities or doing something better, faster, or with less computational requirements than other offerings.
+The group also discussed separating quantitative measures (e.g., number of
+citations) from the value of a work in order to give credit, moving towards
+qualitative or anecdotal evidence of value. Other topics brought up included a
+form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and
+using market penetration or adoption rate in a community as a metric, although
+it was not clear how this would be measured. Finally, the concept a software
+tool's uniqueness or indispensability to a community was mentioned, with value
+being characterized by a particular piece of software either offering unique
+capabilities or doing something better, faster, or with less computational
+requirements than other offerings.
 
-On the second day of WSSSPE3, the working group decided to put aside the taxonomy of contributions and focus on software citations to ensure developers receive credit (regardless of contribution).
-Eventually, once standardizing software citations, the goal would be to return to establishing different roles\slash contributions for this credit.
-Following this decision, the group identified two necessary actions to move forward:
+On the second day of WSSSPE3, the working group decided to put aside the
+taxonomy of contributions and focus on software citations to ensure developers
+receive credit (regardless of contribution). Eventually, once standardizing
+software citations, the goal would be to return to establishing different
+roles\slash contributions for this credit. Following this decision, the group
+identified two necessary actions to move forward:
 \begin{enumerate}
-    \item standardizing a citation file or some other form of metadata associated with software, and
+
+\item standardizing a citation file or some other form of metadata associated
+with software, and
+
 \item standardizing the way to cite software (used directly) in papers.
+
 \end{enumerate}
@@
 
-At a minimum, the metadata needed for software citation (aside from the DOI for the software itself) includes:
+At a minimum, the metadata needed for software citation (aside from the DOI for
+the software itself) includes:
 \begin{itemize}
@@
 This information would then be contained in a citation file, e.g., as part of
-the GitHub repository.
-The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
-metadata schemes in JSON and XML for scientific software and code, and 
-implementing transitive credit via JSON-LD~\cite{wssspe2_katz}.
-Some questions arose about how this information would be stored for 
-closed-source software.
+the GitHub repository. The group also discussed similar efforts such as
+Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an
+attempt to codify minimal metadata schemes in JSON and XML for scientific
+software and code, and implementing transitive credit via
+JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
+would be stored for closed-source software.
 
 As one mechanism for constructing accurate contributor lists from existing
-project contributors, the group discussed associating GitHub accounts---as 
-well as accounts on Bitbucket, CodePlex, and other repositories for 
-open-source scientific software---accounts with ORCID.
-However, a (quick) response from GitHub (via Arfon Smith) indicated that 
-this might not be possible in the near future: ``GitHub doesn't have any 
-plans to allow ORCID accounts to be associated with GitHub user accounts.''
+project contributors, the group discussed associating GitHub accounts---as well
+as accounts on Bitbucket, CodePlex, and other repositories for open-source
+scientific software---accounts with ORCID. However, a (quick) response from
+GitHub (via Arfon Smith) indicated that this might not be possible in the near
+future: ``GitHub doesn't have any plans to allow ORCID accounts to be associated
+with GitHub user accounts.''
 
@@
 publications can be found in various scientific communities---notably,
-representative samples can be found in astronomy~\cite{astronomy_SW_examples} and biology~\cite{Howison2015}.
-The group recommended collecting similar examples from other communities, then 
-developing a software citation principles document in concert with the 
-FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details), 
-following the model of the FORCE11 Data Citation Principles 
-document~\cite{DataCitation2014}.
+representative samples can be found in astronomy~\cite{astronomy_SW_examples}
+and biology~\cite{Howison2015}. The group recommended collecting similar
+examples from other communities, then developing a software citation principles
+document in concert with the FORCE11 Software Citation Working Group (see
+\S\ref{SC:plan} for more details), following the model of the FORCE11 Data
+Citation Principles document~\cite{DataCitation2014}.
 
-The group also discussed briefly whether software used directly in 
-a publication---whether to perform simulation, analysis, or as a dependency 
-for new software developed---should be distinguished from other references 
-due to the dependence of the study on these research artifacts.
-Suggestions included a separate list of citations (with DOIs) for software 
-and other research objects that serve this sort of ``vital'' role.
-Similar recommendations were made by the credit breakout group at 
-WSSSPE2~\cite{WSSSPE2}.
+The group also discussed briefly whether software used directly in a
+publication---whether to perform simulation, analysis, or as a dependency for
+new software developed---should be distinguished from other references due to
+the dependence of the study on these research artifacts. Suggestions included a
+separate list of citations (with DOIs) for software and other research objects
+that serve this sort of ``vital'' role. Similar recommendations were made by the
+credit breakout group at WSSSPE2~\cite{WSSSPE2}.
 
-Finally, although a discrete task from software citations themselves, significant 
-discussion focused on ensuring software citations are indexed in the same 
-manner as publications, allowing the construction of a corresponding 
-software citation network.
-Currently, software releases can receive citable DOIs via 
-Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these 
-citations are not processed by indexers such as Web of Science, Scopus, 
-or Google Scholar.
-Thus, either in parallel or following the primary task, the group will 
-need to reach out to these organizations 
-Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
-during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
-also does not yet have an internal identifier for software or data (but 
-needs\slash plans to add this support).
-Taylor said they prefer a ``software article'' with the usual 
-article metadata (e.g., authors, citations), and mentioned Zenodo as an 
-example -- this proposal seemed to align with group discussions.
-Taylor also mentioned another benefit of the software and associated DOI 
-on GitHub: in addition to a citation, could obtain statistics on 
-usage/downloads/forks, which happens to be what Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.
+Finally, although a discrete task from software citations themselves,
+significant discussion focused on ensuring software citations are indexed in the
+same manner as publications, allowing the construction of a corresponding
+software citation network. Currently, software releases can receive citable DOIs
+via Zenodo~\cite{zenodo-web} and figshare~\cite{figshare-web}; however, these
+citations are not processed by indexers such as Web of Science, Scopus, or
+Google Scholar. Thus, either in parallel or following the primary task, the
+group will need to reach out to these organizations Initial conversations with
+Elsevier\slash Scopus (via Michael Taylor) during WSSSPE3 clarified that Scopus
+is not yet DataCite DOI aware, and also does not yet have an internal identifier
+for software or data (but needs\slash plans to add this support). Taylor said
+they prefer a ``software article'' with the usual article metadata (e.g.,
+authors, citations), and mentioned Zenodo as an example -- this proposal seemed
+to align with group discussions. Taylor also mentioned another benefit of the
+software and associated DOI on GitHub: in addition to a citation, could obtain
+statistics on usage/downloads/forks, which happens to be what
+Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.
 
@@
 
-Modern scientific research relies considerably on software, and ensuring 
-that developers of such scientific software receive credit for their efforts 
-will encourage additional creation and maintenance. \katznote{the prev sentence is good, but feels a bit odd in the middle of the appendix, rather than at the beginning}
-There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
-Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
-Some software providers apply for DOIs but it is still not widely adopted.
-Also, there is no mechanism to cite software dependencies within software.
+Modern scientific research relies considerably on software, and ensuring that
+developers of such scientific software receive credit for their efforts will
+encourage additional creation and maintenance. \katznote{the prev sentence is
+good, but feels a bit odd in the middle of the appendix, rather than at the
+beginning} There currently is no standard mechanism for citing software or
+receiving credit for software (akin to publications/citations). Software is
+eligible for DOI assignment, but DOI metadata fields are not well tuned or
+standardized for software (vs. publications). Some software providers apply for
+DOIs but it is still not widely adopted. Also, there is no mechanism to cite
+software dependencies within software.
 
-Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, Google Scholar) do not currently support software/data document types or DataCite DOIs.
-Therefore, even with universal association of scientific software with DOIs and standardized practices for citing software in publications, software citations will not be indexed in the same manner as traditional publications.
+Major obstacles include the fact that indexers (e.g., Scopus, Web of Science,
+Google Scholar) do not currently support software/data document types or
+DataCite DOIs. Therefore, even with universal association of scientific software
+with DOIs and standardized practices for citing software in publications,
+software citations will not be indexed in the same manner as traditional
+publications.
 
-Although this working group's discussions at WSSSPE3 did not focus much on the topic of tenure, the group recognized that there is no standard policy---generally even within a single university---for software products to be included in promotion and tenure dossiers.
-Thus, it may be difficult to encourage valuing software contributions across the United States or United Kingdom and globally; furthermore, stakeholders are typically not tenured and thus may not be influential enough to change the status quo.
+Although this working group's discussions at WSSSPE3 did not focus much on the
+topic of tenure, the group recognized that there is no standard
+policy---generally even within a single university---for software products to be
+included in promotion and tenure dossiers. Thus, it may be difficult to
+encourage valuing software contributions across the United States or United
+Kingdom and globally; furthermore, stakeholders are typically not tenured and
+thus may not be influential enough to change the status quo.
 
@@
 \begin{enumerate}
-\item Hold virtual meeting to determine group members responsible\slash willing to work on the following tasks, to be organized within one month of the workshop
-\item Compile best practices of software citation across multiple disciplines, including journals and communities of interest\slash practice in the research world, to begin by December 2015
-\item Compile examples of including other products in promotion and tenure dossier, to begin by December 2015
-\item Draft the Software Citation Principles document (including citation metadata file), by April 2016
-\item Publish\slash release the Software Citation Principles document, by August 2016
-\item Reach out to journals, publishers, teachers\slash educators, indexers, and professional societies---likely through meetings with key groups, to begin by September 2016
+
+\item Hold virtual meeting to determine group members responsible\slash willing
+to work on the following tasks, to be organized within one month of the workshop
+
+\item Compile best practices of software citation across multiple disciplines,
+including journals and communities of interest\slash practice in the research
+world, to begin by December 2015
+
+\item Compile examples of including other products in promotion and tenure
+dossier, to begin by December 2015
+
+\item Draft the Software Citation Principles document (including citation
+metadata file), by April 2016
+
+\item Publish\slash release the Software Citation Principles document, by August
+2016
+
+\item Reach out to journals, publishers, teachers\slash educators, indexers, and
+professional societies---likely through meetings with key groups, to begin by
+September 2016
 
@@
 
-The WSSSPE breakout group plans to join efforts related to citing software with the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11 Software Citation Working Group, \url{https://www.force11.org/group/software-citation-working-group}}; Kyle Niemeyer formally requested the merging of these groups following the meeting.
-However, some future plans of the WSSSPE group fall outside the scope of FORCE11-SCWG, which covers software citation practices.
-These activities include working with indexers such as Web of Science and Scopus to index software citations archived on, e.g., Zenodo or FigShare, and pursuing the development of an open indexing service; such plans will be pursued either separately or through the formation of follow-on FORCE11 working groups.
+The WSSSPE breakout group plans to join efforts related to citing software with
+the FORCE11 Software Citation Working Group (FORCE11-SCWG)\footnote{FORCE11
+Software Citation Working Group,
+\url{https://www.force11.org/group/software-citation-working-group}}; Kyle
+Niemeyer formally requested the merging of these groups following the meeting.
+However, some future plans of the WSSSPE group fall outside the scope of
+FORCE11-SCWG, which covers software citation practices. These activities include
+working with indexers such as Web of Science and Scopus to index software
+citations archived on, e.g., Zenodo or FigShare, and pursuing the development of
+an open indexing service; such plans will be pursued either separately or
+through the formation of follow-on FORCE11 working groups.
 
-The group will primarily communicate electronically, with Kyle Niemeyer responsible for ensuring regular progress.
+The group will primarily communicate electronically, with Kyle Niemeyer
+responsible for ensuring regular progress.
 
@@
 
-The near-term actions of the group, focused mainly on software citation, do not require any additional resources.
-However, connections with publishers and indexers will be needed to pursue related activities, although the FORCE11-SCWG may satisfy this need; in addition, some members of the group already reached out to relevant contacts.
-Funding may be needed to organize meetings or for group members to attend relevant meetings, as discussed further below.
+The near-term actions of the group, focused mainly on software citation, do not
+require any additional resources. However, connections with publishers and
+indexers will be needed to pursue related activities, although the FORCE11-SCWG
+may satisfy this need; in addition, some members of the group already reached
+out to relevant contacts. Funding may be needed to organize meetings or for
+group members to attend relevant meetings, as discussed further below.
 
@@
 
-Following the meeting, Kyle Niemeyer formally requested the merging of software citation activities with FORCE11-SCWG.
-Within a month of the meeting, Kyle will organize a virtual meeting of the group and lead the division of responsibilities for compiling existing practices of software citation and including software\slash products in promotion and tenure dossiers.
-Building off of these efforts, the next major milestone is drafting the Software Citation Principles document in collaboration with the SCWG, targeted for April 2016.
-While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the efforts of that group towards the Software Citation Principles document, Kyle will help coordinate contributions from the WSSSPE group members.
+Following the meeting, Kyle Niemeyer formally requested the merging of software
+citation activities with FORCE11-SCWG. Within a month of the meeting, Kyle will
+organize a virtual meeting of the group and lead the division of
+responsibilities for compiling existing practices of software citation and
+including software\slash products in promotion and tenure dossiers. Building off
+of these efforts, the next major milestone is drafting the Software Citation
+Principles document in collaboration with the SCWG, targeted for April 2016.
+While the existing directors of the SCWG, Arfon Smith and Dan Katz, lead the
+efforts of that group towards the Software Citation Principles document, Kyle
+will help coordinate contributions from the WSSSPE group members.
 
@@
 
-Some funding would be useful to support primarily travel to conferences for group meetings (e.g., FORCE2016\footnote{FORCE2016, \url{https://www.force11.org/meetings/force2016}}), and to hold meetings to bring together both group members and key stakeholders (e.g., journals, publishers, professional societies, indexers).
-In addition, funding would be desired to support group members' time to perform work towards the key steps described previously.
+Some funding would be useful to support primarily travel to conferences for
+group meetings (e.g., FORCE2016\footnote{FORCE2016,
+\url{https://www.force11.org/meetings/force2016}}), and to hold meetings to
+bring together both group members and key stakeholders (e.g., journals,
+publishers, professional societies, indexers). In addition, funding would be
+desired to support group members' time to perform work towards the key steps
+described previously.

### 2015-11-26 WSSSPE3/report/wg_appendix_software_credit.tex
Appendix J:  A handful of changes.

@@
 collaborative notes taken during the meeting~\cite{WSSSPE3-SC-google-notes}.
-Please refer to the original sources for the unedited discussions.
+Please refer to the original sources for the unedited discussions if necessary.
 
@@
 contriponents (contributors and components) various weights depending on their
-level of importance; and
+level of importance;
 
@@
 \end{itemize}
-However, as of this writing, only the Project CRediT
+However, as of this writing, only Project CRediT
 roles~\cite{McCall2015_credit,Lin2015_credit} and Contributorship
-Badges~\cite{Mozilla_badges} have been implemented in published papers, though
+Badges~\cite{Mozilla_badges} have been implemented for published papers, and
 both of these only provide a single ``Software'' or ``Computation'' category
-associated with software. In addition, neither of these options allow for the
+associated with software. In addition, neither of these options allows for the
 citation of software itself, but only provide an author contribution related to
@@
 In particular, this issue seemed to be field-dependent. For example, as shown by
-study of Howison and Bullard~\cite{Howison2015}, in the field of biology the
+a study of Howison and Bullard~\cite{Howison2015}, in the field of biology, the
 most-cited papers appear to be those describing scientific software. However,
@@
 (e.g., movies, television shows) and contributors, along with relevant metadata.
-One important use of this system is to track rights and credit for contributors
+One important use of this system is to track rights and credits for contributors
 to entertainment works in order to distribute revenues---similar to the proposed
@@
 taxonomy of contributions and focus on software citations to ensure developers
-receive credit (regardless of contribution). Eventually, once standardizing
-software citations, the goal would be to return to establishing different
+receive credit (regardless of contribution). Eventually, once 
+software citations are standardized, the goal would be to return to establishing different
 roles\slash contributions for this credit. Following this decision, the group
@@
 
-At a minimum, the metadata needed for software citation (aside from the DOI for
-the software itself) includes:
+At a minimum, the metadata needed for software citation %(aside from the DOI for
+%the software itself) \choinote{DOI is in the list below}
+includes:
 \begin{itemize}
@@
 as accounts on Bitbucket, CodePlex, and other repositories for open-source
-scientific software---accounts with ORCID. However, a (quick) response from
+scientific software---with ORCID accounts. However, a (quick) response from
 GitHub (via Arfon Smith) indicated that this might not be possible in the near
@@
 Google Scholar. Thus, either in parallel or following the primary task, the
-group will need to reach out to these organizations Initial conversations with
-Elsevier\slash Scopus (via Michael Taylor) during WSSSPE3 clarified that Scopus
+group will need to reach out to these organizations. Initial conversations with
+Elsevier\slash Scopus via Michael Taylor during WSSSPE3 clarified that Scopus
 is not yet DataCite DOI aware, and also does not yet have an internal identifier
@@
 authors, citations), and mentioned Zenodo as an example -- this proposal seemed
-to align with group discussions. Taylor also mentioned another benefit of the
-software and associated DOI on GitHub: in addition to a citation, could obtain
+to align with our group discussions. Taylor also mentioned another benefit of the
+software and associated DOI on GitHub: in addition to a citation, one could obtain
 statistics on usage/downloads/forks, which happens to be what
@@
 eligible for DOI assignment, but DOI metadata fields are not well tuned or
-standardized for software (vs. publications). Some software providers apply for
-DOIs but it is still not widely adopted. Also, there is no mechanism to cite
+standardized for software (vs.\ publications). Some software providers apply for
+DOIs, but the identifiers are still not widely adopted. Also, there is no mechanism to cite
 software dependencies within software.
@@
 \item Hold virtual meeting to determine group members responsible\slash willing
-to work on the following tasks, to be organized within one month of the workshop
+to work on the following tasks, to be organized within one month of the workshop.
 
@@
 including journals and communities of interest\slash practice in the research
-world, to begin by December 2015
+world, to begin by December 2015.
 
 \item Compile examples of including other products in promotion and tenure
-dossier, to begin by December 2015
+dossier, to begin by December 2015.
 
 \item Draft the Software Citation Principles document (including citation
-metadata file), by April 2016
+metadata file), by April 2016.
 
 \item Publish\slash release the Software Citation Principles document, by August
-2016
+2016.
 
@@
 professional societies---likely through meetings with key groups, to begin by
-September 2016
+September 2016.
 
@@
 Following the meeting, Kyle Niemeyer formally requested the merging of software
-citation activities with FORCE11-SCWG. Within a month of the meeting, Kyle will
+citation activities with FORCE11-SCWG. Within a month of the meeting, Niemeyer will
 organize a virtual meeting of the group and lead the division of

### 2015-11-26 WSSSPE3/report/wg_appendix_software_credit.tex
small edits to edits

@@
 (e.g., movies, television shows) and contributors, along with relevant metadata.
-One important use of this system is to track rights and credits for contributors
+One important use of this system is to track rights and credit for contributors
 to entertainment works in order to distribute revenues---similar to the proposed
@@
 authors, citations), and mentioned Zenodo as an example -- this proposal seemed
-to align with our group discussions. Taylor also mentioned another benefit of the
+to align with our group's discussions. Taylor also mentioned another benefit of the
 software and associated DOI on GitHub: in addition to a citation, one could obtain
@@
 standardized for software (vs.\ publications). Some software providers apply for
-DOIs, but the identifiers are still not widely adopted. Also, there is no mechanism to cite
+DOIs, but this is not widely adopted. Also, there is no mechanism to cite
 software dependencies within software.

### 2015-11-27 WSSSPE3/report/wg_appendix_publishing_SW.tex
Appendix K. No content change.
Appendix K:  A handful of changes.

@@
 
-Steven R.\ Brandt\footnote{email: \href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}} will serve as the point of contact for this working group.
+Steven R.\ Brandt\footnote{email:
+\href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}} will serve as the point
+of contact for this working group.
 
@@
 \begin{itemize}
+
 \item ACM Transactions on Mathematical Software (TOMS): provides the extra step
 of having reviewers validate the code which was submitted with the publication.
+
 \item The Mathematica Journal: publishes Mathematica notebooks (with equations,
 figures, etc.) directly.
-\item O'Reilly Media: announced that it plans to make IPython Notebooks a first-class
- authoring environment for their publishing program alongside their existing mechanisms.
-\item Nature: offering a list of notebooks published alongside more traditional articles,
- and is looking at ways to make these documents more official. There are, in fact, a
- number of journals that offer ``electronic supplements'' to the more traditionally published
- static articles.
-\item IPython: maintained list of ``reproducible academic publications''~\cite{ipython-pubs}.
-\item KBase: narratives build on IPython\slash Jupyter notebooks to build publications that are
-  reproducible, and can be commented or annotated.
+
+\item O'Reilly Media: announced that it plans to make IPython Notebooks a
+first-class authoring environment for their publishing program alongside their
+existing mechanisms.
+
+\item Nature: offering a list of notebooks published alongside more traditional
+articles, and is looking at ways to make these documents more official. There
+are, in fact, a number of journals that offer ``electronic supplements'' to the
+more traditionally published static articles.
+
+\item IPython: maintained a list of ``reproducible academic
+publications''~\cite{ipython-pubs}.
+
+\item KBase: narratives built on IPython\slash Jupyter notebooks; to build
+publications that are reproducible, and can be commented or annotated.
+
 \end{itemize}
 
-The group also discussed future possibilities for this new publication format which might
-provide advantages:
+The group also discussed future possibilities for a new publication format
+which might provide advantages:
 \begin{itemize}
-\item Journals could be built around an existing, widely used framework thereby reducing
-  the burden of studying code on the part of reviewers (common bits of infrastructure
-  which aren't relevant to the science would be automatically excluded).
-\item Journals might be encouraged to use more metadata, making them easier to mine
-  for various analytical purposes.
-\item RIO is an effort to publish fragmentary results that can subsequently be combined
-  into a single content item.
-\item Papers could be made more understandable. Each equation or technical term could
-  offer a link to a document/tutorial explaining its origin and/or derivation.
-\item So many options for publication currently exist that good science may be getting
-  lost in the noise. Would some sort of ``upvote'' mechanism be of value?
-\item Some sort of Replicated Computation Results badge could be made available to
-  publications that have undergone greater scrutiny (this is already done by TOMS).
+
+\item Journals could be built around an existing, widely used framework thereby
+reducing the burden of studying code on the part of reviewers (common bits of
+infrastructure which are not relevant to the science would be automatically
+excluded).
+
+\item Journals might be encouraged to use more metadata, making them easier to
+mine for various analytical purposes.
+
+\item RIO is an effort to publish fragmentary results that can subsequently be
+combined into a single content item. \choinote{What's RIO?}
+
+\item Papers could be made more understandable. Each equation or technical term
+could be linked to a document/tutorial explaining its origin and/or
+derivation.
+
+\item So many options for publication currently exist that good science may be
+getting lost in the noise. Would some sort of ``upvote'' mechanism be of value?
+
+\item Some sort of Replicated Computation Results badge could be made available
+to publications that have undergone greater scrutiny (this is already done by
+TOMS).
+
 \end{itemize}
@@
 
-The opportunity is to collect a list of current executable papers and
-shine a light on the experiments and development efforts currently underway.
+The opportunity is to collect a list of executable papers and shine a
+light on the experiments and development efforts currently underway.
 
 The only obstacle to this is the difficulty in finding and identifying such
-publications. The Software Sustainability Institute was able to do something similar
-for publications about software by making a public page containing a catalog
-of these publications and enlisting the help of the community to grow the list.
+publications. The Software Sustainability Institute was able to do something
+similar for publications about software by making a public page \choinote{where?} containing a
+catalog of these publications and enlisting the help of the community to grow
+the list.
 
@@
 
-Create the first version of the web page to be displayed on the Software Sustainability
-Institute's website: \url{http://www.software.ac.uk}. \katznote{when?}
+Create the first version of the web page to be displayed on the Software
+Sustainability Institute's website: \url{http://www.software.ac.uk}.
+\katznote{when?}
 
@@
 
-Steven R. Brandt will create a first version of the page within a week or so of the WSSSPE3 conference.
+Steven R.\ Brandt will create a first version of the page within a week or so of
+the WSSSPE3 conference.
 

### 2015-11-27 WSSSPE3/report/wg_appendix_publishing_SW.tex
small changes to changes

@@
 
-\item KBase: narratives built on IPython\slash Jupyter notebooks; to build
+\item KBase: narratives build on IPython\slash Jupyter notebooks to build
 publications that are reproducible, and can be commented or annotated.
@@
 The group also discussed future possibilities for a new publication format
-which might provide advantages:
+that might provide advantages:
 \begin{itemize}
@@
 
-\item RIO is an effort to publish fragmentary results that can subsequently be
-combined into a single content item. \choinote{What's RIO?}
+\item The Research Ideas and Outcomes (RIO) journal is an effort to publish fragmentary results that can subsequently be
+combined into a single content item.
 

### 2015-12-01 WSSSPE3/report/wg_appendix_training.tex
Update wg_appendix_training.tex
- resolved \katznote
- shifted timing to early 2016

@@
 
-While little training focuses specifically on this outcome \katznote{sustainable
-software?}, a variety of training activities increase researcher awareness of
+While little training focuses specifically on sustainable
+software, a variety of training activities increase researcher awareness of
 and engagement with software professionals and software engineering practices.
@@
 
-\item During October, Nick Jones and Erin Robinson to draft WSSSPE3 report back.
+\item October through December, Nick Jones and Erin Robinson to draft WSSSPE3 report back.
 
-\item Before December 2015, Nick Jones and Erin Robinson to call a meeting of
+\item Before February 2016, Nick Jones and Erin Robinson to call a meeting of
 the broader group, to review key next steps.
 
-\item First quarter 2016 -- if willing parties are identified, draft workshop proposal
+\item Second quarter 2016 -- if willing parties are identified, draft workshop proposal
 and identify a relevant forum, including future WSSSPE events.

### 2015-12-01 WSSSPE3/report/wg_appendix_software_credit.tex
Moved sentence from SC appendix to main section

@@
 
-Modern scientific research relies considerably on software, and ensuring that
-developers of such scientific software receive credit for their efforts will
-encourage additional creation and maintenance. \katznote{the prev sentence is
-good, but feels a bit odd in the middle of the appendix, rather than at the
-beginning} There currently is no standard mechanism for citing software or
+There currently is no standard mechanism for citing software or
 receiving credit for software (akin to publications/citations). Software is

### 2015-12-01 WSSSPE3/report/wg_main_software_credit.tex
Moved sentence from SC appendix to main section

@@
 
-Science and engineering research increasingly relies on software, but there is
-currently no standard mechanism for citing software or receiving credit for
-developing software akin to receiving credit via citations for writing papers.
+Modern scientific and engineering research relies considerably on software, 
+but currently no standard mechanism exists for citing software or receiving 
+credit for developing software akin to receiving credit via citations for 
+writing papers. Ensuring that developers of such scientific software receive 
+credit for their efforts will encourage additional creation and maintenance. 
 Standardizing software citations offers one route to establishing such a

### 2015-12-01 WSSSPE3/report/wg_main_user_community.tex
adding comment
Start on User Community appendix

@@
 
+\katznote{the next little bit needs some work, as Choi points out}
+
 Several books have been written about software communities: \choinote{only 1 book below}
@@
 ``community'' for software sustainability, what practices and circumstances lead
-to having a community, how can funding help or hinder this process, and perhaps
-most importantly, how can best practices be described and distilled into a
-document that can help new projects.
+to having and maintaining a community, how can funding help or hinder this
+process, and perhaps most importantly, how can best practices be described and
+distilled into a document that can help new projects.
 

### 2015-12-02 WSSSPE3/report/WSSSPE3_review.tex

@@
 been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
-and webinars. In order to contribute to the infrastructure, codes have to adhere
-to specified minimum standards, but ideals are also given to encourage reaching
-the aforementioned goals. These include, e.g., the use of version control,
+and webinars. In order to contribute to the infrastructure, codes must adhere
+to specified minimum standards with the desire that all codes are working toward target standards.
+%, but ideals are also given to encourage reaching the aforementioned goals. 
+These include, e.g., the use of version control,
 certain coding styles, the presence and nature of code tests and documentation,
@@
 cite codes (including non-CIG codes), and only about one eighth specify the
-version used. Compared to other codes, CIG seems to be much better cited. This,
-in part, is attributed to suggested citation forms communicated to the
-community, and runtime tools that help collecting a relevant citation list
-automatically.
+version used. Compared to other codes, CIG seems to be much better cited. 
+This in part is attribute to requiring that donated software provide a citable paper specified in the User Manual. 
+The project is working on tools and methods to generate attribution information automatically.
+%This,
+%in part, is attributed to suggested citation forms communicated to the
+%community, and runtime tools that help collecting a relevant citation list
+%automatically.
 

### 2015-12-02 WSSSPE3/report/wg_appendix_software_credit.tex

--- 2015-11-26 WSSSPE3/report/wg_appendix_software_credit.tex
@@
 products including data) that contributed more significantly than a generic
-citation, including
+citation, including:
 \begin{itemize}
@@
 Although this working group's discussions at WSSSPE3 did not focus much on the
-topic of tenure, the group recognized that there is no standard
+topic of tenure and professional advancement, 
+%we should recognized that this problem goes beyond faculty to recognizing the role of "Research Software Engineers"
+the group recognized that there is no standard
 policy---generally even within a single university---for software products to be

### 2015-12-02 WSSSPE3/report/wg_appendix_user_community.tex
Create wg_appendix_user_community.tex
Former-commit-id: 5c55415aa3f0501218576f3184eab289d523ff83
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{User Community Working Group Discussion}
+\label{sec:appendix_user_community}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{add POC here}
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+
+\begin{itemize}
+\item Ethan Davis -- UCAR Unidata
+\item Dan Gunter -- Lawrence Berkeley National Lab
+\item Liz Jessup -- University of Colorado
+\item Mark Miller -- University of California, San Diego
+\item Lindsey Powers -- The HDF Group
+\item Daniel Ziskin -- NCAR Atmospheric Chemistry Observations and Modeling (ACOM) Laboratory
+\end{itemize}
 
@@
 
+The main opportunity is to increase awareness among scientific
+software developers and project managers of the importance of
+developing a community around their project.
+While this message is fairly well understood in the open source
+community, the scientific community can be more focused on the
+science a software project is supporting rather than the software
+project itself.
+
+As with many of the issues relevant to the sustainability of science
+software, the main challenge here will be changing the culture and
+expectations around scientific software.
+
 \subsection{Key Next Steps}
+
+\begin{itemize}
+\item Survey successful science software projects
+\item Survey community members from the surveyed projects
+\item Distill the survey results and document best practices around community engagement
+\item Look for ways to raise awareness
+\end{itemize}
 

### 2015-12-02 WSSSPE3/report/registrants.tex
Replied to Dan's note.

--- 2015-09-30 WSSSPE3/report/registrants.tex
@@
-Alice Allen & Astrophysics Source Code Library\\
-Gabrielle Allen & NCSA\\
-Janine Aquino & UCAR/NCAR Earth Observing Laboratory\\
-Steven Brandt & Louisiana State University\\
-Jed Brown & CU Boulder\\
-Matthias Bussonnier & UC Berkeley\\
-Jeffrey Carver & University of Alabama\\
-Emily Chen & NCSA\\
-Sou Cheng Choi & NORC at UChicago / IIT\\
-Nancy Collins & NCAR\\
-Ethan Davis & UCAR Unidata\\
-Davide DelVento & NCAR/CISL\\
-Yuhan Ding & Illinois Institute of Technology\\
-Tim Dunne & KnowInnovation \\
-Ward Fisher & UCAR/Unidata\\
-Sandra Gesing & University of Notre Dame\\
-Josh Greenberg & Sloan Foundation\\
-Dan Gunter & LBNL\\
-Ted Habermann & The HDF Group\\
-James Hetherington & University College London\\
-Neil Chue Hong & Software Sustainability Institute\\
-Elisabeth Huffer & Lingua Logica/NASA \\
-Lorraine Hwang & UC Davis - CIG\\
-Raymond Idaszak & RENCI; University of North Carolina at Chapel Hill\\
-Elizabeth Jessup & University of Colorado Boulder\\
-Nick Jones & New Zealand eScience Infrastructure (NeSI)\\
-Daniel Katz & U Chicago & Argonne\\
-Iain Larmour & EPSRC (UK)\\
-Frank Löffler & Louisiana State University\\
-Suresh Marru & Indiana University\\
-Ryan May & UCAR/Unidata\\
-Abigail Cabunoc Mayes & Mozilla Foundation\\
-Jeff McWhirter & Geode Systems\\
-Constantinos Michailidis & Knowinnovation\\
-Don Middleton & NCAR\\
-Mark Miller & SDSC\\
-Pate Motter & University of Colorado\\
-Jaroslaw Nabrzyski & University of Notre Dame\\
-Patrick Nichols & National Center for Atmospheric Research\\
-Kyle Niemeyer & Oregon State University\\
-Laura Owen & NCSA\\
-Abani Patra & Univ at Buffalo\\
-Grace Peng & National Center for Atmospheric Research\\
-Birgit Penzenstadler & California State University Long Beach\\
-Lindsay Powers & The HDF Group\\
-Bernie Randles & UCLA\\
-Erin Robinson & Foundation for Earth Science\\
-Daniel Sellars & CANARIE Inc\\
-Nikolay Simakov & SUNY University at Buffalo\\
-Ian Taylor & Cardiff University\\
-Ilian Todorov & Science & Technology Facilities Council, UK\\
-Benjamin Tovar & University of Notre Dame\\
-Gregory Tucker & University of Colorado at Boulder\\
-Matthew Turk & NCSA\\
-Colin Venters & University of Huddersfield\\
-Alexander Vyushkov & University of Notre Dame\\
-Fraser Watson & National Solar Observatory\\
-Nic Weber & University of Washington\\
-Daniel Ziskin & NCAR - ACOM\\
+Alice Allen & \href{mailto:aallen@ascl.net}{aallen@ascl.net}&Astrophysics Source Code Library\\
+Gabrielle Allen & \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}&NCSA\\
+Janine Aquino & \href{mailto:janine@ucar.edu}{janine@ucar.edu}&UCAR/NCAR Earth Observing Laboratory\\
+Steven Brandt & \href{mailto:sbrandt@cct.lsu.edu}{sbrandt@cct.lsu.edu}&Louisiana State University\\
+Jed Brown & \href{mailto:jed@jedbrown.org}{jed@jedbrown.org}&CU Boulder\\
+Matthias Bussonnier & \href{mailto:bussonniermatthias@gmail.com}{bussonniermatthias@gmail.com}&UC Berkeley\\
+Jeffrey Carver & \href{mailto:carver@cs.ua.edu}{carver@cs.ua.edu}&University of Alabama\\
+Emily Chen & \href{mailto:echen35@illinois.edu}{echen35@illinois.edu}&NCSA\\
+Sou Cheng Choi & \href{mailto:sctchoi@uchicago.edu}{sctchoi@uchicago.edu}&NORC at UChicago / IIT\\
+Nancy Collins & \href{mailto:nancy@ucar.edu}{nancy@ucar.edu}&NCAR\\
+Ethan Davis & \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}&UCAR Unidata\\
+Davide DelVento & \href{mailto:ddvento@ucar.edu}{ddvento@ucar.edu}&NCAR/CISL\\
+Yuhan Ding & \href{mailto:yding2@hawk.iit.edu}{yding2@hawk.iit.edu}&Illinois Institute of Technology\\
+Tim Dunne & \href{mailto:tim.dunne@knowinnovation.com}{tim.dunne@knowinnovation.com}&KnowInnovation \\
+Ward Fisher & \href{mailto:wfisher@ucar.edu}{wfisher@ucar.edu}&UCAR/Unidata\\
+Sandra Gesing & \href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}&University of Notre Dame\\
+Josh Greenberg & \href{mailto:greenberg@sloan.org}{greenberg@sloan.org}&Sloan Foundation\\
+Dan Gunter & \href{mailto:dkgunter@lbl.gov}{dkgunter@lbl.gov}&LBNL\\
+Ted Habermann & \href{mailto:thabermann@hdfgroup.org}{thabermann@hdfgroup.org}&The HDF Group\\
+James Hetherington & \href{mailto:jamespjh@gmail.com}{jamespjh@gmail.com}&University College London\\
+Neil Chue Hong & \href{mailto:N.ChueHong@software.ac.uk}{N.ChueHong@software.ac.uk}&Software Sustainability Institute\\
+Elisabeth Huffer & \href{mailto:beth@lingualogica.net}{beth@lingualogica.net}&Lingua Logica/NASA \\
+Lorraine Hwang & \href{mailto:ljhwang@ucdavis.edu}{ljhwang@ucdavis.edu}&UC Davis - CIG\\
+Raymond Idaszak & \href{mailto:rayi@renci.org}{rayi@renci.org}&RENCI; University of North Carolina at Chapel Hill\\
+Elizabeth Jessup & \href{mailto:jessup@cs.colorado.edu}{jessup@cs.colorado.edu}&University of Colorado Boulder\\
+Nick Jones & \href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}&New Zealand eScience Infrastructure (NeSI)\\
+Daniel Katz & \href{mailto:dsk@uchicago.edu}{dsk@uchicago.edu}&U Chicago \& Argonne\\
+Iain Larmour & \href{mailto:Iain.Larmour@epsrc.ac.uk}{Iain.Larmour@epsrc.ac.uk}&EPSRC (UK)\\
+Frank Löffler & \href{mailto:knarf@cct.lsu.edu}{knarf@cct.lsu.edu}&Louisiana State University\\
+Suresh Marru & \href{mailto:smarru@iu.edu}{smarru@iu.edu}&Indiana University\\
+Ryan May & \href{mailto:rmay@ucar.edu}{rmay@ucar.edu}&UCAR/Unidata\\
+Abigail Cabunoc Mayes & \href{mailto:abigail@mozillafoundation.org}{abigail@mozillafoundation.org}&Mozilla Foundation\\
+Jeff McWhirter & \href{mailto:jeff.mcwhirter@gmail.com}{jeff.mcwhirter@gmail.com}&Geode Systems\\
+Constantinos Michailidis & \href{mailto:costa.michailidis@knowinnovation.com}{costa.michailidis@knowinnovation.com}&Knowinnovation\\
+Don Middleton & \href{mailto:don@ucar.edu}{don@ucar.edu}&NCAR\\
+Mark Miller & \href{mailto:mark.alan.miller2@gmail.com}{mark.alan.miller2@gmail.com}&SDSC\\
+Pate Motter & \href{mailto:pate.motter@colorado.edu}{pate.motter@colorado.edu}&University of Colorado\\
+Jaroslaw Nabrzyski & \href{mailto:naber@nd.edu}{naber@nd.edu}&University of Notre Dame\\
+Patrick Nichols & \href{mailto:pnichols@ucar.edu}{pnichols@ucar.edu}&National Center for Atmospheric Research\\
+Kyle Niemeyer & \href{mailto:Kyle.Niemeyer@oregonstate.edu}{Kyle.Niemeyer@oregonstate.edu}&Oregon State University\\
+Laura Owen & \href{mailto:lowen@illinois.edu}{lowen@illinois.edu}&NCSA\\
+Abani Patra & \href{mailto:abani@buffalo.edu}{abani@buffalo.edu}&Univ at Buffalo\\
+Grace Peng & \href{mailto:grace@ucar.edu}{grace@ucar.edu}&National Center for Atmospheric Research\\
+Birgit Penzenstadler & \href{mailto:Birgit.Penzenstadler@csulb.edu}{Birgit.Penzenstadler@csulb.edu}&California State University Long Beach\\
+Lindsay Powers & \href{mailto:lpowers@hdfgroup.org}{lpowers@hdfgroup.org}&The HDF Group\\
+Bernie Randles & \href{mailto:randles@ucla.edu}{randles@ucla.edu}&UCLA\\
+Erin Robinson & \href{mailto:erinrobinson@esipfed.org}{erinrobinson@esipfed.org}&Foundation for Earth Science\\
+Daniel Sellars & \href{mailto:dan.sellars@canarie.ca}{dan.sellars@canarie.ca}&CANARIE Inc\\
+Nikolay Simakov & \href{mailto:nikolays@buffalo.edu}{nikolays@buffalo.edu}&SUNY University at Buffalo\\
+Ian Taylor & \href{mailto:ian.j.taylor@gmail.com}{ian.j.taylor@gmail.com}&Cardiff University\\
+Ilian Todorov & \href{mailto:ilian.todorov@stfc.ac.uk}{ilian.todorov@stfc.ac.uk}&Science \& Technology Facilities Council, UK\\
+Benjamin Tovar & \href{mailto:btovar@nd.edu}{btovar@nd.edu}&University of Notre Dame\\
+Gregory Tucker & \href{mailto:gtucker@colorado.edu}{gtucker@colorado.edu}&University of Colorado at Boulder\\
+Matthew Turk & \href{mailto:mjturk@illinois.edu}{mjturk@illinois.edu}&NCSA\\
+Colin Venters & \href{mailto:colin.venters@googlemail.com}{colin.venters@googlemail.com}&University of Huddersfield\\
+Alexander Vyushkov & \href{mailto:avyushko@nd.edu}{avyushko@nd.edu}&University of Notre Dame\\
+Fraser Watson & \href{mailto:fwatson@nso.edu}{fwatson@nso.edu}&National Solar Observatory\\
+Nic Weber & \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}&University of Washington\\
+Daniel Ziskin & \href{mailto:ziskin@ucar.edu}{ziskin@ucar.edu}&NCAR - ACOM\\

### 2015-12-02 WSSSPE3/report/wg_main_training.tex
Update wg_main_training.tex
- added third Workshop on related activities.

@@
 
-\subsubsection{Fit with related activities} Two existing venues for discussion
+\subsubsection{Fit with related activities} Three existing venues for discussion
 of related activities are identified:
@@
 Engineering (SEHPCCSE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
+\katznote{fix URL and change to cite}
+
+\item Workshop on Software Engineering for Sustainable Systems\footnote{\url{http://sustainabilitydesign.org/initiatives/se4susy/}}
 \katznote{fix URL and change to cite}

### 2015-12-03 WSSSPE3/report/wg_main_training.tex
Update wg_main_training.tex
- added \cite refs

@@
 \item Workshop on Sustainable Software for Science: Practice and
-Experiences (WSSSPE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
-\katznote{change to cite}
+Experiences (WSSSPE)~\cite{WSSSPE}
 
@@
 Performance Computing in Computational Science and
-Engineering (SEHPCCSE)\footnote{\url{http://wssspe.researchcomputing.org.uk/}}
-\katznote{fix URL and change to cite}
+Engineering (SEHPCCSE)~\cite{SEHPCCSE}
 
-\item Workshop on Software Engineering for Sustainable Systems\footnote{\url{http://sustainabilitydesign.org/initiatives/se4susy/}}
-\katznote{fix URL and change to cite}
+\item Workshop on Software Engineering for Sustainable Systems~\cite{se4susy}
 

### 2015-12-03 WSSSPE3/report/WSSSPE3_review.tex
edits to edits
Replied to Dan's note.
changed single quote marks to double for keynote quotes

@@
 \definecolor{blueish}{rgb}{0.2,0.2,0.8}
+\definecolor{darkblue}{rgb}{0.1,0.1,0.9}
 
@@
 \newcommand{\choinote}[1]{ {\textcolor{orange}    {***Choi: #1 }}}
+\newcommand{\stevenote}[1]{ {\textcolor{darkblue}    {***Steve: #1 }}}
 
@@
 Turk then moved onto various different definitions of sustainability. His
-favorite one was `keeping up with bug reports,' where even if no new features
+favorite one was ``keeping up with bug reports,'' where even if no new features
 were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was `adding of new features', or `maintaining the
-software for a long period of time' such as the cases of TeX or LaTeX with
+sustainability Turk mentioned was ``adding of new features,'' or ``maintaining the
+software for a long period of time'' such as the cases of TeX or LaTeX with
 community help. A notion Turk heard often at supercomputing conferences was that
-sustainable software `continues to work on new architectures'. Yet another
-metric was `people continuing to be able to learn how to use and apply the
-software'. A funder Turk heard talked about sustainability as `continuing to get
-funded'. Turk also recalled that Greg Wilson, among others, said in WSSSPE1.1
-that his view of sustainable software was software that `continued to give the
-same results over time'. A last measure of sustainability Turk presented was
-`the ability to transition between different people developing and using a piece
-of software.'
+sustainable software ``continues to work on new architectures.'' Yet another
+metric was ``people continuing to be able to learn how to use and apply the
+software.'' A funder Turk heard talked about sustainability as ``continuing to get
+funded.'' Turk also recalled that Greg Wilson, among others, said in WSSSPE1.1
+that his view of sustainable software was software that ``continued to give the
+same results over time.'' A last measure of sustainability Turk presented was
+``the ability to transition between different people developing and using a piece
+of software.''
 
@@
 version used. Compared to other codes, CIG seems to be much better cited. 
-This in part is attribute to requiring that donated software provide a citable paper specified in the User Manual. 
+In part, this is attributed to the fact that CIG requires that donated software provide a citable paper specified in the User Manual. 
 The project is working on tools and methods to generate attribution information automatically.

### 2015-12-03 WSSSPE3/report/registrants.tex
revised again

--- 2015-09-30 WSSSPE3/report/registrants.tex
@@
 Emily Chen & NCSA\\
-Sou Cheng Choi & NORC at UChicago / IIT\\
+Sou Cheng Choi &  NORC at the University of Chicago \&  Illinois Institute of Technology\\
 Nancy Collins & NCAR\\

### 2015-12-09 WSSSPE3/report/WSSSPE3_review.tex
update abstract

@@
 \todo{need to update this abstract for WSSSPE3}
-This technical report records and discusses the Second Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE2). 
+This technical report records and discusses the Third Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE3). 
 %The workshop used an
@@
 %divided across five topic areas: 
-The report includes a description of the alternative, experimental submission
-and review process, two workshop keynote presentations, a series of lightning
-talks, a discussion on sustainability, and five discussions from the topic areas
-of exploring sustainability; software development experiences; credit \&
-incentives; reproducibility \& reuse \& sharing; and code testing \& code
-review. For each topic, the report includes a list of tangible actions that were
-proposed and that would lead to potential change.
-%
-The workshop recognized that reliance on scientific software
-is pervasive in all areas of world-leading research today. The workshop
-participants then proceeded to explore different perspectives on the concept of
-sustainability. Key enablers and barriers of sustainable scientific software
-were identified from their experiences. In addition,
-recommendations with new requirements such as software credit files and software
-prize frameworks were outlined for improving practices in sustainable software
-engineering.
-%
-There was also broad consensus that formal
-training in software development or engineering was rare among the
-practitioners. Significant strides need to be made in building a sense of
-community via training in software and technical practices, on increasing their
-size and scope, and on better integrating them directly into graduate education
-programs.
-%
-Finally, journals can define and publish policies to improve reproducibility, whereas
-reviewers can insist that authors provide sufficient information and access to
-data and software to allow them reproduce the results in the paper. Hence a list of
-criteria is compiled for  journals to provide to reviewers so as to make it easier to
-review software submitted for publication as a ``Software Paper.''
+The report includes a description of keynote presentation and a set of lightning
+talks, and focuses on the discussions, future steps, and future organization for
+a set of self-organized working groups.
+For each group, the report also includes a point of contact and a landing
+page that can be used by those who want to join that group's future activities.
+\todo{think about if there are more general conclusions we want to highlight from WSSSPE3}
+%%
+%The workshop recognized that reliance on scientific software
+%is pervasive in all areas of world-leading research today. The workshop
+%participants then proceeded to explore different perspectives on the concept of
+%sustainability. Key enablers and barriers of sustainable scientific software
+%were identified from their experiences. In addition,
+%recommendations with new requirements such as software credit files and software
+%prize frameworks were outlined for improving practices in sustainable software
+%engineering.
+%%
+%There was also broad consensus that formal
+%training in software development or engineering was rare among the
+%practitioners. Significant strides need to be made in building a sense of
+%community via training in software and technical practices, on increasing their
+%size and scope, and on better integrating them directly into graduate education
+%programs.
+%%
+%Finally, journals can define and publish policies to improve reproducibility, whereas
+%reviewers can insist that authors provide sufficient information and access to
+%data and software to allow them reproduce the results in the paper. Hence a list of
+%criteria is compiled for  journals to provide to reviewers so as to make it easier to
+%review software submitted for publication as a ``Software Paper.''
 

### 2015-12-09 WSSSPE3/report/WSSSPE3_review.tex
small fixes in abstract comments

@@
 \begin{abstract}
-\todo{need to update this abstract for WSSSPE3}
 This technical report records and discusses the Third Workshop on Sustainable
@@
 page that can be used by those who want to join that group's future activities.
-\todo{think about if there are more general conclusions we want to highlight from WSSSPE3}
+\todo{think about if there are any general conclusions we want to highlight from WSSSPE3}
 %%

### 2015-12-09 WSSSPE3/report/wg_appendix_publishing_SW.tex
Replied to Dan's note.
Added text stating that the executable paper website going live in January and mentioned the need for ongoing curation.
small fixes

@@
 publications. The Software Sustainability Institute was able to do something
-similar for publications about software by making a public page \choinote{where?} containing a
+similar for publications about software by making a public page on the Software
+Sustainability Institute's website (\url{http://www.software.ac.uk}) containing a
 catalog of these publications and enlisting the help of the community to grow
@@
 Sustainability Institute's website: \url{http://www.software.ac.uk}.
-\katznote{when?}
+We expect the page to be live in early January of 2016.
+
+An ongoing effort to update the page should follow.
 
@@
 
-Steven R.\ Brandt will create a first version of the page within a week or so of
-the WSSSPE3 conference.
-
+Steven R.\ Brandt has created a first version of the page, and it is in the process
+of being posted on the Software
+Sustainability Institute's website: \url{http://www.software.ac.uk}. 
 Neil Chue Hong will take responsibility for the page once it is up.

### 2015-12-11 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
very small edits to edits to edits to edits
Appendix H:  one-character change.
Changes to fix "Choinotes"

@@
 \item Lorraine Hwang -- UC Davis
-\item Simon Trigger \choinote{TBD}
+\item Simon Trigger -- BioTeam, Inc.
 \item Nancy Wilkins-Diehr -- San Diego Supercomputer Center
@@
 programming coming from local groups. This has been very helpful in providing
-funding to establish local groups. It works best when these groups develop good
+funding to establish local groups. These groups work best when they develop good
 collaborations with national cyberinfrastructure pools. When an organization
@@
 We will also look for examples of groups which have successfully become self-
-sustaining following. \choinote{following what?}
+sustaining following initial seed funding.
 

### 2015-12-11 WSSSPE3/report/wg_main_user_community.tex
shortened main and expanded appendix

@@
 
-Mozilla Science maintains a ``Working Open Project
-Guide''~\cite{working-open-wssspe3}, the introduction of which states:
-\begin{quote}
-Working openly with contributors enables your community to learn how to build
-and collaborate together. This document is a guideline on how to work openly and
-involve others in your projects with Mozilla. We want to help you engage your
-community in a way that encourages contributors and builds other leaders.
- \end{quote}
-
-\katznote{the next little bit needs some work, as Choi points out}
-
-Several books have been written about software communities: \choinote{only 1 book below}
-\begin{itemize}
-
-\item ``Art of Community'' by Jono Bacon~\cite{art-of-community}. We could
-consider distilling this for scientific software.
-
-\item Iain Larmour, from EPSRC in the UK (EPSRC: https://www.epsrc.ac.uk/) (not
-sure who from mentioned UK Collaborative Computational Projects (CCP:
-http://ccp.ac.uk) \choinote{Who is ``not sure who'' supposed to be?}
-
-\end{itemize}
+There are a number of activities already in progress that are targeted at improving
+the user community for open-source software, including Mozilla Science's ``Working Open Project
+Guide''~\cite{working-open-wssspe3} and
+``UK Collaborative Computational Projects'' (CCP: http://www.ccp.ac.uk), or 
+ books such as ``Art of Community'' by Jono Bacon~\cite{art-of-community}.
 
@@
 
-The benefits of having a community that were brought up were considered largely
-obvious. In addition to having advocates for the software, and a possible source
-of ``free'' contributions to the codebase, the community becomes a good source
-for requirements, feedback, and metrics. The software community can also act as
-``cheerleaders'' who convince funders or other potential users to fund/use the
-software, and thus help sustain the software.
-
-Practices and circumstances that lead to a community are first, that the
-software offers value. But in addition to this, a community will be much more
-likely to form if they receive (expert) support when they have questions.
-Additional contributing factors are good usability (not always needed), and an
-open development process such as IPython developer meetings on YouTube. It was
-also pointed out that an evangelist for the project, not necessarily but often
-one of the developers, can often make a big difference.
-
-Funding can help the process by encouraging both value to the community and
-high-quality user support. Only providing funding for the software development
-may create good software, but with less likelihood to have a real community. It
-was discussed that federal laboratories are a good incubator for software
-communities, and that a general facility like EarthCube is too dispersed to
-really make a community. Also, domain-specific groups within laboratories or
-universities might provide as an incubator for software communities.
-
-In describing best practices, the group discussed the different modes for
-starting a scientific software project: building on an existing product that
-needs improving, recognizing an unsatisfied need of an existing community, or
-creating a new solution to a need not yet recognized by the community. The group
-also thought that the existing books on software communities would need to be
-evaluated in light of differences between Science Software projects and general
-OSS projects in terms of scale, science, acknowledgement and credit, and funding
-models.
+Everyone agreed on a few points: software must not only offer value, but there
+must be some support for users; and funding can help pay for that support, in
+addition to the usual funding for software development. Openness is generally 
+a virtue. An evangelist, either in the form of a single person or some
+domain-specific group of users, is often the key factor.
 
@@
 replicate this type of project, and look at end-of-life on a community project.
-Inputs to this document would include a software community survey of highly
-functioning communities such as R Open Science, Python SciPy, OPeNDAP, and
-Unidata, with analysis of factors that feed into their success. Also references
-lik the ``Art of Community'' could be adapted and summarized for the science
-software community.
-
-Another next step would be increasing recognition of need for science software
-projects to focus on building and supporting their user communities. Good software
-engineering practices are not enough, and popular training like Software
-Carpentry does not currently address this issue head on.
+Another next step would be better training to increase recognition of need for science software
+projects to focus on building and supporting their user communities.
 
 \subsubsection{Landing Page}
-\todo{link to landing page}
+
+This group doesn't have a landing page yet. Please send requests to join and contribute to both
+Dan Gunter <dkgunter@lbl.gov> and Ethan Davis <edavis@ucar.edu>.

### 2015-12-11 WSSSPE3/report/wg_main_best_practices.tex
commented out non-functional and badly formatted citation

@@
 %\subsubsection{Why it is important}
-Reviewing multiple past articles and talks at different meetings like WSSSPEx\cite{WSSSPE papers, heroux paper, ...} and 
+Reviewing multiple past articles and talks at different meetings like WSSSPEx
+%\cite{WSSSPE papers, heroux paper, ...} 
+and 
 analyzing and promoting sustainable scientific software makes it clear that there 

### 2015-12-11 WSSSPE3/report/wg_appendix_user_community.tex

@@
 
-\todo{add POC here}
+POC:
+Dan Gunter <dkgunter@lbl.gov> and 
+Ethan Davis <edavis@lbl.gov>
 
@@
 \subsection{Summary of Discussion}
+
+Discussion revolved around a few questions: what is the benefit of having a
+``community'' for software sustainability, what practices and circumstances lead
+to having and maintaining a community, how can funding help or hinder this
+process, and perhaps most importantly, how can best practices be described and
+distilled into a document that can help new projects.
+
+The benefits of having a community that were brought up were considered largely
+obvious. In addition to having advocates for the software, and a possible source
+of ``free'' contributions to the codebase, the community becomes a good source
+for requirements, feedback, and metrics. The software community can also act as
+``cheerleaders'' who convince funders or other potential users to fund/use the
+software, and thus help sustain the software.
+
+Practices and circumstances that lead to a community are first, that the
+software offers value. But in addition to this, a community will be much more
+likely to form if they receive (expert) support when they have questions.
+Additional contributing factors are good usability (not always needed), and an
+open development process such as IPython developer meetings on YouTube. It was
+also pointed out that an evangelist for the project, not necessarily but often
+one of the developers, can often make a big difference.
+
+Funding can help the process by encouraging both value to the community and
+high-quality user support. Only providing funding for the software development
+may create good software, but with less likelihood to have a real community. It
+was discussed that federal laboratories are a good incubator for software
+communities, and that a general facility like EarthCube is too dispersed to
+really make a community. Also, domain-specific groups within laboratories or
+universities might provide as an incubator for software communities.
+
+In describing best practices, the group discussed the different modes for
+starting a scientific software project: building on an existing product that
+needs improving, recognizing an unsatisfied need of an existing community, or
+creating a new solution to a need not yet recognized by the community. The group
+also thought that the existing books on software communities would need to be
+evaluated in light of differences between Science Software projects and general
+OSS projects in terms of scale, science, acknowledgement and credit, and funding
+models.
 
@@
 
+The most important next steps is a ``Best Practice'' document, which would
+describe what successful projects with engaged communities look like, how to
+replicate this type of project, and look at end-of-life on a community project.
+Inputs to this document would include a software community survey of highly
+functioning communities such as R Open Science, Python SciPy, OPeNDAP, and
+Unidata, with analysis of factors that feed into their success. Also references
+like the ``Art of Community'' could be adapted and summarized for the science
+software community.
+
+More specifically, the group would like to take the following steps:
+
 \begin{itemize}
@@
 
+Another next step would be increasing recognition of need for science software
+projects to focus on building and supporting their user communities. Good software
+engineering practices are not enough, and popular training like Software
+Carpentry does not currently address this issue head on.
+
 \subsection{Plan for Future Organization}
+
+No definite plans were agreed upon for future organizaton. The major ideas discussed
+were coordinating with another group or adapting some existing text.
+
+Collaboration within the framework of an existing organization seems a good initial
+path. Mozilla Science maintains a ``Working Open Project
+Guide''~\cite{working-open-wssspe3}, the introduction of which states:
+\begin{quote}
+Working openly with contributors enables your community to learn how to build
+and collaborate together. This document is a guideline on how to work openly and
+involve others in your projects with Mozilla. We want to help you engage your
+community in a way that encourages contributors and builds other leaders.
+ \end{quote}
+
+ Another idea is to form a group that could adapt existing commercial-oriented 
+ guidelines for the world of scientific software and top-down funding structures.
+ For example, to distill the ``Art of Community'' by Jono Bacon~\cite{art-of-community}
+for scientific software.
 
@@
 
+The group had many points of agreement, but there is not currently a dedicated core group
+of people who have committed to producing the key milestones. Coordination via phone or
+online would be necessary to build this ``community'' of contributors.
+
 \subsection{Key Milestones and Responsible Parties}
 
+The key milestones for the group's activities align closely with the Key Next Steps above:
+
+\begin{itemize}
+\item Complete and write up a survey of project members, and community members, for successful science software projects
+\item Distill the survey results and document best practices around community engagement
+\end{itemize}
+
 \subsection{Description of Funding Needed}
+
+With a small amount of seed funding, it is possible that members of this group or other parties could
+spend the time necessary to devise a survey of existing projects and deploy this, probably including traveli to
+meetings and workshops for the various software communities.

### 2015-12-11 WSSSPE3/report/wg_appendix_user_community.tex
making Ethan's changes

@@
 Dan Gunter <dkgunter@lbl.gov> and 
-Ethan Davis <edavis@lbl.gov>
+Ethan Davis <edavis@ucar.edu>
 
@@
 With a small amount of seed funding, it is possible that members of this group or other parties could
-spend the time necessary to devise a survey of existing projects and deploy this, probably including traveli to
+spend the time necessary to devise a survey of existing projects and deploy this, probably including travel to
 meetings and workshops for the various software communities.

### 2015-12-11 WSSSPE3/report/wg_main_best_practices.tex
changed comment to note to make it more actionable

@@
 %\subsubsection{Why it is important}
-Reviewing multiple past articles and talks at different meetings like WSSSPEx
-%\cite{WSSSPE papers, heroux paper, ...} 
+Reviewing multiple past articles and talks at different meetings like WSSSPEx \todo{add cites to WSSSPE papers, heroux paper, ...} 
 and 

### 2015-12-11 WSSSPE3/report/wg_main_metrics.tex

@@
 group members. Similarly, it was suggested that a software metrics survey be
-sent to the UK SFTF and TRDF software projects to ask them what metrics would be
-useful to report. The remainder of the discussion focused mainly on the creation
+sent to the UK SFTF (Software For The Future, led by the Engineering and Physical Sciences Research Council) and TRDF (Tools and Resources Development Fund, led by the Biotechnology and Biological Sciences Research Council) software projects to ask them what metrics would be useful to report.
+The remainder of the discussion focused mainly on the creation
 of a white paper on this topic. This resulted in a paper outline and writing
-assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE,
-or JORS. More information about the group discussion is available in
-Appendix~\ref{sec:appendix_metrics}. \choinote{Spell out SFTF, TRDF, CISE, JORS?}
+assignments with the goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of Electrical and Electronic Engineers Computing in Science and Engineering magazine), or JORS (Journal of Open Research Software).
+More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}
 

### 2015-12-15 WSSSPE3/report/wg_main_best_practices.tex

@@
 
-Core questions that will need to be explored are in knowledge management, 
-(transitions between people), reliability (reproducibility), usability, and how a software tool becomes part of the core workflow of well identified users (stakeholders)
-relating to tool success and hence sustainability. \katznote{prev sentence is complex and awkward} Ideas 
-that may need to be explored include:
+Core questions that will need to be explored are in reliability (reproducibility), usability, extensibility, knowledge management and continuity (transitions between people).
+Answers to these will guide us on how a software tool becomes part of the core workflow of well identified users (stakeholders) relating to tool success and hence sustainability.
+\katznote{prev sentence is complex and awkward} % see fix
+Ideas  that may need to be explored include:
 \begin{itemize}
@@
 \item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
-\item Dimensions of sustainability -- economic, technical, environmental, 
-declining interest in primary application area), \katznote{not sure what the prev. comment goes with} social.
+\item Dimensions of sustainability -- economic, technical, environmental and
+obsolescence. 
+\katznote{not sure what the prev. comment goes with} %social.
 \end{itemize}

### 2015-12-15 WSSSPE3/report/wg_appendix_best_practices.tex
Update best practices appendix with POC
removing duplicate entry
accepting change and replicating to appendix

@@
 
-\todo{add POC here}
+Sandra Gesing\footnote{email:\href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}} will serve as
+the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
 
@@
 \item Abigail Cabunoc Mayes -- Mozilla Foundation
-\item Jeff Carver -- University of Alabama
 \item Frank L\"{o}ffler -- Louisiana State University 
@@
 
-Core questions that will need to be explored are in knowledge management, 
-(transitions between people), reliability (reproducibility), usability, and how a software tool becomes part of the core workflow of well identified users (stakeholders)
-relating to tool success and hence sustainability. \katznote{prev sentence is complex and awkward} Ideas 
-that may need to be explored include:
+Core questions that will need to be explored are in reliability (reproducibility), usability, extensibility, knowledge management and continuity (transitions between people).
+Answers to these will guide us on how a software tool becomes part of the core workflow of well identified users (stakeholders) relating to tool success and hence sustainability.
+%\katznote{prev sentence is complex and awkward} % see fix
+Ideas that may need to be explored include:
 \begin{itemize}
@@
 \item Dimensions of sustainability -- economic, technical, environmental,
-declining interest in primary application area), \katznote{not sure what the
-prev. comment goes with} social.
+and obsolescence.
+%declining interest in primary application area), \katznote{not sure what the
+%prev. comment goes with} social.
 

### 2015-12-20 WSSSPE3/report/WSSSPE3_review.tex
fixed question for D.Gunter lightning talk
missing word in abstract

@@
 %divided across five topic areas: 
-The report includes a description of keynote presentation and a set of lightning
+The report includes a description of the keynote presentation and a set of lightning
 talks, and focuses on the discussions, future steps, and future organization for
@@
 activities like publishing, performance studies, graduating students, or
-protecting ideas before publication. Sustainability is not of big concern there,
-compared to these. \katznote{prev sentence is unclear} Instead of this more
-traditional approach, KBase uses the following for large, distributed teams.
+protecting ideas before publication. Instead of this more
+traditional approach, KBase uses a variation on the ``Scrum'' methodology.
 After picking projects and team members, 4-5 teams work on projects about two

### 2015-12-20 WSSSPE3/report/WSSSPE3_review.tex
some general cleanup/consistency

@@
 traditional approach, KBase uses a variation on the ``Scrum'' methodology.
-After picking projects and team members, 4-5 teams work on projects about two
-weeks before a 1-week evaluation by the executive committee, and after which new
+After picking projects and team members, four to five teams work on projects for about two
+weeks before a one-week evaluation by the executive committee, and after which new
 teams and projects might be chosen to restart the ``agile'' development cycle.

### 2015-12-20 WSSSPE3/report/wg_main_best_practices.tex
accepting change and replicating to appendix
some general cleanup/consistency

@@
 Answers to these will guide us on how a software tool becomes part of the core workflow of well identified users (stakeholders) relating to tool success and hence sustainability.
-\katznote{prev sentence is complex and awkward} % see fix
+%\katznote{prev sentence is complex and awkward} % see fix
 Ideas  that may need to be explored include:
@@
 \item When should software ``die''?
-\item Catering to disruptive developments in environment e.g.(new hardware, new methodology) ;
+\item Catering to disruptive developments in environment e.g.(new hardware, new methodology);
 \item Dimensions of sustainability -- economic, technical, environmental and
 obsolescence. 
-\katznote{not sure what the prev. comment goes with} %social.
+%\katznote{not sure what the prev. comment goes with} %social.
 \end{itemize}

### 2015-12-20 WSSSPE3/report/wg_appendix_best_practices.tex
moved all emails to footnotes for consistency

@@
 
-Sandra Gesing\footnote{email:\href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}} will serve as
+Sandra Gesing\footnote{email: \href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}} will serve as
 the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.

### 2015-12-20 WSSSPE3/report/wg_main_metrics.tex
some general cleanup/consistency
moved all emails to footnotes for consistency

@@
 assignments with the goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of Electrical and Electronic Engineers Computing in Science and Engineering magazine), or JORS (Journal of Open Research Software).
-More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}
+More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
 
@@
 In lieu of a landing page, the Useful Metrics for Scientific Software working
-group requests an email be sent to Gabrielle Allen
-\href{mailto:gdallen@illinois.edu}{(gdallen@illinois.edu)} to find out more
+group requests an email be sent to Gabrielle Allen\footnote{email: 
+\href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} to find out more
 about the group's efforts and how to participate.

### 2015-12-20 WSSSPE3/report/wg_main_training.tex
small fix to match bib change
some general cleanup/consistency
moved all emails to footnotes for consistency

@@
 
-\item Workshop on Sustainable Software for Science: Practice and
-Experiences (WSSSPE)~\cite{WSSSPE}
+\item Working towards Sustainable Software for Science: Practice and
+Experiences (WSSSPE) workshops~\cite{WSSSPE}
 
@@
 establishing a community committed to increasing the degree of coordination
-across training projects.
-
-See Appendix~\ref{sec:appendix_training} for more details about the discussion.
+across training projects. See Appendix~\ref{sec:appendix_training} for more details about the discussion.
 
@@
 
-The Training working group requests an email be sent to Nick Jones
-\href{mailto:nick.jones@nesi.org.nz}{(nick.jones@nesi.org.nz)} to find out more
+The Training working group requests an email be sent to Nick Jones\footnote{email: 
+\href{mailto:nick.jones@nesi.org.nz}{nick.jones@nesi.org.nz}} to find out more
 about the group's efforts and how to participate.

### 2015-12-20 WSSSPE3/report/wg_main_industry_interaction.tex
tried converting text to better paragraph form

@@
 
-\textbf{Goal}
+First, we discussed the \textbf{goals} of this research, attempting to answer the 
+question \emph{What is the goal of doing research on transition pathways?}
+A number of research questions arose:  Can we identify collaborations that have 
+occurred and try to understand which were successful, which were unsuccessful, 
+and what factors contributed to these successes/failures? Can we determine what 
+each partner wants to get out of such a collaboration? For example, why would 
+industry be interested in collaborating with academia? Or why would academia 
+be interested in collaborating with industry? How could we design a study that 
+focused on the impact of the software in undergoing this type of transition?
 
-\emph{What is the goal of doing research on transition pathways?}
-
-Can we identify collaborations that have occurred and try to understand which were successful, which were unsuccessful, and what factors contributed to these successes/failures?
-
-Can we determine what each partner wants to get out of such a collaboration?
-For example, why would industry be interested in collaborating with academia? 
-Or why would academia be interested in collaborating with industry?
-
-How could we design a study that focused on the impact of the software in
-undergoing this type of transition?
-
-\textbf{Potential outcomes}
-
-A set of case studies that look at successful and unsuccessful
+Next, we imagined \textbf{potential outcomes} of research on this topic, involving 
+a set of case studies that look at successful and unsuccessful
 transitions of researchers between academia and industry. This might address 
@@
 described as those that lead to either weak or strong sustainability (also
-defined below).
+defined below). In addition, the results from this research might help create a 
+generalizable framework that might allow for the study of different transition 
+pathways (other than academia to industry).
 
-Create a generalizable framework that might allow for the study of different
-transition pathways (other than academia to industry).
-
-\textbf{General Definitions}
-
-We characterize transitions in the following ways:
+Finally, we created some \textbf{general definitions} for these concepts; we 
+characterize transitions in the following ways:
 \begin{itemize}

### 2015-12-20 WSSSPE3/report/wg_main_eng_design.tex
Added reference to eng design main section
added some todo comments about moving text to appendices

@@
 
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design \todo{add reference}:
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}:
 \begin{itemize}
@@
 \end{itemize}
+\todo{should these sets of bullet points be moved to the appendix? or moved into paragraph form, since they are relatively short items?}
 

### 2015-12-20 WSSSPE3/report/wg_main_best_practices.tex
added some todo comments about moving text to appendices
small fix to fix - thanks Kyle!
moving outline to appendix

@@
 \item When should software ``die''?
-\item Catering to disruptive developments in environment e.g.(new hardware, new methodology);
+\item Catering to disruptive developments in environment (e.g., new hardware, new methodology);
 \item Dimensions of sustainability -- economic, technical, environmental and
@@
 the general purpose PeTSC toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.
-We also established a timeline and a rough outline for the report.
+We also established a timeline and a rough outline (see Appendix~\ref{sec:appendix_best_practices}) for the report.
 
@@
 
-\noindent{\bf Outline:}
-\begin{enumerate}
-\item Introduction and Scope of White Paper 
-\item Related Work
-\item Case Studies
-\begin{enumerate} 
-\item PeTSC
-\item NWChem
-\item CIG
-\end{enumerate}
-\item Community Related Practices
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Governance and management
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Funding Related
-\begin{enumerate} 
-\item Findings
-\item Recommendations
-\end{enumerate}
-\item Metrics for sustainability
-\item Tools
-\item Conclusions
-\end{enumerate}
-
 \subsubsection{Landing Page}

### 2015-12-21 WSSSPE3/report/wg_main_funding_spec_expert.tex
Section 5.2 and Section 5.3: No content change
Update link to international list of RSE groups curated by UKRSE.

@@
 
-Research Software Engineers (RSEs) -- those who contribute to science and scholarship
-through software development -- are an important part of the team needed to
-deliver 21st century research. However, existing academic structures and systems
-of funding do not effectively fund and sustain these skills. The resulting high
-levels of turnover and inappropriate incentives are significant contributing
-factors to low levels of reliability and readability observed in scientific
-software. Moreover, the absence of skilled and experienced developers retards
-progress in key projects, and at times causes important projects to fail
+Research Software Engineers (RSEs) -- those who contribute to science and
+scholarship through software development -- are an important part of the team
+needed to deliver 21st century research. However, existing academic structures
+and systems of funding do not effectively fund and sustain these skills. The
+resulting high levels of turnover and inappropriate incentives are significant
+contributing factors to low levels of reliability and readability observed in
+scientific software. Moreover, the absence of skilled and experienced developers
+retards progress in key projects, and at times causes important projects to fail
 completely.
@@
 research domain sufficiently to build meaningful software at a reasonable pace.
-This requires a collaborative approach -- where developers who are fully
-engaged or invested in the research context are co-developing software with domain
+This requires a collaborative approach -- where developers who are fully engaged
+or invested in the research context are co-developing software with domain
 academics.
@@
 a wider solution. There remains the problem of how to ``sell'' the value of this
-investment to investigators within a university. This is an issue best
-addressed by the individual organizations that embark on the plan.
+investment to investigators within a university. This is an issue best addressed
+by the individual organizations that embark on the plan.
 
-For more details on the discussion, see Appendix~\ref{sec:appendix_funding_spec_expert}.
+For more details on the discussion, see
+Appendix~\ref{sec:appendix_funding_spec_expert}.
 
@@
 funded. For example, how many research universities currently fund people in the
-RSE track, whether they bear the RSE moniker or not. Are these developers
-paid by the university or through a program supported by research
-grants/individual PIs? How did they bootstrap the developer track to get this
-started? How successful is the university in getting investigators to pay for
-fractional RSEs? We will author a report describing our findings, should funding
-be available to conduct the investigation.
+RSE track, whether they bear the RSE moniker or not. Are these developers paid
+by the university or through a program supported by research grants/individual
+PIs? How did they bootstrap the developer track to get this started? How
+successful is the university in getting investigators to pay for fractional
+RSEs? We will author a report describing our findings, should funding be
+available to conduct the investigation.
 
@@
 
-To find more information about this group, or join it, see \url{http://www.rse.ac.uk/groups}.
+To see our list of international research software engineering groups so far
+visit \url{http://www.rse.ac.uk/international}. For the UK groups, see
+\url{http://www.rse.ac.uk/groups}. Too add your group to the list, please make a pull request to
+\url{https://github.com/UKRSE/UKRSE.github.io/blob/master/international.md}

### 2015-12-21 WSSSPE3/report/wg_main_funding_spec_expert.tex
small edits

--- 2015-11-15 WSSSPE3/report/wg_main_funding_spec_expert.tex
@@
 
-To find more information about this group, or join it, see \url{http://www.rse.ac.uk/groups}.
+A list of known UK research software engineering groups is available at  
+\url{http://www.rse.ac.uk/groups}, and a list for the rest of the world
+is at \url{http://www.rse.ac.uk/international}.
+To add another group to the list, please make a pull request as requested
+on either of these pages.

### 2015-12-22 WSSSPE3/report/wg_appendix_best_practices.tex
moving outline to appendix
Updated timeline best practices

@@
 
+\noindent{\bf White Paper Outline:}
+\begin{enumerate}
+\item Introduction and Scope of White Paper 
+\item Related Work
+\item Case Studies
+\begin{enumerate} 
+\item PeTSC
+\item NWChem
+\item CIG
+\end{enumerate}
+\item Community Related Practices
+\begin{enumerate} 
+\item Findings
+\item Recommendations
+\end{enumerate}
+\item Governance and management
+\begin{enumerate} 
+\item Findings
+\item Recommendations
+\end{enumerate}
+\item Funding Related
+\begin{enumerate} 
+\item Findings
+\item Recommendations
+\end{enumerate}
+\item Metrics for sustainability
+\item Tools
+\item Conclusions
+\end{enumerate}
+
 \subsection{Key Next Steps}
@@
 \begin{itemize}
-\item 15 Nov: Introduction and scope finished (Abani Patra/Sandra Gesing)
-\item 15 Nov: Sections assigned (Abani Patra/Sandra Gesing)
+\item 28 Dec: Introduction and scope finished (Abani Patra/Sandra Gesing)
+\item 06 Jan: Sections assigned (Abani Patra/Sandra Gesing)
 \item 31 Jan: Analyzing funding possibilities for survey

### 2015-12-23 WSSSPE3/report/wg_main_best_practices.tex
Update wg_main_best_practices.tex

@@
 %\subsubsection{Why it is important}
-Reviewing multiple past articles and talks at different meetings like WSSSPEx \todo{add cites to WSSSPE papers, heroux paper, ...} 
+Reviewing multiple past articles and talks at different meetings like WSSSPEx \cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} 
 and 
@@
 %
-\item 15 Nov: Introduction and scope finished
-\item 15 Nov: Sections assigned
+\item 28 Dec: Introduction and scope finished
+\item 06 Jan: Sections assigned
 \item 31 Jan: Analyzing funding possibilities for survey

### 2015-12-23 WSSSPE3/report/WSSSPE3_review.tex
adding some thoughts to the abstract and conclusions - comments/edits welcome

@@
 page that can be used by those who want to join that group's future activities.
-\todo{think about if there are any general conclusions we want to highlight from WSSSPE3}
+The main challenge left by the workshop is to see if the groups will carry
+out these activities that they have scheduled, and how the WSSSPE community
+can encourage its members to do so.
 %%
@@
 The working groups met for a small part of the first half day and all of the
-second day, excepting some times for the groups to report back to the collected
-workshop attendees. \todo{need to say some more about what came of this}
+second day, with the exception of some short periods for the groups to report
+back to the collected workshop attendees. Each group determined a set
+of activities that the members could do to advance sustainable software in
+a particular area.
 
-\todo{this paragraph is from the wssspe2 report - to be updated...} However, the
-challenge that we have discovered since WSSSPE2 is that it is very hard to
-continue the breakout groups' activities. The WSSSPE2 participants were willing
+The results of these group sessions made it clear that
+there are many interlinked challenges in sustainable software, and that
+while these challenges can be addressed, doing so is difficult because they
+generally are not the full-time job of any of the attendees. 
+As was the case in WSSSPE2 as well, the participants were willing
 to dedicate their time to the groups while they were at the meeting, but
-afterwards, they went back to their (paid) jobs. We need to determine how
+afterwards, they went back to their (paid) jobs.
+
+We need to determine how
 to tie the WSSSPE breakout activities to people's jobs, so that they feel that
@@
 getting the workshop participants to agree to a specific schedule of activities
-during the workshop.
+during the workshop as we have tried to do in WSSSPE3.  It remains to be
+seen, however, if the participants will meet the schedules they set.
+
+The overall challenge
+left to the sustainable software community is perhaps one of organization: how
+to combine the small partial efforts of a large number of people to impact a much
+larger number of people: those who develop and use scientific software.  While
+WSSSPE might help focus the actions of the groups, something more is needed
+to incentivize the wider community, which is a generalization of the sustainable
+software problem itself.
 

### 2015-12-29 WSSSPE2_report/WSSSPE2_review.tex
slight text change in 4.1, changed tables in 5 into figures both as requested by JORS reviewers

@@
 could be challenging to keep a balance between what have to be captured and what
-would be ideal if we do not want to lose user engagement as in the case of the
-old Harvard Dataverse, finding often only the first four fields of three pages
-of metadata were filled out.
+would be ideal if we do not want to lose user engagement, which he said he had
+experienced with the
+Harvard Dataverse~\cite{dataverse} where a large amount (three pages) of metadata
+was requested, but often only the first four fields were filled out.
 
@@
 attendee votes are shown in
-{\tablename}s~\ref{tb:software_sustainability_enablers} and
-\ref{tb:software_sustainability_barriers}, respectively.\footnote{A few other
+%{\tablename}s~\ref{tb:software_sustainability_enablers} 
+{\figurename}s~\ref{fig:software_sustainability_enablers}
+and
+%\ref{tb:software_sustainability_barriers}
+\ref{fig:software_sustainability_barriers}, respectively.\footnote{A few other
 items were suggested as barriers, but were not voted on due to lack of time in
@@
 
-\begin{table*}[ht]
+%\begin{table*}[ht]
+%\centering
+%\caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
+%indicating the fraction of attendees who voted for an item as important.}
+%\label{tb:software_sustainability_enablers}
+%  \begin{small}
+%  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
+%   %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
+%    \hline
+%  Importance & Item \\ \hline \hline
+%********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
+%********** & designing for growth and extension -- open development \\ \hline
+%******* & culture in community for reuse \\ \hline
+%**** & portability \\ \hline
+%**** & culture in development community to support transition between developers \\ \hline
+%*** & interdisciplinary people: science + IT experience \\ \hline
+%** & planning for end of life \\ \hline
+%** & make smart choices about dependencies \\ \hline
+%* & thinking of software as product lines -- long term vs.~short term view \\ \hline
+% & not all communities need new software \\ \hline
+% & converting use into resources \\ \hline
+%    \end{tabular}
+%    \end{small}
+%\end{table*}
+
+\begin{figure}[!ht]
   \centering
-\caption{Enablers of software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.}
-\label{tb:software_sustainability_enablers}
-  \begin{small}
-  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
-   %  \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
-    \hline
-  Importance & Item \\ \hline \hline
-********** & healthy and vibrant communities; vibrant community to champion software \\ \hline
-********** & designing for growth and extension -- open development \\ \hline
-******* & culture in community for reuse \\ \hline
-**** & portability \\ \hline
-**** & culture in development community to support transition between developers \\ \hline
-*** & interdisciplinary people: science + IT experience \\ \hline
-** & planning for end of life \\ \hline
-** & make smart choices about dependencies \\ \hline
-* & thinking of software as product lines -- long term vs.~short term view \\ \hline
- & not all communities need new software \\ \hline
- & converting use into resources \\ \hline
-    \end{tabular}
-    \end{small}
-\end{table*}
-
-\begin{table}[ht]
-\caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
-indicating the fraction of attendees who voted for an item as important.}
-\label{tb:software_sustainability_barriers}
+    \includegraphics[width=1.0\textwidth]{figures/enablers}
+  \caption{Enablers of software sustainability, showing the fraction of attendees (rounded to the nearest 10\%) who voted for an item as important.\label{fig:software_sustainability_enablers}}
+\end{figure}
+
+\begin{figure}[!ht]
   \centering
-  \begin{small}
-  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
-% \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
-   \hline
-  Importance & Item \\ \hline \hline
-******* & lack of incentives, including promotion and tenure process; promotion
-and tenure process in academic is incompatible with sustainability \\ \hline
-*****  & absent or poor documentation \\ \hline
-***** & funding to ensure sustainability is difficult to obtain \\ \hline
-**** & developers are not computer scientists; don't have software engineering
-practices (in particular, those needed to scale-up projects to support and be
-developed by a large sustainable community) \\ \hline
-*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test''
-in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number
-of key people a project would need to lose to become non-viable (the larger the
-number, the healthier the project).} \\ \hline
-** & rate of change of underlying technologies \\ \hline
-** & lack of business models for sustainability \\ \hline
-* & lack of training for how to build sustainability into the system \\ \hline
- & maintenance for software is not visible, appears to ``just happen'' \\ \hline
- & licensing issues \\ \hline
- & staff turnover -- lack of continuity \\ \hline
-\end{tabular}
- \end{small}
-\end{table}
+    \includegraphics[width=1.0\textwidth]{figures/barriers}
+  \caption{Barriers to software sustainability, showing the fraction of attendees (rounded to the nearest 10\%) who voted for an item as important.
+{\footnotesize \newline (1) This item also included the fact that that the promotion and tenure process in academic is incompatible with sustainability.
+\newline  (2) The software engineering practices were those in particular that are needed to scale-up projects to support and be developed by a large sustainable community).
+\newline  (3) The smallest number of key people a project would need to lose to become
+non-viable -- the larger the number, the healthier the project.}
+\label{fig:software_sustainability_barriers}}
+\end{figure}
+
+%\begin{table}[ht]
+%\caption{Barriers to software sustainability, with 0 to 10 `*'s roughly
+%indicating the fraction of attendees who voted for an item as important.}
+%\label{tb:software_sustainability_barriers}
+% \centering
+%  \begin{small}
+%  \begin{tabular}{ | p{1.65cm} | p{12.0cm} |}
+%% \begin{tabular}{ | p{1.9cm} | p{11.5cm} |}
+%   \hline
+%  Importance & Item \\ \hline \hline
+%I & lack of incentives, including promotion and tenure process; promotion
+%and tenure process in academic is incompatible with sustainability \\ \hline
+%*****  & absent or poor documentation \\ \hline
+%***** & funding to ensure sustainability is difficult to obtain \\ \hline
+%**** & developers are not computer scientists; don't have software engineering
+%practices (in particular, those needed to scale-up projects to support and be
+%developed by a large sustainable community) \\ \hline
+%*** & overreliance on one or two people -- the `bus test'\footnote{``Bus test''
+%in Table~\ref{tb:software_sustainability_barriers} refers to the smallest number
+%of key people a project would need to lose to become non-viable (the larger the
+%number, the healthier the project).} \\ \hline
+%** & rate of change of underlying technologies \\ \hline
+%** & lack of business models for sustainability \\ \hline
+%* & lack of training for how to build sustainability into the system \\ \hline
+% & maintenance for software is not visible, appears to ``just happen'' \\ \hline
+% & licensing issues \\ \hline
+% & staff turnover -- lack of continuity \\ \hline
+%\end{tabular}
+% \end{small}
+%\end{table}
 

### 2015-12-29 WSSSPE2_report/WSSSPE2_review.tex
Remove unmatched right parenthesis

@@
 {\footnotesize \newline (1) This item also included the fact that that the promotion and tenure process in academic is incompatible with sustainability.
-\newline  (2) The software engineering practices were those in particular that are needed to scale-up projects to support and be developed by a large sustainable community).
+\newline  (2) The software engineering practices were those in particular that are needed to scale-up projects to support and be developed by a large sustainable community.
 \newline  (3) The smallest number of key people a project would need to lose to become

### 2016-01-04 WSSSPE3/report/WSSSPE3_review.tex
Acknowledge SSI and RCUK
adding authors by number&lines of addtions committed to report

@@
 
-\author{tbd by writing and organizing}
+\author{Daniel S. Katz$^{(1)}$,
+Kyle Niemeyer$^{(2)}$,
+Sou-Cheng T. Choi$^{(3)}$,
+James Hetherington$^{(4)}$,
+Frank L\"{o}ffler$^{(5)}$,
+Dan Gunter$^{(6)}$,
+Ray Idaszak$^{(7)}$,
+Steven R. Brandt$^{(5)}$,
+Mark A. Miller$^{(7)}$,
+Sandra Gessing$^{(9)}$,
+Nick D. Jones$^{(10)}$,
+Nic Weber$^{(11)}$,
+Gabrielle Allen$^{(12)}$,
+Colin C. Venters$^{(13)}$,
+Ethan Davis$^{(14)}$,
+Lorraine Hwang$^{(15)}$,
+Ilian Todorov$^{(16)}$,
+Abani Patra$^{(17)}$,
+Miguel de Val-Borro$^{(18)}$,
+}
 
-%\author{Daniel S. Katz$^{(1)}$, Sou-Cheng T. Choi$^{(2)}$, Nancy Wilkins-Diehr$^{(3)}$, Neil Chue Hong$^{(4)}$,
-%\\Colin C. Venters$^{(5)}$, James Howison$^{(6)}$, Frank Seinstra$^{(7)}$, Matthew Jones$^{(8)}$,
-%\\Karen Cranston$^{(9)}$, Thomas L. Clune$^{(10)}$, Miguel de Val-Borro$^{(11)}$, Richard Littauer$^{(12)}$}
-%%
-%\thanks{{}$^{(1)}$ Computation Institute, 
-%University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; \url{dsk@uchicago.edu}}
-%%
-%\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; \url{sctchoi@uchicago.edu}}
-%%
-%\thanks{{}$^{(3)}$ University of California-San Diego, San Diego, CA, USA; \url{wilkinsn@sdsc.edu}}
-%%
-%\thanks{{}$^{(4)}$ Software Sustainability Institute, 
-%University of Edinburgh, Edinburgh, UK; \url{N.ChueHong@software.ac.uk}}
-%%
-%\thanks{{}$^{(5)}$ University of Huddersfield, School of Computing and Engineering, Huddersfield, UK; \url{C.Venters@hud.ac.uk}}
-%%
-%\thanks{{}$^{(6)}$ University of Texas at Austin, Austin, TX, USA; \url{jhowison@ischool.utexas.edu}}
-%%
-%\thanks{{}$^{(7)}$ Netherlands eScience Centre, Amsterdam, Netherlands; \url{F.Seinstra@esciencecenter.nl}}
-%%
-%\thanks{{}$^{(8)}$ National Center for Ecological Analysis and Synthesis, Santa Barbara, CA, USA; \url{jones@nceas.ucsb.edu}}
-%%
-%\thanks{{}$^{(9)}$ National Evolutionary Synthesis Center, Durham, NC, USA; \url{karen.cranston@nescent.org}}
-%%
-%\thanks{{}$^{(10)}$ NASA Goddard Space Flight Center, Greenbelt, MD, USA; \url{Thomas.L.Clune@nasa.gov}}
-%%
-%\thanks{{}$^{(11)}$ Department of Astrophysical Sciences, 
-%Princeton University, Princeton, NJ, USA; \url{mdevalbo@astro.princeton.edu}}
-%%
-%\thanks{{}$^{(12)}$ University of Saarland, Germany; \url{richard.littauer@gmail.com}}
-%%
+%
+\thanks{{}$^{(1)}$ Computation Institute, 
+University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; d.katz@ieee.org}
+%
+\thanks{{}$^{(2)}$ institution, city, country; email}
+%
+\thanks{{}$^{(3)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
+%
+\thanks{{}$^{(4)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(5)}$ Center for Computation \& Technology, Louisiana State University, Baton Rouge, LA, USA; \{knarf,sbrandt\}@cct.lsu.edu}
+%
+\thanks{{}$^{(6)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(7)}$ RENCI, Chapel Hill, NC, USA; rayi@renci.org}
+%
+\thanks{{}$^{(8)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(9)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(10)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(11)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(12)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA, country; gdallen@illinois.edu}
+%
+\thanks{{}$^{(13)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; C.Venters@hud.ac.uk}
+%
+\thanks{{}$^{(14)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(15)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(16)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(17)}$ unit, institution, city, country; email}
+%
+\thanks{{}$^{(18)}$ Department of Astrophysical Sciences, 
+Princeton University, Princeton, NJ, USA; mdevalbo@astro.princeton.edu}
+%
 

### 2016-01-04 WSSSPE3/report/WSSSPE3_review.tex
fixing typo
JH author details
Added Niemeyer author info

@@
 \author{Daniel S. Katz$^{(1)}$,
-Kyle Niemeyer$^{(2)}$,
+Kyle E.\ Niemeyer$^{(2)}$,
 Sou-Cheng T. Choi$^{(3)}$,
@@
 Mark A. Miller$^{(7)}$,
-Sandra Gessing$^{(9)}$,
+Sandra Gesing$^{(9)}$,
 Nick D. Jones$^{(10)}$,
@@
 %
-\thanks{{}$^{(2)}$ institution, city, country; email}
+\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
+Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
 %
@@
 %
-\thanks{{}$^{(4)}$ unit, institution, city, country; email}
+\thanks{{}$^{(4)}$ Research Software Development Group, University College London, UK; j.hetherington@ucl.ac.uk}
 %
@@
 in part by the National Science Foundation research grant DMS-1522687 and
-a WSSSPE3 travel award.
+a WSSSPE3 travel award. Hetherington was funded by
+the Software Sustainability Institute, RCUK grants EP/H043160/1 and EP/N006410/1.
 

### 2016-01-05 WSSSPE3/report/WSSSPE3_review.tex
adding Mark Miller info

@@
 Steven R. Brandt$^{(5)}$,
-Mark A. Miller$^{(7)}$,
+Mark A. Miller$^{(8)}$,
 Sandra Gesing$^{(9)}$,
@@
 %
-\thanks{{}$^{(8)}$ unit, institution, city, country; email}
+\thanks{{}$^{(8)}$ University of California, San Diego, CA, USA; mmiller@sdsc.edu}
 %
@@
 %
-\thanks{{}$^{(10)}$ unit, institution, city, country; email}
+\thanks{{}$^{(10)}$ New Zealand eScience Infrastructure (NeSI), University of Auckland, Auckland, NZ; nick.jones@nesi.org.nz}
 %
@@
 %
-\thanks{{}$^{(12)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA, country; gdallen@illinois.edu}
+\thanks{{}$^{(12)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA; gdallen@illinois.edu}
 %

### 2016-01-05 WSSSPE3/report/wg_appendix_user_community.tex
moved all emails to footnotes for consistency

@@
 POC:
-Dan Gunter <dkgunter@lbl.gov> and 
-Ethan Davis <edavis@ucar.edu>
+Dan Gunter\footnote{email: \href{mailto:dkgunter@lbl.gov}{dkgunter@lbl.gov}} and 
+Ethan Davis\footnote{email: \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}}
 
@@
 
-No definite plans were agreed upon for future organizaton. The major ideas discussed
+No definite plans were agreed upon for future organization. The major ideas discussed
 were coordinating with another group or adapting some existing text.

### 2016-01-05 WSSSPE3/report/wg_main_publishing_SW.tex
Replied to Dan's note.

@@
 
-The plan is to create a web page describing executable papers, their value, and
-a list of what publishers support them. \katznote{when?}
+The plan is to create and curate a web page describing executable papers, their value, and
+a list of what publishers support them. We expect the page to be available in
+early January of 2016 on the Software Sustainability Institute's website.
 

### 2016-01-05 WSSSPE3/report/wg_main_user_community.tex

@@
 This group doesn't have a landing page yet. Please send requests to join and contribute to both
-Dan Gunter <dkgunter@lbl.gov> and Ethan Davis <edavis@ucar.edu>.
+Dan Gunter\footnote{email: \href{mailto:dkgunter@lbl.gov}{dkgunter@lbl.gov}} and Ethan Davis\footnote{email: \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}}.

### 2016-01-05 WSSSPE3/report/WSSSPE3_review.tex
author info for myself (dan gunter)
added acknowledgments for Dan Gunter

@@
 %
-\thanks{{}$^{(6)}$ unit, institution, city, country; email}
+\thanks{{}$^{(6)}$ Lawrence Berkeley National Laboratory, Berkeley, USA; dkgunter@lbl.gov}
 %
@@
 %
-\thanks{{}$^{(15)}$ unit, institution, city, country; email}
+\thanks{{}$^{(15)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
 %
@@
 
+Work by Gunter was supported by the Office of Science, Office of Biological and 
+Environmental Research, of the US Department of Energy under Award Numbers 
+DE-AC02-05CH11231, DE-AC02-06CH11357, DE-AC05-00OR22725, and DE-AC02-98CH10886, 
+as part of the DOE Systems Biology Knowledgebase and by the Office of Science, 
+Office of Advanced Scientific Computing Research (ASCR) of the U.S. Department 
+of Energy under Contract Number DE-AC02-05CH11231 as part of the Template
+Interfaces for Agile Parallel Data-Intensive Science (TIGRES) project.
+
 \todo{feel free to add stuff here}

### 2016-01-05 WSSSPE3/report/WSSSPE3_review.tex
removing some line feeds
Small changes

@@
 computing and optimization issues in mind; but the SciPy community thinks that
-software is becoming better, with emerging tools such as Jupiter and
+software is becoming better, with emerging tools such as Jupyter and
 productivity packages for research workflows. Admitting such comparisons are
@@
 community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
-Often times, a person funded to work full time on a scientific project can spend
+Oftentimes, a person funded to work full time on a scientific project can spend
 a small amount of time for working on a piece of software necessary for that
@@
 the Software Sustainability Institute, RCUK grants EP/H043160/1 and EP/N006410/1.
-
+%
 Work by Gunter was supported by the Office of Science, Office of Biological and 
@@
 Interfaces for Agile Parallel Data-Intensive Science (TIGRES) project.
-
+%
 \todo{feel free to add stuff here}

### 2016-01-05 WSSSPE3/report/WSSSPE3_review.tex

@@
 the open source software
-community\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}.
+community\footnote{\url{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}}.
 Oftentimes, a person funded to work full time on a scientific project can spend

### 2016-01-06 WSSSPE3/report/WSSSPE3_review.tex
Update with affiliation/email
minor change

@@
 
-\item How can we reduce systemic bias, which goes back to the blog post of
-Dryden especially on how ethics of unpaid labor disproportionately affect
+\item How can we reduce systemic bias, which goes back to Dryden's blog post
+especially on how ethics of unpaid labor disproportionately affect
 underrepresented communities?

### 2016-01-06 WSSSPE3/report/wg_main_funding_spec_expert.tex
remove extra space

@@
 way. However, to bootstrap such a function to a level where it has sufficient
-reputation and client base to be self -sustaining will generally require seed
+reputation and client base to be self-sustaining will generally require seed
 investment.

### 2016-01-06 WSSSPE3/report/WSSSPE3_review.tex

--- 2016-01-05 WSSSPE3/report/WSSSPE3_review.tex
@@
 %
-\thanks{{}$^{(9)}$ unit, institution, city, country; email}
+\thanks{{}$^{(9)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
 %
@@
 has extensive experiences working on large collaborative projects employing
-modern computing tools. He also co-organizes and champions WSSSPE events.
+modern computing tools \cite{2011ApJS..192....9T}. He also co-organizes and
+champions WSSSPE events.
 

### 2016-01-06 WSSSPE3/report/WSSSPE3_review.tex
tweaks to KBase lightning talk description

@@
 
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, and
-Sergei Maslov: \textit{Challenges of a Sustainable Software Platform for
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham,
+Sergei Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software Platform for
 Predictive Biology: Lessons Learned on the KBase Project}} In this talk, Dan
@@
 context of knowledge in the field. Being a large, multi-institutional project,
-one of the big challenges is to agree on standard to enable a single,
+one of the big challenges is to agree on standards to enable a single,
 maintainable system. Working in isolation does not work (anymore) within this
-field, and the community in the field also didn't have standards for software
-engineering. This is contrasted to computer science research, where these
-approaches shorten design cycles, leading to more time for highly rewarded
+field, and the community in the field also does not have standards for software
+engineering. This is contrasted to computer science research, where software
+engineering standards shorten design cycles, leading to more time for highly rewarded
 activities like publishing, performance studies, graduating students, or
@@
 After picking projects and team members, four to five teams work on projects for about two
-weeks before a one-week evaluation by the executive committee, and after which new
-teams and projects might be chosen to restart the ``agile'' development cycle.
+weeks before a one-week evaluation by the executive committee. Based on their assessment new
+teams and projects might be chosen before iteratively restarting the ``agile'' development cycle.
 This process is intentionally open and documented.

### 2016-01-07 WSSSPE3/report/registrants.tex

--- 2015-09-30 WSSSPE3/report/registrants.tex
@@
 Iain Larmour & EPSRC (UK)\\
-Frank Löffler & Louisiana State University\\
+Frank L\"offler & Louisiana State University\\
 Suresh Marru & Indiana University\\

### 2016-01-07 WSSSPE3/report/WSSSPE3_review.tex
removed extra comma after last author
adding more orgs/emails

@@
 Abani Patra$^{(17)}$,
-Miguel de Val-Borro$^{(18)}$,
+Miguel de Val-Borro$^{(18)}$
 }
@@
 %
-\thanks{{}$^{(14)}$ unit, institution, city, country; email}
+\thanks{{}$^{(14)}$ UCAR/Unidata, Boulder, CO, USA; edavis@unidata.ucar.edu}
 %
@@
 %
-\thanks{{}$^{(16)}$ unit, institution, city, country; email}
+\thanks{{}$^{(16)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
 %
-\thanks{{}$^{(17)}$ unit, institution, city, country; email}
+\thanks{{}$^{(17)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
 %

### 2016-01-11 WSSSPE3/report/WSSSPE3_review.tex
small change in author affiliation

@@
 %
-\thanks{{}$^{(14)}$ UCAR/Unidata, Boulder, CO, USA; edavis@unidata.ucar.edu}
+\thanks{{}$^{(14)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
 %
@@
 \thanks{{}$^{(18)}$ Department of Astrophysical Sciences, 
-Princeton University, Princeton, NJ, USA; mdevalbo@astro.princeton.edu}
+Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
 %

### 2016-01-12 WSSSPE3/report/wg_appendix_industry_interaction.tex
Create wg_appendix_industry_interaction.tex
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Nic's changes, plus a couple of small edits

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Industry Interaction Working Group Discussion}
+\label{sec:appendix_industry_interaction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Nic Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} will serve as the point of contact for this working group.
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+
+\begin{itemize}
+\item Nic Weber -- University of Washington
+\item Suresh Marru -- Indiana University
+\item Jeffrey Carver -- University of Alabama
+\item Davide DelVento -- NCAR/CISL
+\item Steven Brandt -- Louisiana State University
+\end{itemize}
 
 \subsection{Summary of Discussion}
+
+Our initial broad question was, "What makes for successful transitions of scientific software from academia to industry?"
+
+There are a number of potential funding transitions that may occur:  
+%
+\begin{itemize}
+
+\item A project could be \textbf{refunded}, and development or maintenance of
+the software continue as planned.
+
+\item A project might locate a \textbf{new source of funding} in which case the
+software may be further developed or simply maintained as before.
+
+\item The project could transition to a \textbf{community supported model}
+whereby ownership, maintenance, and stewardship of the software become similar
+to peer-production models in open-source (e.g., see
+Howison~\cite{howison_sustaining_2015}).
+
+\item The project could receive some form of industry sponsorship in which case
+ownership of the intellectual property, licensing, maintenance activities,
+hosting, etc.\ may change significantly.
+
+\end{itemize}
+
+We characterized each of these potential changes in funding as ``transition
+pathways'' to sustainable software (see similar work by Geels and
+Schot~\cite{Geels:2007}).
+
+Our work at WSSSPE3 included the following activities (described in detail
+below): (1) brainstorming goals for this type of research, (2) imagining
+potential outcomes of completing a set of case studies on this topic, and (3)
+generating a set of working definitions for some of the broad concepts we are
+describing.
+
+First, we discussed the \textbf{goals} of this research, attempting to answer the 
+question \emph{What is the goal of doing research on transition pathways?}
+A number of research questions arose:  Can we identify collaborations that have 
+occurred and try to understand which were successful, which were unsuccessful, 
+and what factors contributed to these successes/failures? Can we determine what 
+each partner wants to get out of such a collaboration? For example, why would 
+industry be interested in collaborating with academia? Or why would academia 
+be interested in collaborating with industry? How could we design a study that 
+focused on the impact of the software in undergoing this type of transition?
+
+Next, we imagined \textbf{potential outcomes} of research on this topic, involving 
+a set of case studies that look at successful and unsuccessful
+transitions of researchers between academia and industry. This might address 
+each of the transition types (described below). Successful transitions are
+described as those that lead to either weak or strong sustainability (also
+defined below). In addition, the results from this research might help create a 
+generalizable framework that might allow for the study of different transition 
+pathways (other than academia to industry).
+
+Finally, we created some \textbf{general definitions} for these concepts; we 
+characterize transitions in the following ways:
+\begin{itemize}
+
+\item Handoff model: academia initially writes the software, industry (for-profit 
+or nonprofit) then takes over the project.
+
+\item Co-Production Model: industry and academia interact throughout development
+of the project.
+
+\item Sponsorship Model: academia writes and maintains the software; 
+industry contributes funding for the development\slash maintenance of software.
+In this example, industry is also likely a user of the software.
+
+\item Spinoff model: transition to a for-profit or non-profit company owned by or in
+collaboration with original developers.
+
+\end{itemize}
+
+We characterized sustainability in the following ways:
+\begin{itemize}
+
+\item Weak Sustainability: Software continues to be accessible, useful, and
+usable.
+
+\item Strong Sustainability: Software meets criteria above, but is also able to
+be reused for further innovation (i.e., issued non-restrictive open-source
+license).
+
+\end{itemize}
+We refer readers to Becker et al.~\cite{Becker:2014} for an extended discussion of weak versus strong
+sustainability.
 
@@
 \subsection{Key Next Steps}
+
+Identify projects that are collaborative, perhaps by reviewing funded projects from programs specifically geared towards industry academic collaborations.
+
+Develop a systematic process for conducting case studies (what kind of data are being gathered about each case)
 

### 2016-01-12 WSSSPE3/report/wg_main_industry_interaction.tex
Nic's changes, plus a couple of small edits

--- 2015-11-15 WSSSPE3/report/wg_main_industry_interaction.tex
@@
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \subsection{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration}
+\label{sec:industry_interaction}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-There are a number of potential funding transitions that may occur:  
-%
-\begin{itemize}
-
-\item A project could be \textbf{refunded}, and development or maintenance of
-the software continue as planned.
-
-\item A project might locate a \textbf{new source of funding} in which case the
-software may be further developed or simply maintained as before.
-
-\item The project could transition to a \textbf{community supported model}
-whereby ownership, maintenance, and stewardship of the software become similar
-to peer-production models in open-source (e.g., see
-Howison~\cite{howison_sustaining_2015}).
-
-\item The project could receive some form of industry sponsorship in which case
-ownership of the intellectual property, licensing, maintenance activities,
-hosting, etc.\ may change significantly.
-
-\end{itemize}
-
-We characterize each of these potential changes in funding as ``transition
-pathways'' to sustainable software (see similar work by Geels and
-Schot~\cite{Geels:2007}).
-
 At WSSSPE3 our group was interested in better understanding successful pathways
@@
 
-Although transitions are often studied under the broad umbrella of ``technology
+\subsubsection{Fit with related activities}
+
+Most previous research and discussion of industry and academic collaboration, sharing, and funding of research software has focused on the impact of such arrangements.
+Examples of these types of reports are as follows:
+
+\begin{itemize}
+\item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
+\item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
+\item Dowling Review from the UK: addresses complexity of work between these two
+communities: \url{http://www.raeng.org.uk/policy/dowling-review}
+\item Pathway to Impact - UK report: two pages of grant proposals are asked to
+forecast what impact they might have (including environmental, academic, economic).
+\end{itemize}
+
+\subsubsection{Discussion}
+
+Although sustainability transitions are often studied under the broad umbrella of ``technology
 transfer,'' we believe there are likely to be a number of different ways in which
-a pathway from initial production to long-term maintenance and secure funding are
+a pathway from initial production to long-term maintenance and secure funding is
 achieved. In short, industry sponsorship is an important aspect of sustaining
 scientific software, but our current understanding of these transitions focuses
-narrowly on commercial successes\slash failures.
-
-\subsubsection{Fit with related activities}
+narrowly on commercial successes or failures of those collaborations.
 
 In looking at existing literature that addresses industry transitions, many
-reports focus on benefits that accrue to the private sector, or to a government
+reports (such as those listed above) focus on benefits that accrue to the private sector, or to a government
 that originally sponsored the research project. This literature does not address
@@
 
-Examples of the former scenario (benefits accruing to private sector) are as
-follows:
-\begin{itemize}
-
-\item REF Impact Case Studies: \url{http://impact.ref.ac.uk/CaseStudies/}
-
-\item Background of projects funded in the UK: \url{http://gtr.rcuk.ac.uk/}
-
-\item Dowling Review from the UK: addresses complexity of work between these two
-communities: \url{http://www.raeng.org.uk/policy/dowling-review}
-
-\item Pathway to Impact - UK report: two pages of grant proposals are asked to
-forecast what impact they might have (including environmental, academic, economic).
-
-\end{itemize}
-
-\subsubsection{Discussion}
-
-Our work at WSSSPE3 included the following activities (described in detail
-below): (1) brainstorming goals for this type of research, (2) imagining
-potential outcomes of completing a set of case studies on this topic, and (3)
-generating a set of working definitions for some of the broad concepts we are
-describing.
-
-\textbf{Goal}
-
-\emph{What is the goal of doing research on transition pathways?}
-
-Can we identify collaborations that have occurred and try to understand which were successful, which were unsuccessful, and what factors contributed to these successes/failures?
-
-Can we determine what each partner wants to get out of such a collaboration?
-Why would industry be interested in collaborating with academia? 
-Why would academia be interested in collaborating with industry?
-
-How could we design a study that focused on the impact of the software in
-undergoing this type of transition?
-
-\textbf{Potential outcomes}
-
-Complete a set of case studies which look at successful and unsuccessful
-transitions of researchers between academia and industry. This might address 
-each of the transition types (described below). Successful transitions are
-described as those that lead to either weak or strong sustainability (also
-defined below).
-
-Create a generalizable framework that might allow for the study of different
-transition pathways (other than academia to industry).
-
-\textbf{General Definitions}
-
-We characterize transitions in the following ways:
-\begin{itemize}
-
-\item Handoff model: academia initially writes the software, industry (for-profit 
-or nonprofit) then takes over the project.
-
-\item Co-Production Model: industry and academia interact throughout development
-of the project.
-
-\item Sponsorship Model: academia writes and maintains the software; the
-industry contributes funding for the development\slash maintenance of software.
-In this example, industry is also likely a user of the software.
-
-\item Spinoff model: transition to for- or non-profit company owned by or in
-collaboration with original developers.
-
-\end{itemize}
-
-We characterize sustainability in the following ways:
-\begin{itemize}
-
-\item Weak Sustainability: Software continues to be accessible, useful, and
-usable.
-
-\item Strong Sustainability: Software meets criteria above, but is also able to
-be reused for further innovation (i.e., issued non-restrictive open-source
-license).
-
-\end{itemize}
-We refer readers to Becker et al.~\cite{Becker:2014} for extended discussion of weak versus strong
-sustainability. For more detail on the group's discussion, see
+For more detail on the group's discussion, see
 Appendix~\ref{sec:appendix_industry_interaction}.
@@
 
+The broad goals that we would like to accomplish are:
+
+\begin{enumerate}
+\item To complete a set of case studies which look at successful and unsuccessful transitions between academic researchers and industry
+\item To create a generalizable framework, which might allow for a broader study of different transition pathways (other than between academia and industry)
+\end{enumerate}
+
+The main plan for the group going forward is the creation of a white paper on the topic of sustainability transitions.
+
 \subsubsection{Landing Page}
 
-To find more information about this group, or join it, see \todo{where should
-someone go who want to know more about this and perhaps wants to contribute?}
+In lieu of a landing page, the Transitions Pathways
+group requests an email be sent to Nic Weber\footnote{email: 
+\href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} to find out more
+about the group's efforts and how to participate.

### 2016-01-15 WSSSPE3/report/WSSSPE3_review.tex
adding final author info

@@
 %
-\thanks{{}$^{(11)}$ unit, institution, city, country; email}
+\thanks{{}$^{(11)}$ University of Washington, Seattle, WA, USA; nmweber@uw.edu}
 %

### 2016-01-15 WSSSPE3/report/WSSSPE3_review.tex
Update WSSSPE3_review.tex
Minor update. Added Birgit Penzenstadler as an author.

@@
 Gabrielle Allen$^{(12)}$,
+Birgit Penzenstadler$^{(19)}$,
 Colin C. Venters$^{(13)}$,
@@
 %
-\thanks{{}$^{(13)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; C.Venters@hud.ac.uk}
+\thanks{{}$^{(13)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; c.venters@hud.ac.uk}
 %
@@
 %
-\thanks{{}$^{(18)}$ Department of Astrophysical Sciences, 
-Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
+\thanks{{}$^{(18)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
+%
+\thanks{{}$^{(19)}$ College of Computer Engineering \& Computer Science, California State University, Long Beach, CA, USA; birgit.penzenstadler@csulb.edu}
 %

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Create wg_appendix_eng_design.tex
Former-commit-id: b358137b07892ef97a717f838b4d279978b4900d
Reformatted group members to itemized list to ensure they appear in the right place.
Former-commit-id: e4584de998d67127bdd327661163fca316f83c01
adding POC request to appendices
Former-commit-id: 42aa40a42eb82bf40df6e482de5fb48dc9085b8d
Update wg_appendix_eng_design.tex
Group members added.

--- 2015-09-30 WSSSPE3/report/wg_appendix_template.tex
@@
-\section{Software Credit Working Group Discussion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Engineering Design Group Discussion}
+\label{sec:appendix_eng_design}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{add POC here}
 
 \subsection{Group Members}
-{\small
-\begin{longtable}{ll}
-   Alice Allen           &  Astrophysics Source Code Library (ASCL)
-\\ Daniel S. Katz        &  University of Chicago \& Argonne National Laboratory
-\end{longtable}
-}
+
+\begin{itemize}
+\item Birgit Penzenstadler -- California State University, CA, USA
+\item Colin C. Venters -- University of Huddersfield, Huddersfield, UK
+\item Matthias Bussonnier -- UC Berkeley
+\item Jeff McWhirter -- Geode Systems
+\item Patrick Nichols -- National Center for Atmospheric Research
+\item Ilian Todorov -- Science \& Technology Facilities Council, UK\\
+\item Ian Taylor -- Cardiff University
+\item Alexander Vyushkov -- University of Notre Dame
+
+\end{itemize}
 

### 2016-01-15 WSSSPE3/report/WSSSPE3_review.tex
keep author names together, fixing order of affilations

--- 2016-01-15 WSSSPE3/report/WSSSPE3_review.tex
@@
 James Hetherington$^{(4)}$,
-Frank L\"{o}ffler$^{(5)}$,
+Frank~L\"{o}ffler$^{(5)}$,
 Dan Gunter$^{(6)}$,
@@
 Gabrielle Allen$^{(12)}$,
-Colin C. Venters$^{(13)}$,
-Ethan Davis$^{(14)}$,
-Lorraine Hwang$^{(15)}$,
-Ilian Todorov$^{(16)}$,
-Abani Patra$^{(17)}$,
-Miguel de Val-Borro$^{(18)}$
+Birgit~Penzenstadler$^{(13)}$,
+Colin C. Venters$^{(14)}$,
+Ethan Davis$^{(15)}$,
+Lorraine Hwang$^{(16)}$,
+Ilian~Todorov$^{(17)}$,
+Abani Patra$^{(18)}$,
+Miguel de Val-Borro$^{(19)}$
 }
@@
 %
-\thanks{{}$^{(13)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; C.Venters@hud.ac.uk}
+\thanks{{}$^{(13)}$ College of Computer Engineering \& Computer Science, California State University, Long Beach, CA, USA; birgit.penzenstadler@csulb.edu}
 %
-\thanks{{}$^{(14)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
+\thanks{{}$^{(14)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; c.venters@hud.ac.uk}
 %
-\thanks{{}$^{(15)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
+\thanks{{}$^{(15)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
 %
-\thanks{{}$^{(16)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
+\thanks{{}$^{(16)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
 %
-\thanks{{}$^{(17)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
+\thanks{{}$^{(17)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
 %
-\thanks{{}$^{(18)}$ Department of Astrophysical Sciences, 
-Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
+\thanks{{}$^{(18)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
+%
+\thanks{{}$^{(19)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
 %

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
POC added

@@
 
-\todo{add POC here}
+Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
+Venters\footnote{email: \href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as
+the points of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
 
@@
 \item Colin C. Venters -- University of Huddersfield, Huddersfield, UK
-\item Matthias Bussonnier -- UC Berkeley
+\item Matthias Bussonnier -- UC Berkeley, CA, USA
 \item Jeff McWhirter -- Geode Systems 
-\item Patrick Nichols -- National Center for Atmospheric Research
-\item Ilian Todorov -- Science \& Technology Facilities Council, UK\\
-\item Ian Taylor -- Cardiff University
-\item Alexander Vyushkov -- University of Notre Dame
-
+\item Patrick Nichols -- National Center for Atmospheric Research, CO, USA
+\item Ilian Todorov -- Science \& Technology Facilities Council, UK
+\item Ian Taylor -- Cardiff University, UK
+\item Alexander Vyushkov -- University of Notre Dame, IN, USA
 \end{itemize}
@@
 \subsection{Summary of Discussion}
+
+This group discussed the principles of engineering design for sustainable software and their application in various domains.
+The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
+Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion.
+There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community.
+Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
+It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
+The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK):
 

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Main text added to the discussion section.

@@
 Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
-Venters\footnote{email: \href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as
-the points of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+Venters\footnote{email: \href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
 
@@
 
-This group discussed the principles of engineering design for sustainable software and their application in various domains.
+Software engineering principles form the basis of methods, techniques, methodologies and tools.
+This group discussed the principles of software engineering design for sustainable software and their application in various domains.
 The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
@@
 It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK):
+The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}:
+
+Software design principles:
+\begin{itemize}
+\item Abstraction;
+\item Coupling and cohesion;
+\item Decomposition and modularization;
+\item Encapsulation and information hiding;
+\item Separation of interface and implementation;
+\item Sufficiency completeness & primitiveness;
+\item Separation of concerns.
+\end{itemize}
+
+User interface design principles:
+\begin{itemize}
+\item Learnability;
+\item User familiarity;
+\item Consistency;
+\item Minimal surprise;
+\item Recoverability;
+\item User guidance;
+\item User diversity
+\end{itemize}
+
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{karlskrona,becker2015}:
+\begin{itemize}
+\item Sustainability is systemic;
+\item ...is multidimensional
+\item ...is interdisciplinary;
+\item ...transcends the system’s purpose;
+\item ...applies to both a system and its wider contexts;
+\item ...requires action on multiple levels;
+\item ...requires multiple timescales;
+\item Changing design to take into account long-term effects doesn’t automatically imply sacrifices;
+\item System visibility is a precondition for and enabler of sustainability design.
+\end{itemize}
+
+This congregated list is an initial collection of principles that could be extended by adding from further related work from separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
+The group identified two example systems to discuss the application of the principles.
+The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
+The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
+The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
+The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community.
+In addition, two challenges were pointed out: mapping of the
+principles to best practice, and demonstrating the return on investment of those best practices.
 
@@
 \subsection{Key Next Steps}
+The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
 

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Updates to main body of text

@@
 \subsection{Summary of Discussion}
-
 Software engineering principles form the basis of methods, techniques, methodologies and tools.
@@
 The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community.
-In addition, two challenges were pointed out: mapping of the
-principles to best practice, and demonstrating the return on investment of those best practices.
 
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community.
+In addition, two challenges were pointed out:
+\begin{itemize}
+\item Mapping of the principles to best practice.
+\item Demonstrating the return on investment of those best practices.
+\end{itemize}
 

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
fix to make latex build
fixing quotes and apostrophes

@@
 \subsection{Description of Opportunity, Challenges, and Obstacles}
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into “bite sized” chunks for the Computational Science and Engineering Community.
+The opportunity was identified to distill existing software engineering and sustainability design knowledge into ``bite sized'' chunks for the Computational Science and Engineering Community.
 In addition, two challenges were pointed out:

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
changing cite to match .bib entry
Update wg_appendix_eng_design.tex
Minor updates.

@@
 
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{karlskrona,becker2015}:
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{karlskrona,SEIS2015}:
 \begin{itemize}
@@
 The opportunity was identified to distill existing software engineering and sustainability design knowledge into ``bite sized'' chunks for the Computational Science and Engineering Community.
-In addition, two challenges were pointed out:
+In addition, two primary challenges were identified:
 \begin{itemize}
@@
 The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
+Preliminary work will be carried out through undergraduate or post-graduate student projects.
 

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
More updates to main body.

@@
 
+A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including:
+
+\begin{itemize}
+\item Continual Refinement of Product and Project Practices
+\item A Working Product at All Times
+\item A Continual Emphasis on Design
+\item Value Defect Prevention over Defect Detection
+\end{itemize}
+
 This congregated list is an initial collection of principles that could be extended by adding from further related work from separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
-The group identified two example systems to discuss the application of the principles.
+This riased the question, What do the principles mean for scientific and computational software?
+
+The group then identified two example systems to discuss the application of the principles.
 The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.

### 2016-01-15 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Minor update

@@
 \subsection{Description of Opportunity, Challenges, and Obstacles}
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into ``bite sized'' chunks for the Computational Science and Engineering Community.
+The opportunity was identified to distill existing software engineering and sustainability design knowledge into ''bite sized'' chunks for the Computational Science and Engineering Community.
 In addition, two primary challenges were identified:

### 2016-01-15 WSSSPE3/report/wg_main_best_practices.tex
added missing periods

@@
 \subsubsection{Landing Page}
-The landing page with instructions, timeline and the white paper is here: \url{https://drive.google.com/drive/folders/0B7KZv1TRi06fbnFkZjQ0ZEJKckk}
-Discussions can be also continued in \url{https://github.com/WSSSPE/meetings/issues/42}
+The landing page with instructions, timeline and the white paper is here: \url{https://drive.google.com/drive/folders/0B7KZv1TRi06fbnFkZjQ0ZEJKckk}.
+Discussions can be also continued in \url{https://github.com/WSSSPE/meetings/issues/42}.

### 2016-01-16 WSSSPE3/report/wg_appendix_industry_interaction.tex
change in appendix title to match main section
beefing up 5.3 appendix

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Industry Interaction Working Group Discussion}
+\section{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration Working Group Discussion}
 \label{sec:appendix_industry_interaction}
@@
 
-Our initial broad question was, "What makes for successful transitions of scientific software from academia to industry?"
-
-There are a number of potential funding transitions that may occur:  
+Our initial broad question was, "What makes for successful transitions of scientific software from academia to industry?" There are a number of potential funding transitions that may occur:  
 %
@@
 whereby ownership, maintenance, and stewardship of the software become similar
-to peer-production models in open-source (e.g., see
-Howison~\cite{howison_sustaining_2015}).
+to peer-production models in open-source (e.g., see Howison~\cite{howison_sustaining_2015}).
 
@@
 hosting, etc.\ may change significantly.
+
+\item The project could gain attention from a industry use case who would potentially make in-kind
+contributions by having paid staff contribute to the software.
 
@@
 \subsection{Description of Opportunity, Challenges, and Obstacles}
+The opportunity is to create a catalog of success/failure for current and future software projects to be prepared for transitions and achieve sustainability of the software.
+
+The obstacle is more superficial, in finding a champion to gather such information.
+It will be a challenge to keep this information and surveys updated.
+With changing rapidly changing industry landscapes, an obsolete survey could be of less or no use.
 
@@
 \subsection{Plan for Future Organization}
+No concrete plans were made at this point. If community can rally behind, some momentum could be built. If you are interested post at \url{https://github.com/WSSSPE/meetings/issues/46}
 
 \subsection{What Else is Needed?}
+Nothing at the moment.
 
 \subsection{Key Milestones and Responsible Parties}
+A key portion of this effort will require focused surveys of projects which have succeeded and failed in transition.
+Both these categories will yield good learning on what works and what does not work.
+The group has identified what needs to be studied further, but has not identified responsible parties to conduct them.
+
+Community could help in gathering data by means of Interviews; Historical documents / documentation and Surveys
+
+An example data collection will be: 
+\begin{itemize}
+\item Origin: Where did project start? 
+\item People involved: How many people in original project were involved in transition/collaboration? 
+\item Specs on software
+\item Language
+\item Size
+\item Hardness (age)
+\item Lead-up to Transition: How long was project in development before it began transition?
+\item Motivation for Transition: Why was transition initiated? By whom? 
+\end{itemize}
 
 \subsection{Description of Funding Needed}
+Concrete funding needs were not discussed in this working group but a general impression was a seed funding will motivate members of this group or others in community to launch a survey effort.

### 2016-01-16 WSSSPE3/report/wg_main_industry_interaction.tex
beefing up 5.3 appendix

@@
 
-Most scientific software is produced as a part of grant-funded research projects
+Most scientific software is produced as a part of grant-funded research projects typically 
 sponsored by federal governments. If we are interested in the sustainability of
@@
 sponsorship ends. More than likely, the project and its resulting software will
-need to undergo some kind of transition in funding and consequently management.
+need to undergo some kind of transition in funding and consequently governance.
 
 At WSSSPE3 our group was interested in better understanding successful pathways
-for scientific software to ``transition'' from government-funded research
+for scientific software to ``transition'' from grant-funded research
 projects to industry sponsorship. (This may be an initially awkward
@@
 a pathway from initial production to long-term maintenance and secure funding is
-achieved. In short, industry sponsorship is an important aspect of sustaining
+achieved. In short, industry sponsorship and/or direct participation is an important aspect of sustaining
 scientific software, but our current understanding of these transitions focuses
@@
 \subsubsection{Landing Page}
-
-In lieu of a landing page, the Transitions Pathways
-group requests an email be sent to Nic Weber\footnote{email: 
-\href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} to find out more
-about the group's efforts and how to participate.
+Transitions Pathways discussions can be posted at \url{https://github.com/WSSSPE/meetings/issues/46} or an email be sent to Nic Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} to find out more about the group's efforts and how to participate.

### 2016-01-16 WSSSPE3/report/WSSSPE3_review.tex
adding author

@@
 Nic Weber$^{(11)}$,
-Gabrielle Allen$^{(12)}$,
-Birgit~Penzenstadler$^{(13)}$,
-Colin C. Venters$^{(14)}$,
-Ethan Davis$^{(15)}$,
-Lorraine Hwang$^{(16)}$,
-Ilian~Todorov$^{(17)}$,
-Abani Patra$^{(18)}$,
-Miguel de Val-Borro$^{(19)}$
+Suresh Marru$^{(12)}$,
+Gabrielle Allen$^{(13)}$,
+Birgit~Penzenstadler$^{(14)}$,
+Colin C. Venters$^{(15)}$,
+Ethan Davis$^{(16)}$,
+Lorraine Hwang$^{(17)}$,
+Ilian~Todorov$^{(18)}$,
+Abani Patra$^{(19)}$,
+Miguel de Val-Borro$^{(20)}$
 }
@@
 %
-\thanks{{}$^{(12)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA; gdallen@illinois.edu}
+\thanks{{}$^{(12)}$ Indiana University, Bloomington, IN, USA; smarru@iu.edu}
 %
-\thanks{{}$^{(13)}$ College of Computer Engineering \& Computer Science, California State University, Long Beach, CA, USA; birgit.penzenstadler@csulb.edu}
+\thanks{{}$^{(13)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA; gdallen@illinois.edu}
 %
-\thanks{{}$^{(14)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; c.venters@hud.ac.uk}
+\thanks{{}$^{(14)}$ College of Computer Engineering \& Computer Science, California State University, Long Beach, CA, USA; birgit.penzenstadler@csulb.edu}
 %
-\thanks{{}$^{(15)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
+\thanks{{}$^{(15)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; c.venters@hud.ac.uk}
 %
-\thanks{{}$^{(16)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
+\thanks{{}$^{(16)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
 %
-\thanks{{}$^{(17)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
+\thanks{{}$^{(17)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
 %
-\thanks{{}$^{(18)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
+\thanks{{}$^{(18)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
 %
-\thanks{{}$^{(19)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
+\thanks{{}$^{(19)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
+%
+\thanks{{}$^{(20)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
 %

### 2016-01-16 WSSSPE3/report/wg_appendix_industry_interaction.tex
minor wording changes

@@
 \subsection{Plan for Future Organization}
-No concrete plans were made at this point. If community can rally behind, some momentum could be built. If you are interested post at \url{https://github.com/WSSSPE/meetings/issues/46}
+No concrete plans have been made at this point.
+If the community can rally behind this topic, some momentum could be built.
+Those interested should post at \url{https://github.com/WSSSPE/meetings/issues/46}
 
@@
 \subsection{Description of Funding Needed}
-Concrete funding needs were not discussed in this working group but a general impression was a seed funding will motivate members of this group or others in community to launch a survey effort.
+Concrete funding needs were not discussed in this working group but a general impression was that some seed funding would motivate members of this group or others in community to launch a survey effort.

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Text updated in the correct file.

@@
-\subsection{Principles for Engineering Design for Sustainable Software}
+\subsection{Principles for Software Engineering Design for Sustainable Software}
 
 \subsubsection{Why it is important}
-\todo{short text here}
+
+Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{}.
+However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{}.
+Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{}.
 
 \subsubsection{Fit with related activities}
-\todo{short text here - can include links/cites}
+The group discussion focused on identifying existing principles of software engineering design that could be adopted by the computational science and engineering communities.
 
 \subsubsection{Discussion}
-\todo{short-ish text here}
-This group discussed the principles of engineering design for sustainable software and their application in various domains.
+
 The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
@@
 It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK):
+The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}:
 
@@
 
+A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including:
+
+\begin{itemize}
+\item Continual Refinement of Product and Project Practices
+\item A Working Product at All Times
+\item A Continual Emphasis on Design
+\item Value Defect Prevention over Defect Detection
+\end{itemize}
+
 This congregated list is an initial collection of principles that could be extended by adding from further 
@@
 The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
+This raised the question, What do the principles mean for scientific and computational software?
+Similalry, how do the principles relate to non-functional requirements?
+In addition, what do the five dimensions of sustainability mean for scientific software i.e.
+environmental, economic, social, techincal and individual?
 
 \subsubsection{Landing Page}
-\todo{link to landing page}
+In lieu of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
+Venters\ to find out more about the group's efforts and how to participate.

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor additions and deletions

@@
 analysis of what the developers did in practice against how the principles could have made a 
-difference. The opportunity was identified to distill existing software engineering and sustainability 
-design knowledge into ``bite sized'' chunks for the Computational Science and Engineering 
-Community. In addition, two challenges were pointed out: mapping of the principles to best practice, 
-and  demonstrating the return on investment of those best practices.
+difference.
 
 \subsubsection{Plans}
-The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
-This raised the question, What do the principles mean for scientific and computational software?
-Similalry, how do the principles relate to non-functional requirements?
-In addition, what do the five dimensions of sustainability mean for scientific software i.e.
-environmental, economic, social, techincal and individual?
+The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a set of guidelines on the principles and how they exemplary apply to scientific software system.
 
 \subsubsection{Landing Page}
-In lieu of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
+In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
 Venters\ to find out more about the group's efforts and how to participate.

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor updates

@@
 It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}:
+The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
 
-Software design principles:
-\begin{itemize}
-\item Abstraction;
-\item Coupling and cohesion;
-\item Decomposition and modularization;
-\item Encapsulation and information hiding;
-\item Separation of interface and implementation;
-\item Sufficiency completeness \& primitiveness;
-\item Separation of concerns.
-\end{itemize}
+Software design principles included: Abstraction; Coupling and cohesion; Decomposition and modularization; Encapsulation and information hiding; Separation of interface and implementation; Sufficiency completeness \& primitiveness; and Separation of concerns.
+Similalry, user interface design principles included: Learnability; User familiarity; Consistency; Minimal surprise; Recoverability; User guidance; and User diversity.
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}.
+The maifestio states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
+%\todo{should these sets of bullet points be moved to the appendix? or moved into paragraph form, since they are relatively short items?}
+A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
 
-User interface design principles:
-\begin{itemize}
-\item Learnability;
-\item User familiarity;
-\item Consistency;
-\item Minimal surprise;
-\item Recoverability;
-\item User guidance;
-\item User diversity.
-\end{itemize}
-
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}:
-\begin{itemize}
-\item Sustainability is systemic;
-\item ...is multidimensional
-\item ...is interdisciplinary;
-\item ...transcends the system's purpose;
-\item ...applies to both a system and its wider contexts;
-\item ...requires action on multiple levels;
-\item ...requires multiple timescales;
-\item Changing design to take into account long-term effects doesn't automatically imply sacrifices;
-\item System visibility is a precondition for and enabler of sustainability design.
-\end{itemize}
-\todo{should these sets of bullet points be moved to the appendix? or moved into paragraph form, since they are relatively short items?}
-
-A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including:
-
-\begin{itemize}
-\item Continual Refinement of Product and Project Practices
-\item A Working Product at All Times
-\item A Continual Emphasis on Design
-\item Value Defect Prevention over Defect Detection
-\end{itemize}
-
-This congregated list is an initial collection of principles that could be extended by adding from further 
-related work form separate disciplines within the field of software engineering, including requirements 
-engineering, software architecture, and testing. The group identified two example systems to discuss 
-the application of the principles. The first one was a quantum chemistry system that allows the analysis 
-of the characteristics and capabilities of molecules and solids. The second one was a modeling system 
-for malaria that permitted biologists to analyze a range of datasets across geography, biology, and 
-epidemiology, and add their own datasets. They then examined the principles and took a retrospective 
-analysis of what the developers did in practice against how the principles could have made a 
-difference.
+This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
+The group identified two example systems to discuss the application of the principles.
+The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
+The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
+The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
 

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor update

@@
 
+It was suggested that at the very minimum, the genetic building blocks of [technically] sustainable software should address two core quality attributes:
+\begin{itemize}
+\item Extensibility: the software’s ability to be extended and the level of effort required to implement the extension;
+\item Maintainability: the effort required to locate and fix an error in operational software.
+\end{itemize}
+These fundamental building blocks could then be extended to include other quality attributes such as portability, resuability, scalability, usability, and energy efficiency etc.
+Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software.
+This raised the question, What do the principles mean for scientific and computational software?
+Similalry, how do the principles relate to non-functional requirements?
+In addition, what do the five dimensions of sustainability mean for scientific software i.e.
+environmental, economic, social, techincal and individual?
+
 \subsubsection{Plans}

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor updates

@@
 The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
-
-It was suggested that at the very minimum, the genetic building blocks of [technically] sustainable software should address two core quality attributes:
+This raised the question, What do the principles mean for  computational scientific and engineering software?
+Similalry, how do the principles relate to non-functional requirements?
+It was suggested that at the very minimum, that sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software:
 \begin{itemize}
@@
 Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software.
-This raised the question, What do the principles mean for scientific and computational software?
-Similalry, how do the principles relate to non-functional requirements?
 In addition, what do the five dimensions of sustainability mean for scientific software i.e.

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Minor changes

@@
 This group discussed the principles of software engineering design for sustainable software and their application in various domains.
-The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
-Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion.
-There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community.
-Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
-It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}:
-
-Software design principles:
-\begin{itemize}
-\item Abstraction;
-\item Coupling and cohesion;
-\item Decomposition and modularization;
-\item Encapsulation and information hiding;
-\item Separation of interface and implementation;
-\item Sufficiency completeness \& primitiveness;
-\item Separation of concerns.
-\end{itemize}
-
-User interface design principles:
-\begin{itemize}
-\item Learnability;
-\item User familiarity;
-\item Consistency;
-\item Minimal surprise;
-\item Recoverability;
-\item User guidance;
-\item User diversity
-\end{itemize}
-
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{karlskrona,SEIS2015}:
-\begin{itemize}
-\item Sustainability is systemic;
-\item ...is multidimensional
-\item ...is interdisciplinary;
-\item ...transcends the system's purpose;
-\item ...applies to both a system and its wider contexts;
-\item ...requires action on multiple levels;
-\item ...requires multiple timescales;
-\item Changing design to take into account long-term effects doesn't automatically imply sacrifices;
-\item System visibility is a precondition for and enabler of sustainability design.
-\end{itemize}
-
-A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including:
-
-\begin{itemize}
-\item Continual Refinement of Product and Project Practices
-\item A Working Product at All Times
-\item A Continual Emphasis on Design
-\item Value Defect Prevention over Defect Detection
-\end{itemize}
-
-This congregated list is an initial collection of principles that could be extended by adding from further related work from separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
-This riased the question, What do the principles mean for scientific and computational software?
-
-The group then identified two example systems to discuss the application of the principles.
-The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
-The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
-The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
 
@@
 \subsection{Key Next Steps}
-The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a guideline on the principles and how they exemplary apply to scientific software system.
-Preliminary work will be carried out through undergraduate or post-graduate student projects.
 
 \subsection{Plan for Future Organization}
+The following plan for future organization was discussed:
+\begin{itemize}
+\item 
+\item Organizing coordinating online calls via Google Hangout.
+\end{itemize}
 
 \subsection{What Else is Needed?}
+The following list of what else is required :
+
+\begin{itemize}
+\item Ethic committee approval required for data collection.
+\item 
+\item 
+\end{itemize}
 
 \subsection{Key Milestones and Responsible Parties}
+The following key milestones were discussed as a roadmap for the set of guidelines on software engineering principles:
+\begin{itemize}
+\item Oct/Nov 2015: Study design and interview guideline
+\item Jan/Feb 2016: Interviews conducted and transcribed
+\item Mar/Apr 2016: Analysis complete
+\item May 2016: Report written
+\end{itemize}
 
 \subsection{Description of Funding Needed}
+Specific funding was not discussed in this working group.
+However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initatives.

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Updates to main sections

@@
 Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group discussed the principles of software engineering design for sustainable software and their application in various domains.
+This group discussed the principles of software engineering design for sustainable software and their application in various domains including quantum chemistry and epidemiology.
 
@@
 \begin{itemize}
-\item 
+\item Identify suitable undergraduate or post-graduate students.
 \item Organizing coordinating online calls via Google Hangout.
@@
 \item Ethic committee approval required for data collection.
-\item 
-\item 
 \end{itemize}

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
Update wg_appendix_eng_design.tex
Minor updates to sub-sections

@@
 \subsection{Key Next Steps}
+In order to achieve (1) the systematic analysis a number of example systems from different scientific domains with regards to the identified principles, (2) the identification of the commonalities and gaps in applying principles to different scientific systems, and (3) the proposal of a set of guidelines on the principles, the following next steps were discussed:
 
@@
 \item Identify suitable undergraduate or post-graduate students.
+\item Design and pilot study.
 \item Organizing coordinating online calls via Google Hangout.
@@
 The following list of what else is required :
-
 \begin{itemize}
-\item Ethic committee approval required for data collection.
+\item Ethics committee review panel approval required for data collection.
 \end{itemize}

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
moving main discussion text to appendix, with short summary in main section adding notes

--- 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex
@@
 \subsection{Summary of Discussion}
-Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group discussed the principles of software engineering design for sustainable software and their application in various domains including quantum chemistry and epidemiology.
+
+The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
+Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion.
+There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community.
+Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
+It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
+The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
+
+Software design principles included: abstraction; coupling and cohesion; decomposition and modularization; encapsulation and information hiding; separation of interface and implementation; sufficiency completeness \& primitiveness; and separation of concerns.
+Similarly, user interface design principles included: learnability; user familiarity; consistency; minimal surprise; recoverability; user guidance; and user diversity.
+The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}.
+The manifesto states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
+A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
+
+This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
+The group identified two example systems to discuss the application of the principles.
+The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
+The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
+The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
+This raised the question, what do the principles mean for  computational scientific and engineering software?
+Similarly, how do the principles relate to non-functional requirements?
+It was suggested that at the very minimum, that sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software:
+\begin{itemize}
+\item Extensibility: the software's ability to be extended and the level of effort required to implement the extension;
+\item Maintainability: the effort required to locate and fix an error in operational software.
+\end{itemize}
+These fundamental building blocks could then be extended to include other quality attributes such as portability, reusability, scalability, usability, and energy efficiency etc.
+Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software.
+In addition, what do the five dimensions of sustainability mean for scientific software, i.e., environmental, economic, social, technical and individual?
 
@@
 \begin{itemize}
-\item Mapping of the principles to best practice.
+\item Mapping of the principles to best practices.
 \item Demonstrating the return on investment of those best practices.
@@
 \subsection{Key Next Steps}
+These was not determined at the workshop. \todo{is this correct, or should something go here?}
 
@@
 \subsection{What Else is Needed?}
-The following list of what else is required :
 
@@
 Specific funding was not discussed in this working group.
-However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initatives.
+However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initiatives.

### 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor update
moving main discussion text to appendix, with short summary in main section adding notes

--- 2016-01-16 WSSSPE3/report/wg_main_eng_design.tex
@@
 
-\subsubsection{Why it is important}
+%\subsubsection{Why it is important}
 
+\todo{fill in cites in this paragraph}
 Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{}.
@@
 
-The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
-Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion.
-There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community.
-Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
-It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
+Software engineering principles form the basis of methods, techniques, methodologies and tools.
+This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}, Tate~\cite{tate2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
+The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements?
+It
+appeared that the sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software: extensibility, the software's ability to be extended and the level of effort required to implement the extension; and
+maintainability: the effort required to locate and fix an error in operational software.
 
-Software design principles included: Abstraction; Coupling and cohesion; Decomposition and modularization; Encapsulation and information hiding; Separation of interface and implementation; Sufficiency completeness \& primitiveness; and Separation of concerns.
-Similalry, user interface design principles included: Learnability; User familiarity; Consistency; Minimal surprise; Recoverability; User guidance; and User diversity.
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}.
-The maifestio states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
-%\todo{should these sets of bullet points be moved to the appendix? or moved into paragraph form, since they are relatively short items?}
-A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
-
-This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
-The group identified two example systems to discuss the application of the principles.
-The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
-The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
-The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
+For more information about the discussion, see Appendix~\ref{sec:appendix_eng_design}.
 
@@
 The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a set of guidelines on the principles and how they exemplary apply to scientific software system.
+Preliminary work will be carried out through undergraduate or post-graduate student projects.
 

### 2016-01-16 WSSSPE3/report/wg_appendix_eng_design.tex

@@
 \subsection{Key Next Steps}
-These was not determined at the workshop. \todo{is this correct, or should something go here?}
+In order to achieve (1) the systematic analysis a number of example systems from different scientific domains with regards to the identified principles, (2) the identification of the commonalities and gaps in applying principles to different scientific systems, and (3) the proposal of a set of guidelines on the principles, the following next steps were discussed:
 
@@
 \subsection{What Else is Needed?}
-
 \begin{itemize}
-\item Ethic committee approval required for data collection.
+\item Ethics committee review panel approval required for data collection.
 \end{itemize}

### 2016-01-18 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Missing References added

@@
 \todo{fill in cites in this paragraph}
-Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{}.
-However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{}.
-Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{}.
+Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{Vliet:2008}.
+However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{Merali:2010,hettrick:2014}.
+Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{Becker:2016}.
 
@@
 Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}, Tate~\cite{tate2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
+This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}, Tate~\cite{tate:2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
 The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements?

### 2016-01-18 WSSSPE3/report/wg_main_eng_design.tex
Update wg_main_eng_design.tex
Minor update

@@
 Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}, Tate~\cite{tate:2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
+This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015}, Tate~\cite{tate:2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
 The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements?

### 2016-01-19 WSSSPE3/report/wg_appendix_eng_design.tex
fixes to new eng design section and appendix

@@
 The manifesto states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
-A number of sustainable software engineering principles proposed by Tate~\cite{tate2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
+A number of sustainable software engineering principles proposed by Tate~\cite{tate:2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
 

### 2016-01-19 WSSSPE3/report/wg_main_eng_design.tex
fixes to new eng design section and appendix

@@
 
-\todo{fill in cites in this paragraph}
 Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{Vliet:2008}.
-However, there is often a mismatch between software engineering theory and practice particulalry in the fields of compuational science and engineering, which can lead to the development of unsustainable software~\cite{Merali:2010,hettrick:2014}.
+However, there is often a mismatch between software engineering theory and practice particularly in the fields of computational science and engineering, which can lead to the development of unsustainable software~\cite{Merali:2010,hettrick:2014}.
 Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{Becker:2016}.

### 2016-01-19 WSSSPE3/report/wg_appendix_metrics.tex
fix latex so document compiles
Small changes and line breaks
Fix typos
Appendix H: Line breaks. One question.
removing question

@@
 \begin{itemize}
-\item Gabrielle Allen -- National Center for Supercomputing Applications
+\item Gabrielle Allen -- University of Illinois at Urbana-Champaign
 \item Emily Chen -- University of Illinois at Urbana-Champaign
@@
 
-The following summary of the group's discussion represents the Useful Metrics for Scientific Software working group's discussion during the WSSSPE3 workshop.
-The group discussion began by agreeing on the common purpose of creating a set of guidance giving examples of specific metrics for the success of scientific software in use, why they were chosen, what they are useful to measure, and any challenges and pitfalls; then publish this as a white paper.
-The group discussed many questions related to useful metrics for scientific software as follows:
-
+The following summary of the group's discussion represents the Useful Metrics
+for Scientific Software working group's discussion during the WSSSPE3 workshop.
+The group discussion began by agreeing on the common purpose of creating a set
+of guidance giving examples of specific metrics for the success of scientific
+software in use, why they were chosen, what they are useful to measure, and any
+challenges and pitfalls; then publish this as a white paper. The group discussed
+many questions related to useful metrics for scientific software as follows:
 \begin{itemize}
@@
 \item
-Can we fit metrics into a common template (i.e. for collection, for description)
+Can we fit metrics into a common template (i.e., for collection, for description)
 
@@
 \item
-        However, there is no commonly used standard (DOAP is a good standard but not widely adopted)
+However, there is no commonly used standard (DOAP is a good standard but not
+widely adopted)
 
@@
 \item
-Number of users could be affected by other metrics e.g. by accessibility
+Number of users could be affected by other metrics e.g., by accessibility
 
@@
 \item
-        Ideal situation is the absolute minimum, so that people can decide on their own what the cost versus usefulness tipping point is
+Ideal situation is the absolute minimum, so that people can decide on their own
+what the cost versus usefulness tipping point is
+
 \end{itemize}
@@
 \item
-        Where should funders place funding for best impact (e.g. funding two-star software versus three-star) and where there are gaps
+        Where should funders place funding for best impact (e.g., funding two-star software versus three-star) and where there are gaps
 
@@
 \item
-Can we use metrics to help people identify the best codes as part of a community effort
+Can we use metrics to help people identify the best codes as part of a community
+effort
 
@@
 \noindent
-Next, a roadmap for how to proceed was discussed including creating a set of milestones and tasks as follows:
-
+Next, a roadmap for how to proceed was discussed including creating a set of
+milestones and tasks as follows:
 \begin{itemize}
@@
 \item
-Go to NSF Software Infrastructure for Sustained Innovation (SI2) projects asking them what metrics they defined, and how useful they were
+Go to NSF Software Infrastructure for Sustained Innovation (SI2) projects asking
+them what metrics they defined, and how useful they were
 
@@
 Ask SI2 PIs to say what metrics they said they would use (copied from proposal)
-
 \item
@@
 \item
-Do something similar for UK SFTF and TRDF software projects to ask them what would be useful metrics to report; also eCSE projects
+Do something similar for UK SFTF and TRDF software projects to ask them what
+would be useful metrics to report; also eCSE projects
 
@@
 \item
-        Give some examples from group members projects, and aim to build out some of the measurement continuum
+Give some examples from group members projects, and aim to build out some of the
+measurement continuum
 
@@
 \noindent
-The idea was put forth for the group to interact with the organizing committee of the 2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI workshop in order to email out a software metrics survey to all SI2 and related awardees as a targeted and relevant set of stakeholders.
-This survey would be created by one of the student group members.
-Similarly, it was suggested that a software metrics survey be sent to the UK SFTF and TRDF software projects to ask them what metrics would be useful to report.
-The remainder of the discussion focused mainly on the creation of a white paper on this topic.
-This resulted in a paper outline and writing assignments with the goal of publishing in venues including WSSSPE4, IEEE CISE, or JORS.
+The idea was put forth for the group to interact with the organizing committee
+of the 2016 NSF Software Infrastructure for Sustained Innovation (SI2) PI
+workshop in order to email out a software metrics survey to all SI2 and related
+awardees as a targeted and relevant set of stakeholders. This survey would be
+created by one of the student group members. Similarly, it was suggested that a
+software metrics survey be sent to the UK SFTF and TRDF software projects to ask
+them what metrics would be useful to report. The remainder of the discussion
+focused mainly on the creation of a white paper on this topic. This resulted in
+a paper outline and writing assignments with the goal of publishing in venues
+including WSSSPE4, IEEE CISE, or JORS.
 
@@
 \item
-Get started on IRB at University of Illinois Urbana-Champaign in anticipation of SI2 project survey (may need more thought into survey)
+Get started on IRB at University of Illinois Urbana-Champaign in anticipation of
+SI2 project survey (may need more thought into survey)
 
@@
 \item
-Our group has created a white paper outline with sections assigned to the above individuals, plus see Section 2 response above for timeline
+Our group has created a white paper outline with sections assigned to the above
+individuals, plus see Section 2 response above for timeline
 
@@
 \item
-Coordination with 2016 NSF SI2 PI workshop organizing committee to possibly piggyback on this event to offer survey to attendees in advance
+Coordination with 2016 NSF SI2 PI workshop organizing committee to possibly
+piggyback on this event to offer survey to attendees in advance
 
 \item
-Coordination (mail communication, info page etc), via WSSSPE github or?
+Coordination (mail communication, info page etc), via WSSSPE GitHub or other?
 
@@
 
-The following items were discussed as a roadmap for the production of a white paper:
-
+The following items were discussed as a roadmap for the production of a white
+paper:
 \begin{enumerate}
 \item
-October -- November 2015: IRB paperwork as appropriate completed (Gabrielle Allen and Emily Chen)
+October -- November 2015: IRB paperwork as appropriate completed (Gabrielle
+Allen and Emily Chen)
 
 \item
-October -- December 2015: Draft white paper sections 1-3 (the paper outline has initial writing assignments)
+October -- December 2015: Draft white paper sections 1-3 (the paper outline has
+initial writing assignments)
 
@@
 Piggyback on planning for 2016 NSF SI2 PIs meeting to be held Feb 16-17, 2016
+
 \end{enumerate}
@@
 
-Funding needs were not discussed in this working group and it was thought that this could potentially be revisited down the road.
+Funding needs were not discussed in this working group and it was thought that
+this could potentially be revisited down the road.

### 2016-01-19 WSSSPE3/report/wg_appendix_industry_interaction.tex
fixed quotation marks

@@
 
-Our initial broad question was, "What makes for successful transitions of scientific software from academia to industry?" There are a number of potential funding transitions that may occur:  
+Our initial broad question was, ``What makes for successful transitions of scientific software from academia to industry?'' There are a number of potential funding transitions that may occur:  
 %

### 2016-01-19 WSSSPE3/report/WSSSPE3_review.tex
added nice TeX\LaTeX symbols

@@
 sustainability Turk mentioned was ``adding of new features,'' or ``maintaining the
-software for a long period of time'' such as the cases of TeX or LaTeX with
+software for a long period of time'' such as the cases of \TeX\ or \LaTeX\ with
 community help. A notion Turk heard often at supercomputing conferences was that

### 2016-01-19 WSSSPE3/report/wg_appendix_best_practices.tex
Update wg_appendix_best_practices.tex
Minor update to group participants
misc symbol corrections

@@
 \item Gregory Tucker -- University of Colorado at Boulder
-\item Birgit Penzens -- California State University Long Beach
+\item Birgit Penzenstadler -- California State University Long Beach
 \item Abigail Cabunoc Mayes -- Mozilla Foundation
 \item Frank L\"{o}ffler -- Louisiana State University 
-\item Colin Venter --  University of Huddersfield
+\item Colin C. Venters --  University of Huddersfield
 \item Lorraine Hwang -- UC Davis 
@@
 
-\item Dimensions of sustainability -- economic, technical, environmental,
+\item Dimensions of sustainability: economic, technical, environmental,
 and obsolescence.

### 2016-01-19 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
misc symbol corrections

@@
 us sit at different points on that spectrum, and move along it during our
-careers (usually from RA to RSE -- examples of movement in the other direction
+careers (usually from RA to RSE---examples of movement in the other direction
 from readers would be welcomed). Either way, the label ``Research Software

### 2016-01-19 WSSSPE3/report/wg_main_eng_design.tex
misc symbol corrections

@@
 \subsubsection{Landing Page}
-In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
-Venters\ to find out more about the group's efforts and how to participate.
+In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.\ Venters\ to find out more about the group's efforts and how to participate.

### 2016-01-19 WSSSPE3/report/wg_main_eng_design.tex
added missing email footnote

--- 2016-01-19 WSSSPE3/report/wg_main_eng_design.tex
@@
 \subsubsection{Landing Page}
-In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
-Venters\ to find out more about the group's efforts and how to participate.
+In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.\ Venters\footnote{email:\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} to find out more about the group's efforts and how to participate.

### 2016-01-20 WSSSPE3/report/wg_appendix_eng_design.tex
use opening quotation marks

@@
 \subsection{Description of Opportunity, Challenges, and Obstacles}
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into ''bite sized'' chunks for the Computational Science and Engineering Community.
+The opportunity was identified to distill existing software engineering and sustainability design knowledge into ``bite sized'' chunks for the Computational Science and Engineering Community.
 In addition, two primary challenges were identified:

### 2016-01-21 WSSSPE3/report/wg_main_best_practices.tex
fixed en-dash to colon
Fix typos.

@@
 \item Catering to disruptive developments in environment (e.g., new hardware, new methodology);
-\item Dimensions of sustainability -- economic, technical, environmental and
+\item Dimensions of sustainability: economic, technical, environmental and
 obsolescence. 
@@
 Some of the tools identified for such analysis are
-the general purpose PeTSC toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.
+the general purpose PETSc toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.
 We also established a timeline and a rough outline (see Appendix~\ref{sec:appendix_best_practices}) for the report.

### 2016-01-22 WSSSPE3/report/wg_appendix_software_credit.tex
Small changes to Software Credit working group
* Use my full name
* clarify description of Mozilla Science's Contributorship Badges for Science & relationship with Project CRediT
* Codemeta -> CodeMeta

@@
 \item Frank L\"{o}ffler -- Louisiana State University
-\item Abby Cabunoc Mayes -- Mozilla Science Lab
+\item Abigail Cabunoc Mayes -- Mozilla Science Lab
 \item Kyle E.\ Niemeyer -- Oregon State University
@@
 Science~\cite{Mozilla_badges}, which provide a badge---associated with
-ORCID~\cite{orcid}---based on author contributions in a similar manner to
-Project CRediT.
+ORCID~\cite{orcid}---for author contributions based on the taxonomy
+outlined in Project CRediT.
 
@@
 the GitHub repository. The group also discussed similar efforts such as
-Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an
+CodeMeta\footnote{CodeMeta: \url{https://github.com/codemeta/codemeta}}, an
 attempt to codify minimal metadata schemes in JSON and XML for scientific
@@
 topic of tenure and professional advancement, 
-%we should recognized that this problem goes beyond faculty to recognizing the role of "Research Software Engineers"
+%we should recognized that this problem goes beyond faculty to recognizing the role of ``Research Software Engineers''
 the group recognized that there is no standard
@@
 thus may not be influential enough to change the status quo.
+However, as discussed in Section~\ref{RSE}, this is changing for Research Software Engineers, at least in the UK.
 

### 2016-01-24 WSSSPE3/report/WSSSPE3_review.tex
Update WSSSPE3_review.tex
Added "University of North Carolina at Chapel Hill" to RENCI entry
p1: Clean up
Fix typos.

@@
 %
-\thanks{{}$^{(7)}$ RENCI, Chapel Hill, NC, USA; rayi@renci.org}
+\thanks{{}$^{(7)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org}
 %
@@
 Software for Science: Practice and Experiences (WSSSPE3). 
-%The workshop used an
-%alternative submission and peer-review process, which led to a set of papers
-%divided across five topic areas: 
 The report includes a description of the keynote presentation and a set of lightning
@@
 can encourage its members to do so.
-%%
-%The workshop recognized that reliance on scientific software
-%is pervasive in all areas of world-leading research today. The workshop
-%participants then proceeded to explore different perspectives on the concept of
-%sustainability. Key enablers and barriers of sustainable scientific software
-%were identified from their experiences. In addition,
-%recommendations with new requirements such as software credit files and software
-%prize frameworks were outlined for improving practices in sustainable software
-%engineering.
-%%
-%There was also broad consensus that formal
-%training in software development or engineering was rare among the
-%practitioners. Significant strides need to be made in building a sense of
-%community via training in software and technical practices, on increasing their
-%size and scope, and on better integrating them directly into graduate education
-%programs.
-%%
-%Finally, journals can define and publish policies to improve reproducibility, whereas
-%reviewers can insist that authors provide sufficient information and access to
-%data and software to allow them reproduce the results in the paper. Hence a list of
-%criteria is compiled for  journals to provide to reviewers so as to make it easier to
-%review software submitted for publication as a ``Software Paper.''
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\katznote{example comment by Dan}
-%
-%\gabnote{example comment by Gabrielle}
-%
-%\nchnote{example comment by Neil}
-%
-%\manishnote{example comment by Manish}
-%
-%\davidnote{example comment by David}
 
@@
 satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
-his wish-list of questions to be answered in the future:
+his wish list of questions to be answered in the future:
 %

### 2016-01-24 WSSSPE3/report/wg_appendix_metrics.tex
minor typo fix in metrics appendix
Fix typos.

@@
 \item
-Number of users could be affected by other metrics e.g., by accessibility
+Number of users could be affected by other metrics including, e.g., accessibility
 
@@
 \item
-No commonly-used standard for collecting or presenting metrics
+No commonly used standard for collecting or presenting metrics
 

### 2016-01-24 WSSSPE3/report/wg_main_eng_design.tex
Fix typos.

@@
 Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015}, Tate~\cite{tate:2005}, and the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
+This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015}, Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
 The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements?

### 2016-01-24 WSSSPE3/report/WSSSPE3_review.tex
Move label outside section command

@@
 
+\choinote{Suggestion: insert somewhere a paragraph, say, about the two
+facilitators who were actively engaging participants with variously tools,
+activities, and reminders}
 \end{abstract}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote \label{sec:keynote}}
+\section{Keynote}\label{sec:keynote}
 %\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
@@
 to Improve Productivity of Legacy Tools}} Abani Patra presented an example of
-how the community using Titan2D, a Geoflow Simulation Software, increased the
+how the community using Titan2D, a geoflow simulation software, increased the
 productivity of their tools by improving both code and data layout. The main
@@
 been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
-and webinars. In order to contribute to the infrastructure, codes must adhere
-to specified minimum standards with the desire that all codes are working toward target standards.
+and webinars. In order to contribute to the infrastructure, codes must adhere to
+specified minimum standards with the desire that all codes are working toward
+target standards.
 %, but ideals are also given to encourage reaching the aforementioned goals. 
-These include, e.g., the use of version control,
-certain coding styles, the presence and nature of code tests and documentation,
-and certain user workflows.
+These include, e.g., the use of version control, certain coding styles, the
+presence and nature of code tests and documentation, and certain user workflows.
 
@@
 cite codes (including non-CIG codes), and only about one eighth specify the
-version used. Compared to other codes, CIG seems to be much better cited. 
-In part, this is attributed to the fact that CIG requires that donated software provide a citable paper specified in the User Manual. 
-The project is working on tools and methods to generate attribution information automatically.
+version used. Compared to other codes, CIG seems to be much better cited. In
+part, this is attributed to the fact that CIG requires that donated software
+provide a citable paper specified in the User Manual. The project is working on
+tools and methods to generate attribution information automatically.
 %This,
@@
 
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham,
-Sergei Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software Platform for
-Predictive Biology: Lessons Learned on the KBase Project}} In this talk, Dan
-Gunter presented experiences and lessons learned as part of KBase, an open
-software and data platform for addressing the grand challenge of systems
+\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
+Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
+Platform for Predictive Biology: Lessons Learned on the KBase Project}} In this
+talk, Dan Gunter presented experiences and lessons learned as part of KBase, an
+open software and data platform for addressing the grand challenge of systems
 biology: predicting and designing biological function. KBase is a unified system
@@
 engineering. This is contrasted to computer science research, where software
-engineering standards shorten design cycles, leading to more time for highly rewarded
-activities like publishing, performance studies, graduating students, or
-protecting ideas before publication. Instead of this more
-traditional approach, KBase uses a variation on the ``Scrum'' methodology.
-After picking projects and team members, four to five teams work on projects for about two
-weeks before a one-week evaluation by the executive committee. Based on their assessment new
-teams and projects might be chosen before iteratively restarting the ``agile'' development cycle.
-This process is intentionally open and documented.
+engineering standards shorten design cycles, leading to more time for highly
+rewarded activities like publishing, performance studies, graduating students,
+or protecting ideas before publication. Instead of this more traditional
+approach, KBase uses a variation on the ``Scrum'' methodology. After picking
+projects and team members, four to five teams work on projects for about two
+weeks before a one-week evaluation by the executive committee. Based on their
+assessment new teams and projects might be chosen before iteratively restarting
+the ``agile'' development cycle. This process is intentionally open and
+documented.
 

### 2016-01-25 WSSSPE3/report/wg_main_funding_spec_expert.tex
Fix typos and minor rewriting.

@@
 
-Research Software Engineers (RSEs) -- those who contribute to science and
-scholarship through software development -- are an important part of the team
+Research Software Engineers (RSEs)---those who contribute to science and
+scholarship through software development---are an important part of the team
 needed to deliver 21st century research. However, existing academic structures
@@
 research domain sufficiently to build meaningful software at a reasonable pace.
-This requires a collaborative approach -- where developers who are fully engaged
+This requires a collaborative approach---where developers who are fully engaged
 or invested in the research context are co-developing software with domain
@@
 software projects at national laboratories or other non-educational
-institutions. This solution is in place and working well for many well-used
+institutions. This solution is in place and working well for many well used
 scientific codebases. However, this strategy has limited application, as much of

### 2016-01-25 WSSSPE3/report/wg_main_metrics.tex
Fix typos and minor rewriting.

@@
 support towards strategic objectives, and making a case for new or continued
-funding. However, there is no commonly-used standard for collecting or
+funding. However, there is no commonly used standard for collecting or
 presenting metrics, nor is it known if there is a common set of metrics for

### 2016-01-25 WSSSPE3/report/wg_main_training.tex
Fix typos and minor rewriting.

@@
 skills taught which might impact on sustainable software are limited at best,
-often lacking a clear and coherent development pathway. This growing array of
-training opportunities could usefully be coordinated by bringing together those
-involved in leading relevant initiatives on a regular basis.
+often lacking a clear and coherent development pathway. Bringing together those
+involved in leading relevant initiatives on a regular basis could helpfully
+coordinate this growing array of training opportunities.
 

### 2016-01-25 WSSSPE3/report/wg_appendix_eng_design.tex
Fix typos.

@@
 It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the SoftWare Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
+The group identified a number of software engineering principles drawn from the Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
 
@@
 \subsection{Description of Funding Needed}
-Specific funding was not discussed in this working group.
-However, this is a open topic that can be be discussed in relation to emerging funding calls from National agencies or grant proposal initiatives.
+Specific funding was not discussed in this working group. However, this is a
+open topic that can be discussed in relation to emerging funding calls from
+National agencies or grant proposal initiatives.

### 2016-01-25 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Fix typos.

@@
 recommend that funders give explicit guidance to grant holders and institutions
-that such arrangement are favorable. Framework agreements permitting this to go
+that such arrangement is favorable. Framework agreements permitting this to go
 ahead without checking back every time with funders and/or grant panels would
@@
 that 40 percent is ideal; two projects per developer, with some time for
-training and infrastructure work. Two developers per project seems to be ideal,
+training and infrastructure work. Having two developers per project seems to be ideal,
 in the sense that software development is enhanced by two pairs of eyes.

### 2016-01-25 WSSSPE3/report/wg_appendix_metrics.tex
Fix typos.

@@
 \item
-Coordination (mail communication, info page etc), via WSSSPE GitHub or other?
+Coordination (mail communication, info page, etc.) via WSSSPE GitHub or other
+means?
 
@@
 \item
-Est. July 2016: By time of next CFP for WSSSPE have complete draft of white paper
+Est.\ July 2016: By the time of next CFP for WSSSPE, to have a complete draft of the white paper
 
 \item
-Est. Sept -- Oct 2016: Responses to white paper submitted to WSSSPE4
+Est.\ Sept -- Oct 2016: Responses to white paper to be submitted to WSSSPE4
 

### 2016-01-25 WSSSPE3/report/wg_appendix_software_credit.tex
Fix typos.

@@
 working with indexers such as Web of Science and Scopus to index software
-citations archived on, e.g., Zenodo or FigShare, and pursuing the development of
+citations archived on, e.g., Zenodo or figshare, and pursuing the development of
 an open indexing service; such plans will be pursued either separately or

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Fix typos and minor rewriting.
Add a TODO note.
Added line breaks. No content change.
Minor changes to abstract.

@@
 The report includes a description of the keynote presentation and a set of lightning
-talks, and focuses on the discussions, future steps, and future organization for
+talks. It also focuses on the discussions, future steps, and future organization for
 a set of self-organized working groups.
-For each group, the report also includes a point of contact and a landing
+For each group, we include a point of contact and a landing
 page that can be used by those who want to join that group's future activities.
-The main challenge left by the workshop is to see if the groups will carry
-out these activities that they have scheduled, and how the WSSSPE community
-can encourage its members to do so.
+The main challenge left by the workshop is to see if the groups will execute 
+these activities that they have scheduled, and how the WSSSPE community
+can encourage its members to follow suit.
 
@@
 \maketitle
-%\newpage
+\newpage
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote}\label{sec:keynote}
+\section{Keynote \label{sec:keynote}}
 %\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 1: Minor changes.

@@
 The WSSSPE3 workshop included multiple mechanisms for participation and
-encouraged team building around solutions. It strongly encouraged participation
-of early-career scientists, postdoctoral researchers, and graduate students,
+encouraged team building around solutions. WSSSPE3 strongly encouraged participation
+of early-career scientists, postdoctoral researchers, graduate students,  
+early-stage researchers, and those from underrepresented groups,
 with funds provided to the conference organizers by the Moore Foundation, the
-National Science Foundation (NSF), and the Software Sustainability Institute, to
+National Science Foundation (NSF), and the Software Sustainability Institute (SSI) to
 support the travel of potential participants who would not otherwise be able to
-attend, early-stage researchers, and those from underrepresented groups. These
-funds allowed 16 additional participants to attend.
+attend the workshop. These
+funds allowed 16 additional participants.
 

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 2: Minor rewriting.

@@
 with the goal of contributing to the discussion of one or more working groups.
-There were~24 lightning talks submitted, and, after a peer-review process, 16 of
+There were~24 lightning talks submitted; after a peer-review process, 16 of
 those were accepted, as discussed further in Section~\ref{sec:lightning}.

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 3: touch up.

@@
 WSSSPE3 left the traditional Supercomputing Conference environment this year,
-and in Turk's words, this spoke to the fact that scientific software comes from
+and in Turk's words, this change spoke to the fact that scientific software comes from
 many different types of inquiries, deployment, strategies for maintenance,
@@
 represent a number of measurable factors such as number of citations; growth of
-a community and number of contributors; amount of funding; prestigious prizes
-awarded; stability of the community in terms of leadership transitions, serving
-community needs, not breaking test suites, and performance on new architectures.
+a community; number of contributors; amount of funding; prestigious prizes
+awarded; stability of the community in terms of leadership transitions; serving
+community needs; not breaking test suites; and performance on new architectures.
 But all these metrics are strictly speaking \emph{proxies} for the values and
@@
 ``sustained'' projects can suffocate new projects, so we need to make sure we
-don't cause novel ideas and packages to suffer at the hands of the status quo.
+do not cause novel ideas and packages to suffer at the hands of the status quo.
 
@@
 \item How can productized software balance its future versus its past, or the
-new needs of the customers versus the existing needs of the development
+emerging needs of the customers versus the existing needs of the development
 community?

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 4: Beautified.

@@
 the field of computing, which is evidenced by the increase in the number of
-events that focus on the topic. Nevertheless, it isn't well understood yet.
+events that focus on the topic. Nevertheless, it is not well understood yet.
 Birgit Penzenstadler argued that we often define sustainability too narrowly.
@@
 through Best Practices in Open Source Software: Computational Infrastructure for
-Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
-Infrastructure for Geodynamics, a community software with a worldwide user base.
+Geodynamics}} Lorraine Hwang presented experiences with the Computational
+Infrastructure for Geodynamics (CIG), a community software with a worldwide user base.
 Like others, their main goals include high usability, sustainability, and
-reproducibility. In order to achieve these goals, communication channels have
+reproducibility. As a means to achieve these goals, various communication channels have
 been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
@@
 in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
-Hwang mentioned various ways to cite efforts within the framework, including
+Hwang mentioned a variety of ways to cite efforts within the framework, including
 science papers, code papers, user manuals, and the CIG website. An analysis of
@@
 located at Notre Dame and Chicago, that links the high energy physics effort to
-biology, astrophysics, digital curation, and other disciplines. It includes
+other disciplines such as biology, astrophysics, and digital curation. It includes
 physicists, digital librarians, as well as computer scientists, and aims to
@@
 needs to be transparent to their workflows, i.e., not requiring additional
-effort to preserve ``research objects''.
+effort to preserve ``research objects.''
 
@@
 Hetherington started by listing some of the problems research software faces,
-which include low levels of reuse and poor standard of verification. For a long
+which include poor standard of verification and  low levels of reuse. For a long
 time, technical solutions to such technical problems were focus of the eResearch
@@
 the possibility of training in reproducible computational research, providing
-collaborations for researchers who don't want to be programmers, and synergies
+collaborations for researchers who do not want to serve as programmers, and creating synergies
 with research computing platforms. The success of such a group would be measured
-by the output and quality of the research software, and could be part of
-Research Computing or faculty. Members would not be independent researchers, but
+by the output and quality of the research software. Members could be part of
+Research Computing or faculty. They would not be independent researchers, but
 would have to have a research background. An attempt to form a community of RSE
@@
 the UK Research Council. An open question is whether this approach can be
-adopted in other countries, including the USA.
+adopted in other countries, including the~USA.
 
@@
 question was, ``What is needed to develop sustainable software?''. Beyond the
-usual suspects of funding, great developers, good design, and software
+usual suspects of funding, proficient developers, good design, and software
 engineering practices, Dan placed the users. He explained this using a
@@
 Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
-Platform for Predictive Biology: Lessons Learned on the KBase Project}} In this
-talk, Dan Gunter presented experiences and lessons learned as part of KBase, an
+Platform for Predictive Biology: Lessons Learned on the KBase Project}} 
+Dan Gunter presented experiences and lessons learned as part of KBase, an
 open software and data platform for addressing the grand challenge of systems
@@
 Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
-approach to overcome some hurdles in scientific publishing: disconnects between
+approach to overcome some hurdles in scientific publishing: disconnects among
 experimental data, research software, and publications. A current effort in the
@@
 includes being publicly available under open licenses, and providing meta-data.
-Last, but not least, it uses digital scholarship elements like persistent
+Last but not least, it uses digital scholarship elements like persistent
 identifiers for software, data, and authors, and it cites both data and
-software. OntoSoft is a tool to help this effort, providing software stewardship
+software. OntoSoft is a tool for helping with this effort, providing software stewardship
 for the Geosciences. As part of this initiative, a special issue of a journal in
@@
 analysis. Problems with this approach include obtaining agreement from the user,
-legal issues, increased maintenance (servers), ethical questions, and also the
+legal issues, increased maintenance (of servers), ethical questions, and also the
 lack of a common infrastructure. Some of these problems are of a social nature
@@
 Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
-Developing Community}} The effort behind this talk by Ray Idaszak started from a
-break-out group at an NSF SI2 workshop in 2015, and centers around creating a
-framework for creation of metrics for scientific software. This framework would
+Developing Community}} The effort behind this talk given by Ray Idaszak started from a
+breakout group at an NSF SI2 workshop in 2015, and centers around building a
+framework for creating metrics for scientific software. This framework would
 improve both the metrics and the software it evaluates, and could also serve as
-a tool for building a community around the idea. With especially the latter idea
+a tool for building a community around the idea. With especially the last-mentioned idea
 (building a community) in mind, a software ``peer review group'' would be

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 4: Beautified.

@@
 compilation flags, and that the internal data layout was not suitable for modern
-technologies (vectorization, accelerators). The approach of the Titan2D
+technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
 developers included reinforcing the code structure using multiple layers of

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
1 change: didn't -> did not
No content change. Just wrap up long lines.
No content change. Just link up some short lines.
First page: non-breaking spaces for middle initials and updated author order.

@@
 
-\author{Daniel S. Katz$^{(1)}$,
-Kyle E.\ Niemeyer$^{(2)}$,
-Sou-Cheng T. Choi$^{(3)}$,
+\author{Daniel S.\ Katz$^{(1)}$,
+Sou-Cheng T.\ Choi$^{(2)}$,
+Kyle E.\ Niemeyer$^{(3)}$,
 James Hetherington$^{(4)}$,
@@
 Ray Idaszak$^{(7)}$,
-Steven R. Brandt$^{(5)}$,
-Mark A. Miller$^{(8)}$,
+Steven R.\ Brandt$^{(5)}$,
+Mark A.\ Miller$^{(8)}$,
 Sandra Gesing$^{(9)}$,
-Nick D. Jones$^{(10)}$,
+Nick D.\ Jones$^{(10)}$,
 Nic Weber$^{(11)}$,
@@
 Birgit~Penzenstadler$^{(14)}$,
-Colin C. Venters$^{(15)}$,
+Colin C.\ Venters$^{(15)}$,
 Ethan Davis$^{(16)}$,
@@
 %
-\thanks{{}$^{(1)}$ Computation Institute, 
+\thanks{{}$^{(1)}$ \hspace{-1ex}Computation Institute, 
 University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; d.katz@ieee.org}
 %
-\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
+%
+\thanks{{}$^{(3)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
 Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
-%
-\thanks{{}$^{(3)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
 %
@@
 %
-\thanks{{}$^{(5)}$ Center for Computation \& Technology, Louisiana State University, Baton Rouge, LA, USA; \{knarf,sbrandt\}@cct.lsu.edu}
+\thanks{{}$^{(5)}$ \hspace{-1.5ex}Center for Computation \& Technology, Louisiana State University, Baton Rouge, LA, USA; \{knarf,sbrandt\}@cct.lsu.edu}
 %
@@
 an imagined future for software development and usage in science and
-engineering. But WSSSPE2 didn't have a good way to enact those mechanisms, or to
+engineering. But WSSSPE2 did not have a good way to enact those mechanisms, or to
 encourage the attendees to follow through on their intentions.
@@
 WSSSPE3 left the traditional Supercomputing Conference environment this year,
-and in Turk's words, this change spoke to the fact that scientific software comes from
-many different types of inquiries, deployment, strategies for maintenance,
-users, and ways of measuring the value of a piece of software. It appeared to
-Turk that the supercomputing community generally adopts some top-down
-approaches, whereas the SciPy community more often than not uses more bottom-up
-systems. The essential messages perceived were also often bipolar: the
+and in Turk's words, this change spoke to the fact that scientific software
+comes from many different types of inquiries, deployment, strategies for
+maintenance, users, and ways of measuring the value of a piece of software. It
+appeared to Turk that the supercomputing community generally adopts some
+top-down approaches, whereas the SciPy community more often than not uses more
+bottom-up systems. The essential messages perceived were also often bipolar: the
 supercomputing community thinks that software is getting harder, with exascale
@@
 were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was ``adding of new features,'' or ``maintaining the
-software for a long period of time'' such as the cases of \TeX\ or \LaTeX\ with
-community help. A notion Turk heard often at supercomputing conferences was that
-sustainable software ``continues to work on new architectures.'' Yet another
-metric was ``people continuing to be able to learn how to use and apply the
-software.'' A funder Turk heard talked about sustainability as ``continuing to get
-funded.'' Turk also recalled that Greg Wilson, among others, said in WSSSPE1.1
-that his view of sustainable software was software that ``continued to give the
-same results over time.'' A last measure of sustainability Turk presented was
-``the ability to transition between different people developing and using a piece
-of software.''
+sustainability Turk mentioned was ``adding of new features,'' or ``maintaining
+the software for a long period of time'' such as the cases of \TeX\ or \LaTeX\
+with community help. A notion Turk heard often at supercomputing conferences was
+that sustainable software ``continues to work on new architectures.'' Yet
+another metric was ``people continuing to be able to learn how to use and apply
+the software.'' A funder Turk heard talked about sustainability as ``continuing
+to get funded.'' Turk also recalled that Greg Wilson, among others, said in
+WSSSPE1.1 that his view of sustainable software was software that ``continued to
+give the same results over time.'' A last measure of sustainability Turk
+presented was ``the ability to transition between different people developing
+and using a piece of software.''
 
@@
 Geodynamics}} Lorraine Hwang presented experiences with the Computational
-Infrastructure for Geodynamics (CIG), a community software with a worldwide user base.
-Like others, their main goals include high usability, sustainability, and
-reproducibility. As a means to achieve these goals, various communication channels have
-been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
-and webinars. In order to contribute to the infrastructure, codes must adhere to
-specified minimum standards with the desire that all codes are working toward
-target standards.
+Infrastructure for Geodynamics (CIG), a community software with a worldwide user
+base. Like others, their main goals include high usability, sustainability, and
+reproducibility. As a means to achieve these goals, various communication
+channels have been developed, such as mailing lists, wikis, workshops,
+hackathons, tutorials, and webinars. In order to contribute to the
+infrastructure, codes must adhere to specified minimum standards with the desire
+that all codes are working toward target standards.
 %, but ideals are also given to encourage reaching the aforementioned goals. 
@@
 the possibility of training in reproducible computational research, providing
-collaborations for researchers who do not want to serve as programmers, and creating synergies
-with research computing platforms. The success of such a group would be measured
-by the output and quality of the research software. Members could be part of
-Research Computing or faculty. They would not be independent researchers, but
-would have to have a research background. An attempt to form a community of RSE
-groups within the UK has been underway for several years, including funding from
-the UK Research Council. An open question is whether this approach can be
-adopted in other countries, including the~USA.
+collaborations for researchers who do not want to serve as programmers, and
+creating synergies with research computing platforms. The success of such a
+group would be measured by the output and quality of the research software.
+Members could be part of Research Computing or faculty. They would not be
+independent researchers, but would have to have a research background. An
+attempt to form a community of RSE groups within the UK has been underway for
+several years, including funding from the UK Research Council. An open question
+is whether this approach can be adopted in other countries, including the~USA.
 
@@
 Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
-Developing Community}} The effort behind this talk given by Ray Idaszak started from a
-breakout group at an NSF SI2 workshop in 2015, and centers around building a
-framework for creating metrics for scientific software. This framework would
-improve both the metrics and the software it evaluates, and could also serve as
-a tool for building a community around the idea. With especially the last-mentioned idea
-(building a community) in mind, a software ``peer review group'' would be
-created, representing stakeholders who will self-review software created by
-their respective communities, and will concurrently develop metrics. The whole
-project should be community-governed, without a single institution overseeing
-the activities or infrastructure, with the hope to evolving community-generated
-and adopted standards. The generation of metrics would be tied to the actual
-evaluation of software, creating an incentive by improving the evaluated
-software itself during this process. The framework code would provide
-infrastructure for the creation of metrics and evaluation, and forums for
-generation of software success metrics. It would also support code reviews of
-the evaluated software. An open question is whether it is possible to fit the
-resulting metrics in a common template. So far, this is still in a design phase,
-with a white paper at the CSESSP workshop 2015 and this talk, but the WSSSPE
-workshops are seen as a forum for the community to assemble and act, and is
-planned to be used also in the future to build this community and framework.
+Developing Community}} The effort behind this talk given by Ray Idaszak started
+from a breakout group at an NSF SI2 workshop in 2015, and centers around
+building a framework for creating metrics for scientific software. This
+framework would improve both the metrics and the software it evaluates, and
+could also serve as a tool for building a community around the idea. With
+especially the last-mentioned idea (building a community) in mind, a software
+``peer review group'' would be created, representing stakeholders who will
+self-review software created by their respective communities, and will
+concurrently develop metrics. The whole project should be community-governed,
+without a single institution overseeing the activities or infrastructure, with
+the hope to evolving community-generated and adopted standards. The generation
+of metrics would be tied to the actual evaluation of software, creating an
+incentive by improving the evaluated software itself during this process. The
+framework code would provide infrastructure for the creation of metrics and
+evaluation, and forums for generation of software success metrics. It would also
+support code reviews of the evaluated software. An open question is whether it
+is possible to fit the resulting metrics in a common template. So far, this is
+still in a design phase, with a white paper at the CSESSP workshop 2015 and this
+talk, but the WSSSPE workshops are seen as a forum for the community to assemble
+and act, and is planned to be used also in the future to build this community
+and framework.
 
@@
 
-The results of these group sessions made it clear that
-there are many interlinked challenges in sustainable software, and that
-while these challenges can be addressed, doing so is difficult because they
-generally are not the full-time job of any of the attendees. 
-As was the case in WSSSPE2 as well, the participants were willing
-to dedicate their time to the groups while they were at the meeting, but
-afterwards, they went back to their (paid) jobs.
+The results of these group sessions made it clear that there are many
+interlinked challenges in sustainable software, and that while these challenges
+can be addressed, doing so is difficult because they generally are not the
+full-time job of any of the attendees. As was the case in WSSSPE2 as well, the
+participants were willing to dedicate their time to the groups while they were
+at the meeting, but afterwards, they went back to their (paid) jobs.
 
-We need to determine how
-to tie the WSSSPE breakout activities to people's jobs, so that they feel that
-continuing them is a higher priority than it is now, perhaps through funding the
-participants, or through funding coordinators for each activity, or perhaps by
-getting the workshop participants to agree to a specific schedule of activities
-during the workshop as we have tried to do in WSSSPE3.  It remains to be
-seen, however, if the participants will meet the schedules they set.
+We need to determine how to tie the WSSSPE breakout activities to people's jobs,
+so that they feel that continuing them is a higher priority than it is now,
+perhaps through funding the participants, or through funding coordinators for
+each activity, or perhaps by getting the workshop participants to agree to a
+specific schedule of activities during the workshop as we have tried to do in
+WSSSPE3. It remains to be seen, however, if the participants will meet the
+schedules they set.
 
-The overall challenge
-left to the sustainable software community is perhaps one of organization: how
-to combine the small partial efforts of a large number of people to impact a much
-larger number of people: those who develop and use scientific software.  While
-WSSSPE might help focus the actions of the groups, something more is needed
-to incentivize the wider community, which is a generalization of the sustainable
-software problem itself.
+The overall challenge left to the sustainable software community is perhaps one
+of organization: how to combine the small partial efforts of a large number of
+people to impact a much larger number of people: those who develop and use
+scientific software. While WSSSPE might help focus the actions of the groups,
+something more is needed to incentivize the wider community, which is a
+generalization of the sustainable software problem itself.
 

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
Acknowledgements: touched up.

@@
 we can collaboratively build a workshop agenda and turn that into an ongoing
-community activity. The success or failure of this effort will only
-become apparent over time.
+community activity. The success or failure of this effort will only become
+apparent over time.
 
@@
 their attendance; and second, as a way to get new ideas across to all the
-attendees. We broke with the tradition of requiring the lightning talk submitters
-to self-publish their papers, and instead used a common peer-review
-platform\footnote{\url{http://easychair.org}}, choosing to publish their
-slides on the workshop website instead.
+attendees. We broke with the tradition of requiring the lightning talk
+submitters to self-publish their papers, and instead used a common peer-review
+platform\footnote{\url{http://easychair.org}}, choosing to publish their slides
+on the workshop website instead.
 
@@
 second day, with the exception of some short periods for the groups to report
-back to the collected workshop attendees. Each group determined a set
-of activities that the members could do to advance sustainable software in
-a particular area.
+back to the collected workshop attendees. Each group determined a set of
+activities that the members could do to advance sustainable software in a
+particular area.
 
@@
 expressed in this material are those of the author(s) and do not necessarily
-reflect the views of the National Science Foundation. Choi's work is supported
-in part by the National Science Foundation research grant DMS-1522687 and
-a WSSSPE3 travel award. Hetherington was funded by
-the Software Sustainability Institute, RCUK grants EP/H043160/1 and EP/N006410/1.
+reflect the views of the National Science Foundation.
+%
+Choi's work was supported in part by the National Science Foundation research
+grant DMS-1522687 and a WSSSPE3 travel award. She thanks the encouragement and
+discussion with Fred Hickernell.
+%
+Hetherington was funded by the Software Sustainability Institute, RCUK grants
+EP/H043160/1 and EP/N006410/1.
 %
 Work by Gunter was supported by the Office of Science, Office of Biological and
-Environmental Research, of the US Department of Energy under Award Numbers
-DE-AC02-05CH11231, DE-AC02-06CH11357, DE-AC05-00OR22725, and DE-AC02-98CH10886,
-as part of the DOE Systems Biology Knowledgebase and by the Office of Science,
-Office of Advanced Scientific Computing Research (ASCR) of the U.S. Department
-of Energy under Contract Number DE-AC02-05CH11231 as part of the Template
-Interfaces for Agile Parallel Data-Intensive Science (TIGRES) project.
+Environmental Research, of the U.S.\ Department of Energy (DOE) under Award
+Numbers DE-AC02-05CH11231, DE-AC02-06CH11357, DE-AC05-00OR22725, and
+DE-AC02-98CH10886, as part of the DOE Systems Biology Knowledgebase and by the
+Office of Science, Office of Advanced Scientific Computing Research (ASCR) of
+the U.S.\ Department of Energy under Contract Number DE-AC02-05CH11231 as part
+of the Template Interfaces for Agile Parallel Data-Intensive Science (TIGRES)
+project.
 %
-\todo{feel free to add stuff here}
+%\todo{feel free to add stuff here}
 
@@
 %\todo{Do we want email addresses here?}
+
 The following is the list of organizers of WSSSPE3.

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
Conclusions: Adding a few sentences for the TODO note about Tim and Costa.

@@
 can encourage its members to follow suit.
-
-\choinote{Suggestion: insert somewhere a paragraph, say, about the two
-facilitators who were actively engaging participants with variously tools,
-activities, and reminders}
 \end{abstract}
@@
 we can collaboratively build a workshop agenda and turn that into an ongoing
-community activity. The success or failure of this effort will only become
-apparent over time.
+community activity. For the first time in WSSSPE meetings, two professional
+event organizers and facilitators were engaged using funding available to the
+WSSSPE3 organizing committe. They assisted the organizing committee members
+planning the workshop agenda during the months leading up to the conference.
+Throughout the workshop, they actively engaged participants with variously
+tools, activities, and reminders. The success or failure of these efforts will
+only become apparent over time.
 

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
Conclusions: Fix a typo.

@@
 event organizers and facilitators were engaged using funding available to the
-WSSSPE3 organizing committe. They assisted the organizing committee members
+WSSSPE3 organizing committee. They assisted the organizing committee members
 planning the workshop agenda during the months leading up to the conference.

### 2016-02-03 WSSSPE3/report/wg_main_best_practices.tex
Section 5.1: Add line breaks. Add \medskip for better spacing.

@@
 %\subsubsection{Why it is important}
-Reviewing multiple past articles and talks at different meetings like WSSSPEx \cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} 
-and 
-analyzing and promoting sustainable scientific software makes it clear that there 
-are several common and recurring ideas that underpin success in developing sustainable software. However, outside of a small community, this knowledge is not widely
-shared. This is  especially true for the large community of scientists who generate most of the software used by scientists but are not primarily software developers. In this
-scenario, a clear and precise exposition of these best practices collected from many sources and open collaboration among all in the community
-in a single source (e.g., journal paper, tutorial)  that can be widely disseminated is necessary and likely to be very valuable.
+Reviewing multiple past articles and talks at different meetings like WSSSPEx
+\cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} and analyzing and
+promoting sustainable scientific software makes it clear that there are several
+common and recurring ideas that underpin success in developing sustainable
+software. However, outside of a small community, this knowledge is not widely
+shared. This is especially true for the large community of scientists who
+generate most of the software used by scientists but are not primarily software
+developers. In this scenario, a clear and precise exposition of these best
+practices collected from many sources and open collaboration among all in the
+community in a single source (e.g., journal paper, tutorial) that can be widely
+disseminated is necessary and likely to be very valuable.
 
@@
 
-The creation of such a ``best practices'' document will build upon the range of activities and topics discussed at WSSSPE3 and associated prior meetings. We will attempt to
-distill the emerging body of knowledge into this document.
-The large number of  articles from the NSF funded SI2 projects (SSE and SSI), ``lightning talks'', ``white papers,'' and reports from different workshops have created a large if somewhat diffuse source for this report.
+The creation of such a ``best practices'' document will build upon the range of
+activities and topics discussed at WSSSPE3 and associated prior meetings. We
+will attempt to distill the emerging body of knowledge into this document. The
+large number of articles from the NSF funded SI2 projects (SSE and SSI),
+``lightning talks'', ``white papers,'' and reports from different workshops have
+created a large if somewhat diffuse source for this report.
 
@@
 
-Core questions that will need to be explored are in reliability (reproducibility), usability, extensibility, knowledge management and continuity (transitions between people).
-Answers to these will guide us on how a software tool becomes part of the core workflow of well identified users (stakeholders) relating to tool success and hence sustainability.
+Core questions that will need to be explored are in reliability
+(reproducibility), usability, extensibility, knowledge management and continuity
+(transitions between people). Answers to these will guide us on how a software
+tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability.
 %\katznote{prev sentence is complex and awkward} % see fix
@@
 \item When should software ``die''?
-\item Catering to disruptive developments in environment (e.g., new hardware, new methodology);
+\item Catering to disruptive developments in environment (e.g., new hardware,
+new methodology);
 \item Dimensions of sustainability: economic, technical, environmental and
@@
 
-Sustainability requires community participation in code development and/or a wide adoption of software.
-The larger the community base is using a piece of software, the better are the funding possibilities and thus also the sustainability options.
-Additionally developer commitment to an application is essential and experience shows that software packages with an evangelist imposing strong inspiration and discipline are more likely to achieve sustainability.
-While a single person can push sustainability to a certain level, open source software also needs sustained commitment from the developer community.
-Such sustained commitments include diverse tasks and roles, which can be fulfilled by diverse developers with different knowledge levels.
-Besides developing software and appropriate software management with measures for extensibility and scalability of the software, active (expertise) support for users via a user forum with a quick turnaround is crucial.
-The barrier to entry for the community as users as well as developers has to be as low as possible.
+Sustainability requires community participation in code development and/or a
+wide adoption of software. The larger the community base is using a piece of
+software, the better are the funding possibilities and thus also the
+sustainability options. Additionally developer commitment to an application is
+essential and experience shows that software packages with an evangelist
+imposing strong inspiration and discipline are more likely to achieve
+sustainability. While a single person can push sustainability to a certain
+level, open source software also needs sustained commitment from the developer
+community. Such sustained commitments include diverse tasks and roles, which can
+be fulfilled by diverse developers with different knowledge levels. Besides
+developing software and appropriate software management with measures for
+extensibility and scalability of the software, active (expertise) support for
+users via a user forum with a quick turnaround is crucial. The barrier to entry
+for the community as users as well as developers has to be as low as possible.
 
-For additional information about the discussion, see Appendix~\ref{sec:appendix_best_practices}.
+For additional information about the discussion, see
+Appendix~\ref{sec:appendix_best_practices}.
 
@@
 %\todo{short text here - not bullets}
-The creation of a best-practices document needs a large and diverse community involved. We have enlisted over ten contributors from the attendees at the WSSSPE3 and 
-those on the mailing list.
-The primary mechanism for developing this document will be to examine and analyze the success of several well known community scientific
-software and organizations supporting scientific software.
-We will attempt then to abstract general principles and best practices.
-Some of the tools identified for such analysis are
-the general purpose PETSc toolkit for linear system solution, NWChem for computational chemistry and the CIG (Computational Infrastructure for Geodynamics) organization dedicated to supporting an ensemble of related tools for the geodynamics community.
-We also established a timeline and a rough outline (see Appendix~\ref{sec:appendix_best_practices}) for the report.
+The creation of a best-practices document needs a large and diverse community
+involved. We have enlisted over ten contributors from the attendees at the
+WSSSPE3 and those on the mailing list. The primary mechanism for developing this
+document will be to examine and analyze the success of several well known
+community scientific software and organizations supporting scientific software.
+We will attempt then to abstract general principles and best practices. Some of
+the tools identified for such analysis are the general purpose PETSc toolkit for
+linear system solution, NWChem for computational chemistry and the CIG
+(Computational Infrastructure for Geodynamics) organization dedicated to
+supporting an ensemble of related tools for the geodynamics community. We also
+established a timeline and a rough outline (see
+Appendix~\ref{sec:appendix_best_practices}) for the report.
 
+\medskip
 \noindent{\bf Timeline:}

### 2016-02-03 WSSSPE3/report/wg_appendix_best_practices.tex
Fix typos.
Section 5.1 and Appendix D: touched up.

@@
 
-Sandra Gesing\footnote{email: \href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}} will serve as
-the point of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+Sandra Gesing\footnote{email:
+\href{mailto:sandra.gesing@nd.edu}{sandra.gesing@nd.edu}} will serve as the
+point of contact for this working group, and be responsible for ensuring timely
+progress of the planned actions.
 
@@
 
-Core questions that will need to be explored are in reliability (reproducibility), usability, extensibility, knowledge management and continuity (transitions between people).
-Answers to these will guide us on how a software tool becomes part of the core workflow of well identified users (stakeholders) relating to tool success and hence sustainability.
+Core questions that will need to be explored are in reliability,
+reproducibility, usability, extensibility, knowledge management, and continuity
+(transitions between people). Answers to these will guide us on how a software
+tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability.
 %\katznote{prev sentence is complex and awkward} % see fix
@@
 
-The opportunity lies in collaboration on a white paper, which will be revisited
+There is an opportunity to collaborate on a white paper, which will be revisited
 regularly for further improvements, to enhance knowledge of the state of best
@@
 
+\medskip
 \noindent{\bf White Paper Outline:}
@@
 \begin{enumerate} 
-\item PeTSC
+\item PETSc
 \item NWChem

### 2016-02-03 WSSSPE3/report/wg_main_best_practices.tex

@@
 
-Core questions that will need to be explored are in reliability
-(reproducibility), usability, extensibility, knowledge management and continuity
-(transitions between people). Answers to these will guide us on how a software
+Core questions that will need to be explored are in reproducibility, 
+reliability,  usability, extensibility, knowledge management, as well as continuity
+(transitions between people). Answers to these questions will guide us on how a software
 tool becomes part of the core workflow of well identified users (stakeholders)
@@
 %\todo{short text here - not bullets}
-The creation of a best-practices document needs a large and diverse community
+The creation of a document on best practices needs a large and diverse community
 involved. We have enlisted over ten contributors from the attendees at the

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
Section 2: More natural spacing without quotes.
editing changes from last 2 weeks - just starting
slight update to abstrcact

@@
 these activities that they have scheduled, and how the WSSSPE community
-can encourage its members to follow suit.
+can encourage this to happen.
 \end{abstract}
@@
 subsequent calls for participation:
-
-\begin{quote}
+%
+%\begin{quote}
 \begin{itemize}
@@
 \end{itemize}
-\end{quote}
+%\end{quote}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote \label{sec:keynote}}
+\section{Keynote}\label{sec:keynote}
 %\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
@@
 satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
-his wish list of questions to be answered in the future:
+his wishlist of questions to be answered in the future:
 %

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
minor change to changes in S1
changes in 2 and 3

@@
 attend the workshop. These
-funds allowed 16 additional participants.
+funds allowed 16 additional people to attend and participate.
 
@@
 with the goal of contributing to the discussion of one or more working groups.
-There were~24 lightning talks submitted; after a peer-review process, 16 of
-those were accepted, as discussed further in Section~\ref{sec:lightning}.
+There were~24 lightning talks submitted; after a peer-review process, 16 
+were accepted, as discussed further in Section~\ref{sec:lightning}.
 
@@
 represent a number of measurable factors such as number of citations; growth of
-a community; number of contributors; amount of funding; prestigious prizes
-awarded; stability of the community in terms of leadership transitions; serving
-community needs; not breaking test suites; and performance on new architectures.
+a community and number of contributors; amount of funding; prestigious prizes
+awarded; stability of the community in terms of leadership transitions, serving
+community needs, not breaking test suites, and performance on new architectures.
 But all these metrics are strictly speaking \emph{proxies} for the values and
@@
 ``sustained'' projects can suffocate new projects, so we need to make sure we
-do not cause novel ideas and packages to suffer at the hands of the status quo.
+don't cause novel ideas and packages to suffer at the hands of the status quo.
 

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
moving authors back to original order

@@
 \author{Daniel S.\ Katz$^{(1)}$,
-Sou-Cheng T.\ Choi$^{(2)}$,
-Kyle E.\ Niemeyer$^{(3)}$,
+Kyle E.\ Niemeyer$^{(2)}$,
+Sou-Cheng T.\ Choi$^{(3)}$,
 James Hetherington$^{(4)}$,
@@
 %
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
+\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
+Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
 %
-\thanks{{}$^{(3)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
-Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
+\thanks{{}$^{(3)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
 %

### 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
moving note about facilitators to intro

@@
 funds allowed 16 additional people to attend and participate.
+
+WSSSPE3 also included two professional
+event organizers/facilitators from Knowinnovation who helped the organizing committee members
+plan the workshop agenda, and during the workshop, they actively engaged
+participants with variously
+tools, activities, and reminders.
 
@@
 we can collaboratively build a workshop agenda and turn that into an ongoing
-community activity. For the first time in WSSSPE meetings, two professional
-event organizers and facilitators were engaged using funding available to the
-WSSSPE3 organizing committee. They assisted the organizing committee members
-planning the workshop agenda during the months leading up to the conference.
-Throughout the workshop, they actively engaged participants with variously
-tools, activities, and reminders. The success or failure of these efforts will
+community activity. The success or failure of these efforts will
 only become apparent over time.

### 2016-02-05 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
Added line breaks. No content change.
more edits...
Section 5.2 and Appendix E: minor changes.

@@
 
-``Are you an RSE or a RA?'' is not properly a binary question. Most of
-us sit at different points on that spectrum, and move along it during our
-careers (usually from RA to RSE---examples of movement in the other direction
-from readers would be welcomed). Either way, the label ``Research Software
-Engineer'' is now starting to have some power. Many scientists do not want to be
-writing code; some do, to varying degrees. These groups can usefully support
-each other.
+``Are you an RSE or a RA?'' is not properly a binary question. Most of us sit at
+different points on that spectrum, and move along it during our careers (usually
+from RA to RSE---examples of movement in the other direction from readers would
+be welcomed). Either way, the label ``Research Software Engineer'' is now
+starting to have some power. Many scientists do not want to be writing code;
+some do, to varying degrees. These groups can usefully support each other.
 
-What is the power of the label? How can we get the word out about RSE support using the label?
+What is the power of the label? How can we get the word out about RSE support
+using the label?
 
@@
 case that the new generation of computer and software savvy scientists will be
-so comfortable in developing their own code that the professional developers will
-not be needed? And this brings up the flip side question, ``Do scientists really
-want to be writing code?''
+so comfortable in developing their own code that the professional developers
+will not be needed? And this brings up the flip side question, ``Do scientists
+really want to be writing code?''
 
-We also had a little discussion about how to make a career path for
-research developers. It need not be solely an academic enterprise, but in the
-past tenure has often been problematic for people of this class.
+We also had a little discussion about how to make a career path for research
+developers. It need not be solely an academic enterprise, but in the past tenure
+has often been problematic for people of this class.
 
 Skills and resources may vary between teams. To help resolve this, maintaining
-high levels of communication between groups will be valuable. In the United Kingdom (UK), there
-are plans to permit resource sharing between institutional RSE groups. Perhaps
-there are circumstances under which an RSE skill exchange could be arranged,
-either formally or informally.
+high levels of communication between groups will be valuable. In the United
+Kingdom (UK), there are plans to permit resource sharing between institutional
+RSE groups. Perhaps there are circumstances under which an RSE skill exchange
+could be arranged, either formally or informally.
 
-Collaborative funding can be important for RSE groups, to ensure that research
-leadership remains with the domain scientists. At NCAR, university partnerships
+Collaborative funding can be crucial to RSE groups, to ensure that research
+leadership remains with the domain scientists. As an example, at NCAR, university partnerships
 are required for submission of proposals, so collaboration is an essential part
@@
 goal is to make the available stack more robust and richer for everyone. They
-offer short cycles of funding for creating some useful
-functionality that shows a diversity of input and draws from across disciplines as a key metric,  If this metric is met successfully, 
-then more funding may follow. This can apply within or across institutions.
+offer short cycles of funding for creating some useful functionality that shows
+a diversity of input and draws from across disciplines as a key metric, If this
+metric is met successfully, then more funding may follow. This can apply within
+or across institutions.
 
@@
 groups will become established. The Sloan Foundation is currently funding data
-science engineers, who work in the context of other software developers at the University
-of Washington. These scientists work in the e-Science Studio/Data Science
-Studio, and they help a group of graduate students in solving their problems in
-data science and data management. During Fall and Spring, a 10-week incubator
-program allows students to work two days a week on a data-intensive
+science engineers, who work in the context of other software developers at the
+University of Washington. These scientists work in the e-Science Studio/Data
+Science Studio, and they help a group of graduate students in solving their
+problems in data science and data management. During Fall and Spring, a 10-week
+incubator program allows students to work two days a week on a data-intensive
 science project. Some fraction of the developer time is dedicated to the
@@
 recommend that funders give explicit guidance to grant holders and institutions
-that such arrangement is favorable. Framework agreements permitting this to go
+that such an arrangement is favorable. Framework agreements permitting this to go
 ahead without checking back every time with funders and/or grant panels would
@@
 
-There is some question about the most effective duration and percent of full
+There is some question about the most effective duration and percentage of full
 time for a programmer's work on a project. At least three months is necessary
@@
 
-In the UK, another source of funding for research software has been the
+In the UK, another source of funding for research software is the
 Collaborative Computational Projects (CCPs): domain specific communities put
@@
 on short-term papers. This can cause infrastructure developers to focus more on
-publications in their areas of mathematics and computer science, the domain developers on the
-shorter-term needs of their own fields. Genuine collaborative co-construction is
-harder to foster.
+publications in their areas of mathematics and computer science, the domain
+developers on the shorter-term needs of their own fields. Genuine collaborative
+co-construction is harder to foster.
 
-It can be a more difficult to help leading domain scientists see the value of
+It can be more difficult to help leading domain scientists see the value of
 engineering effort than those in their teams who are forced to work with
@@
 
-RSEs provide a useful contribution to their universities' teaching
-missions, as well as research, as they are well placed to deliver the
-research programming training that many scientists now need. In the longer term,
-with programming skills taught to all through their careers, we hope specialist
-scientific developers will be less needed.
+RSEs provide a useful contribution to their universities' teaching missions, as
+well as research, as they are well placed to deliver the research programming
+training that many scientists now need. In the longer term, with programming
+skills taught to all through their careers, we hope specialist scientific
+developers will be less needed.
 
@@
 
-Financial support for an initial conference to bring together research software
+Financial support for an initial conference that brings together research software
 groups to form an organization and create a resource sharing structure would

### 2016-02-05 WSSSPE3/report/wg_main_funding_spec_expert.tex
more review of changes
Section 5.2 and Appendix E: minor changes.

@@
 
-While development of new mechanisms in the current academic reward system is a
+While developing new mechanisms in the current academic reward system is a
 worthy aspirational goal, such a dramatic change in this structure does not seem
@@
 software projects at national laboratories or other non-educational
-institutions. This solution is in place and working well for many well used
+institutions. This solution is in place and working well for many well-used
 scientific codebases. However, this strategy has limited application, as much of

### 2016-02-05 WSSSPE3/report/wg_appendix_industry_interaction.tex
Section 5.3 and Appendix F:  spacing. No content change.

@@
 
-Nic Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} will serve as the point of contact for this working group.
+Nic Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} will
+serve as the point of contact for this working group.
 
@@
 
-Our initial broad question was, ``What makes for successful transitions of scientific software from academia to industry?'' There are a number of potential funding transitions that may occur:  
+Our initial broad question was, ``What makes for successful transitions of
+scientific software from academia to industry?'' There are a number of potential
+funding transitions that may occur:
 %
@@
 whereby ownership, maintenance, and stewardship of the software become similar
-to peer-production models in open-source (e.g., see Howison~\cite{howison_sustaining_2015}).
+to peer-production models in open-source (e.g., see
+Howison~\cite{howison_sustaining_2015}).
 
@@
 
-\item The project could gain attention from a industry use case who would potentially make in-kind
-contributions by having paid staff contribute to the software.
+\item The project could gain attention from a industry use case who would
+potentially make in-kind contributions by having paid staff contribute to the
+software.
 
@@
 \end{itemize}
-We refer readers to Becker et al.~\cite{Becker:2014} for an extended discussion of weak versus strong
-sustainability.
+We refer readers to Becker et al.~\cite{Becker:2014} for an extended discussion
+of weak versus strong sustainability.
 
 \subsection{Description of Opportunity, Challenges, and Obstacles}
-The opportunity is to create a catalog of success/failure for current and future software projects to be prepared for transitions and achieve sustainability of the software.
 
-The obstacle is more superficial, in finding a champion to gather such information.
-It will be a challenge to keep this information and surveys updated.
-With changing rapidly changing industry landscapes, an obsolete survey could be of less or no use.
+The opportunity is to create a catalog of success/failure for current and future
+software projects to be prepared for transitions and achieve sustainability of
+the software.
+
+The obstacle is more superficial, in finding a champion to gather such
+information. It will be a challenge to keep this information and surveys
+updated. With changing rapidly changing industry landscapes, an obsolete survey
+could be of less or no use.
 
@@
 
-Identify projects that are collaborative, perhaps by reviewing funded projects from programs specifically geared towards industry academic collaborations.
+Identify projects that are collaborative, perhaps by reviewing funded projects
+from programs specifically geared towards industry academic collaborations.
 
-Develop a systematic process for conducting case studies (what kind of data are being gathered about each case)
+Develop a systematic process for conducting case studies (what kind of data are
+being gathered about each case)
 
 \subsection{Plan for Future Organization}
-No concrete plans have been made at this point.
-If the community can rally behind this topic, some momentum could be built.
-Those interested should post at \url{https://github.com/WSSSPE/meetings/issues/46}
+
+No concrete plans have been made at this point. If the community can rally
+behind this topic, some momentum could be built. Those interested should post at
+\url{https://github.com/WSSSPE/meetings/issues/46}
 
@@
 \subsection{Key Milestones and Responsible Parties}
-A key portion of this effort will require focused surveys of projects which have succeeded and failed in transition.
-Both these categories will yield good learning on what works and what does not work.
-The group has identified what needs to be studied further, but has not identified responsible parties to conduct them.
 
-Community could help in gathering data by means of Interviews; Historical documents / documentation and Surveys
+A key portion of this effort will require focused surveys of projects which have
+succeeded and failed in transition. Both these categories will yield good
+learning on what works and what does not work. The group has identified what
+needs to be studied further, but has not identified responsible parties to
+conduct them.
+
+Community could help in gathering data by means of Interviews; Historical
+documents / documentation and Surveys
 
@@
 \item Origin: Where did project start? 
-\item People involved: How many people in original project were involved in transition/collaboration? 
+\item People involved: How many people in original project were involved in
+transition/collaboration?
 \item Specs on software
@@
 \subsection{Description of Funding Needed}
-Concrete funding needs were not discussed in this working group but a general impression was that some seed funding would motivate members of this group or others in community to launch a survey effort.
+
+Concrete funding needs were not discussed in this working group but a general
+impression was that some seed funding would motivate members of this group or
+others in community to launch a survey effort.

### 2016-02-05 WSSSPE3/report/wg_main_industry_interaction.tex
Section 5.3 and Appendix F:  spacing. No content change.

@@
 
-Most scientific software is produced as a part of grant-funded research projects typically 
-sponsored by federal governments. If we are interested in the sustainability of
-scientific software, then we need to understand what exactly happens when that
-sponsorship ends. More than likely, the project and its resulting software will
-need to undergo some kind of transition in funding and consequently governance.
+Most scientific software is produced as a part of grant-funded research projects
+typically sponsored by federal governments. If we are interested in the
+sustainability of scientific software, then we need to understand what exactly
+happens when that sponsorship ends. More than likely, the project and its
+resulting software will need to undergo some kind of transition in funding and
+consequently governance.
 
 At WSSSPE3 our group was interested in better understanding successful pathways
-for scientific software to ``transition'' from grant-funded research
-projects to industry sponsorship. (This may be an initially awkward
-phrase---some software projects will begin their life being sponsored by
-industry, or result in collaboration between industry and academia. In such
-cases, there is still a need to understand how IP and how maintenance of the
-software is sustained over time.)
+for scientific software to ``transition'' from grant-funded research projects to
+industry sponsorship. (This may be an initially awkward phrase---some software
+projects will begin their life being sponsored by industry, or result in
+collaboration between industry and academia. In such cases, there is still a
+need to understand how IP and how maintenance of the software is sustained over
+time.)
 
@@
 
-Most previous research and discussion of industry and academic collaboration, sharing, and funding of research software has focused on the impact of such arrangements.
-Examples of these types of reports are as follows:
+Most previous research and discussion of industry and academic collaboration,
+sharing, and funding of research software has focused on the impact of such
+arrangements. Examples of these types of reports are as follows:
 
@@
 
-Although sustainability transitions are often studied under the broad umbrella of ``technology
-transfer,'' we believe there are likely to be a number of different ways in which
-a pathway from initial production to long-term maintenance and secure funding is
-achieved. In short, industry sponsorship and/or direct participation is an important aspect of sustaining
-scientific software, but our current understanding of these transitions focuses
-narrowly on commercial successes or failures of those collaborations.
+Although sustainability transitions are often studied under the broad umbrella
+of ``technology transfer,'' we believe there are likely to be a number of
+different ways in which a pathway from initial production to long-term
+maintenance and secure funding is achieved. In short, industry sponsorship
+and/or direct participation is an important aspect of sustaining scientific
+software, but our current understanding of these transitions focuses narrowly on
+commercial successes or failures of those collaborations.
 
 In looking at existing literature that addresses industry transitions, many
-reports (such as those listed above) focus on benefits that accrue to the private sector, or to a government
-that originally sponsored the research project. This literature does not address
-the impact that these transitions have on the accessibility or usability of the
-software, or the impact that these transitions have on the career of the
-researchers involved.
+reports (such as those listed above) focus on benefits that accrue to the
+private sector, or to a government that originally sponsored the research
+project. This literature does not address the impact that these transitions have
+on the accessibility or usability of the software, or the impact that these
+transitions have on the career of the researchers involved.
 
@@
 \begin{enumerate}
-\item To complete a set of case studies which look at successful and unsuccessful transitions between academic researchers and industry
-\item To create a generalizable framework, which might allow for a broader study of different transition pathways (other than between academia and industry)
+
+\item To complete a set of case studies which look at successful and
+unsuccessful transitions between academic researchers and industry
+
+\item To create a generalizable framework, which might allow for a broader study
+of different transition pathways (other than between academia and industry)
+
 \end{enumerate}
 
-The main plan for the group going forward is the creation of a white paper on the topic of sustainability transitions.
+The main plan for the group going forward is the creation of a white paper on
+the topic of sustainability transitions.
 
 \subsubsection{Landing Page}
-Transitions Pathways discussions can be posted at \url{https://github.com/WSSSPE/meetings/issues/46} or an email be sent to Nic Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} to find out more about the group's efforts and how to participate.
+
+Transitions Pathways discussions can be posted at
+\url{https://github.com/WSSSPE/meetings/issues/46} or an email be sent to Nic
+Weber\footnote{email: \href{mailto:nmweber@uw.edu}{nmweber@uw.edu}} to find out
+more about the group's efforts and how to participate.

### 2016-02-05 WSSSPE3/report/wg_appendix_industry_interaction.tex
Section 5.3 and Appendix F: minor changes.

@@
 
-Our initial broad question was, ``What makes for successful transitions of
+The group's initial broad question was, ``What makes for successful transitions of
 scientific software from academia to industry?'' There are a number of potential
@@
 \item The project could transition to a \textbf{community supported model}
-whereby ownership, maintenance, and stewardship of the software become similar
+whereby the software's ownership, maintenance, and stewardship  become similar
 to peer-production models in open-source (e.g., see
@@
 
-We characterized each of these potential changes in funding as ``transition
+We characterized each of the above potential changes in funding as ``transition
 pathways'' to sustainable software (see similar work by Geels and
@@
 
-Our work at WSSSPE3 included the following activities (described in detail
+Our work at WSSSPE3 included the following three activities (described in more detail
 below): (1) brainstorming goals for this type of research, (2) imagining
@@
 transitions of researchers between academia and industry. This might address 
-each of the transition types (described below). Successful transitions are
+each of the transition types (to be described below). Successful transitions are
 described as those that lead to either weak or strong sustainability (also
@@
 
-Community could help in gathering data by means of Interviews; Historical
-documents / documentation and Surveys
+Community members could help in gathering data by means of interviews, historical
+documents or documentation, and surveys.
 
-An example data collection will be: 
+An example of data collection will be: 
 \begin{itemize}

### 2016-02-05 WSSSPE3/report/wg_main_industry_interaction.tex
Section 5.3 and Appendix F: minor changes.

@@
 communities: \url{http://www.raeng.org.uk/policy/dowling-review}
-\item Pathway to Impact - UK report: two pages of grant proposals are asked to
+\item Pathway to Impact -- UK report: two pages of grant proposals are asked to
 forecast what impact they might have (including environmental, academic, economic).

### 2016-02-05 WSSSPE3/report/wg_appendix_eng_design.tex
Added line breaks. No content change.
Section 5.5 and Appendix G: spacing.

@@
 
-Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.
-Venters\footnote{email: \href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points of contact for this working group, and be responsible for ensuring timely progress of the planned actions.
+Birgit Penzenstadler\footnote{email:
+\href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}}
+and Colin C. Venters\footnote{email:
+\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points
+of contact for this working group, and be responsible for ensuring timely
+progress of the planned actions.
 
@@
 
-The group included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy.
-Each participant was invited to give their perspective on the topic area and what they thought were the crucial points for discussion.
-There was a general consensus that there was a need for relating principles to practice for the computational science and engineering community.
-Furthermore, various members of the group expressed their interest in tools and best practices for facilitating the maintenance and evolution of scientific software systems.
-It was agreed to identify principles from software engineering and from sustainability design and, based on those lists, discuss what each of those would mean applied to specific example systems from the expert domains of some of the group members.
-The group identified a number of software engineering principles drawn from the Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
+The group included members from different backgrounds, including quantum
+chemistry, epidemiology, computer science, software engineering, and microscopy.
+Each participant was invited to give their perspective on the topic area and
+what they thought were the crucial points for discussion. There was a general
+consensus that there was a need for relating principles to practice for the
+computational science and engineering community. Furthermore, various members of
+the group expressed their interest in tools and best practices for facilitating
+the maintenance and evolution of scientific software systems. It was agreed to
+identify principles from software engineering and from sustainability design
+and, based on those lists, discuss what each of those would mean applied to
+specific example systems from the expert domains of some of the group members.
+The group identified a number of software engineering principles drawn from the
+Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}.
 
-Software design principles included: abstraction; coupling and cohesion; decomposition and modularization; encapsulation and information hiding; separation of interface and implementation; sufficiency completeness \& primitiveness; and separation of concerns.
-Similarly, user interface design principles included: learnability; user familiarity; consistency; minimal surprise; recoverability; user guidance; and user diversity.
-The sustainability design principles were drawn from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2014}.
-The manifesto states that sustainability is systemic; multidimensional; interdisciplinary; transcends the system's purpose; applies to both a system and its wider contexts; requires action on multiple levels; requires multiple timescales; changing design to take into account long-term effects doesn't automatically imply sacrifices; system visibility is a precondition for and enabler of sustainability design.
-A number of sustainable software engineering principles proposed by Tate~\cite{tate:2005} were also considered including: continual refinement of product and project practices; a working product at all times; continual emphasis on design; and value defect prevention over defect detection.
+Software design principles included: abstraction; coupling and cohesion;
+decomposition and modularization; encapsulation and information hiding;
+separation of interface and implementation; sufficiency completeness \&
+primitiveness; and separation of concerns. Similarly, user interface design
+principles included: learnability; user familiarity; consistency; minimal
+surprise; recoverability; user guidance; and user diversity. The sustainability
+design principles were drawn from the Karlskrona Manifesto on Sustainability
+Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic;
+multidimensional; interdisciplinary; transcends the system's purpose; applies to
+both a system and its wider contexts; requires action on multiple levels;
+requires multiple timescales; changing design to take into account long-term
+effects doesn't automatically imply sacrifices; system visibility is a
+precondition for and enabler of sustainability design. A number of sustainable
+software engineering principles proposed by Tate~\cite{tate:2005} were also
+considered including: continual refinement of product and project practices; a
+working product at all times; continual emphasis on design; and value defect
+prevention over defect detection.
 
-This congregated list is an initial collection of principles that could be extended by adding from further related work form separate disciplines within the field of software engineering, including requirements engineering, software architecture, and testing.
-The group identified two example systems to discuss the application of the principles.
-The first one was a quantum chemistry system that allows the analysis of the characteristics and capabilities of molecules and solids.
-The second one was a modeling system for malaria that permitted biologists to analyze a range of datasets across geography, biology, and epidemiology, and add their own datasets.
-The group then examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference.
-This raised the question, what do the principles mean for  computational scientific and engineering software?
-Similarly, how do the principles relate to non-functional requirements?
-It was suggested that at the very minimum, that sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software:
+This congregated list is an initial collection of principles that could be
+extended by adding from further related work form separate disciplines within
+the field of software engineering, including requirements engineering, software
+architecture, and testing. The group identified two example systems to discuss
+the application of the principles. The first one was a quantum chemistry system
+that allows the analysis of the characteristics and capabilities of molecules
+and solids. The second one was a modeling system for malaria that permitted
+biologists to analyze a range of datasets across geography, biology, and
+epidemiology, and add their own datasets. The group then examined the principles
+and took a retrospective analysis of what the developers did in practice against
+how the principles could have made a difference. This raised the question, what
+do the principles mean for computational scientific and engineering software?
+Similarly, how do the principles relate to non-functional requirements? It was
+suggested that at the very minimum, that sustainable software engineering
+principles should be mapped to two core quality attributes that underpin
+technically sustainable software:
 \begin{itemize}
-\item Extensibility: the software's ability to be extended and the level of effort required to implement the extension;
-\item Maintainability: the effort required to locate and fix an error in operational software.
+\item Extensibility: the software's ability to be extended and the level of
+effort required to implement the extension;
+
+\item Maintainability: the effort required to locate and fix an error in
+operational software.
+
 \end{itemize}
-These fundamental building blocks could then be extended to include other quality attributes such as portability, reusability, scalability, usability, and energy efficiency etc.
-Nevertheless, this raises the question of what metrics and measures are suitable to demonstrate the sustainability of the software.
-In addition, what do the five dimensions of sustainability mean for scientific software, i.e., environmental, economic, social, technical and individual?
+These fundamental building blocks could then be extended to include other
+quality attributes such as portability, reusability, scalability, usability, and
+energy efficiency etc. Nevertheless, this raises the question of what metrics
+and measures are suitable to demonstrate the sustainability of the software. In
+addition, what do the five dimensions of sustainability mean for scientific
+software, i.e., environmental, economic, social, technical and individual?
 
 \subsection{Description of Opportunity, Challenges, and Obstacles}
-The opportunity was identified to distill existing software engineering and sustainability design knowledge into ``bite sized'' chunks for the Computational Science and Engineering Community.
-In addition, two primary challenges were identified:
+The opportunity was identified to distill existing software engineering and
+sustainability design knowledge into ``bite sized'' chunks for the Computational
+Science and Engineering Community. In addition, two primary challenges were
+identified:
 \begin{itemize}
@@
 \subsection{Key Next Steps}
-In order to achieve (1) the systematic analysis a number of example systems from different scientific domains with regards to the identified principles, (2) the identification of the commonalities and gaps in applying principles to different scientific systems, and (3) the proposal of a set of guidelines on the principles, the following next steps were discussed:
+In order to achieve (1) the systematic analysis a number of example systems from
+different scientific domains with regards to the identified principles, (2) the
+identification of the commonalities and gaps in applying principles to different
+scientific systems, and (3) the proposal of a set of guidelines on the
+principles, the following next steps were discussed:
 
@@
 \subsection{Key Milestones and Responsible Parties}
-The following key milestones were discussed as a roadmap for the set of guidelines on software engineering principles:
+The following key milestones were discussed as a roadmap for the set of
+guidelines on software engineering principles:
 \begin{itemize}

### 2016-02-05 WSSSPE3/report/wg_main_eng_design.tex
Section 5.5 and Appendix G: spacing.

@@
 
-Software engineering principles form the basis of methods, techniques, methodologies and tools~\cite{Vliet:2008}.
-However, there is often a mismatch between software engineering theory and practice particularly in the fields of computational science and engineering, which can lead to the development of unsustainable software~\cite{Merali:2010,hettrick:2014}.
-Understanding and applying software engineering principles is essential in order to create and maintain sustainable software~\cite{Becker:2016}.
+Software engineering principles form the basis of methods, techniques,
+methodologies and tools~\cite{Vliet:2008}. However, there is often a mismatch
+between software engineering theory and practice particularly in the fields of
+computational science and engineering, which can lead to the development of
+unsustainable software~\cite{Merali:2010,hettrick:2014}. Understanding and
+applying software engineering principles is essential in order to create and
+maintain sustainable software~\cite{Becker:2016}.
 
 \subsubsection{Fit with related activities}
-The group discussion focused on identifying existing principles of software engineering design that could be adopted by the computational science and engineering communities.
+
+The group discussion focused on identifying existing principles of software
+engineering design that could be adopted by the computational science and
+engineering communities.
 
@@
 
-Software engineering principles form the basis of methods, techniques, methodologies and tools.
-This group, which included members from different backgrounds, including quantum chemistry, epidemiology, computer science, software engineering, and microscopy, discussed the principles of software engineering design for sustainable software (starting with principles from the Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015}, Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge (SWEBOK)~\cite{swebokv3}) and their application in various domains including quantum chemistry and epidemiology.
-The group examined the principles and took a retrospective analysis of what the developers did in practice against how the principles could have made a difference, and asked, what do the principles mean for  computational scientific and engineering software, and how do the principles relate to non-functional requirements?
-It
-appeared that the sustainable software engineering principles should be mapped to two core quality attributes that underpin technically sustainable software: extensibility, the software's ability to be extended and the level of effort required to implement the extension; and
-maintainability: the effort required to locate and fix an error in operational software.
+Software engineering principles form the basis of methods, techniques,
+methodologies and tools. This group, which included members from different
+backgrounds, including quantum chemistry, epidemiology, computer science,
+software engineering, and microscopy, discussed the principles of software
+engineering design for sustainable software (starting with principles from the
+Karlskrona Manifesto on Sustainability Design~\cite{Becker:2015},
+Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
+(SWEBOK)~\cite{swebokv3}) and their application in various domains including
+quantum chemistry and epidemiology. The group examined the principles and took a
+retrospective analysis of what the developers did in practice against how the
+principles could have made a difference, and asked, what do the principles mean
+for computational scientific and engineering software, and how do the principles
+relate to non-functional requirements? It appeared that the sustainable software
+engineering principles should be mapped to two core quality attributes that
+underpin technically sustainable software: extensibility, the software's ability
+to be extended and the level of effort required to implement the extension; and
+maintainability: the effort required to locate and fix an error in operational
+software.
 
@@
 \subsubsection{Plans}
-The next steps in this endeavor are to (1) Systematically analyze a number of example systems from different scientific domains with regards to the identified principles, to (2) Identify the commonalities and gaps in applying those principles to different scientific systems, and to (3) Propose a set of guidelines on the principles and how they exemplary apply to scientific software system.
-Preliminary work will be carried out through undergraduate or post-graduate student projects.
+
+The next steps in this endeavor are to (1) Systematically analyze a number of
+example systems from different scientific domains with regards to the identified
+principles, to (2) Identify the commonalities and gaps in applying those
+principles to different scientific systems, and to (3) Propose a set of
+guidelines on the principles and how they exemplary apply to scientific software
+system. Preliminary work will be carried out through undergraduate or
+post-graduate student projects.
 
 \subsubsection{Landing Page}
-In the absence of a landing page, the Principles for Software Engineering Design for Sustainable Software working group requests an email be sent to Birgit Penzenstadler\footnote{email: \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}} and Colin C.\ Venters\footnote{email:\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} to find out more about the group's efforts and how to participate.
+
+In the absence of a landing page, the Principles for Software Engineering Design
+for Sustainable Software working group requests an email be sent to Birgit
+Penzenstadler\footnote{email:
+\href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}}
+and Colin C.\
+Venters\footnote{email:\href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}}
+to find out more about the group's efforts and how to participate.

### 2016-02-05 WSSSPE3/report/wg_appendix_eng_design.tex
Section 5.5 and Appendix G: tiny changes.

@@
 \href{mailto:birgit.penzenstadler@csulb.edu}{birgit.penzenstadler@csulb.edu}}
-and Colin C. Venters\footnote{email:
+and Colin C.\ Venters\footnote{email:
 \href{mailto:c.venters@hud.ac.uk}{c.venters@hud.ac.uk}} will serve as the points
@@
 \item Birgit Penzenstadler -- California State University, CA, USA
-\item Colin C. Venters -- University of Huddersfield, Huddersfield, UK
+\item Colin C.\ Venters -- University of Huddersfield, Huddersfield, UK
 \item Matthias Bussonnier -- UC Berkeley, CA, USA
@@
 
-The group included members from different backgrounds, including quantum
-chemistry, epidemiology, computer science, software engineering, and microscopy.
+This group is comprised  of members from different backgrounds, including quantum
+chemistry, epidemiology, microscopy, computer science, and software engineering.
 Each participant was invited to give their perspective on the topic area and
@@
 
-Software design principles included: abstraction; coupling and cohesion;
+Software design principles include abstraction; coupling and cohesion;
 decomposition and modularization; encapsulation and information hiding;
-separation of interface and implementation; sufficiency completeness \&
+separation of interface and implementation; sufficiency completeness and
 primitiveness; and separation of concerns. Similarly, user interface design
-principles included: learnability; user familiarity; consistency; minimal
+principles include learnability; user familiarity; consistency; minimal
 surprise; recoverability; user guidance; and user diversity. The sustainability
 design principles were drawn from the Karlskrona Manifesto on Sustainability
-Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic;
-multidimensional; interdisciplinary; transcends the system's purpose; applies to
+Design~\cite{Becker:2014}. The manifesto states that sustainability is systemic,
+multidimensional, and interdisciplinary; transcends the system's purpose; applies to
 both a system and its wider contexts; requires action on multiple levels;
 requires multiple timescales; changing design to take into account long-term
-effects doesn't automatically imply sacrifices; system visibility is a
+effects does not automatically imply sacrifices; system visibility is a
 precondition for and enabler of sustainability design. A number of sustainable
@@
 \subsection{Key Next Steps}
-In order to achieve (1) the systematic analysis a number of example systems from
+
+In order to achieve the following three goals: (1) a systematic analysis of a number of example systems from
 different scientific domains with regards to the identified principles, (2) the
-identification of the commonalities and gaps in applying principles to different
-scientific systems, and (3) the proposal of a set of guidelines on the
-principles, the following next steps were discussed:
+identification of the commonalities and gaps in applying the principles to different
+scientific systems, and (3) a proposal of a set of guidelines on the
+principles, the next steps were discussed in the following subsections.
 
 \subsection{Plan for Future Organization}
+
 The following plan for future organization was discussed:
@@
 \item Design and pilot study.
-\item Organizing coordinating online calls via Google Hangout.
+\item Organize coordinating online calls via Google Hangout.
 \end{itemize}
@@
 \subsection{What Else is Needed?}
+
 \begin{itemize}
@@
 \subsection{Key Milestones and Responsible Parties}
+
 The following key milestones were discussed as a roadmap for the set of
@@
 \subsection{Description of Funding Needed}
+
 Specific funding was not discussed in this working group. However, this is a

### 2016-02-05 WSSSPE3/report/wg_main_eng_design.tex
Section 5.5 and Appendix G: tiny changes.

@@
 
-Software engineering principles form the basis of methods, techniques,
+Principles for software engineering form the basis of methods, techniques,
 methodologies and tools~\cite{Vliet:2008}. However, there is often a mismatch
@@
 
-Software engineering principles form the basis of methods, techniques,
-methodologies and tools. This group, which included members from different
+Software engineering principles form the foundation of methods, techniques,
+methodologies, and tools.  Consisting of members from different
 backgrounds, including quantum chemistry, epidemiology, computer science,
-software engineering, and microscopy, discussed the principles of software
+software engineering, and microscopy, this group discussed the principles of software
 engineering design for sustainable software (starting with principles from the
@@
 Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
-(SWEBOK)~\cite{swebokv3}) and their application in various domains including
+(SWEBOK)~\cite{swebokv3}) and their applications in various domains including
 quantum chemistry and epidemiology. The group examined the principles and took a
@@
 engineering principles should be mapped to two core quality attributes that
-underpin technically sustainable software: extensibility, the software's ability
-to be extended and the level of effort required to implement the extension; and
-maintainability: the effort required to locate and fix an error in operational
+underpin technically sustainable software: \emph{extensibility}, the software's ability
+to be extended and the effort level required to implement the extension; and
+\emph{maintainability}: the effort required to locate and fix an error in operational
 software.
@@
 principles to different scientific systems, and to (3) Propose a set of
-guidelines on the principles and how they exemplary apply to scientific software
-system. Preliminary work will be carried out through undergraduate or
+guidelines on the principles and examine how they exemplarily apply to scientific software
+systems. Preliminary work will be carried out through undergraduate or
 post-graduate student projects.

### 2016-02-05 WSSSPE3/report/wg_appendix_eng_design.tex
small updates

@@
 
-This group is comprised  of members from different backgrounds, including quantum
+This group was comprised  of members from different backgrounds, including quantum
 chemistry, epidemiology, microscopy, computer science, and software engineering.
@@
 
-Software design principles include abstraction; coupling and cohesion;
-decomposition and modularization; encapsulation and information hiding;
-separation of interface and implementation; sufficiency completeness and
-primitiveness; and separation of concerns. Similarly, user interface design
-principles include learnability; user familiarity; consistency; minimal
-surprise; recoverability; user guidance; and user diversity. The sustainability
+Software design principles include abstraction, coupling and cohesion,
+decomposition and modularization, encapsulation and information hiding,
+separation of interface and implementation, sufficiency completeness and
+primitiveness, and separation of concerns. Similarly, user interface design
+principles include learnability, user familiarity, consistency, minimal
+surprise, recoverability, user guidance, and user diversity. The sustainability
 design principles were drawn from the Karlskrona Manifesto on Sustainability
@@
 scientific systems, and (3) a proposal of a set of guidelines on the
-principles, the next steps were discussed in the following subsections.
+principles, the following next steps were discussed.
 

### 2016-02-05 WSSSPE3/report/wg_main_eng_design.tex
small updates

@@
 Tate~\cite{tate:2005}, and the Software Engineering Body of Knowledge
-(SWEBOK)~\cite{swebokv3}) and their applications in various domains including
+(SWEBOK)~\cite{swebokv3}) and their application in various domains including
 quantum chemistry and epidemiology. The group examined the principles and took a

### 2016-02-05 WSSSPE3/report/wg_main_metrics.tex
Added line breaks. No content change.
more review of changes
Section 5.6: 1 change.

--- 2015-12-20 WSSSPE3/report/wg_main_metrics.tex
@@
 
-Metrics for scientific software are important for tenure and promotion,
+Metrics for scientific software are important for many purposes: tenure and promotion,
 scientific impact, discovery, reducing duplication, serving as a basis for
@@
 elements released under the NSF SI2 program all included metrics as a required
-component with submitters requested to include {\it ``a list of tangible metrics,
-with end user involvement, to be used to measure the success of the software
-element developed, \dots''}. These metrics are then reported as part of annual
-reports to NSF by the projects. Although neither the proposal text describing
-the metrics nor the reported metric results are publicly available, there is
-reason to believe that the community will be willing to provide this information
-through a survey mechanism. This survey would be created by one of the student
-group members. Similarly, it was suggested that a software metrics survey be
-sent to the UK SFTF (Software For The Future, led by the Engineering and Physical Sciences Research Council) and TRDF (Tools and Resources Development Fund, led by the Biotechnology and Biological Sciences Research Council) software projects to ask them what metrics would be useful to report.
-The remainder of the discussion focused mainly on the creation
-of a white paper on this topic. This resulted in a paper outline and writing
-assignments with the goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of Electrical and Electronic Engineers Computing in Science and Engineering magazine), or JORS (Journal of Open Research Software).
-More information about the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
+component with submitters requested to include {\it ``a list of tangible
+metrics, with end user involvement, to be used to measure the success of the
+software element developed, \dots''}. These metrics are then reported as part of
+annual reports to NSF by the projects. Although neither the proposal text
+describing the metrics nor the reported metric results are publicly available,
+there is reason to believe that the community will be willing to provide this
+information through a survey mechanism. This survey would be created by one of
+the student group members. Similarly, it was suggested that a software metrics
+survey be sent to the UK SFTF (Software For The Future, led by the Engineering
+and Physical Sciences Research Council) and TRDF (Tools and Resources
+Development Fund, led by the Biotechnology and Biological Sciences Research
+Council) software projects to ask them what metrics would be useful to report.
+The remainder of the discussion focused mainly on the creation of a white paper
+on this topic. This resulted in a paper outline and writing assignments with the
+goal of publishing in venues including WSSSPE4, IEEE CiSE (Institute of
+Electrical and Electronic Engineers Computing in Science and Engineering
+magazine), or JORS (Journal of Open Research Software). More information about
+the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
 

### 2016-02-05 WSSSPE3/report/wg_main_metrics.tex
small change to change in 5.6

--- 2015-12-20 WSSSPE3/report/wg_main_metrics.tex
@@
 
-Metrics for scientific software are important for tenure and promotion,
+Metrics for scientific software are important for many purposes, including tenure and promotion,
 scientific impact, discovery, reducing duplication, serving as a basis for

### 2016-02-05 WSSSPE3/report/wg_appendix_metrics.tex
Added line breaks. No content change.
editing changes from last 2 weeks - just starting
Section 5.7, Appendix H and I: small changes.

@@
 
-Gabrielle Allen\footnote{email: \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} will serve as the point of contact for this working group.
+Gabrielle Allen\footnote{email: \href{mailto:gdallen@illinois.edu}{gdallen@illinois.edu}} 
+will serve as the point of contact for this working group.
 
@@
 
-The following summary of the group's discussion represents the Useful Metrics
+The following is a summary of the Useful Metrics
 for Scientific Software working group's discussion during the WSSSPE3 workshop.
@@
 \item
-However, there is no commonly used standard (DOAP is a good standard but not
+However, there is no commonly-used standard (DOAP is a good standard but not
 widely adopted)
@@
 \item
-        Where should funders place funding for best impact (e.g., funding two-star software versus three-star) and where there are gaps
+Where should funders place funding for best impact (e.g., funding two-star
+software versus three-star) and where there are gaps
 
@@
 The following opportunities, challenges, and obstacles were discussed:
-
+%
 \begin{itemize}
@@
 \item
-We don't know if there is a common set of metrics
+We do not know if there is a common set of metrics
 
@@
 The following next steps were discussed:
-
+%
 \begin{itemize}
@@
 The following plan for future organization was discussed:
-
+%
 \begin{itemize}
@@
 Our group has created a white paper outline with sections assigned to the above
-individuals, plus see Section 2 response above for timeline
+individuals; see timeline below. %plus see Section 2 response above for timeline
 
 \item
-Organizing coordinating phone calls
+Organize coordinating phone calls.
 
@@
 \item
-Est.\ July 2016: By the time of next CFP for WSSSPE, to have a complete draft of the white paper
+Est.\ July 2016: By the time of next CFP for WSSSPE, to have a complete draft of
+the white paper
 

### 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
Section 5.7, Appendix H and I: small changes.

@@
 While little training focuses specifically on sustainable
-software, a variety of training activities increase researcher awareness of
-and engagement with software professionals and software engineering practices.
+software, a variety of training activities increase researcher awareness 
+and engagement with software engineering practices and software professionals.
 Research Software Engineers are being recognized as critical contributors to
@@
 less often incorporated as explicit learning outcomes. With software (and
-similarly, data) often the only tangible artifact of a research method or
+similarly, data) often being the only tangible artifact of a research method or
 protocol, the dependency between software applications and the quality of

### 2016-02-05 WSSSPE3/report/wg_main_training.tex
Added line breaks. No content change.
Section 5.7, Appendix H and I: small changes.

@@
 
-\subsubsection{Fit with related activities} Three existing venues for discussion
-of related activities are identified:
+\subsubsection{Fit with related activities}
 
+Three existing venues for discussion of related events are
+identified:
+%
 \begin{itemize}
@@
 
-Next steps have been identified to quickly test whether there is interest in
+Next steps were identified to quickly test whether there is interest in
 establishing a community committed to increasing the degree of coordination
-across training projects. See Appendix~\ref{sec:appendix_training} for more details about the discussion.
+across training projects. See Appendix~\ref{sec:appendix_training} for more
+details about the discussion.
 

### 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
changes to changes in 5.7, H, and I

@@
 While little training focuses specifically on sustainable
-software, a variety of training activities increase researcher awareness 
+software, a variety of training activities increase researcher awareness of
 and engagement with software engineering practices and software professionals.

### 2016-02-05 WSSSPE3/report/wg_main_training.tex
changes to changes in 5.7, H, and I

@@
 
-Next steps were identified to quickly test whether there is interest in
+Some next steps were identified to quickly test whether there is interest in
 establishing a community committed to increasing the degree of coordination

### 2016-02-05 WSSSPE3/report/wg_appendix_software_credit.tex
Added line breaks. No content change.
Section 5.8, Appendix I and J: small changes.

@@
 \subsection{Group Members}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% g
 
@@
 
-Initial discussions focused on both various mechanisms for, and the
+Initial discussions mainly focused on both various mechanisms for, and the
 philosophical approach behind, crediting software in scientific papers. These
@@
 
-\item Mozilla Science Lab's recently introduced Contributorship Badges for
-Science~\cite{Mozilla_badges}, which provide a badge---associated with
-ORCID~\cite{orcid}---for author contributions based on the taxonomy
+\item Mozilla Science Lab has recently introduced Contributorship Badges for
+Science~\cite{Mozilla_badges}, which provides a badge---associated with
+ORCID~\cite{orcid}---for identifying author contributions in a paper based on the taxonomy
 outlined in Project CRediT.
@@
 reliance of a particular work on prior science, materials, or software---in
-other words, is there a difference between depending on prior scientific
-advances and depending on particular software (or experimental equipment)?
+other words, whether there is a difference between depending on prior scientific
+advances and depending on certain software (or experimental equipment).
 Multiple contributors converged on the conclusion that unique capabilities
@@
 
-During the first day of WSSSPE3 breakout sessions, the group discussed the
+During the first day of WSSSPE3 breakout sessions, the group discussed and deliberated over the
 Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
@@
 (e.g., movies, television shows) and contributors, along with relevant metadata.
-One important use of this system is to track rights and credit for contributors
+One important use of the EIDR system is to track rights and credits for contributors
 to entertainment works in order to distribute revenues---similar to the proposed
@@
 citations) from the value of a work in order to give credit, moving towards
-qualitative or anecdotal evidence of value. Other topics brought up included a
+qualitative or anecdotal evidence of value. Other topics that were brought up included a
 form of PageRank~\cite{Brin1998} for citations, based on number of mentions, and
@@
 
-On the second day of WSSSPE3, the working group decided to put aside the
+On the second day of WSSSPE3, the group decided to put aside the
 taxonomy of contributions and focus on software citations to ensure developers
@@
 software citations are standardized, the goal would be to return to establishing different
-roles\slash contributions for this credit. Following this decision, the group
+roles or contributions for this credit. Following this decision, the group
 identified two necessary actions to move forward:
@@
 
-At a minimum, the metadata needed for software citation %(aside from the DOI for
-%the software itself) \choinote{DOI is in the list below}
-includes:
+At a minimum, the metadata required for software citation includes:
 \begin{itemize}
@@
 and biology~\cite{Howison2015}. The group recommended collecting similar
-examples from other communities, then developing a software citation principles
+examples from other communities, and then developing a set of software citation principles
 document in concert with the FORCE11 Software Citation Working Group (see
@@
 
-The group also discussed briefly whether software used directly in a
-publication---whether to perform simulation, analysis, or as a dependency for
-new software developed---should be distinguished from other references due to
+The group further discussed briefly whether software used directly in a
+publication---whether to perform simulation or analysis, or as a dependency for
+new software to be developed---should be distinguished from other references due to
 the dependence of the study on these research artifacts. Suggestions included a
@@
 
-Finally, although a discrete task from software citations themselves,
+Finally, although a discrete task from software citations,
 significant discussion focused on ensuring software citations are indexed in the
@@
 There currently is no standard mechanism for citing software or
-receiving credit for software (akin to publications/citations). Software is
+receiving credit for software (akin to citations for publications). Software is
 eligible for DOI assignment, but DOI metadata fields are not well tuned or
@@
 %we should recognized that this problem goes beyond faculty to recognizing the role of ``Research Software Engineers''
-the group recognized that there is no standard
-policy---generally even within a single university---for software products to be
-included in promotion and tenure dossiers. Thus, it may be difficult to
-encourage valuing software contributions across the United States or United
-Kingdom and globally; furthermore, stakeholders are typically not tenured and
-thus may not be influential enough to change the status quo.
-However, as discussed in Section~\ref{RSE}, this is changing for Research Software Engineers, at least in the UK.
+the group recognized that there is no standard policy---generally even within a
+single university---for software products to be included in promotion and tenure
+dossiers. Thus, it may be difficult to encourage valuing software contributions
+across the United States or United Kingdom and globally; furthermore,
+stakeholders are typically not tenured and thus may not be influential enough to
+change the status quo. However, as discussed in Section~\ref{RSE}, this is
+changing for Research Software Engineers, at least in the UK.
 
@@
 citation activities with FORCE11-SCWG. Within a month of the meeting, Niemeyer will
-organize a virtual meeting of the group and lead the division of
+organize a virtual meeting of the group and manage the division of
 responsibilities for compiling existing practices of software citation and

### 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
Section 5.8, Appendix I and J: small changes.

--- 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
@@
 While little training focuses specifically on sustainable
-software, a variety of training activities increase researcher awareness 
+software, a variety of training activities could increase researcher awareness of
 and engagement with software engineering practices and software professionals.

### 2016-02-05 WSSSPE3/report/wg_main_software_credit.tex
Section 5.8, Appendix I and J: small changes.

@@
 
-Modern scientific and engineering research relies considerably on software, 
+Modern scientific and engineering research often relies considerably on software, 
 but currently no standard mechanism exists for citing software or receiving 
@@
 metadata necessary for software to be cited and (2) the mechanism for citing
-software in publications. In addition, discussions also oriented around the
+software in publications. Moreover, discussions also oriented around the
 indexing of software citations necessary for establishing a software citation
@@
 (SCWG), and their efforts will focus (over the next six to nine months) on
-developing a document describing principles for software citation. Following
-publication of that document, the working group will focus on outreach to key
+developing a document describing principles for software citation. Following the
+publication of that document, the group will focus on outreach to key
 groups (e.g., journals, publishers, indexers, professional societies).

### 2016-02-05 WSSSPE3/report/wg_appendix_publishing_SW.tex
Section 5.9 and Appendix K: small changes.

@@
 \begin{itemize}
-\item Steven R. Brandt -- Louisiana State University
+\item Steven R.\ Brandt -- Louisiana State University
 \item Daniel Gunter -- LBNL
@@
 
-\item ACM Transactions on Mathematical Software (TOMS): provides the extra step
+\item ACM Transactions on Mathematical Software (TOMS): provides the additional step
 of having reviewers validate the code which was submitted with the publication.
@@
 
-\item Nature: offering a list of notebooks published alongside more traditional
+\item Nature: offers a list of notebooks published alongside more traditional
 articles, and is looking at ways to make these documents more official. There
@@
 
-\item KBase: narratives build on IPython\slash Jupyter notebooks to build
+\item KBase: offers narratives built on IPython or Jupyter notebooks for assembling
 publications that are reproducible, and can be commented or annotated.
@@
 
-\item The Research Ideas and Outcomes (RIO) journal is an effort to publish fragmentary results that can subsequently be
-combined into a single content item.
+\item The Research Ideas and Outcomes (RIO) journal is an effort to publish
+fragmentary results that can subsequently be combined into a single content
+item.
 
@@
 Create the first version of the web page to be displayed on the Software
-Sustainability Institute's website: \url{http://www.software.ac.uk}.
+Sustainability Institute's website. %: \url{http://www.software.ac.uk}. \choinote{already appeared above)
 We expect the page to be live in early January of 2016.
@@
 of being posted on the Software
-Sustainability Institute's website: \url{http://www.software.ac.uk}. 
+Sustainability Institute's website. %: \url{http://www.software.ac.uk}.  \choinote{already appeared above)
 Neil Chue Hong will take responsibility for the page once it is up.

### 2016-02-05 WSSSPE3/report/wg_main_publishing_SW.tex
Section 5.9 and Appendix K: small changes.

@@
 
-This group explored the value of executable papers (papers whose content
+This working group explored the value of executable papers (papers whose content
 includes the code needed to produce their own results), and other forms of
@@
 easier to trace even if appropriately worded credit for software is not
-provided. In addition, these notebooks make it possible to provide/define
+provided. In addition, some software notebooks (i.e., active or executable papers; 
+see Appendix~\ref{sec:appendix_publishing_SW} for examples) 
+make it possible to provide or define
 additional metadata to make the tracing of credit clearer. Finally, attributions
@@
 The group felt that the best way to encourage the use of these new publishing
-concepts would be to create and curate a list identifying publishing venues that
+concepts would be to create and curate a list of publishing venues that
 support them. The Software Sustainability Institute agreed to host this list.

### 2016-02-05 WSSSPE3/report/wg_appendix_user_community.tex
Section 5.10 and Appendix L: small changes. DONE for today's submission deadline.

@@
 
-POC:
+Point of contact:
 Dan Gunter\footnote{email: \href{mailto:dkgunter@lbl.gov}{dkgunter@lbl.gov}} and 
-Ethan Davis\footnote{email: \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}}
+Ethan Davis\footnote{email: \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}}.
 
@@
 also thought that the existing books on software communities would need to be
-evaluated in light of differences between Science Software projects and general
-OSS projects in terms of scale, science, acknowledgement and credit, and funding
+evaluated in light of differences between science software projects and general
+open-source software (OSS) projects in terms of scale, science, acknowledgement and credit, and funding
 models.
@@
 engineering practices are not enough, and popular training like Software
-Carpentry does not currently address this issue head on.
+Carpentry does not currently address this issue head-on.
 
@@
 With a small amount of seed funding, it is possible that members of this group or other parties could
-spend the time necessary to devise a survey of existing projects and deploy this, probably including travel to
+spend the time necessary to devise a survey of existing projects and deploy this, probably by traveling to
 meetings and workshops for the various software communities.

### 2016-02-05 WSSSPE3/report/wg_main_user_community.tex
Section 5.10 and Appendix L: small changes. DONE for today's submission deadline.

--- 2015-12-11 WSSSPE3/report/wg_main_user_community.tex
@@
 
-Discussion revolved around a few questions: what is the benefit of having a
-``community'' for software sustainability, what practices and circumstances lead
-to having and maintaining a community, how can funding help or hinder this
-process, and perhaps most importantly, how can best practices be described and
+Discussion revolved around a few questions: what are the benefits of having a
+``community'' for software sustainability; what practices and circumstances may lead
+to having and maintaining a community; how can funding help or hinder this
+process; and perhaps most importantly, how can best practices be described and
 distilled into a document that can help new projects.
@@
 describe what successful projects with engaged communities look like, how to
-replicate this type of project, and look at end-of-life on a community project.
+replicate this type of project, and look at the end of life of a community project.
 Another next step would be better training to increase recognition of need for science software
@@
 
-This group doesn't have a landing page yet. Please send requests to join and contribute to both
-Dan Gunter <dkgunter@lbl.gov> and Ethan Davis <edavis@ucar.edu>.
+This group does not have a landing page yet. Please send requests to join and
+contribute by writing to both Dan Gunter\footnote{email:
+\href{mailto:dkgunter@lbl.gov}{dkgunter@lbl.gov}} and Ethan
+Davis\footnote{email: \href{mailto:edavis@ucar.edu}{edavis@ucar.edu}}.

### 2016-02-05 WSSSPE3/report/WSSSPE3_review.tex
Attempt to enrich the abstract.

@@
 This technical report records and discusses the Third Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE3). 
-The report includes a description of the keynote presentation and a set of lightning
-talks. It also focuses on the discussions, future steps, and future organization for
-a set of self-organized working groups.
-For each group, we include a point of contact and a landing
-page that can be used by those who want to join that group's future activities.
-The main challenge left by the workshop is to see if the groups will execute 
-these activities that they have scheduled, and how the WSSSPE community
-can encourage this to happen.
+Software for Science: Practice and Experiences (WSSSPE3). The report includes a
+description of the keynote presentation and a set of lightning talks. It also
+focuses on the discussions, future steps, and future organization for a set of
+self-organized working groups on topics including developing pathways to funding
+scientific software; constructing useful common metrics for crediting software
+stakeholders; identifying principles for sustainable software engineering
+design; reaching out to research software organizations around the globe; building
+communities for software sustainability. For each group, we include a point of
+contact and a landing page that can be used by those who want to join that
+group's future activities. The main challenge left by the workshop is to see if
+the groups will execute these activities that they have scheduled, and how the
+WSSSPE community can encourage this to happen.
 \end{abstract}

### 2016-02-05 WSSSPE3/report/WSSSPE3_review.tex
Attempt to enrich the abstract.

--- 2016-02-03 WSSSPE3/report/WSSSPE3_review.tex
@@
 This technical report records and discusses the Third Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE3). 
-The report includes a description of the keynote presentation and a set of lightning
-talks. It also focuses on the discussions, future steps, and future organization for
-a set of self-organized working groups.
-For each group, we include a point of contact and a landing
-page that can be used by those who want to join that group's future activities.
-The main challenge left by the workshop is to see if the groups will execute 
-these activities that they have scheduled, and how the WSSSPE community
-can encourage this to happen.
+Software for Science: Practice and Experiences (WSSSPE3). The report includes a
+description of the keynote presentation of the workshop, which served as an overview of sustainable
+scientific software. It also summarizes a set of lightning
+talks in which speakers highlighted to-the-point lessons and challenges
+pertaining to sustaining scientific software. Lastly, the report focuses on the
+discussions, future steps, and future organization for a set of self-organized
+working groups on topics including developing pathways to funding scientific
+software; constructing useful common metrics for crediting software
+stakeholders; identifying principles for sustainable software engineering
+design; reaching out to research software organizations as a global effort;
+building communities for software sustainability. For each group, we include a
+point of contact and a landing page that can be used by those who want to join
+that group's future activities. The main challenge left by the workshop is to
+see if the groups will execute these activities that they have scheduled, and
+how the WSSSPE community can encourage this to happen.
+
 \end{abstract}

### 2016-02-05 WSSSPE3/report/wg_main_publishing_SW.tex
A few more changes that seem to make sense.

@@
 
-\textbf{Reproducibility}: Part of the purpose of these executable paper venues
-is to (at least partially) address the reproducibility issue by making paper
+\begin{itemize}
+\item \textbf{Reproducibility}: Part of the purpose of these executable paper venues
+is to (at least partially) address the reproducibility issue by making papers
 recompute their own results.
 
-\textbf{Software Credit (\S\ref{sec:software-credit})}: Since these forms of
+\item\textbf{Software Credit (\S\ref{sec:software-credit})}: Since these forms of
 publishing must make their sources explicit in order to execute, they should be
 easier to trace even if appropriately worded credit for software is not
-provided. In addition, some software notebooks (i.e., active or executable papers; 
-see Appendix~\ref{sec:appendix_publishing_SW} for examples) 
+provided. In addition, some  notebooks, i.e., active or executable papers,
+(see Appendix~\ref{sec:appendix_publishing_SW} for some concrete examples) 
 make it possible to provide or define
@@
 
-\textbf{Best Practices (\S\ref{sec:best-practices})}: Because an executable
+\item \textbf{Best Practices (\S\ref{sec:best-practices})}: Because an executable
 paper showcases the code, and the code itself is subject to the review process,
@@
 more likely to be achieved.
+\end{itemize}
 

### 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
Appendix I: small changes.

--- 2016-02-05 WSSSPE3/report/wg_appendix_training.tex
@@
 While little training focuses specifically on sustainable
-software, a variety of training activities increase researcher awareness 
-and engagement with software engineering practices and software professionals.
+software, a variety of training activities could increase researcher awareness of
+and engagement with software professionals and software engineering practices.
 Research Software Engineers are being recognized as critical contributors to

### 2016-02-06 WSSSPE3/report/wg_appendix_publishing_SW.tex
changes to changes in later sections

@@
 Create the first version of the web page to be displayed on the Software
-Sustainability Institute's website. %: \url{http://www.software.ac.uk}. \choinote{already appeared above)
+Sustainability Institute's website: \url{http://www.software.ac.uk}.
+% \choinote{already appeared above) /katznote{but important to have it here for poeple who are scanning/looking at this part of each section}
 We expect the page to be live in early January of 2016.
@@
 of being posted on the Software
-Sustainability Institute's website. %: \url{http://www.software.ac.uk}.  \choinote{already appeared above)
+Sustainability Institute's website: \url{http://www.software.ac.uk}. % \choinote{already appeared above)
 Neil Chue Hong will take responsibility for the page once it is up.

### 2016-02-06 WSSSPE3/report/wg_appendix_software_credit.tex
Appendix J: Remove an unintended change in comment.
changes to changes in later sections

@@
 \subsection{Group Members}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% g
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-Initial discussions mainly focused on both various mechanisms for, and the
+Initial discussions focused on both various mechanisms for, and the
 philosophical approach behind, crediting software in scientific papers. These
@@
 
-\item Mozilla Science Lab has recently introduced Contributorship Badges for
-Science~\cite{Mozilla_badges}, which provides a badge---associated with
-ORCID~\cite{orcid}---for identifying author contributions in a paper based on the taxonomy
+\item Mozilla Science Lab' recently introduced Contributorship Badges for
+Science~\cite{Mozilla_badges}, which provide a badge---associated with an
+ORCID~\cite{orcid}---that recognizes author contributions using the taxonomy
 outlined in Project CRediT.
@@
 
-During the first day of WSSSPE3 breakout sessions, the group discussed and deliberated over the
+In the breakout sessions on the first day of WSSSPE3, the group discussed and deliberated over the
 Entertainment Identifier Registry (EIDR)~\cite{EIDR} as a potential model for
@@
 software citations are standardized, the goal would be to return to establishing different
-roles or contributions for this credit. Following this decision, the group
+roles\slash contributions for this credit. Following this decision, the group
 identified two necessary actions to move forward:
@@
 and biology~\cite{Howison2015}. The group recommended collecting similar
-examples from other communities, and then developing a set of software citation principles
+examples from other communities, and then developing a software citation principles
 document in concert with the FORCE11 Software Citation Working Group (see
@@
 publication---whether to perform simulation or analysis, or as a dependency for
-new software to be developed---should be distinguished from other references due to
+newly developed software---should be distinguished from other references due to
 the dependence of the study on these research artifacts. Suggestions included a

### 2016-02-06 WSSSPE3/report/wg_main_publishing_SW.tex
changes to changes in later sections

@@
 easier to trace even if appropriately worded credit for software is not
-provided. In addition, some  notebooks, i.e., active or executable papers,
-(see Appendix~\ref{sec:appendix_publishing_SW} for some concrete examples) 
+provided. In addition, they
 make it possible to provide or define

### 2016-02-06 WSSSPE3/report/WSSSPE3_review.tex
A few more changes that seem to make sense.
abstract updates

@@
 talks in which speakers highlighted to-the-point lessons and challenges
-pertaining to sustaining scientific software. Lastly, the report focuses on the
+pertaining to sustaining scientific software. The final and main contribution of the report is a summary of the
 discussions, future steps, and future organization for a set of self-organized
@@
 stakeholders; identifying principles for sustainable software engineering
-design; reaching out to research software organizations as a global effort;
+design; reaching out to research software organizations around the world; and
 building communities for software sustainability. For each group, we include a
@@
 
+\newpage
 \appendix

### 2016-02-06 WSSSPE3/report/WSSSPE3_review.tex
final change of author order to recognize lots of editing work from SC Choi

@@
 \author{Daniel S.\ Katz$^{(1)}$,
-Kyle E.\ Niemeyer$^{(2)}$,
-Sou-Cheng T.\ Choi$^{(3)}$,
+Sou-Cheng T.\ Choi$^{(2)}$,
+Kyle E.\ Niemeyer$^{(3)}$,
 James Hetherington$^{(4)}$,
@@
 %
-\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
+\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
+%
+\thanks{{}$^{(3)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
 Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
-%
-\thanks{{}$^{(3)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
 %

### 2016-02-06 WSSSPE3/report/wg_appendix_software_credit.tex
added missing possessive s

@@
 
-\item Mozilla Science Lab' recently introduced Contributorship Badges for
+\item Mozilla Science Lab's recently introduced Contributorship Badges for
 Science~\cite{Mozilla_badges}, which provide a badge---associated with an

### 2016-05-16 WSSSPE3/best_practices/main.tex
White paper on best practices

--- 2016-02-06 WSSSPE3/report/WSSSPE3_review.tex
@@
 %
-
-\begin{abstract}
-This technical report records and discusses the Third Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE3). The report includes a
-description of the keynote presentation of the workshop, which served as an overview of sustainable
-scientific software. It also summarizes a set of lightning
-talks in which speakers highlighted to-the-point lessons and challenges
-pertaining to sustaining scientific software. The final and main contribution of the report is a summary of the
-discussions, future steps, and future organization for a set of self-organized
-working groups on topics including developing pathways to funding scientific
-software; constructing useful common metrics for crediting software
-stakeholders; identifying principles for sustainable software engineering
-design; reaching out to research software organizations around the world; and
-building communities for software sustainability. For each group, we include a
-point of contact and a landing page that can be used by those who want to join
-that group's future activities. The main challenge left by the workshop is to
-see if the groups will execute these activities that they have scheduled, and
-how the WSSSPE community can encourage this to happen.
-
-\end{abstract}
-
-\maketitle
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction} \label{sec:intro}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The Third Workshop on Sustainable Software for Science: Practice and Experiences
-(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
-held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
-WSSSPE series are
-WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with SC13;
-WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
-focused workshop organized jointly with the SciPy
-conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
-WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
-held in conjunction with SC14; and
-WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
-focused workshop organized again jointly with
-SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
-
-Progress in scientific research is dependent on the quality and accessibility of
-software at all levels. Hence it is critical to address challenges related to
-development, deployment, maintenance, and overall sustainability of reusable
-software as well as education around software practices. These challenges can be
-technological, policy based, organizational, and educational; and are of
-interest to developers (the software community), users (science disciplines),
-software-engineering researchers, and researchers studying the conduct of
-science (science of team science, science of organizations, science of science
-and innovation policy, and social science communities). The WSSSPE1 workshop
-engaged a broad scientific community to identify challenges and best practices
-in areas of interest to creating sustainable scientific software. WSSSPE2
-invited the community to propose and discuss specific mechanisms to move towards
-an imagined future for software development and usage in science and
-engineering. But WSSSPE2 did not have a good way to enact those mechanisms, or to
-encourage the attendees to follow through on their intentions.
-
-The WSSSPE3 workshop included multiple mechanisms for participation and
-encouraged team building around solutions. WSSSPE3 strongly encouraged participation
-of early-career scientists, postdoctoral researchers, graduate students,  
-early-stage researchers, and those from underrepresented groups,
-with funds provided to the conference organizers by the Moore Foundation, the
-National Science Foundation (NSF), and the Software Sustainability Institute (SSI) to
-support the travel of potential participants who would not otherwise be able to
-attend the workshop. These
-funds allowed 16 additional people to attend and participate.
-
-WSSSPE3 also included two professional
-event organizers/facilitators from Knowinnovation who helped the organizing committee members
-plan the workshop agenda, and during the workshop, they actively engaged
-participants with variously
-tools, activities, and reminders.
-
-This report is based on collaborative notes taken during the workshop, which
-were linked from the GitHub issues that represented the potential and actual
-working
-groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
-Overall, the report discusses the organization work done before the workshop
-(\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
-lightning talks (\S\ref{sec:lightning}). The report also gives summaries of
-action plans proposed by the working groups (\S\ref{sec:WGs}), and some
-conclusions (\S\ref{sec:conclusions}). Lists of the organizing committee
-(Appendix~\ref{sec:orgcom}), the registered attendees
-(Appendix~\ref{sec:attendees}), and the travel award recipients
-(Appendix~\ref{sec:awardees}) are compiled. Finally, the report includes longer
-descriptions of the activities that occurred in each of the working groups that
-made substantial progress
-(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Calls for Participation} \label{sec:preworkshop}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting
-a process to make progress in sustainable software, as the calls for
-participation said:
-
-\begin{quote} The WSSSPE1 workshop engaged the broad scientific community to
-identify challenges and best practices in areas relevant to sustainable
-scientific software. WSSSPE2 invited the community to propose and discuss
-specific mechanisms to move towards an imagined future practice of software
-development and usage in science and engineering. WSSSPE3 will organize
-self-directed teams that will collaborate prior to and during the workshop to
-create vision documents, proposals, papers, and action plans that will help the
-scientific software community produce software that is more sustainable,
-including developing sustainable career paths for community members. These teams
-are intended to lead into working groups that will be active after the workshop,
-if appropriate, working collaboratively to achieve their goals, and seeking
-funding to do so if needed. \end{quote}
-
-The first call for participation requested lightning talks, where each author
-could make a brief statement about work that either had been done or was needed,
-with the goal of contributing to the discussion of one or more working groups.
-There were~24 lightning talks submitted; after a peer-review process, 16 
-were accepted, as discussed further in Section~\ref{sec:lightning}.
-
-The first call also discussed the potential action topics that came out of
-WSSSPE2, and requested additional suggestions. The combination of existing and
-new topics led to the following 18 potential topics that were advertised in the
-subsequent calls for participation:
-%
-%\begin{quote}
-\begin{itemize}
-\renewcommand{\labelenumi}{\textbf{\theenumi}.}
-\setlength{\rightmargin}{1em}
-
-\item Development and Community
-\begin{itemize}
-\item Writing a white paper/review paper about best practices in developing
-sustainable software
-\item Documenting successful models for funding specialist expertise in software
-collaborations
-\item Creating and curating catalogs for software tools that aid sustainability
-(perhaps categorized by domain, programming languages, architectures, and/or
-functions, e.g., for code testing, documentation)
-\item Documenting case studies for academia/industry interaction
-\item Determining effective strategies for refactoring/improving legacy
-scientific software
-\item Determining principles for engineering design for sustainable software
-\item Create a set of guidance giving examples of specific metrics for the
-success of scientific software in use, why they are chosen, what they are
-useful to measure, and any challenges/pitfalls; then publish this as a white
-paper
-\end{itemize}
-
-\item Training
-\begin{itemize}
-\item Writing a white paper on training for developing sustainable software, and
-coordinating multiple ongoing training-oriented projects
-\item Developing curriculum for software sustainability, and ideas about where
-such curriculum would be presented, such as a summer training institute
-\end{itemize}
-
-\item Credit
-\begin{itemize}
-\item Hacking the credit and citation ecosystem (making it work, or work better,
-for software)
-\item Developing a taxonomy of contributorship/guidelines for including software
-contributions in tenure review
-\item Documenting case studies of receiving credit for software contributions
-\item Developing a system of awards and recognitions to encourage sustainable
-software
-\end{itemize}
-
-\item Publishing
-\begin{itemize}
-\item Developing a categorization of journals that publish software papers
-(building on existing work), and case studies of alternative publishing
-mechanisms that have been shown to improve software discoverability/reuse, e.g.,
-popular blogs/websites
-\item Determining what journals that publish software paper should provide to
-their reviewers (e.g., guidelines, mechanisms, metadata standards)
-\end{itemize}
-
-\item Reproducibility and Testing
-\begin{itemize}
-\item Building a toolkit that could allow conference organizers to easily add a
-reproducibility track
-\item Documenting best practices for code testing and code review
-\end{itemize}
-
-\item Documentation
-\begin{itemize}
-\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the
-community to easily find up-to-date information on a WSSSPE topic (e.g.,
-software credit, scientific software metrics, testing scientific software).
-\end{itemize}
-
-\end{itemize}
-%\end{quote}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote}\label{sec:keynote}
-%\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
-Department of Astronomy, University of Illinois, titled \emph{Why Sustain
-Scientific Software?}. Turk is a prolific scientific software practitioner and
-has extensive experiences working on large collaborative projects employing
-modern computing tools \cite{2011ApJS..192....9T}. He also co-organizes and
-champions WSSSPE events.
-
-In his keynote address, Turk recapped the course of development of WSSSPE
-workshops over the past few years, alongside his career development from a
-postdoc to an academic. The first WSSSPE workshop was at the Supercomputing
-conference (SC13) in 2013, but he observed that the notion of sustainable
-scientific software drew in an audience beyond supercomputing. In the following
-year, WSSSPE1.1 at SciPy had speakers talking about how software has been
-sustained inside the scientific Python community. WSSSPE2 at SC14 had breakout
-group discussions coming up with actionable items, and WSSSPE2.1 at SciPy 2015
-was similar. Turk noted the different atmosphere of the surrounding large
-conferences, despite similar WSSSPE participants.
-
-WSSSPE3 left the traditional Supercomputing Conference environment this year,
-and in Turk's words, this change spoke to the fact that scientific software
-comes from many different types of inquiries, deployment, strategies for
-maintenance, users, and ways of measuring the value of a piece of software. It
-appeared to Turk that the supercomputing community generally adopts some
-top-down approaches, whereas the SciPy community more often than not uses more
-bottom-up systems. The essential messages perceived were also often bipolar: the
-supercomputing community thinks that software is getting harder, with exascale
-computing and optimization issues in mind; but the SciPy community thinks that
-software is becoming better, with emerging tools such as Jupyter and
-productivity packages for research workflows. Admitting such comparisons are
-somewhat unfair generalizations, Turk reminded the audience that the different
-approaches bring different types of ideas to the table, and he welcomed WSSSPE3
-being conducted outside existing preconceptions.
-
-Returning to the topic of his talk, Turk invited the audience to picture
-scientific software as a flower on a landscape under the Sun, which may
-represent a number of measurable factors such as number of citations; growth of
-a community and number of contributors; amount of funding; prestigious prizes
-awarded; stability of the community in terms of leadership transitions, serving
-community needs, not breaking test suites, and performance on new architectures.
-But all these metrics are strictly speaking \emph{proxies} for the values and
-the impact scientific software bears. What we can measure does not give us
-direct insight---it just gives us proxies of insight.
-
-Turk then moved onto various different definitions of sustainability. His
-favorite one was ``keeping up with bug reports,'' where even if no new features
-were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was ``adding of new features,'' or ``maintaining
-the software for a long period of time'' such as the cases of \TeX\ or \LaTeX\
-with community help. A notion Turk heard often at supercomputing conferences was
-that sustainable software ``continues to work on new architectures.'' Yet
-another metric was ``people continuing to be able to learn how to use and apply
-the software.'' A funder Turk heard talked about sustainability as ``continuing
-to get funded.'' Turk also recalled that Greg Wilson, among others, said in
-WSSSPE1.1 that his view of sustainable software was software that ``continued to
-give the same results over time.'' A last measure of sustainability Turk
-presented was ``the ability to transition between different people developing
-and using a piece of software.''
-
-At WSSSPE1, several models were presented for ensuring sustainability. Turk
-considered that a familiar one was a funded piece of software where an external
-agency provided funds to a group who are not necessarily exclusively working on
-and developing the software, keeps it going, and provides it to the scientific
-community. The model of productized software, in which a piece of software has
-grown to the point that research groups or people are willing to support it with
-some amount of funding, for instance, a subscription to use cloud services that
-deploy a piece of software, or purchase of a piece of software. A final model
-Turk felt conflicted about is a volunteer model that is traditional
-old-school---not modern-day open source---development.
-
-Turk discussed whether productizing scientific software was synonymous with
-being sustainable and self-sufficient. He thought it was not necessarily the
-case and furthermore, it could lead to a divergence of interests between users
-and developers.
-
-Turk reminded the audience that the volunteer model means unpaid labor. On this
-note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
-the open source software
-community\footnote{\url{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}}.
-Oftentimes, a person funded to work full time on a scientific project can spend
-a small amount of time for working on a piece of software necessary for that
-project. However, researchers' abilities to participate in that volunteer
-community are not always the same and may not always be aligned with their
-research projects. From Turk's experience, we cannot always rely on unpaid labor
-and volunteer time to sustain a piece of software---this came down to the
-notions of the top-down and the bottom-up approaches, i.e., the funded versus
-the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
-projects can be just as large-scale as a top-down software development project.
-
-Turk said that sustaining scientific software really meant to him conducting
-scientific inquiries, often by some specific software, and sustaining the people
-we care about, our careers, and the future of our fields. According to Turk, we
-all have an invested stake in sustaining scientific software. Hence, having
-``sustained'' projects can suffocate new projects, so we need to make sure we
-don't cause novel ideas and packages to suffer at the hands of the status quo.
-
-Turk talked about possible reasons why we want to sustain scientific software:
-devotion to science and interests in pursuing the next stage of research; fun
-and creative thrill in writing codes and papers; usefulness with measurable
-impacts, for example, LINPACK and HDF groups providing data storage to
-satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
-his wishlist of questions to be answered in the future:
-%
-\begin{itemize}
-
-\item How do we ship a product on time when dealing with a mix of funding models
-and motivations especially when we rely on volunteers?
-
-\item How do we know when it is time to end some software and move on? For
-example, should we stop sustaining Python and switch to Julia and Javascript?
-
-\item How can productized software balance its future versus its past, or the
-emerging needs of the customers versus the existing needs of the development
-community?
-
-\item How can we help avoid burnout and retain the joy in the communities?
-
-\item How can we reduce systemic bias, which goes back to Dryden's blog post
-especially on how ethics of unpaid labor disproportionately affect
-underrepresented communities?
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Lightning Talks} \label{sec:lightning}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{comment}
-\note{
-\href{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}{Slides.}}
-\end{comment}
-
-The lightning talks were intended to give an opportunity for attendees to
-quickly highlight an important issue or a potential solution.
-%
-\begin{enumerate}
-\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
-Facilitating Research Training While Supporting Scientific Research}} Benjamin
-Tovar presented a case study of ``The Cooperative Computing Lab'' at the
-University of Notre Dame, which is a small group of individuals whose main tasks
-are collaborating with people that have large-scale computing problems,
-operating various parallel computer systems, conducting computer science
-research, and developing open source software. One of the main challenges they
-face is finding a balance between flexibility/training and stability/quality.
-Their current solution for ensuring the latter was to add a software engineer
-(the presenter) to the existing team of faculty and students, who now also
-serves as a ``spring'' between flexibility and stability.
-
-\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
-Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
-Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
-Sustainability}} The concept of sustainability has become a topic of interest in
-the field of computing, which is evidenced by the increase in the number of
-events that focus on the topic. Nevertheless, it is not well understood yet.
-Birgit Penzenstadler argued that we often define sustainability too narrowly.
-Instead, sustainability at its heart is a systemic concept and must be viewed
-from a range of different dimensions including environmental, economic,
-individual, social, and technical. She introduced the Karlskrona Manifesto on
-Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
-of related work on the topic of sustainability into a set of (mis-)perceptions
-and principles. The manifesto does not proclaim that there is an easy,
-one-size-fits-all solution around the corner, but rather points out that
-sustainability is a ``wicked problem'' and is often misunderstood. Due to these
-misperceptions, even though sustainability's importance is increasingly
-recognized, many software systems are unsustainable. Even more alarming is that
-most software systems' broader impacts on sustainability are unknown. To change
-this, the Karlskrona Manifesto proposes nine principles and commitments. These
-commitments are not dogmatic laws, but rather commitments to rethink, to move
-beyond the silo mentality, and to analyze in more depth. As such, they do not
-restrict, but rather open up a space for discussion.
-
-\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
-and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
-to Improve Productivity of Legacy Tools}} Abani Patra presented an example of
-how the community using Titan2D, a geoflow simulation software, increased the
-productivity of their tools by improving both code and data layout. The main
-obstacles in this change were the non-existence of a common version control
-system for the source code, coupled with multiple versions of the same code
-base, the fixed format of input files, that many input values were set as
-compilation flags, and that the internal data layout was not suitable for modern
-technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
-developers included reinforcing the code structure using multiple layers of
-Python and C++ interfaces, and a redesign of the data layout to be more suitable
-for modern CPUs and accelerators.
-
-\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
-Thaney: \textit{Collaborative Software Development as Sustainable Software:
-Lessons from Open Source}} Abigail Cabunoc Mayes combines two properties of open
-source that, together, create a suitable habitat for sustainable software. The
-first of the two properties, public, does not only mean public code. It also
-includes public discussions, a public process of including contributions, and an
-open license. The second property, participatory, stresses the importance of
-reaching out to the community and helping potential new members by providing
-better documentation and learning experiences, like code review and examples of
-good first bug reports. Together, Abigail concluded, these two properties not
-only lead to higher quality, reusability, and ease of understanding, but also
-eventually, to sustainability.
-
-\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
-through Best Practices in Open Source Software: Computational Infrastructure for
-Geodynamics}} Lorraine Hwang presented experiences with the Computational
-Infrastructure for Geodynamics (CIG), a community software with a worldwide user
-base. Like others, their main goals include high usability, sustainability, and
-reproducibility. As a means to achieve these goals, various communication
-channels have been developed, such as mailing lists, wikis, workshops,
-hackathons, tutorials, and webinars. In order to contribute to the
-infrastructure, codes must adhere to specified minimum standards with the desire
-that all codes are working toward target standards.
-%, but ideals are also given to encourage reaching the aforementioned goals. 
-These include, e.g., the use of version control, certain coding styles, the
-presence and nature of code tests and documentation, and certain user workflows.
-
-\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
-in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
-Hwang mentioned a variety of ways to cite efforts within the framework, including
-science papers, code papers, user manuals, and the CIG website. An analysis of
-the resulting citations showed that $80\%$ of papers that use CIG codes mention
-the code name, and about the same number includes a citation. Only about $20\%$
-acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
-cite codes (including non-CIG codes), and only about one eighth specify the
-version used. Compared to other codes, CIG seems to be much better cited. In
-part, this is attributed to the fact that CIG requires that donated software
-provide a citable paper specified in the User Manual. The project is working on
-tools and methods to generate attribution information automatically.
-%This,
-%in part, is attributed to suggested citation forms communicated to the
-%community, and runtime tools that help collecting a relevant citation list
-%automatically.
-
-\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
-Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
-Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
-located at Notre Dame and Chicago, that links the high energy physics effort to
-other disciplines such as biology, astrophysics, and digital curation. It includes
-physicists, digital librarians, as well as computer scientists, and aims to
-achieve some commonality across disciplines. Examples are meta-data descriptions
-of archived data, computational descriptions, descriptions of how data was
-processed, questions such as whether computation replication can be automated,
-and what the impact of access policies on the preservation infrastructure is.
-One of the products of this effort is a suite of tools that deals with this
-preservation, and the questions was posed whether that software itself is
-sustainable. Points that were brought up included the need for a user community
-depending on a given software, and the need to provide added value for its
-users. An important method to achieve this was to work with the user community
-from the start, and to budget that way. For the specific example of the
-preservation software, this means that besides adding value to the community, it
-needs to be transparent to their workflows, i.e., not requiring additional
-effort to preserve ``research objects.''
-
-\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
-Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
-Susheel Varma: \textit{An update from UK Research Software Engineers}} James
-Hetherington started by listing some of the problems research software faces,
-which include poor standard of verification and low levels of reuse. For a long
-time, technical solutions to such technical problems were focus of the eResearch
-community, including research software distributions, grids, middleware and
-workflows. Some limited adoption can be seen today in research communities, but
-the main problems have not been solved to a sufficient level. Hetherington
-hypothesized that instead of technical solutions, social innovation is needed: a
-new role in the academic system focused on research software that combines the
-best parts of a craftsperson and a scholar. However, social innovation in
-centuries-old institutions is hard. Alternatives to such a new role would have
-to include rewards for good research software, recognition of software as
-academic output, and rejecting submissions based on irreproducible computational
-results. Some advantages of research software engineering (RSE) groups include
-the possibility of training in reproducible computational research, providing
-collaborations for researchers who do not want to serve as programmers, and
-creating synergies with research computing platforms. The success of such a
-group would be measured by the output and quality of the research software.
-Members could be part of Research Computing or faculty. They would not be
-independent researchers, but would have to have a research background. An
-attempt to form a community of RSE groups within the UK has been underway for
-several years, including funding from the UK Research Council. An open question
-is whether this approach can be adopted in other countries, including the~USA.
-
-\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
-the User into Building Sustainable Software for Science}} Dan Gunter's main
-question was, ``What is needed to develop sustainable software?''. Beyond the
-usual suspects of funding, proficient developers, good design, and software
-engineering practices, Dan placed the users. He explained this using a
-traditional software-development model starting from gathering requirements, and
-reaching release through a design, development, and testing phase. The main
-deficiency with this approach was pointed out to be the too-late interaction
-with users. Instead, an alternative approach was proposed that, at first, skips
-the development phase and repeatedly goes through requirement gathering, design
-and user interaction/learning phases, and only eventually starts development
-once an agreement is reached, leading to an increased user satisfaction, higher
-adoption, and eventually to sustained software.
-
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
-Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
-Platform for Predictive Biology: Lessons Learned on the KBase Project}} 
-Dan Gunter presented experiences and lessons learned as part of KBase, an
-open software and data platform for addressing the grand challenge of systems
-biology: predicting and designing biological function. KBase is a unified system
-that integrates data and analytical tools for comparative functional genomics of
-microbes, plants, and their communities. However, it is also a collaborative
-environment for sharing methods and results, and placing those results in the
-context of knowledge in the field. Being a large, multi-institutional project,
-one of the big challenges is to agree on standards to enable a single,
-maintainable system. Working in isolation does not work (anymore) within this
-field, and the community in the field also does not have standards for software
-engineering. This is contrasted to computer science research, where software
-engineering standards shorten design cycles, leading to more time for highly
-rewarded activities like publishing, performance studies, graduating students,
-or protecting ideas before publication. Instead of this more traditional
-approach, KBase uses a variation on the ``Scrum'' methodology. After picking
-projects and team members, four to five teams work on projects for about two
-weeks before a one-week evaluation by the executive committee. Based on their
-assessment new teams and projects might be chosen before iteratively restarting
-the ``agile'' development cycle. This process is intentionally open and
-documented.
-
-\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
-Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
-Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
-approach to overcome some hurdles in scientific publishing: disconnects among
-experimental data, research software, and publications. A current effort in the
-Geosciences is the ``Geoscience Paper of the Future,'' which includes four
-elements. First, it forms a modern paper including text, data, and pointers to
-supplementary materials. Second, it is reproducible, including data processing,
-workflow, and visualization tools. Third, it is part of open science, which
-includes being publicly available under open licenses, and providing meta-data.
-Last but not least, it uses digital scholarship elements like persistent
-identifiers for software, data, and authors, and it cites both data and
-software. OntoSoft is a tool for helping with this effort, providing software stewardship
-for the Geosciences. As part of this initiative, a special issue of a journal in
-Geoscience areas is planned to include only Geoscience papers of the future,
-with submissions open until the end of 2015. In addition, training sessions are
-provided to geoscientists to learn best practices in software and data sharing,
-provenance documentation, and scholarly publication.
-
-\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
-Accreditation Framework}} Neil Chue Hong presented a proposal to build a
-scientific software accreditation framework. One of the aims of such a framework
-would be to measure how ``good'' a given piece of software is, and to evaluate
-how this can be effectively measured in the first place. This can be compared to
-the effort of the standardized, easy to read, and understandable nutritional
-labeling of food, which only contain a small set of categories. However, such a
-framework for software would be more difficult due to different existing
-community norms. The challenges such a framework faces include that many
-measurements are subjective, that many metrics are too costly, and that
-self-assessment needs to be encouraged. Possible categories would include
-availability, usability, transferability, among others. Such a framework could
-enable both improvement of specific software, as well as comparisons of similar
-software. An accreditation by such a framework could then be part of software
-management plans, ensuring that software is accessible and reusable throughout
-the proposed project and beyond.
-
-\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
-Support for Sustainable Scientific Software}} Carver argued that for scientific
-software to be truly sustainable, there is a need for developers to use
-appropriate software engineering practices. His experience interacting with
-scientific teams indicates that choosing and tailoring these practices is not a
-trivial exercise. There is a general culture clash between software engineering
-and science that hinders our ability to communicate and choose appropriate
-methods. In addition, many experienced scientific software developers appear to
-be unaware of software engineering practices that may be beneficial to them. The
-most appropriate software engineering practices are those that are lightweight,
-properly tailored, and focus on the key software development problems faced by
-scientists. In order to increase the use of software engineering in science, we
-need more documented success stories. These successes need to be socialized 
-%\choinote{not sure if ``socialize'' can be used in such a way}
-within the scientific community through workshops like the Software Engineering
-for Science workshop
-series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in \textit{Computing in Science and
-Engineering} magazine.
-
-\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
-This talk highlighted an attempt to solve the common problem for open source
-development: it is difficult to collect information about how many people use
-particular software, how often, which version, which parts of the software, or
-on which operating system. Current solutions include surveys, but these have
-high uncertainties. A different approach is based on automatic ``call-backs''
-that collect these information at runtime and send it to a central place for
-analysis. Problems with this approach include obtaining agreement from the user,
-legal issues, increased maintenance (of servers), ethical questions, and also the
-lack of a common infrastructure. Some of these problems are of a social nature
-and have to be solved as such, but the last problem (a missing common
-infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
-which is developing common APIs and a library implementation for common,
-repeating tasks such as obtaining user consent. Results are uploaded not
-directly to project servers, but to neutral third parties that only publish
-aggregated statistics to projects.
-
-\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}} Alice Allen described the Astrophysics Source Code Library
-(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
-software that is indexed by indexing services and can be cited. ASCL offers
-clones of existing infrastructure, provides server space and computing
-resources, shares innovations, and permits moves elsewhere. Users provide a
-domain name, then control and configure their site and use it as intended,
-gather their codes as they wish, share innovations, and protect the provided
-computing environment.
-
-\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
-Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
-Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
-Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
-Developing Community}} The effort behind this talk given by Ray Idaszak started
-from a breakout group at an NSF SI2 workshop in 2015, and centers around
-building a framework for creating metrics for scientific software. This
-framework would improve both the metrics and the software it evaluates, and
-could also serve as a tool for building a community around the idea. With
-especially the last-mentioned idea (building a community) in mind, a software
-``peer review group'' would be created, representing stakeholders who will
-self-review software created by their respective communities, and will
-concurrently develop metrics. The whole project should be community-governed,
-without a single institution overseeing the activities or infrastructure, with
-the hope to evolving community-generated and adopted standards. The generation
-of metrics would be tied to the actual evaluation of software, creating an
-incentive by improving the evaluated software itself during this process. The
-framework code would provide infrastructure for the creation of metrics and
-evaluation, and forums for generation of software success metrics. It would also
-support code reviews of the evaluated software. An open question is whether it
-is possible to fit the resulting metrics in a common template. So far, this is
-still in a design phase, with a white paper at the CSESSP workshop 2015 and this
-talk, but the WSSSPE workshops are seen as a forum for the community to assemble
-and act, and is planned to be used also in the future to build this community
-and framework.
-
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Working Groups} \label{sec:WGs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}
-
-\input{wg_main_best_practices}
-\input{wg_main_funding_spec_expert}
-%\input{wg_main_catalogs}
-\input{wg_main_industry_interaction}
-\input{wg_main_legacy_SW}
-\input{wg_main_eng_design}
-\input{wg_main_metrics}
-\input{wg_main_training}
-\input{wg_main_software_credit}
-\input{wg_main_publishing_SW}
-\input{wg_main_user_community}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions} \label{sec:conclusions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-In WSSSPE3, we attempted to take what we learned from WSSSPE1 and WSSSPE2 in how
-we can collaboratively build a workshop agenda and turn that into an ongoing
-community activity. The success or failure of these efforts will
-only become apparent over time.
-
-The workshop had two components, presentations and working groups. The
-presentations, in the first half day of the workshop, included an inspirational
-keynote and a set of lightning talks. We used lightning talks for two reasons:
-first, the need of some participants to have a slot on the agenda to justify
-their attendance; and second, as a way to get new ideas across to all the
-attendees. We broke with the tradition of requiring the lightning talk
-submitters to self-publish their papers, and instead used a common peer-review
-platform\footnote{\url{http://easychair.org}}, choosing to publish their slides
-on the workshop website instead.
-
-The working groups met for a small part of the first half day and all of the
-second day, with the exception of some short periods for the groups to report
-back to the collected workshop attendees. Each group determined a set of
-activities that the members could do to advance sustainable software in a
-particular area.
-
-The results of these group sessions made it clear that there are many
-interlinked challenges in sustainable software, and that while these challenges
-can be addressed, doing so is difficult because they generally are not the
-full-time job of any of the attendees. As was the case in WSSSPE2 as well, the
-participants were willing to dedicate their time to the groups while they were
-at the meeting, but afterwards, they went back to their (paid) jobs.
-
-We need to determine how to tie the WSSSPE breakout activities to people's jobs,
-so that they feel that continuing them is a higher priority than it is now,
-perhaps through funding the participants, or through funding coordinators for
-each activity, or perhaps by getting the workshop participants to agree to a
-specific schedule of activities during the workshop as we have tried to do in
-WSSSPE3. It remains to be seen, however, if the participants will meet the
-schedules they set.
-
-The overall challenge left to the sustainable software community is perhaps one
-of organization: how to combine the small partial efforts of a large number of
-people to impact a much larger number of people: those who develop and use
-scientific software. While WSSSPE might help focus the actions of the groups,
-something more is needed to incentivize the wider community, which is a
-generalization of the sustainable software problem itself.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgments} \label{sec:acks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Work by Katz was supported by the National Science Foundation while working at
-the Foundation. Any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not necessarily
-reflect the views of the National Science Foundation.
-%
-Choi's work was supported in part by the National Science Foundation research
-grant DMS-1522687 and a WSSSPE3 travel award. She thanks the encouragement and
-discussion with Fred Hickernell.
-%
-Hetherington was funded by the Software Sustainability Institute, RCUK grants
-EP/H043160/1 and EP/N006410/1.
-%
-Work by Gunter was supported by the Office of Science, Office of Biological and
-Environmental Research, of the U.S.\ Department of Energy (DOE) under Award
-Numbers DE-AC02-05CH11231, DE-AC02-06CH11357, DE-AC05-00OR22725, and
-DE-AC02-98CH10886, as part of the DOE Systems Biology Knowledgebase and by the
-Office of Science, Office of Advanced Scientific Computing Research (ASCR) of
-the U.S.\ Department of Energy under Contract Number DE-AC02-05CH11231 as part
-of the Template Interfaces for Agile Parallel Data-Intensive Science (TIGRES)
-project.
-%
-%\todo{feel free to add stuff here}
-
-\newpage
-\appendix
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Organizing Committee}  \label{sec:orgcom}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Do we want email addresses here?}
-
-The following is the list of organizers of WSSSPE3.
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{orgcom}
-\end{longtable}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Attendees}  \label{sec:attendees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Do we want email addresses here?}
-The following is a list of participants registered for the WSSSPE3 workshop.
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{registrants}
-\end{longtable}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Travel Award Recipients}  \label{sec:awardees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Do we want email addresses here?}
-The following is the list of travel award recipients for the WSSSPE3 workshop.
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{awardees}
-\end{longtable}
-}
-
-\input{wg_appendix_best_practices}
-\input{wg_appendix_funding_spec_expert}
-%\input{wg_appendix_catalogs}
-\input{wg_appendix_industry_interaction}
-%\input{wg_appendix_legacy_SW}
-\input{wg_appendix_eng_design}
-\input{wg_appendix_metrics}
-\input{wg_appendix_training}
-\input{wg_appendix_software_credit}
-\input{wg_appendix_publishing_SW}
-\input{wg_appendix_user_community}
-
-\bibliographystyle{vancouver}
-
-\bibliography{wssspe}
-\end{document}

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex

@@
 
-\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)}
+\title[]{Best Practices for Developing Sustainable Scientific Software}
 
-\author{Daniel S.\ Katz$^{(1)}$,
-Sou-Cheng T.\ Choi$^{(2)}$,
-Kyle E.\ Niemeyer$^{(3)}$,
-James Hetherington$^{(4)}$,
-Frank~L\"{o}ffler$^{(5)}$,
-Dan Gunter$^{(6)}$,
-Ray Idaszak$^{(7)}$,
-Steven R.\ Brandt$^{(5)}$,
-Mark A.\ Miller$^{(8)}$,
-Sandra Gesing$^{(9)}$,
-Nick D.\ Jones$^{(10)}$,
-Nic Weber$^{(11)}$,
-Suresh Marru$^{(12)}$,
-Gabrielle Allen$^{(13)}$,
-Birgit~Penzenstadler$^{(14)}$,
-Colin C.\ Venters$^{(15)}$,
-Ethan Davis$^{(16)}$,
-Lorraine Hwang$^{(17)}$,
-Ilian~Todorov$^{(18)}$,
-Abani Patra$^{(19)}$,
-Miguel de Val-Borro$^{(20)}$
+\author{Sandra Gesing$^{(1)}$,
+Abani Patra$^{(2)}$,
+Daniel S.\ Katz$^{(3)}$,
 }
@@
 %
-\thanks{{}$^{(1)}$ \hspace{-1ex}Computation Institute, 
+\thanks{{}$^{(1)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
+%
+\thanks{{}$^{(2)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
+%
+\thanks{{}$^{(3)}$ \hspace{-1ex}Computation Institute, 
 University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; d.katz@ieee.org}
 %
-\thanks{{}$^{(2)}$ NORC at the University of Chicago and Illinois Institute of Technology, Chicago, IL, USA; sctchoi@uchicago.edu}
-%
-\thanks{{}$^{(3)}$ School of Mechanical, Industrial, and Manufacturing Engineering, 
-Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
-%
-\thanks{{}$^{(4)}$ Research Software Development Group, University College London, UK; j.hetherington@ucl.ac.uk}
-%
-\thanks{{}$^{(5)}$ \hspace{-1.5ex}Center for Computation \& Technology, Louisiana State University, Baton Rouge, LA, USA; \{knarf,sbrandt\}@cct.lsu.edu}
-%
-\thanks{{}$^{(6)}$ Lawrence Berkeley National Laboratory, Berkeley, USA; dkgunter@lbl.gov}
-%
-\thanks{{}$^{(7)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org}
-%
-\thanks{{}$^{(8)}$ University of California, San Diego, CA, USA; mmiller@sdsc.edu}
-%
-\thanks{{}$^{(9)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
-%
-\thanks{{}$^{(10)}$ New Zealand eScience Infrastructure (NeSI), University of Auckland, Auckland, NZ; nick.jones@nesi.org.nz}
-%
-\thanks{{}$^{(11)}$ University of Washington, Seattle, WA, USA; nmweber@uw.edu}
-%
-\thanks{{}$^{(12)}$ Indiana University, Bloomington, IN, USA; smarru@iu.edu}
-%
-\thanks{{}$^{(13)}$ National Center for Supercomputing Applications, University of Illinois at Urbana-Champaign, Urbana, IL, USA; gdallen@illinois.edu}
-%
-\thanks{{}$^{(14)}$ College of Computer Engineering \& Computer Science, California State University, Long Beach, CA, USA; birgit.penzenstadler@csulb.edu}
-%
-\thanks{{}$^{(15)}$ School of Computing and Engineering, University of Huddersfield, Huddersfield, UK; c.venters@hud.ac.uk}
-%
-\thanks{{}$^{(16)}$ UCAR Unidata, Boulder, CO, USA; edavis@ucar.edu}
-%
-\thanks{{}$^{(17)}$  University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
-%
-\thanks{{}$^{(18)}$ Science \& Technology Facilities Council, UK; ilian.todorov@stfc.ac.uk}
-%
-\thanks{{}$^{(19)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
-%
-\thanks{{}$^{(20)}$ Department of Astrophysical Sciences, Princeton University, Princeton, NJ, USA; valborro@princeton.edu}
-%
+
+\begin{abstract}
+
+\end{abstract}
+
+\maketitle
+\newpage
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction} \label{sec:intro}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\input{related_work}
+\input{case_studies}
+\input{community_related_practices}
+\input{government_and_management}
+\input{funding}
+\input{metrics}
+\input{tools}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Conclusion} \label{sec:conclusion}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section*{Acknowledgments} \label{sec:acks}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\bibliographystyle{vancouver}
+
+\bibliography{wssspe}
+\end{document}

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex

@@
 \newcommand{\note}[1]{ {\textcolor{blueish}    { ***Note:      #1 }}}
-\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
-\newcommand{\clunenote}[1]{ {\textcolor{orange}    { ***Tom:      #1 }}}
-\newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
-\newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
-\newcommand{\manishnote}[1]{  {\textcolor{violet}     { ***Manish: #1 }}}
-\newcommand{\davidnote}[1]{  {\textcolor{darkgreen}      { ***David: #1 }}}
-\newcommand{\colinnote}[1]{ {\textcolor{red}    {***Colin: #1 }}}
-\newcommand{\choinote}[1]{ {\textcolor{orange}    {***Choi: #1 }}}
-\newcommand{\stevenote}[1]{ {\textcolor{darkblue}    {***Steve: #1 }}}
 
@@
 %
-\thanks{{}$^{(1)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
+\thanks{{}$^{(1)}$ \hspace{-1ex}Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
 %
@@
 %
-\thanks{{}$^{(3)}$ \hspace{-1ex}Computation Institute, 
-University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; d.katz@ieee.org}
+\thanks{{}$^{(3)}$ University of Illinois, IL, USA; d.katz@ieee.org}
 %

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex

@@
 
+Scientific software is an integral part of research in a plethora of areas such as applied natural sciences, science of management and humanities - not only increasing the efficiency to find answers to research questions but also enhancing methods to answer some research questions at all.
+While the importance of developing scientific software is evident in the large number of software package, methods and algorithms applied in science, sustaining it and thus make it easily usable and re-usable for the user communities entails challenges immanent in being part of the highly dynamic environment in research.
+Reviewing multiple past articles and talks at different meetings like WSSSPEx
+\cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} and analyzing and
+promoting sustainable scientific software makes it clear that there are several
+common and recurring ideas that underpin success in developing sustainable
+software. However, outside of a small community, this knowledge is not widely
+shared. This is especially true for the large community of scientists who
+generate most of the software used by scientists but are not primarily software
+developers. In this scenario, a clear and precise exposition of these best
+practices collected from many sources and open collaboration among all in the
+community in a single source (e.g., journal paper, tutorial) that can be widely
+disseminated is necessary and likely to be very valuable.
+
+Core questions that will need to be explored are in reproducibility, 
+reliability,  usability, extensibility, knowledge management, as well as continuity
+(transitions between people). Answers to these questions will guide us on how a software
+tool becomes part of the core workflow of well identified users (stakeholders)
+relating to tool success and hence sustainability.
+%\katznote{prev sentence is complex and awkward} % see fix
+Ideas  that may need to be explored include:
+\begin{itemize}
+\item Requirements engineering to create tools with immediate uptake;
+\item When should software ``die''?
+\item Catering to disruptive developments in environment (e.g., new hardware,
+new methodology);
+\item Dimensions of sustainability: economic, technical, environmental and
+obsolescence. 
+%\katznote{not sure what the prev. comment goes with} %social.
+\end{itemize}
+
+Sustainability requires community participation in code development and/or a
+wide adoption of software. The larger the community base is using a piece of
+software, the better are the funding possibilities and thus also the
+sustainability options. Additionally developer commitment to an application is
+essential and experience shows that software packages with an evangelist
+imposing strong inspiration and discipline are more likely to achieve
+sustainability. While a single person can push sustainability to a certain
+level, open source software also needs sustained commitment from the developer
+community. Such sustained commitments include diverse tasks and roles, which can
+be fulfilled by diverse developers with different knowledge levels. Besides
+developing software and appropriate software management with measures for
+extensibility and scalability of the software, active (expertise) support for
+users via a user forum with a quick turnaround is crucial. The barrier to entry
+for the community as users as well as developers has to be as low as possible.
+
 \input{related_work}

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex

@@
 While the importance of developing scientific software is evident in the large number of software package, methods and algorithms applied in science, sustaining it and thus make it easily usable and re-usable for the user communities entails challenges immanent in being part of the highly dynamic environment in research.
+\note{Need rewording - copied from the report}
 Reviewing multiple past articles and talks at different meetings like WSSSPEx
@@
 software, the better are the funding possibilities and thus also the
-sustainability options. Additionally developer commitment to an application is
+sustainability options. Additionally, developer commitment to an application is
 essential and experience shows that software packages with an evangelist
@@
 for the community as users as well as developers has to be as low as possible.
+\note{End of the copied text}
+
+Sustainability of software can be investigated under many diverse aspects and dimensions: economic, technical, environmental and social are only a few.
+The scope of this paper will focus on scientific environments and their implications and the target group are especially developer and project manager of scientific software.
+To set the stage we picked case studies, which have been successful to some degree and analyze why they have been successful.
+In Section~\ref{sec:comm} are practices described, which are partly resulting of the means available in and demands resulting of a scientific environment.
+The subsequent section focus on the social aspects of governing and managing sustainable software, while~\ref{sec:funding} is concerned with economic aspects.
+To be able to measure sustainability of software, it is important to define metrics as reference and guideline to be able to create such software.
+Additionally, we will present characteristics of available tools (see~\ref{sec:tools}) for managing software but also for analysing and delivering metrics of sustainability.
+Finally, we will conclude with an outlook and suggestion to improve sustainability in a scientific context.
 

### 2016-05-16 WSSSPE3/best_practices/main.tex
Update main.tex
updating affiliation

@@
 %
-\thanks{{}$^{(1)}$ \hspace{-1ex}Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
+\thanks{{}$^{(1)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
 %
@@
 %
-\thanks{{}$^{(3)}$ University of Illinois, IL, USA; d.katz@ieee.org}
+\thanks{{}$^{(3)}$ National Center for Supercomputing Applications (NCSA), Graduate School of Library and Information Science, and Electrical and Computer Engineering, University of Illinois, Urbana-Champaign, IL, USA; d.katz@ieee.org}
 %
@@
 
-Sustainability of software can be investigated under many diverse aspects and dimensions: economic, technical, environmental and social are only a few.
-The scope of this paper will focus on scientific environments and their implications and the target group are especially developer and project manager of scientific software.
-To set the stage we picked case studies, which have been successful to some degree and analyze why they have been successful.
-In Section~\ref{sec:comm} are practices described, which are partly resulting of the means available in and demands resulting of a scientific environment.
-The subsequent section focus on the social aspects of governing and managing sustainable software, while~\ref{sec:funding} is concerned with economic aspects.
-To be able to measure sustainability of software, it is important to define metrics as reference and guideline to be able to create such software.
-Additionally, we will present characteristics of available tools (see~\ref{sec:tools}) for managing software but also for analysing and delivering metrics of sustainability.
-Finally, we will conclude with an outlook and suggestion to improve sustainability in a scientific context.
+Sustainability of software can be investigated under many diverse aspects and dimensions: 
+economic, technical, environmental and social are only a few. The scope of this paper will focus 
+on scientific environments and their implications and the target group are especially developer 
+and project manager of scientific software. To set the stage we picked case studies, which have been 
+successful to some degree and analyze why they have been successful. In Section~\ref{sec:comm} are 
+practices described, which are partly resulting of the means available in and demands resulting of a scientific environment. 
+The subsequent section focus on the social aspects of governing and managing sustainable software, 
+while~\ref{sec:funding} is concerned with economic aspects. To be able to measure sustainability of 
+software, it is important to define metrics as reference and guideline to be able to create such 
+software. Additionally, we will present characteristics of available tools (see~\ref{sec:tools}) 
+for managing software but also for analysing and delivering metrics of sustainability. Finally, 
+we will conclude with an outlook and suggestion to improve sustainability in a scientific context.
 

### 2016-05-19 WSSSPE3/best_practices/main.tex
added Kyle to author list

@@
 Daniel S.\ Katz$^{(3)}$,
+Kyle E.\ Niemeyer$^{(4)}$,
 }
@@
 \thanks{{}$^{(3)}$ National Center for Supercomputing Applications (NCSA), Graduate School of Library and Information Science, and Electrical and Computer Engineering, University of Illinois, Urbana-Champaign, IL, USA; d.katz@ieee.org}
+%
+\thanks{{}$^{(4)}$ School of Mechanical, Industrial, and Manufacturing Engineering, Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
 %

### 2016-05-19 WSSSPE3/best_practices/main.tex
intersted to contribute from apache perspective

@@
 Kyle E.\ Niemeyer$^{(4)}$,
+Suresh Marru$^{(5)}$,
 }
@@
 \thanks{{}$^{(4)}$ School of Mechanical, Industrial, and Manufacturing Engineering, Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
+%
+\thanks{{}$^{(5)}$ Pervasive Technology Institute, Indiana University, USA; smarru@iu.edu}
 %

### 2016-05-19 WSSSPE3/best_practices/main.tex
Update main.tex

@@
 Suresh Marru$^{(5)}$,
+Dan Gunter$^{(6)}$
 }
@@
 \thanks{{}$^{(5)}$ Pervasive Technology Institute, Indiana University, USA; smarru@iu.edu}
+%
+\thanks{{}$^{(6)}$ Lawrence Berkeley National Laboratory, Berkeley, CA, USA; dkgunter@lbl.gov}
 %

### 2016-05-19 WSSSPE3/best_practices/main.tex
catching up

--- 2016-05-19 WSSSPE3/best_practices/main.tex
@@
 Daniel S.\ Katz$^{(3)}$,
-Kyle E.\ Niemeyer$^{(4)}$,
 }
@@
 \thanks{{}$^{(3)}$ National Center for Supercomputing Applications (NCSA), Graduate School of Library and Information Science, and Electrical and Computer Engineering, University of Illinois, Urbana-Champaign, IL, USA; d.katz@ieee.org}
-%
-\thanks{{}$^{(4)}$ School of Mechanical, Industrial, and Manufacturing Engineering, Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
 %
@@
 The subsequent section focus on the social aspects of governing and managing sustainable software, 
-while~\ref{sec:funding} is concerned with economic aspects. To be able to measure sustainability of 
+while Section~\ref{sec:funding} is concerned with economic aspects. To be able to measure sustainability of 
 software, it is important to define metrics as reference and guideline to be able to create such 
-software. Additionally, we will present characteristics of available tools (see~\ref{sec:tools}) 
-for managing software but also for analysing and delivering metrics of sustainability. Finally, 
+software. Additionally, we will present characteristics of available tools (see Section~\ref{sec:tools}) 
+for managing software but also for analyzing and delivering metrics of sustainability. Finally, 
 we will conclude with an outlook and suggestion to improve sustainability in a scientific context.

### 2016-05-19 WSSSPE3/best_practices/main.tex

@@
 Daniel S.\ Katz$^{(3)}$,
+Kyle E.\ Niemeyer$^{(4)}$,
+Suresh Marru$^{(5)}$,
+Dan Gunter$^{(6)}$
 }
@@
 \thanks{{}$^{(3)}$ National Center for Supercomputing Applications (NCSA), Graduate School of Library and Information Science, and Electrical and Computer Engineering, University of Illinois, Urbana-Champaign, IL, USA; d.katz@ieee.org}
+%
+\thanks{{}$^{(4)}$ School of Mechanical, Industrial, and Manufacturing Engineering, Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu}
+%
+\thanks{{}$^{(5)}$ Pervasive Technology Institute, Indiana University, USA; smarru@iu.edu}
+%
+\thanks{{}$^{(6)}$ Lawrence Berkeley National Laboratory, Berkeley, CA, USA; dkgunter@lbl.gov}
 %

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
added WSSSPE4 proposal, based on WSSSPE2

\documentclass[11pt]{article}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{enumitem}

 \geometry{
 letterpaper,
 total={215.9mm,279.4mm},
 left=20mm,
 right=20mm,
 top=20mm,
 bottom=20mm,
 bindingoffset=0mm
 }

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\title{The 4th Workshop on Sustainable Software: Best Practices and Experiences (WSSSPE4)
}
\author{Gabrielle Allen, University of Illinois Urbana--Champaign \\
	Daniel S.\ Katz, University of Illinois Urbana--Champaign \\
	Kyle E.\ Niemeyer, Oregon State University
	}
\date{}                                           % Activate to display a given date or no date

\pagestyle{empty}

\parskip .5cm
\parindent 0cm

\begin{document}
\maketitle
%\section{}
%\subsection{}

{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment and maintenance of reusable software for academic research and education. These challenges include the software development process, the support and maintenance of software, governance and business models particularly for sustainability, the role of software in building science communities, the need for software to lead to science which is verifiable and reproducible, policy issues such as how to measure software impact, choose licensing, or assign software credit, and the education and nurturing of early stage researchers in scientific software. 
Based on a successful WSSSPE 1 meeting at SC13 in Denver in November 2013, the organizers expect over 60 submitted papers and around 150 attendees at the WSSSPE 2 meeting that is anticipated to be held at SC14 in New Orleans in November 2014 (a proposal for a SC14 workshop has been submitted).  As with WSSSPE 1, WSSSPE 2 will provide an open and dynamic forum for the community to discuss experiences and approaches in the areas listed above. This proposal requests funds to support attendance at WSSSPE 2 for US based researchers, particularly targeted at encouraging attendance by early stage researchers with attention to diversity.

{\bf \noindent Intellectual Merit:} The participants will be selected using a formal application procedure, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g. students, postdoctoral researchers, early stage faculty). The workshop will include presentations, panels and discussions driven from short peer reviewed papers that will advance understanding in this new area of attention.

{\bf \noindent Broader Impacts:} All material from the workshop will be openly archived on public popular sites (e.g. Figshare, Slideshare, arXiv), it is anticipated that the workshop will again be Tweeted and Documented live including contributions from external participants. It is expected that the workshop will lead to a special issue of an open journal. The organizers take seriously the responsibility to encourage participation from underrepresented groups as workshop leaders as well as attendees, and NSF funds will be used to encourage such participation from the US community.

\newpage

\section{Goals for WSSSPE4}

The first WSSSPE workshop (see Appendix for a description of WSSSPE 1) was held as part of Supercomputing 2013 in Denver during November 2013. The aim on the workshop was to bring together researchers and practitioners to discuss challenges and issues around the development, deployment and maintenance of reusable software for academic research and education. The next workshop, WSSSPE 2, will continue these discussions, focusing on some of the themes identified during WSSSPE~1. WSSSPE~2 is anticipated to be held as a workshop as part of the International IEEE Supercomputing 2014 conference in New Orleans in November 2014. A proposal has been submitted to SC14 which will be reviewed in the next month.

The goals of WSSSPE 2 are
\begin{itemize}
\item Provide an effective mechanism to further discussions and identify research topics in the development, deployment and maintenance of reusable software
\item Proactively encourage the inclusion of issues related to education and involvement of industry
\item Develop a diverse community of engaged researchers, software developers, industrial representatives, students and educators
\item Make the event as accessible and inclusive as possible, leveraging social media and other technologies
\item Motivate the community in providing outputs including publications and reports
\end{itemize}

\section{Workshop Organization and Planning}

The WSSSPE~2 organizing committee comprises 
\begin{itemize}
\item Co-chairs: Gabrielle Allen (Professor, University of Illinois), Daniel S. Katz (Researcher, University of Chicago)
\item From WSSSPE~1 organizing committee: Manish Parashar (Professor, Rutgers University), Neil Chue Hong (Director, University of Edinburgh), David Proctor (Currently AAAS Fellow, NSF). 
\item New organizing members for WSSSPE 2: Colin Venters (Researcher, University of Leads), Nancy Wilkins-Diehr (Researcher, UCSD), and  Karen Cranston (National Evolutionary Synthesis Center)
\end{itemize}

The organizing committee will in the next months be discussing changes to the program based on last year’s experience, but it is anticipated that the one-day workshop will include
\begin{itemize}
\item Two invited keynote speakers
\item Peer reviewed short papers that will form the basis for discussions or panels
\item Targeted panel discussions
\item Breakout groups to discuss on feed back on identified issues
\end{itemize}

The WSSSPE 1 workshop used some novel features that are described in the arXiv report (http://arxiv.org/format/1311.3523v2 and Appendix) and will probably be repeated for WSSSPE 2. These included
\begin{itemize}
\item Authors submitted papers by placing them on an open archive of their choice and submitting the URL for the document
\item Instead of individual author presentations, the accepted papers where grouped into three panels, with panelists presenting the findings from a set of papers followed by a general discussion with all participants.
\end{itemize}

The tentative agenda for the Workshop is

\begin{tabular}{|l|l|} \hline
9.00 --- 9.15 & Introduction, review of the day's activities \\ \hline
9.15 --- 10.15 & Keynotes \\ \hline
10.15 --- 10.30 & Break \\ \hline
10.30 --- 11.15 & Panel 1 and Audience Discussion \\ \hline
11.15 --- 12.00 & Panel 2 and Audience Discussion \\ \hline
12.00 --- 1.00 & Lunch \\ \hline
1.00 --- 2.30 & Led breakout sessions \\ \hline
2.30 --- 3.30 & Summaries and discussion from breakout discussions \\ \hline
3.30 --- 4.30 & Panel 3 and Audience Discussion \\ \hline
4.30 --- 5.00 & Organize follow on activities  \\ \hline
\end{tabular}

\section{Participant Support}

We are requesting funding from NSF to be able to encourage attendance at the meeting from US participants particularly by underrepresented groups, early stage researchers (students, postdocs, beginning faculty), and to assist in recruiting a diverse set of panelists and keynote speakers.

We are seeking funding early enough to be able to include a description of the opportunity for travel support in our call for papers and outreach material for the event to make sure that lack of funding is not a barrier to submitting a paper, and to make it clear that participation by early stage researchers is welcomed and encouraged.

The participants will be selected using a formal application procedure which will be described and advertised on the workshop web page, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g. students, postdoctoral researchers, early stage faculty).

The WSSSPE 2 organizing committee will review the applications and select the participants to be funded through the award.

\section{Award Administration}

The National Center for Supercomputing Applications (NCSA) at the University of Illinois (U of I) will be the lead fiscal institution for this proposal to the Office of Advanced Cyberinfrastructure (ACI) at the National Science Foundation. NCSA staff will manage the travel for the participants and presenters funded through this proposal. US participants and presenters will arrange their flights through the travel agency of NCSA/University of Illinois, with reimbursed by NCSA/University of Illinois. NCSA/ University of Illinois will cover the hotel and meals for all US and European participants and presenters. The WSSSPE~2 organizing committee will review the applications and select the participants to be funded through this award.

\section{Broader Impacts}

The focus of the WSSSPE~2 workshop on sustainable software for science and engineering is a topic of very broad relevance and interest that cuts across many different disciplines and different categories of people from students, developers, faculty, etc. Areas of interest and focus in the workshop include education and career paths of those involved in the development of software.

This proposal is focused primarily on encouraging diverse participation at the workshop, particularly from underrepresented groups and early stage researchers. Funds will be used to help facilitate a diverse set of participants, including panel and keynote speakers.

\newpage

\section*{Appendix}

The following is a report on the WSSSPE~1 workshop that is available at the arXiv (http://arxiv.org/abs/1311.3523)

%\newpage
%
%{\bf UNIVERSITY OF ILLINOIS AT URBANA-CHAMPAIGN}
%
%{\bf BUDGET JUSTIFICATION}
%
%{\bf WSSSPE2}
%
%{\bf Principal Investigator: Gabrielle Allen}
%
%\begin{enumerate}[label=\bf \Alph*.,leftmargin=*]
%
%\item {\bf Senior Personnel: } No funds requested.
%
%\item {\bf Other Personnel: } No funds requested.
%\item {\bf Fringe Benefits: }  No funds requested. 
%\item {\bf Equipment: }  No funds requested. 
%\item {\bf Travel: } \$3,000
%
%{\it National travel for the US based organizers to attend the workshop in New Orleans. Funds will be used to contribute towards airfares and/or the two hotel nights needed for the workshop}
%
%\item {\bf Participant Support: } \$20,000
%?{\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans. Estimated costs include airfare, 2 nights hotel and workshop fees.}
%
%\item {\bf Other Direct Costs: } No funds requested. 
%\item {\bf Indirect Costs: } \$1,755
%
%{\it Indirect costs are assessed at a rate of 58.5\% of Modified Total Direct Costs (MTDC). MTDC is direct costs less equipment, tuition remission, and subawards in excess of \$25,000. }
%
%\item {\bf Total Costs: } \$24,755 
%
%{\it Total funds requested by this proposal. }
%
%
%\end{enumerate}
%
%\newpage
%
%{\bf Data Management Plan}
%
%As for WSSSPE1, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms (e.g. Slideshare, ArXiv, Google Docs, Twitter). The WSSSPE web page and wiki will be additionally used.
%
%Reviews of paper submissions will not be open. 
%

\end{document}

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
starting to work through changes from 2 to 4

@@
 
-{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment and maintenance of reusable software for academic research and education.
+{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment, maintenance, and sustainabiity of reusable software for academic research and education.
 These challenges include the software development process, the support and maintenance of software, governance and business models particularly for sustainability, the role of software in building science communities, the need for software to lead to science which is verifiable and reproducible, policy issues such as how to measure software impact, choose licensing, or assign software credit, and the education and nurturing of early stage researchers in scientific software.
-Based on a successful WSSSPE 1 meeting at SC13 in Denver in November 2013, the organizers expect over 60 submitted papers and around 150 attendees at the WSSSPE 2 meeting that is anticipated to be held at SC14 in New Orleans in November 2014 (a proposal for a SC14 workshop has been submitted).
-As with WSSSPE 1, WSSSPE 2 will provide an open and dynamic forum for the community to discuss experiences and approaches in the areas listed above.
-This proposal requests funds to support attendance at WSSSPE 2 for US based researchers, particularly targeted at encouraging attendance by early stage researchers with attention to diversity.
+Based on successful WSSSPE1-WSSSPE3 meetings at SC13, SC14, and NCAR in 2013, 2014, and 2015, the organizers expect over 60 submitted papers and around 75 attendees at the WSSSPE4 meeting that will be held immediately preceding the RSE conference in Manchester, UK, in September 2016.
+As with previous events, WSSSPE4 will provide an open and dynamic forum for the community to discuss experiences and approaches in the areas listed above.
+This proposal requests funds to support attendance at WSSSPE4 for US based researchers, particularly targeted at encouraging attendance by early stage researchers with attention to diversity.
 
@@
 {\bf \noindent Broader Impacts:} All material from the workshop will be openly archived on public popular sites (e.g.
-Figshare, Slideshare, arXiv), it is anticipated that the workshop will again be Tweeted and Documented live including contributions from external participants.
+Figshare, Slideshare, arXiv), it is anticipated that the workshop will again be tweeted and documented live including contributions from external participants.
 It is expected that the workshop will lead to a special issue of an open journal.
@@
 
-%\newpage
-%
-%{\bf UNIVERSITY OF ILLINOIS AT URBANA-CHAMPAIGN}
-%
-%{\bf BUDGET JUSTIFICATION}
-%
-%{\bf WSSSPE2}
-%
-%{\bf Principal Investigator: Gabrielle Allen}
-%
-%\begin{enumerate}[label=\bf \Alph*.,leftmargin=*]
-%
-%\item {\bf Senior Personnel: } No funds requested.
-%
-%\item {\bf Other Personnel: } No funds requested.
-%\item {\bf Fringe Benefits: }  No funds requested. 
-%\item {\bf Equipment: }  No funds requested. 
-%\item {\bf Travel: } \$3,000
-%
-%{\it National travel for the US based organizers to attend the workshop in New Orleans. Funds will be used to contribute towards airfares and/or the two hotel nights needed for the workshop}
-%
-%\item {\bf Participant Support: } \$20,000
-%?{\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans.
+\newpage
+
+{\bf UNIVERSITY OF ILLINOIS AT URBANA-CHAMPAIGN}
+
+{\bf BUDGET JUSTIFICATION}
+
+{\bf WSSSPE2}
+
+{\bf Principal Investigator: Gabrielle Allen}
+
+\begin{enumerate}[label=\bf \Alph*.,leftmargin=*]
+
+\item {\bf Senior Personnel: } No funds requested.
+
+\item {\bf Other Personnel: } No funds requested.
+\item {\bf Fringe Benefits: }  No funds requested. 
+\item {\bf Equipment: }  No funds requested. 
+\item {\bf Travel: } \$3,000
+
+{\it National travel for the US based organizers to attend the workshop in New Orleans. Funds will be used to contribute towards airfares and/or the two hotel nights needed for the workshop}
+
+\item {\bf Participant Support: } \$20,000
+{\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans.
 Estimated costs include airfare, 2 nights hotel and workshop fees.}
-%
-%\item {\bf Other Direct Costs: } No funds requested. 
-%\item {\bf Indirect Costs: } \$1,755
-%
-%{\it Indirect costs are assessed at a rate of 58.5\% of Modified Total Direct Costs (MTDC). MTDC is direct costs less equipment, tuition remission, and subawards in excess of \$25,000. }
-%
-%\item {\bf Total Costs: } \$24,755 
-%
-%{\it Total funds requested by this proposal. }
-%
-%
-%\end{enumerate}
-%
-%\newpage
-%
-%{\bf Data Management Plan}
-%
-%As for WSSSPE1, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms (e.g.
+
+\item {\bf Other Direct Costs: } No funds requested. 
+\item {\bf Indirect Costs: } \$1,755
+
+{\it Indirect costs are assessed at a rate of 58.5\% of Modified Total Direct Costs (MTDC). MTDC is direct costs less equipment, tuition remission, and subawards in excess of \$25,000. }
+
+\item {\bf Total Costs: } \$24,755
+
+{\it Total funds requested by this proposal. }
+
+\end{enumerate}
+
+\newpage
+
+{\bf Data Management Plan}
+
+As for WSSSPE1, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms (e.g.
 Slideshare, ArXiv, Google Docs, Twitter).
 The WSSSPE web page and wiki will be additionally used.
-%
-%Reviews of paper submissions will not be open. 
-%
+
+Reviews of paper submissions will not be open.
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
change in summary and intro

@@
 students, postdoctoral researchers, early stage faculty).
-The workshop will include presentations, panels and discussions driven from short peer reviewed papers that will advance understanding in this new area of attention.
+The workshop will include vision and position papers as well as lightning talks and demos, aimed at improving sustainable scientific software today as well as defining and working towards the its future.
 
-{\bf \noindent Broader Impacts:} All material from the workshop will be openly archived on public popular sites (e.g.
-Figshare, Slideshare, arXiv), it is anticipated that the workshop will again be tweeted and documented live including contributions from external participants.
-It is expected that the workshop will lead to a special issue of an open journal.
+{\bf \noindent Broader Impacts:} WSSSPE4 accepted submissions (except presentations of previously published work) are planned to be published by WSSSPE as a special collection in http://ceur-ws.org/, which is fully indexed.
+Other material from the workshop will be openly archived on public popular sites (e.g.
+Figshare, Slideshare, arXiv).
+It is anticipated that the workshop will again be tweeted and documented live including contributions from external participants.
 The organizers take seriously the responsibility to encourage participation from underrepresented groups as workshop leaders as well as attendees, and NSF funds will be used to encourage such participation from the US community.
@@
 
-The first WSSSPE workshop (see Appendix for a description of WSSSPE 1) was held as part of Supercomputing 2013 in Denver during November 2013.
-The aim on the workshop was to bring together researchers and practitioners to discuss challenges and issues around the development, deployment and maintenance of reusable software for academic research and education.
-The next workshop, WSSSPE 2, will continue these discussions, focusing on some of the themes identified during WSSSPE~1.
-WSSSPE~2 is anticipated to be held as a workshop as part of the International IEEE Supercomputing 2014 conference in New Orleans in November 2014.
-A proposal has been submitted to SC14 which will be reviewed in the next month.
+Progress in scientific research is dependent on the quality and accessibility of research software at all levels.
+It is now critical to address many new challenges related to the development, deployment, maintenance, and sustainability of open-use research software: the software upon which specific research results rely.
+Open-use software means that the software is widely accessible (whether open source, shareware, or commercial).
+Research software means that the choice of software is essential to specific research results; using different software could produce different results.
+In addition, it is essential that scientists, researchers, and students are able to learn and adopt a new set of software-related skills and methodologies.
+Established researchers are already acquiring some of these skills, and in particular, a specialized class of software developers is emerging in academic environments who are an integral and embedded part of successful research teams.
+WSSSPE provides a forum for discussion of these challenges, including both positions and experiences, and a forum for the community to assemble and act.
 
-The goals of WSSSPE 2 are
+This is the fourth major event in the WSSSPE series.
+The first two WSSSPE workshops were held as part of SC in 2013 and 2014.
+These one-day workshops brought together researchers and practitioners to discuss challenges and issues around the development, deployment and maintenance of reusable software for academic research and education, and discussed potential solutions, moving from a presentation to a presentation and discussion format.
+WSSSPE3, held at NCAR in 2015, grew out of the previous events, and focused almost entirely on facilitated discussions, with the aim of starting working groups to make changes in the practice of scientific software development leading to more sustainable software.
+Two of these groups have continued
+working together, leading to the FORCE11 Software Citation Principles and an in-progress paper
+on best practices for sustainable scientific software development.
+
+In WSSSPE4, held over 2 1/2 days in Manchester, UK, we will include elements from the
+previous workshops, with most of the workshop divided into two tracks:
+
+{\bf Track 1 – Building a sustainable future for open-use research software} has the goals of defining a vision of the future of open-use research software, and in the workshop, initiating the activities that are needed to get there.
+The idea of this track is to first think about where we want to be 5 to 10 years from now, without being too concerned with where we are today, and then to determine how we can move to this future.
+
+{\bf Track 2 – Practices \& experiences in sustainable scientific software} has the goal of improving the quality of today’s research software and the experiences of its developers by sharing practices and experiences.
+This track is focused on the current state of scientific software and what we can do to improve it in the short term, starting with where we are today.
+
+Topics of interest include but are not limited to:
 \begin{itemize}
-\item Provide an effective mechanism to further discussions and identify research topics in the development, deployment and maintenance of reusable software
-\item Proactively encourage the inclusion of issues related to education and involvement of industry
-\item Develop a diverse community of engaged researchers, software developers, industrial representatives, students and educators
-\item Make the event as accessible and inclusive as possible, leveraging social media and other technologies
-\item Motivate the community in providing outputs including publications and reports
+\item Development and Community
+\begin{itemize}
+\item Best practices for developing sustainable software
+\item Models for funding specialist expertise in software collaborations
+\item Software tools that aid sustainability
+\item Academia/industry interaction
+\item Refactoring/improving legacy scientific software
+\item Engineering design for sustainable software
+\item Metrics for the success of scientific software
+\item Adaptation of mainstream software practices for scientific software
 \end{itemize}
+\item Professionalization
+\begin{itemize}
+\item Career paths
+\item RSE as a brand
+\item RSE outside of the UK or Europe
+\item Increase incentives in publishing, funding and promotion for better software
+\end{itemize}
+\item Training
+\begin{itemize}
+\item Training for developing sustainable software
+\item Curriculum for software sustainability
+\end{itemize}
+\item Credit
+\begin{itemize}
+\item Making the existing credit and citation ecosystem work better for software
+\item Future credit and citation ecosystem
+\item Software contributions as a part of tenure review
+\item Case studies of receiving credit for software contributions
+\item Awards and recognition that encourage sustainable software
+\end{itemize}
+\item Software publishing
+\begin{itemize}
+\item Journals and alternative venues for publishing software
+\item Review processes for published software
+\end{itemize}
+\item Software discoverability/reuse
+\begin{itemize}
+\item Proposals and case studies
+\end{itemize}
+\item Reproducibility and testing
+\begin{itemize}
+\item Reproducibility in conferences and journals
+\item Best practices for code testing and code review
+\end{itemize}
+\end{itemize}
+
+Joint sessions of the workshop will include a keynote, a set of lightning talks on topics related to either track, and a panel on best practices for scientific software development.
+Tracks 1 and 2 will be held in parallel with each other.
+
+Track 1 will consist of facilitated discussions, both as a single group and as breakouts.  It will also include discussion of Idea papers that presents implementable proposals related to the track.
+Track 1 participants will build on the proposals and ideas in these papers, with the goal of initiating the planning, development, and execution of some of the ideas during the workshop itself.
+Given the magnitude and importance of the task at hand, the WSSSPE4 organizing committee encourages these proposals to be developed on an open, public, and inclusive basis.
+Submitters are invited to present a vision of some aspect of the future of open-use research software, and a plan of activities to gather and organize the resources needed to get there.
+
+\begin{quote}
+Example idea paper topics include:
+\begin{itemize}
+\item Adaptation of industrial software engineering principles into the research software community with a plan to fund the work
+\item Funding and scaling software carpentry style training in advanced topics
+Infrastructure and funding for community maintenance of open use research software
+\item Scaling the SSI beyond the UK
+\item Specific proposals of how to bridging/network the various research software engineering communities in scalable manner without destroying independence and unique foci of each community
+\end{itemize}
+\end{quote}
+
+Track 2 will consist mostly of presentations, with some discussions, including position papers, experience papers, presentations of previously published work, and demos.  The intent of this
+track is for current scientific software developers and those interested in scientific software development to learn from each other improve the current state of the practice in this area.
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
updates throughout

@@
 \usepackage{enumitem}
+\usepackage{url}
+\usepackage{color}
 
@@
 
-{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment, maintenance, and sustainabiity of reusable software for academic research and education.
+{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment, maintenance, and sustainability of reusable software for academic research and education.
 These challenges include the software development process, the support and maintenance of software, governance and business models particularly for sustainability, the role of software in building science communities, the need for software to lead to science which is verifiable and reproducible, policy issues such as how to measure software impact, choose licensing, or assign software credit, and the education and nurturing of early stage researchers in scientific software.
-Based on successful WSSSPE1-WSSSPE3 meetings at SC13, SC14, and NCAR in 2013, 2014, and 2015, the organizers expect over 60 submitted papers and around 75 attendees at the WSSSPE4 meeting that will be held immediately preceding the RSE conference in Manchester, UK, in September 2016.
+Based on successful WSSSPE1-WSSSPE3 meetings at SC13, SC14, and NCAR in 2013, 2014, and 2015, the organizers expect over 60 submitted papers and around 75 attendees at the WSSSPE4 meeting that will be held immediately preceding the RSE (Research Software Engineer) Conference in Manchester, UK, in September 2016.
 As with previous events, WSSSPE4 will provide an open and dynamic forum for the community to discuss experiences and approaches in the areas listed above.
@@
 These one-day workshops brought together researchers and practitioners to discuss challenges and issues around the development, deployment and maintenance of reusable software for academic research and education, and discussed potential solutions, moving from a presentation to a presentation and discussion format.
+Reports from these workshops, and set of extended papers from them, have been published in special collections of the Journal of Open Research Software \cite{WSSSPE1-collection, WSSSPE2-collection, WSSSPE1, WSSSPE2}.
 WSSSPE3, held at NCAR in 2015, grew out of the previous events, and focused almost entirely on facilitated discussions, with the aim of starting working groups to make changes in the practice of scientific software development leading to more sustainable software.
@@
 on best practices for sustainable scientific software development.
+The report from WSSSPE3 has been submitted to the Journal of Open Research Software and is currently available as a preprint on arXiv~\cite{WSSSPE3}.
 
@@
 
+\begin{quote}
 {\bf Track 1 – Building a sustainable future for open-use research software} has the goals of defining a vision of the future of open-use research software, and in the workshop, initiating the activities that are needed to get there.
@@
 This track is focused on the current state of scientific software and what we can do to improve it in the short term, starting with where we are today.
+\end{quote}
 
@@
 
-The WSSSPE~2 organizing committee comprises 
+The WSSSPE4 organizing committee comprises 
 \begin{itemize}
-\item Co-chairs: Gabrielle Allen (Professor, University of Illinois), Daniel S. Katz (Researcher, University of Chicago)
-\item From WSSSPE~1 organizing committee: Manish Parashar (Professor, Rutgers University), Neil Chue Hong (Director, University of Edinburgh), David Proctor (Currently AAAS Fellow, NSF). 
-\item New organizing members for WSSSPE 2: Colin Venters (Researcher, University of Leads), Nancy Wilkins-Diehr (Researcher, UCSD), and  Karen Cranston (National Evolutionary Synthesis Center)
+\item Gabrielle Allen, University of Illinois Urbana-Champaign, USA
+\item Jeffrey Carver, University of Alabama, USA
+\item Sou-Cheng T. Choi, Illinois Institute of Technology, USA
+\item Tom Crick, Cardiff Metropolitan University, UK
+\item Michael R. Crusoe, Common Workflow Language project
+\item Sandra Gesing, University of Notre Dame, USA
+\item Robert Haines, University of Manchester, UK
+\item Michael Heroux, Sandia National Laboratories, USA
+\item Lorraine J. Hwang, University of California, Davis, USA
+\item Daniel S. Katz, University of Illinois Urbana-Champaign, USA
+\item Kyle E. Niemeyer, Oregon State University, USA
+\item Manish Parashar, Rutgers University, USA
+\item Colin C. Venters, University of Huddersfield, UK
 \end{itemize}
 
-The organizing committee will in the next months be discussing changes to the program based on last year’s experience, but it is anticipated that the one-day workshop will include
+The program committee currently has about 60 global members, from academia, industry, and national labs, with more invitations outstanding.
+See the web site (\url{http://wssspe.researchcomputing.org.uk/wssspe4/}) for the specific names.
+
+The tentative agenda for the workshop is
+
+Sept 12 (pm) --- Joint session of both tracks
+
 \begin{itemize}
-\item Two invited keynote speakers
-\item Peer reviewed short papers that will form the basis for discussions or panels
-\item Targeted panel discussions
-\item Breakout groups to discuss on feed back on identified issues
+\item Introduction
+\item Keynote
+\item Lightning talks
+\item Updates on actions and activities from WSSSPE3 working groups
+\item Discussion and planning for the remainder of WSSSPE4
 \end{itemize}
 
-The WSSSPE 1 workshop used some novel features that are described in the arXiv report (http://arxiv.org/format/1311.3523v2 and Appendix) and will probably be repeated for WSSSPE 2. These included
+Sept 13 (all day) and 14 (am) --- Parallel tracks
+
 \begin{itemize}
-\item Authors submitted papers by placing them on an open archive of their choice and submitting the URL for the document
-\item Instead of individual author presentations, the accepted papers where grouped into three panels, with panelists presenting the findings from a set of papers followed by a general discussion with all participants.
+\item Track 1: This will be a set of working sessions with a facilitated discussion, breakout sessions, report backs, and active writing towards the track goal of defining a vision of the future of open-use research software, and a plan of activities that are needed to get there.
+\item Track 2: Presentations of position papers, experience papers, previously published works, and demos; and breakout sessions or unconference sessions.
 \end{itemize}
 
-The tentative agenda for the Workshop is
+Sept 14 (1:30 pm -- 5 pm) --- Joint session of both tracks
 
-\begin{tabular}{|l|l|} \hline
-9.00 --- 9.15 & Introduction, review of the day's activities \\ \hline
-9.15 --- 10.15 & Keynotes \\ \hline
-10.15 --- 10.30 & Break \\ \hline
-10.30 --- 11.15 & Panel 1 and Audience Discussion \\ \hline
-11.15 --- 12.00 & Panel 2 and Audience Discussion \\ \hline
-12.00 --- 1.00 & Lunch \\ \hline
-1.00 --- 2.30 & Led breakout sessions \\ \hline
-2.30 --- 3.30 & Summaries and discussion from breakout discussions \\ \hline
-3.30 --- 4.30 & Panel 3 and Audience Discussion \\ \hline
-4.30 --- 5.00 & Organize follow on activities  \\ \hline
-\end{tabular}
+\begin{itemize}
+\item Panel on best practices
+\item Summary and discussion of each tracks’ progress
+\item Planning for future events
+\end{itemize}
 
@@
 
-We are seeking funding early enough to be able to include a description of the opportunity for travel support in our call for papers and outreach material for the event to make sure that lack of funding is not a barrier to submitting a paper, and to make it clear that participation by early stage researchers is welcomed and encouraged.
+We are currently including a description of the opportunity for travel support in our call for papers and outreach material for the event to make sure that lack of funding is not a barrier to submitting a paper, and to make it clear that participation by early stage researchers is welcomed and encouraged.
 
@@
 
-The WSSSPE 2 organizing committee will review the applications and select the participants to be funded through the award.
+The WSSSPE4 organizing committee will review the applications and select the participants to be funded through the award.
+
+Awarded funding is expected to be reimbursement of allowable expenses (travel, lodging, sustenance) of up to \$1,500 for participants traveling from outside Europe to the UK, up to \$1,000 for participants traveling from Europe outside the UK to the UK, and up to \$500 for participants traveling from within the UK.
 
@@
 US participants and presenters will arrange their flights through the travel agency of NCSA/University of Illinois, with reimbursed by NCSA/University of Illinois.
-NCSA/ University of Illinois will cover the hotel and meals for all US and European participants and presenters.
 The WSSSPE~2 organizing committee will review the applications and select the participants to be funded through this award.
@@
 
-The focus of the WSSSPE~2 workshop on sustainable software for science and engineering is a topic of very broad relevance and interest that cuts across many different disciplines and different categories of people from students, developers, faculty, etc.
+The focus of the WSSSPE4 workshop on sustainable software for science and engineering is a topic of very broad relevance and interest that cuts across many different disciplines and different categories of people from students, developers, faculty, etc.
 Areas of interest and focus in the workshop include education and career paths of those involved in the development of software.
@@
 
+\section{Prior Support}
+
+{\bf Daniel S. Katz}
+was most recently supported by the NSF as a Program Officer under NSF ACI-1239751, made to the University of Chicago for \$999,966 for the period 3/26/2012 to 3/25/2016.
+A large number of publications were produced under the PI's IRD time in this award, including the WSSSPE1 and WSSSPE2 reports~\cite{WSSSPE1, WSSSPE2}.
+In addition, a number of NSF programs and official documents were produced under this award~\cite{NSF_software_vision}.
+
+{\bf Gabrielle D. Allen}
+\textcolor{red}{todo...}
+
+{\bf Kyle E. Neimeyer}
+\textcolor{red}{todo...}
+
 \newpage
 
-\section*{Appendix}
-
-The following is a report on the WSSSPE~1 workshop that is available at the arXiv (http://arxiv.org/abs/1311.3523)
+\bibliographystyle{abbrv}
+\bibliography{wssspe}
 
@@
 
-{\bf WSSSPE2}
+{\bf WSSSPE4}
 
-{\bf Principal Investigator: Gabrielle Allen}
+{\bf Principal Investigators: Daniel Katz, Gabrielle Allen, Kyle Neimeyer}
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
updates

@@
 \item {\bf Equipment: }  No funds requested. 
-\item {\bf Travel: } \$3,000
+\item {\bf Travel: } \$6,000
 
-{\it National travel for the US based organizers to attend the workshop in New Orleans. Funds will be used to contribute towards airfares and/or the two hotel nights needed for the workshop}
+{\it International travel for the U Illinois based organizers to attend the meeting in Manchester, UK.
+Funds will be used to contribute towards airfares and/or the hotel nights needed for the workshop}
 
-\item {\bf Participant Support: } \$20,000
+\item {\bf Participant Support: } \$40,000
 {\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans.
@@
 
-As for WSSSPE1, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms (e.g.
+As for previous WSSSPE meetings, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms and journals (e.g.
 Slideshare, ArXiv, Google Docs, Twitter).

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
fixed Neimeyer -> Niemeyer
fixing budget

@@
 
-\item {\bf Participant Support: } \$40,000
-{\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans.
-Estimated costs include airfare, 2 nights hotel and workshop fees.}
+\item {\bf Participant Support: } \$41,500
+{\it Funds are requested to partially support travel for 25 participants and presenters (at \$1,500 per participant) from US institutions across the country from the US to Manchester, UK, and 3 participants/presenters (at \$1,000 each)  from US institutions who are already in Europe to travel to Manchester, and 2 participants/presenters  (at \$500 each)  from US institutions who are already in the UK to travel to Manchester.
+Reimbursable costs include airfare, hotel, and sustenance.}
 
 \item {\bf Other Direct Costs: } No funds requested. 
-\item {\bf Indirect Costs: } \$1,755
+\item {\bf Indirect Costs: } \$1,524
 
@@
 
-\item {\bf Total Costs: } \$24,755
+\item {\bf Total Costs: } \$49,024
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
Added my funding
and fixed a typo

@@
 NCSA staff will manage the travel for the participants and presenters funded through this proposal.
-US participants and presenters will arrange their flights through the travel agency of NCSA/University of Illinois, with reimbursed by NCSA/University of Illinois.
-The WSSSPE~2 organizing committee will review the applications and select the participants to be funded through this award.
+U.S.
+participants and presenters will arrange their flights through the travel agency of NCSA/University of Illinois, with reimbursed by NCSA/University of Illinois.
+The WSSSPE4 organizing committee will review the applications and select the participants to be funded through this award.
 
@@
 {\bf Gabrielle D. Allen}
-\textcolor{red}{todo...}
+was the PI for two recent workshop awards related to sustainable software.
+NSF CCF-1551592, made to the University of Illinois for \$88,739 for the period 09/15/2015 to 5/31/2017 supported the workshop on Computational Science \& Engineering Software Sustainability and Productivity Challenges (CSSSPE) which was hosted by NITRD and supported by NSF, DOE and DOD and brought together a very diverse number of agency representatives, software developers and software engineers spanning academia, national labs and industry.
+A report from this workshop is imminent.
+Allen was also PI for a previous awards supporting WSSSPE2 and 3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
+The award funded participation at WSSSPE2 and WSSSPE3 by early career researchers and underrepresented groups.
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
fixing margins and list spacing

@@
  total={215.9mm,279.4mm},
- left=20mm,
- right=20mm,
- top=20mm,
- bottom=20mm,
+ left=25.4mm,
+ right=25.4mm,
+ top=25.4mm,
+ bottom=25.4mm,
  bindingoffset=0mm
+ }
+
+ \newenvironment{shortlist}{
+        \vspace*{-0.8em}
+  \begin{itemize}
+  \setlength{\itemsep}{-0.3em}
+}{
+  \end{itemize}
+        \vspace*{-0.6em}
+}
+
+ \newenvironment{shortlist2}{
+        \vspace*{-0.5em}
+  \begin{itemize}
+  \setlength{\itemsep}{-0.1em}
+}{
+  \end{itemize}
+        \vspace*{-0.3em}
+}
+ \newenvironment{shortenum}{
+        \vspace*{-0.8em}
+  \begin{enumerate}
+  \setlength{\itemsep}{-0.3em}
+}{
+  \end{enumerate}
+        \vspace*{-0.6em}
 }
@@
 }
-\author{Gabrielle Allen, University of Illinois Urbana--Champaign \\
-	Daniel S.\ Katz, University of Illinois Urbana--Champaign \\
+\author{Daniel S.\ Katz, University of Illinois Urbana--Champaign \\
+        Gabrielle Allen, University of Illinois Urbana--Champaign \\
 	Kyle E.\ Niemeyer, Oregon State University
@@
 Topics of interest include but are not limited to:
-\begin{itemize}
+\begin{shortlist}
 \item Development and Community
-\begin{itemize}
+\begin{shortlist}
 \item Best practices for developing sustainable software
@@
 \item Adaptation of mainstream software practices for scientific software
-\end{itemize}
+\end{shortlist}
 \item Professionalization
-\begin{itemize}
+\begin{shortlist}
 \item Career paths
@@
 \item Increase incentives in publishing, funding and promotion for better software
-\end{itemize}
+\end{shortlist}
 \item Training
-\begin{itemize}
+\begin{shortlist}
 \item Training for developing sustainable software
 \item Curriculum for software sustainability
-\end{itemize}
+\end{shortlist}
 \item Credit
-\begin{itemize}
+\begin{shortlist}
 \item Making the existing credit and citation ecosystem work better for software
@@
 \item Awards and recognition that encourage sustainable software
-\end{itemize}
+\end{shortlist}
 \item Software publishing
-\begin{itemize}
+\begin{shortlist}
 \item Journals and alternative venues for publishing software
 \item Review processes for published software
-\end{itemize}
+\end{shortlist}
 \item Software discoverability/reuse
-\begin{itemize}
+\begin{shortlist}
 \item Proposals and case studies
-\end{itemize}
+\end{shortlist}
 \item Reproducibility and testing
-\begin{itemize}
+\begin{shortlist}
 \item Reproducibility in conferences and journals
 \item Best practices for code testing and code review
-\end{itemize}
-\end{itemize}
+\end{shortlist}
+\end{shortlist}
 
@@
 Example idea paper topics include:
-\begin{itemize}
+\begin{shortlist}
 \item Adaptation of industrial software engineering principles into the research software community with a plan to fund the work
@@
 \item Specific proposals of how to bridging/network the various research software engineering communities in scalable manner without destroying independence and unique foci of each community
-\end{itemize}
+\end{shortlist}
 \end{quote}
@@
 The WSSSPE4 organizing committee comprises 
-\begin{itemize}
+\begin{shortlist}
 \item Gabrielle Allen, University of Illinois Urbana-Champaign, USA
@@
 \item Colin C. Venters, University of Huddersfield, UK
-\end{itemize}
+\end{shortlist}
 
@@
 
-\begin{itemize}
+\begin{shortlist}
 \item Introduction
@@
 \item Discussion and planning for the remainder of WSSSPE4
-\end{itemize}
+\end{shortlist}
 
@@
 
-\begin{itemize}
+\begin{shortlist}
 \item Track 1: This will be a set of working sessions with a facilitated discussion, breakout sessions, report backs, and active writing towards the track goal of defining a vision of the future of open-use research software, and a plan of activities that are needed to get there.
 \item Track 2: Presentations of position papers, experience papers, previously published works, and demos; and breakout sessions or unconference sessions.
-\end{itemize}
+\end{shortlist}
 
@@
 
-\begin{itemize}
+\begin{shortlist}
 \item Panel on best practices
@@
 \item Planning for future events
-\end{itemize}
+\end{shortlist}
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
added some details for prior NSF support from Kyle

--- 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
@@
 \usepackage{enumitem}
-\usepackage{url}
+\usepackage{url,hyperref}
 \usepackage{color}
@@
 
-{\bf Kyle E. Neimeyer}
+{\bf Kyle E. Niemeyer}
+is currently supported by the NSF under NSF ACI-1535065, ``SI2-SSE: Collaborative Research: An Intelligent and Adaptive Parallel CPU\slash GPU Co-Processing Software Library for Accelerating Reactive-Flow Simulations,'' made to Oregon State University for \$285,487 for the period of 1 Sept.~2015 to 31 Aug.~2018 (estimated).
+
+This award has thus far resulted in one journal article under review~\cite{Niemeyer:2016}, one conference paper~\cite{Niemeyer:2015}, one meeting poster~\cite{SI2-poster:2016}, one software package~\cite{pyJac-v1.01}, and one manuscript in preparation~\cite{Curtis:2016}.
+
+\textbf{Intellectual Merit:} 
+\textbf{Broader Impacts:} 
 \textcolor{red}{todo...}
@@
 
-{\bf Principal Investigators: Daniel Katz, Gabrielle Allen, Kyle Neimeyer}
+{\bf Principal Investigators: Daniel Katz, Gabrielle Allen, Kyle Niemeyer}
 
@@
 \item {\bf Equipment: }  No funds requested. 
-\item {\bf Travel: } \$6,000
+\item {\bf Travel: } \$3,000
 
-{\it International travel for the U Illinois based organizers to attend the meeting in Manchester, UK.
-Funds will be used to contribute towards airfares and/or the hotel nights needed for the workshop}
+{\it National travel for the US based organizers to attend the workshop in New Orleans. Funds will be used to contribute towards airfares and/or the two hotel nights needed for the workshop}
 
-\item {\bf Participant Support: } \$41,500
-{\it Funds are requested to partially support travel for 25 participants and presenters (at \$1,500 per participant) from US institutions across the country from the US to Manchester, UK, and 3 participants/presenters (at \$1,000 each)  from US institutions who are already in Europe to travel to Manchester, and 2 participants/presenters  (at \$500 each)  from US institutions who are already in the UK to travel to Manchester.
-Reimbursable costs include airfare, hotel, and sustenance.}
+\item {\bf Participant Support: } \$20,000
+{\it Funds are requested to support travel for 18 participants and 2 presenters from US institutions across the country to travel to New Orleans.
+Estimated costs include airfare, 2 nights hotel and workshop fees.}
 
 \item {\bf Other Direct Costs: } No funds requested. 
-\item {\bf Indirect Costs: } \$1,524
+\item {\bf Indirect Costs: } \$1,755
 
@@
 
-\item {\bf Total Costs: } \$49,024
+\item {\bf Total Costs: } \$24,755
 
@@
 
-As for previous WSSSPE meetings, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms and journals (e.g.
+As for WSSSPE1, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms (e.g.
 Slideshare, ArXiv, Google Docs, Twitter).

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
adding my IM and BI in my prior work

@@
 was most recently supported by the NSF as a Program Officer under NSF ACI-1239751, made to the University of Chicago for \$999,966 for the period 3/26/2012 to 3/25/2016.
-A large number of publications were produced under the PI's IRD time in this award, including the WSSSPE1 and WSSSPE2 reports~\cite{WSSSPE1, WSSSPE2}.
-In addition, a number of NSF programs and official documents were produced under this award~\cite{NSF_software_vision}.
+{\bf Intellectual Merit:} A large number of publications were produced under the PI's IRD time in this award, including the WSSSPE1 and WSSSPE2 reports~\cite{WSSSPE1, WSSSPE2}.
+These documents and this work have impacted our understanding of the issues related to sustainable scientific software.
+{\bf Broader Impacts:}
+A number of NSF programs (e.g., SI2, CDS\&E) and official documents were produced under this award (e.g., \cite{NSF_software_vision}).
+These programs have added and developed an emphasis on software as infrastructure at NSF, and the software produced under these programs have impacted both education and industry.
 
@@
 A report from this workshop is imminent.
-Allen was also PI for a previous awards supporting WSSSPE2 and 3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
+Allen was also PI for a previous awards supporting WSSSPE2 and WSSSPE3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
 The award funded participation at WSSSPE2 and WSSSPE3 by early career researchers and underrepresented groups.

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
finished out Kyle's prior NSF support; possibly too long

--- 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
@@
 is currently supported by the NSF under NSF ACI-1535065, ``SI2-SSE: Collaborative Research: An Intelligent and Adaptive Parallel CPU\slash GPU Co-Processing Software Library for Accelerating Reactive-Flow Simulations,'' made to Oregon State University for \$285,487 for the period of 1 Sept.~2015 to 31 Aug.~2018 (estimated).
-
+The overall goal of this project is creating a software library to accelerate the solution of chemical kinetics equations in reactive-flow simulations on hybrid computing hardware.
+This will be achieved through the development of software components that include (1) integration algorithms for operation on parallel processors (e.g., vectorized CPUs, GPUs), (2) a stiffness detection metric used to adaptively select the most efficient solver based on local conditions, and (3) a computational cell clustering strategy to group similar spatial locations.
+These will be incorporated into a usable, sustainable software library: SLACKHA (Software Library for Accelerating Chemical Kinetics on Hybrid Architectures).
 This award has thus far resulted in one journal article under review~\cite{Niemeyer:2016}, one conference paper~\cite{Niemeyer:2015}, one meeting poster~\cite{SI2-poster:2016}, one software package~\cite{pyJac-v1.01}, and one manuscript in preparation~\cite{Curtis:2016}.
-
-\textbf{Intellectual Merit:} 
-\textbf{Broader Impacts:} 
-\textcolor{red}{todo...}
+\textbf{Intellectual Merit:} The complexity of detailed chemical models prevents their use in high-fidelity computational fluid dynamics.
+The software library developed through this project will allow studies that are currently impractical due to computational demands; it will also shorten the design cycle of engines and combustors by reducing the time needed for design simulations.
+\textbf{Broader Impacts:} This work will provide important software tools for designers to help create next-generation clean and efficient engine technology, and for researchers to study difficult problems in combustion.
+In addition, the software elements developed in this program are widely applicable to computational modeling problems outside of combustion, including astrophysics, atmospheric chemistry, biochemical networks, and even cardiac electrophysiology.
+%By enabling the implementation of accurate and realistic chemical models into simulations, new concepts and solutions to existing problems may be explored rapidly and at lower cost.
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
some text fixes throughout, and BI addition

@@
 
-\title{The 4th Workshop on Sustainable Software: Best Practices and Experiences (WSSSPE4)
+\title{\vspace{-2em}The 4th Workshop on Sustainable Software: Best Practices and Experiences (WSSSPE4)
 }
@@
 
-{\bf \noindent Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment, maintenance, and sustainability of reusable software for academic research and education.
+\noindent \textbf{Summary:} The WSSSPE4 workshop is part of a community driven effort to address the many new challenges that are evident for the development, deployment, maintenance, and sustainability of reusable software for academic research and education.
 These challenges include the software development process, the support and maintenance of software, governance and business models particularly for sustainability, the role of software in building science communities, the need for software to lead to science which is verifiable and reproducible, policy issues such as how to measure software impact, choose licensing, or assign software credit, and the education and nurturing of early stage researchers in scientific software.
-Based on successful WSSSPE1-WSSSPE3 meetings at SC13, SC14, and NCAR in 2013, 2014, and 2015, the organizers expect over 60 submitted papers and around 75 attendees at the WSSSPE4 meeting that will be held immediately preceding the RSE (Research Software Engineer) Conference in Manchester, UK, in September 2016.
+Based on successful WSSSPE1--WSSSPE3 meetings at SC13, SC14, and NCAR in 2013, 2014, and 2015, the organizers expect over 60 submitted papers and around 75 attendees at the WSSSPE4 meeting that will be held immediately preceding the RSE (Research Software Engineer) Conference in Manchester, UK, in September 2016.
 As with previous events, WSSSPE4 will provide an open and dynamic forum for the community to discuss experiences and approaches in the areas listed above.
@@
 
-{\bf \noindent Intellectual Merit:} The participants will be selected using a formal application procedure, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g.
-students, postdoctoral researchers, early stage faculty).
+\noindent \textbf{Intellectual Merit:} The participants will be selected using a formal application procedure, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g., students, postdoctoral researchers, early stage faculty).
 The workshop will include vision and position papers as well as lightning talks and demos, aimed at improving sustainable scientific software today as well as defining and working towards the its future.
 
-{\bf \noindent Broader Impacts:} WSSSPE4 accepted submissions (except presentations of previously published work) are planned to be published by WSSSPE as a special collection in http://ceur-ws.org/, which is fully indexed.
-Other material from the workshop will be openly archived on public popular sites (e.g.
-Figshare, Slideshare, arXiv).
+\noindent \textbf{Broader Impacts:} WSSSPE4 accepted submissions (except presentations of previously published work) are planned to be published by WSSSPE as a special collection in \url{http://ceur-ws.org/}, which is fully indexed.
+Other material from the workshop will be openly archived on public popular sites (e.g., Figshare, Slideshare, arXiv).
 It is anticipated that the workshop will again be tweeted and documented live including contributions from external participants.
 The organizers take seriously the responsibility to encourage participation from underrepresented groups as workshop leaders as well as attendees, and NSF funds will be used to encourage such participation from the US community.
+Strategies for reaching\slash recruiting such participants will be solicited from members of the Program Committee with experience supporting students, researchers, and research software engineers with such backgrounds.
 
@@
 These one-day workshops brought together researchers and practitioners to discuss challenges and issues around the development, deployment and maintenance of reusable software for academic research and education, and discussed potential solutions, moving from a presentation to a presentation and discussion format.
-Reports from these workshops, and set of extended papers from them, have been published in special collections of the Journal of Open Research Software \cite{WSSSPE1-collection, WSSSPE2-collection, WSSSPE1, WSSSPE2}.
+Reports from these workshops, and set of extended papers from them, have been published in special collections of the \textit{Journal of Open Research Software} \cite{WSSSPE1-collection, WSSSPE2-collection, WSSSPE1, WSSSPE2}.
 WSSSPE3, held at NCAR in 2015, grew out of the previous events, and focused almost entirely on facilitated discussions, with the aim of starting working groups to make changes in the practice of scientific software development leading to more sustainable software.
@@
 on best practices for sustainable scientific software development.
-The report from WSSSPE3 has been submitted to the Journal of Open Research Software and is currently available as a preprint on arXiv~\cite{WSSSPE3}.
+The report from WSSSPE3 has been submitted to the \textit{Journal of Open Research Software} and is currently available as a preprint on arXiv~\cite{WSSSPE3}.
 
@@
 
-The participants will be selected using a formal application procedure which will be described and advertised on the workshop web page, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g.
-students, postdoctoral researchers, early stage faculty).
+The participants will be selected using a formal application procedure which will be described and advertised on the workshop web page, with selection criteria based around (1) the potential of their attendance to contribute towards their own research and educational goals; (2) their potential contribution to the workshop; (3) providing a diverse set of attendees with priority to early stage researchers (e.g., students, postdoctoral researchers, early stage faculty).
 

### 2016-06-10 WSSSPE4_proposal/WSSSPE4_proposal.tex
added 'for Science' to title...
added more to BI section in proposal
added section on assessment; may need more

@@
 
-\title{\vspace{-2em}The 4th Workshop on Sustainable Software: Best Practices and Experiences (WSSSPE4)
+\title{\vspace{-2em}The 4th Workshop on Sustainable Software for Science: Best Practices and Experiences (WSSSPE4)
 }
@@
 \item Panel on best practices
-\item Summary and discussion of each tracks’ progress
+\item Summary and discussion of each tracks' progress
 \item Planning for future events
@@
 
+\section{Assessment}
+
+Reports describing the activities of WSSSPE and WSSSPE2 culminated in peer-reviewed articles~\cite{WSSSPE1,WSSSPE2}, while the report from WSSSPE3 is currently undergoing peer review (and is available via the arXiv)~\cite{WSSSPE3}; all reports are publicly available via publication in an open-access journal.
+A similar report will be prepared at the conclusion of WSSSPE4, and following the practices of past workshops will be written collaboratively and openly on GitHub.
+
+The organizers of WSSSPE4 take seriously the assessment of workshop activities, in particular the progress of each working group where applicable.
+A portion of the first day's session is planned to discuss and evaluate the status\slash success of the activities proposed during WSSSPE3~\cite{WSSSPE3}.
+
 \section{Broader Impacts}
@@
 Funds will be used to help facilitate a diverse set of participants, including panel and keynote speakers.
+Strategies for reaching\slash recruiting such participants will be solicited from members of the Program Committee with experience supporting students, researchers, and research software engineers with such backgrounds.
 

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
various minor changes

@@
 \documentclass[11pt]{article}
-\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
-\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
-%\geometry{landscape}                % Activate for for rotated page geometry
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+
+\usepackage[letterpaper, margin=1in]{geometry}
+
+\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
+%\microtypecontext{spacing=nonfrench}
+\usepackage{xspace}
+
 %\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
@@
 
- \geometry{
- letterpaper,
- total={215.9mm,279.4mm},
- left=25.4mm,
- right=25.4mm,
- top=25.4mm,
- bottom=25.4mm,
- bindingoffset=0mm
- }
+%better printing of numbers
+\usepackage[english]{babel}
+\usepackage{textcomp}
+\usepackage{csquotes}
 
@@
 \begin{quote}
-{\bf Track 1 – Building a sustainable future for open-use research software} has the goals of defining a vision of the future of open-use research software, and in the workshop, initiating the activities that are needed to get there.
+{\bf Track 1: Building a sustainable future for open-use research software} has the goals of defining a vision of the future of open-use research software, and in the workshop, initiating the activities that are needed to get there.
 The idea of this track is to first think about where we want to be 5 to 10 years from now, without being too concerned with where we are today, and then to determine how we can move to this future.
 
-{\bf Track 2 – Practices \& experiences in sustainable scientific software} has the goal of improving the quality of today’s research software and the experiences of its developers by sharing practices and experiences.
+{\bf Track 2: Practices \& experiences in sustainable scientific software} has the goal of improving the quality of today's research software and the experiences of its developers by sharing practices and experiences.
 This track is focused on the current state of scientific software and what we can do to improve it in the short term, starting with where we are today.
@@
 \item Jeffrey Carver, University of Alabama, USA
-\item Sou-Cheng T. Choi, Illinois Institute of Technology, USA
+\item Sou-Cheng T.~Choi, Illinois Institute of Technology, USA
 \item Tom Crick, Cardiff Metropolitan University, UK
-\item Michael R. Crusoe, Common Workflow Language project
+\item Michael R.~Crusoe, Common Workflow Language project
 \item Sandra Gesing, University of Notre Dame, USA
@@
 \item Michael Heroux, Sandia National Laboratories, USA
-\item Lorraine J. Hwang, University of California, Davis, USA
-\item Daniel S. Katz, University of Illinois Urbana-Champaign, USA
-\item Kyle E. Niemeyer, Oregon State University, USA
+\item Lorraine J.~Hwang, University of California, Davis, USA
+\item Daniel S.~Katz, University of Illinois Urbana-Champaign, USA
+\item Kyle E.~Niemeyer, Oregon State University, USA
 \item Manish Parashar, Rutgers University, USA
-\item Colin C. Venters, University of Huddersfield, UK
+\item Colin C.~Venters, University of Huddersfield, UK
 \end{shortlist}
@@
 
-The tentative agenda for the workshop is
+The tentative agenda for the workshop is:
 
@@
 
-Sept 14 (1:30 pm -- 5 pm) --- Joint session of both tracks
+Sept 14 (1:30--5 pm) --- Joint session of both tracks
 
@@
 
-{\bf Daniel S. Katz}
+\textbf{Daniel S.~Katz}
 was most recently supported by the NSF as a Program Officer under NSF ACI-1239751, made to the University of Chicago for \$999,966 for the period 3/26/2012 to 3/25/2016.
@@
 
-{\bf Gabrielle D. Allen}
+\textbf{Gabrielle D.~Allen}
 was the PI for two recent workshop awards related to sustainable software.
@@
 
-{\bf Kyle E. Niemeyer}
+\textbf{Kyle E.~Niemeyer}
 is currently supported by the NSF under NSF ACI-1535065, ``SI2-SSE: Collaborative Research: An Intelligent and Adaptive Parallel CPU\slash GPU Co-Processing Software Library for Accelerating Reactive-Flow Simulations,'' made to Oregon State University for \$285,487 for the period of 1 Sept.~2015 to 31 Aug.~2018 (estimated).
@@
 
-As for previous WSSSPE meetings, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms and journals (e.g.
-Slideshare, ArXiv, Google Docs, Twitter).
-The WSSSPE web page and wiki will be additionally used.
+As for previous WSSSPE meetings, all material (including publications, presentations, meeting minutes, real time discussions) will be open, hosted and archived on existing public platforms and journals (e.g., Slideshare, arXiv, Google Docs, Twitter).
+The WSSSPE web page and wiki at \url{http://wssspe.researchcomputing.org.uk/} will be additionally used to direct interested parties towards material.
 

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
added research product availability to Kyle prior NSF support

@@
 This award has thus far resulted in one journal article under review~\cite{Niemeyer:2016}, one conference paper~\cite{Niemeyer:2015}, one meeting poster~\cite{SI2-poster:2016}, one software package~\cite{pyJac-v1.01}, and one manuscript in preparation~\cite{Curtis:2016}.
+All research products from this project have been\slash will be made available openly: journal articles are either published in open-access journals or made available as preprints on arXiv~\cite{Niemeyer:2016}, conference papers are made available via figShare if not otherwise accessible~\cite{Niemeyer:2015}, and software is released openly on GitHub\slash Zenodo~\cite{pyJac-v1.01}.
 \textbf{Intellectual Merit:} The complexity of detailed chemical models prevents their use in high-fidelity computational fluid dynamics.

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
removed warning

@@
 \usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
-%\microtypecontext{spacing=nonfrench}
+\microtypecontext{spacing=nonfrench}
 \usepackage{xspace}

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
re-enabled page numbers
adding to assessment

@@
 
-\pagestyle{empty}
+%\pagestyle{empty}
 
@@
 
-Reports describing the activities of WSSSPE and WSSSPE2 culminated in peer-reviewed articles~\cite{WSSSPE1,WSSSPE2}, while the report from WSSSPE3 is currently undergoing peer review (and is available via the arXiv)~\cite{WSSSPE3}; all reports are publicly available via publication in an open-access journal.
+Reports describing the activities of WSSSPE and WSSSPE2 culminated in peer-reviewed articles~\cite{WSSSPE1,WSSSPE2}, while the report from WSSSPE3 is currently undergoing peer review (and is available via arXiv)~\cite{WSSSPE3}; all reports are publicly available via publication in an open-access journal.
 A similar report will be prepared at the conclusion of WSSSPE4, and following the practices of past workshops will be written collaboratively and openly on GitHub.
@@
 A portion of the first day's session is planned to discuss and evaluate the status\slash success of the activities proposed during WSSSPE3~\cite{WSSSPE3}.
+
+\textcolor{red}{How will we know if this is successful? - for track 2, survey attendees?  for track 1, see if we have proposals submitted or activities underway six months later?}
 

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
some work in assessment

@@
 
+We will assess the success of WSSSPE4 with two different mechanisms.
+For Track 1, the activities and\slash or deliverables proposed during the workshop (and documented in the report) will be evaluated for progress or completion six months following the workshop.
+This could include submission of proposals or papers, or the creation of other useful products (e.g., tutorials, wiki, forum).
+Track 2 activities will be assessed by surveying attendees following the workshop with questions such as ``What software development practices are you aware of now that you were not aware of prior to WSSSPE4?'' and ``What practices do you intend on implementing in your own software development work following WSSSPE4?''.
+In addition, the submission of a workshop report for peer review within six months will be another metric for success.
+
 \textcolor{red}{How will we know if this is successful? - for track 2, survey attendees?  for track 1, see if we have proposals submitted or activities underway six months later?}
+\textcolor{blue}{I tried fleshing those suggestions about a bit; what about assessment of efforts to get people from underrepresented groups involved?}
 

### 2016-06-11 WSSSPE4_proposal/WSSSPE4_proposal.tex
more work in assessment

@@
 
-The program committee currently has about 60 global members, from academia, industry, and national labs, with more invitations outstanding.
+The program committee currently has about 65 global members, from academia, industry, and national labs, with more invitations outstanding.
 See the web site (\url{http://wssspe.researchcomputing.org.uk/wssspe4/}) for the specific names.
@@
 Reports describing the activities of WSSSPE and WSSSPE2 culminated in peer-reviewed articles~\cite{WSSSPE1,WSSSPE2}, while the report from WSSSPE3 is currently undergoing peer review (and is available via arXiv)~\cite{WSSSPE3}; all reports are publicly available via publication in an open-access journal.
-A similar report will be prepared at the conclusion of WSSSPE4, and following the practices of past workshops will be written collaboratively and openly on GitHub.
+A similar report will be prepared at the conclusion of WSSSPE4, and following the practices of past workshops, will be written collaboratively and openly on GitHub.
 
@@
 Track 2 activities will be assessed by surveying attendees following the workshop with questions such as ``What software development practices are you aware of now that you were not aware of prior to WSSSPE4?'' and ``What practices do you intend on implementing in your own software development work following WSSSPE4?''.
-In addition, the submission of a workshop report for peer review within six months will be another metric for success.
+In addition, the peer review process for a report on the workshop, to be submitted within six months from the end of the conference, will be another metric for success.
 
-\textcolor{red}{How will we know if this is successful? - for track 2, survey attendees?  for track 1, see if we have proposals submitted or activities underway six months later?}
-\textcolor{blue}{I tried fleshing those suggestions about a bit; what about assessment of efforts to get people from underrepresented groups involved?}
+In our final report, we will discuss the demographics of the supported attendees, the compete set of workshop attendees, speakers, breakout group leads, and the leaders of activities that grow out the workshop.
+In all cases, while our ideal would be to have these demographics match the US or global population averages, our goal will be to significantly improve over academic STEM and tech industry norms.
 

### 2016-06-12 WSSSPE4_proposal/WSSSPE4_proposal.tex
Update prior support

@@
 A report from this workshop is imminent.
-Allen was also PI for a previous awards supporting WSSSPE2 and WSSSPE3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
-The award funded participation at WSSSPE2 and WSSSPE3 by early career researchers and underrepresented groups.
+Allen was also PI for a previous award supporting WSSSPE2 and WSSSPE3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
+{\bf Intellectual Merit:} Publications  resulting from the WSSSPE2 and WSSSPE3 workshop include \cite{WSSSPE2-collection, WSSSPE2, WSSSPE3}.
+{\bf Broader Impacts:} The award funded participation at WSSSPE2 and WSSSPE3 by early career researchers and underrepresented groups.
 

### 2016-06-12 WSSSPE4_proposal/WSSSPE4_proposal.tex
edit in Allen prior support

@@
 was most recently supported by the NSF as a Program Officer under NSF ACI-1239751, made to the University of Chicago for \$999,966 for the period 3/26/2012 to 3/25/2016.
-{\bf Intellectual Merit:} A large number of publications were produced under the PI's IRD time in this award, including the WSSSPE1 and WSSSPE2 reports~\cite{WSSSPE1, WSSSPE2}.
+{\bf Intellectual Merit:} A large number of publications were produced under the PI's IRD time in this award, including the WSSSPE1-WSSSPE3 reports~\cite{WSSSPE1, WSSSPE2, WSSSPE3}.
 These documents and this work have impacted our understanding of the issues related to sustainable scientific software.
@@
 Allen was also PI for a previous award supporting WSSSPE2 and WSSSPE3, NSF ACI-1434218, made to the University of Illinois for \$24,758 for the period 5/1/2014 to 4/30/2016.
-{\bf Intellectual Merit:} Publications  resulting from the WSSSPE2 and WSSSPE3 workshop include \cite{WSSSPE2-collection, WSSSPE2, WSSSPE3}.
-{\bf Broader Impacts:} The award funded participation at WSSSPE2 and WSSSPE3 by early career researchers and underrepresented groups.
+{\bf Intellectual Merit:} The workshops helped us gain knowledge about sustainable software and in particular, sustainable software in science, includes examining the roles of formal software engineering, metics, economics, workforce development, and sociology in scientific software development.
+Publications  resulting from the WSSSPE2 and WSSSPE3 workshop include \cite{WSSSPE2-collection, WSSSPE2, WSSSPE3}.
+{\bf Broader Impacts:} The awards funded participation at CSESSP and WSSSPE2 \& WSSSPE3 by early career researchers and underrepresented groups, diversifying the pool of researchers interested and active in working towards sustainable software.
 

### 2016-06-17 WSSSPE3/report/WSSSPE3_review.tex
changing word in abstract
fixing a typo

@@
 \begin{abstract}
-This technical report records and discusses the Third Workshop on Sustainable
+This report records and discusses the Third Workshop on Sustainable
 Software for Science: Practice and Experiences (WSSSPE3). The report includes a
@@
 plan the workshop agenda, and during the workshop, they actively engaged
-participants with variously
+participants with various
 tools, activities, and reminders.

### 2016-08-26 WSSSPE3/report/JORS-reviews-responses.txt
adding JORS reviews, a document for our responses, and making a couple of the easy changes

We thank the reviewers for their insightful comments, which follow, along
with our responses and the changes we've made.

Reviewer A:

Comments to the authors: 
	The paper summarizes the agenda, discussion, and outputs of the WSSSPE3.0
workshop on scientific software sustainability. The paper provides good
insight into the status of that discourse, very relevant to the JoRS venue.
Each section provides useful literature background and identifies people and
groups that readers can follow up with if they are interested in particular
topics. The inclusion of the longer appendices is very useful.  I have two
comments that would improve the article, neither of which I think are
crucial to publication but would be useful.

The paper lays out it's structure at the start, but each section could end
with additional signposts ("having done x, and before doing y, we will
review z"). I think those help the reader know where they are in a long
document.

RESPONSE:

Time has passed since the submission was prepared. I am conscious that my
late review played a role here. However it does create an opportunity to
check in with each of the groups and see if any progress has been made. For
example, it seems as if the effort to create the "best practices" paper has
flagged (at least nothing has been done since November 2015 at the gdoc
linked from the paper. Perhaps ask the conveners/contacts of each working
group if they have any updates?

RESPONSE:

Just a few typos I spotted (very few overall):

p 27 "refunded" should be "re-funded" or "funded again"

RESPONSE: We've changed the word to re-funded.

p2 "Overall, the report discusses the organization work done before the
workshop (§2); the keynote (§3); a series of lightning talks (§4)." needs
some work. Perhaps "Overall, the report discusses the organizational work
done before the workshop (§2); the keynote (§3) and summarizes the
lightning talks presented at the meeting (§4)."

RESPONSE: we've changed this text to
Overall, the report discusses the organization work done before the
workshop (§2), the keynote (§3), and the lightning talks presented
at the meeting (§4).

------------------------------------------------------

------------------------------------------------------
Reviewer B:

Comments to the authors: 
	General Comments:

The authors have provided a dense, lengthy but complete record of the
WSSSPE3 conference.  It follows on the WSSSPE2 report published in JORS, and
is balanced, trustworthy, and valuable. The authors’ limited
editorializing is simultaneously welcome in a historical report and renders
the lengthy document difficult to read or interpret since it lacks a strong
narrative form. To put that a bit more harshly, the manuscript feels like a
network graph with a goal of encompassing all the historical facts around
WSSSPE3. I think this stems from having a dual goal of tracing accurately
what happened at this conference and trying to encode action items and
practical outcomes into an authoritative document about what WSSSPE as an
entity is trying to become. I would recommend that this aspect of the
manuscript could be improved. Most of my concern is centered on Section 5.

Section 5 “Working Groups” beings without any narrative explaining where
these working groups came from and how they might have been culled from a
list of potential/actual working groups. The 10 working groups do not
correlate precisely with the “18 potential topics” on page 3 nor the 6
categories superimposed on the 18 topics. I suppose the working groups could
have been drawn from them though.  So this sets an unanswered question: were
these working groups decided upon by WSSSPE3 conference members or by a
WSSSPE body before the event? (Aside: does a WSSSPE entity exist today? Did
it exist before WSSSPE1? After WSSSPE2?)

RESPONSE:

Before Section 5 the last time the reader was introduced to the working
group model was on page 3, with the rather diffuse text, “The first call
for participation requested lightning talks, where each author could make a
brief statement about work that either had been done or was needed, with the
goal of contributing to the discussion of one or more working groups.” I
think what is lacking is some threading through of where these working
groups came from and how they are the focus of WSSSPE# going forward.

RESPONSE:

It is also distracting that each subsection of Section 5 represents a
subset/working group speaking with that voice, e.g., Section 5.3, “At
WSSSPE3 our group was interested in better understanding successful pathways
for scientific software to “transition” from grant-funded research
projects to industry sponsorship.” I think it makes clear sense for
individual appendices to speak with a “group” voice, but that the main
document should speak with a single authoritative voice.

RESPONSE:

To fix my perceived problem with Section 5, some background would need to be
added and some of the summary material about working groups that is
presently in the Conclusions section could be moved to introduce the working
groups better.

RESPONSE:

I just have one more comment on the narrative and analysis in this
summarizing manuscript. I suspect given the years WSSSPE has spent digesting
these topics and broadening and maturing its participants that it would be
worthwhile analyzing the following problem posed in the Conclusions a little
bit more in the current article: “We need to determine how to tie the
WSSSPE breakout activities to people’s jobs, so that they feel that
continuing them is a higher priority than it is now, perhaps through funding
the participants, or through funding coordinators for each activity, or
perhaps by getting the workshop participants to agree to a specific schedule
of activities during the workshop as we have tried to do in WSSSPE3.” Is
such a problem  symptomatic of all interdisciplinary projects with or
without limited funding? Would it be more successful if its efforts were
tied directly to the domain specific activities of the individual entities
represented by WSSSPE members? Is there a tension between the (idealistic?)
cross-domain goals of such a broad effort as WSSSPE and the practicalities
of getting people to work on someone else’s stuff?  I think there is an
opportunity for the authors to dig into their self assessed problem a little
bit more than they have.

RESPONSE:

Minor Comments:
1. I would avoid using the ableist term “bipolar” even if lifted
directly from the keynote;

RESPONSE: we've reworded this

2. I wonder if would be better to enhance the linking and referencing of
online materials in this document even beyond its already rich set of
footnoted links. For instance, all of the lightning talks refer to standing
proposals or active projects, but there are almost no reference URLs for
these projects or proposals;

RESPONSE:

3. There are many outlines in the appendices, and I’d like to note that
some outlines in the WSSSPE2 JORS article were negatively affected by poor
typesetting. You might want to make sure JORS gets them all right.

RESPONSE:

### 2016-08-26 WSSSPE3/report/WSSSPE3_review.tex
adding JORS reviews, a document for our responses, and making a couple of the easy changes

@@
 Overall, the report discusses the organization work done before the workshop
-(\S\ref{sec:preworkshop}); the keynote (\S\ref{sec:keynote}); a series of
-lightning talks (\S\ref{sec:lightning}). The report also gives summaries of
-action plans proposed by the working groups (\S\ref{sec:WGs}), and some
-conclusions (\S\ref{sec:conclusions}). Lists of the organizing committee
+(\S\ref{sec:preworkshop}), the keynote (\S\ref{sec:keynote}), and the lightning talks
+presented at the meeting (\S\ref{sec:lightning}). The report also gives summaries of
+action plans proposed by the working groups (\S\ref{sec:WGs}), and provides some
+conclusions (\S\ref{sec:conclusions}). The first three appendices contain lists of the organizing committee
 (Appendix~\ref{sec:orgcom}), the registered attendees
 (Appendix~\ref{sec:attendees}), and the travel award recipients
-(Appendix~\ref{sec:awardees}) are compiled. Finally, the report includes longer
+(Appendix~\ref{sec:awardees}). Finally, the report includes longer
 descriptions of the activities that occurred in each of the working groups that
@@
 top-down approaches, whereas the SciPy community more often than not uses more
-bottom-up systems. The essential messages perceived were also often bipolar: the
+bottom-up systems. According to Turk, there is a divergence in views about
+progress in software: the
 supercomputing community thinks that software is getting harder, with exascale

### 2016-08-26 WSSSPE3/report/wg_appendix_industry_interaction.tex
minor changes
adding JORS reviews, a document for our responses, and making a couple of the easy changes

@@
 
-\item A project could be \textbf{refunded}, and development or maintenance of
+\item A project could be \textbf{re-funded}, and development or maintenance of
 the software continue as planned.
@@
 transitions of researchers between academia and industry. This might address 
-each of the transition types (to be described below). Successful transitions are
+each of the transition types (as described below). Successful transitions are
 described as those that lead to either weak or strong sustainability (also
@@
 
-An example of data collection will be: 
+An example of data collection is: 
 \begin{itemize}

### 2016-08-29 WSSSPE3/report/JORS-reviews-responses.txt
more work in addressing reviewer comments

@@
 
-RESPONSE:
+RESPONSE:  We feel that this report should stick with where things were at
+WSSSPE3, since we otherwise will never be done updating it. But for the
+reviewer’s benefit, some groups have continued to work and have made good
+progress, others have basically stopped work, and some are in the middle.
+Some discussion of this can be found in
+http://www.slideshare.net/danielskatz/summary-of-wssspe-focused-on-wssspe3
 
@@
 Comments to the authors:
+
 General Comments:
@@
 
-RESPONSE:
+RESPONSE: This is a very good observation.  We have added a few paragraphs to
+the start of Section 5 to explain how the 10 initial working groups came about.
+Regarding the second part of this comment, what has happened after WSSSPE3, we
+feel that this report should stick with where things were at WSSSPE3, since
+we otherwise will never be done updating it. But for the reviewer’s benefit,
+some groups have continued to work and have made good progress, others have
+basically stopped work, and some are in the middle.  Some discussion of this
+can be found in http://www.slideshare.net/danielskatz/summary-of-wssspe-focused-on-wssspe3
 
@@
 
-RESPONSE:
+RESPONSE: The new paragraphs at the start of Section 5 should resolve
+this concern.
 
@@
 
-RESPONSE:
+RESPONSE:  Again, the new paragraphs at the start of Section 5 provide
+the requested background.
 

### 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
more work in addressing reviewer comments

@@
 
+After the keynote and lightning talks, the workshop facilitators led an exercise in which the
+attendees first suggested additional topics beyond those in the call for participation (as listed in
+\S\ref{sec:preworkshop}).  One topic that was introduced at this point was ``Building Sustainable
+User Communities for Scientific Software.''  The full set of topics were then placed on
+flipchart-sized pieces of paper around the walls, and the attendees voted on which topics
+they were strongly interested in working on at WSSSPE3, and which topics they were 
+generally interested in contributing to, but not as strongly.  This led to some topics being
+taken out of the mix for the rest of the workshop, since not enough people wanted to contribute
+to them to lead to a useful discussion.  Additionally, some topics were brought together by the
+participants who felt they were closely linked and that a group could address multiple of them
+in a single discussion.
+
+After this, the attendees broke up into small working groups to discuss
+the remaining topics during most of the remaining 1 1/2 days.  A high-level summary
+of each topic and group's work can be found in the subsections of this section, and for
+all but one group, 
+more detailed notes on each group's discussions can be found in Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.
+
+Midway through the first afternoon and between the two days, each group
+had a chance to talk about what progress they had made.  As discussed below (in \S\ref{sec:legacy}), the group that formed to discuss Legacy Software dissolved after the first session
+as group members left to join other groups.
+In addition, in the morning of the
+second day, a small set of reviewers/advisors (external to the groups but chosen by the group
+members) visited each group to listen to what they group was planning and to provide feedback.
+
 %\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}

### 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
adding periods after Lightning talk titles

@@
 \item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
-Facilitating Research Training While Supporting Scientific Research}} Benjamin
+Facilitating Research Training While Supporting Scientific Research}}. Benjamin
 Tovar presented a case study of ``The Cooperative Computing Lab'' at the
@@
 Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
-Sustainability}} The concept of sustainability has become a topic of interest in
+Sustainability}}. The concept of sustainability has become a topic of interest in
 the field of computing, which is evidenced by the increase in the number of
@@
 and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
-to Improve Productivity of Legacy Tools}} Abani Patra presented an example of
+to Improve Productivity of Legacy Tools}}. Abani Patra presented an example of
 how the community using Titan2D, a geoflow simulation software, increased the
@@
 Thaney: \textit{Collaborative Software Development as Sustainable Software:
-Lessons from Open Source}} Abigail Cabunoc Mayes combines two properties of open
+Lessons from Open Source}}. Abigail Cabunoc Mayes combines two properties of open
 source that, together, create a suitable habitat for sustainable software. The
@@
 through Best Practices in Open Source Software: Computational Infrastructure for
-Geodynamics}} Lorraine Hwang presented experiences with the Computational
+Geodynamics}}. Lorraine Hwang presented experiences with the Computational
 Infrastructure for Geodynamics (CIG), a community software with a worldwide user
@@
 Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
-in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
+in the Computational Infrastructure for Geodynamics}}. In a second talk, Lorraine
 Hwang mentioned a variety of ways to cite efforts within the framework, including
@@
 Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
-Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
+Open Science (DASPOS)}}. DASPOS is an NSF-funded multi-disciplinary effort,
 located at Notre Dame and Chicago, that links the high energy physics effort to
@@
 Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
-Susheel Varma: \textit{An update from UK Research Software Engineers}} James
+Susheel Varma: \textit{An update from UK Research Software Engineers}}. James
 Hetherington started by listing some of the problems research software faces,
@@
 \item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
-the User into Building Sustainable Software for Science}} Dan Gunter's main
+the User into Building Sustainable Software for Science}}. Dan Gunter's main
 question was, ``What is needed to develop sustainable software?''. Beyond the
@@
 Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
-Platform for Predictive Biology: Lessons Learned on the KBase Project}} 
+Platform for Predictive Biology: Lessons Learned on the KBase Project}}. 
 Dan Gunter presented experiences and lessons learned as part of KBase, an
@@
 Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
-Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
+Scientists in Best Practices of Software Sharing}}. Erin Robinson presented an
 approach to overcome some hurdles in scientific publishing: disconnects among
@@
 \item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
-Accreditation Framework}} Neil Chue Hong presented a proposal to build a
+Accreditation Framework}}. Neil Chue Hong presented a proposal to build a
 scientific software accreditation framework. One of the aims of such a framework
@@
 \item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
-Support for Sustainable Scientific Software}} Carver argued that for scientific
+Support for Sustainable Scientific Software}}. Carver argued that for scientific
 software to be truly sustainable, there is a need for developers to use
@@
 
-\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
+\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}.
 This talk highlighted an attempt to solve the common problem for open source
@@
 \item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}} Alice Allen described the Astrophysics Source Code Library
+included)}}. Alice Allen described the Astrophysics Source Code Library
 (ascl.net), an increasingly used way to obtain a unique ID for astrophysics
@@
 Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
-Developing Community}} The effort behind this talk given by Ray Idaszak started
+Developing Community}}. The effort behind this talk given by Ray Idaszak started
 from a breakout group at an NSF SI2 workshop in 2015, and centers around

### 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
adding refs for Carver lightning talk

@@
 scientific teams indicates that choosing and tailoring these practices is not a
-trivial exercise. There is a general culture clash between software engineering
+trivial exercise~\cite{Nanthaamornphong-Carver:2015, Heaton-Carver:2015, Basili-etal:2008, 4222616}.
+There is a general culture clash between software engineering
 and science that hinders our ability to communicate and choose appropriate

### 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
changing in-line reference to footnote with full clickable URL

@@
 \item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}}. Alice Allen described the Astrophysics Source Code Library
-(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
+included)}}. Alice Allen described the Astrophysics Source Code Library\footnote{\url{http://ascl.net}}, an increasingly used way to obtain a unique ID for astrophysics
 software that is indexed by indexing services and can be cited. ASCL offers

### 2016-08-30 WSSSPE3/report/WSSSPE3_review.tex
adding ASCL reference/cite

--- 2016-08-29 WSSSPE3/report/WSSSPE3_review.tex
@@
 \item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}}. Alice Allen described the Astrophysics Source Code Library
-(ascl.net), an increasingly used way to obtain a unique ID for astrophysics
-software that is indexed by indexing services and can be cited. ASCL offers
+included)}}. Alice Allen described the Astrophysics Source Code
+Library\footnote{\url{http://ascl.net}}, an increasingly used way to obtain a unique ID for astrophysics
+software that is indexed by indexing services and can be cited~\cite{wssspe2_allen_jors}.
+ASCL offers
 clones of existing infrastructure, provides server space and computing

### 2016-08-30 WSSSPE3/report/WSSSPE3_review.tex
adding RSE cites/references

@@
 new role in the academic system focused on research software that combines the
-best parts of a craftsperson and a scholar. However, social innovation in
+best parts of a craftsperson and a scholar~\cite{rse1,rse2}. However, social innovation in
 centuries-old institutions is hard. Alternatives to such a new role would have

### 2016-08-30 WSSSPE3/report/WSSSPE3_review.tex
adding DASPOS references/cites

@@
 Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
-Open Science (DASPOS)}}. DASPOS is an NSF-funded multi-disciplinary effort,
+Open Science (DASPOS)}}. DASPOS\footnote{\url{http://daspos.org}} is an NSF-funded multi-disciplinary effort,
 located at Notre Dame and Chicago, that links the high energy physics effort to
@@
 One of the products of this effort is a suite of tools that deals with this
-preservation, and the questions was posed whether that software itself is
+preservation~\cite{Meng2015137,Meng:2015:UPE:2755979.2755982,1742-6596-664-3-032022,nd4},
+and the questions was posed whether that software itself is
 sustainable. Points that were brought up included the need for a user community

### 2016-08-30 WSSSPE3/report/WSSSPE3_review.tex
adding CCL URL

@@
 Facilitating Research Training While Supporting Scientific Research}}. Benjamin
-Tovar presented a case study of ``The Cooperative Computing Lab'' at the
+Tovar presented a case study of the Cooperative Computing Lab (CCL)\footnote{\url{http://ccl.cse.nd.edu}} at the
 University of Notre Dame, which is a small group of individuals whose main tasks

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
adding ref/cite for Ahalt lightning talk

@@
 from a breakout group at an NSF SI2 workshop in 2015, and centers around
-building a framework for creating metrics for scientific software. This
+building a framework for creating metrics for scientific software~\cite{ahalt-csessp}. This
 framework would improve both the metrics and the software it evaluates, and

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
adding footnote URLs for Hwang talks

@@
 Geodynamics}}. Lorraine Hwang presented experiences with the Computational
-Infrastructure for Geodynamics (CIG), a community software with a worldwide user
+Infrastructure for Geodynamics (CIG)\footnote{\url{http://geodynamics.org/}}, a community software with a worldwide user
 base. Like others, their main goals include high usability, sustainability, and
@@
 in the Computational Infrastructure for Geodynamics}}. In a second talk, Lorraine
-Hwang mentioned a variety of ways to cite efforts within the framework, including
+Hwang mentioned a variety of ways to cite efforts within the SAGA framework\footnote{\url{https://geodynamics.org/cig/projects/saga/}}, including
 science papers, code papers, user manuals, and the CIG website. An analysis of

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
adding text at the start of 4 and response saying so

@@
 
-The lightning talks were intended to give an opportunity for attendees to
+After the keynote, WSSSPE3 continued with lightning talks.
+These short talks were intended to give an opportunity for attendees to
 quickly highlight an important issue or a potential solution.

### 2016-08-31 WSSSPE3/report/JORS-reviews-responses.txt
adding response for references for lightning talks
adding text at the start of 4 and response saying so
adding another response

@@
 
-RESPONSE:
+RESPONSE: We’ve added some text at the start of Sections 4 and 5 to better
+explain the workshop and report structure.
 
@@
 
-RESPONSE:
+RESPONSE: We have added a number of URLs (footnotes) and references to the
+lightning talk paragraphs.
 
@@
 
-RESPONSE:
+RESPONSE: The JORS typesetting process is a bit different than that of
+traditional journals, but we will strive to pay attention to this and ensure
+that it is correct.

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
adding Patra ref/cite

@@
 how the community using Titan2D, a geoflow simulation software, increased the
-productivity of their tools by improving both code and data layout. The main
+productivity of their tools by improving both code and data layout~\cite{Patra20051}. The main
 obstacles in this change were the non-existence of a common version control

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
small fix in 5 intro, removing first person in 5.1 for group stuf

@@
 
-After the keynote and lightning talks, the workshop facilitators led an exercise in which the
+After the keynote and lightning talks, the workshop facilitators led an exercise to create working
+groups for the rest of the meeting.  The
 attendees first suggested additional topics beyond those in the call for participation (as listed in

### 2016-08-31 WSSSPE3/report/wg_main_best_practices.tex
small fix in 5 intro, removing first person in 5.1 for group stuf

@@
 The creation of such a ``best practices'' document will build upon the range of
-activities and topics discussed at WSSSPE3 and associated prior meetings. We
+activities and topics discussed at WSSSPE3 and associated prior meetings. This
+working group
 will attempt to distill the emerging body of knowledge into this document. The
 large number of articles from the NSF funded SI2 projects (SSE and SSI),
-``lightning talks'', ``white papers,'' and reports from different workshops have
+``lightning talks,'' ``white papers,'' and reports from different workshops have
 created a large if somewhat diffuse source for this report.
@@
 reliability,  usability, extensibility, knowledge management, as well as continuity
-(transitions between people). Answers to these questions will guide us on how a software
-tool becomes part of the core workflow of well identified users (stakeholders)
+(transitions between people). Answers to these questions will guide the group
+to learn how a software
+tool becomes part of the core workflow of well-identified users (stakeholders)
 relating to tool success and hence sustainability.
@@
 The creation of a document on best practices needs a large and diverse community
-involved. We have enlisted over ten contributors from the attendees at the
+involved. The group has enlisted over ten contributors from the attendees at the
 WSSSPE3 and those on the mailing list. The primary mechanism for developing this
@@
 community scientific software and organizations supporting scientific software.
-We will attempt then to abstract general principles and best practices. Some of
+The group will attempt then to abstract general principles and best practices. Some of
 the tools identified for such analysis are the general purpose PETSc toolkit for
@@
 (Computational Infrastructure for Geodynamics) organization dedicated to
-supporting an ensemble of related tools for the geodynamics community. We also
+supporting an ensemble of related tools for the geodynamics community. The group also
 established a timeline and a rough outline (see

### 2016-08-31 WSSSPE3/report/wg_main_eng_design.tex
changing subsections of 5 to not be first person

@@
 
-The group discussion focused on identifying existing principles of software
+This group's discussion focused on identifying existing principles of software
 engineering design that could be adopted by the computational science and

### 2016-08-31 WSSSPE3/report/wg_main_funding_spec_expert.tex
changing subsections of 5 to not be first person

@@
 
-The solution we envision entails creating an environment where software
+The solution this group envisions entails creating an environment where software
 developers are a stable part of a research team. Such an environment mitigates
@@
 specific projects as well as about software development for today's research.
-Our vision is to find a way to promote a university/research institute
+The group's vision is to find a way to promote a university/research institute
 environment where software developers are stable components of research project
@@
 One strategy to promote stability is implementing a mechanism for developers to
-obtain academic credit for software development work. With such a mechanism in
+obtain academic credit for software development work (see \S\ref{sec:software-credit}.) With such a mechanism in
 place, traditional academic funding models and career tracks could properly
@@
 worthy aspirational goal, such a dramatic change in this structure does not seem
-likely in a time scale relevant to this working group. Accordingly, our working
-party sought alternative solutions that may be achievable within the context of
+likely in a time scale relevant to this working group. Accordingly, the working
+group sought alternative solutions that may be achievable within the context of
 existing academic structures. The group felt that developing dedicated research
-software engineering roles within the university, and finding stable funding for
+software engineering roles within the university and finding stable funding for
 those individuals is the most promising mechanism for creating a stable software
@@
 case to the university for continued funding. Research software engineers will
-not be measured by publications, we hope, but by other metrics. Middle-author
+hopefully not be measured by publications, but by other metrics. Middle-author
 publications are common for RSEs. Most RSEs welcome co-authorship on papers
-where the PIs think that the contribution deserves it.
+when the PIs think that the contribution deserves it.
 
@@
 the body of software is created and maintained in research universities.
-Therefore, we argue that research institutions should develop hybrid
+Therefore, the group argues that research institutions should develop hybrid
 academic-technical tracks for this capacity, where employees in this track work
@@
 
-Having convinced ourselves that this would be a positive innovation, we were
+Having convinced themselves that this would be a positive innovation, the group members were
 then faced with the specific question of how to fund the initiation of this
@@
 creation of the group in University College London), but more likely, seed
-funding needs to come from research councils (as with the Research Software
+funding needs to come from research councils or other funding bodies (as with the Research Software
 Engineering Fellowship provided by the UK Engineering and Physical Sciences
-Research Council). We therefore recommend that funding organizations consider
+Research Council). The group therefore recommends that funding organizations consider
 how they might provide such seed funding.
@@
 
-We finally noted that the institutional case for such groups is made easier by
+The working group finally noted that the institutional case for such groups is made easier by
 having successful examples to point to. In the UK, a collective effort to
@@
 other organizations to see if there may be a widespread community of RSEs who do
-not identify themselves as such at this time. We will collect information about
+not identify themselves as such at this time. This working group will collect information about
 the organizational models under which these groups function, and how they are
@@
 successful is the university in getting investigators to pay for fractional
-RSEs? We will author a report describing our findings, should funding be
+RSEs? The group will author a report describing their findings, should funding be
 available to conduct the investigation.

### 2016-08-31 WSSSPE3/report/wg_main_industry_interaction.tex
minor changes
changing subsections of 5 to not be first person

@@
 
-At WSSSPE3 our group was interested in better understanding successful pathways
+At WSSSPE3, this working group was interested in better understanding successful pathways
 for scientific software to ``transition'' from grant-funded research projects to
@@
 sharing, and funding of research software has focused on the impact of such
-arrangements. Examples of these types of reports are as follows:
+arrangements. Examples of these types of reports are:
 
@@
 Although sustainability transitions are often studied under the broad umbrella
-of ``technology transfer,'' we believe there are likely to be a number of
+of ``technology transfer,'' the group believes there are likely to be a number of
 different ways in which a pathway from initial production to long-term
@@
 Plans for carrying forward are currently unclear---this project would require
-sustained attention and effort from our team, and at least some amount of
-funding in order for us to be involved for extended periods of time.
+sustained attention and effort from the group members, and at least some amount of
+funding in order for those members to be involved for extended periods of time.
 
-The broad goals that we would like to accomplish are:
+The broad goals that the group would like to accomplish are:
 

### 2016-08-31 WSSSPE3/report/wg_main_metrics.tex
changing subsections of 5 to not be first person

@@
 assess this, how are metrics monitored, and many more. A more complete bulleted
-list of these questions can be found in Appendix H. Next, a roadmap for how to
-proceed was discussed including creating a set of milestones and tasks. The idea
+list of these questions can be found in Appendix~\ref{sec:appendix_metrics}. Next, a roadmap for how to
+proceed was discussed, including creating a set of milestones and tasks. The idea
 was put forth for the group to interact with the organizing committee of the

### 2016-08-31 WSSSPE3/report/wg_main_software_credit.tex
changing subsections of 5 to not be first person

@@
 
-Since near-term efforts will be shifting to the FORCE11-SCWG, we direct
-interested readers to that group's existing landing page\footnote{FORCE11-SCWG
+Since near-term efforts will be shifting to the FORCE11-SCWG, 
+interested readers should go to that group's existing landing page\footnote{FORCE11-SCWG
 landing page,

### 2016-08-31 WSSSPE3/report/wg_main_user_community.tex
changing subsections of 5 to not be first person

@@
 Guide''~\cite{working-open-wssspe3} and
-``UK Collaborative Computational Projects'' (CCP: http://www.ccp.ac.uk), or 
+``UK Collaborative Computational Projects'' (CCP)\footnote{\url{http://www.ccp.ac.uk)}}, or 
  books such as ``Art of Community'' by Jono Bacon~\cite{art-of-community}.
@@
 
-Everyone agreed on a few points: software must not only offer value, but there
+All the group members agreed on a few points: software must not only offer value, but there
 must be some support for users; and funding can help pay for that support, in

### 2016-08-31 WSSSPE3/report/WSSSPE3_review.tex
adding ref/cite for Gil talk

@@
 experimental data, research software, and publications. A current effort in the
-Geosciences is the ``Geoscience Paper of the Future,'' which includes four
-elements. First, it forms a modern paper including text, data, and pointers to
+Geosciences is the ``Geoscience Paper of the Future,''\footnote{\url{http://www.ontosoft.org/gpf}} which includes four
+elements~\cite{ESS2:ESS284}. First, it forms a modern paper including text, data, and pointers to
 supplementary materials. Second, it is reproducible, including data processing,
@@
 Geoscience areas is planned to include only Geoscience papers of the future,
-with submissions open until the end of 2015. In addition, training sessions are
+with submissions open until the end of 2015~\cite{GPFI}. In addition, training sessions are
 provided to geoscientists to learn best practices in software and data sharing,

### 2016-09-01 WSSSPE3/report/WSSSPE3_review.tex
adding KBase URL

@@
 Platform for Predictive Biology: Lessons Learned on the KBase Project}}. 
-Dan Gunter presented experiences and lessons learned as part of KBase, an
+Dan Gunter presented experiences and lessons learned as part of KBase\footnote{\url{http://kbase.us}}, an
 open software and data platform for addressing the grand challenge of systems

### 2016-09-01 WSSSPE3/report/wg_main_user_community.tex
fixing a typo

@@
 Guide''~\cite{working-open-wssspe3} and
-``UK Collaborative Computational Projects'' (CCP)\footnote{\url{http://www.ccp.ac.uk)}}, or 
+``UK Collaborative Computational Projects'' (CCP)\footnote{\url{http://www.ccp.ac.uk}}, or 
  books such as ``Art of Community'' by Jono Bacon~\cite{art-of-community}.

### 2016-09-01 WSSSPE3/report/WSSSPE3_review.tex
"set of topics were" to "set of topics was"

@@
 \S\ref{sec:preworkshop}).  One topic that was introduced at this point was ``Building Sustainable
-User Communities for Scientific Software.''  The full set of topics were then placed on
+User Communities for Scientific Software.''  The full set of topics was then placed on
 flipchart-sized pieces of paper around the walls, and the attendees voted on which topics

### 2016-09-01 WSSSPE3/report/WSSSPE3_review.tex
topics were "brought together" -> "combined"

--- 2016-09-01 WSSSPE3/report/WSSSPE3_review.tex
@@
 taken out of the mix for the rest of the workshop, since not enough people wanted to contribute
-to them to lead to a useful discussion.  Additionally, some topics were brought together by the
+to them to lead to a useful discussion.  Additionally, some topics were combined by the
 participants who felt they were closely linked and that a group could address multiple of them

### 2016-09-02 WSSSPE3/report/wg_appendix_software_credit.tex
fixing bib to more recent version

@@
 
-\item Transitive credit~\cite{wssspe2_katz,Katz:2014_tc}, which assigns
+\item Transitive credit~\cite{Katz:2014_tc,katz-smith-tc}, which assigns
 contriponents (contributors and components) various weights depending on their
@@
 software and code, and implementing transitive credit via
-JSON-LD~\cite{wssspe2_katz}. Some questions arose about how this information
+JSON-LD~\cite{katz-smith-tc}. Some questions arose about how this information
 would be stored for closed-source software.

### 2016-09-02 WSSSPE3/report/wg_main_software_credit.tex
fixing bib to more recent version

@@
 of credit could be refined based on concepts such as transitive
-credit~\cite{wssspe2_katz,Katz:2014_tc}.
+credit~\cite{Katz:2014_tc,katz-smith-tc}.
 

### 2016-09-14 WSSSPE4_report/orgcom.tex
added organizing committee

Gabrielle Allen &  University of Illinois Urbana--Champaign, USA \\
Jeffrey Carver  & University of Alabama, USA \\
Sou-Cheng (Terrya) Choi &  Illinois Institute of Technology, USA \\
Tom Crick & Cardiff Metropolitan University, UK \\
Michael R.~Crusoe & Common Workflow Language Project \\
Sandra Gesing & University of Notre Dame, USA \\
Robert Haines & University of Manchester, UK \\
Michael Heroux & Sandia National Laboratories, USA \\
Lorraine J.~Hwang & University of California, Davis, USA \\
Daniel S.~Katz & University of Illinois Urbana--Champaign, USA \\
Kyle E.~Niemeyer & Oregon State University, USA \\
Manish Parashar & Rutgers University, USA \\
Colin C.~Venters & University of Huddersfield, UK

### 2016-09-14 WSSSPE4_report/insert_registrants.py
added registrants parsing script

@@ -1,0 +1,25 @@
+#!/usr/bin/env python
+import re, operator
+
+with open("registrants.txt", 'r') as f:
+  people = []
+  person = {}
+  for line in f:
+    if 'institution: ' in line:
+      person['institution'] = re.search('institution: (.*)', line).group(1).replace('&', r'\&')
+    elif 'email: ' in line:
+      person['email'] = re.search('email: (.*)', line).group(1)
+    elif len(line) > 2:
+      person['name'] = line[:-1].replace('&', r'\&')
+    else:
+      people.append(person)
+      person = {}
+if 'name' in person:
+  people.append(person)
+
+for person in sorted(people, key=lambda x: x['name'].split()[-1]):
+  #print(person['name'] + ' & ' + person['institution'] + r'\\')
+  print(person['name'] + ' & ' +
+        r'\href{mailto:' + person['email'] + r'}{' + person['email'] + r'}&' +
+        person['institution'] + r'\\'
+       )

### 2016-09-14 WSSSPE4_report/awardees.tex
added awardees list

Alice Allen & Astrophysics Source Code Library \\
Wolfgang Bangerth & Colorado State University, USA \\
Roman Bauer & Newcastle University, UK \\
Steven Brandt & Louisiana State University, USA \\
Emily Chen & University of Illinois Urbana-Champaign, USA \\
Bruce Childers & University of Pittsburgh, USA \\
Michael R.~Crusoe & Common Workflow Language Project \\
Stephan Druskat & Humboldt-Universit\"{a}t zu Berlin, Germany \\
Anshu Dubey & Argonne National Laboratory, USA \\
Iain Emsley & Oxford University, UK \\
Christopher Gwilliams & Cardiff University, UK \\
Lorraine Hwang & University of California Davis, USA \\
Ray Idaszak & RENCI, University of North Carolina at Chapel Hill, USA \\
Xinlian Liu & Hood College, USA \\
Frank Löffler & Louisiana State University, USA \\
Tengyu Ma & Vanderbilt University, USA \\
Robert H.~McDonald & Indiana University, USA \\
Jonah Miller & Perimeter Institute, Canada
Daniel Mosse & University of Pittsburgh, USA \\
Kyle E.~Niemeyer & Oregon State University, USA \\
Santiago N\'{u}\~{n}ez-Corrales & University of Illinois Urbana-Champaign, USA \\
Dominic Orchard & University of Kent, UK \\
Francisco Queiroz & Pontifical Catholic University of Rio de Janeiro, Brazil \\
Jean Salac & University of Virginia, USA \\
Janos Sallai & Vanderbilt University, USA \\
Eric Seidel & University of California San Diego, USA \\
Prabhjyot Singh Saluja & Southwest Oklahoma State University, USA \\
Karan Shah & Georgia Institute of Technology, USA \\
Sameer Shende & University of Oregon, USA \\

### 2016-09-14 WSSSPE4_report/insert_registrants.py
changed registrants to participants

@@ -1,7 +1,7 @@
 #!/usr/bin/env python
 import re, operator
 
-with open("registrants.txt", 'r') as f:
+with open("participants.txt", 'r') as f:
   people = []
   person = {}
   for line in f:

### 2016-09-14 WSSSPE4_report/insert_participants.py
changed participant script a bit

@@ -20,6 +20,6 @@ if 'name' in person:
 for person in sorted(people, key=lambda x: x['name'].split()[-1]):
   #print(person['name'] + ' & ' + person['institution'] + r'\\')
   print(person['name'] + ' & ' +
-        r'\href{mailto:' + person['email'] + r'}{' + person['email'] + r'}&' +
+        r'\href{mailto:' + person['email'] + r'}{' + person['email'] + r'} & ' +
         person['institution'] + r'\\'
        )

### 2016-09-14 WSSSPE4_report/awardees.tex
added missing newline to awardees

@@
 Robert H.~McDonald & Indiana University, USA \\
-Jonah Miller & Perimeter Institute, Canada
+Jonah Miller & Perimeter Institute, Canada \\
 Daniel Mosse & University of Pittsburgh, USA \\
@@
 Karan Shah & Georgia Institute of Technology, USA \\
-Sameer Shende & University of Oregon, USA \\
+Sameer Shende & University of Oregon, USA

### 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
empty file for open research index section

--- 2016-05-19 WSSSPE3/best_practices/main.tex
@@
-\documentclass[11pt, oneside]{amsart}
-\pdfoutput=1
-
-\usepackage{amsmath}
-\usepackage{amssymb}
-
-\usepackage{color}
-\usepackage{dcolumn}
-\usepackage{float}
-\usepackage{graphicx}
-\usepackage[utf8]{inputenc}
-\usepackage[T1]{fontenc}
-\usepackage{lmodern}
-\usepackage{multirow}
-\usepackage{rotating}
-\usepackage{subfigure}
-\usepackage{psfrag}
-\usepackage{tabularx}
-\usepackage[hyphens]{url}
-\usepackage{wrapfig}
-\usepackage{longtable}
-\usepackage{verbatim}
-\usepackage{booktabs,multicol}
-
-% The following three lines are used for displaying footnote in tables.
-\usepackage{footnote}
-\makesavenoteenv{tabular}
-\makesavenoteenv{table}
-
-\usepackage{enumitem}
-\setlist{leftmargin=7mm}
-
-%\setcounter{secnumdepth}{3}
-%\setcounter{tocdepth}{3}
-
-\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
-\usepackage[all]{hypcap}
-\urlstyle{rm}
-
-\definecolor{orange}{rgb}{1.0,0.3,0.0}
-\definecolor{violet}{rgb}{0.75,0,1}
-\definecolor{darkgreen}{rgb}{0,0.6,0}
-\definecolor{cyan}{rgb}{0.2,0.7,0.7}
-\definecolor{blueish}{rgb}{0.2,0.2,0.8}
-\definecolor{darkblue}{rgb}{0.1,0.1,0.9}
-
-\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
-\newcommand{\note}[1]{ {\textcolor{blueish}    { ***Note:      #1 }}}
-
-% Don't use tt font for urls
-\urlstyle{rm}
-
-% 15 characters / 2.5 cm => 100 characters / line
-% Using 11 pt => 94 characters / line
-\setlength{\paperwidth}{216 mm}
-% 6 lines / 2.5 cm => 55 lines / page
-% Using 11pt => 48 lines / pages
-\setlength{\paperheight}{279 mm}
-\usepackage[top=2.5cm, bottom=2.5cm, left=2.5cm, right=2.5cm]{geometry}
-% You can use a baselinestretch of down to 0.9
-\renewcommand{\baselinestretch}{0.96}
-
-\sloppypar
-
-\begin{document}
-
-\title[]{Best Practices for Developing Sustainable Scientific Software}
-
-\author{Sandra Gesing$^{(1)}$,
-Abani Patra$^{(2)}$,
-Daniel S.\ Katz$^{(3)}$,
-}
-
-%
-\thanks{{}$^{(1)}$ Center for Research Computing, University of Notre Dame, Notre Dame, IN, USA; sandra.gesing@nd.edu}
-%
-\thanks{{}$^{(2)}$ Mechanical and Aerospace Engineering, University at Buffalo, Buffalo, NY, USA; abani@buffalo.edu}
-%
-\thanks{{}$^{(3)}$ National Center for Supercomputing Applications (NCSA), Graduate School of Library and Information Science, and Electrical and Computer Engineering, University of Illinois, Urbana-Champaign, IL, USA; d.katz@ieee.org}
-%
-
-\begin{abstract}
-
-\end{abstract}
-
-\maketitle
-\newpage
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction} \label{sec:intro}
+\subsection{Open Research Index}
+\label{sec:open-research-index}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Scientific software is an integral part of research in a plethora of areas such as applied natural sciences, science of management and humanities - not only increasing the efficiency to find answers to research questions but also enhancing methods to answer some research questions at all.
-While the importance of developing scientific software is evident in the large number of software package, methods and algorithms applied in science, sustaining it and thus make it easily usable and re-usable for the user communities entails challenges immanent in being part of the highly dynamic environment in research.
-\note{Need rewording - copied from the report}
-Reviewing multiple past articles and talks at different meetings like WSSSPEx
-\cite{WSSSPE1, WSSSPE2, 5069157, Blatt_WSSSPE, Ahern_WSSSPE} and analyzing and
-promoting sustainable scientific software makes it clear that there are several
-common and recurring ideas that underpin success in developing sustainable
-software. However, outside of a small community, this knowledge is not widely
-shared. This is especially true for the large community of scientists who
-generate most of the software used by scientists but are not primarily software
-developers. In this scenario, a clear and precise exposition of these best
-practices collected from many sources and open collaboration among all in the
-community in a single source (e.g., journal paper, tutorial) that can be widely
-disseminated is necessary and likely to be very valuable.
-
-Core questions that will need to be explored are in reproducibility, 
-reliability,  usability, extensibility, knowledge management, as well as continuity
-(transitions between people). Answers to these questions will guide us on how a software
-tool becomes part of the core workflow of well identified users (stakeholders)
-relating to tool success and hence sustainability.
-%\katznote{prev sentence is complex and awkward} % see fix
-Ideas  that may need to be explored include:
-\begin{itemize}
-\item Requirements engineering to create tools with immediate uptake;
-\item When should software ``die''?
-\item Catering to disruptive developments in environment (e.g., new hardware,
-new methodology);
-\item Dimensions of sustainability: economic, technical, environmental and
-obsolescence. 
-%\katznote{not sure what the prev. comment goes with} %social.
-\end{itemize}
-
-Sustainability requires community participation in code development and/or a
-wide adoption of software. The larger the community base is using a piece of
-software, the better are the funding possibilities and thus also the
-sustainability options. Additionally, developer commitment to an application is
-essential and experience shows that software packages with an evangelist
-imposing strong inspiration and discipline are more likely to achieve
-sustainability. While a single person can push sustainability to a certain
-level, open source software also needs sustained commitment from the developer
-community. Such sustained commitments include diverse tasks and roles, which can
-be fulfilled by diverse developers with different knowledge levels. Besides
-developing software and appropriate software management with measures for
-extensibility and scalability of the software, active (expertise) support for
-users via a user forum with a quick turnaround is crucial. The barrier to entry
-for the community as users as well as developers has to be as low as possible.
-\note{End of the copied text}
-
-Sustainability of software can be investigated under many diverse aspects and dimensions: 
-economic, technical, environmental and social are only a few. The scope of this paper will focus 
-on scientific environments and their implications and the target group are especially developer 
-and project manager of scientific software. To set the stage we picked case studies, which have been 
-successful to some degree and analyze why they have been successful. In Section~\ref{sec:comm} are 
-practices described, which are partly resulting of the means available in and demands resulting of a scientific environment. 
-The subsequent section focus on the social aspects of governing and managing sustainable software, 
-while Section~\ref{sec:funding} is concerned with economic aspects. To be able to measure sustainability of 
-software, it is important to define metrics as reference and guideline to be able to create such 
-software. Additionally, we will present characteristics of available tools (see Section~\ref{sec:tools}) 
-for managing software but also for analyzing and delivering metrics of sustainability. Finally, 
-we will conclude with an outlook and suggestion to improve sustainability in a scientific context.
-
-\input{related_work}
-\input{case_studies}
-\input{community_related_practices}
-\input{government_and_management}
-\input{funding}
-\input{metrics}
-\input{tools}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusion} \label{sec:conclusion}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgments} \label{sec:acks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\bibliographystyle{vancouver}
-
-\bibliography{wssspe}
-\end{document}

### 2016-09-14 WSSSPE4_report/insert_participants.py
added email options to insert_participants

@@ -1,7 +1,12 @@
 #!/usr/bin/env python
 import re, operator
 
-with open("participants.txt", 'r') as f:
+# set to 'name' for hyperlinking email address with name,
+# 'column' for a column, and
+# 'off' for no email.
+email = 'name'
+
+with open('participants.txt', 'r') as f:
   people = []
   person = {}
   for line in f:
@@ -18,8 +23,16 @@ if 'name' in person:
   people.append(person)
 
 for person in sorted(people, key=lambda x: x['name'].split()[-1]):
-  #print(person['name'] + ' & ' + person['institution'] + r'\\')
-  print(person['name'] + ' & ' +
-        r'\href{mailto:' + person['email'] + r'}{' + person['email'] + r'} & ' +
-        person['institution'] + r'\\'
-       )
+    if email == 'column':
+      print(person['name'] + ' & ' +
+            r'\href{mailto:' + person['email'] + r'}{' +
+            person['email'] + r'} & ' +
+            person['institution'] + r'\\'
+           )
+    elif email == 'off':
+        print(person['name'] + ' & ' + person['institution'] + r'\\')
+    elif email == 'name':
+        print(r'\href{mailto:' + person['email'] + r'}{' +
+              person['name'] + r'} & ' +
+              person['institution'] + r'\\'
+             )

### 2016-09-14 WSSSPE4_report/WSSSPE4_report.tex
barebones draft of WSSSPE4 report

\documentclass[11pt, oneside]{amsart}
\pdfoutput=1

\usepackage{amsmath,amssymb,amsmath}

\usepackage{color}
\usepackage{dcolumn}
\usepackage{float,caption}
\usepackage{graphicx}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\usepackage{xspace}

\usepackage{rotating}
\usepackage{subfigure}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{verbatim}
\usepackage{booktabs,multicol,multirow}

%better printing of numbers
\usepackage[english]{babel}
\usepackage{textcomp}
\usepackage{csquotes}

% The following three lines are used for displaying footnote in tables.
\usepackage{footnote}
\makesavenoteenv{tabular}
\makesavenoteenv{table}

\usepackage{enumitem}
\setlist{leftmargin=7mm}

%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}

\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
\usepackage[all]{hypcap}
\urlstyle{rm}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}
\definecolor{darkblue}{rgb}{0.1,0.1,0.9}

\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
\newcommand{\note}[1]{ {\textcolor{blueish}    { ***Note:      #1 }}}
\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan:      #1 }}}
\newcommand{\gabnote}[1]{ {\textcolor{cyan}    { ***Gabrielle:     #1 }}}
\newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}

% Don't use tt font for urls
\urlstyle{rm}

\usepackage[letterpaper, margin=1in]{geometry}
% You can use a baselinestretch of down to 0.9
\renewcommand{\baselinestretch}{0.96}

\sloppypar

\begin{document}

\title[]{Report on the Fourth Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE4) workshop}

\author{Daniel S.\ Katz$^{(1)}$,
Kyle E.\ Niemeyer$^{(2)}$,
}

%
\thanks{{}$^{(1)}$ \hspace{-1ex}Computation Institute,
University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org}
%
\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu}
%

\begin{abstract}
This report records and discusses the Fourth Working towards Sustainable
Software for Science: Practice and Experiences (WSSSPE4) workshop. The report
includes a description of the keynote presentation of the workshop,
which % TODO
% It also summarizes a set of lightning
% talks in which speakers highlighted to-the-point lessons and challenges
% pertaining to sustaining scientific software.
% The final and main contribution of the report is a summary of the
% discussions, future steps, and future organization for a set of self-organized
% working groups on topics including developing pathways to funding scientific
% software; constructing useful common metrics for crediting software
% stakeholders; identifying principles for sustainable software engineering
% design; reaching out to research software organizations around the world; and
% building communities for software sustainability. For each group, we include a
% point of contact and a landing page that can be used by those who want to join
% that group's future activities. The main challenge left by the workshop is to
% see if the groups will execute these activities that they have scheduled, and
% how the WSSSPE community can encourage this to happen.

\end{abstract}

\maketitle
\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The Fourth Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
held on 12--14 September 2016 in Manchester, England, UK. Previous events in
the WSSSPE series are
WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
held in conjunction with SC13;
WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
focused workshop organized jointly with the SciPy
conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
held in conjunction with SC14;
WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
focused workshop organized again jointly with
SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}};
and WSSSPE3\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}}~\cite{WSSSPE3},
held in Boulder, Colorado, USA.
Note that the WSSSPE series changed its name from ``Workshop on Sustainable
Software for Science: Practice and Experiences'' to ``Working towards
Sustainable Software for Science: Practice and Experiences'' to better reflect
the movement towards a community in addition to the associated workshop series.

\todo{Introduction}

The WSSSPE4 workshop included multiple mechanisms for participation and
encouraged team building around solutions. WSSSPE4 strongly encouraged participation
of early-career scientists, postdoctoral researchers, graduate students,
early-stage researchers, and those from underrepresented groups,
with funds provided to the conference organizers by the National Science
Foundation (NSF), the Alfred P.~Sloan Foundation, and the Software
Sustainability Institute (SSI) to support the travel of potential participants
who would not otherwise be able to attend the workshop. These
funds allowed 29 additional people to attend and participate. A subset of the
organizing committee reviewed 44 applications for travel support and
competitively selected the awardees, including 11 students and five early-career
researchers. In addition, the travel award subcommittee tried to increase the
diversity of applicants by forwarding the notice of travel support to
organizations including PyLadies, Django Girls, Women Who Code, and
Women in HPC.\todo{add links}
\todo{did we reach out to any others?}

WSSSPE4 also included two professional event organizers/facilitators from
KnowInnovation who helped the organizing committee plan the workshop agenda,
and during the workshop, they actively engaged participants with various
tools, activities, and reminders.

This report is based on
\todo{what report is based on}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Calls for participation} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

WSSSPE4 was based on the work done in WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
at \todo{what aimed at?}
as the calls for participation said:

two tracks:

\begin{quote}
    call for participation
\end{quote}

\todo{update}
The calls for participation requested papers describing lightning talks,
where each author could make a brief statement about work that either had been
done or was needed;
idea papers, which present implementable proposals related to Track 1;
position papers, which are longer, not previously published papers related to
Track 2 specifically discussing what we can do to improve sustainable scientific
software in the short term, starting with where we are today;
experience papers, longer papers related to Track 2 that discuss current
practices and experiences and how they have been used to improve the quality of
today’s research software and/or the experiences of its developers;
and demos, brief papers describing a 10--15 minute demonstration of a tool or
process relevant to Track 2 that improves the quality of today's research
software and\slash or the experiences of its developers.

There were~24 lightning talks submitted; after a peer-review process, 16
were accepted, as discussed further in Section~\ref{sec:lightning}.
\todo{update}

%
%\begin{quote}
\begin{itemize}
\renewcommand{\labelenumi}{\textbf{\theenumi}.}
\setlength{\rightmargin}{1em}

\item Development and Community
\begin{itemize}
    \item Best practices for developing sustainable software
    \item Models for funding specialist expertise in software collaborations
    \item Software tools that aid sustainability
    \item Academia/industry interaction
    \item Refactoring/improving legacy scientific software
    \item Engineering design for sustainable software
    \item Metrics for the success of scientific software
    \item Adaptation of mainstream software practices for scientific software
\end{itemize}

\item Professionalization
\begin{itemize}
    \item Career paths
    \item RSE as a brand
    \item RSE outside of the UK or Europe
    \item Increase incentives in publishing, funding and promotion for better software
\end{itemize}

\item Training
\begin{itemize}
    \item Training for developing sustainable software
    \item Curriculum for software sustainability
\end{itemize}

\item Credit
\begin{itemize}
    \item Making the existing credit and citation ecosystem work better for software
    \item Future credit and citation ecosystem
    \item Software contributions as a part of tenure review
    \item Case studies of receiving credit for software contributions
    \item Awards and recognition that encourage sustainable software
\end{itemize}

\item Software publishing
\begin{itemize}
    \item Journals and alternative venues for publishing software
    \item Review processes for published software
\end{itemize}

\item Software discoverability/reuse
\begin{itemize}
    \item Proposals and case studies
\end{itemize}

\item Reproducibility and testing
\begin{itemize}
    \item Reproducibility in conferences and journals
    \item Best practices for code testing and code review
\end{itemize}

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote}\label{sec:keynote}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Lightning talks} \label{sec:lightning}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{comment}
\note{
\href{http://wssspe.researchcomputing.org.uk/wssspe4/agenda/}{Slides.}}
\end{comment}

Intro and description
%
\begin{enumerate}
\item \textbf{Names: \emph{title}}.
Description

\end{enumerate}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Panel discussion} \label{sec:panel}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{description of panel discussion on best practices}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Working groups} \label{sec:WGs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix}

\input{wg_open_research_index}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In WSSSPE4, we

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledgments} \label{sec:acks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% NSF grant for WSSSPE4
This material is based upon work supported by the National Science
Foundation under grant ACI-1648293, and by the Alfred P.~Sloan Foundation
under grant G-2016-7214.

%
\todo{feel free to add stuff here}

\newpage
\appendix
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Organizing committee}  \label{sec:orgcom}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Do we want email addresses here?}

The following is the list of WSSSPE4 organizers.

{\scriptsize
\begin{longtable}{lll}
\input{orgcom}
\end{longtable}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Attendees}  \label{sec:attendees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The following is a list of attendees at WSSSPE4.

{\scriptsize
\begin{longtable}{lll}
\input{participants}
\end{longtable}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Travel award recipients}  \label{sec:awardees}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\todo{Do we want email addresses here?}
The following is the list of WSSSPE4 travel award recipients.

{\scriptsize
\begin{longtable}{lll}
\input{awardees}
\end{longtable}
}

%\input{wg_appendix_best_practices}

\bibliographystyle{vancouver}

\bibliography{wssspe}
\end{document}

### 2016-09-14 WSSSPE4_report/awardees.tex
added superscripts for students and early-career awardees

@@
 Wolfgang Bangerth & Colorado State University, USA \\
-Roman Bauer & Newcastle University, UK \\
+Roman Bauer\textsuperscript{\textdagger} & Newcastle University, UK \\
 Steven Brandt & Louisiana State University, USA \\
-Emily Chen & University of Illinois Urbana-Champaign, USA \\
+Emily Chen* & University of Illinois Urbana-Champaign, USA \\
 Bruce Childers & University of Pittsburgh, USA \\
-Michael R.~Crusoe & Common Workflow Language Project \\
-Stephan Druskat & Humboldt-Universit\"{a}t zu Berlin, Germany \\
+Michael R.~Crusoe\textsuperscript{\textdagger} & Common Workflow Language Project \\
+Stephan Druskat\textsuperscript{\textdagger} & Humboldt-Universit\"{a}t zu Berlin, Germany \\
 Anshu Dubey & Argonne National Laboratory, USA \\
-Iain Emsley & Oxford University, UK \\
-Christopher Gwilliams & Cardiff University, UK \\
+Iain Emsley* & Oxford University, UK \\
+Christopher Gwilliams\textsuperscript{\textdagger} & Cardiff University, UK \\
 Lorraine Hwang & University of California Davis, USA \\
@@
 Frank Löffler & Louisiana State University, USA \\
-Tengyu Ma & Vanderbilt University, USA \\
+Tengyu Ma* & Vanderbilt University, USA \\
 Robert H.~McDonald & Indiana University, USA \\
-Jonah Miller & Perimeter Institute, Canada \\
+Jonah Miller* & Perimeter Institute, Canada \\
 Daniel Mosse & University of Pittsburgh, USA \\
-Kyle E.~Niemeyer & Oregon State University, USA \\
-Santiago N\'{u}\~{n}ez-Corrales & University of Illinois Urbana-Champaign, USA \\
+Kyle E.~Niemeyer\textsuperscript{\textdagger} & Oregon State University, USA \\
+Santiago N\'{u}\~{n}ez-Corrales* & University of Illinois Urbana-Champaign, USA \\
 Dominic Orchard & University of Kent, UK \\
-Francisco Queiroz & Pontifical Catholic University of Rio de Janeiro, Brazil \\
-Jean Salac & University of Virginia, USA \\
+Francisco Queiroz* & Pontifical Catholic University of Rio de Janeiro, Brazil \\
+Jean Salac* & University of Virginia, USA \\
 Janos Sallai & Vanderbilt University, USA \\
-Eric Seidel & University of California San Diego, USA \\
-Prabhjyot Singh Saluja & Southwest Oklahoma State University, USA \\
-Karan Shah & Georgia Institute of Technology, USA \\
-Sameer Shende & University of Oregon, USA
+Prabhjyot (Singh) Saluja* & Southwest Oklahoma State University, USA \\
+Eric Seidel* & University of California San Diego, USA \\
+Karan Shah* & Georgia Institute of Technology, USA \\
+Sameer Shende* & University of Oregon, USA

### 2016-09-14 WSSSPE4_report/awardees.tex
changed * to superscript in awardees list

--- 2016-09-14 WSSSPE4_report/awardees.tex
@@
 Wolfgang Bangerth & Colorado State University, USA \\
-Roman Bauer & Newcastle University, UK \\
+Roman Bauer\textsuperscript{\textdagger} & Newcastle University, UK \\
 Steven Brandt & Louisiana State University, USA \\
-Emily Chen & University of Illinois Urbana-Champaign, USA \\
+Emily Chen\textsuperscript{*} & University of Illinois Urbana-Champaign, USA \\
 Bruce Childers & University of Pittsburgh, USA \\
-Michael R.~Crusoe & Common Workflow Language Project \\
-Stephan Druskat & Humboldt-Universit\"{a}t zu Berlin, Germany \\
+Michael R.~Crusoe\textsuperscript{\textdagger} & Common Workflow Language Project \\
+Stephan Druskat\textsuperscript{\textdagger} & Humboldt-Universit\"{a}t zu Berlin, Germany \\
 Anshu Dubey & Argonne National Laboratory, USA \\
-Iain Emsley & Oxford University, UK \\
-Christopher Gwilliams & Cardiff University, UK \\
+Iain Emsley\textsuperscript{*} & Oxford University, UK \\
+Christopher Gwilliams\textsuperscript{\textdagger} & Cardiff University, UK \\
 Lorraine Hwang & University of California Davis, USA \\
@@
 Frank Löffler & Louisiana State University, USA \\
-Tengyu Ma & Vanderbilt University, USA \\
+Tengyu Ma\textsuperscript{*} & Vanderbilt University, USA \\
 Robert H.~McDonald & Indiana University, USA \\
-Jonah Miller & Perimeter Institute, Canada \\
+Jonah Miller\textsuperscript{*} & Perimeter Institute, Canada \\
 Daniel Mosse & University of Pittsburgh, USA \\
-Kyle E.~Niemeyer & Oregon State University, USA \\
-Santiago N\'{u}\~{n}ez-Corrales & University of Illinois Urbana-Champaign, USA \\
+Kyle E.~Niemeyer\textsuperscript{\textdagger} & Oregon State University, USA \\
+Santiago N\'{u}\~{n}ez-Corrales\textsuperscript{*} & University of Illinois Urbana-Champaign, USA \\
 Dominic Orchard & University of Kent, UK \\
-Francisco Queiroz & Pontifical Catholic University of Rio de Janeiro, Brazil \\
-Jean Salac & University of Virginia, USA \\
+Francisco Queiroz\textsuperscript{*} & Pontifical Catholic University of Rio de Janeiro, Brazil \\
+Jean Salac\textsuperscript{*} & University of Virginia, USA \\
 Janos Sallai & Vanderbilt University, USA \\
-Eric Seidel & University of California San Diego, USA \\
-Prabhjyot Singh Saluja & Southwest Oklahoma State University, USA \\
-Karan Shah & Georgia Institute of Technology, USA \\
-Sameer Shende & University of Oregon, USA
+Prabhjyot (Singh) Saluja\textsuperscript{*} & Southwest Oklahoma State University, USA \\
+Eric Seidel\textsuperscript{*} & University of California San Diego, USA \\
+Karan Shah\textsuperscript{*} & Georgia Institute of Technology, USA \\
+Sameer Shende\textsuperscript{*} & University of Oregon, USA

### 2016-09-14 WSSSPE4_report/WSSSPE4_report.tex
added description of superscripts in awardees table
added mention of RSE re: date & location

@@
 (WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
-held on 12--14 September 2016 in Manchester, England, UK. Previous events in
-the WSSSPE series are
+held on 12--14 September 2016 in Manchester, England, UK.
+This location and date was selected so that WSSSPE4 immediately preceded the
+First Conference of Research Software Engineers.
+Previous events in the WSSSPE series include
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
@@
 %\todo{Do we want email addresses here?}
-The following is the list of WSSSPE4 travel award recipients.
+The following table contains the list of WSSSPE4 travel award recipients, where
+\textsuperscript{*} and \textsuperscript{\textdagger} indicate students and
+early-career researchers, respectively.
 

### 2016-09-14 WSSSPE4_report/WSSSPE4_report.tex
added section for survey response discussion
added files and mention of program committee
fixed Dan's affiliation

@@
 %
-\thanks{{}$^{(1)}$ \hspace{-1ex}Computation Institute,
+\thanks{{}$^{(1)}$ \hspace{-1ex}National Center for Supercomputing Applications (NCSA) \&
+Electrical and Computer Engineering (ECE) Department \&
+School of Information Sciences (iSchool),
 University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Survey responses \label{sec:survey}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{discuss survey responses}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions} \label{sec:conclusions}
@@
 In WSSSPE4, we
+
+\todo{discuss Twitter activity}
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Program committee}  \label{sec:progcom}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The following table lists the WSSSPE4 program committee.
+
+{\scriptsize
+\begin{longtable}{lll}
+\input{program_committee}
+\end{longtable}
+}
+
 %\input{wg_appendix_best_practices}

### 2016-09-15 WSSSPE4_report/WSSSPE4_report.tex
various updates

@@
 
-\title[]{Report on the Fourth Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE4) workshop}
+\title[]{Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}
 
@@
 \begin{abstract}
-This report records and discusses the Fourth Working towards Sustainable
-Software for Science: Practice and Experiences (WSSSPE4) workshop. The report
+This report records and discusses the Fourth Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE4). The report
 includes a description of the keynote presentation of the workshop,
@@
 This location and date was selected so that WSSSPE4 immediately preceded the
-First Conference of Research Software Engineers.
+First Research Software Engineers (RSE) Conference.
 Previous events in the WSSSPE series include
@@
 Sustainable Software for Science: Practice and Experiences'' to better reflect
-the movement towards a community in addition to the associated workshop series.
+the movement towards a community in addition to the associated workshop series.\katznote{really?
+I must have missed that.
+I remember the first workshop as being `working towards' and then future workshops as being `workshop on'.}
 
@@
 with funds provided to the conference organizers by the National Science
-Foundation (NSF), the Alfred P.~Sloan Foundation, and the Software
+Foundation (NSF), the Gordon and Betty Moore Foundation, the Alfred P.~Sloan Foundation, and the Software
 Sustainability Institute (SSI) to support the travel of potential participants
@@
 This material is based upon work supported by the National Science
-Foundation under grant ACI-1648293, and by the Alfred P.~Sloan Foundation
+Foundation under grant ACI-1648293, by the Gordon and Betty Moore Foundation under gift \todo{check to see if there's a number}, and by the Alfred P.~Sloan Foundation
 under grant G-2016-7214.

### 2016-09-15 WSSSPE3/report/WSSSPE3_review.tex
moving long discussions of groups from appendices to new section 6, and changing my affiliation

@@
 %
-\thanks{{}$^{(1)}$ \hspace{-1ex}Computation Institute, 
-University of Chicago \& Argonne National Laboratory, Chicago, IL, USA; d.katz@ieee.org}
+\thanks{{}$^{(1)}$ \hspace{-1ex}National Center for Supercomputing Applications (NCSA) \&
+Electrical and Computer Engineering (ECE) Department \&
+School of Information Sciences (iSchool),
+University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org}
 %
@@
 presented at the meeting (\S\ref{sec:lightning}). The report also gives summaries of
-action plans proposed by the working groups (\S\ref{sec:WGs}), and provides some
-conclusions (\S\ref{sec:conclusions}). The first three appendices contain lists of the organizing committee
+action plans proposed by the working groups (\S\ref{sec:WGs}), 
+then gives longer
+descriptions of the activities that occurred in each of the working groups that
+made substantial progress
+(\S\ref{sec:group_details}), and provides some
+conclusions (\S\ref{sec:conclusions}). The appendices contain lists of the organizing committee
 (Appendix~\ref{sec:orgcom}), the registered attendees
 (Appendix~\ref{sec:attendees}), and the travel award recipients
-(Appendix~\ref{sec:awardees}). Finally, the report includes longer
-descriptions of the activities that occurred in each of the working groups that
-made substantial progress
-(Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}).
+(Appendix~\ref{sec:awardees}).
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Working Groups} \label{sec:WGs}
+\section{Working Groups: Summaries} \label{sec:WGs}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 all but one group, 
-more detailed notes on each group's discussions can be found in Appendices~\ref{sec:appendix_best_practices}--\ref{sec:appendix_user_community}.
+more detailed notes on each group's discussions can be found in \S\ref{sec:group_details}.
 
@@
 \input{wg_main_user_community}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Working Groups: Details} \label{sec:group_details}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\input{wg_appendix_best_practices}
+\input{wg_appendix_funding_spec_expert}
+%\input{wg_appendix_catalogs}
+\input{wg_appendix_industry_interaction}
+%\input{wg_appendix_legacy_SW}
+\input{wg_appendix_eng_design}
+\input{wg_appendix_metrics}
+\input{wg_appendix_training}
+\input{wg_appendix_software_credit}
+\input{wg_appendix_publishing_SW}
+\input{wg_appendix_user_community}
 
@@
 
-\input{wg_appendix_best_practices}
-\input{wg_appendix_funding_spec_expert}
-%\input{wg_appendix_catalogs}
-\input{wg_appendix_industry_interaction}
-%\input{wg_appendix_legacy_SW}
-\input{wg_appendix_eng_design}
-\input{wg_appendix_metrics}
-\input{wg_appendix_training}
-\input{wg_appendix_software_credit}
-\input{wg_appendix_publishing_SW}
-\input{wg_appendix_user_community}
+\newpage
 

### 2016-09-15 WSSSPE3/report/wg_appendix_best_practices.tex
moving long discussions of groups from appendices to new section 6, and changing my affiliation

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Best Practices Group Discussion}
+\subsection{Best Practices Group Discussion}
 \label{sec:appendix_best_practices}
@@
 
-\subsection{Group Members}
+\subsubsection{Group Members}
 
@@
 
-\subsection{Summary of Discussion}
+\subsubsection{Summary of Discussion}
 
@@
 
-\subsection{Description of Opportunity, Challenges, and Obstacles}
+\subsubsection{Description of Opportunity, Challenges, and Obstacles}
 
@@
 
-\subsection{Key Next Steps}
+\subsubsection{Key Next Steps}
 
@@
 
-\subsection{Plan for Future Organization}
+\subsubsection{Plan for Future Organization}
 
@@
 
-\subsection{What Else is Needed?}
+\subsubsection{What Else is Needed?}
 
@@
 
-\subsection{Key Milestones and Responsible Parties}
+\subsubsection{Key Milestones and Responsible Parties}
 \begin{itemize}
@@
 
-\subsection{Description of Funding Needed}
+\subsubsection{Description of Funding Needed}
 We might need funding for a journal publication (open-access options).

### 2016-09-15 WSSSPE3/report/wg_appendix_industry_interaction.tex
moving long discussions of groups from appendices to new section 6, and changing my affiliation

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration Working Group Discussion}
+\subsection{Transition Pathways to Sustainable Software: Industry \& Academic Collaboration Working Group Discussion}
 \label{sec:appendix_industry_interaction}
@@
 
-\subsection{What Else is Needed?}
+\subsubsection{What Else is Needed?}
 Nothing at the moment.
 
-\subsection{Key Milestones and Responsible Parties}
+\subsubsection{Key Milestones and Responsible Parties}
 
@@
 
-\subsection{Description of Funding Needed}
+\subsubsection{Description of Funding Needed}
 

### 2016-09-15 WSSSPE3/report/WSSSPE3_review.tex
a few more changes to explain the restructured paper

@@
 
+This section captures detailed reports from each working group that made significant
+progress.  Each subsection records the discussion of
+a group, as written by that group at that time (in the first person and in the present\slash future tense.)
+Thus, the subsections are records of what the groups did and planned, as of the end of the WSSSPE3 workshop.
+
 \input{wg_appendix_best_practices}

### 2016-09-15 WSSSPE4_report/WSSSPE4_report.tex
added descriptions of the two tracks

@@
 \begin{quote}
-    call for participation
+    \textbf{Track 1 -- Building a sustainable future for open-use research
+    software} has the goals of defining a vision of the future of open-use
+    research software, and in the workshop, initiating the activities that are
+    needed to get there. The idea of this track is to first think about where
+    we want to be 5 to 10 years from now, without being too concerned with
+    where we are today, and then to determine how we can move to this future.
+
+    \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
+    has the goal of improving the quality of today's research software and the
+    experiences of its developers by sharing practices and experiences.
+    This track is focused on the current state of scientific software and what
+    we can do to improve it in the short term, starting with where we are today.
 \end{quote}

### 2016-09-15 WSSSPE4_report/WSSSPE4_report.tex
added draft section on Code of Conduct

@@
 
+This workshop also introduced a Code of Conduct (CoC)
+\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
+for both WSSSPE4 and interactions in the greater WSSSPE community (e.g., on
+Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
+FORCE11 conference CoC~\cite{FORCE11:CoC}, in turn based on the Code4Lib
+CoC~\cite{Code4Lib:CoC}.
+The main guidelines of the CoC are:
+\begin{quote}
+    WSSSPE events are community events intended for networking and collaboration
+    as well as learning. We value the participation of every member of the
+    community and want all attendees to have an enjoyable and fulfilling
+    experience. Accordingly, all attendees are expected to show respect and
+    courtesy to other attendees throughout the event and in interactions online
+    associated with the event.
+
+    The WSSSPE event organizers are dedicated to providing a harassment-free
+    experience for everyone, regardless of gender, gender identity and
+    expression, age, sexual orientation, disability, physical appearance,
+    body size, race, ethnicity, religion (or lack thereof), technology choices,
+    or other group status.
+
+    To make clear what is expected, everyone taking part in WSSSPE events and
+    discussions---speakers, helpers, organizers, and participants---is required
+    to conform to the following Code of Conduct.
+
+    \begin{itemize}
+    \item Communicate appropriately for a professional audience including
+    people of many different backgrounds. Sexual language and imagery are not
+    appropriate for any event.
+
+    \item Be kind to others. Do not insult or put down other attendees. Be
+    mindful of jargon, which can sometimes exclude others from engaging in the
+    discussion.
+
+    \item Behave professionally. Remember that harassment and sexist, racist,
+    ageist, or exclusionary behavior are not appropriate.
+    \end{itemize}
+\end{quote}
+
+The CoC was introduced at the beginning of WSSSPE4, with the CoC subcommittee
+and a general email address introduced for reporting concerns or incidents, or
+asking questions.
+\todo{mention any/lack of violations/reports?}
+
 This report is based on
@@
 at \todo{what aimed at?}
-as the calls for participation said:
 

### 2016-09-17 WSSSPE4_report/WSSSPE4_report.tex
minor changes, adding papers section

@@
 held in Boulder, Colorado, USA.
-Note that the WSSSPE series changed its name from ``Workshop on Sustainable
-Software for Science: Practice and Experiences'' to ``Working towards
-Sustainable Software for Science: Practice and Experiences'' to better reflect
-the movement towards a community in addition to the associated workshop series.\katznote{really?
-I must have missed that.
-I remember the first workshop as being `working towards' and then future workshops as being `workshop on'.}
+Note that the first WSSSPE workshop was named
+``Working towards
+Sustainable Software for Science: Practice and Experiences,'' which remains the meaning
+of the WSSSPE group, but the workshops after that were named 
+``Workshop on Sustainable
+Software for Science: Practice and Experiences.'' Together these reflect
+that WSSSPE is both a community in addition to a set of associated workshops.
 
@@
 and a general email address introduced for reporting concerns or incidents, or
-asking questions.
-\todo{mention any/lack of violations/reports?}
+asking questions.  There was one concern mentioned to the CoC subcommittee
+after the first 1/2 day,
+regarding how part of the workshop was being run, and we changed the workshop
+to address this.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Papers and demos} \label{sec:papers}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Intro and description \katznote{not sure enum is right here}
+%
+\begin{enumerate}
+\item \textbf{Names: \emph{title}}.
+Description
+
+\end{enumerate}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Lightning talks} \label{sec:lightning}
@@
 
-Intro and description
+Intro and description \katznote{not sure enum is right here}
 %

### 2016-09-17 WSSSPE4_report/WSSSPE4_report.tex
more changes

--- 2016-09-15 WSSSPE4_report/WSSSPE4_report.tex
@@
 Kyle E.\ Niemeyer$^{(2)}$, 
+Author order to be determined based on contributions
 }
@@
 (WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
-held on 12--14 September 2016 in Manchester, England, UK.
+held over 2 1/2 days on 12--14 September 2016 in Manchester, England, UK.
 This location and date was selected so that WSSSPE4 immediately preceded the
@@
 held in Boulder, Colorado, USA.
-Note that the WSSSPE series changed its name from ``Workshop on Sustainable
-Software for Science: Practice and Experiences'' to ``Working towards
-Sustainable Software for Science: Practice and Experiences'' to better reflect
-the movement towards a community in addition to the associated workshop series.\katznote{really?
-I must have missed that.
-I remember the first workshop as being `working towards' and then future workshops as being `workshop on'.}
+Note that the first WSSSPE workshop was named
+``Working towards
+Sustainable Software for Science: Practice and Experiences,'' which remains the meaning
+of the WSSSPE group, but the workshops after that were named 
+``Workshop on Sustainable
+Software for Science: Practice and Experiences.'' Together these reflect
+that WSSSPE is both a community and a set of workshops.
 
@@
 and a general email address introduced for reporting concerns or incidents, or
-asking questions.
-\todo{mention any/lack of violations/reports?}
+asking questions.  There was one concern mentioned to the CoC subcommittee
+after the first half day,
+regarding how part of the workshop was being run, and we changed the workshop
+to address this.
 
-This report is based on
-\todo{what report is based on}
+This report is based on the events at the workshop and the submitted materials.
+There events included discussion of the WSSSPE mission and vision, a keynote,
+a set of presentations, a panel, and a number of working groups.  About a day of the
+workshop was spent with participants in the working groups, which occurred in parallel
+with each other.
 
@@
 WSSSPE4 was based on the work done in WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
-at \todo{what aimed at?}
-
-two tracks:
+at producing working groups that better continued working after the workshop ended.
+In addition, based on feedback after WSSSPE3, it became clear that WSSSPE attendees
+had two different motivations in participating.  One motivation was to make a better future
+for research software, and the other was to immediately do better research software development.
+This led to the idea of WSSSPE4 being partially divided into two tracks:
 
@@
 
-\todo{update}
-The calls for participation requested papers describing lightning talks,
-where each author could make a brief statement about work that either had been
-done or was needed;
+The call for participation requested 
 idea papers, which present implementable proposals related to Track 1;
@@
 practices and experiences and how they have been used to improve the quality of
-today’s research software and/or the experiences of its developers;
-and demos, brief papers describing a 10--15 minute demonstration of a tool or
-process relevant to Track 2 that improves the quality of today's research
-software and\slash or the experiences of its developers.
+today's research software and/or the experiences of its developers;
+demos, brief papers describing a tool or
+process that would be demonstrated, relevant to Track 2 that improves the quality of today's research
+software and\slash or the experiences of its developers; and
+extended abstracts describing lightning talks,
+where each author could make a brief statement about work that either had been
+done or was needed.
 
@@
 were accepted, as discussed further in Section~\ref{sec:lightning}.
-\todo{update}
+\todo{update, add paper stats too}
 
@@
 
+\todo{need to mention/link to the proceedings somewhere, though maybe not here.}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{need a person to write this}
+
+The keynote was given by Patricia Lago and entitled ``\todo{title}''.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Mission and vision}\label{sec:mission}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Papers and demos} \label{sec:papers}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{need a person to write this}
+
+Intro and description \katznote{not sure enum list is right here}
+%
+\begin{enumerate}
+\item \textbf{Authors: \emph{title}}.
+Description
+
+\end{enumerate}
 
@@
 
-Intro and description
+\note{need a person to write this}
+
+Intro and description \katznote{not sure enum list is right here}
 %
 \begin{enumerate}
-\item \textbf{Names: \emph{title}}.
+\item \textbf{Authors: \emph{title}}.
 Description
@@
 
-\todo{description of panel discussion on best practices}
+\todo{description of panel discussion on best practices} \note{asked Simon to write}
 
@@
 
-\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix}
+\todo{describe process for assembling people into working groups: vision, gaps, projects exercise, then soapboxes}
+
+\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix} \todo{though think about if an appendix is the right thing if we decide to publish in JORS again}
 
@@
 
-\todo{discuss survey responses}
+\note{need a person to write this - Dan asked Lorraine}
+
+\todo{discuss survey responses; maybe put detailed responses in an appendix}
 

### 2016-09-18 WSSSPE4_report/WSSSPE4_report.tex
adding authors, some assignments

@@
 Kyle E.\ Niemeyer$^{(2)}$, 
+Sandra Gesing$^{(3)}$, 
+Lorraine Hwang$^{(4)}$, 
 Author order to be determined based on contributions
@@
 %
+\thanks{{}$^{(3)}$ \note{info needed}}
+%
+\thanks{{}$^{(4)}$ \note{info needed}}
 
@@
 
+\note{Lorraine and Dan will write this}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-\note{need a person to write this}
+\note{Sandra Gesing will write this}
 
@@
 
-\note{need a person to write this - Dan asked Lorraine}
+\note{Lorraine will write this}
 

### 2016-09-18 WSSSPE4_report/WSSSPE4_report.tex
adding moore gift number

@@
 This material is based upon work supported by the National Science
-Foundation under grant ACI-1648293, by the Gordon and Betty Moore Foundation under gift \todo{check to see if there's a number}, and by the Alfred P.~Sloan Foundation
-under grant G-2016-7214.
+Foundation (ACI-1648293), by the Gordon and Betty Moore Foundation  (GBMF\#5620), and by the Alfred P.~Sloan Foundation (G-2016-7214).
 

### 2016-10-05 WSSSPE3/report/WSSSPE3_appendix.tex

--- 2016-09-15 WSSSPE3/report/WSSSPE3_review.tex
@@
 
-\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)}
+\title[]{Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3): Appendices}
 
@@
 \begin{abstract}
-This report records and discusses the Third Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE3). The report includes a
-description of the keynote presentation of the workshop, which served as an overview of sustainable
-scientific software. It also summarizes a set of lightning
-talks in which speakers highlighted to-the-point lessons and challenges
-pertaining to sustaining scientific software. The final and main contribution of the report is a summary of the
-discussions, future steps, and future organization for a set of self-organized
-working groups on topics including developing pathways to funding scientific
-software; constructing useful common metrics for crediting software
-stakeholders; identifying principles for sustainable software engineering
-design; reaching out to research software organizations around the world; and
-building communities for software sustainability. For each group, we include a
-point of contact and a landing page that can be used by those who want to join
-that group's future activities. The main challenge left by the workshop is to
-see if the groups will execute these activities that they have scheduled, and
-how the WSSSPE community can encourage this to happen.
+This document contains three appendices to the report on the Third Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE3): Appendix A, The list of organizers; Appendix B, the list of
+attendees; and Appendix C, the list of travel award recipients.
 
@@
 \maketitle
-\newpage
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Introduction} \label{sec:intro}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-The Third Workshop on Sustainable Software for Science: Practice and Experiences
-(WSSSPE3)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}} was
-held on 28--29 September 2015 in Boulder, Colorado, USA. Previous events in the
-WSSSPE series are
-WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with SC13;
-WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
-focused workshop organized jointly with the SciPy
-conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
-WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
-held in conjunction with SC14; and
-WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
-focused workshop organized again jointly with
-SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}}.
-
-Progress in scientific research is dependent on the quality and accessibility of
-software at all levels. Hence it is critical to address challenges related to
-development, deployment, maintenance, and overall sustainability of reusable
-software as well as education around software practices. These challenges can be
-technological, policy based, organizational, and educational; and are of
-interest to developers (the software community), users (science disciplines),
-software-engineering researchers, and researchers studying the conduct of
-science (science of team science, science of organizations, science of science
-and innovation policy, and social science communities). The WSSSPE1 workshop
-engaged a broad scientific community to identify challenges and best practices
-in areas of interest to creating sustainable scientific software. WSSSPE2
-invited the community to propose and discuss specific mechanisms to move towards
-an imagined future for software development and usage in science and
-engineering. But WSSSPE2 did not have a good way to enact those mechanisms, or to
-encourage the attendees to follow through on their intentions.
-
-The WSSSPE3 workshop included multiple mechanisms for participation and
-encouraged team building around solutions. WSSSPE3 strongly encouraged participation
-of early-career scientists, postdoctoral researchers, graduate students,  
-early-stage researchers, and those from underrepresented groups,
-with funds provided to the conference organizers by the Moore Foundation, the
-National Science Foundation (NSF), and the Software Sustainability Institute (SSI) to
-support the travel of potential participants who would not otherwise be able to
-attend the workshop. These
-funds allowed 16 additional people to attend and participate.
-
-WSSSPE3 also included two professional
-event organizers/facilitators from Knowinnovation who helped the organizing committee members
-plan the workshop agenda, and during the workshop, they actively engaged
-participants with various
-tools, activities, and reminders.
-
-This report is based on collaborative notes taken during the workshop, which
-were linked from the GitHub issues that represented the potential and actual
-working
-groups\footnote{\url{https://github.com/issues?q=label\%3A\%22WSSSPE3+activity\%22}}.
-Overall, the report discusses the organization work done before the workshop
-(\S\ref{sec:preworkshop}), the keynote (\S\ref{sec:keynote}), and the lightning talks
-presented at the meeting (\S\ref{sec:lightning}). The report also gives summaries of
-action plans proposed by the working groups (\S\ref{sec:WGs}), 
-then gives longer
-descriptions of the activities that occurred in each of the working groups that
-made substantial progress
-(\S\ref{sec:group_details}), and provides some
-conclusions (\S\ref{sec:conclusions}). The appendices contain lists of the organizing committee
-(Appendix~\ref{sec:orgcom}), the registered attendees
-(Appendix~\ref{sec:attendees}), and the travel award recipients
-(Appendix~\ref{sec:awardees}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Calls for Participation} \label{sec:preworkshop}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-WSSSPE3 was based on the work done in WSSSPE1 and WSSSPE2, but aimed at starting
-a process to make progress in sustainable software, as the calls for
-participation said:
-
-\begin{quote} The WSSSPE1 workshop engaged the broad scientific community to
-identify challenges and best practices in areas relevant to sustainable
-scientific software. WSSSPE2 invited the community to propose and discuss
-specific mechanisms to move towards an imagined future practice of software
-development and usage in science and engineering. WSSSPE3 will organize
-self-directed teams that will collaborate prior to and during the workshop to
-create vision documents, proposals, papers, and action plans that will help the
-scientific software community produce software that is more sustainable,
-including developing sustainable career paths for community members. These teams
-are intended to lead into working groups that will be active after the workshop,
-if appropriate, working collaboratively to achieve their goals, and seeking
-funding to do so if needed. \end{quote}
-
-The first call for participation requested lightning talks, where each author
-could make a brief statement about work that either had been done or was needed,
-with the goal of contributing to the discussion of one or more working groups.
-There were~24 lightning talks submitted; after a peer-review process, 16 
-were accepted, as discussed further in Section~\ref{sec:lightning}.
-
-The first call also discussed the potential action topics that came out of
-WSSSPE2, and requested additional suggestions. The combination of existing and
-new topics led to the following 18 potential topics that were advertised in the
-subsequent calls for participation:
-%
-%\begin{quote}
-\begin{itemize}
-\renewcommand{\labelenumi}{\textbf{\theenumi}.}
-\setlength{\rightmargin}{1em}
-
-\item Development and Community
-\begin{itemize}
-\item Writing a white paper/review paper about best practices in developing
-sustainable software
-\item Documenting successful models for funding specialist expertise in software
-collaborations
-\item Creating and curating catalogs for software tools that aid sustainability
-(perhaps categorized by domain, programming languages, architectures, and/or
-functions, e.g., for code testing, documentation)
-\item Documenting case studies for academia/industry interaction
-\item Determining effective strategies for refactoring/improving legacy
-scientific software
-\item Determining principles for engineering design for sustainable software
-\item Create a set of guidance giving examples of specific metrics for the
-success of scientific software in use, why they are chosen, what they are
-useful to measure, and any challenges/pitfalls; then publish this as a white
-paper
-\end{itemize}
-
-\item Training
-\begin{itemize}
-\item Writing a white paper on training for developing sustainable software, and
-coordinating multiple ongoing training-oriented projects
-\item Developing curriculum for software sustainability, and ideas about where
-such curriculum would be presented, such as a summer training institute
-\end{itemize}
-
-\item Credit
-\begin{itemize}
-\item Hacking the credit and citation ecosystem (making it work, or work better,
-for software)
-\item Developing a taxonomy of contributorship/guidelines for including software
-contributions in tenure review
-\item Documenting case studies of receiving credit for software contributions
-\item Developing a system of awards and recognitions to encourage sustainable
-software
-\end{itemize}
-
-\item Publishing
-\begin{itemize}
-\item Developing a categorization of journals that publish software papers
-(building on existing work), and case studies of alternative publishing
-mechanisms that have been shown to improve software discoverability/reuse, e.g.,
-popular blogs/websites
-\item Determining what journals that publish software paper should provide to
-their reviewers (e.g., guidelines, mechanisms, metadata standards)
-\end{itemize}
-
-\item Reproducibility and Testing
-\begin{itemize}
-\item Building a toolkit that could allow conference organizers to easily add a
-reproducibility track
-\item Documenting best practices for code testing and code review
-\end{itemize}
-
-\item Documentation
-\begin{itemize}
-\item Develop landing pages on the WSSSPE website (or elsewhere) that enable the
-community to easily find up-to-date information on a WSSSPE topic (e.g.,
-software credit, scientific software metrics, testing scientific software).
-\end{itemize}
-
-\end{itemize}
-%\end{quote}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote}\label{sec:keynote}
-%\note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-WSSSPE3 began with a keynote speech delivered by Professor Matthew Turk from the
-Department of Astronomy, University of Illinois, titled \emph{Why Sustain
-Scientific Software?}. Turk is a prolific scientific software practitioner and
-has extensive experiences working on large collaborative projects employing
-modern computing tools \cite{2011ApJS..192....9T}. He also co-organizes and
-champions WSSSPE events.
-
-In his keynote address, Turk recapped the course of development of WSSSPE
-workshops over the past few years, alongside his career development from a
-postdoc to an academic. The first WSSSPE workshop was at the Supercomputing
-conference (SC13) in 2013, but he observed that the notion of sustainable
-scientific software drew in an audience beyond supercomputing. In the following
-year, WSSSPE1.1 at SciPy had speakers talking about how software has been
-sustained inside the scientific Python community. WSSSPE2 at SC14 had breakout
-group discussions coming up with actionable items, and WSSSPE2.1 at SciPy 2015
-was similar. Turk noted the different atmosphere of the surrounding large
-conferences, despite similar WSSSPE participants.
-
-WSSSPE3 left the traditional Supercomputing Conference environment this year,
-and in Turk's words, this change spoke to the fact that scientific software
-comes from many different types of inquiries, deployment, strategies for
-maintenance, users, and ways of measuring the value of a piece of software. It
-appeared to Turk that the supercomputing community generally adopts some
-top-down approaches, whereas the SciPy community more often than not uses more
-bottom-up systems. According to Turk, there is a divergence in views about
-progress in software: the
-supercomputing community thinks that software is getting harder, with exascale
-computing and optimization issues in mind; but the SciPy community thinks that
-software is becoming better, with emerging tools such as Jupyter and
-productivity packages for research workflows. Admitting such comparisons are
-somewhat unfair generalizations, Turk reminded the audience that the different
-approaches bring different types of ideas to the table, and he welcomed WSSSPE3
-being conducted outside existing preconceptions.
-
-Returning to the topic of his talk, Turk invited the audience to picture
-scientific software as a flower on a landscape under the Sun, which may
-represent a number of measurable factors such as number of citations; growth of
-a community and number of contributors; amount of funding; prestigious prizes
-awarded; stability of the community in terms of leadership transitions, serving
-community needs, not breaking test suites, and performance on new architectures.
-But all these metrics are strictly speaking \emph{proxies} for the values and
-the impact scientific software bears. What we can measure does not give us
-direct insight---it just gives us proxies of insight.
-
-Turk then moved onto various different definitions of sustainability. His
-favorite one was ``keeping up with bug reports,'' where even if no new features
-were added, the software remains sustainable. Another definition of
-sustainability Turk mentioned was ``adding of new features,'' or ``maintaining
-the software for a long period of time'' such as the cases of \TeX\ or \LaTeX\
-with community help. A notion Turk heard often at supercomputing conferences was
-that sustainable software ``continues to work on new architectures.'' Yet
-another metric was ``people continuing to be able to learn how to use and apply
-the software.'' A funder Turk heard talked about sustainability as ``continuing
-to get funded.'' Turk also recalled that Greg Wilson, among others, said in
-WSSSPE1.1 that his view of sustainable software was software that ``continued to
-give the same results over time.'' A last measure of sustainability Turk
-presented was ``the ability to transition between different people developing
-and using a piece of software.''
-
-At WSSSPE1, several models were presented for ensuring sustainability. Turk
-considered that a familiar one was a funded piece of software where an external
-agency provided funds to a group who are not necessarily exclusively working on
-and developing the software, keeps it going, and provides it to the scientific
-community. The model of productized software, in which a piece of software has
-grown to the point that research groups or people are willing to support it with
-some amount of funding, for instance, a subscription to use cloud services that
-deploy a piece of software, or purchase of a piece of software. A final model
-Turk felt conflicted about is a volunteer model that is traditional
-old-school---not modern-day open source---development.
-
-Turk discussed whether productizing scientific software was synonymous with
-being sustainable and self-sufficient. He thought it was not necessarily the
-case and furthermore, it could lead to a divergence of interests between users
-and developers.
-
-Turk reminded the audience that the volunteer model means unpaid labor. On this
-note, he recommended Ashe Dryden's blog post on the ethics of unpaid labor and
-the open source software
-community\footnote{\url{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}}.
-Oftentimes, a person funded to work full time on a scientific project can spend
-a small amount of time for working on a piece of software necessary for that
-project. However, researchers' abilities to participate in that volunteer
-community are not always the same and may not always be aligned with their
-research projects. From Turk's experience, we cannot always rely on unpaid labor
-and volunteer time to sustain a piece of software---this came down to the
-notions of the top-down and the bottom-up approaches, i.e., the funded versus
-the grassroots. However, Turk pointed out that bottom-up, volunteer-driven
-projects can be just as large-scale as a top-down software development project.
-
-Turk said that sustaining scientific software really meant to him conducting
-scientific inquiries, often by some specific software, and sustaining the people
-we care about, our careers, and the future of our fields. According to Turk, we
-all have an invested stake in sustaining scientific software. Hence, having
-``sustained'' projects can suffocate new projects, so we need to make sure we
-don't cause novel ideas and packages to suffer at the hands of the status quo.
-
-Turk talked about possible reasons why we want to sustain scientific software:
-devotion to science and interests in pursuing the next stage of research; fun
-and creative thrill in writing codes and papers; usefulness with measurable
-impacts, for example, LINPACK and HDF groups providing data storage to
-satellites, which goes beyond usefulness to necessity. Lastly, Turk presented
-his wishlist of questions to be answered in the future:
-%
-\begin{itemize}
-
-\item How do we ship a product on time when dealing with a mix of funding models
-and motivations especially when we rely on volunteers?
-
-\item How do we know when it is time to end some software and move on? For
-example, should we stop sustaining Python and switch to Julia and Javascript?
-
-\item How can productized software balance its future versus its past, or the
-emerging needs of the customers versus the existing needs of the development
-community?
-
-\item How can we help avoid burnout and retain the joy in the communities?
-
-\item How can we reduce systemic bias, which goes back to Dryden's blog post
-especially on how ethics of unpaid labor disproportionately affect
-underrepresented communities?
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Lightning Talks} \label{sec:lightning}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{comment}
-\note{
-\href{http://wssspe.researchcomputing.org.uk/wssspe3/agenda/}{Slides.}}
-\end{comment}
-
-After the keynote, WSSSPE3 continued with lightning talks.
-These short talks were intended to give an opportunity for attendees to
-quickly highlight an important issue or a potential solution.
-%
-\begin{enumerate}
-\item \textbf{Benjamin Tovar and Douglas Thain: \textit{Freedom vs.\ Stability:
-Facilitating Research Training While Supporting Scientific Research}}. Benjamin
-Tovar presented a case study of the Cooperative Computing Lab (CCL)\footnote{\url{http://ccl.cse.nd.edu}} at the
-University of Notre Dame, which is a small group of individuals whose main tasks
-are collaborating with people that have large-scale computing problems,
-operating various parallel computer systems, conducting computer science
-research, and developing open source software. One of the main challenges they
-face is finding a balance between flexibility/training and stability/quality.
-Their current solution for ensuring the latter was to add a software engineer
-(the presenter) to the existing team of faculty and students, who now also
-serves as a ``spring'' between flexibility and stability.
-
-\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
-Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
-Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
-Sustainability}}. The concept of sustainability has become a topic of interest in
-the field of computing, which is evidenced by the increase in the number of
-events that focus on the topic. Nevertheless, it is not well understood yet.
-Birgit Penzenstadler argued that we often define sustainability too narrowly.
-Instead, sustainability at its heart is a systemic concept and must be viewed
-from a range of different dimensions including environmental, economic,
-individual, social, and technical. She introduced the Karlskrona Manifesto on
-Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
-of related work on the topic of sustainability into a set of (mis-)perceptions
-and principles. The manifesto does not proclaim that there is an easy,
-one-size-fits-all solution around the corner, but rather points out that
-sustainability is a ``wicked problem'' and is often misunderstood. Due to these
-misperceptions, even though sustainability's importance is increasingly
-recognized, many software systems are unsustainable. Even more alarming is that
-most software systems' broader impacts on sustainability are unknown. To change
-this, the Karlskrona Manifesto proposes nine principles and commitments. These
-commitments are not dogmatic laws, but rather commitments to rethink, to move
-beyond the silo mentality, and to analyze in more depth. As such, they do not
-restrict, but rather open up a space for discussion.
-
-\item \textbf{Abani Patra, Hossein Aghakhani, Nikolay Simakov, Matthew Jones,
-and Tevfik Kosar: \textit{Integrating New Functionality Using Smart Interfaces
-to Improve Productivity of Legacy Tools}}. Abani Patra presented an example of
-how the community using Titan2D, a geoflow simulation software, increased the
-productivity of their tools by improving both code and data layout~\cite{Patra20051}. The main
-obstacles in this change were the non-existence of a common version control
-system for the source code, coupled with multiple versions of the same code
-base, the fixed format of input files, that many input values were set as
-compilation flags, and that the internal data layout was not suitable for modern
-technologies (e.g., vectorization, accelerators). The approaches of the Titan2D
-developers included reinforcing the code structure using multiple layers of
-Python and C++ interfaces, and a redesign of the data layout to be more suitable
-for modern CPUs and accelerators.
-
-\item \textbf{Abigail Cabunoc Mayes, Bill Mills, Arliss Collins, and Kaitlin
-Thaney: \textit{Collaborative Software Development as Sustainable Software:
-Lessons from Open Source}}. Abigail Cabunoc Mayes combines two properties of open
-source that, together, create a suitable habitat for sustainable software. The
-first of the two properties, public, does not only mean public code. It also
-includes public discussions, a public process of including contributions, and an
-open license. The second property, participatory, stresses the importance of
-reaching out to the community and helping potential new members by providing
-better documentation and learning experiences, like code review and examples of
-good first bug reports. Together, Abigail concluded, these two properties not
-only lead to higher quality, reusability, and ease of understanding, but also
-eventually, to sustainability.
-
-\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
-through Best Practices in Open Source Software: Computational Infrastructure for
-Geodynamics}}. Lorraine Hwang presented experiences with the Computational
-Infrastructure for Geodynamics (CIG)\footnote{\url{http://geodynamics.org/}}, a community software with a worldwide user
-base. Like others, their main goals include high usability, sustainability, and
-reproducibility. As a means to achieve these goals, various communication
-channels have been developed, such as mailing lists, wikis, workshops,
-hackathons, tutorials, and webinars. In order to contribute to the
-infrastructure, codes must adhere to specified minimum standards with the desire
-that all codes are working toward target standards.
-%, but ideals are also given to encourage reaching the aforementioned goals. 
-These include, e.g., the use of version control, certain coding styles, the
-presence and nature of code tests and documentation, and certain user workflows.
-
-\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
-Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
-in the Computational Infrastructure for Geodynamics}}. In a second talk, Lorraine
-Hwang mentioned a variety of ways to cite efforts within the SAGA framework\footnote{\url{https://geodynamics.org/cig/projects/saga/}}, including
-science papers, code papers, user manuals, and the CIG website. An analysis of
-the resulting citations showed that $80\%$ of papers that use CIG codes mention
-the code name, and about the same number includes a citation. Only about $20\%$
-acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
-cite codes (including non-CIG codes), and only about one eighth specify the
-version used. Compared to other codes, CIG seems to be much better cited. In
-part, this is attributed to the fact that CIG requires that donated software
-provide a citable paper specified in the User Manual. The project is working on
-tools and methods to generate attribution information automatically.
-%This,
-%in part, is attributed to suggested citation forms communicated to the
-%community, and runtime tools that help collecting a relevant citation list
-%automatically.
-
-\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
-Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
-Open Science (DASPOS)}}. DASPOS\footnote{\url{http://daspos.org}} is an NSF-funded multi-disciplinary effort,
-located at Notre Dame and Chicago, that links the high energy physics effort to
-other disciplines such as biology, astrophysics, and digital curation. It includes
-physicists, digital librarians, as well as computer scientists, and aims to
-achieve some commonality across disciplines. Examples are meta-data descriptions
-of archived data, computational descriptions, descriptions of how data was
-processed, questions such as whether computation replication can be automated,
-and what the impact of access policies on the preservation infrastructure is.
-One of the products of this effort is a suite of tools that deals with this
-preservation~\cite{Meng2015137,Meng:2015:UPE:2755979.2755982,1742-6596-664-3-032022,nd4},
-and the questions was posed whether that software itself is
-sustainable. Points that were brought up included the need for a user community
-depending on a given software, and the need to provide added value for its
-users. An important method to achieve this was to work with the user community
-from the start, and to budget that way. For the specific example of the
-preservation software, this means that besides adding value to the community, it
-needs to be transparent to their workflows, i.e., not requiring additional
-effort to preserve ``research objects.''
-
-\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
-Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
-Susheel Varma: \textit{An update from UK Research Software Engineers}}. James
-Hetherington started by listing some of the problems research software faces,
-which include poor standard of verification and low levels of reuse. For a long
-time, technical solutions to such technical problems were focus of the eResearch
-community, including research software distributions, grids, middleware and
-workflows. Some limited adoption can be seen today in research communities, but
-the main problems have not been solved to a sufficient level. Hetherington
-hypothesized that instead of technical solutions, social innovation is needed: a
-new role in the academic system focused on research software that combines the
-best parts of a craftsperson and a scholar~\cite{rse1,rse2}. However, social innovation in
-centuries-old institutions is hard. Alternatives to such a new role would have
-to include rewards for good research software, recognition of software as
-academic output, and rejecting submissions based on irreproducible computational
-results. Some advantages of research software engineering (RSE) groups include
-the possibility of training in reproducible computational research, providing
-collaborations for researchers who do not want to serve as programmers, and
-creating synergies with research computing platforms. The success of such a
-group would be measured by the output and quality of the research software.
-Members could be part of Research Computing or faculty. They would not be
-independent researchers, but would have to have a research background. An
-attempt to form a community of RSE groups within the UK has been underway for
-several years, including funding from the UK Research Council. An open question
-is whether this approach can be adopted in other countries, including the~USA.
-
-\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
-the User into Building Sustainable Software for Science}}. Dan Gunter's main
-question was, ``What is needed to develop sustainable software?''. Beyond the
-usual suspects of funding, proficient developers, good design, and software
-engineering practices, Dan placed the users. He explained this using a
-traditional software-development model starting from gathering requirements, and
-reaching release through a design, development, and testing phase. The main
-deficiency with this approach was pointed out to be the too-late interaction
-with users. Instead, an alternative approach was proposed that, at first, skips
-the development phase and repeatedly goes through requirement gathering, design
-and user interaction/learning phases, and only eventually starts development
-once an agreement is reached, leading to an increased user satisfaction, higher
-adoption, and eventually to sustained software.
-
-\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
-Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
-Platform for Predictive Biology: Lessons Learned on the KBase Project}}. 
-Dan Gunter presented experiences and lessons learned as part of KBase\footnote{\url{http://kbase.us}}, an
-open software and data platform for addressing the grand challenge of systems
-biology: predicting and designing biological function. KBase is a unified system
-that integrates data and analytical tools for comparative functional genomics of
-microbes, plants, and their communities. However, it is also a collaborative
-environment for sharing methods and results, and placing those results in the
-context of knowledge in the field. Being a large, multi-institutional project,
-one of the big challenges is to agree on standards to enable a single,
-maintainable system. Working in isolation does not work (anymore) within this
-field, and the community in the field also does not have standards for software
-engineering. This is contrasted to computer science research, where software
-engineering standards shorten design cycles, leading to more time for highly
-rewarded activities like publishing, performance studies, graduating students,
-or protecting ideas before publication. Instead of this more traditional
-approach, KBase uses a variation on the ``Scrum'' methodology. After picking
-projects and team members, four to five teams work on projects for about two
-weeks before a one-week evaluation by the executive committee. Based on their
-assessment new teams and projects might be chosen before iteratively restarting
-the ``agile'' development cycle. This process is intentionally open and
-documented.
-
-\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
-Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
-Scientists in Best Practices of Software Sharing}}. Erin Robinson presented an
-approach to overcome some hurdles in scientific publishing: disconnects among
-experimental data, research software, and publications. A current effort in the
-Geosciences is the ``Geoscience Paper of the Future,''\footnote{\url{http://www.ontosoft.org/gpf}} which includes four
-elements~\cite{ESS2:ESS284}. First, it forms a modern paper including text, data, and pointers to
-supplementary materials. Second, it is reproducible, including data processing,
-workflow, and visualization tools. Third, it is part of open science, which
-includes being publicly available under open licenses, and providing meta-data.
-Last but not least, it uses digital scholarship elements like persistent
-identifiers for software, data, and authors, and it cites both data and
-software. OntoSoft is a tool for helping with this effort, providing software stewardship
-for the Geosciences. As part of this initiative, a special issue of a journal in
-Geoscience areas is planned to include only Geoscience papers of the future,
-with submissions open until the end of 2015~\cite{GPFI}. In addition, training sessions are
-provided to geoscientists to learn best practices in software and data sharing,
-provenance documentation, and scholarly publication.
-
-\item \textbf{Neil Chue Hong: \textit{Building a Scientific Software
-Accreditation Framework}}. Neil Chue Hong presented a proposal to build a
-scientific software accreditation framework. One of the aims of such a framework
-would be to measure how ``good'' a given piece of software is, and to evaluate
-how this can be effectively measured in the first place. This can be compared to
-the effort of the standardized, easy to read, and understandable nutritional
-labeling of food, which only contain a small set of categories. However, such a
-framework for software would be more difficult due to different existing
-community norms. The challenges such a framework faces include that many
-measurements are subjective, that many metrics are too costly, and that
-self-assessment needs to be encouraged. Possible categories would include
-availability, usability, transferability, among others. Such a framework could
-enable both improvement of specific software, as well as comparisons of similar
-software. An accreditation by such a framework could then be part of software
-management plans, ensuring that software is accessible and reusable throughout
-the proposed project and beyond.
-
-\item \textbf{Jeffrey Carver: \textit{On the Need for Software Engineering
-Support for Sustainable Scientific Software}}. Carver argued that for scientific
-software to be truly sustainable, there is a need for developers to use
-appropriate software engineering practices. His experience interacting with
-scientific teams indicates that choosing and tailoring these practices is not a
-trivial exercise~\cite{Nanthaamornphong-Carver:2015, Heaton-Carver:2015, Basili-etal:2008, 4222616}.
-There is a general culture clash between software engineering
-and science that hinders our ability to communicate and choose appropriate
-methods. In addition, many experienced scientific software developers appear to
-be unaware of software engineering practices that may be beneficial to them. The
-most appropriate software engineering practices are those that are lightweight,
-properly tailored, and focus on the key software development problems faced by
-scientists. In order to increase the use of software engineering in science, we
-need more documented success stories. These successes need to be socialized 
-%\choinote{not sure if ``socialize'' can be used in such a way}
-within the scientific community through workshops like the Software Engineering
-for Science workshop
-series\footnote{\href{http://www.SE4Science.org/workshops}{http://www.SE4Science.org/workshops}}
-and the new Software Engineering track in \textit{Computing in Science and
-Engineering} magazine.
-
-\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}.
-This talk highlighted an attempt to solve the common problem for open source
-development: it is difficult to collect information about how many people use
-particular software, how often, which version, which parts of the software, or
-on which operating system. Current solutions include surveys, but these have
-high uncertainties. A different approach is based on automatic ``call-backs''
-that collect these information at runtime and send it to a central place for
-analysis. Problems with this approach include obtaining agreement from the user,
-legal issues, increased maintenance (of servers), ethical questions, and also the
-lack of a common infrastructure. Some of these problems are of a social nature
-and have to be solved as such, but the last problem (a missing common
-infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
-which is developing common APIs and a library implementation for common,
-repeating tasks such as obtaining user consent. Results are uploaded not
-directly to project servers, but to neutral third parties that only publish
-aggregated statistics to projects.
-
-\item \textbf{Alice Allen: \textit{We're giving away the store! (Merchandise not
-included)}}. Alice Allen described the Astrophysics Source Code
-Library\footnote{\url{http://ascl.net}}, an increasingly used way to obtain a unique ID for astrophysics
-software that is indexed by indexing services and can be cited~\cite{wssspe2_allen_jors}.
-ASCL offers
-clones of existing infrastructure, provides server space and computing
-resources, shares innovations, and permits moves elsewhere. Users provide a
-domain name, then control and configure their site and use it as intended,
-gather their codes as they wish, share innovations, and protect the provided
-computing environment.
-
-\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
-Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
-Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
-Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
-Developing Community}}. The effort behind this talk given by Ray Idaszak started
-from a breakout group at an NSF SI2 workshop in 2015, and centers around
-building a framework for creating metrics for scientific software~\cite{ahalt-csessp}. This
-framework would improve both the metrics and the software it evaluates, and
-could also serve as a tool for building a community around the idea. With
-especially the last-mentioned idea (building a community) in mind, a software
-``peer review group'' would be created, representing stakeholders who will
-self-review software created by their respective communities, and will
-concurrently develop metrics. The whole project should be community-governed,
-without a single institution overseeing the activities or infrastructure, with
-the hope to evolving community-generated and adopted standards. The generation
-of metrics would be tied to the actual evaluation of software, creating an
-incentive by improving the evaluated software itself during this process. The
-framework code would provide infrastructure for the creation of metrics and
-evaluation, and forums for generation of software success metrics. It would also
-support code reviews of the evaluated software. An open question is whether it
-is possible to fit the resulting metrics in a common template. So far, this is
-still in a design phase, with a white paper at the CSESSP workshop 2015 and this
-talk, but the WSSSPE workshops are seen as a forum for the community to assemble
-and act, and is planned to be used also in the future to build this community
-and framework.
-
-\end{enumerate}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Working Groups: Summaries} \label{sec:WGs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-After the keynote and lightning talks, the workshop facilitators led an exercise to create working
-groups for the rest of the meeting.  The
-attendees first suggested additional topics beyond those in the call for participation (as listed in
-\S\ref{sec:preworkshop}).  One topic that was introduced at this point was ``Building Sustainable
-User Communities for Scientific Software.''  The full set of topics was then placed on
-flipchart-sized pieces of paper around the walls, and the attendees voted on which topics
-they were strongly interested in working on at WSSSPE3, and which topics they were 
-generally interested in contributing to, but not as strongly.  This led to some topics being
-taken out of the mix for the rest of the workshop, since not enough people wanted to contribute
-to them to lead to a useful discussion.  Additionally, some topics were combined by the
-participants who felt they were closely linked and that a group could address multiple of them
-in a single discussion.
-
-After this, the attendees broke up into small working groups to discuss
-the remaining topics during most of the remaining 1 1/2 days.  A high-level summary
-of each topic and group's work can be found in the subsections of this section, and for
-all but one group, 
-more detailed notes on each group's discussions can be found in \S\ref{sec:group_details}.
-
-Midway through the first afternoon and between the two days, each group
-had a chance to talk about what progress they had made.  As discussed below (in \S\ref{sec:legacy}), the group that formed to discuss Legacy Software dissolved after the first session
-as group members left to join other groups.
-In addition, in the morning of the
-second day, a small set of reviewers/advisors (external to the groups but chosen by the group
-members) visited each group to listen to what they group was planning and to provide feedback.
-
-%\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix - done in line for Software Credit group already}
-
-\input{wg_main_best_practices}
-\input{wg_main_funding_spec_expert}
-%\input{wg_main_catalogs}
-\input{wg_main_industry_interaction}
-\input{wg_main_legacy_SW}
-\input{wg_main_eng_design}
-\input{wg_main_metrics}
-\input{wg_main_training}
-\input{wg_main_software_credit}
-\input{wg_main_publishing_SW}
-\input{wg_main_user_community}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Working Groups: Details} \label{sec:group_details}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\input{wg_appendix_best_practices}
-\input{wg_appendix_funding_spec_expert}
-%\input{wg_appendix_catalogs}
-\input{wg_appendix_industry_interaction}
-%\input{wg_appendix_legacy_SW}
-\input{wg_appendix_eng_design}
-\input{wg_appendix_metrics}
-\input{wg_appendix_training}
-\input{wg_appendix_software_credit}
-\input{wg_appendix_publishing_SW}
-\input{wg_appendix_user_community}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Conclusions} \label{sec:conclusions}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-In WSSSPE3, we attempted to take what we learned from WSSSPE1 and WSSSPE2 in how
-we can collaboratively build a workshop agenda and turn that into an ongoing
-community activity. The success or failure of these efforts will
-only become apparent over time.
-
-The workshop had two components, presentations and working groups. The
-presentations, in the first half day of the workshop, included an inspirational
-keynote and a set of lightning talks. We used lightning talks for two reasons:
-first, the need of some participants to have a slot on the agenda to justify
-their attendance; and second, as a way to get new ideas across to all the
-attendees. We broke with the tradition of requiring the lightning talk
-submitters to self-publish their papers, and instead used a common peer-review
-platform\footnote{\url{http://easychair.org}}, choosing to publish their slides
-on the workshop website instead.
-
-The working groups met for a small part of the first half day and all of the
-second day, with the exception of some short periods for the groups to report
-back to the collected workshop attendees. Each group determined a set of
-activities that the members could do to advance sustainable software in a
-particular area.
-
-The results of these group sessions made it clear that there are many
-interlinked challenges in sustainable software, and that while these challenges
-can be addressed, doing so is difficult because they generally are not the
-full-time job of any of the attendees. As was the case in WSSSPE2 as well, the
-participants were willing to dedicate their time to the groups while they were
-at the meeting, but afterwards, they went back to their (paid) jobs.
-
-We need to determine how to tie the WSSSPE breakout activities to people's jobs,
-so that they feel that continuing them is a higher priority than it is now,
-perhaps through funding the participants, or through funding coordinators for
-each activity, or perhaps by getting the workshop participants to agree to a
-specific schedule of activities during the workshop as we have tried to do in
-WSSSPE3. It remains to be seen, however, if the participants will meet the
-schedules they set.
-
-The overall challenge left to the sustainable software community is perhaps one
-of organization: how to combine the small partial efforts of a large number of
-people to impact a much larger number of people: those who develop and use
-scientific software. While WSSSPE might help focus the actions of the groups,
-something more is needed to incentivize the wider community, which is a
-generalization of the sustainable software problem itself.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Acknowledgments} \label{sec:acks}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Work by Katz was supported by the National Science Foundation while working at
-the Foundation. Any opinion, finding, and conclusions or recommendations
-expressed in this material are those of the author(s) and do not necessarily
-reflect the views of the National Science Foundation.
-%
-Choi's work was supported in part by the National Science Foundation research
-grant DMS-1522687 and a WSSSPE3 travel award. She thanks the encouragement and
-discussion with Fred Hickernell.
-%
-Hetherington was funded by the Software Sustainability Institute, RCUK grants
-EP/H043160/1 and EP/N006410/1.
-%
-Work by Gunter was supported by the Office of Science, Office of Biological and
-Environmental Research, of the U.S.\ Department of Energy (DOE) under Award
-Numbers DE-AC02-05CH11231, DE-AC02-06CH11357, DE-AC05-00OR22725, and
-DE-AC02-98CH10886, as part of the DOE Systems Biology Knowledgebase and by the
-Office of Science, Office of Advanced Scientific Computing Research (ASCR) of
-the U.S.\ Department of Energy under Contract Number DE-AC02-05CH11231 as part
-of the Template Interfaces for Agile Parallel Data-Intensive Science (TIGRES)
-project.
-%
-%\todo{feel free to add stuff here}
 
@@
 
-\newpage
-
-\bibliographystyle{vancouver}
-
-\bibliography{wssspe}
 \end{document}

### 2016-10-05 WSSSPE3/report/registrants.tex

--- 2015-12-03 WSSSPE3/report/registrants.tex
@@
 Emily Chen & NCSA\\
-Sou Cheng Choi &  NORC at the University of Chicago \&  Illinois Institute of Technology\\
+Sou-Cheng Choi &  NORC at the University of Chicago \&  Illinois Institute of Technology\\
 Nancy Collins & NCAR\\

### 2016-10-05 WSSSPE3/report/wg_appendix_eng_design.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Engineering Design Group Discussion}
+\subsection{Engineering Design Group Discussion}
 \label{sec:appendix_eng_design}
@@
 
-\subsection{Description of Opportunity, Challenges, and Obstacles}
+\subsubsection{Description of Opportunity, Challenges, and Obstacles}
 The opportunity was identified to distill existing software engineering and
@@
 
-\subsection{Key Milestones and Responsible Parties}
+\subsubsection{Key Milestones and Responsible Parties}
 

### 2016-10-05 WSSSPE3/report/wg_appendix_funding_spec_expert.tex
A few more changes that seem to make sense.
last change

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Funding Research Programmer Expertise Group Discussion}
+\subsection{Funding Research Programmer Expertise Group Discussion}
 \label{sec:appendix_funding_spec_expert}
@@
 RSE losing their flexibility, becoming so engaged in one project that when that
-project ends, they find it hard to transfer. For this reason, we also recommend
-that 40 percent is ideal; two projects per developer, with some time for
+project ends, they find it hard to transfer. For this reason, we recommend
+that 40\% is ideal; two projects per developer, with some time for
 training and infrastructure work. Having two developers per project seems to be ideal,
@@
 
-\subsection{Plan for Future Organization and Future Needs}
+\subsubsection{Plan for Future Organization and Future Needs}
 

### 2016-10-05 WSSSPE3/report/wg_main_best_practices.tex

@@
 will attempt to distill the emerging body of knowledge into this document. The
-large number of articles from the NSF funded SI2 projects (SSE and SSI),
+large number of articles from the NSF-funded SI2 projects (SSE and SSI),
 ``lightning talks,'' ``white papers,'' and reports from different workshops have
@@
 For additional information about the discussion, see
-Appendix~\ref{sec:appendix_best_practices}.
+Section~\ref{sec:appendix_best_practices}.
 
@@
 established a timeline and a rough outline (see
-Appendix~\ref{sec:appendix_best_practices}) for the report.
+Section~\ref{sec:appendix_best_practices}) for the report.
 

### 2016-10-05 WSSSPE3/report/wg_main_metrics.tex

@@
 assess this, how are metrics monitored, and many more. A more complete bulleted
-list of these questions can be found in Appendix~\ref{sec:appendix_metrics}. Next, a roadmap for how to
+list of these questions can be found in Section~\ref{sec:appendix_metrics}. Next, a roadmap for how to
 proceed was discussed, including creating a set of milestones and tasks. The idea
@@
 magazine), or JORS (Journal of Open Research Software). More information about
-the group discussion is available in Appendix~\ref{sec:appendix_metrics}.
+the group discussion is available in Section~\ref{sec:appendix_metrics}.
 

### 2016-10-05 WSSSPE3/report/wg_main_publishing_SW.tex
changing subsections of 5 to not be first person

@@
 
-See Appendix~\ref{sec:appendix_publishing_SW} for more details about the
+See Section~\ref{sec:appendix_publishing_SW} for more details about the
 discussion.
@@
 The plan is to create and curate a web page describing executable papers, their value, and
-a list of what publishers support them. We expect the page to be available in
+a list of what publishers support them. The group expects the page to be available in
 early January of 2016 on the Software Sustainability Institute's website.

### 2016-10-05 WSSSPE3/report/wg_main_software_credit.tex

@@
 19, following the conclusion of WSSSPE3. See
-Appendix~\ref{sec:appendix_SW_credit} for more detailed discussion of related
+Section~\ref{sec:appendix_SW_credit} for more detailed discussion of related
 activities.
@@
 network either integrated with the existing paper citation ecosystem or
-complementary to it. See Appendix~\ref{sec:appendix_SW_credit} for a more
+complementary to it. See Section~\ref{sec:appendix_SW_credit} for a more
 detailed summary of the working group's discussion on these topics.

### 2016-10-05 WSSSPE4_proposal/WSSSPE4_proposal.tex

@@
 Awarded funding is expected to be reimbursement of allowable expenses (travel, lodging, sustenance) of up to \$1,500 for participants traveling from outside Europe to the UK, up to \$1,000 for participants traveling from Europe outside the UK to the UK, and up to \$500 for participants traveling from within the UK.
+We plan to support the travel of \textbf{at least 10 students}, which comprises \$15,000 of the overall request.
 
@@
 In all cases, while our ideal would be to have these demographics match the US or global population averages, our goal will be to significantly improve over academic STEM and tech industry norms.
+In addition, our goal is to support the travel of at least 10 students.
 
@@
 {\it Funds are requested to partially support travel for 25 participants and presenters (at \$1,500 per participant) from US institutions across the country from the US to Manchester, UK, and 3 participants/presenters (at \$1,000 each)  from US institutions who are already in Europe to travel to Manchester, and 2 participants/presenters  (at \$500 each)  from US institutions who are already in the UK to travel to Manchester.
-Reimbursable costs include airfare, hotel, and sustenance.}
+Reimbursable costs include airfare, hotel, and sustenance.
+Out of the 25 participants, 10 positions will be reserved for students.}
 

### 2016-10-05 WSSSPE4_report/awardees.tex

--- 2016-09-14 WSSSPE4_report/awardees.tex
@@
 Frank Löffler & Louisiana State University, USA \\
-Tengyu Ma & Vanderbilt University, USA \\
+Tengyu Ma\textsuperscript{*} & Vanderbilt University, USA \\
 Robert H.~McDonald & Indiana University, USA \\
-Jonah Miller & Perimeter Institute, Canada \\
+Jonah Miller\textsuperscript{*} & Perimeter Institute, Canada \\
 Daniel Mosse & University of Pittsburgh, USA \\
-Kyle E.~Niemeyer & Oregon State University, USA \\
-Santiago N\'{u}\~{n}ez-Corrales & University of Illinois Urbana-Champaign, USA \\
+Kyle E.~Niemeyer\textsuperscript{\textdagger} & Oregon State University, USA \\
+Santiago N\'{u}\~{n}ez-Corrales\textsuperscript{*} & University of Illinois Urbana-Champaign, USA \\
 Dominic Orchard & University of Kent, UK \\
-Francisco Queiroz & Pontifical Catholic University of Rio de Janeiro, Brazil \\
-Jean Salac & University of Virginia, USA \\
+Francisco Queiroz\textsuperscript{*} & Pontifical Catholic University of Rio de Janeiro, Brazil \\
+Jean Salac\textsuperscript{*} & University of Virginia, USA \\
 Janos Sallai & Vanderbilt University, USA \\
-Eric Seidel & University of California San Diego, USA \\
-Prabhjyot Singh Saluja & Southwest Oklahoma State University, USA \\
-Karan Shah & Georgia Institute of Technology, USA \\
+Prabhjyot (Singh) Saluja\textsuperscript{*} & Southwest Oklahoma State University, USA \\
+Eric Seidel\textsuperscript{*} & University of California San Diego, USA \\
+Karan Shah\textsuperscript{*} & Georgia Institute of Technology, USA \\
 Sameer Shende & University of Oregon, USA

### 2016-10-05 WSSSPE4_report/wg_open_research_index.tex

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Dan to write this}
+
+Introduction to group here, including the overall objective of work in this area.
+
+\subsubsection{Participants}
+
+members of the working group
+
+\subsubsection{Working group objective}
+
+Specific things the working group wants to accomplish in the context of the larger objective.
+
+\subsubsection{Gap or challenge}
+
+What is the gap or challenge being addressed?
+
+\subsubsection{Relevant people and resources}
+
+What people, groups, or resources are needed.
+
+\subsubsection{Plans}
+
+What tasks will the working group undertake
+
+\subsubsection{SMART steps}
+
+What are the first SMART steps proposed?
+
+\subsubsection{More information \& joining instructions}
+
+How could a reader get more information or get more involved?

### 2016-10-31 WSSSPE4_report/WSSSPE4_report.tex
added Kyle for lightning talk todo
adding initial guess for sections and section leads

@@
 
-\note{need a person to write this}
+\note{Kyle Niemeyer will write this}
 
@@
 
-\todo{go through subsections that have both content here and appendix, and add pointer from text in the subsection here to the corresponding appendix} \todo{though think about if an appendix is the right thing if we decide to publish in JORS again}
+Sections: \todo{copy wg\_open\_research\_index for each, add input command for each new file}
+
+\begin{itemize}
+\item Verifying best practices \& metrics for sustainable software [Ray Idaszak]
+\item Software Sustainability Alliance [Neil \& Jean]
+\item Scientific Software Prototyping Infrastructure (S2PI) [Santiago Nunez-Corrales] 
+\item CodeMeta working group [Alice Allen]
+\item White paper on developing sustainable software [Sandra Gesing]
+\item Social Science of Scientific Software [Stuart Geiger]
+\item Software development best practices [Jonah Miller]
+\item Meaningful metrics for sustainable software [Emily Chen]
+\item Coordinating access to CI for research software [Mark Abraham]
+\item Software engineering processes tailed for research software [Anshu Dubey]
+\item Open research index [Dan]
+\item Letters of evaluation for computational scientists [Wolfgang Bangerth]
+\end{itemize}
 

### 2016-10-31 WSSSPE4_report/WSSSPE4_report.tex
fixing case in open research index title; adding best practices section as confirmed by Ray
adding Software Best Practices for Undergraduates section; fixing other best practices section
adding mission info

@@
 
+\todo{write info about process}
+
+After this process, the final statements about WSSSPE are:
+
+{\bf Mission}
+WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
+
+{\bf Vision}
+We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
+
+{\bf Focus areas}
+WSSSPE promotes sustainable research software by positively impacting:
+\begin{itemize}
+\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
+\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
+\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
+\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
+\end{itemize}
+
+{\bf Definitions}
+
+Sustainable software has the capacity to endure such that it will continue to be available in the future, on new platforms, meeting new needs.
+
+The research software lifecycle includes:
+\begin{itemize}
+\item acquiring and assembling resources (including funding and people) into teams and communities
+\item developing software
+\item using software
+\item recognizing contributions to and of software
+\item maintaining software
+\end{itemize}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \begin{itemize}
-\item Verifying best practices \& metrics for sustainable software [Ray Idaszak]
+\item Verifying best practices \& metrics for sustainable research software [Ray Idaszak]
 \item Software Sustainability Alliance [Neil \& Jean]
@@
 \item Social Science of Scientific Software [Stuart Geiger]
-\item Software development best practices [Jonah Miller]
+\item Software Best Practices for Undergraduates [Jonah Miller]
 \item Meaningful metrics for sustainable software [Emily Chen]
@@
 
+\input{wg_best_practices_sustainability}
+\input{wg_best_practices_undergrads}
 \input{wg_open_research_index}

### 2016-11-04 WSSSPE4_report/WSSSPE4_report.tex
adding structure for CodeMeta working group
adding metrics section for Emily
adding three new report sections

@@
 
-Sections: \todo{copy wg\_open\_research\_index for each, add input command for each new file}
+\note{list of sections that is in the process of being confirmed}
+
+Sections:
 
@@
 \item Verifying best practices \& metrics for sustainable research software [Ray Idaszak]
-\item Software Sustainability Alliance [Neil \& Jean]
+\item \todo{to confirm} Software Sustainability Alliance [Neil \& Jean]
 \item Scientific Software Prototyping Infrastructure (S2PI) [Santiago Nunez-Corrales] 
-\item CodeMeta working group [Alice Allen]
+\item CodeMeta [Alice Allen]
 \item White paper on developing sustainable software [Sandra Gesing]
-\item Social Science of Scientific Software [Stuart Geiger]
+\item \todo{to confirm} Social Science of Scientific Software [Stuart Geiger]
 \item Software Best Practices for Undergraduates [Jonah Miller]
 \item Meaningful metrics for sustainable software [Emily Chen]
-\item Coordinating access to CI for research software [Mark Abraham]
-\item Software engineering processes tailed for research software [Anshu Dubey]
+\item \todo{to confirm} Coordinating access to CI for research software [Mark Abraham]
+\item Software engineering processes tailored for research software [Anshu Dubey]
 \item Open research index [Dan]
-\item Letters of evaluation for computational scientists [Wolfgang Bangerth]
+\item \todo{to confirm} Letters of evaluation for computational scientists [Wolfgang Bangerth]
 \end{itemize}
@@
 \input{wg_best_practices_sustainability}
+\input{wg_prototyping}
+\input{wg_codemeta}
+\input{wg_best_practices_developing}
 \input{wg_best_practices_undergrads}
+\input{wg_metrics}
+\input{wg_soft_eng}
 \input{wg_open_research_index}

### 2016-11-05 WSSSPE4_report/WSSSPE4_report.tex
adding mission and vision text, mostly from Lorraine

@@
 %
-\thanks{{}$^{(4)}$ \note{info needed}}
+\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
 
@@
 
-\note{Lorraine and Dan will write this}
+\note{by Lorraine and Dan}
 
-\todo{write info about process}
+Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
+The organizers developed a strawhorse, which was presented to the participants early in the meeting.
+The presentation included guidelines as well as examples from other similar communities.
+The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; Its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
 
-After this process, the final statements about WSSSPE are:
+The participants were given time after the presentation to write down their comments.
+Seven people volunteered to work on revising the mission and vision statements, incorporating these comments and presenting back to the community.
+Based on this feedback, the committee redrafted the mission and vision.
+The committee added focus areas to clarify the organizations breadth and to keep the mission and vision simple and long-lasting.
+Comments after the close of the meeting were incorporated into the statements.
+
+A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
+via the WSSSPE mailing list, Facebook group, and Twitter.
+After 2 weeks without suggested changes, the final statements, as shown below, were added
+to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
 

### 2016-11-05 WSSSPE4_report/WSSSPE4_report.tex
adding survey summary (section 9) and details (appendix E), content from Lorraine

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Survey responses \label{sec:survey}}
+\section{Attendee survey \label{sec:survey}}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Lorraine will write this}
+\note{from Lorraine}
 
-\todo{discuss survey responses; maybe put detailed responses in an appendix}
+All participants present on the last day of the meeting, September 14, 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
+44 responses were recorded, with one respondent indicating that they took the survey twice.
+The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
+Hence, the total number of unique individuals responding to the survey is likely 44.
+The survey contained 9 multiple choice and open-text answer questions.
+Text responses have been alphabetize to preserve anonymity.
+The survey had a 100\% completion rate.
+Appendix~\ref{sec:survey_details} contains the detailed survey results.
+
+In general, the respondents were highly satisfied with the meeting and interested in continuing and participating in WSSSPE activities.
+The mix of topics and types of interactions---talks, panels and discussion---was well balanced with several indicating a desire for time for questions and answers following talks as well for discussions in general.
+Many respondents indicated that they will remain engaged in WSSSPE working group initiatives.
+However, demand for a professional organization encompassing WSSSPE interests was weaker with over half interested in joining and a large percentage willing to consider the idea.
+Respondents were grateful for the opportunity to network, explore new collaborations and for travel support to the conference.
+
+The survey results indicate that future WSSSPE conferences should consider the balance of the attendees.
+This year's meeting had many first time and early career participants who would have benefited from a review of basic concepts and terminology including what it means for software to be sustainable and the roles of research software engineers.
+Additional emphasis and topics to explore would be the inclusion of more case studies, focus on the decision making process in developing and using software, deeper dives into selected topics, tutorials, software training within and outside of STEM, lowering the barriers to implementing best practices in software development, and progress towards executing WSSSPE's vision.
 
@@
 % NSF grant for WSSSPE4
-This material is based upon work supported by the National Science
-Foundation (ACI-1648293), by the Gordon and Betty Moore Foundation  (GBMF\#5620), and by the Alfred P.~Sloan Foundation (G-2016-7214).
+This material is based upon work supported by the National Science Foundation (ACI-1648293),
+by the Gordon and Betty Moore Foundation (GBMF\#5620),
+and by the Alfred P.~Sloan Foundation (G-2016-7214).
 
-%
 \todo{feel free to add stuff here}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Detailed survey results}  \label{sec:survey_details}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\input{survey_details}
+
 %\input{wg_appendix_best_practices}

### 2016-11-05 WSSSPE4_report/survey_details.tex
adding survey summary (section 9) and details (appendix E), content from Lorraine

Note: typos and spelling errors have been corrected, and spelling has been changed from UK standard to US standard in some cases, but no other changes have been made to the attendee responses.

\noindent {\bf 1. Please indicate how strongly you agree or disagree with the following statements}

\todo{make this table prettier! fix vertical spacing}

\begin{table}[h!]
\centering
\caption{General questions about the conference program}
\label{tab:survey_program}
{\scriptsize
\begin{tabular}{|l|cccccc|}
\hline
{\bf Question} &
{\bf strongly} &
{\bf somewhat}  &
{\bf agree} &
{\bf somewhat}  &
{\bf strongly}  &
{\bf Total} \\
&
{\bf disagree} &
{\bf disagree} &
&
{\bf agree} &
{\bf agree} &
\\ \hline
The plenary sessions had a good&
5 &
3 &
8 &
11 &
18 &
45 \\
\hspace{.2cm}mix of topics and speakers. &
 &
 &
 &
 &
 &
 \\
The balance was right between  &
6 &
2 &
7 &
9 &
21 &
45 \\
\hspace{.2cm}talks, panels, and lightning talks. &
&
 &
 &
 &
 &
 \\
There was sufficient time for discussion. &
6 &
6 &
8 &
9 &
16 &
45 \\
\hline
\end{tabular}
}
\end{table}

\noindent {\bf 2. What part of the meeting was the most valuable? select all that apply}

\begin{table}[h!]
\centering
\caption{Value of meeting parts}
\label{tab:survey_parts}
\begin{tabular}{|l|c|}
\hline
{\bf Answer} &
{\bf Count} \\ \hline
Plenary &
19 \\
Panel &
14 \\
Working Groups &
36 \\
Other &
5 \\
Total &
74 \\
\hline
\end{tabular}
\end{table}

\noindent {\bf 3. What additional topics would you like to see discussed at the next meeting?}

\begin{itemize}
\item Concrete reports about best practices
\item considerations of software licensing, build systems, delivery (source vs binary vs whatever) - but of course these are "later" issues... first one needs the software to not be hidden, to have version control, have a README, etc.
\item Container sustainability practices
\item education, training interns
\item Explicit follow-up at WSSSPE5 on all the activities that were created at WSSPE4.  These should be at least lightning talks, perhaps with their own session.
\item financial models for sustaining software
\item Frameworks for building software
\item have a short beginning presentation for each working group and a short ending presentation
\item How software development training can be adapted to different disciplines beyond STEM
\item How to low the barrier for non-tech heavy researchers to follow best software practices.
\item I would like to have more discussions about research, if people can share their research or some of the interesting projects that they're working on, this will help in knowing the problems in sustaining software.
\item I would like to learn more about sustainability issues in industry.
\item I would like to see more grounded advice and less blue-sky thinking. A lot of stuff in the talks was 'this would be nice' but there were not enough talks that were directly applicable to day-to-day programming.
\item Like to see more examples from the real world.  
\item Mechanisms of improving software reusability
\item More concrete projects, concrete possibilities and showcases for collaboration with results
\item Practical ways of achieving some of the goals rather than the discussion of more attempts on the writing of white papers
\item Reproducibility through containerization (I guess there is another list of best practices needed there)
\item RSE award, organizational questions regarding setting up RSE teams within universities
\item Software testing
\item Some theory on what makes for sustainable software, which is not the same as best practice for research software engineering. If this had been talked about before then consider that 80\% of us were new.
\item Specific scientific software sustainability factors
\item Testing methods that work. Ways of promoting organizational change.
\item The progress of our current project.
\item The use of scientific software at different stages of research / work 
\item We discussed scientific software. Should we also discuss software FOR scientists (like Dan's google scholar replacement)?
\end{itemize}

\noindent {\bf 4. What could be improved?}

\begin{itemize}
\item A bit less fly by night scheduling though things worked out just fine.  
\item A lot of the speakers just seemed to be there either to plug their own projects (e.g. Hydroshare, BioDynaMo) or spent the majority of their talk waffling (e.g. Manifesto for Personal Responsibility in the Engineering of Academic Software). Also, the icebreakers were not very useful.
\item A single room for the duration of the meeting. (Particular to this WSSSPE4, the "collaboratory" room with the monitors, was actually perfect for the full meeting.)
\item Allotting more time for questions \& discussions in between talks
\item close to perfect workshop - if I would change one thing that give more time for discussions
\item consistency of participants
\item felt locked in to the breakout group, not sure if that was intended or not
\item Finding rooms a bit confusing at the meeting outset
\item Having more time for the demos.
\item Having more time to discuss some of the talks
\item I was disappointed in the fruit plate because it only had melon.
\item I would like to be able to ask questions after talks. By the time we can discuss the talks, I've forgotten what I wanted to say
\item Identify a local pub/restaurant/cafe to which people can head afterwards for casual meetup for dinner/drinks/discussion/whatever
\item If feedback to people's projects and lighting talks etc is posted online, then we would have also benefits from those thoughts
\item It was a great event overall I feel like if we can bring some real life project integration with best practices that will be helpful 
\item it would be great to see some specific case studies/success stories
\item Keep the online agenda up to date. Better air-conditioning.
\item Less talks
\item more time for discussion and working groups
\item not sure, it looks great to me. I would like to see in plenary the perspective from closed software and open source software (for cross-fertilization)
\item Some longer talks that do deeper dives.  Good that the parallel track idea was abandoned so we could hear all talks.
\item The identification of what RSE and RS are, especially for newcomers
\item The keynote address could have been better targeted for this community beforehand.
\item The workshop was great!!!
\item There is lots of vision here, so we need to put more effort in our project to move forward.
\item Think about a little more planning on the BOF Dinners to make it easier to get groups together. Maybe a Google doc to encourage it etc.
\item To have a few speakers (2-3) that talk about concrete scientific software, showing the simulations and talking about their scientific results, and why they chose the software they did (as compared to other available ones)
\item Try and add some extra Q\&A time after talks (unless that'd mean less talks)
\item We could have additional time to work in working groups
\item WSSPE vision prior to start would have helped
\end{itemize}

\noindent {\bf 5. What would make you more likely to invest time in future WSSSPE initiatives?}

\begin{itemize}
\item A centralized place/organization to see and discuss WSSSPE initiatives. At the moment, WSSSPE organization is scattered across many places. One website with a list of activities and contributors would help.
\item A tutorial session.
\item Being able to contribute to something going forward, rather than a talking shop for the initiatives. 
\item connect to people
\item For meetings: again the availability of travel awards.
\item For myself, I anticipate looking for ways to tie WSSPE initiatives with my current projects such that the outcome benefits a/the broader community.
\item Free lunch and double coffee capacity? Also - seriously - less filler talks and filler activities.
\item Funded support resources
\item funding; focus on more concrete aspects of sustainability 
\item Having more time to invest in future WSSSPE initiatives
\item Having more time to invest in future WSSSPE initiatives; I'd definitely do more if I could. 
\item Honestly, having the time available in my own schedule (which I don't think is something WSSSPE organizers can change)
\item I already plan to do it :)
\item I'm not sure. I'm not very interested in fostering WSSSPE as an thing, but I am interested in WSSSPE as a means to network to find people to work on these issues with me. 
\item If I can see the value of the initiative to me and my value to the initiative
\item if I had funding to specifically work on that
\item If my daily workload would be reduced
\item if the community is and remains active
\item Keeping up your generous travel support alone is enough for me to not even think about missing any further WSSSPE events
\item Know that the work started in one edition can be presented on the following one.
\item Likelihood of useful new contributors.
\item more clearly defined groups
\item More followups between the conferences.
\item More funding opportunities to be able to devote appropriate attention to the topic
\item More senior participants with (i) experience, (ii) influence
\item More time. Some funding.
\item Potential a
\item Prospects for collaborations leading to citable artifacts, e.g. papers
\item some documented forward momentum between sessions
\item The networking is incredible for this meeting and should be encouraged to continue. Having published deliverables out of the community will help this as well.
\item The travel support really helps
\item This topic is related to a large component of what I envision as my professional career path. In particular, the project designed during the sessions will be fleshed out into a demo, and if adequate, into some form of larger piece of software.
\item To establish a metrics for research software.
\item Understanding the landscape of funding and potential for publication
\item Yes, definitely I met with some very interesting people through this platform and would like to be a part of it. It'll be an honor to collaborate to this initiative.
\end{itemize}

\noindent {\bf 6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}

\begin{table}[h!]
\centering
\caption{A WSSSPE society?}
\label{tab:survey_society}
\begin{tabular}{|l|c|}
\hline
{\bf Answer} &
{\bf Count} \\ \hline
Yes &
24 \\
No &
3 \\
Maybe &
18 \\
Total &
45 \\
\hline
\end{tabular}
\end{table}

\noindent {\bf 7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}

\begin{table}[h!]
\centering
\caption{Continuing working groups?}
\label{tab:survey_continuing_wgs}
\begin{tabular}{|l|c|}
\hline
{\bf Answer} &
{\bf Count} \\ \hline
Yes &
32 \\
No &
4 \\
Maybe &
9 \\
Total &
45 \\
\hline
\end{tabular}
\end{table}

\noindent {\bf 8. Working Group Name}

\begin{itemize}
\item "Citations / CodeMeta"  and  "Software Practices White Paper"
\item Best Practices White Paper and Best Practices Undergraduate Course
\item CodeMeta
\item CodeMeta
\item CodeMeta
\item CodeMeta, but I also supplied information to other working groups -- mostly connections to others -- that I hope will be of assistance.
\item Continuous CI for research software on uncommon hardware
\item Continuous Integration
\item Continuous integration for research software; Testing of scientific software
\item Curriculum for an undergraduate course on software best practices for domain scientists
\item Domain Specific Undergraduate Courses on Best Practices for Software Development
\item Domain Specific Undergraduate Courses on Best Practices for Software Development AND International Collaborations between institutions.
\item integration of software testing through Debian 
\item Meaningful Metrics for Scientific Software
\item Meaningful Metrics in Software
\item metadata
\item multiple
\item Open Research Index, Metrics, Social Science, and maybe others
\item Prototypes
\item research software studies
\item Scientific Software Prototyping Infrastructure (S2PI), WG-Best-Practices
\item Setting up Software Sustainability Alliance
\item Software Sustainability Alliance
\item Software Sustainability Alliance; Metrics; White Paper
\item sustainability metrics; best practices
\item Sustainable Software Alliance
\item Sustainable Software Alliance
\item Tenure Letters
\item Testing in Sciences
\item Testing in Scientific Software
\item Verify Best Practices and Metrics for Sustainable Software
\item Verify Best Practices Case Studies
\item White Paper WG, Open Research Index WG
\end{itemize}

\noindent {\bf 9. Do you have any other comments or suggestions or ideas?}

\begin{itemize}
\item As a student, coming to WSSSPE4 has been very rewarding experience. Sustainable software is not my particularly sub-field of interest, but it was very interesting to me to apply the skills I had learned in my research in a different field. It definitely taught me a different perspective and expanded my original perception of where my skills can be applied.
\item As a student, this has helped me explore the different computational research careers and has made me more aware of scientific software best practices.  I found the plenary talks particularly valuable for learning about different areas of study, and it was great how willing all of the WSSSPE participants were to share more and help each other. I also found it valuable that there were other undergraduates at this WSSSPE, and it created another small community and encouraged me to participate more.
\item buddy me-sheet and workshop planning and moderation was excellent
\item Don't start at 8.45am if you're asking people to socialize and connect in the evenings! Work on getting a better gender, geographical and discipline balance. The use of travel bursaries has helped (particularly in getting early career participants).
\item Great organization, enjoyed being here very much, it was surely a success!
\item Great to meet colleagues from the UK and Germany at this event and enjoyed this format for working, engaging, and producing with the community while at the workshop that was facilitated by KnowInnovation. 
\item I enjoyed the brainstorming session on Tuesday morning - well organized
\item I had hoped to have some discussion of building the sustainable communities beyond the software writing stage. The HPC focus, though expected slightly, was too heavy and I think ignores some of the other long term issues, such as preservation and addressing first principles for new technologies, or smaller, individual projects that will also support science. Perhaps it would be good to have a student session as well? (But I do like the integration). 
\item I think the mixing of students, staff, faculty etc in the working groups is very effective and will contribute to  outcomes outside of the working group. It was great to see that we have undergraduates leading working groups here. There should be more attention on diversity. 
\item I wish there were a project on training interns to work at academic environment as programmers
\item I would like to thank the committee for selecting me. As an undergraduate student this was a great experience I got to meet professors from the universities which I wanted to join. Now I exactly know how to prepare my application. I learned some really good technical skills and will definitely apply it to my research. So happy to be a part of this wonderful group. I will design a  logo for WSSSPE next year. Hope I can attend this event next year also.  
\item Is there something better than sharpies (few of running around the room with a permanent marker)? Travel Awardee: Discussions here will and have benefitted my organization and community.
\item It was great to connect with people doing research software engineering in various positions, contexts, and universities. I felt like I learned a lot. Moving between the two rooms was a bit confusing.
\item Maybe involve more researchers from low-income countries if funds are available.
\item no. I think the moderator and the format are both great.
\item Nothing comes to mind
\item Receiving a travel award was crucial for me to come to WSSSPE4 this time, especially because WSSSPE4 was held outside the US. On the other hand, I see the great benefit this had on the geographic diversity of the participants, so I do think that meetings not only in the US are of great value. And again, I would like to emphasize, that travel awards are crucial for this type of meeting, as most participants are not directly supported for this type of activity - one reason to have a WSSSP5. (on another note: have a multi-line text field for comments next time)
\item The issue of software sustainability is on par with the challenges of building scientific instruments. Part of what I am for my doctoral dissertation is finding better abstractions, which leads to sustainability through better mathematics. This workshop raised my awareness in several other aspects, which will be considered for this and all future projects I may be related to.
\item The travel award made it possible for me to come to WSSSPE-4 and broaden my horizons.  Many issues pertaining research software and sustainable software were uncovered, and WSSSPE-4 brought to light the importance of the issues.  I'd encourage other travel awards to people, like me, who do not belong to the community.
\item There aren't lots of researchers knowing such kind of concept, sustainable software. We need to let more RSE know it.
\item There is a lot of discussion, but it's hard to see what actually will come of it. The discussion of the 9 working groups from last year at the beginning of the workshop clearly showed this: IIRC, not one had produced anything, all seemed to be stalled somewhere. We ought to find ways to actually make things *happen* after we talk about it (and admittedly have many good ideas).
\item This has been my first WSSSPE, and as someone coming from the humanities with an interest in RSE it has been an enormously rewarding experience to be able to interact with and learn from individuals and initiatives that have been part of the greater discussion around WSSSPE topics more in both quantity and quality. For me personally, it made me aware that I really need to refactor a lot of my basic assumptions for a PhD thesis, but that's a great thing because it means that I was able to draw on the exchanges and experiences from WSSSPE already. All in all, the workshop has made me more determined to become more involved with this community. And last but not least i just wanna say THANK YOU for the generous travel support I've received, which meant that my PI and dept. didn't have to have any doubts whether I'd drain their resources which are meant to be used towards achieving a completely different research goal.
\item This was an excellent mix of people, timed/structured activities, and untimed events.
\item This was an excellent mix of people, timed/structured activities, and untimed events. I had the opportunity to talk to numerous people about projects, issues, and ways to improve science/scientific software, and leave with many new ideas and possible collaborations/cooperative efforts with complementary projects. The working groups activities were excellent, a good mix of projects, and I saw a lot of progress on several I'm very interested in. (Please group this survey with my earlier one; I hit "end" too soon on the other! Thanks!)
\item We have activity on slack, github, and twitter. And working groups have their own stuff. Can we index these in a single place?
\item What percentage of the working groups actually end up producing anything? It seems to me as though most come up with an idea and then find out they don't have the resources to implement it - nice as the ideas are. Also, I assume someone is going to attempt to reinvent metrics or best practices every single year, I guess that's unavoidable.
\item WSSPE seems to be evolving nicely from year to year.  I hope that in WSSSPE5 there are lots of results to report from the outcome of WSSSPE4 working groups.
\item You could try unconferences instead of tracks. 
\end{itemize}

### 2016-11-07 WSSSPE4_report/wg_codemeta.tex
adding structure for CodeMeta working group

--- 2016-10-05 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{CodeMeta}
+\label{sec:CodeMeta}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Dan to write this}
+\note{Alice to write this}
 
@@
 
+Something in here about this group having formed in reaction to one of the project suggestions made on the second day of the meeting.
+
+On the second day of WSSSPE4, participants were challenged to share their ideas for ideal future world, list gaps that exist between now and future possibilities, and generate possible projects to bridge existing gaps to move us closer to the ideal vision of attendees.
+One of the possible projects listed was to "Define a standard for metadata to be used in current repos including authors, dependencies, ...
+repo url." Several attendees have been participating in a project, CodeMeta, that seeks to do just that.
+Interest in this possible was very high and an enthusiastic group formed with the object of fostering adoption of CodeMeta.
+
 \subsubsection{Participants}
 
-members of the working group
+Gabrielle Allen <gdallen@illinois.edu>
+Stephan Druskat <stephan.druskat@hu-berlin.de>
+Iain Emsley <iain.emsley@oerc.ox.ac.uk>
+Carole Goble <carole.goble@manchester.ac.uk>
+Chris Gwilliams <gwilliamsc@cardiff.ac.uk>
+Rafael Jimenez <rafael.jimenez@elixir-europe.org>
+Frank Löffler <knarf@cct.lsu.edu>
+Kyle Niemeyer <Kyle.Niemeyer@oregonstate.edu>
+Thomas Robitaille <thomas.robitaille@gmail.com>
+Rob Welch <py12rw@leeds.ac.uk>
+Alice Allen <aallen@ascl.net>
 
@@
 
-Specific things the working group wants to accomplish in the context of the larger objective.
+Engage existing members with WSSSPE4 CodeMeta Working Group
+Crosswalk Schema.org types 
+Determine how to engage the community and provide ongoing social engagement and structure
+Assist in implementation of the specifications and brush up documentation 
+Define a list of services to be added and match terms (add to csv) 
+Provide an outsider’s view and contribute suggestions for more understandable docs 
+Find better ways to present crosswalk table
 
@@
 
+Original CodeMeta members
+Create mailing list for WSSSPE CodeMeta participants
+Engage through email
+
+Useful links
+\url{https://github.com/codemeta/codemeta}
+\url{https://github.com/codemeta/codemeta/milestones}
+\url{https://gitter.im/codemeta/codemeta}
+
 \subsubsection{Plans}
@@
 
-What are the first SMART steps proposed?
+Write to managers of CodeMeta project to start dialog with the two groups - Alice (done)
+Post questions generated by our discussion as issues on CodeMeta repo - Iain, Rafa, Stephan, others as needed (done)
+Crosswalk Schema.org types - Rafa, Carole
+Assist in implementation of the specs and brush up documentation - Rob
+Define a list of services to be added and match terms (add to csv) - Chris
+Identify two roles and information on CodeMeta that would be useful to these people for engaging with the project - Stephan (done)
+Think about better ways to present crosswalk table - Tom, Frank
+Create mailing list for WSSSPE CodeMeta participants - Kyle (30 Sep 16)
 
@@
 
-How could a reader get more information or get more involved?
+Because of the work done at WSSSPE4, the CodeMeta project README file\citep{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} was greatly expanded to include a description of the project that is geared to those with little or no prior knowledge of the project and including a list of contributors, information on how one can get involved, a bit of the project history and who is managing the project, and links to additional information.

### 2016-11-07 WSSSPE4_report/survey_details.tex

@@
 \item Concrete reports about best practices
-\item considerations of software licensing, build systems, delivery (source vs binary vs whatever) - but of course these are "later" issues...
+\item considerations of software licensing, build systems, delivery (source vs binary vs whatever) - but of course these are ``later'' issues...
 first one needs the software to not be hidden, to have version control, have a README, etc.
@@
 \item I would like to see more grounded advice and less blue-sky thinking.
-A lot of stuff in the talks was 'this would be nice' but there were not enough talks that were directly applicable to day-to-day programming.
+A lot of stuff in the talks was `this would be nice' but there were not enough talks that were directly applicable to day-to-day programming.
 \item Like to see more examples from the real world.  
@@
 Also, the icebreakers were not very useful.
-\item A single room for the duration of the meeting. (Particular to this WSSSPE4, the "collaboratory" room with the monitors, was actually perfect for the full meeting.)
+\item A single room for the duration of the meeting. (Particular to this WSSSPE4, the ``collaboratory'' room with the monitors, was actually perfect for the full meeting.)
 \item Allotting more time for questions \& discussions in between talks
@@
 \begin{itemize}
-\item "Citations / CodeMeta"  and  "Software Practices White Paper"
+\item ``Citations / CodeMeta''  and  ``Software Practices White Paper''
 \item Best Practices White Paper and Best Practices Undergraduate Course
@@
 The working groups activities were excellent, a good mix of projects, and I saw a lot of progress on several I'm very interested in.
-(Please group this survey with my earlier one; I hit "end" too soon on the other!
+(Please group this survey with my earlier one; I hit ``end'' too soon on the other!
 Thanks!)

### 2016-11-07 WSSSPE4_report/wg_codemeta.tex

@@
 
-Because of the work done at WSSSPE4, the CodeMeta project README file\citep{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} was greatly expanded to include a description of the project that is geared to those with little or no prior knowledge of the project and including a list of contributors, information on how one can get involved, a bit of the project history and who is managing the project, and links to additional information.
+Because of the work done at WSSSPE4, the CodeMeta project README file\footnote{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} was greatly expanded to include a description of the project that is geared to those with little or no prior knowledge of the project and including a list of contributors, information on how one can get involved, a bit of the project history and who is managing the project, and links to additional information.

### 2016-11-08 WSSSPE3/report/WSSSPE3_review.tex
old change...

@@
 Scientific Software?}. Turk is a prolific scientific software practitioner and
-has extensive experiences working on large collaborative projects employing
+has extensive experience working on large collaborative projects employing
 modern computing tools \cite{2011ApJS..192....9T}. He also co-organizes and
@@
 Thaney: \textit{Collaborative Software Development as Sustainable Software:
-Lessons from Open Source}}. Abigail Cabunoc Mayes combines two properties of open
+Lessons from Open Source}}. Abigail Cabunoc Mayes combined two properties of open
 source that, together, create a suitable habitat for sustainable software. The
@@
 is possible to fit the resulting metrics in a common template. So far, this is
-still in a design phase, with a white paper at the CSESSP workshop 2015 and this
+still in a design phase, with a white paper at the 2015 CSESSP workshop and this
 talk, but the WSSSPE workshops are seen as a forum for the community to assemble
@@
 \newpage
+
+\bibliographystyle{vancouver}
+
+\bibliography{wssspe}
+
+\newpage
 \appendix

### 2016-11-08 WSSSPE4_report/survey_details.tex
moving survey figures, fixing figure latex
adding some captions to figures and todos for someone with latex skills and time

@@
 
+%Figures have been inserted with little/no formatting.  
+%Figures are in *.png but can be remade to other formats 
+\todo{should remove these 3 figure borders in the figure source, then in LaTeX place them all in row of fix to use multifig}
+\begin{figure}[h!]
+\includegraphics[width=0.3\textwidth]{figures/SurveyFig1}
+\caption{The plenary sessions had a good mix of topics and speakers.
+\label{fig:SFig1}}
+\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=0.3\textwidth]{figures/SurveyFig2}
+\caption{The balance was right between talks, panels, and lightning talks.
+\label{fig:SFig2}}
+\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=0.3\textwidth]{figures/SurveyFig3}
+\caption{There was sufficient time for discussion.
+\label{fig:SFig3}}
+\end{figure}
+
 \noindent {\bf 2. What part of the meeting was the most valuable? select all that apply}
+
+\todo{move table and figure to be in a single row}
 
@@
 \end{table}
+
+\todo{remove figure border}
+\begin{figure}[h!]
+\includegraphics{figures/SurveyFig4}
+\caption{What part of the meeting was the most valuable?
+\label{fig:SFig4}}
+\end{figure}
 
@@
 
+\todo{move table and figure to be in a single row}
+
 \begin{table}[h!]
@@
 
+\todo{remove figure border}
+\begin{figure}[h]
+\includegraphics{figures/SurveyFig5}
+\caption{Would you be interested in joining a professional society focused around the topics of WSSSPE?
+\label{fig:SFig5}}
+\end{figure}
+
 \noindent {\bf 7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
+
+\todo{move table and figure to be in a single row}
 
@@
 \end{table}
+
+\todo{remove figure border}
+\begin{figure}[h]
+\includegraphics{figures/SurveyFig6}
+\caption{Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE?
+\label{fig:SFig6}}
+\end{figure}
 

### 2016-11-09 WSSSPE4_report/WSSSPE4_report.tex
adding social science section for Stuart to write

@@
 \item White paper on developing sustainable software [Sandra Gesing]
-\item \todo{to confirm} Social Science of Scientific Software [Stuart Geiger]
+\item Social Science of Scientific Software [Stuart Geiger]
 \item Software Best Practices for Undergraduates [Jonah Miller]
@@
 \item Open research index [Dan]
-\item \todo{to confirm} Letters of evaluation for computational scientists [Wolfgang Bangerth]
+\item Letters of evaluation for computational scientists [Wolfgang Bangerth]
 \end{itemize}
@@
 \input{wg_best_practices_developing}
+\input{wg_social}
 \input{wg_best_practices_undergrads}
@@
 \input{wg_open_research_index}
+\input{wg_letters}
 

### 2016-11-16 WSSSPE4_report/insert_participants.py
generating emails

@@ -4,10 +4,11 @@ import re, operator
 # set to 'name' for hyperlinking email address with name,
 # 'column' for a column, and
 # 'off' for no email.
-email = 'name'
+email = 'email'
 
 with open('participants.txt', 'r') as f:
   people = []
+  addresses = []
   person = {}
   for line in f:
     if 'institution: ' in line:
@@ -31,8 +32,13 @@ for person in sorted(people, key=lambda x: x['name'].split()[-1]):
            )
     elif email == 'off':
         print(person['name'] + ' & ' + person['institution'] + r'\\')
+    elif email == 'email':
+        addresses.append(person['email'] + r', ')
     elif email == 'name':
         print(r'\href{mailto:' + person['email'] + r'}{' +
               person['name'] + r'} & ' +
               person['institution'] + r'\\'
              )
+
+if email == 'email':
+    print(''.join(map(str, addresses)))

### 2016-11-16 WSSSPE4_report/insert_participants.py
fixed insert_participants for email list

@@ -6,22 +6,21 @@ import re, operator
 # 'off' for no email.
 email = 'email'
 
+people = []
 with open('participants.txt', 'r') as f:
-  people = []
-  addresses = []
-  person = {}
-  for line in f:
-    if 'institution: ' in line:
-      person['institution'] = re.search('institution: (.*)', line).group(1).replace('&', r'\&')
-    elif 'email: ' in line:
-      person['email'] = re.search('email: (.*)', line).group(1)
-    elif len(line) > 2:
-      person['name'] = line[:-1].replace('&', r'\&')
-    else:
-      people.append(person)
-      person = {}
-if 'name' in person:
-  people.append(person)
+    person = {}
+    for line in f:
+        if 'institution: ' in line:
+            person['institution'] = re.search('institution: (.*)', line).group(1).replace('&', r'\&')
+        elif 'email: ' in line:
+            person['email'] = re.search('email: (.*)', line).group(1)
+        elif len(line) > 2:
+            person['name'] = line[:-1].replace('&', r'\&')
+        else:
+            people.append(person)
+            person = {}
+    if 'name' in person:
+        people.append(person)
 
 for person in sorted(people, key=lambda x: x['name'].split()[-1]):
     if email == 'column':
@@ -33,12 +32,9 @@ for person in sorted(people, key=lambda x: x['name'].split()[-1]):
     elif email == 'off':
         print(person['name'] + ' & ' + person['institution'] + r'\\')
     elif email == 'email':
-        addresses.append(person['email'] + r', ')
+        print(person['email'] + ',')
     elif email == 'name':
         print(r'\href{mailto:' + person['email'] + r'}{' +
               person['name'] + r'} & ' +
               person['institution'] + r'\\'
              )
-
-if email == 'email':
-    print(''.join(map(str, addresses)))

### 2016-11-16 WSSSPE4_report/insert_participants.py
documented new feature of insert_participants

@@ -3,7 +3,8 @@ import re, operator
 
 # set to 'name' for hyperlinking email address with name,
 # 'column' for a column, and
-# 'off' for no email.
+# 'off' for no email, and
+# 'email' to just print a list of emails
 email = 'email'
 
 people = []

### 2016-11-16 WSSSPE4_report/WSSSPE4_report.tex
minor comments
adding section for Alliance working group
adding access subsection

@@
 
-\todo{description of panel discussion on best practices} \note{asked Simon to write}
+\todo{description of panel discussion on best practices} \note{Simon will write this}
 
@@
 
-\note{list of sections that is in the process of being confirmed}
-
-Sections:
-
-\begin{itemize}
-\item Verifying best practices \& metrics for sustainable research software [Ray Idaszak]
-\item \todo{to confirm} Software Sustainability Alliance [Neil \& Jean]
-\item Scientific Software Prototyping Infrastructure (S2PI) [Santiago Nunez-Corrales] 
-\item CodeMeta [Alice Allen]
-\item White paper on developing sustainable software [Sandra Gesing]
-\item Social Science of Scientific Software [Stuart Geiger]
-\item Software Best Practices for Undergraduates [Jonah Miller]
-\item Meaningful metrics for sustainable software [Emily Chen]
-\item \todo{to confirm} Coordinating access to CI for research software [Mark Abraham]
-\item Software engineering processes tailored for research software [Anshu Dubey]
-\item Open research index [Dan]
-\item Letters of evaluation for computational scientists [Wolfgang Bangerth]
-\end{itemize}
+\todo{need to confirm who will write Access subsection (\S\ref{sec:access})}
 
 \input{wg_best_practices_sustainability}
+\input{wg_alliance}
 \input{wg_prototyping}
@@
 \input{wg_metrics}
+\input{wg_access}
 \input{wg_soft_eng}

### 2016-11-21 WSSSPE4_report/wg_open_research_index.tex
starting to work on open research index working group report section

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
+\subsection{Open research index}
 \label{sec:open-research-index}
@@
 
-Introduction to group here, including the overall objective of work in this area.
+The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
+Our goal is not to eliminate commercial products, but to build on what?s there and provide data and services that are missing.
+
+The Open Research Index should take in all research products (papers, software, datasets, workflows, etc.) from their publishers and recorders (journals, societies, domain repositories, government [open access] repositories, preprint servers, general repositories [e.g., figshare, zenodo]) and other services (CrossRef, ORCID).
+Each product should list authors and citations and allow people to search the resulting network.
+Users should also be able to interact with their own record and edit it, like Google Scholar allows.
 
@@
 
-members of the working group
+The members of this group at WSSSPE4 were:
+Gabrielle Allen,
+Bruce Childers,
+Robert Haines,
+Caroline Jay,
+Dan Katz,
+Daniel Mosse,
+Kyle Niemeyer,
+and
+Robert McDonald.
 
@@
 
-Specific things the working group wants to accomplish in the context of the larger objective.
+The working group's plans are relatively simple to express, though quite complex to undertake:
+
+\begin{enumerate}
+\item Determine a plan to build an open research index that allows various stakeholders to satisfy their needs
+\item Then determine if the plan is feasible
+\item If so, then obtain resources
+\item If successful, then build the index
+\end{enumerate}
 
@@
 
-What is the gap or challenge being addressed?
+Google and others provide some services now.
+But these services (and the underlying data) could be removed at any time.
+And the community cannot build new services.
 
@@
 
-What people, groups, or resources are needed.
+Names of people willing to contribute (or expertise needed), and how they?ll contribute
+- Could bring in someone from ORCID, CrossRef, DataCite, ImpactStory/depsy, altmetrics.org, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, softwareheritage.org, DBLP, eSTEP (CWO)
+- Work with some initial publishers who don?t have competing services (domain societies), PubMed, 
+- Need to discuss with potential funders
+   - What data/services would they like to have
+      - Are they willing to invest in this?
+      - E.g. current Arnold Foundation Open Science call
+- Who would actually lead and coordinate this activity?
+
+What external resources are needed, and how they can be brought in?
+- Money
+- Time
 
@@
 
-What tasks will the working group undertake
+Currently unclear if any of us have the time/energy to pursue this
+
+If we do, plan could be:
+- Obtain funding for conversations
+- Conversations with potential partners (publishers, orgs, funders) (could be combined with meeting below)
+- Conversation with Google Scholar developer for better understanding of what they did and are doing
+- Funding for a meeting (perhaps from NSF) (Dagstuhl meeting?)
+- Meeting to define a plan
+- Funding to implement the plan
+
+Could also build a mailing list for us and discuss further, depending on how receptive others are to this idea
+
+Possible Plans
+
+Look at curriculum lattes (Brazil) and/or Researchfish (UK).  It has some problems that people don?t like, could check what features it has and what the issues are
+
+Look at what CrossRef data is
+
+What research products should we track?
+
+What model should be used?
+- Crawling and obtaining data
+- Consortium model
+- Need analysis and decision
+
+Need agreements with publishers, repositories, services
+- Some of these will have costs, though some may be free
+- Need to determine costs
+- Perhaps publishers would pay us to be listed eventually, as we gain power
+- To some extent, we would be competing with some of the publishers? products
+
+If crawling model:
+- How to crawl all of these products?
+- What data (metadata) should be the output of the crawls?
+    - What happens when the products being crawled and/or their metadata change?
+- How to store all of this data (metadata)?
+
+What services to provide on this data (metadata)?
+
+Who will create these services?
+
+What services will the publishers allow to be provided?
+
+Need to develop a plan that shows incremental progress and successes
+
+Consider applying to Google project call to get Google interest and support?
 
@@
 
+Community mailing list/forum (Kyle, today)
+done - wg-open-research-idx on wssspe slack channel
+
+Find a PI (we suggest Neil Chue Hong)
+
+Apply for initial funding
+- For the US, consider Arnold Foundation Open Science award as a planning activity (who, due Dec 15)
+- For the UK,
+
+Determine project name, initial vision, initial mission, logo (who, when)
+
+Identify initial use cases (who, when)
+
+Identify potential stakeholders and partners (who, when)
+
+Discuss this idea with potential partners (who, when)
+- Get their feedback \& buy-in
+- While doing so, seek suggestions for the right project manager to coordinate this
+
+Build up an advisory board (who, when)
+
+Build up a leadership committee and determine a leader (who, when)
+
+Hire a project manager (who, when)
+
+Iterate use case with stakeholders (who, when)
+- E.g. funders, Researchfish, researchers, ?
+
+Iterate vision \& mission with leadership committee and advisory board (who, when)
+
+Develop initial development plan (who, when)
+
 \subsubsection{More information \& joining instructions}
@@
 How could a reader get more information or get more involved?
+
+add into about joining WSSSPE slack and then this channel.

### 2016-11-23 WSSSPE4_report/WSSSPE4_report.tex
working on open research index section of report

@@
 
+In each group's discussion, a number of Specific, Measurable, Achievable, Relevant and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.
+In many cases, this was done through Slack (see \S\ref{sec:slack}.)
+
 \todo{need to confirm who will write Access subsection (\S\ref{sec:access})}
@@
 \input{wg_letters}
+
+\section{Slack team and channels}\label{sec:slack}
+
+As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE.
+To join it, go to \url{https://wssspe.signup.team/}.
+A number of the working groups also created channels within the team; see the subsections in \S\ref{sec:WGs} for the names of these channels.
 

### 2016-11-23 WSSSPE4_report/wg_open_research_index.tex
working on open research index section of report

@@
 
-Names of people willing to contribute (or expertise needed), and how they?ll contribute
-- Could bring in someone from ORCID, CrossRef, DataCite, ImpactStory/depsy, altmetrics.org, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, softwareheritage.org, DBLP, eSTEP (CWO)
-- Work with some initial publishers who don?t have competing services (domain societies), PubMed, 
-- Need to discuss with potential funders
-   - What data/services would they like to have
-      - Are they willing to invest in this?
-      - E.g. current Arnold Foundation Open Science call
-- Who would actually lead and coordinate this activity?
+People/organizations who might be willing to contribute (or whose expertise is needed), and how they will contribute:
 
-What external resources are needed, and how they can be brought in?
-- Money
-- Time
+\begin{itemize}
+\item ORCID, CrossRef, DataCite, ImpactStory/depsy, altmetrics.org, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, softwareheritage.org, DBLP, eSTEP (CWO)
+\item We could work with some initial publishers who don't have competing services (domain societies), PubMed
+\item We need to discuss this idea with potential funders, and ask what data/services would they like to have?  And are they willing to invest in this?
+\end{itemize}
+
+The external resources that are needed are money and time; it is unclear how they can be brought in.
 
@@
 
-Currently unclear if any of us have the time/energy to pursue this
+At the time of the meeting, and today as well, it is unclear who has the time and energy to pursue this idea.
 
-If we do, plan could be:
-- Obtain funding for conversations
-- Conversations with potential partners (publishers, orgs, funders) (could be combined with meeting below)
-- Conversation with Google Scholar developer for better understanding of what they did and are doing
-- Funding for a meeting (perhaps from NSF) (Dagstuhl meeting?)
-- Meeting to define a plan
-- Funding to implement the plan
+If we identify a leader (PI), a plan could be:
+\begin{enumerate}
+\item Obtain funding for a set of initial discussions (or find a group who feel that this is important enough that they will take the time to do it anyhow)
+\item Talk with potential partners (publishers, orgs, funders) (could be combined with meeting below)
+\item Talk with the Google Scholar developer for better understanding of what they did and are planning in the future
+\item Obtain funding for a meeting (perhaps from NSF, perhaps a Dagstuhl meeting)
+\item Hold a community meeting to define a plan (which should be large enough to include representative of all stakeholder groups, and small enough that we still have a good open discussion)
+\item Obtain funding to implement the plan, including contributions and commitments from the stakeholders
+\end{enumerate}
 
-Could also build a mailing list for us and discuss further, depending on how receptive others are to this idea
+Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
+At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \#wg-open-research-idx.
+(See \S\ref{sec:slack} to join the Slack WSSSPE team.)
 
-Possible Plans
+\paragraph{Possible Plans}
 
@@
 
-What are the first SMART steps proposed?
+The working group discussed a number of SMART steps, though most of them did not fulfill all the SMART criteria:
 
-Community mailing list/forum (Kyle, today)
-done - wg-open-research-idx on wssspe slack channel
+\begin{enumerate}
+\item Develop a community mailing list/forum (Kyle, during the workshop).  This was completed; see \S\ref{sec:wg-open-research-index-list}.
 
@@
 - For the US, consider Arnold Foundation Open Science award as a planning activity (who, due Dec 15)
-- For the UK,
+- For the UK, ?
 
@@
 
-\subsubsection{More information \& joining instructions}
+\end{enumerate}
 
-How could a reader get more information or get more involved?
+\subsubsection{More information \& joining instructions}\label{sec:wg-open-research-index-list}
 
-add into about joining WSSSPE slack and then this channel.
+If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \#wg-open-research-idx.  (See \S\ref{sec:slack} to join the Slack WSSSPE team.)

### 2016-11-23 WSSSPE4_report/wg_open_research_index.tex
finishing first draft of open research index working group section

@@
 
-\note{Dan to write this}
+%\note{Dan to write this}
 
@@
 Dan Katz,
+Robert McDonald,
 Daniel Mosse,
-Kyle Niemeyer,
 and 
-Robert McDonald.
+Kyle Niemeyer.
 
@@
 At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \#wg-open-research-idx.
-(See \S\ref{sec:slack} to join the Slack WSSSPE team.)
+(See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
 
-\paragraph{Possible Plans}
+We also discussed some possible more detailed plans:
 
-Look at curriculum lattes (Brazil) and/or Researchfish (UK).  It has some problems that people don?t like, could check what features it has and what the issues are
+\begin{enumerate}
+\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people don't like, we could check what features they has and what the issues are.
 
-Look at what CrossRef data is
+\item Look at what CrossRef data is
 
-What research products should we track?
+\item What research products should we track?
 
-What model should be used?
+\item What model should be used?
 - Crawling and obtaining data
@@
 
-Need agreements with publishers, repositories, services
+\item Need agreements with publishers, repositories, services
 - Some of these will have costs, though some may be free
@@
 
-If crawling model:
+\item If crawling model:
 - How to crawl all of these products?
@@
 
-What services to provide on this data (metadata)?
+\item What services to provide on this data (metadata)?
 
-Who will create these services?
+\item Who will create these services?
 
-What services will the publishers allow to be provided?
+\item What services will the publishers allow to be provided?
 
-Need to develop a plan that shows incremental progress and successes
+\item Need to develop a plan that shows incremental progress and successes
 
-Consider applying to Google project call to get Google interest and support?
+\item Consider applying to Google project call to get Google interest and support?
+\end{enumerate}
 
@@
 
-The working group discussed a number of SMART steps, though most of them did not fulfill all the SMART criteria:
+The working group discussed a number of SMART-like steps, though most of them did not fulfill all the SMART criteria, such as identifying who would do them and when:
 
@@
 
-Find a PI (we suggest Neil Chue Hong)
+\item Find a PI (we suggest Neil Chue Hong)
 
-Apply for initial funding
-- For the US, consider Arnold Foundation Open Science award as a planning activity (who, due Dec 15)
-- For the UK, ?
+\item Apply for initial funding. For the US, we discussed the Arnold Foundation Open Science award as a planning activity (who, due Dec 15), though without a PI, this seems unlikely to be done.
+For the UK, we didn't have a suggestion.
 
-Determine project name, initial vision, initial mission, logo (who, when)
+\item Determine project name, initial vision, initial mission, logo (who, when)
 
-Identify initial use cases (who, when)
+\item Identify initial use cases (who, when)
 
-Identify potential stakeholders and partners (who, when)
+\item Identify potential stakeholders and partners (who, when)
 
-Discuss this idea with potential partners (who, when)
-- Get their feedback \& buy-in
-- While doing so, seek suggestions for the right project manager to coordinate this
+\item Discuss this idea with potential partners (who, when).
+Then hgt their feedback \& buy-in.
+While doing so, seek suggestions for the right project manager to coordinate this
 
-Build up an advisory board (who, when)
+\item Build up an advisory board (who, when)
 
-Build up a leadership committee and determine a leader (who, when)
+\item Build up a leadership committee and determine a leader (who, when)
 
-Hire a project manager (who, when)
+\item Hire a project manager (who, when)
 
-Iterate use case with stakeholders (who, when)
-- E.g. funders, Researchfish, researchers, ?
+\item Iterate use case with stakeholders (who, when),
+e.g., funders, Researchfish, researchers, ?
 
-Iterate vision \& mission with leadership committee and advisory board (who, when)
+\item Iterate vision \& mission with leadership committee and advisory board (who, when)
 
-Develop initial development plan (who, when)
+\item Develop initial development plan (who, when)
 
@@
 
-If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \#wg-open-research-idx.  (See \S\ref{sec:slack} to join the Slack WSSSPE team.)
+If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \#wg-open-research-idx.
+(See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

### 2016-12-12 WSSSPE4_report/wg_best_practices_undergrads.tex
adding Software Best Practices for Undergraduates section; fixing other best practices section
added course for undergraduates

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Software best practices for undergraduates}
+\label{sec:best-practices-undergrads}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Jonah Miller to write this}
+
+% Introduction to group here, including the overall objective of work
+% in this area.
+
+This working group was motivated by the perceived prevalence of
+so-called ``dark code'' in scientific communities: code written by
+individual researchers in an unsustainable way that is never shared
+with the larger community. Participants recalled their own experience
+working with colleagues who write dark code or inheriting
+unsustainable dark code from collaborators.
+
+The question is, then, how to prevent researchers from writing dark
+code? The participants hypothesized that the best strategy is to catch
+researchers while they are still in training and teach them software
+best-practices. Therefore, this working group was formed with the goal
+of developing courses on software best practices aimed at
+undergraduate students studying domain science. The program might be
+similar to a software carpentry or data carpentry workshop but with a
+focus for domain scientists.
+
+\subsubsection{Participants}
+
+The participants in this working group are/were Jonah Miller,
+Aleksandra Nenadic, Raniere Silva, Francisco Queiroz, Hans Fangohr,
+and Prabhjyot Sing. Aleksandra Nenadic and Raniere Silva are involved
+in software and data carpentry.
+
+\subsubsection{Working group objective}
+
+Implementation of a course aimed at each domain science is a long term
+goal. However, a short-term, achievable goal, is to develop a
+curriculum for a course aimed at a single domain science. Since the
+participants in the working group have expertise in physics, this is a
+natural target.
+
+\subsubsection{Gap or challenge}
+
+The challenge is that people are writing unsustainable scientific
+software without ever learning that there is a better way. Here we
+focus on training in good software design and engineering in this
+context.
+
+\subsubsection{Relevant people and resources}
+
+The development of a successful curriculum relies on the expertise of
+software engineers to describe the best practices, domain experts to
+describe model problems and work-flow, and instructors to formulate
+the pedagogy. Ideally these people will be brought together for a
+short workshop or hackathon with the goal of drafting the
+curriculum. Later, organizational partners will be required to
+actually implement the program.
+
+\subsubsection{Plans}
+
+The working group will seek funding to organize a hackathon to develop
+the curriculum and seek the required expertise to invite. The
+hackathon will be implemented and a draft of the curriculum will be
+written within the next year.
+
+\subsubsection{SMART steps}
+
+The working group will accomplish or has accomplished the following
+smart steps:
+
+\begin{enumerate}
+\item Create channels of communication including a mailing list, a
+  slack channel, and a git repository
+\item Review literature on best and good-enough practices for
+  scientific computing as well as literature on pedagogy for
+  scientific and research computing
+\item Organize a telecon to organize a workshop to write the
+  curriculum and decide how to bring in outside experts
+\item Write a draft of the curriculum
+\item Invite the broader community to contribute to a more complete
+  curriculum
+\end{enumerate}
+
+\subsubsection{More information \& joining instructions}
+
+Anyone interested in contributing should get in touch via one of the
+following methods:
+
+\begin{itemize}
+\item Join our google group/mailing list
+  \cite{WSSSPEUndergradGoogleGroup}.
+\item Join the WSSSPE Slack \cite{WSSSPESlack} and sign into the
+  \texttt{\#wg-undergraduatecourse} channel and say hi.
+\item Ask to join the WSSSPE4-undergraduate-course organization
+  \cite{WSSSPEUndergradGithub}.
+\end{itemize}

### 2016-12-13 WSSSPE4_report/wg_best_practices_developing.tex
adding three new report sections
Update wg_best_practices_developing.tex
Add information to subsections Relevant people and resources, SMART steps, and More information & joining instructions

--- 2016-10-05 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{White paper on developing sustainable softwares}
+\label{sec:best-practices-developing}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Dan to write this}
+\note{Sandra to write this}
 
@@
 
-What people, groups, or resources are needed.
+The list of contributors, as of November 10th 2016, includes (in alphabetical order): \\
+Jeffrey C. Carver <carver@cs.ua.edu>, 
+Neil Chue Hong <N.ChueHong@software.ac.uk>, 
+Tom Crick <tcrick@cardiffmet.ac.uk>, 
+Miguel de~Val-Borro <valborro@princeton.edu>, 
+Hans Fangohr <fangohr@soton.ac.uk>
+Sandra Gesing <sandra.gesing@nd.edu>, 
+Derek Groen <Derek.Groen@brunel.ac.uk>, 
+Dan Gunter <DKGunter@lbl.gov>,
+Daniel S. Katz <d.katz@ieee.org>,
+Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
+Frank Löffler <knarf@cct.lsu.edu>,
+Suresh Marru <smarru@iu.edu>,
+Kyle E. Niemeyer <kyle.niemeyer@oregonstate.edu>,
+Abani Patra <abani@buffalo.edu> 
+and 
+Francisco Queiroz <chico@puc-rio.br>
 
@@
 
-What are the first SMART steps proposed?
+First eight SMART steps proposed: \\
+\begin{itemize}
+\item List existing contributors and open GitHub for new ones - Sandra Gesing (done). 
+\item Establish preliminary timeline  - Sandra Gesing (done). 
+\item Organize related work - Francisco Queiroz (done). 
+\item Define scope of the paper - Sandra Gesing (done). 
+\item Suggest new sections - Francisco Queiroz (done). 
+\item Define leading authos and contributors for sections - Francisco Queiroz (done). 
+\item Finalize first version - Sandra Gesing (in progress - due 31st of January, 2016) 
+\item Distribute to WSSSPE community and collect feedback - Sandra Gesing (in progress - due 28th of February, 2016) 
+\end{itemize}
 
@@
 
-How could a reader get more information or get more involved?
+The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more infomation and requests, join \#wg-best-practices channel at WSSSPE's Slack team.

### 2016-12-16 WSSSPE4_report/wg_best_practices_undergrads.tex

@@
 \item Create channels of communication including a mailing list, a
-  slack channel, and a git repository
+  slack channel, and a git repository (done)
 \item Review literature on best and good-enough practices for
   scientific computing as well as literature on pedagogy for
-  scientific and research computing
+  scientific and research computing (by February 1, 2017)
 \item Organize a telecon to organize a workshop to write the
-  curriculum and decide how to bring in outside experts
-\item Write a draft of the curriculum
+  curriculum and decide how to bring in outside experts (by February 1,
+  2017)
+\item Write a draft of the curriculum (by April 1, 2017)
 \item Invite the broader community to contribute to a more complete
-  curriculum
+  curriculum (by April 30, 2017)
 \end{enumerate}

### 2016-12-18 WSSSPE4_report/wg_soft_eng.tex
starting to fill in the content for the SE specific to science codes

--- 2016-10-05 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Software engineering processes tailored for research software}
+\label{sec:soft-eng}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Dan to write this}
+\note{Anshu to write this}
 
-Introduction to group here, including the overall objective of work in this area.
+%Introduction to group here, including the overall objective of work in this area.
 
 \subsubsection{Participants}
-
-members of the working group
-
+\begin{itemize}
+\item Mark Abraham
+\item Anshu Dubey - is a computer scientist at the Mathematics and
+  Computer Science Division at Argonne National Laboratory. She has
+  been deeply involved with the development of multiphysics scientific
+  software in multiple domains which has lead to an interest in, and
+  understanding of, software engineering concerns in scientific
+  computing. 
+\item Hans Fangohr
+\item Dominic Kempf
+\item Eric Seidel
+\end{itemize}
 \subsubsection{Working group objective}
-
-Specific things the working group wants to accomplish in the context of the larger objective.
+Scientific computing software lags behind commercial software in
+adoption of software engineering practices. Causes range from lack of
+exposure to the practices to distrust of adopting practices because
+they do not meet the needs of the teams developing such software. This
+gap is particularly acute in the area of software testing,
+verification and validation where the standard practices are
+simulataneously inadequate and over-onerous. The objective of this
+working group is to (1) conduct literature survey to guage the extent
+of awareness of issue in general, (2) generate content useful for
+the community where needed, and (3) curate the collected and added
+content for the use of the community.
 
@@
 
-What is the gap or challenge being addressed?
+%What is the gap or challenge being addressed?
 

### 2016-12-28 WSSSPE4_report/wg_soft_eng.tex
added the section of report on software engineering tailored for computational science

@@
 %Introduction to group here, including the overall objective of work in this area.
+This working group concerns itself with identify processes that are not well
+covered by general software engineering adequately. The current focus
+is on testing because there is growing awareness about the lack of trust
+and reproducibility in many computational results. A rigorous
+verification process which also involves ongoing regular testing is
+fundamental to addressing the issue of trust. Regular automated
+testing is the first step towards software process that can lead to
+provenance and reproducibility, the hallmarks of quality science.
+
+Computational science and engineering applications have many moving
+parts that need to interoperate with one another. The accuracy and
+reliability of results produced by the scientific software depends not
+only on the individual components behaving correctly, but also on the
+validity of their interactions. Therefore, a rigorous verification
+process and a robust testing regime are critical requirements for
+scientific software.  There have been instances where inadequate
+verification has resulted in publication of wrong results and
+retraction of publications later. As scientific understanding grows,
+models in the corresponding computational software are refined leading
+to more complex codes. Increasing complexity makes them more prone to
+defects not only in individual code units, but also in interaction
+among units. Therefore, a strong verification process combined with a
+rigorous testing regime play a critical role in the prevention of
+wrong scientific results being obtained. However, most science teams
+struggle to find a good solution for themselves. Causes range from
+lack of exposure to the practices to distrust of adopting practices
+because they do not meet the needs of the teams developing such
+software.
 
@@
 \end{itemize}
+
 \subsubsection{Working group objective}
 Scientific computing software lags behind commercial software in
-adoption of software engineering practices. Causes range from lack of
-exposure to the practices to distrust of adopting practices because
-they do not meet the needs of the teams developing such software. This
+adoption of software engineering practices. This
 gap is particularly acute in the area of software testing,
@@
 \subsubsection{Gap or challenge}
-
 %What is the gap or challenge being addressed?
+Computational science code developers often lack of exposure to
+regular testing and its benefits. Good developers will test the code to
+verify that it operates as expected, however, they may not appreciate
+that without regular testing defects can be introduced
+inadvertently. An even bigger challenge is that those who understand
+the importance of regular testing do not often find much help from
+software engineering literature. There is significant gap between
+testing gospel and its applicability to computational science. This
+gap leads to frustration and abandonment of good with the bad. Some
+relevant literature exists, in particular experiences from
+practitioners in computational science who developed their own
+solutions. However, this literature is scattered among many different
+forums, and can be challenging to find. Our working group aims to
+address this gap by curating the existing content and contirbuting
+content where none exists.
 
@@
 
-What people, groups, or resources are needed.
+%What people, groups, or resources are needed.
+The working group will benefit from a wide participation by developers
+of large computational science codes. The reason is that the
+management of such codes becomes intractable without adopting some
+software process and testing regime. The experiences and
+customizations vary, and the community will benefit from hearing about
+as many as possible. The seed resources required are fairly
+minimal. We have started a git repository for collecting the
+existing references. That, and a few volunteers reading through the
+references is all we need in the beginning. As we gather more
+knowledge and pinpoint gaps, we may need more resources to reach out
+to a wider group of developers to collect more information.
 
-\subsubsection{Plans}
+%\subsubsection{Plans}
 
-What tasks will the working group undertake
+%What tasks will the working group undertake
 
@@
 
-What are the first SMART steps proposed?
+%What are the first SMART steps proposed?
+Our first few SMART steps are : 
+\begin{itemize}
+\item Create a channel in wssspe.slack.com {\em done}
+\item Create github repository
+  {\url{http://github.com/wssspe/testing-in-science}} {\em done}
+\item Find and gather existing publications in repository {\em ongoing}
+\item Review and summarise material. Decide whether we consider this
+  sufficient. If yes, then we will put brief report together and
+  conclude the working group.
+\item If we do not consider the material adequate, we start research
+  and gather methodologies for testing in science 
+\item Write a document accessible to computational science and
+  software engineering community, and publish document at a citeable forum.
+\end{itemize}
 
@@
 
-How could a reader get more information or get more involved?
+%How could a reader get more information or get more involved?
+Readers interested in getting more information should get in touch
+with a member of the working group. The working group
+has a slack channel under WSSSPE organization. The channel is called
+{\it wg-testing-in-science}. Readers interested in getting more
+involved can ask for inclusion on the channel. Additionally, a git
+repository exists for contributing content and reference to, and
+curation of the existing literature on this topic.

### 2016-12-30 WSSSPE4_report/wg_soft_eng.tex
edits to new text, added 2 comments

--- 2016-10-05 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Software engineering processes tailored for research software}
+\label{sec:soft-eng}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Dan to write this}
+\note{Anshu to write this}
 
-Introduction to group here, including the overall objective of work in this area.
+%Introduction to group here, including the overall objective of work in this area.
+This working group concerned itself with identifying processes that are not adequately
+covered by general software engineering. The group focused
+on testing because there is a growing awareness about the lack of trust
+and reproducibility in many computational results. A rigorous
+verification process that also involves ongoing regular testing is
+fundamental to addressing the issue of trust. Regular automated
+testing is the first step towards software processes that can lead to
+provenance and reproducibility, the hallmarks of quality science.
+
+\katznote{this paragraph repeats some of the previous one - they probably should be merged}
+Computational science and engineering applications have many moving
+parts that need to interoperate with one another. The accuracy and
+reliability of results produced by the scientific software depends not
+only on the individual components behaving correctly, but also on the
+validity of their interactions. Therefore, a rigorous verification
+process and a robust testing regime are critical requirements for
+scientific software.  There have been instances where inadequate
+verification has resulted in publication of wrong results and later
+retraction of publications. As scientific understanding grows,
+the corresponding computational software models are refined, leading
+to more complex codes. Increasing complexity makes them more prone to
+defects, not only in individual code units, but also in interaction
+among units. Therefore, a strong verification process combined with a
+rigorous testing regime plays a critical role in the prevention of
+generating incorrect scientific results. However, most science teams
+struggle to find a good solution for themselves. Causes range from
+lack of exposure to the practices, to distrust of adopting practices
+because they do not meet the needs of the teams developing such
+software.
 
 \subsubsection{Participants}
-
-members of the working group
+\begin{itemize}
+\item Mark Abraham
+\item Anshu Dubey %- is a computer scientist at the Mathematics and
+%  Computer Science Division at Argonne National Laboratory. She has
+%  been deeply involved with the development of multiphysics scientific
+%  software in multiple domains which has lead to an interest in, and
+%  understanding of, software engineering concerns in scientific
+%  computing. 
+\item Hans Fangohr
+\item Dominic Kempf
+\item Eric Seidel
+\end{itemize}
 
 \subsubsection{Working group objective}
-
-Specific things the working group wants to accomplish in the context of the larger objective.
+Scientific computing software lags behind commercial software in
+adoption of software engineering practices. This
+gap is particularly acute in the area of software testing,
+verification, and validation, where the standard practices are
+simultaneously inadequate and overly onerous.
+This group aims to close the gap.
 
 \subsubsection{Gap or challenge}
-
-What is the gap or challenge being addressed?
+%What is the gap or challenge being addressed?
+Computational science code developers often lack of exposure to
+regular testing and its benefits. Good developers will test their code to
+verify that it operates as expected; however, they may not appreciate
+that without regular testing, defects can be introduced
+inadvertently. An even bigger challenge is that those who understand
+the importance of regular testing do not often find much help from
+software engineering literature. There is a significant gap between
+the testing gospel and its applicability to computational science. This
+gap leads to frustration and abandonment of the good with the bad. Some
+relevant literature exists, in particular experiences from
+practitioners in computational science who developed their own
+solutions. However, this literature is scattered among many different
+forums, and can be challenging to find. Our working group aims to
+address this gap by curating the existing content and contributing
+content where none exists.
 
@@
 
-What people, groups, or resources are needed.
+%What people, groups, or resources are needed.
+The working group will benefit from wide participation by developers
+of large computational science codes. The reason is that the
+management of such codes becomes intractable without adopting some
+software process and a testing regime. The experiences and
+customizations vary, and the community will benefit from hearing about
+as many as possible. The seed resources required are fairly
+minimal. We have started a git repository for collecting the
+existing references. That, and a few volunteers reading through the
+references is all we need in the beginning. As we gather more
+knowledge and pinpoint gaps, we may need more resources to reach out
+to a wider group of developers to collect more information.
 
@@
 
-What tasks will the working group undertake
+%What tasks will the working group undertake
+This working group will (1) conduct a literature survey to gauge the extent
+of awareness of the issue in general, (2) generate content useful for
+the community where needed, and (3) curate the collected and added
+content for the use of the community.
 
@@
 
-What are the first SMART steps proposed?
+%What are the first SMART steps proposed?
+Our first few SMART steps are : 
+\begin{itemize}
+\item Create a channel in wssspe.slack.com -- {\em done}
+\item Create github repository
+  {\url{http://github.com/wssspe/testing-in-science}} -- {\em done}
+\item Find and gather existing publications in repository -- {\em ongoing}
+\item Review and summarize material. Decide whether we consider this
+  sufficient. If yes, then we will put brief report together and
+  conclude the working group.
+\item If we do not consider the material adequate, we start research
+  and gather methodologies for testing in science 
+\item Write a document accessible to computational science and
+  software engineering community, and publish document at a citeable forum.
+\end{itemize}
 
@@
 
-How could a reader get more information or get more involved?
+%How could a reader get more information or get more involved?
+Readers interested in getting more information should get in touch
+with a member of the working group. The working group
+has a channel in the WSSSPE Slack. The channel is called
+{\it wg-testing-in-science}.
+(See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
+Additionally, a git
+repository \katznote{add url} exists for contributing content and reference to, and
+curation of the existing literature on this topic.

### 2017-01-03 WSSSPE4_report/wg_soft_eng.tex
addressed Dan's comments

@@
 
-\note{Anshu to write this}
+%\note{Anshu to write this}
 
@@
 This working group concerned itself with identifying processes that are not adequately
-covered by general software engineering. The group focused
-on testing because there is a growing awareness about the lack of trust
-and reproducibility in many computational results. A rigorous
-verification process that also involves ongoing regular testing is
-fundamental to addressing the issue of trust. Regular automated
-testing is the first step towards software processes that can lead to
-provenance and reproducibility, the hallmarks of quality science.
-
-\katznote{this paragraph repeats some of the previous one - they probably should be merged}
+covered by general software engineering. Verification and testing is
+the first subtopic that we address.  
+% there is a growing awareness about the lack of trust
+% and reproducibility in many computational results. 
+%\katznote{this paragraph repeats some of the previous one - they probably should be merged}
 Computational science and engineering applications have many moving
@@
 only on the individual components behaving correctly, but also on the
-validity of their interactions. Therefore, a rigorous verification
-process and a robust testing regime are critical requirements for
-scientific software.  There have been instances where inadequate
-verification has resulted in publication of wrong results and later
-retraction of publications. As scientific understanding grows,
+validity of their interactions. 
+% Therefore, a rigorous verification
+% process and a robust testing regime are critical requirements for
+% scientific software.  There have been instances where inadequate
+% verification has resulted in publication of wrong results and later
+% retraction of publications.
+As scientific understanding grows
 the corresponding computational software models are refined, leading
@@
 because they do not meet the needs of the teams developing such
-software.
+software.  The current focus of our effort is on testing because it is the first step
+towards building a software processes that can lead to provenance and
+reproducibility, the hallmarks of quality science.
 
@@
 \item Mark Abraham
-\item Anshu Dubey %- is a computer scientist at the Mathematics and
-%  Computer Science Division at Argonne National Laboratory. She has
-%  been deeply involved with the development of multiphysics scientific
-%  software in multiple domains which has lead to an interest in, and
-%  understanding of, software engineering concerns in scientific
-%  computing. 
+\item Anshu Dubey 
 \item Hans Fangohr
@@
 Additionally, a git
-repository \katznote{add url} exists for contributing content and reference to, and
+repository (\url{https://github.com/WSSSPE/WG-Best-Practices.git}) exists for contributing content and reference to, and
 curation of the existing literature on this topic.

### 2017-01-03 WSSSPE4_report/wg_soft_eng.tex
minor changes

@@
 parts that need to interoperate with one another. The accuracy and
-reliability of results produced by the scientific software depends not
+reliability of results produced by scientific software depends not
 only on the individual components behaving correctly, but also on the
@@
 % retraction of publications.
-As scientific understanding grows
+As scientific understanding grows,
 the corresponding computational software models are refined, leading

### 2017-01-04 WSSSPE4_report/wg_best_practices_developing.tex
spelling and LaTeX changes

@@
 Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
-Frank Löffler <knarf@cct.lsu.edu>,
+Frank L\"offler <knarf@cct.lsu.edu>,
 Suresh Marru <smarru@iu.edu>,
@@
 \item Suggest new sections - Francisco Queiroz (done). 
-\item Define leading authos and contributors for sections - Francisco Queiroz (done). 
+\item Define leading authors and contributors for sections - Francisco Queiroz (done). 
 \item Finalize first version - Sandra Gesing (in progress - due 31st of January, 2016) 
@@
 
-The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more infomation and requests, join \#wg-best-practices channel at WSSSPE's Slack team.
+The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more information and requests, join \#wg-best-practices channel at WSSSPE's Slack team.

### 2017-01-04 WSSSPE4_report/wg_alliance.tex
adding section for Alliance working group
alliance section from Jean and Neil

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Software Sustainability Alliance}
+\label{sec:alliance}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Neil and Jean wrote this}
+
+The Software Sustainability Alliance working group aims to establish an alliance between organizations interested in improving the sustainability of research software.
+Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
+We are currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
+
+\subsubsection{Participants}
+
+The participants are:
+Neil Chue Hong <N.ChueHong@software.ac.uk>,
+Jean Salac <jeansalac@virginia.edu>,
+Radovan Bast <radovan.bast@uit.no>,
+Lorraine Hwang <ljhwang@ucdavis.edu>,
+Karthik Ram <karthik.ram@berkeley.edu>,
+and
+Peter Elmer <Peter.Elmer@cern.ch>.
+
+\subsubsection{Working group objective}
+
+The overall objective of the Software Sustainability Alliance working group is to develop the steps necessary for establishing an alliance between organizations willing to engage in mutually reinforcing activities to advance the sustainability of software used in research.
+These organizations are funded groups or teams that aim to advance research software sustainability beyond their local university or community.
+More specifically, this working group aims to define the scope of this alliance and provide clear distinction with WSSSPE.
+We also seek to understand the incentives for members of this alliance and identify its key activities.
+
+\subsubsection{Gap or challenge}
+
+Currently, point-to-point collaboration exists between organizations, but this inadvertently results in competition or redundancy within the sustainable software community.
+An alliance of software sustainability organizations would ease inter-organization collaboration and the promotion of software sustainability.
+This alliance would also improve the pooling of competencies and the sharing of expertise.
+Furthermore, with the international scope of this alliance, it could support an organization who wants to hold an event in a country where another organization exists.
+
+\subsubsection{Relevant people and resources}
+
+Our working group is looking for ideas for which organizations should be consulted or invited to join the Software Sustainability Alliance.
+We are also looking for feedback on the aims and scope for joining the Software Sustainability Alliance.
+
+\subsubsection{Plans}
+
+Throughout the scoping period, our working group will continue to refine the Software Sustainability Alliance as we collect feedback on potential members, aims and scopes.
+We will also update our website to reflect these changes.
+Next steps and further tasks will be determined based on the feedback from the scoping period.
+
+\subsubsection{SMART steps}
+
+The first SMART steps of our working group are outlined below:
+\begin{itemize}
+\item Create a draft text to explain context and incentives for joining the Software Sustainability Alliance
+\item Come up with a list of potential invitees
+\item Update www.softwaresustainability.org to provide information
+\item Draft a letter to go out to potential alliance members
+\item Create a timeline and framework for initial consultation
+\end{itemize}
+
+\subsubsection{More information \& joining instructions}
+
+If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (N.ChueHong@software.ac.uk) and Jean Salac (jeansalac@virginia.edu)

### 2017-01-09 WSSSPE4_report/survey_details.tex
Remove borders on Survey Figures
All figures are now designated rev(used)
moving and shrinking figures
minor comments
playiung with figure spacing

--- 2016-11-07 WSSSPE4_report/survey_details.tex
@@
 
+%Figures have been inserted with little/no formatting.  
+%Figures are in *.png but can be remade to other formats 
+%\todo{Place 3 figures in 1 row with Latex}
+%\begin{figure}[h!]
+%\includegraphics[width=0.3\textwidth]{figures/SurveyFig1rev}
+%\caption{The plenary sessions had a good mix of topics and speakers.
+%\label{fig:SFig1}}
+%\end{figure}
+%\begin{figure}[h!]
+%\includegraphics[width=0.3\textwidth]{figures/SurveyFig2rev}
+%\caption{The balance was right between talks, panels, and lightning talks.
+%\label{fig:SFig2}}
+%\end{figure}
+%\begin{figure}[h!]
+%\includegraphics[width=0.3\textwidth]{figures/SurveyFig3rev}
+%\caption{There was sufficient time for discussion.
+%\label{fig:SFig3}}
+%\end{figure}
+
+\begin{figure}[!htb]
+\minipage{0.32\textwidth}
+  \includegraphics[width=\linewidth]{figures/SurveyFig1rev}
+  \caption{The plenary sessions had a good mix of topics and speakers.\label{fig:SFig1}}
+\endminipage\hfill
+\minipage{0.32\textwidth}
+  \includegraphics[width=\linewidth]{figures/SurveyFig2rev}
+  \caption{The balance was right between talks, panels, and lightning talks.\label{fig:SFig2}}
+\endminipage\hfill
+\minipage{0.35\textwidth}%
+  \includegraphics[width=\linewidth]{figures/SurveyFig3rev}
+  \caption{There was sufficient time for discussion.\label{fig:SFig3}}
+\endminipage
+\end{figure}
+
 \noindent {\bf 2. What part of the meeting was the most valuable? select all that apply}
+
+\todo{move table and figure to be in a single row}
 
@@
 \end{table}
+
+\begin{figure}[h!]
+\includegraphics[width=0.4\textwidth]{figures/SurveyFig4rev}
+\caption{What part of the meeting was the most valuable?
+\label{fig:SFig4}}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+\includegraphics[width=0.4\textwidth]{figures/SurveyFig5rev}
+\caption{Would you be interested in joining a professional society focused around the topics of WSSSPE?
+\label{fig:SFig5}}
+\end{figure}
+
 \noindent {\bf 7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
+
+\todo{move table and figure to be in a single row}
 
@@
 \end{table}
+
+\begin{figure}[h]
+\includegraphics[width=0.4\textwidth]{figures/SurveyFig6rev}
+\caption{Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE?
+\label{fig:SFig6}}
+\end{figure}
 

### 2017-01-10 WSSSPE4_report/WSSSPE4_report.tex
playiung with figure spacing
adding authors of sections that have appeared - info (affil, email, ORCID) needed

@@
 \usepackage{lmodern}
+\usepackage[font=small]{caption}
+\usepackage{varwidth}
 
@@
 Lorraine Hwang$^{(4)}$,
+Alice Allen$^{(5)}$,
+Neil Chue Hong$^{(6)}$,
+Anshu Dubey$^{(7)}$,
+Jonah Miller$^{(8)}$,
+Jean Salac$^{(9)}$,
 Author order to be determined based on contributions
@@
 School of Information Sciences (iSchool),
-University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org}
+University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
 %
 \thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
-Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu}
+Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: ?}
 %
@@
 %
-\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu}
+\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: ?}
+%
+\thanks{{}$^{(5)}$ \note{info needed}}
+%
+\thanks{{}$^{(6)}$ \note{info needed}}
+%
+\thanks{{}$^{(7)}$ \note{info needed}}
+%
+\thanks{{}$^{(8)}$ \note{info needed}}
+%
+\thanks{{}$^{(9)}$ \note{info needed}}
 

### 2017-01-14 WSSSPE4_report/WSSSPE4_report.tex
adding note about keynote summary potential authors
adding social science working group draft

@@
 Anshu Dubey$^{(7)}$,
-Jonah Miller$^{(8)}$,
-Jean Salac$^{(9)}$,
+R. Stuart Geiger$^(8)$,
+Jonah Miller$^{(9)}$,
+Jean Salac$^{(10)}$,
 Author order to be determined based on contributions
@@
 \thanks{{}$^{(9)}$ \note{info needed}}
+%
+\thanks{{}$^{(10)}$ \note{info needed}}
 
@@
 
-\note{need a person to write this}
+\note{need a person to write this - Dan has asked Patricia if she will, if not, perhaps Dan will do it}
 

### 2017-01-14 WSSSPE4_report/wg_social.tex
adding social science working group draft

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Social science for scientific software}
+\label{sec:social}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Stuart to write this}
+
+Introduction to group here, including the overall objective of work in this area. \todo{}
+
+\subsubsection{Participants}
+
+Stuart Geiger <stuart@stuartgeiger.com>, Lorraine Hwang <ljhwang@ucdavis.edu>, Robert McDonald <rhmcdona@indiana.edu>
+
+\subsubsection{Working group objective}
+
+To give social scientists and practitioners working on scientific software, software sustainability, and open source and a better understanding of each other's work, as well as help them connect and coordinate on specific projects of mutual interest.
+
+\subsubsection{Gap or challenge}
+
+There is more and more academic research being done on topics related to software sustainability, including work on software engineering practices and management of open source projects.
+However, academic research in general is often siloed for many reasons, and work on topics relevant to software sustainability is no exception.
+There are many academic studies and projects which may be relevant for practitioners working in this area, and many projects and initiatives by practitioners that may be relevant for social scientists.
+However, there is a gap between these two research and practice.
+Furthermore, we also recognize that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
+
+\subsubsection{Relevant people and resources}
+
+People: a network of invested social scientists and research software engineers who care about research on software engineering practices
+Resources: a mailing list and a central repository for collecting literature and documenting lessons learned
+
+\subsubsection{Plans}
+
+What tasks will the working group undertake \todo{}
+
+\subsubsection{SMART steps}
+
+\begin{itemize}
+\item Collect research questions RSEs have for social scientists (done)
+\item Create a mailing list for people interested in this topic (done) \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
+\item Create a central repository for collecting literature and documenting lessons learned
+\item Publicize mailing list and collect (more) research questions/topics/literature/researchers
+\item Write up synthesis document 
+\end{itemize}
+
+\subsubsection{More information \& joining instructions}
+
+Join the mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}

### 2017-01-16 WSSSPE4_report/WSSSPE4_report.tex
adding prototyping section draft

@@
 Alice Allen$^{(5)}$,
-Neil Chue Hong$^{(6)}$,
-Anshu Dubey$^{(7)}$,
-R. Stuart Geiger$^(8)$,
-Jonah Miller$^{(9)}$,
-Jean Salac$^{(10)}$,
+Jean Salac$^{(6)}$,
+Neil Chue Hong$^{(7)}$,
+Santiago Nunez-Corrales$^{(8)}$
+Anshu Dubey$^{(9)}$,
+R. Stuart Geiger$^(10)$,
+Jonah Miller$^{11)}$,
 Author order to be determined based on contributions
@@
 \thanks{{}$^{(10)}$ \note{info needed}}
+%
+\thanks{{}$^{(11)}$ \note{info needed}}
 

### 2017-01-16 WSSSPE4_report/wg_prototyping.tex
adding prototyping section draft

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Scientific Software Prototyping Infrastructure (S2PI)}
+\label{sec:prototyping}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%from Santiago
+There is a productivity bottleneck---yet to be solved---in HPC from the human
+perspective, first identified in the 1980s~\cite{barstow1982automatic}. Of all
+domain-specific scientists, only a percentage of those use simulations due to
+perceptions mostly related to the difficulty of using and developing those tools.
+XSEDE and other resources have become a gateway for successfully increasing the
+basis of HPC scientific users by facilitating their access to infrastructure and
+tools~\cite{towns2014xsede}, but the development challenge remains unaddressed.
+Of all scientists who use simulations, only a small percentage know how
+to develop prototypes or full applications that can be latter scaled by computer
+scientists and engineers. Writing new parallel applications is unusual for
+domain scientists and thus has created a complex dependency on specialized
+scientific programmers, who are scarce~\cite{post2005computational}. However,
+ the capacity to quickly envision and prototype applications by domain experts
+ ---later to be fully implemented into by scientific programmers---is critical for
+ acquiring maturity and proficiency in the grand challenges the Exascale
+ attempts to solve, and moreover for allowing scientists to develop their own
+ ideas quickly and productively~\cite{vinter2015prototyping}. The latter, in
+ addition, would increase service utilization time in HPC systems, a critical
+measure of efficiency. Considering the existing code base in scientific
+computing, software infrastructures that allow a clean transition from legacy
+systems to new Exascale platforms while preserving flexible prototyping
+ towards the future are central~\cite{hwu2015transitioning}.
+
+This group attempts to tackle the latter challenge through the development of
+novel software prototyping infrastructure with an emphasis at a higher degree
+of abstraction. The group deems the ability to rapidly construct software artifacts that can be
+trusted in terms of the methods from their design up, easily discarded when
+wrong and extended when right at low human and computational cost to be one
+aspect of scientific software sustainability. In essence,
+a motto for scientific software prototyping is {\it fail hard, fail fast}.
+
+\subsubsection{Participants}
+
+\begin{itemize}
+  \item Santiago N\'u\~nez-Corrales, University of Illinois Urbana--Champaign
+        <nunezco2@illinois.edu>
+  \item Chris Sweet, University of Notre Dame <chris.sweet@nd.edu>
+  \item Steven Brandt, Louisiana State University <sbrandt@cct.lsu.edu>
+  \item Dominic Orchard, University of Kent <d.a.orchard@kent.ac.uk>
+\end{itemize}
+
+\subsubsection{Working group objective}
+
+The group's objective is development of a prototype tool that allows domain scientists to generate
+simulation prototypes through a simple, yet expressive declarative
+strongly-typed programming language close to equational expressions. The result
+of that specification is both an executable artifact as well as a specification
+for scientific programmers to later flesh out completely and adapt to particular
+infrastructures. This tool will start in the domain of computational physics,
+and (probably) extend to other domains, and the code will be open source.
+
+The underlying hypothesis of this project is that a strongly typed, declarative
+problem-specification language with adequate constructs, to be implemented
+through stencils and algorithmic skeletons tied to numerical methods, captures a
+ subset of useful and interesting problems in physics and other domains.
+
+\subsubsection{Gap or challenge}
+
+The gap addressed by the project is described by
+following four statements:
+
+\begin{enumerate}
+  \item Domain specific scientists need to strike a balance between being
+  productive in their science areas and having some coding skills.
+  \item Being a research software engineer is a career in its own right which
+  has limited depth in the science basis, especially while serving large
+  communities.
+  \item Collaboration in science is made (mostly informally) through natural
+  language and mathematics, but both present lots of opportunity for potential
+  ambiguity towards software development.
+  \item An unambiguous (formal) language that sets a neutral ground is missing,
+  capable of lifting many of the complexities of constructing useful (and
+  correct) software artifacts that may later evolve into more elaborate codes.
+\end{enumerate}
+
+The latter enumeration of elements is sustained in the fact that scientific
+software packages are digital representations of partial knowledge architectures
+behind research processes.
+
+\subsubsection{Relevant people and resources}
+
+The project is mostly self-contained and only requires development time from
+participants, as well as a Git repository (active). As for skills required, the
+following distribution between project participants was identified:
+
+\begin{description}
+  \item[Steven] parsing expression grammars, prototype stencil
+  \item[Santiago] functional programming, scientific programming
+  \item[Dominic] semantics, types, compilers, languages
+  \item[Chris] numerical methods
+\end{description}
+
+In terms of practice and experience, a sample scientific community is required
+in order to work around the capabilities of the tool. In addition, use cases
+are required (e.g. micromagnetics~\cite{fischbacher2007systematic}).
+
+\subsubsection{Plans}
+
+Our general plan is to develop a functional (meta-)prototype before WSSSPE5, and
+if useful, depart from it towards a larger software infrastructure tied to
+particular cyberinfrastructure resources.
+
+\subsubsection{SMART steps}
+
+\begin{enumerate}
+  \item Define or choose the description language depending on the potential user
+  base.
+  \item Develop the language interpreter/compiler for the equational description
+  language.
+  \item Define a set of fundamental software patterns to translate to algorithmic
+  skeletons.
+  \item Implement a translation mechanism from the language to the patterns.
+  \item Test the system with three known solvable problems and evaluate with a
+  set of functional and non-functional metrics:
+  \begin{enumerate}
+    \item Wave equation
+    \item Newtonian evolution
+    \item Boussinesq equations
+  \end{enumerate}
+\end{enumerate}
+
+\subsubsection{More information \& joining instructions}
+
+S. Brandt has shared the following background links:
+
+\begin{itemize}
+  \item Piraha parser code: \url{https://github.com/stevenrbrandt/piraha-peg/blob/master/piraha.jar}
+  \item A stripped down peg file (grammar) which describes a sequence of equations:
+\url{https://www.cct.lsu.edu/~sbrandt/eqns.peg}
+  \item An example equation file: \url{https://www.cct.lsu.edu/~sbrandt/wave.eq}
+\end{itemize}
+
+For joining to this project, please visit the GitHub repository at
+\url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel
+\texttt{\#wg-sci-soft-proto}.

### 2017-01-19 WSSSPE4_report/wg_letters.tex
adding new section for Wolfgang

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Letters of evaluation for computational scientists}
+\label{sec:letters}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%% See
+%% https://docs.google.com/document/d/1prkkOddRNqIJ4n9qT0XeQIf5TJHWnESdoLFy6rUGL28/edit#heading=h.z17pqk0fi6d
+%% for the document we drew up at the workshop
+
+Letters of reference used for hiring, tenure, and promotion purposes
+are typically first read by departmental committees that have
+expertise in the core areas represented by the department. On the
+other hand, computational scientists and in particular researchers
+working on scientific software -- more or less by definition --
+are typically interdisciplinary, and the achievements that are
+assessed in such letters are consequently often meaningless or at
+least hard to evaluate for disciplinary committees.
+
+Both committees and letter writers need to be aware of this. Providing
+best practice examples, and training both writers and recipients of
+letters may be necessary to level the playing field for computational
+scientists.
+
+\subsubsection{Participants}
+
+Alice Allen, Astrophysics Source Code Library, \url{aallen@ascl.net};
+Gabrielle Allen, NCSA, \url{University of Illinois, gdallen@illinois.edu};
+Wolfgang Bangerth, Colorado State University, \url{bangerth@colostate.edu};
+Bruce Childers, University of Pittsburgh, \url{childers@cs.pitt.edu };
+Lorraine Hwang, University of California, \url{Davis, ljhwang@ucdavis.edu };
+Dan Katz, University of Illinois, \url{dskatz@illinois.edu };
+Frank L{\"o}ffler, Louisiana State University, \url{knarf@cct.lsu.edu};
+Kyle Niemeyer, Oregon State University, \url{kyle.niemeyer@oregonstate.edu};
+Janos Sallai, Vanderbilt University, \url{janos.sallai@vanderbilt.edu }.
+
+\subsubsection{Working group objective}
+
+Scientists working on scientific software are often located in
+disciplinary departments, depending on whether their software
+originates from the mathematical, physical, chemical, or other
+disciplines. As a consequence, they are frequently outside the core
+areas of their science, and their contributions are typically to both
+the research activity their software enables, as well as on
+algorithmic and implementational aspects. This presents issues when
+letters of evaluation for hiring, tenure, and promotion do not specifically
+cover how this is relevant to the discipline.%
+\footnote{The extensive use of such letters, and the problems that are
+  associated with it, may be an issue specific to the US.}
+
+We believe that the authors of scientific software provide important
+services to departments that are no less than strictly disciplinary
+research. Consequently, in order to level the playing field with more
+disciplinary candidates for hires, tenure, and promotion requires that
+letter writers are aware of how their letters will be read by
+committees. It also requires that committees are aware that such
+letters often look different and may provide a different perspective
+of how a candidate's achievements should be assessed. For example, in
+mathematics a typical candidate would be evaluated on the difficulty
+and depth of the statements she may have proven in their papers. On
+the other hand, an author of mathematical software would likely be
+evaluated based on the impact of her software, or the number of citations of
+the publications that describe it. She may also be evaluated by how
+widely the software is used \textit{outside} mathematics, a criterion
+that is rarely used for more disciplinary mathematicians.
+
+\subsubsection{Gap or challenge}
+
+Addressing this problem likely requires that letter writers pay
+particular attention to who exactly the audience of such letters is,
+and tailor the message by specifically highlighting how the work of a
+candidate benefits the department and discipline that is considering a
+candidate. On the other hand, committee members also need to pay
+attention that there are areas that are important to the mission of
+their department and professional community in which different
+standards for evaluation hold.
+
+\subsubsection{Relevant people and resources}
+
+Affecting letter writers and readers essentially requires raising
+awareness of the issue. Various organizations have attempted to do so
+through letters to the community, adjusting definitions of what counts
+in science, etc. Specifically, we are aware of the following
+resources:
+
+\begin{itemize}
+\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue.
+\item The National Science Foundation has produced a ``Dear Colleague'' letter that defines a ``product'' for the purposes of 2-page biosketches that does not include only publications, but also software.
+\item The ACM published an article in the CACM (one of the co-authors
+  was David Patterson) that explained that conference proceedings are
+  \textit{the} venue in Computer Science, and this has been used to
+  convince deans and provosts of the value of these proceedings
+  publications. This article may serve as a template for computational
+  science.
+\item There is also a 1995 report by the National Academies about
+  Computer Science that addresses some of these issues, specifically
+  on how writing software should be considered in comparison to more
+  theoretical research.
+\end{itemize}
+
+\subsubsection{Plans}
+
+There are essentially two important strategies that build on each
+other: (i) raising awareness of the problem beyond just those who are
+affected by it (namely, computational scientists working on scientific
+software), and (ii) providing letter writers, letter readers, and
+evaluating committees with guidance on what criteria are relevant in
+assessing scientific software authors.
+
+Concrete guidance -- for example in the form of ``Dear Colleague''
+letters as mentioned in the previous section -- is most valuable if it
+comes from respected bodies such as professional societies or
+established and respected organizations. Getting these to act
+will only be possible if the problem is widely acknowledged, and so
+the first of the goals above should be the current focus.
+
+We will attempt to address it by organizing sessions at conferences
+that address the career problem, as well as writing editorials that
+can be published in the magazines of our professional societies. These
+editorials ought to outline best practices for letter writers that
+specifically (i) make clear the contribution of a candidate to their
+interdisciplinary area, (ii) the relevance to their home discipline,
+(iii) why writing software is good for the discipline itself.
+
+\subsubsection{SMART steps}
+
+We have identified a few next steps, even if they may not satisfy the
+exact criteria of the SMART system. Specifically:
+\begin{itemize}
+  \item Find and review the documents listed above, send out summary
+    (Wolfgang Bangerth)
+  \item Gather feedback from the other members of the group
+  \item Write editorial for SIAM (Wolfgang)
+  \item Write 2 emails to each other group members listed above about
+    also writing editorials in their communities.
+\end{itemize}
+In the longer term we hope to gather others in our home disciplines to
+let professional organizations weigh in.
+
+\subsubsection{More information \& joining instructions}
+
+The section on resources above lists a number of letters and reports
+that are worth reading. Ultimately, building a community large enough
+to affect change is important; contact Wolfgang Bangerth at
+\url{bangerth@colostate.edu} if you are interested in helping.

### 2017-01-19 WSSSPE4_report/wg_letters.tex
small edits in letters section, adding Bangerth as author

@@
 the research activity their software enables, as well as on
-algorithmic and implementational aspects. This presents issues when
+algorithm and implementation aspects. This presents issues when
 letters of evaluation for hiring, tenure, and promotion do not specifically
@@
 disciplinary candidates for hires, tenure, and promotion requires that
-letter writers are aware of how their letters will be read by
-committees. It also requires that committees are aware that such
+letter writers be aware of how their letters will be read by
+committees. It also requires that committees be aware that such
 letters often look different and may provide a different perspective
@@
 candidate. On the other hand, committee members also need to pay
-attention that there are areas that are important to the mission of
+attention to the fact that there are areas that are important to the mission of
 their department and professional community in which different
@@
 \begin{itemize}
-\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue.
+\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue. \katznote{add citation}
 \item The National Science Foundation has produced a ``Dear Colleague'' letter that defines a ``product'' for the purposes of 2-page biosketches that does not include only publications, but also software.
+\katznote{add citation or link?}
 \item The ACM published an article in the CACM (one of the co-authors
-  was David Patterson) that explained that conference proceedings are
+  was David Patterson) \katznote{why is Patterson significant here?} that explained that conference proceedings are
   \textit{the} venue in Computer Science, and this has been used to
@@
   on how writing software should be considered in comparison to more
-  theoretical research.
+  theoretical research. \katznote{add citation -- \url{https://www.nap.edu/catalog/2236/academic-careers-for-experimental-computer-scientists-and-engineers}}
 \end{itemize}
@@
 
-Concrete guidance -- for example in the form of ``Dear Colleague''
-letters as mentioned in the previous section -- is most valuable if it
+Concrete guidance---for example in the form of ``Dear Colleague''
+letters as mentioned in the previous section---is most valuable if it
 comes from respected bodies such as professional societies or
@@
 specifically (i) make clear the contribution of a candidate to their
-interdisciplinary area, (ii) the relevance to their home discipline,
+interdisciplinary area, (ii) the relevance to their home discipline, and
 (iii) why writing software is good for the discipline itself.

### 2017-01-31 WSSSPE4_report/WSSSPE4_report.tex
small edits in letters section, adding Bangerth as author
changes in section 2 to discuss submission numbers and acceptances

@@
 Jonah Miller$^{11)}$,
+Wolfgang Bangerth$^{12)}$,
 Author order to be determined based on contributions
@@
 \thanks{{}$^{(11)}$ \note{info needed}}
+%
+\thanks{{}$^{(12)}$ \note{info needed}}
 
@@
 
-There were~24 lightning talks submitted; after a peer-review process, 16
-were accepted, as discussed further in Section~\ref{sec:lightning}.
-\todo{update, add paper stats too}
+The following list of contribution topics was suggested, though this was not intended to limit potential submissions.
 
@@
 
-\todo{need to mention/link to the proceedings somewhere, though maybe not here.}
+Submissions to WSSSPE4 comprised
+20 lightning talks,
+4 idea papers,
+3 position paper,
+5 experience papers,
+and
+3 demos.
+Most of the submissions were accepted, since the goal of WSSSPE is always to
+take in as many relevant inputs as possible, and to encourage their authors to
+participate in sharing and implementing their ideas.
+Specially, 
+18 lightning talks,
+4 idea papers,
+3 position paper,
+3 experience papers,
+and
+3 demos
+were accepted.
+The papers are discussed in Section~\ref{sec:papers},
+and the lightning talks are discussed in Section~\ref{sec:lightning}.
+
+\todo{need to mention/link to the proceedings here.}
 

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
added ORCID for KN

@@
 \thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
-Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: ?}
+Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
 %

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
added lightning talk authors and titles

@@
 \newcommand{\nchnote}[1]{  {\textcolor{orange}      { ***Neil: #1 }}}
+\newcommand{\kylenote}[1]{  {\textcolor{cyan}      { ***Kyle: #1 }}}
 
@@
 
-Intro and description \katznote{not sure enum list is right here}
+Intro and description.
+A \textsuperscript{\textasteriskcentered} indicates which author presented the
+lightning talk, for those with multiple authors.
+\katznote{not sure enum list is right here}
+\kylenote{I'm fine with itemize, but they were in a particular order, right?}
 %
 \begin{enumerate}
-\item \textbf{Authors: \emph{title}}.
+    \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
+    programme; supporting the social side of research software}}.
+    Description
+    %Didn’t want to “preach” to the community, wanted to be integrated with the
+    % community, hence the fellows programme. Five attending WSSSPE this year.
+
+    \item \textbf{Sandra Gesing, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
+    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
+    \emph{The Science Gateways Community Institute - Supporting Communities to
+    Achieve Sustainability for Their Science Gateways}}.
+    Description
+
+    \item \textbf{Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
+    and Scott Chamberlain: \emph{A model for
+    peer review and onboarding research software}}.
+    Description
+    %Some people greatly encouraged to only submit to a journal code that has been peer reviewed
+
+    \item \textbf{Frank Löffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
+    \emph{A vision of computing in 10+ years}}.
+    Description
+    %“We (need to) decouple science from numerical methods and optimization” Hard open problem.
+
+    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
+    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
+    \emph{Software Citation: Process, Principles, and Implementation}}.
+    Description
+
+    \item \textbf{Ray Idaszak\textsuperscript{\textasteriskcentered},
+    David G.~Tarboton, Hong Yi, Michael Stealey,
+    Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
+    Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
+    Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
+    Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
+    Best Practices and Culture Change for Developing Sustainable Community Software}}.
+    Description
+
+    \item \textbf{Lorraine Hwang\textsuperscript{\textasteriskcentered},
+    Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
+    \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source Community}}.
+    Description
+
+    \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
+    User-Producer Maturity Review}}.
+    Description
+
+    \item \textbf{Stephan Druskat: \emph{A proposal for the measurement and
+    documentation of research software sustainability in interactive metadata
+    repositories}}.
+    Description
+
+    \item \textbf{Mistral Contrastin, Matthew Danish,
+    Dominic Orchard\textsuperscript{\textasteriskcentered}, and
+    Andrew Rice: \emph{Supporting software sustainability with
+    lightweight specifications}}.
+    Description
+
+    \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
+    Andrei Chiș, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
+    Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
+    Robert Haines, Robert Hirschfeld, James Howison,
+    Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
+    Ralf Lämmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
+    and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
+    Responsibility in the Engineering of Academic Software}}.
+    Description
+
+    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
+    and Gabrielle Allen: \emph{Bringing Techniques from Software Engineering
+    into Scientific Software}}.
     Description

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
added lightning talk refs to paper

--- 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
@@
 
-Intro and description \katznote{not sure enum list is right here}
+Intro and description.
+A \textsuperscript{\textasteriskcentered} indicates which author presented the
+lightning talk, for those with multiple authors.
+\katznote{not sure enum list is right here}
+\kylenote{I'm fine with itemize, but they were in a particular order, right?}
 %
 \begin{enumerate}
-\item \textbf{Authors: \emph{title}}.
+    \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
+    programme; supporting the social side of research software}}~\cite{Sufi:2016ws}.
+    Description
+    %Didn’t want to “preach” to the community, wanted to be integrated with the
+    % community, hence the fellows programme. Five attending WSSSPE this year.
+
+    \item \textbf{Sandra Gesing, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
+    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
+    \emph{The Science Gateways Community Institute - Supporting Communities to
+    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
+    Description
+
+    \item \textbf{Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
+    and Scott Chamberlain: \emph{A model for
+    peer review and onboarding research software}}~\cite{Ram:2016ws}.
+    Description
+    %Some people greatly encouraged to only submit to a journal code that has been peer reviewed
+
+    \item \textbf{Frank Löffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
+    \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
+    Description
+    %“We (need to) decouple science from numerical methods and optimization” Hard open problem.
+
+    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
+    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
+    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
+    Description
+
+    \item \textbf{Ray Idaszak\textsuperscript{\textasteriskcentered},
+    David G.~Tarboton, Hong Yi, Michael Stealey,
+    Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
+    Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
+    Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
+    Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
+    Best Practices and Culture Change for Developing Sustainable Community
+    Software}}~\cite{Idaszak:2016ws}.
+    Description
+
+    \item \textbf{Lorraine Hwang\textsuperscript{\textasteriskcentered},
+    Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
+    \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
+    Community}}~\cite{Hwang:2016ws}.
+    Description
+
+    \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
+    User-Producer Maturity Review}}~\cite{Goble:2016ws}.
+    Description
+
+    \item \textbf{Stephan Druskat: \emph{A proposal for the measurement and
+    documentation of research software sustainability in interactive metadata
+    repositories}}~\cite{Druskat:2016ws}.
+    Description
+
+    \item \textbf{Mistral Contrastin, Matthew Danish,
+    Dominic Orchard\textsuperscript{\textasteriskcentered}, and
+    Andrew Rice: \emph{Supporting software sustainability with
+    lightweight specifications}}~\cite{Contrastin:2016ws}.
+    Description
+
+    \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
+    Andrei Chiș, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
+    Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
+    Robert Haines, Robert Hirschfeld, James Howison,
+    Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
+    Ralf Lämmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
+    and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
+    Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
+    Description
+
+    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
+    and Gabrielle Allen: \emph{Bringing Techniques from Software Engineering
+    into Scientific Software}}~\cite{Seidel:2016ws}.
     Description

### 2017-02-08 WSSSPE4_report/survey_details.tex
some work improving appendix E tables & figures (not done)

@@
 \noindent {\bf 1. Please indicate how strongly you agree or disagree with the following statements}
-
-\todo{make this table prettier! fix vertical spacing}
 
@@
 {\scriptsize
-\begin{tabular}{|l|cccccc|}
-\hline
-{\bf Question} &
-{\bf strongly} &
-{\bf somewhat}  &
-{\bf agree} &
-{\bf somewhat}  &
-{\bf strongly}  &
-{\bf Total} \\
-&
-{\bf disagree} &
-{\bf disagree} &
-&
-{\bf agree} &
-{\bf agree} &
-\\ \hline
-The plenary sessions had a good&
-5 &
-3 &
-8 &
-11 &
-18 &
-45 \\
-\hspace{.2cm}mix of topics and speakers. &
- &
- &
- &
- &
- &
- \\
+\begin{tabular}{@{}lcccccc@{}}
+\toprule
+\multirow{2}{*}{Question} & strongly & somewhat & agree & somewhat & strongly &
+\multirow{2}{*}{Total} \\
+& disagree & disagree & & agree & agree & \\
+\midrule
+\rowcolor{lightgray} The plenary sessions had a good & & & & & & \\
+\rowcolor{lightgray} mix of topics and speakers. &
+\multirow{-2}{*}{5} &
+\multirow{-2}{*}{3} &
+\multirow{-2}{*}{8} &
+\multirow{-2}{*}{11} &
+\multirow{-2}{*}{18} &
+\multirow{-2}{*}{45} \\
 The balance was right between &
-6 &
-2 &
-7 &
-9 &
-21 &
-45 \\
-\hspace{.2cm}talks, panels, and lightning talks. &
-&
- &
- &
- &
- &
- \\
-There was sufficient time for discussion. &
+\multirow{2}{*}{6} &
+\multirow{2}{*}{2} &
+\multirow{2}{*}{7} &
+\multirow{2}{*}{9} &
+\multirow{2}{*}{21} &
+\multirow{2}{*}{45} \\
+talks, panels, and lightning talks. & & & & & & \\
+\rowcolor{lightgray} There was sufficient time for discussion. &
 6 &
@@
 45 \\
-\hline
+\bottomrule
 \end{tabular}
@@
 \begin{figure}[!htb]
-\minipage{0.32\textwidth}
+    \begin{subfigure}{0.34\textwidth}
     \includegraphics[width=\linewidth]{figures/SurveyFig1rev}
-  \caption{The plenary sessions had a good mix of topics and speakers.\label{fig:SFig1}}
-\endminipage\hfill
-\minipage{0.32\textwidth}
+    \caption{The plenary sessions had a good mix of topics and speakers.}
+    \label{fig:SFig1}
+    \end{subfigure}
+    ~
+    \begin{subfigure}{0.32\textwidth}
     \includegraphics[width=\linewidth]{figures/SurveyFig2rev}
-  \caption{The balance was right between talks, panels, and lightning talks.\label{fig:SFig2}}
-\endminipage\hfill
-\minipage{0.35\textwidth}%
+    \caption{The balance was right between talks, panels, and lightning talks.}
+    \label{fig:SFig2}
+    \end{subfigure}
+    \begin{subfigure}{0.25\textwidth}
     \includegraphics[width=\linewidth]{figures/SurveyFig3rev}
-  \caption{There was sufficient time for discussion.\label{fig:SFig3}}
-\endminipage
+    \caption{There was sufficient time for discussion.}
+    \label{fig:SFig3}
+    \end{subfigure}
+    \caption{Responses to generate questions about the conference program}
 \end{figure}
@@
 \begin{table}[h!]
+\rowcolors{1}{}{lightgray}
 \centering

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
some work improving appendix E tables & figures (not done)
fixing an incorrect word and some special characters

@@
 \usepackage{color}
+\usepackage[table]{xcolor}
 \usepackage{dcolumn}
@@
 \usepackage{rotating}
-\usepackage{subfigure}
+\usepackage{caption,subcaption}
 \usepackage{psfrag}
@@
 \usepackage{verbatim}
-\usepackage{booktabs,multicol,multirow}
+\usepackage{array,booktabs,multicol,multirow}
 
@@
 \definecolor{darkblue}{rgb}{0.1,0.1,0.9}
+\definecolor{lightgray}{gray}{0.9}
 
@@
 participate in sharing and implementing their ideas.
-Specially,
+Specifically,
 18 lightning talks,
@@
 
-    \item \textbf{Frank Löffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
+    \item \textbf{Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
     \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
@@
     \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
-    Andrei Chiș, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
+    Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
     Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
@@
     Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
-    Ralf Lämmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
+    Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
     and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
note about lightning talk list type and order
minor formatting fix

@@
 
-    \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
+    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
     has the goal of improving the quality of today's research software and the
@@
 \kylenote{I'm fine with itemize, but they were in a particular order, right?}
+\katznote{yes, let's go with itemize, but keep them in the same order}
 %

### 2017-02-08 WSSSPE4_report/WSSSPE4_report.tex
metrics section copied from Emily's google doc

@@
 
-\author{Daniel S.\ Katz$^{(1)}$,
-Kyle E.\ Niemeyer$^{(2)}$,
-Sandra Gesing$^{(3)}$,
-Lorraine Hwang$^{(4)}$,
-Alice Allen$^{(5)}$,
-Jean Salac$^{(6)}$,
-Neil Chue Hong$^{(7)}$,
-Santiago Nunez-Corrales$^{(8)}$
-Anshu Dubey$^{(9)}$,
-R. Stuart Geiger$^(10)$,
-Jonah Miller$^{11)}$,
-Wolfgang Bangerth$^{12)}$,
-Author order to be determined based on contributions
+\author{Daniel~S.\~Katz$^{(1)}$,
+Kyle~E.\~Niemeyer$^{(2)}$,
+Sandra~Gesing$^{(3)}$,
+Lorraine~Hwang$^{(4)}$,
+Alice~Allen$^{(5)}$,
+Jean~Salac$^{(6)}$,
+Neil~Chue~Hong$^{(7)}$,
+Santiago~Nunez-Corrales$^{(8)}$
+Anshu~Dubey$^{(9)}$,
+R.~Stuart~Geiger$^(10)$,
+Jonah~Miller$^{11)}$,
+Emily~Chen$^{12)}$,
+Wolfgang~Bangerth$^{13)}$,
+\\Author order to be determined based on contributions
 }
@@
 \thanks{{}$^{(12)}$ \note{info needed}}
+%
+\thanks{{}$^{(13)}$ \note{info needed}}
 

### 2017-02-08 WSSSPE4_report/survey_details.tex
remaining formatting fixes for survey appendix

@@
 
-\noindent {\bf 1. Please indicate how strongly you agree or disagree with the following statements}
+\noindent \textbf{1. Please indicate how strongly you agree or disagree with the following statements}
 
@@
 \begin{figure}[!htb]
-    \begin{subfigure}{0.34\textwidth}
-    \includegraphics[width=\linewidth]{figures/SurveyFig1rev}
-    \caption{The plenary sessions had a good mix of topics and speakers.}
+    \captionsetup[subfigure]{position=b}
+    \centering
+    \subcaptionbox{
+    The plenary sessions had a good mix of topics and speakers.
     \label{fig:SFig1}
-    \end{subfigure}
+    }{\includegraphics[width=.34\linewidth]{figures/SurveyFig1rev}}
     ~
-    \begin{subfigure}{0.32\textwidth}
-    \includegraphics[width=\linewidth]{figures/SurveyFig2rev}
-    \caption{The balance was right between talks, panels, and lightning talks.}
+    \subcaptionbox{
+    The balance was right between talks, panels, and lightning talks.
     \label{fig:SFig2}
-    \end{subfigure}
-    \begin{subfigure}{0.25\textwidth}
-    \includegraphics[width=\linewidth]{figures/SurveyFig3rev}
-    \caption{There was sufficient time for discussion.}
+    }{\includegraphics[width=.32\linewidth]{figures/SurveyFig2rev}}
+    ~
+    \subcaptionbox{
+    There was sufficient time for discussion.
     \label{fig:SFig3}
-    \end{subfigure}
+    }{\includegraphics[width=.25\linewidth]{figures/SurveyFig3rev}}
     \caption{Responses to generate questions about the conference program}
@@
 
-\noindent {\bf 2. What part of the meeting was the most valuable? select all that apply}
-
-\todo{move table and figure to be in a single row}
-
-\begin{table}[h!]
-\rowcolors{1}{}{lightgray}
-\centering
-\caption{Value of meeting parts}
-\label{tab:survey_parts}
-\begin{tabular}{|l|c|}
-\hline
-{\bf Answer} &
-{\bf Count} \\ \hline
-Plenary &
-19 \\
-Panel &
-14 \\
-Working Groups &
-36 \\
-Other &
-5 \\
-Total &
-74 \\
-\hline
-\end{tabular}
-\end{table}
+\noindent \textbf{2. What part of the meeting was the most valuable? select all that apply}
 
 \begin{figure}[h!]
-\includegraphics[width=0.4\textwidth]{figures/SurveyFig4rev}
+    \CenterFloatBoxes
+    \begin{floatrow}
+    \ffigbox[\FBwidth]{%
+        \includegraphics[width=0.35\textwidth]{figures/SurveyFig4rev}%
+    }{%
+        \captionsetup{width=\linewidth}
         \caption{What part of the meeting was the most valuable?
-\label{fig:SFig4}}
+        \label{fig:SFig4}}%
+    }
+    \hfill
+    \killfloatstyle\capbtabbox[\linewidth]{%
+    \rowcolors{1}{}{lightgray}
+    \begin{tabular}{@{}ll@{}}
+    \toprule
+    Answer & Count \\
+    \midrule
+    Plenary & 19 \\
+    Panel & 14 \\
+    Working Groups & 36 \\
+    Other & 5 \\
+    Total & 74 \\
+    \bottomrule
+    \end{tabular}
+    }{%
+        \caption{Value of meeting parts}%
+    }
+    \end{floatrow}
 \end{figure}
 
-\noindent {\bf 3. What additional topics would you like to see discussed at the next meeting?}
+\noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}
 
@@
 
-\noindent {\bf 4. What could be improved?}
+\noindent \textbf{4. What could be improved?}
 
@@
 
-\noindent {\bf 5. What would make you more likely to invest time in future WSSSPE initiatives?}
+\noindent \textbf{5. What would make you more likely to invest time in future WSSSPE initiatives?}
 
@@
 
-\noindent {\bf 6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}
+\noindent \textbf{6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}
 
-\todo{move table and figure to be in a single row}
-
-\begin{table}[h!]
+\begin{figure}[h!]
+    \CenterFloatBoxes
     \centering
-\caption{A WSSSPE society?}
+    \begin{floatrow}
+    \ffigbox[\FBwidth]{%
+        \includegraphics[width=0.35\textwidth]{figures/SurveyFig5rev}%
+    }{%
+        \captionsetup{width=\linewidth}
+        \caption{Would you be interested in joining a professional society
+        focused around the topics of WSSSPE?
+        \label{fig:SFig5}}%
+    }
+    %\hfill
+    ~
+    \killfloatstyle\capbtabbox[\linewidth]{%
+    \rowcolors{1}{}{lightgray}
+    \begin{tabular}{@{}ll@{}}
+    \toprule
+    Answer & Count \\
+    \midrule
+    Yes & 24 \\
+    No & 3 \\
+    Maybe & 18 \\
+    Total & 45 \\
+    \bottomrule
+    \end{tabular}
+    }{%
+        \caption{A WSSSPE society?}%
         \label{tab:survey_society}
-\begin{tabular}{|l|c|}
-\hline
-{\bf Answer} &
-{\bf Count} \\ \hline
-Yes &
-24 \\
-No &
-3 \\
-Maybe &
-18 \\
-Total &
-45 \\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{figure}[h]
-\includegraphics[width=0.4\textwidth]{figures/SurveyFig5rev}
-\caption{Would you be interested in joining a professional society focused around the topics of WSSSPE?
-\label{fig:SFig5}}
+    }
+    \end{floatrow}
 \end{figure}
 
-\noindent {\bf 7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
+% \begin{table}[h!]
+% \centering
+% \caption{A WSSSPE society?}
+% \label{tab:survey_society}
+% \begin{tabular}{|l|c|}
+% \hline
+% {\bf Answer} &
+% {\bf Count} \\ \hline
+% Yes &
+% 24 \\
+% No &
+% 3 \\
+% Maybe &
+% 18 \\
+% Total &
+% 45 \\
+% \hline
+% \end{tabular}
+% \end{table}
+%
+% \begin{figure}[h]
+% \includegraphics[width=0.4\textwidth]{figures/SurveyFig5rev}
+% \caption{Would you be interested in joining a professional society focused around the topics of WSSSPE?
+% \label{fig:SFig5}}
+% \end{figure}
 
-\todo{move table and figure to be in a single row}
+\noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
 
-\begin{table}[h!]
+\begin{figure}[h!]
+    \CenterFloatBoxes
     \centering
-\caption{Continuing working groups?}
+    \begin{floatrow}
+    \ffigbox[\FBwidth]{%
+        \includegraphics[width=0.35\textwidth]{figures/SurveyFig6rev}%
+    }{%
+        \captionsetup{width=\linewidth}
+        \caption{Will you be part of a working group  with colleagues going
+        forward  on a topic addressed at WSSSPE?
+        \label{fig:SFig6}}%
+    }
+    ~
+    \killfloatstyle\capbtabbox[\linewidth]{%
+    \rowcolors{1}{}{lightgray}
+    \begin{tabular}{@{}ll@{}}
+    \toprule
+    Answer & Count \\
+    \midrule
+    Yes & 32 \\
+    No & 4 \\
+    Maybe & 9 \\
+    Total & 45 \\
+    \bottomrule
+    \end{tabular}
+    }{%
+        \caption{Continuing working groups?}%
         \label{tab:survey_continuing_wgs}
-\begin{tabular}{|l|c|}
-\hline
-{\bf Answer} &
-{\bf Count} \\ \hline
-Yes &
-32 \\
-No &
-4 \\
-Maybe &
-9 \\
-Total &
-45 \\
-\hline
-\end{tabular}
-\end{table}
-
-\begin{figure}[h]
-\includegraphics[width=0.4\textwidth]{figures/SurveyFig6rev}
-\caption{Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE?
-\label{fig:SFig6}}
+    }
+    \end{floatrow}
 \end{figure}
 
-\noindent {\bf 8. Working Group Name}
+% \begin{table}[h!]
+% \centering
+% \caption{Continuing working groups?}
+% \label{tab:survey_continuing_wgs}
+% \begin{tabular}{@{}ll@{}}
+% \toprule
+% Answer & Count \\
+% \midrule
+% Yes & 32 \\
+% No & 4 \\
+% Maybe & 9 \\
+% Total & 45 \\
+% \bottomrule
+% \end{tabular}
+% \end{table}
+
+% \begin{figure}[h]
+% \includegraphics[width=0.4\textwidth]{figures/SurveyFig6rev}
+% \caption{Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE?
+% \label{fig:SFig6}}
+% \end{figure}
+
+\noindent \textbf{8. Working Group Name}
 
@@
 
-\noindent {\bf 9. Do you have any other comments or suggestions or ideas?}
+\noindent \textbf{9. Do you have any other comments or suggestions or ideas?}
 

### 2017-02-08 WSSSPE4_report/survey_details.tex
small caption changes
fixed formatting for combined figure & table

@@
 \begin{figure}[h!]
-    \CenterFloatBoxes
-    \begin{floatrow}
-    \ffigbox[\FBwidth]{%
+    \centering
+    \begin{tabular}{@{}c@{}}
         \includegraphics[width=0.35\textwidth]{figures/SurveyFig4rev}%
-    }{%
-        \captionsetup{width=\linewidth}
-        \caption{What part of the meeting was the most valuable?
-        \label{fig:SFig4}}%
-    }
-    \hfill
-    \killfloatstyle\capbtabbox[\linewidth]{%
+    \end{tabular}
+    \qquad
     \rowcolors{1}{}{lightgray}
-    \begin{tabular}{@{}ll@{}}
+    \begin{tabular}{@{}lc@{}}
     \toprule
@@
     \end{tabular}
-    }{%
-        \caption{Value of meeting parts}%
-    }
-    \end{floatrow}
+    \caption{Responses to ``What part of the meeting was the most valuable?''}
+    \label{fig:SFig4}
 \end{figure}
@@
 \begin{figure}[h!]
-    \CenterFloatBoxes
     \centering
-    \begin{floatrow}
-    \ffigbox[\FBwidth]{%
+    \begin{tabular}{@{}c@{}}
         \includegraphics[width=0.35\textwidth]{figures/SurveyFig5rev}%
-    }{%
-        \captionsetup{width=\linewidth}
-        \caption{Would you be interested in joining a professional society
-        focused around the topics of WSSSPE?
-        \label{fig:SFig5}}%
-    }
-    %\hfill
-    ~
-    \killfloatstyle\capbtabbox[\linewidth]{%
+    \end{tabular}
+    \qquad
     \rowcolors{1}{}{lightgray}
-    \begin{tabular}{@{}ll@{}}
+    \begin{tabular}{@{}lc@{}}
     \toprule
@@
     \end{tabular}
-    }{%
-        \caption{A WSSSPE society?}%
-        \label{tab:survey_society}
-    }
-    \end{floatrow}
+    \caption{Responses to ``Would you be interested in joining a professional
+    society focused around the topics of WSSSPE?''}
+    \label{fig:SFig5}
 \end{figure}
@@
 \begin{figure}[h!]
-    \CenterFloatBoxes
     \centering
-    \begin{floatrow}
-    \ffigbox[\FBwidth]{%
+    \begin{tabular}{@{}c@{}}
         \includegraphics[width=0.35\textwidth]{figures/SurveyFig6rev}%
-    }{%
-        \captionsetup{width=\linewidth}
-        \caption{Will you be part of a working group  with colleagues going
-        forward  on a topic addressed at WSSSPE?
-        \label{fig:SFig6}}%
-    }
-    ~
-    \killfloatstyle\capbtabbox[\linewidth]{%
+    \end{tabular}
+    \qquad
     \rowcolors{1}{}{lightgray}
-    \begin{tabular}{@{}ll@{}}
+    \begin{tabular}{@{}lc@{}}
     \toprule
@@
     \end{tabular}
-    }{%
-        \caption{Continuing working groups?}%
-        \label{tab:survey_continuing_wgs}
-    }
-    \end{floatrow}
+    \caption{Responses to ``Will you be part of a working group  with colleagues
+    going forward  on a topic addressed at WSSSPE?''}
+    \label{fig:SFig6}
 \end{figure}

### 2017-02-09 WSSSPE4_report/WSSSPE4_report.tex
added changes to main report file missing from last commit
changed lightning talks to itemize
fixed formatting for combined figure & table
final formatting changes in survey details, new page for references
adding proceedings to body

@@
 \usepackage{dcolumn}
-\usepackage{float,caption}
+\usepackage{caption}
 \usepackage{graphicx}
@@
 \usepackage{caption,subcaption}
+\usepackage{float}
 \usepackage{psfrag}
@@
 
-\author{Daniel~S.\~Katz$^{(1)}$,
-Kyle~E.\~Niemeyer$^{(2)}$,
+\author{Daniel~S.~Katz$^{(1)}$,
+Kyle~E.~Niemeyer$^{(2)}$,
 Sandra~Gesing$^{(3)}$,
@@
 and the lightning talks are discussed in Section~\ref{sec:lightning}.
-
-\todo{need to mention/link to the proceedings here.}
+The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
 
@@
 
-\note{by Lorraine and Dan}
+%\note{by Lorraine and Dan}
 
@@
 lightning talk, for those with multiple authors.
-\katznote{not sure enum list is right here}
-\kylenote{I'm fine with itemize, but they were in a particular order, right?}
-\katznote{yes, let's go with itemize, but keep them in the same order}
 %
-\begin{enumerate}
+\begin{itemize}
     \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
@@
 
-\end{enumerate}
+\end{itemize}
 
@@
 
+\newpage
 \bibliographystyle{vancouver}

### 2017-02-11 WSSSPE4_report/WSSSPE4_report.tex
add group formation process

@@
 
-Intro and description \katznote{not sure enum list is right here}
+Intro and description 
 %
-\begin{enumerate}
+\begin{itemize}
 \item \textbf{Authors: \emph{title}}.
@@
 
-\todo{describe process for assembling people into working groups: vision, gaps, projects exercise, then soapboxes}
+After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
+attendees use sticky flip charts and sticky notes on the walls to suggest 
+\begin{enumerate}
+\item A vision on any aspect of the work.
+\item A gap or challenge
+\item A project idea
+\end{enumerate}
+(with one area of the room for each topic.)
+
+Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall.
+\todo{find photos of flip charts, put in repo?} We asked them to work in Google Docs, \todo{find and cite these docs?} using a template for Vision, Gaps, and Projects.
+For those folk who worked on Vision and Gaps, after an hour, we encouraged them to finish their Vision or Gap Google Doc, and to find and join or start working on a project with a group.
+This led to the 12 groups that are discussed in the rest of this section, specifically:
+\begin{enumerate}
+\item Verifying best practices \& metrics for sustainable research software
+\item Software Sustainability Alliance
+\item Scientific Software Prototyping Infrastructure (S2PI)
+\item Standard metadata for software (CodeMeta)
+\item White paper on developing sustainable software
+\item Social science for scientific software
+\item Software best practices for undergraduates
+\item Meaningful metrics for sustainable software
+\item Coordinating access to CI for research software
+\item Software engineering processes tailored for research software
+\item Open research index
+\item Letters of evaluation for computational scientists
+\end{enumerate}
 
@@
 
-As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE.
+As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE~\cite{WSSSPESlack}.
 To join it, go to \url{https://wssspe.signup.team/}.

### 2017-02-14 WSSSPE4_report/WSSSPE4_report.tex
added an initial lightning talk description.

--- 2017-02-09 WSSSPE4_report/WSSSPE4_report.tex
@@
 
-Intro and description \katznote{not sure enum list is right here}
+Intro and description
 %
-\begin{enumerate}
+\begin{itemize}
 \item \textbf{Authors: \emph{title}}.
@@
     programme; supporting the social side of research software}}~\cite{Sufi:2016ws}.
-    Description
-    %Didn’t want to “preach” to the community, wanted to be integrated with the
-    % community, hence the fellows programme. Five attending WSSSPE this year.
+    Sufi, Community Lead at the Software Sustainability Institute (SSI), discussed
+    the SSI Fellowship program and the fellows, including their
+    demographic breakdown and domains of practice. This program came about
+    based on the institute's desire to integrate with the research community---thus
+    inspiring change from within---rather than preach to it. Five SSI fellows
+    attended WSSSPE4.
 

### 2017-02-15 WSSSPE4_report/wg_letters.tex
Address the open comments.

@@
 \footnote{The extensive use of such letters, and the problems that are
-  associated with it, may be an issue specific to the US.}
+  associated with it, may be an issue specific to the United States.}
 
@@
 \begin{itemize}
-\item The Research Computing Association has produced a ``Dear Colleague'' letter on the issue. \katznote{add citation}
-\item The National Science Foundation has produced a ``Dear Colleague'' letter that defines a ``product'' for the purposes of 2-page biosketches that does not include only publications, but also software.
-\katznote{add citation or link?}
-\item The ACM published an article in the CACM (one of the co-authors
-  was David Patterson) \katznote{why is Patterson significant here?} that explained that conference proceedings are
+\item The Computing Research Association has produced a ``Best
+  practices memo'' on the issue \cite{PSU99}. Among other points,
+  the memo also clarified that conference proceedings are
   \textit{the} venue in Computer Science, and this has been used to
   convince deans and provosts of the value of these proceedings
-  publications. This article may serve as a template for computational
+  publications. The article may serve as a template for computational
   science.
+\item The National Science Foundation has produced a number of ``Dear
+  Colleague'' letter that consider the issue of software and, taken
+  together, define a ``product'' for the purposes of
+  2-page biosketches in a way that does not include only publications, but also
+  software.%
+  \footnote{The most pertinent such letter is
+    \href{https://www.nsf.gov/pubs/2014/nsf14059/nsf14059.jsp}{NSF
+      14-059, ``Dear Colleague Letter - Supporting Scientific
+      Discovery through Norms and Practices for Software and Data
+      Citation and Attribution''}.}
+  This definition is ultimately codified in the National Science
+  Foundation's ``Proposal and Award Policies and Procedures Guide''
+  (often abbreviated as GPG) which in 2013 renamed the
+  ``Publications'' section of 2-page biosketches to ``Products'' (see
+  Chapter II.C.2.f(i)(c)) and now explicitly allows software to be
+  referenced in this section.
 \item There is also a 1995 report by the National Academies about
@@
   on how writing software should be considered in comparison to more
-  theoretical research. \katznote{add citation -- \url{https://www.nap.edu/catalog/2236/academic-careers-for-experimental-computer-scientists-and-engineers}}
+  theoretical research \cite{NRC-careers-94}. 
 \end{itemize}

### 2017-02-22 WSSSPE4_report/WSSSPE4_report.tex
Add my affiliation.
some more work on lightning talk descriptions... still not done

@@
     Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
-    Description
+    \note{Description}
 
@@
     peer review and onboarding research software}}~\cite{Ram:2016ws}.
-    Description
+    \note{Description}
     %Some people greatly encouraged to only submit to a journal code that has been peer reviewed
@@
     \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
-    Description
+    \note{Description}
     %“We (need to) decouple science from numerical methods and optimization” Hard open problem.
@@
     \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
-    Description
+    This talk described the timeline of activities of the FORCE11 Software Citation
+    Working Group and the principles of software citation established by the group.
+    Niemeyer, one of the working group cochairs, also discussed the paper published
+    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
+    working group before sunsetting it and spinning up a new Software Citation
+    Implementation Working Group.
 
@@
     Software}}~\cite{Idaszak:2016ws}.
-    Description
+    \note{Description}
 
@@
     Community}}~\cite{Hwang:2016ws}.
-    Description
+    \note{Description}
 
@@
     User-Producer Maturity Review}}~\cite{Goble:2016ws}.
-    Description
+    \note{Description}
 
@@
     repositories}}~\cite{Druskat:2016ws}.
-    Description
+    \note{Description}
 
@@
     lightweight specifications}}~\cite{Contrastin:2016ws}.
-    Description
+    \note{Description}
 
@@
     Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
-    Description
+    \note{Description}
 
@@
     into Scientific Software}}~\cite{Seidel:2016ws}.
-    Description
+    \note{Description}
 

### 2017-03-08 WSSSPE4_report/wg_letters.tex

@@
   software.%
-  \footnote{The most pertinent such letter is
-    \href{https://www.nsf.gov/pubs/2014/nsf14059/nsf14059.jsp}{NSF
+  \footnote{The most pertinent such letter is NSF
       14-059, ``Dear Colleague Letter - Supporting Scientific
       Discovery through Norms and Practices for Software and Data
-      Citation and Attribution''}.}
+      Citation and Attribution''~\cite{nsf-dcl-citation}.}
   This definition is ultimately codified in the National Science

### 2017-03-08 WSSSPE4_report/wg_access.tex
adding access subsection
initial draft of access subsection

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Coordinating access to CI for research software}
+\label{sec:access}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{%Mark Abraham to write this?  Needs confirmation or another volunteer - 
+Dan starting this for now - may poll other members of group to make progress}
+
+%Introduction to group here, including the overall objective of work in this area.
+Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code must acquire, setup, and maintain their own continuous integration systems.
+For example, software groups that develop BIOUNO, CI4SI, GROMACS) have done this.
+
+Note that Debian provides a testing infrastructure (\href{https://ci.debian.net/doc/}{http://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.
+
+\subsubsection{Participants}
+
+%members of the working group
+Xinlian Liu, Mark Abraham, Sameer Shende, James Hetherington, Dominik Kempf, Michael R. Crusoe, Radovan Bast
+
+\subsubsection{Working group objective}
+
+%Specific things the working group wants to accomplish in the context of the larger objective.
+This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems, by coordinating and sharing this burden across multiple projects.
+
+\subsubsection{Gap or challenge}
+
+%What is the gap or challenge being addressed?
+The scope of the group's interest is any type of testing, though interactive access for troubleshooting would be of particular interest beyond just automated testing.
+
+However, there are a lot of open issues:
+
+\begin{itemize}
+\item Some similar work has already been done.  How can this group apply and/or learn from that work, rather than reinventing the wheel?
+\item How to ensure this will work across disciplines?
+\item Meaningful CI for large projects may need hundreds of CPU hours per day
+\end{itemize}
+
+\subsubsection{Relevant people and resources}
+
+%What people, groups, or resources are needed.
+This would need to be picked up by a set of projects, potentially both small and large.
+
+Possibly working with the RSE community would be useful.
+
+\subsubsection{Plans}
+
+%What tasks will the working group undertake
+Some possible goals include:
+\begin{itemize}
+\item Acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to e.g. Debian's testing infrastructure or to a shared Jenkins-based infrastructure
+\item Extending Debian's scope to include published but not mature software
+\item Implementing the same interface but with specialized hardware or available to non-public codes.
+\item \url{https://reproducible-builds.org} but for CI 
+\end{itemize}
+
+\subsubsection{SMART steps}
+
+%What are the first SMART steps proposed?
+\begin{enumerate}
+\item Learn how donate machines to Debian
+\item Find funding, perhaps via NEIC? (\url{https://wiki.neic.no/wiki/NeIC_Community_Wiki}) 
+\item Build community
+\end{enumerate}
+
+\subsubsection{More information \& joining instructions}
+
+How could a reader get more information or get more involved?
+
+\todo{?}

### 2017-03-08 WSSSPE4_report/WSSSPE4_report.tex
removing old TODO
discuss google drive with WG notes and outputs

@@
 %
-\thanks{{}$^{(13)}$ \note{info needed}}
+\thanks{{}$^{(13)}$ Department of Mathematics, Colorado State
+  University, Fort Collins, CO 80523, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
 
@@
 Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall.
-\todo{find photos of flip charts, put in repo?} We asked them to work in Google Docs, \todo{find and cite these docs?} using a template for Vision, Gaps, and Projects.
+\todo{find photos of flip charts, put in repo?} We asked them to work in Google Docs, \todo{find and cite these docs?
+Is this really true, or was this all done in flip charts?} using a template for Vision, Gaps, and Projects.
 For those folk who worked on Vision and Gaps, after an hour, we encouraged them to finish their Vision or Gap Google Doc, and to find and join or start working on a project with a group.
@@
 \item Standard metadata for software (CodeMeta)
-\item White paper on developing sustainable software
+\item White paper on developing sustainable softwarez
 \item Social science for scientific software
@@
 
+The groups worked in Google docs, and eventually created Google slides to show their work back to the larger groups.
+The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}.
 In each group's discussion, a number of Specific, Measurable, Achievable, Relevant and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.
@@
 
-\todo{need to confirm who will write Access subsection (\S\ref{sec:access})}
+%\todo{need to confirm who will write Access subsection (\S\ref{sec:access})}
 

### 2017-03-08 WSSSPE4_report/WSSSPE4_report.tex
added a few more lightning talk descriptions

--- 2017-02-22 WSSSPE4_report/WSSSPE4_report.tex
@@
 %
-\begin{itemize}
+\begin{itemize}[itemsep=1ex]
     \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
@@
     Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
-    \note{Description}
+    In this talk, Gesing introduced the concept of science gateways: end-to-end software
+    solutions for a community that hide underlying complexity. She also posed the need
+    for gateways, which include the increased complexity of research and associated
+    hardware\slash software, a greater need for reproducibility and openness, and the
+    opportunity to integrate research with teaching for better workforce development.
+    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
+    and 1.4 times as many users accessed XSEDE via gateways compared with
+    login in March 2016. The Science Gateways Community Institute offers support
+    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
+    Software Collaborative, Community Engagement and Exchange, and Workforce
+    Development.
 
@@
     peer review and onboarding research software}}~\cite{Ram:2016ws}.
-    \note{Description}
-    %Some people greatly encouraged to only submit to a journal code that has been peer reviewed
+    This talk described the rOpenSci community's review process for research
+    software packages. Submission of a package starts with a pre-submission inquiry
+    for fit with the community criteria, then peer reviewers evaluate based on
+    usability, quality, and style. Upon acceptance, packages are given an rOpenSci
+    README badge and added to the system. Important components include a code of
+    conduct, open and non-adversarial review, and that no submission is
+    rejected---just out of scope. Remaining challenges include scaling, author
+    incentives, and automation of the process.
 
@@
     \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
-    \note{Description}
-    %“We (need to) decouple science from numerical methods and optimization” Hard open problem.
+    This talk began by describing the challenges facing science, and particularly
+    computational science, in recent years: the amount and complexity of knowledge
+    has increased dramatically, making it difficult for individuals or even teams to
+    manage modern, complex problems.
+    Brandt submitted an argument for decoupling science simulation and
+    the solution of computational problems away from numerical methods and low-level
+    optimization. This may be possible with domain-specific languages, or the use
+    of modern features in existing languages. This vision also includes reuse of
+    basic computing infrastructure, including common data structures, parameter files,
+    and high-performance I\slash O. This computing future necessarily requires
+    better recognition and reward systems for research software developers (i.e.,
+    research software engineers), and ways to make increasingly complex scientific
+    papers more understandable.
 

### 2017-03-08 WSSSPE4_report/WSSSPE4_report.tex
all but three lightning talks done now...

@@
     Software}}~\cite{Idaszak:2016ws}.
-    \note{Description}
+    HydroShare is a five-year-old online, collaborative system that supports
+    open sharing of hydrologic data, analytical tools, and computer models. The
+    project involves 10 collaborating institutions that produce tested and reviewed
+    community code every two to three weeks, following software engineering best
+    practices. Undergraduate and graduate students involved in the project not
+    only adopt the best practices exhibited by HydroShare, but see the open,
+    collaborative approach to research software development as the norm.
+    The talk made recommendations to other community efforts, including the need to
+    identify committed, hands-on PIs; evidence-based software engineering for science;
+    continuous integration; the importance of professional software development and
+    operations practices, even for domain scientists; and to identify technology
+    decisions that overlap with career paths.
 
@@
     Community}}~\cite{Hwang:2016ws}.
-    \note{Description}
+    In this talk, Hwang first described the ASPECT (Advance Solver for Problems in
+    Earth ConvecTion) open-source, parallel, extensible finite element code
+    used to simulate thermal convection. ASPECT began development in 2011 as a
+    community-driven project under the Computational Infrastructure for Geodynamics
+    (CIG). Community-building activities around ASPECT focused on the traditional
+    workshops, tutorials, and webinars, but also on slightly nontraditional,
+    multi-day hackathons---which turned out to be one of the most effective
+    forms of fostering a community. These start with leaders establishing project
+    conventions, demonstrating workflows and relevant technical skills, and
+    identifying starter projects. Daily activities involve work plans and
+    science\slash coding briefs, with mini-tutorials given as needed. One important
+    requirement is that each participant must submit at least one commit. Hwang
+    also summarized lessons learned, including that hackathons should last 7--10
+    days, include participants with multidisciplinary expertise and at both novice
+    and expert levels, have at least one dedicated code reviewer per 10 participants,
+    hold daily group updates, organize activities to promote group interactions,
+    and enforce free time.
 
     \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
-    User-Producer Maturity Review}}~\cite{Goble:2016ws}.
-    \note{Description}
+    User-Producer Reciprocity Review}}~\cite{Goble:2016ws}.
+    Goble first described some existing frameworks for measuring the maturity of
+    software, including the Software Sustainability Maturity Model, OSS Watch
+    Openness Rating, NASA Reuse Readiness Rating, Capability Maturity Model, and
+    Qualification and Selection of Open Source software. Then, she proposed a new,
+    simpler framework for project reciprocity, drawing on the intentions of users
+    and producers of the software, from the viewpoint of the producer.
+    Next steps include further definition of the model's characteristics based on
+    both theoretical and empirical investigation.
 
@@
     repositories}}~\cite{Druskat:2016ws}.
-    \note{Description}
+    In this talk, Druskat first identified two technical barriers to software
+    sustainability: identification of good software, and software discoverability.
+    Some of his perspectives were based on experiences in the humanities, a lack
+    of funding for computational infrastructure results in little experience with
+    sustainability (although this may differ in other domains). Druskat proposed
+    metadata repositories as a possible solution, as they can measure and document
+    technical sustainability, and document resulting metrics and more general
+    features of software. Technical sustainability goals include ensuring software
+    existence, preserving potential operation of software, and creating\slash
+    retaining possibilities for further development and maintenance of software.
+    The next steps of such a system include compiling measurable criteria for
+    the sustainability goals (based on qualitative concepts of, e.g.,
+    ``sustainability,'' ``maintainability,'' and ``usability''), and
+    crowd-sourcing a weighting of these criteria.
 
@@
     lightweight specifications}}~\cite{Contrastin:2016ws}.
-    \note{Description}
+    This talk introduced the tool CamFort for verifying Fortran code based on
+    lightweight specifications, which aid in software maintainability and reuse
+    through high-level indications of programmer intent.
+    Currently, CamFort supports specification (and
+    verification) of units and stencil operations. Plans for future specifications
+    include automatic test generation based on user-supplied properties, dependency
+    specifications, and software contracts to check (e.g.) expected value ranges
+    and program behavior.
 
@@
     Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
+    This talk described the activities of the weeklong Dagstuhl Perspectives Workshop
+    manifesto on ``Engineering Academic Software,'' which produced a manifesto
+    describing a pledge affirming the scholarly value of research software and
+    the personal responsibility to act in accordance with that value. The draft
+    (personal) manifesto includes 20 pledges, organized into three areas:
+    recognition of academic software, academic software development processes,
+    and the intellectual content of academic software. The talk concluded with a
+    survey for participants to give feedback on the pledges and their importance.
+
+    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
+    and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
+    into Scientific Software}}~\cite{Seidel:2016ws}.
+    This talk, given by a senior graduate student in Computer Science\slash
+    programming languages research, described several areas of software engineering
+    concepts that benefit the development of research software. In particular,
+    Seidel focused on software correctness, starting from high-level
+    specifications. Correctness can be demonstrated by verifying or proving
+    that the software satisfies the specification, testing against a range of
+    inputs, or synthesis of code from the specification. However, all of these
+    present technical challenges, in addition to the social challenges of
+    encouraging collaboration between software engineering and research software
+    communities.
+
+    \item \textbf{Iain Emsley\textsuperscript{\textasteriskcentered} and David
+    De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
+    community}}~\cite{Emsley:2016ws}.
+    In this talk, Emsley discussed efforts to develop the community around the
+    Drupal content management system, which is supported through the DrupalCon
+    series of conferences, user groups, and an association.
+    The talk described the use of social coding and software credit to develop
+    the Drupal software community, and social documentation to provide
+    documentation and tutorials. By analyzing identities of committers, they
+    found that the larger community is actually a loosely coupled set of
+    heterogeneous communities that use (and sustain themselves) through different
+    social mechanisms.
+
+    \item \textbf{Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
+    D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
+    and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
+    BLAS Routines}}~\cite{Dongarra:2016ws}.
     \note{Description}
 
-    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
-    and Gabrielle Allen: \emph{Bringing Techniques from Software Engineering
-    into Scientific Software}}~\cite{Seidel:2016ws}.
+    \item \textbf{Lukas Breitwieser\textsuperscript{\textasteriskcentered},
+    Roman Bauer, Alberto Di Meglio, Leonard Johard,
+    Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
+    Max Talanov: \textit{The BioDynaMo Project}}~\cite{Bauer:2016ws}.
+    \note{Description}
+
+    \item \textbf{Aseel Aldabjan, Robert Haines, and
+    Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
+    measure the relationship between code quality and software
+    sustainability?}}~\cite{Aldabjan:2016ws}.
     \note{Description}

### 2017-03-09 WSSSPE4_report/WSSSPE4_report.tex
adding comment in intro to lightning talks
Added section on papers and demos

@@
 %
-\thanks{{}$^{(3)}$ \note{info needed}}
+\thanks{{}$^{(3)}$ \note{Center for Research Computing \& Department of Computer Science and Engineering,
+University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}}
 %
@@
 4 idea papers,
-3 position paper,
-3 experience papers,
+3 position papers,
+2 experience papers,
 and
@@
 
-Intro and description
-%
+The 12 excellent submissions resulting in 10-min talks (four idea papers, three position papers, two experience papers and three demos) address a wide range of topics around sustainability for software in science.
+They cover three main areas:
+\begin{enumerate}
+\item The specifics of the academic environment for software development and for so-called research programmers or research software engineers;
+\item The characteristics and needs of research software in general such as  the lifecycle of software in science, preservation, curation, sharing and reproducibility challenges;
+\item Lessons learned from or visions for use cases of scientific software and communities working with and/or on a software package.
+\end{enumerate}
+
 \begin{itemize}
-\item \textbf{Authors: \emph{title}}.
-Description
+\item \textbf{Author: Michael Heroux \emph{Idea paper: Sustainable & Productive: Improving Incentives for Quality Software}}.
+The author presents that improving quality of scientific software is a measure  to address the increased complexity in both modeling and software.
+To support quality, it is essential to provide Computational Science and Engineering teams with publication venues, funding, and  professional recognition.
+Broad improvements in sustainability and productivity can be only achieved when publishers, funding  agencies  and  employers  raise  their  expectations for software quality.
+Additionally, software community leaders are in the unique position to positively impact software quality by working on establishing incentives that will spur creative and novel  approaches to improve developer productivity  and software sustainability.
+
+\item \textbf{Author: Gabrielle Allen \emph{Idea Paper: Establishing a Professional Society for Research Software}}.
+This paper advocates to create a professional society or extend the scope of an existing organization in research software.
+The author presents the motivation, a variety of potential roles of such a society and diverse potential funding mechanisms to be able to establish a sustainable organization.
+She gives an overview on similar activities as well as existing organizations and aims at stimulating discussion among potential members and stakeholders.
+
+\item \textbf{Authors: Olivier Philippe, Simon Hettrick and Neil Chue Hong \emph{Experience Paper: Preliminary analysis of a survey of UK Research Software Engineers}}.
+This paper presents results from a survey conducted on the new role of research software engineers in academia.
+The  survey  provides   demographic  information  about  the  education,  field,  gender,  job  satisfaction  and career  plans  of  the  people  in this community.
+the outcome is that members of the  community  is  found  to be highly educated, derive mainly from the hard sciences, and to be predominantly male.
+Respondents report satisfaction in their jobs,  but  indicate  that  career  progression  is  both  difficult  and opaque.
+The authors urge a  continued  discussion  about  the  experience  of research software engineers and  recommend further  investigation  into  this
+important  community.
+
+\item \textbf{Author: Christopher Gwilliams \emph{Demo: Using Industrial Engagement to Create and Develop Research Ties within Academia}}.
+This demo focuses on the benefits of including a  new  software engineering degree that is influenced by, and works directly with,
+industry.  Cardiff  University  introduced  a  new  industry-focused  degree, BSc Applied Software Engineering (ASE) in September
+2015. The engagement with industry allows students to learn through live
+projects  with  real-world  applications and forge links for career opportunities.
+Also industry benefits from this course that provides an entry point for industry to engage with the university and to develop a network of contacts between industry and academia, which is also a benefit for the university and allows to  identify  potential  opportunities for new research projects and collaborators.
+
+\item \textbf{Authors: Anshu Dubey and Katherine Riley \emph{Experience Paper: Software Engineering and Community Codes Track in ATPESC}}.
+The Argonne Training Program in Extreme Scale
+Computing (ATPESC) was started by Argonne National Laboratory with the objective of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
+One of the unique aspects of the program was inclusion of a
+software engineering and community codes track. The inclusion
+was motivated by the observation that the projects with a good
+software process were better able to meet their scientific goals.
+The paper presents the experience in running the software
+track, the motivation, reception, and evolution of the track over the years.  
+The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.
+
+\item \textbf{Author: Neil Chue Hong \emph{Position Paper: Why do we need to compare research software, and how should we do it?}}.
+This position paper sets out the reasons for why different  stakeholders, from users to developers 
+to funders, might wish to undertake the difficult task to compare research software.
+The author describes a proposed framework for doing so (based around measures of accessibility, usability, maintainability and portability) which takes into account the possibility of variation between different communities about how they prioritise different aspects of research software.
+
+\item \textbf{Authors: Anshu Dubey and Lois Curfman McInnes \emph{Idea Paper: The Lifecycle of Software for Scientific Simulation}}.
+The authors discuss the specifics of the lifecycle of scientific software. While the software lifecycle is a well-researched topic, one class of projects, research software, has received relatively
+little attention from lifecycle researchers. This
+paper formalizes a lifecycle model
+for end-to-end scientific application software, featuring loose
+coupling betweensubmodels for development of infrastructure
+and scientific capability. The authors also invite input from stakeholders
+to converge on a model that captures the complexity of this
+development processes and provides needed lifecycle guidance to
+the scientific software community.
+
+\item \textbf{Authors: Francisco Queiroz and Rejane Spitz \emph{Position Paper: Collaborative Gamification Design for Scientific Software}}. 
+The paper focuses on usability design and presents that gamification, a design trend for improving scientific software usability, for scientific software should be facilitated by an open, collaborative design process supported by
+conversational media. This approach is compatible with 
+qualities often attributed to computational science community 
+regarding openness and collaboration between members of varied 
+professional backgrounds.
+They exemplify the use of conversational media for collaborative design and expect the synergy between collaborators to result in better usability,  greater user acceptance, and adequacy to requirements,
+obtaining optimal design solutions in a sustainable way.
+
+\item \textbf{Authors: Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous and David Wilkinson \emph{Position Paper: Active Curation of Artifacts is Changing the Way Digital Libraries will Operate}}.
+The paper presents the implications of “active curation systems” that provide executable access to
+artifacts and experiments behind published results and enables
+their reuse. These systems allow changing and repeating experiments to understand how an innovation behaves in conditions
+beyond the ones described in a paper and address reproducibility challenges. Active curation systems
+also  enable accountability and accelerate research progress  by
+giving access to complete experimental details. As these systems
+take  hold,  it  is  important  to  understand  their  capabilities  and
+how digital libraries should be integrated with them. The authors describe a study underway to explore how best
+to do this integration.
+
+\item \textbf{Authors: Debashis Ganguly, William C. Garrison III, David Wilkinson, Bruce R. Childers, Adam J. Lee, and Daniel Mosse \emph{Demo: Composing, Reproducing, and Sharing Simulations}}.
+The authors demonstrate their approach of sharing,
+reproducing, and composing simulations toward accelerating
+research productivity while also improving accountability and
+credibility. Specifically, they have developed a case study in
+which they compose and share access control simulations in the
+form of shareable data store units for cloud systems. They share, compose
+and repeat simulations  through a collaborative repository
+(OCCAM) and suggest a general simulation framework (SST) that can
+accelerate the efforts for the whole community.
+
+\item \textbf{Authors: Sameer Shende and Allen Malony. \emph{Demo: Using TAU for Performance Evaluation of Scientific Software}}.
+This demo presents the  TAU Performance System for performance evaluation of Scientific 
+Software written in C++, C, and Fortran. The system addresses performance 
+technology problems  at three levels: instrumentation, 
+measurement, and analysis. The TAU framework supports the 
+configuration and integration of these layers. The authors advocate that effective 
+exploration of performance will necessarily
+require prudent selection from the range of alternative methods TAU provides 
+to  assemble meaningful performance experiments.
+
+\item \textbf{Authors: Janos Sallai, Christopher Iacovella, Christoph Klein and Tengyu Ma.
+\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield Atom-Typing for Molecular Simulation}}.
+
+The authors present an essential and error-prone task in the area of Molecular Dynamics simulations: the efficient selection of correct parameter values for specific molecules, especially forcefields.
+They  present  a  framework  that  aims  to  solve  this
+data  management  issue,  proposing  a  common  format  for
+forcefields that is self-documenting with machine readable,
+declarative usage rules. They investigate processes and tools
+that are commonly used today in software development (e.g,
+unit testing, verification and validation, continuous integra-
+tion,  and  version  control)  and reason that they are, with proper infrastructure
+support, applicable to forcefield development, as well. The
+paper describes how such an infrastructure can tackle man-
+aging and evolving forcefields by the MD community, and
+proposes a way to encourage and incentivize involvement
+by the stakeholders.
 
@@
 
-Intro and description.
+There were 18 \katznote{this is what I found on EasyChair and wrote in \S2, but I think there are 16 listed below.  I wonder what the difference is?} lightning talks presented at WSSSPE4, as follows.
+%
 A \textsuperscript{\textasteriskcentered} indicates which author presented the

### 2017-03-09 WSSSPE4_report/WSSSPE4_report.tex
small fixes to new papers section and a todo

@@
 
+\todo{group by type - idea, position, experience, demo}
 \begin{itemize}
-\item \textbf{Author: Michael Heroux \emph{Idea paper: Sustainable & Productive: Improving Incentives for Quality Software}}.
+\item \textbf{Michael Heroux \emph{Idea paper: Sustainable \& Productive: Improving Incentives for Quality Software}}.
 The author presents that improving quality of scientific software is a measure  to address the increased complexity in both modeling and software.
@@
 
-\item \textbf{Author: Gabrielle Allen \emph{Idea Paper: Establishing a Professional Society for Research Software}}.
+\item \textbf{Gabrielle Allen \emph{Idea Paper: Establishing a Professional Society for Research Software}}.
 This paper advocates to create a professional society or extend the scope of an existing organization in research software.
@@
 
-\item \textbf{Authors: Olivier Philippe, Simon Hettrick and Neil Chue Hong \emph{Experience Paper: Preliminary analysis of a survey of UK Research Software Engineers}}.
+\item \textbf{Olivier Philippe, Simon Hettrick and Neil Chue Hong \emph{Experience Paper: Preliminary analysis of a survey of UK Research Software Engineers}}.
 This paper presents results from a survey conducted on the new role of research software engineers in academia.
@@
 
-\item \textbf{Author: Christopher Gwilliams \emph{Demo: Using Industrial Engagement to Create and Develop Research Ties within Academia}}.
+\item \textbf{Christopher Gwilliams \emph{Demo: Using Industrial Engagement to Create and Develop Research Ties within Academia}}.
 This demo focuses on the benefits of including a  new  software engineering degree that is influenced by, and works directly with,
-industry.  Cardiff  University  introduced  a  new  industry-focused  degree, BSc Applied Software Engineering (ASE) in September
+industry.  Cardiff  University  introduced  a  new  industry-focused  degree, Bachelor of Science in Applied Software Engineering (ASE) in September
 2015. The engagement with industry allows students to learn through live
@@
 
-\item \textbf{Authors: Anshu Dubey and Katherine Riley \emph{Experience Paper: Software Engineering and Community Codes Track in ATPESC}}.
+\item \textbf{Anshu Dubey and Katherine Riley \emph{Experience Paper: Software Engineering and Community Codes Track in ATPESC}}.
 The Argonne Training Program in Extreme Scale
@@
 
-\item \textbf{Author: Neil Chue Hong \emph{Position Paper: Why do we need to compare research software, and how should we do it?}}.
+\item \textbf{Neil Chue Hong \emph{Position Paper: Why do we need to compare research software, and how should we do it?}}.
 This position paper sets out the reasons for why different  stakeholders, from users to developers 
 to funders, might wish to undertake the difficult task to compare research software.
-The author describes a proposed framework for doing so (based around measures of accessibility, usability, maintainability and portability) which takes into account the possibility of variation between different communities about how they prioritise different aspects of research software.
+The author describes a proposed framework for doing so (based around measures of accessibility, usability, maintainability and portability) which takes into account the possibility of variation between different communities about how they prioritize different aspects of research software.
 
-\item \textbf{Authors: Anshu Dubey and Lois Curfman McInnes \emph{Idea Paper: The Lifecycle of Software for Scientific Simulation}}.
+\item \textbf{Anshu Dubey and Lois Curfman McInnes \emph{Idea Paper: The Lifecycle of Software for Scientific Simulation}}.
 The authors discuss the specifics of the lifecycle of scientific software. While the software lifecycle is a well-researched topic, one class of projects, research software, has received relatively
@@
 for end-to-end scientific application software, featuring loose
-coupling betweensubmodels for development of infrastructure
+coupling between submodels for development of infrastructure
 and scientific capability. The authors also invite input from stakeholders
@@
 
-\item \textbf{Authors: Francisco Queiroz and Rejane Spitz \emph{Position Paper: Collaborative Gamification Design for Scientific Software}}. 
+\item \textbf{Francisco Queiroz and Rejane Spitz \emph{Position Paper: Collaborative Gamification Design for Scientific Software}}. 
 The paper focuses on usability design and presents that gamification, a design trend for improving scientific software usability, for scientific software should be facilitated by an open, collaborative design process supported by
@@
 
-\item \textbf{Authors: Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous and David Wilkinson \emph{Position Paper: Active Curation of Artifacts is Changing the Way Digital Libraries will Operate}}.
-The paper presents the implications of “active curation systems” that provide executable access to
+\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous and David Wilkinson \emph{Position Paper: Active Curation of Artifacts is Changing the Way Digital Libraries will Operate}}.
+The paper presents the implications of ``active curation systems'' that provide executable access to
 artifacts and experiments behind published results and enables
@@
 
-\item \textbf{Authors: Debashis Ganguly, William C. Garrison III, David Wilkinson, Bruce R. Childers, Adam J. Lee, and Daniel Mosse \emph{Demo: Composing, Reproducing, and Sharing Simulations}}.
+\item \textbf{Debashis Ganguly, William C. Garrison III, David Wilkinson, Bruce R. Childers, Adam J. Lee, and Daniel Mosse \emph{Demo: Composing, Reproducing, and Sharing Simulations}}.
 The authors demonstrate their approach of sharing,
@@
 
-\item \textbf{Authors: Sameer Shende and Allen Malony. \emph{Demo: Using TAU for Performance Evaluation of Scientific Software}}.
+\item \textbf{Sameer Shende and Allen Malony. \emph{Demo: Using TAU for Performance Evaluation of Scientific Software}}.
 This demo presents the  TAU Performance System for performance evaluation of Scientific 
@@
 
-\item \textbf{Authors: Janos Sallai, Christopher Iacovella, Christoph Klein and Tengyu Ma.
+\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein and Tengyu Ma.
 \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield Atom-Typing for Molecular Simulation}}.
@@
 that are commonly used today in software development (e.g,
-unit testing, verification and validation, continuous integra-
-tion,  and  version  control)  and reason that they are, with proper infrastructure
+unit testing, verification and validation, continuous integration,
+and  version  control)  and reason that they are, with proper infrastructure
 support, applicable to forcefield development, as well. The

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
a few edits to paper section

@@
 
-The 12 excellent submissions resulting in 10-min talks (four idea papers, three position papers, two experience papers and three demos) address a wide range of topics around sustainability for software in science.
-They cover three main areas:
+The 12 excellent submissions resulted in 10-min talks (four idea papers, three position papers,
+two experience papers, and three demos) that addressed a wide range of topics around
+sustainability for software in science. They covered three main areas:
 \begin{enumerate}
-\item The specifics of the academic environment for software development and for so-called research programmers or research software engineers;
-\item The characteristics and needs of research software in general such as  the lifecycle of software in science, preservation, curation, sharing and reproducibility challenges;
-\item Lessons learned from or visions for use cases of scientific software and communities working with and/or on a software package.
+\item The specifics of the academic environment for software development and for so-called
+research programmers or research software engineers;
+\item The characteristics and needs of research software in general such as the lifecycle of
+software in science, preservation, curation, sharing, and reproducibility challenges; and
+\item Lessons learned from or visions for use cases of scientific software and communities
+working with and\slash or on a software package.
 \end{enumerate}

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
added refs for papers, and formatting fixes

@@
 \begin{itemize}
-\item \textbf{Michael Heroux \emph{Idea paper: Sustainable \& Productive: Improving Incentives for Quality Software}}.
-The author presents that improving quality of scientific software is a measure  to address the increased complexity in both modeling and software.
-To support quality, it is essential to provide Computational Science and Engineering teams with publication venues, funding, and  professional recognition.
-Broad improvements in sustainability and productivity can be only achieved when publishers, funding  agencies  and  employers  raise  their  expectations for software quality.
-Additionally, software community leaders are in the unique position to positively impact software quality by working on establishing incentives that will spur creative and novel  approaches to improve developer productivity  and software sustainability.
+\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
+Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
+The author presents that improving quality of scientific software is a measure
+to address the increased complexity in both modeling and software. To support
+quality, it is essential to provide Computational Science and Engineering teams
+with publication venues, funding, and  professional recognition. Broad
+improvements in sustainability and productivity can be only achieved when
+publishers, funding agencies and employers raise their expectations for
+software quality. Additionally, software community leaders are in the unique
+position to positively impact software quality by working on establishing
+incentives that will spur creative and novel approaches to improve developer
+productivity  and software sustainability.
 
-\item \textbf{Gabrielle Allen \emph{Idea Paper: Establishing a Professional Society for Research Software}}.
-This paper advocates to create a professional society or extend the scope of an existing organization in research software.
-The author presents the motivation, a variety of potential roles of such a society and diverse potential funding mechanisms to be able to establish a sustainable organization.
-She gives an overview on similar activities as well as existing organizations and aims at stimulating discussion among potential members and stakeholders.
+\item \textbf{Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
+Society for Research Software}}~\cite{GAllen:2016ws}.
+This paper advocates to create a professional society or extend the scope of an
+existing organization in research software. The author presents the motivation,
+a variety of potential roles of such a society and diverse potential funding
+mechanisms to be able to establish a sustainable organization. She gives an
+overview on similar activities as well as existing organizations and aims at
+stimulating discussion among potential members and stakeholders.
 
-\item \textbf{Olivier Philippe, Simon Hettrick and Neil Chue Hong \emph{Experience Paper: Preliminary analysis of a survey of UK Research Software Engineers}}.
-This paper presents results from a survey conducted on the new role of research software engineers in academia.
-The  survey  provides   demographic  information  about  the  education,  field,  gender,  job  satisfaction  and career  plans  of  the  people  in this community.
-the outcome is that members of the  community  is  found  to be highly educated, derive mainly from the hard sciences, and to be predominantly male.
-Respondents report satisfaction in their jobs,  but  indicate  that  career  progression  is  both  difficult  and opaque.
-The authors urge a  continued  discussion  about  the  experience  of research software engineers and  recommend further  investigation  into  this
+\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
+\emph{Experience Paper: Preliminary analysis of a survey of UK Research
+Software Engineers}}~\cite{Philippe:2016ws}.
+This paper presents results from a survey conducted on the new role of research
+software engineers in academia. The survey provides demographic information
+about the education, field, gender, job satisfaction and career plans of the people
+in this community. The outcome is that members of the community  were found to be
+highly educated, derive mainly from the hard sciences, and are predominantly male.
+Respondents reported satisfaction in their jobs, but indicated that career progression
+is both difficult and opaque. The authors urge a continued discussion about the experience
+of research software engineers and recommend further investigation into this
 important community.
 
-\item \textbf{Christopher Gwilliams \emph{Demo: Using Industrial Engagement to Create and Develop Research Ties within Academia}}.
-This demo focuses on the benefits of including a  new  software engineering degree that is influenced by, and works directly with,
-industry.  Cardiff  University  introduced  a  new  industry-focused  degree, Bachelor of Science in Applied Software Engineering (ASE) in September
+\item \textbf{Christopher Gwilliams: \emph{Demo: Using industrial engagement to
+create and develop research ties within academia}}~\cite{Gwilliams:2016ws}.
+This demo focuses on the benefits of including a  new  software engineering
+degree that is influenced by, and works directly with,
+industry.  Cardiff  University  introduced  a  new  industry-focused  degree,
+Bachelor of Science in Applied Software Engineering (ASE) in September
 2015. The engagement with industry allows students to learn through live
 projects  with  real-world  applications and forge links for career opportunities.
-Also industry benefits from this course that provides an entry point for industry to engage with the university and to develop a network of contacts between industry and academia, which is also a benefit for the university and allows to  identify  potential  opportunities for new research projects and collaborators.
+Also, industry benefits from this course that provides an entry point for industry
+to engage with the university and to develop a network of contacts between industry
+and academia, which is also a benefit for the university and allows to identify
+potential opportunities for new research projects and collaborators.
 
-\item \textbf{Anshu Dubey and Katherine Riley \emph{Experience Paper: Software Engineering and Community Codes Track in ATPESC}}.
+\item \textbf{Anshu Dubey and Katherine Riley: \emph{Experience Paper: Software
+Engineering and Community Codes Track in ATPESC}}~\cite{Dubey:2016ws}.
 The Argonne Training Program in Extreme Scale
-Computing (ATPESC) was started by Argonne National Laboratory with the objective of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
+Computing (ATPESC) was started by Argonne National Laboratory with the objective
+of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
 One of the unique aspects of the program was inclusion of a
@@
 
-\item \textbf{Neil Chue Hong \emph{Position Paper: Why do we need to compare research software, and how should we do it?}}.
-This position paper sets out the reasons for why different  stakeholders, from users to developers
-to funders, might wish to undertake the difficult task to compare research software.
-The author describes a proposed framework for doing so (based around measures of accessibility, usability, maintainability and portability) which takes into account the possibility of variation between different communities about how they prioritize different aspects of research software.
+\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
+research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
+This position paper sets out the reasons for why different stakeholders,
+from users to developers to funders, might wish to undertake the difficult task
+to compare research software. The author describes a proposed framework for doing
+so (based around measures of accessibility, usability, maintainability and
+portability) which takes into account the possibility of variation between
+different communities about how they prioritize different aspects of research software.
 
-\item \textbf{Anshu Dubey and Lois Curfman McInnes \emph{Idea Paper: The Lifecycle of Software for Scientific Simulation}}.
-The authors discuss the specifics of the lifecycle of scientific software. While the software lifecycle is a well-researched topic, one class of projects, research software, has received relatively
+\item \textbf{Anshu Dubey and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
+of Software for Scientific Simulation}}~\cite{Dubey:2016ws}.
+The authors discuss the specifics of the lifecycle of scientific software.
+While the software lifecycle is a well-researched topic, one class of projects,
+research software, has received relatively
 little attention from lifecycle researchers. This
@@
 
-\item \textbf{Francisco Queiroz and Rejane Spitz \emph{Position Paper: Collaborative Gamification Design for Scientific Software}}.
-The paper focuses on usability design and presents that gamification, a design trend for improving scientific software usability, for scientific software should be facilitated by an open, collaborative design process supported by
+\item \textbf{Francisco Queiroz and Rejane Spitz: \emph{Position Paper:
+Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
+The paper focuses on usability design and presents that gamification, a design
+trend for improving scientific software usability, for scientific software should
+be facilitated by an open, collaborative design process supported by
 conversational media. This approach is compatible with
@@
 regarding openness and collaboration between members of varied
-professional backgrounds.
-They exemplify the use of conversational media for collaborative design and expect the synergy between collaborators to result in better usability,  greater user acceptance, and adequacy to requirements,
+professional backgrounds. They exemplify the use of conversational media for
+collaborative design and expect the synergy between collaborators to result in
+better usability, greater user acceptance, and adequacy to requirements,
 obtaining optimal design solutions in a sustainable way.
 
-\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous and David Wilkinson \emph{Position Paper: Active Curation of Artifacts is Changing the Way Digital Libraries will Operate}}.
-The paper presents the implications of ``active curation systems'' that provide executable access to
+\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous, and
+David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
+the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
+The paper presents the implications of ``active curation systems'' that provide
+executable access to
 artifacts and experiments behind published results and enables
-their reuse. These systems allow changing and repeating experiments to understand how an innovation behaves in conditions
-beyond the ones described in a paper and address reproducibility challenges. Active curation systems
+their reuse. These systems allow changing and repeating experiments to understand
+how an innovation behaves in conditions beyond the ones described in a paper and
+address reproducibility challenges. Active curation systems
 also  enable accountability and accelerate research progress  by
@@
 take  hold,  it  is  important  to  understand  their  capabilities  and
-how digital libraries should be integrated with them. The authors describe a study underway to explore how best
+how digital libraries should be integrated with them. The authors describe a
+study underway to explore how best
 to do this integration.
 
-\item \textbf{Debashis Ganguly, William C. Garrison III, David Wilkinson, Bruce R. Childers, Adam J. Lee, and Daniel Mosse \emph{Demo: Composing, Reproducing, and Sharing Simulations}}.
+\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
+Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse: \emph{Demo: Composing,
+Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
 The authors demonstrate their approach of sharing,
@@
 
-\item \textbf{Sameer Shende and Allen Malony. \emph{Demo: Using TAU for Performance Evaluation of Scientific Software}}.
+\item \textbf{Sameer Shende and Allen Malony: \emph{Demo: Using TAU for
+Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
 This demo presents the  TAU Performance System for performance evaluation of Scientific
@@
 
-\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein and Tengyu Ma.
-\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield Atom-Typing for Molecular Simulation}}.
-
-The authors present an essential and error-prone task in the area of Molecular Dynamics simulations: the efficient selection of correct parameter values for specific molecules, especially forcefields.
-They  present  a  framework  that  aims  to  solve  this
-data  management  issue,  proposing  a  common  format  for
+\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
+\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
+Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
+The authors present an essential and error-prone task in the area of Molecular
+Dynamics simulations: the efficient selection of correct parameter values for
+specific molecules, especially forcefields. They present a framework that
+aims to solve this data management issue, proposing a common format for
 forcefields that is self-documenting with machine readable,
 declarative usage rules. They investigate processes and tools
-that are commonly used today in software development (e.g,
+that are commonly used today in software development (e.g.,
 unit testing, verification and validation, continuous integration,
@@
 support, applicable to forcefield development, as well. The
-paper describes how such an infrastructure can tackle man-
-aging and evolving forcefields by the MD community, and
+paper describes how such an infrastructure can tackle managing and evolving
+forcefields by the molecular dynamics community, and
 proposes a way to encourage and incentivize involvement

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
added all but two lightning talks

@@
 
+    \item \textbf{Neil Chue Hong\textsuperscript{\textasteriskcentered}:
+    \textit{Making it easier to understand research software impact}}~\cite{ChueHong:2016wsb}.
+    Chue Hong proposed that the WSSSPE community collaboratively develop a framework
+    to better understand (and measure the impact of) research software. This could
+    be based on a workflow involving mining the literature for mentions of software,
+    identifying linkages between software, and visualizing dependencies between
+    software or between software and other research artifacts. Many tools already
+    exist in this ecosystem, so Chue Hong proposed that a WSSSPE working group
+    collect these tools, summarize use cases, identify missing pieces, and find
+    low-hanging fruit to integrate tools or build new ones.
+
+    \item \textbf{Willem Robert van Hage\textsuperscript{\textasteriskcentered},
+    Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
+    Netherlands eScience Center}}~\cite{vanHage:2016ws}.
+    \note{Description}
+
+    \item \textbf{Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
+    Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
+    Research Software}}~\cite{GAllen:2016wsb}.
+    \note{Description}
+
     \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
@@
     BLAS Routines}}~\cite{Dongarra:2016ws}.
-    \note{Description}
+    This talk described Batched BLAS (BBLAS), a new standard for computing multiple
+    BLAS (Basic Linear Algebra Subproblems) in parallel. Unlike existing parallel
+    BLAS algorithms provided by, e.g., Intel MKL, cuBLAS, or MAGMA, BBLAS is
+    optimized for smaller matrices, which apply to a variety of problems in areas
+    such as machine learning, multifrontal solvers, image processing, fluid
+    dynamics, and astrophysics. BBLAS offers nearly a factor of two improvement
+    in computing speed compared to non-batched operations. To develop a sustainable
+    BBLAS library, the developers have focused on proposing a standard API,
+    holding workshops between academics and vendors to reach consensus, and
+    performing research into API design decisions on performance.
 
@@
     Max Talanov: \textit{The BioDynaMo Project}}~\cite{Bauer:2016ws}.
-    \note{Description}
+    Breitwieser described the BioDynaMo project, a large-scale platform for
+    biological simulations that provides access to computational resources while
+    hiding the complexity of distributed computing. It also promotes computational
+    reproducibility through shared open-access data. The project began as a
+    modernization initiative porting an existing simulation software, Cx3D, from
+    Java to C++, but also incorporated many modern software engineering practices
+    such as hosting and version control on GitHub, automated tests and continuous
+    integration via Travis-CI, code coverage reports, and an automated build system.
+    In addition, over 1000 students participated in development via the Intel
+    Modern Code Developer Challenge competition---the winner achieved a
+    performance speedup of 320 times. The project plans to continue improving
+    the development process with (e.g.) static code checkers and Git commit hooks,
+    continued architecture redesign, and verification.
 
@@
     sustainability?}}~\cite{Aldabjan:2016ws}.
-    \note{Description}
+    This talk explored the relationship between code quality and the sustainability
+    of research software; however, the latter has no concrete definition, thus making
+    it challenging to provide guidance on how to achieve sustainability. The authors
+    proposed a ``software sustainment'' metric based on the time difference between
+    the most recent and initial commits\slash contributions to the codebase.
+    Using this metric, they analyzed 3113 Java projects hosted on GitHub, created
+    in 2009, and found that 22\% and 35\% had sustainment metrics of 0 and less
+    than seven, respectively. Interestingly, the authors also found high
+    code quality (measured using a variety of static analytic metrics) across
+    all these projects, with little correlation to sustainability.
 

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
fixing name collision

@@
 \item \textbf{Anshu Dubey and Katherine Riley: \emph{Experience Paper: Software
-Engineering and Community Codes Track in ATPESC}}~\cite{Dubey:2016ws}.
+Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
 The Argonne Training Program in Extreme Scale
@@
 \item \textbf{Anshu Dubey and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
-of Software for Scientific Simulation}}~\cite{Dubey:2016ws}.
+of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
 The authors discuss the specifics of the lifecycle of scientific software.

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
fixing numbers of submission and lightning talks

@@
 Submissions to WSSSPE4 comprised
-20 lightning talks,
+18 lightning talks,
 4 idea papers,
@@
 Specifically,
-18 lightning talks,
+16 lightning talks,
 4 idea papers,
@@
 
-The 12 excellent submissions resulted in 10-min talks (four idea papers, three position papers,
+WSSSPE included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
 two experience papers, and three demos) that addressed a wide range of topics around
@@
 
-There were 18 \katznote{this is what I found on EasyChair and wrote in \S2, but I think there are 16 listed below.  I wonder what the difference is?} lightning talks presented at WSSSPE4, as follows.
+There were 16 lightning talks presented at WSSSPE4, as follows.
 %

### 2017-03-10 WSSSPE4_report/WSSSPE4_report.tex
adding link for flipchart photos

@@
 
-Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall.
-\todo{find photos of flip charts, put in repo?} We asked them to work in Google Docs, \todo{find and cite these docs?
+Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}.
+We asked them to work in Google Docs, \todo{find and cite these docs?
 Is this really true, or was this all done in flip charts?} using a template for Vision, Gaps, and Projects.

### 2017-03-14 WSSSPE4_report/wg_access.tex
clarify & links to Google Group

@@
 %Introduction to group here, including the overall objective of work in this area.
-Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code must acquire, setup, and maintain their own continuous integration systems.
+Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
 For example, software groups that develop BIOUNO, CI4SI, GROMACS) have done this.
@@
 %Specific things the working group wants to accomplish in the context of the larger objective.
-This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems, by coordinating and sharing this burden across multiple projects.
+This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems (when publicly available CI are not a fit), by coordinating and sharing this burden across multiple projects.
 
@@
 
-How could a reader get more information or get more involved?
+https://groups.google.com/forum/#!forum/continuous-integration-for-research-software
 

### 2017-03-15 WSSSPE4_report/WSSSPE4_report.tex
more accurate description of projects process
added description of NLeSC lightning talk

@@
     Netherlands eScience Center}}~\cite{vanHage:2016ws}.
-    \note{Description}
+    This talk described eStep (eScience Technology Platform) at the Netherlands
+    eScience Center, which includes three components: a software catalogue
+    of eScience software, interfaces, libraries,
+    tools workflows, applications, etc.; a knowledge base of guides, reports, and
+    recommendations for scientific software development; and eScience research
+    documents including such as scientific publications, book chapters, and demos.
+    The eStep website currently presents automatically-generated information about
+    software based on metadata, with a plan for displaying measures of impact and
+    quality.
 
@@
 Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}.
-We asked them to work in Google Docs, \todo{find and cite these docs?
-Is this really true, or was this all done in flip charts?} using a template for Vision, Gaps, and Projects.
 For those folk who worked on Vision and Gaps, after an hour, we encouraged them to finish their Vision or Gap Google Doc, and to find and join or start working on a project with a group.
@@
 \item Standard metadata for software (CodeMeta)
-\item White paper on developing sustainable softwarez
+\item White paper on developing sustainable software
 \item Social science for scientific software

### 2017-03-15 WSSSPE4_report/WSSSPE4_report.tex
adding links to diverse orgs

@@
 diversity of applicants by forwarding the notice of travel support to
-organizations including PyLadies, Django Girls, Women Who Code, and
-Women in HPC.\todo{add links}
+organizations including PyLadies\footnote{\url{http://www.pyladies.com}},
+Django Girls\footnote{\url{https://djangogirls.org}},
+Women Who Code\footnote{\url{https://www.womenwhocode.com}}, and
+Women in HPC\footnote{\url{http://www.womeninhpc.org}}.
 \todo{did we reach out to any others?}

### 2017-03-16 WSSSPE4_report/WSSSPE4_report.tex
added description of metrics lightning talk

@@
     Research Software}}~\cite{GAllen:2016wsb}.
-    \note{Description}
+    In this talk, Chen first summarized some of the outcomes of the WSSSPE3
+    Metrics Breakout Group, who discussed that metrics are important for scientific
+    impact, discovery, reducing duplication of efforts, and evaluating software.
+    The group also summarized barriers to widespread use of metrics: no common
+    set of metrics collected, a need for usefulness to be obvious for projects,
+    and no common standard for collecting\slash presenting metrics. To find out
+    more about practices related to metrics, the authors ran a survey of National
+    Science Foundation Software Infrastructure for Sustained Innovation (SI2)
+    program investigators, with 19 participants. Important results include the
+    observation that most use citations and acknowledgements of software in
+    funding proposals, while download numbers are both challenging to collect
+    and not useful. Overall, participants generally agreed that metrics are
+    important, but said useful metrics are difficult or tedious to collect.
 

### 2017-03-16 WSSSPE4_report/WSSSPE4_report.tex
updated number of lightning talks (19)

@@
 Specifically,
-16 lightning talks,
+19 lightning talks,
 4 idea papers,
@@
 
-There were 16 lightning talks presented at WSSSPE4, as follows.
+There were 19 lightning talks presented at WSSSPE4, as follows.
 %

### 2017-03-16 WSSSPE4_report/WSSSPE4_report.tex
... updated lightning talks again (19 submissions, if 19 presented)
lightning talk section done!

@@
 Submissions to WSSSPE4 comprised
-18 lightning talks,
+19 lightning talks,
 4 idea papers,
@@
 
-\note{Kyle Niemeyer will write this}
-
-There were 19 lightning talks presented at WSSSPE4, as follows.
+There were 19 lightning talks presented at WSSSPE4, as follows (in order of
+presentation).
 %

### 2017-03-17 WSSSPE4_report/WSSSPE4_report.tex
adding update for social group (8.6) from Stuart

@@
 %
-\thanks{{}$^{(3)}$ \note{Center for Research Computing \& Department of Computer Science and Engineering,
-University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}}
+\thanks{{}$^{(3)}$ Center for Research Computing \& Department of Computer Science and Engineering,
+University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}
 %
@@
 %
-\thanks{{}$^{(10)}$ \note{info needed}}
+\thanks{{}$^{(10)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
 %

### 2017-03-17 WSSSPE4_report/wg_social.tex
adding update for social group (8.6) from Stuart

@@
 
-\note{Stuart to write this}
+\note{from Stuart}
 
-Introduction to group here, including the overall objective of work in this area. \todo{}
+%Introduction to group here, including the overall objective of work in this area.
+
+This working group was motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability.
+In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges.
+This group met to discuss existing research projects and findings that might be relevant for RSEs and others in the software sustainability domain, then identified several gaps and challenges to tackle in bringing research and practice together.
 
@@
 
-What tasks will the working group undertake \todo{}
+%What tasks will the working group undertake
+The working group planned to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
+This plan resulted in the creation of the following SMART steps listed in the next section.
 

### 2017-03-18 WSSSPE4_report/wg_best_practices_sustainability.tex
adding Software Best Practices for Undergraduates section; fixing other best practices section
Adding content for Section 8.1

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Verifying best practices \& metrics for sustainable research software}
+\label{sec:best-practices-sustainable}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{Ray to write this}
+
+Many open source projects for research software document their best practices that contributed to the sustainability of the software.
+Many open source projects also document software metrics they use to define their research software project’s success.
+This project endeavors to aggregate several sources of these best practices into a consolidated list of best practices, as and to also do the same for software metrics in another consolidated list.
+Our team will then create a workflow to evaluate how open source projects stand up to these lists.
+
+\subsubsection{Participants}
+
+\begin{itemize}
+\item Ray Idaszak <rayi@renci.org>
+\item August Muench <august.muench@aas.org>
+\item Jonah Miller <jmiller@perimeterinstitute.ca>
+\item Lorraine Hwang <ljhwang@ucdavis.edu>
+\end{itemize}
+
+\subsubsection{Working group objective}
+
+The overall objective of the Verify Best Practices and Metrics for Sustainable Software working group is to take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
+This will allow us to identify gaps on both sides.
+This approach will also allow the group to  hypothesize how successful open source projects can be further improved, verify that recommended approaches for software engineering for science/academia are sufficient and valid, and that metrics for software engineering for science/academia are relevant and useful.
+The group has a specific objective of getting a good cross-sampling of disparate software including community model codes, community cyberinfrastructure, community analytic tools, etc.
+
+\subsubsection{Gap or challenge}
+
+The gap addressed by the project is described by the following:
+\begin{itemize}
+\item Are the suggested software engineering for science/academia best practices verified when evaluated against open source research software projects that currently use modern software engineering?
+\item Are the suggested metrics for science/academia verified to be relevant when evaluated against open source research software projects that currently use modern software engineering?
+\item Do research groups have role models to follow for successful best practices in research software?
+\end{itemize}
+
+\subsubsection{Relevant people and resources}
+
+The people spearheading this effort at the time of the writing of this workshop report are Ray Idaszak <rayi@renci.org>, August Muench <august.muench@aas.org>, Jonah Miller <jmiller@perimeterinstitute.ca>, Lorraine Hwang <ljhwang@ucdavis.edu>, Peter Elmer <peter.elmer@cern.ch>, and Hans Fangohr <fangohr@soton.ac.uk>.
+
+\noindent
+Resources include research projects suggested from WSSSPE community including:
+\begin{itemize}
+\item FLASH -- \url{http://flash.uchicago.edu/site/flashcode/}
+\item GROMACS -- http://www.gromacs.org/
+\item CIG -- \url{http://geodynamics.org/}
+\item CUAHSI -- \url{https://www.cuahsi.org/}
+\item CSDMS -- \url{https://csdms.colorado.edu/wiki/Main_Page}
+\item OntoSoft -- \url{http://www.ontosoft.org/}
+\item Anything -- from ASCL (astronomy) \href{https://ui.adsabs.harvard.edu/#search/q=bibstem%3A%22ASCL%22&sort=citation_count%20desc%2C%20bibcode%20desc}{ ASCL entries sorted by citation count}
+\item Yt-project -- (astronomy) \url{http://yt-project.org}
+\item Einstein Toolkit -- (physics) \url{https://einsteintoolkit.org/}
+\item SpEC -- (physics) \url{https://www.black-holes.org/SpEC.html}
+\item pyCLOUDY -- (astrophysics) \url{http://ascl.net/1304.020}
+\end{itemize}
+
+\noindent
+Additional resources include these sources for best practices to be explored:
+
+\begin{itemize}
+\item Good Enough Practices in Scientific Computing, Wilson et al. 2016
+
+\begin{itemize}
+\item \url{https://swcarpentry.github.io/good-enough-practices-in-scientific-computing/}
+\item \url{http://arxiv.org/abs/1609.00037}
+\end{itemize}
+
+\item Best Practices for Scientific Computing, Wilson et al. 2014
+\begin{itemize}
+\item \url{http://dx.doi.org/10.1371/journal.pbio.1001745}
+\item 8 high level categories 
+\end{itemize}
+
+\item Butterfly: a paradigm towards stable bio & neuro informatics tools development
+\begin{itemize}
+\item \url{http://www.frontiersin.org/10.3389/conf.fnins.2015.91.00009/event_abstract}
+\end{itemize}
+
+\item Working towards Sustainable Software for Science: Practice and Experiences
+\begin{itemize}
+\item JORS collection
+\item \url{http://openresearchsoftware.metajnl.com/collections/special/working-towards-sustainable-software-for-science/}
+\end{itemize}
+
+\item CIG Software Development Best Practices
+\begin{itemize}
+\item \url{https://geodynamics.org/cig/dev/best-practices/}
+\item 6 categories; 3 Levels of detail within each: minimum, standard, target
+\end{itemize}
+
+\item Best Practices for Scientific Computing
+\begin{itemize}
+\item \url{http://arxiv.org/abs/1210.0530v1}
+\end{itemize}
+
+\item Software Sustainability Institute
+\begin{itemize}
+\item \url{https://www.software.ac.uk}
+\end{itemize}
+
+\item WSSSPE papers from past WSSSPE workshops
+\begin{itemize}
+\item \url{http://wssspe.researchcomputing.org.uk/}
+\end{itemize}
+
+\item JOSS Peer Review & rOpenSci onboarding
+\begin{itemize} 
+\item \url{http://joss.theoj.org/about#reviewer_guidelines} 
+\item E.g., \url{https://github.com/openjournals/joss-reviews/issues/43}
+\item \url{https://github.com/ropensci/onboarding}
+\item \url{http://ropensci.org/blog/2016/03/28/software-review}
+\end{itemize}
+
+\item RSE 2016 Talk:
+\begin{itemize}
+\item \href{http://www.rse.ac.uk/conf2016_talks#T1.1}{InterMine: Best Practices for Open Source Software}
+\end{itemize}
+
+\item ANNIS corpus linguistic analysis tool
+\begin{itemize}
+\item \url{http://corpus-tools.org/annis/} (humanities code)
+\end{itemize}
+
+\end{itemize}
+
+\subsubsection{Plans}
+
+Our plans are for each of the Verify Best Practices and Metrics for Sustainable Research Software to volunteer to take on one to two projects each, and as incentive the projects that individual take on can be one they are interested in.
+This would lead to an estimated 5-10 projects to be evaluated within this project.
+In terms of increasing this sample size, it is our hope that once we document the workflow by which these evaluations are performed, others can be self-sufficient in following this workflow and contribute their own evaluations of software they wish to evaluate and have interest in.
+
+\subsubsection{SMART steps}
+
+\begin{itemize}
+\item Create mailing list for our project titled:  Verify Best Practices and Metrics for Sustainable Research Software.
+\item Identify where to obtain representative recommended software engineering best practices and metrics within the WSSSPE4 community including single points-of-contact.
+Obtain said best practices and metrics in raw form.
+\item Create GitHub site for project.
+\item Identify where to obtain representative successful open source projects in some manageable number of domains that also successfully use modern software engineering.
+The minimum number of projects per domain should be 1.
+Obtain said list of representative projects.
+
+\begin{itemize}
+\item Example sources: Astronomy source code library; high citation software papers; CIG software list; personal domain knowledge.
+\end{itemize}
+
+\item Create a case study sheet made of questions to be posed about existing research groups.
+\item Perform review.
+\item Aggregate results and document.  
+\item Document workflow of getting to these results so it can be repeated for additional domains.
+\end{itemize}
+
+\subsubsection{More information \& joining instructions}
+
+To join this group or obtain more information about it, please send an email to: \href{wssspe4-verify-best-practices@googlegroups.com}{Verifying best practices \& metrics for sustainable research software.}

### 2017-03-18 WSSSPE4_report/WSSSPE4_report.tex
adds mention of lightning talks rejected & experience papers shifted to lightning

@@
 3 demos
-were accepted.
+were accepted. (Two lightning talks were rejected, while two experience papers
+were accepted as lightning talks.)
 The papers are discussed in Section~\ref{sec:papers},
@@
 
-The keynote was given by Patricia Lago and entitled ``\todo{title}''.
+The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
 

### 2017-03-18 WSSSPE4_report/WSSSPE4_report.tex
small tweaks to new language

@@
 3 demos
-were accepted. (Two lightning talks were rejected, while two experience papers
+were accepted. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
 were accepted as lightning talks.)

### 2017-03-21 WSSSPE4_report/wg_codemeta.tex
update from Alice

--- 2016-10-05 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{CodeMeta}
+\label{sec:CodeMeta}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-\note{Dan to write this}
+\note{Alice to write this}
 
-Introduction to group here, including the overall objective of work in this area.
+"Define a standard for metadata to be used in current repos including authors, dependencies, ...
+repo url" had been listed as a possible project for a working group; as several attendees have been participating in a project, CodeMeta,\footnote{\url{http://codemeta.github.io/}} that is currently working on software metadata, an enthusiastic group formed to work on advancing the CodeMeta project.
 
@@
 
-members of the working group
+\begin{itemize}
+\item Gabrielle Allen <gdallen@illinois.edu>
+\item Stephan Druskat <stephan.druskat@hu-berlin.de>
+\item Iain Emsley <iain.emsley@oerc.ox.ac.uk>
+\item Carole Goble <carole.goble@manchester.ac.uk>
+\item Chris Gwilliams <gwilliamsc@cardiff.ac.uk>
+\item Rafael Jimenez <rafael.jimenez@elixir-europe.org>
+\item Frank Löffler <knarf@cct.lsu.edu>
+\item Kyle Niemeyer <Kyle.Niemeyer@oregonstate.edu>
+\item Thomas Robitaille <thomas.robitaille@gmail.com>
+\item Rob Welch <py12rw@leeds.ac.uk>
+\item Alice Allen <aallen@ascl.net>
+\end{itemize}
 
@@
 
-Specific things the working group wants to accomplish in the context of the larger objective.
+The primary objective of this working group was to find ways to help the CodeMeta project come to fruition.
+CodeMeta seeks in part to create a "Rosetta Stone" for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
+The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable.\citep{CodeMeta_schema}
 
@@
 
-What is the gap or challenge being addressed?
+Research software is often not shared; that that is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides.
+CodeMeta wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software and increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
 
@@
 
-What people, groups, or resources are needed.
+The CodeMeta team members (as of the WSSSPE4 meeting) are listed below; the CodeMeta project leads are shown in bold, and team members who also part of the WSSSPE working group are shown in italics :
+
+\begin{itemize}
+\item {\bf Carl Boettiger}, UC Berkeley
+\item {\bf Matt Jones}, NCEAS
+\item {\bf Arfon Smith}, GitHub
+\item {\bf Abby Mayes}, Mozilla Science Lab
+\item Yolanda Gil, USC ISI
+\item Peter Slaughter, NCEAS
+\item Patricia Cruse, DataCite
+\item Neil Chue Hong SSI
+\item Mercè Crosas, Harvard IQSS
+\item Martin Fenner, DataCite
+\item Mark Hahnel, figshare
+\item Luke Coy, RIT & MSL
+\item {\em Kyle Niemeyer}, Oregon State
+\item Krzysztof Nowak, Zenodo
+\item Daniel Katz, NCSA
+\item {\em Carole Goble}, University of Manchester
+\item Ashley Sands, UCLA
+\item {\em Alice Allen}, ASCL
+
+\end {itemize}
+
+Resources already established by the CodeMeta team can be found online:
+
+\begin{itemize}
+\item Main website and meeting information (\url{http://codemeta.github.io/})
+\item Github repository (\url{https://github.com/codemeta/codemeta})
+\item List of milestones (\url{https://github.com/codemeta/codemeta/milestones})
+\item Gitter discussion site (\url{https://gitter.im/codemeta/codemeta})
+\end {itemize}
+
+Other resources to draw on are the software metadata vocabularies of Schema.org (\url{https://schema.org/SoftwareSourceCode}) and DOAP (\url{https://github.com/ewilderj/doap}), and the Force11 Software Citation Principles\citep{Smith:2016sc}.
+Relevant documents on research software, including the Guidelines for persistently identifying software using DataCite\citep{Gent2016} are available on the UK's Research Data Network's Research Software web page (\url{https://research-data-network.readme.io/docs/research-software}).
 
@@
 
-What tasks will the working group undertake
+Our plans are for the working group to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
+Further, the group wants to assist in the implementation of the specifications and, by providing an outsider’s view, contribute suggestions for more understandable project documentation.
+Finally, the WG seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.
 
@@
 
-What are the first SMART steps proposed?
+With a comparatively large working group, the use of the established CodeMeta Github repository's issue tracking, and the quick responsiveness of CodeMeta lead Matt Jones to a barrage of questions, comments, and logged issues, several of our SMART steps were completed at WSSSPE; these are identified below in italics.
+
+\begin{itemize}
+\item {\em Write to the managers of CodeMeta project to start dialog with the two groups} 
+\item {\em Post questions generated by our discussion as issues on the CodeMeta repository}
+\item Look at the crosswalk table and Schema.org data elements for common elements
+\item Identify text in the project documentation that is unclear and suggest changes
+\item Define a list of services to be added to the crosswalk table and match terms
+\item {\em Identify two roles and information on CodeMeta that would be useful to these people for engaging with the project}
+\item Determine a better way or ways to present the crosswalk table
+\item Create a mailing list for WSSSPE CodeMeta participants
+\end{itemize}
 
@@
 
-How could a reader get more information or get more involved?
+Notes for the working group were taken in real time in a Google document\footnote{\url{https://docs.google.com/document/d/1UxlHIoBRgVWB8NAXYf4Q0yS7PAqa-EYwFfsNuTKSPbs/edit}} and include email replies in response to WG questions from one of the CodeMeta project leads.
+
+Because of the work done at WSSSPE4, the CodeMeta project README file\footnote{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} was greatly expanded to include a description of the project that is geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
+Though a Google group mailing list has been established for the working group, the easiest way to engage with the CodeMeta project is through its Github repository\footnote{\url{https://github.com/codemeta/codemeta}}.

### 2017-03-21 WSSSPE4_report/wg_access.tex
fixing URL, removing todo
updating participants lists in working groups to be consistent

@@
 %members of the working group
-Xinlian Liu, Mark Abraham, Sameer Shende, James Hetherington, Dominik Kempf, Michael R. Crusoe, Radovan Bast
+\begin{itemize}
+  \item Xinlian Liu <liu@hood.edu>
+  \item Mark Abraham <mjab@kth.se>
+  \item Sameer Shende <sameer@cs.uoregon.edu>
+  \item James Hetherington <j.hetherington@ucl.ac.uk>
+  \item Dominik Kempf <dominic.kempf@iwr.uni-heidelberg.de>
+  \item Michael R. Crusoe <michael.crusoe@gmail.com>
+  \item Radovan Bast <radovan.bast@uit.no>
+\end{itemize}
 
@@
 
-https://groups.google.com/forum/#!forum/continuous-integration-for-research-software
+For more information, see \url{https://groups.google.com/forum/#!forum/continuous-integration-for-research-software}
 
-\todo{?}
+%\todo{?}

### 2017-03-21 WSSSPE4_report/wg_alliance.tex
updating participants lists in working groups to be consistent

@@
 
-The participants are:
-Neil Chue Hong <N.ChueHong@software.ac.uk>,
-Jean Salac <jeansalac@virginia.edu>,
-Radovan Bast <radovan.bast@uit.no>,
-Lorraine Hwang <ljhwang@ucdavis.edu>,
-Karthik Ram <karthik.ram@berkeley.edu>,
-and
-Peter Elmer <Peter.Elmer@cern.ch>.
+\begin{itemize}
+\item Neil Chue Hong <N.ChueHong@software.ac.uk>
+\item Jean Salac <jeansalac@virginia.edu>
+\item Radovan Bast <radovan.bast@uit.no>
+\item Lorraine Hwang <ljhwang@ucdavis.edu>
+\item Karthik Ram <karthik.ram@berkeley.edu>
+\item Peter Elmer <Peter.Elmer@cern.ch>
+\end{itemize}
 

### 2017-03-21 WSSSPE4_report/wg_prototyping.tex
updating participants lists in working groups to be consistent

@@
 \begin{itemize}
-  \item Santiago N\'u\~nez-Corrales, University of Illinois Urbana--Champaign
-        <nunezco2@illinois.edu>
-  \item Chris Sweet, University of Notre Dame <chris.sweet@nd.edu>
-  \item Steven Brandt, Louisiana State University <sbrandt@cct.lsu.edu>
-  \item Dominic Orchard, University of Kent <d.a.orchard@kent.ac.uk>
+  \item Santiago N\'u\~nez-Corrales <nunezco2@illinois.edu>
+  \item Chris Sweet <chris.sweet@nd.edu>
+  \item Steven Brandt <sbrandt@cct.lsu.edu>
+  \item Dominic Orchard <d.a.orchard@kent.ac.uk>
 \end{itemize}

### 2017-03-21 WSSSPE4_report/wg_access.tex
rm leftover )
also time limits may force groups to own CI

@@
 %Introduction to group here, including the overall objective of work in this area.
-Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
+Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code or tests which exceed time limits for free plans must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
 For example, software groups that develop BIOUNO, CI4SI, GROMACS) have done this.

### 2017-03-21 WSSSPE4_report/wg_access.tex
give more details about NeIC work

--- 2017-03-21 WSSSPE4_report/wg_access.tex
@@
 %Introduction to group here, including the overall objective of work in this area.
-Each developer of software with uncommon needs (hardware, software, libraries, data sets) or non-public code must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
-For example, software groups that develop BIOUNO, CI4SI, GROMACS) have done this.
+Each developer of software with uncommon needs (hardware, software, libraries, data sets), non-public code, or tests that exceed time limits for free plans must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
+For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
 
@@
 \begin{enumerate}
-\item Learn how donate machines to Debian
+\item Learn how to donate machines to Debian
 \item Find funding, perhaps via NEIC? (\url{https://wiki.neic.no/wiki/NeIC_Community_Wiki}) 

### 2017-03-27 WSSSPE4_report/WSSSPE4_report.tex
adding partial Alice Allen info
adding ORCID
updating participants lists in working groups to be consistent
adding panel info

@@
 Lorraine~Hwang$^{(4)}$,
-Alice~Allen$^{(5)}$,
-Jean~Salac$^{(6)}$,
-Neil~Chue~Hong$^{(7)}$,
-Santiago~Nunez-Corrales$^{(8)}$
-Anshu~Dubey$^{(9)}$,
-R.~Stuart~Geiger$^(10)$,
-Jonah~Miller$^{11)}$,
-Emily~Chen$^{12)}$,
-Wolfgang~Bangerth$^{13)}$,
+Simon~Hettrick$^{(5)}$,
+Alice~Allen$^{(6)}$,
+Jean~Salac$^{(7)}$,
+Neil~Chue~Hong$^{(8)}$,
+Santiago~Nunez-Corrales$^{(9)}$
+Anshu~Dubey$^{(10)}$,
+R.~Stuart~Geiger$^(11)$,
+Jonah~Miller$^{12)}$,
+Emily~Chen$^{123}$,
+Wolfgang~Bangerth$^{14)}$,
 \\Author order to be determined based on contributions
@@
 %
-\thanks{{}$^{(6)}$ \note{info needed}}
+\thanks{{}$^{(6)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
 %
@@
 %
-\thanks{{}$^{(10)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
-%
-\thanks{{}$^{(11)}$ \note{info needed}}
+\thanks{{}$^{(10)}$ \note{info needed}}
+%
+\thanks{{}$^{(11)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
 %
@@
 %
-\thanks{{}$^{(13)}$ Department of Mathematics, Colorado State
+\thanks{{}$^{(13)}$ \note{info needed}}
+%
+\thanks{{}$^{(14)}$ Department of Mathematics, Colorado State
   University, Fort Collins, CO 80523, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
@@
 
-\todo{Introduction}
+%\todo{Introduction}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Panel discussion} \label{sec:panel}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{description of panel discussion on best practices} \note{Simon will write this}
+\section{Best practices for scientific software panel discussion} \label{sec:panel}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{from Simon}
+
+To investigate best practices for scientific software we put together a panel of five people with different perspectives on ``Best Practices for Scientific Software''.
+Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
+Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester are both Research Software Engineers with decades of experience of writing code for researchers.
+Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business.
+Finally, Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
+
+The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that does not span the Atlantic.
+After translating this into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved in, not just using, but also writing code.
+Hence, the following discussions would \todo{what goes here?}
+
+The panel opened by noting that the world of scientific computing is changing rapidly, and the panel asked which practices have emerged recently that, if not handled correctly, might be the next generation's problem?
+This quickly led to a discussion about the dangers of containerization.
+These technologies are pushed by some---a minority it is hoped---as a panacea for sustainability.
+The panel and attendees generally agreed that containerization is useful, but should not be seen as the solution to sustainability.
+After all, these technologies are just as liable as any to decay.
+
+A later question looked at the common practices from software engineering that the scientific community should be using.
+The discussion that followed picked up on the key elements of sustainability, version control, documentation and ensuring that communities are active and engaged.
+When asked whether there was resistance from the research community to adopting these practices, few negative experiences were reported.
+It would appear that the problem lies in getting the message out about these technologies rather than trying sell researchers on the benefit of them.
+
+The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?
+
+\todo{} <WILL ADD HERE ONCE I'VE TALKED TO PEOPLE>
 

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
formatting
fixing affiliation info

@@
 \thanks{{}$^{(1)}$ \hspace{-1ex}National Center for Supercomputing Applications (NCSA) \&
-Electrical and Computer Engineering (ECE) Department \&
+Department of Computer Science  \&
+Department of Electrical and Computer Engineering  \&
 School of Information Sciences (iSchool),
-University of Illinois at Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
+University of Illinois, Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525}
 %
 \thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
-Oregon State University, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
+Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
 %

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
Adding my affiliation details
Added final sentence and completed hanging sentence.

@@
 %
-\thanks{{}$^{(5)}$ \note{info needed}}
+\thanks{{}$^{(5)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
 %
@@
 
-To investigate best practices for scientific software we put together a panel of five people with different perspectives on ``Best Practices for Scientific Software''.
-Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
-Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester are both Research Software Engineers with decades of experience of writing code for researchers.
+To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software''.
+These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
+Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers.
 Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business.
@@
 
-The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that does not span the Atlantic.
-After translating this into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved in, not just using, but also writing code.
-Hence, the following discussions would \todo{what goes here?}
-
-The panel opened by noting that the world of scientific computing is changing rapidly, and the panel asked which practices have emerged recently that, if not handled correctly, might be the next generation's problem?
+The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that has not crossed the Atlantic.
+After translating this question into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
+Hence, the panel were going to have to deal with questions from a highly experienced audience.
+
+The panel opened by noting that the world of scientific computing is changing rapidly.
+Each panelist was asked for their views on practices which had recently emerged and, if not handled correctly, would become the basis for the next generation's software problems?
 This quickly led to a discussion about the dangers of containerization.
-These technologies are pushed by some---a minority it is hoped---as a panacea for sustainability.
+These technologies are pushed by some---a minority it is hoped---as a panacea for software sustainability.
 The panel and attendees generally agreed that containerization is useful, but should not be seen as the solution to sustainability.
@@
 A later question looked at the common practices from software engineering that the scientific community should be using.
-The discussion that followed picked up on the key elements of sustainability, version control, documentation and ensuring that communities are active and engaged.
+The discussion that followed picked up on the key elements of basic sustainable practice: version control, documentation and ensuring that communities are active and engaged.
 When asked whether there was resistance from the research community to adopting these practices, few negative experiences were reported.
-It would appear that the problem lies in getting the message out about these technologies rather than trying sell researchers on the benefit of them.
-
-The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?
-
-\todo{} <WILL ADD HERE ONCE I'VE TALKED TO PEOPLE>
+From the panel's point of view, it would appear that the problem in getting researchers to employ sustainable practicies lies more in getting the message out to researchers, rather than selling them on the benefits.
+
+The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?" A few good responses were discussed, but the biggest laugh went to Mike Croucher who said "Containerize everything!".
 

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
Update WSSSPE4_report.tex

@@
 
-The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?" A few good responses were discussed, but the biggest laugh went to Mike Croucher who said "Containerize everything!".
+The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?"
+A few good responses were discussed, but the biggest laugh went to Mike Croucher who said "Containerize everything!"
 

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
spelling

@@
 When asked whether there was resistance from the research community to adopting these practices, few negative experiences were reported.
-From the panel's point of view, it would appear that the problem in getting researchers to employ sustainable practicies lies more in getting the message out to researchers, rather than selling them on the benefits.
+From the panel's point of view, it would appear that the problem in getting researchers to employ sustainable practices lies more in getting the message out to researchers, rather than selling them on the benefits.
 

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
fixing quotes

@@
 
-To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software''.
+To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
 These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
@@
 
-The panel rounded up with a final question about "If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?"
-A few good responses were discussed, but the biggest laugh went to Mike Croucher who said "Containerize everything!"
+The panel rounded up with a final question about ``If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?''
+A few good responses were discussed, but the biggest laugh went to Mike Croucher who said ``Containerize everything!''
 

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
Added affiliation
very minor space fixing

@@
 %
-\thanks{{}$^{(1)}$ \hspace{-1ex}National Center for Supercomputing Applications (NCSA) \&
+\thanks{{}$^{(1)}$ National Center for Supercomputing Applications (NCSA) \&
 Department of Computer Science  \&
@@
 %
-\thanks{{}$^{(12)}$ \note{info needed}}
+\thanks{{}$^{(12)}$ Perimeter Institute for Theoretical Physics \&
+Department of Physics, University of Guelph, 
+Guelph, ON, Canada; 
+jmiller@perimeterinstitute.ca; ORCID: 0000-0001-6432-7860}
 %

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
more little changes

@@
 Anshu~Dubey$^{(10)}$,
-R.~Stuart~Geiger$^(11)$,
-Jonah~Miller$^{12)}$,
-Emily~Chen$^{123}$,
-Wolfgang~Bangerth$^{14)}$,
+R.~Stuart~Geiger$^{(11)}$,
+Jonah~Miller$^{(12)}$,
+Emily~Chen$^{(13)}$,
+Wolfgang~Bangerth$^{(14)}$,
 \\Author order to be determined based on contributions

### 2017-03-29 WSSSPE4_report/WSSSPE4_report.tex
adding ORCID
adding partial affil info

@@
 %
-\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: ?}
+\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: 0000-0002-1021-3101}
 %
@@
 %
-\thanks{{}$^{(13)}$ \note{info needed}}
+\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: \note{info needed}}
 %

### 2017-03-30 WSSSPE4_report/WSSSPE4_report.tex
adding orcid, removing zip

@@
 %
-\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: \note{info needed}}
+\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
 %
 \thanks{{}$^{(14)}$ Department of Mathematics, Colorado State
-  University, Fort Collins, CO 80523, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
+  University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
 

### 2017-03-31 WSSSPE4_report/WSSSPE4_report.tex
adding more affil details

@@
 %
-\thanks{{}$^{(9)}$ \note{info needed}}
-%
-\thanks{{}$^{(10)}$ \note{info needed}}
+\thanks{{}$^{(9)}$ Illinois Informatics Institute (I3) and National Center for Supercomputing Applications (NCSA), University of Illinois at Urbana-Champaign, IL, USA;
+nunezco2@illinois.edu;
+ORCID: 0000-0003-4342-6223}
+%
+\thanks{{}$^{(10)}$ Mathematics and Computer Science Division,
+Argonne National Laboratory,
+Lemont, IL, USA \& 
+University of Chicago, 
+Chicago, IL, USA;
+Adubey@anl.gov;
+ORCID: 0000-0003-3299-7426}
 %

### 2017-04-08 WSSSPE4_report/WSSSPE4_report.tex
another affil update
adding last missing affiliation
additions to abstract and conclusions

@@
 Emily~Chen$^{(13)}$,
-Wolfgang~Bangerth$^{(14)}$,
-\\Author order to be determined based on contributions
+Wolfgang~Bangerth$^{(14)}$%,
+%\\Author order to be determined based on contributions
 }
@@
 %
-\thanks{{}$^{(7)}$ \note{info needed}}
-%
-\thanks{{}$^{(8)}$ \note{info needed}}
+\thanks{{}$^{(7)}$ University of Virginia, Charlottesville, VA, USA; jeansalac@virginia.edu; ORCID: 0000-0001-5971-9333}
+%
+\thanks{{}$^{(8)}$ University of Edinburgh, UK; N.ChueHong@epcc.ed.ac.uk;
+ORCID: 0000-0002-8876-7606}
 %
@@
 Software for Science: Practice and Experiences (WSSSPE4). The report
-includes a description of the keynote presentation of the workshop,
-which % TODO
+includes a description of the keynote presentation of the workshop, the
+mission and vision statements that were drafted at the workshop and 
+finalized shortly after it, a set of idea papers, position papers, experience
+papers, demos, and lightning talks, and a panel discussion.  The main part of the report
+covers the set of working groups that formed during the meeting, and for each,
+discusses the participant, the objective and goal, and how the objective can
+be reached, along with contact information for readers who may want to join
+the group. Finally, we present results from a survey of the workshop attendees.
+
 % It also summarizes a set of lightning
@@
 There events included discussion of the WSSSPE mission and vision, a keynote,
-a set of presentations, a panel, and a number of working groups.  About a day of the
+a set of presentations, a panel, and a number of working groups.  One full day of the
 workshop was spent with participants in the working groups, which occurred in parallel
-with each other.
+with each other.  And each of the working groups left with a plan for how they could move
+forward.
 
@@
 
-In WSSSPE4, we
-
-\todo{discuss Twitter activity}
+In WSSSPE4, we heard about a number of interesting projects and ideas, and used those
+ideas to create and form working groups, intended to start at the workshop and then
+continue afterwards, to address challenges that arose from the workshop ideas.  This workshop
+has reinforced the lesson from WSSSPE3 that it is relatively easy to get motivated people
+to attend a meeting and productively spend their time there both doing work and planning
+more work, but it is very hard to get that additional work after the meeting to take place.
+The main problem seems to be one of time.  Once the attendees have agree to spend
+their time at the workshop, they put their energy into doing so productively, but they haven't
+really committed themselves to anything more than this, so their energy and effort trails
+off, as all of their other commitments (particularly, those they are funded to do as part of
+their jobs) come back to the fore. Without a process to resolve this concern, the utility of
+further multi-day WSSSPE workshops is unclear.
+
+%\todo{discuss Twitter activity}
 

### 2017-04-08 WSSSPE4_report/WSSSPE4_report.tex
author changes

@@
 Simon~Hettrick$^{(5)}$,
-Alice~Allen$^{(6)}$,
+Ray Idaszak$^{(6)}$,
 Jean~Salac$^{(7)}$,
 Neil~Chue~Hong$^{(8)}$,
-Santiago~Nunez-Corrales$^{(9)}$
-Anshu~Dubey$^{(10)}$,
+Santiago~Nunez-Corrales$^{(9)}$,
+Alice~Allen$^{(10)}$,
 R.~Stuart~Geiger$^{(11)}$,
@@
 Emily~Chen$^{(13)}$,
-Wolfgang~Bangerth$^{(14)}$%,
+Anshu~Dubey$^{(14)}$,
+Wolfgang~Bangerth$^{(15)}$%,
 %\\Author order to be determined based on contributions
@@
 %
-\thanks{{}$^{(6)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
+\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: \todo{}}
 %
@@
 %
-\thanks{{}$^{(10)}$ Mathematics and Computer Science Division,
+\thanks{{}$^{(10)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
+%
+\thanks{{}$^{(11)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
+%
+\thanks{{}$^{(12)}$ Perimeter Institute for Theoretical Physics \&
+Department of Physics, University of Guelph, 
+Guelph, ON, Canada; 
+jmiller@perimeterinstitute.ca; ORCID: 0000-0001-6432-7860}
+%
+\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
+%
+\thanks{{}$^{(14)}$ Mathematics and Computer Science Division,
 Argonne National Laboratory,
@@
 %
-\thanks{{}$^{(11)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
-%
-\thanks{{}$^{(12)}$ Perimeter Institute for Theoretical Physics \&
-Department of Physics, University of Guelph, 
-Guelph, ON, Canada; 
-jmiller@perimeterinstitute.ca; ORCID: 0000-0001-6432-7860}
-%
-\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
-%
-\thanks{{}$^{(14)}$ Department of Mathematics, Colorado State
+\thanks{{}$^{(15)}$ Department of Mathematics, Colorado State
   University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}

### 2017-04-09 WSSSPE4_report/WSSSPE4_report.tex
adding final ORCID
adding keynote

@@
 %
-\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: \todo{}}
+\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID:0000-0002-3444-7615}
 %
@@
 
-\note{need a person to write this - Dan has asked Patricia if she will, if not, perhaps Dan will do it}
+\note{from Patricia}
 
 The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
+%
+Sustainability is broadly associated with mother nature and ecologic issues.
+When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}.
+In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.
+
+Supporting technical sustainability in (scientific) software is very near to supporting properties like longevity and resilience, i.e., the ability of a software system or application to fulfill its intent~\cite{Huisman2016} as planned independently from time passing or from changes in its operational context.
+Technical sustainability has never been easy, but it is even more of a challenge nowadays with the increasing demands on software solutions to support many aspects of our daily lives.
+
+The keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability.
+This provided a complementary angle to the workshop discussions.
+It addressed software energy efficiency (i.e., how to develop software that consumes less energy, like mobile apps that should consume less device-battery) and software energy-awareness (i.e., how to develop software supporting features that in turn have a lesser ecologic impact).
+
+The keynote used few myths to elaborate on such challenges and related research. Among them:
+
+\begin{itemize}
+\item[] {\bf Myth \#1: Energy-efficient hardware will solve the issue}.
+At the implementation level, to make software use resources effectively, we need to first understand the impact of different coding practices on the way software exploits computing resources (hence energy to feed them).
+In addition, when we have to deal with complex software or new projects, we must understand ``the big picture,'' hence we must reason at the design or architecture level.
+Maybe the software is not implemented yet, or maybe we must identify which design decisions are more or less beneficial for energy consumption.
+To this aim, we capture best practices in reusable patterns and so-called architectural tactics.
+These document generic designs, specific implementations, metrics, and measures.\\
+
+\item[] {\bf Myth \#2: Extensive experimentation will provide the needed know how}.
+While empirical methods to design and execute software engineering experiments are mature and well known, they are insufficient to handle the too many variables and uncertainties for modern software to be sustainable.
+We need new empirical methods that help us gather evidence faster~\cite{Procaccianti2015}.
+This would probably entail relaxing the too many constraints that govern current rules for mitigating threats to validity, and (equally importantly) collaboration- and experience-sharing to build a common knowledge base.
+In addition, we should exploit better (and more truthful) visualizations of experimentation results, to communicate on advances (and less promising directions.)\\
+
+\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments we do not need to invest in a ``greener cloud'' anymore}.
+The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}.
+Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability.
+We need to build smarter software that both manages resources more efficiently in the cloud, and better understands the changes in the software operational context (including user requirements) and dynamically adapt to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\
+
+\end{itemize}
+
+While these points are valid in any type of software, scientific software is even more sensitive to sustainability intents, being in charge of supporting science.
 

### 2017-04-09 WSSSPE4_report/WSSSPE4_report.tex
adding Patricia as author

@@
 Anshu~Dubey$^{(14)}$,
-Wolfgang~Bangerth$^{(15)}$%,
+Wolfgang~Bangerth$^{(15)}$,
+Patricia~Lago$^{(16)}$
 %\\Author order to be determined based on contributions
@@
   University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
+%
+\thanks{{}$^{(16)}$ Vrije Universiteit Amsterdam, The Netherlands; p.lago@vu.nl; ORCID: 0000-0002-2234-0845}
 

### 2017-04-09 WSSSPE4_report/survey_details.tex
final formatting changes in survey details, new page for references

@@
 Note: typos and spelling errors have been corrected, and spelling has been changed from UK standard to US standard in some cases, but no other changes have been made to the attendee responses.
+
+\hspace{0.2cm}
 
@@
     \label{fig:SFig1}
-    }{\includegraphics[width=.34\linewidth]{figures/SurveyFig1rev}}
+    }{\includegraphics[width=.33\linewidth]{figures/SurveyFig1rev}}
     ~
@@
     \label{fig:SFig3}
-    }{\includegraphics[width=.25\linewidth]{figures/SurveyFig3rev}}
-    \caption{Responses to generate questions about the conference program}
+    }{\includegraphics[width=.26\linewidth]{figures/SurveyFig3rev}}
+    \caption{Responses to general questions about the conference program}
 \end{figure}
@@
 
+\newpage
 \noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}
@@
 
+\newpage
 \noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
@@
 \noindent \textbf{8. Working Group Name}
+
+\vspace{0.2cm}
 

### 2017-04-09 WSSSPE4_report/wg_best_practices_sustainability.tex

@@
 Many open source projects for research software document their best practices that contributed to the sustainability of the software.
-Many open source projects also document software metrics they use to define their research software project’s success.
+Many open source projects also document software metrics they use to define their research software project's success.
 This project endeavors to aggregate several sources of these best practices into a consolidated list of best practices, as and to also do the same for software metrics in another consolidated list.
@@
 
-The overall objective of the Verify Best Practices and Metrics for Sustainable Software working group is to take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
+The overall objective of the Verifying Best Practices and Metrics for Sustainable Software working group is to take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
 This will allow us to identify gaps on both sides.
@@
 \begin{itemize}
-\item FLASH -- \url{http://flash.uchicago.edu/site/flashcode/}
-\item GROMACS -- http://www.gromacs.org/
-\item CIG -- \url{http://geodynamics.org/}
-\item CUAHSI -- \url{https://www.cuahsi.org/}
-\item CSDMS -- \url{https://csdms.colorado.edu/wiki/Main_Page}
-\item OntoSoft -- \url{http://www.ontosoft.org/}
-\item Anything -- from ASCL (astronomy) \href{https://ui.adsabs.harvard.edu/#search/q=bibstem%3A%22ASCL%22&sort=citation_count%20desc%2C%20bibcode%20desc}{ ASCL entries sorted by citation count}
-\item Yt-project -- (astronomy) \url{http://yt-project.org}
-\item Einstein Toolkit -- (physics) \url{https://einsteintoolkit.org/}
-\item SpEC -- (physics) \url{https://www.black-holes.org/SpEC.html}
-\item pyCLOUDY -- (astrophysics) \url{http://ascl.net/1304.020}
+\item FLASH (astrophysics, \url{http://flash.uchicago.edu/site/flashcode/})
+\item GROMACS (chemistry, \url{http://www.gromacs.org/})
+\item CIG (Earth science, \url{http://geodynamics.org/})
+\item CUAHSI (Earth science, \url{https://www.cuahsi.org/})
+\item CSDMS (Earth science, \url{https://csdms.colorado.edu/wiki/Main_Page})
+\item OntoSoft (biology, \url{http://www.ontosoft.org/})
+\item ASCL (astronomy, \url{http://ascl.net}), specifically ASCL entries sorted by citation count\footnote{\url{https://ui.adsabs.harvard.edu/\#search/q=bibstem\%3A\%22ASCL\%22&sort=citation_count\%20desc\%2C\%20bibcode\%20desc}}
+\item yt-project (astronomy,  \url{http://yt-project.org})
+\item Einstein Toolkit (physics, \url{https://einsteintoolkit.org/})
+\item SpEC (physics, \url{https://www.black-holes.org/SpEC.html})
+\item pyCLOUDY (astrophysics, \url{http://ascl.net/1304.020})
 \end{itemize}
@@
 \begin{itemize}
-\item Good Enough Practices in Scientific Computing, Wilson et al. 2016
 
+\item Good Enough Practices in Scientific Computing, Wilson et al. 2016~\cite{DBLP:journals/corr/WilsonBCKNT16}
 \begin{itemize}
 \item \url{https://swcarpentry.github.io/good-enough-practices-in-scientific-computing/}
-\item \url{http://arxiv.org/abs/1609.00037}
 \end{itemize}
 
-\item Best Practices for Scientific Computing, Wilson et al. 2014
+\item Best Practices for Scientific Computing, Wilson et al. 2014~\cite{bestprSC}
 \begin{itemize}
-\item \url{http://dx.doi.org/10.1371/journal.pbio.1001745}
 \item 8 high level categories 
@@
 
-\item Butterfly: a paradigm towards stable bio & neuro informatics tools development
-\begin{itemize}
-\item \url{http://www.frontiersin.org/10.3389/conf.fnins.2015.91.00009/event_abstract}
-\end{itemize}
-
-\item Working towards Sustainable Software for Science: Practice and Experiences
-\begin{itemize}
-\item JORS collection
-\item \url{http://openresearchsoftware.metajnl.com/collections/special/working-towards-sustainable-software-for-science/}
-\end{itemize}
+\item Butterfly: a paradigm towards stable bio and neuro informatics tools development~\cite{10.3389/conf.fnins.2015.91.00009}
 
@@
 
-\item Best Practices for Scientific Computing
-\begin{itemize}
-\item \url{http://arxiv.org/abs/1210.0530v1}
-\end{itemize}
+%same as 2nd in list
+%\item Best Practices for Scientific Computing
+%\begin{itemize}
+%\item \url{http://arxiv.org/abs/1210.0530v1}
+%\end{itemize}
 
@@
 \item \url{http://wssspe.researchcomputing.org.uk/}
+\item published as JORS collection (\url{http://openresearchsoftware.metajnl.com/collections/special/working-towards-sustainable-software-for-science/})
 \end{itemize}
 
-\item JOSS Peer Review & rOpenSci onboarding
+\item JOSS Peer Review and rOpenSci onboarding
 \begin{itemize} 
 \item \url{http://joss.theoj.org/about#reviewer_guidelines} 
-\item E.g., \url{https://github.com/openjournals/joss-reviews/issues/43}
+\item Example JOSS review: \url{https://github.com/openjournals/joss-reviews/issues/43}
 \item \url{https://github.com/ropensci/onboarding}
@@
 \begin{itemize}
-\item \href{http://www.rse.ac.uk/conf2016_talks#T1.1}{InterMine: Best Practices for Open Source Software}
+\item InterMine: Best Practices for Open Source Software (\url{http://www.rse.ac.uk/conf2016_talks#T1.1})
 \end{itemize}
@@
 \begin{itemize}
-\item \url{http://corpus-tools.org/annis/} (humanities code)
+\item humanities code (\url{http://corpus-tools.org/annis/})
 \end{itemize}
@@
 Our plans are for each of the Verify Best Practices and Metrics for Sustainable Research Software to volunteer to take on one to two projects each, and as incentive the projects that individual take on can be one they are interested in.
-This would lead to an estimated 5-10 projects to be evaluated within this project.
+This would lead to an estimated 5--10 projects to be evaluated within this project.
 In terms of increasing this sample size, it is our hope that once we document the workflow by which these evaluations are performed, others can be self-sufficient in following this workflow and contribute their own evaluations of software they wish to evaluate and have interest in.
@@
 
-To join this group or obtain more information about it, please send an email to: \href{wssspe4-verify-best-practices@googlegroups.com}{Verifying best practices \& metrics for sustainable research software.}
+To join this group or obtain more information about it, please send an email to the Verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.

### 2017-04-09 WSSSPE4_report/wg_best_practices_undergrads.tex

@@
 
-The participants in this working group are/were Jonah Miller,
-Aleksandra Nenadic, Raniere Silva, Francisco Queiroz, Hans Fangohr,
-and Prabhjyot Sing. Aleksandra Nenadic and Raniere Silva are involved
+\begin{itemize}
+  \item Jonah Miller <jmiller@perimeterinstitute.ca>
+  \item Aleksandra Nenadic <a.nenadic@manchester.ac.uk>
+  \item Raniere Silva <raniere.silva@software.ac.uk>
+  \item Francisco Queiroz <chico@puc-rio.br>
+  \item Hans Fangohr <fangohr@soton.ac.uk>
+  \item Prabhjyot Sing <prabhjyot10@gmail.com>
+\end{itemize}
+
+\noindent Aleksandra Nenadic and Raniere Silva are involved
 in software and data carpentry.

### 2017-04-09 WSSSPE4_report/wg_codemeta.tex

@@
 
-"Define a standard for metadata to be used in current repos including authors, dependencies, ...
-repo url" had been listed as a possible project for a working group; as several attendees have been participating in a project, CodeMeta,\footnote{\url{http://codemeta.github.io/}} that is currently working on software metadata, an enthusiastic group formed to work on advancing the CodeMeta project.
+This group, which included people who had previously been working on the CodeMeta project\footnote{\url{http://codemeta.github.io/}}, formed to address the proposed project,
+ ``Define a standard for metadata to be used in current repos including authors, dependencies, ... repo url.''
 
@@
 \item Rafael Jimenez <rafael.jimenez@elixir-europe.org>
-\item Frank Löffler <knarf@cct.lsu.edu>
+\item Frank L\"{o}ffler <knarf@cct.lsu.edu>
 \item Kyle Niemeyer <Kyle.Niemeyer@oregonstate.edu>
@@
 The primary objective of this working group was to find ways to help the CodeMeta project come to fruition.
-CodeMeta seeks in part to create a "Rosetta Stone" for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
-The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable.\citep{CodeMeta_schema}
+CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
+The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.
 
@@
 \item {\bf Matt Jones}, NCEAS
-\item {\bf Arfon Smith}, GitHub
+\item {\bf Arfon Smith}, Space Telescope Science Institute
 \item {\bf Abby Mayes}, Mozilla Science Lab
@@
 \item Neil Chue Hong SSI
-\item Mercè Crosas, Harvard IQSS
+\item Merc\`{e} Crosas, Harvard IQSS
 \item Martin Fenner, DataCite
 \item Mark Hahnel, figshare
-\item Luke Coy, RIT & MSL
+\item Luke Coy, RIT \& MSL
 \item {\em Kyle Niemeyer}, Oregon State
 \item Krzysztof Nowak, Zenodo
-\item Daniel Katz, NCSA
+\item Daniel S. Katz, NCSA
 \item {\em Carole Goble}, University of Manchester
@@
 
-Other resources to draw on are the software metadata vocabularies of Schema.org (\url{https://schema.org/SoftwareSourceCode}) and DOAP (\url{https://github.com/ewilderj/doap}), and the Force11 Software Citation Principles\citep{Smith:2016sc}.
-Relevant documents on research software, including the Guidelines for persistently identifying software using DataCite\citep{Gent2016} are available on the UK's Research Data Network's Research Software web page (\url{https://research-data-network.readme.io/docs/research-software}).
+Other resources to draw on are the software metadata vocabularies of Schema.org\footnote{\url{https://schema.org/SoftwareSourceCode}} and DOAP\footnote{\url{https://github.com/ewilderj/doap}}, and the FORCE11 Software Citation Principles~\cite{Smith:2016sc}.
+Relevant documents on research software, including the Guidelines for persistently identifying software using DataCite~\cite{Gent2015} are available on the UK's Research Data Network's Research Software web page\footnote{\url{https://research-data-network.readme.io/docs/research-software}}.
 
@@
 Our plans are for the working group to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
-Further, the group wants to assist in the implementation of the specifications and, by providing an outsider’s view, contribute suggestions for more understandable project documentation.
+Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation.
 Finally, the WG seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.

### 2017-04-09 WSSSPE4_report/wg_letters.tex

@@
 
-Alice Allen, Astrophysics Source Code Library, \url{aallen@ascl.net};
-Gabrielle Allen, NCSA, \url{University of Illinois, gdallen@illinois.edu};
-Wolfgang Bangerth, Colorado State University, \url{bangerth@colostate.edu};
-Bruce Childers, University of Pittsburgh, \url{childers@cs.pitt.edu };
-Lorraine Hwang, University of California, \url{Davis, ljhwang@ucdavis.edu };
-Dan Katz, University of Illinois, \url{dskatz@illinois.edu };
-Frank L{\"o}ffler, Louisiana State University, \url{knarf@cct.lsu.edu};
-Kyle Niemeyer, Oregon State University, \url{kyle.niemeyer@oregonstate.edu};
-Janos Sallai, Vanderbilt University, \url{janos.sallai@vanderbilt.edu }.
+\begin{itemize}
+\item Alice Allen <aallen@ascl.net>
+\item Gabrielle Allen <gdallen@illinois.edu>
+\item Wolfgang Bangerth <bangerth@colostate.edu>
+\item Bruce Childers <childers@cs.pitt.edu>
+\item Lorraine Hwang <ljhwang@ucdavis.edu>
+\item Daniel S. Katz <d.katz@ieee.org>
+\item Frank L{\"o}ffler <knarf@cct.lsu.edu>
+\item Kyle Niemeyer <Kyle.Niemeyer@oregonstate.edu>
+\item Janos Sallai <janos.sallai@vanderbilt.edu>
+\end{itemize}
 

### 2017-04-09 WSSSPE4_report/wg_metrics.tex

--- 2016-09-14 WSSSPE4_report/wg_open_research_index.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open Research Index}
-\label{sec:open-research-index}
+\subsection{Meaningful metrics for sustainable software}
+\label{sec:metrics}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\note{content from Emily}
+
+%Introduction to group here, including the overall objective of work in this area.
+
+Meaningful Metrics for Sustainable Scientific Software aims to increase the visibility on the quality of scientific software, facilitate the reusability of scientific software, and promote the best software practices by standardizing metrics via interviews with scientific software developers.
+This working group believes improving the current software metrics system will increase software sustainability.
+Currently, there are inefficiencies regarding software duplication, sustainability, and selection, as well as others, within the scientific software community.
+In order to address these inefficiencies, Meaningful Metrics for Sustainable Scientific Software aims to create a goal-oriented method to collecting productive metrics by focusing on the developer side of software.
+
+\subsubsection{Participants}
+
+%members of the working group
+
+\begin{itemize}
+\item Emily Chen <echen35@illinois.edu>
+\item Patricia Lago <pdotlago@gmail.com>
+\item Udit Nangia <unangi2@illinois.edu>
+\item Tengyu Ma <tengyuma10717@gmail.com>
+\item Aseel Aldabjan <a.dabjan@hotmail.com>
+\end{itemize}
+
+\subsubsection{Working group objective}
+
+%Specific things the working group wants to accomplish in the context of the larger objective.
+
+In the context of the larger objective, Meaningful Metrics for Sustainable Scientific Software hopes to find an efficient solution for streamlining the process of collecting and utilizing metrics to benefit software sustainability, as well as minimize the current inefficiencies within the scientific software community.
+Finding meaningful metrics will improve software evaluation and comparison, thus reducing the effort spent on seeking scientific software.
+
+\subsubsection{Gap or challenge}
+
+%What is the gap or challenge being addressed?
+
+The gap being addressed is the lack of a standard for collecting and presenting metrics.
+This gap delays workflow and creates a multitude of tedious tasks, including searching for the best-fit software, unknowingly duplicating software, and other related busy-work.
+The need for metric standardization stems from the abundance of scientific, both ?dark?
+and open source, software and the difficulties of ensuring the software is sustainable.
+
+\subsubsection{Relevant people and resources}
+
+%What people, groups, or resources are needed.
+
+Relevant people and resources include scientific software developers, researchers that use scientific software, scientific software funding institutions.
+
+\subsubsection{Plans}
+
+%What tasks will the working group undertake
+
+This working group plans on interviewing scientific software developers to form metrics from the goals they have for their software.
+
+\subsubsection{SMART steps}
+
+%What are the first SMART steps proposed?
+
+The SMART steps Meaningful Metrics for Scientific Software has proposed are:
+\begin{enumerate}
+\item Identify the population of scientific software developers who are willing to be interviewed
+\item Define the interview questions and organize them into categories
+\item Interview the participants and map the survey results on goals
+\item Convert the goals to feasible, meaningful metrics
+\item Analyze the collected metrics
+\end{enumerate}
+
+\subsubsection{More information \& joining instructions}
+
+%How could a reader get more information or get more involved?
+
+For more information, please contact Emily Chen at <echen35@illinois.edu>.

### 2017-04-09 WSSSPE4_report/wg_open_research_index.tex

@@
 
-The members of this group at WSSSPE4 were:
-Gabrielle Allen,
-Bruce Childers,
-Robert Haines,
-Caroline Jay,
-Dan Katz,
-Robert McDonald,
-Daniel Mosse,
-and 
-Kyle Niemeyer.
+\begin{itemize}
+\item Gabrielle Allen <gdallen@illinois.edu>
+\item Bruce Childers <childers@pitt.edu>
+\item Robert Haines <robert.haines@manchester.ac.uk>
+\item Caroline Jay <caroline.jay@manchester.ac.uk>
+\item Daniel S. Katz <d.katz@ieee.org>
+\item Robert McDonald <rhmcdona@indiana.edu>
+\item Daniel Mosse <mosse@cs.pitt.edu>
+\item Kyle Niemeyer <Kyle.Niemeyer@oregonstate.edu>
+\end{itemize}
 

### 2017-04-09 WSSSPE4_report/wg_best_practices_sustainability.tex
Misc edits

@@
 Many open source projects also document software metrics they use to define their research software project's success.
-This project endeavors to aggregate several sources of these best practices into a consolidated list of best practices, as and to also do the same for software metrics in another consolidated list.
+This project endeavors to aggregate several sources of these best practices into a consolidated list of best practices, and similar for software metrics.
 Our team will then create a workflow to evaluate how open source projects stand up to these lists.

### 2017-04-09 WSSSPE4_report/wg_open_research_index.tex
Misc edits

@@
 The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
-Our goal is not to eliminate commercial products, but to build on what?s there and provide data and services that are missing.
+Our goal is not to eliminate commercial products, but to build on what's there and provide data and services that are missing.
 
@@
 \item Iterate use case with stakeholders (who, when),
-e.g., funders, Researchfish, researchers, ?
+e.g., funders, Researchfish, researchers ?
 

### 2017-04-09 WSSSPE4_report/wg_prototyping.tex
Misc edits

@@
  ---later to be fully implemented into by scientific programmers---is critical for
- acquiring maturity and proficiency in the grand challenges the Exascale
+ acquiring maturity and proficiency in the grand challenges that Exascale
  attempts to solve, and moreover for allowing scientists to develop their own

### 2017-04-09 WSSSPE4_report/WSSSPE4_report.tex

@@
 
-\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments we do not need to invest in a ``greener cloud'' anymore}.
+\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments, we do not need to invest in a ``greener cloud'' anymore}.
 The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}.
 Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability.
-We need to build smarter software that both manages resources more efficiently in the cloud, and better understands the changes in the software operational context (including user requirements) and dynamically adapt to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\
+We need to build smarter software that both manages resources more efficiently in the cloud, better understands the changes in the software operational context (including user requirements), and dynamically adapts to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\
 
@@
 The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; Its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
+The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
 
@@
 
-{\bf Mission}
+{\bf Mission.}
 WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
 
-{\bf Vision}
+{\bf Vision.}
 We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
 
-{\bf Focus areas}
+{\bf Focus areas.}
 WSSSPE promotes sustainable research software by positively impacting:
@@
 software engineers in academia. The survey provides demographic information
-about the education, field, gender, job satisfaction and career plans of the people
+about the education, field, gender, job satisfaction, and career plans of the people
 in this community. The outcome is that members of the community  were found to be
@@
 to compare research software. The author describes a proposed framework for doing
-so (based around measures of accessibility, usability, maintainability and
+so (based around measures of accessibility, usability, maintainability, and
 portability) which takes into account the possibility of variation between
@@
     Working Group and the principles of software citation established by the group.
-    Niemeyer, one of the working group cochairs, also discussed the paper published
+    Niemeyer, one of the working group co-chairs, also discussed the paper published
     describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
@@
 To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
-These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software.
-Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers.
-Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business.
-Finally, Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
+These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software;
+Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
+Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
+Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
 
 The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that has not crossed the Atlantic.
-After translating this question into the more open ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
+After translating this question into the more open, ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
 Hence, the panel were going to have to deal with questions from a highly experienced audience.
@@
 The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}.
-In each group's discussion, a number of Specific, Measurable, Achievable, Relevant and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.
+In each group's discussion, a number of Specific, Measurable, Achievable, Relevant, and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.
 In many cases, this was done through Slack (see \S\ref{sec:slack}.)

### 2017-04-09 WSSSPE4_report/wg_codemeta.tex

@@
 
-Research software is often not shared; that that is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides.
-CodeMeta wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software and increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
+Research software is often not shared; that which is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides.
+CodeMeta wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software, increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
 

### 2017-04-10 WSSSPE4_report/WSSSPE4_report.tex
adding comments for @kyleniemeyer

@@
     include automatic test generation based on user-supplied properties, dependency
-    specifications, and software contracts to check (e.g.) expected value ranges
+    specifications, and software contracts to check (e.g.) \todo{Kyle, should something be here?}expected value ranges
     and program behavior.
@@
     performance speedup of 320 times. The project plans to continue improving
-    the development process with (e.g.) static code checkers and Git commit hooks,
+    the development process with (e.g.) \todo{Kyle, again an e.g. with no example?} static code checkers and Git commit hooks,
     continued architecture redesign, and verification.

### 2017-04-10 WSSSPE4_report/wg_metrics.tex
fixing quotes

@@
 This gap delays workflow and creates a multitude of tedious tasks, including searching for the best-fit software, unknowingly duplicating software, and other related busy-work.
-The need for metric standardization stems from the abundance of scientific, both ?dark?
-and open source, software and the difficulties of ensuring the software is sustainable.
+The need for metric standardization stems from the abundance of scientific, both `dark' and open source, software and the difficulties of ensuring the software is sustainable.
 
@@
 
-The SMART steps Meaningful Metrics for Scientific Software has proposed are:
+The SMART steps Meaningful Metrics for Scientific Software proposed are:
 \begin{enumerate}
@@
 \item Define the interview questions and organize them into categories
-\item Interview the participants and map the survey results on goals
+\item Interview the participants and map the survey results to goals
 \item Convert the goals to feasible, meaningful metrics

### 2017-04-10 WSSSPE4_report/WSSSPE4_report.tex
resolving comments from Dan

--- 2017-04-09 WSSSPE4_report/WSSSPE4_report.tex
@@
     include automatic test generation based on user-supplied properties, dependency
-    specifications, and software contracts to check (e.g.) expected value ranges
+    specifications, and software contracts to check, e.g., expected value ranges
     and program behavior.
@@
     performance speedup of 320 times. The project plans to continue improving
-    the development process with (e.g.) static code checkers and Git commit hooks,
+    the development process with, e.g., static code checkers and Git commit hooks,
     continued architecture redesign, and verification.

### 2017-04-11 WSSSPE4_report/WSSSPE4_report.tex
fixing ñ in name

@@
 Neil~Chue~Hong$^{(8)}$,
-Santiago~Nunez-Corrales$^{(9)}$,
+Santiago~Nu\~nez-Corrales$^{(9)}$,
 Alice~Allen$^{(10)}$,

### 2017-04-11 WSSSPE4_report/participants.tex
updating names

\href{mailto:mjab@kth.se}{Mark Abraham} & SciLifeLab, KTH, Sweden\\
\href{mailto:Aseel.aldabjan@student.manchester.ac.uk}{Aseel Aldabjan} & University of Manchester\\
\href{mailto:aallen@ascl.net}{Alice Allen} & Astrophysics Source Code Library\\
\href{mailto:gdallen@illinois.edu}{Gabrielle Allen} & NCSA/University of Illinois\\
\href{mailto:pau.andrio@bsc.es}{Pau Andrio} & Barcelona Supercomputing Center\\
\href{mailto:bangerth@colostate.edu}{Wolfgang Bangerth} & Colorado State University\\
\href{mailto:radovan.bast@uit.no}{Radovan Bast} & UiT The Arctic University of Norway, Nordic e-Infrastructure Collaboration\\
\href{mailto:roman.bauer@newcastle.ac.uk}{Roman Bauer} & Newcastle University\\
\href{mailto:sbrandt@cct.lsu.edu}{Steven Brandt} & Louisiana State University\\
\href{mailto:lukas.johannes.breitwieser@cern.ch}{Lukas Johannes Breitwieser} & CERN\\
\href{mailto:echen35@illinois.edu}{Emily Chen} & NCSA/University of Illinois Urbana--Champaign\\
\href{mailto:childers@pitt.edu}{Bruce Childers} & University of Pittsburgh\\
\href{mailto:M.Croucher@Sheffield.ac.uk}{Mike Croucher} & University of Sheffield\\
\href{mailto:michael.crusoe@gmail.com}{Michael R. Crusoe} & Common Workflow Language Project\\
\href{mailto:stephan.druskat@hu-berlin.de}{Stephan Druskat} & Humboldt-Universit\"{a}t zu Berlin\\
\href{mailto:adubey@anl.gov}{Anshu Dubey} & Argonne National Laboratory\\
\href{mailto:tim.dunne@knowinnovation.com}{Tim Dunne} & KnowInnovation\\
\href{mailto:Peter.Elmer@cern.ch}{Peter Elmer} & Princeton University\\
\href{mailto:iain.emsley@oerc.ox.ac.uk}{Iain Emsley} & University of Oxford\\
\href{mailto:fangohr@soton.ac.uk}{Hans Fangohr} & University of Southampton\\
\href{mailto:stuart@stuartgeiger.com}{Stuart Geiger} & Berkeley Institute for Data Science\\
\href{mailto:sandra.gesing@nd.edu}{Sandra Gesing} & University of Notre Dame\\
\href{mailto:carole.goble@manchester.ac.uk}{Carole Goble} & University of Manchester, Software Sustainability Institute\\
\href{mailto:goodale@gmail.com}{Tom Goodale} & Independent\\
\href{mailto:gwilliamsc@cardiff.ac.uk}{Christopher Gwilliams} & Cardiff University\\
\href{mailto:w.vanhage@esciencecenter.nl}{Willem van Hage} & Netherlands eScience Center\\
\href{mailto:robert.haine@manchester.ac.uk}{Robert Haines} & University of Manchester\\
\href{mailto:j.hetherington@ucl.ac.uk}{James Hetherington} & University College London\\
\href{mailto:s.hettrick@software.ac.uk}{Simon Hettrick} & Software Sustainability Institute\\
\href{mailto:N.ChueHong@software.ac.uk}{Neil Chue Hong} & Software Sustainability Institute\\
\href{mailto:ljhwang@ucdavis.edu}{Lorraine Hwang} & UC Davis, CIG\\
\href{mailto:rayi@renci.org}{Ray Idaszak} & RENCI, University of North Carolina at Chapel Hill\\
\href{mailto:caroline.jay@manchester.ac.uk}{Caroline Jay} & University of Manchester\\
\href{mailto:bjimenez@bsc.es}{Brian Jimenez} & Barcelona Supercomputing Center\\
\href{mailto:rafael.jimenez@elixir-europe.org}{Rafael C Jimenez} & ELIXIR Hub, UK\\
\href{mailto:d.katz@ieee.org}{Daniel S Katz} & University of Illinois Urbana--Champaign\\
\href{mailto:dominic.kempf@iwr.uni-heidelberg.de}{Dominic Kempf} & Heidelberg University\\
\href{mailto:p.lago@vu.nl}{Patricia Lago} & Vrije Universiteit Amsterdam\\
\href{mailto:liu@hood.edu}{Xinlian Liu} & Hood College\\
\href{mailto:hannes.loeffler@stfc.ac.uk}{Hannes Loeffler} & STFC\\
\href{mailto:knarf@cct.lsu.edu}{Frank L\"{o}ffler} & Louisiana State University\\
\href{mailto:tengyu.ma@vanderbilt.edu}{Tengyu Ma} & Vanderbilt University\\
\href{mailto:rhmcdona@indiana.edu}{Robert H McDonald} & Indiana University\\
\href{mailto:chris.mentzel@moore.org}{Chris Mentzel} & Moore Foundation\\
\href{mailto:costa.michailidis@knowinnovation.com}{Constantinos Michailidis} & KnowInnovation\\
\href{mailto:jmiller@perimeterinstitute.ca}{Jonah Miller} & Perimeter Institute of Theoretical Physics\\
\href{mailto:mosse@cs.pitt.edu}{Daniel Mosse} & University of Pittsburgh\\
\href{mailto:august.muench@aas.org}{August (Gus) Muench} & American Astronomical Society\\
\href{mailto:unangi2@illinois.edu}{Udit Nangia} & University of Illinois Urbana--Champaign\\
\href{mailto:a.nenadic@manchester.ac.uk}{Aleksandra Nenadic} & University of Manchester, SSI\\
\href{mailto:Kyle.Niemeyer@oregonstate.edu}{Kyle Niemeyer} & Oregon State University\\
\href{mailto:nunezco2@illinois.edu}{Santiago N\'u\~nez-Corrales} & University of Illinois Urbana--Champaign\\
\href{mailto:d.a.orchard@kent.ac.uk}{Dominic Orchard} & University of Kent\\
\href{mailto:chico@puc-rio.br}{Francisco Queiroz} & Pontifical Catholic University of Rio de Janeiro\\
\href{mailto:fons.rademakers@cern.ch}{Fons Rademakers} & CERN\\
\href{mailto:karthik.ram@berkeley.edu}{Karthik Ram} & University of California, Berkeley\\
\href{mailto:samuel.relton@manchester.ac.uk}{Sam Relton} & University Of Manchester\\
\href{mailto:thomas.robitaille@gmail.com}{Tom Robitaille} & Freelance\\
\href{mailto:jeansalac@virginia.edu}{Jean Salac} & University of Virginia\\
\href{mailto:janos.sallai@vanderbilt.edu}{Janos Sallai} & Vanderbilt University\\
\href{mailto:prabhjyot10@gmail.com}{Prabhjyot (Singh) Saluja} & Southwestern Oklahoma State University\\
\href{mailto:eseidel@cs.ucsd.edu}{Eric Seidel} & University of California, San Diego\\
\href{mailto:sameer@cs.uoregon.edu}{Sameer Shende} & University of Oregon\\
\href{mailto:raniere.silva@software.ac.uk}{Raniere Silva} & University of Manchester\\
\href{mailto:shoaib.sufi@software.ac.uk}{Shoaib Sufi} & University of Manchester, Software Sustainability Institute\\
\href{mailto:chris.sweet@nd.edu}{Chris Sweet} & University of Notre Dame\\
\href{mailto:py12rw@leeds.ac.uk}{Rob Welch} & University of Leeds\\
\href{mailto:Christopher.Woods@bristol.ac.uk}{Christopher Woods} & University of Bristol\\

### 2017-04-11 WSSSPE4_report/WSSSPE4_report.tex
name fix

--- 2017-04-10 WSSSPE4_report/WSSSPE4_report.tex
@@
 Neil~Chue~Hong$^{(8)}$,
-Santiago~Nunez-Corrales$^{(9)}$,
+Santiago~N\'u\~nez-Corrales$^{(9)}$,
 Alice~Allen$^{(10)}$,

### 2017-04-11 WSSSPE4_report/program_committee.tex
more name fix

David Abramson & University of Queensland, Australia\\
Lorena A. Barba & George Washington University, USA\\
Ross Bartlett & Sandia National Laboratories, USA\\
Christoph Becker & University of Toronto, Canada\\
Ewout van den Berg & IBM Watson, USA\\
David Bernholdt & Oak Ridge National Laboratory, USA\\
Stefanie Betz & Karlsruhe Institute of Technology, Germany\\
Coral Calero & Universidad Castilla La Mancha, Spain\\
Ishwar Chandramouli & National Cancer Institute, National Institutes of Health, USA\\
Ruzanna Chitchyan & University of Leicester, UK\\
Karen Cranston & Duke University, USA\\
Ewa Deelman & Information Sciences Institute, University of Southern California, USA\\
Charlie E. Dibsdale & O-Sys, Rolls Royce PLC, UK\\
Anshu Dubey & Argonne National Laboratory, USA\\
Nadia Eghbal & Independent Researcher (via Ford Foundation), USA\\
Peter Elmer & CERN, Switzerland\\
Martin Fenner & DataCite, Germany\\
David Gavaghan & University of Oxford, UK\\
Mike Glass & Sandia National Laboratories, USA\\
Carole Goble & University of Manchester, UK\\
Joshua Greenberg & Alfred P. Sloan Foundation, USA\\
Michael K Griffiths & University of Sheffield, UK\\
Sarah Harris & University of Leeds, UK\\
James Hetherington & University College London, UK\\
Fred J. Hickernell & Illinois Institute of Technology, USA\\
Neil Chue Hong & Software Sustainability Institute, University of Edinburgh, UK\\
Caroline Jay & University of Manchester, UK\\
Rafael C. Jimenez & ELIXER, Cambridge, UK\\
Matthew B. Jones & University of California Santa Barbara, USA\\
Nick Jones & New Zealand eScience Infrastructure (NeSI), NZ\\
Sedef Akinli Kocak & Ryerson University, Canada\\
Jong-Suk Ruth Lee & National Institute of Supercomputing and Networking, KISTI, Korea\\
James Lin & Shanghai Jiao Tong University, China\\
Frank Löffler & Louisiana State University, USA\\
Gregory Madey & University of Notre Dame, USA\\
Ketan Maheshwari & University of Pittsburgh, USA\\
Steven Manos & University of Melbourne, Australia\\
Chris A. Mattmann & NASA JPL; University of Southern California, USA\\
Abigail Cabunoc Mayes & Mozilla Science Lab, USA\\
Robert H. McDonald & Indiana University, USA\\
Lois Curfman McInnes & Argonne National Laboratory, USA\\
Alberto Di Meglio & CERN, Switzerland\\
Chris Mentzel & Gordon and Betty Moore Foundation, USA\\
Peter Murray-Rust & University of Cambridge, UK\\
Christopher R. Myers & Cornell University, USA\\
Jarek Nabrzyski & University of Notre Dame, USA\\
Cameron Neylon & Curtin University, Australia\\
Aleksandra Pawlik & New Zealand eScience Infrastructure (NeSI), NZ\\
Fernando Perez & Lawrence Berkeley National Laboratory; University of California, Berkeley, USA\\
Marian Petre & The Open University, UK\\
Marlon Pierce & Indiana University, USA\\
Andreas Prlic & University of California, San Diego, USA\\
Karthik Ram & University of California, Berkeley, USA\\
Morris Riedel & Juelich Supercomputing Centre, Germany\\
Dave De Roure & Oxford e-Research Centre, University of Oxford, UK\\
Norbert Seyff & University of Zurich, Switzerland\\
Arfon Smith & GitHub Inc, USA\\
Borja Sotomayor & University of Chicago, USA\\
Edgar Spalding & University of Wisconsin, USA\\
Maria Spichkova & RMIT University, Australia\\
Victoria Stodden & University of Illinois Urbana--Champaign, USA\\
Matthew Turk & University of Illinois Urbana--Champaign, USA\\
Nancy Wilkins-Diehr & San Diego Supercomputer Center, University of California, San Diego, USA\\
James Willenbring & Sandia National Laboratories, USA\\
Scott Wilson & Cetis LLP, UK\\
Theresa Windus & Iowa State University and Ames Laboratory, USA\\

### 2017-04-17 WSSSPE4_report/WSSSPE4_report.tex
some misc cleanup/consistency fixes

@@
 \makesavenoteenv{table}
+
+\usepackage{placeins}
 
@@
 Hence, the total number of unique individuals responding to the survey is likely 44.
-The survey contained 9 multiple choice and open-text answer questions.
+The survey contained nine multiple choice and open-text answer questions.
 Text responses have been alphabetize to preserve anonymity.
@@
 The main problem seems to be one of time.  Once the attendees have agree to spend
-their time at the workshop, they put their energy into doing so productively, but they haven't
+their time at the workshop, they put their energy into doing so productively, but they have not
 really committed themselves to anything more than this, so their energy and effort trails

### 2017-04-17 WSSSPE4_report/survey_details.tex
some misc cleanup/consistency fixes

@@
 
-\hspace{0.2cm}
-
-\noindent \textbf{1. Please indicate how strongly you agree or disagree with the following statements}
+\begin{enumerate}
+\item \textbf{Please indicate how strongly you agree or disagree with the following statements}
+%\noindent \textbf{1. Please indicate how strongly you agree or disagree with the following statements}
 
@@
 
-\noindent \textbf{2. What part of the meeting was the most valuable? select all that apply}
+%\noindent \textbf{2. What part of the meeting was the most valuable? select all that apply}
+\item \textbf{What part of the meeting was the most valuable? select all that apply}
 
@@
 
-\newpage
-\noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}
+%\newpage
+%\noindent \textbf{3. What additional topics would you like to see discussed at the next meeting?}
+\item \textbf{What additional topics would you like to see discussed at the next meeting?}
 
@@
 
-\noindent \textbf{4. What could be improved?}
+%\noindent \textbf{4. What could be improved?}
+\item \textbf{What could be improved?}
 
@@
 
-\noindent \textbf{5. What would make you more likely to invest time in future WSSSPE initiatives?}
+%\noindent \textbf{5. What would make you more likely to invest time in future WSSSPE initiatives?}
+\item \textbf{What would make you more likely to invest time in future WSSSPE initiatives?}
 
@@
 
-\noindent \textbf{6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}
+%\noindent \textbf{6. Would you be interested in joining a professional society focused around the topics of WSSSPE?}
+\item \textbf{Would you be interested in joining a professional society focused around the topics of WSSSPE?}
 
@@
 
-\newpage
-\noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
+%\newpage
+%\noindent \textbf{7. Will you be part of a working group  with colleagues going forward  on a topic addressed at WSSSPE, and if so, which one?}
+\item \textbf{Will you be part of a working group with colleagues going forward on a topic addressed at WSSSPE, and if so, which one?}
 
@@
 
-\noindent \textbf{8. Working Group Name}
+\FloatBarrier
+%\noindent \textbf{8. Working Group Name}
+\item \textbf{Working group name}
 
-\vspace{0.2cm}
+%\vspace{0.2cm}
 
@@
 
-\noindent \textbf{9. Do you have any other comments or suggestions or ideas?}
+%\noindent \textbf{9. Do you have any other comments or suggestions or ideas?}
+\item \textbf{Do you have any other comments or suggestions or ideas?}
 
@@
 \end{itemize}
+
+\end{enumerate}

### 2017-04-17 WSSSPE4_report/wg_access.tex
some misc cleanup/consistency fixes

@@
 
-Note that Debian provides a testing infrastructure (\href{https://ci.debian.net/doc/}{http://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.
+Note that Debian provides a testing infrastructure (\url{https://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.
 
@@
 \item Learn how to donate machines to Debian
-\item Find funding, perhaps via NEIC? (\url{https://wiki.neic.no/wiki/NeIC_Community_Wiki}) 
+\item Find funding -- NeIC (\url{https://neic.nordforsk.org}) through the CodeRefinery project (\url{http://coderefinery.org}) is building a CI infrastructure (in progress) that will be provided to selected projects (through an open call)
 \item Build community

### 2017-04-17 WSSSPE4_report/wg_alliance.tex
some misc cleanup/consistency fixes

@@
 \item Come up with a list of potential invitees
-\item Update www.softwaresustainability.org to provide information
+\item Update \url{www.softwaresustainability.org} to provide information
 \item Draft a letter to go out to potential alliance members
@@
 
-If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (N.ChueHong@software.ac.uk) and Jean Salac (jeansalac@virginia.edu)
+If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)

### 2017-04-17 WSSSPE4_report/wg_best_practices_developing.tex
some misc cleanup/consistency fixes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{White paper on developing sustainable softwares}
+\subsection{White paper on developing sustainable software}
 \label{sec:best-practices-developing}
@@
 
-Introduction to group here, including the overall objective of work in this area.
+\note{Introduction to group here, including the overall objective of work in this area.}
 
@@
 
-members of the working group
+\note{members of the working group}
 
@@
 
-Specific things the working group wants to accomplish in the context of the larger objective.
+\note{Specific things the working group wants to accomplish in the context of the larger objective.}
 
@@
 
-What tasks will the working group undertake
+\note{What tasks will the working group undertake}
 
@@
 
-The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}. For more information and requests, join \#wg-best-practices channel at WSSSPE's Slack team.
+The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
+For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.

### 2017-04-17 WSSSPE4_report/wg_best_practices_undergrads.tex
some misc cleanup/consistency fixes

@@
 undergraduate students studying domain science. The program might be
-similar to a software carpentry or data carpentry workshop but with a
+similar to a Software Carpentry or Data Carpentry workshop but with a
 focus for domain scientists.
@@
 \noindent Aleksandra Nenadic and Raniere Silva are involved
-in software and data carpentry.
+in Software and Data Carpentry.
 
@@
 \begin{itemize}
-\item Join our google group/mailing list
+\item Join our Google Group/mailing list
   \cite{WSSSPEUndergradGoogleGroup}.

### 2017-04-17 WSSSPE4_report/wg_codemeta.tex
some misc cleanup/consistency fixes

@@
 This group, which included people who had previously been working on the CodeMeta project\footnote{\url{http://codemeta.github.io/}}, formed to address the proposed project,
- ``Define a standard for metadata to be used in current repos including authors, dependencies, ... repo url.''
+ ``Define a standard for metadata to be used in current repos including authors, dependencies, \dots{}, repo url.''
 

### 2017-04-17 WSSSPE4_report/wg_letters.tex
some misc cleanup/consistency fixes

@@
   \item Write editorial for SIAM (Wolfgang)
-  \item Write 2 emails to each other group members listed above about
+  \item Write two emails to each other group members listed above about
     also writing editorials in their communities.
@@
 to affect change is important; contact Wolfgang Bangerth at
-\url{bangerth@colostate.edu} if you are interested in helping.
+<bangerth@colostate.edu> if you are interested in helping.

### 2017-04-17 WSSSPE4_report/wg_metrics.tex
some misc cleanup/consistency fixes

@@
 This gap delays workflow and creates a multitude of tedious tasks, including searching for the best-fit software, unknowingly duplicating software, and other related busy-work.
-The need for metric standardization stems from the abundance of scientific, both `dark' and open source, software and the difficulties of ensuring the software is sustainable.
+The need for metric standardization stems from the abundance of scientific, both ``dark'' and open source, software and the difficulties of ensuring the software is sustainable.
 

### 2017-04-17 WSSSPE4_report/wg_open_research_index.tex
fixing typo
some misc cleanup/consistency fixes

@@
 The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
-Our goal is not to eliminate commercial products, but to build on what's there and provide data and services that are missing.
+Our goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.
 
@@
 \begin{itemize}
-\item ORCID, CrossRef, DataCite, ImpactStory/depsy, altmetrics.org, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, softwareheritage.org, DBLP, eSTEP (CWO)
+\item ORCID, CrossRef, DataCite, ImpactStory/Depsy, \url{altmetrics.org}, Plum Analytics, GitHub, Open AIRE, CHORUS, FORCE11, COS, SHARE, CASRAI, Portico, \url{softwareheritage.org}, DBLP, eSTEP (CWO)
 \item We could work with some initial publishers who don't have competing services (domain societies), PubMed
@@
 Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
-At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \#wg-open-research-idx.
+At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}.
 (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
@@
 \begin{enumerate}
-\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people don't like, we could check what features they has and what the issues are.
+\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, we could check what features they has and what the issues are.
 
@@
 \item What model should be used?
-- Crawling and obtaining data
-- Consortium model
-- Need analysis and decision
+\begin{itemize}
+    \item Crawling and obtaining data
+    \item Consortium model
+    \item Need analysis and decision
+\end{itemize}
 
 \item Need agreements with publishers, repositories, services
-- Some of these will have costs, though some may be free
-- Need to determine costs
-- Perhaps publishers would pay us to be listed eventually, as we gain power
-- To some extent, we would be competing with some of the publishers? products
+\begin{itemize}
+    \item Some of these will have costs, though some may be free
+    \item Need to determine costs
+    \item Perhaps publishers would pay us to be listed eventually, as we gain power
+    \item To some extent, we would be competing with some of the publishers? products
+\end{itemize}
 
 \item If crawling model:
-- How to crawl all of these products?
-- What data (metadata) should be the output of the crawls?
-    - What happens when the products being crawled and/or their metadata change?
-- How to store all of this data (metadata)?
+\begin{itemize}
+    \item How to crawl all of these products?
+    \item What data (metadata) should be the output of the crawls? What happens when the products being crawled and/or their metadata change?
+    \item How to store all of this data (metadata)?
+\end{itemize}
 
@@
 \item Discuss this idea with potential partners (who, when).
-Then hgt their feedback \& buy-in.
+Then get their feedback \& buy-in.
 While doing so, seek suggestions for the right project manager to coordinate this
@@
 
-If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \#wg-open-research-idx.
+If you are interested in more information, please join the WSSSPE team's Slack channel for this working group: \texttt{\#wg-open-research-idx}.
 (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

### 2017-04-17 WSSSPE4_report/wg_prototyping.tex
some misc cleanup/consistency fixes

@@
 aspect of scientific software sustainability. In essence,
-a motto for scientific software prototyping is {\it fail hard, fail fast}.
+a motto for scientific software prototyping is \textit{fail hard, fail fast}.
 
@@
 in order to work around the capabilities of the tool. In addition, use cases
-are required (e.g. micromagnetics~\cite{fischbacher2007systematic}).
+are required (e.g., micromagnetics~\cite{fischbacher2007systematic}).
 

### 2017-04-28 WSSSPE4_report/WSSSPE4_report.tex
removing author notes, other small fixes

@@
 Women in HPC\footnote{\url{http://www.womeninhpc.org}}.
-\todo{did we reach out to any others?}
+%\todo{did we reach out to any others?}
 
@@
 
-\note{from Patricia}
+%\note{from Patricia}
 
@@
 %
-A \textsuperscript{\textasteriskcentered} indicates which author presented the
-lightning talk, for those with multiple authors.
+For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
 %
@@
 
-\note{from Simon}
+%\note{from Simon}
 

### 2017-04-28 WSSSPE4_report/wg_codemeta.tex
removing more author name notes

--- 2017-04-09 WSSSPE4_report/wg_codemeta.tex
@@
 
-\note{Alice to write this}
+%\note{Alice to write this}
 
 This group, which included people who had previously been working on the CodeMeta project\footnote{\url{http://codemeta.github.io/}}, formed to address the proposed project,
- ``Define a standard for metadata to be used in current repos including authors, dependencies, ... repo url.''
+ ``Define a standard for metadata to be used in current repos including authors, dependencies, \dots{}, repo URL.''
 

### 2017-04-28 WSSSPE4_report/wg_alliance.tex
removing author notes, other small fixes
changing to present tense, third person

@@
 
-\note{Neil and Jean wrote this}
+%\note{Neil and Jean wrote this}
 
@@
 Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
-We are currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
+It is currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
 
@@
 More specifically, this working group aims to define the scope of this alliance and provide clear distinction with WSSSPE.
-We also seek to understand the incentives for members of this alliance and identify its key activities.
+The group also seeks to understand the incentives for members of this alliance and identify its key activities.
 
@@
 
-Our working group is looking for ideas for which organizations should be consulted or invited to join the Software Sustainability Alliance.
-We are also looking for feedback on the aims and scope for joining the Software Sustainability Alliance.
+The working group is looking for ideas for which organizations should be consulted or invited to join the Software Sustainability Alliance.
+It is also looking for feedback on the aims and scope for joining the Software Sustainability Alliance.
 
@@
 
-Throughout the scoping period, our working group will continue to refine the Software Sustainability Alliance as we collect feedback on potential members, aims and scopes.
-We will also update our website to reflect these changes.
+Throughout the scoping period, the working group will continue to refine the Software Sustainability Alliance as it collects feedback on potential members, aims and scopes.
+It will also update its website to reflect these changes.
 Next steps and further tasks will be determined based on the feedback from the scoping period.
@@
 
-The first SMART steps of our working group are outlined below:
+The first SMART steps of the working group are outlined below:
 \begin{itemize}

### 2017-04-28 WSSSPE4_report/wg_best_practices_sustainability.tex
removing author notes, other small fixes
changing to present tense, third person

@@
 
-\note{Ray to write this}
+%\note{Ray to write this}
 
@@
 Many open source projects also document software metrics they use to define their research software project's success.
-This project endeavors to aggregate several sources of these best practices into a consolidated list of best practices, and similar for software metrics.
-Our team will then create a workflow to evaluate how open source projects stand up to these lists.
+This group's goal is to aggregate several sources of these best practices into a consolidated list of best practices, and similar for software metrics.
+The team then plans to create a workflow to evaluate how open source projects stand up to these lists.
 
@@
 The overall objective of the Verifying Best Practices and Metrics for Sustainable Software working group is to take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
-This will allow us to identify gaps on both sides.
+This will allow the group to identify gaps on both sides.
 This approach will also allow the group to  hypothesize how successful open source projects can be further improved, verify that recommended approaches for software engineering for science/academia are sufficient and valid, and that metrics for software engineering for science/academia are relevant and useful.
@@
 
-Our plans are for each of the Verify Best Practices and Metrics for Sustainable Research Software to volunteer to take on one to two projects each, and as incentive the projects that individual take on can be one they are interested in.
+The group's plans are for each of group members to volunteer to take on one to two projects.
+As an incentive, the projects that each individual takes on should be ones in which they are already interested.
 This would lead to an estimated 5--10 projects to be evaluated within this project.
-In terms of increasing this sample size, it is our hope that once we document the workflow by which these evaluations are performed, others can be self-sufficient in following this workflow and contribute their own evaluations of software they wish to evaluate and have interest in.
+In terms of increasing this sample size, the group hopes that once it documents the workflow by which these evaluations are performed, others can be self-sufficient in following this workflow and contribute their own evaluations of software they wish to evaluate.
 
@@
 \begin{itemize}
-\item Create mailing list for our project titled:  Verify Best Practices and Metrics for Sustainable Research Software.
+\item Create mailing list for project titled:  Verify Best Practices and Metrics for Sustainable Research Software.
 \item Identify where to obtain representative recommended software engineering best practices and metrics within the WSSSPE4 community including single points-of-contact.

### 2017-04-28 WSSSPE4_report/wg_codemeta.tex
tense and person changes

@@
 
-The primary objective of this working group was to find ways to help the CodeMeta project come to fruition.
+The primary objective of this working group is to find ways to help the CodeMeta project come to fruition.
 CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
@@
 
-Our plans are for the working group to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
+The working group plans to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
 Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation.
@@
 
-With a comparatively large working group, the use of the established CodeMeta Github repository's issue tracking, and the quick responsiveness of CodeMeta lead Matt Jones to a barrage of questions, comments, and logged issues, several of our SMART steps were completed at WSSSPE; these are identified below in italics.
+With a comparatively large working group, the use of the established CodeMeta Github repository's issue tracking, and the quick responsiveness of CodeMeta lead Matt Jones to a barrage of questions, comments, and logged issues, several of the group's SMART steps were completed at WSSSPE; these are identified below in italics.
 

### 2017-04-28 WSSSPE4_report/wg_letters.tex
tense and person changes

@@
 
-We believe that the authors of scientific software provide important
+This group believes that the authors of scientific software provide important
 services to departments that are no less than strictly disciplinary
@@
 through letters to the community, adjusting definitions of what counts
-in science, etc. Specifically, we are aware of the following
+in science, etc. Specifically, the group is aware of the following
 resources:
@@
 
-We will attempt to address it by organizing sessions at conferences
+The group will attempt to address it by organizing sessions at conferences
 that address the career problem, as well as writing editorials that
-can be published in the magazines of our professional societies. These
+can be published in the magazines of professional societies. These
 editorials ought to outline best practices for letter writers that
@@
 
-We have identified a few next steps, even if they may not satisfy the
+The group has identified a few next steps, even if they may not satisfy the
 exact criteria of the SMART system. Specifically:
@@
 \end{itemize}
-In the longer term we hope to gather others in our home disciplines to
+In the longer term, the group hopes to gather others in the home disciplines of its members to
 let professional organizations weigh in.

### 2017-04-28 WSSSPE4_report/wg_open_research_index.tex
tense and person changes

@@
 The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
-Our goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.
+Its goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.
 
@@
 
-Alternatively or additionally, we could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
-At WSSSPE4, there seemed to be enough interest to do this, so we set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}.
+Alternatively or additionally, the group could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
+At WSSSPE4, there seemed to be enough interest to do this, so the group set up a Slack channel within the WSSSPE team, called \texttt{\#wg-open-research-idx}.
 (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)
@@
 \begin{enumerate}
-\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, we could check what features they has and what the issues are.
+\item Look at curriculum lattes (Brazil) and/or Researchfish (UK).  They have some problems that people do not like, the group could check what features they have and what the issues are.
 
@@
 
-\item What research products should we track?
+\item What research products should the index track?
 
@@
     \item Perhaps publishers would pay us to be listed eventually, as we gain power
-    \item To some extent, we would be competing with some of the publishers? products
+    \item To some extent, the group would be competing with some of the publishers' products
 \end{itemize}
@@
 
-\item Apply for initial funding. For the US, we discussed the Arnold Foundation Open Science award as a planning activity (who, due Dec 15), though without a PI, this seems unlikely to be done.
-For the UK, we didn't have a suggestion.
+\item Apply for initial funding. For the US, the group discussed the Arnold Foundation Open Science award as a planning activity (who, due Dec 15), though without a PI, this seems unlikely to be done.
+For the UK, there wasn't a specific suggestion.
 

### 2017-04-28 WSSSPE4_report/wg_prototyping.tex
tense and person changes

@@
 
-Our general plan is to develop a functional (meta-)prototype before WSSSPE5, and
+The group's general plan is to develop a functional (meta-)prototype before WSSSPE5, and
 if useful, depart from it towards a larger software infrastructure tied to

### 2017-04-28 WSSSPE4_report/wg_social.tex
updating participants lists in working groups to be consistent
removing more author name notes
tense and person changes

@@
 
-\note{from Stuart}
+%\note{from Stuart}
 
@@
 
-This working group was motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability.
+This working group is motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability.
 In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges.
@@
 
-Stuart Geiger <stuart@stuartgeiger.com>, Lorraine Hwang <ljhwang@ucdavis.edu>, Robert McDonald <rhmcdona@indiana.edu>
+\begin{itemize}
+  \item Stuart Geiger <stuart@stuartgeiger.com>
+  \item Lorraine Hwang <ljhwang@ucdavis.edu>
+  \item Robert McDonald <rhmcdona@indiana.edu>
+\end{itemize}
 
@@
 %What tasks will the working group undertake
-The working group planned to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
+The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
 This plan resulted in the creation of the following SMART steps listed in the next section.

### 2017-04-28 WSSSPE4_report/wg_soft_eng.tex
updating participants lists in working groups to be consistent
some misc cleanup/consistency fixes
tense and person changes

@@
 %Introduction to group here, including the overall objective of work in this area.
-This working group concerned itself with identifying processes that are not adequately
+This working group is concerned with identifying processes that are not adequately
 covered by general software engineering. Verification and testing is
-the first subtopic that we address.  
+the first subtopic that it addresses.
 % there is a growing awareness about the lack of trust
@@
 \begin{itemize}
-\item Mark Abraham
-\item Anshu Dubey 
-\item Hans Fangohr
-\item Dominic Kempf
-\item Eric Seidel
+\item Mark Abraham <mjab@kth.se>
+\item Anshu Dubey <adubey@anl.gov>
+\item Hans Fangohr <fangohr@soton.ac.uk>
+\item Dominic Kempf <dominic.kempf@iwr.uni-heidelberg.de>
+\item Eric Seidel <eseidel@cs.ucsd.edu>
 \end{itemize}
@@
 solutions. However, this literature is scattered among many different
-forums, and can be challenging to find. Our working group aims to
+forums, and can be challenging to find. This working group aims to
 address this gap by curating the existing content and contributing
@@
 as many as possible. The seed resources required are fairly
-minimal. We have started a git repository for collecting the
+minimal. The group has started a git repository for collecting the
 existing references. That, and a few volunteers reading through the
-references is all we need in the beginning. As we gather more
-knowledge and pinpoint gaps, we may need more resources to reach out
+references is all the group needs in the beginning. As it gathers more
+knowledge and pinpoint gaps, it may need more resources to reach out
 to a wider group of developers to collect more information.
@@
 \item Review and summarize material. Decide whether we consider this
-  sufficient. If yes, then we will put brief report together and
+  sufficient. If yes, then the group will put a brief report together and
   conclude the working group.
-\item If we do not consider the material adequate, we start research
+\item If the group does not consider the material adequate, it will start research
   and gather methodologies for testing in science
 \item Write a document accessible to computational science and
-  software engineering community, and publish document at a citeable forum.
+  software engineering community, and publish document at a citable forum.
 \end{itemize}
@@
 has a channel in the WSSSPE Slack. The channel is called
-{\it wg-testing-in-science}.
+\texttt{\#wg-testing-in-science}.
 (See \S\ref{sec:slack} for instructions on how to join the Slack WSSSPE team.)

### 2017-05-02 WSSSPE4_report/wg_best_practices_sustainability.tex
trimmed people section in best practices/sustainability section

@@
 
-The people spearheading this effort at the time of the writing of this workshop report are Ray Idaszak <rayi@renci.org>, August Muench <august.muench@aas.org>, Jonah Miller <jmiller@perimeterinstitute.ca>, Lorraine Hwang <ljhwang@ucdavis.edu>, Peter Elmer <peter.elmer@cern.ch>, and Hans Fangohr <fangohr@soton.ac.uk>.
+The people spearheading this effort at the time of the writing of this workshop report are the
+working group participants listed previously along with Peter Elmer <peter.elmer@cern.ch> and Hans Fangohr <fangohr@soton.ac.uk>.
 
-\noindent
 Resources include research projects suggested from WSSSPE community including:

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
removing more author name notes
Better explain the purpose of the CoC.

@@
 
-This workshop also introduced a Code of Conduct (CoC)
+At the workshop, we also introduced a Code of Conduct (CoC).%
 \footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
-for both WSSSPE4 and interactions in the greater WSSSPE community (e.g., on
+The CoC was conceived for the workshop itself; however, we intend it
+to also be a guideline for the community of scientists which WSSSPE
+supports, and their personal and online interactions (e.g., on
 Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
@@
 
-\note{from Lorraine}
+%\note{from Lorraine}
 

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
Fix a typo.
Use standard colors for links and references.
More random edits in the first few sections.

--- 2017-04-28 WSSSPE4_report/WSSSPE4_report.tex
@@
 
-\usepackage[bookmarks, bookmarksopen, bookmarksnumbered]{hyperref}
+\usepackage[bookmarks, bookmarksopen, bookmarksnumbered, colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
 \usepackage[all]{hypcap}
@@
 covers the set of working groups that formed during the meeting, and for each,
-discusses the participant, the objective and goal, and how the objective can
+discusses the participants, the objective and goal, and how the objective can
 be reached, along with contact information for readers who may want to join
@@
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with SC13;
+held in conjunction with the Supercomputing 2013 (SC13) conference;
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
@@
 
-WSSSPE4 was based on the work done in WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
+WSSSPE4 was based on the work done at WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
 at producing working groups that better continued working after the workshop ended.
@@
     has the goal of improving the quality of today's research software and the
-    experiences of its developers by sharing practices and experiences.
+    experiences of its developers by sharing best practices and experiences.
     This track is focused on the current state of scientific software and what
@@
 The call for participation requested
-idea papers, which present implementable proposals related to Track 1;
+idea papers, which present implementable proposals related to Track~1;
 position papers, which are longer, not previously published papers related to
-Track 2 specifically discussing what we can do to improve sustainable scientific
+Track~2 specifically discussing what we can do to improve sustainable scientific
 software in the short term, starting with where we are today;
-experience papers, longer papers related to Track 2 that discuss current
+experience papers, longer papers related to Track~2 that discuss current
 practices and experiences and how they have been used to improve the quality of
@@
 demos, brief papers describing a tool or
-process that would be demonstrated, relevant to Track 2 that improves the quality of today's research
+process that would be demonstrated, relevant to Track~2 that improves the quality of today's research
 software and\slash or the experiences of its developers; and
@@
 
-The following list of contribution topics was suggested, though this was not intended to limit potential submissions.
+The following list of contribution topics was suggested, though this was not intended to limit potential submissions:
 
@@
 %
-Sustainability is broadly associated with mother nature and ecologic issues.
+Sustainability is broadly associated with how humans use natural ecosystem services.
 When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}.
@@
 
-Supporting technical sustainability in (scientific) software is very near to supporting properties like longevity and resilience, i.e., the ability of a software system or application to fulfill its intent~\cite{Huisman2016} as planned independently from time passing or from changes in its operational context.
+Supporting technical sustainability in (scientific) software is very near to supporting properties like longevity and resilience, i.e., the ability of a software system or application to fulfill its intent~\cite{Huisman2016}, as planned independently from time passing or from changes in its operational context.
 Technical sustainability has never been easy, but it is even more of a challenge nowadays with the increasing demands on software solutions to support many aspects of our daily lives.
@@
 This provided a complementary angle to the workshop discussions.
-It addressed software energy efficiency (i.e., how to develop software that consumes less energy, like mobile apps that should consume less device-battery) and software energy-awareness (i.e., how to develop software supporting features that in turn have a lesser ecologic impact).
-
-The keynote used few myths to elaborate on such challenges and related research. Among them:
+It addressed software energy efficiency (i.e., how to develop software that consumes less energy, like mobile apps that should consume less device-battery) and software energy-awareness (i.e., how to develop software that can actively decide between energy efficiency and other goals such as performance).
+
+The keynote used a few myths to elaborate on such challenges and related research. Among them:
 

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
Remove a sentence that is not logical.
Specifically, it seems to suggest that science is important because, well, duh, science.

--- 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
@@
 This report is based on the events at the workshop and the submitted materials.
-There events included discussion of the WSSSPE mission and vision, a keynote,
+These events included discussion of the WSSSPE mission and vision, a keynote,
 a set of presentations, a panel, and a number of working groups.  One full day of the

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
undoing a change
SC is just SC, not Supercomputing.  The formal conference name is now very long...

--- 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
@@
 WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with the Supercomputing 2013 (SC13) conference;
+held in conjunction with the SC13 conference;
 WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
@@
 %
-Sustainability is broadly associated with how humans use natural ecosystem services.
+Sustainability is broadly associated with natural ecologic systems.
 When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}.

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex

@@
 
-While these points are valid in any type of software, scientific software is even more sensitive to sustainability intents, being in charge of supporting science.
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-05-03 WSSSPE4_report/participants.tex
turning off email of participants

@@
-\href{mailto:mjab@kth.se}{Mark Abraham} & SciLifeLab, KTH, Sweden\\
-\href{mailto:Aseel.aldabjan@student.manchester.ac.uk}{Aseel Aldabjan} & University of Manchester\\
-\href{mailto:aallen@ascl.net}{Alice Allen} & Astrophysics Source Code Library\\
-\href{mailto:gdallen@illinois.edu}{Gabrielle Allen} & NCSA/University of Illinois\\
-\href{mailto:pau.andrio@bsc.es}{Pau Andrio} & Barcelona Supercomputing Center\\
-\href{mailto:bangerth@colostate.edu}{Wolfgang Bangerth} & Colorado State University\\
-\href{mailto:radovan.bast@uit.no}{Radovan Bast} & UiT The Arctic University of Norway, Nordic e-Infrastructure Collaboration\\
-\href{mailto:roman.bauer@newcastle.ac.uk}{Roman Bauer} & Newcastle University\\
-\href{mailto:sbrandt@cct.lsu.edu}{Steven Brandt} & Louisiana State University\\
-\href{mailto:lukas.johannes.breitwieser@cern.ch}{Lukas Johannes Breitwieser} & CERN\\
-\href{mailto:echen35@illinois.edu}{Emily Chen} & NCSA/University of Illinois Urbana--Champaign\\
-\href{mailto:childers@pitt.edu}{Bruce Childers} & University of Pittsburgh\\
-\href{mailto:M.Croucher@Sheffield.ac.uk}{Mike Croucher} & University of Sheffield\\
-\href{mailto:michael.crusoe@gmail.com}{Michael R. Crusoe} & Common Workflow Language Project\\
-\href{mailto:stephan.druskat@hu-berlin.de}{Stephan Druskat} & Humboldt-Universit\"{a}t zu Berlin\\
-\href{mailto:adubey@anl.gov}{Anshu Dubey} & Argonne National Laboratory\\
-\href{mailto:tim.dunne@knowinnovation.com}{Tim Dunne} & KnowInnovation\\
-\href{mailto:Peter.Elmer@cern.ch}{Peter Elmer} & Princeton University\\
-\href{mailto:iain.emsley@oerc.ox.ac.uk}{Iain Emsley} & University of Oxford\\
-\href{mailto:fangohr@soton.ac.uk}{Hans Fangohr} & University of Southampton\\
-\href{mailto:stuart@stuartgeiger.com}{Stuart Geiger} & Berkeley Institute for Data Science\\
-\href{mailto:sandra.gesing@nd.edu}{Sandra Gesing} & University of Notre Dame\\
-\href{mailto:carole.goble@manchester.ac.uk}{Carole Goble} & University of Manchester, Software Sustainability Institute\\
-\href{mailto:goodale@gmail.com}{Tom Goodale} & Independent\\
-\href{mailto:gwilliamsc@cardiff.ac.uk}{Christopher Gwilliams} & Cardiff University\\
-\href{mailto:w.vanhage@esciencecenter.nl}{Willem van Hage} & Netherlands eScience Center\\
-\href{mailto:robert.haine@manchester.ac.uk}{Robert Haines} & University of Manchester\\
-\href{mailto:j.hetherington@ucl.ac.uk}{James Hetherington} & University College London\\
-\href{mailto:s.hettrick@software.ac.uk}{Simon Hettrick} & Software Sustainability Institute\\
-\href{mailto:N.ChueHong@software.ac.uk}{Neil Chue Hong} & Software Sustainability Institute\\
-\href{mailto:ljhwang@ucdavis.edu}{Lorraine Hwang} & UC Davis, CIG\\
-\href{mailto:rayi@renci.org}{Ray Idaszak} & RENCI, University of North Carolina at Chapel Hill\\
-\href{mailto:caroline.jay@manchester.ac.uk}{Caroline Jay} & University of Manchester\\
-\href{mailto:bjimenez@bsc.es}{Brian Jimenez} & Barcelona Supercomputing Center\\
-\href{mailto:rafael.jimenez@elixir-europe.org}{Rafael C Jimenez} & ELIXIR Hub, UK\\
-\href{mailto:d.katz@ieee.org}{Daniel S Katz} & University of Illinois Urbana--Champaign\\
-\href{mailto:dominic.kempf@iwr.uni-heidelberg.de}{Dominic Kempf} & Heidelberg University\\
-\href{mailto:p.lago@vu.nl}{Patricia Lago} & Vrije Universiteit Amsterdam\\
-\href{mailto:liu@hood.edu}{Xinlian Liu} & Hood College\\
-\href{mailto:hannes.loeffler@stfc.ac.uk}{Hannes Loeffler} & STFC\\
-\href{mailto:knarf@cct.lsu.edu}{Frank L\"{o}ffler} & Louisiana State University\\
-\href{mailto:tengyu.ma@vanderbilt.edu}{Tengyu Ma} & Vanderbilt University\\
-\href{mailto:rhmcdona@indiana.edu}{Robert H McDonald} & Indiana University\\
-\href{mailto:chris.mentzel@moore.org}{Chris Mentzel} & Moore Foundation\\
-\href{mailto:costa.michailidis@knowinnovation.com}{Constantinos Michailidis} & KnowInnovation\\
-\href{mailto:jmiller@perimeterinstitute.ca}{Jonah Miller} & Perimeter Institute of Theoretical Physics\\
-\href{mailto:mosse@cs.pitt.edu}{Daniel Mosse} & University of Pittsburgh\\
-\href{mailto:august.muench@aas.org}{August (Gus) Muench} & American Astronomical Society\\
-\href{mailto:unangi2@illinois.edu}{Udit Nangia} & University of Illinois Urbana--Champaign\\
-\href{mailto:a.nenadic@manchester.ac.uk}{Aleksandra Nenadic} & University of Manchester, SSI\\
-\href{mailto:Kyle.Niemeyer@oregonstate.edu}{Kyle Niemeyer} & Oregon State University\\
-\href{mailto:nunezco2@illinois.edu}{Santiago N\'u\~nez-Corrales} & University of Illinois Urbana--Champaign\\
-\href{mailto:d.a.orchard@kent.ac.uk}{Dominic Orchard} & University of Kent\\
-\href{mailto:chico@puc-rio.br}{Francisco Queiroz} & Pontifical Catholic University of Rio de Janeiro\\
-\href{mailto:fons.rademakers@cern.ch}{Fons Rademakers} & CERN\\
-\href{mailto:karthik.ram@berkeley.edu}{Karthik Ram} & University of California, Berkeley\\
-\href{mailto:samuel.relton@manchester.ac.uk}{Sam Relton} & University Of Manchester\\
-\href{mailto:thomas.robitaille@gmail.com}{Tom Robitaille} & Freelance\\
-\href{mailto:jeansalac@virginia.edu}{Jean Salac} & University of Virginia\\
-\href{mailto:janos.sallai@vanderbilt.edu}{Janos Sallai} & Vanderbilt University\\
-\href{mailto:prabhjyot10@gmail.com}{Prabhjyot (Singh) Saluja} & Southwestern Oklahoma State University\\
-\href{mailto:eseidel@cs.ucsd.edu}{Eric Seidel} & University of California, San Diego\\
-\href{mailto:sameer@cs.uoregon.edu}{Sameer Shende} & University of Oregon\\
-\href{mailto:raniere.silva@software.ac.uk}{Raniere Silva} & University of Manchester\\
-\href{mailto:shoaib.sufi@software.ac.uk}{Shoaib Sufi} & University of Manchester, Software Sustainability Institute\\
-\href{mailto:chris.sweet@nd.edu}{Chris Sweet} & University of Notre Dame\\
-\href{mailto:py12rw@leeds.ac.uk}{Rob Welch} & University of Leeds\\
-\href{mailto:Christopher.Woods@bristol.ac.uk}{Christopher Woods} & University of Bristol\\
+Mark Abraham & SciLifeLab, KTH, Sweden\\
+Aseel Aldabjan & University of Manchester\\
+Alice Allen & Astrophysics Source Code Library\\
+Gabrielle Allen & NCSA/University of Illinois\\
+Pau Andrio & Barcelona Supercomputing Center\\
+Wolfgang Bangerth & Colorado State University\\
+Radovan Bast & UiT The Arctic University of Norway, Nordic e-Infrastructure Collaboration\\
+Roman Bauer & Newcastle University\\
+Steven Brandt & Louisiana State University\\
+Lukas Johannes Breitwieser & CERN\\
+Emily Chen & NCSA/University of Illinois Urbana--Champaign\\
+Bruce Childers & University of Pittsburgh\\
+Mike Croucher & University of Sheffield\\
+Michael R. Crusoe & Common Workflow Language Project\\
+Stephan Druskat & Humboldt-Universit\"{a}t zu Berlin\\
+Anshu Dubey & Argonne National Laboratory\\
+Tim Dunne & KnowInnovation\\
+Peter Elmer & Princeton University\\
+Iain Emsley & University of Oxford\\
+Hans Fangohr & University of Southampton\\
+Stuart Geiger & Berkeley Institute for Data Science\\
+Sandra Gesing & University of Notre Dame\\
+Carole Goble & University of Manchester, Software Sustainability Institute\\
+Tom Goodale & Independent\\
+Christopher Gwilliams & Cardiff University\\
+Willem van Hage & Netherlands eScience Center\\
+Robert Haines & University of Manchester\\
+James Hetherington & University College London\\
+Simon Hettrick & Software Sustainability Institute\\
+Neil Chue Hong & Software Sustainability Institute\\
+Lorraine Hwang & UC Davis, CIG\\
+Ray Idaszak & RENCI, University of North Carolina at Chapel Hill\\
+Caroline Jay & University of Manchester\\
+Brian Jimenez & Barcelona Supercomputing Center\\
+Rafael C Jimenez & ELIXIR Hub, UK\\
+Daniel S Katz & University of Illinois Urbana--Champaign\\
+Dominic Kempf & Heidelberg University\\
+Frank L\"{o}ffler & Louisiana State University\\
+Patricia Lago & Vrije Universiteit Amsterdam\\
+Xinlian Liu & Hood College\\
+Hannes Loeffler & STFC\\
+Tengyu Ma & Vanderbilt University\\
+Robert H McDonald & Indiana University\\
+Chris Mentzel & Moore Foundation\\
+Constantinos Michailidis & KnowInnovation\\
+Jonah Miller & Perimeter Institute of Theoretical Physics\\
+Daniel Mosse & University of Pittsburgh\\
+August (Gus) Muench & American Astronomical Society\\
+Santiago N\'u\~nez-Corrales & University of Illinois Urbana--Champaign\\
+Udit Nangia & University of Illinois Urbana--Champaign\\
+Aleksandra Nenadic & University of Manchester, SSI\\
+Kyle Niemeyer & Oregon State University\\
+Dominic Orchard & University of Kent\\
+Francisco Queiroz & Pontifical Catholic University of Rio de Janeiro\\
+Fons Rademakers & CERN\\
+Karthik Ram & University of California, Berkeley\\
+Sam Relton & University Of Manchester\\
+Tom Robitaille & Freelance\\
+Jean Salac & University of Virginia\\
+Janos Sallai & Vanderbilt University\\
+Prabhjyot (Singh) Saluja & Southwestern Oklahoma State University\\
+Eric Seidel & University of California, San Diego\\
+Sameer Shende & University of Oregon\\
+Raniere Silva & University of Manchester\\
+Shoaib Sufi & University of Manchester, Software Sustainability Institute\\
+Chris Sweet & University of Notre Dame\\
+Rob Welch & University of Leeds\\
+Christopher Woods & University of Bristol\\

### 2017-05-03 WSSSPE4_report/insert_participants.py

@@ -5,7 +5,7 @@ import re, operator
 # 'column' for a column, and
 # 'off' for no email, and
 # 'email' to just print a list of emails
-email = 'email'
+email = 'off'
 
 people = []
 with open('participants.txt', 'r') as f:

### 2017-05-03 WSSSPE4_report/WSSSPE4_report.tex
Changes for section on papers and demos

@@
 
-\todo{group by type - idea, position, experience, demo}
+%
+For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
+%
+%\todo{group by type - idea, position, experience, demo}
+
+The following four papers belong to the first area and tackle diverse aspects from incentives for quality software to advocating a professional society for research software to roles and degrees for research software engineers.
 \begin{itemize}
@@
 
-\item \textbf{Olivier Philippe, Simon Hettrick, and Neil Chue Hong:
+\item \textbf{Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
 \emph{Experience Paper: Preliminary analysis of a survey of UK Research
@@
 
-\item \textbf{Anshu Dubey and Katherine Riley: \emph{Experience Paper: Software
+The second main area concerned with characteristics and needs of research software includes the following five papers.
+
+\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
 Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
@@
 
-\item \textbf{Anshu Dubey and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
+\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
 of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
@@
 
-\item \textbf{Francisco Queiroz and Rejane Spitz: \emph{Position Paper:
+\item \textbf{Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
 Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
@@
 
-\item \textbf{Bruce Childers, Jack Davidson, Wayne Graves, Bernard Rous, and
+\item \textbf{Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
 David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
@@
 
+Three contributions present work on the third main area and elucidate about lessons learned from or visions for use cases of scientific software and communities working with and\slash or on a software package.
+
 \item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
-Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse: \emph{Demo: Composing,
+Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
 Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
@@
 
-\item \textbf{Sameer Shende and Allen Malony: \emph{Demo: Using TAU for
+\item \textbf{Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
 Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
@@
 
-\item \textbf{Janos Sallai, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
+\item \textbf{Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
 \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
@@
 
-    \item \textbf{Sandra Gesing, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
+    \item \textbf{Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
     Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
@@
 % NSF grant for WSSSPE4
-This material is based upon work supported by the National Science Foundation (ACI-1648293),
+This material is based upon work supported by the National Science Foundation (ACI-1648293, ACI-1547611),
 by the Gordon and Betty Moore Foundation (GBMF\#5620),

### 2017-05-03 WSSSPE4_report/wg_best_practices_developing.tex
Update wg_best_practices_developing.tex

@@
 
+Devloping sustainable software can be investigated under many diverse aspects and dimensions: 
+economic, technical, environmental and social are only a few. The white paper will focus 
+on scientific environments and their implications and the target group are especially developer 
+and project manager of scientific software. Due to the complexity of this field it is important to select
+a subset of sustainability aspects for the white paper. The idea is to create a series of papers instead of trying
+to tackle all important topics in one paper.
+For the first white paper, we aim to set the stage with successful use cases and analyze why they have been successful.
+Further topics include community-related practices, government and management, funding, metrics, tools and usability.
+We will collect feedback from the WSSSPE community on the white paper and extend it to a journal paper.
+
 \subsubsection{Participants}
 
-\note{members of the working group}
+Jeffrey C. Carver <carver@cs.ua.edu>,
+Neil Chue Hong <N.ChueHong@software.ac.uk>,
+Tom Crick <tcrick@cardiffmet.ac.uk>,
+Miguel de~Val-Borro <valborro@princeton.edu>,
+Hans Fangohr <fangohr@soton.ac.uk>
+Sandra Gesing <sandra.gesing@nd.edu>,
+Derek Groen <Derek.Groen@brunel.ac.uk>,
+Dan Gunter <DKGunter@lbl.gov>,
+Daniel S. Katz <d.katz@ieee.org>,
+Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
+Frank L\"offler <knarf@cct.lsu.edu>,
+Suresh Marru <smarru@iu.edu>,
+Kyle E. Niemeyer <kyle.niemeyer@oregonstate.edu>,
+Abani Patra <abani@buffalo.edu>
+and
+Francisco Queiroz <chico@puc-rio.br>
 
@@
 
+The working group aims at starting a series of papers and support developers and project managers of scientific software.
+While there are already a few papers available on best practices and sustainability of scientific software in general, the goal is to achieve a series of papers that leads to consensus in the community, tackle many diverse aspects and stays up-to-date with novel trends.
+The goal is quite ambitious and we hope to attract more authors over time who would like to contribute to specific aspects and take the lead on those.
+
 \subsubsection{Gap or challenge}
 
-What is the gap or challenge being addressed?
+The challenge for the group is to accomplish a first version of the white paper to kick off the series.
+The complexity of the topic has led in 2016 to be a bit too ambitious with trying to cover a wide range of topics.
+The distribution of topics over time seems to be a better strategy to get a paper finalized.
+The series would fill also a gap on best practices that the community can contribute to.
 
@@
 \note{What tasks will the working group undertake}
+The resulting white paper will be distributed via the WSSSPE email list.
+Via collecting feedback on it, the further plans are to attract a wider community that contributes to the journal paper and investigate whether more people would like to be involved in the paper series.
 
@@
 \item Define leading authors and contributors for sections - Francisco Queiroz (done).
-\item Finalize first version - Sandra Gesing (in progress - due 31st of January, 2016)
-\item Distribute to WSSSPE community and collect feedback - Sandra Gesing (in progress - due 28th of February, 2016)
+\item Finalize first version - Sandra Gesing (in progress)
+\item Distribute to WSSSPE community and collect feedback - Sandra Gesing (in progress)
 \end{itemize}

### 2017-05-04 WSSSPE4_report/WSSSPE4_report.tex
minor edits to Sandra's changes

@@
 %
+%
+%\todo{group by type - idea, position, experience, demo}
+
+The following four papers belong to the first area, the academic environment.
+They address diverse aspects ranging from incentives for quality software to advocating a professional society for research software to roles and degrees for research software engineers.
 For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
-%
-%\todo{group by type - idea, position, experience, demo}
-
-The following four papers belong to the first area and tackle diverse aspects from incentives for quality software to advocating a professional society for research software to roles and degrees for research software engineers.
+
 \begin{itemize}
@@
 and academia, which is also a benefit for the university and allows to identify
-potential opportunities for new research projects and collaborators.
-
-The second main area concerned with characteristics and needs of research software includes the following five papers.
+potential opportunities for new research projects and collaborators. \\
+
+The second main area, concerned with characteristics and needs of research software, includes the following five papers.
 
@@
 study underway to explore how best
-to do this integration.
-
-Three contributions present work on the third main area and elucidate about lessons learned from or visions for use cases of scientific software and communities working with and\slash or on a software package.
+to do this integration.\\
+
+Three contributions focused on the third main area, elucidating lessons learned from or visions for use cases of scientific software and communities working with and\slash or on a software package.
 
@@
 %
-For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
+For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.
 %

### 2017-05-04 WSSSPE4_report/wg_best_practices_developing.tex
edits to new section

--- 2017-04-17 WSSSPE4_report/wg_best_practices_developing.tex
@@
 
-\note{Sandra to write this}
+%\note{Sandra to write this}
 
-\note{Introduction to group here, including the overall objective of work in this area.}
+%\note{Introduction to group here, including the overall objective of work in this area.}
+
+Many diverse aspects and dimensions of developing sustainable software can be investigated,
+such as economic, technical, environmental, and social. This group aims writing white papers that
+will focus 
+on scientific environments and their implications, targeted at developers 
+and project managers of scientific software. Given the complexity of this field, it is important to select
+a subset of sustainability aspects for the white paper. The idea is to create a series of papers instead of trying
+to tackle all important topics in one paper.
+For the first white paper, the group aims to set the stage with successful use cases and an analysis of why they have been successful.
+Further topics will include community-related practices, government and management, funding, metrics, tools and usability.
+The group will collect feedback from the WSSSPE community on this white paper and extend it to a journal paper.
 
@@
 
-\note{members of the working group}
+\begin{itemize}
+\item Jeffrey C. Carver <carver@cs.ua.edu>,
+\item Neil Chue Hong <N.ChueHong@software.ac.uk>,
+\item Tom Crick <tcrick@cardiffmet.ac.uk>,
+\item Miguel de~Val-Borro <valborro@princeton.edu>,
+\item Hans Fangohr <fangohr@soton.ac.uk>
+\item Sandra Gesing <sandra.gesing@nd.edu>,
+\item Derek Groen <Derek.Groen@brunel.ac.uk>,
+\item Dan Gunter <DKGunter@lbl.gov>,
+\item Daniel S. Katz <d.katz@ieee.org>,
+\item Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
+\item Frank L\"offler <knarf@cct.lsu.edu>,
+\item Suresh Marru <smarru@iu.edu>,
+\item Kyle E. Niemeyer <kyle.niemeyer@oregonstate.edu>,
+\item Abani Patra <abani@buffalo.edu>
+\item Francisco Queiroz <chico@puc-rio.br>
+\end{itemize}
 
@@
 
-\note{Specific things the working group wants to accomplish in the context of the larger objective.}
+%\note{Specific things the working group wants to accomplish in the context of the larger objective.}
+
+The working group aims at starting a series of papers and supporting developers and project managers of scientific software.
+While there are already a few papers available on best practices and sustainability of scientific software in general, the group's goal is to create a series of papers that lead to consensus in the community, tackle many diverse aspects, and stay up-to-date with new trends.
+This goal is quite ambitious, and the group hopes to attract more authors over time who would like to contribute to specific aspects and take the lead on them.
 
@@
 
-What is the gap or challenge being addressed?
+The challenge for the group is to accomplish a first version of the white paper to kick off the series.
+The complexity of the topic has led to a first attempt in 2016 that may have been a bit too ambitious, trying to cover a wide range of topics.
+The addressing of different topics over time seems to be a better strategy to finalize a paper.
+The series would fill also a gap on best practices to which the community can contribute.
 
@@
 
-The list of contributors, as of November 10th 2016, includes (in alphabetical order): \\
-Jeffrey C. Carver <carver@cs.ua.edu>,
-Neil Chue Hong <N.ChueHong@software.ac.uk>,
-Tom Crick <tcrick@cardiffmet.ac.uk>,
-Miguel de~Val-Borro <valborro@princeton.edu>,
-Hans Fangohr <fangohr@soton.ac.uk>
-Sandra Gesing <sandra.gesing@nd.edu>,
-Derek Groen <Derek.Groen@brunel.ac.uk>,
-Dan Gunter <DKGunter@lbl.gov>,
-Daniel S. Katz <d.katz@ieee.org>,
-Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
-Frank L\"offler <knarf@cct.lsu.edu>,
-Suresh Marru <smarru@iu.edu>,
-Kyle E. Niemeyer <kyle.niemeyer@oregonstate.edu>,
-Abani Patra <abani@buffalo.edu>
-and
-Francisco Queiroz <chico@puc-rio.br>
+The list of contributors, as of November 10th 2016, is the same as the list of participants above.
 
@@
 
-\note{What tasks will the working group undertake}
+%\note{What tasks will the working group undertake}
+The resulting draft white paper will be distributed via the WSSSPE email list.
+After collecting feedback on it, the group's plans are to attract a wider community to contribute to an extended journal paper version and to investigate whether more people would like to be involved in the white paper series.
 
@@
 \begin{itemize}
-\item List existing contributors and open GitHub for new ones - Sandra Gesing (done).
-\item Establish preliminary timeline  - Sandra Gesing (done).
-\item Organize related work - Francisco Queiroz (done).
-\item Define scope of the paper - Sandra Gesing (done).
-\item Suggest new sections - Francisco Queiroz (done).
-\item Define leading authors and contributors for sections - Francisco Queiroz (done).
-\item Finalize first version - Sandra Gesing (in progress - due 31st of January, 2016)
-\item Distribute to WSSSPE community and collect feedback - Sandra Gesing (in progress - due 28th of February, 2016)
+\item List existing contributors and open GitHub for new ones -- Sandra Gesing (done)
+\item Establish preliminary timeline  -- Sandra Gesing (done)
+\item Organize related work -- Francisco Queiroz (done)
+\item Define scope of the paper -- Sandra Gesing (done)
+\item Suggest new sections -- Francisco Queiroz (done)
+\item Define leading authors and contributors for sections -- Francisco Queiroz (done)
+\item Finalize first version -- Sandra Gesing (in progress)
+\item Distribute to WSSSPE community and collect feedback -- Sandra Gesing (in progress)
 \end{itemize}
@@
 
-The GitHub repository for the White Paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
+The GitHub repository for the white paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
 For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.

### 2017-05-04 WSSSPE4_report/WSSSPE4_report.tex
removing note
minor edits to papers/demos section

@@
 
-\note{Sandra Gesing will write this}
+%\note{Sandra Gesing will write this}
 
@@
 two experience papers, and three demos) that addressed a wide range of topics around
-sustainability for software in science. They covered three main areas:
+sustainability for software in science. These talks covered three main areas:
 \begin{enumerate}
@@
 
-The following four papers belong to the first area, the academic environment.
-They address diverse aspects ranging from incentives for quality software to advocating a professional society for research software to roles and degrees for research software engineers.
-For talks with multiple authors, an \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
+The following four papers belong to the first area: the academic environment.
+They address diverse aspects ranging from incentives for quality software to
+dvocating a professional society for research software to roles and degrees for
+research software engineers. For talks with multiple authors, an
+\textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
 
@@
 
-The second main area, concerned with characteristics and needs of research software, includes the following five papers.
+The second main area, concerned with characteristics and needs of research software, includes the following five papers:
 
@@
 
-Three contributions focused on the third main area, elucidating lessons learned from or visions for use cases of scientific software and communities working with and\slash or on a software package.
+Three contributions focused on the third main area: elucidating lessons learned
+from or visions for use cases of scientific software and communities working
+with and\slash or on a software package.
 

### 2017-05-04 WSSSPE4_report/wg_best_practices_developing.tex
minor edits in sections 8.1 and 8.6

@@
 Many diverse aspects and dimensions of developing sustainable software can be investigated,
-such as economic, technical, environmental, and social. This group aims writing white papers that
-will focus 
-on scientific environments and their implications, targeted at developers 
+such as economic, technical, environmental, and social. This group aims to write white papers
+that will focus on scientific environments and their implications, targeted at developers
 and project managers of scientific software. Given the complexity of this field, it is important to select
-a subset of sustainability aspects for the white paper. The idea is to create a series of papers instead of trying
-to tackle all important topics in one paper.
-For the first white paper, the group aims to set the stage with successful use cases and an analysis of why they have been successful.
-Further topics will include community-related practices, government and management, funding, metrics, tools and usability.
-The group will collect feedback from the WSSSPE community on this white paper and extend it to a journal paper.
+a subset of sustainability aspects for the white papers. The idea is to create a series of papers instead of trying
+to tackle all important topics in one paper. For the first white paper, the group aims to set the
+stage with successful use cases and an analysis of why they have been successful.
+Further topics will include community-related practices, government and management, funding,
+metrics, tools, and usability. The group will collect feedback from the WSSSPE
+community on this first white paper and extend it to a journal paper.
 
@@
 \begin{itemize}
-\item Jeffrey C. Carver <carver@cs.ua.edu>,
-\item Neil Chue Hong <N.ChueHong@software.ac.uk>,
-\item Tom Crick <tcrick@cardiffmet.ac.uk>,
-\item Miguel de~Val-Borro <valborro@princeton.edu>,
+\item Jeffrey C. Carver <carver@cs.ua.edu>
+\item Neil Chue Hong <N.ChueHong@software.ac.uk>
+\item Tom Crick <tcrick@cardiffmet.ac.uk>
+\item Miguel de~Val-Borro <valborro@princeton.edu>
 \item Hans Fangohr <fangohr@soton.ac.uk>
-\item Sandra Gesing <sandra.gesing@nd.edu>,
-\item Derek Groen <Derek.Groen@brunel.ac.uk>,
-\item Dan Gunter <DKGunter@lbl.gov>,
-\item Daniel S. Katz <d.katz@ieee.org>,
-\item Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>,
-\item Frank L\"offler <knarf@cct.lsu.edu>,
-\item Suresh Marru <smarru@iu.edu>,
-\item Kyle E. Niemeyer <kyle.niemeyer@oregonstate.edu>,
+\item Sandra Gesing <sandra.gesing@nd.edu>
+\item Derek Groen <Derek.Groen@brunel.ac.uk>
+\item Dan Gunter <DKGunter@lbl.gov>
+\item Daniel S. Katz <d.katz@ieee.org>
+\item Alexander Konovalov <alexander.konovalov@st-andrews.ac.uk>
+\item Frank L\"offler <knarf@cct.lsu.edu>
+\item Suresh Marru <smarru@iu.edu>
+\item Kyle E.~Niemeyer <kyle.niemeyer@oregonstate.edu>
 \item Abani Patra <abani@buffalo.edu>
@@
 
-The working group aims at starting a series of papers and supporting developers and project managers of scientific software.
+The working group aims to start a series of papers and supporting developers and project managers of scientific software.
 While there are already a few papers available on best practices and sustainability of scientific software in general, the group's goal is to create a series of papers that lead to consensus in the community, tackle many diverse aspects, and stay up-to-date with new trends.
@@
 The challenge for the group is to accomplish a first version of the white paper to kick off the series.
-The complexity of the topic has led to a first attempt in 2016 that may have been a bit too ambitious, trying to cover a wide range of topics.
-The addressing of different topics over time seems to be a better strategy to finalize a paper.
-The series would fill also a gap on best practices to which the community can contribute.
+The complexity of the topic has led to a first attempt in 2016 that may have been a bit too ambitious,
+by trying to cover a wide range of topics. Addressing different topics over time seems to
+be a better strategy to finalize a paper. The series would fill also a gap on best practices to which the community can contribute.
 
@@
 
-The list of contributors, as of November 10th 2016, is the same as the list of participants above.
+The list of contributors, as of 10 November 2016, is the same as the list of participants above.
 

### 2017-05-07 WSSSPE4_report/WSSSPE4_report.tex
added space after "ORCID:"
adding a list of contents the end of the intro

@@
 %
-\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID:0000-0002-3444-7615}
+\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: 0000-0002-3444-7615}
 %
@@
 This report is based on the events at the workshop and the submitted materials.
-These events included discussion of the WSSSPE mission and vision, a keynote,
-a set of presentations, a panel, and a number of working groups.  One full day of the
+These events included discussion of the call for papers (\S\ref{sec:preworkshop}), the WSSSPE mission and vision (\S\ref{sec:mission}), a keynote (\S\ref{sec:keynote}),
+a set of presentations (\S\ref{sec:papers} and \S\ref{sec:lightning}), a panel (\S\ref{sec:panel}), and a number of working groups (\S\ref{sec:WGs}).  One full day of the
 workshop was spent with participants in the working groups, which occurred in parallel
@@
 forward.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Calls for participation} \label{sec:preworkshop}
+This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
+Appendices list the organizers (\S\ref{sec:orgcom}), attendees(\S\ref{sec:attendees}), travel award recipients (\S\ref{sec:awardees}), and program committee (\S\ref{sec:progcom}), and detail the attendee survey (\S\ref{sec:survey_details}).
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Call for participation} \label{sec:preworkshop}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Mission and vision}\label{sec:mission}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{by Lorraine and Dan}
+
+Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
+The organizers developed a strawhorse, which was presented to the participants early in the meeting.
+The presentation included guidelines as well as examples from other similar communities.
+The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
+
+The participants were given time after the presentation to write down their comments.
+Seven people volunteered to work on revising the mission and vision statements, incorporating these comments and presenting back to the community.
+Based on this feedback, the committee redrafted the mission and vision.
+The committee added focus areas to clarify the organizations breadth and to keep the mission and vision simple and long-lasting.
+Comments after the close of the meeting were incorporated into the statements.
+
+A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
+via the WSSSPE mailing list, Facebook group, and Twitter.
+After 2 weeks without suggested changes, the final statements, as shown below, were added
+to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
+
+{\bf Mission.}
+WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
+
+{\bf Vision.}
+We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
+
+{\bf Focus areas.}
+WSSSPE promotes sustainable research software by positively impacting:
+\begin{itemize}
+\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
+\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
+\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
+\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
+\end{itemize}
+
+{\bf Definitions}
+
+Sustainable software has the capacity to endure such that it will continue to be available in the future, on new platforms, meeting new needs.
+
+The research software lifecycle includes:
+\begin{itemize}
+\item acquiring and assembling resources (including funding and people) into teams and communities
+\item developing software
+\item using software
+\item recognizing contributions to and of software
+\item maintaining software
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Keynote}\label{sec:keynote}
@@
 
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Mission and vision}\label{sec:mission}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{by Lorraine and Dan}
-
-Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
-The organizers developed a strawhorse, which was presented to the participants early in the meeting.
-The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
-
-The participants were given time after the presentation to write down their comments.
-Seven people volunteered to work on revising the mission and vision statements, incorporating these comments and presenting back to the community.
-Based on this feedback, the committee redrafted the mission and vision.
-The committee added focus areas to clarify the organizations breadth and to keep the mission and vision simple and long-lasting.
-Comments after the close of the meeting were incorporated into the statements.
-
-A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
-via the WSSSPE mailing list, Facebook group, and Twitter.
-After 2 weeks without suggested changes, the final statements, as shown below, were added
-to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
-
-{\bf Mission.}
-WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
-
-{\bf Vision.}
-We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
-
-{\bf Focus areas.}
-WSSSPE promotes sustainable research software by positively impacting:
-\begin{itemize}
-\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
-\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
-\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
-\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
-\end{itemize}
-
-{\bf Definitions}
-
-Sustainable software has the capacity to endure such that it will continue to be available in the future, on new platforms, meeting new needs.
-
-The research software lifecycle includes:
-\begin{itemize}
-\item acquiring and assembling resources (including funding and people) into teams and communities
-\item developing software
-\item using software
-\item recognizing contributions to and of software
-\item maintaining software
 \end{itemize}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Slack team and channels}\label{sec:slack}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 

### 2017-05-07 WSSSPE4_report/WSSSPE4_report.tex
moving Wolfgang up in the author order

@@
 Lorraine~Hwang$^{(4)}$,
-Simon~Hettrick$^{(5)}$,
-Ray Idaszak$^{(6)}$,
-Jean~Salac$^{(7)}$,
-Neil~Chue~Hong$^{(8)}$,
-Santiago~N\'u\~nez-Corrales$^{(9)}$,
-Alice~Allen$^{(10)}$,
-R.~Stuart~Geiger$^{(11)}$,
-Jonah~Miller$^{(12)}$,
-Emily~Chen$^{(13)}$,
-Anshu~Dubey$^{(14)}$,
-Wolfgang~Bangerth$^{(15)}$,
+Wolfgang~Bangerth$^{(5)}$,
+Simon~Hettrick$^{(6)}$,
+Ray Idaszak$^{(7)}$,
+Jean~Salac$^{(8)}$,
+Neil~Chue~Hong$^{(9)}$,
+Santiago~N\'u\~nez-Corrales$^{(10)}$,
+Alice~Allen$^{(11)}$,
+R.~Stuart~Geiger$^{(12)}$,
+Jonah~Miller$^{(13)}$,
+Emily~Chen$^{(14)}$,
+Anshu~Dubey$^{(15)}$,
 Patricia~Lago$^{(16)}$
@@
 %
-\thanks{{}$^{(5)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
-%
-\thanks{{}$^{(6)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: 0000-0002-3444-7615}
-%
-\thanks{{}$^{(7)}$ University of Virginia, Charlottesville, VA, USA; jeansalac@virginia.edu; ORCID: 0000-0001-5971-9333}
-%
-\thanks{{}$^{(8)}$ University of Edinburgh, UK; N.ChueHong@epcc.ed.ac.uk;
+\thanks{{}$^{(5)}$ Department of Mathematics, Colorado State
+  University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
+%
+\thanks{{}$^{(6)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
+%
+\thanks{{}$^{(7)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: 0000-0002-3444-7615}
+%
+\thanks{{}$^{(8)}$ University of Virginia, Charlottesville, VA, USA; jeansalac@virginia.edu; ORCID: 0000-0001-5971-9333}
+%
+\thanks{{}$^{(9)}$ University of Edinburgh, UK; N.ChueHong@epcc.ed.ac.uk;
 ORCID: 0000-0002-8876-7606}
 %
-\thanks{{}$^{(9)}$ Illinois Informatics Institute (I3) and National Center for Supercomputing Applications (NCSA), University of Illinois at Urbana-Champaign, IL, USA;
+\thanks{{}$^{(10)}$ Illinois Informatics Institute (I3) and National Center for Supercomputing Applications (NCSA), University of Illinois at Urbana-Champaign, IL, USA;
 nunezco2@illinois.edu;
@@
 %
-\thanks{{}$^{(10)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
-%
-\thanks{{}$^{(11)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
-%
-\thanks{{}$^{(12)}$ Perimeter Institute for Theoretical Physics \&
+\thanks{{}$^{(11)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
+%
+\thanks{{}$^{(12)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
+%
+\thanks{{}$^{(13)}$ Perimeter Institute for Theoretical Physics \&
 Department of Physics, University of Guelph,
@@
 %
-\thanks{{}$^{(13)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
-%
-\thanks{{}$^{(14)}$ Mathematics and Computer Science Division,
+\thanks{{}$^{(14)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
+%
+\thanks{{}$^{(15)}$ Mathematics and Computer Science Division,
 Argonne National Laboratory,
@@
 ORCID: 0000-0003-3299-7426}
-%
-\thanks{{}$^{(15)}$ Department of Mathematics, Colorado State
-  University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
 %

### 2017-05-10 WSSSPE4_report/WSSSPE4_paper.tex
starting WSSSPE4 paper, mostly placeholder
starting to shorten removing survey appendix shortening 8.1

--- 2017-05-07 WSSSPE4_report/WSSSPE4_report.tex
@@
 
-\title[]{Report on the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}
+\title[]{Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\todo{cite~\cite{WSSSPE4-report}}
 
@@
 
-\input{wg_best_practices_sustainability}
+\subsection{Verifying best practices \& metrics for sustainable research software}
+\label{sec:best-practices-sustainable}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Ray to write this}
+
+Many open source projects for research software document their best practices that
+contributed to the sustainability of the software.  Many of these projects also
+document software metrics they use to define their research software project's success.
+%
+This group will take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
+This will allow the group to identify gaps on both sides.
+This approach will also allow the group to hypothesize how successful open source projects can be further improved, verify that recommended approaches for software engineering for science/academia are sufficient and valid, and that metrics for software engineering for science/academia are relevant and useful.
+The group has a specific objective of getting a good cross-sampling of disparate software including community model codes, community cyberinfrastructure, community analytic tools, etc.
+
+Each group member will volunteer to take on one to two projects, ideally ones in which they are already interested.
+This effort will evaluate an initial 5--10 projects.
+Once the group documents the workflow by which these evaluations are performed, other people can choose to follow this workflow and contribute their own evaluations of additional software.
+
+To join this group or obtain more information about it, please send an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
+
 \input{wg_alliance}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Detailed survey results}  \label{sec:survey_details}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\input{survey_details}
+%\section{Detailed survey results}  \label{sec:survey_details}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\input{survey_details}
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
shortening another wg section

@@
 
-\input{wg_alliance}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Software Sustainability Alliance}
+\label{sec:alliance}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Neil and Jean wrote this}
+
+The Software Sustainability Alliance working group aims to establish an alliance between organizations interested in improving the sustainability of research software.
+Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
+These organizations are funded groups or teams that aim to advance research software sustainability beyond their local university or community.
+More specifically, this working group aims to define the scope of this alliance and provide clear distinction with WSSSPE.
+The group also seeks to understand the incentives for members of this alliance and identify its key activities.
+It is currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
+
+Currently, point-to-point collaboration exists between organizations, but this inadvertently results in competition or redundancy within the sustainable software community.
+An alliance of software sustainability organizations would ease inter-organization collaboration and the promotion of software sustainability.
+This alliance would also improve the pooling of competencies and the sharing of expertise.
+Furthermore, with the international scope of this alliance, it could support an organization who wants to hold an event in a country where another organization exists.
+
+If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
+
 \input{wg_prototyping}

### 2017-05-11 WSSSPE4_report/WSSSPE4_report.tex
fixing a typo and some formatting - will update in arXiv later

@@
 They address diverse aspects ranging from incentives for quality software to
-dvocating a professional society for research software to roles and degrees for
+advocating a professional society for research software to roles and degrees for
 research software engineers. For talks with multiple authors, an
@@
 
+\end{itemize}
+
 The second main area, concerned with characteristics and needs of research software, includes the following five papers:
+
+\begin{itemize}
 
@@
 
+\end{itemize}
+
 Three contributions focused on the third main area: elucidating lessons learned
@@
 with and\slash or on a software package.
+
+\begin{itemize}
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_report.tex
fixing more small wording; again will update arXiv later

@@
 To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
-These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software;
+These were: Alice Allen, Editor, Astrophysics Source Code Library, who brought an understanding of the difficulties of organizing a community and curating their software;
 Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
-Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
-Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
+Patricia Lago from Vrije Universiteit Amsterdam, who presented a keynote at the workshop (\S\ref{sec:keynote})  and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
+Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
more shortening removing most appendices removing talk descriptions and combining all talks into one section shortening another working group subsection

@@
 papers, demos, and lightning talks, and a panel discussion.  The main part of the report
-covers the set of working groups that formed during the meeting, and for each,
-discusses the participants, the objective and goal, and how the objective can
-be reached, along with contact information for readers who may want to join
+covers the set of working groups that formed during the meeting, along with contact information for readers who may want to join
 the group. Finally, we present results from a survey of the workshop attendees.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\todo{cite~\cite{WSSSPE4-report}}
 
@@
 
-WSSSPE4 also included two professional event organizers/facilitators from
-KnowInnovation who helped the organizing committee plan the workshop agenda,
-and during the workshop, they actively engaged participants with various
-tools, activities, and reminders.
+WSSSPE4 also included two professional event organizers/facilitators who helped 
+plan the workshop agenda,
+and who engaged with participants during the workshop.
 
@@
 
-This report is based on the events at the workshop and the submitted materials.
-These events included discussion of the call for papers (\S\ref{sec:preworkshop}), the WSSSPE mission and vision (\S\ref{sec:mission}), a keynote (\S\ref{sec:keynote}),
-a set of presentations (\S\ref{sec:papers} and \S\ref{sec:lightning}), a panel (\S\ref{sec:panel}), and a number of working groups (\S\ref{sec:WGs}).  One full day of the
+This paper is a summary of a longer report~\cite{WSSSPE4-report} on the events at the workshop and the submitted materials.
+The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).
+One full day of the
 workshop was spent with participants in the working groups, which occurred in parallel
@@
 This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
-Appendices list the organizers (\S\ref{sec:orgcom}), attendees(\S\ref{sec:attendees}), travel award recipients (\S\ref{sec:awardees}), and program committee (\S\ref{sec:progcom}), and detail the attendee survey (\S\ref{sec:survey_details}).
+An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).
 
@@
 3 demos
-were accepted. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
+were accepted, as mentioned in Section~\ref{sec:presentations}. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
 were accepted as lightning talks.)
-The papers are discussed in Section~\ref{sec:papers},
-and the lightning talks are discussed in Section~\ref{sec:lightning}.
 The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Keynote}\label{sec:keynote}
+\section{Presentations}\label{sec:presentations}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 In addition, software sustainability adds a fourth dimension to environmental, social and economic aspects: technical sustainability, and hence higher complexity~\cite{Lago2015}.
-
-Supporting technical sustainability in (scientific) software is very near to supporting properties like longevity and resilience, i.e., the ability of a software system or application to fulfill its intent~\cite{Huisman2016}, as planned independently from time passing or from changes in its operational context.
-Technical sustainability has never been easy, but it is even more of a challenge nowadays with the increasing demands on software solutions to support many aspects of our daily lives.
-
-The keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability.
-This provided a complementary angle to the workshop discussions.
-It addressed software energy efficiency (i.e., how to develop software that consumes less energy, like mobile apps that should consume less device-battery) and software energy-awareness (i.e., how to develop software that can actively decide between energy efficiency and other goals such as performance).
-
-The keynote used a few myths to elaborate on such challenges and related research. Among them:
-
-\begin{itemize}
-\item[] {\bf Myth \#1: Energy-efficient hardware will solve the issue}.
-At the implementation level, to make software use resources effectively, we need to first understand the impact of different coding practices on the way software exploits computing resources (hence energy to feed them).
-In addition, when we have to deal with complex software or new projects, we must understand ``the big picture,'' hence we must reason at the design or architecture level.
-Maybe the software is not implemented yet, or maybe we must identify which design decisions are more or less beneficial for energy consumption.
-To this aim, we capture best practices in reusable patterns and so-called architectural tactics.
-These document generic designs, specific implementations, metrics, and measures.\\
-
-\item[] {\bf Myth \#2: Extensive experimentation will provide the needed know how}.
-While empirical methods to design and execute software engineering experiments are mature and well known, they are insufficient to handle the too many variables and uncertainties for modern software to be sustainable.
-We need new empirical methods that help us gather evidence faster~\cite{Procaccianti2015}.
-This would probably entail relaxing the too many constraints that govern current rules for mitigating threats to validity, and (equally importantly) collaboration- and experience-sharing to build a common knowledge base.
-In addition, we should exploit better (and more truthful) visualizations of experimentation results, to communicate on advances (and less promising directions.)\\
-
-\item[] {\bf Myth \#3: With massive migration of software systems to cloud-based environments, we do not need to invest in a ``greener cloud'' anymore}.
-The use of cloud-based services promises large-scale optimizations in terms of the resources used by our software applications and related investments~\cite{green-cloud-sw}.
-Unfortunately, cloud provisioning heavily relies on over-capacity rather than intelligent governance, and so far has invested very little in innovation for sustainability.
-We need to build smarter software that both manages resources more efficiently in the cloud, better understands the changes in the software operational context (including user requirements), and dynamically adapts to address sustainability intents~\cite{Procaccianti2016-2} (both in the cloud and on customer's premises.)\\
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Papers and demos} \label{sec:papers}
+Lago's keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability, providing a complementary angle to the workshop discussions.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Papers and demos} \label{sec:papers}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 two experience papers, and three demos) that addressed a wide range of topics around
-sustainability for software in science. These talks covered three main areas:
-\begin{enumerate}
-\item The specifics of the academic environment for software development and for so-called
-research programmers or research software engineers;
-\item The characteristics and needs of research software in general such as the lifecycle of
-software in science, preservation, curation, sharing, and reproducibility challenges; and
-\item Lessons learned from or visions for use cases of scientific software and communities
-working with and\slash or on a software package.
-\end{enumerate}
-
-%
-%
-%\todo{group by type - idea, position, experience, demo}
+sustainability for software in science. These talks covered three main areas.
 
@@
 \begin{itemize}
-\item \textbf{Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
-Improving Incentives for Quality Software}}~\cite{Heroux:2016ws}.
-The author presents that improving quality of scientific software is a measure
-to address the increased complexity in both modeling and software. To support
-quality, it is essential to provide Computational Science and Engineering teams
-with publication venues, funding, and  professional recognition. Broad
-improvements in sustainability and productivity can be only achieved when
-publishers, funding agencies and employers raise their expectations for
-software quality. Additionally, software community leaders are in the unique
-position to positively impact software quality by working on establishing
-incentives that will spur creative and novel approaches to improve developer
-productivity  and software sustainability.
-
-\item \textbf{Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
-Society for Research Software}}~\cite{GAllen:2016ws}.
-This paper advocates to create a professional society or extend the scope of an
-existing organization in research software. The author presents the motivation,
-a variety of potential roles of such a society and diverse potential funding
-mechanisms to be able to establish a sustainable organization. She gives an
-overview on similar activities as well as existing organizations and aims at
-stimulating discussion among potential members and stakeholders.
-
-\item \textbf{Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
+\item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
+Improving Incentives for Quality Software}~\cite{Heroux:2016ws}.
+
+\item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
+Society for Research Software}~\cite{GAllen:2016ws}.
+
+\item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
 \emph{Experience Paper: Preliminary analysis of a survey of UK Research
-Software Engineers}}~\cite{Philippe:2016ws}.
-This paper presents results from a survey conducted on the new role of research
-software engineers in academia. The survey provides demographic information
-about the education, field, gender, job satisfaction, and career plans of the people
-in this community. The outcome is that members of the community  were found to be
-highly educated, derive mainly from the hard sciences, and are predominantly male.
-Respondents reported satisfaction in their jobs, but indicated that career progression
-is both difficult and opaque. The authors urge a continued discussion about the experience
-of research software engineers and recommend further investigation into this
-important community.
-
-\item \textbf{Christopher Gwilliams: \emph{Demo: Using industrial engagement to
-create and develop research ties within academia}}~\cite{Gwilliams:2016ws}.
-This demo focuses on the benefits of including a  new  software engineering
-degree that is influenced by, and works directly with,
-industry.  Cardiff  University  introduced  a  new  industry-focused  degree,
-Bachelor of Science in Applied Software Engineering (ASE) in September
-2015. The engagement with industry allows students to learn through live
-projects  with  real-world  applications and forge links for career opportunities.
-Also, industry benefits from this course that provides an entry point for industry
-to engage with the university and to develop a network of contacts between industry
-and academia, which is also a benefit for the university and allows to identify
-potential opportunities for new research projects and collaborators. \\
-
-The second main area, concerned with characteristics and needs of research software, includes the following five papers:
-
-\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
-Engineering and Community Codes Track in ATPESC}}~\cite{Dubey1:2016ws}.
-The Argonne Training Program in Extreme Scale
-Computing (ATPESC) was started by Argonne National Laboratory with the objective
-of expanding the ranks of better prepared users of high-performance computing (HPC) machines.
-One of the unique aspects of the program was inclusion of a
-software engineering and community codes track. The inclusion
-was motivated by the observation that the projects with a good
-software process were better able to meet their scientific goals.
-The paper presents the experience in running the software
-track, the motivation, reception, and evolution of the track over the years.
-The authors recommend the inclusion of similar tracks in other HPC-oriented training programs.
-
-\item \textbf{Neil Chue Hong: \emph{Position Paper: Why do we need to compare
-research software, and how should we do it?}}~\cite{ChueHong:2016ws}.
-This position paper sets out the reasons for why different stakeholders,
-from users to developers to funders, might wish to undertake the difficult task
-to compare research software. The author describes a proposed framework for doing
-so (based around measures of accessibility, usability, maintainability, and
-portability) which takes into account the possibility of variation between
-different communities about how they prioritize different aspects of research software.
-
-\item \textbf{Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
-of Software for Scientific Simulation}}~\cite{Dubey2:2016ws}.
-The authors discuss the specifics of the lifecycle of scientific software.
-While the software lifecycle is a well-researched topic, one class of projects,
-research software, has received relatively
-little attention from lifecycle researchers. This
-paper formalizes a lifecycle model
-for end-to-end scientific application software, featuring loose
-coupling between submodels for development of infrastructure
-and scientific capability. The authors also invite input from stakeholders
-to converge on a model that captures the complexity of this
-development processes and provides needed lifecycle guidance to
-the scientific software community.
-
-\item \textbf{Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
-Collaborative Gamification Design for Scientific Software}}~\cite{Queiroz:2016ws}.
-The paper focuses on usability design and presents that gamification, a design
-trend for improving scientific software usability, for scientific software should
-be facilitated by an open, collaborative design process supported by
-conversational media. This approach is compatible with
-qualities often attributed to computational science community
-regarding openness and collaboration between members of varied
-professional backgrounds. They exemplify the use of conversational media for
-collaborative design and expect the synergy between collaborators to result in
-better usability, greater user acceptance, and adequacy to requirements,
-obtaining optimal design solutions in a sustainable way.
-
-\item \textbf{Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
+Software Engineers}~\cite{Philippe:2016ws}.
+
+\item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
+create and develop research ties within academia}~\cite{Gwilliams:2016ws}.
+
+\end{itemize}
+
+\noindent The second main area, concerned with characteristics and needs of research software, includes the following five papers:
+
+\begin{itemize}
+
+\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
+Engineering and Community Codes Track in ATPESC}~\cite{Dubey1:2016ws}.
+
+\item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
+research software, and how should we do it?}~\cite{ChueHong:2016ws}.
+
+\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
+of Software for Scientific Simulation}~\cite{Dubey2:2016ws}.
+
+\item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
+Collaborative Gamification Design for Scientific Software}~\cite{Queiroz:2016ws}.
+
+\item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
 David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
-the Way Digital Libraries will Operate}}~\cite{Childers:2016ws}.
-The paper presents the implications of ``active curation systems'' that provide
-executable access to
-artifacts and experiments behind published results and enables
-their reuse. These systems allow changing and repeating experiments to understand
-how an innovation behaves in conditions beyond the ones described in a paper and
-address reproducibility challenges. Active curation systems
-also  enable accountability and accelerate research progress  by
-giving access to complete experimental details. As these systems
-take  hold,  it  is  important  to  understand  their  capabilities  and
-how digital libraries should be integrated with them. The authors describe a
-study underway to explore how best
-to do this integration.\\
-
-Three contributions focused on the third main area: elucidating lessons learned
+the Way Digital Libraries will Operate}~\cite{Childers:2016ws}.
+
+\end{itemize}
+
+\noindent Three contributions focused on the third main area: elucidating lessons learned
 from or visions for use cases of scientific software and communities working
@@
 
-\item \textbf{Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
+\begin{itemize}
+
+\item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
 Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
-Reproducing, and Sharing Simulations}}~\cite{Ganguly:2016ws}.
-The authors demonstrate their approach of sharing,
-reproducing, and composing simulations toward accelerating
-research productivity while also improving accountability and
-credibility. Specifically, they have developed a case study in
-which they compose and share access control simulations in the
-form of shareable data store units for cloud systems. They share, compose
-and repeat simulations  through a collaborative repository
-(OCCAM) and suggest a general simulation framework (SST) that can
-accelerate the efforts for the whole community.
-
-\item \textbf{Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
-Performance Evaluation of Scientific Software}}~\cite{Shende:2016ws}.
-This demo presents the  TAU Performance System for performance evaluation of Scientific
-Software written in C++, C, and Fortran. The system addresses performance
-technology problems  at three levels: instrumentation,
-measurement, and analysis. The TAU framework supports the
-configuration and integration of these layers. The authors advocate that effective
-exploration of performance will necessarily
-require prudent selection from the range of alternative methods TAU provides
-to  assemble meaningful performance experiments.
-
-\item \textbf{Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
+Reproducing, and Sharing Simulations}~\cite{Ganguly:2016ws}.
+
+\item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
+Performance Evaluation of Scientific Software}~\cite{Shende:2016ws}.
+
+\item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
 \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
-Atom-Typing for Molecular Simulation}}~\cite{Sallai:2016ws}.
-The authors present an essential and error-prone task in the area of Molecular
-Dynamics simulations: the efficient selection of correct parameter values for
-specific molecules, especially forcefields. They present a framework that
-aims to solve this data management issue, proposing a common format for
-forcefields that is self-documenting with machine readable,
-declarative usage rules. They investigate processes and tools
-that are commonly used today in software development (e.g.,
-unit testing, verification and validation, continuous integration,
-and version control) and reason that they are, with proper infrastructure
-support, applicable to forcefield development, as well. The
-paper describes how such an infrastructure can tackle managing and evolving
-forcefields by the molecular dynamics community, and
-proposes a way to encourage and incentivize involvement
-by the stakeholders.
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Lightning talks} \label{sec:lightning}
+Atom-Typing for Molecular Simulation}~\cite{Sallai:2016ws}.
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Lightning talks} \label{sec:lightning}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-There were 19 lightning talks presented at WSSSPE4, as follows (in order of
+\noindent There were 19 lightning talks presented at WSSSPE4, as follows (in order of
 presentation).
 %
-For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.
-%
-\begin{itemize}[itemsep=1ex]
-    \item \textbf{Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
-    programme; supporting the social side of research software}}~\cite{Sufi:2016ws}.
-    Sufi, Community Lead at the Software Sustainability Institute (SSI), discussed
-    the SSI Fellowship program and the fellows, including their
-    demographic breakdown and domains of practice. This program came about
-    based on the institute's desire to integrate with the research community---thus
-    inspiring change from within---rather than preach to it. Five SSI fellows
-    attended WSSSPE4.
-
-    \item \textbf{Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
+%For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.
+
+\begin{itemize}%[itemsep=1ex]
+    \item Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
+    programme; supporting the social side of research software}~\cite{Sufi:2016ws}.
+
+    \item Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
     Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
     \emph{The Science Gateways Community Institute - Supporting Communities to
-    Achieve Sustainability for Their Science Gateways}}~\cite{Gesing:2016ws}.
-    In this talk, Gesing introduced the concept of science gateways: end-to-end software
-    solutions for a community that hide underlying complexity. She also posed the need
-    for gateways, which include the increased complexity of research and associated
-    hardware\slash software, a greater need for reproducibility and openness, and the
-    opportunity to integrate research with teaching for better workforce development.
-    In fact, gateway users surpassed login-based users in XSEDE systems in 2013,
-    and 1.4 times as many users accessed XSEDE via gateways compared with
-    login in March 2016. The Science Gateways Community Institute offers support
-    related to gateways in five areas: Incubator, Extended Developer Support, Scientific
-    Software Collaborative, Community Engagement and Exchange, and Workforce
-    Development.
-
-    \item \textbf{Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
+    Achieve Sustainability for Their Science Gateways}~\cite{Gesing:2016ws}.
+
+    \item Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
     and Scott Chamberlain: \emph{A model for
-    peer review and onboarding research software}}~\cite{Ram:2016ws}.
-    This talk described the rOpenSci community's review process for research
-    software packages. Submission of a package starts with a pre-submission inquiry
-    for fit with the community criteria, then peer reviewers evaluate based on
-    usability, quality, and style. Upon acceptance, packages are given an rOpenSci
-    README badge and added to the system. Important components include a code of
-    conduct, open and non-adversarial review, and that no submission is
-    rejected---just out of scope. Remaining challenges include scaling, author
-    incentives, and automation of the process.
-
-    \item \textbf{Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
-    \emph{A vision of computing in 10+ years}}~\cite{Loffler:2016ws}.
-    This talk began by describing the challenges facing science, and particularly
-    computational science, in recent years: the amount and complexity of knowledge
-    has increased dramatically, making it difficult for individuals or even teams to
-    manage modern, complex problems.
-    Brandt submitted an argument for decoupling science simulation and
-    the solution of computational problems away from numerical methods and low-level
-    optimization. This may be possible with domain-specific languages, or the use
-    of modern features in existing languages. This vision also includes reuse of
-    basic computing infrastructure, including common data structures, parameter files,
-    and high-performance I\slash O. This computing future necessarily requires
-    better recognition and reward systems for research software developers (i.e.,
-    research software engineers), and ways to make increasingly complex scientific
-    papers more understandable.
-
-    \item \textbf{Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
+    peer review and onboarding research software}~\cite{Ram:2016ws}.
+
+    \item Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
+    \emph{A vision of computing in 10+ years}~\cite{Loffler:2016ws}.
+
+    \item Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
     Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
-    \emph{Software Citation: Process, Principles, and Implementation}}~\cite{Katz:2016ws}.
-    This talk described the timeline of activities of the FORCE11 Software Citation
-    Working Group and the principles of software citation established by the group.
-    Niemeyer, one of the working group co-chairs, also discussed the paper published
-    describing the principles~\cite{Smith:2016sc}, and the remaining steps of the
-    working group before sunsetting it and spinning up a new Software Citation
-    Implementation Working Group.
-
-    \item \textbf{Ray Idaszak\textsuperscript{\textasteriskcentered},
+    \emph{Software Citation: Process, Principles, and Implementation}~\cite{Katz:2016ws}.
+
+    \item Ray Idaszak\textsuperscript{\textasteriskcentered},
     David G.~Tarboton, Hong Yi, Michael Stealey,
@@
     Best Practices and Culture Change for Developing Sustainable Community
-    Software}}~\cite{Idaszak:2016ws}.
-    HydroShare is a five-year-old online, collaborative system that supports
-    open sharing of hydrologic data, analytical tools, and computer models. The
-    project involves 10 collaborating institutions that produce tested and reviewed
-    community code every two to three weeks, following software engineering best
-    practices. Undergraduate and graduate students involved in the project not
-    only adopt the best practices exhibited by HydroShare, but see the open,
-    collaborative approach to research software development as the norm.
-    The talk made recommendations to other community efforts, including the need to
-    identify committed, hands-on PIs; evidence-based software engineering for science;
-    continuous integration; the importance of professional software development and
-    operations practices, even for domain scientists; and to identify technology
-    decisions that overlap with career paths.
-
-    \item \textbf{Lorraine Hwang\textsuperscript{\textasteriskcentered},
+    Software}~\cite{Idaszak:2016ws}.
+
+    \item Lorraine Hwang\textsuperscript{\textasteriskcentered},
     Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
     \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
-    Community}}~\cite{Hwang:2016ws}.
-    In this talk, Hwang first described the ASPECT (Advance Solver for Problems in
-    Earth ConvecTion) open-source, parallel, extensible finite element code
-    used to simulate thermal convection. ASPECT began development in 2011 as a
-    community-driven project under the Computational Infrastructure for Geodynamics
-    (CIG). Community-building activities around ASPECT focused on the traditional
-    workshops, tutorials, and webinars, but also on slightly nontraditional,
-    multi-day hackathons---which turned out to be one of the most effective
-    forms of fostering a community. These start with leaders establishing project
-    conventions, demonstrating workflows and relevant technical skills, and
-    identifying starter projects. Daily activities involve work plans and
-    science\slash coding briefs, with mini-tutorials given as needed. One important
-    requirement is that each participant must submit at least one commit. Hwang
-    also summarized lessons learned, including that hackathons should last 7--10
-    days, include participants with multidisciplinary expertise and at both novice
-    and expert levels, have at least one dedicated code reviewer per 10 participants,
-    hold daily group updates, organize activities to promote group interactions,
-    and enforce free time.
-
-    \item \textbf{Carole Goble: \emph{A Simple Profiling Framework for Software
-    User-Producer Reciprocity Review}}~\cite{Goble:2016ws}.
-    Goble first described some existing frameworks for measuring the maturity of
-    software, including the Software Sustainability Maturity Model, OSS Watch
-    Openness Rating, NASA Reuse Readiness Rating, Capability Maturity Model, and
-    Qualification and Selection of Open Source software. Then, she proposed a new,
-    simpler framework for project reciprocity, drawing on the intentions of users
-    and producers of the software, from the viewpoint of the producer.
-    Next steps include further definition of the model's characteristics based on
-    both theoretical and empirical investigation.
-
-    \item \textbf{Stephan Druskat: \emph{A proposal for the measurement and
+    Community}~\cite{Hwang:2016ws}.
+
+    \item Carole Goble: \emph{A Simple Profiling Framework for Software
+    User-Producer Reciprocity Review}~\cite{Goble:2016ws}.
+
+    \item Stephan Druskat: \emph{A proposal for the measurement and
     documentation of research software sustainability in interactive metadata
-    repositories}}~\cite{Druskat:2016ws}.
-    In this talk, Druskat first identified two technical barriers to software
-    sustainability: identification of good software, and software discoverability.
-    Some of his perspectives were based on experiences in the humanities, a lack
-    of funding for computational infrastructure results in little experience with
-    sustainability (although this may differ in other domains). Druskat proposed
-    metadata repositories as a possible solution, as they can measure and document
-    technical sustainability, and document resulting metrics and more general
-    features of software. Technical sustainability goals include ensuring software
-    existence, preserving potential operation of software, and creating\slash
-    retaining possibilities for further development and maintenance of software.
-    The next steps of such a system include compiling measurable criteria for
-    the sustainability goals (based on qualitative concepts of, e.g.,
-    ``sustainability,'' ``maintainability,'' and ``usability''), and
-    crowd-sourcing a weighting of these criteria.
-
-    \item \textbf{Mistral Contrastin, Matthew Danish,
+    repositories}~\cite{Druskat:2016ws}.
+
+    \item Mistral Contrastin, Matthew Danish,
     Dominic Orchard\textsuperscript{\textasteriskcentered}, and
     Andrew Rice: \emph{Supporting software sustainability with
-    lightweight specifications}}~\cite{Contrastin:2016ws}.
-    This talk introduced the tool CamFort for verifying Fortran code based on
-    lightweight specifications, which aid in software maintainability and reuse
-    through high-level indications of programmer intent.
-    Currently, CamFort supports specification (and
-    verification) of units and stencil operations. Plans for future specifications
-    include automatic test generation based on user-supplied properties, dependency
-    specifications, and software contracts to check, e.g., expected value ranges
-    and program behavior.
-
-    \item \textbf{Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
+    lightweight specifications}~\cite{Contrastin:2016ws}.
+
+    \item Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
     Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
@@
     and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
-    Responsibility in the Engineering of Academic Software}}~\cite{AAllen:2016ws}.
-    This talk described the activities of the weeklong Dagstuhl Perspectives Workshop
-    manifesto on ``Engineering Academic Software,'' which produced a manifesto
-    describing a pledge affirming the scholarly value of research software and
-    the personal responsibility to act in accordance with that value. The draft
-    (personal) manifesto includes 20 pledges, organized into three areas:
-    recognition of academic software, academic software development processes,
-    and the intellectual content of academic software. The talk concluded with a
-    survey for participants to give feedback on the pledges and their importance.
-
-    \item \textbf{Neil Chue Hong\textsuperscript{\textasteriskcentered}:
-    \textit{Making it easier to understand research software impact}}~\cite{ChueHong:2016wsb}.
-    Chue Hong proposed that the WSSSPE community collaboratively develop a framework
-    to better understand (and measure the impact of) research software. This could
-    be based on a workflow involving mining the literature for mentions of software,
-    identifying linkages between software, and visualizing dependencies between
-    software or between software and other research artifacts. Many tools already
-    exist in this ecosystem, so Chue Hong proposed that a WSSSPE working group
-    collect these tools, summarize use cases, identify missing pieces, and find
-    low-hanging fruit to integrate tools or build new ones.
-
-    \item \textbf{Willem Robert van Hage\textsuperscript{\textasteriskcentered},
+    Responsibility in the Engineering of Academic Software}~\cite{AAllen:2016ws}.
+
+    \item Neil Chue Hong\textsuperscript{\textasteriskcentered}:
+    \textit{Making it easier to understand research software impact}~\cite{ChueHong:2016wsb}.
+
+    \item Willem Robert van Hage\textsuperscript{\textasteriskcentered},
     Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
-    Netherlands eScience Center}}~\cite{vanHage:2016ws}.
-    This talk described eStep (eScience Technology Platform) at the Netherlands
-    eScience Center, which includes three components: a software catalogue
-    of eScience software, interfaces, libraries,
-    tools workflows, applications, etc.; a knowledge base of guides, reports, and
-    recommendations for scientific software development; and eScience research
-    documents including such as scientific publications, book chapters, and demos.
-    The eStep website currently presents automatically-generated information about
-    software based on metadata, with a plan for displaying measures of impact and
-    quality.
-
-    \item \textbf{Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
+    Netherlands eScience Center}~\cite{vanHage:2016ws}.
+
+    \item Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
     Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
-    Research Software}}~\cite{GAllen:2016wsb}.
-    In this talk, Chen first summarized some of the outcomes of the WSSSPE3
-    Metrics Breakout Group, who discussed that metrics are important for scientific
-    impact, discovery, reducing duplication of efforts, and evaluating software.
-    The group also summarized barriers to widespread use of metrics: no common
-    set of metrics collected, a need for usefulness to be obvious for projects,
-    and no common standard for collecting\slash presenting metrics. To find out
-    more about practices related to metrics, the authors ran a survey of National
-    Science Foundation Software Infrastructure for Sustained Innovation (SI2)
-    program investigators, with 19 participants. Important results include the
-    observation that most use citations and acknowledgements of software in
-    funding proposals, while download numbers are both challenging to collect
-    and not useful. Overall, participants generally agreed that metrics are
-    important, but said useful metrics are difficult or tedious to collect.
-
-    \item \textbf{Eric L.~Seidel\textsuperscript{\textasteriskcentered}
+    Research Software}~\cite{GAllen:2016wsb}.
+
+    \item Eric L.~Seidel\textsuperscript{\textasteriskcentered}
     and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
-    into Scientific Software}}~\cite{Seidel:2016ws}.
-    This talk, given by a senior graduate student in Computer Science\slash
-    programming languages research, described several areas of software engineering
-    concepts that benefit the development of research software. In particular,
-    Seidel focused on software correctness, starting from high-level
-    specifications. Correctness can be demonstrated by verifying or proving
-    that the software satisfies the specification, testing against a range of
-    inputs, or synthesis of code from the specification. However, all of these
-    present technical challenges, in addition to the social challenges of
-    encouraging collaboration between software engineering and research software
-    communities.
-
-    \item \textbf{Iain Emsley\textsuperscript{\textasteriskcentered} and David
+    into Scientific Software}~\cite{Seidel:2016ws}.
+
+    \item Iain Emsley\textsuperscript{\textasteriskcentered} and David
     De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
-    community}}~\cite{Emsley:2016ws}.
-    In this talk, Emsley discussed efforts to develop the community around the
-    Drupal content management system, which is supported through the DrupalCon
-    series of conferences, user groups, and an association.
-    The talk described the use of social coding and software credit to develop
-    the Drupal software community, and social documentation to provide
-    documentation and tutorials. By analyzing identities of committers, they
-    found that the larger community is actually a loosely coupled set of
-    heterogeneous communities that use (and sustain themselves) through different
-    social mechanisms.
-
-    \item \textbf{Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
+    community}~\cite{Emsley:2016ws}.
+
+    \item Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
     D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
     and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
-    BLAS Routines}}~\cite{Dongarra:2016ws}.
-    This talk described Batched BLAS (BBLAS), a new standard for computing multiple
-    BLAS (Basic Linear Algebra Subproblems) in parallel. Unlike existing parallel
-    BLAS algorithms provided by, e.g., Intel MKL, cuBLAS, or MAGMA, BBLAS is
-    optimized for smaller matrices, which apply to a variety of problems in areas
-    such as machine learning, multifrontal solvers, image processing, fluid
-    dynamics, and astrophysics. BBLAS offers nearly a factor of two improvement
-    in computing speed compared to non-batched operations. To develop a sustainable
-    BBLAS library, the developers have focused on proposing a standard API,
-    holding workshops between academics and vendors to reach consensus, and
-    performing research into API design decisions on performance.
-
-    \item \textbf{Lukas Breitwieser\textsuperscript{\textasteriskcentered},
+    BLAS Routines}~\cite{Dongarra:2016ws}.
+
+    \item Lukas Breitwieser\textsuperscript{\textasteriskcentered},
     Roman Bauer, Alberto Di Meglio, Leonard Johard,
     Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
-    Max Talanov: \textit{The BioDynaMo Project}}~\cite{Bauer:2016ws}.
-    Breitwieser described the BioDynaMo project, a large-scale platform for
-    biological simulations that provides access to computational resources while
-    hiding the complexity of distributed computing. It also promotes computational
-    reproducibility through shared open-access data. The project began as a
-    modernization initiative porting an existing simulation software, Cx3D, from
-    Java to C++, but also incorporated many modern software engineering practices
-    such as hosting and version control on GitHub, automated tests and continuous
-    integration via Travis-CI, code coverage reports, and an automated build system.
-    In addition, over 1000 students participated in development via the Intel
-    Modern Code Developer Challenge competition---the winner achieved a
-    performance speedup of 320 times. The project plans to continue improving
-    the development process with, e.g., static code checkers and Git commit hooks,
-    continued architecture redesign, and verification.
-
-    \item \textbf{Aseel Aldabjan, Robert Haines, and
+    Max Talanov: \textit{The BioDynaMo Project}~\cite{Bauer:2016ws}.
+
+    \item Aseel Aldabjan, Robert Haines, and
     Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
     measure the relationship between code quality and software
-    sustainability?}}~\cite{Aldabjan:2016ws}.
-    This talk explored the relationship between code quality and the sustainability
-    of research software; however, the latter has no concrete definition, thus making
-    it challenging to provide guidance on how to achieve sustainability. The authors
-    proposed a ``software sustainment'' metric based on the time difference between
-    the most recent and initial commits\slash contributions to the codebase.
-    Using this metric, they analyzed 3113 Java projects hosted on GitHub, created
-    in 2009, and found that 22\% and 35\% had sustainment metrics of 0 and less
-    than seven, respectively. Interestingly, the authors also found high
-    code quality (measured using a variety of static analytic metrics) across
-    all these projects, with little correlation to sustainability.
-
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Best practices for scientific software panel discussion} \label{sec:panel}
+    sustainability?}~\cite{Aldabjan:2016ws}.
+
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Best practices for scientific software panel discussion} \label{sec:panel}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-To investigate best practices for scientific software we brought together a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
-These were: Alice Allen, Editor, Astrophysics Source Code Library, brought an understanding of the difficulties of organizing a community and curating their software;
+The final WSSSPE4 presentation was a panel %To investigate best practices for scientific software we brought together a panel 
+of five experts with different perspectives on ``Best Practices for Scientific Software.''
+These were: Alice Allen, Editor, Astrophysics Source Code Library, who brought an understanding of the difficulties of organizing a community and curating their software;
 Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
-Patricia Lago from Vrije Universiteit Amsterdam presented a keynote at the workshop (\S\ref{sec:keynote}) and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
-Karthik Ram from the Berkeley Institute for Data Science brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
-
-The panel opened with a question to the floor ``Who in the audience still codes in anger?'' which caused some confusion as, it turns out, that this is a turn of phrase that has not crossed the Atlantic.
-After translating this question into the more open, ``Who in the audience still writes code as part of their day-to-day work?'' we were able to confirm the suspicion that the WSSSPE audience is heavily involved writing, not just using, code.
-Hence, the panel were going to have to deal with questions from a highly experienced audience.
-
-The panel opened by noting that the world of scientific computing is changing rapidly.
-Each panelist was asked for their views on practices which had recently emerged and, if not handled correctly, would become the basis for the next generation's software problems?
-This quickly led to a discussion about the dangers of containerization.
-These technologies are pushed by some---a minority it is hoped---as a panacea for software sustainability.
-The panel and attendees generally agreed that containerization is useful, but should not be seen as the solution to sustainability.
-After all, these technologies are just as liable as any to decay.
-
-A later question looked at the common practices from software engineering that the scientific community should be using.
-The discussion that followed picked up on the key elements of basic sustainable practice: version control, documentation and ensuring that communities are active and engaged.
-When asked whether there was resistance from the research community to adopting these practices, few negative experiences were reported.
-From the panel's point of view, it would appear that the problem in getting researchers to employ sustainable practices lies more in getting the message out to researchers, rather than selling them on the benefits.
-
-The panel rounded up with a final question about ``If you wanted to have the most disastrous effect on scientific software, what one practice would you advise people to keep doing?''
-A few good responses were discussed, but the biggest laugh went to Mike Croucher who said ``Containerize everything!''
+Patricia Lago from Vrije Universiteit Amsterdam, who presented a keynote at the workshop  and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
+Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
 
@@
 \begin{enumerate}
-\item A vision on any aspect of the work.
+\item A vision on any aspect of the work
 \item A gap or challenge
@@
 (with one area of the room for each topic.)
-
-Once we had the walls covered in ideas, we had people organically form groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}.
-For those folk who worked on Vision and Gaps, after an hour, we encouraged them to finish their Vision or Gap Google Doc, and to find and join or start working on a project with a group.
-This led to the 12 groups that are discussed in the rest of this section, specifically:
-\begin{enumerate}
-\item Verifying best practices \& metrics for sustainable research software
-\item Software Sustainability Alliance
-\item Scientific Software Prototyping Infrastructure (S2PI)
-\item Standard metadata for software (CodeMeta)
-\item White paper on developing sustainable software
-\item Social science for scientific software
-\item Software best practices for undergraduates
-\item Meaningful metrics for sustainable software
-\item Coordinating access to CI for research software
-\item Software engineering processes tailored for research software
-\item Open research index
-\item Letters of evaluation for computational scientists
-\end{enumerate}
-
-The groups worked in Google docs, and eventually created Google slides to show their work back to the larger groups.
+%
+Next, attendees organically formed groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}.
+This led to the 12 groups that are discussed in the rest of this section.
+The groups then worked in Google docs, and eventually created Google slides to show their work back to the larger groups.
 The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}.
-In each group's discussion, a number of Specific, Measurable, Achievable, Relevant, and Timely (SMART) steps were created, and ideally, at least the first one (to set up some communication mechanism) was completed.
-In many cases, this was done through Slack (see \S\ref{sec:slack}.)
 
@@
 \input{wg_prototyping}
-\input{wg_codemeta}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{CodeMeta}
+\label{sec:CodeMeta}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Alice to write this}
+
+Research software is often not shared; that which is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides.
+The CodeMeta project\footnote{\url{http://codemeta.github.io/}} wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software, increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
+CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
+The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.
+
+The primary objective of this working group, which included people who had previously been working on the CodeMeta project, is to find ways to help the project come to fruition.
+%
+The working group plans to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
+Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation.
+Finally, the group seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.
+%
+While at WSSSPE4, the group greatly expanded the CodeMeta project README file\footnote{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
+
+Though a Google group mailing list has been established for the working group, the easiest way to engage with the CodeMeta project is through its Github repository\footnote{\url{https://github.com/codemeta/codemeta}}.
+
 \input{wg_best_practices_developing}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Organizing committee}  \label{sec:orgcom}
+\section{Organization}  \label{sec:orgcom}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-The following is the list of WSSSPE4 organizers.
+WSSSPE4 organizers were:
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Attendees}  \label{sec:attendees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The following is a list of attendees at WSSSPE4.
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{participants}
-\end{longtable}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Travel award recipients}  \label{sec:awardees}
+%\section{Attendees}  \label{sec:attendees}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%The following is a list of attendees at WSSSPE4.
+%
+%{\scriptsize
+%\begin{longtable}{lll}
+%\input{participants}
+%\end{longtable}
+%}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Travel award recipients}  \label{sec:awardees}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %\todo{Do we want email addresses here?}
-The following table contains the list of WSSSPE4 travel award recipients, where
-\textsuperscript{*} and \textsuperscript{\textdagger} indicate students and
-early-career researchers, respectively.
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{awardees}
-\end{longtable}
-}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Program committee}  \label{sec:progcom}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-The following table lists the WSSSPE4 program committee.
+%The following table contains the list of WSSSPE4 travel award recipients, where
+%\textsuperscript{*} and \textsuperscript{\textdagger} indicate students and
+%early-career researchers, respectively.
+%
+%{\scriptsize
+%\begin{longtable}{lll}
+%\input{awardees}
+%\end{longtable}
+%}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Program committee}  \label{sec:progcom}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+The WSSSPE4 program committee comprised:
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
typo

@@
 covers the set of working groups that formed during the meeting, along with contact information for readers who may want to join
-the group. Finally, we present results from a survey of the workshop attendees.
+a group. Finally, we present results from a survey of the workshop attendees.
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Attendees}  \label{sec:attendees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%The following is a list of attendees at WSSSPE4.
-%
-%{\scriptsize
-%\begin{longtable}{lll}
-%\input{participants}
-%\end{longtable}
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Travel award recipients}  \label{sec:awardees}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Do we want email addresses here?}
-%The following table contains the list of WSSSPE4 travel award recipients, where
-%\textsuperscript{*} and \textsuperscript{\textdagger} indicate students and
-%early-career researchers, respectively.
-%
-%{\scriptsize
-%\begin{longtable}{lll}
-%\input{awardees}
-%\end{longtable}
-%}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Program committee}  \label{sec:progcom}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 The WSSSPE4 program committee comprised:
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Detailed survey results}  \label{sec:survey_details}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\input{survey_details}
-
-%\input{wg_appendix_best_practices}
-
 \newpage

### 2017-05-11 WSSSPE4_report/wg_prototyping.tex
another arXiv fix, to be submitted later
another fix for arXiv

--- 2017-04-17 WSSSPE4_report/wg_prototyping.tex
@@
 scientific programmers, who are scarce~\cite{post2005computational}. However,
- the capacity to quickly envision and prototype applications by domain experts
- ---later to be fully implemented into by scientific programmers---is critical for
+the capacity to quickly envision and prototype applications by domain
+experts---later to be fully implemented into by scientific programmers---is critical for
 acquiring maturity and proficiency in the grand challenges that Exascale
@@
 
-Our general plan is to develop a functional (meta-)prototype before WSSSPE5, and
-if useful, depart from it towards a larger software infrastructure tied to
+The group's general plan is to develop a functional (meta-)prototype before WSSSPE5, and
+if useful, then develop a larger software infrastructure tied to
 particular cyberinfrastructure resources.
@@
 
-For joining to this project, please visit the GitHub repository at
+To join this project, please visit the GitHub repository at
 \url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel

### 2017-05-11 WSSSPE4_report/wg_social.tex
another report fix for arXiv

@@
 However, there is a gap between these two research and practice.
-Furthermore, we also recognize that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
+Furthermore, the group also recognize that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
shortening paper by reducing WG discussion

@@
 
-\input{wg_prototyping}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Scientific Software Prototyping Infrastructure (S2PI)}
+\label{sec:prototyping}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%from Santiago
+There is a productivity bottleneck---yet to be solved---in HPC from the human
+perspective, first identified in the 1980s~\cite{barstow1982automatic}. Of all
+domain-specific scientists, only a percentage of those use simulations due to
+perceptions mostly related to the difficulty of using and developing those tools.
+XSEDE and other resources have become a gateway for successfully increasing the
+basis of HPC scientific users by facilitating their access to infrastructure and
+tools~\cite{towns2014xsede}, but the development challenge remains.
+Of all scientists who use simulations, only a small percentage know how
+to develop prototypes or full applications that can be latter scaled by computer
+scientists and engineers. Writing new parallel applications is unusual for
+domain scientists and thus has created a complex dependency on specialized
+scientific programmers, who are scarce~\cite{post2005computational}. However,
+the capacity to quickly envision and prototype applications by domain
+experts---later to be fully implemented into by scientific programmers---is critical for
+acquiring maturity and proficiency in the grand challenges that Exascale
+attempts to solve, and moreover for allowing scientists to develop their own
+ideas quickly and productively~\cite{vinter2015prototyping}. The latter, in
+addition, would increase service utilization time in HPC systems, a critical
+measure of efficiency. Considering the existing code base in scientific
+computing, software infrastructures that allow a clean transition from legacy
+systems to new Exascale platforms while preserving flexible prototyping
+towards the future are central~\cite{hwu2015transitioning}.
+
+This group deems the ability to rapidly construct software artifacts that can be
+trusted in terms of the methods from their design up, easily discarded when
+wrong and extended when right at low human and computational cost to be one
+aspect of scientific software sustainability. In essence,
+a motto for scientific software prototyping is \textit{fail hard, fail fast}.
+%
+The group's objective is development of a prototype tool that allows domain scientists to generate
+simulation prototypes through a simple, yet expressive declarative
+strongly-typed programming language close to equational expressions. The result
+of that specification is both an executable artifact as well as a specification
+for scientific programmers to later flesh out completely and adapt to particular
+infrastructures. This tool will start in the domain of computational physics,
+and (probably) extend to other domains, and the code will be open source.
+%
+The underlying hypothesis of this project is that a strongly typed, declarative
+problem-specification language with adequate constructs, to be implemented
+through stencils and algorithmic skeletons tied to numerical methods, captures a
+ subset of useful and interesting problems in physics and other domains.
+%
+The group plans to develop a functional (meta-)prototype before WSSSPE5, and
+if useful, then develop a larger software infrastructure tied to
+particular cyberinfrastructure resources.
+
+To join this project, please visit the GitHub repository at
+\url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel
+\texttt{\#wg-sci-soft-proto}.
 
@@
 
-\input{wg_best_practices_developing}
-\input{wg_social}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{White paper on developing sustainable software}
+\label{sec:best-practices-developing}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Sandra to write this}
+
+Many diverse aspects and dimensions (such as economic, technical, environmental, and social) of developing sustainable software can be investigated,
+This group aims to write white papers
+that will focus on scientific environments and their implications, targeted at developers
+and project managers of scientific software. Given the complexity of this field, it is important to select
+a subset of sustainability aspects for the white papers. The idea is to create a series of papers instead of trying
+to tackle all important topics in one paper. 
+%
+While there are already a few papers available on best practices and sustainability of scientific software in general, the group's goal is to create a series of papers that lead to consensus in the community, tackle many diverse aspects, and stay up-to-date with new trends.
+%
+For the first white paper, the group aims to set the
+stage with successful use cases and an analysis of why they have been successful.
+Further topics will include community-related practices, government and management, funding,
+metrics, tools, and usability. 
+The resulting draft white paper will be distributed via the WSSSPE email list.
+After collecting feedback on it, the group's plans are to attract a wider community to contribute to an extended journal paper version and to investigate whether more people would like to be involved in the white paper series.
+
+The GitHub repository for the white paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
+For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Social science for scientific software}
+\label{sec:social}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{from Stuart}
+
+%Introduction to group here, including the overall objective of work in this area.
+
+There is more and more academic research being done on topics related to software sustainability, including work on software engineering practices and management of open source projects.
+However, academic research in general is often siloed for many reasons, and work on topics relevant to software sustainability is no exception.
+There are many academic studies and projects which may be relevant for practitioners working in this area, and many projects and initiatives by practitioners that may be relevant for social scientists.
+However, there is a gap between these two research and practice.
+%
+This working group is motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability.
+%
+ In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges. 
+The group aims to give social scientists and practitioners working on scientific software, software sustainability, and open source and a better understanding of each other's work, as well as help them connect and coordinate on specific projects of mutual interest.
+%
+Furthermore, the group also recognize that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
+
+The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
+This plan resulted in the creation of the following SMART steps listed in the next section.
+
+To join the group, join its mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
+
 \input{wg_best_practices_undergrads}

### 2017-05-11 WSSSPE4_report/wg_metrics.tex
removing more author name notes
small report fix for arXiv

@@
 
-\note{content from Emily}
+%\note{content from Emily}
 
@@
 
-Meaningful Metrics for Sustainable Scientific Software aims to increase the visibility on the quality of scientific software, facilitate the reusability of scientific software, and promote the best software practices by standardizing metrics via interviews with scientific software developers.
+Meaningful Metrics for Sustainable Scientific Software aims to increase the visibility of the quality of scientific software, facilitate the reusability of scientific software, and promote the best software practices by standardizing metrics via interviews with scientific software developers.
 This working group believes improving the current software metrics system will increase software sustainability.

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
removing note about detailed survey appendix
more shortening of WG subsections

@@
 
-\input{wg_best_practices_undergrads}
-\input{wg_metrics}
-\input{wg_access}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Software best practices for undergraduates}
+\label{sec:best-practices-undergrads}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Jonah Miller to write this}
+
+This working group was motivated by the perceived prevalence of
+so-called ``dark code'' in scientific communities: code written by
+individual researchers in an unsustainable way that is never shared
+with the larger community. Participants recalled their own experience
+working with colleagues who write dark code or inheriting
+unsustainable dark code from collaborators.
+%
+The question is, then, how to prevent researchers from writing dark
+code? The participants hypothesized that the best strategy is to catch
+researchers while they are still in training and teach them software
+best-practices. Therefore, this working group's goal
+is to develop courses on software best practices aimed at
+undergraduate students studying domain science. The program might be
+similar to a Software Carpentry or Data Carpentry workshop but 
+focused for domain scientists.
+
+While implementation of a course aimed at each domain science is a long term
+goal, a short-term, achievable goal, is to develop a
+curriculum for a course aimed at a single domain science. Since the
+participants in the working group have expertise in physics, this is a
+natural target.
+%
+The development of a successful curriculum relies on the expertise of
+software engineers to describe the best practices, domain experts to
+describe model problems and work-flow, and instructors to formulate
+the pedagogy. Ideally these people will be brought together for a
+short workshop or hackathon with the goal of drafting the
+curriculum. The
+hackathon will be implemented and a draft of the curriculum will be
+written within the next year.
+Later, organizational partners will be required to
+actually implement the program.
+
+Anyone interested in contributing should get in touch via one of the
+following methods:
+
+\begin{itemize}
+\item Join our Google Group/mailing list
+  \cite{WSSSPEUndergradGoogleGroup}.
+\item Join the WSSSPE Slack (\S\ref{sec:slack}) and the
+  \texttt{\#wg-undergraduatecourse} channel.
+\item Ask to join the WSSSPE4-undergraduate-course organization
+  \cite{WSSSPEUndergradGithub}.
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Meaningful metrics for sustainable software}
+\label{sec:metrics}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{content from Emily}
+
+This group aims to increase the visibility of the quality of scientific software, facilitate the reusability of scientific software, and promote the best software practices by standardizing metrics via interviews with scientific software developers.
+This working group believes improving the current software metrics system will increase software sustainability.
+Currently, there are inefficiencies regarding software duplication, sustainability, and selection, as well as others, within the scientific software community.
+In order to address these inefficiencies, the group aims to create a goal-oriented method to collecting productive metrics by focusing on the developer side of software.
+
+In the context of this aim, the group hopes to find an efficient solution for streamlining the process of collecting and utilizing metrics to benefit software sustainability, as well as minimize the current inefficiencies within the scientific software community.
+This will improve software evaluation and comparison, thus reducing the effort spent on seeking scientific software.
+%
+The group plans on surveying and interviewing scientific software developers to form metrics from the goals they have for their software, then to publicize the results.
+
+For more information, please contact Emily Chen at <echen35@illinois.edu>.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Coordinating access to CI for research software}
+\label{sec:access}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{%Mark Abraham to write this?  Needs confirmation or another volunteer -
+%Dan starting this for now - may poll other members of group to make progress}
+
+Each developer of software with uncommon needs (hardware, software, libraries, data sets), non-public code, or tests that exceed time limits for free plans must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
+For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
+(Note that Debian provides a testing infrastructure (\url{https://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.)
+
+This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems (when publicly available CI are not a fit), by coordinating and sharing this burden across multiple projects.
+%
+The scope of the group's interest is any type of testing, though interactive access for troubleshooting would be of particular interest beyond just automated testing.
+
+However, there are a lot of open issues:
+\begin{itemize}
+\item Some similar work has already been done.  How can this group apply and/or learn from that work, rather than reinventing the wheel?
+\item How to ensure this will work across disciplines?
+\item Meaningful CI for large projects may need hundreds of CPU hours per day
+\end{itemize}
+
+Some possible goals include:
+\begin{itemize}
+\item Acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to e.g. Debian's testing infrastructure or to a shared Jenkins-based infrastructure
+\item Extending Debian's scope to include published but not mature software
+\item Implementing the same interface but with specialized hardware or available to non-public codes.
+\item \url{https://reproducible-builds.org} but for CI
+\end{itemize}
+
+For more information, see \url{https://groups.google.com/forum/#!forum/continuous-integration-for-research-software}
+
 \input{wg_soft_eng}
@@
 The survey had a 100\% completion rate.
-Appendix~\ref{sec:survey_details} contains the detailed survey results.
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
finished first draft of short working group sections

@@
 
-\input{wg_soft_eng}
-\input{wg_open_research_index}
-\input{wg_letters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Software engineering processes tailored for research software}
+\label{sec:soft-eng}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Anshu to write this}
+
+This working group is concerned with identifying processes that are not adequately
+covered by general software engineering, starting with verification and testing.
+Computational science and engineering applications have many moving
+parts that need to interoperate with one another. The accuracy and
+reliability of results produced by scientific software depends not
+only on the individual components behaving correctly, but also on the
+validity of their interactions.
+As scientific understanding grows,
+the corresponding computational software models are refined, leading
+to more complex codes. Increasing complexity makes them more prone to
+defects, not only in individual code units, but also in interaction
+among units. Therefore, a strong verification process combined with a
+rigorous testing regime plays a critical role in the prevention of
+generating incorrect scientific results. However, most science teams
+struggle to find a good solution for themselves.
+
+One cause is lack of exposure to the practices. While
+good developers will test their code to
+verify that it operates as expected, they may not appreciate
+that without regular testing, defects can be introduced
+inadvertently. 
+%
+An even bigger challenge is that those who understand
+the importance of regular testing do not often find much help from
+software engineering literature. There is a significant gap between
+the testing gospel and its applicability to computational science. This
+gap leads to frustration and abandonment of the good with the bad. Some
+relevant literature exists, in particular experiences from
+practitioners in computational science who developed their own
+solutions. However, this literature is scattered among many different
+forums, and can be challenging to find. This working group aims to
+address this gap by curating the existing content and contributing
+content where none exists.
+
+This working group will (1) conduct a literature survey to gauge the extent
+of awareness of the issue in general, (2) generate content useful for
+the community where needed, and (3) curate the collected and added
+content for the use of the community.
+
+Readers interested in getting more information should use the working group'
+channel in the WSSSPE Slack (\S\ref{sec:slack}), called
+\texttt{\#wg-testing-in-science}.
+Additionally, a git
+repository (\url{https://github.com/WSSSPE/WG-Best-Practices.git}) exists for contributing content and reference to, and
+curation of the existing literature on this topic.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Open research index}
+\label{sec:open-research-index}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%\note{Dan to write this}
+
+The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
+Its goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.
+%
+The Open Research Index should take in all research products (papers, software, datasets, workflows, etc.) from their publishers and recorders (journals, societies, domain repositories, government [open access] repositories, preprint servers, general repositories [e.g., figshare, zenodo]) and other services (CrossRef, ORCID).
+Each product should list authors and citations and allow people to search the resulting network.
+Users should also be able to interact with their own record and edit it, like Google Scholar allows.
+
+The working group's plans are relatively simple to express, though quite complex to undertake.
+They are to first determine a plan to build an open research index that allows various stakeholders to satisfy their needs, then to determine if the plan is feasible.
+At the time of the meeting, and today as well, it is unclear who has the time and energy to pursue this idea. Thus, a leader needs to be identified.
+With a leader and feasible plan, the group would then obtain resources, and then build the index.
+%
+Because Google and others provide some similar services today, though these services (and the underlying data) could be removed at any time and the community cannot build new services, these companies would ideally be involved in the activities of the group.
+
+Alternatively or additionally, the group could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
+At WSSSPE4, there seemed to be enough interest to do this, so the group set up a Slack channel within the WSSSPE team (\S\ref{sec:slack}), called \texttt{\#wg-open-research-idx} for its members and any future members.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\subsection{Letters of evaluation for computational scientists}
+\label{sec:letters}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Scientists working on scientific software are often located in
+disciplinary departments, depending on whether their software
+originates from the mathematical, physical, chemical, or other
+disciplines. As a consequence, they are frequently outside the core
+areas of their science, and their contributions are typically to both
+the research activity their software enables, as well as on
+algorithm and implementation aspects. This presents issues when
+letters of evaluation for hiring, tenure, and promotion do not specifically
+cover how this is relevant to the discipline.\footnote{The extensive use of such letters, and the problems that are
+  associated with it, may be an issue specific to the United States.}
+
+This group believes that the authors of scientific software provide important
+services to departments that are no less than strictly disciplinary
+research. Consequently, in order to level the playing field with more
+disciplinary candidates for hires, tenure, and promotion requires that
+letter writers be aware of how their letters will be read by
+committees. It also requires that committees be aware that such
+letters often look different and may provide a different perspective
+of how a candidate's achievements should be assessed. For example, in
+mathematics a typical candidate would be evaluated on the difficulty
+and depth of the statements she may have proven in their papers. On
+the other hand, an author of mathematical software would likely be
+evaluated based on the impact of her software, or the number of citations of
+the publications that describe it. She may also be evaluated by how
+widely the software is used \textit{outside} mathematics, a criterion
+that is rarely used for more disciplinary mathematicians.
+
+Addressing this problem likely requires that letter writers pay
+particular attention to who exactly the audience of such letters is,
+and tailor the message by specifically highlighting how the work of a
+candidate benefits the department and discipline that is considering a
+candidate. On the other hand, committee members also need to pay
+attention to the fact that there are areas that are important to the mission of
+their department and professional community in which different
+standards for evaluation hold.
+%
+There are essentially two important strategies that build on each
+other: (i) raising awareness of the problem beyond just those who are
+affected by it (namely, computational scientists working on scientific
+software), and (ii) providing letter writers, letter readers, and
+evaluating committees with guidance on what criteria are relevant in
+assessing scientific software authors.
+
+Concrete guidance is most valuable if it
+comes from respected bodies such as professional societies or
+established and respected organizations. Getting these to act
+will only be possible if the problem is widely acknowledged, and so
+the first of the goals above should be the current focus.
+%
+The group will attempt to address it by organizing sessions at conferences
+that address the career problem, as well as writing editorials that
+can be published in the magazines of professional societies. These
+editorials ought to outline best practices for letter writers that
+specifically (i) make clear the contribution of a candidate to their
+interdisciplinary area, (ii) the relevance to their home discipline, and
+(iii) why writing software is good for the discipline itself.
+
+Ultimately, building a community large enough
+to affect change is important; contact Wolfgang Bangerth at
+<bangerth@colostate.edu> if you are interested in helping.
 

### 2017-05-11 WSSSPE4_report/WSSSPE4_paper.tex
small tweaks

@@
 After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
-attendees use sticky flip charts and sticky notes on the walls to suggest
-\begin{enumerate}
-\item A vision on any aspect of the work
-\item A gap or challenge
-\item A project idea
-\end{enumerate}
-(with one area of the room for each topic.)
-%
-Next, attendees organically formed groups around the flip charts on the wall\footnote{Photos of the flipcharts: \url{https://github.com/WSSSPE/media/tree/master/WSSSPE4-photos}}.
-This led to the 12 groups that are discussed in the rest of this section.
-The groups then worked in Google docs, and eventually created Google slides to show their work back to the larger groups.
-The docs and slides were stored in a common Google Drive folder\footnote{\url{https://drive.google.com/drive/u/0/folders/0BzIpChHwv-q5X3lRaWhYMXJZaUE}}.
-
-%\todo{need to confirm who will write Access subsection (\S\ref{sec:access})}
+attendees use sticky flip charts and sticky notes on the walls to suggest: (1) a vision on any aspect of the work, (2) a gap or challenge, or (3) a project idea.
+Next, attendees organically formed groups around the flip charts on the walls.  This led to the 12 groups that are discussed in the rest of this section.
+Summaries of each group's activities follow; for more details
+about the group discussions see \cite{WSSSPE4-report}.
 
@@
 \begin{itemize}
-\item Join our Google Group/mailing list
+\item Join the Google Group/mailing list
   \cite{WSSSPEUndergradGoogleGroup}.
@@
 For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
-(Note that Debian provides a testing infrastructure (\url{https://ci.debian.net/doc/}) that is mature and supports hardware and other requirement specifications.)
+(Note that Debian provides a testing infrastructure\footnote{\url{https://ci.debian.net/doc/}} that is mature and supports hardware and other requirement specifications.)
 

### 2017-05-13 WSSSPE4_report/WSSSPE4_paper.tex
trying to make "Definitions" bit look nicer
updating abstract

@@
 \begin{abstract}
-This report records and discusses the Fourth Workshop on Sustainable
-Software for Science: Practice and Experiences (WSSSPE4). The report
-includes a description of the keynote presentation of the workshop, the
-mission and vision statements that were drafted at the workshop and
-finalized shortly after it, a set of idea papers, position papers, experience
-papers, demos, and lightning talks, and a panel discussion.  The main part of the report
-covers the set of working groups that formed during the meeting, along with contact information for readers who may want to join
-a group. Finally, we present results from a survey of the workshop attendees.
+
+This article summarizes motivations, organization, and activities of the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4).
+The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
+This article includes the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
+The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group.
+Finally, it discusses a survey of the workshop attendees.
 
@@
 
-{\bf Definitions}
-
-Sustainable software has the capacity to endure such that it will continue to be available in the future, on new platforms, meeting new needs.
-
-The research software lifecycle includes:
+\textbf{Definitions:}
+\emph{Sustainable software} has the capacity to endure such that it will continue to
+be available in the future, on new platforms, meeting new needs.
+The \emph{research software lifecycle} includes:
 \begin{itemize}

### 2017-05-13 WSSSPE4_report/WSSSPE4_paper.tex
more minor shortening

@@
 The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
+The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
 
@@
 The \emph{research software lifecycle} includes:
-\begin{itemize}
-\item acquiring and assembling resources (including funding and people) into teams and communities
-\item developing software
-\item using software
-\item recognizing contributions to and of software
-\item maintaining software
-\end{itemize}
+acquiring and assembling resources (including funding and people) into teams and communities,
+developing software,
+using software,
+recognizing contributions to and of software,
+and
+maintaining software.
 
@@
 After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
-attendees use sticky flip charts and sticky notes on the walls to suggest: (1) a vision on any aspect of the work, (2) a gap or challenge, or (3) a project idea.
-Next, attendees organically formed groups around the flip charts on the walls.  This led to the 12 groups that are discussed in the rest of this section.
+attendees use sticky flip charts and sticky notes on the walls to suggest a vision on any aspect of the work, a gap or challenge, or a project idea.
+Next, attendees organically formed groups around the flip charts, and 12 groups emerged.
 Summaries of each group's activities follow; for more details

### 2017-05-13 WSSSPE4_report/WSSSPE4_paper.tex
fixes to problems found by Alice

@@
 were accepted as lightning talks.)
-The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
+The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
 
@@
 The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
+The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world would look like if the organization is successful; and from which, focus areas could be used to establish the scope of the organization along with its goals.
 
@@
 Of all scientists who use simulations, only a small percentage know how
-to develop prototypes or full applications that can be latter scaled by computer
+to develop prototypes or full applications that can be later scaled by computer
 scientists and engineers. Writing new parallel applications is unusual for
@@
 There are many academic studies and projects which may be relevant for practitioners working in this area, and many projects and initiatives by practitioners that may be relevant for social scientists.
-However, there is a gap between these two research and practice.
+However, there is a gap between these two in research and practice.
 %
@@
 This working group was motivated by the perceived prevalence of
-so-called ``dark code'' in scientific communities: code written by
+so-called ``hidden code'' in scientific communities: code written by
 individual researchers in an unsustainable way that is never shared
 with the larger community. Participants recalled their own experience
-working with colleagues who write dark code or inheriting
-unsustainable dark code from collaborators.
-%
-The question is, then, how to prevent researchers from writing dark
+working with colleagues who write hidden code or inheriting
+unsustainable hidden code from collaborators.
+%
+The question is, then, how to prevent researchers from writing hidden
 code? The participants hypothesized that the best strategy is to catch
@@
 services to departments that are no less than strictly disciplinary
-research. Consequently, in order to level the playing field with more
+research. Consequently, leveling the playing field with more
 disciplinary candidates for hires, tenure, and promotion requires that

### 2017-05-13 WSSSPE4_report/WSSSPE4_report.tex
updated definitions formatting
changes for arXiv report

@@
 and the lightning talks are discussed in Section~\ref{sec:lightning}.
-The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
+The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
 
@@
 The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
+The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world would look like if the organization is successful; and from which, focus areas could be used to establish the scope of the organization along with its goals.
 

### 2017-05-16 WSSSPE4_report/WSSSPE4_paper.tex
one more change from Alice
a bit more shortening, based on suggestions from Anshu

@@
 held in Boulder, Colorado, USA.
-Note that the first WSSSPE workshop was named
+The first WSSSPE workshop was named
 ``Working towards
@@
 %\todo{did we reach out to any others?}
-
+%
 WSSSPE4 also included two professional event organizers/facilitators who helped
@@
 
-At the workshop, we also introduced a Code of Conduct (CoC).%
+At the workshop, we introduced a Code of Conduct (CoC).%
 \footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
-The CoC was conceived for the workshop itself; however, we intend it
-to also be a guideline for the community of scientists which WSSSPE
-supports, and their personal and online interactions (e.g., on
+The CoC was conceived for the workshop and a guideline for the community of scientists which WSSSPE
+supports, including their personal and online interactions (e.g., on
 Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
-FORCE11 conference CoC~\cite{FORCE11:CoC}, in turn based on the Code4Lib
+FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
 CoC~\cite{Code4Lib:CoC}.
@@
 
-WSSSPE4 was based on the work done at WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
+WSSSPE4 was based on the work done at previous WSSSPE meetings, but aimed
 at producing working groups that better continued working after the workshop ended.
-In addition, based on feedback after WSSSPE3, it became clear that WSSSPE attendees
-had two different motivations in participating.  One motivation was to make a better future
-for research software, and the other was to immediately do better research software development.
+In addition, based on feedback after WSSSPE3, it became clear that attendees
+had two distinct motivations:  to make a better future
+for research software, and to immediately do better research software development.
 This led to the idea of WSSSPE4 being partially divided into two tracks:
@@
     \textbf{Track 1 -- Building a sustainable future for open-use research
-    software} has the goals of defining a vision of the future of open-use
+    software}: defining a vision of the future of open-use
     research software, and in the workshop, initiating the activities that are
-    needed to get there. The idea of this track is to first think about where
-    we want to be 5 to 10 years from now, without being too concerned with
-    where we are today, and then to determine how we can move to this future.
-
-    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
-    has the goal of improving the quality of today's research software and the
+    needed to get there.
+
+    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
+    improving the quality of today's research software and the
     experiences of its developers by sharing best practices and experiences.
-    This track is focused on the current state of scientific software and what
-    we can do to improve it in the short term, starting with where we are today.
 \end{quote}
 
-The call for participation requested
-idea papers, which present implementable proposals related to Track~1;
-position papers, which are longer, not previously published papers related to
-Track~2 specifically discussing what we can do to improve sustainable scientific
-software in the short term, starting with where we are today;
-experience papers, longer papers related to Track~2 that discuss current
-practices and experiences and how they have been used to improve the quality of
+The call for participation requested:
+idea papers, implementable proposals related to Track~1;
+position papers, longer, not previously published papers
+discussing what we can do to improve sustainable scientific
+software in the short term;
+experience papers, longer papers that discuss current
+experiences and how they have been used to improve the quality of
 today's research software and/or the experiences of its developers;
 demos, brief papers describing a tool or
-process that would be demonstrated, relevant to Track~2 that improves the quality of today's research
+process that would be demonstrated that improves the quality of today's research
 software and\slash or the experiences of its developers; and
@@
 
-The following list of contribution topics was suggested, though this was not intended to limit potential submissions:
-
-%
-%\begin{quote}
+Suggested contribution topics included:
+
 \begin{itemize}
@@
 
-\item Development and Community
-\begin{itemize}
-    \item Best practices for developing sustainable software
-    \item Models for funding specialist expertise in software collaborations
-    \item Software tools that aid sustainability
-    \item Academia/industry interaction
-    \item Refactoring/improving legacy scientific software
-    \item Engineering design for sustainable software
-    \item Metrics for the success of scientific software
-    \item Adaptation of mainstream software practices for scientific software
-\end{itemize}
-
-\item Professionalization
-\begin{itemize}
-    \item Career paths
-    \item RSE as a brand
-    \item RSE outside of the UK or Europe
-    \item Increase incentives in publishing, funding and promotion for better software
-\end{itemize}
-
-\item Training
-\begin{itemize}
-    \item Training for developing sustainable software
-    \item Curriculum for software sustainability
-\end{itemize}
-
-\item Credit
-\begin{itemize}
-    \item Making the existing credit and citation ecosystem work better for software
-    \item Future credit and citation ecosystem
-    \item Software contributions as a part of tenure review
-    \item Case studies of receiving credit for software contributions
-    \item Awards and recognition that encourage sustainable software
-\end{itemize}
-
-\item Software publishing
-\begin{itemize}
-    \item Journals and alternative venues for publishing software
-    \item Review processes for published software
-\end{itemize}
-
-\item Software discoverability/reuse
-\begin{itemize}
-    \item Proposals and case studies
-\end{itemize}
-
-\item Reproducibility and testing
-\begin{itemize}
-    \item Reproducibility in conferences and journals
-    \item Best practices for code testing and code review
-\end{itemize}
+\item Development and Community:
+best practices for developing sustainable software;
+models for funding specialist expertise in software collaborations;
+software tools that aid sustainability;
+academia/industry interaction;
+refactoring/improving legacy scientific software;
+engineering design for sustainable software;
+metrics for the success of scientific software; and
+adaptation of mainstream software practices for scientific software
+
+\item Professionalization:
+career paths;
+RSE as a brand;
+RSE outside of the UK or Europe; and
+Increase incentives in publishing, funding and promotion for better software
+
+\item Training:
+taining for developing sustainable software; and 
+curriculum for software sustainability
+
+\item Credit:
+making the existing credit and citation ecosystem work better for software;
+future credit and citation ecosystem;
+software contributions as a part of tenure review;
+case studies of receiving credit for software contributions; and
+awards and recognition that encourage sustainable software
+
+\item Software publishing:
+journals and alternative venues for publishing software; and
+review processes for published software
+
+\item Software discoverability/reuse:
+proposals and case studies
+
+\item Reproducibility and testing:
+reproducibility in conferences and journals; and
+best practices for code testing and code review
 
@@
 
-Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
+Going into WSSSPE4, WSSSPE had no formal mission or vision statement.
 The organizers developed a strawhorse, which was presented to the participants early in the meeting.
@@
 
-A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
+A final draft was put on GitHub and advertised
 via the WSSSPE mailing list, Facebook group, and Twitter.
-After 2 weeks without suggested changes, the final statements, as shown below, were added
+After two weeks without suggested changes, the final statements, as shown below, were added
 to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
@@
 %
-While at WSSSPE4, the group greatly expanded the CodeMeta project README file\footnote{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
+While at WSSSPE4, the group greatly expanded the CodeMeta project README file to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Coordinating access to CI for research software}
+\subsection{Coordinating access to continuous integration (CI) for research softwaree}
 \label{sec:access}

### 2017-05-17 WSSSPE4_report/WSSSPE4_paper.tex
small fixes for both paper and report

@@
 (WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
-held over 2 1/2 days on 12--14 September 2016 in Manchester, England, UK.
+held over 2 1/2 days on 12--14 September 2016 in Manchester, England.
 This location and date was selected so that WSSSPE4 immediately preceded the
@@
 \footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
-The CoC was conceived for the workshop and a guideline for the community of scientists which WSSSPE
+The CoC was conceived for the workshop and a guideline for the community of scientists that WSSSPE
 supports, including their personal and online interactions (e.g., on

### 2017-05-17 WSSSPE4_report/WSSSPE4_report.tex
small fixes for both paper and report

@@
 The CoC was conceived for the workshop itself; however, we intend it
-to also be a guideline for the community of scientists which WSSSPE
+to also be a guideline for the community of scientists that WSSSPE
 supports, and their personal and online interactions (e.g., on

### 2017-05-17 WSSSPE4_report/WSSSPE4_paper.tex
fixing typo
another error

@@
 Based on this feedback, the committee redrafted the mission and vision.
-The committee added focus areas to clarify the organizations breadth and to keep the mission and vision simple and long-lasting.
+The committee added focus areas to clarify the organization's breadth and to keep the mission and vision simple and long-lasting.
 Comments after the close of the meeting were incorporated into the statements.
@@
 
-WSSSPE included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
+WSSSPE4 included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
 two experience papers, and three demos) that addressed a wide range of topics around

### 2017-05-18 WSSSPE4_report/WSSSPE4_paper.tex
another typo
minor date formatting change

--- 2017-05-16 WSSSPE4_report/WSSSPE4_paper.tex
@@
 \begin{abstract}
-
-This article summarizes motivations, organization, and activities of the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4).
-The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
-This article includes the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
-The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group.
-Finally, it discusses a survey of the workshop attendees.
+This report records and discusses the Fourth Workshop on Sustainable
+Software for Science: Practice and Experiences (WSSSPE4). The report
+includes a description of the keynote presentation of the workshop, the
+mission and vision statements that were drafted at the workshop and
+finalized shortly after it, a set of idea papers, position papers, experience
+papers, demos, and lightning talks, and a panel discussion.  The main part of the report
+covers the set of working groups that formed during the meeting, along with contact information for readers who may want to join
+a group. Finally, we present results from a survey of the workshop attendees.
 
@@
 held in Boulder, Colorado, USA.
-The first WSSSPE workshop was named
+Note that the first WSSSPE workshop was named
 ``Working towards
@@
 %\todo{did we reach out to any others?}
-%
+
 WSSSPE4 also included two professional event organizers/facilitators who helped
@@
 
-At the workshop, we introduced a Code of Conduct (CoC).%
+At the workshop, we also introduced a Code of Conduct (CoC).%
 \footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
-The CoC was conceived for the workshop and a guideline for the community of scientists which WSSSPE
-supports, including their personal and online interactions (e.g., on
+The CoC was conceived for the workshop itself; however, we intend it
+to also be a guideline for the community of scientists which WSSSPE
+supports, and their personal and online interactions (e.g., on
 Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
-FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
+FORCE11 conference CoC~\cite{FORCE11:CoC}, in turn based on the Code4Lib
 CoC~\cite{Code4Lib:CoC}.
@@
 
-WSSSPE4 was based on the work done at previous WSSSPE meetings, but aimed
+WSSSPE4 was based on the work done at WSSSPE1, WSSSPE2, and WSSSPE3, but aimed
 at producing working groups that better continued working after the workshop ended.
-In addition, based on feedback after WSSSPE3, it became clear that attendees
-had two distinct motivations:  to make a better future
-for research software, and to immediately do better research software development.
+In addition, based on feedback after WSSSPE3, it became clear that WSSSPE attendees
+had two different motivations in participating.  One motivation was to make a better future
+for research software, and the other was to immediately do better research software development.
 This led to the idea of WSSSPE4 being partially divided into two tracks:
@@
     \textbf{Track 1 -- Building a sustainable future for open-use research
-    software}: defining a vision of the future of open-use
+    software} has the goals of defining a vision of the future of open-use
     research software, and in the workshop, initiating the activities that are
-    needed to get there.
-
-    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
-    improving the quality of today's research software and the
+    needed to get there. The idea of this track is to first think about where
+    we want to be 5 to 10 years from now, without being too concerned with
+    where we are today, and then to determine how we can move to this future.
+
+    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}
+    has the goal of improving the quality of today's research software and the
     experiences of its developers by sharing best practices and experiences.
+    This track is focused on the current state of scientific software and what
+    we can do to improve it in the short term, starting with where we are today.
 \end{quote}
 
-The call for participation requested:
-idea papers, implementable proposals related to Track~1;
-position papers, longer, not previously published papers
-discussing what we can do to improve sustainable scientific
-software in the short term;
-experience papers, longer papers that discuss current
-experiences and how they have been used to improve the quality of
+The call for participation requested
+idea papers, which present implementable proposals related to Track~1;
+position papers, which are longer, not previously published papers related to
+Track~2 specifically discussing what we can do to improve sustainable scientific
+software in the short term, starting with where we are today;
+experience papers, longer papers related to Track~2 that discuss current
+practices and experiences and how they have been used to improve the quality of
 today's research software and/or the experiences of its developers;
 demos, brief papers describing a tool or
-process that would be demonstrated that improves the quality of today's research
+process that would be demonstrated, relevant to Track~2 that improves the quality of today's research
 software and\slash or the experiences of its developers; and
@@
 
-Suggested contribution topics included:
-
+The following list of contribution topics was suggested, though this was not intended to limit potential submissions:
+
+%
+%\begin{quote}
 \begin{itemize}
@@
 
-\item Development and Community:
-best practices for developing sustainable software;
-models for funding specialist expertise in software collaborations;
-software tools that aid sustainability;
-academia/industry interaction;
-refactoring/improving legacy scientific software;
-engineering design for sustainable software;
-metrics for the success of scientific software; and
-adaptation of mainstream software practices for scientific software
-
-\item Professionalization:
-career paths;
-RSE as a brand;
-RSE outside of the UK or Europe; and
-Increase incentives in publishing, funding and promotion for better software
-
-\item Training:
-taining for developing sustainable software; and 
-curriculum for software sustainability
-
-\item Credit:
-making the existing credit and citation ecosystem work better for software;
-future credit and citation ecosystem;
-software contributions as a part of tenure review;
-case studies of receiving credit for software contributions; and
-awards and recognition that encourage sustainable software
-
-\item Software publishing:
-journals and alternative venues for publishing software; and
-review processes for published software
-
-\item Software discoverability/reuse:
-proposals and case studies
-
-\item Reproducibility and testing:
-reproducibility in conferences and journals; and
-best practices for code testing and code review
+\item Development and Community
+\begin{itemize}
+    \item Best practices for developing sustainable software
+    \item Models for funding specialist expertise in software collaborations
+    \item Software tools that aid sustainability
+    \item Academia/industry interaction
+    \item Refactoring/improving legacy scientific software
+    \item Engineering design for sustainable software
+    \item Metrics for the success of scientific software
+    \item Adaptation of mainstream software practices for scientific software
+\end{itemize}
+
+\item Professionalization
+\begin{itemize}
+    \item Career paths
+    \item RSE as a brand
+    \item RSE outside of the UK or Europe
+    \item Increase incentives in publishing, funding and promotion for better software
+\end{itemize}
+
+\item Training
+\begin{itemize}
+    \item Training for developing sustainable software
+    \item Curriculum for software sustainability
+\end{itemize}
+
+\item Credit
+\begin{itemize}
+    \item Making the existing credit and citation ecosystem work better for software
+    \item Future credit and citation ecosystem
+    \item Software contributions as a part of tenure review
+    \item Case studies of receiving credit for software contributions
+    \item Awards and recognition that encourage sustainable software
+\end{itemize}
+
+\item Software publishing
+\begin{itemize}
+    \item Journals and alternative venues for publishing software
+    \item Review processes for published software
+\end{itemize}
+
+\item Software discoverability/reuse
+\begin{itemize}
+    \item Proposals and case studies
+\end{itemize}
+
+\item Reproducibility and testing
+\begin{itemize}
+    \item Reproducibility in conferences and journals
+    \item Best practices for code testing and code review
+\end{itemize}
 
@@
 were accepted as lightning talks.)
-The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
+The paper and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
 
@@
 
-Going into WSSSPE4, WSSSPE had no formal mission or vision statement.
+Going into WSSSPE4, the WSSSPE organization had not had a formal mission or vision statement.
 The organizers developed a strawhorse, which was presented to the participants early in the meeting.
 The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world would look like if the organization is successful; and from which, focus areas could be used to establish the scope of the organization along with its goals.
+The guidelines were that in general, an organization's mission should stand the test of time and state what is wrong and how the organization is going to fix it; its vision should imagine what the world look like if the organization is successful; and from which, focus areas could be use to establish the scope of the organization along with its goals.
 
@@
 
-A final draft was put on GitHub and advertised
+A final draft was put on GitHub (\url{https://github.com/WSSSPE/mission-vision}) and advertised
 via the WSSSPE mailing list, Facebook group, and Twitter.
@@
 The \emph{research software lifecycle} includes:
-acquiring and assembling resources (including funding and people) into teams and communities,
-developing software,
-using software,
-recognizing contributions to and of software,
-and
-maintaining software.
+\begin{itemize}
+\item acquiring and assembling resources (including funding and people) into teams and communities
+\item developing software
+\item using software
+\item recognizing contributions to and of software
+\item maintaining software
+\end{itemize}
 
@@
 After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
-attendees use sticky flip charts and sticky notes on the walls to suggest a vision on any aspect of the work, a gap or challenge, or a project idea.
-Next, attendees organically formed groups around the flip charts, and 12 groups emerged.
+attendees use sticky flip charts and sticky notes on the walls to suggest: (1) a vision on any aspect of the work, (2) a gap or challenge, or (3) a project idea.
+Next, attendees organically formed groups around the flip charts on the walls.  This led to the 12 groups that are discussed in the rest of this section.
 Summaries of each group's activities follow; for more details
@@
 Of all scientists who use simulations, only a small percentage know how
-to develop prototypes or full applications that can be later scaled by computer
+to develop prototypes or full applications that can be latter scaled by computer
 scientists and engineers. Writing new parallel applications is unusual for
@@
 %
-While at WSSSPE4, the group greatly expanded the CodeMeta project README file to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
+While at WSSSPE4, the group greatly expanded the CodeMeta project README file\footnote{\url{https://github.com/codemeta/codemeta/blob/master/README.md}} to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
 
@@
 There are many academic studies and projects which may be relevant for practitioners working in this area, and many projects and initiatives by practitioners that may be relevant for social scientists.
-However, there is a gap between these two in research and practice.
+However, there is a gap between these two research and practice.
 %
@@
 This working group was motivated by the perceived prevalence of
-so-called ``hidden code'' in scientific communities: code written by
+so-called ``dark code'' in scientific communities: code written by
 individual researchers in an unsustainable way that is never shared
 with the larger community. Participants recalled their own experience
-working with colleagues who write hidden code or inheriting
-unsustainable hidden code from collaborators.
-%
-The question is, then, how to prevent researchers from writing hidden
+working with colleagues who write dark code or inheriting
+unsustainable dark code from collaborators.
+%
+The question is, then, how to prevent researchers from writing dark
 code? The participants hypothesized that the best strategy is to catch
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Coordinating access to continuous integration (CI) for research softwaree}
+\subsection{Coordinating access to CI for research software}
 \label{sec:access}
@@
 services to departments that are no less than strictly disciplinary
-research. Consequently, leveling the playing field with more
+research. Consequently, in order to level the playing field with more
 disciplinary candidates for hires, tenure, and promotion requires that
@@
 
-All participants present on the last day of the meeting, September 14, 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
+All participants present on the last day of the meeting, 14 September 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
 44 responses were recorded, with one respondent indicating that they took the survey twice.

### 2017-05-18 WSSSPE4_report/WSSSPE4_paper.tex

--- 2017-05-17 WSSSPE4_report/WSSSPE4_paper.tex
@@
 Summaries of each group's activities follow; for more details
-about the group discussions see \cite{WSSSPE4-report}.
+about the group discussions, see \cite{WSSSPE4-report}.
 

### 2017-05-18 WSSSPE4_report/wg_best_practices_undergrads.tex
removing more author name notes
small fix for arXiv

@@
 
-\note{Jonah Miller to write this}
+%\note{Jonah Miller to write this}
 
@@
 This working group was motivated by the perceived prevalence of
-so-called ``dark code'' in scientific communities: code written by
+so-called ``hidden code'' in scientific communities: code written by
 individual researchers in an unsustainable way that is never shared
 with the larger community. Participants recalled their own experience
-working with colleagues who write dark code or inheriting
-unsustainable dark code from collaborators.
+working with colleagues who write hidden code or inheriting
+unsustainable hidden code from collaborators.
 
-The question is, then, how to prevent researchers from writing dark
+The question is, then, how to prevent researchers from writing hidden
 code? The participants hypothesized that the best strategy is to catch
@@
   \cite{WSSSPEUndergradGoogleGroup}.
-\item Join the WSSSPE Slack \cite{WSSSPESlack} and sign into the
-  \texttt{\#wg-undergraduatecourse} channel and say hi.
+\item Join the WSSSPE Slack (\S\ref{sec:slack}) and the
+  \texttt{\#wg-undergraduatecourse} channel.
 \item Ask to join the WSSSPE4-undergraduate-course organization

### 2017-11-29 WSSSPE4_report/WSSSPE4_paper.tex
fixing typo
3 fixes (2 typos & # of attendees) suggested by JORS reviewer B

@@
 (WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
-held over 2 1/2 days on 12--14 September 2016 in Manchester, England.
+held over 2 1/2 days on 12--14 September 2016 in Manchester, England, with 68 attendees.
 This location and date was selected so that WSSSPE4 immediately preceded the
@@
 the capacity to quickly envision and prototype applications by domain
-experts---later to be fully implemented into by scientific programmers---is critical for
+experts---later to be fully implemented by scientific programmers---is critical for
 acquiring maturity and proficiency in the grand challenges that Exascale
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Coordinating access to continuous integration (CI) for research softwaree}
+\subsection{Coordinating access to continuous integration (CI) for research software}
 \label{sec:access}
@@
 The survey contained nine multiple choice and open-text answer questions.
-Text responses have been alphabetize to preserve anonymity.
+Text responses have been alphabetized to preserve anonymity.
 The survey had a 100\% completion rate.

### 2017-12-17 WSSSPE4_report/WSSSPE4_paper.tex
incorporated some referee comments

@@
 who would not otherwise be able to attend the workshop. These
-funds allowed 29 additional people to attend and participate. A subset of the
-organizing committee reviewed 44 applications for travel support and
-competitively selected the awardees, including 11 students and five early-career
-researchers. In addition, the travel award subcommittee tried to increase the
-diversity of applicants by forwarding the notice of travel support to
-organizations including PyLadies\footnote{\url{http://www.pyladies.com}},
-Django Girls\footnote{\url{https://djangogirls.org}},
-Women Who Code\footnote{\url{https://www.womenwhocode.com}}, and
-Women in HPC\footnote{\url{http://www.womeninhpc.org}}.
+funds allowed 29 additional people % to attend and participate. A subset of the
+% organizing committee reviewed 
+out of 44 applicants which 
+% ions for travel support and
+% competitively selected the awardees, 
+included 11 students and five early-career
+researchers.
+%  In addition, the travel award subcommittee tried to increase the
+% diversity of applicants by forwarding the notice of travel support to
+% organizations including PyLadies\footnote{\url{http://www.pyladies.com}},
+% Django Girls\footnote{\url{https://djangogirls.org}},
+% Women Who Code\footnote{\url{https://www.womenwhocode.com}}, and
+% Women in HPC\footnote{\url{http://www.womeninhpc.org}}.
 %\todo{did we reach out to any others?}
@@
     discussions---speakers, helpers, organizers, and participants---is required
-    to conform to the following Code of Conduct.
-
-    \begin{itemize}
-    \item Communicate appropriately for a professional audience including
-    people of many different backgrounds. Sexual language and imagery are not
-    appropriate for any event.
-
-    \item Be kind to others. Do not insult or put down other attendees. Be
-    mindful of jargon, which can sometimes exclude others from engaging in the
-    discussion.
-
-    \item Behave professionally. Remember that harassment and sexist, racist,
-    ageist, or exclusionary behavior are not appropriate.
-    \end{itemize}
+    to conform to the Code of Conduct, listed on the community website \cite{}.
+
+%     \begin{itemize}
+%     \item Communicate appropriately for a professional audience including
+%     people of many different backgrounds. Sexual language and imagery are not
+%     appropriate for any event.
+
+%     \item Be kind to others. Do not insult or put down other attendees. Be
+%     mindful of jargon, which can sometimes exclude others from engaging in the
+%     discussion.
+
+%     \item Behave professionally. Remember that harassment and sexist, racist,
+%     ageist, or exclusionary behavior are not appropriate.
+%     \end{itemize}
  \end{quote}
@@
 The CoC was introduced at the beginning of WSSSPE4, with the CoC subcommittee
-and a general email address introduced for reporting concerns or incidents, or
-asking questions.  There was one concern mentioned to the CoC subcommittee
-after the first half day,
-regarding how part of the workshop was being run, and we changed the workshop
+and a general email address for reporting concerns or incidents, or
+asking questions.  One concern was raised after the first half day, 
+% regarding how part of the workshop was being run, 
+and we changed the workshop
 to address this.
@@
 participate in sharing and implementing their ideas.
-Specifically,
-19 lightning talks,
-4 idea papers,
-3 position papers,
-2 experience papers,
-and
-3 demos
-were accepted, as mentioned in Section~\ref{sec:presentations}. (Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
+% Specifically,
+% 19 lightning talks,
+% 4 idea papers,
+% 3 position papers,
+% 2 experience papers,
+% and
+% 3 demos
+% were accepted, as mentioned in Section~\ref{sec:presentations}.
+(Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
 were accepted as lightning talks.)
@@
 
-The participants were given time after the presentation to write down their comments.
-Seven people volunteered to work on revising the mission and vision statements, incorporating these comments and presenting back to the community.
-Based on this feedback, the committee redrafted the mission and vision.
+The participants were 
+%given time after the presentation to write down their  
+invited to comment,
+% Seven people volunteered to work on 
+and 
+%revising 
+the mission and vision statements, were revised 
+%incorporating these comments and presenting back to the community.
+based on this feedback.
+% the committee redrafted the mission and vision.
 The committee added focus areas to clarify the organization's breadth and to keep the mission and vision simple and long-lasting.
-Comments after the close of the meeting were incorporated into the statements.
-
+Comments after the close of the meeting were also incorporated into the statements.
 A final draft was put on GitHub and advertised
 via the WSSSPE mailing list, Facebook group, and Twitter.
-After two weeks without suggested changes, the final statements, as shown below, were added
+After two weeks without suggested changes, the final statements, 
+%as shown below, 
+%were added
 to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
 
-{\bf Mission.}
-WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
-
-{\bf Vision.}
-We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
-
-{\bf Focus areas.}
-WSSSPE promotes sustainable research software by positively impacting:
-\begin{itemize}
-\item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
-\item {\bf Careers}. Developing and supporting career paths in research software development and engineering
-\item {\bf Learning}. Engaging in activities to promote peer learning and interaction
-\item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
-\end{itemize}
-
-\textbf{Definitions:}
-\emph{Sustainable software} has the capacity to endure such that it will continue to
-be available in the future, on new platforms, meeting new needs.
-The \emph{research software lifecycle} includes:
-acquiring and assembling resources (including funding and people) into teams and communities,
-developing software,
-using software,
-recognizing contributions to and of software,
-and
-maintaining software.
+% {\bf Mission.}
+% WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
+
+% {\bf Vision.}
+% We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
+
+% {\bf Focus areas.}
+% WSSSPE promotes sustainable research software by positively impacting:
+% \begin{itemize}
+% \item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
+% \item {\bf Careers}. Developing and supporting career paths in research software development and engineering
+% \item {\bf Learning}. Engaging in activities to promote peer learning and interaction
+% \item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
+% \end{itemize}
+
+% \textbf{Definitions:}
+% \emph{Sustainable software} has the capacity to endure such that it will continue to
+% be available in the future, on new platforms, meeting new needs.
+% The \emph{research software lifecycle} includes:
+% acquiring and assembling resources (including funding and people) into teams and communities,
+% developing software,
+% using software,
+% recognizing contributions to and of software,
+% and
+% maintaining software.
 
@@
 
-The following four papers belong to the first area: the academic environment.
+% The following four papers belong to the first area: the academic environment.
+%For talks with multiple authors, an
+% \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
+Four papers, 
+% \begin{itemize}
+% \item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
+% Improving Incentives for Quality Software}
+~\cite{Heroux:2016ws}, 
+% \item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
+% Society for Research Software}
+~\cite{GAllen:2016ws}, 
+% \item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
+% \emph{Experience Paper: Preliminary analysis of a survey of UK Research
+% Software Engineers}
+~\cite{Philippe:2016ws}, and 
+% \item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
+% create and develop research ties within academia}
+~\cite{Gwilliams:2016ws} belong to the first area: the academic environment.
 They address diverse aspects ranging from incentives for quality software to
 advocating a professional society for research software to roles and degrees for
-research software engineers. For talks with multiple authors, an
-\textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
-
-\begin{itemize}
-\item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
-Improving Incentives for Quality Software}~\cite{Heroux:2016ws}.
-
-\item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
-Society for Research Software}~\cite{GAllen:2016ws}.
-
-\item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
-\emph{Experience Paper: Preliminary analysis of a survey of UK Research
-Software Engineers}~\cite{Philippe:2016ws}.
-
-\item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
-create and develop research ties within academia}~\cite{Gwilliams:2016ws}.
-
-\end{itemize}
-
-\noindent The second main area, concerned with characteristics and needs of research software, includes the following five papers:
-
-\begin{itemize}
-
-\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
-Engineering and Community Codes Track in ATPESC}~\cite{Dubey1:2016ws}.
-
-\item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
-research software, and how should we do it?}~\cite{ChueHong:2016ws}.
-
-\item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
-of Software for Scientific Simulation}~\cite{Dubey2:2016ws}.
-
-\item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
-Collaborative Gamification Design for Scientific Software}~\cite{Queiroz:2016ws}.
-
-\item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
-David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
-the Way Digital Libraries will Operate}~\cite{Childers:2016ws}.
-
-\end{itemize}
-
-\noindent Three contributions focused on the third main area: elucidating lessons learned
+research software engineers.
+%\end{itemize}
+The second main area, concerned with characteristics and needs of
+research software, includes five papers:
+~\cite{Dubey1:2016ws},  ~\cite{ChueHong:2016ws},
+~\cite{Dubey2:2016ws}, ~\cite{Queiroz:2016ws}, and ~\cite{Childers:2016ws}.
+% \begin{itemize}
+% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
+% Engineering and Community Codes Track in ATPESC}
+% \item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
+% research software, and how should we do it?}
+% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
+% of Software for Scientific Simulation}
+% \item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
+% Collaborative Gamification Design for Scientific Software}
+% \item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
+% David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
+% the Way Digital Libraries will Operate}
+% \end{itemize}
+%\noindent 
+Three contributions, ~\cite{Ganguly:2016ws}, ~\cite{Shende:2016ws},
+and ~\cite{Sallai:2016ws} , focused on the third main area: elucidating lessons learned
 from or visions for use cases of scientific software and communities working
-with and\slash or on a software package.
-
-\begin{itemize}
-
-\item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
-Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
-Reproducing, and Sharing Simulations}~\cite{Ganguly:2016ws}.
-
-\item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
-Performance Evaluation of Scientific Software}~\cite{Shende:2016ws}.
-
-\item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
-\emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
-Atom-Typing for Molecular Simulation}~\cite{Sallai:2016ws}.
-
-\end{itemize}
+with and\slash or on a software package. There were 19 lightning talks presented at WSSSPE4, (in order of
+presentation): ~\cite{Sufi:2016ws}, ~\cite{Gesing:2016ws},
+~\cite{Ram:2016ws}, ~\cite{Loffler:2016ws}, ~\cite{Katz:2016ws},
+~\cite{Idaszak:2016ws}, ~\cite{Hwang:2016ws}, ~\cite{Goble:2016ws},
+~\cite{Druskat:2016ws}, ~\cite{Contrastin:2016ws},
+~\cite{AAllen:2016ws}, ~\cite{ChueHong:2016wsb},
+~\cite{vanHage:2016ws}, ~\cite{GAllen:2016wsb}, ~\cite{Seidel:2016ws}, 
+  ~\cite{Emsley:2016ws}, ~\cite{Dongarra:2016ws},
+  ~\cite{Bauer:2016ws},  and ~\cite{Aldabjan:2016ws}.
+
+% \begin{itemize}
+% \item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
+% Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
+% Reproducing, and Sharing Simulations}
+% \item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
+% Performance Evaluation of Scientific Software}
+% \item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
+% \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
+% Atom-Typing for Molecular Simulation}
+% \end{itemize}
 
@@
 
-\noindent There were 19 lightning talks presented at WSSSPE4, as follows (in order of
-presentation).
 %
@@
 
-\begin{itemize}%[itemsep=1ex]
-    \item Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
-    programme; supporting the social side of research software}~\cite{Sufi:2016ws}.
-
-    \item Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
-    Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
-    \emph{The Science Gateways Community Institute - Supporting Communities to
-    Achieve Sustainability for Their Science Gateways}~\cite{Gesing:2016ws}.
-
-    \item Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
-    and Scott Chamberlain: \emph{A model for
-    peer review and onboarding research software}~\cite{Ram:2016ws}.
-
-    \item Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
-    \emph{A vision of computing in 10+ years}~\cite{Loffler:2016ws}.
-
-    \item Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
-    Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
-    \emph{Software Citation: Process, Principles, and Implementation}~\cite{Katz:2016ws}.
-
-    \item Ray Idaszak\textsuperscript{\textasteriskcentered},
-    David G.~Tarboton, Hong Yi, Michael Stealey,
-    Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
-    Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
-    Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
-    Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
-    Best Practices and Culture Change for Developing Sustainable Community
-    Software}~\cite{Idaszak:2016ws}.
-
-    \item Lorraine Hwang\textsuperscript{\textasteriskcentered},
-    Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
-    \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
-    Community}~\cite{Hwang:2016ws}.
-
-    \item Carole Goble: \emph{A Simple Profiling Framework for Software
-    User-Producer Reciprocity Review}~\cite{Goble:2016ws}.
-
-    \item Stephan Druskat: \emph{A proposal for the measurement and
-    documentation of research software sustainability in interactive metadata
-    repositories}~\cite{Druskat:2016ws}.
-
-    \item Mistral Contrastin, Matthew Danish,
-    Dominic Orchard\textsuperscript{\textasteriskcentered}, and
-    Andrew Rice: \emph{Supporting software sustainability with
-    lightweight specifications}~\cite{Contrastin:2016ws}.
-
-    \item Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
-    Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
-    Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
-    Robert Haines, Robert Hirschfeld, James Howison,
-    Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
-    Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
-    and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
-    Responsibility in the Engineering of Academic Software}~\cite{AAllen:2016ws}.
-
-    \item Neil Chue Hong\textsuperscript{\textasteriskcentered}:
-    \textit{Making it easier to understand research software impact}~\cite{ChueHong:2016wsb}.
-
-    \item Willem Robert van Hage\textsuperscript{\textasteriskcentered},
-    Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
-    Netherlands eScience Center}~\cite{vanHage:2016ws}.
-
-    \item Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
-    Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
-    Research Software}~\cite{GAllen:2016wsb}.
-
-    \item Eric L.~Seidel\textsuperscript{\textasteriskcentered}
-    and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
-    into Scientific Software}~\cite{Seidel:2016ws}.
-
-    \item Iain Emsley\textsuperscript{\textasteriskcentered} and David
-    De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
-    community}~\cite{Emsley:2016ws}.
-
-    \item Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
-    D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
-    and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
-    BLAS Routines}~\cite{Dongarra:2016ws}.
-
-    \item Lukas Breitwieser\textsuperscript{\textasteriskcentered},
-    Roman Bauer, Alberto Di Meglio, Leonard Johard,
-    Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
-    Max Talanov: \textit{The BioDynaMo Project}~\cite{Bauer:2016ws}.
-
-    \item Aseel Aldabjan, Robert Haines, and
-    Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
-    measure the relationship between code quality and software
-    sustainability?}~\cite{Aldabjan:2016ws}.
-
-\end{itemize}
+% \begin{itemize}%[itemsep=1ex]
+%     \item Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
+%     programme; supporting the social side of research software}
+
+%     \item Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
+%     Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
+%     \emph{The Science Gateways Community Institute - Supporting Communities to
+%     Achieve Sustainability for Their Science Gateways}
+
+%     \item Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
+%     and Scott Chamberlain: \emph{A model for
+%     peer review and onboarding research software}
+%     \item Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
+%     \emph{A vision of computing in 10+ years}
+%     \item Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
+%     Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
+%     \emph{Software Citation: Process, Principles, and
+%       Implementation}
+
+%     \item Ray Idaszak\textsuperscript{\textasteriskcentered},
+%     David G.~Tarboton, Hong Yi, Michael Stealey,
+%     Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
+%     Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
+%     Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
+%     Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
+%     Best Practices and Culture Change for Developing Sustainable Community
+%     Software}
+
+%     \item Lorraine Hwang\textsuperscript{\textasteriskcentered},
+%     Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
+%     \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
+%     Community}
+
+%     \item Carole Goble: \emph{A Simple Profiling Framework for Software
+%     User-Producer Reciprocity Review}
+%     \item Stephan Druskat: \emph{A proposal for the measurement and
+%     documentation of research software sustainability in interactive metadata
+%     repositories}
+
+%     \item Mistral Contrastin, Matthew Danish,
+%     Dominic Orchard\textsuperscript{\textasteriskcentered}, and
+%     Andrew Rice: \emph{Supporting software sustainability with
+%     lightweight specifications}
+
+%     \item Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
+%     Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
+%     Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
+%     Robert Haines, Robert Hirschfeld, James Howison,
+%     Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
+%     Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
+%     and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
+%     Responsibility in the Engineering of Academic
+%     Software}
+
+%     \item Neil Chue Hong\textsuperscript{\textasteriskcentered}:
+%     \textit{Making it easier to understand research software
+%       impact}
+
+%     \item Willem Robert van Hage\textsuperscript{\textasteriskcentered},
+%     Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
+%     Netherlands eScience Center}
+
+%     \item Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
+%     Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
+%     Research Software}
+
+%     \item Eric L.~Seidel\textsuperscript{\textasteriskcentered}
+%     and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
+%     into Scientific Software}
+
+%     \item Iain Emsley\textsuperscript{\textasteriskcentered} and David
+%     De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
+%     community}
+
+%     \item Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
+%     D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
+%     and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
+%     BLAS Routines}
+
+%     \item Lukas Breitwieser\textsuperscript{\textasteriskcentered},
+%     Roman Bauer, Alberto Di Meglio, Leonard Johard,
+%     Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
+%     Max Talanov: \textit{The BioDynaMo Project}
+
+%     \item Aseel Aldabjan, Robert Haines, and
+%     Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
+%     measure the relationship between code quality and software
+%     sustainability?}
+
+% \end{itemize}
 
@@
 Many open source projects for research software document their best practices that
-contributed to the sustainability of the software.  Many of these projects also
+contribute to the sustainability of the software.  Many of these projects also
 document software metrics they use to define their research software project's success.
@@
 
-To join this group or obtain more information about it, please send an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
+Anyone can join this group or obtain more information about it, by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
 
@@
 
-If interested, visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
+If interested, one can visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
 
@@
 
-To join this project, please visit the GitHub repository at
+To join this project, one can visit the GitHub repository at
 \url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel
@@
 %
-Furthermore, the group also recognize that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
+Furthermore, the group also recognizes that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
 
@@
 
-To join the group, join its mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
+To join the group, one can join its mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
 
@@
 
-For more information, please contact Emily Chen at <echen35@illinois.edu>.
+For more information, one can contact Emily Chen at <echen35@illinois.edu>.
 

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
adding PDF of submission, initial version of response letter accepting changes in intro from Anshu removing appendix of organizers and PC

@@
 \setlist{leftmargin=7mm}
+
+\makeatletter
+\newcommand\footnoteref[1]{\protected@xdef\@thefnmark{\ref{#1}}\@footnotemark}
+\makeatother
 
@@
 At the workshop, we introduced a Code of Conduct (CoC).%
-\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
+\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
 The CoC was conceived for the workshop and a guideline for the community of scientists that WSSSPE
@@
     discussions---speakers, helpers, organizers, and participants---is required
-    to conform to the Code of Conduct, listed on the community website \cite{}.
+    to conform to the Code of Conduct\footnoteref{footnote:CoC}.
 
@@
 This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
-An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).
+%An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).
 
@@
 
-\newpage
-\appendix
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Organization}  \label{sec:orgcom}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\todo{Do we want email addresses here?}
-
-WSSSPE4 organizers were:
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{orgcom}
-\end{longtable}
-}
-
-The WSSSPE4 program committee comprised:
-
-{\scriptsize
-\begin{longtable}{lll}
-\input{program_committee}
-\end{longtable}
-}
+%\newpage
+%\appendix
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\section{Organization}  \label{sec:orgcom}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%\todo{Do we want email addresses here?}
+%
+%WSSSPE4 organizers were:
+%
+%{\scriptsize
+%\begin{longtable}{lll}
+%\input{orgcom}
+%\end{longtable}
+%}
+%
+%
+%The WSSSPE4 program committee comprised:
+%
+%{\scriptsize
+%\begin{longtable}{lll}
+%\input{program_committee}
+%\end{longtable}
+%}
 

### 2017-12-18 WSSSPE4_report/jors-response.tex
adding PDF of submission, initial version of response letter accepting changes in intro from Anshu removing appendix of organizers and PC

\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{csquotes}
\usepackage[english]{babel}
\usepackage{textcomp}
\usepackage{lmodern}
\usepackage{ulem}

\usepackage[letterpaper, margin=1in]{geometry}

\usepackage{xcolor}
\usepackage{fancyhdr}
\usepackage{lastpage}

\usepackage[hyphens]{url}
\usepackage[breaklinks=true,linkcolor=blue, citecolor=blue, urlcolor=blue, colorlinks=true]{hyperref}

\usepackage{todonotes}

\usepackage[
    backend=biber,
    style=numeric,
    citestyle=numeric-comp,
    sorting=none,
    bibencoding=UTF-8,
    giveninits=true,
    maxbibnames=1000,
]{biblatex}
\addbibresource{refs.bib}

\usepackage{parskip} % no indentation, space between paras

\usepackage[font={itshape}]{quoting}

\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}
\cfoot{Page \thepage{}~of~\pageref{LastPage}}

\newcommand\joss{\textit{JOSS}}

\definecolor{darkgreen}{rgb}{0,0.5,0}

\usepackage[xcolor]{changebar}
% % *** Be very precise and careful about including whitespace and punctuation in your edits ***
% \newcommand{\addone}[1]{{\sloppy\cbcolor{teal}\textcolor{teal}{\cbstart {#1}\cbend}}}  % add
% \newcommand{\addtwo}[1]{{\sloppy\cbcolor{red}\textcolor{red}{\cbstart {#1}\cbend}}}  % add
% \newcommand{\addthree}[1]{{\sloppy\cbcolor{darkgreen}\textcolor{darkgreen}{\cbstart {#1}\cbend}}}  % add
% \newcommand{\deleteone}[1]{\sloppy\cbcolor{teal}\textcolor{teal}{\cbdelete\sout{#1}}}
% \newcommand{\deletetwo}[1]{\sloppy\cbcolor{red}\textcolor{red}{\cbdelete \sout{#1}}}
% \newcommand{\deletethree}[1]{\sloppy\cbcolor{darkgreen}\textcolor{darkgreen}{\cbdelete \sout{#1}}}
% \newcommand{\addnoul}[1]{{\textcolor{teal}{#1}}}     % add with no underline
% \newcommand{\deletenoso}[1]{{\textcolor{red}{#1}}}    % delete with no strikeout

\newcommand{\add}[1]{{\sloppy\textcolor{teal}{#1}}}  % add
\newcommand{\delete}[1]{\sloppy\textcolor{red}{\sout{#1}}}
\newcommand{\addnoul}[1]{{\textcolor{teal}{#1}}}     % add with no underline
\newcommand{\deletenoso}[1]{{\textcolor{red}{#1}}}    % delete with no strikeout
%% For the final version, use these four commands instead
% \renewcommand{\add}[1]{#1}
% \renewcommand{\addnoul}[1]{#1}
% \renewcommand{\delete}[1]{}
% \renewcommand{\deletenoso}[1]{}

\title{WSSSPE4-paper-reviewer-response}
\author{Daniel S. Katz}
\date{December 2017}

\begin{document}

\today\\

Samuel Moore\\
\textit{Journal of Open Research Software}\\

Subject: WSSSPE4 paper review\\

Dear Mr. Moore,

We enclose revisions for our manuscript ``Fourth Workshop on Sustainable Software for Science: Practice and Experiences
(WSSSPE4)'' and address the reviewers' comments in the following remarks.

\section*{Reviewer A}

\begin{quoting}
Comments to the authors: 
This article provides a comprehensive review of the 4th WSSSPE.  It covers information that will be of interest to the community and gives a good background to the series of workshops.  It tackles issues which are important to the community of research software engineers and to the improvement and sustainability of research software.
The comprehensive write-up of the working groups is useful, outlining the set of actions to be carried out moving forward. The links for those who wish to be further involved in this work will be useful to the reader.
The article is, however, 3 times longer than the stipulated maximum word count of 3500. 
\end{quoting}

In discussions with the editor-in-chief before submission, we were told:

``The APC is \pounds300 up to 8,000 words (approximately 12 pages including
figures and references). Additional APC charges will be \pounds10 for every
additional 750 words ($\approx$1 page).
So 16 pages would be \pounds340 - \pounds350. I would probably not be wanting to
go longer than 16 pages unless there was a good reason.''

Following these instructions, our initial submission had a body of 13 pages, an appendix of 2 pages, and 3 pages of references.

Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.  The revised version we are resubmitting has been reduced to \textcolor{red}{XX PAGES} of body, and 3 pages of references.

\begin{quoting}
There are several places where the length could be reduced, for example the list of presentations seems unnecessary when the references are also included

\end{quoting}

\textcolor{red}{NEED TO ADDRESS}

\begin{quoting}

and details of the Code of Conduct could be omitted as the link for this is provided.

\end{quoting}

The Code of Conduct has been removed.

\begin{quoting}

 In some places there is maybe detail which is not useful to the reader, for example detail of the survey responses.  I would recommend a shortened version of this article for publication but must reject it in its current format.
\end{quoting}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Reviewer B}

\begin{quoting}
Comments to the authors: 
This article reviews a workshop that focused on sustainable software for science. While the topic of this workshop is undoubtedly important, and a review that reports on the events of the workshop, provides a bit of background about the challenges that such a workshop seeks to address, and synthesizes some of the broad conclusions of the workshop would be quite valuable, the current paper focuses quite closely on the first of these, and provides less in terms of the latter two. I believe this is a missed opportunity to address some of the issues that the authors themselves recognize in section 7 of the article, including some level-setting in terms of terminology and context for the meeting.

As the authors acknowledge in the Introduction, this paper is a summary of a longer more detailed report on the events of the workshop, but the article as it is right now still reads as a detailed enumeration of the events of the conference. For example, while the introduction provides a list of the WSSSPE events that preceded this meeting, a reader that is not familiar with this community would receive very little background knowledge about these meetings, and the impetus for holding them, from reading this Introduction. The second section provides a bit more in this sense, and it might be that using some of the content from the call for papers earlier on in the article would make it easier to read the paper for someone who is less familiar with the workshop.

On the other hand, the level of detail is sometimes overwhelming, and distracts from the main thrust of the paper. For example, while several lines of text are devoted to reporting on the number of attendees who filled out a survey about the event, the completion rates of the survey, and the distribution mechanisms of the survey, only rather cursory and qualitative survey results are reported. More detail on the latter, and less detail on the former would improve this section, in my opinion.  For example, if one of the stated goals of the workshop was to improve research software, by imparting best practices on participants, was this goal met, according to the survey results?

Similarly, while the code of conduct is provided in full, no context is provided for establishing a code of conduct for such an event. Why do events such as this benefit from a code of conduct? What is the social context in which codes of conduct in these types of events become established? In what ways does a code of conduct benefit this event? Instead of discussing these more general issues (maybe only some of them are needed here) an enigmatic comment about an event that required some form of intervention from the organizers is mentioned, but left unexplained. Would reporting more fully about what happened be beneficial to the reader?s understanding of the workshop? If not, maybe shortening this section would be better?

Finally, in a similar vein, the summary and conclusion offers very little in terms of synthesis of main themes that arose during the meeting. The reader is left to synthesize the diversity of sections 4 and 5, that discuss the main activities of the workshop itself, on their own. It may very well be that the workshop is diverse enough in terms of its topics (and indeed, it does seem like there are a lot of very interesting topics, and a lot of interesting work done as part of the workshop!), but surely there is some over-arching themes that could be summarized? Instead, the conclusion is that a workshop such as this is unlikely to produce tangible work in the long run, because participants are reluctant to commit to additional activities, above and beyond these undertaken during the workshop. While this seems like an honest (if disappointing?) conclusion, it is also difficult to reconcile with the multiple places in which the authors ask interested parties to join Slack channels and mailing lists. Maybe the expectation that a few hours of joint work at a workshop would lead to a committed collaboration with other participants is too high? Rather, expectations from conferences and workshops are usually much lower: exchanging ideas, establishing and broadening a professional network, and learning about work that is being done in the field. Maybe it's enough if these goals are achieved at WSSSPE as well?

Minor:

1. On page 8, the sentence fragment: ``...later to be fully implemented into by scientific programmers?'' seems to be missing a word or words. Alternatively the word ``into'' should be dropped.

2. It would probably be appropriate to mention the number of participants early on. From reading section 7, I surmise that there were 44 participants (or maybe it was 43, accounting for the participant who filled the survey twice? A detail the paper could probably do without?). Or does the completion rate refer to how many questions each of the 44 (43?) attendees.

3. In section 7: ``Text responses have been alphabetize'' => ``Text responses have been alphabetized''.
\end{quoting}

We have made these three requested changes.

%\printbibliography

\end{document}

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
minor changes in sections 2 and 4, more work in response.

@@
 3 demos.
-Most of the submissions were accepted, since the goal of WSSSPE is always to
+Most submissions were accepted, since the goal of WSSSPE is always to
 take in as many relevant inputs as possible, and to encourage their authors to
-participate in sharing and implementing their ideas.
+participate in sharing and implementing their ideas. Two submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
+were accepted as lightning talks.
 % Specifically,
@@
 % were accepted, as mentioned in Section~\ref{sec:presentations}.
-(Two of the submitted lightning talks and one submitted experience paper were rejected, while two more submitted experience papers
-were accepted as lightning talks.)
+%
 The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
@@
 sustainability for software in science. These talks covered three main areas.
-
+%
 % The following four papers belong to the first area: the academic environment.
@@
 % \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
-Four papers, 
+%Four papers %,  
 % \begin{itemize}
@@
 % Improving Incentives for Quality Software}
-~\cite{Heroux:2016ws}, 
+%~\cite{Heroux:2016ws}, 
 % \item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
 % Society for Research Software}
-~\cite{GAllen:2016ws}, 
+%~\cite{GAllen:2016ws}, 
 % \item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
@@
 % Software Engineers}
-~\cite{Philippe:2016ws}, and 
+%~\cite{Philippe:2016ws}, and 
 % \item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
 % create and develop research ties within academia}
-~\cite{Gwilliams:2016ws} belong to the first area: the academic environment.
-They address diverse aspects ranging from incentives for quality software to
+%~\cite{Gwilliams:2016ws} 
+%belong to 
+The first area was the academic environment, with four papers
+that address diverse aspects ranging from incentives for quality software to
 advocating a professional society for research software to roles and degrees for
-research software engineers.
+research software engineers~\cite{Heroux:2016ws, GAllen:2016ws, Philippe:2016ws, Gwilliams:2016ws}.
 %\end{itemize}
-The second main area, concerned with characteristics and needs of
-research software, includes five papers:
-~\cite{Dubey1:2016ws},  ~\cite{ChueHong:2016ws},
-~\cite{Dubey2:2016ws}, ~\cite{Queiroz:2016ws}, and ~\cite{Childers:2016ws}.
+The second area was concerned with characteristics and needs of
+research software and included five papers~\cite{Dubey1:2016ws, ChueHong:2016ws, Dubey2:2016ws, Queiroz:2016ws, Childers:2016ws}.
 % \begin{itemize}
@@
 %\noindent 
-Three contributions, ~\cite{Ganguly:2016ws}, ~\cite{Shende:2016ws},
-and ~\cite{Sallai:2016ws} , focused on the third main area: elucidating lessons learned
+%Three contributions, ~\cite{Ganguly:2016ws}, ~\cite{Shende:2016ws},
+%and ~\cite{Sallai:2016ws} , focused on 
+The third  area focused on elucidating lessons learned
 from or visions for use cases of scientific software and communities working
-with and\slash or on a software package. There were 19 lightning talks presented at WSSSPE4, (in order of
-presentation): ~\cite{Sufi:2016ws}, ~\cite{Gesing:2016ws},
-~\cite{Ram:2016ws}, ~\cite{Loffler:2016ws}, ~\cite{Katz:2016ws},
-~\cite{Idaszak:2016ws}, ~\cite{Hwang:2016ws}, ~\cite{Goble:2016ws},
-~\cite{Druskat:2016ws}, ~\cite{Contrastin:2016ws},
-~\cite{AAllen:2016ws}, ~\cite{ChueHong:2016wsb},
-~\cite{vanHage:2016ws}, ~\cite{GAllen:2016wsb}, ~\cite{Seidel:2016ws}, 
-  ~\cite{Emsley:2016ws}, ~\cite{Dongarra:2016ws},
-  ~\cite{Bauer:2016ws},  and ~\cite{Aldabjan:2016ws}.
+with and\slash or on a software package, with three papers~\cite{Ganguly:2016ws, Shende:2016ws, Sallai:2016ws}.
+
+In addition, 19 lightning talks were presented at WSSSPE4~\cite{Sufi:2016ws, Gesing:2016ws, Ram:2016ws, Loffler:2016ws, Katz:2016ws, Idaszak:2016ws, Hwang:2016ws, Goble:2016ws, Druskat:2016ws, Contrastin:2016ws, AAllen:2016ws, ChueHong:2016wsb, vanHage:2016ws, GAllen:2016wsb, Seidel:2016ws, Emsley:2016ws, Dongarra:2016ws, Bauer:2016ws, Aldabjan:2016ws}.
 

### 2017-12-18 WSSSPE4_report/jors-response.tex
minor changes in sections 2 and 4, more work in response.

@@
 figures and references). Additional APC charges will be \pounds10 for every
-additional 750 words ($\approx$1 page).
+additional 750 words ($\sim$1 page).
 So 16 pages would be \pounds340 - \pounds350. I would probably not be wanting to
@@
 
-Following these instructions, our initial submission had a body of 13 pages, an appendix of 2 pages, and 3 pages of references.
+Following these instructions, our initial submission had a body of 13 pages, an appendix of 2 pages, and 3 pages of references (as JORS publishes appendices as separate PDF attachments, and does not typeset them.)
 
 Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.
-The revised version we are resubmitting has been reduced to \textcolor{red}{XX PAGES} of body, and 3 pages of references.
+The revised version we are resubmitting has been reduced to \textcolor{red}{11} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5} of the original 18 pages.
+
+\textcolor{red}{Can we remove 2 more pages?}
 
@@
 
-\textcolor{red}{NEED TO ADDRESS}
+We have removed the list of presentations.
 
@@
  In some places there is maybe detail which is not useful to the reader, for example detail of the survey responses.
+
+\end{quoting}
+
+\textcolor{red}{NEED TO RESPOND}
+
+\begin{quoting}
 I would recommend a shortened version of this article for publication but must reject it in its current format.

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
work in intro - move CFP topics here, reduce description of previous events, move CoC to new section (3), work on response letter to match

@@
 First Research Software Engineers (RSE) Conference.
-Previous events in the WSSSPE series include
-WSSSPE1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1/}}~\cite{WSSSPE1-pre-report,WSSSPE1},
-held in conjunction with the SC13 conference;
-WSSSPE1.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}}, a
-focused workshop organized jointly with the SciPy
-conference\footnote{\url{https://conference.scipy.org/scipy2014/participate/wssspe/}};
-WSSSPE2\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2/}}~\cite{WSSSPE2-pre-report,WSSSPE2},
-held in conjunction with SC14;
-WSSSPE2.1\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}}, a
-focused workshop organized again jointly with
-SciPy\footnote{\url{http://scipy2015.scipy.org/ehome/115969/286469/}};
-and WSSSPE3\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe3/}}~\cite{WSSSPE3},
-held in Boulder, Colorado, USA.
-The first WSSSPE workshop was named
+Previous WSSSPE events\footnote{The first WSSSPE workshop was named
 ``Working towards
@@
 Software for Science: Practice and Experiences.'' Together these reflect
-that WSSSPE is both a community and a set of workshops.
+that WSSSPE is both a community and a set of workshops.}, all of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1, WSSSPE2-pre-report,WSSSPE2},
+two half-day workshops\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}, \url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}} held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, and a 1-and-1/2-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}.
+%
+Based on the work done at these previous WSSSPE meetings, WSSSPE4 was aimed
+at producing working groups that would better continue working after the workshop ended.
+
+Specific topics that were discussed in previous meetings and suggested for WSSSPE4 included:
+
+\begin{itemize}
+\renewcommand{\labelenumi}{\textbf{\theenumi}.}
+\setlength{\rightmargin}{1em}
+
+\item Development and Community:
+best practices for developing sustainable software;
+models for funding specialist expertise in software collaborations;
+software tools that aid sustainability;
+academia/industry interaction;
+refactoring/improving legacy scientific software;
+engineering design for sustainable software;
+metrics for the success of scientific software; and
+adaptation of mainstream software practices for scientific software
+
+\item Professionalization:
+career paths;
+RSE as a brand;
+RSE outside of the UK or Europe; and
+Increase incentives in publishing, funding and promotion for better software
+
+\item Training:
+training for developing sustainable software; and 
+curriculum for software sustainability
+
+\item Credit:
+making the existing credit and citation ecosystem work better for software;
+future credit and citation ecosystem;
+software contributions as a part of tenure review;
+case studies of receiving credit for software contributions; and
+awards and recognition that encourage sustainable software
+
+\item Software publishing:
+journals and alternative venues for publishing software; and
+review processes for published software
+
+\item Software discoverability/reuse:
+proposals and case studies
+
+\item Reproducibility and testing:
+reproducibility in conferences and journals; and
+best practices for code testing and code review
+
+\end{itemize}
 
@@
 
-The WSSSPE4 workshop included multiple mechanisms for participation and
-encouraged team building around solutions. WSSSPE4 strongly encouraged participation
+WSSSPE4 included multiple mechanisms for participation and
+encouraged team building around solutions. It strongly encouraged participation
 of early-career scientists, postdoctoral researchers, graduate students,
@@
 who would not otherwise be able to attend the workshop. These
-funds allowed 29 additional people % to attend and participate. A subset of the
+funds allowed 29 additional attendees % to attend and participate. A subset of the
 % organizing committee reviewed 
-out of 44 applicants which 
+(of 44 applicants), 
 % ions for travel support and
 % competitively selected the awardees, 
-included 11 students and five early-career
+including 11 students and five early-career
 researchers.
@@
 
-At the workshop, we introduced a Code of Conduct (CoC).%
-\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}
-The CoC was conceived for the workshop and a guideline for the community of scientists that WSSSPE
-supports, including their personal and online interactions (e.g., on
-Twitter, in email lists, in the Slack team). The WSSSPE4 CoC is based on the
-FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
-CoC~\cite{Code4Lib:CoC}.
-The main guidelines of the CoC are:
-\begin{quote}
-    WSSSPE events are community events intended for networking and collaboration
-    as well as learning. We value the participation of every member of the
-    community and want all attendees to have an enjoyable and fulfilling
-    experience. Accordingly, all attendees are expected to show respect and
-    courtesy to other attendees throughout the event and in interactions online
-    associated with the event.
-
-    The WSSSPE event organizers are dedicated to providing a harassment-free
-    experience for everyone, regardless of gender, gender identity and
-    expression, age, sexual orientation, disability, physical appearance,
-    body size, race, ethnicity, religion (or lack thereof), technology choices,
-    or other group status.
-
-    To make clear what is expected, everyone taking part in WSSSPE events and
-    discussions---speakers, helpers, organizers, and participants---is required
-    to conform to the Code of Conduct\footnoteref{footnote:CoC}.
-
-%     \begin{itemize}
-%     \item Communicate appropriately for a professional audience including
-%     people of many different backgrounds. Sexual language and imagery are not
-%     appropriate for any event.
-
-%     \item Be kind to others. Do not insult or put down other attendees. Be
-%     mindful of jargon, which can sometimes exclude others from engaging in the
-%     discussion.
-
-%     \item Behave professionally. Remember that harassment and sexist, racist,
-%     ageist, or exclusionary behavior are not appropriate.
-%     \end{itemize}
- \end{quote}
-
-The CoC was introduced at the beginning of WSSSPE4, with the CoC subcommittee
-and a general email address for reporting concerns or incidents, or
-asking questions.  One concern was raised after the first half day, 
-% regarding how part of the workshop was being run, 
-and we changed the workshop
-to address this.
-
-This paper is a summary of a longer report~\cite{WSSSPE4-report} on the events at the workshop and the submitted materials.
-The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).
+This paper is a summary of a longer report~\cite{WSSSPE4-report} on the workshop and the submitted materials.
+The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).
 One full day of the
@@
 
-WSSSPE4 was based on the work done at previous WSSSPE meetings, but aimed
-at producing working groups that better continued working after the workshop ended.
-In addition, based on feedback after WSSSPE3, it became clear that attendees
-had two distinct motivations:  to make a better future
-for research software, and to immediately do better research software development.
-This led to the idea of WSSSPE4 being partially divided into two tracks:
+Post-WSSSPE3 feedback said that attendees
+had two distinct motivations for attending:  to make a better future
+for research software, and to immediately do better research software development. Thus,
+WSSSPE4 was %partially 
+planned as two tracks:
 
@@
 
-Suggested contribution topics included:
-
-\begin{itemize}
-\renewcommand{\labelenumi}{\textbf{\theenumi}.}
-\setlength{\rightmargin}{1em}
-
-\item Development and Community:
-best practices for developing sustainable software;
-models for funding specialist expertise in software collaborations;
-software tools that aid sustainability;
-academia/industry interaction;
-refactoring/improving legacy scientific software;
-engineering design for sustainable software;
-metrics for the success of scientific software; and
-adaptation of mainstream software practices for scientific software
-
-\item Professionalization:
-career paths;
-RSE as a brand;
-RSE outside of the UK or Europe; and
-Increase incentives in publishing, funding and promotion for better software
-
-\item Training:
-taining for developing sustainable software; and 
-curriculum for software sustainability
-
-\item Credit:
-making the existing credit and citation ecosystem work better for software;
-future credit and citation ecosystem;
-software contributions as a part of tenure review;
-case studies of receiving credit for software contributions; and
-awards and recognition that encourage sustainable software
-
-\item Software publishing:
-journals and alternative venues for publishing software; and
-review processes for published software
-
-\item Software discoverability/reuse:
-proposals and case studies
-
-\item Reproducibility and testing:
-reproducibility in conferences and journals; and
-best practices for code testing and code review
-
-\end{itemize}
+%Suggested contribution topics included:
+%
+%\begin{itemize}
+%\renewcommand{\labelenumi}{\textbf{\theenumi}.}
+%\setlength{\rightmargin}{1em}
+%
+%\item Development and Community:
+%best practices for developing sustainable software;
+%models for funding specialist expertise in software collaborations;
+%software tools that aid sustainability;
+%academia/industry interaction;
+%refactoring/improving legacy scientific software;
+%engineering design for sustainable software;
+%metrics for the success of scientific software; and
+%adaptation of mainstream software practices for scientific software
+%
+%\item Professionalization:
+%career paths;
+%RSE as a brand;
+%RSE outside of the UK or Europe; and
+%Increase incentives in publishing, funding and promotion for better software
+%
+%\item Training:
+%taining for developing sustainable software; and 
+%curriculum for software sustainability
+%
+%\item Credit:
+%making the existing credit and citation ecosystem work better for software;
+%future credit and citation ecosystem;
+%software contributions as a part of tenure review;
+%case studies of receiving credit for software contributions; and
+%awards and recognition that encourage sustainable software
+%
+%\item Software publishing:
+%journals and alternative venues for publishing software; and
+%review processes for published software
+%
+%\item Software discoverability/reuse:
+%proposals and case studies
+%
+%\item Reproducibility and testing:
+%reproducibility in conferences and journals; and
+%best practices for code testing and code review
+%
+%\end{itemize}
 
@@
 The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Code of Conduct}\label{sec:CoC}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+WSSSPE4 introduced a Code of Conduct (CoC)\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}, which was conceived for the workshop and as guidance for the community of scientists that WSSSPE
+supports, including their personal and online interactions (e.g., on
+Twitter, in email lists, in Slack). This CoC is based on the
+FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
+CoC~\cite{Code4Lib:CoC}.
+The main guidelines are:
+\begin{quote}
+    WSSSPE events are community events intended for networking and collaboration
+    as well as learning. We value the participation of every member of the
+    community and want all attendees to have an enjoyable and fulfilling
+    experience. Accordingly, all attendees are expected to show respect and
+    courtesy to other attendees throughout the event and in interactions online
+    associated with the event.
+
+    The WSSSPE event organizers are dedicated to providing a harassment-free
+    experience for everyone, regardless of gender, gender identity and
+    expression, age, sexual orientation, disability, physical appearance,
+    body size, race, ethnicity, religion (or lack thereof), technology choices,
+    or other group status.
+
+    To make clear what is expected, everyone taking part in WSSSPE events and
+    discussions---speakers, helpers, organizers, and participants---is required
+    to conform to the Code of Conduct\footnoteref{footnote:CoC}.
+
+%     \begin{itemize}
+%     \item Communicate appropriately for a professional audience including
+%     people of many different backgrounds. Sexual language and imagery are not
+%     appropriate for any event.
+
+%     \item Be kind to others. Do not insult or put down other attendees. Be
+%     mindful of jargon, which can sometimes exclude others from engaging in the
+%     discussion.
+
+%     \item Behave professionally. Remember that harassment and sexist, racist,
+%     ageist, or exclusionary behavior are not appropriate.
+%     \end{itemize}
+ \end{quote}
+
+The CoC was discussed at the beginning of WSSSPE4, with the CoC subcommittee
+and a general email address for reporting concerns or incidents, or
+asking questions.  One concern was raised after the first half day, 
+% regarding how part of the workshop was being run, 
+and we changed the workshop
+to address this.
 

### 2017-12-18 WSSSPE4_report/jors-response.tex
work in intro - move CFP topics here, reduce description of previous events, move CoC to new section (3), work on response letter to match

@@
 
-Following these instructions, our initial submission had a body of 13 pages, an appendix of 2 pages, and 3 pages of references (as JORS publishes appendices as separate PDF attachments, and does not typeset them.)
+Following these instructions, our initial submission had 1 title page, a body of 12 pages, an appendix of 2 pages, and 3 pages of references.
+(Note that JORS publishes appendices as separate PDF attachments, and does not typeset them.)
 
 Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.
-The revised version we are resubmitting has been reduced to \textcolor{red}{11} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5} of the original 18 pages.
+The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{10} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5} of the original 18 pages.
 
@@
 
+\end{quoting}
+We've reduced the list of previous events at the start to make this material more of a summary, and less of a list, and moved a little of Section 2 into Section 1 to make the connection of the events and the motivation for WSSSPE4 more strong, including the CFP topics.
+We've also moved the Code of Conduct discussion from the introduction to a new Section (3) to better focus the introduction.
+\begin{quoting}
+
 On the other hand, the level of detail is sometimes overwhelming, and distracts from the main thrust of the paper.

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
minor change in mission and vision section

@@
 via the WSSSPE mailing list, Facebook group, and Twitter.
-After two weeks without suggested changes, the final statements, 
+After two weeks without suggested changes, the final statements\footnote{\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}}
 %as shown below, 
-%were added
-to the WSSSPE web page (\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}).
+were added
+to the WSSSPE website.
 

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
minor grammer - finished reviewed Anshu's changes (Thanks again Anshu!)

@@
 
-Anyone can join this group or obtain more information about it, by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
+Anyone can join this group or obtain more information about it by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
 

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
changing slack section to subsection reducing details about survey mechanisms matching comments in response letter

@@
 forward.
-This report also mentions the Slack channel created for further discussions (\S\ref{sec:slack}), and it highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
+This report also highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
 %An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).
@@
 and we changed the workshop
-to address this.
+to address this concern.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Slack team and channels}\label{sec:slack}
+\subsection{Slack team and channels}\label{sec:slack}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 To join it, go to \url{https://wssspe.signup.team/}.
-A number of the working groups also created channels within the team; see the subsections in \S\ref{sec:WGs} for the names of these channels.
+A number of the working groups also created channels within the team; see the other subsections in \S\ref{sec:WGs} for the names of these channels.
 
@@
 
-All participants present on the last day of the meeting, 14 September 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
-44 responses were recorded, with one respondent indicating that they took the survey twice.
-The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
-Hence, the total number of unique individuals responding to the survey is likely 44.
-The survey contained nine multiple choice and open-text answer questions.
-Text responses have been alphabetized to preserve anonymity.
-The survey had a 100\% completion rate.
-
+At the end of WSSSPE4, all remaining participants completed an on-line survey.
+%
+%All participants present on the last day of the meeting, 14 September 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
+%44 responses were recorded, with one respondent indicating that they took the survey twice.
+%The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
+%Hence, the total number of unique individuals responding to the survey is likely 44.
+%The survey contained nine multiple choice and open-text answer questions.
+%Text responses have been alphabetized to preserve anonymity.
+%The survey had a 100\% completion rate.
+%
 In general, the respondents were highly satisfied with the meeting and interested in continuing and participating in WSSSPE activities.
@@
 Many respondents indicated that they will remain engaged in WSSSPE working group initiatives.
-However, demand for a professional organization encompassing WSSSPE interests was weaker with over half interested in joining and a large percentage willing to consider the idea.
+There was weaker demand for a professional organization encompassing WSSSPE interests, with over half interested in joining and a large percentage willing to consider the idea.
 Respondents were grateful for the opportunity to network, explore new collaborations and for travel support to the conference.
@@
 The survey results indicate that future WSSSPE conferences should consider the balance of the attendees.
-This year's meeting had many first time and early career participants who would have benefited from a review of basic concepts and terminology including what it means for software to be sustainable and the roles of research software engineers.
+WSSSPE4 had many first time and early career participants who would have benefited from a review of basic concepts and terminology including what it means for software to be sustainable and the roles of research software engineers.
 Additional emphasis and topics to explore would be the inclusion of more case studies, focus on the decision making process in developing and using software, deeper dives into selected topics, tutorials, software training within and outside of STEM, lowering the barriers to implementing best practices in software development, and progress towards executing WSSSPE's vision.

### 2017-12-18 WSSSPE4_report/jors-response.tex
changing slack section to subsection reducing details about survey mechanisms matching comments in response letter

@@
 Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.
-The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{10} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5} of the original 18 pages.
+The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{9.5} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5.5} of the original 18 pages.
 
-\textcolor{red}{Can we remove 2 more pages?}
+\textcolor{red}{Can we remove 1.5 more pages?}
 
@@
 
-\textcolor{red}{NEED TO RESPOND}
+A paragraph in the survey section (7) has been removed.
 
@@
 
+\end{quoting}
+We've removed the discussion about the survey respondents and rate, etc.
+
+While we agree that the reviewer has stated a workshop goal of improving research software, this is not something that we intended to do during the workshop itself, so we cannot measure how it has changed over the 2 1/2 days of the workshop.
+We are trying to change the course of the community by pushing on the wheel of the ship, and it will take time to see how the ship responds.
+\begin{quoting}
+
 Similarly, while the code of conduct is provided in full, no context is provided for establishing a code of conduct for such an event.
@@
 In what ways does a code of conduct benefit this event?
+
+\end{quoting}
+We've removed the Code of Conduct itself.
+
+We feel that many events in current society (see the \#metoo movement for example, or one of many discussions about the behavior of scientists at field sites) show how what many might think are common sense rules of conduct are not actually commonly followed.
+At least anecdotally, and clearly in today's society, stating these rules makes in difference in behavior, and makes clear what is unacceptable.
+\begin{quoting}
+
 Instead of discussing these more general issues (maybe only some of them are needed here) an enigmatic comment about an event that required some form of intervention from the organizers is mentioned, but left unexplained.
-Would reporting more fully about what happened be beneficial to the reader?s understanding of the workshop?
+Would reporting more fully about what happened be beneficial to the reader's understanding of the workshop?
 If not, maybe shortening this section would be better?
+
+\end{quoting}
+
+We don't think that reporting on a specific incident is proper.
+
+\begin{quoting}
 
@@
 
+\end{quoting}
+
+\textcolor{red}{This paragraph is the main thing that has not been addressed yet. (Note that what the reviewer calls 4 and 5 are now 5 and 6.)}
+
+\begin{quoting}
+
 Minor:
@@
 1. On page 8, the sentence fragment: ``...later to be fully implemented into by scientific programmers?'' seems to be missing a word or words. Alternatively the word ``into'' should be dropped.
+
+\end{quoting}
+
+We dropped ``into''.
+
+\begin{quoting}
 
@@
 
-3. In section 7: ``Text responses have been alphabetize'' => ``Text responses have been alphabetized''.
 \end{quoting}
 
-We have made these three requested changes.
+We removed the detail about the survey, and initially mention the 68 workshop participants.
 
-%\printbibliography
+\begin{quoting}
+
+3. In section 7: ``Text responses have been alphabetize'' => ``Text responses have been alphabetized''.
+
+\end{quoting}
+
+We've made this change.
 

### 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
fix to abstract to match current paper, minor reduction in presentations section

@@
 The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
-This article includes the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
+This article includes the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
 The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group.
@@
 Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
-Patricia Lago from Vrije Universiteit Amsterdam, who presented a keynote at the workshop  and brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
+Patricia Lago from Vrije Universiteit Amsterdam, who brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
 Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.

### 2018-01-05 WSSSPE4_report/WSSSPE4_paper.tex
adding comments from Reviewer B in conclusions - still need to deal with them

@@
 
+\textcolor{red}{comments from reviewer B to discuss: Finally, in a similar vein, the summary and conclusion offers very little in terms of synthesis of main themes that arose during the meeting.
+The reader is left to synthesize the diversity of sections 4 and 5, that discuss the main activities of the workshop itself, on their own.
+It may very well be that the workshop is diverse enough in terms of its topics (and indeed, it does seem like there are a lot of very interesting topics, and a lot of interesting work done as part of the workshop!), but surely there is some over- arching themes that could be summarized?
+Instead, the conclusion is that a workshop such as this is unlikely to produce tangible work in the long run, because participants are reluctant to commit to additional activities, above and beyond these undertaken during the workshop.
+While this seems like an honest (if disappointing?) conclusion, it is also difficult to reconcile with the multiple places in which the authors ask interested parties to join Slack channels and mailing lists.
+Maybe the expectation that a few hours of joint work at a workshop would lead to a committed collaboration with other participants is too high?
+Rather, expectations from conferences and workshops are usually much lower: exchanging ideas, establishing and broadening a professional network, and learning about work that is being done in the field.
+Maybe it?s enough if these goals are achieved at WSSSPE as well?}
+
 In WSSSPE4, we heard about a number of interesting projects and ideas, and used those

### 2018-01-05 WSSSPE4_report/WSSSPE4_paper.tex
addressing reviewer comment about conclusion section

--- 2017-12-18 WSSSPE4_report/WSSSPE4_paper.tex
@@
 their jobs) come back to the fore. Without a process to resolve this concern, the utility of
-further multi-day WSSSPE workshops is unclear.
+further multi-day WSSSPE workshops is unclear, at least to start new activities.  On a more
+positive note, WSSSPE4 was successful as a gathering place to discuss scientific
+software sustainability, and for groups that are already in place or that can be composed
+of related funded activities to meet.  Future WSSSPE meetings will likely combine these
+functions: a meeting place for like-minded individuals, a place to share experiences and
+lessons learned, and a place for both existing groups and sets of related projects to meet
+in person.  Other parts of the WSSSPE infrastructure (the email list and the Slack) will
+be used both for general discussions, as well as for groups that choose to focus
+on specific problems.
+
+After WSSSPE4, we also have learned enough to summarize the areas of active work
+in sustainable scientific software, based on the WSSSPE4 mission and vision statements, the WSSSPE4 call for participation (based on topics in previous WSSSPE meetings), and the working groups formed in WSSSPE4.
+Following the example of the recently funded US project,
+``Conceptualizing a US Research Software Sustainability Institute (URSSI)\footnote{\url{https://www.nsf.gov/awardsearch/showAward?AWD_ID=1743188}},'' the high-level
+version of these areas is: functioning of 1) the individual and team, 2) the research software, and 3) the research field itself, with six specific topics, as shown in Table~\ref{tab:topics}
+
+\begin{table}[h]
+\centering
+\caption{Scientific Software Sustainability Research Topics and Areas}
+\label{tab:topics}
+\begin{tabular}{|l|c|c|c|}
+ \hline
+Topics \textbackslash\ Areas & Individual and team &  Research software & Research field  \\
+ \hline
+Principles and Best Practices & \checkmark & \checkmark & \checkmark  \\
+ \hline
+Careers & \checkmark &  &   \\
+ \hline
+Learning & \checkmark &  &   \\
+ \hline
+Credit & \checkmark &  & \checkmark  \\
+ \hline
+Organization and community & \checkmark &  &  \checkmark \\
+ \hline
+Common Infrastructure &  & \checkmark &   \\
+ \hline
+\end{tabular}
+\end{table}
 

### 2018-01-05 WSSSPE4_report/jors-response.tex
addressing reviewer comment about conclusion section

@@
 \author{Daniel S. Katz}
-\date{December 2017}
+\date{January 2018}
 
@@
 Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.
-The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{9.5} pages of body, and 3 pages of references, a reduction of \textcolor{red}{5.5} of the original 18 pages.
+The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{10} pages of body, and 3 pages of references, a reduction of \textcolor{red}{4} of the original 18 pages.
 
-\textcolor{red}{Can we remove 1.5 more pages?}
+\textcolor{red}{Can we remove more pages? Sandra is working on Section 6.}
 
@@
 
-\textcolor{red}{This paragraph is the main thing that has not been addressed yet. (Note that what the reviewer calls 4 and 5 are now 5 and 6.)}
+We particularly thank the reviewer for this point.  We realized that this was a weakness, but were unsure how to address it.  We've now followed this useful suggestion, and added to the
+conclusions paragraph.
 

### 2018-01-05 WSSSPE4_report/WSSSPE4_paper.tex
Added commas around "e.g."

@@
 \begin{itemize}
-\item Acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to e.g. Debian's testing infrastructure or to a shared Jenkins-based infrastructure
+\item Acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to, e.g., Debian's testing infrastructure or to a shared Jenkins-based infrastructure
 \item Extending Debian's scope to include published but not mature software

### 2018-01-08 WSSSPE4_report/WSSSPE4_paper.tex
Update WSSSPE4_paper.tex

--- 2018-01-05 WSSSPE4_report/WSSSPE4_paper.tex
@@
 
-Many open source projects for research software document their best practices that
-contribute to the sustainability of the software.  Many of these projects also
-document software metrics they use to define their research software project's success.
+Many open source projects for research software document their best practices and metrics that
+contribute to the sustainability of the software as well as to define their research software project's success.
 %
@@
 This will allow the group to identify gaps on both sides.
-This approach will also allow the group to hypothesize how successful open source projects can be further improved, verify that recommended approaches for software engineering for science/academia are sufficient and valid, and that metrics for software engineering for science/academia are relevant and useful.
-The group has a specific objective of getting a good cross-sampling of disparate software including community model codes, community cyberinfrastructure, community analytic tools, etc.
-
-Each group member will volunteer to take on one to two projects, ideally ones in which they are already interested.
-This effort will evaluate an initial 5--10 projects.
-Once the group documents the workflow by which these evaluations are performed, other people can choose to follow this workflow and contribute their own evaluations of additional software.
-
 Anyone can join this group or obtain more information about it by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
@@
 Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
-These organizations are funded groups or teams that aim to advance research software sustainability beyond their local university or community.
-More specifically, this working group aims to define the scope of this alliance and provide clear distinction with WSSSPE.
-The group also seeks to understand the incentives for members of this alliance and identify its key activities.
 It is currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
-
 Currently, point-to-point collaboration exists between organizations, but this inadvertently results in competition or redundancy within the sustainable software community.
-An alliance of software sustainability organizations would ease inter-organization collaboration and the promotion of software sustainability.
-This alliance would also improve the pooling of competencies and the sharing of expertise.
-Furthermore, with the international scope of this alliance, it could support an organization who wants to hold an event in a country where another organization exists.
-
+An alliance of software sustainability organizations would ease inter-organization collaboration, the promotion of software sustainability as well as would improve the pooling of competencies and the sharing of expertise.
 If interested, one can visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
@@
 tools~\cite{towns2014xsede}, but the development challenge remains.
-Of all scientists who use simulations, only a small percentage know how
-to develop prototypes or full applications that can be later scaled by computer
-scientists and engineers. Writing new parallel applications is unusual for
-domain scientists and thus has created a complex dependency on specialized
-scientific programmers, who are scarce~\cite{post2005computational}. However,
-the capacity to quickly envision and prototype applications by domain
-experts---later to be fully implemented by scientific programmers---is critical for
-acquiring maturity and proficiency in the grand challenges that Exascale
-attempts to solve, and moreover for allowing scientists to develop their own
-ideas quickly and productively~\cite{vinter2015prototyping}. The latter, in
-addition, would increase service utilization time in HPC systems, a critical
-measure of efficiency. Considering the existing code base in scientific
-computing, software infrastructures that allow a clean transition from legacy
-systems to new Exascale platforms while preserving flexible prototyping
-towards the future are central~\cite{hwu2015transitioning}.
-
-This group deems the ability to rapidly construct software artifacts that can be
-trusted in terms of the methods from their design up, easily discarded when
-wrong and extended when right at low human and computational cost to be one
-aspect of scientific software sustainability. In essence,
-a motto for scientific software prototyping is \textit{fail hard, fail fast}.
 %
@@
 for scientific programmers to later flesh out completely and adapt to particular
-infrastructures. This tool will start in the domain of computational physics,
-and (probably) extend to other domains, and the code will be open source.
-%
-The underlying hypothesis of this project is that a strongly typed, declarative
-problem-specification language with adequate constructs, to be implemented
-through stencils and algorithmic skeletons tied to numerical methods, captures a
- subset of useful and interesting problems in physics and other domains.
-%
-The group plans to develop a functional (meta-)prototype before WSSSPE5, and
-if useful, then develop a larger software infrastructure tied to
-particular cyberinfrastructure resources.
+infrastructures.
 
@@
 The CodeMeta project\footnote{\url{http://codemeta.github.io/}} wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software, increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
-CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
-The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.
-
-The primary objective of this working group, which included people who had previously been working on the CodeMeta project, is to find ways to help the project come to fruition.
-%
-The working group plans to engage with those already working on CodeMeta, to examine the existing CodeMeta crosswalk table to see what improvements and additions might be made, and to determine how to engage the community and provide ongoing social engagement and structure.
-Further, the group wants to assist in the implementation of the specifications and, by providing an outsider's view, contribute suggestions for more understandable project documentation.
-Finally, the group seeks a better way or ways to present the crosswalk table to make it more easily understood and consumable by research software communities.
 %
@@
 %
-For the first white paper, the group aims to set the
-stage with successful use cases and an analysis of why they have been successful.
-Further topics will include community-related practices, government and management, funding,
-metrics, tools, and usability.
-The resulting draft white paper will be distributed via the WSSSPE email list.
-After collecting feedback on it, the group's plans are to attract a wider community to contribute to an extended journal paper version and to investigate whether more people would like to be involved in the white paper series.
-
 The GitHub repository for the white paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
@@
 However, academic research in general is often siloed for many reasons, and work on topics relevant to software sustainability is no exception.
-There are many academic studies and projects which may be relevant for practitioners working in this area, and many projects and initiatives by practitioners that may be relevant for social scientists.
-However, there is a gap between these two in research and practice.
 %
@@
 %
- In any domain, bringing research and practice closer together a mutually beneficial goal, but also has its challenges.
-The group aims to give social scientists and practitioners working on scientific software, software sustainability, and open source and a better understanding of each other's work, as well as help them connect and coordinate on specific projects of mutual interest.
-%
-Furthermore, the group also recognizes that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
-
 The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
@@
 individual researchers in an unsustainable way that is never shared
-with the larger community. Participants recalled their own experience
-working with colleagues who write hidden code or inheriting
-unsustainable hidden code from collaborators.
-%
-The question is, then, how to prevent researchers from writing hidden
-code? The participants hypothesized that the best strategy is to catch
+with the larger community. 
+%
+The participants hypothesized that the best strategy is to catch
 researchers while they are still in training and teach them software
@@
 
-While implementation of a course aimed at each domain science is a long term
-goal, a short-term, achievable goal, is to develop a
-curriculum for a course aimed at a single domain science. Since the
-participants in the working group have expertise in physics, this is a
-natural target.
-%
 The development of a successful curriculum relies on the expertise of
 software engineers to describe the best practices, domain experts to
-describe model problems and work-flow, and instructors to formulate
+describe model problems and workflow, and instructors to formulate
 the pedagogy. Ideally these people will be brought together for a
 short workshop or hackathon with the goal of drafting the
-curriculum. The
-hackathon will be implemented and a draft of the curriculum will be
-written within the next year.
-Later, organizational partners will be required to
-actually implement the program.
-
+curriculum. 
 Anyone interested in contributing should get in touch via one of the
@@
 
-In the context of this aim, the group hopes to find an efficient solution for streamlining the process of collecting and utilizing metrics to benefit software sustainability, as well as minimize the current inefficiencies within the scientific software community.
-This will improve software evaluation and comparison, thus reducing the effort spent on seeking scientific software.
-%
-The group plans on surveying and interviewing scientific software developers to form metrics from the goals they have for their software, then to publicize the results.
-
 For more information, one can contact Emily Chen at <echen35@illinois.edu>.
@@
 For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
-(Note that Debian provides a testing infrastructure\footnote{\url{https://ci.debian.net/doc/}} that is mature and supports hardware and other requirement specifications.)
 
@@
 %
-The scope of the group's interest is any type of testing, though interactive access for troubleshooting would be of particular interest beyond just automated testing.
-
-However, there are a lot of open issues:
-\begin{itemize}
-\item Some similar work has already been done.  How can this group apply and/or learn from that work, rather than reinventing the wheel?
-\item How to ensure this will work across disciplines?
-\item Meaningful CI for large projects may need hundreds of CPU hours per day
-\end{itemize}
-
-Some possible goals include:
-\begin{itemize}
-\item Acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to e.g. Debian's testing infrastructure or to a shared Jenkins-based infrastructure
-\item Extending Debian's scope to include published but not mature software
-\item Implementing the same interface but with specialized hardware or available to non-public codes.
-\item \url{https://reproducible-builds.org} but for CI
-\end{itemize}
+
+However, there are a lot of open issues such as how can the group learn from existing similar work without reinventing the wheel.
+Some possible goals include to acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to, e.g., Debian's testing infrastructure or to extend Debian's scope to include published but not mature software.
 
@@
 validity of their interactions.
-As scientific understanding grows,
-the corresponding computational software models are refined, leading
-to more complex codes. Increasing complexity makes them more prone to
-defects, not only in individual code units, but also in interaction
-among units. Therefore, a strong verification process combined with a
-rigorous testing regime plays a critical role in the prevention of
-generating incorrect scientific results. However, most science teams
-struggle to find a good solution for themselves.
-
-One cause is lack of exposure to the practices. While
-good developers will test their code to
-verify that it operates as expected, they may not appreciate
-that without regular testing, defects can be introduced
-inadvertently.
-%
-An even bigger challenge is that those who understand
+A big challenge is that those who understand
 the importance of regular testing do not often find much help from
@@
 the testing gospel and its applicability to computational science. This
-gap leads to frustration and abandonment of the good with the bad. Some
-relevant literature exists, in particular experiences from
-practitioners in computational science who developed their own
-solutions. However, this literature is scattered among many different
-forums, and can be challenging to find. This working group aims to
-address this gap by curating the existing content and contributing
-content where none exists.
+gap leads to frustration and abandonment of the good with the bad.
 
@@
 
-The working group's plans are relatively simple to express, though quite complex to undertake.
-They are to first determine a plan to build an open research index that allows various stakeholders to satisfy their needs, then to determine if the plan is feasible.
-At the time of the meeting, and today as well, it is unclear who has the time and energy to pursue this idea. Thus, a leader needs to be identified.
+The working group's plans are relatively simple to express, though quite complex to undertake. A leader needs to be identified.
 With a leader and feasible plan, the group would then obtain resources, and then build the index.
@@
 
-This group believes that the authors of scientific software provide important
-services to departments that are no less than strictly disciplinary
-research. Consequently, leveling the playing field with more
-disciplinary candidates for hires, tenure, and promotion requires that
-letter writers be aware of how their letters will be read by
+Consequently, letter writers neet to be aware of how their letters will be read by
 committees. It also requires that committees be aware that such
 letters often look different and may provide a different perspective
-of how a candidate's achievements should be assessed. For example, in
-mathematics a typical candidate would be evaluated on the difficulty
-and depth of the statements she may have proven in their papers. On
-the other hand, an author of mathematical software would likely be
-evaluated based on the impact of her software, or the number of citations of
-the publications that describe it. She may also be evaluated by how
-widely the software is used \textit{outside} mathematics, a criterion
-that is rarely used for more disciplinary mathematicians.
-
-Addressing this problem likely requires that letter writers pay
-particular attention to who exactly the audience of such letters is,
-and tailor the message by specifically highlighting how the work of a
-candidate benefits the department and discipline that is considering a
-candidate. On the other hand, committee members also need to pay
-attention to the fact that there are areas that are important to the mission of
-their department and professional community in which different
-standards for evaluation hold.
+of how a candidate's achievements should be assessed.
+
 %
@@
 other: (i) raising awareness of the problem beyond just those who are
-affected by it (namely, computational scientists working on scientific
-software), and (ii) providing letter writers, letter readers, and
+affected by it, and (ii) providing letter writers, letter readers, and
 evaluating committees with guidance on what criteria are relevant in
 assessing scientific software authors.
-
-Concrete guidance is most valuable if it
-comes from respected bodies such as professional societies or
-established and respected organizations. Getting these to act
-will only be possible if the problem is widely acknowledged, and so
-the first of the goals above should be the current focus.
-%
-The group will attempt to address it by organizing sessions at conferences
-that address the career problem, as well as writing editorials that
-can be published in the magazines of professional societies. These
-editorials ought to outline best practices for letter writers that
-specifically (i) make clear the contribution of a candidate to their
-interdisciplinary area, (ii) the relevance to their home discipline, and
-(iii) why writing software is good for the discipline itself.
-
+%
 Ultimately, building a community large enough

### 2018-01-08 WSSSPE4_report/WSSSPE4_paper.tex
adding back a little of Sandra removed removing 6.13, moving Slack description to start of 6 updating reviewer response letter to match

@@
 Summaries of each group's activities follow; for more details
-about the group discussions, see \cite{WSSSPE4-report}.
+about these discussions, see \cite{WSSSPE4-report}.
+There is a Slack team for WSSSPE~\cite{WSSSPESlack} that one can join at \url{https://wssspe.signup.team/}; some working groups also created channels within this team.
 
@@
 
-Many open source projects for research software document their best practices and metrics that
-contribute to the sustainability of the software as well as to define their research software project's success.
-%
-This group will take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
-This will allow the group to identify gaps on both sides.
-Anyone can join this group or obtain more information about it by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
+Many open source research software projects document their best practices that
+contribute to the sustainability of the software as well as the metrics that define their research software project's success.
+%
+This group will take the outputs of the WSSSPE efforts to identify best development practices and to identify metrics for sustainable research software, and cross-reference them with current open source research software that successfully uses modern software engineering.
+This will allow the group to identify gaps on both sides and to hypothesize how successful projects can be further improved.
+One can join this group or obtain more information about it by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
 
@@
 
-The Software Sustainability Alliance working group aims to establish an alliance between organizations interested in improving the sustainability of research software.
+The Software Sustainability Alliance working group aims to establish an alliance between organizations interested in improving the sustainability of research software at a larger scale than their local university or community, and to distinguish between this group and WSSSPE.
 Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
-It is currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
+It is currently seeking feedback on potential activities and member organizations, as well as the aims and scope of this alliance.
 Currently, point-to-point collaboration exists between organizations, but this inadvertently results in competition or redundancy within the sustainable software community.
-An alliance of software sustainability organizations would ease inter-organization collaboration, the promotion of software sustainability as well as would improve the pooling of competencies and the sharing of expertise.
-If interested, one can visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
+An alliance of software sustainability organizations would ease inter-organization collaboration and the promotion of software sustainability, andwould improve the pooling of competencies and the sharing of expertise.
+If interested, one can visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<salac@uchicago.edu>)
 
@@
 There is a productivity bottleneck---yet to be solved---in HPC from the human
-perspective, first identified in the 1980s~\cite{barstow1982automatic}. Of all
-domain-specific scientists, only a percentage of those use simulations due to
-perceptions mostly related to the difficulty of using and developing those tools.
+perspective, first identified in the 1980s~\cite{barstow1982automatic}. Only a fraction of
+domain-specific scientists use simulations, mostly due to
+perceptions about the difficulty of using and developing such tools.
 XSEDE and other resources have become a gateway for successfully increasing the
@@
 tools~\cite{towns2014xsede}, but the development challenge remains.
+%
+This group deems the ability to rapidly construct software artifacts that can be
+trusted in terms of the methods from their design up, easily discarded when
+wrong and extended when right at low human and computational cost to be one
+aspect of scientific software sustainability.
 %
@@
 infrastructures. 
-
-To join this project, one can visit the GitHub repository at
-\url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel
-\texttt{\#wg-sci-soft-proto}.
+%
+To join this project, one can visit the GitHub repository 
+(\url{https://github.com/nunezco2/S2PI}) or join \texttt{\#wg-sci-soft-proto} in the WSSSPE Slack.
 
@@
 %
+CodeMeta seeks in part to create a ``Rosetta Stone'' for software metadata to facilitate retaining such metadata between repositories, services, registries, indexers, publishers, citation managers, and other entities that create, ingest, use, and/or store metadata about software.
+The project also wants to establish a JSON-LD schema as a tool for making metadata machine-readable~\cite{CodeMeta_schema}.
+% 
 While at WSSSPE4, the group greatly expanded the CodeMeta project README file to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
-
-Though a Google group mailing list has been established for the working group, the easiest way to engage with the CodeMeta project is through its Github repository\footnote{\url{https://github.com/codemeta/codemeta}}.
+%
+Though a Google group mailing list has been established for the working group, the easiest way to engage with the CodeMeta project is through its Github repository: \url{https://github.com/codemeta/codemeta}.
 
@@
 
-Many diverse aspects and dimensions (such as economic, technical, environmental, and social) of developing sustainable software can be investigated,
+Many diverse aspects and dimensions (e.g., economic, technical, environmental, social) of developing sustainable software can be investigated.
 This group aims to write white papers
-that will focus on scientific environments and their implications, targeted at developers
+that focus on scientific environments and their implications, targeted at developers
 and project managers of scientific software. Given the complexity of this field, it is important to select
@@
 The GitHub repository for the white paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
-For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.
+For more information and requests, join the \texttt{\#wg-best-practices} channel in the WSSSPE Slack.
 
@@
 %
+Furthermore, the group also recognizes that academic social science and research software engineering are not monolithic, and there is a need to connect people who care about research-driven best practices inside of these two domains with each other as well.
+%
 The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
-This plan resulted in the creation of the following SMART steps listed in the next section.
-
+%
 To join the group, one can join its mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
@@
 researchers while they are still in training and teach them software
-best-practices. Therefore, this working group's goal
+best practices. Therefore, the working group's goal
 is to develop courses on software best practices aimed at
 undergraduate students studying domain science. The program might be
-similar to a Software Carpentry or Data Carpentry workshop but
+similar to Software and Data Carpentry workshops but
 focused for domain scientists.
@@
 curriculum. 
-Anyone interested in contributing should get in touch via one of the
-following methods:
-
-\begin{itemize}
-\item Join the Google Group/mailing list
-  \cite{WSSSPEUndergradGoogleGroup}.
-\item Join the WSSSPE Slack (\S\ref{sec:slack}) and the
-  \texttt{\#wg-undergraduatecourse} channel.
-\item Ask to join the WSSSPE4-undergraduate-course organization
-  \cite{WSSSPEUndergradGithub}.
-\end{itemize}
+Anyone interested in contributing can join the Google Group/mailing list (\url{https://groups.google.com/forum/#!forum/
+wssspe4-software-best-practices-for-undergraduates}), 
+join the \texttt{\#wg-undergraduatecourse} channel in the WSSSPEs Slack,
+or ask to join the WSSSPE4-undergraduate-course organization (\url{https://github.com/orgs/WSSSPE4-
+undergraduate-course}).
 
@@
 In order to address these inefficiencies, the group aims to create a goal-oriented method to collecting productive metrics by focusing on the developer side of software.
-
+The group plans to survey and interview scientific software developers to form metrics from the goals they have for their software, then to publicize the results
+%
 For more information, one can contact Emily Chen at <echen35@illinois.edu>.
@@
 For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
-
+%
 This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems (when publicly available CI are not a fit), by coordinating and sharing this burden across multiple projects.
 %
-
 However, there are a lot of open issues such as how can the group learn from existing similar work without reinventing the wheel.
 Some possible goals include to acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to, e.g., Debian's testing infrastructure or to extend Debian's scope to include published but not mature software.
-
-For more information, see \url{https://groups.google.com/forum/#!forum/continuous-integration-for-research-software}
+%
+For more information, see \url{https://groups.google.com/forum/#!forum/continuous-integration-for-research-software}.
 
@@
 gap leads to frustration and abandonment of the good with the bad. 
-
+%
 This working group will (1) conduct a literature survey to gauge the extent
@@
 content for the use of the community.
-
-Readers interested in getting more information should use the working group'
-channel in the WSSSPE Slack (\S\ref{sec:slack}), called
-\texttt{\#wg-testing-in-science}.
+%
+To join this effort, one can use the \texttt{\#wg-testing-in-science}
+channel in the WSSSPE Slack.
 Additionally, a git
@@
 %
-The Open Research Index should take in all research products (papers, software, datasets, workflows, etc.) from their publishers and recorders (journals, societies, domain repositories, government [open access] repositories, preprint servers, general repositories [e.g., figshare, zenodo]) and other services (CrossRef, ORCID).
-Each product should list authors and citations and allow people to search the resulting network.
-Users should also be able to interact with their own record and edit it, like Google Scholar allows.
-
-The working group's plans are relatively simple to express, though quite complex to undertake. A leader needs to be identified.
+The Open Research Index would take in all research products (papers, software, datasets, workflows, etc.) from their publishers and recorders (journals, societies, domain repositories, government [open access] repositories, preprint servers, general repositories [e.g., figshare, zenodo]) and other services (CrossRef, ORCID).
+Each product would list authors and citations and allow people to search the resulting network.
+Users would also be able to interact with their own record and edit it, as Google Scholar allows.
+%
+The group's plans are relatively simple to express, though quite complex to undertake; the first is to identify a leader.
 With a leader and feasible plan, the group would then obtain resources, and then build the index.
@@
 Because Google and others provide some similar services today, though these services (and the underlying data) could be removed at any time and the community cannot build new services, these companies would ideally be involved in the activities of the group.
-
+%
 Alternatively or additionally, the group could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
-At WSSSPE4, there seemed to be enough interest to do this, so the group set up a Slack channel within the WSSSPE team (\S\ref{sec:slack}), called \texttt{\#wg-open-research-idx} for its members and any future members.
+At WSSSPE4, there seemed to be enough interest to do this, so the group set up a \texttt{\#wg-open-research-idx} channel within the WSSSPE team Slack for its members and any future members.
 
@@
 letters of evaluation for hiring, tenure, and promotion do not specifically
-cover how this is relevant to the discipline.\footnote{The extensive use of such letters, and the problems that are
-  associated with it, may be an issue specific to the United States.}
-
-Consequently, letter writers neet to be aware of how their letters will be read by
+cover how this is relevant to the discipline.\footnote{The extensive use of such letters, and the problems that are associated with it, may be an issue specific to the United States.}
+%
+Consequently, letter writers need to be aware of how their letters will be read by
 committees. It also requires that committees be aware that such
@@
 of how a candidate's achievements should be assessed. 
-
-%
-There are essentially two important strategies that build on each
+%
+This group has two important strategies that build on each
 other: (i) raising awareness of the problem beyond just those who are
@@
 Ultimately, building a community large enough
-to affect change is important; contact Wolfgang Bangerth at
-<bangerth@colostate.edu> if you are interested in helping.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Slack team and channels}\label{sec:slack}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE~\cite{WSSSPESlack}.
-To join it, go to \url{https://wssspe.signup.team/}.
-A number of the working groups also created channels within the team; see the other subsections in \S\ref{sec:WGs} for the names of these channels.
+to affect change is important; one may contact Wolfgang Bangerth at
+<bangerth@colostate.edu> if interested in helping.
 

### 2018-01-08 WSSSPE4_report/jors-response.tex
adding back a little of Sandra removed removing 6.13, moving Slack description to start of 6 updating reviewer response letter to match

@@
 Nonetheless, we agree with the reviewers that this can be reduced, and have done so by removing the appendix and as advised by the reviewers, as further discussed below.
-The revised version we are resubmitting has been reduced to 1 title page, \textcolor{red}{10} pages of body, and 3 pages of references, a reduction of \textcolor{red}{4} of the original 18 pages.
-
-\textcolor{red}{Can we remove more pages? Sandra is working on Section 6.}
+The revised version we are resubmitting has been reduced to 1 title page, 7.5 pages of body, and 2.75 pages of references, a reduction of 6.75 of the original 18 pages.
 

### 2018-02-01 WSSSPE4_report/WSSSPE4_paper.tex
2 changes as suggested in email by Wolfgang
small changes related to proofreading for JORS

@@
 \usepackage{csquotes}
+
+\usepackage{cite}
 
@@
 %
-\thanks{{}$^{(8)}$ University of Virginia, Charlottesville, VA, USA; jeansalac@virginia.edu; ORCID: 0000-0001-5971-9333}
+\thanks{{}$^{(8)}$ University of Chicago, Chicago, IL, USA; salac@uchicago.edu; ORCID: 0000-0001-5971-9333}
 %
@@
 The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
-This article includes the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
+This article discusses the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
 The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group.
@@
 CoC~\cite{Code4Lib:CoC}.
-The main guidelines are:
+In introducing the CoC, we asked participants to agree to the following main guidelines:
 \begin{quote}

### 2018-03-28 WSSSPE5.1_report/WSSSPE5.1_paper.tex
adding start of WSSSPE5.1 report

--- 2018-01-08 WSSSPE4_report/WSSSPE4_paper.tex
@@
 
-\title[]{Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4)}
+\title[]{Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) \& The State of Sustainable Software -- TITLE NEEDS WORK}
 
 \author{Daniel~S.~Katz$^{(1)}$,
-Kyle~E.~Niemeyer$^{(2)}$,
-Sandra~Gesing$^{(3)}$,
-Lorraine~Hwang$^{(4)}$,
-Wolfgang~Bangerth$^{(5)}$,
-Simon~Hettrick$^{(6)}$,
-Ray Idaszak$^{(7)}$,
-Jean~Salac$^{(8)}$,
-Neil~Chue~Hong$^{(9)}$,
-Santiago~N\'u\~nez-Corrales$^{(10)}$,
-Alice~Allen$^{(11)}$,
-R.~Stuart~Geiger$^{(12)}$,
-Jonah~Miller$^{(13)}$,
-Emily~Chen$^{(14)}$,
-Anshu~Dubey$^{(15)}$,
-Patricia~Lago$^{(16)}$
+Caroline $^{(2)}$,
+Rob $^{(3)}$,
+Alexander $^{(4)}$,
+Stephan\ $^{(5)}$\
+please add full names, affiliations, orcids
+%Kyle~E.~Niemeyer$^{(2)}$,
+%Sandra~Gesing$^{(3)}$,
+%Lorraine~Hwang$^{(4)}$,
+%Wolfgang~Bangerth$^{(5)}$,
+%Simon~Hettrick$^{(6)}$,
+%Ray Idaszak$^{(7)}$,
+%Jean~Salac$^{(8)}$,
+%Neil~Chue~Hong$^{(9)}$,
+%Santiago~N\'u\~nez-Corrales$^{(10)}$,
+%Alice~Allen$^{(11)}$,
+%R.~Stuart~Geiger$^{(12)}$,
+%Jonah~Miller$^{(13)}$,
+%Emily~Chen$^{(14)}$,
+%Anshu~Dubey$^{(15)}$,
+%Patricia~Lago$^{(16)}$
 %\\Author order to be determined based on contributions
@@
 %
-\thanks{{}$^{(2)}$ School of Mechanical, Industrial, and Manufacturing Engineering,
-Oregon State University, Corvallis, OR, USA; kyle.niemeyer@oregonstate.edu; ORCID: 0000-0003-4425-7097}
+\thanks{{}$^{(2)}$ department/school, city, country; email; ORCID: orcid}
 %
-\thanks{{}$^{(3)}$ Center for Research Computing \& Department of Computer Science and Engineering,
-University of Notre Dame, IN, USA; sandra.gesing@nd.edu; ORCID: 0000-0002-6051-0673}
+\thanks{{}$^{(3)}$ department/school, city, country; email; ORCID: orcid}
 %
-\thanks{{}$^{(4)}$ University of California, Davis, CA, USA; ljhwang@ucdavis.edu; ORCID: 0000-0002-1021-3101}
+\thanks{{}$^{(4)}$ department/school, city, country; email; ORCID: orcid}
 %
-\thanks{{}$^{(5)}$ Department of Mathematics, Colorado State
-  University, Fort Collins, CO, USA; bangerth@colostate.edu; ORCID: 0000-0003-2311-9402}
+\thanks{{}$^{(5)}$ department/school, city, country; email; ORCID: orcid}
 %
-\thanks{{}$^{(6)}$ The Software Sustainability Institute, Southampton, UK; s.hettrick@software.ac.uk; ORCID: 0000-0002-6809-5195}
-%
-\thanks{{}$^{(7)}$ RENCI, University of North Carolina at Chapel Hill, Chapel Hill, NC, USA; rayi@renci.org; ORCID: 0000-0002-3444-7615}
-%
-\thanks{{}$^{(8)}$ University of Virginia, Charlottesville, VA, USA; jeansalac@virginia.edu; ORCID: 0000-0001-5971-9333}
-%
-\thanks{{}$^{(9)}$ University of Edinburgh, UK; N.ChueHong@epcc.ed.ac.uk;
-ORCID: 0000-0002-8876-7606}
-%
-\thanks{{}$^{(10)}$ Illinois Informatics Institute (I3) and National Center for Supercomputing Applications (NCSA), University of Illinois at Urbana-Champaign, IL, USA;
-nunezco2@illinois.edu;
-ORCID: 0000-0003-4342-6223}
-%
-\thanks{{}$^{(11)}$ Astrophysics Source Code Library, Astronomy Department, University of Maryland, College Park, MD, USA; aallen@ascl.net; ORCID: 0000-0003-3477-2845}
-%
-\thanks{{}$^{(12)}$ University of California, Berkeley, Berkeley Institute for Data Science (BIDS), Berkeley, CA, USA; stuart@stuartgeiger.com; ORCID: 0000-0001-7215-0532}
-%
-\thanks{{}$^{(13)}$ Perimeter Institute for Theoretical Physics \&
-Department of Physics, University of Guelph,
-Guelph, ON, Canada;
-jmiller@perimeterinstitute.ca; ORCID: 0000-0001-6432-7860}
-%
-\thanks{{}$^{(14)}$ University of Illinois at Urbana-Champaign, USA; echen35@illinois.edu; ORCID: 0000-0003-2578-0296}
-%
-\thanks{{}$^{(15)}$ Mathematics and Computer Science Division,
-Argonne National Laboratory,
-Lemont, IL, USA \&
-University of Chicago,
-Chicago, IL, USA;
-Adubey@anl.gov;
-ORCID: 0000-0003-3299-7426}
-%
-\thanks{{}$^{(16)}$ Vrije Universiteit Amsterdam, The Netherlands; p.lago@vu.nl; ORCID: 0000-0002-2234-0845}
 
 \begin{abstract}
-
+\todo{Need to update}
 This article summarizes motivations, organization, and activities of the Fourth Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE4).
 The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
-This article includes the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
+This article discusses the code of conduct; the mission and vision statements that were drafted at the workshop and finalized shortly after it; the keynote and idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop; and a panel discussion on best practices.
 The main part of the article discusses the set of working groups that formed during the meeting, along with contact information for readers who may want to join a group.
@@
 
-The Fourth Workshop on Sustainable Software for Science: Practice and Experiences
-(WSSSPE4)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe4/}} was
-held over 2 1/2 days on 12--14 September 2016 in Manchester, England, with 68 attendees.
-This location and date was selected so that WSSSPE4 immediately preceded the
-First Research Software Engineers (RSE) Conference.
+\todo{Dan}
+
+The Workshop on Sustainable Software for Science: Practice and Experiences
+(WSSSPE5.1)\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe5/}} \katznote{check URL} was a one-day meeting held in September 2017 in Manchester, England immediately preceding the
+Second Research Software Engineers (RSE) Conference, so that RSE attendees could also attend WSSSPE5.1.  \katznote{number of attendees?}
+
 Previous WSSSPE events\footnote{The first WSSSPE workshop was named
@@
 Software for Science: Practice and Experiences.'' Together these reflect
-that WSSSPE is both a community and a set of workshops.}, all of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1, WSSSPE2-pre-report,WSSSPE2},
-two half-day workshops\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}, \url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}} held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, and a 1-and-1/2-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}.
+that WSSSPE is both a community and a set of workshops.}, all but the last of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1, WSSSPE2-pre-report,WSSSPE2},
+two half-day workshops\footnote{\url{http://wssspe.researchcomputing.org.uk/wssspe1-1/}, \url{http://wssspe.researchcomputing.org.uk/wssspe2-1/}} held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, a 1-and-1/2-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}, and a 2-and-1/2-day workshop that immediately preceded the
+First Research Software Engineers (RSE) Conference in Manchester, England.
+~\cite{WSSSPE4}
 %
@@
 
-This paper is a summary of a longer report~\cite{WSSSPE4-report} on the workshop and the submitted materials.
 The remainder of the paper includes the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); a set of presentations that included a keynote, papers, lighting talks, and a panel (\S\ref{sec:presentations}); and the activities of a set of working groups (\S\ref{sec:WGs}).
@@
 This report also highlights an attendee survey (\S\ref{sec:survey}) before concluding (\S\ref{sec:conclusions}).
-%An appendix lists the organizers and program committee members (\S\ref{sec:orgcom}).
+
+\katznote{new text below, need to replace lots of what is above:}
+
+In October 2017, a group of people interested in sustainable research software came together at the Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) meeting in Manchester, UK.
+This workshop built on six previous WSSSPE workshops, all of which dedicated significant portions of the meeting to group discussions about problems and potential solutions in the sustainable research software space.
+WSSSPE5.1 used the speed blog methodology to generate 8 reports on different views of the space.
+These blogs, previously published by the UK Software Sustainability Institute (SSI) on its website, form the second portion of this paper.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Call for participation} \label{sec:preworkshop}
+\section{Call for participation and response} \label{sec:preworkshop}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-Post-WSSSPE3 feedback said that attendees
-had two distinct motivations for attending:  to make a better future
-for research software, and to immediately do better research software development. Thus,
-WSSSPE4 was %partially 
-planned as two tracks:
+\todo{Stephan, maybe you can put something short here?}
 
-\begin{quote}
-    \textbf{Track 1 -- Building a sustainable future for open-use research
-    software}: defining a vision of the future of open-use
-    research software, and in the workshop, initiating the activities that are
-    needed to get there.
-
-    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
-    improving the quality of today's research software and the
-    experiences of its developers by sharing best practices and experiences.
-\end{quote}
-
-The call for participation requested:
-idea papers, implementable proposals related to Track~1;
-position papers, longer, not previously published papers
-discussing what we can do to improve sustainable scientific
-software in the short term;
-experience papers, longer papers that discuss current
-experiences and how they have been used to improve the quality of
-today's research software and/or the experiences of its developers;
-demos, brief papers describing a tool or
-process that would be demonstrated that improves the quality of today's research
-software and\slash or the experiences of its developers; and
-extended abstracts describing lightning talks,
-where each author could make a brief statement about work that either had been
-done or was needed.
+%Post-WSSSPE3 feedback said that attendees
+%had two distinct motivations for attending:  to make a better future
+%for research software, and to immediately do better research software development. Thus,
+%WSSSPE4 was planned as two tracks:
+%
+%\begin{quote}
+%    \textbf{Track 1 -- Building a sustainable future for open-use research
+%    software}: defining a vision of the future of open-use
+%    research software, and in the workshop, initiating the activities that are
+%    needed to get there.
+%
+%    \noindent \textbf{Track 2 -- Practices \& experiences in sustainable scientific software}:
+%    improving the quality of today's research software and the
+%    experiences of its developers by sharing best practices and experiences.
+%\end{quote}
+%
+%The call for participation requested:
+%idea papers, implementable proposals related to Track~1;
+%position papers, longer, not previously published papers
+%discussing what we can do to improve sustainable scientific
+%software in the short term;
+%experience papers, longer papers that discuss current
+%experiences and how they have been used to improve the quality of
+%today's research software and/or the experiences of its developers;
+%demos, brief papers describing a tool or
+%process that would be demonstrated that improves the quality of today's research
+%software and\slash or the experiences of its developers; and
+%extended abstracts describing lightning talks,
+%where each author could make a brief statement about work that either had been
+%done or was needed.
 
@@
 were accepted as lightning talks.
-% Specifically,
-% 19 lightning talks,
-% 4 idea papers,
-% 3 position papers,
-% 2 experience papers,
-% and
-% 3 demos
-% were accepted, as mentioned in Section~\ref{sec:presentations}.
-%
 The papers and lightning talks have been published as a volume in the CEUR Workshop Proceedings~\cite{WSSSPE4-proceedings}.
@@
 
-WSSSPE4 introduced a Code of Conduct (CoC)\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}}, which was conceived for the workshop and as guidance for the community of scientists that WSSSPE
-supports, including their personal and online interactions (e.g., on
+\todo{do we want to keep some discussion of this here?  maybe fold in into the intro?}
+
+WSSSPE5.1 included a Code of Conduct (CoC)\footnote{\label{footnote:CoC}\url{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}} as guidance for the community of scientists that WSSSPE
+supports, including the workshop and their personal and online interactions (e.g., on
 Twitter, in email lists, in Slack). This CoC is based on the
@@
 
-%     \begin{itemize}
-%     \item Communicate appropriately for a professional audience including
-%     people of many different backgrounds. Sexual language and imagery are not
-%     appropriate for any event.
-
-%     \item Be kind to others. Do not insult or put down other attendees. Be
-%     mindful of jargon, which can sometimes exclude others from engaging in the
-%     discussion.
-
-%     \item Behave professionally. Remember that harassment and sexist, racist,
-%     ageist, or exclusionary behavior are not appropriate.
-%     \end{itemize}
  \end{quote}
 
-The CoC was discussed at the beginning of WSSSPE4, with the CoC subcommittee
+The CoC was discussed at the beginning of WSSSPE5.1, with the CoC subcommittee
 and a general email address for reporting concerns or incidents, or
-asking questions.  One concern was raised after the first half day, 
-% regarding how part of the workshop was being run, 
-and we changed the workshop
-to address this concern.
+asking questions.
 
@@
 
-%\note{by Lorraine and Dan}
+\todo{should we take any of this into the intro, or drop it all?}
 
-Going into WSSSPE4, WSSSPE had no formal mission or vision statement.
-The organizers developed a strawhorse, which was presented to the participants early in the meeting.
-The presentation included guidelines as well as examples from other similar communities.
-The guidelines were that in general, an organization's mission should stand the test of time and state what the organization does; its vision should imagine what the world would look like if the organization is successful; and from which, focus areas could be used to establish the scope of the organization along with its goals.
+ {\bf Mission.}
+ WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
 
-The participants were 
-%given time after the presentation to write down their  
-invited to comment,
-% Seven people volunteered to work on 
+ {\bf Vision.}
+ We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
+
+ {\bf Focus areas.}
+ WSSSPE promotes sustainable research software by positively impacting:
+ \begin{itemize}
+ \item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
+ \item {\bf Careers}. Developing and supporting career paths in research software development and engineering
+ \item {\bf Learning}. Engaging in activities to promote peer learning and interaction
+ \item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
+ \end{itemize}
+
+ \textbf{Definitions:}
+ \emph{Sustainable software} has the capacity to endure such that it will continue to
+ be available in the future, on new platforms, meeting new needs.
+ The \emph{research software lifecycle} includes:
+ acquiring and assembling resources (including funding and people) into teams and communities,
+ developing software,
+ using software,
+ recognizing contributions to and of software,
  and
-%revising 
-the mission and vision statements, were revised 
-%incorporating these comments and presenting back to the community.
-based on this feedback.
-% the committee redrafted the mission and vision.
-The committee added focus areas to clarify the organization's breadth and to keep the mission and vision simple and long-lasting.
-Comments after the close of the meeting were also incorporated into the statements.
-A final draft was put on GitHub and advertised
-via the WSSSPE mailing list, Facebook group, and Twitter.
-After two weeks without suggested changes, the final statements\footnote{\url{http://wssspe.researchcomputing.org.uk/about-wssspe/}}
-%as shown below, 
-were added
-to the WSSSPE website.
-
-% {\bf Mission.}
-% WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
-
-% {\bf Vision.}
-% We envision a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.
-
-% {\bf Focus areas.}
-% WSSSPE promotes sustainable research software by positively impacting:
-% \begin{itemize}
-% \item {\bf Principles and Best Practices}. Promoting best practices in sustainable software
-% \item {\bf Careers}. Developing and supporting career paths in research software development and engineering
-% \item {\bf Learning}. Engaging in activities to promote peer learning and interaction
-% \item {\bf Credit}. Ensuring recognition of research software as an intellectual contribution equal to other research products
-% \end{itemize}
-
-% \textbf{Definitions:}
-% \emph{Sustainable software} has the capacity to endure such that it will continue to
-% be available in the future, on new platforms, meeting new needs.
-% The \emph{research software lifecycle} includes:
-% acquiring and assembling resources (including funding and people) into teams and communities,
-% developing software,
-% using software,
-% recognizing contributions to and of software,
-% and
-% maintaining software.
+ maintaining software.
 
@@
 
-%\note{from Patricia}
+\todo{Stephan}
 
-The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
-%
-Sustainability is broadly associated with natural ecologic systems.
+%The keynote was given by Patricia Lago and entitled ``The legacy of unsustainable software''.
+%Sustainability is broadly associated with natural ecologic systems.
 When we translate the notion of sustainability to software solutions, however, we often confuse {\em impact in a certain point in time} with {\em balanced and durable effects}.
@@
 Lago's keynote discussed the challenges (and some related ongoing research) of combining technical and environmental sustainability, providing a complementary angle to the workshop discussions.
+%
+%WSSSPE4 included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
+%two experience papers, and three demos) that addressed a wide range of topics around
+%sustainability for software in science. These talks covered three main areas.
+%The first area was the academic environment, with four papers
+%that address diverse aspects ranging from incentives for quality software to
+%advocating a professional society for research software to roles and degrees for
+%research software engineers~\cite{Heroux:2016ws, GAllen:2016ws, Philippe:2016ws, Gwilliams:2016ws}.
+%The second area was concerned with characteristics and needs of
+%research software and included five papers~\cite{Dubey1:2016ws, ChueHong:2016ws, Dubey2:2016ws, Queiroz:2016ws, Childers:2016ws}.
+%The third  area focused on elucidating lessons learned
+%from or visions for use cases of scientific software and communities working
+%with and\slash or on a software package, with three papers~\cite{Ganguly:2016ws, Shende:2016ws, Sallai:2016ws}.
+%
+%In addition, 19 lightning talks were presented at WSSSPE4~\cite{Sufi:2016ws, Gesing:2016ws, Ram:2016ws, Loffler:2016ws, Katz:2016ws, Idaszak:2016ws, Hwang:2016ws, Goble:2016ws, Druskat:2016ws, Contrastin:2016ws, AAllen:2016ws, ChueHong:2016wsb, vanHage:2016ws, GAllen:2016wsb, Seidel:2016ws, Emsley:2016ws, Dongarra:2016ws, Bauer:2016ws, Aldabjan:2016ws}.
+%
+%The final WSSSPE4 presentation was a panel of five experts with different perspectives on ``Best Practices for Scientific Software.''
+%These were: Alice Allen, Editor, Astrophysics Source Code Library, who brought an understanding of the difficulties of organizing a community and curating their software;
+%Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
+%Patricia Lago from Vrije Universiteit Amsterdam, who brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
+%Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Papers and demos} \label{sec:papers}
+\section{Speed blogs} \label{sec:speed_blogs}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\note{Sandra Gesing will write this}
+\todo{Alexander}
 
-WSSSPE4 included the presentation of 12 10-minute talks (based on four idea papers, three position papers,
-two experience papers, and three demos) that addressed a wide range of topics around
-sustainability for software in science. These talks covered three main areas.
-%
-% The following four papers belong to the first area: the academic environment.
-%For talks with multiple authors, an
-% \textsuperscript{\textasteriskcentered} indicates the author who presented the talk.
-%Four papers %,  
-% \begin{itemize}
-% \item Michael Heroux: \emph{Idea paper: Sustainable \& Productive:
-% Improving Incentives for Quality Software}
-%~\cite{Heroux:2016ws}, 
-% \item Gabrielle Allen: \emph{Idea Paper: Establishing a Professional
-% Society for Research Software}
-%~\cite{GAllen:2016ws}, 
-% \item Olivier Philippe\textsuperscript{\textasteriskcentered}, Simon Hettrick, and Neil Chue Hong:
-% \emph{Experience Paper: Preliminary analysis of a survey of UK Research
-% Software Engineers}
-%~\cite{Philippe:2016ws}, and 
-% \item Christopher Gwilliams: \emph{Demo: Using industrial engagement to
-% create and develop research ties within academia}
-%~\cite{Gwilliams:2016ws} 
-%belong to 
-The first area was the academic environment, with four papers
-that address diverse aspects ranging from incentives for quality software to
-advocating a professional society for research software to roles and degrees for
-research software engineers~\cite{Heroux:2016ws, GAllen:2016ws, Philippe:2016ws, Gwilliams:2016ws}.
-%\end{itemize}
-The second area was concerned with characteristics and needs of
-research software and included five papers~\cite{Dubey1:2016ws, ChueHong:2016ws, Dubey2:2016ws, Queiroz:2016ws, Childers:2016ws}.
-% \begin{itemize}
-% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Katherine Riley: \emph{Experience Paper: Software
-% Engineering and Community Codes Track in ATPESC}
-% \item Neil Chue Hong: \emph{Position Paper: Why do we need to compare
-% research software, and how should we do it?}
-% \item Anshu Dubey\textsuperscript{\textasteriskcentered} and Lois Curfman McInnes: \emph{Idea Paper: The Lifecycle
-% of Software for Scientific Simulation}
-% \item Francisco Queiroz\textsuperscript{\textasteriskcentered} and Rejane Spitz: \emph{Position Paper:
-% Collaborative Gamification Design for Scientific Software}
-% \item Bruce Childers\textsuperscript{\textasteriskcentered}, Jack Davidson, Wayne Graves, Bernard Rous, and
-% David Wilkinson: \emph{Position Paper: Active Curation of Artifacts is Changing
-% the Way Digital Libraries will Operate}
-% \end{itemize}
-%\noindent 
-%Three contributions, ~\cite{Ganguly:2016ws}, ~\cite{Shende:2016ws},
-%and ~\cite{Sallai:2016ws} , focused on 
-The third  area focused on elucidating lessons learned
-from or visions for use cases of scientific software and communities working
-with and\slash or on a software package, with three papers~\cite{Ganguly:2016ws, Shende:2016ws, Sallai:2016ws}.
+\note{explain the process of speed blogging}
 
-In addition, 19 lightning talks were presented at WSSSPE4~\cite{Sufi:2016ws, Gesing:2016ws, Ram:2016ws, Loffler:2016ws, Katz:2016ws, Idaszak:2016ws, Hwang:2016ws, Goble:2016ws, Druskat:2016ws, Contrastin:2016ws, AAllen:2016ws, ChueHong:2016wsb, vanHage:2016ws, GAllen:2016wsb, Seidel:2016ws, Emsley:2016ws, Dongarra:2016ws, Bauer:2016ws, Aldabjan:2016ws}.
-
-% \begin{itemize}
-% \item Debashis Ganguly, William C.\ Garrison III, David Wilkinson,
-% Bruce R.\ Childers, Adam J.\ Lee, and Daniel Mosse\textsuperscript{\textasteriskcentered}: \emph{Demo: Composing,
-% Reproducing, and Sharing Simulations}
-% \item Sameer Shende\textsuperscript{\textasteriskcentered} and Allen Malony: \emph{Demo: Using TAU for
-% Performance Evaluation of Scientific Software}
-% \item Janos Sallai\textsuperscript{\textasteriskcentered}, Christopher Iacovella, Christoph Klein, and Tengyu Ma:
-% \emph{Idea Paper: Development of a Software Framework for Formalizing Forcefield
-% Atom-Typing for Molecular Simulation}
-% \end{itemize}
+\note{summarize the speed blogs themselves, with citations/pointers to them on the SSI page}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Lightning talks} \label{sec:lightning}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{comment}
-\note{
-\href{http://wssspe.researchcomputing.org.uk/wssspe4/agenda/}{Slides.}}
-\end{comment}
-
-%
-%For talks with multiple authors, an \textsuperscript{\textasteriskcentered} again indicates the author who presented the talk.
-
-% \begin{itemize}%[itemsep=1ex]
-%     \item Shoaib Sufi: \emph{The Software Sustainability Institute Fellowship
-%     programme; supporting the social side of research software}
-
-%     \item Sandra Gesing\textsuperscript{\textasteriskcentered}, Maytal Dahan, Linda B.~Hayden, Katherine Lawrence,
-%     Suresh Marru, Marlon Pierce, Nancy Wilkins-Diehr, and Michael Zentner:
-%     \emph{The Science Gateways Community Institute - Supporting Communities to
-%     Achieve Sustainability for Their Science Gateways}
-
-%     \item Karthik Ram\textsuperscript{\textasteriskcentered}, Noam Ross,
-%     and Scott Chamberlain: \emph{A model for
-%     peer review and onboarding research software}
-%     \item Frank L\"{o}ffler and Steven Brandt\textsuperscript{\textasteriskcentered}:
-%     \emph{A vision of computing in 10+ years}
-%     \item Daniel S.~Katz, Kyle E.~Niemeyer\textsuperscript{\textasteriskcentered},
-%     Arfon M.~Smith, and the FORCE11 Software Citation Working Group:
-%     \emph{Software Citation: Process, Principles, and
-%       Implementation}
-
-%     \item Ray Idaszak\textsuperscript{\textasteriskcentered},
-%     David G.~Tarboton, Hong Yi, Michael Stealey,
-%     Pabitra Dash, Alva Couch, Daniel P.~Ames, Jeffery S.~Horsburgh, Tony Castronova,
-%     Jon Goodall, Mohamed Morsy, Venkatesh Merwade, Mauriel Ramirez, Tian Gan,
-%     Drew (Zhiyu) Li, Jeff Sadler, Shawn Crawley, Zhaokun Xue, Lan Zhao, Carol Song,
-%     Christina Bandaragoda: \emph{HydroShare - A Case Study in Software Engineering
-%     Best Practices and Culture Change for Developing Sustainable Community
-%     Software}
-
-%     \item Lorraine Hwang\textsuperscript{\textasteriskcentered},
-%     Wolfgang Bangerth, Timo Heister, and Louise Kellogg:
-%     \emph{ASPECT: Hackathons as an Example of Sustaining an Open Source
-%     Community}
-
-%     \item Carole Goble: \emph{A Simple Profiling Framework for Software
-%     User-Producer Reciprocity Review}
-%     \item Stephan Druskat: \emph{A proposal for the measurement and
-%     documentation of research software sustainability in interactive metadata
-%     repositories}
-
-%     \item Mistral Contrastin, Matthew Danish,
-%     Dominic Orchard\textsuperscript{\textasteriskcentered}, and
-%     Andrew Rice: \emph{Supporting software sustainability with
-%     lightweight specifications}
-
-%     \item Alice Allen, Cecilia Aragon, Christoph Becker, Jeffrey Carver,
-%     Andrei Chi\c{s}, Benoit Combemale, Mike Croucher, Kevin Crowston, Daniel Garijo,
-%     Ashish Gehani, Carole Goble\textsuperscript{\textasteriskcentered},
-%     Robert Haines, Robert Hirschfeld, James Howison,
-%     Kathryn Huff, Caroline Jay, Daniel S.~Katz, Claude Kirchner, Kateryna Kuksenok,
-%     Ralf L\"{a}mmel, Oscar Nierstrasz, Matt Turk, Rob van Nieuwpoort, Matthew Vaughn,
-%     and Jurgen Vinju: \emph{``I solemnly pledge'': A Manifesto for Personal
-%     Responsibility in the Engineering of Academic
-%     Software}
-
-%     \item Neil Chue Hong\textsuperscript{\textasteriskcentered}:
-%     \textit{Making it easier to understand research software
-%       impact}
-
-%     \item Willem Robert van Hage\textsuperscript{\textasteriskcentered},
-%     Jason Maassen and Rob van Nieuwpoort: \textit{Software Impact Measurement at the
-%     Netherlands eScience Center}
-
-%     \item Gabrielle Allen, Emily Chen\textsuperscript{\textasteriskcentered},
-%     Ray Idaszak, and Daniel S.\ Katz: \textit{Report on Software Metrics for
-%     Research Software}
-
-%     \item Eric L.~Seidel\textsuperscript{\textasteriskcentered}
-%     and Gabrielle Allen: \textit{Bringing Techniques from Software Engineering
-%     into Scientific Software}
-
-%     \item Iain Emsley\textsuperscript{\textasteriskcentered} and David
-%     De Roure: \textit{Sustaining the social: Connecting the lives of Drupal
-%     community}
-
-%     \item Jack Dongarra, Sven Hammarling, Nicholas J.\ Higham, Samuel
-%     D.\ Relton\textsuperscript{\textasteriskcentered}, Pedro Valero-Lara,
-%     and Mawussi Zounon: \textit{Creating a Standardised Set of Batched
-%     BLAS Routines}
-
-%     \item Lukas Breitwieser\textsuperscript{\textasteriskcentered},
-%     Roman Bauer, Alberto Di Meglio, Leonard Johard,
-%     Marcus Kaiser, Marco Manca, Manuel Mazzara, Fons Rademakers,
-%     Max Talanov: \textit{The BioDynaMo Project}
-
-%     \item Aseel Aldabjan, Robert Haines, and
-%     Caroline Jay\textsuperscript{\textasteriskcentered}: \textit{How should we
-%     measure the relationship between code quality and software
-%     sustainability?}
-
-% \end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Best practices for scientific software panel discussion} \label{sec:panel}
+\section{Analyzing the speed blogs} \label{sec:speed_blog_analysis}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%\note{from Simon}
+\todo{Caroline and Rob}
 
-The final WSSSPE4 presentation was a panel %To investigate best practices for scientific software we brought together a panel
-of five experts with different perspectives on ``Best Practices for Scientific Software.''
-These were: Alice Allen, Editor, Astrophysics Source Code Library, who brought an understanding of the difficulties of organizing a community and curating their software;
-Mike Croucher from the University of Sheffield and Rob Haines from the University of Manchester who are both Research Software Engineers with decades of experience of writing code for researchers;
-Patricia Lago from Vrije Universiteit Amsterdam, who brought a fresh perspective on software sustainability from the point of view of its impact on society and business; and
-Karthik Ram from the Berkeley Institute for Data Science, who brought his perspective on the practicalities of using scientific software to conduct his research as a data scientist.
+\note{explain the method?}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Working groups} \label{sec:WGs}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\note{show the date?}
 
-After most of the lightning talks and other presentations, WSSSPE used three areas of a large room to let
-attendees use sticky flip charts and sticky notes on the walls to suggest a vision on any aspect of the work, a gap or challenge, or a project idea.
-Next, attendees organically formed groups around the flip charts, and 12 groups emerged.
-Summaries of each group's activities follow; for more details
-about the group discussions, see \cite{WSSSPE4-report}.
+\note{show a figure?}
 
-\subsection{Verifying best practices \& metrics for sustainable research software}
-\label{sec:best-practices-sustainable}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\note{explain the results?}
 
-%\note{Ray to write this}
+\katznote{here's some stuff I wrote previously - see if you want to keep any of it:}
 
-Many open source projects for research software document their best practices and metrics that
-contribute to the sustainability of the software as well as to define their research software project's success.
-%
-This group will take the outputs of the WSSSPE efforts to identify best practices for creation of sustainable software for science/academia, and also the outputs of WSSSPE efforts to identify metrics for sustainable software for science/academia and cross-reference these with current open source research software that successfully uses modern software engineering.
-This will allow the group to identify gaps on both sides.
-Anyone can join this group or obtain more information about it by sending an email to the verifying best practices \& metrics for sustainable research software group: <wssspe4-verify-best-practices@googlegroups.com>.
+However, to frame the context of these reports, we first examine the overall space, which can be viewed in many ways.
+The high-level topics in the call for participation in the previous meeting, WSSSPE4, were: development and community, professionalization, training, credit, software publishing, software discoverability and reuse, and reproducibility and testing.
+The set of topics initially proposed in WSSSPE5.1 were: social factors, technical choices, funding and careers, management and guidance, citation and credit, training, reproducible research, software quality, legacy code, and containers and packaging.
+WSSSPE5.1 also mentioned that there seemed to be sufficient writing already on citation and credit, and containers and packaging, so while these topics that fit under the WSSSPE umbrella, they were not suggested as topics for speed blogs.
+In addition, the lead author of this paper has given talks to summarize this space previously, based on topics that emerged from the previous WSSSPE meetings before WSSSPE4.
+Those topics are: software engineering; portability; training and education; incentives, citation/credit models, and metrics; intellectual property; publication and peer review; software communities and sociology; sustainability and funding models; career paths; software dissemination, catalogs, search, and review; multi-disciplinary science; and reproducibility.
+All three of these ways of looking at the space have value; none are wrong, none are right, all are useful.
+They each try to look at the problems we have.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Software Sustainability Alliance}
-\label{sec:alliance}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Neil and Jean wrote this}
-
-The Software Sustainability Alliance working group aims to establish an alliance between organizations interested in improving the sustainability of research software.
-Such an alliance would ease the collaboration between member organizations to improve the sharing of expertise, resources and best practices.
-It is currently seeking feedback on potential member organizations, as well as the aims and scope of this alliance.
-Currently, point-to-point collaboration exists between organizations, but this inadvertently results in competition or redundancy within the sustainable software community.
-An alliance of software sustainability organizations would ease inter-organization collaboration, the promotion of software sustainability as well as would improve the pooling of competencies and the sharing of expertise.
-If interested, one can visit \url{http://softwaresustainability.org/} or email Neil Chue Hong (<N.ChueHong@software.ac.uk>) and Jean Salac (<jeansalac@virginia.edu>)
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Scientific Software Prototyping Infrastructure (S2PI)}
-\label{sec:prototyping}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%from Santiago
-There is a productivity bottleneck---yet to be solved---in HPC from the human
-perspective, first identified in the 1980s~\cite{barstow1982automatic}. Of all
-domain-specific scientists, only a percentage of those use simulations due to
-perceptions mostly related to the difficulty of using and developing those tools.
-XSEDE and other resources have become a gateway for successfully increasing the
-basis of HPC scientific users by facilitating their access to infrastructure and
-tools~\cite{towns2014xsede}, but the development challenge remains.
-%
-The group's objective is development of a prototype tool that allows domain scientists to generate
-simulation prototypes through a simple, yet expressive declarative
-strongly-typed programming language close to equational expressions. The result
-of that specification is both an executable artifact as well as a specification
-for scientific programmers to later flesh out completely and adapt to particular
-infrastructures.
-
-To join this project, one can visit the GitHub repository at
-\url{https://github.com/nunezco2/S2PI} or join the WSSSPE Slack channel
-\texttt{\#wg-sci-soft-proto}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{CodeMeta}
-\label{sec:CodeMeta}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Alice to write this}
-
-Research software is often not shared; that which is shared may not have much metadata associated with it, and that which does exist often does not travel further than the website on which the software resides.
-The CodeMeta project\footnote{\url{http://codemeta.github.io/}} wants to incentivize software developers to release their software, encourage the development of metadata for it, enable credit assignment and citation of research software, increase its discoverability, more easily track dependencies, and enable reuse of software metadata, all goals that WSSSPE attendees have great interest in supporting.
-%
-While at WSSSPE4, the group greatly expanded the CodeMeta project README file to include a description of the project geared to those with little or no prior knowledge of the project, a list of contributors, information on how one can get involved, a brief project history and who is managing the project, and links to additional information.
-
-Though a Google group mailing list has been established for the working group, the easiest way to engage with the CodeMeta project is through its Github repository\footnote{\url{https://github.com/codemeta/codemeta}}.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{White paper on developing sustainable software}
-\label{sec:best-practices-developing}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Sandra to write this}
-
-Many diverse aspects and dimensions (such as economic, technical, environmental, and social) of developing sustainable software can be investigated,
-This group aims to write white papers
-that will focus on scientific environments and their implications, targeted at developers
-and project managers of scientific software. Given the complexity of this field, it is important to select
-a subset of sustainability aspects for the white papers. The idea is to create a series of papers instead of trying
-to tackle all important topics in one paper.
-%
-While there are already a few papers available on best practices and sustainability of scientific software in general, the group's goal is to create a series of papers that lead to consensus in the community, tackle many diverse aspects, and stay up-to-date with new trends.
-%
-The GitHub repository for the white paper can be found at \url{https://github.com/WSSSPE/WG-Best-Practices}.
-For more information and requests, join the \texttt{\#wg-best-practices} channel at WSSSPE's Slack team.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Social science for scientific software}
-\label{sec:social}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{from Stuart}
-
-%Introduction to group here, including the overall objective of work in this area.
-
-There is more and more academic research being done on topics related to software sustainability, including work on software engineering practices and management of open source projects.
-However, academic research in general is often siloed for many reasons, and work on topics relevant to software sustainability is no exception.
-%
-This working group is motivated by the goal of building better connections between academic researchers who are studying topics in or relating to software sustainability with practitioners, managers, and administrators who are working in the area of software sustainability.
-%
-The working group plans to briefly survey existing literature to get a better sense of the academic research landscape, facilitate some initial dialog between academic researchers and practitioners at WSSSPE4, then identify needed actions that would be mutually beneficial to researchers and practitioners.
-This plan resulted in the creation of the following SMART steps listed in the next section.
-
-To join the group, one can join its mailing list at \url{https://groups.google.com/forum/#!forum/researchsoftwarestudies}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Software best practices for undergraduates}
-\label{sec:best-practices-undergrads}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Jonah Miller to write this}
-
-This working group was motivated by the perceived prevalence of
-so-called ``hidden code'' in scientific communities: code written by
-individual researchers in an unsustainable way that is never shared
-with the larger community. 
-%
-The participants hypothesized that the best strategy is to catch
-researchers while they are still in training and teach them software
-best-practices. Therefore, this working group's goal
-is to develop courses on software best practices aimed at
-undergraduate students studying domain science. The program might be
-similar to a Software Carpentry or Data Carpentry workshop but
-focused for domain scientists.
-
-The development of a successful curriculum relies on the expertise of
-software engineers to describe the best practices, domain experts to
-describe model problems and workflow, and instructors to formulate
-the pedagogy. Ideally these people will be brought together for a
-short workshop or hackathon with the goal of drafting the
-curriculum. 
-Anyone interested in contributing should get in touch via one of the
-following methods:
-
-\begin{itemize}
-\item Join the Google Group/mailing list
-  \cite{WSSSPEUndergradGoogleGroup}.
-\item Join the WSSSPE Slack (\S\ref{sec:slack}) and the
-  \texttt{\#wg-undergraduatecourse} channel.
-\item Ask to join the WSSSPE4-undergraduate-course organization
-  \cite{WSSSPEUndergradGithub}.
-\end{itemize}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Meaningful metrics for sustainable software}
-\label{sec:metrics}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{content from Emily}
-
-This group aims to increase the visibility of the quality of scientific software, facilitate the reusability of scientific software, and promote the best software practices by standardizing metrics via interviews with scientific software developers.
-This working group believes improving the current software metrics system will increase software sustainability.
-Currently, there are inefficiencies regarding software duplication, sustainability, and selection, as well as others, within the scientific software community.
-In order to address these inefficiencies, the group aims to create a goal-oriented method to collecting productive metrics by focusing on the developer side of software.
-
-For more information, one can contact Emily Chen at <echen35@illinois.edu>.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Coordinating access to continuous integration (CI) for research software}
-\label{sec:access}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{%Mark Abraham to write this?  Needs confirmation or another volunteer -
-%Dan starting this for now - may poll other members of group to make progress}
-
-Each developer of software with uncommon needs (hardware, software, libraries, data sets), non-public code, or tests that exceed time limits for free plans must acquire, setup, and maintain their own continuous integration systems because their needs make them ineligible for popular free services such as Travis CI.
-For example, software groups that develop BIOUNO, CI4SI, or GROMACS have done this.
-
-This group is interested in reducing the burden of different projects having to build and maintain their own continuous integration systems (when publicly available CI are not a fit), by coordinating and sharing this burden across multiple projects.
-%
-
-However, there are a lot of open issues such as how can the group learn from existing similar work without reinventing the wheel.
-Some possible goals include to acquire additional hardware such as GPUs, Xeon PHI, FPGAs and add/share them to, e.g., Debian's testing infrastructure or to extend Debian's scope to include published but not mature software.
-
-For more information, see \url{https://groups.google.com/forum/#!forum/continuous-integration-for-research-software}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Software engineering processes tailored for research software}
-\label{sec:soft-eng}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Anshu to write this}
-
-This working group is concerned with identifying processes that are not adequately
-covered by general software engineering, starting with verification and testing.
-Computational science and engineering applications have many moving
-parts that need to interoperate with one another. The accuracy and
-reliability of results produced by scientific software depends not
-only on the individual components behaving correctly, but also on the
-validity of their interactions.
-A big challenge is that those who understand
-the importance of regular testing do not often find much help from
-software engineering literature. There is a significant gap between
-the testing gospel and its applicability to computational science. This
-gap leads to frustration and abandonment of the good with the bad.
-
-This working group will (1) conduct a literature survey to gauge the extent
-of awareness of the issue in general, (2) generate content useful for
-the community where needed, and (3) curate the collected and added
-content for the use of the community.
-
-Readers interested in getting more information should use the working group'
-channel in the WSSSPE Slack (\S\ref{sec:slack}), called
-\texttt{\#wg-testing-in-science}.
-Additionally, a git
-repository (\url{https://github.com/WSSSPE/WG-Best-Practices.git}) exists for contributing content and reference to, and
-curation of the existing literature on this topic.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Open research index}
-\label{sec:open-research-index}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{Dan to write this}
-
-The aim of this group is to investigate the building of an index of research products in an open sustainable manner.
-Its goal is not to eliminate commercial products, but to build on what is there and provide data and services that are missing.
-%
-The Open Research Index should take in all research products (papers, software, datasets, workflows, etc.) from their publishers and recorders (journals, societies, domain repositories, government [open access] repositories, preprint servers, general repositories [e.g., figshare, zenodo]) and other services (CrossRef, ORCID).
-Each product should list authors and citations and allow people to search the resulting network.
-Users should also be able to interact with their own record and edit it, like Google Scholar allows.
-
-The working group's plans are relatively simple to express, though quite complex to undertake. A leader needs to be identified.
-With a leader and feasible plan, the group would then obtain resources, and then build the index.
-%
-Because Google and others provide some similar services today, though these services (and the underlying data) could be removed at any time and the community cannot build new services, these companies would ideally be involved in the activities of the group.
-
-Alternatively or additionally, the group could build a mailing list for us and discuss further, depending on how receptive others are to this idea.
-At WSSSPE4, there seemed to be enough interest to do this, so the group set up a Slack channel within the WSSSPE team (\S\ref{sec:slack}), called \texttt{\#wg-open-research-idx} for its members and any future members.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Letters of evaluation for computational scientists}
-\label{sec:letters}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-Scientists working on scientific software are often located in
-disciplinary departments, depending on whether their software
-originates from the mathematical, physical, chemical, or other
-disciplines. As a consequence, they are frequently outside the core
-areas of their science, and their contributions are typically to both
-the research activity their software enables, as well as on
-algorithm and implementation aspects. This presents issues when
-letters of evaluation for hiring, tenure, and promotion do not specifically
-cover how this is relevant to the discipline.\footnote{The extensive use of such letters, and the problems that are
-  associated with it, may be an issue specific to the United States.}
-
-Consequently, letter writers neet to be aware of how their letters will be read by
-committees. It also requires that committees be aware that such
-letters often look different and may provide a different perspective
-of how a candidate's achievements should be assessed.
-
-%
-There are essentially two important strategies that build on each
-other: (i) raising awareness of the problem beyond just those who are
-affected by it, and (ii) providing letter writers, letter readers, and
-evaluating committees with guidance on what criteria are relevant in
-assessing scientific software authors.
-%
-Ultimately, building a community large enough
-to affect change is important; contact Wolfgang Bangerth at
-<bangerth@colostate.edu> if you are interested in helping.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Slack team and channels}\label{sec:slack}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-As part of the discussion at WSSSPE4, we created a Slack team for WSSSPE~\cite{WSSSPESlack}.
-To join it, go to \url{https://wssspe.signup.team/}.
-A number of the working groups also created channels within the team; see the other subsections in \S\ref{sec:WGs} for the names of these channels.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Attendee survey \label{sec:survey}}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\note{from Lorraine}
-
-At the end of WSSSPE4, all remaining participants completed an on-line survey.
-%
-%All participants present on the last day of the meeting, 14 September 2016, were asked to complete the online survey administered through Qualtrics\footnote{\url{https://www.qualtrics.com}.}.
-%44 responses were recorded, with one respondent indicating that they took the survey twice.
-%The survey URL was also later distributed to attendees by email and advertised on twitter for attendees who were not there on the last day to have a chance to respond, and one additional response was recorded after the meeting.
-%Hence, the total number of unique individuals responding to the survey is likely 44.
-%The survey contained nine multiple choice and open-text answer questions.
-%Text responses have been alphabetized to preserve anonymity.
-%The survey had a 100\% completion rate.
-%
-In general, the respondents were highly satisfied with the meeting and interested in continuing and participating in WSSSPE activities.
-The mix of topics and types of interactions---talks, panels and discussion---was well balanced with several indicating a desire for time for questions and answers following talks as well for discussions in general.
-Many respondents indicated that they will remain engaged in WSSSPE working group initiatives.
-There was weaker demand for a professional organization encompassing WSSSPE interests, with over half interested in joining and a large percentage willing to consider the idea.
-Respondents were grateful for the opportunity to network, explore new collaborations and for travel support to the conference.
-
-The survey results indicate that future WSSSPE conferences should consider the balance of the attendees.
-WSSSPE4 had many first time and early career participants who would have benefited from a review of basic concepts and terminology including what it means for software to be sustainable and the roles of research software engineers.
-Additional emphasis and topics to explore would be the inclusion of more case studies, focus on the decision making process in developing and using software, deeper dives into selected topics, tutorials, software training within and outside of STEM, lowering the barriers to implementing best practices in software development, and progress towards executing WSSSPE's vision.
+Another choice is to look at this at a different level: that of actors.
+We can say that we are concerned about software, people, organizations, and the field itself.
+And we could center our view through the lens of one or more of these actors.
 
@@
 
-In WSSSPE4, we heard about a number of interesting projects and ideas, and used those
-ideas to create and form working groups, intended to start at the workshop and then
-continue afterwards, to address challenges that arose from the workshop ideas. This workshop
-has reinforced the lesson from WSSSPE3 that it is relatively easy to get motivated people
-to attend a meeting and productively spend their time there both doing work and planning
-more work, but it is very hard to get that additional work after the meeting to take place.
-The main problem seems to be one of time.  Once the attendees have agree to spend
-their time at the workshop, they put their energy into doing so productively, but they have not
-really committed themselves to anything more than this, so their energy and effort trails
-off, as all of their other commitments (particularly, those they are funded to do as part of
-their jobs) come back to the fore. Without a process to resolve this concern, the utility of
-further multi-day WSSSPE workshops is unclear, at least to start new activities.  On a more
-positive note, WSSSPE4 was successful as a gathering place to discuss scientific
-software sustainability, and for groups that are already in place or that can be composed
-of related funded activities to meet.  Future WSSSPE meetings will likely combine these
-functions: a meeting place for like-minded individuals, a place to share experiences and
-lessons learned, and a place for both existing groups and sets of related projects to meet
-in person.  Other parts of the WSSSPE infrastructure (the email list and the Slack) will
-be used both for general discussions, as well as for groups that choose to focus
-on specific problems.
-
-After WSSSPE4, we also have learned enough to summarize the areas of active work
-in sustainable scientific software, based on the WSSSPE4 mission and vision statements, the WSSSPE4 call for participation (based on topics in previous WSSSPE meetings), and the working groups formed in WSSSPE4.
-Following the example of the recently funded US project,
-``Conceptualizing a US Research Software Sustainability Institute (URSSI)\footnote{\url{https://www.nsf.gov/awardsearch/showAward?AWD_ID=1743188}},'' the high-level
-version of these areas is: functioning of 1) the individual and team, 2) the research software, and 3) the research field itself, with six specific topics, as shown in Table~\ref{tab:topics}
-
-\begin{table}[h]
-\centering
-\caption{Scientific Software Sustainability Research Topics and Areas}
-\label{tab:topics}
-\begin{tabular}{|l|c|c|c|}
- \hline
-Topics \textbackslash\ Areas & Individual and team &  Research software & Research field  \\
- \hline
-Principles and Best Practices & \checkmark & \checkmark & \checkmark  \\
- \hline
-Careers & \checkmark &  &   \\
- \hline
-Learning & \checkmark &  &   \\
- \hline
-Credit & \checkmark &  & \checkmark  \\
- \hline
-Organization and community & \checkmark &  &  \checkmark \\
- \hline
-Common Infrastructure &  & \checkmark &   \\
- \hline
-\end{tabular}
-\end{table}
-
-%\todo{discuss Twitter activity}
+\todo{anyone?}
 
@@
 
-% NSF grant for WSSSPE4
-This material is based upon work supported by the National Science Foundation (ACI-1648293, ACI-1547611),
-by the Gordon and Betty Moore Foundation (GBMF\#5620),
-and by the Alfred P.~Sloan Foundation (G-2016-7214).
-
-%\todo{feel free to add stuff here}
-
-%\newpage
-%\appendix
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section{Organization}  \label{sec:orgcom}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%\todo{Do we want email addresses here?}
-%
-%WSSSPE4 organizers were:
-%
-%{\scriptsize
-%\begin{longtable}{lll}
-%\input{orgcom}
-%\end{longtable}
-%}
-%
-%
-%The WSSSPE4 program committee comprised:
-%
-%{\scriptsize
-%\begin{longtable}{lll}
-%\input{program_committee}
-%\end{longtable}
-%}
+\todo{feel free to add stuff here}
 

### 2018-03-30 WSSSPE5.1_report/WSSSPE5.1_paper.tex
Kill whitespace errors in a separate commit.
I have a rather strict editor setup.
Add Rob's full details.

@@
 Caroline $^{(2)}$,
-Rob $^{(3)}$,
+Robert~Haines$^{(3)}$,
 Alexander $^{(4)}$,
@@
 %
-\thanks{{}$^{(3)}$ department/school, city, country; email; ORCID: orcid}
+\thanks{{}$^{(3)}$ University of Manchester, Manchester, UK; robert.haines@manchester.ac.uk; ORCID: 0000-0002-9538-7919}
 %

### 2018-07-21 WSSSPE5.1_report/main.tex
updating with version published on arXiv

\documentclass[11pt,letterpaper]{article}
%\usepackage{f1000_styles}

\usepackage[numbers]{natbib}

\pdfoutput=1

%\usepackage{amsmath,amssymb,amsmath}

\usepackage{color}
\usepackage[table]{xcolor}
\usepackage{dcolumn}
\usepackage{caption}
\usepackage{graphicx}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[font=small]{caption}
\usepackage{varwidth}

\usepackage[activate={true,nocompatibility},final,tracking=true,kerning=true,spacing=true,factor=1100,stretch=10,shrink=10]{microtype}
\microtypecontext{spacing=nonfrench}
\usepackage{xspace}

\usepackage{rotating}
\usepackage{caption,subcaption}
\usepackage{float}
\usepackage{psfrag}
\usepackage{tabularx}
\usepackage[hyphens]{url}
\usepackage{wrapfig}
\usepackage{longtable}
\usepackage{verbatim}
\usepackage{array,booktabs,multicol,multirow}

%better printing of numbers
\usepackage[english]{babel}
\usepackage{textcomp}
\usepackage{csquotes}

%\usepackage{cite}

% The following three lines are used for displaying footnote in tables.
\usepackage{footnote}
\makesavenoteenv{tabular}
\makesavenoteenv{table}

\usepackage{placeins}

\usepackage{makecell}

\usepackage{enumitem}
\setlist{leftmargin=7mm}

\makeatletter
\newcommand\footnoteref[1]{\protected@xdef\@thefnmark{\ref{#1}}\@footnotemark}
\makeatother

%\setcounter{secnumdepth}{3}
%\setcounter{tocdepth}{3}

\usepackage[bookmarks, bookmarksopen, bookmarksnumbered, colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
\usepackage[all]{hypcap}
\urlstyle{rm}

\definecolor{orange}{rgb}{1.0,0.3,0.0}
\definecolor{violet}{rgb}{0.75,0,1}
\definecolor{darkgreen}{rgb}{0,0.6,0}
\definecolor{cyan}{rgb}{0.2,0.7,0.7}
\definecolor{blueish}{rgb}{0.2,0.2,0.8}
\definecolor{darkblue}{rgb}{0.1,0.1,0.9}
\definecolor{lightgray}{gray}{0.9}

%\newcommand{\todo}[1]{{\color{blue}$\blacksquare$~\textsf{[TODO: #1]}}}
\newcommand{\note}[1]{ {\textcolor{blueish}        { ***Note: #1 }}}
\newcommand{\katznote}[1]{ {\textcolor{magenta}    { ***Dan: #1 }}}
\newcommand{\sdnote}[1]{ {\textcolor{darkgreen}    { ***Stephan: #1 }}}
\newcommand{\rhnote}[1]{ {\textcolor{cyan}         { ***Rob: #1 }}}
\newcommand{\LXnote}[1]{  {\textcolor{orange}      { ***Alex: #1 }}}

\newcommand*\rot{\rotatebox{90}}

% Don't use tt font for urls
\urlstyle{rm}

%\usepackage[letterpaper, margin=1in]{geometry}
% You can use a baselinestretch of down to 0.9
%\renewcommand{\baselinestretch}{0.96}

\sloppypar

\begin{document}

\title{The State of Sustainable Research Software:
Results from the Workshop on Sustainable Software for Science: Practice
and Experiences (WSSSPE5.1)}
\author{Daniel S.~Katz\thanks{National Center for Supercomputing Applications (NCSA) \& Department of Computer Science  \& Department of Electrical and Computer Engineering  \& School of Information Sciences (iSchool), University of Illinois, Urbana--Champaign, IL, USA; d.katz@ieee.org; ORCID: 0000-0001-5934-7525},
Stephan Druskat\thanks{Department of German Studies and Linguistics, Humboldt-Universität zu Berlin, Berlin, Germany; stephan.druskat@hu-berlin.de; ORCID: 0000-0003-4925-7248},
Robert Haines\thanks{Research IT, University of Manchester, Manchester, UK; robert.haines@manchester.ac.uk; ORCID: 0000-0002-9538-7919},\\
Caroline Jay\thanks{School of Computer Science, University of Manchester, Manchester, UK; caroline.jay@manchester.ac.uk; ORCID: 0000-0002-6080-1382},
Alexander Struck\thanks{Cluster of Excellence Image Knowledge Gestaltung at Humboldt-Universität zu Berlin; Germany; Alexander.Struck@hu-berlin.de; ORCID: 0000-0002-1173-9228}}

\maketitle
%\thispagestyle{fancy}

\begin{abstract}
This article summarizes motivations, organization, and activities of the Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) held in Manchester, UK in September 2017. The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit. This article discusses the Code of Conduct, idea papers, position papers, experience papers, demos, and lightning talks presented during the workshop. The main part of the article discusses the speed-blogging groups that formed during the meeting, along with the outputs of those sessions. % Finally, it discusses a survey of the workshop attendees.

% It also summarizes a set of lightning
% talks in which speakers highlighted to-the-point lessons and challenges
% pertaining to sustaining scientific software.
% The final and main contribution of the report is a summary of the
% discussions, future steps, and future organization for a set of self-organized
% working groups on topics including developing pathways to funding scientific
% software; constructing useful common metrics for crediting software
% stakeholders; identifying principles for sustainable software engineering
% design; reaching out to research software organizations around the world; and
% building communities for software sustainability. For each group, we include a
% point of contact and a landing page that can be used by those who want to join
% that group's future activities. The main challenge left by the workshop is to
% see if the groups will execute these activities that they have scheduled, and
% how the WSSSPE community can encourage this to happen.

\end{abstract}

\section*{Keywords}

sustainable software, research software engineering

\clearpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction} \label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{Dan}

In September 2017, 37 people interested in sustainable research software came together at the Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE5.1, \href{http://wssspe.researchcomputing.org.uk/wssspe5-1/}{wssspe.researchcomputing.org.uk/wssspe5-1/}) meeting in Manchester, UK. This immediately preceded the Second Research Software Engineers (RSE) Conference, so that RSE attendees could also attend WSSSPE5.1.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Mission and vision}\label{sec:mission}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{should we take any of this into the intro, or drop it all?}

%\sdnote{I'm for keeping it for ``newcomers''}

%{\bf Mission.}
 WSSSPE is an international community-driven organization that promotes sustainable research software by addressing challenges related to the full lifecycle of research software through shared learning and community action.
%
% {\bf Vision.}
 It envisions a world where research software is accessible, robust, sustained, and recognized as a scholarly research product critical to the advancement of knowledge, learning, and discovery.

% {\bf Focus areas.}
 WSSSPE promotes sustainable research software by positively impacting:
 \begin{itemize}
 \item Principles and Best Practices. Promoting best practices in sustainable software
 \item Careers. Developing and supporting career paths in research software development and engineering
 \item Learning. Engaging in activities to promote peer learning and interaction
 \item Credit. Ensuring recognition of research software as an intellectual contribution equal to other research products
 \end{itemize}

% \textbf{Definitions:}
WSSSPE defines \emph{Sustainable software} as software that has the capacity to endure such that it will continue to
 be available in the future, on new platforms, meeting new needs.
 The \emph{research software lifecycle} includes:
 acquiring and assembling resources (including funding and people) into teams and communities,
 developing software,
 using software,
 recognizing contributions to and of software,
 and
 maintaining software.

Previous WSSSPE events\footnote{The first WSSSPE workshop was named ``Working towards Sustainable Software for Science: Practice and Experiences,'' which remains the meaning of the WSSSPE group, but the workshops after that were named ``Workshop on Sustainable Software for Science: Practice and Experiences.'' Together these reflect that WSSSPE is both a community and a set of workshops.},
all but the last of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1,WSSSPE2-pre-report,WSSSPE2}, two half-day workshops (\href{http://wssspe.researchcomputing.org.uk/wssspe1-1/}{wssspe.researchcomputing.org.uk/wssspe1-1/}, \href{http://wssspe.researchcomputing.org.uk/wssspe2-1/}{wssspe.researchcomputing.org.uk/wssspe2-1/}) held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, a one-and-a-half-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}, and a two-and-a-half-day workshop that immediately preceded the First Research Software Engineers (RSE) Conference in Manchester, England~\cite{WSSSPE4-report}.

These workshops dedicated a significant portion of their time to group discussions about problems and potential solutions in the sustainable research software space.  WSSSPE5.1 used the speed blog methodology to generate eight reports on different views of the space. The blogs were published by the UK Software Sustainability Institute (SSI) on its website.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Call for participation and response} \label{sec:preworkshop}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\sdnote{Have left out most of the reasoning behind the call, and call details. Could be long enough to warrrant a heading, but can likewise be merged with intro.}

Submissions to WSSSPE5.1 were made to two tracks, each of which accepted lightning talks (up to two pages) and papers (up to four pages).

\begin{itemize}
\item Track 1 -- ``The state of the art in sustainable research software'' - called for submissions that report and evaluate projects which created or currently create foundations, ecologies and tools for sustainable research software.
\item Track 2 -- ``Towards a sustainable future for research software'' - called for submissions that identify new or revisit existing implementable proposals for making research software sustainable.
\end{itemize}

Submissions comprised six papers (four in Track 1; two in Track 2) and eight lightning talks (six in Track 1; two in  Track 2).
Of these, four papers (three in Track 1; one in Track 2) and seven lightning talks (five in Track 1; two in Track 2) were accepted.
Similar to previous years, most submissions were accepted in order to include as many relevant inputs as possible, and to encourage their authors to share and implement their input.

All papers and lightning talks have been published as a figshare collection~\cite{WSSSPE5_1_proceedings_2017}. Slides for the given talks have been published on the WSSSPE5.1 website (\href{http://wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}{wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Code of Conduct}\label{sec:CoC}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{do we want to keep some discussion of this here?  maybe fold in into the intro?}

WSSSPE5.1 included a Code of Conduct (CoC, \href{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}{wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}) as guidance for the community of scientists that WSSSPE
supports, including the workshop and their personal and online interactions (e.g., on
Twitter, in email lists, in Slack). This CoC is based on the
FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
CoC~\cite{Code4Lib:CoC}.
In introducing the CoC, we asked participants to agree to the following main guidelines:
\begin{quote}
    WSSSPE events are community events intended for networking and collaboration
    as well as learning. We value the participation of every member of the
    community and want all attendees to have an enjoyable and fulfilling
    experience. Accordingly, all attendees are expected to show respect and
    courtesy to other attendees throughout the event and in interactions online
    associated with the event.

    The WSSSPE event organizers are dedicated to providing a harassment-free
    experience for everyone, regardless of gender, gender identity and
    expression, age, sexual orientation, disability, physical appearance,
    body size, race, ethnicity, religion (or lack thereof), technology choices,
    or other group status.

    To make clear what is expected, everyone taking part in WSSSPE events and
    discussions---speakers, helpers, organizers, and participants---is required
    to conform to the Code of Conduct.

 \end{quote}

The CoC was discussed at the beginning of WSSSPE5.1, including presenting the CoC subcommittee
and a general email address for reporting concerns or incidents, or
asking questions.

\subsection{Outline}
The remainder of this paper includes %the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); 
a set of presentations from the accepted papers and lighting talks (\S\ref{sec:presentations}); outputs from the speed-blogging groups (\S\ref{sec:speed_blogs};  thematic analysis of the resulting blogs (\S\ref{sec:speed_blog_analysis}), and discussion about attendee opinions of the meeting (\S\ref{sec:survey}), before concluding (\S\ref{sec:conclusions}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Presentations}\label{sec:presentations}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\sdnote{Kept this abstract to present overview of topics covered rather than re-hashing presentations. IMHO blogs are perhaps the more important outcome. Let me know if I should also give synopses or a broad overview of topics.} \katznote{thanks for doing this - I generally like it}

WSSSPE5.1 included the presentation of four 12-minute talks based on accepted papers \cite{nangia_track_2017,haupt_track_2017,queiroz_track_2017,mulholland_track_2017},
and seven 6-minute lightning talks \cite{silva_track_2017,struck_track_2017,washbrook_track_2017,dasler_track_2017,alhozaimy_track_2017,maassen_track_2017,druskat_track_2017}.

\begin{figure}[h!]
  \centering
  \includegraphics[width=\textwidth]{sustainability-schematic-weighted.pdf}
  \caption{Research software sustainability space schematic; topic weights based on WSSSPE5.1 presentations.}
  \label{fig:schematic}
\end{figure}

Figure~\ref{fig:schematic} shows a version of a schematic of the research 
software sustainability space 
as introduced by Daniel S.~Katz \cite{katz_research_2018}.
In this version, both node outlines and edge labels are weighted to show the distribution of
their references from the workshop presentations \cite{druskat_activity}.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{sustainability-actions-plot.pdf}
  \caption{Distribution of action coverage over presentations.}
  \label{fig:actions}
\end{figure}

Figure~\ref{fig:actions} shows the distribution of actions over all presentations, where an action is a labeled edge from a node in the sustainability schematic (Figure~\ref{fig:schematic}) to another node. The source node represents the actor, and the target node the actee of the respective action.
%
The weights and action distribution show a clear overall primary focus of presentations on software development (\textit{People develop Software}). More generally, the people involved in research software take prominent focus as well as software engineering principles, and research software itself. However, across all presentations most actors, actees, and actions within the space have been the topic of a presentation as actor or actee, with the exception of hardware and underlying software. Future workshops could take care to address this topic specifically in their calls for submissions in order to close gaps in research, discussion and progress.

%This coverage of topics continues in the speed blogs, which were partly based on presentations, as discussed in a presentation by Daniel S.~Katz \cite{katz_research_2018}.

Table~\ref{tab:topics} shows the distribution of combined actor and actee reference from the
research software sustainability space over workshop
presentations.

\begin{table}[h!]
\centering
\footnotesize
\begin{tabular}{ccccccccc}
%	\hline
    	\rot{\shortstack[l]{Presentation}} & \rot{Communities} & \rot{\shortstack[l]{Funding \\ organizations}} & \rot{\shortstack[l]{Hardware \& \\ underlying \\ software}} & \rot{\shortstack[l]{Hiring \\ organizations}} & \rot{People} & \rot{\shortstack[l]{Publishers, \\ repositories, \\ indices}} & \rot{Software} & \rot{\shortstack[l]{Software \\ engineering \\ processes}} \\
        \hline
\cite{nangia_track_2017} &                                 &                                          &                                                    &                                         & •                          &                                                     & •                            &                                                   \\ 
\hline
\cite{haupt_track_2017} & •                               &                                          &                                                    & •                                       & •                          &                                                     & •                            & •                                                 \\ 
\hline
\cite{queiroz_track_2017} &                                 &                                          &                                                    &                                         & •                          &                                                     & •                            & •                                                 \\ 
\hline
\cite{mulholland_track_2017} &                                 &                                          &                                                    & •                                       & •                          &                                                     & •                            & •                                                 \\ 
\hline
\cite{silva_track_2017} &                                 & •                                        &                                                    & •                                       & •                          &                                                     & •                            &                                                   \\ 
\hline
\cite{struck_track_2017} &                                 & •                                        &                                                    & •                                       & •                          &                                                     & •                            & •                                                 \\ 
\hline
\cite{washbrook_track_2017} &                                 &                                          &                                                    & •                                       & •                          &                                                     & •                            & •                                                 \\ 
\hline
\cite{dasler_track_2017} &                                 &                                          &                                                    & •                                       &                            & •                                                   & •                            &                                                   \\ 
\hline
\cite{alhozaimy_track_2017} &                                 &                                          &                                                    &                                         & •                          & •                                                   & •                            & •                                                 \\ 
\hline
\cite{maassen_track_2017} & •                               &                                          &                                                    & •                                       & •                          & •                                                   & •                            & •                                                 \\ 
\hline
\cite{druskat_track_2017} & •                               & •                                        &                                                    & •                                       & •                          & •                                                   & •                            & •                     \\
\hline
\end{tabular}
\caption{Distribution of topics from the research software sustainability
space over workshop presentations for actors and actees (combined).}
\label{tab:topics}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Speed blogs} \label{sec:speed_blogs}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{Alexander}

%\note{explain the process of speed blogging}\\
% https://www.software.ac.uk/term/speed-blogging
The Software Sustainability Institute (SSI) discusses speed blogging at \href{https://www.software.ac.uk/term/speed-blogging}{www.software.ac.uk/term/speed-blogging}. They provide helpful approaches and pointers to some valuable resources on academic writing.
%
The goal of speed blogging is to preserve as much content and context as possible from working groups, and to publish results in an easily digestible form. The output is usually published as blog posts, but other formats, such as an audio podcast or an informative table, could also serve the same purpose. During the process it is important to reserve at least of half the available working group time for creating the desired output format. The recommended best practice is to work collaboratively on such a document and review assigned chapters.
The SSI says,
\begin{quotation}
Speed blogs can be completed and be publication ready during the allotted time at an event or they can be near complete [at] the event and then tidied up and made ready for publication soon after the event. Thus they are time bound, time sensitive and are a fixed task that does not represent an ongoing commitment.
\end{quotation}

%\note{summarize the speed blogs themselves, with citations/pointers to them on the SSI page}\\
The rest of this section summarizes the WSSSPE5.1 blog posts, which are accessible at \href{http://wssspe.researchcomputing.org.uk/wssspe5-1/}{wssspe.researchcomputing.org.uk/wssspe5-1/}. Indented text indicated quotations from the blogs.

\begin{itemize}
\item The Research Software Project Manager (\href{https://www.software.ac.uk/blog/2017-12-04-research-software-project-manager}{www.software.ac.uk/blog/2017-12-04-research-software-project-manager})
\begin{quote}
For many, the role of research software project manager (RSPM) may be an accidental calling. The career path for this role isn’t well-established, and research software development in academia may itself be something of a haphazard, nigh-accidental byproduct of conducting domain research. Individuals approaching this role may have little to no wider industry experience, instead approaching the project manager role from research or research software engineering.
\end{quote}
The authors present the challenges for the RSPM role, noting that is inherently different from similar positions in the industry. The focus of work is on enabling research as an iterative process, prone to many changes along the way. While many academics excel at winning grants and pushing the scientific frontiers, only a few are trained in software project management. And those who are may focus on technical aspects, discarding parts of the overall picture. This raises many questions, as documented in the blog post, and offers answers to a few of them, include strengthening the role of the RSPM and considering positions dedicated to project management. Such a delegation of tasks would smooth the process of reporting to administration and free up necessary resources for actual research.

\item Looking for software use in research (\href{https://www.software.ac.uk/blog/2017-12-05-looking-software-use-research}
{www.software.ac.uk/blog/2017-12-05-looking-software-use-research}
\begin{quote}
Nowadays, software is used in most research. But how the software is created, used, and what it depends on are not well understood questions. The importance of such knowledge varies based on the motivation of the reader. On one side, we could be interested in the impact of the software, how many times it has been used and by who. This type of analysis could come, for example, from funding bodies and organisations to reward the creation of something and help its sustainability, from institutions who hire people behind that software, or from the software authors to get an understanding of the needs of their users or simply to get credit for their work. Another motivation may be trying to understand the research being carried out with a particular software or set of tools either for purely academic purposes (e.g., by historians and scholars of science) or with a commercial perspective (such as by intellectual property teams from universities for the monetisation of the software). %lx% should be shortened
\end{quote}
The blog post reports on methods for finding software. Researchers usually consult their favorite search engines and programming language package archives. In the future, Current Research Information Systems (CRIS) or repositories maintained by funders may hold information about software developed during research. Software as a research output is increasingly registered with DOIs \cite{Fenner2018DOI} but identifying software in written text still has its challenges \cite{Li2017R}. Repositories such as \href{https://www.bioconductor.org/}{Bioconductor} offer some domain-specific overview of available tools. General purpose repositories, such as \href{http://github.com}{GitHub.com}, track usage as in forks on their platforms, whereas services like \href{http://dpesy.org}{Depsy.org} and \href{https://libraries.io/}{libraries.io} provide aggregated software usage information from many sources. %lx% put links in there
Software, considered by some to be a subset of research data, could also be identified resolving PIDs like DOIs, minted by services like \href{http://datacite.org}{DataCite.org} that point to published data and software. 
%Efforts describing relevant research software repositories in a meta catalog have been undertaken by, e.g., \href{http://re3data.org}{re3data.org}, \katznote{re3data is really a catalog of repositories that meet some guidelines, so I'm not sure if makes sense to discuss it here, or maybe I'm misunderstanding} but currently fail to provide consistent indexing and comprehensive search functionality. \LXnote{It's not a discussion but rather mention of a failed effort to make software findable} \katznote{Are you saying that an intent of the re3data.org project was to make software findable, or that after re3data.org started for other purposes, it had an effort to make software findable?} \LXnote{by definition of the funding agency (DFG), research software is a subset of research data, therefore the intent of re3data.org is to make research data AND software findable ... by cataloging repositories that hold data and/or software. The resulting platform has a lot of room for improvement but is mentioned here as a different way of pointing researchers to software. If that is too confusing for the reader, let me know and I will remove the sentence.} \katznote{I see what you are saying, but it's very indirect and confusing to me without a longer explanation, such as you have provided in the comments - maybe remove this?}

\item Towards Reproducibility in Research Software (\href{https://www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software}{www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software})
\begin{quote}
Ensuring reproducibility of research has been identified as one of the challenges in scientific research. While reproducibility of results is a concern in all fields of science, the emphasis of this group is in the area of computer software reuse and the reproduction of results. The availability of complete descriptions, ideally including program source code, documentation and archives of all necessary components and input datasets would be a major step to resolving research reproducibility concerns.
\end{quote}
A short review of relevant services and platforms is provided by the authors. It includes concepts from software engineering, e.g., version control, being introduced in programmes such as \href{https://software-carpentry.org}{Software Carpentry}.  %lx% link
Platforms and tools, such as \href{https://www.overleaf.com}{Overleaf} or \href{http://jupyter.org}{Jupyter notebooks}, ease collaborative review of research and reproducibility of code respectively. While the uptake of such best practices is slow and the lack of documentation sometimes hinders reproducibility, we may see more of them in the future, especially if incentives are set right.

\item Why research software engineers should have permanent contracts (\href{https://www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts}{www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts})

\noindent At present, only a few research institutions employ research software engineers and make their resources available to the whole organization. This blog post discusses some success stories motivating long term contracts for RSEs. In general ``better software [leads to] better research'' (\href{https://software.ac.uk/}{SSI}).
\begin{quote}
Higher quality output is on every university's wish list, as it leads to a potential increase in QR funding (\href{http://www.hefce.ac.uk/rsrch/funding/mainstream/}{www.hefce.ac.uk/rsrch/funding/mainstream/}), while reducing the reputational risk associated with substandard research practices.
Reproducibility is notoriously difficult to achieve and RSEs are an essential part of enabling this [\ldots] leading to higher citations and greater research impact.
\end{quote}
Skilled RSE are sought after and organizations get what they pay for. The ``stability and the potential for some form of career progression'' that permanent contracts offer should be a consideration when hiring such staff. And the costs do not necessarily reduce the university's baseline funds as research projects are encouraged to ``cost in'' consulting by centrally pooled RSEs. Long-term benefits for organizations are the spreading of best practices in software development and an institutional memory provided by the central RSE group who have worked in many projects. 
\begin{quote}
An RSE team that is permanently employed can be truly agile. Recruiting an expert for a short period of time in an academic institution is virtually impossible. As long as we rely on fixed-term contracts for RSEs, a lot of important work will fail to be done, and funds will not be spent as effectively as they could be.
\end{quote}

\item A standard format for CITATION files (\href{https://www.software.ac.uk/blog/2017-12-12-standard-format-citation-files}{www.software.ac.uk/blog/2017-12-12-standard-format-citation-files})
\begin{quote}
The citation of research software has a number of purposes, most importantly attribution and credit, but also the provision of impact metrics for funding proposals, job interviews, etc. Stringent software citation practices [\ldots] therefore include the citation of a software version itself, rather than a paper about the software. Direct software citation also enables reproducibility of research results as the exact version can be retrieved from the citation. 
\end{quote}
While a diverse range of citation recommendations exist for software projects, many projects still proceed without a proper citation. In order to unify the various existing  procedures, the authors proposed the inclusion of a CITATION file in a software repository, holding standardized information, e.g., ORCID to enable linking authors with their respective ID from \href{https://orcid.org}{orcid.org}. It shall be easy to read and write in order to enable fast and straightforward creation of citation information. For this purpose, the site \href{https://research-software.org/citation/}{research-software.org/citation/} was created and is actively maintained to cover all aspects of research software citation. Widespread use of the Citation File Format (CFF) may enable transitive credit information provision and better impact metrics next to reusability by other actors.

\item Overcoming barriers to adopting software best practices in research (\href{https://www.software.ac.uk/blog/2017-12-07-overcoming-barriers-adopting-software-best-practices-research}{www.software.ac.uk/blog/2017-12-07-overcoming-barriers-adopting-software-best-practices-research})

\noindent Research careers create a wide spectrum of skill levels (compared to corporate environments). Training a group of researchers on a new topic is challenging, given the lack of a peer group at some research frontiers or missing institutional support for training and the significant upfront cost involved in learning e.\,g., best development practices. Researchers who mostly work on code in isolation rarely have an opportunity to have code reviews (with a notable exception as described in \href{https://www.software.ac.uk/blog/2018-05-18-code-review-academia}{www.software.ac.uk/blog/2018-05-18-code-review-academia}), need to learn on their own, and are often not recognized for writing code. Paper publication in a ``high-impact peer-reviewed journal'' still pushes careers. A way to overcome barriers is to introduce, e.g., ``industry standards'' in an manner adapted to research environments. The blog post discusses how the ``SCRUM'' approach may not work, but ``agile'' or a ``maturity model'' approach may better fit the current research practices.
\begin{quote}
As a general concept: start small and then go as far as necessary. Reaching for the perfect software development approach is intimidating and overwhelming, and it is not the task of a researcher nor necessary for most research projects. A maturity model can help researchers identify where they are and where they should be [\ldots]. Restricting the use of tech jargon to a minimum and offering explanations where necessary can help, too.
\end{quote}

\item Encouraging good software development practice in research teams (\href{https://www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams}{www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams}

\noindent ``Training in good software development skills is vital for the uptake and maintenance of good practice in a research community.'' This requires that resources, tools, and sometimes management support the availability of, and that the trainee be motivated. The latter could be achieved when the ``connection between good practice and increased publication rate and quality, as well as funding availability'' are demonstrated. A ``consistent approach to software development [\ldots] can benefit collaboration and the longevity of projects.'' Success story from the DLR (\href{https://www.dlr.de}{www.dlr.de}) and the EMBL (\href{https://www.embl.de/}{www.embl.de}) are reported in the blog post, which goes into detail on how to build a community and keep it alive. The authors propose the concept of ``inner source,'' where open source principles are utilized collaboratively inside an organization to develop ``common scientific frameworks.'' GitLab is praised as a beneficial tool that eases adoption of good practices. The authors recognize that rollout of training, tools, and services should happen iteratively and is always highly dependent on the particular environment. Therefore they are interested to learn from others and encourage feedback via e-mail.

\item Overcoming Entry Barriers to Motivate Better Practice in Research Software Engineering (\href{https://www.software.ac.uk/blog/2017-12-14-overcoming-entry-barriers-motivate-better-practice-research-software-engineering}{www.software.ac.uk/blog/2017-12-14-overcoming-entry-barriers-motivate-better-practice-research-software-engineering})
\begin{quote}
What can be termed as ``coding'' is a subset of wider software engineering practices such as version control, continuous integration and good software design. Coding is prevalent in academia but practices that allow sustainable software to be produced are frequently overlooked.  Motivating the uptake of the approaches, methods and tools, and highlighting the benefit they deliver, by engaging with researchers who develop software is the first step in spreading best practice in our community.
\end{quote}
The authors point out the benefit of using online systems such as GitLab to reduce entry barriers and motivating the use of, e.g., version control and continuous integration (CI). Other software engineering principles such as pair-programming and code review are also encouraged early in (graduate) students training in order to demonstrate the benefits for future use and bridging gaps between disciplines. If these research software management practices become a requirement in grants application and reporting, widespread deployment is inevitable. Increasing recognition of software as a valid research output could provide further motivation, along with better reproducibility and reduced duplication of effort.

\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analyzing the speed blogs} \label{sec:speed_blog_analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\sdnote{Alexander, all, would it be interesting to look into topic distribution for the speed blogs as well? Dan has had a slide on it at another talk (referenced in section above), but could possibly be done in more detail.}
%\katznote{I was thinking that might be part of what Caroline and Rob would do in this section.}

%\todo{Caroline and Rob}

%\note{explain the results?}

The speed blog topics were determined using an unconference format---where participants chose what to cover as a group---and therefore provided a snapshot of issues that were particularly timely and relevant to the WSSSPE community. Participants were free to choose which speed blogging group to join, depending on their own particular interests and goals. In this section, we treat the speed blogs as qualitative data, which we systematically analyse to determine the prevalence of particular themes.

\subsection{Method} 
We used a hybrid thematic/framework analytic approach, where we used the schematic of the research software sustainability space shown in Figure~\ref{fig:schematic} as a starting point for our analysis, and then refined this as we familiarized ourselves with the data. The schematic nodes formed the categories: \emph{funders} (`funding organizations' in the schematic); \emph{employers} (`hiring organizations'); \emph{publishers, repositories, indices}; \emph{research software} (`software'); \emph{software engineering processes}; \emph{communities}. The edges provided the categories: \emph{reward \& recognition} (`recognize, reward'); \emph{training}; \emph{standardization} (`standardize'); \emph{reproducibility} (`reproduce'). Based on a bottom-up analysis of the data, we broke the schematic `people' node down further into \emph{users}, \emph{research software engineers} and \emph{researchers}, and added a further category of \emph{software infrastructure}.

Authors Jay and Haines coded the blogposts independently, recording for each blogpost whether or not the theme was present. This process resulted in agreement of 79\%. Disagreements were then resolved via discussion. The original data set (blogposts and individual and joint coding scores) are available for further analysis~\cite{data:speed-blog-analysis}.

\subsection{Results}
All of the eight blogs mentioned research software, and researchers (i.e., domain specialists rather than RSEs). The blogs also all mentioned reward and recognition, either for software itself, or for people writing software. The prevalence of all of the categories across the blog posts can be seen in Figure~\ref{fig:speedblogs}. During the discussion of the speed blog topics it was decided that there seemed to be sufficient writing already on citation and credit, so while this topic fit under the WSSSPE umbrella, it was not suggested as a topic for speed blogs. The fact that it was mentioned in all the blogs indicates that it is still an important theme for the community.

\begin{figure}[h!]
  \includegraphics[width=\textwidth]{speed_blogs_categories_distribution.png}
  \caption{The number of speed blogs mentioning each of the themes.}
  \label{fig:speedblogs}
\end{figure}

%The high-level topics in the call for participation in the previous meeting, WSSSPE4, were: development and community, professionalization, training, credit, software publishing, software discoverability and reuse, and reproducibility and testing. The set of topics initially proposed in WSSSPE5.1 were: social factors, technical choices, funding and careers, management and guidance, citation and credit, training, reproducible research, software quality, legacy code, and containers and packaging. WSSSPE5.1 also mentioned that there seemed to be sufficient writing already on citation and credit, and containers and packaging, so while these topics that fit under the WSSSPE umbrella, they were not suggested as topics for speed blogs. In addition, the lead author of this paper has given talks to summarize this space previously, based on topics that emerged from the previous WSSSPE meetings before WSSSPE4. Those topics are: software engineering; portability; training and education; incentives, citation/credit models, and metrics; intellectual property; publication and peer review; software communities and sociology; sustainability and funding models; career paths; software dissemination, catalogs, search, and review; multi-disciplinary science; and reproducibility. All three of these ways of looking at the space have value; none are wrong, none are right, all are useful. They each try to look at the problems we have.

%Another choice is to look at this at a different level: that of actors. We can say that we are concerned about software, people, organizations, and the field itself. And we could center our view through the lens of one or more of these actors.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Attendee survey} \label{sec:survey}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

At the end of WSSSPE5.1, participants were asked to complete an online survey, and 35 of the 37 did \cite{druskat_survey}.
%
The answers indicate that respondents were very satisfied with the workshop in general. Similarly to WSSSPE4\cite{WSSSPE4-report}, the mixture of topics and speakers as well as the proportions of longer presentations and lightning talks was deemed well-balanced, while some responses indicate that more time for discussion would be helpful, which may include an extension of the workshop length. The majority of respondents (73\%) found the discussion/speed blogging session more valuable than the presentation session, which seems to support the hands-on approach of the workshop format. In general, respondents found the purpose of the workshop sufficiently clear, and generally agree that its purpose was fulfilled.

Similar to the results of the WSSSPE4 attendee survey \cite{WSSSPE4-report}, the results of the current survey indicate that future WSSSPE events should include a review of basic concepts and terminology: 11\% of respondents somewhat disagree that the concept of ``software sustainability'' as presented at the workshop is well-defined. Similarly, the overall vision and progress of WSSSPE as well as the organization of concrete actions between WSSSPE workshops should be defined and communicated more strongly. Additional topics to explore and include in future events were case studies (also mentioned in the WSSSPE4 survey results); the organization of Research Software Engineering teams; funding for software fellowships; measurement and composition of communities, as well as tools, services and user engagement practices; impact metrics for sustainable software; and guidance on making the case for sustainability.

Some participants found the length of the lightning talk (six minutes) to be odd. They suggested that their length should be reduced to more typical talking times for this type of presentation (two-to-three minutes), and that the time saved as a result should be used for keynotes at the beginning of the workshop. Alternative suggestions were that all presentations should be of equal length (12-15 minutes). Other suggestions included making the workshop available virtually or provide recordings, and repeatedly, to make it longer than one day.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions} \label{sec:conclusions}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\todo{more needed?}
In summary, the community recognizes the need for improving software engineering practices and takes action. A set of initial success stories have been reported from a handful of organizations. Recognition for work, including via citations, remains a topic of interest.

Based on the presentations and working groups at WSSSPE, we have presented a set of topics and mapped the WSSSPE5.1 presentations and speed blogs onto these topics. The presentations and speed blogs cover most of the topics and do not have subjects that are not in the list, indicating this that topics (or themes, a more fine-grained mapping of the space as discussed in Section~\ref{sec:speed_blog_analysis}) are a good representation of the space.

%The discussions and hands-on sessions were the most valued parts of the workshop by the participants, who requested that the overall duration of the workshop should be lengthened. \LXnote{thinks that this could work as conclusions.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section*{Acknowledgments} \label{sec:acks}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\katznote{feel free to add stuff here}

\newpage
\bibliographystyle{vancouver}

\bibliography{wssspe.bib}
\end{document}

### 2018-07-27 WSSSPE5.1_report/wssspe51jors.tex
adding JORS submission for WSSSPE5.1 paper

--- 2018-07-21 WSSSPE5.1_report/main.tex
@@
 \title{The State of Sustainable Research Software:
-Results from the Workshop on Sustainable Software for Science: Practice
+Learning from the Workshop on Sustainable Software for Science: Practice
 and Experiences (WSSSPE5.1)}
@@
 \begin{abstract}
-This article summarizes motivations, organization, and activities of the Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) held in Manchester, UK in September 2017.
+This paper uses the accepted submissions from the Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE5.1) held in Manchester, UK in September 2017 and the speed blogs written during the event to examine the state of research software.
+It presents a schematic of the space, then examines coverage in terms of topics, actors, actees, and themes by both the submissions and the blogs.
+
+%.
 The WSSSPE series promotes sustainable research software by positively impacting principles and best practices, careers, learning, and credit.
@@
 
-\section*{Keywords}
-
-sustainable software, research software engineering
+%\section*{Keywords}
+%sustainable software, research software engineering
 
@@
 In September 2017, 37 people interested in sustainable research software came together at the Working towards Sustainable Software for Science: Practice and Experiences (WSSSPE5.1, \href{http://wssspe.researchcomputing.org.uk/wssspe5-1/}{wssspe.researchcomputing.org.uk/wssspe5-1/}) meeting in Manchester, UK.
-This immediately preceded the Second Research Software Engineers (RSE) Conference, so that RSE attendees could also attend WSSSPE5.1.
+WSSSPE5.1 immediately preceded the Second Research Software Engineers (RSE) Conference, so that RSE attendees could also attend WSSSPE5.1.
 
@@
 
-Previous WSSSPE events\footnote{The first WSSSPE workshop was named ``Working towards Sustainable Software for Science: Practice and Experiences,'' which remains the meaning of the WSSSPE group, but the workshops after that were named ``Workshop on Sustainable Software for Science: Practice and Experiences.'' Together these reflect that WSSSPE is both a community and a set of workshops.},
-all but the last of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1,WSSSPE2-pre-report,WSSSPE2}, two half-day workshops (\href{http://wssspe.researchcomputing.org.uk/wssspe1-1/}{wssspe.researchcomputing.org.uk/wssspe1-1/}, \href{http://wssspe.researchcomputing.org.uk/wssspe2-1/}{wssspe.researchcomputing.org.uk/wssspe2-1/}) held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, a one-and-a-half-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}, and a two-and-a-half-day workshop that immediately preceded the First Research Software Engineers (RSE) Conference in Manchester, England~\cite{WSSSPE4-report}.
-
-These workshops dedicated a significant portion of their time to group discussions about problems and potential solutions in the sustainable research software space.
-WSSSPE5.1 used the speed blog methodology to generate eight reports on different views of the space.
+Six previous WSSSPE events\footnote{The first WSSSPE workshop was named ``Working towards Sustainable Software for Science: Practice and Experiences,'' which remains the meaning of the WSSSPE group, but the workshops after that were named ``Workshop on Sustainable Software for Science: Practice and Experiences.'' Together these reflect that WSSSPE is both a community and a set of workshops.}~\cite{WSSSPE1-pre-report,WSSSPE1,WSSSPE2-pre-report,WSSSPE2}
+%,
+%all but the last of which were in the US, were two general, presentation-focused, one-day workshops held with the SC13 and SC14 conferences~\cite{WSSSPE1-pre-report,WSSSPE1,WSSSPE2-pre-report,WSSSPE2}, two half-day workshops (\href{http://wssspe.researchcomputing.org.uk/wssspe1-1/}{wssspe.researchcomputing.org.uk/wssspe1-1/}, \href{http://wssspe.researchcomputing.org.uk/wssspe2-1/}{wssspe.researchcomputing.org.uk/wssspe2-1/}) held with SciPy 2014 and 2015 that contained presentations about specific sustainable Python software packages, a one-and-a-half-day workshop that included teams that self-assembled and discussed focused software sustainability topics~\cite{WSSSPE3}, and a two-and-a-half-day workshop that immediately preceded the First Research Software Engineers (RSE) Conference in Manchester, England~\cite{WSSSPE4-report}.
+%
+%These workshops dedicated a significant portion of their time to 
+included 
+group discussions about problems and potential solutions in the sustainable research software space.
+WSSSPE5.1 used the speed blog methodology to generate eight reports on different views of this space.
 The blogs were published by the UK Software Sustainability Institute (SSI) on its website.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Call for participation and response} \label{sec:preworkshop}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%Submissions to WSSSPE5.1 were made to two tracks, either as lightning talks (up to two pages) or papers (up to four pages).
+Track 1 -- ``The state of the art in sustainable research software'' - called for submissions that report and evaluate past and ongoing projects; Track 2 -- ``Towards a sustainable future for research software'' - called for submissions that identify new proposals for making research software sustainable.
 
-%\sdnote{Have left out most of the reasoning behind the call, and call details. Could be long enough to warrrant a heading, but can likewise be merged with intro.}
-
-Submissions to WSSSPE5.1 were made to two tracks, each of which accepted lightning talks (up to two pages) and papers (up to four pages).
-
-\begin{itemize}
-\item Track 1 -- ``The state of the art in sustainable research software'' - called for submissions that report and evaluate projects which created or currently create foundations, ecologies and tools for sustainable research software.
-\item Track 2 -- ``Towards a sustainable future for research software'' - called for submissions that identify new or revisit existing implementable proposals for making research software sustainable.
-\end{itemize}
-
-Submissions comprised six papers (four in Track 1; two in Track 2) and eight lightning talks (six in Track 1; two in  Track 2).
-Of these, four papers (three in Track 1; one in Track 2) and seven lightning talks (five in Track 1; two in Track 2) were accepted.
-Similar to previous years, most submissions were accepted in order to include as many relevant inputs as possible, and to encourage their authors to share and implement their input.
-
-All papers and lightning talks have been published as a figshare collection~\cite{WSSSPE5_1_proceedings_2017}.
-Slides for the given talks have been published on the WSSSPE5.1 website (\href{http://wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}{wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}).
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\subsection{Code of Conduct}\label{sec:CoC}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%\todo{do we want to keep some discussion of this here?  maybe fold in into the intro?}
-
-WSSSPE5.1 included a Code of Conduct (CoC, \href{http://wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}{wssspe.researchcomputing.org.uk/wssspe4/code-of-conduct/}) as guidance for the community of scientists that WSSSPE
-supports, including the workshop and their personal and online interactions (e.g., on
-Twitter, in email lists, in Slack). This CoC is based on the
-FORCE11 conference CoC~\cite{FORCE11:CoC}, which is in turn based on the Code4Lib
-CoC~\cite{Code4Lib:CoC}.
-In introducing the CoC, we asked participants to agree to the following main guidelines:
-\begin{quote}
-    WSSSPE events are community events intended for networking and collaboration
-    as well as learning. We value the participation of every member of the
-    community and want all attendees to have an enjoyable and fulfilling
-    experience. Accordingly, all attendees are expected to show respect and
-    courtesy to other attendees throughout the event and in interactions online
-    associated with the event.
-
-    The WSSSPE event organizers are dedicated to providing a harassment-free
-    experience for everyone, regardless of gender, gender identity and
-    expression, age, sexual orientation, disability, physical appearance,
-    body size, race, ethnicity, religion (or lack thereof), technology choices,
-    or other group status.
-
-    To make clear what is expected, everyone taking part in WSSSPE events and
-    discussions---speakers, helpers, organizers, and participants---is required
-    to conform to the Code of Conduct.
-
- \end{quote}
-
-The CoC was discussed at the beginning of WSSSPE5.1, including presenting the CoC subcommittee
-and a general email address for reporting concerns or incidents, or
-asking questions.
-
-\subsection{Outline}
 The remainder of this paper includes %the call for papers (\S\ref{sec:preworkshop}); the WSSSPE Code of Conduct (\S\ref{sec:CoC}) and mission and vision (\S\ref{sec:mission}); 
-a set of presentations from the accepted papers and lighting talks (\S\ref{sec:presentations}); outputs from the speed-blogging groups (\S\ref{sec:speed_blogs};  thematic analysis of the resulting blogs (\S\ref{sec:speed_blog_analysis}), and discussion about attendee opinions of the meeting (\S\ref{sec:survey}), before concluding (\S\ref{sec:conclusions}).
+a set of presentations from the accepted papers and lighting talks (\S\ref{sec:presentations}), outputs from the speed-blogging groups (\S\ref{sec:speed_blogs}), and thematic analysis of the resulting blogs (\S\ref{sec:speed_blog_analysis}),
+%and discussion about attendee opinions of the meeting (\S\ref{sec:survey}),
+before concluding (\S\ref{sec:conclusions}).
 
@@
 
-WSSSPE5.1 included the presentation of four 12-minute talks based on accepted papers \cite{nangia_track_2017,haupt_track_2017,queiroz_track_2017,mulholland_track_2017},
-and seven 6-minute lightning talks \cite{silva_track_2017,struck_track_2017,washbrook_track_2017,dasler_track_2017,alhozaimy_track_2017,maassen_track_2017,druskat_track_2017}.
+Submissions to WSSSPE5.1 comprised six papers %(four in Track 1; two in Track 2) 
+and eight lightning talks. %(six in Track 1; two in  Track 2).
+Of these, four papers \cite{nangia_track_2017,haupt_track_2017,queiroz_track_2017,mulholland_track_2017} %(three in Track 1; one in Track 2) 
+and seven lightning talks \cite{silva_track_2017,struck_track_2017,washbrook_track_2017,dasler_track_2017,alhozaimy_track_2017,maassen_track_2017,druskat_track_2017} %(five in Track 1; two in Track 2)
+were accepted.
+All papers and lightning talks were published as a figshare collection~\cite{WSSSPE5_1_proceedings_2017}.
+Slides for the given talks have been published on the WSSSPE5.1 website (\href{http://wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}{wssspe.researchcomputing.org.uk/wssspe5-1/wssspe5-1-agenda/}).
+
+%WSSSPE5.1 included the presentation of four 12-minute talks based on accepted papers \cite{nangia_track_2017,haupt_track_2017,queiroz_track_2017,mulholland_track_2017}, and seven 6-minute lightning talks \cite{silva_track_2017,struck_track_2017,washbrook_track_2017,dasler_track_2017,alhozaimy_track_2017,maassen_track_2017,druskat_track_2017}.
 
@@
   \centering
-  \includegraphics[width=\textwidth]{sustainability-schematic-weighted.pdf}
+  \includegraphics[width=\textwidth]{sustainability-schematic-weighted-jors.pdf}
   \caption{Research software sustainability space schematic; topic weights based on WSSSPE5.1 presentations.}
@@
 software sustainability space 
-as introduced by Daniel S.~Katz \cite{katz_research_2018}.
+as introduced by Katz \cite{katz_research_2018}.
 In this version, both node outlines and edge labels are weighted to show the distribution of
@@
 \begin{figure}[h!]
-  \includegraphics[width=\textwidth]{sustainability-actions-plot.pdf}
+  \includegraphics[width=\textwidth]{sustainability-actions-plot-resize-jors.pdf}
   \caption{Distribution of action coverage over presentations.}
@@
 The Software Sustainability Institute (SSI) discusses speed blogging at \href{https://www.software.ac.uk/term/speed-blogging}{www.software.ac.uk/term/speed-blogging}.
-They provide helpful approaches and pointers to some valuable resources on academic writing.
+%They provide helpful approaches and pointers to some valuable resources on academic writing.
 %
-The goal of speed blogging is to preserve as much content and context as possible from working groups, and to publish results in an easily digestible form.
-The output is usually published as blog posts, but other formats, such as an audio podcast or an informative table, could also serve the same purpose.
+The goal of speed blogging is to preserve as much content and context as possible from working groups, and to publish results in an easily digestible form, usually blog posts. 
+%The output is usually published as blog posts, but other formats, such as an audio podcast or an informative table, could also serve the same purpose.
 During the process it is important to reserve at least of half the available working group time for creating the desired output format.
 The recommended best practice is to work collaboratively on such a document and review assigned chapters.
-The SSI says,
-\begin{quotation}
-Speed blogs can be completed and be publication ready during the allotted time at an event or they can be near complete [at] the event and then tidied up and made ready for publication soon after the event.
+Half the available working group time at WSSSPE5.1 was allocated to discussion, and the other half to writing the blogs. 
+%The SSI says,
+%\begin{quotation}
+%Speed blogs can be completed and be publication ready during the allotted time at an event or they can be near complete [at] the event and then tidied up and made ready for publication soon after the event.
 Thus they are time bound, time sensitive and are a fixed task that does not represent an ongoing commitment.
-\end{quotation}
-
+%\end{quotation}
+%
+%
 %\note{summarize the speed blogs themselves, with citations/pointers to them on the SSI page}\\
@@
 \begin{itemize}
-\item The Research Software Project Manager (\href{https://www.software.ac.uk/blog/2017-12-04-research-software-project-manager}{www.software.ac.uk/blog/2017-12-04-research-software-project-manager})
+\item The Research Software Project Manager\\(\href{https://www.software.ac.uk/blog/2017-12-04-research-software-project-manager}{www.software.ac.uk/blog/2017-12-04-research-software-project-manager})
 \begin{quote}
@@
 \end{quote}
-The authors present the challenges for the RSPM role, noting that is inherently different from similar positions in the industry.
+The authors present the challenges for the RSPM role, noting that it is inherently different from similar positions in the industry.
 The focus of work is on enabling research as an iterative process, prone to many changes along the way.
-While many academics excel at winning grants and pushing the scientific frontiers, only a few are trained in software project management.
-And those who are may focus on technical aspects, discarding parts of the overall picture.
-This raises many questions, as documented in the blog post, and offers answers to a few of them, include strengthening the role of the RSPM and considering positions dedicated to project management.
+While many academics excel at winning grants and pushing the scientific frontiers, only a few are trained in software project management, and those who are may focus on technical aspects, discarding parts of the overall picture.
+The blogpost raises a number of questions about this situation, and suggests answers to a few of them, including strengthening the role of the RSPM and considering positions dedicated to project management.
 Such a delegation of tasks would smooth the process of reporting to administration and free up necessary resources for actual research.
 
-\item Looking for software use in research (\href{https://www.software.ac.uk/blog/2017-12-05-looking-software-use-research}
-{www.software.ac.uk/blog/2017-12-05-looking-software-use-research}
+\item Looking for software use in research \\(\href{https://www.software.ac.uk/blog/2017-12-05-looking-software-use-research}
+{www.software.ac.uk/blog/2017-12-05-looking-software-use-research})
 \begin{quote}
@@
 The blog post reports on methods for finding software.
-Researchers usually consult their favorite search engines and programming language package archives.
+Today, researchers usually consult search engines and programming language package archives.
 In the future, Current Research Information Systems (CRIS) or repositories maintained by funders may hold information about software developed during research.
 Software as a research output is increasingly registered with DOIs \cite{Fenner2018DOI} but identifying software in written text still has its challenges \cite{Li2017R}.
-Repositories such as \href{https://www.bioconductor.org/}{Bioconductor} offer some domain-specific overview of available tools.
-General purpose repositories, such as \href{http://github.com}{GitHub.com}, track usage as in forks on their platforms, whereas services like \href{http://dpesy.org}{Depsy.org} and \href{https://libraries.io/}{libraries.io} provide aggregated software usage information from many sources.
+Repositories such as \href{https://www.bioconductor.org/}{Bioconductor} offer a domain-specific overview of available tools.
+General purpose repositories, such as \href{http://github.com}{GitHub.com}, track usage as forks on their platforms, whereas services like \href{http://dpesy.org}{Depsy.org} and \href{https://libraries.io/}{libraries.io} provide aggregated software usage information from many sources.
 %lx% put links in there
-Software, considered by some to be a subset of research data, could also be identified resolving PIDs like DOIs, minted by services like \href{http://datacite.org}{DataCite.org} that point to published data and software.
+Software, considered by some to be a subset of research data, could also be identified via DOIs, minted by services like \href{http://datacite.org}{DataCite.org} that point to published data and software.
 %Efforts describing relevant research software repositories in a meta catalog have been undertaken by, e.g., \href{http://re3data.org}{re3data.org}, \katznote{re3data is really a catalog of repositories that meet some guidelines, so I'm not sure if makes sense to discuss it here, or maybe I'm misunderstanding} but currently fail to provide consistent indexing and comprehensive search functionality.
@@
 
-\item Towards Reproducibility in Research Software (\href{https://www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software}{www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software})
+\item Towards Reproducibility in Research Software\\ (\href{https://www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software}{www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software})
 \begin{quote}
@@
 
-\item Why research software engineers should have permanent contracts (\href{https://www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts}{www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts})
+\item Why research software engineers should have permanent contracts \\(\href{https://www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts}{www.software.ac.uk/blog/2017-12-11-why-research-software-engineers-should-have-permanent-contracts})
 
@@
 \end{quote}
-Skilled RSE are sought after and organizations get what they pay for.
+Skilled RSEs are sought after and organizations get what they pay for.
 The ``stability and the potential for some form of career progression'' that permanent contracts offer should be a consideration when hiring such staff.
@@
 
-\item A standard format for CITATION files (\href{https://www.software.ac.uk/blog/2017-12-12-standard-format-citation-files}{www.software.ac.uk/blog/2017-12-12-standard-format-citation-files})
+\item A standard format for CITATION files\\(\href{https://www.software.ac.uk/blog/2017-12-12-standard-format-citation-files}{www.software.ac.uk/blog/2017-12-12-standard-format-citation-files})
 \begin{quote}
@@
 Paper publication in a ``high-impact peer-reviewed journal'' still pushes careers.
-A way to overcome barriers is to introduce, e.g., ``industry standards'' in an manner adapted to research environments.
+A way to overcome barriers is to introduce, e.g., ``industry standards'' in a manner adapted to research environments.
 The blog post discusses how the ``SCRUM'' approach may not work, but ``agile'' or a ``maturity model'' approach may better fit the current research practices.
@@
 \begin{figure}[h!]
-  \includegraphics[width=\textwidth]{speed_blogs_categories_distribution.png}
+  \includegraphics[width=\textwidth]{speed_blogs_categories_distribution-jors.png}
   \caption{The number of speed blogs mentioning each of the themes.}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section{Attendee survey} \label{sec:survey}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-At the end of WSSSPE5.1, participants were asked to complete an online survey, and 35 of the 37 did \cite{druskat_survey}.
-%
-The answers indicate that respondents were very satisfied with the workshop in general.
-Similarly to WSSSPE4\cite{WSSSPE4-report}, the mixture of topics and speakers as well as the proportions of longer presentations and lightning talks was deemed well-balanced, while some responses indicate that more time for discussion would be helpful, which may include an extension of the workshop length.
-The majority of respondents (73\%) found the discussion/speed blogging session more valuable than the presentation session, which seems to support the hands-on approach of the workshop format.
-In general, respondents found the purpose of the workshop sufficiently clear, and generally agree that its purpose was fulfilled.
-
-Similar to the results of the WSSSPE4 attendee survey \cite{WSSSPE4-report}, the results of the current survey indicate that future WSSSPE events should include a review of basic concepts and terminology: 11\% of respondents somewhat disagree that the concept of ``software sustainability'' as presented at the workshop is well-defined.
-Similarly, the overall vision and progress of WSSSPE as well as the organization of concrete actions between WSSSPE workshops should be defined and communicated more strongly.
-Additional topics to explore and include in future events were case studies (also mentioned in the WSSSPE4 survey results); the organization of Research Software Engineering teams; funding for software fellowships; measurement and composition of communities, as well as tools, services and user engagement practices; impact metrics for sustainable software; and guidance on making the case for sustainability.
-
-Some participants found the length of the lightning talk (six minutes) to be odd.
-They suggested that their length should be reduced to more typical talking times for this type of presentation (two-to-three minutes), and that the time saved as a result should be used for keynotes at the beginning of the workshop.
-Alternative suggestions were that all presentations should be of equal length (12-15 minutes).
-Other suggestions included making the workshop available virtually or provide recordings, and repeatedly, to make it longer than one day.
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \section{Conclusions} \label{sec:conclusions}
@@
 Recognition for work, including via citations, remains a topic of interest.
-
-Based on the presentations and working groups at WSSSPE, we have presented a set of topics and mapped the WSSSPE5.1 presentations and speed blogs onto these topics.
-The presentations and speed blogs cover most of the topics and do not have subjects that are not in the list, indicating this that topics (or themes, a more fine-grained mapping of the space as discussed in Section~\ref{sec:speed_blog_analysis}) are a good representation of the space.
+%
+Based on the presentations and working groups at WSSSPE5.1, we have presented a set of topics and mapped the %WSSSPE5.1
+presentations and speed blogs onto these topics.
+The presentations and speed blogs cover most of the topics and do not have subjects that are not in the list, indicating that the topics (or themes, a more fine-grained mapping of the space as discussed in Section~\ref{sec:speed_blog_analysis}) are a good representation of the space.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\section*{Acknowledgments} \label{sec:acks}
+\section*{Acknowledgments} \label{sec:acks}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+S. Druskat would like to acknowledge funding assistance from the Software Sustainability Institute. The Software Sustainability Institute is supported by the EPSRC, BBSRC and ESRC Grant EP/N006410/1.
 
@@
 
-\bibliography{wssspe.bib}
+\bibliography{wssspe51jors.bib}
 \end{document}

### 2018-12-04 WSSSPE5.1_report/wssspe51jors.tex

@@
 
-Figure~\ref{fig:schematic} shows a version of a schematic of the research 
+The "research software sustainability space" can be described as a set of activities which impact the sustainability of research software in different ways and on different levels, e.g., how a research software is developed by developers, how it is funded, how and where it is published, etc.
+These activities consist of "actions", which are undertaken by "agents", where an "actor" acts on an "actee".
+Activities in the space can thus be modeled as a directed graph, and represented in respective schematics.
+In this section, we aim to evaluate which subset of activities has been represented in the presentations given at WSSSPE5.1.
+
+Figure~\ref{fig:schematic} shows a version of a schematic of activities within the research 
 software sustainability space 
 as introduced by Katz \cite{katz_research_2018}.
-In this version, both node outlines and edge labels are weighted to show the distribution of
-their references from the workshop presentations \cite{druskat_activity}.
+The original schematic has been a result of introspective work based on experience rather than quantitative research.
+Author Druskat has remodeled the original schematic manually, using \cite{drawio} run on \href{https://www.draw.io}{draw.io}, to produce Figure~\ref{fig:schematic}.
+In order to adapt it for purposes of workshop evaluation, both node outlines and edge labels have been weighted to show the distribution of recorded agents and actions across workshop presentations \cite{druskat_activity}.
+This work was done based on the presentation abstracts \cite{WSSSPE5_1_proceedings_2017} rather than the actual presentations to ensure traceability.
+
+To calculate the applicable weights, the schematic was resolved into unique activities.
+A unique activity is a single edge label verb (the action) going from a node (the actor) to another node (the actee).
+The presentations were then coded, recording for each presentation whether or not an activity was present.
+Based on the total of occurrences of an activity across presentations ($x$) and the range of occurrence totals of all activities across presentations ($r$), the font size in point (pt) for the edge labels representing single actions has been normalized to target scaling range $t = 12-24$, yielding the scaled font size $x'$:
+
+\[x' =  \frac{x - r_{min}}{r_{max} - r_{min}} \cdot \left(t_{min} - t_{min}\right) + t_{min}\]
+
+Similarly, the total of occurrences of an agent across presentations, either as actor or actee of an activity, has been normalized to target scaling range $t = 2-20$, yielding the scaled width of the node outline in pt.
 
@@
   \includegraphics[width=\textwidth]{sustainability-actions-plot-resize-jors.pdf}
-  \caption{Distribution of action coverage over presentations.}
+  \caption{Distribution of activity coverage over presentations.}
   \label{fig:actions}
@@
 
-Figure~\ref{fig:actions} shows the distribution of actions over all presentations, where an action is a labeled edge from a node in the sustainability schematic (Figure~\ref{fig:schematic}) to another node.
+Figure~\ref{fig:actions} shows the distribution of actions over all presentations, based on \cite{druskat_activity}, where an action is a labeled edge from a node in the sustainability schematic (Figure~\ref{fig:schematic}) to another node.
 The source node represents the actor, and the target node the actee of the respective action.
@@
 Future workshops could take care to address this topic specifically in their calls for submissions in order to close gaps in research, discussion and progress.
+
+Taken for itself, these figures only represent a snapshot of activities around research software sustainability.
+In order to represent the continuum of efforts and make quantitatively informed statements about both general progress, and the specific insights and outcomes from the particular workshop reported on here, a larger corpus of workshop products (abstracts, presentations, blogs, etc.) would have to be analyzed.
+While such a larger-scaled analysis is out of scope for this report, work has started within the WSSSPE community to create an ontology of activities in the research software sustainability space \cite{druskat2018mapping,druskat_mapping_nodate}, which will enable analyses of this kind in future work.
 
@@
 The rest of this section summarizes the WSSSPE5.1 blog posts, which are accessible at \href{http://wssspe.researchcomputing.org.uk/wssspe5-1/}{wssspe.researchcomputing.org.uk/wssspe5-1/}.
-Indented text indicated quotations from the blogs.
+Indented italicized text indicates quotations from the blogs.
 
@@
 \item The Research Software Project Manager\\(\href{https://www.software.ac.uk/blog/2017-12-04-research-software-project-manager}{www.software.ac.uk/blog/2017-12-04-research-software-project-manager})
-\begin{quote}
+\begin{quote}\textit{
 For many, the role of research software project manager (RSPM) may be an accidental calling.
 The career path for this role isn’t well-established, and research software development in academia may itself be something of a haphazard, nigh-accidental byproduct of conducting domain research.
-Individuals approaching this role may have little to no wider industry experience, instead approaching the project manager role from research or research software engineering.
+Individuals approaching this role may have little to no wider industry experience, instead approaching the project manager role from research or research software engineering.}
 \end{quote}
@@
 {www.software.ac.uk/blog/2017-12-05-looking-software-use-research})
-\begin{quote}
+\begin{quote}\textit{
 Nowadays, software is used in most research.
@@
 This type of analysis could come, for example, from funding bodies and organisations to reward the creation of something and help its sustainability, from institutions who hire people behind that software, or from the software authors to get an understanding of the needs of their users or simply to get credit for their work.
-Another motivation may be trying to understand the research being carried out with a particular software or set of tools either for purely academic purposes (e.g., by historians and scholars of science) or with a commercial perspective (such as by intellectual property teams from universities for the monetisation of the software).
-%lx% should be shortened
+Another motivation may be trying to understand the research being carried out with a particular software or set of tools either for purely academic purposes (e.g., by historians and scholars of science) or with a commercial perspective (such as by intellectual property teams from universities for the monetisation of the software).} %lx% should be shortened
 \end{quote}
@@
 \item Towards Reproducibility in Research Software\\ (\href{https://www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software}{www.software.ac.uk/blog/2017-12-06-towards-reproducibility-research-software})
-\begin{quote}
+\begin{quote}\textit{
 Ensuring reproducibility of research has been identified as one of the challenges in scientific research.
 While reproducibility of results is a concern in all fields of science, the emphasis of this group is in the area of computer software reuse and the reproduction of results.
-The availability of complete descriptions, ideally including program source code, documentation and archives of all necessary components and input datasets would be a major step to resolving research reproducibility concerns.
+The availability of complete descriptions, ideally including program source code, documentation and archives of all necessary components and input datasets would be a major step to resolving research reproducibility concerns.}
 \end{quote}
@@
 In general ``better software [leads to] better research'' (\href{https://software.ac.uk/}{SSI}).
-\begin{quote}
+\begin{quote}\textit{
 Higher quality output is on every university's wish list, as it leads to a potential increase in QR funding (\href{http://www.hefce.ac.uk/rsrch/funding/mainstream/}{www.hefce.ac.uk/rsrch/funding/mainstream/}), while reducing the reputational risk associated with substandard research practices.
-Reproducibility is notoriously difficult to achieve and RSEs are an essential part of enabling this [\ldots] leading to higher citations and greater research impact.
+Reproducibility is notoriously difficult to achieve and RSEs are an essential part of enabling this [\ldots] leading to higher citations and greater research impact.}
 \end{quote}
@@
 Long-term benefits for organizations are the spreading of best practices in software development and an institutional memory provided by the central RSE group who have worked in many projects.
-\begin{quote}
+\begin{quote}\textit{
 An RSE team that is permanently employed can be truly agile.
 Recruiting an expert for a short period of time in an academic institution is virtually impossible.
-As long as we rely on fixed-term contracts for RSEs, a lot of important work will fail to be done, and funds will not be spent as effectively as they could be.
+As long as we rely on fixed-term contracts for RSEs, a lot of important work will fail to be done, and funds will not be spent as effectively as they could be.}
 \end{quote}
@@
 \item A standard format for CITATION files\\(\href{https://www.software.ac.uk/blog/2017-12-12-standard-format-citation-files}{www.software.ac.uk/blog/2017-12-12-standard-format-citation-files})
-\begin{quote}
+\begin{quote}\textit{
 The citation of research software has a number of purposes, most importantly attribution and credit, but also the provision of impact metrics for funding proposals, job interviews, etc.
@@
 Direct software citation also enables reproducibility of research results as the exact version can be retrieved from the citation.
+}
 \end{quote}
@@
 The blog post discusses how the ``SCRUM'' approach may not work, but ``agile'' or a ``maturity model'' approach may better fit the current research practices.
-\begin{quote}
+\begin{quote}\textit{
 As a general concept: start small and then go as far as necessary.
@@
 A maturity model can help researchers identify where they are and where they should be [\ldots].
-Restricting the use of tech jargon to a minimum and offering explanations where necessary can help, too.
+Restricting the use of tech jargon to a minimum and offering explanations where necessary can help, too.}
 \end{quote}
 
-\item Encouraging good software development practice in research teams (\href{https://www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams}{www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams}
+\item Encouraging good software development practice in research teams (\href{https://www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams}{www.software.ac.uk/blog/2017-12-13-encouraging-good-software-development-practice-research-teams})
 
@@
 \item Overcoming Entry Barriers to Motivate Better Practice in Research Software Engineering (\href{https://www.software.ac.uk/blog/2017-12-14-overcoming-entry-barriers-motivate-better-practice-research-software-engineering}{www.software.ac.uk/blog/2017-12-14-overcoming-entry-barriers-motivate-better-practice-research-software-engineering})
-\begin{quote}
+\begin{quote}\textit{
 What can be termed as ``coding'' is a subset of wider software engineering practices such as version control, continuous integration and good software design.
 Coding is prevalent in academia but practices that allow sustainable software to be produced are frequently overlooked.
-Motivating the uptake of the approaches, methods and tools, and highlighting the benefit they deliver, by engaging with researchers who develop software is the first step in spreading best practice in our community.
+Motivating the uptake of the approaches, methods and tools, and highlighting the benefit they deliver, by engaging with researchers who develop software is the first step in spreading best practice in our community.}
 \end{quote}
@@
 %\todo{more needed?} 
-In summary, the community recognizes the need for improving software engineering practices and takes action.
+In summary, the community recognizes the need for improving software engineering practices and has been starting to take action around this.
 A set of initial success stories have been reported from a handful of organizations.
-Recognition for work, including via citations, remains a topic of interest.
+Recognition for work, including via citations, remains a topic of interest, with less progress that is needed.
 %
@@
 The presentations and speed blogs cover most of the topics and do not have subjects that are not in the list, indicating that the topics (or themes, a more fine-grained mapping of the space as discussed in Section~\ref{sec:speed_blog_analysis}) are a good representation of the space.
+
+While the existing set of topics may serve as a broad overview of issues in the research software sustainability space, we cannot yet use them to make broader, informed statements about how the space itself and the activities taking place within it have developed, or whether -- and which -- activities have helped to solve issues, and whether there are gaps in activities that the community should aim to fill.
+
+Overall, as we have stated previously~\cite{WSSSPE4-paper}, we have learned from the first four years of WSSSPE that it is relatively easy to get motivated people to attend a meeting and productively spend their time there both doing work and planning more work, but it is very hard to get that additional work after the meeting to take place.
+Given this, we have turned WSSSPE meetings (including this one and another in 2017, and one in 2018) into gathering places to discuss scientific software sustainability, and for groups that are already in place or that can be composed of related funded activities to meet.
 
